---
title: "Spike-in vs No Spike-in: Significant mRNA Counts (Diverging Bar Charts)"
author: "Kate"
date: "2026-07-20"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

First notebook in the spike-in vs literature comparison subsection. Counts how
many genes are significantly up- and down-regulated (sictrl hypoxia vs
normoxia, |log2FC| >= 0.5 and padj < 0.05) at each timepoint (1/4/24hr), for
RNA, Ribo, and TE separately, comparing spike-in-normalized DESeq2 size
factors against standard (no spike-in) size factors. Each of RNA/Ribo/TE gets
its own bar chart, faceted by timepoint, with Up and Down shown as separate
bars (both plotted as positive counts) grouped by direction and dodged by
spike-in status, styled after the eIF3e/eIF3d TE reference figure. Later
notebooks in this subsection will bring in published literature values for
comparison.

**With-spike-in results are not recomputed here.** `rna_ribo_te_analysis.Rmd`
already ran `deFunction(mode = c("RNA","RIBO","TE"))` with spike-in
normalization for the sictrl hypoxia-vs-normoxia contrast at all three
timepoints and saved `rna_lfc`/`ribo_lfc`/`te_lfc` and matching `*_padj`
columns to `output/translation_categories_{1,4,24}hr.csv` (see lines 92-141,
197-246, 302-351 of that notebook) — those are loaded directly below, for all
three modes.

**TE no-spike-in results are also not recomputed.** `timecourse_qc.Rmd` already
saved these to `output/deseq_results_{1,4,24}hr_nospike.csv`
(control=normoxia, test=hypoxia, treatment=sictrl, mode="TE", spikeInMatrix =
NULL). That file is filtered to `!is.na(symbol)`, but the annotation table
(`accessories/human/gene_anno_hs_dm_v49_r111.tsv`) has zero NA symbols, and
row counts were verified to match `translation_categories_*hr.csv` exactly at
all three timepoints (11,879 genes each) — so the filter is a no-op here and
this file is a safe, complete source.

**RNA and Ribo no-spike-in results have no saved file anywhere in the repo**
(confirmed by exhaustive grep for `nospike`/`no_spike` across all `.Rmd`/`.R`
files and `find` for matching filenames — the only other no-spike-in files
are TE-mode si3d/si3e-vs-sictrl contrasts in `si3e_si3d_normoxia_1and4hr_timecourse.Rmd`
and external MDA-MB-231/MCF7-SIX1 TE files from a different repo in
`eif3d_geneset_cellline_comparison.Rmd`, none of which are the sictrl
hypoxia-vs-normoxia RNA/Ribo contrast needed here). `analyzing_spikeins.Rmd`
hits the same gap and recomputes RNA/Ribo no-spike itself for that reason. We
follow the same approach: only RNA and Ribo no-spike-in are computed fresh,
from the same count matrix and metadata per timepoint.

```{r load_libraries}
library(here)
library(tidyverse)
library(DESeq2)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded deFunction, createNormMatrix\n")
```

# Load With-Spike-In Results (precomputed, RNA/Ribo/TE)

```{r load_with_spike_results}
translation_categories_files <- c(
  "1hr" = here("output", "translation_categories_1hr.csv"),
  "4hr" = here("output", "translation_categories_4hr.csv"),
  "24hr" = here("output", "translation_categories_24hr.csv")
)
stopifnot("translation_categories_*hr.csv missing" = all(file.exists(translation_categories_files)))

modes <- c("RNA", "Ribo", "TE")
mode_cols <- list(
  RNA  = c(lfc = "rna_lfc",  padj = "rna_padj"),
  Ribo = c(lfc = "ribo_lfc", padj = "ribo_padj"),
  TE   = c(lfc = "te_lfc",   padj = "te_padj")
)

with_spike_results <- lapply(translation_categories_files, function(f) {
  raw <- read_csv(f, show_col_types = FALSE)
  setNames(lapply(modes, function(m) {
    cols <- mode_cols[[m]]
    raw %>% dplyr::select(gene_id, log2FoldChange = all_of(cols[["lfc"]]), padj = all_of(cols[["padj"]]))
  }), modes)
})

for (tp in names(with_spike_results)) {
  for (m in modes) {
    cat(tp, m, "with-spike results loaded:", nrow(with_spike_results[[tp]][[m]]), "genes\n")
  }
}
```

# Load No-Spike-In TE Results (precomputed)

```{r load_te_nospike}
te_nospike_files <- c(
  "1hr" = here("output", "deseq_results_1hr_nospike.csv"),
  "4hr" = here("output", "deseq_results_4hr_nospike.csv"),
  "24hr" = here("output", "deseq_results_24hr_nospike.csv")
)
stopifnot("deseq_results_*hr_nospike.csv missing" = all(file.exists(te_nospike_files)))

te_nospike_results <- lapply(te_nospike_files, function(f) {
  read_csv(f, show_col_types = FALSE) %>%
    dplyr::select(gene_id, log2FoldChange, padj)
})

for (tp in names(te_nospike_results)) {
  stopifnot(
    "TE no-spike gene universe does not match with-spike gene universe" =
      nrow(te_nospike_results[[tp]]) == nrow(with_spike_results[[tp]][["TE"]])
  )
  cat(tp, "TE no-spike results loaded:", nrow(te_nospike_results[[tp]]), "genes\n")
}
```

# Compute No-Spike-In RNA/Ribo/TE Results

No saved file covers RNA/Ribo no-spike-in, so all three modes (RNA, Ribo, and
TE) are computed here using `createNormMatrix(..., spikein = NULL)` +
`deFunction(mode = ..., spikeInMatrix = NULL)`, matching the pattern in
`analyzing_spikeins.Rmd`. TE is recomputed as well, purely as a sanity check
against the saved `deseq_results_*hr_nospike.csv` loaded above — the plotted
TE no-spike numbers still come from the saved file, not this recomputation.

```{r load_raw_counts_for_nospike}
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("allCounts:", dim(allCounts), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

```{r run_nospike_de}
timepoints <- c(1, 4, 24)

run_de_nospike <- function(hours_val) {
  metadata_tp <- metadata %>%
    filter(treatment == "sictrl", hours == hours_val) %>%
    mutate(comparison_variable = condition)

  sample_cols <- metadata_tp$sample_name
  counts_tp <- allCounts[, sample_cols]

  normMatrix_noSpike <- createNormMatrix(
    allCounts = counts_tp,
    metadata = metadata_tp,
    control_condition = "normoxia",
    spikein = NULL,
    keepSpikes = NULL
  )

  setNames(lapply(c("RNA", "RIBO", "TE"), function(m) {
    deFunction(
      countsMatrix = counts_tp,
      allCounts_Norm = normMatrix_noSpike,
      metadata = metadata_tp,
      control_condition = "normoxia",
      test_condition = "hypoxia",
      spikeInMatrix = NULL,
      keepSpikes = NULL,
      mode = m
    )
  }), c("RNA", "Ribo", "TE"))
}

no_spike_results_fresh <- setNames(lapply(timepoints, run_de_nospike), paste0(timepoints, "hr"))
cat("Fresh no-spike-in RNA/Ribo/TE computed for:", paste(names(no_spike_results_fresh), collapse = ", "), "\n")
```

## Sanity check: fresh TE no-spike vs saved TE no-spike

```{r sanity_check_te_nospike}
validate_te_nospike <- function(saved_df, fresh_df, label) {
  cat("\n---", label, "---\n")
  cat("Saved genes:", nrow(saved_df), " Fresh genes:", nrow(fresh_df), "\n")

  stopifnot(
    "Gene ID sets differ between saved and freshly-recomputed TE no-spike results" =
      setequal(saved_df$gene_id, fresh_df$gene_id)
  )

  joined <- inner_join(
    saved_df %>% dplyr::select(gene_id, saved_lfc = log2FoldChange, saved_padj = padj),
    fresh_df %>% dplyr::select(gene_id, fresh_lfc = log2FoldChange, fresh_padj = padj),
    by = "gene_id"
  )
  stopifnot("Join on gene_id dropped rows unexpectedly" = nrow(joined) == nrow(saved_df))

  lfc_cor <- cor(joined$saved_lfc, joined$fresh_lfc, use = "complete.obs")
  cat("LFC correlation (saved vs fresh):", round(lfc_cor, 6), "\n")

  stopifnot(
    "LFC values do not match saved TE no-spike results within tolerance" =
      isTRUE(all.equal(joined$saved_lfc, joined$fresh_lfc, tolerance = 1e-8)),
    "padj values do not match saved TE no-spike results within tolerance" =
      isTRUE(all.equal(joined$saved_padj, joined$fresh_padj, tolerance = 1e-8))
  )

  cat("PASSED: fresh TE no-spike recomputation matches saved deseq_results_*hr_nospike.csv\n")
  invisible(TRUE)
}

for (tp in names(te_nospike_results)) {
  validate_te_nospike(te_nospike_results[[tp]], no_spike_results_fresh[[tp]][["TE"]], tp)
}

cat("\n✓ SANITY CHECK PASSED: saved TE no-spike results match a fresh recomputation at all timepoints\n")
```

```{r assemble_de_results}
no_spike_results <- lapply(names(with_spike_results), function(tp) {
  list(
    RNA  = no_spike_results_fresh[[tp]][["RNA"]],
    Ribo = no_spike_results_fresh[[tp]][["Ribo"]],
    TE   = te_nospike_results[[tp]]
  )
})
names(no_spike_results) <- names(with_spike_results)

de_results <- lapply(names(with_spike_results), function(tp) {
  setNames(lapply(modes, function(m) {
    list(with_spike = with_spike_results[[tp]][[m]], no_spike = no_spike_results[[tp]][[m]])
  }), modes)
})
names(de_results) <- names(with_spike_results)
cat("DE ready for timepoints:", paste(names(de_results), collapse = ", "), "\n")
```

## Computational checks

```{r checks}
for (tp in names(de_results)) {
  for (m in modes) {
    for (spike_status in c("with_spike", "no_spike")) {
      res <- de_results[[tp]][[m]][[spike_status]]
      stopifnot(
        "log2FoldChange must be finite" =
          all(is.finite(res$log2FoldChange[!is.na(res$log2FoldChange)])),
        "padj must be within [0, 1]" =
          all(res$padj >= 0 & res$padj <= 1, na.rm = TRUE)
      )
    }
  }
}
cat("All log2FoldChange/padj range checks passed\n")
```

# Count significant genes (|log2FC| >= 0.5, padj < 0.05)

```{r count_sig_genes}
lfc_cutoff <- 0.5
padj_cutoff <- 0.05

count_sig <- function(res) {
  sig <- res %>% filter(!is.na(padj), padj < padj_cutoff, abs(log2FoldChange) >= lfc_cutoff)
  tibble(
    Up = sum(sig$log2FoldChange > 0),
    Down = sum(sig$log2FoldChange < 0)
  )
}

sig_counts <- map_dfr(names(de_results), function(tp) {
  map_dfr(modes, function(m) {
    map_dfr(c(with_spike = "with_spike", no_spike = "no_spike"), function(spike_status) {
      count_sig(de_results[[tp]][[m]][[spike_status]])
    }, .id = "spike_status") %>%
      mutate(Mode = m)
  }) %>%
    mutate(Timepoint = tp)
})

sig_counts_long <- sig_counts %>%
  pivot_longer(cols = c(Up, Down), names_to = "Direction", values_to = "n") %>%
  mutate(
    spike_status = recode(spike_status,
                           with_spike = "With spike-in",
                           no_spike = "No spike-in"),
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Direction = factor(Direction, levels = c("Up", "Down")),
    Mode = factor(Mode, levels = c("RNA", "Ribo", "TE"))
  )

print(sig_counts_long)
```

# Bar charts (one per mode: RNA, Ribo, TE)

Up and Down are both plotted as positive bar heights (no sign flip), grouped
side by side under their own Direction tick and dodged by spike-in status,
faceted by timepoint (one panel per 1/4/24hr) — following the
labeled-bar-chart concept from the eIF3e/eIF3d TE reference figure. Direction
sets the base hue (Up = Okabe-Ito vermillion, Down = Okabe-Ito dark blue);
spike-in status sets the shade (with spike-in = full saturation, no spike-in
= a lighter tint of the same hue), mirroring how the reference figure shades
pre- vs. post-spike-in bars of the same gene. Each bar is additionally
labeled "Pre"/"Post" (spike-in normalization), matching the "Pre spike-in" /
"Post spike-in" terminology already used elsewhere in this repo (e.g.
`analyzing_spikeins.Rmd`, `si3e_si3d_normoxia_1and4hr_timecourse.Rmd`): Pre =
no spike-in, Post = with spike-in.

```{r diverging_bar_plots, fig.width=7, fig.height=4.5}
up_dark <- "#D55E00"
down_dark <- "#0072B2"
lighten <- function(hex, amount = 0.55) colorRampPalette(c(hex, "white"))(101)[round(amount * 100) + 1]
up_light <- lighten(up_dark)
down_light <- lighten(down_dark)

fill_values <- c(
  "Up - With spike-in" = up_dark,
  "Up - No spike-in" = up_light,
  "Down - With spike-in" = down_dark,
  "Down - No spike-in" = down_light
)

make_diverging_bar <- function(df, mode_label) {
  plot_df <- df %>%
    mutate(
      fill_group = factor(paste(Direction, spike_status, sep = " - "), levels = names(fill_values)),
      spike_status = factor(spike_status, levels = c("No spike-in", "With spike-in")),
      prepost_label = if_else(spike_status == "With spike-in", "Post", "Pre")
    )

  dodge <- position_dodge(width = 0.8)
  y_max <- 9000
  label_y <- -0.06 * y_max

  ggplot(plot_df, aes(x = Direction, y = n, fill = fill_group, group = spike_status)) +
    geom_col(width = 0.7, position = dodge) +
    geom_text(aes(label = n), position = dodge, vjust = -0.4, size = 3.5, fontface = "bold") +
    geom_text(aes(label = prepost_label, y = label_y), position = dodge,
              size = 3, fontface = "bold", color = "grey30") +
    geom_hline(yintercept = 0, linewidth = 1) +
    scale_fill_manual(values = fill_values) +
    scale_y_continuous(limits = c(label_y * 1.3, y_max * 1.15)) +
    facet_wrap(~Timepoint, nrow = 1) +
    labs(title = paste0(mode_label, " sig genes: sictrl hypoxia vs normoxia"),
         x = NULL, y = "# sig genes (padj < 0.05, |log2FC| > 0.5)", fill = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "none",
          axis.line = element_line(linewidth = 0.9),
          axis.ticks = element_line(linewidth = 0.9),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 15))
}

diverging_bar_plots <- setNames(lapply(modes, function(m) {
  make_diverging_bar(sig_counts_long %>% filter(Mode == m), m)
}), modes)

diverging_bar_plots[["RNA"]]
ggsave(here("plots/01_divergingbar_sig_rna_spike_vs_nospike.pdf"), plot = diverging_bar_plots[["RNA"]], width = 7, height = 4.5)

diverging_bar_plots[["Ribo"]]
ggsave(here("plots/01_divergingbar_sig_ribo_spike_vs_nospike.pdf"), plot = diverging_bar_plots[["Ribo"]], width = 7, height = 4.5)

diverging_bar_plots[["TE"]]
ggsave(here("plots/01_divergingbar_sig_te_spike_vs_nospike.pdf"), plot = diverging_bar_plots[["TE"]], width = 7, height = 4.5)
```

# Save results

```{r save_results}
write_csv(sig_counts_long, here("output/spikein_literature_comparison/01_sig_gene_counts_spike_vs_nospike.csv"))
cat("Saved significant gene counts (RNA/Ribo/TE) to output/spikein_literature_comparison/\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
