---
title: "Spike-in vs No Spike-in vs Literature: Significant mRNA Counts (Lollipop Charts)"
author: "Kate"
date: "2026-07-20"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

Second notebook in the spike-in vs literature comparison subsection. Part 1
re-renders `01_lollipop_sig_genes_spike_vs_nospike.Rmd`'s significant-gene
counts as horizontal bar charts (flipped orientation from 01's vertical
bars), reusing 01's already-saved output with no recomputation. Part 2
documents candidate published ribosome profiling / translation-efficiency
hypoxia datasets located for comparison — these are **not yet plotted**,
pending Kate's decision on which are actually comparable (see caveats noted
per dataset below).

```{r load_libraries}
library(here)
library(tidyverse)
```

# Part 1: Lollipop version of notebook 01

## Load 01's saved significant-gene counts

No recomputation here — `01_lollipop_sig_genes_spike_vs_nospike.Rmd` already
computed and saved these counts (padj < 0.05, |log2FC| >= 0.5, sictrl hypoxia
vs normoxia, RNA/Ribo/TE, with vs without spike-in, at 1/4/24hr) to
`output/spikein_literature_comparison/01_sig_gene_counts_spike_vs_nospike.csv`.

```{r load_01_results}
sig_counts_file <- here("output/spikein_literature_comparison/01_sig_gene_counts_spike_vs_nospike.csv")
stopifnot("01_sig_gene_counts_spike_vs_nospike.csv missing — knit 01 first" = file.exists(sig_counts_file))

sig_counts_long <- read_csv(sig_counts_file, show_col_types = FALSE) %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Direction = factor(Direction, levels = c("Up", "Down")),
    Mode = factor(Mode, levels = c("RNA", "Ribo", "TE"))
  )

modes <- levels(sig_counts_long$Mode)
print(sig_counts_long)
```

## Bar charts (one per mode: RNA, Ribo, TE)

Same information as before (Up = Okabe-Ito vermillion, Down = Okabe-Ito dark
blue; with spike-in = full saturation, no spike-in = a lighter tint of the
same hue; faceted by timepoint; dodged by spike-in status), but drawn as
horizontal bars instead of lollipops: Direction now sits on the y-axis and
the count runs left to right on the x-axis (flipped from the original
vertical orientation).

```{r bar_plots, fig.width=7, fig.height=4.5}
up_dark <- "#D55E00"
down_dark <- "#0072B2"
lighten <- function(hex, amount = 0.55) colorRampPalette(c(hex, "white"))(101)[round(amount * 100) + 1]
up_light <- lighten(up_dark)
down_light <- lighten(down_dark)

color_values <- c(
  "Up - With spike-in" = up_dark,
  "Up - No spike-in" = up_light,
  "Down - With spike-in" = down_dark,
  "Down - No spike-in" = down_light
)

make_bar <- function(df, mode_label) {
  plot_df <- df %>%
    mutate(
      color_group = factor(paste(Direction, spike_status, sep = " - "), levels = names(color_values)),
      spike_status = factor(spike_status, levels = c("No spike-in", "With spike-in"))
    )

  dodge <- position_dodge(width = 0.7)
  x_max <- 9000

  ggplot(plot_df, aes(y = Direction, x = n, fill = color_group)) +
    geom_col(position = dodge, width = 0.6) +
    geom_text(aes(label = n), position = dodge, hjust = -0.2, size = 3.5, fontface = "bold") +
    geom_vline(xintercept = 0, linewidth = 1) +
    scale_fill_manual(values = color_values) +
    scale_x_continuous(limits = c(0, x_max * 1.15)) +
    facet_wrap(~Timepoint, nrow = 1) +
    labs(title = paste0(mode_label, " sig genes: sictrl hypoxia vs normoxia"),
         x = "# sig genes (padj < 0.05, |log2FC| > 0.5)", y = NULL, fill = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top",
          axis.line.x = element_line(linewidth = 0.9),
          axis.ticks.x = element_line(linewidth = 0.9),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 15))
}

bar_plots <- setNames(lapply(modes, function(m) {
  make_bar(sig_counts_long %>% filter(Mode == m), m)
}), modes)

bar_plots[["RNA"]]
ggsave(here("plots/02_bar_sig_rna_spike_vs_nospike.pdf"), plot = bar_plots[["RNA"]], width = 7, height = 4.5)

bar_plots[["Ribo"]]
ggsave(here("plots/02_bar_sig_ribo_spike_vs_nospike.pdf"), plot = bar_plots[["Ribo"]], width = 7, height = 4.5)

bar_plots[["TE"]]
ggsave(here("plots/02_bar_sig_te_spike_vs_nospike.pdf"), plot = bar_plots[["TE"]], width = 7, height = 4.5)
```

# Part 2: Candidate literature ribosome-profiling hypoxia datasets

**Status: research only, nothing plotted yet.** These are candidates located
for Kate to review; inclusion depends on her judgment about comparability to
our own condition (MDA-MB-231, sictrl, hypoxia vs normoxia, 1/4/24hr — exact
O2% and chamber duration for our own experiment were not found documented
anywhere in this repo and need to be confirmed before any "same condition"
claims are made in a figure). Citations use `Author et al. YYYY` format per
Kate's request, to sit alongside our own "Pre spike-in" / "Post spike-in"
bars in a future combined plot.

| Citation | Journal | Cell line | O2% hypoxia (vs normoxia) | Duration(s) | Method | Reported sig. gene/mRNA counts | Comparability notes |
|---|---|---|---|---|---|---|---|
| Jiang et al. 2017 ([10.1186/s12864-017-3996-8](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-3996-8), PMID 28826393, PMC5563900) — Jiang Z, Yang J, Dai A, Wang Y, Li W, Xie Z. "Ribosome profiling reveals translational regulation of mammalian cells in response to hypoxic stress." Raw data: SRA **SRP094893** | BMC Genomics | ARPE-19 (human retinal pigment epithelium) | 1% (vs 21%) | 1h, 2h, 4h | **Simultaneous RNA-seq + Ribo-seq** (not RNA-seq only). edgeR. DEG/DTG cutoff: \|log2FC\| >= 1 AND mean logRPKM >= 3.3; DTE cutoff: FC > 2 AND mean RPKM > 5 | DEGs (mRNA): 92 (1h), 192 (2h), 212 (4h); DTGs (RPF): 128, 183, 293; DTE mRNAs: 419, 239, 158 — **direction (up vs down) is not split out at any timepoint** | Best O2%/timepoint match to ours (1% O2, 1h/4h) of any candidate. Non-cancer cell line. **BLOCKED for this figure: their significance calls use fold-change plus an RPKM abundance floor with NO adjusted p-value anywhere in the criterion** — so Additional file 2's per-gene table cannot be re-filtered at our padj < 0.05, and the RPKM floor cannot be undone. Harder block than Shen et al. (which at least reports a raw p). Would require reprocessing SRP094893 from raw reads through our own DESeq2. |
| Ho et al. 2016 | Cell Reports (10.1016/j.celrep.2016.01.036) | U87MG (glioblastoma), 786-O (renal clear cell carcinoma) | 1% (vs 21%) | 24h (+ 6h pretreatment/24h labeling variant) | Ribosome-footprint profiling + quantitative MS | No up/down significant-gene counts reported in the padj/FC-cutoff sense used here — genes instead binned into Class I/II/III translation-efficiency-ratio categories (~25%/~60%/~15%) | Duration matches our 24h timepoint well. Not a compatible summary statistic (no discrete "N up / N down" figure) — would need raw data reprocessing to get a comparable count, not just a citation. |
| Ho et al. 2020 | Nature Communications (10.1038/s41467-020-16504-1) | U87MG (glioblastoma) | 1% (0.1% used only in a *C. elegans* validation arm) | 24h primary, 72h confirmatory | Polysome profiling + TMT-pSILAC + RNA-seq (not ribo-seq) | HuR-knockdown-dependent: ~3,435 mRNAs affected at TE level vs. 1,323 at RNA level (2-fold threshold) — this is an RBP-knockdown effect, not a plain hypoxia-vs-normoxia comparison | Different assay (polysome profiling, not ribo-seq) and different comparison (HuR KD, not hypoxia vs normoxia) — weakest match of the four; flagging mainly because it's from the same group/cell line as Ho et al. 2016. |
| Andreev et al. 2015 | Genome Biology (10.1186/s13059-015-0651-z) | PC12 (rat pheochromocytoma, neural) | 0% O2 (anoxia) **+ glucose deprivation**, vs 21% O2 | 20min, 40min, 1h, 2h, 4h | Ribo-seq, Z-score + 10% FDR | At 40 min: 3,202 DE genes by ribo-seq occupancy; 2,556 DE genes by TE — **up/down not separated in the summary**; other timepoints not fully extracted here | Oxygen-glucose deprivation (OGD), not pure hypoxia — confounded by glucose withdrawal. Anoxia (0% O2) is far more severe than typical 1% hypoxia protocols. Different species (rat) and cell type (neural, not epithelial/cancer). Weak match on condition despite good ribo-seq method match. |

**Not pursued further / lower priority:**
- Erhard et al. 2025 (Nat Commun, HIF1α nucleolar ribosome biogenesis, breast cancer, 1% vs 20% O2) — breast-cancer relevant and O2% plausibly comparable, but the paper's focus is rRNA transcription/ribosome biogenesis, not a ribo-seq TE up/down gene-count comparison; would need a closer read of methods/supplement to see if a comparable count exists.
- van den Beucken et al. (RNA-binding-protein/anaerobic-metabolism hypoxia translation work) — repeatedly surfaced in searches but could not confirm a ribo-seq-based up/down significant-gene count in the time spent; flagging the author/topic in case Kate recognizes the specific paper she has in mind.

**Resolved:** our own hypoxia condition is 1% O2, chamber duration = the
sample timepoint itself (1/4/24hr). Confirmed by Kate; not documented
anywhere in this repo prior to this notebook. Under this condition, Jiang et
al. 2017 (1% O2, 1h/2h/4h) is the closest O2%/duration match among the
candidates above.

**Resolved (2026-08-03), Jiang et al. 2017 — checked against the paper, not
usable here.** This row previously carried author-year only, with no DOI, and
had been questioned as possibly being an RNA-seq-only study. Both points are
now settled against the published article: it *is* simultaneous RNA-seq +
ribosome profiling (see the title), first author Zhiwen Jiang, and the DTE
counts recorded above (419/239/158) match the paper exactly. **However, it
cannot contribute a bar to any plot in this notebook**: Jiang et al. call
significance with edgeR using fold-change plus a mean-RPKM abundance floor and
no FDR/adjusted p-value at all, so their supplementary per-gene table cannot be
re-filtered under our padj < 0.05 the way Hettiarachchi et al.'s was. Getting a
comparable count would mean reprocessing SRA SRP094893 from raw reads through
our own DESeq2 — out of scope for this figure. Do not re-open this without new
information; the blocker is the absent FDR, not a missing file.

**Net result: Hettiarachchi et al. 2019 (TE) and Shen et al. 2020 (RNA, Ribo)
are the only published datasets plotted in this notebook, and the only ones of
those surveyed that can be.** Note that neither used spike-in normalization,
so every literature bar in Parts 4-6 is effectively a no-spike-in measurement
and can only speak to whether our no-spike-in numbers are plausible — it cannot
adjudicate our with-spike-in counts. Neither is a cancer cell line (human
primary hepatocytes; rat H9C2 cardiomyoblasts), so there is no cancer-line
comparator in this figure at all.

# Part 3: Hettiarachchi et al. 2019 comparison (human primary hepatocytes)

Hettiarachchi GK, et al. "Translational and transcriptional responses in
human primary hepatocytes under hypoxia." *Am J Physiol Gastrointest Liver
Physiol* 2019. DOI: 10.1152/ajpgi.00331.2018 (PMC6620582).

- **Cell type:** pooled human primary hepatocytes from 5 donors (Thermo
  Fisher Scientific), mixed age/sex/phenotype.
- **O2%:** 1% hypoxia vs. ambient (~21%) normoxia — **matches our own
  condition exactly.**
- **Timepoints:** 30 min and 240 min of hypoxia exposure (chamber: Plas-Labs
  glove box, 37°C). **240 min = 4hr matches our 4hr timepoint exactly.** 30
  min does not correspond exactly to any of our timepoints (our earliest is
  1hr = 60min) — noted, not silently aligned.
- **Method:** RNA-seq + Ribo-seq + TE (ratio of ribosome-protected fragments
  to mRNA fragments), matching our own assay type.
- **Their own significance cutoffs:** |log2FC| > 1 and padj (BH-FDR) < 0.10
  — looser padj, stricter LFC than ours (padj < 0.05, |log2FC| >= 0.5).
- **Their reported totals (Fig. in paper, not split by up/down):** 760
  differential-TE genes at 30 min; 841 at 240 min.
- Supplied by Kate: `accessories/hettiarachchi riboseq/supplemental file
  s1.xls`, with per-comparison tabs including `"0 vs 30 min TE"` and
  `"0 vs 240 min TE"`.

```{r load_libraries_hettiarachchi}
library(readxl)
```

```{r load_hettiarachchi_te}
hetti_file <- here("accessories", "hettiarachchi riboseq", "supplemental file s1.xls")
stopifnot("Hettiarachchi supplemental file not found" = file.exists(hetti_file))

hetti_sheets <- c("30min" = "0 vs 30 min TE", "240min" = "0 vs 240 min TE")

hetti_te <- lapply(hetti_sheets, function(sheet) {
  read_excel(hetti_file, sheet = sheet) %>%
    dplyr::rename(symbol = 1)
})

for (tp in names(hetti_te)) {
  cat(tp, "TE sheet rows:", nrow(hetti_te[[tp]]), "\n")
}
```

## Important caveat: the supplied TE tabs are already pre-filtered to the authors' own significant gene set

```{r check_hetti_prefiltered}
for (tp in names(hetti_te)) {
  d <- hetti_te[[tp]]
  stopifnot(
    "Expected every row to already pass the paper's own |log2FC| > 1 cutoff" =
      all(abs(d$log2FoldChange) > 1, na.rm = TRUE),
    "Expected every row to already pass the paper's own padj < 0.10 cutoff" =
      all(d$padj < 0.10, na.rm = TRUE)
  )
}
cat("Confirmed: both TE tabs contain only the", nrow(hetti_te[["30min"]]),
    "/", nrow(hetti_te[["240min"]]), "genes the authors themselves called\n")
cat("significant (|log2FC| > 1, padj < 0.10) — NOT the full genome-wide DESeq2 output.\n")
```

Row counts (760, 841) match the paper's reported totals exactly, confirming
these tabs are the authors' final significant-gene lists, not unfiltered
results. **Consequence:** applying our own thresholds below only re-filters
*within* this already-significant set. Any gene with 0.5 <= |log2FC| < 1
that would have passed our padj < 0.05 cutoff on the full dataset is absent
from this file entirely — so the counts below are a **lower bound**, not the
true count we'd get from the full unfiltered Hettiarachchi et al. DESeq2
results (which were not supplied).

## Reproducibility check: applying their exact cutoffs to their own data reproduces their reported totals

```{r hetti_reproduce_their_totals}
paper_reported_total <- c("30min" = 760, "240min" = 841)

for (tp in names(hetti_te)) {
  d <- hetti_te[[tp]]
  n_matching_their_cutoffs <- sum(abs(d$log2FoldChange) > 1 & d$padj < 0.10, na.rm = TRUE)
  cat(tp, ": applying |log2FC| > 1 & padj < 0.10 to their data gives",
      n_matching_their_cutoffs, "genes (they reported", paper_reported_total[[tp]], ")\n")
  stopifnot(
    "Reapplying the paper's own cutoffs to their own data did not reproduce their reported total" =
      n_matching_their_cutoffs == paper_reported_total[[tp]]
  )
}
cat("Confirmed: their cutoffs reproduce their reported totals exactly at both timepoints.\n")
```

## Apply our thresholds (padj < 0.05, |log2FC| >= 0.5) and compare to their reported totals

```{r hetti_count_sig}
count_sig_hetti <- function(d) {
  sig <- d %>% filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) >= 0.5)
  tibble(Up = sum(sig$log2FoldChange > 0), Down = sum(sig$log2FoldChange < 0))
}

hetti_counts <- map_dfr(names(hetti_te), function(tp) {
  count_sig_hetti(hetti_te[[tp]]) %>% mutate(Timepoint = tp)
}) %>%
  mutate(Our_threshold_total = Up + Down)

hetti_comparison <- hetti_counts %>%
  mutate(
    Paper_reported_total = c(`30min` = 760, `240min` = 841)[Timepoint],
    Paper_cutoffs = "|log2FC| > 1, padj < 0.10 (not split by direction)"
  ) %>%
  dplyr::select(Timepoint, Up, Down, Our_threshold_total, Paper_reported_total, Paper_cutoffs)

print(hetti_comparison)

write_csv(hetti_comparison, here("output/spikein_literature_comparison/02_hettiarachchi_te_comparison.csv"))
cat("Saved Hettiarachchi et al. 2019 TE comparison to output/spikein_literature_comparison/\n")
```

Our-threshold total will be <= the paper's reported total at each timepoint
(subset of an already-filtered set, per the caveat above) — this is expected
and not evidence of a discrepancy in either analysis.

# Part 4: TE comparison (1/4/24hr) — with spike-in vs no spike-in vs Hettiarachchi et al. 2019

Reorganized to match the layout used for the final combined plot in
`04_bar_te_si3d_si3e_with_nospike_vs_literature.Rmd`: source labels on the
x-axis, bars going up (count on the y-axis), axis text rotated 90° and
wrapped, dotted/dashed vertical separators grouping the categories instead
of interleaving them.

Left to right: **with spike-in**, all three timepoints (1hr, 4hr, 24hr) —
one set of Up/Down bars per timepoint — then a **dotted** vertical line,
then **no spike-in**, all three timepoints — another set of Up/Down bars per
timepoint (6 "ours" sets total), then a **dashed** vertical line, then
Hettiarachchi et al. 2019's own Up/Down counts, in order of the hypoxia
timepoint used (30min, then 240min). Unlike the previous version of this
plot, our three timepoints are no longer individually paired against a
matched Hettiarachchi timepoint — all "ours" bars are grouped together, all
literature bars are grouped together, same as notebook 04.

All "ours" counts use **our** thresholds (padj < 0.05, |log2FC| >= 0.5), not
the original publication's cutoffs — the Hettiarachchi bars remain a lower
bound (see Part 3 caveat) since their supplied data was pre-filtered under
their own looser/stricter mixed cutoffs before we ever applied ours. The
Hettiarachchi counts reused here (`hetti_counts`, from Part 3) were already
reproducibility-checked against the paper's own reported totals in the
`hetti_reproduce_their_totals` chunk above (760 at 30min, 841 at 240min,
both `stopifnot`-verified) — this section adds no new, unchecked
Hettiarachchi numbers. Our 24hr timepoint has no Hettiarachchi match (the
paper only reports 30min/240min) — it still gets its own with-spike/no-spike
bars in the "ours" block, just no literature counterpart.

Bar fill follows notebook 04's convention: "ours" bars (all six) share one
Okabe-Ito color, literature bars are grey; Up/Down is conveyed by the bold
vermillion/blue count label above each bar, not by fill.

```{r combine_te_and_hetti}
ours_color <- "#009E73"
literature_color <- "grey50"

our_te_all <- sig_counts_long %>%
  filter(Mode == "TE") %>%
  mutate(spike_status = factor(spike_status, levels = c("With spike-in", "No spike-in"))) %>%
  transmute(
    source = paste0("**", spike_status, " (", Timepoint, ")**"),
    Direction, n, spike_status, Timepoint
  )

hetti_30min_long <- hetti_counts %>%
  filter(Timepoint == "30min") %>%
  dplyr::select(Up, Down) %>%
  pivot_longer(everything(), names_to = "Direction", values_to = "n") %>%
  mutate(source = "Hettiarachchi et al. 2019<br>(30min, 1% O2,<br>human primary hepatocytes)")

hetti_240min_long <- hetti_counts %>%
  filter(Timepoint == "240min") %>%
  dplyr::select(Up, Down) %>%
  pivot_longer(everything(), names_to = "Direction", values_to = "n") %>%
  mutate(source = "Hettiarachchi et al. 2019<br>(240min/4hr, 1% O2,<br>human primary hepatocytes)")

source_levels <- c(
  "**With spike-in (1hr)**", "**With spike-in (4hr)**", "**With spike-in (24hr)**",
  "**No spike-in (1hr)**", "**No spike-in (4hr)**", "**No spike-in (24hr)**",
  "Hettiarachchi et al. 2019<br>(30min, 1% O2,<br>human primary hepatocytes)",
  "Hettiarachchi et al. 2019<br>(240min/4hr, 1% O2,<br>human primary hepatocytes)"
)

n_ours <- 6
n_with_spike <- 3

source_colors <- setNames(
  c(rep(ours_color, n_ours), rep(literature_color, length(source_levels) - n_ours)),
  source_levels
)

te_vs_hetti <- bind_rows(
  our_te_all %>% dplyr::select(source, Direction, n),
  hetti_30min_long %>% dplyr::select(source, Direction, n),
  hetti_240min_long %>% dplyr::select(source, Direction, n)
) %>%
  mutate(
    Direction = factor(Direction, levels = c("Up", "Down")),
    source = factor(source, levels = source_levels)
  )

print(te_vs_hetti)
```

```{r plot_te_vs_hetti, fig.width=12, fig.height=10}
library(ggtext)

y_pad <- 0.20 * max(te_vs_hetti$n)
dodge <- position_dodge(width = 0.7)

p_te_vs_hetti <- ggplot(te_vs_hetti, aes(x = source, y = n, fill = source, group = Direction)) +
  geom_col(position = dodge, width = 0.6) +
  geom_text(aes(label = n, color = Direction), position = dodge,
            vjust = -0.4, size = 8, fontface = "bold", key_glyph = "point") +
  geom_hline(yintercept = 0, linewidth = 1) +
  geom_vline(xintercept = n_with_spike + 0.5, linetype = "dotted", linewidth = 0.7, color = "grey30") +
  geom_vline(xintercept = n_ours + 0.5, linetype = "dashed", linewidth = 0.7, color = "grey30") +
  scale_fill_manual(values = source_colors, guide = "none") +
  scale_color_manual(values = c(Up = up_dark, Down = down_dark), name = NULL) +
  scale_y_continuous(limits = c(0, max(te_vs_hetti$n) + y_pad)) +
  scale_x_discrete(limits = source_levels) +
  guides(color = guide_legend(override.aes = list(size = 7))) +
  labs(title = "TE changes: ours (1/4/24hr, with/without spike-in)\nvs Hettiarachchi et al. 2019",
       subtitle = paste0(
         "All counts use OUR cutoffs (padj < 0.05, |log2FC| >= 0.5), not the original publication's cutoffs.\n",
         "Hettiarachchi bars are a LOWER BOUND: their supplied data was pre-filtered to |log2FC| > 1, so genes at\n",
         "0.5 <= |log2FC| < 1 that would pass our cutoffs are absent from the file entirely (see Part 3).\n",
         "Ours = Okabe-Ito; literature = grey."
       ),
       x = NULL, y = "# sig genes") +
  theme_classic(base_size = 24) +
  theme(legend.position = "top",
        legend.text = element_text(size = 21),
        plot.title = element_text(size = 25),
        plot.subtitle = element_text(color = "grey40", size = 15),
        axis.line.y = element_line(linewidth = 0.9),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_text(size = 22),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_markdown(size = 18, angle = 90, hjust = 1, vjust = 0.5))

p_te_vs_hetti
ggsave(here("plots/02_bar_te_all_timepoints_spike_vs_nospike_vs_hettiarachchi.pdf"), plot = p_te_vs_hetti, width = 12, height = 10)
```

# Part 5: RNA & Ribo comparison vs Shen et al. 2020 (hypoxic rat cardiomyocytes)

Shen JJ, et al. "Translatome and Transcriptome Profiling of Hypoxic-Induced
Rat Cardiomyocytes." *Molecular Therapy Nucleic Acids* 2020. DOI:
10.1016/j.omtn.2020.09.010 (PMC7689039).

Unlike Part 4 (TE), this comparison uses **RNA and Ribo separately** (ours,
4hr only — the closest of our timepoints to Shen et al.'s 8h — with and
without spike-in) against Shen et al.'s **RNA-seq (transcriptome)** and
**polyribo-seq (translatome)** DEG counts — matching metric to metric
instead of comparing TE to a non-TE metric. Only the **8h vs normoxia**
comparison is used for both RNA-seq and Ribo-seq (a true hypoxia-vs-normoxia
comparison, like ours); the paper's 24h-vs-8h comparisons (severe vs. mild
hypoxia, NOT vs. normoxia — see prior discussion) are excluded from this
plot entirely, for both modalities, since that isn't the comparison type
this plot is built around.

- **Cell type:** rat cardiac myoblast **H9C2** cell line (in vitro; the
  paper's in vivo rat left-ventricle arm is not used here).
- **O2%:** 1% O2 (94% N2, 5% CO2) vs standard culture normoxia.
- **Timepoint:** 8h hypoxia. Not an exact match to our 4hr timepoint
  (longer).
- **No spike-in normalization used.** Confirmed directly: Shen et al.
  quantify with StringTie FPKM values and call differential expression with
  Ballgown — no ERCC or exogenous RNA spike-ins anywhere in their pipeline.
  So every Shen et al. bar in this plot is effectively a "no spike-in"
  condition, unlike our own bars (which show both).
- **Their cutoffs:** |log2FC| >= 1 and **raw p < 0.05** (Ballgown, not
  BH-adjusted padj) — looser statistical correction than ours (padj < 0.05).
- **No supplementary per-gene file used for a re-filter under our own
  cutoffs.** Ballgown's output does not report an `lfcSE` column, so even
  with the raw supplementary tables (Table S3 = RNA-seq, Table S4 =
  polyribo-seq DEGs) there is no valid way to propagate error into a padj
  for a derived quantity — the counts below are the paper's own reported
  totals, taken as-is, not recomputed under our thresholds.

```{r combine_rna_ribo_and_shen}
shen_groups <- tribble(
  ~source, ~Direction, ~n,
  "Shen et al. 2020<br>(H9C2, RNA-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)", "Up", 1527,
  "Shen et al. 2020<br>(H9C2, RNA-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)", "Down", 1398,
  "Shen et al. 2020<br>(H9C2, Ribo-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)", "Up", 536,
  "Shen et al. 2020<br>(H9C2, Ribo-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)", "Down", 429
)

our_rna_ribo <- sig_counts_long %>%
  filter(Mode %in% c("RNA", "Ribo"), Timepoint == "4hr") %>%
  transmute(
    modality = as.character(Mode),
    source = paste0("**", Mode, ": ", spike_status, " (", Timepoint, ")**"),
    Direction, n
  )

rna_source_levels <- c(
  "**RNA: With spike-in (4hr)**", "**RNA: No spike-in (4hr)**",
  "Shen et al. 2020<br>(H9C2, RNA-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)"
)
ribo_source_levels <- c(
  "**Ribo: With spike-in (4hr)**", "**Ribo: No spike-in (4hr)**",
  "Shen et al. 2020<br>(H9C2, Ribo-seq, 8hr, 1% O2, |log2FC| >= 1, p < 0.05)"
)

rna_vs_shen <- bind_rows(
  our_rna_ribo %>% filter(modality == "RNA") %>% dplyr::select(source, Direction, n),
  shen_groups %>% filter(grepl("RNA-seq", source))
) %>%
  mutate(Direction = factor(Direction, levels = c("Up", "Down")),
         source = factor(source, levels = rna_source_levels))

ribo_vs_shen <- bind_rows(
  our_rna_ribo %>% filter(modality == "Ribo") %>% dplyr::select(source, Direction, n),
  shen_groups %>% filter(grepl("Ribo-seq", source))
) %>%
  mutate(Direction = factor(Direction, levels = c("Up", "Down")),
         source = factor(source, levels = ribo_source_levels))

print(rna_vs_shen)
print(ribo_vs_shen)
```

```{r plot_rna_ribo_vs_shen, fig.width=7.5, fig.height=4.5}
make_rna_ribo_bar <- function(df, source_levels_local, modality_label) {
  x_pad_local <- 0.15 * max(df$n)
  ggplot(df, aes(y = source, x = n, fill = Direction)) +
    geom_col(position = dodge, width = 0.6) +
    geom_text(aes(label = n), position = dodge, hjust = -0.25, size = 3.5, fontface = "bold") +
    geom_vline(xintercept = 0, linewidth = 1) +
    scale_fill_manual(values = c(Up = up_dark, Down = down_dark)) +
    scale_x_continuous(limits = c(0, max(df$n) + x_pad_local)) +
    scale_y_discrete(limits = rev(source_levels_local)) +
    labs(title = paste0(modality_label, " changes: ours (with/without spike-in, 4hr only) vs Shen et al. 2020"),
         subtitle = "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Shen et al. counts are their own reported 8hr-vs-normoxia totals (|log2FC| >= 1, raw p < 0.05, not BH-adjusted), not recomputed under our cutoffs.",
         x = "# sig genes", y = NULL, fill = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top",
          plot.subtitle = element_text(color = "grey40", size = 8.5),
          axis.line.x = element_line(linewidth = 0.9),
          axis.ticks.x = element_line(linewidth = 0.9),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.y = element_markdown(size = 9.5))
}

p_rna_vs_shen <- make_rna_ribo_bar(rna_vs_shen, rna_source_levels, "RNA")
p_rna_vs_shen
ggsave(here("plots/02_bar_rna_vs_shen.pdf"), plot = p_rna_vs_shen, width = 7.5, height = 4.5)

p_ribo_vs_shen <- make_rna_ribo_bar(ribo_vs_shen, ribo_source_levels, "Ribo")
p_ribo_vs_shen
ggsave(here("plots/02_bar_ribo_vs_shen.pdf"), plot = p_ribo_vs_shen, width = 7.5, height = 4.5)
```

# Part 6: RNA & Ribo versions of the Part 4 (TE) plot, one modality per plot

Same 04-style layout as Part 4's TE-vs-Hettiarachchi plot (source labels on
the x-axis, bars up, dotted line splitting with-spike/no-spike "ours"
blocks, dashed line before literature, ours = Okabe-Ito, literature = grey)
— but for RNA and Ribo instead of TE, each as its own plot rather than one
combined chart, since Shen et al. 2020's RNA-seq and Ribo-seq numbers are
different modalities and don't belong on the same "# sig genes" axis as
each other (or as our TE bars). Each plot pairs our own RNA (or Ribo)
with/without spike-in counts at all three timepoints (1/4/24hr) against
Shen et al. 2020's matching modality (their only reported hypoxia-vs-normoxia
timepoint, 8h) — see Part 5 above for full sourcing/caveats on the Shen et
al. numbers reused here.

```{r plot_04style_rna_ribo, fig.width=12, fig.height=10}
make_ours_vs_lit_plot <- function(mode, literature_df, literature_levels, mode_label,
                                  lit_label_short, subtitle_text) {
  ours_levels <- c(
    paste0("**With spike-in (", levels(sig_counts_long$Timepoint), ")**"),
    paste0("**No spike-in (", levels(sig_counts_long$Timepoint), ")**")
  )
  n_ours <- length(ours_levels)
  n_with_spike <- length(levels(sig_counts_long$Timepoint))

  our_data <- sig_counts_long %>%
    filter(Mode == mode) %>%
    mutate(spike_status = factor(spike_status, levels = c("With spike-in", "No spike-in"))) %>%
    transmute(source = paste0("**", spike_status, " (", Timepoint, ")**"), Direction, n)

  source_levels_local <- c(ours_levels, literature_levels)

  plot_df <- bind_rows(our_data, literature_df) %>%
    mutate(Direction = factor(Direction, levels = c("Up", "Down")),
           source = factor(source, levels = source_levels_local))

  source_colors_local <- setNames(
    c(rep(ours_color, n_ours), rep(literature_color, length(literature_levels))),
    source_levels_local
  )

  y_pad_local <- 0.20 * max(plot_df$n)
  dodge_local <- position_dodge(width = 0.7)

  ggplot(plot_df, aes(x = source, y = n, fill = source, group = Direction)) +
    geom_col(position = dodge_local, width = 0.6) +
    geom_text(aes(label = n, color = Direction), position = dodge_local,
              vjust = -0.4, size = 8, fontface = "bold", key_glyph = "point") +
    geom_hline(yintercept = 0, linewidth = 1) +
    geom_vline(xintercept = n_with_spike + 0.5, linetype = "dotted", linewidth = 0.7, color = "grey30") +
    geom_vline(xintercept = n_ours + 0.5, linetype = "dashed", linewidth = 0.7, color = "grey30") +
    scale_fill_manual(values = source_colors_local, guide = "none") +
    scale_color_manual(values = c(Up = up_dark, Down = down_dark), name = NULL) +
    scale_y_continuous(limits = c(0, max(plot_df$n) + y_pad_local)) +
    scale_x_discrete(limits = source_levels_local) +
    guides(color = guide_legend(override.aes = list(size = 7))) +
    labs(title = paste0(mode_label, " changes: ours (1/4/24hr, with/without spike-in)\nvs ", lit_label_short),
         subtitle = subtitle_text,
         x = NULL, y = "# sig genes") +
    theme_classic(base_size = 24) +
    theme(legend.position = "top",
          legend.text = element_text(size = 21),
          plot.title = element_text(size = 25),
          plot.subtitle = element_text(color = "grey40", size = 15),
          axis.line.y = element_line(linewidth = 0.9),
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.title.y = element_text(size = 22),
          axis.line.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_markdown(size = 18, angle = 90, hjust = 1, vjust = 0.5))
}

shen_rna_literature <- tribble(
  ~source, ~Direction, ~n,
  "Shen et al. 2020<br>(H9C2, RNA-seq, 8hr,<br>1% O2, |log2FC| >= 1, p < 0.05)", "Up", 1527,
  "Shen et al. 2020<br>(H9C2, RNA-seq, 8hr,<br>1% O2, |log2FC| >= 1, p < 0.05)", "Down", 1398
)
shen_rna_levels <- unique(shen_rna_literature$source)

# Shen et al.'s bars are their OWN reported totals under their OWN cutoffs, not
# re-filtered under ours (Ballgown reports no lfcSE, so no padj can be derived
# for a re-count) - see Part 5. Their threshold is stricter on LFC (>= 1 vs our
# >= 0.5) but looser on multiple testing (raw p vs our BH padj), so the net
# direction of the bias is not knowable; say so rather than implying a
# common-cutoff comparison.
shen_subtitle <- paste0(
  "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Shen et al. bars are their own reported 8hr-vs-normoxia totals\n",
  "(|log2FC| >= 1, raw p < 0.05, NOT BH-adjusted), taken as-is and not recomputed under our cutoffs.\n",
  "Ours = Okabe-Ito; literature = grey."
)

shen_ribo_literature <- tribble(
  ~source, ~Direction, ~n,
  "Shen et al. 2020<br>(H9C2, Ribo-seq, 8hr,<br>1% O2, |log2FC| >= 1, p < 0.05)", "Up", 536,
  "Shen et al. 2020<br>(H9C2, Ribo-seq, 8hr,<br>1% O2, |log2FC| >= 1, p < 0.05)", "Down", 429
)
shen_ribo_levels <- unique(shen_ribo_literature$source)

p_rna_04style <- make_ours_vs_lit_plot(
  mode = "RNA", literature_df = shen_rna_literature, literature_levels = shen_rna_levels,
  mode_label = "RNA", lit_label_short = "Shen et al. 2020",
  subtitle_text = shen_subtitle
)
p_rna_04style
ggsave(here("plots/02_bar_rna_all_timepoints_spike_vs_nospike_vs_shen.pdf"), plot = p_rna_04style, width = 12, height = 10)

p_ribo_04style <- make_ours_vs_lit_plot(
  mode = "Ribo", literature_df = shen_ribo_literature, literature_levels = shen_ribo_levels,
  mode_label = "Ribo", lit_label_short = "Shen et al. 2020",
  subtitle_text = shen_subtitle
)
p_ribo_04style
ggsave(here("plots/02_bar_ribo_all_timepoints_spike_vs_nospike_vs_shen.pdf"), plot = p_ribo_04style, width = 12, height = 10)
```

# Session Info

```{r session_info}
sessionInfo()
```
