---
title: "eIF3d/eIF3e Knockdown TE Changes vs Literature"
author: "Kate"
date: "2026-07-21"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

Third notebook in the spike-in vs literature comparison subsection. Very
similar analysis to `01_lollipop_sig_genes_spike_vs_nospike.Rmd` /
`02_lollipop_sig_genes_spike_vs_nospike_and_literature.Rmd`, but the
comparison is **si3d vs sictrl** and **si3e vs sictrl** (eIF3d/eIF3e
knockdown effect on TE), in **normoxia** — not the sictrl hypoxia-vs-normoxia
comparison those notebooks used.

**All our comparisons in this notebook use the 1+4hr time-regressed data**,
not the three separate 1/4/24hr timepoints. `deFunction_time_normalized()`
(hours as a covariate, `~ hours + Condition + SeqType + Condition:SeqType`)
combines the 1hr and 4hr samples into one regression per treatment, giving a
single, more highly powered si3d-vs-sictrl and si3e-vs-sictrl TE estimate —
this is also the timepoint choice suggested on the original eIF3e/eIF3d TE
reference figure this whole subsection was built from ("perhaps change to
using 1_4hr time regression"). Since we're now using one combined estimate
per treatment, **the three separate normoxia timepoints are not shown** in
this notebook.

**No DESeq2 is run in this notebook.** Per Kate: these si3d-vs-sictrl /
si3e-vs-sictrl TE comparisons are already computed and saved (spike-in
normalized, time-regressed) — reuse them as-is, don't recompute.

- **si3d vs sictrl, normoxia, 1+4hr time-regressed:**
  `output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv`
- **si3e vs sictrl, normoxia, 1+4hr time-regressed:**
  `output/te_si3e_vs_sictrl_normoxia_1and4hr.csv` (TE-only columns; note the
  different filename pattern from si3d's — both produced by the same
  `eif3e_eif3d_normoxia_and_hypoxia.Rmd` "1+4hr (time-regressed)" sections,
  confirmed via `deFunction_time_normalized(..., mode = "TE")` with spike-in).

**Sign convention** (per `CLAUDE.md`): despite the filenames, the underlying
DESeq2 contrast runs **sictrl as `test_condition`, si3d/si3e as
`control_condition`** — the extracted coefficient is log2(sictrl / si3d) or
log2(sictrl / si3e). So **positive `te_lfc` = sictrl TE > knockdown TE =
knockdown reduces TE = eIF3d/eIF3e promotes that gene's translation**;
negative = eIF3d/eIF3e inhibits. "Up" in this notebook therefore means
"eIF3d/eIF3e-promoted", not literally "translation went up in the knockdown."

```{r load_libraries}
library(here)
library(tidyverse)
library(ggtext)
```

# Load si3d/si3e vs sictrl TE (normoxia, 1+4hr time-regressed, spike-in normalized, precomputed)

```{r load_si3d_si3e_te}
te_files <- tribble(
  ~treatment, ~file,
  "si3d", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv",
  "si3e", "te_si3e_vs_sictrl_normoxia_1and4hr.csv"
)

stopifnot("One or more si3d/si3e 1+4hr TE files missing" =
  all(file.exists(here("output", te_files$file))))

te_data <- pmap(te_files, function(treatment, file) {
  read_csv(here("output", file), show_col_types = FALSE) %>%
    dplyr::select(gene_id, log2FoldChange = te_lfc, padj = te_padj) %>%
    mutate(treatment = treatment)
}) %>% list_rbind()

for (i in seq_len(nrow(te_files))) {
  n <- sum(te_data$treatment == te_files$treatment[i])
  cat(te_files$treatment[i], "1+4hr TE rows:", n, "\n")
}
```

## Computational checks

```{r checks}
stopifnot(
  "log2FoldChange must be finite" =
    all(is.finite(te_data$log2FoldChange[!is.na(te_data$log2FoldChange)])),
  "padj must be within [0, 1]" =
    all(te_data$padj >= 0 & te_data$padj <= 1, na.rm = TRUE)
)
cat("log2FoldChange/padj range checks passed\n")
```

# Count significant genes (|log2FC| >= 0.5, padj < 0.05)

```{r count_sig_genes}
lfc_cutoff <- 0.5
padj_cutoff <- 0.05

sig_counts_si3d_si3e <- te_data %>%
  filter(!is.na(padj), padj < padj_cutoff, abs(log2FoldChange) >= lfc_cutoff) %>%
  mutate(Direction = if_else(log2FoldChange > 0, "Up", "Down")) %>%
  dplyr::count(treatment, Direction, name = "n") %>%
  mutate(
    Direction = factor(Direction, levels = c("Up", "Down")),
    treatment = factor(treatment, levels = c("si3d", "si3e"))
  )

print(sig_counts_si3d_si3e)
```

# Bar chart: si3d vs si3e TE changes (1+4hr time-regressed)

Direction on the y-axis (Up = Okabe-Ito vermillion = eIF3d/eIF3e-promoted;
Down = Okabe-Ito dark blue = eIF3d/eIF3e-inhibited), dodged by treatment
(si3d vs si3e) — same visual language as notebooks 01/02, with treatment
taking the place spike-in status held there.

```{r plot_si3d_si3e_bar, fig.width=6, fig.height=3.5}
up_dark <- "#D55E00"
down_dark <- "#0072B2"
lighten <- function(hex, amount = 0.55) colorRampPalette(c(hex, "white"))(101)[round(amount * 100) + 1]
up_light <- lighten(up_dark)
down_light <- lighten(down_dark)

color_values <- c(
  "Up - si3d" = up_dark,
  "Up - si3e" = up_light,
  "Down - si3d" = down_dark,
  "Down - si3e" = down_light
)

plot_df <- sig_counts_si3d_si3e %>%
  mutate(color_group = factor(paste(Direction, treatment, sep = " - "), levels = names(color_values)))

dodge <- position_dodge(width = 0.7)
x_max <- max(plot_df$n) * 1.15

p_si3d_si3e <- ggplot(plot_df, aes(y = Direction, x = n, fill = color_group)) +
  geom_col(position = dodge, width = 0.6) +
  geom_text(aes(label = n), position = dodge, hjust = -0.2, size = 3.5, fontface = "bold") +
  geom_vline(xintercept = 0, linewidth = 1) +
  scale_fill_manual(values = color_values) +
  scale_x_continuous(limits = c(0, x_max)) +
  labs(title = "eIF3d/eIF3e knockdown TE changes: si3d/si3e vs sictrl, normoxia (1+4hr)",
       x = "# sig genes (padj < 0.05, |log2FC| >= 0.5)", y = NULL, fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        axis.line.x = element_line(linewidth = 0.9),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

p_si3d_si3e
ggsave(here("plots/03_bar_te_si3d_si3e_normoxia.pdf"), plot = p_si3d_si3e, width = 6, height = 3.5)
```

# Save results

```{r save_results}
write_csv(sig_counts_si3d_si3e, here("output/spikein_literature_comparison/03_sig_te_counts_si3d_si3e.csv"))
cat("Saved si3d/si3e TE sig gene counts to output/spikein_literature_comparison/\n")
```

# Literature comparison: Lin et al. 2020 (MCF-10A, eIF3e knockdown)

Lin Y, et al. "eIF3 Associates with 80S Ribosomes to Promote Translation
Elongation, Mitochondrial Homeostasis, and Muscle Health." *Molecular Cell*
2020. DOI: 10.1016/j.molcel.2020.06.003. Ribosome profiling data: GEO
GSE131074.

- **Cell type:** MCF-10A (human non-tumorigenic breast epithelial cell
  line) — not a hypoxia experiment; this is a baseline (normoxic) eIF3e
  knockdown, so it compares against our si3e-vs-sictrl normoxia (1+4hr)
  number above, not a hypoxia condition.
- **Knockdown:** si-eIF3e (siRNA), 72h, vs si-Ctrl, triplicate.
- **Method:** ribosome profiling (RPF) + total RNA-seq (matched design to
  our RNA/Ribo/TE setup).
- **Their significance criteria** (results text, quoted directly by Kate
  from the paper): "several hundred significant changes in translational
  efficiency (TE) defined as changes in ribosome protected fragments (RPFs,
  p < 0.01) but not in total mRNA (p > 0.01)." **No log2FC magnitude
  threshold** and **raw p, not padj/FDR** — a materially different, more
  idiosyncratic scheme than ours (`padj < 0.05, |log2FC| >= 0.5`) or
  Hettiarachchi/Shen's (padj/FDR + |log2FC| magnitude cutoffs). Counts below
  are their own reported totals, not recomputed under our thresholds.
- **Group identity, confirmed from Figure 1C itself** (Kate provided the
  actual figure image): y-axis = `log2 RPF (si-eIF3e/si-Ctrl)`. The
  **orange group is "220 up"** (positive y = RPF higher in the KD),
  enriched for Transmembrane/Mitochondrion/Glycoprotein/ER-Golgi
  transport/Translation. The **blue group is "240 down"** (negative y =
  RPF lower in the KD), enriched for Cell-cell adhesion/mRNA
  processing/Pentose phosphate pathway/Exonuclease/Ubiquitin ligase
  activity. This matches the paper's results-text prose, not the
  figure-legend-style text Kate found earlier (which had misattributed
  240↔orange and 220↔blue relative to the actual figure).
- **Why this is not a simple reciprocal sign-flip.** A naive flip (their
  fold change is KD-over-control, ours is control-over-KD, so just negate)
  would call the 220 "up-in-RPF" mitochondrial/transmembrane group
  eIF3e-**inhibited**. But the paper's own results section is explicitly
  titled *"Loss of eIF3e Curbs the Synthesis of Mitochondrial Proteins
  Despite Paradoxical Effects on Ribosome Occupancy"* — i.e., RPF
  occupancy going **up** for this group in the KD is a documented
  **paradox** caused by ribosome stalling/queueing during defective
  elongation, not increased productive translation. The paper validates
  this with pulsed-SILAC: actual mitochondrial protein *synthesis*
  **decreases** in the KD, confirming eIF3e **promotes** this group,
  the opposite of what the raw RPF-ratio sign alone would suggest. The
  paper's Highlights state this directly: "eIF3-via eIF3e-promotes mRNA
  selective early translation elongation" of "membrane, secretory, and
  organelle targeted proteins" — exactly the 220/orange group.
- **Mapping used here:**
  - **220 mRNAs** (orange, "up" in RPF, mitochondrial/transmembrane) =
    eIF3e-**promoted** per the paper's SILAC-validated conclusion = our
    **Up**.
  - **240 mRNAs** (blue, "down" in RPF, cell adhesion/mRNA processing) =
    our **Down** (eIF3e-inhibited/not part of the elongation-promoted
    group).

```{r combine_si3e_and_lin}
lin_groups <- tribble(
  ~source, ~Direction, ~n,
  "Lin et al. 2020<br>(MCF-10A, si-eIF3e 72h, RPF p<0.01 & mRNA p>0.01)", "Up", 220,
  "Lin et al. 2020<br>(MCF-10A, si-eIF3e 72h, RPF p<0.01 & mRNA p>0.01)", "Down", 240
)

our_si3e <- sig_counts_si3d_si3e %>%
  filter(treatment == "si3e") %>%
  transmute(source = "**si3e (1+4hr)**", Direction, n = n)

si3e_source_levels <- c(
  "**si3e (1+4hr)**",
  "Lin et al. 2020<br>(MCF-10A, si-eIF3e 72h, RPF p<0.01 & mRNA p>0.01)"
)

si3e_vs_lin <- bind_rows(our_si3e, lin_groups) %>%
  mutate(Direction = factor(Direction, levels = c("Up", "Down")),
         source = factor(source, levels = si3e_source_levels))

print(si3e_vs_lin)
```

```{r plot_si3e_vs_lin, fig.width=7.5, fig.height=3.5}
dodge2 <- position_dodge(width = 0.7)
x_pad3 <- 0.15 * max(si3e_vs_lin$n)

p_si3e_vs_lin <- ggplot(si3e_vs_lin, aes(y = source, x = n, fill = Direction)) +
  geom_col(position = dodge2, width = 0.6) +
  geom_text(aes(label = n), position = dodge2, hjust = -0.25, size = 3.5, fontface = "bold") +
  geom_vline(xintercept = 0, linewidth = 1) +
  scale_fill_manual(values = c(Up = up_dark, Down = down_dark)) +
  scale_x_continuous(limits = c(0, max(si3e_vs_lin$n) + x_pad3)) +
  scale_y_discrete(limits = rev(si3e_source_levels)) +
  labs(title = "eIF3e TE changes: ours (si3e vs sictrl, normoxia, 1+4hr) vs Lin et al. 2020 (MCF-10A)",
       subtitle = "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Lin et al. counts are their own reported totals (RPF p < 0.01, mRNA p > 0.01, no LFC threshold), sign-flipped to match our eIF3e-promotes convention.",
       x = "# sig genes", y = NULL, fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(color = "grey40", size = 8),
        axis.line.x = element_line(linewidth = 0.9),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_markdown(size = 9.5))

p_si3e_vs_lin
ggsave(here("plots/03_bar_si3e_vs_lin.pdf"), plot = p_si3e_vs_lin, width = 7.5, height = 3.5)
```

# Literature comparison: Okubo et al. 2025 (human iPSC, EIF3D knockdown)

Okubo C, et al. "EIF3D safeguards the homeostasis of key signaling pathways
in human primed pluripotency." *Science Advances* 2025;11:eadq5484. PMC
PMC11980838.

- **Cell type:** human iPSC lines **WTB6** and **1B4** (CRISPRi Gen 1 clone
  4 derived from WTB6) — not a hypoxia experiment; baseline (normoxic)
  EIF3D loss-of-function, so it compares against our si3d-vs-sictrl
  normoxia (1+4hr) number, not a hypoxia condition.
- **Knockdown:** CRISPRi (doxycycline-inducible), harvested day 3
  post-doxycycline.
- **No spike-in normalization used** — confirmed directly: "Read counting
  and normalization were performed using fp-count and DESeq2,
  respectively" (standard library-size normalization, no ERCC or
  exogenous spike-ins).
- **Their cutoffs:** padj < 0.05 AND |log2FC| > 1 — same padj threshold as
  ours, but double the LFC magnitude (ours is |log2FC| >= 0.5).
- **Sign convention — flipped to match ours.** Their fold-change formula,
  quoted directly from methods: `Fold change_i = TE_i(sgEIF3D) /
  TE_i(control)` — i.e., **KD-over-control**, confirmed explicit (not a
  paradox case like Lin et al. above — no stalling/occupancy caveat noted
  in this paper). Ours is `log2(sictrl / si3d)` (control-over-KD), the
  reciprocal direction. So:
  - Their **402 genes, increased in the KD** = TE higher in sgEIF3D = our
    `te_lfc < 0` = our **Down** (eIF3d-inhibited).
  - Their **919 genes, decreased in the KD** = TE lower in sgEIF3D = our
    `te_lfc > 0` = our **Up** (eIF3d-promoted).

```{r combine_si3d_and_okubo}
okubo_groups <- tribble(
  ~source, ~Direction, ~n,
  "Okubo et al. 2025<br>(human iPSC, EIF3D CRISPRi day 3, padj<0.05 & |log2FC|>1)", "Down", 402,
  "Okubo et al. 2025<br>(human iPSC, EIF3D CRISPRi day 3, padj<0.05 & |log2FC|>1)", "Up", 919
)

our_si3d <- sig_counts_si3d_si3e %>%
  filter(treatment == "si3d") %>%
  transmute(source = "**si3d (1+4hr)**", Direction, n = n)

si3d_source_levels <- c(
  "**si3d (1+4hr)**",
  "Okubo et al. 2025<br>(human iPSC, EIF3D CRISPRi day 3, padj<0.05 & |log2FC|>1)"
)

si3d_vs_okubo <- bind_rows(our_si3d, okubo_groups) %>%
  mutate(Direction = factor(Direction, levels = c("Up", "Down")),
         source = factor(source, levels = si3d_source_levels))

print(si3d_vs_okubo)
```

```{r plot_si3d_vs_okubo, fig.width=7.5, fig.height=3.5}
dodge3 <- position_dodge(width = 0.7)
x_pad4 <- 0.15 * max(si3d_vs_okubo$n)

p_si3d_vs_okubo <- ggplot(si3d_vs_okubo, aes(y = source, x = n, fill = Direction)) +
  geom_col(position = dodge3, width = 0.6) +
  geom_text(aes(label = n), position = dodge3, hjust = -0.25, size = 3.5, fontface = "bold") +
  geom_vline(xintercept = 0, linewidth = 1) +
  scale_fill_manual(values = c(Up = up_dark, Down = down_dark)) +
  scale_x_continuous(limits = c(0, max(si3d_vs_okubo$n) + x_pad4)) +
  scale_y_discrete(limits = rev(si3d_source_levels)) +
  labs(title = "eIF3d TE changes: ours (si3d vs sictrl, normoxia, 1+4hr) vs Okubo et al. 2025 (human iPSC)",
       subtitle = "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Okubo et al. counts are their own reported totals (padj < 0.05, |log2FC| > 1), sign-flipped to match our eIF3d-promotes convention.",
       x = "# sig genes", y = NULL, fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(color = "grey40", size = 8),
        axis.line.x = element_line(linewidth = 0.9),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_markdown(size = 9.5))

p_si3d_vs_okubo
ggsave(here("plots/03_bar_si3d_vs_okubo.pdf"), plot = p_si3d_vs_okubo, width = 7.5, height = 3.5)
```

# Literature comparison: Herrmannová et al. 2024 (HeLa, eIF3d and eIF3e knockdown)

Herrmannová A, et al. "Perturbations in eIF3 subunit stoichiometry alter
expression of ribosomal proteins and key components of the MAPK signaling
pathways." *eLife* 2024;13:e95846. DOI: 10.7554/eLife.95846. PMC PMC11534336.

- **Cell type:** HeLa cells — not a hypoxia experiment; baseline (normoxic)
  siRNA knockdown, so it compares against our si3d/si3e-vs-sictrl normoxia
  (1+4hr) numbers.
- **Knockdown:** siRNA against eIF3d or eIF3e (also eIF3h, not used here),
  vs non-targeting (NT) control; duration not explicitly stated in the
  text retrieved (16-32-fold mRNA knockdown efficiency by qPCR).
- **No spike-in normalization found** in the methods text retrieved — DESeq2
  standard library-size normalization ("Using the R DESeq2 package, we
  identified transcripts with significantly altered TE"). The full methods
  section could not be completely retrieved, so this is reported with
  slightly less confidence than the explicit statements for Lin et al. and
  Okubo et al. above.
- **Their cutoffs:** padj < 0.05 only — **no log2FC magnitude threshold**
  stated, unlike ours (`padj < 0.05, |log2FC| >= 0.5`).
- **Sign convention — flipped to match ours.** "Up"/"down" in their reported
  DTEG (differential TE gene) counts refer to increased/decreased TE **in
  the knockdown relative to the NT control** (confirmed directly), the same
  KD-over-control direction as Okubo et al. above (no stalling/paradox
  caveat noted for this paper). So:
  - Their **eIF3e-KD: 1180 DTEGs total (622 down, 558 up in the KD).**
    558 up-in-KD → our `te_lfc < 0` → our **Down** = 558. 622 down-in-KD →
    our `te_lfc > 0` → our **Up** = 622.
  - Their **eIF3d-KD: 1971 DTEGs total (1040 down, 931 up in the KD).**
    931 up-in-KD → our **Down** = 931. 1040 down-in-KD → our **Up** = 1040.

```{r combine_si3d_si3e_and_herrmannova}
herrmannova_groups <- tribble(
  ~source, ~target, ~Direction, ~n,
  "Herrmannova et al. 2024<br>(HeLa, siEIF3D, padj<0.05, no LFC threshold)", "eIF3d", "Up", 1040,
  "Herrmannova et al. 2024<br>(HeLa, siEIF3D, padj<0.05, no LFC threshold)", "eIF3d", "Down", 931,
  "Herrmannova et al. 2024<br>(HeLa, siEIF3E, padj<0.05, no LFC threshold)", "eIF3e", "Up", 622,
  "Herrmannova et al. 2024<br>(HeLa, siEIF3E, padj<0.05, no LFC threshold)", "eIF3e", "Down", 558
)

our_si3d_si3e_labeled <- sig_counts_si3d_si3e %>%
  mutate(
    target = recode(as.character(treatment), si3d = "eIF3d", si3e = "eIF3e"),
    source = paste0("**", as.character(treatment), " (1+4hr)**")
  ) %>%
  dplyr::select(source, target, Direction, n)

target_levels <- c("eIF3d", "eIF3e")
si3d_si3e_source_levels <- c(
  "**si3d (1+4hr)**", "Herrmannova et al. 2024<br>(HeLa, siEIF3D, padj<0.05, no LFC threshold)",
  "**si3e (1+4hr)**", "Herrmannova et al. 2024<br>(HeLa, siEIF3E, padj<0.05, no LFC threshold)"
)

si3d_si3e_vs_herrmannova <- bind_rows(our_si3d_si3e_labeled, herrmannova_groups) %>%
  mutate(
    Direction = factor(Direction, levels = c("Up", "Down")),
    target = factor(target, levels = target_levels),
    source = factor(source, levels = si3d_si3e_source_levels)
  )

print(si3d_si3e_vs_herrmannova)
```

```{r plot_si3d_si3e_vs_herrmannova, fig.width=7.5, fig.height=5.5}
dodge4 <- position_dodge(width = 0.7)
x_pad5 <- 0.15 * max(si3d_si3e_vs_herrmannova$n)

p_si3d_si3e_vs_herrmannova <- ggplot(si3d_si3e_vs_herrmannova, aes(y = source, x = n, fill = Direction)) +
  geom_col(position = dodge4, width = 0.6) +
  geom_text(aes(label = n), position = dodge4, hjust = -0.25, size = 3.5, fontface = "bold") +
  geom_vline(xintercept = 0, linewidth = 1) +
  scale_fill_manual(values = c(Up = up_dark, Down = down_dark)) +
  scale_x_continuous(limits = c(0, max(si3d_si3e_vs_herrmannova$n) + x_pad5)) +
  scale_y_discrete(limits = rev(si3d_si3e_source_levels)) +
  facet_wrap(~target, ncol = 1, scales = "free_y") +
  labs(title = "eIF3d/eIF3e TE changes: ours (normoxia, 1+4hr) vs Herrmannova et al. 2024 (HeLa)",
       subtitle = "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Herrmannova et al. counts are their own reported totals (padj < 0.05, no LFC threshold), sign-flipped to match our eIF3d/eIF3e-promotes convention.",
       x = "# sig genes", y = NULL, fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(color = "grey40", size = 8),
        axis.line.x = element_line(linewidth = 0.9),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_markdown(size = 9.5),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 13))

p_si3d_si3e_vs_herrmannova
ggsave(here("plots/03_bar_si3d_si3e_vs_herrmannova.pdf"), plot = p_si3d_si3e_vs_herrmannova, width = 7.5, height = 5.5)
```

# Combined plot: all sources, separate plots for eIF3d and eIF3e

Everything from this notebook, as two separate figures (one per target) —
ours (si3d/si3e, 1+4hr) plus Lin et al. 2020, Okubo et al. 2025, and
Herrmannova et al. 2024. Each source's own caveats (cutoffs, spike-in
status, sign-flip logic) are documented in the sections above; these plots
do not repeat them, only the comparison itself.

```{r combine_all_sources}
lin_groups_labeled <- lin_groups %>% mutate(target = "eIF3e")
okubo_groups_labeled <- okubo_groups %>% mutate(target = "eIF3d")

all_source_levels <- c(
  "**si3d (1+4hr)**",
  "Okubo et al. 2025<br>(human iPSC, EIF3D CRISPRi day 3, padj<0.05 & |log2FC|>1)",
  "Herrmannova et al. 2024<br>(HeLa, siEIF3D, padj<0.05, no LFC threshold)",
  "**si3e (1+4hr)**",
  "Lin et al. 2020<br>(MCF-10A, si-eIF3e 72h, RPF p<0.01 & mRNA p>0.01)",
  "Herrmannova et al. 2024<br>(HeLa, siEIF3E, padj<0.05, no LFC threshold)"
)

all_vs_literature <- bind_rows(
  our_si3d_si3e_labeled,
  lin_groups_labeled,
  okubo_groups_labeled,
  herrmannova_groups
) %>%
  mutate(
    Direction = factor(Direction, levels = c("Up", "Down")),
    target = factor(target, levels = target_levels),
    source = factor(source, levels = all_source_levels)
  )

print(all_vs_literature)
```

Color now maps to **source/dataset** (Okabe-Ito palette) instead of
Direction, so it's immediately visible whose data each bar is: ours (si3d =
orange, si3e = sky blue) vs. each literature source in its own remaining
Okabe-Ito color. Since color no longer distinguishes Up from Down, that's
carried by transparency instead (Up = solid, Down = faded) plus the value
labels.

```{r plot_all_vs_literature, fig.width=8, fig.height=4}
source_colors <- c(
  "**si3d (1+4hr)**" = "#E69F00",
  "**si3e (1+4hr)**" = "#56B4E9",
  "Okubo et al. 2025<br>(human iPSC, EIF3D CRISPRi day 3, padj<0.05 & |log2FC|>1)" = "#009E73",
  "Lin et al. 2020<br>(MCF-10A, si-eIF3e 72h, RPF p<0.01 & mRNA p>0.01)" = "#CC79A7",
  "Herrmannova et al. 2024<br>(HeLa, siEIF3D, padj<0.05, no LFC threshold)" = "#D55E00",
  "Herrmannova et al. 2024<br>(HeLa, siEIF3E, padj<0.05, no LFC threshold)" = "#D55E00"
)

make_target_plot <- function(df, target_label) {
  df <- droplevels(df)
  source_levels_local <- intersect(all_source_levels, levels(df$source))
  dodge_local <- position_dodge(width = 0.7)
  x_pad_local <- 0.15 * max(df$n)

  ggplot(df, aes(y = source, x = n, fill = source, group = Direction)) +
    geom_col(position = dodge_local, width = 0.6) +
    geom_text(aes(label = n, color = Direction), position = dodge_local,
              hjust = -0.25, size = 3.5, fontface = "bold", key_glyph = "point") +
    geom_vline(xintercept = 0, linewidth = 1) +
    scale_fill_manual(values = source_colors, guide = "none") +
    scale_color_manual(values = c(Up = up_dark, Down = down_dark), name = NULL) +
    scale_x_continuous(limits = c(0, max(df$n) + x_pad_local)) +
    scale_y_discrete(limits = rev(source_levels_local)) +
    guides(color = guide_legend(override.aes = list(size = 4))) +
    labs(title = paste0(target_label, " TE changes: ours (normoxia, 1+4hr) vs all literature comparisons"),
         subtitle = "Ours uses OUR cutoffs (padj < 0.05, |log2FC| >= 0.5). Each literature source uses its own reported totals and cutoffs (see sections above), sign-flipped to match our promotes convention. Bar color = source; label color = Up (vermillion) / Down (blue).",
         x = "# sig genes", y = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top",
          plot.subtitle = element_text(color = "grey40", size = 8),
          axis.line.x = element_line(linewidth = 0.9),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.y = element_markdown(size = 9.5))
}

p_all_vs_literature_eif3d <- make_target_plot(all_vs_literature %>% filter(target == "eIF3d"), "eIF3d")
p_all_vs_literature_eif3d
ggsave(here("plots/03_bar_all_vs_literature_eif3d.pdf"), plot = p_all_vs_literature_eif3d, width = 8, height = 4)

p_all_vs_literature_eif3e <- make_target_plot(all_vs_literature %>% filter(target == "eIF3e"), "eIF3e")
p_all_vs_literature_eif3e
ggsave(here("plots/03_bar_all_vs_literature_eif3e.pdf"), plot = p_all_vs_literature_eif3e, width = 8, height = 4)
```

# Session Info

```{r session_info}
sessionInfo()
```
