---
title: "Acute Hypoxic Response vs 3e/3d KD Sensitivity"
author: "Kate"
date: "2025-12-12"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(here)
library(janitor)
library(tidyverse)
library(ggvenn)
library(ggpubr)
library(DT)
library(cowplot)
library(GeneOverlap)
library(writexl)
library(reshape2)
library(VennDiagram)
```

```{r functions}
source(here("code/functions.R"))
cat("Available functions: createNormMatrix, deFunction\n")
```

## Read in gene lists

### Acute hypoxic response genes (sensitive to 3e/3d KD)

```{r read-acute-hypoxic-response}
# Read in the combined signatures file
acute_sigs <- read_csv(here("counts", "all_combined_sigs_12_9_25.csv"))

# Check the structure
head(acute_sigs)
colnames(acute_sigs)
table(acute_sigs$gs_name)
```

### 3e/3d KD gene signatures in normoxia

```{r read-normoxia-3d-signatures}
# 3d signatures - normoxia
normoxia_3d_promotes_1hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_1hr.csv"))
normoxia_3d_inhibits_1hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_1hr.csv"))

normoxia_3d_promotes_4hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_4hr.csv"))
normoxia_3d_inhibits_4hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_4hr.csv"))

normoxia_3d_promotes_24hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_24hr.csv"))
normoxia_3d_inhibits_24hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_24hr.csv"))

# Combined timepoints
normoxia_3d_promotes_combined <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_combined.csv"))
normoxia_3d_inhibits_combined <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_combined.csv"))
```

```{r read-normoxia-3e-signatures}
# 3e signatures - normoxia
normoxia_3e_promotes_1hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_1hr.csv"))
normoxia_3e_inhibits_1hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_1hr.csv"))

normoxia_3e_promotes_4hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_4hr.csv"))
normoxia_3e_inhibits_4hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_4hr.csv"))

normoxia_3e_promotes_24hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_24hr.csv"))
normoxia_3e_inhibits_24hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_24hr.csv"))

# Combined timepoints
normoxia_3e_promotes_combined <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_combined.csv"))
normoxia_3e_inhibits_combined <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_combined.csv"))
```

### 3e/3d KD gene signatures in hypoxia

```{r read-hypoxia-3d-signatures}
# 3d signatures - hypoxia
hypoxia_3d_promotes_1hr <- read_csv(here("output/genesets/hypoxia_3d_promotes_TE_1hr.csv"))
hypoxia_3d_inhibits_1hr <- read_csv(here("output/genesets/hypoxia_3d_inhibits_TE_1hr.csv"))

hypoxia_3d_promotes_4hr <- read_csv(here("output/genesets/hypoxia_3d_promotes_TE_4hr.csv"))
hypoxia_3d_inhibits_4hr <- read_csv(here("output/genesets/hypoxia_3d_inhibits_TE_4hr.csv"))

hypoxia_3d_promotes_24hr <- read_csv(here("output/genesets/hypoxia_3d_promotes_TE_24hr.csv"))
hypoxia_3d_inhibits_24hr <- read_csv(here("output/genesets/hypoxia_3d_inhibits_TE_24hr.csv"))
```

```{r read-hypoxia-3e-signatures}
# 3e signatures - hypoxia
hypoxia_3e_promotes_1hr <- read_csv(here("output/genesets/hypoxia_3e_promotes_TE_1hr.csv"))
hypoxia_3e_inhibits_1hr <- read_csv(here("output/genesets/hypoxia_3e_inhibits_TE_1hr.csv"))

hypoxia_3e_promotes_4hr <- read_csv(here("output/genesets/hypoxia_3e_promotes_TE_4hr.csv"))
hypoxia_3e_inhibits_4hr <- read_csv(here("output/genesets/hypoxia_3e_inhibits_TE_4hr.csv"))

hypoxia_3e_promotes_24hr <- read_csv(here("output/genesets/hypoxia_3e_promotes_TE_24hr.csv"))
hypoxia_3e_inhibits_24hr <- read_csv(here("output/genesets/hypoxia_3e_inhibits_TE_24hr.csv"))
```

## Summary of gene lists

```{r summary}
# Create a summary table of gene counts
summary_df <- data.frame(
  condition = c(
    # Normoxia 3d
    "normoxia_3d_promotes_1hr", "normoxia_3d_inhibits_1hr",
    "normoxia_3d_promotes_4hr", "normoxia_3d_inhibits_4hr",
    "normoxia_3d_promotes_24hr", "normoxia_3d_inhibits_24hr",
    "normoxia_3d_promotes_combined", "normoxia_3d_inhibits_combined",
    # Normoxia 3e
    "normoxia_3e_promotes_1hr", "normoxia_3e_inhibits_1hr",
    "normoxia_3e_promotes_4hr", "normoxia_3e_inhibits_4hr",
    "normoxia_3e_promotes_24hr", "normoxia_3e_inhibits_24hr",
    "normoxia_3e_promotes_combined", "normoxia_3e_inhibits_combined",
    # Hypoxia 3d
    "hypoxia_3d_promotes_1hr", "hypoxia_3d_inhibits_1hr",
    "hypoxia_3d_promotes_4hr", "hypoxia_3d_inhibits_4hr",
    "hypoxia_3d_promotes_24hr", "hypoxia_3d_inhibits_24hr",
    # Hypoxia 3e
    "hypoxia_3e_promotes_1hr", "hypoxia_3e_inhibits_1hr",
    "hypoxia_3e_promotes_4hr", "hypoxia_3e_inhibits_4hr",
    "hypoxia_3e_promotes_24hr", "hypoxia_3e_inhibits_24hr"
  ),
  n_genes = c(
    nrow(normoxia_3d_promotes_1hr), nrow(normoxia_3d_inhibits_1hr),
    nrow(normoxia_3d_promotes_4hr), nrow(normoxia_3d_inhibits_4hr),
    nrow(normoxia_3d_promotes_24hr), nrow(normoxia_3d_inhibits_24hr),
    nrow(normoxia_3d_promotes_combined), nrow(normoxia_3d_inhibits_combined),
    nrow(normoxia_3e_promotes_1hr), nrow(normoxia_3e_inhibits_1hr),
    nrow(normoxia_3e_promotes_4hr), nrow(normoxia_3e_inhibits_4hr),
    nrow(normoxia_3e_promotes_24hr), nrow(normoxia_3e_inhibits_24hr),
    nrow(normoxia_3e_promotes_combined), nrow(normoxia_3e_inhibits_combined),
    nrow(hypoxia_3d_promotes_1hr), nrow(hypoxia_3d_inhibits_1hr),
    nrow(hypoxia_3d_promotes_4hr), nrow(hypoxia_3d_inhibits_4hr),
    nrow(hypoxia_3d_promotes_24hr), nrow(hypoxia_3d_inhibits_24hr),
    nrow(hypoxia_3e_promotes_1hr), nrow(hypoxia_3e_inhibits_1hr),
    nrow(hypoxia_3e_promotes_4hr), nrow(hypoxia_3e_inhibits_4hr),
    nrow(hypoxia_3e_promotes_24hr), nrow(hypoxia_3e_inhibits_24hr)
  )
)

datatable(summary_df)
```

## Extract gene sets from combined signatures file

```{r extract-acute-hypoxic-3d-sensitive}
# Extract genes upregulated in acute hypoxic response that are inhibited by 3d KD
acute_up_3d_sensitive_1hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd") %>%
  pull(db_gene_symbol)

acute_up_3d_sensitive_4hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd") %>%
  pull(db_gene_symbol)

acute_up_3d_sensitive_24hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd") %>%
  pull(db_gene_symbol)

# Extract genes promoted by 3d in hypoxia
hypoxia_3d_promotes_1hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3d_promotes_TE_1hr") %>%
  pull(db_gene_symbol)

hypoxia_3d_promotes_4hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3d_promotes_TE_4hr") %>%
  pull(db_gene_symbol)

hypoxia_3d_promotes_24hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3d_promotes_TE_24hr") %>%
  pull(db_gene_symbol)

# Print counts
cat("Genes upregulated in acute hypoxic response and sensitive to 3d KD:\n")
cat("  1hr:", length(acute_up_3d_sensitive_1hr), "\n")
cat("  4hr:", length(acute_up_3d_sensitive_4hr), "\n")
cat("  24hr:", length(acute_up_3d_sensitive_24hr), "\n\n")

cat("Genes promoted by 3d in hypoxia:\n")
cat("  1hr:", length(hypoxia_3d_promotes_1hr_genes), "\n")
cat("  4hr:", length(hypoxia_3d_promotes_4hr_genes), "\n")
cat("  24hr:", length(hypoxia_3d_promotes_24hr_genes), "\n")
```

## Comparison: Acute hypoxic response (3d-sensitive) vs 3d promotes in hypoxia

# 3d

### 1 hour timepoint

```{r venn-3d-1hr}
# Create Venn diagram
venn_list_1hr <- list(
  "3d-sensitive in 1hr hypoxic response" = acute_up_3d_sensitive_1hr,
  "3d promotes TE in 1hr hypoxia" = hypoxia_3d_promotes_1hr_genes
)

p_venn_1hr <- ggvenn(
  venn_list_1hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_1hr


ggsave(here("plots/venn_acute_hypoxic_3d_sensitive_vs_3d_promotes_1hr.pdf"),
       p_venn_1hr, width = 6, height = 5)
```

```{r overlap-stats-3d-1hr}
# Calculate overlap statistics
overlap_1hr <- intersect(acute_up_3d_sensitive_1hr, hypoxia_3d_promotes_1hr_genes)
only_acute_1hr <- setdiff(acute_up_3d_sensitive_1hr, hypoxia_3d_promotes_1hr_genes)
only_promotes_1hr <- setdiff(hypoxia_3d_promotes_1hr_genes, acute_up_3d_sensitive_1hr)

cat("1 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_1hr), "\n")
cat("  Only in acute hypoxic (3d-sensitive):", length(only_acute_1hr), "\n")
cat("  Only in 3d promotes:", length(only_promotes_1hr), "\n")
cat("  % of acute hypoxic that are 3d-promoted:",
    round(100 * length(overlap_1hr) / length(acute_up_3d_sensitive_1hr), 1), "%\n")
cat("  % of 3d-promoted that are acute hypoxic:",
    round(100 * length(overlap_1hr) / length(hypoxia_3d_promotes_1hr_genes), 1), "%\n")

# GeneOverlap test
go_1hr <- newGeneOverlap(acute_up_3d_sensitive_1hr,
                         hypoxia_3d_promotes_1hr_genes,
                         genome.size = 20000)
go_1hr <- testGeneOverlap(go_1hr)
cat("  Fisher's exact test p-value:", format(getPval(go_1hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_1hr), 3), "\n")
```

### 4 hour timepoint

```{r venn-3d-4hr}
# Create Venn diagram
venn_list_4hr <- list(
  "3d-sensitive in 4hr hypoxic response" = acute_up_3d_sensitive_4hr,
  "3d promotes TE in 4hr hypoxia" = hypoxia_3d_promotes_4hr_genes
)

p_venn_4hr <- ggvenn(
  venn_list_4hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_4hr


ggsave(here("plots/venn_acute_hypoxic_3d_sensitive_vs_3d_promotes_4hr.pdf"),
       p_venn_4hr, width = 6, height = 5)
```

```{r overlap-stats-3d-4hr}
# Calculate overlap statistics
overlap_4hr <- intersect(acute_up_3d_sensitive_4hr, hypoxia_3d_promotes_4hr_genes)
only_acute_4hr <- setdiff(acute_up_3d_sensitive_4hr, hypoxia_3d_promotes_4hr_genes)
only_promotes_4hr <- setdiff(hypoxia_3d_promotes_4hr_genes, acute_up_3d_sensitive_4hr)

cat("4 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_4hr), "\n")
cat("  Only in acute hypoxic (3d-sensitive):", length(only_acute_4hr), "\n")
cat("  Only in 3d promotes:", length(only_promotes_4hr), "\n")
cat("  % of acute hypoxic that are 3d-promoted:",
    round(100 * length(overlap_4hr) / length(acute_up_3d_sensitive_4hr), 1), "%\n")
cat("  % of 3d-promoted that are acute hypoxic:",
    round(100 * length(overlap_4hr) / length(hypoxia_3d_promotes_4hr_genes), 1), "%\n")

# GeneOverlap test
go_4hr <- newGeneOverlap(acute_up_3d_sensitive_4hr,
                         hypoxia_3d_promotes_4hr_genes,
                         genome.size = 20000)
go_4hr <- testGeneOverlap(go_4hr)
cat("  Fisher's exact test p-value:", format(getPval(go_4hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_4hr), 3), "\n")
```

### 24 hour timepoint

```{r venn-3d-24hr}
# Create Venn diagram
venn_list_24hr <- list(
  "3d-sensitive in 24hr hypoxic response" = acute_up_3d_sensitive_24hr,
  "3d promotes TE in 24hr hypoxia" = hypoxia_3d_promotes_24hr_genes
)

p_venn_24hr <- ggvenn(
  venn_list_24hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_24hr


ggsave(here("plots/venn_acute_hypoxic_3d_sensitive_vs_3d_promotes_24hr.pdf"),
       p_venn_24hr, width = 6, height = 5)
```

```{r overlap-stats-3d-24hr}
# Calculate overlap statistics
overlap_24hr <- intersect(acute_up_3d_sensitive_24hr, hypoxia_3d_promotes_24hr_genes)
only_acute_24hr <- setdiff(acute_up_3d_sensitive_24hr, hypoxia_3d_promotes_24hr_genes)
only_promotes_24hr <- setdiff(hypoxia_3d_promotes_24hr_genes, acute_up_3d_sensitive_24hr)

cat("4 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_24hr), "\n")
cat("  Only in acute hypoxic (3d-sensitive):", length(only_acute_24hr), "\n")
cat("  Only in 3d promotes:", length(only_promotes_24hr), "\n")
cat("  % of acute hypoxic that are 3d-promoted:",
    round(100 * length(overlap_24hr) / length(acute_up_3d_sensitive_24hr), 1), "%\n")
cat("  % of 3d-promoted that are acute hypoxic:",
    round(100 * length(overlap_24hr) / length(hypoxia_3d_promotes_24hr_genes), 1), "%\n")

# GeneOverlap test
go_24hr <- newGeneOverlap(acute_up_3d_sensitive_24hr,
                         hypoxia_3d_promotes_24hr_genes,
                         genome.size = 20000)
go_24hr <- testGeneOverlap(go_24hr)
cat("  Fisher's exact test p-value:", format(getPval(go_24hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_24hr), 3), "\n")
```


### Summary across timepoints

```{r summary-3d-overlap}
# Create summary table
overlap_summary <- data.frame(
  timepoint = c("1hr", "4hr", "24hr"),
  acute_hypoxic_3d_sensitive = c(
    length(acute_up_3d_sensitive_1hr),
    length(acute_up_3d_sensitive_4hr),
    length(acute_up_3d_sensitive_24hr)
  ),
  hypoxia_3d_promotes = c(
    length(hypoxia_3d_promotes_1hr_genes),
    length(hypoxia_3d_promotes_4hr_genes),
    length(hypoxia_3d_promotes_24hr_genes)
  ),
  overlap = c(
    length(overlap_1hr),
    length(overlap_4hr),
    length(overlap_24hr)
  ),
  pct_acute_in_promotes = c(
    round(100 * length(overlap_1hr) / length(acute_up_3d_sensitive_1hr), 1),
    round(100 * length(overlap_4hr) / length(acute_up_3d_sensitive_4hr), 1),
    round(100 * length(overlap_24hr) / length(acute_up_3d_sensitive_24hr), 1)
  ),
  pct_promotes_in_acute = c(
    round(100 * length(overlap_1hr) / length(hypoxia_3d_promotes_1hr_genes), 1),
    round(100 * length(overlap_4hr) / length(hypoxia_3d_promotes_4hr_genes), 1),
    round(100 * length(overlap_24hr) / length(hypoxia_3d_promotes_24hr_genes), 1)
  ),
  jaccard_index = c(
    round(getJaccard(go_1hr), 3),
    round(getJaccard(go_4hr), 3),
    round(getJaccard(go_24hr), 3)
  ),
  fisher_pval = c(
    getPval(go_1hr),
    getPval(go_4hr),
    getPval(go_24hr)
  )
)

datatable(overlap_summary,
          options = list(pageLength = 5),
          caption = "Overlap between acute hypoxic response (3d-sensitive) and 3d promotes in hypoxia")
```


# 3e
## Extract gene sets from combined signatures file

```{r extract-acute-hypoxic-3e-sensitive}
# Extract genes upregulated in acute hypoxic response that are inhibited by 3e KD
acute_up_3e_sensitive_1hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd") %>%
  pull(db_gene_symbol)

acute_up_3e_sensitive_4hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd") %>%
  pull(db_gene_symbol)

acute_up_3e_sensitive_24hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd") %>%
  pull(db_gene_symbol)

# Extract genes promoted by 3e in hypoxia
hypoxia_3e_promotes_1hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3e_promotes_TE_1hr") %>%
  pull(db_gene_symbol)

hypoxia_3e_promotes_4hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3e_promotes_TE_4hr") %>%
  pull(db_gene_symbol)

hypoxia_3e_promotes_24hr_genes <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_3e_promotes_TE_24hr") %>%
  pull(db_gene_symbol)

# Print counts
cat("Genes upregulated in acute hypoxic response and sensitive to 3e KD:\n")
cat("  1hr:", length(acute_up_3e_sensitive_1hr), "\n")
cat("  4hr:", length(acute_up_3e_sensitive_4hr), "\n")
cat("  24hr:", length(acute_up_3e_sensitive_24hr), "\n\n")

cat("Genes promoted by 3e in hypoxia:\n")
cat("  1hr:", length(hypoxia_3e_promotes_1hr_genes), "\n")
cat("  4hr:", length(hypoxia_3e_promotes_4hr_genes), "\n")
cat("  24hr:", length(hypoxia_3e_promotes_24hr_genes), "\n")
```

## Comparison: Acute hypoxic response (3e-sensitive) vs 3e promotes in hypoxia

### 1 hour timepoint

```{r venn-3e-1hr}
# Create Venn diagram
venn_list_1hr <- list(
  "3e-sensitive in 1hr hypoxic response" = acute_up_3e_sensitive_1hr,
  "3e promotes TE in 1hr hypoxia" = hypoxia_3e_promotes_1hr_genes
)

p_venn_1hr <- ggvenn(
  venn_list_1hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_1hr


ggsave(here("plots/venn_acute_hypoxic_3e_sensitive_vs_3e_promotes_1hr.pdf"),
       p_venn_1hr, width = 6, height = 5)
```

```{r overlap-stats-3e-1hr}
# Calculate overlap statistics
overlap_1hr <- intersect(acute_up_3e_sensitive_1hr, hypoxia_3e_promotes_1hr_genes)
only_acute_1hr <- setdiff(acute_up_3e_sensitive_1hr, hypoxia_3e_promotes_1hr_genes)
only_promotes_1hr <- setdiff(hypoxia_3e_promotes_1hr_genes, acute_up_3e_sensitive_1hr)

cat("1 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_1hr), "\n")
cat("  Only in acute hypoxic (3e-sensitive):", length(only_acute_1hr), "\n")
cat("  Only in 3e promotes:", length(only_promotes_1hr), "\n")
cat("  % of acute hypoxic that are 3e-promoted:",
    round(100 * length(overlap_1hr) / length(acute_up_3e_sensitive_1hr), 1), "%\n")
cat("  % of 3e-promoted that are acute hypoxic:",
    round(100 * length(overlap_1hr) / length(hypoxia_3e_promotes_1hr_genes), 1), "%\n")

# GeneOverlap test
go_1hr <- newGeneOverlap(acute_up_3e_sensitive_1hr,
                         hypoxia_3e_promotes_1hr_genes,
                         genome.size = 20000)
go_1hr <- testGeneOverlap(go_1hr)
cat("  Fisher's exact test p-value:", format(getPval(go_1hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_1hr), 3), "\n")
```

### 4 hour timepoint

```{r venn-3e-4hr}
# Create Venn diagram
venn_list_4hr <- list(
  "3e-sensitive in 4hr hypoxic response" = acute_up_3e_sensitive_4hr,
  "3e promotes TE in 4hr hypoxia" = hypoxia_3e_promotes_4hr_genes
)

p_venn_4hr <- ggvenn(
  venn_list_4hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_4hr


ggsave(here("plots/venn_acute_hypoxic_3e_sensitive_vs_3e_promotes_4hr.pdf"),
       p_venn_4hr, width = 6, height = 5)
```

```{r overlap-stats-3e-4hr}
# Calculate overlap statistics
overlap_4hr <- intersect(acute_up_3e_sensitive_4hr, hypoxia_3e_promotes_4hr_genes)
only_acute_4hr <- setdiff(acute_up_3e_sensitive_4hr, hypoxia_3e_promotes_4hr_genes)
only_promotes_4hr <- setdiff(hypoxia_3e_promotes_4hr_genes, acute_up_3e_sensitive_4hr)

cat("4 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_4hr), "\n")
cat("  Only in acute hypoxic (3e-sensitive):", length(only_acute_4hr), "\n")
cat("  Only in 3e promotes:", length(only_promotes_4hr), "\n")
cat("  % of acute hypoxic that are 3e-promoted:",
    round(100 * length(overlap_4hr) / length(acute_up_3e_sensitive_4hr), 1), "%\n")
cat("  % of 3e-promoted that are acute hypoxic:",
    round(100 * length(overlap_4hr) / length(hypoxia_3e_promotes_4hr_genes), 1), "%\n")

# GeneOverlap test
go_4hr <- newGeneOverlap(acute_up_3e_sensitive_4hr,
                         hypoxia_3e_promotes_4hr_genes,
                         genome.size = 20000)
go_4hr <- testGeneOverlap(go_4hr)
cat("  Fisher's exact test p-value:", format(getPval(go_4hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_4hr), 3), "\n")
```

### 24 hour timepoint

```{r venn-3e-24hr}
# Create Venn diagram
venn_list_24hr <- list(
  "3e-sensitive in 24hr hypoxic response" = acute_up_3e_sensitive_24hr,
  "3e promotes TE in 24hr hypoxia" = hypoxia_3e_promotes_24hr_genes
)

p_venn_24hr <- ggvenn(
  venn_list_24hr,
  fill_color = c("#E76F51", "#2A9D8F"),
  stroke_size = 0.5,
  set_name_size = 4,
  text_size = 3
)

p_venn_24hr



ggsave(here("plots/venn_acute_hypoxic_3e_sensitive_vs_3e_promotes_24hr.pdf"),
       p_venn_24hr, width = 6, height = 5)
```

```{r overlap-stats-3e-24hr}
# Calculate overlap statistics
overlap_24hr <- intersect(acute_up_3e_sensitive_24hr, hypoxia_3e_promotes_24hr_genes)
only_acute_24hr <- setdiff(acute_up_3e_sensitive_24hr, hypoxia_3e_promotes_24hr_genes)
only_promotes_24hr <- setdiff(hypoxia_3e_promotes_24hr_genes, acute_up_3e_sensitive_24hr)

cat("4 hour overlap statistics:\n")
cat("  Overlap:", length(overlap_24hr), "\n")
cat("  Only in acute hypoxic (3e-sensitive):", length(only_acute_24hr), "\n")
cat("  Only in 3e promotes:", length(only_promotes_24hr), "\n")
cat("  % of acute hypoxic that are 3e-promoted:",
    round(100 * length(overlap_24hr) / length(acute_up_3e_sensitive_24hr), 1), "%\n")
cat("  % of 3e-promoted that are acute hypoxic:",
    round(100 * length(overlap_24hr) / length(hypoxia_3e_promotes_24hr_genes), 1), "%\n")

# GeneOverlap test
go_24hr <- newGeneOverlap(acute_up_3e_sensitive_24hr,
                         hypoxia_3e_promotes_24hr_genes,
                         genome.size = 20000)
go_24hr <- testGeneOverlap(go_24hr)
cat("  Fisher's exact test p-value:", format(getPval(go_24hr), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_24hr), 3), "\n")
```

# ECDF plots: Distribution of TE changes in 3d KD

## 1 hour timepoint - 3d KD in hypoxia

```{r read-3d-hypoxia-1hr-full-results}
# Read in full translation categories data for si3d vs sictrl in hypoxia at 1hr
te_3d_hypoxia_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))

# Check the data
head(te_3d_hypoxia_1hr)
cat("Total genes:", nrow(te_3d_hypoxia_1hr), "\n")
```

```{r prepare-gene-sets-for-ecdf-1hr}
# Extract genes that are up/down in acute hypoxic response and sensitive to 3d KD
# These gene lists are already extracted above as:
# - acute_up_3d_sensitive_1hr (genes upregulated in hypoxic response, inhibited by 3d KD)

# Also need genes downregulated in acute hypoxic response and sensitive to 3d KD
acute_down_3d_sensitive_1hr <- acute_sigs %>%
  filter(gs_name == "MDA-MB-231_hypoxia_response_decreases_TE_1hr_inhibited_by_3d") %>%
  pull(db_gene_symbol)

cat("Genes upregulated in acute hypoxic response (3d-sensitive):",
    length(acute_up_3d_sensitive_1hr), "\n")
cat("Genes downregulated in acute hypoxic response (3d-sensitive):",
    length(acute_down_3d_sensitive_1hr), "\n")

# Create a dataframe with gene categories for plotting
te_3d_hypoxia_1hr_plot <- te_3d_hypoxia_1hr %>%
  mutate(
    gene_category = case_when(
      symbol %in% acute_up_3d_sensitive_1hr ~ "Acute hypoxic up\n(3d-sensitive)",
      symbol %in% acute_down_3d_sensitive_1hr ~ "Acute hypoxic down\n(3d-sensitive)",
      TRUE ~ "Other genes"
    )
  ) %>%
  # Remove genes with NA TE LFC
  filter(!is.na(te_lfc))

# Count genes in each category
table(te_3d_hypoxia_1hr_plot$gene_category)
```

```{r ecdf-te-3d-hypoxia-1hr}
# Create ECDF plot for TE LFC distribution
p_ecdf_3d_1hr <- ggplot(te_3d_hypoxia_1hr_plot,
                         aes(x = te_lfc, color = gene_category)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c(
      "Acute hypoxic up\n(3d-sensitive)" = "#E76F51",
      "Acute hypoxic down\n(3d-sensitive)" = "#264653",
      "Other genes" = "grey70"
    ),
    name = "Gene category"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", alpha = 0.5) +
  labs(
    x = "TE log2 fold change (si3d vs siCtrl, 1hr hypoxia)",
    y = "Cumulative probability",
    title = "Distribution of TE changes upon 3d KD in 1hr hypoxia"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

p_ecdf_3d_1hr

ggsave(here("plots/ecdf_te_3d_hypoxia_1hr_acute_hypoxic_response.pdf"),
       p_ecdf_3d_1hr, width = 8, height = 5)
```

```{r ecdf-stats-3d-1hr}
# Statistical comparison using Kolmogorov-Smirnov test
# Compare acute hypoxic up (3d-sensitive) vs other genes
ks_up_vs_other <- ks.test(
  te_3d_hypoxia_1hr_plot %>%
    filter(gene_category == "Acute hypoxic up\n(3d-sensitive)") %>%
    pull(te_lfc),
  te_3d_hypoxia_1hr_plot %>%
    filter(gene_category == "Other genes") %>%
    pull(te_lfc)
)

# Compare acute hypoxic down (3d-sensitive) vs other genes
ks_down_vs_other <- ks.test(
  te_3d_hypoxia_1hr_plot %>%
    filter(gene_category == "Acute hypoxic down\n(3d-sensitive)") %>%
    pull(te_lfc),
  te_3d_hypoxia_1hr_plot %>%
    filter(gene_category == "Other genes") %>%
    pull(te_lfc)
)

cat("Kolmogorov-Smirnov test results:\n")
cat("\nAcute hypoxic UP (3d-sensitive) vs Other genes:\n")
cat("  D-statistic:", ks_up_vs_other$statistic, "\n")
cat("  p-value:", format(ks_up_vs_other$p.value, scientific = TRUE), "\n")

cat("\nAcute hypoxic DOWN (3d-sensitive) vs Other genes:\n")
cat("  D-statistic:", ks_down_vs_other$statistic, "\n")
cat("  p-value:", format(ks_down_vs_other$p.value, scientific = TRUE), "\n")

# Summary statistics for each category
summary_stats <- te_3d_hypoxia_1hr_plot %>%
  group_by(gene_category) %>%
  summarize(
    n = n(),
    mean_te_lfc = mean(te_lfc, na.rm = TRUE),
    median_te_lfc = median(te_lfc, na.rm = TRUE),
    sd_te_lfc = sd(te_lfc, na.rm = TRUE),
    .groups = "drop"
  )

datatable(summary_stats,
          caption = "Summary statistics of TE LFC by gene category (1hr, 3d KD in hypoxia)")
```
