---
title: "timecourse 231 acute hypoxic response"
author: "Kate"
date: "2025-11_30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(EnhancedVolcano)
library(msigdbr)
library(viridis)
library(gridExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(survival)
library(DESeq2)
library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(plotly)
library(ggokabeito)
library(ggvenn)
library(ggpubr)
library(DT)
library(edgeR)
library(cowplot)
library(ggExtra)
library(ggprism)
library(GeneOverlap)
library(writexl)
library(survminer)
library(readxl)
library(reshape2)
library(pheatmap)
library(ggridges)
```

```{r}
source(here("code/functions.R"))
cat("Available functions: createNormMatrix, deFunction\n")
```

# Load data from timecourse_qc analysis

```{r load_data}
# Load count matrices
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

cat("Loaded count matrices:\n")
cat("allCounts dimensions:", dim(allCounts), "\n")
cat("normMatrix dimensions:", dim(normMatrix), "\n")
cat("spikein dimensions:", dim(spikein), "\n")
cat("Number of spike-in genes:", length(top500_ids), "\n")
```

# Load DESeq results (with spike-in normalization)

```{r load_deseq_results}
# Load classified DESeq results for all timepoints
deseq_1hr <- read_csv(here("output", "deseq_results_1hr_spike.csv"))
deseq_4hr <- read_csv(here("output", "deseq_results_4hr_spike.csv"))
deseq_24hr <- read_csv(here("output", "deseq_results_24hr_spike.csv"))

cat("Loaded DESeq results:\n")
cat("1hr results:", nrow(deseq_1hr), "genes\n")
cat("4hr results:", nrow(deseq_4hr), "genes\n")
cat("24hr results:", nrow(deseq_24hr), "genes\n")

# Summary of significant genes per timepoint
cat("\n1hr significant genes:\n")
table(deseq_1hr$class)

cat("\n4hr significant genes:\n")
table(deseq_4hr$class)

cat("\n24hr significant genes:\n")
table(deseq_24hr$class)
```



# Which genes are hypoxia up one hour? which overlap in all three?


```{r}
unique(deseq_1hr$class)

up_1_hr <- deseq_1hr |> 
  filter(class== "Hypoxia up") |> pull(symbol)
up_4_hr <- deseq_4hr |> 
  filter(class== "Hypoxia up")|> pull(symbol)

up_24_hr <- deseq_24hr |>
  filter(class== "Hypoxia up")|> pull(symbol)

# Also get downregulated genes for sanity check
down_1_hr <- deseq_1hr |>
  filter(class== "Hypoxia down") |> pull(symbol)
down_4_hr <- deseq_4hr |>
  filter(class== "Hypoxia down")|> pull(symbol)
down_24_hr <- deseq_24hr |>
  filter(class== "Hypoxia down")|> pull(symbol)

```

## Analyze overlap of upregulated genes across timepoints

```{r}
# Calculate intersections
int_1_4 <- intersect(up_1_hr, up_4_hr)
int_1_24 <- intersect(up_1_hr, up_24_hr)
int_4_24 <- intersect(up_4_hr, up_24_hr)
int_all_three <- intersect(int_1_4, up_24_hr)

# Count genes at each timepoint
cat("Number of upregulated genes at each timepoint:\n")
cat("1hr:", length(up_1_hr), "genes\n")
cat("4hr:", length(up_4_hr), "genes\n")
cat("24hr:", length(up_24_hr), "genes\n\n")

# Count overlaps
cat("Overlap between timepoints:\n")
cat("1hr & 4hr:", length(int_1_4), "genes\n")
cat("1hr & 24hr:", length(int_1_24), "genes\n")
cat("4hr & 24hr:", length(int_4_24), "genes\n")
cat("All three timepoints:", length(int_all_three), "genes\n\n")

# Calculate unique genes per timepoint
unique_1hr <- setdiff(up_1_hr, union(up_4_hr, up_24_hr))
unique_4hr <- setdiff(up_4_hr, union(up_1_hr, up_24_hr))
unique_24hr <- setdiff(up_24_hr, union(up_1_hr, up_4_hr))

cat("Unique genes (not shared with other timepoints):\n")
cat("Only 1hr:", length(unique_1hr), "genes\n")
cat("Only 4hr:", length(unique_4hr), "genes\n")
cat("Only 24hr:", length(unique_24hr), "genes\n\n")

# What percentage of 1hr genes remain upregulated at later timepoints?
pct_1hr_in_4hr <- round(length(int_1_4) / length(up_1_hr) * 100, 1)
pct_1hr_in_24hr <- round(length(int_1_24) / length(up_1_hr) * 100, 1)
pct_1hr_in_all <- round(length(int_all_three) / length(up_1_hr) * 100, 1)

cat("Of the", length(up_1_hr), "genes upregulated at 1hr:\n")
cat(pct_1hr_in_4hr, "% are also upregulated at 4hr\n")
cat(pct_1hr_in_24hr, "% are also upregulated at 24hr\n")
cat(pct_1hr_in_all, "% are upregulated at all three timepoints\n")
```

## Visualize gene overlap with Venn diagram

```{r venn_diagram}
# Create list for Venn diagram
venn_list <- list(
  "1hr" = up_1_hr,
  "4hr" = up_4_hr,
  "24hr" = up_24_hr
)

# Create Venn diagram
venn_plot <- ggvenn(venn_list,
                    fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                    stroke_size = 0.5,
                    set_name_size = 4,
                    text_size = 3) +
  ggtitle("Upregulated genes across hypoxia timepoints (control cells)")

venn_plot

ggsave(here("plots/hypoxia_upregulated_venn.pdf"), venn_plot, width = 6, height = 5)
```

## Summary table of gene categories

```{r summary_table}
# Create comprehensive summary
summary_df <- data.frame(
  Category = c(
    "All three timepoints",
    "1hr & 4hr only",
    "1hr & 24hr only",
    "4hr & 24hr only",
    "1hr only",
    "4hr only",
    "24hr only"
  ),
  Count = c(
    length(int_all_three),
    length(setdiff(int_1_4, up_24_hr)),
    length(setdiff(int_1_24, up_4_hr)),
    length(setdiff(int_4_24, up_1_hr)),
    length(unique_1hr),
    length(unique_4hr),
    length(unique_24hr)
  )
)

summary_df <- summary_df %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

datatable(summary_df,
          caption = "Distribution of upregulated genes across timepoints",
          options = list(pageLength = 10))
```

## Export gene lists for each category

```{r export_gene_lists}
# Create gene lists with additional information from DESeq results
genes_all_three <- deseq_1hr %>%
  filter(symbol %in% int_all_three) %>%
  dplyr::select(gene_id, symbol) %>%
  left_join(deseq_1hr %>% dplyr::select(gene_id, padj_1hr = padj, log2FC_1hr = log2FoldChange), by = "gene_id") %>%
  left_join(deseq_4hr %>% dplyr::select(gene_id, padj_4hr = padj, log2FC_4hr = log2FoldChange), by = "gene_id") %>%
  left_join(deseq_24hr %>% dplyr::select(gene_id, padj_24hr = padj, log2FC_24hr = log2FoldChange), by = "gene_id") %>%
  arrange(desc(log2FC_1hr))

genes_1hr_only <- deseq_1hr %>%
  filter(symbol %in% unique_1hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

genes_4hr_only <- deseq_4hr %>%
  filter(symbol %in% unique_4hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

genes_24hr_only <- deseq_24hr %>%
  filter(symbol %in% unique_24hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

# Display genes upregulated at all three timepoints
cat("\nTop genes upregulated at all three timepoints:\n")
datatable(genes_all_three,
          caption = paste("Genes upregulated at 1hr, 4hr, and 24hr (n =", length(int_all_three), ")"),
          options = list(pageLength = 10))

# Export to Excel
gene_lists <- list(
  "All_three_timepoints" = genes_all_three,
  "1hr_only" = genes_1hr_only,
  "4hr_only" = genes_4hr_only,
  "24hr_only" = genes_24hr_only,
  "1hr_and_4hr" = deseq_1hr %>% filter(symbol %in% setdiff(int_1_4, up_24_hr)) %>%
    dplyr::select(gene_id, symbol, padj_1hr = padj, log2FC_1hr = log2FoldChange),
  "1hr_and_24hr" = deseq_1hr %>% filter(symbol %in% setdiff(int_1_24, up_4_hr)) %>%
    dplyr::select(gene_id, symbol, padj_1hr = padj, log2FC_1hr = log2FoldChange),
  "4hr_and_24hr" = deseq_4hr %>% filter(symbol %in% setdiff(int_4_24, up_1_hr)) %>%
    dplyr::select(gene_id, symbol, padj_4hr = padj, log2FC_4hr = log2FoldChange)
)

write_xlsx(gene_lists, here("output/hypoxia_upregulated_gene_lists.xlsx"))
cat("\nGene lists exported to: output/hypoxia_upregulated_gene_lists.xlsx\n")
```

# GSEA: Hypoxia Response Pathway Analysis

Perform Gene Set Enrichment Analysis (GSEA) on the ranked list of genes based on RNA log2 fold changes (hypoxia vs normoxia) at each timepoint, following the approach used in ribopipe-shiny.

## Load MSigDB gene sets

```{r load_msigdb}
# Get Hallmark gene sets from MSigDB
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark$gs_name)), "\n")
```

## GSEA: 1hr Hypoxia

```{r gsea_1hr}
# Create ranked list based on RNA log2FC (hypoxia vs normoxia)
ranked_list_1hr <- deseq_1hr$log2FoldChange
names(ranked_list_1hr) <- deseq_1hr$symbol

# Sort by LFC (descending) and remove duplicates
ranked_list_1hr <- sort(ranked_list_1hr, decreasing = TRUE)
ranked_list_1hr <- ranked_list_1hr[!duplicated(names(ranked_list_1hr))]

cat("1hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_1hr), "\n")
cat("LFC range:", round(min(ranked_list_1hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_1hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_1hr <- GSEA(
  geneList = ranked_list_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_1hr_results <- gsea_1hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

# Round numeric columns
gsea_1hr_results[] <- lapply(gsea_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 1hr - Top pathways:\n")
print(gsea_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot for significant pathways
p_gsea_1hr <- ggplot(gsea_1hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 1hr Hypoxia (Hallmark Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 9))

print(p_gsea_1hr)
ggsave(here("plots/gsea_hallmark_1hr.pdf"), p_gsea_1hr, width = 10, height = 8)
```

## GSEA: 4hr Hypoxia

```{r gsea_4hr}
# Create ranked list
ranked_list_4hr <- deseq_4hr$log2FoldChange
names(ranked_list_4hr) <- deseq_4hr$symbol
ranked_list_4hr <- sort(ranked_list_4hr, decreasing = TRUE)
ranked_list_4hr <- ranked_list_4hr[!duplicated(names(ranked_list_4hr))]

cat("4hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_4hr), "\n")
cat("LFC range:", round(min(ranked_list_4hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_4hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_4hr <- GSEA(
  geneList = ranked_list_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_4hr_results <- gsea_4hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_4hr_results[] <- lapply(gsea_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 4hr - Top pathways:\n")
print(gsea_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_4hr <- ggplot(gsea_4hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 4hr Hypoxia (Hallmark Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 9))

print(p_gsea_4hr)
ggsave(here("plots/gsea_hallmark_4hr.pdf"), p_gsea_4hr, width = 10, height = 8)
```

## GSEA: 24hr Hypoxia

```{r gsea_24hr}
# Create ranked list
ranked_list_24hr <- deseq_24hr$log2FoldChange
names(ranked_list_24hr) <- deseq_24hr$symbol
ranked_list_24hr <- sort(ranked_list_24hr, decreasing = TRUE)
ranked_list_24hr <- ranked_list_24hr[!duplicated(names(ranked_list_24hr))]

cat("24hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_24hr), "\n")
cat("LFC range:", round(min(ranked_list_24hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_24hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_24hr <- GSEA(
  geneList = ranked_list_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_24hr_results <- gsea_24hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_24hr_results[] <- lapply(gsea_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 24hr - Top pathways:\n")
print(gsea_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_24hr <- ggplot(gsea_24hr_results %>% filter(p.adjust < 0.05),
                       aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 24hr Hypoxia (Hallmark Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 9))

print(p_gsea_24hr)
ggsave(here("plots/gsea_hallmark_24hr.pdf"), p_gsea_24hr, width = 10, height = 8)
```

## Export GSEA results

```{r export_gsea}
# Compile all GSEA results (exclude leading_edge column for Excel export)
gsea_results_export <- list(
  "GSEA_1hr" = gsea_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_4hr" = gsea_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_24hr" = gsea_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_results_export, here("output/hypoxia_gsea_hallmark.xlsx"))
cat("\nGSEA results exported to: output/hypoxia_gsea_hallmark.xlsx\n")
```

## Additional Visualizations

### 1. Heatmap: Pathway NES across timepoints

Shows how pathway enrichment changes over the timecourse.

```{r gsea_heatmap}
# Combine results from all timepoints
gsea_combined <- bind_rows(
  gsea_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_pathways <- gsea_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
gsea_wide <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  dplyr::select(ID, NES, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
gsea_wide <- gsea_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_matrix <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match gsea_wide
sig_matrix <- sig_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches between gsea_wide and sig_matrix
sig_matrix <- sig_matrix[rownames(gsea_wide), ]

# Create heatmap
if (nrow(gsea_wide) > 0) {
  p_heatmap <- pheatmap(
    gsea_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    display_numbers = sig_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "GSEA NES Across Timepoints (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/gsea_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_heatmap)
  dev.off()

  cat("Heatmap shows", nrow(gsea_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for heatmap\n")
}
```

# ORA: Hallmark Pathway Enrichment for Upregulated Genes

Perform Over-Representation Analysis (ORA) to test if upregulated genes at each timepoint are enriched for Hallmark pathways.

## ORA: 1hr Upregulated Genes

```{r ora_hallmark_1hr}
# Run ORA on upregulated genes at 1hr
ora_1hr <- enricher(
  gene = up_1_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_1hr_results <- ora_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_1hr_results[] <- lapply(ora_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 1hr - Enriched pathways:\n")
print(ora_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_1hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_1hr <- ggplot(ora_1hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 1hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_1hr)
  ggsave(here("plots/ora_hallmark_1hr.pdf"), p_ora_1hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 1hr upregulated genes\n")
}
```

## ORA: 4hr Upregulated Genes

```{r ora_hallmark_4hr}
# Run ORA on upregulated genes at 4hr
ora_4hr <- enricher(
  gene = up_4_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_4hr_results <- ora_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_4hr_results[] <- lapply(ora_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 4hr - Enriched pathways:\n")
print(ora_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_4hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_4hr <- ggplot(ora_4hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 4hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_4hr)
  ggsave(here("plots/ora_hallmark_4hr.pdf"), p_ora_4hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 4hr upregulated genes\n")
}
```

## ORA: 24hr Upregulated Genes

```{r ora_hallmark_24hr}
# Run ORA on upregulated genes at 24hr
ora_24hr <- enricher(
  gene = up_24_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_24hr_results <- ora_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_24hr_results[] <- lapply(ora_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 24hr - Enriched pathways:\n")
print(ora_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_24hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_24hr <- ggplot(ora_24hr_results %>% filter(p.adjust < 0.05),
                         aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                             fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 24hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_24hr)
  ggsave(here("plots/ora_hallmark_24hr.pdf"), p_ora_24hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 24hr upregulated genes\n")
}
```

## Export ORA results

```{r export_ora}
# Compile all ORA results
ora_results_export <- list(
  "ORA_1hr" = ora_1hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_4hr" = ora_4hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_24hr" = ora_24hr_results %>% dplyr::select(-GeneRatio_numeric)
)

write_xlsx(ora_results_export, here("output/hypoxia_ora_hallmark.xlsx"))
cat("\nORA results exported to: output/hypoxia_ora_hallmark.xlsx\n")
```

## ORA Heatmap: Pathway enrichment across timepoints

```{r ora_heatmap}
# Combine results from all timepoints
ora_combined <- bind_rows(
  ora_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_ora_pathways <- ora_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
ora_wide <- ora_combined %>%
  filter(ID %in% sig_ora_pathways) %>%
  dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
ora_wide <- ora_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_ora_matrix <- ora_combined %>%
  filter(ID %in% sig_ora_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match ora_wide
sig_ora_matrix <- sig_ora_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_ora_matrix <- sig_ora_matrix[rownames(ora_wide), ]

# Create heatmap
if (nrow(ora_wide) > 0) {
  p_ora_heatmap <- pheatmap(
    ora_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "red"))(100),
    breaks = seq(0, max(ora_wide), length.out = 101),
    display_numbers = sig_ora_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA Gene Ratio Across Timepoints - Hallmark (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/ora_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_ora_heatmap)
  dev.off()

  cat("ORA heatmap shows", nrow(ora_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for ORA heatmap\n")
}
```

# ORA: Hallmark Pathway Enrichment for Downregulated Genes (Sanity Check)

Perform Over-Representation Analysis (ORA) on downregulated genes to check for expected pathway depletion patterns.

## ORA: 1hr Downregulated Genes

```{r ora_hallmark_down_1hr}
# Run ORA on downregulated genes at 1hr
ora_down_1hr <- enricher(
  gene = down_1_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_1hr_results <- ora_down_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_1hr_results[] <- lapply(ora_down_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 1hr Downregulated - Enriched pathways:\n")
print(ora_down_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_1hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_1hr <- ggplot(ora_down_1hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 1hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_1hr)
  ggsave(here("plots/ora_hallmark_down_1hr.pdf"), p_ora_down_1hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 1hr downregulated genes\n")
}
```

## ORA: 4hr Downregulated Genes

```{r ora_hallmark_down_4hr}
# Run ORA on downregulated genes at 4hr
ora_down_4hr <- enricher(
  gene = down_4_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_4hr_results <- ora_down_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_4hr_results[] <- lapply(ora_down_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 4hr Downregulated - Enriched pathways:\n")
print(ora_down_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_4hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_4hr <- ggplot(ora_down_4hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 4hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_4hr)
  ggsave(here("plots/ora_hallmark_down_4hr.pdf"), p_ora_down_4hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 4hr downregulated genes\n")
}
```

## ORA: 24hr Downregulated Genes

```{r ora_hallmark_down_24hr}
# Run ORA on downregulated genes at 24hr
ora_down_24hr <- enricher(
  gene = down_24_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_24hr_results <- ora_down_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_24hr_results[] <- lapply(ora_down_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 24hr Downregulated - Enriched pathways:\n")
print(ora_down_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_24hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_24hr <- ggplot(ora_down_24hr_results %>% filter(p.adjust < 0.05),
                         aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                             fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 24hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_24hr)
  ggsave(here("plots/ora_hallmark_down_24hr.pdf"), p_ora_down_24hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 24hr downregulated genes\n")
}
```

## Export ORA results (Downregulated)

```{r export_ora_down}
# Compile all ORA results for downregulated genes
ora_down_results_export <- list(
  "ORA_Down_1hr" = ora_down_1hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_Down_4hr" = ora_down_4hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_Down_24hr" = ora_down_24hr_results %>% dplyr::select(-GeneRatio_numeric)
)

write_xlsx(ora_down_results_export, here("output/hypoxia_ora_hallmark_downregulated.xlsx"))
cat("\nORA results for downregulated genes exported to: output/hypoxia_ora_hallmark_downregulated.xlsx\n")
```

## ORA Heatmap: Pathway enrichment across timepoints (Downregulated)

```{r ora_heatmap_down}
# Combine results from all timepoints
ora_down_combined <- bind_rows(
  ora_down_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_down_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_down_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_ora_down_pathways <- ora_down_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
ora_down_wide <- ora_down_combined %>%
  filter(ID %in% sig_ora_down_pathways) %>%
  dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
ora_down_wide <- ora_down_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_ora_down_matrix <- ora_down_combined %>%
  filter(ID %in% sig_ora_down_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match ora_down_wide
sig_ora_down_matrix <- sig_ora_down_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_ora_down_matrix <- sig_ora_down_matrix[rownames(ora_down_wide), ]

# Create heatmap
if (nrow(ora_down_wide) > 0) {
  p_ora_down_heatmap <- pheatmap(
    ora_down_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "blue"))(100),
    breaks = seq(0, max(ora_down_wide), length.out = 101),
    display_numbers = sig_ora_down_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA Gene Ratio Across Timepoints - Downregulated (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/ora_heatmap_downregulated_timecourse.pdf"), width = 8, height = 10)
  print(p_ora_down_heatmap)
  dev.off()

  cat("ORA heatmap for downregulated genes shows", nrow(ora_down_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for ORA heatmap (downregulated genes)\n")
}
```

# GSEA: KEGG Pathways

Perform the same GSEA analysis using KEGG pathways instead of Hallmark pathways.

## Load KEGG gene sets

```{r load_kegg}
# Get KEGG pathways from MSigDB
kegg_pathways_df <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB KEGG gene sets:\n")
cat("Number of pathways:", length(unique(kegg_pathways_df$gs_name)), "\n")
```

## GSEA: 1hr Hypoxia (KEGG)

```{r gsea_kegg_1hr}
# Run GSEA with KEGG pathways
set.seed(42)
gsea_kegg_1hr <- GSEA(
  geneList = ranked_list_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_1hr_results <- gsea_kegg_1hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

# Round numeric columns
gsea_kegg_1hr_results[] <- lapply(gsea_kegg_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 1hr - Top pathways:\n")
print(gsea_kegg_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot for significant pathways
p_gsea_kegg_1hr <- ggplot(gsea_kegg_1hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 1hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_1hr)
ggsave(here("plots/gsea_kegg_1hr.pdf"), p_gsea_kegg_1hr, width = 10, height = 8)
```

## GSEA: 4hr Hypoxia (KEGG)

```{r gsea_kegg_4hr}
# Run GSEA
set.seed(42)
gsea_kegg_4hr <- GSEA(
  geneList = ranked_list_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_4hr_results <- gsea_kegg_4hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_kegg_4hr_results[] <- lapply(gsea_kegg_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 4hr - Top pathways:\n")
print(gsea_kegg_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_kegg_4hr <- ggplot(gsea_kegg_4hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 4hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_4hr)
ggsave(here("plots/gsea_kegg_4hr.pdf"), p_gsea_kegg_4hr, width = 10, height = 8)
```

## GSEA: 24hr Hypoxia (KEGG)

```{r gsea_kegg_24hr}
# Run GSEA
set.seed(42)
gsea_kegg_24hr <- GSEA(
  geneList = ranked_list_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_24hr_results <- gsea_kegg_24hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_kegg_24hr_results[] <- lapply(gsea_kegg_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 24hr - Top pathways:\n")
print(gsea_kegg_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_kegg_24hr <- ggplot(gsea_kegg_24hr_results %>% filter(p.adjust < 0.05),
                       aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 24hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_24hr)
ggsave(here("plots/gsea_kegg_24hr.pdf"), p_gsea_kegg_24hr, width = 10, height = 8)
```

## Export KEGG GSEA results

```{r export_gsea_kegg}
# Compile all KEGG GSEA results
gsea_kegg_results_export <- list(
  "GSEA_KEGG_1hr" = gsea_kegg_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_KEGG_4hr" = gsea_kegg_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_KEGG_24hr" = gsea_kegg_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_kegg_results_export, here("output/hypoxia_gsea_kegg.xlsx"))
cat("\nKEGG GSEA results exported to: output/hypoxia_gsea_kegg.xlsx\n")
```

## KEGG Heatmap: Pathway NES across timepoints

```{r gsea_kegg_heatmap}
# Combine results from all timepoints
gsea_kegg_combined <- bind_rows(
  gsea_kegg_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_kegg_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_kegg_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_kegg_pathways <- gsea_kegg_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
gsea_kegg_wide <- gsea_kegg_combined %>%
  filter(ID %in% sig_kegg_pathways) %>%
  dplyr::select(ID, NES, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
gsea_kegg_wide <- gsea_kegg_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_kegg_matrix <- gsea_kegg_combined %>%
  filter(ID %in% sig_kegg_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match gsea_kegg_wide
sig_kegg_matrix <- sig_kegg_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_kegg_matrix <- sig_kegg_matrix[rownames(gsea_kegg_wide), ]

# Create heatmap
if (nrow(gsea_kegg_wide) > 0) {
  p_kegg_heatmap <- pheatmap(
    gsea_kegg_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    display_numbers = sig_kegg_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "GSEA NES Across Timepoints - KEGG (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/gsea_kegg_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_kegg_heatmap)
  dev.off()

  cat("KEGG heatmap shows", nrow(gsea_kegg_wide), "significant pathways\n")
} else {
  cat("No significant KEGG pathways found for heatmap\n")
}
```

# Comprehensive Analysis: Hypoxia Response with Knockdown Effects


## Load Translation Category Data

```{r load_translation_categories}
# Load hypoxia response categories
hyp_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
hyp_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
hyp_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load eIF3d knockdown categories
si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))

# Load eIF3e knockdown categories
si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"))
si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"))
si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"))
si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))

# Get gene symbols from deseq_1hr (already loaded)
gene_symbols <- deseq_1hr %>% dplyr::select(gene_id, symbol)

cat("Loaded all translation category data\n")
```

## Get TE-Regulated Genes from Hypoxia Response

```{r get_te_regulated_genes}
# Get genes with TE regulation (Exclusive, Buffered, TE_only) at any timepoint
te_regulated_genes <- bind_rows(
  hyp_1hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id),
  hyp_4hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id),
  hyp_24hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id)
)

# Get unique gene list
te_regulated_gene_ids <- unique(te_regulated_genes$gene_id)

cat("Total genes with TE regulation in hypoxia response:", length(te_regulated_gene_ids), "\n")
cat("Breakdown by timepoint:\n")
cat("  1hr:", sum(hyp_1hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
cat("  4hr:", sum(hyp_4hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
cat("  24hr:", sum(hyp_24hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
```

## Create Helper Function

```{r helper_function}
# Function to determine if TE is significantly changed
# TE is significant in categories: Exclusive, Buffered, TE_only
# TE is NOT significant in: Forwarded, Not_sig
is_te_not_significant <- function(category) {
  category %in% c("Forwarded", "Not_sig")
}
```

## Build Comprehensive Table

```{r build_table}
# Start with TE-regulated genes from hypoxia response
analysis_table <- data.frame(gene_id = te_regulated_gene_ids)

colnames(analysis_table)

# Add gene symbols
analysis_table <- analysis_table %>%
  left_join(gene_symbols, by = "gene_id")

# Add hypoxia response data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(hyp_1hr %>% 
            rename(hyp_1hr_rna_lfc = rna_lfc, hyp_1hr_ribo_lfc = ribo_lfc, hyp_1hr_te_lfc = te_lfc,
                   hyp_1hr_category = category,
                   te_padj_1hr = te_padj),
            by = "gene_id") %>%
  left_join(hyp_4hr %>%
            rename(hyp_4hr_rna_lfc = rna_lfc, hyp_4hr_ribo_lfc = ribo_lfc, hyp_4hr_te_lfc = te_lfc,
                    hyp_4hr_category = category,
                    te_padj_4hr = te_padj),
            by = "gene_id") %>%
  left_join(hyp_24hr %>%
            rename(hyp_24hr_rna_lfc = rna_lfc, hyp_24hr_ribo_lfc = ribo_lfc, hyp_24hr_te_lfc = te_lfc,
                 hyp_24hr_category = category,
                 te_padj_24hr = te_padj),
            by = "gene_id")

```

```{r add_si3d_data}
# Add eIF3d knockdown data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(si3d_norm_1hr %>% 
            rename(si3d_norm_1hr_rna_lfc = rna_lfc, si3d_norm_1hr_ribo_lfc = ribo_lfc, si3d_norm_1hr_te_lfc = te_lfc,
                   si3d_norm_1hr_category = category), by = "gene_id") %>%
  left_join(si3d_norm_4hr %>% 
            rename(si3d_norm_4hr_rna_lfc = rna_lfc, si3d_norm_4hr_ribo_lfc = ribo_lfc, si3d_norm_4hr_te_lfc = te_lfc,
                   si3d_norm_4hr_category = category), by = "gene_id") %>%
  left_join(si3d_norm_24hr %>% 
            rename(si3d_norm_24hr_rna_lfc = rna_lfc, si3d_norm_24hr_ribo_lfc = ribo_lfc, si3d_norm_24hr_te_lfc = te_lfc,
                   si3d_norm_24hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_1hr %>% 
            rename(si3d_hyp_1hr_rna_lfc = rna_lfc, si3d_hyp_1hr_ribo_lfc = ribo_lfc, si3d_hyp_1hr_te_lfc = te_lfc,
                   si3d_hyp_1hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_4hr %>% 
            rename(si3d_hyp_4hr_rna_lfc = rna_lfc, si3d_hyp_4hr_ribo_lfc = ribo_lfc, si3d_hyp_4hr_te_lfc = te_lfc,
                    si3d_hyp_4hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_24hr %>% 
            rename(si3d_hyp_24hr_rna_lfc = rna_lfc, si3d_hyp_24hr_ribo_lfc = ribo_lfc, si3d_hyp_24hr_te_lfc = te_lfc,
                   si3d_hyp_24hr_category = category), by = "gene_id")

```

```{r add_si3e_data}
# Add eIF3e knockdown data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(si3e_norm_1hr %>% 
            rename(si3e_norm_1hr_rna_lfc = rna_lfc, si3e_norm_1hr_ribo_lfc = ribo_lfc, si3e_norm_1hr_te_lfc = te_lfc,
                  si3e_norm_1hr_category = category,
                  si3e_norm_1hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_norm_4hr %>% 
            rename(si3e_norm_4hr_rna_lfc = rna_lfc, si3e_norm_4hr_ribo_lfc = ribo_lfc, si3e_norm_4hr_te_lfc = te_lfc,
                   si3e_norm_4hr_category = category,
                   si3e_norm_4hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_norm_24hr %>% 
            rename(si3e_norm_24hr_rna_lfc = rna_lfc, si3e_norm_24hr_ribo_lfc = ribo_lfc, si3e_norm_24hr_te_lfc = te_lfc,
                  si3e_norm_24hr_category = category,
                  si3e_norm_24hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_hyp_1hr %>% 
            rename(si3e_hyp_1hr_rna_lfc = rna_lfc, si3e_hyp_1hr_ribo_lfc = ribo_lfc, si3e_hyp_1hr_te_lfc = te_lfc,
                    si3e_hyp_1hr_category = category), by = "gene_id") %>%
  left_join(si3e_hyp_4hr %>% 
            rename(si3e_hyp_4hr_rna_lfc = rna_lfc, si3e_hyp_4hr_ribo_lfc = ribo_lfc, si3e_hyp_4hr_te_lfc = te_lfc,
                   si3e_hyp_4hr_category = category,
                   si3e_hyp_4hr_padj=te_padj), by = "gene_id") %>%
  left_join(si3e_hyp_24hr %>% 
            rename(si3e_hyp_24hr_rna_lfc = rna_lfc, si3e_hyp_24hr_ribo_lfc = ribo_lfc, si3e_hyp_24hr_te_lfc = te_lfc,
                   si3e_hyp_24hr_category = category,
                   si3e_hyp_24hr_padj=te_padj), by = "gene_id")

cat("Total columns in analysis table:", ncol(analysis_table), "\n")
```

## Add Interpretable Columns

```{r add_interpretable_columns}
# Add columns indicating TE regulation status
analysis_table <- analysis_table %>%
  mutate(
    # Hypoxia response TE direction
    hyp_1hr_te_dir = case_when(
      is.na(hyp_1hr_te_lfc) ~ "NA",
      hyp_1hr_te_lfc > 0 ~ "Up",
      hyp_1hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),
    hyp_4hr_te_dir = case_when(
      is.na(hyp_4hr_te_lfc) ~ "NA",
      hyp_4hr_te_lfc > 0 ~ "Up",
      hyp_4hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),
    hyp_24hr_te_dir = case_when(
      is.na(hyp_24hr_te_lfc) ~ "NA",
      hyp_24hr_te_lfc > 0 ~ "Up",
      hyp_24hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),

    # eIF3d knockdown - TE not significant?
    si3d_norm_1hr_te_not_sig = is_te_not_significant(si3d_norm_1hr_category),
    si3d_norm_4hr_te_not_sig = is_te_not_significant(si3d_norm_4hr_category),
    si3d_norm_24hr_te_not_sig = is_te_not_significant(si3d_norm_24hr_category),
    si3d_hyp_1hr_te_not_sig = is_te_not_significant(si3d_hyp_1hr_category),
    si3d_hyp_4hr_te_not_sig = is_te_not_significant(si3d_hyp_4hr_category),
    si3d_hyp_24hr_te_not_sig = is_te_not_significant(si3d_hyp_24hr_category),

    # eIF3e knockdown - TE not significant?
    si3e_norm_1hr_te_not_sig = is_te_not_significant(si3e_norm_1hr_category),
    si3e_norm_4hr_te_not_sig = is_te_not_significant(si3e_norm_4hr_category),
    si3e_norm_24hr_te_not_sig = is_te_not_significant(si3e_norm_24hr_category),
    si3e_hyp_1hr_te_not_sig = is_te_not_significant(si3e_hyp_1hr_category),
    si3e_hyp_4hr_te_not_sig = is_te_not_significant(si3e_hyp_4hr_category),
    si3e_hyp_24hr_te_not_sig = is_te_not_significant(si3e_hyp_24hr_category)
  )

cat("Added interpretable columns\n")
```


genes_of_interest


```{r}
genes_of_interest <- analysis_table |> 
  filter(hyp_1hr_category=="Exclusive" & hyp_1hr_te_lfc>0) |> 
  filter(hyp_4hr_te_lfc > 0) |> 
    filter(hyp_24hr_te_lfc > 0)


```



# Effect of knockdown on hypoxic response

Read in deseq results

```{r}
#sictrl
hyp_response_deseq_1hr <-read_csv(here("output", "translation_categories_1hr.csv")) |> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
  rename(sictrl_1hr_te_lfc = te_lfc, sictrl_1hr_te_padj = te_padj)
hyp_response_deseq_4hr<-read_csv(here("output", "translation_categories_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(sictrl_4hr_te_lfc = te_lfc, sictrl_4hr_te_padj = te_padj)

hyp_response_deseq_24hr<-read_csv(here("output", "translation_categories_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(sictrl_24hr_te_lfc = te_lfc, sictrl_24hr_te_padj = te_padj)


#3e
hyp_response_deseq_1hr_3e_kd <- read_csv(here("output", "translation_categories_si3e_1hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(si3e_1hr_te_lfc = te_lfc, si3e_1hr_te_padj = te_padj)

hyp_response_deseq_4hr_3e_kd<- read_csv(here("output", "translation_categories_si3e_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3e_4hr_te_lfc = te_lfc, si3e_4hr_te_padj = te_padj)

hyp_response_deseq_24hr_3e_kd<- read_csv(here("output", "translation_categories_si3e_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3e_24hr_te_lfc = te_lfc, si3e_24hr_te_padj = te_padj)


#3d
hyp_response_deseq_1hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_1hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3d_1hr_te_lfc = te_lfc, si3d_1hr_te_padj = te_padj)

hyp_response_deseq_4hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
        rename(si3d_4hr_te_lfc = te_lfc, si3d_4hr_te_padj = te_padj)

hyp_response_deseq_24hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
        rename(si3d_24hr_te_lfc = te_lfc, si3d_24hr_te_padj = te_padj)


all_hypoxic_response <-hyp_response_deseq_1hr |> 
  left_join(hyp_response_deseq_4hr) |> 
    left_join(hyp_response_deseq_24hr) |> 
    left_join(hyp_response_deseq_1hr_3e_kd) |> 
    left_join(hyp_response_deseq_4hr_3e_kd) |> 
    left_join(hyp_response_deseq_24hr_3e_kd) |> 
    left_join(hyp_response_deseq_1hr_3d_kd) |> 
    left_join(hyp_response_deseq_4hr_3d_kd) |> 
    left_join(hyp_response_deseq_24hr_3d_kd)

  

head(all_hypoxic_response)
```




# 1 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_1hr <- all_hypoxic_response %>%
  filter(sictrl_1hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_1hr %>%
  filter(si3e_1hr_te_lfc < 0) %>%
  filter(si3e_1hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_1hr %>%
  filter(si3e_1hr_te_lfc > 0) %>%
  filter(si3e_1hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_1hr_plot <- ggplot(data = sig_sictrl_1hr, aes(x = sictrl_1hr_te_lfc, y = si3e_1hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_1hr_plot <- ggMarginal(scatter_3e_1hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_1hr <- cowplot::plot_grid(scatter_3e_1hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_1hr

ggsave(filename= here("plots/si3e_1hr_response.pdf"), scatter_3e_1hr, height=4, width=4)



```




# 1 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_1hr <- all_hypoxic_response %>%
  filter(sictrl_1hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_1hr %>%
  filter(si3d_1hr_te_lfc < 0) %>%
  filter(si3d_1hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_1hr %>%
  filter(si3d_1hr_te_lfc > 0) %>%
  filter(si3d_1hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_1hr_plot <- ggplot(data = sig_sictrl_1hr, aes(x = sictrl_1hr_te_lfc, y = si3d_1hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_1hr_plot <- ggMarginal(scatter_3d_1hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_1hr <- cowplot::plot_grid(scatter_3d_1hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_1hr

ggsave(filename= here("plots/si3d_1hr_response.pdf"), scatter_3d_1hr, height=4, width=4)



```


# 4 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_4hr <- all_hypoxic_response %>%
  filter(sictrl_4hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_4hr %>%
  filter(si3e_4hr_te_lfc < 0) %>%
  filter(si3e_4hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_4hr %>%
  filter(si3e_4hr_te_lfc > 0) %>%
  filter(si3e_4hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_4hr_plot <- ggplot(data = sig_sictrl_4hr, aes(x = sictrl_4hr_te_lfc, y = si3e_4hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_4hr_plot <- ggMarginal(scatter_3e_4hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_4hr <- cowplot::plot_grid(scatter_3e_4hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_4hr

ggsave(filename= here("plots/si3e_4hr_response.pdf"), scatter_3e_4hr, height=4, width=4)



```




# 4 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_4hr <- all_hypoxic_response %>%
  filter(sictrl_4hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_4hr %>%
  filter(si3d_4hr_te_lfc < 0) %>%
  filter(si3d_4hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_4hr %>%
  filter(si3d_4hr_te_lfc > 0) %>%
  filter(si3d_4hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_4hr_plot <- ggplot(data = sig_sictrl_4hr, aes(x = sictrl_4hr_te_lfc, y = si3d_4hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_4hr_plot <- ggMarginal(scatter_3d_4hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_4hr <- cowplot::plot_grid(scatter_3d_4hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_4hr

ggsave(filename= here("plots/si3d_4hr_response.pdf"), scatter_3d_4hr, height=4, width=4)



```


# 24 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_24hr <- all_hypoxic_response %>%
  filter(sictrl_24hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_24hr %>%
  filter(si3e_24hr_te_lfc < 0) %>%
  filter(si3e_24hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_24hr %>%
  filter(si3e_24hr_te_lfc > 0) %>%
  filter(si3e_24hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_24hr_plot <- ggplot(data = sig_sictrl_24hr, aes(x = sictrl_24hr_te_lfc, y = si3e_24hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_24hr_plot <- ggMarginal(scatter_3e_24hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_24hr <- cowplot::plot_grid(scatter_3e_24hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_24hr

ggsave(filename= here("plots/si3e_24hr_response.pdf"), scatter_3e_24hr, height=4, width=4)



```




# 24 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_24hr <- all_hypoxic_response %>%
  filter(sictrl_24hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_24hr %>%
  filter(si3d_24hr_te_lfc < 0) %>%
  filter(si3d_24hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_24hr %>%
  filter(si3d_24hr_te_lfc > 0) %>%
  filter(si3d_24hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_24hr_plot <- ggplot(data = sig_sictrl_24hr, aes(x = sictrl_24hr_te_lfc, y = si3d_24hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_24hr_plot <- ggMarginal(scatter_3d_24hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_24hr <- cowplot::plot_grid(scatter_3d_24hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_24hr

ggsave(filename= here("plots/si3d_24hr_response.pdf"), scatter_3d_24hr, height=4, width=4)



```
