---
title: "timecourse 231 acute hypoxic response"
author: "Kate"
date: "2025-11_30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(EnhancedVolcano)
library(msigdbr)
library(viridis)
library(gridExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(survival)
library(DESeq2)
library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(plotly)
library(ggokabeito)
library(ggvenn)
library(ggpubr)
library(DT)
library(edgeR)
library(cowplot)
library(ggExtra)
library(ggprism)
library(GeneOverlap)
library(writexl)
library(survminer)
library(readxl)
library(reshape2)
library(pheatmap)
library(ggridges)
```

```{r}
source(here("code/functions.R"))
cat("Available functions: createNormMatrix, deFunction\n")
```

# Load data from timecourse_qc analysis

```{r load_data}
# Load count matrices
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

cat("Loaded count matrices:\n")
cat("allCounts dimensions:", dim(allCounts), "\n")
cat("normMatrix dimensions:", dim(normMatrix), "\n")
cat("spikein dimensions:", dim(spikein), "\n")
cat("Number of spike-in genes:", length(top500_ids), "\n")
```

# Load DESeq results (with spike-in normalization)

```{r load_deseq_results}
# Load classified DESeq results for all timepoints
deseq_1hr <- read_csv(here("output", "deseq_results_1hr_spike.csv"))
deseq_4hr <- read_csv(here("output", "deseq_results_4hr_spike.csv"))
deseq_24hr <- read_csv(here("output", "deseq_results_24hr_spike.csv"))

cat("Loaded DESeq results:\n")
cat("1hr results:", nrow(deseq_1hr), "genes\n")
cat("4hr results:", nrow(deseq_4hr), "genes\n")
cat("24hr results:", nrow(deseq_24hr), "genes\n")

# Summary of significant genes per timepoint
cat("\n1hr significant genes:\n")
table(deseq_1hr$class)

cat("\n4hr significant genes:\n")
table(deseq_4hr$class)

cat("\n24hr significant genes:\n")
table(deseq_24hr$class)
```


# Distributions

## RNA log2FoldChange distribution across timepoints

```{r rna_lfc_distribution_across_timepoints, fig.width=8, fig.height=6}
# Prepare data for density plot
# Combine RNA log2FoldChange from all three timepoints into long format
rna_lfc_data <- bind_rows(
  read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, rna_lfc) %>%
    mutate(Timepoint = "1hr"),
  read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, rna_lfc) %>%
    mutate(Timepoint = "4hr"),
  read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, rna_lfc) %>%
    mutate(Timepoint = "24hr")
) %>%
  # Factor timepoint for proper ordering in legend
  mutate(Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")))

# Summary statistics for each timepoint
cat("Summary statistics for RNA log2FoldChange distributions:\n\n")
rna_lfc_data %>%
  group_by(Timepoint) %>%
  summarise(
    n_genes = n(),
    mean_lfc = mean(rna_lfc, na.rm = TRUE),
    median_lfc = median(rna_lfc, na.rm = TRUE),
    sd_lfc = sd(rna_lfc, na.rm = TRUE)
  ) %>%
  print()

hr_24 <- rna_lfc_data |> 
  filter(Timepoint=="24hr")
hist(hr_24$rna_lfc)



# Create density plot with three overlaid curves
density_plot <- ggplot(rna_lfc_data, aes(x = rna_lfc, color = Timepoint)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "RNA Expression Changes Across Hypoxia Timecourse",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (RNA)",
    y = "Density",
    color = "Timepoint"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-10, 10)) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold")
  )



# Display the plot
print(density_plot)

# Save the plot
ggsave(
  here("output", "rna_lfc_density_timecourse.pdf"),
  plot = density_plot,
  width = 5,
  height = 4
)

cat("\nDensity plot created and saved to output/rna_lfc_density_timecourse.pdf\n")
```

## RNA log2FoldChange distribution - Faceted by timepoint

```{r rna_lfc_distribution_faceted, fig.width=10, fig.height=4}
# Create faceted density plot for RNA - one panel per timepoint
density_plot_rna_faceted <- ggplot(rna_lfc_data, aes(x = rna_lfc, fill = Timepoint, color = Timepoint)) +
  geom_density(alpha = 0.5, size = 1.2) +
  scale_fill_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "RNA Expression Changes by Timepoint",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (RNA)",
    y = "Density"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~Timepoint, ncol = 3, scales = "free") +
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

# Display the plot
print(density_plot_rna_faceted)

# Save the plot
ggsave(
  here("output", "rna_lfc_density_faceted.pdf"),
  plot = density_plot_rna_faceted,
  width = 10,
  height = 4
)

cat("\nFaceted RNA density plot created and saved to output/rna_lfc_density_faceted.pdf\n")
```

## Ribo log2FoldChange distribution across timepoints

```{r ribo_lfc_distribution_across_timepoints, fig.width=8, fig.height=6}
# Load translation categories data to get ribo_lfc
ribo_lfc_data <- bind_rows(
  read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, ribo_lfc) %>%
    mutate(Timepoint = "1hr"),
  read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, ribo_lfc) %>%
    mutate(Timepoint = "4hr"),
  read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, ribo_lfc) %>%
    mutate(Timepoint = "24hr")
) %>%
  # Factor timepoint for proper ordering in legend
  mutate(Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")))

# Summary statistics for each timepoint
cat("Summary statistics for Ribo log2FoldChange distributions:\n\n")
ribo_lfc_data %>%
  group_by(Timepoint) %>%
  summarise(
    n_genes = n(),
    mean_lfc = mean(ribo_lfc, na.rm = TRUE),
    median_lfc = median(ribo_lfc, na.rm = TRUE),
    sd_lfc = sd(ribo_lfc, na.rm = TRUE)
  ) %>%
  print()

# Create density plot with three overlaid curves
density_plot_ribo <- ggplot(ribo_lfc_data, aes(x = ribo_lfc, color = Timepoint)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Ribosome Occupancy Changes Across Hypoxia Timecourse",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (Ribo)",
    y = "Density",
    color = "Timepoint"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-10, 10)) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold")
  )

# Display the plot
print(density_plot_ribo)

# Save the plot
ggsave(
  here("output", "ribo_lfc_density_timecourse.pdf"),
  plot = density_plot_ribo,
  width = 5,
  height = 4
)

cat("\nDensity plot created and saved to output/ribo_lfc_density_timecourse.pdf\n")
```

## Ribo log2FoldChange distribution - Faceted by timepoint

```{r ribo_lfc_distribution_faceted, fig.width=10, fig.height=4}
# Create faceted density plot for Ribo - one panel per timepoint
density_plot_ribo_faceted <- ggplot(ribo_lfc_data, aes(x = ribo_lfc, fill = Timepoint, color = Timepoint)) +
  geom_density(alpha = 0.5, size = 1.2) +
  scale_fill_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Ribosome Occupancy Changes by Timepoint",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (Ribo)",
    y = "Density"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~Timepoint, ncol = 3, scales = "free") +
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

# Display the plot
print(density_plot_ribo_faceted)

# Save the plot
ggsave(
  here("output", "ribo_lfc_density_faceted.pdf"),
  plot = density_plot_ribo_faceted,
  width = 10,
  height = 4
)

cat("\nFaceted Ribo density plot created and saved to output/ribo_lfc_density_faceted.pdf\n")
```

## TE log2FoldChange distribution across timepoints

```{r te_lfc_distribution_across_timepoints, fig.width=8, fig.height=6}
# Load translation categories data to get te_lfc
te_lfc_data <- bind_rows(
  read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, te_lfc) %>%
    mutate(Timepoint = "1hr"),
  read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, te_lfc) %>%
    mutate(Timepoint = "4hr"),
  read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE) %>%
    select(gene_id, te_lfc) %>%
    mutate(Timepoint = "24hr")
) %>%
  # Factor timepoint for proper ordering in legend
  mutate(Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")))

# Summary statistics for each timepoint
cat("Summary statistics for TE log2FoldChange distributions:\n\n")
te_lfc_data %>%
  group_by(Timepoint) %>%
  summarise(
    n_genes = n(),
    mean_lfc = mean(te_lfc, na.rm = TRUE),
    median_lfc = median(te_lfc, na.rm = TRUE),
    sd_lfc = sd(te_lfc, na.rm = TRUE)
  ) %>%
  print()

# Create density plot with three overlaid curves
density_plot_te <- ggplot(te_lfc_data, aes(x = te_lfc, color = Timepoint)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Translation Efficiency Changes Across Hypoxia Timecourse",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (TE)",
    y = "Density",
    color = "Timepoint"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-10, 10)) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold")
  )

# Display the plot
print(density_plot_te)

# Save the plot
ggsave(
  here("output", "te_lfc_density_timecourse.pdf"),
  plot = density_plot_te,
  width = 5,
  height = 4
)

cat("\nDensity plot created and saved to output/te_lfc_density_timecourse.pdf\n")
```

## TE log2FoldChange distribution - Faceted by timepoint

```{r te_lfc_distribution_faceted, fig.width=10, fig.height=4}
# Create faceted density plot for TE - one panel per timepoint
density_plot_te_faceted <- ggplot(te_lfc_data, aes(x = te_lfc, fill = Timepoint, color = Timepoint)) +
  geom_density(alpha = 0.5, size = 1.2) +
  scale_fill_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Translation Efficiency Changes by Timepoint",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (TE)",
    y = "Density"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~Timepoint, ncol = 3, scales = "free") +
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

# Display the plot
print(density_plot_te_faceted)

# Save the plot
ggsave(
  here("output", "te_lfc_density_faceted.pdf"),
  plot = density_plot_te_faceted,
  width = 10,
  height = 4
)

cat("\nFaceted TE density plot created and saved to output/te_lfc_density_faceted.pdf\n")
```

## Combined RNA, Ribo, and TE distributions (faceted)

```{r combined_lfc_distribution_faceted, fig.width=8, fig.height=10}
# Combine all three data types into one dataframe
combined_lfc_data <- bind_rows(
  rna_lfc_data %>%
    rename(log2FC = rna_lfc) %>%
    mutate(Data_Type = "RNA"),
  ribo_lfc_data %>%
    rename(log2FC = ribo_lfc) %>%
    mutate(Data_Type = "Ribo"),
  te_lfc_data %>%
    rename(log2FC = te_lfc) %>%
    mutate(Data_Type = "TE")
) %>%
  # Factor Data_Type for proper ordering in facets
  mutate(Data_Type = factor(Data_Type, levels = c("RNA", "Ribo", "TE")))

# Create faceted density plot
combined_density_plot <- ggplot(combined_lfc_data, aes(x = log2FC, color = Timepoint)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Distribution of Changes Across Hypoxia Timecourse",
    subtitle = "RNA, Ribosome Occupancy, and Translation Efficiency",
    x = "log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density",
    color = "Timepoint"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-10, 10), ylim = c(0, 3)) +
  facet_grid(Data_Type ~ ., scales = "free_y") +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

# Display the plot
print(combined_density_plot)

# Save the plot
ggsave(
  here("output", "combined_lfc_density_timecourse_faceted.pdf"),
  plot = combined_density_plot,
  width = 3,
  height = 5
)

cat("\nCombined faceted density plot created and saved to output/combined_lfc_density_timecourse_faceted.pdf\n")
```

## Combined RNA, Ribo, and TE distributions (faceted, zoomed)

```{r combined_lfc_distribution_faceted_zoomed, fig.width=8, fig.height=10}
# Create zoomed faceted density plot with narrower x-axis
combined_density_plot_zoomed <- ggplot(combined_lfc_data, aes(x = log2FC, color = Timepoint)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "1hr" = "#E69F00",   # Orange - acute response
      "4hr" = "#CC79A7",   # Mauve/pink - intermediate
      "24hr" = "#56B4E9"   # Light blue - sustained response
    )
  ) +
  labs(
    title = "Distribution of Changes Across Hypoxia Timecourse (Zoomed)",
    subtitle = "RNA, Ribosome Occupancy, and Translation Efficiency",
    x = "log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density",
    color = "Timepoint"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(0, 3)) +
  facet_grid(Data_Type ~ ., scales = "free_y") +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

# Display the plot
print(combined_density_plot_zoomed)

# Save the plot
ggsave(
  here("output", "combined_lfc_density_timecourse_faceted_zoomed.pdf"),
  plot = combined_density_plot_zoomed,
  width = 3,
  height = 5
)

cat("\nCombined faceted density plot (zoomed) created and saved to output/combined_lfc_density_timecourse_faceted_zoomed.pdf\n")
```







# Which genes are hypoxia up one hour? which overlap in all three?


```{r}
unique(deseq_1hr$class)

up_1_hr <- deseq_1hr |> 
  filter(class== "Hypoxia up") |> pull(symbol)
up_4_hr <- deseq_4hr |> 
  filter(class== "Hypoxia up")|> pull(symbol)

up_24_hr <- deseq_24hr |>
  filter(class== "Hypoxia up")|> pull(symbol)

# Also get downregulated genes for sanity check
down_1_hr <- deseq_1hr |>
  filter(class== "Hypoxia down") |> pull(symbol)
down_4_hr <- deseq_4hr |>
  filter(class== "Hypoxia down")|> pull(symbol)
down_24_hr <- deseq_24hr |>
  filter(class== "Hypoxia down")|> pull(symbol)

```

## Analyze overlap of upregulated genes across timepoints

```{r}
# Calculate intersections
int_1_4 <- intersect(up_1_hr, up_4_hr)
int_1_24 <- intersect(up_1_hr, up_24_hr)
int_4_24 <- intersect(up_4_hr, up_24_hr)
int_all_three <- intersect(int_1_4, up_24_hr)

# Count genes at each timepoint
cat("Number of upregulated genes at each timepoint:\n")
cat("1hr:", length(up_1_hr), "genes\n")
cat("4hr:", length(up_4_hr), "genes\n")
cat("24hr:", length(up_24_hr), "genes\n\n")

# Count overlaps
cat("Overlap between timepoints:\n")
cat("1hr & 4hr:", length(int_1_4), "genes\n")
cat("1hr & 24hr:", length(int_1_24), "genes\n")
cat("4hr & 24hr:", length(int_4_24), "genes\n")
cat("All three timepoints:", length(int_all_three), "genes\n\n")

# Calculate unique genes per timepoint
unique_1hr <- setdiff(up_1_hr, union(up_4_hr, up_24_hr))
unique_4hr <- setdiff(up_4_hr, union(up_1_hr, up_24_hr))
unique_24hr <- setdiff(up_24_hr, union(up_1_hr, up_4_hr))

cat("Unique genes (not shared with other timepoints):\n")
cat("Only 1hr:", length(unique_1hr), "genes\n")
cat("Only 4hr:", length(unique_4hr), "genes\n")
cat("Only 24hr:", length(unique_24hr), "genes\n\n")

# What percentage of 1hr genes remain upregulated at later timepoints?
pct_1hr_in_4hr <- round(length(int_1_4) / length(up_1_hr) * 100, 1)
pct_1hr_in_24hr <- round(length(int_1_24) / length(up_1_hr) * 100, 1)
pct_1hr_in_all <- round(length(int_all_three) / length(up_1_hr) * 100, 1)

cat("Of the", length(up_1_hr), "genes upregulated at 1hr:\n")
cat(pct_1hr_in_4hr, "% are also upregulated at 4hr\n")
cat(pct_1hr_in_24hr, "% are also upregulated at 24hr\n")
cat(pct_1hr_in_all, "% are upregulated at all three timepoints\n")
```

## Visualize gene overlap with Venn diagram

```{r venn_diagram}
# Create list for Venn diagram
venn_list <- list(
  "1hr" = up_1_hr,
  "4hr" = up_4_hr,
  "24hr" = up_24_hr
)

# Create Venn diagram
venn_plot <- ggvenn(venn_list,
                    fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                    stroke_size = 0.5,
                    set_name_size = 4,
                    text_size = 3) +
  ggtitle("Upregulated genes across hypoxia timepoints (control cells)")

venn_plot

ggsave(here("plots/hypoxia_upregulated_venn.pdf"), venn_plot, width = 6, height = 5)
```

## Summary table of gene categories

```{r summary_table}
# Create comprehensive summary
summary_df <- data.frame(
  Category = c(
    "All three timepoints",
    "1hr & 4hr only",
    "1hr & 24hr only",
    "4hr & 24hr only",
    "1hr only",
    "4hr only",
    "24hr only"
  ),
  Count = c(
    length(int_all_three),
    length(setdiff(int_1_4, up_24_hr)),
    length(setdiff(int_1_24, up_4_hr)),
    length(setdiff(int_4_24, up_1_hr)),
    length(unique_1hr),
    length(unique_4hr),
    length(unique_24hr)
  )
)

summary_df <- summary_df %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

datatable(summary_df,
          caption = "Distribution of upregulated genes across timepoints",
          options = list(pageLength = 10))
```

## Export gene lists for each category

```{r export_gene_lists}
# Create gene lists with additional information from DESeq results
genes_all_three <- deseq_1hr %>%
  filter(symbol %in% int_all_three) %>%
  dplyr::select(gene_id, symbol) %>%
  left_join(deseq_1hr %>% dplyr::select(gene_id, padj_1hr = padj, log2FC_1hr = log2FoldChange), by = "gene_id") %>%
  left_join(deseq_4hr %>% dplyr::select(gene_id, padj_4hr = padj, log2FC_4hr = log2FoldChange), by = "gene_id") %>%
  left_join(deseq_24hr %>% dplyr::select(gene_id, padj_24hr = padj, log2FC_24hr = log2FoldChange), by = "gene_id") %>%
  arrange(desc(log2FC_1hr))

genes_1hr_only <- deseq_1hr %>%
  filter(symbol %in% unique_1hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

genes_4hr_only <- deseq_4hr %>%
  filter(symbol %in% unique_4hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

genes_24hr_only <- deseq_24hr %>%
  filter(symbol %in% unique_24hr) %>%
  dplyr::select(gene_id, symbol, padj, log2FoldChange) %>%
  arrange(desc(log2FoldChange))

# Display genes upregulated at all three timepoints
cat("\nTop genes upregulated at all three timepoints:\n")
datatable(genes_all_three,
          caption = paste("Genes upregulated at 1hr, 4hr, and 24hr (n =", length(int_all_three), ")"),
          options = list(pageLength = 10))

# Export to Excel
gene_lists <- list(
  "All_three_timepoints" = genes_all_three,
  "1hr_only" = genes_1hr_only,
  "4hr_only" = genes_4hr_only,
  "24hr_only" = genes_24hr_only,
  "1hr_and_4hr" = deseq_1hr %>% filter(symbol %in% setdiff(int_1_4, up_24_hr)) %>%
    dplyr::select(gene_id, symbol, padj_1hr = padj, log2FC_1hr = log2FoldChange),
  "1hr_and_24hr" = deseq_1hr %>% filter(symbol %in% setdiff(int_1_24, up_4_hr)) %>%
    dplyr::select(gene_id, symbol, padj_1hr = padj, log2FC_1hr = log2FoldChange),
  "4hr_and_24hr" = deseq_4hr %>% filter(symbol %in% setdiff(int_4_24, up_1_hr)) %>%
    dplyr::select(gene_id, symbol, padj_4hr = padj, log2FC_4hr = log2FoldChange)
)

write_xlsx(gene_lists, here("output/hypoxia_upregulated_gene_lists.xlsx"))
cat("\nGene lists exported to: output/hypoxia_upregulated_gene_lists.xlsx\n")
```

# GSEA: RNA-seq Changes (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (hypoxia vs normoxia) at each timepoint in control cells.

## Load RNA differential expression results

```{r load_rna_results}
# Load RNA results from translation analysis (already computed in rna_ribo_te_analysis.Rmd)
rna_results_1hr <- read_csv(here("output", "translation_categories_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_4hr <- read_csv(here("output", "translation_categories_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_24hr <- read_csv(here("output", "translation_categories_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

# Add gene symbols
gene_symbols <- deseq_1hr %>% dplyr::select(gene_id, symbol)

res_rna_1hr <- rna_results_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_4hr <- rna_results_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_24hr <- rna_results_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

cat("RNA results loaded from translation_categories files:\n")
cat("1hr:", nrow(res_rna_1hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_1hr$padj < 0.05, na.rm = TRUE), "\n")
cat("4hr:", nrow(res_rna_4hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_4hr$padj < 0.05, na.rm = TRUE), "\n")
cat("24hr:", nrow(res_rna_24hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Load Hallmark gene sets

```{r load_hallmark_rna}
# Get Hallmark gene sets from MSigDB
hallmark_rna <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark_rna$gs_name)), "\n")
```

## GSEA: RNA 1hr

```{r gsea_rna_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list_rna_1hr <- res_rna_1hr$log2FoldChange
names(ranked_list_rna_1hr) <- res_rna_1hr$symbol

# Sort and remove duplicates
ranked_list_rna_1hr <- sort(ranked_list_rna_1hr, decreasing = TRUE)
ranked_list_rna_1hr <- ranked_list_rna_1hr[!duplicated(names(ranked_list_rna_1hr))]

cat("1hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_1hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_1hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_1hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_1hr <- GSEA(
  geneList = ranked_list_rna_1hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_1hr@result) > 0) {
  gsea_rna_1hr_results <- gsea_rna_1hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_1hr_results[] <- lapply(gsea_rna_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA 1hr - Top pathways:\n")
  print(gsea_rna_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  # Create empty results dataframe with expected structure
  gsea_rna_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA 1hr - No pathways enriched\n")
}
```

## GSEA: RNA 4hr

```{r gsea_rna_4hr}
# Create ranked list (using all genes)
ranked_list_rna_4hr <- res_rna_4hr$log2FoldChange
names(ranked_list_rna_4hr) <- res_rna_4hr$symbol
ranked_list_rna_4hr <- sort(ranked_list_rna_4hr, decreasing = TRUE)
ranked_list_rna_4hr <- ranked_list_rna_4hr[!duplicated(names(ranked_list_rna_4hr))]

cat("4hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_4hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_4hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_4hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_4hr <- GSEA(
  geneList = ranked_list_rna_4hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_4hr@result) > 0) {
  gsea_rna_4hr_results <- gsea_rna_4hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_4hr_results[] <- lapply(gsea_rna_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA 4hr - Top pathways:\n")
  print(gsea_rna_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA 4hr - No pathways enriched\n")
}
```

## GSEA: RNA 24hr

```{r gsea_rna_24hr}
# Create ranked list (using all genes)
ranked_list_rna_24hr <- res_rna_24hr$log2FoldChange
names(ranked_list_rna_24hr) <- res_rna_24hr$symbol
ranked_list_rna_24hr <- sort(ranked_list_rna_24hr, decreasing = TRUE)
ranked_list_rna_24hr <- ranked_list_rna_24hr[!duplicated(names(ranked_list_rna_24hr))]

cat("24hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_24hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_24hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_24hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_24hr <- GSEA(
  geneList = ranked_list_rna_24hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_24hr@result) > 0) {
  gsea_rna_24hr_results <- gsea_rna_24hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_24hr_results[] <- lapply(gsea_rna_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA 24hr - Top pathways:\n")
  print(gsea_rna_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA 24hr - No pathways enriched\n")
}
```

## Export RNA GSEA results

```{r export_gsea_rna}
# Compile all RNA GSEA results
gsea_rna_results_export <- list(
  "GSEA_RNA_1hr" = gsea_rna_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_4hr" = gsea_rna_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_24hr" = gsea_rna_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_results_export, here("output/hypoxia_gsea_rna_hallmark.xlsx"))
cat("\nRNA GSEA results exported to: output/hypoxia_gsea_rna_hallmark.xlsx\n")
```

## RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_heatmap}
# Combine results from all timepoints - only if they have data
gsea_rna_combined <- bind_rows(
  if(nrow(gsea_rna_1hr_results) > 0) gsea_rna_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_4hr_results) > 0) gsea_rna_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_24hr_results) > 0) gsea_rna_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_rna_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_rna_pathways <- gsea_rna_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_rna_pathways) > 0) {
    # Reshape for heatmap
    gsea_rna_wide <- gsea_rna_combined %>%
      filter(ID %in% sig_rna_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_rna_wide)) {
        gsea_rna_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_rna_wide <- gsea_rna_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_rna_matrix <- gsea_rna_combined %>%
      filter(ID %in% sig_rna_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_rna_matrix)) {
        sig_rna_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_rna_matrix <- sig_rna_matrix[rownames(gsea_rna_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_rna_heatmap <- pheatmap(
      gsea_rna_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_rna_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES Across Timepoints - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_heatmap_timecourse.pdf"), width = 8, height = 10)
    print(p_rna_heatmap)
    dev.off()

    cat("RNA heatmap shows", nrow(gsea_rna_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for RNA heatmap\n")
}
```

# GSEA: RNA-seq Changes in si3e Treatment (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (hypoxia vs normoxia) at each timepoint in eIF3e knockdown cells.

## Load si3e RNA differential expression results

```{r load_rna_results_si3e}
# Load RNA results from translation analysis for si3e treatment
rna_results_si3e_1hr <- read_csv(here("output", "translation_categories_si3e_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_si3e_4hr <- read_csv(here("output", "translation_categories_si3e_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_si3e_24hr <- read_csv(here("output", "translation_categories_si3e_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

# Add gene symbols
res_rna_si3e_1hr <- rna_results_si3e_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_si3e_4hr <- rna_results_si3e_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_si3e_24hr <- rna_results_si3e_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

cat("si3e RNA results loaded:\n")
cat("1hr:", nrow(res_rna_si3e_1hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
cat("4hr:", nrow(res_rna_si3e_4hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
cat("24hr:", nrow(res_rna_si3e_24hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## GSEA: si3e RNA 1hr

```{r gsea_rna_si3e_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list_rna_si3e_1hr <- res_rna_si3e_1hr$log2FoldChange
names(ranked_list_rna_si3e_1hr) <- res_rna_si3e_1hr$symbol

# Sort and remove duplicates
ranked_list_rna_si3e_1hr <- sort(ranked_list_rna_si3e_1hr, decreasing = TRUE)
ranked_list_rna_si3e_1hr <- ranked_list_rna_si3e_1hr[!duplicated(names(ranked_list_rna_si3e_1hr))]

cat("si3e 1hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3e_1hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3e_1hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3e_1hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3e_1hr <- GSEA(
  geneList = ranked_list_rna_si3e_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3e_1hr@result) > 0) {
  gsea_rna_si3e_1hr_results <- gsea_rna_si3e_1hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_1hr_results[] <- lapply(gsea_rna_si3e_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e 1hr - Top pathways:\n")
  print(gsea_rna_si3e_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e 1hr - No pathways enriched\n")
}
```

## GSEA: si3e RNA 4hr

```{r gsea_rna_si3e_4hr}
# Create ranked list (using all genes)
ranked_list_rna_si3e_4hr <- res_rna_si3e_4hr$log2FoldChange
names(ranked_list_rna_si3e_4hr) <- res_rna_si3e_4hr$symbol
ranked_list_rna_si3e_4hr <- sort(ranked_list_rna_si3e_4hr, decreasing = TRUE)
ranked_list_rna_si3e_4hr <- ranked_list_rna_si3e_4hr[!duplicated(names(ranked_list_rna_si3e_4hr))]

cat("si3e 4hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3e_4hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3e_4hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3e_4hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3e_4hr <- GSEA(
  geneList = ranked_list_rna_si3e_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3e_4hr@result) > 0) {
  gsea_rna_si3e_4hr_results <- gsea_rna_si3e_4hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_4hr_results[] <- lapply(gsea_rna_si3e_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e 4hr - Top pathways:\n")
  print(gsea_rna_si3e_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e 4hr - No pathways enriched\n")
}
```

## GSEA: si3e RNA 24hr

```{r gsea_rna_si3e_24hr}
# Create ranked list (using all genes)
ranked_list_rna_si3e_24hr <- res_rna_si3e_24hr$log2FoldChange
names(ranked_list_rna_si3e_24hr) <- res_rna_si3e_24hr$symbol
ranked_list_rna_si3e_24hr <- sort(ranked_list_rna_si3e_24hr, decreasing = TRUE)
ranked_list_rna_si3e_24hr <- ranked_list_rna_si3e_24hr[!duplicated(names(ranked_list_rna_si3e_24hr))]

cat("si3e 24hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3e_24hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3e_24hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3e_24hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3e_24hr <- GSEA(
  geneList = ranked_list_rna_si3e_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3e_24hr@result) > 0) {
  gsea_rna_si3e_24hr_results <- gsea_rna_si3e_24hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_24hr_results[] <- lapply(gsea_rna_si3e_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e 24hr - Top pathways:\n")
  print(gsea_rna_si3e_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e 24hr - No pathways enriched\n")
}
```

## Export si3e RNA GSEA results

```{r export_gsea_rna_si3e}
# Compile all si3e RNA GSEA results
gsea_rna_si3e_results_export <- list(
  "GSEA_RNA_si3e_1hr" = gsea_rna_si3e_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_4hr" = gsea_rna_si3e_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_24hr" = gsea_rna_si3e_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3e_results_export, here("output/hypoxia_gsea_rna_si3e_hallmark.xlsx"))
cat("\nsi3e RNA GSEA results exported to: output/hypoxia_gsea_rna_si3e_hallmark.xlsx\n")
```

## si3e RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3e_heatmap}
# Combine results from all timepoints - only if they have data
gsea_rna_si3e_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_1hr_results) > 0) gsea_rna_si3e_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_4hr_results) > 0) gsea_rna_si3e_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_24hr_results) > 0) gsea_rna_si3e_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_rna_si3e_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_rna_si3e_pathways <- gsea_rna_si3e_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_rna_si3e_pathways) > 0) {
    # Reshape for heatmap
    gsea_rna_si3e_wide <- gsea_rna_si3e_combined %>%
      filter(ID %in% sig_rna_si3e_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_rna_si3e_wide)) {
        gsea_rna_si3e_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_rna_si3e_wide <- gsea_rna_si3e_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_rna_si3e_matrix <- gsea_rna_si3e_combined %>%
      filter(ID %in% sig_rna_si3e_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_rna_si3e_matrix)) {
        sig_rna_si3e_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_rna_si3e_matrix <- sig_rna_si3e_matrix[rownames(gsea_rna_si3e_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_rna_si3e_heatmap <- pheatmap(
      gsea_rna_si3e_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_rna_si3e_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES Across Timepoints - si3e - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3e_heatmap_timecourse.pdf"), width = 8, height = 10)
    print(p_rna_si3e_heatmap)
    dev.off()

    cat("si3e RNA heatmap shows", nrow(gsea_rna_si3e_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3e RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3e RNA heatmap\n")
}
```

# GSEA: RNA-seq Changes in si3d Treatment (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (hypoxia vs normoxia) at each timepoint in eIF3d knockdown cells.

## Load si3d RNA differential expression results

```{r load_rna_results_si3d}
# Load RNA results from translation analysis for si3d treatment
rna_results_si3d_1hr <- read_csv(here("output", "translation_categories_si3d_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_si3d_4hr <- read_csv(here("output", "translation_categories_si3d_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

rna_results_si3d_24hr <- read_csv(here("output", "translation_categories_si3d_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, rna_padj)

# Add gene symbols
res_rna_si3d_1hr <- rna_results_si3d_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_si3d_4hr <- rna_results_si3d_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

res_rna_si3d_24hr <- rna_results_si3d_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = rna_lfc, padj = rna_padj)

cat("si3d RNA results loaded:\n")
cat("1hr:", nrow(res_rna_si3d_1hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
cat("4hr:", nrow(res_rna_si3d_4hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
cat("24hr:", nrow(res_rna_si3d_24hr), "genes, Significant (padj < 0.05):",
    sum(res_rna_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## GSEA: si3d RNA 1hr

```{r gsea_rna_si3d_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list_rna_si3d_1hr <- res_rna_si3d_1hr$log2FoldChange
names(ranked_list_rna_si3d_1hr) <- res_rna_si3d_1hr$symbol

# Sort and remove duplicates
ranked_list_rna_si3d_1hr <- sort(ranked_list_rna_si3d_1hr, decreasing = TRUE)
ranked_list_rna_si3d_1hr <- ranked_list_rna_si3d_1hr[!duplicated(names(ranked_list_rna_si3d_1hr))]

cat("si3d 1hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3d_1hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3d_1hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3d_1hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3d_1hr <- GSEA(
  geneList = ranked_list_rna_si3d_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3d_1hr@result) > 0) {
  gsea_rna_si3d_1hr_results <- gsea_rna_si3d_1hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_1hr_results[] <- lapply(gsea_rna_si3d_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d 1hr - Top pathways:\n")
  print(gsea_rna_si3d_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d 1hr - No pathways enriched\n")
}
```

## GSEA: si3d RNA 4hr

```{r gsea_rna_si3d_4hr}
# Create ranked list (using all genes)
ranked_list_rna_si3d_4hr <- res_rna_si3d_4hr$log2FoldChange
names(ranked_list_rna_si3d_4hr) <- res_rna_si3d_4hr$symbol
ranked_list_rna_si3d_4hr <- sort(ranked_list_rna_si3d_4hr, decreasing = TRUE)
ranked_list_rna_si3d_4hr <- ranked_list_rna_si3d_4hr[!duplicated(names(ranked_list_rna_si3d_4hr))]

cat("si3d 4hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3d_4hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3d_4hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3d_4hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3d_4hr <- GSEA(
  geneList = ranked_list_rna_si3d_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3d_4hr@result) > 0) {
  gsea_rna_si3d_4hr_results <- gsea_rna_si3d_4hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_4hr_results[] <- lapply(gsea_rna_si3d_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d 4hr - Top pathways:\n")
  print(gsea_rna_si3d_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d 4hr - No pathways enriched\n")
}
```

## GSEA: si3d RNA 24hr

```{r gsea_rna_si3d_24hr}
# Create ranked list (using all genes)
ranked_list_rna_si3d_24hr <- res_rna_si3d_24hr$log2FoldChange
names(ranked_list_rna_si3d_24hr) <- res_rna_si3d_24hr$symbol
ranked_list_rna_si3d_24hr <- sort(ranked_list_rna_si3d_24hr, decreasing = TRUE)
ranked_list_rna_si3d_24hr <- ranked_list_rna_si3d_24hr[!duplicated(names(ranked_list_rna_si3d_24hr))]

cat("si3d 24hr RNA ranked list:\n")
cat("Total genes:", length(ranked_list_rna_si3d_24hr), "\n")
cat("LFC range:", round(min(ranked_list_rna_si3d_24hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_rna_si3d_24hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_rna_si3d_24hr <- GSEA(
  geneList = ranked_list_rna_si3d_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_rna_si3d_24hr@result) > 0) {
  gsea_rna_si3d_24hr_results <- gsea_rna_si3d_24hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_24hr_results[] <- lapply(gsea_rna_si3d_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d 24hr - Top pathways:\n")
  print(gsea_rna_si3d_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d 24hr - No pathways enriched\n")
}
```

## Export si3d RNA GSEA results

```{r export_gsea_rna_si3d}
# Compile all si3d RNA GSEA results
gsea_rna_si3d_results_export <- list(
  "GSEA_RNA_si3d_1hr" = gsea_rna_si3d_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_4hr" = gsea_rna_si3d_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_24hr" = gsea_rna_si3d_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3d_results_export, here("output/hypoxia_gsea_rna_si3d_hallmark.xlsx"))
cat("\nsi3d RNA GSEA results exported to: output/hypoxia_gsea_rna_si3d_hallmark.xlsx\n")
```

## si3d RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3d_heatmap}
# Combine results from all timepoints - only if they have data
gsea_rna_si3d_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_1hr_results) > 0) gsea_rna_si3d_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_4hr_results) > 0) gsea_rna_si3d_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_24hr_results) > 0) gsea_rna_si3d_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_rna_si3d_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_rna_si3d_pathways <- gsea_rna_si3d_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_rna_si3d_pathways) > 0) {
    # Reshape for heatmap
    gsea_rna_si3d_wide <- gsea_rna_si3d_combined %>%
      filter(ID %in% sig_rna_si3d_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_rna_si3d_wide)) {
        gsea_rna_si3d_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_rna_si3d_wide <- gsea_rna_si3d_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_rna_si3d_matrix <- gsea_rna_si3d_combined %>%
      filter(ID %in% sig_rna_si3d_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_rna_si3d_matrix)) {
        sig_rna_si3d_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_rna_si3d_matrix <- sig_rna_si3d_matrix[rownames(gsea_rna_si3d_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_rna_si3d_heatmap <- pheatmap(
      gsea_rna_si3d_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_rna_si3d_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES Across Timepoints - si3d - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3d_heatmap_timecourse.pdf"), width = 8, height = 10)
    print(p_rna_si3d_heatmap)
    dev.off()

    cat("si3d RNA heatmap shows", nrow(gsea_rna_si3d_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3d RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3d RNA heatmap\n")
}
```

# GSEA: TE Changes (Hypoxia Response Pathway Analysis)

Perform Gene Set Enrichment Analysis (GSEA) on the ranked list of genes based on TE log2 fold changes (hypoxia vs normoxia) at each timepoint, following the approach used in ribopipe-shiny.

## Load MSigDB gene sets

```{r load_msigdb}
# Get Hallmark gene sets from MSigDB
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark$gs_name)), "\n")
```

## GSEA: 1hr Hypoxia

```{r gsea_1hr}
# Create ranked list based on RNA log2FC (hypoxia vs normoxia)
ranked_list_1hr <- deseq_1hr$log2FoldChange
names(ranked_list_1hr) <- deseq_1hr$symbol

# Sort by LFC (descending) and remove duplicates
ranked_list_1hr <- sort(ranked_list_1hr, decreasing = TRUE)
ranked_list_1hr <- ranked_list_1hr[!duplicated(names(ranked_list_1hr))]

cat("1hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_1hr), "\n")
cat("LFC range:", round(min(ranked_list_1hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_1hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_1hr <- GSEA(
  geneList = ranked_list_1hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_1hr_results <- gsea_1hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

# Round numeric columns
gsea_1hr_results[] <- lapply(gsea_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 1hr - Top pathways:\n")
print(gsea_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
```

## GSEA: 4hr Hypoxia

```{r gsea_4hr}
# Create ranked list
ranked_list_4hr <- deseq_4hr$log2FoldChange
names(ranked_list_4hr) <- deseq_4hr$symbol
ranked_list_4hr <- sort(ranked_list_4hr, decreasing = TRUE)
ranked_list_4hr <- ranked_list_4hr[!duplicated(names(ranked_list_4hr))]

cat("4hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_4hr), "\n")
cat("LFC range:", round(min(ranked_list_4hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_4hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_4hr <- GSEA(
  geneList = ranked_list_4hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_4hr_results <- gsea_4hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_4hr_results[] <- lapply(gsea_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 4hr - Top pathways:\n")
print(gsea_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
```

## GSEA: 24hr Hypoxia

```{r gsea_24hr}
# Create ranked list
ranked_list_24hr <- deseq_24hr$log2FoldChange
names(ranked_list_24hr) <- deseq_24hr$symbol
ranked_list_24hr <- sort(ranked_list_24hr, decreasing = TRUE)
ranked_list_24hr <- ranked_list_24hr[!duplicated(names(ranked_list_24hr))]

cat("24hr Hypoxia ranked list:\n")
cat("Total genes:", length(ranked_list_24hr), "\n")
cat("LFC range:", round(min(ranked_list_24hr, na.rm = TRUE), 2), "to",
    round(max(ranked_list_24hr, na.rm = TRUE), 2), "\n\n")

# Run GSEA
gsea_24hr <- GSEA(
  geneList = ranked_list_24hr,
  eps = 0,
    exponent = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark
)

# Extract results
gsea_24hr_results <- gsea_24hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_24hr_results[] <- lapply(gsea_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA 24hr - Top pathways:\n")
print(gsea_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
```

## Export GSEA results

```{r export_gsea}
# Compile all GSEA results (exclude leading_edge column for Excel export)
gsea_results_export <- list(
  "GSEA_1hr" = gsea_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_4hr" = gsea_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_24hr" = gsea_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_results_export, here("output/hypoxia_gsea_hallmark.xlsx"))
cat("\nGSEA results exported to: output/hypoxia_gsea_hallmark.xlsx\n")
```

## Additional Visualizations

### 1. Heatmap: Pathway NES across timepoints

Shows how pathway enrichment changes over the timecourse.

```{r gsea_heatmap}
# Combine results from all timepoints
gsea_combined <- bind_rows(
  gsea_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_pathways <- gsea_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
gsea_wide <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  dplyr::select(ID, NES, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
gsea_wide <- gsea_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_matrix <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match gsea_wide
sig_matrix <- sig_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches between gsea_wide and sig_matrix
sig_matrix <- sig_matrix[rownames(gsea_wide), ]

# Create heatmap
if (nrow(gsea_wide) > 0) {
  p_heatmap <- pheatmap(
    gsea_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    display_numbers = sig_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "GSEA NES Across Timepoints (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/gsea_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_heatmap)
  dev.off()

  cat("Heatmap shows", nrow(gsea_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for heatmap\n")
}
```

# ORA: Hallmark Pathway Enrichment for Upregulated Genes

Perform Over-Representation Analysis (ORA) to test if upregulated genes at each timepoint are enriched for Hallmark pathways.

## ORA: 1hr Upregulated Genes

```{r ora_hallmark_1hr}
# Run ORA on upregulated genes at 1hr
ora_1hr <- enricher(
  gene = up_1_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_1hr_results <- ora_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_1hr_results[] <- lapply(ora_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 1hr - Enriched pathways:\n")
print(ora_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_1hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_1hr <- ggplot(ora_1hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 1hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_1hr)
  ggsave(here("plots/ora_hallmark_1hr.pdf"), p_ora_1hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 1hr upregulated genes\n")
}
```

## ORA: 4hr Upregulated Genes

```{r ora_hallmark_4hr}
# Run ORA on upregulated genes at 4hr
ora_4hr <- enricher(
  gene = up_4_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_4hr_results <- ora_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_4hr_results[] <- lapply(ora_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 4hr - Enriched pathways:\n")
print(ora_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_4hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_4hr <- ggplot(ora_4hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 4hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_4hr)
  ggsave(here("plots/ora_hallmark_4hr.pdf"), p_ora_4hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 4hr upregulated genes\n")
}
```

## ORA: 24hr Upregulated Genes

```{r ora_hallmark_24hr}
# Run ORA on upregulated genes at 24hr
ora_24hr <- enricher(
  gene = up_24_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_24hr_results <- ora_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_24hr_results[] <- lapply(ora_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 24hr - Enriched pathways:\n")
print(ora_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_24hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_24hr <- ggplot(ora_24hr_results %>% filter(p.adjust < 0.05),
                         aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                             fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 24hr Upregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_24hr)
  ggsave(here("plots/ora_hallmark_24hr.pdf"), p_ora_24hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 24hr upregulated genes\n")
}
```

## Export ORA results

```{r export_ora}
# Compile all ORA results
ora_results_export <- list(
  "ORA_1hr" = ora_1hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_4hr" = ora_4hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_24hr" = ora_24hr_results %>% dplyr::select(-GeneRatio_numeric)
)

write_xlsx(ora_results_export, here("output/hypoxia_ora_hallmark.xlsx"))
cat("\nORA results exported to: output/hypoxia_ora_hallmark.xlsx\n")
```

## ORA Heatmap: Pathway enrichment across timepoints

```{r ora_heatmap}
# Combine results from all timepoints
ora_combined <- bind_rows(
  ora_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_ora_pathways <- ora_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
ora_wide <- ora_combined %>%
  filter(ID %in% sig_ora_pathways) %>%
  dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
ora_wide <- ora_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_ora_matrix <- ora_combined %>%
  filter(ID %in% sig_ora_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match ora_wide
sig_ora_matrix <- sig_ora_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_ora_matrix <- sig_ora_matrix[rownames(ora_wide), ]

# Create heatmap
if (nrow(ora_wide) > 0) {
  p_ora_heatmap <- pheatmap(
    ora_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "red"))(100),
    breaks = seq(0, max(ora_wide), length.out = 101),
    display_numbers = sig_ora_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA Gene Ratio Across Timepoints - Hallmark (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/ora_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_ora_heatmap)
  dev.off()

  cat("ORA heatmap shows", nrow(ora_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for ORA heatmap\n")
}
```

# ORA: Hallmark Pathway Enrichment for Downregulated Genes (Sanity Check)

Perform Over-Representation Analysis (ORA) on downregulated genes to check for expected pathway depletion patterns.

## ORA: 1hr Downregulated Genes

```{r ora_hallmark_down_1hr}
# Run ORA on downregulated genes at 1hr
ora_down_1hr <- enricher(
  gene = down_1_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_1hr_results <- ora_down_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_1hr_results[] <- lapply(ora_down_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 1hr Downregulated - Enriched pathways:\n")
print(ora_down_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_1hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_1hr <- ggplot(ora_down_1hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 1hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_1hr)
  ggsave(here("plots/ora_hallmark_down_1hr.pdf"), p_ora_down_1hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 1hr downregulated genes\n")
}
```

## ORA: 4hr Downregulated Genes

```{r ora_hallmark_down_4hr}
# Run ORA on downregulated genes at 4hr
ora_down_4hr <- enricher(
  gene = down_4_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_4hr_results <- ora_down_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_4hr_results[] <- lapply(ora_down_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 4hr Downregulated - Enriched pathways:\n")
print(ora_down_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_4hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_4hr <- ggplot(ora_down_4hr_results %>% filter(p.adjust < 0.05),
                        aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                            fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 4hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_4hr)
  ggsave(here("plots/ora_hallmark_down_4hr.pdf"), p_ora_down_4hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 4hr downregulated genes\n")
}
```

## ORA: 24hr Downregulated Genes

```{r ora_hallmark_down_24hr}
# Run ORA on downregulated genes at 24hr
ora_down_24hr <- enricher(
  gene = down_24_hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_down_24hr_results <- ora_down_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_down_24hr_results[] <- lapply(ora_down_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA Hallmark 24hr Downregulated - Enriched pathways:\n")
print(ora_down_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(p.adjust) %>% head(10))

# Dot plot
if (nrow(ora_down_24hr_results %>% filter(p.adjust < 0.05)) > 0) {
  p_ora_down_24hr <- ggplot(ora_down_24hr_results %>% filter(p.adjust < 0.05),
                         aes(y = reorder(ID, GeneRatio_numeric), x = GeneRatio_numeric,
                             fill = p.adjust, size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(2, 10)) +
    scale_fill_viridis(discrete = FALSE, direction = -1) +
    labs(y = "Pathways",
         x = "Gene Ratio",
         title = "ORA: 24hr Downregulated Genes (Hallmark Pathways)",
         size = "Gene Count",
         fill = "Adjusted P-value") +
    theme_cowplot() +
    theme(axis.text.y = element_text(size = 9))

  print(p_ora_down_24hr)
  ggsave(here("plots/ora_hallmark_down_24hr.pdf"), p_ora_down_24hr, width = 10, height = 8)
} else {
  cat("No significant pathways for 24hr downregulated genes\n")
}
```

## Export ORA results (Downregulated)

```{r export_ora_down}
# Compile all ORA results for downregulated genes
ora_down_results_export <- list(
  "ORA_Down_1hr" = ora_down_1hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_Down_4hr" = ora_down_4hr_results %>% dplyr::select(-GeneRatio_numeric),
  "ORA_Down_24hr" = ora_down_24hr_results %>% dplyr::select(-GeneRatio_numeric)
)

write_xlsx(ora_down_results_export, here("output/hypoxia_ora_hallmark_downregulated.xlsx"))
cat("\nORA results for downregulated genes exported to: output/hypoxia_ora_hallmark_downregulated.xlsx\n")
```

## ORA Heatmap: Pathway enrichment across timepoints (Downregulated)

```{r ora_heatmap_down}
# Combine results from all timepoints
ora_down_combined <- bind_rows(
  ora_down_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_down_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_down_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_ora_down_pathways <- ora_down_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
ora_down_wide <- ora_down_combined %>%
  filter(ID %in% sig_ora_down_pathways) %>%
  dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
ora_down_wide <- ora_down_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_ora_down_matrix <- ora_down_combined %>%
  filter(ID %in% sig_ora_down_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match ora_down_wide
sig_ora_down_matrix <- sig_ora_down_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_ora_down_matrix <- sig_ora_down_matrix[rownames(ora_down_wide), ]

# Create heatmap
if (nrow(ora_down_wide) > 0) {
  p_ora_down_heatmap <- pheatmap(
    ora_down_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "blue"))(100),
    breaks = seq(0, max(ora_down_wide), length.out = 101),
    display_numbers = sig_ora_down_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA Gene Ratio Across Timepoints - Downregulated (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/ora_heatmap_downregulated_timecourse.pdf"), width = 8, height = 10)
  print(p_ora_down_heatmap)
  dev.off()

  cat("ORA heatmap for downregulated genes shows", nrow(ora_down_wide), "significant pathways\n")
} else {
  cat("No significant pathways found for ORA heatmap (downregulated genes)\n")
}
```

# GSEA: KEGG Pathways

Perform the same GSEA analysis using KEGG pathways instead of Hallmark pathways.

## Load KEGG gene sets

```{r load_kegg}
# Get KEGG pathways from MSigDB
kegg_pathways_df <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB KEGG gene sets:\n")
cat("Number of pathways:", length(unique(kegg_pathways_df$gs_name)), "\n")
```

## GSEA: 1hr Hypoxia (KEGG)

```{r gsea_kegg_1hr}
# Run GSEA with KEGG pathways
set.seed(42)
gsea_kegg_1hr <- GSEA(
  geneList = ranked_list_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_1hr_results <- gsea_kegg_1hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

# Round numeric columns
gsea_kegg_1hr_results[] <- lapply(gsea_kegg_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 1hr - Top pathways:\n")
print(gsea_kegg_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot for significant pathways
p_gsea_kegg_1hr <- ggplot(gsea_kegg_1hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 1hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_1hr)
ggsave(here("plots/gsea_kegg_1hr.pdf"), p_gsea_kegg_1hr, width = 10, height = 8)
```

## GSEA: 4hr Hypoxia (KEGG)

```{r gsea_kegg_4hr}
# Run GSEA
set.seed(42)
gsea_kegg_4hr <- GSEA(
  geneList = ranked_list_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_4hr_results <- gsea_kegg_4hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_kegg_4hr_results[] <- lapply(gsea_kegg_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 4hr - Top pathways:\n")
print(gsea_kegg_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_kegg_4hr <- ggplot(gsea_kegg_4hr_results %>% filter(p.adjust < 0.05),
                      aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 4hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_4hr)
ggsave(here("plots/gsea_kegg_4hr.pdf"), p_gsea_kegg_4hr, width = 10, height = 8)
```

## GSEA: 24hr Hypoxia (KEGG)

```{r gsea_kegg_24hr}
# Run GSEA
set.seed(42)
gsea_kegg_24hr <- GSEA(
  geneList = ranked_list_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_pathways_df
)

# Extract results
gsea_kegg_24hr_results <- gsea_kegg_24hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_kegg_24hr_results[] <- lapply(gsea_kegg_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA KEGG 24hr - Top pathways:\n")
print(gsea_kegg_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))

# Dot plot
p_gsea_kegg_24hr <- ggplot(gsea_kegg_24hr_results %>% filter(p.adjust < 0.05),
                       aes(y = reorder(ID, NES), x = NES, fill = p.adjust, size = leading_edge_count)) +
  geom_point(alpha = 0.7, shape = 21, color = "black") +
  scale_size(range = c(2, 10)) +
  scale_fill_viridis(discrete = FALSE, direction = -1) +
  labs(y = "Pathways",
       x = "Normalized Enrichment Score (NES)",
       title = "GSEA: 24hr Hypoxia (KEGG Pathways)",
       size = "Number of Genes\nin Leading Edge",
       fill = "Adjusted P-value") +
  theme_cowplot() +
  theme(axis.text.y = element_text(size = 8))

print(p_gsea_kegg_24hr)
ggsave(here("plots/gsea_kegg_24hr.pdf"), p_gsea_kegg_24hr, width = 10, height = 8)
```

## Export KEGG GSEA results

```{r export_gsea_kegg}
# Compile all KEGG GSEA results
gsea_kegg_results_export <- list(
  "GSEA_KEGG_1hr" = gsea_kegg_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_KEGG_4hr" = gsea_kegg_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_KEGG_24hr" = gsea_kegg_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_kegg_results_export, here("output/hypoxia_gsea_kegg.xlsx"))
cat("\nKEGG GSEA results exported to: output/hypoxia_gsea_kegg.xlsx\n")
```

## KEGG Heatmap: Pathway NES across timepoints

```{r gsea_kegg_heatmap}
# Combine results from all timepoints
gsea_kegg_combined <- bind_rows(
  gsea_kegg_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_kegg_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_kegg_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_kegg_pathways <- gsea_kegg_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

# Reshape for heatmap
gsea_kegg_wide <- gsea_kegg_combined %>%
  filter(ID %in% sig_kegg_pathways) %>%
  dplyr::select(ID, NES, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
gsea_kegg_wide <- gsea_kegg_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Create significance annotation matrix
sig_kegg_matrix <- gsea_kegg_combined %>%
  filter(ID %in% sig_kegg_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match gsea_kegg_wide
sig_kegg_matrix <- sig_kegg_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_kegg_matrix <- sig_kegg_matrix[rownames(gsea_kegg_wide), ]

# Create heatmap
if (nrow(gsea_kegg_wide) > 0) {
  p_kegg_heatmap <- pheatmap(
    gsea_kegg_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    display_numbers = sig_kegg_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "GSEA NES Across Timepoints - KEGG (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots/gsea_kegg_heatmap_timecourse.pdf"), width = 8, height = 10)
  print(p_kegg_heatmap)
  dev.off()

  cat("KEGG heatmap shows", nrow(gsea_kegg_wide), "significant pathways\n")
} else {
  cat("No significant KEGG pathways found for heatmap\n")
}
```

# GSEA: Custom MDA-MB-231 Gene Sets

Perform GSEA analysis using custom gene sets derived from MDA-MB-231 cell line experiments, including eIF3d/eIF3e knockdown effects on translation efficiency under normoxia and hypoxia conditions.

## Load Custom Gene Sets

```{r load_custom_genesets}
# Load custom gene sets from CSV
custom_genesets_raw <- read_csv(here("accessories", "all_combined_sigs_12_9_25.csv"))

# Filter to MDA-MB-231 gene sets and create TERM2GENE format
custom_genesets <- custom_genesets_raw %>%
  filter(str_starts(gs_name, "MDA-MB-231_")) %>%
  # Exclude overlap gene sets
  filter(!str_detect(gs_name, "overlap")) %>%
  # Exclude genes where db_gene_symbol is just an ENSG ID (no standard gene symbol)
  filter(!str_starts(db_gene_symbol, "ENSG")) %>%
  # Remove any NA values in gene symbols
  filter(!is.na(db_gene_symbol), db_gene_symbol != "") %>%
  dplyr::select(gs_name, db_gene_symbol) %>%
  distinct()

cat("Loaded custom MDA-MB-231 gene sets:\n")
cat("Number of gene sets:", length(unique(custom_genesets$gs_name)), "\n")
cat("Total gene-pathway associations:", nrow(custom_genesets), "\n")

# Show gene set size distribution
geneset_sizes <- custom_genesets %>%
  group_by(gs_name) %>%
  summarise(size = n()) %>%
  arrange(size)

cat("\nGene set size distribution:\n")
cat("Min:", min(geneset_sizes$size), "\n")
cat("Max:", max(geneset_sizes$size), "\n")
cat("Mean:", round(mean(geneset_sizes$size), 1), "\n")
cat("Median:", median(geneset_sizes$size), "\n")
```

## GSEA: 1hr Hypoxia (Custom Gene Sets)

```{r gsea_custom_1hr}
# Run GSEA with custom gene sets (using pre-existing ranked list)
set.seed(42)
gsea_custom_1hr <- GSEA(
  geneList = ranked_list_1hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 10,     # Lower than Hallmark/KEGG to capture smaller custom sets
  maxGSSize = 1000,   # Filter out very large sets (2500-gene sets)
  TERM2GENE = custom_genesets
)

# Extract results
gsea_custom_1hr_results <- gsea_custom_1hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

# Round numeric columns
gsea_custom_1hr_results[] <- lapply(gsea_custom_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA Custom 1hr - Top pathways:\n")
cat("Total pathways tested:", nrow(gsea_custom_1hr_results), "\n")
cat("Significant pathways (p.adjust < 0.05):",
    sum(gsea_custom_1hr_results$p.adjust < 0.05), "\n\n")
print(gsea_custom_1hr_results %>%
        filter(p.adjust < 0.05) %>%
        arrange(desc(abs(NES))) %>%
        head(15))
```

## GSEA: 4hr Hypoxia (Custom Gene Sets)

```{r gsea_custom_4hr}
# Run GSEA with custom gene sets
set.seed(42)
gsea_custom_4hr <- GSEA(
  geneList = ranked_list_4hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 1000,
  TERM2GENE = custom_genesets
)

# Extract results
gsea_custom_4hr_results <- gsea_custom_4hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_custom_4hr_results[] <- lapply(gsea_custom_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA Custom 4hr - Top pathways:\n")
cat("Total pathways tested:", nrow(gsea_custom_4hr_results), "\n")
cat("Significant pathways (p.adjust < 0.05):",
    sum(gsea_custom_4hr_results$p.adjust < 0.05), "\n\n")
print(gsea_custom_4hr_results %>%
        filter(p.adjust < 0.05) %>%
        arrange(desc(abs(NES))) %>%
        head(15))
```

## GSEA: 24hr Hypoxia (Custom Gene Sets)

```{r gsea_custom_24hr}
# Run GSEA with custom gene sets
set.seed(42)
gsea_custom_24hr <- GSEA(
  geneList = ranked_list_24hr,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 1000,
  TERM2GENE = custom_genesets
)

# Extract results
gsea_custom_24hr_results <- gsea_custom_24hr@result %>%
  dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
  rename("leading_edge" = core_enrichment) %>%
  mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
         leading_edge_fraction = leading_edge_count / setSize)

gsea_custom_24hr_results[] <- lapply(gsea_custom_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("GSEA Custom 24hr - Top pathways:\n")
cat("Total pathways tested:", nrow(gsea_custom_24hr_results), "\n")
cat("Significant pathways (p.adjust < 0.05):",
    sum(gsea_custom_24hr_results$p.adjust < 0.05), "\n\n")
print(gsea_custom_24hr_results %>%
        filter(p.adjust < 0.05) %>%
        arrange(desc(abs(NES))) %>%
        head(15))
```

## Export Custom GSEA Results

```{r export_gsea_custom}
# Compile all custom GSEA results
gsea_custom_results_export <- list(
  "GSEA_Custom_1hr" = gsea_custom_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_Custom_4hr" = gsea_custom_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_Custom_24hr" = gsea_custom_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_custom_results_export, here("output/hypoxia_gsea_custom.xlsx"))
cat("\nCustom GSEA results exported to: output/hypoxia_gsea_custom.xlsx\n")
```

## Custom Gene Sets Heatmap: Pathway NES across timepoints

```{r gsea_custom_heatmap}
# Combine results from all timepoints
gsea_custom_combined <- bind_rows(
  gsea_custom_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_custom_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_custom_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_custom_pathways <- gsea_custom_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

cat("Found", length(sig_custom_pathways), "pathways significant in at least one timepoint\n")

# Reshape for heatmap
gsea_custom_wide <- gsea_custom_combined %>%
  filter(ID %in% sig_custom_pathways) %>%
  dplyr::select(ID, NES, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
  column_to_rownames("ID")

# Reorder columns
gsea_custom_wide <- gsea_custom_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Sort rows by 1hr NES (groups positive and negative enrichments together)
gsea_custom_wide <- gsea_custom_wide[order(gsea_custom_wide$`1hr`, decreasing = TRUE), ]

# Create significance annotation matrix
sig_custom_matrix <- gsea_custom_combined %>%
  filter(ID %in% sig_custom_pathways) %>%
  dplyr::select(ID, p.adjust, timepoint) %>%
  mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
  dplyr::select(ID, timepoint, sig) %>%
  pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
  column_to_rownames("ID")

# Reorder columns to match gsea_custom_wide
sig_custom_matrix <- sig_custom_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

# Ensure row order matches
sig_custom_matrix <- sig_custom_matrix[rownames(gsea_custom_wide), ]

# Create heatmap with dynamic height
if (nrow(gsea_custom_wide) > 0) {
  # Calculate height based on number of pathways (12 pixels per row + margin)
  heatmap_height <- max(8, min(20, 0.25 * nrow(gsea_custom_wide) + 2))

  p_custom_heatmap <- pheatmap(
    gsea_custom_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(100),
    breaks = seq(-3, 3, length.out = 101),
    display_numbers = sig_custom_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "GSEA NES Across Timepoints - Custom MDA-MB-231 Gene Sets (* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save with dynamic height
  pdf(here("plots/gsea_custom_heatmap_timecourse.pdf"), width = 8, height = heatmap_height)
  print(p_custom_heatmap)
  dev.off()

  cat("Custom gene sets heatmap shows", nrow(gsea_custom_wide), "significant pathways\n")
  cat("Heatmap saved to: plots/gsea_custom_heatmap_timecourse.pdf\n")
} else {
  cat("No significant custom pathways found for heatmap\n")
}
```

# Comprehensive Analysis: Hypoxia Response with Knockdown Effects


## Load Translation Category Data

```{r load_translation_categories}
# Load hypoxia response categories
hyp_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
hyp_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
hyp_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load eIF3d knockdown categories
si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))

# Load eIF3e knockdown categories
si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"))
si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"))
si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"))
si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))

# Get gene symbols from deseq_1hr (already loaded)
gene_symbols <- deseq_1hr %>% dplyr::select(gene_id, symbol)

cat("Loaded all translation category data\n")
```

## Summary of RNA, Ribo, and TE Changes Across Timepoints

```{r summary_changes_hypoxia}
# Function to count significant changes
count_changes <- function(df, lfc_threshold = 0) {
  data.frame(
    RNA_up = sum(df$rna_padj < 0.05 & df$rna_lfc > lfc_threshold, na.rm = TRUE),
    RNA_down = sum(df$rna_padj < 0.05 & df$rna_lfc < -lfc_threshold, na.rm = TRUE),
    Ribo_up = sum(df$ribo_padj < 0.05 & df$ribo_lfc > lfc_threshold, na.rm = TRUE),
    Ribo_down = sum(df$ribo_padj < 0.05 & df$ribo_lfc < -lfc_threshold, na.rm = TRUE),
    TE_up = sum(df$te_padj < 0.05 & df$te_lfc > lfc_threshold, na.rm = TRUE),
    TE_down = sum(df$te_padj < 0.05 & df$te_lfc < -lfc_threshold, na.rm = TRUE)
  )
}

# Count with padj < 0.05 and any LFC (LFC != 0)
cat("=== Changes with padj < 0.05 and LFC > 0 or < 0 ===\n\n")
changes_1hr <- count_changes(hyp_1hr, lfc_threshold = 0)
changes_4hr <- count_changes(hyp_4hr, lfc_threshold = 0)
changes_24hr <- count_changes(hyp_24hr, lfc_threshold = 0)

summary_table <- rbind(
  data.frame(Timepoint = "1hr", changes_1hr),
  data.frame(Timepoint = "4hr", changes_4hr),
  data.frame(Timepoint = "24hr", changes_24hr)
)

print(summary_table)

cat("\n=== Changes with padj < 0.05 and LFC > 0.5 or < -0.5 ===\n\n")
changes_1hr_lfc05 <- count_changes(hyp_1hr, lfc_threshold = 0.5)
changes_4hr_lfc05 <- count_changes(hyp_4hr, lfc_threshold = 0.5)
changes_24hr_lfc05 <- count_changes(hyp_24hr, lfc_threshold = 0.5)

summary_table_lfc05 <- rbind(
  data.frame(Timepoint = "1hr", changes_1hr_lfc05),
  data.frame(Timepoint = "4hr", changes_4hr_lfc05),
  data.frame(Timepoint = "24hr", changes_24hr_lfc05)
)

print(summary_table_lfc05)
```

### Visualize Changes as Barplots

```{r barplots_changes}
# Reshape data for plotting - LFC > 0 (RNA and Ribo only)
plot_data_rna_ribo <- summary_table %>%
  pivot_longer(cols = -Timepoint, names_to = "Change_Type", values_to = "Count") %>%
  separate(Change_Type, into = c("Data_Type", "Direction"), sep = "_") %>%
  filter(Data_Type %in% c("RNA", "Ribo")) %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Data_Type = factor(Data_Type, levels = c("RNA", "Ribo")),
    Direction = factor(Direction, levels = c("up", "down")),
    Direction = recode(Direction, "up" = "Up", "down" = "Down"),
    Type_Direction = interaction(Data_Type, Direction, sep = "\n"),
    Type_Direction = factor(Type_Direction, levels = c("RNA\nUp", "RNA\nDown", "Ribo\nUp", "Ribo\nDown"))
  )

# Create barplot for RNA and Ribo - LFC > 0
p1_rna_ribo <- ggplot(plot_data_rna_ribo, aes(x = Type_Direction, y = Count, fill = Data_Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = Count), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("RNA" = "#CC79A7", "Ribo" = "#0072B2")) +
  facet_wrap(~Timepoint, ncol = 3) +
  labs(
    title = "RNA and Ribo Changes in Hypoxia (padj < 0.05, LFC > 0)",
    x = "",
    y = "Number of Genes",
    fill = "Data Type"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 9)
  )

print(p1_rna_ribo)

# Reshape data for plotting - LFC > 0 (TE only)
plot_data_te <- summary_table %>%
  pivot_longer(cols = -Timepoint, names_to = "Change_Type", values_to = "Count") %>%
  separate(Change_Type, into = c("Data_Type", "Direction"), sep = "_") %>%
  filter(Data_Type == "TE") %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Direction = factor(Direction, levels = c("up", "down")),
    Direction = recode(Direction, "up" = "Up", "down" = "Down"),
    Type_Direction = interaction(Data_Type, Direction, sep = "\n"),
    Type_Direction = factor(Type_Direction, levels = c("TE\nUp", "TE\nDown"))
  )

# Create barplot for TE - LFC > 0
p1_te <- ggplot(plot_data_te, aes(x = Type_Direction, y = Count, fill = Data_Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = Count), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("TE" = "#009E73")) +
  facet_wrap(~Timepoint, ncol = 3) +
  labs(
    title = "TE Changes in Hypoxia (padj < 0.05, LFC > 0)",
    x = "",
    y = "Number of Genes",
    fill = "Data Type"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 9)
  )

print(p1_te)

# Reshape data for plotting - LFC > 0.5 (RNA and Ribo only)
plot_data_rna_ribo_lfc05 <- summary_table_lfc05 %>%
  pivot_longer(cols = -Timepoint, names_to = "Change_Type", values_to = "Count") %>%
  separate(Change_Type, into = c("Data_Type", "Direction"), sep = "_") %>%
  filter(Data_Type %in% c("RNA", "Ribo")) %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Data_Type = factor(Data_Type, levels = c("RNA", "Ribo")),
    Direction = factor(Direction, levels = c("up", "down")),
    Direction = recode(Direction, "up" = "Up", "down" = "Down"),
    Type_Direction = interaction(Data_Type, Direction, sep = "\n"),
    Type_Direction = factor(Type_Direction, levels = c("RNA\nUp", "RNA\nDown", "Ribo\nUp", "Ribo\nDown"))
  )

# Create barplot for RNA and Ribo - LFC > 0.5
p2_rna_ribo <- ggplot(plot_data_rna_ribo_lfc05, aes(x = Type_Direction, y = Count, fill = Data_Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = Count), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("RNA" = "#CC79A7", "Ribo" = "#0072B2")) +
  facet_wrap(~Timepoint, ncol = 3) +
  labs(
    title = "RNA and Ribo Changes in Hypoxia (padj < 0.05, |LFC| > 0.5)",
    x = "",
    y = "Number of Genes",
    fill = "Data Type"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 9)
  )

print(p2_rna_ribo)

# Reshape data for plotting - LFC > 0.5 (TE only)
plot_data_te_lfc05 <- summary_table_lfc05 %>%
  pivot_longer(cols = -Timepoint, names_to = "Change_Type", values_to = "Count") %>%
  separate(Change_Type, into = c("Data_Type", "Direction"), sep = "_") %>%
  filter(Data_Type == "TE") %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("1hr", "4hr", "24hr")),
    Direction = factor(Direction, levels = c("up", "down")),
    Direction = recode(Direction, "up" = "Up", "down" = "Down"),
    Type_Direction = interaction(Data_Type, Direction, sep = "\n"),
    Type_Direction = factor(Type_Direction, levels = c("TE\nUp", "TE\nDown"))
  )

# Create barplot for TE - LFC > 0.5
p2_te <- ggplot(plot_data_te_lfc05, aes(x = Type_Direction, y = Count, fill = Data_Type)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = Count), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("TE" = "#009E73")) +
  facet_wrap(~Timepoint, ncol = 3) +
  labs(
    title = "TE Changes in Hypoxia (padj < 0.05, |LFC| > 0.5)",
    x = "",
    y = "Number of Genes",
    fill = "Data Type"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 9)
  )

print(p2_te)

# Save plots
ggsave(here("plots/hypoxia_rna_ribo_changes_lfc0_barplot.pdf"), p1_rna_ribo, width = 12, height = 5)
ggsave(here("plots/hypoxia_te_changes_lfc0_barplot.pdf"), p1_te, width = 12, height = 5)
ggsave(here("plots/hypoxia_rna_ribo_changes_lfc05_barplot.pdf"), p2_rna_ribo, width = 12, height = 5)
ggsave(here("plots/hypoxia_te_changes_lfc05_barplot.pdf"), p2_te, width = 12, height = 5)
```

## Get TE-Regulated Genes from Hypoxia Response

```{r get_te_regulated_genes}
# Get genes with TE regulation (Exclusive, Buffered, TE_only) at any timepoint
te_regulated_genes <- bind_rows(
  hyp_1hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id),
  hyp_4hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id),
  hyp_24hr %>% filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>% dplyr::select(gene_id)
)

# Get unique gene list
te_regulated_gene_ids <- unique(te_regulated_genes$gene_id)

cat("Total genes with TE regulation in hypoxia response:", length(te_regulated_gene_ids), "\n")
cat("Breakdown by timepoint:\n")
cat("  1hr:", sum(hyp_1hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
cat("  4hr:", sum(hyp_4hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
cat("  24hr:", sum(hyp_24hr$category %in% c("Exclusive", "Buffered", "TE_only")), "genes\n")
```

## Create Helper Function

```{r helper_function}
# Function to determine if TE is significantly changed
# TE is significant in categories: Exclusive, Buffered, TE_only
# TE is NOT significant in: Forwarded, Not_sig
is_te_not_significant <- function(category) {
  category %in% c("Forwarded", "Not_sig")
}
```

## Build Comprehensive Table

```{r build_table}
# Start with TE-regulated genes from hypoxia response
analysis_table <- data.frame(gene_id = te_regulated_gene_ids)

colnames(analysis_table)

# Add gene symbols
analysis_table <- analysis_table %>%
  left_join(gene_symbols, by = "gene_id")

# Add hypoxia response data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(hyp_1hr %>% 
            rename(hyp_1hr_rna_lfc = rna_lfc, hyp_1hr_ribo_lfc = ribo_lfc, hyp_1hr_te_lfc = te_lfc,
                   hyp_1hr_category = category,
                   te_padj_1hr = te_padj),
            by = "gene_id") %>%
  left_join(hyp_4hr %>%
            rename(hyp_4hr_rna_lfc = rna_lfc, hyp_4hr_ribo_lfc = ribo_lfc, hyp_4hr_te_lfc = te_lfc,
                    hyp_4hr_category = category,
                    te_padj_4hr = te_padj),
            by = "gene_id") %>%
  left_join(hyp_24hr %>%
            rename(hyp_24hr_rna_lfc = rna_lfc, hyp_24hr_ribo_lfc = ribo_lfc, hyp_24hr_te_lfc = te_lfc,
                 hyp_24hr_category = category,
                 te_padj_24hr = te_padj),
            by = "gene_id")

```

```{r add_si3d_data}
# Add eIF3d knockdown data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(si3d_norm_1hr %>% 
            rename(si3d_norm_1hr_rna_lfc = rna_lfc, si3d_norm_1hr_ribo_lfc = ribo_lfc, si3d_norm_1hr_te_lfc = te_lfc,
                   si3d_norm_1hr_category = category), by = "gene_id") %>%
  left_join(si3d_norm_4hr %>% 
            rename(si3d_norm_4hr_rna_lfc = rna_lfc, si3d_norm_4hr_ribo_lfc = ribo_lfc, si3d_norm_4hr_te_lfc = te_lfc,
                   si3d_norm_4hr_category = category), by = "gene_id") %>%
  left_join(si3d_norm_24hr %>% 
            rename(si3d_norm_24hr_rna_lfc = rna_lfc, si3d_norm_24hr_ribo_lfc = ribo_lfc, si3d_norm_24hr_te_lfc = te_lfc,
                   si3d_norm_24hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_1hr %>% 
            rename(si3d_hyp_1hr_rna_lfc = rna_lfc, si3d_hyp_1hr_ribo_lfc = ribo_lfc, si3d_hyp_1hr_te_lfc = te_lfc,
                   si3d_hyp_1hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_4hr %>% 
            rename(si3d_hyp_4hr_rna_lfc = rna_lfc, si3d_hyp_4hr_ribo_lfc = ribo_lfc, si3d_hyp_4hr_te_lfc = te_lfc,
                    si3d_hyp_4hr_category = category), by = "gene_id") %>%
  left_join(si3d_hyp_24hr %>% 
            rename(si3d_hyp_24hr_rna_lfc = rna_lfc, si3d_hyp_24hr_ribo_lfc = ribo_lfc, si3d_hyp_24hr_te_lfc = te_lfc,
                   si3d_hyp_24hr_category = category), by = "gene_id")

```

```{r add_si3e_data}
# Add eIF3e knockdown data (include LFC and TE padj only)
analysis_table <- analysis_table %>%
  left_join(si3e_norm_1hr %>% 
            rename(si3e_norm_1hr_rna_lfc = rna_lfc, si3e_norm_1hr_ribo_lfc = ribo_lfc, si3e_norm_1hr_te_lfc = te_lfc,
                  si3e_norm_1hr_category = category,
                  si3e_norm_1hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_norm_4hr %>% 
            rename(si3e_norm_4hr_rna_lfc = rna_lfc, si3e_norm_4hr_ribo_lfc = ribo_lfc, si3e_norm_4hr_te_lfc = te_lfc,
                   si3e_norm_4hr_category = category,
                   si3e_norm_4hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_norm_24hr %>% 
            rename(si3e_norm_24hr_rna_lfc = rna_lfc, si3e_norm_24hr_ribo_lfc = ribo_lfc, si3e_norm_24hr_te_lfc = te_lfc,
                  si3e_norm_24hr_category = category,
                  si3e_norm_24hr_padj = te_padj), by = "gene_id") %>%
  left_join(si3e_hyp_1hr %>% 
            rename(si3e_hyp_1hr_rna_lfc = rna_lfc, si3e_hyp_1hr_ribo_lfc = ribo_lfc, si3e_hyp_1hr_te_lfc = te_lfc,
                    si3e_hyp_1hr_category = category), by = "gene_id") %>%
  left_join(si3e_hyp_4hr %>% 
            rename(si3e_hyp_4hr_rna_lfc = rna_lfc, si3e_hyp_4hr_ribo_lfc = ribo_lfc, si3e_hyp_4hr_te_lfc = te_lfc,
                   si3e_hyp_4hr_category = category,
                   si3e_hyp_4hr_padj=te_padj), by = "gene_id") %>%
  left_join(si3e_hyp_24hr %>% 
            rename(si3e_hyp_24hr_rna_lfc = rna_lfc, si3e_hyp_24hr_ribo_lfc = ribo_lfc, si3e_hyp_24hr_te_lfc = te_lfc,
                   si3e_hyp_24hr_category = category,
                   si3e_hyp_24hr_padj=te_padj), by = "gene_id")

cat("Total columns in analysis table:", ncol(analysis_table), "\n")
```

## Add Interpretable Columns

```{r add_interpretable_columns}
# Add columns indicating TE regulation status
analysis_table <- analysis_table %>%
  mutate(
    # Hypoxia response TE direction
    hyp_1hr_te_dir = case_when(
      is.na(hyp_1hr_te_lfc) ~ "NA",
      hyp_1hr_te_lfc > 0 ~ "Up",
      hyp_1hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),
    hyp_4hr_te_dir = case_when(
      is.na(hyp_4hr_te_lfc) ~ "NA",
      hyp_4hr_te_lfc > 0 ~ "Up",
      hyp_4hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),
    hyp_24hr_te_dir = case_when(
      is.na(hyp_24hr_te_lfc) ~ "NA",
      hyp_24hr_te_lfc > 0 ~ "Up",
      hyp_24hr_te_lfc < 0 ~ "Down",
      TRUE ~ "No change"
    ),

    # eIF3d knockdown - TE not significant?
    si3d_norm_1hr_te_not_sig = is_te_not_significant(si3d_norm_1hr_category),
    si3d_norm_4hr_te_not_sig = is_te_not_significant(si3d_norm_4hr_category),
    si3d_norm_24hr_te_not_sig = is_te_not_significant(si3d_norm_24hr_category),
    si3d_hyp_1hr_te_not_sig = is_te_not_significant(si3d_hyp_1hr_category),
    si3d_hyp_4hr_te_not_sig = is_te_not_significant(si3d_hyp_4hr_category),
    si3d_hyp_24hr_te_not_sig = is_te_not_significant(si3d_hyp_24hr_category),

    # eIF3e knockdown - TE not significant?
    si3e_norm_1hr_te_not_sig = is_te_not_significant(si3e_norm_1hr_category),
    si3e_norm_4hr_te_not_sig = is_te_not_significant(si3e_norm_4hr_category),
    si3e_norm_24hr_te_not_sig = is_te_not_significant(si3e_norm_24hr_category),
    si3e_hyp_1hr_te_not_sig = is_te_not_significant(si3e_hyp_1hr_category),
    si3e_hyp_4hr_te_not_sig = is_te_not_significant(si3e_hyp_4hr_category),
    si3e_hyp_24hr_te_not_sig = is_te_not_significant(si3e_hyp_24hr_category)
  )

cat("Added interpretable columns\n")
```


genes_of_interest


```{r}
genes_of_interest <- analysis_table |> 
  filter(hyp_1hr_category=="Exclusive" & hyp_1hr_te_lfc>0) |> 
  filter(hyp_4hr_te_lfc > 0) |> 
    filter(hyp_24hr_te_lfc > 0)


```



# Effect of knockdown on hypoxic response

Read in deseq results

```{r}
#sictrl
hyp_response_deseq_1hr <-read_csv(here("output", "translation_categories_1hr.csv")) |> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
  rename(sictrl_1hr_te_lfc = te_lfc, sictrl_1hr_te_padj = te_padj)
hyp_response_deseq_4hr<-read_csv(here("output", "translation_categories_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(sictrl_4hr_te_lfc = te_lfc, sictrl_4hr_te_padj = te_padj)

hyp_response_deseq_24hr<-read_csv(here("output", "translation_categories_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(sictrl_24hr_te_lfc = te_lfc, sictrl_24hr_te_padj = te_padj)


#3e
hyp_response_deseq_1hr_3e_kd <- read_csv(here("output", "translation_categories_si3e_1hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
    rename(si3e_1hr_te_lfc = te_lfc, si3e_1hr_te_padj = te_padj)

hyp_response_deseq_4hr_3e_kd<- read_csv(here("output", "translation_categories_si3e_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3e_4hr_te_lfc = te_lfc, si3e_4hr_te_padj = te_padj)

hyp_response_deseq_24hr_3e_kd<- read_csv(here("output", "translation_categories_si3e_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3e_24hr_te_lfc = te_lfc, si3e_24hr_te_padj = te_padj)


#3d
hyp_response_deseq_1hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_1hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
      rename(si3d_1hr_te_lfc = te_lfc, si3d_1hr_te_padj = te_padj)

hyp_response_deseq_4hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_4hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
        rename(si3d_4hr_te_lfc = te_lfc, si3d_4hr_te_padj = te_padj)

hyp_response_deseq_24hr_3d_kd<- read_csv(here("output", "translation_categories_si3d_24hr.csv"))|> 
  dplyr::select(gene_id, te_lfc, te_padj) |> 
        rename(si3d_24hr_te_lfc = te_lfc, si3d_24hr_te_padj = te_padj)


all_hypoxic_response <-hyp_response_deseq_1hr |> 
  left_join(hyp_response_deseq_4hr) |> 
    left_join(hyp_response_deseq_24hr) |> 
    left_join(hyp_response_deseq_1hr_3e_kd) |> 
    left_join(hyp_response_deseq_4hr_3e_kd) |> 
    left_join(hyp_response_deseq_24hr_3e_kd) |> 
    left_join(hyp_response_deseq_1hr_3d_kd) |> 
    left_join(hyp_response_deseq_4hr_3d_kd) |> 
    left_join(hyp_response_deseq_24hr_3d_kd)

  

head(all_hypoxic_response)
```




# 1 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_1hr <- all_hypoxic_response %>%
  filter(sictrl_1hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_1hr %>%
  filter(si3e_1hr_te_lfc < 0) %>%
  filter(si3e_1hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_1hr %>%
  filter(si3e_1hr_te_lfc > 0) %>%
  filter(si3e_1hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_1hr_plot <- ggplot(data = sig_sictrl_1hr, aes(x = sictrl_1hr_te_lfc, y = si3e_1hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_1hr_plot <- ggMarginal(scatter_3e_1hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_1hr <- cowplot::plot_grid(scatter_3e_1hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_1hr

ggsave(filename= here("plots/si3e_1hr_response.pdf"), scatter_3e_1hr, height=4, width=4)



```




# 1 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_1hr <- all_hypoxic_response %>%
  filter(sictrl_1hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_1hr %>%
  filter(sictrl_1hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_1hr %>%
  filter(si3d_1hr_te_lfc < 0) %>%
  filter(si3d_1hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_1hr %>%
  filter(si3d_1hr_te_lfc > 0) %>%
  filter(si3d_1hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_1hr_plot <- ggplot(data = sig_sictrl_1hr, aes(x = sictrl_1hr_te_lfc, y = si3d_1hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_1hr_plot <- ggMarginal(scatter_3d_1hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_1hr <- cowplot::plot_grid(scatter_3d_1hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_1hr

ggsave(filename= here("plots/si3d_1hr_response.pdf"), scatter_3d_1hr, height=4, width=4)



```


# 4 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_4hr <- all_hypoxic_response %>%
  filter(sictrl_4hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_4hr %>%
  filter(si3e_4hr_te_lfc < 0) %>%
  filter(si3e_4hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_4hr %>%
  filter(si3e_4hr_te_lfc > 0) %>%
  filter(si3e_4hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_4hr_plot <- ggplot(data = sig_sictrl_4hr, aes(x = sictrl_4hr_te_lfc, y = si3e_4hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_4hr_plot <- ggMarginal(scatter_3e_4hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_4hr <- cowplot::plot_grid(scatter_3e_4hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_4hr

ggsave(filename= here("plots/si3e_4hr_response.pdf"), scatter_3e_4hr, height=4, width=4)



```




# 4 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_4hr <- all_hypoxic_response %>%
  filter(sictrl_4hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_4hr %>%
  filter(sictrl_4hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_4hr %>%
  filter(si3d_4hr_te_lfc < 0) %>%
  filter(si3d_4hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_4hr %>%
  filter(si3d_4hr_te_lfc > 0) %>%
  filter(si3d_4hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_4hr_plot <- ggplot(data = sig_sictrl_4hr, aes(x = sictrl_4hr_te_lfc, y = si3d_4hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_4hr_plot <- ggMarginal(scatter_3d_4hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_4hr <- cowplot::plot_grid(scatter_3d_4hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_4hr

ggsave(filename= here("plots/si3d_4hr_response.pdf"), scatter_3d_4hr, height=4, width=4)



```


# 24 hr - 3e KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_24hr <- all_hypoxic_response %>%
  filter(sictrl_24hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc > 0) %>%
  nrow()




#counts si3e
si3e_down <- sig_sictrl_24hr %>%
  filter(si3e_24hr_te_lfc < 0) %>%
  filter(si3e_24hr_te_padj < 0.05) %>%
  nrow()

si3e_up <- sig_sictrl_24hr %>%
  filter(si3e_24hr_te_lfc > 0) %>%
  filter(si3e_24hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3e` = c(si3e_up, si3e_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3e_24hr_plot <- ggplot(data = sig_sictrl_24hr, aes(x = sictrl_24hr_te_lfc, y = si3e_24hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3e LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3e_24hr_plot <- ggMarginal(scatter_3e_24hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#56B4E9")
           )

# Combine plot and table
scatter_3e_24hr <- cowplot::plot_grid(scatter_3e_24hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3e_24hr

ggsave(filename= here("plots/si3e_24hr_response.pdf"), scatter_3e_24hr, height=4, width=4)



```




# 24 hr - 3d KD

```{r}

colnames(all_hypoxic_response)

sig_sictrl_24hr <- all_hypoxic_response %>%
  filter(sictrl_24hr_te_padj <0.05)


#counts sig_sictrl
ctrl_down <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc < 0) %>%
  nrow()

ctrl_up <- sig_sictrl_24hr %>%
  filter(sictrl_24hr_te_lfc > 0) %>%
  nrow()




#counts si3d
si3d_down <- sig_sictrl_24hr %>%
  filter(si3d_24hr_te_lfc < 0) %>%
  filter(si3d_24hr_te_padj < 0.05) %>%
  nrow()

si3d_up <- sig_sictrl_24hr %>%
  filter(si3d_24hr_te_lfc > 0) %>%
  filter(si3d_24hr_te_padj < 0.05) %>%
  nrow()


# Create summary table (transposed with up/down as rows)
summary_table <- data.frame(
  Direction = c("Up", "Down"),
  `si-CTRL` = c(ctrl_up, ctrl_down),
  `si-eIF3d` = c(si3d_up, si3d_down),
  check.names = FALSE
)

# Create table grob
table_grob <- gridExtra::tableGrob(summary_table, rows = NULL, theme = gridExtra::ttheme_minimal(base_size = 9))

scatter_3d_24hr_plot <- ggplot(data = sig_sictrl_24hr, aes(x = sictrl_24hr_te_lfc, y = si3d_24hr_te_lfc)) +
  ylim(-2,2) +
  xlim(-2,2) +
  geom_hline(yintercept = 0, color = "grey") +
  geom_vline(xintercept = 0, color = "grey") +
  geom_point(alpha=.25) +
  ylab(bquote(Log[2]~"si-eIF3d LFC"~"TE: hypoxia vs normoxia")) +
  xlab(bquote(Log[2]~"si-CTRL LFC"~"TE: hypoxia vs normoxia")) +
    # theme_cowplot(font_size = 9)+
  theme_prism()

scatter_3d_24hr_plot <- ggMarginal(scatter_3d_24hr_plot, type = "histogram",
           xparams = list(binwidth = .1, fill = "grey"),
           yparams = list(binwidth = .1, fill = "#E69F00")
           )

# Combine plot and table
scatter_3d_24hr <- cowplot::plot_grid(scatter_3d_24hr_plot, table_grob, ncol = 1, rel_heights = c(1, 0.15))

scatter_3d_24hr

ggsave(filename= here("plots/si3d_24hr_response.pdf"), scatter_3d_24hr, height=4, width=4)



```
