---
title: "Spike-In Normalization Comparison"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(DESeq2)
library(here)
library(cowplot)
library(ggprism)
```

# Source Functions

```{r source_functions}
source(here("code", "functions.R"))
```

# Load Data

```{r load_data}
# Load human gene counts
allCounts <- read_csv(here("output", "human_allCounts_raw.csv"), show_col_types = FALSE) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("Human gene counts loaded:", nrow(allCounts), "genes x", ncol(allCounts), "samples\n")

# Load spike-in counts
spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv"), show_col_types = FALSE) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("Spike-in counts loaded:", nrow(spikein), "genes x", ncol(spikein), "samples\n")

# Load metadata
metadata_all <- read_csv(here("accessories", "metadata_231_timecourse.csv"), show_col_types = FALSE)

cat("Metadata loaded:", nrow(metadata_all), "samples\n")

# Load gene annotations
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"),
                      show_col_types = FALSE)

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotations loaded:", nrow(gene_symbols), "unique genes\n")

# Define top 500 spike-in genes
top500_ids <- spikein %>%
  as.data.frame() %>%
  mutate(total = rowSums(.)) %>%
  arrange(desc(total)) %>%
  head(500) %>%
  rownames()

cat("Top 500 spike-in genes selected\n")
```

# Run Nospike Analyses

This section runs differential expression analyses for RNA, RIBO, and TE modes WITHOUT spike-in normalization, for each of the three timepoints (1hr, 4hr, 24hr).

## 1hr Timepoint

```{r run_nospike_1hr}
cat("\n=== Processing 1hr timepoint ===\n")

# Filter to 1hr siCtrl samples
metadata_1hr <- metadata_all %>%
  filter(hours == 1, treatment == "siCTRL") %>%
  mutate(
    library = case_when(
      type == "Input_Ribo_RNAseq" ~ "rna",
      type == "RiboSeq" ~ "ribo"
    ),
    sample_name = id,
    comparison_variable = condition
  )

cat("1hr samples:", nrow(metadata_1hr), "\n")

allCounts_1hr <- allCounts[, metadata_1hr$sample_name]
spikein_1hr <- spikein[, metadata_1hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_1hr_nospike <- createNormMatrix(
  allCounts = allCounts_1hr,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## 4hr Timepoint

```{r run_nospike_4hr}
cat("\n=== Processing 4hr timepoint ===\n")

# Filter to 4hr siCtrl samples
metadata_4hr <- metadata_all %>%
  filter(hours == 4, treatment == "siCTRL") %>%
  mutate(
    library = case_when(
      type == "Input_Ribo_RNAseq" ~ "rna",
      type == "RiboSeq" ~ "ribo"
    ),
    sample_name = id,
    comparison_variable = condition
  )

cat("4hr samples:", nrow(metadata_4hr), "\n")

allCounts_4hr <- allCounts[, metadata_4hr$sample_name]
spikein_4hr <- spikein[, metadata_4hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_4hr_nospike <- createNormMatrix(
  allCounts = allCounts_4hr,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## 24hr Timepoint

```{r run_nospike_24hr}
cat("\n=== Processing 24hr timepoint ===\n")

# Filter to 24hr siCtrl samples
metadata_24hr <- metadata_all %>%
  filter(hours == 24, treatment == "siCTRL") %>%
  mutate(
    library = case_when(
      type == "Input_Ribo_RNAseq" ~ "rna",
      type == "RiboSeq" ~ "ribo"
    ),
    sample_name = id,
    comparison_variable = condition
  )

cat("24hr samples:", nrow(metadata_24hr), "\n")

allCounts_24hr <- allCounts[, metadata_24hr$sample_name]
spikein_24hr <- spikein[, metadata_24hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_24hr_nospike <- createNormMatrix(
  allCounts = allCounts_24hr,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

# Load POST Spike-in Data

Load the spike-in normalized results that were previously generated by `rna_ribo_te_analysis.Rmd`.

```{r load_spike_data}
cat("\n=== Loading POST spike-in data ===\n")

spike_1hr <- read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE)
spike_4hr <- read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE)
spike_24hr <- read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE)

cat("Loaded POST spike-in data:\n")
cat("  1hr:", nrow(spike_1hr), "genes\n")
cat("  4hr:", nrow(spike_4hr), "genes\n")
cat("  24hr:", nrow(spike_24hr), "genes\n")
```

# Prepare Combined Datasets

Combine PRE spike-in (nospike) and POST spike-in data for each metric (RNA, Ribo, TE).

```{r prepare_combined_data}
cat("\n=== Preparing combined datasets ===\n")

# Combine RNA data
rna_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_rna_nospike_1hr$gene_id,
    lfc = res_rna_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_rna_nospike_4hr$gene_id,
    lfc = res_rna_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_rna_nospike_24hr$gene_id,
    lfc = res_rna_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    rename(lfc = rna_lfc),
  spike_4hr %>%
    select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    rename(lfc = rna_lfc),
  spike_24hr %>%
    select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    rename(lfc = rna_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

# Combine RIBO data
ribo_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_ribo_nospike_1hr$gene_id,
    lfc = res_ribo_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_ribo_nospike_4hr$gene_id,
    lfc = res_ribo_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_ribo_nospike_24hr$gene_id,
    lfc = res_ribo_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    rename(lfc = ribo_lfc),
  spike_4hr %>%
    select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    rename(lfc = ribo_lfc),
  spike_24hr %>%
    select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    rename(lfc = ribo_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

# Combine TE data
te_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_te_nospike_1hr$gene_id,
    lfc = res_te_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_te_nospike_4hr$gene_id,
    lfc = res_te_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_te_nospike_24hr$gene_id,
    lfc = res_te_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    rename(lfc = te_lfc),
  spike_4hr %>%
    select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    rename(lfc = te_lfc),
  spike_24hr %>%
    select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    rename(lfc = te_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

cat("Combined LFC data prepared:\n")
cat("  RNA LFC:", nrow(rna_lfc_combined), "observations\n")
cat("  Ribo LFC:", nrow(ribo_lfc_combined), "observations\n")
cat("  TE LFC:", nrow(te_lfc_combined), "observations\n")
```

# Create Density Plots

## RNA log2FoldChange Distribution

```{r rna_density_plot, fig.width=12, fig.height=5}
p_rna <- ggplot(rna_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    title = "RNA Expression Changes: Spike-In Normalization Comparison",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (RNA)",
    y = "Density",
    fill = "Normalization"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  xlim(-4, 4) +
  facet_wrap(~timepoint, ncol = 3, scales = "free_y") +
  theme_prism() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "top"
  )

print(p_rna)

ggsave(
  here("plots", "rna_lfc_spike_comparison.pdf"),
  plot = p_rna,
  width = 12,
  height = 5
)

cat("\nRNA density plot saved to plots/rna_lfc_spike_comparison.pdf\n")
```

## Ribosome Occupancy log2FoldChange Distribution

```{r ribo_density_plot, fig.width=12, fig.height=5}
p_ribo <- ggplot(ribo_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    title = "Ribosome Occupancy Changes: Spike-In Normalization Comparison",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (Ribo)",
    y = "Density",
    fill = "Normalization"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  xlim(-4, 4) +
  facet_wrap(~timepoint, ncol = 3, scales = "free_y") +
  theme_prism() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "top"
  )

print(p_ribo)

ggsave(
  here("plots", "ribo_lfc_spike_comparison.pdf"),
  plot = p_ribo,
  width = 12,
  height = 5
)

cat("\nRibo density plot saved to plots/ribo_lfc_spike_comparison.pdf\n")
```

## Translation Efficiency log2FoldChange Distribution

```{r te_density_plot, fig.width=12, fig.height=5}
p_te <- ggplot(te_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    title = "Translation Efficiency Changes: Spike-In Normalization Comparison",
    subtitle = "Distribution of log2FoldChange (Hypoxia vs Normoxia)",
    x = "log2 Fold Change (TE)",
    y = "Density",
    fill = "Normalization"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  xlim(-4, 4) +
  facet_wrap(~timepoint, ncol = 3, scales = "free_y") +
  theme_prism() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "top"
  )

print(p_te)

ggsave(
  here("plots", "te_lfc_spike_comparison.pdf"),
  plot = p_te,
  width = 12,
  height = 5
)

cat("\nTE density plot saved to plots/te_lfc_spike_comparison.pdf\n")
```

# Summary

This analysis compares the effect of spike-in normalization on RNA expression, ribosome occupancy, and translation efficiency measurements across three timepoints (1hr, 4hr, 24hr) of hypoxia exposure.

**Key findings:**
- PRE spike-in (nospike): Standard DESeq2 normalization
- POST spike-in (spike): Normalization using top 500 Drosophila spike-in genes

**Plots generated:**
1. `rna_lfc_spike_comparison.pdf` - RNA expression changes
2. `ribo_lfc_spike_comparison.pdf` - Ribosome occupancy changes
3. `te_lfc_spike_comparison.pdf` - Translation efficiency changes

Each plot shows density distributions faceted by timepoint, with pre vs post spike-in normalization overlaid for direct comparison.
