---
title: "Spike-In Normalization Comparison"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(DESeq2)
library(here)
library(cowplot)
library(ggprism)
library(dplyr)  # reload after DESeq2 to prevent AnnotationDbi masking dplyr::select
```

# Source Functions

```{r source_functions}
source(here("code", "functions.R"))
```

# Load Data

```{r load_data}
# Load human gene counts
allCounts <- read_csv(here("output", "human_allCounts_raw.csv"), show_col_types = FALSE) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("Human gene counts loaded:", nrow(allCounts), "genes x", ncol(allCounts), "samples\n")

# Load spike-in counts
spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv"), show_col_types = FALSE) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("Spike-in counts loaded:", nrow(spikein), "genes x", ncol(spikein), "samples\n")

# Load metadata
metadata_all <- read_csv(here("accessories", "metadata_231_timecourse.csv"), show_col_types = FALSE)

cat("Metadata loaded:", nrow(metadata_all), "samples\n")

# Preprocess metadata to match column names in allCounts
metadata_all <- metadata_all %>%
  mutate(
    treatment = recode(treatment, "siCTRL" = "sictrl", "si3e" = "si3e", "si3d" = "si3d"),
    rep = recode(rep, RepA = "a", RepB = "b", RepC = "c"),
    library = case_when(
      type == "Input_Ribo_RNAseq" ~ "rna",
      type == "RiboSeq" ~ "ribo"
    ),
    sample_name = paste(condition, treatment, hours, rep, library, id, sep = "_"),
    comparison_variable = condition
  )

cat("Metadata preprocessed with sample_name column\n")

# Load gene annotations
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"),
                      show_col_types = FALSE)

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotations loaded:", nrow(gene_symbols), "unique genes\n")

# Define top 500 spike-in genes
top500_ids <- spikein %>%
  as.data.frame() %>%
  mutate(total = rowSums(.)) %>%
  arrange(desc(total)) %>%
  head(500) %>%
  rownames()

cat("Top 500 spike-in genes selected\n")
```

# Run Nospike Analyses

This section runs differential expression analyses for RNA, RIBO, and TE modes WITHOUT spike-in normalization, for each of the three timepoints (1hr, 4hr, 24hr).

## 1hr Timepoint

```{r run_nospike_1hr}
cat("\n=== Processing 1hr timepoint ===\n")

# Filter to 1hr siCtrl samples
metadata_1hr <- metadata_all %>%
  filter(hours == 1, treatment == "sictrl")

cat("1hr samples:", nrow(metadata_1hr), "\n")

allCounts_1hr <- allCounts[, metadata_1hr$sample_name]
spikein_1hr <- spikein[, metadata_1hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_1hr_nospike <- createNormMatrix(
  allCounts = allCounts_1hr,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## 4hr Timepoint

```{r run_nospike_4hr}
cat("\n=== Processing 4hr timepoint ===\n")

# Filter to 4hr siCtrl samples
metadata_4hr <- metadata_all %>%
  filter(hours == 4, treatment == "sictrl")

cat("4hr samples:", nrow(metadata_4hr), "\n")

allCounts_4hr <- allCounts[, metadata_4hr$sample_name]
spikein_4hr <- spikein[, metadata_4hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_4hr_nospike <- createNormMatrix(
  allCounts = allCounts_4hr,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")
cat("  TE (nospike) up   (LFC>0.5,  padj<0.05):",
    sum(res_te_nospike_4hr$log2FoldChange > 0.5 & res_te_nospike_4hr$padj < 0.05, na.rm=TRUE), "\n")
cat("  TE (nospike) down (LFC<-0.5, padj<0.05):",
    sum(res_te_nospike_4hr$log2FoldChange < -0.5 & res_te_nospike_4hr$padj < 0.05, na.rm=TRUE), "\n")
```

## 24hr Timepoint

```{r run_nospike_24hr}
cat("\n=== Processing 24hr timepoint ===\n")

# Filter to 24hr siCtrl samples
metadata_24hr <- metadata_all %>%
  filter(hours == 24, treatment == "sictrl")

cat("24hr samples:", nrow(metadata_24hr), "\n")

allCounts_24hr <- allCounts[, metadata_24hr$sample_name]
spikein_24hr <- spikein[, metadata_24hr$sample_name]

# Create nospike normalization matrix
allCountsNorm_24hr_nospike <- createNormMatrix(
  allCounts = allCounts_24hr,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  spikein = NULL,
  keepSpikes = NULL
)

cat("Normalization matrix created (nospike)\n")

# Run deFunction for RNA nospike
cat("\nRunning RNA analysis (nospike)...\n")
res_rna_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for RIBO nospike
cat("\nRunning RIBO analysis (nospike)...\n")
res_ribo_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "RIBO"
)

cat("RIBO analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")

# Run deFunction for TE nospike
cat("\nRunning TE analysis (nospike)...\n")
res_te_nospike_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

# Load POST Spike-in Data

Load the spike-in normalized results that were previously generated by `rna_ribo_te_analysis.Rmd`.

```{r load_spike_data}
cat("\n=== Loading POST spike-in data ===\n")

spike_1hr <- read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE)
spike_4hr <- read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE)
spike_24hr <- read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE)

cat("Loaded POST spike-in data:\n")
cat("  1hr:", nrow(spike_1hr), "genes\n")
cat("  4hr:", nrow(spike_4hr), "genes\n")
cat("  24hr:", nrow(spike_24hr), "genes\n")
```

# Sanity Check: Verify Nospike Results Match Original

Load the original nospike TE results from `timecourse_qc.Rmd` and verify they match our newly generated results.

```{r sanity_check_nospike}
cat("\n=== Sanity Check: Comparing nospike TE results ===\n")

# Load original nospike results from timecourse_qc.Rmd
original_1hr <- read_csv(here("output", "deseq_results_1hr_nospike.csv"), show_col_types = FALSE)
original_4hr <- read_csv(here("output", "deseq_results_4hr_nospike.csv"), show_col_types = FALSE)
original_24hr <- read_csv(here("output", "deseq_results_24hr_nospike.csv"), show_col_types = FALSE)

# Compare 1hr results
cat("\n--- 1hr Timepoint ---\n")
cat("Original dimensions:", dim(original_1hr), "\n")
cat("New dimensions:", dim(res_te_nospike_1hr), "\n")
cat("Original sig genes (padj < 0.05):", sum(original_1hr$padj < 0.05, na.rm = TRUE), "\n")
cat("New sig genes (padj < 0.05):", sum(res_te_nospike_1hr$padj < 0.05, na.rm = TRUE), "\n")

# Check if gene IDs match
genes_match_1hr <- all.equal(original_1hr$gene_id, res_te_nospike_1hr$gene_id)
cat("Gene IDs match:", isTRUE(genes_match_1hr), "\n")

# Check LFC correlation
lfc_cor_1hr <- cor(original_1hr$log2FoldChange, res_te_nospike_1hr$log2FoldChange, use = "complete.obs")
cat("LFC correlation:", round(lfc_cor_1hr, 6), "\n")

# Check if values are identical
lfc_identical_1hr <- all.equal(original_1hr$log2FoldChange, res_te_nospike_1hr$log2FoldChange, tolerance = 1e-10)
padj_identical_1hr <- all.equal(original_1hr$padj, res_te_nospike_1hr$padj, tolerance = 1e-10)
cat("LFC values identical:", isTRUE(lfc_identical_1hr), "\n")
cat("padj values identical:", isTRUE(padj_identical_1hr), "\n")

# Compare 4hr results
cat("\n--- 4hr Timepoint ---\n")
cat("Original dimensions:", dim(original_4hr), "\n")
cat("New dimensions:", dim(res_te_nospike_4hr), "\n")
cat("Original sig genes (padj < 0.05):", sum(original_4hr$padj < 0.05, na.rm = TRUE), "\n")
cat("New sig genes (padj < 0.05):", sum(res_te_nospike_4hr$padj < 0.05, na.rm = TRUE), "\n")

genes_match_4hr <- all.equal(original_4hr$gene_id, res_te_nospike_4hr$gene_id)
cat("Gene IDs match:", isTRUE(genes_match_4hr), "\n")

lfc_cor_4hr <- cor(original_4hr$log2FoldChange, res_te_nospike_4hr$log2FoldChange, use = "complete.obs")
cat("LFC correlation:", round(lfc_cor_4hr, 6), "\n")

lfc_identical_4hr <- all.equal(original_4hr$log2FoldChange, res_te_nospike_4hr$log2FoldChange, tolerance = 1e-10)
padj_identical_4hr <- all.equal(original_4hr$padj, res_te_nospike_4hr$padj, tolerance = 1e-10)
cat("LFC values identical:", isTRUE(lfc_identical_4hr), "\n")
cat("padj values identical:", isTRUE(padj_identical_4hr), "\n")

# Compare 24hr results
cat("\n--- 24hr Timepoint ---\n")
cat("Original dimensions:", dim(original_24hr), "\n")
cat("New dimensions:", dim(res_te_nospike_24hr), "\n")
cat("Original sig genes (padj < 0.05):", sum(original_24hr$padj < 0.05, na.rm = TRUE), "\n")
cat("New sig genes (padj < 0.05):", sum(res_te_nospike_24hr$padj < 0.05, na.rm = TRUE), "\n")

genes_match_24hr <- all.equal(original_24hr$gene_id, res_te_nospike_24hr$gene_id)
cat("Gene IDs match:", isTRUE(genes_match_24hr), "\n")

lfc_cor_24hr <- cor(original_24hr$log2FoldChange, res_te_nospike_24hr$log2FoldChange, use = "complete.obs")
cat("LFC correlation:", round(lfc_cor_24hr, 6), "\n")

lfc_identical_24hr <- all.equal(original_24hr$log2FoldChange, res_te_nospike_24hr$log2FoldChange, tolerance = 1e-10)
padj_identical_24hr <- all.equal(original_24hr$padj, res_te_nospike_24hr$padj, tolerance = 1e-10)
cat("LFC values identical:", isTRUE(lfc_identical_24hr), "\n")
cat("padj values identical:", isTRUE(padj_identical_24hr), "\n")

# Final verification status
all_checks_pass <- isTRUE(genes_match_1hr) && isTRUE(lfc_identical_1hr) && isTRUE(padj_identical_1hr) &&
                   isTRUE(genes_match_4hr) && isTRUE(lfc_identical_4hr) && isTRUE(padj_identical_4hr) &&
                   isTRUE(genes_match_24hr) && isTRUE(lfc_identical_24hr) && isTRUE(padj_identical_24hr)

if (all_checks_pass) {
  cat("\n✓ SANITY CHECK PASSED: All nospike results match original analysis\n")
} else {
  cat("\n✗ SANITY CHECK WARNING: Some differences detected in nospike results\n")
}
```

# Prepare Combined Datasets

Combine PRE spike-in (nospike) and POST spike-in data for each metric (RNA, Ribo, TE).

```{r prepare_combined_data}
cat("\n=== Preparing combined datasets ===\n")

# Combine RNA data
rna_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_rna_nospike_1hr$gene_id,
    lfc = res_rna_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_rna_nospike_4hr$gene_id,
    lfc = res_rna_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_rna_nospike_24hr$gene_id,
    lfc = res_rna_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    dplyr::select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    dplyr::rename(lfc = rna_lfc),
  spike_4hr %>%
    dplyr::select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    dplyr::rename(lfc = rna_lfc),
  spike_24hr %>%
    dplyr::select(gene_id, rna_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    dplyr::rename(lfc = rna_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

# Combine RIBO data
ribo_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_ribo_nospike_1hr$gene_id,
    lfc = res_ribo_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_ribo_nospike_4hr$gene_id,
    lfc = res_ribo_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_ribo_nospike_24hr$gene_id,
    lfc = res_ribo_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    dplyr::select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    dplyr::rename(lfc = ribo_lfc),
  spike_4hr %>%
    dplyr::select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    dplyr::rename(lfc = ribo_lfc),
  spike_24hr %>%
    dplyr::select(gene_id, ribo_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    dplyr::rename(lfc = ribo_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

# Combine TE data
te_lfc_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = res_te_nospike_1hr$gene_id,
    lfc = res_te_nospike_1hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = res_te_nospike_4hr$gene_id,
    lfc = res_te_nospike_4hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = res_te_nospike_24hr$gene_id,
    lfc = res_te_nospike_24hr$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in (from CSV files)
  spike_1hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    dplyr::rename(lfc = te_lfc),
  spike_4hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    dplyr::rename(lfc = te_lfc),
  spike_24hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    dplyr::rename(lfc = te_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

cat("Combined LFC data prepared:\n")
cat("  RNA LFC:", nrow(rna_lfc_combined), "observations\n")
cat("  Ribo LFC:", nrow(ribo_lfc_combined), "observations\n")
cat("  TE LFC:", nrow(te_lfc_combined), "observations\n")
```

# Create Density Plots

## RNA log2FoldChange Distribution

```{r rna_density_plot, fig.width=9, fig.height=2.5}
library(ggh4x)

p_rna <- ggplot(rna_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    x = "RNA log2FC",
    y = "Density",
    fill = NULL
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~timepoint, ncol = 3, scales = "free") +
  facetted_pos_scales(
    x = list(
      timepoint == "1hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "4hr" ~ scale_x_continuous(limits = c(-4, 4)),
      timepoint == "24hr" ~ scale_x_continuous(limits = c(-4, 4))
    )
  ) +
  theme_prism() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "top",
    legend.text = element_text(size = 9),
    legend.margin = margin(b = -5),
    plot.margin = margin(t = 5, r = 5, b = 2, l = 5)
  )

print(p_rna)

ggsave(
  here("plots", "rna_lfc_spike_comparison.pdf"),
  plot = p_rna,
  width = 5,
  height = 2
)

cat("\nRNA density plot saved to plots/rna_lfc_spike_comparison.pdf\n")
```

## Ribosome Occupancy log2FoldChange Distribution

```{r ribo_density_plot, fig.width=9, fig.height=2.5}
p_ribo <- ggplot(ribo_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    x = "Ribo log2FC",
    y = "Density",
    fill = NULL
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~timepoint, ncol = 3, scales = "free") +
  facetted_pos_scales(
    x = list(
      timepoint == "1hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "4hr" ~ scale_x_continuous(limits = c(-4, 4)),
      timepoint == "24hr" ~ scale_x_continuous(limits = c(-4, 4))
    )
  ) +
  theme_prism() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "none",
    plot.margin = margin(t = 2, r = 5, b = 2, l = 5)
  )

print(p_ribo)

ggsave(
  here("plots", "ribo_lfc_spike_comparison.pdf"),
  plot = p_ribo,
  width = 7,
  height = 4
)

cat("\nRibo density plot saved to plots/ribo_lfc_spike_comparison.pdf\n")
```

## Translation Efficiency log2FoldChange Distribution

```{r te_density_plot, fig.width=9, fig.height=2.5}
p_te <- ggplot(te_lfc_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#0072B2",   # Okabe Ito dark blue (index 5)
      "Post spike-in" = "#009E73"   # Okabe Ito bluish green (index 3)
    )
  ) +
  labs(
    x = "TE log2FC",
    y = "Density",
    fill = NULL
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~timepoint, ncol = 3, scales = "free") +
  facetted_pos_scales(
    x = list(
      timepoint == "1hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "4hr" ~ scale_x_continuous(limits = c(-4, 4)),
      timepoint == "24hr" ~ scale_x_continuous(limits = c(-4, 4))
    )
  ) +
  theme_prism() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "grey90"),
    legend.position = "none",
    plot.margin = margin(t = 2, r = 5, b = 5, l = 5)
  )

print(p_te)

ggsave(
  here("plots", "te_lfc_spike_comparison.pdf"),
  plot = p_te,
  width = 5,
  height = 1.8
)

cat("\nTE density plot saved to plots/te_lfc_spike_comparison.pdf\n")
```

# Summary

This analysis compares the effect of spike-in normalization on RNA expression, ribosome occupancy, and translation efficiency measurements across three timepoints (1hr, 4hr, 24hr) of hypoxia exposure.

**Key findings:**
- PRE spike-in (nospike): Standard DESeq2 normalization
- POST spike-in (spike): Normalization using top 500 Drosophila spike-in genes

**Plots generated:**
1. `rna_lfc_spike_comparison.pdf` - RNA expression changes
2. `ribo_lfc_spike_comparison.pdf` - Ribosome occupancy changes
3. `te_lfc_spike_comparison.pdf` - Translation efficiency changes

Each plot shows density distributions faceted by timepoint, with pre vs post spike-in normalization overlaid for direct comparison.

# Effect of eIF3e and eIF3d on General Translation (Normoxia)

This section examines the effect of eIF3e and eIF3d on global translation efficiency in normoxia conditions, comparing spike-in vs no spike-in normalization.

## Run No Spike-in Analyses for eIF3e and eIF3d Comparisons

### Load Clean Metadata

```{r load_clean_metadata}
# Use metadata_clean.csv which matches the allCounts column names
metadata_clean <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE)
cat("Loaded clean metadata:", nrow(metadata_clean), "samples\n")
```

### eIF3e (si3e vs siCTRL) - No Spike-in

```{r run_nospike_si3e_normoxia}
cat("\n=== Running si3e vs siCTRL TE analysis (no spike-in, normoxia) ===\n")

# Process each timepoint
si3e_nospike_results <- list()

for (tp in c(1, 4, 24)) {
  cat("\n--- Processing", tp, "hr timepoint ---\n")

  # Filter to normoxia samples with si3e and sictrl (using clean metadata)
  metadata_tp <- metadata_clean %>%
    filter(hours == tp, condition == "normoxia", treatment %in% c("sictrl", "si3e"))

  cat("Samples:", nrow(metadata_tp), "\n")

  allCounts_tp <- allCounts[, metadata_tp$sample_name]

  # Create nospike normalization matrix
  metadata_tp_mod <- metadata_tp %>%
    mutate(comparison_variable = treatment)

  allCountsNorm_tp_nospike <- createNormMatrix(
    allCounts = allCounts_tp,
    metadata = metadata_tp_mod,
    control_condition = "sictrl",
    spikein = NULL,
    keepSpikes = NULL
  )

  # Run TE analysis
  res_te <- deFunction(
    countsMatrix = allCounts_tp,
    allCounts_Norm = allCountsNorm_tp_nospike,
    metadata = metadata_tp_mod,
    control_condition = "sictrl",
    test_condition = "si3e",
    spikeInMatrix = NULL,
    keepSpikes = NULL,
    mode = "TE"
  )

  si3e_nospike_results[[paste0(tp, "hr")]] <- res_te
  cat("TE analysis complete. Significant genes (padj < 0.05):",
      sum(res_te$padj < 0.05, na.rm = TRUE), "\n")
}
```

### eIF3d (si3d vs siCTRL) - No Spike-in

```{r run_nospike_si3d_normoxia}
cat("\n=== Running si3d vs siCTRL TE analysis (no spike-in, normoxia) ===\n")

si3d_nospike_results <- list()

for (tp in c(1, 4, 24)) {
  cat("\n--- Processing", tp, "hr timepoint ---\n")

  # Filter to normoxia samples with si3d and sictrl (using clean metadata)
  metadata_tp <- metadata_clean %>%
    filter(hours == tp, condition == "normoxia", treatment %in% c("sictrl", "si3d"))

  cat("Samples:", nrow(metadata_tp), "\n")

  allCounts_tp <- allCounts[, metadata_tp$sample_name]

  # Create nospike normalization matrix
  metadata_tp_mod <- metadata_tp %>%
    mutate(comparison_variable = treatment)

  allCountsNorm_tp_nospike <- createNormMatrix(
    allCounts = allCounts_tp,
    metadata = metadata_tp_mod,
    control_condition = "sictrl",
    spikein = NULL,
    keepSpikes = NULL
  )

  # Run TE analysis
  res_te <- deFunction(
    countsMatrix = allCounts_tp,
    allCounts_Norm = allCountsNorm_tp_nospike,
    metadata = metadata_tp_mod,
    control_condition = "sictrl",
    test_condition = "si3d",
    spikeInMatrix = NULL,
    keepSpikes = NULL,
    mode = "TE"
  )

  si3d_nospike_results[[paste0(tp, "hr")]] <- res_te
  cat("TE analysis complete. Significant genes (padj < 0.05):",
      sum(res_te$padj < 0.05, na.rm = TRUE), "\n")
}
```

## Load Spike-in Normalized Data

```{r load_spikein_kd_data}
cat("\n=== Loading spike-in normalized eIF3e and eIF3d data (normoxia) ===\n")

# Load si3e vs siCTRL normoxia data (spike-in normalized)
si3e_spike_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"), show_col_types = FALSE)
si3e_spike_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"), show_col_types = FALSE)
si3e_spike_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"), show_col_types = FALSE)

cat("Loaded si3e vs siCTRL normoxia data (spike-in):\n")
cat("  1hr:", nrow(si3e_spike_1hr), "genes\n")
cat("  4hr:", nrow(si3e_spike_4hr), "genes\n")
cat("  24hr:", nrow(si3e_spike_24hr), "genes\n")

# Load si3d vs siCTRL normoxia data (spike-in normalized)
si3d_spike_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"), show_col_types = FALSE)
si3d_spike_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"), show_col_types = FALSE)
si3d_spike_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"), show_col_types = FALSE)

cat("\nLoaded si3d vs siCTRL normoxia data (spike-in):\n")
cat("  1hr:", nrow(si3d_spike_1hr), "genes\n")
cat("  4hr:", nrow(si3d_spike_4hr), "genes\n")
cat("  24hr:", nrow(si3d_spike_24hr), "genes\n")
```

## Prepare Combined Data for Density Plots

```{r prepare_kd_combined_data}
# Combine si3e TE data (pre vs post spike-in)
si3e_te_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = si3e_nospike_results[["1hr"]]$gene_id,
    lfc = si3e_nospike_results[["1hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = si3e_nospike_results[["4hr"]]$gene_id,
    lfc = si3e_nospike_results[["4hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = si3e_nospike_results[["24hr"]]$gene_id,
    lfc = si3e_nospike_results[["24hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in
  si3e_spike_1hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    dplyr::rename(lfc = te_lfc),
  si3e_spike_4hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    dplyr::rename(lfc = te_lfc),
  si3e_spike_24hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    dplyr::rename(lfc = te_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

# Combine si3d TE data (pre vs post spike-in)
si3d_te_combined <- bind_rows(
  # Pre spike-in (nospike)
  data.frame(
    gene_id = si3d_nospike_results[["1hr"]]$gene_id,
    lfc = si3d_nospike_results[["1hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "1hr"
  ),
  data.frame(
    gene_id = si3d_nospike_results[["4hr"]]$gene_id,
    lfc = si3d_nospike_results[["4hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "4hr"
  ),
  data.frame(
    gene_id = si3d_nospike_results[["24hr"]]$gene_id,
    lfc = si3d_nospike_results[["24hr"]]$log2FoldChange,
    normalization = "Pre spike-in",
    timepoint = "24hr"
  ),
  # Post spike-in
  si3d_spike_1hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "1hr") %>%
    dplyr::rename(lfc = te_lfc),
  si3d_spike_4hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "4hr") %>%
    dplyr::rename(lfc = te_lfc),
  si3d_spike_24hr %>%
    dplyr::select(gene_id, te_lfc) %>%
    mutate(normalization = "Post spike-in", timepoint = "24hr") %>%
    dplyr::rename(lfc = te_lfc)
) %>%
  filter(!is.na(lfc)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in"))
  )

cat("Combined TE data prepared:\n")
cat("  si3e TE:", nrow(si3e_te_combined), "observations\n")
cat("  si3d TE:", nrow(si3d_te_combined), "observations\n")
```

## Effect of eIF3e on TE (Normoxia)

```{r si3e_te_density_plot, fig.width=9, fig.height=2.5}
p_si3e_te <- ggplot(si3e_te_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#366885",
      "Post spike-in" = "#56B4E9"
    )
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~timepoint, ncol = 3, scales = "free") +
  facetted_pos_scales(
    x = list(
      timepoint == "1hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "4hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "24hr" ~ scale_x_continuous(limits = c(-2, 2))
    )
  ) +
  labs(
    title = "Effect of eIF3e on Translation Efficiency (Normoxia)",
    x = "TE log2FC",
    y = "Density",
    fill = NULL
  ) +
  theme_prism() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 9),
    legend.margin = margin(b = -5),
    plot.margin = margin(t = 5, r = 10, b = 5, l = 5)
  )

print(p_si3e_te)

ggsave(
  here("plots", "si3e_te_spike_comparison_normoxia.pdf"),
  plot = p_si3e_te,
  width = 6,
  height = 3
)

cat("\neIF3e TE density plot saved to plots/si3e_te_spike_comparison_normoxia.pdf\n")
```

## Effect of eIF3d on TE (Normoxia)

```{r si3d_te_density_plot, fig.width=9, fig.height=2.5}
p_si3d_te <- ggplot(si3d_te_combined, aes(x = lfc, fill = normalization)) +
  geom_density(alpha = 0.6, size = 0.8) +
  scale_fill_manual(
    values = c(
      "Pre spike-in" = "#87610e",
      "Post spike-in" = "#E69F00"
    )
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  facet_wrap(~timepoint, ncol = 3, scales = "free") +
  facetted_pos_scales(
    x = list(
      timepoint == "1hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "4hr" ~ scale_x_continuous(limits = c(-2, 2)),
      timepoint == "24hr" ~ scale_x_continuous(limits = c(-2, 2))
    )
  ) +
  labs(
    title = "Effect of eIF3d on Translation Efficiency (Normoxia)",
    x = "TE log2FC",
    y = "Density",
    fill = NULL
  ) +
  theme_prism() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    strip.text = element_text(face = "bold", size = 10),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(size = 11, face = "bold"),
    legend.position = "none",
    plot.margin = margin(t = 5, r = 10, b = 5, l = 5)
  )

print(p_si3d_te)

ggsave(
  here("plots", "si3d_te_spike_comparison_normoxia.pdf"),
  plot = p_si3d_te,
  width = 6,
  height = 2.5
)

cat("\neIF3d TE density plot saved to plots/si3d_te_spike_comparison_normoxia.pdf\n")
```

# Significant Gene Counts: Pre vs Post Spike-in (Normoxia 1hr)

Bar plots showing the number of significantly changed TE genes (padj < 0.05) split by direction, comparing pre vs post spike-in normalization at the 1hr timepoint.

## eIF3e (si3e vs siCTRL) — Normoxia 1hr

```{r si3e_barplot_1hr, fig.width=3.5, fig.height=3.5}
si3e_bar_data <- data.frame(
  normalization = rep(c("Pre spike-in", "Post spike-in"), each = 2),
  direction     = rep(c("Up", "Down"), 2),
  count         = c(
    sum(si3e_nospike_results[["1hr"]]$padj < 0.05 & si3e_nospike_results[["1hr"]]$log2FoldChange > 0.58, na.rm = TRUE),
    -sum(si3e_nospike_results[["1hr"]]$padj < 0.05 & si3e_nospike_results[["1hr"]]$log2FoldChange < -0.58, na.rm = TRUE),
    sum(si3e_spike_1hr$te_padj < 0.05 & si3e_spike_1hr$te_lfc > 0.58, na.rm = TRUE),
    -sum(si3e_spike_1hr$te_padj < 0.05 & si3e_spike_1hr$te_lfc < -0.58, na.rm = TRUE)
  )
) %>%
  mutate(
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in")),
    direction     = factor(direction, levels = c("Up", "Down")),
    fill_group    = paste(normalization, direction)
  )

# Up: original density plot colors; Down: lighter/muted versions of same blues
si3e_colors <- c(
  "Pre spike-in Up"   = "#366885",
  "Post spike-in Up"  = "#56B4E9",
  "Pre spike-in Down" = "#8FAFC5",
  "Post spike-in Down"= "#BBDAEE"
)

p_si3e_bar <- ggplot(si3e_bar_data, aes(x = normalization, y = count, fill = fill_group)) +
  geom_col(position = "identity", width = 0.6) +
  geom_text(
    aes(label = abs(count), vjust = ifelse(count >= 0, -0.4, 1.4)),
    size = 3.5
  ) +
  scale_fill_manual(values = si3e_colors) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
  labs(
    title    = "eIF3e TE \u2014 Normoxia 1hr",
    subtitle = "padj < 0.05, |log2FC| > 0.58",
    x = NULL,
    y = "# sig genes"
  ) +
  theme_prism() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 9),
    plot.title    = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 8, color = "gray40")
  )

print(p_si3e_bar)

ggsave(
  here("plots", "si3e_te_sigcounts_spike_normoxia_1hr.pdf"),
  plot = p_si3e_bar,
  width = 3.5,
  height = 3.5
)
cat("eIF3e bar plot saved\n")
```

## eIF3d (si3d vs siCTRL) — Normoxia 1hr

```{r si3d_barplot_1hr, fig.width=3.5, fig.height=3.5}
si3d_bar_data <- data.frame(
  normalization = rep(c("Pre spike-in", "Post spike-in"), each = 2),
  direction     = rep(c("Up", "Down"), 2),
  count         = c(
    sum(si3d_nospike_results[["1hr"]]$padj < 0.05 & si3d_nospike_results[["1hr"]]$log2FoldChange > 0.58, na.rm = TRUE),
    -sum(si3d_nospike_results[["1hr"]]$padj < 0.05 & si3d_nospike_results[["1hr"]]$log2FoldChange < -0.58, na.rm = TRUE),
    sum(si3d_spike_1hr$te_padj < 0.05 & si3d_spike_1hr$te_lfc > 0.58, na.rm = TRUE),
    -sum(si3d_spike_1hr$te_padj < 0.05 & si3d_spike_1hr$te_lfc < -0.58, na.rm = TRUE)
  )
) %>%
  mutate(
    normalization = factor(normalization, levels = c("Pre spike-in", "Post spike-in")),
    direction     = factor(direction, levels = c("Up", "Down")),
    fill_group    = paste(normalization, direction)
  )

# Up: original density plot colors; Down: lighter/muted versions of same oranges
si3d_colors <- c(
  "Pre spike-in Up"   = "#87610e",
  "Post spike-in Up"  = "#E69F00",
  "Pre spike-in Down" = "#C0955A",
  "Post spike-in Down"= "#F0CC6A"
)

p_si3d_bar <- ggplot(si3d_bar_data, aes(x = normalization, y = count, fill = fill_group)) +
  geom_col(position = "identity", width = 0.6) +
  geom_text(
    aes(label = abs(count), vjust = ifelse(count >= 0, -0.4, 1.4)),
    size = 3.5
  ) +
  scale_fill_manual(values = si3d_colors) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
  labs(
    title    = "eIF3d TE \u2014 Normoxia 1hr",
    subtitle = "padj < 0.05, |log2FC| > 0.58",
    x = NULL,
    y = "# sig genes"
  ) +
  theme_prism() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 9),
    plot.title    = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 8, color = "gray40")
  )

print(p_si3d_bar)

ggsave(
  here("plots", "si3d_te_sigcounts_spike_normoxia_1hr.pdf"),
  plot = p_si3d_bar,
  width = 3.5,
  height = 3.5
)
cat("eIF3d bar plot saved\n")
```

## Combined — eIF3e and eIF3d, Normoxia 1hr

```{r combined_barplot_1hr, fig.width=5, fig.height=3.5}
combined_bar_data <- bind_rows(
  si3e_bar_data %>% mutate(gene = "eIF3e"),
  si3d_bar_data %>% mutate(gene = "eIF3d")
) %>%
  mutate(
    x_label = factor(
      paste0(gene, "\n", normalization),
      levels = c("eIF3e\nPre spike-in", "eIF3e\nPost spike-in",
                 "eIF3d\nPre spike-in", "eIF3d\nPost spike-in")
    ),
    fill_group = paste(gene, normalization, direction)
  )

combined_colors <- c(
  "eIF3e Pre spike-in Up"    = "#366885",
  "eIF3e Post spike-in Up"   = "#56B4E9",
  "eIF3e Pre spike-in Down"  = "#8FAFC5",
  "eIF3e Post spike-in Down" = "#BBDAEE",
  "eIF3d Pre spike-in Up"    = "#87610e",
  "eIF3d Post spike-in Up"   = "#E69F00",
  "eIF3d Pre spike-in Down"  = "#C0955A",
  "eIF3d Post spike-in Down" = "#F0CC6A"
)

p_combined_bar <- ggplot(combined_bar_data, aes(x = x_label, y = count, fill = fill_group)) +
  geom_col(position = "identity", width = 0.6) +
  geom_text(
    aes(label = abs(count), vjust = ifelse(count >= 0, -0.4, 1.4)),
    size = 3.5
  ) +
  scale_fill_manual(values = combined_colors) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
  labs(
    title    = "eIF3e and eIF3d TE \u2014 Normoxia 1hr",
    subtitle = "padj < 0.05, |log2FC| > 0.58",
    x = NULL,
    y = "# sig genes"
  ) +
  theme_prism() +
  theme(
    legend.position = "none",
    axis.text.x  = element_text(size = 9),
    plot.title   = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 8, color = "gray40")
  )

print(p_combined_bar)

ggsave(
  here("plots", "combined_te_sigcounts_spike_normoxia_1hr.pdf"),
  plot = p_combined_bar,
  width = 5,
  height = 3.5
)
cat("Combined bar plot saved\n")
```
