---
title: "Calculate Translation Efficiency from Spike-in Normalized Counts - All Treatments"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

# Overview

This analysis calculates translation efficiency (TE) directly from spike-in normalized count matrices for ALL treatments (sictrl, si3e, si3d) and conditions (normoxia, hypoxia).

TE is calculated as: TE = log2(ribo + 1) - log2(rna + 1)

**Important:** TE values are NOT centered because we use spike-in normalization. The spike-ins (Drosophila genes) provide an absolute reference scale, allowing us to preserve real biological differences in global translation efficiency between conditions (e.g., global translation suppression during hypoxia).

The spike-in normalized count matrices are loaded from CSVs created by eif3e_eif3d_normoxia_and_hypoxia.Rmd.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load_libraries}
library(here)
library(tidyverse)
library(reshape2)
library(ggpubr)
```

# Load Gene Annotation

```{r load_gene_annotation}
# Load gene annotation file
geneInfo <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"), col_names = FALSE)
colnames(geneInfo) <- c("gene_id", "symbol", "biotype", "species", "version")

# Create gene symbol lookup (keep only unique gene_id entries)
gene_symbols <- geneInfo %>%
  filter(species == "human") %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  dplyr::select(gene_id, symbol, biotype)

cat("Gene annotation loaded\n")
cat("Total genes:", nrow(gene_symbols), "\n")
```

# 1 Hour Timepoint - ALL Treatments and Conditions

## Load Spike-in Normalized Counts

```{r load_norm_matrix_all_1hr}
# Load spike-in normalized count matrix created by eif3e_eif3d_normoxia_and_hypoxia.Rmd
# This includes ALL treatments (sictrl, si3e, si3d) and both conditions (normoxia, hypoxia)
# Spike-in genes are already removed in eif3e_eif3d_normoxia_and_hypoxia.Rmd
allCountsNorm_all_1hr_SPIKE <- read_csv(here("output", "allCountsNorm_all_1hr_SPIKE.csv")) %>%
  column_to_rownames("gene_id")

cat("Loaded spike-in normalized matrix for ALL samples at 1hr\n")
cat("Dimensions:", dim(allCountsNorm_all_1hr_SPIKE), "\n")
cat("Sample names:\n")
print(colnames(allCountsNorm_all_1hr_SPIKE))
```

## Calculate TE for 1hr

```{r calculate_te_all_1hr}
# Extract ribo and rna columns and log2 transform
riboN_all_1hr <- log2(1 + as.matrix(allCountsNorm_all_1hr_SPIKE[, grep(pattern = "ribo", colnames(allCountsNorm_all_1hr_SPIKE))]))
rnaN_all_1hr <- log2(1 + as.matrix(allCountsNorm_all_1hr_SPIKE[, grep(pattern = "rna", colnames(allCountsNorm_all_1hr_SPIKE))]))

# Ensure we only use samples that have BOTH ribo and rna data
# Extract sample identifiers (everything before _ribo or _rna)
ribo_samples <- gsub("_ribo_.*", "", colnames(riboN_all_1hr))
rna_samples <- gsub("_rna_.*", "", colnames(rnaN_all_1hr))
paired_samples <- intersect(ribo_samples, rna_samples)

# Subset to paired samples only
ribo_idx <- match(paired_samples, ribo_samples)
rna_idx <- match(paired_samples, rna_samples)
riboN_all_1hr <- riboN_all_1hr[, ribo_idx]
rnaN_all_1hr <- rnaN_all_1hr[, rna_idx]

cat("1hr: Found", length(paired_samples), "paired ribo/rna samples\n")
if (length(setdiff(c(ribo_samples, rna_samples), paired_samples)) > 0) {
  cat("Excluded samples:", setdiff(c(ribo_samples, rna_samples), paired_samples), "\n")
}

# Calculate TE (NOT centered - spike-ins provide absolute reference)
TE_all_1hr <- riboN_all_1hr - rnaN_all_1hr

# Calculate median for reference
median_te_all_1hr <- median(as.data.frame(TE_all_1hr) %>% melt() %>% pull(value))
cat("Median TE (all 1hr):", median_te_all_1hr, "\n")

# Rename columns to indicate TE
colnames(TE_all_1hr) <- gsub(pattern = "ribo", replacement = "TE", x = colnames(TE_all_1hr))

cat("Calculated TE for all samples at 1hr\n")
```

## Visualize TE Distribution (1hr)

```{r plot_te_dist_all_1hr}
# TE distribution (not centered - preserves global translation changes)
p1 <- gghistogram(
  as.data.frame(TE_all_1hr) %>% melt(),
  x = "value",
  bins = 50,
  fill = "steelblue",
  title = "TE Distribution - All 1hr Samples (Spike-in Normalized)",
  xlab = "TE (log2)"
)

print(p1)
```

## Create Complete Data Table (1hr)

```{r complete_table_all_1hr}
# Combine TE, RNA, and Ribo data
full_te_list_all_1hr <- TE_all_1hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::select(symbol, gene_id, biotype, everything())

full_rna_list_all_1hr <- rnaN_all_1hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

full_ribo_list_all_1hr <- riboN_all_1hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

# Combine all data
complete_list_all_1hr <- full_te_list_all_1hr %>%
  left_join(full_rna_list_all_1hr, by = "gene_id") %>%
  left_join(full_ribo_list_all_1hr, by = "gene_id")

# Save to CSV
write_csv(complete_list_all_1hr, here("output", "complete_te_list_all_1hr.csv"))
cat("Saved complete TE list for all 1hr samples to output/complete_te_list_all_1hr.csv\n")
```

## Reformat to Long Format (1hr)

```{r reformat_long_all_1hr}
all_long_1hr <- complete_list_all_1hr %>%
  melt(id.vars = c("symbol", "gene_id", "biotype")) %>%
  separate(variable, into = c("condition", "treatment", "timepoint", "rep", "type", "id"), sep = "_")

head(all_long_1hr)
cat("Total rows in long format (1hr):", nrow(all_long_1hr), "\n")
```

# 4 Hour Timepoint - ALL Treatments and Conditions

## Load Spike-in Normalized Counts

```{r load_norm_matrix_all_4hr}
# Load spike-in normalized count matrix created by eif3e_eif3d_normoxia_and_hypoxia.Rmd
# Spike-in genes are already removed in eif3e_eif3d_normoxia_and_hypoxia.Rmd
allCountsNorm_all_4hr_SPIKE <- read_csv(here("output", "allCountsNorm_all_4hr_SPIKE.csv")) %>%
  column_to_rownames("gene_id")

cat("Loaded spike-in normalized matrix for ALL samples at 4hr\n")
cat("Dimensions:", dim(allCountsNorm_all_4hr_SPIKE), "\n")
```

## Calculate TE for 4hr

```{r calculate_te_all_4hr}
# Extract ribo and rna columns and log2 transform
riboN_all_4hr <- log2(1 + as.matrix(allCountsNorm_all_4hr_SPIKE[, grep(pattern = "ribo", colnames(allCountsNorm_all_4hr_SPIKE))]))
rnaN_all_4hr <- log2(1 + as.matrix(allCountsNorm_all_4hr_SPIKE[, grep(pattern = "rna", colnames(allCountsNorm_all_4hr_SPIKE))]))

# Ensure we only use samples that have BOTH ribo and rna data
# Extract sample identifiers (everything before _ribo or _rna)
ribo_samples <- gsub("_ribo_.*", "", colnames(riboN_all_4hr))
rna_samples <- gsub("_rna_.*", "", colnames(rnaN_all_4hr))
paired_samples <- intersect(ribo_samples, rna_samples)

# Subset to paired samples only
ribo_idx <- match(paired_samples, ribo_samples)
rna_idx <- match(paired_samples, rna_samples)
riboN_all_4hr <- riboN_all_4hr[, ribo_idx]
rnaN_all_4hr <- rnaN_all_4hr[, rna_idx]

cat("4hr: Found", length(paired_samples), "paired ribo/rna samples\n")
if (length(setdiff(c(ribo_samples, rna_samples), paired_samples)) > 0) {
  cat("Excluded samples:", setdiff(c(ribo_samples, rna_samples), paired_samples), "\n")
}

# Calculate TE (NOT centered - spike-ins provide absolute reference)
TE_all_4hr <- riboN_all_4hr - rnaN_all_4hr

# Calculate median for reference
median_te_all_4hr <- median(as.data.frame(TE_all_4hr) %>% melt() %>% pull(value))
cat("Median TE (all 4hr):", median_te_all_4hr, "\n")

# Rename columns
colnames(TE_all_4hr) <- gsub(pattern = "ribo", replacement = "TE", x = colnames(TE_all_4hr))

cat("Calculated TE for all samples at 4hr\n")
```

## Visualize TE Distribution (4hr)

```{r plot_te_dist_all_4hr}
# TE distribution (not centered - preserves global translation changes)
p1 <- gghistogram(
  as.data.frame(TE_all_4hr) %>% melt(),
  x = "value",
  bins = 50,
  fill = "steelblue",
  title = "TE Distribution - All 4hr Samples (Spike-in Normalized)",
  xlab = "TE (log2)"
)

print(p1)
```

## Create Complete Data Table (4hr)

```{r complete_table_all_4hr}
# Combine TE, RNA, and Ribo data
full_te_list_all_4hr <- TE_all_4hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::select(symbol, gene_id, biotype, everything())

full_rna_list_all_4hr <- rnaN_all_4hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

full_ribo_list_all_4hr <- riboN_all_4hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

# Combine all data
complete_list_all_4hr <- full_te_list_all_4hr %>%
  left_join(full_rna_list_all_4hr, by = "gene_id") %>%
  left_join(full_ribo_list_all_4hr, by = "gene_id")

# Save to CSV
write_csv(complete_list_all_4hr, here("output", "complete_te_list_all_4hr.csv"))
cat("Saved complete TE list for all 4hr samples to output/complete_te_list_all_4hr.csv\n")
```

## Reformat to Long Format (4hr)

```{r reformat_long_all_4hr}
all_long_4hr <- complete_list_all_4hr %>%
  melt(id.vars = c("symbol", "gene_id", "biotype")) %>%
  separate(variable, into = c("condition", "treatment", "timepoint", "rep", "type", "id"), sep = "_")

head(all_long_4hr)
cat("Total rows in long format (4hr):", nrow(all_long_4hr), "\n")
```

# 24 Hour Timepoint - ALL Treatments and Conditions

## Load Spike-in Normalized Counts

```{r load_norm_matrix_all_24hr}
# Load spike-in normalized count matrix created by eif3e_eif3d_normoxia_and_hypoxia.Rmd
# Spike-in genes are already removed in eif3e_eif3d_normoxia_and_hypoxia.Rmd
allCountsNorm_all_24hr_SPIKE <- read_csv(here("output", "allCountsNorm_all_24hr_SPIKE.csv")) %>%
  column_to_rownames("gene_id")

cat("Loaded spike-in normalized matrix for ALL samples at 24hr\n")
cat("Dimensions:", dim(allCountsNorm_all_24hr_SPIKE), "\n")
```

## Calculate TE for 24hr

```{r calculate_te_all_24hr}
# Extract ribo and rna columns and log2 transform
riboN_all_24hr <- log2(1 + as.matrix(allCountsNorm_all_24hr_SPIKE[, grep(pattern = "ribo", colnames(allCountsNorm_all_24hr_SPIKE))]))
rnaN_all_24hr <- log2(1 + as.matrix(allCountsNorm_all_24hr_SPIKE[, grep(pattern = "rna", colnames(allCountsNorm_all_24hr_SPIKE))]))

# Ensure we only use samples that have BOTH ribo and rna data
# Extract sample identifiers (everything before _ribo or _rna)
ribo_samples <- gsub("_ribo_.*", "", colnames(riboN_all_24hr))
rna_samples <- gsub("_rna_.*", "", colnames(rnaN_all_24hr))
paired_samples <- intersect(ribo_samples, rna_samples)

# Subset to paired samples only
ribo_idx <- match(paired_samples, ribo_samples)
rna_idx <- match(paired_samples, rna_samples)
riboN_all_24hr <- riboN_all_24hr[, ribo_idx]
rnaN_all_24hr <- rnaN_all_24hr[, rna_idx]

cat("24hr: Found", length(paired_samples), "paired ribo/rna samples\n")
if (length(setdiff(c(ribo_samples, rna_samples), paired_samples)) > 0) {
  cat("Excluded samples:", setdiff(c(ribo_samples, rna_samples), paired_samples), "\n")
}

# Calculate TE (NOT centered - spike-ins provide absolute reference)
TE_all_24hr <- riboN_all_24hr - rnaN_all_24hr

# Calculate median for reference
median_te_all_24hr <- median(as.data.frame(TE_all_24hr) %>% melt() %>% pull(value))
cat("Median TE (all 24hr):", median_te_all_24hr, "\n")

# Rename columns
colnames(TE_all_24hr) <- gsub(pattern = "ribo", replacement = "TE", x = colnames(TE_all_24hr))

cat("Calculated TE for all samples at 24hr\n")
```

## Visualize TE Distribution (24hr)

```{r plot_te_dist_all_24hr}
# TE distribution (not centered - preserves global translation changes)
p1 <- gghistogram(
  as.data.frame(TE_all_24hr) %>% melt(),
  x = "value",
  bins = 50,
  fill = "steelblue",
  title = "TE Distribution - All 24hr Samples (Spike-in Normalized)",
  xlab = "TE (log2)"
)

print(p1)
```

## Create Complete Data Table (24hr)

```{r complete_table_all_24hr}
# Combine TE, RNA, and Ribo data
full_te_list_all_24hr <- TE_all_24hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::select(symbol, gene_id, biotype, everything())

full_rna_list_all_24hr <- rnaN_all_24hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

full_ribo_list_all_24hr <- riboN_all_24hr %>%
  as.data.frame() %>%
  rownames_to_column("gene_id")

# Combine all data
complete_list_all_24hr <- full_te_list_all_24hr %>%
  left_join(full_rna_list_all_24hr, by = "gene_id") %>%
  left_join(full_ribo_list_all_24hr, by = "gene_id")

# Save to CSV
write_csv(complete_list_all_24hr, here("output", "complete_te_list_all_24hr.csv"))
cat("Saved complete TE list for all 24hr samples to output/complete_te_list_all_24hr.csv\n")
```

## Reformat to Long Format (24hr)

```{r reformat_long_all_24hr}
all_long_24hr <- complete_list_all_24hr %>%
  melt(id.vars = c("symbol", "gene_id", "biotype")) %>%
  separate(variable, into = c("condition", "treatment", "timepoint", "rep", "type", "id"), sep = "_")

head(all_long_24hr)
cat("Total rows in long format (24hr):", nrow(all_long_24hr), "\n")
```

# Sanity Checks

## Check eIF3e RNA levels (1hr, normoxia)

```{r check_eif3e_rna_1hr}
rna_3e_formatted_1hr <- all_long_1hr %>%
  filter(symbol == "EIF3E") %>%
  filter(condition == "normoxia") %>%
  filter(type == "rna")

ggplot(rna_3e_formatted_1hr, aes(x = treatment, y = value)) +
  geom_point(size = 3) +
  labs(title = "EIF3E RNA Levels - 1hr Normoxia (All Treatments)",
       x = "Treatment",
       y = "Log2(Normalized Counts + 1)") +
  theme_minimal()
```

## Check eIF3d RNA levels (1hr, normoxia)

```{r check_eif3d_rna_1hr}
rna_3d_formatted_1hr <- all_long_1hr %>%
  filter(symbol == "EIF3D") %>%
  filter(condition == "normoxia") %>%
  filter(type == "rna")

ggplot(rna_3d_formatted_1hr, aes(x = treatment, y = value)) +
  geom_point(size = 3) +
  labs(title = "EIF3D RNA Levels - 1hr Normoxia (All Treatments)",
       x = "Treatment",
       y = "Log2(Normalized Counts + 1)") +
  theme_minimal()
```

## Example: ATF4 RNA, Ribo, and TE across treatments (1hr, normoxia)

```{r atf4_example_1hr_norm}
atf4_data_1hr_norm <- all_long_1hr %>%
  filter(symbol == "ATF4") %>%
  filter(condition == "normoxia")

ggplot(atf4_data_1hr_norm, aes(x = treatment, y = value, color = treatment)) +
  geom_point(size = 3) +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "ATF4 in Normoxia - 1hr (All Treatments)",
       x = "Treatment",
       y = "Value") +
  theme_minimal()
```

## Example: ATF4 RNA, Ribo, and TE - Hypoxia vs Normoxia in sictrl (1hr)

```{r atf4_hypoxia_sictrl_1hr}
atf4_data_sictrl_1hr <- all_long_1hr %>%
  filter(symbol == "ATF4") %>%
  filter(treatment == "sictrl")

ggplot(atf4_data_sictrl_1hr, aes(x = condition, y = value, color = condition)) +
  geom_point(size = 3) +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "ATF4 in sictrl - 1hr (Normoxia vs Hypoxia)",
       x = "Condition",
       y = "Value") +
  scale_color_manual(values = c("normoxia" = "blue", "hypoxia" = "red")) +
  theme_minimal()
```

## Example: ATF4 across ALL treatments and conditions (1hr)

```{r atf4_all_1hr}
atf4_data_all_1hr <- all_long_1hr %>%
  filter(symbol == "ATF4")

ggplot(atf4_data_all_1hr, aes(x = treatment, y = value, color = condition)) +
  geom_point(size = 3, position = position_dodge(width = 0.3)) +
  facet_wrap(~type, scales = "free_y") +
  labs(title = "ATF4 - 1hr (All Treatments and Conditions)",
       x = "Treatment",
       y = "Value") +
  scale_color_manual(values = c("normoxia" = "blue", "hypoxia" = "red")) +
  theme_minimal()
```

# Summary Statistics

## Number of samples per treatment and condition

```{r summary_stats}
# 1hr
# summary_1hr <- all_long_1hr %>%
#   distinct(condition, treatment, rep, type, id) %>%
#   count(condition, treatment, type)

# cat("1hr sample counts:\n")
# print(summary_1hr)

# # 4hr
# summary_4hr <- all_long_4hr %>%
#   distinct(condition, treatment, rep, type, id) %>%
#   count(condition, treatment, type)

# cat("\n4hr sample counts:\n")
# print(summary_4hr)

# # 24hr
# summary_24hr <- all_long_24hr %>%
#   distinct(condition, treatment, rep, type, id) %>%
#   count(condition, treatment, type)

# cat("\n24hr sample counts:\n")
# print(summary_24hr)
```

# TE Correlation Across Timepoints

## Combine TE data from all timepoints

```{r combine_timepoints}
# Extract TE values for normoxia sictrl across all timepoints
# Get average TE per gene per timepoint

te_1hr_norm_sictrl <- all_long_1hr %>%
  filter(condition == "normoxia", treatment == "sictrl", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_1hr = mean(value, na.rm = TRUE), .groups = "drop")

te_4hr_norm_sictrl <- all_long_4hr %>%
  filter(condition == "normoxia", treatment == "sictrl", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_4hr = mean(value, na.rm = TRUE), .groups = "drop")

te_24hr_norm_sictrl <- all_long_24hr %>%
  filter(condition == "normoxia", treatment == "sictrl", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_24hr = mean(value, na.rm = TRUE), .groups = "drop")

# Get mean RNA expression across all timepoints and samples for filtering
rna_expr_all <- bind_rows(
  all_long_1hr %>% filter(condition == "normoxia", treatment == "sictrl", type == "rna"),
  all_long_4hr %>% filter(condition == "normoxia", treatment == "sictrl", type == "rna"),
  all_long_24hr %>% filter(condition == "normoxia", treatment == "sictrl", type == "rna")
) %>%
  group_by(gene_id) %>%
  summarize(mean_rna = mean(value, na.rm = TRUE), .groups = "drop")

# Combine all TE data
te_combined <- te_1hr_norm_sictrl %>%
  left_join(te_4hr_norm_sictrl, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(te_24hr_norm_sictrl, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(rna_expr_all, by = "gene_id") %>%
  filter(!is.na(TE_1hr) & !is.na(TE_4hr) & !is.na(TE_24hr))

cat("Combined TE data for normoxia sictrl:\n")
cat("Total genes:", nrow(te_combined), "\n")
head(te_combined)
```

## TE Correlations - All Genes

```{r te_correlations_all}
# Calculate correlations
cor_1hr_4hr <- cor(te_combined$TE_1hr, te_combined$TE_4hr, use = "complete.obs")
cor_1hr_24hr <- cor(te_combined$TE_1hr, te_combined$TE_24hr, use = "complete.obs")
cor_4hr_24hr <- cor(te_combined$TE_4hr, te_combined$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_all <- bind_rows(
  te_combined %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                         cor_val = cor_1hr_4hr),
  te_combined %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                         cor_val = cor_1hr_24hr),
  te_combined %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                         cor_val = cor_4hr_24hr)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_all_faceted <- ggplot(te_pairs_all, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_all %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia sictrl) - All Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_all_faceted)

# Save plot
ggsave(here("plots", "te_correlations_all_genes_faceted.pdf"),
       p_all_faceted, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_all_genes_faceted.pdf\n")

cat("\nCorrelation summary (all genes):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr, 3), "\n")
```

## Filter to 500 Most Expressed Genes

```{r top500_genes}
# Identify top 500 most expressed genes based on mean RNA expression
top500_genes <- te_combined %>%
  arrange(desc(mean_rna)) %>%
  head(500) %>%
  pull(gene_id)

te_top500 <- te_combined %>%
  filter(gene_id %in% top500_genes)

cat("Top 500 most expressed genes:\n")
cat("Number of genes:", nrow(te_top500), "\n")
cat("Mean RNA range:", range(te_top500$mean_rna), "\n")
```

## TE Correlations - Top 500 Most Expressed Genes

```{r te_correlations_top500}
# Calculate correlations
cor_1hr_4hr_top500 <- cor(te_top500$TE_1hr, te_top500$TE_4hr, use = "complete.obs")
cor_1hr_24hr_top500 <- cor(te_top500$TE_1hr, te_top500$TE_24hr, use = "complete.obs")
cor_4hr_24hr_top500 <- cor(te_top500$TE_4hr, te_top500$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_top500 <- bind_rows(
  te_top500 %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                       cor_val = cor_1hr_4hr_top500),
  te_top500 %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                       cor_val = cor_1hr_24hr_top500),
  te_top500 %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                       cor_val = cor_4hr_24hr_top500)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_top500_faceted <- ggplot(te_pairs_top500, aes(x = x, y = y)) +
  geom_point(alpha = 0.5, size = 1.5, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_top500 %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia sictrl) - Top 500 Most Expressed Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_top500_faceted)

# Save plot
ggsave(here("plots", "te_correlations_top500_genes_faceted.pdf"),
       p_top500_faceted, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_top500_genes_faceted.pdf\n")

cat("\nCorrelation summary (top 500 most expressed genes):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr_top500, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr_top500, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr_top500, 3), "\n")
```

## Save Combined TE Data - sictrl

```{r save_combined_te_sictrl}
# Save combined TE data for all genes
write_csv(te_combined, here("output", "te_combined_normoxia_sictrl.csv"))
cat("Saved combined TE data to output/te_combined_normoxia_sictrl.csv\n")

# Save top 500 genes
write_csv(te_top500, here("output", "te_top500_normoxia_sictrl.csv"))
cat("Saved top 500 genes to output/te_top500_normoxia_sictrl.csv\n")
```

# TE Correlation Across Timepoints - si3d

## Combine TE data from all timepoints - si3d

```{r combine_timepoints_si3d}
# Extract TE values for normoxia si3d across all timepoints
te_1hr_norm_si3d <- all_long_1hr %>%
  filter(condition == "normoxia", treatment == "si3d", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_1hr = mean(value, na.rm = TRUE), .groups = "drop")

te_4hr_norm_si3d <- all_long_4hr %>%
  filter(condition == "normoxia", treatment == "si3d", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_4hr = mean(value, na.rm = TRUE), .groups = "drop")

te_24hr_norm_si3d <- all_long_24hr %>%
  filter(condition == "normoxia", treatment == "si3d", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_24hr = mean(value, na.rm = TRUE), .groups = "drop")

# Get mean RNA expression across all timepoints and samples
rna_expr_si3d <- bind_rows(
  all_long_1hr %>% filter(condition == "normoxia", treatment == "si3d", type == "rna"),
  all_long_4hr %>% filter(condition == "normoxia", treatment == "si3d", type == "rna"),
  all_long_24hr %>% filter(condition == "normoxia", treatment == "si3d", type == "rna")
) %>%
  group_by(gene_id) %>%
  summarize(mean_rna = mean(value, na.rm = TRUE), .groups = "drop")

# Combine all TE data
te_combined_si3d <- te_1hr_norm_si3d %>%
  left_join(te_4hr_norm_si3d, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(te_24hr_norm_si3d, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(rna_expr_si3d, by = "gene_id") %>%
  filter(!is.na(TE_1hr) & !is.na(TE_4hr) & !is.na(TE_24hr))

cat("Combined TE data for normoxia si3d:\n")
cat("Total genes:", nrow(te_combined_si3d), "\n")
```

## TE Correlations - All Genes (si3d)

```{r te_correlations_all_si3d}
# Calculate correlations
cor_1hr_4hr_si3d <- cor(te_combined_si3d$TE_1hr, te_combined_si3d$TE_4hr, use = "complete.obs")
cor_1hr_24hr_si3d <- cor(te_combined_si3d$TE_1hr, te_combined_si3d$TE_24hr, use = "complete.obs")
cor_4hr_24hr_si3d <- cor(te_combined_si3d$TE_4hr, te_combined_si3d$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_all_si3d <- bind_rows(
  te_combined_si3d %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                              cor_val = cor_1hr_4hr_si3d),
  te_combined_si3d %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                              cor_val = cor_1hr_24hr_si3d),
  te_combined_si3d %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                              cor_val = cor_4hr_24hr_si3d)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_all_faceted_si3d <- ggplot(te_pairs_all_si3d, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_all_si3d %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia si3d) - All Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_all_faceted_si3d)

# Save plot
ggsave(here("plots", "te_correlations_all_genes_faceted_si3d.pdf"),
       p_all_faceted_si3d, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_all_genes_faceted_si3d.pdf\n")

cat("\nCorrelation summary (all genes, si3d):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr_si3d, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr_si3d, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr_si3d, 3), "\n")
```

## Filter to 500 Most Expressed Genes (si3d)

```{r top500_genes_si3d}
# Identify top 500 most expressed genes
top500_genes_si3d <- te_combined_si3d %>%
  arrange(desc(mean_rna)) %>%
  head(500) %>%
  pull(gene_id)

te_top500_si3d <- te_combined_si3d %>%
  filter(gene_id %in% top500_genes_si3d)

cat("Top 500 most expressed genes (si3d):\n")
cat("Number of genes:", nrow(te_top500_si3d), "\n")
cat("Mean RNA range:", range(te_top500_si3d$mean_rna), "\n")
```

## TE Correlations - Top 500 Most Expressed Genes (si3d)

```{r te_correlations_top500_si3d}
# Calculate correlations
cor_1hr_4hr_top500_si3d <- cor(te_top500_si3d$TE_1hr, te_top500_si3d$TE_4hr, use = "complete.obs")
cor_1hr_24hr_top500_si3d <- cor(te_top500_si3d$TE_1hr, te_top500_si3d$TE_24hr, use = "complete.obs")
cor_4hr_24hr_top500_si3d <- cor(te_top500_si3d$TE_4hr, te_top500_si3d$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_top500_si3d <- bind_rows(
  te_top500_si3d %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                            cor_val = cor_1hr_4hr_top500_si3d),
  te_top500_si3d %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                            cor_val = cor_1hr_24hr_top500_si3d),
  te_top500_si3d %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                            cor_val = cor_4hr_24hr_top500_si3d)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_top500_faceted_si3d <- ggplot(te_pairs_top500_si3d, aes(x = x, y = y)) +
  geom_point(alpha = 0.5, size = 1.5, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_top500_si3d %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia si3d) - Top 500 Most Expressed Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_top500_faceted_si3d)

# Save plot
ggsave(here("plots", "te_correlations_top500_genes_faceted_si3d.pdf"),
       p_top500_faceted_si3d, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_top500_genes_faceted_si3d.pdf\n")

cat("\nCorrelation summary (top 500 most expressed genes, si3d):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr_top500_si3d, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr_top500_si3d, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr_top500_si3d, 3), "\n")
```

## Save Combined TE Data - si3d

```{r save_combined_te_si3d}
# Save combined TE data for all genes
write_csv(te_combined_si3d, here("output", "te_combined_normoxia_si3d.csv"))
cat("Saved combined TE data to output/te_combined_normoxia_si3d.csv\n")

# Save top 500 genes
write_csv(te_top500_si3d, here("output", "te_top500_normoxia_si3d.csv"))
cat("Saved top 500 genes to output/te_top500_normoxia_si3d.csv\n")
```

# TE Correlation Across Timepoints - si3e

## Combine TE data from all timepoints - si3e

```{r combine_timepoints_si3e}
# Extract TE values for normoxia si3e across all timepoints
te_1hr_norm_si3e <- all_long_1hr %>%
  filter(condition == "normoxia", treatment == "si3e", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_1hr = mean(value, na.rm = TRUE), .groups = "drop")

te_4hr_norm_si3e <- all_long_4hr %>%
  filter(condition == "normoxia", treatment == "si3e", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_4hr = mean(value, na.rm = TRUE), .groups = "drop")

te_24hr_norm_si3e <- all_long_24hr %>%
  filter(condition == "normoxia", treatment == "si3e", type == "TE") %>%
  group_by(symbol, gene_id, biotype) %>%
  summarize(TE_24hr = mean(value, na.rm = TRUE), .groups = "drop")

# Get mean RNA expression across all timepoints and samples
rna_expr_si3e <- bind_rows(
  all_long_1hr %>% filter(condition == "normoxia", treatment == "si3e", type == "rna"),
  all_long_4hr %>% filter(condition == "normoxia", treatment == "si3e", type == "rna"),
  all_long_24hr %>% filter(condition == "normoxia", treatment == "si3e", type == "rna")
) %>%
  group_by(gene_id) %>%
  summarize(mean_rna = mean(value, na.rm = TRUE), .groups = "drop")

# Combine all TE data
te_combined_si3e <- te_1hr_norm_si3e %>%
  left_join(te_4hr_norm_si3e, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(te_24hr_norm_si3e, by = c("symbol", "gene_id", "biotype")) %>%
  left_join(rna_expr_si3e, by = "gene_id") %>%
  filter(!is.na(TE_1hr) & !is.na(TE_4hr) & !is.na(TE_24hr))

cat("Combined TE data for normoxia si3e:\n")
cat("Total genes:", nrow(te_combined_si3e), "\n")
```

## TE Correlations - All Genes (si3e)

```{r te_correlations_all_si3e}
# Calculate correlations
cor_1hr_4hr_si3e <- cor(te_combined_si3e$TE_1hr, te_combined_si3e$TE_4hr, use = "complete.obs")
cor_1hr_24hr_si3e <- cor(te_combined_si3e$TE_1hr, te_combined_si3e$TE_24hr, use = "complete.obs")
cor_4hr_24hr_si3e <- cor(te_combined_si3e$TE_4hr, te_combined_si3e$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_all_si3e <- bind_rows(
  te_combined_si3e %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                              cor_val = cor_1hr_4hr_si3e),
  te_combined_si3e %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                              cor_val = cor_1hr_24hr_si3e),
  te_combined_si3e %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                              cor_val = cor_4hr_24hr_si3e)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_all_faceted_si3e <- ggplot(te_pairs_all_si3e, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_all_si3e %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia si3e) - All Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_all_faceted_si3e)

# Save plot
ggsave(here("plots", "te_correlations_all_genes_faceted_si3e.pdf"),
       p_all_faceted_si3e, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_all_genes_faceted_si3e.pdf\n")

cat("\nCorrelation summary (all genes, si3e):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr_si3e, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr_si3e, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr_si3e, 3), "\n")
```

## Filter to 500 Most Expressed Genes (si3e)

```{r top500_genes_si3e}
# Identify top 500 most expressed genes
top500_genes_si3e <- te_combined_si3e %>%
  arrange(desc(mean_rna)) %>%
  head(500) %>%
  pull(gene_id)

te_top500_si3e <- te_combined_si3e %>%
  filter(gene_id %in% top500_genes_si3e)

cat("Top 500 most expressed genes (si3e):\n")
cat("Number of genes:", nrow(te_top500_si3e), "\n")
cat("Mean RNA range:", range(te_top500_si3e$mean_rna), "\n")
```

## TE Correlations - Top 500 Most Expressed Genes (si3e)

```{r te_correlations_top500_si3e}
# Calculate correlations
cor_1hr_4hr_top500_si3e <- cor(te_top500_si3e$TE_1hr, te_top500_si3e$TE_4hr, use = "complete.obs")
cor_1hr_24hr_top500_si3e <- cor(te_top500_si3e$TE_1hr, te_top500_si3e$TE_24hr, use = "complete.obs")
cor_4hr_24hr_top500_si3e <- cor(te_top500_si3e$TE_4hr, te_top500_si3e$TE_24hr, use = "complete.obs")

# Reshape data for faceted plot
te_pairs_top500_si3e <- bind_rows(
  te_top500_si3e %>% mutate(comparison = "1hr vs 4hr", x = TE_1hr, y = TE_4hr,
                            cor_val = cor_1hr_4hr_top500_si3e),
  te_top500_si3e %>% mutate(comparison = "1hr vs 24hr", x = TE_1hr, y = TE_24hr,
                            cor_val = cor_1hr_24hr_top500_si3e),
  te_top500_si3e %>% mutate(comparison = "4hr vs 24hr", x = TE_4hr, y = TE_24hr,
                            cor_val = cor_4hr_24hr_top500_si3e)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

# Create faceted plot
p_top500_faceted_si3e <- ggplot(te_pairs_top500_si3e, aes(x = x, y = y)) +
  geom_point(alpha = 0.5, size = 1.5, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison, scales = "free") +
  geom_text(data = te_pairs_top500_si3e %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 4, color = "black") +
  labs(title = "TE Correlations Across Timepoints (normoxia si3e) - Top 500 Most Expressed Genes",
       x = "TE (earlier timepoint)",
       y = "TE (later timepoint)") +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))

print(p_top500_faceted_si3e)

# Save plot
ggsave(here("plots", "te_correlations_top500_genes_faceted_si3e.pdf"),
       p_top500_faceted_si3e, width = 12, height = 4)
cat("Saved plot to plots/te_correlations_top500_genes_faceted_si3e.pdf\n")

cat("\nCorrelation summary (top 500 most expressed genes, si3e):\n")
cat("1hr vs 4hr: R =", round(cor_1hr_4hr_top500_si3e, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr_top500_si3e, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr_top500_si3e, 3), "\n")
```

## Save Combined TE Data - si3e

```{r save_combined_te_si3e}
# Save combined TE data for all genes
write_csv(te_combined_si3e, here("output", "te_combined_normoxia_si3e.csv"))
cat("Saved combined TE data to output/te_combined_normoxia_si3e.csv\n")

# Save top 500 genes
write_csv(te_top500_si3e, here("output", "te_top500_normoxia_si3e.csv"))
cat("Saved top 500 genes to output/te_top500_normoxia_si3e.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
