---
title: "CLIP Binding Sites Analysis - Lamper Cap-Binding Data"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(cowplot)
library(readxl)
library(janitor)
library(rstatix)
```

## Overview

This analysis examines whether eIF3d cap-binding targets (from Lamper et al. glucose deprivation experiments) show differential hypoxic response compared to other genes.

## Load TE Changes from Hypoxia Timecourse

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

# Add gene symbols to TE data and prepare for analysis
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Loaded TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes\n")
cat("4hr:", nrow(te_4hr_gsea), "genes\n")
cat("24hr:", nrow(te_24hr_gsea), "genes\n")
```

## Load Lamper Cap-Binding Data

```{r load_lamper_data}
# Load Lamper et al. cap-binding data (glucose deprivation cap binders)
lamper_data <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/lamper-s3.xlsx") %>%
  clean_names()

cat("Lamper data columns after clean_names():\n")
print(colnames(lamper_data))

cat("\nLamper data dimensions:", nrow(lamper_data), "rows x", ncol(lamper_data), "columns\n")
head(lamper_data)
```

## Define Cap-Binding Gene Sets

Using thresholds from original analysis:
- Glucose deprivation cap binders: `log2fold_change_glu > 2 & padj_glu < 0.005`
- Complete media cap binders: `log2fold_change_cm > 2 & padj_cm < 0.005`

```{r define_cap_binders}
# Convert columns to numeric (they may be stored as character in Excel)
lamper_data <- lamper_data %>%
  mutate(
    log2fold_change_cm = as.numeric(log2fold_change_cm),
    log2fold_change_glu = as.numeric(log2fold_change_glu),
    padj_cm = as.numeric(padj_cm),
    padj_glu = as.numeric(padj_glu)
  )

# Define glucose deprivation cap binders (GLU_HIT)
glu_cap_binders <- lamper_data %>%
  filter(log2fold_change_glu > 2 & padj_glu < 0.005) %>%
  pull(ensembl_gene_id)

# Define complete media cap binders (CM_HIT)
cm_cap_binders <- lamper_data %>%
  filter(log2fold_change_cm > 2 & padj_cm < 0.005) %>%
  pull(ensembl_gene_id)

cat("Number of glucose deprivation cap binders (GLU_HIT):", length(glu_cap_binders), "\n")
cat("Number of complete media cap binders (CM_HIT):", length(cm_cap_binders), "\n")
cat("Overlap between GLU and CM:", length(intersect(glu_cap_binders, cm_cap_binders)), "\n")
```

## Classify Genes by Cap-Binding Status

```{r classify_genes}
# Strip version numbers from gene IDs for matching
# TE data has versioned IDs (e.g., ENSG00000000003.17)
# Lamper data may have unversioned IDs

# Clean gene IDs in TE data
te_1hr_cap_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

te_4hr_cap_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

te_24hr_cap_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

# Report overlap statistics
cat("\nOverlap with TE data:\n")
cat("1hr - Glucose cap binders found:", sum(te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
cat("4hr - Glucose cap binders found:", sum(te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
cat("24hr - Glucose cap binders found:", sum(te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (1 Hour)

```{r ecdf_1hr, fig.width=6, fig.height=5}
# Create ECDF plot for 1 hour
p_1hr <- ggplot(te_1hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 1 Hour",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_1hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_1hr.pdf"),
  plot = p_1hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (4 Hours)

```{r ecdf_4hr, fig.width=6, fig.height=5}
# Create ECDF plot for 4 hours
p_4hr <- ggplot(te_4hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 4 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_4hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_4hr.pdf"),
  plot = p_4hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (24 Hours)

```{r ecdf_24hr, fig.width=6, fig.height=5}
# Create ECDF plot for 24 hours
p_24hr <- ggplot(te_24hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 24 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_24hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_24hr.pdf"),
  plot = p_24hr,
  width = 5,
  height = 4
)
```

## Statistical Tests

```{r statistical_tests}
# Perform KS tests comparing cap binders vs other genes
ks_1hr <- ks.test(
  te_1hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_1hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

ks_4hr <- ks.test(
  te_4hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_4hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

ks_24hr <- ks.test(
  te_24hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_24hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

# Perform Wilcoxon tests using rstatix
wilcox_1hr <- te_1hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

wilcox_4hr <- te_4hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

wilcox_24hr <- te_24hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

# Compile statistics
stats_results <- data.frame(
  timepoint = c("1hr", "4hr", "24hr"),
  n_cap_binders = c(
    sum(te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"),
    sum(te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"),
    sum(te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder")
  ),
  n_other = c(
    sum(te_1hr_cap_status$cap_status == "Other"),
    sum(te_4hr_cap_status$cap_status == "Other"),
    sum(te_24hr_cap_status$cap_status == "Other")
  ),
  median_cap_binders = c(
    median(te_1hr_cap_status$log2FoldChange[te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"]),
    median(te_4hr_cap_status$log2FoldChange[te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"]),
    median(te_24hr_cap_status$log2FoldChange[te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder"])
  ),
  median_other = c(
    median(te_1hr_cap_status$log2FoldChange[te_1hr_cap_status$cap_status == "Other"]),
    median(te_4hr_cap_status$log2FoldChange[te_4hr_cap_status$cap_status == "Other"]),
    median(te_24hr_cap_status$log2FoldChange[te_24hr_cap_status$cap_status == "Other"])
  ),
  ks_statistic = c(ks_1hr$statistic, ks_4hr$statistic, ks_24hr$statistic),
  ks_pvalue = c(ks_1hr$p.value, ks_4hr$p.value, ks_24hr$p.value),
  wilcox_statistic = c(wilcox_1hr$statistic, wilcox_4hr$statistic, wilcox_24hr$statistic),
  wilcox_pvalue = c(wilcox_1hr$p, wilcox_4hr$p, wilcox_24hr$p)
)

cat("\nStatistical Test Results:\n")
print(stats_results)

# Export statistics
write_csv(stats_results, here("output", "cap_binder_statistics.csv"))
cat("\nStatistics exported to: output/cap_binder_statistics.csv\n")
```

## Faceted All-Timepoints Plot

```{r faceted_plot, fig.width=10, fig.height=4}
# Combine all timepoints for faceted plot
te_combined <- bind_rows(
  te_1hr_cap_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_cap_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_cap_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels <- te_combined %>%
  group_by(timepoint, cap_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      cap_status == "Glucose Dep. Cap Binder" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot
p_faceted <- ggplot(te_combined, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels, aes(x = x, y = y, label = label, color = cap_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_faceted)

# Save the faceted plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_all_timepoints.pdf"),
  plot = p_faceted,
  width = 10,
  height = 4
)
```

## Complete Media Cap Binders Analysis

For comparison, also analyze complete media cap binders.

```{r cm_cap_binders, fig.width=10, fig.height=4}
# Classify genes by CM cap-binding status
te_1hr_cm_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

te_4hr_cm_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

te_24hr_cm_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

cat("\nComplete Media Cap Binder overlap with TE data:\n")
cat("1hr:", sum(te_1hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")
cat("4hr:", sum(te_4hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")
cat("24hr:", sum(te_24hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")

# Combine all timepoints for faceted plot
te_cm_combined <- bind_rows(
  te_1hr_cm_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_cm_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_cm_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels_cm <- te_cm_combined %>%
  group_by(timepoint, cap_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      cap_status == "Complete Media Cap Binder" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot for CM cap binders
p_cm_faceted <- ggplot(te_cm_combined, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Complete Media Cap Binder" = "#0072B2", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels_cm, aes(x = x, y = y, label = label, color = cap_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper - Complete Media)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_cm_faceted)

# Save the CM faceted plot
ggsave(
  here("plots", "ecdf_cap_binders_cm_all_timepoints.pdf"),
  plot = p_cm_faceted,
  width = 10,
  height = 4
)
```

## Lee PAR-CLIP eIF3d Binding Analysis

Analyze eIF3d binding targets from Lee et al. 2015 PAR-CLIP data.

```{r load_lee_parclip}
# Load Lee 2015 PAR-CLIP data
lee_parclip <- read_csv("/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv") %>%
  clean_names()

cat("Lee PAR-CLIP data columns:\n")
print(colnames(lee_parclip))

cat("\nSubunit distribution:\n")
print(table(lee_parclip$subunit))

# Filter for eIF3d-specific binding and get unique gene IDs
lee_eif3d_genes <- lee_parclip %>%
  filter(subunit == "d") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")  # Strip version numbers if present

cat("\nNumber of eIF3d-bound genes (Lee PAR-CLIP):", length(lee_eif3d_genes), "\n")
```

```{r classify_lee_binding}
# Classify genes by Lee eIF3d binding status
te_1hr_lee_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

te_4hr_lee_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

te_24hr_lee_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

# Report overlap statistics
cat("\nLee eIF3d-bound genes overlap with TE data:\n")
cat("1hr - eIF3d-bound genes found:", sum(te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
cat("4hr - eIF3d-bound genes found:", sum(te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
cat("24hr - eIF3d-bound genes found:", sum(te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
```

```{r lee_ecdf_plot, fig.width=10, fig.height=4}
# Combine all timepoints for faceted plot
te_lee_combined <- bind_rows(
  te_1hr_lee_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_lee_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_lee_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels_lee <- te_lee_combined %>%
  group_by(timepoint, binding_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      binding_status == "eIF3d-Bound (Lee)" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot with green color for Lee data
p_lee_faceted <- ggplot(te_lee_combined, aes(x = log2FoldChange, color = binding_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("eIF3d-Bound (Lee)" = "#009E73", "Other" = "#808080"),
    name = "eIF3d Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels_lee, aes(x = x, y = y, label = label, color = binding_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Binding Status (Lee PAR-CLIP)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_lee_faceted)

# Save the Lee PAR-CLIP faceted plot
ggsave(
  here("plots", "ecdf_eif3d_lee_parclip_all_timepoints.pdf"),
  plot = p_lee_faceted,
  width = 10,
  height = 4
)
```

```{r lee_statistical_tests}
# Perform KS tests comparing Lee eIF3d-bound vs other genes
ks_lee_1hr <- ks.test(
  te_1hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_1hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

ks_lee_4hr <- ks.test(
  te_4hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_4hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

ks_lee_24hr <- ks.test(
  te_24hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_24hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

# Perform Wilcoxon tests using rstatix
wilcox_lee_1hr <- te_1hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

wilcox_lee_4hr <- te_4hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

wilcox_lee_24hr <- te_24hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

# Compile Lee PAR-CLIP statistics
lee_stats_results <- data.frame(
  timepoint = c("1hr", "4hr", "24hr"),
  n_eif3d_bound = c(
    sum(te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"),
    sum(te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"),
    sum(te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)")
  ),
  n_other = c(
    sum(te_1hr_lee_status$binding_status == "Other"),
    sum(te_4hr_lee_status$binding_status == "Other"),
    sum(te_24hr_lee_status$binding_status == "Other")
  ),
  median_eif3d_bound = c(
    median(te_1hr_lee_status$log2FoldChange[te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"]),
    median(te_4hr_lee_status$log2FoldChange[te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"]),
    median(te_24hr_lee_status$log2FoldChange[te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)"])
  ),
  median_other = c(
    median(te_1hr_lee_status$log2FoldChange[te_1hr_lee_status$binding_status == "Other"]),
    median(te_4hr_lee_status$log2FoldChange[te_4hr_lee_status$binding_status == "Other"]),
    median(te_24hr_lee_status$log2FoldChange[te_24hr_lee_status$binding_status == "Other"])
  ),
  ks_statistic = c(ks_lee_1hr$statistic, ks_lee_4hr$statistic, ks_lee_24hr$statistic),
  ks_pvalue = c(ks_lee_1hr$p.value, ks_lee_4hr$p.value, ks_lee_24hr$p.value),
  wilcox_statistic = c(wilcox_lee_1hr$statistic, wilcox_lee_4hr$statistic, wilcox_lee_24hr$statistic),
  wilcox_pvalue = c(wilcox_lee_1hr$p, wilcox_lee_4hr$p, wilcox_lee_24hr$p)
)

cat("\nLee PAR-CLIP Statistical Test Results:\n")
print(lee_stats_results)

# Export Lee PAR-CLIP statistics
write_csv(lee_stats_results, here("output", "lee_parclip_statistics.csv"))
cat("\nLee PAR-CLIP statistics exported to: output/lee_parclip_statistics.csv\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
