---
title: "CLIP Binding Sites Analysis - Lamper Cap-Binding Data"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(cowplot)
library(readxl)
library(janitor)
library(rstatix)
library(ggvenn) 
```

## Overview

This analysis examines whether eIF3d cap-binding targets (from Lamper et al. glucose deprivation experiments) show differential hypoxic response compared to other genes.

## Load TE Changes from Hypoxia Timecourse

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

# Add gene symbols to TE data and prepare for analysis
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Loaded TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes\n")
cat("4hr:", nrow(te_4hr_gsea), "genes\n")
cat("24hr:", nrow(te_24hr_gsea), "genes\n")
```

## Load Lamper Cap-Binding Data

```{r load_lamper_data}
# Load Lamper et al. cap-binding data (glucose deprivation cap binders)
lamper_data <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/lamper-s3.xlsx") %>%
  clean_names()

cat("Lamper data columns after clean_names():\n")
print(colnames(lamper_data))

cat("\nLamper data dimensions:", nrow(lamper_data), "rows x", ncol(lamper_data), "columns\n")
head(lamper_data)
```

## Define Cap-Binding Gene Sets

Using thresholds from original analysis:
- Glucose deprivation cap binders: `log2fold_change_glu > 2 & padj_glu < 0.005`
- Complete media cap binders: `log2fold_change_cm > 2 & padj_cm < 0.005`

```{r define_cap_binders}
# Convert columns to numeric (they may be stored as character in Excel)
lamper_data <- lamper_data %>%
  mutate(
    log2fold_change_cm = as.numeric(log2fold_change_cm),
    log2fold_change_glu = as.numeric(log2fold_change_glu),
    padj_cm = as.numeric(padj_cm),
    padj_glu = as.numeric(padj_glu)
  )

# Define glucose deprivation cap binders (GLU_HIT)
glu_cap_binders <- lamper_data %>%
  filter(log2fold_change_glu > 2 & padj_glu < 0.005) %>%
  pull(ensembl_gene_id)

# Define complete media cap binders (CM_HIT)
cm_cap_binders <- lamper_data %>%
  filter(log2fold_change_cm > 2 & padj_cm < 0.005) %>%
  pull(ensembl_gene_id)

cat("Number of glucose deprivation cap binders (GLU_HIT):", length(glu_cap_binders), "\n")
cat("Number of complete media cap binders (CM_HIT):", length(cm_cap_binders), "\n")
cat("Overlap between GLU and CM:", length(intersect(glu_cap_binders, cm_cap_binders)), "\n")
```

## Classify Genes by Cap-Binding Status

```{r classify_genes}
# Strip version numbers from gene IDs for matching
# TE data has versioned IDs (e.g., ENSG00000000003.17)
# Lamper data may have unversioned IDs

# Clean gene IDs in TE data
te_1hr_cap_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

te_4hr_cap_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

te_24hr_cap_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glucose Dep. Cap Binder",
      TRUE ~ "Other"
    )
  )

# Report overlap statistics
cat("\nOverlap with TE data:\n")
cat("1hr - Glucose cap binders found:", sum(te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
cat("4hr - Glucose cap binders found:", sum(te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
cat("24hr - Glucose cap binders found:", sum(te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder"), "\n")
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (1 Hour)

```{r ecdf_1hr, fig.width=6, fig.height=5}
# Create ECDF plot for 1 hour
p_1hr <- ggplot(te_1hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 1 Hour",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_1hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_1hr.pdf"),
  plot = p_1hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (4 Hours)

```{r ecdf_4hr, fig.width=6, fig.height=5}
# Create ECDF plot for 4 hours
p_4hr <- ggplot(te_4hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 4 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_4hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_4hr.pdf"),
  plot = p_4hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by Cap-Binding Status (24 Hours)

```{r ecdf_24hr, fig.width=6, fig.height=5}
# Create ECDF plot for 24 hours
p_24hr <- ggplot(te_24hr_cap_status, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    subtitle = "Translation Efficiency Changes at 24 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_24hr)

# Save the plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_24hr.pdf"),
  plot = p_24hr,
  width = 5,
  height = 4
)
```

## Statistical Tests

```{r statistical_tests}
# Perform KS tests comparing cap binders vs other genes
ks_1hr <- ks.test(
  te_1hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_1hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

ks_4hr <- ks.test(
  te_4hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_4hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

ks_24hr <- ks.test(
  te_24hr_cap_status %>% filter(cap_status == "Glucose Dep. Cap Binder") %>% pull(log2FoldChange),
  te_24hr_cap_status %>% filter(cap_status == "Other") %>% pull(log2FoldChange)
)

# Perform Wilcoxon tests using rstatix
wilcox_1hr <- te_1hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

wilcox_4hr <- te_4hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

wilcox_24hr <- te_24hr_cap_status %>%
  wilcox_test(log2FoldChange ~ cap_status)

# Compile statistics
stats_results <- data.frame(
  timepoint = c("1hr", "4hr", "24hr"),
  n_cap_binders = c(
    sum(te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"),
    sum(te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"),
    sum(te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder")
  ),
  n_other = c(
    sum(te_1hr_cap_status$cap_status == "Other"),
    sum(te_4hr_cap_status$cap_status == "Other"),
    sum(te_24hr_cap_status$cap_status == "Other")
  ),
  median_cap_binders = c(
    median(te_1hr_cap_status$log2FoldChange[te_1hr_cap_status$cap_status == "Glucose Dep. Cap Binder"]),
    median(te_4hr_cap_status$log2FoldChange[te_4hr_cap_status$cap_status == "Glucose Dep. Cap Binder"]),
    median(te_24hr_cap_status$log2FoldChange[te_24hr_cap_status$cap_status == "Glucose Dep. Cap Binder"])
  ),
  median_other = c(
    median(te_1hr_cap_status$log2FoldChange[te_1hr_cap_status$cap_status == "Other"]),
    median(te_4hr_cap_status$log2FoldChange[te_4hr_cap_status$cap_status == "Other"]),
    median(te_24hr_cap_status$log2FoldChange[te_24hr_cap_status$cap_status == "Other"])
  ),
  ks_statistic = c(ks_1hr$statistic, ks_4hr$statistic, ks_24hr$statistic),
  ks_pvalue = c(ks_1hr$p.value, ks_4hr$p.value, ks_24hr$p.value),
  wilcox_statistic = c(wilcox_1hr$statistic, wilcox_4hr$statistic, wilcox_24hr$statistic),
  wilcox_pvalue = c(wilcox_1hr$p, wilcox_4hr$p, wilcox_24hr$p)
)

cat("\nStatistical Test Results:\n")
print(stats_results)

# Export statistics
write_csv(stats_results, here("output", "cap_binder_statistics.csv"))
cat("\nStatistics exported to: output/cap_binder_statistics.csv\n")
```

## Faceted All-Timepoints Plot

```{r faceted_plot, fig.width=10, fig.height=4}
# Combine all timepoints for faceted plot
te_combined <- bind_rows(
  te_1hr_cap_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_cap_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_cap_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels <- te_combined %>%
  group_by(timepoint, cap_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      cap_status == "Glucose Dep. Cap Binder" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot
p_faceted <- ggplot(te_combined, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Glucose Dep. Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels, aes(x = x, y = y, label = label, color = cap_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_faceted)

# Save the faceted plot
ggsave(
  here("plots", "ecdf_cap_binders_glu_all_timepoints.pdf"),
  plot = p_faceted,
  width = 10,
  height = 4
)
```

## Complete Media Cap Binders Analysis

For comparison, also analyze complete media cap binders.

```{r cm_cap_binders, fig.width=10, fig.height=4}
# Classify genes by CM cap-binding status
te_1hr_cm_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

te_4hr_cm_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

te_24hr_cm_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    cap_status = case_when(
      gene_id_clean %in% cm_cap_binders ~ "Complete Media Cap Binder",
      TRUE ~ "Other"
    )
  )

cat("\nComplete Media Cap Binder overlap with TE data:\n")
cat("1hr:", sum(te_1hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")
cat("4hr:", sum(te_4hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")
cat("24hr:", sum(te_24hr_cm_status$cap_status == "Complete Media Cap Binder"), "\n")

# Combine all timepoints for faceted plot
te_cm_combined <- bind_rows(
  te_1hr_cm_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_cm_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_cm_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels_cm <- te_cm_combined %>%
  group_by(timepoint, cap_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      cap_status == "Complete Media Cap Binder" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot for CM cap binders
p_cm_faceted <- ggplot(te_cm_combined, aes(x = log2FoldChange, color = cap_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("Complete Media Cap Binder" = "#0072B2", "Other" = "#808080"),
    name = "Cap-Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels_cm, aes(x = x, y = y, label = label, color = cap_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Cap-Binding Status (Lamper - Complete Media)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_cm_faceted)

# Save the CM faceted plot
ggsave(
  here("plots", "ecdf_cap_binders_cm_all_timepoints.pdf"),
  plot = p_cm_faceted,
  width = 10,
  height = 4
)
```

## Lee PAR-CLIP eIF3d Binding Analysis

Analyze eIF3d binding targets from Lee et al. 2015 PAR-CLIP data.

```{r load_lee_parclip}
# Load Lee 2015 PAR-CLIP data
lee_parclip <- read_csv("/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv") %>%
  clean_names()

cat("Lee PAR-CLIP data columns:\n")
print(colnames(lee_parclip))

cat("\nSubunit distribution:\n")
print(table(lee_parclip$subunit))

# Filter for eIF3d-specific binding and get unique gene IDs
lee_eif3d_genes <- lee_parclip %>%
  filter(subunit == "d") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")  # Strip version numbers if present

cat("\nNumber of eIF3d-bound genes (Lee PAR-CLIP):", length(lee_eif3d_genes), "\n")
```

```{r classify_lee_binding}
# Classify genes by Lee eIF3d binding status
te_1hr_lee_status <- te_1hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

te_4hr_lee_status <- te_4hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

te_24hr_lee_status <- te_24hr_gsea %>%
  mutate(gene_id_clean = str_replace(gene_id, "\\..*", "")) %>%
  mutate(
    binding_status = case_when(
      gene_id_clean %in% lee_eif3d_genes ~ "eIF3d-Bound (Lee)",
      TRUE ~ "Other"
    )
  )

# Report overlap statistics
cat("\nLee eIF3d-bound genes overlap with TE data:\n")
cat("1hr - eIF3d-bound genes found:", sum(te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
cat("4hr - eIF3d-bound genes found:", sum(te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
cat("24hr - eIF3d-bound genes found:", sum(te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)"), "\n")
```

```{r lee_ecdf_plot, fig.width=10, fig.height=4}
# Combine all timepoints for faceted plot
te_lee_combined <- bind_rows(
  te_1hr_lee_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_lee_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_lee_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels_lee <- te_lee_combined %>%
  group_by(timepoint, binding_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      binding_status == "eIF3d-Bound (Lee)" ~ 0.15,
      TRUE ~ 0.05
    )
  )

# Create faceted ECDF plot with green color for Lee data
p_lee_faceted <- ggplot(te_lee_combined, aes(x = log2FoldChange, color = binding_status)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("eIF3d-Bound (Lee)" = "#009E73", "Other" = "#808080"),
    name = "eIF3d Binding Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels_lee, aes(x = x, y = y, label = label, color = binding_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by eIF3d Binding Status (Lee PAR-CLIP)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_lee_faceted)

# Save the Lee PAR-CLIP faceted plot
ggsave(
  here("plots", "ecdf_eif3d_lee_parclip_all_timepoints.pdf"),
  plot = p_lee_faceted,
  width = 10,
  height = 4
)
```

```{r lee_statistical_tests}
# Perform KS tests comparing Lee eIF3d-bound vs other genes
ks_lee_1hr <- ks.test(
  te_1hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_1hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

ks_lee_4hr <- ks.test(
  te_4hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_4hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

ks_lee_24hr <- ks.test(
  te_24hr_lee_status %>% filter(binding_status == "eIF3d-Bound (Lee)") %>% pull(log2FoldChange),
  te_24hr_lee_status %>% filter(binding_status == "Other") %>% pull(log2FoldChange)
)

# Perform Wilcoxon tests using rstatix
wilcox_lee_1hr <- te_1hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

wilcox_lee_4hr <- te_4hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

wilcox_lee_24hr <- te_24hr_lee_status %>%
  wilcox_test(log2FoldChange ~ binding_status)

# Compile Lee PAR-CLIP statistics
lee_stats_results <- data.frame(
  timepoint = c("1hr", "4hr", "24hr"),
  n_eif3d_bound = c(
    sum(te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"),
    sum(te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"),
    sum(te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)")
  ),
  n_other = c(
    sum(te_1hr_lee_status$binding_status == "Other"),
    sum(te_4hr_lee_status$binding_status == "Other"),
    sum(te_24hr_lee_status$binding_status == "Other")
  ),
  median_eif3d_bound = c(
    median(te_1hr_lee_status$log2FoldChange[te_1hr_lee_status$binding_status == "eIF3d-Bound (Lee)"]),
    median(te_4hr_lee_status$log2FoldChange[te_4hr_lee_status$binding_status == "eIF3d-Bound (Lee)"]),
    median(te_24hr_lee_status$log2FoldChange[te_24hr_lee_status$binding_status == "eIF3d-Bound (Lee)"])
  ),
  median_other = c(
    median(te_1hr_lee_status$log2FoldChange[te_1hr_lee_status$binding_status == "Other"]),
    median(te_4hr_lee_status$log2FoldChange[te_4hr_lee_status$binding_status == "Other"]),
    median(te_24hr_lee_status$log2FoldChange[te_24hr_lee_status$binding_status == "Other"])
  ),
  ks_statistic = c(ks_lee_1hr$statistic, ks_lee_4hr$statistic, ks_lee_24hr$statistic),
  ks_pvalue = c(ks_lee_1hr$p.value, ks_lee_4hr$p.value, ks_lee_24hr$p.value),
  wilcox_statistic = c(wilcox_lee_1hr$statistic, wilcox_lee_4hr$statistic, wilcox_lee_24hr$statistic),
  wilcox_pvalue = c(wilcox_lee_1hr$p, wilcox_lee_4hr$p, wilcox_lee_24hr$p)
)

cat("\nLee PAR-CLIP Statistical Test Results:\n")
print(lee_stats_results)

# Export Lee PAR-CLIP statistics
write_csv(lee_stats_results, here("output", "lee_parclip_statistics.csv"))
cat("\nLee PAR-CLIP statistics exported to: output/lee_parclip_statistics.csv\n")
```

## Session Info

```{r session_info}
sessionInfo()
```

## CLIP vs eIF3d/eIF3e TE Regulation Overlap Analysis

This section analyzes the overlap between eIF3d CLIP binding targets and genes whose translation efficiency (TE) is promoted or inhibited by eIF3d/eIF3e.

### Load TE Regulation Gene Sets

```{r load_te_regulation_data}
# Helper function to clean gene IDs (strip version numbers)
clean_gene_id <- function(gene_id) {
  str_replace(gene_id, "\\..*", "")
}

# Load gene sets from combined signatures file
all_sigs <- read_csv(here("accessories", "all_combined_sigs_1_22_26.csv"))

# Filter for MDA-MB-231 lfc0.5 gene sets
mda_sigs <- all_sigs %>%
  filter(str_detect(gs_name, "^MDA-MB-231")) %>%
  filter(str_detect(gs_name, "lfc0.5"))

cat("Loaded", nrow(mda_sigs), "total gene-signature pairs\n")
cat("Unique signatures:", length(unique(mda_sigs$gs_name)), "\n")
```

### Define TE Regulation Gene Sets

Extract gene sets for both promotes and inhibits directions from the combined signatures file.

```{r define_te_regulation_genes}
# Helper function to extract gene set by pattern matching gs_name
get_geneset <- function(sigs, pattern) {
  sigs %>%
    filter(str_detect(gs_name, pattern)) %>%
    pull(ensembl_gene) %>%
    unique()
}

# Extract all 16 gene sets (2 factors x 2 directions x (3 hypoxia + 1 normoxia))

# eIF3d promotes TE - hypoxia
te_3d_promotes_hypoxia_1hr <- get_geneset(mda_sigs, "hypoxia_3d_promotes_TE_1hr_lfc0.5")
te_3d_promotes_hypoxia_4hr <- get_geneset(mda_sigs, "hypoxia_3d_promotes_TE_4hr_lfc0.5")
te_3d_promotes_hypoxia_24hr <- get_geneset(mda_sigs, "hypoxia_3d_promotes_TE_24hr_lfc0.5")

# eIF3d inhibits TE - hypoxia
te_3d_inhibits_hypoxia_1hr <- get_geneset(mda_sigs, "hypoxia_3d_inhibits_TE_1hr_lfc0.5")
te_3d_inhibits_hypoxia_4hr <- get_geneset(mda_sigs, "hypoxia_3d_inhibits_TE_4hr_lfc0.5")
te_3d_inhibits_hypoxia_24hr <- get_geneset(mda_sigs, "hypoxia_3d_inhibits_TE_24hr_lfc0.5")

# eIF3e promotes TE - hypoxia
te_3e_promotes_hypoxia_1hr <- get_geneset(mda_sigs, "hypoxia_3e_promotes_TE_1hr_lfc0.5")
te_3e_promotes_hypoxia_4hr <- get_geneset(mda_sigs, "hypoxia_3e_promotes_TE_4hr_lfc0.5")
te_3e_promotes_hypoxia_24hr <- get_geneset(mda_sigs, "hypoxia_3e_promotes_TE_24hr_lfc0.5")

# eIF3e inhibits TE - hypoxia
te_3e_inhibits_hypoxia_1hr <- get_geneset(mda_sigs, "hypoxia_3e_inhibits_TE_1hr_lfc0.5")
te_3e_inhibits_hypoxia_4hr <- get_geneset(mda_sigs, "hypoxia_3e_inhibits_TE_4hr_lfc0.5")
te_3e_inhibits_hypoxia_24hr <- get_geneset(mda_sigs, "hypoxia_3e_inhibits_TE_24hr_lfc0.5")

# eIF3d normoxia (alltimepoints)
te_3d_promotes_normoxia <- get_geneset(mda_sigs, "normoxia_3d_promotes_TE_alltimepoints_lfc0.5")
te_3d_inhibits_normoxia <- get_geneset(mda_sigs, "normoxia_3d_inhibits_TE_alltimepoints_lfc0.5")

# eIF3e normoxia (alltimepoints)
te_3e_promotes_normoxia <- get_geneset(mda_sigs, "normoxia_3e_promotes_TE_alltimepoints_lfc0.5")
te_3e_inhibits_normoxia <- get_geneset(mda_sigs, "normoxia_3e_inhibits_TE_alltimepoints_lfc0.5")

cat("Number of genes in each TE regulation gene set:\n\n")
cat("eIF3d promotes TE:\n")
cat("  hypoxia 1hr:", length(te_3d_promotes_hypoxia_1hr), "\n")
cat("  hypoxia 4hr:", length(te_3d_promotes_hypoxia_4hr), "\n")
cat("  hypoxia 24hr:", length(te_3d_promotes_hypoxia_24hr), "\n")
cat("  normoxia:", length(te_3d_promotes_normoxia), "\n")

cat("\neIF3d inhibits TE:\n")
cat("  hypoxia 1hr:", length(te_3d_inhibits_hypoxia_1hr), "\n")
cat("  hypoxia 4hr:", length(te_3d_inhibits_hypoxia_4hr), "\n")
cat("  hypoxia 24hr:", length(te_3d_inhibits_hypoxia_24hr), "\n")
cat("  normoxia:", length(te_3d_inhibits_normoxia), "\n")

cat("\neIF3e promotes TE:\n")
cat("  hypoxia 1hr:", length(te_3e_promotes_hypoxia_1hr), "\n")
cat("  hypoxia 4hr:", length(te_3e_promotes_hypoxia_4hr), "\n")
cat("  hypoxia 24hr:", length(te_3e_promotes_hypoxia_24hr), "\n")
cat("  normoxia:", length(te_3e_promotes_normoxia), "\n")

cat("\neIF3e inhibits TE:\n")
cat("  hypoxia 1hr:", length(te_3e_inhibits_hypoxia_1hr), "\n")
cat("  hypoxia 4hr:", length(te_3e_inhibits_hypoxia_4hr), "\n")
cat("  hypoxia 24hr:", length(te_3e_inhibits_hypoxia_24hr), "\n")
cat("  normoxia:", length(te_3e_inhibits_normoxia), "\n")
```

### Define Overlap Analysis Function

```{r overlap_analysis_function}
# Helper function to run overlap analysis
run_overlap_analysis <- function(clip_genes, te_genes, universe,
                                  clip_label, te_label) {
  # Get overlapping genes
  overlap_genes <- intersect(clip_genes, te_genes)

  # Build 2x2 contingency table
  # Rows: CLIP (yes/no), Cols: TE-regulated (yes/no)
  a <- length(overlap_genes)  # CLIP+ and TE-regulated+
  b <- length(setdiff(clip_genes, te_genes))  # CLIP+ and TE-regulated-
  c <- length(setdiff(te_genes, clip_genes))  # CLIP- and TE-regulated+
  d <- length(universe) - a - b - c  # CLIP- and TE-regulated-

  contingency_table <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)

  # Run Fisher's exact test
  fisher_result <- fisher.test(contingency_table)

  # Return results as a tibble
  tibble(
    clip_dataset = clip_label,
    te_condition = te_label,
    n_clip = length(clip_genes),
    n_te_regulated = length(te_genes),
    n_overlap = a,
    n_universe = length(universe),
    odds_ratio = fisher_result$estimate,
    conf_low = fisher_result$conf.int[1],
    conf_high = fisher_result$conf.int[2],
    pvalue = fisher_result$p.value,
    overlap_genes = list(overlap_genes)
  )
}
```

### Run All Overlap Comparisons

```{r run_overlap_comparisons}
# Define the universe as all unique genes in the MDA-MB-231 signatures
universe <- mda_sigs %>%
  pull(ensembl_gene) %>%
  unique()

cat("Universe size:", length(universe), "genes\n")

# Define CLIP gene sets (already defined earlier, but ensure they're clean)
glu_cap_clean <- unique(glu_cap_binders)
cm_cap_clean <- unique(cm_cap_binders)
lee_eif3d_clean <- unique(lee_eif3d_genes)

# Filter CLIP genes to those in universe
glu_cap_in_universe <- intersect(glu_cap_clean, universe)
cm_cap_in_universe <- intersect(cm_cap_clean, universe)
lee_eif3d_in_universe <- intersect(lee_eif3d_clean, universe)

cat("\nCLIP genes in universe:\n")
cat("Glucose dep cap binders:", length(glu_cap_in_universe), "\n")
cat("Complete media cap binders:", length(cm_cap_in_universe), "\n")
cat("Lee PAR-CLIP eIF3d:", length(lee_eif3d_in_universe), "\n")

# Define TE regulation gene sets list (16 gene sets: 2 factors x 2 directions x 4 conditions)
te_regulation_list <- list(
  # eIF3d promotes TE
  "3d_promotes_hypoxia_1hr" = te_3d_promotes_hypoxia_1hr,
  "3d_promotes_hypoxia_4hr" = te_3d_promotes_hypoxia_4hr,
  "3d_promotes_hypoxia_24hr" = te_3d_promotes_hypoxia_24hr,
  "3d_promotes_normoxia" = te_3d_promotes_normoxia,
  # eIF3d inhibits TE
  "3d_inhibits_hypoxia_1hr" = te_3d_inhibits_hypoxia_1hr,
  "3d_inhibits_hypoxia_4hr" = te_3d_inhibits_hypoxia_4hr,
  "3d_inhibits_hypoxia_24hr" = te_3d_inhibits_hypoxia_24hr,
  "3d_inhibits_normoxia" = te_3d_inhibits_normoxia,
  # eIF3e promotes TE
  "3e_promotes_hypoxia_1hr" = te_3e_promotes_hypoxia_1hr,
  "3e_promotes_hypoxia_4hr" = te_3e_promotes_hypoxia_4hr,
  "3e_promotes_hypoxia_24hr" = te_3e_promotes_hypoxia_24hr,
  "3e_promotes_normoxia" = te_3e_promotes_normoxia,
  # eIF3e inhibits TE
  "3e_inhibits_hypoxia_1hr" = te_3e_inhibits_hypoxia_1hr,
  "3e_inhibits_hypoxia_4hr" = te_3e_inhibits_hypoxia_4hr,
  "3e_inhibits_hypoxia_24hr" = te_3e_inhibits_hypoxia_24hr,
  "3e_inhibits_normoxia" = te_3e_inhibits_normoxia
)

# Define CLIP datasets list
clip_list <- list(
  glu_cap_binders = glu_cap_in_universe,
  cm_cap_binders = cm_cap_in_universe,
  lee_parclip = lee_eif3d_in_universe
)

# Run all comparisons (3 CLIP x 16 TE = 48 comparisons)
all_results <- tibble()

for (clip_name in names(clip_list)) {
  for (te_name in names(te_regulation_list)) {
    result <- run_overlap_analysis(
      clip_genes = clip_list[[clip_name]],
      te_genes = te_regulation_list[[te_name]],
      universe = universe,
      clip_label = clip_name,
      te_label = te_name
    )
    all_results <- bind_rows(all_results, result)
  }
}

# Apply BH multiple testing correction
all_results <- all_results %>%
  mutate(padj = p.adjust(pvalue, method = "BH"))

# Add parsed condition info for plotting
all_results <- all_results %>%
  mutate(
    factor = case_when(
      str_detect(te_condition, "3d") ~ "eIF3d",
      str_detect(te_condition, "3e") ~ "eIF3e"
    ),
    regulation = case_when(
      str_detect(te_condition, "promotes") ~ "Promotes TE",
      str_detect(te_condition, "inhibits") ~ "Inhibits TE"
    ),
    condition = case_when(
      str_detect(te_condition, "hypoxia") ~ "Hypoxia",
      str_detect(te_condition, "normoxia") ~ "Normoxia"
    ),
    timepoint = case_when(
      str_detect(te_condition, "_1hr") ~ "1hr",
      str_detect(te_condition, "_4hr") ~ "4hr",
      str_detect(te_condition, "_24hr") ~ "24hr",
      str_detect(te_condition, "normoxia") ~ "All"
    )
  )

# Display results
cat("\nOverlap analysis results:\n")
print(all_results %>% dplyr::select(-overlap_genes))
```

### Generate Venn Diagrams

```{r generate_venn_diagrams, fig.width=5, fig.height=5}
# Generate Venn diagrams for each CLIP/TE regulation combination
for (clip_name in names(clip_list)) {
  for (te_name in names(te_regulation_list)) {
    # Create Venn list
    venn_list <- list(
      CLIP = clip_list[[clip_name]],
      TE_regulated = te_regulation_list[[te_name]]
    )

    # Create nice labels
    clip_label <- case_when(
      clip_name == "glu_cap_binders" ~ "Glu Cap Binders",
      clip_name == "cm_cap_binders" ~ "CM Cap Binders",
      clip_name == "lee_parclip" ~ "Lee PAR-CLIP"
    )

    # Parse TE regulation label
    te_label <- te_name %>%
      str_replace("3d_", "eIF3d ") %>%
      str_replace("3e_", "eIF3e ") %>%
      str_replace("_", " ") %>%
      str_replace("_", " ")

    names(venn_list) <- c(clip_label, paste0("TE\n(", te_label, ")"))

    # Create plot
    p_venn <- ggvenn(
      venn_list,
      fill_color = c("#D55E00", "#0072B2"),
      stroke_size = 0.5,
      set_name_size = 3,
      text_size = 4
    ) +
      ggtitle(paste0(clip_label, " vs ", te_label)) +
      theme(plot.title = element_text(hjust = 0.5, size = 10))

    # Save plot
    filename <- paste0("venn_", clip_name, "_vs_", te_name, ".pdf")
    ggsave(
      here("plots", filename),
      plot = p_venn,
      width = 5,
      height = 5
    )
  }
}

cat("Venn diagrams saved to plots/ directory\n")
```

### Generate Odds Ratio Bar Charts

```{r odds_ratio_bar_charts, fig.width=12, fig.height=8}
# Format p-value for display
format_pval <- function(p) {
  case_when(
    p < 0.001 ~ formatC(p, format = "e", digits = 1),
    p < 0.01 ~ formatC(p, format = "f", digits = 3),
    TRUE ~ formatC(p, format = "f", digits = 2)
  )
}

# Prepare plot data with formatted p-values and significance-based fill
plot_data <- all_results %>%
  mutate(
    pval_label = format_pval(padj),
    n_genes_label = as.character(n_te_regulated),
    significant = padj < 0.05,
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    factor = factor(factor, levels = c("eIF3d", "eIF3e")),
    regulation = factor(regulation, levels = c("Promotes TE", "Inhibits TE")),
    # Create fill category: factor name if significant, "ns" if not
    fill_group = ifelse(significant, as.character(factor), "ns")
  )

# Create separate plots for each CLIP dataset
for (clip_name in unique(plot_data$clip_dataset)) {
  clip_label <- case_when(
    clip_name == "glu_cap_binders" ~ "Glucose Deprivation Cap Binders (Lamper)",
    clip_name == "cm_cap_binders" ~ "Complete Media Cap Binders (Lamper)",
    clip_name == "lee_parclip" ~ "eIF3d PAR-CLIP (Lee)"
  )

  # Get the number of CLIP genes for the subtitle
  n_clip_genes <- plot_data %>%
    filter(clip_dataset == clip_name) %>%
    pull(n_clip) %>%
    unique()

  plot_subset <- plot_data %>%
    filter(clip_dataset == clip_name)

  p_or <- ggplot(plot_subset, aes(x = timepoint, y = odds_ratio, fill = fill_group, color = factor, group = factor)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, linewidth = 1) +
    geom_text(
      aes(y = odds_ratio + 0.1, label = n_genes_label, group = factor),
      position = position_dodge(width = 0.8),
      size = 2.5,
      vjust = 0,
      color = "black",
      show.legend = FALSE
    ) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    facet_grid(regulation ~ condition, scales = "free_x") +
    scale_fill_manual(
      values = c("eIF3d" = "#E69F00", "eIF3e" = "#56B4E9", "ns" = "white"),
      guide = "none"
    ) +
    scale_color_manual(
      values = c("eIF3d" = "#E69F00", "eIF3e" = "#56B4E9"),
      name = "Factor"
    ) +
    labs(
      title = paste0("CLIP Binding vs TE Regulation Overlap: ", clip_label),
      subtitle = paste0("CLIP gene set: n = ", n_clip_genes, " genes | Numbers above bars = TE gene set size | Filled = padj < 0.05"),
      x = "Timepoint",
      y = "Odds Ratio"
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
      plot.subtitle = element_text(hjust = 0.5, size = 9),
      strip.text = element_text(face = "bold")
    )

  print(p_or)

  # Save plot
  filename <- paste0("odds_ratio_", clip_name, ".pdf")
  ggsave(
    here("plots", filename),
    plot = p_or,
    width = 12,
    height = 8
  )
}

cat("Odds ratio bar charts saved to plots/ directory\n")
```

### Export Statistics

```{r export_overlap_statistics}
# Prepare export data (remove list column)
export_data <- all_results %>%
  dplyr::select(-overlap_genes)

# Export to CSV
write_csv(export_data, here("output", "clip_te_regulation_overlap_statistics.csv"))

cat("\nStatistics exported to: output/clip_te_regulation_overlap_statistics.csv\n")
cat("\nSummary of significant overlaps (padj < 0.05):\n")
significant_results <- export_data %>%
  filter(padj < 0.05) %>%
  arrange(padj)
print(significant_results)
```

## Lee PAR-CLIP Regional Binding Analysis

This section analyzes the overlap between eIF3d binding at different mRNA regions (5' UTR, CDS, 3' UTR) and TE regulation gene sets.

### Load and Process Regional Binding Data

```{r load_regional_binding}
# Load Lee PAR-CLIP data with region annotations
lee_parclip_regions <- read_csv("/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv") %>%
  clean_names()

# Filter for eIF3d and get unique genes per region
lee_3d_5utr <- lee_parclip_regions %>%
  filter(subunit == "d" & annotation == "5' UTR") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")

lee_3d_cds <- lee_parclip_regions %>%
  filter(subunit == "d" & annotation == "CDS exon") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")

lee_3d_3utr <- lee_parclip_regions %>%
  filter(subunit == "d" & annotation == "3' UTR") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")

cat("eIF3d binding by region (unique genes):\n")
cat("5' UTR:", length(lee_3d_5utr), "genes\n")
cat("CDS:", length(lee_3d_cds), "genes\n")
cat("3' UTR:", length(lee_3d_3utr), "genes\n")

# Filter to genes in universe
lee_3d_5utr_in_universe <- intersect(lee_3d_5utr, universe)
lee_3d_cds_in_universe <- intersect(lee_3d_cds, universe)
lee_3d_3utr_in_universe <- intersect(lee_3d_3utr, universe)

cat("\neIF3d binding in universe:\n")
cat("5' UTR:", length(lee_3d_5utr_in_universe), "genes\n")
cat("CDS:", length(lee_3d_cds_in_universe), "genes\n")
cat("3' UTR:", length(lee_3d_3utr_in_universe), "genes\n")
```

### Run Regional Overlap Analysis

```{r regional_overlap_analysis}
# Define regional CLIP gene sets
regional_clip_list <- list(
  "5' UTR" = lee_3d_5utr_in_universe,
  "CDS" = lee_3d_cds_in_universe,
  "3' UTR" = lee_3d_3utr_in_universe
)

# Run all comparisons (3 regions x 16 TE gene sets = 48 comparisons)
regional_results <- tibble()

for (region_name in names(regional_clip_list)) {
  for (te_name in names(te_regulation_list)) {
    result <- run_overlap_analysis(
      clip_genes = regional_clip_list[[region_name]],
      te_genes = te_regulation_list[[te_name]],
      universe = universe,
      clip_label = region_name,
      te_label = te_name
    )
    regional_results <- bind_rows(regional_results, result)
  }
}

# Apply BH multiple testing correction
regional_results <- regional_results %>%
  mutate(padj = p.adjust(pvalue, method = "BH"))

# Add parsed condition info for plotting
regional_results <- regional_results %>%
  mutate(
    region = clip_dataset,
    factor = case_when(
      str_detect(te_condition, "3d") ~ "eIF3d",
      str_detect(te_condition, "3e") ~ "eIF3e"
    ),
    regulation = case_when(
      str_detect(te_condition, "promotes") ~ "Promotes TE",
      str_detect(te_condition, "inhibits") ~ "Inhibits TE"
    ),
    condition = case_when(
      str_detect(te_condition, "hypoxia") ~ "Hypoxia",
      str_detect(te_condition, "normoxia") ~ "Normoxia"
    ),
    timepoint = case_when(
      str_detect(te_condition, "_1hr") ~ "1hr",
      str_detect(te_condition, "_4hr") ~ "4hr",
      str_detect(te_condition, "_24hr") ~ "24hr",
      str_detect(te_condition, "normoxia") ~ "All"
    )
  )

cat("\nRegional overlap analysis results:\n")
print(regional_results %>% dplyr::select(-overlap_genes))
```

### Generate Regional Odds Ratio Bar Charts

```{r regional_odds_ratio_charts, fig.width=14, fig.height=10}
# Prepare plot data
regional_plot_data <- regional_results %>%
  mutate(
    pval_label = format_pval(padj),
    n_genes_label = as.character(n_te_regulated),
    significant = padj < 0.05,
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    factor = factor(factor, levels = c("eIF3d", "eIF3e")),
    regulation = factor(regulation, levels = c("Promotes TE", "Inhibits TE")),
    region = factor(region, levels = c("5' UTR", "CDS", "3' UTR")),
    # Create fill category: region name if significant, "ns" if not
    fill_group = ifelse(significant, as.character(region), "ns")
  )

# Get region gene counts for subtitle
region_counts <- paste0(
  "5' UTR: ", length(lee_3d_5utr_in_universe), " genes | ",
  "CDS: ", length(lee_3d_cds_in_universe), " genes | ",
  "3' UTR: ", length(lee_3d_3utr_in_universe), " genes"
)

# Create plot for eIF3d-regulated genes only (since we're looking at eIF3d binding)
plot_3d_regional <- regional_plot_data %>%
  filter(factor == "eIF3d")

p_regional_3d <- ggplot(plot_3d_regional, aes(x = timepoint, y = odds_ratio, fill = fill_group, color = region, group = region)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, linewidth = 1) +
  geom_text(
    aes(y = odds_ratio + 0.1, label = n_genes_label, group = region),
    position = position_dodge(width = 0.8),
    size = 2.5,
    vjust = 0,
    color = "black",
    show.legend = FALSE
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  facet_grid(regulation ~ condition, scales = "free_x") +
  scale_fill_manual(
    values = c("5' UTR" = "#009E73", "CDS" = "#CC79A7", "3' UTR" = "#F0E442", "ns" = "white"),
    guide = "none"
  ) +
  scale_color_manual(
    values = c("5' UTR" = "#009E73", "CDS" = "#CC79A7", "3' UTR" = "#F0E442"),
    name = "Binding Region"
  ) +
  labs(
    title = "eIF3d Binding Region vs eIF3d TE Regulation Overlap (Lee PAR-CLIP)",
    subtitle = paste0(region_counts, " | Numbers = TE gene set size | Filled = padj < 0.05"),
    x = "Timepoint",
    y = "Odds Ratio"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    strip.text = element_text(face = "bold")
  )

print(p_regional_3d)

ggsave(
  here("plots", "odds_ratio_lee_parclip_regional_3d.pdf"),
  plot = p_regional_3d,
  width = 14,
  height = 8
)

# Also create plot for eIF3e-regulated genes (to see if 3d binding affects 3e targets)
plot_3e_regional <- regional_plot_data %>%
  filter(factor == "eIF3e")

p_regional_3e <- ggplot(plot_3e_regional, aes(x = timepoint, y = odds_ratio, fill = fill_group, color = region, group = region)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7, linewidth = 1) +
  geom_text(
    aes(y = odds_ratio + 0.1, label = n_genes_label, group = region),
    position = position_dodge(width = 0.8),
    size = 2.5,
    vjust = 0,
    color = "black",
    show.legend = FALSE
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  facet_grid(regulation ~ condition, scales = "free_x") +
  scale_fill_manual(
    values = c("5' UTR" = "#009E73", "CDS" = "#CC79A7", "3' UTR" = "#F0E442", "ns" = "white"),
    guide = "none"
  ) +
  scale_color_manual(
    values = c("5' UTR" = "#009E73", "CDS" = "#CC79A7", "3' UTR" = "#F0E442"),
    name = "Binding Region"
  ) +
  labs(
    title = "eIF3d Binding Region vs eIF3e TE Regulation Overlap (Lee PAR-CLIP)",
    subtitle = paste0(region_counts, " | Numbers = TE gene set size | Filled = padj < 0.05"),
    x = "Timepoint",
    y = "Odds Ratio"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    plot.subtitle = element_text(hjust = 0.5, size = 9),
    strip.text = element_text(face = "bold")
  )

print(p_regional_3e)

ggsave(
  here("plots", "odds_ratio_lee_parclip_regional_3e.pdf"),
  plot = p_regional_3e,
  width = 14,
  height = 8
)

cat("Regional odds ratio bar charts saved to plots/ directory\n")
```

### Export Regional Statistics

```{r export_regional_statistics}
# Prepare export data
regional_export <- regional_results %>%
  dplyr::select(-overlap_genes)

# Export to CSV
write_csv(regional_export, here("output", "lee_parclip_regional_overlap_statistics.csv"))

cat("\nRegional statistics exported to: output/lee_parclip_regional_overlap_statistics.csv\n")
cat("\nSummary of significant regional overlaps (padj < 0.05):\n")
significant_regional <- regional_export %>%
  filter(padj < 0.05) %>%
  arrange(padj)
print(significant_regional)
```
