---
title: "Comparison of DESeq2 Strategies: Rmd vs Shiny App"
author: "Kate"
date: "2025-12-09"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

This analysis compares two DESeq2 strategies:
1. The deFunction approach used in the rna_ribo_te_analysis.Rmd
2. The approach built into the Shiny app

We're testing siCtrl vs si3e in normoxia at 1 hour, comparing:
- TE analysis WITH spike-ins
- TE analysis WITHOUT spike-ins

We count significantly regulated genes (padj < 0.05, no LFC cutoff) and compare to the Shiny app results.

```{r load_libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(janitor)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded deFunction and related functions\n")
```

# Load Data

## Load raw counts from Shiny app

```{r load_raw_counts}
# Load the raw filtered counts from the Shiny app
raw_counts <- read_csv("~/Downloads/raw_filtered_counts_2025-12-09.csv") %>%
  column_to_rownames("gene_id")

cat("Raw counts dimensions:", dim(raw_counts), "\n")
```

## Extract human genes from raw counts

The raw counts file from the Shiny app only contains human genes (drosophila spike-ins are in a separate file).

```{r separate_genes}
# The raw counts should only have human genes (ENSG IDs)
gene_ids <- rownames(raw_counts)
is_human <- grepl("^ENSG", gene_ids)

cat("Total genes in raw counts:", length(gene_ids), "\n")
cat("Human genes (ENSG):", sum(is_human), "\n")

# Extract human genes (should be all of them)
human_counts <- raw_counts[is_human, ]

cat("\nHuman counts dimensions:", dim(human_counts), "\n")
cat("All genes are human:", all(is_human), "\n")
```

## Load and filter metadata for siCtrl vs si3e comparison at 1hr normoxia

```{r create_metadata}
# Load the full metadata file
metadata_full <- read_csv(here("output", "metadata_clean.csv"))

# Filter for sictrl and si3e at normoxia 1hr
metadata_comparison <- metadata_full %>%
  filter(treatment %in% c("sictrl", "si3e"),
         condition == "normoxia",
         hours == 1)

# The raw counts have column names like: siCTRL_normoxia_1_a_ribo
# We need to create sample names that match this format AND add the sample ID
# Format: {TREATMENT}_{condition}_{hours}_{rep}_{library}_{id}

metadata_comparison <- metadata_comparison %>%
  mutate(
    # Convert treatment to match raw counts format (sictrl -> siCTRL)
    treatment_formatted = case_when(
      treatment == "sictrl" ~ "siCTRL",
      treatment == "si3e" ~ "si3e",
      treatment == "si3d" ~ "si3d",
      TRUE ~ treatment
    ),
    # Create sample name matching raw counts format with ID at the end
    # Use .data$rep to avoid conflict with rep() function
    sample_name_with_id = paste0(treatment_formatted, "_", condition, "_", hours, "_", .data$rep, "_", library, "_", id),
    # Also keep the raw counts column name for matching
    sample_name_raw = paste0(treatment_formatted, "_", condition, "_", hours, "_", .data$rep, "_", library)
  )

print(metadata_comparison %>% select(id, treatment, library, rep, sample_name_raw, sample_name_with_id))

# Get the sample names (with IDs) for downstream analysis
sample_names <- metadata_comparison$sample_name_with_id

# But use the raw format names to extract from the raw counts
sample_names_raw <- metadata_comparison$sample_name_raw

# CRITICAL: Match Shiny app behavior - filter out genes with rowSums <= 1 ACROSS ALL SAMPLES
# The Shiny app applies this filter to ALL samples first, then subsets to comparison samples
# This is done in server.R line 1997, but raw_counts_mat has ALL samples at that point
cat("\nHuman counts dimensions (before rowSums filter):", dim(human_counts), "\n")

# Remove symbol column if present, we only need numeric counts for filtering
if ("symbol" %in% colnames(human_counts)) {
  human_counts_numeric <- human_counts %>% select(-symbol)
} else {
  human_counts_numeric <- human_counts
}

# Make rownames unique and filter by rowSums across ALL samples
rownames(human_counts_numeric) <- make.unique(rownames(human_counts_numeric))
human_counts_filtered <- human_counts_numeric[rowSums(human_counts_numeric) > 1, ]

cat("Human counts dimensions (after removing rowSums <= 1 across all samples):", dim(human_counts_filtered), "\n")

# NOW filter to just the comparison samples
counts_comparison <- human_counts_filtered[, sample_names_raw]

cat("Filtered counts dimensions (after subsetting to comparison samples):", dim(counts_comparison), "\n")
cat("Column names in counts:\n")
print(colnames(counts_comparison))
cat("Counts contain only human genes (no drosophila):", !any(grepl("^FBgn", rownames(counts_comparison))), "\n")

# Update metadata to use the raw sample names (without IDs) for matching with counts
# CRITICAL: Set comparison_variable to match Shiny app format: treatment_condition_hours
# This matches what the Shiny app does (e.g., "siCTRL_normoxia_1" or "si3e_normoxia_1")
metadata_comparison <- metadata_comparison %>%
  select(id, library, replicate = rep, treatment, treatment_formatted, condition, hours, sample_name = sample_name_raw) %>%
  mutate(comparison_variable = paste(treatment_formatted, condition, hours, sep = "_")) %>%
  select(-treatment_formatted)

cat("\nMetadata structure:\n")
print(metadata_comparison)
cat("\nComparison variable values:", unique(metadata_comparison$comparison_variable), "\n")
cat("These should be: siCTRL_normoxia_1 and si3e_normoxia_1\n")
```

## Prepare spike-in data (top 500 drosophila genes)

```{r load_spikein}
# The raw_filtered_counts file doesn't contain drosophila genes
# (they're removed by the Shiny app's raw_counts() reactive)
# Instead, load the spike-in matrix that was exported by the Shiny app
# This contains the 500 drosophila spike-in genes used for normalization

# Load spike-in matrix from Shiny app export
spikein_matrix_full <- read_csv("/Users/katematlin/github/ribopipe-shiny/spikeInMatrix.csv") %>%
  column_to_rownames("gene_id")

cat("Spike-in matrix dimensions (all samples):", dim(spikein_matrix_full), "\n")

# Filter to our comparison samples
spikein_comparison <- spikein_matrix_full[, sample_names_raw]

cat("Spike-in matrix dimensions (comparison samples):", dim(spikein_comparison), "\n")

# Get the spike-in IDs (should be 500)
top500_ids <- rownames(spikein_comparison)

cat("Number of spike-in genes:", length(top500_ids), "\n")

# CRITICAL: Ensure spike-in matrix columns match the filtered counts_comparison
if (!identical(colnames(spikein_comparison), colnames(counts_comparison))) {
  cat("WARNING: Spike-in columns don't match counts columns. Reordering...\n")
  spikein_comparison <- spikein_comparison[, colnames(counts_comparison), drop = FALSE]
}

cat("All spike-ins are drosophila:", all(grepl("^FBgn", top500_ids)), "\n")
cat("Spike-in sample names match count matrix:", identical(colnames(spikein_comparison), colnames(counts_comparison)), "\n")

# Verify these are drosophila genes (should start with "FBgn")
cat("First few spike-in gene IDs:", head(top500_ids), "\n")
```

## Create normalized matrix for both strategies

We need a normalized matrix for deFunction. We'll create two versions:
1. WITH spike-ins: Uses top 500 drosophila genes for size factor estimation
2. WITHOUT spike-ins: Standard DESeq2 normalization (no drosophila)

```{r create_norm_matrices}
# Create normalized matrix WITH spike-ins
# This will combine human counts with drosophila spike-ins, use spike-ins for normalization
# Use the SAME control_condition format as the Shiny app
normMatrix_withSpikes <- createNormMatrix(
  allCounts = counts_comparison,
  metadata = metadata_comparison,
  control_condition = "siCTRL_normoxia_1",
  spikein = spikein_comparison,
  keepSpikes = top500_ids
)

cat("Normalized matrix WITH spike-ins dimensions:", dim(normMatrix_withSpikes), "\n")
cat("Contains drosophila genes:", any(grepl("^FBgn", rownames(normMatrix_withSpikes))), "\n")

# Create normalized matrix WITHOUT spike-ins
# This will only have human genes, no drosophila
# Use the SAME control_condition format as the Shiny app
normMatrix_noSpikes <- createNormMatrix(
  allCounts = counts_comparison,
  metadata = metadata_comparison,
  control_condition = "siCTRL_normoxia_1",
  spikein = NULL,
  keepSpikes = NULL
)

is_dros_withSpikes <- grepl("^FBgn", rownames(normMatrix_withSpikes))
  if (any(is_dros_withSpikes)) {
    normMatrix_withSpikes <- normMatrix_withSpikes[!is_dros_withSpikes, , drop = FALSE]
    cat("Removed", sum(is_dros_withSpikes), "drosophila genes from spike-in normalized matrix\n")
  }


cat("\nNormalized matrix WITHOUT spike-ins dimensions:", dim(normMatrix_noSpikes), "\n")
cat("Contains drosophila genes:", any(grepl("^FBgn", rownames(normMatrix_noSpikes))), "\n")

cat("\nNormalized matrices created successfully\n")
```

# Run DESeq2 Analysis

## Strategy 1: WITH Spike-ins

```{r deseq_with_spikes}
# Use the SAME comparison_variable format as the Shiny app
res_te_withSpikes <- deFunction(
  countsMatrix = counts_comparison,
  allCounts_Norm = normMatrix_withSpikes,
  metadata = metadata_comparison,
  control_condition = "siCTRL_normoxia_1",
  test_condition = "si3e_normoxia_1",
  spikeInMatrix = spikein_comparison,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("\n=== WITH SPIKE-INS RESULTS ===\n")
cat("Total genes:", nrow(res_te_withSpikes), "\n")

# Count significant genes with padj < 0.05 (no LFC cutoff)
sig_withSpikes <- res_te_withSpikes %>%
  filter(!is.na(padj), padj < 0.05)

n_sig_withSpikes <- nrow(sig_withSpikes)
n_up_withSpikes <- sum(sig_withSpikes$log2FoldChange > 0)
n_down_withSpikes <- sum(sig_withSpikes$log2FoldChange < 0)

cat("Significant genes (padj < 0.05):", n_sig_withSpikes, "\n")
cat("  Up-regulated:", n_up_withSpikes, "\n")
cat("  Down-regulated:", n_down_withSpikes, "\n")
```

## Strategy 2: WITHOUT Spike-ins

```{r deseq_without_spikes}
# Use the SAME comparison_variable format as the Shiny app
res_te_noSpikes <- deFunction(
  countsMatrix = counts_comparison,
  allCounts_Norm = normMatrix_noSpikes,
  metadata = metadata_comparison,
  control_condition = "siCTRL_normoxia_1",
  test_condition = "si3e_normoxia_1",
  spikeInMatrix = NULL,
  keepSpikes = NULL,
  mode = "TE"
)

cat("\n=== WITHOUT SPIKE-INS RESULTS ===\n")
cat("Total genes:", nrow(res_te_noSpikes), "\n")

# Count significant genes with padj < 0.05 (no LFC cutoff)
sig_noSpikes <- res_te_noSpikes %>%
  filter(!is.na(padj), padj < 0.05)

n_sig_noSpikes <- nrow(sig_noSpikes)
n_up_noSpikes <- sum(sig_noSpikes$log2FoldChange > 0)
n_down_noSpikes <- sum(sig_noSpikes$log2FoldChange < 0)

cat("Significant genes (padj < 0.05):", n_sig_noSpikes, "\n")
cat("  Up-regulated:", n_up_noSpikes, "\n")
cat("  Down-regulated:", n_down_noSpikes, "\n")
```

# Load Shiny App Results

```{r load_app_results}
# Load TE results from Shiny app
app_spike <- read_csv("~/Downloads/spike_delta_te.csv")
app_nospike <- read_csv("~/Downloads/no_spikedelta_te.csv")

cat("\n=== SHINY APP RESULTS ===\n")

# Count significant genes from app (padj < 0.05, no LFC cutoff)
app_spike_sig <- app_spike %>%
  filter(!is.na(padj), padj < 0.05)

n_app_spike_sig <- nrow(app_spike_sig)
n_app_spike_up <- sum(app_spike_sig$log2FoldChange > 0)
n_app_spike_down <- sum(app_spike_sig$log2FoldChange < 0)

cat("\nApp WITH spike-ins:\n")
cat("  Total significant:", n_app_spike_sig, "\n")
cat("  Up-regulated:", n_app_spike_up, "\n")
cat("  Down-regulated:", n_app_spike_down, "\n")

app_nospike_sig <- app_nospike %>%
  filter(!is.na(padj), padj < 0.05)

n_app_nospike_sig <- nrow(app_nospike_sig)
n_app_nospike_up <- sum(app_nospike_sig$log2FoldChange > 0)
n_app_nospike_down <- sum(app_nospike_sig$log2FoldChange < 0)

cat("\nApp WITHOUT spike-ins:\n")
cat("  Total significant:", n_app_nospike_sig, "\n")
cat("  Up-regulated:", n_app_nospike_up, "\n")
cat("  Down-regulated:", n_app_nospike_down, "\n")
```

# Comparison Summary

```{r comparison_table}
# Create comparison table
comparison_df <- data.frame(
  Method = c("Rmd with spikes", "Rmd no spikes", "App with spikes", "App no spikes"),
  Total_Significant = c(n_sig_withSpikes, n_sig_noSpikes, n_app_spike_sig, n_app_nospike_sig),
  Up_Regulated = c(n_up_withSpikes, n_up_noSpikes, n_app_spike_up, n_app_nospike_up),
  Down_Regulated = c(n_down_withSpikes, n_down_noSpikes, n_app_spike_down, n_app_nospike_down)
)

cat("\n=== COMPARISON SUMMARY ===\n")
print(comparison_df)

# Calculate differences
cat("\n=== DIFFERENCES ===\n")
cat("Rmd WITH spikes vs App WITH spikes:\n")
cat("  Difference in total significant:", n_sig_withSpikes - n_app_spike_sig, "\n")
cat("  Difference in up-regulated:", n_up_withSpikes - n_app_spike_up, "\n")
cat("  Difference in down-regulated:", n_down_withSpikes - n_app_spike_down, "\n")

cat("\nRmd NO spikes vs App NO spikes:\n")
cat("  Difference in total significant:", n_sig_noSpikes - n_app_nospike_sig, "\n")
cat("  Difference in up-regulated:", n_up_noSpikes - n_app_nospike_up, "\n")
cat("  Difference in down-regulated:", n_down_noSpikes - n_app_nospike_down, "\n")
```

# Visualization

```{r plot_comparison}
library(ggplot2)

# Reshape for plotting
comparison_long <- comparison_df %>%
  pivot_longer(cols = c(Up_Regulated, Down_Regulated),
               names_to = "Direction",
               values_to = "Count")

# Create bar plot
ggplot(comparison_long, aes(x = Method, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Up_Regulated" = "#E74C3C", "Down_Regulated" = "#3498DB")) +
  labs(title = "Comparison of DESeq2 Strategies: siCtrl vs si3e (normoxia 1hr)",
       subtitle = "TE Analysis: Significant genes with padj < 0.05",
       x = "Method",
       y = "Number of Genes",
       fill = "Direction") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("plots/deseq_strategy_comparison.pdf"), width = 8, height = 6)
```

# Detailed Gene-Level Comparison

Let's look at which genes are called significant in each method:

```{r gene_overlap}
# Get gene lists
genes_rmd_spikes <- res_te_withSpikes %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene_id)

genes_rmd_nospikes <- res_te_noSpikes %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene_id)

genes_app_spikes <- app_spike %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene_id)

genes_app_nospikes <- app_nospike %>%
  filter(!is.na(padj), padj < 0.05) %>%
  pull(gene_id)

# Compare overlap for WITH spikes
overlap_spikes <- length(intersect(genes_rmd_spikes, genes_app_spikes))
only_rmd_spikes <- length(setdiff(genes_rmd_spikes, genes_app_spikes))
only_app_spikes <- length(setdiff(genes_app_spikes, genes_rmd_spikes))

cat("\n=== OVERLAP: WITH SPIKE-INS ===\n")
cat("Common to both methods:", overlap_spikes, "\n")
cat("Only in Rmd method:", only_rmd_spikes, "\n")
cat("Only in App method:", only_app_spikes, "\n")

# Compare overlap for WITHOUT spikes
overlap_nospikes <- length(intersect(genes_rmd_nospikes, genes_app_nospikes))
only_rmd_nospikes <- length(setdiff(genes_rmd_nospikes, genes_app_nospikes))
only_app_nospikes <- length(setdiff(genes_app_nospikes, genes_rmd_nospikes))

cat("\n=== OVERLAP: WITHOUT SPIKE-INS ===\n")
cat("Common to both methods:", overlap_nospikes, "\n")
cat("Only in Rmd method:", only_rmd_nospikes, "\n")
cat("Only in App method:", only_app_nospikes, "\n")
```

# Save Results

```{r save_results}
# Save comparison summary
write_csv(comparison_df, here("output", "deseq_strategy_comparison_summary.csv"))

# Save detailed results
write_csv(res_te_withSpikes, here("output", "te_results_rmd_withSpikes_sictrl_vs_si3e_1hr.csv"))
write_csv(res_te_noSpikes, here("output", "te_results_rmd_noSpikes_sictrl_vs_si3e_1hr.csv"))

cat("\nResults saved to output/\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
