---
title: "DAP5 Polysome Profiling Analysis"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(cowplot)
library(janitor)
library(rstatix)
library(DESeq2)
```

## Overview

This analysis examines whether DAP5-dependent genes (from polysome profiling data, GSE115142) show differential hypoxic response compared to other genes. DAP5 (eIF4G2) is an alternative translation initiation factor that can promote cap-independent translation.

We use DESeq2 with an interaction model to identify genes with significantly different translation efficiency (TE) between DAP5 shRNA knockdown and non-silencing (NS) control conditions.

## Load Polysome Data

```{r load_polysome_data}
# Load polysome profiling data from GSE115142
# Columns: gene symbol, then count data for DAP5 overexpression and control (NS) conditions
polysome_raw <- read_tsv(
  here("accessories", "GSE115142_quant.featurecounts.counts.rev.cleaned.trDap5_alternative.txt"),
  col_names = TRUE,
  col_types = paste0("c", strrep("d", 14)),
  show_col_types = FALSE
)

# The first column is unnamed - it contains gene symbols
colnames(polysome_raw)[1] <- "symbol"

cat("Polysome data dimensions:", nrow(polysome_raw), "genes x", ncol(polysome_raw), "columns\n")
cat("\nColumn names:\n")
print(colnames(polysome_raw))
```

## Fix Excel-Corrupted Gene Symbols

Excel auto-converts gene symbols that resemble dates (e.g. MARCH1 → Mar-1, SEPT1 → Sep-1).
Detect and restore them before any downstream steps.

```{r fix_excel_symbols}
# Known Excel date-to-gene mappings (Mon-N format)
# MARCH family: Mar-1 .. Mar-12  -> MARCH1 .. MARCH12
# SEPT family:  Sep-1 .. Sep-15  -> SEPT1  .. SEPT15
# DEC family:   Dec-1, Dec-2     -> DEC1, DEC2

excel_symbol_map <- c(
  setNames(paste0("MARCH", 1:12), paste0(1:12, "-Mar")),
  setNames(paste0("SEPT",  1:15), paste0(1:15, "-Sep")),
  setNames(paste0("DEC",   1:2),  paste0(1:2,  "-Dec"))
)

# Detect any corrupted symbols present in the data
corrupted <- polysome_raw[["symbol"]][polysome_raw[["symbol"]] %in% names(excel_symbol_map)]
cat("Corrupted symbols detected:", length(corrupted), "\n")
if (length(corrupted) > 0) {
  cat("  Found:", paste(sort(corrupted), collapse = ", "), "\n")
  cat("  Fixed:", paste(excel_symbol_map[sort(corrupted)], collapse = ", "), "\n")
}

# Apply fix
polysome_raw[["symbol"]] <- ifelse(
  polysome_raw[["symbol"]] %in% names(excel_symbol_map),
  excel_symbol_map[polysome_raw[["symbol"]]],
  polysome_raw[["symbol"]]
)

cat("\nSymbol column class after fix:", class(polysome_raw[["symbol"]]), "\n")

# Check library sizes
lib_sizes <- colSums(polysome_raw %>% select(where(is.numeric)), na.rm = TRUE)
cat("\nLibrary sizes:\n")
print(lib_sizes)
cat("\nLibrary size fold difference (max/min):", max(lib_sizes) / min(lib_sizes), "\n")
```

## DESeq2 Analysis for Differential Translation Efficiency

We use DESeq2 with an interaction model to test for genes where the Heavy/Total ratio differs between DAP5 and control conditions.

Design: ~ condition + fraction + condition:fraction

The interaction term (condition:fraction) tests for differential TE.

```{r deseq2_setup}
# Select only Heavy (HD) and Total (T) fractions for TE analysis
# DAP5: TR_R4_Dap5_HD, TR_R4_Dap5_T, TR_R5_Dap5_HD, TR_R5_Dap5_T
# NS: TR_R11_NS_HD, TR_R11_NS_T, TR_R3_NS_HD, TR_R3_NS_T

# Create count matrix with only HD and T samples
count_cols <- c("TR_R4_Dap5_HD", "TR_R4_Dap5_T",
                "TR_R5_Dap5_HD", "TR_R5_Dap5_T",
                "TR_R11_NS_HD", "TR_R11_NS_T",
                "TR_R3_NS_HD", "TR_R3_NS_T")

# Check for duplicate gene symbols (base R to avoid S4Vectors dispatch on %in%)
sym_vec <- polysome_raw[["symbol"]]
dup_syms <- unique(sym_vec[duplicated(sym_vec)])
cat("Duplicate symbols found:", length(dup_syms), "\n")
if (length(dup_syms) > 0) {
  cat("\nDuplicated entries:\n")
  dup_rows <- polysome_raw[sym_vec %in% dup_syms, c("symbol", count_cols)]
  print(dup_rows[order(dup_rows[["symbol"]]), ])
}

# Aggregate duplicate gene symbols by summing counts
polysome_aggregated <- polysome_raw %>%
  dplyr::select(symbol, all_of(count_cols)) %>%
  group_by(symbol) %>%
  summarise(across(all_of(count_cols), sum)) %>%
  ungroup()

count_matrix <- polysome_aggregated %>%
  column_to_rownames("symbol") %>%
  as.matrix()

# Create sample metadata
sample_info <- data.frame(
  sample = count_cols,
  condition = c("DAP5", "DAP5", "DAP5", "DAP5", "NS", "NS", "NS", "NS"),
  fraction = c("HD", "T", "HD", "T", "HD", "T", "HD", "T"),
  replicate = c("R4", "R4", "R5", "R5", "R11", "R11", "R3", "R3"),
  row.names = count_cols
)

# Set factor levels (DAP5 shKD as reference for condition so results are NS/control vs KD,
# i.e. positive log2FC = higher in control = DAP5-Promoted; T as reference for fraction)
sample_info$condition <- factor(sample_info$condition, levels = c("DAP5", "NS"))
sample_info$fraction <- factor(sample_info$fraction, levels = c("T", "HD"))

cat("Sample metadata:\n")
print(sample_info)
```

```{r deseq2_run}
# Create DESeq2 dataset with interaction design
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = sample_info,
  design = ~ condition + fraction + condition:fraction
)

# Filter low-count genes (at least 10 reads total across samples)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
cat("\nGenes after filtering:", nrow(dds), "\n")

# Run DESeq2
dds <- DESeq(dds)

# Get results for the interaction term (differential TE)
# Comparison: NS control vs DAP5 shKD (positive log2FC = higher TE in control = DAP5-Promoted)
res_interaction <- results(dds, name = "conditionNS.fractionHD")

# Convert to data frame and add gene symbols
deseq_results <- as.data.frame(res_interaction) %>%
  rownames_to_column("symbol") %>%
  arrange(padj)

cat("\nDESeq2 results summary:\n")
summary(res_interaction)

cat("\nTop genes by adjusted p-value:\n")
head(deseq_results, 10)
```

## Validation: Known DAP5-Sensitive and Insensitive Genes

Check that DESeq2 results recapitulate known biology from de la Parra et al. 2018:
sensitive genes should have positive log2FC (TE higher in control than KD);
moderately sensitive genes should have intermediate log2FC;
insensitive genes should cluster near zero.

```{r validate_known_genes, fig.width=7, fig.height=5}
# Gene groups from de la Parra et al. 2018
sensitive_genes <- c(
  "ETS1", "OSMR", "MYCL", "ITGAV", "TP53BP1", "MMP1"
)
moderate_genes <- c(
  "EGFR", "SERPINE2"
)
insensitive_genes <- c(
  "LMNA", "HNRNPD", "GAPDH"
)

# Display labels for plot
gene_labels <- c(
  ETS1 = "ETS1", OSMR = "OSMR", MYCL = "L-Myc",
  ITGAV = "ITGaV", TP53BP1 = "p53-BP1", MMP1 = "MMP1",
  EGFR = "EGFR", SERPINE2 = "SERPINE2",
  LMNA = "Lamin A", HNRNPD = "AUF1", GAPDH = "GAPDH"
)

validation_genes <- c(sensitive_genes, moderate_genes, insensitive_genes)

validation_df <- deseq_results %>%
  filter(symbol %in% validation_genes) %>%
  mutate(
    group = case_when(
      symbol %in% sensitive_genes   ~ "Sensitive",
      symbol %in% moderate_genes    ~ "Moderately sensitive",
      symbol %in% insensitive_genes ~ "Not sensitive"
    ),
    group = factor(group, levels = c("Sensitive", "Moderately sensitive", "Not sensitive")),
    label = gene_labels[symbol],
    label = factor(label, levels = gene_labels[rev(validation_genes)])
  )

cat("Validation gene results:\n")
validation_df %>%
  select(symbol, label, log2FoldChange, lfcSE, padj, group) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  print()

# Bar plot
p_validation <- ggplot(validation_df,
    aes(x = log2FoldChange, y = label, fill = group)) +
  geom_col() +
  geom_errorbarh(aes(xmin = log2FoldChange - lfcSE,
                     xmax = log2FoldChange + lfcSE),
                 height = 0.3, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_text(aes(label = ifelse(is.na(padj), "NA",
                               ifelse(padj < 0.001, "<0.001", round(padj, 3))),
                x = log2FoldChange + ifelse(log2FoldChange >= 0, 0.05, -0.05)),
            hjust = ifelse(validation_df$log2FoldChange >= 0, 0, 1),
            size = 3.5, color = "gray20") +
  scale_fill_manual(
    values = c("Sensitive"           = "#D55E00",
               "Moderately sensitive" = "#E69F00",
               "Not sensitive"        = "#808080"),
    name = NULL
  ) +
  labs(
    title = "TE Change: NS Control vs DAP5 shKD",
    subtitle = "Positive log2FC = higher in control = DAP5-Promoted",
    x = "log2FC (NS control / DAP5 shKD)",
    y = NULL
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_validation)

ggsave(
  here("plots", "validation_dap5_known_genes.pdf"),
  plot = p_validation,
  width = 7, height = 5
)
```

## Define DAP5-Dependent Gene Sets

Using DESeq2 results with thresholds:
- **padj < 0.05** (statistically significant)
- **|log2FC| > 0.5** (~1.4-fold change in TE)

- **DAP5-Promoted**: genes with significant positive interaction (TE increased by DAP5)
- **DAP5-Inhibited**: genes with significant negative interaction (TE decreased by DAP5)

```{r define_dap5_genesets}
# Define thresholds
padj_threshold <- 0.05
log2fc_threshold <- 0.5

# Classify genes by DAP5 dependence using DESeq2 results
deseq_classified <- deseq_results %>%
  mutate(
    dap5_status = case_when(
      !is.na(padj) & padj < padj_threshold & log2FoldChange > log2fc_threshold ~ "DAP5-Promoted",
      !is.na(padj) & padj < padj_threshold & log2FoldChange < -log2fc_threshold ~ "DAP5-Inhibited",
      TRUE ~ "Other"
    )
  )

# Report gene counts
cat("\nDAP5 classification counts (DESeq2, padj <", padj_threshold, "& |log2FC| >", log2fc_threshold, "):\n")
print(table(deseq_classified$dap5_status))

# Extract gene sets
dap5_promoted_genes <- deseq_classified %>%
  filter(dap5_status == "DAP5-Promoted") %>%
  pull(symbol)

dap5_inhibited_genes <- deseq_classified %>%
  filter(dap5_status == "DAP5-Inhibited") %>%
  pull(symbol)

cat("\nDAP5-Promoted genes:", length(dap5_promoted_genes), "\n")
cat("DAP5-Inhibited genes:", length(dap5_inhibited_genes), "\n")

# Show distribution of log2FC for significant genes
cat("\nlog2FC distribution for DAP5-Promoted genes:\n")
print(summary(deseq_classified$log2FoldChange[deseq_classified$dap5_status == "DAP5-Promoted"]))

cat("\nlog2FC distribution for DAP5-Inhibited genes:\n")
print(summary(deseq_classified$log2FoldChange[deseq_classified$dap5_status == "DAP5-Inhibited"]))
```

## Volcano Plot of DESeq2 Results

```{r volcano_plot, fig.width=7, fig.height=6}
# Create volcano plot
p_volcano <- ggplot(deseq_classified, aes(x = log2FoldChange, y = -log10(padj), color = dap5_status)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(
    values = c("DAP5-Promoted" = "#D55E00", "DAP5-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = c(-log2fc_threshold, log2fc_threshold), linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(padj_threshold), linetype = "dashed", color = "gray40") +
  labs(
    title = "DAP5 Effect on Translation Efficiency",
    subtitle = paste0("DESeq2 interaction term (padj < ", padj_threshold, ", |log2FC| > ", log2fc_threshold, ")"),
    x = "log2 Fold Change (TE: NS control / DAP5 shKD)\nright = higher in control = DAP5-Promoted",
    y = "-log10(adjusted p-value)"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-4, 4)

print(p_volcano)

ggsave(
  here("plots", "volcano_dap5_polysome_deseq2.pdf"),
  plot = p_volcano,
  width = 7,
  height = 6
)
```

## Load Hypoxia TE Data

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE)
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"), show_col_types = FALSE)
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"), show_col_types = FALSE)

# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"),
                      show_col_types = FALSE)

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

# Add gene symbols to TE data and prepare for analysis
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
# Note: Using dplyr:: prefix to avoid conflicts with DESeq2/S4Vectors packages
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  dplyr::slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  dplyr::slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  dplyr::slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Loaded TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes\n")
cat("4hr:", nrow(te_4hr_gsea), "genes\n")
cat("24hr:", nrow(te_24hr_gsea), "genes\n")
```

## Classify Genes by DAP5 Status

```{r classify_genes}
# Classify genes by DAP5 dependence status (using DESeq2-defined gene sets)
te_1hr_dap5_status <- te_1hr_gsea %>%
  mutate(
    dap5_status = case_when(
      symbol %in% dap5_promoted_genes ~ "DAP5-Promoted",
      symbol %in% dap5_inhibited_genes ~ "DAP5-Inhibited",
      TRUE ~ "Other"
    )
  )

te_4hr_dap5_status <- te_4hr_gsea %>%
  mutate(
    dap5_status = case_when(
      symbol %in% dap5_promoted_genes ~ "DAP5-Promoted",
      symbol %in% dap5_inhibited_genes ~ "DAP5-Inhibited",
      TRUE ~ "Other"
    )
  )

te_24hr_dap5_status <- te_24hr_gsea %>%
  mutate(
    dap5_status = case_when(
      symbol %in% dap5_promoted_genes ~ "DAP5-Promoted",
      symbol %in% dap5_inhibited_genes ~ "DAP5-Inhibited",
      TRUE ~ "Other"
    )
  )

# Report overlap statistics
cat("\nOverlap with hypoxia TE data:\n")
cat("1hr - DAP5-Promoted found:", sum(te_1hr_dap5_status$dap5_status == "DAP5-Promoted"), "\n")
cat("1hr - DAP5-Inhibited found:", sum(te_1hr_dap5_status$dap5_status == "DAP5-Inhibited"), "\n")
cat("4hr - DAP5-Promoted found:", sum(te_4hr_dap5_status$dap5_status == "DAP5-Promoted"), "\n")
cat("4hr - DAP5-Inhibited found:", sum(te_4hr_dap5_status$dap5_status == "DAP5-Inhibited"), "\n")
cat("24hr - DAP5-Promoted found:", sum(te_24hr_dap5_status$dap5_status == "DAP5-Promoted"), "\n")
cat("24hr - DAP5-Inhibited found:", sum(te_24hr_dap5_status$dap5_status == "DAP5-Inhibited"), "\n")
```

## ECDF Plot: Hypoxic Response by DAP5 Status (1 Hour)

```{r ecdf_1hr, fig.width=6, fig.height=5}
# Create ECDF plot for 1 hour
p_1hr <- ggplot(te_1hr_dap5_status, aes(x = log2FoldChange, color = dap5_status)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(
    values = c("DAP5-Promoted" = "#D55E00", "DAP5-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Dependence",
    subtitle = "Translation Efficiency Changes at 1 Hour",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_1hr)

# Save the plot
ggsave(
  here("plots", "ecdf_dap5_polysome_1hr.pdf"),
  plot = p_1hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by DAP5 Status (4 Hours)

```{r ecdf_4hr, fig.width=6, fig.height=5}
# Create ECDF plot for 4 hours
p_4hr <- ggplot(te_4hr_dap5_status, aes(x = log2FoldChange, color = dap5_status)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(
    values = c("DAP5-Promoted" = "#D55E00", "DAP5-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Dependence",
    subtitle = "Translation Efficiency Changes at 4 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_4hr)

# Save the plot
ggsave(
  here("plots", "ecdf_dap5_polysome_4hr.pdf"),
  plot = p_4hr,
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by DAP5 Status (24 Hours)

```{r ecdf_24hr, fig.width=6, fig.height=5}
# Create ECDF plot for 24 hours
p_24hr <- ggplot(te_24hr_dap5_status, aes(x = log2FoldChange, color = dap5_status)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(
    values = c("DAP5-Promoted" = "#D55E00", "DAP5-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Dependence",
    subtitle = "Translation Efficiency Changes at 24 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

print(p_24hr)

# Save the plot
ggsave(
  here("plots", "ecdf_dap5_polysome_24hr.pdf"),
  plot = p_24hr,
  width = 5,
  height = 4
)
```

## Faceted All-Timepoints Plot

```{r faceted_plot, fig.width=10, fig.height=4}
# Combine all timepoints for faceted plot
te_combined <- bind_rows(
  te_1hr_dap5_status %>% mutate(timepoint = "1 Hour"),
  te_4hr_dap5_status %>% mutate(timepoint = "4 Hours"),
  te_24hr_dap5_status %>% mutate(timepoint = "24 Hours")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

# Calculate counts for annotations
count_labels <- te_combined %>%
  group_by(timepoint, dap5_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = paste0("n=", n)) %>%
  mutate(
    x = 0.7,
    y = case_when(
      dap5_status == "DAP5-Promoted" ~ 0.20,
      dap5_status == "DAP5-Inhibited" ~ 0.10,
      TRUE ~ 0.03
    )
  )

# Create faceted ECDF plot
p_faceted <- ggplot(te_combined, aes(x = log2FoldChange, color = dap5_status)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(
    values = c("DAP5-Promoted" = "#D55E00", "DAP5-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = count_labels, aes(x = x, y = y, label = label, color = dap5_status),
            size = 3, show.legend = FALSE) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Hypoxic Response by DAP5 Dependence (GSE115142, DESeq2)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-3, 1)

print(p_faceted)

# Save the faceted plot
ggsave(
  here("plots", "ecdf_dap5_polysome_all_timepoints.pdf"),
  plot = p_faceted,
  width = 10,
  height = 4
)
```

## Statistical Tests

```{r statistical_tests}
# Helper function to run statistical tests for a given comparison
run_stats_tests <- function(te_data, group1_name, group2_name) {
  group1_vals <- te_data %>% filter(dap5_status == group1_name) %>% pull(log2FoldChange)
  group2_vals <- te_data %>% filter(dap5_status == group2_name) %>% pull(log2FoldChange)

  if (length(group1_vals) < 2 || length(group2_vals) < 2) {
    return(list(
      ks_stat = NA, ks_p = NA,
      wilcox_stat = NA, wilcox_p = NA,
      n1 = length(group1_vals), n2 = length(group2_vals),
      median1 = ifelse(length(group1_vals) > 0, median(group1_vals), NA),
      median2 = ifelse(length(group2_vals) > 0, median(group2_vals), NA)
    ))
  }

  ks_result <- ks.test(group1_vals, group2_vals)
  wilcox_result <- wilcox.test(group1_vals, group2_vals)

  return(list(
    ks_stat = ks_result$statistic,
    ks_p = ks_result$p.value,
    wilcox_stat = wilcox_result$statistic,
    wilcox_p = wilcox_result$p.value,
    n1 = length(group1_vals),
    n2 = length(group2_vals),
    median1 = median(group1_vals),
    median2 = median(group2_vals)
  ))
}

# Run tests for DAP5-Promoted vs Other
stats_promoted_1hr <- run_stats_tests(te_1hr_dap5_status, "DAP5-Promoted", "Other")
stats_promoted_4hr <- run_stats_tests(te_4hr_dap5_status, "DAP5-Promoted", "Other")
stats_promoted_24hr <- run_stats_tests(te_24hr_dap5_status, "DAP5-Promoted", "Other")

# Run tests for DAP5-Inhibited vs Other
stats_inhibited_1hr <- run_stats_tests(te_1hr_dap5_status, "DAP5-Inhibited", "Other")
stats_inhibited_4hr <- run_stats_tests(te_4hr_dap5_status, "DAP5-Inhibited", "Other")
stats_inhibited_24hr <- run_stats_tests(te_24hr_dap5_status, "DAP5-Inhibited", "Other")

# Compile statistics
stats_results <- data.frame(
  timepoint = rep(c("1hr", "4hr", "24hr"), 2),
  comparison = c(rep("DAP5-Promoted vs Other", 3), rep("DAP5-Inhibited vs Other", 3)),
  n_test_group = c(
    stats_promoted_1hr$n1, stats_promoted_4hr$n1, stats_promoted_24hr$n1,
    stats_inhibited_1hr$n1, stats_inhibited_4hr$n1, stats_inhibited_24hr$n1
  ),
  n_other = c(
    stats_promoted_1hr$n2, stats_promoted_4hr$n2, stats_promoted_24hr$n2,
    stats_inhibited_1hr$n2, stats_inhibited_4hr$n2, stats_inhibited_24hr$n2
  ),
  median_test_group = c(
    stats_promoted_1hr$median1, stats_promoted_4hr$median1, stats_promoted_24hr$median1,
    stats_inhibited_1hr$median1, stats_inhibited_4hr$median1, stats_inhibited_24hr$median1
  ),
  median_other = c(
    stats_promoted_1hr$median2, stats_promoted_4hr$median2, stats_promoted_24hr$median2,
    stats_inhibited_1hr$median2, stats_inhibited_4hr$median2, stats_inhibited_24hr$median2
  ),
  ks_statistic = c(
    stats_promoted_1hr$ks_stat, stats_promoted_4hr$ks_stat, stats_promoted_24hr$ks_stat,
    stats_inhibited_1hr$ks_stat, stats_inhibited_4hr$ks_stat, stats_inhibited_24hr$ks_stat
  ),
  ks_pvalue = c(
    stats_promoted_1hr$ks_p, stats_promoted_4hr$ks_p, stats_promoted_24hr$ks_p,
    stats_inhibited_1hr$ks_p, stats_inhibited_4hr$ks_p, stats_inhibited_24hr$ks_p
  ),
  wilcox_statistic = c(
    stats_promoted_1hr$wilcox_stat, stats_promoted_4hr$wilcox_stat, stats_promoted_24hr$wilcox_stat,
    stats_inhibited_1hr$wilcox_stat, stats_inhibited_4hr$wilcox_stat, stats_inhibited_24hr$wilcox_stat
  ),
  wilcox_pvalue = c(
    stats_promoted_1hr$wilcox_p, stats_promoted_4hr$wilcox_p, stats_promoted_24hr$wilcox_p,
    stats_inhibited_1hr$wilcox_p, stats_inhibited_4hr$wilcox_p, stats_inhibited_24hr$wilcox_p
  )
)

cat("\nStatistical Test Results:\n")
print(stats_results)

# Export statistics
write_csv(stats_results, here("output", "dap5_polysome_statistics.csv"))
cat("\nStatistics exported to: output/dap5_polysome_statistics.csv\n")
```

## Export Gene Sets and DESeq2 Results

```{r export_genesets}
# Export full DESeq2 results
write_csv(deseq_results, here("output", "dap5_deseq2_full_results.csv"))
cat("Full DESeq2 results exported to: output/dap5_deseq2_full_results.csv\n")

# Create gene set data frame (only significant genes, |log2FC| > 0.5)
geneset_df <- deseq_classified %>%
  filter(dap5_status != "Other") %>%
  select(symbol, log2FoldChange, padj, dap5_status) %>%
  arrange(dap5_status, desc(abs(log2FoldChange)))

cat("\nGene set summary (DESeq2, padj <", padj_threshold, "& |log2FC| >", log2fc_threshold, "):\n")
print(table(geneset_df$dap5_status))

# Export gene sets
write_csv(geneset_df, here("output", "genesets", "dap5_polysome_genesets.csv"))
cat("\nGene sets exported to: output/genesets/dap5_polysome_genesets.csv\n")

# Create stricter gene set with |log2FC| > 1 threshold
log2fc_threshold_strict <- 1.0

geneset_strict_df <- deseq_results %>%
  mutate(
    dap5_status = case_when(
      !is.na(padj) & padj < padj_threshold & log2FoldChange > log2fc_threshold_strict ~ "DAP5-Promoted",
      !is.na(padj) & padj < padj_threshold & log2FoldChange < -log2fc_threshold_strict ~ "DAP5-Inhibited",
      TRUE ~ "Other"
    )
  ) %>%
  filter(dap5_status != "Other") %>%
  select(symbol, log2FoldChange, padj, dap5_status) %>%
  arrange(dap5_status, desc(abs(log2FoldChange)))

cat("\nStricter gene set summary (DESeq2, padj <", padj_threshold, "& |log2FC| >", log2fc_threshold_strict, "):\n")
print(table(geneset_strict_df$dap5_status))

# Export stricter gene sets
write_csv(geneset_strict_df, here("output", "genesets", "dap5_polysome_genesets_lfc1.csv"))
cat("\nStricter gene sets exported to: output/genesets/dap5_polysome_genesets_lfc1.csv\n")
```

## eIF3d Promotes Gene Set: DAP5 Polysome Score Distribution

### Load gene sets

```{r load_eif3d_genesets}
norm_promotes_1hr <- read_csv(here("output", "genesets", "normoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                              show_col_types = FALSE) %>%
  pull(db_gene_symbol)

hyp_promotes_1hr  <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                              show_col_types = FALSE) %>%
  pull(db_gene_symbol)

cat("Normoxia promotes-3d (1hr, lfc>0.5):", length(norm_promotes_1hr), "genes\n")
cat("Hypoxia promotes-3d (1hr, lfc>0.5):", length(hyp_promotes_1hr), "genes\n")
```

### Helper function (mirrors teleman_comparison.Rmd ip_density_plot)

```{r dap5_density_fn}
dap5_density_plot <- function(dap5_data, gene_syms, cat_label, color, title_str, outfile) {
  plot_data <- dap5_data %>%
    filter(!is.na(log2FoldChange)) %>%
    mutate(category = ifelse(symbol %in% gene_syms, cat_label, "All other genes"))

  target_lfc <- plot_data %>% filter(category == cat_label) %>% pull(log2FoldChange)
  other_lfc  <- plot_data %>% filter(category == "All other genes") %>% pull(log2FoldChange)

  ks_result <- suppressWarnings(ks.test(target_lfc, other_lfc))

  plot_data <- plot_data %>%
    mutate(category_labeled = ifelse(
      category == cat_label,
      sprintf("%s (n=%d)", cat_label, length(target_lfc)),
      sprintf("All other genes (n=%d)", length(other_lfc))
    ))

  color_vals <- setNames(
    c(color, "grey50"),
    c(sprintf("%s (n=%d)", cat_label, length(target_lfc)),
      sprintf("All other genes (n=%d)", length(other_lfc)))
  )

  p <- ggplot(plot_data, aes(x = log2FoldChange, color = category_labeled)) +
    geom_density(linewidth = 1) +
    scale_color_manual(values = color_vals) +
    labs(
      title    = title_str,
      subtitle = sprintf("KS test p-value = %.2e", ks_result$p.value),
      x        = "log2FC (TE: NS control / DAP5 shKD)\nright = higher in control = DAP5-Promoted",
      y        = "Density",
      color    = NULL
    ) +
    theme_cowplot() +
    theme(legend.position = "top")

  print(p)
  ggsave(here("plots", outfile), width = 8, height = 6)
  cat(title_str, "- KS p:", ks_result$p.value, "\n")
}
```

### Normoxia promotes-3d (1hr, lfc>0.5)

```{r dap5_density_norm, fig.width=8, fig.height=6}
dap5_density_plot(
  deseq_results,
  norm_promotes_1hr,
  "Upregulated by eIF3d normoxia (1hr, lfc>0.5)",
  "#E69F00",
  "1hr normoxia: DAP5 Polysome Score (LFC > 0.5)",
  "dap5_polysome_density_3d_norm_1hr_lfc05.pdf"
)
```

### Hypoxia promotes-3d (1hr, lfc>0.5)

```{r dap5_density_hyp, fig.width=8, fig.height=6}
dap5_density_plot(
  deseq_results,
  hyp_promotes_1hr,
  "Upregulated by eIF3d hypoxia (1hr, lfc>0.5)",
  "#E69F00",
  "1hr hypoxia: DAP5 Polysome Score (LFC > 0.5)",
  "dap5_polysome_density_3d_hyp_1hr_lfc05.pdf"
)
```

## Venn Diagram: eIF3d Promotes vs DAP5 Gene Sets

```{r venn_eif3d_dap5, fig.width=12, fig.height=5}
library(ggVennDiagram)

venn_norm <- list(
  "eIF3d promotes\nnormoxia (1hr)" = norm_promotes_1hr,
  "DAP5-promoted"                  = dap5_promoted_genes,
  "DAP5-inhibited"                 = dap5_inhibited_genes
)

venn_hyp <- list(
  "eIF3d promotes\nhypoxia (1hr)" = hyp_promotes_1hr,
  "DAP5-promoted"                 = dap5_promoted_genes,
  "DAP5-inhibited"                = dap5_inhibited_genes
)

p_venn_norm <- ggVennDiagram(venn_norm, label_alpha = 0, set_color = c("#D55E00", "grey40", "grey40")) +
  scale_fill_gradient(low = "white", high = "grey80") +
  labs(title = "Normoxia eIF3d promotes vs DAP5 sets") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

p_venn_hyp <- ggVennDiagram(venn_hyp, label_alpha = 0, set_color = c("#009E73", "grey40", "grey40")) +
  scale_fill_gradient(low = "white", high = "grey80") +
  labs(title = "Hypoxia eIF3d promotes vs DAP5 sets") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

p_venn_combined <- plot_grid(p_venn_norm, p_venn_hyp, nrow = 1, labels = c("A", "B"))
print(p_venn_combined)

ggsave(here("plots", "venn_eif3d_dap5_genesets.pdf"),
       plot = p_venn_combined, width = 12, height = 5)
```

## Session Info

```{r session_info}
sessionInfo()
```
