---
title: "ECDF Plots: eIF3d and eIF3e Gene Sets"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis creates ECDF plots to visualize translational changes during hypoxia in control cells, colored by custom eIF3d/eIF3e gene sets.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggplot2)
library(ggnewscale)
```

# Load TE Data

Load translation efficiency changes in siCtrl cells (hypoxia vs normoxia) at all three timepoints.

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

cat("Loaded TE data:\n")
cat("1hr:", nrow(te_1hr), "genes\n")
cat("4hr:", nrow(te_4hr), "genes\n")
cat("24hr:", nrow(te_24hr), "genes\n")
```

# Load 3d KD Comparison Data

Load the si3d vs sictrl comparison data for each hypoxia timepoint (for LFC color scaling).

```{r load_3d_kd_data}
# Load si3d vs sictrl comparisons (for gradient coloring)
si3d_hypoxia_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

si3d_hypoxia_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

si3d_hypoxia_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

cat("Loaded 3d KD comparison data:\n")
cat("1hr:", nrow(si3d_hypoxia_1hr), "genes\n")
cat("4hr:", nrow(si3d_hypoxia_4hr), "genes\n")
cat("24hr:", nrow(si3d_hypoxia_24hr), "genes\n")

# Load si3e vs sictrl comparisons (for gradient coloring)
si3e_hypoxia_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

si3e_hypoxia_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

si3e_hypoxia_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

cat("\nLoaded 3e KD comparison data:\n")
cat("1hr:", nrow(si3e_hypoxia_1hr), "genes\n")
cat("4hr:", nrow(si3e_hypoxia_4hr), "genes\n")
cat("24hr:", nrow(si3e_hypoxia_24hr), "genes\n")
```

# Load Gene Annotation

```{r load_gene_annotation}
# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

# Prepare TE Data with Gene Symbols

```{r prepare_te_data}
# Add gene symbols to TE data
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Prepared TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes with symbols\n")
cat("4hr:", nrow(te_4hr_gsea), "genes with symbols\n")
cat("24hr:", nrow(te_24hr_gsea), "genes with symbols\n")
```

# Load Custom Gene Sets

```{r load_custom_genesets}
# Load all custom gene sets
custom_genesets <- read_csv(here("accessories", "all_combined_sigs_1_22_26.csv"))

cat("Loaded custom gene sets:\n")
cat("Total entries:", nrow(custom_genesets), "\n")
cat("Columns:", paste(colnames(custom_genesets), collapse = ", "), "\n")
```

```{r summarize_genesets}
# Summarize available gene sets
geneset_summary <- custom_genesets %>%
  group_by(gs_name) %>%
  summarize(n_genes = n(), .groups = "drop") %>%
  arrange(desc(n_genes))

cat("\nAvailable gene sets (", nrow(geneset_summary), " total):\n", sep = "")
print(geneset_summary, n = 30)
```

```{r filter_mda_genesets}
# Filter to MDA-MB-231 gene sets for easy reference
mda_genesets <- geneset_summary %>%
  filter(str_detect(gs_name, "^MDA-MB-231"))

cat("\nMDA-MB-231 gene sets (", nrow(mda_genesets), " total):\n", sep = "")
print(mda_genesets, n = 50)
```

# Helper Function to Extract Gene Sets

```{r helper_function}
# Function to extract genes from a specific gene set
get_geneset_genes <- function(geneset_name) {
  custom_genesets %>%
    filter(gs_name == geneset_name) %>%
    pull(db_gene_symbol) %>%
    unique()
}

# Example usage:
# genes_3d_promotes_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_1hr")
```

# ECDF Plots

## ECDF: 3d-Promoted Genes (Normoxia Combined, LFC>0.5)

```{r ecdf_3d_promotes_combined, fig.width=10, fig.height=4}
# Get genes promoted by 3d in normoxia (combined, LFC>0.5)
genes_3d_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3d_promotes_TE_combined_lfc0.5")
cat("Genes in 3d promotes set:", length(genes_3d_promotes), "\n")

# Combine all timepoints into long format
te_combined <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3d_promotes, "3d Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3d_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3d_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3d_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes",
    subtitle = "Normoxia 3d promotes TE (combined, LFC>0.5)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3d_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```

## ECDF: 3d-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r ecdf_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets
genes_3d_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_1hr_lfc0.5")
genes_3d_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_4hr_lfc0.5")
genes_3d_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_24hr_lfc0.5")

cat("Genes per timepoint:\n")
cat("1hr:", length(genes_3d_1hr), "\n")
cat("4hr:", length(genes_3d_4hr), "\n")
cat("24hr:", length(genes_3d_24hr), "\n")

# Combine timepoints with timepoint-matched gene classification
te_combined_matched <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_4hr, "3d Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_24hr, "3d Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Create label data for gene counts
label_data <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3d_1hr)),
    paste0("n = ", length(genes_3d_4hr)),
    paste0("n = ", length(genes_3d_24hr))
  )
)

# Create faceted ECDF plot
p <- ggplot(te_combined_matched, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, .5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p)

ggsave(
  here("plots", "ecdf_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p,
  width = 10,
  height = 4
)
```


## Density Plot: 3d-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r density_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Create faceted density plot
p_density <- ggplot(te_combined_matched, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density)

ggsave(
  here("plots", "density_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_density,
  width = 10,
  height = 4
)
```

# 3d promotes with time as covariate
```{r}
# Get genes promoted by 3d in normoxia (combined)
genes_3d_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3d_promotes_TE_alltimepoints_lfc0.5")
cat("Genes in 3d promotes set:", length(genes_3d_promotes), "\n")

# Combine all timepoints into long format
te_combined_3d <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3d_promotes, "3d Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3d_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3d_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined_3d, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3d_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes",
    subtitle = "Normoxia 3d promotes TE (timepoint excluded as covariate)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3d_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```




## ECDF: 3d-Promoted Genes at 1hr (Fixed Across Timecourse)

```{r ecdf_3d_promotes_1hr_fixed, fig.width=10, fig.height=4}
# Use the SAME 1hr gene set for all timepoints (already defined as genes_3d_1hr)
te_combined_3d_1hr_fixed <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes (1hr)", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes (1hr)", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes (1hr)", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data (same gene count for all timepoints)
label_data_3d_1hr_fixed <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3d_1hr))
)

# ECDF plot
p_ecdf_3d_1hr_fixed <- ggplot(te_combined_3d_1hr_fixed, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes (1hr)" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3d_1hr_fixed, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes (1hr Definition)",
    subtitle = "Hypoxia 3d promotes TE at 1hr (LFC>0.5), tracked across timecourse",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_ecdf_3d_1hr_fixed)

ggsave(
  here("plots", "ecdf_3d_promotes_1hr_fixed_across_timecourse.pdf"),
  plot = p_ecdf_3d_1hr_fixed,
  width = 10,
  height = 4
)
```

## Volcano Plot: 3d-Promoted Genes (Hypoxia, Timepoint-Matched)

```{r volcano_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Create combined data with significance and gene group coloring
te_volcano <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_4hr, "3d Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_24hr, "3d Promotes", "Other"))
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj),
    significant = padj < 0.05,
    color_group = case_when(
      gene_group == "3d Promotes" & significant ~ "3d Promotes (sig)",
      gene_group == "3d Promotes" & !significant ~ "3d Promotes (ns)",
      gene_group == "Other" & significant ~ "Other (sig)",
      TRUE ~ "Other (ns)"
    )
  ) %>%
  # Arrange so pink dots are plotted last (on top)
  arrange(gene_group != "3d Promotes")

# Create volcano plot
p_volcano <- ggplot(te_volcano, aes(x = log2FoldChange, y = neg_log10_padj, color = color_group)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(
    values = c(
      "3d Promotes (sig)" = "#E69F00",
      "3d Promotes (ns)" = "#E69F00",
      "Other (sig)" = "#808080",
      "Other (ns)" = "#C0C0C0"
    ),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = 0, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Volcano: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_volcano)

ggsave(
  here("plots", "volcano_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_volcano,
  width = 10,
  height = 4
)
```

## Volcano Plot: All Genes Colored by 3d KD TE LFC (Divergent)

```{r volcano_all_genes_3d_lfc, fig.width=10, fig.height=4}
# Create volcano data with 3d KD LFC values for coloring (all genes)
te_volcano_3d <- bind_rows(
  te_1hr_gsea %>%
    left_join(si3d_hypoxia_1hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "1hr"),
  te_4hr_gsea %>%
    left_join(si3d_hypoxia_4hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "4hr"),
  te_24hr_gsea %>%
    left_join(si3d_hypoxia_24hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj)
  )

# Create volcano plot with divergent color scale (blue = inhibits, orange = promotes)
p_volcano_3d <- ggplot(te_volcano_3d, aes(x = log2FoldChange, y = neg_log10_padj, fill = si3d_te_lfc)) +
  geom_point(alpha = 0.6, size = 1.5, shape = 21, color = "black", stroke = 0.2) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#E69F00",
                       midpoint = 0, name = "3d KD\nTE LFC") +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  labs(
    title = "Volcano: Hypoxia TE Response Colored by 3d KD Effect",
    subtitle = "Orange = 3d promotes TE, Blue = 3d inhibits TE",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

print(p_volcano_3d)

ggsave(
  here("plots", "volcano_all_genes_3d_lfc.pdf"),
  plot = p_volcano_3d,
  width = 10,
  height = 4
)
```

## Volcano Plot: All Genes Colored by 3e KD TE LFC (Divergent)

```{r volcano_all_genes_3e_lfc, fig.width=10, fig.height=4}
# Create volcano data with 3e KD LFC values for coloring (all genes)
te_volcano_3e <- bind_rows(
  te_1hr_gsea %>%
    left_join(si3e_hypoxia_1hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "1hr"),
  te_4hr_gsea %>%
    left_join(si3e_hypoxia_4hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "4hr"),
  te_24hr_gsea %>%
    left_join(si3e_hypoxia_24hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj)
  )

# Create volcano plot with divergent color scale (orange = inhibits, blue = promotes)
p_volcano_3e <- ggplot(te_volcano_3e, aes(x = log2FoldChange, y = neg_log10_padj, fill = si3e_te_lfc)) +
  geom_point(alpha = 0.6, size = 1.5, shape = 21, color = "black", stroke = 0.2) +
  scale_fill_gradient2(low = "#E69F00", mid = "white", high = "#56B4E9",
                       midpoint = 0, name = "3e KD\nTE LFC") +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  labs(
    title = "Volcano: Hypoxia TE Response Colored by 3e KD Effect",
    subtitle = "Blue = 3e promotes TE, Orange = 3e inhibits TE",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

print(p_volcano_3e)

ggsave(
  here("plots", "volcano_all_genes_3e_lfc.pdf"),
  plot = p_volcano_3e,
  width = 10,
  height = 4
)
```

# eIF3e ECDF Plots

## ECDF: 3e-Promoted Genes (Normoxia Combined)

```{r ecdf_3e_promotes_combined, fig.width=10, fig.height=4}
# Get genes promoted by 3e in normoxia (combined)
genes_3e_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3e_promotes_TE_combined")
cat("Genes in 3e promotes set:", length(genes_3e_promotes), "\n")

# Combine all timepoints into long format
te_combined_3e <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3e_promotes, "3e Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3e_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3e_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined_3e, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes",
    subtitle = "Normoxia 3e promotes TE (combined)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3e_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```

# Normoxia with timepoint covariate excluded

```{r}
# Get genes promoted by 3e in normoxia (combined)
genes_3e_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3e_promotes_TE_alltimepoints_lfc0.5")
cat("Genes in 3e promotes set:", length(genes_3e_promotes), "\n")

# Combine all timepoints into long format
te_combined_3e <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3e_promotes, "3e Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3e_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3e_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined_3e, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes",
    subtitle = "Normoxia 3e promotes TE (timepoint excluded as covariate)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3e_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```





## ECDF: 3e-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r ecdf_3e_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets for 3e
genes_3e_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_1hr_lfc0.5")
genes_3e_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_4hr_lfc0.5")
genes_3e_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_24hr_lfc0.5")

cat("Genes per timepoint:\n")
cat("1hr:", length(genes_3e_1hr), "\n")
cat("4hr:", length(genes_3e_4hr), "\n")
cat("24hr:", length(genes_3e_24hr), "\n")

# Combine timepoints with timepoint-matched gene classification
te_combined_matched_3e <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3e_4hr, "3e Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3e_24hr, "3e Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Create label data for gene counts
label_data_3e <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3e_1hr)),
    paste0("n = ", length(genes_3e_4hr)),
    paste0("n = ", length(genes_3e_24hr))
  )
)

# Create faceted ECDF plot
p_3e <- ggplot(te_combined_matched_3e, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3e promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e)

ggsave(
  here("plots", "ecdf_3e_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_3e,
  width = 10,
  height = 4
)
```

## ECDF: 3e-Promoted Genes at 1hr (Fixed Across Timecourse)

```{r ecdf_3e_promotes_1hr_fixed, fig.width=10, fig.height=4}
# Use the SAME 1hr gene set for all timepoints (already defined as genes_3e_1hr)
te_combined_3e_1hr_fixed <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes (1hr)", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes (1hr)", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes (1hr)", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data (same gene count for all timepoints)
label_data_3e_1hr_fixed <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3e_1hr))
)

# ECDF plot
p_ecdf_3e_1hr_fixed <- ggplot(te_combined_3e_1hr_fixed, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes (1hr)" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e_1hr_fixed, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes (1hr Definition)",
    subtitle = "Hypoxia 3e promotes TE at 1hr (LFC>0.5), tracked across timecourse",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_ecdf_3e_1hr_fixed)

ggsave(
  here("plots", "ecdf_3e_promotes_1hr_fixed_across_timecourse.pdf"),
  plot = p_ecdf_3e_1hr_fixed,
  width = 10,
  height = 4
)
```

as density

```{r}
ggplot(te_combined_matched_3e, aes(x = log2FoldChange, color = gene_group)) +
  geom_density()+
  facet_wrap(~timepoint)+
  theme_cowplot()
```

# ECDF Plots: 3d/3e TE Regulation by CLIP Binding Status

This section creates ECDF plots showing the distribution of eIF3d/eIF3e TE regulation (si3d/si3e vs siCtrl TE LFC), colored by whether genes are cap binders or PAR-CLIP bound.

## Load CLIP Binding Data

```{r load_clip_data}
library(readxl)
library(janitor)

# Load Lamper et al. cap-binding data (glucose deprivation cap binders)
lamper_data <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/lamper-s3.xlsx") %>%
  clean_names()

# Convert columns to numeric
lamper_data <- lamper_data %>%
  mutate(
    log2fold_change_cm = as.numeric(log2fold_change_cm),
    log2fold_change_glu = as.numeric(log2fold_change_glu),
    padj_cm = as.numeric(padj_cm),
    padj_glu = as.numeric(padj_glu)
  )

# Define glucose deprivation cap binders (GLU_HIT)
glu_cap_binders <- lamper_data %>%
  filter(log2fold_change_glu > 2 & padj_glu < 0.005) %>%
  pull(ensembl_gene_id)

# Define complete media cap binders (CM_HIT)
cm_cap_binders <- lamper_data %>%
  filter(log2fold_change_cm > 2 & padj_cm < 0.005) %>%
  pull(ensembl_gene_id)

cat("Lamper cap binders:\n")
cat("Glucose deprivation cap binders:", length(glu_cap_binders), "\n")
cat("Complete media cap binders:", length(cm_cap_binders), "\n")

# Load Lee 2015 PAR-CLIP data
lee_parclip <- read_csv("/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv") %>%
  clean_names()

# Filter for eIF3d-specific binding and get unique gene IDs
lee_eif3d_genes <- lee_parclip %>%
  filter(subunit == "d") %>%
  pull(ensembl_id) %>%
  unique() %>%
  str_replace("\\..*", "")  # Strip version numbers if present

cat("Lee PAR-CLIP eIF3d-bound genes:", length(lee_eif3d_genes), "\n")
```

## Prepare 3d/3e TE Data with CLIP Status

```{r prepare_clip_te_data}
# Helper function to clean gene IDs (strip version numbers)
clean_gene_id <- function(gene_id) {
  str_replace(gene_id, "\\..*", "")
}

# Load full si3d and si3e comparison data with gene IDs
si3d_1hr_full <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))
si3d_4hr_full <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))
si3d_24hr_full <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))

si3e_1hr_full <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))
si3e_4hr_full <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))
si3e_24hr_full <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))

# Also load normoxia data
si3d_normoxia_full <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))
si3e_normoxia_full <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_alltimepoints.csv")) %>%
  mutate(gene_id_clean = clean_gene_id(gene_id))

cat("Loaded si3d/si3e comparison data with gene IDs\n")
```

## ECDF: eIF3d Regulation of Cap Binders (Glucose Deprivation)

```{r ecdf_3d_glu_cap_binders, fig.width=10, fig.height=8}
# Combine all si3d timepoints with cap binder classification
si3d_combined_glu <- bind_rows(
  si3d_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3d_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3d_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3d_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    cap_status = ifelse(gene_id_clean %in% glu_cap_binders, "Glu Cap Binder", "Other")
  )

# Count genes in each group per timepoint
cap_counts_glu <- si3d_combined_glu %>%
  group_by(timepoint, condition, cap_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(cap_status == "Glu Cap Binder")

# Create ECDF plot
p_3d_glu <- ggplot(si3d_combined_glu, aes(x = te_lfc, color = cap_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("Glu Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = cap_counts_glu,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#D55E00") +
  labs(
    title = "eIF3d Regulation of Glucose Deprivation Cap Binders",
    subtitle = "TE LFC (si3d vs siCtrl) - positive = 3d promotes TE",
    x = "TE log2 Fold Change (si3d vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3d_glu)

ggsave(
  here("plots", "ecdf_3d_regulation_glu_cap_binders.pdf"),
  plot = p_3d_glu,
  width = 10,
  height = 6
)
```

## ECDF: eIF3e Regulation of Cap Binders (Glucose Deprivation)

```{r ecdf_3e_glu_cap_binders, fig.width=10, fig.height=8}
# Combine all si3e timepoints with cap binder classification
si3e_combined_glu <- bind_rows(
  si3e_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3e_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3e_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3e_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    cap_status = ifelse(gene_id_clean %in% glu_cap_binders, "Glu Cap Binder", "Other")
  )

# Count genes in each group per timepoint
cap_counts_glu_3e <- si3e_combined_glu %>%
  group_by(timepoint, condition, cap_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(cap_status == "Glu Cap Binder")

# Create ECDF plot
p_3e_glu <- ggplot(si3e_combined_glu, aes(x = te_lfc, color = cap_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("Glu Cap Binder" = "#D55E00", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = cap_counts_glu_3e,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#D55E00") +
  labs(
    title = "eIF3e Regulation of Glucose Deprivation Cap Binders",
    subtitle = "TE LFC (si3e vs siCtrl) - positive = 3e promotes TE",
    x = "TE log2 Fold Change (si3e vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e_glu)

ggsave(
  here("plots", "ecdf_3e_regulation_glu_cap_binders.pdf"),
  plot = p_3e_glu,
  width = 10,
  height = 6
)
```

## ECDF: eIF3d Regulation of Complete Media Cap Binders

```{r ecdf_3d_cm_cap_binders, fig.width=10, fig.height=8}
# Add CM cap binder classification
si3d_combined_cm <- bind_rows(
  si3d_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3d_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3d_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3d_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    cap_status = ifelse(gene_id_clean %in% cm_cap_binders, "CM Cap Binder", "Other")
  )

# Count genes
cap_counts_cm <- si3d_combined_cm %>%
  group_by(timepoint, condition, cap_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(cap_status == "CM Cap Binder")

# Create ECDF plot
p_3d_cm <- ggplot(si3d_combined_cm, aes(x = te_lfc, color = cap_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("CM Cap Binder" = "#0072B2", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = cap_counts_cm,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#0072B2") +
  labs(
    title = "eIF3d Regulation of Complete Media Cap Binders",
    subtitle = "TE LFC (si3d vs siCtrl) - positive = 3d promotes TE",
    x = "TE log2 Fold Change (si3d vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3d_cm)

ggsave(
  here("plots", "ecdf_3d_regulation_cm_cap_binders.pdf"),
  plot = p_3d_cm,
  width = 10,
  height = 6
)
```

## ECDF: eIF3e Regulation of Complete Media Cap Binders

```{r ecdf_3e_cm_cap_binders, fig.width=10, fig.height=8}
# Add CM cap binder classification
si3e_combined_cm <- bind_rows(
  si3e_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3e_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3e_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3e_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    cap_status = ifelse(gene_id_clean %in% cm_cap_binders, "CM Cap Binder", "Other")
  )

# Count genes
cap_counts_cm_3e <- si3e_combined_cm %>%
  group_by(timepoint, condition, cap_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(cap_status == "CM Cap Binder")

# Create ECDF plot
p_3e_cm <- ggplot(si3e_combined_cm, aes(x = te_lfc, color = cap_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("CM Cap Binder" = "#0072B2", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = cap_counts_cm_3e,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#0072B2") +
  labs(
    title = "eIF3e Regulation of Complete Media Cap Binders",
    subtitle = "TE LFC (si3e vs siCtrl) - positive = 3e promotes TE",
    x = "TE log2 Fold Change (si3e vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e_cm)

ggsave(
  here("plots", "ecdf_3e_regulation_cm_cap_binders.pdf"),
  plot = p_3e_cm,
  width = 10,
  height = 6
)
```

## ECDF: eIF3d Regulation of Lee PAR-CLIP Bound Genes

```{r ecdf_3d_lee_parclip, fig.width=10, fig.height=8}
# Add Lee PAR-CLIP classification
si3d_combined_lee <- bind_rows(
  si3d_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3d_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3d_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3d_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    binding_status = ifelse(gene_id_clean %in% lee_eif3d_genes, "eIF3d-Bound (Lee)", "Other")
  )

# Count genes
lee_counts <- si3d_combined_lee %>%
  group_by(timepoint, condition, binding_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(binding_status == "eIF3d-Bound (Lee)")

# Create ECDF plot
p_3d_lee <- ggplot(si3d_combined_lee, aes(x = te_lfc, color = binding_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("eIF3d-Bound (Lee)" = "#009E73", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = lee_counts,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#009E73") +
  labs(
    title = "eIF3d Regulation of Lee PAR-CLIP Bound Genes",
    subtitle = "TE LFC (si3d vs siCtrl) - positive = 3d promotes TE",
    x = "TE log2 Fold Change (si3d vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3d_lee)

ggsave(
  here("plots", "ecdf_3d_regulation_lee_parclip.pdf"),
  plot = p_3d_lee,
  width = 10,
  height = 6
)
```

## ECDF: eIF3e Regulation of Lee PAR-CLIP Bound Genes

```{r ecdf_3e_lee_parclip, fig.width=10, fig.height=8}
# Add Lee PAR-CLIP classification
si3e_combined_lee <- bind_rows(
  si3e_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3e_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3e_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3e_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    binding_status = ifelse(gene_id_clean %in% lee_eif3d_genes, "eIF3d-Bound (Lee)", "Other")
  )

# Count genes
lee_counts_3e <- si3e_combined_lee %>%
  group_by(timepoint, condition, binding_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  filter(binding_status == "eIF3d-Bound (Lee)")

# Create ECDF plot
p_3e_lee <- ggplot(si3e_combined_lee, aes(x = te_lfc, color = binding_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("eIF3d-Bound (Lee)" = "#009E73", "Other" = "#808080"),
    name = ""
  ) +
  facet_grid(condition ~ timepoint, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = lee_counts_3e,
            aes(x = -1, y = 0.15, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 3, color = "#009E73") +
  labs(
    title = "eIF3e Regulation of Lee PAR-CLIP Bound Genes",
    subtitle = "TE LFC (si3e vs siCtrl) - positive = 3e promotes TE",
    x = "TE log2 Fold Change (si3e vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e_lee)

ggsave(
  here("plots", "ecdf_3e_regulation_lee_parclip.pdf"),
  plot = p_3e_lee,
  width = 10,
  height = 6
)
```

## Combined Plot: All CLIP Gene Sets for eIF3d

```{r ecdf_3d_all_clip_combined, fig.width=12, fig.height=6}
# Create combined data with all CLIP classifications
si3d_all_clip <- bind_rows(
  si3d_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3d_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3d_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3d_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    clip_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glu Cap Binder",
      gene_id_clean %in% cm_cap_binders ~ "CM Cap Binder",
      gene_id_clean %in% lee_eif3d_genes ~ "Lee PAR-CLIP",
      TRUE ~ "Other"
    ),
    clip_status = factor(clip_status, levels = c("Glu Cap Binder", "CM Cap Binder", "Lee PAR-CLIP", "Other"))
  )

# Hypoxia only for cleaner plot
si3d_hypoxia_clip <- si3d_all_clip %>%
  filter(condition == "Hypoxia")

# Create ECDF plot with all CLIP gene sets
p_3d_all_clip <- ggplot(si3d_hypoxia_clip, aes(x = te_lfc, color = clip_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("Glu Cap Binder" = "#D55E00", "CM Cap Binder" = "#0072B2",
               "Lee PAR-CLIP" = "#009E73", "Other" = "#808080"),
    name = "CLIP Status"
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "eIF3d Regulation by CLIP Binding Status (Hypoxia)",
    subtitle = "TE LFC (si3d vs siCtrl) - positive = 3d promotes TE",
    x = "TE log2 Fold Change (si3d vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3d_all_clip)

ggsave(
  here("plots", "ecdf_3d_regulation_all_clip_hypoxia.pdf"),
  plot = p_3d_all_clip,
  width = 12,
  height = 5
)
```

## Combined Plot: All CLIP Gene Sets for eIF3e

```{r ecdf_3e_all_clip_combined, fig.width=12, fig.height=6}
# Create combined data with all CLIP classifications
si3e_all_clip <- bind_rows(
  si3e_1hr_full %>% mutate(timepoint = "1hr", condition = "Hypoxia"),
  si3e_4hr_full %>% mutate(timepoint = "4hr", condition = "Hypoxia"),
  si3e_24hr_full %>% mutate(timepoint = "24hr", condition = "Hypoxia"),
  si3e_normoxia_full %>% mutate(timepoint = "All", condition = "Normoxia")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr", "All")),
    clip_status = case_when(
      gene_id_clean %in% glu_cap_binders ~ "Glu Cap Binder",
      gene_id_clean %in% cm_cap_binders ~ "CM Cap Binder",
      gene_id_clean %in% lee_eif3d_genes ~ "Lee PAR-CLIP",
      TRUE ~ "Other"
    ),
    clip_status = factor(clip_status, levels = c("Glu Cap Binder", "CM Cap Binder", "Lee PAR-CLIP", "Other"))
  )

# Hypoxia only for cleaner plot
si3e_hypoxia_clip <- si3e_all_clip %>%
  filter(condition == "Hypoxia")

# Create ECDF plot with all CLIP gene sets
p_3e_all_clip <- ggplot(si3e_hypoxia_clip, aes(x = te_lfc, color = clip_status)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("Glu Cap Binder" = "#D55E00", "CM Cap Binder" = "#0072B2",
               "Lee PAR-CLIP" = "#009E73", "Other" = "#808080"),
    name = "CLIP Status"
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "eIF3e Regulation by CLIP Binding Status (Hypoxia)",
    subtitle = "TE LFC (si3e vs siCtrl) - positive = 3e promotes TE",
    x = "TE log2 Fold Change (si3e vs siCtrl)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 2) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e_all_clip)

ggsave(
  here("plots", "ecdf_3e_regulation_all_clip_hypoxia.pdf"),
  plot = p_3e_all_clip,
  width = 12,
  height = 5
)
```

## Statistical Tests: KS Tests for CLIP Binding Status

```{r clip_ks_tests}
# Function to run KS test for a specific condition/timepoint
run_ks_test <- function(data, clip_group, clip_col) {
  clip_genes <- data %>% filter(!!sym(clip_col) == clip_group) %>% pull(te_lfc)
  other_genes <- data %>% filter(!!sym(clip_col) != clip_group & !!sym(clip_col) != "Other") %>% pull(te_lfc)
  other_genes <- data %>% filter(!!sym(clip_col) == "Other") %>% pull(te_lfc)

  if (length(clip_genes) < 5) {
    return(tibble(
      n_clip = length(clip_genes),
      n_other = length(other_genes),
      median_clip = NA,
      median_other = NA,
      ks_statistic = NA,
      ks_pvalue = NA
    ))
  }

  ks_result <- ks.test(clip_genes, other_genes)

  tibble(
    n_clip = length(clip_genes),
    n_other = length(other_genes),
    median_clip = median(clip_genes, na.rm = TRUE),
    median_other = median(other_genes, na.rm = TRUE),
    ks_statistic = ks_result$statistic,
    ks_pvalue = ks_result$p.value
  )
}

# Run tests for all combinations
clip_tests_results <- bind_rows(
  # si3d tests
  si3d_combined_glu %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "Glu Cap Binder", "cap_status")) %>%
    mutate(factor = "eIF3d", clip_set = "Glu Cap Binder"),

  si3d_combined_cm %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "CM Cap Binder", "cap_status")) %>%
    mutate(factor = "eIF3d", clip_set = "CM Cap Binder"),

  si3d_combined_lee %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "eIF3d-Bound (Lee)", "binding_status")) %>%
    mutate(factor = "eIF3d", clip_set = "Lee PAR-CLIP"),

  # si3e tests
  si3e_combined_glu %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "Glu Cap Binder", "cap_status")) %>%
    mutate(factor = "eIF3e", clip_set = "Glu Cap Binder"),

  si3e_combined_cm %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "CM Cap Binder", "cap_status")) %>%
    mutate(factor = "eIF3e", clip_set = "CM Cap Binder"),

  si3e_combined_lee %>%
    group_by(condition, timepoint) %>%
    group_modify(~run_ks_test(.x, "eIF3d-Bound (Lee)", "binding_status")) %>%
    mutate(factor = "eIF3e", clip_set = "Lee PAR-CLIP")
)

# Apply BH correction
clip_tests_results <- clip_tests_results %>%
  mutate(padj = p.adjust(ks_pvalue, method = "BH"))

cat("\nKS Test Results for CLIP Binding vs TE Regulation:\n")
print(clip_tests_results %>% arrange(padj))

# Export results
write_csv(clip_tests_results, here("output", "clip_te_regulation_ks_tests.csv"))
cat("\nResults exported to: output/clip_te_regulation_ks_tests.csv\n")
```