---
title: "ECDF Plots: eIF3d and eIF3e Gene Sets"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis creates ECDF plots to visualize translational changes during hypoxia in control cells, colored by custom eIF3d/eIF3e gene sets.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggplot2)
library(ggnewscale)
```

# Load TE Data

Load translation efficiency changes in siCtrl cells (hypoxia vs normoxia) at all three timepoints.

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

cat("Loaded TE data:\n")
cat("1hr:", nrow(te_1hr), "genes\n")
cat("4hr:", nrow(te_4hr), "genes\n")
cat("24hr:", nrow(te_24hr), "genes\n")
```

# Load 3d KD Comparison Data

Load the si3d vs sictrl comparison data for each hypoxia timepoint (for LFC color scaling).

```{r load_3d_kd_data}
# Load si3d vs sictrl comparisons (for gradient coloring)
si3d_hypoxia_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

si3d_hypoxia_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

si3d_hypoxia_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  select(gene_id, symbol, si3d_te_lfc = te_lfc)

cat("Loaded 3d KD comparison data:\n")
cat("1hr:", nrow(si3d_hypoxia_1hr), "genes\n")
cat("4hr:", nrow(si3d_hypoxia_4hr), "genes\n")
cat("24hr:", nrow(si3d_hypoxia_24hr), "genes\n")

# Load si3e vs sictrl comparisons (for gradient coloring)
si3e_hypoxia_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

si3e_hypoxia_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

si3e_hypoxia_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  select(gene_id, symbol, si3e_te_lfc = te_lfc)

cat("\nLoaded 3e KD comparison data:\n")
cat("1hr:", nrow(si3e_hypoxia_1hr), "genes\n")
cat("4hr:", nrow(si3e_hypoxia_4hr), "genes\n")
cat("24hr:", nrow(si3e_hypoxia_24hr), "genes\n")
```

# Load Gene Annotation

```{r load_gene_annotation}
# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

# Prepare TE Data with Gene Symbols

```{r prepare_te_data}
# Add gene symbols to TE data
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Prepared TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes with symbols\n")
cat("4hr:", nrow(te_4hr_gsea), "genes with symbols\n")
cat("24hr:", nrow(te_24hr_gsea), "genes with symbols\n")
```

# Load Custom Gene Sets

```{r load_custom_genesets}
# Load all custom gene sets
custom_genesets <- read_csv(here("accessories", "all_combined_sigs_1_19_26.csv"))

cat("Loaded custom gene sets:\n")
cat("Total entries:", nrow(custom_genesets), "\n")
cat("Columns:", paste(colnames(custom_genesets), collapse = ", "), "\n")
```

```{r summarize_genesets}
# Summarize available gene sets
geneset_summary <- custom_genesets %>%
  group_by(gs_name) %>%
  summarize(n_genes = n(), .groups = "drop") %>%
  arrange(desc(n_genes))

cat("\nAvailable gene sets (", nrow(geneset_summary), " total):\n", sep = "")
print(geneset_summary, n = 30)
```

```{r filter_mda_genesets}
# Filter to MDA-MB-231 gene sets for easy reference
mda_genesets <- geneset_summary %>%
  filter(str_detect(gs_name, "^MDA-MB-231"))

cat("\nMDA-MB-231 gene sets (", nrow(mda_genesets), " total):\n", sep = "")
print(mda_genesets, n = 50)
```

# Helper Function to Extract Gene Sets

```{r helper_function}
# Function to extract genes from a specific gene set
get_geneset_genes <- function(geneset_name) {
  custom_genesets %>%
    filter(gs_name == geneset_name) %>%
    pull(db_gene_symbol) %>%
    unique()
}

# Example usage:
# genes_3d_promotes_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_1hr")
```

# ECDF Plots

## ECDF: 3d-Promoted Genes (Normoxia Combined, LFC>0.5)

```{r ecdf_3d_promotes_combined, fig.width=10, fig.height=4}
# Get genes promoted by 3d in normoxia (combined, LFC>0.5)
genes_3d_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3d_promotes_TE_combined_lfc0.5")
cat("Genes in 3d promotes set:", length(genes_3d_promotes), "\n")

# Combine all timepoints into long format
te_combined <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3d_promotes, "3d Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3d_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3d_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3d_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes",
    subtitle = "Normoxia 3d promotes TE (combined, LFC>0.5)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3d_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```

## ECDF: 3d-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r ecdf_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets
genes_3d_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_1hr_lfc0.5")
genes_3d_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_4hr_lfc0.5")
genes_3d_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_24hr_lfc0.5")

cat("Genes per timepoint:\n")
cat("1hr:", length(genes_3d_1hr), "\n")
cat("4hr:", length(genes_3d_4hr), "\n")
cat("24hr:", length(genes_3d_24hr), "\n")

# Combine timepoints with timepoint-matched gene classification
te_combined_matched <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_4hr, "3d Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_24hr, "3d Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Create label data for gene counts
label_data <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3d_1hr)),
    paste0("n = ", length(genes_3d_4hr)),
    paste0("n = ", length(genes_3d_24hr))
  )
)

# Create faceted ECDF plot
p <- ggplot(te_combined_matched, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, .5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p)

ggsave(
  here("plots", "ecdf_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p,
  width = 10,
  height = 4
)
```


## Density Plot: 3d-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r density_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Create faceted density plot
p_density <- ggplot(te_combined_matched, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic Response: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density)

ggsave(
  here("plots", "density_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_density,
  width = 10,
  height = 4
)
```

## Volcano Plot: 3d-Promoted Genes (Hypoxia, Timepoint-Matched)

```{r volcano_3d_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Create combined data with significance and gene group coloring
te_volcano <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_4hr, "3d Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_24hr, "3d Promotes", "Other"))
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj),
    significant = padj < 0.05,
    color_group = case_when(
      gene_group == "3d Promotes" & significant ~ "3d Promotes (sig)",
      gene_group == "3d Promotes" & !significant ~ "3d Promotes (ns)",
      gene_group == "Other" & significant ~ "Other (sig)",
      TRUE ~ "Other (ns)"
    )
  ) %>%
  # Arrange so pink dots are plotted last (on top)
  arrange(gene_group != "3d Promotes")

# Create volcano plot
p_volcano <- ggplot(te_volcano, aes(x = log2FoldChange, y = neg_log10_padj, color = color_group)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(
    values = c(
      "3d Promotes (sig)" = "#E69F00",
      "3d Promotes (ns)" = "#E69F00",
      "Other (sig)" = "#808080",
      "Other (ns)" = "#C0C0C0"
    ),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  geom_text(data = label_data, aes(x = 0, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Volcano: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_volcano)

ggsave(
  here("plots", "volcano_3d_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_volcano,
  width = 10,
  height = 4
)
```

## Volcano Plot: All Genes Colored by 3d KD TE LFC (Divergent)

```{r volcano_all_genes_3d_lfc, fig.width=10, fig.height=4}
# Create volcano data with 3d KD LFC values for coloring (all genes)
te_volcano_3d <- bind_rows(
  te_1hr_gsea %>%
    left_join(si3d_hypoxia_1hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "1hr"),
  te_4hr_gsea %>%
    left_join(si3d_hypoxia_4hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "4hr"),
  te_24hr_gsea %>%
    left_join(si3d_hypoxia_24hr %>% select(symbol, si3d_te_lfc), by = "symbol") %>%
    mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj)
  )

# Create volcano plot with divergent color scale (blue = inhibits, orange = promotes)
p_volcano_3d <- ggplot(te_volcano_3d, aes(x = log2FoldChange, y = neg_log10_padj, fill = si3d_te_lfc)) +
  geom_point(alpha = 0.6, size = 1.5, shape = 21, color = "black", stroke = 0.2) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#E69F00",
                       midpoint = 0, name = "3d KD\nTE LFC") +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  labs(
    title = "Volcano: Hypoxia TE Response Colored by 3d KD Effect",
    subtitle = "Orange = 3d promotes TE, Blue = 3d inhibits TE",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

print(p_volcano_3d)

ggsave(
  here("plots", "volcano_all_genes_3d_lfc.pdf"),
  plot = p_volcano_3d,
  width = 10,
  height = 4
)
```

## Volcano Plot: All Genes Colored by 3e KD TE LFC (Divergent)

```{r volcano_all_genes_3e_lfc, fig.width=10, fig.height=4}
# Create volcano data with 3e KD LFC values for coloring (all genes)
te_volcano_3e <- bind_rows(
  te_1hr_gsea %>%
    left_join(si3e_hypoxia_1hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "1hr"),
  te_4hr_gsea %>%
    left_join(si3e_hypoxia_4hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "4hr"),
  te_24hr_gsea %>%
    left_join(si3e_hypoxia_24hr %>% select(symbol, si3e_te_lfc), by = "symbol") %>%
    mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = -log10(padj)
  )

# Create volcano plot with divergent color scale (orange = inhibits, blue = promotes)
p_volcano_3e <- ggplot(te_volcano_3e, aes(x = log2FoldChange, y = neg_log10_padj, fill = si3e_te_lfc)) +
  geom_point(alpha = 0.6, size = 1.5, shape = 21, color = "black", stroke = 0.2) +
  scale_fill_gradient2(low = "#E69F00", mid = "white", high = "#56B4E9",
                       midpoint = 0, name = "3e KD\nTE LFC") +
  facet_wrap(~timepoint, nrow = 1, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  labs(
    title = "Volcano: Hypoxia TE Response Colored by 3e KD Effect",
    subtitle = "Blue = 3e promotes TE, Orange = 3e inhibits TE",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "-log10(padj)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

print(p_volcano_3e)

ggsave(
  here("plots", "volcano_all_genes_3e_lfc.pdf"),
  plot = p_volcano_3e,
  width = 10,
  height = 4
)
```

# eIF3e ECDF Plots

## ECDF: 3e-Promoted Genes (Normoxia Combined)

```{r ecdf_3e_promotes_combined, fig.width=10, fig.height=4}
# Get genes promoted by 3e in normoxia (combined)
genes_3e_promotes <- get_geneset_genes("MDA-MB-231_normoxia_3e_promotes_TE_combined")
cat("Genes in 3e promotes set:", length(genes_3e_promotes), "\n")

# Combine all timepoints into long format
te_combined_3e <- bind_rows(
  te_1hr_gsea %>% mutate(timepoint = "1hr"),
  te_4hr_gsea %>% mutate(timepoint = "4hr"),
  te_24hr_gsea %>% mutate(timepoint = "24hr")
) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    gene_group = ifelse(symbol %in% genes_3e_promotes, "3e Promotes", "Other")
  )

# Label data for gene count (same for all facets since it's combined)
label_data_3e_combined <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = paste0("n = ", length(genes_3e_promotes))
)

# Create faceted ECDF plot
ggplot(te_combined_3e, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e_combined, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes",
    subtitle = "Normoxia 3e promotes TE (combined)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(
  here("plots", "ecdf_3e_promotes_normoxia_combined.pdf"),
  width = 10,
  height = 4
)
```

## ECDF: 3e-Promoted Genes (Hypoxia, Timepoint-Matched, LFC>0.5)

```{r ecdf_3e_promotes_timepoint_matched, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets for 3e
genes_3e_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_1hr_lfc0.5")
genes_3e_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_4hr_lfc0.5")
genes_3e_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_24hr_lfc0.5")

cat("Genes per timepoint:\n")
cat("1hr:", length(genes_3e_1hr), "\n")
cat("4hr:", length(genes_3e_4hr), "\n")
cat("24hr:", length(genes_3e_24hr), "\n")

# Combine timepoints with timepoint-matched gene classification
te_combined_matched_3e <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3e_4hr, "3e Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3e_24hr, "3e Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Create label data for gene counts
label_data_3e <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3e_1hr)),
    paste0("n = ", length(genes_3e_4hr)),
    paste0("n = ", length(genes_3e_24hr))
  )
)

# Create faceted ECDF plot
p_3e <- ggplot(te_combined_matched_3e, aes(x = log2FoldChange, color = gene_group)) +
  stat_ecdf(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e, aes(x = 0, y = 0.05, label = label),
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic Response: 3e-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3e promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_3e)

ggsave(
  here("plots", "ecdf_3e_promotes_hypoxia_timepoint_matched.pdf"),
  plot = p_3e,
  width = 10,
  height = 4
)
```

as density

```{r}
ggplot(te_combined_matched_3e, aes(x = log2FoldChange, color = gene_group)) +
  geom_density()+
  facet_wrap(~timepoint)+
  theme_cowplot()
```