---
title: "Effect of knockdown on hypoxic response"
author: "Kate"
date: "2025-01-16"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(ggExtra)
library(ggprism)
library(cowplot)
```

# Effect of knockdown on hypoxic response

Each panel compares the hypoxia vs normoxia TE change under si-CTRL (x) against the
same contrast under si-eIF3d or si-eIF3e (y), restricted to genes with a significant
si-CTRL hypoxic response.

## Read in DESeq results

```{r load-te-tables, message=FALSE}
# One reader for all nine tables. Each file is keyed on a unique gene_id, so the
# joins below are 1:1; the stopifnot after the join asserts that.
read_te <- function(file, prefix) {
  read_csv(here("output", file), show_col_types = FALSE) |>
    dplyr::select(gene_id, te_lfc, te_padj) |>
    dplyr::rename(
      !!paste0(prefix, "_te_lfc")  := te_lfc,
      !!paste0(prefix, "_te_padj") := te_padj
    )
}

te_files <- tibble::tribble(
  ~prefix,         ~file,
  "sictrl_1hr",    "translation_categories_1hr.csv",
  "sictrl_4hr",    "translation_categories_4hr.csv",
  "sictrl_24hr",   "translation_categories_24hr.csv",
  "si3e_1hr",      "translation_categories_si3e_1hr.csv",
  "si3e_4hr",      "translation_categories_si3e_4hr.csv",
  "si3e_24hr",     "translation_categories_si3e_24hr.csv",
  "si3d_1hr",      "translation_categories_si3d_1hr.csv",
  "si3d_4hr",      "translation_categories_si3d_4hr.csv",
  "si3d_24hr",     "translation_categories_si3d_24hr.csv"
)

all_hypoxic_response <- purrr::reduce(
  purrr::map2(te_files$file, te_files$prefix, read_te),
  \(x, y) dplyr::left_join(x, y, by = "gene_id")
)

# Validation: keys unique (no row multiplication from the joins), all expected
# columns present, and every LFC finite where not NA.
expected_cols <- c("gene_id",
                   paste0(te_files$prefix, "_te_lfc"),
                   paste0(te_files$prefix, "_te_padj"))
stopifnot(
  "join multiplied rows: gene_id is not unique" =
    nrow(all_hypoxic_response) == dplyr::n_distinct(all_hypoxic_response$gene_id),
  "missing expected TE columns" =
    all(expected_cols %in% colnames(all_hypoxic_response))
)
lfc_cols <- paste0(te_files$prefix, "_te_lfc")
stopifnot("non-finite TE log fold changes" = all(vapply(
  lfc_cols,
  \(cc) all(is.finite(all_hypoxic_response[[cc]][!is.na(all_hypoxic_response[[cc]])])),
  logical(1)
)))

cat("Loaded", nrow(all_hypoxic_response), "genes across",
    nrow(te_files), "TE tables\n")
head(all_hypoxic_response)
```

## Shared helpers

```{r panel-helpers}
LFC_CUT  <- 0.5
PADJ_CUT <- 0.05
AXIS_LIM <- 2

GENE_FILL <- c(si3e = "#56B4E9", si3d = "#E69F00")
GENE_LAB  <- c(si3e = "si-eIF3e", si3d = "si-eIF3d")

# Genes with a significant si-CTRL hypoxic TE response at a given timepoint,
# returned as a two-column frame (x = si-CTRL LFC, y = knockdown LFC) plus the
# knockdown padj needed for the bar counts.
response_pair <- function(df, gene, tp, lfc_cut = LFC_CUT, padj_cut = PADJ_CUT) {
  ctrl_lfc  <- paste0("sictrl_", tp, "_te_lfc")
  ctrl_padj <- paste0("sictrl_", tp, "_te_padj")
  kd_lfc    <- paste0(gene, "_", tp, "_te_lfc")
  kd_padj   <- paste0(gene, "_", tp, "_te_padj")

  df |>
    dplyr::filter(.data[[ctrl_padj]] < padj_cut,
                  abs(.data[[ctrl_lfc]]) > lfc_cut) |>
    dplyr::transmute(
      gene_id,
      x       = .data[[ctrl_lfc]],
      y       = .data[[kd_lfc]],
      y_padj  = .data[[kd_padj]]
    )
}

# Scatter with marginal histograms.
#
# coord_cartesian (not xlim/ylim) is deliberate: xlim/ylim are scale limits and
# DELETE out-of-range rows before the geom and the marginal histograms see them,
# which at 1hr silently removed ~19% of genes, all of them the largest hypoxic
# responders. coord_cartesian zooms without dropping data.
make_response_panel <- function(df, gene, tp, lim = AXIS_LIM) {
  d <- response_pair(df, gene, tp) |>
    dplyr::filter(!is.na(x), !is.na(y))

  # R and n are computed on every gene, including those the zoom puts off-panel.
  # The off-panel count is printed so the axis range is never mistaken for the
  # full data range.
  r      <- cor(d$x, d$y, method = "pearson")
  n_off  <- sum(abs(d$x) > lim | abs(d$y) > lim)
  lab    <- sprintf("Pearson R = %.2f, n = %d", r, nrow(d))
  if (n_off > 0) lab <- paste0(lab, "\n", n_off, " outside axis range")

  p <- ggplot(d, aes(x = x, y = y)) +
    geom_hline(yintercept = 0, color = "grey") +
    geom_vline(xintercept = 0, color = "grey") +
    geom_point(alpha = .25) +
    annotate("text", x = Inf, y = -Inf, label = lab,
             hjust = 1.05, vjust = -0.4, size = 3, lineheight = 0.95) +
    coord_cartesian(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
    xlab(bquote(Log[2] ~ "si-CTRL LFC" ~ "TE: hypoxia vs normoxia")) +
    ylab(bquote(Log[2] ~ .(GENE_LAB[[gene]]) ~ "LFC" ~ "TE: hypoxia vs normoxia")) +
    theme_prism(base_size = 11)

  ggMarginal(p, type = "histogram",
             xparams = list(binwidth = .1, fill = "grey"),
             yparams = list(binwidth = .1, fill = GENE_FILL[[gene]]))
}

# Direction-matched counts.
#
# The knockdown bars count only genes that keep the si-CTRL direction, so si-KD Up
# is a strict subset of si-CTRL Up (same for Down) and the bars are comparable.
# Counting the knockdown sign on its own conflated retained response with outright
# reversal: at 24hr si3e that put 1115 genes in the "Up" bar of which 1111 were
# si-CTRL DOWN. Reversals are counted separately and plotted below.
te_bar_counts <- function(df, gene, tp, lfc_cut = LFC_CUT, padj_cut = PADJ_CUT) {
  d <- response_pair(df, gene, tp, lfc_cut, padj_cut)

  c_up   <- d$x >  lfc_cut
  c_down <- d$x < -lfc_cut
  k_up   <- d$y >  lfc_cut & d$y_padj < padj_cut
  k_down <- d$y < -lfc_cut & d$y_padj < padj_cut

  tibble::tibble(
    ctrl_up        = sum(c_up,             na.rm = TRUE),
    ctrl_down      = sum(c_down,           na.rm = TRUE),
    kd_up          = sum(c_up   & k_up,    na.rm = TRUE),
    kd_down        = sum(c_down & k_down,  na.rm = TRUE),
    rev_up_to_down = sum(c_up   & k_down,  na.rm = TRUE),
    rev_down_to_up = sum(c_down & k_up,    na.rm = TRUE)
  )
}

BAR_FILL <- c(Up = "#D55E00", Down = "#0072B2")
BAR_SUBTITLE <- paste0("padj < ", PADJ_CUT, ", |LFC| > ", LFC_CUT,
                       "; si-KD bars are direction matched subsets of si-CTRL")

make_bar_plot <- function(counts, title) {
  bar_data <- tibble::tibble(
    Condition = factor(rep(c("si-CTRL", "si-KD"), each = 2),
                       levels = c("si-CTRL", "si-KD")),
    Direction = factor(rep(c("Up", "Down"), 2), levels = c("Up", "Down")),
    Count     = c(counts$ctrl_up, counts$ctrl_down,
                  counts$kd_up,   counts$kd_down)
  )

  ggplot(bar_data, aes(x = Condition, y = Count, fill = Direction)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.9),
              vjust = -0.5, size = 5) +
    scale_fill_manual(values = BAR_FILL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_prism() +
    theme(legend.position = "right") +
    labs(title = title, subtitle = BAR_SUBTITLE)
}

panel_spec <- tidyr::expand_grid(
  timepoint = c("1hr", "4hr", "24hr"),
  gene      = c("si3e", "si3d")
)
```

## Per-timepoint panels

```{r response-panels, fig.width=4, fig.height=4, message=FALSE, warning=FALSE}
# Scatters and bar plots for all six gene x timepoint combinations. Previously six
# copy-pasted ~90 line chunks; the loop keeps them guaranteed identical.
#
# Scatter filenames carry a "knockdown_" prefix: the unprefixed
# si3{d,e}_{tp}_response.pdf names are also written by
# code/acute_to_long_hypoxic_response.Rmd, so whichever notebook knit last silently
# won the file. That notebook keeps the old names.
all_counts <- vector("list", nrow(panel_spec))

for (i in seq_len(nrow(panel_spec))) {
  gene <- panel_spec$gene[i]
  tp   <- panel_spec$timepoint[i]

  scatter <- make_response_panel(all_hypoxic_response, gene, tp)
  print(scatter)
  ggsave(here("plots", paste0("knockdown_", gene, "_", tp, "_response.pdf")),
         scatter, height = 4.5, width = 4.5)

  counts <- te_bar_counts(all_hypoxic_response, gene, tp)
  all_counts[[i]] <- counts |>
    dplyr::mutate(gene = gene, timepoint = tp, .before = 1)

  bar <- make_bar_plot(counts, paste0("TE Changes - ", tp, " ", GENE_LAB[[gene]]))
  print(bar)
  ggsave(here("plots", paste0("knockdown_", gene, "_", tp, "_barplot.pdf")),
         bar, height = 4.5, width = 4.5)
}

counts_tbl <- dplyr::bind_rows(all_counts)
knitr::kable(counts_tbl, caption = "Direction matched counts and reversals")
```

## Combined faceted bar plot: eIF3e (top) and eIF3d (bottom) x timepoint

```{r combined-faceted-barplot, fig.width=10, fig.height=7}
combined_data <- counts_tbl |>
  tidyr::pivot_longer(
    c(ctrl_up, ctrl_down, kd_up, kd_down),
    names_to = c("Condition", "Direction"),
    names_sep = "_",
    values_to = "Count"
  ) |>
  dplyr::mutate(
    Gene      = factor(dplyr::recode(gene, si3e = "eIF3e", si3d = "eIF3d"),
                       levels = c("eIF3e", "eIF3d")),
    Timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    Condition = factor(dplyr::recode(Condition, ctrl = "si-CTRL", kd = "si-KD"),
                       levels = c("si-CTRL", "si-KD")),
    Direction = factor(stringr::str_to_title(Direction), levels = c("Up", "Down"))
  )

make_faceted_bar <- function(d, title, facets) {
  ggplot(d, aes(x = Condition, y = Count, fill = Direction)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.9),
              vjust = -0.5, size = 5) +
    scale_fill_manual(values = BAR_FILL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    facets +
    theme_prism(base_size = 16) +
    theme(legend.position = "right",
          strip.text    = element_text(size = 16),
          axis.text     = element_text(size = 14),
          axis.title    = element_text(size = 16),
          legend.text   = element_text(size = 14),
          legend.title  = element_text(size = 16),
          plot.title    = element_text(size = 18),
          plot.subtitle = element_text(size = 13)) +
    labs(title = title, subtitle = BAR_SUBTITLE, x = "Condition", y = "Count")
}

p_combined <- make_faceted_bar(
  combined_data,
  "TE Changes: Hypoxic Response and Effect of eIF3 Knockdown",
  facet_wrap(~ Gene + Timepoint, scales = "free_y", nrow = 2)
)
p_combined
ggsave(here("plots", "te_changes_combined_eif3_hypoxia.pdf"), p_combined,
       width = 10, height = 7)

p_3e <- make_faceted_bar(
  combined_data |> dplyr::filter(Gene == "eIF3e"),
  "TE Changes: eIF3e Knockdown Hypoxic Response",
  facet_wrap(~ Timepoint, scales = "free_y", nrow = 1)
)
p_3e
ggsave(here("plots", "te_changes_eif3e_hypoxia.pdf"), p_3e, width = 10, height = 4)

p_3d <- make_faceted_bar(
  combined_data |> dplyr::filter(Gene == "eIF3d"),
  "TE Changes: eIF3d Knockdown Hypoxic Response",
  facet_wrap(~ Timepoint, scales = "free_y", nrow = 1)
)
p_3d
ggsave(here("plots", "te_changes_eif3d_hypoxia.pdf"), p_3d, width = 10, height = 4)
```

## Direction reversals under knockdown

```{r reversal-barplot, fig.width=10, fig.height=4}
# Genes whose si-CTRL hypoxic response flips sign under knockdown. These used to be
# folded into the si-KD Up/Down bars above, which made the 24hr eIF3e panel read as
# a 50-fold increase in up-regulated genes when it is almost entirely reversal.
reversal_data <- counts_tbl |>
  tidyr::pivot_longer(c(rev_up_to_down, rev_down_to_up),
                      names_to = "Reversal", values_to = "Count") |>
  dplyr::mutate(
    Gene      = factor(dplyr::recode(gene, si3e = "eIF3e", si3d = "eIF3d"),
                       levels = c("eIF3e", "eIF3d")),
    Timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    Reversal  = factor(dplyr::recode(Reversal,
                                     rev_up_to_down = "si-CTRL up, KD down",
                                     rev_down_to_up = "si-CTRL down, KD up"),
                       levels = c("si-CTRL up, KD down", "si-CTRL down, KD up"))
  )

p_reversal <- ggplot(reversal_data, aes(x = Reversal, y = Count, fill = Reversal)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("si-CTRL up, KD down" = "#0072B2",
                               "si-CTRL down, KD up" = "#D55E00")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  facet_wrap(~ Gene + Timepoint, scales = "free_y", nrow = 2) +
  theme_prism(base_size = 14) +
  theme(legend.position = "right",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.subtitle = element_text(size = 12)) +
  labs(title = "Hypoxic TE response reversed by eIF3 knockdown",
       subtitle = paste0("Genes significant in both si-CTRL and si-KD (padj < ",
                         PADJ_CUT, ", |LFC| > ", LFC_CUT,
                         ") whose response flips sign. Note free y scales."),
       x = NULL, y = "Count")

p_reversal
ggsave(here("plots", "te_changes_reversals_eif3_hypoxia.pdf"), p_reversal,
       width = 10, height = 7.5)
```

## Combined stacked figure: eIF3e (top) over eIF3d (bottom)

```{r combined-barplot-plotgrid, fig.width=10, fig.height=8}
# Reuse p_3e/p_3d built above rather than rebuilding them. Row label ("eIF3e"/"eIF3d")
# replaces "Count" as the y-axis title; only the top (eIF3e) panel keeps the shared
# Up/Down legend since both rows use the identical Direction color mapping. Swapping
# in theme_classic() (replacing theme_prism()'s boxed panel) drops the full border
# and grey strip background, leaving plain bold facet labels and L-shaped axes.
#
# The y axis ticks and labels are dropped here (this figure only, not the standalone
# te_changes_eif3{d,e}_hypoxia.pdf versions): every bar carries its count as a
# geom_text label, so the scale is redundant. Note this also hides the fact that the
# facets use free y scales, so bar heights are NOT comparable across timepoints.
as_stacked_row <- function(p, row_label, legend) {
  p +
    theme_classic(base_size = 16) +
    labs(title = NULL, subtitle = NULL, y = row_label) +
    theme(
      axis.title.y = element_text(face = "bold", size = 20),
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 16),
      legend.position = legend
    )
}

# The per-row titles are stripped above so the two panels stack flush; the title and
# subtitle are drawn once for the whole figure instead. The second subtitle line is
# needed here specifically because this figure has no y axis: without it there is
# nothing on the page to signal that the facets use free scales.
stacked_header <- ggdraw() +
  draw_label("TE Changes: Hypoxic Response and Effect of eIF3 Knockdown",
             fontface = "bold", size = 18, x = 0.01, hjust = 0, y = 0.78) +
  draw_label(BAR_SUBTITLE,
             size = 12, colour = "grey25", x = 0.01, hjust = 0, y = 0.42) +
  draw_label("Free y scales per panel; bar heights are not comparable across timepoints",
             size = 12, colour = "grey25", x = 0.01, hjust = 0, y = 0.12)

stacked_bar_panels <- plot_grid(
  stacked_header,
  plot_grid(
    as_stacked_row(p_3e, "eIF3e", "right"),
    as_stacked_row(p_3d, "eIF3d", "none"),
    ncol = 1,
    align = "v",
    axis = "lr"
  ),
  ncol = 1,
  rel_heights = c(0.13, 1)
)

print(stacked_bar_panels)

ggsave(
  here("plots", "combined_eif3e_eif3d_sig_counts.pdf"),
  plot = stacked_bar_panels,
  width = 10,
  height = 8
)

cat("Combined eIF3e/eIF3d bar figure saved to plots/combined_eif3e_eif3d_sig_counts.pdf\n")
```

## Per-timepoint figures: eIF3e (left) and eIF3d (right)

The stacked figure above puts all three timepoints on one page with free y scales,
which makes the eIF3d panel unreadable next to the 6,195/8,825 eIF3e bars. Here the
same counts are split into one figure per timepoint, eIF3e first and eIF3d second,
side by side. Within a timepoint the two panels still use free y scales for the same
reason (the eIF3e counts run far above eIF3d at 4hr and 24hr), so bar heights are
comparable within a panel, not across the pair.

```{r per-timepoint-barplots, fig.width=3.6, fig.height=2.4}
make_timepoint_bar <- function(d, gene_label, show_legend, show_ylab = TRUE) {
  ggplot(d, aes(x = Condition, y = Count, fill = Direction)) +
    # Bar geometry. With `width = W` split across n = 2 dodged groups each bar draws
    # W/2 wide; the within-pair gap is D/2 - W/2 and the between-condition gap is
    # 1 - D/2 - W/2. Dropping the dodge span D to just above W closes the within-pair
    # gap to ~0.015 so the Up/Down bars nearly touch. That necessarily reopens the
    # condition gap (the two trade off against each other at fixed W), so the panel is
    # narrowed below instead: the x expansion tightens to 0.32, which more than repays
    # it - total x span goes 1.90 -> 1.64 categories.
    geom_bar(stat = "identity", position = position_dodge(width = 0.53),
             width = 0.5) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.53),
              vjust = -0.5, size = 3.2) +
    scale_fill_manual(values = BAR_FILL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    # Default discrete expansion pads 0.6 of a category either side. The outer bar
    # edges now sit at 0.745 / 2.255, so 0.32 keeps a visible margin rather than
    # parking the first bar flush on the y axis.
    scale_x_discrete(expand = expansion(add = 0.32)) +
    theme_classic(base_size = 16) +
    theme(
      axis.title.y     = element_text(face = "bold", size = 12),
      axis.text.y      = element_blank(),
      axis.ticks.y     = element_blank(),
      axis.text.x      = element_text(size = 9),
      plot.title       = element_text(face = "bold", size = 11, hjust = 0.5),
      legend.position  = if (show_legend) "right" else "none",
      legend.text      = element_text(size = 9),
      legend.title     = element_text(size = 10)
    ) +
    labs(title = gene_label, x = NULL,
         y = if (show_ylab) "Count" else NULL)
}

timepoint_figs <- lapply(levels(combined_data$Timepoint), function(tp) {
  d_tp <- dplyr::filter(combined_data, Timepoint == tp)

  # eIF3e first (left), eIF3d second (right) - the requested order.
  # Both panels count the same thing, so "Count" is drawn once on the left; dropping
  # it on the right hands that strip of page back to the eIF3d panel.
  p_e <- make_timepoint_bar(dplyr::filter(d_tp, Gene == "eIF3e"), "eIF3e", FALSE,
                            show_ylab = TRUE)
  p_d <- make_timepoint_bar(dplyr::filter(d_tp, Gene == "eIF3d"), "eIF3d", FALSE,
                            show_ylab = FALSE)

  # Both panels share the identical Direction mapping, so the legend is drawn once
  # and parked to the right of the pair. Keeping it inside a panel would wedge it
  # between eIF3e and eIF3d.
  bar_legend <- get_legend(make_timepoint_bar(d_tp, "eIF3e", TRUE))

  header <- ggdraw() +
    draw_label(paste0("TE Changes: Hypoxic Response and Effect of eIF3 Knockdown - ", tp),
               fontface = "bold", size = 10, x = 0.01, hjust = 0, y = 0.72) +
    draw_label(BAR_SUBTITLE,
               size = 7, colour = "grey25", x = 0.01, hjust = 0, y = 0.28)

  fig <- plot_grid(
    header,
    plot_grid(p_e, p_d, bar_legend, nrow = 1, align = "h", axis = "tb",
              rel_widths = c(1, 1, 0.42)),
    ncol = 1,
    rel_heights = c(0.18, 1)
  )

  print(fig)

  ggsave(here("plots", paste0("te_changes_eif3e_eif3d_", tp, ".pdf")),
         plot = fig, width = 3.6, height = 2.4)

  fig
})
names(timepoint_figs) <- levels(combined_data$Timepoint)

# Every timepoint must contribute both knockdowns x both conditions x both directions.
stopifnot("expected 3 per-timepoint figures" = length(timepoint_figs) == 3)
stopifnot("each timepoint needs 8 bars (2 genes x 2 conditions x 2 directions)" =
  all(table(combined_data$Timepoint) == 8))

cat("Saved plots/te_changes_eif3e_eif3d_{1hr,4hr,24hr}.pdf\n")
```

## Per-timepoint figures: si-CTRL, si-eIF3e and si-eIF3d in one panel

The figures above split eIF3e and eIF3d into two panels, which repeats the si-CTRL
bars once per panel. Here the same counts are drawn as three condition groups on a
single axis - si-CTRL, then si-eIF3e, then si-eIF3d - one figure per timepoint.

This is only well defined because the si-CTRL bars do not depend on which knockdown
they are being compared against: `te_bar_counts()` derives `ctrl_up`/`ctrl_down` from
the si-CTRL contrast alone, so the two rows of `counts_tbl` at a given timepoint carry
the same pair. That is asserted rather than assumed before the two rows are collapsed
to one group.

One consequence of putting all three groups on one axis: unlike the paired figures,
there is now a single shared y scale, and at 4hr/24hr the knockdown bars are a few
percent of the si-CTRL bars (505 vs 6,195; 2,619 vs 8,825) so they render as slivers.
The printed counts carry that information instead. The eIF3e bars at 4hr and 24hr
(3/1/2 and 4/1/2) are small enough to be invisible as bars at any shared linear scale.

```{r ctrl-3e-3d-barplots, fig.width=5, fig.height=3}
# The si-CTRL group is shared by both knockdowns, so it must be identical across the
# two rows of counts_tbl at each timepoint before they can be collapsed into one bar.
ctrl_consistency <- counts_tbl |>
  dplyr::group_by(timepoint) |>
  dplyr::summarise(n_up   = dplyr::n_distinct(ctrl_up),
                   n_down = dplyr::n_distinct(ctrl_down),
                   .groups = "drop")
stopifnot(
  "si-CTRL counts differ between the eIF3e and eIF3d rows" =
    all(ctrl_consistency$n_up == 1 & ctrl_consistency$n_down == 1)
)

three_way_data <- dplyr::bind_rows(
  counts_tbl |>
    dplyr::distinct(timepoint, ctrl_up, ctrl_down) |>
    dplyr::transmute(timepoint, Condition = "si-CTRL",
                     Up = ctrl_up, Down = ctrl_down),
  counts_tbl |>
    dplyr::transmute(timepoint,
                     Condition = dplyr::recode(gene, si3e = "si-eIF3e",
                                               si3d = "si-eIF3d"),
                     Up = kd_up, Down = kd_down)
) |>
  tidyr::pivot_longer(c(Up, Down), names_to = "Direction", values_to = "Count") |>
  dplyr::mutate(
    Timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    Condition = factor(Condition,
                       levels = c("si-CTRL", "si-eIF3e", "si-eIF3d")),
    Direction = factor(Direction, levels = c("Up", "Down"))
  )

make_three_way_bar <- function(d, tp) {
  ggplot(d, aes(x = Condition, y = Count, fill = Direction)) +
    # Same dodge geometry as the paired figures: a dodge span just above the bar
    # width closes the Up/Down gap to ~0.03 of a category so each pair reads as one
    # condition, and the x expansion is left at the default here because three
    # groups already fill the panel.
    geom_bar(stat = "identity", position = position_dodge(width = 0.55),
             width = 0.5) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.55),
              vjust = -0.5, size = 3) +
    scale_fill_manual(values = BAR_FILL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_classic(base_size = 16) +
    theme(
      axis.title.y    = element_text(face = "bold", size = 12),
      # The bar labels carry the counts and the small knockdown bars are unreadable
      # against the si-CTRL scale, so the tick labels add nothing.
      axis.text.y     = element_blank(),
      axis.ticks.y    = element_blank(),
      axis.text.x     = element_text(size = 9),
      plot.title      = element_text(face = "bold", size = 10),
      plot.subtitle   = element_text(size = 7, colour = "grey25"),
      legend.position = "right",
      legend.text     = element_text(size = 9),
      legend.title    = element_text(size = 10)
    ) +
    labs(
      title = paste0("TE Changes: Hypoxic Response and Effect of eIF3 Knockdown - ", tp),
      subtitle = BAR_SUBTITLE,
      x = NULL, y = "Count"
    )
}

three_way_figs <- lapply(levels(three_way_data$Timepoint), function(tp) {
  fig <- make_three_way_bar(dplyr::filter(three_way_data, Timepoint == tp), tp)
  print(fig)
  ggsave(here("plots", paste0("te_changes_ctrl_si3e_si3d_", tp, ".pdf")),
         plot = fig, width = 5, height = 3)
  fig
})
names(three_way_figs) <- levels(three_way_data$Timepoint)

# Six bars per timepoint (3 conditions x 2 directions), and every knockdown bar is a
# direction matched subset of its si-CTRL bar so it can never exceed it.
stopifnot(
  "expected 3 per-timepoint figures" = length(three_way_figs) == 3,
  "each timepoint needs 6 bars (3 conditions x 2 directions)" =
    all(table(three_way_data$Timepoint) == 6),
  "a knockdown bar exceeds its si-CTRL bar" =
    all(counts_tbl$kd_up <= counts_tbl$ctrl_up &
        counts_tbl$kd_down <= counts_tbl$ctrl_down)
)

knitr::kable(
  tidyr::pivot_wider(three_way_data, id_cols = c(Timepoint, Condition),
                     names_from = Direction, values_from = Count),
  caption = "Counts behind the three condition bars"
)

cat("Saved plots/te_changes_ctrl_si3e_si3d_{1hr,4hr,24hr}.pdf\n")
```

## Fate of the si-CTRL hypoxic response under knockdown

Every gene with a significant si-CTRL hypoxic response falls into exactly one of three
fates under knockdown, so the three add to 100% and the composition has a denominator.
The count plots above report reversals without one, which hides the result: at 24hr the
si-CTRL down response is reversed 10x more often than it is maintained under si3e, and
the opposite under si3d.

```{r fate-composition, fig.width=11, fig.height=7.5}
# Per-gene fate. Maintained / Reversed / Below threshold are mutually exclusive and
# exhaustive within the si-CTRL significant set, which is what makes the stack a
# genuine composition. NA knockdown values fall to "Below threshold" rather than
# dropping out, so the denominator stays the full si-CTRL significant set.
FATE_LEVELS <- c("Maintained", "Reversed", "Below threshold")

# Deliberately NOT the Up/Down palette used above: this is a different encoding and
# reusing #D55E00/#0072B2 here would read as direction.
FATE_FILL <- c("Maintained"      = "#009E73",
               "Reversed"        = "#CC79A7",
               "Below threshold" = "grey80")

response_fate <- function(df, gene, tp, lfc_cut = LFC_CUT, padj_cut = PADJ_CUT) {
  response_pair(df, gene, tp, lfc_cut, padj_cut) |>
    dplyr::mutate(
      ctrl_dir = ifelse(x > lfc_cut, "Up", "Down"),
      k_up     = !is.na(y) & !is.na(y_padj) & y >  lfc_cut & y_padj < padj_cut,
      k_down   = !is.na(y) & !is.na(y_padj) & y < -lfc_cut & y_padj < padj_cut,
      fate = dplyr::case_when(
        (ctrl_dir == "Up" & k_up)   | (ctrl_dir == "Down" & k_down) ~ "Maintained",
        (ctrl_dir == "Up" & k_down) | (ctrl_dir == "Down" & k_up)   ~ "Reversed",
        .default = "Below threshold"
      )
    )
}

fate_all <- purrr::pmap_dfr(
  list(panel_spec$gene, panel_spec$timepoint),
  \(g, tp) response_fate(all_hypoxic_response, g, tp) |>
    dplyr::mutate(gene = g, timepoint = tp)
) |>
  dplyr::mutate(
    Gene      = factor(dplyr::recode(gene, si3e = "eIF3e", si3d = "eIF3d"),
                       levels = c("eIF3e", "eIF3d")),
    Timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    ctrl_dir  = factor(ctrl_dir, levels = c("Up", "Down")),
    fate      = factor(fate, levels = FATE_LEVELS)
  )

fate_summary <- fate_all |>
  dplyr::count(Gene, Timepoint, ctrl_dir, fate, .drop = FALSE) |>
  dplyr::group_by(Gene, Timepoint, ctrl_dir) |>
  dplyr::mutate(total = sum(n), prop = n / total) |>
  dplyr::ungroup() |>
  dplyr::filter(total > 0)

# Composition must sum to 1 within every bar, otherwise the fates are not exhaustive.
stopifnot("fate proportions do not sum to 1" = all(abs(
  fate_summary |>
    dplyr::group_by(Gene, Timepoint, ctrl_dir) |>
    dplyr::summarise(s = sum(prop), .groups = "drop") |>
    dplyr::pull(s) - 1
) < 1e-9))

fate_totals <- fate_summary |>
  dplyr::distinct(Gene, Timepoint, ctrl_dir, total)

p_fate <- ggplot(fate_summary, aes(x = ctrl_dir, y = prop, fill = fate)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = ifelse(prop >= 0.06,
                               paste0(n, "\n", scales::percent(prop, accuracy = 0.1)),
                               "")),
            position = position_stack(vjust = 0.5), size = 3.1, lineheight = 0.95) +
  geom_text(data = fate_totals, inherit.aes = FALSE,
            aes(x = ctrl_dir, y = 1.07, label = paste0("n = ", total)), size = 3.4) +
  scale_fill_manual(values = FATE_FILL, name = "Fate under knockdown") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.12),
                     breaks = seq(0, 1, 0.25), expand = c(0, 0)) +
  facet_wrap(~ Gene + Timepoint, nrow = 2) +
  theme_classic(base_size = 14) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, colour = "grey25")) +
  labs(title = "Fate of the si-CTRL hypoxic TE response under eIF3 knockdown",
       subtitle = paste0("Denominator is genes significant in si-CTRL (padj < ",
                         PADJ_CUT, ", |LFC| > ", LFC_CUT,
                         "); segments labelled where they exceed 6%"),
       x = "si-CTRL hypoxic response direction",
       y = "Share of si-CTRL significant genes")

p_fate
ggsave(here("plots", "knockdown_fate_composition_eif3.pdf"), p_fate,
       width = 11, height = 7.5)

knitr::kable(
  fate_summary |>
    dplyr::mutate(pct = round(100 * prop, 2)) |>
    dplyr::select(Gene, Timepoint, ctrl_dir, fate, n, total, pct),
  caption = "Fate composition of the si-CTRL hypoxic response"
)
```

## Reversals highlighted on the response scatter

```{r reversal-scatter, fig.width=11, fig.height=7.5}
# Same axes as the per-panel scatters, but coloured by fate so reversals are visible
# as points in the off-diagonal quadrants. Unlike any count plot this shows the
# magnitude of each reversal, which is what distinguishes a marginal 0.6 flip from a
# genuine multi-log swing. Points are drawn grey first and reversals last so they are
# not buried under the bulk.
fate_plot_df <- fate_all |>
  dplyr::filter(!is.na(x), !is.na(y)) |>
  dplyr::arrange(fate == "Maintained", fate == "Reversed")

rev_labels <- fate_plot_df |>
  dplyr::group_by(Gene, Timepoint) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_rev   = sum(fate == "Reversed"),
    n_off   = sum(abs(x) > AXIS_LIM | abs(y) > AXIS_LIM),
    .groups = "drop"
  ) |>
  dplyr::mutate(lab = paste0("n = ", n_total, ", reversed = ", n_rev,
                             "\n", n_off, " outside axis range"))

p_rev_scatter <- ggplot(fate_plot_df, aes(x = x, y = y, colour = fate)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_point(alpha = 0.45, size = 0.9) +
  geom_text(data = rev_labels, inherit.aes = FALSE,
            aes(x = Inf, y = -Inf, label = lab),
            hjust = 1.03, vjust = -0.3, size = 3, lineheight = 0.95) +
  scale_colour_manual(values = FATE_FILL, name = "Fate under knockdown") +
  coord_cartesian(xlim = c(-AXIS_LIM, AXIS_LIM), ylim = c(-AXIS_LIM, AXIS_LIM)) +
  facet_wrap(~ Gene + Timepoint, nrow = 2) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3))) +
  theme_classic(base_size = 14) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, colour = "grey25")) +
  labs(title = "Hypoxic TE response under si-CTRL vs eIF3 knockdown, coloured by fate",
       subtitle = paste0("Reversals are the off-diagonal quadrants (padj < ", PADJ_CUT,
                         ", |LFC| > ", LFC_CUT, "). Axes are zoomed, not filtered"),
       x = bquote(Log[2] ~ "si-CTRL LFC" ~ "TE: hypoxia vs normoxia"),
       y = bquote(Log[2] ~ "si-KD LFC" ~ "TE: hypoxia vs normoxia"))

p_rev_scatter
ggsave(here("plots", "knockdown_reversal_scatter_eif3.pdf"), p_rev_scatter,
       width = 11, height = 7.5)
```

## Global correlation across all mRNAs

Unlike every plot above, this one applies no si-CTRL significance filter: it is all
~11,900 genes with a TE estimate in both conditions. The regression slope is the
statistic of interest. A slope of 1 would mean knockdown leaves the hypoxic response
untouched; every slope here is well below 1, i.e. the response is globally blunted.

```{r global-correlation, fig.width=11, fig.height=7.5}
COR_LIM <- 3

# No si-CTRL padj/LFC filter here, unlike response_pair().
response_pair_all <- function(df, gene, tp) {
  df |>
    dplyr::transmute(
      gene_id,
      x = .data[[paste0("sictrl_", tp, "_te_lfc")]],
      y = .data[[paste0(gene, "_", tp, "_te_lfc")]]
    ) |>
    dplyr::filter(!is.na(x), !is.na(y))
}

cor_df <- purrr::pmap_dfr(
  list(panel_spec$gene, panel_spec$timepoint),
  \(g, tp) response_pair_all(all_hypoxic_response, g, tp) |>
    dplyr::mutate(gene = g, timepoint = tp)
) |>
  dplyr::mutate(
    Gene      = factor(dplyr::recode(gene, si3e = "eIF3e", si3d = "eIF3d"),
                       levels = c("eIF3e", "eIF3d")),
    Timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr"))
  )

# Pearson and Spearman are both reported because they disagree substantially in
# places (si3d 1hr: 0.35 vs 0.55), which means outliers are pulling Pearson around.
# All statistics use every gene, including those the axis zoom puts off-panel.
cor_stats <- cor_df |>
  dplyr::group_by(Gene, Timepoint) |>
  dplyr::summarise(
    n     = dplyr::n(),
    r     = cor(x, y),
    rho   = cor(x, y, method = "spearman"),
    slope = unname(coef(lm(y ~ x))[2]),
    n_off = sum(abs(x) > COR_LIM | abs(y) > COR_LIM),
    .groups = "drop"
  ) |>
  dplyr::mutate(lab = sprintf(
    "Pearson R = %.2f, Spearman rho = %.2f\nslope = %.2f, n = %s\n%s outside axis range",
    r, rho, slope, format(n, big.mark = ","), format(n_off, big.mark = ",")
  ))

stopifnot("correlation stats missing a panel" = nrow(cor_stats) == nrow(panel_spec))

p_cor <- ggplot(cor_df, aes(x = x, y = y)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_point(colour = "#0072B2", alpha = 0.15, size = 0.45) +
  # Dashed = y = x (no effect of knockdown); solid = actual fit, over all genes.
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey35", linewidth = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "#D55E00", linewidth = 0.8) +
  geom_text(data = cor_stats, inherit.aes = FALSE,
            aes(x = -Inf, y = Inf, label = lab),
            hjust = -0.04, vjust = 1.15, size = 3, lineheight = 0.95) +
  coord_cartesian(xlim = c(-COR_LIM, COR_LIM), ylim = c(-COR_LIM, COR_LIM)) +
  facet_wrap(~ Gene + Timepoint, nrow = 2) +
  theme_classic(base_size = 14) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, colour = "grey25")) +
  labs(title = "Hypoxic TE response, si-CTRL vs eIF3 knockdown, all mRNAs",
       subtitle = paste0("No significance filter. Dashed line is y = x; ",
                         "orange is the fit. Slope below 1 means a blunted response"),
       x = bquote(Log[2] ~ "si-CTRL LFC" ~ "TE: hypoxia vs normoxia"),
       y = bquote(Log[2] ~ "si-KD LFC" ~ "TE: hypoxia vs normoxia"))

p_cor
ggsave(here("plots", "knockdown_global_correlation_eif3.pdf"), p_cor,
       width = 11, height = 7.5)

knitr::kable(
  cor_stats |>
    dplyr::mutate(dplyr::across(c(r, rho, slope), \(v) round(v, 3))) |>
    dplyr::select(Gene, Timepoint, n, r, rho, slope, n_off),
  caption = "All-mRNA correlation between si-CTRL and knockdown hypoxic TE response"
)
```

### Same correlation as a stacked grid

```{r global-correlation-grid, fig.width=11, fig.height=8.5}
# Stacked layout matching combined_eif3e_eif3d_sig_counts.pdf: eIF3e over eIF3d, gene
# name as the bold side label, one shared header.
#
# The y axis ticks and labels are KEPT here, unlike the bar version. There the count
# sat on top of every bar so the scale was redundant; here nothing labels a point, and
# the dashed y = x reference is only interpretable against a readable scale.
make_cor_row <- function(gene_lab, show_x_title) {
  ggplot(cor_df |> dplyr::filter(Gene == gene_lab), aes(x = x, y = y)) +
    geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
    geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
    geom_point(colour = "#0072B2", alpha = 0.15, size = 0.45) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey35", linewidth = 0.5) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                colour = "#D55E00", linewidth = 0.8) +
    geom_text(data = cor_stats |> dplyr::filter(Gene == gene_lab),
              inherit.aes = FALSE, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.04, vjust = 1.15, size = 3, lineheight = 0.95) +
    coord_cartesian(xlim = c(-COR_LIM, COR_LIM), ylim = c(-COR_LIM, COR_LIM)) +
    facet_wrap(~ Timepoint, nrow = 1) +
    theme_classic(base_size = 16) +
    theme(
      axis.title.y = element_text(face = "bold", size = 20),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 16)
    ) +
    labs(
      x = if (show_x_title) {
        bquote(Log[2] ~ "si-CTRL LFC" ~ "TE: hypoxia vs normoxia")
      } else NULL,
      y = gene_lab
    )
}

cor_header <- ggdraw() +
  draw_label("Hypoxic TE response, si-CTRL vs eIF3 knockdown, all mRNAs",
             fontface = "bold", size = 18, x = 0.01, hjust = 0, y = 0.78) +
  draw_label(paste0("No significance filter, n = ",
                    format(max(cor_stats$n), big.mark = ","),
                    " genes. Dashed line is y = x; orange is the fit"),
             size = 12, colour = "grey25", x = 0.01, hjust = 0, y = 0.42) +
  draw_label(paste0("Axes are log2 TE LFC (hypoxia vs normoxia); ",
                    "slope below 1 means the response is blunted under knockdown"),
             size = 12, colour = "grey25", x = 0.01, hjust = 0, y = 0.12)

cor_grid <- plot_grid(
  cor_header,
  plot_grid(
    make_cor_row("eIF3e", show_x_title = FALSE),
    make_cor_row("eIF3d", show_x_title = TRUE),
    ncol = 1,
    align = "v",
    axis = "lr"
  ),
  ncol = 1,
  rel_heights = c(0.12, 1)
)

print(cor_grid)

ggsave(here("plots", "knockdown_global_correlation_eif3_grid.pdf"), cor_grid,
       width = 11, height = 8.5)

cat("Stacked correlation grid saved to plots/knockdown_global_correlation_eif3_grid.pdf\n")
```
