---
title: "Treatment Comparison Analysis - sictrl vs si3d/si3e"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
library(ggrepel)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

```{r load_gene_symbols}
# Load gene symbols from DESeq results
# These files were generated in the timecourse_qc analysis
deseq_1hr <- read_csv(here("output", "deseq_results_1hr_spike.csv"))
gene_symbols <- deseq_1hr %>% dplyr::select(gene_id, symbol)

cat("Loaded gene symbols for", nrow(gene_symbols), "genes\n")
```

# Create Spike-in Normalized Count Matrices for ALL Samples

This section creates spike-in normalized count matrices for ALL treatments and conditions at each timepoint.

```{r create_norm_matrices_all_1hr}
# 1 Hour - ALL treatments and conditions
metadata_all_1hr <- metadata %>%
  filter(hours == 1)

sample_cols_all_1hr <- metadata_all_1hr$sample_name
allCounts_all_1hr <- allCounts[, sample_cols_all_1hr]
spikein_all_1hr <- spikein[, sample_cols_all_1hr]

# Create spike-in normalized count matrix for ALL samples at 1hr
allCountsNorm_all_1hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_1hr,
  metadata = metadata_all_1hr,
  control_condition = "normoxia",
  spikein = spikein_all_1hr,
  keepSpikes = top500_ids
)

# Save spike-in normalized matrix
write_csv(
  allCountsNorm_all_1hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_1hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_1hr_SPIKE.csv\n")
cat("Dimensions:", dim(allCountsNorm_all_1hr_SPIKE), "\n")
```

```{r create_norm_matrices_all_4hr}
# 4 Hour - ALL treatments and conditions
metadata_all_4hr <- metadata %>%
  filter(hours == 4)

sample_cols_all_4hr <- metadata_all_4hr$sample_name
allCounts_all_4hr <- allCounts[, sample_cols_all_4hr]
spikein_all_4hr <- spikein[, sample_cols_all_4hr]

# Create spike-in normalized count matrix for ALL samples at 4hr
allCountsNorm_all_4hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_4hr,
  metadata = metadata_all_4hr,
  control_condition = "normoxia",
  spikein = spikein_all_4hr,
  keepSpikes = top500_ids
)

# Save spike-in normalized matrix
write_csv(
  allCountsNorm_all_4hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_4hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_4hr_SPIKE.csv\n")
cat("Dimensions:", dim(allCountsNorm_all_4hr_SPIKE), "\n")
```

```{r create_norm_matrices_all_24hr}
# 24 Hour - ALL treatments and conditions
metadata_all_24hr <- metadata %>%
  filter(hours == 24)

sample_cols_all_24hr <- metadata_all_24hr$sample_name
allCounts_all_24hr <- allCounts[, sample_cols_all_24hr]
spikein_all_24hr <- spikein[, sample_cols_all_24hr]

# Create spike-in normalized count matrix for ALL samples at 24hr
allCountsNorm_all_24hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_24hr,
  metadata = metadata_all_24hr,
  control_condition = "normoxia",
  spikein = spikein_all_24hr,
  keepSpikes = top500_ids
)

# Save spike-in normalized matrix
write_csv(
  allCountsNorm_all_24hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_24hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_24hr_SPIKE.csv\n")
cat("Dimensions:", dim(allCountsNorm_all_24hr_SPIKE), "\n")
```

# sictrl vs si3d in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_norm_1hr}
metadata_si3d_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_1hr}
res_rna_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_1hr}
res_ribo_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_1hr}
res_te_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_1hr}
results_si3d_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_1hr,
  res_ribo = res_ribo_si3d_norm_1hr,
  res_te = res_te_si3d_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_1hr$summary_table)
print(results_si3d_norm_1hr$plot + ggtitle("sictrl vs si3d in normoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_1hr$categorized_table %>%
  left_join(res_rna_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3d_norm_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_norm_4hr}
metadata_si3d_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_4hr}
res_rna_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_4hr}
res_ribo_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_4hr}
res_te_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_4hr}
results_si3d_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_4hr,
  res_ribo = res_ribo_si3d_norm_4hr,
  res_te = res_te_si3d_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_4hr$summary_table)
print(results_si3d_norm_4hr$plot + ggtitle("sictrl vs si3d in normoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_4hr$categorized_table %>%
  left_join(res_rna_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3d_norm_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_norm_24hr}
metadata_si3d_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_24hr}
res_rna_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_24hr}
res_ribo_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_24hr}
res_te_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_24hr}
results_si3d_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_24hr,
  res_ribo = res_ribo_si3d_norm_24hr,
  res_te = res_te_si3d_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_24hr$summary_table)
print(results_si3d_norm_24hr$plot + ggtitle("sictrl vs si3d in normoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_24hr$categorized_table %>%
  left_join(res_rna_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3d_norm_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_24hr.csv"))
```

## Combined Normoxia (All Timepoints)

### Prepare metadata
```{r metadata_si3d_norm_combined}
metadata_si3d_norm_combined <- metadata %>%
  filter(condition == "normoxia", hours %in% c(1, 4, 24)) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_combined$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (combined) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_combined}
res_rna_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_combined}
res_ribo_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_combined}
res_te_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_combined}
results_si3d_norm_combined <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_combined,
  res_ribo = res_ribo_si3d_norm_combined,
  res_te = res_te_si3d_norm_combined,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_combined$summary_table)
print(results_si3d_norm_combined$plot + ggtitle("sictrl vs si3d in normoxia (combined)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_combined.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_combined}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_combined$categorized_table %>%
  left_join(res_rna_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_combined.csv"))
write_csv(results_si3d_norm_combined$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_combined.csv"))
```

# sictrl vs si3d in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_hyp_1hr}
metadata_si3d_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_1hr}
res_rna_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_1hr}
res_ribo_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_1hr}
res_te_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_1hr}
results_si3d_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_1hr,
  res_ribo = res_ribo_si3d_hyp_1hr,
  res_te = res_te_si3d_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_1hr$summary_table)
print(results_si3d_hyp_1hr$plot + ggtitle("sictrl vs si3d in hypoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_1hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3d_hyp_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_hyp_4hr}
metadata_si3d_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_4hr}
res_rna_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_4hr}
res_ribo_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_4hr}
res_te_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_4hr}
results_si3d_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_4hr,
  res_ribo = res_ribo_si3d_hyp_4hr,
  res_te = res_te_si3d_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_4hr$summary_table)
print(results_si3d_hyp_4hr$plot + ggtitle("sictrl vs si3d in hypoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_4hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3d_hyp_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_hyp_24hr}
metadata_si3d_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_24hr}
res_rna_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_24hr}
res_ribo_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_24hr}
res_te_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_24hr}
results_si3d_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_24hr,
  res_ribo = res_ribo_si3d_hyp_24hr,
  res_te = res_te_si3d_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_24hr$summary_table)
print(results_si3d_hyp_24hr$plot + ggtitle("sictrl vs si3d in hypoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_24hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3d_hyp_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# sictrl vs si3e in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_norm_1hr}
metadata_si3e_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_1hr}
res_rna_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_1hr}
res_ribo_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_1hr}
res_te_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_1hr}
results_si3e_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_1hr,
  res_ribo = res_ribo_si3e_norm_1hr,
  res_te = res_te_si3e_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_1hr$summary_table)
print(results_si3e_norm_1hr$plot + ggtitle("sictrl vs si3e in normoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_1hr$categorized_table %>%
  left_join(res_rna_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3e_norm_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_norm_4hr}
metadata_si3e_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_4hr}
res_rna_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_4hr}
res_ribo_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_4hr}
res_te_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_4hr}
results_si3e_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_4hr,
  res_ribo = res_ribo_si3e_norm_4hr,
  res_te = res_te_si3e_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_4hr$summary_table)
print(results_si3e_norm_4hr$plot + ggtitle("sictrl vs si3e in normoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_4hr$categorized_table %>%
  left_join(res_rna_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3e_norm_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_norm_24hr}
metadata_si3e_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_24hr}
res_rna_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_24hr}
res_ribo_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_24hr}
res_te_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_24hr}
results_si3e_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_24hr,
  res_ribo = res_ribo_si3e_norm_24hr,
  res_te = res_te_si3e_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_24hr$summary_table)
print(results_si3e_norm_24hr$plot + ggtitle("sictrl vs si3e in normoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_24hr$categorized_table %>%
  left_join(res_rna_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3e_norm_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_24hr.csv"))
```

# sictrl vs si3e in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_hyp_1hr}
metadata_si3e_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_1hr}
res_rna_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_1hr}
res_ribo_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_1hr}
res_te_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_1hr}
results_si3e_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_1hr,
  res_ribo = res_ribo_si3e_hyp_1hr,
  res_te = res_te_si3e_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_1hr$summary_table)
print(results_si3e_hyp_1hr$plot + ggtitle("sictrl vs si3e in hypoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_1hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3e_hyp_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_hyp_4hr}
metadata_si3e_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_4hr}
res_rna_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_4hr}
res_ribo_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_4hr}
res_te_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_4hr}
results_si3e_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_4hr,
  res_ribo = res_ribo_si3e_hyp_4hr,
  res_te = res_te_si3e_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_4hr$summary_table)
print(results_si3e_hyp_4hr$plot + ggtitle("sictrl vs si3e in hypoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_4hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3e_hyp_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_hyp_24hr}
metadata_si3e_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_24hr}
res_rna_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_24hr}
res_ribo_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_24hr}
res_te_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_24hr}
results_si3e_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_24hr,
  res_ribo = res_ribo_si3e_hyp_24hr,
  res_te = res_te_si3e_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_24hr$summary_table)
print(results_si3e_hyp_24hr$plot + ggtitle("sictrl vs si3e in hypoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_24hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3e_hyp_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_24hr.csv"))
```

# sictrl vs si3e in NORMOXIA - Combined

## Combined Normoxia (All Timepoints)

### Prepare metadata
```{r metadata_si3e_norm_combined}
metadata_si3e_norm_combined <- metadata %>%
  filter(condition == "normoxia", hours %in% c(1, 4, 24)) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_combined$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (combined) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_combined}
res_rna_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_combined}
res_ribo_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_combined}
res_te_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_combined}
results_si3e_norm_combined <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_combined,
  res_ribo = res_ribo_si3e_norm_combined,
  res_te = res_te_si3e_norm_combined,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_combined$summary_table)
print(results_si3e_norm_combined$plot + ggtitle("sictrl vs si3e in normoxia (combined)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_combined.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_combined}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_combined$categorized_table %>%
  left_join(res_rna_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_combined.csv"))
write_csv(results_si3e_norm_combined$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_combined.csv"))
```

# Combined Visualizations

## sictrl vs si3d - All Conditions and Timepoints

```{r combined_si3d}
# Combine all si3d categorized tables with condition and timepoint labels
combined_si3d <- bind_rows(
  results_si3d_norm_1hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "1hr"),
  results_si3d_norm_4hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "4hr"),
  results_si3d_norm_24hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "24hr"),
  results_si3d_hyp_1hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "1hr"),
  results_si3d_hyp_4hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "4hr"),
  results_si3d_hyp_24hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "24hr")
) %>%
  # Add gene symbols
  left_join(gene_symbols, by = "gene_id")

# Set factor levels for proper ordering
combined_si3d$condition <- factor(combined_si3d$condition, levels = c("normoxia", "hypoxia"))
combined_si3d$timepoint <- factor(combined_si3d$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_si3d <- combined_si3d %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create faceted plot with TE count annotations
combined_si3d_plot <- ggplot(combined_si3d %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(condition ~ timepoint) +
  geom_text(data = te_counts_si3d,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("sictrl vs si3d - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_si3d_plot)

ggsave(here("plots/combined_si3d_all_conditions_timepoints.pdf"), width=9, height=5.5)
```

## sictrl vs si3e - All Conditions and Timepoints

```{r combined_si3e}
# Combine all si3e categorized tables with condition and timepoint labels
combined_si3e <- bind_rows(
  results_si3e_norm_1hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "1hr"),
  results_si3e_norm_4hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "4hr"),
  results_si3e_norm_24hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "24hr"),
  results_si3e_hyp_1hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "1hr"),
  results_si3e_hyp_4hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "4hr"),
  results_si3e_hyp_24hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "24hr")
) %>%
  # Add gene symbols
  left_join(gene_symbols, by = "gene_id")

# Set factor levels for proper ordering
combined_si3e$condition <- factor(combined_si3e$condition, levels = c("normoxia", "hypoxia"))
combined_si3e$timepoint <- factor(combined_si3e$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_si3e <- combined_si3e %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create faceted plot with TE count annotations
combined_si3e_plot <- ggplot(combined_si3e %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(condition ~ timepoint) +
  geom_text(data = te_counts_si3e,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("sictrl vs si3e - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_si3e_plot)

ggsave(here("plots/combined_si3e_all_conditions_timepoints.pdf"), width=9, height=5.5)
```

## Combined si3d and si3e Visualization

```{r combined_si3d_si3e}
# Combine si3d and si3e data with treatment labels
combined_all <- bind_rows(
  combined_si3d %>% mutate(treatment = "si3d"),
  combined_si3e %>% mutate(treatment = "si3e")
)

# Set factor levels for proper ordering
combined_all$treatment <- factor(combined_all$treatment, levels = c("si3d", "si3e"))
combined_all$condition <- factor(combined_all$condition, levels = c("normoxia", "hypoxia"))
combined_all$timepoint <- factor(combined_all$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations (including treatment)
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_combined <- combined_all %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(treatment, condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create combined faceted plot
# 4 rows (2 treatments × 2 conditions) × 3 columns (timepoints) = 12 panels
combined_all_plot <- ggplot(combined_all %>% filter(category != "Not_sig"),
                            aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(treatment + condition ~ timepoint) +
  geom_text(data = te_counts_combined,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("si3d and sictrl vs si3e - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_all_plot)

ggsave(here("plots/combined_si3d_si3e_all_conditions_timepoints.pdf"), width=10, height=8)
```

## Highlight Gene of Interest Across All Conditions

```{r highlight_gene_of_interest}
# Define gene of interest
gene_of_interest <- "TEAD1"

# Check if gene was found in data
gene_found <- combined_all %>%
  filter(symbol == gene_of_interest, category != "Not_sig")

if(nrow(gene_found) > 0) {
  cat("Found", gene_of_interest, "in", nrow(gene_found), "condition(s)\n")

  # Define colors (same as combined_all_plot)
  myCols <- c("Not_sig" = "grey80",
              "Forwarded" = "blue",
              "Exclusive" = "red",
              "Buffered" = "green",
              "TE_only" = "purple")

  # Create plot identical to combined_all_plot with gene label added
  highlighted_plot <- ggplot(combined_all %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
    # ALL genes (including TEAD1) colored by category - same as combined_all_plot
    geom_point(alpha = 0.6, size = 0.8) +
    scale_color_manual(values = myCols) +
    xlim(-5, 5) + ylim(-5, 5) +
    facet_grid(treatment + condition ~ timepoint) +
    geom_text(data = te_counts_combined,
              aes(x = -4, y = 4, label = label),
              inherit.aes = FALSE,
              size = 3,
              hjust = 0,
              vjust = 1,
              color = "black",
              fontface = "bold") +
    # ADD ONLY THIS: Label for gene of interest
    geom_text_repel(data = gene_found,
                    aes(label = symbol),
                    size = 3.5,
                    fontface = "bold",
                    box.padding = 0.5,
                    point.padding = 0.3,
                    segment.size = 0.5,
                    max.overlaps = 20,
                    show.legend = FALSE) +
    xlab("RNA-seq log2FC") +
    ylab("Ribo-seq log2FC") +
    ggtitle(paste0("Highlighting ", gene_of_interest, " - si3d and sictrl vs si3e")) +
    theme_minimal(base_size = 14) +
    theme(legend.title = element_blank(),
          legend.position = "bottom",
          strip.text = element_text(face = "bold"))

  print(highlighted_plot)

  ggsave(here("plots", paste0("highlighted_", gene_of_interest, "_all_conditions.pdf")),
         width=10, height=8)

  # Print summary table for gene of interest
  cat("\nSummary for", gene_of_interest, ":\n")
  gene_summary <- gene_found %>%
    dplyr::select(treatment, condition, timepoint, rna_lfc, ribo_lfc, te_lfc, category) %>%
    arrange(treatment, condition, timepoint)

  print(gene_summary)

} else {
  cat("Warning: Gene", gene_of_interest, "not found in the dataset\n")
}
```

```{r}

```