---
title: "Treatment Comparison Analysis - si3d/si3e vs sictrl"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

# si3d vs sictrl in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_norm_1hr}
metadata_si3d_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_1hr}
res_rna_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_1hr}
res_ribo_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_1hr}
res_te_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_1hr}
results_si3d_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_1hr,
  res_ribo = res_ribo_si3d_norm_1hr,
  res_te = res_te_si3d_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_1hr$summary_table)
print(results_si3d_norm_1hr$plot + ggtitle("si3d vs sictrl in normoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_1hr}
write_csv(results_si3d_norm_1hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3d_norm_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_norm_4hr}
metadata_si3d_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_4hr}
res_rna_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_4hr}
res_ribo_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_4hr}
res_te_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_4hr}
results_si3d_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_4hr,
  res_ribo = res_ribo_si3d_norm_4hr,
  res_te = res_te_si3d_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_4hr$summary_table)
print(results_si3d_norm_4hr$plot + ggtitle("si3d vs sictrl in normoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_4hr}
write_csv(results_si3d_norm_4hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3d_norm_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_norm_24hr}
metadata_si3d_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_24hr}
res_rna_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_24hr}
res_ribo_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_24hr}
res_te_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_24hr}
results_si3d_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_24hr,
  res_ribo = res_ribo_si3d_norm_24hr,
  res_te = res_te_si3d_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_24hr$summary_table)
print(results_si3d_norm_24hr$plot + ggtitle("si3d vs sictrl in normoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_24hr}
write_csv(results_si3d_norm_24hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3d_norm_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_24hr.csv"))
```

# si3d vs sictrl in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_hyp_1hr}
metadata_si3d_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_1hr}
res_rna_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_1hr}
res_ribo_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_1hr}
res_te_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_1hr}
results_si3d_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_1hr,
  res_ribo = res_ribo_si3d_hyp_1hr,
  res_te = res_te_si3d_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_1hr$summary_table)
print(results_si3d_hyp_1hr$plot + ggtitle("si3d vs sictrl in hypoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_1hr}
write_csv(results_si3d_hyp_1hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3d_hyp_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_hyp_4hr}
metadata_si3d_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_4hr}
res_rna_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_4hr}
res_ribo_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_4hr}
res_te_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_4hr}
results_si3d_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_4hr,
  res_ribo = res_ribo_si3d_hyp_4hr,
  res_te = res_te_si3d_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_4hr$summary_table)
print(results_si3d_hyp_4hr$plot + ggtitle("si3d vs sictrl in hypoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_4hr}
write_csv(results_si3d_hyp_4hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3d_hyp_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_hyp_24hr}
metadata_si3d_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3d vs sictrl in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_24hr}
res_rna_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_24hr}
res_ribo_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_24hr}
res_te_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3d",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_24hr}
results_si3d_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_24hr,
  res_ribo = res_ribo_si3d_hyp_24hr,
  res_te = res_te_si3d_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_24hr$summary_table)
print(results_si3d_hyp_24hr$plot + ggtitle("si3d vs sictrl in hypoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_24hr}
write_csv(results_si3d_hyp_24hr$categorized_table,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3d_hyp_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# si3e vs sictrl in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_norm_1hr}
metadata_si3e_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_1hr}
res_rna_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_1hr}
res_ribo_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_1hr}
res_te_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_1hr}
results_si3e_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_1hr,
  res_ribo = res_ribo_si3e_norm_1hr,
  res_te = res_te_si3e_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_1hr$summary_table)
print(results_si3e_norm_1hr$plot + ggtitle("si3e vs sictrl in normoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_1hr}
write_csv(results_si3e_norm_1hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3e_norm_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_norm_4hr}
metadata_si3e_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_4hr}
res_rna_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_4hr}
res_ribo_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_4hr}
res_te_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_4hr}
results_si3e_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_4hr,
  res_ribo = res_ribo_si3e_norm_4hr,
  res_te = res_te_si3e_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_4hr$summary_table)
print(results_si3e_norm_4hr$plot + ggtitle("si3e vs sictrl in normoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_4hr}
write_csv(results_si3e_norm_4hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3e_norm_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_norm_24hr}
metadata_si3e_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_24hr}
res_rna_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_24hr}
res_ribo_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_24hr}
res_te_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_24hr}
results_si3e_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_24hr,
  res_ribo = res_ribo_si3e_norm_24hr,
  res_te = res_te_si3e_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_24hr$summary_table)
print(results_si3e_norm_24hr$plot + ggtitle("si3e vs sictrl in normoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_24hr}
write_csv(results_si3e_norm_24hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3e_norm_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_24hr.csv"))
```

# si3e vs sictrl in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_hyp_1hr}
metadata_si3e_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_1hr}
res_rna_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_1hr}
res_ribo_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_1hr}
res_te_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_1hr}
results_si3e_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_1hr,
  res_ribo = res_ribo_si3e_hyp_1hr,
  res_te = res_te_si3e_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_1hr$summary_table)
print(results_si3e_hyp_1hr$plot + ggtitle("si3e vs sictrl in hypoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_1hr}
write_csv(results_si3e_hyp_1hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3e_hyp_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_hyp_4hr}
metadata_si3e_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_4hr}
res_rna_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_4hr}
res_ribo_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_4hr}
res_te_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_4hr}
results_si3e_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_4hr,
  res_ribo = res_ribo_si3e_hyp_4hr,
  res_te = res_te_si3e_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_4hr$summary_table)
print(results_si3e_hyp_4hr$plot + ggtitle("si3e vs sictrl in hypoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_4hr}
write_csv(results_si3e_hyp_4hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3e_hyp_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_hyp_24hr}
metadata_si3e_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("si3e vs sictrl in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_24hr}
res_rna_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_24hr}
res_ribo_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_24hr}
res_te_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "sictrl",
  test_condition = "si3e",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_24hr}
results_si3e_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_24hr,
  res_ribo = res_ribo_si3e_hyp_24hr,
  res_te = res_te_si3e_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_24hr$summary_table)
print(results_si3e_hyp_24hr$plot + ggtitle("si3e vs sictrl in hypoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_24hr}
write_csv(results_si3e_hyp_24hr$categorized_table,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3e_hyp_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_24hr.csv"))
```
