---
title: "Treatment Comparison Analysis - sictrl vs si3d/si3e"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
library(ggrepel)
library(clusterProfiler)
library(msigdbr)
library(pheatmap)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

```{r load_gene_symbols}
# Load gene symbols from DESeq results
# These files were generated in the timecourse_qc analysis
deseq_1hr <- read_csv(here("output", "deseq_results_1hr_spike.csv"))
gene_symbols <- deseq_1hr %>% dplyr::select(gene_id, symbol)

cat("Loaded gene symbols for", nrow(gene_symbols), "genes\n")
```

# Create Spike-in Normalized Count Matrices for ALL Samples

This section creates spike-in normalized count matrices for ALL treatments and conditions at each timepoint.

```{r create_norm_matrices_all_1hr}
# 1 Hour - ALL treatments and conditions
metadata_all_1hr <- metadata %>%
  filter(hours == 1)

sample_cols_all_1hr <- metadata_all_1hr$sample_name
allCounts_all_1hr <- allCounts[, sample_cols_all_1hr]
spikein_all_1hr <- spikein[, sample_cols_all_1hr]

# Create spike-in normalized count matrix for ALL samples at 1hr
allCountsNorm_all_1hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_1hr,
  metadata = metadata_all_1hr,
  control_condition = "normoxia",
  spikein = spikein_all_1hr,
  keepSpikes = top500_ids
)

cat("Before removing spike-ins - Dimensions:", dim(allCountsNorm_all_1hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_all_1hr_SPIKE <- allCountsNorm_all_1hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("After removing spike-ins - Dimensions:", dim(allCountsNorm_all_1hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_all_1hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_1hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_1hr_SPIKE.csv\n")
```

```{r create_norm_matrices_all_4hr}
# 4 Hour - ALL treatments and conditions
metadata_all_4hr <- metadata %>%
  filter(hours == 4)

sample_cols_all_4hr <- metadata_all_4hr$sample_name
allCounts_all_4hr <- allCounts[, sample_cols_all_4hr]
spikein_all_4hr <- spikein[, sample_cols_all_4hr]

# Create spike-in normalized count matrix for ALL samples at 4hr
allCountsNorm_all_4hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_4hr,
  metadata = metadata_all_4hr,
  control_condition = "normoxia",
  spikein = spikein_all_4hr,
  keepSpikes = top500_ids
)

cat("Before removing spike-ins - Dimensions:", dim(allCountsNorm_all_4hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_all_4hr_SPIKE <- allCountsNorm_all_4hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("After removing spike-ins - Dimensions:", dim(allCountsNorm_all_4hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_all_4hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_4hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_4hr_SPIKE.csv\n")
```

```{r create_norm_matrices_all_24hr}
# 24 Hour - ALL treatments and conditions
metadata_all_24hr <- metadata %>%
  filter(hours == 24)

sample_cols_all_24hr <- metadata_all_24hr$sample_name
allCounts_all_24hr <- allCounts[, sample_cols_all_24hr]
spikein_all_24hr <- spikein[, sample_cols_all_24hr]

# Create spike-in normalized count matrix for ALL samples at 24hr
allCountsNorm_all_24hr_SPIKE <- createNormMatrix(
  allCounts = allCounts_all_24hr,
  metadata = metadata_all_24hr,
  control_condition = "normoxia",
  spikein = spikein_all_24hr,
  keepSpikes = top500_ids
)

cat("Before removing spike-ins - Dimensions:", dim(allCountsNorm_all_24hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_all_24hr_SPIKE <- allCountsNorm_all_24hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("After removing spike-ins - Dimensions:", dim(allCountsNorm_all_24hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_all_24hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_all_24hr_SPIKE.csv")
)
cat("Saved allCountsNorm_all_24hr_SPIKE.csv\n")
```

# sictrl vs si3d in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_norm_1hr}
metadata_si3d_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_1hr}
res_rna_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_1hr}
res_ribo_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_1hr}
res_te_si3d_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_1hr}
results_si3d_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_1hr,
  res_ribo = res_ribo_si3d_norm_1hr,
  res_te = res_te_si3d_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_1hr$summary_table)
print(results_si3d_norm_1hr$plot + ggtitle("sictrl vs si3d in normoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_1hr$categorized_table %>%
  left_join(res_rna_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3d_norm_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_norm_4hr}
metadata_si3d_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_4hr}
res_rna_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_4hr}
res_ribo_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_4hr}
res_te_si3d_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_4hr}
results_si3d_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_4hr,
  res_ribo = res_ribo_si3d_norm_4hr,
  res_te = res_te_si3d_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_4hr$summary_table)
print(results_si3d_norm_4hr$plot + ggtitle("sictrl vs si3d in normoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_4hr$categorized_table %>%
  left_join(res_rna_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3d_norm_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_norm_24hr}
metadata_si3d_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_24hr}
res_rna_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_24hr}
res_ribo_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_24hr}
res_te_si3d_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_24hr}
results_si3d_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_24hr,
  res_ribo = res_ribo_si3d_norm_24hr,
  res_te = res_te_si3d_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_24hr$summary_table)
print(results_si3d_norm_24hr$plot + ggtitle("sictrl vs si3d in normoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_24hr$categorized_table %>%
  left_join(res_rna_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3d_norm_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_24hr.csv"))
```

## Combined Normoxia (All Timepoints)

### Prepare metadata
```{r metadata_si3d_norm_combined}
metadata_si3d_norm_combined <- metadata %>%
  filter(condition == "normoxia", hours %in% c(1, 4, 24)) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_combined$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in normoxia (combined) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_norm_combined}
res_rna_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_norm_combined}
res_ribo_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_norm_combined}
res_te_si3d_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_norm_combined,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_norm_combined}
results_si3d_norm_combined <- categorize_translation_changes(
  res_rna = res_rna_si3d_norm_combined,
  res_ribo = res_ribo_si3d_norm_combined,
  res_te = res_te_si3d_norm_combined,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_norm_combined$summary_table)
print(results_si3d_norm_combined$plot + ggtitle("sictrl vs si3d in normoxia (combined)"))

ggsave(here("plots/si3d_vs_sictrl_normoxia_combined.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_norm_combined}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_norm_combined$categorized_table %>%
  left_join(res_rna_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_normoxia_combined.csv"))
write_csv(results_si3d_norm_combined$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_normoxia_combined.csv"))
```

# sictrl vs si3d in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3d_hyp_1hr}
metadata_si3d_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_1hr}
res_rna_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_1hr}
res_ribo_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_1hr}
res_te_si3d_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_1hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_1hr}
results_si3d_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_1hr,
  res_ribo = res_ribo_si3d_hyp_1hr,
  res_te = res_te_si3d_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_1hr$summary_table)
print(results_si3d_hyp_1hr$plot + ggtitle("sictrl vs si3d in hypoxia (1hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_1hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3d_hyp_1hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3d_hyp_4hr}
metadata_si3d_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_4hr}
res_rna_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_4hr}
res_ribo_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_4hr}
res_te_si3d_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_4hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_4hr}
results_si3d_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_4hr,
  res_ribo = res_ribo_si3d_hyp_4hr,
  res_te = res_te_si3d_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_4hr$summary_table)
print(results_si3d_hyp_4hr$plot + ggtitle("sictrl vs si3d in hypoxia (4hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_4hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3d_hyp_4hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3d_hyp_24hr}
metadata_si3d_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3d in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3d_hyp_24hr}
res_rna_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3d_hyp_24hr}
res_ribo_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3d_hyp_24hr}
res_te_si3d_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_hyp_24hr,
  control_condition = "si3d",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3d_hyp_24hr}
results_si3d_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_hyp_24hr,
  res_ribo = res_ribo_si3d_hyp_24hr,
  res_te = res_te_si3d_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3d_hyp_24hr$summary_table)
print(results_si3d_hyp_24hr$plot + ggtitle("sictrl vs si3d in hypoxia (24hr)"))

ggsave(here("plots/si3d_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3d_hyp_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3d_hyp_24hr$categorized_table %>%
  left_join(res_rna_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3d_hyp_24hr$summary_table,
          here("output", "translation_summary_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# sictrl vs si3e in NORMOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_norm_1hr}
metadata_si3e_norm_1hr <- metadata %>%
  filter(condition == "normoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_1hr}
res_rna_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_1hr}
res_ribo_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_1hr}
res_te_si3e_norm_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_1hr}
results_si3e_norm_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_1hr,
  res_ribo = res_ribo_si3e_norm_1hr,
  res_te = res_te_si3e_norm_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_1hr$summary_table)
print(results_si3e_norm_1hr$plot + ggtitle("sictrl vs si3e in normoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_1hr$categorized_table %>%
  left_join(res_rna_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"))
write_csv(results_si3e_norm_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_norm_4hr}
metadata_si3e_norm_4hr <- metadata %>%
  filter(condition == "normoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_4hr}
res_rna_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_4hr}
res_ribo_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_4hr}
res_te_si3e_norm_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_4hr}
results_si3e_norm_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_4hr,
  res_ribo = res_ribo_si3e_norm_4hr,
  res_te = res_te_si3e_norm_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_4hr$summary_table)
print(results_si3e_norm_4hr$plot + ggtitle("sictrl vs si3e in normoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_4hr$categorized_table %>%
  left_join(res_rna_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv"))
write_csv(results_si3e_norm_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_norm_24hr}
metadata_si3e_norm_24hr <- metadata %>%
  filter(condition == "normoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_24hr}
res_rna_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_24hr}
res_ribo_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_24hr}
res_te_si3e_norm_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_24hr}
results_si3e_norm_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_24hr,
  res_ribo = res_ribo_si3e_norm_24hr,
  res_te = res_te_si3e_norm_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_24hr$summary_table)
print(results_si3e_norm_24hr$plot + ggtitle("sictrl vs si3e in normoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_24hr$categorized_table %>%
  left_join(res_rna_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv"))
write_csv(results_si3e_norm_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_24hr.csv"))
```

# sictrl vs si3e in HYPOXIA

## 1 Hour

### Prepare metadata
```{r metadata_si3e_hyp_1hr}
metadata_si3e_hyp_1hr <- metadata %>%
  filter(condition == "hypoxia", hours == 1) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (1hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_1hr}
res_rna_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_1hr}
res_ribo_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_1hr}
res_te_si3e_hyp_1hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_1hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_1hr}
results_si3e_hyp_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_1hr,
  res_ribo = res_ribo_si3e_hyp_1hr,
  res_te = res_te_si3e_hyp_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_1hr$summary_table)
print(results_si3e_hyp_1hr$plot + ggtitle("sictrl vs si3e in hypoxia (1hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_1hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_1hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_1hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_1hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
write_csv(results_si3e_hyp_1hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_1hr.csv"))
```

## 4 Hours

### Prepare metadata
```{r metadata_si3e_hyp_4hr}
metadata_si3e_hyp_4hr <- metadata %>%
  filter(condition == "hypoxia", hours == 4) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (4hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_4hr}
res_rna_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_4hr}
res_ribo_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_4hr}
res_te_si3e_hyp_4hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_4hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_4hr}
results_si3e_hyp_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_4hr,
  res_ribo = res_ribo_si3e_hyp_4hr,
  res_te = res_te_si3e_hyp_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_4hr$summary_table)
print(results_si3e_hyp_4hr$plot + ggtitle("sictrl vs si3e in hypoxia (4hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_4hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_4hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_4hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_4hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
write_csv(results_si3e_hyp_4hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_4hr.csv"))
```

## 24 Hours

### Prepare metadata
```{r metadata_si3e_hyp_24hr}
metadata_si3e_hyp_24hr <- metadata %>%
  filter(condition == "hypoxia", hours == 24) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_hyp_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in hypoxia (24hr) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_hyp_24hr}
res_rna_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_hyp_24hr}
res_ribo_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_hyp_24hr}
res_te_si3e_hyp_24hr <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_hyp_24hr,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_hyp_24hr}
results_si3e_hyp_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_hyp_24hr,
  res_ribo = res_ribo_si3e_hyp_24hr,
  res_te = res_te_si3e_hyp_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_hyp_24hr$summary_table)
print(results_si3e_hyp_24hr$plot + ggtitle("sictrl vs si3e in hypoxia (24hr)"))

ggsave(here("plots/si3e_vs_sictrl_hypoxia_24hr.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_hyp_24hr}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_hyp_24hr$categorized_table %>%
  left_join(res_rna_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_hyp_24hr %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))
write_csv(results_si3e_hyp_24hr$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_hypoxia_24hr.csv"))
```

# sictrl vs si3e in NORMOXIA - Combined

## Combined Normoxia (All Timepoints)

### Prepare metadata
```{r metadata_si3e_norm_combined}
metadata_si3e_norm_combined <- metadata %>%
  filter(condition == "normoxia", hours %in% c(1, 4, 24)) %>%
  filter(treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_combined$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

cat("sictrl vs si3e in normoxia (combined) - Samples:", length(sample_cols), "\n")
```

### Run RNA analysis
```{r rna_si3e_norm_combined}
res_rna_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RNA"
)
```

### Run Ribo analysis
```{r ribo_si3e_norm_combined}
res_ribo_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "RIBO"
)
```

### Run TE analysis
```{r te_si3e_norm_combined}
res_te_si3e_norm_combined <- deFunction(
  countsMatrix = allCounts_filtered,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_norm_combined,
  control_condition = "si3e",
  test_condition = "sictrl",
  spikeInMatrix = spikein_filtered,
  keepSpikes = top500_ids,
  mode = "TE"
)
```

### Categorize translation changes
```{r categorize_si3e_norm_combined}
results_si3e_norm_combined <- categorize_translation_changes(
  res_rna = res_rna_si3e_norm_combined,
  res_ribo = res_ribo_si3e_norm_combined,
  res_te = res_te_si3e_norm_combined,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

print(results_si3e_norm_combined$summary_table)
print(results_si3e_norm_combined$plot + ggtitle("sictrl vs si3e in normoxia (combined)"))

ggsave(here("plots/si3e_vs_sictrl_normoxia_combined.pdf"), width=4.5, height=3)
```

### Save results
```{r save_si3e_norm_combined}
# Add padj values and gene symbols to categorized table
categorized_with_padj <- results_si3e_norm_combined$categorized_table %>%
  left_join(res_rna_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_norm_combined %>% dplyr::select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id") %>%
  left_join(gene_symbols, by = "gene_id")

write_csv(categorized_with_padj,
          here("output", "translation_categories_si3e_vs_sictrl_normoxia_combined.csv"))
write_csv(results_si3e_norm_combined$summary_table,
          here("output", "translation_summary_si3e_vs_sictrl_normoxia_combined.csv"))
```

# Bar Plots: TE Changes by Timepoint

This section creates bar plots showing the number of genes with significant TE changes (up and down) for eIF3d and eIF3e knockdowns across normoxia and hypoxia timepoints.

## Helper Function for Counting Significant TE Changes

```{r count_te_changes_function}
# Helper function to count significant TE changes
count_te_changes <- function(res_te, padj_cutoff = 0.05, lfc_cutoff = 0.5) {
  up <- sum(res_te$padj < padj_cutoff & res_te$log2FoldChange > lfc_cutoff, na.rm = TRUE)
  down <- sum(res_te$padj < padj_cutoff & res_te$log2FoldChange < -lfc_cutoff, na.rm = TRUE)
  return(c(up = up, down = down))
}
```

## eIF3d-induced TE Changes

```{r te_barplot_eif3d}
# Count TE changes for si3d across all timepoints
te_counts_3d <- data.frame(
  Condition = factor(rep(c("Norm", "Norm", "Norm", "Hyp", "Hyp", "Hyp"), each = 2),
                     levels = c("Norm", "Hyp")),
  Hour = factor(rep(c("1", "4", "24", "1", "4", "24"), each = 2),
                levels = c("1", "4", "24")),
  Direction = rep(c("Up", "Down"), 6),
  Count = c(
    count_te_changes(res_te_si3d_norm_1hr),
    count_te_changes(res_te_si3d_norm_4hr),
    count_te_changes(res_te_si3d_norm_24hr),
    count_te_changes(res_te_si3d_hyp_1hr),
    count_te_changes(res_te_si3d_hyp_4hr),
    count_te_changes(res_te_si3d_hyp_24hr)
  )
)

te_barplot_3d <- ggplot(te_counts_3d, aes(x = Hour, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Up" = "#D55E00", "Down" = "#0072B2")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_grid(~ Condition, switch = "x") +
  theme_prism() +
  theme(
    legend.position = "top",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "eIF3d-induced TE changes",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    y = "Number of genes",
    x = NULL
  )

print(te_barplot_3d)
ggsave(here("plots/eif3d_te_changes_barplot.pdf"), te_barplot_3d, width = 7, height = 5)
```

## eIF3e-induced TE Changes

```{r te_barplot_eif3e}
# Count TE changes for si3e across all timepoints
te_counts_3e <- data.frame(
  Condition = factor(rep(c("Norm", "Norm", "Norm", "Hyp", "Hyp", "Hyp"), each = 2),
                     levels = c("Norm", "Hyp")),
  Hour = factor(rep(c("1", "4", "24", "1", "4", "24"), each = 2),
                levels = c("1", "4", "24")),
  Direction = rep(c("Up", "Down"), 6),
  Count = c(
    count_te_changes(res_te_si3e_norm_1hr),
    count_te_changes(res_te_si3e_norm_4hr),
    count_te_changes(res_te_si3e_norm_24hr),
    count_te_changes(res_te_si3e_hyp_1hr),
    count_te_changes(res_te_si3e_hyp_4hr),
    count_te_changes(res_te_si3e_hyp_24hr)
  )
)

te_barplot_3e <- ggplot(te_counts_3e, aes(x = Hour, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Up" = "#D55E00", "Down" = "#0072B2")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_grid(~ Condition, switch = "x") +
  theme_prism() +
  theme(
    legend.position = "top",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "eIF3e-induced TE changes",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    y = "Number of genes",
    x = NULL
  )

print(te_barplot_3e)
ggsave(here("plots/eif3e_te_changes_barplot.pdf"), te_barplot_3e, width = 7, height = 5)
```

## eIF3d-induced RNA and Ribo Changes

```{r rna_ribo_barplot_eif3d}
# Count RNA and Ribo changes for si3d across all timepoints
rna_ribo_counts_3d <- data.frame(
  Condition = factor(rep(c("Norm", "Norm", "Norm", "Hyp", "Hyp", "Hyp"), each = 2, times = 2),
                     levels = c("Norm", "Hyp")),
  Hour = factor(rep(c("1", "4", "24", "1", "4", "24"), each = 2, times = 2),
                levels = c("1", "4", "24")),
  Measurement = factor(rep(c("RNA", "Ribo"), each = 12),
                       levels = c("RNA", "Ribo")),
  Direction = rep(c("Up", "Down"), 12),
  Count = c(
    # RNA counts
    count_te_changes(res_rna_si3d_norm_1hr),
    count_te_changes(res_rna_si3d_norm_4hr),
    count_te_changes(res_rna_si3d_norm_24hr),
    count_te_changes(res_rna_si3d_hyp_1hr),
    count_te_changes(res_rna_si3d_hyp_4hr),
    count_te_changes(res_rna_si3d_hyp_24hr),
    # Ribo counts
    count_te_changes(res_ribo_si3d_norm_1hr),
    count_te_changes(res_ribo_si3d_norm_4hr),
    count_te_changes(res_ribo_si3d_norm_24hr),
    count_te_changes(res_ribo_si3d_hyp_1hr),
    count_te_changes(res_ribo_si3d_hyp_4hr),
    count_te_changes(res_ribo_si3d_hyp_24hr)
  )
)

rna_ribo_barplot_3d <- ggplot(rna_ribo_counts_3d, aes(x = Hour, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Up" = "#D55E00", "Down" = "#0072B2")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_grid(Measurement ~ Condition, switch = "x") +
  theme_prism() +
  theme(
    legend.position = "top",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "eIF3d-induced RNA and Ribo changes",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    y = "Number of genes",
    x = NULL
  )

print(rna_ribo_barplot_3d)
ggsave(here("plots/eif3d_rna_ribo_changes_barplot.pdf"), rna_ribo_barplot_3d, width = 7, height = 6)
```

## eIF3e-induced RNA and Ribo Changes

```{r rna_ribo_barplot_eif3e}
# Count RNA and Ribo changes for si3e across all timepoints
rna_ribo_counts_3e <- data.frame(
  Condition = factor(rep(c("Norm", "Norm", "Norm", "Hyp", "Hyp", "Hyp"), each = 2, times = 2),
                     levels = c("Norm", "Hyp")),
  Hour = factor(rep(c("1", "4", "24", "1", "4", "24"), each = 2, times = 2),
                levels = c("1", "4", "24")),
  Measurement = factor(rep(c("RNA", "Ribo"), each = 12),
                       levels = c("RNA", "Ribo")),
  Direction = rep(c("Up", "Down"), 12),
  Count = c(
    # RNA counts
    count_te_changes(res_rna_si3e_norm_1hr),
    count_te_changes(res_rna_si3e_norm_4hr),
    count_te_changes(res_rna_si3e_norm_24hr),
    count_te_changes(res_rna_si3e_hyp_1hr),
    count_te_changes(res_rna_si3e_hyp_4hr),
    count_te_changes(res_rna_si3e_hyp_24hr),
    # Ribo counts
    count_te_changes(res_ribo_si3e_norm_1hr),
    count_te_changes(res_ribo_si3e_norm_4hr),
    count_te_changes(res_ribo_si3e_norm_24hr),
    count_te_changes(res_ribo_si3e_hyp_1hr),
    count_te_changes(res_ribo_si3e_hyp_4hr),
    count_te_changes(res_ribo_si3e_hyp_24hr)
  )
)

rna_ribo_barplot_3e <- ggplot(rna_ribo_counts_3e, aes(x = Hour, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Up" = "#D55E00", "Down" = "#0072B2")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_grid(Measurement ~ Condition, switch = "x") +
  theme_prism() +
  theme(
    legend.position = "top",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(0.5, "lines"),
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "eIF3e-induced RNA and Ribo changes",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    y = "Number of genes",
    x = NULL
  )

print(rna_ribo_barplot_3e)
ggsave(here("plots/eif3e_rna_ribo_changes_barplot.pdf"), rna_ribo_barplot_3e, width = 7, height = 6)
```

# Venn Diagrams: Overlap of Significant TE Changes Across Timepoints

This section creates Venn diagrams showing the overlap of genes with significant TE changes (padj < 0.05, |LFC| > 0.5) across the three timepoints (1hr, 4hr, 24hr).

## Helper Function for Extracting Significant Gene IDs

```{r get_sig_genes_function}
# Helper function to get gene IDs with significant TE changes
get_sig_te_genes <- function(res_te, padj_cutoff = 0.05, lfc_cutoff = 0.5) {
  res_te %>%
    filter(padj < padj_cutoff & abs(log2FoldChange) > lfc_cutoff) %>%
    pull(gene_id)
}
```

## eIF3e-induced TE Changes: Normoxia

```{r venn_te_si3e_normoxia}
# Get significant genes at each timepoint
si3e_norm_1hr_genes <- get_sig_te_genes(res_te_si3e_norm_1hr)
si3e_norm_4hr_genes <- get_sig_te_genes(res_te_si3e_norm_4hr)
si3e_norm_24hr_genes <- get_sig_te_genes(res_te_si3e_norm_24hr)

# Create list for Venn diagram
venn_list_si3e_norm <- list(
  "1hr" = si3e_norm_1hr_genes,
  "4hr" = si3e_norm_4hr_genes,
  "24hr" = si3e_norm_24hr_genes
)

# Create Venn diagram
venn_si3e_norm <- ggvenn(venn_list_si3e_norm,
                         fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                         stroke_size = 0.5,
                         set_name_size = 5,
                         text_size = 4) +
  ggtitle("eIF3e-induced TE changes in Normoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_si3e_norm)
ggsave(here("plots/venn_te_si3e_normoxia.pdf"), venn_si3e_norm, width = 6, height = 5)
```

## eIF3e-induced TE Changes: Hypoxia

```{r venn_te_si3e_hypoxia}
# Get significant genes at each timepoint
si3e_hyp_1hr_genes <- get_sig_te_genes(res_te_si3e_hyp_1hr)
si3e_hyp_4hr_genes <- get_sig_te_genes(res_te_si3e_hyp_4hr)
si3e_hyp_24hr_genes <- get_sig_te_genes(res_te_si3e_hyp_24hr)

# Create list for Venn diagram
venn_list_si3e_hyp <- list(
  "1hr" = si3e_hyp_1hr_genes,
  "4hr" = si3e_hyp_4hr_genes,
  "24hr" = si3e_hyp_24hr_genes
)

# Create Venn diagram
venn_si3e_hyp <- ggvenn(venn_list_si3e_hyp,
                        fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                        stroke_size = 0.5,
                        set_name_size = 5,
                        text_size = 4) +
  ggtitle("eIF3e-induced TE changes in Hypoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_si3e_hyp)
ggsave(here("plots/venn_te_si3e_hypoxia.pdf"), venn_si3e_hyp, width = 6, height = 5)
```

## eIF3d-induced TE Changes: Normoxia

```{r venn_te_si3d_normoxia}
# Get significant genes at each timepoint
si3d_norm_1hr_genes <- get_sig_te_genes(res_te_si3d_norm_1hr)
si3d_norm_4hr_genes <- get_sig_te_genes(res_te_si3d_norm_4hr)
si3d_norm_24hr_genes <- get_sig_te_genes(res_te_si3d_norm_24hr)

# Create list for Venn diagram
venn_list_si3d_norm <- list(
  "1hr" = si3d_norm_1hr_genes,
  "4hr" = si3d_norm_4hr_genes,
  "24hr" = si3d_norm_24hr_genes
)

# Create Venn diagram
venn_si3d_norm <- ggvenn(venn_list_si3d_norm,
                         fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                         stroke_size = 0.5,
                         set_name_size = 5,
                         text_size = 4) +
  ggtitle("eIF3d-induced TE changes in Normoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_si3d_norm)
ggsave(here("plots/venn_te_si3d_normoxia.pdf"), venn_si3d_norm, width = 6, height = 5)
```

## eIF3d-induced TE Changes: Hypoxia

```{r venn_te_si3d_hypoxia}
# Get significant genes at each timepoint
si3d_hyp_1hr_genes <- get_sig_te_genes(res_te_si3d_hyp_1hr)
si3d_hyp_4hr_genes <- get_sig_te_genes(res_te_si3d_hyp_4hr)
si3d_hyp_24hr_genes <- get_sig_te_genes(res_te_si3d_hyp_24hr)

# Create list for Venn diagram
venn_list_si3d_hyp <- list(
  "1hr" = si3d_hyp_1hr_genes,
  "4hr" = si3d_hyp_4hr_genes,
  "24hr" = si3d_hyp_24hr_genes
)

# Create Venn diagram
venn_si3d_hyp <- ggvenn(venn_list_si3d_hyp,
                        fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                        stroke_size = 0.5,
                        set_name_size = 5,
                        text_size = 4) +
  ggtitle("eIF3d-induced TE changes in Hypoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_si3d_hyp)
ggsave(here("plots/venn_te_si3d_hypoxia.pdf"), venn_si3d_hyp, width = 6, height = 5)
```

# Venn Diagrams: TE Normoxia vs Hypoxia Overlap by Timepoint

This section creates Venn diagrams showing the overlap of significant TE changes between normoxia and hypoxia at each timepoint, with Up and Down genes shown separately.

## Helper Functions for Up/Down Gene Extraction

```{r get_up_down_genes_function}
# Helper function to get gene IDs with significant UP TE changes
get_sig_te_up <- function(res_te, padj_cutoff = 0.05, lfc_cutoff = 0.5) {
  res_te %>%
    filter(padj < padj_cutoff & log2FoldChange > lfc_cutoff) %>%
    pull(gene_id)
}

# Helper function to get gene IDs with significant DOWN TE changes
get_sig_te_down <- function(res_te, padj_cutoff = 0.05, lfc_cutoff = 0.5) {
  res_te %>%
    filter(padj < padj_cutoff & log2FoldChange < -lfc_cutoff) %>%
    pull(gene_id)
}
```

## eIF3e: TE Normoxia vs Hypoxia Overlap

```{r venn_te_si3e_norm_vs_hyp}
# Create individual Venn diagrams for each timepoint and direction

# 1hr Up
venn_3e_1hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3e_norm_1hr),
       "Hyp" = get_sig_te_up(res_te_si3e_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 1hr Down
venn_3e_1hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3e_norm_1hr),
       "Hyp" = get_sig_te_down(res_te_si3e_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Up
venn_3e_4hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3e_norm_4hr),
       "Hyp" = get_sig_te_up(res_te_si3e_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Down
venn_3e_4hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3e_norm_4hr),
       "Hyp" = get_sig_te_down(res_te_si3e_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Up
venn_3e_24hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3e_norm_24hr),
       "Hyp" = get_sig_te_up(res_te_si3e_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Down
venn_3e_24hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3e_norm_24hr),
       "Hyp" = get_sig_te_down(res_te_si3e_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Combine into grid using patchwork
venn_grid_3e <- (venn_3e_1hr_up | venn_3e_1hr_down) /
                (venn_3e_4hr_up | venn_3e_4hr_down) /
                (venn_3e_24hr_up | venn_3e_24hr_down) +
  plot_annotation(
    title = "eIF3e-induced TE changes: Normoxia vs Hypoxia",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11)
    )
  )

print(venn_grid_3e)
ggsave(here("plots/venn_te_si3e_norm_vs_hyp_grid.pdf"), venn_grid_3e, width = 8, height = 10)
```

## eIF3d: TE Normoxia vs Hypoxia Overlap

```{r venn_te_si3d_norm_vs_hyp}
# Create individual Venn diagrams for each timepoint and direction

# 1hr Up
venn_3d_1hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3d_norm_1hr),
       "Hyp" = get_sig_te_up(res_te_si3d_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 1hr Down
venn_3d_1hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3d_norm_1hr),
       "Hyp" = get_sig_te_down(res_te_si3d_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Up
venn_3d_4hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3d_norm_4hr),
       "Hyp" = get_sig_te_up(res_te_si3d_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Down
venn_3d_4hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3d_norm_4hr),
       "Hyp" = get_sig_te_down(res_te_si3d_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Up
venn_3d_24hr_up <- ggvenn(
  list("Norm" = get_sig_te_up(res_te_si3d_norm_24hr),
       "Hyp" = get_sig_te_up(res_te_si3d_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Down
venn_3d_24hr_down <- ggvenn(
  list("Norm" = get_sig_te_down(res_te_si3d_norm_24hr),
       "Hyp" = get_sig_te_down(res_te_si3d_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Combine into grid using patchwork
venn_grid_3d <- (venn_3d_1hr_up | venn_3d_1hr_down) /
                (venn_3d_4hr_up | venn_3d_4hr_down) /
                (venn_3d_24hr_up | venn_3d_24hr_down) +
  plot_annotation(
    title = "eIF3d-induced TE changes: Normoxia vs Hypoxia",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11)
    )
  )

print(venn_grid_3d)
ggsave(here("plots/venn_te_si3d_norm_vs_hyp_grid.pdf"), venn_grid_3d, width = 8, height = 10)
```

# Density Plots: TE Changes Normoxia vs Hypoxia

This section creates density plots showing the distribution of TE log2 fold changes, comparing normoxia and hypoxia at each timepoint.

## eIF3e: TE Density Plots

```{r density_te_si3e}
# Create density plot for 1hr
density_3e_1hr <- ggplot() +
  geom_density(data = res_te_si3e_norm_1hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3e_hyp_1hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "top") +
  labs(title = "1hr", x = "TE log2FC", y = "Density")

# Create density plot for 4hr
density_3e_4hr <- ggplot() +
  geom_density(data = res_te_si3e_norm_4hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3e_hyp_4hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "4hr", x = "TE log2FC", y = "Density")

# Create density plot for 24hr
density_3e_24hr <- ggplot() +
  geom_density(data = res_te_si3e_norm_24hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3e_hyp_24hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#56B4E9", "Hypoxia" = "#004080"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "24hr", x = "TE log2FC", y = "Density")

# Combine into column using patchwork
density_grid_3e <- density_3e_1hr / density_3e_4hr / density_3e_24hr +
  plot_annotation(
    title = "eIF3e-induced TE changes: Normoxia vs Hypoxia",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  )

print(density_grid_3e)
ggsave(here("plots/density_te_si3e_norm_vs_hyp.pdf"), density_grid_3e, width = 6, height = 9)
```

## eIF3d: TE Density Plots

```{r density_te_si3d}
# Create density plot for 1hr
density_3d_1hr <- ggplot() +
  geom_density(data = res_te_si3d_norm_1hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_1hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "top") +
  labs(title = "1hr", x = "TE log2FC", y = "Density")

# Create density plot for 4hr
density_3d_4hr <- ggplot() +
  geom_density(data = res_te_si3d_norm_4hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_4hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "4hr", x = "TE log2FC", y = "Density")

# Create density plot for 24hr
density_3d_24hr <- ggplot() +
  geom_density(data = res_te_si3d_norm_24hr, aes(x = log2FoldChange, color = "Normoxia", fill = "Normoxia"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_24hr, aes(x = log2FoldChange, color = "Hypoxia", fill = "Hypoxia"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  scale_fill_manual(values = c("Normoxia" = "#E69F00", "Hypoxia" = "#8B4000"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-3, 3)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "24hr", x = "TE log2FC", y = "Density")

# Combine into column using patchwork
density_grid_3d <- density_3d_1hr / density_3d_4hr / density_3d_24hr +
  plot_annotation(
    title = "eIF3d-induced TE changes: Normoxia vs Hypoxia",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  )

print(density_grid_3d)
ggsave(here("plots/density_te_si3d_norm_vs_hyp.pdf"), density_grid_3d, width = 6, height = 9)
```

# Venn Diagrams: eIF3e vs eIF3d Overlap

This section creates Venn diagrams showing the overlap of significant TE changes between eIF3e and eIF3d at each timepoint, with Up and Down genes shown separately.

## Normoxia: eIF3e vs eIF3d Overlap

```{r venn_3e_vs_3d_normoxia}
# 1hr Up
venn_norm_1hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_norm_1hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_norm_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 1hr Down
venn_norm_1hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_norm_1hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_norm_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Up
venn_norm_4hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_norm_4hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_norm_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Down
venn_norm_4hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_norm_4hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_norm_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Up
venn_norm_24hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_norm_24hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_norm_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Down
venn_norm_24hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_norm_24hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_norm_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Combine into grid
venn_grid_3e_3d_norm <- (venn_norm_1hr_up | venn_norm_1hr_down) /
                        (venn_norm_4hr_up | venn_norm_4hr_down) /
                        (venn_norm_24hr_up | venn_norm_24hr_down) +
  plot_annotation(
    title = "eIF3e vs eIF3d TE changes: Normoxia",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11)
    )
  )

print(venn_grid_3e_3d_norm)
ggsave(here("plots/venn_te_3e_vs_3d_normoxia.pdf"), venn_grid_3e_3d_norm, width = 8, height = 10)
```

## Hypoxia: eIF3e vs eIF3d Overlap

```{r venn_3e_vs_3d_hypoxia}
# 1hr Up
venn_hyp_1hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_hyp_1hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 1hr Down
venn_hyp_1hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_hyp_1hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_hyp_1hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("1hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Up
venn_hyp_4hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_hyp_4hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 4hr Down
venn_hyp_4hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_hyp_4hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_hyp_4hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("4hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Up
venn_hyp_24hr_up <- ggvenn(
  list("eIF3e" = get_sig_te_up(res_te_si3e_hyp_24hr),
       "eIF3d" = get_sig_te_up(res_te_si3d_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Up") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# 24hr Down
venn_hyp_24hr_down <- ggvenn(
  list("eIF3e" = get_sig_te_down(res_te_si3e_hyp_24hr),
       "eIF3d" = get_sig_te_down(res_te_si3d_hyp_24hr)),
  fill_color = c("#56B4E9", "#E69F00"),
  stroke_size = 0.5, set_name_size = 4, text_size = 4) +
  ggtitle("24hr - Down") +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Combine into grid
venn_grid_3e_3d_hyp <- (venn_hyp_1hr_up | venn_hyp_1hr_down) /
                       (venn_hyp_4hr_up | venn_hyp_4hr_down) /
                       (venn_hyp_24hr_up | venn_hyp_24hr_down) +
  plot_annotation(
    title = "eIF3e vs eIF3d TE changes: Hypoxia",
    subtitle = "padj < 0.05, |LFC| > 0.5",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 11)
    )
  )

print(venn_grid_3e_3d_hyp)
ggsave(here("plots/venn_te_3e_vs_3d_hypoxia.pdf"), venn_grid_3e_3d_hyp, width = 8, height = 10)
```

# Density Plots: eIF3e vs eIF3d TE Changes in Hypoxia

This section creates density plots comparing the distribution of TE log2 fold changes between eIF3e and eIF3d knockdowns in hypoxia at each timepoint.

```{r density_te_3e_vs_3d_hypoxia}
# Create density plot for 1hr
density_hyp_1hr <- ggplot() +
  geom_density(data = res_te_si3e_hyp_1hr, aes(x = log2FoldChange, color = "eIF3e", fill = "eIF3e"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_1hr, aes(x = log2FoldChange, color = "eIF3d", fill = "eIF3d"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  scale_fill_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-1.6, 1.6)) +
  theme_prism() +
  theme(legend.position = "top") +
  labs(title = "1hr", x = "TE log2FC", y = "Density")

# Create density plot for 4hr
density_hyp_4hr <- ggplot() +
  geom_density(data = res_te_si3e_hyp_4hr, aes(x = log2FoldChange, color = "eIF3e", fill = "eIF3e"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_4hr, aes(x = log2FoldChange, color = "eIF3d", fill = "eIF3d"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  scale_fill_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-1.6, 1.6)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "4hr", x = "TE log2FC", y = "Density")

# Create density plot for 24hr
density_hyp_24hr <- ggplot() +
  geom_density(data = res_te_si3e_hyp_24hr, aes(x = log2FoldChange, color = "eIF3e", fill = "eIF3e"), size = 1, alpha = 0.2) +
  geom_density(data = res_te_si3d_hyp_24hr, aes(x = log2FoldChange, color = "eIF3d", fill = "eIF3d"), size = 1, alpha = 0.2) +
  scale_color_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  scale_fill_manual(values = c("eIF3e" = "#56B4E9", "eIF3d" = "#E69F00"), name = "") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  coord_cartesian(xlim = c(-1.6, 1.6)) +
  theme_prism() +
  theme(legend.position = "none") +
  labs(title = "24hr", x = "TE log2FC", y = "Density")
# Combine into column using patchwork
density_grid_3e_3d_hyp <- density_hyp_1hr / density_hyp_4hr / density_hyp_24hr +
  plot_annotation(
    title = "eIF3e vs eIF3d TE changes in Hypoxia",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  )

print(density_grid_3e_3d_hyp)
ggsave(here("plots/density_te_3e_vs_3d_hypoxia.pdf"), density_grid_3e_3d_hyp, width = 6, height = 9)
```

# GSEA: Pathway Analysis

This section performs Gene Set Enrichment Analysis (GSEA) on RNA and TE changes for eIF3e and eIF3d knockdowns in normoxia and hypoxia conditions.

## Load MSigDB Hallmark Gene Sets

```{r load_hallmark}
# Load MSigDB Hallmark gene sets for human
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# Create gene set format for clusterProfiler (TERM2GENE)
hallmark_rna <- hallmark_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded", length(unique(hallmark_rna$gs_name)), "Hallmark pathways\n")
```

# GSEA: RNA Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA RNA si3e Normoxia - 1hr

```{r gsea_rna_si3e_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_norm_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_1hr_results[] <- lapply(gsea_rna_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 4hr

```{r gsea_rna_si3e_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_norm_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_4hr_results[] <- lapply(gsea_rna_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 24hr

```{r gsea_rna_si3e_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_norm_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_24hr_results[] <- lapply(gsea_rna_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia RNA GSEA results

```{r export_gsea_rna_si3e_norm}
# Compile all si3e normoxia RNA GSEA results
gsea_rna_si3e_norm_export <- list(
  "GSEA_RNA_si3e_norm_1hr" = gsea_rna_si3e_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_norm_4hr" = gsea_rna_si3e_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_norm_24hr" = gsea_rna_si3e_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3e_norm_export, here("output/gsea_rna_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia RNA GSEA results exported to: output/gsea_rna_si3e_normoxia_hallmark.xlsx\n")
```

## si3e Normoxia RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3e_norm_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_norm_1hr_results) > 0) gsea_rna_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_4hr_results) > 0) gsea_rna_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_24hr_results) > 0) gsea_rna_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia RNA heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3e Normoxia RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3e Normoxia RNA heatmap\n")
}
```

# GSEA: RNA Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA RNA si3e Hypoxia - 1hr

```{r gsea_rna_si3e_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_hyp_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_1hr_results[] <- lapply(gsea_rna_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 4hr

```{r gsea_rna_si3e_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_hyp_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_4hr_results[] <- lapply(gsea_rna_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 24hr

```{r gsea_rna_si3e_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3e_hyp_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_24hr_results[] <- lapply(gsea_rna_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia RNA GSEA results

```{r export_gsea_rna_si3e_hyp}
# Compile all si3e hypoxia RNA GSEA results
gsea_rna_si3e_hyp_export <- list(
  "GSEA_RNA_si3e_hyp_1hr" = gsea_rna_si3e_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_hyp_4hr" = gsea_rna_si3e_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_hyp_24hr" = gsea_rna_si3e_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3e_hyp_export, here("output/gsea_rna_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia RNA GSEA results exported to: output/gsea_rna_si3e_hypoxia_hallmark.xlsx\n")
```

## si3e Hypoxia RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3e_hyp_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_hyp_1hr_results) > 0) gsea_rna_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_4hr_results) > 0) gsea_rna_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_24hr_results) > 0) gsea_rna_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia RNA heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3e Hypoxia RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3e Hypoxia RNA heatmap\n")
}
```

# GSEA: RNA Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA RNA si3d Normoxia - 1hr

```{r gsea_rna_si3d_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_norm_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_1hr_results[] <- lapply(gsea_rna_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 4hr

```{r gsea_rna_si3d_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_norm_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_4hr_results[] <- lapply(gsea_rna_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 24hr

```{r gsea_rna_si3d_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_norm_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_24hr_results[] <- lapply(gsea_rna_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia RNA GSEA results

```{r export_gsea_rna_si3d_norm}
# Compile all si3d normoxia RNA GSEA results
gsea_rna_si3d_norm_export <- list(
  "GSEA_RNA_si3d_norm_1hr" = gsea_rna_si3d_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_norm_4hr" = gsea_rna_si3d_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_norm_24hr" = gsea_rna_si3d_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3d_norm_export, here("output/gsea_rna_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia RNA GSEA results exported to: output/gsea_rna_si3d_normoxia_hallmark.xlsx\n")
```

## si3d Normoxia RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3d_norm_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_norm_1hr_results) > 0) gsea_rna_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_4hr_results) > 0) gsea_rna_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_24hr_results) > 0) gsea_rna_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia RNA heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3d Normoxia RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3d Normoxia RNA heatmap\n")
}
```

# GSEA: RNA Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA RNA si3d Hypoxia - 1hr

```{r gsea_rna_si3d_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_hyp_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_1hr_results[] <- lapply(gsea_rna_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 4hr

```{r gsea_rna_si3d_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_hyp_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_4hr_results[] <- lapply(gsea_rna_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 24hr

```{r gsea_rna_si3d_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
res_with_symbols <- res_rna_si3d_hyp_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_24hr_results[] <- lapply(gsea_rna_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia RNA GSEA results

```{r export_gsea_rna_si3d_hyp}
# Compile all si3d hypoxia RNA GSEA results
gsea_rna_si3d_hyp_export <- list(
  "GSEA_RNA_si3d_hyp_1hr" = gsea_rna_si3d_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_hyp_4hr" = gsea_rna_si3d_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_hyp_24hr" = gsea_rna_si3d_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3d_hyp_export, here("output/gsea_rna_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia RNA GSEA results exported to: output/gsea_rna_si3d_hypoxia_hallmark.xlsx\n")
```

## si3d Hypoxia RNA GSEA Heatmap: Pathway NES across timepoints

```{r gsea_rna_si3d_hyp_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_hyp_1hr_results) > 0) gsea_rna_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_4hr_results) > 0) gsea_rna_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_24hr_results) > 0) gsea_rna_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_rna_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia RNA heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3d Hypoxia RNA heatmap\n")
  }
} else {
  cat("No GSEA results available for si3d Hypoxia RNA heatmap\n")
}
```

# GSEA: TE Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on TE log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA TE si3e Normoxia - 1hr

```{r gsea_te_si3e_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_norm_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_1hr_results[] <- lapply(gsea_te_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 4hr

```{r gsea_te_si3e_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_norm_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_4hr_results[] <- lapply(gsea_te_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 24hr

```{r gsea_te_si3e_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_norm_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_24hr_results[] <- lapply(gsea_te_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia TE GSEA results

```{r export_gsea_te_si3e_norm}
# Compile all si3e normoxia TE GSEA results
gsea_te_si3e_norm_export <- list(
  "GSEA_TE_si3e_norm_1hr" = gsea_te_si3e_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_norm_4hr" = gsea_te_si3e_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_norm_24hr" = gsea_te_si3e_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3e_norm_export, here("output/gsea_te_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia TE GSEA results exported to: output/gsea_te_si3e_normoxia_hallmark.xlsx\n")
```

## si3e Normoxia TE GSEA Heatmap: Pathway NES across timepoints

```{r gsea_te_si3e_norm_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_norm_1hr_results) > 0) gsea_te_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_4hr_results) > 0) gsea_te_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_24hr_results) > 0) gsea_te_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Calculate actual NES range (symmetric around 0)
    nes_max <- max(abs(gsea_wide), na.rm = TRUE)
    nes_breaks <- seq(-nes_max, nes_max, length.out = 101)
    legend_vals <- round(seq(-nes_max, nes_max, length.out = 5), 2)

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = nes_breaks,
      legend_breaks = legend_vals,
      legend_labels = c(legend_vals[1:4], paste0(legend_vals[5], "\nNES")),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_te_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia TE heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3e Normoxia TE heatmap\n")
  }
} else {
  cat("No GSEA results available for si3e Normoxia TE heatmap\n")
}
```

# GSEA: TE Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on TE log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA TE si3e Hypoxia - 1hr

```{r gsea_te_si3e_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_hyp_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_1hr_results[] <- lapply(gsea_te_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 4hr

```{r gsea_te_si3e_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_hyp_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_4hr_results[] <- lapply(gsea_te_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 24hr

```{r gsea_te_si3e_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3e_hyp_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_24hr_results[] <- lapply(gsea_te_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia TE GSEA results

```{r export_gsea_te_si3e_hyp}
# Compile all si3e hypoxia TE GSEA results
gsea_te_si3e_hyp_export <- list(
  "GSEA_TE_si3e_hyp_1hr" = gsea_te_si3e_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_hyp_4hr" = gsea_te_si3e_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_hyp_24hr" = gsea_te_si3e_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3e_hyp_export, here("output/gsea_te_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia TE GSEA results exported to: output/gsea_te_si3e_hypoxia_hallmark.xlsx\n")
```

## si3e Hypoxia TE GSEA Heatmap: Pathway NES across timepoints

```{r gsea_te_si3e_hyp_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_hyp_1hr_results) > 0) gsea_te_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_4hr_results) > 0) gsea_te_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_24hr_results) > 0) gsea_te_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Calculate actual NES range (symmetric around 0)
    nes_max <- max(abs(gsea_wide), na.rm = TRUE)
    nes_breaks <- seq(-nes_max, nes_max, length.out = 101)
    legend_vals <- round(seq(-nes_max, nes_max, length.out = 5), 2)

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = nes_breaks,
      legend_breaks = legend_vals,
      legend_labels = c(legend_vals[1:4], paste0(legend_vals[5], "\nNES")),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_te_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia TE heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3e Hypoxia TE heatmap\n")
  }
} else {
  cat("No GSEA results available for si3e Hypoxia TE heatmap\n")
}
```

# GSEA: TE Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on TE log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA TE si3d Normoxia - 1hr

```{r gsea_te_si3d_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_norm_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_1hr_results[] <- lapply(gsea_te_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 4hr

```{r gsea_te_si3d_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_norm_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_4hr_results[] <- lapply(gsea_te_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 24hr

```{r gsea_te_si3d_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_norm_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_24hr_results[] <- lapply(gsea_te_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia TE GSEA results

```{r export_gsea_te_si3d_norm}
# Compile all si3d normoxia TE GSEA results
gsea_te_si3d_norm_export <- list(
  "GSEA_TE_si3d_norm_1hr" = gsea_te_si3d_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_norm_4hr" = gsea_te_si3d_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_norm_24hr" = gsea_te_si3d_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3d_norm_export, here("output/gsea_te_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia TE GSEA results exported to: output/gsea_te_si3d_normoxia_hallmark.xlsx\n")
```

## si3d Normoxia TE GSEA Heatmap: Pathway NES across timepoints

```{r gsea_te_si3d_norm_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_norm_1hr_results) > 0) gsea_te_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_4hr_results) > 0) gsea_te_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_24hr_results) > 0) gsea_te_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Calculate actual NES range (symmetric around 0)
    nes_max <- max(abs(gsea_wide), na.rm = TRUE)
    nes_breaks <- seq(-nes_max, nes_max, length.out = 101)
    legend_vals <- round(seq(-nes_max, nes_max, length.out = 5), 2)

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = nes_breaks,
      legend_breaks = legend_vals,
      legend_labels = c(legend_vals[1:4], paste0(legend_vals[5], "\nNES")),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_te_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia TE heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3d Normoxia TE heatmap\n")
  }
} else {
  cat("No GSEA results available for si3d Normoxia TE heatmap\n")
}
```

# GSEA: TE Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on TE log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA TE si3d Hypoxia - 1hr

```{r gsea_te_si3d_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_hyp_1hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_1hr_results[] <- lapply(gsea_te_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 4hr

```{r gsea_te_si3d_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_hyp_4hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_4hr_results[] <- lapply(gsea_te_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 24hr

```{r gsea_te_si3d_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
res_with_symbols <- res_te_si3d_hyp_24hr %>%
  left_join(gene_symbols, by = "gene_id")
ranked_list <- res_with_symbols$log2FoldChange
names(ranked_list) <- res_with_symbols$symbol

# Remove NA symbols and sort
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_24hr_results[] <- lapply(gsea_te_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia TE GSEA results

```{r export_gsea_te_si3d_hyp}
# Compile all si3d hypoxia TE GSEA results
gsea_te_si3d_hyp_export <- list(
  "GSEA_TE_si3d_hyp_1hr" = gsea_te_si3d_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_hyp_4hr" = gsea_te_si3d_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_hyp_24hr" = gsea_te_si3d_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3d_hyp_export, here("output/gsea_te_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia TE GSEA results exported to: output/gsea_te_si3d_hypoxia_hallmark.xlsx\n")
```

## si3d Hypoxia TE GSEA Heatmap: Pathway NES across timepoints

```{r gsea_te_si3d_hyp_heatmap}
# Combine results from all timepoints - only if they have data
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_hyp_1hr_results) > 0) gsea_te_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_4hr_results) > 0) gsea_te_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_24hr_results) > 0) gsea_te_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

# Check if we have any results at all
if (nrow(gsea_combined) > 0) {
  # Filter for pathways significant in at least one timepoint
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    # Reshape for heatmap
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    # Reorder columns
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    # Create significance annotation matrix
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist in sig matrix
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    # Reorder columns and ensure row order matches
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    # Calculate actual NES range (symmetric around 0)
    nes_max <- max(abs(gsea_wide), na.rm = TRUE)
    nes_breaks <- seq(-nes_max, nes_max, length.out = 101)
    legend_vals <- round(seq(-nes_max, nes_max, length.out = 5), 2)

    # Create heatmap
    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = nes_breaks,
      legend_breaks = legend_vals,
      legend_labels = c(legend_vals[1:4], paste0(legend_vals[5], "\nNES")),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    # Save
    pdf(here("plots/gsea_te_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia TE heatmap shows", nrow(gsea_wide), "significant pathways\n")
  } else {
    cat("No significant pathways found for si3d Hypoxia TE heatmap\n")
  }
} else {
  cat("No GSEA results available for si3d Hypoxia TE heatmap\n")
}
```

# Combined Visualizations

## sictrl vs si3d - All Conditions and Timepoints

```{r combined_si3d}
# Combine all si3d categorized tables with condition and timepoint labels
combined_si3d <- bind_rows(
  results_si3d_norm_1hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "1hr"),
  results_si3d_norm_4hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "4hr"),
  results_si3d_norm_24hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "24hr"),
  results_si3d_hyp_1hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "1hr"),
  results_si3d_hyp_4hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "4hr"),
  results_si3d_hyp_24hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "24hr")
) %>%
  # Add gene symbols
  left_join(gene_symbols, by = "gene_id")

# Set factor levels for proper ordering
combined_si3d$condition <- factor(combined_si3d$condition, levels = c("normoxia", "hypoxia"))
combined_si3d$timepoint <- factor(combined_si3d$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_si3d <- combined_si3d %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create faceted plot with TE count annotations
combined_si3d_plot <- ggplot(combined_si3d %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(condition ~ timepoint) +
  geom_text(data = te_counts_si3d,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("sictrl vs si3d - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_si3d_plot)

ggsave(here("plots/combined_si3d_all_conditions_timepoints.pdf"), width=9, height=5.5)
```

## sictrl vs si3e - All Conditions and Timepoints

```{r combined_si3e}
# Combine all si3e categorized tables with condition and timepoint labels
combined_si3e <- bind_rows(
  results_si3e_norm_1hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "1hr"),
  results_si3e_norm_4hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "4hr"),
  results_si3e_norm_24hr$categorized_table %>% mutate(condition = "normoxia", timepoint = "24hr"),
  results_si3e_hyp_1hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "1hr"),
  results_si3e_hyp_4hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "4hr"),
  results_si3e_hyp_24hr$categorized_table %>% mutate(condition = "hypoxia", timepoint = "24hr")
) %>%
  # Add gene symbols
  left_join(gene_symbols, by = "gene_id")

# Set factor levels for proper ordering
combined_si3e$condition <- factor(combined_si3e$condition, levels = c("normoxia", "hypoxia"))
combined_si3e$timepoint <- factor(combined_si3e$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_si3e <- combined_si3e %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create faceted plot with TE count annotations
combined_si3e_plot <- ggplot(combined_si3e %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(condition ~ timepoint) +
  geom_text(data = te_counts_si3e,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("sictrl vs si3e - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_si3e_plot)

ggsave(here("plots/combined_si3e_all_conditions_timepoints.pdf"), width=9, height=5.5)
```

## Combined si3d and si3e Visualization

```{r combined_si3d_si3e}
# Combine si3d and si3e data with treatment labels
combined_all <- bind_rows(
  combined_si3d %>% mutate(treatment = "si3d"),
  combined_si3e %>% mutate(treatment = "si3e")
)

# Set factor levels for proper ordering
combined_all$treatment <- factor(combined_all$treatment, levels = c("si3d", "si3e"))
combined_all$condition <- factor(combined_all$condition, levels = c("normoxia", "hypoxia"))
combined_all$timepoint <- factor(combined_all$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations (including treatment)
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts_combined <- combined_all %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(treatment, condition, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")

# Create combined faceted plot
# 4 rows (2 treatments × 2 conditions) × 3 columns (timepoints) = 12 panels
combined_all_plot <- ggplot(combined_all %>% filter(category != "Not_sig"),
                            aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(treatment + condition ~ timepoint) +
  geom_text(data = te_counts_combined,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  ggtitle("si3d and sictrl vs si3e - All Conditions and Timepoints") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_all_plot)

ggsave(here("plots/combined_si3d_si3e_all_conditions_timepoints.pdf"), width=10, height=8)
```

## Highlight Gene of Interest Across All Conditions

```{r highlight_gene_of_interest}
# Define gene of interest
gene_of_interest <- "TEAD1"

# Check if gene was found in data
gene_found <- combined_all %>%
  filter(symbol == gene_of_interest, category != "Not_sig")

if(nrow(gene_found) > 0) {
  cat("Found", gene_of_interest, "in", nrow(gene_found), "condition(s)\n")

  # Define colors (same as combined_all_plot)
  myCols <- c("Not_sig" = "grey80",
              "Forwarded" = "blue",
              "Exclusive" = "red",
              "Buffered" = "green",
              "TE_only" = "purple")

  # Create plot identical to combined_all_plot with gene label added
  highlighted_plot <- ggplot(combined_all %>% filter(category != "Not_sig"),
                             aes(x = rna_lfc, y = ribo_lfc, color = category)) +
    # ALL genes (including TEAD1) colored by category - same as combined_all_plot
    geom_point(alpha = 0.6, size = 0.8) +
    scale_color_manual(values = myCols) +
    xlim(-5, 5) + ylim(-5, 5) +
    facet_grid(treatment + condition ~ timepoint) +
    geom_text(data = te_counts_combined,
              aes(x = -4, y = 4, label = label),
              inherit.aes = FALSE,
              size = 3,
              hjust = 0,
              vjust = 1,
              color = "black",
              fontface = "bold") +
    # ADD ONLY THIS: Label for gene of interest
    geom_text_repel(data = gene_found,
                    aes(label = symbol),
                    size = 3.5,
                    fontface = "bold",
                    box.padding = 0.5,
                    point.padding = 0.3,
                    segment.size = 0.5,
                    max.overlaps = 20,
                    show.legend = FALSE) +
    xlab("RNA-seq log2FC") +
    ylab("Ribo-seq log2FC") +
    ggtitle(paste0("Highlighting ", gene_of_interest, " - si3d and sictrl vs si3e")) +
    theme_minimal(base_size = 14) +
    theme(legend.title = element_blank(),
          legend.position = "bottom",
          strip.text = element_text(face = "bold"))

  print(highlighted_plot)

  ggsave(here("plots", paste0("highlighted_", gene_of_interest, "_all_conditions.pdf")),
         width=10, height=8)

  # Print summary table for gene of interest
  cat("\nSummary for", gene_of_interest, ":\n")
  gene_summary <- gene_found %>%
    dplyr::select(treatment, condition, timepoint, rna_lfc, ribo_lfc, te_lfc, category) %>%
    arrange(treatment, condition, timepoint)

  print(gene_summary)

} else {
  cat("Warning: Gene", gene_of_interest, "not found in the dataset\n")
}
```

```{r}

```