---
title: "eIF3d-Promoting Gene Set Overlap Across Cell Lines"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Overview

This analysis identifies genes whose translation is promoted by eIF3d (TE LFC > 0.5, padj < 0.05) across three independent datasets and visualizes the overlap between gene sets.

**Datasets:**
- **MCF7-SIX1 (no spike-in)**: 2024 eIF3e/3d hypoxia project
- **MDA-MB-231 (no spike-in)**: 2024 eIF3e/3d hypoxia project
- **MDA-MB-231 (spike-in)**: Current project, spike-in normalized, 1hr timepoint

**Directionality convention (standardized to positive = eIF3d promotes):**
- Datasets 1 & 2: original LFC = log2(si3d / sictrl); `te_lfc_standardized = -log2FoldChange`
- Dataset 3: LFC = log2(sictrl / si3d); positive already means eIF3d promotes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 6)
```

```{r libraries}
library(tidyverse)
library(here)
library(ggvenn)
library(GeneOverlap)
library(knitr)
```

---

# Load Data

## Datasets 1 & 2: 2024 eIF3e/3d Project

These files have columns: `gene_id, symbol, baseMean, log2FoldChange, lfcSE, pvalue, padj`

The DESeq2 design used `sictrl` as reference, so `log2FoldChange` = log2(si3d / sictrl).
Positive = eIF3d suppresses; **negative = eIF3d promotes**.
We negate to standardize: `te_lfc = -log2FoldChange` so positive = eIF3d promotes.

```{r load-2024-data}
# MCF7-SIX1 no spike-in normoxia
mcf7six1_nospike_norm_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/normoxia_sictrl_3d.csv"
) %>%
  mutate(
    dataset = "MCF7-SIX1 (no spike-in)",
    condition = "normoxia",
    te_lfc = -log2FoldChange,
    te_padj = padj
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

# MCF7-SIX1 no spike-in hypoxia
mcf7six1_nospike_hyp_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/hypoxia_si_ctrl_3d.csv"
) %>%
  mutate(
    dataset = "MCF7-SIX1 (no spike-in)",
    condition = "hypoxia",
    te_lfc = -log2FoldChange,
    te_padj = padj
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

# MDA-MB-231 no spike-in normoxia
mda231_nospike_norm_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/counts_231/normoxia_si_ctrl_3d.csv"
) %>%
  mutate(
    dataset = "MDA-MB-231 (no spike-in)",
    condition = "normoxia",
    te_lfc = -log2FoldChange,
    te_padj = padj
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

# MDA-MB-231 no spike-in hypoxia
mda231_nospike_hyp_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/counts_231/hypoxia_si_ctrl_3d.csv"
) %>%
  mutate(
    dataset = "MDA-MB-231 (no spike-in)",
    condition = "hypoxia",
    te_lfc = -log2FoldChange,
    te_padj = padj
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

cat("2024 project data loaded:\n")
cat("  MCF7-SIX1 no spike-in normoxia:", nrow(mcf7six1_nospike_norm_raw), "genes\n")
cat("  MCF7-SIX1 no spike-in hypoxia:", nrow(mcf7six1_nospike_hyp_raw), "genes\n")
cat("  MDA-MB-231 no spike-in normoxia:", nrow(mda231_nospike_norm_raw), "genes\n")
cat("  MDA-MB-231 no spike-in hypoxia:", nrow(mda231_nospike_hyp_raw), "genes\n")
```

## Dataset 3: Current Project (MDA-MB-231 Spike-in Normalized, 1hr)

These files have columns: `gene_id, rna_lfc, ribo_lfc, te_lfc, category, rna_padj, ribo_padj, te_padj, symbol`

LFC = log2(sictrl / si3d); positive te_lfc already means eIF3d promotes.

```{r load-spikein-data}
# MDA-MB-231 spike-in normoxia 1hr
mda231_spike_norm_raw <- read_csv(
  here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")
) %>%
  mutate(
    dataset = "MDA-MB-231 (spike-in)",
    condition = "normoxia"
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

# MDA-MB-231 spike-in hypoxia 1hr
mda231_spike_hyp_raw <- read_csv(
  here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")
) %>%
  mutate(
    dataset = "MDA-MB-231 (spike-in)",
    condition = "hypoxia"
  ) %>%
  dplyr::select(symbol, te_lfc, te_padj, dataset, condition)

cat("Spike-in project data loaded:\n")
cat("  MDA-MB-231 spike-in normoxia 1hr:", nrow(mda231_spike_norm_raw), "genes\n")
cat("  MDA-MB-231 spike-in hypoxia 1hr:", nrow(mda231_spike_hyp_raw), "genes\n")
```

---

# Define Gene Sets

Filter to genes with `te_lfc > 0.5` and `te_padj < 0.05` (eIF3d promotes translation).

```{r define-gene-sets}
# Helper function to extract promoted gene symbols
get_promoted_genes <- function(df) {
  df %>%
    filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>%
    dplyr::select(symbol) %>%
    distinct() %>%
    pull(symbol)
}

promoted_mcf7six1_nospike_norm <- get_promoted_genes(mcf7six1_nospike_norm_raw)
promoted_mcf7six1_nospike_hyp  <- get_promoted_genes(mcf7six1_nospike_hyp_raw)
promoted_231_nospike_norm      <- get_promoted_genes(mda231_nospike_norm_raw)
promoted_231_nospike_hyp       <- get_promoted_genes(mda231_nospike_hyp_raw)
promoted_231_spike_norm        <- get_promoted_genes(mda231_spike_norm_raw)
promoted_231_spike_hyp         <- get_promoted_genes(mda231_spike_hyp_raw)

cat("Gene counts (te_lfc > 0.5, padj < 0.05):\n")
cat("  MCF7-SIX1 no spike-in normoxia:", length(promoted_mcf7six1_nospike_norm), "\n")
cat("  MCF7-SIX1 no spike-in hypoxia:", length(promoted_mcf7six1_nospike_hyp), "\n")
cat("  MDA-MB-231 no spike-in normoxia:", length(promoted_231_nospike_norm), "\n")
cat("  MDA-MB-231 no spike-in hypoxia:", length(promoted_231_nospike_hyp), "\n")
cat("  MDA-MB-231 spike-in normoxia:", length(promoted_231_spike_norm), "\n")
cat("  MDA-MB-231 spike-in hypoxia:", length(promoted_231_spike_hyp), "\n")
```

---

# Normoxia Overlap

## Venn Diagram

```{r venn-normoxia, fig.width=7, fig.height=6}
venn_list_norm <- list(
  "MCF7-SIX1\n(no spike)" = promoted_mcf7six1_nospike_norm,
  "231\n(no spike)"        = promoted_231_nospike_norm,
  "231\n(spike)"           = promoted_231_spike_norm
)

venn_norm <- ggvenn(venn_list_norm,
                    fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                    stroke_size = 0.5,
                    set_name_size = 4,
                    text_size = 4) +
  ggtitle("eIF3d-Promoted Genes: Normoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_norm)
ggsave(here("plots/venn_eif3d_promoted_normoxia_celllines.pdf"), venn_norm,
       width = 7, height = 6)
```

## GeneOverlap Statistical Tests

```{r geneoverlap-normoxia}
genome_size <- 20000

# MCF7-SIX1 (no spike) vs MDA-MB-231 (no spike)
go_norm_mcf7six1_231nospike <- newGeneOverlap(promoted_mcf7six1_nospike_norm,
                                              promoted_231_nospike_norm,
                                              genome.size = genome_size)
go_norm_mcf7six1_231nospike <- testGeneOverlap(go_norm_mcf7six1_231nospike)

# MCF7-SIX1 (no spike) vs MDA-MB-231 (spike)
go_norm_mcf7six1_231spike <- newGeneOverlap(promoted_mcf7six1_nospike_norm,
                                            promoted_231_spike_norm,
                                            genome.size = genome_size)
go_norm_mcf7six1_231spike <- testGeneOverlap(go_norm_mcf7six1_231spike)

# MDA-MB-231 (no spike) vs MDA-MB-231 (spike)
go_norm_231nospike_231spike <- newGeneOverlap(promoted_231_nospike_norm,
                                              promoted_231_spike_norm,
                                              genome.size = genome_size)
go_norm_231nospike_231spike <- testGeneOverlap(go_norm_231nospike_231spike)

cat("=== Normoxia GeneOverlap Results ===\n\n")

cat("MCF7-SIX1 (no spike) vs MDA-MB-231 (no spike):\n")
cat("  Overlap:", length(getIntersection(go_norm_mcf7six1_231nospike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_norm_mcf7six1_231nospike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_norm_mcf7six1_231nospike), 3), "\n\n")

cat("MCF7-SIX1 (no spike) vs MDA-MB-231 (spike):\n")
cat("  Overlap:", length(getIntersection(go_norm_mcf7six1_231spike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_norm_mcf7six1_231spike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_norm_mcf7six1_231spike), 3), "\n\n")

cat("MDA-MB-231 (no spike) vs MDA-MB-231 (spike):\n")
cat("  Overlap:", length(getIntersection(go_norm_231nospike_231spike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_norm_231nospike_231spike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_norm_231nospike_231spike), 3), "\n")
```

## Intersection Gene Tables

```{r intersection-tables-normoxia}
# All three datasets
norm_all3 <- Reduce(intersect, list(promoted_mcf7six1_nospike_norm,
                                    promoted_231_nospike_norm,
                                    promoted_231_spike_norm))

# Pairwise intersections (excluding all-three)
norm_mcf7six1_231nospike_only <- setdiff(
  intersect(promoted_mcf7six1_nospike_norm, promoted_231_nospike_norm),
  promoted_231_spike_norm
)
norm_mcf7six1_231spike_only <- setdiff(
  intersect(promoted_mcf7six1_nospike_norm, promoted_231_spike_norm),
  promoted_231_nospike_norm
)
norm_231nospike_231spike_only <- setdiff(
  intersect(promoted_231_nospike_norm, promoted_231_spike_norm),
  promoted_mcf7six1_nospike_norm
)

cat("Normoxia — All three datasets (", length(norm_all3), "genes):\n")
if (length(norm_all3) > 0) cat(paste(sort(norm_all3), collapse = ", "), "\n\n")

cat("Normoxia — MCF7-SIX1 (no spike) & 231 (no spike) only (", length(norm_mcf7six1_231nospike_only), "genes):\n")
if (length(norm_mcf7six1_231nospike_only) > 0) cat(paste(sort(norm_mcf7six1_231nospike_only), collapse = ", "), "\n\n")

cat("Normoxia — MCF7-SIX1 (no spike) & 231 (spike) only (", length(norm_mcf7six1_231spike_only), "genes):\n")
if (length(norm_mcf7six1_231spike_only) > 0) cat(paste(sort(norm_mcf7six1_231spike_only), collapse = ", "), "\n\n")

cat("Normoxia — 231 (no spike) & 231 (spike) only (", length(norm_231nospike_231spike_only), "genes):\n")
if (length(norm_231nospike_231spike_only) > 0) cat(paste(sort(norm_231nospike_231spike_only), collapse = ", "), "\n\n")
```

```{r norm-all3-table}
if (length(norm_all3) > 0) {
  norm_all3_df <- tibble(symbol = norm_all3) %>%
    left_join(
      mcf7six1_nospike_norm_raw %>%
        dplyr::select(symbol, mcf7six1_nospike_norm_lfc = te_lfc, mcf7six1_nospike_norm_padj = te_padj),
      by = "symbol"
    ) %>%
    left_join(
      mda231_nospike_norm_raw %>%
        dplyr::select(symbol, mda231_nospike_norm_lfc = te_lfc, mda231_nospike_norm_padj = te_padj),
      by = "symbol"
    ) %>%
    left_join(
      mda231_spike_norm_raw %>%
        dplyr::select(symbol, mda231_spike_norm_lfc = te_lfc, mda231_spike_norm_padj = te_padj),
      by = "symbol"
    ) %>%
    arrange(symbol)

  kable(norm_all3_df, digits = 3,
        caption = "Genes promoted by eIF3d in normoxia across all three datasets")
}
```

---

# Hypoxia Overlap

## Venn Diagram

```{r venn-hypoxia, fig.width=7, fig.height=6}
venn_list_hyp <- list(
  "MCF7-SIX1\n(no spike)" = promoted_mcf7six1_nospike_hyp,
  "231\n(no spike)"        = promoted_231_nospike_hyp,
  "231\n(spike)"           = promoted_231_spike_hyp
)

venn_hyp <- ggvenn(venn_list_hyp,
                   fill_color = c("#E69F00", "#56B4E9", "#009E73"),
                   stroke_size = 0.5,
                   set_name_size = 4,
                   text_size = 4) +
  ggtitle("eIF3d-Promoted Genes: Hypoxia") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(venn_hyp)
ggsave(here("plots/venn_eif3d_promoted_hypoxia_celllines.pdf"), venn_hyp,
       width = 7, height = 6)
```

## GeneOverlap Statistical Tests

```{r geneoverlap-hypoxia}
# MCF7-SIX1 (no spike) vs MDA-MB-231 (no spike)
go_hyp_mcf7six1_231nospike <- newGeneOverlap(promoted_mcf7six1_nospike_hyp,
                                             promoted_231_nospike_hyp,
                                             genome.size = genome_size)
go_hyp_mcf7six1_231nospike <- testGeneOverlap(go_hyp_mcf7six1_231nospike)

# MCF7-SIX1 (no spike) vs MDA-MB-231 (spike)
go_hyp_mcf7six1_231spike <- newGeneOverlap(promoted_mcf7six1_nospike_hyp,
                                           promoted_231_spike_hyp,
                                           genome.size = genome_size)
go_hyp_mcf7six1_231spike <- testGeneOverlap(go_hyp_mcf7six1_231spike)

# MDA-MB-231 (no spike) vs MDA-MB-231 (spike)
go_hyp_231nospike_231spike <- newGeneOverlap(promoted_231_nospike_hyp,
                                             promoted_231_spike_hyp,
                                             genome.size = genome_size)
go_hyp_231nospike_231spike <- testGeneOverlap(go_hyp_231nospike_231spike)

cat("=== Hypoxia GeneOverlap Results ===\n\n")

cat("MCF7-SIX1 (no spike) vs MDA-MB-231 (no spike):\n")
cat("  Overlap:", length(getIntersection(go_hyp_mcf7six1_231nospike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_hyp_mcf7six1_231nospike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_hyp_mcf7six1_231nospike), 3), "\n\n")

cat("MCF7-SIX1 (no spike) vs MDA-MB-231 (spike):\n")
cat("  Overlap:", length(getIntersection(go_hyp_mcf7six1_231spike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_hyp_mcf7six1_231spike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_hyp_mcf7six1_231spike), 3), "\n\n")

cat("MDA-MB-231 (no spike) vs MDA-MB-231 (spike):\n")
cat("  Overlap:", length(getIntersection(go_hyp_231nospike_231spike)), "genes\n")
cat("  Fisher's p-value:", format(getPval(go_hyp_231nospike_231spike), scientific = TRUE), "\n")
cat("  Jaccard index:", round(getJaccard(go_hyp_231nospike_231spike), 3), "\n")
```

## Intersection Gene Tables

```{r intersection-tables-hypoxia}
# All three datasets
hyp_all3 <- Reduce(intersect, list(promoted_mcf7six1_nospike_hyp,
                                   promoted_231_nospike_hyp,
                                   promoted_231_spike_hyp))

# Pairwise intersections (excluding all-three)
hyp_mcf7six1_231nospike_only <- setdiff(
  intersect(promoted_mcf7six1_nospike_hyp, promoted_231_nospike_hyp),
  promoted_231_spike_hyp
)
hyp_mcf7six1_231spike_only <- setdiff(
  intersect(promoted_mcf7six1_nospike_hyp, promoted_231_spike_hyp),
  promoted_231_nospike_hyp
)
hyp_231nospike_231spike_only <- setdiff(
  intersect(promoted_231_nospike_hyp, promoted_231_spike_hyp),
  promoted_mcf7six1_nospike_hyp
)

cat("Hypoxia — All three datasets (", length(hyp_all3), "genes):\n")
if (length(hyp_all3) > 0) cat(paste(sort(hyp_all3), collapse = ", "), "\n\n")

cat("Hypoxia — MCF7-SIX1 (no spike) & 231 (no spike) only (", length(hyp_mcf7six1_231nospike_only), "genes):\n")
if (length(hyp_mcf7six1_231nospike_only) > 0) cat(paste(sort(hyp_mcf7six1_231nospike_only), collapse = ", "), "\n\n")

cat("Hypoxia — MCF7-SIX1 (no spike) & 231 (spike) only (", length(hyp_mcf7six1_231spike_only), "genes):\n")
if (length(hyp_mcf7six1_231spike_only) > 0) cat(paste(sort(hyp_mcf7six1_231spike_only), collapse = ", "), "\n\n")

cat("Hypoxia — 231 (no spike) & 231 (spike) only (", length(hyp_231nospike_231spike_only), "genes):\n")
if (length(hyp_231nospike_231spike_only) > 0) cat(paste(sort(hyp_231nospike_231spike_only), collapse = ", "), "\n\n")
```

```{r hyp-all3-table}
if (length(hyp_all3) > 0) {
  hyp_all3_df <- tibble(symbol = hyp_all3) %>%
    left_join(
      mcf7six1_nospike_hyp_raw %>%
        dplyr::select(symbol, mcf7six1_nospike_hyp_lfc = te_lfc, mcf7six1_nospike_hyp_padj = te_padj),
      by = "symbol"
    ) %>%
    left_join(
      mda231_nospike_hyp_raw %>%
        dplyr::select(symbol, mda231_nospike_hyp_lfc = te_lfc, mda231_nospike_hyp_padj = te_padj),
      by = "symbol"
    ) %>%
    left_join(
      mda231_spike_hyp_raw %>%
        dplyr::select(symbol, mda231_spike_hyp_lfc = te_lfc, mda231_spike_hyp_padj = te_padj),
      by = "symbol"
    ) %>%
    arrange(symbol)

  kable(hyp_all3_df, digits = 3,
        caption = "Genes promoted by eIF3d in hypoxia across all three datasets")
}
```

---

# Summary Table

```{r summary-table}
summary_df <- tribble(
  ~Comparison, ~Condition, ~Set1, ~N1, ~Set2, ~N2, ~Overlap, ~Pvalue, ~Jaccard,
  "MCF7-SIX1 (no spike) vs 231 (no spike)", "Normoxia",
    "MCF7-SIX1 (no spike)", length(promoted_mcf7six1_nospike_norm),
    "231 (no spike)",        length(promoted_231_nospike_norm),
    length(getIntersection(go_norm_mcf7six1_231nospike)),
    format(getPval(go_norm_mcf7six1_231nospike), scientific = TRUE),
    round(getJaccard(go_norm_mcf7six1_231nospike), 3),
  "MCF7-SIX1 (no spike) vs 231 (spike)",   "Normoxia",
    "MCF7-SIX1 (no spike)", length(promoted_mcf7six1_nospike_norm),
    "231 (spike)",           length(promoted_231_spike_norm),
    length(getIntersection(go_norm_mcf7six1_231spike)),
    format(getPval(go_norm_mcf7six1_231spike), scientific = TRUE),
    round(getJaccard(go_norm_mcf7six1_231spike), 3),
  "231 (no spike) vs 231 (spike)",          "Normoxia",
    "231 (no spike)",        length(promoted_231_nospike_norm),
    "231 (spike)",           length(promoted_231_spike_norm),
    length(getIntersection(go_norm_231nospike_231spike)),
    format(getPval(go_norm_231nospike_231spike), scientific = TRUE),
    round(getJaccard(go_norm_231nospike_231spike), 3),
  "MCF7-SIX1 (no spike) vs 231 (no spike)", "Hypoxia",
    "MCF7-SIX1 (no spike)", length(promoted_mcf7six1_nospike_hyp),
    "231 (no spike)",        length(promoted_231_nospike_hyp),
    length(getIntersection(go_hyp_mcf7six1_231nospike)),
    format(getPval(go_hyp_mcf7six1_231nospike), scientific = TRUE),
    round(getJaccard(go_hyp_mcf7six1_231nospike), 3),
  "MCF7-SIX1 (no spike) vs 231 (spike)",   "Hypoxia",
    "MCF7-SIX1 (no spike)", length(promoted_mcf7six1_nospike_hyp),
    "231 (spike)",           length(promoted_231_spike_hyp),
    length(getIntersection(go_hyp_mcf7six1_231spike)),
    format(getPval(go_hyp_mcf7six1_231spike), scientific = TRUE),
    round(getJaccard(go_hyp_mcf7six1_231spike), 3),
  "231 (no spike) vs 231 (spike)",          "Hypoxia",
    "231 (no spike)",        length(promoted_231_nospike_hyp),
    "231 (spike)",           length(promoted_231_spike_hyp),
    length(getIntersection(go_hyp_231nospike_231spike)),
    format(getPval(go_hyp_231nospike_231spike), scientific = TRUE),
    round(getJaccard(go_hyp_231nospike_231spike), 3)
)

kable(summary_df,
      col.names = c("Comparison", "Condition", "Set 1", "N1", "Set 2", "N2",
                    "Overlap", "Fisher p-value", "Jaccard"),
      caption = "Summary of eIF3d-promoted gene set overlaps across cell lines (te_lfc > 0.5, padj < 0.05)")
```

```{r all3-summary}
cat("Genes promoted by eIF3d in ALL THREE datasets:\n")
cat("  Normoxia:", length(norm_all3), "genes\n")
cat("  Hypoxia:", length(hyp_all3), "genes\n")

both_conditions_all3 <- intersect(norm_all3, hyp_all3)
cat("  Both normoxia and hypoxia (all three):", length(both_conditions_all3), "genes\n")
if (length(both_conditions_all3) > 0) {
  cat("  Genes:", paste(sort(both_conditions_all3), collapse = ", "), "\n")
}
```

---

# Teleman RiboSeq Distribution

```{r load-teleman}
res_riboseq_symbol <- read_csv(
  "/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv"
)

# Deduplicate: keep highest baseMean per symbol (matches teleman_comparison.Rmd pattern)
teleman_riboseq <- res_riboseq_symbol %>%
  filter(!is.na(symbol), !is.na(log2FoldChange)) %>%
  group_by(symbol) %>%
  slice_max(baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Teleman RiboSeq genes (deduplicated):", nrow(teleman_riboseq), "\n")
```

```{r teleman-distribution-plots}
plot_teleman_dist <- function(teleman_df, genes, label, color, title, filename) {
  n <- sum(teleman_df$symbol %in% genes)
  gene_label <- paste0(label, " (n=", n, ")")

  plot_df <- teleman_df %>%
    mutate(group = if_else(symbol %in% genes, gene_label, "Other"))

  p <- ggplot() +
    geom_density(data = plot_df, aes(x = log2FoldChange), fill = "grey80", color = "grey60") +
    geom_density(data = filter(plot_df, group == gene_label),
                 aes(x = log2FoldChange, fill = group, color = group), alpha = 0.4) +
    scale_fill_manual(values  = setNames(color, gene_label)) +
    scale_color_manual(values = setNames(color, gene_label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "Teleman RiboSeq log2FC (4E inhibitor vs vehicle)",
         y = "Density", fill = NULL, color = NULL, title = title) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top", plot.title = element_text(hjust = 0.5))

  print(p)
  ggsave(here(filename), p, width = 7, height = 5)
  invisible(p)
}

# MDA-MB-231 spike-in
plot_teleman_dist(teleman_riboseq, promoted_231_spike_hyp,  "Hypoxia-promoted",  "#2a9d8f",
                  "MDA-MB-231 (spike-in) \u2014 eIF3d-promoted in hypoxia",
                  "plots/teleman_dist_231_spike_hyp.pdf")
plot_teleman_dist(teleman_riboseq, promoted_231_spike_norm, "Normoxia-promoted", "#e76f51",
                  "MDA-MB-231 (spike-in) \u2014 eIF3d-promoted in normoxia",
                  "plots/teleman_dist_231_spike_norm.pdf")

# MDA-MB-231 no spike-in
plot_teleman_dist(teleman_riboseq, promoted_231_nospike_hyp,  "Hypoxia-promoted",  "#2a9d8f",
                  "MDA-MB-231 (no spike-in) \u2014 eIF3d-promoted in hypoxia",
                  "plots/teleman_dist_231_nospike_hyp.pdf")
plot_teleman_dist(teleman_riboseq, promoted_231_nospike_norm, "Normoxia-promoted", "#e76f51",
                  "MDA-MB-231 (no spike-in) \u2014 eIF3d-promoted in normoxia",
                  "plots/teleman_dist_231_nospike_norm.pdf")

# MCF7-SIX1 no spike-in
plot_teleman_dist(teleman_riboseq, promoted_mcf7six1_nospike_hyp,  "Hypoxia-promoted",  "#2a9d8f",
                  "MCF7-SIX1 (no spike-in) \u2014 eIF3d-promoted in hypoxia",
                  "plots/teleman_dist_mcf7six1_nospike_hyp.pdf")
plot_teleman_dist(teleman_riboseq, promoted_mcf7six1_nospike_norm, "Normoxia-promoted", "#e76f51",
                  "MCF7-SIX1 (no spike-in) \u2014 eIF3d-promoted in normoxia",
                  "plots/teleman_dist_mcf7six1_nospike_norm.pdf")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
