---
title: "eIF3d and eIF3e Effects on Hypoxic Response"
author: "Kate"
date: "2026-01-12"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis compares the hypoxic response (hypoxia vs normoxia at 1 hour) across three treatments:

- **siCtrl**: Control siRNA
- **si3d**: eIF3d knockdown
- **si3e**: eIF3e knockdown

We visualize the distribution of log2 fold changes for:

- **RNA**: mRNA expression changes
- **Ribo**: Ribosome-protected fragment (RPF) changes
- **TE**: Translation efficiency changes

The goal is to understand how eIF3d and eIF3e knockdowns affect the cellular response to hypoxia.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggplot2)
library(ggpubr)
library(ggokabeito)
library(DT)
```

```{r source_functions}
source(here("code/functions.R"))
cat("Loaded functions from functions.R\n")
```

# Load and Prepare Data

Load the translation categories files for all three treatments. These files contain DESeq2 results comparing hypoxia vs normoxia at 1 hour.

```{r load_data}
# Load siCtrl hypoxia vs normoxia (1hr)
lfc_data_sictrl <- read_csv(here("output", "translation_categories_1hr.csv")) %>%
  select(gene_id, rna_lfc, ribo_lfc, te_lfc) %>%
  mutate(Treatment = "siCtrl")

# Load si3d hypoxia vs normoxia (1hr)
lfc_data_si3d <- read_csv(here("output", "translation_categories_si3d_1hr.csv")) %>%
  select(gene_id, rna_lfc, ribo_lfc, te_lfc) %>%
  mutate(Treatment = "si3d")

# Load si3e hypoxia vs normoxia (1hr)
lfc_data_si3e <- read_csv(here("output", "translation_categories_si3e_1hr.csv")) %>%
  select(gene_id, rna_lfc, ribo_lfc, te_lfc) %>%
  mutate(Treatment = "si3e")

cat("Loaded data for:\n")
cat("  siCtrl:", nrow(lfc_data_sictrl), "genes\n")
cat("  si3d:", nrow(lfc_data_si3d), "genes\n")
cat("  si3e:", nrow(lfc_data_si3e), "genes\n")
```

```{r combine_data}
# Combine all treatments
combined_data <- bind_rows(lfc_data_sictrl, lfc_data_si3d, lfc_data_si3e) %>%
  mutate(Treatment = factor(Treatment, levels = c("siCtrl", "si3d", "si3e")))

cat("\nCombined data:", nrow(combined_data), "total rows\n")
cat("Treatments:", levels(combined_data$Treatment), "\n")
```

```{r transform_to_long}
# Transform to long format for faceted plotting
combined_long <- combined_data %>%
  pivot_longer(
    cols = c(rna_lfc, ribo_lfc, te_lfc),
    names_to = "Measurement",
    values_to = "log2FC"
  ) %>%
  mutate(
    Measurement = factor(
      Measurement,
      levels = c("rna_lfc", "ribo_lfc", "te_lfc"),
      labels = c("RNA", "Ribo", "TE")
    )
  )

cat("\nLong format data:", nrow(combined_long), "rows\n")
cat("Measurements:", levels(combined_long$Measurement), "\n")
```

# Summary Statistics

Calculate summary statistics for each treatment and measurement combination.

```{r summary_statistics}
summary_stats <- combined_long %>%
  group_by(Treatment, Measurement) %>%
  summarise(
    n_genes = n(),
    mean_lfc = mean(log2FC, na.rm = TRUE),
    median_lfc = median(log2FC, na.rm = TRUE),
    sd_lfc = sd(log2FC, na.rm = TRUE),
    q25 = quantile(log2FC, 0.25, na.rm = TRUE),
    q75 = quantile(log2FC, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Display as formatted table
DT::datatable(
  summary_stats,
  options = list(pageLength = 9),
  caption = "Summary Statistics: Hypoxic Response by Treatment and Measurement"
) %>%
  DT::formatRound(columns = c("mean_lfc", "median_lfc", "sd_lfc", "q25", "q75"), digits = 3)
```

```{r save_summary_stats}
# Save to CSV
write_csv(summary_stats, here("output", "hypoxic_response_summary_stats_1hr.csv"))
cat("Saved summary statistics to output/hypoxic_response_summary_stats_1hr.csv\n")
```

# Density Plots: Compare siCtrl vs si3d

These plots compare the hypoxic response between siCtrl (grey) and si3d (orange) for each measurement type.

## RNA Expression Changes: siCtrl vs si3d

```{r density_rna_si3d, fig.width=8, fig.height=5}
plot_rna_si3d <- combined_long %>%
  filter(Measurement == "RNA", Treatment %in% c("siCtrl", "si3d")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3d" = "#D55E00"     # Orange
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-.2, .2)) +
  labs(
    title = "RNA Expression Changes: siCtrl vs si3d",
    subtitle = "Distribution of RNA log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "RNA log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )


print(plot_rna_si3d)

ggsave(
  here("plots", "hypoxic_response_rna_sictrl_vs_si3d_1hr.pdf"),
  plot = plot_rna_si3d,
  width = 5,
  height = 3,
  device = "pdf"
)
```

## Ribosome Occupancy Changes: siCtrl vs si3d

```{r density_ribo_si3d, fig.width=8, fig.height=5}
plot_ribo_si3d <- combined_long %>%
  filter(Measurement == "Ribo", Treatment %in% c("siCtrl", "si3d")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3d" = "#D55E00"     # Orange
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-2, 2)) +
  labs(
    title = "Ribosome Occupancy Changes: siCtrl vs si3d",
    subtitle = "Distribution of Ribo log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "Ribo log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

print(plot_ribo_si3d)

ggsave(
  here("plots", "hypoxic_response_ribo_sictrl_vs_si3d_1hr.pdf"),
  plot = plot_ribo_si3d,
  width = 5,
  height = 4,
  device = "pdf"
)
```

## Translation Efficiency Changes: siCtrl vs si3d

```{r density_te_si3d, fig.width=8, fig.height=5}
plot_te_si3d <- combined_long %>%
  filter(Measurement == "TE", Treatment %in% c("siCtrl", "si3d")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3d" = "#D55E00"     # Orange
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-2, 2)) +
  labs(
    title = "Translation Efficiency Changes: siCtrl vs si3d",
    subtitle = "Distribution of TE log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "TE log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

print(plot_te_si3d)

ggsave(
  here("plots", "hypoxic_response_te_sictrl_vs_si3d_1hr.pdf"),
  plot = plot_te_si3d,
  width = 5,
  height = 4,
  device = "pdf"
)
```

# Density Plots: Compare siCtrl vs si3e

These plots compare the hypoxic response between siCtrl (grey) and si3e (blue) for each measurement type.

## RNA Expression Changes: siCtrl vs si3e

```{r density_rna_si3e, fig.width=8, fig.height=5}
plot_rna_si3e <- combined_long %>%
  filter(Measurement == "RNA", Treatment %in% c("siCtrl", "si3e")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3e" = "#0072B2"     # Blue
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    title = "RNA Expression Changes: siCtrl vs si3e",
    subtitle = "Distribution of RNA log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "RNA log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

print(plot_rna_si3e)

ggsave(
  here("plots", "hypoxic_response_rna_sictrl_vs_si3e_1hr.pdf"),
  plot = plot_rna_si3e,
  width = 8,
  height = 5,
  device = "pdf"
)
```

## Ribosome Occupancy Changes: siCtrl vs si3e

```{r density_ribo_si3e, fig.width=8, fig.height=5}
plot_ribo_si3e <- combined_long %>%
  filter(Measurement == "Ribo", Treatment %in% c("siCtrl", "si3e")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3e" = "#0072B2"     # Blue
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    title = "Ribosome Occupancy Changes: siCtrl vs si3e",
    subtitle = "Distribution of Ribo log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "Ribo log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

print(plot_ribo_si3e)

ggsave(
  here("plots", "hypoxic_response_ribo_sictrl_vs_si3e_1hr.pdf"),
  plot = plot_ribo_si3e,
  width = 8,
  height = 5,
  device = "pdf"
)
```

## Translation Efficiency Changes: siCtrl vs si3e

```{r density_te_si3e, fig.width=8, fig.height=5}
plot_te_si3e <- combined_long %>%
  filter(Measurement == "TE", Treatment %in% c("siCtrl", "si3e")) %>%
  ggplot(aes(x = log2FC, color = Treatment)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "siCtrl" = "#808080",  # Grey
      "si3e" = "#0072B2"     # Blue
    ),
    name = "Treatment"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    title = "Translation Efficiency Changes: siCtrl vs si3e",
    subtitle = "Distribution of TE log2FC (Hypoxia vs Normoxia) at 1 Hour",
    x = "TE log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

print(plot_te_si3e)

ggsave(
  here("plots", "hypoxic_response_te_sictrl_vs_si3e_1hr.pdf"),
  plot = plot_te_si3e,
  width = 8,
  height = 5,
  device = "pdf"
)
```

# Density Plot: Compare Measurements by Treatment

This plot shows the distribution of RNA, Ribo, and TE log2 fold changes within each treatment. This visualization makes it easy to see the relationship between measurements (e.g., whether TE changes are driven by RNA or Ribo changes) within each treatment context.

```{r density_by_treatment, fig.width=12, fig.height=4}
plot_by_treatment <- ggplot(combined_long, aes(x = log2FC, color = Measurement)) +
  geom_density(size = 1.2, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "RNA" = "#E69F00",   # Orange
      "Ribo" = "#CC79A7",  # Pink
      "TE" = "#56B4E9"     # Light blue
    ),
    name = "Measurement"
  ) +
  facet_wrap(~Treatment, ncol = 3, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", size = 0.5) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    title = "RNA, Ribo, and TE Changes in Hypoxic Response",
    subtitle = "Comparison across siCtrl, si3d, and si3e Treatments at 1 Hour",
    x = "log2 Fold Change",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "grey90")
  )

print(plot_by_treatment)

ggsave(
  here("plots", "hypoxic_response_density_by_treatment_1hr.pdf"),
  plot = plot_by_treatment,
  width = 12,
  height = 4,
  device = "pdf"
)
```

# Interpretation Guide

## What to look for in these plots:

### siCtrl vs si3d Plots (Grey vs Orange):
- **Distribution shifts**: Does si3d shift the distribution left (attenuated response) or right (enhanced response) compared to siCtrl?
- **Peak positions**: Is the si3d peak at a different position than siCtrl?
- **Spread differences**: Does si3d knockdown increase or decrease variability in the hypoxic response?
- **Measurement-specific effects**: Does si3d have similar effects on RNA, Ribo, and TE, or are some measurements more affected?

### siCtrl vs si3e Plots (Grey vs Blue):
- **Distribution shifts**: Does si3e shift the distribution left (attenuated response) or right (enhanced response) compared to siCtrl?
- **Peak positions**: Is the si3e peak at a different position than siCtrl?
- **Spread differences**: Does si3e knockdown increase or decrease variability in the hypoxic response?
- **Measurement-specific effects**: Does si3e have similar effects on RNA, Ribo, and TE, or are some measurements more affected?

### Comparing si3d vs si3e:
- **Similarities**: Do both knockdowns shift distributions in the same direction?
- **Differences**: Are there measurement types where si3d and si3e have opposite effects?
- **Magnitude**: Which knockdown has a stronger effect on the hypoxic response?

### Plot by Treatment (Faceted):
- **RNA-Ribo coupling**: Do RNA and Ribo distributions overlap (forwarded regulation) or differ (translational buffering)?
- **TE patterns**: TE distribution shows the net translational regulation independent of mRNA changes
- **Treatment-specific patterns**: How does the RNA-Ribo-TE relationship differ between siCtrl, si3d, and si3e?

## Key Questions:
1. Does knocking down eIF3d or eIF3e dampen or enhance the hypoxic response?
2. Do the knockdowns primarily affect RNA levels, ribosome occupancy, or translation efficiency?
3. Are eIF3d and eIF3e knockdowns similar or different in their effects?
4. Does the hypoxic response become more or less variable with the knockdowns?
5. Which measurement type (RNA, Ribo, or TE) shows the most dramatic differences between treatments?

# Session Info

```{r session_info}
sessionInfo()
```
