---
title: "Generate TE Signature Gene Sets"
author: "Kate"
date: "2025-12-03"
output: html_document
editor_options:
  chunk_output_type: console
---

Generate TE signature gene sets for eIF3d and eIF3e across all conditions and timepoints.
This follows the format from the 2024_eIF3e_hypoxia project (intersect.Rmd).

Gene sets represent genes whose translation is promoted or inhibited by eIF3d/eIF3e.
The DESeq2 comparison is sictrl vs si3d/si3e (control - knockdown), so:
- Positive LFC = higher in control = eIF3d/e PROMOTES translation
- Negative LFC = lower in control = eIF3d/e INHIBITS translation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r load_libraries}
library(here)
library(tidyverse)
library(DT)

cat("Libraries loaded successfully\n")
```

# si3d Signatures

## si3d Normoxia

### 1 Hour

```{r si3d_normoxia_1hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")
)

# Create PROMOTES gene set (positive LFC)
normoxia_3d_promotes_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_promotes_TE_1hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_promotes_TE_1hr$ensembl_gene)

normoxia_3d_promotes_TE_1hr <- normoxia_3d_promotes_TE_1hr %>%
  mutate(
    gs_name = "normoxia_3d_promotes_TE_1hr",
    gs_description = "genes whose translation is promoted by eIF3d in normoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_promotes_TE_1hr:", nrow(normoxia_3d_promotes_TE_1hr), "genes\n")

# Create INHIBITS gene set (negative LFC)
normoxia_3d_inhibits_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_inhibits_TE_1hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_inhibits_TE_1hr$ensembl_gene)

normoxia_3d_inhibits_TE_1hr <- normoxia_3d_inhibits_TE_1hr %>%
  mutate(
    gs_name = "normoxia_3d_inhibits_TE_1hr",
    gs_description = "genes whose translation is inhibited by eIF3d in normoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_inhibits_TE_1hr:", nrow(normoxia_3d_inhibits_TE_1hr), "genes\n")
```

### 4 Hours

```{r si3d_normoxia_4hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")
)

# Create PROMOTES gene set
normoxia_3d_promotes_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_promotes_TE_4hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_promotes_TE_4hr$ensembl_gene)

normoxia_3d_promotes_TE_4hr <- normoxia_3d_promotes_TE_4hr %>%
  mutate(
    gs_name = "normoxia_3d_promotes_TE_4hr",
    gs_description = "genes whose translation is promoted by eIF3d in normoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_promotes_TE_4hr:", nrow(normoxia_3d_promotes_TE_4hr), "genes\n")

# Create INHIBITS gene set
normoxia_3d_inhibits_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_inhibits_TE_4hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_inhibits_TE_4hr$ensembl_gene)

normoxia_3d_inhibits_TE_4hr <- normoxia_3d_inhibits_TE_4hr %>%
  mutate(
    gs_name = "normoxia_3d_inhibits_TE_4hr",
    gs_description = "genes whose translation is inhibited by eIF3d in normoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_inhibits_TE_4hr:", nrow(normoxia_3d_inhibits_TE_4hr), "genes\n")
```

### 24 Hours

```{r si3d_normoxia_24hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")
)

# Create PROMOTES gene set
normoxia_3d_promotes_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_promotes_TE_24hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_promotes_TE_24hr$ensembl_gene)

normoxia_3d_promotes_TE_24hr <- normoxia_3d_promotes_TE_24hr %>%
  mutate(
    gs_name = "normoxia_3d_promotes_TE_24hr",
    gs_description = "genes whose translation is promoted by eIF3d in normoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_promotes_TE_24hr:", nrow(normoxia_3d_promotes_TE_24hr), "genes\n")

# Create INHIBITS gene set
normoxia_3d_inhibits_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_inhibits_TE_24hr$ensembl_gene <- sub("\\..*", "", normoxia_3d_inhibits_TE_24hr$ensembl_gene)

normoxia_3d_inhibits_TE_24hr <- normoxia_3d_inhibits_TE_24hr %>%
  mutate(
    gs_name = "normoxia_3d_inhibits_TE_24hr",
    gs_description = "genes whose translation is inhibited by eIF3d in normoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_inhibits_TE_24hr:", nrow(normoxia_3d_inhibits_TE_24hr), "genes\n")
```

### Combined (All Timepoints)

```{r si3d_normoxia_combined}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_combined.csv")
)

# Create PROMOTES gene set (positive LFC)
normoxia_3d_promotes_TE_combined <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_promotes_TE_combined$ensembl_gene <- sub("\\..*", "", normoxia_3d_promotes_TE_combined$ensembl_gene)

normoxia_3d_promotes_TE_combined <- normoxia_3d_promotes_TE_combined %>%
  mutate(
    gs_name = "normoxia_3d_promotes_TE_combined",
    gs_description = "genes whose translation is promoted by eIF3d in normoxia, pooled across 3 timepoints (1hr, 4hr, 24hr), in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_promotes_TE_combined:", nrow(normoxia_3d_promotes_TE_combined), "genes\n")

# Create INHIBITS gene set (negative LFC)
normoxia_3d_inhibits_TE_combined <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3d_inhibits_TE_combined$ensembl_gene <- sub("\\..*", "", normoxia_3d_inhibits_TE_combined$ensembl_gene)

normoxia_3d_inhibits_TE_combined <- normoxia_3d_inhibits_TE_combined %>%
  mutate(
    gs_name = "normoxia_3d_inhibits_TE_combined",
    gs_description = "genes whose translation is inhibited by eIF3d in normoxia, pooled across 3 timepoints (1hr, 4hr, 24hr), in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3d_inhibits_TE_combined:", nrow(normoxia_3d_inhibits_TE_combined), "genes\n")
```

## si3d Hypoxia

### 1 Hour

```{r si3d_hypoxia_1hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")
)

# Create PROMOTES gene set
hypoxia_3d_promotes_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_promotes_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_promotes_TE_1hr$ensembl_gene)

hypoxia_3d_promotes_TE_1hr <- hypoxia_3d_promotes_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_3d_promotes_TE_1hr",
    gs_description = "genes whose translation is promoted by eIF3d in hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_promotes_TE_1hr:", nrow(hypoxia_3d_promotes_TE_1hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3d_inhibits_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_inhibits_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_inhibits_TE_1hr$ensembl_gene)

hypoxia_3d_inhibits_TE_1hr <- hypoxia_3d_inhibits_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_3d_inhibits_TE_1hr",
    gs_description = "genes whose translation is inhibited by eIF3d in hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_inhibits_TE_1hr:", nrow(hypoxia_3d_inhibits_TE_1hr), "genes\n")
```

### 4 Hours

```{r si3d_hypoxia_4hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")
)

# Create PROMOTES gene set
hypoxia_3d_promotes_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_promotes_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_promotes_TE_4hr$ensembl_gene)

hypoxia_3d_promotes_TE_4hr <- hypoxia_3d_promotes_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_3d_promotes_TE_4hr",
    gs_description = "genes whose translation is promoted by eIF3d in hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_promotes_TE_4hr:", nrow(hypoxia_3d_promotes_TE_4hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3d_inhibits_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_inhibits_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_inhibits_TE_4hr$ensembl_gene)

hypoxia_3d_inhibits_TE_4hr <- hypoxia_3d_inhibits_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_3d_inhibits_TE_4hr",
    gs_description = "genes whose translation is inhibited by eIF3d in hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_inhibits_TE_4hr:", nrow(hypoxia_3d_inhibits_TE_4hr), "genes\n")
```

### 24 Hours

```{r si3d_hypoxia_24hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")
)

# Create PROMOTES gene set
hypoxia_3d_promotes_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_promotes_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_promotes_TE_24hr$ensembl_gene)

hypoxia_3d_promotes_TE_24hr <- hypoxia_3d_promotes_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_3d_promotes_TE_24hr",
    gs_description = "genes whose translation is promoted by eIF3d in hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_promotes_TE_24hr:", nrow(hypoxia_3d_promotes_TE_24hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3d_inhibits_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3d_inhibits_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_3d_inhibits_TE_24hr$ensembl_gene)

hypoxia_3d_inhibits_TE_24hr <- hypoxia_3d_inhibits_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_3d_inhibits_TE_24hr",
    gs_description = "genes whose translation is inhibited by eIF3d in hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3d_inhibits_TE_24hr:", nrow(hypoxia_3d_inhibits_TE_24hr), "genes\n")
```

# si3e Signatures

## si3e Normoxia

### 1 Hour

```{r si3e_normoxia_1hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")
)

# Create PROMOTES gene set
normoxia_3e_promotes_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_promotes_TE_1hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_promotes_TE_1hr$ensembl_gene)

normoxia_3e_promotes_TE_1hr <- normoxia_3e_promotes_TE_1hr %>%
  mutate(
    gs_name = "normoxia_3e_promotes_TE_1hr",
    gs_description = "genes whose translation is promoted by eIF3e in normoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_promotes_TE_1hr:", nrow(normoxia_3e_promotes_TE_1hr), "genes\n")

# Create INHIBITS gene set
normoxia_3e_inhibits_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_inhibits_TE_1hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_inhibits_TE_1hr$ensembl_gene)

normoxia_3e_inhibits_TE_1hr <- normoxia_3e_inhibits_TE_1hr %>%
  mutate(
    gs_name = "normoxia_3e_inhibits_TE_1hr",
    gs_description = "genes whose translation is inhibited by eIF3e in normoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_inhibits_TE_1hr:", nrow(normoxia_3e_inhibits_TE_1hr), "genes\n")
```

### 4 Hours

```{r si3e_normoxia_4hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")
)

# Create PROMOTES gene set
normoxia_3e_promotes_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_promotes_TE_4hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_promotes_TE_4hr$ensembl_gene)

normoxia_3e_promotes_TE_4hr <- normoxia_3e_promotes_TE_4hr %>%
  mutate(
    gs_name = "normoxia_3e_promotes_TE_4hr",
    gs_description = "genes whose translation is promoted by eIF3e in normoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_promotes_TE_4hr:", nrow(normoxia_3e_promotes_TE_4hr), "genes\n")

# Create INHIBITS gene set
normoxia_3e_inhibits_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_inhibits_TE_4hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_inhibits_TE_4hr$ensembl_gene)

normoxia_3e_inhibits_TE_4hr <- normoxia_3e_inhibits_TE_4hr %>%
  mutate(
    gs_name = "normoxia_3e_inhibits_TE_4hr",
    gs_description = "genes whose translation is inhibited by eIF3e in normoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_inhibits_TE_4hr:", nrow(normoxia_3e_inhibits_TE_4hr), "genes\n")
```

### 24 Hours

```{r si3e_normoxia_24hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")
)

# Create PROMOTES gene set
normoxia_3e_promotes_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_promotes_TE_24hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_promotes_TE_24hr$ensembl_gene)

normoxia_3e_promotes_TE_24hr <- normoxia_3e_promotes_TE_24hr %>%
  mutate(
    gs_name = "normoxia_3e_promotes_TE_24hr",
    gs_description = "genes whose translation is promoted by eIF3e in normoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_promotes_TE_24hr:", nrow(normoxia_3e_promotes_TE_24hr), "genes\n")

# Create INHIBITS gene set
normoxia_3e_inhibits_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_inhibits_TE_24hr$ensembl_gene <- sub("\\..*", "", normoxia_3e_inhibits_TE_24hr$ensembl_gene)

normoxia_3e_inhibits_TE_24hr <- normoxia_3e_inhibits_TE_24hr %>%
  mutate(
    gs_name = "normoxia_3e_inhibits_TE_24hr",
    gs_description = "genes whose translation is inhibited by eIF3e in normoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_inhibits_TE_24hr:", nrow(normoxia_3e_inhibits_TE_24hr), "genes\n")
```

### Combined (All Timepoints)

```{r si3e_normoxia_combined}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_normoxia_combined.csv")
)

# Create PROMOTES gene set (positive LFC)
normoxia_3e_promotes_TE_combined <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_promotes_TE_combined$ensembl_gene <- sub("\\..*", "", normoxia_3e_promotes_TE_combined$ensembl_gene)

normoxia_3e_promotes_TE_combined <- normoxia_3e_promotes_TE_combined %>%
  mutate(
    gs_name = "normoxia_3e_promotes_TE_combined",
    gs_description = "genes whose translation is promoted by eIF3e in normoxia, pooled across 3 timepoints (1hr, 4hr, 24hr), in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_promotes_TE_combined:", nrow(normoxia_3e_promotes_TE_combined), "genes\n")

# Create INHIBITS gene set (negative LFC)
normoxia_3e_inhibits_TE_combined <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

normoxia_3e_inhibits_TE_combined$ensembl_gene <- sub("\\..*", "", normoxia_3e_inhibits_TE_combined$ensembl_gene)

normoxia_3e_inhibits_TE_combined <- normoxia_3e_inhibits_TE_combined %>%
  mutate(
    gs_name = "normoxia_3e_inhibits_TE_combined",
    gs_description = "genes whose translation is inhibited by eIF3e in normoxia, pooled across 3 timepoints (1hr, 4hr, 24hr), in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("normoxia_3e_inhibits_TE_combined:", nrow(normoxia_3e_inhibits_TE_combined), "genes\n")
```

## si3e Hypoxia

### 1 Hour

```{r si3e_hypoxia_1hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")
)

# Create PROMOTES gene set
hypoxia_3e_promotes_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_promotes_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_promotes_TE_1hr$ensembl_gene)

hypoxia_3e_promotes_TE_1hr <- hypoxia_3e_promotes_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_3e_promotes_TE_1hr",
    gs_description = "genes whose translation is promoted by eIF3e in hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_promotes_TE_1hr:", nrow(hypoxia_3e_promotes_TE_1hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3e_inhibits_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_inhibits_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_inhibits_TE_1hr$ensembl_gene)

hypoxia_3e_inhibits_TE_1hr <- hypoxia_3e_inhibits_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_3e_inhibits_TE_1hr",
    gs_description = "genes whose translation is inhibited by eIF3e in hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_inhibits_TE_1hr:", nrow(hypoxia_3e_inhibits_TE_1hr), "genes\n")
```

### 4 Hours

```{r si3e_hypoxia_4hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")
)

# Create PROMOTES gene set
hypoxia_3e_promotes_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_promotes_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_promotes_TE_4hr$ensembl_gene)

hypoxia_3e_promotes_TE_4hr <- hypoxia_3e_promotes_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_3e_promotes_TE_4hr",
    gs_description = "genes whose translation is promoted by eIF3e in hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_promotes_TE_4hr:", nrow(hypoxia_3e_promotes_TE_4hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3e_inhibits_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_inhibits_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_inhibits_TE_4hr$ensembl_gene)

hypoxia_3e_inhibits_TE_4hr <- hypoxia_3e_inhibits_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_3e_inhibits_TE_4hr",
    gs_description = "genes whose translation is inhibited by eIF3e in hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_inhibits_TE_4hr:", nrow(hypoxia_3e_inhibits_TE_4hr), "genes\n")
```

### 24 Hours

```{r si3e_hypoxia_24hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")
)

# Create PROMOTES gene set
hypoxia_3e_promotes_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_promotes_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_promotes_TE_24hr$ensembl_gene)

hypoxia_3e_promotes_TE_24hr <- hypoxia_3e_promotes_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_3e_promotes_TE_24hr",
    gs_description = "genes whose translation is promoted by eIF3e in hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_promotes_TE_24hr:", nrow(hypoxia_3e_promotes_TE_24hr), "genes\n")

# Create INHIBITS gene set
hypoxia_3e_inhibits_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_3e_inhibits_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_3e_inhibits_TE_24hr$ensembl_gene)

hypoxia_3e_inhibits_TE_24hr <- hypoxia_3e_inhibits_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_3e_inhibits_TE_24hr",
    gs_description = "genes whose translation is inhibited by eIF3e in hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_3e_inhibits_TE_24hr:", nrow(hypoxia_3e_inhibits_TE_24hr), "genes\n")
```

# Hypoxia Response Signatures

## 1 Hour

```{r hypoxia_response_1hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_1hr.csv")
)

# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_1hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data %>%
  left_join(gene_symbols, by = "gene_id")

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_1hr$ensembl_gene)

hypoxia_response_increases_TE_1hr <- hypoxia_response_increases_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_1hr",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_1hr:", nrow(hypoxia_response_increases_TE_1hr), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_1hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_1hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_1hr$ensembl_gene)

hypoxia_response_decreases_TE_1hr <- hypoxia_response_decreases_TE_1hr %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_1hr",
    gs_description = "genes whose translation efficiency is decreased by hypoxia at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_1hr:", nrow(hypoxia_response_decreases_TE_1hr), "genes\n")
```

## 4 Hours

```{r hypoxia_response_4hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_4hr.csv")
)

# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_4hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data %>%
  left_join(gene_symbols, by = "gene_id")

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_4hr$ensembl_gene)

hypoxia_response_increases_TE_4hr <- hypoxia_response_increases_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_4hr",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_4hr:", nrow(hypoxia_response_increases_TE_4hr), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_4hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_4hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_4hr$ensembl_gene)

hypoxia_response_decreases_TE_4hr <- hypoxia_response_decreases_TE_4hr %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_4hr",
    gs_description = "genes whose translation efficiency is decreased by hypoxia at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_4hr:", nrow(hypoxia_response_decreases_TE_4hr), "genes\n")
```

## 24 Hours

```{r hypoxia_response_24hr}
# Load data
translation_data <- read_csv(
  here("output", "translation_categories_24hr.csv")
)

# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_24hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data %>%
  left_join(gene_symbols, by = "gene_id")

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_24hr$ensembl_gene)

hypoxia_response_increases_TE_24hr <- hypoxia_response_increases_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_24hr",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_24hr:", nrow(hypoxia_response_increases_TE_24hr), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_24hr <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_24hr$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_24hr$ensembl_gene)

hypoxia_response_decreases_TE_24hr <- hypoxia_response_decreases_TE_24hr %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_24hr",
    gs_description = "genes whose translation efficiency is decreased by hypoxia at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_24hr:", nrow(hypoxia_response_decreases_TE_24hr), "genes\n")
```

####
# Hypoxic response that is unchanged with 3d/d KD


## 1 Hour

```{r hypoxia_response_1hr kd}
# Load data
translation_data_sictrl <- read_csv(
  here("output", "translation_categories_1hr.csv")
)

translation_data_3dkd <- read_csv(
  here("output", "translation_categories_si3d_1hr.csv")
)

translation_data_3ekd <- read_csv(
  here("output", "translation_categories_si3e_1hr.csv")
)

unchanged_3d <- translation_data_3dkd |>
  filter(te_padj > 0.05) |> pull(gene_id)
unchanged_3e <- translation_data_3ekd |>
  filter(te_padj > 0.05) |> pull(gene_id)



# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_1hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data_sictrl %>%
  left_join(gene_symbols, by = "gene_id")

# 3d KD

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd$ensembl_gene)

hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd <- hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 1hr but inhibited with 3d KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd:", nrow(hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_1hr_inhibited_by_3d <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_1hr_inhibited_by_3d$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_1hr_inhibited_by_3d$ensembl_gene)

hypoxia_response_decreases_TE_1hr_inhibited_by_3d <- hypoxia_response_decreases_TE_1hr_inhibited_by_3d %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_1hr_inhibited_by_3d",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3d KD at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_1hr_inhibited_by_3d:", nrow(hypoxia_response_decreases_TE_1hr_inhibited_by_3d), "genes\n")




# 3e KD

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd$ensembl_gene)

hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd <- hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 1hr but inhibited with 3e KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd:", nrow(hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_1hr_inhibited_by_3e <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_1hr_inhibited_by_3e$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_1hr_inhibited_by_3e$ensembl_gene)

hypoxia_response_decreases_TE_1hr_inhibited_by_3e <- hypoxia_response_decreases_TE_1hr_inhibited_by_3e %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_1hr_inhibited_by_3e",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3e KD at 1hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_1hr_inhibited_by_3e:", nrow(hypoxia_response_decreases_TE_1hr_inhibited_by_3e), "genes\n")
```

## 4 hr

```{r hypoxia_response_4hr kd}
# Load data
translation_data_sictrl <- read_csv(
  here("output", "translation_categories_4hr.csv")
)

translation_data_3dkd <- read_csv(
  here("output", "translation_categories_si3d_4hr.csv")
)

translation_data_3ekd <- read_csv(
  here("output", "translation_categories_si3e_4hr.csv")
)

unchanged_3d <- translation_data_3dkd |>
  filter(te_padj > 0.05) |> pull(gene_id)
unchanged_3e <- translation_data_3ekd |>
  filter(te_padj > 0.05) |> pull(gene_id)



# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_4hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data_sictrl %>%
  left_join(gene_symbols, by = "gene_id")

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd$ensembl_gene)

hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd <- hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 4hr but inhibited with 3d KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd:", nrow(hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_4hr_inhibited_by_3d <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_4hr_inhibited_by_3d$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_4hr_inhibited_by_3d$ensembl_gene)

hypoxia_response_decreases_TE_4hr_inhibited_by_3d <- hypoxia_response_decreases_TE_4hr_inhibited_by_3d %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_4hr_inhibited_by_3d",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3d KD at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_4hr_inhibited_by_3d:", nrow(hypoxia_response_decreases_TE_4hr_inhibited_by_3d), "genes\n")

# 3e

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd$ensembl_gene)

hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd <- hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 4hr but inhibited with 3e KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd:", nrow(hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_4hr_inhibited_by_3e <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_4hr_inhibited_by_3e$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_4hr_inhibited_by_3e$ensembl_gene)

hypoxia_response_decreases_TE_4hr_inhibited_by_3e <- hypoxia_response_decreases_TE_4hr_inhibited_by_3e %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_4hr_inhibited_by_3e",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3e KD at 4hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_4hr_inhibited_by_3e:", nrow(hypoxia_response_decreases_TE_4hr_inhibited_by_3e), "genes\n")
```


## 24 hr

```{r hypoxia_response_24hr kd}
# Load data
translation_data_sictrl <- read_csv(
  here("output", "translation_categories_24hr.csv")
)

translation_data_3dkd <- read_csv(
  here("output", "translation_categories_si3d_24hr.csv")
)

translation_data_3ekd <- read_csv(
  here("output", "translation_categories_si3e_24hr.csv")
)

unchanged_3d <- translation_data_3dkd |>
  filter(te_padj > 0.05) |> pull(gene_id)
unchanged_3e <- translation_data_3ekd |>
  filter(te_padj > 0.05) |> pull(gene_id)



# Add gene symbols (need to join from existing gene symbol mapping)
gene_symbols <- read_csv(here("output", "deseq_results_24hr_spike.csv")) %>%
  select(gene_id, symbol)

translation_data <- translation_data_sictrl %>%
  left_join(gene_symbols, by = "gene_id")

## 3d

# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd$ensembl_gene)

hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd <- hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 24hr but inhibited with 3d KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd:", nrow(hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_24hr_inhibited_by_3d <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3d) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_24hr_inhibited_by_3d$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_24hr_inhibited_by_3d$ensembl_gene)

hypoxia_response_decreases_TE_24hr_inhibited_by_3d <- hypoxia_response_decreases_TE_24hr_inhibited_by_3d %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_24hr_inhibited_by_3d",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3d KD at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_24hr_inhibited_by_3d:", nrow(hypoxia_response_decreases_TE_24hr_inhibited_by_3d), "genes\n")

## 3e


# Create INCREASES gene set (positive LFC)
hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd$ensembl_gene <- sub("\\..*", "", hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd$ensembl_gene)

hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd <- hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd %>%
  mutate(
    gs_name = "hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd",
    gs_description = "genes whose translation efficiency is increased by hypoxia at 24hr but inhibited with 3e KD, in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd:", nrow(hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd), "genes\n")

# Create DECREASES gene set (negative LFC)
hypoxia_response_decreases_TE_24hr_inhibited_by_3e <- translation_data %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
    filter(gene_id %in% unchanged_3e) |> 
  arrange(desc(abs(te_lfc))) %>%
  slice_head(n = 2500) %>%
  select(gene_id, symbol) %>%
  rename(ensembl_gene = gene_id)

hypoxia_response_decreases_TE_24hr_inhibited_by_3e$ensembl_gene <- sub("\\..*", "", hypoxia_response_decreases_TE_24hr_inhibited_by_3e$ensembl_gene)

hypoxia_response_decreases_TE_24hr_inhibited_by_3e <- hypoxia_response_decreases_TE_24hr_inhibited_by_3e %>%
  mutate(
    gs_name = "hypoxia_response_decreases_TE_24hr_inhibited_by_3e",
    gs_description = "genes whose translation efficiency is decreased by hypoxia but inhibited by 3e KD at 24hr in MDA-MB-231 cells, spike-in normalized",
    gs_source_species = "homo sapiens"
  ) %>%
  rename(db_gene_symbol = symbol) %>%
  select(db_gene_symbol, ensembl_gene, gs_name, gs_description, gs_source_species)

cat("hypoxia_response_decreases_TE_24hr_inhibited_by_3e:", nrow(hypoxia_response_decreases_TE_24hr_inhibited_by_3e), "genes\n")
```






# Export All Gene Sets

## Export Individual Files

```{r export_individual_files}
# Create output directory if it doesn't exist
output_dir <- here("output", "genesets")
if(!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Created directory:", output_dir, "\n")
}

# List of all gene sets to export
gene_sets <- list(
  normoxia_3d_promotes_TE_combined = normoxia_3d_promotes_TE_combined,
  normoxia_3d_inhibits_TE_combined = normoxia_3d_inhibits_TE_combined,
  hypoxia_3d_promotes_TE_1hr = hypoxia_3d_promotes_TE_1hr,
  hypoxia_3d_inhibits_TE_1hr = hypoxia_3d_inhibits_TE_1hr,
  hypoxia_3d_promotes_TE_4hr = hypoxia_3d_promotes_TE_4hr,
  hypoxia_3d_inhibits_TE_4hr = hypoxia_3d_inhibits_TE_4hr,
  hypoxia_3d_promotes_TE_24hr = hypoxia_3d_promotes_TE_24hr,
  hypoxia_3d_inhibits_TE_24hr = hypoxia_3d_inhibits_TE_24hr,
  normoxia_3e_promotes_TE_combined = normoxia_3e_promotes_TE_combined,
  normoxia_3e_inhibits_TE_combined = normoxia_3e_inhibits_TE_combined,
  hypoxia_3e_promotes_TE_1hr = hypoxia_3e_promotes_TE_1hr,
  hypoxia_3e_inhibits_TE_1hr = hypoxia_3e_inhibits_TE_1hr,
  hypoxia_3e_promotes_TE_4hr = hypoxia_3e_promotes_TE_4hr,
  hypoxia_3e_inhibits_TE_4hr = hypoxia_3e_inhibits_TE_4hr,
  hypoxia_3e_promotes_TE_24hr = hypoxia_3e_promotes_TE_24hr,
  hypoxia_3e_inhibits_TE_24hr = hypoxia_3e_inhibits_TE_24hr,
  hypoxia_response_increases_TE_1hr = hypoxia_response_increases_TE_1hr,
  hypoxia_response_decreases_TE_1hr = hypoxia_response_decreases_TE_1hr,
  hypoxia_response_increases_TE_4hr = hypoxia_response_increases_TE_4hr,
  hypoxia_response_decreases_TE_4hr = hypoxia_response_decreases_TE_4hr,
  hypoxia_response_increases_TE_24hr = hypoxia_response_increases_TE_24hr,
  hypoxia_response_decreases_TE_24hr = hypoxia_response_decreases_TE_24hr,
  hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd = hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd,
  hypoxia_response_decreases_TE_1hr_inhibited_by_3d = hypoxia_response_decreases_TE_1hr_inhibited_by_3d,
  hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd = hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd,
  hypoxia_response_decreases_TE_1hr_inhibited_by_3e = hypoxia_response_decreases_TE_1hr_inhibited_by_3e,
  hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd = hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd,
  hypoxia_response_decreases_TE_4hr_inhibited_by_3d = hypoxia_response_decreases_TE_4hr_inhibited_by_3d,
  hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd = hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd,
  hypoxia_response_decreases_TE_4hr_inhibited_by_3e = hypoxia_response_decreases_TE_4hr_inhibited_by_3e,
  hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd = hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd,
  hypoxia_response_decreases_TE_24hr_inhibited_by_3d = hypoxia_response_decreases_TE_24hr_inhibited_by_3d,
  hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd = hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd,
  hypoxia_response_decreases_TE_24hr_inhibited_by_3e = hypoxia_response_decreases_TE_24hr_inhibited_by_3e
)

# Export each gene set
cat("\n=== EXPORTING INDIVIDUAL FILES ===\n")
for (name in names(gene_sets)) {
  file_path <- here(output_dir, paste0(name, ".csv"))
  write_csv(gene_sets[[name]], file_path)
  cat("Exported", name, "to", file_path, "(", nrow(gene_sets[[name]]), "genes )\n")
}
```



## Create Combined File

```{r create_combined_file}
# Combine all gene sets into one data frame
all_signatures <- bind_rows(gene_sets)

# Export combined file
combined_file <- here(output_dir, "all_te_signatures.csv")
write_csv(all_signatures, combined_file)
cat("\n=== COMBINED FILE ===\n")
cat("Exported combined gene sets to:", combined_file, "\n")
cat("Total rows:", nrow(all_signatures), "\n")
```

## Create Summary File

```{r create_summary}
# Create summary with gene set names, descriptions, and gene counts
summary_df <- data.frame(
  gene_set = names(gene_sets),
  n_genes = sapply(gene_sets, nrow),
  description = sapply(gene_sets, function(gs) unique(gs$gs_description)[1]),
  stringsAsFactors = FALSE
)

# Export summary
summary_file <- here(output_dir, "geneset_summary.csv")
write_csv(summary_df, summary_file)

cat("\n=== SUMMARY FILE ===\n")
cat("Exported summary to:", summary_file, "\n")

cat("\n=== EXPORT COMPLETE ===\n")
cat("Total gene sets created: 34\n")
cat("Individual files: 34\n")
cat("Combined file: 1\n")
cat("Summary file: 1\n")
```



Read in genesets from MCF7-six1 and overlap

combine with all_signatures
```{r}
cell_line_overlap_all <- read_csv("/Users/katematlin/github/six1_231_extra_analysis/output/genesets/all_te_lists.csv")

mcf7six1_only_all <- read_csv("/Users/katematlin/github/six1_231_extra_analysis/output/genesets/mcf7six1_only/all_mcf7six1_signatures.csv")

```

Export all lists

```{r}
# Add cell line prefix to gs_name for each dataset
all_signatures_with_cellline <- all_signatures %>%
  mutate(gs_name = paste0("MDA-MB-231_", gs_name))

cell_line_overlap_all_with_cellline <- cell_line_overlap_all %>%
  mutate(gs_name = paste0("MDA-MB-231_MCF7-SIX1_overlap_", gs_name))

mcf7six1_only_all_with_cellline <- mcf7six1_only_all %>%
  mutate(gs_name = paste0("MCF7-SIX1_", gs_name))

# Combine all datasets
all_combined_sigs_12_9_25 <- rbind(
  all_signatures_with_cellline,
  cell_line_overlap_all_with_cellline,
  mcf7six1_only_all_with_cellline
)

write_csv(all_combined_sigs_12_9_25, here("all_combined_sigs_12_9_25.csv"))
```