---
title: "GSEA: Antigen Presentation – Hypoxic Response & si3d Effect"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Focused fgsea analysis of macrophage antigen presentation and MHC pathways. Two comparisons displayed side by side:

1. **Hypoxic response** (siCtrl hypoxia vs normoxia): 1hr, 4hr, 24hr
2. **si3d effect** (si3d vs siCtrl): Norm 1&4hr, Hyp 1hr, Hyp 4hr, Hyp 24hr

Pathway sets: Reactome antigen presentation (cross-presentation, MHC-I/II), GO antigen processing terms, KEGG antigen processing, Hallmark IFN/allograft sets, and filtered C7 ImmuneSigDB macrophage/APC signatures.

# Load Libraries

```{r load_libraries}
library(fgsea)
library(msigdbr)
library(pheatmap)
library(tidyverse)
library(here)
```

# Load Data

```{r load_data}
# Gene annotation (needed for hypoxic response files, which lack symbol column)
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup()

# Hypoxic response: siCtrl hypoxia vs normoxia (need symbol join)
te_hyp_1hr  <- read_csv(here("output", "translation_categories_1hr.csv"))  %>% left_join(gene_symbols, by = "gene_id")
te_hyp_4hr  <- read_csv(here("output", "translation_categories_4hr.csv"))  %>% left_join(gene_symbols, by = "gene_id")
te_hyp_24hr <- read_csv(here("output", "translation_categories_24hr.csv")) %>% left_join(gene_symbols, by = "gene_id")

# si3d effect: si3d vs siCtrl (symbol already present)
te_si3d_norm  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
te_si3d_hyp1  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
te_si3d_hyp4  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
te_si3d_hyp24 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# Sanity Checks

```{r sanity_checks}
summarise_te <- function(df, label) {
  sig <- df %>% filter(te_padj < 0.05, abs(te_lfc) > 0.5, !is.na(symbol))
  cat(label, "\n")
  cat("  Total genes:", nrow(df), "\n")
  cat("  With symbol:", sum(!is.na(df$symbol)), "\n")
  cat("  TE sig (padj < 0.05 & |lfc| > 0.5):", nrow(sig),
      "– up:", sum(sig$te_lfc > 0.5),
      "| down:", sum(sig$te_lfc < -0.5), "\n\n")
}

cat("=== Hypoxic response (siCtrl hypoxia vs normoxia) ===\n\n")
summarise_te(te_hyp_1hr,  "Hypoxia 1hr")
summarise_te(te_hyp_4hr,  "Hypoxia 4hr")
summarise_te(te_hyp_24hr, "Hypoxia 24hr")

cat("=== si3d effect (si3d vs siCtrl) ===\n\n")
summarise_te(te_si3d_norm,  "Normoxia 1&4hr")
summarise_te(te_si3d_hyp1,  "Hypoxia 1hr")
summarise_te(te_si3d_hyp4,  "Hypoxia 4hr")
summarise_te(te_si3d_hyp24, "Hypoxia 24hr")
```

# Build Ranked Lists

```{r ranked_lists}
make_ranked <- function(df) {
  df <- df %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(te_lfc), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranked <- df$te_lfc
  names(ranked) <- df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked[!duplicated(names(ranked))]
}

# Hypoxic response
ranked_hyp_1hr  <- make_ranked(te_hyp_1hr)
ranked_hyp_4hr  <- make_ranked(te_hyp_4hr)
ranked_hyp_24hr <- make_ranked(te_hyp_24hr)

# si3d effect
ranked_si3d_norm  <- make_ranked(te_si3d_norm)
ranked_si3d_hyp1  <- make_ranked(te_si3d_hyp1)
ranked_si3d_hyp4  <- make_ranked(te_si3d_hyp4)
ranked_si3d_hyp24 <- make_ranked(te_si3d_hyp24)

cat("Ranked list sizes:\n")
cat("  Hypoxic response – 1hr:", length(ranked_hyp_1hr),
    "| 4hr:", length(ranked_hyp_4hr),
    "| 24hr:", length(ranked_hyp_24hr), "\n")
cat("  si3d effect – Norm:", length(ranked_si3d_norm),
    "| Hyp 1hr:", length(ranked_si3d_hyp1),
    "| Hyp 4hr:", length(ranked_si3d_hyp4),
    "| Hyp 24hr:", length(ranked_si3d_hyp24), "\n")
```

# Build Curated Pathway List

```{r curated_pathways}
to_pathways <- function(db) {
  db %>%
    dplyr::select(gs_name, gene_symbol) %>%
    group_by(gs_name) %>%
    summarise(genes = list(gene_symbol), .groups = "drop") %>%
    deframe()
}

# --- Reactome: specific antigen presentation pathways by R-HSA ID ---
reactome_ids <- c(
  "R-HSA-1236975",  # Antigen processing - Cross presentation
  "R-HSA-2132295",  # MHC class II antigen presentation
  "R-HSA-1236974",  # ER-Phagosome pathway
  "R-HSA-983170"    # Antigen Presentation: Folding, assembly and peptide loading of class I MHC
)

reactome_raw <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
pathways_reactome_curated <- to_pathways(
  reactome_raw %>% filter(gs_exact_source %in% reactome_ids)
)

# --- GO:BP: antigen processing and presentation terms by GO ID ---
go_ids <- c(
  "GO:0019882",  # antigen processing and presentation
  "GO:0019885",  # antigen processing and presentation of exogenous peptide antigen via MHC class I
  "GO:0019886"   # antigen processing and presentation of exogenous peptide antigen via MHC class II
)

gobp_raw <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
pathways_go_curated <- to_pathways(
  gobp_raw %>% filter(gs_exact_source %in% go_ids)
)

# --- KEGG: antigen processing and presentation ---
kegg_ids <- c("hsa04612")

kegg_raw <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY")
pathways_kegg_curated <- to_pathways(
  kegg_raw %>% filter(gs_exact_source %in% kegg_ids)
)

# --- Hallmark: IFN and allograft sets ---
hallmark_names <- c(
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_ALLOGRAFT_REJECTION"
)

hallmark_raw <- msigdbr(species = "Homo sapiens", category = "H")
pathways_hallmark_curated <- to_pathways(
  hallmark_raw %>% filter(gs_name %in% hallmark_names)
)

# --- C7 ImmuneSigDB: macrophage/APC-relevant signatures ---
# Filter by keyword patterns in gene set name
apc_keywords <- "ANTIGEN|MHC|_APC_|MACROPHAGE|DENDRITIC|INTERFERON|ALLOGRAFT"

c7_raw <- msigdbr(species = "Homo sapiens", category = "C7", subcategory = "IMMUNESIGDB")
pathways_c7_curated <- to_pathways(
  c7_raw %>% filter(grepl(apc_keywords, gs_name, ignore.case = FALSE))
)

# --- Combine all into a single curated pathway list ---
pathways_curated <- c(
  pathways_reactome_curated,
  pathways_go_curated,
  pathways_kegg_curated,
  pathways_hallmark_curated,
  pathways_c7_curated
)

# Remove any duplicate names (keep first occurrence)
pathways_curated <- pathways_curated[!duplicated(names(pathways_curated))]

cat("Curated pathway counts:\n")
cat("  Reactome:", length(pathways_reactome_curated), "\n")
cat("  GO:BP:", length(pathways_go_curated), "\n")
cat("  KEGG:", length(pathways_kegg_curated), "\n")
cat("  Hallmark:", length(pathways_hallmark_curated), "\n")
cat("  C7 ImmuneSigDB:", length(pathways_c7_curated), "\n")
cat("  Total curated:", length(pathways_curated), "\n\n")
cat("Pathway names:\n")
cat(paste(" -", names(pathways_curated)), sep = "\n")
```

# Run fgsea

```{r run_fgsea}
run_fgsea <- function(ranked, pathways, condition_label) {
  set.seed(9)
  res <- fgsea(
    pathways  = pathways,
    stats     = ranked,
    minSize   = 5,
    maxSize   = 500,
    gseaParam = 0,
    nproc     = 1
  )
  res %>%
    as_tibble() %>%
    dplyr::select(pathway, NES, pval, padj) %>%
    mutate(condition = condition_label)
}

# Hypoxic response
fgsea_hyp_1hr  <- run_fgsea(ranked_hyp_1hr,  pathways_curated, "Hyp 1hr")
fgsea_hyp_4hr  <- run_fgsea(ranked_hyp_4hr,  pathways_curated, "Hyp 4hr")
fgsea_hyp_24hr <- run_fgsea(ranked_hyp_24hr, pathways_curated, "Hyp 24hr")

# si3d effect
fgsea_si3d_norm  <- run_fgsea(ranked_si3d_norm,  pathways_curated, "si3d Norm")
fgsea_si3d_hyp1  <- run_fgsea(ranked_si3d_hyp1,  pathways_curated, "si3d Hyp1")
fgsea_si3d_hyp4  <- run_fgsea(ranked_si3d_hyp4,  pathways_curated, "si3d Hyp4")
fgsea_si3d_hyp24 <- run_fgsea(ranked_si3d_hyp24, pathways_curated, "si3d Hyp24")
```

# Combined Heatmap

```{r combined_heatmap}
col_order <- c("Hyp 1hr", "Hyp 4hr", "Hyp 24hr",
               "si3d Norm", "si3d Hyp1", "si3d Hyp4", "si3d Hyp24")

all_results <- bind_rows(
  fgsea_hyp_1hr, fgsea_hyp_4hr, fgsea_hyp_24hr,
  fgsea_si3d_norm, fgsea_si3d_hyp1, fgsea_si3d_hyp4, fgsea_si3d_hyp24
)

# NES matrix: pivot first, then intersect with curated pathway list
# (pathways filtered out by fgsea minSize won't be in all_results)
wide <- all_results %>%
  dplyr::select(pathway, NES, condition) %>%
  pivot_wider(names_from = condition, values_from = NES, values_fill = 0) %>%
  column_to_rownames("pathway")
for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0

# Keep only curated pathways that are actually present in fgsea results
all_pathways <- intersect(names(pathways_curated), rownames(wide))
n_dropped <- length(names(pathways_curated)) - length(all_pathways)
if (n_dropped > 0) message(n_dropped, " curated pathways dropped (too few genes in ranked list for fgsea minSize filter)")
message(length(all_pathways), " pathways shown in heatmap")

wide <- wide[all_pathways, col_order, drop = FALSE]

# Significance matrix
sig_mat <- all_results %>%
  mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
  dplyr::select(pathway, condition, sig) %>%
  pivot_wider(names_from = condition, values_from = sig, values_fill = "") %>%
  column_to_rownames("pathway")
for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
sig_mat <- sig_mat[all_pathways, col_order, drop = FALSE]

max_abs <- max(abs(wide), na.rm = TRUE)
max_abs <- max(max_abs, 0.1)

p <- pheatmap(
  wide,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  scale           = "none",
  color           = colorRampPalette(c("blue", "white", "red"))(100),
  breaks          = seq(-max_abs, max_abs, length.out = 101),
  display_numbers = sig_mat,
  number_color    = "black",
  fontsize_number = 10,
  main            = "Antigen Presentation Pathways – Hypoxic Response & si3d Effect (* = padj < 0.05)",
  fontsize_row    = 6,
  fontsize_col    = 9,
  cellwidth       = 38,
  cellheight      = 10,
  gaps_col        = 3  # visual gap between hypoxic response and si3d columns
)

outfile <- here("plots", "gsea_antigen_presentation_combined.pdf")
pdf(outfile, width = 14, height = max(5, nrow(wide) * 0.15 + 4))
print(p)
dev.off()
message("Saved: ", outfile, " (", nrow(wide), " pathways)")
```

# Combined Heatmap – No GSE Signatures

Same as above but excluding C7 ImmuneSigDB signatures (GSE_ prefixed names), keeping only the curated Reactome, GO, KEGG, and Hallmark pathways.

```{r combined_heatmap_no_gse}
core_pathways <- all_pathways[!grepl("^GSE", all_pathways)]
message(length(core_pathways), " core pathways (", length(all_pathways) - length(core_pathways), " GSE_ signatures removed)")

wide_core   <- wide[core_pathways, col_order, drop = FALSE]
sig_mat_core <- sig_mat[core_pathways, col_order, drop = FALSE]

max_abs_core <- max(abs(wide_core), na.rm = TRUE)
max_abs_core <- max(max_abs_core, 0.1)

p_core <- pheatmap(
  wide_core,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  scale           = "none",
  color           = colorRampPalette(c("blue", "white", "red"))(100),
  breaks          = seq(-max_abs_core, max_abs_core, length.out = 101),
  display_numbers = sig_mat_core,
  number_color    = "black",
  fontsize_number = 10,
  main            = "Antigen Presentation Pathways (core sets) – Hypoxic Response & si3d Effect (* = padj < 0.05)",
  fontsize_row    = 7,
  fontsize_col    = 9,
  cellwidth       = 38,
  cellheight      = 12,
  gaps_col        = 3
)

outfile_core <- here("plots", "gsea_antigen_presentation_core.pdf")
pdf(outfile_core, width = 14, height = max(5, nrow(wide_core) * 0.2 + 3))
print(p_core)
dev.off()
message("Saved: ", outfile_core, " (", nrow(wide_core), " pathways)")
```

# Enrichment Plots for Key Pathways

Individual GSEA enrichment plots across all 7 conditions for two key pathways: KEGG antigen processing and Reactome MHC class II.

```{r enrichment_plots}
# Identify the exact names as stored in our curated pathway list
kegg_ap_name   <- grep("KEGG.*ANTIGEN", names(pathways_curated), value = TRUE)[1]
reactome_mhc2  <- grep("REACTOME.*MHC.*CLASS.*II|REACTOME.*MHC_CLASS_II", names(pathways_curated), value = TRUE)[1]

cat("KEGG antigen processing pathway:", kegg_ap_name, "\n")
cat("Reactome MHC class II pathway:", reactome_mhc2, "\n")

ranked_list <- list(
  "Hyp 1hr"    = ranked_hyp_1hr,
  "Hyp 4hr"    = ranked_hyp_4hr,
  "Hyp 24hr"   = ranked_hyp_24hr,
  "si3d Norm"  = ranked_si3d_norm,
  "si3d Hyp1"  = ranked_si3d_hyp1,
  "si3d Hyp4"  = ranked_si3d_hyp4,
  "si3d Hyp24" = ranked_si3d_hyp24
)

make_enrichment_panels <- function(pathway_name, ranked_list, pathways, outfile) {
  if (is.na(pathway_name) || !pathway_name %in% names(pathways)) {
    message("Pathway not found: ", pathway_name)
    return(invisible(NULL))
  }
  # Each condition gets its own page in the PDF (one per print call)
  pdf(outfile, width = 6, height = 4)
  for (cond in names(ranked_list)) {
    p <- plotEnrichment(pathways[[pathway_name]], ranked_list[[cond]]) +
      ggtitle(paste0(cond, " | ", pathway_name)) +
      theme(plot.title = element_text(size = 8))
    print(p)
  }
  dev.off()
  message("Saved: ", outfile, " (", length(ranked_list), " pages)")
}

if (!is.na(kegg_ap_name)) {
  make_enrichment_panels(
    kegg_ap_name, ranked_list, pathways_curated,
    here("plots", "gsea_antigen_presentation_kegg_enrichment.pdf")
  )
}

if (!is.na(reactome_mhc2)) {
  make_enrichment_panels(
    reactome_mhc2, ranked_list, pathways_curated,
    here("plots", "gsea_antigen_presentation_reactome_mhc2_enrichment.pdf")
  )
}
```

# Gene-level TE LFC Heatmap: Macrophage Polarization & Antigen Presentation

TE LFC values for curated genes relevant to macrophage pro-inflammatory (M1), anti-inflammatory (M2), and antigen presentation programs. Rows annotated by category; columns are the same 7 conditions as the pathway heatmaps.

```{r gene_heatmap}
# --- Curated gene list by category ---
gene_categories <- c(
  # Immune checkpoint / immunosuppressive ligands
  CD274    = "Checkpoint", PDCD1LG2 = "Checkpoint", VSIG4  = "Checkpoint",
  LGALS9   = "Checkpoint", HAVCR2   = "Checkpoint",

  # M2 / anti-inflammatory macrophage markers
  ARG1   = "M2/Anti-inflam", MRC1   = "M2/Anti-inflam", CD163  = "M2/Anti-inflam",
  MSR1   = "M2/Anti-inflam", CHI3L3 = "M2/Anti-inflam", CHI3L1 = "M2/Anti-inflam",
  SPP1   = "M2/Anti-inflam", IL10   = "M2/Anti-inflam", TGFB1  = "M2/Anti-inflam",
  IL6    = "M2/Anti-inflam", CSF1   = "M2/Anti-inflam",

  # MHC antigen presentation machinery
  `HLA-DRA` = "MHC/APC", `HLA-DRB1` = "MHC/APC",
  `HLA-DPA1` = "MHC/APC", `HLA-DPB1` = "MHC/APC", `HLA-DQA1` = "MHC/APC",
  `HLA-A`  = "MHC/APC", `HLA-B`  = "MHC/APC", `HLA-C`  = "MHC/APC",
  B2M    = "MHC/APC", TAP1   = "MHC/APC", TAP2   = "MHC/APC",
  PSMB8  = "MHC/APC", PSMB9  = "MHC/APC", NLRC5  = "MHC/APC",

  # Lysosomal / endosomal antigen processing
  CTSB   = "Lysosomal", CTSD   = "Lysosomal", CTSL   = "Lysosomal",
  LAMP1  = "Lysosomal", LAMP2  = "Lysosomal", NPC2   = "Lysosomal"
)

gene_symbols_of_interest <- names(gene_categories)

# --- Extract te_lfc per condition ---
get_te_lfc <- function(df, condition_label) {
  df %>%
    filter(symbol %in% gene_symbols_of_interest, !is.na(te_lfc)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(te_lfc), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(symbol, te_lfc) %>%
    mutate(condition = condition_label)
}

gene_results <- bind_rows(
  get_te_lfc(te_hyp_1hr,   "Hyp 1hr"),
  get_te_lfc(te_hyp_4hr,   "Hyp 4hr"),
  get_te_lfc(te_hyp_24hr,  "Hyp 24hr"),
  get_te_lfc(te_si3d_norm, "si3d Norm"),
  get_te_lfc(te_si3d_hyp1, "si3d Hyp1"),
  get_te_lfc(te_si3d_hyp4, "si3d Hyp4"),
  get_te_lfc(te_si3d_hyp24,"si3d Hyp24")
)

# --- Build matrix ---
gene_col_order <- c("Hyp 1hr", "Hyp 4hr", "Hyp 24hr",
                    "si3d Norm", "si3d Hyp1", "si3d Hyp4", "si3d Hyp24")

gene_wide <- gene_results %>%
  pivot_wider(names_from = condition, values_from = te_lfc, values_fill = NA) %>%
  column_to_rownames("symbol")
for (cn in gene_col_order) if (!cn %in% colnames(gene_wide)) gene_wide[[cn]] <- NA
gene_wide <- gene_wide[, gene_col_order, drop = FALSE]

# Keep only genes present in at least one condition
genes_present <- rownames(gene_wide)[rowSums(!is.na(gene_wide)) > 0]
n_missing <- length(gene_symbols_of_interest) - length(genes_present)
if (n_missing > 0) {
  missing <- setdiff(gene_symbols_of_interest, genes_present)
  message(n_missing, " genes not found in TE data: ", paste(missing, collapse = ", "))
}
gene_wide <- gene_wide[genes_present, , drop = FALSE]

# Replace remaining NAs with 0 for display
gene_wide[is.na(gene_wide)] <- 0

# --- Row annotation by category ---
row_anno <- data.frame(
  Category = gene_categories[genes_present],
  row.names = genes_present
)

anno_colors <- list(
  Category = c(
    "Checkpoint"     = "#7b2d8b",
    "M2/Anti-inflam" = "#4575b4",
    "MHC/APC"        = "#1a9850",
    "Lysosomal"      = "#d97b00"
  )
)

# Order rows by category then alphabetically
row_order <- order(row_anno$Category, rownames(gene_wide))
gene_wide  <- gene_wide[row_order, , drop = FALSE]
row_anno   <- row_anno[row_order, , drop = FALSE]

gene_max_abs <- max(abs(gene_wide), na.rm = TRUE)
gene_max_abs <- max(gene_max_abs, 0.1)

p_genes <- pheatmap(
  gene_wide,
  cluster_rows    = FALSE,
  cluster_cols    = FALSE,
  scale           = "none",
  color           = colorRampPalette(c("blue", "white", "red"))(100),
  breaks          = seq(-gene_max_abs, gene_max_abs, length.out = 101),
  annotation_row  = row_anno,
  annotation_colors = anno_colors,
  main            = "TE LFC – Macrophage Polarization & Antigen Presentation Genes",
  fontsize_row    = 8,
  fontsize_col    = 9,
  cellwidth       = 38,
  cellheight      = 12,
  gaps_col        = 3,
  na_col          = "grey90"
)

outfile_genes <- here("plots", "gsea_macrophage_gene_heatmap.pdf")
pdf(outfile_genes, width = 14, height = max(5, nrow(gene_wide) * 0.2 + 3))
print(p_genes)
dev.off()
message("Saved: ", outfile_genes, " (", nrow(gene_wide), " genes)")
```

```{r}

```