---
title: "GSEA: eIF3d KD vs siCtrl – KEGG, GO:BP, Reactome"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

fgsea analysis of the eIF3d knockdown effect (si3d vs siCtrl) ranked by TE LFC. Conditions: normoxia 1&4hr (time-regressed), hypoxia 1hr, 4hr, and 24hr. Pathway databases: KEGG, GO:BP, and Reactome.

# Load Libraries

```{r load_libraries}
library(fgsea)
library(msigdbr)
library(pheatmap)
library(tidyverse)
library(here)
```

# Load Data

```{r load_data}
te_norm  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
te_hyp1  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
te_hyp4  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
te_hyp24 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# Sanity Checks

```{r sanity_checks}
summarise_te <- function(df, label) {
  sig <- df %>% filter(te_padj < 0.05, abs(te_lfc) > 0.5, !is.na(symbol))
  cat(label, "\n")
  cat("  Total genes:", nrow(df), "\n")
  cat("  With symbol:", sum(!is.na(df$symbol)), "\n")
  cat("  TE sig (padj < 0.05 & |lfc| > 0.5):", nrow(sig),
      "– up:", sum(sig$te_lfc > 0.5),
      "| down:", sum(sig$te_lfc < -0.5), "\n\n")
}

summarise_te(te_norm,  "Normoxia 1&4hr (time-regressed)")
summarise_te(te_hyp1,  "Hypoxia 1hr")
summarise_te(te_hyp4,  "Hypoxia 4hr")
summarise_te(te_hyp24, "Hypoxia 24hr")
```

# Build Ranked Lists

```{r ranked_lists}
make_ranked <- function(df) {
  df <- df %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(te_lfc), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranked <- df$te_lfc
  names(ranked) <- df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked[!duplicated(names(ranked))]
}

ranked_norm  <- make_ranked(te_norm)
ranked_hyp1  <- make_ranked(te_hyp1)
ranked_hyp4  <- make_ranked(te_hyp4)
ranked_hyp24 <- make_ranked(te_hyp24)

cat("Ranked list sizes: Norm 1&4hr =", length(ranked_norm),
    "| Hyp 1hr =", length(ranked_hyp1),
    "| Hyp 4hr =", length(ranked_hyp4),
    "| Hyp 24hr =", length(ranked_hyp24), "\n")
```

# Load Pathway Databases

```{r load_pathways}
to_pathways <- function(db) {
  db %>%
    dplyr::select(gs_name, gene_symbol) %>%
    group_by(gs_name) %>%
    summarise(genes = list(gene_symbol), .groups = "drop") %>%
    deframe()
}

pathways_kegg     <- to_pathways(msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY"))
pathways_gobp     <- to_pathways(msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP"))
pathways_reactome <- to_pathways(msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME"))
pathways_hallmark <- to_pathways(msigdbr(species = "Homo sapiens", category = "H"))

cat("Pathway counts – KEGG:", length(pathways_kegg),
    "| GO:BP:", length(pathways_gobp),
    "| Reactome:", length(pathways_reactome),
    "| Hallmark:", length(pathways_hallmark), "\n")
```

# Run fgsea

```{r run_fgsea}
run_fgsea <- function(ranked, pathways, condition_label) {
  set.seed(9)
  res <- fgsea(
    pathways  = pathways,
    stats     = ranked,
    minSize   = 10,
    maxSize   = 500,
    gseaParam = 1,
    nproc     = 1
  )
  res %>%
    as_tibble() %>%
    dplyr::select(pathway, NES, pval, padj, leadingEdge) %>%
    mutate(timepoint = condition_label)
}

# KEGG
fgsea_kegg_norm  <- run_fgsea(ranked_norm,  pathways_kegg, "Norm 1&4hr")
fgsea_kegg_hyp1  <- run_fgsea(ranked_hyp1,  pathways_kegg, "Hyp 1hr")
fgsea_kegg_hyp4  <- run_fgsea(ranked_hyp4,  pathways_kegg, "Hyp 4hr")
fgsea_kegg_hyp24 <- run_fgsea(ranked_hyp24, pathways_kegg, "Hyp 24hr")

# GO:BP
fgsea_gobp_norm  <- run_fgsea(ranked_norm,  pathways_gobp, "Norm 1&4hr")
fgsea_gobp_hyp1  <- run_fgsea(ranked_hyp1,  pathways_gobp, "Hyp 1hr")
fgsea_gobp_hyp4  <- run_fgsea(ranked_hyp4,  pathways_gobp, "Hyp 4hr")
fgsea_gobp_hyp24 <- run_fgsea(ranked_hyp24, pathways_gobp, "Hyp 24hr")

# Reactome
fgsea_reactome_norm  <- run_fgsea(ranked_norm,  pathways_reactome, "Norm 1&4hr")
fgsea_reactome_hyp1  <- run_fgsea(ranked_hyp1,  pathways_reactome, "Hyp 1hr")
fgsea_reactome_hyp4  <- run_fgsea(ranked_hyp4,  pathways_reactome, "Hyp 4hr")
fgsea_reactome_hyp24 <- run_fgsea(ranked_hyp24, pathways_reactome, "Hyp 24hr")

# Hallmark
fgsea_hallmark_norm  <- run_fgsea(ranked_norm,  pathways_hallmark, "Norm 1&4hr")
fgsea_hallmark_hyp1  <- run_fgsea(ranked_hyp1,  pathways_hallmark, "Hyp 1hr")
fgsea_hallmark_hyp4  <- run_fgsea(ranked_hyp4,  pathways_hallmark, "Hyp 4hr")
fgsea_hallmark_hyp24 <- run_fgsea(ranked_hyp24, pathways_hallmark, "Hyp 24hr")
```

# Heatmap Helper

```{r heatmap_helper}
make_gsea_heatmap <- function(results_list, title, outfile, up_only = FALSE) {
  combined <- bind_rows(results_list)

  col_order <- c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

  sig_pathways <- combined %>%
    filter(padj < 0.05) %>%
    pull(pathway) %>%
    unique()

  if (up_only) {
    up_pathways <- combined %>%
      filter(padj < 0.05, NES > 0) %>%
      pull(pathway) %>%
      unique()
    sig_pathways <- intersect(sig_pathways, up_pathways)
  }

  if (length(sig_pathways) == 0) {
    message("No significant pathways (padj < 0.05) for: ", title)
    return(invisible(NULL))
  }

  wide <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    dplyr::select(pathway, NES, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
    dplyr::select(pathway, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(wide), na.rm = TRUE)
  max_abs <- max(max_abs, 0.1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("blue", "white", "red"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = title,
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(outfile, width = 10, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()

  message("Saved: ", outfile, " (", length(sig_pathways), " pathways)")
  invisible(p)
}
```

# Hallmark Heatmap

```{r heatmap_hallmark}
make_gsea_heatmap(
  list(fgsea_hallmark_norm, fgsea_hallmark_hyp1, fgsea_hallmark_hyp4, fgsea_hallmark_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – Hallmark (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_hallmark.pdf")
)
```

# KEGG Heatmap

```{r heatmap_kegg}
make_gsea_heatmap(
  list(fgsea_kegg_norm, fgsea_kegg_hyp1, fgsea_kegg_hyp4, fgsea_kegg_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – KEGG (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_kegg.pdf")
)
```

# GO:BP Heatmap

```{r heatmap_gobp}
make_gsea_heatmap(
  list(fgsea_gobp_norm, fgsea_gobp_hyp1, fgsea_gobp_hyp4, fgsea_gobp_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – GO:BP (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_gobp.pdf")
)
```

## GO:BP – Up Only

```{r heatmap_gobp_up}
make_gsea_heatmap(
  list(fgsea_gobp_norm, fgsea_gobp_hyp1, fgsea_gobp_hyp4, fgsea_gobp_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – GO:BP, up-regulated only (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_gobp_up.pdf"),
  up_only = TRUE
)
```

# Reactome Heatmap

```{r heatmap_reactome}
make_gsea_heatmap(
  list(fgsea_reactome_norm, fgsea_reactome_hyp1, fgsea_reactome_hyp4, fgsea_reactome_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – Reactome (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_reactome.pdf")
)
```

## Reactome – Up Only

```{r heatmap_reactome_up}
make_gsea_heatmap(
  list(fgsea_reactome_norm, fgsea_reactome_hyp1, fgsea_reactome_hyp4, fgsea_reactome_hyp24),
  title   = "TE GSEA – siCtrl vs si3d – Reactome, up-regulated only (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_reactome_up.pdf"),
  up_only = TRUE
)
```

# ORA – Genes Promoted by eIF3d (TE LFC > 0.5, padj < 0.05)

## Define gene sets and print counts

```{r ora_gene_sets}
up_norm  <- te_norm  %>% filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol) %>% unique()
up_hyp1  <- te_hyp1  %>% filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol) %>% unique()
up_hyp4  <- te_hyp4  %>% filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol) %>% unique()
up_hyp24 <- te_hyp24 %>% filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol) %>% unique()

cat("eIF3d-promoted genes (TE LFC > 0.5, padj < 0.05):\n")
cat("  Norm 1&4hr:", length(up_norm),  "\n")
cat("  Hyp 1hr:   ", length(up_hyp1),  "\n")
cat("  Hyp 4hr:   ", length(up_hyp4),  "\n")
cat("  Hyp 24hr:  ", length(up_hyp24), "\n")
```

## Run ORA

```{r run_ora}
universe <- unique(c(
  te_norm  %>% filter(!is.na(symbol)) %>% pull(symbol),
  te_hyp1  %>% filter(!is.na(symbol)) %>% pull(symbol),
  te_hyp4  %>% filter(!is.na(symbol)) %>% pull(symbol),
  te_hyp24 %>% filter(!is.na(symbol)) %>% pull(symbol)
))

run_ora <- function(genes, pathways, condition_label) {
  fora(pathways = pathways, genes = genes, universe = universe,
       minSize = 10, maxSize = 500) %>%
    as_tibble() %>%
    mutate(
      timepoint        = condition_label,
      enrichment_ratio = (overlap / length(genes)) / (size / length(universe)),
      log2_er          = log2(enrichment_ratio)
    ) %>%
    dplyr::select(pathway, pval, padj, overlap, size, enrichment_ratio, log2_er, timepoint)
}

ora_hallmark_norm  <- run_ora(up_norm,  pathways_hallmark, "Norm 1&4hr")
ora_hallmark_hyp1  <- run_ora(up_hyp1,  pathways_hallmark, "Hyp 1hr")
ora_hallmark_hyp4  <- run_ora(up_hyp4,  pathways_hallmark, "Hyp 4hr")
ora_hallmark_hyp24 <- run_ora(up_hyp24, pathways_hallmark, "Hyp 24hr")

ora_kegg_norm  <- run_ora(up_norm,  pathways_kegg, "Norm 1&4hr")
ora_kegg_hyp1  <- run_ora(up_hyp1,  pathways_kegg, "Hyp 1hr")
ora_kegg_hyp4  <- run_ora(up_hyp4,  pathways_kegg, "Hyp 4hr")
ora_kegg_hyp24 <- run_ora(up_hyp24, pathways_kegg, "Hyp 24hr")
```

## ORA Heatmap Helper

```{r ora_heatmap_helper}
make_ora_heatmap <- function(results_list, title, outfile) {
  combined <- bind_rows(results_list)

  sig_pathways <- combined %>%
    filter(padj < 0.05) %>%
    pull(pathway) %>%
    unique()

  if (length(sig_pathways) == 0) {
    message("No significant pathways (padj < 0.05) for: ", title)
    return(invisible(NULL))
  }

  col_order <- c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

  wide <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    dplyr::select(pathway, log2_er, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = log2_er, values_fill = 0) %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
    dplyr::select(pathway, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(as.matrix(wide)), na.rm = TRUE) + 0.01
  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("blue", "white", "red"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = title,
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(outfile, width = 10, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()

  message("Saved: ", outfile, " (", length(sig_pathways), " pathways)")
  invisible(p)
}
```

## ORA Heatmaps

```{r ora_heatmap_hallmark}
make_ora_heatmap(
  list(ora_hallmark_norm, ora_hallmark_hyp1, ora_hallmark_hyp4, ora_hallmark_hyp24),
  title   = "ORA – eIF3d-promoted genes – Hallmark (* = padj < 0.05)",
  outfile = here("plots", "ora_si3d_promoted_hallmark.pdf")
)
```

```{r ora_heatmap_kegg}
make_ora_heatmap(
  list(ora_kegg_norm, ora_kegg_hyp1, ora_kegg_hyp4, ora_kegg_hyp24),
  title   = "ORA – eIF3d-promoted genes – KEGG (* = padj < 0.05)",
  outfile = here("plots", "ora_si3d_promoted_kegg.pdf")
)
```

# Positive Control Check

```{r positive_control}
# HALLMARK_HYPOXIA: si3d KD at hypoxia timepoints should show enrichment if
# eIF3d promotes translation of hypoxia-response genes (expect NES > 0 or consistent direction)
# HALLMARK_EMT: primary pathway of interest for leading edge extraction
bind_rows(fgsea_hallmark_norm, fgsea_hallmark_hyp1, fgsea_hallmark_hyp4, fgsea_hallmark_hyp24) %>%
  filter(pathway %in% c("HALLMARK_HYPOXIA", "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")) %>%
  dplyr::select(pathway, timepoint, NES, padj) %>%
  arrange(pathway, timepoint)
```

# EMT Leading Edge Genes

```{r emt_leading_edge}
extract_emt_leading_edge <- function(fgsea_list, db_label) {
  bind_rows(fgsea_list) %>%
    filter(grepl("EMT|EPITHELIAL_MESENCHYMAL", pathway, ignore.case = TRUE),
           !is.na(leadingEdge)) %>%
    transmute(pathway, NES, padj, timepoint, database = db_label,
              gene = leadingEdge) %>%
    unnest(gene)
}

emt_si3d <- bind_rows(
  extract_emt_leading_edge(list(fgsea_hallmark_norm, fgsea_hallmark_hyp1, fgsea_hallmark_hyp4, fgsea_hallmark_hyp24), "Hallmark"),
  extract_emt_leading_edge(list(fgsea_kegg_norm,     fgsea_kegg_hyp1,     fgsea_kegg_hyp4,     fgsea_kegg_hyp24),     "KEGG")
)

write_csv(emt_si3d, here("output", "gsea_emt_leading_edge_si3d_vs_sictrl.csv"))

emt_si3d %>%
  count(database, timepoint, pathway, NES, padj) %>%
  arrange(database, timepoint) %>%
  print(n = Inf)
```

# EMT Leading Edge – Genes per Condition

```{r emt_genes_per_timepoint}
for (tp in c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")) {
  genes <- emt_si3d %>%
    filter(database == "Hallmark", timepoint == tp) %>%
    pull(gene) %>%
    sort()
  cat("siCtrl vs si3d –", tp, "\n")
  cat(paste(genes, collapse = ", "))
  cat("\n\n")
}
```

# EMT Leading Edge – Overlap with Hypoxic Response (Timepoint-Matched)

```{r emt_overlap_hypoxia}
emt_hypoxia_resp <- read_csv(here("output", "gsea_emt_leading_edge_hypoxia_vs_normoxia.csv"))

tp_map <- c("Hyp 1hr" = "1hr", "Hyp 4hr" = "4hr", "Hyp 24hr" = "24hr")

for (tp_si3d in names(tp_map)) {
  tp_hyp <- tp_map[[tp_si3d]]

  genes_si3d <- emt_si3d %>%
    filter(database == "Hallmark", timepoint == tp_si3d) %>%
    pull(gene)

  genes_hyp <- emt_hypoxia_resp %>%
    filter(database == "Hallmark", timepoint == tp_hyp) %>%
    pull(gene)

  overlap <- sort(intersect(genes_si3d, genes_hyp))

  cat(tp_hyp, "hypoxic response ∩ siCtrl vs si3d", tp_si3d, "\n")
  cat("n =", length(overlap), "\n")
  cat(paste(overlap, collapse = ", "))
  cat("\n\n")
}
```

# EMT Leading Edge – Promoted by eIF3d (TE LFC > 0.5, padj < 0.05)

```{r emt_promoted_by_3d}
emt_genes_hallmark <- emt_si3d %>%
  filter(database == "Hallmark") %>%
  pull(gene) %>%
  unique()

conditions <- list(
  "Norm 1&4hr" = te_norm,
  "Hyp 1hr"    = te_hyp1,
  "Hyp 4hr"    = te_hyp4,
  "Hyp 24hr"   = te_hyp24
)

emt_promoted <- map_dfr(names(conditions), function(cond) {
  conditions[[cond]] %>%
    filter(symbol %in% emt_genes_hallmark, te_lfc > 0.5, te_padj < 0.05) %>%
    mutate(condition = cond) %>%
    dplyr::select(symbol, te_lfc, te_padj, condition)
})

print(emt_promoted %>% arrange(condition, desc(te_lfc)), n = Inf)
```

# OXPHOS Leading Edge – LFC and padj by Condition

```{r oxphos_leading_edge}
conditions <- list(
  "Norm 1&4hr" = te_norm,
  "Hyp 1hr"    = te_hyp1,
  "Hyp 4hr"    = te_hyp4,
  "Hyp 24hr"   = te_hyp24
)

extract_oxphos_leading_edge <- function(fgsea_list, db_label) {
  bind_rows(fgsea_list) %>%
    filter(grepl("OXIDATIVE_PHOSPHORYLATION", pathway, ignore.case = TRUE),
           !is.na(leadingEdge)) %>%
    transmute(pathway, NES, padj, timepoint, database = db_label,
              gene = leadingEdge) %>%
    unnest(gene)
}

oxphos_si3d <- bind_rows(
  extract_oxphos_leading_edge(list(fgsea_hallmark_norm, fgsea_hallmark_hyp1, fgsea_hallmark_hyp4, fgsea_hallmark_hyp24), "Hallmark"),
  extract_oxphos_leading_edge(list(fgsea_kegg_norm,     fgsea_kegg_hyp1,     fgsea_kegg_hyp4,     fgsea_kegg_hyp24),     "KEGG")
)

oxphos_si3d %>%
  count(database, timepoint, pathway, NES, padj) %>%
  arrange(database, timepoint) %>%
  print(n = Inf)

for (db in c("Hallmark", "KEGG")) {
  for (cond in c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")) {
    genes <- oxphos_si3d %>%
      filter(database == db, timepoint == cond) %>%
      pull(gene)
    if (length(genes) == 0) next

    result <- conditions[[cond]] %>%
      filter(symbol %in% genes) %>%
      dplyr::select(symbol, te_lfc, te_padj) %>%
      arrange(desc(te_lfc))

    cat("\n", db, "OXPHOS –", cond, "\n")
    print(result, n = Inf)
  }
}
```
