---
title: "GSEA: eIF3d KD vs siCtrl – KEGG, GO:BP, Reactome"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

fgsea analysis of the eIF3d knockdown effect (si3d vs siCtrl) ranked by TE LFC. Conditions: normoxia 1&4hr (time-regressed), hypoxia 1hr, 4hr, and 24hr. Pathway databases: KEGG, GO:BP, and Reactome.

# Load Libraries

```{r load_libraries}
library(fgsea)
library(msigdbr)
library(pheatmap)
library(tidyverse)
library(here)
```

# Load Data

```{r load_data}
te_norm  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
te_hyp1  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
te_hyp4  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
te_hyp24 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))
```

# Sanity Checks

```{r sanity_checks}
summarise_te <- function(df, label) {
  sig <- df %>% filter(te_padj < 0.05, abs(te_lfc) > 0.5, !is.na(symbol))
  cat(label, "\n")
  cat("  Total genes:", nrow(df), "\n")
  cat("  With symbol:", sum(!is.na(df$symbol)), "\n")
  cat("  TE sig (padj < 0.05 & |lfc| > 0.5):", nrow(sig),
      "– up:", sum(sig$te_lfc > 0.5),
      "| down:", sum(sig$te_lfc < -0.5), "\n\n")
}

summarise_te(te_norm,  "Normoxia 1&4hr (time-regressed)")
summarise_te(te_hyp1,  "Hypoxia 1hr")
summarise_te(te_hyp4,  "Hypoxia 4hr")
summarise_te(te_hyp24, "Hypoxia 24hr")
```

# Build Ranked Lists

```{r ranked_lists}
make_ranked <- function(df) {
  df <- df %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(te_lfc), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranked <- df$te_lfc
  names(ranked) <- df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked[!duplicated(names(ranked))]
}

ranked_norm  <- make_ranked(te_norm)
ranked_hyp1  <- make_ranked(te_hyp1)
ranked_hyp4  <- make_ranked(te_hyp4)
ranked_hyp24 <- make_ranked(te_hyp24)

cat("Ranked list sizes: Norm 1&4hr =", length(ranked_norm),
    "| Hyp 1hr =", length(ranked_hyp1),
    "| Hyp 4hr =", length(ranked_hyp4),
    "| Hyp 24hr =", length(ranked_hyp24), "\n")
```

# Load Pathway Databases

```{r load_pathways}
to_pathways <- function(db) {
  db %>%
    dplyr::select(gs_name, gene_symbol) %>%
    group_by(gs_name) %>%
    summarise(genes = list(gene_symbol), .groups = "drop") %>%
    deframe()
}

pathways_kegg     <- to_pathways(msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY"))
pathways_gobp     <- to_pathways(msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP"))
pathways_reactome <- to_pathways(msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME"))

cat("Pathway counts – KEGG:", length(pathways_kegg),
    "| GO:BP:", length(pathways_gobp),
    "| Reactome:", length(pathways_reactome), "\n")
```

# Run fgsea

```{r run_fgsea}
run_fgsea <- function(ranked, pathways, condition_label) {
  set.seed(9)
  res <- fgsea(
    pathways  = pathways,
    stats     = ranked,
    minSize   = 10,
    maxSize   = 500,
    gseaParam = 0,
    nproc     = 1
  )
  res %>%
    as_tibble() %>%
    dplyr::select(pathway, NES, pval, padj) %>%
    mutate(timepoint = condition_label)
}

# KEGG
fgsea_kegg_norm  <- run_fgsea(ranked_norm,  pathways_kegg, "Norm 1&4hr")
fgsea_kegg_hyp1  <- run_fgsea(ranked_hyp1,  pathways_kegg, "Hyp 1hr")
fgsea_kegg_hyp4  <- run_fgsea(ranked_hyp4,  pathways_kegg, "Hyp 4hr")
fgsea_kegg_hyp24 <- run_fgsea(ranked_hyp24, pathways_kegg, "Hyp 24hr")

# GO:BP
fgsea_gobp_norm  <- run_fgsea(ranked_norm,  pathways_gobp, "Norm 1&4hr")
fgsea_gobp_hyp1  <- run_fgsea(ranked_hyp1,  pathways_gobp, "Hyp 1hr")
fgsea_gobp_hyp4  <- run_fgsea(ranked_hyp4,  pathways_gobp, "Hyp 4hr")
fgsea_gobp_hyp24 <- run_fgsea(ranked_hyp24, pathways_gobp, "Hyp 24hr")

# Reactome
fgsea_reactome_norm  <- run_fgsea(ranked_norm,  pathways_reactome, "Norm 1&4hr")
fgsea_reactome_hyp1  <- run_fgsea(ranked_hyp1,  pathways_reactome, "Hyp 1hr")
fgsea_reactome_hyp4  <- run_fgsea(ranked_hyp4,  pathways_reactome, "Hyp 4hr")
fgsea_reactome_hyp24 <- run_fgsea(ranked_hyp24, pathways_reactome, "Hyp 24hr")
```

# Heatmap Helper

```{r heatmap_helper}
make_gsea_heatmap <- function(results_list, title, outfile, up_only = FALSE) {
  combined <- bind_rows(results_list)

  col_order <- c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

  sig_pathways <- combined %>%
    filter(padj < 0.05) %>%
    pull(pathway) %>%
    unique()

  if (up_only) {
    up_pathways <- combined %>%
      filter(padj < 0.05, NES > 0) %>%
      pull(pathway) %>%
      unique()
    sig_pathways <- intersect(sig_pathways, up_pathways)
  }

  if (length(sig_pathways) == 0) {
    message("No significant pathways (padj < 0.05) for: ", title)
    return(invisible(NULL))
  }

  wide <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    dplyr::select(pathway, NES, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
    dplyr::select(pathway, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(wide), na.rm = TRUE)
  max_abs <- max(max_abs, 0.1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("blue", "white", "red"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = title,
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(outfile, width = 10, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()

  message("Saved: ", outfile, " (", length(sig_pathways), " pathways)")
  invisible(p)
}
```

# KEGG Heatmap

```{r heatmap_kegg}
make_gsea_heatmap(
  list(fgsea_kegg_norm, fgsea_kegg_hyp1, fgsea_kegg_hyp4, fgsea_kegg_hyp24),
  title   = "TE GSEA – si3d vs siCtrl – KEGG (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_kegg.pdf")
)
```

# GO:BP Heatmap

```{r heatmap_gobp}
make_gsea_heatmap(
  list(fgsea_gobp_norm, fgsea_gobp_hyp1, fgsea_gobp_hyp4, fgsea_gobp_hyp24),
  title   = "TE GSEA – si3d vs siCtrl – GO:BP (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_gobp.pdf")
)
```

## GO:BP – Up Only

```{r heatmap_gobp_up}
make_gsea_heatmap(
  list(fgsea_gobp_norm, fgsea_gobp_hyp1, fgsea_gobp_hyp4, fgsea_gobp_hyp24),
  title   = "TE GSEA – si3d vs siCtrl – GO:BP, up-regulated only (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_gobp_up.pdf"),
  up_only = TRUE
)
```

# Reactome Heatmap

```{r heatmap_reactome}
make_gsea_heatmap(
  list(fgsea_reactome_norm, fgsea_reactome_hyp1, fgsea_reactome_hyp4, fgsea_reactome_hyp24),
  title   = "TE GSEA – si3d vs siCtrl – Reactome (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_reactome.pdf")
)
```

## Reactome – Up Only

```{r heatmap_reactome_up}
make_gsea_heatmap(
  list(fgsea_reactome_norm, fgsea_reactome_hyp1, fgsea_reactome_hyp4, fgsea_reactome_hyp24),
  title   = "TE GSEA – si3d vs siCtrl – Reactome, up-regulated only (* = padj < 0.05)",
  outfile = here("plots", "gsea_si3d_reactome_up.pdf"),
  up_only = TRUE
)
```
