---
title: "GSEA Analysis - eIF3e and eIF3d Knockdowns"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This analysis performs Gene Set Enrichment Analysis (GSEA) on RNA-seq and Translation Efficiency (TE) changes in eIF3e and eIF3d knockdown cells under normoxia and hypoxia conditions.

**Comparisons:**
- sictrl vs si3e in normoxia
- sictrl vs si3e in hypoxia
- sictrl vs si3d in normoxia
- sictrl vs si3d in hypoxia

**Timepoints:** 1hr, 4hr, 24hr

**Data types:** RNA-seq log2 fold changes and TE log2 fold changes

```{r load_libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(pheatmap)
library(writexl)
library(enrichplot)
library(patchwork)

cat("Libraries loaded successfully\n")
```

# Load Hallmark Gene Sets

```{r load_hallmark}
# Load MSigDB Hallmark gene sets for human
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# Create gene set format for clusterProfiler (TERM2GENE)
hallmark_rna <- hallmark_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark_rna$gs_name)), "\n")
```

# Load KEGG Gene Sets

```{r load_kegg}
# Load MSigDB KEGG gene sets for human
kegg_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY")

# Create gene set format for clusterProfiler (TERM2GENE)
kegg_rna <- kegg_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB KEGG gene sets:\n")
cat("Number of pathways:", length(unique(kegg_rna$gs_name)), "\n")
```

# Load RNA and TE Results

## si3e Normoxia (sictrl vs si3e in normoxia)

```{r load_si3e_norm}
# Load RNA results
rna_si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3e Normoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3e_norm_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3e_norm_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3e_norm_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3e_norm_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3e_norm_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3e_norm_24hr), "genes\n")
```

## si3e Hypoxia (sictrl vs si3e in hypoxia)

```{r load_si3e_hyp}
# Load RNA results
rna_si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3e Hypoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3e_hyp_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3e_hyp_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3e_hyp_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3e_hyp_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3e_hyp_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3e_hyp_24hr), "genes\n")
```

## si3d Normoxia (sictrl vs si3d in normoxia)

```{r load_si3d_norm}
# Load RNA results
rna_si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3d Normoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3d_norm_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3d_norm_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3d_norm_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3d_norm_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3d_norm_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3d_norm_24hr), "genes\n")
```

## si3d Hypoxia (sictrl vs si3d in hypoxia)

```{r load_si3d_hyp}
# Load RNA results
rna_si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3d Hypoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3d_hyp_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3d_hyp_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3d_hyp_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3d_hyp_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3d_hyp_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3d_hyp_24hr), "genes\n")
```

# GSEA: RNA Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA RNA si3e Normoxia - 1hr

```{r gsea_rna_si3e_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_1hr_results[] <- lapply(gsea_rna_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 4hr

```{r gsea_rna_si3e_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_4hr_results[] <- lapply(gsea_rna_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 24hr

```{r gsea_rna_si3e_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_24hr_results[] <- lapply(gsea_rna_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia RNA GSEA results

```{r export_gsea_rna_si3e_norm}
gsea_rna_si3e_norm_export <- list(
  "GSEA_RNA_si3e_norm_1hr" = gsea_rna_si3e_norm_1hr_results,
  "GSEA_RNA_si3e_norm_4hr" = gsea_rna_si3e_norm_4hr_results,
  "GSEA_RNA_si3e_norm_24hr" = gsea_rna_si3e_norm_24hr_results)

write_xlsx(gsea_rna_si3e_norm_export, here("output/gsea_rna_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia RNA GSEA results exported\n")
```

## si3e Normoxia RNA GSEA Heatmap

```{r gsea_rna_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_norm_1hr_results) > 0) gsea_rna_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_4hr_results) > 0) gsea_rna_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_24hr_results) > 0) gsea_rna_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA RNA si3e Hypoxia - 1hr

```{r gsea_rna_si3e_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_1hr_results[] <- lapply(gsea_rna_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 4hr

```{r gsea_rna_si3e_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_4hr_results[] <- lapply(gsea_rna_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 24hr

```{r gsea_rna_si3e_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_24hr_results[] <- lapply(gsea_rna_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia RNA GSEA results

```{r export_gsea_rna_si3e_hyp}
gsea_rna_si3e_hyp_export <- list(
  "GSEA_RNA_si3e_hyp_1hr" = gsea_rna_si3e_hyp_1hr_results,
  "GSEA_RNA_si3e_hyp_4hr" = gsea_rna_si3e_hyp_4hr_results,
  "GSEA_RNA_si3e_hyp_24hr" = gsea_rna_si3e_hyp_24hr_results)

write_xlsx(gsea_rna_si3e_hyp_export, here("output/gsea_rna_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia RNA GSEA results exported\n")
```

## si3e Hypoxia RNA GSEA Heatmap

```{r gsea_rna_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_hyp_1hr_results) > 0) gsea_rna_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_4hr_results) > 0) gsea_rna_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_24hr_results) > 0) gsea_rna_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA RNA si3d Normoxia - 1hr

```{r gsea_rna_si3d_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_1hr_results[] <- lapply(gsea_rna_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 4hr

```{r gsea_rna_si3d_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_4hr_results[] <- lapply(gsea_rna_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 24hr

```{r gsea_rna_si3d_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_24hr_results[] <- lapply(gsea_rna_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia RNA GSEA results

```{r export_gsea_rna_si3d_norm}
gsea_rna_si3d_norm_export <- list(
  "GSEA_RNA_si3d_norm_1hr" = gsea_rna_si3d_norm_1hr_results,
  "GSEA_RNA_si3d_norm_4hr" = gsea_rna_si3d_norm_4hr_results,
  "GSEA_RNA_si3d_norm_24hr" = gsea_rna_si3d_norm_24hr_results)

write_xlsx(gsea_rna_si3d_norm_export, here("output/gsea_rna_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia RNA GSEA results exported\n")
```

## si3d Normoxia RNA GSEA Heatmap

```{r gsea_rna_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_norm_1hr_results) > 0) gsea_rna_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_4hr_results) > 0) gsea_rna_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_24hr_results) > 0) gsea_rna_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA RNA si3d Hypoxia - 1hr

```{r gsea_rna_si3d_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_1hr_results[] <- lapply(gsea_rna_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 4hr

```{r gsea_rna_si3d_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_4hr_results[] <- lapply(gsea_rna_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 24hr

```{r gsea_rna_si3d_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_24hr_results[] <- lapply(gsea_rna_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia RNA GSEA results

```{r export_gsea_rna_si3d_hyp}
gsea_rna_si3d_hyp_export <- list(
  "GSEA_RNA_si3d_hyp_1hr" = gsea_rna_si3d_hyp_1hr_results,
  "GSEA_RNA_si3d_hyp_4hr" = gsea_rna_si3d_hyp_4hr_results,
  "GSEA_RNA_si3d_hyp_24hr" = gsea_rna_si3d_hyp_24hr_results)

write_xlsx(gsea_rna_si3d_hyp_export, here("output/gsea_rna_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia RNA GSEA results exported\n")
```

## si3d Hypoxia RNA GSEA Heatmap

```{r gsea_rna_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_hyp_1hr_results) > 0) gsea_rna_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_4hr_results) > 0) gsea_rna_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_24hr_results) > 0) gsea_rna_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA TE si3e Normoxia - 1hr

```{r gsea_te_si3e_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_1hr_results[] <- lapply(gsea_te_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 4hr

```{r gsea_te_si3e_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_4hr_results[] <- lapply(gsea_te_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 24hr

```{r gsea_te_si3e_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_24hr_results[] <- lapply(gsea_te_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia TE GSEA results

```{r export_gsea_te_si3e_norm}
gsea_te_si3e_norm_export <- list(
  "GSEA_TE_si3e_norm_1hr" = gsea_te_si3e_norm_1hr_results,
  "GSEA_TE_si3e_norm_4hr" = gsea_te_si3e_norm_4hr_results,
  "GSEA_TE_si3e_norm_24hr" = gsea_te_si3e_norm_24hr_results)

write_xlsx(gsea_te_si3e_norm_export, here("output/gsea_te_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia TE GSEA results exported\n")
```

## si3e Normoxia TE GSEA Heatmap

```{r gsea_te_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_norm_1hr_results) > 0) gsea_te_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_4hr_results) > 0) gsea_te_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_24hr_results) > 0) gsea_te_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA TE si3e Hypoxia - 1hr

```{r gsea_te_si3e_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_1hr_results[] <- lapply(gsea_te_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 4hr

```{r gsea_te_si3e_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_4hr_results[] <- lapply(gsea_te_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 24hr

```{r gsea_te_si3e_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_24hr_results[] <- lapply(gsea_te_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia TE GSEA results

```{r export_gsea_te_si3e_hyp}
gsea_te_si3e_hyp_export <- list(
  "GSEA_TE_si3e_hyp_1hr" = gsea_te_si3e_hyp_1hr_results,
  "GSEA_TE_si3e_hyp_4hr" = gsea_te_si3e_hyp_4hr_results,
  "GSEA_TE_si3e_hyp_24hr" = gsea_te_si3e_hyp_24hr_results)

write_xlsx(gsea_te_si3e_hyp_export, here("output/gsea_te_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia TE GSEA results exported\n")
```

## si3e Hypoxia TE GSEA Heatmap

```{r gsea_te_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_hyp_1hr_results) > 0) gsea_te_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_4hr_results) > 0) gsea_te_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_24hr_results) > 0) gsea_te_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA TE si3d Normoxia - 1hr

```{r gsea_te_si3d_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_1hr_results[] <- lapply(gsea_te_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 4hr

```{r gsea_te_si3d_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_4hr_results[] <- lapply(gsea_te_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 24hr

```{r gsea_te_si3d_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_24hr_results[] <- lapply(gsea_te_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia TE GSEA results

```{r export_gsea_te_si3d_norm}
gsea_te_si3d_norm_export <- list(
  "GSEA_TE_si3d_norm_1hr" = gsea_te_si3d_norm_1hr_results,
  "GSEA_TE_si3d_norm_4hr" = gsea_te_si3d_norm_4hr_results,
  "GSEA_TE_si3d_norm_24hr" = gsea_te_si3d_norm_24hr_results)

write_xlsx(gsea_te_si3d_norm_export, here("output/gsea_te_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia TE GSEA results exported\n")
```

## si3d Normoxia TE GSEA Heatmap

```{r gsea_te_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_norm_1hr_results) > 0) gsea_te_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_4hr_results) > 0) gsea_te_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_24hr_results) > 0) gsea_te_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA TE si3d Hypoxia - 1hr

```{r gsea_te_si3d_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_1hr_results[] <- lapply(gsea_te_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 4hr

```{r gsea_te_si3d_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_4hr_results[] <- lapply(gsea_te_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 24hr

```{r gsea_te_si3d_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_24hr_results[] <- lapply(gsea_te_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia TE GSEA results

```{r export_gsea_te_si3d_hyp}
gsea_te_si3d_hyp_export <- list(
  "GSEA_TE_si3d_hyp_1hr" = gsea_te_si3d_hyp_1hr_results,
  "GSEA_TE_si3d_hyp_4hr" = gsea_te_si3d_hyp_4hr_results,
  "GSEA_TE_si3d_hyp_24hr" = gsea_te_si3d_hyp_24hr_results)

write_xlsx(gsea_te_si3d_hyp_export, here("output/gsea_te_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia TE GSEA results exported\n")
```

## si3d Hypoxia TE GSEA Heatmap

```{r gsea_te_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_hyp_1hr_results) > 0) gsea_te_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_4hr_results) > 0) gsea_te_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_24hr_results) > 0) gsea_te_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```
# Hallmark Hypoxia Pathway - Running Enrichment Plots

This section creates classic GSEA running enrichment plots specifically for the HALLMARK_HYPOXIA pathway, showing TE-based enrichment/depletion across treatments and timepoints.

## Normoxia - TE Changes in Hallmark Hypoxia

```{r hallmark_hypoxia_te_normoxia, fig.width=18, fig.height=9}
# Pathway of interest
pathway_id <- "HALLMARK_HYPOXIA"

# Create individual plots for each treatment/timepoint
plots_list <- list()

# si3e normoxia - 1hr
if (pathway_id %in% gsea_te_si3e_norm_1hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_norm_1hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_1hr"]] <- gseaplot2(gsea_te_si3e_norm_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 1hr",
                                        color = "green3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 1hr") +
    theme_void()
}

# si3e normoxia - 4hr
if (pathway_id %in% gsea_te_si3e_norm_4hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_norm_4hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_4hr"]] <- gseaplot2(gsea_te_si3e_norm_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 4hr",
                                        color = "green3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 4hr") +
    theme_void()
}

# si3e normoxia - 24hr
if (pathway_id %in% gsea_te_si3e_norm_24hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_norm_24hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_24hr"]] <- gseaplot2(gsea_te_si3e_norm_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3e - 24hr",
                                         color = "green3",
                                         base_size = 11,
                                         subplots = 1,
                                         pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 24hr") +
    theme_void()
}

# si3d normoxia - 1hr
if (pathway_id %in% gsea_te_si3d_norm_1hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_norm_1hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_1hr"]] <- gseaplot2(gsea_te_si3d_norm_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 1hr",
                                        color = "purple3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 1hr") +
    theme_void()
}

# si3d normoxia - 4hr
if (pathway_id %in% gsea_te_si3d_norm_4hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_norm_4hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_4hr"]] <- gseaplot2(gsea_te_si3d_norm_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 4hr",
                                        color = "purple3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 4hr") +
    theme_void()
}

# si3d normoxia - 24hr
if (pathway_id %in% gsea_te_si3d_norm_24hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_norm_24hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_24hr"]] <- gseaplot2(gsea_te_si3d_norm_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3d - 24hr",
                                         color = "purple3",
                                         base_size = 11,
                                         subplots = 1,
                                         pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 24hr") +
    theme_void()
}

# Combine plots - 2 rows (treatments) × 3 columns (timepoints)
combined_normoxia <- wrap_plots(
  plots_list[["si3e_1hr"]], plots_list[["si3e_4hr"]], plots_list[["si3e_24hr"]],
  plots_list[["si3d_1hr"]], plots_list[["si3d_4hr"]], plots_list[["si3d_24hr"]],
  ncol = 3, nrow = 2
)

print(combined_normoxia)

# Save
ggsave(here("plots/hallmark_hypoxia_te_normoxia.pdf"),
       combined_normoxia,
       width = 18, height = 9)

cat("Normoxia Hallmark Hypoxia plot created and saved\n")
```

## Hypoxia - TE Changes in Hallmark Hypoxia

```{r hallmark_hypoxia_te_hypoxia, fig.width=18, fig.height=9}
# Pathway of interest
pathway_id <- "HALLMARK_HYPOXIA"

# Create individual plots for each treatment/timepoint
plots_list <- list()

# si3e hypoxia - 1hr
if (pathway_id %in% gsea_te_si3e_hyp_1hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_hyp_1hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_1hr"]] <- gseaplot2(gsea_te_si3e_hyp_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 1hr",
                                        color = "green3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 1hr") +
    theme_void()
}

# si3e hypoxia - 4hr
if (pathway_id %in% gsea_te_si3e_hyp_4hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_hyp_4hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_4hr"]] <- gseaplot2(gsea_te_si3e_hyp_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 4hr",
                                        color = "green3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 4hr") +
    theme_void()
}

# si3e hypoxia - 24hr
if (pathway_id %in% gsea_te_si3e_hyp_24hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3e_hyp_24hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3e_24hr"]] <- gseaplot2(gsea_te_si3e_hyp_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3e - 24hr",
                                         color = "green3",
                                         base_size = 11,
                                         subplots = 1,
                                         pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3e_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 24hr") +
    theme_void()
}

# si3d hypoxia - 1hr
if (pathway_id %in% gsea_te_si3d_hyp_1hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_hyp_1hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_1hr"]] <- gseaplot2(gsea_te_si3d_hyp_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 1hr",
                                        color = "purple3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 1hr") +
    theme_void()
}

# si3d hypoxia - 4hr
if (pathway_id %in% gsea_te_si3d_hyp_4hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_hyp_4hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_4hr"]] <- gseaplot2(gsea_te_si3d_hyp_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 4hr",
                                        color = "purple3",
                                        base_size = 11,
                                        subplots = 1,
                                        pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 4hr") +
    theme_void()
}

# si3d hypoxia - 24hr
if (pathway_id %in% gsea_te_si3d_hyp_24hr@result$ID) {
  # Extract stats
  pathway_stats <- gsea_te_si3d_hyp_24hr@result %>% filter(ID == pathway_id)
  nes_val <- round(pathway_stats$NES, 3)
  padj_val <- formatC(pathway_stats$p.adjust, format = "e", digits = 2)

  plots_list[["si3d_24hr"]] <- gseaplot2(gsea_te_si3d_hyp_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3d - 24hr",
                                         color = "purple3",
                                         base_size = 11,
                                         subplots = 1,
                                         pvalue_table = FALSE) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("NES: ", nes_val, "\np.adjust: ", padj_val),
             hjust = 1.1, vjust = 1.5, size = 3.5)
} else {
  plots_list[["si3d_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 24hr") +
    theme_void()
}

# Combine plots - 2 rows (treatments) × 3 columns (timepoints)
combined_hypoxia <- wrap_plots(
  plots_list[["si3e_1hr"]], plots_list[["si3e_4hr"]], plots_list[["si3e_24hr"]],
  plots_list[["si3d_1hr"]], plots_list[["si3d_4hr"]], plots_list[["si3d_24hr"]],
  ncol = 3, nrow = 2
)

print(combined_hypoxia)

# Save
ggsave(here("plots/hallmark_hypoxia_te_hypoxia.pdf"),
       combined_hypoxia,
       width = 18, height = 9)

cat("Hypoxia Hallmark Hypoxia plot created and saved\n")
```

# KEGG GSEA Analysis

# GSEA: RNA Changes in si3e Normoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in normoxia using KEGG pathways.

## GSEA RNA si3e Normoxia - 1hr (KEGG)

```{r gsea_rna_kegg_si3e_norm_1hr}
ranked_list <- rna_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_norm_1hr_results[] <- lapply(gsea_rna_kegg_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Normoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_norm_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 4hr (KEGG)

```{r gsea_rna_kegg_si3e_norm_4hr}
ranked_list <- rna_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_norm_4hr_results[] <- lapply(gsea_rna_kegg_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Normoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_norm_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 24hr (KEGG)

```{r gsea_rna_kegg_si3e_norm_24hr}
ranked_list <- rna_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_norm_24hr_results[] <- lapply(gsea_rna_kegg_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Normoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_norm_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3e Normoxia RNA GSEA results (KEGG)

```{r export_gsea_rna_kegg_si3e_norm}
gsea_rna_kegg_si3e_norm_export <- list(
  "GSEA_RNA_si3e_norm_1hr" = gsea_rna_kegg_si3e_norm_1hr_results,
  "GSEA_RNA_si3e_norm_4hr" = gsea_rna_kegg_si3e_norm_4hr_results,
  "GSEA_RNA_si3e_norm_24hr" = gsea_rna_kegg_si3e_norm_24hr_results)
write_xlsx(gsea_rna_kegg_si3e_norm_export, here("output/gsea_rna_si3e_normoxia_kegg.xlsx"))
cat("\nsi3e Normoxia RNA GSEA results (KEGG) exported\n")
```

## si3e Normoxia RNA GSEA Heatmap (KEGG)

```{r gsea_rna_kegg_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_kegg_si3e_norm_1hr_results) > 0) gsea_rna_kegg_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3e_norm_4hr_results) > 0) gsea_rna_kegg_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3e_norm_24hr_results) > 0) gsea_rna_kegg_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "RNA GSEA NES - si3e Normoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_rna_si3e_normoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3e Normoxia RNA KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3e Hypoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia using KEGG pathways.

## GSEA RNA si3e Hypoxia - 1hr (KEGG)

```{r gsea_rna_kegg_si3e_hyp_1hr}
ranked_list <- rna_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_hyp_1hr_results[] <- lapply(gsea_rna_kegg_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Hypoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_hyp_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 4hr (KEGG)

```{r gsea_rna_kegg_si3e_hyp_4hr}
ranked_list <- rna_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_hyp_4hr_results[] <- lapply(gsea_rna_kegg_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Hypoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_hyp_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 24hr (KEGG)

```{r gsea_rna_kegg_si3e_hyp_24hr}
ranked_list <- rna_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3e_hyp_24hr_results[] <- lapply(gsea_rna_kegg_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3e Hypoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3e_hyp_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3e Hypoxia RNA GSEA results (KEGG)

```{r export_gsea_rna_kegg_si3e_hyp}
gsea_rna_kegg_si3e_hyp_export <- list(
  "GSEA_RNA_si3e_hyp_1hr" = gsea_rna_kegg_si3e_hyp_1hr_results,
  "GSEA_RNA_si3e_hyp_4hr" = gsea_rna_kegg_si3e_hyp_4hr_results,
  "GSEA_RNA_si3e_hyp_24hr" = gsea_rna_kegg_si3e_hyp_24hr_results)
write_xlsx(gsea_rna_kegg_si3e_hyp_export, here("output/gsea_rna_si3e_hypoxia_kegg.xlsx"))
cat("\nsi3e Hypoxia RNA GSEA results (KEGG) exported\n")
```

## si3e Hypoxia RNA GSEA Heatmap (KEGG)

```{r gsea_rna_kegg_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_kegg_si3e_hyp_1hr_results) > 0) gsea_rna_kegg_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3e_hyp_4hr_results) > 0) gsea_rna_kegg_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3e_hyp_24hr_results) > 0) gsea_rna_kegg_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "RNA GSEA NES - si3e Hypoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_rna_si3e_hypoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3e Hypoxia RNA KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Normoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in normoxia using KEGG pathways.

## GSEA RNA si3d Normoxia - 1hr (KEGG)

```{r gsea_rna_kegg_si3d_norm_1hr}
ranked_list <- rna_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_norm_1hr_results[] <- lapply(gsea_rna_kegg_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Normoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_norm_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 4hr (KEGG)

```{r gsea_rna_kegg_si3d_norm_4hr}
ranked_list <- rna_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_norm_4hr_results[] <- lapply(gsea_rna_kegg_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Normoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_norm_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 24hr (KEGG)

```{r gsea_rna_kegg_si3d_norm_24hr}
ranked_list <- rna_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_norm_24hr_results[] <- lapply(gsea_rna_kegg_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Normoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_norm_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3d Normoxia RNA GSEA results (KEGG)

```{r export_gsea_rna_kegg_si3d_norm}
gsea_rna_kegg_si3d_norm_export <- list(
  "GSEA_RNA_si3d_norm_1hr" = gsea_rna_kegg_si3d_norm_1hr_results,
  "GSEA_RNA_si3d_norm_4hr" = gsea_rna_kegg_si3d_norm_4hr_results,
  "GSEA_RNA_si3d_norm_24hr" = gsea_rna_kegg_si3d_norm_24hr_results)
write_xlsx(gsea_rna_kegg_si3d_norm_export, here("output/gsea_rna_si3d_normoxia_kegg.xlsx"))
cat("\nsi3d Normoxia RNA GSEA results (KEGG) exported\n")
```

## si3d Normoxia RNA GSEA Heatmap (KEGG)

```{r gsea_rna_kegg_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_kegg_si3d_norm_1hr_results) > 0) gsea_rna_kegg_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3d_norm_4hr_results) > 0) gsea_rna_kegg_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3d_norm_24hr_results) > 0) gsea_rna_kegg_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "RNA GSEA NES - si3d Normoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_rna_si3d_normoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3d Normoxia RNA KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Hypoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia using KEGG pathways.

## GSEA RNA si3d Hypoxia - 1hr (KEGG)

```{r gsea_rna_kegg_si3d_hyp_1hr}
ranked_list <- rna_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_hyp_1hr_results[] <- lapply(gsea_rna_kegg_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Hypoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_hyp_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 4hr (KEGG)

```{r gsea_rna_kegg_si3d_hyp_4hr}
ranked_list <- rna_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_hyp_4hr_results[] <- lapply(gsea_rna_kegg_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Hypoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_hyp_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 24hr (KEGG)

```{r gsea_rna_kegg_si3d_hyp_24hr}
ranked_list <- rna_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_rna_kegg_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_rna_kegg_si3d_hyp_24hr_results[] <- lapply(gsea_rna_kegg_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA RNA si3d Hypoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_rna_kegg_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_kegg_si3d_hyp_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3d Hypoxia RNA GSEA results (KEGG)

```{r export_gsea_rna_kegg_si3d_hyp}
gsea_rna_kegg_si3d_hyp_export <- list(
  "GSEA_RNA_si3d_hyp_1hr" = gsea_rna_kegg_si3d_hyp_1hr_results,
  "GSEA_RNA_si3d_hyp_4hr" = gsea_rna_kegg_si3d_hyp_4hr_results,
  "GSEA_RNA_si3d_hyp_24hr" = gsea_rna_kegg_si3d_hyp_24hr_results)
write_xlsx(gsea_rna_kegg_si3d_hyp_export, here("output/gsea_rna_si3d_hypoxia_kegg.xlsx"))
cat("\nsi3d Hypoxia RNA GSEA results (KEGG) exported\n")
```

## si3d Hypoxia RNA GSEA Heatmap (KEGG)

```{r gsea_rna_kegg_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_kegg_si3d_hyp_1hr_results) > 0) gsea_rna_kegg_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3d_hyp_4hr_results) > 0) gsea_rna_kegg_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_kegg_si3d_hyp_24hr_results) > 0) gsea_rna_kegg_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "RNA GSEA NES - si3d Hypoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_rna_si3d_hypoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3d Hypoxia RNA KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Normoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in normoxia using KEGG pathways.

## GSEA TE si3e Normoxia - 1hr (KEGG)

```{r gsea_te_kegg_si3e_norm_1hr}
ranked_list <- te_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 1hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_norm_1hr_results[] <- lapply(gsea_te_kegg_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Normoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_norm_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 4hr (KEGG)

```{r gsea_te_kegg_si3e_norm_4hr}
ranked_list <- te_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 4hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_norm_4hr_results[] <- lapply(gsea_te_kegg_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Normoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_norm_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 24hr (KEGG)

```{r gsea_te_kegg_si3e_norm_24hr}
ranked_list <- te_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Normoxia 24hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_norm_24hr_results[] <- lapply(gsea_te_kegg_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Normoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_norm_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3e Normoxia TE GSEA results (KEGG)

```{r export_gsea_te_kegg_si3e_norm}
gsea_te_kegg_si3e_norm_export <- list(
  "GSEA_TE_si3e_norm_1hr" = gsea_te_kegg_si3e_norm_1hr_results,
  "GSEA_TE_si3e_norm_4hr" = gsea_te_kegg_si3e_norm_4hr_results,
  "GSEA_TE_si3e_norm_24hr" = gsea_te_kegg_si3e_norm_24hr_results)
write_xlsx(gsea_te_kegg_si3e_norm_export, here("output/gsea_te_si3e_normoxia_kegg.xlsx"))
cat("\nsi3e Normoxia TE GSEA results (KEGG) exported\n")
```

## si3e Normoxia TE GSEA Heatmap (KEGG)

```{r gsea_te_kegg_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_kegg_si3e_norm_1hr_results) > 0) gsea_te_kegg_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3e_norm_4hr_results) > 0) gsea_te_kegg_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3e_norm_24hr_results) > 0) gsea_te_kegg_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "TE GSEA NES - si3e Normoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_te_si3e_normoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3e Normoxia TE KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Hypoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia using KEGG pathways.

## GSEA TE si3e Hypoxia - 1hr (KEGG)

```{r gsea_te_kegg_si3e_hyp_1hr}
ranked_list <- te_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 1hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_hyp_1hr_results[] <- lapply(gsea_te_kegg_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Hypoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_hyp_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 4hr (KEGG)

```{r gsea_te_kegg_si3e_hyp_4hr}
ranked_list <- te_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 4hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_hyp_4hr_results[] <- lapply(gsea_te_kegg_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Hypoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_hyp_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 24hr (KEGG)

```{r gsea_te_kegg_si3e_hyp_24hr}
ranked_list <- te_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3e Hypoxia 24hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3e_hyp_24hr_results[] <- lapply(gsea_te_kegg_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3e Hypoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3e_hyp_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3e Hypoxia TE GSEA results (KEGG)

```{r export_gsea_te_kegg_si3e_hyp}
gsea_te_kegg_si3e_hyp_export <- list(
  "GSEA_TE_si3e_hyp_1hr" = gsea_te_kegg_si3e_hyp_1hr_results,
  "GSEA_TE_si3e_hyp_4hr" = gsea_te_kegg_si3e_hyp_4hr_results,
  "GSEA_TE_si3e_hyp_24hr" = gsea_te_kegg_si3e_hyp_24hr_results)
write_xlsx(gsea_te_kegg_si3e_hyp_export, here("output/gsea_te_si3e_hypoxia_kegg.xlsx"))
cat("\nsi3e Hypoxia TE GSEA results (KEGG) exported\n")
```

## si3e Hypoxia TE GSEA Heatmap (KEGG)

```{r gsea_te_kegg_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_kegg_si3e_hyp_1hr_results) > 0) gsea_te_kegg_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3e_hyp_4hr_results) > 0) gsea_te_kegg_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3e_hyp_24hr_results) > 0) gsea_te_kegg_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "TE GSEA NES - si3e Hypoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_te_si3e_hypoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3e Hypoxia TE KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Normoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in normoxia using KEGG pathways.

## GSEA TE si3d Normoxia - 1hr (KEGG)

```{r gsea_te_kegg_si3d_norm_1hr}
ranked_list <- te_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 1hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_norm_1hr_results[] <- lapply(gsea_te_kegg_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Normoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_norm_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 4hr (KEGG)

```{r gsea_te_kegg_si3d_norm_4hr}
ranked_list <- te_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 4hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_norm_4hr_results[] <- lapply(gsea_te_kegg_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Normoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_norm_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 24hr (KEGG)

```{r gsea_te_kegg_si3d_norm_24hr}
ranked_list <- te_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Normoxia 24hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_norm_24hr_results[] <- lapply(gsea_te_kegg_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Normoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_norm_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3d Normoxia TE GSEA results (KEGG)

```{r export_gsea_te_kegg_si3d_norm}
gsea_te_kegg_si3d_norm_export <- list(
  "GSEA_TE_si3d_norm_1hr" = gsea_te_kegg_si3d_norm_1hr_results,
  "GSEA_TE_si3d_norm_4hr" = gsea_te_kegg_si3d_norm_4hr_results,
  "GSEA_TE_si3d_norm_24hr" = gsea_te_kegg_si3d_norm_24hr_results)
write_xlsx(gsea_te_kegg_si3d_norm_export, here("output/gsea_te_si3d_normoxia_kegg.xlsx"))
cat("\nsi3d Normoxia TE GSEA results (KEGG) exported\n")
```

## si3d Normoxia TE GSEA Heatmap (KEGG)

```{r gsea_te_kegg_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_kegg_si3d_norm_1hr_results) > 0) gsea_te_kegg_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3d_norm_4hr_results) > 0) gsea_te_kegg_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3d_norm_24hr_results) > 0) gsea_te_kegg_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "TE GSEA NES - si3d Normoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_te_si3d_normoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3d Normoxia TE KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Hypoxia (KEGG)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia using KEGG pathways.

## GSEA TE si3d Hypoxia - 1hr (KEGG)

```{r gsea_te_kegg_si3d_hyp_1hr}
ranked_list <- te_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_1hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 1hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_hyp_1hr_results[] <- lapply(gsea_te_kegg_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Hypoxia 1hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_hyp_1hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 1hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 4hr (KEGG)

```{r gsea_te_kegg_si3d_hyp_4hr}
ranked_list <- te_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_4hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 4hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_hyp_4hr_results[] <- lapply(gsea_te_kegg_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Hypoxia 4hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_hyp_4hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 4hr (KEGG) - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 24hr (KEGG)

```{r gsea_te_kegg_si3d_hyp_24hr}
ranked_list <- te_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_24hr$symbol
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]
cat("si3d Hypoxia 24hr TE - Total genes:", length(ranked_list), "\n")

gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = kegg_rna
)

if (nrow(gsea_result@result) > 0) {
  gsea_te_kegg_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)
  gsea_te_kegg_si3d_hyp_24hr_results[] <- lapply(gsea_te_kegg_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })
  cat("GSEA TE si3d Hypoxia 24hr (KEGG) - Top pathways:\n")
  print(gsea_te_kegg_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_kegg_si3d_hyp_24hr_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(), p.adjust = numeric(),
    setSize = integer(), leading_edge = character(), leading_edge_count = integer(),
    leading_edge_fraction = numeric(), stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 24hr (KEGG) - No pathways enriched\n")
}
```

## Export si3d Hypoxia TE GSEA results (KEGG)

```{r export_gsea_te_kegg_si3d_hyp}
gsea_te_kegg_si3d_hyp_export <- list(
  "GSEA_TE_si3d_hyp_1hr" = gsea_te_kegg_si3d_hyp_1hr_results,
  "GSEA_TE_si3d_hyp_4hr" = gsea_te_kegg_si3d_hyp_4hr_results,
  "GSEA_TE_si3d_hyp_24hr" = gsea_te_kegg_si3d_hyp_24hr_results)
write_xlsx(gsea_te_kegg_si3d_hyp_export, here("output/gsea_te_si3d_hypoxia_kegg.xlsx"))
cat("\nsi3d Hypoxia TE GSEA results (KEGG) exported\n")
```

## si3d Hypoxia TE GSEA Heatmap (KEGG)

```{r gsea_te_kegg_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_kegg_si3d_hyp_1hr_results) > 0) gsea_te_kegg_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3d_hyp_4hr_results) > 0) gsea_te_kegg_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_kegg_si3d_hyp_24hr_results) > 0) gsea_te_kegg_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(gsea_wide)) gsea_wide[[tp]] <- 0 }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]
    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")
    for (tp in c("1hr", "4hr", "24hr")) { if (!tp %in% colnames(sig_matrix)) sig_matrix[[tp]] <- "" }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
    p_heatmap <- pheatmap(
      gsea_wide, cluster_rows = FALSE, cluster_cols = FALSE, scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix, number_color = "black", fontsize_number = 14,
      main = "TE GSEA NES - si3d Hypoxia - KEGG (* = p.adjust < 0.05)",
      fontsize_row = 8, fontsize_col = 10, cellwidth = 40, cellheight = 12
    )
    pdf(here("plots/gsea_te_si3d_hypoxia_kegg_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()
    cat("si3d Hypoxia TE KEGG heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# ORA: EIF3D-Promoted Genes (Hypoxia 1hr, LFC > 0.5)

Over-representation analysis of the 1,604 EIF3D-promoted genes against Hallmark and KEGG gene sets.

```{r ora_eif3d_load_genes}
# Load EIF3D-promoted gene list
eif3d_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv")) %>%
  pull(db_gene_symbol) %>%
  unique() %>%
  na.omit() %>%
  as.character()

cat("EIF3D-promoted genes loaded:", length(eif3d_genes), "\n")

# Background: all tested genes from TE data (si3d hypoxia 1hr)
background_genes <- te_si3d_hyp_1hr$symbol
background_genes <- background_genes[!is.na(background_genes)]
cat("Background genes:", length(background_genes), "\n")
```

## ORA: Hallmark

```{r ora_eif3d_hallmark}
ora_hallmark <- enricher(
  gene = eif3d_genes,
  TERM2GENE = hallmark_rna,
  universe = background_genes,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

if (!is.null(ora_hallmark) && nrow(ora_hallmark@result) > 0) {
  ora_hallmark_results <- ora_hallmark@result %>%
    mutate(GeneRatio_numeric = sapply(GeneRatio, function(x) {
      parts <- strsplit(x, "/")[[1]]
      as.numeric(parts[1]) / as.numeric(parts[2])
    })) %>%
    arrange(p.adjust)

  cat("Hallmark ORA - Significant pathways (p.adjust < 0.05):\n")
  print(ora_hallmark_results %>% filter(p.adjust < 0.05) %>% dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count))
} else {
  ora_hallmark_results <- data.frame(
    ID = character(), GeneRatio = character(), BgRatio = character(),
    pvalue = numeric(), p.adjust = numeric(), qvalue = numeric(),
    geneID = character(), Count = integer(), GeneRatio_numeric = numeric(),
    stringsAsFactors = FALSE
  )
  cat("Hallmark ORA - No pathways enriched\n")
}
```

## ORA: KEGG

```{r ora_eif3d_kegg}
ora_kegg <- enricher(
  gene = eif3d_genes,
  TERM2GENE = kegg_rna,
  universe = background_genes,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

if (!is.null(ora_kegg) && nrow(ora_kegg@result) > 0) {
  ora_kegg_results <- ora_kegg@result %>%
    mutate(GeneRatio_numeric = sapply(GeneRatio, function(x) {
      parts <- strsplit(x, "/")[[1]]
      as.numeric(parts[1]) / as.numeric(parts[2])
    })) %>%
    arrange(p.adjust)

  cat("KEGG ORA - Significant pathways (p.adjust < 0.05):\n")
  print(ora_kegg_results %>% filter(p.adjust < 0.05) %>% dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count))
} else {
  ora_kegg_results <- data.frame(
    ID = character(), GeneRatio = character(), BgRatio = character(),
    pvalue = numeric(), p.adjust = numeric(), qvalue = numeric(),
    geneID = character(), Count = integer(), GeneRatio_numeric = numeric(),
    stringsAsFactors = FALSE
  )
  cat("KEGG ORA - No pathways enriched\n")
}
```

## ORA: GO Terms (BP, MF, CC)

```{r ora_eif3d_go}
library(org.Hs.eg.db)

run_go_ora <- function(ont_label) {
  res <- enrichGO(
    gene = eif3d_genes,
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = ont_label,
    universe = background_genes,
    pAdjustMethod = "fdr",
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 10,
    maxGSSize = 500
  )
  res
}

ora_go_bp <- run_go_ora("BP")
ora_go_mf <- run_go_ora("MF")
ora_go_cc <- run_go_ora("CC")

extract_go_results <- function(ora_obj, label) {
  if (!is.null(ora_obj) && nrow(ora_obj@result) > 0) {
    res <- ora_obj@result %>%
      mutate(GeneRatio_numeric = sapply(GeneRatio, function(x) {
        parts <- strsplit(x, "/")[[1]]
        as.numeric(parts[1]) / as.numeric(parts[2])
      })) %>%
      arrange(p.adjust)
    cat(label, "- Significant pathways (p.adjust < 0.05):", sum(res$p.adjust < 0.05), "\n")
    res
  } else {
    cat(label, "- No pathways enriched\n")
    data.frame(
      ID = character(), Description = character(), GeneRatio = character(),
      BgRatio = character(), pvalue = numeric(), p.adjust = numeric(),
      qvalue = numeric(), geneID = character(), Count = integer(),
      GeneRatio_numeric = numeric(), stringsAsFactors = FALSE
    )
  }
}

ora_go_bp_results <- extract_go_results(ora_go_bp, "GO BP")
ora_go_mf_results <- extract_go_results(ora_go_mf, "GO MF")
ora_go_cc_results <- extract_go_results(ora_go_cc, "GO CC")
```

## Export ORA Results

```{r export_ora_eif3d}
ora_export <- list(
  "Hallmark" = ora_hallmark_results,
  "KEGG" = ora_kegg_results,
  "GO_BP" = ora_go_bp_results,
  "GO_MF" = ora_go_mf_results,
  "GO_CC" = ora_go_cc_results
)
write_xlsx(ora_export, here("output/ora_eif3d_promoted_hypoxia_1hr_lfc0.5.xlsx"))
cat("ORA results exported to output/ora_eif3d_promoted_hypoxia_1hr_lfc0.5.xlsx\n")
```

## ORA Dot Plots

```{r ora_eif3d_dotplots}
# Hallmark dot plot - significant pathways only
if (!is.null(ora_hallmark) && sum(ora_hallmark@result$p.adjust < 0.05) > 0) {
  ora_hallmark_sig <- ora_hallmark
  ora_hallmark_sig@result <- ora_hallmark_sig@result %>% filter(p.adjust < 0.05)
  p_ora_hallmark <- dotplot(ora_hallmark_sig, showCategory = nrow(ora_hallmark_sig@result),
                            title = "ORA Hallmark - EIF3D-Promoted Genes (Hypoxia 1hr, LFC > 0.5)")
  print(p_ora_hallmark)
  ggsave(here("plots/ora_eif3d_hallmark_dotplot.pdf"), p_ora_hallmark, width = 10, height = 8)
  cat("Hallmark ORA dotplot saved\n")
} else {
  cat("No significant Hallmark pathways to plot\n")
}

# KEGG dot plot - significant pathways only
if (!is.null(ora_kegg) && sum(ora_kegg@result$p.adjust < 0.05) > 0) {
  ora_kegg_sig <- ora_kegg
  ora_kegg_sig@result <- ora_kegg_sig@result %>% filter(p.adjust < 0.05)
  p_ora_kegg <- dotplot(ora_kegg_sig, showCategory = nrow(ora_kegg_sig@result),
                        title = "ORA KEGG - EIF3D-Promoted Genes (Hypoxia 1hr, LFC > 0.5)")
  print(p_ora_kegg)
  ggsave(here("plots/ora_eif3d_kegg_dotplot.pdf"), p_ora_kegg, width = 10, height = 8)
  cat("KEGG ORA dotplot saved\n")
} else {
  cat("No significant KEGG pathways to plot\n")
}

# GO dot plots - significant pathways only
plot_go_dotplot <- function(ora_obj, label, filename) {
  if (!is.null(ora_obj) && sum(ora_obj@result$p.adjust < 0.05) > 0) {
    ora_sig <- ora_obj
    ora_sig@result <- ora_sig@result %>% filter(p.adjust < 0.05)
    p <- dotplot(ora_sig, showCategory = nrow(ora_sig@result),
                 title = paste0("ORA ", label, " - EIF3D-Promoted Genes (Hypoxia 1hr, LFC > 0.5)"))
    print(p)
    ggsave(here("plots", filename), p, width = 10, height = 8)
    cat(label, "ORA dotplot saved\n")
  } else {
    cat("No significant", label, "pathways to plot\n")
  }
}

plot_go_dotplot(ora_go_bp, "GO BP", "ora_eif3d_go_bp_dotplot.pdf")
plot_go_dotplot(ora_go_mf, "GO MF", "ora_eif3d_go_mf_dotplot.pdf")
plot_go_dotplot(ora_go_cc, "GO CC", "ora_eif3d_go_cc_dotplot.pdf")
```
