---
title: "GSEA Analysis - eIF3e and eIF3d Knockdowns"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This analysis performs Gene Set Enrichment Analysis (GSEA) on RNA-seq and Translation Efficiency (TE) changes in eIF3e and eIF3d knockdown cells under normoxia and hypoxia conditions.

**Comparisons:**
- sictrl vs si3e in normoxia
- sictrl vs si3e in hypoxia
- sictrl vs si3d in normoxia
- sictrl vs si3d in hypoxia

**Timepoints:** 1hr, 4hr, 24hr

**Data types:** RNA-seq log2 fold changes and TE log2 fold changes

```{r load_libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(pheatmap)
library(writexl)
library(enrichplot)
library(patchwork)

cat("Libraries loaded successfully\n")
```

# Load Hallmark Gene Sets

```{r load_hallmark}
# Load MSigDB Hallmark gene sets for human
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# Create gene set format for clusterProfiler (TERM2GENE)
hallmark_rna <- hallmark_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark_rna$gs_name)), "\n")
```

# Load RNA and TE Results

## si3e Normoxia (sictrl vs si3e in normoxia)

```{r load_si3e_norm}
# Load RNA results
rna_si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3e_norm_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_norm_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_norm_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3e Normoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3e_norm_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3e_norm_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3e_norm_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3e_norm_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3e_norm_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3e_norm_24hr), "genes\n")
```

## si3e Hypoxia (sictrl vs si3e in hypoxia)

```{r load_si3e_hyp}
# Load RNA results
rna_si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3e_hyp_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_hyp_4hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3e_hyp_24hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3e Hypoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3e_hyp_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3e_hyp_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3e_hyp_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3e_hyp_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3e_hyp_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3e_hyp_24hr), "genes\n")
```

## si3d Normoxia (sictrl vs si3d in normoxia)

```{r load_si3d_norm}
# Load RNA results
rna_si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3d Normoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3d_norm_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3d_norm_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3d_norm_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3d_norm_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3d_norm_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3d_norm_24hr), "genes\n")
```

## si3d Hypoxia (sictrl vs si3d in hypoxia)

```{r load_si3d_hyp}
# Load RNA results
rna_si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

rna_si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, rna_lfc, symbol) %>%
  rename(log2FoldChange = rna_lfc)

# Load TE results
te_si3d_hyp_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_hyp_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_hyp_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3d Hypoxia data loaded:\n")
cat("RNA - 1hr:", nrow(rna_si3d_hyp_1hr), "genes\n")
cat("RNA - 4hr:", nrow(rna_si3d_hyp_4hr), "genes\n")
cat("RNA - 24hr:", nrow(rna_si3d_hyp_24hr), "genes\n")
cat("TE - 1hr:", nrow(te_si3d_hyp_1hr), "genes\n")
cat("TE - 4hr:", nrow(te_si3d_hyp_4hr), "genes\n")
cat("TE - 24hr:", nrow(te_si3d_hyp_24hr), "genes\n")
```

# GSEA: RNA Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA RNA si3e Normoxia - 1hr

```{r gsea_rna_si3e_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_1hr_results[] <- lapply(gsea_rna_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 4hr

```{r gsea_rna_si3e_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_4hr_results[] <- lapply(gsea_rna_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Normoxia - 24hr

```{r gsea_rna_si3e_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_norm_24hr_results[] <- lapply(gsea_rna_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia RNA GSEA results

```{r export_gsea_rna_si3e_norm}
gsea_rna_si3e_norm_export <- list(
  "GSEA_RNA_si3e_norm_1hr" = gsea_rna_si3e_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_norm_4hr" = gsea_rna_si3e_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_norm_24hr" = gsea_rna_si3e_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3e_norm_export, here("output/gsea_rna_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia RNA GSEA results exported\n")
```

## si3e Normoxia RNA GSEA Heatmap

```{r gsea_rna_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_norm_1hr_results) > 0) gsea_rna_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_4hr_results) > 0) gsea_rna_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_norm_24hr_results) > 0) gsea_rna_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA RNA si3e Hypoxia - 1hr

```{r gsea_rna_si3e_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_1hr_results[] <- lapply(gsea_rna_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 4hr

```{r gsea_rna_si3e_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_4hr_results[] <- lapply(gsea_rna_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3e Hypoxia - 24hr

```{r gsea_rna_si3e_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3e_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3e_hyp_24hr_results[] <- lapply(gsea_rna_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia RNA GSEA results

```{r export_gsea_rna_si3e_hyp}
gsea_rna_si3e_hyp_export <- list(
  "GSEA_RNA_si3e_hyp_1hr" = gsea_rna_si3e_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_hyp_4hr" = gsea_rna_si3e_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3e_hyp_24hr" = gsea_rna_si3e_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3e_hyp_export, here("output/gsea_rna_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia RNA GSEA results exported\n")
```

## si3e Hypoxia RNA GSEA Heatmap

```{r gsea_rna_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3e_hyp_1hr_results) > 0) gsea_rna_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_4hr_results) > 0) gsea_rna_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3e_hyp_24hr_results) > 0) gsea_rna_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA RNA si3d Normoxia - 1hr

```{r gsea_rna_si3d_norm_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_1hr_results[] <- lapply(gsea_rna_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 4hr

```{r gsea_rna_si3d_norm_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_4hr_results[] <- lapply(gsea_rna_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Normoxia - 24hr

```{r gsea_rna_si3d_norm_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_norm_24hr_results[] <- lapply(gsea_rna_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia RNA GSEA results

```{r export_gsea_rna_si3d_norm}
gsea_rna_si3d_norm_export <- list(
  "GSEA_RNA_si3d_norm_1hr" = gsea_rna_si3d_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_norm_4hr" = gsea_rna_si3d_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_norm_24hr" = gsea_rna_si3d_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3d_norm_export, here("output/gsea_rna_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia RNA GSEA results exported\n")
```

## si3d Normoxia RNA GSEA Heatmap

```{r gsea_rna_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_norm_1hr_results) > 0) gsea_rna_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_4hr_results) > 0) gsea_rna_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_norm_24hr_results) > 0) gsea_rna_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: RNA Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on RNA-seq log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA RNA si3d Hypoxia - 1hr

```{r gsea_rna_si3d_hyp_1hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_1hr_results[] <- lapply(gsea_rna_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 4hr

```{r gsea_rna_si3d_hyp_4hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_4hr_results[] <- lapply(gsea_rna_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA RNA si3d Hypoxia - 24hr

```{r gsea_rna_si3d_hyp_24hr}
# Create ranked list based on RNA log2FC (using all genes)
ranked_list <- rna_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- rna_si3d_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_rna_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_rna_si3d_hyp_24hr_results[] <- lapply(gsea_rna_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA RNA si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_rna_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_rna_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA RNA si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia RNA GSEA results

```{r export_gsea_rna_si3d_hyp}
gsea_rna_si3d_hyp_export <- list(
  "GSEA_RNA_si3d_hyp_1hr" = gsea_rna_si3d_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_hyp_4hr" = gsea_rna_si3d_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_RNA_si3d_hyp_24hr" = gsea_rna_si3d_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_rna_si3d_hyp_export, here("output/gsea_rna_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia RNA GSEA results exported\n")
```

## si3d Hypoxia RNA GSEA Heatmap

```{r gsea_rna_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_rna_si3d_hyp_1hr_results) > 0) gsea_rna_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_4hr_results) > 0) gsea_rna_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_rna_si3d_hyp_24hr_results) > 0) gsea_rna_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "RNA GSEA NES - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_rna_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia RNA heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in normoxia.

## GSEA TE si3e Normoxia - 1hr

```{r gsea_te_si3e_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_1hr_results[] <- lapply(gsea_te_si3e_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 4hr

```{r gsea_te_si3e_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_4hr_results[] <- lapply(gsea_te_si3e_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Normoxia - 24hr

```{r gsea_te_si3e_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3e_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_norm_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_norm_24hr_results[] <- lapply(gsea_te_si3e_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Normoxia TE GSEA results

```{r export_gsea_te_si3e_norm}
gsea_te_si3e_norm_export <- list(
  "GSEA_TE_si3e_norm_1hr" = gsea_te_si3e_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_norm_4hr" = gsea_te_si3e_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_norm_24hr" = gsea_te_si3e_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3e_norm_export, here("output/gsea_te_si3e_normoxia_hallmark.xlsx"))
cat("\nsi3e Normoxia TE GSEA results exported\n")
```

## si3e Normoxia TE GSEA Heatmap

```{r gsea_te_si3e_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_norm_1hr_results) > 0) gsea_te_si3e_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_4hr_results) > 0) gsea_te_si3e_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_norm_24hr_results) > 0) gsea_te_si3e_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3e Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3e_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Normoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3e Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3e) at each timepoint in hypoxia.

## GSEA TE si3e Hypoxia - 1hr

```{r gsea_te_si3e_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_1hr_results[] <- lapply(gsea_te_si3e_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 4hr

```{r gsea_te_si3e_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_4hr_results[] <- lapply(gsea_te_si3e_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3e Hypoxia - 24hr

```{r gsea_te_si3e_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3e_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3e_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3e Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3e_hyp_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3e_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3e_hyp_24hr_results[] <- lapply(gsea_te_si3e_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3e Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3e_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3e_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3e Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3e Hypoxia TE GSEA results

```{r export_gsea_te_si3e_hyp}
gsea_te_si3e_hyp_export <- list(
  "GSEA_TE_si3e_hyp_1hr" = gsea_te_si3e_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_hyp_4hr" = gsea_te_si3e_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3e_hyp_24hr" = gsea_te_si3e_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3e_hyp_export, here("output/gsea_te_si3e_hypoxia_hallmark.xlsx"))
cat("\nsi3e Hypoxia TE GSEA results exported\n")
```

## si3e Hypoxia TE GSEA Heatmap

```{r gsea_te_si3e_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3e_hyp_1hr_results) > 0) gsea_te_si3e_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_4hr_results) > 0) gsea_te_si3e_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3e_hyp_24hr_results) > 0) gsea_te_si3e_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3e Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3e_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3e Hypoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Normoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in normoxia.

## GSEA TE si3d Normoxia - 1hr

```{r gsea_te_si3d_norm_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_1hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_1hr_results[] <- lapply(gsea_te_si3d_norm_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_norm_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 4hr

```{r gsea_te_si3d_norm_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_4hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_4hr_results[] <- lapply(gsea_te_si3d_norm_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_norm_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Normoxia - 24hr

```{r gsea_te_si3d_norm_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_norm_24hr$log2FoldChange
names(ranked_list) <- te_si3d_norm_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Normoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_norm_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_norm_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_norm_24hr_results[] <- lapply(gsea_te_si3d_norm_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Normoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_norm_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_norm_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Normoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Normoxia TE GSEA results

```{r export_gsea_te_si3d_norm}
gsea_te_si3d_norm_export <- list(
  "GSEA_TE_si3d_norm_1hr" = gsea_te_si3d_norm_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_norm_4hr" = gsea_te_si3d_norm_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_norm_24hr" = gsea_te_si3d_norm_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3d_norm_export, here("output/gsea_te_si3d_normoxia_hallmark.xlsx"))
cat("\nsi3d Normoxia TE GSEA results exported\n")
```

## si3d Normoxia TE GSEA Heatmap

```{r gsea_te_si3d_norm_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_norm_1hr_results) > 0) gsea_te_si3d_norm_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_4hr_results) > 0) gsea_te_si3d_norm_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_norm_24hr_results) > 0) gsea_te_si3d_norm_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3d Normoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3d_normoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Normoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```

# GSEA: TE Changes in si3d Hypoxia (Hallmark Pathways)

Perform Gene Set Enrichment Analysis (GSEA) on Translation Efficiency (TE) log2 fold changes (sictrl vs si3d) at each timepoint in hypoxia.

## GSEA TE si3d Hypoxia - 1hr

```{r gsea_te_si3d_hyp_1hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_1hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_1hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 1hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_1hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_1hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_1hr_results[] <- lapply(gsea_te_si3d_hyp_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 1hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_1hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 1hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 4hr

```{r gsea_te_si3d_hyp_4hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_4hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_4hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 4hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_4hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_4hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_4hr_results[] <- lapply(gsea_te_si3d_hyp_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 4hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_4hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 4hr - No pathways enriched\n")
}
```

## GSEA TE si3d Hypoxia - 24hr

```{r gsea_te_si3d_hyp_24hr}
# Create ranked list based on TE log2FC (using all genes)
ranked_list <- te_si3d_hyp_24hr$log2FoldChange
names(ranked_list) <- te_si3d_hyp_24hr$symbol

# Remove NA symbols and duplicates
ranked_list <- ranked_list[!is.na(names(ranked_list))]
ranked_list <- sort(ranked_list, decreasing = TRUE)
ranked_list <- ranked_list[!duplicated(names(ranked_list))]

cat("si3d Hypoxia 24hr - Total genes:", length(ranked_list), "\n")

# Run GSEA
gsea_result <- GSEA(
  geneList = ranked_list,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize = 20,
  maxGSSize = 1000,
  TERM2GENE = hallmark_rna
)

# Store GSEA object for plotting
gsea_te_si3d_hyp_24hr <- gsea_result

# Extract results - handle case where no enrichments found
if (nrow(gsea_result@result) > 0) {
  gsea_te_si3d_hyp_24hr_results <- gsea_result@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_si3d_hyp_24hr_results[] <- lapply(gsea_te_si3d_hyp_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE si3d Hypoxia 24hr - Top pathways:\n")
  print(gsea_te_si3d_hyp_24hr_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(10))
} else {
  gsea_te_si3d_hyp_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = integer(),
    leading_edge = character(),
    leading_edge_count = integer(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE si3d Hypoxia 24hr - No pathways enriched\n")
}
```

## Export si3d Hypoxia TE GSEA results

```{r export_gsea_te_si3d_hyp}
gsea_te_si3d_hyp_export <- list(
  "GSEA_TE_si3d_hyp_1hr" = gsea_te_si3d_hyp_1hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_hyp_4hr" = gsea_te_si3d_hyp_4hr_results %>% dplyr::select(-leading_edge),
  "GSEA_TE_si3d_hyp_24hr" = gsea_te_si3d_hyp_24hr_results %>% dplyr::select(-leading_edge)
)

write_xlsx(gsea_te_si3d_hyp_export, here("output/gsea_te_si3d_hypoxia_hallmark.xlsx"))
cat("\nsi3d Hypoxia TE GSEA results exported\n")
```

## si3d Hypoxia TE GSEA Heatmap

```{r gsea_te_si3d_hyp_heatmap}
gsea_combined <- bind_rows(
  if(nrow(gsea_te_si3d_hyp_1hr_results) > 0) gsea_te_si3d_hyp_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_4hr_results) > 0) gsea_te_si3d_hyp_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL,
  if(nrow(gsea_te_si3d_hyp_24hr_results) > 0) gsea_te_si3d_hyp_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint) else NULL
)

if (nrow(gsea_combined) > 0) {
  sig_pathways <- gsea_combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways) > 0) {
    gsea_wide <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, NES, timepoint) %>%
      pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
      column_to_rownames("ID")

    # Ensure all three timepoint columns exist
    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(gsea_wide)) {
        gsea_wide[[tp]] <- 0
      }
    }
    gsea_wide <- gsea_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

    sig_matrix <- gsea_combined %>%
      filter(ID %in% sig_pathways) %>%
      dplyr::select(ID, p.adjust, timepoint) %>%
      mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
      dplyr::select(ID, timepoint, sig) %>%
      pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
      column_to_rownames("ID")

    for (tp in c("1hr", "4hr", "24hr")) {
      if (!tp %in% colnames(sig_matrix)) {
        sig_matrix[[tp]] <- ""
      }
    }
    sig_matrix <- sig_matrix[rownames(gsea_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

    p_heatmap <- pheatmap(
      gsea_wide,
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("blue", "white", "red"))(100),
      breaks = seq(-3, 3, length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 14,
      main = "TE GSEA NES - si3d Hypoxia - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 8,
      fontsize_col = 10,
      cellwidth = 40,
      cellheight = 12
    )

    pdf(here("plots/gsea_te_si3d_hypoxia_heatmap.pdf"), width = 8, height = 10)
    print(p_heatmap)
    dev.off()

    cat("si3d Hypoxia TE heatmap created\n")
  } else {
    cat("No significant pathways found\n")
  }
} else {
  cat("No GSEA results available\n")
}
```
# Hallmark Hypoxia Pathway - Running Enrichment Plots

This section creates classic GSEA running enrichment plots specifically for the HALLMARK_HYPOXIA pathway, showing TE-based enrichment/depletion across treatments and timepoints.

## Normoxia - TE Changes in Hallmark Hypoxia

```{r hallmark_hypoxia_te_normoxia, fig.width=14, fig.height=9}
# Pathway of interest
pathway_id <- "HALLMARK_HYPOXIA"

# Create individual plots for each treatment/timepoint
plots_list <- list()

# si3e normoxia - 1hr
if (pathway_id %in% gsea_te_si3e_norm_1hr@result$ID) {
  plots_list[["si3e_1hr"]] <- gseaplot2(gsea_te_si3e_norm_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 1hr",
                                        color = "green3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3e_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 1hr") +
    theme_void()
}

# si3e normoxia - 4hr
if (pathway_id %in% gsea_te_si3e_norm_4hr@result$ID) {
  plots_list[["si3e_4hr"]] <- gseaplot2(gsea_te_si3e_norm_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 4hr",
                                        color = "green3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3e_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 4hr") +
    theme_void()
}

# si3e normoxia - 24hr
if (pathway_id %in% gsea_te_si3e_norm_24hr@result$ID) {
  plots_list[["si3e_24hr"]] <- gseaplot2(gsea_te_si3e_norm_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3e - 24hr",
                                         color = "green3",
                                         base_size = 11,
                                         pvalue_table = TRUE)
} else {
  plots_list[["si3e_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 24hr") +
    theme_void()
}

# si3d normoxia - 1hr
if (pathway_id %in% gsea_te_si3d_norm_1hr@result$ID) {
  plots_list[["si3d_1hr"]] <- gseaplot2(gsea_te_si3d_norm_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 1hr",
                                        color = "purple3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3d_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 1hr") +
    theme_void()
}

# si3d normoxia - 4hr
if (pathway_id %in% gsea_te_si3d_norm_4hr@result$ID) {
  plots_list[["si3d_4hr"]] <- gseaplot2(gsea_te_si3d_norm_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 4hr",
                                        color = "purple3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3d_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 4hr") +
    theme_void()
}

# si3d normoxia - 24hr
if (pathway_id %in% gsea_te_si3d_norm_24hr@result$ID) {
  plots_list[["si3d_24hr"]] <- gseaplot2(gsea_te_si3d_norm_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3d - 24hr",
                                         color = "purple3",
                                         base_size = 11,
                                         pvalue_table = TRUE)
} else {
  plots_list[["si3d_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 24hr") +
    theme_void()
}

# Combine plots - 2 rows (treatments) × 3 columns (timepoints)
combined_normoxia <- wrap_plots(
  plots_list[["si3e_1hr"]], plots_list[["si3e_4hr"]], plots_list[["si3e_24hr"]],
  plots_list[["si3d_1hr"]], plots_list[["si3d_4hr"]], plots_list[["si3d_24hr"]],
  ncol = 3, nrow = 2
)

print(combined_normoxia)

# Save
ggsave(here("plots/hallmark_hypoxia_te_normoxia.pdf"),
       combined_normoxia,
       width = 14, height = 9)

cat("Normoxia Hallmark Hypoxia plot created and saved\n")
```

## Hypoxia - TE Changes in Hallmark Hypoxia

```{r hallmark_hypoxia_te_hypoxia, fig.width=14, fig.height=9}
# Pathway of interest
pathway_id <- "HALLMARK_HYPOXIA"

# Create individual plots for each treatment/timepoint
plots_list <- list()

# si3e hypoxia - 1hr
if (pathway_id %in% gsea_te_si3e_hyp_1hr@result$ID) {
  plots_list[["si3e_1hr"]] <- gseaplot2(gsea_te_si3e_hyp_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 1hr",
                                        color = "green3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3e_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 1hr") +
    theme_void()
}

# si3e hypoxia - 4hr
if (pathway_id %in% gsea_te_si3e_hyp_4hr@result$ID) {
  plots_list[["si3e_4hr"]] <- gseaplot2(gsea_te_si3e_hyp_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3e - 4hr",
                                        color = "green3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3e_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 4hr") +
    theme_void()
}

# si3e hypoxia - 24hr
if (pathway_id %in% gsea_te_si3e_hyp_24hr@result$ID) {
  plots_list[["si3e_24hr"]] <- gseaplot2(gsea_te_si3e_hyp_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3e - 24hr",
                                         color = "green3",
                                         base_size = 11,
                                         pvalue_table = TRUE)
} else {
  plots_list[["si3e_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3e - 24hr") +
    theme_void()
}

# si3d hypoxia - 1hr
if (pathway_id %in% gsea_te_si3d_hyp_1hr@result$ID) {
  plots_list[["si3d_1hr"]] <- gseaplot2(gsea_te_si3d_hyp_1hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 1hr",
                                        color = "purple3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3d_1hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 1hr") +
    theme_void()
}

# si3d hypoxia - 4hr
if (pathway_id %in% gsea_te_si3d_hyp_4hr@result$ID) {
  plots_list[["si3d_4hr"]] <- gseaplot2(gsea_te_si3d_hyp_4hr,
                                        geneSetID = pathway_id,
                                        title = "si3d - 4hr",
                                        color = "purple3",
                                        base_size = 11,
                                        pvalue_table = TRUE)
} else {
  plots_list[["si3d_4hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 4hr") +
    theme_void()
}

# si3d hypoxia - 24hr
if (pathway_id %in% gsea_te_si3d_hyp_24hr@result$ID) {
  plots_list[["si3d_24hr"]] <- gseaplot2(gsea_te_si3d_hyp_24hr,
                                         geneSetID = pathway_id,
                                         title = "si3d - 24hr",
                                         color = "purple3",
                                         base_size = 11,
                                         pvalue_table = TRUE)
} else {
  plots_list[["si3d_24hr"]] <- ggplot() +
    annotate("text", x = 0.5, y = 0.5, label = "Pathway not found", size = 5) +
    ggtitle("si3d - 24hr") +
    theme_void()
}

# Combine plots - 2 rows (treatments) × 3 columns (timepoints)
combined_hypoxia <- wrap_plots(
  plots_list[["si3e_1hr"]], plots_list[["si3e_4hr"]], plots_list[["si3e_24hr"]],
  plots_list[["si3d_1hr"]], plots_list[["si3d_4hr"]], plots_list[["si3d_24hr"]],
  ncol = 3, nrow = 2
)

print(combined_hypoxia)

# Save
ggsave(here("plots/hallmark_hypoxia_te_hypoxia.pdf"),
       combined_hypoxia,
       width = 14, height = 9)

cat("Hypoxia Hallmark Hypoxia plot created and saved\n")
```

