---
title: "GSEA — si3d Normoxia (all timepoints, time-regressed): Reactome & Literature Translation Pathways (TE)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

GSEA on Translation Efficiency (TE) changes in si3d knockdown cells under normoxia, using all 9 replicates (1hr, 4hr, 24hr) with timepoint regressed out as a covariate.

**Comparison:** si3d vs sictrl in normoxia, all timepoints (time-regressed)
**Data type:** TE log2 fold changes only
**Direction:** positive NES = promoted by eIF3d (te_lfc = sictrl / si3d)

**Gene set collections:**
1. Reactome pathways (MSigDB C2:CP:REACTOME)
2. 25 literature translation signatures

```{r load_libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(pheatmap)
library(writexl)
library(enrichplot)
library(patchwork)

cat("Libraries loaded successfully\n")
```

# Load Gene Sets

## Reactome Pathways

```{r load_reactome}
reactome_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
reactome_term2gene <- reactome_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Reactome gene sets:\n")
cat("Number of pathways:", length(unique(reactome_term2gene$gs_name)), "\n")
```

## Literature Translation Signatures

```{r load_lit_sigs}
lit_sigs <- read_csv(here("accessories", "translation_signatures_literature_1_9_25.csv"))
lit_term2gene <- lit_sigs %>%
  dplyr::select(gs_name, db_gene_symbol) %>%
  rename(gene_symbol = db_gene_symbol)

cat("Loaded literature translation signatures:\n")
cat("Number of gene sets:", length(unique(lit_term2gene$gs_name)), "\n")
print(unique(lit_term2gene$gs_name))
```

# Load Data

## All Timepoints (Time-Regressed)

```{r load_data}
te_si3d_norm <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("si3d Normoxia (all timepoints) TE data loaded:", nrow(te_si3d_norm), "genes\n")
```

## Individual Timepoints

```{r load_data_timepoints}
te_si3d_norm_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_4hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

te_si3d_norm_24hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  dplyr::select(gene_id, te_lfc, symbol) %>%
  rename(log2FoldChange = te_lfc)

cat("1hr:", nrow(te_si3d_norm_1hr), "genes | 4hr:", nrow(te_si3d_norm_4hr), "genes | 24hr:", nrow(te_si3d_norm_24hr), "genes\n")
```

# Build TE Ranked Lists

```{r build_ranked_list}
make_ranked <- function(df) {
  r <- df$log2FoldChange
  names(r) <- df$symbol
  r <- r[!is.na(names(r))]
  r <- sort(r, decreasing = TRUE)
  r[!duplicated(names(r))]
}

te_ranked       <- make_ranked(te_si3d_norm)
te_ranked_1hr   <- make_ranked(te_si3d_norm_1hr)
te_ranked_4hr   <- make_ranked(te_si3d_norm_4hr)
te_ranked_24hr  <- make_ranked(te_si3d_norm_24hr)

cat("Ranked list sizes — all:", length(te_ranked),
    "| 1hr:", length(te_ranked_1hr),
    "| 4hr:", length(te_ranked_4hr),
    "| 24hr:", length(te_ranked_24hr), "\n")
```

# GSEA — TE × Reactome

## All Timepoints (Time-Regressed)

```{r gsea_te_reactome}
set.seed(9)
gsea_te_reactome <- GSEA(
  geneList     = te_ranked,
  eps          = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  minGSSize    = 20,
  maxGSSize    = 1000,
  TERM2GENE    = reactome_term2gene
)

if (nrow(gsea_te_reactome@result) > 0) {
  gsea_te_reactome_results <- gsea_te_reactome@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename(leading_edge = core_enrichment) %>%
    mutate(
      leading_edge_count    = sapply(strsplit(leading_edge, "/"), length),
      leading_edge_fraction = leading_edge_count / setSize
    )

  gsea_te_reactome_results[] <- lapply(gsea_te_reactome_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE × Reactome — results:\n")
  cat("Total pathways tested:", nrow(gsea_te_reactome_results), "\n")
  cat("Significant (FDR < 0.05):", sum(gsea_te_reactome_results$p.adjust < 0.05), "\n\n")
  print(gsea_te_reactome_results %>% filter(p.adjust < 0.05) %>% arrange(desc(abs(NES))) %>% head(15))
} else {
  gsea_te_reactome_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(),
    p.adjust = numeric(), setSize = integer(), leading_edge = character(),
    leading_edge_count = integer(), leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE × Reactome — No pathways enriched\n")
}
```

# GSEA — TE × Literature Translation Signatures

```{r gsea_te_lit}
set.seed(9)
gsea_te_lit <- GSEA(
  geneList      = te_ranked,
  eps           = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff  = 1,
  minGSSize     = 5,
  maxGSSize     = 5000,
  TERM2GENE     = lit_term2gene
)

if (nrow(gsea_te_lit@result) > 0) {
  gsea_te_lit_results <- gsea_te_lit@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename(leading_edge = core_enrichment) %>%
    mutate(
      leading_edge_count    = sapply(strsplit(leading_edge, "/"), length),
      leading_edge_fraction = leading_edge_count / setSize
    )

  gsea_te_lit_results[] <- lapply(gsea_te_lit_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE × Literature — results:\n")
  cat("Total gene sets tested:", nrow(gsea_te_lit_results), "\n")
  cat("Significant (FDR < 0.05):", sum(gsea_te_lit_results$p.adjust < 0.05), "\n\n")
  print(gsea_te_lit_results %>% arrange(desc(NES)))
} else {
  gsea_te_lit_results <- data.frame(
    ID = character(), NES = numeric(), pvalue = numeric(),
    p.adjust = numeric(), setSize = integer(), leading_edge = character(),
    leading_edge_count = integer(), leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE × Literature — No gene sets enriched\n")
}
```

# Visualizations

## Reactome: Top Significant Pathways

```{r plot_reactome, fig.width=12, fig.height=14}
if (nrow(gsea_te_reactome_results) > 0) {
  plot_data <- gsea_te_reactome_results %>%
    filter(abs(NES) >= 1, p.adjust < 0.05) %>%
    arrange(NES) %>%
    mutate(
      label = str_replace_all(ID, "REACTOME_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
      label = fct_inorder(label)
    )

  if (nrow(plot_data) > 0) {
    p_reactome <- ggplot(plot_data, aes(x = NES, y = label)) +
      geom_col(fill = "#D73027") +
      geom_vline(xintercept = 0, linewidth = 0.4) +
      labs(
        title    = "GSEA TE × Reactome — si3d Normoxia (all timepoints)",
        subtitle = "|NES| ≥ 1 and FDR < 0.05",
        x        = "Normalized Enrichment Score (NES)",
        y        = NULL
      ) +
      theme_classic(base_size = 11) +
      theme(axis.text.y = element_text(size = 9))
    print(p_reactome)
    ggsave(here("plots", "gsea_te_si3d_normoxia_alltimepoints_reactome_bar.pdf"), p_reactome, width = 12, height = 14)
  } else {
    cat("No Reactome pathways with |NES| >= 1.5 to plot.\n")
  }
} else {
  cat("No Reactome GSEA results to plot.\n")
}

```

## Reactome GSEA — Individual Timepoints

```{r gsea_te_reactome_timepoints}
run_gsea_reactome <- function(ranked, label) {
  set.seed(9)
  res <- GSEA(
    geneList      = ranked,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 20,
    maxGSSize     = 1000,
    TERM2GENE     = reactome_term2gene
  )
  if (nrow(res@result) > 0) {
    out <- res@result %>%
      dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
      rename(leading_edge = core_enrichment) %>%
      mutate(
        leading_edge_count    = sapply(strsplit(leading_edge, "/"), length),
        leading_edge_fraction = leading_edge_count / setSize
      )
    out[] <- lapply(out, function(x) if (is.numeric(x)) round(x, 4) else x)
    cat(label, "— significant (FDR < 0.05):", sum(out$p.adjust < 0.05), "\n")
    out
  } else {
    cat(label, "— no pathways enriched\n")
    data.frame(ID = character(), NES = numeric(), pvalue = numeric(),
               p.adjust = numeric(), setSize = integer(), leading_edge = character(),
               leading_edge_count = integer(), leading_edge_fraction = numeric(),
               stringsAsFactors = FALSE)
  }
}

gsea_te_reactome_1hr_results  <- run_gsea_reactome(te_ranked_1hr,  "1hr")
gsea_te_reactome_4hr_results  <- run_gsea_reactome(te_ranked_4hr,  "4hr")
gsea_te_reactome_24hr_results <- run_gsea_reactome(te_ranked_24hr, "24hr")
```

```{r plot_reactome_timepoints, fig.width=12, fig.height=14}
plot_reactome_bar <- function(results, title) {
  plot_data <- results %>%
    filter(abs(NES) >= 1, p.adjust < 0.05) %>%
    arrange(NES) %>%
    mutate(
      label = str_replace_all(ID, "REACTOME_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
      label = fct_inorder(label)
    )

  if (nrow(plot_data) == 0) {
    cat("No pathways with |NES| >= 1.5 and FDR < 0.05 for:", title, "\n")
    return(invisible(NULL))
  }

  ggplot(plot_data, aes(x = NES, y = label)) +
    geom_col(fill = "#D73027") +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    labs(
      title    = paste("GSEA TE × Reactome —", title),
      subtitle = "|NES| ≥ 1 and FDR < 0.05",
      x        = "Normalized Enrichment Score (NES)",
      y        = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
}

p_1hr  <- plot_reactome_bar(gsea_te_reactome_1hr_results,  "si3d Normoxia 1hr")
p_4hr  <- plot_reactome_bar(gsea_te_reactome_4hr_results,  "si3d Normoxia 4hr")
p_24hr <- plot_reactome_bar(gsea_te_reactome_24hr_results, "si3d Normoxia 24hr")

if (!is.null(p_1hr))  { print(p_1hr);  ggsave(here("plots", "gsea_te_si3d_normoxia_1hr_reactome_bar.pdf"),  p_1hr,  width = 12, height = 14) }
if (!is.null(p_4hr))  { print(p_4hr);  ggsave(here("plots", "gsea_te_si3d_normoxia_4hr_reactome_bar.pdf"),  p_4hr,  width = 12, height = 14) }
if (!is.null(p_24hr)) { print(p_24hr); ggsave(here("plots", "gsea_te_si3d_normoxia_24hr_reactome_bar.pdf"), p_24hr, width = 12, height = 14) }
```

## Reactome: Dot Plot

```{r plot_reactome_dot, fig.width=11, fig.height=8}
if (nrow(gsea_te_reactome_results) > 0) {
  sig_reactome <- gsea_te_reactome_results %>%
    filter(abs(NES) >= 1, p.adjust < 0.05) %>%
    arrange(desc(abs(NES)))

  if (nrow(sig_reactome) > 0) {
    p_dot <- ggplot(sig_reactome,
                    aes(x = NES,
                        y = fct_reorder(str_replace_all(ID, "REACTOME_", "") %>%
                                          str_replace_all("_", " ") %>%
                                          str_to_title(), NES),
                        size = leading_edge_fraction,
                        color = p.adjust)) +
      geom_point() +
      scale_color_gradient(low = "#D73027", high = "#abd9e9", name = "FDR") +
      scale_size_continuous(name = "Leading-edge\nfraction", range = c(2, 8)) +
      geom_vline(xintercept = 0, linewidth = 0.4, linetype = "dashed") +
      labs(
        title = "GSEA TE × Reactome — si3d Normoxia (all timepoints)",
        x     = "NES",
        y     = NULL
      ) +
      theme_classic(base_size = 11) +
      theme(axis.text.y = element_text(size = 9))
    print(p_dot)
    ggsave(here("plots", "gsea_te_si3d_normoxia_alltimepoints_reactome_dot.pdf"), p_dot, width = 11, height = 8)
  }
}
```

## Literature: All Timepoints (Time-Regressed)

```{r plot_lit, fig.width=10, fig.height=6}
plot_lit_bar <- function(results, title, outfile) {
  plot_data <- results %>%
    filter(abs(NES) >= 1, p.adjust < 0.05) %>%
    arrange(NES) %>%
    mutate(ID = fct_inorder(ID))

  if (nrow(plot_data) == 0) {
    cat("No literature gene sets with |NES| >= 1 and FDR < 0.05 for:", title, "\n")
    return(invisible(NULL))
  }

  p <- ggplot(plot_data, aes(x = NES, y = ID)) +
    geom_col(fill = "#D73027") +
    geom_vline(xintercept = 0, linewidth = 0.4) +
    labs(
      title    = paste("GSEA TE × Literature —", title),
      subtitle = "|NES| ≥ 1 and FDR < 0.05",
      x        = "Normalized Enrichment Score (NES)",
      y        = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
  print(p)
  ggsave(here("plots", outfile), p, width = 10, height = 6)
  p
}

if (nrow(gsea_te_lit_results) > 0) {
  plot_lit_bar(gsea_te_lit_results, "si3d Normoxia (all timepoints)",
               "gsea_te_si3d_normoxia_alltimepoints_lit.pdf")
} else {
  cat("No literature GSEA results to plot.\n")
}
```

## Literature: Individual Timepoints

```{r gsea_te_lit_timepoints}
run_gsea_lit <- function(ranked, label) {
  set.seed(9)
  res <- GSEA(
    geneList      = ranked,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 5,
    maxGSSize     = 5000,
    TERM2GENE     = lit_term2gene
  )
  if (nrow(res@result) > 0) {
    out <- res@result %>%
      dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
      rename(leading_edge = core_enrichment) %>%
      mutate(
        leading_edge_count    = sapply(strsplit(leading_edge, "/"), length),
        leading_edge_fraction = leading_edge_count / setSize
      )
    out[] <- lapply(out, function(x) if (is.numeric(x)) round(x, 4) else x)
    cat(label, "— significant (FDR < 0.05):", sum(out$p.adjust < 0.05), "\n")
    list(results = out, object = res)
  } else {
    cat(label, "— no gene sets enriched\n")
    list(
      results = data.frame(ID = character(), NES = numeric(), pvalue = numeric(),
                           p.adjust = numeric(), setSize = integer(), leading_edge = character(),
                           leading_edge_count = integer(), leading_edge_fraction = numeric(),
                           stringsAsFactors = FALSE),
      object = res
    )
  }
}

tmp_1hr  <- run_gsea_lit(te_ranked_1hr,  "1hr")
tmp_4hr  <- run_gsea_lit(te_ranked_4hr,  "4hr")
tmp_24hr <- run_gsea_lit(te_ranked_24hr, "24hr")

gsea_te_lit_1hr_results  <- tmp_1hr$results
gsea_te_lit_4hr_results  <- tmp_4hr$results
gsea_te_lit_24hr_results <- tmp_24hr$results

gsea_te_lit_1hr  <- tmp_1hr$object
gsea_te_lit_4hr  <- tmp_4hr$object
gsea_te_lit_24hr <- tmp_24hr$object
```

```{r plot_lit_timepoints, fig.width=10, fig.height=6}
plot_lit_bar(gsea_te_lit_1hr_results,  "si3d Normoxia 1hr",  "gsea_te_si3d_normoxia_1hr_lit.pdf")
plot_lit_bar(gsea_te_lit_4hr_results,  "si3d Normoxia 4hr",  "gsea_te_si3d_normoxia_4hr_lit.pdf")
plot_lit_bar(gsea_te_lit_24hr_results, "si3d Normoxia 24hr", "gsea_te_si3d_normoxia_24hr_lit.pdf")
```

## Running Enrichment Plots — TOP Motif mRNAs

```{r enrichment_plots_top_motif, fig.width=8, fig.height=5}
top_motif_id <- "TOP motif mRNAs"

top_motif_id <- "TOP motif mRNAs"

# All timepoints (time-regressed)
p_top_all <- gseaplot2(gsea_te_lit, geneSetID = top_motif_id, subplots = 1:2, pvalue_table = FALSE,
                       title = sprintf("TOP motif mRNAs — si3d Normoxia (all timepoints)\nNES = %.3f | FDR = %.3f",
                                       gsea_te_lit@result$NES[gsea_te_lit@result$ID == top_motif_id],
                                       gsea_te_lit@result$p.adjust[gsea_te_lit@result$ID == top_motif_id]))
print(p_top_all)
ggsave(here("plots", "gsea_top_motif_si3d_norm_alltimepoints.pdf"), p_top_all, width = 8, height = 5)

# 1hr
p_top_1hr <- gseaplot2(gsea_te_lit_1hr, geneSetID = top_motif_id, subplots = 1:2, pvalue_table = FALSE,
                       title = sprintf("TOP motif mRNAs — si3d Normoxia 1hr\nNES = %.3f | FDR = %.3f",
                                       gsea_te_lit_1hr@result$NES[gsea_te_lit_1hr@result$ID == top_motif_id],
                                       gsea_te_lit_1hr@result$p.adjust[gsea_te_lit_1hr@result$ID == top_motif_id]))
print(p_top_1hr)
ggsave(here("plots", "gsea_top_motif_si3d_norm_1hr.pdf"), p_top_1hr, width = 8, height = 5)

# 4hr
p_top_4hr <- gseaplot2(gsea_te_lit_4hr, geneSetID = top_motif_id, subplots = 1:2, pvalue_table = FALSE,
                       title = sprintf("TOP motif mRNAs — si3d Normoxia 4hr\nNES = %.3f | FDR = %.3f",
                                       gsea_te_lit_4hr@result$NES[gsea_te_lit_4hr@result$ID == top_motif_id],
                                       gsea_te_lit_4hr@result$p.adjust[gsea_te_lit_4hr@result$ID == top_motif_id]))
print(p_top_4hr)
ggsave(here("plots", "gsea_top_motif_si3d_norm_4hr.pdf"), p_top_4hr, width = 8, height = 5)

# 24hr
p_top_24hr <- gseaplot2(gsea_te_lit_24hr, geneSetID = top_motif_id, subplots = 1:2, pvalue_table = FALSE,
                        title = sprintf("TOP motif mRNAs — si3d Normoxia 24hr\nNES = %.3f | FDR = %.3f",
                                        gsea_te_lit_24hr@result$NES[gsea_te_lit_24hr@result$ID == top_motif_id],
                                        gsea_te_lit_24hr@result$p.adjust[gsea_te_lit_24hr@result$ID == top_motif_id]))
print(p_top_24hr)
ggsave(here("plots", "gsea_top_motif_si3d_norm_24hr.pdf"), p_top_24hr, width = 8, height = 5)
```

# Overlap Analysis — eIF3d-Promoted Genes × Literature Signatures

Hypergeometric test (ORA) asking whether each literature gene set is enriched among genes significantly promoted by eIF3d in the time-regressed normoxia analysis (TE LFC > 0.5, FDR < 0.05).

```{r overlap_promoted_lit}
# Load full data with te_padj
te_si3d_norm_full <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv"),
  show_col_types = FALSE
) %>%
  filter(!is.na(symbol))

universe_genes  <- te_si3d_norm_full$symbol
promoted_genes  <- te_si3d_norm_full %>%
  filter(te_lfc > 0.5, te_padj < 0.05) %>%
  pull(symbol)

cat("Universe:", length(universe_genes), "genes\n")
cat("eIF3d-promoted (LFC > 0.5, FDR < 0.05):", length(promoted_genes), "genes\n")

ora_lit <- enricher(
  gene          = promoted_genes,
  universe      = universe_genes,
  TERM2GENE     = lit_term2gene,
  pAdjustMethod = "fdr",
  pvalueCutoff  = 1,
  qvalueCutoff  = 1,
  minGSSize     = 5,
  maxGSSize     = 5000
)

ora_lit_results <- ora_lit@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
  mutate(
    gene_ratio_num = sapply(strsplit(GeneRatio, "/"), function(x) as.integer(x[1]) / as.integer(x[2])),
    bg_ratio_num   = sapply(strsplit(BgRatio,   "/"), function(x) as.integer(x[1]) / as.integer(x[2])),
    fold_enrichment = gene_ratio_num / bg_ratio_num
  )

ora_lit_results[] <- lapply(ora_lit_results, function(x) if (is.numeric(x)) round(x, 4) else x)

cat("\nSignificant (FDR < 0.05):", sum(ora_lit_results$p.adjust < 0.05), "\n")
print(ora_lit_results %>% arrange(p.adjust) %>% dplyr::select(ID, Count, fold_enrichment, pvalue, p.adjust))
```

```{r plot_overlap_lit, fig.width=10, fig.height=6}
plot_data_ora <- ora_lit_results %>%
  filter(p.adjust < 0.05) %>%
  arrange(fold_enrichment) %>%
  mutate(ID = fct_inorder(ID))

if (nrow(plot_data_ora) == 0) {
  cat("No significant overlaps (FDR < 0.05).\n")
} else {

  # 1. Dot plot: fold enrichment × FDR × overlap count
  p_dot <- ggplot(plot_data_ora, aes(x = fold_enrichment, y = ID, size = Count, color = p.adjust)) +
    geom_point() +
    scale_color_gradient(low = "#D73027", high = "#abd9e9", name = "FDR") +
    scale_size_continuous(name = "Overlap\n(genes)", range = c(3, 9)) +
    geom_vline(xintercept = 1, linewidth = 0.4, linetype = "dashed") +
    labs(
      title    = "ORA — eIF3d-promoted genes × Literature Signatures",
      subtitle = "x = fold enrichment over background | size = overlap count | color = FDR",
      x        = "Fold enrichment",
      y        = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
  print(p_dot)
  ggsave(here("plots", "ora_promoted_lit_dot.pdf"), p_dot, width = 10, height = 6)

  # 2. GeneRatio dot plot (heatmap.Rmd style)
  p_generatio <- ggplot(plot_data_ora,
                        aes(y = reorder(ID, gene_ratio_num),
                            x = gene_ratio_num,
                            fill = p.adjust,
                            size = Count)) +
    geom_point(alpha = 0.7, shape = 21, color = "black") +
    scale_size(range = c(3, 10), name = "Gene count") +
    scale_fill_viridis_c(direction = -1, name = "FDR") +
    labs(
      title    = "ORA — eIF3d-promoted genes × Literature Signatures",
      subtitle = "x = fraction of eIF3d-promoted genes in each set | FDR < 0.05",
      x        = "Gene ratio",
      y        = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 9),
          plot.title  = element_text(face = "bold"))
  print(p_generatio)
  ggsave(here("plots", "ora_promoted_lit_generatio.pdf"), p_generatio, width = 10, height = 6)

  # 4. Bar plot: -log10(FDR), bars sorted by significance
  p_fdr <- ora_lit_results %>%
    arrange(p.adjust) %>%
    mutate(
      ID  = fct_inorder(ID),
      sig = p.adjust < 0.05
    ) %>%
    ggplot(aes(x = -log10(p.adjust), y = ID, fill = sig)) +
    geom_col() +
    scale_fill_manual(values = c("TRUE" = "#D73027", "FALSE" = "grey70"), guide = "none") +
    geom_vline(xintercept = -log10(0.05), linewidth = 0.4, linetype = "dashed") +
    labs(
      title    = "ORA — eIF3d-promoted genes × Literature Signatures",
      subtitle = "Dashed line = FDR 0.05 | red = significant",
      x        = "-log10(FDR)",
      y        = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
  print(p_fdr)
  ggsave(here("plots", "ora_promoted_lit_fdr_bar.pdf"), p_fdr, width = 10, height = 6)

  # 5. Bar plot: overlap count for significant sets
  p_count <- plot_data_ora %>%
    arrange(Count) %>%
    mutate(ID = fct_inorder(ID)) %>%
    ggplot(aes(x = Count, y = ID)) +
    geom_col(fill = "#D73027") +
    labs(
      title    = "ORA — eIF3d-promoted genes × Literature Signatures",
      subtitle = "Number of overlapping genes (FDR < 0.05 sets only)",
      x        = "Overlapping genes (n)",
      y        = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.y = element_text(size = 9))
  print(p_count)
  ggsave(here("plots", "ora_promoted_lit_count_bar.pdf"), p_count, width = 10, height = 6)

}
```

# Save Outputs

```{r save_outputs}
write_xlsx(
  list(
    TE_Reactome = gsea_te_reactome_results,
    TE_Lit      = gsea_te_lit_results
  ),
  here("output", "gsea_te_si3d_normoxia_alltimepoints_reactome_lit.xlsx")
)

cat("Results saved to: output/gsea_te_si3d_normoxia_alltimepoints_reactome_lit.xlsx\n")
cat("Sheets: TE_Reactome (", nrow(gsea_te_reactome_results), "rows),",
    "TE_Lit (", nrow(gsea_te_lit_results), "rows)\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
