---
title: "TE Heatmap - All Conditions and Timepoints"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

# Overview

This analysis creates a heatmap of translation efficiency (TE) values across all conditions, treatments, and timepoints. Z-scores are calculated for each gene across all samples to show relative changes in TE.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load_libraries, message=TRUE, warning=TRUE}
library(here)
library(tidyverse)
library(pheatmap)
library(clusterProfiler)
library(msigdbr)
library(writexl)

cat("Libraries loaded successfully\n")
cat("clusterProfiler version:", as.character(packageVersion("clusterProfiler")), "\n")
```

# Load Hallmark Gene Sets

Load MSigDB Hallmark gene sets for pathway enrichment analysis.

```{r load_hallmark, message=TRUE, warning=TRUE}
# Load MSigDB Hallmark gene sets for human
cat("Loading Hallmark gene sets...\n")
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")
cat("Hallmark sets loaded, rows:", nrow(hallmark_sets), "\n")

# Create gene set format for clusterProfiler (TERM2GENE)
hallmark_term2gene <- hallmark_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("Created TERM2GENE format\n")
cat("Number of pathways:", length(unique(hallmark_term2gene$gs_name)), "\n")
cat("Total gene-pathway pairs:", nrow(hallmark_term2gene), "\n")
```

# Load TE Data

Load the complete TE lists for all timepoints.

```{r load_te_data}
# Load TE data for all timepoints (spike-ins already removed in calculate_te_from_normalized_counts.Rmd)
te_1hr <- read_csv(here("output", "complete_te_list_all_1hr.csv"))
te_4hr <- read_csv(here("output", "complete_te_list_all_4hr.csv"))
te_24hr <- read_csv(here("output", "complete_te_list_all_24hr.csv"))

cat("Loaded TE data:\n")
cat("1hr:", nrow(te_1hr), "genes\n")
cat("4hr:", nrow(te_4hr), "genes\n")
cat("24hr:", nrow(te_24hr), "genes\n")
```

# Extract TE Columns

Extract only the TE columns from each timepoint and combine them.

```{r extract_te_columns}
# Get TE columns (those containing "_TE_")
te_cols_1hr <- grep("_TE_", colnames(te_1hr), value = TRUE)
te_cols_4hr <- grep("_TE_", colnames(te_4hr), value = TRUE)
te_cols_24hr <- grep("_TE_", colnames(te_24hr), value = TRUE)

# Extract TE values
te_matrix_1hr <- te_1hr %>%
  dplyr::select(gene_id, all_of(te_cols_1hr)) %>%
  column_to_rownames("gene_id")

te_matrix_4hr <- te_4hr %>%
  dplyr::select(gene_id, all_of(te_cols_4hr)) %>%
  column_to_rownames("gene_id")

te_matrix_24hr <- te_24hr %>%
  dplyr::select(gene_id, all_of(te_cols_24hr)) %>%
  column_to_rownames("gene_id")

cat("TE matrix dimensions:\n")
cat("1hr:", dim(te_matrix_1hr), "\n")
cat("4hr:", dim(te_matrix_4hr), "\n")
cat("24hr:", dim(te_matrix_24hr), "\n")
```

# Order Columns

Order columns in the specified order: sictrl -> si3d -> si3e, within each: normoxia -> hypoxia, within each: 1hr -> 4hr -> 24hr.

```{r order_columns}
# Combine all timepoints first
# Ensure all matrices have the same genes
common_genes <- Reduce(intersect, list(rownames(te_matrix_1hr),
                                        rownames(te_matrix_4hr),
                                        rownames(te_matrix_24hr)))

te_matrix_1hr <- te_matrix_1hr[common_genes, ]
te_matrix_4hr <- te_matrix_4hr[common_genes, ]
te_matrix_24hr <- te_matrix_24hr[common_genes, ]

cat("Number of common genes:", length(common_genes), "\n")

# Create a combined data frame with all column info
all_cols <- data.frame(
  col_name = c(colnames(te_matrix_1hr), colnames(te_matrix_4hr), colnames(te_matrix_24hr)),
  timepoint = c(rep("1hr", ncol(te_matrix_1hr)),
                rep("4hr", ncol(te_matrix_4hr)),
                rep("24hr", ncol(te_matrix_24hr))),
  stringsAsFactors = FALSE
)

# Parse column names to extract condition and treatment
# Format: condition_treatment_timepoint_rep_type_id
all_cols$condition <- sapply(strsplit(all_cols$col_name, "_"), `[`, 1)
all_cols$treatment <- sapply(strsplit(all_cols$col_name, "_"), `[`, 2)
all_cols$rep <- sapply(strsplit(all_cols$col_name, "_"), `[`, 4)

# Set factor levels for ordering
all_cols$treatment <- factor(all_cols$treatment, levels = c("sictrl", "si3d", "si3e"))
all_cols$condition <- factor(all_cols$condition, levels = c("normoxia", "hypoxia"))
all_cols$timepoint <- factor(all_cols$timepoint, levels = c("1hr", "4hr", "24hr"))

# Order by treatment, then condition, then timepoint, then replicate
all_cols_ordered <- all_cols %>%
  arrange(treatment, condition, timepoint, rep)

# Create matrices for each timepoint with ordered columns
te_matrix_1hr_ordered <- te_matrix_1hr[, all_cols_ordered$col_name[all_cols_ordered$timepoint == "1hr"]]
te_matrix_4hr_ordered <- te_matrix_4hr[, all_cols_ordered$col_name[all_cols_ordered$timepoint == "4hr"]]
te_matrix_24hr_ordered <- te_matrix_24hr[, all_cols_ordered$col_name[all_cols_ordered$timepoint == "24hr"]]

# Combine into single matrix in the correct order
te_matrix_all <- cbind(te_matrix_1hr_ordered, te_matrix_4hr_ordered, te_matrix_24hr_ordered)

# Reorder the entire matrix according to all_cols_ordered
te_matrix_all <- te_matrix_all[, all_cols_ordered$col_name]

cat("Combined TE matrix dimensions:", dim(te_matrix_all), "\n")
cat("Column order preview (first 6):\n")
print(head(colnames(te_matrix_all)))
```

# Calculate Z-scores

Calculate z-scores for each gene across all samples.

```{r calculate_zscores}
# Calculate z-scores row-wise (for each gene across all samples)
te_matrix_zscore <- t(scale(t(te_matrix_all)))

# Check for genes with zero variance (will have NaN z-scores)
genes_with_nan <- rownames(te_matrix_zscore)[rowSums(is.na(te_matrix_zscore)) > 0]
cat("Number of genes with NaN z-scores (zero variance):", length(genes_with_nan), "\n")

# Remove genes with NaN z-scores
if (length(genes_with_nan) > 0) {
  te_matrix_zscore <- te_matrix_zscore[!rownames(te_matrix_zscore) %in% genes_with_nan, ]
  cat("Removed", length(genes_with_nan), "genes with zero variance\n")
}

cat("Final matrix for heatmap:", dim(te_matrix_zscore), "\n")
```

# Create Column Annotations

Create annotations to show the groupings in the heatmap.

```{r create_annotations}
# Create annotation data frame using the ordered column info
col_anno <- data.frame(
  Treatment = all_cols_ordered$treatment,
  Condition = all_cols_ordered$condition,
  Timepoint = all_cols_ordered$timepoint,
  row.names = all_cols_ordered$col_name
)

# Ensure factors are in the correct order
col_anno$Treatment <- factor(col_anno$Treatment, levels = c("sictrl", "si3d", "si3e"))
col_anno$Condition <- factor(col_anno$Condition, levels = c("normoxia", "hypoxia"))
col_anno$Timepoint <- factor(col_anno$Timepoint, levels = c("1hr", "4hr", "24hr"))

# Define colors for annotations
anno_colors <- list(
  Treatment = c("sictrl" = "#808080", "si3d" = "#E69F00", "si3e" = "#56B4E9"),
  Condition = c("normoxia" = "#009E73", "hypoxia" = "#CC79A7"),
  Timepoint = c("1hr" = "#E0E0E0", "4hr" = "#A0A0A0", "24hr" = "#606060")
)

cat("Column annotations created\n")
cat("Annotation structure:\n")
print(head(col_anno))
```

# Create Heatmap

Generate the heatmap with pheatmap.

```{r create_heatmap, fig.width=12, fig.height=10}
# Create heatmap
pheatmap(
  te_matrix_zscore,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores Across All Conditions and Timepoints",
  fontsize = 10,
  border_color = NA
)
```

# Save Heatmap

Save the heatmap as a PDF.

```{r save_heatmap}
# Save as PDF
pdf(here("plots", "te_heatmap_all_conditions.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores Across All Conditions and Timepoints",
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Heatmap saved to plots/te_heatmap_all_conditions.pdf\n")
```

# Optional: Filter to Top Variable Genes

Create a heatmap with only the top most variable genes for better visualization.

```{r filter_top_variable}
# Calculate variance for each gene
gene_vars <- apply(te_matrix_zscore, 1, var, na.rm = TRUE)

# Select top 1000 most variable genes
top_n_genes <- 1000
top_genes <- names(sort(gene_vars, decreasing = TRUE)[1:top_n_genes])

te_matrix_zscore_top <- te_matrix_zscore[top_genes, ]

cat("Selected top", top_n_genes, "most variable genes\n")
```

```{r heatmap_top_variable, fig.width=12, fig.height=10}
# Create heatmap with top variable genes
pheatmap(
  te_matrix_zscore_top,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", top_n_genes, " Most Variable Genes"),
  fontsize = 10,
  border_color = NA
)
```

```{r save_heatmap_top_variable}
# Save as PDF
pdf(here("plots", "te_heatmap_top_variable_genes.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_top,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", top_n_genes, " Most Variable Genes"),
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Top variable genes heatmap saved to plots/te_heatmap_top_variable_genes.pdf\n")
```

# Filter for Genes Most Changed in Acute Hypoxic Response (siCtrl 1hr)

Identify genes with the largest TE changes between normoxia and hypoxia at 1hr in sictrl condition.

```{r filter_acute_hypoxia_response}
# Get normoxia sictrl 1hr samples
normoxia_sictrl_1hr_cols <- all_cols_ordered %>%
  filter(treatment == "sictrl", condition == "normoxia", timepoint == "1hr") %>%
  pull(col_name)

# Get hypoxia sictrl 1hr samples
hypoxia_sictrl_1hr_cols <- all_cols_ordered %>%
  filter(treatment == "sictrl", condition == "hypoxia", timepoint == "1hr") %>%
  pull(col_name)

# Calculate mean TE for each condition
mean_te_normoxia_1hr <- rowMeans(te_matrix_all[, normoxia_sictrl_1hr_cols], na.rm = TRUE)
mean_te_hypoxia_1hr <- rowMeans(te_matrix_all[, hypoxia_sictrl_1hr_cols], na.rm = TRUE)

# Calculate TE change (hypoxia - normoxia)
te_change_1hr <- mean_te_hypoxia_1hr - mean_te_normoxia_1hr

# Get top changed genes
n_top_changed <- 500
top_increased <- names(sort(te_change_1hr, decreasing = TRUE)[1:n_top_changed])
top_decreased <- names(sort(te_change_1hr, decreasing = FALSE)[1:n_top_changed])
top_changed_genes <- c(top_increased, top_decreased)

cat("Genes with largest TE increase in hypoxia (1hr sictrl):", length(top_increased), "\n")
cat("Genes with largest TE decrease in hypoxia (1hr sictrl):", length(top_decreased), "\n")
cat("Total genes for heatmap:", length(top_changed_genes), "\n")

# Create summary data frame
te_change_summary <- data.frame(
  gene_id = names(te_change_1hr),
  te_change = te_change_1hr,
  mean_normoxia = mean_te_normoxia_1hr,
  mean_hypoxia = mean_te_hypoxia_1hr
) %>%
  arrange(desc(abs(te_change)))

cat("\nTop 10 most changed genes:\n")
print(head(te_change_summary, 10))

# Save the TE change summary
write_csv(te_change_summary, here("output", "te_change_sictrl_normoxia_vs_hypoxia_1hr.csv"))
cat("\nSaved TE change summary to output/te_change_sictrl_normoxia_vs_hypoxia_1hr.csv\n")
```

```{r heatmap_acute_hypoxia_response, fig.width=12, fig.height=10}
# Filter z-score matrix to top changed genes
te_matrix_zscore_acute <- te_matrix_zscore[top_changed_genes, ]

# Create heatmap
pheatmap(
  te_matrix_zscore_acute,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", length(top_changed_genes),
                " Genes Changed in Acute Hypoxic Response (siCtrl 1hr)"),
  fontsize = 10,
  border_color = NA
)
```

```{r save_heatmap_acute_hypoxia}
# Save as PDF
pdf(here("plots", "te_heatmap_acute_hypoxia_response.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_acute,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", length(top_changed_genes),
                " Genes Changed in Acute Hypoxic Response (siCtrl 1hr)"),
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Acute hypoxia response heatmap saved to plots/te_heatmap_acute_hypoxia_response.pdf\n")
```

# Filter for Genes with Increased TE in Hypoxia (siCtrl 1hr)

Create a heatmap showing only genes that increase TE in response to hypoxia at 1hr in sictrl.

```{r filter_te_increase_hypoxia}
# Get genes with increased TE in hypoxia (top N)
n_increased <- 500
genes_increased_te <- names(sort(te_change_1hr, decreasing = TRUE)[1:n_increased])

cat("Number of genes with increased TE in hypoxia:", length(genes_increased_te), "\n")
cat("TE change range:", range(te_change_1hr[genes_increased_te]), "\n")

# Create summary for increased genes
te_increase_summary <- te_change_summary %>%
  filter(gene_id %in% genes_increased_te) %>%
  arrange(desc(te_change))

cat("\nTop 10 genes with increased TE in hypoxia:\n")
print(head(te_increase_summary, 10))

# Save the increased TE genes
write_csv(te_increase_summary, here("output", "te_increased_in_hypoxia_sictrl_1hr.csv"))
cat("\nSaved increased TE genes to output/te_increased_in_hypoxia_sictrl_1hr.csv\n")
```

```{r heatmap_te_increase_hypoxia, fig.width=12, fig.height=10}
# Filter z-score matrix to genes with increased TE
te_matrix_zscore_increased <- te_matrix_zscore[genes_increased_te, ]

# Create heatmap

heatmap <- 
  pheatmap(
  te_matrix_zscore_increased,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", n_increased,
                " Genes with Increased TE in Hypoxia (siCtrl 1hr)"),
  fontsize = 10,
  border_color = NA
)
print(heatmap)
```

```{r save_heatmap_te_increase}
# Save as PDF
pdf(here("plots", "te_heatmap_increased_in_hypoxia.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_increased,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - Top ", n_increased,
                " Genes with Increased TE in Hypoxia (siCtrl 1hr)"),
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Increased TE in hypoxia heatmap saved to plots/te_heatmap_increased_in_hypoxia.pdf\n")
```

# Filter for Genes with Hypoxic Response Affected by si3d/si3e

Identify genes that change in hypoxia in sictrl but have reduced/no response in si3d or si3e knockdowns.

```{r calculate_hypoxia_response_all_treatments}
# Get sample columns for each treatment/condition combination at 1hr
normoxia_sictrl_1hr <- all_cols_ordered %>%
  filter(treatment == "sictrl", condition == "normoxia", timepoint == "1hr") %>%
  pull(col_name)

hypoxia_sictrl_1hr <- all_cols_ordered %>%
  filter(treatment == "sictrl", condition == "hypoxia", timepoint == "1hr") %>%
  pull(col_name)

normoxia_si3d_1hr <- all_cols_ordered %>%
  filter(treatment == "si3d", condition == "normoxia", timepoint == "1hr") %>%
  pull(col_name)

hypoxia_si3d_1hr <- all_cols_ordered %>%
  filter(treatment == "si3d", condition == "hypoxia", timepoint == "1hr") %>%
  pull(col_name)

normoxia_si3e_1hr <- all_cols_ordered %>%
  filter(treatment == "si3e", condition == "normoxia", timepoint == "1hr") %>%
  pull(col_name)

hypoxia_si3e_1hr <- all_cols_ordered %>%
  filter(treatment == "si3e", condition == "hypoxia", timepoint == "1hr") %>%
  pull(col_name)

# Calculate hypoxia response (hypoxia - normoxia) for each treatment
hypoxia_response_sictrl <- rowMeans(te_matrix_all[, hypoxia_sictrl_1hr], na.rm = TRUE) -
                           rowMeans(te_matrix_all[, normoxia_sictrl_1hr], na.rm = TRUE)

hypoxia_response_si3d <- rowMeans(te_matrix_all[, hypoxia_si3d_1hr], na.rm = TRUE) -
                         rowMeans(te_matrix_all[, normoxia_si3d_1hr], na.rm = TRUE)

hypoxia_response_si3e <- rowMeans(te_matrix_all[, hypoxia_si3e_1hr], na.rm = TRUE) -
                         rowMeans(te_matrix_all[, normoxia_si3e_1hr], na.rm = TRUE)

# Calculate the difference in hypoxia response
# Positive values = hypoxia response is blunted in knockdown
response_diff_si3d <- hypoxia_response_sictrl - hypoxia_response_si3d
response_diff_si3e <- hypoxia_response_sictrl - hypoxia_response_si3e

cat("Calculated hypoxia response for all treatments\n")
```

```{r filter_blunted_response}
# Find genes with strong hypoxic response in sictrl that is blunted in si3d or si3e
# Criteria:
# 1. Strong hypoxic response in sictrl (TE change > threshold)
# 2. Reduced response in si3d or si3e (difference in response > threshold)

min_sictrl_response <- 0.5  # Minimum TE change in sictrl hypoxia response
min_blunted_diff <- 0.3     # Minimum difference showing blunted response

# Genes with blunted response in si3d
genes_blunted_si3d <- names(which(
  abs(hypoxia_response_sictrl) > min_sictrl_response &
  response_diff_si3d > min_blunted_diff
))

# Genes with blunted response in si3e
genes_blunted_si3e <- names(which(
  abs(hypoxia_response_sictrl) > min_sictrl_response &
  response_diff_si3e > min_blunted_diff
))

# Genes blunted in either si3d or si3e
genes_blunted_either <- unique(c(genes_blunted_si3d, genes_blunted_si3e))

cat("Genes with blunted hypoxic response in si3d:", length(genes_blunted_si3d), "\n")
cat("Genes with blunted hypoxic response in si3e:", length(genes_blunted_si3e), "\n")
cat("Genes with blunted response in either knockdown:", length(genes_blunted_either), "\n")

# Create summary data frame
blunted_response_summary <- data.frame(
  gene_id = names(hypoxia_response_sictrl),
  sictrl_response = hypoxia_response_sictrl,
  si3d_response = hypoxia_response_si3d,
  si3e_response = hypoxia_response_si3e,
  diff_si3d = response_diff_si3d,
  diff_si3e = response_diff_si3e
) %>%
  filter(gene_id %in% genes_blunted_either) %>%
  arrange(desc(pmax(diff_si3d, diff_si3e)))

cat("\nTop 10 genes with most blunted hypoxic response:\n")
print(head(blunted_response_summary, 10))

# Save summary
write_csv(blunted_response_summary, here("output", "genes_with_blunted_hypoxia_response.csv"))
cat("\nSaved blunted response genes to output/genes_with_blunted_hypoxia_response.csv\n")
```

```{r heatmap_blunted_response, fig.width=12, fig.height=10}
# Filter z-score matrix to genes with blunted response
te_matrix_zscore_blunted <- te_matrix_zscore[genes_blunted_either, ]

# Create heatmap
heatmap_blunted <- pheatmap(
  te_matrix_zscore_blunted,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - ", length(genes_blunted_either),
                " Genes with Blunted Hypoxic Response in si3d/si3e"),
  fontsize = 10,
  border_color = NA
)
print(heatmap_blunted)
```

```{r save_heatmap_blunted_response}
# Save as PDF
pdf(here("plots", "te_heatmap_blunted_hypoxia_response.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_blunted,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - ", length(genes_blunted_either),
                " Genes with Blunted Hypoxic Response in si3d/si3e"),
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Blunted hypoxia response heatmap saved to plots/te_heatmap_blunted_hypoxia_response.pdf\n")
```

# Blunted Response Heatmap - Normoxia vs Hypoxia Split

Create the same heatmap but with all normoxia samples on the left and all hypoxia samples on the right.

```{r reorder_columns_by_condition}
# Reorder columns: condition first, then treatment, then timepoint
all_cols_by_condition <- all_cols_ordered %>%
  arrange(condition, treatment, timepoint, rep)

# Reorder the z-score matrix for blunted genes
te_matrix_zscore_blunted_reordered <- te_matrix_zscore_blunted[, all_cols_by_condition$col_name]

# Create new column annotations
col_anno_by_condition <- data.frame(
  Condition = all_cols_by_condition$condition,
  Treatment = all_cols_by_condition$treatment,
  Timepoint = all_cols_by_condition$timepoint,
  row.names = all_cols_by_condition$col_name
)

# Ensure factors are in the correct order
col_anno_by_condition$Condition <- factor(col_anno_by_condition$Condition, levels = c("normoxia", "hypoxia"))
col_anno_by_condition$Treatment <- factor(col_anno_by_condition$Treatment, levels = c("sictrl", "si3d", "si3e"))
col_anno_by_condition$Timepoint <- factor(col_anno_by_condition$Timepoint, levels = c("1hr", "4hr", "24hr"))

cat("Reordered matrix by condition\n")
cat("Column order: normoxia samples (all treatments/timepoints), then hypoxia samples (all treatments/timepoints)\n")
```

```{r heatmap_blunted_by_condition, fig.width=12, fig.height=10}
# Create heatmap with condition-based ordering
heatmap_blunted_cond <- pheatmap(
  te_matrix_zscore_blunted_reordered,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_condition,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - ", length(genes_blunted_either),
                " Genes with Blunted Hypoxic Response (Normoxia | Hypoxia)"),
  fontsize = 10,
  border_color = NA
)
print(heatmap_blunted_cond)
```

```{r save_heatmap_blunted_by_condition}
# Save as PDF
pdf(here("plots", "te_heatmap_blunted_hypoxia_response_by_condition.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_blunted_reordered,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_condition,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = paste0("TE Z-scores - ", length(genes_blunted_either),
                " Genes with Blunted Hypoxic Response (Normoxia | Hypoxia)"),
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("Condition-split blunted hypoxia response heatmap saved to plots/te_heatmap_blunted_hypoxia_response_by_condition.pdf\n")
```

# All Genes Heatmap - Normoxia vs Hypoxia Split

Create a heatmap with all genes, organized with normoxia samples on the left and hypoxia samples on the right.

```{r reorder_all_genes_by_condition}
# Reorder the full z-score matrix by condition
te_matrix_zscore_all_reordered <- te_matrix_zscore[, all_cols_by_condition$col_name]

cat("Reordered full matrix by condition\n")
cat("Total genes:", nrow(te_matrix_zscore_all_reordered), "\n")
cat("Column order: normoxia (all treatments/timepoints) | hypoxia (all treatments/timepoints)\n")
```

```{r heatmap_all_genes_by_condition, fig.width=12, fig.height=10}
# Create heatmap with all genes
heatmap_all_cond <- pheatmap(
  te_matrix_zscore_all_reordered,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_condition,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes (Normoxia | Hypoxia)",
  fontsize = 10,
  border_color = NA
)
print(heatmap_all_cond)
```

```{r save_heatmap_all_genes_by_condition}
# Save as PDF
pdf(here("plots", "te_heatmap_all_genes_by_condition.pdf"), width = 12, height = 10)
pheatmap(
  te_matrix_zscore_all_reordered,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_condition,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes (Normoxia | Hypoxia)",
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("All genes heatmap by condition saved to plots/te_heatmap_all_genes_by_condition.pdf\n")
```

# All Genes Heatmap - All Timepoints, Clustered by siCtrl Response

Create a heatmap with all genes and all timepoints, organized by treatment with normoxia and hypoxia side-by-side, clustered by sictrl hypoxic response across the timecourse.

```{r reorder_all_timepoints_by_treatment}
# Order all samples by treatment, then condition, then timepoint
cols_all_by_treatment <- all_cols_ordered %>%
  arrange(treatment, condition, timepoint, rep)

# Reorder the full z-score matrix
te_matrix_zscore_by_treatment <- te_matrix_zscore[, cols_all_by_treatment$col_name]

# Create column annotations
col_anno_by_treatment <- data.frame(
  Treatment = cols_all_by_treatment$treatment,
  Condition = cols_all_by_treatment$condition,
  Timepoint = cols_all_by_treatment$timepoint,
  row.names = cols_all_by_treatment$col_name
)

col_anno_by_treatment$Treatment <- factor(col_anno_by_treatment$Treatment, levels = c("sictrl", "si3d", "si3e"))
col_anno_by_treatment$Condition <- factor(col_anno_by_treatment$Condition, levels = c("normoxia", "hypoxia"))
col_anno_by_treatment$Timepoint <- factor(col_anno_by_treatment$Timepoint, levels = c("1hr", "4hr", "24hr"))

cat("Reordered all samples by treatment\n")
cat("Total genes:", nrow(te_matrix_zscore_by_treatment), "\n")
cat("Total samples:", ncol(te_matrix_zscore_by_treatment), "\n")
cat("Column order: sictrl (norm/hyp: 1hr, 4hr, 24hr) | si3d (norm/hyp: 1hr, 4hr, 24hr) | si3e (norm/hyp: 1hr, 4hr, 24hr)\n")
```

```{r cluster_by_sictrl_timecourse}
# Cluster genes based on sictrl samples across all timepoints and conditions
sictrl_all_cols <- cols_all_by_treatment %>%
  filter(treatment == "sictrl") %>%
  pull(col_name)

# Extract sictrl data for clustering
te_matrix_sictrl_all <- te_matrix_zscore_by_treatment[, sictrl_all_cols]

# Perform hierarchical clustering on sictrl timecourse data
row_clustering_timecourse <- hclust(dist(te_matrix_sictrl_all))

cat("Clustered genes based on sictrl response across all timepoints and conditions\n")
```

# Elbow Plot for K-means Clustering

Determine the optimal number of clusters for k-means clustering by examining the within-cluster sum of squares (WSS) for different k values.

```{r elbow_plot}
# Test k-means with different numbers of clusters
k_values <- 2:15
wss <- numeric(length(k_values))

# Calculate WSS for each k
set.seed(123)  # For reproducibility
for (i in seq_along(k_values)) {
  k <- k_values[i]
  kmeans_result <- kmeans(te_matrix_sictrl_all, centers = k, nstart = 25)
  wss[i] <- kmeans_result$tot.withinss
  cat("k =", k, "WSS =", round(wss[i], 2), "\n")
}

# Create elbow plot
elbow_data <- data.frame(k = k_values, wss = wss)

ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Elbow Plot for K-means Clustering",
    subtitle = "Based on siCtrl samples across all timepoints and conditions",
    x = "Number of Clusters (k)",
    y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 12)
  ) +
  scale_x_continuous(breaks = k_values)
```

```{r heatmap_all_genes_all_timepoints, fig.width=14, fig.height=10}
# Create heatmap with all genes and timepoints, using sictrl-based clustering
heatmap_all_timecourse <- pheatmap(
  te_matrix_zscore_by_treatment,
  cluster_rows = row_clustering_timecourse,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_treatment,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes, All Timepoints (Clustered by siCtrl Timecourse)",
  fontsize = 10,
  border_color = NA
)
print(heatmap_all_timecourse)
```

```{r save_heatmap_all_genes_all_timepoints}
# Save as PDF
pdf(here("plots", "te_heatmap_all_genes_all_timepoints_by_treatment.pdf"), width = 14, height = 10)
pheatmap(
  te_matrix_zscore_by_treatment,
  cluster_rows = row_clustering_timecourse,
  cluster_cols = FALSE,
  annotation_col = col_anno_by_treatment,
  annotation_colors = anno_colors,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes, All Timepoints (Clustered by siCtrl Timecourse)",
  fontsize = 10,
  border_color = NA
)
dev.off()

cat("All genes all timepoints heatmap saved to plots/te_heatmap_all_genes_all_timepoints_by_treatment.pdf\n")
```

# All Genes Heatmap - K-means Clustering with 8 Clusters

Create the same heatmap as above but using k-means clustering with k=8 clusters instead of hierarchical clustering.

```{r kmeans_clustering_k8}
# Perform k-means clustering with k=8 on sictrl data
set.seed(123)  # For reproducibility
kmeans_k8 <- kmeans(te_matrix_sictrl_all, centers = 8, nstart = 25)

cat("K-means clustering with k=8 completed\n")
cat("Cluster sizes:\n")
cluster_sizes <- table(kmeans_k8$cluster)
print(cluster_sizes)

# Order genes by cluster
cluster_order <- order(kmeans_k8$cluster)

# Reorder columns by condition (normoxia | hypoxia)
te_matrix_zscore_by_condition_kmeans <- te_matrix_zscore_by_treatment[cluster_order, all_cols_by_condition$col_name]

# Create row annotation with cluster assignments
row_anno <- data.frame(
  Cluster = paste0("C", kmeans_k8$cluster[cluster_order],
                   " (n=", cluster_sizes[kmeans_k8$cluster[cluster_order]], ")"),
  row.names = rownames(te_matrix_zscore_by_condition_kmeans)
)

# Make cluster a factor to maintain order
row_anno$Cluster <- factor(row_anno$Cluster,
                           levels = unique(row_anno$Cluster))

# Define colors for clusters
cluster_colors <- setNames(
  rainbow(8, s = 0.6, v = 0.8),
  unique(row_anno$Cluster)
)

# Calculate positions for gaps between clusters
cluster_boundaries <- cumsum(as.numeric(cluster_sizes[order(as.numeric(names(cluster_sizes)))]))
gaps_positions <- cluster_boundaries[-length(cluster_boundaries)]

cat("\nCluster boundaries for gaps:\n")
print(gaps_positions)
```

```{r heatmap_kmeans_k8, fig.width=14, fig.height=10}
# Create heatmap with k-means clustering (k=8)

dev.off()
heatmap_kmeans_k8 <- pheatmap(
  te_matrix_zscore_by_condition_kmeans,
  cluster_rows = FALSE,  # Already ordered by cluster
  cluster_cols = FALSE,
  annotation_row = row_anno,
  annotation_col = col_anno_by_condition,
  annotation_colors = list(
    Cluster = cluster_colors,
    Condition = anno_colors$Condition,
    Treatment = anno_colors$Treatment,
    Timepoint = anno_colors$Timepoint
  ),
  gaps_row = gaps_positions,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes, All Timepoints (K-means k=8, Normoxia | Hypoxia)",
  fontsize = 10,
  fontsize_row = 8,
  border_color = NA
)
print(heatmap_kmeans_k8)
```

```{r save_heatmap_kmeans_k8}
# Save as PDF
pdf(here("plots", "te_heatmap_all_genes_kmeans_k8.pdf"), width = 14, height = 10)
pheatmap(
  te_matrix_zscore_by_condition_kmeans,
  cluster_rows = FALSE,  # Already ordered by cluster
  cluster_cols = FALSE,
  annotation_row = row_anno,
  annotation_col = col_anno_by_condition,
  annotation_colors = list(
    Cluster = cluster_colors,
    Condition = anno_colors$Condition,
    Treatment = anno_colors$Treatment,
    Timepoint = anno_colors$Timepoint
  ),
  gaps_row = gaps_positions,
  show_rownames = FALSE,
  show_colnames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  main = "TE Z-scores - All Genes, All Timepoints (K-means k=8, Normoxia | Hypoxia)",
  fontsize = 10,
  fontsize_row = 8,
  border_color = NA
)
dev.off()

cat("K-means (k=8) heatmap saved to plots/te_heatmap_all_genes_kmeans_k8.pdf\n")
```

# ORA: Pathway Enrichment for K-means Clusters

Perform Over-Representation Analysis (ORA) using Hallmark gene sets for each of the 8 k-means clusters.

## Extract Genes from Each Cluster

```{r extract_cluster_genes}
# Get genes for each cluster
cluster_genes_list <- list()

for (i in 1:8) {
  # Get genes in this cluster
  cluster_genes <- names(kmeans_k8$cluster[kmeans_k8$cluster == i])
  cluster_genes_list[[paste0("Cluster_", i)]] <- cluster_genes

  cat("Cluster", i, ":", length(cluster_genes), "genes\n")

  # Save gene list to file
  write.csv(
    data.frame(gene_id = cluster_genes),
    here("output", paste0("cluster_", i, "_genes.csv")),
    row.names = FALSE
  )
}

cat("\nGene lists saved to output/cluster_*_genes.csv\n")
```

## ORA for Each Cluster

```{r ora_clusters}
# Initialize list to store ORA results
ora_cluster_results <- list()

# Perform ORA for each cluster
for (i in 1:8) {
  cat("\n=== Cluster", i, "===\n")

  # Run ORA
  ora_result <- enricher(
    gene = cluster_genes_list[[paste0("Cluster_", i)]],
    TERM2GENE = hallmark_term2gene,
    pAdjustMethod = "fdr",
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 15,
    maxGSSize = 500
  )

  # Extract and format results
  if (nrow(ora_result@result) > 0) {
    ora_df <- ora_result@result %>%
      dplyr::select(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
      mutate(
        GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"),
                                   function(x) as.numeric(x[1])/as.numeric(x[2])),
        Cluster = i
      )

    # Round numeric columns
    ora_df <- ora_df %>%
      mutate(
        pvalue = round(pvalue, 6),
        p.adjust = round(p.adjust, 6),
        GeneRatio_numeric = round(GeneRatio_numeric, 4)
      )

    ora_cluster_results[[paste0("Cluster_", i)]] <- ora_df

    # Print significant pathways
    sig_pathways <- ora_df %>% filter(p.adjust < 0.05)
    cat("Significant pathways (p.adjust < 0.05):", nrow(sig_pathways), "\n")

    if (nrow(sig_pathways) > 0) {
      cat("\nTop pathways:\n")
      print(sig_pathways %>%
              arrange(p.adjust) %>%
              dplyr::select(ID, GeneRatio, p.adjust, Count) %>%
              head(5))
    }
  } else {
    cat("No enriched pathways found\n")
    ora_cluster_results[[paste0("Cluster_", i)]] <- data.frame()
  }
}
```

## Visualize ORA Results - Individual Cluster Plots

### Cluster 1

```{r ora_cluster_1_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_1"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 1 (", length(cluster_genes_list[["Cluster_1"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_1.pdf"), p, width = 10, height = 8)
    cat("Cluster 1 plot saved\n")
  } else {
    cat("Cluster 1: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 1: No ORA results\n")
}
```

### Cluster 2

```{r ora_cluster_2_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_2"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 2 (", length(cluster_genes_list[["Cluster_2"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_2.pdf"), p, width = 10, height = 8)
    cat("Cluster 2 plot saved\n")
  } else {
    cat("Cluster 2: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 2: No ORA results\n")
}
```

### Cluster 3

```{r ora_cluster_3_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_3"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 3 (", length(cluster_genes_list[["Cluster_3"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_3.pdf"), p, width = 10, height = 8)
    cat("Cluster 3 plot saved\n")
  } else {
    cat("Cluster 3: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 3: No ORA results\n")
}
```

### Cluster 4

```{r ora_cluster_4_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_4"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 4 (", length(cluster_genes_list[["Cluster_4"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_4.pdf"), p, width = 10, height = 8)
    cat("Cluster 4 plot saved\n")
  } else {
    cat("Cluster 4: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 4: No ORA results\n")
}
```

### Cluster 5

```{r ora_cluster_5_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_5"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 5 (", length(cluster_genes_list[["Cluster_5"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_5.pdf"), p, width = 10, height = 8)
    cat("Cluster 5 plot saved\n")
  } else {
    cat("Cluster 5: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 5: No ORA results\n")
}
```

### Cluster 6

```{r ora_cluster_6_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_6"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 6 (", length(cluster_genes_list[["Cluster_6"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_6.pdf"), p, width = 10, height = 8)
    cat("Cluster 6 plot saved\n")
  } else {
    cat("Cluster 6: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 6: No ORA results\n")
}
```

### Cluster 7

```{r ora_cluster_7_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_7"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 7 (", length(cluster_genes_list[["Cluster_7"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_7.pdf"), p, width = 10, height = 8)
    cat("Cluster 7 plot saved\n")
  } else {
    cat("Cluster 7: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 7: No ORA results\n")
}
```

### Cluster 8

```{r ora_cluster_8_plot, fig.width=10, fig.height=8}
ora_df <- ora_cluster_results[["Cluster_8"]]

if (nrow(ora_df) > 0) {
  sig_pathways <- ora_df %>% filter(p.adjust < 0.05)

  if (nrow(sig_pathways) > 0) {
    plot_data <- sig_pathways %>%
      arrange(p.adjust) %>%
      head(20)

    p <- ggplot(plot_data,
                aes(y = reorder(ID, GeneRatio_numeric),
                    x = GeneRatio_numeric,
                    fill = p.adjust,
                    size = Count)) +
      geom_point(alpha = 0.7, shape = 21, color = "black") +
      scale_size(range = c(3, 10)) +
      scale_fill_viridis_c(direction = -1) +
      labs(y = "Hallmark Pathways",
           x = "Gene Ratio",
           title = paste0("ORA: Cluster 8 (", length(cluster_genes_list[["Cluster_8"]]), " genes)"),
           size = "Gene Count",
           fill = "Adjusted\nP-value") +
      theme_minimal() +
      theme(
        axis.text.y = element_text(size = 9),
        plot.title = element_text(face = "bold")
      )

    print(p)
    ggsave(here("plots", "ora_cluster_8.pdf"), p, width = 10, height = 8)
    cat("Cluster 8 plot saved\n")
  } else {
    cat("Cluster 8: No significant pathways to plot\n")
  }
} else {
  cat("Cluster 8: No ORA results\n")
}
```

## Combined ORA Heatmap Across All Clusters

```{r ora_combined_heatmap, fig.width=12, fig.height=10}
# Combine all ORA results
ora_all_clusters <- bind_rows(ora_cluster_results)

if (nrow(ora_all_clusters) > 0) {
  # Filter for pathways significant in at least one cluster
  sig_pathways_any <- ora_all_clusters %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>%
    unique()

  if (length(sig_pathways_any) > 0) {
    # Create wide format for heatmap (pathways x clusters)
    ora_wide <- ora_all_clusters %>%
      filter(ID %in% sig_pathways_any) %>%
      dplyr::select(ID, Cluster, GeneRatio_numeric) %>%
      pivot_wider(names_from = Cluster,
                  values_from = GeneRatio_numeric,
                  names_prefix = "Cluster_") %>%
      column_to_rownames("ID")

    # Replace NA with 0
    ora_wide[is.na(ora_wide)] <- 0

    # Create significance matrix for annotation
    sig_matrix <- ora_all_clusters %>%
      filter(ID %in% sig_pathways_any) %>%
      dplyr::select(ID, Cluster, p.adjust) %>%
      pivot_wider(names_from = Cluster,
                  values_from = p.adjust,
                  names_prefix = "Cluster_") %>%
      column_to_rownames("ID")

    sig_matrix[is.na(sig_matrix)] <- 1
    sig_matrix <- ifelse(sig_matrix < 0.05, "*", "")
    sig_matrix <- sig_matrix[rownames(ora_wide), colnames(ora_wide)]

    # Create heatmap
    p_ora_heatmap <- pheatmap(
      ora_wide,
      cluster_rows = TRUE,
      cluster_cols = FALSE,
      scale = "none",
      color = colorRampPalette(c("white", "orange", "red"))(100),
      breaks = seq(0, max(ora_wide), length.out = 101),
      display_numbers = sig_matrix,
      number_color = "black",
      fontsize_number = 12,
      main = "ORA Gene Ratio Across K-means Clusters - Hallmark (* = p.adjust < 0.05)",
      fontsize_row = 9,
      fontsize_col = 10,
      cellwidth = 30,
      cellheight = 12,
      border_color = "grey60"
    )

    print(p_ora_heatmap)

    # Save heatmap
    pdf(here("plots", "ora_kmeans_clusters_heatmap.pdf"), width = 12, height = 10)
    print(p_ora_heatmap)
    dev.off()

    cat("\nORA heatmap shows", nrow(ora_wide), "pathways across 8 clusters\n")
    cat("Heatmap saved to plots/ora_kmeans_clusters_heatmap.pdf\n")
  } else {
    cat("No significant pathways found across any cluster\n")
  }
} else {
  cat("No ORA results to visualize\n")
}
```

## Export ORA Results

```{r export_ora_results}
# Prepare export list (remove geneID column for readability in Excel)
ora_export_list <- list()

for (i in 1:8) {
  if (nrow(ora_cluster_results[[paste0("Cluster_", i)]]) > 0) {
    ora_export_list[[paste0("Cluster_", i)]] <-
      ora_cluster_results[[paste0("Cluster_", i)]] %>%
      dplyr::select(-GeneRatio_numeric, -Cluster, -geneID) %>%
      arrange(p.adjust)
  }
}

# Export to Excel
if (length(ora_export_list) > 0) {
  write_xlsx(ora_export_list, here("output", "ora_kmeans_clusters_hallmark.xlsx"))
  cat("\nORA results exported to: output/ora_kmeans_clusters_hallmark.xlsx\n")
}

# Print summary statistics
cat("\n=== ORA Summary ===\n")
for (i in 1:8) {
  n_genes <- length(cluster_genes_list[[paste0("Cluster_", i)]])

  if (nrow(ora_cluster_results[[paste0("Cluster_", i)]]) > 0) {
    n_sig <- ora_cluster_results[[paste0("Cluster_", i)]] %>%
      filter(p.adjust < 0.05) %>%
      nrow()

    top_pathway <- ora_cluster_results[[paste0("Cluster_", i)]] %>%
      filter(p.adjust < 0.05) %>%
      arrange(p.adjust) %>%
      dplyr::slice(1) %>%
      pull(ID)

    cat("Cluster", i, ":", n_genes, "genes,", n_sig, "significant pathways")
    if (length(top_pathway) > 0) {
      cat(", Top:", top_pathway[1])
    }
    cat("\n")
  } else {
    cat("Cluster", i, ":", n_genes, "genes, 0 significant pathways\n")
  }
}
```

# Session Info

```{r session_info}
sessionInfo()
```
