---
title: "Hypoxic RBP Knockdown Analysis"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(cowplot)
library(readxl)
library(janitor)
library(rstatix)
```

## Overview

This analysis examines how RNA-binding protein (RBP) knockdowns affect protein production in hypoxia, based on TMT-pSILAC proteomics data from Ho et al. 2020 ("A network of RNA-binding proteins controls translation efficiency to activate anaerobic metabolism", Nature Communications).

We compare RBP-dependent genes' translation efficiency response across hypoxia timepoints using ECDF plots.

## Load and Process RBP Knockdown Data

```{r load_rbp_data}
# Load TMT-pSILAC proteomics data from Ho et al. 2020
# Sheet "Figure 2b" contains 1,726 proteins with abundance measurements
rbp_data <- read_xlsx("/Users/katematlin/Downloads/41467_2020_16504_MOESM3_ESM.xlsx",
                      sheet = "Figure 2b") %>%
  clean_names()

cat("RBP knockdown data columns after clean_names():\n")
print(colnames(rbp_data))

cat("\nRBP knockdown data dimensions:", nrow(rbp_data), "rows x", ncol(rbp_data), "columns\n")
head(rbp_data)
```

```{r calculate_log2fc}
# After clean_names(), the columns are:
# - gene_symbol
# - normoxia_abundance_126
# - hypoxia_abundance_average_127c
# - hypoxia_pcbp1_si_rna_abundance_128c
# - hypoxia_pcbp2_si_rna_abundance_129c
# - hypoxia_ptbp1_si_rna_abundance_130n
# - hypoxia_hn_rnp_a2_b1_si_rna_abundance_130c
# - hypoxia_hu_r_si_rna_abundance_131

# Calculate log2 fold changes: log2(KD abundance / Hypoxia control abundance)
# Filter for proteins with ≥2 unique peptides for confident quantification
min_peptides <- 2

rbp_log2fc <- rbp_data %>%
  mutate(
    symbol = gene_symbol,
    n_unique_peptides = as.numeric(number_unique_peptides),
    normoxia = as.numeric(normoxia_abundance_126),
    hypoxia_ctrl = as.numeric(hypoxia_abundance_average_127c),
    pcbp1_abundance = as.numeric(hypoxia_pcbp1_si_rna_abundance_128c),
    pcbp2_abundance = as.numeric(hypoxia_pcbp2_si_rna_abundance_129c),
    ptbp1_abundance = as.numeric(hypoxia_ptbp1_si_rna_abundance_130n),
    hnrnpa2b1_abundance = as.numeric(hypoxia_hn_rnp_a2_b1_si_rna_abundance_130c),
    hur_abundance = as.numeric(hypoxia_hu_r_si_rna_abundance_131)
  ) %>%
  # Filter for minimum peptide count
  filter(n_unique_peptides >= min_peptides) %>%
  # Calculate log2FC for each RBP knockdown vs hypoxia control
  mutate(
    log2fc_pcbp1 = log2(pcbp1_abundance / hypoxia_ctrl),
    log2fc_pcbp2 = log2(pcbp2_abundance / hypoxia_ctrl),
    log2fc_ptbp1 = log2(ptbp1_abundance / hypoxia_ctrl),
    log2fc_hnrnpa2b1 = log2(hnrnpa2b1_abundance / hypoxia_ctrl),
    log2fc_hur = log2(hur_abundance / hypoxia_ctrl)
  ) %>%
  dplyr::select(symbol, n_unique_peptides, normoxia, hypoxia_ctrl,
                pcbp1_abundance, pcbp2_abundance, ptbp1_abundance,
                hnrnpa2b1_abundance, hur_abundance,
                log2fc_pcbp1, log2fc_pcbp2, log2fc_ptbp1,
                log2fc_hnrnpa2b1, log2fc_hur) %>%
  filter(!is.na(symbol))

cat("\nApplied filter: ≥", min_peptides, "unique peptides\n")

cat("\nProcessed RBP log2FC data:\n")
cat("Number of genes:", nrow(rbp_log2fc), "\n")
head(rbp_log2fc)

# Summary statistics for each RBP
cat("\nLog2FC summary for each RBP knockdown:\n")
cat("\nPCBP1:\n")
print(summary(rbp_log2fc$log2fc_pcbp1))
cat("\nPCBP2:\n")
print(summary(rbp_log2fc$log2fc_pcbp2))
cat("\nPTBP1:\n")
print(summary(rbp_log2fc$log2fc_ptbp1))
cat("\nhnRNP A2/B1:\n")
print(summary(rbp_log2fc$log2fc_hnrnpa2b1))
cat("\nHuR:\n")
print(summary(rbp_log2fc$log2fc_hur))
```

## Define RBP-Dependent Gene Sets

**Filtering criteria:**
- Minimum 2 unique peptides (for confident protein quantification)
- |log2FC| > 0.5 threshold for RBP dependency

**Categories:**
- **RBP-Promoted**: log2FC < -0.5 (protein DOWN when RBP knocked down, meaning RBP promotes its production)
- **RBP-Inhibited**: log2FC > 0.5 (protein UP when RBP knocked down, meaning RBP inhibits its production)

*Note: The TMT-pSILAC proteomics data provides abundance ratios without per-protein p-values, which is common for this type of experiment. We use a peptide count filter plus fold change threshold as quality criteria.*

```{r define_genesets}
# Define threshold
lfc_threshold <- 0.5

# Function to classify genes by RBP dependency
classify_rbp_dependency <- function(log2fc, threshold = 0.5) {
  case_when(
    log2fc < -threshold ~ "RBP-Promoted",
    log2fc > threshold ~ "RBP-Inhibited",
    TRUE ~ "Other"
  )
}

# Classify genes for each RBP
rbp_classified <- rbp_log2fc %>%
  mutate(
    pcbp1_status = classify_rbp_dependency(log2fc_pcbp1, lfc_threshold),
    pcbp2_status = classify_rbp_dependency(log2fc_pcbp2, lfc_threshold),
    ptbp1_status = classify_rbp_dependency(log2fc_ptbp1, lfc_threshold),
    hnrnpa2b1_status = classify_rbp_dependency(log2fc_hnrnpa2b1, lfc_threshold),
    hur_status = classify_rbp_dependency(log2fc_hur, lfc_threshold)
  )

# Count genes in each category for each RBP
cat("\nGene counts by RBP dependency (|log2FC| > 0.5):\n\n")

rbp_counts <- data.frame(
  RBP = c("PCBP1", "PCBP2", "PTBP1", "hnRNP A2/B1", "HuR"),
  Promoted = c(
    sum(rbp_classified$pcbp1_status == "RBP-Promoted", na.rm = TRUE),
    sum(rbp_classified$pcbp2_status == "RBP-Promoted", na.rm = TRUE),
    sum(rbp_classified$ptbp1_status == "RBP-Promoted", na.rm = TRUE),
    sum(rbp_classified$hnrnpa2b1_status == "RBP-Promoted", na.rm = TRUE),
    sum(rbp_classified$hur_status == "RBP-Promoted", na.rm = TRUE)
  ),
  Inhibited = c(
    sum(rbp_classified$pcbp1_status == "RBP-Inhibited", na.rm = TRUE),
    sum(rbp_classified$pcbp2_status == "RBP-Inhibited", na.rm = TRUE),
    sum(rbp_classified$ptbp1_status == "RBP-Inhibited", na.rm = TRUE),
    sum(rbp_classified$hnrnpa2b1_status == "RBP-Inhibited", na.rm = TRUE),
    sum(rbp_classified$hur_status == "RBP-Inhibited", na.rm = TRUE)
  ),
  Other = c(
    sum(rbp_classified$pcbp1_status == "Other", na.rm = TRUE),
    sum(rbp_classified$pcbp2_status == "Other", na.rm = TRUE),
    sum(rbp_classified$ptbp1_status == "Other", na.rm = TRUE),
    sum(rbp_classified$hnrnpa2b1_status == "Other", na.rm = TRUE),
    sum(rbp_classified$hur_status == "Other", na.rm = TRUE)
  )
)

print(rbp_counts)
```

## Load Hypoxia TE Data

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

# Add gene symbols to TE data and prepare for analysis
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Loaded TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes\n")
cat("4hr:", nrow(te_4hr_gsea), "genes\n")
cat("24hr:", nrow(te_24hr_gsea), "genes\n")
```

## ECDF Analysis Functions

```{r ecdf_functions}
# Function to create ECDF plot for a single RBP at a single timepoint
create_ecdf_plot <- function(te_data, rbp_status_col, rbp_name, timepoint_label) {

  # Merge TE data with RBP classification
  merged_data <- te_data %>%
    inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp_status_col)),
               by = "symbol")

  # Rename status column for consistent plotting
  merged_data <- merged_data %>%
    rename(rbp_status = !!sym(rbp_status_col))

  # Count genes in each category
  n_promoted <- sum(merged_data$rbp_status == "RBP-Promoted", na.rm = TRUE)
  n_inhibited <- sum(merged_data$rbp_status == "RBP-Inhibited", na.rm = TRUE)
  n_other <- sum(merged_data$rbp_status == "Other", na.rm = TRUE)

  # Create plot
  p <- ggplot(merged_data, aes(x = log2FoldChange, color = rbp_status)) +
    stat_ecdf(linewidth = 1.2) +
    scale_color_manual(
      values = c("RBP-Promoted" = "#D55E00", "RBP-Inhibited" = "#0072B2", "Other" = "#808080"),
      name = paste0(rbp_name, " Status"),
      labels = c(
        "RBP-Promoted" = paste0("RBP-Promoted (n=", n_promoted, ")"),
        "RBP-Inhibited" = paste0("RBP-Inhibited (n=", n_inhibited, ")"),
        "Other" = paste0("Other (n=", n_other, ")")
      )
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0("Hypoxic TE Response by ", rbp_name, " Dependency"),
      subtitle = paste0("Translation Efficiency Changes at ", timepoint_label),
      x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
      y = "Cumulative Proportion"
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5)
    ) +
    xlim(-1, 1)

  return(list(plot = p, data = merged_data))
}

# Function to create faceted plot for all timepoints
create_faceted_ecdf_plot <- function(rbp_status_col, rbp_name, rbp_short_name) {

  # Merge TE data with RBP classification for each timepoint
  merged_1hr <- te_1hr_gsea %>%
    inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp_status_col)),
               by = "symbol") %>%
    rename(rbp_status = !!sym(rbp_status_col)) %>%
    mutate(timepoint = "1 Hour")

  merged_4hr <- te_4hr_gsea %>%
    inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp_status_col)),
               by = "symbol") %>%
    rename(rbp_status = !!sym(rbp_status_col)) %>%
    mutate(timepoint = "4 Hours")

  merged_24hr <- te_24hr_gsea %>%
    inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp_status_col)),
               by = "symbol") %>%
    rename(rbp_status = !!sym(rbp_status_col)) %>%
    mutate(timepoint = "24 Hours")

  # Combine all timepoints
  te_combined <- bind_rows(merged_1hr, merged_4hr, merged_24hr) %>%
    mutate(timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")))

  # Calculate counts for annotations
  count_labels <- te_combined %>%
    group_by(timepoint, rbp_status) %>%
    summarise(n = n(), .groups = "drop") %>%
    mutate(label = paste0("n=", n)) %>%
    mutate(
      x = 0.7,
      y = case_when(
        rbp_status == "RBP-Promoted" ~ 0.25,
        rbp_status == "RBP-Inhibited" ~ 0.15,
        TRUE ~ 0.05
      )
    )

  # Create faceted ECDF plot
  p <- ggplot(te_combined, aes(x = log2FoldChange, color = rbp_status)) +
    stat_ecdf(linewidth = 1.2) +
    scale_color_manual(
      values = c("RBP-Promoted" = "#D55E00", "RBP-Inhibited" = "#0072B2", "Other" = "#808080"),
      name = paste0(rbp_name, " Status")
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    geom_text(data = count_labels, aes(x = x, y = y, label = label, color = rbp_status),
              size = 2.5, show.legend = FALSE) +
    facet_wrap(~timepoint, nrow = 1) +
    labs(
      title = paste0("Hypoxic TE Response by ", rbp_name, " Dependency"),
      x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
      y = "Cumulative Proportion"
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90"),
      strip.text = element_text(face = "bold")
    ) +
    xlim(-3, 1)

  return(list(plot = p, data = te_combined))
}
```

## ECDF Plots for Each RBP

### PCBP1

```{r ecdf_pcbp1, fig.width=10, fig.height=4}
pcbp1_result <- create_faceted_ecdf_plot("pcbp1_status", "PCBP1", "PCBP1")
print(pcbp1_result$plot)

ggsave(
  here("plots", "ecdf_rbp_PCBP1_all_timepoints.pdf"),
  plot = pcbp1_result$plot,
  width = 10,
  height = 4
)
```

### PCBP2

```{r ecdf_pcbp2, fig.width=10, fig.height=4}
pcbp2_result <- create_faceted_ecdf_plot("pcbp2_status", "PCBP2", "PCBP2")
print(pcbp2_result$plot)

ggsave(
  here("plots", "ecdf_rbp_PCBP2_all_timepoints.pdf"),
  plot = pcbp2_result$plot,
  width = 10,
  height = 4
)
```

### PTBP1

```{r ecdf_ptbp1, fig.width=10, fig.height=4}
ptbp1_result <- create_faceted_ecdf_plot("ptbp1_status", "PTBP1", "PTBP1")
print(ptbp1_result$plot)

ggsave(
  here("plots", "ecdf_rbp_PTBP1_all_timepoints.pdf"),
  plot = ptbp1_result$plot,
  width = 10,
  height = 4
)
```

### hnRNP A2/B1

```{r ecdf_hnrnpa2b1, fig.width=10, fig.height=4}
hnrnpa2b1_result <- create_faceted_ecdf_plot("hnrnpa2b1_status", "hnRNP A2/B1", "hnRNPA2B1")
print(hnrnpa2b1_result$plot)

ggsave(
  here("plots", "ecdf_rbp_hnRNPA2B1_all_timepoints.pdf"),
  plot = hnrnpa2b1_result$plot,
  width = 10,
  height = 4
)
```

### HuR

```{r ecdf_hur, fig.width=10, fig.height=4}
hur_result <- create_faceted_ecdf_plot("hur_status", "HuR", "HuR")
print(hur_result$plot)

ggsave(
  here("plots", "ecdf_rbp_HuR_all_timepoints.pdf"),
  plot = hur_result$plot,
  width = 10,
  height = 4
)
```

## Statistical Tests

```{r statistical_tests}
# Function to perform statistical tests for one RBP at one timepoint
perform_tests <- function(te_data, rbp_status_col, rbp_name, timepoint) {

  merged_data <- te_data %>%
    inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp_status_col)),
               by = "symbol") %>%
    rename(rbp_status = !!sym(rbp_status_col))

  # Get vectors for each group
  promoted <- merged_data %>%
    filter(rbp_status == "RBP-Promoted") %>%
    pull(log2FoldChange)
  inhibited <- merged_data %>%
    filter(rbp_status == "RBP-Inhibited") %>%
    pull(log2FoldChange)
  other <- merged_data %>%
    filter(rbp_status == "Other") %>%
    pull(log2FoldChange)

  results <- data.frame(
    rbp = rbp_name,
    timepoint = timepoint,
    n_promoted = length(promoted),
    n_inhibited = length(inhibited),
    n_other = length(other),
    median_promoted = ifelse(length(promoted) > 0, median(promoted), NA),
    median_inhibited = ifelse(length(inhibited) > 0, median(inhibited), NA),
    median_other = ifelse(length(other) > 0, median(other), NA)
  )

  # KS test: Promoted vs Other
  if (length(promoted) >= 3 && length(other) >= 3) {
    ks_promoted <- ks.test(promoted, other)
    results$ks_promoted_vs_other_stat <- ks_promoted$statistic
    results$ks_promoted_vs_other_pval <- ks_promoted$p.value
  } else {
    results$ks_promoted_vs_other_stat <- NA
    results$ks_promoted_vs_other_pval <- NA
  }

  # KS test: Inhibited vs Other
  if (length(inhibited) >= 3 && length(other) >= 3) {
    ks_inhibited <- ks.test(inhibited, other)
    results$ks_inhibited_vs_other_stat <- ks_inhibited$statistic
    results$ks_inhibited_vs_other_pval <- ks_inhibited$p.value
  } else {
    results$ks_inhibited_vs_other_stat <- NA
    results$ks_inhibited_vs_other_pval <- NA
  }

  # Wilcoxon test: Promoted vs Other
  if (length(promoted) >= 3 && length(other) >= 3) {
    wilcox_promoted <- wilcox.test(promoted, other)
    results$wilcox_promoted_vs_other_stat <- wilcox_promoted$statistic
    results$wilcox_promoted_vs_other_pval <- wilcox_promoted$p.value
  } else {
    results$wilcox_promoted_vs_other_stat <- NA
    results$wilcox_promoted_vs_other_pval <- NA
  }

  # Wilcoxon test: Inhibited vs Other
  if (length(inhibited) >= 3 && length(other) >= 3) {
    wilcox_inhibited <- wilcox.test(inhibited, other)
    results$wilcox_inhibited_vs_other_stat <- wilcox_inhibited$statistic
    results$wilcox_inhibited_vs_other_pval <- wilcox_inhibited$p.value
  } else {
    results$wilcox_inhibited_vs_other_stat <- NA
    results$wilcox_inhibited_vs_other_pval <- NA
  }

  return(results)
}

# Define RBP info for iteration
rbp_info <- list(
  list(col = "pcbp1_status", name = "PCBP1"),
  list(col = "pcbp2_status", name = "PCBP2"),
  list(col = "ptbp1_status", name = "PTBP1"),
  list(col = "hnrnpa2b1_status", name = "hnRNP_A2B1"),
  list(col = "hur_status", name = "HuR")
)

# Perform tests for all RBPs and timepoints
all_stats <- data.frame()

for (rbp in rbp_info) {
  for (tp in list(list(data = te_1hr_gsea, label = "1hr"),
                  list(data = te_4hr_gsea, label = "4hr"),
                  list(data = te_24hr_gsea, label = "24hr"))) {
    stats <- perform_tests(tp$data, rbp$col, rbp$name, tp$label)
    all_stats <- bind_rows(all_stats, stats)
  }
}

cat("\nStatistical Test Results:\n")
print(all_stats %>%
        dplyr::select(rbp, timepoint, n_promoted, n_inhibited,
                      median_promoted, median_inhibited, median_other,
                      ks_promoted_vs_other_pval, ks_inhibited_vs_other_pval))
```

## Summary Comparison Plot

```{r summary_heatmap, fig.width=10, fig.height=6}
# Create summary data for heatmap
# Show -log10(p-value) for KS test, with sign indicating direction

summary_data <- all_stats %>%
  mutate(
    # Calculate effect direction based on median difference
    promoted_effect = median_promoted - median_other,
    inhibited_effect = median_inhibited - median_other,
    # Create signed -log10 p-values for visualization
    promoted_score = -log10(ks_promoted_vs_other_pval) * sign(promoted_effect),
    inhibited_score = -log10(ks_inhibited_vs_other_pval) * sign(inhibited_effect)
  ) %>%
  dplyr::select(rbp, timepoint, promoted_score, inhibited_score,
                ks_promoted_vs_other_pval, ks_inhibited_vs_other_pval)

# Reshape for heatmap
summary_long <- summary_data %>%
  pivot_longer(
    cols = c(promoted_score, inhibited_score),
    names_to = "comparison",
    values_to = "score"
  ) %>%
  mutate(
    comparison = case_when(
      comparison == "promoted_score" ~ "RBP-Promoted",
      comparison == "inhibited_score" ~ "RBP-Inhibited"
    ),
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    rbp = factor(rbp, levels = c("PCBP1", "PCBP2", "PTBP1", "hnRNP_A2B1", "HuR"))
  )

# Get corresponding p-values
summary_long <- summary_long %>%
  mutate(
    pval = case_when(
      comparison == "RBP-Promoted" ~ summary_data$ks_promoted_vs_other_pval[match(paste(rbp, timepoint), paste(summary_data$rbp, summary_data$timepoint))],
      comparison == "RBP-Inhibited" ~ summary_data$ks_inhibited_vs_other_pval[match(paste(rbp, timepoint), paste(summary_data$rbp, summary_data$timepoint))]
    ),
    significance = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01 ~ "**",
      pval < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

# Create heatmap
p_heatmap <- ggplot(summary_long, aes(x = timepoint, y = rbp, fill = score)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#0072B2", mid = "white", high = "#D55E00",
    midpoint = 0,
    name = "Signed\n-log10(p)",
    limits = c(-3, 3),
    oob = scales::squish
  ) +
  facet_wrap(~comparison, nrow = 1) +
  labs(
    title = "RBP Dependency Effects on Hypoxic TE Response",
    subtitle = "Positive: higher TE in hypoxia | Negative: lower TE in hypoxia",
    x = "Hypoxia Timepoint",
    y = "RBP Knockdown"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  )

print(p_heatmap)

ggsave(
  here("plots", "ecdf_rbp_summary_heatmap.pdf"),
  plot = p_heatmap,
  width = 8,
  height = 5
)
```

```{r bar_plot_summary, fig.width=12, fig.height=5}
# Create bar plot showing median TE shifts
median_summary <- all_stats %>%
  dplyr::select(rbp, timepoint, median_promoted, median_inhibited, median_other) %>%
  pivot_longer(
    cols = c(median_promoted, median_inhibited),
    names_to = "category",
    values_to = "median_te"
  ) %>%
  mutate(
    category = case_when(
      category == "median_promoted" ~ "RBP-Promoted",
      category == "median_inhibited" ~ "RBP-Inhibited"
    ),
    # Calculate difference from "Other"
    delta_median = median_te - median_other,
    timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")),
    rbp = factor(rbp, levels = c("PCBP1", "PCBP2", "PTBP1", "hnRNP_A2B1", "HuR"))
  )

p_bar <- ggplot(median_summary, aes(x = rbp, y = delta_median, fill = category)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(
    values = c("RBP-Promoted" = "#D55E00", "RBP-Inhibited" = "#0072B2"),
    name = "Gene Category"
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  labs(
    title = "Median TE Shift of RBP-Dependent Genes vs Other Genes",
    subtitle = "Delta = Median(RBP-category) - Median(Other)",
    x = "RBP",
    y = "Delta Median TE log2FC"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )

print(p_bar)

ggsave(
  here("plots", "ecdf_rbp_median_shift_barplot.pdf"),
  plot = p_bar,
  width = 12,
  height = 5
)
```

## Combined All-RBPs Faceted Plot

```{r combined_all_rbps, fig.width=15, fig.height=12}
# Create a comprehensive faceted plot with all RBPs and timepoints

# Prepare combined data
all_rbp_combined <- data.frame()

for (rbp in rbp_info) {
  for (tp in list(list(data = te_1hr_gsea, label = "1 Hour"),
                  list(data = te_4hr_gsea, label = "4 Hours"),
                  list(data = te_24hr_gsea, label = "24 Hours"))) {
    merged <- tp$data %>%
      inner_join(rbp_classified %>% dplyr::select(symbol, !!sym(rbp$col)),
                 by = "symbol") %>%
      rename(rbp_status = !!sym(rbp$col)) %>%
      mutate(
        rbp_name = rbp$name,
        timepoint = tp$label
      )
    all_rbp_combined <- bind_rows(all_rbp_combined, merged)
  }
}

all_rbp_combined <- all_rbp_combined %>%
  mutate(
    timepoint = factor(timepoint, levels = c("1 Hour", "4 Hours", "24 Hours")),
    rbp_name = factor(rbp_name, levels = c("PCBP1", "PCBP2", "PTBP1", "hnRNP_A2B1", "HuR"))
  )

# Create comprehensive faceted plot
p_all_rbps <- ggplot(all_rbp_combined, aes(x = log2FoldChange, color = rbp_status)) +
  stat_ecdf(linewidth = 0.8) +
  scale_color_manual(
    values = c("RBP-Promoted" = "#D55E00", "RBP-Inhibited" = "#0072B2", "Other" = "#808080"),
    name = "RBP Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  facet_grid(rbp_name ~ timepoint) +
  labs(
    title = "Hypoxic TE Response by RBP Dependency",
    subtitle = "All RBPs across all timepoints",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold")
  ) +
  xlim(-1, 1)

print(p_all_rbps)

ggsave(
  here("plots", "ecdf_rbp_all_timepoints_faceted.pdf"),
  plot = p_all_rbps,
  width = 12,
  height = 12
)
```

## Export Results

```{r export_results}
# Export full processed log2FC data
write_csv(rbp_log2fc, here("output", "rbp_knockdown_log2fc.csv"))
cat("Exported: output/rbp_knockdown_log2fc.csv\n")

# Export statistics table
write_csv(all_stats, here("output", "rbp_knockdown_statistics.csv"))
cat("Exported: output/rbp_knockdown_statistics.csv\n")

# Export gene sets for each RBP
geneset_output <- rbp_classified %>%
  dplyr::select(symbol,
                pcbp1_status, log2fc_pcbp1,
                pcbp2_status, log2fc_pcbp2,
                ptbp1_status, log2fc_ptbp1,
                hnrnpa2b1_status, log2fc_hnrnpa2b1,
                hur_status, log2fc_hur)

write_csv(geneset_output, here("output", "genesets", "rbp_dependent_genesets.csv"))
cat("Exported: output/genesets/rbp_dependent_genesets.csv\n")

cat("\nAll output files exported successfully.\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
