---
title: "Hypoxic 3d Gene Set Characterization: Leading Edge Analysis"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis characterizes the leading edge genes from eIF3d and eIF3e gene sets when tested against the siCtrl hypoxic TE response. Unlike the previous top-20 analysis, this extracts the **full leading edge** directly from fgsea's `leadingEdge` column.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggplot2)
library(fgsea)
library(writexl)
library(clusterProfiler)
library(msigdbr)
library(pheatmap)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(rstatix)
library(tximport)
library(scales)
library(FeatureReachR)
```

# Load Data

## Gene Annotation

```{r load_gene_annotation}
# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

## TE Data (siCtrl Hypoxic Response)

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

cat("Loaded TE data:\n")
cat("1hr:", nrow(te_1hr), "genes\n")
cat("4hr:", nrow(te_4hr), "genes\n")
cat("24hr:", nrow(te_24hr), "genes\n")
```

## Prepare TE Data with Gene Symbols

```{r prepare_te_data}
# Add gene symbols to TE data
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Prepared TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes with symbols\n")
cat("4hr:", nrow(te_4hr_gsea), "genes with symbols\n")
cat("24hr:", nrow(te_24hr_gsea), "genes with symbols\n")
```

## Load Custom Gene Sets

```{r load_custom_genesets}
# Load custom gene sets from CSV
custom_genesets_raw <- read_csv(here("accessories", "all_combined_sigs_1_19_26.csv"))

# Filter to MDA-MB-231 gene sets
custom_genesets <- custom_genesets_raw %>%
  filter(str_starts(gs_name, "MDA-MB-231_")) %>%
  # Exclude overlap gene sets
  filter(!str_detect(gs_name, "overlap")) %>%
  # Exclude genes where db_gene_symbol is just an ENSG ID (no standard gene symbol)
  filter(!str_starts(db_gene_symbol, "ENSG")) %>%
  # Remove any NA values in gene symbols
  filter(!is.na(db_gene_symbol), db_gene_symbol != "") %>%
  dplyr::select(gs_name, db_gene_symbol) %>%
  distinct()

cat("Loaded custom MDA-MB-231 gene sets:\n")
cat("Number of gene sets:", length(unique(custom_genesets$gs_name)), "\n")
cat("Total gene-pathway associations:", nrow(custom_genesets), "\n")

# Convert to fgsea format (named list of gene vectors)
pathways_fgsea <- custom_genesets %>%
  group_by(gs_name) %>%
  summarise(genes = list(db_gene_symbol), .groups = "drop") %>%
  deframe()

cat("\nConverted to fgsea format:", length(pathways_fgsea), "pathways\n")
```

## Helper Function to Extract Gene Sets

```{r helper_function}
# Function to extract genes from a specific gene set
get_geneset_genes <- function(geneset_name) {
  custom_genesets %>%
    filter(gs_name == geneset_name) %>%
    pull(db_gene_symbol) %>%
    unique()
}
```

# Create Ranked Lists (TE LFC)

```{r create_ranked_lists}
# Create ranked list from TE LFC (hypoxia vs normoxia in siCtrl)
create_ranked_list <- function(te_df) {
  ranked <- te_df$log2FoldChange
  names(ranked) <- te_df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked <- ranked[!duplicated(names(ranked))]
  return(ranked)
}

ranked_list_1hr <- create_ranked_list(te_1hr_gsea)
ranked_list_4hr <- create_ranked_list(te_4hr_gsea)
ranked_list_24hr <- create_ranked_list(te_24hr_gsea)

cat("Created ranked lists (TE LFC):\n")
cat("1hr:", length(ranked_list_1hr), "genes, range:",
    round(min(ranked_list_1hr), 2), "to", round(max(ranked_list_1hr), 2), "\n")
cat("4hr:", length(ranked_list_4hr), "genes, range:",
    round(min(ranked_list_4hr), 2), "to", round(max(ranked_list_4hr), 2), "\n")
cat("24hr:", length(ranked_list_24hr), "genes, range:",
    round(min(ranked_list_24hr), 2), "to", round(max(ranked_list_24hr), 2), "\n")
```

# Run fgsea Analysis

```{r run_fgsea}
# Run fgsea for each timepoint
set.seed(42)

fgsea_1hr <- fgsea(
  pathways = pathways_fgsea,
  stats = ranked_list_1hr,
  minSize = 1,
  maxSize = 5000,
  gseaParam = 0,
  nproc = 1
)

fgsea_4hr <- fgsea(
  pathways = pathways_fgsea,
  stats = ranked_list_4hr,
  minSize = 1,
  maxSize = 5000,
  gseaParam = 0,
  nproc = 1
)

fgsea_24hr <- fgsea(
  pathways = pathways_fgsea,
  stats = ranked_list_24hr,
  minSize = 1,
  maxSize = 5000,
  gseaParam = 0,
  nproc = 1
)

cat("fgsea complete:\n")
cat("1hr: tested", nrow(fgsea_1hr), "pathways\n")
cat("4hr: tested", nrow(fgsea_4hr), "pathways\n")
cat("24hr: tested", nrow(fgsea_24hr), "pathways\n")
```

# Extract Full Leading Edge Genes

```{r extract_leading_edge}
# Function to extract full leading edge genes for a specific gene set
extract_leading_edge <- function(fgsea_result, geneset_name, timepoint) {
  result <- fgsea_result %>%
    filter(pathway == geneset_name)

  if (nrow(result) == 0) {
    warning(paste("Gene set not found:", geneset_name))
    return(NULL)
  }

  # The leadingEdge column is a list column containing gene names
  leading_edge_genes <- result$leadingEdge[[1]]

  data.frame(
    geneset = geneset_name,
    timepoint = timepoint,
    gene = leading_edge_genes,
    NES = result$NES,
    pval = result$pval,
    padj = result$padj,
    setSize = result$size,
    leading_edge_count = length(leading_edge_genes)
  )
}

# Extract leading edge for 3d promotes at each timepoint (LFC > 0.5)
le_3d_1hr <- extract_leading_edge(fgsea_1hr, "MDA-MB-231_hypoxia_3d_promotes_TE_1hr_lfc0.5", "1hr")
le_3d_4hr <- extract_leading_edge(fgsea_4hr, "MDA-MB-231_hypoxia_3d_promotes_TE_4hr_lfc0.5", "4hr")
le_3d_24hr <- extract_leading_edge(fgsea_24hr, "MDA-MB-231_hypoxia_3d_promotes_TE_24hr_lfc0.5", "24hr")

# Extract leading edge for 3e promotes at each timepoint (LFC > 0.5)
le_3e_1hr <- extract_leading_edge(fgsea_1hr, "MDA-MB-231_hypoxia_3e_promotes_TE_1hr_lfc0.5", "1hr")
le_3e_4hr <- extract_leading_edge(fgsea_4hr, "MDA-MB-231_hypoxia_3e_promotes_TE_4hr_lfc0.5", "4hr")
le_3e_24hr <- extract_leading_edge(fgsea_24hr, "MDA-MB-231_hypoxia_3e_promotes_TE_24hr_lfc0.5", "24hr")

# Combine 3d results
le_3d_combined <- bind_rows(le_3d_1hr, le_3d_4hr, le_3d_24hr)

# Combine 3e results
le_3e_combined <- bind_rows(le_3e_1hr, le_3e_4hr, le_3e_24hr)

cat("\n3d Leading Edge Summary:\n")
cat("1hr:", nrow(le_3d_1hr), "genes (NES:", round(le_3d_1hr$NES[1], 2),
    ", padj:", format(le_3d_1hr$padj[1], scientific = TRUE, digits = 2), ")\n")
cat("4hr:", nrow(le_3d_4hr), "genes (NES:", round(le_3d_4hr$NES[1], 2),
    ", padj:", format(le_3d_4hr$padj[1], scientific = TRUE, digits = 2), ")\n")
cat("24hr:", nrow(le_3d_24hr), "genes (NES:", round(le_3d_24hr$NES[1], 2),
    ", padj:", format(le_3d_24hr$padj[1], scientific = TRUE, digits = 2), ")\n")

cat("\n3e Leading Edge Summary:\n")
cat("1hr:", nrow(le_3e_1hr), "genes (NES:", round(le_3e_1hr$NES[1], 2),
    ", padj:", format(le_3e_1hr$padj[1], scientific = TRUE, digits = 2), ")\n")
cat("4hr:", nrow(le_3e_4hr), "genes (NES:", round(le_3e_4hr$NES[1], 2),
    ", padj:", format(le_3e_4hr$padj[1], scientific = TRUE, digits = 2), ")\n")
cat("24hr:", nrow(le_3e_24hr), "genes (NES:", round(le_3e_24hr$NES[1], 2),
    ", padj:", format(le_3e_24hr$padj[1], scientific = TRUE, digits = 2), ")\n")
```

# Density Plots

## 3d-Promoted Genes (Timepoint-Matched)

```{r density_3d_promotes, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets
genes_3d_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_1hr_lfc0.5")
genes_3d_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_4hr_lfc0.5")
genes_3d_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3d_promotes_TE_24hr_lfc0.5")

cat("3d Promotes gene counts:\n")
cat("1hr:", length(genes_3d_1hr), "\n")
cat("4hr:", length(genes_3d_4hr), "\n")
cat("24hr:", length(genes_3d_24hr), "\n")

# Combine with timepoint-matched classification
te_combined_3d <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3d_1hr, "3d Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3d_4hr, "3d Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3d_24hr, "3d Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data for gene counts
label_data_3d <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3d_1hr)),
    paste0("n = ", length(genes_3d_4hr)),
    paste0("n = ", length(genes_3d_24hr))
  )
)

# Create density plot
p_density_3d <- ggplot(te_combined_3d, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("3d Promotes" = "#E69F00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3d, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#E69F00") +
  labs(
    title = "Hypoxic TE Response: 3d-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3d promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density_3d)

ggsave(
  here("plots", "density_3d_promotes_hypoxia_characterization.pdf"),
  plot = p_density_3d,
  width = 10,
  height = 4
)
```

## 3e-Promoted Genes (Timepoint-Matched)

```{r density_3e_promotes, fig.width=10, fig.height=4}
# Get timepoint-specific gene sets
genes_3e_1hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_1hr_lfc0.5")
genes_3e_4hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_4hr_lfc0.5")
genes_3e_24hr <- get_geneset_genes("MDA-MB-231_hypoxia_3e_promotes_TE_24hr_lfc0.5")

cat("3e Promotes gene counts:\n")
cat("1hr:", length(genes_3e_1hr), "\n")
cat("4hr:", length(genes_3e_4hr), "\n")
cat("24hr:", length(genes_3e_24hr), "\n")

# Combine with timepoint-matched classification
te_combined_3e <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% genes_3e_1hr, "3e Promotes", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% genes_3e_4hr, "3e Promotes", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% genes_3e_24hr, "3e Promotes", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data for gene counts
label_data_3e <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(genes_3e_1hr)),
    paste0("n = ", length(genes_3e_4hr)),
    paste0("n = ", length(genes_3e_24hr))
  )
)

# Create density plot
p_density_3e <- ggplot(te_combined_3e, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("3e Promotes" = "#56B4E9", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_3e, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#56B4E9") +
  labs(
    title = "Hypoxic TE Response: 3e-Promoted Genes (Timepoint-Matched)",
    subtitle = "Hypoxia 3e promotes TE (LFC>0.5) at each timepoint",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density_3e)

ggsave(
  here("plots", "density_3e_promotes_hypoxia_characterization.pdf"),
  plot = p_density_3e,
  width = 10,
  height = 4
)
```

## Leading Edge Genes Only (3d Promotes)

This plot shows the distribution of ONLY the leading edge genes (those contributing to the enrichment signal) vs all other genes.

```{r density_leading_edge_3d, fig.width=10, fig.height=4}
# Get leading edge gene lists
le_genes_3d_1hr <- le_3d_1hr$gene
le_genes_3d_4hr <- le_3d_4hr$gene
le_genes_3d_24hr <- le_3d_24hr$gene

cat("3d Leading Edge gene counts:\n")
cat("1hr:", length(le_genes_3d_1hr), "\n")
cat("4hr:", length(le_genes_3d_4hr), "\n")
cat("24hr:", length(le_genes_3d_24hr), "\n")

# Combine TE data with leading edge classification
te_combined_le_3d <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% le_genes_3d_1hr, "Leading Edge", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% le_genes_3d_4hr, "Leading Edge", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% le_genes_3d_24hr, "Leading Edge", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data
label_data_le_3d <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(le_genes_3d_1hr)),
    paste0("n = ", length(le_genes_3d_4hr)),
    paste0("n = ", length(le_genes_3d_24hr))
  )
)

# Density plot for leading edge
p_density_le_3d <- ggplot(te_combined_le_3d, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("Leading Edge" = "#D55E00", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_le_3d, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#D55E00") +
  labs(
    title = "Hypoxic TE Response: Leading Edge Genes from 3d Promotes Enrichment",
    subtitle = "Full leading edge extracted from fgsea (hypoxia 3d promotes TE, LFC>0.5)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density_le_3d)

ggsave(
  here("plots", "density_leading_edge_3d_promotes_hypoxia.pdf"),
  plot = p_density_le_3d,
  width = 10,
  height = 4
)
```

## Leading Edge Genes Only (3e Promotes)

```{r density_leading_edge_3e, fig.width=10, fig.height=4}
# Get leading edge gene lists
le_genes_3e_1hr <- le_3e_1hr$gene
le_genes_3e_4hr <- le_3e_4hr$gene
le_genes_3e_24hr <- le_3e_24hr$gene

cat("3e Leading Edge gene counts:\n")
cat("1hr:", length(le_genes_3e_1hr), "\n")
cat("4hr:", length(le_genes_3e_4hr), "\n")
cat("24hr:", length(le_genes_3e_24hr), "\n")

# Combine TE data with leading edge classification
te_combined_le_3e <- bind_rows(
  te_1hr_gsea %>%
    mutate(timepoint = "1hr",
           gene_group = ifelse(symbol %in% le_genes_3e_1hr, "Leading Edge", "Other")),
  te_4hr_gsea %>%
    mutate(timepoint = "4hr",
           gene_group = ifelse(symbol %in% le_genes_3e_4hr, "Leading Edge", "Other")),
  te_24hr_gsea %>%
    mutate(timepoint = "24hr",
           gene_group = ifelse(symbol %in% le_genes_3e_24hr, "Leading Edge", "Other"))
) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Label data
label_data_le_3e <- data.frame(
  timepoint = factor(c("1hr", "4hr", "24hr"), levels = c("1hr", "4hr", "24hr")),
  label = c(
    paste0("n = ", length(le_genes_3e_1hr)),
    paste0("n = ", length(le_genes_3e_4hr)),
    paste0("n = ", length(le_genes_3e_24hr))
  )
)

# Density plot for leading edge
p_density_le_3e <- ggplot(te_combined_le_3e, aes(x = log2FoldChange, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(
    values = c("Leading Edge" = "#0072B2", "Other" = "#808080"),
    name = ""
  ) +
  facet_wrap(~timepoint, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(data = label_data_le_3e, aes(x = -1.5, y = Inf, label = label), vjust = 1.5,
            inherit.aes = FALSE, size = 4, color = "#0072B2") +
  labs(
    title = "Hypoxic TE Response: Leading Edge Genes from 3e Promotes Enrichment",
    subtitle = "Full leading edge extracted from fgsea (hypoxia 3e promotes TE, LFC>0.5)",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Density"
  ) +
  xlim(-2, 0.5) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_density_le_3e)

ggsave(
  here("plots", "density_leading_edge_3e_promotes_hypoxia.pdf"),
  plot = p_density_le_3e,
  width = 10,
  height = 4
)
```

# Leading Edge Summary Table

```{r leading_edge_summary_table}
# Create summary table of leading edge genes for 3d promotes
le_summary_3d <- data.frame(
  Timepoint = c("1hr", "4hr", "24hr"),
  Leading_Edge_Genes = c(
    length(le_genes_3d_1hr),
    length(le_genes_3d_4hr),
    length(le_genes_3d_24hr)
  ),
  Gene_Set_Size = c(
    le_3d_1hr$setSize[1],
    le_3d_4hr$setSize[1],
    le_3d_24hr$setSize[1]
  ),
  NES = c(
    round(le_3d_1hr$NES[1], 3),
    round(le_3d_4hr$NES[1], 3),
    round(le_3d_24hr$NES[1], 3)
  ),
  pval = c(
    format(le_3d_1hr$pval[1], scientific = TRUE, digits = 2),
    format(le_3d_4hr$pval[1], scientific = TRUE, digits = 2),
    format(le_3d_24hr$pval[1], scientific = TRUE, digits = 2)
  ),
  padj = c(
    format(le_3d_1hr$padj[1], scientific = TRUE, digits = 2),
    format(le_3d_4hr$padj[1], scientific = TRUE, digits = 2),
    format(le_3d_24hr$padj[1], scientific = TRUE, digits = 2)
  )
)

cat("\n3d Promotes Leading Edge Summary:\n")
print(le_summary_3d)
knitr::kable(le_summary_3d, caption = "3d Promotes: Leading Edge Summary by Timepoint")
```

# ORA Analysis: Leading Edge Genes with Hallmark Genesets

## Load Hallmark Genesets

```{r load_hallmark}
# Load MSigDB Hallmark gene sets
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB Hallmark gene sets:\n")
cat("Number of pathways:", length(unique(hallmark$gs_name)), "\n")
```

## Load KEGG Pathways

```{r load_kegg}
# Load MSigDB KEGG pathways
kegg <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>%
  dplyr::select(gs_name, gene_symbol)

cat("Loaded MSigDB KEGG pathways:\n")
cat("Number of pathways:", length(unique(kegg$gs_name)), "\n")
```

## ORA: 1hr Leading Edge Genes

```{r ora_1hr}
# Run ORA on 1hr leading edge genes
ora_1hr <- enricher(
  gene = le_genes_3d_1hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_1hr_results <- ora_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_1hr_results[] <- lapply(ora_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA 1hr - Top pathways (p.adjust < 0.05):\n")
print(ora_1hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

## ORA: 4hr Leading Edge Genes

```{r ora_4hr}
# Run ORA on 4hr leading edge genes
ora_4hr <- enricher(
  gene = le_genes_3d_4hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_4hr_results <- ora_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_4hr_results[] <- lapply(ora_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA 4hr - Top pathways (p.adjust < 0.05):\n")
print(ora_4hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

## ORA: 24hr Leading Edge Genes

```{r ora_24hr}
# Run ORA on 24hr leading edge genes
ora_24hr <- enricher(
  gene = le_genes_3d_24hr,
  TERM2GENE = hallmark,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_24hr_results <- ora_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_24hr_results[] <- lapply(ora_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("ORA 24hr - Top pathways (p.adjust < 0.05):\n")
print(ora_24hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

# ORA Analysis: Leading Edge Genes with KEGG Pathways

## KEGG ORA: 1hr Leading Edge Genes

```{r ora_kegg_1hr}
# Run ORA on 1hr leading edge genes with KEGG pathways
ora_kegg_1hr <- enricher(
  gene = le_genes_3d_1hr,
  TERM2GENE = kegg,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_kegg_1hr_results <- ora_kegg_1hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_kegg_1hr_results[] <- lapply(ora_kegg_1hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("KEGG ORA 1hr - Top pathways (p.adjust < 0.05):\n")
print(ora_kegg_1hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

## KEGG ORA: 4hr Leading Edge Genes

```{r ora_kegg_4hr}
# Run ORA on 4hr leading edge genes with KEGG pathways
ora_kegg_4hr <- enricher(
  gene = le_genes_3d_4hr,
  TERM2GENE = kegg,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_kegg_4hr_results <- ora_kegg_4hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_kegg_4hr_results[] <- lapply(ora_kegg_4hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("KEGG ORA 4hr - Top pathways (p.adjust < 0.05):\n")
print(ora_kegg_4hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

## KEGG ORA: 24hr Leading Edge Genes

```{r ora_kegg_24hr}
# Run ORA on 24hr leading edge genes with KEGG pathways
ora_kegg_24hr <- enricher(
  gene = le_genes_3d_24hr,
  TERM2GENE = kegg,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize = 10,
  maxGSSize = 500
)

# Extract results
ora_kegg_24hr_results <- ora_kegg_24hr@result %>%
  dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count) %>%
  mutate(GeneRatio_numeric = sapply(strsplit(GeneRatio, "/"), function(x) as.numeric(x[1])/as.numeric(x[2])))

ora_kegg_24hr_results[] <- lapply(ora_kegg_24hr_results, function(x) {
  if (is.numeric(x)) round(x, 4) else x
})

cat("KEGG ORA 24hr - Top pathways (p.adjust < 0.05):\n")
print(ora_kegg_24hr_results %>% filter(p.adjust < 0.05) %>% head(10))
```

## ORA Heatmap: Leading Edge Genes Across Timepoints

```{r ora_heatmap, fig.width=8, fig.height=10}
# Combine results from all timepoints
ora_combined <- bind_rows(
  ora_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_ora_pathways <- ora_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

cat("Pathways significant in at least one timepoint:", length(sig_ora_pathways), "\n")

if (length(sig_ora_pathways) > 0) {
  # Reshape for heatmap
  ora_wide <- ora_combined %>%
    filter(ID %in% sig_ora_pathways) %>%
    dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
    column_to_rownames("ID")

  # Reorder columns
  ora_wide <- ora_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

  # Create significance annotation matrix
  sig_ora_matrix <- ora_combined %>%
    filter(ID %in% sig_ora_pathways) %>%
    dplyr::select(ID, p.adjust, timepoint) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(ID, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")

  # Reorder columns to match ora_wide
  sig_ora_matrix <- sig_ora_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

  # Ensure row order matches
  sig_ora_matrix <- sig_ora_matrix[rownames(ora_wide), ]

  # Clean up pathway names (remove HALLMARK_ prefix)
  rownames(ora_wide) <- gsub("HALLMARK_", "", rownames(ora_wide))
  rownames(sig_ora_matrix) <- gsub("HALLMARK_", "", rownames(sig_ora_matrix))

  # Create heatmap
  p_ora_heatmap <- pheatmap(
    ora_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "red"))(100),
    breaks = seq(0, max(ora_wide), length.out = 101),
    display_numbers = sig_ora_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA: 3d Leading Edge Genes - Hallmark Pathways\n(* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots", "ora_heatmap_3d_leading_edge.pdf"), width = 8, height = 10)
  print(p_ora_heatmap)
  dev.off()

  cat("\nSaved: plots/ora_heatmap_3d_leading_edge.pdf\n")
} else {
  cat("No pathways significant at p.adjust < 0.05\n")
}
```

## KEGG ORA Heatmap: Leading Edge Genes Across Timepoints

```{r ora_kegg_heatmap, fig.width=8, fig.height=10}
# Combine KEGG results from all timepoints
ora_kegg_combined <- bind_rows(
  ora_kegg_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_kegg_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint),
  ora_kegg_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, GeneRatio_numeric, p.adjust, timepoint)
)

# Filter for pathways significant in at least one timepoint
sig_kegg_pathways <- ora_kegg_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>%
  unique()

cat("KEGG pathways significant in at least one timepoint:", length(sig_kegg_pathways), "\n")

if (length(sig_kegg_pathways) > 0) {
  # Reshape for heatmap
  ora_kegg_wide <- ora_kegg_combined %>%
    filter(ID %in% sig_kegg_pathways) %>%
    dplyr::select(ID, GeneRatio_numeric, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = GeneRatio_numeric, values_fill = 0) %>%
    column_to_rownames("ID")

  # Reorder columns
  ora_kegg_wide <- ora_kegg_wide %>% dplyr::select(`1hr`, `4hr`, `24hr`)

  # Create significance annotation matrix
  sig_kegg_matrix <- ora_kegg_combined %>%
    filter(ID %in% sig_kegg_pathways) %>%
    dplyr::select(ID, p.adjust, timepoint) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(ID, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")

  # Reorder columns to match ora_kegg_wide
  sig_kegg_matrix <- sig_kegg_matrix %>% dplyr::select(`1hr`, `4hr`, `24hr`)

  # Ensure row order matches
  sig_kegg_matrix <- sig_kegg_matrix[rownames(ora_kegg_wide), ]

  # Clean up pathway names (remove KEGG_ prefix)
  rownames(ora_kegg_wide) <- gsub("KEGG_", "", rownames(ora_kegg_wide))
  rownames(sig_kegg_matrix) <- gsub("KEGG_", "", rownames(sig_kegg_matrix))

  # Create heatmap
  p_ora_kegg_heatmap <- pheatmap(
    ora_kegg_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("white", "red"))(100),
    breaks = seq(0, max(ora_kegg_wide), length.out = 101),
    display_numbers = sig_kegg_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "ORA: 3d Leading Edge Genes - KEGG Pathways\n(* = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 12
  )

  # Save
  pdf(here("plots", "ora_heatmap_kegg_3d_leading_edge.pdf"), width = 8, height = 10)
  print(p_ora_kegg_heatmap)
  dev.off()

  cat("\nSaved: plots/ora_heatmap_kegg_3d_leading_edge.pdf\n")
} else {
  cat("No KEGG pathways significant at p.adjust < 0.05\n")
}
```

# 5' UTR Characterization of Leading Edge Genes

Characterize the 5' UTRs of the 1hr leading edge genes (3d promotes) by analyzing length and GC content.

## Load TxDb and Gene Info

```{r load_txdb}
# Load pre-built TxDb from Gencode v26
txdb <- loadDb("/Users/katematlin/github/mpra_preliminary/accessories/txdb.gencode26.sqlite")

# Load gene info for transcript-to-symbol mapping
geneInfo <- read_csv("/Users/katematlin/github/mpra_preliminary/accessories/gencode.v26.primary.info.csv.gz",
                     col_names = FALSE) %>%
  filter(!str_detect(X2, "pre_"))

colnames(geneInfo) <- c("gene_id", "transcript_id", "biotype", "symbol")

# Create transcript to gene mapping
tx2gene <- geneInfo[, c("transcript_id", "gene_id")]
gene_symbol_map <- geneInfo %>% dplyr::select(gene_id, symbol) %>% distinct()

cat("Loaded TxDb and gene info\n")
cat("Transcripts in geneInfo:", nrow(geneInfo), "\n")
```

## Load Salmon Quantification

```{r load_salmon}
# Get list of salmon quant files
salmon_dir <- here("2025_MDAcells_riboseq", "salmon")
salmon_samples <- list.dirs(salmon_dir, recursive = FALSE, full.names = FALSE)

# Build file paths
salmon_files <- file.path(salmon_dir, salmon_samples, "quant.sf.gz")
names(salmon_files) <- salmon_samples

# Filter to files that exist
salmon_files <- salmon_files[file.exists(salmon_files)]
cat("Found", length(salmon_files), "Salmon quant files\n")

# Import with tximport (transcript-level)
txi_transcripts <- tximport(salmon_files, type = "salmon", tx2gene = tx2gene, txOut = TRUE)

# Get transcript abundances and find most abundant transcript per gene
transcript_abundance <- txi_transcripts$abundance %>%
  as.data.frame() %>%
  rownames_to_column(var = "transcript_id") %>%
  left_join(tx2gene, by = "transcript_id")

# Sum counts across all samples and select most abundant transcript per gene
most_abundant_tx <- transcript_abundance %>%
  mutate(total_abundance = rowSums(across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::select(transcript_id, gene_id, total_abundance) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  group_by(gene_id_clean) %>%
  slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Selected most abundant transcript for", nrow(most_abundant_tx), "genes\n")
```

## Extract 5' UTR Regions

```{r extract_utr5}
# Extract 5' UTR regions from TxDb
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)

# Calculate 5' UTR widths
utr5_widths <- sum(width(utr5)) %>%
  as.data.frame() %>%
  rownames_to_column("transcript_id")
colnames(utr5_widths) <- c("transcript_id", "utr5_width")

cat("Extracted 5' UTR widths for", nrow(utr5_widths), "transcripts\n")
```

## Extract 5' UTR Sequences

```{r extract_utr5_sequences}
# Get genome sequences
genome <- BSgenome.Hsapiens.UCSC.hg38

# Extract 5' UTR sequences
utr5_sequences <- extractTranscriptSeqs(genome, utr5) %>% as.data.frame()

# Format sequences
utr5_seq <- utr5_sequences %>%
  rownames_to_column("transcript_id") %>%
  dplyr::rename("sequence" = "x") %>%
  mutate(seq_length = str_length(sequence))

cat("Extracted sequences for", nrow(utr5_seq), "5' UTRs\n")
```

## Calculate GC Content

```{r calculate_gc}
# GC content function
gc_content <- function(seq) {
  seq <- str_to_upper(seq)
  gc_count <- str_count(seq, "[GC]")
  total_length <- str_length(seq)
  return((gc_count / total_length) * 100)
}

# Add GC content to sequences
utr5_seq <- utr5_seq %>%
  mutate(gc_percent = gc_content(sequence))

cat("Calculated GC content\n")
cat("Median GC content:", round(median(utr5_seq$gc_percent, na.rm = TRUE), 2), "%\n")
```

## Merge UTR Data with Leading Edge Genes

```{r merge_utr_leading_edge}
# Map leading edge gene symbols to transcripts
# First, add gene symbols to most_abundant_tx
most_abundant_tx <- most_abundant_tx %>%
  left_join(gene_symbol_map %>% mutate(gene_id_clean = sub("\\..*", "", gene_id)),
            by = "gene_id_clean")

# Merge UTR data with transcript selection
utr_data <- most_abundant_tx %>%
  left_join(utr5_widths, by = "transcript_id") %>%
  left_join(utr5_seq %>% dplyr::select(transcript_id, gc_percent), by = "transcript_id") %>%
  filter(!is.na(utr5_width), !is.na(gc_percent))

# Classify genes: Leading Edge (1hr) vs Other
utr_data <- utr_data %>%
  mutate(gene_group = ifelse(symbol %in% le_genes_3d_1hr, "Leading Edge (1hr)", "Other"))

cat("\nUTR data merged:\n")
cat("Total genes with UTR data:", nrow(utr_data), "\n")
cat("Leading Edge (1hr) genes:", sum(utr_data$gene_group == "Leading Edge (1hr)"), "\n")
cat("Other genes:", sum(utr_data$gene_group == "Other"), "\n")
```

## Density Plot: 5' UTR Length

```{r density_utr_length, fig.width=8, fig.height=5}
# Calculate medians
median_le <- median(utr_data$utr5_width[utr_data$gene_group == "Leading Edge (1hr)"], na.rm = TRUE)
median_other <- median(utr_data$utr5_width[utr_data$gene_group == "Other"], na.rm = TRUE)

# Count genes in each group
n_le <- sum(utr_data$gene_group == "Leading Edge (1hr)")
n_other <- sum(utr_data$gene_group == "Other")

cat("Median 5' UTR length:\n")
cat("Leading Edge (1hr):", round(median_le, 1), "nt (n =", n_le, ")\n")
cat("Other:", round(median_other, 1), "nt (n =", n_other, ")\n")

# Density plot with log2 scale
p_utr_length <- ggplot(utr_data, aes(x = utr5_width, color = gene_group)) +
  geom_density(size = 1) +
  scale_x_continuous(trans = "log2",
                     breaks = trans_breaks("log2", function(x) 2^x, n = 10),
                     labels = trans_format("log2", math_format(2^.x))) +
  scale_color_manual(values = c("Leading Edge (1hr)" = "#E69F00", "Other" = "#808080"),
                     name = "") +
  geom_vline(xintercept = median_le, color = "#E69F00", linetype = "dashed", alpha = 0.7) +
  geom_vline(xintercept = median_other, color = "#808080", linetype = "dashed", alpha = 0.7) +
  annotate("text", x = 16, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Leading Edge: n = ", n_le), color = "#E69F00", size = 4) +
  annotate("text", x = 16, y = Inf, vjust = 3, hjust = 0,
           label = paste0("Other: n = ", n_other), color = "#808080", size = 4) +
  labs(
    title = "5' UTR Length: 3d Leading Edge (1hr) vs Other Genes",
    x = "5' UTR Length (nt, log2 scale)",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_utr_length)

ggsave(here("plots", "density_utr5_length_leading_edge.pdf"), p_utr_length, width = 8, height = 5)

# Wilcoxon test
wilcox_length <- utr_data %>%
  wilcox_test(utr5_width ~ gene_group, ref.group = "Other")

cat("\nWilcoxon test (5' UTR length):\n")
print(wilcox_length)
```

## Density Plot: GC Content

```{r density_gc_content, fig.width=8, fig.height=5}
# Calculate medians
median_gc_le <- median(utr_data$gc_percent[utr_data$gene_group == "Leading Edge (1hr)"], na.rm = TRUE)
median_gc_other <- median(utr_data$gc_percent[utr_data$gene_group == "Other"], na.rm = TRUE)

cat("Median 5' UTR GC content:\n")
cat("Leading Edge (1hr):", round(median_gc_le, 2), "% (n =", n_le, ")\n")
cat("Other:", round(median_gc_other, 2), "% (n =", n_other, ")\n")

# Density plot
p_gc_content <- ggplot(utr_data, aes(x = gc_percent, color = gene_group)) +
  geom_density(size = 1) +
  scale_color_manual(values = c("Leading Edge (1hr)" = "#E69F00", "Other" = "#808080"),
                     name = "") +
  geom_vline(xintercept = median_gc_le, color = "#E69F00", linetype = "dashed", alpha = 0.7) +
  geom_vline(xintercept = median_gc_other, color = "#808080", linetype = "dashed", alpha = 0.7) +
  annotate("text", x = 20, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Leading Edge: n = ", n_le), color = "#E69F00", size = 4) +
  annotate("text", x = 20, y = Inf, vjust = 3, hjust = 0,
           label = paste0("Other: n = ", n_other), color = "#808080", size = 4) +
  labs(
    title = "5' UTR GC Content: 3d Leading Edge (1hr) vs Other Genes",
    x = "GC Content (%)",
    y = "Density"
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

print(p_gc_content)

ggsave(here("plots", "density_utr5_gc_leading_edge.pdf"), p_gc_content, width = 8, height = 5)

# Wilcoxon test
wilcox_gc <- utr_data %>%
  wilcox_test(gc_percent ~ gene_group, ref.group = "Other")

cat("\nWilcoxon test (GC content):\n")
print(wilcox_gc)
```

## Summary Table: UTR Characteristics

```{r utr_summary_table}
# Summary statistics
utr_summary <- utr_data %>%
  group_by(gene_group) %>%
  summarize(
    n_genes = n(),
    median_length = median(utr5_width, na.rm = TRUE),
    mean_length = mean(utr5_width, na.rm = TRUE),
    median_gc = median(gc_percent, na.rm = TRUE),
    mean_gc = mean(gc_percent, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n5' UTR Characteristics Summary:\n")
print(utr_summary)
knitr::kable(utr_summary, digits = 2, caption = "5' UTR Characteristics: Leading Edge vs Other Genes")
```

# 5' UTR Sequence Motif Analysis

Compare sequence motifs in 5' UTRs of Leading Edge (1hr) genes versus size-matched control genes.

## Prepare Sequence Sets for Motif Analysis

```{r prepare_motif_sequences}
# Get sequences for leading edge genes
le_sequences <- utr_data %>%
  filter(gene_group == "Leading Edge (1hr)") %>%
  left_join(utr5_seq %>% dplyr::select(transcript_id, sequence), by = "transcript_id") %>%
  filter(!is.na(sequence), sequence != "")

le_seq_set <- DNAStringSet(le_sequences$sequence)

# Get control sequences (size-matched random sample from "Other" genes)
other_sequences <- utr_data %>%
  filter(gene_group == "Other") %>%
  left_join(utr5_seq %>% dplyr::select(transcript_id, sequence), by = "transcript_id") %>%
  filter(!is.na(sequence), sequence != "")

set.seed(42)
n_le <- nrow(le_sequences)
control_sample <- other_sequences[sample(nrow(other_sequences), n_le), ]
control_seq_set <- DNAStringSet(control_sample$sequence)

cat("Sequence sets prepared:\n")
cat("Leading Edge sequences:", length(le_seq_set), "\n")
cat("Control sequences:", length(control_seq_set), "\n")
```

## RBP Motif Analysis with RBNS_PWM (RNA Bind-N-Seq)

```{r rbns_motif_analysis, fig.width=10, fig.height=8}
# Run motif comparison using RBNS PWMs
rbns_results <- motif_compare(RBNS_PWM, le_seq_set, control_seq_set)

# Filter significant results
rbns_sig <- rbns_results %>%
  filter(p_adj < 0.05)

cat("RBNS PWM Analysis:\n")
cat("Total motifs tested:", nrow(rbns_results), "\n")
cat("Significant motifs (p.adj < 0.05):", nrow(rbns_sig), "\n")

# Visualize enriched motifs
if (nrow(rbns_sig %>% filter(log2FC > 0)) > 0) {
  p_rbns <- ggplot(rbns_sig %>% filter(log2FC > 0), aes(x = reorder(motif, log2FC), y = log2FC)) +
    geom_bar(stat = "identity", fill = "#E69F00") +
    coord_flip() +
    theme_cowplot() +
    labs(title = "Enriched RBP Motifs in 3d Leading Edge 5' UTRs",
         subtitle = "RNA Bind-N-Seq PWMs, p.adj < 0.05",
         x = "", y = "log2 Fold Change")

  print(p_rbns)

  ggsave(here("plots", "motif_rbns_3d_leading_edge.pdf"), p_rbns, width = 10, height = 8)
  cat("\nSaved: plots/motif_rbns_3d_leading_edge.pdf\n")
} else {
  cat("No significantly enriched RBNS motifs found\n")
}

# Show significant results
if (nrow(rbns_sig) > 0) {
  cat("\nSignificant RBNS motifs:\n")
  print(rbns_sig %>% arrange(desc(log2FC)))
}
```

## RBP Motif Analysis with CISBPRNA_hs_PWM (CLIP-seq)

```{r cisbp_motif_analysis, fig.width=10, fig.height=8}
# Run motif comparison using CISBP-RNA PWMs
cisbp_results <- motif_compare(CISBPRNA_hs_PWM, le_seq_set, control_seq_set)

# Filter significant results
cisbp_sig <- cisbp_results %>%
  filter(p_adj < 0.05)

cat("CISBP-RNA PWM Analysis:\n")
cat("Total motifs tested:", nrow(cisbp_results), "\n")
cat("Significant motifs (p.adj < 0.05):", nrow(cisbp_sig), "\n")

# Visualize enriched motifs
if (nrow(cisbp_sig %>% filter(log2FC > 0)) > 0) {
  p_cisbp <- ggplot(cisbp_sig %>% filter(log2FC > 0), aes(x = reorder(motif, log2FC), y = log2FC)) +
    geom_bar(stat = "identity", fill = "#E69F00") +
    coord_flip() +
    theme_cowplot() +
    labs(title = "Enriched RBP Motifs in 3d Leading Edge 5' UTRs",
         subtitle = "CISBP-RNA CLIP-seq PWMs, p.adj < 0.05",
         x = "", y = "log2 Fold Change")

  print(p_cisbp)

  ggsave(here("plots", "motif_cisbp_3d_leading_edge.pdf"), p_cisbp, width = 10, height = 8)
  cat("\nSaved: plots/motif_cisbp_3d_leading_edge.pdf\n")
} else {
  cat("No significantly enriched CISBP-RNA motifs found\n")
}

# Show significant results
if (nrow(cisbp_sig) > 0) {
  cat("\nSignificant CISBP-RNA motifs:\n")
  print(cisbp_sig %>% arrange(desc(log2FC)))
}
```

## 6-mer Enrichment Analysis

```{r sixmer_analysis, fig.width=10, fig.height=8}
# Count 6-mers in each set
sixmers <- data.frame(
  oligonucleotideFrequency(le_seq_set, width = 6, simplify.as = "matrix") %>% colSums(),
  oligonucleotideFrequency(control_seq_set, width = 6, simplify.as = "matrix") %>% colSums()
)
colnames(sixmers) <- c("le_count", "control_count")

# Calculate frequency and enrichment
sixmer_enrich <- sixmers %>%
  mutate(le_freq = le_count / sum(le_count),
         control_freq = control_count / sum(control_count),
         log2FC = log2(le_freq / control_freq)) %>%
  rownames_to_column("motif") %>%
  mutate(le_NOT_count = sum(le_count) - le_count,
         control_NOT_count = sum(control_count) - control_count)

# Fisher's exact test for each 6-mer
fisher_results <- data.frame()
for (i in sixmer_enrich$motif) {
  test_row <- sixmer_enrich %>% filter(motif == i)
  conting <- data.frame(
    "case" = c(test_row$le_count, test_row$le_NOT_count),
    "control" = c(test_row$control_count, test_row$control_NOT_count)
  )
  fisher_p <- fisher_test(conting, detailed = TRUE)$p
  fisher_results <- rbind(fisher_results, data.frame(motif = i, pval = fisher_p))
}

sixmer_final <- sixmer_enrich %>%
  left_join(fisher_results, by = "motif") %>%
  mutate(padj = p.adjust(pval, method = "fdr"))

# Filter significant enriched 6-mers
sig_sixmers <- sixmer_final %>%
  filter(padj < 0.05, log2FC > 0) %>%
  arrange(desc(log2FC))

cat("6-mer Enrichment Analysis:\n")
cat("Total 6-mers tested:", nrow(sixmer_final), "\n")
cat("Significant enriched 6-mers (FDR < 0.05, log2FC > 0):", nrow(sig_sixmers), "\n")

# Visualize top enriched 6-mers
if (nrow(sig_sixmers) > 0) {
  p_sixmer <- ggplot(sig_sixmers %>% head(20), aes(x = reorder(motif, log2FC), y = log2FC)) +
    size=6+
    geom_bar(stat = "identity", fill = "#E69F00") +
    coord_flip() +
    theme_cowplot() +
    labs(title = "Top Enriched 6-mers in 3d Leading Edge 5' UTRs",
         subtitle = "Fisher's exact test, FDR < 0.05",
         x = "", y = "log2 Fold Change over Control")

  print(p_sixmer)

  ggsave(here("plots", "sixmer_enrichment_3d_leading_edge.pdf"), p_sixmer, width = 10, height = 8)
  cat("\nSaved: plots/sixmer_enrichment_3d_leading_edge.pdf\n")

  cat("\nTop 20 enriched 6-mers:\n")
  print(sig_sixmers %>% head(20) %>% dplyr::select(motif, log2FC, le_count, control_count, padj))
} else {
  cat("No significantly enriched 6-mers found\n")
}
```

# Export Results

```{r export_results}
# Export leading edge gene lists
leading_edge_export <- list(
  "3d_1hr" = le_3d_1hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count),
  "3d_4hr" = le_3d_4hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count),
  "3d_24hr" = le_3d_24hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count),
  "3e_1hr" = le_3e_1hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count),
  "3e_4hr" = le_3e_4hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count),
  "3e_24hr" = le_3e_24hr %>% dplyr::select(gene, NES, pval, padj, setSize, leading_edge_count)
)

write_xlsx(leading_edge_export, here("output", "3d_3e_promotes_leading_edge_genes.xlsx"))

# Export fgsea summary for 3d and 3e promotes genesets
fgsea_summary <- bind_rows(
  fgsea_1hr %>%
    filter(str_detect(pathway, "3d_promotes|3e_promotes")) %>%
    filter(str_detect(pathway, "lfc0.5")) %>%
    mutate(ranked_list_timepoint = "1hr"),
  fgsea_4hr %>%
    filter(str_detect(pathway, "3d_promotes|3e_promotes")) %>%
    filter(str_detect(pathway, "lfc0.5")) %>%
    mutate(ranked_list_timepoint = "4hr"),
  fgsea_24hr %>%
    filter(str_detect(pathway, "3d_promotes|3e_promotes")) %>%
    filter(str_detect(pathway, "lfc0.5")) %>%
    mutate(ranked_list_timepoint = "24hr")
) %>%
  mutate(leading_edge_count = sapply(leadingEdge, length)) %>%
  dplyr::select(pathway, ranked_list_timepoint, NES, pval, padj, size, leading_edge_count) %>%
  arrange(pathway, ranked_list_timepoint)

write_csv(fgsea_summary, here("output", "3d_3e_promotes_fgsea_summary.csv"))

cat("\nExported results:\n")
cat("- output/3d_3e_promotes_leading_edge_genes.xlsx\n")
cat("- output/3d_3e_promotes_fgsea_summary.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
