---
title: "Condition-Level Translation Efficiency (Single DESeq2 Run)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

# Overview

This notebook calculates **condition-level TE** — one log2(ribo/rna) value per gene per
group — using a single DESeq2 run on all samples together. Groups are defined as every
combination of condition × treatment × timepoint × library. Replicates within each group
estimate the variance; the contrast always extracts ribo vs rna within a given context.

Spike-in size factors are used to normalize across libraries.

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(ashr)
library(ggplot2)
library(cowplot)
```

```{r load_functions}
source(here("code/functions.R"))
```

---

# 1. Load Data

```{r load_counts}
allCounts <- read_csv(here("output", "human_allCounts_raw.csv"),
                      show_col_types = FALSE) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv"),
                    show_col_types = FALSE) %>%
  column_to_rownames("gene_id")

metadata <- read_csv(here("output", "metadata_clean.csv"),
                     show_col_types = FALSE)

cat("allCounts:", dim(allCounts), "\n")
cat("spikein:", dim(spikein), "\n")
cat("metadata:", nrow(metadata), "samples\n")
```

```{r load_gene_anno}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"),
                      show_col_types = FALSE)

gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol) %>%
  arrange(gene_id) %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup()

cat("Gene symbols loaded:", nrow(gene_symbols), "unique genes\n")
```

---

# 2. Spike-In Selection

Select top 500 spike-in genes by total expression for size factor estimation.

```{r select_spikes}
top500_ids <- spikein %>%
  rowSums() %>%
  sort(decreasing = TRUE) %>%
  head(500) %>%
  names()

cat("Using", length(top500_ids), "spike-in control genes\n")
```

---

# 3. Build DESeq2 Object

## 3.1 Combine counts and build group factor

```{r build_metadata}
# Align spike-in columns to allCounts
spikein_aligned <- spikein[, colnames(allCounts), drop = FALSE]

combined_counts <- rbind(allCounts, spikein_aligned)
spike_idx <- which(rownames(combined_counts) %in% top500_ids)

cat("Combined count matrix:", dim(combined_counts), "\n")
cat("Spike-in row indices:", length(spike_idx), "\n")

# Build group factor: condition_treatment_hours_library
# Coerce hours to character to avoid numeric formatting issues
metadata_dds <- metadata %>%
  mutate(group = paste(condition, treatment, as.character(hours), library, sep = "_")) %>%
  column_to_rownames("sample_name")

# Verify colnames match rownames
stopifnot(all(colnames(combined_counts) %in% rownames(metadata_dds)))
metadata_dds <- metadata_dds[colnames(combined_counts), ]

cat("\nGroup levels in data:\n")
print(sort(unique(metadata_dds$group)))
cat("\nTotal groups:", length(unique(metadata_dds$group)), "\n")
```

## 3.2 Create DESeq2 dataset, estimate size factors, remove spike-ins

```{r build_dds}
dds <- DESeqDataSetFromMatrix(
  countData = round(combined_counts),
  colData   = metadata_dds,
  design    = ~ group
)

# Spike-in size factors
dds <- estimateSizeFactors(dds, controlGenes = spike_idx)

cat("\nSize factors (first 10 samples):\n")
print(round(head(sizeFactors(dds), 10), 3))

# Remove spike-ins AFTER size factor estimation
dds <- dds[!grepl("^FBgn", rownames(dds)), ]
cat("\nGenes remaining after spike-in removal:", nrow(dds), "\n")
```

## 3.3 Run DESeq2

This may take several minutes with all samples.

```{r run_deseq, cache=TRUE}
dds <- DESeq(dds)
cat("DESeq2 complete.\n")
print(head(resultsNames(dds)))
```

---

# 4. Extract TE Contrasts

For every combination of condition × treatment × timepoint, compare ribo vs rna to get TE.

```{r define_groups}
conditions  <- c("normoxia", "hypoxia")
treatments  <- c("sictrl", "si3e", "si3d")
timepoints  <- c(1, 4, 24)

# All expected group levels
expected_groups <- expand.grid(
  condition = conditions,
  treatment = treatments,
  hours     = timepoints,
  library   = c("ribo", "rna"),
  stringsAsFactors = FALSE
) %>%
  mutate(group = paste(condition, treatment, as.character(hours), library, sep = "_")) %>%
  pull(group)

available_groups <- levels(dds$group)
missing_groups   <- setdiff(expected_groups, available_groups)

if (length(missing_groups) > 0) {
  cat("WARNING: The following expected groups are missing from the model:\n")
  print(missing_groups)
} else {
  cat("All expected group levels are present in the model.\n")
}
```

```{r extract_te}
# Build a table of every contrast we want to extract:
# one row per condition × treatment × timepoint combination.
# For each row, the contrast is ribo vs rna within that group context.
contrast_grid <- expand.grid(
  condition = conditions,
  treatment = treatments,
  hours     = timepoints,
  stringsAsFactors = FALSE
) %>%
  mutate(
    ribo_grp = paste(condition, treatment, as.character(hours), "ribo", sep = "_"),
    rna_grp  = paste(condition, treatment, as.character(hours), "rna",  sep = "_"),
    # Flag rows where a group is absent from the DESeq2 model (e.g. missing replicate)
    skip     = !ribo_grp %in% available_groups | !rna_grp %in% available_groups
  )

cat("Contrasts to extract:", nrow(contrast_grid), "\n")
if (any(contrast_grid$skip)) {
  cat("Skipping (group missing from model):\n")
  print(contrast_grid %>% filter(skip) %>% dplyr::select(condition, treatment, hours, ribo_grp, rna_grp))
}

# Iterate over each contrast row and collect results into a list
results_list <- vector("list", nrow(contrast_grid))

for (i in seq_len(nrow(contrast_grid))) {
  row <- contrast_grid[i, ]

  if (row$skip) next

  message("Extracting TE [", i, "/", nrow(contrast_grid), "]: ",
          row$ribo_grp, " vs ", row$rna_grp)

  # Step 1: get raw DESeq2 results for this ribo-vs-rna contrast
  res_raw <- results(dds,
                     contrast = c("group", row$ribo_grp, row$rna_grp),
                     independentFiltering = FALSE)

  # Step 2: apply ashr LFC shrinkage
  # apeglm cannot be used here because it requires a single named model coefficient
  # (e.g. from resultsNames(dds)), and fits a penalised GLM around that one term.
  # Our contrasts are arbitrary linear combinations of group levels — c("group", ribo, rna) —
  # which are not directly represented as named coefficients in the model.
  # ashr works from the distribution of observed LFCs and their standard errors,
  # so it can shrink any contrast without needing a named coefficient.
  res_shrunk <- lfcShrink(dds,
                          contrast = c("group", row$ribo_grp, row$rna_grp),
                          res      = res_raw,
                          type     = "ashr")

  # Step 3: convert to data frame and attach the group labels
  results_list[[i]] <- as.data.frame(res_shrunk) %>%
    rownames_to_column("gene_id") %>%
    mutate(condition = row$condition,
           treatment = row$treatment,
           hours     = row$hours)
}

# Combine all per-contrast results into one long data frame
te_results <- bind_rows(results_list)

cat("\nTE results dimensions:", dim(te_results), "\n")
cat("Unique condition × treatment × hours combinations:\n")
print(te_results %>% distinct(condition, treatment, hours) %>% arrange(condition, treatment, hours))
```

---

# 5. Join Gene Symbols and Save

```{r join_symbols}
te_results_anno <- te_results %>%
  left_join(gene_symbols, by = "gene_id") %>%
  dplyr::select(gene_id, symbol, condition, treatment, hours,
                log2FoldChange, lfcSE, pvalue, padj, baseMean)

cat("Final table dimensions:", dim(te_results_anno), "\n")
```

```{r save_output}
write_csv(te_results_anno, here("output", "condition_level_te.csv"))
cat("Saved: output/condition_level_te.csv\n")
```

---

# 6. Sanity Check Plots

## Check 1: Model structure

```{r check_model_structure}
cat("Group levels in DESeq2 model:\n")
grp_levels <- levels(dds$group)
print(grp_levels)
cat("\nTotal:", length(grp_levels), "groups (expected 36 = 2 conditions × 3 treatments × 3 timepoints × 2 libraries)\n")

cat("\nSize factors summary:\n")
sf_df <- data.frame(
  sample     = names(sizeFactors(dds)),
  size_factor = sizeFactors(dds)
) %>%
  left_join(metadata %>% dplyr::select(sample_name, library),
            by = c("sample" = "sample_name"))

ggplot(sf_df, aes(x = library, y = size_factor, color = library)) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_color_manual(values = c("rna" = "steelblue", "ribo" = "tomato")) +
  labs(title = "Spike-in size factors by library type",
       y = "Size factor", x = "Library") +
  theme_cowplot(12) +
  theme(legend.position = "none")
```

## Check 2: Global TE distribution (sictrl)

Hypoxia sictrl should show globally suppressed TE compared to normoxia sictrl.

```{r check_te_distribution}
te_sictrl <- te_results_anno %>%
  filter(treatment == "sictrl", !is.na(log2FoldChange))

# Median TE per condition × timepoint
te_medians <- te_sictrl %>%
  group_by(condition, hours) %>%
  summarise(median_te = median(log2FoldChange, na.rm = TRUE), .groups = "drop")

cat("Median TE by condition and timepoint (sictrl):\n")
print(te_medians)

ggplot(te_sictrl, aes(x = log2FoldChange, color = condition)) +
  geom_density() +
  facet_wrap(~ hours, labeller = label_both) +
  scale_color_manual(values = c("normoxia" = "steelblue", "hypoxia" = "tomato")) +
  xlim(-5, 5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "TE distribution: sictrl, normoxia vs hypoxia",
       x = "log2 TE (ribo / rna)", y = "Density") +
  theme_cowplot(12)
```

## Check 3: ATF4 sanity check

ATF4 TE should be elevated in hypoxia vs normoxia (especially 1hr and 4hr sictrl) — a known translational responder.

```{r check_atf4}
atf4_te <- te_results_anno %>%
  filter(symbol == "ATF4")

cat("ATF4 TE results:\n")
print(atf4_te %>%
        dplyr::select(condition, treatment, hours, log2FoldChange, padj) %>%
        arrange(treatment, hours, condition))

ggplot(atf4_te, aes(x = factor(hours), y = log2FoldChange,
                    fill = condition, group = condition)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  facet_wrap(~ treatment) +
  scale_fill_manual(values = c("normoxia" = "steelblue", "hypoxia" = "tomato")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "ATF4 Translation Efficiency",
       x = "Hours", y = "log2 TE (ribo / rna)") +
  theme_cowplot(12)
```

## Check 4: Correlation with per-sample TE

Cross-validate DESeq2 group-level TE against the mean of per-sample TE values from `complete_te_list_all_*hr.csv`.

```{r check_correlation}
# Load existing per-sample TE tables and extract TE columns only
load_persample_te <- function(hr) {
  fname <- here("output", paste0("complete_te_list_all_", hr, "hr.csv"))
  df <- read_csv(fname, show_col_types = FALSE)

  # TE columns are labeled like: normoxia_sictrl_1_a_TE_NM2025_XXXX
  te_cols <- grep("_TE_", colnames(df), value = TRUE)

  df %>%
    dplyr::select(gene_id, all_of(te_cols)) %>%
    pivot_longer(cols = all_of(te_cols),
                 names_to  = "sample_label",
                 values_to = "per_sample_te") %>%
    # Parse condition, treatment, hours from sample label
    # Format: condition_treatment_hours_rep_TE_id
    mutate(
      parts     = str_split(sample_label, "_"),
      condition = map_chr(parts, 1),
      treatment = map_chr(parts, 2),
      hours     = as.numeric(map_chr(parts, 3)),
      rep       = map_chr(parts, 4)
    ) %>%
    dplyr::select(gene_id, condition, treatment, hours, rep, per_sample_te)
}

persample_te <- map_dfr(c(1, 4, 24), load_persample_te)

# Average per-sample TE per gene × condition × treatment × hours
mean_persample_te <- persample_te %>%
  group_by(gene_id, condition, treatment, hours) %>%
  summarise(mean_per_sample_te = mean(per_sample_te, na.rm = TRUE), .groups = "drop")

# Join with DESeq2 group-level TE
te_compare <- te_results_anno %>%
  dplyr::select(gene_id, condition, treatment, hours, deseq2_te = log2FoldChange) %>%
  inner_join(mean_persample_te, by = c("gene_id", "condition", "treatment", "hours")) %>%
  filter(!is.na(deseq2_te), !is.na(mean_per_sample_te))

cat("Genes available for cross-validation:", nrow(te_compare), "\n")

# Correlation per condition × treatment × hours
corr_summary <- te_compare %>%
  group_by(condition, treatment, hours) %>%
  summarise(r = cor(mean_per_sample_te, deseq2_te, use = "complete.obs"),
            n = n(),
            .groups = "drop")

cat("\nCorrelation between per-sample mean TE and DESeq2 group TE:\n")
print(corr_summary %>% arrange(condition, treatment, hours))

# Per-replicate TE vs DESeq2 group TE, faceted by timepoint × condition × treatment
te_compare_reps <- persample_te %>%
  inner_join(
    te_results_anno %>% dplyr::select(gene_id, condition, treatment, hours,
                                       deseq2_te = log2FoldChange),
    by = c("gene_id", "condition", "treatment", "hours")
  ) %>%
  filter(!is.na(per_sample_te), !is.na(deseq2_te))

r_labels <- te_compare %>%
  group_by(condition, treatment, hours) %>%
  summarise(r = round(cor(mean_per_sample_te, deseq2_te, use = "complete.obs"), 3),
            .groups = "drop") %>%
  mutate(label = paste0("r = ", r),
         mean_per_sample_te = 5.5, deseq2_te = -4.5)

ggplot(te_compare, aes(x = mean_per_sample_te, y = deseq2_te)) +
  geom_point(alpha = 0.2, size = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_text(data = r_labels, aes(label = label), hjust = 1, vjust = 0, size = 2.5) +
  facet_grid(hours ~ condition + treatment, labeller = label_value) +
  xlim(-6, 6) + ylim(-6, 6) +
  labs(x = "Mean per-replicate TE (log2 ribo/rna)",
       y = "DESeq2 group TE (log2FC)") +
  theme_cowplot(9) +
  theme(strip.text = element_text(size = 7))

# Per-replicate scatter for sictrl 1hr normoxia
rep_norm_1hr <- te_compare_reps %>%
  filter(treatment == "sictrl", condition == "normoxia", hours == 1)

r_labels_norm <- rep_norm_1hr %>%
  group_by(rep) %>%
  summarise(r = round(cor(per_sample_te, deseq2_te, use = "complete.obs"), 3), .groups = "drop") %>%
  mutate(label = paste0("r = ", r), per_sample_te = 5.5, deseq2_te = -4.5)

ggplot(rep_norm_1hr, aes(x = per_sample_te, y = deseq2_te)) +
  geom_point(alpha = 0.3, size = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_text(data = r_labels_norm, aes(label = label), hjust = 1, vjust = 0, size = 3) +
  facet_wrap(~ rep, nrow = 1) +
  xlim(-6, 6) + ylim(-6, 6) +
  labs(title = "sictrl normoxia 1hr — per replicate",
       x = "Per-replicate TE (log2 ribo/rna)",
       y = "DESeq2 group TE (log2FC)") +
  theme_cowplot(11)

rep_norm_1hr %>%
  pivot_longer(cols = c(per_sample_te, deseq2_te),
               names_to = "source", values_to = "te") %>%
  mutate(source = recode(source,
                         per_sample_te = "Per-replicate TE",
                         deseq2_te     = "DESeq2 group TE")) %>%
  ggplot(aes(x = te, fill = source, color = source)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ rep, nrow = 1) +
  xlim(-6, 6) +
  scale_fill_manual(values = c("Per-replicate TE" = "steelblue", "DESeq2 group TE" = "grey40")) +
  scale_color_manual(values = c("Per-replicate TE" = "steelblue", "DESeq2 group TE" = "grey40")) +
  labs(title = "sictrl normoxia 1hr — TE distributions per replicate",
       x = "TE (log2)", y = "Density", fill = NULL, color = NULL) +
  theme_cowplot(11) +
  theme(legend.position = "bottom")

# Per-replicate scatter for sictrl 1hr hypoxia
rep_hyp_1hr <- te_compare_reps %>%
  filter(treatment == "sictrl", condition == "hypoxia", hours == 1)

r_labels_hyp <- rep_hyp_1hr %>%
  group_by(rep) %>%
  summarise(r = round(cor(per_sample_te, deseq2_te, use = "complete.obs"), 3), .groups = "drop") %>%
  mutate(label = paste0("r = ", r), per_sample_te = 5.5, deseq2_te = -4.5)

ggplot(rep_hyp_1hr, aes(x = per_sample_te, y = deseq2_te)) +
  geom_point(alpha = 0.3, size = 0.6, color = "tomato") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_text(data = r_labels_hyp, aes(label = label), hjust = 1, vjust = 0, size = 3) +
  facet_wrap(~ rep, nrow = 1) +
  xlim(-6, 6) + ylim(-6, 6) +
  labs(title = "sictrl hypoxia 1hr — per replicate",
       x = "Per-replicate TE (log2 ribo/rna)",
       y = "DESeq2 group TE (log2FC)") +
  theme_cowplot(11)

rep_hyp_1hr %>%
  pivot_longer(cols = c(per_sample_te, deseq2_te),
               names_to = "source", values_to = "te") %>%
  mutate(source = recode(source,
                         per_sample_te = "Per-replicate TE",
                         deseq2_te     = "DESeq2 group TE")) %>%
  ggplot(aes(x = te, fill = source, color = source)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ rep, nrow = 1) +
  xlim(-6, 6) +
  scale_fill_manual(values = c("Per-replicate TE" = "tomato", "DESeq2 group TE" = "grey40")) +
  scale_color_manual(values = c("Per-replicate TE" = "tomato", "DESeq2 group TE" = "grey40")) +
  labs(title = "sictrl hypoxia 1hr — TE distributions per replicate",
       x = "TE (log2)", y = "Density", fill = NULL, color = NULL) +
  theme_cowplot(11) +
  theme(legend.position = "bottom")
```

## Check 5: Knockdown validation

Verify that siRNA knockdowns behave as expected at the TE level.

```{r check_knockdowns}
kd_genes <- te_results_anno %>%
  filter(symbol %in% c("EIF3E", "EIF3D"))

cat("EIF3E and EIF3D TE across conditions:\n")
print(kd_genes %>%
        dplyr::select(symbol, condition, treatment, hours, log2FoldChange, padj) %>%
        arrange(symbol, treatment, hours, condition))

ggplot(kd_genes, aes(x = factor(hours), y = log2FoldChange,
                     color = condition, shape = treatment, group = interaction(condition, treatment))) +
  geom_point(size = 3, position = position_dodge(0.4)) +
  geom_line(position = position_dodge(0.4), alpha = 0.5) +
  facet_wrap(~ symbol) +
  scale_color_manual(values = c("normoxia" = "steelblue", "hypoxia" = "tomato")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "EIF3D / EIF3E Translation Efficiency (knockdown validation)",
       x = "Hours", y = "log2 TE (ribo / rna)") +
  theme_cowplot(12)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
