---
title: "Individual Gene Visualization"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This notebook provides functions to visualize individual gene expression (RNA, Ribo, and TE) across normoxia/hypoxia conditions with and without eIF3d/3e knockdowns. Data is spike-in normalized.

## Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggplot2)
library(cowplot)
library(ggprism)

# Source custom functions
source(here("code/functions.R"))
```

## Load Data

```{r load_data}
# Load spike-in normalized counts for all timepoints
counts_1hr <- read_csv(here("output", "allCountsNorm_all_1hr_SPIKE.csv"))
counts_4hr <- read_csv(here("output", "allCountsNorm_all_4hr_SPIKE.csv"))
counts_24hr <- read_csv(here("output", "allCountsNorm_all_24hr_SPIKE.csv"))

cat("Loaded normalized counts:\n")
cat("1hr:", ncol(counts_1hr) - 1, "samples,", nrow(counts_1hr), "genes\n")
cat("4hr:", ncol(counts_4hr) - 1, "samples,", nrow(counts_4hr), "genes\n")
cat("24hr:", ncol(counts_24hr) - 1, "samples,", nrow(counts_24hr), "genes\n")
```

```{r load_gene_annotation}
# Load gene annotation for symbol lookup
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, desc(type == "protein_coding")) %>%
  group_by(gene_id) %>%
  slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "unique gene IDs\n")
```

## Helper Functions

```{r parse_sample_names_function}
#' Parse sample column names into metadata components
#'
#' @param col_names Character vector of column names (excluding gene_id)
#' @return Data frame with parsed sample metadata
parse_sample_names <- function(col_names) {
  # Pattern: {condition}_{treatment}_{timepoint}_{replicate}_{library}_{sampleID}
  # Example: hypoxia_sictrl_1_a_ribo_NM2025_0076


  parsed <- tibble(sample_name = col_names) %>%
    separate(
      sample_name,
      into = c("condition", "treatment", "timepoint", "replicate", "library", "sample_id"),
      sep = "_",
      remove = FALSE,
      extra = "merge"  # Handle sample_id with underscore (NM2025_XXXX)
    ) %>%
    mutate(
      timepoint = paste0(timepoint, "hr"),
      condition = factor(condition, levels = c("normoxia", "hypoxia")),
      treatment = factor(treatment, levels = c("sictrl", "si3e", "si3d")),
      library = factor(library, levels = c("rna", "ribo"))
    )

  return(parsed)
}
```

## Main Visualization Function

```{r visualize_gene_function}
#' Visualize individual gene expression across conditions
#'
#' Creates four plots:
#' 1. RNA & Ribo combined - bars side by side for each condition
#' 2. TE separate - translation efficiency
#' 3. RNA only - RNA counts per treatment
#' 4. Ribo only - Ribo counts per treatment
#'
#' @param gene_input Gene ID (ENSG...) or gene symbol
#' @param counts_1hr Data frame of 1hr normalized counts
#' @param counts_4hr Data frame of 4hr normalized counts
#' @param counts_24hr Data frame of 24hr normalized counts
#' @param gene_symbols Data frame mapping gene_id to symbol
#' @param log_transform Logical, whether to log2(x+1) transform counts (default: TRUE)
#' @return A list with four ggplot objects: $rna_ribo, $te, $rna, and $ribo
visualize_gene <- function(gene_input,
                           counts_1hr,
                           counts_4hr,
                           counts_24hr,
                           gene_symbols,
                           log_transform = TRUE) {

  # Resolve gene_id from input (could be symbol or ENSG ID)
  if (grepl("^ENSG", gene_input)) {
    gene_id <- gene_input
    gene_symbol <- gene_symbols %>%
      filter(gene_id == !!gene_input) %>%
      pull(symbol) %>%
      dplyr::first()
    if (is.na(gene_symbol)) gene_symbol <- gene_input
  } else {
    # Input is a symbol - find corresponding gene_id
    gene_symbol <- gene_input
    gene_id <- gene_symbols %>%
      filter(symbol == !!gene_input) %>%
      pull(gene_id) %>%
      dplyr::first()

    if (is.na(gene_id)) {
      stop("Gene symbol '", gene_input, "' not found in annotation.")
    }
  }

  cat("Visualizing gene:", gene_symbol, "(", gene_id, ")\n")

  # Extract gene data from each timepoint
  extract_gene_data <- function(counts_df, timepoint_label) {
    gene_row <- counts_df %>% filter(gene_id == !!gene_id)

    if (nrow(gene_row) == 0) {
      warning("Gene ", gene_id, " not found in ", timepoint_label, " data.")
      return(NULL)
    }

    # Pivot to long format
    gene_long <- gene_row %>%
      pivot_longer(
        cols = -gene_id,
        names_to = "sample_name",
        values_to = "count"
      )

    # Parse sample metadata from column names
    sample_meta <- parse_sample_names(gene_long$sample_name)

    gene_long <- gene_long %>%
      left_join(sample_meta, by = "sample_name") %>%
      mutate(timepoint = timepoint_label)

    return(gene_long)
  }

  # Combine all timepoints
  gene_data <- bind_rows(
    extract_gene_data(counts_1hr, "1hr"),
    extract_gene_data(counts_4hr, "4hr"),
    extract_gene_data(counts_24hr, "24hr")
  ) %>%
    mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

  # Separate RNA and Ribo data
  rna_data <- gene_data %>%
    filter(library == "rna") %>%
    dplyr::select(condition, treatment, timepoint, replicate, rna_count = count)

  ribo_data <- gene_data %>%
    filter(library == "ribo") %>%
    dplyr::select(condition, treatment, timepoint, replicate, ribo_count = count)

  # Merge RNA and Ribo by sample identifiers to calculate TE
  te_data <- inner_join(
    rna_data,
    ribo_data,
    by = c("condition", "treatment", "timepoint", "replicate")
  ) %>%
    mutate(
      # TE = log2(Ribo + 1) - log2(RNA + 1)
      te_value = log2(ribo_count + 1) - log2(rna_count + 1)
    )

  # Prepare combined RNA/Ribo data for plotting
  if (log_transform) {
    rna_ribo_data <- bind_rows(
      rna_data %>% mutate(value = log2(rna_count + 1), library = "RNA"),
      ribo_data %>% mutate(value = log2(ribo_count + 1), library = "Ribo")
    )
    y_label <- "Spike-in Normalized Counts (log2)"
  } else {
    rna_ribo_data <- bind_rows(
      rna_data %>% mutate(value = rna_count, library = "RNA"),
      ribo_data %>% mutate(value = ribo_count, library = "Ribo")
    )
    y_label <- "Spike-in Normalized Counts"
  }

  rna_ribo_data <- rna_ribo_data %>%
    mutate(
      library = factor(library, levels = c("RNA", "Ribo")),
      treatment = factor(treatment, levels = c("sictrl", "si3d", "si3e")),
      # Create combined facet variable: normoxia timepoints first, then hypoxia
      facet_label = paste0(condition, "\n", timepoint),
      facet_label = factor(facet_label, levels = c(
        "normoxia\n1hr", "normoxia\n4hr", "normoxia\n24hr",
        "hypoxia\n1hr", "hypoxia\n4hr", "hypoxia\n24hr"
      )),
      # Create treatment-library interaction for coloring
      treat_lib = interaction(treatment, library, sep = "-")
    )

  # Define color palette: lighter for RNA, darker for Ribo
  # sictrl = grey, si3d = orange, si3e = blue
  rna_ribo_colors <- c(
    "sictrl-RNA" = "grey70",
    "sictrl-Ribo" = "grey40",
    "si3d-RNA" = "#F5C566",      # lighter orange
    "si3d-Ribo" = "#E69F00",     # Okabe-Ito orange
    "si3e-RNA" = "#89CFF0",      # lighter blue
    "si3e-Ribo" = "#56B4E9"      # Okabe-Ito light blue
  )

  # Summary statistics for RNA/Ribo
  rna_ribo_summary <- rna_ribo_data %>%
    group_by(treatment, library, facet_label, treat_lib) %>%
    summarize(
      mean_value = mean(value, na.rm = TRUE),
      se_value = sd(value, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    )

  # Calculate y-axis minimum (bottom of graph) for RNA/Ribo
  y_min_rna_ribo <- min(rna_ribo_data$value, na.rm = TRUE) - 0.1

  # Create RNA/Ribo combined plot
  p_rna_ribo <- ggplot() +
    # Bars extending from bottom of graph to mean
    geom_rect(
      data = rna_ribo_summary,
      aes(xmin = as.numeric(treatment) + ifelse(library == "RNA", -0.35, 0.02),
          xmax = as.numeric(treatment) + ifelse(library == "RNA", -0.02, 0.35),
          ymin = y_min_rna_ribo,
          ymax = mean_value,
          fill = treat_lib),
      alpha = 0.7
    ) +
    # Error bars (pointing up from mean)
    geom_errorbar(
      data = rna_ribo_summary,
      aes(x = as.numeric(treatment) + ifelse(library == "RNA", -0.185, 0.185),
          ymin = mean_value,
          ymax = mean_value + se_value),
      width = 0.15
    ) +
    # Individual replicate points
    geom_point(
      data = rna_ribo_data,
      aes(x = as.numeric(treatment) + ifelse(library == "RNA", -0.185, 0.185),
          y = value, fill = treat_lib, shape = library),
      position = position_jitter(width = 0.08),
      size = 3,
      color = "black",
      stroke = 0.5
    ) +
    scale_shape_manual(values = c("RNA" = 22, "Ribo" = 21), name = "") +
    # Facet by condition and timepoint
    facet_wrap(~facet_label, nrow = 1) +
    # Fix x-axis labels
    scale_x_continuous(
      breaks = 1:3,
      labels = c("sictrl", "si3d", "si3e")
    ) +
    # Styling
    scale_fill_manual(
      values = rna_ribo_colors,
      name = "",
      labels = c(
        "sictrl-RNA" = "sictrl RNA", "sictrl-Ribo" = "sictrl Ribo",
        "si3d-RNA" = "si3d RNA", "si3d-Ribo" = "si3d Ribo",
        "si3e-RNA" = "si3e RNA", "si3e-Ribo" = "si3e Ribo"
      )
    ) +
    labs(
      title = paste0(gene_symbol, " - RNA & Ribo"),
      subtitle = gene_id,
      x = "Treatment",
      y = y_label
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "gray90")
    ) +
    guides(fill = guide_legend(nrow = 1))

  # TE plot - faceted by condition and timepoint, colored by treatment
  # Okabe-Ito colors: grey for sictrl, orange for si3d, light blue for si3e
  te_colors <- c("sictrl" = "grey50", "si3d" = "#E69F00", "si3e" = "#56B4E9")

  te_plot_data <- te_data %>%
    dplyr::select(condition, treatment, timepoint, replicate, value = te_value) %>%
    mutate(
      treatment = factor(treatment, levels = c("sictrl", "si3d", "si3e")),
      # Create combined facet variable: normoxia timepoints first, then hypoxia
      facet_label = paste0(condition, "\n", timepoint),
      facet_label = factor(facet_label, levels = c(
        "normoxia\n1hr", "normoxia\n4hr", "normoxia\n24hr",
        "hypoxia\n1hr", "hypoxia\n4hr", "hypoxia\n24hr"
      ))
    )

  # Summary stats for bars
  te_summary <- te_plot_data %>%
    group_by(treatment, facet_label) %>%
    summarize(
      mean_value = mean(value, na.rm = TRUE),
      se_value = sd(value, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    )

  # Calculate y-axis minimum (bottom of graph)
  y_min <- min(te_plot_data$value, na.rm = TRUE) - 0.1

  p_te <- ggplot() +
    # Bars extending from bottom of graph to mean
    geom_rect(
      data = te_summary,
      aes(xmin = as.numeric(treatment) - 0.35,
          xmax = as.numeric(treatment) + 0.35,
          ymin = y_min,
          ymax = mean_value,
          fill = treatment),
      alpha = 0.7
    ) +
    # Error bars (pointing up from mean)
    geom_errorbar(
      data = te_summary,
      aes(x = treatment,
          ymin = mean_value,
          ymax = mean_value + se_value),
      width = 0.2
    ) +
    # Individual replicate points
    geom_point(
      data = te_plot_data,
      aes(x = treatment, y = value, fill = treatment),
      position = position_jitter(width = 0.2),
      shape = 21,
      size = 3,
      color = "black",
      stroke = 0.5
    ) +
    facet_wrap(~facet_label, nrow = 1) +
    scale_fill_manual(values = te_colors, name = "Treatment") +
    labs(
      title = paste0(gene_symbol, " - Translation Efficiency"),
      subtitle = gene_id,
      x = "Treatment",
      y = "TE (log2 Ribo/RNA)"
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "gray90")
    )

  # RNA-only plot (TE-plot style: single centered bar per treatment)
  rna_only_data <- rna_ribo_data %>% filter(library == "RNA")
  rna_only_summary <- rna_ribo_summary %>% filter(library == "RNA")

  rna_colors <- c("sictrl" = "grey70", "si3d" = "#F5C566", "si3e" = "#89CFF0")

  y_min_rna <- min(rna_only_data$value, na.rm = TRUE) - 0.1

  p_rna <- ggplot() +
    geom_rect(
      data = rna_only_summary,
      aes(xmin = as.numeric(treatment) - 0.35,
          xmax = as.numeric(treatment) + 0.35,
          ymin = y_min_rna,
          ymax = mean_value,
          fill = treatment),
      alpha = 0.7
    ) +
    geom_errorbar(
      data = rna_only_summary,
      aes(x = treatment,
          ymin = mean_value,
          ymax = mean_value + se_value),
      width = 0.2
    ) +
    geom_point(
      data = rna_only_data,
      aes(x = treatment, y = value, fill = treatment),
      position = position_jitter(width = 0.2),
      shape = 22,
      size = 3,
      color = "black",
      stroke = 0.5
    ) +
    facet_wrap(~facet_label, nrow = 1) +
    scale_fill_manual(values = rna_colors, name = "Treatment") +
    labs(
      title = paste0(gene_symbol, " - RNA"),
      subtitle = gene_id,
      x = "Treatment",
      y = y_label
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "gray90")
    )

  # Ribo-only plot (TE-plot style: single centered bar per treatment)
  ribo_only_data <- rna_ribo_data %>% filter(library == "Ribo")
  ribo_only_summary <- rna_ribo_summary %>% filter(library == "Ribo")

  ribo_colors <- c("sictrl" = "grey40", "si3d" = "#E69F00", "si3e" = "#56B4E9")

  y_min_ribo <- min(ribo_only_data$value, na.rm = TRUE) - 0.1

  p_ribo <- ggplot() +
    geom_rect(
      data = ribo_only_summary,
      aes(xmin = as.numeric(treatment) - 0.35,
          xmax = as.numeric(treatment) + 0.35,
          ymin = y_min_ribo,
          ymax = mean_value,
          fill = treatment),
      alpha = 0.7
    ) +
    geom_errorbar(
      data = ribo_only_summary,
      aes(x = treatment,
          ymin = mean_value,
          ymax = mean_value + se_value),
      width = 0.2
    ) +
    geom_point(
      data = ribo_only_data,
      aes(x = treatment, y = value, fill = treatment),
      position = position_jitter(width = 0.2),
      shape = 21,
      size = 3,
      color = "black",
      stroke = 0.5
    ) +
    facet_wrap(~facet_label, nrow = 1) +
    scale_fill_manual(values = ribo_colors, name = "Treatment") +
    labs(
      title = paste0(gene_symbol, " - Ribo"),
      subtitle = gene_id,
      x = "Treatment",
      y = y_label
    ) +
    theme_cowplot() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.background = element_rect(fill = "gray90")
    )

  return(list(rna_ribo = p_rna_ribo, te = p_te, rna = p_rna, ribo = p_ribo))
}
```

## Batch Visualization Function

```{r visualize_multiple_genes_function}
#' Visualize multiple genes and save to PDF
#'
#' @param gene_list Vector of gene IDs or symbols
#' @param output_prefix Prefix for output PDF filename
#' @param ... Additional arguments passed to visualize_gene()
visualize_multiple_genes <- function(gene_list,
                                     output_prefix = "gene_expression",
                                     counts_1hr = counts_1hr,
                                     counts_4hr = counts_4hr,
                                     counts_24hr = counts_24hr,
                                     gene_symbols = gene_symbols,
                                     ...) {

  # Create output directory if needed
  if (!dir.exists(here("plots"))) {
    dir.create(here("plots"))
  }

  all_plots <- list()

  for (gene in gene_list) {
    cat("\n--- Processing:", gene, "---\n")

    tryCatch({
      plots <- visualize_gene(
        gene_input = gene,
        counts_1hr = counts_1hr,
        counts_4hr = counts_4hr,
        counts_24hr = counts_24hr,
        gene_symbols = gene_symbols,
        ...
      )

      all_plots[[gene]] <- plots

      # Combine RNA/Ribo and TE plots vertically
      combined <- plot_grid(
        plots$rna_ribo,
        plots$te,
        ncol = 1,
        align = "v",
        axis = "lr",
        rel_heights = c(1, 1)
      )

      # Save combined plot
      ggsave(
        here("plots", paste0(output_prefix, "_", gene, ".pdf")),
        plot = combined,
        width = 10,
        height = 10
      )

    }, error = function(e) {
      warning("Failed to plot gene ", gene, ": ", e$message)
    })
  }

  return(all_plots)
}
```

## Example: Visualize a Single Gene

```{r example_single_gene, fig.width=10, fig.height=10}
# Specify gene of interest (can use gene symbol or ENSG ID)
gene_of_interest <- "HERPUD1"  # Classic hypoxia-responsive gene

# Generate visualization
plots <- visualize_gene(
  gene_input = gene_of_interest,
  counts_1hr = counts_1hr,
  counts_4hr = counts_4hr,
  counts_24hr = counts_24hr,
  gene_symbols = gene_symbols,
  log_transform = TRUE
)

# Display RNA/Ribo plot
print(plots$rna_ribo)

ggsave(
  here("plots", paste0("gene_expression_rna_ribo_", gene_of_interest, ".pdf")),
  plot = plots$rna_ribo,
  width = 10,
  height = 5
)

# Display TE plot
print(plots$te)

# Save plots to file
combined <- plot_grid(
  plots$rna_ribo,
  plots$te,
  ncol = 1,
  align = "v",
  axis = "lr"
)

ggsave(
  here("plots", paste0("gene_expression_", gene_of_interest, ".pdf")),
  plot = combined,
  width = 12,
  height = 8
)
```

```{r example_separate_rna_ribo, fig.width=10, fig.height=5}
# Display separate RNA and Ribo plots
print(plots$rna)
print(plots$ribo)
```

```{r save_example_plot, eval=FALSE}
# Save plots to file
combined <- plot_grid(
  plots$rna_ribo,
  plots$te,
  ncol = 1,
  align = "v",
  axis = "lr"
)

ggsave(
  here("plots", paste0("gene_expression_", gene_of_interest, ".pdf")),
  plot = combined,
  width = 10,
  height = 10
)
```

## Example: Visualize Multiple Genes

```{r example_multiple_genes, eval=FALSE}
# List of genes of interest
genes_of_interest <- c(
  "VEGFA",    # Classic hypoxia gene
  "HIF1A",    # Hypoxia master regulator
  "LDHA",     # Glycolysis gene
  "BNIP3",    # HIF target
  "EIF3D",    # Translation factor of interest
  "EIF3E"     # Translation factor of interest
)

# Generate all plots
all_plots <- visualize_multiple_genes(
  gene_list = genes_of_interest,
  output_prefix = "hypoxia_gene",
  counts_1hr = counts_1hr,
  counts_4hr = counts_4hr,
  counts_24hr = counts_24hr,
  gene_symbols = gene_symbols,
  log_transform = TRUE
)
```

## Session Info

```{r session_info}
sessionInfo()
```