---
title: "EIF3 IP Mass Spec: HypoDMSO vs NormDMSO"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Label-free quantitative mass spectrometry from an EIF3 complex immunoprecipitation.
Comparing hypoxia + DMSO vehicle (HypoDMSO) vs. normoxia + DMSO vehicle (NormDMSO).
n=1 per condition — fold change only, no p-values.

# Load Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(ggrepel)
library(pheatmap)
library(openxlsx)

# Create output directory if it doesn't exist
dir.create("output", showWarnings = FALSE)
```

# 1. Load Data

The data is an Excel export from Scaffold. We use the `combined_protein` sheet,
which contains protein-level quantification with Protein Probability scores.

```{r load-data}
# Path to the mass spec results file (relative to project root)
ms_file <- here::here("accessories/mass_spec/TTF1_20250924_JLi45922530_results_hydrox.xlsx")

# Load the combined_protein sheet — header is row 1, no metadata to skip
raw_df <- read_excel(ms_file, sheet = "combined_protein")

# Verify the load
cat("Dimensions:", nrow(raw_df), "rows x", ncol(raw_df), "columns\n\n")
cat("Column names:\n")
print(colnames(raw_df))

cat("\nFirst few rows:\n")
print(head(raw_df[, 1:10]))
```

# 2. Identify Intensity Columns

Use `grep()` to find intensity columns by pattern rather than hardcoding names,
excluding any columns belonging to the 8430 sample, TechMix, Hypo8430, or Norm8430.

```{r identify-columns}
all_cols <- colnames(raw_df)

# Patterns to exclude (non-DMSO samples)
exclude_pattern <- "8430|TechMix|Hypo8430|Norm8430"

# Find HypoDMSO intensity column
hypo_candidates <- grep("HypoDMSO", all_cols, value = TRUE)
hypo_candidates <- hypo_candidates[grepl("Intensity", hypo_candidates)]
hypo_candidates <- hypo_candidates[!grepl(exclude_pattern, hypo_candidates)]
hypo_col <- hypo_candidates[1]

# Find NormDMSO intensity column
norm_candidates <- grep("NormDMSO", all_cols, value = TRUE)
norm_candidates <- norm_candidates[grepl("Intensity", norm_candidates)]
norm_candidates <- norm_candidates[!grepl(exclude_pattern, norm_candidates)]
norm_col <- norm_candidates[1]

cat("HypoDMSO intensity column:", hypo_col, "\n")
cat("NormDMSO intensity column:", norm_col, "\n")
```

# 3. Filter & Select

Keep proteins with Protein Probability >= 0.95 and select relevant columns.

```{r filter-select}
# Filter and select key columns
ms_df <- raw_df %>%
  # Filter by protein probability threshold
  filter(`Protein Probability` >= 0.95) %>%
  # Select key columns, renaming intensity columns for clarity
  select(
    Gene,
    Protein,
    Description,
    hypo_intensity = all_of(hypo_col),
    norm_intensity = all_of(norm_col)
  ) %>%
  # Clean up gene names (strip whitespace)
  mutate(gene_name = str_trim(Gene))

cat("Proteins passing Protein Probability >= 0.95:", nrow(ms_df), "\n")
print(head(ms_df))
```

# 4. Imputation & Log2 Transformation

Replace 0 and NA intensity values with half the minimum non-zero intensity
across both conditions, then log2-transform.

```{r impute-log2}
# Compute half the minimum non-zero value across both intensity columns
all_intensities <- c(ms_df$hypo_intensity, ms_df$norm_intensity)
min_nonzero <- min(all_intensities[all_intensities > 0], na.rm = TRUE)
impute_val <- min_nonzero / 2

cat("Minimum non-zero intensity:", min_nonzero, "\n")
cat("Imputation value (half-min):", impute_val, "\n")
cat("Proteins with 0/NA in HypoDMSO:", sum(is.na(ms_df$hypo_intensity) | ms_df$hypo_intensity == 0), "\n")
cat("Proteins with 0/NA in NormDMSO:", sum(is.na(ms_df$norm_intensity) | ms_df$norm_intensity == 0), "\n")

# Impute and log2-transform
ms_df <- ms_df %>%
  mutate(
    hypo_imputed = ifelse(is.na(hypo_intensity) | hypo_intensity == 0, impute_val, hypo_intensity),
    norm_imputed = ifelse(is.na(norm_intensity) | norm_intensity == 0, impute_val, norm_intensity),
    hypo_log2    = log2(hypo_imputed),
    norm_log2    = log2(norm_imputed)
  )
```

# 5. Fold Change & Direction

Calculate log2 fold change (HypoDMSO / NormDMSO). Positive = enriched in hypoxia.

```{r fold-change}
ms_df <- ms_df %>%
  mutate(
    log2FC = hypo_log2 - norm_log2,
    mean_log2 = (hypo_log2 + norm_log2) / 2,  # abundance proxy for plotting
    direction = case_when(
      log2FC >  0.5 ~ "Up in Hypoxia",
      log2FC < -0.5 ~ "Down in Hypoxia",
      TRUE          ~ "Unchanged"
    )
  )

# Summary
table(ms_df$direction)
```

# 6. Export Results

```{r export}
# --- CSV export ---
write_csv(ms_df, "output/eif3_HypoDMSO_vs_NormDMSO_results.csv")
cat("Saved: output/eif3_HypoDMSO_vs_NormDMSO_results.csv\n")

# --- Excel export (two sheets) ---
wb <- createWorkbook()

# Sheet 1: All proteins sorted by log2FC descending
addWorksheet(wb, "All Proteins")
writeData(wb, "All Proteins", ms_df %>% arrange(desc(log2FC)))

# Sheet 2: EIF3 subunits only
eif3_df <- ms_df %>% filter(str_starts(gene_name, "EIF3"))
addWorksheet(wb, "EIF3 Subunits")
writeData(wb, "EIF3 Subunits", eif3_df %>% arrange(desc(log2FC)))

saveWorkbook(wb, "output/eif3_HypoDMSO_vs_NormDMSO_results.xlsx", overwrite = TRUE)
cat("Saved: output/eif3_HypoDMSO_vs_NormDMSO_results.xlsx\n")
```

# 7. Visualizations

## A) Fold Change Scatter Plot (volcano-style)

```{r scatter-plot, fig.width=8, fig.height=6}
# Color scheme
dir_colors <- c(
  "Up in Hypoxia"   = "#D62728",
  "Down in Hypoxia" = "#1F77B4",
  "Unchanged"       = "gray70"
)

# Genes to label (all EIF3 subunits)
label_df <- ms_df %>% filter(str_starts(gene_name, "EIF3"))

p_scatter <- ggplot(ms_df, aes(x = log2FC, y = mean_log2, color = direction)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text_repel(
    data = label_df,
    aes(label = gene_name),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.4,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "gray40", linewidth = 0.5) +
  scale_color_manual(values = dir_colors, name = NULL) +
  labs(
    title = "HypoDMSO vs NormDMSO | EIF3 complex IP",
    x     = "log2 Fold Change (Hypo / Norm)",
    y     = "Mean log2 Intensity"
  ) +
  theme_classic() +
  theme(legend.position = "top")

print(p_scatter)

# Save
ggsave("output/scatter_HypoDMSO_vs_NormDMSO.pdf", p_scatter, width = 8, height = 6)
ggsave("output/scatter_HypoDMSO_vs_NormDMSO.png", p_scatter, width = 8, height = 6, dpi = 300)
cat("Saved scatter plot\n")
```

## B) Heatmap (EIF3 subunits only)

```{r heatmap, fig.width=5, fig.height=7}
# Filter to EIF3 subunits
eif3_heat <- ms_df %>%
  filter(str_starts(gene_name, "EIF3")) %>%
  arrange(gene_name)

# Build intensity matrix (rows = genes, cols = conditions)
heat_mat <- eif3_heat %>%
  select(hypo_log2, norm_log2) %>%
  as.matrix()
rownames(heat_mat) <- eif3_heat$gene_name
colnames(heat_mat) <- c("HypoDMSO", "NormDMSO")

# Z-score scale per row
heat_mat_scaled <- t(scale(t(heat_mat)))

# Display numbers: log2FC (rounded to 1 decimal) in each cell
display_mat <- matrix(
  round(eif3_heat$log2FC, 1),
  nrow = nrow(heat_mat),
  ncol = 2,
  dimnames = list(rownames(heat_mat), colnames(heat_mat))
)

# Draw heatmap (saved via pheatmap filename argument)
pheatmap(
  heat_mat_scaled,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  color           = colorRampPalette(c("blue", "white", "red"))(50),
  display_numbers = display_mat,
  fontsize         = 10,
  main            = "EIF3 subunits — log2 intensity (z-scored)\n(numbers = log2FC)",
  filename        = "output/heatmap_eif3_subunits.pdf",
  width           = 5,
  height          = 7
)

pheatmap(
  heat_mat_scaled,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  color           = colorRampPalette(c("blue", "white", "red"))(50),
  display_numbers = display_mat,
  fontsize         = 10,
  main            = "EIF3 subunits — log2 intensity (z-scored)\n(numbers = log2FC)",
  filename        = "output/heatmap_eif3_subunits.png",
  width           = 5,
  height          = 7
)

cat("Saved heatmap\n")
dev.off()
# Also render inline
pheatmap(
  heat_mat_scaled,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  color           = colorRampPalette(c("blue", "white", "red"))(50),
  display_numbers = display_mat,
  fontsize         = 10,
  main            = "EIF3 subunits — log2 intensity (z-scored)\n(numbers = log2FC)"
)
```

## C) Top 20 Horizontal Bar Chart

```{r bar-chart, fig.width=7, fig.height=7}
# Select top 20 by absolute log2FC
top20 <- ms_df %>%
  arrange(desc(abs(log2FC))) %>%
  slice_head(n = 20) %>%
  mutate(gene_name = fct_reorder(gene_name, log2FC))

p_bar <- ggplot(top20, aes(x = log2FC, y = gene_name, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "gray20") +
  scale_fill_manual(values = dir_colors, name = NULL) +
  labs(
    title = "Top 20 proteins by fold change | HypoDMSO vs NormDMSO",
    x     = "log2 Fold Change (Hypo / Norm)",
    y     = NULL
  ) +
  theme_classic() +
  theme(legend.position = "top")

print(p_bar)

# Save
ggsave("output/barplot_top20_HypoDMSO_vs_NormDMSO.pdf", p_bar, width = 7, height = 7)
ggsave("output/barplot_top20_HypoDMSO_vs_NormDMSO.png", p_bar, width = 7, height = 7, dpi = 300)
cat("Saved bar chart\n")
```

# 8. Console Summary

```{r summary}
cat("=== ANALYSIS SUMMARY ===\n\n")
cat("Total proteins passing filter (Protein Probability >= 0.95):", nrow(ms_df), "\n\n")

cat("Direction breakdown:\n")
dir_counts <- table(ms_df$direction)
print(dir_counts)
cat("\n")

cat("EIF3 subunits detected (sorted by log2FC descending):\n")
ms_df %>%
  filter(str_starts(gene_name, "EIF3")) %>%
  select(gene_name, log2FC, direction, hypo_log2, norm_log2) %>%
  arrange(desc(log2FC)) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  print(n = Inf)
```
