---
title: "Normoxia Timepoint TE Similarity"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

# Overview

This analysis examines how similar TE (translational efficiency) changes are across normoxia timepoints (1hr, 4hr, 24hr) for si3d vs siCtrl, and how the alltimepoints regression set relates to each individual timepoint. This informs interpretation of the random forest model behavior when using the alltimepoints positive class.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(cowplot)
library(pheatmap)
library(ggvenn)
library(DT)
```

# Load Data

```{r load-data}
norm_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"))
norm_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv"))
norm_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv"))
norm_all <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv"))

cat("Gene counts:\n")
cat("  1hr:", nrow(norm_1hr), "\n")
cat("  4hr:", nrow(norm_4hr), "\n")
cat("  24hr:", nrow(norm_24hr), "\n")
cat("  alltimepoints:", nrow(norm_all), "\n")
```

```{r join-data}
# Join all four files into a wide data frame on gene_id
wide <- norm_1hr %>%
  select(gene_id, symbol, te_lfc_1hr = te_lfc, te_padj_1hr = te_padj) %>%
  inner_join(
    norm_4hr %>% select(gene_id, te_lfc_4hr = te_lfc, te_padj_4hr = te_padj),
    by = "gene_id"
  ) %>%
  inner_join(
    norm_24hr %>% select(gene_id, te_lfc_24hr = te_lfc, te_padj_24hr = te_padj),
    by = "gene_id"
  ) %>%
  inner_join(
    norm_all %>% select(gene_id, te_lfc_all = te_lfc, te_padj_all = te_padj),
    by = "gene_id"
  )

cat("Genes after inner join:", nrow(wide), "\n")
```

# 1. Pairwise TE LFC Scatter Plots

```{r sig-status}
# Define significance: te_padj < 0.05 & |te_lfc| > 0.5
sig_threshold_lfc <- 0.5
sig_threshold_padj <- 0.05

wide <- wide %>%
  mutate(
    sig_1hr = te_padj_1hr < sig_threshold_padj & abs(te_lfc_1hr) > sig_threshold_lfc,
    sig_4hr = te_padj_4hr < sig_threshold_padj & abs(te_lfc_4hr) > sig_threshold_lfc,
    sig_24hr = te_padj_24hr < sig_threshold_padj & abs(te_lfc_24hr) > sig_threshold_lfc,
    sig_all = te_padj_all < sig_threshold_padj & abs(te_lfc_all) > sig_threshold_lfc
  )
```

```{r scatter-function}
# Helper: make one scatter panel with Pearson r and Spearman rho annotations
make_scatter <- function(df, x_col, y_col, sig_x_col, sig_y_col,
                         x_label, y_label) {
  # Significance coloring: both, either, neither
  df <- df %>%
    mutate(
      sig_status = case_when(
        .data[[sig_x_col]] & .data[[sig_y_col]] ~ "Both",
        .data[[sig_x_col]] | .data[[sig_y_col]] ~ "Either",
        TRUE ~ "Neither"
      ),
      sig_status = factor(sig_status, levels = c("Neither", "Either", "Both"))
    )

  x_vals <- df[[x_col]]
  y_vals <- df[[y_col]]
  pearson_r <- cor(x_vals, y_vals, use = "complete.obs", method = "pearson")
  spearman_rho <- cor(x_vals, y_vals, use = "complete.obs", method = "spearman")

  ann_label <- sprintf("r = %.2f\nρ = %.2f", pearson_r, spearman_rho)

  ggplot(df, aes(x = .data[[x_col]], y = .data[[y_col]], color = sig_status)) +
    geom_point(size = 0.6, alpha = 0.5) +
    scale_color_manual(
      values = c("Neither" = "grey70", "Either" = "#F4A261", "Both" = "#E63946"),
      name = "Significant"
    ) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
    annotate("text", x = -Inf, y = Inf, label = ann_label,
             hjust = -0.1, vjust = 1.2, size = 3) +
    labs(x = x_label, y = y_label) +
    theme_classic() +
    theme(legend.position = "none",
          axis.title = element_text(size = 9))
}
```

```{r scatter-plots, fig.width=12, fig.height=8}
p_1hr_4hr <- make_scatter(wide, "te_lfc_1hr", "te_lfc_4hr",
                           "sig_1hr", "sig_4hr",
                           "TE LFC (1hr)", "TE LFC (4hr)")

p_1hr_24hr <- make_scatter(wide, "te_lfc_1hr", "te_lfc_24hr",
                            "sig_1hr", "sig_24hr",
                            "TE LFC (1hr)", "TE LFC (24hr)")

p_4hr_24hr <- make_scatter(wide, "te_lfc_4hr", "te_lfc_24hr",
                            "sig_4hr", "sig_24hr",
                            "TE LFC (4hr)", "TE LFC (24hr)")

p_1hr_all <- make_scatter(wide, "te_lfc_1hr", "te_lfc_all",
                           "sig_1hr", "sig_all",
                           "TE LFC (1hr)", "TE LFC (alltimepoints)")

p_4hr_all <- make_scatter(wide, "te_lfc_4hr", "te_lfc_all",
                           "sig_4hr", "sig_all",
                           "TE LFC (4hr)", "TE LFC (alltimepoints)")

p_24hr_all <- make_scatter(wide, "te_lfc_24hr", "te_lfc_all",
                            "sig_24hr", "sig_all",
                            "TE LFC (24hr)", "TE LFC (alltimepoints)")

# Shared legend for sig status
legend_plot <- make_scatter(wide, "te_lfc_1hr", "te_lfc_4hr",
                             "sig_1hr", "sig_4hr", "", "") +
  theme(legend.position = "right")
shared_legend <- get_legend(legend_plot)

scatter_grid <- plot_grid(
  p_1hr_4hr, p_1hr_24hr, p_4hr_24hr,
  p_1hr_all, p_4hr_all, p_24hr_all,
  ncol = 3
)

plot_grid(scatter_grid, shared_legend,
          rel_widths = c(1, 0.12),
          labels = c("Pairwise TE LFC (si3d vs siCtrl, normoxia)", ""),
          label_size = 11)
```

# 2. Correlation Matrix Heatmap

```{r correlation-matrix, fig.width=5, fig.height=4.5}
te_mat <- wide %>%
  select(te_lfc_1hr, te_lfc_4hr, te_lfc_24hr, te_lfc_all) %>%
  as.matrix()

cor_mat <- cor(te_mat, method = "spearman", use = "complete.obs")

colnames(cor_mat) <- rownames(cor_mat) <- c("1hr", "4hr", "24hr", "alltimepoints")

pheatmap(
  cor_mat,
  display_numbers = TRUE,
  number_format = "%.3f",
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(50),
  breaks = seq(0, 1, length.out = 51),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Spearman correlation of TE LFC across normoxia timepoints\n(si3d vs siCtrl)",
  fontsize_number = 10
)
```

# 3. Venn Diagram — Significant TE Upregulation

```{r venn-sig-up, fig.width=7, fig.height=6}
# Significant upregulation: te_lfc > 0.5 & te_padj < 0.05
sig_up_1hr <- wide %>% filter(te_lfc_1hr > 0.5 & te_padj_1hr < 0.05) %>% pull(gene_id)
sig_up_4hr <- wide %>% filter(te_lfc_4hr > 0.5 & te_padj_4hr < 0.05) %>% pull(gene_id)
sig_up_24hr <- wide %>% filter(te_lfc_24hr > 0.5 & te_padj_24hr < 0.05) %>% pull(gene_id)

cat("Significant TE upregulation (te_lfc > 0.5, te_padj < 0.05):\n")
cat("  1hr:", length(sig_up_1hr), "\n")
cat("  4hr:", length(sig_up_4hr), "\n")
cat("  24hr:", length(sig_up_24hr), "\n")
cat("  Shared across all three:", length(Reduce(intersect, list(sig_up_1hr, sig_up_4hr, sig_up_24hr))), "\n")
cat("  Unique to 1hr:", length(setdiff(sig_up_1hr, union(sig_up_4hr, sig_up_24hr))), "\n")
cat("  Unique to 4hr:", length(setdiff(sig_up_4hr, union(sig_up_1hr, sig_up_24hr))), "\n")
cat("  Unique to 24hr:", length(setdiff(sig_up_24hr, union(sig_up_1hr, sig_up_4hr))), "\n")
cat("  1hr ∩ 4hr (not 24hr):", length(setdiff(intersect(sig_up_1hr, sig_up_4hr), sig_up_24hr)), "\n")
cat("  1hr ∩ 24hr (not 4hr):", length(setdiff(intersect(sig_up_1hr, sig_up_24hr), sig_up_4hr)), "\n")
cat("  4hr ∩ 24hr (not 1hr):", length(setdiff(intersect(sig_up_4hr, sig_up_24hr), sig_up_1hr)), "\n")

venn_list_up <- list(
  "1hr" = sig_up_1hr,
  "4hr" = sig_up_4hr,
  "24hr" = sig_up_24hr
)

ggvenn(
  venn_list_up,
  fill_color = c("#457B9D", "#A8DADC", "#E63946"),
  stroke_size = 0.5,
  set_name_size = 5,
  text_size = 4
) +
  labs(title = "Significant TE upregulation by timepoint\n(si3d vs siCtrl, normoxia; te_lfc > 0.5, te_padj < 0.05)")
```

# 4. Jaccard Similarity Table

```{r jaccard-table}
jaccard <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

# Upregulated gene sets
sig_sets <- list(
  "1hr" = sig_up_1hr,
  "4hr" = sig_up_4hr,
  "24hr" = sig_up_24hr,
  "alltimepoints" = wide %>% filter(te_lfc_all > 0.5 & te_padj_all < 0.05) %>% pull(gene_id)
)

pairs <- combn(names(sig_sets), 2, simplify = FALSE)

jaccard_df <- map_dfr(pairs, function(p) {
  tibble(
    Set_A = p[1],
    Set_B = p[2],
    n_A = length(sig_sets[[p[1]]]),
    n_B = length(sig_sets[[p[2]]]),
    n_intersect = length(intersect(sig_sets[[p[1]]], sig_sets[[p[2]]])),
    jaccard = round(jaccard(sig_sets[[p[1]]], sig_sets[[p[2]]]), 3)
  )
})

datatable(jaccard_df,
          caption = "Jaccard similarity of significant TE upregulation gene sets",
          options = list(pageLength = 10))
```

# 5. Rank Correlation Bar Chart

```{r rank-correlation-barchart, fig.width=7, fig.height=4}
# For genes significant in at least one timepoint (excluding alltimepoints)
sig_any <- wide %>%
  filter(sig_1hr | sig_4hr | sig_24hr)

cat("Genes significant in at least one timepoint:", nrow(sig_any), "\n")

# Compute pairwise Spearman rank correlations on te_lfc among these genes
pairs_lfc <- list(
  c("te_lfc_1hr", "te_lfc_4hr"),
  c("te_lfc_1hr", "te_lfc_24hr"),
  c("te_lfc_4hr", "te_lfc_24hr"),
  c("te_lfc_1hr", "te_lfc_all"),
  c("te_lfc_4hr", "te_lfc_all"),
  c("te_lfc_24hr", "te_lfc_all")
)

pair_labels <- c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr",
                 "1hr vs alltimepoints", "4hr vs alltimepoints", "24hr vs alltimepoints")

rank_cor_df <- map2_dfr(pairs_lfc, pair_labels, function(p, lbl) {
  rho <- cor(sig_any[[p[1]]], sig_any[[p[2]]], method = "spearman", use = "complete.obs")
  tibble(pair = lbl, spearman_rho = rho)
})

rank_cor_df <- rank_cor_df %>%
  mutate(pair = factor(pair, levels = pair_labels),
         comparison_type = ifelse(grepl("alltimepoints", pair), "vs alltimepoints", "timepoint vs timepoint"))

ggplot(rank_cor_df, aes(x = pair, y = spearman_rho, fill = comparison_type)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = round(spearman_rho, 3)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("timepoint vs timepoint" = "#457B9D",
                                "vs alltimepoints" = "#E76F51"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = NULL,
    y = "Spearman ρ",
    title = "Rank correlation of TE LFC among genes significant at ≥1 timepoint"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "top")
```

# 6. Summary of Gene Set Sizes

```{r gene-set-summary}
n_1hr <- length(sig_up_1hr)
n_4hr <- length(sig_up_4hr)
n_24hr <- length(sig_up_24hr)
n_all_tp <- length(sig_sets[["alltimepoints"]])

n_1of3 <- wide %>%
  mutate(n_sig = (te_lfc_1hr > 0.5 & te_padj_1hr < 0.05) +
                  (te_lfc_4hr > 0.5 & te_padj_4hr < 0.05) +
                  (te_lfc_24hr > 0.5 & te_padj_24hr < 0.05)) %>%
  filter(n_sig == 1) %>% nrow()

n_2of3 <- wide %>%
  mutate(n_sig = (te_lfc_1hr > 0.5 & te_padj_1hr < 0.05) +
                  (te_lfc_4hr > 0.5 & te_padj_4hr < 0.05) +
                  (te_lfc_24hr > 0.5 & te_padj_24hr < 0.05)) %>%
  filter(n_sig == 2) %>% nrow()

n_3of3 <- wide %>%
  mutate(n_sig = (te_lfc_1hr > 0.5 & te_padj_1hr < 0.05) +
                  (te_lfc_4hr > 0.5 & te_padj_4hr < 0.05) +
                  (te_lfc_24hr > 0.5 & te_padj_24hr < 0.05)) %>%
  filter(n_sig == 3) %>% nrow()

summary_tbl <- tibble(
  Metric = c(
    "Significant at 1hr", "Significant at 4hr", "Significant at 24hr",
    "Significant in alltimepoints regression",
    "Significant in exactly 1-of-3 timepoints",
    "Significant in exactly 2-of-3 timepoints",
    "Significant in all 3 timepoints"
  ),
  N_genes = c(n_1hr, n_4hr, n_24hr, n_all_tp, n_1of3, n_2of3, n_3of3)
)

datatable(summary_tbl,
          caption = "Gene set sizes: significant TE upregulation (te_lfc > 0.5, te_padj < 0.05)",
          options = list(pageLength = 10, dom = "t"))
```
