---
title: "timecourse 231"
author: "Kate"
date: "2025-11_20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(EnhancedVolcano)
library(msigdbr)
library(viridis)
library(gridExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(survival)
library(DESeq2)
library(limma)
library(clusterProfiler)
library(plotly)
library(ggokabeito)
library(ggvenn)
library(ggpubr)
library(DT)
library(edgeR)
library(cowplot)
library(ggExtra)
library(ggprism)
library(GeneOverlap)
library(writexl)
library(survminer)
library(readxl)
library(reshape2)
```



```{r}
source(here("code/functions.R"))

```


pull in normalized matrix
```{r}

norm_allcounts <- read_csv(here("output/normalizedCounts.csv")) |> 
    column_to_rownames("gene_id") |> 
    as.matrix()

dim(norm_allcounts)



```


PCA of all counts looking at all categories

outlier 107 has already been removed
already filtered for pc genes etc

```{r}
# PCA for all samples

# log transforming with pseudocount to avoid undefined numbers
qcinput <- log2(norm_allcounts + 1) 

qcinput <- qcinput[rowSds(qcinput) > 1,] 

dim(qcinput)



pca_all <- prcomp(t(qcinput), scale. = TRUE)

pca_data_all <- as.data.frame(pca_all$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna"),
    treatment = ifelse(
      grepl("ctrl", sample), "ctrl",
      ifelse(grepl("3e", sample), "si3e",
      ifelse(grepl("3d", sample), "si3d", "other"))
    ),
    timepoint = ifelse(
      grepl("_24_", sample), "24 hr",
      ifelse(grepl("_4_", sample), "4 hr",
      ifelse(grepl("_1_", sample), "1 hr", "other"))
    )
  )


pca_plot_all <- ggplot(pca_data_all, aes(x = PC1, y = PC2, color = treatment, shape = type)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_all.pdf"), pca_plot_all, width=8, height = 6)
```

Pca 2 and 3


```{r}
# log transforming with pseudocount to avoid undefined numbers

pca_plot_all <- ggplot(pca_data_all, aes(x = PC2, y = PC3, color = treatment, shape = condition)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC2 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC3 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_2_3_.pdf"), pca_plot_all, width=8, height = 6)
```


RNA only

```{r}
# PCA for RNA samples only

# 1) Identify RNA columns
rna_cols <- grepl("rna", colnames(norm_allcounts))

# quick check
cat("Number of RNA columns:", sum(rna_cols), "\n")

# 2) subset to only RNA columns and ensure numeric matrix
qcinput_rna <- as.matrix(norm_allcounts[, rna_cols])

# if any columns are not numeric this will error — check:
if(!all(sapply(as.data.frame(qcinput_rna), is.numeric))){
  stop("Some RNA columns are not numeric. Convert them first.")
}

# 3) log2 transform
qcinput_rna <- log2(qcinput_rna + 1)

# 4) filter genes by SD using matrixStats::rowSds
if(!requireNamespace("matrixStats", quietly = TRUE)) install.packages("matrixStats")
library(matrixStats)

qcinput_rna_filtered <- qcinput_rna[rowSds(qcinput_rna) > 1, ]
dim(qcinput_rna_filtered)  # see how many genes remain

# 5) PCA
pca_rna <- prcomp(t(qcinput_rna), scale. = TRUE)

pca_data_rna <- as.data.frame(pca_rna$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna"),
    treatment = ifelse(
      grepl("ctrl", sample), "siCtrl",
      ifelse(grepl("3e", sample), "si3e",
      ifelse(grepl("3d", sample), "si3d", "other"))
    ),
    timepoint = ifelse(
      grepl("_24_", sample), "24hr",
      ifelse(grepl("_4_", sample), "4hr",
      ifelse(grepl("_1_", sample), "1hr", "other"))
    ),
    # Create combined treatment_timepoint for color mapping
    treatment_time = paste(treatment, timepoint, sep = "_"),
    # Set factor levels to control legend order
    treatment_time = factor(treatment_time, levels = c(
      "siCtrl_1hr", "siCtrl_4hr", "siCtrl_24hr",
      "si3d_1hr", "si3d_4hr", "si3d_24hr",
      "si3e_1hr", "si3e_4hr", "si3e_24hr"
    ))
  )

# Define Okabe-Ito colors with shading by timepoint
# siCtrl = grey shades, si3d = orange shades, si3e = blue shades
# Lightest shades for 1hr, medium for 4hr, darkest for 24hr
color_mapping_rna <- c(
  "siCtrl_1hr" = "#CCCCCC",   # lightest grey
  "siCtrl_4hr" = "#999999",   # medium grey (Okabe-Ito grey)
  "siCtrl_24hr" = "#666666",  # darkest grey
  "si3d_1hr" = "#FFCC66",     # lightest orange
  "si3d_4hr" = "#E69F00",     # medium orange (Okabe-Ito orange)
  "si3d_24hr" = "#CC8800",    # darkest orange
  "si3e_1hr" = "#99D6F5",     # lightest blue
  "si3e_4hr" = "#56B4E9",     # medium blue (Okabe-Ito sky blue)
  "si3e_24hr" = "#0072B2"     # darkest blue (Okabe-Ito blue)
)

pca_plot_rna <- ggplot(pca_data_rna, aes(x = PC1, y = PC2, color = treatment_time, shape = condition)) +
  geom_point(size = 3) +
  theme_prism() +
  scale_color_manual(
    values = color_mapping_rna,
    labels = c(
      "siCtrl_1hr" = "siCtrl 1hr",
      "siCtrl_4hr" = "siCtrl 4hr",
      "siCtrl_24hr" = "siCtrl 24hr",
      "si3d_1hr" = "si3d 1hr",
      "si3d_4hr" = "si3d 4hr",
      "si3d_24hr" = "si3d 24hr",
      "si3e_1hr" = "si3e 1hr",
      "si3e_4hr" = "si3e 4hr",
      "si3e_24hr" = "si3e 24hr"
    )
  ) +
  scale_shape_manual(values = c("normoxia" = 16, "hypoxia" = 17)) +
  labs(
    x = paste0("PC1 (", round(summary(pca_rna)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_rna)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - RNA Only",
    color = "Treatment & Timepoint",
    shape = "Condition"
  )

pca_plot_rna
ggsave(here("plots/pca_RNA.pdf"), pca_plot_rna, width=8, height = 6)
```


RIBO only

```{r}
# PCA for ribo samples only

# 1) Identify ribo columns
ribo_cols <- grepl("ribo", colnames(norm_allcounts))

# quick check
cat("Number of ribo columns:", sum(ribo_cols), "\n")

# 2) subset to only ribo columns and ensure numeric matrix
qcinput_ribo <- as.matrix(norm_allcounts[, ribo_cols])

# if any columns are not numeric this will error — check:
if(!all(sapply(as.data.frame(qcinput_ribo), is.numeric))){
  stop("Some ribo columns are not numeric. Convert them first.")
}

# 3) log2 transform
qcinput_ribo <- log2(qcinput_ribo + 1)

# 4) filter genes by SD using matrixStats::rowSds
if(!requireNamespace("matrixStats", quietly = TRUE)) install.packages("matrixStats")
library(matrixStats)

qcinput_ribo_filtered <- qcinput_ribo[rowSds(qcinput_ribo) > 1, ]
dim(qcinput_ribo_filtered)  # see how many genes remain

# 5) PCA
pca_ribo <- prcomp(t(qcinput_ribo), scale. = TRUE)

pca_data_ribo <- as.data.frame(pca_ribo$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna"),
    treatment = ifelse(
      grepl("ctrl", sample), "siCtrl",
      ifelse(grepl("3e", sample), "si3e",
      ifelse(grepl("3d", sample), "si3d", "other"))
    ),
    timepoint = ifelse(
      grepl("_24_", sample), "24hr",
      ifelse(grepl("_4_", sample), "4hr",
      ifelse(grepl("_1_", sample), "1hr", "other"))
    ),
    # Create combined treatment_timepoint for color mapping
    treatment_time = paste(treatment, timepoint, sep = "_"),
    # Set factor levels to control legend order
    treatment_time = factor(treatment_time, levels = c(
      "siCtrl_1hr", "siCtrl_4hr", "siCtrl_24hr",
      "si3d_1hr", "si3d_4hr", "si3d_24hr",
      "si3e_1hr", "si3e_4hr", "si3e_24hr"
    ))
  )

# Define Okabe-Ito colors with shading by timepoint
# siCtrl = grey shades, si3d = orange shades, si3e = blue shades
# Lightest shades for 1hr, medium for 4hr, darkest for 24hr
color_mapping_ribo <- c(
  "siCtrl_1hr" = "#CCCCCC",   # lightest grey
  "siCtrl_4hr" = "#999999",   # medium grey (Okabe-Ito grey)
  "siCtrl_24hr" = "#666666",  # darkest grey
  "si3d_1hr" = "#FFCC66",     # lightest orange
  "si3d_4hr" = "#E69F00",     # medium orange (Okabe-Ito orange)
  "si3d_24hr" = "#CC8800",    # darkest orange
  "si3e_1hr" = "#99D6F5",     # lightest blue
  "si3e_4hr" = "#56B4E9",     # medium blue (Okabe-Ito sky blue)
  "si3e_24hr" = "#0072B2"     # darkest blue (Okabe-Ito blue)
)

pca_plot_ribo <- ggplot(pca_data_ribo, aes(x = PC1, y = PC2, color = treatment_time, shape = condition)) +
  geom_point(size = 3) +
  theme_prism() +
  scale_color_manual(
    values = color_mapping_ribo,
    labels = c(
      "siCtrl_1hr" = "siCtrl 1hr",
      "siCtrl_4hr" = "siCtrl 4hr",
      "siCtrl_24hr" = "siCtrl 24hr",
      "si3d_1hr" = "si3d 1hr",
      "si3d_4hr" = "si3d 4hr",
      "si3d_24hr" = "si3d 24hr",
      "si3e_1hr" = "si3e 1hr",
      "si3e_4hr" = "si3e 4hr",
      "si3e_24hr" = "si3e 24hr"
    )
  ) +
  scale_shape_manual(values = c("normoxia" = 16, "hypoxia" = 17)) +
  labs(
    x = paste0("PC1 (", round(summary(pca_ribo)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_ribo)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - Ribo Only",
    color = "Treatment & Timepoint",
    shape = "Condition"
  )

pca_plot_ribo
ggsave(here("plots/pca_RIBO.pdf"), pca_plot_ribo, width=8, height = 6)
```

# Correlation of TE Changes Across Timepoints in Normoxia

Load knockdown vs control comparisons in normoxia to compare how TE changes evolve over time.

```{r}
# Load KD vs control comparisons in NORMOXIA for si3d
si3d_norm_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol) %>%
  rename(te_lfc_1hr = te_lfc, te_padj_1hr = te_padj)

si3d_norm_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_4hr = te_lfc, te_padj_4hr = te_padj)

si3d_norm_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_24hr = te_lfc, te_padj_24hr = te_padj)

# Merge si3d timepoints
si3d_merged <- si3d_norm_1hr %>%
  inner_join(si3d_norm_4hr, by = "gene_id") %>%
  inner_join(si3d_norm_24hr, by = "gene_id") %>%
  mutate(treatment = "si3d")

# Load KD vs control comparisons in NORMOXIA for si3e
si3e_norm_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol) %>%
  rename(te_lfc_1hr = te_lfc, te_padj_1hr = te_padj)

si3e_norm_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_4hr = te_lfc, te_padj_4hr = te_padj)

si3e_norm_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_24hr = te_lfc, te_padj_24hr = te_padj)

# Merge si3e timepoints
si3e_merged <- si3e_norm_1hr %>%
  inner_join(si3e_norm_4hr, by = "gene_id") %>%
  inner_join(si3e_norm_24hr, by = "gene_id") %>%
  mutate(treatment = "si3e")

# Combine both treatments
all_timepoints <- bind_rows(si3d_merged, si3e_merged)
```

Pairwise timepoint comparisons - 1hr vs 4hr:

```{r}
# Calculate correlations
cor_1hr_4hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_1hr, te_lfc_4hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("1hr vs 4hr correlations:")
print(cor_1hr_4hr)

# Plot 1hr vs 4hr
p_1hr_4hr <- ggplot(all_timepoints, aes(x = te_lfc_1hr, y = te_lfc_4hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 4 hr (Control vs KD)",
    title = "TE Changes: 1hr vs 4hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")


p_1hr_4hr
ggsave(here("plots/normoxia_te_1hr_vs_4hr.pdf"), p_1hr_4hr, width = 7, height = 5)
```

Pairwise timepoint comparisons - 1hr vs 24hr:

```{r}
# Calculate correlations
cor_1hr_24hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_1hr, te_lfc_24hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("1hr vs 24hr correlations:")
print(cor_1hr_24hr)

# Plot 1hr vs 24hr
p_1hr_24hr <- ggplot(all_timepoints, aes(x = te_lfc_1hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "TE Changes: 1hr vs 24hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")

p_1hr_24hr
ggsave(here("plots/normoxia_te_1hr_vs_24hr.pdf"), p_1hr_24hr, width = 7, height = 5)
```

Pairwise timepoint comparisons - 4hr vs 24hr:

```{r}
# Calculate correlations
cor_4hr_24hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_4hr, te_lfc_24hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("4hr vs 24hr correlations:")
print(cor_4hr_24hr)

# Plot 4hr vs 24hr
p_4hr_24hr <- ggplot(all_timepoints, aes(x = te_lfc_4hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 4 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "TE Changes: 4hr vs 24hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")

p_4hr_24hr
ggsave(here("plots/normoxia_te_4hr_vs_24hr.pdf"), p_4hr_24hr, width = 7, height = 5)
```

Faceted view of all pairwise comparisons (filtered for significance separately by treatment):

```{r}
# For si3d: filter genes significant in at least one timepoint for each comparison
si3d_pairs <- all_timepoints %>%
  filter(treatment == "si3d") %>%
  {
    bind_rows(
      filter(., te_padj_1hr < 0.05 | te_padj_4hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "4 hr", te_x = te_lfc_1hr, te_y = te_lfc_4hr),
      filter(., te_padj_1hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "24 hr", te_x = te_lfc_1hr, te_y = te_lfc_24hr),
      filter(., te_padj_4hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "4 hr", y_time = "24 hr", te_x = te_lfc_4hr, te_y = te_lfc_24hr)
    )
  } %>%
  mutate(comparison = paste(x_time, "vs", y_time))

# For si3e: filter genes significant in at least one timepoint for each comparison
si3e_pairs <- all_timepoints %>%
  filter(treatment == "si3e") %>%
  {
    bind_rows(
      filter(., te_padj_1hr < 0.05 | te_padj_4hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "4 hr", te_x = te_lfc_1hr, te_y = te_lfc_4hr),
      filter(., te_padj_1hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "24 hr", te_x = te_lfc_1hr, te_y = te_lfc_24hr),
      filter(., te_padj_4hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "4 hr", y_time = "24 hr", te_x = te_lfc_4hr, te_y = te_lfc_24hr)
    )
  } %>%
  mutate(comparison = paste(x_time, "vs", y_time))

# Combine both treatments
timepoint_pairs <- bind_rows(si3d_pairs, si3e_pairs)

# Calculate correlations for faceting
correlations_facet <- timepoint_pairs %>%
  group_by(treatment, comparison) %>%
  summarise(
    cor = cor(te_x, te_y, use = "complete.obs"),
    .groups = "drop"
  )

# Faceted plot
p_faceted <- ggplot(timepoint_pairs, aes(x = te_x, y = te_y, color=treatment)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) +
  facet_grid(treatment ~ comparison) +
  theme_prism() +
  labs(
    x = "TE log2FC (Earlier Timepoint, Control vs KD)",
    y = "TE log2FC (Later Timepoint, Control vs KD)",
    title = "TE Changes Across Timepoints in Normoxia",
    color="Treatment"
  ) +
  geom_text(
    data = correlations_facet,
    aes(label = paste0("r = ", round(cor, 3))),
    x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5, size = 3
  )+
      scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9"))+
      xlim(-10,10)+
      ylim(-10,10)+
        theme(legend.position = "none")



p_faceted
ggsave(here("plots/normoxia_te_timepoint_correlations_faceted.pdf"), p_faceted, width = 12, height = 9)
```

Combined plot showing all comparisons side by side:

```{r}
p_combined_comparisons <- ggplot(timepoint_pairs, aes(x = te_x, y = te_y, color = treatment)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ comparison, nrow = 1) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC (Earlier Timepoint, Control vs KD)",
    y = "TE log2FC (Later Timepoint, Control vs KD)",
    title = "TE Changes Across Timepoints in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson", size = 2.5)

p_combined_comparisons
ggsave(here("plots/normoxia_te_timepoint_correlations_combined.pdf"), p_combined_comparisons, width = 14, height = 4)
```

# Tracking Top 100 Genes Across Timepoints

Identify genes with the highest TE changes at 1hr and track how they evolve at later timepoints.

## si3d: Top 100 genes at 1hr

```{r}
# Filter for significant genes at 1hr and get top 100 by absolute te_lfc
si3d_top100_1hr <- si3d_merged %>%
  filter(te_padj_1hr < 0.05) %>%
  arrange(desc(abs(te_lfc_1hr))) %>%
  head(100)

cat("Number of significant genes at 1hr (si3d):",
    sum(si3d_merged$te_padj_1hr < 0.05, na.rm = TRUE), "\n")
cat("Top 100 genes selected\n\n")

# Summary statistics for these genes across timepoints
si3d_top100_summary <- si3d_top100_1hr %>%
  summarise(
    n_genes = n(),
    # 1hr stats
    mean_lfc_1hr = mean(te_lfc_1hr, na.rm = TRUE),
    median_lfc_1hr = median(te_lfc_1hr, na.rm = TRUE),
    # 4hr stats
    mean_lfc_4hr = mean(te_lfc_4hr, na.rm = TRUE),
    median_lfc_4hr = median(te_lfc_4hr, na.rm = TRUE),
    sig_4hr = sum(te_padj_4hr < 0.05, na.rm = TRUE),
    # 24hr stats
    mean_lfc_24hr = mean(te_lfc_24hr, na.rm = TRUE),
    median_lfc_24hr = median(te_lfc_24hr, na.rm = TRUE),
    sig_24hr = sum(te_padj_24hr < 0.05, na.rm = TRUE)
  )

print("Summary of top 100 genes (si3d) across timepoints:")
print(si3d_top100_summary)

# Correlation of these top genes across timepoints
cat("\nCorrelations for top 100 genes (si3d):\n")
cat("1hr vs 4hr:", cor(si3d_top100_1hr$te_lfc_1hr, si3d_top100_1hr$te_lfc_4hr, use = "complete.obs"), "\n")
cat("1hr vs 24hr:", cor(si3d_top100_1hr$te_lfc_1hr, si3d_top100_1hr$te_lfc_24hr, use = "complete.obs"), "\n")
cat("4hr vs 24hr:", cor(si3d_top100_1hr$te_lfc_4hr, si3d_top100_1hr$te_lfc_24hr, use = "complete.obs"), "\n")
```

Correlation plots for si3d top 100 genes - 1hr vs 4hr:

```{r}
# Plot 1hr vs 4hr for top 100 genes
p_si3d_top100_1hr_4hr <- ggplot(si3d_top100_1hr, aes(x = te_lfc_1hr, y = te_lfc_4hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#E69F00") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 4 hr (Control vs KD)",
    title = "si3d Top 100 Genes: TE Changes 1hr vs 4hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3d_top100_1hr_4hr
ggsave(here("plots/si3d_top100_te_1hr_vs_4hr.pdf"), p_si3d_top100_1hr_4hr, width = 6, height = 5)
```

Correlation plots for si3d top 100 genes - 1hr vs 24hr:

```{r}
# Plot 1hr vs 24hr for top 100 genes
p_si3d_top100_1hr_24hr <- ggplot(si3d_top100_1hr, aes(x = te_lfc_1hr, y = te_lfc_24hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#E69F00") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "si3d Top 100 Genes: TE Changes 1hr vs 24hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3d_top100_1hr_24hr
ggsave(here("plots/si3d_top100_te_1hr_vs_24hr.pdf"), p_si3d_top100_1hr_24hr, width = 6, height = 5)
```

Correlation plots for si3d top 100 genes - 4hr vs 24hr:

```{r}
# Plot 4hr vs 24hr for top 100 genes
p_si3d_top100_4hr_24hr <- ggplot(si3d_top100_1hr, aes(x = te_lfc_4hr, y = te_lfc_24hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#E69F00") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 4 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "si3d Top 100 Genes: TE Changes 4hr vs 24hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3d_top100_4hr_24hr
ggsave(here("plots/si3d_top100_te_4hr_vs_24hr.pdf"), p_si3d_top100_4hr_24hr, width = 6, height = 5)
```

## si3e: Top 100 genes at 1hr

```{r}
# Filter for significant genes at 1hr and get top 100 by absolute te_lfc
si3e_top100_1hr <- si3e_merged %>%
  filter(te_padj_1hr < 0.05) %>%
  arrange(desc(abs(te_lfc_1hr))) %>%
  head(100)

cat("Number of significant genes at 1hr (si3e):",
    sum(si3e_merged$te_padj_1hr < 0.05, na.rm = TRUE), "\n")
cat("Top 100 genes selected\n\n")

# Summary statistics for these genes across timepoints
si3e_top100_summary <- si3e_top100_1hr %>%
  summarise(
    n_genes = n(),
    # 1hr stats
    mean_lfc_1hr = mean(te_lfc_1hr, na.rm = TRUE),
    median_lfc_1hr = median(te_lfc_1hr, na.rm = TRUE),
    # 4hr stats
    mean_lfc_4hr = mean(te_lfc_4hr, na.rm = TRUE),
    median_lfc_4hr = median(te_lfc_4hr, na.rm = TRUE),
    sig_4hr = sum(te_padj_4hr < 0.05, na.rm = TRUE),
    # 24hr stats
    mean_lfc_24hr = mean(te_lfc_24hr, na.rm = TRUE),
    median_lfc_24hr = median(te_lfc_24hr, na.rm = TRUE),
    sig_24hr = sum(te_padj_24hr < 0.05, na.rm = TRUE)
  )

print("Summary of top 100 genes (si3e) across timepoints:")
print(si3e_top100_summary)

# Correlation of these top genes across timepoints
cat("\nCorrelations for top 100 genes (si3e):\n")
cat("1hr vs 4hr:", cor(si3e_top100_1hr$te_lfc_1hr, si3e_top100_1hr$te_lfc_4hr, use = "complete.obs"), "\n")
cat("1hr vs 24hr:", cor(si3e_top100_1hr$te_lfc_1hr, si3e_top100_1hr$te_lfc_24hr, use = "complete.obs"), "\n")
cat("4hr vs 24hr:", cor(si3e_top100_1hr$te_lfc_4hr, si3e_top100_1hr$te_lfc_24hr, use = "complete.obs"), "\n")
```

Correlation plots for si3e top 100 genes - 1hr vs 4hr:

```{r}
# Plot 1hr vs 4hr for top 100 genes
p_si3e_top100_1hr_4hr <- ggplot(si3e_top100_1hr, aes(x = te_lfc_1hr, y = te_lfc_4hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#56B4E9") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 4 hr (Control vs KD)",
    title = "si3e Top 100 Genes: TE Changes 1hr vs 4hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3e_top100_1hr_4hr
ggsave(here("plots/si3e_top100_te_1hr_vs_4hr.pdf"), p_si3e_top100_1hr_4hr, width = 6, height = 5)
```

Correlation plots for si3e top 100 genes - 1hr vs 24hr:

```{r}
# Plot 1hr vs 24hr for top 100 genes
p_si3e_top100_1hr_24hr <- ggplot(si3e_top100_1hr, aes(x = te_lfc_1hr, y = te_lfc_24hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#56B4E9") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "si3e Top 100 Genes: TE Changes 1hr vs 24hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3e_top100_1hr_24hr
ggsave(here("plots/si3e_top100_te_1hr_vs_24hr.pdf"), p_si3e_top100_1hr_24hr, width = 6, height = 5)
```

Correlation plots for si3e top 100 genes - 4hr vs 24hr:

```{r}
# Plot 4hr vs 24hr for top 100 genes
p_si3e_top100_4hr_24hr <- ggplot(si3e_top100_1hr, aes(x = te_lfc_4hr, y = te_lfc_24hr)) +
  geom_point(alpha = 0.5, size = 2, color = "#56B4E9") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_prism() +
  labs(
    x = "TE log2FC at 4 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "si3e Top 100 Genes: TE Changes 4hr vs 24hr"
  ) +
  stat_cor(method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10)

p_si3e_top100_4hr_24hr
ggsave(here("plots/si3e_top100_te_4hr_vs_24hr.pdf"), p_si3e_top100_4hr_24hr, width = 6, height = 5)
```

## Combined comparison of top genes from both treatments

```{r}
# Compare the top gene sets
si3d_top100_genes <- si3d_top100_1hr$gene_id
si3e_top100_genes <- si3e_top100_1hr$gene_id

# Overlap between top genes
overlap_genes <- intersect(si3d_top100_genes, si3e_top100_genes)
cat("Overlap between si3d and si3e top 100 genes at 1hr:", length(overlap_genes), "genes\n")

# Venn diagram
top100_list <- list(
  "si3d top 100" = si3d_top100_genes,
  "si3e top 100" = si3e_top100_genes
)

p_venn_top100 <- ggvenn(top100_list,
                        fill_color = c("#E69F00", "#56B4E9"),
                        stroke_size = 0.5,
                        set_name_size = 4)

p_venn_top100
ggsave(here("plots/top100_1hr_overlap_venn.pdf"), p_venn_top100, width = 6, height = 5)
```

Combined correlation plots for both treatments:

```{r}
# Combine both treatments for faceted comparison
combined_top100 <- bind_rows(
  si3d_top100_1hr %>% mutate(treatment = "si3d"),
  si3e_top100_1hr %>% mutate(treatment = "si3e")
)

# 1hr vs 4hr combined
p_combined_1hr_4hr <- ggplot(combined_top100, aes(x = te_lfc_1hr, y = te_lfc_4hr, color = treatment)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ treatment) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 4 hr (Control vs KD)",
    title = "Top 100 Genes: TE Changes 1hr vs 4hr"
  ) +
  stat_cor(aes(color = treatment), method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10) +
  theme(legend.position = "none")

p_combined_1hr_4hr
ggsave(here("plots/combined_top100_te_1hr_vs_4hr.pdf"), p_combined_1hr_4hr, width = 10, height = 5)

# 1hr vs 24hr combined
p_combined_1hr_24hr <- ggplot(combined_top100, aes(x = te_lfc_1hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ treatment) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "Top 100 Genes: TE Changes 1hr vs 24hr"
  ) +
  stat_cor(aes(color = treatment), method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10) +
  theme(legend.position = "none")

p_combined_1hr_24hr
ggsave(here("plots/combined_top100_te_1hr_vs_24hr.pdf"), p_combined_1hr_24hr, width = 10, height = 5)

# 4hr vs 24hr combined
p_combined_4hr_24hr <- ggplot(combined_top100, aes(x = te_lfc_4hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ treatment) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 4 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "Top 100 Genes: TE Changes 4hr vs 24hr"
  ) +
  stat_cor(aes(color = treatment), method = "pearson") +
  xlim(-10, 10) +
  ylim(-10, 10) +
  theme(legend.position = "none")

p_combined_4hr_24hr
ggsave(here("plots/combined_top100_te_4hr_vs_24hr.pdf"), p_combined_4hr_24hr, width = 10, height = 5)
```

Export top gene lists:

```{r}
# Export lists with full data
write_csv(si3d_top100_1hr, here("output/si3d_top100_genes_1hr.csv"))
write_csv(si3e_top100_1hr, here("output/si3e_top100_genes_1hr.csv"))

# Export overlap genes
if(length(overlap_genes) > 0) {
  overlap_data <- si3d_merged %>%
    filter(gene_id %in% overlap_genes) %>%
    select(gene_id, symbol, te_lfc_1hr, te_padj_1hr, te_lfc_4hr, te_padj_4hr, te_lfc_24hr, te_padj_24hr) %>%
    rename_with(~paste0(., "_si3d"), .cols = starts_with("te_")) %>%
    left_join(
      si3e_merged %>%
        filter(gene_id %in% overlap_genes) %>%
        select(gene_id, te_lfc_1hr, te_padj_1hr, te_lfc_4hr, te_padj_4hr, te_lfc_24hr, te_padj_24hr) %>%
        rename_with(~paste0(., "_si3e"), .cols = starts_with("te_")),
      by = "gene_id"
    )

  write_csv(overlap_data, here("output/top100_overlap_genes_1hr.csv"))
}

cat("\nExported top 100 gene lists and overlap data to output directory\n")
```

# GSEA: 3d KD Normoxia Gene Set Enrichment in 1hr TE Changes

Test whether genes sensitive to eIF3d KD across all timepoints in normoxia are enriched in the ranked list of TE changes at 1 hour normoxia. Analyze separately:
1. Genes whose translation is promoted by eIF3d (positive TE LFC)
2. Genes whose translation is inhibited by eIF3d (negative TE LFC)
3. Genes in the intersect of promotes and inhibits

```{r}
# Load fgsea if not already loaded
if(!require(fgsea)) {
  install.packages("fgsea")
  library(fgsea)
}
```

Load the gene sets and prepare ranked list:

```{r}
# Load normoxia 3d promotes and inhibits gene sets (POOLED across timepoints)
normoxia_3d_promotes_combined <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_combined.csv"))
normoxia_3d_inhibits_combined <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_combined.csv"))

cat("POOLED gene sets (from combined DESeq2 analysis):\n")
cat("normoxia_3d_promotes_combined:", nrow(normoxia_3d_promotes_combined), "genes\n")
cat("normoxia_3d_inhibits_combined:", nrow(normoxia_3d_inhibits_combined), "genes\n")

# Extract gene lists from pooled analysis
normoxia_3d_promotes_pooled <- normoxia_3d_promotes_combined %>% pull(ensembl_gene)
normoxia_3d_inhibits_pooled <- normoxia_3d_inhibits_combined %>% pull(ensembl_gene)

# Load individual timepoint gene sets to create INTERSECTION gene sets
normoxia_3d_promotes_1hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_1hr.csv"))
normoxia_3d_promotes_4hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_4hr.csv"))
normoxia_3d_promotes_24hr <- read_csv(here("output/genesets/normoxia_3d_promotes_TE_24hr.csv"))

normoxia_3d_inhibits_1hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_1hr.csv"))
normoxia_3d_inhibits_4hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_4hr.csv"))
normoxia_3d_inhibits_24hr <- read_csv(here("output/genesets/normoxia_3d_inhibits_TE_24hr.csv"))

# Create INTERSECTION gene sets (genes significant at ALL three timepoints)
normoxia_3d_promotes_intersection <- Reduce(intersect, list(
  normoxia_3d_promotes_1hr %>% pull(ensembl_gene),
  normoxia_3d_promotes_4hr %>% pull(ensembl_gene),
  normoxia_3d_promotes_24hr %>% pull(ensembl_gene)
))

normoxia_3d_inhibits_intersection <- Reduce(intersect, list(
  normoxia_3d_inhibits_1hr %>% pull(ensembl_gene),
  normoxia_3d_inhibits_4hr %>% pull(ensembl_gene),
  normoxia_3d_inhibits_24hr %>% pull(ensembl_gene)
))

cat("\nINTERSECTION gene sets (significant at ALL three individual timepoints):\n")
cat("normoxia_3d_promotes_intersection:", length(normoxia_3d_promotes_intersection), "genes\n")
cat("normoxia_3d_inhibits_intersection:", length(normoxia_3d_inhibits_intersection), "genes\n")

# Print warning if intersection sets are empty
if(length(normoxia_3d_promotes_intersection) == 0) {
  cat("WARNING: No genes in promotes intersection - no genes are significant at all three timepoints\n")
}
if(length(normoxia_3d_inhibits_intersection) == 0) {
  cat("WARNING: No genes in inhibits intersection - no genes are significant at all three timepoints\n")
}
cat("\n")

# Create ranked list from si3d_norm_1hr (already loaded earlier in this file)
# Rank genes by TE LFC (control vs 3d KD)
ranked_list <- si3d_norm_1hr %>%
  filter(!is.na(te_lfc_1hr)) %>%
  # Remove version numbers from gene_id to match geneset format
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc_1hr)) %>%
  select(gene_id_clean, te_lfc_1hr)

# Create named vector for fgsea
ranked_vector <- setNames(ranked_list$te_lfc_1hr, ranked_list$gene_id_clean)

cat("Ranked list contains", length(ranked_vector), "genes\n")
cat("Range of TE LFC:", round(min(ranked_vector, na.rm = TRUE), 2), "to",
    round(max(ranked_vector, na.rm = TRUE), 2), "\n")
```

Run GSEA for pooled and intersection gene sets:

```{r}
# Create gene set list for fgsea (only include non-empty gene sets)
gene_sets <- list(
  "normoxia_3d_promotes_pooled" = normoxia_3d_promotes_pooled,
  "normoxia_3d_inhibits_pooled" = normoxia_3d_inhibits_pooled,
  "normoxia_3d_promotes_intersection" = normoxia_3d_promotes_intersection,
  "normoxia_3d_inhibits_intersection" = normoxia_3d_inhibits_intersection
)

# Remove empty gene sets (0 genes)
gene_sets <- gene_sets[sapply(gene_sets, length) > 0]

cat("Gene sets to analyze:\n")
for(name in names(gene_sets)) {
  cat("  ", name, ": ", length(gene_sets[[name]]), " genes\n", sep = "")
}
cat("\n")

# Run fgsea (no minimum gene set size)
set.seed(42)
fgsea_results <- fgsea(
  pathways = gene_sets,
  stats = ranked_vector,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

# Display results
print(fgsea_results)

cat("\nGSEA Results Summary:\n\n")
for(i in 1:nrow(fgsea_results)) {
  cat("Pathway:", fgsea_results$pathway[i], "\n")
  cat("  NES:", round(fgsea_results$NES[i], 3), "\n")
  cat("  p-value:", format(fgsea_results$pval[i], scientific = TRUE, digits = 3), "\n")
  cat("  Adjusted p-value:", format(fgsea_results$padj[i], scientific = TRUE, digits = 3), "\n")
  cat("  Gene set size:", fgsea_results$size[i], "\n\n")
}
```

## GSEA Plots

### Pooled Gene Sets

Plot enrichment for pooled promotes gene set:

```{r}
p_gsea_promotes_pooled <- plotEnrichment(
  gene_sets[["normoxia_3d_promotes_pooled"]],
  ranked_vector
) +
  theme_prism() +
  labs(
    title = "GSEA: 3d Promotes TE Genes (Pooled Analysis)",
    subtitle = paste0("n = ", length(normoxia_3d_promotes_pooled), " genes, ",
                     "NES = ", round(fgsea_results$NES[fgsea_results$pathway == "normoxia_3d_promotes_pooled"], 3),
                     ", padj = ", format(fgsea_results$padj[fgsea_results$pathway == "normoxia_3d_promotes_pooled"],
                                        scientific = TRUE, digits = 2)),
    x = "Rank in TE LFC (Control vs si3d, 1hr Normoxia)",
    y = "Enrichment Score"
  )

p_gsea_promotes_pooled
ggsave(here("plots/gsea_3d_promotes_pooled_1hr_normoxia.pdf"), p_gsea_promotes_pooled, width = 8, height = 6)
```

Plot enrichment for pooled inhibits gene set:

```{r}
p_gsea_inhibits_pooled <- plotEnrichment(
  gene_sets[["normoxia_3d_inhibits_pooled"]],
  ranked_vector
) +
  theme_prism() +
  labs(
    title = "GSEA: 3d Inhibits TE Genes (Pooled Analysis)",
    subtitle = paste0("n = ", length(normoxia_3d_inhibits_pooled), " genes, ",
                     "NES = ", round(fgsea_results$NES[fgsea_results$pathway == "normoxia_3d_inhibits_pooled"], 3),
                     ", padj = ", format(fgsea_results$padj[fgsea_results$pathway == "normoxia_3d_inhibits_pooled"],
                                        scientific = TRUE, digits = 2)),
    x = "Rank in TE LFC (Control vs si3d, 1hr Normoxia)",
    y = "Enrichment Score"
  )

p_gsea_inhibits_pooled
ggsave(here("plots/gsea_3d_inhibits_pooled_1hr_normoxia.pdf"), p_gsea_inhibits_pooled, width = 8, height = 6)
```

### Intersection Gene Sets

Plot enrichment for intersection promotes gene set:

```{r}
if(length(normoxia_3d_promotes_intersection) > 0 &&
   "normoxia_3d_promotes_intersection" %in% fgsea_results$pathway) {
  p_gsea_promotes_intersection <- plotEnrichment(
    gene_sets[["normoxia_3d_promotes_intersection"]],
    ranked_vector
  ) +
    theme_prism() +
    labs(
      title = "GSEA: 3d Promotes TE Genes (Intersection of All Timepoints)",
      subtitle = paste0("n = ", length(normoxia_3d_promotes_intersection), " genes, ",
                       "NES = ", round(fgsea_results$NES[fgsea_results$pathway == "normoxia_3d_promotes_intersection"], 3),
                       ", padj = ", format(fgsea_results$padj[fgsea_results$pathway == "normoxia_3d_promotes_intersection"],
                                          scientific = TRUE, digits = 2)),
      x = "Rank in TE LFC (Control vs si3d, 1hr Normoxia)",
      y = "Enrichment Score"
    )

  p_gsea_promotes_intersection
  ggsave(here("plots/gsea_3d_promotes_intersection_1hr_normoxia.pdf"), p_gsea_promotes_intersection, width = 8, height = 6)
} else {
  cat("Skipping promotes intersection plot - gene set is empty (0 genes)\n")
}
```

Plot enrichment for intersection inhibits gene set:

```{r}
if(length(normoxia_3d_inhibits_intersection) > 0 &&
   "normoxia_3d_inhibits_intersection" %in% fgsea_results$pathway) {
  p_gsea_inhibits_intersection <- plotEnrichment(
    gene_sets[["normoxia_3d_inhibits_intersection"]],
    ranked_vector
  ) +
    theme_prism() +
    labs(
      title = "GSEA: 3d Inhibits TE Genes (Intersection of All Timepoints)",
      subtitle = paste0("n = ", length(normoxia_3d_inhibits_intersection), " genes, ",
                       "NES = ", round(fgsea_results$NES[fgsea_results$pathway == "normoxia_3d_inhibits_intersection"], 3),
                       ", padj = ", format(fgsea_results$padj[fgsea_results$pathway == "normoxia_3d_inhibits_intersection"],
                                          scientific = TRUE, digits = 2)),
      x = "Rank in TE LFC (Control vs si3d, 1hr Normoxia)",
      y = "Enrichment Score"
    )

  p_gsea_inhibits_intersection
  ggsave(here("plots/gsea_3d_inhibits_intersection_1hr_normoxia.pdf"), p_gsea_inhibits_intersection, width = 8, height = 6)
} else {
  cat("Skipping inhibits intersection plot - gene set is empty (0 genes)\n")
}
```

# GSEA: si3d Normoxia 4hr

```{r}
# Load 4hr TE data for ranking
si3d_4hr_te <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol)

# Create ranked list
ranked_list_4hr <- si3d_4hr_te %>%
  filter(!is.na(te_lfc)) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc)) %>%
  select(gene_id_clean, te_lfc)

ranked_vector_4hr <- setNames(ranked_list_4hr$te_lfc, ranked_list_4hr$gene_id_clean)

# Run GSEA
set.seed(42)
fgsea_results_4hr <- fgsea(
  pathways = gene_sets[sapply(gene_sets, length) > 0],
  stats = ranked_vector_4hr,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

print(fgsea_results_4hr)
```

## Plots - si3d 4hr

```{r}
# Pooled promotes
p_promotes_pooled_4hr <- plotEnrichment(gene_sets[["normoxia_3d_promotes_pooled"]], ranked_vector_4hr) +
  theme_prism() +
  labs(title = "GSEA: 3d Promotes TE (Pooled) - 4hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3d_promotes_pooled), " genes, ",
                        "NES = ", round(fgsea_results_4hr$NES[fgsea_results_4hr$pathway == "normoxia_3d_promotes_pooled"], 3),
                        ", padj = ", format(fgsea_results_4hr$padj[fgsea_results_4hr$pathway == "normoxia_3d_promotes_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3d, 4hr Normoxia)", y = "Enrichment Score")
p_promotes_pooled_4hr
ggsave(here("plots/gsea_3d_promotes_pooled_4hr_normoxia.pdf"), p_promotes_pooled_4hr, width = 8, height = 6)

# Pooled inhibits
p_inhibits_pooled_4hr <- plotEnrichment(gene_sets[["normoxia_3d_inhibits_pooled"]], ranked_vector_4hr) +
  theme_prism() +
  labs(title = "GSEA: 3d Inhibits TE (Pooled) - 4hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3d_inhibits_pooled), " genes, ",
                        "NES = ", round(fgsea_results_4hr$NES[fgsea_results_4hr$pathway == "normoxia_3d_inhibits_pooled"], 3),
                        ", padj = ", format(fgsea_results_4hr$padj[fgsea_results_4hr$pathway == "normoxia_3d_inhibits_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3d, 4hr Normoxia)", y = "Enrichment Score")
p_inhibits_pooled_4hr
ggsave(here("plots/gsea_3d_inhibits_pooled_4hr_normoxia.pdf"), p_inhibits_pooled_4hr, width = 8, height = 6)

# Intersection promotes
if("normoxia_3d_promotes_intersection" %in% fgsea_results_4hr$pathway) {
  p_promotes_int_4hr <- plotEnrichment(gene_sets[["normoxia_3d_promotes_intersection"]], ranked_vector_4hr) +
    theme_prism() +
    labs(title = "GSEA: 3d Promotes TE (Intersection) - 4hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3d_promotes_intersection), " genes, ",
                          "NES = ", round(fgsea_results_4hr$NES[fgsea_results_4hr$pathway == "normoxia_3d_promotes_intersection"], 3),
                          ", padj = ", format(fgsea_results_4hr$padj[fgsea_results_4hr$pathway == "normoxia_3d_promotes_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3d, 4hr Normoxia)", y = "Enrichment Score")
  p_promotes_int_4hr
  ggsave(here("plots/gsea_3d_promotes_intersection_4hr_normoxia.pdf"), p_promotes_int_4hr, width = 8, height = 6)
}

# Intersection inhibits
if("normoxia_3d_inhibits_intersection" %in% fgsea_results_4hr$pathway) {
  p_inhibits_int_4hr <- plotEnrichment(gene_sets[["normoxia_3d_inhibits_intersection"]], ranked_vector_4hr) +
    theme_prism() +
    labs(title = "GSEA: 3d Inhibits TE (Intersection) - 4hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3d_inhibits_intersection), " genes, ",
                          "NES = ", round(fgsea_results_4hr$NES[fgsea_results_4hr$pathway == "normoxia_3d_inhibits_intersection"], 3),
                          ", padj = ", format(fgsea_results_4hr$padj[fgsea_results_4hr$pathway == "normoxia_3d_inhibits_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3d, 4hr Normoxia)", y = "Enrichment Score")
  p_inhibits_int_4hr
  ggsave(here("plots/gsea_3d_inhibits_intersection_4hr_normoxia.pdf"), p_inhibits_int_4hr, width = 8, height = 6)
}
```

# GSEA: si3d Normoxia 24hr

```{r}
# Load 24hr TE data for ranking
si3d_24hr_te <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol)

# Create ranked list
ranked_list_24hr <- si3d_24hr_te %>%
  filter(!is.na(te_lfc)) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc)) %>%
  select(gene_id_clean, te_lfc)

ranked_vector_24hr <- setNames(ranked_list_24hr$te_lfc, ranked_list_24hr$gene_id_clean)

# Run GSEA
set.seed(42)
fgsea_results_24hr <- fgsea(
  pathways = gene_sets[sapply(gene_sets, length) > 0],
  stats = ranked_vector_24hr,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

print(fgsea_results_24hr)
```

## Plots - si3d 24hr

```{r}
# Pooled promotes
p_promotes_pooled_24hr <- plotEnrichment(gene_sets[["normoxia_3d_promotes_pooled"]], ranked_vector_24hr) +
  theme_prism() +
  labs(title = "GSEA: 3d Promotes TE (Pooled) - 24hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3d_promotes_pooled), " genes, ",
                        "NES = ", round(fgsea_results_24hr$NES[fgsea_results_24hr$pathway == "normoxia_3d_promotes_pooled"], 3),
                        ", padj = ", format(fgsea_results_24hr$padj[fgsea_results_24hr$pathway == "normoxia_3d_promotes_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3d, 24hr Normoxia)", y = "Enrichment Score")
p_promotes_pooled_24hr
ggsave(here("plots/gsea_3d_promotes_pooled_24hr_normoxia.pdf"), p_promotes_pooled_24hr, width = 8, height = 6)

# Pooled inhibits
p_inhibits_pooled_24hr <- plotEnrichment(gene_sets[["normoxia_3d_inhibits_pooled"]], ranked_vector_24hr) +
  theme_prism() +
  labs(title = "GSEA: 3d Inhibits TE (Pooled) - 24hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3d_inhibits_pooled), " genes, ",
                        "NES = ", round(fgsea_results_24hr$NES[fgsea_results_24hr$pathway == "normoxia_3d_inhibits_pooled"], 3),
                        ", padj = ", format(fgsea_results_24hr$padj[fgsea_results_24hr$pathway == "normoxia_3d_inhibits_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3d, 24hr Normoxia)", y = "Enrichment Score")
p_inhibits_pooled_24hr
ggsave(here("plots/gsea_3d_inhibits_pooled_24hr_normoxia.pdf"), p_inhibits_pooled_24hr, width = 8, height = 6)

# Intersection promotes
if("normoxia_3d_promotes_intersection" %in% fgsea_results_24hr$pathway) {
  p_promotes_int_24hr <- plotEnrichment(gene_sets[["normoxia_3d_promotes_intersection"]], ranked_vector_24hr) +
    theme_prism() +
    labs(title = "GSEA: 3d Promotes TE (Intersection) - 24hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3d_promotes_intersection), " genes, ",
                          "NES = ", round(fgsea_results_24hr$NES[fgsea_results_24hr$pathway == "normoxia_3d_promotes_intersection"], 3),
                          ", padj = ", format(fgsea_results_24hr$padj[fgsea_results_24hr$pathway == "normoxia_3d_promotes_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3d, 24hr Normoxia)", y = "Enrichment Score")
  p_promotes_int_24hr
  ggsave(here("plots/gsea_3d_promotes_intersection_24hr_normoxia.pdf"), p_promotes_int_24hr, width = 8, height = 6)
}

# Intersection inhibits
if("normoxia_3d_inhibits_intersection" %in% fgsea_results_24hr$pathway) {
  p_inhibits_int_24hr <- plotEnrichment(gene_sets[["normoxia_3d_inhibits_intersection"]], ranked_vector_24hr) +
    theme_prism() +
    labs(title = "GSEA: 3d Inhibits TE (Intersection) - 24hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3d_inhibits_intersection), " genes, ",
                          "NES = ", round(fgsea_results_24hr$NES[fgsea_results_24hr$pathway == "normoxia_3d_inhibits_intersection"], 3),
                          ", padj = ", format(fgsea_results_24hr$padj[fgsea_results_24hr$pathway == "normoxia_3d_inhibits_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3d, 24hr Normoxia)", y = "Enrichment Score")
  p_inhibits_int_24hr
  ggsave(here("plots/gsea_3d_inhibits_intersection_24hr_normoxia.pdf"), p_inhibits_int_24hr, width = 8, height = 6)
}
```

# GSEA: si3e Normoxia (All Timepoints)

Load si3e gene sets:

```{r}
# Load normoxia 3e gene sets (POOLED)
normoxia_3e_promotes_combined <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_combined.csv"))
normoxia_3e_inhibits_combined <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_combined.csv"))

normoxia_3e_promotes_pooled <- normoxia_3e_promotes_combined %>% pull(ensembl_gene)
normoxia_3e_inhibits_pooled <- normoxia_3e_inhibits_combined %>% pull(ensembl_gene)

# Load individual timepoint gene sets for INTERSECTION
normoxia_3e_promotes_1hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_1hr.csv"))
normoxia_3e_promotes_4hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_4hr.csv"))
normoxia_3e_promotes_24hr <- read_csv(here("output/genesets/normoxia_3e_promotes_TE_24hr.csv"))

normoxia_3e_inhibits_1hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_1hr.csv"))
normoxia_3e_inhibits_4hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_4hr.csv"))
normoxia_3e_inhibits_24hr <- read_csv(here("output/genesets/normoxia_3e_inhibits_TE_24hr.csv"))

# Create INTERSECTION gene sets
normoxia_3e_promotes_intersection <- Reduce(intersect, list(
  normoxia_3e_promotes_1hr %>% pull(ensembl_gene),
  normoxia_3e_promotes_4hr %>% pull(ensembl_gene),
  normoxia_3e_promotes_24hr %>% pull(ensembl_gene)
))

normoxia_3e_inhibits_intersection <- Reduce(intersect, list(
  normoxia_3e_inhibits_1hr %>% pull(ensembl_gene),
  normoxia_3e_inhibits_4hr %>% pull(ensembl_gene),
  normoxia_3e_inhibits_24hr %>% pull(ensembl_gene)
))

cat("si3e POOLED gene sets:\n")
cat("  promotes:", length(normoxia_3e_promotes_pooled), "genes\n")
cat("  inhibits:", length(normoxia_3e_inhibits_pooled), "genes\n\n")

cat("si3e INTERSECTION gene sets:\n")
cat("  promotes:", length(normoxia_3e_promotes_intersection), "genes\n")
cat("  inhibits:", length(normoxia_3e_inhibits_intersection), "genes\n\n")

# Create gene set list
gene_sets_3e <- list(
  "normoxia_3e_promotes_pooled" = normoxia_3e_promotes_pooled,
  "normoxia_3e_inhibits_pooled" = normoxia_3e_inhibits_pooled,
  "normoxia_3e_promotes_intersection" = normoxia_3e_promotes_intersection,
  "normoxia_3e_inhibits_intersection" = normoxia_3e_inhibits_intersection
)

# Remove empty gene sets
gene_sets_3e <- gene_sets_3e[sapply(gene_sets_3e, length) > 0]
```

## GSEA: si3e Normoxia 1hr

```{r}
# Load 1hr TE data for ranking
si3e_1hr_te <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol)

# Create ranked list
ranked_list_3e_1hr <- si3e_1hr_te %>%
  filter(!is.na(te_lfc)) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc)) %>%
  select(gene_id_clean, te_lfc)

ranked_vector_3e_1hr <- setNames(ranked_list_3e_1hr$te_lfc, ranked_list_3e_1hr$gene_id_clean)

# Run GSEA
set.seed(42)
fgsea_results_3e_1hr <- fgsea(
  pathways = gene_sets_3e,
  stats = ranked_vector_3e_1hr,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

print(fgsea_results_3e_1hr)
```

### Plots - si3e 1hr

```{r}
# Pooled promotes
p_3e_promotes_pooled_1hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_pooled"]], ranked_vector_3e_1hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Promotes TE (Pooled) - 1hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_promotes_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_1hr$NES[fgsea_results_3e_1hr$pathway == "normoxia_3e_promotes_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_1hr$padj[fgsea_results_3e_1hr$pathway == "normoxia_3e_promotes_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 1hr Normoxia)", y = "Enrichment Score")
p_3e_promotes_pooled_1hr
ggsave(here("plots/gsea_3e_promotes_pooled_1hr_normoxia.pdf"), p_3e_promotes_pooled_1hr, width = 8, height = 6)

# Pooled inhibits
p_3e_inhibits_pooled_1hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_pooled"]], ranked_vector_3e_1hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Inhibits TE (Pooled) - 1hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_inhibits_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_1hr$NES[fgsea_results_3e_1hr$pathway == "normoxia_3e_inhibits_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_1hr$padj[fgsea_results_3e_1hr$pathway == "normoxia_3e_inhibits_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 1hr Normoxia)", y = "Enrichment Score")
p_3e_inhibits_pooled_1hr
ggsave(here("plots/gsea_3e_inhibits_pooled_1hr_normoxia.pdf"), p_3e_inhibits_pooled_1hr, width = 8, height = 6)

# Intersection promotes
if("normoxia_3e_promotes_intersection" %in% fgsea_results_3e_1hr$pathway) {
  p_3e_promotes_int_1hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_intersection"]], ranked_vector_3e_1hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Promotes TE (Intersection) - 1hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_promotes_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_1hr$NES[fgsea_results_3e_1hr$pathway == "normoxia_3e_promotes_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_1hr$padj[fgsea_results_3e_1hr$pathway == "normoxia_3e_promotes_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 1hr Normoxia)", y = "Enrichment Score")
  p_3e_promotes_int_1hr
  ggsave(here("plots/gsea_3e_promotes_intersection_1hr_normoxia.pdf"), p_3e_promotes_int_1hr, width = 8, height = 6)
}

# Intersection inhibits
if("normoxia_3e_inhibits_intersection" %in% fgsea_results_3e_1hr$pathway) {
  p_3e_inhibits_int_1hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_intersection"]], ranked_vector_3e_1hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Inhibits TE (Intersection) - 1hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_inhibits_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_1hr$NES[fgsea_results_3e_1hr$pathway == "normoxia_3e_inhibits_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_1hr$padj[fgsea_results_3e_1hr$pathway == "normoxia_3e_inhibits_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 1hr Normoxia)", y = "Enrichment Score")
  p_3e_inhibits_int_1hr
  ggsave(here("plots/gsea_3e_inhibits_intersection_1hr_normoxia.pdf"), p_3e_inhibits_int_1hr, width = 8, height = 6)
}
```

## GSEA: si3e Normoxia 4hr

```{r}
# Load 4hr TE data for ranking
si3e_4hr_te <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol)

# Create ranked list
ranked_list_3e_4hr <- si3e_4hr_te %>%
  filter(!is.na(te_lfc)) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc)) %>%
  select(gene_id_clean, te_lfc)

ranked_vector_3e_4hr <- setNames(ranked_list_3e_4hr$te_lfc, ranked_list_3e_4hr$gene_id_clean)

# Run GSEA
set.seed(42)
fgsea_results_3e_4hr <- fgsea(
  pathways = gene_sets_3e,
  stats = ranked_vector_3e_4hr,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

print(fgsea_results_3e_4hr)
```

### Plots - si3e 4hr

```{r}
# Pooled promotes
p_3e_promotes_pooled_4hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_pooled"]], ranked_vector_3e_4hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Promotes TE (Pooled) - 4hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_promotes_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_4hr$NES[fgsea_results_3e_4hr$pathway == "normoxia_3e_promotes_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_4hr$padj[fgsea_results_3e_4hr$pathway == "normoxia_3e_promotes_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 4hr Normoxia)", y = "Enrichment Score")
p_3e_promotes_pooled_4hr
ggsave(here("plots/gsea_3e_promotes_pooled_4hr_normoxia.pdf"), p_3e_promotes_pooled_4hr, width = 8, height = 6)

# Pooled inhibits
p_3e_inhibits_pooled_4hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_pooled"]], ranked_vector_3e_4hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Inhibits TE (Pooled) - 4hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_inhibits_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_4hr$NES[fgsea_results_3e_4hr$pathway == "normoxia_3e_inhibits_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_4hr$padj[fgsea_results_3e_4hr$pathway == "normoxia_3e_inhibits_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 4hr Normoxia)", y = "Enrichment Score")
p_3e_inhibits_pooled_4hr
ggsave(here("plots/gsea_3e_inhibits_pooled_4hr_normoxia.pdf"), p_3e_inhibits_pooled_4hr, width = 8, height = 6)

# Intersection promotes
if("normoxia_3e_promotes_intersection" %in% fgsea_results_3e_4hr$pathway) {
  p_3e_promotes_int_4hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_intersection"]], ranked_vector_3e_4hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Promotes TE (Intersection) - 4hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_promotes_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_4hr$NES[fgsea_results_3e_4hr$pathway == "normoxia_3e_promotes_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_4hr$padj[fgsea_results_3e_4hr$pathway == "normoxia_3e_promotes_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 4hr Normoxia)", y = "Enrichment Score")
  p_3e_promotes_int_4hr
  ggsave(here("plots/gsea_3e_promotes_intersection_4hr_normoxia.pdf"), p_3e_promotes_int_4hr, width = 8, height = 6)
}

# Intersection inhibits
if("normoxia_3e_inhibits_intersection" %in% fgsea_results_3e_4hr$pathway) {
  p_3e_inhibits_int_4hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_intersection"]], ranked_vector_3e_4hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Inhibits TE (Intersection) - 4hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_inhibits_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_4hr$NES[fgsea_results_3e_4hr$pathway == "normoxia_3e_inhibits_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_4hr$padj[fgsea_results_3e_4hr$pathway == "normoxia_3e_inhibits_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 4hr Normoxia)", y = "Enrichment Score")
  p_3e_inhibits_int_4hr
  ggsave(here("plots/gsea_3e_inhibits_intersection_4hr_normoxia.pdf"), p_3e_inhibits_int_4hr, width = 8, height = 6)
}
```

## GSEA: si3e Normoxia 24hr

```{r}
# Load 24hr TE data for ranking
si3e_24hr_te <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol)

# Create ranked list
ranked_list_3e_24hr <- si3e_24hr_te %>%
  filter(!is.na(te_lfc)) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  arrange(desc(te_lfc)) %>%
  select(gene_id_clean, te_lfc)

ranked_vector_3e_24hr <- setNames(ranked_list_3e_24hr$te_lfc, ranked_list_3e_24hr$gene_id_clean)

# Run GSEA
set.seed(42)
fgsea_results_3e_24hr <- fgsea(
  pathways = gene_sets_3e,
  stats = ranked_vector_3e_24hr,
  minSize = 1,
  maxSize = 5000,
  nperm = 10000
)

print(fgsea_results_3e_24hr)
```

### Plots - si3e 24hr

```{r}
# Pooled promotes
p_3e_promotes_pooled_24hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_pooled"]], ranked_vector_3e_24hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Promotes TE (Pooled) - 24hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_promotes_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_24hr$NES[fgsea_results_3e_24hr$pathway == "normoxia_3e_promotes_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_24hr$padj[fgsea_results_3e_24hr$pathway == "normoxia_3e_promotes_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 24hr Normoxia)", y = "Enrichment Score")
p_3e_promotes_pooled_24hr
ggsave(here("plots/gsea_3e_promotes_pooled_24hr_normoxia.pdf"), p_3e_promotes_pooled_24hr, width = 8, height = 6)

# Pooled inhibits
p_3e_inhibits_pooled_24hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_pooled"]], ranked_vector_3e_24hr) +
  theme_prism() +
  labs(title = "GSEA: 3e Inhibits TE (Pooled) - 24hr Normoxia",
       subtitle = paste0("n = ", length(normoxia_3e_inhibits_pooled), " genes, ",
                        "NES = ", round(fgsea_results_3e_24hr$NES[fgsea_results_3e_24hr$pathway == "normoxia_3e_inhibits_pooled"], 3),
                        ", padj = ", format(fgsea_results_3e_24hr$padj[fgsea_results_3e_24hr$pathway == "normoxia_3e_inhibits_pooled"], scientific = TRUE, digits = 2)),
       x = "Rank in TE LFC (Control vs si3e, 24hr Normoxia)", y = "Enrichment Score")
p_3e_inhibits_pooled_24hr
ggsave(here("plots/gsea_3e_inhibits_pooled_24hr_normoxia.pdf"), p_3e_inhibits_pooled_24hr, width = 8, height = 6)

# Intersection promotes
if("normoxia_3e_promotes_intersection" %in% fgsea_results_3e_24hr$pathway) {
  p_3e_promotes_int_24hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_promotes_intersection"]], ranked_vector_3e_24hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Promotes TE (Intersection) - 24hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_promotes_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_24hr$NES[fgsea_results_3e_24hr$pathway == "normoxia_3e_promotes_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_24hr$padj[fgsea_results_3e_24hr$pathway == "normoxia_3e_promotes_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 24hr Normoxia)", y = "Enrichment Score")
  p_3e_promotes_int_24hr
  ggsave(here("plots/gsea_3e_promotes_intersection_24hr_normoxia.pdf"), p_3e_promotes_int_24hr, width = 8, height = 6)
}

# Intersection inhibits
if("normoxia_3e_inhibits_intersection" %in% fgsea_results_3e_24hr$pathway) {
  p_3e_inhibits_int_24hr <- plotEnrichment(gene_sets_3e[["normoxia_3e_inhibits_intersection"]], ranked_vector_3e_24hr) +
    theme_prism() +
    labs(title = "GSEA: 3e Inhibits TE (Intersection) - 24hr Normoxia",
         subtitle = paste0("n = ", length(normoxia_3e_inhibits_intersection), " genes, ",
                          "NES = ", round(fgsea_results_3e_24hr$NES[fgsea_results_3e_24hr$pathway == "normoxia_3e_inhibits_intersection"], 3),
                          ", padj = ", format(fgsea_results_3e_24hr$padj[fgsea_results_3e_24hr$pathway == "normoxia_3e_inhibits_intersection"], scientific = TRUE, digits = 2)),
         x = "Rank in TE LFC (Control vs si3e, 24hr Normoxia)", y = "Enrichment Score")
  p_3e_inhibits_int_24hr
  ggsave(here("plots/gsea_3e_inhibits_intersection_24hr_normoxia.pdf"), p_3e_inhibits_int_24hr, width = 8, height = 6)
}
```

