---
title: "timecourse 231"
author: "Kate"
date: "2025-11_20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(EnhancedVolcano)
library(msigdbr)
library(viridis)
library(gridExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(survival)
library(DESeq2)
library(limma)
library(clusterProfiler)
library(plotly)
library(ggokabeito)
library(ggvenn)
library(ggpubr)
library(DT)
library(edgeR)
library(cowplot)
library(ggExtra)
library(ggprism)
library(GeneOverlap)
library(writexl)
library(survminer)
library(readxl)
library(reshape2)
```



```{r}
source(here("code/functions.R"))

```


pull in normalized matrix
```{r}

norm_allcounts <- read_csv(here("output/normalizedCounts.csv")) |> 
    column_to_rownames("gene_id") |> 
    as.matrix()

dim(norm_allcounts)



```


PCA of all counts looking at all categories

outlier 107 has already been removed
already filtered for pc genes etc

```{r}
# PCA for all samples

# log transforming with pseudocount to avoid undefined numbers
qcinput <- log2(norm_allcounts + 1) 

qcinput <- qcinput[rowSds(qcinput) > 1,] 

dim(qcinput)



pca_all <- prcomp(t(qcinput), scale. = TRUE)

pca_data_all <- as.data.frame(pca_all$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna"),
    treatment = ifelse(
      grepl("ctrl", sample), "ctrl",
      ifelse(grepl("3e", sample), "si3e",
      ifelse(grepl("3d", sample), "si3d", "other"))
    ),
    timepoint = ifelse(
      grepl("_24_", sample), "24 hr",
      ifelse(grepl("_4_", sample), "4 hr",
      ifelse(grepl("_1_", sample), "1 hr", "other"))
    )
  )


pca_plot_all <- ggplot(pca_data_all, aes(x = PC1, y = PC2, color = treatment, shape = type)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_all.pdf"), pca_plot_all, width=8, height = 6)
```

Pca 2 and 3


```{r}
# log transforming with pseudocount to avoid undefined numbers

pca_plot_all <- ggplot(pca_data_all, aes(x = PC2, y = PC3, color = treatment, shape = condition)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC2 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC3 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_2_3_.pdf"), pca_plot_all, width=8, height = 6)
```


RIBO only

```{r}
# PCA for all samples

# log transforming with pseudocount to avoid undefined numbers
# 1) logical vector of ribo columns
ribo_only <- norm_allcounts[, grepl("ribo", names(norm_allcounts))]

# quick check
table(ribo_only)    # how many TRUE / FALSE

# 2) subset to only ribo columns and ensure numeric matrix
qcinput <- as.matrix(norm_allcounts[, ribo_cols])

# if any columns are not numeric this will error — check:
if(!all(sapply(as.data.frame(qcinput), is.numeric))){
  stop("Some ribo columns are not numeric. Convert them first.")
}

# 3) log2 transform
qcinput <- log2(qcinput + 1)

# 4) filter genes by SD using matrixStats::rowSds
if(!requireNamespace("matrixStats", quietly = TRUE)) install.packages("matrixStats")
library(matrixStats)

qcinput_filtered <- qcinput[rowSds(qcinput) > 1, ]
dim(qcinput_filtered)  # see how many genes remain


dim(qcinput)



pca_all <- prcomp(t(qcinput), scale. = TRUE)

pca_data_all <- as.data.frame(pca_all$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna"),
    treatment = ifelse(
      grepl("ctrl", sample), "ctrl",
      ifelse(grepl("3e", sample), "si3e",
      ifelse(grepl("3d", sample), "si3d", "other"))
    ),
    timepoint = ifelse(
      grepl("_24_", sample), "24 hr",
      ifelse(grepl("_4_", sample), "4 hr",
      ifelse(grepl("_1_", sample), "1 hr", "other"))
    )
  )


pca_plot_all <- ggplot(pca_data_all, aes(x = PC1, y = PC2, color = timepoint, shape = type)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_RIBO.pdf"), pca_plot_all, width=8, height = 6)
```

# Correlation of TE Changes Across Timepoints in Normoxia

Load knockdown vs control comparisons in normoxia to compare how TE changes evolve over time.

```{r}
# Load KD vs control comparisons in NORMOXIA for si3d
si3d_norm_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol) %>%
  rename(te_lfc_1hr = te_lfc, te_padj_1hr = te_padj)

si3d_norm_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_4hr = te_lfc, te_padj_4hr = te_padj)

si3d_norm_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_24hr = te_lfc, te_padj_24hr = te_padj)

# Merge si3d timepoints
si3d_merged <- si3d_norm_1hr %>%
  inner_join(si3d_norm_4hr, by = "gene_id") %>%
  inner_join(si3d_norm_24hr, by = "gene_id") %>%
  mutate(treatment = "si3d")

# Load KD vs control comparisons in NORMOXIA for si3e
si3e_norm_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  select(gene_id, te_lfc, te_padj, symbol) %>%
  rename(te_lfc_1hr = te_lfc, te_padj_1hr = te_padj)

si3e_norm_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_4hr = te_lfc, te_padj_4hr = te_padj)

si3e_norm_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  select(gene_id, te_lfc, te_padj) %>%
  rename(te_lfc_24hr = te_lfc, te_padj_24hr = te_padj)

# Merge si3e timepoints
si3e_merged <- si3e_norm_1hr %>%
  inner_join(si3e_norm_4hr, by = "gene_id") %>%
  inner_join(si3e_norm_24hr, by = "gene_id") %>%
  mutate(treatment = "si3e")

# Combine both treatments
all_timepoints <- bind_rows(si3d_merged, si3e_merged)
```

Pairwise timepoint comparisons - 1hr vs 4hr:

```{r}
# Calculate correlations
cor_1hr_4hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_1hr, te_lfc_4hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("1hr vs 4hr correlations:")
print(cor_1hr_4hr)

# Plot 1hr vs 4hr
p_1hr_4hr <- ggplot(all_timepoints, aes(x = te_lfc_1hr, y = te_lfc_4hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 4 hr (Control vs KD)",
    title = "TE Changes: 1hr vs 4hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")


p_1hr_4hr
ggsave(here("plots/normoxia_te_1hr_vs_4hr.pdf"), p_1hr_4hr, width = 7, height = 5)
```

Pairwise timepoint comparisons - 1hr vs 24hr:

```{r}
# Calculate correlations
cor_1hr_24hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_1hr, te_lfc_24hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("1hr vs 24hr correlations:")
print(cor_1hr_24hr)

# Plot 1hr vs 24hr
p_1hr_24hr <- ggplot(all_timepoints, aes(x = te_lfc_1hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 1 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "TE Changes: 1hr vs 24hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")

p_1hr_24hr
ggsave(here("plots/normoxia_te_1hr_vs_24hr.pdf"), p_1hr_24hr, width = 7, height = 5)
```

Pairwise timepoint comparisons - 4hr vs 24hr:

```{r}
# Calculate correlations
cor_4hr_24hr <- all_timepoints %>%
  group_by(treatment) %>%
  summarise(
    cor = cor(te_lfc_4hr, te_lfc_24hr, use = "complete.obs"),
    n = n(),
    .groups = "drop"
  )

print("4hr vs 24hr correlations:")
print(cor_4hr_24hr)

# Plot 4hr vs 24hr
p_4hr_24hr <- ggplot(all_timepoints, aes(x = te_lfc_4hr, y = te_lfc_24hr, color = treatment)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC at 4 hr (Control vs KD)",
    y = "TE log2FC at 24 hr (Control vs KD)",
    title = "TE Changes: 4hr vs 24hr in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson")+
  facet_wrap(~treatment)+
  xlim(-10,10)+
  ylim(-10,10)+
  theme(legend.position = "none")

p_4hr_24hr
ggsave(here("plots/normoxia_te_4hr_vs_24hr.pdf"), p_4hr_24hr, width = 7, height = 5)
```

Faceted view of all pairwise comparisons (filtered for significance separately by treatment):

```{r}
# For si3d: filter genes significant in at least one timepoint for each comparison
si3d_pairs <- all_timepoints %>%
  filter(treatment == "si3d") %>%
  {
    bind_rows(
      filter(., te_padj_1hr < 0.05 | te_padj_4hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "4 hr", te_x = te_lfc_1hr, te_y = te_lfc_4hr),
      filter(., te_padj_1hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "24 hr", te_x = te_lfc_1hr, te_y = te_lfc_24hr),
      filter(., te_padj_4hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "4 hr", y_time = "24 hr", te_x = te_lfc_4hr, te_y = te_lfc_24hr)
    )
  } %>%
  mutate(comparison = paste(x_time, "vs", y_time))

# For si3e: filter genes significant in at least one timepoint for each comparison
si3e_pairs <- all_timepoints %>%
  filter(treatment == "si3e") %>%
  {
    bind_rows(
      filter(., te_padj_1hr < 0.05 | te_padj_4hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "4 hr", te_x = te_lfc_1hr, te_y = te_lfc_4hr),
      filter(., te_padj_1hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "1 hr", y_time = "24 hr", te_x = te_lfc_1hr, te_y = te_lfc_24hr),
      filter(., te_padj_4hr < 0.05 | te_padj_24hr < 0.05) %>%
        mutate(x_time = "4 hr", y_time = "24 hr", te_x = te_lfc_4hr, te_y = te_lfc_24hr)
    )
  } %>%
  mutate(comparison = paste(x_time, "vs", y_time))

# Combine both treatments
timepoint_pairs <- bind_rows(si3d_pairs, si3e_pairs)

# Calculate correlations for faceting
correlations_facet <- timepoint_pairs %>%
  group_by(treatment, comparison) %>%
  summarise(
    cor = cor(te_x, te_y, use = "complete.obs"),
    .groups = "drop"
  )

# Faceted plot
p_faceted <- ggplot(timepoint_pairs, aes(x = te_x, y = te_y, color=treatment)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) +
  facet_grid(treatment ~ comparison) +
  theme_prism() +
  labs(
    x = "TE log2FC (Earlier Timepoint, Control vs KD)",
    y = "TE log2FC (Later Timepoint, Control vs KD)",
    title = "TE Changes Across Timepoints in Normoxia",
    color="Treatment"
  ) +
  geom_text(
    data = correlations_facet,
    aes(label = paste0("r = ", round(cor, 3))),
    x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5, size = 3
  )+
      scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9"))+
      xlim(-10,10)+
      ylim(-10,10)+
        theme(legend.position = "none")



p_faceted
ggsave(here("plots/normoxia_te_timepoint_correlations_faceted.pdf"), p_faceted, width = 12, height = 9)
```

Combined plot showing all comparisons side by side:

```{r}
p_combined_comparisons <- ggplot(timepoint_pairs, aes(x = te_x, y = te_y, color = treatment)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ comparison, nrow = 1) +
  theme_prism() +
  scale_color_manual(values = c("si3d" = "#E69F00", "si3e" = "#56B4E9")) +
  labs(
    x = "TE log2FC (Earlier Timepoint, Control vs KD)",
    y = "TE log2FC (Later Timepoint, Control vs KD)",
    title = "TE Changes Across Timepoints in Normoxia",
    color = "Treatment"
  ) +
  stat_cor(aes(color = treatment), method = "pearson", size = 2.5)

p_combined_comparisons
ggsave(here("plots/normoxia_te_timepoint_correlations_combined.pdf"), p_combined_comparisons, width = 14, height = 4)
```

# Reconciling TE Changes vs Absolute TE Correlations

The analysis above shows TE **CHANGES** (KD vs control) have low correlations across timepoints. Let's compare this with absolute TE values to understand the biology.

## Load absolute TE data from spike-in normalized analysis

```{r load_absolute_te}
# Load absolute TE values calculated from spike-in normalized counts
te_sictrl <- read_csv(here("output", "te_combined_normoxia_sictrl.csv"))
te_si3d <- read_csv(here("output", "te_combined_normoxia_si3d.csv"))
te_si3e <- read_csv(here("output", "te_combined_normoxia_si3e.csv"))

cat("Loaded absolute TE values for all treatments\n")
```

## Calculate TE changes (simple difference: KD - control)

```{r calculate_simple_te_changes}
# Calculate TE changes as simple differences (not DESeq2 interaction term)
# si3d - sictrl
te_diff_si3d <- te_si3d %>%
  select(gene_id, symbol, TE_1hr, TE_4hr, TE_24hr) %>%
  rename(TE_si3d_1hr = TE_1hr, TE_si3d_4hr = TE_4hr, TE_si3d_24hr = TE_24hr) %>%
  left_join(
    te_sictrl %>% select(gene_id, TE_1hr, TE_4hr, TE_24hr) %>%
      rename(TE_ctrl_1hr = TE_1hr, TE_ctrl_4hr = TE_4hr, TE_ctrl_24hr = TE_24hr),
    by = "gene_id"
  ) %>%
  mutate(
    te_diff_1hr = TE_si3d_1hr - TE_ctrl_1hr,
    te_diff_4hr = TE_si3d_4hr - TE_ctrl_4hr,
    te_diff_24hr = TE_si3d_24hr - TE_ctrl_24hr
  ) %>%
  filter(!is.na(te_diff_1hr) & !is.na(te_diff_4hr) & !is.na(te_diff_24hr))

# si3e - sictrl
te_diff_si3e <- te_si3e %>%
  select(gene_id, symbol, TE_1hr, TE_4hr, TE_24hr) %>%
  rename(TE_si3e_1hr = TE_1hr, TE_si3e_4hr = TE_4hr, TE_si3e_24hr = TE_24hr) %>%
  left_join(
    te_sictrl %>% select(gene_id, TE_1hr, TE_4hr, TE_24hr) %>%
      rename(TE_ctrl_1hr = TE_1hr, TE_ctrl_4hr = TE_4hr, TE_ctrl_24hr = TE_24hr),
    by = "gene_id"
  ) %>%
  mutate(
    te_diff_1hr = TE_si3e_1hr - TE_ctrl_1hr,
    te_diff_4hr = TE_si3e_4hr - TE_ctrl_4hr,
    te_diff_24hr = TE_si3e_24hr - TE_ctrl_24hr
  ) %>%
  filter(!is.na(te_diff_1hr) & !is.na(te_diff_4hr) & !is.na(te_diff_24hr))

cat("Calculated simple TE differences (KD - control)\n")
```

## Compare simple TE differences vs DESeq2 te_lfc

```{r compare_methods}
# For si3d: merge simple differences with DESeq2 LFCs
si3d_comparison <- te_diff_si3d %>%
  select(gene_id, symbol, te_diff_1hr, te_diff_4hr, te_diff_24hr) %>%
  left_join(
    si3d_merged %>% select(gene_id, te_lfc_1hr, te_lfc_4hr, te_lfc_24hr),
    by = "gene_id"
  )

# Check correlation between methods at each timepoint
cor_1hr_methods_si3d <- cor(si3d_comparison$te_diff_1hr, si3d_comparison$te_lfc_1hr, use = "complete.obs")
cor_4hr_methods_si3d <- cor(si3d_comparison$te_diff_4hr, si3d_comparison$te_lfc_4hr, use = "complete.obs")
cor_24hr_methods_si3d <- cor(si3d_comparison$te_diff_24hr, si3d_comparison$te_lfc_24hr, use = "complete.obs")

cat("\nCorrelation between simple TE difference and DESeq2 te_lfc (si3d):\n")
cat("1hr:  R =", round(cor_1hr_methods_si3d, 3), "\n")
cat("4hr:  R =", round(cor_4hr_methods_si3d, 3), "\n")
cat("24hr: R =", round(cor_24hr_methods_si3d, 3), "\n")

# Plot comparison for 1hr
p_method_comp_1hr <- ggplot(si3d_comparison, aes(x = te_diff_1hr, y = te_lfc_1hr)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue") +
  theme_prism() +
  labs(
    x = "Simple TE difference (si3d - sictrl)",
    y = "DESeq2 TE log2FC (interaction term)",
    title = paste0("si3d 1hr: TE Difference vs DESeq2 LFC\nR = ", round(cor_1hr_methods_si3d, 3))
  )

print(p_method_comp_1hr)
ggsave(here("plots/te_difference_vs_deseq_si3d_1hr.pdf"), p_method_comp_1hr, width = 6, height = 5)
```

## Correlations of simple TE differences across timepoints

```{r simple_diff_correlations}
# si3d simple differences across time
cor_diff_1hr_4hr_si3d <- cor(te_diff_si3d$te_diff_1hr, te_diff_si3d$te_diff_4hr, use = "complete.obs")
cor_diff_1hr_24hr_si3d <- cor(te_diff_si3d$te_diff_1hr, te_diff_si3d$te_diff_24hr, use = "complete.obs")
cor_diff_4hr_24hr_si3d <- cor(te_diff_si3d$te_diff_4hr, te_diff_si3d$te_diff_24hr, use = "complete.obs")

# si3e simple differences across time
cor_diff_1hr_4hr_si3e <- cor(te_diff_si3e$te_diff_1hr, te_diff_si3e$te_diff_4hr, use = "complete.obs")
cor_diff_1hr_24hr_si3e <- cor(te_diff_si3e$te_diff_1hr, te_diff_si3e$te_diff_24hr, use = "complete.obs")
cor_diff_4hr_24hr_si3e <- cor(te_diff_si3e$te_diff_4hr, te_diff_si3e$te_diff_24hr, use = "complete.obs")

# Plot simple differences across time for si3d
te_diff_pairs_si3d <- bind_rows(
  te_diff_si3d %>% mutate(comparison = "1hr vs 4hr", x = te_diff_1hr, y = te_diff_4hr,
                          cor_val = cor_diff_1hr_4hr_si3d),
  te_diff_si3d %>% mutate(comparison = "1hr vs 24hr", x = te_diff_1hr, y = te_diff_24hr,
                          cor_val = cor_diff_1hr_24hr_si3d),
  te_diff_si3d %>% mutate(comparison = "4hr vs 24hr", x = te_diff_4hr, y = te_diff_24hr,
                          cor_val = cor_diff_4hr_24hr_si3d)
) %>%
  mutate(comparison = factor(comparison, levels = c("1hr vs 4hr", "1hr vs 24hr", "4hr vs 24hr")))

p_simple_diff_si3d <- ggplot(te_diff_pairs_si3d, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  facet_wrap(~comparison) +
  geom_text(data = te_diff_pairs_si3d %>%
              group_by(comparison) %>%
              summarize(cor_val = unique(cor_val),
                       x = min(x, na.rm = TRUE) + 0.1 * (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)),
                       y = max(y, na.rm = TRUE) - 0.1 * (max(y, na.rm = TRUE) - min(y, na.rm = TRUE)),
                       .groups = "drop"),
            aes(x = x, y = y, label = paste0("R = ", round(cor_val, 3))),
            hjust = 0, vjust = 1, size = 3.5) +
  labs(title = "Simple TE Differences (si3d - sictrl) Across Timepoints",
       x = "TE difference (earlier timepoint)",
       y = "TE difference (later timepoint)") +
  theme_prism()

print(p_simple_diff_si3d)
ggsave(here("plots/simple_te_differences_si3d_timepoints.pdf"), p_simple_diff_si3d, width = 12, height = 4)

cat("\n=== COMPARISON SUMMARY ===\n\n")
cat("Simple TE difference correlations across time (si3d):\n")
cat("1hr vs 4hr:  R =", round(cor_diff_1hr_4hr_si3d, 3), "\n")
cat("1hr vs 24hr: R =", round(cor_diff_1hr_24hr_si3d, 3), "\n")
cat("4hr vs 24hr: R =", round(cor_diff_4hr_24hr_si3d, 3), "\n\n")

cat("DESeq2 te_lfc correlations across time (si3d):\n")
cat("1hr vs 4hr:  R =", round(cor_1hr_4hr$cor[cor_1hr_4hr$treatment == "si3d"], 3), "\n")
cat("1hr vs 24hr: R =", round(cor_1hr_24hr$cor[cor_1hr_24hr$treatment == "si3d"], 3), "\n")
cat("4hr vs 24hr: R =", round(cor_4hr_24hr$cor[cor_4hr_24hr$treatment == "si3d"], 3), "\n")
```

## Interpretation

```{r interpretation}
cat("\n=== KEY FINDINGS ===\n\n")

cat("1. ABSOLUTE TE VALUES (from calculate_te_from_normalized_counts.Rmd):\n")
cat("   - High correlations across timepoints within each treatment\n")
cat("   - Indicates baseline TE is stable over time\n\n")

cat("2. TE CHANGES (KD vs control, both methods):\n")
cat("   - Lower correlations across timepoints\n")
cat("   - Simple difference and DESeq2 methods give similar results\n")
cat("   - Indicates KD affects different genes at different times\n\n")

cat("3. BIOLOGICAL INTERPRETATION:\n")
cat("   - Genes maintain their relative TE over time (high absolute TE correlation)\n")
cat("   - BUT the EFFECT of KD varies by timepoint (low change correlation)\n")
cat("   - Example: A gene might have high TE at all timepoints in BOTH ctrl and KD,\n")
cat("     but the DIFFERENCE (KD effect) changes over time\n\n")

cat("4. METHODOLOGY:\n")
cat("   - DESeq2 interaction term (te_lfc) is statistically appropriate\n")
cat("   - Simple difference gives similar biological conclusions\n")
cat("   - Both methods show KD effects evolve over time\n")
```

