---
title: "ORA: C2 Canonical Pathways — si3d and si3e, all conditions"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Over-representation analysis (ORA) using MSigDB C2 Canonical Pathways (C2:CP) for genes
with significant translational efficiency (TE) changes:

- **Promoted**: TE LFC > 0.5, te_padj < 0.05
- **Repressed**: TE LFC < −0.5, te_padj < 0.05

**Factors:** si3d, si3e
**Conditions:** Normoxia 1+4hr (time-regressed), Hypoxia 1hr, 4hr, 24hr

```{r load_libraries}
library(here)
library(tidyverse)
library(clusterProfiler)
library(msigdbr)
library(writexl)
library(enrichplot)
library(pheatmap)

cat("Libraries loaded\n")
```

# Load C2 Canonical Pathway Gene Sets

```{r load_c2cp}
c2_cp_sets <- msigdbr(species = "Homo sapiens", category = "C2") %>%
  filter(grepl("^CP", gs_subcat))

c2_cp_term2gene <- c2_cp_sets %>%
  dplyr::select(gs_name, gene_symbol)

cat("C2:CP pathways loaded:", length(unique(c2_cp_term2gene$gs_name)), "\n")
```

# Load Input Files

```{r load_data}
# si3d files
si3d_norm <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
si3d_hyp1 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
si3d_hyp4 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
si3d_hyp24 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))

# si3e files
si3e_norm <- read_csv(here("output", "te_si3e_vs_sictrl_normoxia_1and4hr.csv"))
si3e_hyp1 <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
si3e_hyp4 <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
si3e_hyp24 <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))

cat("All input files loaded\n")

# Summarize gene counts
for (nm in c("si3d_norm", "si3d_hyp1", "si3d_hyp4", "si3d_hyp24",
             "si3e_norm", "si3e_hyp1", "si3e_hyp4", "si3e_hyp24")) {
  df <- get(nm)
  cat(nm, "- rows:", nrow(df),
      "| promoted:", sum(df$te_lfc > 0.5 & df$te_padj < 0.05, na.rm = TRUE),
      "| repressed:", sum(df$te_lfc < -0.5 & df$te_padj < 0.05, na.rm = TRUE), "\n")
}
```

# ORA Function

```{r ora_function}
run_ora <- function(df, direction, term2gene, label) {
  bg <- df %>% filter(!is.na(symbol)) %>% pull(symbol)
  genes <- if (direction == "promoted") {
    df %>% filter(te_lfc > 0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol)
  } else {
    df %>% filter(te_lfc < -0.5, te_padj < 0.05, !is.na(symbol)) %>% pull(symbol)
  }
  cat(label, direction, "- genes:", length(genes), "| background:", length(bg), "\n")
  if (length(genes) < 5) {
    cat("  Too few genes, skipping\n")
    return(NULL)
  }
  enricher(
    gene = genes,
    universe = bg,
    TERM2GENE = term2gene,
    pAdjustMethod = "fdr",
    pvalueCutoff = 1,
    qvalueCutoff = 1,
    minGSSize = 10,
    maxGSSize = 500
  )
}

extract_results <- function(ora_res, label) {
  if (is.null(ora_res) || nrow(ora_res@result) == 0) {
    cat("  No results for:", label, "\n")
    return(NULL)
  }
  ora_res@result %>%
    dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID) %>%
    mutate(
      gene_ratio_num = sapply(strsplit(GeneRatio, "/"), function(x) as.integer(x[1]) / as.integer(x[2])),
      bg_ratio_num   = sapply(strsplit(BgRatio,   "/"), function(x) as.integer(x[1]) / as.integer(x[2])),
      fold_enrichment = round(gene_ratio_num / bg_ratio_num, 4),
      pvalue = round(pvalue, 6),
      p.adjust = round(p.adjust, 6)
    ) %>%
    dplyr::select(-gene_ratio_num, -bg_ratio_num) %>%
    arrange(p.adjust)
}
```

# Run ORA — si3d

## si3d Normoxia 1+4hr

```{r ora_si3d_norm}
ora_si3d_norm_prom <- run_ora(si3d_norm, "promoted",  c2_cp_term2gene, "si3d_normoxia")
ora_si3d_norm_repr <- run_ora(si3d_norm, "repressed", c2_cp_term2gene, "si3d_normoxia")

res_si3d_norm_prom <- extract_results(ora_si3d_norm_prom, "si3d_norm_promoted")
res_si3d_norm_repr <- extract_results(ora_si3d_norm_repr, "si3d_norm_repressed")

cat("si3d normoxia promoted - significant:", if (!is.null(res_si3d_norm_prom)) sum(res_si3d_norm_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d normoxia repressed - significant:", if (!is.null(res_si3d_norm_repr)) sum(res_si3d_norm_repr$p.adjust < 0.05) else 0, "\n")
```

## si3d Hypoxia 1hr

```{r ora_si3d_hyp1}
ora_si3d_hyp1_prom <- run_ora(si3d_hyp1, "promoted",  c2_cp_term2gene, "si3d_hyp1hr")
ora_si3d_hyp1_repr <- run_ora(si3d_hyp1, "repressed", c2_cp_term2gene, "si3d_hyp1hr")

res_si3d_hyp1_prom <- extract_results(ora_si3d_hyp1_prom, "si3d_hyp1_promoted")
res_si3d_hyp1_repr <- extract_results(ora_si3d_hyp1_repr, "si3d_hyp1_repressed")

cat("si3d hyp 1hr promoted - significant:", if (!is.null(res_si3d_hyp1_prom)) sum(res_si3d_hyp1_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 1hr repressed - significant:", if (!is.null(res_si3d_hyp1_repr)) sum(res_si3d_hyp1_repr$p.adjust < 0.05) else 0, "\n")
```

## si3d Hypoxia 4hr

```{r ora_si3d_hyp4}
ora_si3d_hyp4_prom <- run_ora(si3d_hyp4, "promoted",  c2_cp_term2gene, "si3d_hyp4hr")
ora_si3d_hyp4_repr <- run_ora(si3d_hyp4, "repressed", c2_cp_term2gene, "si3d_hyp4hr")

res_si3d_hyp4_prom <- extract_results(ora_si3d_hyp4_prom, "si3d_hyp4_promoted")
res_si3d_hyp4_repr <- extract_results(ora_si3d_hyp4_repr, "si3d_hyp4_repressed")

cat("si3d hyp 4hr promoted - significant:", if (!is.null(res_si3d_hyp4_prom)) sum(res_si3d_hyp4_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 4hr repressed - significant:", if (!is.null(res_si3d_hyp4_repr)) sum(res_si3d_hyp4_repr$p.adjust < 0.05) else 0, "\n")
```

## si3d Hypoxia 24hr

```{r ora_si3d_hyp24}
ora_si3d_hyp24_prom <- run_ora(si3d_hyp24, "promoted",  c2_cp_term2gene, "si3d_hyp24hr")
ora_si3d_hyp24_repr <- run_ora(si3d_hyp24, "repressed", c2_cp_term2gene, "si3d_hyp24hr")

res_si3d_hyp24_prom <- extract_results(ora_si3d_hyp24_prom, "si3d_hyp24_promoted")
res_si3d_hyp24_repr <- extract_results(ora_si3d_hyp24_repr, "si3d_hyp24_repressed")

cat("si3d hyp 24hr promoted - significant:", if (!is.null(res_si3d_hyp24_prom)) sum(res_si3d_hyp24_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 24hr repressed - significant:", if (!is.null(res_si3d_hyp24_repr)) sum(res_si3d_hyp24_repr$p.adjust < 0.05) else 0, "\n")
```

# Run ORA — si3e

## si3e Normoxia 1+4hr

```{r ora_si3e_norm}
ora_si3e_norm_prom <- run_ora(si3e_norm, "promoted",  c2_cp_term2gene, "si3e_normoxia")
ora_si3e_norm_repr <- run_ora(si3e_norm, "repressed", c2_cp_term2gene, "si3e_normoxia")

res_si3e_norm_prom <- extract_results(ora_si3e_norm_prom, "si3e_norm_promoted")
res_si3e_norm_repr <- extract_results(ora_si3e_norm_repr, "si3e_norm_repressed")

cat("si3e normoxia promoted - significant:", if (!is.null(res_si3e_norm_prom)) sum(res_si3e_norm_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e normoxia repressed - significant:", if (!is.null(res_si3e_norm_repr)) sum(res_si3e_norm_repr$p.adjust < 0.05) else 0, "\n")
```

## si3e Hypoxia 1hr

```{r ora_si3e_hyp1}
ora_si3e_hyp1_prom <- run_ora(si3e_hyp1, "promoted",  c2_cp_term2gene, "si3e_hyp1hr")
ora_si3e_hyp1_repr <- run_ora(si3e_hyp1, "repressed", c2_cp_term2gene, "si3e_hyp1hr")

res_si3e_hyp1_prom <- extract_results(ora_si3e_hyp1_prom, "si3e_hyp1_promoted")
res_si3e_hyp1_repr <- extract_results(ora_si3e_hyp1_repr, "si3e_hyp1_repressed")

cat("si3e hyp 1hr promoted - significant:", if (!is.null(res_si3e_hyp1_prom)) sum(res_si3e_hyp1_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 1hr repressed - significant:", if (!is.null(res_si3e_hyp1_repr)) sum(res_si3e_hyp1_repr$p.adjust < 0.05) else 0, "\n")
```

## si3e Hypoxia 4hr

```{r ora_si3e_hyp4}
ora_si3e_hyp4_prom <- run_ora(si3e_hyp4, "promoted",  c2_cp_term2gene, "si3e_hyp4hr")
ora_si3e_hyp4_repr <- run_ora(si3e_hyp4, "repressed", c2_cp_term2gene, "si3e_hyp4hr")

res_si3e_hyp4_prom <- extract_results(ora_si3e_hyp4_prom, "si3e_hyp4_promoted")
res_si3e_hyp4_repr <- extract_results(ora_si3e_hyp4_repr, "si3e_hyp4_repressed")

cat("si3e hyp 4hr promoted - significant:", if (!is.null(res_si3e_hyp4_prom)) sum(res_si3e_hyp4_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 4hr repressed - significant:", if (!is.null(res_si3e_hyp4_repr)) sum(res_si3e_hyp4_repr$p.adjust < 0.05) else 0, "\n")
```

## si3e Hypoxia 24hr

```{r ora_si3e_hyp24}
ora_si3e_hyp24_prom <- run_ora(si3e_hyp24, "promoted",  c2_cp_term2gene, "si3e_hyp24hr")
ora_si3e_hyp24_repr <- run_ora(si3e_hyp24, "repressed", c2_cp_term2gene, "si3e_hyp24hr")

res_si3e_hyp24_prom <- extract_results(ora_si3e_hyp24_prom, "si3e_hyp24_promoted")
res_si3e_hyp24_repr <- extract_results(ora_si3e_hyp24_repr, "si3e_hyp24_repressed")

cat("si3e hyp 24hr promoted - significant:", if (!is.null(res_si3e_hyp24_prom)) sum(res_si3e_hyp24_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 24hr repressed - significant:", if (!is.null(res_si3e_hyp24_repr)) sum(res_si3e_hyp24_repr$p.adjust < 0.05) else 0, "\n")
```

# Dotplots

```{r plot_function}
make_dotplot <- function(ora_res, label, direction, outfile) {
  if (is.null(ora_res)) {
    cat("No results for:", label, direction, "\n")
    return(invisible(NULL))
  }
  sig <- ora_res@result %>% filter(p.adjust < 0.05)
  if (nrow(sig) == 0) {
    cat("No significant pathways (p.adjust < 0.05) for:", label, direction, "\n")
    return(invisible(NULL))
  }
  cat("Plotting", nrow(sig), "significant pathways for:", label, direction, "\n")
  p <- dotplot(ora_res, showCategory = nrow(sig)) +
    ggtitle(paste(label, direction))
  pdf(outfile, width = 10, height = max(4, nrow(sig) * 0.35 + 2))
  print(p)
  dev.off()
  cat("Saved:", outfile, "\n")
  p
}
```

## si3d Plots

```{r plot_si3d_norm}
make_dotplot(ora_si3d_norm_prom, "si3d_normoxia", "promoted",
             here("plots", "ora_c2cp_si3d_normoxia_promoted.pdf"))
make_dotplot(ora_si3d_norm_repr, "si3d_normoxia", "repressed",
             here("plots", "ora_c2cp_si3d_normoxia_repressed.pdf"))
```

```{r plot_si3d_hyp1}
make_dotplot(ora_si3d_hyp1_prom, "si3d_hyp1hr", "promoted",
             here("plots", "ora_c2cp_si3d_hyp1hr_promoted.pdf"))
make_dotplot(ora_si3d_hyp1_repr, "si3d_hyp1hr", "repressed",
             here("plots", "ora_c2cp_si3d_hyp1hr_repressed.pdf"))
```

```{r plot_si3d_hyp4}
make_dotplot(ora_si3d_hyp4_prom, "si3d_hyp4hr", "promoted",
             here("plots", "ora_c2cp_si3d_hyp4hr_promoted.pdf"))
make_dotplot(ora_si3d_hyp4_repr, "si3d_hyp4hr", "repressed",
             here("plots", "ora_c2cp_si3d_hyp4hr_repressed.pdf"))
```

```{r plot_si3d_hyp24}
make_dotplot(ora_si3d_hyp24_prom, "si3d_hyp24hr", "promoted",
             here("plots", "ora_c2cp_si3d_hyp24hr_promoted.pdf"))
make_dotplot(ora_si3d_hyp24_repr, "si3d_hyp24hr", "repressed",
             here("plots", "ora_c2cp_si3d_hyp24hr_repressed.pdf"))
```

## si3e Plots

```{r plot_si3e_norm}
make_dotplot(ora_si3e_norm_prom, "si3e_normoxia", "promoted",
             here("plots", "ora_c2cp_si3e_normoxia_promoted.pdf"))
make_dotplot(ora_si3e_norm_repr, "si3e_normoxia", "repressed",
             here("plots", "ora_c2cp_si3e_normoxia_repressed.pdf"))
```

```{r plot_si3e_hyp1}
make_dotplot(ora_si3e_hyp1_prom, "si3e_hyp1hr", "promoted",
             here("plots", "ora_c2cp_si3e_hyp1hr_promoted.pdf"))
make_dotplot(ora_si3e_hyp1_repr, "si3e_hyp1hr", "repressed",
             here("plots", "ora_c2cp_si3e_hyp1hr_repressed.pdf"))
```

```{r plot_si3e_hyp4}
make_dotplot(ora_si3e_hyp4_prom, "si3e_hyp4hr", "promoted",
             here("plots", "ora_c2cp_si3e_hyp4hr_promoted.pdf"))
make_dotplot(ora_si3e_hyp4_repr, "si3e_hyp4hr", "repressed",
             here("plots", "ora_c2cp_si3e_hyp4hr_repressed.pdf"))
```

```{r plot_si3e_hyp24}
make_dotplot(ora_si3e_hyp24_prom, "si3e_hyp24hr", "promoted",
             here("plots", "ora_c2cp_si3e_hyp24hr_promoted.pdf"))
make_dotplot(ora_si3e_hyp24_repr, "si3e_hyp24hr", "repressed",
             here("plots", "ora_c2cp_si3e_hyp24hr_repressed.pdf"))
```

# Export to Excel

```{r export_xlsx}
# Helper to get significant results or empty df for a sheet
sig_or_empty <- function(res) {
  empty <- data.frame(
    ID = character(), GeneRatio = character(), BgRatio = character(),
    pvalue = numeric(), p.adjust = numeric(), Count = integer(),
    geneID = character(), fold_enrichment = numeric(),
    stringsAsFactors = FALSE
  )
  if (is.null(res)) return(empty)
  out <- res %>% filter(p.adjust < 0.05)
  if (nrow(out) == 0) return(empty)
  out %>% dplyr::select(ID, GeneRatio, BgRatio, pvalue, p.adjust, Count, geneID, fold_enrichment)
}

sheets <- list(
  si3d_normoxia_promoted  = sig_or_empty(res_si3d_norm_prom),
  si3d_normoxia_repressed = sig_or_empty(res_si3d_norm_repr),
  si3d_hyp_1hr_promoted   = sig_or_empty(res_si3d_hyp1_prom),
  si3d_hyp_1hr_repressed  = sig_or_empty(res_si3d_hyp1_repr),
  si3d_hyp_4hr_promoted   = sig_or_empty(res_si3d_hyp4_prom),
  si3d_hyp_4hr_repressed  = sig_or_empty(res_si3d_hyp4_repr),
  si3d_hyp_24hr_promoted  = sig_or_empty(res_si3d_hyp24_prom),
  si3d_hyp_24hr_repressed = sig_or_empty(res_si3d_hyp24_repr),
  si3e_normoxia_promoted  = sig_or_empty(res_si3e_norm_prom),
  si3e_normoxia_repressed = sig_or_empty(res_si3e_norm_repr),
  si3e_hyp_1hr_promoted   = sig_or_empty(res_si3e_hyp1_prom),
  si3e_hyp_1hr_repressed  = sig_or_empty(res_si3e_hyp1_repr),
  si3e_hyp_4hr_promoted   = sig_or_empty(res_si3e_hyp4_prom),
  si3e_hyp_4hr_repressed  = sig_or_empty(res_si3e_hyp4_repr),
  si3e_hyp_24hr_promoted  = sig_or_empty(res_si3e_hyp24_prom),
  si3e_hyp_24hr_repressed = sig_or_empty(res_si3e_hyp24_repr)
)

write_xlsx(sheets, here("output", "ora_c2cp_3d_3e_all_conditions.xlsx"))
cat("Saved: output/ora_c2cp_3d_3e_all_conditions.xlsx\n")
cat("Sheets:", paste(names(sheets), collapse = ", "), "\n")

# Summary of significant results per sheet
cat("\nSignificant pathways per condition:\n")
for (nm in names(sheets)) {
  cat(sprintf("  %-35s: %d\n", nm, nrow(sheets[[nm]])))
}
```

# Heatmaps

Cross-condition heatmaps colored by signed fold-enrichment (promoted = positive, repressed = negative).

```{r heatmap_helpers}
# Combine promoted/repressed ORA results for one condition into a signed df.
# If a pathway appears in both, keep the row with the lower p.adjust.
build_signed_ora_df <- function(res_prom, res_repr, cond_label) {
  rows <- list()
  if (!is.null(res_prom) && nrow(res_prom) > 0)
    rows[["prom"]] <- res_prom %>% mutate(signed_fe = fold_enrichment, direction = "promoted")
  if (!is.null(res_repr) && nrow(res_repr) > 0)
    rows[["repr"]] <- res_repr %>% mutate(signed_fe = -fold_enrichment, direction = "repressed")
  if (length(rows) == 0) return(NULL)
  bind_rows(rows) %>%
    group_by(ID) %>%
    slice_min(p.adjust, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(condition = cond_label) %>%
    dplyr::select(ID, signed_fe, p.adjust, condition)
}

# Build and save a pheatmap for one factor (si3d or si3e).
make_ora_heatmap <- function(cond_list, factor_label, outfile) {
  combined <- bind_rows(cond_list)
  if (is.null(combined) || nrow(combined) == 0) {
    cat("No data for:", factor_label, "\n")
    return(invisible(NULL))
  }

  sig_pathways <- combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) == 0) {
    cat("No significant pathways for:", factor_label, "\n")
    return(invisible(NULL))
  }
  cat(factor_label, "- significant pathways:", length(sig_pathways), "\n")

  col_order <- c("Normoxia", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

  # Wide matrix — signed fold_enrichment; 0 for untested/non-significant cells
  wide <- combined %>%
    filter(ID %in% sig_pathways) %>%
    dplyr::select(ID, signed_fe, condition) %>%
    pivot_wider(names_from = condition, values_from = signed_fe, values_fill = 0) %>%
    column_to_rownames("ID")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  # Significance matrix — asterisk where p.adjust < 0.05
  sig_mat <- combined %>%
    filter(ID %in% sig_pathways) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(ID, sig, condition) %>%
    pivot_wider(names_from = condition, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(wide), na.rm = TRUE)
  max_abs <- max(max_abs, 1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("blue", "white", "red"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = paste0("C2:CP ORA \u2014 ", factor_label,
                             "\n(signed fold-enrichment; * = p.adjust < 0.05)"),
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(outfile, width = 14, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()
  cat("Saved:", outfile, "\n")
  p
}

# Heatmap for a single direction (promoted or repressed).
# res_list: named list of 4 extract_results() dfs (one per condition).
# Colors by fold_enrichment on a white-to-red scale (all values positive).
make_ora_heatmap_dir <- function(res_list, factor_label, direction_label, outfile) {
  combined <- bind_rows(
    lapply(names(res_list), function(cond) {
      res <- res_list[[cond]]
      if (is.null(res) || nrow(res) == 0) return(NULL)
      res %>% mutate(condition = cond) %>%
        dplyr::select(ID, fold_enrichment, p.adjust, condition)
    })
  )
  if (is.null(combined) || nrow(combined) == 0) {
    cat("No data for:", factor_label, direction_label, "\n")
    return(invisible(NULL))
  }

  sig_pathways <- combined %>% filter(p.adjust < 0.05) %>% pull(ID) %>% unique()
  if (length(sig_pathways) == 0) {
    cat("No significant pathways for:", factor_label, direction_label, "\n")
    return(invisible(NULL))
  }
  cat(factor_label, direction_label, "- significant pathways:", length(sig_pathways), "\n")

  col_order <- c("Normoxia", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

  wide <- combined %>%
    filter(ID %in% sig_pathways) %>%
    dplyr::select(ID, fold_enrichment, condition) %>%
    pivot_wider(names_from = condition, values_from = fold_enrichment, values_fill = 0) %>%
    column_to_rownames("ID")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- combined %>%
    filter(ID %in% sig_pathways) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(ID, sig, condition) %>%
    pivot_wider(names_from = condition, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_fe <- max(wide, na.rm = TRUE)
  max_fe <- max(max_fe, 1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("white", "red"))(100),
    breaks          = seq(0, max_fe, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = paste0("ORA \u2014 ", factor_label, " (", direction_label, ")",
                             "\n(fold-enrichment; * = p.adjust < 0.05)"),
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(outfile, width = 14, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()
  cat("Saved:", outfile, "\n")
  p
}
```

## si3d Heatmap

```{r heatmap_si3d}
si3d_cond_list <- list(
  "Normoxia" = build_signed_ora_df(res_si3d_norm_prom,  res_si3d_norm_repr,  "Normoxia"),
  "Hyp 1hr"  = build_signed_ora_df(res_si3d_hyp1_prom,  res_si3d_hyp1_repr,  "Hyp 1hr"),
  "Hyp 4hr"  = build_signed_ora_df(res_si3d_hyp4_prom,  res_si3d_hyp4_repr,  "Hyp 4hr"),
  "Hyp 24hr" = build_signed_ora_df(res_si3d_hyp24_prom, res_si3d_hyp24_repr, "Hyp 24hr")
)
make_ora_heatmap(si3d_cond_list, "si3d", here("plots", "ora_c2cp_si3d_heatmap.pdf"))
```

## si3e Heatmap

```{r heatmap_si3e}
si3e_cond_list <- list(
  "Normoxia" = build_signed_ora_df(res_si3e_norm_prom,  res_si3e_norm_repr,  "Normoxia"),
  "Hyp 1hr"  = build_signed_ora_df(res_si3e_hyp1_prom,  res_si3e_hyp1_repr,  "Hyp 1hr"),
  "Hyp 4hr"  = build_signed_ora_df(res_si3e_hyp4_prom,  res_si3e_hyp4_repr,  "Hyp 4hr"),
  "Hyp 24hr" = build_signed_ora_df(res_si3e_hyp24_prom, res_si3e_hyp24_repr, "Hyp 24hr")
)
make_ora_heatmap(si3e_cond_list, "si3e", here("plots", "ora_c2cp_si3e_heatmap.pdf"))
```

## C2:CP Directional Heatmaps

Separate heatmaps per direction (fold-enrichment, white–red scale).

```{r heatmap_c2cp_si3d_dir}
make_ora_heatmap_dir(
  list("Normoxia" = res_si3d_norm_prom, "Hyp 1hr" = res_si3d_hyp1_prom,
       "Hyp 4hr"  = res_si3d_hyp4_prom, "Hyp 24hr" = res_si3d_hyp24_prom),
  "si3d C2:CP", "promoted",
  here("plots", "ora_c2cp_si3d_promoted_heatmap.pdf"))

make_ora_heatmap_dir(
  list("Normoxia" = res_si3d_norm_repr, "Hyp 1hr" = res_si3d_hyp1_repr,
       "Hyp 4hr"  = res_si3d_hyp4_repr, "Hyp 24hr" = res_si3d_hyp24_repr),
  "si3d C2:CP", "repressed",
  here("plots", "ora_c2cp_si3d_repressed_heatmap.pdf"))
```

```{r heatmap_c2cp_si3e_dir}
make_ora_heatmap_dir(
  list("Normoxia" = res_si3e_norm_prom, "Hyp 1hr" = res_si3e_hyp1_prom,
       "Hyp 4hr"  = res_si3e_hyp4_prom, "Hyp 24hr" = res_si3e_hyp24_prom),
  "si3e C2:CP", "promoted",
  here("plots", "ora_c2cp_si3e_promoted_heatmap.pdf"))

make_ora_heatmap_dir(
  list("Normoxia" = res_si3e_norm_repr, "Hyp 1hr" = res_si3e_hyp1_repr,
       "Hyp 4hr"  = res_si3e_hyp4_repr, "Hyp 24hr" = res_si3e_hyp24_repr),
  "si3e C2:CP", "repressed",
  here("plots", "ora_c2cp_si3e_repressed_heatmap.pdf"))
```

# KEGG ORA

## Load KEGG Gene Sets

```{r load_kegg}
# MSigDB renamed KEGG subcategory in v7.2+ (CP:KEGG_LEGACY); use grepl to catch any version
cat("C2:CP subcategories available:", paste(unique(c2_cp_sets$gs_subcat), collapse = ", "), "\n")

kegg_term2gene <- c2_cp_sets %>%
  filter(grepl("KEGG", gs_subcat, ignore.case = TRUE)) %>%
  dplyr::select(gs_name, gene_symbol)

cat("KEGG pathways loaded:", length(unique(kegg_term2gene$gs_name)), "\n")
if (nrow(kegg_term2gene) == 0) stop("No KEGG pathways found — check gs_subcat names above")
```

## Run ORA — si3d KEGG

```{r kegg_ora_si3d_norm}
kegg_ora_si3d_norm_prom <- run_ora(si3d_norm, "promoted",  kegg_term2gene, "si3d_normoxia_kegg")
kegg_ora_si3d_norm_repr <- run_ora(si3d_norm, "repressed", kegg_term2gene, "si3d_normoxia_kegg")

kegg_res_si3d_norm_prom <- extract_results(kegg_ora_si3d_norm_prom, "si3d_norm_promoted_kegg")
kegg_res_si3d_norm_repr <- extract_results(kegg_ora_si3d_norm_repr, "si3d_norm_repressed_kegg")

cat("si3d normoxia promoted - significant:", if (!is.null(kegg_res_si3d_norm_prom)) sum(kegg_res_si3d_norm_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d normoxia repressed - significant:", if (!is.null(kegg_res_si3d_norm_repr)) sum(kegg_res_si3d_norm_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3d_hyp1}
kegg_ora_si3d_hyp1_prom <- run_ora(si3d_hyp1, "promoted",  kegg_term2gene, "si3d_hyp1hr_kegg")
kegg_ora_si3d_hyp1_repr <- run_ora(si3d_hyp1, "repressed", kegg_term2gene, "si3d_hyp1hr_kegg")

kegg_res_si3d_hyp1_prom <- extract_results(kegg_ora_si3d_hyp1_prom, "si3d_hyp1_promoted_kegg")
kegg_res_si3d_hyp1_repr <- extract_results(kegg_ora_si3d_hyp1_repr, "si3d_hyp1_repressed_kegg")

cat("si3d hyp 1hr promoted - significant:", if (!is.null(kegg_res_si3d_hyp1_prom)) sum(kegg_res_si3d_hyp1_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 1hr repressed - significant:", if (!is.null(kegg_res_si3d_hyp1_repr)) sum(kegg_res_si3d_hyp1_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3d_hyp4}
kegg_ora_si3d_hyp4_prom <- run_ora(si3d_hyp4, "promoted",  kegg_term2gene, "si3d_hyp4hr_kegg")
kegg_ora_si3d_hyp4_repr <- run_ora(si3d_hyp4, "repressed", kegg_term2gene, "si3d_hyp4hr_kegg")

kegg_res_si3d_hyp4_prom <- extract_results(kegg_ora_si3d_hyp4_prom, "si3d_hyp4_promoted_kegg")
kegg_res_si3d_hyp4_repr <- extract_results(kegg_ora_si3d_hyp4_repr, "si3d_hyp4_repressed_kegg")

cat("si3d hyp 4hr promoted - significant:", if (!is.null(kegg_res_si3d_hyp4_prom)) sum(kegg_res_si3d_hyp4_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 4hr repressed - significant:", if (!is.null(kegg_res_si3d_hyp4_repr)) sum(kegg_res_si3d_hyp4_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3d_hyp24}
kegg_ora_si3d_hyp24_prom <- run_ora(si3d_hyp24, "promoted",  kegg_term2gene, "si3d_hyp24hr_kegg")
kegg_ora_si3d_hyp24_repr <- run_ora(si3d_hyp24, "repressed", kegg_term2gene, "si3d_hyp24hr_kegg")

kegg_res_si3d_hyp24_prom <- extract_results(kegg_ora_si3d_hyp24_prom, "si3d_hyp24_promoted_kegg")
kegg_res_si3d_hyp24_repr <- extract_results(kegg_ora_si3d_hyp24_repr, "si3d_hyp24_repressed_kegg")

cat("si3d hyp 24hr promoted - significant:", if (!is.null(kegg_res_si3d_hyp24_prom)) sum(kegg_res_si3d_hyp24_prom$p.adjust < 0.05) else 0, "\n")
cat("si3d hyp 24hr repressed - significant:", if (!is.null(kegg_res_si3d_hyp24_repr)) sum(kegg_res_si3d_hyp24_repr$p.adjust < 0.05) else 0, "\n")
```

## Run ORA — si3e KEGG

```{r kegg_ora_si3e_norm}
kegg_ora_si3e_norm_prom <- run_ora(si3e_norm, "promoted",  kegg_term2gene, "si3e_normoxia_kegg")
kegg_ora_si3e_norm_repr <- run_ora(si3e_norm, "repressed", kegg_term2gene, "si3e_normoxia_kegg")

kegg_res_si3e_norm_prom <- extract_results(kegg_ora_si3e_norm_prom, "si3e_norm_promoted_kegg")
kegg_res_si3e_norm_repr <- extract_results(kegg_ora_si3e_norm_repr, "si3e_norm_repressed_kegg")

cat("si3e normoxia promoted - significant:", if (!is.null(kegg_res_si3e_norm_prom)) sum(kegg_res_si3e_norm_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e normoxia repressed - significant:", if (!is.null(kegg_res_si3e_norm_repr)) sum(kegg_res_si3e_norm_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3e_hyp1}
kegg_ora_si3e_hyp1_prom <- run_ora(si3e_hyp1, "promoted",  kegg_term2gene, "si3e_hyp1hr_kegg")
kegg_ora_si3e_hyp1_repr <- run_ora(si3e_hyp1, "repressed", kegg_term2gene, "si3e_hyp1hr_kegg")

kegg_res_si3e_hyp1_prom <- extract_results(kegg_ora_si3e_hyp1_prom, "si3e_hyp1_promoted_kegg")
kegg_res_si3e_hyp1_repr <- extract_results(kegg_ora_si3e_hyp1_repr, "si3e_hyp1_repressed_kegg")

cat("si3e hyp 1hr promoted - significant:", if (!is.null(kegg_res_si3e_hyp1_prom)) sum(kegg_res_si3e_hyp1_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 1hr repressed - significant:", if (!is.null(kegg_res_si3e_hyp1_repr)) sum(kegg_res_si3e_hyp1_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3e_hyp4}
kegg_ora_si3e_hyp4_prom <- run_ora(si3e_hyp4, "promoted",  kegg_term2gene, "si3e_hyp4hr_kegg")
kegg_ora_si3e_hyp4_repr <- run_ora(si3e_hyp4, "repressed", kegg_term2gene, "si3e_hyp4hr_kegg")

kegg_res_si3e_hyp4_prom <- extract_results(kegg_ora_si3e_hyp4_prom, "si3e_hyp4_promoted_kegg")
kegg_res_si3e_hyp4_repr <- extract_results(kegg_ora_si3e_hyp4_repr, "si3e_hyp4_repressed_kegg")

cat("si3e hyp 4hr promoted - significant:", if (!is.null(kegg_res_si3e_hyp4_prom)) sum(kegg_res_si3e_hyp4_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 4hr repressed - significant:", if (!is.null(kegg_res_si3e_hyp4_repr)) sum(kegg_res_si3e_hyp4_repr$p.adjust < 0.05) else 0, "\n")
```

```{r kegg_ora_si3e_hyp24}
kegg_ora_si3e_hyp24_prom <- run_ora(si3e_hyp24, "promoted",  kegg_term2gene, "si3e_hyp24hr_kegg")
kegg_ora_si3e_hyp24_repr <- run_ora(si3e_hyp24, "repressed", kegg_term2gene, "si3e_hyp24hr_kegg")

kegg_res_si3e_hyp24_prom <- extract_results(kegg_ora_si3e_hyp24_prom, "si3e_hyp24_promoted_kegg")
kegg_res_si3e_hyp24_repr <- extract_results(kegg_ora_si3e_hyp24_repr, "si3e_hyp24_repressed_kegg")

cat("si3e hyp 24hr promoted - significant:", if (!is.null(kegg_res_si3e_hyp24_prom)) sum(kegg_res_si3e_hyp24_prom$p.adjust < 0.05) else 0, "\n")
cat("si3e hyp 24hr repressed - significant:", if (!is.null(kegg_res_si3e_hyp24_repr)) sum(kegg_res_si3e_hyp24_repr$p.adjust < 0.05) else 0, "\n")
```

## KEGG Heatmaps

```{r kegg_heatmap_si3d}
kegg_si3d_cond_list <- list(
  "Normoxia" = build_signed_ora_df(kegg_res_si3d_norm_prom,  kegg_res_si3d_norm_repr,  "Normoxia"),
  "Hyp 1hr"  = build_signed_ora_df(kegg_res_si3d_hyp1_prom,  kegg_res_si3d_hyp1_repr,  "Hyp 1hr"),
  "Hyp 4hr"  = build_signed_ora_df(kegg_res_si3d_hyp4_prom,  kegg_res_si3d_hyp4_repr,  "Hyp 4hr"),
  "Hyp 24hr" = build_signed_ora_df(kegg_res_si3d_hyp24_prom, kegg_res_si3d_hyp24_repr, "Hyp 24hr")
)
make_ora_heatmap(kegg_si3d_cond_list, "si3d (KEGG)", here("plots", "ora_kegg_si3d_heatmap.pdf"))
```

```{r kegg_heatmap_si3e}
kegg_si3e_cond_list <- list(
  "Normoxia" = build_signed_ora_df(kegg_res_si3e_norm_prom,  kegg_res_si3e_norm_repr,  "Normoxia"),
  "Hyp 1hr"  = build_signed_ora_df(kegg_res_si3e_hyp1_prom,  kegg_res_si3e_hyp1_repr,  "Hyp 1hr"),
  "Hyp 4hr"  = build_signed_ora_df(kegg_res_si3e_hyp4_prom,  kegg_res_si3e_hyp4_repr,  "Hyp 4hr"),
  "Hyp 24hr" = build_signed_ora_df(kegg_res_si3e_hyp24_prom, kegg_res_si3e_hyp24_repr, "Hyp 24hr")
)
make_ora_heatmap(kegg_si3e_cond_list, "si3e (KEGG)", here("plots", "ora_kegg_si3e_heatmap.pdf"))
```

## KEGG Directional Heatmaps

```{r kegg_heatmap_si3d_dir}
make_ora_heatmap_dir(
  list("Normoxia" = kegg_res_si3d_norm_prom, "Hyp 1hr" = kegg_res_si3d_hyp1_prom,
       "Hyp 4hr"  = kegg_res_si3d_hyp4_prom, "Hyp 24hr" = kegg_res_si3d_hyp24_prom),
  "si3d KEGG", "promoted",
  here("plots", "ora_kegg_si3d_promoted_heatmap.pdf"))

make_ora_heatmap_dir(
  list("Normoxia" = kegg_res_si3d_norm_repr, "Hyp 1hr" = kegg_res_si3d_hyp1_repr,
       "Hyp 4hr"  = kegg_res_si3d_hyp4_repr, "Hyp 24hr" = kegg_res_si3d_hyp24_repr),
  "si3d KEGG", "repressed",
  here("plots", "ora_kegg_si3d_repressed_heatmap.pdf"))
```

```{r kegg_heatmap_si3e_dir}
make_ora_heatmap_dir(
  list("Normoxia" = kegg_res_si3e_norm_prom, "Hyp 1hr" = kegg_res_si3e_hyp1_prom,
       "Hyp 4hr"  = kegg_res_si3e_hyp4_prom, "Hyp 24hr" = kegg_res_si3e_hyp24_prom),
  "si3e KEGG", "promoted",
  here("plots", "ora_kegg_si3e_promoted_heatmap.pdf"))

make_ora_heatmap_dir(
  list("Normoxia" = kegg_res_si3e_norm_repr, "Hyp 1hr" = kegg_res_si3e_hyp1_repr,
       "Hyp 4hr"  = kegg_res_si3e_hyp4_repr, "Hyp 24hr" = kegg_res_si3e_hyp24_repr),
  "si3e KEGG", "repressed",
  here("plots", "ora_kegg_si3e_repressed_heatmap.pdf"))
```

```{r}

```