---
title: "eIF3d/e RNA signatures in patient survival data (METABRIC & TCGA)"
author: "Kate"
date: "2026-03-15"
output: html_document
editor_options:
  chunk_output_type: console
---

RNA signatures derived from DESeq2 (si3e/si3d vs sictrl) and applied to METABRIC
and TCGA breast cancer patient datasets to generate survival curves.

Signatures:
- Normoxia: 1hr + 4hr timepoints combined, design = ~timepoint + treatment
- Hypoxia: 1hr only, design = ~treatment

Sign convention: levels = c("siX", "sictrl") → result is sictrl_vs_siX
  Positive LFC = higher in sictrl = eIF3e/3d PROMOTES this gene
  up_genes (LFC > 1) = eIF3e/3d-promoted genes
  down_genes (LFC < -1) = eIF3e/3d-suppressed genes
  total_z_score = mean(up) - mean(down) → high score = high eIF3 activity

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(survminer)
library(survival)
library(ggprism)
library(readxl)
```

```{r colors}
theme_set(theme_prism())

si3e_col <- "#56B4E9"
si3d_col <- "#E69F00"
```

---

# Load data

```{r gene_info}
geneInfo <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = FALSE
) %>%
  filter(!str_detect(X3, "pre_"))

colnames(geneInfo) <- c("gene_id", "symbol", "biotype", "species", "version")

gene_symbol <- geneInfo %>%
  dplyr::select(gene_id, symbol) %>%
  distinct(gene_id, .keep_all = TRUE)
```

```{r raw_counts}
metadata <- read_csv(here("output", "metadata_clean.csv"))

raw_counts <- read_csv(here("counts", "raw_filtered_allCounts.csv")) %>%
  column_to_rownames("gene_id")

# All RNA columns
rna_mat <- as.matrix(raw_counts[, grep("_rna_", colnames(raw_counts))])
```

```{r spikein}
cts_dros <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

# Remove outlier sample
cts_dros <- cts_dros[, !grepl("0107", colnames(cts_dros))]

# Top 500 expressed Drosophila spike-in genes
top500_ids <- cts_dros %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  mutate(total = rowSums(across(-gene_id), na.rm = TRUE)) %>%
  slice_max(total, n = 500, with_ties = FALSE) %>%
  pull(gene_id)

spikein <- cts_dros[top500_ids, , drop = FALSE] %>% round(., 0)
```

---

# Generate RNA signatures via DESeq2

## Helper: run DESeq2 and return full results joined with gene symbols

```{r deseq_helper}
# run_deseq: mirrors deFunction spike-in normalization from functions.R
# 1. Appends spike-in rows to count matrix
# 2. Estimates size factors using spike-in controlGenes
# 3. Removes spike-ins, preserves size factors, then runs DESeq2
run_deseq <- function(count_mat, sample_info, design_formula, result_coef) {
  sample_info <- sample_info[match(colnames(count_mat), sample_info$sample_name), ]
  rownames(sample_info) <- sample_info$sample_name

  # Match spike-in columns to the count matrix samples
  spike_sub <- spikein[, colnames(count_mat), drop = FALSE]

  # Combine human counts + spike-ins
  counts_combined <- rbind(round(count_mat, 0), round(spike_sub, 0))
  spike_indices   <- which(rownames(counts_combined) %in% top500_ids)

  dds <- DESeqDataSetFromMatrix(
    countData = counts_combined,
    colData   = sample_info,
    design    = design_formula
  )

  # Estimate size factors using spike-ins as control genes
  dds <- estimateSizeFactors(dds, controlGenes = spike_indices)
  message("Size factors estimated using ", length(spike_indices), " spike-in genes")

  # Remove spike-ins but preserve size factors
  sf <- sizeFactors(dds)
  dds <- dds[-spike_indices, ]
  sizeFactors(dds) <- sf

  dds <- DESeq(dds)
  message("Result names: ", paste(resultsNames(dds), collapse = ", "))

  res <- results(dds, name = result_coef, cooksCutoff = FALSE, independentFiltering = FALSE)
  lfcShrink(dds = dds, coef = result_coef, res = res, type = "apeglm") %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    inner_join(gene_symbol, ., by = "gene_id")
}
```

## si3e Normoxia (1hr + 4hr combined, ~timepoint + treatment)

```{r sig_3e_norm}
norm_3e_info <- metadata %>%
  filter(library == "rna", condition == "normoxia",
         treatment %in% c("sictrl", "si3e"), hours %in% c(1, 4)) %>%
  mutate(
    treatment = factor(treatment, levels = c("si3e", "sictrl")),
    timepoint = factor(hours)
  )

norm_3e_mat <- rna_mat[, norm_3e_info$sample_name]

res_3e_norm <- run_deseq(
  norm_3e_mat, norm_3e_info,
  design_formula = ~ timepoint + treatment,
  result_coef    = "treatment_sictrl_vs_si3e"
)

# Confirm direction: EIF3E should be positive (more in sictrl than si3e KD)
res_3e_norm %>% filter(symbol == "EIF3E") %>% dplyr::select(symbol, log2FoldChange, padj)
```

```{r sig_3e_norm_split}
sig_3e_norm <- res_3e_norm %>% filter(padj < 0.05)
cat("Total significant genes (3e normoxia):", nrow(sig_3e_norm), "\n")

up_3e_norm   <- sig_3e_norm %>% filter(log2FoldChange >  1) %>% pull(symbol)
down_3e_norm <- sig_3e_norm %>% filter(log2FoldChange < -1) %>% pull(symbol)
cat("LFC1  — Up:", length(up_3e_norm), "  Down:", length(down_3e_norm), "\n")

up_3e_norm_lfc15   <- sig_3e_norm %>% filter(log2FoldChange >  1.5) %>% pull(symbol)
down_3e_norm_lfc15 <- sig_3e_norm %>% filter(log2FoldChange < -1.5) %>% pull(symbol)
cat("LFC1.5 — Up:", length(up_3e_norm_lfc15), "  Down:", length(down_3e_norm_lfc15), "\n")
```

## si3d Normoxia (1hr + 4hr combined, ~timepoint + treatment)

```{r sig_3d_norm}
norm_3d_info <- metadata %>%
  filter(library == "rna", condition == "normoxia",
         treatment %in% c("sictrl", "si3d"), hours %in% c(1, 4)) %>%
  mutate(
    treatment = factor(treatment, levels = c("si3d", "sictrl")),
    timepoint = factor(hours)
  )

norm_3d_mat <- rna_mat[, norm_3d_info$sample_name]

res_3d_norm <- run_deseq(
  norm_3d_mat, norm_3d_info,
  design_formula = ~ timepoint + treatment,
  result_coef    = "treatment_sictrl_vs_si3d"
)

res_3d_norm %>% filter(symbol == "EIF3D") %>% dplyr::select(symbol, log2FoldChange, padj)
```

```{r sig_3d_norm_split}
sig_3d_norm <- res_3d_norm %>% filter(padj < 0.05)
cat("Total significant genes (3d normoxia):", nrow(sig_3d_norm), "\n")

up_3d_norm   <- sig_3d_norm %>% filter(log2FoldChange >  1) %>% pull(symbol)
down_3d_norm <- sig_3d_norm %>% filter(log2FoldChange < -1) %>% pull(symbol)
cat("LFC1  — Up:", length(up_3d_norm), "  Down:", length(down_3d_norm), "\n")

up_3d_norm_lfc15   <- sig_3d_norm %>% filter(log2FoldChange >  1.5) %>% pull(symbol)
down_3d_norm_lfc15 <- sig_3d_norm %>% filter(log2FoldChange < -1.5) %>% pull(symbol)
cat("LFC1.5 — Up:", length(up_3d_norm_lfc15), "  Down:", length(down_3d_norm_lfc15), "\n")
```

## si3e Hypoxia 1hr (~treatment)

```{r sig_3e_hyp}
hyp_3e_info <- metadata %>%
  filter(library == "rna", condition == "hypoxia",
         treatment %in% c("sictrl", "si3e"), hours == 1) %>%
  mutate(treatment = factor(treatment, levels = c("si3e", "sictrl")))

hyp_3e_mat <- rna_mat[, hyp_3e_info$sample_name]

res_3e_hyp <- run_deseq(
  hyp_3e_mat, hyp_3e_info,
  design_formula = ~ treatment,
  result_coef    = "treatment_sictrl_vs_si3e"
)

res_3e_hyp %>% filter(symbol == "EIF3E") %>% dplyr::select(symbol, log2FoldChange, padj)
```

```{r sig_3e_hyp_split}
sig_3e_hyp <- res_3e_hyp %>% filter(padj < 0.05)
cat("Total significant genes (3e hypoxia 1hr):", nrow(sig_3e_hyp), "\n")

up_3e_hyp   <- sig_3e_hyp %>% filter(log2FoldChange >  1) %>% pull(symbol)
down_3e_hyp <- sig_3e_hyp %>% filter(log2FoldChange < -1) %>% pull(symbol)
cat("LFC1  — Up:", length(up_3e_hyp), "  Down:", length(down_3e_hyp), "\n")

up_3e_hyp_lfc15   <- sig_3e_hyp %>% filter(log2FoldChange >  1.5) %>% pull(symbol)
down_3e_hyp_lfc15 <- sig_3e_hyp %>% filter(log2FoldChange < -1.5) %>% pull(symbol)
cat("LFC1.5 — Up:", length(up_3e_hyp_lfc15), "  Down:", length(down_3e_hyp_lfc15), "\n")

write_csv(sig_3e_hyp %>% filter(abs(log2FoldChange) > 1),
          here("output", "rna_sig_3e_hypoxia_1hr_lfc1.csv"))
```

## si3d Hypoxia 1hr (~treatment)

```{r sig_3d_hyp}
hyp_3d_info <- metadata %>%
  filter(library == "rna", condition == "hypoxia",
         treatment %in% c("sictrl", "si3d"), hours == 1) %>%
  mutate(treatment = factor(treatment, levels = c("si3d", "sictrl")))

hyp_3d_mat <- rna_mat[, hyp_3d_info$sample_name]

res_3d_hyp <- run_deseq(
  hyp_3d_mat, hyp_3d_info,
  design_formula = ~ treatment,
  result_coef    = "treatment_sictrl_vs_si3d"
)

res_3d_hyp %>% filter(symbol == "EIF3D") %>% dplyr::select(symbol, log2FoldChange, padj)
```

```{r sig_3d_hyp_split}
sig_3d_hyp <- res_3d_hyp %>% filter(padj < 0.05)
cat("Total significant genes (3d hypoxia 1hr):", nrow(sig_3d_hyp), "\n")

up_3d_hyp   <- sig_3d_hyp %>% filter(log2FoldChange >  1) %>% pull(symbol)
down_3d_hyp <- sig_3d_hyp %>% filter(log2FoldChange < -1) %>% pull(symbol)
cat("LFC1  — Up:", length(up_3d_hyp), "  Down:", length(down_3d_hyp), "\n")

up_3d_hyp_lfc15   <- sig_3d_hyp %>% filter(log2FoldChange >  1.5) %>% pull(symbol)
down_3d_hyp_lfc15 <- sig_3d_hyp %>% filter(log2FoldChange < -1.5) %>% pull(symbol)
cat("LFC1.5 — Up:", length(up_3d_hyp_lfc15), "  Down:", length(down_3d_hyp_lfc15), "\n")

write_csv(sig_3d_hyp %>% filter(abs(log2FoldChange) > 1),
          here("output", "rna_sig_3d_hypoxia_1hr_lfc1.csv"))
```

---

# METABRIC survival analysis

METABRIC: Nature 2012 and Nat Commun 2016, downloaded from cBioPortal.

```{r metabric_load, cache=TRUE}
cbioportal_path <- "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/cbioportal_data"

zscores_mb <- read.table(
  file.path(cbioportal_path, "brca_metabric/data_mrna_illumina_microarray_zscores_ref_diploid_samples.txt"),
  sep = "\t", header = TRUE
)

# Deduplicate genes by averaging — rowsum() is much faster than summarise(across())
tmp_mb  <- zscores_mb %>% dplyr::select(-Entrez_Gene_Id)
mat_mb  <- as.matrix(tmp_mb[, -1])   # numeric columns only
genes_mb <- tmp_mb$Hugo_Symbol
rs_mb   <- rowsum(mat_mb, group = genes_mb, na.rm = TRUE)
n_mb    <- table(genes_mb)
merged_zscores_mb <- as.data.frame(rs_mb / as.vector(n_mb[rownames(rs_mb)]))
```

```{r metabric_clin}
clin_mb <- read_excel(
  file.path(cbioportal_path, "brca_metabric/data_clinical_patient_edit.xlsx")
)
clin_mb$OS_MONTHS <- as.numeric(clin_mb$OS_MONTHS)
clin_mb$OS_STATUS[which(clin_mb$OS_STATUS == "1:DECEASED" & clin_mb$OS_MONTHS > 120)] <- "0:LIVING"
clin_mb$OS_STATUS <- clin_mb$OS_STATUS == "1:DECEASED"
clin_mb$OS_YEARS  <- clin_mb$OS_MONTHS / 12
clin_mb$OS_YEARS[which(clin_mb$OS_YEARS >= 10)] <- 10
```

```{r compute_zscore_fn}
# Compute composite z-score for a signature in an expression matrix (genes x patients)
compute_sig_zscores <- function(expr_mat, up_genes, down_genes, sig_all_genes) {
  # Filter matrix to signature genes present in data
  expr_sub <- expr_mat[rownames(expr_mat) %in% sig_all_genes, , drop = FALSE]

  zdf <- data.frame(
    patient      = colnames(expr_sub),
    up_z_score   = NA_real_,
    down_z_score = NA_real_
  )
  for (i in seq_len(nrow(zdf))) {
    up_present   <- intersect(up_genes,   rownames(expr_sub))
    down_present <- intersect(down_genes, rownames(expr_sub))
    zdf$up_z_score[i]   <- mean(expr_sub[up_present,   i], na.rm = TRUE)
    zdf$down_z_score[i] <- mean(expr_sub[down_present, i], na.rm = TRUE)
  }
  zdf %>%
    mutate(total_z_score = up_z_score - down_z_score)
}

# Assign quartiles and clean patient IDs for METABRIC
assign_quartiles_mb <- function(zdf) {
  qbreaks <- quantile(zdf$total_z_score, probs = seq(0, 1, 0.25), na.rm = TRUE)
  zdf$quantile <- cut(zdf$total_z_score, breaks = qbreaks, include.lowest = TRUE, labels = FALSE)
  zdf$patient  <- sub("-[^-]*$", "", zdf$patient)
  zdf$patient  <- gsub("\\.", "-", zdf$patient)
  zdf
}

# Assign quartiles and clean patient IDs for TCGA
assign_quartiles_tcga <- function(zdf) {
  qbreaks <- quantile(zdf$total_z_score, probs = seq(0, 1, 0.25), na.rm = TRUE)
  zdf$quantile <- cut(zdf$total_z_score, breaks = qbreaks, include.lowest = TRUE, labels = FALSE)
  zdf$patient  <- gsub("\\.", "-", as.character(zdf$patient))
  zdf$patient  <- sub("-[^-]*$", "", zdf$patient)
  zdf
}
```

```{r surv_plot_fn}
# Plot survival curve comparing Q1 vs Q4 and return HR / p-value
plot_surv <- function(clin_df, zscores_df, label_low, label_high, color_high,
                      title = "", save_name = NULL, save_path = here("plots")) {

  merged <- left_join(clin_df, zscores_df, by = c("PATIENT_ID" = "patient")) %>%
    filter(!is.na(quantile)) %>%
    filter(quantile %in% c(1, 4))

  fit    <- survfit(Surv(OS_YEARS, OS_STATUS) ~ quantile, data = merged)
  coxfit <- coxph(Surv(OS_YEARS, OS_STATUS) ~ quantile, data = merged)
  s      <- summary(coxfit)

  hr   <- round(s$coefficients[1, "exp(coef)"], 2)
  pval <- format.pval(s$coefficients[1, "Pr(>|z|)"], digits = 2, eps = 0.001)

  colors <- grDevices::colorRampPalette(c("grey", color_high))(2)

  p <- ggsurvplot(
    fit,
    data         = merged,
    size         = 1,
    palette      = colors,
    legend.labs  = c(label_low, label_high),
    break.time.by = 1,
    ylab         = "Overall Survival",
    xlab         = "Time in years",
    xlim         = c(0, 10),
    font.x       = 20,
    font.y       = 20,
    font.tickslab = 16,
    font.legend  = 16,
    conf.int     = FALSE,
    pval         = FALSE,
    risk.table   = FALSE,
    ggtheme      = theme_prism()
  )

  final_plot <- p$plot +
    annotate("text", x = 3, y = 0.2,
             label = paste0("HR = ", hr, "\np = ", pval),
             size = 6, color = "black") +
    ggtitle(title)

  print(final_plot)

  if (!is.null(save_name)) {
    ggsave(save_name, path = save_path, width = 8, height = 6)
  }

  invisible(list(hr = hr, pval = pval, n = nrow(merged)))
}
```

## METABRIC: eIF3e normoxia signature

```{r mb_3e_norm}
zdf_3e_norm_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3e_norm, down_3e_norm, sig_3e_norm$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3e_norm_mb,
  label_low  = "Low 3e (normoxia)",
  label_high = "High 3e (normoxia)",
  color_high = si3e_col,
  title      = "METABRIC: si3e normoxia LFC1",
  save_name  = "mb_3e_norm_lfc1_surv.pdf"
)
```

```{r mb_3e_norm_lfc15}
zdf_3e_norm_lfc15_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3e_norm_lfc15, down_3e_norm_lfc15, sig_3e_norm$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3e_norm_lfc15_mb,
  label_low  = "Low 3e (normoxia)",
  label_high = "High 3e (normoxia)",
  color_high = si3e_col,
  title      = "METABRIC: si3e normoxia LFC1.5",
  save_name  = "mb_3e_norm_lfc15_surv.pdf"
)
```

## METABRIC: eIF3d normoxia signature

```{r mb_3d_norm}
zdf_3d_norm_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3d_norm, down_3d_norm, sig_3d_norm$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3d_norm_mb,
  label_low  = "Low 3d (normoxia)",
  label_high = "High 3d (normoxia)",
  color_high = si3d_col,
  title      = "METABRIC: si3d normoxia LFC1",
  save_name  = "mb_3d_norm_lfc1_surv.pdf"
)
```

```{r mb_3d_norm_lfc15}
zdf_3d_norm_lfc15_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3d_norm_lfc15, down_3d_norm_lfc15, sig_3d_norm$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3d_norm_lfc15_mb,
  label_low  = "Low 3d (normoxia)",
  label_high = "High 3d (normoxia)",
  color_high = si3d_col,
  title      = "METABRIC: si3d normoxia LFC1.5",
  save_name  = "mb_3d_norm_lfc15_surv.pdf"
)
```

## METABRIC: eIF3e hypoxia 1hr signature

```{r mb_3e_hyp}
zdf_3e_hyp_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3e_hyp, down_3e_hyp, sig_3e_hyp$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3e_hyp_mb,
  label_low  = "Low 3e (hypoxia)",
  label_high = "High 3e (hypoxia)",
  color_high = si3e_col,
  title      = "METABRIC: si3e hypoxia 1hr LFC1",
  save_name  = "mb_3e_hyp_lfc1_surv.pdf"
)
```

```{r mb_3e_hyp_lfc15}
zdf_3e_hyp_lfc15_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3e_hyp_lfc15, down_3e_hyp_lfc15, sig_3e_hyp$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3e_hyp_lfc15_mb,
  label_low  = "Low 3e (hypoxia)",
  label_high = "High 3e (hypoxia)",
  color_high = si3e_col,
  title      = "METABRIC: si3e hypoxia 1hr LFC1.5",
  save_name  = "mb_3e_hyp_lfc15_surv.pdf"
)
```

## METABRIC: eIF3d hypoxia 1hr signature

```{r mb_3d_hyp}
zdf_3d_hyp_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3d_hyp, down_3d_hyp, sig_3d_hyp$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3d_hyp_mb,
  label_low  = "Low 3d (hypoxia)",
  label_high = "High 3d (hypoxia)",
  color_high = si3d_col,
  title      = "METABRIC: si3d hypoxia 1hr LFC1",
  save_name  = "mb_3d_hyp_lfc1_surv.pdf"
)
```

```{r mb_3d_hyp_lfc15}
zdf_3d_hyp_lfc15_mb <- compute_sig_zscores(
  merged_zscores_mb, up_3d_hyp_lfc15, down_3d_hyp_lfc15, sig_3d_hyp$symbol
) %>% assign_quartiles_mb()

plot_surv(
  clin_mb, zdf_3d_hyp_lfc15_mb,
  label_low  = "Low 3d (hypoxia)",
  label_high = "High 3d (hypoxia)",
  color_high = si3d_col,
  title      = "METABRIC: si3d hypoxia 1hr LFC1.5",
  save_name  = "mb_3d_hyp_lfc15_surv.pdf"
)
```

---

# TCGA survival analysis

TCGA Firehose Legacy, downloaded from cBioPortal (1,108 samples).

```{r tcga_load, cache=TRUE}
zscores_tcga <- read.table(
  file.path(cbioportal_path, "brca_tcga/tcga_data_mrna_seq_v2_rsem_zscores_ref_diploid_samples.txt"),
  sep = "\t", header = TRUE
)

tmp_tcga  <- zscores_tcga %>% dplyr::select(-Entrez_Gene_Id)
mat_tcga  <- as.matrix(tmp_tcga[, -1])
genes_tcga <- tmp_tcga$Hugo_Symbol
rs_tcga   <- rowsum(mat_tcga, group = genes_tcga, na.rm = TRUE)
n_tcga    <- table(genes_tcga)
merged_zscores_tcga <- as.data.frame(rs_tcga / as.vector(n_tcga[rownames(rs_tcga)]))
```

```{r tcga_clin}
clin_tcga <- read_excel(
  file.path(cbioportal_path, "brca_tcga/new_tcga_clinical.xlsx")
) %>%
  dplyr::select(OS_STATUS, OS_MONTHS, PATIENT_ID) %>%
  filter(OS_MONTHS != "[Not Available]", OS_MONTHS > 0) %>%
  mutate(OS_MONTHS = as.numeric(OS_MONTHS))

clin_tcga$OS_STATUS[which(clin_tcga$OS_STATUS == "1:DECEASED" & clin_tcga$OS_MONTHS > 120)] <- "0:LIVING"
clin_tcga$OS_STATUS <- clin_tcga$OS_STATUS == "1:DECEASED"
clin_tcga$OS_YEARS  <- clin_tcga$OS_MONTHS / 12
clin_tcga$OS_YEARS[which(clin_tcga$OS_YEARS >= 10)] <- 10
```

## TCGA: eIF3e normoxia signature

```{r tcga_3e_norm}
zdf_3e_norm_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3e_norm, down_3e_norm, sig_3e_norm$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3e_norm_tcga,
  label_low  = "Low 3e (normoxia)",
  label_high = "High 3e (normoxia)",
  color_high = si3e_col,
  title      = "TCGA: si3e normoxia LFC1",
  save_name  = "tcga_3e_norm_lfc1_surv.pdf"
)
```

```{r tcga_3e_norm_lfc15}
zdf_3e_norm_lfc15_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3e_norm_lfc15, down_3e_norm_lfc15, sig_3e_norm$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3e_norm_lfc15_tcga,
  label_low  = "Low 3e (normoxia)",
  label_high = "High 3e (normoxia)",
  color_high = si3e_col,
  title      = "TCGA: si3e normoxia LFC1.5",
  save_name  = "tcga_3e_norm_lfc15_surv.pdf"
)
```

## TCGA: eIF3d normoxia signature

```{r tcga_3d_norm}
zdf_3d_norm_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3d_norm, down_3d_norm, sig_3d_norm$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3d_norm_tcga,
  label_low  = "Low 3d (normoxia)",
  label_high = "High 3d (normoxia)",
  color_high = si3d_col,
  title      = "TCGA: si3d normoxia LFC1",
  save_name  = "tcga_3d_norm_lfc1_surv.pdf"
)
```

```{r tcga_3d_norm_lfc15}
zdf_3d_norm_lfc15_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3d_norm_lfc15, down_3d_norm_lfc15, sig_3d_norm$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3d_norm_lfc15_tcga,
  label_low  = "Low 3d (normoxia)",
  label_high = "High 3d (normoxia)",
  color_high = si3d_col,
  title      = "TCGA: si3d normoxia LFC1.5",
  save_name  = "tcga_3d_norm_lfc15_surv.pdf"
)
```

## TCGA: eIF3e hypoxia 1hr signature

```{r tcga_3e_hyp}
zdf_3e_hyp_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3e_hyp, down_3e_hyp, sig_3e_hyp$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3e_hyp_tcga,
  label_low  = "Low 3e (hypoxia)",
  label_high = "High 3e (hypoxia)",
  color_high = si3e_col,
  title      = "TCGA: si3e hypoxia 1hr LFC1",
  save_name  = "tcga_3e_hyp_lfc1_surv.pdf"
)
```

```{r tcga_3e_hyp_lfc15}
zdf_3e_hyp_lfc15_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3e_hyp_lfc15, down_3e_hyp_lfc15, sig_3e_hyp$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3e_hyp_lfc15_tcga,
  label_low  = "Low 3e (hypoxia)",
  label_high = "High 3e (hypoxia)",
  color_high = si3e_col,
  title      = "TCGA: si3e hypoxia 1hr LFC1.5",
  save_name  = "tcga_3e_hyp_lfc15_surv.pdf"
)
```

## TCGA: eIF3d hypoxia 1hr signature

```{r tcga_3d_hyp}
zdf_3d_hyp_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3d_hyp, down_3d_hyp, sig_3d_hyp$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3d_hyp_tcga,
  label_low  = "Low 3d (hypoxia)",
  label_high = "High 3d (hypoxia)",
  color_high = si3d_col,
  title      = "TCGA: si3d hypoxia 1hr LFC1",
  save_name  = "tcga_3d_hyp_lfc1_surv.pdf"
)
```

```{r tcga_3d_hyp_lfc15}
zdf_3d_hyp_lfc15_tcga <- compute_sig_zscores(
  merged_zscores_tcga, up_3d_hyp_lfc15, down_3d_hyp_lfc15, sig_3d_hyp$symbol
) %>% assign_quartiles_tcga()

plot_surv(
  clin_tcga, zdf_3d_hyp_lfc15_tcga,
  label_low  = "Low 3d (hypoxia)",
  label_high = "High 3d (hypoxia)",
  color_high = si3d_col,
  title      = "TCGA: si3d hypoxia 1hr LFC1.5",
  save_name  = "tcga_3d_hyp_lfc15_surv.pdf"
)
```
