---
title: "Prepare Data for ORFquant"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(here)
library(tidyverse)
```

```{r load-metadata}
metadata_231 <- read_csv(here("accessories", "metadata_231_timecourse.csv"))
```

```{r inspect-metadata}
# View structure and summary
glimpse(metadata_231)
head(metadata_231)
```

## Generate BAM Merge Scripts for Cluster

Generate text files listing BAM paths for each group (replicates merged) and bsub scripts to run `samtools merge` on the cluster.

```{r define-cluster-paths}
# Define cluster paths
bam_base_path <- "/beevol/home/bartholk/kateData_ribo/2025_MDAcells_riboseq2/STAR"
cluster_base_path <- "/beevol/home/matlinka/eif3e_hypoxia/ORFquant/orfquant_hypoxia_timecourse/merge_bams"
merged_output_path <- file.path(cluster_base_path, "merged_bams")
bam_lists_path <- file.path(cluster_base_path, "bam_lists")
```

```{r create-output-directories}
# Create output directories
dir.create(here("cluster_scripts", "bam_lists"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("cluster_scripts", "merge_bams"), recursive = TRUE, showWarnings = FALSE)
```

```{r generate-merge-scripts}
# Track group names for master script
group_names <- c()

# Group by condition, treatment, hours and write files (RiboSeq only)
metadata_231 %>%
  filter(type == "RiboSeq") %>%
  mutate(bam_path = file.path(bam_base_path, paste0(id, ".bam"))) %>%
  group_by(condition, treatment, hours) %>%
  group_walk(~{
    group_name <- paste(.y$condition, .y$treatment, "RiboSeq", paste0(.y$hours, "h"), sep = "_")
    group_names <<- c(group_names, group_name)

    # Write BAM list file
    writeLines(.x$bam_path, here("cluster_scripts", "bam_lists", paste0(group_name, ".txt")))

    # Write bsub script
    bsub_content <- glue::glue('#!/bin/bash
#BSUB -q rna
#BSUB -R "select[mem>30] rusage[mem=30]"
#BSUB -R "span[hosts=1]"
#BSUB -o {cluster_base_path}/logs/merge_{group_name}_%J.log
#BSUB -e {cluster_base_path}/logs/merge_{group_name}_%J.out

module load samtools

samtools merge {merged_output_path}/{group_name}_merged.bam -b {bam_lists_path}/{group_name}.txt
')
    writeLines(bsub_content, here("cluster_scripts", "merge_bams", paste0("merge_", group_name, ".bsub")))
  })

# Report number of groups processed
cat("Generated scripts for", length(group_names), "groups\n")
```

```{r generate-master-submit-script}
# Write master submit script
submit_lines <- c("#!/bin/bash", "", paste0("cd ", cluster_base_path, "/merged_bams"), "")
submit_lines <- c(submit_lines, paste0("bsub < merge_", group_names, ".bsub"))
writeLines(submit_lines, here("cluster_scripts", "submit_all_merges.bsub"))

cat("Master submit script created at cluster_scripts/submit_all_merges.bsub\n")
```

```{r verify-outputs}
# Verify outputs
bam_list_files <- list.files(here("cluster_scripts", "bam_lists"), pattern = "\\.txt$")
bsub_files <- list.files(here("cluster_scripts", "merge_bams"), pattern = "\\.bsub$")
master_script_exists <- file.exists(here("cluster_scripts", "submit_all_merges.bsub"))

cat("BAM list files created:", length(bam_list_files), "\n")
cat("bsub scripts created:", length(bsub_files), "\n")
cat("Master submit script exists:", master_script_exists, "\n")

# Show example BAM list file
example_bam_list <- bam_list_files[1]
cat("\nExample BAM list file (", example_bam_list, "):\n", sep = "")
cat(readLines(here("cluster_scripts", "bam_lists", example_bam_list)), sep = "\n")

# Show example bsub script
example_bsub <- bsub_files[1]
cat("\nExample bsub script (", example_bsub, "):\n", sep = "")
cat(readLines(here("cluster_scripts", "merge_bams", example_bsub)), sep = "\n")
```

## Process BAM Files for RiboseQC

The merged BAM files from riboPipe use `hs_X` chromosome naming (e.g., hs_1, hs_2) but the GENCODE annotation uses `chr` naming (chr1, chr2). We need to:
1. Rename chromosomes from `hs_X` to `chr` format
2. Filter to main chromosomes only (chr1-22, X, Y, M) to match the annotation

```{r define-riboseqc-paths}
# Define RiboseQC cluster paths
riboseqc_base_path <- "/beevol/home/matlinka/eif3e_hypoxia/ORFquant/orfquant_hypoxia_timecourse/riboseqc"
gtf_file <- "/beevol/home/bartholk/riboPipe_comps/star_index/v45/gencode.v45.annotation.gtf.gz"
genome_seq <- "/beevol/home/bartholk/riboPipe_comps/star_index/v45/GRCh38.primary_assembly.genome.fa"
annotation_file <- file.path(riboseqc_base_path, "gencode.v45.annotation.gtf.gz_Rannot")
twobit_file <- file.path(riboseqc_base_path, "GRCh38.primary_assembly.genome.2bit")
bams_filtered_path <- file.path(riboseqc_base_path, "bams_filtered")
```

```{r create-riboseqc-directories}
# Create output directories for RiboseQC scripts
dir.create(here("cluster_scripts", "riboseqc", "r_scripts"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("cluster_scripts", "riboseqc", "bsub_scripts"), recursive = TRUE, showWarnings = FALSE)
```

```{r generate-2bit-conversion-script}
# Generate script to convert FASTA to 2bit format (run first on cluster)
convert_2bit_content <- glue::glue('#!/usr/bin/env bash
#BSUB -q rna
#BSUB -o {riboseqc_base_path}/logs/convert_2bit_%J.log
#BSUB -e {riboseqc_base_path}/logs/convert_2bit_%J.out

. /usr/share/Modules/init/bash
module load ucsc/v349

faToTwoBit /beevol/home/bartholk/riboPipe_comps/star_index/v45/GRCh38.primary_assembly.genome.fa \\
  {twobit_file}
')
writeLines(convert_2bit_content, here("cluster_scripts", "riboseqc", "convert_to_2bit.bsub"))
cat("Created convert_to_2bit.bsub\n")
```

```{r generate-annotation-prep-scripts}
# Generate R script to prepare RiboseQC annotation
prepare_annotation_r <- glue::glue('library(RiboseQC)

prepare_annotation_files(
  annotation_directory = "{riboseqc_base_path}",
  twobit_file = "{twobit_file}",
  gtf_file = "{gtf_file}",
  genome_seq = "{genome_seq}",
  scientific_name = "Homo.sapiens",
  annotation_name = "gencode.v45",
  export_bed_tables_TxDb = FALSE,
  forge_BSgenome = FALSE,
  create_TxDb = TRUE
)
')
writeLines(prepare_annotation_r, here("cluster_scripts", "riboseqc", "prepare_annotation.R"))

# Generate bsub script to run annotation preparation
prepare_annotation_bsub <- glue::glue('#!/usr/bin/env bash
#BSUB -n 10
#BSUB -q rna
#BSUB -R "select[mem>30] rusage[mem=30]"
#BSUB -R "span[hosts=1]"
#BSUB -o {riboseqc_base_path}/logs/prepare_annotation_%J.log
#BSUB -e {riboseqc_base_path}/logs/prepare_annotation_%J.out

. /usr/share/Modules/init/bash
source ~/.bashrc
conda activate working_orf

cd {riboseqc_base_path}
Rscript prepare_annotation.R
')
writeLines(prepare_annotation_bsub, here("cluster_scripts", "riboseqc", "prepare_annotation.bsub"))
cat("Created prepare_annotation.R and prepare_annotation.bsub\n")
```

## Process BAM Chromosomes

The riboPipe BAM files use `hs_X` chromosome naming but GENCODE uses `chr`. This script:
1. Renames chromosomes from hs_X to chr format
2. Filters to main chromosomes (chr1-22, X, Y, M) to avoid seqlevel mismatches

```{r generate-bam-processing-script}
# Generate script to process BAM chromosomes
bam_process_content <- glue::glue('#!/bin/bash
# Process merged BAMs: rename chromosomes and filter to main chroms
# Run this AFTER merging BAMs and BEFORE running RiboseQC

cd {riboseqc_base_path}
module load samtools

# Create output directories
mkdir -p bams_chr bams_filtered

# Create header with chr naming (from first merged BAM)
FIRST_BAM=$(ls {merged_output_path}/*_merged.bam | grep -v tmp | head -1)
samtools view -H $FIRST_BAM | sed "s/SN:hs_/SN:chr/" > new_header.sam

# Create filtered header with only main chromosomes
samtools view -H $FIRST_BAM | grep -v "^@SQ" > header_filtered.sam
samtools view -H $FIRST_BAM | grep "^@SQ" | sed "s/SN:hs_/SN:chr/" | grep -E "SN:chr([0-9]+|X|Y|M)\\b" >> header_filtered.sam

# Define main chromosomes
CHROMS="chr1 chr2 chr3 chr4 chr5 chr6 chr7 chr8 chr9 chr10 chr11 chr12 chr13 chr14 chr15 chr16 chr17 chr18 chr19 chr20 chr21 chr22 chrX chrY chrM"

# Process each merged BAM
for bam in {merged_output_path}/hypoxia_*_merged.bam {merged_output_path}/normoxia_*_merged.bam; do
  name=$(basename "$bam")
  echo "Processing $name..."

  # Rename chromosomes
  samtools reheader new_header.sam "$bam" > bams_chr/"$name"
  samtools index bams_chr/"$name"

  # Filter to main chromosomes and apply filtered header
  samtools view -h -b bams_chr/"$name" $CHROMS > bams_filtered/"$name".tmp
  samtools reheader header_filtered.sam bams_filtered/"$name".tmp > bams_filtered/"$name"
  rm bams_filtered/"$name".tmp
  samtools index bams_filtered/"$name"
done

echo "Done! Filtered BAMs are in bams_filtered/"
')
writeLines(bam_process_content, here("cluster_scripts", "riboseqc", "process_bam_chromosomes.sh"))
cat("Created process_bam_chromosomes.sh\n")
```

## Generate RiboseQC Scripts

```{r generate-riboseqc-scripts}
# Generate R and bsub scripts for each group
# Reuses group_names from the merge section above
# Uses filtered BAMs with chr naming and main chromosomes only

for (group_name in group_names) {
  # R script for RiboseQC analysis
  r_script_content <- glue::glue('library(devtools)
library("RiboseQC")

RiboseQC_analysis(
  annotation_file = "{annotation_file}",
  bam_files = "{bams_filtered_path}/{group_name}_merged.bam",
  report_file = "{group_name}_riboseqc.html",
  write_tmp_files = TRUE
)
')
  writeLines(r_script_content, here("cluster_scripts", "riboseqc", "r_scripts", paste0("riboseqc_", group_name, ".R")))

  # bsub script
  bsub_content <- glue::glue('#!/usr/bin/env bash
#BSUB -n 10
#BSUB -q rna
#BSUB -R "select[mem>30] rusage[mem=30]"
#BSUB -R "span[hosts=1]"
#BSUB -o {riboseqc_base_path}/logs/riboseqc_{group_name}_%J.log
#BSUB -e {riboseqc_base_path}/logs/riboseqc_{group_name}_%J.out

. /usr/share/Modules/init/bash
source ~/.bashrc
conda activate working_orf

cd {riboseqc_base_path}
Rscript r_scripts/riboseqc_{group_name}.R
')
  writeLines(bsub_content, here("cluster_scripts", "riboseqc", "bsub_scripts", paste0("riboseqc_", group_name, ".bsub")))
}

cat("Generated RiboseQC scripts for", length(group_names), "groups\n")
```

```{r generate-riboseqc-master-submit}
# Write master submit script for all RiboseQC jobs
submit_lines <- c(
  "#!/bin/bash",
  "",
  paste0("cd ", riboseqc_base_path),
  ""
)
submit_lines <- c(submit_lines, paste0("bsub < bsub_scripts/riboseqc_", group_names, ".bsub"))
writeLines(submit_lines, here("cluster_scripts", "riboseqc", "submit_all_riboseqc.bsub"))

cat("Master submit script created at cluster_scripts/riboseqc/submit_all_riboseqc.bsub\n")
```

```{r verify-riboseqc-outputs}
# Verify RiboseQC outputs
r_script_files <- list.files(here("cluster_scripts", "riboseqc", "r_scripts"), pattern = "\\.R$")
riboseqc_bsub_files <- list.files(here("cluster_scripts", "riboseqc", "bsub_scripts"), pattern = "\\.bsub$")
convert_script_exists <- file.exists(here("cluster_scripts", "riboseqc", "convert_to_2bit.bsub"))
prep_annotation_exists <- file.exists(here("cluster_scripts", "riboseqc", "prepare_annotation.R"))
riboseqc_master_exists <- file.exists(here("cluster_scripts", "riboseqc", "submit_all_riboseqc.bsub"))

cat("RiboseQC R scripts created:", length(r_script_files), "\n")
cat("RiboseQC bsub scripts created:", length(riboseqc_bsub_files), "\n")
cat("convert_to_2bit.bsub exists:", convert_script_exists, "\n")
cat("prepare_annotation.R exists:", prep_annotation_exists, "\n")
cat("Master submit script exists:", riboseqc_master_exists, "\n")

# Show example R script
example_r_script <- r_script_files[1]
cat("\nExample R script (", example_r_script, "):\n", sep = "")
cat(readLines(here("cluster_scripts", "riboseqc", "r_scripts", example_r_script)), sep = "\n")

# Show example bsub script
example_riboseqc_bsub <- riboseqc_bsub_files[1]
cat("\nExample bsub script (", example_riboseqc_bsub, "):\n", sep = "")
cat(readLines(here("cluster_scripts", "riboseqc", "bsub_scripts", example_riboseqc_bsub)), sep = "\n")
```

## Execution Order on Cluster

Run these steps in order:

1. **Merge BAMs**: `bash submit_all_merges.bsub`
2. **Convert FASTA to 2bit**: `bsub < convert_to_2bit.bsub`
3. **Prepare annotation**: `bsub < prepare_annotation.bsub` (wait for 2bit to complete)
4. **Process BAM chromosomes**: `bash process_bam_chromosomes.sh` (wait for merges to complete)
5. **Run RiboseQC**: `bash submit_all_riboseqc.bsub` (wait for annotation and BAM processing)

## Generate ORFquant Scripts

After RiboseQC completes, run ORFquant for translation quantification.

```{r define-orfquant-paths}
# Define ORFquant cluster paths
orfquant_base_path <- "/beevol/home/matlinka/eif3e_hypoxia/ORFquant/orfquant_hypoxia_timecourse/orfquant"
```

```{r create-orfquant-directories}
# Create output directories for ORFquant scripts
dir.create(here("cluster_scripts", "orfquant", "r_scripts"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("cluster_scripts", "orfquant", "bsub_scripts"), recursive = TRUE, showWarnings = FALSE)
```

```{r generate-orfquant-scripts}
# Generate R and bsub scripts for each group

for (group_name in group_names) {
  # R script for ORFquant analysis
  r_script_content <- glue::glue('library(ORFquant)

run_ORFquant(
  for_ORFquant_file = "{bams_filtered_path}/{group_name}_merged.bam_for_ORFquant",
  annotation_file = "{annotation_file}",
  n_cores = 10,
  prefix = "{group_name}"
)
')
  writeLines(r_script_content, here("cluster_scripts", "orfquant", "r_scripts", paste0("orfquant_", group_name, ".R")))

  # bsub script
  bsub_content <- glue::glue('#!/usr/bin/env bash
#BSUB -n 10
#BSUB -q rna
#BSUB -R "select[mem>50] rusage[mem=50]"
#BSUB -R "span[hosts=1]"
#BSUB -o {orfquant_base_path}/logs/orfquant_{group_name}_%J.log
#BSUB -e {orfquant_base_path}/logs/orfquant_{group_name}_%J.out

. /usr/share/Modules/init/bash
source ~/.bashrc
conda activate working_orf

cd {orfquant_base_path}
Rscript r_scripts/orfquant_{group_name}.R
')
  writeLines(bsub_content, here("cluster_scripts", "orfquant", "bsub_scripts", paste0("orfquant_", group_name, ".bsub")))
}

cat("Generated ORFquant scripts for", length(group_names), "groups\n")
```

```{r generate-orfquant-master-submit}
# Write master submit script for all ORFquant jobs
submit_lines <- c(
  "#!/bin/bash",
  "",
  paste0("cd ", orfquant_base_path),
  ""
)
submit_lines <- c(submit_lines, paste0("bsub < bsub_scripts/orfquant_", group_names, ".bsub"))
writeLines(submit_lines, here("cluster_scripts", "orfquant", "submit_all_orfquant.bsub"))

cat("Master submit script created at cluster_scripts/orfquant/submit_all_orfquant.bsub\n")
```

```{r verify-orfquant-outputs}
# Verify ORFquant outputs
orfquant_r_files <- list.files(here("cluster_scripts", "orfquant", "r_scripts"), pattern = "\\.R$")
orfquant_bsub_files <- list.files(here("cluster_scripts", "orfquant", "bsub_scripts"), pattern = "\\.bsub$")

cat("ORFquant R scripts created:", length(orfquant_r_files), "\n")
cat("ORFquant bsub scripts created:", length(orfquant_bsub_files), "\n")
```
