---
title: "PROGENy Pathway Activity: si3d, si3e, and siCtrl Hypoxic TE Response"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

PROGENy pathway-activity analog of `gsea_si3d_vs_sictrl.Rmd`, `gsea_si3e_vs_sictrl.Rmd`, and
`gsea_hypoxic_response_kegg_go_reactome.Rmd`: same TE-ranked-list inputs and same
pathway x timepoint heatmap style, but scored against PROGENy's 14 canonical signaling
pathways instead of MSigDB Hallmark/KEGG/GO:BP/Reactome gene sets.

PROGENy pathways are weighted, signed gene footprints, not plain gene sets, so this notebook
does not use `fgsea`/`fora`. Instead, `decoupleR::run_mlm()` fits a multivariate linear model
of each condition's TE log2FC against the PROGENy weight matrix, producing a per-pathway
activity `score` (the multiple-regression t-statistic, directly analogous to fgsea's `NES`) and
a `p_value`, for every pathway in a single call per contrast.

`progeny::getModel()` is used instead of `decoupleR::get_progeny()` because the latter requires
the `OmnipathR` package (not installed) and a live network call to fetch the model from
omnipathdb.org; `progeny::getModel()` ships the same model data offline inside the `progeny`
package.

Three sections below mirror the three sibling notebooks' three TE contrasts: si3d vs sictrl,
si3e vs sictrl, and the siCtrl-only hypoxia-vs-normoxia baseline. Only one heatmap per contrast
is produced (PROGENy has no database-collection split), showing all 14 pathways always (no
significance-based row filtering, unlike the MSigDB heatmaps -- at 14 rows there's no
decluttering need).

# Load Libraries

```{r load_libraries}
library(progeny)
library(decoupleR)
library(pheatmap)
library(tidyverse)
library(here)

source(here("code", "functions.R"))
```

# Build PROGENy Network

```{r build_network}
progeny_net <- build_progeny_network(top = 100)

cat("PROGENy pathways:", length(unique(progeny_net$source)), "\n")
print(sort(unique(progeny_net$source)))

# Sanity check (CLAUDE.md "Integrating External Data from Papers" convention): confirm the
# Hypoxia pathway is present and its footprint includes canonical HIF target genes.
hypoxia_genes <- progeny_net %>% filter(source == "Hypoxia") %>% pull(target)
stopifnot(
  "Hypoxia pathway must be present in PROGENy model" = "Hypoxia" %in% progeny_net$source,
  "Hypoxia footprint must include canonical HIF target genes" =
    sum(c("VEGFA", "SLC2A1", "CA9", "PDK1") %in% hypoxia_genes) >= 3
)
cat("Hypoxia pathway spot-check passed:", sum(c("VEGFA", "SLC2A1", "CA9", "PDK1") %in% hypoxia_genes),
    "/ 4 canonical HIF targets found\n")
```

# Ranked-List Helper

Identical to `make_ranked()` in the sibling GSEA notebooks.

```{r ranked_list_helper}
make_ranked <- function(df) {
  df <- df %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(te_lfc), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranked <- df$te_lfc
  names(ranked) <- df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked[!duplicated(names(ranked))]
}
```

# PROGENy Heatmap Helper

Same visual style as the sibling notebooks' `make_gsea_heatmap()` (Okabe-Ito diverging color,
`cluster_rows = TRUE`, significance stars), adapted for `run_mlm()`'s output schema and always
showing all 14 pathways.

```{r progeny_heatmap_helper}
make_progeny_heatmap <- function(results, title, outfile, col_order) {
  wide <- results %>%
    dplyr::select(pathway, NES, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = NES) %>%
    column_to_rownames("pathway")
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- results %>%
    mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
    dplyr::select(pathway, timepoint, sig) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("pathway")
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(wide), na.rm = TRUE)
  max_abs <- max(max_abs, 0.1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("#0072B2", "white", "#D55E00"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = title,
    fontsize_row    = 9,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 16
  )

  pdf(outfile, width = 8, height = 6)
  print(p)
  dev.off()

  message("Saved: ", outfile)
  invisible(p)
}

# Runs decoupleR::run_mlm on a matrix built from the intersection of genes present across all
# supplied ranked-list conditions (avoids NA in the response matrix; PROGENy's model genes are
# a small (~1300-gene) subset of the assayed transcriptome, so intersecting across conditions
# still retains essentially all model genes -- verified: minimum per-pathway representation
# across si3d's 4 conditions is 17 genes, well above run_mlm's minsize=5 default).
run_progeny_mlm <- function(ranked_list, col_order) {
  common_genes <- Reduce(intersect, lapply(ranked_list, names))
  stopifnot(
    "must retain the large majority of PROGENy model genes after intersecting conditions" =
      sum(unique(progeny_net$target) %in% common_genes) / length(unique(progeny_net$target)) > 0.4
  )

  mat <- sapply(ranked_list, function(x) x[common_genes])
  rownames(mat) <- common_genes

  run_mlm(mat = mat, network = progeny_net, .source = source, .target = target,
          .mor = mor, minsize = 5) %>%
    dplyr::rename(pathway = source, NES = score, timepoint = condition) %>%
    group_by(timepoint) %>%
    mutate(padj = p.adjust(p_value, method = "fdr")) %>%
    ungroup() %>%
    mutate(timepoint = factor(timepoint, levels = col_order)) %>%
    dplyr::select(pathway, NES, p_value, padj, timepoint)
}
```

# si3d vs sictrl

```{r load_data_si3d}
te_si3d_norm  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
te_si3d_hyp1  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
te_si3d_hyp4  <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"))
te_si3d_hyp24 <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"))

col_order_kd <- c("Norm 1&4hr", "Hyp 1hr", "Hyp 4hr", "Hyp 24hr")

ranked_si3d <- list(
  "Norm 1&4hr" = make_ranked(te_si3d_norm),
  "Hyp 1hr"    = make_ranked(te_si3d_hyp1),
  "Hyp 4hr"    = make_ranked(te_si3d_hyp4),
  "Hyp 24hr"   = make_ranked(te_si3d_hyp24)
)

cat("si3d ranked list sizes:", paste(names(ranked_si3d), lengths(ranked_si3d), sep = "=", collapse = " | "), "\n")
```

```{r progeny_si3d}
progeny_si3d <- run_progeny_mlm(ranked_si3d, col_order_kd)

print(progeny_si3d %>% arrange(timepoint, pathway), n = Inf)

make_progeny_heatmap(
  progeny_si3d,
  title     = "PROGENy activity, siCtrl vs si3d (* = padj < 0.05)",
  outfile   = here("plots", "progeny_si3d_vs_sictrl.pdf"),
  col_order = col_order_kd
)
```

```{r positive_control_si3d}
progeny_si3d %>%
  filter(pathway == "Hypoxia") %>%
  arrange(timepoint) %>%
  print()
```

# si3e vs sictrl

Mirrors the si3d section above. Note the normoxia 1+4hr time-regressed file for si3e does not
follow the standard `translation_categories_` naming used everywhere else (including si3d's own
normoxia file) -- it's `te_si3e_vs_sictrl_normoxia_1and4hr.csv` (fewer columns: gene_id, te_lfc,
te_padj, symbol -- but that's all `make_ranked()` needs).

```{r load_data_si3e}
te_si3e_norm  <- read_csv(here("output", "te_si3e_vs_sictrl_normoxia_1and4hr.csv"))
te_si3e_hyp1  <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))
te_si3e_hyp4  <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"))
te_si3e_hyp24 <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"))

ranked_si3e <- list(
  "Norm 1&4hr" = make_ranked(te_si3e_norm),
  "Hyp 1hr"    = make_ranked(te_si3e_hyp1),
  "Hyp 4hr"    = make_ranked(te_si3e_hyp4),
  "Hyp 24hr"   = make_ranked(te_si3e_hyp24)
)

cat("si3e ranked list sizes:", paste(names(ranked_si3e), lengths(ranked_si3e), sep = "=", collapse = " | "), "\n")
```

```{r progeny_si3e}
progeny_si3e <- run_progeny_mlm(ranked_si3e, col_order_kd)

print(progeny_si3e %>% arrange(timepoint, pathway), n = Inf)

make_progeny_heatmap(
  progeny_si3e,
  title     = "PROGENy activity, siCtrl vs si3e (* = padj < 0.05)",
  outfile   = here("plots", "progeny_si3e_vs_sictrl.pdf"),
  col_order = col_order_kd
)
```

```{r positive_control_si3e}
progeny_si3e %>%
  filter(pathway == "Hypoxia") %>%
  arrange(timepoint) %>%
  print()
```

# siCtrl Hypoxia vs Normoxia (Baseline)

Mirrors `gsea_hypoxic_response_kegg_go_reactome.Rmd`: siCtrl-only hypoxia-vs-normoxia TE
response, no knockdown contrast. `translation_categories_{1,4,24}hr.csv` don't have a `symbol`
column pre-joined (unlike the si3d/si3e vs sictrl files above), so gene symbols are joined from
`gene_anno_hs_dm_v49_r111.tsv` first, exactly as the sibling notebook does.

```{r load_data_hypoxic_response}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, desc(type == "protein_coding")) %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup()

te_hyp_resp_1hr  <- read_csv(here("output", "translation_categories_1hr.csv"))  %>% left_join(gene_symbols, by = "gene_id")
te_hyp_resp_4hr  <- read_csv(here("output", "translation_categories_4hr.csv"))  %>% left_join(gene_symbols, by = "gene_id")
te_hyp_resp_24hr <- read_csv(here("output", "translation_categories_24hr.csv")) %>% left_join(gene_symbols, by = "gene_id")

col_order_baseline <- c("1hr", "4hr", "24hr")

ranked_hyp_resp <- list(
  "1hr"  = make_ranked(te_hyp_resp_1hr),
  "4hr"  = make_ranked(te_hyp_resp_4hr),
  "24hr" = make_ranked(te_hyp_resp_24hr)
)

cat("Hypoxic response ranked list sizes:", paste(names(ranked_hyp_resp), lengths(ranked_hyp_resp), sep = "=", collapse = " | "), "\n")
```

```{r progeny_hypoxic_response}
progeny_hyp_resp <- run_progeny_mlm(ranked_hyp_resp, col_order_baseline)

print(progeny_hyp_resp %>% arrange(timepoint, pathway), n = Inf)

make_progeny_heatmap(
  progeny_hyp_resp,
  title     = "PROGENy activity, siCtrl hypoxia vs normoxia (* = padj < 0.05)",
  outfile   = here("plots", "progeny_hypoxic_response.pdf"),
  col_order = col_order_baseline
)
```

## Positive Control: Hypoxia Pathway

**Important caveat on this whole notebook's TE-ranked PROGENy scores:** PROGENy's model weights
were fit on RNA-abundance perturbation signatures, so PROGENy pathway activity is fundamentally
an RNA-level readout. TE (ribo/rna ratio) is specifically *normalized against* RNA abundance, so
a pathway whose footprint is dominated by transcriptional induction (like `Hypoxia` itself) is
not expected to show up strongly at the TE level even when the underlying biology is real and
strong at the RNA level -- TE isolates the translational-efficiency layer on top of whatever RNA
changes already happened, which is a different (and by design largely orthogonal) signal.

Empirically here: `Hypoxia` pathway activity ranked by **TE** log2FC is not significant at any
timepoint (see table below), while the identical pipeline ranked by **RNA** log2FC instead gives
a very strong, highly significant positive score, confirming the network/scoring machinery itself
is correct and that the null TE-level result is a real, expected consequence of what TE measures
-- not a pipeline bug. This mirrors the sibling GSEA notebooks' `positive_control` chunks
(`HALLMARK_HYPOXIA` check), which are similarly descriptive, not asserted with `stopifnot`.

```{r positive_control_hypoxic_response}
hypoxia_check_te <- progeny_hyp_resp %>% filter(pathway == "Hypoxia") %>% arrange(timepoint)
cat("Hypoxia pathway activity, ranked by TE log2FC (this notebook's metric):\n")
print(hypoxia_check_te)

ranked_rna_1hr <- te_hyp_resp_1hr %>%
  filter(!is.na(symbol), !is.na(rna_lfc)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(rna_lfc), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  { setNames(.$rna_lfc, .$symbol) } %>%
  sort(decreasing = TRUE)
ranked_rna_1hr <- ranked_rna_1hr[!duplicated(names(ranked_rna_1hr))]

hypoxia_check_rna <- run_progeny_mlm(list("RNA 1hr" = ranked_rna_1hr), "RNA 1hr") %>%
  filter(pathway == "Hypoxia")
cat("\nPipeline sanity check -- same network/scoring, ranked by RNA log2FC instead of TE\n",
    "(RNA is expected to show strong Hypoxia activation; confirms the PROGENy network and\n",
    "run_mlm() call are working correctly):\n")
print(hypoxia_check_rna)

stopifnot(
  "pipeline sanity check: RNA-ranked Hypoxia activity must be significantly positive at 1hr (confirms scoring machinery works)" =
    hypoxia_check_rna$NES > 0 & hypoxia_check_rna$padj < 0.05
)
cat("\nPipeline sanity check passed (RNA-ranked Hypoxia activity is significantly positive, as expected)\n")
```
