---
title: "TE Distribution: Promotes-3D Geneset vs Control"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

## Overview

This notebook visualizes where genes in the "promotes eIF3d in hypoxia" geneset
fall within the global TE distribution at hypoxia 1hr. We compare the
promotes-3d set against a matched-size control set derived from genes
non-regulated in all 6 si3d vs sictrl comparisons.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)

set.seed(42)
```

---

## 1. Load TE data

```{r load-te}
te_all <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE)

te_hypoxia_1hr <- te_all %>%
  filter(condition == "hypoxia", treatment == "sictrl", hours == 1) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

message("Genes in hypoxia sictrl 1hr TE data: ", nrow(te_hypoxia_1hr))
message("Genes with valid log2FoldChange: ", sum(!is.na(te_hypoxia_1hr$log2FoldChange)))
```

---

## 2. Define genesets

### Promotes-3D set

```{r load-promotes}
promotes_genes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)

number_promotes <- nrow(promotes_genes)

message("Promotes-3d geneset size: ", nrow(promotes_genes))
message("Columns: ", paste(names(promotes_genes), collapse = ", "))
```

### Control pool from translation_categories

```{r control-pool}
tc_conditions <- c("normoxia_1hr", "normoxia_4hr", "normoxia_24hr",
                   "hypoxia_1hr",  "hypoxia_4hr",  "hypoxia_24hr")

tc_all <- map(tc_conditions, ~read_csv(
  here("output", paste0("translation_categories_si3d_vs_sictrl_", .x, ".csv")),
  show_col_types = FALSE
))

# Genes passing non-regulated criteria in each comparison
nonreg_per_file <- map(tc_all, ~.x %>%
  filter(abs(te_lfc) < 0.5, te_padj > 0.5) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  pull(ensembl_gene))

# Keep only genes that pass in ALL 6 comparisons
control_pool_ids <- Reduce(intersect, nonreg_per_file)
message("Non-regulated pool (all 6 comparisons): ", length(control_pool_ids), " genes")
```

```{r control-filter}
# Filter to genes present in TE data, with valid values, excluding promotes set
control_pool <- te_hypoxia_1hr %>%
  filter(
    ensembl_gene %in% control_pool_ids,
    !ensembl_gene %in% promotes_genes$ensembl_gene,
    !is.na(log2FoldChange)
  )

message("Control pool after TE data filter: ", nrow(control_pool), " genes")

# How many promotes genes have valid TE at hypoxia 1hr?
n_promotes <- te_hypoxia_1hr %>%
  filter(ensembl_gene %in% promotes_genes$ensembl_gene, !is.na(log2FoldChange)) %>%
  nrow()
message("Promotes genes with valid TE: ", n_promotes)
```

```{r control-sample}
# Check if pool is large enough; relax criteria if needed
if (nrow(control_pool) < n_promotes) {
  message("Pool too small (", nrow(control_pool), " < ", n_promotes,
          "). Relaxing to |te_lfc| < 0.5 only (dropping te_padj filter).")

  nonreg_relaxed <- map(tc_all, ~.x %>%
    filter(abs(te_lfc) < 0.5) %>%
    mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
    pull(ensembl_gene))

  control_pool_ids_relaxed <- Reduce(intersect, nonreg_relaxed)
  message("Relaxed non-regulated pool (all 6 comparisons): ",
          length(control_pool_ids_relaxed), " genes")

  control_pool <- te_hypoxia_1hr %>%
    filter(
      ensembl_gene %in% control_pool_ids_relaxed,
      !ensembl_gene %in% promotes_genes$ensembl_gene,
      !is.na(log2FoldChange)
    )
  message("Relaxed control pool after TE data filter: ", nrow(control_pool), " genes")
}

control_sample <- control_pool %>% slice_sample(n = n_promotes)
message("Control sample size: ", nrow(control_sample))
```

---

## 3. Verification checks

```{r verification}
# 1. Sizes match
stopifnot(nrow(control_sample) == n_promotes)
message("CHECK 1 PASSED: control_sample (", nrow(control_sample),
        ") == n_promotes (", n_promotes, ")")

# 2. No overlap between promotes and control
overlap <- intersect(promotes_genes$ensembl_gene, control_sample$ensembl_gene)
stopifnot(length(overlap) == 0)
message("CHECK 2 PASSED: No gene overlap between promotes and control sets")
```

---

## 4. Annotate TE data by group

```{r annotate}
te_annotated <- te_hypoxia_1hr %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_genes$ensembl_gene ~ "promotes_3d",
    ensembl_gene %in% control_sample$ensembl_gene ~ "control",
    TRUE                                           ~ "other"
  )) %>%
  filter(!is.na(log2FoldChange))

message("Group counts:")
# print(count(te_annotated, group))
```

---

## 5. Visualizations

### Plot 1 — Density overlay

```{r plot-density, fig.width=8, fig.height=5}
# Prepare foreground data (non-background groups)
te_fg <- te_annotated %>%
  filter(group != "other") %>%
  mutate(group = factor(group, levels = c("promotes_3d", "control")))

group_colors <- c(promotes_3d = "#2a9d8f", control = "#e76f51")
group_labels <- c(promotes_3d = paste0("Promotes-3D (n=", n_promotes, ")"),
                  control     = paste0("Control (n=", nrow(control_sample), ")"))

ggplot() +
  # Background: all genes
  geom_density(
    data = te_annotated,
    aes(x = log2FoldChange),
    fill = "grey80", color = "grey60", alpha = 0.6
  ) +
  # Foreground: promotes and control
  geom_density(
    data = te_fg,
    aes(x = log2FoldChange, color = group, fill = group),
    alpha = 0.35, linewidth = 0.9
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = group_colors, labels = group_labels) +
  scale_fill_manual(values  = group_colors, labels = group_labels) +
  labs(
    title    = "TE distribution: promotes-3D vs control (hypoxia, sictrl, 1hr)",
    subtitle = paste0("Background = all genes with valid TE (n=", nrow(te_annotated), ")"),
    x        = "Absolute TE (ribo vs rna)",
    y        = "Density",
    color    = NULL,
    fill     = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

### Plot 2 — Empirical CDF

```{r plot-ecdf, fig.width=8, fig.height=5}
# Include all three groups for ECDF
te_ecdf <- te_annotated %>%
  mutate(group = factor(group, levels = c("promotes_3d", "control", "other")))

ecdf_colors <- c(promotes_3d = "#e76f51", control = "#2a9d8f", other = "grey60")
ecdf_labels <- c(
  promotes_3d = paste0("Promotes-3D (n=", n_promotes, ")"),
  control     = paste0("Control (n=", nrow(control_sample), ")"),
  other       = paste0("All other genes (n=", sum(te_annotated$group == "other"), ")")
)

ggplot(te_ecdf, aes(x = log2FoldChange, color = group)) +
  stat_ecdf(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = ecdf_colors, labels = ecdf_labels) +
  labs(
    title    = "ECDF of TE: promotes-3D vs control (hypoxia, sictrl, 1hr)",
    x        = "Absolute TE (ribo vs rna)",
    y        = "Cumulative proportion",
    color    = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")+
  xlim(-4.5,2.5)

ggsave(here("output/3d_vs_sictrl_hyp.pdf"), width=5, height=4)
 ```

### Plot 3 — Summary statistics table

```{r summary-stats}
summary_stats <- te_annotated %>%
  group_by(group) %>%
  summarise(
    n          = n(),
    median_TE  = round(median(log2FoldChange), 3),
    IQR_TE     = round(IQR(log2FoldChange), 3),
    pct_TE_pos = round(100 * mean(log2FoldChange > 0), 1),
    .groups    = "drop"
  ) %>%
  arrange(desc(group == "promotes_3d"), desc(group == "control"))

knitr::kable(
  summary_stats,
  col.names = c("Group", "n", "Median TE", "IQR", "% TE > 0"),
  caption   = "Summary statistics per group (hypoxia sictrl 1hr)"
)
```

---

## 6. Absolute TE in normoxia sictrl 1hr

Are promotes-3D genes already distinctively translated at baseline, before any
hypoxia or knockdown? Here we use the ribo/rna ratio within normoxia sictrl 1hr
as an absolute TE measure.

### Filter normoxia sictrl 1hr

```{r norm-filter}
te_normoxia_1hr <- te_all %>%
  filter(condition == "normoxia", treatment == "sictrl", hours == 1) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

message("Genes in normoxia sictrl 1hr TE data: ", nrow(te_normoxia_1hr))
message("Genes with valid log2FoldChange: ", sum(!is.na(te_normoxia_1hr$log2FoldChange)))
```

### Annotate by group

```{r norm-annotate}
te_norm_annotated <- te_normoxia_1hr %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_genes$ensembl_gene ~ "promotes_3d",
    ensembl_gene %in% control_sample$ensembl_gene ~ "control",
    TRUE                                           ~ "other"
  )) %>%
  filter(!is.na(log2FoldChange))

message("promotes_3d in normoxia data: ",
        sum(te_norm_annotated$group == "promotes_3d"))
message("control in normoxia data: ",
        sum(te_norm_annotated$group == "control"))
```

### Plot 1 — Density overlay

```{r norm-density, fig.width=8, fig.height=5}
te_norm_fg <- te_norm_annotated %>%
  filter(group != "other") %>%
  mutate(group = factor(group, levels = c("promotes_3d", "control")))

norm_group_colors <- c(promotes_3d = "#2a9d8f", control = "#e76f51")
norm_group_labels <- c(
  promotes_3d = paste0("Promotes-3D (n=", sum(te_norm_annotated$group == "promotes_3d"), ")"),
  control     = paste0("Control (n=", sum(te_norm_annotated$group == "control"), ")")
)

ggplot() +
  geom_density(
    data = te_norm_annotated,
    aes(x = log2FoldChange),
    fill = "grey80", color = "grey60", alpha = 0.6
  ) +
  geom_density(
    data = te_norm_fg,
    aes(x = log2FoldChange, color = group, fill = group),
    alpha = 0.35, linewidth = 0.9
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = norm_group_colors, labels = norm_group_labels) +
  scale_fill_manual(values  = norm_group_colors, labels = norm_group_labels) +
  coord_cartesian(xlim = c(-4.5, 2.5)) +
  labs(
    title    = "TE distribution: promotes-3D vs control (normoxia, sictrl, 1hr)",
    subtitle = paste0("Background = all genes with valid TE (n=", nrow(te_norm_annotated), ")"),
    x        = "Absolute TE (ribo vs rna)",
    y        = "Density",
    color    = NULL,
    fill     = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

### Plot 2 — ECDF

```{r norm-ecdf, fig.width=8, fig.height=5}
te_norm_ecdf <- te_norm_annotated %>%
  mutate(group = factor(group, levels = c("promotes_3d", "control", "other")))

norm_ecdf_colors <- c(promotes_3d = "#e76f51", control = "#2a9d8f", other = "grey60")
norm_ecdf_labels <- c(
  promotes_3d = paste0("Promotes-3D (n=", sum(te_norm_annotated$group == "promotes_3d"), ")"),
  control     = paste0("Control (n=", sum(te_norm_annotated$group == "control"), ")"),
  other       = paste0("All other genes (n=", sum(te_norm_annotated$group == "other"), ")")
)

ggplot(te_norm_ecdf, aes(x = log2FoldChange, color = group)) +
  stat_ecdf(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = norm_ecdf_colors, labels = norm_ecdf_labels) +
  labs(
    title    = "ECDF of TE: promotes-3D vs control (normoxia, sictrl, 1hr)",
    x        = "Absolute TE (ribo vs rna)",
    y        = "Cumulative proportion",
    color    = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top") +
  xlim(-4.5, 2.5)

ggsave(here("output/3d_vs_sictrl_norm.pdf"), width = 5, height = 4)
```

### Plot 3 — Summary statistics table

```{r norm-summary-stats}
norm_summary_stats <- te_norm_annotated %>%
  group_by(group) %>%
  summarise(
    n          = n(),
    median_TE  = round(median(log2FoldChange), 3),
    IQR_TE     = round(IQR(log2FoldChange), 3),
    pct_TE_pos = round(100 * mean(log2FoldChange > 0), 1),
    .groups    = "drop"
  ) %>%
  arrange(desc(group == "promotes_3d"), desc(group == "control"))

knitr::kable(
  norm_summary_stats,
  col.names = c("Group", "n", "Median TE", "IQR", "% TE > 0"),
  caption   = "Summary statistics per group (normoxia sictrl 1hr)"
)
```


