---
title: "Translational Regulation Analysis - RNA/Ribo/TE Categorization"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

```{r load_gene_symbols}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"),
                      show_col_types = FALSE)

gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene symbols loaded:", nrow(gene_symbols), "unique genes\n")
```

# 1 Hour Analysis

## Prepare metadata for 1hr

```{r metadata_1hr}
metadata_1hr <- metadata %>%
  filter(treatment == "sictrl", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_1hr <- metadata_1hr$sample_name
allCounts_1hr <- allCounts[, sample_cols_1hr]
spikein_1hr <- spikein[, sample_cols_1hr]

cat("1hr samples:", length(sample_cols_1hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_1hr}
res_rna_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_1hr}
res_ribo_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_1hr}
res_te_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_1hr}
results_1hr <- categorize_translation_changes(
  res_rna = res_rna_1hr,
  res_ribo = res_ribo_1hr,
  res_te = res_te_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_1hr$summary_table)

# Show plot
print(results_1hr$plot + ggtitle("sictrl 1 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_1hr}
# Add padj values to categorized table
categorized_with_padj_1hr <- results_1hr$categorized_table %>%
  left_join(res_rna_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_1hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_1hr, here("output", "translation_categories_1hr.csv"))
write_csv(results_1hr$summary_table, here("output", "translation_summary_1hr.csv"))
cat("Saved 1hr categorization results to output/\n")
```

# 4 Hour Analysis

## Prepare metadata for 4hr

```{r metadata_4hr}
metadata_4hr <- metadata %>%
  filter(treatment == "sictrl", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_4hr <- metadata_4hr$sample_name
allCounts_4hr <- allCounts[, sample_cols_4hr]
spikein_4hr <- spikein[, sample_cols_4hr]

cat("4hr samples:", length(sample_cols_4hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_4hr}
res_rna_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_4hr}
res_ribo_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_4hr}
res_te_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_4hr}
results_4hr <- categorize_translation_changes(
  res_rna = res_rna_4hr,
  res_ribo = res_ribo_4hr,
  res_te = res_te_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_4hr$summary_table)

# Show plot
print(results_4hr$plot + ggtitle("sictrl 4 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_4hr}
# Add padj values to categorized table
categorized_with_padj_4hr <- results_4hr$categorized_table %>%
  left_join(res_rna_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_4hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_4hr, here("output", "translation_categories_4hr.csv"))
write_csv(results_4hr$summary_table, here("output", "translation_summary_4hr.csv"))
cat("Saved 4hr categorization results to output/\n")
```

# 24 Hour Analysis

## Prepare metadata for 24hr

```{r metadata_24hr}
metadata_24hr <- metadata %>%
  filter(treatment == "sictrl", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_24hr <- metadata_24hr$sample_name
allCounts_24hr <- allCounts[, sample_cols_24hr]
spikein_24hr <- spikein[, sample_cols_24hr]

cat("24hr samples:", length(sample_cols_24hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_24hr}
res_rna_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_24hr}
res_ribo_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_24hr}
res_te_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_24hr}
results_24hr <- categorize_translation_changes(
  res_rna = res_rna_24hr,
  res_ribo = res_ribo_24hr,
  res_te = res_te_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_24hr$summary_table)

# Show plot
print(results_24hr$plot + ggtitle("sictrl 24 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_24hr}
# Add padj values to categorized table
categorized_with_padj_24hr <- results_24hr$categorized_table %>%
  left_join(res_rna_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_24hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_24hr, here("output", "translation_categories_24hr.csv"))
write_csv(results_24hr$summary_table, here("output", "translation_summary_24hr.csv"))
cat("Saved 24hr categorization results to output/\n")
```


# Combined Comparison Across Timepoints

## Compare category distributions

```{r compare_timepoints}
# Combine summary tables
summary_combined <- bind_rows(
  results_1hr$summary_table %>% mutate(Timepoint = "1hr"),
  results_4hr$summary_table %>% mutate(Timepoint = "4hr"),
  results_24hr$summary_table %>% mutate(Timepoint = "24hr")
)

summary_combined$Timepoint <- factor(summary_combined$Timepoint, levels=c("1hr", "4hr", "24hr"))

# Bar plot of category counts
ggplot(summary_combined %>% filter(Category != "Not_sig"),
       aes(x = Timepoint, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Forwarded" = "blue", "Exclusive" = "red",
                               "Buffered" = "green", "TE_only" = "purple")) +
  theme_minimal() +
  labs(title = "Translational Regulation Categories Across Timepoints",
       y = "Number of Genes")

       ggsave(here("plots/combined_numbers_sictrl.pdf"), width=4, height=2)
```




# si3e Treatment Analysis

## si3e 1 Hour Analysis

### Prepare metadata for si3e 1hr

```{r metadata_si3e_1hr}
metadata_si3e_1hr <- metadata %>%
  filter(treatment == "si3e", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_1hr <- metadata_si3e_1hr$sample_name
allCounts_si3e_1hr <- allCounts[, sample_cols_si3e_1hr]
spikein_si3e_1hr <- spikein[, sample_cols_si3e_1hr]

cat("si3e 1hr samples:", length(sample_cols_si3e_1hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_1hr}
res_rna_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_1hr}
res_ribo_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_1hr}
res_te_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_1hr}
results_si3e_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_1hr,
  res_ribo = res_ribo_si3e_1hr,
  res_te = res_te_si3e_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_1hr$summary_table)

# Show plot
print(results_si3e_1hr$plot + ggtitle("si3e 1 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_1hr}
# Add padj values to categorized table
categorized_with_padj_si3e_1hr <- results_si3e_1hr$categorized_table %>%
  left_join(res_rna_si3e_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_1hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_1hr, here("output", "translation_categories_si3e_1hr.csv"))
write_csv(results_si3e_1hr$summary_table, here("output", "translation_summary_si3e_1hr.csv"))
cat("Saved si3e 1hr categorization results to output/\n")
```

## si3e 4 Hour Analysis

### Prepare metadata for si3e 4hr

```{r metadata_si3e_4hr}
metadata_si3e_4hr <- metadata %>%
  filter(treatment == "si3e", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_4hr <- metadata_si3e_4hr$sample_name
allCounts_si3e_4hr <- allCounts[, sample_cols_si3e_4hr]
spikein_si3e_4hr <- spikein[, sample_cols_si3e_4hr]

cat("si3e 4hr samples:", length(sample_cols_si3e_4hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_4hr}
res_rna_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_4hr}
res_ribo_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_4hr}
res_te_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_4hr}
results_si3e_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_4hr,
  res_ribo = res_ribo_si3e_4hr,
  res_te = res_te_si3e_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_4hr$summary_table)

# Show plot
print(results_si3e_4hr$plot + ggtitle("si3e 4 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_4hr}
# Add padj values to categorized table
categorized_with_padj_si3e_4hr <- results_si3e_4hr$categorized_table %>%
  left_join(res_rna_si3e_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_4hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_4hr, here("output", "translation_categories_si3e_4hr.csv"))
write_csv(results_si3e_4hr$summary_table, here("output", "translation_summary_si3e_4hr.csv"))
cat("Saved si3e 4hr categorization results to output/\n")
```

## si3e 24 Hour Analysis

### Prepare metadata for si3e 24hr

```{r metadata_si3e_24hr}
metadata_si3e_24hr <- metadata %>%
  filter(treatment == "si3e", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_24hr <- metadata_si3e_24hr$sample_name
allCounts_si3e_24hr <- allCounts[, sample_cols_si3e_24hr]
spikein_si3e_24hr <- spikein[, sample_cols_si3e_24hr]

cat("si3e 24hr samples:", length(sample_cols_si3e_24hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_24hr}
res_rna_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_24hr}
res_ribo_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_24hr}
res_te_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_24hr}
results_si3e_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_24hr,
  res_ribo = res_ribo_si3e_24hr,
  res_te = res_te_si3e_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_24hr$summary_table)

# Show plot
print(results_si3e_24hr$plot + ggtitle("si3e 24 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_24hr}
# Add padj values to categorized table
categorized_with_padj_si3e_24hr <- results_si3e_24hr$categorized_table %>%
  left_join(res_rna_si3e_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_24hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_24hr, here("output", "translation_categories_si3e_24hr.csv"))
write_csv(results_si3e_24hr$summary_table, here("output", "translation_summary_si3e_24hr.csv"))
cat("Saved si3e 24hr categorization results to output/\n")
```

# si3d Treatment Analysis

## si3d 1 Hour Analysis

### Prepare metadata for si3d 1hr

```{r metadata_si3d_1hr}
metadata_si3d_1hr <- metadata %>%
  filter(treatment == "si3d", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_1hr <- metadata_si3d_1hr$sample_name
allCounts_si3d_1hr <- allCounts[, sample_cols_si3d_1hr]
spikein_si3d_1hr <- spikein[, sample_cols_si3d_1hr]

cat("si3d 1hr samples:", length(sample_cols_si3d_1hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_1hr}
res_rna_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_1hr}
res_ribo_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_1hr}
res_te_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_1hr}
results_si3d_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_1hr,
  res_ribo = res_ribo_si3d_1hr,
  res_te = res_te_si3d_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_1hr$summary_table)

# Show plot
print(results_si3d_1hr$plot + ggtitle("si3d 1 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_1hr}
# Add padj values to categorized table
categorized_with_padj_si3d_1hr <- results_si3d_1hr$categorized_table %>%
  left_join(res_rna_si3d_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_1hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_1hr, here("output", "translation_categories_si3d_1hr.csv"))
write_csv(results_si3d_1hr$summary_table, here("output", "translation_summary_si3d_1hr.csv"))
cat("Saved si3d 1hr categorization results to output/\n")
```

## si3d 4 Hour Analysis

### Prepare metadata for si3d 4hr

```{r metadata_si3d_4hr}
metadata_si3d_4hr <- metadata %>%
  filter(treatment == "si3d", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_4hr <- metadata_si3d_4hr$sample_name
allCounts_si3d_4hr <- allCounts[, sample_cols_si3d_4hr]
spikein_si3d_4hr <- spikein[, sample_cols_si3d_4hr]

cat("si3d 4hr samples:", length(sample_cols_si3d_4hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_4hr}
res_rna_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_4hr}
res_ribo_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_4hr}
res_te_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_4hr}
results_si3d_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_4hr,
  res_ribo = res_ribo_si3d_4hr,
  res_te = res_te_si3d_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_4hr$summary_table)

# Show plot
print(results_si3d_4hr$plot + ggtitle("si3d 4 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_4hr}
# Add padj values to categorized table
categorized_with_padj_si3d_4hr <- results_si3d_4hr$categorized_table %>%
  left_join(res_rna_si3d_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_4hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_4hr, here("output", "translation_categories_si3d_4hr.csv"))
write_csv(results_si3d_4hr$summary_table, here("output", "translation_summary_si3d_4hr.csv"))
cat("Saved si3d 4hr categorization results to output/\n")
```

## si3d 24 Hour Analysis

### Prepare metadata for si3d 24hr

```{r metadata_si3d_24hr}
metadata_si3d_24hr <- metadata %>%
  filter(treatment == "si3d", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_24hr <- metadata_si3d_24hr$sample_name
allCounts_si3d_24hr <- allCounts[, sample_cols_si3d_24hr]
spikein_si3d_24hr <- spikein[, sample_cols_si3d_24hr]

cat("si3d 24hr samples:", length(sample_cols_si3d_24hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_24hr}
res_rna_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_24hr}
res_ribo_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_24hr}
res_te_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_24hr}
results_si3d_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_24hr,
  res_ribo = res_ribo_si3d_24hr,
  res_te = res_te_si3d_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_24hr$summary_table)

# Show plot
print(results_si3d_24hr$plot + ggtitle("si3d 24 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_24hr}
# Add padj values to categorized table
categorized_with_padj_si3d_24hr <- results_si3d_24hr$categorized_table %>%
  left_join(res_rna_si3d_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_24hr %>% dplyr::transmute(gene_id, te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_24hr, here("output", "translation_categories_si3d_24hr.csv"))
write_csv(results_si3d_24hr$summary_table, here("output", "translation_summary_si3d_24hr.csv"))
cat("Saved si3d 24hr categorization results to output/\n")
```

# Combined Faceted Plot: All Treatments and Timepoints


```{r combined_facet_plot}
# Combine all categorized tables with treatment and timepoint labels
combined_all <- bind_rows(
  results_1hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "1hr"),
  results_4hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "4hr"),
  results_24hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "24hr"),
  results_si3e_1hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "1hr"),
  results_si3e_4hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "4hr"),
  results_si3e_24hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "24hr"),
  results_si3d_1hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "1hr"),
  results_si3d_4hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "4hr"),
  results_si3d_24hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "24hr")
)

# Set factor levels for proper ordering
combined_all$treatment <- factor(combined_all$treatment, levels = c("sictrl", "si3e", "si3d"))
combined_all$timepoint <- factor(combined_all$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts <- combined_all %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(treatment, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")


### add symbol

combined_all2 <- combined_all |> 
left_join(gene_symbols)

####



# Create faceted plot with TE count annotations
combined_facet <- ggplot(combined_all2 %>% filter(category != "Not_sig"),
                         aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(treatment ~ timepoint) +
  geom_text(data = te_counts,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_facet)

ggsave(here("plots/combined_all_treatments_timepoints.pdf"), width=9, height=7)
```


```{r combined_facet_rna_ribo_counts}
# Combine all categorized tables with padj values for RNA and Ribo significance counts
combined_with_padj <- bind_rows(
  results_1hr$categorized_table %>%
    mutate(treatment = "sictrl", timepoint = "1hr") %>%
    left_join(res_rna_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_4hr$categorized_table %>%
    mutate(treatment = "sictrl", timepoint = "4hr") %>%
    left_join(res_rna_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_24hr$categorized_table %>%
    mutate(treatment = "sictrl", timepoint = "24hr") %>%
    left_join(res_rna_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3e_1hr$categorized_table %>%
    mutate(treatment = "si3e", timepoint = "1hr") %>%
    left_join(res_rna_si3e_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3e_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3e_4hr$categorized_table %>%
    mutate(treatment = "si3e", timepoint = "4hr") %>%
    left_join(res_rna_si3e_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3e_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3e_24hr$categorized_table %>%
    mutate(treatment = "si3e", timepoint = "24hr") %>%
    left_join(res_rna_si3e_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3e_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3d_1hr$categorized_table %>%
    mutate(treatment = "si3d", timepoint = "1hr") %>%
    left_join(res_rna_si3d_1hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3d_1hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3d_4hr$categorized_table %>%
    mutate(treatment = "si3d", timepoint = "4hr") %>%
    left_join(res_rna_si3d_4hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3d_4hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id"),
  results_si3d_24hr$categorized_table %>%
    mutate(treatment = "si3d", timepoint = "24hr") %>%
    left_join(res_rna_si3d_24hr %>% dplyr::transmute(gene_id, rna_padj = padj), by = "gene_id") %>%
    left_join(res_ribo_si3d_24hr %>% dplyr::transmute(gene_id, ribo_padj = padj), by = "gene_id")
)

# Set factor levels for proper ordering (si3d above si3e)
combined_with_padj$treatment <- factor(combined_with_padj$treatment, levels = c("sictrl", "si3d", "si3e"))
combined_with_padj$timepoint <- factor(combined_with_padj$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate RNA and Ribo significance counts (padj < 0.05) for genes shown in plot
rna_ribo_counts <- combined_with_padj %>%
  filter(category != "Not_sig") %>%
  group_by(treatment, timepoint) %>%
  summarise(
    rna_up = sum(rna_padj < 0.05 & rna_lfc > 0, na.rm = TRUE),
    rna_down = sum(rna_padj < 0.05 & rna_lfc < 0, na.rm = TRUE),
    ribo_up = sum(ribo_padj < 0.05 & ribo_lfc > 0, na.rm = TRUE),
    ribo_down = sum(ribo_padj < 0.05 & ribo_lfc < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("RNA: ", rna_up, " up, ", rna_down, " dn\n",
                        "Ribo: ", ribo_up, " up, ", ribo_down, " dn"))

# Create faceted plot with RNA/Ribo significance annotations
combined_facet_rna_ribo <- ggplot(combined_with_padj %>% filter(category != "Not_sig"),
                                   aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(treatment ~ timepoint) +
  geom_text(data = rna_ribo_counts,
            aes(x = -4, y = -4, label = label),
            inherit.aes = FALSE,
            size = 2.5,
            hjust = 0,
            vjust = 0,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_facet_rna_ribo)

ggsave(here("plots/combined_all_treatments_timepoints_rna_ribo_counts.pdf"), width=9, height=7)
```


```{r sictrl_scatter_with_barplot}
# Filter to sictrl only
sictrl_data <- combined_with_padj %>%
  filter(treatment == "sictrl")

# Calculate RNA and Ribo significance counts for sictrl
sictrl_counts <- sictrl_data %>%
  filter(category != "Not_sig") %>%
  group_by(timepoint) %>%
  summarise(
    rna_up = sum(rna_padj < 0.05 & rna_lfc > 0, na.rm = TRUE),
    rna_down = sum(rna_padj < 0.05 & rna_lfc < 0, na.rm = TRUE),
    ribo_up = sum(ribo_padj < 0.05 & ribo_lfc > 0, na.rm = TRUE),
    ribo_down = sum(ribo_padj < 0.05 & ribo_lfc < 0, na.rm = TRUE),
    .groups = "drop"
  )

# Create scatter plot (sictrl only, faceted by timepoint)
scatter_plot <- ggplot(sictrl_data %>% filter(category != "Not_sig"),
                       aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  coord_fixed(ratio = 1, xlim = c(-5, 5), ylim = c(-5, 5)) +
  facet_wrap(~ timepoint, nrow = 1) +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# Reshape counts for barplot (long format)
bar_data <- sictrl_counts %>%
  pivot_longer(cols = c(rna_up, rna_down, ribo_up, ribo_down),
               names_to = "type", values_to = "count") %>%
  mutate(
    assay = ifelse(grepl("rna", type), "RNA", "Ribo"),
    direction = ifelse(grepl("up", type), "Up", "Down")
  )

# Create barplot with counts on top (grouped by assay, colored by direction)
bar_data$assay <- factor(bar_data$assay, levels = c("RNA", "Ribo"))
bar_data$direction <- factor(bar_data$direction, levels = c("Up", "Down"))

bar_plot <- ggplot(bar_data, aes(x = assay, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = count), position = position_dodge(width = 0.8), vjust = -0.3, size = 3) +
  facet_wrap(~ timepoint, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = c("Up" = "firebrick", "Down" = "steelblue")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  xlab("") + ylab("Number of genes") +
  theme_cowplot() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        strip.text = element_blank())

# Combine plots using cowplot
combined_plot <- plot_grid(scatter_plot, bar_plot, ncol = 1, rel_heights = c(2, 1), align = "v")

print(combined_plot)

ggsave(here("plots/sictrl_scatter_with_barplot.pdf"), combined_plot, width = 9, height = 7)
```

```{r sictrl_rna_ribo_scatter_sig}
library(ggExtra)
library(ggprism)
library(grid)

sictrl_sig <- combined_with_padj %>%
  filter(treatment == "sictrl") %>%
  filter(rna_padj < 0.05 | ribo_padj < 0.05)

sictrl_sig$timepoint <- factor(sictrl_sig$timepoint, levels = c("1hr", "4hr", "24hr"))

tp_plots <- lapply(c("1hr", "4hr", "24hr"), function(tp) {
  df <- sictrl_sig[sictrl_sig$timepoint == tp, ]

  p <- ggplot(data = df, aes(x = rna_lfc, y = ribo_lfc)) +
    ylim(-5, 5) +
    xlim(-5, 5) +
    geom_hline(yintercept = 0, color = "grey") +
    geom_vline(xintercept = 0, color = "grey") +
    geom_point(alpha = 0.25) +
    ylab(bquote(Log[2] ~ "si-CTRL LFC" ~ "RIBO: hypoxia vs normoxia")) +
    xlab(bquote(Log[2] ~ "si-CTRL LFC" ~ "RNA: hypoxia vs normoxia")) +
    ggtitle(tp) +
    theme_prism()

  ggMarginal(p, type = "histogram",
             xparams = list(binwidth = 0.1, fill = "grey"),
             yparams = list(binwidth = 0.1, fill = "red"))
})

tp_plots

# Combine: capture each ggMarginal output as a grob then arrange side-by-side
grobs <- lapply(tp_plots, function(p) grid.grabExpr(print(p), warn = 0))

pdf(here("plots/sictrl_rna_ribo_scatter_sig.pdf"), width = 13.5, height = 5)
gridExtra::grid.arrange(grobs = grobs, nrow = 1)
dev.off()
```

```{r sictrl_sig_counts}
# Counts of sig up / down for RNA and Ribo per timepoint (sictrl, padj < 0.05)
sictrl_dir_counts <- combined_with_padj %>%
  filter(treatment == "sictrl") %>%
  group_by(timepoint) %>%
  summarise(
    RNA_Up    = sum(rna_padj  < 0.05 & rna_lfc  > 0, na.rm = TRUE),
    RNA_Down  = sum(rna_padj  < 0.05 & rna_lfc  < 0, na.rm = TRUE),
    Ribo_Up   = sum(ribo_padj < 0.05 & ribo_lfc > 0, na.rm = TRUE),
    Ribo_Down = sum(ribo_padj < 0.05 & ribo_lfc < 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

print(sictrl_dir_counts)

sictrl_dir_long <- sictrl_dir_counts %>%
  pivot_longer(-timepoint,
               names_to  = c("assay", "direction"),
               names_sep = "_",
               values_to = "count") %>%
  mutate(direction = factor(direction, levels = c("Up", "Down")),
         assay     = factor(assay, levels = c("RNA", "Ribo")))

ggplot(sictrl_dir_long, aes(x = timepoint, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3.5) +
  facet_wrap(~ assay, nrow = 1) +
  scale_fill_manual(values = c("Up" = "firebrick", "Down" = "steelblue")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Number of significant genes (padj < 0.05)", fill = NULL) +
  theme_cowplot() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

ggsave(here("plots/sictrl_sig_counts_barplot.pdf"), width = 7, height = 4)
```


```{r}
library(ggrepel)

genes_to_highlight <- c("TEAD1")

combined_facet2 <- combined_facet +
  geom_point(
    data = combined_all2 %>% filter(symbol %in% genes_to_highlight),
    aes(x = rna_lfc, y = ribo_lfc),
    color = "black", size = 2.5
  ) +
  geom_text_repel(
    data = combined_all2 %>% filter(symbol %in% genes_to_highlight),
    aes(label = symbol),
    size = 3,
    fontface = "bold",
    box.padding = 0.4,
    point.padding = 0.3,
    max.overlaps = Inf
  )

  combined_facet2

  ggsave(here("plots/combined_all_treatments_timepoints_highlighted gene.pdf"), width=9, height=7)


```

genes exclusive at 1 hr or 4 hr

```{r}

head(combined_all2)
combined_all3 <- combined_all2 |> 
  filter(timepoint=="1hr"| timepoint=="4hr") |> 
  filter(category=="Exclusive") |> 
  filter(te_lfc > 0) |> 
  filter(treatment=="sictrl") |> 
  pull(symbol)

```

