---
title: "Translational Regulation Analysis - RNA/Ribo/TE Categorization"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

# 1 Hour Analysis

## Prepare metadata for 1hr

```{r metadata_1hr}
metadata_1hr <- metadata %>%
  filter(treatment == "sictrl", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_1hr <- metadata_1hr$sample_name
allCounts_1hr <- allCounts[, sample_cols_1hr]
spikein_1hr <- spikein[, sample_cols_1hr]

cat("1hr samples:", length(sample_cols_1hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_1hr}
res_rna_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_1hr}
res_ribo_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_1hr}
res_te_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_1hr}
results_1hr <- categorize_translation_changes(
  res_rna = res_rna_1hr,
  res_ribo = res_ribo_1hr,
  res_te = res_te_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_1hr$summary_table)

# Show plot
print(results_1hr$plot + ggtitle("sictrl 1 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_1hr}
# Add padj values to categorized table
categorized_with_padj_1hr <- results_1hr$categorized_table %>%
  left_join(res_rna_1hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_1hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_1hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_1hr, here("output", "translation_categories_1hr.csv"))
write_csv(results_1hr$summary_table, here("output", "translation_summary_1hr.csv"))
cat("Saved 1hr categorization results to output/\n")
```

# 4 Hour Analysis

## Prepare metadata for 4hr

```{r metadata_4hr}
metadata_4hr <- metadata %>%
  filter(treatment == "sictrl", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_4hr <- metadata_4hr$sample_name
allCounts_4hr <- allCounts[, sample_cols_4hr]
spikein_4hr <- spikein[, sample_cols_4hr]

cat("4hr samples:", length(sample_cols_4hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_4hr}
res_rna_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_4hr}
res_ribo_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_4hr}
res_te_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_4hr}
results_4hr <- categorize_translation_changes(
  res_rna = res_rna_4hr,
  res_ribo = res_ribo_4hr,
  res_te = res_te_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_4hr$summary_table)

# Show plot
print(results_4hr$plot + ggtitle("sictrl 4 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_4hr}
# Add padj values to categorized table
categorized_with_padj_4hr <- results_4hr$categorized_table %>%
  left_join(res_rna_4hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_4hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_4hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_4hr, here("output", "translation_categories_4hr.csv"))
write_csv(results_4hr$summary_table, here("output", "translation_summary_4hr.csv"))
cat("Saved 4hr categorization results to output/\n")
```

# 24 Hour Analysis

## Prepare metadata for 24hr

```{r metadata_24hr}
metadata_24hr <- metadata %>%
  filter(treatment == "sictrl", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_24hr <- metadata_24hr$sample_name
allCounts_24hr <- allCounts[, sample_cols_24hr]
spikein_24hr <- spikein[, sample_cols_24hr]

cat("24hr samples:", length(sample_cols_24hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_24hr}
res_rna_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_24hr}
res_ribo_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_24hr}
res_te_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_24hr}
results_24hr <- categorize_translation_changes(
  res_rna = res_rna_24hr,
  res_ribo = res_ribo_24hr,
  res_te = res_te_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_24hr$summary_table)

# Show plot
print(results_24hr$plot + ggtitle("sictrl 24 hour hypoxia vs normoxia"))

ggsave(here("plots/sictrl_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

## Save results

```{r save_24hr}
# Add padj values to categorized table
categorized_with_padj_24hr <- results_24hr$categorized_table %>%
  left_join(res_rna_24hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_24hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_24hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_24hr, here("output", "translation_categories_24hr.csv"))
write_csv(results_24hr$summary_table, here("output", "translation_summary_24hr.csv"))
cat("Saved 24hr categorization results to output/\n")
```


# Combined Comparison Across Timepoints

## Compare category distributions

```{r compare_timepoints}
# Combine summary tables
summary_combined <- bind_rows(
  results_1hr$summary_table %>% mutate(Timepoint = "1hr"),
  results_4hr$summary_table %>% mutate(Timepoint = "4hr"),
  results_24hr$summary_table %>% mutate(Timepoint = "24hr")
)

summary_combined$Timepoint <- factor(summary_combined$Timepoint, levels=c("1hr", "4hr", "24hr"))

# Bar plot of category counts
ggplot(summary_combined %>% filter(Category != "Not_sig"),
       aes(x = Timepoint, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Forwarded" = "blue", "Exclusive" = "red",
                               "Buffered" = "green", "TE_only" = "purple")) +
  theme_minimal() +
  labs(title = "Translational Regulation Categories Across Timepoints",
       y = "Number of Genes")

       ggsave(here("plots/combined_numbers_sictrl.pdf"), width=4, height=2)
```




# si3e Treatment Analysis

## si3e 1 Hour Analysis

### Prepare metadata for si3e 1hr

```{r metadata_si3e_1hr}
metadata_si3e_1hr <- metadata %>%
  filter(treatment == "si3e", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_1hr <- metadata_si3e_1hr$sample_name
allCounts_si3e_1hr <- allCounts[, sample_cols_si3e_1hr]
spikein_si3e_1hr <- spikein[, sample_cols_si3e_1hr]

cat("si3e 1hr samples:", length(sample_cols_si3e_1hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_1hr}
res_rna_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_1hr}
res_ribo_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_1hr}
res_te_si3e_1hr <- deFunction(
  countsMatrix = allCounts_si3e_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_1hr}
results_si3e_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_1hr,
  res_ribo = res_ribo_si3e_1hr,
  res_te = res_te_si3e_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_1hr$summary_table)

# Show plot
print(results_si3e_1hr$plot + ggtitle("si3e 1 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_1hr}
# Add padj values to categorized table
categorized_with_padj_si3e_1hr <- results_si3e_1hr$categorized_table %>%
  left_join(res_rna_si3e_1hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_1hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_1hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_1hr, here("output", "translation_categories_si3e_1hr.csv"))
write_csv(results_si3e_1hr$summary_table, here("output", "translation_summary_si3e_1hr.csv"))
cat("Saved si3e 1hr categorization results to output/\n")
```

## si3e 4 Hour Analysis

### Prepare metadata for si3e 4hr

```{r metadata_si3e_4hr}
metadata_si3e_4hr <- metadata %>%
  filter(treatment == "si3e", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_4hr <- metadata_si3e_4hr$sample_name
allCounts_si3e_4hr <- allCounts[, sample_cols_si3e_4hr]
spikein_si3e_4hr <- spikein[, sample_cols_si3e_4hr]

cat("si3e 4hr samples:", length(sample_cols_si3e_4hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_4hr}
res_rna_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_4hr}
res_ribo_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_4hr}
res_te_si3e_4hr <- deFunction(
  countsMatrix = allCounts_si3e_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_4hr}
results_si3e_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_4hr,
  res_ribo = res_ribo_si3e_4hr,
  res_te = res_te_si3e_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_4hr$summary_table)

# Show plot
print(results_si3e_4hr$plot + ggtitle("si3e 4 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_4hr}
# Add padj values to categorized table
categorized_with_padj_si3e_4hr <- results_si3e_4hr$categorized_table %>%
  left_join(res_rna_si3e_4hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_4hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_4hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_4hr, here("output", "translation_categories_si3e_4hr.csv"))
write_csv(results_si3e_4hr$summary_table, here("output", "translation_summary_si3e_4hr.csv"))
cat("Saved si3e 4hr categorization results to output/\n")
```

## si3e 24 Hour Analysis

### Prepare metadata for si3e 24hr

```{r metadata_si3e_24hr}
metadata_si3e_24hr <- metadata %>%
  filter(treatment == "si3e", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3e_24hr <- metadata_si3e_24hr$sample_name
allCounts_si3e_24hr <- allCounts[, sample_cols_si3e_24hr]
spikein_si3e_24hr <- spikein[, sample_cols_si3e_24hr]

cat("si3e 24hr samples:", length(sample_cols_si3e_24hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3e_24hr}
res_rna_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3e_24hr}
res_ribo_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3e_24hr}
res_te_si3e_24hr <- deFunction(
  countsMatrix = allCounts_si3e_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3e_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3e_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3e_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3e_24hr}
results_si3e_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3e_24hr,
  res_ribo = res_ribo_si3e_24hr,
  res_te = res_te_si3e_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3e_24hr$summary_table)

# Show plot
print(results_si3e_24hr$plot + ggtitle("si3e 24 hour hypoxia vs normoxia"))

ggsave(here("plots/si3e_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3e_24hr}
# Add padj values to categorized table
categorized_with_padj_si3e_24hr <- results_si3e_24hr$categorized_table %>%
  left_join(res_rna_si3e_24hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3e_24hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3e_24hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3e_24hr, here("output", "translation_categories_si3e_24hr.csv"))
write_csv(results_si3e_24hr$summary_table, here("output", "translation_summary_si3e_24hr.csv"))
cat("Saved si3e 24hr categorization results to output/\n")
```

# si3d Treatment Analysis

## si3d 1 Hour Analysis

### Prepare metadata for si3d 1hr

```{r metadata_si3d_1hr}
metadata_si3d_1hr <- metadata %>%
  filter(treatment == "si3d", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_1hr <- metadata_si3d_1hr$sample_name
allCounts_si3d_1hr <- allCounts[, sample_cols_si3d_1hr]
spikein_si3d_1hr <- spikein[, sample_cols_si3d_1hr]

cat("si3d 1hr samples:", length(sample_cols_si3d_1hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_1hr}
res_rna_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_1hr}
res_ribo_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_1hr}
res_te_si3d_1hr <- deFunction(
  countsMatrix = allCounts_si3d_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_1hr}
results_si3d_1hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_1hr,
  res_ribo = res_ribo_si3d_1hr,
  res_te = res_te_si3d_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_1hr$summary_table)

# Show plot
print(results_si3d_1hr$plot + ggtitle("si3d 1 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_1hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_1hr}
# Add padj values to categorized table
categorized_with_padj_si3d_1hr <- results_si3d_1hr$categorized_table %>%
  left_join(res_rna_si3d_1hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_1hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_1hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_1hr, here("output", "translation_categories_si3d_1hr.csv"))
write_csv(results_si3d_1hr$summary_table, here("output", "translation_summary_si3d_1hr.csv"))
cat("Saved si3d 1hr categorization results to output/\n")
```

## si3d 4 Hour Analysis

### Prepare metadata for si3d 4hr

```{r metadata_si3d_4hr}
metadata_si3d_4hr <- metadata %>%
  filter(treatment == "si3d", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_4hr <- metadata_si3d_4hr$sample_name
allCounts_si3d_4hr <- allCounts[, sample_cols_si3d_4hr]
spikein_si3d_4hr <- spikein[, sample_cols_si3d_4hr]

cat("si3d 4hr samples:", length(sample_cols_si3d_4hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_4hr}
res_rna_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_4hr}
res_ribo_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_4hr}
res_te_si3d_4hr <- deFunction(
  countsMatrix = allCounts_si3d_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_4hr}
results_si3d_4hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_4hr,
  res_ribo = res_ribo_si3d_4hr,
  res_te = res_te_si3d_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_4hr$summary_table)

# Show plot
print(results_si3d_4hr$plot + ggtitle("si3d 4 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_4hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_4hr}
# Add padj values to categorized table
categorized_with_padj_si3d_4hr <- results_si3d_4hr$categorized_table %>%
  left_join(res_rna_si3d_4hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_4hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_4hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_4hr, here("output", "translation_categories_si3d_4hr.csv"))
write_csv(results_si3d_4hr$summary_table, here("output", "translation_summary_si3d_4hr.csv"))
cat("Saved si3d 4hr categorization results to output/\n")
```

## si3d 24 Hour Analysis

### Prepare metadata for si3d 24hr

```{r metadata_si3d_24hr}
metadata_si3d_24hr <- metadata %>%
  filter(treatment == "si3d", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_si3d_24hr <- metadata_si3d_24hr$sample_name
allCounts_si3d_24hr <- allCounts[, sample_cols_si3d_24hr]
spikein_si3d_24hr <- spikein[, sample_cols_si3d_24hr]

cat("si3d 24hr samples:", length(sample_cols_si3d_24hr), "\n")
```

### Run RNA analysis

```{r rna_analysis_si3d_24hr}
res_rna_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run Ribo analysis

```{r ribo_analysis_si3d_24hr}
res_ribo_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Run TE analysis

```{r te_analysis_si3d_24hr}
res_te_si3d_24hr <- deFunction(
  countsMatrix = allCounts_si3d_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_si3d_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_si3d_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_si3d_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

### Categorize genes

```{r categorize_si3d_24hr}
results_si3d_24hr <- categorize_translation_changes(
  res_rna = res_rna_si3d_24hr,
  res_ribo = res_ribo_si3d_24hr,
  res_te = res_te_si3d_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.0
)

# Display summary
print(results_si3d_24hr$summary_table)

# Show plot
print(results_si3d_24hr$plot + ggtitle("si3d 24 hour hypoxia vs normoxia"))

ggsave(here("plots/si3d_24hr_norm_to_hyp.pdf"), width=4.5, height=3)
```

### Save results

```{r save_si3d_24hr}
# Add padj values to categorized table
categorized_with_padj_si3d_24hr <- results_si3d_24hr$categorized_table %>%
  left_join(res_rna_si3d_24hr %>% select(gene_id, padj) %>% rename(rna_padj = padj), by = "gene_id") %>%
  left_join(res_ribo_si3d_24hr %>% select(gene_id, padj) %>% rename(ribo_padj = padj), by = "gene_id") %>%
  left_join(res_te_si3d_24hr %>% select(gene_id, padj) %>% rename(te_padj = padj), by = "gene_id")

write_csv(categorized_with_padj_si3d_24hr, here("output", "translation_categories_si3d_24hr.csv"))
write_csv(results_si3d_24hr$summary_table, here("output", "translation_summary_si3d_24hr.csv"))
cat("Saved si3d 24hr categorization results to output/\n")
```

# Combined Faceted Plot: All Treatments and Timepoints


```{r combined_facet_plot}
# Combine all categorized tables with treatment and timepoint labels
combined_all <- bind_rows(
  results_1hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "1hr"),
  results_4hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "4hr"),
  results_24hr$categorized_table %>% mutate(treatment = "sictrl", timepoint = "24hr"),
  results_si3e_1hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "1hr"),
  results_si3e_4hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "4hr"),
  results_si3e_24hr$categorized_table %>% mutate(treatment = "si3e", timepoint = "24hr"),
  results_si3d_1hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "1hr"),
  results_si3d_4hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "4hr"),
  results_si3d_24hr$categorized_table %>% mutate(treatment = "si3d", timepoint = "24hr")
)

# Set factor levels for proper ordering
combined_all$treatment <- factor(combined_all$treatment, levels = c("sictrl", "si3e", "si3d"))
combined_all$timepoint <- factor(combined_all$timepoint, levels = c("1hr", "4hr", "24hr"))

# Calculate TE change counts for annotations
# Note: Categories Exclusive, Buffered, and TE_only guarantee all genes have:
#   - TE padj < 0.05 (significance threshold)
#   - abs(te_lfc) > 0.0 (LFC threshold from lfc_cutoff parameter)
# These criteria are enforced by categorize_translation_changes()
te_counts <- combined_all %>%
  filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
  group_by(treatment, timepoint) %>%
  summarise(
    up = sum(te_lfc > 0, na.rm = TRUE),      # TE up-regulated (increased translation efficiency)
    down = sum(te_lfc < 0, na.rm = TRUE),    # TE down-regulated (decreased translation efficiency)
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Up: ", up, "\n", "Down: ", down))

# Define colors
myCols <- c("Not_sig" = "grey80",
            "Forwarded" = "blue",
            "Exclusive" = "red",
            "Buffered" = "green",
            "TE_only" = "purple")


### add symbol

combined_all2 <- combined_all |> 
left_join(gene_symbols)

####



# Create faceted plot with TE count annotations
combined_facet <- ggplot(combined_all2 %>% filter(category != "Not_sig"),
                         aes(x = rna_lfc, y = ribo_lfc, color = category)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = myCols) +
  xlim(-5, 5) + ylim(-5, 5) +
  facet_grid(treatment ~ timepoint) +
  geom_text(data = te_counts,
            aes(x = -4, y = 4, label = label),
            inherit.aes = FALSE,
            size = 3,
            hjust = 0,
            vjust = 1,
            color = "black",
            fontface = "bold") +
  xlab("RNA-seq log2FC") +
  ylab("Ribo-seq log2FC") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

print(combined_facet)

ggsave(here("plots/combined_all_treatments_timepoints.pdf"), width=9, height=7)
```


```{r}
library(ggrepel)

genes_to_highlight <- c("TEAD1")

combined_facet2 <- combined_facet +
  geom_point(
    data = combined_all2 %>% filter(symbol %in% genes_to_highlight),
    aes(x = rna_lfc, y = ribo_lfc),
    color = "black", size = 2.5
  ) +
  geom_text_repel(
    data = combined_all2 %>% filter(symbol %in% genes_to_highlight),
    aes(label = symbol),
    size = 3,
    fontface = "bold",
    box.padding = 0.4,
    point.padding = 0.3,
    max.overlaps = Inf
  )

  combined_facet2

  ggsave(here("plots/combined_all_treatments_timepoints_highlighted gene.pdf"), width=9, height=7)


```

genes exclusive at 1 hr or 4 hr

```{r}

head(combined_all2)
combined_all3 <- combined_all2 |> 
  filter(timepoint=="1hr"| timepoint=="4hr") |> 
  filter(category=="Exclusive") |> 
  filter(te_lfc > 0) |> 
  filter(treatment=="sictrl") |> 
  pull(symbol)

```

