---
title: "Translational Regulation Analysis - RNA/Ribo/TE Categorization"
author: "Kate"
date: "2025-11-30"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(here)
library(janitor)
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(DT)
library(writexl)
library(cowplot)
```

```{r load_functions}
source(here("code/functions.R"))
cat("Loaded functions: createNormMatrix, deFunction, categorize_translation_changes\n")
```

# Load Data from timecourse_qc Analysis

```{r load_data}
# Load count matrices from timecourse_qc analysis
allCounts <- read_csv(here("output", "human_allCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

normMatrix <- read_csv(here("output", "normalizedCounts.csv")) %>%
  column_to_rownames("gene_id")

spikein <- read_csv(here("output", "dros_spikeInCounts_raw.csv")) %>%
  column_to_rownames("gene_id")

top500_ids <- spikein %>%
  rownames_to_column("gene_id") %>%
  pull(gene_id)

# Load metadata
metadata <- read_csv(here("output", "metadata_clean.csv"))

cat("Data loaded successfully\n")
cat("allCounts:", dim(allCounts), "\n")
cat("normMatrix:", dim(normMatrix), "\n")
cat("Spike-in genes:", length(top500_ids), "\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

# 1 Hour Analysis

## Prepare metadata for 1hr

```{r metadata_1hr}
metadata_1hr <- metadata %>%
  filter(treatment == "sictrl", hours == 1) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_1hr <- metadata_1hr$sample_name
allCounts_1hr <- allCounts[, sample_cols_1hr]
spikein_1hr <- spikein[, sample_cols_1hr]

cat("1hr samples:", length(sample_cols_1hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_1hr}
res_rna_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_1hr}
res_ribo_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_1hr}
res_te_1hr <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_1hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_1hr}
results_1hr <- categorize_translation_changes(
  res_rna = res_rna_1hr,
  res_ribo = res_ribo_1hr,
  res_te = res_te_1hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58496250072
)

# Display summary
print(results_1hr$summary_table)

# Show plot
print(results_1hr$plot + ggtitle("1 Hour: Translational Regulation Categories"))
```

## Save results

```{r save_1hr}
write_csv(results_1hr$categorized_table, here("output", "translation_categories_1hr.csv"))
write_csv(results_1hr$summary_table, here("output", "translation_summary_1hr.csv"))
cat("Saved 1hr categorization results to output/\n")
```

# 4 Hour Analysis

## Prepare metadata for 4hr

```{r metadata_4hr}
metadata_4hr <- metadata %>%
  filter(treatment == "sictrl", hours == 4) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_4hr <- metadata_4hr$sample_name
allCounts_4hr <- allCounts[, sample_cols_4hr]
spikein_4hr <- spikein[, sample_cols_4hr]

cat("4hr samples:", length(sample_cols_4hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_4hr}
res_rna_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_4hr}
res_ribo_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_4hr}
res_te_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_4hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_4hr}
results_4hr <- categorize_translation_changes(
  res_rna = res_rna_4hr,
  res_ribo = res_ribo_4hr,
  res_te = res_te_4hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58496250072
)

# Display summary
print(results_4hr$summary_table)

# Show plot
print(results_4hr$plot + ggtitle("4 Hours: Translational Regulation Categories"))
```

## Save results

```{r save_4hr}
write_csv(results_4hr$categorized_table, here("output", "translation_categories_4hr.csv"))
write_csv(results_4hr$summary_table, here("output", "translation_summary_4hr.csv"))
cat("Saved 4hr categorization results to output/\n")
```

# 24 Hour Analysis

## Prepare metadata for 24hr

```{r metadata_24hr}
metadata_24hr <- metadata %>%
  filter(treatment == "sictrl", hours == 24) %>%
  mutate(comparison_variable = condition)

# Create timepoint-specific count matrices
sample_cols_24hr <- metadata_24hr$sample_name
allCounts_24hr <- allCounts[, sample_cols_24hr]
spikein_24hr <- spikein[, sample_cols_24hr]

cat("24hr samples:", length(sample_cols_24hr), "\n")
```

## Run RNA analysis

```{r rna_analysis_24hr}
res_rna_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RNA"
)

cat("RNA analysis complete. Significant genes (padj < 0.05):",
    sum(res_rna_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run Ribo analysis

```{r ribo_analysis_24hr}
res_ribo_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "RIBO"
)

cat("Ribo analysis complete. Significant genes (padj < 0.05):",
    sum(res_ribo_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Run TE analysis

```{r te_analysis_24hr}
res_te_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = normMatrix,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

cat("TE analysis complete. Significant genes (padj < 0.05):",
    sum(res_te_24hr$padj < 0.05, na.rm = TRUE), "\n")
```

## Categorize genes

```{r categorize_24hr}
results_24hr <- categorize_translation_changes(
  res_rna = res_rna_24hr,
  res_ribo = res_ribo_24hr,
  res_te = res_te_24hr,
  padj_cutoff = 0.05,
  lfc_cutoff = 0.58496250072
)

# Display summary
print(results_24hr$summary_table)

# Show plot
print(results_24hr$plot + ggtitle("24 Hours: Translational Regulation Categories"))
```

## Save results

```{r save_24hr}
write_csv(results_24hr$categorized_table, here("output", "translation_categories_24hr.csv"))
write_csv(results_24hr$summary_table, here("output", "translation_summary_24hr.csv"))
cat("Saved 24hr categorization results to output/\n")
```

# Combined Comparison Across Timepoints

## Compare category distributions

```{r compare_timepoints}
# Combine summary tables
summary_combined <- bind_rows(
  results_1hr$summary_table %>% mutate(Timepoint = "1hr"),
  results_4hr$summary_table %>% mutate(Timepoint = "4hr"),
  results_24hr$summary_table %>% mutate(Timepoint = "24hr")
)

# Bar plot of category counts
ggplot(summary_combined %>% filter(Category != "Not_sig"),
       aes(x = Timepoint, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Forwarded" = "blue", "Exclusive" = "red",
                               "Buffered" = "green", "TE_only" = "purple")) +
  theme_minimal() +
  labs(title = "Translational Regulation Categories Across Timepoints",
       y = "Number of Genes")
```

## Create combined plots

```{r combined_plots}
p1 <- results_1hr$plot + ggtitle("1 Hour")
p2 <- results_4hr$plot + ggtitle("4 Hours")
p3 <- results_24hr$plot + ggtitle("24 Hours")

plot_grid(p1, p2, p3, ncol = 3, align = "hv")
```

## Display combined summary table

```{r combined_summary}
datatable(
  summary_combined,
  caption = "Summary of Translational Regulation Categories Across All Timepoints",
  options = list(pageLength = 15)
)
```
