---
title: "Teleman 4E Resistance Overlap with MDA-MB-231 Timecourse"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

Compare genes resistant to eIF4E inhibition (Teleman RiboSeq dataset) with genes upregulated by eIF3d in MDA-MB-231 cells across a hypoxia timecourse (1hr, 4hr, 24hr).

**Reference analysis**: `/Users/katematlin/github/mpra_preliminary/code/resistant_4e_overlap2.Rmd`

**Key question**: Do genes that maintain translation when eIF4E is inhibited overlap with genes whose translation is promoted by eIF3d during hypoxic stress?

## Load Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(DESeq2)
library(janitor)
library(ggVennDiagram)
library(cowplot)
library(GeneOverlap)
library(rstatix)
library(VennDiagram)
library(ggvenn)
library(writexl)
library(readxl)
library(cowplot)
library(ggpubr)
```

## Read in Teleman riboseq data
This was analyzed by me previously
teleman riboseq = res riboseq symbol
```{r}
res_riboseq_symbol <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv")
```


## Define 4E-Resistant Genes

```{r define-resistant-4e}
# Genes with positive log2FC and padj < 0.05
# Positive = resistant to 4E inhibition (still translated when eIF4E is inhibited)
resistant_4E <- res_riboseq_symbol %>%
  filter(log2FoldChange > 1) %>%
  filter(padj < 0.05) %>%
  filter(!is.na(symbol))

cat("Number of 4E-resistant genes (before deduplication):", nrow(resistant_4E), "\n")

# Check for duplicates
n_duplicates <- sum(duplicated(resistant_4E$symbol))
cat("Duplicate gene symbols:", n_duplicates, "\n")

# Deduplicate based on highest baseMean (keep gene with highest expression)
if(n_duplicates > 0) {
  resistant_4E <- resistant_4E %>%
    group_by(symbol) %>%
    slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
    ungroup()

  cat("After deduplication:", nrow(resistant_4E), "genes\n")
}

# Create gene set for overlap analysis - use pull() to extract vector
resistant_set <- data.frame(
  gs_name = "Resistant_4E",
  gene = resistant_4E %>% pull(symbol)
)

# Show top resistant genes
resistant_4E %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(symbol, log2FoldChange, padj, baseMean) %>%
  head(20)
```

# Load Timecourse Data

## Import eIF3d/3e Data - 1hr Timepoint

```{r import-1hr-data}
hyp_3d_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "1hr")

norm_3d_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "1hr")

hyp_3e_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "1hr")

norm_3e_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "1hr")

cat("1hr data loaded:\n")
cat("  hyp_3d_1hr:", nrow(hyp_3d_1hr), "genes\n")
cat("  norm_3d_1hr:", nrow(norm_3d_1hr), "genes\n")
cat("  hyp_3e_1hr:", nrow(hyp_3e_1hr), "genes\n")
cat("  norm_3e_1hr:", nrow(norm_3e_1hr), "genes\n")
```

## Import eIF3d/3e Data - 4hr Timepoint

```{r import-4hr-data}
hyp_3d_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "4hr")

norm_3d_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "4hr")

hyp_3e_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "4hr")

norm_3e_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "4hr")

cat("4hr data loaded:\n")
cat("  hyp_3d_4hr:", nrow(hyp_3d_4hr), "genes\n")
cat("  norm_3d_4hr:", nrow(norm_3d_4hr), "genes\n")
cat("  hyp_3e_4hr:", nrow(hyp_3e_4hr), "genes\n")
cat("  norm_3e_4hr:", nrow(norm_3e_4hr), "genes\n")
```

## Import eIF3d/3e Data - 24hr Timepoint

```{r import-24hr-data}
hyp_3d_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "24hr")

norm_3d_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "24hr")

hyp_3e_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "24hr")

norm_3e_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "24hr")

cat("24hr data loaded:\n")
cat("  hyp_3d_24hr:", nrow(hyp_3d_24hr), "genes\n")
  cat("  norm_3d_24hr:", nrow(norm_3d_24hr), "genes\n")
  cat("  hyp_3e_24hr:", nrow(hyp_3e_24hr), "genes\n")
  cat("  norm_3e_24hr:", nrow(norm_3e_24hr), "genes\n")
```

## Verify Data Structure

```{r verify-data-structure}
# Check column names
cat("Columns in translation categories files:\n")
colnames(hyp_3d_1hr)

# Preview data
head(hyp_3d_1hr)
```

# Define Gene Lists - 1hr Timepoint

**Key interpretation**:
- Files contain siCTRL - si3d comparison
- **Positive te_lfc** = higher TE in siCTRL = eIF3d promotes translation
- These are "down_with_3dkd" genes (analogous to reference analysis)

## 1hr - eIF3d Regulated Genes

```{r genelists-1hr-3d}
# Genes upregulated by eIF3d in hypoxia (te_lfc > 0 means higher in siCTRL)
upreg_by_3d_hypoxia_1hr <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_1hr$gs_name <- rep("upreg_by_3d_hypoxia_1hr", nrow(upreg_by_3d_hypoxia_1hr))

cat("1hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_1hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_1hr_lfc05 <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_1hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_1hr_lfc05", nrow(upreg_by_3d_hypoxia_1hr_lfc05))

cat("1hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_1hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_1hr <- norm_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_1hr$gs_name <- rep("upreg_by_3d_normoxia_1hr", nrow(upreg_by_3d_normoxia_1hr))

cat("1hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_1hr), "\n")

# Genes downregulated by eIF3d in hypoxia (optional, for comparison)
up_with_3dkd_hypoxia_1hr <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

cat("1hr hypoxia - genes downregulated by eIF3d:", nrow(up_with_3dkd_hypoxia_1hr), "\n")
```

## 1hr - eIF3e Regulated Genes

```{r genelists-1hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_1hr <- hyp_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_1hr$gs_name <- rep("upreg_by_3e_hypoxia_1hr", nrow(upreg_by_3e_hypoxia_1hr))

cat("1hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_1hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_1hr_lfc05 <- hyp_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_1hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_1hr_lfc05", nrow(upreg_by_3e_hypoxia_1hr_lfc05))

cat("1hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_1hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_1hr <- norm_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_1hr$gs_name <- rep("upreg_by_3e_normoxia_1hr", nrow(upreg_by_3e_normoxia_1hr))

cat("1hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_1hr), "\n")
```

# Define Gene Lists - 4hr Timepoint

## 4hr - eIF3d Regulated Genes

```{r genelists-4hr-3d}
# Genes upregulated by eIF3d in hypoxia
upreg_by_3d_hypoxia_4hr <- hyp_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_4hr$gs_name <- rep("upreg_by_3d_hypoxia_4hr", nrow(upreg_by_3d_hypoxia_4hr))

cat("4hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_4hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_4hr_lfc05 <- hyp_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_4hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_4hr_lfc05", nrow(upreg_by_3d_hypoxia_4hr_lfc05))

cat("4hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_4hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_4hr <- norm_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_4hr$gs_name <- rep("upreg_by_3d_normoxia_4hr", nrow(upreg_by_3d_normoxia_4hr))

cat("4hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_4hr), "\n")
```

## 4hr - eIF3e Regulated Genes

```{r genelists-4hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_4hr <- hyp_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_4hr$gs_name <- rep("upreg_by_3e_hypoxia_4hr", nrow(upreg_by_3e_hypoxia_4hr))

cat("4hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_4hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_4hr_lfc05 <- hyp_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_4hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_4hr_lfc05", nrow(upreg_by_3e_hypoxia_4hr_lfc05))

cat("4hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_4hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_4hr <- norm_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_4hr$gs_name <- rep("upreg_by_3e_normoxia_4hr", nrow(upreg_by_3e_normoxia_4hr))

cat("4hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_4hr), "\n")
```

# Define Gene Lists - 24hr Timepoint

## 24hr - eIF3d Regulated Genes

```{r genelists-24hr-3d}
# Genes upregulated by eIF3d in hypoxia
upreg_by_3d_hypoxia_24hr <- hyp_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_24hr$gs_name <- rep("upreg_by_3d_hypoxia_24hr", nrow(upreg_by_3d_hypoxia_24hr))

cat("24hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_24hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_24hr_lfc05 <- hyp_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_24hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_24hr_lfc05", nrow(upreg_by_3d_hypoxia_24hr_lfc05))

cat("24hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_24hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_24hr <- norm_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_24hr$gs_name <- rep("upreg_by_3d_normoxia_24hr", nrow(upreg_by_3d_normoxia_24hr))

cat("24hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_24hr), "\n")
```

## 24hr - eIF3e Regulated Genes

```{r genelists-24hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_24hr <- hyp_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_24hr$gs_name <- rep("upreg_by_3e_hypoxia_24hr", nrow(upreg_by_3e_hypoxia_24hr))

cat("24hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_24hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_24hr_lfc05 <- hyp_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_24hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_24hr_lfc05", nrow(upreg_by_3e_hypoxia_24hr_lfc05))

cat("24hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_24hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_24hr <- norm_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_24hr$gs_name <- rep("upreg_by_3e_normoxia_24hr", nrow(upreg_by_3e_normoxia_24hr))

cat("24hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_24hr), "\n")
```

## Summary Table of Gene Counts

```{r genelist-summary}
geneset_counts <- data.frame(
  timepoint = rep(c("1hr", "4hr", "24hr"), each = 4),
  condition_treatment = rep(c("hypoxia_3d", "normoxia_3d", "hypoxia_3e", "normoxia_3e"), 3),
  n_genes = c(
    nrow(upreg_by_3d_hypoxia_1hr),
    nrow(upreg_by_3d_normoxia_1hr),
    nrow(upreg_by_3e_hypoxia_1hr),
    nrow(upreg_by_3e_normoxia_1hr),
    nrow(upreg_by_3d_hypoxia_4hr),
    nrow(upreg_by_3d_normoxia_4hr),
    nrow(upreg_by_3e_hypoxia_4hr),
    nrow(upreg_by_3e_normoxia_4hr),
    nrow(upreg_by_3d_hypoxia_24hr),
    nrow(upreg_by_3d_normoxia_24hr),
    nrow(upreg_by_3e_hypoxia_24hr),
    nrow(upreg_by_3e_normoxia_24hr)
  )
)

geneset_counts
```



Distribution of translation after 4E inhbition
positive= still regulated upon 4E inhibition
negative=inhibited, sensitive to 4E inhibition


```{r}
# Genes with positive log2FC and padj < 0.05
# Positive = resistant to 4E inhibition (still translated when eIF4E is inhibited)

head(res_riboseq_symbol)

ggplot(res_riboseq_symbol, aes(x=log2FoldChange))+
    geom_density()+
    theme_cowplot()

#now color by categories of resistance

```

# Distribution Analysis: eIF3d/3e Upregulated Genes in Teleman Data

## 1hr - eIF3d Upregulated Genes

```{r density-3d-1hr}
# Prepare data with category labels
res_riboseq_3d_1hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_lfc <- res_riboseq_3d_1hr %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_lfc <- res_riboseq_3d_1hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_1hr <- ks.test(upreg_3d_1hr_lfc, other_1hr_lfc)

# Add sample sizes to labels
res_riboseq_3d_1hr <- res_riboseq_3d_1hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr) (n=%d)", length(upreg_3d_1hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_lfc))))

# Create color mapping
color_values_3d_1hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr) (n=%d)", length(upreg_3d_1hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_1hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3d_1hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_1hr) +
  labs(title = "1hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_1hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_1hr.pdf"), width = 8, height = 6)

cat("1hr eIF3d - KS test p-value:", ks_test_3d_1hr$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes

```{r density-3d-4hr}
# Prepare data with category labels
res_riboseq_3d_4hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_lfc <- res_riboseq_3d_4hr %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_lfc <- res_riboseq_3d_4hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_4hr <- ks.test(upreg_3d_4hr_lfc, other_4hr_lfc)

# Add sample sizes to labels
res_riboseq_3d_4hr <- res_riboseq_3d_4hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr) (n=%d)", length(upreg_3d_4hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_lfc))))

# Create color mapping
color_values_3d_4hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr) (n=%d)", length(upreg_3d_4hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_4hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3d_4hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_4hr) +
  labs(title = "4hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_4hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_4hr.pdf"), width = 8, height = 6)

cat("4hr eIF3d - KS test p-value:", ks_test_3d_4hr$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes

```{r density-3d-24hr}
# Prepare data with category labels
res_riboseq_3d_24hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_lfc <- res_riboseq_3d_24hr %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_lfc <- res_riboseq_3d_24hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_24hr <- ks.test(upreg_3d_24hr_lfc, other_24hr_lfc)

# Add sample sizes to labels
res_riboseq_3d_24hr <- res_riboseq_3d_24hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr) (n=%d)", length(upreg_3d_24hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_lfc))))

# Create color mapping
color_values_3d_24hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr) (n=%d)", length(upreg_3d_24hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_24hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3d_24hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_24hr) +
  labs(title = "24hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_24hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_24hr.pdf"), width = 8, height = 6)

cat("24hr eIF3d - KS test p-value:", ks_test_3d_24hr$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes

```{r density-3e-1hr}
# Prepare data with category labels
res_riboseq_3e_1hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_lfc <- res_riboseq_3e_1hr %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_lfc <- res_riboseq_3e_1hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_1hr <- ks.test(upreg_3e_1hr_lfc, other_3e_1hr_lfc)

# Add sample sizes to labels
res_riboseq_3e_1hr <- res_riboseq_3e_1hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr) (n=%d)", length(upreg_3e_1hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc))))

# Create color mapping
color_values_3e_1hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr) (n=%d)", length(upreg_3e_1hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3e_1hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_1hr) +
  labs(title = "1hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_1hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_1hr.pdf"), width = 8, height = 6)

cat("1hr eIF3e - KS test p-value:", ks_test_3e_1hr$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes

```{r density-3e-4hr}
# Prepare data with category labels
res_riboseq_3e_4hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_lfc <- res_riboseq_3e_4hr %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_lfc <- res_riboseq_3e_4hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_4hr <- ks.test(upreg_3e_4hr_lfc, other_3e_4hr_lfc)

# Add sample sizes to labels
res_riboseq_3e_4hr <- res_riboseq_3e_4hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr) (n=%d)", length(upreg_3e_4hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc))))

# Create color mapping
color_values_3e_4hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr) (n=%d)", length(upreg_3e_4hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3e_4hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_4hr) +
  labs(title = "4hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_4hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_4hr.pdf"), width = 8, height = 6)

cat("4hr eIF3e - KS test p-value:", ks_test_3e_4hr$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes

```{r density-3e-24hr}
# Prepare data with category labels
res_riboseq_3e_24hr <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_lfc <- res_riboseq_3e_24hr %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_lfc <- res_riboseq_3e_24hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_24hr <- ks.test(upreg_3e_24hr_lfc, other_3e_24hr_lfc)

# Add sample sizes to labels
res_riboseq_3e_24hr <- res_riboseq_3e_24hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr) (n=%d)", length(upreg_3e_24hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc))))

# Create color mapping
color_values_3e_24hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr) (n=%d)", length(upreg_3e_24hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc)))
)

# Density plot
ggplot(res_riboseq_3e_24hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_24hr) +
  labs(title = "24hr: Distribution of 4E-Resistance (Teleman Data)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_24hr$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_24hr.pdf"), width = 8, height = 6)

cat("24hr eIF3e - KS test p-value:", ks_test_3e_24hr$p.value, "\n")
```

## Summary of Statistical Tests

```{r summary-ks-tests}
ks_summary <- data.frame(
  timepoint = rep(c("1hr", "4hr", "24hr"), 2),
  treatment = c(rep("eIF3d", 3), rep("eIF3e", 3)),
  ks_pvalue = c(
    ks_test_3d_1hr$p.value,
    ks_test_3d_4hr$p.value,
    ks_test_3d_24hr$p.value,
    ks_test_3e_1hr$p.value,
    ks_test_3e_4hr$p.value,
    ks_test_3e_24hr$p.value
  ),
  ks_statistic = c(
    ks_test_3d_1hr$statistic,
    ks_test_3d_4hr$statistic,
    ks_test_3d_24hr$statistic,
    ks_test_3e_1hr$statistic,
    ks_test_3e_4hr$statistic,
    ks_test_3e_24hr$statistic
  )
)

ks_summary

write_csv(ks_summary, here("output/teleman_ks_test_summary.csv"))
```

# Distribution Analysis with Stricter Cutoff (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-1hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_lfc_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_lfc_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_1hr_lfc05 <- ks.test(upreg_3d_1hr_lfc_lfc05, other_1hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_1hr_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_lfc_lfc05))))

# Create color mapping
color_values_3d_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_1hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_1hr_lfc05) +
  labs(title = "1hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-4hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_lfc_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_lfc_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_4hr_lfc05 <- ks.test(upreg_3d_4hr_lfc_lfc05, other_4hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_4hr_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_lfc_lfc05))))

# Create color mapping
color_values_3d_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_4hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_4hr_lfc05) +
  labs(title = "4hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-24hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_lfc_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_lfc_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_24hr_lfc05 <- ks.test(upreg_3d_24hr_lfc_lfc05, other_24hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_24hr_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_lfc_lfc05))))

# Create color mapping
color_values_3d_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_24hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_24hr_lfc05) +
  labs(title = "24hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-1hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_lfc_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_lfc_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_1hr_lfc05 <- ks.test(upreg_3e_1hr_lfc_lfc05, other_3e_1hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_1hr_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc_lfc05))))

# Create color mapping
color_values_3e_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_1hr_lfc05) +
  labs(title = "1hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-4hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_lfc_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_lfc_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_4hr_lfc05 <- ks.test(upreg_3e_4hr_lfc_lfc05, other_3e_4hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_4hr_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc_lfc05))))

# Create color mapping
color_values_3e_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_4hr_lfc05) +
  labs(title = "4hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-24hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_lfc_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_lfc_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_24hr_lfc05 <- ks.test(upreg_3e_24hr_lfc_lfc05, other_3e_24hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_24hr_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc_lfc05))))

# Create color mapping
color_values_3e_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_24hr_lfc05) +
  labs(title = "24hr: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_24hr_lfc05$p.value, "\n")
```


I don't really think this is showing anything stark

Let's look at binding

## Do mRNAs upregulated by 3d lose 4E binding upon 4E inhibition?

# Read in teleman 4E IP data
fig 4E

eIF4E-IP

log2((4ebpI cells : eIF4E-IP/Input) / (control cells: eIF4E IP /Total))

Indeed, transcriptome-wide, we observed a negative correlation between the change in translation efficiency of mRNAs upon eIF4E inhibition and their binding to eIF4E (Fig. 4E).

```{r}
teleman_IP_4E_inhib_v_vehicle <- read_csv("/Users/katematlin/github/mpra_preliminary/res_ip_4E_inhib_v_vehicle.csv")

teleman_IP_4E_vehicle <- read_csv("/Users/katematlin/github/mpra_preliminary/res_ip_vehicle.csv")

teleman_IP_4E_inhib <- read_csv("/Users/katematlin/github/mpra_preliminary/res_4E_ip_inhib.csv")

```

Look primarily at teleman IP 4E inibitor vs vehicle


```{r}
ggplot(teleman_IP_4E_inhib_v_vehicle, aes(x=log2FoldChange)) +
    geom_density()+
    theme_cowplot()
```

# Distribution Analysis: eIF3d/3e Upregulated Genes in 4E-IP Data

## 1hr - eIF3d Upregulated Genes (4E-IP)

```{r ip-density-3d-1hr}
# Prepare data with category labels
ip_4E_3d_1hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_ip <- ip_4E_3d_1hr %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_ip <- ip_4E_3d_1hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_1hr <- ks.test(upreg_3d_1hr_ip, other_1hr_ip)

# Add sample sizes to labels
ip_4E_3d_1hr <- ip_4E_3d_1hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr) (n=%d)", length(upreg_3d_1hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_ip))))

# Create color mapping
color_values_ip_3d_1hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr) (n=%d)", length(upreg_3d_1hr_ip)),
    sprintf("All other genes (n=%d)", length(other_1hr_ip)))
)

# Density plot
ggplot(ip_4E_3d_1hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_1hr) +
  labs(title = "1hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_1hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_1hr.pdf"), width = 8, height = 6)

cat("1hr eIF3d - 4E-IP KS test p-value:", ks_test_ip_3d_1hr$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes (4E-IP)

```{r ip-density-3d-4hr}
# Prepare data with category labels
ip_4E_3d_4hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_ip <- ip_4E_3d_4hr %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_ip <- ip_4E_3d_4hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_4hr <- ks.test(upreg_3d_4hr_ip, other_4hr_ip)

# Add sample sizes to labels
ip_4E_3d_4hr <- ip_4E_3d_4hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr) (n=%d)", length(upreg_3d_4hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_ip))))

# Create color mapping
color_values_ip_3d_4hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr) (n=%d)", length(upreg_3d_4hr_ip)),
    sprintf("All other genes (n=%d)", length(other_4hr_ip)))
)

# Density plot
ggplot(ip_4E_3d_4hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_4hr) +
  labs(title = "4hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_4hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_4hr.pdf"), width = 8, height = 6)

cat("4hr eIF3d - 4E-IP KS test p-value:", ks_test_ip_3d_4hr$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes (4E-IP)

```{r ip-density-3d-24hr}
# Prepare data with category labels
ip_4E_3d_24hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_ip <- ip_4E_3d_24hr %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_ip <- ip_4E_3d_24hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_24hr <- ks.test(upreg_3d_24hr_ip, other_24hr_ip)

# Add sample sizes to labels
ip_4E_3d_24hr <- ip_4E_3d_24hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr) (n=%d)", length(upreg_3d_24hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_ip))))

# Create color mapping
color_values_ip_3d_24hr <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr) (n=%d)", length(upreg_3d_24hr_ip)),
    sprintf("All other genes (n=%d)", length(other_24hr_ip)))
)

# Density plot
ggplot(ip_4E_3d_24hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_24hr) +
  labs(title = "24hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_24hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_24hr.pdf"), width = 8, height = 6)

cat("24hr eIF3d - 4E-IP KS test p-value:", ks_test_ip_3d_24hr$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes (4E-IP)

```{r ip-density-3e-1hr}
# Prepare data with category labels
ip_4E_3e_1hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_ip <- ip_4E_3e_1hr %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_ip <- ip_4E_3e_1hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_1hr <- ks.test(upreg_3e_1hr_ip, other_3e_1hr_ip)

# Add sample sizes to labels
ip_4E_3e_1hr <- ip_4E_3e_1hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr) (n=%d)", length(upreg_3e_1hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_ip))))

# Create color mapping
color_values_ip_3e_1hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr) (n=%d)", length(upreg_3e_1hr_ip)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_ip)))
)

# Density plot
ggplot(ip_4E_3e_1hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_1hr) +
  labs(title = "1hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_1hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_1hr.pdf"), width = 8, height = 6)

cat("1hr eIF3e - 4E-IP KS test p-value:", ks_test_ip_3e_1hr$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes (4E-IP)

```{r ip-density-3e-4hr}
# Prepare data with category labels
ip_4E_3e_4hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_ip <- ip_4E_3e_4hr %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_ip <- ip_4E_3e_4hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_4hr <- ks.test(upreg_3e_4hr_ip, other_3e_4hr_ip)

# Add sample sizes to labels
ip_4E_3e_4hr <- ip_4E_3e_4hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr) (n=%d)", length(upreg_3e_4hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_ip))))

# Create color mapping
color_values_ip_3e_4hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr) (n=%d)", length(upreg_3e_4hr_ip)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_ip)))
)

# Density plot
ggplot(ip_4E_3e_4hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_4hr) +
  labs(title = "4hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_4hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_4hr.pdf"), width = 8, height = 6)

cat("4hr eIF3e - 4E-IP KS test p-value:", ks_test_ip_3e_4hr$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes (4E-IP)

```{r ip-density-3e-24hr}
# Prepare data with category labels
ip_4E_3e_24hr <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_ip <- ip_4E_3e_24hr %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_ip <- ip_4E_3e_24hr %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_24hr <- ks.test(upreg_3e_24hr_ip, other_3e_24hr_ip)

# Add sample sizes to labels
ip_4E_3e_24hr <- ip_4E_3e_24hr %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr) (n=%d)", length(upreg_3e_24hr_ip)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_ip))))

# Create color mapping
color_values_ip_3e_24hr <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr) (n=%d)", length(upreg_3e_24hr_ip)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_ip)))
)

# Density plot
ggplot(ip_4E_3e_24hr, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_24hr) +
  labs(title = "24hr: 4E-IP Change upon 4E Inhibition",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_24hr$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_24hr.pdf"), width = 8, height = 6)

cat("24hr eIF3e - 4E-IP KS test p-value:", ks_test_ip_3e_24hr$p.value, "\n")
```

# 4E-IP Distribution Analysis with Stricter Cutoff (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-1hr-lfc05}
# Prepare data with category labels
ip_4E_3d_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_ip_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_ip_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_1hr_lfc05 <- ks.test(upreg_3d_1hr_ip_lfc05, other_1hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_1hr_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_1hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_1hr_lfc05) +
  labs(title = "1hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-4hr-lfc05}
# Prepare data with category labels
ip_4E_3d_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_ip_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_ip_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_4hr_lfc05 <- ks.test(upreg_3d_4hr_ip_lfc05, other_4hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_4hr_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_4hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_4hr_lfc05) +
  labs(title = "4hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-24hr-lfc05}
# Prepare data with category labels
ip_4E_3d_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_ip_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_ip_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_24hr_lfc05 <- ks.test(upreg_3d_24hr_ip_lfc05, other_24hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_24hr_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_24hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_24hr_lfc05) +
  labs(title = "24hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-1hr-lfc05}
# Prepare data with category labels
ip_4E_3e_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_ip_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_ip_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_1hr_lfc05 <- ks.test(upreg_3e_1hr_ip_lfc05, other_3e_1hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_1hr_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_1hr_lfc05) +
  labs(title = "1hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-4hr-lfc05}
# Prepare data with category labels
ip_4E_3e_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_ip_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_ip_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_4hr_lfc05 <- ks.test(upreg_3e_4hr_ip_lfc05, other_3e_4hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_4hr_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_4hr_lfc05) +
  labs(title = "4hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-24hr-lfc05}
# Prepare data with category labels
ip_4E_3e_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_ip_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_ip_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_24hr_lfc05 <- ks.test(upreg_3e_24hr_ip_lfc05, other_3e_24hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_24hr_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_24hr_lfc05) +
  labs(title = "24hr: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_24hr_lfc05$p.value, "\n")
```

## Summary of 4E-IP Statistical Tests

```{r summary-ip-ks-tests}
ks_ip_summary <- data.frame(
  timepoint = rep(c("1hr", "4hr", "24hr"), 2),
  treatment = c(rep("eIF3d", 3), rep("eIF3e", 3)),
  ks_pvalue = c(
    ks_test_ip_3d_1hr$p.value,
    ks_test_ip_3d_4hr$p.value,
    ks_test_ip_3d_24hr$p.value,
    ks_test_ip_3e_1hr$p.value,
    ks_test_ip_3e_4hr$p.value,
    ks_test_ip_3e_24hr$p.value
  ),
  ks_statistic = c(
    ks_test_ip_3d_1hr$statistic,
    ks_test_ip_3d_4hr$statistic,
    ks_test_ip_3d_24hr$statistic,
    ks_test_ip_3e_1hr$statistic,
    ks_test_ip_3e_4hr$statistic,
    ks_test_ip_3e_24hr$statistic
  )
)

ks_ip_summary

write_csv(ks_ip_summary, here("output/teleman_ip_ks_test_summary.csv"))
```