---
title: "Teleman 4E Resistance Overlap with MDA-MB-231 Timecourse"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

Compare genes resistant to eIF4E inhibition (Teleman RiboSeq dataset) with genes upregulated by eIF3d in MDA-MB-231 cells across a hypoxia timecourse (1hr, 4hr, 24hr).

**Reference analysis**: `/Users/katematlin/github/mpra_preliminary/code/resistant_4e_overlap2.Rmd`

**Key question**: Do genes that maintain translation when eIF4E is inhibited overlap with genes whose translation is promoted by eIF3d during hypoxic stress?

## Load Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(DESeq2)
library(janitor)
library(ggVennDiagram)
library(cowplot)
library(GeneOverlap)
library(rstatix)
library(VennDiagram)
library(ggvenn)
library(writexl)
library(readxl)
library(cowplot)
library(ggpubr)
```

## Read in Teleman riboseq data
This was analyzed by me previously
teleman riboseq = res riboseq symbol
```{r}
res_riboseq_symbol <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv")
```


## Define 4E-Resistant Genes

```{r define-resistant-4e}
# Genes with positive log2FC and padj < 0.05
# Positive = resistant to 4E inhibition (still translated when eIF4E is inhibited)
resistant_4E <- res_riboseq_symbol %>%
  filter(log2FoldChange > 1) %>%
  filter(padj < 0.05) %>%
  filter(!is.na(symbol))

cat("Number of 4E-resistant genes (before deduplication):", nrow(resistant_4E), "\n")

# Check for duplicates
n_duplicates <- sum(duplicated(resistant_4E$symbol))
cat("Duplicate gene symbols:", n_duplicates, "\n")

# Deduplicate based on highest baseMean (keep gene with highest expression)
if(n_duplicates > 0) {
  resistant_4E <- resistant_4E %>%
    group_by(symbol) %>%
    slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
    ungroup()

  cat("After deduplication:", nrow(resistant_4E), "genes\n")
}

# Create gene set for overlap analysis - use pull() to extract vector
resistant_set <- data.frame(
  gs_name = "Resistant_4E",
  gene = resistant_4E %>% pull(symbol)
)

# Show top resistant genes
resistant_4E %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(symbol, log2FoldChange, padj, baseMean) %>%
  head(20)
```

# Load Timecourse Data

## Import eIF3d/3e Data - 1hr Timepoint

```{r import-1hr-data}
hyp_3d_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "1hr")

norm_3d_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "1hr")

hyp_3e_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "1hr")

norm_3e_1hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "1hr")

cat("1hr data loaded:\n")
cat("  hyp_3d_1hr:", nrow(hyp_3d_1hr), "genes\n")
cat("  norm_3d_1hr:", nrow(norm_3d_1hr), "genes\n")
cat("  hyp_3e_1hr:", nrow(hyp_3e_1hr), "genes\n")
cat("  norm_3e_1hr:", nrow(norm_3e_1hr), "genes\n")
```

## Import eIF3d/3e Data - 4hr Timepoint

```{r import-4hr-data}
hyp_3d_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "4hr")

norm_3d_4hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_4hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "4hr")

hyp_3e_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "4hr")

norm_3e_4hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "4hr")

cat("4hr data loaded:\n")
cat("  hyp_3d_4hr:", nrow(hyp_3d_4hr), "genes\n")
cat("  norm_3d_4hr:", nrow(norm_3d_4hr), "genes\n")
cat("  hyp_3e_4hr:", nrow(hyp_3e_4hr), "genes\n")
cat("  norm_3e_4hr:", nrow(norm_3e_4hr), "genes\n")
```

## Import eIF3d/3e Data - 24hr Timepoint

```{r import-24hr-data}
hyp_3d_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "24hr")

norm_3d_24hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_24hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "24hr")

hyp_3e_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")) %>%
  mutate(condition = "hypoxia", timepoint = "24hr")

norm_3e_24hr <- read_csv(here("output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv")) %>%
  mutate(condition = "normoxia", timepoint = "24hr")

cat("24hr data loaded:\n")
cat("  hyp_3d_24hr:", nrow(hyp_3d_24hr), "genes\n")
  cat("  norm_3d_24hr:", nrow(norm_3d_24hr), "genes\n")
  cat("  hyp_3e_24hr:", nrow(hyp_3e_24hr), "genes\n")
  cat("  norm_3e_24hr:", nrow(norm_3e_24hr), "genes\n")
```

## Verify Data Structure

```{r verify-data-structure}
# Check column names
cat("Columns in translation categories files:\n")
colnames(hyp_3d_1hr)

# Preview data
head(hyp_3d_1hr)
```

# Define Gene Lists - 1hr Timepoint

**Key interpretation**:
- Files contain siCTRL - si3d comparison
- **Positive te_lfc** = higher TE in siCTRL = eIF3d promotes translation
- These are "down_with_3dkd" genes (analogous to reference analysis)

## 1hr - eIF3d Regulated Genes

```{r genelists-1hr-3d}
# Genes upregulated by eIF3d in hypoxia (te_lfc > 0 means higher in siCTRL)
upreg_by_3d_hypoxia_1hr <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_1hr$gs_name <- rep("upreg_by_3d_hypoxia_1hr", nrow(upreg_by_3d_hypoxia_1hr))

cat("1hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_1hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_1hr_lfc05 <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_1hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_1hr_lfc05", nrow(upreg_by_3d_hypoxia_1hr_lfc05))

cat("1hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_1hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_1hr <- norm_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_1hr$gs_name <- rep("upreg_by_3d_normoxia_1hr", nrow(upreg_by_3d_normoxia_1hr))

cat("1hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_1hr), "\n")

# Genes upregulated by eIF3d in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_normoxia_1hr_lfc05 <- norm_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_1hr_lfc05$gs_name <- rep("upreg_by_3d_normoxia_1hr_lfc05", nrow(upreg_by_3d_normoxia_1hr_lfc05))

cat("1hr normoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_normoxia_1hr_lfc05), "\n")

# Genes downregulated by eIF3d in hypoxia (optional, for comparison)
up_with_3dkd_hypoxia_1hr <- hyp_3d_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc < 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

cat("1hr hypoxia - genes downregulated by eIF3d:", nrow(up_with_3dkd_hypoxia_1hr), "\n")
```

## 1hr - eIF3e Regulated Genes

```{r genelists-1hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_1hr <- hyp_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_1hr$gs_name <- rep("upreg_by_3e_hypoxia_1hr", nrow(upreg_by_3e_hypoxia_1hr))

cat("1hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_1hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_1hr_lfc05 <- hyp_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_1hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_1hr_lfc05", nrow(upreg_by_3e_hypoxia_1hr_lfc05))

cat("1hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_1hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_1hr <- norm_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_1hr$gs_name <- rep("upreg_by_3e_normoxia_1hr", nrow(upreg_by_3e_normoxia_1hr))

cat("1hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_1hr), "\n")

# Genes upregulated by eIF3e in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_normoxia_1hr_lfc05 <- norm_3e_1hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_1hr_lfc05$gs_name <- rep("upreg_by_3e_normoxia_1hr_lfc05", nrow(upreg_by_3e_normoxia_1hr_lfc05))

cat("1hr normoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_normoxia_1hr_lfc05), "\n")
```

# Define Gene Lists - 4hr Timepoint

## 4hr - eIF3d Regulated Genes

```{r genelists-4hr-3d}
# Genes upregulated by eIF3d in hypoxia
upreg_by_3d_hypoxia_4hr <- hyp_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_4hr$gs_name <- rep("upreg_by_3d_hypoxia_4hr", nrow(upreg_by_3d_hypoxia_4hr))

cat("4hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_4hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_4hr_lfc05 <- hyp_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_4hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_4hr_lfc05", nrow(upreg_by_3d_hypoxia_4hr_lfc05))

cat("4hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_4hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_4hr <- norm_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_4hr$gs_name <- rep("upreg_by_3d_normoxia_4hr", nrow(upreg_by_3d_normoxia_4hr))

cat("4hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_4hr), "\n")

# Genes upregulated by eIF3d in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_normoxia_4hr_lfc05 <- norm_3d_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_4hr_lfc05$gs_name <- rep("upreg_by_3d_normoxia_4hr_lfc05", nrow(upreg_by_3d_normoxia_4hr_lfc05))

cat("4hr normoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_normoxia_4hr_lfc05), "\n")
```

## 4hr - eIF3e Regulated Genes

```{r genelists-4hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_4hr <- hyp_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_4hr$gs_name <- rep("upreg_by_3e_hypoxia_4hr", nrow(upreg_by_3e_hypoxia_4hr))

cat("4hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_4hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_4hr_lfc05 <- hyp_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_4hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_4hr_lfc05", nrow(upreg_by_3e_hypoxia_4hr_lfc05))

cat("4hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_4hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_4hr <- norm_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_4hr$gs_name <- rep("upreg_by_3e_normoxia_4hr", nrow(upreg_by_3e_normoxia_4hr))

cat("4hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_4hr), "\n")

# Genes upregulated by eIF3e in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_normoxia_4hr_lfc05 <- norm_3e_4hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_4hr_lfc05$gs_name <- rep("upreg_by_3e_normoxia_4hr_lfc05", nrow(upreg_by_3e_normoxia_4hr_lfc05))

cat("4hr normoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_normoxia_4hr_lfc05), "\n")
```

# Define Gene Lists - 24hr Timepoint

## 24hr - eIF3d Regulated Genes

```{r genelists-24hr-3d}
# Genes upregulated by eIF3d in hypoxia
upreg_by_3d_hypoxia_24hr <- hyp_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_24hr$gs_name <- rep("upreg_by_3d_hypoxia_24hr", nrow(upreg_by_3d_hypoxia_24hr))

cat("24hr hypoxia - genes upregulated by eIF3d (lfc > 0):", nrow(upreg_by_3d_hypoxia_24hr), "\n")

# Genes upregulated by eIF3d in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_hypoxia_24hr_lfc05 <- hyp_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_hypoxia_24hr_lfc05$gs_name <- rep("upreg_by_3d_hypoxia_24hr_lfc05", nrow(upreg_by_3d_hypoxia_24hr_lfc05))

cat("24hr hypoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_hypoxia_24hr_lfc05), "\n")

# Genes upregulated by eIF3d in normoxia
upreg_by_3d_normoxia_24hr <- norm_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_24hr$gs_name <- rep("upreg_by_3d_normoxia_24hr", nrow(upreg_by_3d_normoxia_24hr))

cat("24hr normoxia - genes upregulated by eIF3d:", nrow(upreg_by_3d_normoxia_24hr), "\n")

# Genes upregulated by eIF3d in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3d_normoxia_24hr_lfc05 <- norm_3d_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3d_normoxia_24hr_lfc05$gs_name <- rep("upreg_by_3d_normoxia_24hr_lfc05", nrow(upreg_by_3d_normoxia_24hr_lfc05))

cat("24hr normoxia - genes upregulated by eIF3d (lfc > 0.5):", nrow(upreg_by_3d_normoxia_24hr_lfc05), "\n")
```

## 24hr - eIF3e Regulated Genes

```{r genelists-24hr-3e}
# Genes upregulated by eIF3e in hypoxia
upreg_by_3e_hypoxia_24hr <- hyp_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_24hr$gs_name <- rep("upreg_by_3e_hypoxia_24hr", nrow(upreg_by_3e_hypoxia_24hr))

cat("24hr hypoxia - genes upregulated by eIF3e (lfc > 0):", nrow(upreg_by_3e_hypoxia_24hr), "\n")

# Genes upregulated by eIF3e in hypoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_hypoxia_24hr_lfc05 <- hyp_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_hypoxia_24hr_lfc05$gs_name <- rep("upreg_by_3e_hypoxia_24hr_lfc05", nrow(upreg_by_3e_hypoxia_24hr_lfc05))

cat("24hr hypoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_hypoxia_24hr_lfc05), "\n")

# Genes upregulated by eIF3e in normoxia
upreg_by_3e_normoxia_24hr <- norm_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_24hr$gs_name <- rep("upreg_by_3e_normoxia_24hr", nrow(upreg_by_3e_normoxia_24hr))

cat("24hr normoxia - genes upregulated by eIF3e:", nrow(upreg_by_3e_normoxia_24hr), "\n")

# Genes upregulated by eIF3e in normoxia with stricter cutoff (te_lfc > 0.5)
upreg_by_3e_normoxia_24hr_lfc05 <- norm_3e_24hr %>%
  filter(te_padj < 0.05) %>%
  filter(te_lfc > 0.5) %>%
  filter(!is.na(symbol)) %>%
  dplyr::select(symbol) %>%
  distinct()

upreg_by_3e_normoxia_24hr_lfc05$gs_name <- rep("upreg_by_3e_normoxia_24hr_lfc05", nrow(upreg_by_3e_normoxia_24hr_lfc05))

cat("24hr normoxia - genes upregulated by eIF3e (lfc > 0.5):", nrow(upreg_by_3e_normoxia_24hr_lfc05), "\n")
```

## Summary Table of Gene Counts

```{r genelist-summary}
geneset_counts <- data.frame(
  timepoint = rep(c("1hr", "4hr", "24hr"), each = 4),
  condition_treatment = rep(c("hypoxia_3d", "normoxia_3d", "hypoxia_3e", "normoxia_3e"), 3),
  n_genes = c(
    nrow(upreg_by_3d_hypoxia_1hr),
    nrow(upreg_by_3d_normoxia_1hr),
    nrow(upreg_by_3e_hypoxia_1hr),
    nrow(upreg_by_3e_normoxia_1hr),
    nrow(upreg_by_3d_hypoxia_4hr),
    nrow(upreg_by_3d_normoxia_4hr),
    nrow(upreg_by_3e_hypoxia_4hr),
    nrow(upreg_by_3e_normoxia_4hr),
    nrow(upreg_by_3d_hypoxia_24hr),
    nrow(upreg_by_3d_normoxia_24hr),
    nrow(upreg_by_3e_hypoxia_24hr),
    nrow(upreg_by_3e_normoxia_24hr)
  )
)

geneset_counts
```



Distribution of translation after 4E inhbition
positive= still regulated upon 4E inhibition
negative=inhibited, sensitive to 4E inhibition


```{r}
# Genes with positive log2FC and padj < 0.05
# Positive = resistant to 4E inhibition (still translated when eIF4E is inhibited)

head(res_riboseq_symbol)

ggplot(res_riboseq_symbol, aes(x=log2FoldChange))+
    geom_density()+
    theme_cowplot()

#now color by categories of resistance

```

# Distribution Analysis (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-1hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_lfc_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_lfc_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_1hr_lfc05 <- ks.test(upreg_3d_1hr_lfc_lfc05, other_1hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_1hr_lfc05 <- res_riboseq_3d_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_lfc_lfc05))))

# Create color mapping
color_values_3d_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_1hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_1hr_lfc05) +
  labs(title = "1hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-4hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_lfc_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_lfc_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_4hr_lfc05 <- ks.test(upreg_3d_4hr_lfc_lfc05, other_4hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_4hr_lfc05 <- res_riboseq_3d_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_lfc_lfc05))))

# Create color mapping
color_values_3d_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_4hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_4hr_lfc05) +
  labs(title = "4hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes (LFC > 0.5)

```{r density-3d-24hr-lfc05}
# Prepare data with category labels
res_riboseq_3d_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_lfc_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_lfc_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_24hr_lfc05 <- ks.test(upreg_3d_24hr_lfc_lfc05, other_24hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3d_24hr_lfc05 <- res_riboseq_3d_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_lfc_lfc05))))

# Create color mapping
color_values_3d_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_24hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3d_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_24hr_lfc05) +
  labs(title = "24hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-1hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_lfc_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_lfc_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_1hr_lfc05 <- ks.test(upreg_3e_1hr_lfc_lfc05, other_3e_1hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_1hr_lfc05 <- res_riboseq_3e_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc_lfc05))))

# Create color mapping
color_values_3e_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_1hr_lfc05) +
  labs(title = "1hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-4hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_lfc_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_lfc_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_4hr_lfc05 <- ks.test(upreg_3e_4hr_lfc_lfc05, other_3e_4hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_4hr_lfc05 <- res_riboseq_3e_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc_lfc05))))

# Create color mapping
color_values_3e_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_4hr_lfc05) +
  labs(title = "4hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes (LFC > 0.5)

```{r density-3e-24hr-lfc05}
# Prepare data with category labels
res_riboseq_3e_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_lfc_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_lfc_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_24hr_lfc05 <- ks.test(upreg_3e_24hr_lfc_lfc05, other_3e_24hr_lfc_lfc05)

# Add sample sizes to labels
res_riboseq_3e_24hr_lfc05 <- res_riboseq_3e_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_lfc_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc_lfc05))))

# Create color mapping
color_values_3e_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_lfc_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_lfc_lfc05)))
)

# Density plot
ggplot(res_riboseq_3e_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_24hr_lfc05) +
  labs(title = "24hr hypoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_24hr_lfc05$p.value, "\n")
```


# Distribution Analysis - Normoxia (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3d-norm-1hr-lfc05}
res_riboseq_3d_norm_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (1hr)",
                           "All other genes"))

upreg_3d_norm_1hr_lfc <- res_riboseq_3d_norm_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (1hr)") %>%
  pull(log2FoldChange)

other_norm_1hr_lfc <- res_riboseq_3d_norm_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_norm_1hr_lfc05 <- ks.test(upreg_3d_norm_1hr_lfc, other_norm_1hr_lfc)

res_riboseq_3d_norm_1hr_lfc05 <- res_riboseq_3d_norm_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (1hr)",
                                   sprintf("Upregulated by eIF3d normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_1hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_1hr_lfc))))

color_values_3d_norm_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_1hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_1hr_lfc)))
)

ggplot(res_riboseq_3d_norm_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_norm_1hr_lfc05) +
  labs(title = "1hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_norm_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_norm_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr normoxia eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_norm_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3d-norm-4hr-lfc05}
res_riboseq_3d_norm_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (4hr)",
                           "All other genes"))

upreg_3d_norm_4hr_lfc <- res_riboseq_3d_norm_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (4hr)") %>%
  pull(log2FoldChange)

other_norm_4hr_lfc <- res_riboseq_3d_norm_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_norm_4hr_lfc05 <- ks.test(upreg_3d_norm_4hr_lfc, other_norm_4hr_lfc)

res_riboseq_3d_norm_4hr_lfc05 <- res_riboseq_3d_norm_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (4hr)",
                                   sprintf("Upregulated by eIF3d normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_4hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_4hr_lfc))))

color_values_3d_norm_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_4hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_4hr_lfc)))
)

ggplot(res_riboseq_3d_norm_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_norm_4hr_lfc05) +
  labs(title = "4hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_norm_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_norm_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr normoxia eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_norm_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3d-norm-24hr-lfc05}
res_riboseq_3d_norm_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (24hr)",
                           "All other genes"))

upreg_3d_norm_24hr_lfc <- res_riboseq_3d_norm_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (24hr)") %>%
  pull(log2FoldChange)

other_norm_24hr_lfc <- res_riboseq_3d_norm_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3d_norm_24hr_lfc05 <- ks.test(upreg_3d_norm_24hr_lfc, other_norm_24hr_lfc)

res_riboseq_3d_norm_24hr_lfc05 <- res_riboseq_3d_norm_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (24hr)",
                                   sprintf("Upregulated by eIF3d normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_24hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_24hr_lfc))))

color_values_3d_norm_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_24hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_24hr_lfc)))
)

ggplot(res_riboseq_3d_norm_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3d_norm_24hr_lfc05) +
  labs(title = "24hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3d_norm_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3d_norm_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr normoxia eIF3d (lfc>0.5) - KS test p-value:", ks_test_3d_norm_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3e-norm-1hr-lfc05}
res_riboseq_3e_norm_1hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (1hr)",
                           "All other genes"))

upreg_3e_norm_1hr_lfc <- res_riboseq_3e_norm_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (1hr)") %>%
  pull(log2FoldChange)

other_3e_norm_1hr_lfc <- res_riboseq_3e_norm_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_norm_1hr_lfc05 <- ks.test(upreg_3e_norm_1hr_lfc, other_3e_norm_1hr_lfc)

res_riboseq_3e_norm_1hr_lfc05 <- res_riboseq_3e_norm_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (1hr)",
                                   sprintf("Upregulated by eIF3e normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_1hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_1hr_lfc))))

color_values_3e_norm_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_1hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_1hr_lfc)))
)

ggplot(res_riboseq_3e_norm_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_norm_1hr_lfc05) +
  labs(title = "1hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_norm_1hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_norm_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr normoxia eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_norm_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3e-norm-4hr-lfc05}
res_riboseq_3e_norm_4hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (4hr)",
                           "All other genes"))

upreg_3e_norm_4hr_lfc <- res_riboseq_3e_norm_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (4hr)") %>%
  pull(log2FoldChange)

other_3e_norm_4hr_lfc <- res_riboseq_3e_norm_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_norm_4hr_lfc05 <- ks.test(upreg_3e_norm_4hr_lfc, other_3e_norm_4hr_lfc)

res_riboseq_3e_norm_4hr_lfc05 <- res_riboseq_3e_norm_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (4hr)",
                                   sprintf("Upregulated by eIF3e normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_4hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_4hr_lfc))))

color_values_3e_norm_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_4hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_4hr_lfc)))
)

ggplot(res_riboseq_3e_norm_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_norm_4hr_lfc05) +
  labs(title = "4hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_norm_4hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_norm_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr normoxia eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_norm_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5)

```{r density-3e-norm-24hr-lfc05}
res_riboseq_3e_norm_24hr_lfc05 <- res_riboseq_symbol %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (24hr)",
                           "All other genes"))

upreg_3e_norm_24hr_lfc <- res_riboseq_3e_norm_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (24hr)") %>%
  pull(log2FoldChange)

other_3e_norm_24hr_lfc <- res_riboseq_3e_norm_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_3e_norm_24hr_lfc05 <- ks.test(upreg_3e_norm_24hr_lfc, other_3e_norm_24hr_lfc)

res_riboseq_3e_norm_24hr_lfc05 <- res_riboseq_3e_norm_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (24hr)",
                                   sprintf("Upregulated by eIF3e normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_24hr_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_24hr_lfc))))

color_values_3e_norm_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_24hr_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_24hr_lfc)))
)

ggplot(res_riboseq_3e_norm_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_3e_norm_24hr_lfc05) +
  labs(title = "24hr normoxia: Distribution of 4E-Resistance (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_3e_norm_24hr_lfc05$p.value),
       x = "log2FC (4E inhibition)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  xlim(-2, 2) +
  theme(legend.position = "top")

ggsave(here("output/teleman_density_3e_norm_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr normoxia eIF3e (lfc>0.5) - KS test p-value:", ks_test_3e_norm_24hr_lfc05$p.value, "\n")
```

I don't really think this is showing anything stark

Let's look at binding

## Do mRNAs upregulated by 3d lose 4E binding upon 4E inhibition?

# Read in teleman 4E IP data
fig 4E

eIF4E-IP

log2((4ebpI cells : eIF4E-IP/Input) / (control cells: eIF4E IP /Total))

Indeed, transcriptome-wide, we observed a negative correlation between the change in translation efficiency of mRNAs upon eIF4E inhibition and their binding to eIF4E (Fig. 4E).

```{r}
teleman_IP_4E_inhib_v_vehicle <- read_csv("/Users/katematlin/github/mpra_preliminary/res_ip_4E_inhib_v_vehicle.csv")

teleman_IP_4E_vehicle <- read_csv("/Users/katematlin/github/mpra_preliminary/res_ip_vehicle.csv")

teleman_IP_4E_inhib <- read_csv("/Users/katematlin/github/mpra_preliminary/res_4E_ip_inhib.csv")

```

Look primarily at teleman IP 4E inibitor vs vehicle


```{r}
ggplot(teleman_IP_4E_inhib_v_vehicle, aes(x=log2FoldChange)) +
    geom_density()+
    theme_cowplot()
```

# Distribution Analysis: eIF3d/3e Upregulated Genes in 4E-IP Data

# 4E-IP Distribution Analysis (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-1hr-lfc05}
# Prepare data with category labels
ip_4E_3d_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d (1hr)",
                           "All other genes"))

# Statistical test
upreg_3d_1hr_ip_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (1hr)") %>%
  pull(log2FoldChange)

other_1hr_ip_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_1hr_lfc05 <- ks.test(upreg_3d_1hr_ip_lfc05, other_1hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_1hr_lfc05 <- ip_4E_3d_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (1hr)",
                                   sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_1hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (1hr, lfc>0.5) (n=%d)", length(upreg_3d_1hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_1hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_1hr_lfc05) +
  labs(title = "1hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-4hr-lfc05}
# Prepare data with category labels
ip_4E_3d_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d (4hr)",
                           "All other genes"))

# Statistical test
upreg_3d_4hr_ip_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (4hr)") %>%
  pull(log2FoldChange)

other_4hr_ip_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_4hr_lfc05 <- ks.test(upreg_3d_4hr_ip_lfc05, other_4hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_4hr_lfc05 <- ip_4E_3d_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (4hr)",
                                   sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_4hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (4hr, lfc>0.5) (n=%d)", length(upreg_3d_4hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_4hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_4hr_lfc05) +
  labs(title = "4hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3d-24hr-lfc05}
# Prepare data with category labels
ip_4E_3d_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d (24hr)",
                           "All other genes"))

# Statistical test
upreg_3d_24hr_ip_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d (24hr)") %>%
  pull(log2FoldChange)

other_24hr_ip_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_24hr_lfc05 <- ks.test(upreg_3d_24hr_ip_lfc05, other_24hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3d_24hr_lfc05 <- ip_4E_3d_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d (24hr)",
                                   sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_24hr_ip_lfc05))))

# Create color mapping
color_values_ip_3d_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d (24hr, lfc>0.5) (n=%d)", length(upreg_3d_24hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_24hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3d_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_24hr_lfc05) +
  labs(title = "24hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-1hr-lfc05}
# Prepare data with category labels
ip_4E_3e_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e (1hr)",
                           "All other genes"))

# Statistical test
upreg_3e_1hr_ip_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (1hr)") %>%
  pull(log2FoldChange)

other_3e_1hr_ip_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_1hr_lfc05 <- ks.test(upreg_3e_1hr_ip_lfc05, other_3e_1hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_1hr_lfc05 <- ip_4E_3e_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (1hr)",
                                   sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_1hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (1hr, lfc>0.5) (n=%d)", length(upreg_3e_1hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_1hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_1hr_lfc05) +
  labs(title = "1hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-4hr-lfc05}
# Prepare data with category labels
ip_4E_3e_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e (4hr)",
                           "All other genes"))

# Statistical test
upreg_3e_4hr_ip_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (4hr)") %>%
  pull(log2FoldChange)

other_3e_4hr_ip_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_4hr_lfc05 <- ks.test(upreg_3e_4hr_ip_lfc05, other_3e_4hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_4hr_lfc05 <- ip_4E_3e_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (4hr)",
                                   sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_4hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (4hr, lfc>0.5) (n=%d)", length(upreg_3e_4hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_4hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_4hr_lfc05) +
  labs(title = "4hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes (LFC > 0.5) - 4E-IP

```{r ip-density-3e-24hr-lfc05}
# Prepare data with category labels
ip_4E_3e_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_hypoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e (24hr)",
                           "All other genes"))

# Statistical test
upreg_3e_24hr_ip_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e (24hr)") %>%
  pull(log2FoldChange)

other_3e_24hr_ip_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_24hr_lfc05 <- ks.test(upreg_3e_24hr_ip_lfc05, other_3e_24hr_ip_lfc05)

# Add sample sizes to labels
ip_4E_3e_24hr_lfc05 <- ip_4E_3e_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e (24hr)",
                                   sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_ip_lfc05)),
                                   sprintf("All other genes (n=%d)", length(other_3e_24hr_ip_lfc05))))

# Create color mapping
color_values_ip_3e_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e (24hr, lfc>0.5) (n=%d)", length(upreg_3e_24hr_ip_lfc05)),
    sprintf("All other genes (n=%d)", length(other_3e_24hr_ip_lfc05)))
)

# Density plot
ggplot(ip_4E_3e_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_24hr_lfc05) +
  labs(title = "24hr hypoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_24hr_lfc05$p.value, "\n")
```


# 4E-IP Distribution Analysis - Normoxia (LFC > 0.5)

## 1hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3d-norm-1hr-lfc05}
ip_4E_3d_norm_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (1hr)",
                           "All other genes"))

upreg_3d_norm_1hr_ip_lfc <- ip_4E_3d_norm_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (1hr)") %>%
  pull(log2FoldChange)

other_norm_1hr_ip_lfc <- ip_4E_3d_norm_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_norm_1hr_lfc05 <- ks.test(upreg_3d_norm_1hr_ip_lfc, other_norm_1hr_ip_lfc)

ip_4E_3d_norm_1hr_lfc05 <- ip_4E_3d_norm_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (1hr)",
                                   sprintf("Upregulated by eIF3d normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_1hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_1hr_ip_lfc))))

color_values_ip_3d_norm_1hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_1hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_1hr_ip_lfc)))
)

ggplot(ip_4E_3d_norm_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_norm_1hr_lfc05) +
  labs(title = "1hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_norm_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_norm_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr normoxia eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_norm_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3d-norm-4hr-lfc05}
ip_4E_3d_norm_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (4hr)",
                           "All other genes"))

upreg_3d_norm_4hr_ip_lfc <- ip_4E_3d_norm_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (4hr)") %>%
  pull(log2FoldChange)

other_norm_4hr_ip_lfc <- ip_4E_3d_norm_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_norm_4hr_lfc05 <- ks.test(upreg_3d_norm_4hr_ip_lfc, other_norm_4hr_ip_lfc)

ip_4E_3d_norm_4hr_lfc05 <- ip_4E_3d_norm_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (4hr)",
                                   sprintf("Upregulated by eIF3d normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_4hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_4hr_ip_lfc))))

color_values_ip_3d_norm_4hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_4hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_4hr_ip_lfc)))
)

ggplot(ip_4E_3d_norm_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_norm_4hr_lfc05) +
  labs(title = "4hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_norm_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_norm_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr normoxia eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_norm_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3d Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3d-norm-24hr-lfc05}
ip_4E_3d_norm_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3d_normoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3d normoxia (24hr)",
                           "All other genes"))

upreg_3d_norm_24hr_ip_lfc <- ip_4E_3d_norm_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3d normoxia (24hr)") %>%
  pull(log2FoldChange)

other_norm_24hr_ip_lfc <- ip_4E_3d_norm_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3d_norm_24hr_lfc05 <- ks.test(upreg_3d_norm_24hr_ip_lfc, other_norm_24hr_ip_lfc)

ip_4E_3d_norm_24hr_lfc05 <- ip_4E_3d_norm_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3d normoxia (24hr)",
                                   sprintf("Upregulated by eIF3d normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_24hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_norm_24hr_ip_lfc))))

color_values_ip_3d_norm_24hr_lfc05 <- setNames(
  c("#E69F00", "grey50"),
  c(sprintf("Upregulated by eIF3d normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3d_norm_24hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_norm_24hr_ip_lfc)))
)

ggplot(ip_4E_3d_norm_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3d_norm_24hr_lfc05) +
  labs(title = "24hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3d_norm_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3d_norm_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr normoxia eIF3d (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3d_norm_24hr_lfc05$p.value, "\n")
```

## 1hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3e-norm-1hr-lfc05}
ip_4E_3e_norm_1hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_1hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (1hr)",
                           "All other genes"))

upreg_3e_norm_1hr_ip_lfc <- ip_4E_3e_norm_1hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (1hr)") %>%
  pull(log2FoldChange)

other_3e_norm_1hr_ip_lfc <- ip_4E_3e_norm_1hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_norm_1hr_lfc05 <- ks.test(upreg_3e_norm_1hr_ip_lfc, other_3e_norm_1hr_ip_lfc)

ip_4E_3e_norm_1hr_lfc05 <- ip_4E_3e_norm_1hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (1hr)",
                                   sprintf("Upregulated by eIF3e normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_1hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_1hr_ip_lfc))))

color_values_ip_3e_norm_1hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (1hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_1hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_1hr_ip_lfc)))
)

ggplot(ip_4E_3e_norm_1hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_norm_1hr_lfc05) +
  labs(title = "1hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_norm_1hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_norm_1hr_lfc05.pdf"), width = 8, height = 6)

cat("1hr normoxia eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_norm_1hr_lfc05$p.value, "\n")
```

## 4hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3e-norm-4hr-lfc05}
ip_4E_3e_norm_4hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_4hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (4hr)",
                           "All other genes"))

upreg_3e_norm_4hr_ip_lfc <- ip_4E_3e_norm_4hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (4hr)") %>%
  pull(log2FoldChange)

other_3e_norm_4hr_ip_lfc <- ip_4E_3e_norm_4hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_norm_4hr_lfc05 <- ks.test(upreg_3e_norm_4hr_ip_lfc, other_3e_norm_4hr_ip_lfc)

ip_4E_3e_norm_4hr_lfc05 <- ip_4E_3e_norm_4hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (4hr)",
                                   sprintf("Upregulated by eIF3e normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_4hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_4hr_ip_lfc))))

color_values_ip_3e_norm_4hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (4hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_4hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_4hr_ip_lfc)))
)

ggplot(ip_4E_3e_norm_4hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_norm_4hr_lfc05) +
  labs(title = "4hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_norm_4hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_norm_4hr_lfc05.pdf"), width = 8, height = 6)

cat("4hr normoxia eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_norm_4hr_lfc05$p.value, "\n")
```

## 24hr - eIF3e Upregulated Genes in Normoxia (LFC > 0.5) - 4E-IP

```{r ip-density-3e-norm-24hr-lfc05}
ip_4E_3e_norm_24hr_lfc05 <- teleman_IP_4E_inhib_v_vehicle %>%
  mutate(category = ifelse(symbol %in% upreg_by_3e_normoxia_24hr_lfc05$symbol,
                           "Upregulated by eIF3e normoxia (24hr)",
                           "All other genes"))

upreg_3e_norm_24hr_ip_lfc <- ip_4E_3e_norm_24hr_lfc05 %>%
  filter(category == "Upregulated by eIF3e normoxia (24hr)") %>%
  pull(log2FoldChange)

other_3e_norm_24hr_ip_lfc <- ip_4E_3e_norm_24hr_lfc05 %>%
  filter(category == "All other genes") %>%
  pull(log2FoldChange)

ks_test_ip_3e_norm_24hr_lfc05 <- ks.test(upreg_3e_norm_24hr_ip_lfc, other_3e_norm_24hr_ip_lfc)

ip_4E_3e_norm_24hr_lfc05 <- ip_4E_3e_norm_24hr_lfc05 %>%
  mutate(category_labeled = ifelse(category == "Upregulated by eIF3e normoxia (24hr)",
                                   sprintf("Upregulated by eIF3e normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_24hr_ip_lfc)),
                                   sprintf("All other genes (n=%d)", length(other_3e_norm_24hr_ip_lfc))))

color_values_ip_3e_norm_24hr_lfc05 <- setNames(
  c("#56B4E9", "grey50"),
  c(sprintf("Upregulated by eIF3e normoxia (24hr, lfc>0.5) (n=%d)", length(upreg_3e_norm_24hr_ip_lfc)),
    sprintf("All other genes (n=%d)", length(other_3e_norm_24hr_ip_lfc)))
)

ggplot(ip_4E_3e_norm_24hr_lfc05, aes(x = log2FoldChange, color = category_labeled)) +
  geom_density(size = 1) +
  scale_color_manual(values = color_values_ip_3e_norm_24hr_lfc05) +
  labs(title = "24hr normoxia: 4E-IP Change upon 4E Inhibition (LFC > 0.5)",
       subtitle = sprintf("KS test p-value = %.2e", ks_test_ip_3e_norm_24hr_lfc05$p.value),
       x = "log2FC (4E-IP: inhibitor vs vehicle)",
       y = "Density",
       color = NULL) +
  theme_cowplot() +
  theme(legend.position = "top")

ggsave(here("output/teleman_ip_density_3e_norm_24hr_lfc05.pdf"), width = 8, height = 6)

cat("24hr normoxia eIF3e (lfc>0.5) - 4E-IP KS test p-value:", ks_test_ip_3e_norm_24hr_lfc05$p.value, "\n")
```

# 4E-IP Distribution Analysis - Vehicle Condition

```{r ip-density-vehicle-helper}
ip_density_plot <- function(ip_data, gene_symbols, cat_label, color, title_str, xlabel, outfile, xlim_range = NULL) {
  plot_data <- ip_data %>%
    mutate(category = ifelse(symbol %in% gene_symbols, cat_label, "All other genes"))

  upreg_lfc <- plot_data %>% filter(category == cat_label) %>% pull(log2FoldChange)
  other_lfc  <- plot_data %>% filter(category == "All other genes") %>% pull(log2FoldChange)

  ks_result <- suppressWarnings(ks.test(upreg_lfc, other_lfc))

  plot_data <- plot_data %>%
    mutate(category_labeled = ifelse(category == cat_label,
                                     sprintf("%s (n=%d)", cat_label, length(upreg_lfc)),
                                     sprintf("All other genes (n=%d)", length(other_lfc))))

  color_vals <- setNames(
    c(color, "grey50"),
    c(sprintf("%s (n=%d)", cat_label, length(upreg_lfc)),
      sprintf("All other genes (n=%d)", length(other_lfc)))
  )

  p <- ggplot(plot_data, aes(x = log2FoldChange, color = category_labeled)) +
    geom_density(size = 1) +
    scale_color_manual(values = color_vals) +
    labs(title = title_str,
         subtitle = sprintf("KS test p-value = %.2e", ks_result$p.value),
         x = xlabel,
         y = "Density",
         color = NULL) +
    theme_cowplot() +
    theme(legend.position = "top")

  if (!is.null(xlim_range)) p <- p + xlim(xlim_range[1], xlim_range[2])

  print(p)
  ggsave(here(paste0("output/", outfile)), width = 8, height = 6)
  cat(title_str, "- KS p:", ks_result$p.value, "\n")
}
```


## Hypoxia - eIF3d/3e Upregulated Genes (vehicle, LFC > 0.5)

```{r ip-density-vehicle-3d-hyp-1hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_hypoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3d (1hr, lfc>0.5)", "#E69F00",
  "1hr hypoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_1hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3d-hyp-4hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_hypoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3d (4hr, lfc>0.5)", "#E69F00",
  "4hr hypoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_4hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3d-hyp-24hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_hypoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3d (24hr, lfc>0.5)", "#E69F00",
  "24hr hypoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_24hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-hyp-1hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_hypoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3e (1hr, lfc>0.5)", "#56B4E9",
  "1hr hypoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_1hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-hyp-4hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_hypoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3e (4hr, lfc>0.5)", "#56B4E9",
  "4hr hypoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_4hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-hyp-24hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_hypoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3e (24hr, lfc>0.5)", "#56B4E9",
  "24hr hypoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_24hr_lfc05_vehicle.pdf")
```

## Normoxia - eIF3d/3e Upregulated Genes (vehicle, LFC > 0.5)

```{r ip-density-vehicle-3d-norm-1hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_normoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (1hr, lfc>0.5)", "#E69F00",
  "1hr normoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_norm_1hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3d-norm-4hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_normoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (4hr, lfc>0.5)", "#E69F00",
  "4hr normoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_norm_4hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3d-norm-24hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3d_normoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (24hr, lfc>0.5)", "#E69F00",
  "24hr normoxia: 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_norm_24hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-norm-1hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_normoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (1hr, lfc>0.5)", "#56B4E9",
  "1hr normoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_norm_1hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-norm-4hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_normoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (4hr, lfc>0.5)", "#56B4E9",
  "4hr normoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_norm_4hr_lfc05_vehicle.pdf")
```

```{r ip-density-vehicle-3e-norm-24hr-lfc05}
ip_density_plot(teleman_IP_4E_vehicle, upreg_by_3e_normoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (24hr, lfc>0.5)", "#56B4E9",
  "24hr normoxia: 4E-IP Vehicle (eIF3e, LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3e_norm_24hr_lfc05_vehicle.pdf")
```

## Normoxia - eIF3d Time-Regressed 1+4hr (vehicle, LFC > 0.5)

```{r ip-density-vehicle-3d-norm-1and4hr-lfc05}
norm_1and4hr_ids <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) %>%
  filter(te_lfc > 0.5, te_padj < 0.05) %>%
  pull(symbol)

ip_density_plot(teleman_IP_4E_vehicle, norm_1and4hr_ids,
  "Upregulated by eIF3d normoxia (1+4hr time-regressed, lfc>0.5)", "#E69F00",
  "Normoxia 1+4hr (time-regressed): 4E-IP Vehicle (LFC > 0.5)", "log2FC (4E-IP: vehicle)",
  "teleman_ip_density_3d_norm_1and4hr_lfc05_vehicle.pdf")
```

# 4E-IP Distribution Analysis - Inhibitor Condition


## Hypoxia - eIF3d/3e Upregulated Genes (inhibitor, LFC > 0.5)

```{r ip-density-inhib-3d-hyp-1hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_hypoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3d (1hr, lfc>0.5)", "#E69F00",
  "1hr hypoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_1hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3d-hyp-4hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_hypoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3d (4hr, lfc>0.5)", "#E69F00",
  "4hr hypoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_4hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3d-hyp-24hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_hypoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3d (24hr, lfc>0.5)", "#E69F00",
  "24hr hypoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_24hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-hyp-1hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_hypoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3e (1hr, lfc>0.5)", "#56B4E9",
  "1hr hypoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_1hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-hyp-4hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_hypoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3e (4hr, lfc>0.5)", "#56B4E9",
  "4hr hypoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_4hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-hyp-24hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_hypoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3e (24hr, lfc>0.5)", "#56B4E9",
  "24hr hypoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_24hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

## Normoxia - eIF3d/3e Upregulated Genes (inhibitor, LFC > 0.5)

```{r ip-density-inhib-3d-norm-1hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_normoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (1hr, lfc>0.5)", "#E69F00",
  "1hr normoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_norm_1hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3d-norm-4hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_normoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (4hr, lfc>0.5)", "#E69F00",
  "4hr normoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_norm_4hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3d-norm-24hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3d_normoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3d normoxia (24hr, lfc>0.5)", "#E69F00",
  "24hr normoxia: 4E-IP Inhibitor (LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3d_norm_24hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-norm-1hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_normoxia_1hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (1hr, lfc>0.5)", "#56B4E9",
  "1hr normoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_norm_1hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-norm-4hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_normoxia_4hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (4hr, lfc>0.5)", "#56B4E9",
  "4hr normoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_norm_4hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

```{r ip-density-inhib-3e-norm-24hr-lfc05}
ip_density_plot(teleman_IP_4E_inhib, upreg_by_3e_normoxia_24hr_lfc05$symbol,
  "Upregulated by eIF3e normoxia (24hr, lfc>0.5)", "#56B4E9",
  "24hr normoxia: 4E-IP Inhibitor (eIF3e, LFC > 0.5)", "log2FC (4E-IP: inhibitor)",
  "teleman_ip_density_3e_norm_24hr_lfc05_inhib.pdf", xlim_range = c(-2, 2))
```

# 3d te_lfc Distribution Colored by 4E Status

For each condition, plot the full distribution of eIF3d te_lfc values for all genes, colored
by their 4E inhibition status from Teleman: resistant (LFC > 1, padj < 0.05),
sensitive (LFC < -1, padj < 0.05), or other.

## Define 4E status categories

```{r define-4e-status}
# Sensitive defined symmetrically to resistant: LFC < -1, padj < 0.05
sensitive_4E <- res_riboseq_symbol %>%
  filter(log2FoldChange < -1) %>%
  filter(padj < 0.05) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("4E-resistant genes:", nrow(resistant_4E), "\n")
cat("4E-sensitive genes:", nrow(sensitive_4E), "\n")

# Build a per-symbol lookup table
teleman_4e_status <- res_riboseq_symbol %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(status_4e = case_when(
    log2FoldChange > 1  & padj < 0.05 ~ "4E-resistant",
    log2FoldChange < -1 & padj < 0.05 ~ "4E-sensitive",
    TRUE ~ "Other"
  )) %>%
  dplyr::select(symbol, status_4e)

cat("4E status counts:\n")
print(table(teleman_4e_status$status_4e))
```

## Helper function

```{r 4e-status-density-function}
plot_3d_telfc_by_4e_status <- function(data_3d, title_str, outfile) {
  status_levels <- c("4E-resistant", "4E-sensitive", "Other")

  plot_data <- data_3d %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    left_join(teleman_4e_status, by = "symbol") %>%
    mutate(status_4e = replace_na(status_4e, "Other"),
           status_4e = factor(status_4e, levels = status_levels))

  n_resist <- sum(plot_data$status_4e == "4E-resistant")
  n_sensit <- sum(plot_data$status_4e == "4E-sensitive")
  n_other  <- sum(plot_data$status_4e == "Other")

  ks_res_vs_other <- suppressWarnings(
    ks.test(
      plot_data$te_lfc[plot_data$status_4e == "4E-resistant"],
      plot_data$te_lfc[plot_data$status_4e != "4E-resistant"]
    )
  )

  ks_sens_vs_other <- suppressWarnings(
    ks.test(
      plot_data$te_lfc[plot_data$status_4e == "4E-sensitive"],
      plot_data$te_lfc[plot_data$status_4e != "4E-sensitive"]
    )
  )

  plot_data <- plot_data %>%
    mutate(status_labeled = case_when(
      status_4e == "4E-resistant" ~ sprintf("4E-resistant (n=%d)", n_resist),
      status_4e == "4E-sensitive" ~ sprintf("4E-sensitive (n=%d)", n_sensit),
      TRUE                        ~ sprintf("Other (n=%d)", n_other)
    ),
    status_labeled = factor(status_labeled,
      levels = c(sprintf("4E-resistant (n=%d)", n_resist),
                 sprintf("4E-sensitive (n=%d)", n_sensit),
                 sprintf("Other (n=%d)", n_other))))

  color_vals <- setNames(
    c("#009E73", "#D55E00", "grey70"),
    c(sprintf("4E-resistant (n=%d)", n_resist),
      sprintf("4E-sensitive (n=%d)", n_sensit),
      sprintf("Other (n=%d)", n_other))
  )

  p <- ggplot(plot_data, aes(x = te_lfc, color = status_labeled)) +
    stat_ecdf(size = 1) +
    scale_color_manual(values = color_vals) +
    labs(title = title_str,
         subtitle = sprintf("KS: resistant vs other p = %.2e; sensitive vs other p = %.2e",
                            ks_res_vs_other$p.value, ks_sens_vs_other$p.value),
         x = "te_lfc (eIF3d effect on translation efficiency)",
         y = "Cumulative proportion",
         color = "4E inhibition status") +
    theme_cowplot() +
    theme(legend.position = "top")+
    xlim(-1,1)

  print(p)
  ggsave(here(paste0("output/", outfile)), width = 8, height = 6)

  cat(title_str, "- KS resistant vs other p:", ks_res_vs_other$p.value, "\n")
  cat(title_str, "- KS sensitive vs other p:", ks_sens_vs_other$p.value, "\n")
  cat("  Median te_lfc resistant:", median(plot_data$te_lfc[plot_data$status_4e == "4E-resistant"], na.rm = TRUE), "\n")
  cat("  Median te_lfc sensitive:", median(plot_data$te_lfc[plot_data$status_4e == "4E-sensitive"], na.rm = TRUE), "\n")
  cat("  Median te_lfc other:",     median(plot_data$te_lfc[plot_data$status_4e == "Other"],         na.rm = TRUE), "\n")
}
```

## Hypoxia - eIF3d

```{r 4e-status-3d-hypoxia-1hr}
plot_3d_telfc_by_4e_status(hyp_3d_1hr,  "Hypoxia 1hr: eIF3d te_lfc by 4E status",  "telfc_4estatus_3d_hyp_1hr.pdf")
```

```{r 4e-status-3d-hypoxia-4hr}
plot_3d_telfc_by_4e_status(hyp_3d_4hr,  "Hypoxia 4hr: eIF3d te_lfc by 4E status",  "telfc_4estatus_3d_hyp_4hr.pdf")
```

```{r 4e-status-3d-hypoxia-24hr}
plot_3d_telfc_by_4e_status(hyp_3d_24hr, "Hypoxia 24hr: eIF3d te_lfc by 4E status", "telfc_4estatus_3d_hyp_24hr.pdf")
```

## Normoxia - eIF3d

```{r 4e-status-3d-normoxia-1hr}
plot_3d_telfc_by_4e_status(norm_3d_1hr,  "Normoxia 1hr: eIF3d te_lfc by 4E status",  "telfc_4estatus_3d_norm_1hr.pdf")
```

```{r 4e-status-3d-normoxia-4hr}
plot_3d_telfc_by_4e_status(norm_3d_4hr,  "Normoxia 4hr: eIF3d te_lfc by 4E status",  "telfc_4estatus_3d_norm_4hr.pdf")
```

```{r 4e-status-3d-normoxia-24hr}
plot_3d_telfc_by_4e_status(norm_3d_24hr, "Normoxia 24hr: eIF3d te_lfc by 4E status", "telfc_4estatus_3d_norm_24hr.pdf")
```

# ECDF of te_lfc colored by IP enrichment status

## Build IP status lookup tables

```{r ip-status-inhib-v-vehicle}
ip_status_inhib_v_vehicle <- teleman_IP_4E_inhib_v_vehicle %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ip_status = case_when(
    log2FoldChange > 1  & padj < 0.05 ~ "IP-enriched",
    log2FoldChange < -1 & padj < 0.05 ~ "IP-depleted",
    TRUE ~ "Other"
  )) %>%
  dplyr::select(symbol, ip_status)

cat("IP status counts (inhib vs vehicle):\n")
print(table(ip_status_inhib_v_vehicle$ip_status))
```

```{r ip-status-vehicle}
ip_status_vehicle <- teleman_IP_4E_vehicle %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ip_status = case_when(
    log2FoldChange > 1  & padj < 0.05 ~ "IP-enriched",
    log2FoldChange < -1 & padj < 0.05 ~ "IP-depleted",
    TRUE ~ "Other"
  )) %>%
  dplyr::select(symbol, ip_status)

cat("IP status counts (vehicle):\n")
print(table(ip_status_vehicle$ip_status))
```

```{r ip-status-inhib}
ip_status_inhib <- teleman_IP_4E_inhib %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(ip_status = case_when(
    log2FoldChange > 1  & padj < 0.05 ~ "IP-enriched",
    log2FoldChange < -1 & padj < 0.05 ~ "IP-depleted",
    TRUE ~ "Other"
  )) %>%
  dplyr::select(symbol, ip_status)

cat("IP status counts (inhib):\n")
print(table(ip_status_inhib$ip_status))
```

## Helper function

```{r ip-status-ecdf-function}
plot_telfc_by_ip_status <- function(data_3d, ip_status_tbl, title_str, outfile) {
  status_levels <- c("IP-enriched", "IP-depleted", "Other")

  plot_data <- data_3d %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    left_join(ip_status_tbl, by = "symbol") %>%
    mutate(ip_status = replace_na(ip_status, "Other"),
           ip_status = factor(ip_status, levels = status_levels))

  n_enriched <- sum(plot_data$ip_status == "IP-enriched")
  n_depleted <- sum(plot_data$ip_status == "IP-depleted")
  n_other    <- sum(plot_data$ip_status == "Other")

  ks_enrich_vs_other <- suppressWarnings(
    ks.test(
      plot_data$te_lfc[plot_data$ip_status == "IP-enriched"],
      plot_data$te_lfc[plot_data$ip_status != "IP-enriched"]
    )
  )

  ks_deplete_vs_other <- suppressWarnings(
    ks.test(
      plot_data$te_lfc[plot_data$ip_status == "IP-depleted"],
      plot_data$te_lfc[plot_data$ip_status != "IP-depleted"]
    )
  )

  plot_data <- plot_data %>%
    mutate(status_labeled = case_when(
      ip_status == "IP-enriched" ~ sprintf("IP-enriched (n=%d)", n_enriched),
      ip_status == "IP-depleted" ~ sprintf("IP-depleted (n=%d)", n_depleted),
      TRUE                       ~ sprintf("Other (n=%d)", n_other)
    ),
    status_labeled = factor(status_labeled,
      levels = c(sprintf("IP-enriched (n=%d)", n_enriched),
                 sprintf("IP-depleted (n=%d)", n_depleted),
                 sprintf("Other (n=%d)", n_other))))

  color_vals <- setNames(
    c("#009E73", "#D55E00", "grey70"),
    c(sprintf("IP-enriched (n=%d)", n_enriched),
      sprintf("IP-depleted (n=%d)", n_depleted),
      sprintf("Other (n=%d)", n_other))
  )

  p <- ggplot(plot_data, aes(x = te_lfc, color = status_labeled)) +
    stat_ecdf(size = 1) +
    scale_color_manual(values = color_vals) +
    labs(title = title_str,
         subtitle = sprintf("KS: enriched vs other p = %.2e; depleted vs other p = %.2e",
                            ks_enrich_vs_other$p.value, ks_deplete_vs_other$p.value),
         x = "te_lfc (eIF3d effect on translation efficiency)",
         y = "Cumulative proportion",
         color = "IP enrichment status") +
    theme_cowplot() +
    theme(legend.position = "top") +
    xlim(-1, 1)

  print(p)
  ggsave(here(paste0("output/", outfile)), width = 8, height = 6)

  cat(title_str, "- KS enriched vs other p:", ks_enrich_vs_other$p.value, "\n")
  cat(title_str, "- KS depleted vs other p:", ks_deplete_vs_other$p.value, "\n")
  cat("  Median te_lfc enriched:", median(plot_data$te_lfc[plot_data$ip_status == "IP-enriched"], na.rm = TRUE), "\n")
  cat("  Median te_lfc depleted:", median(plot_data$te_lfc[plot_data$ip_status == "IP-depleted"], na.rm = TRUE), "\n")
  cat("  Median te_lfc other:",    median(plot_data$te_lfc[plot_data$ip_status == "Other"],        na.rm = TRUE), "\n")
}
```

## Hypoxia - eIF3d × IP status

### Inhibitor vs vehicle

```{r ip-status-inhibvvehicle-hyp-1hr}
plot_telfc_by_ip_status(hyp_3d_1hr,  ip_status_inhib_v_vehicle, "Hypoxia 1hr: eIF3d te_lfc by IP status (inhib vs vehicle)",  "telfc_ipstatus_inhibvvehicle_3d_hyp_1hr.pdf")
```

```{r ip-status-inhibvvehicle-hyp-4hr}
plot_telfc_by_ip_status(hyp_3d_4hr,  ip_status_inhib_v_vehicle, "Hypoxia 4hr: eIF3d te_lfc by IP status (inhib vs vehicle)",  "telfc_ipstatus_inhibvvehicle_3d_hyp_4hr.pdf")
```

```{r ip-status-inhibvvehicle-hyp-24hr}
plot_telfc_by_ip_status(hyp_3d_24hr, ip_status_inhib_v_vehicle, "Hypoxia 24hr: eIF3d te_lfc by IP status (inhib vs vehicle)", "telfc_ipstatus_inhibvvehicle_3d_hyp_24hr.pdf")
```

### Vehicle

```{r ip-status-vehicle-hyp-1hr}
plot_telfc_by_ip_status(hyp_3d_1hr,  ip_status_vehicle, "Hypoxia 1hr: eIF3d te_lfc by IP status (vehicle)",  "telfc_ipstatus_vehicle_3d_hyp_1hr.pdf")
```

```{r ip-status-vehicle-hyp-4hr}
plot_telfc_by_ip_status(hyp_3d_4hr,  ip_status_vehicle, "Hypoxia 4hr: eIF3d te_lfc by IP status (vehicle)",  "telfc_ipstatus_vehicle_3d_hyp_4hr.pdf")
```

```{r ip-status-vehicle-hyp-24hr}
plot_telfc_by_ip_status(hyp_3d_24hr, ip_status_vehicle, "Hypoxia 24hr: eIF3d te_lfc by IP status (vehicle)", "telfc_ipstatus_vehicle_3d_hyp_24hr.pdf")
```

### Inhibitor

```{r ip-status-inhib-hyp-1hr}
plot_telfc_by_ip_status(hyp_3d_1hr,  ip_status_inhib, "Hypoxia 1hr: eIF3d te_lfc by IP status (inhib)",  "telfc_ipstatus_inhib_3d_hyp_1hr.pdf")
```

```{r ip-status-inhib-hyp-4hr}
plot_telfc_by_ip_status(hyp_3d_4hr,  ip_status_inhib, "Hypoxia 4hr: eIF3d te_lfc by IP status (inhib)",  "telfc_ipstatus_inhib_3d_hyp_4hr.pdf")
```

```{r ip-status-inhib-hyp-24hr}
plot_telfc_by_ip_status(hyp_3d_24hr, ip_status_inhib, "Hypoxia 24hr: eIF3d te_lfc by IP status (inhib)", "telfc_ipstatus_inhib_3d_hyp_24hr.pdf")
```

## Normoxia - eIF3d × IP status

### Inhibitor vs vehicle

```{r ip-status-inhibvvehicle-norm-1hr}
plot_telfc_by_ip_status(norm_3d_1hr,  ip_status_inhib_v_vehicle, "Normoxia 1hr: eIF3d te_lfc by IP status (inhib vs vehicle)",  "telfc_ipstatus_inhibvvehicle_3d_norm_1hr.pdf")
```

```{r ip-status-inhibvvehicle-norm-4hr}
plot_telfc_by_ip_status(norm_3d_4hr,  ip_status_inhib_v_vehicle, "Normoxia 4hr: eIF3d te_lfc by IP status (inhib vs vehicle)",  "telfc_ipstatus_inhibvvehicle_3d_norm_4hr.pdf")
```

```{r ip-status-inhibvvehicle-norm-24hr}
plot_telfc_by_ip_status(norm_3d_24hr, ip_status_inhib_v_vehicle, "Normoxia 24hr: eIF3d te_lfc by IP status (inhib vs vehicle)", "telfc_ipstatus_inhibvvehicle_3d_norm_24hr.pdf")
```

### Vehicle

```{r ip-status-vehicle-norm-1hr}
plot_telfc_by_ip_status(norm_3d_1hr,  ip_status_vehicle, "Normoxia 1hr: eIF3d te_lfc by IP status (vehicle)",  "telfc_ipstatus_vehicle_3d_norm_1hr.pdf")
```

```{r ip-status-vehicle-norm-4hr}
plot_telfc_by_ip_status(norm_3d_4hr,  ip_status_vehicle, "Normoxia 4hr: eIF3d te_lfc by IP status (vehicle)",  "telfc_ipstatus_vehicle_3d_norm_4hr.pdf")
```

```{r ip-status-vehicle-norm-24hr}
plot_telfc_by_ip_status(norm_3d_24hr, ip_status_vehicle, "Normoxia 24hr: eIF3d te_lfc by IP status (vehicle)", "telfc_ipstatus_vehicle_3d_norm_24hr.pdf")
```

### Inhibitor

```{r ip-status-inhib-norm-1hr}
plot_telfc_by_ip_status(norm_3d_1hr,  ip_status_inhib, "Normoxia 1hr: eIF3d te_lfc by IP status (inhib)",  "telfc_ipstatus_inhib_3d_norm_1hr.pdf")
```

```{r ip-status-inhib-norm-4hr}
plot_telfc_by_ip_status(norm_3d_4hr,  ip_status_inhib, "Normoxia 4hr: eIF3d te_lfc by IP status (inhib)",  "telfc_ipstatus_inhib_3d_norm_4hr.pdf")
```

```{r ip-status-inhib-norm-24hr}
plot_telfc_by_ip_status(norm_3d_24hr, ip_status_inhib, "Normoxia 24hr: eIF3d te_lfc by IP status (inhib)", "telfc_ipstatus_inhib_3d_norm_24hr.pdf")
```