---
title: "timecourse 231"
author: "Kate"
date: "2025-11_20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(EnhancedVolcano)
library(msigdbr)
library(viridis)
library(gridExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(survival)
library(DESeq2)
library(limma)
library(clusterProfiler)
library(plotly)
library(ggokabeito)
library(ggvenn)
library(ggpubr)
library(DT)
library(edgeR)
library(cowplot)
library(ggExtra)
library(ggprism)
library(GeneOverlap)
library(writexl)
library(survminer)
library(readxl)
library(reshape2)
```

```{r create_output_dir}
# Create output directory if it doesn't exist
if(!dir.exists(here("output"))) {
  dir.create(here("output"))
  cat("Created output directory\n")
} else {
  cat("Output directory already exists\n")
}
```

# Functions

```{r load_functions}
cat("Loading functions from functions.R...\n")
source(here("code/functions.R"))
cat("Functions loaded successfully!\n")
cat("Available functions: createNormMatrix, deFunction\n")

# createNormMatrix <- function(allCounts, spikein = NULL, keepSpikes = NULL) {
#   # 1) coerce & round base counts
#   allCounts <- round(as.matrix(allCounts), 0)

#   # 2) detect empty/absent spike-ins
#   has_spike <- !is.null(spikein) &&
#                ((is.matrix(spikein) || is.data.frame(spikein)) && nrow(spikein) > 0)

#   if (has_spike) {
#     spikein <- round(as.matrix(spikein), 0)

#     # ensure sample columns align
#     if (!identical(colnames(allCounts), colnames(spikein))) {
#       if (!all(colnames(allCounts) %in% colnames(spikein)))
#         stop("spikein is missing some sample columns found in allCounts.")
#       spikein <- spikein[, colnames(allCounts), drop = FALSE]
#     }

#     # combine human + spike-ins (rows)
#     counts_all <- rbind(allCounts, spikein)

#     # indices of spike-in rows, if provided
#     Spikeindices <- if (!is.null(keepSpikes) && length(keepSpikes) > 0) {
#       which(rownames(counts_all) %in% keepSpikes)
#     } else {
#       NULL
#     }
#   } else {
#     # no spike-ins: use human-only counts and default normalization
#     counts_all   <- allCounts
#     Spikeindices <- NULL
#   }

#   # 3) sample metadata (kept as close to yours as possible)
#   sample_info <- metadata %>%
#     dplyr::select(type, tx, rep, cell) %>%
#     unique()
#   colnames(sample_info) <- c("SeqType","Condition","Batch","Cell")
#   sample_info$Condition <- stats::relevel(as.factor(sample_info$Condition), ref = "CTRL")
#   sample_info$SeqType   <- stats::relevel(as.factor(sample_info$SeqType),   ref = "rna")
#   # (DESeq2 will set rownames(colData) to colnames(counts_all) if they’re NULL)

#   # 4) DESeq2 object + size factors
#   dds <- DESeq2::DESeqDataSetFromMatrix(
#     countData = counts_all,
#     colData   = sample_info,
#     design    = ~ Condition + SeqType + Condition:SeqType
#   )

#   # IMPORTANT: default = median-ratio (no poscounts)
#   if (is.null(Spikeindices) || length(Spikeindices) == 0) {
#     dds <- DESeq2::estimateSizeFactors(dds)  # type="ratio" by default
#   } else {
#     dds <- DESeq2::estimateSizeFactors(dds, controlGenes = Spikeindices)  # still "ratio"
#   }

#   # 5) normalized counts (DESeq2 normalization, not CPM)
#   norm <- DESeq2::counts(dds, normalized = TRUE)

#   plotData <- norm %>%
#     as.data.frame() %>%
#     tibble::rownames_to_column(var = "gene_id") %>%
#     dplyr::inner_join(gene_symbol, .)

#   return(norm)
# }

```
# Initial QC

```{r}

ribo <- read_csv(file = "2025_MDAcells_riboseq/QC_summary_ribo.csv") %>% dplyr::select(Library_ID, Total_Reads_Start,Percent_Original_Reads_Used, "Frags_Mapped")

colnames(ribo) <- c("id","reads","percent_reads_used", "frags_mapped")

# total_ribo_reads_used <- ribo |> 
#     mutate(total_reads_used = reads * (percent_reads_used*.01))



ribo_long <- reshape2::melt(ribo, variable.name = "stage")

p_ribo_reads <- ggbarplot(data = ribo_long |> filter(stage == "reads") , x = "id", y = "value", position = position_dodge()) +
  ylab("ribo-seq reads") +
  theme_prism() +
  rotate_x_text(angle = 90) +
scale_y_log10(breaks = 10^seq(1:8), labels = c("10", "100", "1K", "10K", "100K", "1M", "10M", "100M")) +
annotate("rect", ymin = 10^7, ymax = Inf, xmin = 0, xmax = Inf, fill = "green", alpha = 0.2) +
annotate("rect", ymin = 10^6, ymax = 10^7, xmin = 0, xmax = Inf, fill = "orange", alpha = 0.2) + 
annotate("rect", ymin = 0, ymax = 10^6, xmin = 0, xmax = Inf, fill = "red", alpha = 0.2)


p_ribo_reads + ggtitle("ribo-seq reads")
ggsave(here("plots/riboseq_reads.png"), width=11, height = 5)

p_ribo_percent <- ggbarplot(data = ribo_long |> filter(stage == "percent_reads_used") , x = "id", y = "value", position = position_dodge()) + 
  ylab("% ribo-seq reads") +
  theme_prism() +
  rotate_x_text(angle = 90) +
  annotate("rect", ymin = 20, ymax = Inf, xmin = 0, xmax = Inf, fill = "green", alpha = 0.2) + 
    annotate("rect", ymin = 0, ymax = 20, xmin = 0, xmax = Inf, fill = "red", alpha = 0.2)


p_ribo_percent + ggtitle("ribo-seq % used")
ggsave(here("plots/riboseq_percent.png"), width=11, height = 5)



# reads into salmon

p_ribo_reads_salmon <- ggbarplot(data = ribo_long |> filter(stage == "frags_mapped") , x = "id", y = "value", position = position_dodge()) +
  ylab("ribo-seq reads into salmon") +
  theme_prism() +
  rotate_x_text(angle = 90) +
scale_y_log10(breaks = 10^seq(1:8), labels = c("10", "100", "1K", "10K", "100K", "1M", "10M", "100M")) 
# +
# annotate("rect", ymin = 10^7, ymax = Inf, xmin = 0, xmax = Inf, fill = "green", alpha = 0.2) +
# annotate("rect", ymin = 10^6, ymax = 10^7, xmin = 0, xmax = Inf, fill = "orange", alpha = 0.2) + 
# annotate("rect", ymin = 0, ymax = 10^6, xmin = 0, xmax = Inf, fill = "red", alpha = 0.2)


p_ribo_reads_salmon + ggtitle("frags_mapped")
ggsave(here("plots/ribo_seq_frags_mapped.png"), width=11, height = 5)


```

```{r}

rna <- read_csv(file = "2025_MDAcells_riboseq/QC_summary_rna.csv") %>% dplyr::select(Library_ID, Total_Reads_Start, Percent_Original_Reads_Used)

colnames(rna) <- c("id","reads","percent_reads_used")

rna_long <- reshape2::melt(rna, variable.name = "stage")

p_rna_reads <- ggbarplot(data = rna_long |> filter(stage != "percent_reads_used") , x = "id", y = "value", position = position_dodge()) +
  ylab("rna-seq reads") +
  theme_prism() +
  rotate_x_text(angle = 90) +
scale_y_log10(breaks = 10^seq(1:8), labels = c("10", "100", "1K", "10K", "100K", "1M", "10M", "100M")) +
annotate("rect", ymin = 10^7, ymax = Inf, xmin = 0, xmax = Inf, fill = "green", alpha = 0.2) +
annotate("rect", ymin = 10^6, ymax = 10^7, xmin = 0, xmax = Inf, fill = "orange", alpha = 0.2) + 
annotate("rect", ymin = 0, ymax = 10^6, xmin = 0, xmax = Inf, fill = "red", alpha = 0.2)

p_rna_reads + ggtitle("rna-seq reads")
ggsave(here("plots/rnaseq_reads.png"), width=11, height = 5)


p_rna_percent <- ggbarplot(data = rna_long |> filter(stage == "percent_reads_used") , x = "id", y = "value", position = position_dodge()) + 
  ylab("% rna-seq reads") +
  theme_prism() +
  rotate_x_text(angle = 90) +
  annotate("rect", ymin = 70, ymax = Inf, xmin = 0, xmax = Inf, fill = "green", alpha = 0.2) +
  annotate("rect", ymin = 50, ymax = 70, xmin = 0, xmax = Inf, fill = "orange", alpha = 0.2) + 
  annotate("rect", ymin = 0, ymax = 50, xmin = 0, xmax = Inf, fill = "red", alpha = 0.2)


p_rna_percent + ggtitle("rna-seq % used")
ggsave(here("plots/rnaseq_percent.png"), width=11, height = 5)




```


```{r}
si3e_col = "#56B4E9"
si3d_col = "#E69F00"
```

## Load transcript-level data

Importing metadata and all salmon counts
```{r meta data}
metadata <- read_csv(here("accessories","metadata_231_timecourse.csv"), skip = 0) %>% 
  clean_names()

head(metadata)
unique(metadata$type)

metadata <- metadata |>
  rename(library = `type`)


metadata <- metadata |>
  dplyr::mutate(type = dplyr::recode(library,
                                      "RiboSeq"="ribo")) |> 
                                          dplyr::mutate(library = dplyr::recode(type,
                                      "Input_Ribo_RNAseq"="rna"))

#ribosome lysate = ribosome seq
#rRNA depletion = RNA seq
#rename variables remove outlier

#edit factor levels
metadata$library <- relevel(x = factor(metadata$library), ref = "rna")

metadata$treatment <- relevel(x = factor(metadata$treatment), ref = "siCTRL")
metadata$treatment <- recode(metadata$treatment, "siCTRL" = "sictrl")

metadata$condition <- relevel(x = factor(metadata$condition), ref = "normoxia")
metadata$rep <- recode(metadata$rep, RepA = "a", RepB = "b", RepC = "c")

```

```{r import gene info}

#import key on RNA type, remove premRNA
geneInfo <- read_tsv(here("accessories", "human","gene_anno_hs_dm_v49_r111.tsv.gz"), col_names = F) 
head(geneInfo)
colnames(geneInfo) <- c("gene_id","symbol","biotype","species", "version")

geneInfo <- geneInfo %>%
  filter(!str_detect(biotype, "pre_"))


tx2gene <- read_tsv(here("accessories/human/tx2gene_hs_dm_v49_r111.tsv.gz"), col_names="FALSE")
colnames(tx2gene) <- c("transcript", "gene")

#import transcripts for gene to transcript mapping

gene_symbol <- geneInfo %>% dplyr::select(gene_id, symbol) %>% unique()
head(gene_symbol)


# Check for duplicate gene_ids
duplicate_genes <- gene_symbol %>%
  group_by(gene_id) %>%
  filter(n() > 1) %>%
  arrange(gene_id)

if(nrow(duplicate_genes) > 0) {
  cat("Found", nrow(duplicate_genes), "duplicate gene_id entries in gene_symbol:\n")
  print(duplicate_genes)
  # Keep only first occurrence of each gene_id
  gene_symbol <- gene_symbol %>%
    distinct(gene_id, .keep_all = TRUE)
  cat("After removing duplicates: Total genes:", nrow(gene_symbol), "\n")
} else {
  cat("No duplicate gene_ids found in gene_symbol\n")
}

# Verify gene_symbol has unique gene_ids
cat("Total genes in gene_symbol:", nrow(gene_symbol), "\n")
cat("Unique gene_ids:", length(unique(gene_symbol$gene_id)), "\n")

```

```{r txi import, include=FALSE}
#name files to import
myquantfiles <- paste(here(), "/2025_MDAcells_riboseq/salmon/",
                      metadata$id,
                      "/quant.sf.gz",
                      sep = "")

names(myquantfiles) <- paste(metadata$condition, metadata$treatment, metadata$hours, metadata$rep, metadata$library, metadata$id, sep = "_")

#import salmon files
myTxi <- tximport(myquantfiles, type = "salmon", tx2gene = tx2gene)

# Check for duplicate rownames in counts
cat("Rows in myTxi$counts:", nrow(myTxi$counts), "\n")
cat("Unique rownames:", length(unique(rownames(myTxi$counts))), "\n")

if(any(duplicated(rownames(myTxi$counts)))) {
  dup_genes <- rownames(myTxi$counts)[duplicated(rownames(myTxi$counts))]
  cat("WARNING: Found", length(dup_genes), "duplicate gene IDs in myTxi$counts:\n")
  print(head(dup_genes, 20))
}
```

separate out human and dros genes


```{r}
#separate dros and human counts
counts <- myTxi$counts

is_dros <- grepl("^FB", rownames(counts))

cts_dros  <- counts[ is_dros, , drop = FALSE] # <- keep for separate heatmap
cts_human <- counts[!is_dros, , drop = FALSE]


# Separate RNA and Ribo samples
rna_samples <- grep("rna", colnames(cts_human), ignore.case = TRUE)
ribo_samples <- grep("ribo", colnames(cts_human), ignore.case = TRUE)

cts_human_rna <- cts_human[, rna_samples, drop = FALSE]
cts_human_ribo <- cts_human[, ribo_samples, drop = FALSE]

# Look at distribution of sum of counts for RNA
# to determine threshold for calling gene expressed
hist(log2(rowSums(cts_human_rna)), breaks = 60, main = "RNA: log2(rowSums) distribution")

# Look at distribution of sum of counts for Ribo
hist(log2(rowSums(cts_human_ribo)), breaks = 60, main = "Ribo: log2(rowSums) distribution")

# Determine genes that pass threshold in RNA and Ribo separately
keepGenes_rna <- rownames(cts_human_rna[log2(rowSums(cts_human_rna)) > 9,])
keepGenes_ribo <- rownames(cts_human_ribo[log2(rowSums(cts_human_ribo)) > 8,])

# Keep only genes that pass in BOTH RNA and Ribo
keepGenes <- intersect(keepGenes_rna, keepGenes_ribo)

cat("Genes passing threshold in RNA:", length(keepGenes_rna), "\n")
cat("Genes passing threshold in Ribo:", length(keepGenes_ribo), "\n")
cat("Genes passing threshold in BOTH:", length(keepGenes), "\n")


pc <-geneInfo %>%
  filter(biotype =="protein_coding") %>% 
  pull(gene_id) %>% 
  unique()

pc_keep <- pc[pc %in% keepGenes]

nonpolyA <- readxl::read_xlsx(path = here("accessories","polyAminus.xlsx"), sheet = 1) %>% 
   pull(Gene)


nonpolyA_ID <- geneInfo %>%
  filter(symbol %in% nonpolyA) %>% 
  pull(gene_id) %>% 
  unique()

shortgenes <- read_csv(here("accessories", "human", "short_genes_under_200nt_human_v49.csv.zip")) %>% pull(gene_id)
#Neel requests that we remove short genes! 


# retrieve counts from the myTxi and remove noncoding and non-pA genes
#NOTE: Neel requests that we only remove the non-pA genes and keep annotation open. 

allCounts <- cts_human %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene") %>%
  filter(gene %in% keepGenes) %>%
  filter(!gene %in% nonpolyA_ID) %>%
  filter(!gene %in% shortgenes) %>%
  column_to_rownames(var = "gene") %>%
  as.matrix() %>%
  round(.,0)


#let's require a protein_coding biotype for dros genes
keepDros <- geneInfo %>% filter(species == "fly") %>% filter(biotype == "protein_coding") %>% pull(gene_id)

cts_dros <- cts_dros[rownames(cts_dros) %in% keepDros, , drop = FALSE]


cts_dros <- cts_dros[rowSums(cts_dros) > 0, ] #let's require non-zero values
#hist(log2(1 + rowSums(cts_dros)), breaks = 60) #won't include now

#We have previously determined that we want to only use the top 500 expressed spikeIn genes. 

top500_ids <- cts_dros %>%
  as.data.frame() %>%                     # matrix -> df
  rownames_to_column("gene_id") %>%       # keep gene IDs
  mutate(total = rowSums(across(-gene_id), na.rm = TRUE)) %>%
  slice_max(total, n = 500, with_ties = FALSE) %>%  # exactly 500
  pull(gene_id)

# subset in that order
spikein <- cts_dros[top500_ids, , drop = FALSE] %>% round(.,0)

```


```{r standard deviation filtering}
# log transforming with pseudocount to avoid undefined numbers
qcinput <- log2(myTxi$abundance[keepGenes,] + 1) 

qcinput <- qcinput[rowSds(qcinput) > 1,] 

dim(qcinput)
```



```{r}

mycor <- cor(qcinput, method = "pearson")

heatmap <- pheatmap::pheatmap(
  mat = mycor,
  border_color = "black",  # Black borders around cells
   #color = colorRampPalette(c("blue", "white", "red"))(100),  # Custom color gradient
   scale = "none",  
  cluster_rows = TRUE,  
  cluster_cols = TRUE,  
  fontsize = 8 
)

heatmap

ggsave(here("plots/qc_all.pdf"), heatmap, width=15, height = 14)
```

```{r}
# PCA for all samples
pca_all <- prcomp(t(qcinput), scale. = TRUE)

pca_data_all <- as.data.frame(pca_all$x) %>%
  rownames_to_column("sample") %>%
  mutate(
    condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia"),
    type = ifelse(grepl("ribo", sample), "ribo", "rna")
  )

pca_plot_all <- ggplot(pca_data_all, aes(x = PC1, y = PC2, color = condition, shape = type)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_all)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_all)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - All Samples"
  )

pca_plot_all
ggsave(here("plots/pca_all.pdf"), pca_plot_all, width=8, height = 6)
```

# 107 is an outlier

## RNA only

```{r}
qcinput_rna <- qcinput[, grep("rna", colnames(qcinput))]

mycor_rna <- cor(qcinput_rna, method = "pearson")

heatmap_rna <- pheatmap::pheatmap(
  mat = mycor_rna,
  border_color = "black",
  scale = "none",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize = 8
)

heatmap_rna
ggsave(here("plots/qc_rna_only.pdf"), heatmap_rna, width=10, height = 9)
```

```{r}
# PCA for RNA only
pca_rna <- prcomp(t(qcinput_rna), scale. = TRUE)

pca_data_rna <- as.data.frame(pca_rna$x) %>%
  rownames_to_column("sample") %>%
  mutate(condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia")) |> 
    mutate(treatment = ifelse(grepl("ctrl", sample), "control", 
  ifelse(grepl("3d", sample), "si3d",
    ifelse(grepl("3e", sample), "si3e",
    "other"))))


pca_plot_rna <- ggplot(pca_data_rna, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_rna)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_rna)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - RNA Only"
  )

pca_plot_rna
ggsave(here("plots/pca_rna_only.pdf"), pca_plot_rna, width=8, height = 6)
```

## Ribo only

```{r}
qcinput_ribo <- qcinput[, grep("ribo", colnames(qcinput))]

mycor_ribo <- cor(qcinput_ribo, method = "pearson")

heatmap_ribo <- pheatmap::pheatmap(
  mat = mycor_ribo,
  border_color = "black",
  scale = "none",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize = 8
)

heatmap_ribo
ggsave(here("plots/qc_ribo_only.pdf"), heatmap_ribo, width=10, height = 9)
```

```{r}
# PCA for Ribo only
pca_ribo <- prcomp(t(qcinput_ribo), scale. = TRUE)

pca_data_ribo <- as.data.frame(pca_ribo$x) %>%
  rownames_to_column("sample") %>%
  mutate(condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia")) |> 
      mutate(treatment = ifelse(grepl("ctrl", sample), "control", 
  ifelse(grepl("3d", sample), "si3d",
    ifelse(grepl("3e", sample), "si3e",
    "other"))))

pca_plot_ribo <- ggplot(pca_data_ribo, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_ribo)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_ribo)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - Ribo Only"
  )

pca_plot_ribo
ggsave(here("plots/pca_ribo_only.pdf"), pca_plot_ribo, width=8, height = 6)
```

## Ribo only without outlier 107

```{r}
qcinput_ribo_no107 <- qcinput_ribo[, !grepl("0107", colnames(qcinput_ribo))]

mycor_ribo_no107 <- cor(qcinput_ribo_no107, method = "pearson")

heatmap_ribo_no107 <- pheatmap::pheatmap(
  mat = mycor_ribo_no107,
  border_color = "black",
  scale = "none",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize = 8
)

heatmap_ribo_no107
ggsave(here("plots/qc_ribo_no107.pdf"), heatmap_ribo_no107, width=10, height = 9)
```

```{r}
# PCA for Ribo only without outlier 107
pca_ribo_no107 <- prcomp(t(qcinput_ribo_no107), scale. = TRUE)

pca_data_ribo_no107 <- as.data.frame(pca_ribo_no107$x) %>%
  rownames_to_column("sample") %>%
  mutate(condition = ifelse(grepl("normoxia", sample), "normoxia", "hypoxia")) |> 
      mutate(treatment = ifelse(grepl("ctrl", sample), "control", 
  ifelse(grepl("3d", sample), "si3d",
    ifelse(grepl("3e", sample), "si3e",
    "other"))))

pca_plot_ribo_no107 <- ggplot(pca_data_ribo_no107, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 3) +
  theme_prism() +
  labs(
    x = paste0("PC1 (", round(summary(pca_ribo_no107)$importance[2,1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_ribo_no107)$importance[2,2] * 100, 1), "%)"),
    title = "PCA - Ribo Only (no 107)"
  )

pca_plot_ribo_no107
ggsave(here("plots/pca_ribo_no107.pdf"), pca_plot_ribo_no107, width=8, height = 6)
```


QC looks good! Remove outlier from downstream filtering

```{r remove outlier}
# Remove outlier sample 0107 from allCounts
allCounts <- allCounts[, !grepl("0107", colnames(allCounts))]

# Remove outlier sample 0107 from spikein
spikein <- spikein[, !grepl("0107", colnames(spikein))]

# Remove outlier sample 0107 from metadata
metadata <- metadata %>%
  filter(!grepl("0107", id)) %>%
  dplyr::select(-type)

cat("After removing outlier 0107:\n")
cat("allCounts dimensions:", dim(allCounts), "\n")
cat("spikein dimensions:", dim(spikein), "\n")
cat("metadata rows:", nrow(metadata), "\n")
cat("metadata columns:", colnames(metadata), "\n")
```

```{r save_csvs}
# Save raw human counts for downstream analysis
write_csv(
  allCounts %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "human_allCounts_raw.csv")
)
cat("Saved allCounts to output/human_allCounts_raw.csv\n")

# Save spike-in counts for downstream analysis
write_csv(
  spikein %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "dros_spikeInCounts_raw.csv")
)
cat("Saved spikein to output/dros_spikeInCounts_raw.csv\n")
cat("Note: top500_ids can be derived from this CSV using: top500_ids <- spikein %>% rownames_to_column('gene_id') %>% pull(gene_id)\n")

# Create global normalized matrix with spike-in for ALL samples
# Prepare metadata for all samples
metadata_all <- metadata %>%
  mutate(
    sample_name = paste(condition, treatment, hours, rep, library, id, sep = "_"),
    comparison_variable = condition
  )

# Create normalized matrix using spike-in normalization
normMatrix <- createNormMatrix(
  allCounts = allCounts,
  metadata = metadata_all,
  control_condition = "normoxia",
  spikein = which(rownames(rbind(allCounts, spikein)) %in% top500_ids),
  keepSpikes = top500_ids
)

cat("Dimensions (before removing spike-ins):", dim(normMatrix), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
normMatrix <- normMatrix %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("Dimensions (after removing spike-ins):", dim(normMatrix), "\n")

# Save normalized matrix (without spike-in genes)
write_csv(
  normMatrix %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "normalizedCounts.csv")
)
cat("Saved normalized counts to output/normalizedCounts.csv\n")

# Save cleaned metadata for downstream analysis
write_csv(metadata_all, here("output", "metadata_clean.csv"))
cat("Saved metadata to output/metadata_clean.csv\n")
```


# Spike-in normalization comparison: Normoxia CTRL 1hr vs Hypoxia CTRL 1hr

## Setup for comparison analysis

```{r spike comparison setup}
# Prepare metadata with columns expected by deFunction
# Add sample_name and comparison_variable columns
metadata_prep <- metadata %>%
  mutate(
    sample_name = paste(condition, treatment, hours, rep, library, id, sep = "_"),
    comparison_variable = condition
  )

# Filter metadata for 1 hour control samples
metadata_1hr <- metadata_prep %>%
  filter(treatment == "sictrl", hours == 1)

# Get sample names for filtering
sample_cols <- metadata_1hr$sample_name



# Verify all column names exist in allCounts
missing_cols <- sample_cols[!sample_cols %in% colnames(allCounts)]
if(length(missing_cols) > 0) {
  cat("\nWARNING: Missing columns:\n")
  cat(missing_cols, sep = "\n")
}

# Subset allCounts and spikein using the defined column names
allCounts_1hr <- allCounts[, sample_cols]
spikein_1hr <- spikein[, sample_cols]

cat("\nSelected", ncol(allCounts_1hr), "columns for analysis\n")

# Combine human and spike-in counts
allCounts_combined <- rbind(allCounts_1hr, spikein_1hr)
Spikeindices <- which(rownames(allCounts_combined) %in% top500_ids)

cat("\nCombined counts dimensions:", dim(allCounts_combined), "\n")
cat("Number of spike-in genes:", length(Spikeindices), "\n")
cat("Spike-in gene range in matrix: rows", min(Spikeindices), "to", max(Spikeindices), "\n")

# Setup sample info for DESeq2
sample_info_1hr <- metadata_1hr %>%
  dplyr::select(library, condition, rep) %>%
  unique() |> 
  as.data.frame()

# Set rownames to match column names in allCounts_combined
rownames(sample_info_1hr) <- sample_cols

# Verify order matches
if(!identical(rownames(sample_info_1hr), colnames(allCounts_combined))) {
  cat("WARNING: Row order mismatch!\n")
  # Reorder to match
  sample_info_1hr <- sample_info_1hr[colnames(allCounts_combined), ]
}

sample_info_1hr$condition <- relevel(as.factor(sample_info_1hr$condition), ref = "normoxia")
sample_info_1hr$library <- relevel(as.factor(sample_info_1hr$library), ref = "rna")

cat("\nSample info head:\n")
print(head(sample_info_1hr))

# Create CPM normalized counts (no spike-in)
allCountsNorm_1hr_nospike <- createNormMatrix(allCounts = allCounts_1hr, metadata=metadata_1hr, control_condition="normoxia", spikein = NULL, keepSpikes = NULL)

cat("allCountsNorm_1hr dimensions:", dim(allCountsNorm_1hr_nospike), "\n")


allCountsNorm_1hr_SPIKE <- createNormMatrix(allCounts = allCounts_1hr, metadata=metadata_1hr, control_condition="normoxia", spikein = Spikeindices, keepSpikes = top500_ids)

cat("allCountsNorm_1hr dimensions (before removing spike-ins):", dim(allCountsNorm_1hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_1hr_SPIKE <- allCountsNorm_1hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("allCountsNorm_1hr dimensions (after removing spike-ins):", dim(allCountsNorm_1hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_1hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_1hr_SPIKE.csv")
)
cat("Saved allCountsNorm_1hr_SPIKE.csv\n")
```

## Using deFunction from functions.R

```{r deFunction analysis 1hr}
# Run differential expression using deFunction for NO SPIKE
res_te_nospike_func <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_nospike,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,  # No spike-ins for this analysis
  keepSpikes = NULL,
  mode = "TE"
)

# Add gene symbols
res_te_nospike_classified <- res_te_nospike_func %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nNo-spike results using deFunction:\n")
table(res_te_nospike_classified$class)

```

## With spike in

```{r}

# Run differential expression using deFunction for SPIKE-IN
res_te_spike_func <- deFunction(
  countsMatrix = allCounts_1hr,
  allCounts_Norm = allCountsNorm_1hr_SPIKE,
  metadata = metadata_1hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_1hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

# Add gene symbols
res_te_spike_classified <- res_te_spike_func %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nSpike-in results using deFunction:\n")
table(res_te_spike_classified$class)

# Save 1hr spike-in results for downstream analysis
write_csv(
  res_te_spike_classified,
  here("output", "deseq_results_1hr_spike.csv")
)
cat("Saved 1hr spike-in DESeq results to output/deseq_results_1hr_spike.csv\n")
```

```{r no spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_nospike_volcano <- EnhancedVolcano(
  toptable = res_te_nospike_classified,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_nospike_classified$symbol,
  xlim = c(-5, 5),
  title = "No Spike: Normoxia vs Hypoxia CTRL 1hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_nospike_volcano

```

```{r spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_spike_volcano <- EnhancedVolcano(
  toptable = res_te_spike_classified,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_spike_classified$symbol,
  xlim = c(-5, 5),
  title = "Spike: Normoxia vs Hypoxia CTRL 1hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_spike_volcano

```

## Comparison of spike vs no-spike normalization

```{r comparison venn, fig.width=8, fig.height=6}
# Compare counts
myTab <- bind_rows(
  table(res_te_spike_classified$class) %>% as.data.frame(),
  table(res_te_nospike_classified$class) %>% as.data.frame()
) %>%
  filter(Var1 != "ns") %>%
  mutate(spike = c("yes", "yes", "no", "no"))

ggpubr::ggbarplot(
  data = myTab,
  x = "Var1",
  y = "Freq",
  fill = "spike",
  position = position_dodge(0.75),
  palette = palette_okabe_ito()[c(5, 6)],
  label = TRUE
) +
  theme_prism() +
  xlab("") +
  ylab("count") +
  ggtitle("siCTRL +/- spike-in normalization: Hyp vs Norm 1hr")

ggsave(here("plots/hyp_vs_norm_sictrl_1hr_spike.pdf"))


#Create gene lists for Venn diagrams
hypoxia_up <- list(
  NoSpike = res_te_nospike_classified %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id),
  Spike = res_te_spike_classified %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id)
)

hypoxia_down <- list(
  NoSpike = res_te_nospike_classified %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id),
  Spike = res_te_spike_classified %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id)
)

# Venn diagrams
ggvenn(hypoxia_down, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Downregulated in Hypoxia")

ggvenn(hypoxia_up, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Upregulated in Hypoxia")

# Create comparison dataframe
comp_df <- data.frame(
  gene = res_te_spike_classified$gene_id,
  symbol = res_te_spike_classified$symbol,
  spike_LFC = res_te_spike_classified$log2FoldChange,
  spike_SE = res_te_spike_classified$lfcSE,
  spike_padj = res_te_spike_classified$padj,
  nospike_LFC = res_te_nospike_classified$log2FoldChange,
  nospike_SE = res_te_nospike_classified$lfcSE,
  nospike_padj = res_te_nospike_classified$padj
) %>%
  mutate(
    class = case_when(
      gene %in% setdiff(hypoxia_down$NoSpike, hypoxia_down$Spike) ~ "no spike down",
      gene %in% intersect(hypoxia_down$NoSpike, hypoxia_down$Spike) ~ "both down",
      gene %in% setdiff(hypoxia_down$Spike, hypoxia_down$NoSpike) ~ "spike down",
      gene %in% setdiff(hypoxia_up$NoSpike, hypoxia_up$Spike) ~ "no spike up",
      gene %in% intersect(hypoxia_up$NoSpike, hypoxia_up$Spike) ~ "both up",
      gene %in% setdiff(hypoxia_up$Spike, hypoxia_up$NoSpike) ~ "spike up",
      TRUE ~ "ns"
    )
  )

#Summary bar plot
d <- table(comp_df$class) %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  filter(Var1 != "ns")

ggpubr::ggbarplot(
  data = d,
  x = "Var1",
  y = "Freq",
  fill = "Var1",
  palette = "Paired",
  label = TRUE
) +
  xlab("") +
  ylab("count") +
  theme_prism() +
  rotate_x_text(angle = 90, hjust = 1, vjust = NULL)
ggsave(here("plots/spike_summary_1hr.pdf"))
```

## LFC distribution comparison

```{r LFC comparison, fig.width=8, fig.height=6}
comp_df_long <- comp_df %>% reshape2::melt()

# Overall density plot
ggpubr::ggdensity(
  comp_df_long %>% filter(variable %in% c("spike_LFC", "nospike_LFC") & class != "ns"),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0) +
  xlab("LFC") +
  xlim(-4, 4) +
  theme_prism()

ggsave(here("plots/spike_1hr_density.pdf"), width=10, height=7)

# Faceted density plot by class
comp_df_long$class <- factor(
  as.factor(comp_df_long$class),
  levels = c("both down", "spike down", "no spike down", "both up", "spike up", "no spike up", "ns")
)

p_d <- ggpubr::ggdensity(
  comp_df_long %>% filter(variable %in% c("spike_LFC", "nospike_LFC")),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = -1) +
  geom_vline(xintercept = 1) +
  xlab("LFC") +
  xlim(-4, 4)

ggpubr::facet(p = p_d, facet.by = "class", nrow = 3, scales = "free_y") +
  theme_prism()

```

## Scatter plot: spike vs no-spike LFC

```{r scatter LFC, fig.width=8, fig.height=6}
ggpubr::ggscatter(
  comp_df %>% filter(class != "ns"),
  x = "nospike_LFC",
  y = "spike_LFC",
  color = "class",
  palette = palette_okabe_ito(),
  alpha = 0.33
) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  geom_hline(yintercept = 1) +
  geom_hline(yintercept = -1) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_text(x = 3, y = 0, label = "False Pos?") +
  geom_text(x = -3, y = 0, label = "False Pos?") +
  geom_text(x = 0, y = -3, label = "False\nNeg?") +
  geom_text(x = 0, y = 3, label = "False\nNeg?") +
  theme_prism() +
  ggtitle("Spike vs No-Spike LFC: Norm vs Hyp CTRL 1hr")


ggsave(here("plots/false_neg_pos_spikes.pdf"))
```


# 4 hour analysis 

```{r spike comparison setup}
# Prepare metadata with columns expected by deFunction
# Add sample_name and comparison_variable columns


# Filter metadata for 4 hour control samples
metadata_4hr <- metadata_prep %>%
  filter(treatment == "sictrl", hours == 4)

# Get sample names for filtering
sample_cols_4hr <- metadata_4hr$sample_name



# Verify all column names exist in allCounts
missing_cols <- sample_cols_4hr[!sample_cols_4hr %in% colnames(allCounts)]
if(length(missing_cols) > 0) {
  cat("\nWARNING: Missing columns:\n")
  cat(missing_cols, sep = "\n")
}

# Subset allCounts and spikein using the defined column names
allCounts_4hr <- allCounts[, sample_cols_4hr]
spikein_4hr <- spikein[, sample_cols_4hr]

cat("\nSelected", ncol(allCounts_4hr), "columns for analysis\n")

# Combine human and spike-in counts
allCounts_combined_4hr <- rbind(allCounts_4hr, spikein_4hr)
Spikeindices <- which(rownames(allCounts_combined_4hr) %in% top500_ids)

cat("\nCombined counts dimensions:", dim(allCounts_combined_4hr), "\n")
cat("Number of spike-in genes:", length(Spikeindices), "\n")
cat("Spike-in gene range in matrix: rows", min(Spikeindices), "to", max(Spikeindices), "\n")

# Setup sample info for DESeq2
sample_info_4hr <- metadata_4hr %>%
  dplyr::select(library, condition, rep) %>%
  unique() |>
  as.data.frame()

# Set rownames to match column names in allCounts_combined_4hr
rownames(sample_info_4hr) <- sample_cols_4hr

# Verify order matches
if(!identical(rownames(sample_info_4hr), colnames(allCounts_combined_4hr))) {
  cat("WARNING: Row order mismatch!\n")
  # Reorder to match
  sample_info_4hr <- sample_info_4hr[colnames(allCounts_combined_4hr), ]
}

sample_info_4hr$condition <- relevel(as.factor(sample_info_4hr$condition), ref = "normoxia")
sample_info_4hr$library <- relevel(as.factor(sample_info_4hr$library), ref = "rna")

cat("\nSample info head:\n")
print(head(sample_info_4hr))

# Create CPM normalized counts (no spike-in)
allCountsNorm_4hr_nospike <- createNormMatrix(allCounts = allCounts_4hr, metadata=metadata_4hr, control_condition="normoxia", spikein = NULL, keepSpikes = NULL)

cat("allCountsNorm_4hr dimensions:", dim(allCountsNorm_4hr_nospike), "\n")


allCountsNorm_4hr_SPIKE <- createNormMatrix(allCounts = allCounts_4hr, metadata=metadata_4hr, control_condition="normoxia", spikein = Spikeindices, keepSpikes = top500_ids)

cat("allCountsNorm_4hr dimensions (before removing spike-ins):", dim(allCountsNorm_4hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_4hr_SPIKE <- allCountsNorm_4hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("allCountsNorm_4hr dimensions (after removing spike-ins):", dim(allCountsNorm_4hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_4hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_4hr_SPIKE.csv")
)
cat("Saved allCountsNorm_4hr_SPIKE.csv\n")
```

## Using deFunction from functions.R

```{r deFunction analysis 4hr}
# Run differential expression using deFunction for NO SPIKE
res_te_nospike_func_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_nospike,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,  # No spike-ins for this analysis
  keepSpikes = NULL,
  mode = "TE"
)

# Add gene symbols
res_te_nospike_classified_4hr <- res_te_nospike_func_4hr %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nNo-spike results using deFunction:\n")
table(res_te_nospike_classified_4hr$class)

```

## With spike in

```{r}

# Run differential expression using deFunction for SPIKE-IN
res_te_spike_func_4hr <- deFunction(
  countsMatrix = allCounts_4hr,
  allCounts_Norm = allCountsNorm_4hr_SPIKE,
  metadata = metadata_4hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_4hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

# Add gene symbols
res_te_spike_classified_4hr <- res_te_spike_func_4hr %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nSpike-in results using deFunction:\n")
table(res_te_spike_classified_4hr$class)

# Save 4hr spike-in results for downstream analysis
write_csv(
  res_te_spike_classified_4hr,
  here("output", "deseq_results_4hr_spike.csv")
)
cat("Saved 4hr spike-in DESeq results to output/deseq_results_4hr_spike.csv\n")
```

```{r no spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_nospike_volcano_4hr <- EnhancedVolcano(
  toptable = res_te_nospike_classified_4hr,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_nospike_classified_4hr$symbol,
  xlim = c(-5, 5),
  title = "No Spike: Normoxia vs Hypoxia CTRL 4hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_nospike_volcano_4hr

```

```{r spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_spike_volcano_4hr <- EnhancedVolcano(
  toptable = res_te_spike_classified_4hr,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_spike_classified_4hr$symbol,
  xlim = c(-5, 5),
  title = "Spike: Normoxia vs Hypoxia CTRL 4hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_spike_volcano_4hr

```

## Comparison of spike vs no-spike normalization

```{r comparison venn, fig.width=8, fig.height=6}
# Compare counts
myTab_4hr <- bind_rows(
  table(res_te_spike_classified_4hr$class) %>% as.data.frame(),
  table(res_te_nospike_classified_4hr$class) %>% as.data.frame()
) %>%
  filter(Var1 != "ns") %>%
  mutate(spike = c("yes", "yes", "no", "no"))

ggpubr::ggbarplot(
  data = myTab_4hr,
  x = "Var1",
  y = "Freq",
  fill = "spike",
  position = position_dodge(0.75),
  palette = palette_okabe_ito()[c(5, 6)],
  label = TRUE
) +
  theme_prism() +
  xlab("") +
  ylab("count") +
  ggtitle("siCTRL +/- spike-in normalization: Hyp vs Norm 4hr")

ggsave(here("plots/hyp_vs_norm_sictrl_4hr_spike.pdf"))


#Create gene lists for Venn diagrams
hypoxia_up_4hr <- list(
  NoSpike = res_te_nospike_classified_4hr %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id),
  Spike = res_te_spike_classified_4hr %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id)
)

hypoxia_down_4hr <- list(
  NoSpike = res_te_nospike_classified_4hr %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id),
  Spike = res_te_spike_classified_4hr %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id)
)

# Venn diagrams
ggvenn(hypoxia_down_4hr, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Downregulated in Hypoxia")

ggvenn(hypoxia_up_4hr, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Upregulated in Hypoxia")

# Create comparison dataframe
comp_df_4hr <- data.frame(
  gene = res_te_spike_classified_4hr$gene_id,
  symbol = res_te_spike_classified_4hr$symbol,
  spike_LFC = res_te_spike_classified_4hr$log2FoldChange,
  spike_SE = res_te_spike_classified_4hr$lfcSE,
  spike_padj = res_te_spike_classified_4hr$padj,
  nospike_LFC = res_te_nospike_classified_4hr$log2FoldChange,
  nospike_SE = res_te_nospike_classified_4hr$lfcSE,
  nospike_padj = res_te_nospike_classified_4hr$padj
) %>%
  mutate(
    class = case_when(
      gene %in% setdiff(hypoxia_down_4hr$NoSpike, hypoxia_down_4hr$Spike) ~ "no spike down",
      gene %in% intersect(hypoxia_down_4hr$NoSpike, hypoxia_down_4hr$Spike) ~ "both down",
      gene %in% setdiff(hypoxia_down_4hr$Spike, hypoxia_down_4hr$NoSpike) ~ "spike down",
      gene %in% setdiff(hypoxia_up_4hr$NoSpike, hypoxia_up_4hr$Spike) ~ "no spike up",
      gene %in% intersect(hypoxia_up_4hr$NoSpike, hypoxia_up_4hr$Spike) ~ "both up",
      gene %in% setdiff(hypoxia_up_4hr$Spike, hypoxia_up_4hr$NoSpike) ~ "spike up",
      TRUE ~ "ns"
    )
  )

#Summary bar plot
d_4hr <- table(comp_df_4hr$class) %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  filter(Var1 != "ns")

ggpubr::ggbarplot(
  data = d_4hr,
  x = "Var1",
  y = "Freq",
  fill = "Var1",
  palette = "Paired",
  label = TRUE
) +
  xlab("") +
  ylab("count") +
  theme_prism() +
  rotate_x_text(angle = 90, hjust = 1, vjust = NULL)
ggsave(here("plots/spike_summary_4hr.pdf"))
```

## LFC distribution comparison

```{r LFC comparison, fig.width=8, fig.height=6}
comp_df_long_4hr <- comp_df_4hr %>% reshape2::melt()

# Overall density plot
ggpubr::ggdensity(
  comp_df_long_4hr %>% filter(variable %in% c("spike_LFC", "nospike_LFC") & class != "ns"),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0) +
  xlab("LFC") +
  xlim(-4, 4) +
  theme_prism()

ggsave(here("plots/spike_4hr_density.pdf"), width=10, height=7)

# Faceted density plot by class
comp_df_long_4hr$class <- factor(
  as.factor(comp_df_long_4hr$class),
  levels = c("both down", "spike down", "no spike down", "both up", "spike up", "no spike up", "ns")
)

p_d_4hr <- ggpubr::ggdensity(
  comp_df_long_4hr %>% filter(variable %in% c("spike_LFC", "nospike_LFC")),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = -1) +
  geom_vline(xintercept = 1) +
  xlab("LFC") +
  xlim(-4, 4)

ggpubr::facet(p = p_d_4hr, facet.by = "class", nrow = 3, scales = "free_y") +
  theme_prism()

```

## Scatter plot: spike vs no-spike LFC

```{r scatter LFC, fig.width=8, fig.height=6}
ggpubr::ggscatter(
  comp_df_4hr %>% filter(class != "ns"),
  x = "nospike_LFC",
  y = "spike_LFC",
  color = "class",
  palette = palette_okabe_ito(),
  alpha = 0.33
) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  geom_hline(yintercept = 1) +
  geom_hline(yintercept = -1) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_text(x = 3, y = 0, label = "False Pos?") +
  geom_text(x = -3, y = 0, label = "False Pos?") +
  geom_text(x = 0, y = -3, label = "False\nNeg?") +
  geom_text(x = 0, y = 3, label = "False\nNeg?") +
  theme_prism() +
  ggtitle("Spike vs No-Spike LFC: Norm vs Hyp CTRL 4hr")


ggsave(here("plots/false_neg_pos_spikes_4hr.pdf"))
```


# 24 hour analysis

```{r spike comparison setup}
# Prepare metadata with columns expected by deFunction
# Add sample_name and comparison_variable columns


# Filter metadata for 24 hour control samples
metadata_24hr <- metadata_prep %>%
  filter(treatment == "sictrl", hours == 24)

# Get sample names for filtering
sample_cols_24hr <- metadata_24hr$sample_name



# Verify all column names exist in allCounts
missing_cols <- sample_cols_24hr[!sample_cols_24hr %in% colnames(allCounts)]
if(length(missing_cols) > 0) {
  cat("\nWARNING: Missing columns:\n")
  cat(missing_cols, sep = "\n")
}

# Subset allCounts and spikein using the defined column names
allCounts_24hr <- allCounts[, sample_cols_24hr]
spikein_24hr <- spikein[, sample_cols_24hr]

cat("\nSelected", ncol(allCounts_24hr), "columns for analysis\n")

# Combine human and spike-in counts
allCounts_combined_24hr <- rbind(allCounts_24hr, spikein_24hr)
Spikeindices <- which(rownames(allCounts_combined_24hr) %in% top500_ids)

cat("\nCombined counts dimensions:", dim(allCounts_combined_24hr), "\n")
cat("Number of spike-in genes:", length(Spikeindices), "\n")
cat("Spike-in gene range in matrix: rows", min(Spikeindices), "to", max(Spikeindices), "\n")

# Setup sample info for DESeq2
sample_info_24hr <- metadata_24hr %>%
  dplyr::select(library, condition, rep) %>%
  unique() |>
  as.data.frame()

# Set rownames to match column names in allCounts_combined_24hr
rownames(sample_info_24hr) <- sample_cols_24hr

# Verify order matches
if(!identical(rownames(sample_info_24hr), colnames(allCounts_combined_24hr))) {
  cat("WARNING: Row order mismatch!\n")
  # Reorder to match
  sample_info_24hr <- sample_info_24hr[colnames(allCounts_combined_24hr), ]
}

sample_info_24hr$condition <- relevel(as.factor(sample_info_24hr$condition), ref = "normoxia")
sample_info_24hr$library <- relevel(as.factor(sample_info_24hr$library), ref = "rna")

cat("\nSample info head:\n")
print(head(sample_info_24hr))

# Create CPM normalized counts (no spike-in)
allCountsNorm_24hr_nospike <- createNormMatrix(allCounts = allCounts_24hr, metadata=metadata_24hr, control_condition="normoxia", spikein = NULL, keepSpikes = NULL)

cat("allCountsNorm_24hr dimensions:", dim(allCountsNorm_24hr_nospike), "\n")


allCountsNorm_24hr_SPIKE <- createNormMatrix(allCounts = allCounts_24hr, metadata=metadata_24hr, control_condition="normoxia", spikein = Spikeindices, keepSpikes = top500_ids)

cat("allCountsNorm_24hr dimensions (before removing spike-ins):", dim(allCountsNorm_24hr_SPIKE), "\n")

# Remove spike-in genes (ERCC, SIRV, Drosophila, eGFP) from normalized matrix
allCountsNorm_24hr_SPIKE <- allCountsNorm_24hr_SPIKE %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  filter(!grepl("^ERCC-|^ERCC_|^SIRV|^eGFP|^FBgn", gene_id, ignore.case = TRUE)) %>%
  column_to_rownames("gene_id") %>%
  as.matrix()

cat("allCountsNorm_24hr dimensions (after removing spike-ins):", dim(allCountsNorm_24hr_SPIKE), "\n")

# Save spike-in normalized matrix (without spike-in genes)
write_csv(
  allCountsNorm_24hr_SPIKE %>% as.data.frame() %>% rownames_to_column("gene_id"),
  here("output", "allCountsNorm_24hr_SPIKE.csv")
)
cat("Saved allCountsNorm_24hr_SPIKE.csv\n")
```

## Using deFunction from functions.R

```{r deFunction analysis 24hr}
# Run differential expression using deFunction for NO SPIKE
res_te_nospike_func_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_nospike,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = NULL,  # No spike-ins for this analysis
  keepSpikes = NULL,
  mode = "TE"
)

# Add gene symbols
res_te_nospike_classified_24hr <- res_te_nospike_func_24hr %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nNo-spike results using deFunction:\n")
table(res_te_nospike_classified_24hr$class)

```

## With spike in

```{r}

# Run differential expression using deFunction for SPIKE-IN
res_te_spike_func_24hr <- deFunction(
  countsMatrix = allCounts_24hr,
  allCounts_Norm = allCountsNorm_24hr_SPIKE,
  metadata = metadata_24hr,
  control_condition = "normoxia",
  test_condition = "hypoxia",
  spikeInMatrix = spikein_24hr,
  keepSpikes = top500_ids,
  mode = "TE"
)

# Add gene symbols
res_te_spike_classified_24hr <- res_te_spike_func_24hr %>%
  left_join(gene_symbol, by = "gene_id") %>%
  filter(!is.na(symbol)) %>%
  mutate(
    class = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "Hypoxia up",
      padj < 0.05 & log2FoldChange < 0 ~ "Hypoxia down",
      TRUE ~ "ns"
    )
  )

cat("\nSpike-in results using deFunction:\n")
table(res_te_spike_classified_24hr$class)

# Save 24hr spike-in results for downstream analysis
write_csv(
  res_te_spike_classified_24hr,
  here("output", "deseq_results_24hr_spike.csv")
)
cat("Saved 24hr spike-in DESeq results to output/deseq_results_24hr_spike.csv\n")
```

```{r no spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_nospike_volcano_24hr <- EnhancedVolcano(
  toptable = res_te_nospike_classified_24hr,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_nospike_classified_24hr$symbol,
  xlim = c(-5, 5),
  title = "No Spike: Normoxia vs Hypoxia CTRL 24hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_nospike_volcano_24hr

```

```{r spike analysis, fig.width=8, fig.height=8}
# Volcano plot
p_spike_volcano_24hr <- EnhancedVolcano(
  toptable = res_te_spike_classified_24hr,
  colAlpha = 0.5,
  pCutoff = 0.05,
  x = "log2FoldChange",
  y = "padj",
  lab = res_te_spike_classified_24hr$symbol,
  xlim = c(-5, 5),
  title = "Spike: Normoxia vs Hypoxia CTRL 24hr delta TE",
  subtitle = "",
  legendPosition = "bottom",
  col = c('grey', 'black', 'blue', 'red3'),
  legendLabels = c('n.s.', 'Log2FC', 'p-value', 'p-value & Log2FC'),
  max.overlaps = Inf,
  labSize = 3
)

p_spike_volcano_24hr

```

## Comparison of spike vs no-spike normalization

```{r comparison venn, fig.width=8, fig.height=6}
# Compare counts
myTab_24hr <- bind_rows(
  table(res_te_spike_classified_24hr$class) %>% as.data.frame(),
  table(res_te_nospike_classified_24hr$class) %>% as.data.frame()
) %>%
  filter(Var1 != "ns") %>%
  mutate(spike = c("yes", "yes", "no", "no"))

ggpubr::ggbarplot(
  data = myTab_24hr,
  x = "Var1",
  y = "Freq",
  fill = "spike",
  position = position_dodge(0.75),
  palette = palette_okabe_ito()[c(5, 6)],
  label = TRUE
) +
  theme_prism() +
  xlab("") +
  ylab("count") +
  ggtitle("siCTRL +/- spike-in normalization: Hyp vs Norm 24hr")

ggsave(here("plots/hyp_vs_norm_sictrl_24hr_spike.pdf"))


#Create gene lists for Venn diagrams
hypoxia_up_24hr <- list(
  NoSpike = res_te_nospike_classified_24hr %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id),
  Spike = res_te_spike_classified_24hr %>%
    filter(class == "Hypoxia up") %>%
    pull(gene_id)
)

hypoxia_down_24hr <- list(
  NoSpike = res_te_nospike_classified_24hr %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id),
  Spike = res_te_spike_classified_24hr %>%
    filter(class == "Hypoxia down") %>%
    pull(gene_id)
)

# Venn diagrams
ggvenn(hypoxia_down_24hr, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Downregulated in Hypoxia")

ggvenn(hypoxia_up_24hr, auto_scale = FALSE, show_percentage = FALSE) +
  ggtitle("Upregulated in Hypoxia")

# Create comparison dataframe
comp_df_24hr <- data.frame(
  gene = res_te_spike_classified_24hr$gene_id,
  symbol = res_te_spike_classified_24hr$symbol,
  spike_LFC = res_te_spike_classified_24hr$log2FoldChange,
  spike_SE = res_te_spike_classified_24hr$lfcSE,
  spike_padj = res_te_spike_classified_24hr$padj,
  nospike_LFC = res_te_nospike_classified_24hr$log2FoldChange,
  nospike_SE = res_te_nospike_classified_24hr$lfcSE,
  nospike_padj = res_te_nospike_classified_24hr$padj
) %>%
  mutate(
    class = case_when(
      gene %in% setdiff(hypoxia_down_24hr$NoSpike, hypoxia_down_24hr$Spike) ~ "no spike down",
      gene %in% intersect(hypoxia_down_24hr$NoSpike, hypoxia_down_24hr$Spike) ~ "both down",
      gene %in% setdiff(hypoxia_down_24hr$Spike, hypoxia_down_24hr$NoSpike) ~ "spike down",
      gene %in% setdiff(hypoxia_up_24hr$NoSpike, hypoxia_up_24hr$Spike) ~ "no spike up",
      gene %in% intersect(hypoxia_up_24hr$NoSpike, hypoxia_up_24hr$Spike) ~ "both up",
      gene %in% setdiff(hypoxia_up_24hr$Spike, hypoxia_up_24hr$NoSpike) ~ "spike up",
      TRUE ~ "ns"
    )
  )

#Summary bar plot
d_24hr <- table(comp_df_24hr$class) %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  filter(Var1 != "ns")

ggpubr::ggbarplot(
  data = d_24hr,
  x = "Var1",
  y = "Freq",
  fill = "Var1",
  palette = "Paired",
  label = TRUE
) +
  xlab("") +
  ylab("count") +
  theme_prism() +
  rotate_x_text(angle = 90, hjust = 1, vjust = NULL)
ggsave(here("plots/spike_summary_24hr.pdf"))
```

## LFC distribution comparison

```{r LFC comparison, fig.width=8, fig.height=6}
comp_df_long_24hr <- comp_df_24hr %>% reshape2::melt()

# Overall density plot
ggpubr::ggdensity(
  comp_df_long_24hr %>% filter(variable %in% c("spike_LFC", "nospike_LFC") & class != "ns"),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0) +
  xlab("LFC") +
  xlim(-4, 4) +
  theme_prism()

ggsave(here("plots/spike_24hr_density.pdf"), width=10, height=7)

# Faceted density plot by class
comp_df_long_24hr$class <- factor(
  as.factor(comp_df_long_24hr$class),
  levels = c("both down", "spike down", "no spike down", "both up", "spike up", "no spike up", "ns")
)

p_d_24hr <- ggpubr::ggdensity(
  comp_df_long_24hr %>% filter(variable %in% c("spike_LFC", "nospike_LFC")),
  x = "value",
  fill = "variable",
  palette = palette_okabe_ito()[c(1, 2)]
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = -1) +
  geom_vline(xintercept = 1) +
  xlab("LFC") +
  xlim(-4, 4)

ggpubr::facet(p = p_d_24hr, facet.by = "class", nrow = 3, scales = "free_y") +
  theme_prism()

```

## Scatter plot: spike vs no-spike LFC

```{r scatter LFC, fig.width=8, fig.height=6}
ggpubr::ggscatter(
  comp_df_24hr %>% filter(class != "ns"),
  x = "nospike_LFC",
  y = "spike_LFC",
  color = "class",
  palette = palette_okabe_ito(),
  alpha = 0.33
) +
  xlim(-4, 4) +
  ylim(-4, 4) +
  geom_hline(yintercept = 1) +
  geom_hline(yintercept = -1) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_text(x = 3, y = 0, label = "False Pos?") +
  geom_text(x = -3, y = 0, label = "False Pos?") +
  geom_text(x = 0, y = -3, label = "False\nNeg?") +
  geom_text(x = 0, y = 3, label = "False\nNeg?") +
  theme_prism() +
  ggtitle("Spike vs No-Spike LFC: Norm vs Hyp CTRL 24hr")


ggsave(here("plots/false_neg_pos_spikes_24hr.pdf"))
```

# Spike-In Effect on TE Changes: eIF3e and eIF3d Knockdowns

## si3d vs sictrl Normoxia 1hr - Spike-In Comparison

```{r si3d_norm_1hr_spikein_comparison}
# Prepare metadata
metadata_si3d_norm_1hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 1, treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                             test_condition = "si3d", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                              test_condition = "si3d", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                            test_condition = "si3d", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3d_1hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                                test_condition = "si3d", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                                 test_condition = "si3d", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3d_norm_1hr, control_condition = "sictrl",
                               test_condition = "si3d", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3d_1hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3d_1hr <- bind_rows(
  results_spike_si3d_1hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3d", timepoint = "1hr"),
  results_no_spike_si3d_1hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3d", timepoint = "1hr")
)

print(te_counts_si3d_1hr)
```

## si3d vs sictrl Normoxia 4hr - Spike-In Comparison

```{r si3d_norm_4hr_spikein_comparison}
# Prepare metadata
metadata_si3d_norm_4hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 4, treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                             test_condition = "si3d", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                              test_condition = "si3d", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                            test_condition = "si3d", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3d_4hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                                test_condition = "si3d", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                                 test_condition = "si3d", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3d_norm_4hr, control_condition = "sictrl",
                               test_condition = "si3d", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3d_4hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3d_4hr <- bind_rows(
  results_spike_si3d_4hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3d", timepoint = "4hr"),
  results_no_spike_si3d_4hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3d", timepoint = "4hr")
)

print(te_counts_si3d_4hr)
```

## si3d vs sictrl Normoxia 24hr - Spike-In Comparison

```{r si3d_norm_24hr_spikein_comparison}
# Prepare metadata
metadata_si3d_norm_24hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 24, treatment %in% c("sictrl", "si3d")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3d_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                             test_condition = "si3d", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                              test_condition = "si3d", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                            test_condition = "si3d", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3d_24hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                                test_condition = "si3d", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                                 test_condition = "si3d", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3d_norm_24hr, control_condition = "sictrl",
                               test_condition = "si3d", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3d_24hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3d_24hr <- bind_rows(
  results_spike_si3d_24hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3d", timepoint = "24hr"),
  results_no_spike_si3d_24hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3d", timepoint = "24hr")
)

print(te_counts_si3d_24hr)
```

## si3e vs sictrl Normoxia 1hr - Spike-In Comparison

```{r si3e_norm_1hr_spikein_comparison}
# Prepare metadata
metadata_si3e_norm_1hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 1, treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_1hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                             test_condition = "si3e", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                              test_condition = "si3e", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                            test_condition = "si3e", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3e_1hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                                test_condition = "si3e", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                                 test_condition = "si3e", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3e_norm_1hr, control_condition = "sictrl",
                               test_condition = "si3e", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3e_1hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3e_1hr <- bind_rows(
  results_spike_si3e_1hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3e", timepoint = "1hr"),
  results_no_spike_si3e_1hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3e", timepoint = "1hr")
)

print(te_counts_si3e_1hr)
```

## si3e vs sictrl Normoxia 4hr - Spike-In Comparison

```{r si3e_norm_4hr_spikein_comparison}
# Prepare metadata
metadata_si3e_norm_4hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 4, treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_4hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                             test_condition = "si3e", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                              test_condition = "si3e", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                            test_condition = "si3e", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3e_4hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                                test_condition = "si3e", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                                 test_condition = "si3e", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3e_norm_4hr, control_condition = "sictrl",
                               test_condition = "si3e", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3e_4hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3e_4hr <- bind_rows(
  results_spike_si3e_4hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3e", timepoint = "4hr"),
  results_no_spike_si3e_4hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3e", timepoint = "4hr")
)

print(te_counts_si3e_4hr)
```

## si3e vs sictrl Normoxia 24hr - Spike-In Comparison

```{r si3e_norm_24hr_spikein_comparison}
# Prepare metadata
metadata_si3e_norm_24hr <- metadata_all %>%
  filter(condition == "normoxia", hours == 24, treatment %in% c("sictrl", "si3e")) %>%
  mutate(comparison_variable = treatment)

sample_cols <- metadata_si3e_norm_24hr$sample_name
allCounts_filtered <- allCounts[, sample_cols]
spikein_filtered <- spikein[, sample_cols]

## WITH SPIKE-INS
res_rna_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                             metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                             test_condition = "si3e", spikeInMatrix = spikein_filtered,
                             keepSpikes = top500_ids, mode = "RNA")
res_ribo_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                              metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                              test_condition = "si3e", spikeInMatrix = spikein_filtered,
                              keepSpikes = top500_ids, mode = "RIBO")
res_te_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                            metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                            test_condition = "si3e", spikeInMatrix = spikein_filtered,
                            keepSpikes = top500_ids, mode = "TE")

results_spike_si3e_24hr <- categorize_translation_changes(res_rna_spike, res_ribo_spike, res_te_spike,
                                                          padj_cutoff = 0.05, lfc_cutoff = 0.0)

## WITHOUT SPIKE-INS
res_rna_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                                test_condition = "si3e", spikeInMatrix = NULL,
                                keepSpikes = NULL, mode = "RNA")
res_ribo_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                                 metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                                 test_condition = "si3e", spikeInMatrix = NULL,
                                 keepSpikes = NULL, mode = "RIBO")
res_te_no_spike <- deFunction(countsMatrix = allCounts_filtered, allCounts_Norm = normMatrix,
                               metadata = metadata_si3e_norm_24hr, control_condition = "sictrl",
                               test_condition = "si3e", spikeInMatrix = NULL,
                               keepSpikes = NULL, mode = "TE")

results_no_spike_si3e_24hr <- categorize_translation_changes(res_rna_no_spike, res_ribo_no_spike, res_te_no_spike,
                                                             padj_cutoff = 0.05, lfc_cutoff = 0.0)

## Count TE-regulated genes
te_counts_si3e_24hr <- bind_rows(
  results_spike_si3e_24hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "With Spike-Ins", treatment = "si3e", timepoint = "24hr"),
  results_no_spike_si3e_24hr$categorized_table %>%
    filter(category %in% c("Exclusive", "Buffered", "TE_only")) %>%
    summarise(up = sum(te_lfc > 0), down = sum(te_lfc < 0)) %>%
    mutate(method = "Without Spike-Ins", treatment = "si3e", timepoint = "24hr")
)

print(te_counts_si3e_24hr)
```

## Combined Comparison and Plotting

```{r spikein_comparison_plots, fig.width=10, fig.height=6}
# Combine all comparisons
all_te_comparisons <- bind_rows(
  te_counts_si3d_1hr,
  te_counts_si3d_4hr,
  te_counts_si3d_24hr,
  te_counts_si3e_1hr,
  te_counts_si3e_4hr,
  te_counts_si3e_24hr
)

# Reshape for plotting
plot_data <- all_te_comparisons %>%
  pivot_longer(cols = c(up, down), names_to = "direction", values_to = "count") %>%
  mutate(direction = factor(direction, levels = c("up", "down")),
         timepoint = factor(timepoint, levels = c("1hr", "4hr", "24hr")))

# Bar plot si3d
ggplot(plot_data %>% filter(treatment == "si3d"),
       aes(x = timepoint, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c("up" = "red", "down" = "blue")) +
  labs(title = "si3d: Spike-In Effect on TE-Regulated Genes",
       subtitle = "Normoxia, TE up vs down",
       x = "Timepoint", y = "Number of genes", fill = "TE Direction") +
  theme_minimal(base_size = 14)

ggsave(here("plots/spikein_te_comparison_si3d.pdf"), width=8, height=5)

# Bar plot si3e
ggplot(plot_data %>% filter(treatment == "si3e"),
       aes(x = timepoint, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ method) +
  scale_fill_manual(values = c("up" = "red", "down" = "blue")) +
  labs(title = "si3e: Spike-In Effect on TE-Regulated Genes",
       subtitle = "Normoxia, TE up vs down",
       x = "Timepoint", y = "Number of genes", fill = "TE Direction") +
  theme_minimal(base_size = 14)

ggsave(here("plots/spikein_te_comparison_si3e.pdf"), width=8, height=5)

# Combined plot
ggplot(plot_data, aes(x = timepoint, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = count), position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  facet_grid(treatment ~ method) +
  scale_fill_manual(values = c("up" = "red", "down" = "blue")) +
  labs(title = "Spike-In Effect on TE-Regulated Genes",
       subtitle = "si3d and si3e vs sictrl, Normoxia",
       x = "Timepoint", y = "Number of genes", fill = "TE Direction") +
  theme_minimal(base_size = 14)

ggsave(here("plots/spikein_te_comparison_combined.pdf"), width=10, height=6)
```