---
title: "Translation Factors in Hypoxic Response"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(clusterProfiler)
library(enrichplot)
library(pheatmap)
library(writexl)
library(readxl)
```

## Load TE Changes from Hypoxia Timecourse

```{r load_te_data}
# Load TE changes from hypoxia timecourse at all three timepoints
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))
te_4hr <- read_csv(here("output", "translation_categories_4hr.csv"))
te_24hr <- read_csv(here("output", "translation_categories_24hr.csv"))

# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

# Add gene symbols to TE data and prepare for GSEA
# Handle duplicate gene symbols by keeping the gene with highest absolute LFC
te_1hr_gsea <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_4hr_gsea <- te_4hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

te_24hr_gsea <- te_24hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  rename(log2FoldChange = te_lfc, padj = te_padj) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Loaded TE data with gene symbols:\n")
cat("1hr:", nrow(te_1hr_gsea), "genes\n")
cat("  - With symbols:", sum(!is.na(te_1hr_gsea$symbol)), "\n")
cat("  - Without symbols (NA):", sum(is.na(te_1hr_gsea$symbol)), "\n")
cat("4hr:", nrow(te_4hr_gsea), "genes\n")
cat("  - With symbols:", sum(!is.na(te_4hr_gsea$symbol)), "\n")
cat("24hr:", nrow(te_24hr_gsea), "genes\n")
cat("  - With symbols:", sum(!is.na(te_24hr_gsea$symbol)), "\n")
```

## Load Translation Factor Signatures from Literature

```{r load_translation_signatures}
# Load translation factor signatures from literature
translation_sigs <- read_csv(here("accessories", "translation_signatures_literature.csv"))

cat("Loaded translation signatures:\n")
print(colnames(translation_sigs))
cat("\nNumber of genes in signature dataset:", nrow(translation_sigs), "\n")
```

## Load DAP5 Knockdown Data (Weber 2022)

```{r load_dap5_weber}
# Load DAP5 knockdown ribo-seq data from Weber et al. 2022
dap5_weber <- read_xlsx(here("accessories", "dap5kd_riboseq_weber2022.xlsx"))

colnames(dap5_weber)
head(dap5_weber)

# Filter for significant genes and create up/down gene sets
dap5_sig <- dap5_weber %>%
  filter(!is.na(geneIDs)) |> 
  filter(comment != "no")


# Up-regulated upon DAP5 KD (TE increased)
dap5_represses <- dap5_sig %>% 
  filter(comment=="up")

# Down-regulated upon DAP5 KD (TE decreased)
dap5_promotes <- dap5_sig %>% 
  filter(comment=="down")
```

## Load Teleman riboseq data

```{r define-resistant-4e}

res_riboseq_symbol <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv")

# First remove duplicates from the entire dataset (keep gene with highest baseMean)
res_riboseq_dedup <- res_riboseq_symbol %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Total genes after removing duplicates:", nrow(res_riboseq_dedup), "\n")

# Resistant genes: log2FC > 1 and padj < 0.05
# Positive = resistant to 4E inhibition (still translated when eIF4E is inhibited)
resistant_4E <- res_riboseq_dedup %>%
  filter(log2FoldChange > 1) %>%
  filter(padj < 0.05)

cat("Number of 4E-resistant genes (LFC > 1, padj < 0.05):", nrow(resistant_4E), "\n")
if(nrow(resistant_4E) > 0) {
  cat("log2FC range:", round(min(resistant_4E$log2FoldChange), 2), "to",
      round(max(resistant_4E$log2FoldChange), 2), "\n")
}

# Create gene set for overlap analysis - use pull() to extract vector
resistant_set <- data.frame(
  gs_name = "Resistant_4E",
  gene = resistant_4E %>% pull(symbol)
)

# Show top resistant genes
resistant_4E %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(symbol, log2FoldChange, padj, baseMean) %>%
  head(20)
```

## Define sensitive 4E lists

```{r define-sensitive-4e}

# Sensitive genes: log2FC < -1 and padj < 0.05
# Negative = sensitive to 4E inhibition (reduced translation when eIF4E is inhibited)
sensitive_4E <- res_riboseq_dedup %>%
  filter(log2FoldChange < -1) %>%
  filter(padj < 0.05)

cat("Number of 4E-sensitive genes (LFC < -1, padj < 0.05):", nrow(sensitive_4E), "\n")
if(nrow(sensitive_4E) > 0) {
  cat("log2FC range:", round(min(sensitive_4E$log2FoldChange), 2), "to",
      round(max(sensitive_4E$log2FoldChange), 2), "\n")
}

# Create gene set for overlap analysis - use pull() to extract vector
sensitive_set <- data.frame(
  gs_name = "Sensitive_4E",
  gene = sensitive_4E %>% pull(symbol)
)

# Show top sensitive genes (most negative log2FC)
sensitive_4E %>%
  arrange(log2FoldChange) %>%
  dplyr::select(symbol, log2FoldChange, padj, baseMean) %>%
  head(20)
```

## Combine Resistant and Sensitive 4E Gene Sets for GSEA

```{r combine-4e-genesets}
# Combine resistant and sensitive gene sets into TERM2GENE format for GSEA
# This creates two separate gene sets: Resistant_4E and Sensitive_4E
# IMPORTANT: TERM2GENE needs columns in order: term, gene (column names don't matter, just order)
teleman_4E_genesets <- bind_rows(resistant_set, sensitive_set) %>%
  dplyr::select(gs_name, gene)  # Ensure correct column order

cat("Gene sets for GSEA (TERM2GENE format):\n")
cat("Total gene entries:", nrow(teleman_4E_genesets), "\n")
cat("Number of gene sets: 2\n")
cat("  - Resistant_4E:", sum(teleman_4E_genesets$gs_name == "Resistant_4E"), "genes\n")
cat("  - Sensitive_4E:", sum(teleman_4E_genesets$gs_name == "Sensitive_4E"), "genes\n")

# Define gene lists for overlap checking
te_genes_1hr <- unique(te_1hr_gsea$symbol)
te_genes_4hr <- unique(te_4hr_gsea$symbol)
te_genes_24hr <- unique(te_24hr_gsea$symbol)

# Check overlap with TE data
resistant_genes <- unique(resistant_set$gene)
sensitive_genes <- unique(sensitive_set$gene)

cat("\nResistant_4E overlap with TE data:\n")
cat("1hr:", length(intersect(resistant_genes, te_genes_1hr)), "genes\n")
cat("4hr:", length(intersect(resistant_genes, te_genes_4hr)), "genes\n")
cat("24hr:", length(intersect(resistant_genes, te_genes_24hr)), "genes\n\n")

cat("Sensitive_4E overlap with TE data:\n")
cat("1hr:", length(intersect(sensitive_genes, te_genes_1hr)), "genes\n")
cat("4hr:", length(intersect(sensitive_genes, te_genes_4hr)), "genes\n")
cat("24hr:", length(intersect(sensitive_genes, te_genes_24hr)), "genes\n")
```

## ECDF Plot: Hypoxic Response by 4E Sensitivity

```{r ecdf_4e_sensitivity, fig.width=6, fig.height=5}
# Classify genes by 4E sensitivity status
te_1hr_4e_status <- te_1hr_gsea %>%
  mutate(
    status_4E = case_when(
      symbol %in% resistant_set$gene ~ "4E-Resistant",
      symbol %in% sensitive_set$gene ~ "4E-Sensitive",
      TRUE ~ "Other"
    )
  )

# Create ECDF plot with all three categories
ggplot(te_1hr_4e_status, aes(x = log2FoldChange, color = status_4E)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("4E-Resistant" = "#D55E00", "4E-Sensitive" = "#0072B2", "Other" = "#808080"),
    name = "4E Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by 4E Sensitivity",
    subtitle = "Translation Efficiency Changes at 1 Hour",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
    xlim(-1,1)

  # Save the plot
ggsave(
  here("plots", "te_sictrl_telemangroups.pdf"),
  width = 5,
  height = 4
)


# 4 hr

te_4hr_4e_status <- te_4hr_gsea %>%
  mutate(
    status_4E = case_when(
      symbol %in% resistant_set$gene ~ "4E-Resistant",
      symbol %in% sensitive_set$gene ~ "4E-Sensitive",
      TRUE ~ "Other"
    )
  )

# Create ECDF plot with all three categories
ggplot(te_4hr_4e_status, aes(x = log2FoldChange, color = status_4E)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("4E-Resistant" = "#D55E00", "4E-Sensitive" = "#0072B2", "Other" = "#808080"),
    name = "4E Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by 4E Sensitivity",
    subtitle = "Translation Efficiency Changes at 4 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
    xlim(-1,1)

  # Save the plot
ggsave(
  here("plots", "te_sictrl_telemangroups_4hr.pdf"),
  width = 5,
  height = 4
)


# 24 hr

te_24hr_4e_status <- te_24hr_gsea %>%
  mutate(
    status_4E = case_when(
      symbol %in% resistant_set$gene ~ "4E-Resistant",
      symbol %in% sensitive_set$gene ~ "4E-Sensitive",
      TRUE ~ "Other"
    )
  )

# Create ECDF plot with all three categories
ggplot(te_24hr_4e_status, aes(x = log2FoldChange, color = status_4E)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("4E-Resistant" = "#D55E00", "4E-Sensitive" = "#0072B2", "Other" = "#808080"),
    name = "4E Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by 4E Sensitivity",
    subtitle = "Translation Efficiency Changes at 24 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )+
    xlim(-1,1)

  # Save the plot
ggsave(
  here("plots", "te_sictrl_telemangroups_24hr.pdf"),
  width = 5,
  height = 4
)
```

## ECDF Plot: Hypoxic Response by DAP5 Sensitivity (Weber 2022)

```{r ecdf_dap5_sensitivity, fig.width=6, fig.height=5}
# 1 hr
te_1hr_dap5_status <- te_1hr_gsea %>%
  mutate(
    status_DAP5 = case_when(
      symbol %in% dap5_up$symbol ~ "DAP5-Up",
      symbol %in% dap5_down$symbol ~ "DAP5-Down",
      TRUE ~ "Other"
    )
  )

ggplot(te_1hr_dap5_status, aes(x = log2FoldChange, color = status_DAP5)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("DAP5-Up" = "#D55E00", "DAP5-Down" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Sensitivity",
    subtitle = "Translation Efficiency Changes at 1 Hour",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

ggsave(
  here("plots", "te_sictrl_dap5groups_1hr.pdf"),
  width = 5,
  height = 4
)

# 4 hr
te_4hr_dap5_status <- te_4hr_gsea %>%
  mutate(
    status_DAP5 = case_when(
      symbol %in% dap5_up$symbol ~ "DAP5-Up",
      symbol %in% dap5_down$symbol ~ "DAP5-Down",
      TRUE ~ "Other"
    )
  )

ggplot(te_4hr_dap5_status, aes(x = log2FoldChange, color = status_DAP5)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("DAP5-Up" = "#D55E00", "DAP5-Down" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Sensitivity",
    subtitle = "Translation Efficiency Changes at 4 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

ggsave(
  here("plots", "te_sictrl_dap5groups_4hr.pdf"),
  width = 5,
  height = 4
)

# 24 hr
te_24hr_dap5_status <- te_24hr_gsea %>%
  mutate(
    status_DAP5 = case_when(
      symbol %in% dap5_up$symbol ~ "DAP5-Up",
      symbol %in% dap5_down$symbol ~ "DAP5-Down",
      TRUE ~ "Other"
    )
  )

ggplot(te_24hr_dap5_status, aes(x = log2FoldChange, color = status_DAP5)) +
  stat_ecdf(size = 1.2) +
  scale_color_manual(
    values = c("DAP5-Up" = "#D55E00", "DAP5-Down" = "#0072B2", "Other" = "#808080"),
    name = "DAP5 Status"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Hypoxic Response by DAP5 Sensitivity",
    subtitle = "Translation Efficiency Changes at 24 Hours",
    x = "TE log2 Fold Change (Hypoxia vs Normoxia)",
    y = "Cumulative Proportion"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  xlim(-1, 1)

ggsave(
  here("plots", "te_sictrl_dap5groups_24hr.pdf"),
  width = 5,
  height = 4
)
```

## Prepare Translation Signatures for GSEA

```{r prepare-translation-sigs}
# Convert translation_sigs to TERM2GENE format for GSEA
translation_sigs_term2gene <- translation_sigs %>%
  dplyr::select(gs_name, db_gene_symbol) %>%
  filter(!is.na(db_gene_symbol)) %>%
  rename(gene = db_gene_symbol)

# Check gene set sizes and overlaps
cat("\nTranslation Signature Gene Sets:\n")
translation_sigs_summary <- translation_sigs_term2gene %>%
  group_by(gs_name) %>%
  summarize(total_genes = n(), .groups = 'drop') %>%
  arrange(desc(total_genes))
print(translation_sigs_summary)

# Check overlap with TE data
cat("\nOverlap with TE data (1hr):\n")
for(gs in unique(translation_sigs_term2gene$gs_name)) {
  genes <- translation_sigs_term2gene %>% filter(gs_name == gs) %>% pull(gene)
  overlap <- length(intersect(genes, te_genes_1hr))
  cat(sprintf("  %s: %d / %d genes\n", gs, overlap, length(genes)))
}
```

## Prepare Ranked Lists for GSEA

```{r prepare-ranked-lists}
# Create ranked lists based on TE log2FC for each timepoint
ranked_list_te_1hr <- te_1hr_gsea$log2FoldChange
names(ranked_list_te_1hr) <- te_1hr_gsea$symbol
ranked_list_te_1hr <- ranked_list_te_1hr[!is.na(ranked_list_te_1hr)]
ranked_list_te_1hr <- sort(ranked_list_te_1hr, decreasing = TRUE)

ranked_list_te_4hr <- te_4hr_gsea$log2FoldChange
names(ranked_list_te_4hr) <- te_4hr_gsea$symbol
ranked_list_te_4hr <- ranked_list_te_4hr[!is.na(ranked_list_te_4hr)]
ranked_list_te_4hr <- sort(ranked_list_te_4hr, decreasing = TRUE)

ranked_list_te_24hr <- te_24hr_gsea$log2FoldChange
names(ranked_list_te_24hr) <- te_24hr_gsea$symbol
ranked_list_te_24hr <- ranked_list_te_24hr[!is.na(ranked_list_te_24hr)]
ranked_list_te_24hr <- sort(ranked_list_te_24hr, decreasing = TRUE)

cat("Ranked lists prepared:\n")
cat("1hr:", length(ranked_list_te_1hr), "genes\n")
cat("  - Range:", round(min(ranked_list_te_1hr), 3), "to", round(max(ranked_list_te_1hr), 3), "\n")
cat("  - Duplicated names:", sum(duplicated(names(ranked_list_te_1hr))), "\n")
cat("4hr:", length(ranked_list_te_4hr), "genes\n")
cat("24hr:", length(ranked_list_te_24hr), "genes\n")

# Check overlap with gene sets
cat("\nGene set overlap check:\n")
cat("Resistant genes in ranked list 1hr:", sum(resistant_genes %in% names(ranked_list_te_1hr)), "/", length(resistant_genes), "\n")
cat("Sensitive genes in ranked list 1hr:", sum(sensitive_genes %in% names(ranked_list_te_1hr)), "/", length(sensitive_genes), "\n")
```

## Teleman 4E GSEA

## GSEA: TE 1hr

```{r gsea_te_1hr}
# Run GSEA (unweighted)
gsea_te_1hr <- GSEA(
  geneList = ranked_list_te_1hr,
  exponent = 0,  # Use unweighted Kolmogorov-Smirnov statistic
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = teleman_4E_genesets,
  minGSSize = 10,
  maxGSSize = 1000  # Increased to accommodate larger gene sets (574-628 genes)
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_te_1hr@result) > 0) {
  gsea_te_1hr_results <- gsea_te_1hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_1hr_results[] <- lapply(gsea_te_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE 1hr results:\n")
  print(gsea_te_1hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  # Create empty dataframe with correct structure
  gsea_te_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE 1hr - No gene sets met minimum size requirement (minGSSize = 10)\n")
}

# Create enrichment score plots if results exist
if(nrow(gsea_te_1hr_results) > 0) {
  # Create individual plots for each gene set with NES and padj annotations
  for(i in 1:nrow(gsea_te_1hr@result)) {
    gene_set_name <- gsea_te_1hr@result$ID[i]
    nes_val <- round(gsea_te_1hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_1hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    p <- gseaplot2(gsea_te_1hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
  }

  # Save each gene set to separate PDF files
  for(i in 1:nrow(gsea_te_1hr@result)) {
    gene_set_name <- gsea_te_1hr@result$ID[i]
    nes_val <- round(gsea_te_1hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_1hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    cat("Plotting gene set:", gene_set_name, "\n")

    # Save to individual PDF file
    pdf_name <- sprintf("teleman_4E_gsea_enrichment_1hr_%s.pdf", gene_set_name)
    pdf(here("plots", pdf_name), width = 4, height = 3)

    p <- gseaplot2(gsea_te_1hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
    dev.off()
  }

  cat("Total plots created for 1hr:", nrow(gsea_te_1hr@result), "\n")
}
```

## GSEA: TE 4hr

```{r gsea_te_4hr}
# Run GSEA (unweighted)
gsea_te_4hr <- GSEA(
  geneList = ranked_list_te_4hr,
  exponent = 0,  # Use unweighted Kolmogorov-Smirnov statistic
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = teleman_4E_genesets,
  minGSSize = 10,
  maxGSSize = 1000  # Increased to accommodate larger gene sets (574-628 genes)
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_te_4hr@result) > 0) {
  gsea_te_4hr_results <- gsea_te_4hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_4hr_results[] <- lapply(gsea_te_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE 4hr results:\n")
  print(gsea_te_4hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  # Create empty dataframe with correct structure
  gsea_te_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE 4hr - No gene sets met minimum size requirement (minGSSize = 10)\n")
}

# Create enrichment score plots if results exist
if(nrow(gsea_te_4hr_results) > 0) {
  # Create individual plots for each gene set with NES and padj annotations
  for(i in 1:nrow(gsea_te_4hr@result)) {
    gene_set_name <- gsea_te_4hr@result$ID[i]
    nes_val <- round(gsea_te_4hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_4hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    p <- gseaplot2(gsea_te_4hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
  }

  # Save each gene set to separate PDF files
  for(i in 1:nrow(gsea_te_4hr@result)) {
    gene_set_name <- gsea_te_4hr@result$ID[i]
    nes_val <- round(gsea_te_4hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_4hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    cat("Plotting gene set:", gene_set_name, "\n")

    # Save to individual PDF file
    pdf_name <- sprintf("teleman_4E_gsea_enrichment_4hr_%s.pdf", gene_set_name)
    pdf(here("plots", pdf_name), width = 4, height = 3)

    p <- gseaplot2(gsea_te_4hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
    dev.off()
  }

  cat("Total plots created for 4hr:", nrow(gsea_te_4hr@result), "\n")
}
```

## GSEA: TE 24hr

```{r gsea_te_24hr}
# Run GSEA (unweighted)
gsea_te_24hr <- GSEA(
  geneList = ranked_list_te_24hr,
  exponent = 0,  # Use unweighted Kolmogorov-Smirnov statistic
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = teleman_4E_genesets,
  minGSSize = 10,
  maxGSSize = 1000  # Increased to accommodate larger gene sets (574-628 genes)
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_te_24hr@result) > 0) {
  gsea_te_24hr_results <- gsea_te_24hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_te_24hr_results[] <- lapply(gsea_te_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA TE 24hr results:\n")
  print(gsea_te_24hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  # Create empty dataframe with correct structure
  gsea_te_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA TE 24hr - No gene sets met minimum size requirement (minGSSize = 10)\n")
}

# Create enrichment score plots if results exist
if(nrow(gsea_te_24hr_results) > 0) {
  # Create individual plots for each gene set with NES and padj annotations
  for(i in 1:nrow(gsea_te_24hr@result)) {
    gene_set_name <- gsea_te_24hr@result$ID[i]
    nes_val <- round(gsea_te_24hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_24hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    p <- gseaplot2(gsea_te_24hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
  }

  # Save each gene set to separate PDF files
  for(i in 1:nrow(gsea_te_24hr@result)) {
    gene_set_name <- gsea_te_24hr@result$ID[i]
    nes_val <- round(gsea_te_24hr@result$NES[i], 3)
    padj_val <- formatC(gsea_te_24hr@result$p.adjust[i], format = "e", digits = 2)

    plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

    cat("Plotting gene set:", gene_set_name, "\n")

    # Save to individual PDF file
    pdf_name <- sprintf("teleman_4E_gsea_enrichment_24hr_%s.pdf", gene_set_name)
    pdf(here("plots", pdf_name), width = 4, height = 3)

    p <- gseaplot2(gsea_te_24hr,
                   geneSetID = i,
                   title = plot_title,
                   pvalue_table = FALSE,
                   ES_geom = "line")
    print(p)
    dev.off()
  }

  cat("Total plots created for 24hr:", nrow(gsea_te_24hr@result), "\n")
}
```

# GSEA: Translation Signatures from Literature

## GSEA: Translation Sigs 1hr

```{r gsea_translation_sigs_1hr}
# Run GSEA (unweighted)
gsea_trans_1hr <- GSEA(
  geneList = ranked_list_te_1hr,
  exponent = 0,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = translation_sigs_term2gene,
  minGSSize = 10,
  maxGSSize = 2000  # Larger to accommodate sets up to 1,558 genes
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_trans_1hr@result) > 0) {
  gsea_trans_1hr_results <- gsea_trans_1hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_trans_1hr_results[] <- lapply(gsea_trans_1hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA Translation Sigs 1hr results:\n")
  print(gsea_trans_1hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  gsea_trans_1hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA Translation Sigs 1hr - No gene sets met minimum size requirement\n")
}

# Create enrichment plots for significant results
if(nrow(gsea_trans_1hr_results) > 0) {
  sig_results <- gsea_trans_1hr_results %>% filter(p.adjust < 0.05)
  if(nrow(sig_results) > 0) {
    for(i in which(gsea_trans_1hr@result$p.adjust < 0.05)) {
      gene_set_name <- gsea_trans_1hr@result$ID[i]
      nes_val <- round(gsea_trans_1hr@result$NES[i], 3)
      padj_val <- formatC(gsea_trans_1hr@result$p.adjust[i], format = "e", digits = 2)

      plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

      # Save to individual PDF
      pdf_name <- sprintf("translation_sigs_gsea_enrichment_1hr_%s.pdf",
                         gsub(" ", "_", gene_set_name))
      pdf(here("plots", pdf_name), width = 8, height = 4)
      p <- gseaplot2(gsea_trans_1hr,
                     geneSetID = i,
                     title = plot_title,
                     pvalue_table = FALSE,
                     ES_geom = "line")
      print(p)
      dev.off()
    }
    cat("Created", nrow(sig_results), "enrichment plots for significant results\n")
  }
}
```

## GSEA: Translation Sigs 4hr

```{r gsea_translation_sigs_4hr}
# Run GSEA (unweighted)
gsea_trans_4hr <- GSEA(
  geneList = ranked_list_te_4hr,
  exponent = 0,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = translation_sigs_term2gene,
  minGSSize = 10,
  maxGSSize = 2000
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_trans_4hr@result) > 0) {
  gsea_trans_4hr_results <- gsea_trans_4hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_trans_4hr_results[] <- lapply(gsea_trans_4hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA Translation Sigs 4hr results:\n")
  print(gsea_trans_4hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  gsea_trans_4hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA Translation Sigs 4hr - No gene sets met minimum size requirement\n")
}

# Create enrichment plots for significant results
if(nrow(gsea_trans_4hr_results) > 0) {
  sig_results <- gsea_trans_4hr_results %>% filter(p.adjust < 0.05)
  if(nrow(sig_results) > 0) {
    for(i in which(gsea_trans_4hr@result$p.adjust < 0.05)) {
      gene_set_name <- gsea_trans_4hr@result$ID[i]
      nes_val <- round(gsea_trans_4hr@result$NES[i], 3)
      padj_val <- formatC(gsea_trans_4hr@result$p.adjust[i], format = "e", digits = 2)

      plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

      # Save to individual PDF
      pdf_name <- sprintf("translation_sigs_gsea_enrichment_4hr_%s.pdf",
                         gsub(" ", "_", gene_set_name))
      pdf(here("plots", pdf_name), width = 8, height = 4)
      p <- gseaplot2(gsea_trans_4hr,
                     geneSetID = i,
                     title = plot_title,
                     pvalue_table = FALSE,
                     ES_geom = "line")
      print(p)
      dev.off()
    }
    cat("Created", nrow(sig_results), "enrichment plots for significant results\n")
  }
}
```

## GSEA: Translation Sigs 24hr

```{r gsea_translation_sigs_24hr}
# Run GSEA (unweighted)
gsea_trans_24hr <- GSEA(
  geneList = ranked_list_te_24hr,
  exponent = 0,
  eps = 0,
  pAdjustMethod = "fdr",
  pvalueCutoff = 1,
  TERM2GENE = translation_sigs_term2gene,
  minGSSize = 10,
  maxGSSize = 2000
)

# Extract results - ALWAYS extract, even if not significant
if (nrow(gsea_trans_24hr@result) > 0) {
  gsea_trans_24hr_results <- gsea_trans_24hr@result %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
    rename("leading_edge" = core_enrichment) %>%
    mutate(leading_edge_count = sapply(strsplit(leading_edge, "/"), length),
           leading_edge_fraction = leading_edge_count / setSize)

  gsea_trans_24hr_results[] <- lapply(gsea_trans_24hr_results, function(x) {
    if (is.numeric(x)) round(x, 4) else x
  })

  cat("GSEA Translation Sigs 24hr results:\n")
  print(gsea_trans_24hr_results %>% dplyr::select(ID, NES, pvalue, p.adjust, setSize))
} else {
  gsea_trans_24hr_results <- data.frame(
    ID = character(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    setSize = numeric(),
    leading_edge = character(),
    leading_edge_count = numeric(),
    leading_edge_fraction = numeric(),
    stringsAsFactors = FALSE
  )
  cat("GSEA Translation Sigs 24hr - No gene sets met minimum size requirement\n")
}

# Create enrichment plots for significant results
if(nrow(gsea_trans_24hr_results) > 0) {
  sig_results <- gsea_trans_24hr_results %>% filter(p.adjust < 0.05)
  if(nrow(sig_results) > 0) {
    for(i in which(gsea_trans_24hr@result$p.adjust < 0.05)) {
      gene_set_name <- gsea_trans_24hr@result$ID[i]
      nes_val <- round(gsea_trans_24hr@result$NES[i], 3)
      padj_val <- formatC(gsea_trans_24hr@result$p.adjust[i], format = "e", digits = 2)

      plot_title <- sprintf("%s\nNES = %s, padj = %s", gene_set_name, nes_val, padj_val)

      # Save to individual PDF
      pdf_name <- sprintf("translation_sigs_gsea_enrichment_24hr_%s.pdf",
                         gsub(" ", "_", gene_set_name))
      pdf(here("plots", pdf_name), width = 8, height = 4)
      p <- gseaplot2(gsea_trans_24hr,
                     geneSetID = i,
                     title = plot_title,
                     pvalue_table = FALSE,
                     ES_geom = "line")
      print(p)
      dev.off()
    }
    cat("Created", nrow(sig_results), "enrichment plots for significant results\n")
  }
}
```

## Export TE GSEA Results

```{r export_gsea_te}
# Compile all TE GSEA results - handle empty results
prepare_for_export <- function(df) {
  if (nrow(df) > 0 && "leading_edge" %in% colnames(df)) {
    return(df %>% dplyr::select(-leading_edge))
  } else if (nrow(df) > 0) {
    return(df)
  } else {
    # Return empty dataframe with basic structure
    return(data.frame(
      ID = character(),
      NES = numeric(),
      pvalue = numeric(),
      p.adjust = numeric(),
      setSize = numeric(),
      stringsAsFactors = FALSE
    ))
  }
}

gsea_te_results_export <- list(
  "GSEA_TE_1hr" = prepare_for_export(gsea_te_1hr_results),
  "GSEA_TE_4hr" = prepare_for_export(gsea_te_4hr_results),
  "GSEA_TE_24hr" = prepare_for_export(gsea_te_24hr_results)
)

write_xlsx(gsea_te_results_export, here("output/teleman_4E_gsea_te_timecourse.xlsx"))
cat("\nTE GSEA results exported to: output/teleman_4E_gsea_te_timecourse.xlsx\n")
```

## TE GSEA Heatmap: Gene Set NES Across Timepoints (Always Show)

```{r gsea_te_heatmap}
# Combine results from all timepoints - ALWAYS combine, regardless of significance
gsea_te_combined <- bind_rows(
  gsea_te_1hr_results %>% mutate(timepoint = "1hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_te_4hr_results %>% mutate(timepoint = "4hr") %>% dplyr::select(ID, NES, p.adjust, timepoint),
  gsea_te_24hr_results %>% mutate(timepoint = "24hr") %>% dplyr::select(ID, NES, p.adjust, timepoint)
)

# Always create heatmap if we have any results
if (nrow(gsea_te_combined) > 0) {
  # Reshape for heatmap
  gsea_te_wide <- gsea_te_combined %>%
    dplyr::select(ID, NES, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
    column_to_rownames("ID")

  # Ensure all three timepoint columns exist
  for (tp in c("1hr", "4hr", "24hr")) {
    if (!tp %in% colnames(gsea_te_wide)) {
      gsea_te_wide[[tp]] <- 0
    }
  }
  # Reorder columns
  gsea_te_wide <- gsea_te_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

  # Create significance annotation matrix
  sig_te_matrix <- gsea_te_combined %>%
    dplyr::select(ID, p.adjust, timepoint) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(-p.adjust) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")

  # Ensure all three timepoint columns exist
  for (tp in c("1hr", "4hr", "24hr")) {
    if (!tp %in% colnames(sig_te_matrix)) {
      sig_te_matrix[[tp]] <- ""
    }
  }
  # Reorder columns and ensure row order matches
  sig_te_matrix <- sig_te_matrix[rownames(gsea_te_wide), c("1hr", "4hr", "24hr"), drop = FALSE]

  # Create heatmap
  p_te_heatmap <- pheatmap(
    gsea_te_wide,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(50),
    breaks = seq(-3, 3, length.out = 51),
    display_numbers = sig_te_matrix,
    number_color = "black",
    fontsize_number = 14,
    main = "Teleman 4E Gene Sets (LFC Threshold) - TE GSEA NES Across Timepoints (* = p.adjust < 0.05)",
    fontsize_row = 10,
    fontsize_col = 10,
    cellwidth = 40,
    cellheight = 30
  )

  # Save
  pdf(here("plots/teleman_4E_gsea_te_heatmap_timecourse.pdf"), width = 8, height = 6)
  print(p_te_heatmap)
  dev.off()

  cat("TE heatmap shows", nrow(gsea_te_wide), "gene sets\n")
} else {
  cat("No GSEA results available for TE heatmap\n")
}
```

## Translation Signatures GSEA Heatmap and Table

```{r translation_sigs_heatmap_and_table}
# Combine results from all timepoints
gsea_trans_combined <- bind_rows(
  gsea_trans_1hr_results %>% mutate(timepoint = "1hr") %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, timepoint),
  gsea_trans_4hr_results %>% mutate(timepoint = "4hr") %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, timepoint),
  gsea_trans_24hr_results %>% mutate(timepoint = "24hr") %>%
    dplyr::select(ID, NES, pvalue, p.adjust, setSize, timepoint)
)

# Create comprehensive table with NES and p.adjust for all gene sets
gsea_trans_table <- gsea_trans_combined %>%
  pivot_wider(
    id_cols = ID,
    names_from = timepoint,
    values_from = c(NES, p.adjust, setSize),
    names_glue = "{timepoint}_{.value}"
  ) %>%
  arrange(ID)

# Export table
write_xlsx(
  list("Translation_Sigs_GSEA" = gsea_trans_table),
  here("output/translation_sigs_gsea_te_timecourse_table.xlsx")
)

cat("\nTranslation Signatures GSEA Table exported to: output/translation_sigs_gsea_te_timecourse_table.xlsx\n")
print(gsea_trans_table)

# Create heatmap showing ALL pathways
if (nrow(gsea_trans_combined) > 0) {
  # Create heatmap with all gene sets (no filtering)
  gsea_trans_wide <- gsea_trans_combined %>%
    dplyr::select(ID, NES, timepoint) %>%
    pivot_wider(names_from = timepoint, values_from = NES, values_fill = 0) %>%
    column_to_rownames("ID")

  # Ensure all three timepoint columns exist
  for (tp in c("1hr", "4hr", "24hr")) {
    if (!tp %in% colnames(gsea_trans_wide)) {
      gsea_trans_wide[[tp]] <- 0
    }
  }
  gsea_trans_wide <- gsea_trans_wide[, c("1hr", "4hr", "24hr"), drop = FALSE]

  # Create significance annotation matrix for all gene sets
  sig_trans_matrix <- gsea_trans_combined %>%
    dplyr::select(ID, p.adjust, timepoint) %>%
    mutate(sig = ifelse(p.adjust < 0.05, "*", "")) %>%
    dplyr::select(-p.adjust) %>%
    pivot_wider(names_from = timepoint, values_from = sig, values_fill = "") %>%
    column_to_rownames("ID")

  # Ensure all columns exist and match row order
  for (tp in c("1hr", "4hr", "24hr")) {
    if (!tp %in% colnames(sig_trans_matrix)) {
      sig_trans_matrix[[tp]] <- ""
    }
  }
  sig_trans_matrix <- sig_trans_matrix[rownames(gsea_trans_wide), c("1hr", "4hr", "24hr"), drop = FALSE]
dev.off()
  # Create heatmap with all gene sets
  pheatmap(
    gsea_trans_wide,
    cluster_rows = TRUE,  # Cluster rows to group similar patterns
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(50),
    breaks = seq(-3, 3, length.out = 51),
    display_numbers = sig_trans_matrix,
    number_color = "black",
    fontsize_number = 12,
    main = "Translation Signatures - TE GSEA NES Across Timepoints\n(All gene sets, * = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 50,
    cellheight = 20
  )
dev.off()
  # Save heatmap to PDF with dynamic height based on number of gene sets
  pdf_height <- max(8, nrow(gsea_trans_wide) * 0.3 + 2)
  pdf(here("plots/translation_sigs_gsea_te_heatmap_timecourse.pdf"),
      width = 8, height = pdf_height)
  pheatmap(
    gsea_trans_wide,
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    scale = "none",
    color = colorRampPalette(c("blue", "white", "red"))(50),
    breaks = seq(-3, 3, length.out = 51),
    display_numbers = sig_trans_matrix,
    number_color = "black",
    fontsize_number = 12,
    main = "Translation Signatures - TE GSEA NES Across Timepoints\n(All gene sets, * = p.adjust < 0.05)",
    fontsize_row = 8,
    fontsize_col = 10,
    cellwidth = 50,
    cellheight = 20
  )
  dev.off()

  cat("Heatmap shows all", nrow(gsea_trans_wide), "gene sets\n")
} else {
  cat("\nNo GSEA results available for heatmap\n")
}
```

## Export Translation Signatures GSEA Results

```{r export_gsea_translation_sigs}
# Export translation signatures GSEA results for each timepoint
gsea_trans_results_export <- list(
  "GSEA_Translation_1hr" = prepare_for_export(gsea_trans_1hr_results),
  "GSEA_Translation_4hr" = prepare_for_export(gsea_trans_4hr_results),
  "GSEA_Translation_24hr" = prepare_for_export(gsea_trans_24hr_results)
)

write_xlsx(gsea_trans_results_export,
           here("output/translation_sigs_gsea_te_timecourse_detailed.xlsx"))

cat("\nTranslation Signatures GSEA detailed results exported to: output/translation_sigs_gsea_te_timecourse_detailed.xlsx\n")
```

