---
title: "RIP-seq QC: eIF3d, eIF4E, IgG-mock IP"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(janitor)
library(tximport)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(matrixStats)
library(ggpubr)
library(ggprism)
library(ggokabeito)
library(readxl)
```

```{r create_dirs}
dir.create(here("ripseq", "output"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("ripseq", "plots"),  showWarnings = FALSE, recursive = TRUE)
```

# Metadata

```{r metadata}
metadata <- read_csv(here("ripseq", "ripseq2026_metadata.csv")) |>
  clean_names() |>
  dplyr::rename(id = library_id) |>
  mutate(
    condition = tolower(treatment1),
    ip_type   = recode(treatment2,
                  "FLAG-eIF3d_IP" = "flag_eif3d",
                  "eIF4E_IP"      = "eif4e",
                  "IgG-mock_IP"   = "igg_mock"),
    rep       = tolower(replicate),
    hours     = as.factor(time_hr)
  )

metadata$condition <- relevel(factor(metadata$condition), ref = "normoxia")
metadata$ip_type   <- relevel(factor(metadata$ip_type),   ref = "igg_mock")

cat("Samples:", nrow(metadata), "\n")
cat("Conditions:", levels(metadata$condition), "\n")
cat("IP types:", levels(metadata$ip_type), "\n")
cat("Timepoints:", levels(metadata$hours), "hr\n")
cat("Replicates:", unique(metadata$rep), "\n")
```

# Initial Read QC

```{r read_qc_counts}
p_reads <- ggplot(metadata, aes(x = id, y = pf_clusters, fill = ip_type)) +
  geom_col() +
  scale_y_log10(
    breaks = 10^seq(1, 9),
    labels = c("10", "100", "1K", "10K", "100K", "1M", "10M", "100M", "1B")
  ) +
  annotate("rect", ymin = 1e7, ymax = Inf,  xmin = -Inf, xmax = Inf, fill = "green",  alpha = 0.12) +
  annotate("rect", ymin = 1e6, ymax = 1e7,  xmin = -Inf, xmax = Inf, fill = "orange", alpha = 0.12) +
  annotate("rect", ymin = 0,   ymax = 1e6,  xmin = -Inf, xmax = Inf, fill = "red",    alpha = 0.12) +
  scale_fill_okabe_ito() +
  labs(x = NULL, y = "PF clusters (log10)", fill = "IP type",
       title = "RIP-seq: sequencing depth") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p_reads
ggsave(here("ripseq", "plots", "ripseq_pf_clusters.pdf"), p_reads, width = 12, height = 5)
```

```{r read_qc_q30}
p_q30 <- ggplot(metadata, aes(x = id, y = percent_greater_than_equal_to_q30_bases, fill = ip_type)) +
  geom_col() +
  geom_hline(yintercept = 80, linetype = "dashed", color = "grey30", linewidth = 0.8) +
  scale_fill_okabe_ito() +
  labs(x = NULL, y = "% bases >= Q30", fill = "IP type",
       title = "RIP-seq: base quality (Q30)") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p_q30
ggsave(here("ripseq", "plots", "ripseq_q30.pdf"), p_q30, width = 12, height = 5)
```

# Gene Annotations

```{r gene_info}
geneInfo <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
                     col_names = FALSE)
colnames(geneInfo) <- c("gene_id", "symbol", "biotype", "species", "version")
geneInfo <- geneInfo |> filter(!str_detect(biotype, "pre_"))

tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = FALSE)
colnames(tx2gene) <- c("transcript", "gene")

gene_symbol <- geneInfo |>
  dplyr::select(gene_id, symbol) |>
  distinct(gene_id, .keep_all = TRUE)

cat("Genes in annotation:", nrow(gene_symbol), "\n")
```

# tximport

```{r tximport, include=FALSE}
salmon_base <- file.path(here(), "ripseq", "salmon", metadata$id)
myquantfiles <- ifelse(
  file.exists(file.path(salmon_base, "quant.sf.gz")),
  file.path(salmon_base, "quant.sf.gz"),
  file.path(salmon_base, "quant.sf")
)
names(myquantfiles) <- paste(metadata$condition, metadata$ip_type,
                              metadata$hours, metadata$rep, metadata$id, sep = "_")

missing <- myquantfiles[!file.exists(myquantfiles)]
if (length(missing) > 0) {
  cat("MISSING files:\n"); cat(missing, sep = "\n")
  stop("Some quant files not found")
}
cat("Uncompressed samples:", sum(!grepl("\\.gz$", myquantfiles)), "\n")

myTxi <- tximport(myquantfiles, type = "salmon", tx2gene = tx2gene)

cat("Genes in myTxi$counts:", nrow(myTxi$counts), "\n")
cat("Samples:", ncol(myTxi$counts), "\n")

if (any(duplicated(rownames(myTxi$counts)))) {
  dup_genes <- rownames(myTxi$counts)[duplicated(rownames(myTxi$counts))]
  cat("WARNING: Found", length(dup_genes), "duplicate gene IDs\n")
} else {
  cat("No duplicate gene IDs\n")
}
```

# Separate Human from Spike-in

```{r separate_human}
counts  <- myTxi$counts
is_dros <- grepl("^FB", rownames(counts))

# NOTE: No spike-ins were used in this experiment (SpikeIn = NA in metadata).
# Drosophila count should be 0. Spike-in removal from the Salmon reference index
# is currently in progress -- this check will catch any residual mapping.
cat("Drosophila genes detected:", sum(is_dros), "(expect 0 -- no spike-ins used)\n")

cts_human <- counts[!is_dros, , drop = FALSE]
cat("Human genes:", nrow(cts_human), "\n")
```

# Read Distribution and Cutoff

```{r read_distribution}
log2_rowsums <- log2(rowSums(cts_human))
log2_rowsums <- log2_rowsums[is.finite(log2_rowsums)]

hist(log2_rowsums, breaks = 60,
     main = "All samples: log2(rowSums) distribution",
     xlab = "log2(sum of counts across all 36 samples)",
     col = "steelblue", border = "white")
abline(v = 10, col = "red", lty = 2, lwd = 2)
legend("topright", legend = "cutoff = 9", col = "red", lty = 2)
```

```{r apply_cutoff}
keepGenes <- rownames(cts_human)[log2(rowSums(cts_human)) > 10]
cat("Genes passing log2(rowSums) > 10 cutoff:", length(keepGenes), "\n")
```

# Gene Filtering

```{r gene_filtering}
pc <- geneInfo |>
  filter(biotype == "protein_coding") |>
  pull(gene_id) |>
  unique()

nonpolyA_ID <- geneInfo |>
  filter(symbol %in% pull(read_xlsx(here("accessories", "polyAminus.xlsx"), sheet = 1), Gene)) |>
  pull(gene_id) |>
  unique()

shortgenes <- read_csv(here("accessories", "human", "short_genes_under_200nt_human_v49.csv.zip"),
                       show_col_types = FALSE) |>
  pull(gene_id)

allCounts <- cts_human |>
  as.data.frame() |>
  rownames_to_column("gene") |>
  filter(gene %in% keepGenes,
         gene %in% pc,
         !gene %in% nonpolyA_ID,
         !gene %in% shortgenes) |>
  column_to_rownames("gene") |>
  as.matrix() |>
  round(0)

cat("After expression cutoff:", length(keepGenes), "genes\n")
cat("After protein-coding filter:", sum(keepGenes %in% pc), "genes\n")
cat("After all filters:", nrow(allCounts), "genes\n")
```

# RNA-seq Expression Filter

Restrict to genes expressed in the riboseq RNA-seq (Input_Ribo_RNAseq) experiment so the RIP-seq universe is anchored to confirmed transcribed genes.

```{r rnaseq_filter}
rna_meta <- read_csv(here("accessories", "metadata_231_timecourse.csv")) |>
  filter(type == "Input_Ribo_RNAseq")

rna_raw <- read_csv(here("output", "human_allCounts_raw.csv")) |>
  column_to_rownames("gene_id")

# Column names are like "normoxia_sictrl_1_a_rna_NM2025_0085"; extract NM ID to match
col_ids <- sub(".*_(NM\\d+_\\d+)$", "\\1", colnames(rna_raw))
rna_only <- rna_raw[, col_ids %in% rna_meta$id, drop = FALSE]

cat("RNA-seq samples used:", ncol(rna_only), "\n")

keepGenes_rnaseq <- rownames(rna_only)[log2(rowSums(rna_only)) > 9]
allCounts <- allCounts[rownames(allCounts) %in% keepGenes_rnaseq, ]
cat("After RNA-seq expression filter:", nrow(allCounts), "genes\n")
```

# QC Input: SD Filtering

```{r sd_filter}
qcinput <- log2(myTxi$abundance[rownames(allCounts), ] + 1)
qcinput  <- qcinput[matrixStats::rowSds(qcinput) > 1, ]
cat("Genes after SD > 1 filter:", nrow(qcinput), "\n")
```

# Correlation Heatmap

```{r heatmap}
mycor <- cor(qcinput, method = "pearson")

heatmap_all <- pheatmap::pheatmap(
  mat          = mycor,
  border_color = "black",
  scale        = "none",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize     = 7
)

heatmap_all
ggsave(here("ripseq", "plots", "ripseq_qc_heatmap.pdf"), heatmap_all, width = 14, height = 13)
```

# PCA

```{r pca_base}
pca_all <- prcomp(t(qcinput), scale. = TRUE)

pca_data <- as.data.frame(pca_all$x) |>
  rownames_to_column("sample") |>
  mutate(id = sub(".*_(NM\\d+_\\d+)$", "\\1", sample)) |>
  left_join(dplyr::select(metadata, id, condition, ip_type, hours, rep), by = "id")

pc1_var <- round(summary(pca_all)$importance[2, 1] * 100, 1)
pc2_var <- round(summary(pca_all)$importance[2, 2] * 100, 1)
```

## PCA: IP type x condition

```{r pca_iptype}
p_pca_ip <- ggplot(pca_data, aes(x = PC1, y = PC2, color = ip_type, shape = condition)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = paste0("PC1 (", pc1_var, "%)"),
       y = paste0("PC2 (", pc2_var, "%)"),
       color = "IP type", shape = "Condition",
       title = "PCA - all samples: IP type") +
  theme_classic(base_size = 12)

p_pca_ip
ggsave(here("ripseq", "plots", "ripseq_pca_iptype.pdf"), p_pca_ip, width = 7, height = 5)
```

## PCA: timepoint x IP type

```{r pca_timepoint}
p_pca_time <- ggplot(pca_data, aes(x = PC1, y = PC2, color = hours, shape = ip_type)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = paste0("PC1 (", pc1_var, "%)"),
       y = paste0("PC2 (", pc2_var, "%)"),
       color = "Timepoint (hr)", shape = "IP type",
       title = "PCA - all samples: timepoint") +
  theme_classic(base_size = 12)

p_pca_time
ggsave(here("ripseq", "plots", "ripseq_pca_timepoint.pdf"), p_pca_time, width = 7, height = 5)
```

## PCA: replicate

```{r pca_rep}
p_pca_rep <- ggplot(pca_data, aes(x = PC1, y = PC2, color = rep, shape = ip_type)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = paste0("PC1 (", pc1_var, "%)"),
       y = paste0("PC2 (", pc2_var, "%)"),
       color = "Replicate", shape = "IP type",
       title = "PCA - all samples: replicate") +
  theme_classic(base_size = 12)

p_pca_rep
ggsave(here("ripseq", "plots", "ripseq_pca_replicate.pdf"), p_pca_rep, width = 7, height = 5)
```

# PCA: Raw Counts (un-normalized)

```{r pca_raw_base}
qcinput_raw <- log1p(allCounts)
qcinput_raw <- qcinput_raw[matrixStats::rowSds(qcinput_raw) > 1, ]
cat("Genes after SD > 1 filter (raw counts):", nrow(qcinput_raw), "\n")

pca_raw <- prcomp(t(qcinput_raw), scale. = TRUE)

pca_raw_data <- as.data.frame(pca_raw$x) |>
  rownames_to_column("sample") |>
  mutate(id = sub(".*_(NM\\d+_\\d+)$", "\\1", sample)) |>
  left_join(dplyr::select(metadata, id, condition, ip_type, hours, rep), by = "id")

pc1_raw <- round(summary(pca_raw)$importance[2, 1] * 100, 1)
pc2_raw <- round(summary(pca_raw)$importance[2, 2] * 100, 1)
```

```{r pca_raw_iptype}
p_pca_raw_ip <- ggplot(pca_raw_data, aes(x = PC1, y = PC2, color = ip_type, shape = condition)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = paste0("PC1 (", pc1_raw, "%)"),
       y = paste0("PC2 (", pc2_raw, "%)"),
       color = "IP type", shape = "Condition",
       title = "PCA - all samples: IP type (raw counts, log1p)") +
  theme_classic(base_size = 12)

p_pca_raw_ip
ggsave(here("ripseq", "plots", "00_pca_raw_counts.pdf"), p_pca_raw_ip, width = 7, height = 5)
```

# IP CPM vs RNA CPM

Scatter of mean IP CPM (RIP-seq, averaged across replicates per condition/timepoint) against mean RNA CPM (riboseq RNA-seq experiment), faceted by IP type. Positive correlation is expected; deviations highlight enrichment or depletion by the IP.

```{r cor_rna_ip}
# RNA CPM: mean across riboseq RNA-seq samples
shared_genes <- intersect(rownames(allCounts), rownames(rna_only))
rna_sub      <- rna_only[shared_genes, , drop = FALSE]
rna_cpm_mean <- setNames(rowMeans(sweep(rna_sub, 2, colSums(rna_sub), "/") * 1e6),
                         shared_genes)

rip_cpm_mat <- sweep(allCounts[shared_genes, ], 2, colSums(allCounts[shared_genes, ]), "/") * 1e6

rip_long_all <- as.data.frame(rip_cpm_mat) |>
  rownames_to_column("gene_id") |>
  pivot_longer(-gene_id, names_to = "sample_name", values_to = "cpm") |>
  mutate(id = sub(".*_(NM\\d+_\\d+)$", "\\1", sample_name)) |>
  left_join(dplyr::select(metadata, id, condition, ip_type, hours), by = "id") |>
  group_by(gene_id, ip_type, condition, hours) |>
  summarise(mean_ip_cpm = mean(cpm, na.rm = TRUE), .groups = "drop") |>
  mutate(rna_cpm   = rna_cpm_mean[gene_id],
         timepoint = paste0(condition, "_", hours, "hr"))

r2_all <- rip_long_all |>
  filter(!is.na(rna_cpm), !is.na(mean_ip_cpm)) |>
  group_by(ip_type) |>
  summarise(r2    = cor(log10(rna_cpm + 1), log10(mean_ip_cpm + 1))^2,
            .groups = "drop") |>
  mutate(label = paste0("R2 = ", round(r2, 3)))

p_cor_all <- ggplot(rip_long_all, aes(x = rna_cpm + 1, y = mean_ip_cpm + 1, color = timepoint)) +
  geom_point(size = 0.4, alpha = 0.25) +
  geom_text(data = r2_all, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.1, vjust = 1.4, size = 3.5, color = "black") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_okabe_ito() +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(x = "RNA CPM (riboseq, log10)",
       y = "IP CPM (RIP-seq, log10)",
       color = "Condition",
       title = "IP CPM vs RNA CPM by IP type (all samples)") +
  theme_classic(base_size = 12)

p_cor_all
ggsave(here("ripseq", "plots", "00a_ip_vs_rna_cpm_all.pdf"), p_cor_all, width = 14, height = 5)
```

```{r cor_rna_ip_nozero}
rip_long_all_nz <- rip_long_all |> filter(mean_ip_cpm > 0)

r2_all_nz <- rip_long_all_nz |>
  filter(!is.na(rna_cpm)) |>
  group_by(ip_type) |>
  summarise(r2    = cor(log10(rna_cpm + 1), log10(mean_ip_cpm + 1))^2,
            .groups = "drop") |>
  mutate(label = paste0("R2 = ", round(r2, 3)))

p_cor_all_nz <- ggplot(rip_long_all_nz, aes(x = rna_cpm + 1, y = mean_ip_cpm + 1, color = timepoint)) +
  geom_point(size = 0.4, alpha = 0.25) +
  geom_text(data = r2_all_nz, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.1, vjust = 1.4, size = 3.5, color = "black") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_okabe_ito() +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(x = "RNA CPM (riboseq, log10)",
       y = "IP CPM (RIP-seq, log10)",
       color = "Condition",
       title = "IP CPM vs RNA CPM by IP type (all samples, zeros excluded)") +
  theme_classic(base_size = 12)

p_cor_all_nz
ggsave(here("ripseq", "plots", "00a_ip_vs_rna_cpm_all_nozero.pdf"), p_cor_all_nz, width = 14, height = 5)
```

```{r cor_rna_ip_no_outlier}
cts_no_out  <- allCounts[, !grepl("0077", colnames(allCounts))]
meta_no_out <- metadata |> dplyr::filter(!grepl("0077", id))

rip_cpm_mat_no <- sweep(cts_no_out[shared_genes, ], 2, colSums(cts_no_out[shared_genes, ]), "/") * 1e6

rip_long_no <- as.data.frame(rip_cpm_mat_no) |>
  rownames_to_column("gene_id") |>
  pivot_longer(-gene_id, names_to = "sample_name", values_to = "cpm") |>
  mutate(id = sub(".*_(NM\\d+_\\d+)$", "\\1", sample_name)) |>
  left_join(dplyr::select(meta_no_out, id, condition, ip_type, hours), by = "id") |>
  group_by(gene_id, ip_type, condition, hours) |>
  summarise(mean_ip_cpm = mean(cpm, na.rm = TRUE), .groups = "drop") |>
  mutate(rna_cpm   = rna_cpm_mean[gene_id],
         timepoint = paste0(condition, "_", hours, "hr"))

r2_no <- rip_long_no |>
  filter(!is.na(rna_cpm), !is.na(mean_ip_cpm)) |>
  group_by(ip_type) |>
  summarise(r2    = cor(log10(rna_cpm + 1), log10(mean_ip_cpm + 1))^2,
            .groups = "drop") |>
  mutate(label = paste0("R2 = ", round(r2, 3)))

p_cor_no <- ggplot(rip_long_no, aes(x = rna_cpm + 1, y = mean_ip_cpm + 1, color = timepoint)) +
  geom_point(size = 0.4, alpha = 0.25) +
  geom_text(data = r2_no, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.1, vjust = 1.4, size = 3.5, color = "black") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_okabe_ito() +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(x = "RNA CPM (riboseq, log10)",
       y = "IP CPM (RIP-seq, log10)",
       color = "Condition",
       title = "IP CPM vs RNA CPM by IP type (outlier 0077 removed)") +
  theme_classic(base_size = 12)

p_cor_no
ggsave(here("ripseq", "plots", "00a_ip_vs_rna_cpm_no_outlier.pdf"), p_cor_no, width = 14, height = 5)
```

```{r cor_rna_ip_no_outlier_nozero}
rip_long_no_nz <- rip_long_no |> filter(mean_ip_cpm > 0)

r2_no_nz <- rip_long_no_nz |>
  filter(!is.na(rna_cpm)) |>
  group_by(ip_type) |>
  summarise(r2    = cor(log10(rna_cpm + 1), log10(mean_ip_cpm + 1))^2,
            .groups = "drop") |>
  mutate(label = paste0("R2 = ", round(r2, 3)))

p_cor_no_nz <- ggplot(rip_long_no_nz, aes(x = rna_cpm + 1, y = mean_ip_cpm + 1, color = timepoint)) +
  geom_point(size = 0.4, alpha = 0.25) +
  geom_text(data = r2_no_nz, aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.1, vjust = 1.4, size = 3.5, color = "black") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_okabe_ito() +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(x = "RNA CPM (riboseq, log10)",
       y = "IP CPM (RIP-seq, log10)",
       color = "Condition",
       title = "IP CPM vs RNA CPM by IP type (outlier 0077 removed, zeros excluded)") +
  theme_classic(base_size = 12)

p_cor_no_nz
ggsave(here("ripseq", "plots", "00a_ip_vs_rna_cpm_no_outlier_nozero.pdf"), p_cor_no_nz, width = 14, height = 5)
```

## Per-sample scatter (all samples, before outlier exclusion)

```{r cor_rna_ip_per_sample}
rip_long_per_sample <- as.data.frame(rip_cpm_mat) |>
  rownames_to_column("gene_id") |>
  pivot_longer(-gene_id, names_to = "sample_name", values_to = "ip_cpm") |>
  mutate(id = sub(".*_(NM\\d+_\\d+)$", "\\1", sample_name)) |>
  left_join(dplyr::select(metadata, id, condition, ip_type, hours, rep), by = "id") |>
  mutate(rna_cpm     = rna_cpm_mean[gene_id],
         facet_label = paste0(id, " | ", ip_type, " | ", condition, " ", hours, "hr ", rep))

p_cor_per_sample <- ggplot(rip_long_per_sample,
                           aes(x = rna_cpm + 1, y = ip_cpm + 1, color = ip_type)) +
  geom_point(size = 0.3, alpha = 0.2) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c(
    "igg_mock"   = "#009E73",
    "eif4e"      = "#0072B2",
    "flag_eif3d" = "#E69F00"
  )) +
  facet_wrap(~ facet_label, ncol = 9) +
  labs(x = "RNA CPM (riboseq, log10)",
       y = "IP CPM (RIP-seq, log10)",
       color = "IP type",
       title = "IP CPM vs RNA CPM: per sample (all samples, before outlier exclusion)") +
  theme_classic(base_size = 7) +
  theme(strip.text = element_text(size = 6))

p_cor_per_sample
ggsave(here("ripseq", "plots", "00a_ip_vs_rna_cpm_per_sample.pdf"),
       p_cor_per_sample, width = 22, height = 16)
```

# Save Outputs

```{r save_outputs}
saveRDS(allCounts, here("ripseq", "output", "ripseq_allCounts_raw.rds"))

write_csv(
  as.data.frame(allCounts) |> rownames_to_column("gene_id"),
  here("ripseq", "output", "ripseq_allCounts_raw.csv")
)

write_csv(metadata, here("ripseq", "output", "ripseq_metadata_clean.csv"))

cat("Saved:\n")
cat("  ripseq/output/ripseq_allCounts_raw.rds\n")
cat("  ripseq/output/ripseq_allCounts_raw.csv\n")
cat("  ripseq/output/ripseq_metadata_clean.csv\n")
cat("Final dimensions:", nrow(allCounts), "genes x", ncol(allCounts), "samples (outlier NOT removed -- see 01)\n")
```
