---
title: "RIP-seq: Are Top TE Targets Identified in FLAG3D IP?"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
```

```{r load_te}
nor <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
hyp <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))

top_nor <- nor |> filter(te_padj < 0.05) |> arrange(desc(te_lfc)) |> slice_head(n = 500) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
top_hyp <- hyp |> filter(te_padj < 0.05) |> arrange(desc(te_lfc)) |> slice_head(n = 500) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
```

```{r load_rip}
rip_nor_full    <- read_csv(here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg_with_outlier.csv"))
rip_nor_clean   <- read_csv(here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg.csv"))
rip_nor_no_out2 <- read_csv(here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg_no_out2.csv"))
rip_hyp_full    <- read_csv(here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg_with_outlier.csv"))
rip_hyp_clean   <- read_csv(here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg.csv"))
rip_hyp_no_out2 <- read_csv(here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg_no_out2.csv"))
```

## Overlap: top 500 TE targets vs FLAG3D IP hits (LFC > 1, padj < 0.05)

```{r overlap_nor_full}
hits_nor_full <- top_nor |>
  inner_join(rip_nor_full |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Normoxia - with outlier:", nrow(hits_nor_full), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_nor_full, n = Inf)
```

```{r overlap_nor_clean}
hits_nor_clean <- top_nor |>
  inner_join(rip_nor_clean |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Normoxia - outlier removed:", nrow(hits_nor_clean), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_nor_clean, n = Inf)
```

```{r overlap_hyp_full}
hits_hyp_full <- top_hyp |>
  inner_join(rip_hyp_full |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Hypoxia - with outlier:", nrow(hits_hyp_full), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_hyp_full, n = Inf)
```

```{r overlap_hyp_clean}
hits_hyp_clean <- top_hyp |>
  inner_join(rip_hyp_clean |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Hypoxia - outlier removed:", nrow(hits_hyp_clean), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_hyp_clean, n = Inf)
```

```{r overlap_nor_no_out2}
hits_nor_no_out2 <- top_nor |>
  inner_join(rip_nor_no_out2 |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Normoxia - 0065 removed:", nrow(hits_nor_no_out2), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_nor_no_out2, n = Inf)
```

```{r overlap_hyp_no_out2}
hits_hyp_no_out2 <- top_hyp |>
  inner_join(rip_hyp_no_out2 |> filter(log2FoldChange > 1, padj < 0.05) |> dplyr::select(-symbol), by = "gene_id_clean") |>
  dplyr::select(symbol, te_lfc, log2FoldChange, padj)
cat("Hypoxia - 0065 removed:", nrow(hits_hyp_no_out2), "/ 500 TE targets found in FLAG3D IP (LFC > 1, padj < 0.05)\n")
print(hits_hyp_no_out2, n = Inf)
```

## Summary plot

```{r overlap_summary_plot}
summary_df <- tibble(
  condition = c("Normoxia", "Normoxia", "Normoxia", "Hypoxia", "Hypoxia", "Hypoxia"),
  outlier   = c("With outlier", "0077 removed", "0065 removed",
                "With outlier", "0077 removed", "0065 removed"),
  n_hits    = c(nrow(hits_nor_full), nrow(hits_nor_clean), nrow(hits_nor_no_out2),
                nrow(hits_hyp_full), nrow(hits_hyp_clean), nrow(hits_hyp_no_out2))
) |>
  mutate(
    condition = factor(condition, levels = c("Normoxia", "Hypoxia")),
    outlier   = factor(outlier, levels = c("With outlier", "0077 removed", "0065 removed"))
  )

p <- ggplot(summary_df, aes(x = outlier, y = n_hits, fill = outlier)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_hits), vjust = -0.4, size = 4) +
  facet_wrap(~ condition) +
  scale_fill_manual(values = c("With outlier" = "grey60",
                                "0077 removed" = "#E69F00",
                                "0065 removed" = "#009E73")) +
  labs(x = NULL, y = "Top 500 TE targets in FLAG3D IP\n(LFC > 1, padj < 0.05)",
       title = "Overlap of top 500 TE targets with FLAG3D IP hits") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

p
ggsave(here("ripseq", "plots", "07_te_targets_flag3d_overlap.pdf"), p, width = 8, height = 4)
```


```{r}

```