---
title: "RIP-seq: eIF4E and FLAG-eIF3d binding changes, hypoxia vs normoxia (interaction model)"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
```

# eIF3d-promoted gene set (riboseq TE, hypoxia 1hr)

```{r te_data}
te_hyp <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                   show_col_types = FALSE) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

promoted <- te_hyp |>
  filter(te_lfc > 0.5, te_padj < 0.05)

cat("eIF3d hypoxia-promoted genes (te_lfc > 0.5, te_padj < 0.05):", nrow(promoted), "\n")
```

# Load RIP-seq count matrix and metadata

```{r load_rip}
allCounts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
metadata  <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                      show_col_types = FALSE) |>
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia")),
    ip_type   = factor(ip_type,   levels = c("igg_mock", "flag_eif3d", "eif4e")),
    hours     = factor(hours)
  )

gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
                      col_names = c("gene_id", "symbol", "biotype", "species", "version"),
                      show_col_types = FALSE) |>
  filter(species == "human") |>
  dplyr::select(gene_id, symbol) |>
  distinct(gene_id, .keep_all = TRUE) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

OUTLIER_ID <- "0077"

allCounts_filt <- allCounts[, !grepl(OUTLIER_ID, colnames(allCounts))]
metadata_filt  <- metadata |> filter(!grepl(OUTLIER_ID, id))

cat("allCounts:", nrow(allCounts), "genes x", ncol(allCounts), "samples\n")
cat("After outlier removal:", ncol(allCounts_filt), "samples\n")
```

# DESeq2 interaction models: eIF4E delta at each hypoxia timepoint vs normoxia 1hr

LFC = log2(eIF4E_hyp_Xhr / IgG_hyp_Xhr) - log2(eIF4E_nor_1hr / IgG_nor_1hr)
Positive = eIF4E binding increases in hypoxia; negative = decreases.

```{r deseq_4e}
res_4e_list <- list()

for (hyp_hr in c("1", "4", "24")) {
  meta_sub <- metadata_filt |>
    filter((condition == "normoxia" & hours == "1") |
           (condition == "hypoxia"  & hours == hyp_hr),
           ip_type %in% c("eif4e", "igg_mock")) |>
    mutate(ip_type = droplevels(factor(ip_type, levels = c("igg_mock", "eif4e")))) |>
    as.data.frame()

  cts_sub <- do.call(cbind, lapply(meta_sub$id, function(x) {
    allCounts_filt[, grep(x, colnames(allCounts_filt)), drop = FALSE]
  }))
  rownames(meta_sub) <- colnames(cts_sub)

  cat("\n--- eIF4E hypoxia", hyp_hr, "hr vs normoxia 1hr ---\n")
  cat("samples:", ncol(cts_sub), "\n")
  cat("condition levels:", levels(meta_sub$condition), "\n")
  cat("ip_type levels:", levels(meta_sub$ip_type), "\n")

  dds <- DESeqDataSetFromMatrix(cts_sub, meta_sub,
                                 design = ~ condition + ip_type + condition:ip_type)
  dds <- DESeq(dds, quiet = TRUE)

  cat("coefficients:", paste(resultsNames(dds), collapse = ", "), "\n")

  coef_int <- grep("\\.", resultsNames(dds), value = TRUE)
  cat("interaction coef:", coef_int, "\n")

  res <- as.data.frame(lfcShrink(dds, coef = coef_int, type = "apeglm")) |>
    rownames_to_column("gene_id") |>
    left_join(gene_anno, by = "gene_id") |>
    dplyr::rename(delta_eif4e_lfc  = log2FoldChange,
                  delta_eif4e_padj = padj) |>
    mutate(timepoint = paste0("hypoxia_", hyp_hr, "hr_vs_normoxia"))

  tp_label <- paste0(hyp_hr, "hr")
  res_4e_list[[tp_label]] <- res

  write_csv(res,
            here("ripseq", "output",
                 paste0("ripseq_deseq_eif4e_hypoxia_", hyp_hr, "hr_vs_normoxia_delta.csv")))
  cat("saved:", nrow(res), "genes | padj < 0.05:",
      sum(res$delta_eif4e_padj < 0.05, na.rm = TRUE), "\n")
}
```

# DESeq2 interaction models: FLAG-eIF3d delta at each hypoxia timepoint vs normoxia 1hr

LFC = log2(eIF3d_hyp_Xhr / IgG_hyp_Xhr) - log2(eIF3d_nor_1hr / IgG_nor_1hr)

```{r deseq_3d}
res_3d_list <- list()

for (hyp_hr in c("1", "4", "24")) {
  meta_sub <- metadata_filt |>
    filter((condition == "normoxia" & hours == "1") |
           (condition == "hypoxia"  & hours == hyp_hr),
           ip_type %in% c("flag_eif3d", "igg_mock")) |>
    mutate(ip_type = droplevels(factor(ip_type, levels = c("igg_mock", "flag_eif3d")))) |>
    as.data.frame()

  cts_sub <- do.call(cbind, lapply(meta_sub$id, function(x) {
    allCounts_filt[, grep(x, colnames(allCounts_filt)), drop = FALSE]
  }))
  rownames(meta_sub) <- colnames(cts_sub)

  cat("\n--- FLAG-eIF3d hypoxia", hyp_hr, "hr vs normoxia 1hr ---\n")
  cat("samples:", ncol(cts_sub), "\n")
  cat("condition levels:", levels(meta_sub$condition), "\n")
  cat("ip_type levels:", levels(meta_sub$ip_type), "\n")

  dds <- DESeqDataSetFromMatrix(cts_sub, meta_sub,
                                 design = ~ condition + ip_type + condition:ip_type)
  dds <- DESeq(dds, quiet = TRUE)

  cat("coefficients:", paste(resultsNames(dds), collapse = ", "), "\n")

  coef_int <- grep("\\.", resultsNames(dds), value = TRUE)
  cat("interaction coef:", coef_int, "\n")

  res <- as.data.frame(lfcShrink(dds, coef = coef_int, type = "apeglm")) |>
    rownames_to_column("gene_id") |>
    left_join(gene_anno, by = "gene_id") |>
    dplyr::rename(delta_flag3d_lfc  = log2FoldChange,
                  delta_flag3d_padj = padj) |>
    mutate(timepoint = paste0("hypoxia_", hyp_hr, "hr_vs_normoxia"))

  tp_label <- paste0(hyp_hr, "hr")
  res_3d_list[[tp_label]] <- res

  write_csv(res,
            here("ripseq", "output",
                 paste0("ripseq_deseq_flag3d_hypoxia_", hyp_hr, "hr_vs_normoxia_delta.csv")))
  cat("saved:", nrow(res), "genes | padj < 0.05:",
      sum(res$delta_flag3d_padj < 0.05, na.rm = TRUE), "\n")
}
```

# CDF: eIF3d-promoted genes vs eIF4E binding change at hypoxia 1hr

```{r cdf_setup}
group_colors <- c("eIF3d hypoxia promotes" = "#E69F00", "All other genes" = "grey80")

plot_df <- res_4e_list[["1hr"]] |>
  filter(!is.na(delta_eif4e_lfc), !is.na(gene_id_clean)) |>
  mutate(
    group = if_else(gene_id_clean %in% promoted$gene_id_clean,
                    "eIF3d hypoxia promotes", "All other genes"),
    group = factor(group, levels = c("eIF3d hypoxia promotes", "All other genes"))
  )

cat("eIF3d-promoted genes in eIF4E 1hr model:", sum(plot_df$group == "eIF3d hypoxia promotes"), "\n")
cat("All other genes:", sum(plot_df$group == "All other genes"), "\n")
```

```{r cdf_plot}
wt <- wilcox.test(
  delta_eif4e_lfc ~ group,
  data        = plot_df,
  alternative = "less"
)
pval_label <- paste0("Wilcoxon p = ", signif(wt$p.value, 3))

p <- ggplot(plot_df, aes(x = delta_eif4e_lfc, color = group)) +
  stat_ecdf(linewidth = 0.8) +
  scale_color_manual(values = group_colors) +
  annotate("text", x = Inf, y = Inf, label = pval_label,
           hjust = 1.1, vjust = 1.4, size = 3.5) +
  labs(
    x     = "eIF4E binding change, hypoxia vs normoxia (log2FC)",
    y     = "Cumulative proportion",
    color = NULL,
    title = "eIF4E binding change at 1hr: eIF3d-promoted vs all others"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "09_cdf_eif4e_delta_eif3d_promoted.pdf"),
       p, width = 5, height = 4)
p
```

# Volcano: eIF4E binding change (hypoxia vs normoxia), per timepoint

```{r volcano_4e}
vol_4e <- bind_rows(res_4e_list, .id = "tp") |>
  filter(!is.na(delta_eif4e_lfc), !is.na(delta_eif4e_padj)) |>
  mutate(
    tp      = factor(tp, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = pmin(-log10(delta_eif4e_padj), 50),
    sig = if_else(delta_eif4e_padj < 0.05, "sig", "ns"),
    sig = factor(sig, levels = c("sig", "ns"))
  )

vol_colors_4e <- c("sig" = "#0072B2", "ns" = "grey70")

p_vol_4e <- ggplot(vol_4e, aes(x = delta_eif4e_lfc, y = neg_log10_padj, color = sig)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", linewidth = 0.4, color = "grey40") +
  scale_color_manual(values = vol_colors_4e,
                     labels = c("padj < 0.05", "ns")) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    y     = "-log10(padj)",
    color = NULL,
    title = "eIF4E IP enrichment change: hypoxia vs normoxia"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "09_volcano_eif4e_delta.pdf"),
       p_vol_4e, width = 10, height = 4)
p_vol_4e
```

```{r density_4e}
p_dens_4e <- ggplot(vol_4e, aes(x = delta_eif4e_lfc)) +
  geom_density(fill = "#0072B2", color = "#0072B2", alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    y     = "Density",
    title = "eIF4E IP enrichment change: density"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "09_density_eif4e_delta.pdf"),
       p_dens_4e, width = 10, height = 4)
p_dens_4e
```

# Volcano: FLAG-eIF3d binding change (hypoxia vs normoxia), per timepoint

```{r volcano_3d}
vol_3d <- bind_rows(res_3d_list, .id = "tp") |>
  filter(!is.na(delta_flag3d_lfc), !is.na(delta_flag3d_padj)) |>
  mutate(
    tp      = factor(tp, levels = c("1hr", "4hr", "24hr")),
    neg_log10_padj = pmin(-log10(delta_flag3d_padj), 50),
    sig = if_else(delta_flag3d_padj < 0.05, "sig", "ns"),
    sig = factor(sig, levels = c("sig", "ns"))
  )

vol_colors_3d <- c("sig" = "#E69F00", "ns" = "grey70")

p_vol_3d <- ggplot(vol_3d, aes(x = delta_flag3d_lfc, y = neg_log10_padj, color = sig)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", linewidth = 0.4, color = "grey40") +
  scale_color_manual(values = vol_colors_3d,
                     labels = c("padj < 0.05", "ns")) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "FLAG-eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "-log10(padj)",
    color = NULL,
    title = "FLAG-eIF3d IP enrichment change: hypoxia vs normoxia"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "09_volcano_flag3d_delta.pdf"),
       p_vol_3d, width = 10, height = 4)
p_vol_3d
```

```{r density_3d}
p_dens_3d <- ggplot(vol_3d, aes(x = delta_flag3d_lfc)) +
  geom_density(fill = "#E69F00", color = "#E69F00", alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "FLAG-eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "Density",
    title = "FLAG-eIF3d IP enrichment change: density"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "09_density_flag3d_delta.pdf"),
       p_dens_3d, width = 10, height = 4)
p_dens_3d
```
