---
title: "RIP-seq: eIF3d-dependent TE vs eIF4E binding change, per hypoxia timepoint"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
```

Requires the 3 eIF4E delta CSVs from notebook 09.

# Load data

```{r load}
eif4e_paths <- list(
  "1hr"  = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_1hr_vs_normoxia_delta.csv"),
  "4hr"  = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_4hr_vs_normoxia_delta.csv"),
  "24hr" = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_24hr_vs_normoxia_delta.csv")
)

stopifnot("Run notebook 09 first to generate eIF4E delta CSVs" =
  all(file.exists(unlist(eif4e_paths))))

eif4e_all <- imap_dfr(eif4e_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, delta_eif4e_lfc, delta_eif4e_padj) |>
    mutate(tp = tp)
})

te_paths <- list(
  "1hr"  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  "4hr"  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"),
  "24hr" = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")
)

te_all <- imap_dfr(te_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id),
           tp = tp)
})

eif4e_enrich_paths <- list(
  "1hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_eif4e_vs_igg.csv"),
  "4hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_4hr_eif4e_vs_igg.csv"),
  "24hr" = here("ripseq", "output", "ripseq_deseq_hypoxia_24hr_eif4e_vs_igg.csv")
)

eif4e_bound_all <- imap_dfr(eif4e_enrich_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, log2FoldChange, padj) |>
    mutate(tp         = tp,
           eif4e_bound = !is.na(padj) & padj < 0.05 & log2FoldChange > 1)
})

cat("TE genes (all timepoints):", nrow(te_all), "\n")
cat("eIF4E delta genes (all timepoints):", nrow(eif4e_all), "\n")
cat("eIF4E-bound genes per timepoint:\n")
eif4e_bound_all |> dplyr::count(tp, eif4e_bound) |> print()
```

# Join and label

```{r join}
group_colors <- c("eIF3d hypoxia promotes" = "#E69F00", "All other genes" = "grey80")

plot_df <- te_all |>
  inner_join(eif4e_all, by = c("gene_id_clean", "tp")) |>
  left_join(eif4e_bound_all |> dplyr::select(gene_id_clean, tp, eif4e_bound),
            by = c("gene_id_clean", "tp")) |>
  filter(!is.na(te_lfc), !is.na(delta_eif4e_lfc)) |>
  mutate(
    group = if_else(te_lfc > 0.5 & te_padj < 0.05,
                    "eIF3d hypoxia promotes", "All other genes"),
    group = factor(group, levels = c("eIF3d hypoxia promotes", "All other genes")),
    eif4e_bound  = replace_na(eif4e_bound, FALSE),
    eif4e_group  = if_else(eif4e_bound, "eIF4E bound", "All other genes"),
    eif4e_group  = factor(eif4e_group, levels = c("eIF4E bound", "All other genes")),
    tp    = factor(tp, levels = c("1hr", "4hr", "24hr"))
  )

plot_df |>
  dplyr::count(tp, group) |>
  print()
```

# Scatter plot: eIF3d-dependent TE vs eIF4E binding change, faceted by timepoint

```{r scatter}
p_scatter <- ggplot(plot_df, aes(x = te_lfc, y = delta_eif4e_lfc)) +
  geom_point(aes(color = group), alpha = 0.25, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, alpha = 0.15) +
  scale_color_manual(values = group_colors) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "eIF3d-dependent TE (log2FC, hypoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    color = NULL,
    title = "eIF3d TE activity vs eIF4E binding change"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")+
  xlim(-5,5)+
  ylim(-10,10)

ggsave(here("ripseq", "plots", "10_scatter_te_vs_eif4e_per_timepoint.pdf"),
       p_scatter, width = 10, height = 4)
p_scatter
```

# CDF: eIF4E binding change by eIF3d-TE group, faceted by timepoint

```{r cdf}
wt_df <- plot_df |>
  group_by(tp) |>
  summarise(
    p     = wilcox.test(delta_eif4e_lfc ~ group, alternative = "less")$p.value,
    label = paste0("p = ", signif(p, 3)),
    .groups = "drop"
  )

p_cdf <- ggplot(plot_df, aes(x = delta_eif4e_lfc, color = group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data  = wt_df,
            aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(values = group_colors) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    y     = "Cumulative proportion",
    color = NULL,
    title = "eIF4E binding change: eIF3d-promoted vs all others"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "10_cdf_eif4e_delta_by_te_group_per_timepoint.pdf"),
       p_cdf, width = 10, height = 4)
p_cdf
```

# CDF: eIF3d-dependent TE by eIF4E-bound vs all others, faceted by timepoint

```{r cdf_reverse}
eif4e_colors <- c("eIF4E bound" = "#0072B2", "All other genes" = "grey80")

wt_4e_df <- plot_df |>
  group_by(tp) |>
  summarise(
    p     = wilcox.test(te_lfc ~ eif4e_group, alternative = "less")$p.value,
    label = paste0("p = ", signif(p, 3)),
    .groups = "drop"
  )

p_cdf_rev <- ggplot(plot_df, aes(x = te_lfc, color = eif4e_group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_4e_df,
            aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(values = eif4e_colors) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "eIF3d-dependent TE (log2FC, hypoxia)",
    y     = "Cumulative proportion",
    color = NULL,
    title = "eIF3d TE: eIF4E-bound vs all others"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom")+
  xlim(-2,2)

ggsave(here("ripseq", "plots", "10_cdf_te_by_eif4e_bound_per_timepoint.pdf"),
       p_cdf_rev, width = 10, height = 4)
p_cdf_rev
```

# Boxplots: 3d vs 4e IP binding change by TE gene group

```{r load_ip_delta}
hyp_genes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> dplyr::select(gene_id_clean = ensembl_gene)

nor_genes <- read_csv(
  here("output", "genesets", "normoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> dplyr::select(gene_id_clean = ensembl_gene)

gene_group_df <- bind_rows(
  hyp_genes |> mutate(in_hyp = TRUE),
  nor_genes |> mutate(in_nor = TRUE)
) |>
  group_by(gene_id_clean) |>
  summarise(in_hyp = any(!is.na(in_hyp)), in_nor = any(!is.na(in_nor)), .groups = "drop") |>
  mutate(
    gene_group = case_when(
      in_hyp & in_nor  ~ "Both",
      in_hyp & !in_nor ~ "Hypoxia only",
      !in_hyp & in_nor ~ "Normoxia only",
      TRUE             ~ "Neither"
    ),
    gene_group = factor(gene_group, levels = c("Neither", "Normoxia only", "Both", "Hypoxia only"))
  )

flag3d_paths <- list(
  "1hr"  = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_1hr_vs_normoxia_delta.csv"),
  "4hr"  = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_4hr_vs_normoxia_delta.csv"),
  "24hr" = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_24hr_vs_normoxia_delta.csv")
)

flag3d_delta <- imap_dfr(flag3d_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, delta_lfc = delta_flag3d_lfc) |>
    mutate(tp = tp, ip_type = "eIF3d IP")
})

eif4e_delta <- eif4e_all |>
  dplyr::select(gene_id_clean, delta_lfc = delta_eif4e_lfc, tp) |>
  mutate(ip_type = "eIF4E IP")

delta_df <- bind_rows(flag3d_delta, eif4e_delta) |>
  left_join(gene_group_df, by = "gene_id_clean") |>
  mutate(
    gene_group = replace_na(as.character(gene_group), "Neither") |>
      factor(levels = c("Neither", "Normoxia only", "Both", "Hypoxia only")),
    tp      = factor(tp,      levels = c("1hr", "4hr", "24hr")),
    ip_type = factor(ip_type, levels = c("eIF4E IP", "eIF3d IP"))
  )

cat("=== IgG denominator: n per gene group ===\n")
delta_df |>
  distinct(gene_id_clean, gene_group) |>
  dplyr::count(gene_group) |>
  print()

cat("\n=== IgG denominator: n per gene group x timepoint ===\n")
delta_df |>
  distinct(gene_id_clean, tp, gene_group) |>
  dplyr::count(gene_group, tp) |>
  tidyr::pivot_wider(names_from = tp, values_from = n) |>
  print()
```

```{r boxplot_ip_delta_by_group}
group_colors_4 <- c(
  "Neither"       = "grey80",
  "Normoxia only" = "#0072B2",
  "Both"          = "#CC79A7",
  "Hypoxia only"  = "#E69F00"
)

p_box <- ggplot(delta_df, aes(x = tp, y = delta_lfc, fill = gene_group)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  geom_violin(linewidth = 0.6) +
  stat_summary(fun = median, geom = "point", size = 1.5, color = "black",
               position = position_dodge(width = 0.9)) +
  coord_cartesian(ylim = c(-4, 3)) +
  scale_fill_manual(values = group_colors_4) +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(
    x    = "Hypoxia timepoint",
    y    = "IP binding change (log2FC, hypoxia vs normoxia)",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("ripseq", "plots", "10_violin_ip_delta_by_te_group.pdf"),
       p_box, width = 10, height = 5)
p_box
```

```{r cdf_ip_delta_by_group}
p_cdf_ip <- ggplot(delta_df, aes(x = delta_lfc, color = gene_group)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  stat_ecdf(linewidth = 0.8) +
  coord_cartesian(xlim = c(-4, 3)) +
  scale_color_manual(values = group_colors_4) +
  facet_grid(ip_type ~ tp) +
  labs(
    x     = "IP binding change (log2FC, hypoxia vs normoxia)",
    y     = "Cumulative proportion",
    color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "10_cdf_ip_delta_by_te_group.pdf"),
       p_cdf_ip, width = 10, height = 6)
p_cdf_ip
```

# IP/RNA delta: equivalent plots using RNA as denominator

```{r load_ip_rna_delta}
rna_enrich_path <- here("ripseq", "output", "18_ip_over_rna_enrichment.csv")
stopifnot("Run notebook 18 first to generate IP/RNA enrichment CSV" =
  file.exists(rna_enrich_path))

rna_enrich <- read_csv(rna_enrich_path, show_col_types = FALSE)

cat("=== IP/RNA enrichment table ===\n")
cat("Rows:", nrow(rna_enrich), "\n")
cat("ip_types:", paste(unique(rna_enrich$ip_type), collapse = ", "), "\n")
cat("timepoints:", paste(sort(unique(rna_enrich$timepoint)), collapse = ", "), "\n\n")

# Genes per ip_type x timepoint
cat("=== Genes per ip_type x timepoint ===\n")
rna_enrich |> dplyr::count(ip_type, timepoint) |> print()

# Separate normoxia reference from hypoxia timepoints
rna_nor <- rna_enrich |>
  filter(timepoint == "normoxia_1hr") |>
  dplyr::select(gene_id_clean, ip_type, log2_ip_over_rna_nor = log2_ip_over_rna)

rna_hyp <- rna_enrich |>
  filter(timepoint != "normoxia_1hr") |>
  dplyr::select(gene_id_clean, ip_type, timepoint, log2_ip_over_rna_hyp = log2_ip_over_rna)

cat("\n=== Comparisons being computed ===\n")
cat("delta = log2(IP/RNA at hypoxia_Xhr) - log2(IP/RNA at normoxia_1hr)\n")
rna_hyp |>
  distinct(ip_type, timepoint) |>
  mutate(comparison = paste0(timepoint, " minus normoxia_1hr")) |>
  print()

rna_delta_raw <- rna_hyp |>
  inner_join(rna_nor, by = c("gene_id_clean", "ip_type")) |>
  mutate(delta_rna_lfc = log2_ip_over_rna_hyp - log2_ip_over_rna_nor)

# Genes lost due to missing normoxia value
n_hyp   <- nrow(rna_hyp)
n_delta <- nrow(rna_delta_raw)
cat("\n=== Join check ===\n")
cat("Hypoxia rows before join:", n_hyp, "\n")
cat("Rows after joining normoxia reference:", n_delta, "\n")
cat("Dropped (missing normoxia):", n_hyp - n_delta, "\n\n")

cat("=== Delta summary per ip_type x timepoint ===\n")
rna_delta_raw |>
  group_by(ip_type, timepoint) |>
  summarise(
    n      = n(),
    median = round(median(delta_rna_lfc, na.rm = TRUE), 3),
    mean   = round(mean(delta_rna_lfc,   na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  print()

# Compare gene coverage to IgG delta
cat("\n=== Gene coverage vs IgG delta ===\n")
cat("IgG delta genes (3d+4e, all timepoints):", nrow(delta_df), "\n")
cat("RNA delta genes (3d+4e, all timepoints):", nrow(rna_delta_raw), "\n")

# Parse timepoint string to tp label (e.g. "hypoxia_1hr" -> "1hr")
rna_delta_df <- rna_delta_raw |>
  mutate(
    tp = sub("hypoxia_", "", timepoint),
    ip_type = case_when(
      ip_type == "eif4e"      ~ "eIF4E IP",
      ip_type == "flag_eif3d" ~ "eIF3d IP"
    )
  ) |>
  left_join(gene_group_df, by = "gene_id_clean") |>
  mutate(
    gene_group = replace_na(as.character(gene_group), "Neither") |>
      factor(levels = c("Neither", "Normoxia only", "Both", "Hypoxia only")),
    tp      = factor(tp,      levels = c("1hr", "4hr", "24hr")),
    ip_type = factor(ip_type, levels = c("eIF4E IP", "eIF3d IP"))
  )

cat("\n=== RNA denominator: n per gene group ===\n")
rna_delta_df |>
  distinct(gene_id_clean, gene_group) |>
  dplyr::count(gene_group) |>
  print()
```

```{r violin_ip_rna_delta_by_group}
p_violin_rna <- ggplot(rna_delta_df, aes(x = tp, y = delta_rna_lfc, fill = gene_group)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  geom_violin(linewidth = 0.6) +
  stat_summary(fun = median, geom = "point", size = 1.5, color = "black",
               position = position_dodge(width = 0.9)) +
  coord_cartesian(ylim = c(-4, 3)) +
  scale_fill_manual(values = group_colors_4) +
  facet_wrap(~ ip_type, nrow = 1) +
  labs(
    x    = "Hypoxia timepoint",
    y    = "IP/RNA binding change (log2FC, hypoxia vs normoxia)",
    fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(here("ripseq", "plots", "10_violin_ip_rna_delta_by_te_group.pdf"),
       p_violin_rna, width = 10, height = 5)
p_violin_rna
```

```{r cdf_ip_rna_delta_by_group}
p_cdf_rna <- ggplot(rna_delta_df, aes(x = delta_rna_lfc, color = gene_group)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey40") +
  stat_ecdf(linewidth = 0.8) +
  coord_cartesian(xlim = c(-4, 3)) +
  scale_color_manual(values = group_colors_4) +
  facet_grid(ip_type ~ tp) +
  labs(
    x     = "IP/RNA binding change (log2FC, hypoxia vs normoxia)",
    y     = "Cumulative proportion",
    color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "10_cdf_ip_rna_delta_by_te_group.pdf"),
       p_cdf_rna, width = 10, height = 6)
p_cdf_rna
```
