---
title: "RIP-seq: eIF3d vs eIF4E binding trade-off in 1hr hypoxia"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
```

Hypothesis: specific mRNAs switch from cap-dependent (eIF4E) to eIF3d-dependent
translation in 1hr hypoxia, manifesting as a negative correlation between the
hypoxia-induced change in eIF3d RIP binding and the change in eIF4E RIP binding.

Requires delta CSVs from notebook 09.

```{r constants}
RIP_DELTA_PADJ <- 0.05
TE_LFC_CUT     <- 0.5
TE_PADJ_CUT    <- 0.05

group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "All other genes"        = "grey80"
)
```

# Load data

```{r load}
delta_3d_path <- here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_1hr_vs_normoxia_delta.csv")
delta_4e_path <- here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_1hr_vs_normoxia_delta.csv")

stopifnot("Run notebook 09 first to generate delta CSVs" =
  all(file.exists(delta_3d_path, delta_4e_path)))

delta_3d <- read_csv(delta_3d_path, show_col_types = FALSE) |>
  dplyr::select(gene_id_clean, delta_flag3d_lfc, delta_flag3d_padj)

delta_4e <- read_csv(delta_4e_path, show_col_types = FALSE) |>
  dplyr::select(gene_id_clean, delta_eif4e_lfc, delta_eif4e_padj)

te_hyp <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, symbol, te_lfc, te_padj)

te_sictrl_hyp <- read_csv(
  here("output", "translation_categories_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, sictrl_te_lfc = te_lfc, sictrl_te_padj = te_padj)

cat("delta_3d genes:", nrow(delta_3d), "\n")
cat("delta_4e genes:", nrow(delta_4e), "\n")
cat("TE genes (si3d vs sictrl):", nrow(te_hyp), "\n")
cat("sictrl hypoxia-promoted TE (>0.5, padj<0.05):",
    sum(te_sictrl_hyp$sictrl_te_lfc > 0.5 & te_sictrl_hyp$sictrl_te_padj < 0.05, na.rm = TRUE), "\n")
```

# Join and label

```{r join}
plot_df <- inner_join(delta_3d, delta_4e, by = "gene_id_clean") |>
  left_join(te_hyp, by = "gene_id_clean") |>
  left_join(te_sictrl_hyp, by = "gene_id_clean") |>
  filter(!is.na(delta_flag3d_lfc), !is.na(delta_eif4e_lfc)) |>
  mutate(
    group = if_else(
      !is.na(te_lfc) & !is.na(te_padj) & te_lfc > TE_LFC_CUT & te_padj < TE_PADJ_CUT,
      "eIF3d hypoxia promotes", "All other genes"
    ),
    group = factor(group, levels = c("eIF3d hypoxia promotes", "All other genes")),
    sictrl_hyp_group = if_else(
      !is.na(sictrl_te_lfc) & !is.na(sictrl_te_padj) &
        sictrl_te_lfc > TE_LFC_CUT & sictrl_te_padj < TE_PADJ_CUT,
      "sictrl hypoxia promotes", "All other genes"
    ),
    sictrl_hyp_group = factor(sictrl_hyp_group,
                               levels = c("sictrl hypoxia promotes", "All other genes")),
    eif3d_gained = !is.na(delta_flag3d_padj) & delta_flag3d_padj < RIP_DELTA_PADJ &
      delta_flag3d_lfc > 0,
    eif3d_gained_group = if_else(eif3d_gained,
                                  "eIF3d gained in hypoxia", "All other genes"),
    eif3d_gained_group = factor(eif3d_gained_group,
                                 levels = c("eIF3d gained in hypoxia", "All other genes")),
    quadrant = case_when(
      delta_flag3d_lfc >= 0 & delta_eif4e_lfc <  0 ~ "3d up / 4e down",
      delta_flag3d_lfc >= 0 & delta_eif4e_lfc >= 0 ~ "3d up / 4e up",
      delta_flag3d_lfc <  0 & delta_eif4e_lfc <  0 ~ "3d down / 4e down",
      delta_flag3d_lfc <  0 & delta_eif4e_lfc >= 0 ~ "3d down / 4e up"
    ),
    quadrant = factor(quadrant,
                      levels = c("3d up / 4e down", "3d up / 4e up",
                                 "3d down / 4e down", "3d down / 4e up"))
  )

cat("Genes in joined set:", nrow(plot_df), "\n")
cat("eIF3d-promoted TE genes:", sum(plot_df$group == "eIF3d hypoxia promotes"), "\n")
cat("sictrl hypoxia-promoted TE genes:", sum(plot_df$sictrl_hyp_group == "sictrl hypoxia promotes"), "\n")
cat("eIF3d-gained in hypoxia (padj<0.05, LFC>0):", sum(plot_df$eif3d_gained, na.rm = TRUE), "\n")
cat("\nQuadrant counts (all genes):\n")
plot_df |> dplyr::count(quadrant) |> print()
cat("\nQuadrant counts (eIF3d-promoted TE genes):\n")
plot_df |> filter(group == "eIF3d hypoxia promotes") |> dplyr::count(quadrant) |> print()
```

```{r correlations}
cor_all <- cor.test(plot_df$delta_flag3d_lfc, plot_df$delta_eif4e_lfc,
                    method = "spearman", exact = FALSE)
promoted_df <- filter(plot_df, group == "eIF3d hypoxia promotes")
cor_prom <- cor.test(promoted_df$delta_flag3d_lfc, promoted_df$delta_eif4e_lfc,
                     method = "spearman", exact = FALSE)

cat(sprintf("Spearman rho (all genes): %.3f  p = %.2e\n", cor_all$estimate, cor_all$p.value))
cat(sprintf("Spearman rho (eIF3d-promoted): %.3f  p = %.2e\n",
            cor_prom$estimate, cor_prom$p.value))
```

# Scatter: delta eIF3d vs delta eIF4E

```{r scatter}
rho_label <- sprintf("rho = %.2f, p = %.2e", cor_all$estimate, cor_all$p.value)

p_scatter <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc)) +
  geom_point(data = filter(plot_df, group == "All other genes"),
             color = "grey80", alpha = 0.25, size = 0.6) +
  geom_point(data = filter(plot_df, group == "eIF3d hypoxia promotes"),
             color = "#E69F00", alpha = 0.5, size = 0.8) +
  geom_smooth(data = plot_df, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 2.8,
           label = paste0("eIF3d promotes: n = ",
                          sum(plot_df$group == "eIF3d hypoxia promotes")),
           color = "#E69F00", size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "eIF3d vs eIF4E binding trade-off, 1hr hypoxia"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e.pdf"),
       p_scatter, width = 5, height = 5)
p_scatter
```

# Scatter: sictrl hypoxia-promoted TE

```{r scatter_sictrl}
n_sictrl <- sum(plot_df$sictrl_hyp_group == "sictrl hypoxia promotes")

p_scatter_sictrl <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc)) +
  geom_point(data = filter(plot_df, sictrl_hyp_group == "All other genes"),
             color = "grey80", alpha = 0.25, size = 0.6) +
  geom_point(data = filter(plot_df, sictrl_hyp_group == "sictrl hypoxia promotes"),
             color = "#CC0000", alpha = 0.6, size = 0.9) +
  geom_smooth(data = plot_df, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 2.8,
           label = paste0("sictrl promotes: n = ", n_sictrl),
           color = "#CC0000", size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "Hypoxia-induced TE (sictrl) vs binding change, 1hr hypoxia"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_sictrl_hyp.pdf"),
       p_scatter_sictrl, width = 5, height = 5)
p_scatter_sictrl
```

# CDF: delta eIF4E by eIF3d-gained vs all

```{r cdf}
gained_colors <- c("eIF3d gained in hypoxia" = "#0072B2", "All other genes" = "grey80")

wt_gained <- wilcox.test(
  delta_eif4e_lfc ~ eif3d_gained_group,
  data      = plot_df,
  alternative = "less"
)
wt_label <- paste0("p = ", signif(wt_gained$p.value, 3))

p_cdf <- ggplot(plot_df, aes(x = delta_eif4e_lfc, color = eif3d_gained_group)) +
  stat_ecdf(linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = wt_label, size = 3.5) +
  scale_color_manual(values = gained_colors) +
  labs(
    x     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    y     = "Cumulative proportion",
    color = NULL,
    title = "eIF4E binding change: eIF3d-gained vs all others"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "12_cdf_delta4e_by_eif3d_gained.pdf"),
       p_cdf, width = 5, height = 4)
p_cdf
```

# Quadrant bar chart with Fisher enrichment

```{r quadrant}
quad_all <- plot_df |>
  dplyr::count(quadrant, name = "n_all")

quad_prom <- plot_df |>
  filter(group == "eIF3d hypoxia promotes") |>
  dplyr::count(quadrant, name = "n_promoted")

quad_df <- left_join(quad_all, quad_prom, by = "quadrant") |>
  replace_na(list(n_promoted = 0L)) |>
  pivot_longer(c(n_all, n_promoted), names_to = "set", values_to = "n") |>
  mutate(set = if_else(set == "n_all", "All genes", "eIF3d hypoxia promotes"),
         set = factor(set, levels = c("All genes", "eIF3d hypoxia promotes")))

p_quad <- ggplot(quad_df, aes(x = quadrant, y = n, fill = set)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("All genes" = "grey70",
                                "eIF3d hypoxia promotes" = "#E69F00")) +
  labs(
    x    = NULL,
    y    = "Number of genes",
    fill = NULL,
    title = "Gene counts by binding change quadrant"
  ) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")

# Fisher test: enrichment of eIF3d-promoted genes in "3d up / 4e down" vs all other quadrants
n_tradoff_prom  <- quad_prom |> filter(quadrant == "3d up / 4e down") |> pull(n_promoted)
n_tradoff_all   <- quad_all  |> filter(quadrant == "3d up / 4e down") |> pull(n_all)
n_other_prom    <- sum(quad_prom$n_promoted) - n_tradoff_prom
n_other_all     <- sum(quad_all$n_all) - n_tradoff_all

fisher_mat <- matrix(
  c(n_tradoff_prom, n_tradoff_all - n_tradoff_prom,
    n_other_prom,   n_other_all   - n_other_prom),
  nrow = 2,
  dimnames = list(c("promoted", "other"), c("trade-off quadrant", "other quadrants"))
)
fisher_res <- fisher.test(fisher_mat, alternative = "greater")
cat(sprintf(
  "Fisher test (eIF3d-promoted enrichment in '3d up / 4e down'): OR = %.2f, p = %.3e\n",
  fisher_res$estimate, fisher_res$p.value
))

ggsave(here("ripseq", "plots", "12_quadrant_counts.pdf"),
       p_quad, width = 5, height = 4)
p_quad
```
