---
title: "RIP-seq: eIF3d vs eIF4E binding trade-off in 1hr hypoxia"
author: "Kate"
date: "2026-05"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
```

Hypothesis: specific mRNAs switch from cap-dependent (eIF4E) to eIF3d-dependent
translation in 1hr hypoxia, manifesting as a negative correlation between the
hypoxia-induced change in eIF3d RIP binding and the change in eIF4E RIP binding.

Requires delta CSVs from notebook 09.

```{r constants}
RIP_DELTA_PADJ <- 0.05
TE_LFC_CUT     <- 0.5
TE_PADJ_CUT    <- 0.05

group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "All other genes"        = "grey80"
)
```

# Load data

```{r load}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
                     col_names = c("gene_id", "symbol", "biotype", "species", "version"),
                     show_col_types = FALSE) |>
  filter(species == "human") |>
  dplyr::select(gene_id_clean = gene_id, symbol) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id_clean)) |>
  distinct(gene_id_clean, .keep_all = TRUE)

delta_3d_path <- here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_1hr_vs_normoxia_delta.csv")
delta_4e_path <- here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_1hr_vs_normoxia_delta.csv")

stopifnot("Run notebook 09 first to generate delta CSVs" =
  all(file.exists(delta_3d_path, delta_4e_path)))

delta_3d <- read_csv(delta_3d_path, show_col_types = FALSE) |>
  dplyr::select(gene_id_clean, delta_flag3d_lfc, delta_flag3d_padj)

delta_4e <- read_csv(delta_4e_path, show_col_types = FALSE) |>
  dplyr::select(gene_id_clean, delta_eif4e_lfc, delta_eif4e_padj)

te_hyp <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, symbol, te_lfc, te_padj)

te_sictrl_hyp <- read_csv(
  here("output", "translation_categories_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, sictrl_te_lfc = te_lfc, sictrl_te_padj = te_padj)

cat("delta_3d genes:", nrow(delta_3d), "\n")
cat("delta_4e genes:", nrow(delta_4e), "\n")
cat("TE genes (si3d vs sictrl):", nrow(te_hyp), "\n")
cat("sictrl hypoxia-promoted TE (>0.5, padj<0.05):",
    sum(te_sictrl_hyp$sictrl_te_lfc > 0.5 & te_sictrl_hyp$sictrl_te_padj < 0.05, na.rm = TRUE), "\n")
```

# Join and label

```{r join}
plot_df <- inner_join(delta_3d, delta_4e, by = "gene_id_clean") |>
  left_join(te_hyp, by = "gene_id_clean") |>
  left_join(te_sictrl_hyp, by = "gene_id_clean") |>
  left_join(gene_anno |> dplyr::rename(symbol_anno = symbol), by = "gene_id_clean") |>
  mutate(symbol = coalesce(symbol, symbol_anno)) |>
  dplyr::select(-symbol_anno) |>
  filter(!is.na(delta_flag3d_lfc), !is.na(delta_eif4e_lfc)) |>
  mutate(
    group = if_else(
      !is.na(te_lfc) & !is.na(te_padj) & te_lfc > TE_LFC_CUT & te_padj < TE_PADJ_CUT,
      "eIF3d hypoxia promotes", "All other genes"
    ),
    group = factor(group, levels = c("eIF3d hypoxia promotes", "All other genes")),
    sictrl_hyp_group = if_else(
      !is.na(sictrl_te_lfc) & !is.na(sictrl_te_padj) &
        sictrl_te_lfc > TE_LFC_CUT & sictrl_te_padj < TE_PADJ_CUT,
      "sictrl hypoxia promotes", "All other genes"
    ),
    sictrl_hyp_group = factor(sictrl_hyp_group,
                               levels = c("sictrl hypoxia promotes", "All other genes")),
    eif3d_gained = !is.na(delta_flag3d_padj) & delta_flag3d_padj < RIP_DELTA_PADJ &
      delta_flag3d_lfc > 0,
    eif3d_gained_group = if_else(eif3d_gained,
                                  "eIF3d gained in hypoxia", "All other genes"),
    eif3d_gained_group = factor(eif3d_gained_group,
                                 levels = c("eIF3d gained in hypoxia", "All other genes")),
    quadrant = case_when(
      delta_flag3d_lfc >= 0 & delta_eif4e_lfc <  0 ~ "3d up / 4e down",
      delta_flag3d_lfc >= 0 & delta_eif4e_lfc >= 0 ~ "3d up / 4e up",
      delta_flag3d_lfc <  0 & delta_eif4e_lfc <  0 ~ "3d down / 4e down",
      delta_flag3d_lfc <  0 & delta_eif4e_lfc >= 0 ~ "3d down / 4e up"
    ),
    quadrant = factor(quadrant,
                      levels = c("3d up / 4e down", "3d up / 4e up",
                                 "3d down / 4e down", "3d down / 4e up"))
  )

cat("Genes in joined set:", nrow(plot_df), "\n")
cat("eIF3d-promoted TE genes:", sum(plot_df$group == "eIF3d hypoxia promotes"), "\n")
cat("sictrl hypoxia-promoted TE genes:", sum(plot_df$sictrl_hyp_group == "sictrl hypoxia promotes"), "\n")
cat("eIF3d-gained in hypoxia (padj<0.05, LFC>0):", sum(plot_df$eif3d_gained, na.rm = TRUE), "\n")
cat("\nQuadrant counts (all genes):\n")
plot_df |> dplyr::count(quadrant) |> print()
cat("\nQuadrant counts (eIF3d-promoted TE genes):\n")
plot_df |> filter(group == "eIF3d hypoxia promotes") |> dplyr::count(quadrant) |> print()
```

```{r correlations}
cor_all <- cor.test(plot_df$delta_flag3d_lfc, plot_df$delta_eif4e_lfc,
                    method = "spearman", exact = FALSE)
promoted_df <- filter(plot_df, group == "eIF3d hypoxia promotes")
cor_prom <- cor.test(promoted_df$delta_flag3d_lfc, promoted_df$delta_eif4e_lfc,
                     method = "spearman", exact = FALSE)

cat(sprintf("Spearman rho (all genes): %.3f  p = %.2e\n", cor_all$estimate, cor_all$p.value))
cat(sprintf("Spearman rho (eIF3d-promoted): %.3f  p = %.2e\n",
            cor_prom$estimate, cor_prom$p.value))
```

# Scatter: delta eIF3d vs delta eIF4E

```{r scatter}
rho_label <- sprintf("rho = %.2f, p = %.2e", cor_all$estimate, cor_all$p.value)

p_scatter <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc)) +
  geom_point(data = filter(plot_df, group == "All other genes"),
             color = "grey80", alpha = 0.25, size = 0.6) +
  geom_point(data = filter(plot_df, group == "eIF3d hypoxia promotes"),
             color = "#E69F00", alpha = 0.5, size = 0.8) +
  geom_smooth(data = plot_df, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 2.8,
           label = paste0("eIF3d promotes: n = ",
                          sum(plot_df$group == "eIF3d hypoxia promotes")),
           color = "#E69F00", size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "eIF3d vs eIF4E binding trade-off, 1hr hypoxia"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e.pdf"),
       p_scatter, width = 5, height = 5)
p_scatter
```

# Scatter: gradient colored by sictrl TE (1hr hypoxia)

```{r scatter_te_gradient}
te_cut <- 0.5
df_te_grey <- filter(plot_df, is.na(sictrl_te_lfc) | abs(sictrl_te_lfc) < te_cut)
df_te_blue <- filter(plot_df, !is.na(sictrl_te_lfc) & sictrl_te_lfc <= -te_cut)
df_te_red  <- filter(plot_df, !is.na(sictrl_te_lfc) & sictrl_te_lfc >= te_cut)

p_scatter_te_gradient <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc,
                                               color = sictrl_te_lfc)) +
  geom_point(data = df_te_grey, alpha = 0.25, size = 0.6) +
  geom_point(data = df_te_blue, alpha = 0.70, size = 0.8) +
  geom_point(data = df_te_red,  alpha = 0.80, size = 0.9) +
  geom_smooth(data = plot_df, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  scale_color_gradient2(
    low      = "#2C7BB6",
    mid      = "grey85",
    high     = "#D7191C",
    midpoint = 0,
    limits   = c(-2, 2),
    oob      = scales::squish,
    na.value = "grey85",
    name     = "sictrl delta TE\n(1hr hypoxia)"
  ) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "eIF3d vs eIF4E binding, colored by sictrl TE (1hr hypoxia)"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_te_gradient.pdf"),
       p_scatter_te_gradient, width = 5.5, height = 5)
p_scatter_te_gradient
```

# Scatter: gradient colored by sictrl TE rank (1hr hypoxia)

```{r scatter_te_rank}
plot_df_rank <- plot_df |>
  mutate(te_prank = percent_rank(sictrl_te_lfc)) |>
  arrange(te_prank)

p_scatter_te_rank <- ggplot(plot_df_rank,
                             aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc,
                                 color = te_prank)) +
  geom_point(alpha = 0.5, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  scale_color_gradient2(
    low      = "#2C7BB6",
    mid      = "grey85",
    high     = "#D7191C",
    midpoint = 0.5,
    limits   = c(0, 1),
    na.value = "grey85",
    name     = "sictrl TE rank\n(1hr hypoxia)",
    labels   = c("lowest", "", "median", "", "highest"),
    breaks   = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "eIF3d vs eIF4E binding, colored by sictrl TE rank (1hr hypoxia)"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_te_rank.pdf"),
       p_scatter_te_rank, width = 5.5, height = 5)
p_scatter_te_rank
```

# identify dots

what are the dots

```{r identify_4e_drop_3d_neutral}
trade_off_genes <- plot_df |>
  filter(
    abs(delta_flag3d_lfc) < 1,
    delta_eif4e_lfc < -2
  )

cat("Genes with eIF4E drop (< -2) but neutral eIF3d change (|lfc| < 1):",
    nrow(trade_off_genes), "\n")
print(trade_off_genes |> dplyr::select(gene_id_clean, symbol, delta_flag3d_lfc, delta_eif4e_lfc) |>
        arrange(delta_eif4e_lfc))

# top quartile by sictrl TE LFC within trade-off set
te_q75 <- quantile(trade_off_genes$sictrl_te_lfc, 0.75, na.rm = TRUE)
cat("\nTop quartile sictrl TE LFC cutoff:", round(te_q75, 3), "\n")

top_te_trade_off <- trade_off_genes |>
  filter(!is.na(sictrl_te_lfc), sictrl_te_lfc >= te_q75)

cat("Trade-off genes in top sictrl TE quartile:", nrow(top_te_trade_off), "\n")
print(top_te_trade_off |>
        dplyr::select(gene_id_clean, symbol, sictrl_te_lfc, delta_flag3d_lfc, delta_eif4e_lfc) |>
        arrange(desc(sictrl_te_lfc)))

# top quartile by eIF3d-promoted TE LFC (si3d vs sictrl hypoxia) within trade-off set
eif3d_q75 <- quantile(trade_off_genes$te_lfc, 0.75, na.rm = TRUE)
cat("\nTop quartile eIF3d-promoted TE LFC cutoff:", round(eif3d_q75, 3), "\n")

top_eif3d_trade_off <- trade_off_genes |>
  filter(!is.na(te_lfc), te_lfc >= eif3d_q75)

cat("Trade-off genes in top eIF3d-promoted TE quartile:", nrow(top_eif3d_trade_off), "\n")
print(top_eif3d_trade_off |>
        dplyr::select(gene_id_clean, symbol, te_lfc, delta_flag3d_lfc, delta_eif4e_lfc) |>
        arrange(desc(te_lfc)))

# overlap: top quartile in both sictrl TE and eIF3d-promoted TE
overlap_trade_off <- inner_join(
  top_te_trade_off    |> dplyr::select(gene_id_clean),
  top_eif3d_trade_off |> dplyr::select(gene_id_clean),
  by = "gene_id_clean"
) |>
  left_join(trade_off_genes, by = "gene_id_clean")

cat("\nOverlap (top quartile in both sictrl TE and eIF3d-promoted TE):", nrow(overlap_trade_off), "\n")
print(overlap_trade_off |>
        dplyr::select(gene_id_clean, symbol, sictrl_te_lfc, te_lfc, delta_flag3d_lfc, delta_eif4e_lfc) |>
        arrange(desc(te_lfc)))

      overlap_trade_off |> pull(symbol)
```

















# Scatter: sictrl hypoxia-promoted TE

```{r scatter_sictrl}
n_sictrl <- sum(plot_df$sictrl_hyp_group == "sictrl hypoxia promotes")

p_scatter_sictrl <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc)) +
  geom_point(data = filter(plot_df, sictrl_hyp_group == "All other genes"),
             color = "grey80", alpha = 0.25, size = 0.6) +
  geom_point(data = filter(plot_df, sictrl_hyp_group == "sictrl hypoxia promotes"),
             color = "#CC0000", alpha = 0.6, size = 0.9) +
  geom_smooth(data = plot_df, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = rho_label, size = 3.5) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 2.8,
           label = paste0("sictrl promotes: n = ", n_sictrl),
           color = "#CC0000", size = 3.5) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    title = "Hypoxia-induced TE (sictrl) vs binding change, 1hr hypoxia"
  ) +
  theme_classic(base_size = 12)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_sictrl_hyp.pdf"),
       p_scatter_sictrl, width = 5, height = 5)
p_scatter_sictrl
```












# CDF: delta eIF4E by eIF3d-gained vs all

```{r cdf}
gained_colors <- c("eIF3d gained in hypoxia" = "#0072B2", "All other genes" = "grey80")

wt_gained <- wilcox.test(
  delta_eif4e_lfc ~ eif3d_gained_group,
  data      = plot_df,
  alternative = "less"
)
wt_label <- paste0("p = ", signif(wt_gained$p.value, 3))

p_cdf <- ggplot(plot_df, aes(x = delta_eif4e_lfc, color = eif3d_gained_group)) +
  stat_ecdf(linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = wt_label, size = 3.5) +
  scale_color_manual(values = gained_colors) +
  labs(
    x     = "eIF4E binding change (log2FC, hypoxia vs normoxia)",
    y     = "Cumulative proportion",
    color = NULL,
    title = "eIF4E binding change: eIF3d-gained vs all others"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(here("ripseq", "plots", "12_cdf_delta4e_by_eif3d_gained.pdf"),
       p_cdf, width = 5, height = 4)
p_cdf
```

# Quadrant bar chart with Fisher enrichment

```{r quadrant}
quad_all <- plot_df |>
  dplyr::count(quadrant, name = "n_all")

quad_prom <- plot_df |>
  filter(group == "eIF3d hypoxia promotes") |>
  dplyr::count(quadrant, name = "n_promoted")

quad_df <- left_join(quad_all, quad_prom, by = "quadrant") |>
  replace_na(list(n_promoted = 0L)) |>
  pivot_longer(c(n_all, n_promoted), names_to = "set", values_to = "n") |>
  mutate(set = if_else(set == "n_all", "All genes", "eIF3d hypoxia promotes"),
         set = factor(set, levels = c("All genes", "eIF3d hypoxia promotes")))

p_quad <- ggplot(quad_df, aes(x = quadrant, y = n, fill = set)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("All genes" = "grey70",
                                "eIF3d hypoxia promotes" = "#E69F00")) +
  labs(
    x    = NULL,
    y    = "Number of genes",
    fill = NULL,
    title = "Gene counts by binding change quadrant"
  ) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")

# Fisher test: enrichment of eIF3d-promoted genes in "3d up / 4e down" vs all other quadrants
n_tradoff_prom  <- quad_prom |> filter(quadrant == "3d up / 4e down") |> pull(n_promoted)
n_tradoff_all   <- quad_all  |> filter(quadrant == "3d up / 4e down") |> pull(n_all)
n_other_prom    <- sum(quad_prom$n_promoted) - n_tradoff_prom
n_other_all     <- sum(quad_all$n_all) - n_tradoff_all

fisher_mat <- matrix(
  c(n_tradoff_prom, n_tradoff_all - n_tradoff_prom,
    n_other_prom,   n_other_all   - n_other_prom),
  nrow = 2,
  dimnames = list(c("promoted", "other"), c("trade-off quadrant", "other quadrants"))
)
fisher_res <- fisher.test(fisher_mat, alternative = "greater")
cat(sprintf(
  "Fisher test (eIF3d-promoted enrichment in '3d up / 4e down'): OR = %.2f, p = %.3e\n",
  fisher_res$estimate, fisher_res$p.value
))

ggsave(here("ripseq", "plots", "12_quadrant_counts.pdf"),
       p_quad, width = 5, height = 4)
p_quad
```

# Scatter across all hypoxia timepoints

```{r load_all_timepoints}
tp_labels <- c("hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")

delta_paths <- list(
  hypoxia_1hr  = list(
    d3 = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_1hr_vs_normoxia_delta.csv"),
    e4 = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_1hr_vs_normoxia_delta.csv")
  ),
  hypoxia_4hr  = list(
    d3 = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_4hr_vs_normoxia_delta.csv"),
    e4 = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_4hr_vs_normoxia_delta.csv")
  ),
  hypoxia_24hr = list(
    d3 = here("ripseq", "output", "ripseq_deseq_flag3d_hypoxia_24hr_vs_normoxia_delta.csv"),
    e4 = here("ripseq", "output", "ripseq_deseq_eif4e_hypoxia_24hr_vs_normoxia_delta.csv")
  )
)

tp_to_te <- list(
  hypoxia_1hr  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  hypoxia_4hr  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"),
  hypoxia_24hr = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")
)

stopifnot("Missing delta CSVs — run notebook 09 first" =
  all(sapply(unlist(lapply(delta_paths, unlist)), file.exists)))
stopifnot("Missing TE CSVs" =
  all(sapply(unlist(tp_to_te), file.exists)))

# load all loci without deduplication
plot_df_all_raw <- imap_dfr(delta_paths, \(paths, tp) {
  d3 <- read_csv(paths$d3, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, symbol, baseMean, delta_flag3d_lfc)
  e4 <- read_csv(paths$e4, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, baseMean_4e = baseMean, delta_eif4e_lfc)
  te <- read_csv(tp_to_te[[tp]], show_col_types = FALSE) |>
    dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    dplyr::select(gene_id_clean, te_lfc, te_padj)

  inner_join(d3, e4, by = "gene_id_clean") |>
    left_join(te, by = "gene_id_clean") |>
    dplyr::mutate(timepoint = tp)
})

# canonical locus per gene: highest mean baseMean across timepoints
canonical_ids_tp <- plot_df_all_raw |>
  dplyr::group_by(symbol, gene_id_clean) |>
  dplyr::summarise(mean_baseMean = mean(baseMean), .groups = "drop") |>
  dplyr::group_by(symbol) |>
  dplyr::slice_max(order_by = mean_baseMean, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, gene_id_clean)

plot_df_all <- plot_df_all_raw |>
  dplyr::filter(gene_id_clean %in% canonical_ids_tp$gene_id_clean) |>
  dplyr::mutate(
    timepoint = factor(timepoint, levels = tp_labels),
    group = if_else(
      !is.na(te_lfc) & !is.na(te_padj) & te_lfc > TE_LFC_CUT & te_padj < TE_PADJ_CUT,
      "eIF3d hypoxia promotes", "All other genes"
    ),
    group = factor(group, levels = c("eIF3d hypoxia promotes", "All other genes"))
  )

cat("Genes and eIF3d-promoted counts per timepoint:\n")
plot_df_all |> dplyr::count(timepoint, group) |> print()
```

```{r scatter_all_timepoints, fig.width=14, fig.height=5}
rho_df_all <- plot_df_all |>
  dplyr::group_by(timepoint) |>
  dplyr::summarise(
    rho   = cor.test(delta_flag3d_lfc, delta_eif4e_lfc,
                     method = "spearman", exact = FALSE)$estimate,
    pval  = cor.test(delta_flag3d_lfc, delta_eif4e_lfc,
                     method = "spearman", exact = FALSE)$p.value,
    n_prom = sum(group == "eIF3d hypoxia promotes"),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    rho_label  = sprintf("rho = %.2f, p = %.2e", rho, pval),
    prom_label = paste0("eIF3d promotes (", gsub("hypoxia_", "", as.character(timepoint)), "): n = ", n_prom)
  )

p_scatter_all <- ggplot(mapping = aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc)) +
  geom_point(data = filter(plot_df_all, group == "All other genes"),
             color = "grey80", alpha = 0.25, size = 0.5) +
  geom_point(data = filter(plot_df_all, group == "eIF3d hypoxia promotes"),
             color = "#E69F00", alpha = 0.5, size = 0.7) +
  geom_smooth(data = plot_df_all, method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_text(data = rho_df_all, aes(label = rho_label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  geom_text(data = rho_df_all, aes(label = prom_label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 2.8,
            color = "#E69F00", size = 5, inherit.aes = FALSE) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia, timepoint-matched)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia, timepoint-matched)",
    title = "eIF3d vs eIF4E binding trade-off across hypoxia timepoints\n(axes and eIF3d-promoted coloring timepoint-matched)"
  ) +
  theme_classic(base_size = 16)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_all_timepoints.pdf"),
       p_scatter_all, width = 14, height = 5)
p_scatter_all
```

```{r scatter_te_rank_all_timepoints, fig.width=14, fig.height=5}
tp_to_sictrl_te <- list(
  hypoxia_1hr  = here("output", "translation_categories_1hr.csv"),
  hypoxia_4hr  = here("output", "translation_categories_4hr.csv"),
  hypoxia_24hr = here("output", "translation_categories_24hr.csv")
)

stopifnot("Missing sictrl TE CSVs" =
  all(sapply(unlist(tp_to_sictrl_te), file.exists)))

sictrl_te_all <- imap_dfr(tp_to_sictrl_te, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    dplyr::select(gene_id_clean, sictrl_te_lfc = te_lfc) |>
    dplyr::mutate(timepoint = tp)
})

plot_df_rank_all <- plot_df_all |>
  dplyr::select(gene_id_clean, timepoint, delta_flag3d_lfc, delta_eif4e_lfc) |>
  left_join(sictrl_te_all, by = c("gene_id_clean", "timepoint")) |>
  dplyr::group_by(timepoint) |>
  dplyr::mutate(te_prank = percent_rank(sictrl_te_lfc)) |>
  dplyr::ungroup() |>
  dplyr::arrange(te_prank)

rho_df_rank_all <- plot_df_rank_all |>
  dplyr::group_by(timepoint) |>
  dplyr::summarise(
    rho  = cor.test(delta_flag3d_lfc, delta_eif4e_lfc,
                    method = "spearman", exact = FALSE)$estimate,
    pval = cor.test(delta_flag3d_lfc, delta_eif4e_lfc,
                    method = "spearman", exact = FALSE)$p.value,
    .groups = "drop"
  ) |>
  dplyr::mutate(rho_label = sprintf("rho = %.2f, p = %.2e", rho, pval))

p_scatter_te_rank_all <- ggplot(
  plot_df_rank_all,
  aes(x = delta_flag3d_lfc, y = delta_eif4e_lfc, color = te_prank)
) +
  geom_point(alpha = 0.5, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE,
              color = "black", linewidth = 0.7, fill = "grey70") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_text(data = rho_df_rank_all, aes(label = rho_label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_gradient2(
    low      = "#2C7BB6",
    mid      = "grey85",
    high     = "#D7191C",
    midpoint = 0.5,
    limits   = c(0, 1),
    na.value = "grey85",
    name     = "sictrl TE rank\n(timepoint-matched)",
    labels   = c("lowest", "", "median", "", "highest"),
    breaks   = c(0, 0.25, 0.5, 0.75, 1)
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "eIF3d binding change (log2FC, hypoxia vs normoxia, timepoint-matched)",
    y     = "eIF4E binding change (log2FC, hypoxia vs normoxia, timepoint-matched)",
    title = "eIF3d vs eIF4E binding, colored by sictrl TE rank (timepoint-matched)"
  ) +
  theme_classic(base_size = 16)

ggsave(here("ripseq", "plots", "12_scatter_delta3d_vs_delta4e_te_rank_all_timepoints.pdf"),
       p_scatter_te_rank_all, width = 14, height = 5)
p_scatter_te_rank_all
```

# Trade-off genes: 4E binding decrease, 3D neutral, across timepoints

```{r trade_off_tp}
TRADEOFF_4E_CUT <- -2
TRADEOFF_3D_ABS <-  1

trade_off_tp <- plot_df_all |>
  dplyr::filter(
    delta_eif4e_lfc  <  TRADEOFF_4E_CUT,
    abs(delta_flag3d_lfc) < TRADEOFF_3D_ABS
  )

cat("Trade-off gene counts per timepoint (4E drop <", TRADEOFF_4E_CUT,
    ", |3d change| <", TRADEOFF_3D_ABS, "):\n")
trade_off_tp |> dplyr::count(timepoint) |> print()

for (tp in levels(trade_off_tp$timepoint)) {
  cat("\n---", tp, "---\n")
  trade_off_tp |>
    dplyr::filter(timepoint == tp) |>
    dplyr::select(symbol, delta_flag3d_lfc, delta_eif4e_lfc) |>
    dplyr::arrange(delta_eif4e_lfc) |>
    print()
}
```

```{r venn_trade_off_tp, fig.width=6, fig.height=6}
sets_tp <- split(trade_off_tp$gene_id_clean, trade_off_tp$timepoint)

p_venn_tp <- ggvenn(
  sets_tp,
  fill_color  = c("#56B4E9", "#E69F00", "#009E73"),
  stroke_size = 0.5,
  text_size   = 4.5
) +
  ggtitle(paste0(
    "Trade-off genes overlap across timepoints\n",
    "(4E drop < ", TRADEOFF_4E_CUT, ", |3d change| < ", TRADEOFF_3D_ABS, ")"
  )) +
  theme_void(base_size = 14)

ggsave(here("ripseq", "plots", "12_venn_tradeoff_genes_timepoints.pdf"),
       p_venn_tp, width = 6, height = 6)
p_venn_tp
```

```{r trade_off_tp_table}
trade_off_wide <- trade_off_tp |>
  dplyr::select(gene_id_clean, symbol, timepoint) |>
  dplyr::mutate(present = TRUE) |>
  tidyr::pivot_wider(names_from = timepoint, values_from = present,
                     values_fill = FALSE) |>
  dplyr::mutate(n_timepoints = rowSums(dplyr::across(all_of(tp_labels)))) |>
  dplyr::arrange(desc(n_timepoints), symbol)

cat("Genes present at 2+ timepoints:\n")
trade_off_wide |> dplyr::filter(n_timepoints >= 2) |> print()

cat("\nGenes present at all 3 timepoints:\n")
trade_off_wide |> dplyr::filter(n_timepoints == 3) |> print()
```

```{r}

```