---
title: "KEGG GSEA: eIF4E and FLAG-eIF3d RIP-seq"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

GSEA (fgsea) on eIF4E and FLAG-eIF3d RIP-seq data across four conditions: normoxia 1hr, hypoxia 1hr, hypoxia 4hr, and hypoxia 24hr. Genes are ranked by log2FoldChange (IP vs IgG). Heatmaps show NES per condition; pathways included if padj < 0.05 in any condition.

# Load Libraries

```{r load_libraries}
library(fgsea)
library(msigdbr)
library(pheatmap)
library(tidyverse)
library(here)
```

# Load KEGG Pathways

```{r load_pathways}
to_pathways <- function(db) {
  db %>%
    dplyr::select(gs_name, gene_symbol) %>%
    group_by(gs_name) %>%
    summarise(genes = list(gene_symbol), .groups = "drop") %>%
    deframe()
}

pathways_kegg <- to_pathways(
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY")
)

cat("KEGG pathways loaded:", length(pathways_kegg), "\n")
```

# Helper Functions

```{r helpers}
make_ranked <- function(df) {
  df <- df %>%
    filter(!is.na(symbol), !is.na(log2FoldChange)) %>%
    group_by(symbol) %>%
    slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
    ungroup()
  ranked <- df$log2FoldChange
  names(ranked) <- df$symbol
  ranked <- sort(ranked, decreasing = TRUE)
  ranked[!duplicated(names(ranked))]
}

run_fgsea <- function(ranked, pathways, condition_label, gsea_param = 1) {
  set.seed(9)
  res <- fgsea(
    pathways  = pathways,
    stats     = ranked,
    minSize   = 10,
    maxSize   = 500,
    gseaParam = gsea_param,
    nproc     = 1
  )
  res %>%
    as_tibble() %>%
    dplyr::select(pathway, NES, pval, padj, leadingEdge) %>%
    mutate(condition = condition_label)
}

make_gsea_heatmap <- function(results_list, title, outfile,
                              col_order = c("Normoxia 1hr", "Hypoxia 1hr",
                                            "Hypoxia 4hr", "Hypoxia 24hr")) {
  combined <- bind_rows(results_list)

  sig_pathways <- combined %>%
    filter(padj < 0.05) %>%
    pull(pathway) %>%
    unique()

  if (length(sig_pathways) == 0) {
    message("No significant pathways (padj < 0.05) for: ", title)
    return(invisible(NULL))
  }

  wide <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    dplyr::select(pathway, NES, condition) %>%
    pivot_wider(names_from = condition, values_from = NES, values_fill = 0) %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(wide)) wide[[cn]] <- 0
  wide <- wide[, col_order, drop = FALSE]

  sig_mat <- combined %>%
    filter(pathway %in% sig_pathways) %>%
    mutate(sig = ifelse(padj < 0.05, "*", "")) %>%
    dplyr::select(pathway, condition, sig) %>%
    pivot_wider(names_from = condition, values_from = sig, values_fill = "") %>%
    column_to_rownames("pathway")
  for (cn in col_order) if (!cn %in% colnames(sig_mat)) sig_mat[[cn]] <- ""
  sig_mat <- sig_mat[rownames(wide), col_order, drop = FALSE]

  max_abs <- max(abs(wide), na.rm = TRUE)
  max_abs <- max(max_abs, 0.1)

  p <- pheatmap(
    wide,
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    scale           = "none",
    color           = colorRampPalette(c("blue", "white", "red"))(100),
    breaks          = seq(-max_abs, max_abs, length.out = 101),
    display_numbers = sig_mat,
    number_color    = "black",
    fontsize_number = 12,
    main            = title,
    fontsize_row    = 7,
    fontsize_col    = 10,
    cellwidth       = 40,
    cellheight      = 12
  )

  pdf(here("ripseq", "plots", outfile),
      width = 10, height = max(4, length(sig_pathways) * 0.18 + 3))
  print(p)
  dev.off()

  message("Saved: ", outfile, " (", length(sig_pathways), " pathways)")
  invisible(p)
}
```

# eIF4E IP KEGG GSEA

## Load and rank data

```{r eif4e_load}
eif4e_files <- list(
  "Normoxia 1hr" = here("ripseq", "output", "ripseq_deseq_normoxia_1hr_eif4e_vs_igg.csv"),
  "Hypoxia 1hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_eif4e_vs_igg.csv"),
  "Hypoxia 4hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_4hr_eif4e_vs_igg.csv"),
  "Hypoxia 24hr" = here("ripseq", "output", "ripseq_deseq_hypoxia_24hr_eif4e_vs_igg.csv")
)

eif4e_data <- map(eif4e_files, read_csv, show_col_types = FALSE)

# Sanity check
walk2(eif4e_data, names(eif4e_data), function(df, label) {
  n_sym  <- sum(!is.na(df$symbol))
  n_bound <- sum(df$log2FoldChange > 1 & df$padj < 0.05, na.rm = TRUE)
  cat(label, "| genes with symbol:", n_sym, "| bound (log2FC>1, padj<0.05):", n_bound, "\n")
})

eif4e_ranked <- map2(eif4e_data, names(eif4e_data), ~ make_ranked(.x))
cat("\nRanked list sizes:\n")
walk2(eif4e_ranked, names(eif4e_ranked), ~ cat(.y, ":", length(.x), "\n"))
```

## Run fgsea

```{r eif4e_fgsea}
eif4e_gsea <- map2(eif4e_ranked, names(eif4e_ranked),
                   ~ run_fgsea(.x, pathways_kegg, .y))

walk2(eif4e_gsea, names(eif4e_gsea), function(res, label) {
  n_sig <- sum(res$padj < 0.05, na.rm = TRUE)
  cat(label, "| significant KEGG pathways (padj<0.05):", n_sig, "\n")
})
```

## Heatmap

```{r eif4e_heatmap, fig.width=10, fig.height=8}
make_gsea_heatmap(
  results_list = eif4e_gsea,
  title        = "eIF4E RIP KEGG GSEA (NES, * = padj < 0.05)",
  outfile      = "14_kegg_gsea_eif4e_rip.pdf"
)
```

# FLAG-eIF3d IP KEGG GSEA

## Load and rank data

```{r flag3d_load}
flag3d_files <- list(
  "Normoxia 1hr" = here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg.csv"),
  "Hypoxia 1hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg.csv"),
  "Hypoxia 4hr"  = here("ripseq", "output", "ripseq_deseq_hypoxia_4hr_flag_eif3d_vs_igg.csv"),
  "Hypoxia 24hr" = here("ripseq", "output", "ripseq_deseq_hypoxia_24hr_flag_eif3d_vs_igg.csv")
)

flag3d_data <- map(flag3d_files, read_csv, show_col_types = FALSE)

# Sanity check — EIF3D should be a top hit in normoxia 1hr
walk2(flag3d_data, names(flag3d_data), function(df, label) {
  n_sym   <- sum(!is.na(df$symbol))
  n_bound <- sum(df$log2FoldChange > 1 & df$padj < 0.05, na.rm = TRUE)
  cat(label, "| genes with symbol:", n_sym, "| bound (log2FC>1, padj<0.05):", n_bound, "\n")
})

eif3d_lfc_nor1hr <- flag3d_data[["Normoxia 1hr"]] %>%
  filter(symbol == "EIF3D") %>%
  pull(log2FoldChange)
cat("\nEIF3D log2FC in FLAG-eIF3d IP normoxia 1hr:", round(eif3d_lfc_nor1hr, 2),
    "(expected strongly positive)\n")

flag3d_ranked <- map2(flag3d_data, names(flag3d_data), ~ make_ranked(.x))
cat("\nRanked list sizes:\n")
walk2(flag3d_ranked, names(flag3d_ranked), ~ cat(.y, ":", length(.x), "\n"))
```

## Run fgsea

```{r flag3d_fgsea}
flag3d_gsea <- map2(flag3d_ranked, names(flag3d_ranked),
                    ~ run_fgsea(.x, pathways_kegg, .y))

walk2(flag3d_gsea, names(flag3d_gsea), function(res, label) {
  n_sig <- sum(res$padj < 0.05, na.rm = TRUE)
  cat(label, "| significant KEGG pathways (padj<0.05):", n_sig, "\n")
})
```

## Heatmap

```{r flag3d_heatmap, fig.width=10, fig.height=8}
make_gsea_heatmap(
  results_list = flag3d_gsea,
  title        = "FLAG-eIF3d RIP KEGG GSEA (NES, * = padj < 0.05)",
  outfile      = "14_kegg_gsea_flag3d_rip.pdf"
)
```
