---
title: "RIP-seq: eIF3d FLAG-IP enrichment vs Lee 2015 PAR-CLIP targets"
author: "Kate"
date: "2026-06"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(janitor)
library(ggvenn)
library(cowplot)
```

Compares FLAG-eIF3d RIP-seq enrichment (our data, 4 timepoints) to eIF3d PAR-CLIP
targets from Lee et al. 2015. Mirrors the Teleman eIF4E comparison in notebook 03.

```{r constants}
LFC_CUTOFF  <- 0.5
PADJ_CUTOFF <- 0.05
LEE_COLOR   <- "#009E73"
OTHER_COLOR <- "grey40"
TP_LEVELS   <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")
```

# Load data

```{r load_lee}
lee_parclip <- read_csv(
  "/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv",
  show_col_types = FALSE
) |> clean_names()

lee_eif3d_ids <- lee_parclip |>
  filter(subunit == "d") |>
  mutate(ensembl_id = str_replace(ensembl_id, "\\..*", "")) |>
  pull(ensembl_id) |>
  unique()

lee_eif3d_names <- lee_parclip |>
  filter(subunit == "d") |>
  pull(gene_name) |>
  unique()

cat("Lee PAR-CLIP eIF3d-bound genes (Ensembl IDs):", length(lee_eif3d_ids), "\n")
cat("Lee PAR-CLIP eIF3d-bound genes (gene names):", length(lee_eif3d_names), "\n")
```

```{r load_rip}
rip_paths <- list(
  normoxia_1hr = here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg.csv"),
  hypoxia_1hr  = here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg.csv"),
  hypoxia_4hr  = here("ripseq", "output", "ripseq_deseq_hypoxia_4hr_flag_eif3d_vs_igg.csv"),
  hypoxia_24hr = here("ripseq", "output", "ripseq_deseq_hypoxia_24hr_flag_eif3d_vs_igg.csv")
)

stopifnot("Run notebook 01 first to generate FLAG-eIF3d DESeq2 CSVs" =
  all(file.exists(unlist(rip_paths))))

rip_all <- imap_dfr(rip_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    mutate(tp = tp)
})

cat("RIP-seq genes (all timepoints):", nrow(rip_all), "\n")
```

# Join and label

```{r join}
rip_labeled <- rip_all |>
  filter(!is.na(log2FoldChange)) |>
  mutate(
    lee_bound    = gene_id_clean %in% lee_eif3d_ids,
    our_enriched = log2FoldChange > LFC_CUTOFF & !is.na(padj) & padj < PADJ_CUTOFF,
    tp           = factor(tp, levels = TP_LEVELS)
  )

stopifnot("Too few Lee-bound genes after join — check Ensembl ID matching" =
  sum(rip_labeled$lee_bound) > 50)

cat("Lee-bound genes detected per timepoint:\n")
rip_labeled |> dplyr::count(tp, lee_bound) |> print()

cat("\nOur enriched genes per timepoint (LFC >", LFC_CUTOFF, ", padj <", PADJ_CUTOFF, "):\n")
rip_labeled |> dplyr::count(tp, our_enriched) |> print()
```

# CDF: FLAG-eIF3d enrichment by Lee PAR-CLIP status, faceted by timepoint

```{r cdf}
wt_df <- rip_labeled |>
  group_by(tp) |>
  summarise(
    p     = wilcox.test(log2FoldChange ~ lee_bound, alternative = "less")$p.value,
    label = paste0("p = ", signif(p, 3)),
    .groups = "drop"
  )

n_lee  <- sum(rip_labeled$lee_bound[rip_labeled$tp == "normoxia_1hr"])
n_other <- sum(!rip_labeled$lee_bound[rip_labeled$tp == "normoxia_1hr"])

p_cdf <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 3.5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP bound vs all others"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")+
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound.pdf"),
       p_cdf, width = 12, height = 4)
p_cdf
```

# Venn diagrams: our enriched vs Lee targets, per timepoint

```{r venn, fig.width=10, fig.height=10}
venn_plots <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> filter(tp == tp_name)

  our_ids <- df |> filter(our_enriched) |> pull(gene_id_clean)
  lee_ids <- lee_eif3d_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched:", length(our_ids),
      "| Lee bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched" = our_ids,
    "Lee PAR-CLIP\neIF3d bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", LEE_COLOR),
         stroke_size = 0.5,
         text_size   = 3.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_grid <- plot_grid(plotlist = venn_plots, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip.pdf"),
       p_venn_grid, width = 10, height = 10)
p_venn_grid
```

# LFC > 1 threshold

```{r label_lfc1}
LFC_CUTOFF_1 <- 1.0

rip_labeled <- rip_labeled |>
  mutate(our_enriched_lfc1 = log2FoldChange > LFC_CUTOFF_1 & !is.na(padj) & padj < PADJ_CUTOFF)

cat("Our enriched genes per timepoint (LFC >", LFC_CUTOFF_1, ", padj <", PADJ_CUTOFF, "):\n")
rip_labeled |> dplyr::count(tp, our_enriched_lfc1) |> print()
```

# CDF (LFC > 1): FLAG-eIF3d enrichment by Lee PAR-CLIP status, faceted by timepoint

```{r cdf_lfc1}
p_cdf_lfc1 <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 3.5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP bound vs all others (LFC > 1 threshold)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom") +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound_lfc1.pdf"),
       p_cdf_lfc1, width = 12, height = 4)
p_cdf_lfc1
```

# Venn diagrams (LFC > 1): our enriched vs Lee targets, per timepoint

```{r venn_lfc1, fig.width=10, fig.height=10}
venn_plots_lfc1 <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> filter(tp == tp_name)

  our_ids <- df |> filter(our_enriched_lfc1) |> pull(gene_id_clean)
  lee_ids <- lee_eif3d_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched (LFC > 1):", length(our_ids),
      "| Lee bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched" = our_ids,
    "Lee PAR-CLIP\neIF3d bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", LEE_COLOR),
         stroke_size = 0.5,
         text_size   = 3.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_grid_lfc1 <- plot_grid(plotlist = venn_plots_lfc1, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_lfc1.pdf"),
       p_venn_grid_lfc1, width = 10, height = 10)
p_venn_grid_lfc1
```
