---
title: "RIP-seq: eIF3d FLAG-IP enrichment vs Lee 2015 PAR-CLIP targets"
author: "Kate"
date: "2026-06"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(janitor)
library(ggvenn)
library(cowplot)
```

Compares FLAG-eIF3d RIP-seq enrichment (our data, 4 timepoints) to eIF3d PAR-CLIP
targets from Lee et al. 2015. Mirrors the Teleman eIF4E comparison in notebook 03.

```{r constants}
LFC_CUTOFF  <- 0.5
PADJ_CUTOFF <- 0.05
LEE_COLOR   <- "#009E73"
EIF3B_COLOR <- "#F0E442"
OTHER_COLOR <- "grey40"
TP_LEVELS   <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")
```

# Load data

```{r load_lee}
lee_parclip <- read_csv(
  "/Users/katematlin/github/mpra_preliminary/data/lee_parclip.csv",
  show_col_types = FALSE
) |> clean_names()

lee_eif3d_ids <- lee_parclip |>
  dplyr::filter(subunit == "d") |>
  dplyr::mutate(ensembl_id = str_replace(ensembl_id, "\\..*", "")) |>
  dplyr::pull(ensembl_id) |>
  unique()

lee_eif3d_names <- lee_parclip |>
  dplyr::filter(subunit == "d") |>
  dplyr::pull(gene_name) |>
  unique()

cat("Lee PAR-CLIP eIF3d-bound genes (Ensembl IDs):", length(lee_eif3d_ids), "\n")
cat("Lee PAR-CLIP eIF3d-bound genes (gene names):", length(lee_eif3d_names), "\n")
```

```{r load_lee_3b}
lee_eif3b_ids <- lee_parclip |>
  dplyr::filter(subunit == "b") |>
  dplyr::mutate(ensembl_id = str_replace(ensembl_id, "\\..*", "")) |>
  dplyr::pull(ensembl_id) |>
  unique()

cat("Lee PAR-CLIP eIF3b-bound genes (Ensembl IDs):", length(lee_eif3b_ids), "\n")
```

```{r load_rip}
rip_paths <- list(
  normoxia_1hr = here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg.csv"),
  hypoxia_1hr  = here("ripseq", "output", "ripseq_deseq_hypoxia_1hr_flag_eif3d_vs_igg.csv"),
  hypoxia_4hr  = here("ripseq", "output", "ripseq_deseq_hypoxia_4hr_flag_eif3d_vs_igg.csv"),
  hypoxia_24hr = here("ripseq", "output", "ripseq_deseq_hypoxia_24hr_flag_eif3d_vs_igg.csv")
)

stopifnot("Run notebook 01 first to generate FLAG-eIF3d DESeq2 CSVs" =
  all(file.exists(unlist(rip_paths))))

rip_all <- imap_dfr(rip_paths, \(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::mutate(tp = tp)
})

cat("RIP-seq genes (all timepoints):", nrow(rip_all), "\n")
```

# Join and label

```{r join}
rip_labeled <- rip_all |>
  dplyr::filter(!is.na(log2FoldChange)) |>
  dplyr::mutate(
    lee_bound    = gene_id_clean %in% lee_eif3d_ids,
    lee_3b_bound = gene_id_clean %in% lee_eif3b_ids,
    our_enriched = log2FoldChange > LFC_CUTOFF & !is.na(padj) & padj < PADJ_CUTOFF,
    tp           = factor(tp, levels = TP_LEVELS)
  )

stopifnot("Too few Lee-bound genes after join — check Ensembl ID matching" =
  sum(rip_labeled$lee_bound) > 50)
stopifnot("Too few Lee eIF3b-bound genes after join — check Ensembl ID matching" =
  sum(rip_labeled$lee_3b_bound) > 20)

cat("Lee eIF3d-bound genes detected per timepoint:\n")
rip_labeled |> dplyr::count(tp, lee_bound) |> print()

cat("\nLee eIF3b-bound genes detected per timepoint:\n")
rip_labeled |> dplyr::count(tp, lee_3b_bound) |> print()

cat("\nOur enriched genes per timepoint (LFC >", LFC_CUTOFF, ", padj <", PADJ_CUTOFF, "):\n")
rip_labeled |> dplyr::count(tp, our_enriched) |> print()
```

# CDF: FLAG-eIF3d enrichment by Lee PAR-CLIP status, faceted by timepoint

```{r cdf}
wt_df <- rip_labeled |>
  dplyr::group_by(tp) |>
  dplyr::summarise(
    p     = wilcox.test(log2FoldChange ~ lee_bound, alternative = "less")$p.value,
    label = paste0("p = ", signif(p, 3)),
    .groups = "drop"
  )

p_cdf <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP bound vs all others"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom") +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound.pdf"),
       p_cdf, width = 12, height = 4)
p_cdf
```

```{r cdf_normoxia, fig.width=6, fig.height=4}
p_cdf_normoxia <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr") |>
  ggplot(aes(x = log2FoldChange,
             color = factor(lee_bound, levels = c(TRUE, FALSE)))) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = dplyr::filter(wt_df, tp == "normoxia_1hr"), aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment:\nLee PAR-CLIP bound vs all others (normoxia 1hr)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13)) +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound_normoxia.pdf"),
       p_cdf_normoxia, width = 6, height = 4)
p_cdf_normoxia
```

# Venn diagrams: our enriched vs Lee targets, per timepoint

```{r venn, fig.width=10, fig.height=10}
venn_plots <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> dplyr::filter(tp == tp_name)

  our_ids <- df |> dplyr::filter(our_enriched) |> dplyr::pull(gene_id_clean)
  lee_ids <- lee_eif3d_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched:", length(our_ids),
      "| Lee bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched" = our_ids,
    "Lee PAR-CLIP\neIF3d bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", LEE_COLOR),
         stroke_size = 0.5,
         text_size   = 4.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_grid <- plot_grid(plotlist = venn_plots, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip.pdf"),
       p_venn_grid, width = 10, height = 10)
p_venn_grid
```

```{r venn_normoxia, fig.width=5, fig.height=5}
norm_our_ids <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr", our_enriched) |>
  dplyr::pull(gene_id_clean)

p_venn_normoxia <- ggvenn(
  list("FLAG-eIF3d\nenriched" = norm_our_ids,
       "Lee PAR-CLIP\neIF3d bound" = lee_eif3d_ids),
  fill_color  = c("#E69F00", LEE_COLOR),
  stroke_size = 0.5,
  text_size   = 4.5
) +
  ggtitle("normoxia_1hr") +
  theme_void(base_size = 16)

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_normoxia.pdf"),
       p_venn_normoxia, width = 5, height = 5)
p_venn_normoxia
```

# LFC > 1 threshold

```{r label_lfc1}
LFC_CUTOFF_1 <- 1.0

rip_labeled <- rip_labeled |>
  dplyr::mutate(our_enriched_lfc1 = log2FoldChange > LFC_CUTOFF_1 & !is.na(padj) & padj < PADJ_CUTOFF)

cat("Our enriched genes per timepoint (LFC >", LFC_CUTOFF_1, ", padj <", PADJ_CUTOFF, "):\n")
rip_labeled |> dplyr::count(tp, our_enriched_lfc1) |> print()
```

# CDF (LFC > 1): FLAG-eIF3d enrichment by Lee PAR-CLIP status, faceted by timepoint

```{r cdf_lfc1}
p_cdf_lfc1 <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP bound vs all others (LFC > 1 threshold)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom") +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound_lfc1.pdf"),
       p_cdf_lfc1, width = 12, height = 4)
p_cdf_lfc1
```

```{r cdf_lfc1_normoxia, fig.width=6, fig.height=4}
p_cdf_lfc1_normoxia <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr") |>
  ggplot(aes(x = log2FoldChange,
             color = factor(lee_bound, levels = c(TRUE, FALSE)))) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = dplyr::filter(wt_df, tp == "normoxia_1hr"), aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = LEE_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP bound (n = ", length(lee_eif3d_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment:\nLee PAR-CLIP bound vs all others (normoxia 1hr, LFC > 1)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13)) +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_bound_lfc1_normoxia.pdf"),
       p_cdf_lfc1_normoxia, width = 6, height = 4)
p_cdf_lfc1_normoxia
```

# Venn diagrams (LFC > 1): our enriched vs Lee targets, per timepoint

```{r venn_lfc1, fig.width=10, fig.height=10}
venn_plots_lfc1 <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> dplyr::filter(tp == tp_name)

  our_ids <- df |> dplyr::filter(our_enriched_lfc1) |> dplyr::pull(gene_id_clean)
  lee_ids <- lee_eif3d_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched (LFC > 1):", length(our_ids),
      "| Lee bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched" = our_ids,
    "Lee PAR-CLIP\neIF3d bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", LEE_COLOR),
         stroke_size = 0.5,
         text_size   = 4.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_grid_lfc1 <- plot_grid(plotlist = venn_plots_lfc1, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_lfc1.pdf"),
       p_venn_grid_lfc1, width = 10, height = 10)
p_venn_grid_lfc1
```

```{r venn_lfc1_normoxia, fig.width=5, fig.height=5}
norm_our_ids_lfc1 <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr", our_enriched_lfc1) |>
  dplyr::pull(gene_id_clean)

p_venn_lfc1_normoxia <- ggvenn(
  list("FLAG-eIF3d\nenriched" = norm_our_ids_lfc1,
       "Lee PAR-CLIP\neIF3d bound" = lee_eif3d_ids),
  fill_color  = c("#E69F00", LEE_COLOR),
  stroke_size = 0.5,
  text_size   = 4.5
) +
  ggtitle("normoxia_1hr (LFC > 1)") +
  theme_void(base_size = 16)

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_lfc1_normoxia.pdf"),
       p_venn_lfc1_normoxia, width = 5, height = 5)
p_venn_lfc1_normoxia
```

# eIF3b PAR-CLIP: same analysis for Lee eIF3b targets

```{r wt_3b}
wt_3b_df <- rip_labeled |>
  dplyr::group_by(tp) |>
  dplyr::summarise(
    p     = wilcox.test(log2FoldChange ~ lee_3b_bound, alternative = "less")$p.value,
    label = paste0("p = ", signif(p, 3)),
    .groups = "drop"
  )
```

```{r cdf_3b, fig.width=12, fig.height=4}
p_cdf_3b <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_3b_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_3b_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = EIF3B_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP eIF3b bound (n = ", length(lee_eif3b_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP eIF3b bound vs all others"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom") +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_3b_bound.pdf"),
       p_cdf_3b, width = 12, height = 4)
p_cdf_3b
```

```{r cdf_3b_normoxia, fig.width=6, fig.height=4}
p_cdf_3b_normoxia <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr") |>
  ggplot(aes(x = log2FoldChange,
             color = factor(lee_3b_bound, levels = c(TRUE, FALSE)))) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = dplyr::filter(wt_3b_df, tp == "normoxia_1hr"), aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = EIF3B_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP eIF3b bound (n = ", length(lee_eif3b_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment:\nLee PAR-CLIP eIF3b bound vs all others (normoxia 1hr)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13)) +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_3b_bound_normoxia.pdf"),
       p_cdf_3b_normoxia, width = 6, height = 4)
p_cdf_3b_normoxia
```

```{r venn_3b, fig.width=10, fig.height=10}
venn_3b_plots <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> dplyr::filter(tp == tp_name)

  our_ids  <- df |> dplyr::filter(our_enriched) |> dplyr::pull(gene_id_clean)
  lee_ids  <- lee_eif3b_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched:", length(our_ids),
      "| Lee eIF3b bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched"      = our_ids,
    "Lee PAR-CLIP\neIF3b bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", EIF3B_COLOR),
         stroke_size = 0.5,
         text_size   = 4.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_3b_grid <- plot_grid(plotlist = venn_3b_plots, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_3b.pdf"),
       p_venn_3b_grid, width = 10, height = 10)
p_venn_3b_grid
```

```{r venn_3b_normoxia, fig.width=5, fig.height=5}
norm_our_ids_3b <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr", our_enriched) |>
  dplyr::pull(gene_id_clean)

p_venn_3b_normoxia <- ggvenn(
  list("FLAG-eIF3d\nenriched"      = norm_our_ids_3b,
       "Lee PAR-CLIP\neIF3b bound" = lee_eif3b_ids),
  fill_color  = c("#E69F00", EIF3B_COLOR),
  stroke_size = 0.5,
  text_size   = 4.5
) +
  ggtitle("normoxia_1hr") +
  theme_void(base_size = 16)

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_3b_normoxia.pdf"),
       p_venn_3b_normoxia, width = 5, height = 5)
p_venn_3b_normoxia
```

# eIF3b PAR-CLIP: LFC > 1 threshold

```{r cdf_3b_lfc1, fig.width=12, fig.height=4}
p_cdf_3b_lfc1 <- ggplot(
  rip_labeled,
  aes(x     = log2FoldChange,
      color = factor(lee_3b_bound, levels = c(TRUE, FALSE)))
) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = wt_3b_df, aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = EIF3B_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP eIF3b bound (n = ", length(lee_eif3b_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  facet_wrap(~ tp, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment: Lee PAR-CLIP eIF3b bound vs all others (LFC > 1 threshold)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom") +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_3b_bound_lfc1.pdf"),
       p_cdf_3b_lfc1, width = 12, height = 4)
p_cdf_3b_lfc1
```

```{r cdf_3b_lfc1_normoxia, fig.width=6, fig.height=4}
p_cdf_3b_lfc1_normoxia <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr") |>
  ggplot(aes(x = log2FoldChange,
             color = factor(lee_3b_bound, levels = c(TRUE, FALSE)))) +
  stat_ecdf(linewidth = 0.8) +
  geom_text(data = dplyr::filter(wt_3b_df, tp == "normoxia_1hr"), aes(label = label),
            x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
            size = 5, inherit.aes = FALSE) +
  scale_color_manual(
    values = c("TRUE" = EIF3B_COLOR, "FALSE" = OTHER_COLOR),
    labels = c(
      "TRUE"  = paste0("Lee PAR-CLIP eIF3b bound (n = ", length(lee_eif3b_ids), ")"),
      "FALSE" = "All other genes"
    )
  ) +
  labs(
    x     = "log2FC (eIF3d IP / IgG)",
    y     = "Cumulative fraction",
    color = NULL,
    title = "FLAG-eIF3d RIP enrichment:\nLee PAR-CLIP eIF3b bound vs all others (normoxia 1hr, LFC > 1)"
  ) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13)) +
  xlim(-3, 8)

ggsave(here("ripseq", "plots", "15_cdf_rip_by_lee_parclip_3b_bound_lfc1_normoxia.pdf"),
       p_cdf_3b_lfc1_normoxia, width = 6, height = 4)
p_cdf_3b_lfc1_normoxia
```

```{r venn_3b_lfc1, fig.width=10, fig.height=10}
venn_3b_lfc1_plots <- lapply(TP_LEVELS, function(tp_name) {
  df <- rip_labeled |> dplyr::filter(tp == tp_name)

  our_ids <- df |> dplyr::filter(our_enriched_lfc1) |> dplyr::pull(gene_id_clean)
  lee_ids <- lee_eif3b_ids

  n_overlap <- length(intersect(our_ids, lee_ids))
  cat(tp_name, "- our enriched (LFC > 1):", length(our_ids),
      "| Lee eIF3b bound:", length(lee_ids),
      "| overlap:", n_overlap, "\n")

  venn_list <- list(
    "FLAG-eIF3d\nenriched"      = our_ids,
    "Lee PAR-CLIP\neIF3b bound" = lee_ids
  )

  ggvenn(venn_list,
         fill_color  = c("#E69F00", EIF3B_COLOR),
         stroke_size = 0.5,
         text_size   = 4.5) +
    ggtitle(tp_name) +
    theme_void()
})

p_venn_3b_grid_lfc1 <- plot_grid(plotlist = venn_3b_lfc1_plots, ncol = 2, labels = "AUTO")

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_3b_lfc1.pdf"),
       p_venn_3b_grid_lfc1, width = 10, height = 10)
p_venn_3b_grid_lfc1
```

```{r venn_3b_lfc1_normoxia, fig.width=5, fig.height=5}
norm_our_ids_3b_lfc1 <- rip_labeled |>
  dplyr::filter(tp == "normoxia_1hr", our_enriched_lfc1) |>
  dplyr::pull(gene_id_clean)

p_venn_3b_lfc1_normoxia <- ggvenn(
  list("FLAG-eIF3d\nenriched"      = norm_our_ids_3b_lfc1,
       "Lee PAR-CLIP\neIF3b bound" = lee_eif3b_ids),
  fill_color  = c("#E69F00", EIF3B_COLOR),
  stroke_size = 0.5,
  text_size   = 4.5
) +
  ggtitle("normoxia_1hr (LFC > 1)") +
  theme_void(base_size = 16)

ggsave(here("ripseq", "plots", "15_venn_rip_vs_lee_parclip_3b_lfc1_normoxia.pdf"),
       p_venn_3b_lfc1_normoxia, width = 5, height = 5)
p_venn_3b_lfc1_normoxia
```

# Scatter correlations: our eIF3d RIP-seq vs Lee PAR-CLIP binding score

```{r parclip_scores}
lee_parclip_score <- lee_parclip |>
  dplyr::filter(subunit == "d") |>
  dplyr::mutate(gene_id_clean = str_replace(ensembl_id, "\\..*", "")) |>
  dplyr::group_by(gene_id_clean) |>
  dplyr::summarise(parclip_reads = sum(reads, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(log2_parclip = log2(parclip_reads + 1))

cat("eIF3d PAR-CLIP genes with read scores:", nrow(lee_parclip_score), "\n")
```

```{r rna_cpm_normoxia_1hr_3d}
rna_meta_nor1 <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                           show_col_types = FALSE) |>
  dplyr::filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
                condition == "normoxia", hours == 1)

rna_raw <- read_csv(here("output", "human_allCounts_raw.csv"),
                    show_col_types = FALSE) |>
  column_to_rownames("gene_id")

rna_col_ids <- sub(".*_(NM\\d+_\\d+)$", "\\1", colnames(rna_raw))
rna_nor1 <- rna_raw[, rna_col_ids %in% rna_meta_nor1$id, drop = FALSE]
cat("siCTRL normoxia 1hr RNA samples:", ncol(rna_nor1), "\n")

rna_cpm_nor1 <- sweep(rna_nor1, 2, colSums(rna_nor1), "/") * 1e6
rna_mean_nor1 <- tibble(
  gene_id_clean = sub("\\..*", "", rownames(rna_cpm_nor1)),
  rna_cpm       = rowMeans(rna_cpm_nor1)
)
cat("Genes in RNA-seq gene universe:", nrow(rna_mean_nor1), "\n")
```

## eIF3d IP/IgG (normoxia, 1hr) vs Lee PAR-CLIP

```{r scatter_3d_igg_vs_parclip, fig.width=6, fig.height=5}
scatter_3d_igg <- rip_all |>
  dplyr::filter(tp == "normoxia_1hr", gene_id_clean %in% rna_mean_nor1$gene_id_clean,
                !is.na(log2FoldChange)) |>
  dplyr::inner_join(lee_parclip_score, by = "gene_id_clean")

cat("Genes in IP/IgG vs PAR-CLIP scatter:", nrow(scatter_3d_igg), "\n")

r_igg <- cor(scatter_3d_igg$log2FoldChange, scatter_3d_igg$log2_parclip, method = "pearson")
r_label_igg <- paste0("Pearson R = ", round(r_igg, 3))

p_scatter_3d_igg <- ggplot(scatter_3d_igg, aes(x = log2FoldChange, y = log2_parclip)) +
  geom_point(size = 0.5, alpha = 0.3, color = "grey40") +
  geom_smooth(method = "lm", color = "grey30", linewidth = 0.8, se = FALSE) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = r_label_igg, size = 5) +
  labs(
    x = "Our eIF3d IP / IgG log2FC (normoxia, 1hr)",
    y = "Lee PAR-CLIP log2(reads + 1)"
  ) +
  theme_classic(base_size = 16)

p_scatter_3d_igg
ggsave(here("ripseq", "plots", "15_scatter_3d_igg_vs_parclip.pdf"),
       p_scatter_3d_igg, width = 6, height = 5)
```

## eIF3d IP/RNA (siCTRL, normoxia, 1hr) vs Lee PAR-CLIP

```{r scatter_3d_iprna_vs_parclip, fig.width=6, fig.height=5}
rip_counts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
eif3d_nor1_cols <- grep("normoxia_flag_eif3d_1_", colnames(rip_counts), value = TRUE)
cat("eIF3d normoxia 1hr IP samples:", length(eif3d_nor1_cols), "\n")

ip_cpm_3d <- sweep(rip_counts[, eif3d_nor1_cols, drop = FALSE],
                   2, colSums(rip_counts[, eif3d_nor1_cols, drop = FALSE]), "/") * 1e6
ip_mean_3d <- tibble(
  gene_id_clean = sub("\\..*", "", rownames(ip_cpm_3d)),
  ip_cpm        = rowMeans(ip_cpm_3d)
)

scatter_3d_iprna <- inner_join(ip_mean_3d, rna_mean_nor1, by = "gene_id_clean") |>
  dplyr::mutate(log2_ip_over_rna = log2((ip_cpm + 1) / (rna_cpm + 1))) |>
  dplyr::inner_join(lee_parclip_score, by = "gene_id_clean")

cat("Genes in IP/RNA vs PAR-CLIP scatter:", nrow(scatter_3d_iprna), "\n")

r_iprna <- cor(scatter_3d_iprna$log2_ip_over_rna, scatter_3d_iprna$log2_parclip, method = "pearson")
r_label_iprna <- paste0("Pearson R = ", round(r_iprna, 3))

p_scatter_3d_iprna <- ggplot(scatter_3d_iprna, aes(x = log2_ip_over_rna, y = log2_parclip)) +
  geom_point(size = 0.5, alpha = 0.3, color = "grey40") +
  geom_smooth(method = "lm", color = "grey30", linewidth = 0.8, se = FALSE) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = r_label_iprna, size = 5) +
  labs(
    x = "Our eIF3d IP / RNA log2 ratio (siCTRL, normoxia, 1hr)",
    y = "Lee PAR-CLIP log2(reads + 1)"
  ) +
  theme_classic(base_size = 16)

p_scatter_3d_iprna
ggsave(here("ripseq", "plots", "15_scatter_3d_iprna_vs_parclip.pdf"),
       p_scatter_3d_iprna, width = 6, height = 5)
```

```{r}

```