---
title: "RIP-seq: IP vs riboseq sictrl RNA (DESeq2) vs IP/IgG (DESeq2) — correlation with eIF3d-promoted targets"
author: "Kate"
date: "2026-06"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
```

# Overview

Compute three DESeq2-based enrichment metrics for RIP-seq:

1. **IP / riboseq sictrl RNA** (DESeq2): raw RIP-seq IP counts combined with raw,
   condition- and timepoint-matched sictrl RNA-seq counts from the parallel riboseq
   experiment, contrasted with `DESeqDataSetFromMatrix(design = ~ seq_type)` the same
   way IP/IgG is. Caveat: RIP-seq (2026 batch) and RNA-seq (2025 batch) are different
   sequencing experiments with no shared spike-in, so `seq_type` is fully confounded
   with sequencing batch — DESeq2's median-of-ratios normalization cannot separate true
   IP enrichment from batch/library-prep composition differences. This limitation is
   inherent (the prior CPM-ratio method was equally uncorrected for it) and not fixable
   without a shared spike-in, so it is documented rather than "fixed" with an
   unidentifiable batch covariate.

2. **IP / IgG** (existing): log2FC from DESeq2 with outliers removed.

3. **IgG (mock) / riboseq sictrl RNA** (new, Section 1b): same DESeq2 contrast design as
   (1) above, but using IgG-mock (negative-control antibody) pulldown counts as the
   numerator instead of a bait IP -- a background/nonspecific-binding reference. Sample
   `NM2026_0077` (a confirmed real outlier: a normoxia_1hr IgG-mock replicate, rep "c")
   is excluded from the IgG-mock numerator only, via the same `OUTLIER_ID` blanket-`grepl`
   pattern used in `code/01_eif3d_rip_deseq.Rmd`. Because 0077 is normoxia_1hr-only, this
   drops normoxia_1hr IgG-mock from 3 to 2 replicates; hypoxia_1hr/4hr/24hr are unaffected
   and keep all 3 IgG-mock replicates (including `NM2026_0078`, a different hypoxia_1hr
   IgG-mock replicate that is a deliberate, already-settled keep -- distinct from 0077).
   Same RIP-seq (2026) vs RNA-seq (2025) batch-confound caveat as (1) applies here
   identically (no shared spike-in). IgG/RNA does not get a TE scatter plot in this
   notebook (Sections 6/7 below are IP-only); see `19_density_3group_ip_enrichment.Rmd`
   for its CDF plots.

For strategies (1) and (2), scatter plots show enrichment vs eIF3d-promoted TE (te_lfc
from translation_categories_si3d_vs_sictrl; positive = eIF3d promotes, per CLAUDE.md
direction note).

---

# Load gene annotation

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# A gene is protein-coding if ANY transcript row has biotype "protein_coding".
# distinct() before this would arbitrarily pick one biotype per gene and misclassify
# genes whose first listed transcript is NMD or protein_coding_CDS_not_defined.
pc_genes <- gene_anno_all |>
  filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)

# Deduplicate for symbol lookups; prefer protein_coding row where available
gene_anno <- gene_anno_all |>
  arrange(gene_id_clean, desc(biotype == "protein_coding")) |>
  distinct(gene_id_clean, .keep_all = TRUE) |>
  dplyr::select(gene_id_clean, symbol, biotype)

cat("Protein-coding genes (any-transcript rule):", length(pc_genes), "\n")
cat("Unique genes in annotation:", nrow(gene_anno), "\n")
```

---

# Section 1: Compute IP/RNA enrichment (DESeq2)

## Load RIP-seq raw counts (IP samples) and matched sictrl RNA-seq raw counts

```{r load_rip_and_rna_counts}
tp_map <- tribble(
  ~timepoint,     ~condition,  ~hours,
  "normoxia_1hr", "normoxia",  "1",
  "hypoxia_1hr",  "hypoxia",   "1",
  "hypoxia_4hr",  "hypoxia",   "4",
  "hypoxia_24hr", "hypoxia",   "24"
)
tp_levels <- tp_map$timepoint

# RIP-seq: IP samples only (no IgG needed for this contrast)
rip_counts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
rip_meta   <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"), show_col_types = FALSE) |>
  mutate(hours = as.character(hours))
cat("RIP-seq raw counts:", nrow(rip_counts), "genes x", ncol(rip_counts), "samples\n")

# RNA-seq: sictrl only, raw counts (not spike-in normalized -- DESeq2 does its own
# size-factor estimation on the combined matrix). col_select narrows the read to
# gene_id + sictrl RNA columns only (out of 108 total sample columns in the file).
rna_counts_all <- read_csv(
  here("output", "human_allCounts_raw.csv"),
  col_select = c(gene_id, matches("sictrl.*_rna_")),
  show_col_types = FALSE
) |>
  column_to_rownames("gene_id") |>
  as.matrix()
rna_meta <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE) |>
  filter(library == "rna", treatment == "sictrl") |>
  mutate(hours = as.character(hours))
cat("sictrl RNA-seq raw counts:", nrow(rna_counts_all), "genes x", ncol(rna_counts_all), "samples\n")

# RIP-seq genes are a strict subset of the RNA-seq gene set (both GENCODE v49) --
# subset RNA counts to the RIP-seq gene set, no version-stripping needed.
stopifnot("RIP-seq genes must be a subset of RNA-seq genes" =
  all(rownames(rip_counts) %in% rownames(rna_counts_all)))
rna_counts <- rna_counts_all[rownames(rip_counts), , drop = FALSE]

# Protein-coding filter (belt-and-suspenders: rip_counts is already pc-filtered
# upstream in 00_ripseq_qc.Rmd, so this is a no-op in practice)
rip_gene_id_clean <- sub("\\..*", "", rownames(rip_counts))
rip_counts <- rip_counts[rip_gene_id_clean %in% pc_genes, , drop = FALSE]
rna_counts <- rna_counts[rownames(rip_counts), , drop = FALSE]
cat("After protein-coding filter:", nrow(rip_counts), "genes\n")
```

## DESeq2: IP vs sictrl RNA, per ip_type x timepoint

```{r run_deseq_ip_vs_rna}
run_deseq_ip_vs_rna <- function(ip_target, timepoint, condition, hours) {
  ip_ids  <- rip_meta |>
    filter(ip_type == ip_target, condition == !!condition, hours == !!hours) |>
    pull(id)
  rna_ids <- rna_meta |>
    filter(condition == !!condition, hours == !!hours) |>
    pull(id)
  stopifnot(
    "Expected 3 IP replicates" = length(ip_ids) == 3,
    "Expected 3 sictrl RNA replicates" = length(rna_ids) == 3
  )

  ip_cts  <- rip_counts[, grep(paste(ip_ids, collapse = "|"), colnames(rip_counts)), drop = FALSE]
  rna_cts <- rna_counts[, grep(paste(rna_ids, collapse = "|"), colnames(rna_counts)), drop = FALSE]
  stopifnot(
    "grep on IP sample IDs did not resolve to exactly 3 columns" = ncol(ip_cts) == 3,
    "grep on RNA sample IDs did not resolve to exactly 3 columns" = ncol(rna_cts) == 3
  )

  # Genes with zero RNA counts across all 3 sictrl replicates have no real
  # denominator -- DESeq2/apeglm can still report these as large, "significant"
  # log2FC (IP counts alone drive the estimate), which is a zero-denominator
  # artifact, not enrichment. The old CPM-ratio method excluded these via
  # `mean_rna_norm > 0`; reproduce that guard here before fitting DESeq2.
  rna_covered  <- rowSums(rna_cts) > 0
  n_dropped    <- sum(!rna_covered)
  if (n_dropped > 0) {
    cat("  Dropping", n_dropped, "genes with zero RNA coverage across all sictrl reps (",
        ip_target, timepoint, ")\n")
  }
  ip_cts  <- ip_cts[rna_covered, , drop = FALSE]
  rna_cts <- rna_cts[rna_covered, , drop = FALSE]

  combined_cts <- round(cbind(ip_cts, rna_cts))

  seq_type <- factor(c(rep(ip_target, ncol(ip_cts)), rep("rna", ncol(rna_cts))),
                      levels = c("rna", ip_target))
  col_data <- data.frame(seq_type = seq_type, row.names = colnames(combined_cts))

  dds  <- DESeqDataSetFromMatrix(combined_cts, col_data, design = ~ seq_type)
  dds  <- DESeq(dds, quiet = TRUE)
  result_names <- resultsNames(dds)
  coef <- result_names[grep(ip_target, result_names)]
  stopifnot("Expected exactly one matching DESeq2 coefficient" = length(coef) == 1)
  res  <- as.data.frame(lfcShrink(dds, coef = coef, type = "apeglm"))

  res |>
    rownames_to_column("gene_id") |>
    mutate(
      gene_id_clean    = sub("\\..*", "", gene_id),
      ip_type          = ip_target,
      timepoint        = timepoint,
      log2_ip_over_rna = log2FoldChange
    ) |>
    dplyr::select(gene_id_clean, ip_type, timepoint, baseMean, log2_ip_over_rna, padj)
}

enrichment_rna <- purrr::map_dfr(c("flag_eif3d", "eif4e"), function(ip_target) {
  purrr::pmap_dfr(tp_map, function(timepoint, condition, hours) {
    run_deseq_ip_vs_rna(ip_target, timepoint, condition, hours)
  })
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

cat("\nGenes with IP/RNA enrichment per ip_type x timepoint:\n")
enrichment_rna |>
  group_by(ip_type, timepoint) |>
  summarise(n = n(), .groups = "drop") |>
  print()

# Sanity check: EIF3D (the bait) should be in top enriched genes for flag_eif3d
eif3d_check <- enrichment_rna |>
  filter(ip_type == "flag_eif3d", timepoint == "hypoxia_1hr") |>
  arrange(desc(log2_ip_over_rna)) |>
  dplyr::slice(1:10) |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean")
cat("\nTop flag_eif3d IP/RNA enriched genes at hypoxia_1hr (DESeq2):\n")
print(eif3d_check |> dplyr::select(symbol, log2_ip_over_rna, padj))

# Sanity check: EIF4E (the bait) should be in top enriched genes for eif4e
eif4e_check <- enrichment_rna |>
  filter(ip_type == "eif4e", timepoint == "hypoxia_1hr") |>
  arrange(desc(log2_ip_over_rna)) |>
  dplyr::slice(1:10) |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean")
cat("\nTop eif4e IP/RNA enriched genes at hypoxia_1hr (DESeq2):\n")
print(eif4e_check |> dplyr::select(symbol, log2_ip_over_rna, padj))
```

## Save enrichment table

```{r save_enrichment}
out_table <- enrichment_rna |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean") |>
  dplyr::select(gene_id_clean, symbol, ip_type, timepoint, baseMean, log2_ip_over_rna, padj)

write_csv(out_table, here("ripseq", "output", "18_ip_over_rna_enrichment.csv"))
cat("Saved:", here("ripseq", "output", "18_ip_over_rna_enrichment.csv"), "\n")
```

## DESeq2: IgG vs sictrl RNA, per timepoint (background reference)

Background/nonspecific-binding reference: same DESeq2 contrast design as
`run_deseq_ip_vs_rna()` above, but with IgG-mock pulldown counts as the numerator.
`NM2026_0077` (normoxia_1hr, rep "c") is a confirmed outlier and is excluded from the
IgG-mock numerator only (RNA-seq denominator is unaffected); this is a deliberate,
already-settled exclusion distinct from `NM2026_0078` (hypoxia_1hr, rep "c"), which is
retained. Because 0077 only exists at normoxia_1hr, the normoxia_1hr contrast runs on 2
IgG-mock replicates instead of 3; hypoxia_1hr/4hr/24hr are unaffected (0077 doesn't exist
there, so filtering it out is a no-op) and keep all 3 replicates.

```{r outlier_const_igg}
# Same OUTLIER_ID exclusion pattern as code/01_eif3d_rip_deseq.Rmd. NM2026_0078
# (igg_mock, hypoxia_1hr, rep c) is a *different* sample and is intentionally
# retained -- do not add it here.
OUTLIER_ID <- "0077"
```

```{r run_deseq_igg_vs_rna}
run_deseq_igg_vs_rna <- function(timepoint, condition, hours) {
  igg_ids <- rip_meta |>
    filter(ip_type == "igg_mock", condition == !!condition, hours == !!hours,
           !grepl(OUTLIER_ID, id)) |>
    pull(id)
  rna_ids <- rna_meta |>
    filter(condition == !!condition, hours == !!hours) |>
    pull(id)
  stopifnot(
    # 2 or 3, not hardcoded 3: normoxia_1hr drops to 2 reps once NM2026_0077
    # (confirmed outlier) is excluded; hypoxia timepoints stay at 3.
    "Expected 2 or 3 IgG-mock replicates" = length(igg_ids) %in% c(2, 3),
    "Expected 3 sictrl RNA replicates"    = length(rna_ids) == 3
  )

  igg_cts <- rip_counts[, grep(paste(igg_ids, collapse = "|"), colnames(rip_counts)), drop = FALSE]
  rna_cts <- rna_counts[, grep(paste(rna_ids, collapse = "|"), colnames(rna_counts)), drop = FALSE]
  stopifnot(
    "grep on IgG-mock sample IDs did not resolve to the expected column count" =
      ncol(igg_cts) == length(igg_ids),
    "grep on RNA sample IDs did not resolve to exactly 3 columns" = ncol(rna_cts) == 3
  )

  rna_covered <- rowSums(rna_cts) > 0
  n_dropped   <- sum(!rna_covered)
  if (n_dropped > 0) {
    cat("  Dropping", n_dropped, "genes with zero RNA coverage across all sictrl reps ( igg_mock",
        timepoint, ")\n")
  }
  igg_cts <- igg_cts[rna_covered, , drop = FALSE]
  rna_cts <- rna_cts[rna_covered, , drop = FALSE]

  combined_cts <- round(cbind(igg_cts, rna_cts))

  seq_type <- factor(c(rep("igg_mock", ncol(igg_cts)), rep("rna", ncol(rna_cts))),
                      levels = c("rna", "igg_mock"))
  col_data <- data.frame(seq_type = seq_type, row.names = colnames(combined_cts))

  dds  <- DESeqDataSetFromMatrix(combined_cts, col_data, design = ~ seq_type)
  dds  <- DESeq(dds, quiet = TRUE)
  result_names <- resultsNames(dds)
  coef <- result_names[grep("igg_mock", result_names)]
  stopifnot("Expected exactly one matching DESeq2 coefficient" = length(coef) == 1)
  res  <- as.data.frame(lfcShrink(dds, coef = coef, type = "apeglm"))

  res |>
    rownames_to_column("gene_id") |>
    mutate(
      gene_id_clean     = sub("\\..*", "", gene_id),
      ip_type           = "igg_mock",
      timepoint         = timepoint,
      log2_igg_over_rna = log2FoldChange
    ) |>
    dplyr::select(gene_id_clean, ip_type, timepoint, baseMean, log2_igg_over_rna, padj)
}

enrichment_igg_rna <- purrr::pmap_dfr(tp_map, function(timepoint, condition, hours) {
  run_deseq_igg_vs_rna(timepoint, condition, hours)
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

cat("\nGenes with IgG/RNA enrichment per timepoint:\n")
enrichment_igg_rna |>
  group_by(timepoint) |>
  summarise(n = n(), .groups = "drop") |>
  print()
```

```{r sanity_igg_vs_rna}
# IgG-mock has no bait protein, so there is no "self-enrichment" spot check
# analogous to the flag_eif3d / eif4e top-hit checks above. Instead: (a) confirm
# replicate counts resolved exactly as expected per timepoint, and (b) confirm
# gene coverage doesn't vary wildly across timepoints (a large swing would signal
# a broken join/filter rather than real biology).
rep_counts_igg <- rip_meta |>
  filter(ip_type == "igg_mock", !grepl(OUTLIER_ID, id)) |>
  dplyr::count(condition, hours)
print(rep_counts_igg)

stopifnot(
  "normoxia_1hr IgG-mock must have exactly 2 replicates after excluding NM2026_0077" =
    rep_counts_igg$n[rep_counts_igg$condition == "normoxia" & rep_counts_igg$hours == "1"] == 2,
  "hypoxia IgG-mock timepoints must retain all 3 replicates" =
    all(rep_counts_igg$n[rep_counts_igg$condition == "hypoxia"] == 3)
)

n_per_tp_igg <- enrichment_igg_rna |> dplyr::count(timepoint) |> pull(n)
stopifnot(
  "IgG/RNA enrichment table must cover all 4 timepoints" =
    length(unique(enrichment_igg_rna$timepoint)) == 4,
  "gene coverage should not vary by more than ~20% across timepoints" =
    (max(n_per_tp_igg) - min(n_per_tp_igg)) / max(n_per_tp_igg) < 0.2
)
cat("IgG/RNA sanity checks passed\n")
```

## Save IgG/RNA enrichment table

```{r save_igg_enrichment}
out_table_igg <- enrichment_igg_rna |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean") |>
  dplyr::select(gene_id_clean, symbol, ip_type, timepoint, baseMean, log2_igg_over_rna, padj)

write_csv(out_table_igg, here("ripseq", "output", "18_igg_over_rna_enrichment.csv"))
cat("Saved:", here("ripseq", "output", "18_igg_over_rna_enrichment.csv"), "\n")
```

---

# Section 2: Load translation categories (3d-promoted TE)

```{r load_te}
# Direction note (CLAUDE.md): despite file name "si3d_vs_sictrl", the DESeq2
# contrast is log2(sictrl/si3d). Positive te_lfc = eIF3d promotes translation.
te_files <- list(
  normoxia_1hr  = here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"),
  hypoxia_1hr   = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  hypoxia_4hr   = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"),
  hypoxia_24hr  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")
)
stopifnot("TE files missing" = all(file.exists(unlist(te_files))))

te_long <- imap_dfr(te_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id, te_lfc, te_padj, symbol) |>
    mutate(
      gene_id_clean = sub("\\..*", "", gene_id),
      timepoint     = tp
    )
}) |>
  mutate(
    timepoint  = factor(timepoint, levels = tp_levels),
    promoted   = te_lfc > 0.5 & !is.na(te_padj) & te_padj < 0.05
  )

cat("TE entries per timepoint:\n")
te_long |> dplyr::count(timepoint) |> print()
cat("3d-promoted per timepoint:\n")
te_long |> dplyr::count(timepoint, promoted) |> dplyr::filter(promoted) |> print()
```

---

# Section 3: Load existing IP/IgG DESeq2 results

```{r load_igg}
# eIF3d: use _no_out2 versions (both outliers removed)
igg_3d_files <- setNames(
  here("ripseq", "output", paste0(
    "ripseq_deseq_", tp_levels, "_flag_eif3d_vs_igg_no_out2.csv"
  )),
  tp_levels
)
stopifnot("eIF3d IP/IgG files missing" = all(file.exists(igg_3d_files)))

igg_3d <- imap_dfr(igg_3d_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, log2FoldChange, padj, symbol) |>
    mutate(ip_type = "flag_eif3d", timepoint = tp)
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

# eIF4E: standard files (no outlier issue)
igg_4e_files <- setNames(
  here("ripseq", "output", paste0(
    "ripseq_deseq_", tp_levels, "_eif4e_vs_igg.csv"
  )),
  tp_levels
)
stopifnot("eIF4E IP/IgG files missing" = all(file.exists(igg_4e_files)))

igg_4e <- imap_dfr(igg_4e_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, log2FoldChange, padj, symbol) |>
    mutate(ip_type = "eif4e", timepoint = tp)
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

igg_long <- bind_rows(igg_3d, igg_4e)
cat("IP/IgG entries:\n")
igg_long |> dplyr::count(ip_type, timepoint) |> print()
```

---

# Section 3b: Correlation helper

```{r cor_helper}
spearman_label <- function(df, x_col, y_col) {
  df |>
    group_by(timepoint) |>
    summarise(
      rho = cor(.data[[x_col]], .data[[y_col]], method = "spearman", use = "complete.obs"),
      pval = tryCatch(
        cor.test(.data[[x_col]], .data[[y_col]], method = "spearman")$p.value,
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) |>
    mutate(label = paste0("rho = ", round(rho, 2),
                          "\np = ", formatC(pval, format = "e", digits = 1)))
}
```

---

# Section 3c: IP/IgG vs IP/RNA correlation (DESeq2)

Directly compare the two enrichment metrics against each other (not against TE) --
sanity check that the two DESeq2 contrasts broadly agree on which genes are IP-enriched.

```{r build_igg_vs_rna_df}
df_igg_vs_rna <- igg_long |>
  filter(!is.na(log2FoldChange)) |>
  dplyr::select(gene_id_clean, ip_type, timepoint, symbol, log2FoldChange) |>
  inner_join(
    enrichment_rna |> dplyr::select(gene_id_clean, ip_type, timepoint, log2_ip_over_rna),
    by = c("gene_id_clean", "ip_type", "timepoint")
  ) |>
  filter(is.finite(log2FoldChange), is.finite(log2_ip_over_rna))

cat("IP/IgG vs IP/RNA joined rows:\n")
df_igg_vs_rna |> dplyr::count(ip_type, timepoint) |> print()
```

```{r fig_3d_igg_vs_rna, fig.width=14, fig.height=4}
df_3d_igg_vs_rna <- df_igg_vs_rna |> filter(ip_type == "flag_eif3d")

cor_3d_igg_vs_rna <- spearman_label(df_3d_igg_vs_rna, "log2FoldChange", "log2_ip_over_rna")
cat("eIF3d IP/IgG vs IP/RNA Spearman:\n"); print(cor_3d_igg_vs_rna)

p_3d_igg_vs_rna <- ggplot(df_3d_igg_vs_rna, aes(x = log2FoldChange, y = log2_ip_over_rna)) +
  geom_point(size = 0.5, alpha = 0.3, color = "grey40") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_3d_igg_vs_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG, DESeq2)",
    y     = "log2FC (eIF3d IP / sictrl RNA, DESeq2)",
    title = "eIF3d: IP/IgG vs IP/RNA enrichment (DESeq2)"
  ) +
  theme_classic(base_size = 12)

p_3d_igg_vs_rna
ggsave(here("ripseq", "plots", "18_eif3d_igg_vs_rna_correlation.pdf"), p_3d_igg_vs_rna,
       width = 14, height = 5)
```

```{r fig_4e_igg_vs_rna, fig.width=14, fig.height=4}
df_4e_igg_vs_rna <- df_igg_vs_rna |> filter(ip_type == "eif4e")

cor_4e_igg_vs_rna <- spearman_label(df_4e_igg_vs_rna, "log2FoldChange", "log2_ip_over_rna")
cat("eIF4E IP/IgG vs IP/RNA Spearman:\n"); print(cor_4e_igg_vs_rna)

p_4e_igg_vs_rna <- ggplot(df_4e_igg_vs_rna, aes(x = log2FoldChange, y = log2_ip_over_rna)) +
  geom_point(size = 0.5, alpha = 0.3, color = "grey40") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_4e_igg_vs_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF4E IP / IgG, DESeq2)",
    y     = "log2FC (eIF4E IP / sictrl RNA, DESeq2)",
    title = "eIF4E: IP/IgG vs IP/RNA enrichment (DESeq2)"
  ) +
  theme_classic(base_size = 12)

p_4e_igg_vs_rna
ggsave(here("ripseq", "plots", "18_eif4e_igg_vs_rna_correlation.pdf"), p_4e_igg_vs_rna,
       width = 14, height = 5)
```

---

# Section 4: Build plot data frames

```{r build_plot_dfs}
# Join IP/RNA enrichment with TE
df_rna <- enrichment_rna |>
  left_join(
    te_long |> dplyr::select(gene_id_clean, timepoint, te_lfc, te_padj, promoted),
    by = c("gene_id_clean", "timepoint")
  ) |>
  filter(!is.na(te_lfc), is.finite(log2_ip_over_rna))

# Join IP/IgG with TE
df_igg <- igg_long |>
  filter(!is.na(log2FoldChange)) |>
  left_join(
    te_long |> dplyr::select(gene_id_clean, timepoint, te_lfc, te_padj, promoted),
    by = c("gene_id_clean", "timepoint")
  ) |>
  filter(!is.na(te_lfc), is.finite(log2FoldChange))
```

---

# Section 6: Figure 1 — eIF3d IP/RNA vs TE (DESeq2)

```{r fig1_3d_rna, fig.width=14, fig.height=4}
df_3d_rna <- df_rna |> filter(ip_type == "flag_eif3d")

cor_3d_rna <- spearman_label(df_3d_rna, "log2_ip_over_rna", "te_lfc")
cat("eIF3d IP/RNA vs TE Spearman:\n"); print(cor_3d_rna)

p_3d_rna <- ggplot(df_3d_rna, aes(x = log2_ip_over_rna, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_3d_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / sictrl RNA, DESeq2)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF3d IP / riboseq sictrl RNA (DESeq2) vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_3d_rna
ggsave(here("ripseq", "plots", "18_eif3d_ip_over_rna_vs_te.pdf"), p_3d_rna,
       width = 14, height = 5)
```

# Figure 1b — eIF4E IP/RNA vs TE

```{r fig1_4e_rna, fig.width=14, fig.height=4}
df_4e_rna <- df_rna |> filter(ip_type == "eif4e")

cor_4e_rna <- spearman_label(df_4e_rna, "log2_ip_over_rna", "te_lfc")
cat("eIF4E IP/RNA vs TE Spearman:\n"); print(cor_4e_rna)

p_4e_rna <- ggplot(df_4e_rna, aes(x = log2_ip_over_rna, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_4e_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF4E IP / sictrl RNA, DESeq2)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF4E IP / riboseq sictrl RNA (DESeq2) vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_4e_rna
ggsave(here("ripseq", "plots", "18_eif4e_ip_over_rna_vs_te.pdf"), p_4e_rna,
       width = 14, height = 5)
```

---

# Section 7: Figure 2 — eIF3d IP/IgG vs TE (DESeq2)

```{r fig2_3d_igg, fig.width=14, fig.height=4}
df_3d_igg <- df_igg |> filter(ip_type == "flag_eif3d")

cor_3d_igg <- spearman_label(df_3d_igg, "log2FoldChange", "te_lfc")
cat("eIF3d IP/IgG vs TE Spearman:\n"); print(cor_3d_igg)

p_3d_igg <- ggplot(df_3d_igg, aes(x = log2FoldChange, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_3d_igg, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG, outliers removed)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF3d IP / IgG vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_3d_igg
ggsave(here("ripseq", "plots", "18_eif3d_ip_over_igg_vs_te.pdf"), p_3d_igg,
       width = 14, height = 5)
```

# Figure 2b — eIF4E IP/IgG vs TE

```{r fig2_4e_igg, fig.width=14, fig.height=4}
df_4e_igg <- df_igg |> filter(ip_type == "eif4e")

cor_4e_igg <- spearman_label(df_4e_igg, "log2FoldChange", "te_lfc")
cat("eIF4E IP/IgG vs TE Spearman:\n"); print(cor_4e_igg)

p_4e_igg <- ggplot(df_4e_igg, aes(x = log2FoldChange, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_4e_igg, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF4E IP / IgG)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF4E IP / IgG vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_4e_igg
ggsave(here("ripseq", "plots", "18_eif4e_ip_over_igg_vs_te.pdf"), p_4e_igg,
       width = 14, height = 5)
```

---

# Summary: Spearman correlations across normalization strategies

```{r summary_table}
cor_summary <- bind_rows(
  cor_3d_rna |> mutate(ip_type = "flag_eif3d", normalization = "IP / riboseq RNA (DESeq2)"),
  cor_4e_rna |> mutate(ip_type = "eif4e",      normalization = "IP / riboseq RNA (DESeq2)"),
  cor_3d_igg |> mutate(ip_type = "flag_eif3d", normalization = "IP / IgG"),
  cor_4e_igg |> mutate(ip_type = "eif4e",      normalization = "IP / IgG")
) |>
  dplyr::select(normalization, ip_type, timepoint, rho, pval)

print(cor_summary)
write_csv(cor_summary, here("ripseq", "output", "18_spearman_normalization_comparison.csv"))
cat("Saved correlation summary.\n")
```

```{r}

```