---
title: "RIP-seq: IP normalized to riboseq sictrl RNA vs IP/IgG — correlation with eIF3d-promoted targets"
author: "Kate"
date: "2026-06"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
```

# Overview

Compare two normalization strategies for RIP-seq enrichment:

1. **IP / riboseq sictrl RNA** (new): log2(IP CPM / spike-in normalized RNA). RNA is
   condition- and timepoint-matched sictrl from the parallel riboseq experiment.
   Note: the RIP-seq has no spike-ins, so IP is library-size normalized (CPM) while
   the denominator uses spike-in normalized RNA counts.

2. **IP / IgG** (existing): log2FC from DESeq2 with outliers removed.

For each strategy, scatter plots show enrichment vs eIF3d-promoted TE (te_lfc from
translation_categories_si3d_vs_sictrl; positive = eIF3d promotes, per CLAUDE.md
direction note).

---

# Load gene annotation

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# A gene is protein-coding if ANY transcript row has biotype "protein_coding".
# distinct() before this would arbitrarily pick one biotype per gene and misclassify
# genes whose first listed transcript is NMD or protein_coding_CDS_not_defined.
pc_genes <- gene_anno_all |>
  filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)

# Deduplicate for symbol lookups; prefer protein_coding row where available
gene_anno <- gene_anno_all |>
  arrange(gene_id_clean, desc(biotype == "protein_coding")) |>
  distinct(gene_id_clean, .keep_all = TRUE) |>
  dplyr::select(gene_id_clean, symbol, biotype)

cat("Protein-coding genes (any-transcript rule):", length(pc_genes), "\n")
cat("Unique genes in annotation:", nrow(gene_anno), "\n")
```

---

# Section 1: Compute IP/RNA enrichment (new normalization)

## Load RIP-seq raw counts

```{r load_rip_counts}
allCounts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
cat("RIP-seq raw counts:", nrow(allCounts), "genes x", ncol(allCounts), "samples\n")

# Remove sample QC fail (NM2026_0077). The IgG PCA outlier (NM2026_0065) is
# not relevant here since we only use IP samples, not IgG, for this normalization.
allCounts <- allCounts[, !grepl("NM2026_0077", colnames(allCounts))]
cat("After outlier removal:", ncol(allCounts), "samples\n")
```

## Compute CPM per sample, then mean per condition

```{r ip_cpm}
# CPM using total counts per sample as size factor
lib_sizes <- colSums(allCounts)
cpm_mat   <- sweep(allCounts, 2, lib_sizes, "/") * 1e6

# Parse sample names into metadata columns
# Column format: {condition}_{ip_type}_{hours}_{rep}_{NM_ID}
# e.g. normoxia_flag_eif3d_1_a_NM2026_0045
cpm_long <- as.data.frame(cpm_mat) |>
  rownames_to_column("gene_id") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pivot_longer(-c(gene_id, gene_id_clean), names_to = "sample_name", values_to = "cpm") |>
  mutate(
    condition = case_when(
      grepl("^normoxia", sample_name) ~ "normoxia",
      grepl("^hypoxia",  sample_name) ~ "hypoxia",
      TRUE ~ NA_character_
    ),
    ip_type = case_when(
      grepl("flag_eif3d", sample_name) ~ "flag_eif3d",
      grepl("eif4e",      sample_name) ~ "eif4e",
      grepl("igg_mock",   sample_name) ~ "igg_mock",
      TRUE ~ NA_character_
    ),
    hours = as.integer(sapply(strsplit(sample_name, "_"), function(p) {
      rep_idx <- which(nchar(p) == 1 & p %in% c("a", "b", "c"))
      p[rep_idx - 1]
    })),
    timepoint = paste0(condition, "_", hours, "hr")
  )

# Average CPM across replicates per gene x ip_type x timepoint
ip_mean_cpm <- cpm_long |>
  filter(ip_type %in% c("flag_eif3d", "eif4e")) |>
  group_by(gene_id_clean, ip_type, condition, hours, timepoint) |>
  summarise(mean_ip_cpm = mean(cpm, na.rm = TRUE), .groups = "drop")

cat("Gene x ip x timepoint rows:", nrow(ip_mean_cpm), "\n")
```

## Load spike-in normalized sictrl RNA

```{r load_rna}
tp_levels <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")

# Helper: load one SPIKE file, extract sictrl rna columns, return long df
load_spike_rna <- function(file_path, timepoint_hours) {
  raw <- read_csv(file_path, show_col_types = FALSE)
  rna_cols <- grep("sictrl.*_rna_", colnames(raw), value = TRUE)
  raw |>
    dplyr::select(gene_id, all_of(rna_cols)) |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    pivot_longer(-c(gene_id, gene_id_clean), names_to = "sample_name", values_to = "rna_norm") |>
    mutate(
      condition = ifelse(grepl("^normoxia", sample_name), "normoxia", "hypoxia"),
      hours     = timepoint_hours,
      timepoint = paste0(condition, "_", hours, "hr")
    ) |>
    group_by(gene_id_clean, condition, hours, timepoint) |>
    summarise(mean_rna_norm = mean(rna_norm, na.rm = TRUE), .groups = "drop")
}

rna_1hr  <- load_spike_rna(here("output", "allCountsNorm_1hr_SPIKE.csv"),  1)
rna_4hr  <- load_spike_rna(here("output", "allCountsNorm_4hr_SPIKE.csv"),  4)
rna_24hr <- load_spike_rna(here("output", "allCountsNorm_24hr_SPIKE.csv"), 24)

# For normoxia RIP-seq (timepoint = normoxia_1hr), use normoxia_1hr RNA
# For hypoxia timepoints, use matching hypoxia RNA
rna_long <- bind_rows(rna_1hr, rna_4hr, rna_24hr) |>
  filter(
    !(condition == "normoxia" & hours != 1)  # keep only normoxia_1hr (no normoxia RIP-seq at 4/24hr)
  )

cat("RNA long rows:", nrow(rna_long), "\n")
cat("Timepoints in RNA data:", paste(unique(rna_long$timepoint), collapse = ", "), "\n")
```

## Compute log2(IP CPM / RNA norm) per gene per condition

```{r ip_over_rna}
joined <- ip_mean_cpm |>
  left_join(rna_long, by = c("gene_id_clean", "condition", "hours", "timepoint"))
cat("After join — genes:", n_distinct(joined$gene_id_clean), "| rows:", nrow(joined), "\n")
cat("Missing RNA (NA):", sum(is.na(joined$mean_rna_norm)), "rows\n")

after_rna_filter <- joined |> filter(!is.na(mean_rna_norm), mean_rna_norm > 0)
cat("After RNA > 0 filter — genes:", n_distinct(after_rna_filter$gene_id_clean),
    "| rows:", nrow(after_rna_filter), "\n")

cat("\nBiotype breakdown of genes before pc filter:\n")
after_rna_filter |>
  distinct(gene_id_clean) |>
  left_join(dplyr::select(gene_anno, gene_id_clean, biotype), by = "gene_id_clean") |>
  dplyr::count(biotype, sort = TRUE) |>
  print()

enrichment_rna <- after_rna_filter |>
  filter(gene_id_clean %in% pc_genes) |>
  mutate(
    log2_ip_over_rna = log2((mean_ip_cpm + 1) / (mean_rna_norm + 1)),
    timepoint = factor(timepoint, levels = tp_levels)
  )
cat("\nAfter pc_genes filter — genes:", n_distinct(enrichment_rna$gene_id_clean),
    "| rows:", nrow(enrichment_rna), "\n")

cat("\nGenes with IP/RNA enrichment per ip_type x timepoint:\n")
enrichment_rna |>
  group_by(ip_type, timepoint) |>
  summarise(n = n(), .groups = "drop") |>
  print()

# Sanity check: EIF3D (the bait) should be in top enriched genes for flag_eif3d
eif3d_check <- enrichment_rna |>
  filter(ip_type == "flag_eif3d", timepoint == "hypoxia_1hr") |>
  arrange(desc(log2_ip_over_rna)) |>
  dplyr::slice(1:10) |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean")
cat("\nTop flag_eif3d IP/RNA enriched genes at hypoxia_1hr:\n")
print(eif3d_check |> dplyr::select(symbol, log2_ip_over_rna))
```

## Save enrichment table

```{r save_enrichment}
out_table <- enrichment_rna |>
  left_join(dplyr::select(gene_anno, gene_id_clean, symbol), by = "gene_id_clean") |>
  dplyr::select(gene_id_clean, symbol, ip_type, timepoint, mean_ip_cpm, mean_rna_norm, log2_ip_over_rna)

write_csv(out_table, here("ripseq", "output", "18_ip_over_rna_enrichment.csv"))
cat("Saved:", here("ripseq", "output", "18_ip_over_rna_enrichment.csv"), "\n")
```

---

# Section 2: Load translation categories (3d-promoted TE)

```{r load_te}
# Direction note (CLAUDE.md): despite file name "si3d_vs_sictrl", the DESeq2
# contrast is log2(sictrl/si3d). Positive te_lfc = eIF3d promotes translation.
te_files <- list(
  normoxia_1hr  = here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"),
  hypoxia_1hr   = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  hypoxia_4hr   = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv"),
  hypoxia_24hr  = here("output", "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv")
)
stopifnot("TE files missing" = all(file.exists(unlist(te_files))))

te_long <- imap_dfr(te_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id, te_lfc, te_padj, symbol) |>
    mutate(
      gene_id_clean = sub("\\..*", "", gene_id),
      timepoint     = tp
    )
}) |>
  mutate(
    timepoint  = factor(timepoint, levels = tp_levels),
    promoted   = te_lfc > 0.5 & !is.na(te_padj) & te_padj < 0.05
  )

cat("TE entries per timepoint:\n")
te_long |> dplyr::count(timepoint) |> print()
cat("3d-promoted per timepoint:\n")
te_long |> dplyr::count(timepoint, promoted) |> dplyr::filter(promoted) |> print()
```

---

# Section 3: Load existing IP/IgG DESeq2 results

```{r load_igg}
# eIF3d: use _no_out2 versions (both outliers removed)
igg_3d_files <- setNames(
  here("ripseq", "output", paste0(
    "ripseq_deseq_", tp_levels, "_flag_eif3d_vs_igg_no_out2.csv"
  )),
  tp_levels
)
stopifnot("eIF3d IP/IgG files missing" = all(file.exists(igg_3d_files)))

igg_3d <- imap_dfr(igg_3d_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, log2FoldChange, padj, symbol) |>
    mutate(ip_type = "flag_eif3d", timepoint = tp)
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

# eIF4E: standard files (no outlier issue)
igg_4e_files <- setNames(
  here("ripseq", "output", paste0(
    "ripseq_deseq_", tp_levels, "_eif4e_vs_igg.csv"
  )),
  tp_levels
)
stopifnot("eIF4E IP/IgG files missing" = all(file.exists(igg_4e_files)))

igg_4e <- imap_dfr(igg_4e_files, function(path, tp) {
  read_csv(path, show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, log2FoldChange, padj, symbol) |>
    mutate(ip_type = "eif4e", timepoint = tp)
}) |>
  mutate(timepoint = factor(timepoint, levels = tp_levels))

igg_long <- bind_rows(igg_3d, igg_4e)
cat("IP/IgG entries:\n")
igg_long |> dplyr::count(ip_type, timepoint) |> print()
```

---

# Section 4: Build plot data frames

```{r build_plot_dfs}
# Join IP/RNA enrichment with TE
df_rna <- enrichment_rna |>
  left_join(
    te_long |> dplyr::select(gene_id_clean, timepoint, te_lfc, te_padj, promoted),
    by = c("gene_id_clean", "timepoint")
  ) |>
  filter(!is.na(te_lfc), is.finite(log2_ip_over_rna))

# Join IP/IgG with TE
df_igg <- igg_long |>
  filter(!is.na(log2FoldChange)) |>
  left_join(
    te_long |> dplyr::select(gene_id_clean, timepoint, te_lfc, te_padj, promoted),
    by = c("gene_id_clean", "timepoint")
  ) |>
  filter(!is.na(te_lfc), is.finite(log2FoldChange))
```

---

# Section 5: Correlation helper

```{r cor_helper}
spearman_label <- function(df, x_col, y_col) {
  df |>
    group_by(timepoint) |>
    summarise(
      rho = cor(.data[[x_col]], .data[[y_col]], method = "spearman", use = "complete.obs"),
      pval = tryCatch(
        cor.test(.data[[x_col]], .data[[y_col]], method = "spearman")$p.value,
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) |>
    mutate(label = paste0("rho = ", round(rho, 2),
                          "\np = ", formatC(pval, format = "e", digits = 1)))
}
```

---

# Section 6: Figure 1 — eIF3d IP/RNA vs TE (new normalization)

```{r fig1_3d_rna, fig.width=14, fig.height=4}
df_3d_rna <- df_rna |> filter(ip_type == "flag_eif3d")

cor_3d_rna <- spearman_label(df_3d_rna, "log2_ip_over_rna", "te_lfc")
cat("eIF3d IP/RNA vs TE Spearman:\n"); print(cor_3d_rna)

p_3d_rna <- ggplot(df_3d_rna, aes(x = log2_ip_over_rna, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_3d_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2(eIF3d IP CPM / sictrl RNA)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF3d IP / riboseq sictrl RNA vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_3d_rna
ggsave(here("ripseq", "plots", "18_eif3d_ip_over_rna_vs_te.pdf"), p_3d_rna,
       width = 14, height = 5)
```

# Figure 1b — eIF4E IP/RNA vs TE

```{r fig1_4e_rna, fig.width=14, fig.height=4}
df_4e_rna <- df_rna |> filter(ip_type == "eif4e")

cor_4e_rna <- spearman_label(df_4e_rna, "log2_ip_over_rna", "te_lfc")
cat("eIF4E IP/RNA vs TE Spearman:\n"); print(cor_4e_rna)

p_4e_rna <- ggplot(df_4e_rna, aes(x = log2_ip_over_rna, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_4e_rna, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2(eIF4E IP CPM / sictrl RNA)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF4E IP / riboseq sictrl RNA vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_4e_rna
ggsave(here("ripseq", "plots", "18_eif4e_ip_over_rna_vs_te.pdf"), p_4e_rna,
       width = 14, height = 5)
```

---

# Section 7: Figure 2 — eIF3d IP/IgG vs TE (existing normalization)

```{r fig2_3d_igg, fig.width=14, fig.height=4}
df_3d_igg <- df_igg |> filter(ip_type == "flag_eif3d")

cor_3d_igg <- spearman_label(df_3d_igg, "log2FoldChange", "te_lfc")
cat("eIF3d IP/IgG vs TE Spearman:\n"); print(cor_3d_igg)

p_3d_igg <- ggplot(df_3d_igg, aes(x = log2FoldChange, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_3d_igg, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF3d IP / IgG, outliers removed)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF3d IP / IgG vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_3d_igg
ggsave(here("ripseq", "plots", "18_eif3d_ip_over_igg_vs_te.pdf"), p_3d_igg,
       width = 14, height = 5)
```

# Figure 2b — eIF4E IP/IgG vs TE

```{r fig2_4e_igg, fig.width=14, fig.height=4}
df_4e_igg <- df_igg |> filter(ip_type == "eif4e")

cor_4e_igg <- spearman_label(df_4e_igg, "log2FoldChange", "te_lfc")
cat("eIF4E IP/IgG vs TE Spearman:\n"); print(cor_4e_igg)

p_4e_igg <- ggplot(df_4e_igg, aes(x = log2FoldChange, y = te_lfc)) +
  geom_point(aes(color = promoted), size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "grey80") +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_text(data = cor_4e_igg, aes(label = label),
            x = Inf, y = Inf, hjust = 1.05, vjust = 1.2,
            inherit.aes = FALSE, size = 3) +
  scale_color_manual(
    values = c("TRUE" = "#E69F00", "FALSE" = "grey70"),
    labels = c("TRUE" = "3d promoted (TE LFC > 0.5, padj < 0.05)", "FALSE" = "Other"),
    name   = NULL
  ) +
  facet_wrap(~ timepoint, nrow = 1) +
  labs(
    x     = "log2FC (eIF4E IP / IgG)",
    y     = "TE LFC (sictrl / si3d)",
    title = "eIF4E IP / IgG vs eIF3d-promoted TE"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

p_4e_igg
ggsave(here("ripseq", "plots", "18_eif4e_ip_over_igg_vs_te.pdf"), p_4e_igg,
       width = 14, height = 5)
```

---

# Summary: Spearman correlations across normalization strategies

```{r summary_table}
cor_summary <- bind_rows(
  cor_3d_rna |> mutate(ip_type = "flag_eif3d", normalization = "IP / riboseq RNA"),
  cor_4e_rna |> mutate(ip_type = "eif4e",      normalization = "IP / riboseq RNA"),
  cor_3d_igg |> mutate(ip_type = "flag_eif3d", normalization = "IP / IgG"),
  cor_4e_igg |> mutate(ip_type = "eif4e",      normalization = "IP / IgG")
) |>
  dplyr::select(normalization, ip_type, timepoint, rho, pval)

print(cor_summary)
write_csv(cor_summary, here("ripseq", "output", "18_spearman_normalization_comparison.csv"))
cat("Saved correlation summary.\n")
```

```{r}

```