---
title: "RIP-seq: eIF3d/eIF4E IP enrichment by gene group (positive / negative control / other)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

CDF plots comparing three gene groups -- the eIF3d TE-promoted positive set, the shared
negative-control set, and all other genes -- across both IP baits (FLAG-eIF3d and eIF4E)
and both enrichment metrics (IP/IgG and IP/RNA), for each RIP-seq timepoint (hypoxia 1hr,
4hr, 24hr; normoxia 1hr) as standalone (non-faceted) plots.

**Both gene sets are evaluated at every timepoint** (not one gene set per timepoint): the
hypoxia-promoted positive set and the normoxia-promoted positive set are each plotted
against every RIP-seq timepoint's data, so e.g. the hypoxia-1hr-defined target set's
IP/IgG enrichment can be inspected at the normoxia_1hr RIP-seq timepoint too, and vice
versa. That's 2 baits x 2 metrics x 4 timepoints x 2 gene sets = 32 plots, plus 4
timepoints x 2 gene sets = 8 additional IgG-mock/RNA background-reference plots (no bait
dimension -- IgG-mock is a single pseudo-bait, not split eIF3d/eIF4E). 40 plots total.

All values are read directly from `output/predictive_modeling/feature_matrix_ripseq.rds`
-- the exact same RIP-seq feature matrix loaded by
`code/predictive_modeling/07_rf_classification_eif3d_targets_no_boruta.Rmd` via
`feature_matrix_path = "feature_matrix_ripseq.rds"` -- so these plots show the identical
values and gene universe used by the predictive model, not an independent re-derivation.

**Gene sets match exactly what's used in predictive modeling.** The hypoxia positive set
is `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` (the standard `condition="hypoxia",
timepoint="1hr"` run). The normoxia positive set is built from the same "1and4hr"
time-regressed DESeq2 model used by `07_rf_classification_eif3d_targets_no_boruta.Rmd`
when `timepoint="1and4hr"`: `translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv`
filtered to `te_lfc > 0.5 & te_padj < 0.05` (see notebook 07's `create_target` chunk for
the identical filter). Each positive set has its own **size-matched** negative-control
subsample -- not the full `negative_control_genes.csv` pool -- reconstructed exactly as
notebook 07's `create_target` chunk does: same `feature_matrix_ripseq.rds` gene universe,
same `set.seed(9)`, same `sample()` call, run once for the hypoxia positive-set size and
once for the normoxia positive-set size. Which gene set is "Positive" (hypoxia-promoted
vs normoxia-promoted) is now an independent choice from which timepoint's RIP-seq data is
being plotted.

**Caveat on the IP/RNA columns** (`*_rna_lfc_*`): RIP-seq (2026 sequencing batch) and
RNA-seq (2025 batch) share no spike-in, so the IP-vs-RNA contrast is fully confounded
with sequencing batch -- DESeq2 normalization cannot separate true IP enrichment from
batch/library-prep composition differences. Treat these columns as noisier / less
directly interpretable than the IP/IgG columns.

**Caveat on the IgG/RNA plots**: `log2_igg_over_rna` is loaded directly from
`ripseq/output/18_igg_over_rna_enrichment.csv` (produced by
`code/18_ip_input_normalization.Rmd`), not from `feature_matrix_ripseq.rds` -- it is a
background/nonspecific-binding reference, not a predictive-modeling feature. It carries
the same RIP-seq (2026) vs RNA-seq (2025) batch-confound caveat as the IP/RNA columns
above (no shared spike-in). It also has one fewer replicate at normoxia_1hr: sample
`NM2026_0077` (a confirmed outlier, normoxia_1hr IgG-mock rep "c") is excluded from the
IgG-mock numerator, leaving 2 IgG-mock replicates at normoxia_1hr vs 3 at every hypoxia
timepoint. `NM2026_0078` (hypoxia_1hr IgG-mock, a different sample) is unaffected and
retained. The gene sets used for these plots are identical to the rest of this notebook
-- no new positive/negative-control sets are defined for IgG/RNA.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Constants

```{r constants}
library(here)
library(tidyverse)

CONDITIONS <- c("hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr", "normoxia_1hr")

# Display label for each RIP-seq timepoint, used in axis labels / plot titles.
timepoint_labels <- c(
  hypoxia_1hr  = "hypoxia 1hr",
  hypoxia_4hr  = "hypoxia 4hr",
  hypoxia_24hr = "hypoxia 24hr",
  normoxia_1hr = "normoxia 1hr"
)

# Which gene set defines "Positive" -- independent of which timepoint's RIP-seq
# data is being plotted (see intro). Both rows are evaluated at every timepoint.
geneset_defs <- tibble::tibble(
  geneset        = c("hypoxia", "normoxia"),
  word           = c("hypoxia-promoted", "normoxia-promoted"),
  positive_label = c("eIF3d hypoxia promotes", "eIF3d normoxia promotes"),
  tag            = c("hypset", "norset")
)

group_colors <- c(
  "Positive"         = "#E69F00",
  "Negative control" = "#000000",  # intentional override of the repo's default
                                    # "grey40" for negative control, per request
  "Other"            = "grey80"
)
```

---

# Section 1: Load RIP-seq feature matrix

```{r load_feature_matrix}
fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Feature matrix loaded:", nrow(fm_ripseq), "unique genes\n")

value_cols <- c(
  paste0("ripseq_eif3d_igg_lfc_", CONDITIONS),
  paste0("eif4e_ip_igg_lfc_", CONDITIONS),
  paste0("ripseq_eif3d_rna_lfc_", CONDITIONS),
  paste0("eif4e_ip_rna_lfc_", CONDITIONS)
)

stopifnot(
  "all 16 required RIP-seq columns must be present" =
    all(value_cols %in% colnames(fm_ripseq)),
  "required columns must be finite where non-NA" = all(sapply(
    value_cols,
    function(col) all(is.finite(fm_ripseq[[col]][!is.na(fm_ripseq[[col]])]))
  ))
)
cat("Feature matrix column checks passed\n")
```

---

# Section 1b: Load IgG/RNA background-reference enrichment

IgG/RNA is a background/nonspecific-binding reference metric, not a predictive-modeling
feature -- it is not part of `feature_matrix_ripseq.rds` and is instead loaded directly
from its long-format CSV output (one row per gene x timepoint, unlike `fm_ripseq`'s one
row per gene).

```{r load_igg_rna}
igg_rna_long <- read_csv(
  here("ripseq", "output", "18_igg_over_rna_enrichment.csv"),
  show_col_types = FALSE
) |>
  mutate(timepoint = factor(timepoint, levels = CONDITIONS))

cat("IgG/RNA enrichment loaded:", nrow(igg_rna_long), "rows across",
    length(unique(igg_rna_long$timepoint)), "timepoints\n")

stopifnot(
  "required columns must be present in igg_rna_long" =
    all(c("gene_id_clean", "timepoint", "log2_igg_over_rna") %in% colnames(igg_rna_long)),
  "log2_igg_over_rna must be finite where non-NA" =
    all(is.finite(igg_rna_long$log2_igg_over_rna[!is.na(igg_rna_long$log2_igg_over_rna)])),
  "all 4 CONDITIONS timepoints must be present" =
    all(CONDITIONS %in% as.character(igg_rna_long$timepoint))
)
cat("IgG/RNA column checks passed\n")
```

---

# Section 2: Load gene sets

```{r load_genesets}
# Hypoxia positive set: the standard hypoxia_1hr TE-promoted set.
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> pull(ensembl_gene)

# Normoxia positive set: built from the same "1and4hr" time-regressed DESeq2 model
# used by 07_rf_classification_eif3d_targets_no_boruta.Rmd when timepoint="1and4hr"
# -- identical filter as that notebook's create_target chunk.
promotes_nor <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) |>
  dplyr::filter(te_lfc > 0.5, te_padj < 0.05) |>
  dplyr::mutate(ensembl_gene = sub("\\..*", "", gene_id)) |>
  pull(ensembl_gene)

negctrl_clean <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pull(gene_id_clean)

# Size-matched negative control: exactly replicates the sampling in notebook 07's
# create_target chunk (feature_set="ripseq"), so "Negative control" here is
# identical to what the RF model actually trains against, not the full
# negative_control_genes.csv pool. Same feature-matrix universe (fm_ripseq),
# same set.seed(9), same sample() call, same order of operations.
neg_in_features <- negctrl_clean[negctrl_clean %in% fm_ripseq$gene_id_clean]

pos_in_features_hyp <- promotes_hyp[promotes_hyp %in% fm_ripseq$gene_id_clean]
number_to_match_hyp <- min(length(pos_in_features_hyp), length(neg_in_features))
set.seed(9)
neg_hyp <- sample(neg_in_features, number_to_match_hyp)

pos_in_features_nor <- promotes_nor[promotes_nor %in% fm_ripseq$gene_id_clean]
number_to_match_nor <- min(length(pos_in_features_nor), length(neg_in_features))
set.seed(9)
neg_nor <- sample(neg_in_features, number_to_match_nor)

# Keyed by gene-set choice ("hypoxia" / "normoxia"), not by RIP-seq timepoint --
# both gene sets are evaluated at every timepoint (see intro).
pos_sets <- list(hypoxia = promotes_hyp, normoxia = promotes_nor)
neg_sets <- list(hypoxia = neg_hyp,      normoxia = neg_nor)

cat("Positive set (hypoxia):", length(promotes_hyp), "genes\n")
cat("Positive set (normoxia, 1and4hr time-regressed):", length(promotes_nor), "genes\n")
cat("Size-matched negative control (hypoxia):", length(neg_hyp), "genes\n")
cat("Size-matched negative control (normoxia):", length(neg_nor), "genes\n")

stopifnot(
  "positive (hypoxia) and negative-control sets must be mutually exclusive" =
    length(intersect(promotes_hyp, neg_hyp)) == 0,
  "positive (normoxia) and negative-control sets must be mutually exclusive" =
    length(intersect(promotes_nor, neg_nor)) == 0
)
cat("Gene set mutual-exclusivity checks passed\n")
```

---

# Section 3: Group-assignment helper

```{r build_group_df}
build_group_df <- function(fm, value_col, geneset) {
  positive_set <- pos_sets[[geneset]]
  negative_set <- neg_sets[[geneset]]
  fm |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::mutate(
      group = dplyr::case_when(
        gene_id_clean %in% positive_set  ~ "Positive",
        gene_id_clean %in% negative_set  ~ "Negative control",
        TRUE                              ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

`geneset` is `"hypoxia"` or `"normoxia"` -- which positive/negative-control pair to use --
independent of `value_col`, which encodes the bait/metric/timepoint of the data being
grouped. For the **eIF4E-bait** panels, "Positive" is still an eIF3d TE-promoted gene set
(never redefined per bait): this tests whether eIF3d-dependent translational targets are
also enriched in the eIF4E (cap-binding) IP, not an eIF4E-specific gene set.

---

# Section 4: Validation

```{r validation_computational}
for (tp in CONDITIONS) {
  for (col in value_cols[grepl(tp, value_cols)]) {
    for (gs in geneset_defs$geneset) {
      d <- build_group_df(fm_ripseq, col, gs)
      stopifnot(
        "group column must have no NA" = !any(is.na(d$group)),
        "Positive group must be non-empty" = sum(d$group == "Positive") > 0,
        "Negative control group must be non-empty" = sum(d$group == "Negative control") > 0,
        "values must be finite" = all(is.finite(d[[col]]))
      )
    }
  }
}
cat("Computational checks passed\n")
```

```{r validation_computational_igg}
for (tp in CONDITIONS) {
  for (gs in geneset_defs$geneset) {
    d <- igg_rna_long |>
      dplyr::filter(timepoint == tp) |>
      build_group_df("log2_igg_over_rna", gs)
    stopifnot(
      "group column must have no NA (IgG/RNA)" = !any(is.na(d$group)),
      "Positive group must be non-empty (IgG/RNA)" = sum(d$group == "Positive") > 0,
      "Negative control group must be non-empty (IgG/RNA)" =
        sum(d$group == "Negative control") > 0,
      "values must be finite (IgG/RNA)" = all(is.finite(d$log2_igg_over_rna))
    )
  }
}
cat("IgG/RNA computational checks passed\n")
```

```{r validation_biological}
eif3d_self <- fm_ripseq |>
  dplyr::filter(symbol == "EIF3D") |>
  pull(ripseq_eif3d_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF3D self-enrichment in flag_eif3d IP/IgG must be strongly positive" =
    length(eif3d_self) > 0 && all(eif3d_self > 2, na.rm = TRUE)
)
cat("EIF3D self-enrichment (hypoxia_1hr, IP/IgG):", eif3d_self, "\n")

eif4e_self <- fm_ripseq |>
  dplyr::filter(symbol == "EIF4E") |>
  pull(eif4e_ip_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF4E self-enrichment in eif4e IP/IgG must be positive" =
    length(eif4e_self) > 0 && all(eif4e_self > 0, na.rm = TRUE)
)
cat("EIF4E self-enrichment (hypoxia_1hr, IP/IgG):", eif4e_self, "\n")

cat("Biological spot-checks passed\n")
```

---

# Section 5: Plotting helper

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, positive_label, x_label, plot_title, out_name,
                           width = 7, height = 5.5) {
  n_labels <- c(
    "Positive"         = paste0(positive_label, " (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = c(-2, 2)) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 6: Generate plots

Each subsection loops over all 4 RIP-seq timepoints x both gene sets (hypoxia-promoted,
normoxia-promoted), producing 8 plots per bait/metric combination (4 for IgG/RNA, which
has no bait dimension). Output filenames follow
`19_cdf_<bait>_<metric>_<timepoint>_<hypset|norset>.pdf`. Every plot object is also
stashed in `plot_registry` (keyed `<bait>_<metric>_<timepoint>_<geneset>`, e.g.
`"eif3d_igg_hypoxia_1hr_hypoxia"`), used in Section 7 to assemble composite grids.

```{r plot_registry_init}
plot_registry <- list()
```

## eIF3d IP/IgG

```{r plots_eif3d_igg}
for (tp in CONDITIONS) {
  for (i in seq_len(nrow(geneset_defs))) {
    value_col <- paste0("ripseq_eif3d_igg_lfc_", tp)
    df <- build_group_df(fm_ripseq, value_col, geneset_defs$geneset[i])
    p <- make_cdf_plot(
      df, value_col,
      positive_label = geneset_defs$positive_label[i],
      x_label    = paste0("log2FC (eIF3d IP / IgG, ", timepoint_labels[[tp]], ")"),
      plot_title = paste0("eIF3d ", geneset_defs$word[i], " targets: IP/IgG CDF (",
                           timepoint_labels[[tp]], ")"),
      out_name   = paste0("19_cdf_eif3d_igg_", tp, "_", geneset_defs$tag[i], ".pdf")
    )
    plot_registry[[paste("eif3d", "igg", tp, geneset_defs$geneset[i], sep = "_")]] <- p
    print(p)
  }
}
```

## eIF3d IP/RNA

```{r plots_eif3d_iprna}
for (tp in CONDITIONS) {
  for (i in seq_len(nrow(geneset_defs))) {
    value_col <- paste0("ripseq_eif3d_rna_lfc_", tp)
    df <- build_group_df(fm_ripseq, value_col, geneset_defs$geneset[i])
    p <- make_cdf_plot(
      df, value_col,
      positive_label = geneset_defs$positive_label[i],
      x_label    = paste0("log2FC (eIF3d IP / sictrl RNA, DESeq2, ", timepoint_labels[[tp]], ")"),
      plot_title = paste0("eIF3d ", geneset_defs$word[i], " targets: IP/RNA CDF (",
                           timepoint_labels[[tp]], ")"),
      out_name   = paste0("19_cdf_eif3d_iprna_", tp, "_", geneset_defs$tag[i], ".pdf")
    )
    plot_registry[[paste("eif3d", "rna", tp, geneset_defs$geneset[i], sep = "_")]] <- p
    print(p)
  }
}
```

## eIF4E IP/IgG

```{r plots_eif4e_igg}
for (tp in CONDITIONS) {
  for (i in seq_len(nrow(geneset_defs))) {
    value_col <- paste0("eif4e_ip_igg_lfc_", tp)
    df <- build_group_df(fm_ripseq, value_col, geneset_defs$geneset[i])
    p <- make_cdf_plot(
      df, value_col,
      positive_label = geneset_defs$positive_label[i],
      x_label    = paste0("log2FC (eIF4E IP / IgG, ", timepoint_labels[[tp]], ")"),
      plot_title = paste0("eIF3d ", geneset_defs$word[i], " targets in eIF4E IP: IP/IgG CDF (",
                           timepoint_labels[[tp]], ")"),
      out_name   = paste0("19_cdf_eif4e_igg_", tp, "_", geneset_defs$tag[i], ".pdf")
    )
    plot_registry[[paste("eif4e", "igg", tp, geneset_defs$geneset[i], sep = "_")]] <- p
    print(p)
  }
}
```

## eIF4E IP/RNA

```{r plots_eif4e_iprna}
for (tp in CONDITIONS) {
  for (i in seq_len(nrow(geneset_defs))) {
    value_col <- paste0("eif4e_ip_rna_lfc_", tp)
    df <- build_group_df(fm_ripseq, value_col, geneset_defs$geneset[i])
    p <- make_cdf_plot(
      df, value_col,
      positive_label = geneset_defs$positive_label[i],
      x_label    = paste0("log2FC (eIF4E IP / sictrl RNA, DESeq2, ", timepoint_labels[[tp]], ")"),
      plot_title = paste0("eIF3d ", geneset_defs$word[i], " targets in eIF4E IP: IP/RNA CDF (",
                           timepoint_labels[[tp]], ")"),
      out_name   = paste0("19_cdf_eif4e_iprna_", tp, "_", geneset_defs$tag[i], ".pdf")
    )
    plot_registry[[paste("eif4e", "rna", tp, geneset_defs$geneset[i], sep = "_")]] <- p
    print(p)
  }
}
```

## IgG-mock / RNA (background reference)

No bait dimension here -- IgG-mock is a single pseudo-bait shared across baits, not
split eIF3d/eIF4E. "Positive" is still an eIF3d TE-promoted gene set (same convention as
the eIF4E panels above): this tests whether eIF3d-dependent translational targets show
any background/nonspecific enrichment in the IgG-mock pulldown, which they should not if
IP/IgG and IP/RNA enrichment in the bait IPs reflects real binding rather than shared
RNA-abundance artifacts. The normoxia_1hr panels' x-axis labels flag the 2-replicate
caveat.

```{r plots_igg_rna}
for (tp in CONDITIONS) {
  tp_label_x <- timepoint_labels[[tp]]
  if (tp == "normoxia_1hr") tp_label_x <- paste0(tp_label_x, "; n=2 IgG-mock reps")

  for (i in seq_len(nrow(geneset_defs))) {
    df <- igg_rna_long |>
      dplyr::filter(timepoint == tp) |>
      build_group_df("log2_igg_over_rna", geneset_defs$geneset[i])
    p <- make_cdf_plot(
      df, "log2_igg_over_rna",
      positive_label = geneset_defs$positive_label[i],
      x_label    = paste0("log2FC (IgG-mock / sictrl RNA, DESeq2, ", tp_label_x, ")"),
      plot_title = paste0("eIF3d ", geneset_defs$word[i],
                           " targets in IgG-mock background: IgG/RNA CDF (",
                           timepoint_labels[[tp]], ")"),
      out_name   = paste0("19_cdf_igg_rna_", tp, "_", geneset_defs$tag[i], ".pdf")
    )
    plot_registry[[paste("iggrna", tp, geneset_defs$geneset[i], sep = "_")]] <- p
    print(p)
  }
}
```

---

# Section 7: Composite grids (patchwork)

Assembles the individual plots above into composite figures, one per bait x gene-set x
metric combination (single metric row -- IP/IgG or IP/RNA -- 4 timepoint columns ordered
normoxia, hypoxia 1hr, 4hr, 24hr), plus separate single-row IgG-mock/RNA background
composites per gene set (bait-independent, so only 2 needed, not 4). IP/IgG and IP/RNA
are rendered as separate figures (rather than stacked rows of one figure) since they use
different axis scales and are usually consulted independently. To keep each panel legible
in a 4-column grid, panels here drop their individual title/legend/verbose axis label
(redundant once there's a column header and row label) and keep only the curve + the
on-plot n= annotation; one shared legend (color mapping only, no per-panel n=, since
that's already shown per-panel) sits at the bottom of the whole figure.

```{r composite_helper}
library(patchwork)

# Display column order: normoxia first, then hypoxia 1/4/24hr.
col_order <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")
col_headers <- c("Norm", "Hyp - 1", "Hyp - 4", "Hyp - 24")

make_text_panel <- function(label, angle = 0, size = 5) {
  ggplot() +
    theme_void() +
    annotate("text", x = 0, y = 0, label = label, size = size, fontface = "bold", angle = angle)
}

# Strip per-panel clutter for composite use: no title, short generic x-label
# (the column header + row label already convey timepoint/metric), no legend
# (collected once at the bottom instead). The on-plot n= annotation is left
# untouched since it's compact and genuinely varies panel to panel.
strip_for_composite <- function(p) {
  p + labs(title = NULL, x = "log2FC") + theme(legend.position = "none")
}

make_composite_grid <- function(bait, geneset, title, out_name, metrics = c("igg", "rna")) {
  if (bait == "iggrna") {
    row_labels  <- "IgG/RNA"
    row_metrics <- list(NULL)
  } else {
    metric_labels <- c(igg = "IP/IgG", rna = "IP/RNA")
    row_labels  <- unname(metric_labels[metrics])
    row_metrics <- as.list(metrics)
  }
  n_rows <- length(row_labels)

  row_keys <- lapply(row_metrics, function(metric) {
    if (is.null(metric)) paste("iggrna", col_order, geneset, sep = "_")
    else paste(bait, metric, col_order, geneset, sep = "_")
  })
  stopifnot("all requested plots must exist in plot_registry" =
    all(unlist(row_keys) %in% names(plot_registry)))

  # Flat, row-major list of grid cells: blank corner + 4 column headers, then
  # for each metric row: row label + 4 stripped panels. wrap_plots(ncol=5,...)
  # lays this out as a (1 + n_rows) x 5 grid with no cell-spanning needed.
  cells <- c(list(make_text_panel("")), lapply(col_headers, make_text_panel))
  for (r in seq_len(n_rows)) {
    cells <- c(cells,
               list(make_text_panel(row_labels[r], angle = 90)),
               lapply(plot_registry[row_keys[[r]]], strip_for_composite))
  }

  main_grid <- patchwork::wrap_plots(
    cells, ncol = 5,
    widths  = c(0.15, 1, 1, 1, 1),
    heights = c(0.15, rep(1, n_rows))
  )

  # One shared legend (plain Positive/Negative control/Other labels, no n=,
  # since exact n= is already shown per-panel in the on-plot text).
  legend_src <- plot_registry[[row_keys[[1]][1]]] +
    scale_color_manual(values = group_colors, breaks = names(group_colors)) +
    theme(legend.position = "bottom")
  legend_grob <- cowplot::get_legend(legend_src)

  full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
    patchwork::plot_layout(heights = c(1, 0.06)) +
    patchwork::plot_annotation(title = title)

  ggsave(here("ripseq", "plots", out_name), full,
         width = 4 * 4.2 + 1, height = n_rows * 4 + 2, limitsize = FALSE)
  full
}
```

```{r composite_eif3d_hypset_igg, fig.width=18, fig.height=5}
make_composite_grid(
  "eif3d", "hypoxia", metrics = "igg",
  title    = "eIF3d IP/IgG: eIF3d hypoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif3d_hypset_igg.pdf"
)
```

```{r composite_eif3d_hypset_rna, fig.width=18, fig.height=5}
make_composite_grid(
  "eif3d", "hypoxia", metrics = "rna",
  title    = "eIF3d IP/RNA: eIF3d hypoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif3d_hypset_rna.pdf"
)
```

```{r composite_eif3d_norset_igg, fig.width=18, fig.height=5}
make_composite_grid(
  "eif3d", "normoxia", metrics = "igg",
  title    = "eIF3d IP/IgG: eIF3d normoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif3d_norset_igg.pdf"
)
```

```{r composite_eif3d_norset_rna, fig.width=18, fig.height=5}
make_composite_grid(
  "eif3d", "normoxia", metrics = "rna",
  title    = "eIF3d IP/RNA: eIF3d normoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif3d_norset_rna.pdf"
)
```

```{r composite_eif4e_hypset_igg, fig.width=18, fig.height=5}
make_composite_grid(
  "eif4e", "hypoxia", metrics = "igg",
  title    = "eIF4E IP/IgG: eIF3d hypoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif4e_hypset_igg.pdf"
)
```

```{r composite_eif4e_hypset_rna, fig.width=18, fig.height=5}
make_composite_grid(
  "eif4e", "hypoxia", metrics = "rna",
  title    = "eIF4E IP/RNA: eIF3d hypoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif4e_hypset_rna.pdf"
)
```

```{r composite_eif4e_norset_igg, fig.width=18, fig.height=5}
make_composite_grid(
  "eif4e", "normoxia", metrics = "igg",
  title    = "eIF4E IP/IgG: eIF3d normoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif4e_norset_igg.pdf"
)
```

```{r composite_eif4e_norset_rna, fig.width=18, fig.height=5}
make_composite_grid(
  "eif4e", "normoxia", metrics = "rna",
  title    = "eIF4E IP/RNA: eIF3d normoxia-promoted targets, all timepoints",
  out_name = "19_grid_eif4e_norset_rna.pdf"
)
```

```{r composite_iggrna_hypset, fig.width=18, fig.height=5}
make_composite_grid(
  "iggrna", "hypoxia",
  title    = "IgG-mock background: eIF3d hypoxia-promoted targets, all timepoints",
  out_name = "19_grid_iggrna_hypset.pdf"
)
```

```{r composite_iggrna_norset, fig.width=18, fig.height=5}
make_composite_grid(
  "iggrna", "normoxia",
  title    = "IgG-mock background: eIF3d normoxia-promoted targets, all timepoints",
  out_name = "19_grid_iggrna_norset.pdf"
)
```
