---
title: "RIP-seq: eIF3d/eIF4E IP enrichment by gene group (positive / negative control / other)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

CDF plots comparing three gene groups -- the eIF3d TE-promoted positive set,
the shared negative-control set, and all other genes -- across both IP baits
(FLAG-eIF3d and eIF4E) and both enrichment metrics (IP/IgG and IP/RNA), for each
hypoxia RIP-seq timepoint (1hr, 4hr, 24hr) and normoxia (1hr) separately as
standalone (non-faceted) plots. That's 2 baits x 2 metrics x 4 timepoints = 16 plots,
plus 4 additional IgG-mock/RNA background-reference plots (no bait dimension -- IgG-mock
is a single pseudo-bait, not split eIF3d/eIF4E), one per timepoint. 20 plots total.

All values are read directly from `output/predictive_modeling/feature_matrix_ripseq.rds`
-- the exact same RIP-seq feature matrix loaded by
`code/predictive_modeling/07_rf_classification_eif3d_targets_no_boruta.Rmd` via
`feature_matrix_path = "feature_matrix_ripseq.rds"` -- so these plots show the
identical values and gene universe used by the predictive model, not an independent
re-derivation.

**Gene sets match exactly what's used in predictive modeling** (not per-RIP-seq-timepoint
sets): the hypoxia positive set is `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` (the
standard `condition="hypoxia", timepoint="1hr"` run), applied uniformly across all
three hypoxia RIP-seq timepoints (1hr, 4hr, 24hr) -- i.e. we're asking "where do
IP enrichment values for the same fixed hypoxia_1hr-defined target set fall at each
RIP-seq timepoint," not redefining the target set per RIP-seq timepoint. The normoxia
positive set is built from the same "1and4hr" time-regressed DESeq2 model used by
`07_rf_classification_eif3d_targets_no_boruta.Rmd` when `timepoint="1and4hr"`:
`translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv` filtered to
`te_lfc > 0.5 & te_padj < 0.05` (see notebook 07's `create_target` chunk for the
identical filter). The negative-control set is the **size-matched** random subsample
notebook 07 actually trains against -- not the full `negative_control_genes.csv`
pool. This is reconstructed exactly: same `feature_matrix_ripseq.rds` gene universe,
same `set.seed(9)`, same `sample()` call as notebook 07's `create_target` chunk, run
once for the hypoxia positive-set size and once for the normoxia positive-set size
(the two conditions draw different-sized subsamples since their positive sets differ
in size).

**Caveat on the IP/RNA columns** (`*_rna_lfc_*`): RIP-seq (2026 sequencing batch) and
RNA-seq (2025 batch) share no spike-in, so the IP-vs-RNA contrast is fully confounded
with sequencing batch -- DESeq2 normalization cannot separate true IP enrichment from
batch/library-prep composition differences. Treat these columns as noisier / less
directly interpretable than the IP/IgG columns.

**Caveat on the IgG/RNA plots**: `log2_igg_over_rna` is loaded directly from
`ripseq/output/18_igg_over_rna_enrichment.csv` (produced by
`code/18_ip_input_normalization.Rmd`), not from `feature_matrix_ripseq.rds` -- it is a
background/nonspecific-binding reference, not a predictive-modeling feature. It carries
the same RIP-seq (2026) vs RNA-seq (2025) batch-confound caveat as the IP/RNA columns
above (no shared spike-in). It also has one fewer replicate at normoxia_1hr: sample
`NM2026_0077` (a confirmed outlier, normoxia_1hr IgG-mock rep "c") is excluded from the
IgG-mock numerator, leaving 2 IgG-mock replicates at normoxia_1hr vs 3 at every hypoxia
timepoint. `NM2026_0078` (hypoxia_1hr IgG-mock, a different sample) is unaffected and
retained. The gene sets used for these plots are identical to the rest of this notebook
-- no new positive/negative-control sets are defined for IgG/RNA.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Constants

```{r constants}
library(here)
library(tidyverse)

CONDITIONS <- c("hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr", "normoxia_1hr")

group_colors <- c(
  "Positive"         = "#E69F00",
  "Negative control" = "#000000",  # intentional override of the repo's default
                                    # "grey40" for negative control, per request
  "Other"            = "grey80"
)
```

---

# Section 1: Load RIP-seq feature matrix

```{r load_feature_matrix}
fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Feature matrix loaded:", nrow(fm_ripseq), "unique genes\n")

value_cols <- c(
  paste0("ripseq_eif3d_igg_lfc_", CONDITIONS),
  paste0("eif4e_ip_igg_lfc_", CONDITIONS),
  paste0("ripseq_eif3d_rna_lfc_", CONDITIONS),
  paste0("eif4e_ip_rna_lfc_", CONDITIONS)
)

stopifnot(
  "all 16 required RIP-seq columns must be present" =
    all(value_cols %in% colnames(fm_ripseq)),
  "required columns must be finite where non-NA" = all(sapply(
    value_cols,
    function(col) all(is.finite(fm_ripseq[[col]][!is.na(fm_ripseq[[col]])]))
  ))
)
cat("Feature matrix column checks passed\n")
```

---

# Section 1b: Load IgG/RNA background-reference enrichment

IgG/RNA is a background/nonspecific-binding reference metric, not a predictive-modeling
feature -- it is not part of `feature_matrix_ripseq.rds` and is instead loaded directly
from its long-format CSV output (one row per gene x timepoint, unlike `fm_ripseq`'s one
row per gene).

```{r load_igg_rna}
igg_rna_long <- read_csv(
  here("ripseq", "output", "18_igg_over_rna_enrichment.csv"),
  show_col_types = FALSE
) |>
  mutate(timepoint = factor(timepoint, levels = CONDITIONS))

cat("IgG/RNA enrichment loaded:", nrow(igg_rna_long), "rows across",
    length(unique(igg_rna_long$timepoint)), "timepoints\n")

stopifnot(
  "required columns must be present in igg_rna_long" =
    all(c("gene_id_clean", "timepoint", "log2_igg_over_rna") %in% colnames(igg_rna_long)),
  "log2_igg_over_rna must be finite where non-NA" =
    all(is.finite(igg_rna_long$log2_igg_over_rna[!is.na(igg_rna_long$log2_igg_over_rna)])),
  "all 4 CONDITIONS timepoints must be present" =
    all(CONDITIONS %in% as.character(igg_rna_long$timepoint))
)
cat("IgG/RNA column checks passed\n")
```

---

# Section 2: Load gene sets

```{r load_genesets}
# Hypoxia positive set: the standard hypoxia_1hr TE-promoted set, reused as-is
# across all hypoxia RIP-seq timepoints (matches predictive modeling; we are not
# redefining the target set per RIP-seq timepoint).
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> pull(ensembl_gene)

# Normoxia positive set: built from the same "1and4hr" time-regressed DESeq2 model
# used by 07_rf_classification_eif3d_targets_no_boruta.Rmd when timepoint="1and4hr"
# -- identical filter as that notebook's create_target chunk.
promotes_nor <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) |>
  dplyr::filter(te_lfc > 0.5, te_padj < 0.05) |>
  dplyr::mutate(ensembl_gene = sub("\\..*", "", gene_id)) |>
  pull(ensembl_gene)

negctrl_clean <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pull(gene_id_clean)

# Size-matched negative control: exactly replicates the sampling in notebook 07's
# create_target chunk (feature_set="ripseq"), so "Negative control" here is
# identical to what the RF model actually trains against, not the full
# negative_control_genes.csv pool. Same feature-matrix universe (fm_ripseq),
# same set.seed(9), same sample() call, same order of operations.
neg_in_features <- negctrl_clean[negctrl_clean %in% fm_ripseq$gene_id_clean]

pos_in_features_hyp <- promotes_hyp[promotes_hyp %in% fm_ripseq$gene_id_clean]
number_to_match_hyp <- min(length(pos_in_features_hyp), length(neg_in_features))
set.seed(9)
neg_hyp <- sample(neg_in_features, number_to_match_hyp)

pos_in_features_nor <- promotes_nor[promotes_nor %in% fm_ripseq$gene_id_clean]
number_to_match_nor <- min(length(pos_in_features_nor), length(neg_in_features))
set.seed(9)
neg_nor <- sample(neg_in_features, number_to_match_nor)

# Same hypoxia positive/negative sets apply to all three hypoxia RIP-seq
# timepoints; normoxia_1hr is the only RIP-seq normoxia timepoint.
pos_sets <- list(
  hypoxia_1hr  = promotes_hyp,
  hypoxia_4hr  = promotes_hyp,
  hypoxia_24hr = promotes_hyp,
  normoxia_1hr = promotes_nor
)
neg_sets <- list(
  hypoxia_1hr  = neg_hyp,
  hypoxia_4hr  = neg_hyp,
  hypoxia_24hr = neg_hyp,
  normoxia_1hr = neg_nor
)

cat("Positive set (hypoxia, all timepoints):", length(promotes_hyp), "genes\n")
cat("Positive set (normoxia, 1and4hr time-regressed):", length(promotes_nor), "genes\n")
cat("Size-matched negative control (hypoxia):", length(neg_hyp), "genes\n")
cat("Size-matched negative control (normoxia):", length(neg_nor), "genes\n")

stopifnot(
  "positive (hypoxia) and negative-control sets must be mutually exclusive" =
    length(intersect(promotes_hyp, neg_hyp)) == 0,
  "positive (normoxia) and negative-control sets must be mutually exclusive" =
    length(intersect(promotes_nor, neg_nor)) == 0
)
cat("Gene set mutual-exclusivity checks passed\n")
```

---

# Section 3: Group-assignment helper

```{r build_group_df}
build_group_df <- function(fm, value_col, timepoint_val) {
  positive_set <- pos_sets[[timepoint_val]]
  negative_set <- neg_sets[[timepoint_val]]
  fm |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::mutate(
      group = dplyr::case_when(
        gene_id_clean %in% positive_set  ~ "Positive",
        gene_id_clean %in% negative_set  ~ "Negative control",
        TRUE                              ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

For the **eIF4E-bait** panels, "Positive" is still the eIF3d TE-promoted gene set --
this tests whether eIF3d-dependent translational targets are also enriched in the
eIF4E (cap-binding) IP, not an eIF4E-specific gene set.

---

# Section 4: Validation

```{r validation_computational}
for (tp in CONDITIONS) {
  for (col in value_cols[grepl(tp, value_cols)]) {
    d <- build_group_df(fm_ripseq, col, tp)
    stopifnot(
      "group column must have no NA" = !any(is.na(d$group)),
      "Positive group must be non-empty" = sum(d$group == "Positive") > 0,
      "Negative control group must be non-empty" = sum(d$group == "Negative control") > 0,
      "values must be finite" = all(is.finite(d[[col]]))
    )
  }
}
cat("Computational checks passed\n")
```

```{r validation_computational_igg}
for (tp in CONDITIONS) {
  d <- igg_rna_long |>
    dplyr::filter(timepoint == tp) |>
    build_group_df("log2_igg_over_rna", tp)
  stopifnot(
    "group column must have no NA (IgG/RNA)" = !any(is.na(d$group)),
    "Positive group must be non-empty (IgG/RNA)" = sum(d$group == "Positive") > 0,
    "Negative control group must be non-empty (IgG/RNA)" =
      sum(d$group == "Negative control") > 0,
    "values must be finite (IgG/RNA)" = all(is.finite(d$log2_igg_over_rna))
  )
}
cat("IgG/RNA computational checks passed\n")
```

```{r validation_biological}
eif3d_self <- fm_ripseq |>
  dplyr::filter(symbol == "EIF3D") |>
  pull(ripseq_eif3d_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF3D self-enrichment in flag_eif3d IP/IgG must be strongly positive" =
    length(eif3d_self) > 0 && all(eif3d_self > 2, na.rm = TRUE)
)
cat("EIF3D self-enrichment (hypoxia_1hr, IP/IgG):", eif3d_self, "\n")

eif4e_self <- fm_ripseq |>
  dplyr::filter(symbol == "EIF4E") |>
  pull(eif4e_ip_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF4E self-enrichment in eif4e IP/IgG must be positive" =
    length(eif4e_self) > 0 && all(eif4e_self > 0, na.rm = TRUE)
)
cat("EIF4E self-enrichment (hypoxia_1hr, IP/IgG):", eif4e_self, "\n")

cat("Biological spot-checks passed\n")
```

---

# Section 5: Plotting helper

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, positive_label, x_label, plot_title, out_name,
                           width = 7, height = 5.5) {
  n_labels <- c(
    "Positive"         = paste0(positive_label, " (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = c(-2, 2)) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 6: Generate plots

## eIF3d IP/IgG

```{r plot_eif3d_igg_hypoxia_1hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_igg_lfc_hypoxia_1hr", "hypoxia_1hr")
make_cdf_plot(
  df, "ripseq_eif3d_igg_lfc_hypoxia_1hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / IgG, hypoxia 1hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/IgG CDF (hypoxia 1hr)",
  out_name   = "19_cdf_eif3d_igg_hypoxia_1hr.pdf"
)
```

```{r plot_eif3d_igg_hypoxia_4hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_igg_lfc_hypoxia_4hr", "hypoxia_4hr")
make_cdf_plot(
  df, "ripseq_eif3d_igg_lfc_hypoxia_4hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / IgG, hypoxia 4hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/IgG CDF (hypoxia 4hr)",
  out_name   = "19_cdf_eif3d_igg_hypoxia_4hr.pdf"
)
```

```{r plot_eif3d_igg_hypoxia_24hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_igg_lfc_hypoxia_24hr", "hypoxia_24hr")
make_cdf_plot(
  df, "ripseq_eif3d_igg_lfc_hypoxia_24hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / IgG, hypoxia 24hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/IgG CDF (hypoxia 24hr)",
  out_name   = "19_cdf_eif3d_igg_hypoxia_24hr.pdf"
)
```

```{r plot_eif3d_igg_normoxia_1hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_igg_lfc_normoxia_1hr", "normoxia_1hr")
make_cdf_plot(
  df, "ripseq_eif3d_igg_lfc_normoxia_1hr",
  positive_label = "eIF3d normoxia promotes",
  x_label    = "log2FC (eIF3d IP / IgG, normoxia 1hr)",
  plot_title = "eIF3d normoxia-promoted targets: IP/IgG CDF (normoxia 1hr)",
  out_name   = "19_cdf_eif3d_igg_normoxia_1hr.pdf"
)
```

## eIF3d IP/RNA

```{r plot_eif3d_iprna_hypoxia_1hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_rna_lfc_hypoxia_1hr", "hypoxia_1hr")
make_cdf_plot(
  df, "ripseq_eif3d_rna_lfc_hypoxia_1hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / sictrl RNA, DESeq2, hypoxia 1hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/RNA CDF (hypoxia 1hr)",
  out_name   = "19_cdf_eif3d_iprna_hypoxia_1hr.pdf"
)
```

```{r plot_eif3d_iprna_hypoxia_4hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_rna_lfc_hypoxia_4hr", "hypoxia_4hr")
make_cdf_plot(
  df, "ripseq_eif3d_rna_lfc_hypoxia_4hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / sictrl RNA, DESeq2, hypoxia 4hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/RNA CDF (hypoxia 4hr)",
  out_name   = "19_cdf_eif3d_iprna_hypoxia_4hr.pdf"
)
```

```{r plot_eif3d_iprna_hypoxia_24hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_rna_lfc_hypoxia_24hr", "hypoxia_24hr")
make_cdf_plot(
  df, "ripseq_eif3d_rna_lfc_hypoxia_24hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF3d IP / sictrl RNA, DESeq2, hypoxia 24hr)",
  plot_title = "eIF3d hypoxia-promoted targets: IP/RNA CDF (hypoxia 24hr)",
  out_name   = "19_cdf_eif3d_iprna_hypoxia_24hr.pdf"
)
```

```{r plot_eif3d_iprna_normoxia_1hr}
df <- build_group_df(fm_ripseq, "ripseq_eif3d_rna_lfc_normoxia_1hr", "normoxia_1hr")
make_cdf_plot(
  df, "ripseq_eif3d_rna_lfc_normoxia_1hr",
  positive_label = "eIF3d normoxia promotes",
  x_label    = "log2FC (eIF3d IP / sictrl RNA, DESeq2, normoxia 1hr)",
  plot_title = "eIF3d normoxia-promoted targets: IP/RNA CDF (normoxia 1hr)",
  out_name   = "19_cdf_eif3d_iprna_normoxia_1hr.pdf"
)
```

## eIF4E IP/IgG

```{r plot_eif4e_igg_hypoxia_1hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_igg_lfc_hypoxia_1hr", "hypoxia_1hr")
make_cdf_plot(
  df, "eif4e_ip_igg_lfc_hypoxia_1hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / IgG, hypoxia 1hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/IgG CDF (hypoxia 1hr)",
  out_name   = "19_cdf_eif4e_igg_hypoxia_1hr.pdf"
)
```

```{r plot_eif4e_igg_hypoxia_4hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_igg_lfc_hypoxia_4hr", "hypoxia_4hr")
make_cdf_plot(
  df, "eif4e_ip_igg_lfc_hypoxia_4hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / IgG, hypoxia 4hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/IgG CDF (hypoxia 4hr)",
  out_name   = "19_cdf_eif4e_igg_hypoxia_4hr.pdf"
)
```

```{r plot_eif4e_igg_hypoxia_24hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_igg_lfc_hypoxia_24hr", "hypoxia_24hr")
make_cdf_plot(
  df, "eif4e_ip_igg_lfc_hypoxia_24hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / IgG, hypoxia 24hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/IgG CDF (hypoxia 24hr)",
  out_name   = "19_cdf_eif4e_igg_hypoxia_24hr.pdf"
)
```

```{r plot_eif4e_igg_normoxia_1hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_igg_lfc_normoxia_1hr", "normoxia_1hr")
make_cdf_plot(
  df, "eif4e_ip_igg_lfc_normoxia_1hr",
  positive_label = "eIF3d normoxia promotes",
  x_label    = "log2FC (eIF4E IP / IgG, normoxia 1hr)",
  plot_title = "eIF3d normoxia-promoted targets in eIF4E IP: IP/IgG CDF (normoxia 1hr)",
  out_name   = "19_cdf_eif4e_igg_normoxia_1hr.pdf"
)
```

## eIF4E IP/RNA

```{r plot_eif4e_iprna_hypoxia_1hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_rna_lfc_hypoxia_1hr", "hypoxia_1hr")
make_cdf_plot(
  df, "eif4e_ip_rna_lfc_hypoxia_1hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / sictrl RNA, DESeq2, hypoxia 1hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/RNA CDF (hypoxia 1hr)",
  out_name   = "19_cdf_eif4e_iprna_hypoxia_1hr.pdf"
)
```

```{r plot_eif4e_iprna_hypoxia_4hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_rna_lfc_hypoxia_4hr", "hypoxia_4hr")
make_cdf_plot(
  df, "eif4e_ip_rna_lfc_hypoxia_4hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / sictrl RNA, DESeq2, hypoxia 4hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/RNA CDF (hypoxia 4hr)",
  out_name   = "19_cdf_eif4e_iprna_hypoxia_4hr.pdf"
)
```

```{r plot_eif4e_iprna_hypoxia_24hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_rna_lfc_hypoxia_24hr", "hypoxia_24hr")
make_cdf_plot(
  df, "eif4e_ip_rna_lfc_hypoxia_24hr",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (eIF4E IP / sictrl RNA, DESeq2, hypoxia 24hr)",
  plot_title = "eIF3d hypoxia-promoted targets in eIF4E IP: IP/RNA CDF (hypoxia 24hr)",
  out_name   = "19_cdf_eif4e_iprna_hypoxia_24hr.pdf"
)
```

```{r plot_eif4e_iprna_normoxia_1hr}
df <- build_group_df(fm_ripseq, "eif4e_ip_rna_lfc_normoxia_1hr", "normoxia_1hr")
make_cdf_plot(
  df, "eif4e_ip_rna_lfc_normoxia_1hr",
  positive_label = "eIF3d normoxia promotes",
  x_label    = "log2FC (eIF4E IP / sictrl RNA, DESeq2, normoxia 1hr)",
  plot_title = "eIF3d normoxia-promoted targets in eIF4E IP: IP/RNA CDF (normoxia 1hr)",
  out_name   = "19_cdf_eif4e_iprna_normoxia_1hr.pdf"
)
```

## IgG-mock / RNA (background reference)

No bait dimension here -- IgG-mock is a single pseudo-bait shared across baits, not
split eIF3d/eIF4E. "Positive" is still the eIF3d TE-promoted gene set (same convention
as the eIF4E panels above): this tests whether eIF3d-dependent translational targets
show any background/nonspecific enrichment in the IgG-mock pulldown, which they should
not if IP/IgG and IP/RNA enrichment in the bait IPs reflects real binding rather than
shared RNA-abundance artifacts.

```{r plot_igg_rna_hypoxia_1hr}
df <- igg_rna_long |>
  dplyr::filter(timepoint == "hypoxia_1hr") |>
  build_group_df("log2_igg_over_rna", "hypoxia_1hr")
make_cdf_plot(
  df, "log2_igg_over_rna",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (IgG-mock / sictrl RNA, DESeq2, hypoxia 1hr)",
  plot_title = "eIF3d hypoxia-promoted targets in IgG-mock background: IgG/RNA CDF (hypoxia 1hr)",
  out_name   = "19_cdf_igg_rna_hypoxia_1hr.pdf"
)
```

```{r plot_igg_rna_hypoxia_4hr}
df <- igg_rna_long |>
  dplyr::filter(timepoint == "hypoxia_4hr") |>
  build_group_df("log2_igg_over_rna", "hypoxia_4hr")
make_cdf_plot(
  df, "log2_igg_over_rna",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (IgG-mock / sictrl RNA, DESeq2, hypoxia 4hr)",
  plot_title = "eIF3d hypoxia-promoted targets in IgG-mock background: IgG/RNA CDF (hypoxia 4hr)",
  out_name   = "19_cdf_igg_rna_hypoxia_4hr.pdf"
)
```

```{r plot_igg_rna_hypoxia_24hr}
df <- igg_rna_long |>
  dplyr::filter(timepoint == "hypoxia_24hr") |>
  build_group_df("log2_igg_over_rna", "hypoxia_24hr")
make_cdf_plot(
  df, "log2_igg_over_rna",
  positive_label = "eIF3d hypoxia promotes",
  x_label    = "log2FC (IgG-mock / sictrl RNA, DESeq2, hypoxia 24hr)",
  plot_title = "eIF3d hypoxia-promoted targets in IgG-mock background: IgG/RNA CDF (hypoxia 24hr)",
  out_name   = "19_cdf_igg_rna_hypoxia_24hr.pdf"
)
```

```{r plot_igg_rna_normoxia_1hr}
df <- igg_rna_long |>
  dplyr::filter(timepoint == "normoxia_1hr") |>
  build_group_df("log2_igg_over_rna", "normoxia_1hr")
make_cdf_plot(
  df, "log2_igg_over_rna",
  positive_label = "eIF3d normoxia promotes",
  x_label    = "log2FC (IgG-mock / sictrl RNA, DESeq2, normoxia 1hr; n=2 IgG-mock reps)",
  plot_title = "eIF3d normoxia-promoted targets in IgG-mock background: IgG/RNA CDF (normoxia 1hr)",
  out_name   = "19_cdf_igg_rna_normoxia_1hr.pdf"
)
```
