---
title: "RIP-seq: eIF4E IP validation against published Teleman eIF4E-IP (vehicle)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Sanity check: does our own eIF4E RIP-seq recapitulate the published Teleman-lab eIF4E-IP
enrichment/depletion? Three-group CDF plots (Positive / Negative control / Other),
faceted per RIP-seq timepoint, mirroring the design in
`ripseq/code/19_density_3group_ip_enrichment.Rmd` -- but restricted to the eIF4E bait
only (both IgG and RNA denominators), plus the bait-independent IgG-mock/RNA background
reference.

**Gene sets**: "Positive" is the Teleman eIF4E-enriched set (Okabe-Ito reddish
purple/pink `#CC79A7`, matching the color already used for "Teleman vehicle eIF4E-IP" in
`ripseq/code/03_eif4e_rip_analysis.Rmd`). "Negative control" is **not** a random
size-matched sample (unlike notebook 19) -- it's the biologically meaningful set of genes
Teleman found significantly *depleted* from the eIF4E-IP (mirrors the enriched-set
definition, just the opposite tail). Both sets come from the same external "vehicle"
(untreated, i.e. the "normal condition" of that experiment -- as opposed to their
inhibitor-treated condition) DESeq2 result already used in
`ripseq/code/03_eif4e_rip_analysis.Rmd`, with the identical thresholds established there
(`log2FoldChange > 1.0 & padj < 0.05` for enriched; `log2FoldChange < -1.0 & padj < 0.05`
for depleted). Unlike notebook 19, there is only one gene-set choice here (Teleman), not
two, and only one bait (eIF4E), so this notebook does not need the generic
multi-geneset/multi-bait machinery notebook 19 uses.

**Caveats** (same as notebook 19): the IP/RNA and IgG/RNA columns carry the RIP-seq
(2026 batch) vs RNA-seq (2025 batch) no-shared-spike-in batch confound; the IgG-mock/RNA
metric has 2 replicates at normoxia_1hr (sample `NM2026_0077` excluded as a confirmed
outlier) vs 3 at every hypoxia timepoint.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Constants

```{r constants}
library(here)
library(tidyverse)
library(patchwork)

CONDITIONS <- c("hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr", "normoxia_1hr")

timepoint_labels <- c(
  hypoxia_1hr  = "hypoxia 1hr",
  hypoxia_4hr  = "hypoxia 4hr",
  hypoxia_24hr = "hypoxia 24hr",
  normoxia_1hr = "normoxia 1hr"
)

# Display column order for composite grids: normoxia first, then hypoxia 1/4/24hr.
col_order   <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")
col_headers <- c("Norm", "Hyp - 1", "Hyp - 4", "Hyp - 24")

TELEMAN_LFC_CUTOFF  <- 1.0
TELEMAN_PADJ_CUTOFF <- 0.05

group_colors <- c(
  "Positive"         = "#CC79A7",  # Okabe-Ito reddish purple/pink
  "Negative control" = "#000000",
  "Other"            = "grey80"
)
```

---

# Section 1: Load RIP-seq feature matrix

```{r load_feature_matrix}
fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Feature matrix loaded:", nrow(fm_ripseq), "unique genes\n")

value_cols <- c(
  paste0("eif4e_ip_igg_lfc_", CONDITIONS),
  paste0("eif4e_ip_rna_lfc_", CONDITIONS)
)

stopifnot(
  "all 8 required eIF4E RIP-seq columns must be present" =
    all(value_cols %in% colnames(fm_ripseq)),
  "required columns must be finite where non-NA" = all(sapply(
    value_cols,
    function(col) all(is.finite(fm_ripseq[[col]][!is.na(fm_ripseq[[col]])]))
  ))
)
cat("Feature matrix column checks passed\n")
```

---

# Section 1b: Load IgG/RNA background-reference enrichment

```{r load_igg_rna}
igg_rna_long <- read_csv(
  here("ripseq", "output", "18_igg_over_rna_enrichment.csv"),
  show_col_types = FALSE
) |>
  mutate(timepoint = factor(timepoint, levels = CONDITIONS))

cat("IgG/RNA enrichment loaded:", nrow(igg_rna_long), "rows across",
    length(unique(igg_rna_long$timepoint)), "timepoints\n")

stopifnot(
  "required columns must be present in igg_rna_long" =
    all(c("gene_id_clean", "timepoint", "log2_igg_over_rna") %in% colnames(igg_rna_long)),
  "log2_igg_over_rna must be finite where non-NA" =
    all(is.finite(igg_rna_long$log2_igg_over_rna[!is.na(igg_rna_long$log2_igg_over_rna)])),
  "all 4 CONDITIONS timepoints must be present" =
    all(CONDITIONS %in% as.character(igg_rna_long$timepoint))
)
cat("IgG/RNA column checks passed\n")
```

---

# Section 2: Load Teleman eIF4E-IP (vehicle) gene sets

```{r load_teleman}
teleman_vehicle <- read_csv(
  "/Users/katematlin/github/mpra_preliminary/res_eif4e_ip_vehicle.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(!is.na(symbol)) |>
  dplyr::group_by(symbol) |>
  dplyr::slice_max(order_by = baseMean, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

cat("Teleman vehicle genes loaded:", nrow(teleman_vehicle), "\n")

teleman_enriched_symbols <- teleman_vehicle |>
  dplyr::filter(log2FoldChange > TELEMAN_LFC_CUTOFF, padj < TELEMAN_PADJ_CUTOFF) |>
  dplyr::pull(symbol)
teleman_depleted_symbols <- teleman_vehicle |>
  dplyr::filter(log2FoldChange < -TELEMAN_LFC_CUTOFF, padj < TELEMAN_PADJ_CUTOFF) |>
  dplyr::pull(symbol)

cat("Teleman-enriched (LFC >", TELEMAN_LFC_CUTOFF, ", padj <", TELEMAN_PADJ_CUTOFF, "):",
    length(teleman_enriched_symbols), "\n")
cat("Teleman-depleted (LFC < -", TELEMAN_LFC_CUTOFF, ", padj <", TELEMAN_PADJ_CUTOFF, "):",
    length(teleman_depleted_symbols), "\n")

# Map Teleman symbols onto our gene_id_clean universe via fm_ripseq's own symbol column.
positive_set <- fm_ripseq |>
  dplyr::filter(symbol %in% teleman_enriched_symbols) |>
  dplyr::pull(gene_id_clean)
negative_set <- fm_ripseq |>
  dplyr::filter(symbol %in% teleman_depleted_symbols) |>
  dplyr::pull(gene_id_clean)

cat("Teleman-enriched genes in our universe (Positive):", length(positive_set), "\n")
cat("Teleman-depleted genes in our universe (Negative control):", length(negative_set), "\n")

stopifnot(
  "too few Teleman-enriched genes after join -- check file path or symbol matching" =
    length(positive_set) > 50,
  "too few Teleman-depleted genes after join -- check file path or symbol matching" =
    length(negative_set) > 50,
  "Teleman enriched and depleted sets must be mutually exclusive" =
    length(intersect(positive_set, negative_set)) == 0
)
cat("Teleman gene set checks passed\n")
```

---

# Section 3: Group-assignment helper

```{r build_group_df}
build_group_df <- function(fm, value_col) {
  fm |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::mutate(
      group = dplyr::case_when(
        gene_id_clean %in% positive_set ~ "Positive",
        gene_id_clean %in% negative_set ~ "Negative control",
        TRUE                             ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

Only one gene-set choice exists in this notebook (Teleman enriched/depleted), so unlike
notebook 19's `build_group_df()`, this version takes no `geneset`/`timepoint_val`
argument -- `positive_set`/`negative_set` are fixed module-level constants from Section 2.

---

# Section 4: Validation

```{r validation_computational}
for (tp in CONDITIONS) {
  d_igg <- build_group_df(fm_ripseq, paste0("eif4e_ip_igg_lfc_", tp))
  d_rna <- build_group_df(fm_ripseq, paste0("eif4e_ip_rna_lfc_", tp))
  d_iggrna <- igg_rna_long |>
    dplyr::filter(timepoint == tp) |>
    build_group_df("log2_igg_over_rna")
  for (d in list(d_igg, d_rna, d_iggrna)) {
    stopifnot(
      "group column must have no NA" = !any(is.na(d$group)),
      "Positive group must be non-empty" = sum(d$group == "Positive") > 0,
      "Negative control group must be non-empty" = sum(d$group == "Negative control") > 0
    )
  }
}
cat("Computational checks passed\n")
```

```{r validation_biological}
eif4e_self <- fm_ripseq |>
  dplyr::filter(symbol == "EIF4E") |>
  pull(eif4e_ip_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF4E self-enrichment in eif4e IP/IgG must be positive" =
    length(eif4e_self) > 0 && all(eif4e_self > 0, na.rm = TRUE)
)
cat("EIF4E self-enrichment (hypoxia_1hr, IP/IgG):", eif4e_self, "\n")
cat("Biological spot-check passed\n")
```

---

# Section 5: Plotting helper

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, x_label, plot_title, out_name,
                           width = 7, height = 5.5) {
  n_labels <- c(
    "Positive"         = paste0("Teleman eIF4E enriched (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Teleman eIF4E depleted (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = c(-2, 2)) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 6: Generate plots

```{r plot_registry_init}
plot_registry <- list()
```

## eIF4E IP/IgG and IP/RNA vs Teleman enriched/depleted

```{r plots_eif4e}
for (tp in CONDITIONS) {
  df_igg <- build_group_df(fm_ripseq, paste0("eif4e_ip_igg_lfc_", tp))
  p_igg <- make_cdf_plot(
    df_igg, paste0("eif4e_ip_igg_lfc_", tp),
    x_label    = paste0("log2FC (eIF4E IP / IgG, ", timepoint_labels[[tp]], ")"),
    plot_title = paste0("Teleman eIF4E-enriched/depleted targets in eIF4E IP: IP/IgG CDF (",
                         timepoint_labels[[tp]], ")"),
    out_name   = paste0("20_cdf_eif4e_igg_", tp, "_teleman.pdf")
  )
  plot_registry[[paste("eif4e", "igg", tp, sep = "_")]] <- p_igg
  print(p_igg)

  df_rna <- build_group_df(fm_ripseq, paste0("eif4e_ip_rna_lfc_", tp))
  p_rna <- make_cdf_plot(
    df_rna, paste0("eif4e_ip_rna_lfc_", tp),
    x_label    = paste0("log2FC (eIF4E IP / sictrl RNA, DESeq2, ", timepoint_labels[[tp]], ")"),
    plot_title = paste0("Teleman eIF4E-enriched/depleted targets in eIF4E IP: IP/RNA CDF (",
                         timepoint_labels[[tp]], ")"),
    out_name   = paste0("20_cdf_eif4e_iprna_", tp, "_teleman.pdf")
  )
  plot_registry[[paste("eif4e", "rna", tp, sep = "_")]] <- p_rna
  print(p_rna)
}
```

## IgG-mock / RNA (background reference) vs Teleman enriched/depleted

```{r plots_igg_rna}
for (tp in CONDITIONS) {
  tp_label_x <- timepoint_labels[[tp]]
  if (tp == "normoxia_1hr") tp_label_x <- paste0(tp_label_x, "; n=2 IgG-mock reps")

  df_iggrna <- igg_rna_long |>
    dplyr::filter(timepoint == tp) |>
    build_group_df("log2_igg_over_rna")
  p_iggrna <- make_cdf_plot(
    df_iggrna, "log2_igg_over_rna",
    x_label    = paste0("log2FC (IgG-mock / sictrl RNA, DESeq2, ", tp_label_x, ")"),
    plot_title = paste0("Teleman eIF4E-enriched/depleted targets in IgG-mock background: IgG/RNA CDF (",
                         timepoint_labels[[tp]], ")"),
    out_name   = paste0("20_cdf_igg_rna_", tp, "_teleman.pdf")
  )
  plot_registry[[paste("iggrna", tp, sep = "_")]] <- p_iggrna
  print(p_iggrna)
}
```

---

# Section 7: Composite grids (patchwork)

Assembles the individual plots above into two composite figures: eIF4E IP (IgG row +
RNA row, 4 timepoint columns) and the bait-independent IgG-mock/RNA background (single
row, 4 columns). Same decluttered-panel + shared-legend + header/row-label design as
`ripseq/code/19_density_3group_ip_enrichment.Rmd`'s Section 7.

```{r composite_helper}
make_text_panel <- function(label, angle = 0, size = 5) {
  ggplot() +
    theme_void() +
    annotate("text", x = 0, y = 0, label = label, size = size, fontface = "bold", angle = angle)
}

strip_for_composite <- function(p) {
  p + labs(title = NULL, x = "log2FC") + theme(legend.position = "none")
}

make_composite_grid <- function(bait, title, out_name) {
  if (bait == "iggrna") {
    row_labels <- "IgG/RNA"
    row_keys   <- list(paste("iggrna", col_order, sep = "_"))
  } else {
    row_labels <- c("IP/IgG", "IP/RNA")
    row_keys   <- list(
      paste(bait, "igg", col_order, sep = "_"),
      paste(bait, "rna", col_order, sep = "_")
    )
  }
  n_rows <- length(row_labels)
  stopifnot("all requested plots must exist in plot_registry" =
    all(unlist(row_keys) %in% names(plot_registry)))

  cells <- c(list(make_text_panel("")), lapply(col_headers, make_text_panel))
  for (r in seq_len(n_rows)) {
    cells <- c(cells,
               list(make_text_panel(row_labels[r], angle = 90)),
               lapply(plot_registry[row_keys[[r]]], strip_for_composite))
  }

  main_grid <- patchwork::wrap_plots(
    cells, ncol = 5,
    widths  = c(0.15, 1, 1, 1, 1),
    heights = c(0.15, rep(1, n_rows))
  )

  legend_src <- plot_registry[[row_keys[[1]][1]]] +
    scale_color_manual(values = group_colors, breaks = names(group_colors)) +
    theme(legend.position = "bottom")
  legend_grob <- cowplot::get_legend(legend_src)

  full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
    patchwork::plot_layout(heights = c(1, 0.06)) +
    patchwork::plot_annotation(title = title)

  ggsave(here("ripseq", "plots", out_name), full,
         width = 4 * 4.2 + 1, height = n_rows * 4 + 2, limitsize = FALSE)
  full
}
```

```{r composite_eif4e, fig.width=18, fig.height=9}
make_composite_grid(
  "eif4e",
  title    = "eIF4E IP: Teleman eIF4E-enriched/depleted targets, all timepoints",
  out_name = "20_grid_eif4e_teleman.pdf"
)
```

```{r composite_iggrna, fig.width=18, fig.height=5}
make_composite_grid(
  "iggrna",
  title    = "IgG-mock background: Teleman eIF4E-enriched/depleted targets, all timepoints",
  out_name = "20_grid_iggrna_teleman.pdf"
)
```
