---
title: "RIP-seq: eIF4E binding delta-delta (hypoxia vs normoxia) -- CDFs for eIF3d hypoxia-promoted targets"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

This notebook looks at how eIF4E RIP-seq binding **changes** between hypoxia (each
timepoint) and normoxia, rather than raw IP/IgG enrichment at each timepoint separately
(that's `19_density_3group_ip_enrichment.Rmd`). It reuses the DESeq2 interaction-model
"delta-delta" approach `09_rip_delta_deseq_and_cdf.Rmd` introduced in May for eIF4E
IP/IgG, recomputed fresh here, plus a new analogous delta-delta for the IgG/RNA
background-reference metric (never previously computed). Both are genuinely
double-difference ("delta-delta") quantities -- each side of the definition below
(`eIF4E_Xhr / IgG_Xhr`, or `IgG-mock_Xhr / RNA_Xhr`) is already a first-order delta (IP
enrichment over background within one condition/timepoint), and the notebook then takes
the difference of that delta between hypoxia and normoxia. Both delta-deltas are
plotted as 3-group CDFs -- the eIF3d hypoxia-promoted positive set (orange), its
size-matched negative control (black), and all other genes (grey) -- across the three
hypoxia timepoints, same convention as notebook 19.

**Delta-delta definitions:**

- `delta_eif4e_lfc` = log2(eIF4E_hyp_Xhr / IgG_hyp_Xhr) - log2(eIF4E_nor_1hr / IgG_nor_1hr).
  Positive = eIF4E binding increases in hypoxia relative to normoxia. Computed as the
  `condition:ip_type` interaction coefficient from a single joint DESeq2 fit (design
  `~ condition + ip_type + condition:ip_type`), not as two separately-fit log2FCs
  subtracted afterward -- see the independent cross-check below for why that
  distinction matters.
- `delta_igg_rna_lfc` = log2(IgG-mock_hyp_Xhr / RNA_hyp_Xhr) - log2(IgG-mock_nor_1hr / RNA_nor_1hr).
  Background/nonspecific-binding reference: whether the negative-control pulldown shows
  any systematic hypoxia-vs-normoxia shift. Same interaction-coefficient logic, with
  `seq_type` (igg_mock vs. RNA) in place of `ip_type`.

**Staleness note:** the eIF4E delta-delta CSVs notebook 09 wrote in May (`ripseq_deseq_eif4e_hypoxia_*hr_vs_normoxia_delta.csv`)
predate the current `ripseq_allCounts_raw.rds` (regenerated in June), so they're
recomputed fresh here rather than reused. This notebook writes new `23_`-prefixed output
files and does **not** overwrite the old nb09 files, since notebooks 10, 12, and 13
still read those old filenames directly.

**Batch-confound note (IgG/RNA delta-delta):** RIP-seq (2026 batch) and RNA-seq (2025
batch) share no spike-in, so a single-condition IgG/RNA contrast (as in
`18_ip_input_normalization.Rmd`) is fully confounded with sequencing batch. Here,
because both conditions (hypoxia and normoxia) draw from the same two batches, the
`seq_type` batch effect largely cancels out in the `condition:seq_type` interaction
term -- a real reduction in that caveat relative to the single-condition metric.

**Replicate-count caveat:** normoxia_1hr IgG-mock drops from 3 to 2 replicates after
excluding outlier sample `NM2026_0077` (same exclusion as notebooks 01/09/18); all
hypoxia timepoints keep 3 IgG-mock replicates.

**Validation strategy:** each delta-delta gets three layers of checking: (1) computational
range/finite checks, (2) a *consistency* check against the same interaction-model
methodology run previously (for eIF4E, the old nb09 CSV) -- useful for catching gross
recompute errors, but not independent, since it shares the same pipeline, and (3) an
*independent cross-check* against a structurally different DESeq2 design (two separate
per-condition fits reconstructed as `hyp - nor`, rather than one joint interaction fit)
-- this is the check that actually catches a bug specific to the interaction-model
approach itself, and also the check that confirms the joint interaction coefficient and
the naive two-step subtraction are measuring the same underlying quantity (they
correlate strongly but are not numerically identical, since the joint fit shares
dispersion/size-factor estimation across more samples and `apeglm` shrinkage is applied
directly to the interaction term rather than to two terms subtracted afterward). There
is no clean, literature-grounded expectation for how any single gene's RIP-seq binding
*change* should behave a priori (unlike raw enrichment, where e.g. the bait's own
self-enrichment is a reliable invariant), so the independent cross-check also stands in
for a single-gene biological spot-check here.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Section 1: Load inputs

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(patchwork)
library(cowplot)
```

```{r constants}
OUTLIER_ID <- "0077"
HYP_HOURS  <- c("1", "4", "24")

col_headers <- c("Hyp - 1", "Hyp - 4", "Hyp - 24")

group_colors <- c(
  "Positive"         = "#E69F00",
  "Negative control" = "#000000",  # intentional override of the repo's default
                                    # "grey40" for negative control, per request
  "Other"            = "grey80"
)
```

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  dplyr::filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

gene_anno <- gene_anno_all |>
  arrange(gene_id_clean, desc(biotype == "protein_coding")) |>
  distinct(gene_id_clean, .keep_all = TRUE) |>
  dplyr::select(gene_id_clean, symbol)

# A gene is protein-coding if ANY transcript row has biotype "protein_coding"
# (same any-transcript rule as 18_ip_input_normalization.Rmd).
pc_genes <- gene_anno_all |>
  dplyr::filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)
```

```{r load_rip}
allCounts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
metadata  <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                      show_col_types = FALSE) |>
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia")),
    ip_type   = factor(ip_type,   levels = c("igg_mock", "flag_eif3d", "eif4e")),
    hours     = factor(hours)
  )

metadata_filt  <- metadata  |> dplyr::filter(!grepl(OUTLIER_ID, id))
allCounts_filt <- allCounts[, !grepl(OUTLIER_ID, colnames(allCounts))]

# Belt-and-suspenders protein-coding filter (same rationale as
# 18_ip_input_normalization.Rmd): allCounts is already pc-filtered upstream in
# 00_ripseq_qc.Rmd, so this is a no-op in practice, but guards against that
# upstream assumption silently changing.
allCounts_gene_id_clean <- sub("\\..*", "", rownames(allCounts_filt))
allCounts_filt <- allCounts_filt[allCounts_gene_id_clean %in% pc_genes, , drop = FALSE]

cat("RIP-seq counts after outlier removal + protein-coding filter:",
    nrow(allCounts_filt), "genes x", ncol(allCounts_filt), "samples\n")
```

```{r load_rna}
rna_counts_all <- read_csv(
  here("output", "human_allCounts_raw.csv"),
  col_select = c(gene_id, matches("sictrl.*_rna_")),
  show_col_types = FALSE
) |>
  column_to_rownames("gene_id") |>
  as.matrix()

rna_meta <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE) |>
  dplyr::filter(library == "rna", treatment == "sictrl") |>
  mutate(hours = as.character(hours), condition = as.character(condition))

stopifnot("RIP-seq genes must be a subset of RNA-seq genes" =
  all(rownames(allCounts_filt) %in% rownames(rna_counts_all)))
rna_counts <- rna_counts_all[rownames(allCounts_filt), , drop = FALSE]

cat("sictrl RNA-seq counts (subset to RIP-seq genes):", nrow(rna_counts), "genes x",
    ncol(rna_counts), "samples\n")
```

---

# Section 2: Gene sets

Identical construction to `19_density_3group_ip_enrichment.Rmd` Section 2 (same as
notebook 07's `create_target` chunk) -- only the hypoxia-defined positive set is needed
here.

```{r load_genesets}
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> pull(ensembl_gene)

fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

negctrl_clean <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pull(gene_id_clean)

neg_in_features     <- negctrl_clean[negctrl_clean %in% fm_ripseq$gene_id_clean]
pos_in_features_hyp <- promotes_hyp[promotes_hyp %in% fm_ripseq$gene_id_clean]
number_to_match_hyp <- min(length(pos_in_features_hyp), length(neg_in_features))
set.seed(9)
neg_hyp <- sample(neg_in_features, number_to_match_hyp)

cat("Positive set (hypoxia-promoted):", length(promotes_hyp), "genes\n")
cat("Size-matched negative control:", length(neg_hyp), "genes\n")

stopifnot("positive and negative-control sets must be mutually exclusive" =
  length(intersect(promotes_hyp, neg_hyp)) == 0)
cat("Gene set mutual-exclusivity check passed\n")
```

---

# Section 3: eIF4E IP/IgG delta-delta (recomputed)

Adapted from `09_rip_delta_deseq_and_cdf.Rmd`'s `deseq_4e` chunk, run fresh against the
current `ripseq_allCounts_raw.rds`. Design: `~ condition + ip_type + condition:ip_type`,
igg_mock vs eif4e, per hypoxia timepoint vs normoxia_1hr; the interaction coefficient is
the delta-delta.

```{r deseq_4e_delta}
res_4e_list <- list()

for (hyp_hr in HYP_HOURS) {
  meta_sub <- metadata_filt |>
    dplyr::filter((condition == "normoxia" & hours == "1") |
                  (condition == "hypoxia"  & hours == hyp_hr),
                  ip_type %in% c("eif4e", "igg_mock")) |>
    mutate(ip_type = droplevels(factor(ip_type, levels = c("igg_mock", "eif4e")))) |>
    as.data.frame()

  stopifnot(
    "Expected 5 normoxia (2 igg_mock + 3 eif4e) and 6 hypoxia (3 igg_mock + 3 eif4e) samples" =
      sum(meta_sub$condition == "normoxia") == 5 &&
      sum(meta_sub$condition == "hypoxia")  == 6
  )

  # Build cts_sub one sample at a time, in meta_sub row order, with an
  # explicit per-sample column-count check -- catches an id that resolves to
  # zero or multiple columns before it can silently misalign counts with
  # metadata (same rationale as 18_ip_input_normalization.Rmd's explicit
  # ncol(...) == n checks after grep-based column selection).
  cts_cols <- lapply(meta_sub$id, function(x) {
    hits <- grep(x, colnames(allCounts_filt))
    stopifnot("each sample id must resolve to exactly one count-matrix column" =
      length(hits) == 1)
    allCounts_filt[, hits, drop = FALSE]
  })
  cts_sub <- do.call(cbind, cts_cols)
  rownames(meta_sub) <- colnames(cts_sub)

  dds <- DESeqDataSetFromMatrix(cts_sub, meta_sub,
                                 design = ~ condition + ip_type + condition:ip_type)
  dds <- DESeq(dds, quiet = TRUE)

  coef_int <- grep("\\.", resultsNames(dds), value = TRUE)
  stopifnot("Expected exactly one interaction coefficient" = length(coef_int) == 1)

  res <- as.data.frame(lfcShrink(dds, coef = coef_int, type = "apeglm")) |>
    rownames_to_column("gene_id") |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    left_join(gene_anno, by = "gene_id_clean") |>
    dplyr::rename(delta_eif4e_lfc = log2FoldChange, delta_eif4e_padj = padj) |>
    mutate(timepoint = paste0("hypoxia_", hyp_hr, "hr_vs_normoxia"))

  res_4e_list[[hyp_hr]] <- res

  write_csv(res, here("ripseq", "output",
                       paste0("23_eif4e_hypoxia_", hyp_hr, "hr_vs_normoxia_delta.csv")))
  cat(hyp_hr, "hr: samples =", ncol(cts_sub), "| genes =", nrow(res),
      "| padj<0.05 =", sum(res$delta_eif4e_padj < 0.05, na.rm = TRUE), "\n")
}
```

```{r validate_4e_delta}
for (hyp_hr in HYP_HOURS) {
  res <- res_4e_list[[hyp_hr]]
  stopifnot(
    "delta_eif4e_lfc must be finite where non-NA" =
      all(is.finite(res$delta_eif4e_lfc[!is.na(res$delta_eif4e_lfc)])),
    "gene count should be in a reasonable range" = nrow(res) > 5000
  )
}
cat("eIF4E delta-delta computational checks passed\n")
```

**Consistency check** (not a biological check -- see below): Spearman correlation
between the freshly recomputed delta-delta and the old (stale) nb09 CSV. This only confirms
the recompute is a refinement of the same interaction-model methodology, not a break --
it would *not* catch a bug shared by both runs (e.g. the same wrong reference factor
level), since both use the identical `~ condition + ip_type + condition:ip_type`
pipeline.

```{r consistency_check_4e_delta}
old_paths <- setNames(
  here("ripseq", "output",
       paste0("ripseq_deseq_eif4e_hypoxia_", HYP_HOURS, "hr_vs_normoxia_delta.csv")),
  HYP_HOURS
)

for (hyp_hr in HYP_HOURS) {
  stopifnot("expected old nb09 delta-delta CSV to exist for comparison" =
    file.exists(old_paths[[hyp_hr]]))

  old <- read_csv(old_paths[[hyp_hr]], show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, delta_eif4e_lfc_old = delta_eif4e_lfc)

  cmp <- res_4e_list[[hyp_hr]] |>
    dplyr::select(gene_id_clean, delta_eif4e_lfc) |>
    inner_join(old, by = "gene_id_clean") |>
    dplyr::filter(is.finite(delta_eif4e_lfc), is.finite(delta_eif4e_lfc_old))

  rho <- cor(cmp$delta_eif4e_lfc, cmp$delta_eif4e_lfc_old, method = "spearman")
  cat(hyp_hr, "hr: Spearman rho vs old nb09 delta-delta (same methodology) =", round(rho, 3),
      "(n =", nrow(cmp), ")\n")
  stopifnot("recomputed eIF4E delta-delta should correlate strongly with the prior (stale) run" =
    rho > 0.7)
}
cat("eIF4E delta-delta consistency check passed\n")
```

**Independent cross-check**: this is the check that actually stands in for a biological
spot-check here. Rather than asserting an invariant about one specific gene (there is
no clean a priori expectation for how *any individual* gene's eIF4E binding should
change between conditions), this validates the interaction-model delta-delta against a
*structurally independent* computation: `deseq_results_4e_final` (notebook 01's
"final" per-timepoint eIF4E-vs-IgG two-group DESeq2 model, saved to
`ripseq_deseq_{tp}_eif4e_vs_igg.csv`) reconstructed as
`lfc_hyp - lfc_nor`. Unlike the consistency check above, this uses an entirely
different DESeq2 design (two separate two-group fits vs. one joint interaction fit),
so a bug specific to the interaction-model approach (e.g. a flipped reference level in
the `condition:ip_type` term) would show up here even if it also existed in the old
nb09 file.

```{r independent_crosscheck_4e_delta}
nor_1hr_igg <- read_csv(
  here("ripseq", "output", "ripseq_deseq_normoxia_1hr_eif4e_vs_igg.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, lfc_nor = log2FoldChange)

for (hyp_hr in HYP_HOURS) {
  hyp_igg <- read_csv(
    here("ripseq", "output", paste0("ripseq_deseq_hypoxia_", hyp_hr, "hr_eif4e_vs_igg.csv")),
    show_col_types = FALSE
  ) |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    dplyr::select(gene_id_clean, lfc_hyp = log2FoldChange)

  cmp <- res_4e_list[[hyp_hr]] |>
    dplyr::select(gene_id_clean, delta_eif4e_lfc) |>
    inner_join(hyp_igg, by = "gene_id_clean") |>
    inner_join(nor_1hr_igg, by = "gene_id_clean") |>
    mutate(reconstructed = lfc_hyp - lfc_nor) |>
    dplyr::filter(is.finite(delta_eif4e_lfc), is.finite(reconstructed))

  rho <- cor(cmp$delta_eif4e_lfc, cmp$reconstructed, method = "spearman")
  cat(hyp_hr, "hr: Spearman rho vs independently-reconstructed two-model delta-delta =",
      round(rho, 3), "(n =", nrow(cmp), ")\n")
  stopifnot("interaction-model delta-delta should correlate strongly with the independently reconstructed delta-delta" =
    rho > 0.7)
}
cat("eIF4E delta-delta independent cross-check passed\n")
```

---

# Section 4: IgG/RNA delta-delta (new)

Same interaction-model structure as Section 3, but with `seq_type` (igg_mock vs sictrl
RNA) in place of `ip_type`, combining `18_ip_input_normalization.Rmd`'s IgG/RNA counts
setup (igg_mock counts + sictrl RNA counts, zero-RNA-coverage guard) with the
`condition:seq_type` interaction design from Section 3. Per hypoxia timepoint: igg_mock
hyp_Xhr (3 reps) + igg_mock nor_1hr (2 reps, outlier `0077` excluded) + sictrl RNA
hyp_Xhr (3 reps) + sictrl RNA nor_1hr (3 reps) = 11 samples.

```{r deseq_iggrna_delta}
res_iggrna_list <- list()

for (hyp_hr in HYP_HOURS) {
  igg_meta <- metadata_filt |>
    dplyr::filter(ip_type == "igg_mock",
                  (condition == "normoxia" & hours == "1") |
                  (condition == "hypoxia"  & hours == hyp_hr)) |>
    mutate(condition = as.character(condition), seq_type = "igg_mock") |>
    dplyr::select(id, condition, seq_type)

  rna_meta_sub <- rna_meta |>
    dplyr::filter((condition == "normoxia" & hours == "1") |
                  (condition == "hypoxia"  & hours == hyp_hr)) |>
    mutate(seq_type = "rna") |>
    dplyr::select(id, condition, seq_type)

  stopifnot(
    "Expected 2 normoxia + 3 hypoxia IgG-mock replicates" =
      sum(igg_meta$condition == "normoxia") == 2 && sum(igg_meta$condition == "hypoxia") == 3,
    "Expected 3 normoxia + 3 hypoxia sictrl RNA replicates" =
      sum(rna_meta_sub$condition == "normoxia") == 3 && sum(rna_meta_sub$condition == "hypoxia") == 3
  )

  col_data <- bind_rows(igg_meta, rna_meta_sub) |> as.data.frame()

  # Build combined counts one sample at a time, in col_data's row order, so
  # column order is guaranteed to match col_data (igg_mock samples come from
  # allCounts_filt, rna samples from rna_counts). Explicit per-sample
  # column-count check catches an id resolving to zero or multiple columns
  # before it can silently misalign counts with metadata.
  cts_list <- lapply(seq_len(nrow(col_data)), function(i) {
    src  <- if (col_data$seq_type[i] == "igg_mock") allCounts_filt else rna_counts
    hits <- grep(col_data$id[i], colnames(src))
    stopifnot("each sample id must resolve to exactly one count-matrix column" =
      length(hits) == 1)
    src[, hits, drop = FALSE]
  })
  combined_cts <- do.call(cbind, cts_list)
  rownames(col_data) <- colnames(combined_cts)

  # Zero-RNA-coverage guard, same rationale as 18_ip_input_normalization.Rmd:
  # genes with no RNA counts across all sictrl reps have no real denominator.
  rna_covered <- rowSums(combined_cts[, col_data$seq_type == "rna", drop = FALSE]) > 0
  n_dropped   <- sum(!rna_covered)
  if (n_dropped > 0) {
    cat("  Dropping", n_dropped, "genes with zero RNA coverage (igg_mock", hyp_hr, "hr)\n")
  }
  combined_cts <- round(combined_cts[rna_covered, , drop = FALSE])

  col_data <- col_data |>
    mutate(
      condition = factor(condition, levels = c("normoxia", "hypoxia")),
      seq_type  = factor(seq_type,  levels = c("rna", "igg_mock"))
    )

  dds <- DESeqDataSetFromMatrix(combined_cts, col_data,
                                 design = ~ condition + seq_type + condition:seq_type)
  dds <- DESeq(dds, quiet = TRUE)

  coef_int <- grep("\\.", resultsNames(dds), value = TRUE)
  stopifnot("Expected exactly one interaction coefficient" = length(coef_int) == 1)

  res <- as.data.frame(lfcShrink(dds, coef = coef_int, type = "apeglm")) |>
    rownames_to_column("gene_id") |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    left_join(gene_anno, by = "gene_id_clean") |>
    dplyr::rename(delta_igg_rna_lfc = log2FoldChange, delta_igg_rna_padj = padj) |>
    mutate(timepoint = paste0("hypoxia_", hyp_hr, "hr_vs_normoxia"))

  res_iggrna_list[[hyp_hr]] <- res

  write_csv(res, here("ripseq", "output",
                       paste0("23_iggrna_hypoxia_", hyp_hr, "hr_vs_normoxia_delta.csv")))
  cat(hyp_hr, "hr: samples =", ncol(combined_cts), "| genes =", nrow(res),
      "| padj<0.05 =", sum(res$delta_igg_rna_padj < 0.05, na.rm = TRUE), "\n")
}
```

```{r validate_iggrna_delta}
for (hyp_hr in HYP_HOURS) {
  res <- res_iggrna_list[[hyp_hr]]
  stopifnot(
    "delta_igg_rna_lfc must be finite where non-NA" =
      all(is.finite(res$delta_igg_rna_lfc[!is.na(res$delta_igg_rna_lfc)]))
  )
}

n_per_tp <- sapply(res_iggrna_list, nrow)
stopifnot("gene coverage should not vary by more than ~20% across timepoints" =
  (max(n_per_tp) - min(n_per_tp)) / max(n_per_tp) < 0.2)

# IgG-mock has no bait protein, so there is no self-enrichment spot check
# (same reasoning as 18_ip_input_normalization.Rmd). This distributional
# plausibility check confirms the background reference doesn't show a large
# systematic hypoxia-vs-normoxia shift -- useful, but loose (up to 2-fold) and
# insensitive to a per-gene misalignment that leaves the median near zero.
# The independent cross-check below is the stronger validation.
for (hyp_hr in HYP_HOURS) {
  med <- median(res_iggrna_list[[hyp_hr]]$delta_igg_rna_lfc, na.rm = TRUE)
  cat(hyp_hr, "hr: median delta_igg_rna_lfc =", round(med, 3), "\n")
  stopifnot("background-reference delta-delta should not show a large systematic shift" =
    abs(med) < 1)
}
cat("IgG/RNA delta-delta computational + distributional-plausibility checks passed\n")
```

**Independent cross-check**: same rationale as the eIF4E delta-delta's independent
cross-check above. `18_ip_input_normalization.Rmd`'s single-condition IgG/RNA DESeq2
model (`18_igg_over_rna_enrichment.csv`) uses a structurally different design (one
condition per fit, no interaction term) from this notebook's `condition:seq_type`
interaction model, so reconstructing `delta_igg_rna_lfc` as
`val_hyp - val_nor` from that independent computation and checking correlation
catches a bug specific to this notebook's interaction-model approach.

```{r independent_crosscheck_iggrna_delta}
igg_rna_single <- read_csv(
  here("ripseq", "output", "18_igg_over_rna_enrichment.csv"),
  show_col_types = FALSE
)

nor_1hr_iggrna <- igg_rna_single |>
  dplyr::filter(timepoint == "normoxia_1hr") |>
  dplyr::select(gene_id_clean, val_nor = log2_igg_over_rna)

for (hyp_hr in HYP_HOURS) {
  hyp_iggrna <- igg_rna_single |>
    dplyr::filter(timepoint == paste0("hypoxia_", hyp_hr, "hr")) |>
    dplyr::select(gene_id_clean, val_hyp = log2_igg_over_rna)

  cmp <- res_iggrna_list[[hyp_hr]] |>
    dplyr::select(gene_id_clean, delta_igg_rna_lfc) |>
    inner_join(hyp_iggrna, by = "gene_id_clean") |>
    inner_join(nor_1hr_iggrna, by = "gene_id_clean") |>
    mutate(reconstructed = val_hyp - val_nor) |>
    dplyr::filter(is.finite(delta_igg_rna_lfc), is.finite(reconstructed))

  rho <- cor(cmp$delta_igg_rna_lfc, cmp$reconstructed, method = "spearman")
  cat(hyp_hr, "hr: Spearman rho vs independently-reconstructed two-model delta-delta =",
      round(rho, 3), "(n =", nrow(cmp), ")\n")
  stopifnot("interaction-model IgG/RNA delta-delta should correlate strongly with the independently reconstructed delta-delta" =
    rho > 0.7)
}
cat("IgG/RNA delta-delta independent cross-check passed\n")
```

---

# Section 5: Group-assignment + CDF helper

```{r build_group_df}
build_group_df <- function(df, value_col) {
  df |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    mutate(
      group = dplyr::case_when(
        gene_id_clean %in% promotes_hyp ~ "Positive",
        gene_id_clean %in% neg_hyp      ~ "Negative control",
        TRUE                             ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

```{r validation_groups}
for (hyp_hr in HYP_HOURS) {
  d1 <- build_group_df(res_4e_list[[hyp_hr]], "delta_eif4e_lfc")
  d2 <- build_group_df(res_iggrna_list[[hyp_hr]], "delta_igg_rna_lfc")
  stopifnot(
    "Positive group must be non-empty (eIF4E)" = sum(d1$group == "Positive") > 0,
    "Negative control group must be non-empty (eIF4E)" = sum(d1$group == "Negative control") > 0,
    "Positive group must be non-empty (IgG/RNA)" = sum(d2$group == "Positive") > 0,
    "Negative control group must be non-empty (IgG/RNA)" = sum(d2$group == "Negative control") > 0
  )
}
cat("Group-assignment checks passed\n")
```

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, x_label, plot_title, out_name,
                           width = 7, height = 5.5) {
  n_labels <- c(
    "Positive"         = paste0("eIF3d hypoxia promotes (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = c(-2, 2)) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 6: Generate plots

```{r plot_registry_init}
plot_registry <- list()
```

## eIF4E IP/IgG delta-delta

```{r plots_eif4e_delta}
for (hyp_hr in HYP_HOURS) {
  df <- build_group_df(res_4e_list[[hyp_hr]], "delta_eif4e_lfc")
  p <- make_cdf_plot(
    df, "delta_eif4e_lfc",
    x_label    = paste0("eIF4E binding change (log2FC, hypoxia ", hyp_hr, "hr vs normoxia 1hr)"),
    plot_title = paste0("eIF4E IP/IgG delta-delta: eIF3d hypoxia-promoted targets (hypoxia ",
                         hyp_hr, "hr vs normoxia)"),
    out_name   = paste0("23_cdf_eif4e_delta_", hyp_hr, "hr.pdf")
  )
  plot_registry[[paste0("eif4e_", hyp_hr)]] <- p
  print(p)
}
```

## IgG/RNA delta-delta

```{r plots_iggrna_delta}
for (hyp_hr in HYP_HOURS) {
  df <- build_group_df(res_iggrna_list[[hyp_hr]], "delta_igg_rna_lfc")
  p <- make_cdf_plot(
    df, "delta_igg_rna_lfc",
    x_label    = paste0("IgG-mock/RNA background change (log2FC, hypoxia ", hyp_hr,
                         "hr vs normoxia 1hr)"),
    plot_title = paste0("IgG-mock/RNA delta-delta: eIF3d hypoxia-promoted targets (hypoxia ",
                         hyp_hr, "hr vs normoxia)"),
    out_name   = paste0("23_cdf_iggrna_delta_", hyp_hr, "hr.pdf")
  )
  plot_registry[[paste0("iggrna_", hyp_hr)]] <- p
  print(p)
}
```

---

# Section 7: Composite grids (patchwork)

One row (delta-delta) x 3 columns (hypoxia 1hr / 4hr / 24hr) -- no normoxia column,
since delta-delta is inherently relative to the normoxia_1hr baseline. Panels drop
their individual title/legend/verbose axis label for the composite (redundant once
there's a column header); one shared legend sits at the bottom.

```{r composite_helper}
make_text_panel <- function(label, angle = 0, size = 5) {
  ggplot() +
    theme_void() +
    annotate("text", x = 0, y = 0, label = label, size = size, fontface = "bold", angle = angle)
}

strip_for_composite <- function(p) {
  p + labs(title = NULL, x = "log2FC") + theme(legend.position = "none")
}

make_composite_grid <- function(prefix, title, out_name) {
  keys <- paste0(prefix, "_", HYP_HOURS)
  stopifnot("all requested plots must exist in plot_registry" = all(keys %in% names(plot_registry)))

  cells <- c(list(make_text_panel("")), lapply(col_headers, make_text_panel))
  cells <- c(cells,
             list(make_text_panel("delta-delta", angle = 90)),
             lapply(plot_registry[keys], strip_for_composite))

  main_grid <- patchwork::wrap_plots(
    cells, ncol = 4,
    widths  = c(0.15, 1, 1, 1),
    heights = c(0.15, 1)
  )

  legend_src <- plot_registry[[keys[1]]] +
    scale_color_manual(values = group_colors, breaks = names(group_colors)) +
    theme(legend.position = "bottom")
  legend_grob <- cowplot::get_legend(legend_src)

  full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
    patchwork::plot_layout(heights = c(1, 0.06)) +
    patchwork::plot_annotation(title = title)

  ggsave(here("ripseq", "plots", out_name), full, width = 3 * 4.2 + 1, height = 6,
         limitsize = FALSE)
  full
}
```

```{r composite_eif4e, fig.width=14, fig.height=6}
make_composite_grid(
  "eif4e",
  title    = "eIF4E IP/IgG binding delta-delta: eIF3d hypoxia-promoted targets, all hypoxia timepoints",
  out_name = "23_grid_eif4e_delta.pdf"
)
```

```{r composite_iggrna, fig.width=14, fig.height=6}
make_composite_grid(
  "iggrna",
  title    = "IgG-mock/RNA delta-delta (background reference): eIF3d hypoxia-promoted targets, all hypoxia timepoints",
  out_name = "23_grid_iggrna_delta.pdf"
)
```
