---
title: "RIP-seq: FLAG-eIF3d binding delta (hypoxia vs normoxia) -- CDFs for eIF3d hypoxia-promoted targets"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

This notebook is the FLAG-eIF3d counterpart to `23_eif4e_delta_hypoxia_vs_normoxia_cdf.Rmd`:
it looks at how FLAG-eIF3d RIP-seq binding **changes** between hypoxia (each timepoint)
and normoxia, rather than raw IP/IgG enrichment at each timepoint separately (that's
`19_density_3group_ip_enrichment.Rmd`). Only the bait-specific delta (nb23 Section 3
equivalent) is recomputed here -- the IgG-mock/RNA background delta-delta (nb23 Section 4)
is not bait-specific, so it is not re-derived; see `23_eif4e_delta_hypoxia_vs_normoxia_cdf.Rmd`
Section 4 / `23_iggrna_hypoxia_*hr_vs_normoxia_delta.csv` for that reference.

**Delta definition:**

- `delta_flag3d_lfc` = log2(FLAG3d_hyp_Xhr / IgG_hyp_Xhr) - log2(FLAG3d_nor_1hr / IgG_nor_1hr).
  Positive = FLAG-eIF3d binding increases in hypoxia relative to normoxia.

**Staleness note:** the FLAG-3d delta CSVs notebook 09 wrote (`ripseq_deseq_flag3d_hypoxia_*hr_vs_normoxia_delta.csv`)
predate the current `ripseq_allCounts_raw.rds`, so they're recomputed fresh here rather
than reused. This notebook writes new `24_`-prefixed output files and does **not**
overwrite the old nb09 files.

**Replicate-count caveat:** normoxia_1hr IgG-mock drops from 3 to 2 replicates after
excluding outlier sample `NM2026_0077` (same exclusion as notebooks 01/09/18/23); all
hypoxia timepoints keep 3 IgG-mock replicates.

**Validation strategy:** same three-layer approach as nb23: (1) computational range/finite
checks, (2) a *consistency* check against the same interaction-model methodology run
previously (the old nb09 CSV) -- useful for catching gross recompute errors, but not
independent, since it shares the same pipeline, and (3) an *independent cross-check*
against a structurally different DESeq2 design (two separate per-condition fits
reconstructed as `hyp - nor`, rather than one joint interaction fit) -- this is the check
that actually catches a bug specific to the interaction-model approach itself.

**Gene-highlight note:** Grid 2 marks four genes of interest (RPP25, COL12A1, TXNIP,
CDKN2B) on the FLAG-3d delta CDFs with a vertical line + label per gene. CDKN2B is not
present in either the RIP-seq or RNA-seq count matrices in this repo (confirmed absent
upstream, not a filtering artifact), so it is dropped from the highlight grid; only
RPP25, COL12A1, and TXNIP are plotted.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Section 1: Load inputs

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(patchwork)
library(cowplot)
library(ggrepel)
```

```{r constants}
OUTLIER_ID <- "0077"
HYP_HOURS  <- c("1", "4", "24")

col_headers <- c("Hyp - 1", "Hyp - 4", "Hyp - 24")

group_colors <- c(
  "Positive"         = "#E69F00",
  "Negative control" = "#000000",  # intentional override of the repo's default
                                    # "grey40" for negative control, per request
  "Other"            = "grey80"
)

highlight_genes <- c("RPP25", "COL12A1", "TXNIP")
```

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  dplyr::filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

gene_anno <- gene_anno_all |>
  arrange(gene_id_clean, desc(biotype == "protein_coding")) |>
  distinct(gene_id_clean, .keep_all = TRUE) |>
  dplyr::select(gene_id_clean, symbol)

# A gene is protein-coding if ANY transcript row has biotype "protein_coding"
# (same any-transcript rule as 18_ip_input_normalization.Rmd / 23).
pc_genes <- gene_anno_all |>
  dplyr::filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)
```

```{r load_rip}
allCounts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
metadata  <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                      show_col_types = FALSE) |>
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia")),
    ip_type   = factor(ip_type,   levels = c("igg_mock", "flag_eif3d", "eif4e")),
    hours     = factor(hours)
  )

metadata_filt  <- metadata  |> dplyr::filter(!grepl(OUTLIER_ID, id))
allCounts_filt <- allCounts[, !grepl(OUTLIER_ID, colnames(allCounts))]

# Belt-and-suspenders protein-coding filter (same rationale as
# 18_ip_input_normalization.Rmd / 23): allCounts is already pc-filtered
# upstream in 00_ripseq_qc.Rmd, so this is a no-op in practice, but guards
# against that upstream assumption silently changing.
allCounts_gene_id_clean <- sub("\\..*", "", rownames(allCounts_filt))
allCounts_filt <- allCounts_filt[allCounts_gene_id_clean %in% pc_genes, , drop = FALSE]

cat("RIP-seq counts after outlier removal + protein-coding filter:",
    nrow(allCounts_filt), "genes x", ncol(allCounts_filt), "samples\n")
```

---

# Section 2: Gene sets

Identical construction to `23_eif4e_delta_hypoxia_vs_normoxia_cdf.Rmd` Section 2 /
`19_density_3group_ip_enrichment.Rmd` Section 2 (same as notebook 07's `create_target`
chunk) -- only the hypoxia-defined positive set is needed here.

```{r load_genesets}
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> pull(ensembl_gene)

fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

negctrl_clean <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pull(gene_id_clean)

neg_in_features     <- negctrl_clean[negctrl_clean %in% fm_ripseq$gene_id_clean]
pos_in_features_hyp <- promotes_hyp[promotes_hyp %in% fm_ripseq$gene_id_clean]
number_to_match_hyp <- min(length(pos_in_features_hyp), length(neg_in_features))
set.seed(9)
neg_hyp <- sample(neg_in_features, number_to_match_hyp)

cat("Positive set (hypoxia-promoted):", length(promotes_hyp), "genes\n")
cat("Size-matched negative control:", length(neg_hyp), "genes\n")

stopifnot("positive and negative-control sets must be mutually exclusive" =
  length(intersect(promotes_hyp, neg_hyp)) == 0)
cat("Gene set mutual-exclusivity check passed\n")
```

---

# Section 3: FLAG-eIF3d IP/IgG delta (recomputed)

Adapted from `23_eif4e_delta_hypoxia_vs_normoxia_cdf.Rmd`'s `deseq_4e_delta` chunk, with
`flag_eif3d` in place of `eif4e`. Design: `~ condition + ip_type + condition:ip_type`,
igg_mock vs flag_eif3d, per hypoxia timepoint vs normoxia_1hr; the interaction
coefficient is the delta.

```{r deseq_flag3d_delta}
res_flag3d_list <- list()

for (hyp_hr in HYP_HOURS) {
  meta_sub <- metadata_filt |>
    dplyr::filter((condition == "normoxia" & hours == "1") |
                  (condition == "hypoxia"  & hours == hyp_hr),
                  ip_type %in% c("flag_eif3d", "igg_mock")) |>
    mutate(ip_type = droplevels(factor(ip_type, levels = c("igg_mock", "flag_eif3d")))) |>
    as.data.frame()

  stopifnot(
    "Expected 5 normoxia (2 igg_mock + 3 flag_eif3d) and 6 hypoxia (3 igg_mock + 3 flag_eif3d) samples" =
      sum(meta_sub$condition == "normoxia") == 5 &&
      sum(meta_sub$condition == "hypoxia")  == 6
  )

  # Build cts_sub one sample at a time, in meta_sub row order, with an
  # explicit per-sample column-count check -- catches an id that resolves to
  # zero or multiple columns before it can silently misalign counts with
  # metadata (same rationale as 18_ip_input_normalization.Rmd / 23).
  cts_cols <- lapply(meta_sub$id, function(x) {
    hits <- grep(x, colnames(allCounts_filt))
    stopifnot("each sample id must resolve to exactly one count-matrix column" =
      length(hits) == 1)
    allCounts_filt[, hits, drop = FALSE]
  })
  cts_sub <- do.call(cbind, cts_cols)
  rownames(meta_sub) <- colnames(cts_sub)

  dds <- DESeqDataSetFromMatrix(cts_sub, meta_sub,
                                 design = ~ condition + ip_type + condition:ip_type)
  dds <- DESeq(dds, quiet = TRUE)

  coef_int <- grep("\\.", resultsNames(dds), value = TRUE)
  stopifnot("Expected exactly one interaction coefficient" = length(coef_int) == 1)

  res <- as.data.frame(lfcShrink(dds, coef = coef_int, type = "apeglm")) |>
    rownames_to_column("gene_id") |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    left_join(gene_anno, by = "gene_id_clean") |>
    dplyr::rename(delta_flag3d_lfc = log2FoldChange, delta_flag3d_padj = padj) |>
    mutate(timepoint = paste0("hypoxia_", hyp_hr, "hr_vs_normoxia"))

  res_flag3d_list[[hyp_hr]] <- res

  write_csv(res, here("ripseq", "output",
                       paste0("24_flag3d_hypoxia_", hyp_hr, "hr_vs_normoxia_delta.csv")))
  cat(hyp_hr, "hr: samples =", ncol(cts_sub), "| genes =", nrow(res),
      "| padj<0.05 =", sum(res$delta_flag3d_padj < 0.05, na.rm = TRUE), "\n")
}
```

```{r validate_flag3d_delta}
for (hyp_hr in HYP_HOURS) {
  res <- res_flag3d_list[[hyp_hr]]
  stopifnot(
    "delta_flag3d_lfc must be finite where non-NA" =
      all(is.finite(res$delta_flag3d_lfc[!is.na(res$delta_flag3d_lfc)])),
    "gene count should be in a reasonable range" = nrow(res) > 5000
  )
}
cat("FLAG-eIF3d delta computational checks passed\n")
```

**Consistency check** (not a biological check -- see below): Spearman correlation
between the freshly recomputed delta and the old (stale) nb09 CSV. This only confirms
the recompute is a refinement of the same interaction-model methodology, not a break --
it would *not* catch a bug shared by both runs (e.g. the same wrong reference factor
level), since both use the identical `~ condition + ip_type + condition:ip_type`
pipeline.

```{r consistency_check_flag3d_delta}
old_paths <- setNames(
  here("ripseq", "output",
       paste0("ripseq_deseq_flag3d_hypoxia_", HYP_HOURS, "hr_vs_normoxia_delta.csv")),
  HYP_HOURS
)

for (hyp_hr in HYP_HOURS) {
  stopifnot("expected old nb09 delta CSV to exist for comparison" =
    file.exists(old_paths[[hyp_hr]]))

  old <- read_csv(old_paths[[hyp_hr]], show_col_types = FALSE) |>
    dplyr::select(gene_id_clean, delta_flag3d_lfc_old = delta_flag3d_lfc)

  cmp <- res_flag3d_list[[hyp_hr]] |>
    dplyr::select(gene_id_clean, delta_flag3d_lfc) |>
    inner_join(old, by = "gene_id_clean") |>
    dplyr::filter(is.finite(delta_flag3d_lfc), is.finite(delta_flag3d_lfc_old))

  rho <- cor(cmp$delta_flag3d_lfc, cmp$delta_flag3d_lfc_old, method = "spearman")
  cat(hyp_hr, "hr: Spearman rho vs old nb09 delta (same methodology) =", round(rho, 3),
      "(n =", nrow(cmp), ")\n")
  stopifnot("recomputed FLAG-eIF3d delta should correlate strongly with the prior (stale) run" =
    rho > 0.7)
}
cat("FLAG-eIF3d delta consistency check passed\n")
```

**Independent cross-check**: this is the check that actually stands in for a biological
spot-check here. Rather than asserting an invariant about one specific gene (there is
no clean a priori expectation for how *any individual* gene's FLAG-eIF3d binding should
change between conditions), this validates the interaction-model delta against a
*structurally independent* computation: notebook 01's "final" per-timepoint
FLAG-eIF3d-vs-IgG two-group DESeq2 model, saved to
`ripseq_deseq_{tp}_flag_eif3d_vs_igg.csv`, reconstructed as `lfc_hyp - lfc_nor`. Unlike
the consistency check above, this uses an entirely different DESeq2 design (two separate
two-group fits vs. one joint interaction fit), so a bug specific to the interaction-model
approach (e.g. a flipped reference level in the `condition:ip_type` term) would show up
here even if it also existed in the old nb09 file.

```{r independent_crosscheck_flag3d_delta}
nor_1hr_igg <- read_csv(
  here("ripseq", "output", "ripseq_deseq_normoxia_1hr_flag_eif3d_vs_igg.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, lfc_nor = log2FoldChange)

for (hyp_hr in HYP_HOURS) {
  hyp_igg <- read_csv(
    here("ripseq", "output", paste0("ripseq_deseq_hypoxia_", hyp_hr, "hr_flag_eif3d_vs_igg.csv")),
    show_col_types = FALSE
  ) |>
    mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
    dplyr::select(gene_id_clean, lfc_hyp = log2FoldChange)

  cmp <- res_flag3d_list[[hyp_hr]] |>
    dplyr::select(gene_id_clean, delta_flag3d_lfc) |>
    inner_join(hyp_igg, by = "gene_id_clean") |>
    inner_join(nor_1hr_igg, by = "gene_id_clean") |>
    mutate(reconstructed = lfc_hyp - lfc_nor) |>
    dplyr::filter(is.finite(delta_flag3d_lfc), is.finite(reconstructed))

  rho <- cor(cmp$delta_flag3d_lfc, cmp$reconstructed, method = "spearman")
  cat(hyp_hr, "hr: Spearman rho vs independently-reconstructed two-model delta =",
      round(rho, 3), "(n =", nrow(cmp), ")\n")
  stopifnot("interaction-model delta should correlate strongly with the independently reconstructed delta" =
    rho > 0.7)
}
cat("FLAG-eIF3d delta independent cross-check passed\n")
```

---

# Section 4: Group-assignment + CDF helper

```{r build_group_df}
build_group_df <- function(df, value_col) {
  df |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    mutate(
      group = dplyr::case_when(
        gene_id_clean %in% promotes_hyp ~ "Positive",
        gene_id_clean %in% neg_hyp      ~ "Negative control",
        TRUE                             ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

```{r validation_groups}
for (hyp_hr in HYP_HOURS) {
  d1 <- build_group_df(res_flag3d_list[[hyp_hr]], "delta_flag3d_lfc")
  stopifnot(
    "Positive group must be non-empty (FLAG-eIF3d)" = sum(d1$group == "Positive") > 0,
    "Negative control group must be non-empty (FLAG-eIF3d)" = sum(d1$group == "Negative control") > 0
  )
}
cat("Group-assignment checks passed\n")
```

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, x_label, plot_title, out_name,
                           width = 7, height = 5.5) {
  n_labels <- c(
    "Positive"         = paste0("eIF3d hypoxia promotes (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = c(-2, 2)) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 5: Generate plots

```{r plot_registry_init}
plot_registry <- list()
```

## FLAG-eIF3d IP/IgG delta

```{r plots_flag3d_delta}
for (hyp_hr in HYP_HOURS) {
  df <- build_group_df(res_flag3d_list[[hyp_hr]], "delta_flag3d_lfc")
  p <- make_cdf_plot(
    df, "delta_flag3d_lfc",
    x_label    = paste0("FLAG-eIF3d binding change (log2FC, hypoxia ", hyp_hr, "hr vs normoxia 1hr)"),
    plot_title = paste0("FLAG-eIF3d IP/IgG delta: eIF3d hypoxia-promoted targets (hypoxia ",
                         hyp_hr, "hr vs normoxia)"),
    out_name   = paste0("24_cdf_flag3d_delta_", hyp_hr, "hr.pdf")
  )
  plot_registry[[paste0("flag3d_", hyp_hr)]] <- p
  print(p)
}
```

---

# Section 6: Composite grid (patchwork) -- Grid 1

One row (delta) x 3 columns (hypoxia 1hr / 4hr / 24hr) -- no normoxia column, since
delta is inherently relative to the normoxia_1hr baseline. Panels drop their individual
title/legend/verbose axis label for the composite (redundant once there's a column
header); one shared legend sits at the bottom.

```{r composite_helper}
make_text_panel <- function(label, angle = 0, size = 5) {
  ggplot() +
    theme_void() +
    annotate("text", x = 0, y = 0, label = label, size = size, fontface = "bold", angle = angle)
}

strip_for_composite <- function(p) {
  p + labs(title = NULL, subtitle = NULL, x = "log2FC") + theme(legend.position = "none")
}

make_composite_grid <- function(prefix, title, out_name, plots = plot_registry) {
  keys <- paste0(prefix, "_", HYP_HOURS)
  stopifnot("all requested plots must exist in plot registry" = all(keys %in% names(plots)))

  cells <- c(list(make_text_panel("")), lapply(col_headers, make_text_panel))
  cells <- c(cells,
             list(make_text_panel("delta", angle = 90)),
             lapply(plots[keys], strip_for_composite))

  main_grid <- patchwork::wrap_plots(
    cells, ncol = 4,
    widths  = c(0.15, 1, 1, 1),
    heights = c(0.15, 1)
  )

  legend_src <- plots[[keys[1]]] +
    scale_color_manual(values = group_colors, breaks = names(group_colors)) +
    theme(legend.position = "bottom")
  legend_grob <- cowplot::get_legend(legend_src)

  full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
    patchwork::plot_layout(heights = c(1, 0.06)) +
    patchwork::plot_annotation(title = title)

  ggsave(here("ripseq", "plots", out_name), full, width = 3 * 4.2 + 1, height = 6,
         limitsize = FALSE)
  full
}
```

```{r composite_flag3d, fig.width=14, fig.height=6}
make_composite_grid(
  "flag3d",
  title    = "FLAG-eIF3d IP/IgG binding delta: eIF3d hypoxia-promoted targets, all hypoxia timepoints",
  out_name = "24_grid_flag3d_delta.pdf"
)
```

---

# Section 7: Gene-highlight composite grid -- Grid 2

Same 3-group CDF background as Grid 1, but with RPP25, COL12A1, and TXNIP marked as
points sitting directly on the "Other" curve at each gene's own `delta_flag3d_lfc`
value/cumulative-fraction. CDKN2B is excluded (absent from both the RIP-seq and
RNA-seq count matrices in this repo -- confirmed upstream, not a filtering artifact
within this notebook).

An earlier version of this plot marked genes with full-height vertical lines and
on-plot text labels; with two of the three genes landing far outside the `[-2, 2]`
display window at several timepoints, that approach needed increasingly elaborate
stagger/repel logic to keep lines and labels from colliding or crossing, and still
looked cluttered. Points-on-curve + a standard ggplot fill legend sidesteps all of
that: a gene's marker is a filled circle if `delta_flag3d_lfc` falls inside `[-2, 2]`,
or a triangle clamped to the near axis edge if it doesn't (its exact value is then
printed in the subtitle instead of crowding the plot itself).

```{r highlight_gene_lookup}
highlight_ids <- gene_anno |>
  dplyr::filter(symbol %in% highlight_genes) |>
  distinct(symbol, .keep_all = TRUE)

stopifnot("all highlight genes must resolve to a gene_id_clean" =
  nrow(highlight_ids) == length(highlight_genes))
cat("Highlight gene lookup:\n")
print(highlight_ids)
```

```{r make_highlight_cdf_plot}
highlight_colors <- setNames(
  RColorBrewer::brewer.pal(max(3, length(highlight_genes)), "Dark2")[seq_along(highlight_genes)],
  highlight_genes
)

make_highlight_cdf_plot <- function(df, value_col, x_label, plot_title, out_name,
                                     width = 7, height = 5.5, xlim = c(-2, 2)) {
  n_labels <- c(
    "Positive"         = paste0("eIF3d hypoxia promotes (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  # Each gene is placed exactly on the curve: x = its own delta (clamped into
  # the display window if off-scale), y = the empirical CDF's actual
  # cumulative fraction at that (unclamped) value. Off-scale genes get a
  # triangle instead of a circle; their true value goes in the subtitle
  # rather than as an on-plot label, so there is nothing left to collide.
  # Two genes clamped to the same edge would otherwise land at (near-)
  # identical points and hide one another -- stagger those ties slightly.
  ecdf_fun <- ecdf(df[[value_col]])
  gene_pts <- highlight_ids |>
    left_join(dplyr::select(df, gene_id_clean, all_of(value_col)), by = "gene_id_clean") |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::rename(raw_value = all_of(value_col)) |>
    mutate(
      off_scale  = raw_value < xlim[1] | raw_value > xlim[2],
      clamp_side = dplyr::case_when(
        raw_value < xlim[1] ~ "low",
        raw_value > xlim[2] ~ "high",
        TRUE                ~ "none"
      ),
      plot_y = ecdf_fun(raw_value)
    ) |>
    arrange(symbol) |>
    group_by(clamp_side) |>
    mutate(stagger_idx = row_number() - 1L) |>
    ungroup() |>
    mutate(plot_x = dplyr::case_when(
      clamp_side == "low"  ~ xlim[1] + 0.05 + stagger_idx * 0.12,
      clamp_side == "high" ~ xlim[2] - 0.05 - stagger_idx * 0.12,
      TRUE                 ~ raw_value
    ))

  if (nrow(gene_pts) < nrow(highlight_ids)) {
    missing <- setdiff(highlight_ids$symbol, gene_pts$symbol)
    cat("  Note:", paste(missing, collapse = ", "),
        "not finite/present for this timepoint -- omitted from plot\n")
  }

  subtitle_txt <- paste0(
    paste0(gene_pts$symbol, " = ", round(gene_pts$raw_value, 2)), collapse = "   "
  )
  if (any(gene_pts$off_scale)) {
    cat("  Off-scale (marker clamped to axis edge as a triangle, true value in subtitle):",
        paste(gene_pts$symbol[gene_pts$off_scale], collapse = ", "), "\n")
  }

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    geom_point(
      data = gene_pts, aes(x = plot_x, y = plot_y, fill = symbol, shape = off_scale),
      inherit.aes = FALSE, color = "black", size = 3, stroke = 0.6
    ) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    scale_fill_manual(values = highlight_colors, breaks = highlight_genes, name = NULL) +
    scale_shape_manual(values = c(`FALSE` = 21, `TRUE` = 24), guide = "none") +
    guides(fill = guide_legend(override.aes = list(shape = 21))) +
    coord_cartesian(xlim = xlim, ylim = c(0, 1), clip = "off") +
    labs(x = x_label, y = "Cumulative fraction", color = NULL,
         title = plot_title, subtitle = subtitle_txt) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom", legend.box = "vertical")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

```{r plots_highlight}
highlight_registry <- list()

for (hyp_hr in HYP_HOURS) {
  df <- build_group_df(res_flag3d_list[[hyp_hr]], "delta_flag3d_lfc")
  cat("FLAG-eIF3d delta,", hyp_hr, "hr -- gene highlight check:\n")
  p <- make_highlight_cdf_plot(
    df, "delta_flag3d_lfc",
    x_label    = paste0("FLAG-eIF3d binding change (log2FC, hypoxia ", hyp_hr, "hr vs normoxia 1hr)"),
    plot_title = paste0("FLAG-eIF3d delta with gene highlights (hypoxia ", hyp_hr, "hr vs normoxia)"),
    out_name   = paste0("24_cdf_flag3d_delta_highlight_", hyp_hr, "hr.pdf")
  )
  highlight_registry[[paste0("flag3d_", hyp_hr)]] <- p
  print(p)
}
```

```{r composite_highlight, fig.width=14, fig.height=6}
make_composite_grid(
  "flag3d",
  title    = "FLAG-eIF3d IP/IgG binding delta with gene highlights (RPP25, COL12A1, TXNIP)",
  out_name = "24_grid_flag3d_delta_highlight.pdf",
  plots    = highlight_registry
)
```
