---
title: "RIP-seq: eIF4E IP/RNA CDF, our data vs Teleman, side by side"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Side-by-side comparison of our eIF4E IP/RNA-input enrichment against the published
Teleman-lab eIF4E-IP, using the **same** eIF3d hypoxia-promoted gene sets (positive /
negative control / other, as in `19_density_3group_ip_enrichment.Rmd`) as the CDF
groups in every panel. Covers all 4 combinations of:

- **Our RIP-seq timepoint**: hypoxia 1hr (`eif4e_ip_rna_lfc_hypoxia_1hr`) or
  normoxia 1hr (`eif4e_ip_rna_lfc_normoxia_1hr`).
- **Teleman condition**: vehicle/unstressed (`eif4e_ip_vehicle_lfc`, from
  `res_eif4e_ip_vehicle.csv`) or eIF4E-inhibitor-treated
  (`eif4e_ip_inhibitor_only_lfc`, from `res_eif4e_ip_inhibitor_only.csv`).

Unlike notebook 20 (which uses Teleman's own enriched/depleted gene calls as the CDF
groups and puts *our* data on the x-axis), this notebook keeps the group definitions
fixed to our eIF3d hypoxia-promoted positive/negative sets in every panel and instead
puts the *Teleman* or *our* enrichment values on the x-axis -- a direct "does our
result look like theirs" comparison for the same gene groups, across both of our
RIP-seq timepoints and both of Teleman's treatment conditions.

**Gene groups are always the hypoxia-promoted set**, even in the normoxia-timepoint
panels -- same convention as notebook 19 (which gene set is "Positive" is an
independent choice from which timepoint/condition's data is being plotted).

**Label correction (verified against source):** notebooks `03_eif4e_rip_analysis.Rmd`
and `20_eif4e_teleman_validation.Rmd` label the Teleman `log2FoldChange` columns as
"IP/IgG". Tracing `/Users/katematlin/github/mpra_preliminary/code/eif4e_ip_clip_analysis.Rmd`
(the script that writes `res_eif4e_ip_vehicle.csv` and `res_eif4e_ip_inhibitor_only.csv`)
confirms both are `design = ~type` fits (vehicle-only / BP_inhib-only samples,
respectively), contrasting `eIF4E_IPed_RNA` (IP) against `Total_RNA` (input) -- there is
no IgG condition anywhere in the Teleman experiment. So both columns are eIF4E
**IP/RNA-input**, directly analogous to our `eif4e_ip_rna_lfc_*` columns, not IP/IgG.
This notebook uses the corrected "IP/RNA" label; notebooks 03/20 are left as-is per
current scope (label fix was applied here only).

**Interpretation note for the inhibitor panels**: the inhibitor (BP_inhib /
4EBP1-4A-overexpression, per Teleman's naming) disrupts eIF4E's cap-binding/eIF4G
interaction, so eIF4E-bound mRNAs under inhibitor treatment are not the same population
as under vehicle. The inhibitor-paired panels are not replicates of the vehicle-paired
panels -- they ask a different question (does the eIF3d-promoted signature still track
eIF4E binding when cap-dependent binding is disrupted).

**All panels use the same fixed x-axis range** (`xlim = c(-2, 2)`, the repo's
established convention for these RIP-seq LFC metrics -- see notebooks 19/20/23) so any
two panels are directly visually comparable.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Section 1: Load RIP-seq feature matrix

All 4 x-axis metrics used in this notebook -- our `eif4e_ip_rna_lfc_hypoxia_1hr` /
`eif4e_ip_rna_lfc_normoxia_1hr` and Teleman's `eif4e_ip_vehicle_lfc` /
`eif4e_ip_inhibitor_only_lfc` -- are already columns of `feature_matrix_ripseq.rds` (the
Teleman columns joined onto our gene universe by symbol back in
`code/predictive_modeling/01b_feature_extraction_targeted.Rmd`), so no separate raw-file
loading or DESeq2 fitting is needed here.

```{r libraries}
library(here)
library(tidyverse)
library(patchwork)
library(cowplot)
```

```{r constants}
group_colors <- c(
  "Positive"         = "#E69F00",
  "Negative control" = "#000000",
  "Other"            = "grey80"
)

X_LIMITS <- c(-2, 2)

# Our RIP-seq timepoints: value column (IP/RNA) + the IP/IgG column used for the
# EIF4E self-enrichment spot-check (IP/RNA is the noisier, batch-confounded metric
# -- see Section 4).
our_defs <- tibble::tibble(
  our_key     = c("hypoxia_1hr", "normoxia_1hr"),
  our_col     = c("eif4e_ip_rna_lfc_hypoxia_1hr", "eif4e_ip_rna_lfc_normoxia_1hr"),
  our_igg_col = c("eif4e_ip_igg_lfc_hypoxia_1hr", "eif4e_ip_igg_lfc_normoxia_1hr"),
  our_label   = c("Our data: eIF4E IP/RNA (hypoxia, 1hr)", "Our data: eIF4E IP/RNA (normoxia, 1hr)"),
  our_xlab    = c("log2FC (eIF4E IP / sictrl RNA, DESeq2, hypoxia 1hr)",
                  "log2FC (eIF4E IP / sictrl RNA, DESeq2, normoxia 1hr)")
)

# Teleman conditions: value column (IP/RNA-input) + the source CSV's independently
# verified EIF4E self-enrichment (reported in Section 4, not re-derived here).
teleman_defs <- tibble::tibble(
  tele_key   = c("vehicle", "inhibitor"),
  tele_col   = c("eif4e_ip_vehicle_lfc", "eif4e_ip_inhibitor_only_lfc"),
  tele_label = c("Teleman: eIF4E IP/RNA (vehicle, unstressed)",
                 "Teleman: eIF4E IP/RNA (inhibitor-treated)"),
  tele_xlab  = c("log2FC (eIF4E IP / Total RNA input, DESeq2, Teleman vehicle/unstressed)",
                 "log2FC (eIF4E IP / Total RNA input, DESeq2, Teleman inhibitor-treated)")
)
```

```{r load_feature_matrix}
fm_ripseq <- readRDS(here("output", "predictive_modeling", "feature_matrix_ripseq.rds")) |>
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Feature matrix loaded:", nrow(fm_ripseq), "unique genes\n")

value_cols <- c(our_defs$our_col, teleman_defs$tele_col)

stopifnot(
  "all 4 required value columns must be present" = all(value_cols %in% colnames(fm_ripseq)),
  "required columns must be finite where non-NA" = all(sapply(
    value_cols,
    function(col) all(is.finite(fm_ripseq[[col]][!is.na(fm_ripseq[[col]])]))
  ))
)
cat("Feature matrix column checks passed\n")
```

---

# Section 2: Gene sets

Identical construction to `19_density_3group_ip_enrichment.Rmd` Section 2's hypoxia
gene set -- same positive set, same `set.seed(9)` size-matched negative control, so the
"Negative control" group here is exactly the one already used everywhere else in this
repo for the hypoxia-promoted positive set (and what the RF model itself trains
against). Used for every panel in this notebook, regardless of which
timepoint/condition's data is plotted.

```{r load_genesets}
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
) |> pull(ensembl_gene)

negctrl_clean <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  pull(gene_id_clean)

neg_in_features     <- negctrl_clean[negctrl_clean %in% fm_ripseq$gene_id_clean]
pos_in_features_hyp <- promotes_hyp[promotes_hyp %in% fm_ripseq$gene_id_clean]
number_to_match_hyp <- min(length(pos_in_features_hyp), length(neg_in_features))
set.seed(9)
neg_hyp <- sample(neg_in_features, number_to_match_hyp)

cat("Positive set (hypoxia-promoted):", length(promotes_hyp), "genes\n")
cat("Size-matched negative control:", length(neg_hyp), "genes\n")

stopifnot("positive and negative-control sets must be mutually exclusive" =
  length(intersect(promotes_hyp, neg_hyp)) == 0)
cat("Gene set mutual-exclusivity check passed\n")
```

---

# Section 3: Group-assignment helper

```{r build_group_df}
build_group_df <- function(fm, value_col) {
  fm |>
    dplyr::filter(is.finite(.data[[value_col]])) |>
    dplyr::mutate(
      group = dplyr::case_when(
        gene_id_clean %in% promotes_hyp ~ "Positive",
        gene_id_clean %in% neg_hyp      ~ "Negative control",
        TRUE                             ~ "Other"
      ),
      group = factor(group, levels = c("Positive", "Negative control", "Other"))
    )
}
```

---

# Section 4: Validation

```{r validation_computational}
for (col in value_cols) {
  d <- build_group_df(fm_ripseq, col)
  stopifnot(
    "group column must have no NA" = !any(is.na(d$group)),
    "Positive group must be non-empty" = sum(d$group == "Positive") > 0,
    "Negative control group must be non-empty" = sum(d$group == "Negative control") > 0,
    "values must be finite" = all(is.finite(d[[col]]))
  )
}
cat("Computational checks passed\n")
```

EIF4E self-enrichment, checked for all 4 conditions. Our data uses the **IP/IgG**
column at each timepoint (established in notebooks 19/20 as the reliable metric for
this spot-check) rather than IP/RNA -- per the repo's standing caveat, IP/RNA carries a
RIP-seq-vs-RNA-seq batch confound (no shared spike-in) and is noisier/less directly
interpretable (reported below for reference, not gated). Both Teleman columns are
single clean two-group DESeq2 fits (no batch confound) and are gated directly,
independently verified against their source CSVs
(`res_eif4e_ip_vehicle.csv`: LFC = 1.74, padj ~ 1e-28;
`res_eif4e_ip_inhibitor_only.csv`: LFC = 2.95, padj ~ 7.6e-43).

```{r validation_biological}
for (i in seq_len(nrow(our_defs))) {
  self_igg <- fm_ripseq |>
    dplyr::filter(symbol == "EIF4E") |>
    pull(.data[[our_defs$our_igg_col[i]]])
  stopifnot(
    "EIF4E self-enrichment in our eif4e IP/IgG must be positive" =
      length(self_igg) > 0 && all(self_igg > 0, na.rm = TRUE)
  )
  cat("EIF4E self-enrichment (ours, IP/IgG,", our_defs$our_key[i], "):", self_igg, "\n")

  self_rna <- fm_ripseq |>
    dplyr::filter(symbol == "EIF4E") |>
    pull(.data[[our_defs$our_col[i]]])
  cat("EIF4E self-enrichment (ours, IP/RNA,", our_defs$our_key[i],
      ", not gated -- noisy metric):", self_rna, "\n")
}

for (i in seq_len(nrow(teleman_defs))) {
  self_tele <- fm_ripseq |>
    dplyr::filter(symbol == "EIF4E") |>
    pull(.data[[teleman_defs$tele_col[i]]])
  stopifnot(
    "EIF4E self-enrichment in Teleman IP/RNA must be positive" =
      length(self_tele) > 0 && all(self_tele > 0, na.rm = TRUE)
  )
  cat("EIF4E self-enrichment (Teleman,", teleman_defs$tele_key[i], ", IP/RNA):", self_tele, "\n")
}

cat("Biological spot-checks passed\n")
```

Informational only (not gated by `stopifnot`, since there is no established prior for
how strongly these independently-generated eIF4E IP/RNA metrics should correlate,
especially for the inhibitor condition which perturbs the cap-binding population):
Spearman correlation for each of the 4 our-timepoint x Teleman-condition combinations.

```{r consistency_correlation}
for (i in seq_len(nrow(our_defs))) {
  for (j in seq_len(nrow(teleman_defs))) {
    cmp <- fm_ripseq |>
      dplyr::filter(is.finite(.data[[our_defs$our_col[i]]]),
                    is.finite(.data[[teleman_defs$tele_col[j]]]))
    rho <- cor(cmp[[our_defs$our_col[i]]], cmp[[teleman_defs$tele_col[j]]], method = "spearman")
    cat("Spearman rho (ours,", our_defs$our_key[i], "vs Teleman,", teleman_defs$tele_key[j],
        "):", round(rho, 3), "(n =", nrow(cmp), ")\n")
  }
}
```

---

# Section 5: Plotting helper

```{r make_cdf_plot}
make_cdf_plot <- function(df, value_col, x_label, plot_title, out_name,
                           width = 7.5, height = 5.5, xlim = X_LIMITS) {
  n_labels <- c(
    "Positive"         = paste0("eIF3d hypoxia promotes (n=", sum(df$group == "Positive"), ")"),
    "Negative control" = paste0("Negative control (n=", sum(df$group == "Negative control"), ")"),
    "Other"            = paste0("All other genes (n=", sum(df$group == "Other"), ")")
  )

  wt_pos_other <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Other"])
  wt_pos_neg   <- wilcox.test(df[[value_col]][df$group == "Positive"],
                               df[[value_col]][df$group == "Negative control"])
  cat(plot_title, "\n")
  cat("  Positive vs Other p =", signif(wt_pos_other$p.value, 3), "\n")
  cat("  Positive vs Negative p =", signif(wt_pos_neg$p.value, 3), "\n")

  n_label <- paste0(
    "Positive n=", sum(df$group == "Positive"), "\n",
    "Negative control n=", sum(df$group == "Negative control"), "\n",
    "Other n=", sum(df$group == "Other")
  )

  p <- ggplot(df, aes(x = .data[[value_col]], color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
    annotate("text", x = Inf, y = 0, hjust = 1.1, vjust = -0.4, label = n_label, size = 3.5) +
    scale_color_manual(values = group_colors, breaks = names(group_colors), labels = n_labels) +
    coord_cartesian(xlim = xlim) +
    labs(x = x_label, y = "Cumulative fraction", color = NULL, title = plot_title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  ggsave(here("ripseq", "plots", out_name), p, width = width, height = height)
  p
}
```

---

# Section 6: Generate individual panels

Each of the 4 underlying distributions (2 our-timepoints + 2 Teleman-conditions) is
plotted once and cached in `plot_registry`, then reused across composites in Section 7
-- e.g. the "our hypoxia 1hr" panel is identical whether it's paired with Teleman
vehicle or Teleman inhibitor, so it is not recomputed twice.

```{r plot_registry_init}
plot_registry <- list()
```

```{r plots_ours}
for (i in seq_len(nrow(our_defs))) {
  df <- build_group_df(fm_ripseq, our_defs$our_col[i])
  p <- make_cdf_plot(
    df, our_defs$our_col[i],
    x_label    = our_defs$our_xlab[i],
    plot_title = our_defs$our_label[i],
    out_name   = paste0("25_cdf_eif4e_ours_", our_defs$our_key[i], ".pdf")
  )
  plot_registry[[paste0("ours_", our_defs$our_key[i])]] <- p
  print(p)
}
```

```{r plots_teleman}
for (j in seq_len(nrow(teleman_defs))) {
  df <- build_group_df(fm_ripseq, teleman_defs$tele_col[j])
  p <- make_cdf_plot(
    df, teleman_defs$tele_col[j],
    x_label    = teleman_defs$tele_xlab[j],
    plot_title = teleman_defs$tele_label[j],
    out_name   = paste0("25_cdf_eif4e_teleman_", teleman_defs$tele_key[j], ".pdf")
  )
  plot_registry[[paste0("teleman_", teleman_defs$tele_key[j])]] <- p
  print(p)
}
```

---

# Section 7: Side-by-side composites (all 4 combinations)

```{r composite_helper}
strip_legend <- function(p) p + theme(legend.position = "none")

make_side_by_side <- function(our_key, tele_key, title, out_name) {
  p_ours    <- plot_registry[[paste0("ours_", our_key)]]
  p_teleman <- plot_registry[[paste0("teleman_", tele_key)]]
  stopifnot("both panels must exist in plot_registry" = !is.null(p_ours) && !is.null(p_teleman))

  main_grid <- patchwork::wrap_plots(strip_legend(p_ours), strip_legend(p_teleman), ncol = 2)

  legend_grob <- cowplot::get_legend(
    p_ours + scale_color_manual(values = group_colors, breaks = names(group_colors)) +
      theme(legend.position = "bottom")
  )

  full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
    patchwork::plot_layout(heights = c(1, 0.08)) +
    patchwork::plot_annotation(title = title)

  ggsave(here("ripseq", "plots", out_name), full, width = 14, height = 6.5, limitsize = FALSE)
  full
}
```

## Hypoxia 1hr (ours) vs vehicle (Teleman)

```{r composite_hyp_vehicle, fig.width=14, fig.height=6.5}
make_side_by_side(
  "hypoxia_1hr", "vehicle",
  title    = "eIF4E IP/RNA: eIF3d hypoxia-promoted targets, ours (hypoxia 1hr) vs Teleman (vehicle/unstressed)",
  out_name = "25_composite_eif4e_hypoxia1hr_vs_teleman_vehicle.pdf"
)
```

## Hypoxia 1hr (ours) vs inhibitor-treated (Teleman)

```{r composite_hyp_inhibitor, fig.width=14, fig.height=6.5}
make_side_by_side(
  "hypoxia_1hr", "inhibitor",
  title    = "eIF4E IP/RNA: eIF3d hypoxia-promoted targets, ours (hypoxia 1hr) vs Teleman (inhibitor-treated)",
  out_name = "25_composite_eif4e_hypoxia1hr_vs_teleman_inhibitor.pdf"
)
```

## Normoxia 1hr (ours) vs vehicle (Teleman)

```{r composite_nor_vehicle, fig.width=14, fig.height=6.5}
make_side_by_side(
  "normoxia_1hr", "vehicle",
  title    = "eIF4E IP/RNA: eIF3d hypoxia-promoted targets, ours (normoxia 1hr) vs Teleman (vehicle/unstressed)",
  out_name = "25_composite_eif4e_normoxia1hr_vs_teleman_vehicle.pdf"
)
```

## Normoxia 1hr (ours) vs inhibitor-treated (Teleman)

```{r composite_nor_inhibitor, fig.width=14, fig.height=6.5}
make_side_by_side(
  "normoxia_1hr", "inhibitor",
  title    = "eIF4E IP/RNA: eIF3d hypoxia-promoted targets, ours (normoxia 1hr) vs Teleman (inhibitor-treated)",
  out_name = "25_composite_eif4e_normoxia1hr_vs_teleman_inhibitor.pdf"
)
```

---

# Section 8: Delta panels (stress-induced change, ours vs Teleman)

Both perturbations disrupt eIF4E's normal mRNA-binding landscape relative to a matched
baseline -- our hypoxia condition relative to normoxia, and Teleman's eIF4E-inhibitor
(4EBP1-4A overexpression, which blocks eIF4E's cap-binding/eIF4G interaction) relative to
vehicle. Both sides are genuine DESeq2 interaction-model delta-deltas (a single joint fit
with a `condition:seq_type`-style interaction coefficient), not two separately-fit
log2FCs subtracted afterward -- same principle as notebook 23's IP/IgG and IgG/RNA
delta-deltas.

- **Teleman side**: `eif4e_ip_inhib_vs_vehicle_lfc`, already present in
  `feature_matrix_ripseq.rds` (joined in `01b_feature_extraction_targeted.Rmd` from
  `res_eif4e_ip_inhib_vs_vehicle.csv`). That CSV comes from a joint DESeq2 fit
  (`design = ~treatment + type + treatment:type`, `treatment` = vehicle/BP_inhib,
  `type` = input/IP) in `mpra_preliminary/code/eif4e_ip_clip_analysis.Rmd` -- the
  `treatment:type` interaction coefficient is the delta-delta. Negative = binding lost
  under inhibitor (cap-dependence disrupted).
- **Our side**: `delta_ours_lfc`, computed fresh below using the identical
  `~ condition + seq_type + condition:seq_type` design notebook 23 Section 4 uses for
  the IgG/RNA delta-delta, but with actual eIF4E-IP counts (`ip_type == "eif4e"`) in
  place of IgG-mock, restricted to hypoxia_1hr vs normoxia_1hr (the only timepoint pair
  used elsewhere in this notebook). Positive = eIF4E/RNA enrichment increases in
  hypoxia relative to normoxia.

```{r delta_libraries}
library(DESeq2)
```

```{r delta_load_counts}
OUTLIER_ID <- "0077"

gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  dplyr::filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

pc_genes <- gene_anno_all |>
  dplyr::filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)

allCounts    <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
rip_metadata <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                         show_col_types = FALSE)

rip_metadata_filt <- rip_metadata |> dplyr::filter(!grepl(OUTLIER_ID, id))
allCounts_filt    <- allCounts[, !grepl(OUTLIER_ID, colnames(allCounts))]

# Belt-and-suspenders protein-coding filter (same rationale as
# 18_ip_input_normalization.Rmd / 23_eif4e_delta_hypoxia_vs_normoxia_cdf.Rmd).
allCounts_gene_id_clean <- sub("\\..*", "", rownames(allCounts_filt))
allCounts_filt <- allCounts_filt[allCounts_gene_id_clean %in% pc_genes, , drop = FALSE]

rna_counts_all <- read_csv(
  here("output", "human_allCounts_raw.csv"),
  col_select = c(gene_id, matches("sictrl.*_rna_")),
  show_col_types = FALSE
) |>
  column_to_rownames("gene_id") |>
  as.matrix()

rna_meta_delta <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE) |>
  dplyr::filter(library == "rna", treatment == "sictrl") |>
  mutate(hours = as.character(hours), condition = as.character(condition))

stopifnot("RIP-seq genes must be a subset of RNA-seq genes" =
  all(rownames(allCounts_filt) %in% rownames(rna_counts_all)))
rna_counts_delta <- rna_counts_all[rownames(allCounts_filt), , drop = FALSE]

cat("RIP-seq counts after outlier removal + protein-coding filter:",
    nrow(allCounts_filt), "genes x", ncol(allCounts_filt), "samples\n")
cat("sictrl RNA-seq counts (subset to RIP-seq genes):", nrow(rna_counts_delta), "genes x",
    ncol(rna_counts_delta), "samples\n")
```

```{r delta_ours_deseq}
eif4e_meta_delta <- rip_metadata_filt |>
  dplyr::filter(ip_type == "eif4e",
                (condition == "normoxia" & hours == "1") |
                (condition == "hypoxia"  & hours == "1")) |>
  mutate(condition = as.character(condition), seq_type = "eif4e") |>
  dplyr::select(id, condition, seq_type)

rna_meta_sub_delta <- rna_meta_delta |>
  dplyr::filter((condition == "normoxia" & hours == "1") |
                (condition == "hypoxia"  & hours == "1")) |>
  mutate(seq_type = "rna") |>
  dplyr::select(id, condition, seq_type)

stopifnot(
  "Expected 3 normoxia + 3 hypoxia eIF4E-IP replicates" =
    sum(eif4e_meta_delta$condition == "normoxia") == 3 &&
    sum(eif4e_meta_delta$condition == "hypoxia")  == 3,
  "Expected 3 normoxia + 3 hypoxia sictrl RNA replicates" =
    sum(rna_meta_sub_delta$condition == "normoxia") == 3 &&
    sum(rna_meta_sub_delta$condition == "hypoxia")  == 3
)

col_data_delta <- bind_rows(eif4e_meta_delta, rna_meta_sub_delta) |> as.data.frame()

# Build combined counts one sample at a time, in col_data_delta's row order (same
# explicit-per-sample-column-count-check rationale as notebook 23 Section 4).
cts_list_delta <- lapply(seq_len(nrow(col_data_delta)), function(i) {
  src  <- if (col_data_delta$seq_type[i] == "eif4e") allCounts_filt else rna_counts_delta
  hits <- grep(col_data_delta$id[i], colnames(src))
  stopifnot("each sample id must resolve to exactly one count-matrix column" =
    length(hits) == 1)
  src[, hits, drop = FALSE]
})
combined_cts_delta <- do.call(cbind, cts_list_delta)
rownames(col_data_delta) <- colnames(combined_cts_delta)

# Zero-RNA-coverage guard, same rationale as 18_ip_input_normalization.Rmd /
# notebook 23 Section 4: genes with no RNA counts across all sictrl reps have no
# real denominator.
rna_covered_delta <- rowSums(combined_cts_delta[, col_data_delta$seq_type == "rna", drop = FALSE]) > 0
n_dropped_delta   <- sum(!rna_covered_delta)
if (n_dropped_delta > 0) {
  cat("Dropping", n_dropped_delta, "genes with zero RNA coverage\n")
}
combined_cts_delta <- round(combined_cts_delta[rna_covered_delta, , drop = FALSE])

col_data_delta <- col_data_delta |>
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia")),
    seq_type  = factor(seq_type,  levels = c("rna", "eif4e"))
  )

dds_delta <- DESeqDataSetFromMatrix(combined_cts_delta, col_data_delta,
                                     design = ~ condition + seq_type + condition:seq_type)
dds_delta <- DESeq(dds_delta, quiet = TRUE)

coef_int_delta <- grep("\\.", resultsNames(dds_delta), value = TRUE)
stopifnot("Expected exactly one interaction coefficient" = length(coef_int_delta) == 1)

res_ours_delta <- as.data.frame(lfcShrink(dds_delta, coef = coef_int_delta, type = "apeglm")) |>
  rownames_to_column("gene_id") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::rename(delta_ours_lfc = log2FoldChange, delta_ours_padj = padj) |>
  dplyr::select(gene_id_clean, delta_ours_lfc, delta_ours_padj)

write_csv(res_ours_delta, here("ripseq", "output",
                               "25_eif4e_ip_rna_delta_hypoxia1hr_vs_normoxia1hr.csv"))

cat("Ours eIF4E IP/RNA delta-delta: samples =", ncol(combined_cts_delta),
    "| genes =", nrow(res_ours_delta),
    "| padj<0.05 =", sum(res_ours_delta$delta_ours_padj < 0.05, na.rm = TRUE), "\n")
```

```{r delta_ours_validate}
stopifnot(
  "delta_ours_lfc must be finite where non-NA" =
    all(is.finite(res_ours_delta$delta_ours_lfc[!is.na(res_ours_delta$delta_ours_lfc)])),
  "gene count should be in a reasonable range" = nrow(res_ours_delta) > 5000
)
cat("Ours eIF4E IP/RNA delta-delta computational checks passed\n")
```

**Independent cross-check** (same rationale as notebook 23's cross-checks): reconstruct
the delta as two separately-fit per-timepoint log2FCs
(`eif4e_ip_rna_lfc_hypoxia_1hr - eif4e_ip_rna_lfc_normoxia_1hr`, already-validated
columns in `feature_matrix_ripseq.rds`) and confirm it correlates strongly with the
joint interaction-model result. This uses a structurally different DESeq2 design (two
separate two-group fits vs. one joint interaction fit), so it catches bugs specific to
the interaction-model approach itself.

```{r delta_ours_crosscheck}
naive_delta <- fm_ripseq |>
  dplyr::transmute(
    gene_id_clean,
    delta_naive = eif4e_ip_rna_lfc_hypoxia_1hr - eif4e_ip_rna_lfc_normoxia_1hr
  )

cmp_delta <- res_ours_delta |>
  dplyr::select(gene_id_clean, delta_ours_lfc) |>
  inner_join(naive_delta, by = "gene_id_clean") |>
  dplyr::filter(is.finite(delta_ours_lfc), is.finite(delta_naive))

rho_delta <- cor(cmp_delta$delta_ours_lfc, cmp_delta$delta_naive, method = "spearman")
cat("Spearman rho, interaction-model delta vs naive two-fit subtraction:", round(rho_delta, 3),
    "(n =", nrow(cmp_delta), ")\n")
stopifnot("interaction-model delta should correlate strongly with the naive two-fit subtraction" =
  rho_delta > 0.7)
cat("Ours eIF4E IP/RNA delta-delta independent cross-check passed\n")
```

```{r delta_join}
fm_delta <- fm_ripseq |>
  dplyr::left_join(res_ours_delta |> dplyr::select(gene_id_clean, delta_ours_lfc),
                    by = "gene_id_clean") |>
  dplyr::rename(delta_teleman_lfc = eif4e_ip_inhib_vs_vehicle_lfc)

df_ours_delta    <- build_group_df(fm_delta, "delta_ours_lfc")
df_teleman_delta <- build_group_df(fm_delta, "delta_teleman_lfc")

stopifnot(
  "Positive group must be non-empty (ours delta)" = sum(df_ours_delta$group == "Positive") > 0,
  "Negative control group must be non-empty (ours delta)" =
    sum(df_ours_delta$group == "Negative control") > 0,
  "Positive group must be non-empty (Teleman delta)" = sum(df_teleman_delta$group == "Positive") > 0,
  "Negative control group must be non-empty (Teleman delta)" =
    sum(df_teleman_delta$group == "Negative control") > 0
)
cat("Delta group-assignment checks passed\n")
```

Both delta panels use a tighter `xlim = c(-1, 1)` (rather than the repo-standard
`c(-2, 2)`) since these interaction-term coefficients are much smaller in magnitude than
the raw enrichment metrics plotted elsewhere in this notebook.

```{r delta_plots}
p_ours_delta <- make_cdf_plot(
  df_ours_delta, "delta_ours_lfc",
  x_label    = "log2FC (eIF4E IP/RNA delta-delta, hypoxia 1hr vs normoxia 1hr)",
  plot_title = "Our data: eIF4E IP/RNA delta-delta (hypoxia 1hr vs normoxia 1hr)",
  out_name   = "25_cdf_eif4e_delta_ours_hypoxia_vs_normoxia.pdf",
  xlim       = c(-1, 1)
)

p_teleman_delta <- make_cdf_plot(
  df_teleman_delta, "delta_teleman_lfc",
  x_label    = "log2FC (eIF4E IP/RNA delta-delta, Teleman inhibitor-treated vs vehicle)",
  plot_title = "Teleman: eIF4E IP/RNA delta-delta (inhibitor-treated vs vehicle)",
  out_name   = "25_cdf_eif4e_delta_teleman_inhibitor_vs_vehicle.pdf",
  xlim       = c(-1, 1)
)

plot_registry[["delta_ours"]]    <- p_ours_delta
plot_registry[["delta_teleman"]] <- p_teleman_delta
```

## Ours (hypoxia vs normoxia) vs Teleman (inhibitor vs vehicle)

```{r composite_delta, fig.width=14, fig.height=6.5}
main_grid <- patchwork::wrap_plots(strip_legend(p_ours_delta), strip_legend(p_teleman_delta), ncol = 2)

legend_grob <- cowplot::get_legend(
  p_ours_delta + scale_color_manual(values = group_colors, breaks = names(group_colors)) +
    theme(legend.position = "bottom")
)

full <- (main_grid / patchwork::wrap_elements(legend_grob)) +
  patchwork::plot_layout(heights = c(1, 0.08)) +
  patchwork::plot_annotation(
    title = "eIF4E IP/RNA delta: ours (hypoxia 1hr vs normoxia 1hr) vs Teleman (inhibitor-treated vs vehicle)"
  )

ggsave(here("ripseq", "plots", "25_composite_eif4e_delta_ours_vs_teleman.pdf"), full,
       width = 14, height = 6.5, limitsize = FALSE)
full
```
