---
title: "RIP-seq: PCA of eIF4E IP vs sictrl RNA input"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Paper QC figure: **the eIF4E IP worked.** Two claims, both readable off a single PCA:

1. eIF4E IP replicates are tight.
2. The IP is compositionally distinct from total RNA.

The 2026 RIP submission contains **no input libraries** -- all 36 samples are
`FLAG-eIF3d_IP` / `eIF4E_IP` / `IgG-mock_IP`. The "RNA input" here is the
condition- and timepoint-matched **sictrl RNA-seq from the parallel 2025 riboseq
experiment**, the same denominator `18_ip_input_normalization.Rmd` uses.

## Normalization: why we start from raw counts on both sides

The two experiments have incompatible canonical normalizations:

| | RIP-seq 2026 | Timecourse RNA-seq 2025 |
|---|---|---|
| Spike-ins | **none** (`SpikeIn = NA`) | **Drosophila** (`^FB`, split at `timecourse_qc.Rmd:345`) |
| Canonical normalization | DESeq2 median-of-ratios | **spike-in size factors** (`createNormMatrix()`, `code/functions.R:32-66`) |

The RNA arm's spike-in normalization **cannot be transferred to the IP arm**, which
has no spike-ins. Pairing a spike-in-normalized RNA matrix against a TPM or
median-of-ratios IP matrix would put a pure normalization artifact on PC1. So both
arms enter as **raw counts** and are normalized once, jointly, exactly as
`18_ip_input_normalization.Rmd` does.

Pipeline: DESeq2 size factors on the combined matrix -> `vst(blind = TRUE)` ->
`prcomp(scale. = TRUE)`. `blind = TRUE` keeps the design out of the dispersion
estimate, which is what an unsupervised QC figure requires. `scale. = TRUE` matches
the existing project PCAs (`00_ripseq_qc.Rmd`, `01_eif3d_rip_deseq.Rmd`).

## Two caveats that must travel with this figure

**1. Design imbalance.** The eIF4E IP arm has **normoxia at 1hr only** (normoxia 1hr
x3, hypoxia 1/4/24hr x3). Within the IPs, PC2 orders by time, and
"normoxia vs hypoxia" is collinear with "1hr vs later". **No hypoxia-vs-normoxia
claim can be read off a PC here.** This notebook makes no such claim and the figure
legend says so.

**2. Batch.** `18_ip_input_normalization.Rmd` documents the IP-vs-RNA axis as
confounded with sequencing batch (2026 RIP vs 2025 RNA-seq, no shared spike-in) and
calls it unfixable for the enrichment contrast. For a PCA it is partly *testable*:
IgG-mock is the same 2026 batch, kit (`KAPA_RNAEvo`) and prep as the eIF4E IP but is
a non-specific pulldown. Section 6 runs that test.

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(matrixStats)
library(tximport)
library(pheatmap)
library(ggokabeito)
```

```{r params}
# Sample QC outlier, consistent with notebooks 01/09/18/23/24
# (01_eif3d_rip_deseq.Rmd:40). 0077 is an igg_mock normoxia_1hr replicate, so it
# affects only the Section 6 supplement panel.
OUTLIER_ID <- "0077"

# The ripseq PCAs (00_ripseq_qc.Rmd, 01_eif3d_rip_deseq.Rmd) colour with
# scale_color_okabe_ito(). Keep that, but pass an explicit `order` instead of taking
# the default: entries 1 (#E69F00) and 2 (#56B4E9) are reserved for eIF3d/eIF3e per
# CLAUDE.md, and the default order would hand them to these groups.
#   5 = #0072B2 blue, 8 = #999999 grey, 7 = #CC79A7 purple
OKABE_2GRP <- c(8, 5)     # RNA input (grey), eIF4E IP (blue)
OKABE_3GRP <- c(8, 5, 7)  # + IgG mock (purple)
```

---

# Load gene annotation

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# A gene is protein-coding if ANY transcript row has biotype "protein_coding"
# (same rule as 18_ip_input_normalization.Rmd -- distinct() first would arbitrarily
# pick one biotype per gene and misclassify NMD / CDS_not_defined first rows).
pc_genes <- gene_anno_all |>
  filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)

gene_symbol <- gene_anno_all |>
  arrange(gene_id, desc(biotype == "protein_coding")) |>
  distinct(gene_id, .keep_all = TRUE) |>
  dplyr::select(gene_id, symbol)

cat("Protein-coding genes (any-transcript rule):", length(pc_genes), "\n")
```

---

# Section 1: Build the matched IP + RNA count matrix

Both arms enter as raw counts. The RNA arm is subset to the four condition x
timepoint cells the eIF4E IP actually covers, giving balanced 12 v 12 arms.

```{r load_counts}
# The 4 cells the eIF4E IP covers (normoxia exists at 1hr only)
cells <- tibble(
  condition = c("normoxia", "hypoxia", "hypoxia", "hypoxia"),
  hours     = c("1",        "1",       "4",       "24")
)

# --- RIP-seq (2026): raw counts, already pc/polyA/length filtered in 00_ripseq_qc.Rmd
rip_counts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
rip_meta   <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                       show_col_types = FALSE) |>
  mutate(hours = as.character(hours))
cat("RIP-seq raw counts:", nrow(rip_counts), "genes x", ncol(rip_counts), "samples\n")

# --- Timecourse RNA-seq (2025): raw sictrl counts. col_select narrows the read to
# gene_id + sictrl RNA columns (out of 108 total sample columns in the file).
rna_counts_all <- read_csv(
  here("output", "human_allCounts_raw.csv"),
  col_select = c(gene_id, matches("sictrl.*_rna_")),
  show_col_types = FALSE
) |>
  column_to_rownames("gene_id") |>
  as.matrix()
rna_meta_all <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE) |>
  filter(library == "rna", treatment == "sictrl") |>
  mutate(hours = as.character(hours))
cat("sictrl RNA-seq raw counts:", nrow(rna_counts_all), "genes x",
    ncol(rna_counts_all), "samples\n")

# RIP-seq genes are a strict subset of the RNA-seq gene set (both GENCODE v49)
stopifnot("RIP-seq genes must be a subset of RNA-seq genes" =
  all(rownames(rip_counts) %in% rownames(rna_counts_all)))
rna_counts_all <- rna_counts_all[rownames(rip_counts), , drop = FALSE]

# Protein-coding filter (belt-and-suspenders: rip_counts is pc-filtered upstream)
rip_gene_id_clean <- sub("\\..*", "", rownames(rip_counts))
rip_counts     <- rip_counts[rip_gene_id_clean %in% pc_genes, , drop = FALSE]
rna_counts_all <- rna_counts_all[rownames(rip_counts), , drop = FALSE]
cat("After protein-coding filter:", nrow(rip_counts), "genes\n")
```

```{r build_matched_matrix}
# Helper: pull the count columns for a set of sample IDs, asserting the grep resolved
# to exactly one column per ID (sample IDs are embedded in longer column names).
subset_cols <- function(mat, ids, label) {
  cols <- grep(paste(ids, collapse = "|"), colnames(mat))
  stopifnot(
    "grep on sample IDs did not resolve to one column per ID" = length(cols) == length(ids)
  )
  cat("  ", label, ":", length(cols), "samples\n")
  mat[, cols, drop = FALSE]
}

# Recover the NM#####_#### sample ID from a count-matrix column name. Same idiom as
# 00_ripseq_qc.Rmd:246 / 01_eif3d_rip_deseq.Rmd:313; works on both the ripseq column
# names (condition_ip_hours_rep_NM2026_0045) and the RNA ones
# (normoxia_sictrl_1_a_rna_NM2025_0085).
sample_id_from_col <- function(x) sub(".*_(NM\\d+_\\d+)$", "\\1", x)

eif4e_meta <- rip_meta |> filter(ip_type == "eif4e")
rna_meta   <- rna_meta_all |> semi_join(cells, by = c("condition", "hours"))

cat("Building matched 12 v 12 matrix:\n")
ip_cts  <- subset_cols(rip_counts,     eif4e_meta$id, "eIF4E IP")
rna_cts <- subset_cols(rna_counts_all, rna_meta$id,   "RNA input")

stopifnot(
  "Expected 12 eIF4E IP samples"  = ncol(ip_cts)  == 12,
  "Expected 12 sictrl RNA samples" = ncol(rna_cts) == 12
)

cts <- round(cbind(ip_cts, rna_cts))
ids <- sample_id_from_col(colnames(cts))
n_ip <- ncol(ip_cts)

col_data <- data.frame(
  seq_type = factor(
    c(rep("eIF4E IP", n_ip), rep("RNA input", ncol(rna_cts))),
    levels = c("RNA input", "eIF4E IP")   # RNA input as reference
  ),
  condition = c(eif4e_meta$condition[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$condition[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  hours     = c(eif4e_meta$hours[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$hours[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  rep       = c(eif4e_meta$rep[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$rep[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  row.names = colnames(cts)
) |>
  mutate(timepoint = factor(paste0(condition, "_", hours, "hr"),
                            levels = paste0(cells$condition, "_", cells$hours, "hr")))

stopifnot(
  "Metadata join left NA values"  = !anyNA(col_data),
  "Genes shared between arms too few" = nrow(cts) > 10000
)

cat("\nSamples per group x timepoint:\n")
print(table(col_data$seq_type, col_data$timepoint))

# 3 replicates in every occupied cell
stopifnot("Expected 3 replicates per group x timepoint cell" =
  all(table(col_data$seq_type, col_data$timepoint) == 3))
```

```{r depth_summary}
# Depth is comparable between arms -- normalization here is about composition, not depth
lib_sizes <- tibble(
  sample   = colnames(cts),
  seq_type = col_data$seq_type,
  lib_M    = colSums(cts) / 1e6
)
lib_sizes |>
  group_by(seq_type) |>
  summarise(n = n(), min_M = round(min(lib_M), 1), median_M = round(median(lib_M), 1),
            max_M = round(max(lib_M), 1), .groups = "drop") |>
  print()
```

---

# Section 2: Joint normalization

DESeq2 size factors on the combined matrix, then variance-stabilizing transform.

```{r normalize}
dds <- DESeqDataSetFromMatrix(cts, col_data, design = ~ seq_type)
dds <- estimateSizeFactors(dds)

sf <- sizeFactors(dds)
stopifnot("Size factors must be positive and finite" = all(sf > 0 & is.finite(sf)))

cat("Size factors by group:\n")
tibble(seq_type = col_data$seq_type, sf = sf) |>
  group_by(seq_type) |>
  summarise(min = round(min(sf), 2), median = round(median(sf), 2),
            max = round(max(sf), 2), .groups = "drop") |>
  print()

sf_ratio <- median(sf[col_data$seq_type == "RNA input"]) /
            median(sf[col_data$seq_type == "eIF4E IP"])
cat("\nMedian size factor ratio, RNA / IP:", round(sf_ratio, 2), "\n")
cat("NOTE: median-of-ratios assumes most genes are unchanged between groups. That is\n",
    "     only approximately true across an IP/input boundary -- see Section 5 for the\n",
    "     sensitivity check that this choice does not drive the PC structure.\n")

vsd <- vst(dds, blind = TRUE)
v   <- assay(vsd)
stopifnot("VST output must be finite" = all(is.finite(v)))
cat("\nVST matrix:", nrow(v), "genes x", ncol(v), "samples\n")
```

---

# Section 3: Main figure -- eIF4E IP vs RNA input

```{r pca_helpers}
# Run prcomp on a VST matrix and return scores + variance explained.
run_pca <- function(mat, meta) {
  mat <- mat[rowSds(mat) > 0, , drop = FALSE]
  p   <- prcomp(t(mat), scale. = TRUE)
  pv  <- summary(p)$importance[2, ] * 100
  stopifnot(
    "PC1 must explain at least as much variance as PC2" = pv[1] >= pv[2],
    "Variance explained must sum to <= 100" = sum(pv) <= 100 + 1e-6
  )
  list(
    scores = as.data.frame(p$x[, 1:3]) |>
      rownames_to_column("sample") |>
      bind_cols(meta |> dplyr::select(seq_type, condition, hours, rep, timepoint)),
    var_pct = pv,
    n_genes = nrow(mat)
  )
}

# Axis label carrying variance explained, matching 00_ripseq_qc.Rmd
pc_lab <- function(pca, i) paste0("PC", i, " (", round(pca$var_pct[i], 1), "%)")

# Replicate scatter must be measured WITHIN each group x timepoint cell. Pooling a
# whole arm together instead would fold the between-timepoint spread into the
# "replicate" number -- and since PC2 here IS the timepoint axis, that inflates it
# several-fold and measures the wrong thing entirely.
group_spread <- function(pca) {
  pca$scores |>
    group_by(seq_type, timepoint) |>
    # residuals about each cell mean = pure replicate scatter
    mutate(r1 = PC1 - mean(PC1), r2 = PC2 - mean(PC2)) |>
    group_by(seq_type) |>
    summarise(
      n            = n(),
      # pooled within-cell sd, df = n - (number of cells)
      rep_sd_PC1   = round(sqrt(sum(r1^2) / (n() - n_distinct(timepoint))), 1),
      rep_sd_PC2   = round(sqrt(sum(r2^2) / (n() - n_distinct(timepoint))), 1),
      # total spread across the whole arm, for contrast
      total_sd_PC1 = round(sd(PC1), 1),
      total_sd_PC2 = round(sd(PC2), 1),
      .groups = "drop"
    )
}
```

```{r pca_main, fig.width=7, fig.height=5}
pca_main <- run_pca(v, col_data)
cat("Genes used:", pca_main$n_genes, "\n")
cat("Variance explained: PC1", round(pca_main$var_pct[1], 1), "%  PC2",
    round(pca_main$var_pct[2], 1), "%  PC3", round(pca_main$var_pct[3], 1), "%\n\n")

spread_main <- group_spread(pca_main)
print(spread_main)

p_main <- ggplot(pca_main$scores, aes(x = PC1, y = PC2,
                                      color = seq_type, shape = timepoint)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_okabe_ito(order = OKABE_2GRP) +
  labs(
    x = pc_lab(pca_main, 1), y = pc_lab(pca_main, 2),
    color = "Library", shape = "Condition / time",
    title = "PCA - eIF4E IP vs sictrl RNA input",
    # Design-imbalance caveat travels with the figure: without it, PC2 invites a
    # hypoxia reading the design cannot support (see Overview).
    subtitle = paste0("eIF4E IP has normoxia at 1hr only, so structure within the ",
                      "IPs tracks time\nand cannot be read as a hypoxia effect.")
  ) +
  theme_classic(base_size = 12)

p_main
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_input.pdf"),
       p_main, width = 7, height = 5)
```

**Replicate tightness** is the QC claim: within-group standard deviation on PC1 and
PC2 for each arm is printed above and should be small relative to the between-group
separation on PC1.

```{r qc_claim_check}
between_gap <- abs(diff(tapply(pca_main$scores$PC1, pca_main$scores$seq_type, mean)))
cat("Between-group PC1 separation:", round(between_gap, 1), "\n")
cat("Max within-cell replicate sd on PC1:", max(spread_main$rep_sd_PC1), "\n")
stopifnot("Replicate scatter should be well below the group separation" =
  max(spread_main$rep_sd_PC1) < between_gap / 4)

# Decompose the within-arm variance on PC2: how much is timepoint, how much is
# replicate noise? This is what licenses calling PC2 "the timepoint axis".
cat("\nPC2 variance decomposition within each arm (one-way ANOVA on timepoint):\n")
for (g in levels(pca_main$scores$seq_type)) {
  d  <- pca_main$scores |> filter(seq_type == g)
  fit <- summary(aov(PC2 ~ timepoint, data = d))[[1]]
  cat(sprintf("  %-10s timepoint explains %4.1f%% of PC2 variance within the arm (p = %.2g)\n",
              g, 100 * fit[["Sum Sq"]][1] / sum(fit[["Sum Sq"]]), fit[["Pr(>F)"]][1]))
}
```

## PCA coloured by replicate

The ripseq QC PCAs always include a replicate-coloured panel
(`00_ripseq_qc.Rmd`, `01_eif3d_rip_deseq.Rmd`). Here it doubles as the visual form of
the QC claim: replicate letter should show **no** structure on either axis.

```{r pca_replicate, fig.width=7, fig.height=5}
p_rep <- ggplot(pca_main$scores, aes(x = PC1, y = PC2, color = rep, shape = seq_type)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = pc_lab(pca_main, 1), y = pc_lab(pca_main, 2),
       color = "Replicate", shape = "Library",
       title = "PCA - eIF4E IP vs RNA input: replicate") +
  theme_classic(base_size = 12)

p_rep
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_replicate.pdf"),
       p_rep, width = 7, height = 5)

# Replicate must not drive either axis -- the complement of the timepoint ANOVA above.
for (pc in c("PC1", "PC2")) {
  fit <- summary(aov(as.formula(paste(pc, "~ rep")), data = pca_main$scores))[[1]]
  cat(sprintf("%s ~ replicate: %.1f%% of variance, p = %.2f\n",
              pc, 100 * fit[["Sum Sq"]][1] / sum(fit[["Sum Sq"]]), fit[["Pr(>F)"]][1]))
}
```

## Sample-sample correlation heatmap

Uses the project's established QC-heatmap recipe rather than the VST matrix the PCA
runs on: **log2(TPM + 1)**, `rowSds > 1` computed across all samples, then Pearson.
That is what `timecourse_qc.Rmd:437-448` does for the riboseq/RNA heatmaps and what
the ribopipe Shiny app does (`ribopipe_shiny_final/server.R:544-560`), so these
heatmaps are directly comparable to those.

TPM is not saved by any upstream notebook (only raw counts are), so both arms are
re-imported from their salmon quants here. TPM is also a *within-sample*
normalization, so unlike the size-factor route it makes no "most genes unchanged"
assumption across the IP/input boundary.

```{r heatmap_tpm_import}
tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript", "gene"), show_col_types = FALSE)

# The two experiments were quantified against DIFFERENT salmon indices, so they cannot
# go through one tximport() call (it asserts identical transcript vectors). The 2026
# RIP index has the spike-ins removed (517,038 transcripts, 0 Drosophila); the 2025
# index still carries 30,277 Drosophila transcripts. The human transcript set is
# identical between them, so importing separately and joining on gene_id is exact.
import_tpm <- function(paths, nms, label) {
  names(paths) <- nms
  stopifnot("Some salmon quant files not found" = all(file.exists(paths)))
  txi <- tximport(paths, type = "salmon", tx2gene = tx2gene)
  cat("  ", label, ":", nrow(txi$abundance), "genes x", ncol(txi$abundance), "samples\n")
  txi$abundance
}

cat("tximport per arm (different indices):\n")
tpm_ip <- import_tpm(
  {q <- file.path(here(), "ripseq", "salmon", eif4e_meta$id, "quant.sf.gz")
   ifelse(file.exists(q), q, sub("\\.gz$", "", q))},
  paste(eif4e_meta$condition, "eif4e", eif4e_meta$hours,
        eif4e_meta$rep, eif4e_meta$id, sep = "_"),
  "eIF4E IP ")
tpm_rna <- import_tpm(
  file.path(here(), "2025_MDAcells_riboseq", "salmon", rna_meta$id, "quant.sf.gz"),
  rna_meta$sample_name,
  "RNA input")

# Drosophila share differs by arm (0% in the RIP index, ~5.5% in the 2025 one), which
# would deflate every human TPM in the RNA arm by a constant. Drop spike-ins and
# renormalize each sample to 1e6 over human genes so the two arms are on one scale.
# The single-experiment precedents (timecourse_qc.Rmd, ribopipe) never need this --
# their spike-in share is common to all samples and cancels.
drop_spikes_renorm <- function(m) {
  m <- m[!grepl("^FB", rownames(m)), , drop = FALSE]
  t(t(m) / colSums(m)) * 1e6
}
cat("Drosophila TPM share before renormalization -- IP:",
    sprintf("%.2f%%", 100 * mean(colSums(tpm_ip[grepl("^FB", rownames(tpm_ip)), , drop = FALSE]) / colSums(tpm_ip))),
    " RNA:",
    sprintf("%.2f%%\n", 100 * mean(colSums(tpm_rna[grepl("^FB", rownames(tpm_rna)), , drop = FALSE]) / colSums(tpm_rna))))

tpm_ip  <- drop_spikes_renorm(tpm_ip)
tpm_rna <- drop_spikes_renorm(tpm_rna)

shared <- intersect(rownames(tpm_ip), rownames(tpm_rna))
tpm    <- cbind(tpm_ip[shared, , drop = FALSE], tpm_rna[shared, , drop = FALSE])
cat("Joined TPM matrix:", nrow(tpm), "genes x", ncol(tpm), "samples\n")

stopifnot(
  "TPM columns must sum to ~1e6 after renormalization" =
    all(abs(colSums(tpm) - 1e6) < 1e-6 * 1e6),
  "All notebook genes must be present in the TPM matrix" =
    all(rownames(cts) %in% rownames(tpm))
)
```

```

```{r cor_heatmap, fig.width=9, fig.height=8}
# Restrict to the gene set used everywhere else in this notebook, then apply the
# project recipe: log2(TPM + 1) -> rowSds > 1 across all samples -> Pearson.
qcinput <- log2(tpm[rownames(cts), , drop = FALSE] + 1)
qcinput <- qcinput[rowSds(qcinput) > 1, ]
cat("Genes after SD > 1 filter:", nrow(qcinput), "\n")

# Column order follows the quant vectors above, which were built from eif4e_meta then
# rna_meta -- the same order as `cts`, so col_data lines up. Assert rather than trust.
stopifnot("TPM column order must match the counts matrix" =
  identical(sample_id_from_col(colnames(qcinput)), ids))

mycor <- cor(qcinput, method = "pearson")
sample_anno <- data.frame(
  library   = col_data$seq_type,
  timepoint = col_data$timepoint,
  row.names = colnames(qcinput)
)

p_heatmap <- pheatmap::pheatmap(
  mat            = mycor,
  annotation_col = sample_anno,
  border_color   = "black",
  scale          = "none",
  cluster_rows   = TRUE,
  cluster_cols   = TRUE,
  fontsize       = 7
)
ggsave(here("ripseq", "plots", "26_correlation_heatmap.pdf"),
       p_heatmap, width = 9, height = 8)

# Within-arm correlations must exceed between-arm ones -- the heatmap's QC claim
same_arm <- outer(col_data$seq_type, col_data$seq_type, "==")
diag(same_arm) <- NA
cat("Median Pearson r within an arm:",
    round(median(mycor[which(same_arm)], na.rm = TRUE), 3), "\n")
cat("Median Pearson r across arms:  ",
    round(median(mycor[which(!same_arm)], na.rm = TRUE), 3), "\n")
stopifnot("Samples should correlate better within their own arm" =
  median(mycor[which(same_arm)], na.rm = TRUE) >
  median(mycor[which(!same_arm)], na.rm = TRUE))
```

---

# Section 4: Composition diagnostics

The IP and the input differ compositionally by construction. The concern a reviewer
will raise is that this difference -- especially the mitochondrial fraction -- is
driving PC1 as an artifact. Quantify it, then show it is not.

```{r composition}
sym <- gene_symbol$symbol[match(rownames(cts), gene_symbol$gene_id)]

cpm  <- function(x) t(t(x) / colSums(x)) * 1e6
m_ip  <- rowMeans(cpm(ip_cts[rownames(cts), , drop = FALSE]))
m_rna <- rowMeans(cpm(rna_cts[rownames(cts), , drop = FALSE]))

share <- function(x, k) sum(x[k]) / 1e6 * 100
top_share <- function(x, n) sum(sort(x, decreasing = TRUE)[1:n]) / 1e6 * 100

is_mt <- grepl("^MT-", sym)
comp <- tibble(
  metric = c("top-100 gene share (%)", "top-500 gene share (%)", "MT- share (%)"),
  eif4e_IP  = round(c(top_share(m_ip, 100),  top_share(m_ip, 500),  share(m_ip, is_mt)), 1),
  RNA_input = round(c(top_share(m_rna, 100), top_share(m_rna, 500), share(m_rna, is_mt)), 1)
)
print(comp)

cat("\nSpearman, mean logCPM IP vs RNA:",
    round(cor(log2(m_ip + 1), log2(m_rna + 1), method = "spearman"), 3), "\n")
lfc <- log2(m_ip + 1) - log2(m_rna + 1)
cat("Median logFC (IP - RNA):", round(median(lfc), 2),
    " IQR:", round(quantile(lfc, 0.25), 2), "to", round(quantile(lfc, 0.75), 2), "\n")
cat("Genes shifted > 2-fold:", round(mean(abs(lfc) > 1) * 100, 1), "%\n")

# Biological spot-check: the compositional signature this figure depends on.
# MT transcripts are POLRMT products and lack an m7G cap, so a cap-binding IP should
# deplete them strongly relative to rRNA-depleted total RNA.
mt_ip  <- share(m_ip,  is_mt)
mt_rna <- share(m_rna, is_mt)
stopifnot(
  "MT share in RNA input outside expected range" = mt_rna >= 10 & mt_rna <= 25,
  "MT share in eIF4E IP outside expected range"  = mt_ip  >= 0  & mt_ip  <= 5,
  "MT should be depleted in the cap-binding IP relative to input" = mt_ip < mt_rna
)
cat("\nMT depletion in IP vs input:", round(mt_rna / mt_ip, 1), "fold\n")

# Bait sanity check: EIF4E itself must be detected in the IP arm
eif4e_row <- which(sym == "EIF4E")
stopifnot("EIF4E not found in the count matrix" = length(eif4e_row) == 1)
stopifnot("EIF4E has no counts in the IP arm" =
  all(rowSums(ip_cts[rownames(cts), , drop = FALSE][eif4e_row, , drop = FALSE]) > 0))
cat("EIF4E mean CPM -- IP:", round(m_ip[eif4e_row], 1),
    " RNA:", round(m_rna[eif4e_row], 1), "\n")
```

```{r pca_no_mt, fig.width=7, fig.height=5}
# Re-run with MT genes removed. Under VST + scaling every gene contributes equally,
# so 13 MT genes out of ~10,800 cannot dominate -- PC1 should be essentially unchanged.
v_no_mt  <- v[!grepl("^MT-", gene_symbol$symbol[match(rownames(v), gene_symbol$gene_id)]), ]
pca_nomt <- run_pca(v_no_mt, col_data)

cat("MT genes removed:", nrow(v) - nrow(v_no_mt), "\n")
cat("PC1 all genes:", round(pca_main$var_pct[1], 1), "%   PC1 MT removed:",
    round(pca_nomt$var_pct[1], 1), "%\n")
stopifnot("Removing MT genes should not materially change PC1" =
  abs(pca_main$var_pct[1] - pca_nomt$var_pct[1]) < 0.5)

p_nomt <- ggplot(pca_nomt$scores, aes(x = PC1, y = PC2,
                                      color = seq_type, shape = timepoint)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_okabe_ito(order = OKABE_2GRP) +
  labs(x = pc_lab(pca_nomt, 1), y = pc_lab(pca_nomt, 2),
       color = "Library", shape = "Condition / time",
       title = "PCA - eIF4E IP vs RNA input: mitochondrial genes removed") +
  theme_classic(base_size = 12)

p_nomt
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_mt_removed.pdf"),
       p_nomt, width = 7, height = 5)
```

---

# Section 5: Sensitivity -- within-group size factors

Median-of-ratios centres the two arms on a shared median, which is an assumption
across an IP/input boundary. Re-estimate size factors **within** each arm instead
(depth correction only, no cross-arm centring) and confirm the PC structure holds.

```{r sensitivity_sf}
sf_within <- rep(NA_real_, ncol(cts))
for (g in levels(col_data$seq_type)) {
  k <- col_data$seq_type == g
  sf_within[k] <- estimateSizeFactorsForMatrix(cts[, k, drop = FALSE])
}
stopifnot("Within-group size factors must be positive and finite" =
  all(sf_within > 0 & is.finite(sf_within)))

dds_w <- DESeqDataSetFromMatrix(cts, col_data, design = ~ seq_type)
sizeFactors(dds_w) <- sf_within
v_w <- assay(vst(dds_w, blind = TRUE))
stopifnot("VST output must be finite" = all(is.finite(v_w)))

pca_w <- run_pca(v_w, col_data)
cat("Combined size factors  -> PC1", round(pca_main$var_pct[1], 1),
    "%  PC2", round(pca_main$var_pct[2], 1), "%\n")
cat("Within-group size factors -> PC1", round(pca_w$var_pct[1], 1),
    "%  PC2", round(pca_w$var_pct[2], 1), "%\n\n")

cat("Group separation on PC1 preserved:\n")
print(round(tapply(pca_w$scores$PC1, pca_w$scores$seq_type, mean), 1))

# The two normalizations must agree on which samples sit where, up to a sign flip
# (PC sign is arbitrary in prcomp).
cor_pc1 <- cor(pca_main$scores$PC1, pca_w$scores$PC1)
cat("\nCorrelation of PC1 scores between the two normalizations:",
    round(cor_pc1, 3), "\n")
stopifnot("PC1 sample ordering must be robust to the size factor choice" =
  abs(cor_pc1) > 0.95)
```

---

# Section 6 (supplement): IgG-mock as a batch anchor

Is the IP-vs-RNA axis real pulldown specificity, or just 2026-vs-2025 batch?
IgG-mock is the **same batch, kit and prep** as the eIF4E IP but is a non-specific
pulldown. If PC1 were library batch, the two IPs would sit together, away from RNA.

```{r pca_3group, fig.width=7, fig.height=5}
igg_meta <- rip_meta |> filter(ip_type == "igg_mock", !grepl(OUTLIER_ID, id))
cat("IgG-mock samples after dropping", OUTLIER_ID, ":", nrow(igg_meta), "\n")

igg_cts <- subset_cols(rip_counts, igg_meta$id, "IgG mock")
cts3    <- round(cbind(ip_cts, igg_cts, rna_cts))
ids3    <- sample_id_from_col(colnames(cts3))
stopifnot("Outlier was not excluded" = !any(grepl(OUTLIER_ID, ids3)))

meta3 <- bind_rows(
  eif4e_meta |> mutate(seq_type = "eIF4E IP"),
  igg_meta   |> mutate(seq_type = "IgG mock"),
  rna_meta   |> mutate(seq_type = "RNA input")
) |>
  dplyr::select(id, seq_type, condition, hours, rep)

col_data3 <- meta3[match(ids3, meta3$id), ] |>
  mutate(
    seq_type  = factor(seq_type, levels = c("RNA input", "eIF4E IP", "IgG mock")),
    timepoint = factor(paste0(condition, "_", hours, "hr"),
                       levels = paste0(cells$condition, "_", cells$hours, "hr"))
  ) |>
  as.data.frame()
rownames(col_data3) <- colnames(cts3)
stopifnot("Metadata join left NA values" = !anyNA(col_data3))

dds3 <- estimateSizeFactors(DESeqDataSetFromMatrix(cts3, col_data3, design = ~ seq_type))
v3   <- assay(vst(dds3, blind = TRUE))
stopifnot("VST output must be finite" = all(is.finite(v3)))

pca3 <- run_pca(v3, col_data3)
cat("\nVariance explained: PC1", round(pca3$var_pct[1], 1), "%  PC2",
    round(pca3$var_pct[2], 1), "%\n\n")

cat("PC1 mean by group (the batch-anchor test):\n")
pc1_means <- tapply(pca3$scores$PC1, pca3$scores$seq_type, mean)
print(round(pc1_means, 1))

cat("\nReplicate scatter by group:\n")
print(group_spread(pca3))

# The specificity argument: eIF4E must sit closer to RNA input than to IgG-mock,
# despite sharing batch, kit and prep with IgG-mock.
d_to_rna <- abs(pc1_means[["eIF4E IP"]] - pc1_means[["RNA input"]])
d_to_igg <- abs(pc1_means[["eIF4E IP"]] - pc1_means[["IgG mock"]])
cat("\neIF4E PC1 distance to RNA input:", round(d_to_rna, 1),
    " to IgG mock:", round(d_to_igg, 1), "\n")
stopifnot("eIF4E IP should sit closer to input than to its same-batch IgG control" =
  d_to_rna < d_to_igg)
cat("=> PC1 is dominated by pulldown specificity, not sequencing batch.\n")

p_3grp <- ggplot(pca3$scores, aes(x = PC1, y = PC2,
                                  color = seq_type, shape = timepoint)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_okabe_ito(order = OKABE_3GRP) +
  labs(x = pc_lab(pca3, 1), y = pc_lab(pca3, 2),
       color = "Library", shape = "Condition / time",
       title = "PCA - three groups: IgG-mock as a batch anchor") +
  theme_classic(base_size = 12)

p_3grp
ggsave(here("ripseq", "plots", "26_pca_3group_with_igg.pdf"),
       p_3grp, width = 7, height = 5)
```

The large IgG-mock replicate scatter above is a property of the non-specific
pulldown, not of any one sample -- it is essentially unchanged whether or not
outlier `r OUTLIER_ID` is included. That is why IgG-mock is kept out of the main
figure: on a shared axis it compresses the eIF4E and RNA arms into a corner.

---

# Section 7: Export sample coordinates

```{r export}
pca_main$scores |>
  mutate(analysis = "main_eif4e_vs_rna") |>
  bind_rows(pca3$scores |> mutate(analysis = "supplement_3group")) |>
  dplyr::select(analysis, sample, seq_type, condition, hours, rep, timepoint,
                PC1, PC2, PC3) |>
  write_csv(here("ripseq", "output", "26_pca_sample_coordinates.csv"))

cat("Wrote ripseq/output/26_pca_sample_coordinates.csv\n")
```

```{r session_info}
sessionInfo()
```
