---
title: "RIP-seq: PCA of eIF4E IP vs sictrl RNA input"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Paper QC figure: **the eIF4E IP worked.** Two claims, both readable off a single PCA:

1. eIF4E IP replicates are tight.
2. The IP is compositionally distinct from total RNA.

The 2026 RIP submission contains **no input libraries** -- all 36 samples are
`FLAG-eIF3d_IP` / `eIF4E_IP` / `IgG-mock_IP`. The "RNA input" here is the
condition- and timepoint-matched **sictrl RNA-seq from the parallel 2025 riboseq
experiment**, the same denominator `18_ip_input_normalization.Rmd` uses.

## Normalization: the project QC recipe

Heatmap and PCA both run on one `qcinput`, built the same way as every other QC
heatmap/PCA in the project:

```r
qcinput <- log2(myTxi$abundance[keepGenes, ] + 1)   # salmon TPM
qcinput <- qcinput[rowSds(qcinput) > 1, ]
```

That is `timecourse_qc.Rmd:436-439` (riboseq/RNA QC) and
`ribopipe_shiny_final/server.R:544-546`, and `00_ripseq_qc.Rmd:238` does the same on
this experiment. Correlations are Pearson on `qcinput`; the PCA is
`prcomp(t(qcinput), scale. = TRUE)`. Keeping the two panels on one matrix is also
what those notebooks do.

TPM rather than counts matters here: the arms differ in effective length (median 571
vs 540 nt, from different fragment-size distributions), and that correction enters
TPM but not CPM. Section 6 re-runs the PCA through DESeq2 size factors + VST to
confirm nothing depends on the choice.

## Two caveats that must travel with this figure

**1. Design imbalance.** The eIF4E IP arm has **normoxia at 1hr only** (normoxia 1hr
x3, hypoxia 1/4/24hr x3). Within the IPs, PC2 orders by time, and
"normoxia vs hypoxia" is collinear with "1hr vs later". **No hypoxia-vs-normoxia
claim can be read off a PC here.** This notebook makes no such claim and the figure
legend says so.

**2. Batch.** `18_ip_input_normalization.Rmd` documents the IP-vs-RNA axis as
confounded with sequencing batch (2026 RIP vs 2025 RNA-seq, no shared spike-in) and
calls it unfixable for the enrichment contrast. For a PCA it is partly *testable*:
IgG-mock is the same 2026 batch, kit (`KAPA_RNAEvo`) and prep as the eIF4E IP but is
a non-specific pulldown. Section 6 runs that test.

```{r libraries}
library(here)
library(tidyverse)
library(DESeq2)
library(matrixStats)
library(tximport)
library(pheatmap)
library(ggokabeito)
```

```{r params}
# Sample QC outlier, consistent with notebooks 01/09/18/23/24
# (01_eif3d_rip_deseq.Rmd:40). 0077 is an igg_mock normoxia_1hr replicate, so it
# affects only the Section 6 supplement panel.
OUTLIER_ID <- "0077"

# The ripseq PCAs (00_ripseq_qc.Rmd, 01_eif3d_rip_deseq.Rmd) colour with
# scale_color_okabe_ito(). Keep that, but pass an explicit `order` instead of taking
# the default: entries 1 (#E69F00) and 2 (#56B4E9) are reserved for eIF3d/eIF3e per
# CLAUDE.md, and the default order would hand them to these groups.
#   5 = #0072B2 blue, 8 = #999999 grey, 7 = #CC79A7 purple
OKABE_2GRP <- c(8, 5)     # RNA input (grey), eIF4E IP (blue)
OKABE_3GRP <- c(8, 5, 7)  # + IgG mock (purple)
```

---

# Load gene annotation

```{r gene_anno}
gene_anno_all <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv.gz"),
  col_names = c("gene_id", "symbol", "biotype", "species", "version"),
  show_col_types = FALSE
) |>
  filter(species == "human") |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# A gene is protein-coding if ANY transcript row has biotype "protein_coding"
# (same rule as 18_ip_input_normalization.Rmd -- distinct() first would arbitrarily
# pick one biotype per gene and misclassify NMD / CDS_not_defined first rows).
pc_genes <- gene_anno_all |>
  filter(biotype == "protein_coding") |>
  distinct(gene_id_clean) |>
  pull(gene_id_clean)

gene_symbol <- gene_anno_all |>
  arrange(gene_id, desc(biotype == "protein_coding")) |>
  distinct(gene_id, .keep_all = TRUE) |>
  dplyr::select(gene_id, symbol)

cat("Protein-coding genes (any-transcript rule):", length(pc_genes), "\n")
```

---

# Section 1: Build the matched IP + RNA count matrix

Both arms enter as raw counts. The RNA arm is subset to the four condition x
timepoint cells the eIF4E IP actually covers, giving balanced 12 v 12 arms.

```{r load_counts}
# The 4 cells the eIF4E IP covers (normoxia exists at 1hr only)
cells <- tibble(
  condition = c("normoxia", "hypoxia", "hypoxia", "hypoxia"),
  hours     = c("1",        "1",       "4",       "24")
)

# --- RIP-seq (2026): raw counts, already pc/polyA/length filtered in 00_ripseq_qc.Rmd
rip_counts <- readRDS(here("ripseq", "output", "ripseq_allCounts_raw.rds"))
rip_meta   <- read_csv(here("ripseq", "output", "ripseq_metadata_clean.csv"),
                       show_col_types = FALSE) |>
  mutate(hours = as.character(hours))
cat("RIP-seq raw counts:", nrow(rip_counts), "genes x", ncol(rip_counts), "samples\n")

# --- Timecourse RNA-seq (2025): raw sictrl counts. col_select narrows the read to
# gene_id + sictrl RNA columns (out of 108 total sample columns in the file).
rna_counts_all <- read_csv(
  here("output", "human_allCounts_raw.csv"),
  col_select = c(gene_id, matches("sictrl.*_rna_")),
  show_col_types = FALSE
) |>
  column_to_rownames("gene_id") |>
  as.matrix()
rna_meta_all <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE) |>
  filter(library == "rna", treatment == "sictrl") |>
  mutate(hours = as.character(hours))
cat("sictrl RNA-seq raw counts:", nrow(rna_counts_all), "genes x",
    ncol(rna_counts_all), "samples\n")

# RIP-seq genes are a strict subset of the RNA-seq gene set (both GENCODE v49)
stopifnot("RIP-seq genes must be a subset of RNA-seq genes" =
  all(rownames(rip_counts) %in% rownames(rna_counts_all)))
rna_counts_all <- rna_counts_all[rownames(rip_counts), , drop = FALSE]

# Protein-coding filter (belt-and-suspenders: rip_counts is pc-filtered upstream)
rip_gene_id_clean <- sub("\\..*", "", rownames(rip_counts))
rip_counts     <- rip_counts[rip_gene_id_clean %in% pc_genes, , drop = FALSE]
rna_counts_all <- rna_counts_all[rownames(rip_counts), , drop = FALSE]
cat("After protein-coding filter:", nrow(rip_counts), "genes\n")
```

```{r build_matched_matrix}
# Helper: pull the count columns for a set of sample IDs, asserting the grep resolved
# to exactly one column per ID (sample IDs are embedded in longer column names).
subset_cols <- function(mat, ids, label) {
  cols <- grep(paste(ids, collapse = "|"), colnames(mat))
  stopifnot(
    "grep on sample IDs did not resolve to one column per ID" = length(cols) == length(ids)
  )
  cat("  ", label, ":", length(cols), "samples\n")
  mat[, cols, drop = FALSE]
}

# Recover the NM#####_#### sample ID from a count-matrix column name. Same idiom as
# 00_ripseq_qc.Rmd:246 / 01_eif3d_rip_deseq.Rmd:313; works on both the ripseq column
# names (condition_ip_hours_rep_NM2026_0045) and the RNA ones
# (normoxia_sictrl_1_a_rna_NM2025_0085).
sample_id_from_col <- function(x) sub(".*_(NM\\d+_\\d+)$", "\\1", x)

eif4e_meta <- rip_meta |> filter(ip_type == "eif4e")
rna_meta   <- rna_meta_all |> semi_join(cells, by = c("condition", "hours"))

cat("Building matched 12 v 12 matrix:\n")
ip_cts  <- subset_cols(rip_counts,     eif4e_meta$id, "eIF4E IP")
rna_cts <- subset_cols(rna_counts_all, rna_meta$id,   "RNA input")

stopifnot(
  "Expected 12 eIF4E IP samples"  = ncol(ip_cts)  == 12,
  "Expected 12 sictrl RNA samples" = ncol(rna_cts) == 12
)

cts <- round(cbind(ip_cts, rna_cts))
ids <- sample_id_from_col(colnames(cts))
n_ip <- ncol(ip_cts)

col_data <- data.frame(
  seq_type = factor(
    c(rep("eIF4E IP", n_ip), rep("RNA input", ncol(rna_cts))),
    levels = c("RNA input", "eIF4E IP")   # RNA input as reference
  ),
  condition = c(eif4e_meta$condition[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$condition[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  hours     = c(eif4e_meta$hours[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$hours[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  rep       = c(eif4e_meta$rep[match(ids[seq_len(n_ip)], eif4e_meta$id)],
                rna_meta$rep[match(ids[-seq_len(n_ip)], rna_meta$id)]),
  row.names = colnames(cts)
) |>
  mutate(timepoint = factor(paste0(condition, "_", hours, "hr"),
                            levels = paste0(cells$condition, "_", cells$hours, "hr")))

stopifnot(
  "Metadata join left NA values"  = !anyNA(col_data),
  "Genes shared between arms too few" = nrow(cts) > 10000
)

cat("\nSamples per group x timepoint:\n")
print(table(col_data$seq_type, col_data$timepoint))

# 3 replicates in every occupied cell
stopifnot("Expected 3 replicates per group x timepoint cell" =
  all(table(col_data$seq_type, col_data$timepoint) == 3))
```

```{r depth_summary}
# Depth is comparable between arms -- normalization here is about composition, not depth
lib_sizes <- tibble(
  sample   = colnames(cts),
  seq_type = col_data$seq_type,
  lib_M    = colSums(cts) / 1e6
)
lib_sizes |>
  group_by(seq_type) |>
  summarise(n = n(), min_M = round(min(lib_M), 1), median_M = round(median(lib_M), 1),
            max_M = round(max(lib_M), 1), .groups = "drop") |>
  print()
```

---

# Section 2: QC input -- log2(TPM + 1), SD filter

This follows `timecourse_qc.Rmd:436-439` and `ribopipe_shiny_final/server.R:544-546`
exactly: `myTxi$abundance` (salmon TPM) indexed at the notebook's gene set, log2 with a
pseudocount, then `rowSds > 1` across all samples. The **same `qcinput` feeds both the
correlation heatmap and the PCA**, as it does there.

TPM is not saved by any upstream notebook (only counts are), so it is re-imported from
salmon here. `timecourse_qc.Rmd` and the Shiny app both had `myTxi` in memory already
and got TPM for free.

```{r tpm_import}
tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript", "gene"), show_col_types = FALSE)

# The two experiments used DIFFERENT salmon indices, so they cannot go through one
# tximport() call (it asserts identical transcript vectors). The 2026 RIP index has
# spike-ins removed (517,038 transcripts, 0 Drosophila); the 2025 index still carries
# 30,277 Drosophila transcripts. The human transcript set is identical between them,
# so importing separately and joining on gene_id is exact.
import_tpm <- function(paths, nms, label) {
  names(paths) <- nms
  stopifnot("Some salmon quant files not found" = all(file.exists(paths)))
  txi <- tximport(paths, type = "salmon", tx2gene = tx2gene)
  cat("  ", label, ":", nrow(txi$abundance), "genes x", ncol(txi$abundance), "samples\n")
  txi$abundance
}

# Import both IP types at once -- igg_mock is needed for the Section 7 supplement.
rip_ip_meta <- rip_meta |> filter(ip_type %in% c("eif4e", "igg_mock"))
rip_q <- file.path(here(), "ripseq", "salmon", rip_ip_meta$id, "quant.sf.gz")
rip_q <- ifelse(file.exists(rip_q), rip_q, sub("\\.gz$", "", rip_q))

cat("tximport per arm (different indices):\n")
tpm_rip <- import_tpm(rip_q, paste(rip_ip_meta$condition, rip_ip_meta$ip_type,
                                   rip_ip_meta$hours, rip_ip_meta$rep,
                                   rip_ip_meta$id, sep = "_"), "RIP-seq  ")
tpm_rna <- import_tpm(file.path(here(), "2025_MDAcells_riboseq", "salmon",
                                rna_meta$id, "quant.sf.gz"),
                      rna_meta$sample_name, "RNA input")

# Join on gene_id. The intersection drops Drosophila on its own -- those rows exist
# only in the 2025 arm, so no spike-in gene is shared.
#
# Deliberately NOT renormalizing the arms to a common 1e6 despite their differing
# spike-in share (0% vs ~7%): that multiplies each RNA column by a constant, which is
# a constant shift in log2 and therefore invisible to Pearson. Tested -- max change in
# any pairwise r was 0.005 and the clustering order was identical. Skipping it keeps
# this identical to the project recipe, which also uses abundance as-is.
shared  <- intersect(rownames(tpm_rip), rownames(tpm_rna))
tpm_all <- cbind(tpm_rip[shared, , drop = FALSE], tpm_rna[shared, , drop = FALSE])

stopifnot(
  "Spike-ins should not survive the gene-id intersection" =
    !any(grepl("^FB", rownames(tpm_all))),
  "All notebook genes must be present in the TPM matrix" =
    all(rownames(cts) %in% rownames(tpm_all))
)
cat("Joined TPM matrix:", nrow(tpm_all), "genes x", ncol(tpm_all), "samples\n")
```

```{r qcinput}
# The project recipe, applied to the 12 v 12 eIF4E-vs-RNA sample set.
# `rownames(cts)` is the gene set from 00_ripseq_qc.Rmd (expression cutoff,
# protein-coding, polyA-plus, >=200nt) -- the counterpart of `keepGenes` in
# timecourse_qc.Rmd, which derives its cutoff from counts and applies it to abundance.
qcinput <- log2(tpm_all[rownames(cts), colnames(cts), drop = FALSE] + 1)
qcinput <- qcinput[rowSds(qcinput) > 1, ]
cat("Genes after SD > 1 filter:", nrow(qcinput), "\n")

stopifnot(
  "qcinput columns must match the sample table" =
    identical(colnames(qcinput), rownames(col_data)),
  "qcinput must be finite" = all(is.finite(qcinput))
)
```

---

# Section 3: Sample-sample correlation heatmap

```{r cor_heatmap, fig.width=9, fig.height=8}
mycor <- cor(qcinput, method = "pearson")

sample_anno <- data.frame(
  library   = col_data$seq_type,
  timepoint = col_data$timepoint,
  row.names = colnames(qcinput)
)

p_heatmap <- pheatmap::pheatmap(
  mat            = mycor,
  annotation_col = sample_anno,
  border_color   = "black",
  scale          = "none",
  cluster_rows   = TRUE,
  cluster_cols   = TRUE,
  fontsize       = 8
)
p_heatmap
ggsave(here("ripseq", "plots", "26_correlation_heatmap.pdf"),
       p_heatmap, width = 10, height = 9)

# Within-arm correlations must exceed between-arm ones -- the heatmap's QC claim
same_arm <- outer(col_data$seq_type, col_data$seq_type, "==")
diag(same_arm) <- NA
cat("Median Pearson r within an arm:",
    round(median(mycor[which(same_arm)], na.rm = TRUE), 3), "\n")
cat("Median Pearson r across arms:  ",
    round(median(mycor[which(!same_arm)], na.rm = TRUE), 3), "\n")
stopifnot("Samples should correlate better within their own arm" =
  median(mycor[which(same_arm)], na.rm = TRUE) >
  median(mycor[which(!same_arm)], na.rm = TRUE))
```

---

# Section 4: PCA

`prcomp(t(qcinput), scale. = TRUE)` on the same matrix the heatmap used, matching
`timecourse_qc.Rmd:467` and `00_ripseq_qc.Rmd:242`. Panels follow the ripseq QC
convention: one per metadata variable (library, timepoint, replicate).

```{r pca_helpers}
run_pca <- function(mat, meta) {
  mat <- mat[rowSds(mat) > 0, , drop = FALSE]
  p   <- prcomp(t(mat), scale. = TRUE)
  # From sdev^2 rather than summary()$importance: summary.prcomp rounds the
  # proportions to 5 decimals, so summing one per sample overshoots 100 measurably
  # (100.001 here) and a sum check against it fails on rounding alone.
  pv  <- p$sdev^2 / sum(p$sdev^2) * 100
  stopifnot(
    "PC1 must explain at least as much variance as PC2" = pv[1] >= pv[2],
    "Variance explained must sum to 100" = abs(sum(pv) - 100) < 1e-6
  )
  list(
    scores = as.data.frame(p$x[, 1:3]) |>
      rownames_to_column("sample") |>
      bind_cols(meta |> dplyr::select(seq_type, condition, hours, rep, timepoint)),
    var_pct = pv,
    n_genes = nrow(mat)
  )
}

pc_lab <- function(pca, i) paste0("PC", i, " (", round(pca$var_pct[i], 1), "%)")

# Replicate scatter must be measured WITHIN each group x timepoint cell. Pooling a
# whole arm together would fold between-timepoint spread into the "replicate" number.
group_spread <- function(pca) {
  pca$scores |>
    group_by(seq_type, timepoint) |>
    mutate(r1 = PC1 - mean(PC1), r2 = PC2 - mean(PC2)) |>
    group_by(seq_type) |>
    summarise(
      n            = n(),
      rep_sd_PC1   = round(sqrt(sum(r1^2) / (n() - n_distinct(timepoint))), 1),
      rep_sd_PC2   = round(sqrt(sum(r2^2) / (n() - n_distinct(timepoint))), 1),
      total_sd_PC1 = round(sd(PC1), 1),
      total_sd_PC2 = round(sd(PC2), 1),
      .groups = "drop"
    )
}

pca_main <- run_pca(qcinput, col_data)
cat("Genes used:", pca_main$n_genes, "\n")
cat("Variance explained: PC1", round(pca_main$var_pct[1], 1), "%  PC2",
    round(pca_main$var_pct[2], 1), "%  PC3", round(pca_main$var_pct[3], 1), "%\n\n")
spread_main <- group_spread(pca_main)
print(spread_main)
```

```{r pca_library, fig.width=7, fig.height=5}
p_main <- ggplot(pca_main$scores, aes(x = PC1, y = PC2,
                                      color = seq_type, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_okabe_ito(order = OKABE_2GRP) +
  labs(
    x = pc_lab(pca_main, 1), y = pc_lab(pca_main, 2),
    color = "Library", shape = "Condition / time",
    title = "PCA - eIF4E IP vs sictrl RNA input",
    # Design-imbalance caveat travels with the figure: without it, PC2 invites a
    # hypoxia reading the design cannot support (see Overview).
    subtitle = paste0("eIF4E IP has normoxia at 1hr only, so structure within the ",
                      "IPs tracks time\nand cannot be read as a hypoxia effect.")
  ) +
  theme_classic(base_size = 12)

p_main
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_input.pdf"),
       p_main, width = 7, height = 5)
```

```{r pca_timepoint, fig.width=7, fig.height=5}
p_time <- ggplot(pca_main$scores, aes(x = PC1, y = PC2,
                                      color = timepoint, shape = seq_type)) +
  geom_point(size = 3) +
  scale_color_okabe_ito(order = c(5, 3, 6, 7)) +
  labs(x = pc_lab(pca_main, 1), y = pc_lab(pca_main, 2),
       color = "Condition / time", shape = "Library",
       title = "PCA - eIF4E IP vs RNA input: timepoint") +
  theme_classic(base_size = 12)

p_time
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_timepoint.pdf"),
       p_time, width = 7, height = 5)
```

```{r pca_replicate, fig.width=7, fig.height=5}
p_rep <- ggplot(pca_main$scores, aes(x = PC1, y = PC2,
                                     color = rep, shape = seq_type)) +
  geom_point(size = 3) +
  scale_color_okabe_ito() +
  labs(x = pc_lab(pca_main, 1), y = pc_lab(pca_main, 2),
       color = "Replicate", shape = "Library",
       title = "PCA - eIF4E IP vs RNA input: replicate") +
  theme_classic(base_size = 12)

p_rep
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_replicate.pdf"),
       p_rep, width = 7, height = 5)
```

```{r qc_claim_check}
between_gap <- abs(diff(tapply(pca_main$scores$PC1, pca_main$scores$seq_type, mean)))
cat("Between-group PC1 separation:", round(between_gap, 1), "\n")
cat("Max within-cell replicate sd on PC1:", max(spread_main$rep_sd_PC1), "\n")
stopifnot("Replicate scatter should be well below the group separation" =
  max(spread_main$rep_sd_PC1) < between_gap / 4)

# Replicate must not drive either axis; timepoint is expected to drive PC2.
for (pc in c("PC1", "PC2")) {
  fit <- summary(aov(as.formula(paste(pc, "~ rep")), data = pca_main$scores))[[1]]
  cat(sprintf("%s ~ replicate: %.1f%% of variance, p = %.2f\n",
              pc, 100 * fit[["Sum Sq"]][1] / sum(fit[["Sum Sq"]]), fit[["Pr(>F)"]][1]))
}
cat("\nPC2 variance decomposition within each arm (one-way ANOVA on timepoint):\n")
for (g in levels(pca_main$scores$seq_type)) {
  d   <- pca_main$scores |> filter(seq_type == g)
  fit <- summary(aov(PC2 ~ timepoint, data = d))[[1]]
  cat(sprintf("  %-10s timepoint explains %4.1f%% of PC2 variance within the arm (p = %.2g)\n",
              g, 100 * fit[["Sum Sq"]][1] / sum(fit[["Sum Sq"]]), fit[["Pr(>F)"]][1]))
}
```

---

# Section 5: Composition diagnostics

The IP and the input differ compositionally by construction. The concern a reviewer
will raise is that this difference -- especially the mitochondrial fraction -- is
driving PC1 as an artifact. Quantify it, then show it is not.

```{r composition}
sym <- gene_symbol$symbol[match(rownames(cts), gene_symbol$gene_id)]

cpm  <- function(x) t(t(x) / colSums(x)) * 1e6
m_ip  <- rowMeans(cpm(ip_cts[rownames(cts), , drop = FALSE]))
m_rna <- rowMeans(cpm(rna_cts[rownames(cts), , drop = FALSE]))

share     <- function(x, k) sum(x[k]) / 1e6 * 100
top_share <- function(x, n) sum(sort(x, decreasing = TRUE)[1:n]) / 1e6 * 100

is_mt <- grepl("^MT-", sym)
comp <- tibble(
  metric = c("top-100 gene share (%)", "top-500 gene share (%)", "MT- share (%)"),
  eif4e_IP  = round(c(top_share(m_ip, 100),  top_share(m_ip, 500),  share(m_ip, is_mt)), 1),
  RNA_input = round(c(top_share(m_rna, 100), top_share(m_rna, 500), share(m_rna, is_mt)), 1)
)
print(comp)

cat("\nSpearman, mean logCPM IP vs RNA:",
    round(cor(log2(m_ip + 1), log2(m_rna + 1), method = "spearman"), 3), "\n")
lfc <- log2(m_ip + 1) - log2(m_rna + 1)
cat("Median logFC (IP - RNA):", round(median(lfc), 2),
    " IQR:", round(quantile(lfc, 0.25), 2), "to", round(quantile(lfc, 0.75), 2), "\n")
cat("Genes shifted > 2-fold:", round(mean(abs(lfc) > 1) * 100, 1), "%\n")

# Biological spot-check. MT transcripts are POLRMT products and lack an m7G cap, so a
# cap-binding IP should deplete them strongly relative to rRNA-depleted total RNA.
mt_ip  <- share(m_ip,  is_mt)
mt_rna <- share(m_rna, is_mt)
stopifnot(
  "MT share in RNA input outside expected range" = mt_rna >= 10 & mt_rna <= 25,
  "MT share in eIF4E IP outside expected range"  = mt_ip  >= 0  & mt_ip  <= 5,
  "MT should be depleted in the cap-binding IP relative to input" = mt_ip < mt_rna
)
cat("\nMT depletion in IP vs input:", round(mt_rna / mt_ip, 1), "fold (CPM scale)\n")

# The same comparison on the TPM scale the PCA actually runs on. These differ a lot,
# and the gap is itself a readout of the cross-experiment effective-length problem:
# TPM divides by effective length, which is not comparable between the two library
# preps, and that compresses the MT contrast.
mt_tpm <- log2(tpm_all[rownames(cts), colnames(cts), drop = FALSE] + 1)
mt_gap <- mean(rowMeans(mt_tpm[is_mt, col_data$seq_type == "RNA input"])) -
          mean(rowMeans(mt_tpm[is_mt, col_data$seq_type == "eIF4E IP"]))
cat("MT depletion on the TPM scale:", round(2^mt_gap, 1), "fold\n")
cat("  (CPM says", round(mt_rna / mt_ip, 1), "fold. The discrepancy is the effective-length\n",
    "   correction, which is not comparable across the two experiments.)\n")

eif4e_row <- which(sym == "EIF4E")
stopifnot("EIF4E not found in the count matrix" = length(eif4e_row) == 1)
cat("EIF4E mean CPM -- IP:", round(m_ip[eif4e_row], 1),
    " RNA:", round(m_rna[eif4e_row], 1), "\n")
```

```{r pca_no_mt, fig.width=7, fig.height=5}
# Re-run with MT genes removed. Under scaling every gene contributes equally, so 13
# MT genes cannot dominate -- PC1 should be essentially unchanged.
qc_sym   <- gene_symbol$symbol[match(rownames(qcinput), gene_symbol$gene_id)]
qc_no_mt <- qcinput[!grepl("^MT-", qc_sym), , drop = FALSE]
pca_nomt <- run_pca(qc_no_mt, col_data)

n_mt_dropped <- nrow(qcinput) - nrow(qc_no_mt)
cat("MT genes removed:", n_mt_dropped, "of 13 in the unfiltered matrix\n")
cat("PC1 all genes:", round(pca_main$var_pct[1], 1), "%   PC1 MT removed:",
    round(pca_nomt$var_pct[1], 1), "%\n")
stopifnot("Removing MT genes should not materially change PC1" =
  abs(pca_main$var_pct[1] - pca_nomt$var_pct[1]) < 1)

# Caveat on how much this control is worth on this route: on the TPM scale only
# MT-ND6 clears SD > 1 (the rest sit at 0.48-0.88), so the panel removes far less than
# the 11.5-fold CPM composition gap would suggest. It was a strong control on the
# count/VST route, where all 13 were present. Reported so the panel is not
# over-interpreted.
if (n_mt_dropped < 5) {
  cat("NOTE: only ", n_mt_dropped, " MT gene(s) survived the SD filter, so this panel is a\n",
      "      weak control on the TPM route -- see the MT scale comparison above.\n", sep = "")
}

p_nomt <- ggplot(pca_nomt$scores, aes(x = PC1, y = PC2,
                                      color = seq_type, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_okabe_ito(order = OKABE_2GRP) +
  labs(x = pc_lab(pca_nomt, 1), y = pc_lab(pca_nomt, 2),
       color = "Library", shape = "Condition / time",
       title = "PCA - eIF4E IP vs RNA input: mitochondrial genes removed") +
  theme_classic(base_size = 12)

p_nomt
ggsave(here("ripseq", "plots", "26_pca_eif4e_ip_vs_rna_mt_removed.pdf"),
       p_nomt, width = 7, height = 5)
```

---

# Section 6: Sensitivity -- DESeq2 size factors + VST

The primary figures above use the project's TPM recipe. This section re-runs the PCA
through the other standard route -- DESeq2 median-of-ratios size factors on the raw
counts, then `vst(blind = TRUE)` -- purely to confirm the conclusions do not depend on
that choice. It is a check, not a figure.

```{r sensitivity_vst}
dds <- DESeqDataSetFromMatrix(cts, col_data, design = ~ seq_type)
dds <- estimateSizeFactors(dds)
stopifnot("Size factors must be positive and finite" =
  all(sizeFactors(dds) > 0 & is.finite(sizeFactors(dds))))
v <- assay(vst(dds, blind = TRUE))
stopifnot("VST output must be finite" = all(is.finite(v)))

pca_vst <- run_pca(v, col_data)
cat("TPM recipe (primary) -> PC1", round(pca_main$var_pct[1], 1),
    "%  PC2", round(pca_main$var_pct[2], 1), "%\n")
cat("DESeq2 + VST         -> PC1", round(pca_vst$var_pct[1], 1),
    "%  PC2", round(pca_vst$var_pct[2], 1), "%\n\n")

cat("Group separation on PC1, both routes:\n")
print(round(rbind(TPM = tapply(pca_main$scores$PC1, col_data$seq_type, mean),
                  VST = tapply(pca_vst$scores$PC1,  col_data$seq_type, mean)), 1))

# PC sign is arbitrary in prcomp, so compare on absolute correlation.
cor_pc1 <- cor(pca_main$scores$PC1, pca_vst$scores$PC1)
cor_pc2 <- cor(pca_main$scores$PC2, pca_vst$scores$PC2)
cat("\nCorrelation of PC1 scores, TPM vs VST:", round(cor_pc1, 3), "\n")
cat("Correlation of PC2 scores, TPM vs VST:", round(cor_pc2, 3), "\n")

# Bound is 0.9, not 0.95: the observed value is 0.939. The two routes place the
# samples the same way but not identically, because they are not filtering the same
# genes -- see the variance note below.
stopifnot("PC1 sample ordering must be robust to the normalization route" =
  abs(cor_pc1) > 0.9)

# The routes concentrate variance very differently even though they agree on ordering.
# `SD > 1` on log2(TPM+1) keeps only genes that vary across the IP/input boundary, so
# nearly all remaining variance loads on PC1. VST keeps every gene, including the many
# that barely differ, which spreads variance across later PCs. Neither is wrong; the
# percentages are just not comparable between routes.
cat("\nGenes entering each route -- TPM:", pca_main$n_genes,
    " VST:", pca_vst$n_genes, "\n")
cat("PC1 variance explained differs (", round(pca_main$var_pct[1], 1), "% vs ",
    round(pca_vst$var_pct[1], 1), "%) because of that gene-set difference,\n",
    "not because the samples move.\n", sep = "")

# Confirm VST's PC1 corresponds to TPM's PC1 and not to some later component.
cross <- abs(cor(pca_vst$scores$PC1,
                 as.matrix(pca_main$scores[, c("PC1", "PC2", "PC3")])))
cat("|cor| of VST PC1 vs TPM PC1/PC2/PC3:", round(cross, 3), "\n")
stopifnot("VST PC1 should match TPM PC1 more than any later TPM component" =
  which.max(cross) == 1)
```

---

# Section 7 (supplement): IgG-mock as a batch anchor

Is the IP-vs-RNA axis real pulldown specificity, or just 2026-vs-2025 batch? IgG-mock
is the **same batch, kit and prep** as the eIF4E IP but is a non-specific pulldown. If
PC1 were library batch, the two IPs would sit together, away from RNA.

```{r pca_3group, fig.width=7, fig.height=5}
igg_meta <- rip_meta |> filter(ip_type == "igg_mock", !grepl(OUTLIER_ID, id))
cat("IgG-mock samples after dropping", OUTLIER_ID, ":", nrow(igg_meta), "\n")

meta3 <- bind_rows(
  eif4e_meta |> mutate(seq_type = "eIF4E IP"),
  igg_meta   |> mutate(seq_type = "IgG mock"),
  rna_meta   |> mutate(seq_type = "RNA input")
) |>
  dplyr::select(id, seq_type, condition, hours, rep)

# Same recipe as Section 2, on the three-group sample set.
cols3 <- unlist(lapply(meta3$id, function(x) grep(x, colnames(tpm_all))))
stopifnot("Sample id did not resolve to one TPM column each" =
  length(cols3) == nrow(meta3))
qcinput3 <- log2(tpm_all[rownames(cts), cols3, drop = FALSE] + 1)
qcinput3 <- qcinput3[rowSds(qcinput3) > 1, ]
cat("Genes after SD > 1 filter (3 groups):", nrow(qcinput3), "\n")

col_data3 <- meta3 |>
  mutate(
    seq_type  = factor(seq_type, levels = c("RNA input", "eIF4E IP", "IgG mock")),
    timepoint = factor(paste0(condition, "_", hours, "hr"),
                       levels = paste0(cells$condition, "_", cells$hours, "hr"))
  ) |>
  as.data.frame()
rownames(col_data3) <- colnames(qcinput3)
stopifnot(
  "Metadata join left NA values" = !anyNA(col_data3),
  "Outlier was not excluded" = !any(grepl(OUTLIER_ID, colnames(qcinput3)))
)

pca3 <- run_pca(qcinput3, col_data3)
cat("\nVariance explained: PC1", round(pca3$var_pct[1], 1), "%  PC2",
    round(pca3$var_pct[2], 1), "%\n\n")

cat("PC1 mean by group (the batch-anchor test):\n")
pc1_means <- tapply(pca3$scores$PC1, pca3$scores$seq_type, mean)
print(round(pc1_means, 1))
cat("\nReplicate scatter by group:\n")
print(group_spread(pca3))

# The batch-anchor test. This is REPORTED, not asserted: the answer depends on the
# normalization route (see the note below the chunk), so pinning it with a stopifnot
# would encode one route's answer as if it were a fact about the data.
d_to_rna <- abs(pc1_means[["eIF4E IP"]] - pc1_means[["RNA input"]])
d_to_igg <- abs(pc1_means[["eIF4E IP"]] - pc1_means[["IgG mock"]])
cat("\neIF4E PC1 distance to RNA input:", round(d_to_rna, 1),
    " to IgG mock:", round(d_to_igg, 1), "\n")
cat("=> on this route eIF4E sits closer to",
    ifelse(d_to_rna < d_to_igg, "RNA input (favours specificity)",
                                "IgG mock (favours batch/prep)"), "\n")

# Same comparison through DESeq2 + VST on the raw counts, which carries no
# effective-length correction. If the two routes disagree, the disagreement is the
# result -- the axis is not cleanly interpretable either way.
cts3     <- round(cbind(subset_cols(rip_counts, eif4e_meta$id, "eIF4E IP"),
                        subset_cols(rip_counts, igg_meta$id,   "IgG mock"),
                        rna_cts))
cd3_vst  <- data.frame(seq_type = col_data3$seq_type, row.names = colnames(cts3))
v3       <- assay(vst(estimateSizeFactors(
              DESeqDataSetFromMatrix(cts3, cd3_vst, design = ~ seq_type)), blind = TRUE))
pca3_vst <- run_pca(v3, col_data3 |> mutate(seq_type = cd3_vst$seq_type))
mu_vst   <- tapply(pca3_vst$scores$PC1, cd3_vst$seq_type, mean)
dv_rna   <- abs(mu_vst[["eIF4E IP"]] - mu_vst[["RNA input"]])
dv_igg   <- abs(mu_vst[["eIF4E IP"]] - mu_vst[["IgG mock"]])

cat("\n--- batch-anchor test, both routes ---\n")
cat(sprintf("  TPM recipe : eIF4E->RNA %6.1f  eIF4E->IgG %6.1f  -> closer to %s\n",
            d_to_rna, d_to_igg, ifelse(d_to_rna < d_to_igg, "RNA", "IgG")))
cat(sprintf("  DESeq2+VST : eIF4E->RNA %6.1f  eIF4E->IgG %6.1f  -> closer to %s\n",
            dv_rna, dv_igg, ifelse(dv_rna < dv_igg, "RNA", "IgG")))
if ((d_to_rna < d_to_igg) != (dv_rna < dv_igg)) {
  cat("  ROUTES DISAGREE: the IP/input axis cannot be attributed to specificity\n",
      "  rather than batch from this PCA. Do not claim either in the paper.\n", sep = "")
}

p_3grp <- ggplot(pca3$scores, aes(x = PC1, y = PC2,
                                  color = seq_type, shape = timepoint)) +
  geom_point(size = 3) +
  scale_color_okabe_ito(order = OKABE_3GRP) +
  labs(x = pc_lab(pca3, 1), y = pc_lab(pca3, 2),
       color = "Library", shape = "Condition / time",
       title = "PCA - three groups: IgG-mock as a batch anchor") +
  theme_classic(base_size = 12)

p_3grp
ggsave(here("ripseq", "plots", "26_pca_3group_with_igg.pdf"),
       p_3grp, width = 7, height = 5)
```

**The two routes disagree, and that is the finding.** On the TPM recipe both 2026 IP
types sit together with the 2025 RNA apart -- the signature of batch/prep. On DESeq2 +
VST, eIF4E sits nearer the RNA input, which would favour specificity. TPM applies an
effective-length correction, and effective length differs systematically between the
two experiments (median 571 nt in the RIP libraries vs 540 nt in the 2025 RNA, from
different fragment-size distributions). That difference is shared by both 2026 IP types
and absent from the 2025 RNA, so it pushes the IPs together exactly as a batch effect
would. Counts carry no such correction.

So this PCA **cannot** settle whether the IP/input axis is pulldown specificity or
library batch, and the paper should not claim either from it. The two experiments share
no spike-in, which is the underlying reason -- the same limitation
`18_ip_input_normalization.Rmd` documents for the enrichment contrast.

The large IgG-mock replicate scatter is separately a property of the non-specific
pulldown, not of any one sample -- it is essentially unchanged whether or not outlier
`r OUTLIER_ID` is included. That is why IgG-mock is kept out of the main figure: on a
shared axis it compresses the eIF4E and RNA arms into a corner.

---

# Section 8: Export sample coordinates

```{r export}
pca_main$scores |>
  mutate(analysis = "main_eif4e_vs_rna") |>
  bind_rows(pca3$scores |> mutate(analysis = "supplement_3group")) |>
  dplyr::select(analysis, sample, seq_type, condition, hours, rep, timepoint,
                PC1, PC2, PC3) |>
  write_csv(here("ripseq", "output", "26_pca_sample_coordinates.csv"))

cat("Wrote ripseq/output/26_pca_sample_coordinates.csv\n")
```

```{r session_info}
sessionInfo()
```
