---
title: "DHX29 reanalysis 01: Thor-Ribo-seq count distributions (GSE266018)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Raw selective ribosome profiling counts behind Hia et al., *Human DHX29 detects nonoptimal
codon usage to regulate mRNA stability* (GEO **GSE266018**). HEK293T cells expressing
FLAG-DHX29; ribosome footprints from the input and from the FLAG-DHX29 pull-down were
subjected to Thor-Ribo-Seq. Three Input replicates, three DHX29-IP replicates.

The repo already touches this paper through its supplementary tables in
`code/predictive_modeling/23_dhx29_riboseq_slamseq.Rmd` (Table S8 codon enrichment, Table S6
SLAM-seq half-lives). Starting from the raw per-transcript counts instead opens up per-gene
IP/Input enrichment, which the supplementary tables do not provide.

**Scope.** This notebook is an ingest and QC step. It loads all six files, shows the
distribution of both data columns, and then verifies the annotation the files carry. Nothing
here reaches a feature matrix. Per the repository convention on external data, the paper's
codon-level result is already replicated against Table S8 in notebook 23; reproducing that
codon enrichment from these raw counts belongs to the next notebook in this folder, and must
happen before these counts are joined to any feature matrix.

**File format.** Tab-delimited, no header, three columns. GEO's own processing note says
"comma-delimited"; it is wrong. Column 1 is a versioned ENSEMBL transcript ID, column 2 is a
CDS length in codons with a literal `N/A` for non-coding transcripts, column 3 is the raw
footprint count. The section below verifies that column 3 is raw counts; sections 5b and 5d
test the identity of column 2. Neither is taken on the documentation's word.

```{r load_libraries}
library(tidyverse)
library(here)
library(DBI)
library(RSQLite)

assay_colors <- c("Input" = "#0072B2", "DHX29-IP" = "#CC79A7")
```

---

## Section 0: Sample manifest

Typed out from the GEO series metadata.

```{r samples}
samples <- tibble(
  gsm    = c("GSM8239469", "GSM8239470", "GSM8239471",
             "GSM8239472", "GSM8239473", "GSM8239474"),
  sample = c("Input1", "Input2", "Input3", "IP1", "IP2", "IP3"),
  assay  = c("Input", "Input", "Input", "DHX29-IP", "DHX29-IP", "DHX29-IP"),
  rep    = c(1L, 2L, 3L, 1L, 2L, 3L)
) %>%
  mutate(
    path   = here("accessories", "dhx29", "GSE266018_RAW",
                  paste0(gsm, "_", sample, ".txt.gz")),
    assay  = factor(assay, levels = c("Input", "DHX29-IP")),
    sample = factor(sample, levels = c("Input1", "Input2", "Input3",
                                       "IP1", "IP2", "IP3"))
  )

stopifnot("GSE266018 count file missing" = all(file.exists(samples$path)))
print(samples %>% dplyr::select(gsm, sample, assay, rep))
```

---

## Section 1: Load all six files

`read_tsv` reads the vector of gzipped paths in one call and tags each row with its source
file via `id`, so no loop and no helper are needed.

```{r load_counts}
counts <- read_tsv(
  samples$path,
  col_names = c("transcript_id", "cds_codons", "count"),
  na        = "N/A",
  col_types = "cii",
  id        = "path"
) %>%
  left_join(samples, by = "path") %>%
  mutate(transcript_id_clean = str_remove(transcript_id, "\\..*")) %>%
  dplyr::select(transcript_id, transcript_id_clean, cds_codons, count,
                sample, assay, rep, gsm)

cat("Rows loaded:", format(nrow(counts), big.mark = ","), "\n")
cat("Distinct transcripts:",
    format(n_distinct(counts$transcript_id), big.mark = ","), "\n")

# The length column is an annotation, not a measurement: one row per transcript.
tx_annot <- counts %>%
  distinct(transcript_id, transcript_id_clean, cds_codons)

n_tx        <- nrow(tx_annot)
n_coding    <- sum(!is.na(tx_annot$cds_codons))
n_noncoding <- sum(is.na(tx_annot$cds_codons))
n_neg_len   <- sum(tx_annot$cds_codons < 0, na.rm = TRUE)

cat("Transcripts:", format(n_tx, big.mark = ","),
    "| length present:", format(n_coding, big.mark = ","),
    "| N/A:", format(n_noncoding, big.mark = ","),
    "| length < 0:", n_neg_len, "\n")
```

---

### Is column 3 really raw counts?

GEO's processing note says so, but the same sentence calls these files comma-delimited when
they are tab-delimited, so it is corroborated here rather than trusted. Each check below
eliminates a specific alternative.

```{r verify_counts_are_counts}
wide_chk <- counts %>%
  dplyr::select(transcript_id, cds_codons, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count)
sample_cols <- c("Input1", "Input2", "Input3", "IP1", "IP2", "IP3")

# 1. Integer-valued throughout.
n_noninteger <- sum(counts$count != as.integer(counts$count))
cat("1. Non-integer values across all", format(nrow(counts), big.mark = ","),
    "numbers:", n_noninteger, "\n")

# 2. Library sizes differ. CPM and TPM sum to exactly 1e6 in every column.
lib <- colSums(wide_chk[, sample_cols])
cat("2. Column sums span", format(min(lib), big.mark = ","), "to",
    format(max(lib), big.mark = ","),
    sprintf("(%.2f-fold). CPM/TPM would be identical.\n", max(lib) / min(lib)))

# 3. Counts rise with CDS length. TPM, RPKM and FPKM divide length out.
# Non-coding rows carry no length, so restrict to coding transcripts.
coding_chk <- wide_chk %>% filter(!is.na(cds_codons))
rho_len <- cor(coding_chk$Input1, coding_chk$cds_codons, method = "spearman")
cat("3. Spearman(count, CDS codons) =", round(rho_len, 3),
    "over", format(nrow(coding_chk), big.mark = ","),
    "coding transcripts -> not a length-normalised measure\n")

# 4. The smallest non-zero value is the quantum of a count.
cat("4. Smallest non-zero value per sample:",
    paste(sapply(wide_chk[, sample_cols], function(x) min(x[x > 0])), collapse = " "), "\n")

# 5. Mean-variance follows a negative binomial: Poisson at low mean, overdispersed at high.
mv <- tibble(m = rowMeans(coding_chk[, c("Input1", "Input2", "Input3")]),
             v = apply(coding_chk[, c("Input1", "Input2", "Input3")], 1, var)) %>%
  filter(m > 0) %>%
  mutate(bin = cut(m, breaks = c(0, 1, 2, 5, 10, 50, 200, Inf),
                   labels = c("<=1", "1-2", "2-5", "5-10", "10-50", "50-200", ">200")))
mv_tab <- mv %>%
  group_by(mean_bin = bin) %>%
  summarise(n = n(), median_mean = median(m), median_var = median(v), .groups = "drop") %>%
  mutate(var_over_mean = median_var / median_mean,
         implied_dispersion = (var_over_mean - 1) / median_mean)
cat("\n5. Mean-variance across the three Input replicates:\n")
print(as.data.frame(mv_tab %>% mutate(across(where(is.numeric), ~ round(.x, 3)))))

stopifnot(
  "column 3 contains non-integer values, so it is not raw counts" = n_noninteger == 0,
  "library sizes are identical, which would indicate a normalised measure" =
    max(lib) / min(lib) > 1.05,
  "counts do not scale with CDS length, which would indicate length normalisation" =
    rho_len > 0.1,
  "the smallest non-zero value is not 1" =
    all(sapply(wide_chk[, sample_cols], function(x) min(x[x > 0])) == 1),
  "variance does not approach the mean at low counts, so the Poisson limit is absent" =
    between(mv_tab$var_over_mean[mv_tab$mean_bin == "5-10"], 0.8, 1.2),
  "overdispersion does not grow with the mean as a negative binomial requires" =
    mv_tab$var_over_mean[mv_tab$mean_bin == ">200"] >
    mv_tab$var_over_mean[mv_tab$mean_bin == "10-50"]
)
cat("\nAll format checks passed.\n")
```

```{r plot_mean_variance, fig.width = 7, fig.height = 5}
phi <- median(mv_tab$implied_dispersion[mv_tab$median_mean > 10])
curve_df <- tibble(m = 10^seq(log10(min(mv$m)), log10(max(mv$m)), length.out = 200)) %>%
  mutate(poisson = m, nb = m + phi * m^2)

p_mv <- ggplot(mv, aes(m, v)) +
  geom_point(size = 0.3, alpha = 0.12, colour = "grey40") +
  geom_line(data = curve_df, aes(m, poisson), colour = "#0072B2",
            linewidth = 0.9, linetype = "dashed") +
  geom_line(data = curve_df, aes(m, nb), colour = "#CC79A7", linewidth = 0.9) +
  geom_point(data = mv_tab, aes(median_mean, median_var), colour = "black",
             size = 2.5, inherit.aes = FALSE) +
  scale_x_log10() + scale_y_log10() +
  annotate("text", x = min(mv$m), y = max(mv$v),
           label = paste0("dashed: Poisson, variance = mean\n",
                          "solid: negative binomial, dispersion = ",
                          sprintf("%.3f", phi)),
           hjust = 0, vjust = 1, size = 3.2) +
  labs(title = "Column 3 behaves like counts, not like a normalised measure",
       subtitle = str_wrap(paste0(
         "Per-transcript mean against variance across the three Input replicates, n = ",
         format(nrow(mv), big.mark = ","),
         ". Black points are the bin medians. Low counts sit on the Poisson line and ",
         "overdispersion grows with the mean, as a negative binomial requires. The ",
         "horizontal banding at low mean is integer discreteness."), width = 84),
       x = "mean count across Input replicates", y = "variance") +
  theme_classic(base_size = 12)

print(p_mv)
ggsave(here("plots", "dhx29_01_mean_variance.pdf"), p_mv, width = 7, height = 5)
```

The fifth check is the positive evidence; the others eliminate alternatives. Variance over
mean approaches 1 at low counts, the Poisson limit of sampling discrete molecules, then rises
linearly as `var = mean + dispersion * mean^2`. The implied dispersion is roughly constant
near 0.03 across the well-measured bins, which is a single sensible parameter rather than an
arbitrary shape. No normalised quantity reproduces that.

The low-mean bins fall slightly below 1 because a variance estimated from three replicates is
noisy and discreteness bites at counts near zero. That is expected and not evidence against.

---

## Section 2: Per-sample summary

```{r sample_summary}
sample_summary <- counts %>%
  group_by(gsm, sample, assay, rep) %>%
  summarise(
    n_rows                = n(),
    n_length_present      = sum(!is.na(cds_codons)),
    n_detected            = sum(count > 0),
    library_size          = sum(count),
    median_count_detected = median(count[count > 0]),
    p90_count_detected    = quantile(count[count > 0], 0.90),
    .groups = "drop"
  )

print(as.data.frame(sample_summary))

ip_input_ratio <- sum(sample_summary$library_size[sample_summary$assay == "DHX29-IP"]) /
  sum(sample_summary$library_size[sample_summary$assay == "Input"])
cat("\nTotal DHX29-IP / Input library size ratio:", round(ip_input_ratio, 3), "\n")
```

---

## Section 3: Plot 1, footprint count distribution

All rows, `log2(count + 1)`, six curves. The spike at zero is mostly the non-coding
transcripts, which carry no counts at all; the subtitle gives the split so the spike is
readable.

```{r plot_count_distribution, fig.width = 9, fig.height = 4.5}
count_caption <- counts %>%
  group_by(sample) %>%
  summarise(n_det = sum(count > 0), .groups = "drop") %>%
  mutate(lab = paste0(sample, " ", format(n_det, big.mark = ","))) %>%
  pull(lab) %>%
  paste(collapse = "   ")

# Both panels are the same data. The right panel drops nothing from the analysis; it
# only lets the shape above zero be seen past the spike that the left panel is made of.
counts_panelled <- bind_rows(
  counts %>% mutate(panel = "all transcripts"),
  counts %>% filter(count > 0) %>% mutate(panel = "count > 0")
) %>%
  mutate(panel = factor(panel, levels = c("all transcripts", "count > 0")))

p_counts <- ggplot(counts_panelled, aes(x = log2(count + 1),
                                        colour = assay, fill = assay,
                                        group = sample, linetype = factor(rep))) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_colour_manual(values = assay_colors, name = "Assay") +
  scale_fill_manual(values = assay_colors, name = "Assay") +
  scale_linetype_manual(values = c("solid", "dashed", "dotted"), name = "Replicate") +
  labs(
    title    = "Thor-Ribo-seq footprint counts, GSE266018",
    subtitle = str_wrap(paste0(
      "n = ", format(n_tx, big.mark = ","), " transcripts per sample: ",
      format(n_coding, big.mark = ","), " with a CDS length and ",
      format(n_noncoding, big.mark = ","),
      " non-coding, the latter all zero in all six files."), width = 95),
    caption  = str_wrap(paste0("transcripts with count > 0 per sample:  ",
                               count_caption), width = 95),
    x        = "log2(count + 1)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.caption = element_text(hjust = 0))

print(p_counts)
ggsave(here("plots", "dhx29_01_count_distribution.pdf"),
       p_counts, width = 9, height = 4.5)
```

---

## Section 4: Plot 2, CDS length distribution

Column 2 is byte-identical in all six files, so there is nothing per-sample to overlay: this
is one curve over the distinct transcript set. `log2` is undefined on the negative lengths,
so those rows are dropped here; they carry no counts anywhere.

```{r plot_length_distribution, fig.width = 8, fig.height = 4.5}
len_plot_data <- tx_annot %>% filter(!is.na(cds_codons), cds_codons >= 0)

cat("Transcripts in the length plot:",
    format(nrow(len_plot_data), big.mark = ","), "\n")
cat("Dropped for negative length:", n_neg_len,
    "| their total count across all six samples:",
    sum(counts$count[!is.na(counts$cds_codons) & counts$cds_codons < 0]), "\n\n")

# The log2 axis is readable but not interpretable, so report the native units too.
len_median <- median(len_plot_data$cds_codons)
cat("Column 2 in its native units (codons):\n")
print(round(quantile(len_plot_data$cds_codons,
                     c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99))))
cat("mean", round(mean(len_plot_data$cds_codons), 1),
    "| min", min(len_plot_data$cds_codons),
    "| max", max(len_plot_data$cds_codons), "\n")

p_len <- ggplot(len_plot_data, aes(x = log2(cds_codons + 1))) +
  geom_density(alpha = 0.15, linewidth = 0.8,
               colour = "grey20", fill = "grey70") +
  geom_vline(xintercept = log2(len_median + 1), linetype = "dashed",
             colour = "#0072B2") +
  annotate("text", x = log2(len_median + 1), y = Inf,
           label = paste0("median = ", len_median, " codons"),
           hjust = -0.08, vjust = 2, size = 3.2, colour = "#0072B2") +
  labs(
    title    = "CDS length column, GSE266018",
    subtitle = str_wrap(paste0(
      "n = ", format(nrow(len_plot_data), big.mark = ","),
      " transcripts. This column is identical in all six files, so there is one ",
      "curve, not six."), width = 90),
    x        = "log2(CDS codons + 1)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12)

print(p_len)
ggsave(here("plots", "dhx29_01_cds_length_distribution.pdf"),
       p_len, width = 8, height = 4.5)
```

---

## Section 5: Is column 2 really CDS length, and are the transcript versions right

### 5a: Pull CDS lengths independently from the repo transcript database

`accessories/human/txdb.gencode49.sqlite` stores CDS ranges with the stop codon merged in.

```{r txdb_cds_lengths}
con <- dbConnect(RSQLite::SQLite(),
                 here("accessories", "human", "txdb.gencode49.sqlite"))

v49_tx    <- tbl(con, "transcript") %>% dplyr::select(`_tx_id`, tx_name) %>% collect()
v49_splic <- tbl(con, "splicing")   %>% dplyr::select(`_tx_id`, `_cds_id`) %>% collect()
v49_cdsr  <- tbl(con, "cds")        %>% dplyr::select(`_cds_id`, cds_start, cds_end) %>% collect()
v49_gene  <- tbl(con, "gene")       %>% collect() %>% distinct(`_tx_id`, .keep_all = TRUE)

v49_cds <- v49_splic %>%
  filter(!is.na(`_cds_id`)) %>%
  inner_join(v49_cdsr, by = "_cds_id") %>%
  group_by(`_tx_id`) %>%
  summarise(cds_len = sum(cds_end - cds_start + 1), .groups = "drop") %>%
  inner_join(v49_tx, by = "_tx_id") %>%
  left_join(v49_gene, by = "_tx_id") %>%
  mutate(tx_clean = str_remove(tx_name, "\\..*")) %>%
  dplyr::select(tx_name, tx_clean, gene_id, cds_len)

cat("v49 transcripts total:", format(nrow(v49_tx), big.mark = ","), "\n")
cat("v49 transcripts with a CDS:", format(nrow(v49_cds), big.mark = ","), "\n")
```

#### Plot 3: annotated CDS length in nucleotides

The Section 4 figure for the length just pulled here, same construction, native nucleotides.
Restricted to the transcripts that appear in these GEO files, so it describes the same
population rather than all of GENCODE.

```{r plot_cds_length_nt, fig.width = 8, fig.height = 4.5}
nt_plot_data <- tx_annot %>%
  inner_join(v49_cds, by = c("transcript_id_clean" = "tx_clean")) %>%
  filter(cds_len > 0)

nt_median <- median(nt_plot_data$cds_len)

cat("Transcripts in these files with a v49 CDS:",
    format(nrow(nt_plot_data), big.mark = ","), "\n")
cat("Annotated CDS length in nucleotides:\n")
print(round(quantile(nt_plot_data$cds_len,
                     c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99))))
cat("mean", round(mean(nt_plot_data$cds_len), 1),
    "| min", min(nt_plot_data$cds_len),
    "| max", max(nt_plot_data$cds_len), "\n")

p_len_nt <- ggplot(nt_plot_data, aes(x = log2(cds_len + 1))) +
  geom_density(alpha = 0.15, linewidth = 0.8,
               colour = "grey20", fill = "grey70") +
  geom_vline(xintercept = log2(nt_median + 1), linetype = "dashed",
             colour = "#0072B2") +
  annotate("text", x = log2(nt_median + 1), y = Inf,
           label = paste0("median = ", nt_median, " nucleotides"),
           hjust = -0.08, vjust = 2, size = 3.2, colour = "#0072B2") +
  labs(
    title    = "Annotated CDS length, GENCODE v49",
    subtitle = str_wrap(paste0(
      "n = ", format(nrow(nt_plot_data), big.mark = ","),
      " transcripts from these files that carry a v49 CDS. Pulled from the transcript ",
      "database above, not read from the GEO files."), width = 90),
    x        = "log2(CDS nucleotides + 1)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12)

print(p_len_nt)
ggsave(here("plots", "dhx29_01_cds_length_nt_distribution.pdf"),
       p_len_nt, width = 8, height = 4.5)
```


### 5b: Are the `N/A` rows really non-coding

Tested against `accessories/human/tx_biotype_v49.tsv.gz`, a transcript-level biotype table.
It is a separate file read on a separate path from the sqlite database, so this is
independent evidence rather than the same evidence twice.

```{r biotype_check}
tx_biotype <- read_tsv(here("accessories", "human", "tx_biotype_v49.tsv.gz"),
                       col_types = cols(.default = col_character())) %>%
  mutate(
    tx_clean        = str_remove(transcript_id, "\\..*"),
    v49_cds_bearing = transcript_type %in% c("protein_coding", "nonsense_mediated_decay",
                                             "non_stop_decay", "protein_coding_LoF",
                                             "polymorphic_pseudogene") |
                      str_detect(transcript_type, "^(IG|TR)_[VDJC]_gene$")
  ) %>%
  dplyr::select(tx_clean, v49_transcript_type = transcript_type, v49_cds_bearing)

annot_check <- tx_annot %>%
  mutate(
    tx_clean   = transcript_id_clean,
    geo_coding = if_else(is.na(cds_codons), "GEO N/A", "GEO length present")
  ) %>%
  left_join(tx_biotype, by = "tx_clean") %>%
  left_join(v49_cds, by = "tx_clean") %>%
  mutate(
    in_v49         = !is.na(v49_cds_bearing),
    v49_has_cds_row = !is.na(cds_len),
    match_class    = case_when(
      transcript_id %in% v49_tx$tx_name ~ "exact",
      tx_clean %in% str_remove(v49_tx$tx_name, "\\..*") ~ "drift",
      TRUE ~ "absent"
    ),
    match_class = factor(match_class, levels = c("exact", "drift", "absent"))
  )

stopifnot("annot_check gained or lost rows in the joins" = nrow(annot_check) == n_tx)

cat("Transcripts absent from v49 and therefore unjudged:",
    format(sum(!annot_check$in_v49), big.mark = ","), "\n\n")

cat("GEO coding status vs independent v49 biotype (v49-recognised transcripts only):\n")
two_by_two <- annot_check %>%
  filter(in_v49) %>%
  dplyr::count(geo_coding, v49_cds_bearing) %>%
  pivot_wider(names_from = v49_cds_bearing, values_from = n,
              names_prefix = "v49_cds_bearing_", values_fill = 0)
print(as.data.frame(two_by_two))

concord_na <- annot_check %>%
  filter(in_v49, geo_coding == "GEO N/A") %>%
  summarise(n = n(), ok = sum(!v49_cds_bearing)) %>%
  mutate(frac = ok / n)
concord_cd <- annot_check %>%
  filter(in_v49, geo_coding == "GEO length present") %>%
  summarise(n = n(), ok = sum(v49_cds_bearing)) %>%
  mutate(frac = ok / n)

cat("\nN/A rows independently non-coding:  ",
    format(concord_na$ok, big.mark = ","), "/", format(concord_na$n, big.mark = ","),
    " = ", sprintf("%.2f%%", 100 * concord_na$frac), "\n", sep = "")
cat("Length-bearing rows independently CDS-bearing:  ",
    format(concord_cd$ok, big.mark = ","), "/", format(concord_cd$n, big.mark = ","),
    " = ", sprintf("%.2f%%", 100 * concord_cd$frac), "\n\n", sep = "")

cat("Where the disagreement sits, by v49 biotype:\n")
print(as.data.frame(
  annot_check %>%
    filter(in_v49,
           (geo_coding == "GEO N/A") == v49_cds_bearing) %>%
    dplyr::count(geo_coding, v49_transcript_type, sort = TRUE) %>%
    head(12)
))

# Two routes to the same call: the biotype label, and presence of CDS ranges in the database.
route_agreement <- annot_check %>%
  filter(in_v49) %>%
  dplyr::count(v49_cds_bearing, v49_has_cds_row)
cat("\nBiotype label vs presence of CDS ranges in the txdb:\n")
print(as.data.frame(route_agreement))

stopifnot(
  "under 99% of GEO N/A rows are independently non-coding" = concord_na$frac >= 0.99,
  "under 99% of GEO length-bearing rows are independently CDS-bearing" = concord_cd$frac >= 0.99,
  "biotype label and txdb CDS ranges disagree about which transcripts are coding" =
    all(annot_check$v49_cds_bearing[annot_check$in_v49] ==
        annot_check$v49_has_cds_row[annot_check$in_v49])
)
```

#### Is the disagreement annotation vintage, and what does it cost?

If the ~1,780 disagreements are GENCODE moving between the paper's release and v49, they
should concentrate on transcripts whose version also moved. They do, but only partly, and the
remainder is informative: a biotype relabelling changes no sequence, so it carries no version
bump. The second half of the chunk prices the disagreement in footprints, which is what
matters when these counts later meet a v49-keyed table.

```{r biotype_disagreement_vintage}
disagree <- annot_check %>%
  filter(in_v49) %>%
  mutate(agree = (geo_coding == "GEO length present") == v49_cds_bearing,
         match_class = droplevels(match_class))

cat("Disagreement rate by transcript version class:\n")
print(as.data.frame(
  disagree %>%
    dplyr::count(match_class, agree) %>%
    pivot_wider(names_from = agree, values_from = n, values_fill = 0) %>%
    dplyr::rename(n_disagree = `FALSE`, n_agree = `TRUE`) %>%
    mutate(pct_disagree = round(100 * n_disagree / (n_disagree + n_agree), 2))
))

cat("\nDirection of the disagreement, by version class:\n")
print(as.data.frame(disagree %>% filter(!agree) %>% dplyr::count(match_class, geo_coding)))

na_but_coding  <- disagree %>% filter(!agree, geo_coding == "GEO N/A")
coding_but_not <- disagree %>% filter(!agree, geo_coding == "GEO length present")

na_footprints <- sum(counts$count[counts$transcript_id %in% na_but_coding$transcript_id])
exposure <- counts %>%
  filter(transcript_id %in% coding_but_not$transcript_id) %>%
  group_by(transcript_id) %>%
  summarise(total = sum(count), .groups = "drop")

cat("\nWhat the disagreement costs:\n")
cat("  v49-coding transcripts the GEO pipeline never counted: ",
    format(nrow(na_but_coding), big.mark = ","),
    ", carrying ", na_footprints, " footprints.\n",
    "    These are structural zeros. The pipeline saw no CDS, so it counted nothing.\n",
    "    They are not evidence that the transcript is untranslated.\n", sep = "")
cat("  v49-non-coding transcripts that do carry footprints: ",
    format(sum(exposure$total > 0), big.mark = ","), " of ",
    format(nrow(exposure), big.mark = ","), ", ",
    format(sum(exposure$total), big.mark = ","), " footprints, ",
    sprintf("%.3f%%", 100 * sum(exposure$total) / sum(counts$count)),
    " of the library.\n",
    "    Filtering a downstream table on v49 coding status drops these silently.\n", sep = "")

rate <- disagree %>%
  group_by(match_class) %>%
  summarise(r = mean(!agree), .groups = "drop")

stopifnot(
  "version drift does not raise the biotype disagreement rate, so the vintage reading fails" =
    rate$r[rate$match_class == "drift"] > rate$r[rate$match_class == "exact"],
  "a transcript the GEO files call non-coding carries footprints" = na_footprints == 0,
  "more than 1% of footprints sit on transcripts v49 calls non-coding" =
    sum(exposure$total) / sum(counts$count) < 0.01
)
```

Drift raises the rate roughly tenfold, so vintage is real. It is not the whole story: a large
minority of disagreements sit on transcripts with an identical version, and there the
dominant direction is coding becoming non-coding. That is the signature of GENCODE
withdrawing CDS definitions from alternative isoforms and relabelling them, which leaves the
sequence untouched.


### 5c: Transcript version match against v49

```{r version_classes}
cat("Version match class vs GEO coding status:\n")
print(as.data.frame(
  annot_check %>%
    dplyr::count(geo_coding, match_class) %>%
    pivot_wider(names_from = match_class, values_from = n, values_fill = 0)
))

coding_match_rate <- annot_check %>%
  filter(geo_coding == "GEO length present") %>%
  summarise(frac = mean(match_class != "absent")) %>%
  pull(frac)

cat("\nCoding GEO transcripts matching a v49 transcript on the unversioned ID: ",
    sprintf("%.2f%%", 100 * coding_match_rate), "\n", sep = "")
```

### 5d: Test the rule `cds_codons == round(cds_len / 3) - 10`

Rounding matters. About a fifth of v49 transcripts have a CDS length that is not a multiple
of three, and `floor` fails on a third of those while `round` does not. `cds_len / 3` is
never exactly `.5`, so R's round-half-to-even never comes into play.

```{r length_rule}
rule <- annot_check %>%
  filter(!is.na(cds_codons), !is.na(cds_len), match_class != "absent") %>%
  mutate(
    cds_codons_v49 = round(cds_len / 3) - 10,
    delta          = cds_codons - cds_codons_v49
  )

rule_summary <- rule %>%
  group_by(match_class) %>%
  summarise(
    n            = n(),
    agreement    = mean(delta == 0),
    pearson_r    = cor(cds_codons, cds_codons_v49),
    median_delta = median(delta),
    .groups = "drop"
  ) %>%
  filter(n > 0)

cat("Agreement with round(cds_len / 3) - 10, split by transcript version match:\n")
print(as.data.frame(rule_summary %>% mutate(agreement = round(agreement, 4),
                                            pearson_r = round(pearson_r, 4))))

cat("\nCDS lengths not a multiple of 3 in the joined set: ",
    sprintf("%.1f%%", 100 * mean(rule$cds_len %% 3 != 0)), "\n", sep = "")
```

The split is the result. Where the transcript version is identical the rule holds almost
perfectly, which is what identifies the column as CDS length. Where the version drifted the
CDS genuinely changed between the paper's annotation and v49, and the rule degrades in
exactly the way it should.

#### Plot 4: is column 2 a codon count or a nucleotide length?

Codons. The division by three in the rule above is the whole answer, but correlation cannot
show it, because a nucleotide length and a codon count are proportional and so correlate
identically. The **slope** distinguishes them, and so does the offset.

```{r codons_or_nucleotides, fig.width = 8, fig.height = 4.5}
# Exact-version transcripts whose CDS is a whole number of codons, so that nt / 3 is
# exact and the offset can be read without rounding in the way.
codon_test <- rule %>% filter(match_class == "exact", cds_len %% 3 == 0)

slope_nt <- unname(coef(lm(cds_codons ~ cds_len, data = codon_test))[2])
offset   <- codon_test$cds_len / 3 - codon_test$cds_codons

cat("On the ", format(nrow(codon_test), big.mark = ","),
    " exact-version transcripts with a CDS length divisible by 3:\n\n", sep = "")
cat("  median column 2                 ", median(codon_test$cds_codons), " codons\n", sep = "")
cat("  median v49 CDS                  ", median(codon_test$cds_len), " nucleotides\n", sep = "")
cat("  median v49 CDS / 3              ", median(codon_test$cds_len) / 3, "\n", sep = "")
cat("  regression slope on CDS nt      ", sprintf("%.4f", slope_nt),
    "   (1/3 = 0.3333 for codons, 1 for nucleotides)\n", sep = "")
cat("  CDS nt / 3 minus column 2       median ", median(offset),
    ", equal to 10 in ", sprintf("%.2f%%", 100 * mean(offset == 10)),
    " of cases\n", sep = "")

stopifnot(
  "slope on CDS nucleotides is not 1/3, so column 2 is not a codon count" =
    abs(slope_nt - 1/3) < 0.01,
  "the gap between CDS nt / 3 and column 2 is not 10 codons" =
    median(offset) == 10,
  "the 10-codon gap is not near-universal" = mean(offset == 10) > 0.99
)

codon_long <- codon_test %>%
  transmute(
    `GEO column 2`                = cds_codons,
    `v49 CDS in codons (nt / 3)`  = cds_len / 3,
    `v49 CDS in nucleotides`      = cds_len
  ) %>%
  pivot_longer(everything(), names_to = "quantity", values_to = "value") %>%
  mutate(quantity = factor(quantity, levels = c("GEO column 2",
                                                "v49 CDS in codons (nt / 3)",
                                                "v49 CDS in nucleotides")))

p_codon <- ggplot(codon_long, aes(x = log2(value + 1),
                                  colour = quantity, fill = quantity)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_colour_manual(values = c("GEO column 2" = "#0072B2",
                                 "v49 CDS in codons (nt / 3)" = "#CC79A7",
                                 "v49 CDS in nucleotides" = "grey40"),
                      name = NULL) +
  scale_fill_manual(values = c("GEO column 2" = "#0072B2",
                               "v49 CDS in codons (nt / 3)" = "#CC79A7",
                               "v49 CDS in nucleotides" = "grey40"),
                    name = NULL) +
  labs(
    title    = "Column 2 is a codon count, not a nucleotide length",
    subtitle = str_wrap(paste0(
      "n = ", format(nrow(codon_test), big.mark = ","),
      " transcripts. The GEO column overlays the v49 CDS divided by three and sits ",
      "log2(3) = 1.58 units left of the nucleotide length."), width = 82),
    x        = "log2(value + 1)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_codon)
ggsave(here("plots", "dhx29_01_codons_vs_nucleotides.pdf"),
       p_codon, width = 8, height = 4.5)
```

The offset is a flat ten codons rather than a proportion, which says the column counts
codons over a trimmed window rather than the whole CDS. Ten is what you get by dropping five codons
at each end, the usual guard against the initiation and termination peaks in ribosome
profiling. The arithmetic is verified; the reason for the trim is an inference, since GEO
documents neither column.


#### Plot 5: nucleotides per codon

The same conclusion read the simplest possible way. Carry the externally gathered CDS length
alongside column 2 and divide one by the other. If column 2 counts codons the ratio must land
on three.

```{r nt_per_codon, fig.width = 8, fig.height = 4.5}
codon_ratio <- annot_check %>%
  filter(!is.na(cds_codons), !is.na(cds_len), cds_codons > 0) %>%
  left_join(rule %>% dplyr::select(transcript_id, delta), by = "transcript_id") %>%
  mutate(
    cds_length_external        = cds_len,
    nt_per_codon               = cds_length_external / cds_codons,
    nt_per_codon_trim_restored = cds_length_external / (cds_codons + 10)
  )

cat("Transcripts with both a column 2 value and an external CDS length: ",
    format(nrow(codon_ratio), big.mark = ","), "\n", sep = "")
cat("Excluded for a column 2 value of zero or less: ",
    sum(annot_check$cds_codons <= 0, na.rm = TRUE), "\n\n", sep = "")

cat("external CDS nt / column 2, as reported:\n")
print(round(quantile(codon_ratio$nt_per_codon,
                     c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99)), 3))
cat("within 0.1 of 3: ",
    sprintf("%.1f%%", 100 * mean(abs(codon_ratio$nt_per_codon - 3) < 0.1)),
    "\n\n", sep = "")

cat("external CDS nt / (column 2 + 10), the trim added back:\n")
print(round(quantile(codon_ratio$nt_per_codon_trim_restored,
                     c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99)), 4))
cat("within 0.01 of 3: ",
    sprintf("%.1f%%", 100 * mean(abs(codon_ratio$nt_per_codon_trim_restored - 3) < 0.01)),
    "\n\n", sep = "")

# Where the ratio is not near 3 at all, it is never a transcript the rule fits.
cat("Ratios below 3: ", sum(codon_ratio$nt_per_codon < 3),
    " of ", format(nrow(codon_ratio), big.mark = ","),
    ".  Of those, the number where the CDS length rule already fails: ",
    sum(codon_ratio$nt_per_codon < 3 & codon_ratio$delta != 0), "\n\n", sep = "")

print(as.data.frame(
  codon_ratio %>%
    group_by(match_class) %>%
    summarise(n = n(),
              median_as_reported   = round(median(nt_per_codon), 3),
              median_trim_restored = round(median(nt_per_codon_trim_restored), 4),
              .groups = "drop") %>%
    filter(n > 1)
))

stopifnot(
  "a transcript the CDS length rule fits has fewer than 3 nucleotides per codon" =
    all(codon_ratio$nt_per_codon[codon_ratio$delta == 0] >= 3),
  "every sub-3 ratio should be a transcript where the rule already fails" =
    all(codon_ratio$delta[codon_ratio$nt_per_codon < 3] != 0),
  "fewer than 98% of ratios reach 3" =
    mean(codon_ratio$nt_per_codon >= 3) > 0.98,
  "the as-reported ratio does not sit just above 3" =
    between(median(codon_ratio$nt_per_codon), 3, 3.3),
  "adding the 10-codon trim back does not give exactly 3" =
    median(codon_ratio$nt_per_codon_trim_restored) == 3,
  "the trim-restored ratio is not concentrated on 3" =
    mean(abs(codon_ratio$nt_per_codon_trim_restored - 3) < 0.01) > 0.85
)

ratio_long <- codon_ratio %>%
  transmute(`as reported: nt / column 2`            = nt_per_codon,
            `trim added back: nt / (column 2 + 10)` = nt_per_codon_trim_restored) %>%
  pivot_longer(everything(), names_to = "variant", values_to = "ratio") %>%
  mutate(variant = factor(variant,
                          levels = c("as reported: nt / column 2",
                                     "trim added back: nt / (column 2 + 10)")))

# A kernel density would be fit across the full range, which runs past 11, and would
# come out flat inside the window worth looking at. Fixed bins on the window instead.
ratio_window <- ratio_long %>% filter(between(ratio, 2.8, 4.1))
n_outside <- ratio_long %>% filter(!between(ratio, 2.8, 4.1)) %>% nrow()

p_ratio <- ggplot(ratio_window, aes(x = ratio, fill = variant, colour = variant)) +
  geom_histogram(binwidth = 0.02, alpha = 0.55, linewidth = 0.2) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ variant, scales = "free_y") +
  scale_colour_manual(values = c("#0072B2", "#CC79A7"), guide = "none") +
  scale_fill_manual(values = c("#0072B2", "#CC79A7"), guide = "none") +
  labs(
    title    = "Nucleotides per codon",
    subtitle = str_wrap(paste0(
      "n = ", format(nrow(codon_ratio), big.mark = ","),
      " transcripts, of which ", format(n_outside, big.mark = ","),
      " fall outside the 2.8 to 4.1 window shown. Dashed line at 3. Bins of 0.02. ",
      "The as-reported ratio clears 3 and trails right, because a constant 10-codon ",
      "trim costs a short CDS proportionally more."), width = 88),
    x        = "external CDS nucleotides per codon",
    y        = "Transcripts"
  ) +
  theme_classic(base_size = 12)

print(p_ratio)
ggsave(here("plots", "dhx29_01_nt_per_codon.pdf"), p_ratio, width = 8, height = 4.5)
```

So the ratio is three, and neither departure from it is noise. The upward drift is the trim,
re-derived here from a direction the rule test never used. The handful of ratios that fall
below three are, without exception, transcripts whose CDS was reannotated between the paper's
GENCODE and v49, the same population that fails the rule in 5d.


#### Plot 6: codons and nucleotides, side by side

The same comparison as a picture in the style of Section 4: column 2 on the left, the CDS
length pulled from GENCODE v49 on the right, in its own native nucleotides, on a shared axis.
Restricted to transcripts that carry both values, so the two panels describe the same genes.
The units differ on purpose: the offset between the curves is what the next plot measures.

```{r length_column_vs_external, fig.width = 9, fig.height = 4.5}
len_pair <- annot_check %>%
  filter(!is.na(cds_codons), cds_codons >= 0, !is.na(cds_len)) %>%
  left_join(rule %>% dplyr::select(transcript_id, delta), by = "transcript_id") %>%
  mutate(rule_holds = if_else(!is.na(delta) & delta == 0,
                              "CDS length rule holds", "rule fails"))

len_compare <- len_pair %>%
  transmute(rule_holds,
            `column 2 (codons)`     = cds_codons,
            `v49 CDS (nucleotides)` = cds_len) %>%
  pivot_longer(-rule_holds, names_to = "source", values_to = "value") %>%
  mutate(source = factor(source, levels = c("column 2 (codons)",
                                            "v49 CDS (nucleotides)")))

len_medians <- len_compare %>%
  group_by(source) %>%
  summarise(med = median(value), n = n(), .groups = "drop") %>%
  mutate(unit = if_else(source == "column 2 (codons)", "codons", "nucleotides"),
         lab  = paste0("median = ", round(med, 1), " ", unit))

cat("Shared transcripts: ", format(len_medians$n[1], big.mark = ","), "\n\n", sep = "")
print(as.data.frame(
  len_compare %>%
    group_by(source) %>%
    summarise(q25 = round(quantile(value, 0.25), 1),
              median = round(median(value), 1),
              q75 = round(quantile(value, 0.75), 1),
              mean = round(mean(value), 1), .groups = "drop")
))

# The pooled gap between the panel medians is wider than the trim. Split it: the trim
# accounts for the gap wherever the rule holds, and reannotated transcripts do the rest.
gap_split <- len_pair %>%
  group_by(rule_holds) %>%
  summarise(n                  = n(),
            median_col2_nt     = median(cds_codons) * 3,
            median_v49_nt      = median(cds_len),
            .groups = "drop") %>%
  mutate(median_gap_nt = median_v49_nt - median_col2_nt)

cat("\nOn a common footing, column 2 times 3 against the v49 CDS in nucleotides:\n")
print(as.data.frame(gap_split))

stopifnot(
  "where the CDS length rule holds, the median gap is not the 30-nucleotide trim" =
    between(gap_split$median_gap_nt[gap_split$rule_holds == "CDS length rule holds"], 27, 33)
)

p_len_pair <- ggplot(len_compare, aes(x = log2(value + 1))) +
  geom_density(alpha = 0.15, linewidth = 0.8, colour = "grey20", fill = "grey70") +
  geom_vline(data = len_medians, aes(xintercept = log2(med + 1)),
             linetype = "dashed", colour = "#0072B2") +
  geom_text(data = len_medians, aes(x = log2(med + 1), y = Inf, label = lab),
            hjust = -0.08, vjust = 2, size = 3.2, colour = "#0072B2") +
  facet_wrap(~ source) +
  labs(
    title    = "The length column against the CDS length pulled from GENCODE v49",
    subtitle = str_wrap(paste0(
      "n = ", format(len_medians$n[1], big.mark = ","),
      " transcripts carrying both values, same genes in both panels. Left is codons read ",
      "straight from column 2, right is the annotated CDS in nucleotides. The curves sit ",
      "log2(3) = 1.58 apart because of the units, not because the transcripts differ."),
      width = 100),
    x        = "log2(value + 1)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12)

print(p_len_pair)
ggsave(here("plots", "dhx29_01_length_column_vs_external.pdf"),
       p_len_pair, width = 9, height = 4.5)
```


#### Plot 7: codons times three against annotated nucleotides

If column 2 is a codon count over a trimmed window, then multiplying it by three should
recover the annotated CDS in nucleotides, short by a constant. Fitting that line puts a
number on the constant without ever using the rule from 5d.

```{r codons_x3_vs_nt, fig.width = 8, fig.height = 5}
nt_corr <- len_pair %>%
  mutate(nt_from_col2 = cds_codons * 3)

fit_all  <- lm(cds_len ~ nt_from_col2, data = nt_corr)
nt_ok    <- nt_corr %>% filter(rule_holds == "CDS length rule holds")
fit_ok   <- lm(cds_len ~ nt_from_col2, data = nt_ok)

corr_summary <- tibble(
  subset    = c("all shared transcripts", "CDS length rule holds"),
  n         = c(nrow(nt_corr), nrow(nt_ok)),
  pearson_r = c(cor(nt_corr$nt_from_col2, nt_corr$cds_len),
                cor(nt_ok$nt_from_col2, nt_ok$cds_len)),
  spearman  = c(cor(nt_corr$nt_from_col2, nt_corr$cds_len, method = "spearman"),
                cor(nt_ok$nt_from_col2, nt_ok$cds_len, method = "spearman")),
  slope     = c(coef(fit_all)[2], coef(fit_ok)[2]),
  intercept = c(coef(fit_all)[1], coef(fit_ok)[1])
)

print(as.data.frame(corr_summary %>%
                      mutate(across(pearson_r:slope, ~ round(.x, 5)),
                             intercept = round(intercept, 2))))

stopifnot(
  "slope of annotated nucleotides on column 2 times 3 is not 1" =
    abs(coef(fit_ok)[2] - 1) < 0.001,
  "intercept is not the 30-nucleotide trim" =
    abs(coef(fit_ok)[1] - 30) < 1,
  "column 2 times 3 does not track annotated nucleotides" =
    cor(nt_ok$nt_from_col2, nt_ok$cds_len) > 0.9999,
  "rank correlation across all shared transcripts is below 0.95" =
    cor(nt_corr$nt_from_col2, nt_corr$cds_len, method = "spearman") > 0.95
)

win <- 6000
n_out <- sum(nt_corr$nt_from_col2 > win | nt_corr$cds_len > win)

corr_lab <- paste0(
  "n = ", format(nrow(nt_corr), big.mark = ","),
  "\nPearson R = ", sprintf("%.4f", corr_summary$pearson_r[1]),
  "\nSpearman = ", sprintf("%.4f", corr_summary$spearman[1]),
  "\nwhere the rule holds (n = ", format(nrow(nt_ok), big.mark = ","), "):",
  "\n   slope = ", sprintf("%.4f", corr_summary$slope[2]),
  ", intercept = ", sprintf("%.1f", corr_summary$intercept[2]), " nt")

p_nt_corr <- ggplot(nt_corr, aes(x = nt_from_col2, y = cds_len, colour = rule_holds)) +
  geom_point(size = 0.4, alpha = 0.25) +
  geom_abline(slope = 1, intercept = 30, linetype = "dashed",
              colour = "grey15", linewidth = 0.4) +
  annotate("text", x = -Inf, y = Inf, label = corr_lab,
           hjust = -0.05, vjust = 1.1, size = 3) +
  coord_cartesian(xlim = c(0, win), ylim = c(0, win)) +
  scale_colour_manual(values = c("CDS length rule holds" = "#0072B2",
                                 "rule fails" = "#CC79A7"),
                      name = NULL) +
  labs(
    title    = "Column 2 times three against the annotated CDS in nucleotides",
    subtitle = str_wrap(paste0(
      "Dashed line is y = x + 30, the 10-codon trim expressed in nucleotides. ",
      "Axes clipped at ", format(win, big.mark = ","), " nt; ",
      format(n_out, big.mark = ","), " transcripts fall outside and are not drawn, ",
      "though every statistic above uses all of them."), width = 92),
    x        = "column 2 times 3 (nucleotides)",
    y        = "annotated v49 CDS (nucleotides)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_nt_corr)
ggsave(here("plots", "dhx29_01_codons_x3_vs_nucleotides.pdf"),
       p_nt_corr, width = 8, height = 5)
```

Slope one, intercept thirty nucleotides. The trim falls out of the regression as an
intercept, independently of how it was derived in 5d.


### 5e: Plot 8, the CDS length this project computes elsewhere

The feature matrices carry a `cds_length` built by a different code path in
`code/predictive_modeling/01_feature_extraction.Rmd`, on the most abundant transcript per
gene rather than on the full annotation. It is the length these counts would eventually have
to line up with, so it is worth checking directly.

Two caveats on what this does and does not show. It is an independent **code path**, not an
independent **annotation**: both routes trace back to GENCODE v49, and the printout below
confirms they are identical on every shared transcript. So this validates the extraction,
not the annotation. And the feature matrix keeps one transcript per gene, chosen for
abundance, so the shared set is skewed toward stable, well-expressed annotations. That is
why agreement here runs higher than the blended figure in 5d.

```{r feature_matrix_cds}
fm_path <- here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
stopifnot("feature matrix not found; knit the 01x_ chain first" = file.exists(fm_path))

fm_cds <- readRDS(fm_path) %>%
  dplyr::select(transcript_id_clean, cds_length) %>%
  filter(!is.na(cds_length)) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

cat("Feature matrix transcripts with a CDS length:",
    format(nrow(fm_cds), big.mark = ","), "\n")

fm_join <- annot_check %>%
  filter(!is.na(cds_codons)) %>%
  inner_join(fm_cds, by = "transcript_id_clean") %>%
  mutate(
    cds_codons_fm = round(cds_length / 3) - 10,
    delta_fm      = cds_codons - cds_codons_fm
  )

cat("Transcripts shared with the feature matrix:",
    format(nrow(fm_join), big.mark = ","), "\n\n")

fm_overall <- fm_join %>%
  summarise(
    n         = n(),
    agreement = mean(delta_fm == 0),
    pearson_r = cor(cds_codons, cds_codons_fm),
    spearman  = cor(cds_codons, cds_codons_fm, method = "spearman")
  )

cat("GEO column 2 vs feature matrix cds_length, all shared transcripts:\n")
print(as.data.frame(fm_overall %>% mutate(across(agreement:spearman, ~ round(.x, 4)))))

cat("\nSame comparison split by transcript version match against v49:\n")
print(as.data.frame(
  fm_join %>%
    group_by(match_class) %>%
    summarise(n = n(), agreement = round(mean(delta_fm == 0), 4),
              pearson_r = round(cor(cds_codons, cds_codons_fm), 4),
              .groups = "drop") %>%
    filter(n > 1)
))

# The two independent sources against each other, on the transcripts both cover.
both_sources <- fm_join %>% filter(!is.na(cds_len))
cat("\nThe two independent sources against each other (txdb vs feature matrix), n = ",
    format(nrow(both_sources), big.mark = ","), ":  Pearson R = ",
    sprintf("%.4f", cor(both_sources$cds_len, both_sources$cds_length)),
    ", identical in ",
    sprintf("%.2f%%", 100 * mean(both_sources$cds_len == both_sources$cds_length)),
    " of cases\n", sep = "")

stopifnot(
  "correlation with the feature matrix CDS length is below 0.98" =
    fm_overall$pearson_r > 0.98,
  "under 95% exact agreement with the feature matrix CDS length" =
    fm_overall$agreement > 0.95,
  "the two independent CDS length sources disagree with each other" =
    cor(both_sources$cds_len, both_sources$cds_length) > 0.99
)
```

```{r plot_length_vs_feature_matrix, fig.width = 8, fig.height = 4.5}
fm_lab <- paste0("n = ", format(fm_overall$n, big.mark = ","),
                 "\nPearson R = ", sprintf("%.4f", fm_overall$pearson_r),
                 "\nSpearman = ", sprintf("%.4f", fm_overall$spearman),
                 "\nexact match = ", sprintf("%.2f%%", 100 * fm_overall$agreement))

p_fm <- fm_join %>%
  filter(cds_codons >= 0, cds_codons_fm >= 0) %>%
  ggplot(aes(x = log2(cds_codons_fm + 1), y = log2(cds_codons + 1),
             colour = match_class)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.5, alpha = 0.35) +
  annotate("text", x = -Inf, y = Inf, label = fm_lab,
           hjust = -0.1, vjust = 1.15, size = 3) +
  scale_colour_manual(values = c("exact" = "#0072B2", "drift" = "#CC79A7",
                                 "absent" = "grey60"),
                      name = "v49 version match") +
  labs(
    title    = "GEO length column vs the CDS length this project computes elsewhere",
    subtitle = str_wrap(paste0(
      "Feature matrix cds_length from 01_feature_extraction.Rmd, converted with ",
      "round(CDS nt / 3) - 10. Dashed line is y = x."), width = 78),
    x        = "log2(expected codons from the feature matrix + 1)",
    y        = "log2(GEO column 2 + 1)"
  ) +
  theme_classic(base_size = 12)

print(p_fm)
ggsave(here("plots", "dhx29_01_cds_length_vs_feature_matrix.pdf"),
       p_fm, width = 8, height = 4.5)
```

### 5f: Plot 9, observed against expected CDS length

```{r plot_length_vs_v49, fig.width = 7.5, fig.height = 4}
rule_labels <- rule_summary %>%
  mutate(lab = paste0("n = ", format(n, big.mark = ","),
                      "\nR = ", sprintf("%.4f", pearson_r),
                      "\nexact match = ", sprintf("%.2f%%", 100 * agreement)))

p_scatter <- rule %>%
  filter(cds_codons >= 0, cds_codons_v49 >= 0) %>%
  ggplot(aes(x = log2(cds_codons_v49 + 1), y = log2(cds_codons + 1))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "#0072B2") +
  geom_point(size = 0.3, alpha = 0.12, colour = "grey20") +
  geom_text(data = rule_labels, aes(x = -Inf, y = Inf, label = lab),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ match_class) +
  labs(
    title    = "GEO length column vs CDS length from GENCODE v49",
    subtitle = "Expected = round(CDS nt / 3) - 10. Dashed line is y = x.",
    x        = "log2(expected codons from v49 + 1)",
    y        = "log2(GEO column 2 + 1)"
  ) +
  theme_classic(base_size = 12)

print(p_scatter)
ggsave(here("plots", "dhx29_01_cds_length_vs_v49.pdf"),
       p_scatter, width = 7.5, height = 4)
```

### 5g: Plot 10, the residual

`delta` is exactly zero for the overwhelming majority of the exact-version class and has a
long integer tail, so a density curve would smooth the one thing worth seeing. Binned
proportions instead.

```{r plot_length_delta, fig.width = 8, fig.height = 4.5}
delta_buckets <- rule %>%
  mutate(bucket = cut(abs(delta), breaks = c(-1, 0, 9, 99, Inf),
                      labels = c("0", "1 to 9", "10 to 99", "100 or more"))) %>%
  dplyr::count(match_class, bucket) %>%
  group_by(match_class) %>%
  mutate(frac = n / sum(n), n_class = sum(n)) %>%
  ungroup() %>%
  arrange(match_class) %>%
  mutate(class_lab = factor(paste0(match_class, " (n = ",
                                   format(n_class, big.mark = ","), ")"),
                            levels = unique(paste0(match_class, " (n = ",
                                                   format(n_class, big.mark = ","), ")"))))

print(as.data.frame(delta_buckets %>% dplyr::select(match_class, bucket, n) %>%
                      pivot_wider(names_from = bucket, values_from = n, values_fill = 0)))

p_delta <- ggplot(delta_buckets, aes(x = bucket, y = frac, fill = class_lab)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7,
           colour = "grey20", linewidth = 0.3) +
  geom_text(aes(label = format(n, big.mark = ",")),
            position = position_dodge(width = 0.75),
            vjust = -0.4, size = 2.8) +
  scale_y_continuous(limits = c(0, 1.08), breaks = seq(0, 1, 0.25)) +
  scale_fill_manual(values = setNames(c("#0072B2", "#CC79A7"),
                                      levels(delta_buckets$class_lab)),
                    name = "Version match") +
  labs(
    title    = "Residual between the GEO length column and GENCODE v49",
    subtitle = str_wrap(paste0(
      "delta = column 2 minus round(CDS nt / 3) - 10, bucketed by absolute value. ",
      "Bar labels are transcript counts."), width = 90),
    x        = "absolute delta (codons)",
    y        = "Fraction of the class"
  ) +
  theme_classic(base_size = 12)

print(p_delta)
ggsave(here("plots", "dhx29_01_cds_length_delta.pdf"),
       p_delta, width = 8, height = 4.5)
```

### 5h: The disagreements

Printed rather than summarised away, so a systematic offset would be visible.

```{r worst_mismatches}
cat("Largest disagreements within the exact-version class:\n")
print(as.data.frame(
  rule %>%
    filter(match_class == "exact", delta != 0) %>%
    arrange(desc(abs(delta))) %>%
    dplyr::select(transcript_id, gene_id, v49_transcript_type,
                  cds_codons, cds_len, cds_codons_v49, delta) %>%
    head(15)
))

cat("\nDelta values within the exact-version class, most common first:\n")
print(as.data.frame(
  rule %>%
    filter(match_class == "exact") %>%
    dplyr::count(delta, sort = TRUE) %>%
    head(8)
))
```

The 159 exact-version disagreements are dominated by `nonsense_mediated_decay`
transcripts, where GENCODE moved the annotated CDS boundary without bumping the transcript
version. That is annotation churn on the v49 side, not a misread column.

### 5i: Assertions

```{r assertions}
rows_per_sample <- counts %>% dplyr::count(sample) %>% pull(n)

len_by_sample <- counts %>%
  dplyr::select(sample, transcript_id, cds_codons) %>%
  pivot_wider(names_from = sample, values_from = cds_codons)

exact_row <- rule_summary %>% filter(match_class == "exact")
drift_row <- rule_summary %>% filter(match_class == "drift")

stopifnot(
  "samples differ in row count" = n_distinct(rows_per_sample) == 1,
  "unexpected row count per sample" = rows_per_sample[1] == 227264,
  "counts are not finite non-negative integers" =
    all(is.finite(counts$count)) && all(counts$count >= 0),
  "the length column is not identical across all six samples" =
    all(len_by_sample$Input1 == len_by_sample$Input2 &
        len_by_sample$Input1 == len_by_sample$Input3 &
        len_by_sample$Input1 == len_by_sample$IP1 &
        len_by_sample$Input1 == len_by_sample$IP2 &
        len_by_sample$Input1 == len_by_sample$IP3, na.rm = TRUE),
  "N/A rows disagree between samples" =
    all(rowSums(is.na(len_by_sample[, -1])) %in% c(0L, 6L)),
  "a non-coding row carries counts" =
    all(counts$count[is.na(counts$cds_codons)] == 0),
  "a negative-length row carries counts" =
    all(counts$count[!is.na(counts$cds_codons) & counts$cds_codons < 0] == 0),
  "exact-version agreement with the CDS length rule is below 0.99" =
    exact_row$agreement > 0.99,
  "exact-version correlation with v49 CDS length is below 0.99" =
    exact_row$pearson_r > 0.99,
  "version drift did not degrade agreement, so the join is not doing what it claims" =
    drift_row$agreement < exact_row$agreement,
  "under 90% of coding GEO transcripts match a v49 transcript" =
    coding_match_rate > 0.90
)

cat("All Section 5 assertions passed.\n")
```

### 5j: Biological spot-checks

Canonical housekeeping transcripts must sit at the top of the count distribution in every
sample. One that does not would mean the wrong column was read.

```{r housekeeping}
hk <- tibble(
  symbol         = c("ACTB", "GAPDH", "EEF1A1"),
  gene_id_clean  = c("ENSG00000075624", "ENSG00000111640", "ENSG00000156508")
)

hk_check <- counts %>%
  left_join(annot_check %>%
              transmute(transcript_id, gene_id_clean = str_remove(gene_id, "\\..*")),
            by = "transcript_id") %>%
  filter(count > 0) %>%
  group_by(sample) %>%
  mutate(pct = percent_rank(count)) %>%
  ungroup() %>%
  inner_join(hk, by = "gene_id_clean") %>%
  group_by(sample, assay, symbol) %>%
  summarise(top_pct = max(pct), top_count = max(count), .groups = "drop")

print(as.data.frame(
  hk_check %>%
    mutate(top_pct = round(top_pct, 4)) %>%
    pivot_wider(id_cols = c(sample, assay), names_from = symbol,
                values_from = top_pct)
))

stopifnot(
  "a housekeeping transcript is not in the top 5% of counts in every sample" =
    all(hk_check$top_pct > 0.95),
  "housekeeping check did not cover all three genes in all six samples" =
    nrow(hk_check) == 18,
  "DHX29-IP libraries are not smaller than Input, unexpected for a pull-down" =
    ip_input_ratio < 1
)

cat("Housekeeping and library-size checks passed.\n")
```

---

## Section 6: Save

```{r save_outputs}
out_dir <- here("output", "dhx29_reanalysis")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(counts, file.path(out_dir, "dhx29_thor_riboseq_counts_long.rds"))
write_csv(sample_summary, file.path(out_dir, "dhx29_thor_riboseq_sample_summary.csv"))

annot_check %>%
  left_join(rule %>% dplyr::select(transcript_id, cds_codons_v49, delta),
            by = "transcript_id") %>%
  left_join(fm_join %>% dplyr::select(transcript_id, cds_length,
                                      cds_codons_fm, delta_fm),
            by = "transcript_id") %>%
  left_join(codon_ratio %>% dplyr::select(transcript_id, nt_per_codon,
                                          nt_per_codon_trim_restored),
            by = "transcript_id") %>%
  dplyr::select(transcript_id, transcript_id_clean, gene_id,
                cds_codons, cds_length_external = cds_len,
                nt_per_codon, nt_per_codon_trim_restored,
                cds_codons_v49, delta,
                cds_length_feature_matrix = cds_length,
                cds_codons_fm, delta_fm,
                match_class, v49_transcript_type, v49_cds_bearing) %>%
  write_csv(file.path(out_dir, "dhx29_cds_length_vs_v49.csv"))

dbDisconnect(con)

cat("Wrote:\n",
    " ", file.path(out_dir, "dhx29_thor_riboseq_counts_long.rds"), "\n",
    " ", file.path(out_dir, "dhx29_thor_riboseq_sample_summary.csv"), "\n",
    " ", file.path(out_dir, "dhx29_cds_length_vs_v49.csv"), "\n", sep = "")
```

```{r session_info}
sessionInfo()
```
