---
title: "DHX29 reanalysis 02: can transcript-level counts reproduce the paper's codon usage?"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Notebook 01 established that GSE266018 gives **per-transcript** footprint totals over a
trimmed CDS window, and nothing positional. The paper's headline codon result (Table S8) is
an **A-site** statistic: for each sample, the fraction of footprints whose A site sits on
each of the 64 codons. Those are different estimators, so this notebook asks a bounded
question rather than claiming a replication.

**The proxy.** Under a null of uniform footprint density along the counted window, the
expected share of footprints at codon `c` is

```
E(c)  =  sum over transcripts of  count_t * n_ct / L_t
```

where `n_ct` is how often codon `c` occurs in transcript `t`'s counted window and `L_t` is
that window's length in codons. Normalised across codons, `E(c)` is directly comparable to a
Table S8 column, because both are codon shares summing to one.

**What it can and cannot see.** The proxy carries only the *transcript-level* component: are
the mRNAs enriched in the DHX29 pull-down themselves biased toward codon `c`? It is blind to
the *positional* component, which is the paper's actual claim, that DHX29 associates with
ribosomes while they decode a nonoptimal codon. A ribosome redistributing along a fixed set
of mRNAs moves the paper's statistic and leaves the proxy untouched. So agreement is
evidence the mRNA-level bias is real; disagreement in magnitude is expected and refutes
nothing.

**Transcript level, not gene level, and the isoforms are not independent.** These files
carry one row per transcript, not per gene. Reads were assigned to every compatible isoform
rather than apportioned to one, so a gene's isoforms can repeat the same count verbatim.
Section 3 measures how often. Two consequences here: the proxy weights a multi-isoform gene
by the number of isoforms it happens to have annotated, and summing transcripts to a gene
total would over-count. Comparing IP against Input on the *same* transcript is unaffected,
because the same assignment rule produced both.

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)

assay_colors <- c("Input" = "#0072B2", "DHX29-IP" = "#CC79A7")
```

---

## Section 1: Inputs

```{r load_counts}
counts <- readRDS(here("output", "dhx29_reanalysis",
                       "dhx29_thor_riboseq_counts_long.rds"))
stopifnot("notebook 01 output looks wrong" = nrow(counts) == 227264 * 6)
cat("Counts loaded:", format(nrow(counts), big.mark = ","), "rows\n")
```

Table S8, with the replication check the repository convention asks for before external data
is used. The columns are shares, so each sample must sum to one, and `Fold_Change` must be
`IP_Avg / Input_Avg`.

```{r load_s8}
s8 <- read_csv(here("accessories", "csc_data", "hia_dhx29",
                    "science.adw0288_tables_s1_to_s11",
                    "science.adw0288_table_s8.csv"),
               show_col_types = FALSE)

sample_cols <- c("Input1", "Input2", "Input3", "IP1", "IP2", "IP3")

stopifnot(
  "Table S8 does not carry all 64 codons" = nrow(s8) == 64,
  "Table S8 sample columns do not sum to 1" =
    all(abs(colSums(s8[, sample_cols]) - 1) < 1e-6),
  "Table S8 Fold_Change is not IP_Avg / Input_Avg" =
    max(abs(s8$Fold_Change - s8$IP_Avg / s8$Input_Avg)) < 1e-6,
  "Table S8 averages do not match their replicates" =
    max(abs(s8$IP_Avg - rowMeans(s8[, c("IP1", "IP2", "IP3")]))) < 1e-6
)
cat("Table S8 checks passed. Fold_Change range:",
    sprintf("%.3f to %.3f", min(s8$Fold_Change), max(s8$Fold_Change)), "\n")
```

---

## Section 2: Codon composition of the counted window

The window is reconstructed to match the counts exactly: drop five codons from each end of
the CDS, stop codon included. Notebook 01 showed the reported length is `CDS nt / 3 - 10`, so
the reconstructed window length must equal it. That equality is the check that the
composition is being computed over the same nucleotides the footprints were counted on.

```{r codon_composition}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds_seqs <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                  cdsBy(txdb, by = "tx", use.names = TRUE))

usable  <- (width(cds_seqs) %% 3 == 0) & (width(cds_seqs) >= 33)
cds_seqs <- cds_seqs[usable]
window   <- subseq(cds_seqs, start = 16, end = width(cds_seqs) - 15)

cat("CDS sequences with a usable window:", format(length(window), big.mark = ","), "\n")

codon_freq <- trinucleotideFrequency(window, step = 3)
rownames(codon_freq) <- names(window)     # Biostrings drops these
window_len <- rowSums(codon_freq)

stopifnot(
  "reconstructed window is not CDS nt / 3 - 10" =
    all(window_len == width(cds_seqs) / 3 - 10),
  "codon matrix lost its transcript names" =
    all(grepl("^ENST", rownames(codon_freq))),
  "codon matrix is not 64 wide" = ncol(codon_freq) == 64
)

# Biological spot-check: ATG must appear at least once in essentially every window only
# if the start codon survived the trim, which it does not. Use a composition invariant
# instead: the four codon families of a random window must all be represented in bulk.
cat("Median codons per window:", median(window_len), "\n")
cat("Share of windows containing at least one CTG (most used Leu codon):",
    sprintf("%.3f", mean(codon_freq[, "CTG"] > 0)), "\n")
stopifnot("CTG is absent from most windows, composition is wrong" =
            mean(codon_freq[, "CTG"] > 0) > 0.8)
```

---

## Section 3: The proxy

```{r build_proxy}
row_key <- tibble(tx_clean = str_remove(rownames(codon_freq), "\\..*"),
                  row      = seq_len(nrow(codon_freq)))

proxy_long <- map_dfr(levels(counts$sample), function(s) {
  cs <- counts %>%
    filter(sample == s, count > 0, !is.na(cds_codons)) %>%
    transmute(tx_clean = transcript_id_clean, count)
  m <- inner_join(row_key, cs, by = "tx_clean")
  w <- m$count / window_len[m$row]
  e <- as.vector(crossprod(codon_freq[m$row, , drop = FALSE], w))
  tibble(sample = s, codon = colnames(codon_freq),
         expected = e / sum(e), n_tx = nrow(m))
})

cat("Transcripts contributing per sample:\n")
print(as.data.frame(proxy_long %>% distinct(sample, n_tx)))

stopifnot(
  "proxy shares do not sum to 1 within a sample" =
    all(abs((proxy_long %>% group_by(sample) %>%
               summarise(s = sum(expected), .groups = "drop"))$s - 1) < 1e-9)
)

proxy <- proxy_long %>%
  dplyr::select(-n_tx) %>%
  pivot_wider(names_from = sample, values_from = expected)

cmp <- inner_join(proxy, s8, by = "codon", suffix = c("_proxy", "_paper"))
stopifnot("codons did not join" = nrow(cmp) == 64)
```


### How much isoform duplication is in the weighting?

```{r isoform_duplication}
gene_map <- read_csv(here("output", "dhx29_reanalysis", "dhx29_cds_length_vs_v49.csv"),
                     show_col_types = FALSE) %>%
  dplyr::select(transcript_id, gene_id) %>%
  filter(!is.na(gene_id))

dup_check <- counts %>%
  filter(sample == "Input1", count > 0) %>%
  inner_join(gene_map, by = "transcript_id") %>%
  group_by(gene_id) %>%
  filter(n() > 1) %>%
  summarise(n_tx = n(), n_distinct_counts = n_distinct(count), .groups = "drop")

cat("Multi-transcript genes carrying counts:",
    format(nrow(dup_check), big.mark = ","), "\n")
cat("Genes where every isoform repeats one identical count: ",
    format(sum(dup_check$n_distinct_counts == 1), big.mark = ","),
    sprintf(" (%.1f%%)", 100 * mean(dup_check$n_distinct_counts == 1)),
    ", median ", median(dup_check$n_tx[dup_check$n_distinct_counts == 1]),
    " isoforms each\n", sep = "")

stopifnot(
  "no isoform duplication found, which contradicts notebook 01" =
    mean(dup_check$n_distinct_counts == 1) > 0.05,
  "isoform duplication is so pervasive that transcript rows carry no independent signal" =
    mean(dup_check$n_distinct_counts == 1) < 0.5
)
```

Roughly a fifth of multi-isoform genes repeat a single count across all their isoforms, so
those genes enter the proxy with weight multiplied by their annotated isoform count. This
inflates the contribution of heavily annotated genes. It affects IP and Input identically,
so it largely cancels in the fold change, which is the quantity Section 5 compares.

---

## Section 4: Per-sample agreement

```{r per_sample}
per_sample <- map_dfr(sample_cols, function(s) {
  a <- cmp[[paste0(s, "_proxy")]]; b <- cmp[[paste0(s, "_paper")]]
  tibble(sample = s,
         assay    = if_else(str_starts(s, "Input"), "Input", "DHX29-IP"),
         pearson  = cor(a, b),
         spearman = cor(a, b, method = "spearman"))
})
print(as.data.frame(per_sample %>% mutate(across(pearson:spearman, ~ round(.x, 3)))))
```

---

## Section 5: Fold change, the quantity the paper reports

```{r fold_change, fig.width = 7.5, fig.height = 5}
cmp <- cmp %>%
  mutate(
    proxy_fc = rowMeans(across(all_of(paste0(c("IP1", "IP2", "IP3"), "_proxy")))) /
               rowMeans(across(all_of(paste0(c("Input1", "Input2", "Input3"), "_proxy")))),
    paper_fc = Fold_Change,
    wobble   = str_sub(codon, 3, 3),
    is_stop  = codon %in% c("TAA", "TAG", "TGA")
  )

sense <- cmp %>% filter(!is_stop)

fc_test <- cor.test(sense$proxy_fc, sense$paper_fc, method = "spearman",
                    exact = FALSE)

cat("Sense codons compared:", nrow(sense), "\n")
cat(sprintf("Fold change agreement: Pearson %.3f, Spearman %.3f (p = %.2g)\n",
            cor(sense$proxy_fc, sense$paper_fc), fc_test$estimate, fc_test$p.value))
cat(sprintf("Amplitude: proxy spans %.3f to %.3f, paper spans %.3f to %.3f\n",
            min(sense$proxy_fc), max(sense$proxy_fc),
            min(sense$paper_fc), max(sense$paper_fc)))
cat(sprintf("The proxy recovers about %.0f%% of the paper's dynamic range.\n",
            100 * (max(sense$proxy_fc) - min(sense$proxy_fc)) /
                  (max(sense$paper_fc) - min(sense$paper_fc))))

stopifnot(
  "the proxy does not track the paper's codon fold change at all" =
    fc_test$estimate > 0.4,
  "the proxy should not reproduce the paper's amplitude" =
    (max(sense$proxy_fc) - min(sense$proxy_fc)) <
    (max(sense$paper_fc) - min(sense$paper_fc))
)

p_fc <- ggplot(sense, aes(x = paper_fc, y = proxy_fc, colour = wobble)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey60") +
  geom_vline(xintercept = 1, linetype = "dotted", colour = "grey60") +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey30",
              linewidth = 0.5, linetype = "dashed") +
  scale_colour_manual(values = c(A = "#009E73", C = "#0072B2",
                                 G = "#D55E00", `T` = "#CC79A7"),
                      name = "3rd base") +
  labs(
    title    = "Transcript-level proxy against the paper's A-site codon enrichment",
    subtitle = str_wrap(paste0(
      "n = ", nrow(sense), " sense codons. Spearman ",
      sprintf("%.3f", fc_test$estimate), ", p = ",
      sprintf("%.2g", fc_test$p.value),
      ". Note the axis scales: the proxy moves within a few percent while the paper's ",
      "statistic moves by tens of percent."), width = 92),
    x        = "paper Fold_Change (IP / Input, A-site)",
    y        = "proxy fold change (composition weighted)"
  ) +
  theme_classic(base_size = 12)

print(p_fc)
ggsave(here("plots", "dhx29_02_proxy_vs_paper_codon_fc.pdf"),
       p_fc, width = 7.5, height = 5)
```

---

## Section 6: Does the wobble split go the same way?

The paper's claim is enrichment on nonoptimal codons. In human those skew A- and T-ending.
If the mRNA-level component of that is real, the proxy should split the same way even though
it is much smaller.

```{r wobble, fig.width = 7.5, fig.height = 4.5}
wobble_summary <- sense %>%
  mutate(ending = if_else(wobble %in% c("A", "T"), "A or T ending",
                          "G or C ending")) %>%
  group_by(ending) %>%
  summarise(n = n(),
            proxy_median = median(proxy_fc),
            paper_median = median(paper_fc), .groups = "drop")
print(as.data.frame(wobble_summary %>% mutate(across(where(is.numeric), ~ round(.x, 4)))))

at_vs_gc <- sense %>% mutate(at = wobble %in% c("A", "T"))
w_proxy <- wilcox.test(proxy_fc ~ at, data = at_vs_gc)
w_paper <- wilcox.test(paper_fc ~ at, data = at_vs_gc)
cat(sprintf("\nA/T vs G/C ending, Wilcoxon: proxy p = %.3g, paper p = %.3g\n",
            w_proxy$p.value, w_paper$p.value))

same_direction <-
  sign(diff(wobble_summary$proxy_median)) == sign(diff(wobble_summary$paper_median))
cat("Proxy and paper split in the same direction:", same_direction, "\n")

p_wob <- sense %>%
  dplyr::select(codon, wobble, proxy_fc, paper_fc) %>%
  pivot_longer(c(proxy_fc, paper_fc), names_to = "source", values_to = "fc") %>%
  mutate(source = factor(source, levels = c("paper_fc", "proxy_fc"),
                         labels = c("paper, A-site", "proxy, composition")),
         ending = if_else(wobble %in% c("A", "T"), "A or T", "G or C")) %>%
  ggplot(aes(x = ending, y = fc, fill = ending)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey60") +
  geom_boxplot(alpha = 0.35, outlier.shape = NA, width = 0.55) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.7) +
  facet_wrap(~ source, scales = "free_y") +
  scale_fill_manual(values = c("A or T" = "#009E73", "G or C" = "#0072B2"),
                    guide = "none") +
  labs(
    title    = "Third-base split, both estimators",
    subtitle = str_wrap(paste0(
      "n = ", nrow(sense), " sense codons. Free y axes: the two estimators differ ",
      "by an order of magnitude in amplitude, so a shared axis would flatten the ",
      "proxy panel."), width = 92),
    x        = "third codon position",
    y        = "IP / Input fold change"
  ) +
  theme_classic(base_size = 12)

print(p_wob)
ggsave(here("plots", "dhx29_02_wobble_split_both_estimators.pdf"),
       p_wob, width = 7.5, height = 4.5)
```

---

## Section 7: Verdict

```{r verdict}
write_csv(sense %>%
            dplyr::select(codon, wobble, proxy_fc, paper_fc,
                          Input_Avg, IP_Avg),
          here("output", "dhx29_reanalysis", "dhx29_02_proxy_vs_paper_codon_fc.csv"))
cat("Wrote output/dhx29_reanalysis/dhx29_02_proxy_vs_paper_codon_fc.csv\n")
```

The direction replicates and the magnitude does not, which is the expected result rather
than a disappointing one. The rank ordering of codons agrees well above chance, so the
mRNAs that DHX29 pulls down really are biased in codon content in the direction the paper
reports. But the proxy moves within a few percent where the paper's statistic moves by tens
of percent, so the great majority of the paper's signal is positional and lives in where the
ribosome sits, not in which mRNA it is on. Per-transcript totals cannot reach it.

Practically: GSE266018 supports per-transcript DHX29 enrichment and a weak codon-composition
readout. Any codon-resolution claim has to come from Table S8, which
`code/predictive_modeling/23_dhx29_riboseq_slamseq.Rmd` already validates, or from
reprocessing the raw reads with A-site assignment.
