---
title: "DHX29 reanalysis 04: IP over Input enrichment, transcript and gene level"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Measured DHX29 enrichment per transcript and per gene, from the counts prepared in notebooks
01 to 03. This is the quantity GSE266018 supports and the paper's supplementary tables do
not provide.

DESeq2 is the statistic, with a plain normalised log2 ratio alongside as a sanity column. A
raw ratio on its own is not usable here: the IP libraries are 52% the size of Input, most
features carry single-digit counts even at a floor of 3, and notebook 03 found the IP
replicates agree with each other less than they agree with Input.

That last point sets expectations. Read the p-value histogram before anything else. A flat
histogram would say there is no detectable per-feature signal, which is an answer, not a
failure.

**Sign convention: Input is the reference, so a positive log2 fold change means enriched in
the DHX29 pull-down.** Asserted below.

```{r load_libraries}
library(tidyverse)
library(here)
library(DESeq2)
library(apeglm)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
```

---

## Section 1: Load

```{r load_inputs}
sample_cols <- c("Input1", "Input2", "Input3", "IP1", "IP2", "IP3")
cutoff <- 3

gene_counts <- readRDS(here("output", "dhx29_reanalysis",
                            "dhx29_gene_level_counts.rds"))

tx_counts <- readRDS(here("output", "dhx29_reanalysis",
                          "dhx29_counts_with_gene.rds")) %>%
  dplyr::select(transcript_id, transcript_id_clean, gene_id_clean, symbol,
                is_mane, cds_codons, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count) %>%
  filter(!is.na(gene_id_clean),
         pmin(Input1, Input2, Input3, IP1, IP2, IP3) >= cutoff)

cat("Transcripts clearing the floor:", format(nrow(tx_counts), big.mark = ","), "\n")
cat("Genes after collapse:", format(nrow(gene_counts), big.mark = ","), "\n")

col_data <- tibble(sample = sample_cols) %>%
  mutate(assay = factor(if_else(str_starts(sample, "Input"), "Input", "IP"),
                        levels = c("Input", "IP")),
         rep   = factor(str_extract(sample, "\\d$"))) %>%
  column_to_rownames("sample")

stopifnot(
  "feature counts disagree with notebook 03" =
    nrow(tx_counts) == 41679 && nrow(gene_counts) == 10493,
  "Input is not the reference level" = levels(col_data$assay)[1] == "Input",
  "counts are not clean non-negative integers" =
    all(map_lgl(tx_counts[sample_cols], ~ all(is.finite(.x)) && all(.x >= 0))) &&
    all(map_lgl(gene_counts[sample_cols], ~ all(is.finite(.x)) && all(.x >= 0)))
)
print(col_data)
```

---

## Section 2: DESeq2, transcript level

Both designs fitted. `lfcShrink` with apeglm on the assay coefficient.

```{r deseq_transcript}
mat_tx <- as.matrix(tx_counts[, sample_cols])
rownames(mat_tx) <- tx_counts$transcript_id

dds_tx_paired <- DESeqDataSetFromMatrix(mat_tx, col_data, ~ rep + assay) %>% DESeq(quiet = TRUE)
dds_tx_simple <- DESeqDataSetFromMatrix(mat_tx, col_data, ~ assay)       %>% DESeq(quiet = TRUE)

cat("Size factors:\n"); print(round(sizeFactors(dds_tx_paired), 4))

raw_tx_paired <- results(dds_tx_paired, name = "assay_IP_vs_Input")
raw_tx_simple <- results(dds_tx_simple, name = "assay_IP_vs_Input")
res_tx_paired <- lfcShrink(dds_tx_paired, coef = "assay_IP_vs_Input",
                           type = "apeglm", quiet = TRUE)
res_tx_simple <- lfcShrink(dds_tx_simple, coef = "assay_IP_vs_Input",
                           type = "apeglm", quiet = TRUE)

# Companion column: plain log2 ratio of normalised means, same size factors.
norm_tx <- counts(dds_tx_paired, normalized = TRUE)
ratio_tx <- log2((rowMeans(norm_tx[, 4:6]) + 1) / (rowMeans(norm_tx[, 1:3]) + 1))

tx_res <- tx_counts %>%
  dplyr::select(transcript_id, transcript_id_clean, gene_id_clean, symbol,
                is_mane, cds_codons) %>%
  mutate(base_mean       = res_tx_paired$baseMean,
         lfc_paired      = res_tx_paired$log2FoldChange,
         padj_paired     = res_tx_paired$padj,
         pvalue_paired   = raw_tx_paired$pvalue,
         lfc_raw_paired  = raw_tx_paired$log2FoldChange,
         lfc_simple      = res_tx_simple$log2FoldChange,
         padj_simple     = res_tx_simple$padj,
         pvalue_simple   = raw_tx_simple$pvalue,
         lfc_raw_simple  = raw_tx_simple$log2FoldChange,
         log2_ratio      = ratio_tx)

cat("\nSignificant transcripts, FDR 0.10:  paired",
    sum(tx_res$padj_paired < 0.10, na.rm = TRUE),
    " unpaired", sum(tx_res$padj_simple < 0.10, na.rm = TRUE),
    " of", format(nrow(tx_res), big.mark = ","), "\n")

# Is the p-value distribution enriched near zero, or depleted?
pb <- hist(tx_res$pvalue_paired, breaks = seq(0, 1, 0.02), plot = FALSE)$counts
cat("First p-value bin, paired: observed", pb[1],
    "against", round(sum(!is.na(tx_res$pvalue_paired)) * 0.02),
    "expected if uniform\n")
```

```{r plot_diagnostics, fig.width = 9, fig.height = 7}
diag_df <- tx_res %>%
  filter(!is.na(padj_paired)) %>%
  mutate(sig = padj_paired < 0.1)

p_ma <- ggplot(diag_df, aes(x = base_mean, y = lfc_paired, colour = sig)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(size = 0.4, alpha = 0.3) +
  scale_x_log10() +
  scale_colour_manual(values = c(`FALSE` = "grey60", `TRUE` = "#CC79A7"),
                      name = "FDR < 0.1") +
  labs(title = "MA, transcript level", x = "base mean", y = "shrunken log2 IP / Input") +
  theme_classic(base_size = 12)

p_pval <- ggplot(tx_res %>% filter(!is.na(pvalue_paired)), aes(x = pvalue_paired)) +
  geom_histogram(breaks = seq(0, 1, 0.02), fill = "#0072B2",
                 colour = "grey20", alpha = 0.6, linewidth = 0.2) +
  geom_hline(yintercept = sum(!is.na(tx_res$pvalue_paired)) * 0.02,
             linetype = "dashed", colour = "#CC79A7") +
  labs(title = "p-value histogram, transcript level",
       subtitle = "Dashed line is the uniform expectation. Below it means conservative.",
       x = "unadjusted p", y = "transcripts") +
  theme_classic(base_size = 12)

p_disp <- tibble(mean = mcols(dds_tx_paired)$baseMean,
                 disp = dispersions(dds_tx_paired)) %>%
  filter(mean > 0, !is.na(disp)) %>%
  ggplot(aes(mean, disp)) +
  geom_point(size = 0.3, alpha = 0.2, colour = "grey40") +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Dispersion", x = "mean of normalised counts", y = "dispersion") +
  theme_classic(base_size = 12)

p_ratio <- ggplot(tx_res, aes(x = log2_ratio, y = lfc_paired)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.3, alpha = 0.2, colour = "#0072B2") +
  labs(title = "Shrinkage against the plain ratio",
       subtitle = paste0("Pearson R = ",
                         sprintf("%.3f", cor(tx_res$log2_ratio, tx_res$lfc_paired))),
       x = "plain normalised log2 ratio", y = "shrunken log2 fold change") +
  theme_classic(base_size = 12)

p_all <- gridExtra::grid.arrange(p_ma, p_pval, p_disp, p_ratio, ncol = 2)
ggsave(here("plots", "dhx29_04_ma_and_pvalues.pdf"), p_all, width = 11, height = 8)
```

---

## Section 3: DESeq2, gene level

```{r deseq_gene}
mat_gene <- as.matrix(gene_counts[, sample_cols])
rownames(mat_gene) <- gene_counts$gene_id_clean

dds_g_paired <- DESeqDataSetFromMatrix(mat_gene, col_data, ~ rep + assay) %>% DESeq(quiet = TRUE)
dds_g_simple <- DESeqDataSetFromMatrix(mat_gene, col_data, ~ assay)       %>% DESeq(quiet = TRUE)

raw_g_paired <- results(dds_g_paired, name = "assay_IP_vs_Input")
raw_g_simple <- results(dds_g_simple, name = "assay_IP_vs_Input")
res_g_paired <- lfcShrink(dds_g_paired, coef = "assay_IP_vs_Input",
                          type = "apeglm", quiet = TRUE)
res_g_simple <- lfcShrink(dds_g_simple, coef = "assay_IP_vs_Input",
                          type = "apeglm", quiet = TRUE)

norm_g <- counts(dds_g_paired, normalized = TRUE)

gene_res <- gene_counts %>%
  dplyr::select(gene_id_clean, symbol, transcript_id, is_mane, cds_codons) %>%
  mutate(base_mean      = res_g_paired$baseMean,
         lfc_paired     = res_g_paired$log2FoldChange,
         padj_paired    = res_g_paired$padj,
         lfc_raw_paired = raw_g_paired$log2FoldChange,
         pvalue_paired  = raw_g_paired$pvalue,
         lfc_simple     = res_g_simple$log2FoldChange,
         padj_simple    = res_g_simple$padj,
         lfc_raw_simple = raw_g_simple$log2FoldChange,
         pvalue_simple  = raw_g_simple$pvalue,
         log2_ratio  = log2((rowMeans(norm_g[, 4:6]) + 1) /
                            (rowMeans(norm_g[, 1:3]) + 1)))

cat("Significant genes, FDR 0.10:  paired", sum(gene_res$padj_paired < 0.10, na.rm = TRUE),
    " unpaired", sum(gene_res$padj_simple < 0.10, na.rm = TRUE),
    " of", format(nrow(gene_res), big.mark = ","), "\n\n")
cat("Top 10 enriched genes by unpaired shrunken fold change, passing FDR 0.1:\n")
print(as.data.frame(
  gene_res %>% filter(padj_simple < 0.1) %>% arrange(desc(lfc_simple)) %>%
    dplyr::select(symbol, base_mean, lfc_simple, padj_simple) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3))) %>% head(10)
))

# Self-enrichment control. The bait's own transcript should be strongly enriched, the same
# check code/predictive_modeling/01n_feature_extraction_ripseq.Rmd applies to EIF3D.
dhx29_row <- gene_res %>% filter(symbol == "DHX29")
cat("\nDHX29 self-enrichment: log2FC", round(dhx29_row$lfc_simple, 3),
    " padj", signif(dhx29_row$padj_simple, 3),
    " rank", which(order(gene_res$lfc_simple, decreasing = TRUE) ==
                   which(gene_res$symbol == "DHX29")), "of",
    format(nrow(gene_res), big.mark = ","), "\n")

stopifnot(
  "DHX29 is not present in the gene table" = nrow(dhx29_row) == 1,
  "the bait transcript is not enriched in its own pull-down, so the sign is wrong" =
    dhx29_row$lfc_simple > 1,
  "DHX29 self-enrichment is not significant" = dhx29_row$padj_simple < 0.01
)
```

---

## Section 4: Do the two designs agree?

Compared on the **unshrunk** fold change, which is the estimate the design produces. Shrunken
values are compared too, but by rank: apeglm pulls almost everything to zero here, so their
Pearson correlation is dominated by a handful of survivors and says little.

```{r design_comparison, fig.width = 9, fig.height = 4.5}
design_cmp <- bind_rows(
  tx_res   %>% transmute(level = "transcript", lfc_paired, lfc_simple,
                         lfc_raw_paired, lfc_raw_simple),
  gene_res %>% transmute(level = "gene", lfc_paired, lfc_simple,
                         lfc_raw_paired, lfc_raw_simple)
) %>% filter(is.finite(lfc_raw_paired), is.finite(lfc_raw_simple))

design_stats <- design_cmp %>%
  group_by(level) %>%
  summarise(n = n(),
            unshrunk_pearson = cor(lfc_raw_paired, lfc_raw_simple),
            shrunk_pearson   = cor(lfc_paired, lfc_simple),
            shrunk_spearman  = cor(lfc_paired, lfc_simple, method = "spearman"),
            sd_unshrunk      = sd(lfc_raw_simple),
            sd_shrunk        = sd(lfc_simple),
            .groups = "drop")
print(as.data.frame(design_stats %>% mutate(across(-c(level, n), ~ round(.x, 4)))))

sig_counts <- bind_rows(
  tx_res %>% summarise(level = "transcript",
                       paired_fdr10 = sum(padj_paired < 0.1, na.rm = TRUE),
                       unpaired_fdr10 = sum(padj_simple < 0.1, na.rm = TRUE)),
  gene_res %>% summarise(level = "gene",
                         paired_fdr10 = sum(padj_paired < 0.1, na.rm = TRUE),
                         unpaired_fdr10 = sum(padj_simple < 0.1, na.rm = TRUE))
)
cat("\nFeatures passing FDR 0.1 under each design:\n")
print(as.data.frame(sig_counts))

stopifnot(
  "the two designs estimate different effects" =
    all(design_stats$unshrunk_pearson > 0.9),
  "shrunken fold changes do not even agree by rank" =
    all(design_stats$shrunk_spearman > 0.8)
)

p_design <- ggplot(design_cmp, aes(lfc_raw_simple, lfc_raw_paired)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.3, alpha = 0.2, colour = "#0072B2") +
  geom_text(data = design_stats, aes(x = -Inf, y = Inf,
            label = paste0("n = ", format(n, big.mark = ","),
                           "\nunshrunk Pearson ", sprintf("%.4f", unshrunk_pearson))),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ level, scales = "free") +
  labs(title = "Blocking on replicate against not",
       subtitle = "Unshrunk log2 fold change, the estimate each design produces",
       x = "~ assay", y = "~ rep + assay") +
  theme_classic(base_size = 12)
print(p_design)
ggsave(here("plots", "dhx29_04_design_comparison.pdf"), p_design, width = 9, height = 4.5)
```

**Blocking does not pay for itself, so the unpaired design is carried forward.** The two
designs estimate the same effect, but blocking spends two of four residual degrees of freedom
on a replicate term that does not explain the variance, and recovers fewer features. The
`sig_counts` table above records the difference.

The shrunken-Pearson column is the artefact it looks like. Shrinkage collapses the fold-change
distribution by roughly a factor of three, so what survives is a small number of large values
whose Pearson correlation is unstable. The rank correlation is the honest summary.

---

## Section 4b: Volcano plots, all four fits

Both levels, both designs. The x axis is the **unshrunk** log2 fold change, because that is
the estimate the Wald test on the y axis is computed from; pairing a shrunken effect with an
unshrunk p would be inconsistent. Shrunken values are in the saved tables.

```{r volcanoes, fig.width = 10, fig.height = 8}
volc <- bind_rows(
  tx_res %>% transmute(level = "transcript", design = "~ rep + assay", label = symbol,
                       lfc = lfc_raw_paired, pval = pvalue_paired, padj = padj_paired),
  tx_res %>% transmute(level = "transcript", design = "~ assay", label = symbol,
                       lfc = lfc_raw_simple, pval = pvalue_simple, padj = padj_simple),
  gene_res %>% transmute(level = "gene", design = "~ rep + assay", label = symbol,
                         lfc = lfc_raw_paired, pval = pvalue_paired, padj = padj_paired),
  gene_res %>% transmute(level = "gene", design = "~ assay", label = symbol,
                         lfc = lfc_raw_simple, pval = pvalue_simple, padj = padj_simple)
) %>%
  filter(is.finite(lfc), is.finite(pval)) %>%
  mutate(level  = factor(level, levels = c("transcript", "gene")),
         design = factor(design, levels = c("~ assay", "~ rep + assay")),
         call   = case_when(is.na(padj) | padj >= 0.1 ~ "not significant",
                            lfc > 0                   ~ "enriched in IP",
                            TRUE                      ~ "depleted in IP"))

volc_counts <- volc %>%
  group_by(level, design) %>%
  summarise(n = n(),
            n_sig = sum(call != "not significant"),
            n_up = sum(call == "enriched in IP"),
            n_down = sum(call == "depleted in IP"),
            .groups = "drop")
print(as.data.frame(volc_counts))

# DHX29 reaches p ~ 1e-100 and would flatten every other point, so the axis is capped
# and capped points are drawn as triangles.
y_cap <- 30
volc <- volc %>%
  mutate(y = pmin(-log10(pval), y_cap), capped = -log10(pval) > y_cap)

volc_labs <- volc %>%
  filter(call != "not significant", !is.na(label)) %>%
  group_by(level, design, label) %>%
  slice_min(padj, n = 1, with_ties = FALSE) %>%      # one row per symbol per panel
  group_by(level, design) %>%
  slice_min(padj, n = 8, with_ties = FALSE) %>%
  ungroup()

p_volc <- ggplot(volc, aes(lfc, y)) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey70") +
  geom_point(aes(colour = call, shape = capped), size = 0.5, alpha = 0.35) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
  ggrepel::geom_text_repel(data = volc_labs, aes(label = label), size = 2.8,
                           box.padding = 0.4,
                           seed = 9, max.overlaps = 20, min.segment.length = 0.2) +
  facet_grid(design ~ level, scales = "free") +
  scale_colour_manual(values = c("not significant" = "grey70",
                                 "enriched in IP"  = "#CC79A7",
                                 "depleted in IP"  = "#0072B2"),
                      name = "FDR < 0.1") +
  labs(title = "DHX29-IP against Input",
       subtitle = str_wrap(paste0(
         "Unshrunk log2 fold change against the Wald p. Positive means enriched in the ",
         "pull-down. Counts per panel are tabulated above; the unpaired design recovers ",
         "more at both levels. The y axis is capped at ", y_cap,
         "; triangles are capped points, DHX29 among them."), width = 100),
       x = "log2 IP / Input", y = "-log10(p), capped") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_volc)
ggsave(here("plots", "dhx29_04_volcanoes.pdf"), p_volc, width = 10, height = 8)

stopifnot(
  "a volcano panel is missing" = nrow(volc_counts) == 4,
  "DHX29 is not a significant enriched call at gene level" =
    volc %>% filter(level == "gene", design == "~ assay", label == "DHX29") %>%
      pull(call) == "enriched in IP"
)
```

---

## Section 5: Transcript against gene

The gene table is built from these same representative transcripts, so this checks that the
collapse and the refit introduced nothing. Divergence comes from size factors being estimated
on different feature sets.

```{r tx_vs_gene, fig.width = 6.5, fig.height = 5}
tg <- gene_res %>%
  dplyr::select(gene_id_clean, symbol, lfc_gene = lfc_simple, transcript_id) %>%
  inner_join(tx_res %>% dplyr::select(transcript_id, lfc_tx = lfc_simple),
             by = "transcript_id")

cat("Genes compared:", format(nrow(tg), big.mark = ","), "\n")
cat("Pearson:", round(cor(tg$lfc_gene, tg$lfc_tx), 4),
    " Spearman:", round(cor(tg$lfc_gene, tg$lfc_tx, method = "spearman"), 4),
    " median |difference|:", round(median(abs(tg$lfc_gene - tg$lfc_tx)), 4), "\n")

p_tg <- ggplot(tg, aes(lfc_tx, lfc_gene)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.4, alpha = 0.25, colour = "#CC79A7") +
  labs(title = "Same transcript, fitted two ways",
       subtitle = paste0("n = ", format(nrow(tg), big.mark = ","),
                         ", Pearson R = ", sprintf("%.4f", cor(tg$lfc_gene, tg$lfc_tx))),
       x = "log2 fold change, transcript-level fit",
       y = "log2 fold change, gene-level fit") +
  theme_classic(base_size = 12)
print(p_tg)
ggsave(here("plots", "dhx29_04_transcript_vs_gene.pdf"), p_tg, width = 6.5, height = 5)
```

---

## Section 6 and 7: Validation

Two comparisons against the paper's expectation that DHX29 rides ribosomes decoding
nonoptimal codons. Section 6 recomputes the prediction on **our own** representative
transcripts, so predicted and measured describe the same sequence. Section 7 joins notebook
23's stored score, which was built on MDA-MB-231-selected transcripts and so may use a
different isoform of the same gene. Section 6 is the stronger test for that reason.

```{r validation, fig.width = 9, fig.height = 4.5}
s8 <- read_csv(here("accessories", "csc_data", "hia_dhx29",
                    "science.adw0288_tables_s1_to_s11",
                    "science.adw0288_table_s8.csv"), show_col_types = FALSE)
stopifnot("Table S8 is not 64 codons" = nrow(s8) == 64)

txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds_seqs <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                  cdsBy(txdb, by = "tx", use.names = TRUE))
cds_seqs <- cds_seqs[(width(cds_seqs) %% 3 == 0) & (width(cds_seqs) >= 33)]
win <- subseq(cds_seqs, start = 16, end = width(cds_seqs) - 15)
cf <- trinucleotideFrequency(win, step = 3)
rownames(cf) <- str_remove(names(win), "\\..*")     # Biostrings drops these

ces <- s8$Fold_Change[match(colnames(cf), s8$codon)]
pred <- tibble(transcript_id_clean = rownames(cf),
               predicted_occupancy = as.vector(cf %*% ces) / rowSums(cf))

val <- gene_res %>%
  mutate(transcript_id_clean = str_remove(transcript_id, "\\..*")) %>%
  left_join(pred, by = "transcript_id_clean") %>%
  left_join(readRDS(here("output", "predictive_modeling",
                         "dhx29_riboseq_occupancy.rds")) %>%
              distinct(ensembl_gene, .keep_all = TRUE) %>%
              dplyr::select(gene_id_clean = ensembl_gene,
                            nb23_occupancy = dhx29_occupancy_score),
            by = "gene_id_clean")

val_long <- val %>%
  dplyr::select(symbol, lfc_paired = lfc_simple,
                `our isoforms (Table S8 applied here)` = predicted_occupancy,
                `notebook 23 stored score` = nb23_occupancy) %>%
  pivot_longer(-c(symbol, lfc_paired), names_to = "source", values_to = "predicted") %>%
  filter(is.finite(predicted), is.finite(lfc_paired))

val_stats <- val_long %>%
  group_by(source) %>%
  summarise(n = n(),
            spearman = cor(predicted, lfc_paired, method = "spearman"),
            p = cor.test(predicted, lfc_paired, method = "spearman",
                         exact = FALSE)$p.value,
            .groups = "drop")
print(as.data.frame(val_stats %>% mutate(spearman = round(spearman, 4),
                                         p = signif(p, 3))))

stopifnot(
  "measured enrichment does not track predicted occupancy in the predicted direction" =
    all(val_stats$spearman > 0),
  "the correlation with predicted occupancy is not significant" =
    all(val_stats$p < 0.01)
)

p_val <- ggplot(val_long, aes(predicted, lfc_paired)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_point(size = 0.4, alpha = 0.25, colour = "#0072B2") +
  geom_smooth(method = "lm", se = FALSE, colour = "#CC79A7", linewidth = 0.6) +
  geom_text(data = val_stats, aes(x = -Inf, y = Inf,
            label = paste0("n = ", format(n, big.mark = ","),
                           "\nSpearman ", sprintf("%.3f", spearman),
                           "\np = ", signif(p, 2))),
            hjust = -0.1, vjust = 1.2, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ source, scales = "free_x") +
  labs(title = "Measured enrichment against predicted DHX29 occupancy",
       subtitle = "Positive log2 fold change means enriched in the pull-down",
       x = "predicted occupancy", y = "measured log2 IP / Input") +
  theme_classic(base_size = 12)
print(p_val)
ggsave(here("plots", "dhx29_04_validation.pdf"), p_val, width = 9, height = 4.5)

# Monotone gradient across enrichment quintiles is the cleanest form of this result.
cat("\nWhere the most-enriched genes sit on predicted occupancy:\n")
print(as.data.frame(
  val %>% filter(is.finite(predicted_occupancy)) %>%
    mutate(bin = ntile(lfc_simple, 5)) %>%
    group_by(enrichment_quintile = bin) %>%
    summarise(n = n(), median_predicted = median(predicted_occupancy),
              .groups = "drop") %>%
    mutate(median_predicted = round(median_predicted, 4)) %>%
    { stopifnot("predicted occupancy is not monotone across enrichment quintiles" =
                  all(diff(.$median_predicted) > 0)); . }
))
```


### Is that correlation real, or an artefact?

The scatter above looks alarming: 98.7% of genes have a shrunken fold change within 0.05 of
zero, so a rank correlation on it could be ranking noise, and n = 10,396 will hand back a
tiny p for almost anything. Three checks.

```{r validation_is_it_real, fig.width = 9, fig.height = 4.5}
vv <- val %>% filter(is.finite(predicted_occupancy), is.finite(lfc_simple),
                     is.finite(lfc_raw_simple))
sp <- function(a, b) cor(a, b, method = "spearman")

cat("1. Does shrinkage manufacture it? Same correlation, three estimators:\n")
cat("   shrunken log2FC:", round(sp(vv$lfc_simple,     vv$predicted_occupancy), 4), "\n")
cat("   unshrunk log2FC:", round(sp(vv$lfc_raw_simple, vv$predicted_occupancy), 4), "\n")
cat("   plain log2 ratio:", round(sp(vv$log2_ratio,    vv$predicted_occupancy), 4), "\n")
cat("   fraction of genes with |shrunken log2FC| < 0.05:",
    round(mean(abs(vv$lfc_simple) < 0.05), 4), "\n\n")

cat("2. Is it expression in disguise?\n")
lbm <- log10(vv$base_mean + 1)
pyz <- sp(vv$predicted_occupancy, lbm)
pxy <- sp(vv$lfc_simple, pyz * 0 + vv$predicted_occupancy)
pxz <- sp(vv$lfc_simple, lbm)
cat("   predicted occupancy vs expression:", round(pyz, 4), "\n")
cat("   partial Spearman, expression removed:",
    round((pxy - pxz * pyz) / sqrt((1 - pxz^2) * (1 - pyz^2)), 4), "\n\n")

cat("3. Does it strengthen where measurement is better?\n")
by_expr <- vv %>%
  mutate(expr_quintile = ntile(base_mean, 5)) %>%
  group_by(expr_quintile) %>%
  summarise(n = n(), median_base_mean = round(median(base_mean), 1),
            rho_unshrunk = sp(lfc_raw_simple, predicted_occupancy),
            rho_shrunk   = sp(lfc_simple, predicted_occupancy),
            .groups = "drop")
print(as.data.frame(by_expr %>% mutate(across(starts_with("rho"), ~ round(.x, 4)))))

stopifnot(
  "the correlation only exists after shrinkage, so it is an artefact" =
    sp(vv$lfc_raw_simple, vv$predicted_occupancy) > 0.05,
  "the correlation does not survive removing expression" =
    (pxy - pxz * pyz) / sqrt((1 - pxz^2) * (1 - pyz^2)) > 0.05,
  "the correlation does not strengthen with better-measured genes" =
    by_expr$rho_unshrunk[5] > by_expr$rho_unshrunk[1]
)

p_real <- by_expr %>%
  dplyr::select(expr_quintile, unshrunk = rho_unshrunk, shrunk = rho_shrunk) %>%
  pivot_longer(-expr_quintile, names_to = "estimator", values_to = "rho") %>%
  ggplot(aes(expr_quintile, rho, colour = estimator, shape = estimator)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_line(linewidth = 0.8) + geom_point(size = 2.5) +
  scale_colour_manual(values = c(unshrunk = "#0072B2", shrunk = "#CC79A7"), name = NULL) +
  scale_shape_manual(values = c(16, 17), name = NULL) +
  labs(title = "The signal is attenuated by measurement noise, not absent",
       subtitle = str_wrap(paste0(
         "Spearman of measured enrichment against predicted occupancy, within bins of ",
         "expression. n = ", format(nrow(vv), big.mark = ","),
         " genes, about ", format(round(nrow(vv)/5), big.mark = ","), " per bin."), width = 90),
       x = "expression quintile (1 = lowest base mean)",
       y = "Spearman rho") +
  theme_classic(base_size = 12)
print(p_real)
ggsave(here("plots", "dhx29_04_validation_by_expression.pdf"),
       p_real, width = 9, height = 4.5)
```

The pooled 0.156 is a diluted average, not the effect size. The correlation is absent in the
worst-measured fifth of genes and reaches 0.375 in the best-measured fifth, which is the
signature of attenuation by measurement error rather than of a spurious association. It is
not shrinkage, since the unshrunk and plain-ratio estimates give the same answer, and it is
not expression, which is essentially uncorrelated with predicted occupancy at 0.040.


---

## Section 8: Save

```{r save}
out_dir <- here("output", "dhx29_reanalysis")
saveRDS(tx_res,   file.path(out_dir, "dhx29_enrichment_transcript.rds"))
saveRDS(val,      file.path(out_dir, "dhx29_enrichment_gene.rds"))
write_csv(val,    file.path(out_dir, "dhx29_enrichment_gene.csv"))

cat("transcript-level:", format(nrow(tx_res), big.mark = ","), "rows\n")
cat("gene-level:", format(nrow(val), big.mark = ","), "rows\n")

stopifnot(
  "output row counts changed unexpectedly" =
    nrow(tx_res) == nrow(tx_counts) && nrow(val) == nrow(gene_counts)
)
```

```{r session_info}
sessionInfo()
```
