---
title: "DHX29 reanalysis 05: GC by codon position, and DHX29 against eIF3d TE"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Two follow-ons to the enrichment measured in notebook 04.

1. GC content of the most and least DHX29-bound mRNAs, split by codon position. The paper's
   mechanism is wobble-position codon choice, so GC3 should move more than GC1 or GC2 if the
   effect is real.
2. DHX29 binding against eIF3d-dependent translation efficiency.

**Cross-cell-line caveat.** DHX29 enrichment is HEK293T. The eIF3d TE data is MDA-MB-231.
Sections 4 to 6 compare a measurement in one cell line against a measurement in another.

GC is computed on our own HEK293T representative transcripts rather than joined from
`feature_matrix_positional_gc.rds`, which is keyed on MDA-MB-231 dominant transcripts and
would put a different isoform against our measurement.

```{r libs}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)

group_cols <- c("DHX29 top 500" = "#CC79A7", "DHX29 bottom 500" = "#0072B2",
                "all other genes" = "grey70")
```

---

## Section 1: Groups, and whether they are significant

Top and bottom 500 by enrichment, from the genes already clearing the notebook 03 floor of a
minimum count of 3 in every sample. That floor is the expression cutoff; nothing further is
applied.

```{r groups}
enr <- readRDS(here("output", "dhx29_reanalysis", "dhx29_enrichment_gene.rds"))
cat("Genes carried from notebook 04:", format(nrow(enr), big.mark = ","), "\n")

enr <- enr %>%
  arrange(desc(lfc_simple)) %>%
  mutate(rank = row_number(),
         group = case_when(rank <= 500                ~ "DHX29 top 500",
                           rank >  n() - 500          ~ "DHX29 bottom 500",
                           TRUE                       ~ "all other genes"),
         group = factor(group, levels = names(group_cols)))

cat("\nTranscriptome-wide counts at padj < 0.05 and |LFC| > 0.5:\n")
print(as.data.frame(tibble(
  estimator = c("shrunken", "unshrunk"),
  enriched  = c(sum(enr$padj_simple < 0.05 & enr$lfc_simple     >  0.5, na.rm = TRUE),
                sum(enr$padj_simple < 0.05 & enr$lfc_raw_simple >  0.5, na.rm = TRUE)),
  depleted  = c(sum(enr$padj_simple < 0.05 & enr$lfc_simple     < -0.5, na.rm = TRUE),
                sum(enr$padj_simple < 0.05 & enr$lfc_raw_simple < -0.5, na.rm = TRUE))
)))

cat("\nHow extreme are the two groups, and do they pass those cutoffs?\n")
print(as.data.frame(
  enr %>% filter(group != "all other genes") %>%
    group_by(group) %>%
    summarise(n = n(),
              lfc_min = round(min(lfc_simple), 3),
              lfc_max = round(max(lfc_simple), 3),
              passing_cutoff = sum(padj_simple < 0.05 & abs(lfc_simple) > 0.5, na.rm = TRUE),
              .groups = "drop")
))

stopifnot(
  "group sizes are wrong" =
    all(table(enr$group)[c("DHX29 top 500", "DHX29 bottom 500")] == 500),
  "groups overlap" = nrow(enr) == n_distinct(enr$gene_id_clean)
)
```

The groups are extreme by ranking, not by significance. The `passing_cutoff` column says how
few of each 500 clear `padj < 0.05` with `|LFC| > 0.5`, and everything downstream should be
read with that in mind.

---

## Section 2: GC by codon position

Full CDS in frame, stop codon removed, GC taken over the whole CDS and at each codon
position.

```{r gc_position}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds_all <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                 cdsBy(txdb, by = "tx", use.names = TRUE))
cds_all <- cds_all[width(cds_all) %% 3 == 0 & width(cds_all) >= 60]
cds_all <- subseq(cds_all, start = 1, end = width(cds_all) - 3)   # drop the stop codon
names(cds_all) <- str_remove(names(cds_all), "\\..*")

# Per-position GC from codon counts: exact, and avoids a per-sequence loop.
codon_n <- trinucleotideFrequency(cds_all, step = 3)
bases   <- do.call(rbind, strsplit(colnames(codon_n), ""))        # 64 codons x 3 positions
is_gc   <- matrix(bases %in% c("G", "C"), nrow = 64)
n_codon <- rowSums(codon_n)

gc_tbl <- tibble(
  transcript_id_clean = names(cds_all),
  cds_gc = 100 * letterFrequency(cds_all, "GC", as.prob = TRUE)[, 1],
  gc1 = 100 * as.vector(codon_n %*% is_gc[, 1]) / n_codon,
  gc2 = 100 * as.vector(codon_n %*% is_gc[, 2]) / n_codon,
  gc3 = 100 * as.vector(codon_n %*% is_gc[, 3]) / n_codon
)

# Two routes to the same number: the three positions must average to the whole-CDS value.
stopifnot("positional GC does not average to CDS GC" =
            max(abs(rowMeans(as.matrix(gc_tbl[, c("gc1", "gc2", "gc3")])) -
                    gc_tbl$cds_gc)) < 1e-8)

gc_dat <- enr %>%
  mutate(transcript_id_clean = str_remove(transcript_id, "\\..*")) %>%
  inner_join(gc_tbl, by = "transcript_id_clean")

cat("Genes with a usable CDS:", format(nrow(gc_dat), big.mark = ","),
    "of", format(nrow(enr), big.mark = ","), "\n")
print(as.data.frame(
  gc_dat %>% group_by(group) %>%
    summarise(n = n(), across(c(cds_gc, gc1, gc2, gc3), ~ round(median(.x), 2)),
              .groups = "drop")
))

actb <- gc_dat$cds_gc[gc_dat$symbol == "ACTB"]
stopifnot(
  "a GC value is outside [0, 100]" =
    all(map_lgl(gc_dat[c("cds_gc", "gc1", "gc2", "gc3")],
                ~ all(.x >= 0 & .x <= 100, na.rm = TRUE))),
  "GC3 median is outside the expected 40 to 65 percent" =
    between(median(gc_dat$gc3), 40, 65),
  "ACTB CDS GC is outside 50 to 70 percent" =
    length(actb) == 1 && between(actb, 50, 70)
)
cat("ACTB CDS GC:", round(actb, 2), "%\n")
```

---

## Section 3: The figure

```{r gc_figure, fig.width = 8, fig.height = 9}
gc_long <- gc_dat %>%
  dplyr::select(group, `CDS GC` = cds_gc, GC1 = gc1, GC2 = gc2, GC3 = gc3) %>%
  pivot_longer(-group, names_to = "measure", values_to = "gc") %>%
  mutate(measure = factor(measure, levels = c("CDS GC", "GC1", "GC2", "GC3")))

gc_meds <- gc_long %>% group_by(measure, group) %>%
  summarise(med = median(gc), .groups = "drop")

gc_tests <- gc_long %>%
  filter(group != "all other genes") %>%
  group_by(measure) %>%
  summarise(p = wilcox.test(gc ~ droplevels(group))$p.value,
            delta = median(gc[group == "DHX29 top 500"]) -
                    median(gc[group == "DHX29 bottom 500"]),
            .groups = "drop") %>%
  mutate(lab = paste0("top - bottom = ", sprintf("%+.2f", delta),
                      "\nWilcoxon p = ", signif(p, 3)))
print(as.data.frame(gc_tests %>% mutate(p = signif(p, 3), delta = round(delta, 3))))

grp_n <- gc_dat %>% dplyr::count(group) %>%
  mutate(lab = paste0(group, " (n = ", format(n, big.mark = ","), ")"))

p_gc <- ggplot(gc_long, aes(gc, colour = group)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = gc_meds, aes(xintercept = med, colour = group),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  geom_text(data = gc_tests, aes(x = Inf, y = Inf, label = lab),
            hjust = 1.1, vjust = 1.4, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ measure, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = group_cols, labels = grp_n$lab, name = NULL) +
  labs(title = "GC by codon position, most against least DHX29-bound",
       subtitle = str_wrap(paste0(
         "HEK293T representative transcripts. Top and bottom 500 by measured IP over ",
         "Input enrichment among genes clearing a floor of 3. Dashed lines are group ",
         "medians."), width = 88),
       x = "GC (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_gc)
ggsave(here("plots", "dhx29_05_gc_position_density.pdf"), p_gc, width = 8, height = 9)
```

---

## Section 4: eIF3d TE data, with the direction check

`translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv` is named in the opposite order from
the coefficient it contains. The coefficient is `sictrl_vs_si3d`, so every column is
**log2(sictrl / si3d)** and a positive `te_lfc` means eIF3d promotes. Labels below say so.

```{r load_te}
te <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
               show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

promoted <- te %>% filter(te_lfc > 0.5, te_padj < 0.05)
geneset  <- read_csv(here("output", "genesets",
                          "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE)

eif3d_rna <- te$rna_lfc[te$symbol == "EIF3D"]
cat("EIF3D rna_lfc:", round(eif3d_rna, 3),
    "(must be strongly positive; the siRNA depletes its own target)\n")
cat("Genes with te_lfc > 0.5 and te_padj < 0.05:", nrow(promoted), "\n")
cat("Genes in hypoxia_3d_promotes_TE_1hr_lfc0.5.csv:", nrow(geneset), "\n")
cat("Overlap:", length(intersect(promoted$gene_id_clean, geneset$ensembl_gene)), "\n")

stopifnot(
  "EIF3D is not depleted by its own siRNA, so the contrast direction has flipped" =
    length(eif3d_rna) == 1 && eif3d_rna > 2,
  "the promoted set is not 1604 genes" = nrow(promoted) == 1604,
  "the promoted set does not match the stored gene set" =
    length(intersect(promoted$gene_id_clean, geneset$ensembl_gene)) == 1604
)
cat("Spot check passed: 1,604 mRNAs promoted by eIF3d at LFC 0.5, padj 0.05.\n")
```

---

## Section 5: DHX29 enrichment against eIF3d TE

```{r te_correlation, fig.width = 7.5, fig.height = 5.5}
comb <- enr %>%
  inner_join(te %>% dplyr::select(gene_id_clean, te_lfc, te_padj, rna_lfc),
             by = "gene_id_clean") %>%
  mutate(eif3d = if_else(gene_id_clean %in% promoted$gene_id_clean,
                         "eIF3d promoted", "other"),
         eif3d = factor(eif3d, levels = c("other", "eIF3d promoted")))

cat("Genes shared between the DHX29 and TE tables:",
    format(nrow(comb), big.mark = ","), "\n")
cat("  of which eIF3d promoted:", sum(comb$eif3d == "eIF3d promoted"), "\n\n")

ct <- cor.test(comb$lfc_simple, comb$te_lfc, method = "spearman", exact = FALSE)
cat("Pooled Spearman:", round(ct$estimate, 4), " p =", signif(ct$p.value, 3),
    " Pearson:", round(cor(comb$lfc_simple, comb$te_lfc), 4), "\n\n")

cat("Within expression quintiles, since notebook 04 showed attenuation:\n")
print(as.data.frame(
  comb %>% mutate(q = ntile(base_mean, 5)) %>%
    group_by(expr_quintile = q) %>%
    summarise(n = n(),
              spearman = round(cor(lfc_simple, te_lfc, method = "spearman"), 4),
              .groups = "drop")
))

p_cor <- ggplot(comb, aes(te_lfc, lfc_simple)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_point(data = comb %>% filter(eif3d == "other"),
             colour = "grey70", size = 0.5, alpha = 0.4) +
  geom_point(data = comb %>% filter(eif3d == "eIF3d promoted"),
             colour = "#E69F00", size = 0.7, alpha = 0.7) +
  labs(title = "DHX29 ribosome enrichment against eIF3d-dependent TE",
       subtitle = str_wrap(paste0(
         "n = ", format(nrow(comb), big.mark = ","), " genes, Spearman ",
         sprintf("%.4f", ct$estimate), ", p = ", signif(ct$p.value, 3),
         ". Orange marks the ", sum(comb$eif3d == "eIF3d promoted"),
         " eIF3d-promoted genes present here. DHX29 is HEK293T, TE is MDA-MB-231."),
         width = 95),
       x = "TE log2(sictrl / si3d), hypoxia 1hr. Positive means eIF3d promotes",
       y = "DHX29 log2 IP / Input") +
  theme_classic(base_size = 12)

print(p_cor)
ggsave(here("plots", "dhx29_05_te_correlation.pdf"), p_cor, width = 7.5, height = 5.5)
```

Almost the whole distribution sits within a few hundredths of zero, so the plot above is
mostly empty space around a line. Same data, y clipped to plus or minus 0.1.

```{r te_correlation_zoom, fig.width = 7.5, fig.height = 5.5}
n_out <- sum(abs(comb$lfc_simple) > 0.1)

p_zoom <- p_cor +
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  labs(title = "DHX29 enrichment against eIF3d TE, y clipped",
       subtitle = str_wrap(paste0(
         "n = ", format(nrow(comb), big.mark = ","), " genes, of which ",
         format(n_out, big.mark = ","), " fall outside the plus or minus 0.1 window and ",
         "are not drawn. Statistics are unchanged: Spearman ",
         sprintf("%.4f", ct$estimate), ", p = ", signif(ct$p.value, 3),
         ". Orange marks the ", sum(comb$eif3d == "eIF3d promoted"),
         " eIF3d-promoted genes."), width = 95))

print(p_zoom)
ggsave(here("plots", "dhx29_05_te_correlation_zoom.pdf"), p_zoom,
       width = 7.5, height = 5.5)

cat("Genes outside the clipped window:", n_out,
    sprintf("(%.2f%%)", 100 * n_out / nrow(comb)), "\n")
```

---

## Section 6: The gene-set contrast

Most genes have a DHX29 enrichment near zero, so a correlation is a weak instrument. Asking
whether the eIF3d-promoted set differs from the rest tests the same hypothesis with more
power.

```{r geneset_contrast, fig.width = 7.5, fig.height = 4.5}
w <- wilcox.test(lfc_simple ~ eif3d, data = comb)
meds <- comb %>% group_by(eif3d) %>%
  summarise(n = n(), median_lfc = median(lfc_simple), .groups = "drop")
print(as.data.frame(meds %>% mutate(median_lfc = round(median_lfc, 4))))
cat("Wilcoxon p =", signif(w$p.value, 3),
    " difference in medians =",
    round(diff(meds$median_lfc), 4), "\n")

p_set <- ggplot(comb, aes(lfc_simple, colour = eif3d)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = meds, aes(xintercept = median_lfc, colour = eif3d),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("Wilcoxon p = ", signif(w$p.value, 3)),
           hjust = 1.1, vjust = 1.4, size = 3.2) +
  coord_cartesian(xlim = quantile(comb$lfc_simple, c(0.005, 0.995))) +
  scale_colour_manual(values = c("other" = "grey70",
                                 "eIF3d promoted" = "#E69F00"),
                      labels = paste0(meds$eif3d, " (n = ",
                                      format(meds$n, big.mark = ","), ")"),
                      name = NULL) +
  labs(title = "DHX29 enrichment, eIF3d-promoted genes against the rest",
       subtitle = "Axis clipped to the central 99%. Dashed lines are group medians.",
       x = "DHX29 log2 IP / Input", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_set)
ggsave(here("plots", "dhx29_05_te_geneset_contrast.pdf"), p_set, width = 7.5, height = 4.5)
```

### Does the null depend on how "promoted" is defined?

The contrast above is asymmetric: the promoted group requires both `te_lfc > 0.5` and
`te_padj < 0.05`, while the comparator is everything else, which includes genes that clear
the fold-change bar but miss significance. Those belong to neither a clean positive nor a
clean negative set. This section checks whether the null survives dropping the significance
requirement entirely.

```{r definition_sensitivity, fig.width = 8, fig.height = 4.5}
cat("What is actually inside the comparator group:\n")
print(as.data.frame(
  comb %>% filter(eif3d == "other") %>%
    mutate(what = case_when(
      te_lfc >  0.5 & (is.na(te_padj) | te_padj >= 0.05) ~ "promoted by LFC, not significant",
      te_lfc < -0.5 & te_padj < 0.05                     ~ "significantly inhibited",
      te_lfc < -0.5                                      ~ "inhibited by LFC, not significant",
      TRUE                                               ~ "within +/- 0.5 LFC")) %>%
    dplyr::count(what) %>% mutate(pct = round(100 * n / sum(n), 1)) %>% arrange(desc(n))
))

cat("\nPromoted in the overlap: ", sum(comb$te_lfc > 0.5 & comb$te_padj < 0.05, na.rm = TRUE),
    " with significance, ", sum(comb$te_lfc > 0.5, na.rm = TRUE),
    " on fold change alone\n\n", sep = "")

# 1. Every gene, binned by TE fold change, no significance filter anywhere.
binned <- comb %>%
  mutate(bin = cut(te_lfc, breaks = c(-Inf, -1, -0.5, -0.1, 0.1, 0.5, 1, 1.5, 2, Inf),
                   labels = c("< -1", "-1 to -0.5", "-0.5 to -0.1", "-0.1 to 0.1",
                              "0.1 to 0.5", "0.5 to 1", "1 to 1.5",
                              "1.5 to 2", "> 2"))) %>%
  group_by(bin) %>%
  summarise(n = n(), median_dhx29 = median(lfc_simple),
            q25 = quantile(lfc_simple, 0.25), q75 = quantile(lfc_simple, 0.75),
            .groups = "drop")
cat("DHX29 enrichment by TE fold-change bin, all genes:\n")
print(as.data.frame(binned %>% mutate(across(where(is.numeric), ~ signif(.x, 3)))))

# 2. Three disjoint classes on fold change alone.
lfc_only <- comb %>%
  mutate(cls = case_when(te_lfc >  0.5 ~ "promoted (LFC only)",
                         te_lfc < -0.5 ~ "inhibited (LFC only)",
                         TRUE          ~ "unchanged"))
kw <- kruskal.test(lfc_simple ~ cls, data = lfc_only)
cat("\nFold change only, three classes:\n")
print(as.data.frame(
  lfc_only %>% group_by(cls) %>%
    summarise(n = n(), median_dhx29 = signif(median(lfc_simple), 3), .groups = "drop")
))
cat("Kruskal-Wallis p =", signif(kw$p.value, 3), "\n")

p_bin <- ggplot(binned, aes(bin, median_dhx29)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2, colour = "grey50") +
  geom_point(size = 3, colour = "#E69F00") +
  geom_text(aes(label = paste0("n = ", format(n, big.mark = ","))),
            vjust = -1.2, size = 2.8) +
  labs(title = "DHX29 enrichment across the full eIF3d TE range",
       subtitle = str_wrap(paste0(
         "Every gene in the overlap, binned on TE log2(sictrl / si3d) with no ",
         "significance filter. Points are medians, bars the interquartile range. ",
         "Positive TE means eIF3d promotes."), width = 92),
       x = "TE log2(sictrl / si3d), hypoxia 1hr", y = "median DHX29 log2 IP / Input") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(p_bin)
ggsave(here("plots", "dhx29_05_te_bins_all_genes.pdf"), p_bin, width = 8, height = 4.5)
```

The null holds under every definition. Binning all 8,710 genes on fold change alone moves the
median DHX29 enrichment by about 0.003 in log2 units across the whole range, and the ordering
is not monotone. Dropping the significance requirement raises the promoted group from 1,315 to
2,561 genes and changes nothing. The small Kruskal-Wallis p is n again, not effect.

---

### Is the TE result null, or just attenuated?

Notebook 04 used a specific diagnostic: a real signal buried in measurement noise should
strengthen as measurement improves. There, the correlation with predicted codon occupancy
rose monotonically across expression quintiles, from about zero to 0.375.

Neither the quintile table above nor the fold-change bins do that. It is flat and sits near zero throughout, so the
same diagnostic that certified the codon result as real certifies this one as genuinely
absent. The Wilcoxon p of the gene-set contrast is small only because n is 8,710; the
difference in medians is about 0.001 in log2 units, which is nothing.

Read together: DHX29 ribosome occupancy is predicted by codon content, and is unrelated to
eIF3d-dependent translation efficiency. The cross-cell-line design is a genuine limitation
here, since a HEK293T measurement cannot rule out a relationship that exists in MDA-MB-231.

---

## Section 7: Save

```{r save}
out <- here("output", "dhx29_reanalysis")
saveRDS(gc_dat %>% dplyr::select(gene_id_clean, symbol, transcript_id, group,
                                 lfc_simple, padj_simple, cds_gc, gc1, gc2, gc3),
        file.path(out, "dhx29_gc_by_position.rds"))
write_csv(comb %>% dplyr::select(gene_id_clean, symbol, lfc_simple, padj_simple,
                                 base_mean, te_lfc, te_padj, rna_lfc, eif3d),
          file.path(out, "dhx29_vs_te_hypoxia1hr.csv"))
cat("Wrote dhx29_gc_by_position.rds (", nrow(gc_dat), " rows) and ",
    "dhx29_vs_te_hypoxia1hr.csv (", nrow(comb), " rows)\n", sep = "")
```

```{r session_info}
sessionInfo()
```
