---
title: "DHX29 reanalysis 06: gene-level counts by summing transcripts, and DESeq2"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Notebook 03 collapsed to gene level by picking one representative transcript per gene. This
notebook takes the other route: sum every transcript into its gene, pool the six samples into
a single total, and look at that distribution to place a floor.

A fork, not a replacement. Notebooks 03 to 05 continue to use the representative-transcript
table; nothing here feeds them.

**The caveat.** Summing double-counts reads wherever the same footprints were assigned to
more than one isoform. Notebook 03 measured that population: 2.1% of multi-isoform genes
share one identical count across isoforms of differing CDS length, the duplication signature,
and a further 8.5% share a count across isoforms whose CDS lengths are genuinely identical,
which is benign. Section 3 quantifies the combined inflation. The summed table is fine for a
coverage question like where to put a floor. It is not appropriate as a differential input
without this caveat attached, which is why notebook 03 chose a representative instead.

```{r libs}
library(tidyverse)
library(here)
library(DESeq2)
library(apeglm)
library(ggrepel)

sample_cols <- c("Input1", "Input2", "Input3", "IP1", "IP2", "IP3")
# Chosen to sit just above the threshold DESeq2 selects for itself. Notebook 04's
# independent filtering picked a base mean of 11.48, which over six samples is a pooled
# total of about 69, so a floor here of 70 stops being redone downstream. The summed
# equivalent of notebook 03's "3 in every sample" would be 18, which leaves roughly
# 4,000 genes for DESeq2 to discard later anyway.
cutoff_total <- 70
```

---

## Section 1: Aggregate transcripts to genes

```{r aggregate}
tx <- readRDS(here("output", "dhx29_reanalysis", "dhx29_counts_with_gene.rds")) %>%
  filter(!is.na(cds_codons), !is.na(gene_id_clean))

gene_sum <- tx %>%
  group_by(gene_id_clean, sample) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  pivot_wider(names_from = sample, values_from = count) %>%
  mutate(total = rowSums(across(all_of(sample_cols))))

sym <- tx %>% distinct(gene_id_clean, symbol)
gene_sum <- gene_sum %>% left_join(sym, by = "gene_id_clean")

cat("coding transcripts in: ",
    format(n_distinct(tx$transcript_id), big.mark = ","), "\n", sep = "")
cat("genes out:             ", format(nrow(gene_sum), big.mark = ","), "\n", sep = "")
cat("genes carrying counts: ", format(sum(gene_sum$total > 0), big.mark = ","), "\n\n", sep = "")

cat("per-gene total across all six samples:\n")
print(round(quantile(gene_sum$total, c(0.10, 0.25, 0.50, 0.75, 0.90, 0.99))))
```

---

## Section 2: Distribution and the floor

The axis keeps log2 spacing but is labelled in counts, so the line sits on a readable tick.

```{r plot_summed, fig.width = 7.5, fig.height = 4.5}
retention <- tibble(floor = c(3, 18, 30, 50, 70, 100, 150)) %>%
  mutate(genes = map_int(floor, ~ sum(gene_sum$total >= .x)),
         pct   = round(100 * genes / nrow(gene_sum), 1))
cat("genes retained at each candidate floor:\n")
print(as.data.frame(retention))
cat("\nchosen floor:", cutoff_total, "->",
    format(sum(gene_sum$total >= cutoff_total), big.mark = ","), "genes\n")

p_sum <- ggplot(gene_sum, aes(log2(total + 1))) +
  geom_density(linewidth = 0.8, colour = "grey20") +
  geom_vline(xintercept = log2(cutoff_total + 1), linetype = "dashed",
             colour = "grey15", linewidth = 0.5) +
  annotate("text", x = log2(cutoff_total + 1), y = Inf,
           label = paste0("total = ", cutoff_total),
           hjust = -0.1, vjust = 1.8, size = 3.2, colour = "grey15") +
  scale_x_continuous(
    breaks = log2(c(0, 1, 10, 70, 300, 1000, 10000, 1e5) + 1),
    labels = c(0, 1, 10, 70, 300, 1000, 10000, 1e5)) +
  labs(
    title    = "Gene-level footprint totals, all six samples pooled",
    subtitle = str_wrap(paste0(
      "n = ", format(nrow(gene_sum), big.mark = ","),
      " genes, counts summed across every transcript and every sample. The floor of ",
      cutoff_total, " sits just above the base-mean threshold DESeq2 selects for itself."),
      width = 88),
    x = "total footprints per gene (log2 spacing)", y = "Density"
  ) +
  theme_classic(base_size = 12)

print(p_sum)
ggsave(here("plots", "dhx29_06_summed_count_distribution.pdf"),
       p_sum, width = 7.5, height = 4.5)
```

---

## Section 3: How much did summing inflate the counts?

```{r inflation}
max_tx <- tx %>%
  group_by(gene_id_clean, sample) %>%
  summarise(count = max(count), .groups = "drop") %>%
  group_by(gene_id_clean) %>%
  summarise(total_max_tx = sum(count), .groups = "drop")

infl <- gene_sum %>%
  dplyr::select(gene_id_clean, total) %>%
  inner_join(max_tx, by = "gene_id_clean") %>%
  filter(total_max_tx > 0) %>%
  mutate(ratio = total / total_max_tx)

cat("ratio of the gene sum to its largest single transcript, n = ",
    format(nrow(infl), big.mark = ","), "\n", sep = "")
print(round(quantile(infl$ratio, c(0.25, 0.50, 0.75, 0.90, 0.99)), 3))
cat("\ngenes where summing changes nothing (single transcript carries all): ",
    format(sum(infl$ratio == 1), big.mark = ","),
    sprintf(" (%.1f%%)", 100 * mean(infl$ratio == 1)), "\n", sep = "")

stopifnot(
  "summing should never reduce a gene's count below its largest transcript" =
    all(infl$ratio >= 1)
)
```

Part of that ratio is genuine spread across isoforms and part is the same reads counted more
than once. These files cannot separate the two.

---

## Section 4: Checks

```{r checks}
tx_total   <- sum(tx$count)
gene_total <- sum(gene_sum$total)
cat("footprints in the transcript table:", format(tx_total, big.mark = ","), "\n")
cat("footprints in the gene table:      ", format(gene_total, big.mark = ","), "\n")

hk <- tibble(symbol = c("ACTB", "GAPDH", "EEF1A1"),
             gene_id_clean = c("ENSG00000075624", "ENSG00000111640", "ENSG00000156508"))
hk_rank <- gene_sum %>%
  mutate(pct = percent_rank(total)) %>%
  inner_join(hk %>% dplyr::rename(expected = symbol), by = "gene_id_clean")
print(as.data.frame(hk_rank %>%
                      transmute(expected, total, percentile = round(pct, 4))))

stopifnot(
  "aggregation did not conserve footprints" = tx_total == gene_total,
  "counts are not finite non-negative integers" =
    all(map_lgl(gene_sum[sample_cols], ~ all(is.finite(.x)) && all(.x >= 0))),
  "the gene table is not one row per gene" =
    nrow(gene_sum) == n_distinct(gene_sum$gene_id_clean),
  "gene count changed unexpectedly" = nrow(gene_sum) == 20322,
  "a housekeeping gene is not in the top 1% of totals" =
    nrow(hk_rank) == 3 && all(hk_rank$pct > 0.99)
)
cat("\nAll checks passed.\n")
```

The conservation check is the load-bearing one: summing must neither drop nor duplicate a
footprint relative to its own input. It does not, and cannot, detect duplication that was
already present in the GEO files.

---

## Section 6: Gene-level DESeq2 on the summed table

Same contrast as notebook 04, but on genes built by summing transcripts and filtered at a
pooled total of `r cutoff_total`. Both designs fitted so the blocking term can be judged
rather than assumed.

```{r deseq_summed}
keep <- gene_sum %>% filter(total >= cutoff_total)
mat  <- as.matrix(keep[, sample_cols])
rownames(mat) <- keep$gene_id_clean

col_data <- tibble(sample = sample_cols) %>%
  mutate(assay = factor(if_else(str_starts(sample, "Input"), "Input", "IP"),
                        levels = c("Input", "IP")),
         rep   = factor(str_extract(sample, "\\d$"))) %>%
  column_to_rownames("sample")

stopifnot(
  "Input is not the reference level" = levels(col_data$assay)[1] == "Input",
  "counts entering DESeq2 are not clean" =
    all(is.finite(mat)) && all(mat >= 0) && all(mat == round(mat))
)

dds_paired <- DESeqDataSetFromMatrix(mat, col_data, ~ rep + assay) %>% DESeq(quiet = TRUE)
dds_simple <- DESeqDataSetFromMatrix(mat, col_data, ~ assay)       %>% DESeq(quiet = TRUE)

raw_paired <- results(dds_paired, name = "assay_IP_vs_Input", alpha = 0.05)
raw_simple <- results(dds_simple, name = "assay_IP_vs_Input", alpha = 0.05)
shr_paired <- lfcShrink(dds_paired, coef = "assay_IP_vs_Input", type = "apeglm", quiet = TRUE)
shr_simple <- lfcShrink(dds_simple, coef = "assay_IP_vs_Input", type = "apeglm", quiet = TRUE)

res <- keep %>%
  dplyr::select(gene_id_clean, symbol, total) %>%
  mutate(base_mean      = raw_simple$baseMean,
         lfc_raw_paired = raw_paired$log2FoldChange,
         pvalue_paired  = raw_paired$pvalue,
         padj_paired    = raw_paired$padj,
         lfc_paired     = shr_paired$log2FoldChange,
         lfc_raw_simple = raw_simple$log2FoldChange,
         pvalue_simple  = raw_simple$pvalue,
         padj_simple    = raw_simple$padj,
         lfc_simple     = shr_simple$log2FoldChange)

cat("genes tested:", format(nrow(res), big.mark = ","), "\n")
cat("size factors:\n"); print(round(sizeFactors(dds_simple), 4))
cat("\nDESeq2 independent filtering threshold (baseMean):",
    round(metadata(raw_simple)$filterThreshold, 2),
    " -> testable:", format(sum(!is.na(res$padj_simple)), big.mark = ","), "\n\n")

sig <- tibble(
  design   = c("~ assay", "~ rep + assay"),
  fdr_0.05 = c(sum(res$padj_simple < 0.05, na.rm = TRUE),
               sum(res$padj_paired < 0.05, na.rm = TRUE)),
  fdr_0.10 = c(sum(res$padj_simple < 0.10, na.rm = TRUE),
               sum(res$padj_paired < 0.10, na.rm = TRUE)),
  up_0.05_lfc0.5 = c(sum(res$padj_simple < 0.05 & res$lfc_simple >  0.5, na.rm = TRUE),
                     sum(res$padj_paired < 0.05 & res$lfc_paired >  0.5, na.rm = TRUE)),
  down_0.05_lfc0.5 = c(sum(res$padj_simple < 0.05 & res$lfc_simple < -0.5, na.rm = TRUE),
                       sum(res$padj_paired < 0.05 & res$lfc_paired < -0.5, na.rm = TRUE)))
print(as.data.frame(sig))

dhx29 <- res %>% filter(symbol == "DHX29")
cat("\nDHX29 self-enrichment, unpaired: log2FC", round(dhx29$lfc_simple, 3),
    " padj", signif(dhx29$padj_simple, 3), "\n")
stopifnot(
  "the bait is not enriched in its own pull-down, so the sign is wrong" =
    nrow(dhx29) == 1 && dhx29$lfc_simple > 1 && dhx29$padj_simple < 0.01
)
```

---

## Section 7: Do the two designs agree?

Compared on the unshrunk fold change, which is the estimate each design produces and the one
the Wald test on it belongs to.

Spearman throughout. The counts are not normal and their variance is not constant, so a
product-moment correlation has no claim here. It is also unusable in practice: a pooled total
does not guarantee a count in every sample, so a few genes carry a zero in one condition and
their unshrunk estimate blows up to tens of log2 units, which dominates Pearson while barely
touching the ranks.

```{r design_correlation, fig.width = 6.5, fig.height = 5.5}
cmp <- res %>% filter(is.finite(lfc_raw_paired), is.finite(lfc_raw_simple))

zero_in_a_sample <- sum(pmin(keep$Input1, keep$Input2, keep$Input3,
                             keep$IP1, keep$IP2, keep$IP3) == 0)
cat("genes with a zero in at least one sample:", zero_in_a_sample,
    sprintf(" (%.1f%%)", 100 * zero_in_a_sample / nrow(keep)),
    "  <- a pooled floor cannot exclude these\n\n")

stats <- tibble(
  n                 = nrow(cmp),
  unshrunk_spearman = cor(cmp$lfc_raw_paired, cmp$lfc_raw_simple, method = "spearman"),
  shrunk_spearman   = cor(cmp$lfc_paired, cmp$lfc_simple, method = "spearman"),
  sd_unshrunk       = sd(cmp$lfc_raw_simple),
  sd_shrunk         = sd(cmp$lfc_simple))
print(as.data.frame(stats %>% mutate(across(-n, ~ round(.x, 4)))))

fit <- lm(lfc_raw_paired ~ lfc_raw_simple, data = cmp)
cat("\nslope:", round(coef(fit)[2], 4), " intercept:", round(coef(fit)[1], 4), "\n")

# Pearson is not asserted: it is hostage to a handful of blown-up estimates on genes with
# a zero in one condition. Rank agreement is the robust statement.
stopifnot(
  "the two designs do not agree by rank on the unshrunk estimate" =
    stats$unshrunk_spearman > 0.9,
  "shrunken fold changes do not agree by rank" = stats$shrunk_spearman > 0.8
)

p_cmp <- ggplot(cmp, aes(lfc_raw_simple, lfc_raw_paired)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.4, alpha = 0.25, colour = "grey40") +
  geom_smooth(method = "lm", se = TRUE, colour = "#0072B2",
              fill = "#0072B2", alpha = 0.15, linewidth = 0.8) +
  annotate("text", x = -Inf, y = Inf,
           label = paste0("n = ", format(stats$n, big.mark = ","),
                          "\nSpearman ", sprintf("%.4f", stats$unshrunk_spearman),
                          "\nslope ", sprintf("%.3f", coef(fit)[2])),
           hjust = -0.1, vjust = 1.2, size = 3.2) +
  coord_cartesian(xlim = quantile(cmp$lfc_raw_simple, c(0.001, 0.999)),
                  ylim = quantile(cmp$lfc_raw_paired, c(0.001, 0.999))) +
  labs(title = "Blocking on replicate against not",
       subtitle = str_wrap(paste0(
         "Unshrunk log2 IP over Input per gene, summed counts at a floor of ",
         cutoff_total, ". Dashed line is y = x, blue is a linear fit. Axes show the ",
         "central 99.8%; the excluded points are genes with a zero in one condition."),
         width = 80),
       x = "~ assay", y = "~ rep + assay") +
  theme_classic(base_size = 12)

print(p_cmp)
ggsave(here("plots", "dhx29_06_design_correlation.pdf"), p_cmp, width = 6.5, height = 5.5)
```

---

## Section 8: Volcano, both designs

```{r volcano_summed, fig.width = 9, fig.height = 5}
sig_p <- 0.05; sig_lfc <- 0.5; y_cap <- 30

volc <- bind_rows(
  res %>% transmute(design = "~ assay", label = symbol,
                    lfc = lfc_raw_simple, pval = pvalue_simple, padj = padj_simple),
  res %>% transmute(design = "~ rep + assay", label = symbol,
                    lfc = lfc_raw_paired, pval = pvalue_paired, padj = padj_paired)
) %>%
  filter(is.finite(lfc), is.finite(pval)) %>%
  mutate(design = factor(design, levels = c("~ assay", "~ rep + assay")),
         call = case_when(is.na(padj) | padj >= sig_p | abs(lfc) <= sig_lfc ~ "not significant",
                          lfc > sig_lfc ~ "Up", TRUE ~ "Down"),
         call = factor(call, levels = c("Up", "Down", "not significant")),
         y = pmin(-log10(pval), y_cap), capped = -log10(pval) > y_cap)

counts_tbl <- volc %>%
  group_by(design) %>%
  summarise(n = n(), up = sum(call == "Up"), down = sum(call == "Down"), .groups = "drop")
print(as.data.frame(counts_tbl))

# The y axis is the raw Wald p but the call uses padj, so the line must be found per panel.
sig_lines <- volc %>% filter(!is.na(padj), padj < sig_p) %>%
  group_by(design) %>% summarise(yint = -log10(max(pval)), .groups = "drop")
lab_up   <- counts_tbl %>% mutate(lab = paste0("Up: ", up))
lab_down <- counts_tbl %>% mutate(lab = paste0("Down: ", down))

volc_labs <- volc %>%
  filter(call != "not significant", !is.na(label)) %>%
  group_by(design, label) %>% slice_min(padj, n = 1, with_ties = FALSE) %>%
  group_by(design) %>% slice_min(padj, n = 8, with_ties = FALSE) %>% ungroup()

p_volc <- ggplot(volc, aes(lfc, y)) +
  geom_vline(xintercept = c(-sig_lfc, sig_lfc), linetype = "dotted", colour = "grey60") +
  geom_hline(data = sig_lines, aes(yintercept = yint), linetype = "dotted", colour = "grey60") +
  geom_point(aes(colour = call, shape = capped, alpha = call, size = call)) +
  geom_text(data = lab_up, aes(x = Inf, y = Inf, label = lab), hjust = 1.15, vjust = 1.6,
            size = 3.4, colour = "#D55E00", inherit.aes = FALSE) +
  geom_text(data = lab_down, aes(x = -Inf, y = Inf, label = lab), hjust = -0.15, vjust = 1.6,
            size = 3.4, colour = "#0072B2", inherit.aes = FALSE) +
  ggrepel::geom_text_repel(data = volc_labs, aes(label = label), size = 2.8,
                           box.padding = 0.4, seed = 9, max.overlaps = 20,
                           min.segment.length = 0.2) +
  facet_wrap(~ design) +
  scale_colour_manual(values = c("Up" = "#D55E00", "Down" = "#0072B2",
                                 "not significant" = "grey75"), name = NULL) +
  scale_alpha_manual(values = c("Up" = 0.9, "Down" = 0.9, "not significant" = 0.25),
                     guide = "none") +
  scale_size_manual(values = c("Up" = 1.3, "Down" = 1.3, "not significant" = 0.45),
                    guide = "none") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
  coord_cartesian(xlim = quantile(volc$lfc, c(0.001, 0.999))) +
  labs(title = "DHX29-IP against Input, summed gene counts",
       x = "log2 IP / Input", y = paste0("-log10(Wald p), capped at ", y_cap)) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_volc)
ggsave(here("plots", "dhx29_06_volcano_gene_summed.pdf"), p_volc, width = 9, height = 5)

stopifnot(
  "a volcano panel is missing" = nrow(counts_tbl) == 2,
  "DHX29 is not called Up under the unpaired design" =
    volc %>% filter(design == "~ assay", label == "DHX29") %>% pull(call) == "Up"
)
```

---

## Section 5: Save

```{r save}
out <- here("output", "dhx29_reanalysis", "dhx29_gene_summed_counts.rds")
saveRDS(gene_sum %>% dplyr::select(gene_id_clean, symbol, all_of(sample_cols), total), out)
res_out <- here("output", "dhx29_reanalysis", "dhx29_gene_summed_enrichment.rds")
saveRDS(res, res_out)
cat("Wrote", out, "\n  ", format(nrow(gene_sum), big.mark = ","), "genes, one row each\n")
cat("Wrote", res_out, "\n  ", format(nrow(res), big.mark = ","),
    "genes clearing the floor, both designs\n")
```

```{r session_info}
sessionInfo()
```
