---
title: "DHX29 reanalysis 07: summed-gene enrichment against eIF3d TE"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

The two eIF3d comparisons from notebook 05, rebuilt on the summed-gene enrichment from
notebook 06 rather than the representative-transcript enrichment from notebook 04. Same
questions, different gene-level route, so the two can be read side by side.

Spearman throughout. The counts are not normal and their variance is not constant between
conditions, so a product-moment correlation has no claim on this data.

**Cross-cell-line caveat.** DHX29 enrichment is HEK293T, the eIF3d TE data is MDA-MB-231.

```{r libs}
library(tidyverse)
library(here)
```

---

## Section 1: Inputs, with the direction check

`translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv` is named in the opposite order from
the coefficient it contains. The coefficient is `sictrl_vs_si3d`, so every column is
**log2(sictrl / si3d)** and a positive `te_lfc` means eIF3d promotes.

```{r load}
enr <- readRDS(here("output", "dhx29_reanalysis", "dhx29_gene_summed_enrichment.rds"))
cat("genes from notebook 06:", format(nrow(enr), big.mark = ","), "\n")

te <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
               show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

promoted <- te %>% filter(te_lfc > 0.5, te_padj < 0.05)
eif3d_rna <- te$rna_lfc[te$symbol == "EIF3D"]

cat("EIF3D rna_lfc:", round(eif3d_rna, 3), "(must be strongly positive)\n")
cat("eIF3d-promoted genes:", nrow(promoted), "\n")

stopifnot(
  "EIF3D is not depleted by its own siRNA, so the contrast direction has flipped" =
    length(eif3d_rna) == 1 && eif3d_rna > 2,
  "the promoted set is not 1604 genes" = nrow(promoted) == 1604
)

comb <- enr %>%
  inner_join(te %>% dplyr::select(gene_id_clean, te_lfc, te_padj), by = "gene_id_clean") %>%
  filter(is.finite(lfc_simple), is.finite(te_lfc))

cat("genes shared with the TE table:", format(nrow(comb), big.mark = ","), "\n")
```

---

## Section 2: Enrichment against eIF3d TE

```{r scatter, fig.width = 7.5, fig.height = 5.5}
ct <- cor.test(comb$lfc_simple, comb$te_lfc, method = "spearman", exact = FALSE)
cat("Spearman:", round(ct$estimate, 4), " p =", signif(ct$p.value, 3), "\n\n")

cat("within expression quintiles:\n")
print(as.data.frame(
  comb %>% mutate(q = ntile(base_mean, 5)) %>%
    group_by(expr_quintile = q) %>%
    summarise(n = n(), median_total = round(median(total)),
              spearman = round(cor(lfc_simple, te_lfc, method = "spearman"), 4),
              .groups = "drop")
))

p_cor <- ggplot(comb, aes(te_lfc, lfc_simple)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_point(colour = "grey40", size = 0.5, alpha = 0.35) +
  geom_smooth(method = "lm", se = TRUE, colour = "#0072B2",
              fill = "#0072B2", alpha = 0.2, linewidth = 0.7) +
  coord_cartesian(ylim = quantile(comb$lfc_simple, c(0.005, 0.995))) +
  labs(title = "DHX29 enrichment against eIF3d-dependent TE",
       subtitle = str_wrap(paste0(
         "Summed gene counts. n = ", format(nrow(comb), big.mark = ","),
         " genes, Spearman ", sprintf("%.4f", ct$estimate),
         ", p = ", signif(ct$p.value, 3),
         ". Blue is a linear fit with its confidence band."), width = 88),
       x = "TE log2(sictrl / si3d), hypoxia 1hr. Positive means eIF3d promotes",
       y = "DHX29 log2 IP / Input") +
  theme_classic(base_size = 12)

print(p_cor)
ggsave(here("plots", "dhx29_07_te_correlation_summed.pdf"), p_cor, width = 7.5, height = 5.5)
```

---

## Section 3: Enrichment split by eIF3d gene set

```{r cdf, fig.width = 7.5, fig.height = 5}
neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  filter(!gene_id_clean %in% promoted$gene_id_clean,
         gene_id_clean %in% comb$gene_id_clean)

promoted_here <- intersect(promoted$gene_id_clean, comb$gene_id_clean)
set.seed(9)
neg_ctrl <- sample(neg_pool$gene_id_clean,
                   size = min(length(promoted_here), nrow(neg_pool)), replace = FALSE)

cat("eIF3d-promoted present:", length(promoted_here),
    "| negative pool:", nrow(neg_pool),
    "| size-matched draw:", length(neg_ctrl), "\n")

set_cols <- c("eIF3d hypoxia promotes" = "#E69F00",
              "Negative control"       = "black",
              "All other genes"        = "grey70")

set_dat <- comb %>%
  mutate(gene_set = case_when(
    gene_id_clean %in% promoted_here ~ "eIF3d hypoxia promotes",
    gene_id_clean %in% neg_ctrl      ~ "Negative control",
    TRUE                             ~ "All other genes"),
    gene_set = factor(gene_set, levels = names(set_cols)))

set_stats <- set_dat %>%
  group_by(gene_set) %>%
  summarise(n = n(), median_lfc = median(lfc_simple), .groups = "drop")
print(as.data.frame(set_stats %>% mutate(median_lfc = round(median_lfc, 5))))

pv <- set_dat$lfc_simple[set_dat$gene_set == "eIF3d hypoxia promotes"]
nv <- set_dat$lfc_simple[set_dat$gene_set == "Negative control"]
w  <- wilcox.test(pv, nv)
ks <- suppressWarnings(ks.test(pv, nv))
cat("\npromoted vs negative control: Wilcoxon p =", signif(w$p.value, 3),
    " KS p =", signif(ks$p.value, 3),
    " difference in medians =", signif(median(pv) - median(nv), 3), "\n")

stopifnot(
  "the gene sets overlap" = length(intersect(promoted_here, neg_ctrl)) == 0,
  "the negative control was not size-matched" =
    length(neg_ctrl) == min(length(promoted_here), nrow(neg_pool))
)

set_lab <- set_stats %>%
  mutate(lab = paste0(gene_set, " (n = ", format(n, big.mark = ","), ")"))

p_cdf <- ggplot(set_dat, aes(lfc_simple, colour = gene_set)) +
  stat_ecdf(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  coord_cartesian(xlim = c(-0.2, 0.2)) +
  scale_colour_manual(values = set_cols, labels = set_lab$lab, name = NULL) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("promoted vs control\nWilcoxon p = ", signif(w$p.value, 3),
                          "\nKS p = ", signif(ks$p.value, 3)),
           hjust = 1.1, vjust = 1.4, size = 3.2) +
  labs(title = "DHX29 enrichment, split by eIF3d gene set",
       x = "DHX29 log2 IP / Input", y = "Cumulative fraction") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_cdf)
ggsave(here("plots", "dhx29_07_cdf_by_geneset_summed.pdf"), p_cdf, width = 7.5, height = 5)
```

---

## Section 4: Against the representative-transcript route

```{r compare_routes}
rep_route <- readRDS(here("output", "dhx29_reanalysis", "dhx29_enrichment_gene.rds")) %>%
  dplyr::select(gene_id_clean, lfc_rep = lfc_simple)

both <- comb %>%
  dplyr::select(gene_id_clean, lfc_summed = lfc_simple) %>%
  inner_join(rep_route, by = "gene_id_clean")

cat("genes in both routes:", format(nrow(both), big.mark = ","), "\n")
cat("Spearman between the two enrichment estimates:",
    round(cor(both$lfc_summed, both$lfc_rep, method = "spearman"), 4), "\n")
```

```{r save}
write_csv(comb %>% dplyr::select(gene_id_clean, symbol, total, base_mean,
                                 lfc_simple, padj_simple, te_lfc, te_padj),
          here("output", "dhx29_reanalysis", "dhx29_07_summed_vs_te.csv"))
cat("Wrote dhx29_07_summed_vs_te.csv,", format(nrow(comb), big.mark = ","), "genes\n")
```

```{r session_info}
sessionInfo()
```
