---
title: "DHX29 reanalysis 08: SLAM-seq stability, DHX29 binding, and GC3"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Three questions on the Hia 2026 SLAM-seq half-lives (Table S6), which measure mRNA stability
in control against DHX29 knockout HEK293T cells. Positive `logFC` means stabilised on
knockout, so a message DHX29 normally destabilises.

1. Where do eIF3d-promoted genes sit on that axis? The notebook 23 figure, redrawn unfilled.
2. Where do the transcripts we measured as most and least DHX29-bound sit on it?
3. What is the GC3 of the most stabilised and most destabilised messages?

Question 2 uses our own measured enrichment from notebook 04's **transcript-level** table,
which joins natively to Table S6 since both are keyed on transcripts. Spearman throughout.

```{r libs}
library(tidyverse)
library(here)
library(readxl)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)

eif3d_cols <- c("eIF3d hypoxia promotes" = "#E69F00",
                "Negative control"       = "black",
                "All other genes"        = "grey70")
dhx29_cols <- c("DHX29 top 200"    = "#CC79A7",
                "DHX29 bottom 200" = "#0072B2",
                "All other genes"  = "grey70")
stab_cols  <- c("Most stabilized"   = "#D55E00",
                "Most destabilized" = "#0072B2",
                "All other genes"   = "grey70")
n_extreme <- 200
```

---

## Section 1: SLAM-seq half-lives

```{r load_s6}
s6 <- read_excel(here("accessories", "csc_data", "hia_dhx29",
                      "science.adw0288_tables_s1_to_s11",
                      "science.adw0288_table_s6.xlsx")) %>%
  transmute(transcript_id_clean = str_extract(gene_name, "^ENST[0-9]+"),
            ctrl_hl = as.numeric(Ctrl_Half_life),
            ko_hl   = as.numeric(KO_Half_life),
            logfc   = as.numeric(logFC)) %>%
  filter(!is.na(transcript_id_clean), is.finite(ctrl_hl), is.finite(ko_hl),
         is.finite(logfc), ctrl_hl > 0, ko_hl > 0) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

cat("transcripts with a half-life pair:", format(nrow(s6), big.mark = ","), "\n")
cat("median control half-life:", round(median(s6$ctrl_hl), 2), "h\n")
cat("median logFC:", round(median(s6$logfc), 4),
    " | fraction stabilized on KO:", sprintf("%.1f%%", 100 * mean(s6$logfc > 0)), "\n")

stopifnot(
  "logFC is not log2(KO / Ctrl)" =
    max(abs(s6$logfc - log2(s6$ko_hl / s6$ctrl_hl))) < 1e-3,
  "half-lives must be positive" = all(s6$ctrl_hl > 0 & s6$ko_hl > 0),
  "control half-life is outside a plausible range" =
    between(median(s6$ctrl_hl), 1, 20)
)
cat("Table S6 checks passed.\n")
```

---

## Section 2: eIF3d gene sets on the stability axis

The notebook 23 figure, unfilled.

```{r eif3d_density, fig.width = 7.5, fig.height = 5}
tx_gene <- readRDS(here("output", "dhx29_reanalysis", "dhx29_counts_with_gene.rds")) %>%
  distinct(transcript_id_clean, gene_id_clean) %>%
  filter(!is.na(gene_id_clean))

te <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
               show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
promoted <- te %>% filter(te_lfc > 0.5, te_padj < 0.05)
stopifnot("the promoted set is not 1604 genes" = nrow(promoted) == 1604,
          "EIF3D direction has flipped" = te$rna_lfc[te$symbol == "EIF3D"] > 2)

s6g <- s6 %>% inner_join(tx_gene, by = "transcript_id_clean")

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  filter(!gene_id_clean %in% promoted$gene_id_clean,
         gene_id_clean %in% s6g$gene_id_clean)
promoted_here <- intersect(promoted$gene_id_clean, s6g$gene_id_clean)
set.seed(9)
neg_ctrl <- sample(neg_pool$gene_id_clean,
                   size = min(length(promoted_here), nrow(neg_pool)), replace = FALSE)

d_eif3d <- s6g %>%
  mutate(grp = case_when(gene_id_clean %in% promoted_here ~ "eIF3d hypoxia promotes",
                         gene_id_clean %in% neg_ctrl      ~ "Negative control",
                         TRUE                             ~ "All other genes"),
         grp = factor(grp, levels = names(eif3d_cols)))

st <- d_eif3d %>% group_by(grp) %>%
  summarise(n = n(), med = median(logfc), .groups = "drop")
print(as.data.frame(st %>% mutate(med = round(med, 4))))
w1 <- wilcox.test(logfc ~ grp,
                  data = d_eif3d %>% filter(grp != "All other genes") %>% droplevels())
cat("promoted vs negative control: Wilcoxon p =", signif(w1$p.value, 3), "\n")

p1 <- ggplot(d_eif3d, aes(logfc, colour = grp)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = st, aes(xintercept = med, colour = grp),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("promoted vs control\nWilcoxon p = ", signif(w1$p.value, 3))) +
  coord_cartesian(xlim = c(-0.5, 0.5)) +
  scale_colour_manual(values = eif3d_cols,
                      labels = paste0(st$grp, " (n = ", format(st$n, big.mark = ","), ")"),
                      name = NULL) +
  labs(title = "mRNA stabilization upon DHX29 KO: eIF3d gene sets",
       x = "log2(KO / Ctrl) half-life ratio. Positive means stabilized on KO",
       y = "Density") +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
print(p1)
ggsave(here("plots", "dhx29_08_slamseq_by_eif3d_set.pdf"), p1, width = 7.5, height = 5)
```

---

## Section 3: Most and least DHX29-bound transcripts on the same axis

```{r dhx29_density, fig.width = 7.5, fig.height = 5}
tx_enr <- readRDS(here("output", "dhx29_reanalysis", "dhx29_enrichment_transcript.rds")) %>%
  filter(is.finite(lfc_simple)) %>%
  dplyr::select(transcript_id_clean, lfc_simple)

# Rank within the transcripts that carry BOTH measurements. Ranking over the full
# transcript table first would leave only ~27 of each extreme in the SLAM-seq set by
# chance, since 5,581 of 41,679 transcripts have a half-life.
d_dhx29 <- s6 %>%
  inner_join(tx_enr, by = "transcript_id_clean") %>%
  arrange(desc(lfc_simple)) %>%
  mutate(rank = row_number(),
         grp = case_when(rank <= n_extreme       ~ "DHX29 top 200",
                         rank >  n() - n_extreme ~ "DHX29 bottom 200",
                         TRUE                    ~ "All other genes"),
         grp = factor(grp, levels = names(dhx29_cols)))
cat("transcripts with both a half-life and an enrichment estimate:",
    format(nrow(d_dhx29), big.mark = ","), "\n")

st2 <- d_dhx29 %>% group_by(grp) %>%
  summarise(n = n(), med = median(logfc), .groups = "drop") %>% filter(n > 0)
print(as.data.frame(st2 %>% mutate(med = round(med, 4))))
cat("\nSpearman(DHX29 enrichment, half-life logFC) over all shared transcripts:",
    round(cor(d_dhx29$lfc_simple, d_dhx29$logfc, method = "spearman"), 4),
    " n =", format(nrow(d_dhx29), big.mark = ","), "\n")

ex <- d_dhx29 %>% filter(grp != "All other genes") %>% droplevels()
w2 <- if (n_distinct(ex$grp) == 2) wilcox.test(logfc ~ grp, data = ex) else NULL
if (!is.null(w2)) cat("top vs bottom 200: Wilcoxon p =", signif(w2$p.value, 3), "\n")

p2 <- ggplot(d_dhx29, aes(logfc, colour = grp)) +
  stat_ecdf(linewidth = 0.9) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  {if (!is.null(w2)) annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
        label = paste0("top vs bottom\nWilcoxon p = ", signif(w2$p.value, 3)))} +
  coord_cartesian(xlim = c(-0.5, 0.5)) +
  scale_colour_manual(values = dhx29_cols,
                      labels = paste0(st2$grp, " (n = ", format(st2$n, big.mark = ","), ")"),
                      name = NULL) +
  labs(title = "mRNA stabilization upon DHX29 KO, by measured DHX29 binding",
       x = "log2(KO / Ctrl) half-life ratio. Positive means stabilized on KO",
       y = "Cumulative fraction") +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
print(p2)
ggsave(here("plots", "dhx29_08_slamseq_by_dhx29_binding.pdf"), p2, width = 7.5, height = 5)
```

---

## Section 4: GC3 of the most stabilized and destabilized messages

```{r gc3_stability, fig.width = 7.5, fig.height = 5}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                             cdsBy(txdb, by = "tx", use.names = TRUE))
cds <- cds[width(cds) %% 3 == 0 & width(cds) >= 60]
cds <- subseq(cds, 1, width(cds) - 3)              # drop the stop codon
names(cds) <- str_remove(names(cds), "\\..*")

cn <- trinucleotideFrequency(cds, step = 3)
rownames(cn) <- names(cds)
third <- do.call(rbind, strsplit(colnames(cn), ""))[, 3]
gc3_tbl <- tibble(transcript_id_clean = rownames(cn),
                  gc3 = 100 * as.vector(cn %*% as.numeric(third %in% c("G", "C"))) /
                        rowSums(cn))

d_gc3 <- s6 %>%
  arrange(desc(logfc)) %>%
  mutate(rank = row_number(),
         grp = case_when(rank <= n_extreme       ~ "Most stabilized",
                         rank >  n() - n_extreme ~ "Most destabilized",
                         TRUE                    ~ "All other genes"),
         grp = factor(grp, levels = names(stab_cols))) %>%
  inner_join(gc3_tbl, by = "transcript_id_clean")

st3 <- d_gc3 %>% group_by(grp) %>%
  summarise(n = n(), med_gc3 = median(gc3), med_logfc = median(logfc), .groups = "drop")
print(as.data.frame(st3 %>% mutate(across(where(is.numeric), ~ round(.x, 3)))))

ex3 <- d_gc3 %>% filter(grp != "All other genes") %>% droplevels()
w3 <- wilcox.test(gc3 ~ grp, data = ex3)
cat("\nstabilized vs destabilized GC3: Wilcoxon p =", signif(w3$p.value, 3),
    " difference in medians =",
    round(st3$med_gc3[st3$grp == "Most stabilized"] -
          st3$med_gc3[st3$grp == "Most destabilized"], 2), "\n")
cat("Spearman(GC3, half-life logFC) over all transcripts:",
    round(cor(d_gc3$gc3, d_gc3$logfc, method = "spearman"), 4),
    " n =", format(nrow(d_gc3), big.mark = ","), "\n")

p3 <- ggplot(d_gc3, aes(gc3, colour = grp)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = st3, aes(xintercept = med_gc3, colour = grp),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("stabilized vs destabilized\nWilcoxon p = ",
                          signif(w3$p.value, 3))) +
  scale_colour_manual(values = stab_cols,
                      labels = paste0(st3$grp, " (n = ", format(st3$n, big.mark = ","), ")"),
                      name = NULL) +
  labs(title = "GC3 of the most stabilized and destabilized mRNAs on DHX29 KO",
       x = "GC3 (%)", y = "Density") +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
print(p3)
ggsave(here("plots", "dhx29_08_gc3_by_stability.pdf"), p3, width = 7.5, height = 5)
```

```{r save}
write_csv(d_gc3 %>% dplyr::select(transcript_id_clean, ctrl_hl, ko_hl, logfc, gc3, grp),
          here("output", "dhx29_reanalysis", "dhx29_08_stability_gc3.csv"))
cat("Wrote dhx29_08_stability_gc3.csv,", format(nrow(d_gc3), big.mark = ","), "transcripts\n")
```

```{r session_info}
sessionInfo()
```
