---
title: "DHX29 reanalysis 10: SLAM-seq half-life change by GC3 bin"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Hia et al. 2026 (Science, adw0288) show cumulative distributions of the DHX29-depletion
log-fold change split into six fixed GC3 bins, and the GC3-poor bin separates from the rest:
K562 RNA-seq on CRISPRi knockdown (Fig 1F) and HEK293T RNA-seq on knockout (Fig 1G). This
notebook draws the same figure on the paper's **stability** readout, Table S6, which is the
SLAM-seq half-life time course in control against DHX29-knockout HEK293T cells.

A few points fix what this figure is and is not:

- **The contrast is knockout, not knockdown.** Table S6 carries only `Ctrl` and `KO` arms.
  The knockdown data (Table S2) is K562 RNA-seq, a different cell line and a different
  readout, and is not used here.
- **`logFC` is log2(KO half-life / Ctrl half-life).** Positive means the message lives longer
  when DHX29 is gone, so DHX29 normally destabilises it. mRNAs *stabilised by* DHX29 are the
  negative tail. The identity is asserted against the two half-life columns at load.
- **Transcript-level.** Table S6 is keyed on the transcript whose 3'UTR was assayed, so GC3
  is computed from that same transcript's GENCODE v49 CDS. No gene-level aggregation and no
  transcript selected in a different cell line enters the join.

The reason the usual "replicate one number from the paper" check is not what guards this
notebook: the panel being redrawn *is* the claim under test, so asserting it would assert the
answer. Section 1 instead checks the structural identities the table must satisfy whatever
the biology, and Section 2 checks the GC3 computation against an independent recomputation.

```{r libs}
library(tidyverse)
library(here)
library(readxl)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)

# Six-step sequential ramp for the GC3 bins, GC3-rich to GC3-poor. Not the paper's
# red-to-blue rainbow: the project reserves #E69F00 and #56B4E9 for eIF3d and eIF3e, and a
# rainbow through those hues would read as a gene set. Anchored instead on three Okabe-Ito
# colours, vermillion -> reddish purple -> dark blue. The intermediate anchor is what keeps
# the middle bins legible; interpolating vermillion straight to blue passes through grey and
# renders 50-60% and 60-70% nearly identical.
gc3_levels <- c("> 80%", "70 ~ 80%", "60 ~ 70%", "50 ~ 60%", "40 ~ 50%", "< 40%")
gc3_colors <- setNames(colorRampPalette(c("#D55E00", "#CC79A7", "#0072B2"))(6), gc3_levels)
gc3_breaks <- c(-Inf, 40, 50, 60, 70, 80, Inf)
```

---

## Section 1: Table S6, the SLAM-seq half-lives

```{r load_s6}
s6_path <- here("accessories", "csc_data", "hia_dhx29",
                "science.adw0288_tables_s1_to_s11", "science.adw0288_table_s6.xlsx")
stopifnot("Table S6 not found" = file.exists(s6_path))

s6_raw <- read_excel(s6_path)
cat("Table S6 rows as supplied:", format(nrow(s6_raw), big.mark = ","), "\n")

s6 <- s6_raw %>%
  transmute(transcript_id_clean = str_extract(gene_name, "^ENST[0-9]+"),
            ctrl_hl = as.numeric(Ctrl_Half_life),
            ko_hl   = as.numeric(KO_Half_life),
            logfc   = as.numeric(logFC)) %>%
  filter(!is.na(transcript_id_clean), is.finite(ctrl_hl), is.finite(ko_hl),
         is.finite(logfc), ctrl_hl > 0, ko_hl > 0) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

cat("transcripts with a usable half-life pair:", format(nrow(s6), big.mark = ","), "\n")
cat("median control half-life:", round(median(s6$ctrl_hl), 2), "h",
    " | median KO half-life:", round(median(s6$ko_hl), 2), "h\n")
cat("median logFC:", round(median(s6$logfc), 4),
    " | fraction stabilized on KO:", sprintf("%.1f%%", 100 * mean(s6$logfc > 0)), "\n")
```

```{r check_s6}
# Structural identities the table must satisfy regardless of what the biology turns out to
# be. The logFC identity is the important one: it pins the direction of the column, so a
# supplement reissued with the ratio inverted would fail here rather than silently flip
# every conclusion in this notebook.
stopifnot(
  "logFC is not log2(KO / Ctrl) half-life" =
    max(abs(s6$logfc - log2(s6$ko_hl / s6$ctrl_hl))) < 1e-3,
  "half-lives must be positive" = all(s6$ctrl_hl > 0 & s6$ko_hl > 0),
  "logFC must be finite" = all(is.finite(s6$logfc)),
  "control half-life median is outside a plausible mRNA range" =
    between(median(s6$ctrl_hl), 1, 20),
  "transcript IDs must be unique" = !any(duplicated(s6$transcript_id_clean))
)

# Direction check, stated as the paper's own model: DHX29 is proposed to drive decay, so
# losing it should lengthen half-lives for most messages. This is a property of the loaded
# table, not a quoted figure from the text.
stopifnot("most messages are expected to lengthen on DHX29 KO" = median(s6$logfc) > 0)
cat("Table S6 structural checks passed.\n")
```

---

## Section 2: GC3 from the assayed transcript's CDS

```{r gc3}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds  <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                              cdsBy(txdb, by = "tx", use.names = TRUE))
cds  <- cds[width(cds) %% 3 == 0 & width(cds) >= 60]
cds  <- subseq(cds, 1, width(cds) - 3)             # drop the stop codon
names(cds) <- str_remove(names(cds), "\\..*")

cn <- trinucleotideFrequency(cds, step = 3)
# trinucleotideFrequency drops the sequence names, so row identity would silently become
# row number without this.
rownames(cn) <- names(cds)
third <- do.call(rbind, strsplit(colnames(cn), ""))[, 3]

gc3_tbl <- tibble(
  transcript_id_clean = rownames(cn),
  gc3     = 100 * as.vector(cn %*% as.numeric(third %in% c("G", "C"))) / rowSums(cn),
  cds_gc  = 100 * as.vector(letterFrequency(cds, "GC")) / width(cds),
  n_codon = rowSums(cn)
)
cat("transcripts with an in-frame CDS in v49:", format(nrow(gc3_tbl), big.mark = ","), "\n")
```

```{r check_gc3}
# Independent recomputation of GC3 on a random subset, by pulling every third base out of
# the sequence as a string. The matrix route above and this one share no code, so a wrong
# codon-position index or a misaligned frame cannot pass both.
set.seed(9)
idx <- sample(seq_along(cds), 200)
gc3_direct <- vapply(idx, function(i) {
  s  <- as.character(cds[[i]])
  p3 <- substring(s, seq(3, nchar(s), by = 3), seq(3, nchar(s), by = 3))
  100 * mean(p3 %in% c("G", "C"))
}, numeric(1))
max_dev <- max(abs(gc3_direct - gc3_tbl$gc3[idx]))
cat("max deviation, matrix GC3 vs direct string GC3 over 200 transcripts:",
    signif(max_dev, 3), "\n")

stopifnot(
  "GC3 must lie in [0, 100]"     = all(gc3_tbl$gc3    >= 0 & gc3_tbl$gc3    <= 100),
  "CDS GC must lie in [0, 100]"  = all(gc3_tbl$cds_gc >= 0 & gc3_tbl$cds_gc <= 100),
  "matrix GC3 disagrees with a direct per-codon recomputation" = max_dev < 1e-9,
  "median GC3 is outside the expected human range of 40-65%" =
    between(median(gc3_tbl$gc3), 40, 65)
)

# Biological spot-check: ACTB CDS GC is expected at 50-70% (the convention used in 01_).
tx_sym <- readRDS(here("output", "dhx29_reanalysis", "dhx29_counts_with_gene.rds")) %>%
  distinct(transcript_id_clean, symbol) %>%
  filter(!is.na(symbol))
actb_gc <- gc3_tbl %>% inner_join(tx_sym, by = "transcript_id_clean") %>%
  filter(symbol == "ACTB") %>% pull(cds_gc)
cat("ACTB CDS GC% over", length(actb_gc), "coding transcripts:",
    paste(round(range(actb_gc), 1), collapse = " to "), "\n")
stopifnot("ACTB CDS GC% is outside the expected 50-70% window" =
            length(actb_gc) > 0 && all(actb_gc >= 50 & actb_gc <= 70))
cat("GC3 checks passed.\n")
```

```{r join}
d <- s6 %>%
  inner_join(gc3_tbl, by = "transcript_id_clean") %>%
  mutate(gc3_bin = cut(gc3, breaks = gc3_breaks,
                       labels = rev(gc3_levels), right = FALSE),
         gc3_bin = factor(as.character(gc3_bin), levels = gc3_levels))

cat("transcripts with both a half-life and a GC3:", format(nrow(d), big.mark = ","),
    "of", format(nrow(s6), big.mark = ","), "\n")
cat("lost to no in-frame v49 CDS:", nrow(s6) - nrow(d), "\n")

stopifnot(
  "every transcript must fall in exactly one GC3 bin" = !any(is.na(d$gc3_bin)),
  "bin assignment must preserve the row count"        = sum(table(d$gc3_bin)) == nrow(d),
  "each bin needs enough transcripts for a stable ECDF" = all(table(d$gc3_bin) >= 50)
)
```

---

## Section 3: The figure

```{r bin_stats}
bin_stats <- d %>%
  group_by(gc3_bin) %>%
  summarise(n            = n(),
            median_gc3   = round(median(gc3), 1),
            median_logfc = round(median(logfc), 4),
            median_ctrl_hl = round(median(ctrl_hl), 2),
            median_ko_hl   = round(median(ko_hl), 2),
            .groups = "drop") %>%
  arrange(factor(gc3_bin, levels = gc3_levels))
print(as.data.frame(bin_stats))

rho_logfc <- cor(d$gc3, d$logfc, method = "spearman")
lo <- d$logfc[d$gc3_bin == "< 40%"]
hi <- d$logfc[d$gc3_bin == "> 80%"]
w_ext  <- wilcox.test(lo, hi)
ks_ext <- suppressWarnings(ks.test(lo, hi))

cat("\nSpearman(GC3, log2(KO/Ctrl) half-life) =", round(rho_logfc, 4),
    " n =", format(nrow(d), big.mark = ","), "\n")
cat("< 40% vs > 80%: Wilcoxon p =", signif(w_ext$p.value, 3),
    " | KS D =", round(as.numeric(ks_ext$statistic), 4),
    ", p =", signif(ks_ext$p.value, 3), "\n")
cat("difference in medians (<40% minus >80%) =",
    round(median(lo) - median(hi), 4), "log2 units\n")
```

```{r cdf_plot, fig.width = 6.5, fig.height = 5}
leg_labels <- paste0(bin_stats$gc3_bin, " (n=", bin_stats$n, ")")
names(leg_labels) <- as.character(bin_stats$gc3_bin)

p <- ggplot(d, aes(x = logfc, colour = gc3_bin)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60", linewidth = 0.5) +
  scale_colour_manual(values = gc3_colors, labels = leg_labels,
                      breaks = gc3_levels, name = "GC3 content") +
  scale_x_continuous(breaks = seq(-1, 1, 0.25)) +
  coord_cartesian(xlim = c(-1, 1), ylim = c(0, 1)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.6, size = 3,
           label = paste0("Spearman(GC3, logFC) = ", round(rho_logfc, 3),
                          ", n = ", format(nrow(d), big.mark = ","),
                          "\n< 40% vs > 80% Wilcoxon p = ", signif(w_ext$p.value, 3))) +
  labs(title = "HEK293T SLAM-seq: mRNA half-life change on DHX29 loss",
       subtitle = "Table S6. Positive means the message is stabilized when DHX29 is gone",
       x = "log2 (DHX29-KO / Control) half-life", y = "Proportion") +
  theme_classic(base_size = 12) +
  theme(legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = "white", colour = "grey80"),
        legend.key.height = unit(0.8, "lines"))
print(p)

ggsave(here("plots", "dhx29_10_slamseq_logfc_cdf_by_gc3.pdf"), p, width = 6.5, height = 5)
cat("Saved: plots/dhx29_10_slamseq_logfc_cdf_by_gc3.pdf\n")
```

---

## Section 4: Is there any GC3 structure in this table at all?

If the bins separate on the *change*, they should normally also separate on baseline
stability, since GC3 tracks codon optimality. Testing the control arm on its own says
whether the flat result above is specific to the DHX29 contrast or whether this half-life
table carries no GC3 signal of any kind. Nothing here is a claim about the paper's RNA-seq
panels, which are a different readout in a different experiment.

```{r baseline, fig.width = 6.5, fig.height = 5}
rho_ctrl <- cor(d$gc3, d$ctrl_hl, method = "spearman")
rho_ko   <- cor(d$gc3, d$ko_hl,   method = "spearman")
cat("Spearman(GC3, control half-life) =", round(rho_ctrl, 4), "\n")
cat("Spearman(GC3, KO half-life)      =", round(rho_ko, 4), "\n")
cat("n =", format(nrow(d), big.mark = ","), "\n")

p_base <- ggplot(d, aes(x = ctrl_hl, colour = gc3_bin)) +
  stat_ecdf(linewidth = 0.8) +
  scale_colour_manual(values = gc3_colors, labels = leg_labels,
                      breaks = gc3_levels, name = "GC3 content") +
  coord_cartesian(xlim = quantile(d$ctrl_hl, c(0, 0.99)), ylim = c(0, 1)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.6, size = 3,
           label = paste0("Spearman(GC3, half-life) = ", round(rho_ctrl, 3),
                          ", n = ", format(nrow(d), big.mark = ","))) +
  labs(title = "Baseline mRNA stability by GC3 bin, control cells",
       subtitle = "Same transcripts and bins as above, control arm only",
       x = "Control half-life (h)", y = "Proportion") +
  theme_classic(base_size = 12) +
  theme(legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = "white", colour = "grey80"),
        legend.key.height = unit(0.8, "lines"))
print(p_base)

ggsave(here("plots", "dhx29_10_slamseq_baseline_halflife_cdf_by_gc3.pdf"),
       p_base, width = 6.5, height = 5)
cat("Saved: plots/dhx29_10_slamseq_baseline_halflife_cdf_by_gc3.pdf\n")
```

---

## Section 5: Save

```{r save}
out <- d %>%
  dplyr::select(transcript_id_clean, ctrl_hl, ko_hl, logfc, gc3, cds_gc, n_codon, gc3_bin)
write_csv(out, here("output", "dhx29_reanalysis", "dhx29_10_slamseq_gc3_bins.csv"))
write_csv(bin_stats, here("output", "dhx29_reanalysis", "dhx29_10_gc3_bin_summary.csv"))
cat("Wrote dhx29_10_slamseq_gc3_bins.csv,", format(nrow(out), big.mark = ","), "transcripts\n")
cat("Wrote dhx29_10_gc3_bin_summary.csv\n")
```

```{r session_info}
sessionInfo()
```
