---
title: "Feature reconciliation: this repo vs 2026_TranslationResponseFeatureAnalysis"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Audit only. This notebook computes **no new features** and writes **no feature matrix**. It
compares our feature definitions against the collaborator repo
`mukherjeelab/2026_TranslationResponseFeatureAnalysis` and produces the discrepancy table that
drives `01_build_feature_matrix.Rmd`.

Two levels of comparison:

1. **Name level** - which features exist on each side.
2. **Value level** - for quantities both sides compute, recompute ours under *their* definition
   and measure the disagreement. A feature that is nominally shared but disagrees numerically is
   the failure a name-level audit cannot see.

The collaborator pipeline has never been run end to end, so there are no computed values to
pull. What we take from it is the **published reference tables** (checked into their repo,
mirrored here under `accessories/collaborator_reference/`) and their scoring logic.

```{r libraries}
library(tidyverse)
library(here)
```

```{r load_data}
features <- readRDS(here("output", "predictive_modeling",
                         "feature_matrix_dhx29_kd_feature.rds"))
seqs     <- readRDS(here("output", "predictive_modeling",
                         "precomputed_kozak_sequences.rds"))

ref_dir <- here("accessories", "collaborator_reference")

cat("Feature matrix:", nrow(features), "transcripts x", ncol(features), "columns\n")

# Restrict sequences to the transcripts actually in the feature matrix, so every
# number below describes the modelled set rather than the full annotation.
utr5 <- seqs$utr5_seq %>% filter(transcript_id_clean %in% features$transcript_id_clean)
cds  <- seqs$cds_seq  %>% filter(transcript_id_clean %in% features$transcript_id_clean)

stopifnot(
  "not every feature-matrix transcript has a 5'UTR sequence" =
    nrow(utr5) == nrow(features),
  "not every feature-matrix transcript has a CDS sequence" =
    nrow(cds) == nrow(features)
)
cat("Sequences available for all", nrow(cds), "modelled transcripts\n")
```

# 1. Reference tables: provenance and replication

Each external table is checked against a published quantity **before** it is used anywhere,
per the project convention for incorporating data from papers.

```{r reference_tables}
kozak_pub <- read_tsv(file.path(ref_dir, "kozak_1987_table1_frequencies.tsv"),
                      show_col_types = FALSE)
noderer   <- read_tsv(file.path(ref_dir, "noderer_2014_tis_efficiency.tsv"),
                      skip = 1, show_col_types = FALSE)   # line 1 is the table caption
cridge    <- read_tsv(file.path(ref_dir, "cridge2018_termination_contexts.tsv"),
                      show_col_types = FALSE)

# Kozak 1987 (NAR 15:8125) Table 1 - 7 variable positions, frequencies sum to 1 per position
stopifnot(
  "Kozak table is not 7 positions"        = nrow(kozak_pub) == 7,
  "Kozak frequencies do not sum to 1"     =
    all(abs(rowSums(kozak_pub[, c("A_frequency","C_frequency",
                                  "G_frequency","T_frequency")]) - 1) < 1e-8)
)
cat("Kozak 1987: 7 positions, all rows sum to 1.00\n")

# Noderer 2014 (MSB 10:748) Table S2 - all 65,536 -6..+5 contexts, efficiency 12-150
stopifnot(
  "Noderer table is not 65,536 contexts"  = nrow(noderer) == 65536L,
  "Noderer contexts are not 11 nt"        = all(nchar(noderer$sequence) == 11L),
  "Noderer efficiency outside 12-150"     =
    all(noderer$efficiency >= 12 & noderer$efficiency <= 150)
)
cat("Noderer 2014:", nrow(noderer), "contexts, efficiency range",
    paste(range(noderer$efficiency), collapse = "-"), "\n")

# Cridge 2018 (NAR 46:1927) Table S1 - 3 stops x 64 triplets, paper median ~0.087%
cridge_median <- median(cridge$readthrough_measure)
stopifnot(
  "Cridge table is not 192 contexts"      = nrow(cridge) == 192L,
  "Cridge median readthrough off"         = between(cridge_median, 0.08, 0.095),
  "Cridge readthrough not a percentage"   =
    all(cridge$readthrough_measure > 0 & cridge$readthrough_measure < 5)
)
cat("Cridge 2018:", nrow(cridge), "contexts, median readthrough",
    round(cridge_median, 4), "% (paper: ~0.087%)\n")
```

# 2. Kozak PWM: a confirmed error in `01h_`

`01h_feature_extraction_kozak.Rmd` builds its position frequency matrix by hand. Comparing it
against the published Kozak 1987 Table 1 **and** against the base frequencies observed in our
own transcripts gives an independent, three-way check.

```{r kozak_windows}
# Rebuild the Kozak window exactly as 01h_ does: last 6 nt of 5'UTR + first 4 nt of CDS.
kozak_win <- inner_join(utr5, cds, by = "transcript_id_clean") %>%
  mutate(
    utr5_len  = nchar(utr5_sequence),
    utr5_tail = str_sub(utr5_sequence, start = pmax(1L, utr5_len - 5L)),
    cds_head  = str_sub(cds_sequence, 1, 4),
    window    = paste0(utr5_tail, cds_head)
  ) %>%
  filter(nchar(window) == 10L)

# 01h_ never checks that the window is anchored on a real AUG. Transcripts whose annotated
# CDS does not begin with ATG still receive a score, on a window that is not a start context.
kozak_win <- kozak_win %>% mutate(anchored = str_sub(window, 7, 9) == "ATG")

cat("10-nt windows:", nrow(kozak_win), "\n")
cat("NOT anchored on ATG:", sum(!kozak_win$anchored),
    sprintf("(%.2f%%) - these should be NA, 01h_ scores them anyway\n",
            100 * mean(!kozak_win$anchored)))

kozak_ok <- kozak_win %>% filter(anchored)
```

```{r kozak_pwm_three_way}
# Window index of each scored position: -6..-1 are chars 1-6, +4 is char 10.
win_idx <- c("-6" = 1, "-5" = 2, "-4" = 3, "-3" = 4, "-2" = 5, "-1" = 6, "+4" = 10)

# Parse the in-use matrix straight out of 01h_'s source rather than restating it here.
# Re-typing a matrix by hand is what produced the defect this section documents, so the
# audit must not repeat it. Section 2b then proves the parse is faithful.
kozak_src <- read_lines(here("code", "predictive_modeling",
                             "01h_feature_extraction_kozak.Rmd"))
mat_start <- grep("^kozak_pwm <- matrix\\(", kozak_src)
stopifnot("could not locate the kozak_pwm matrix in 01h_" = length(mat_start) == 1)
mat_end   <- grep("^\\)$|byrow = TRUE", kozak_src)
mat_end   <- min(mat_end[mat_end > mat_start])

pwm_nums <- kozak_src[mat_start:mat_end] %>%
  str_remove("#.*$") %>%                       # drop the trailing position comments
  str_extract_all("[0-9]*\\.[0-9]+") %>%
  unlist() %>%
  as.numeric()

stopifnot("expected 28 frequencies (7 positions x 4 bases) in 01h_" =
            length(pwm_nums) == 28L)

pwm_ours <- matrix(pwm_nums, nrow = 7, byrow = TRUE,
  dimnames = list(c("-6","-5","-4","-3","-2","-1","+4"), c("A","C","G","T")))

stopifnot("parsed 01h_ matrix rows do not sum to 1" =
            all(abs(rowSums(pwm_ours) - 1) < 1e-8))
cat("Parsed the in-use PWM from 01h_ source; all 7 rows sum to 1.00\n")

# Key on context_index (1-6 and 10), which is the window position directly. The `position`
# column reads as a number ("+4" -> 4), so it is not safe as a row name.
stopifnot("unexpected Kozak context_index set" =
  setequal(kozak_pub$context_index, c(1:6, 10)))

pwm_publ <- kozak_pub %>%
  arrange(match(context_index, win_idx)) %>%
  dplyr::select(A = A_frequency, C = C_frequency,
                G = G_frequency, T = T_frequency) %>%
  as.matrix()
rownames(pwm_publ) <- names(win_idx)
stopifnot("published PWM rows misaligned" =
  identical(rownames(pwm_publ), rownames(pwm_ours)))

# Base frequencies actually observed at each window position in our own transcripts
pwm_obs <- t(sapply(win_idx, function(i) {
  tab <- table(factor(str_sub(kozak_ok$window, i, i), levels = c("A","C","G","T")))
  as.numeric(tab) / sum(tab)
}))
colnames(pwm_obs) <- c("A","C","G","T")

three_way <- map_dfr(rownames(pwm_ours), function(p) {
  tibble(position = p, base = c("A","C","G","T"),
         ours = pwm_ours[p, ], published = pwm_publ[p, ], observed = pwm_obs[p, ])
})

cat("Argmax base per position (ours / published / observed in our data):\n")
three_way %>%
  group_by(position) %>%
  summarise(ours      = base[which.max(ours)],
            published = base[which.max(published)],
            observed  = base[which.max(observed)],
            .groups = "drop") %>%
  mutate(position = factor(position, levels = rownames(pwm_ours))) %>%
  arrange(position) %>%
  mutate(ours_agrees = ours == published) %>%
  print()
```

## 2b. Is the parsed matrix the one actually in use?

The parse above is only trustworthy if scoring with it reproduces the `kozak_pwm_score` column
that `01h_` already wrote to disk. This closes the loop: nothing about the in-use matrix is
taken on trust, and a mis-parse fails here rather than propagating into the conclusions.

```{r kozak_parse_verification}
lo <- function(m) log2((m + 0.001) / 0.25)    # 01h_'s pseudocount

score_pwm <- function(windows, pwm) {
  logodds <- lo(pwm)
  mat <- str_split_fixed(windows, "", 10)
  rowSums(sapply(seq_along(win_idx), function(k) {
    logodds[names(win_idx)[k], mat[, win_idx[k]]]
  }))
}

saved_kozak <- readRDS(here("output", "predictive_modeling", "feature_matrix_kozak.rds")) %>%
  dplyr::select(transcript_id_clean, kozak_pwm_score)

# 01h_ scores every window of >= 10 nt, anchored on AUG or not, so verify against all of them.
verify <- kozak_win %>%
  mutate(rescored = score_pwm(window, pwm_ours)) %>%
  inner_join(saved_kozak, by = "transcript_id_clean") %>%
  filter(!is.na(kozak_pwm_score))

max_dev <- max(abs(verify$rescored - verify$kozak_pwm_score))
cat("Rescored vs saved kozak_pwm_score: n =", nrow(verify),
    ", max |deviation| =", signif(max_dev, 3), "\n")

stopifnot("parsed PWM does not reproduce the saved kozak_pwm_score column" =
            max_dev < 1e-8)
cat("Verified: the parsed matrix is the one that produced the feature on disk\n")
```

The published table and the frequencies observed in our own transcripts agree at every
position. Our hand-typed matrix does not.

```{r kozak_minus3}
cat("\nPosition -3, the most informative position:\n")
three_way %>% filter(position == "-3") %>%
  dplyr::select(base, ours, published, observed) %>% print()

gap_ours <- lo(pwm_ours)["-3","A"] - lo(pwm_ours)["-3","G"]
gap_publ <- lo(pwm_publ)["-3","A"] - lo(pwm_publ)["-3","G"]

n_g3 <- sum(str_sub(kozak_ok$window, 4, 4) == "G")
cat(sprintf("\nA-vs-G log-odds gap at -3:  ours %.2f bits,  published %.2f bits\n",
            gap_ours, gap_publ))
cat(sprintf("Transcripts with G at -3: %d (%.1f%%)\n",
            n_g3, 100 * n_g3 / nrow(kozak_ok)))
```

Position -3 is the **purine** position - the R in the GCC**R**CCATGG consensus, where A and G
are both favourable. `01h_`'s own `match_kozak()` encodes this correctly, scoring `[AG]` as
equivalent. Its PWM does not: it penalises G at -3 while rewarding A, so the two Kozak features
in the same notebook contradict each other at the position that matters most.

```{r kozak_impact}
# Rescore every window under both matrices and measure the damage.
# score_pwm() and lo() are defined in the verification chunk above.

# 01h_'s ordinal tier, reproduced so the within-tier check below is exact
match_kozak <- function(seq) {
  if (grepl("^(GCCACC|GCCGCC)ATGG$", seq)) return(3L)
  if (grepl("^...[AG]..ATGG$", seq))       return(2L)
  if (grepl("^...[AG]..ATG[ACT]$", seq))   return(1L)
  if (grepl("^...[CT]..ATGG$", seq))       return(1L)
  0L
}

kozak_cmp <- kozak_ok %>%
  mutate(
    pwm_ours  = score_pwm(window, pwm_ours),
    pwm_publ  = score_pwm(window, pwm_publ),
    tier      = map_int(window, match_kozak),
    base_m3   = str_sub(window, 4, 4)
  )

cat("Correlation between the two scorings: r =",
    round(cor(kozak_cmp$pwm_ours, kozak_cmp$pwm_publ), 3),
    ", n =", nrow(kozak_cmp), "\n")

# The check that would have caught this. The existing tier-monotonicity check passes under
# BOTH matrices, because tiers are set mostly by +4 and by [CT]-vs-[AG]: a within-tier error
# at -3 cancels in the tier means. Comparing A-at--3 against G-at--3 INSIDE a tier does not.
cat("\nWithin tier 2 ('strong' = [AG] at -3, G at +4), mean PWM by base at -3:\n")
within_tier <- kozak_cmp %>%
  filter(tier == 2L, base_m3 %in% c("A", "G")) %>%
  group_by(base_m3) %>%
  summarise(n = n(), mean_ours = mean(pwm_ours),
            mean_published = mean(pwm_publ), .groups = "drop")
print(within_tier)

spread_ours <- diff(rev(within_tier$mean_ours))
spread_publ <- diff(rev(within_tier$mean_published))
cat(sprintf("\nA-minus-G spread within one tier: ours %.2f bits, published %.2f bits\n",
            abs(spread_ours), abs(spread_publ)))
cat("Both matrices pass the existing 'mean PWM increases tier 0->3' check;\n")
cat("only the within-tier comparison separates them.\n")
```

```{r kozak_plot, fig.width = 7, fig.height = 6}
p_kozak <- three_way %>%
  mutate(position = factor(position, levels = rownames(pwm_ours))) %>%
  pivot_longer(c(ours, published, observed),
               names_to = "source", values_to = "freq") %>%
  mutate(source = factor(source, levels = c("ours", "published", "observed"),
                         labels = c("01h_ hand-typed", "Kozak 1987 Table 1",
                                    "observed in our transcripts"))) %>%
  ggplot(aes(x = base, y = freq, fill = source)) +
  geom_col(position = "dodge") +
  facet_wrap(~ position, nrow = 2) +
  scale_fill_manual(values = c("01h_ hand-typed" = "#D55E00",
                               "Kozak 1987 Table 1" = "#0072B2",
                               "observed in our transcripts" = "grey70"),
                    name = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom") +
  labs(title = "Kozak position frequency matrix: hand-typed vs published vs observed",
       subtitle = paste0("n = ", nrow(kozak_ok), " AUG-anchored windows"),
       x = "Base", y = "Frequency")

print(p_kozak)
ggsave(here("plots", "00_kozak_pwm_three_way.pdf"), p_kozak, width = 7, height = 6)
```

**Resolution:** replace the hand-typed matrix with `kozak_1987_table1_frequencies.tsv`, NA the
windows that are not AUG-anchored, and add the within-tier purine check to `01_`.

# 3. GC content: units and denominator

```{r gc_audit}
gc_ours <- function(seq) {                       # 01h_/01_ definition
  seq <- str_to_upper(seq)
  100 * str_count(seq, "[GC]") / str_length(seq) # denominator includes ambiguous bases
}
gc_theirs <- function(seq) {                     # collaborator definition
  seq <- str_to_upper(seq)
  canonical <- str_count(seq, "[ACGT]")          # ambiguous bases excluded
  str_count(seq, "[GC]") / canonical             # fraction, not percent
}

gc_cmp <- cds %>%
  transmute(transcript_id_clean,
            ours   = gc_ours(cds_sequence),
            theirs = gc_theirs(cds_sequence),
            theirs_as_pct = 100 * theirs,
            n_ambiguous   = str_count(str_to_upper(cds_sequence), "[^ACGT]"))

cat("CDS GC, n =", nrow(gc_cmp), "\n")
cat("  transcripts containing an ambiguous base:", sum(gc_cmp$n_ambiguous > 0), "\n")
cat("  max absolute difference once units are matched:",
    signif(max(abs(gc_cmp$ours - gc_cmp$theirs_as_pct)), 3), "percentage points\n")
cat("  Pearson r =", round(cor(gc_cmp$ours, gc_cmp$theirs_as_pct), 6), "\n")
```

The denominator difference is immaterial in this annotation: GENCODE v49 CDS sequences are
essentially free of ambiguous bases. **The units are not immaterial.** Ours is a percentage and
theirs is a fraction, a factor of 100 that would silently corrupt any shared threshold.
Keep percent (the project validation convention asserts GC in `[0, 100]`), and record the unit
explicitly in `feature_annotation.csv`.

# 4. GC3: the stop codon

```{r gc3_audit}
gc3_ours <- function(seqs) {          # 01_ definition: every codon, stop included
  vapply(seqs, function(s) {
    n <- nchar(s) %/% 3L
    if (n < 1L) return(NA_real_)
    starts <- seq.int(3L, n * 3L, 3L)
    sum(substr(rep.int(s, n), starts, starts) %in% c("G", "C")) * 100 / n
  }, numeric(1L), USE.NAMES = FALSE)
}
gc3_theirs <- function(seqs) {        # collaborator definition: sense codons only
  vapply(seqs, function(s) {
    n <- nchar(s) %/% 3L
    if (n < 2L) return(NA_real_)
    last <- substr(s, (n - 1L) * 3L + 1L, n * 3L)
    n_sense <- if (last %in% c("TAA", "TAG", "TGA")) n - 1L else n
    starts <- seq.int(3L, n_sense * 3L, 3L)
    sum(substr(rep.int(s, n_sense), starts, starts) %in% c("G", "C")) / n_sense
  }, numeric(1L), USE.NAMES = FALSE)
}

gc3_cmp <- cds %>%
  transmute(transcript_id_clean,
            ours   = gc3_ours(str_to_upper(cds_sequence)),
            theirs = 100 * gc3_theirs(str_to_upper(cds_sequence))) %>%
  mutate(diff = ours - theirs)

cat("GC3, n =", sum(!is.na(gc3_cmp$diff)), "\n")
cat("  Pearson r =", round(cor(gc3_cmp$ours, gc3_cmp$theirs, use = "complete.obs"), 6), "\n")
cat("  median difference (ours - theirs):", round(median(gc3_cmp$diff, na.rm = TRUE), 4),
    "percentage points\n")
cat("  max |difference|:", round(max(abs(gc3_cmp$diff), na.rm = TRUE), 3), "\n")
```

Stop codons are TAA / TAG / TGA, so their third base is A, G or A. Including the stop therefore
pulls GC3 down by a predictable amount that scales as `1 / n_codons` - largest for short CDSs,
which is where it is least affordable. The effect is small but is a **bias, not noise**: it has a
sign. Adopt the sense-codon definition and keep percent units.

```{r gc3_plot, fig.width = 6, fig.height = 5}
p_gc3 <- ggplot(gc3_cmp, aes(x = theirs, y = ours)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.15, size = 0.6, colour = "grey30") +
  theme_classic(base_size = 12) +
  labs(title = "GC3: stop codon included vs excluded",
       subtitle = sprintf("Pearson r = %.5f, n = %d",
                          cor(gc3_cmp$ours, gc3_cmp$theirs, use = "complete.obs"),
                          sum(!is.na(gc3_cmp$diff))),
       x = "GC3, sense codons only (%)", y = "GC3, stop codon included (%)")

print(p_gc3)
ggsave(here("plots", "00_gc3_stop_codon_definition.pdf"), p_gc3, width = 6, height = 5)
```

# 5. log2 length

```{r log2_length_audit}
len_cmp <- features %>%
  filter(!is.na(cds_length), cds_length > 0) %>%
  transmute(cds_length,
            ours   = log2(cds_length + 1),   # our convention, applied to all three regions
            theirs = log2(cds_length))       # theirs, for CDS only

cat("log2 CDS length, n =", nrow(len_cmp), "\n")
cat("  max |difference|:", signif(max(abs(len_cmp$ours - len_cmp$theirs)), 3), "\n")
cat("  shortest CDS in the matrix:", min(len_cmp$cds_length), "nt\n")
```

Numerically negligible, and our convention is the better one: `+1` is applied uniformly to
5'UTR, CDS and 3'UTR, whereas theirs uses `+1` for the UTRs but not the CDS. The `+1` is what
makes a zero-length UTR representable at all, so dropping it only for the CDS buys nothing and
costs internal consistency. **Keep ours.**

# 6. RNA structure: a name collision, not a shared feature

Both sides emit columns named `struct_accessibility_*` from RNAplfold with **identical window
parameters** (`-W 150 -L 100`). They are nonetheless different quantities:

| | ours | theirs |
|---|---|---|
| Folding unit | each region folded in isolation | whole spliced transcript, folded once |
| Retained | unpaired probability per region, plus min / sd / structured-region counts | mean unpaired probability per region, plus low-accessibility window fractions |

Because `-W 150` caps base pairing at a 150-nt span either way, the two can only diverge
**within 150 nt of a region boundary** - a rounding error for long CDSs. Not so for 5'UTRs,
which are frequently shorter than 200 nt: there the whole region sits inside the boundary zone
and isolated folding imposes an artificial free 3' end exactly at the start codon.

**Resolution:** keep ours (the richer summary), rename to `tco_struct_*` so the columns can
never be silently equated with theirs, and record the 5'UTR caveat. No refolding. If
`tco_struct_accessibility_cap_proximal` or `..._aug_context` reaches the top of a SHAP ranking,
revisit before interpreting it.

# 7. Inventory

```{r inventory}
inventory <- tribble(
  ~feature_ours,              ~feature_theirs,                   ~quantity,                  ~status,              ~recommendation,
  "kozak_pwm_score",          "kozak1987_logodds_score",          "Kozak context log-odds",   "definition_differs", "FIX: adopt published Kozak 1987 Table 1; current -3 and -6 rows are wrong",
  "kozak_score",              "kozak_context_class",              "Kozak ordinal tier",       "match",              "keep ours; NA the non-AUG-anchored windows",
  NA,                         "noderer2014_tis_efficiency",       "measured TIS efficiency",  "theirs_only",        "ADOPT; same family as Kozak, never model both together",
  "gc3",                      "gc3_fraction",                     "GC at codon position 3",   "definition_differs", "FIX: exclude the stop codon; keep percent units",
  "cds_gc",                   "cds_gc",                           "CDS GC content",           "definition_differs", "units only (percent vs fraction); keep percent, record the unit",
  "utr5_gc",                  "utr5_gc",                          "5'UTR GC content",         "definition_differs", "as cds_gc",
  "utr3_gc",                  "utr3_gc",                          "3'UTR GC content",         "definition_differs", "as cds_gc",
  "transcript_gc",            "transcript_gc",                    "transcript GC content",    "definition_differs", "ours is a length-weighted mean of regions; theirs is direct. Keep ours, document",
  "log2_cds_length",          "log2_cds_length",                  "log2 CDS length",          "definition_differs", "keep ours (+1 applied uniformly); theirs is internally asymmetric",
  "log2_utr5_length",         "log2_utr5_length",                 "log2 5'UTR length",        "match",              "keep ours",
  "log2_utr3_length",         "log2_utr3_length",                 "log2 3'UTR length",        "match",              "keep ours",
  "struct_accessibility_*",   "struct_accessibility_*",           "RNAplfold accessibility",  "definition_differs", "NAME COLLISION: rename ours to tco_struct_*; keep ours; do not refold",
  "csc",                      "icodon_human_stability_score",     "iCodon codon stability",   "match",              "same tool; verify by value in 01_",
  "csc_q1..q4, csc_first75",  NA,                                 "positional CSC",           "ours_only",          "DROP: collapses the worst correlated block",
  "cai",                      NA,                                 "codon adaptation index",   "ours_only",          "keep, whole-CDS only",
  "fop",                      NA,                                 "fraction optimal codons",  "ours_only",          "keep, whole-CDS only",
  "tai",                      "tai_genomic_internal",             "tRNA adaptation index",    "definition_differs", "ADOPT theirs: explicit wobble table + GtRNAdb gene copy number; keep ours alongside for one release to measure the shift",
  NA,                         "termination_context_minus9_plus9", "stop codon context",       "theirs_only",        "ADOPT: new family, nothing equivalent here",
  NA,                         "cridge2018_readthrough_plus4_to_plus6", "readthrough propensity", "theirs_only",   "ADOPT: 192-context published lookup",
  NA,                         "proline_fraction, PPP_motif_density, max_net_charge_30aa, ...", "nascent peptide", "theirs_only", "ADOPT: pure sequence, no external data",
  NA,                         "max_consecutive_AAA_lys_codons / AAG",  "CDS poly(A) tracks",  "theirs_only",        "ADOPT: keep AAA and AAG separate by design",
  NA,                         "control_ribo_frame0_fraction, control_ribo_cds_coverage_fraction, ...", "detectability", "theirs_only", "ADOPT as a separate 'technical' block",
  "g4mer_*",                  NA,                                 "G-quadruplex propensity",  "ours_only",          "MERGE: already built, currently unused",
  "clip_*",                   NA,                                 "eIF3 CLIP binding",        "ours_only",          "keep as 'external' block",
  "zhu2024_*, hia2026_*, karner2026_*", NA,                        "SLAM-seq / perturbation", "ours_only",          "keep as 'external' block"
)

out_dir <- here("output", "translational_control_overview")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
write_csv(inventory, file.path(out_dir, "feature_inventory_comparison.csv"))

cat("Inventory rows:", nrow(inventory), "\n\n")
inventory %>% count(status, name = "n") %>% arrange(desc(n)) %>% print()

stopifnot(
  "unrecognised status value" =
    all(inventory$status %in% c("match", "definition_differs", "ours_only", "theirs_only")),
  "a row has neither side named" =
    all(!(is.na(inventory$feature_ours) & is.na(inventory$feature_theirs)))
)
cat("\nWritten to", file.path(out_dir, "feature_inventory_comparison.csv"), "\n")
```

# Summary

**One real bug.** `kozak_pwm_score` uses a hand-typed matrix that is wrong at -3 and -6.
Position -3 is the purine position; the published table gives G a frequency of 0.36 and our own
transcripts show 0.36, but `01h_` has 0.13. The result penalises G at -3 by 0.93 bits while
`match_kozak()` in the same notebook treats it as favourable - the two Kozak features
contradict each other for roughly a third of transcripts. The existing tier-monotonicity check
passes under both matrices and cannot detect this; a within-tier purine comparison can, and is
added to `01_`.

**Three definition fixes**, all small and all signed rather than random: GC3 should exclude the
stop codon, GC units must be recorded as percent, and non-AUG-anchored Kozak windows should be
NA rather than scored.

**Two places where ours is better** and stays: `log2_*_length` (uniform `+1`) and the structure
features (richer per-region summary). The structure columns are renamed `tco_struct_*` because
the names collide with a different quantity.

**Five families to adopt** and one to merge, listed in the inventory.
