---
title: "Build the translational-control feature matrix"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  pass_through: FALSE
  mutate_input: ""
  apply_audit_repairs: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Single auditable builder for `code/translational_control_overview/`. Replaces the nine-notebook
chain (`01_` -> `01b_` -> `01d_` -> `22_` -> `23_` -> `01i_` -> `01j_` -> `01k_` -> `87_`), two
of which overwrite `feature_matrix_external_stability.rds` in place. Nothing under
`output/predictive_modeling/` is read-modify-written here; this notebook only reads.

Two artifacts:

| Output | Role |
|---|---|
| `feature_matrix_tco.rds` | the matrix, keyed on `transcript_id_clean` |
| `feature_annotation.csv` | one row per column: `block`, `family`, `source`, `definition` |

**`feature_annotation.csv` is the central artifact.** It replaces notebook 07's eight
`if (feature_set == ...)` branches and its regex cascade with an explicit, inspectable table.
`02_rf_model.Rmd` selects features by joining against it (`include_blocks`), and `03_shap.Rmd`
aggregates SHAP by `family` - which is exact, because SHAP is additive.

## `pass_through`

With `params$pass_through = TRUE` the matrix is copied from the baseline **unchanged** and only
the annotation is new. That is deliberate and is the first thing to run: it lets
`02_rf_model.Rmd` be checked against the frozen baseline before any definition is corrected or
any feature added. Until that reproduction gate passes, no later difference is interpretable -
a fork that silently fails to rebuild the baseline looks exactly like a real feature effect.

Definition fixes (Kozak PWM, GC3 stop codon), the `tco_struct_*` rename and the new feature
families land in later passes, with `pass_through = FALSE`.

```{r libraries}
library(tidyverse)
library(here)
```

```{r load_baseline}
baseline_rds <- "feature_matrix_dhx29_kd_feature.rds"
features <- readRDS(here("output", "predictive_modeling", baseline_rds))

out_dir <- here("output", "translational_control_overview")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Recorded before anything is added, so the gate can always tell baseline-era columns from
# ones this notebook introduced. Self-maintaining: every new family is "tco" automatically.
baseline_cols <- colnames(features)

cat("Baseline matrix:", baseline_rds, "\n")
cat(nrow(features), "transcripts x", ncol(features), "columns\n")

stopifnot(
  "matrix is not keyed on transcript_id_clean" =
    "transcript_id_clean" %in% colnames(features),
  "transcript_id_clean is not unique" =
    !any(duplicated(features$transcript_id_clean)),
  "gene identifier trio is incomplete" =
    all(c("gene_id", "gene_id_clean", "symbol") %in% colnames(features))
)
```

# 1. Corrected feature variants

Three columns are known to be wrong or superseded (audited in `00_`). Each correction is added
as a **parallel column** rather than overwriting the original, so `04_` can measure how much
each one moves the ranking instead of only showing the endpoint. `02_rf_model.Rmd` selects one
of each pair via `variant_set`; the two members of a pair are near-duplicates and must never
enter the same forest, where they would split credit and hide the effect being measured.

| current | corrected | why |
|---|---|---|
| `kozak_pwm_score` | `kozak_pwm_score_v2` | hand-typed PWM is wrong at -6 and -3 |
| `gc3` | `gc3_internal` | initiator and stop codon were both counted |
| `cai` | `cai_internal` | initiator AUG was counted |
| `fop` | `fop_internal` | initiator AUG was counted |
| `csc` | `csc_internal` | iCodon was given the stop-containing CDS |
| `tai` | `tai_gtrnadb` | collaborator variant: explicit wobble table + GtRNAdb copy number |

```{r corrected_features, eval = !params$pass_through}
ref_dir <- here("accessories", "collaborator_reference")
seqs <- readRDS(here("output", "predictive_modeling", "precomputed_kozak_sequences.rds"))

utr5 <- seqs$utr5_seq %>% filter(transcript_id_clean %in% features$transcript_id_clean)
cds  <- seqs$cds_seq  %>% filter(transcript_id_clean %in% features$transcript_id_clean)
stopifnot("sequences missing for some transcripts" =
            nrow(utr5) == nrow(features) && nrow(cds) == nrow(features))

## --- deliberate corruption, for 01c_mutation_test_checks.R --------------------------
# A check that has never been shown to FAIL is uninformative: the 01h_ Kozak PWM passed its
# own tier-monotonicity check while being wrong. mutate_input applies one named corruption to
# an INPUT, so 01c_ can assert that the corresponding check fires. Mutations that slip through
# are recorded as blind spots rather than quietly ignored.
#
# Guard: a mutation run never writes feature_matrix_tco.rds (see the write chunk).
MUT <- params$mutate_input
KNOWN_MUTATIONS <- c("", "noderer_window_shift", "cridge_permute", "term_stop_shuffle",
                     "peptide_frame_shift", "g4_region_swap", "kozak_minus3_bug",
                     # one per defect class the independent audit found
                     "struct_zero_fill", "gene_level_join", "duplicate_feature",
                     "constant_feature", "length_proxy_feature", "golden_drift")
stopifnot("unknown mutate_input" = MUT %in% KNOWN_MUTATIONS)
if (nzchar(MUT)) {
  cat("\n*** MUTATION RUN:", MUT, "- outputs will NOT be written ***\n\n")
  set.seed(9)
}

## --- kozak_pwm_score_v2 -------------------------------------------------------------
# Published Kozak 1987 Table 1, imported as a file. Never transcribed: a hand-typed copy is
# what produced the defect this replaces.
kozak_pub <- read_tsv(file.path(ref_dir, "kozak_1987_table1_frequencies.tsv"),
                      show_col_types = FALSE)
win_idx <- c("-6" = 1, "-5" = 2, "-4" = 3, "-3" = 4, "-2" = 5, "-1" = 6, "+4" = 10)
stopifnot("unexpected Kozak context_index set" =
            setequal(kozak_pub$context_index, c(1:6, 10)))

pwm_publ <- kozak_pub %>%
  arrange(match(context_index, win_idx)) %>%
  dplyr::select(A = A_frequency, C = C_frequency, G = G_frequency, T = T_frequency) %>%
  as.matrix()
rownames(pwm_publ) <- names(win_idx)
if (MUT == "kozak_minus3_bug") {
  # The historical 01h_ defect: G at -3 typed as 0.13 instead of the published ~0.36-0.41.
  pwm_publ["-3", "G"] <- 0.13
  pwm_publ["-3", ]    <- pwm_publ["-3", ] / sum(pwm_publ["-3", ])
}
stopifnot("published Kozak rows do not sum to 1" =
            all(abs(rowSums(pwm_publ) - 1) < 1e-8))

kozak_tbl <- inner_join(utr5, cds, by = "transcript_id_clean") %>%
  mutate(
    utr5_len = nchar(utr5_sequence),
    window   = paste0(str_sub(utr5_sequence, start = pmax(1L, utr5_len - 5L)),
                      str_sub(cds_sequence, 1, 4)),
    # 01h_ scored any >=10 nt window. A window whose CDS does not start with ATG is not a
    # start context at all, so it is NA here rather than silently scored as weak.
    usable   = nchar(window) == 10L & str_sub(window, 7, 9) == "ATG"
  )

logodds <- log2((pwm_publ + 0.001) / 0.25)
score_window <- function(win) {
  mat <- str_split_fixed(win, "", 10)
  rowSums(sapply(seq_along(win_idx), function(k)
    logodds[names(win_idx)[k], mat[, win_idx[k]]]))
}
kozak_tbl$kozak_pwm_score_v2 <- NA_real_
kozak_tbl$kozak_pwm_score_v2[kozak_tbl$usable] <-
  score_window(kozak_tbl$window[kozak_tbl$usable])

cat("kozak_pwm_score_v2: scored", sum(kozak_tbl$usable), "of", nrow(kozak_tbl),
    "; NA (not AUG-anchored or too short):", sum(!kozak_tbl$usable), "\n")

## --- noderer_tis_efficiency ---------------------------------------------------------
# Noderer 2014 (MSB 10:748) Table S2: initiation efficiency MEASURED by FACS-seq for all
# 65,536 AUG contexts, -6..+5. Imported, never transcribed. This is the same quantity
# kozak_pwm_score_v2 estimates from a 1987 frequency table - but measured, not estimated, and
# the two agree only weakly (rho ~0.15, see the validation below), so 02_ keeps both by
# default. Its initiation_score param can still isolate either one.
noderer <- read_tsv(file.path(ref_dir, "noderer_2014_tis_efficiency.tsv"),
                    skip = 1, show_col_types = FALSE) %>%   # line 1 is the table caption
  mutate(context = str_replace_all(sequence, "U", "T"))     # source alphabet is RNA

stopifnot(
  "Noderer table is not 65,536 contexts"  = nrow(noderer) == 65536L,
  "Noderer contexts are not 11 nt"        = all(nchar(noderer$context) == 11L),
  "Noderer contexts are not AUG-anchored" = all(str_sub(noderer$context, 7, 9) == "ATG"),
  "Noderer efficiency outside 12-150"     = all(between(noderer$efficiency, 12, 150)),
  "Noderer contexts are not unique"       = !any(duplicated(noderer$context))
)
cat("Noderer 2014:", nrow(noderer), "contexts, efficiency",
    paste(range(noderer$efficiency), collapse = "-"), "\n")

# -6..+5 is six nt of 5'UTR then the first five of the CDS: one nt longer than the Kozak
# window, so it is built separately rather than reused.
noderer_tbl <- inner_join(utr5, cds, by = "transcript_id_clean") %>%
  mutate(
    utr5_len = nchar(utr5_sequence),
    context  = if (MUT == "noderer_window_shift")
                 paste0(str_sub(utr5_sequence, start = pmax(1L, utr5_len - 4L)),
                        str_sub(cds_sequence, 1, 6))      # -5..+6 instead of -6..+5
               else
                 paste0(str_sub(utr5_sequence, start = pmax(1L, utr5_len - 5L)),
                        str_sub(cds_sequence, 1, 5)),
    usable   = nchar(context) == 11L & str_sub(context, 7, 9) == "ATG"
  ) %>%
  left_join(noderer %>% dplyr::select(context, noderer_tis_efficiency = efficiency),
            by = "context")
noderer_tbl$noderer_tis_efficiency[!noderer_tbl$usable] <- NA_real_

# The lookup is exhaustive over ACGT, so a usable window that fails to match can only contain
# an N. Anything else means the window or the alphabet is wrong.
unmatched <- sum(noderer_tbl$usable & is.na(noderer_tbl$noderer_tis_efficiency))
stopifnot("usable windows failed the exhaustive Noderer lookup" =
            unmatched == sum(noderer_tbl$usable & str_detect(noderer_tbl$context, "[^ACGT]")))
cat("noderer_tis_efficiency: scored",
    sum(!is.na(noderer_tbl$noderer_tis_efficiency)), "of", nrow(noderer_tbl),
    "; NA (not AUG-anchored or too short):", sum(!noderer_tbl$usable),
    "; unmatched (non-ACGT):", unmatched, "\n")

## --- CDS eligibility ----------------------------------------------------------------
# GENCODE carries 3'-incomplete CDS annotations (cds_end_NF): the CDS starts correctly at ATG
# but runs off the end of the annotation, so its length is not a multiple of 3. 01g_ already
# refuses to score these (coRdon requires %%3 == 0), which is why cai/tai/csc/fop are NA for
# them - but gc3 computes on the truncated fragment anyway, giving a median ~8.7 points below
# the rest of the transcriptome. Flag the condition explicitly and apply it uniformly, so a
# codon-derived feature is never computed on a CDS that cannot support codons.
cds <- cds %>%
  mutate(
    sq_upper       = str_to_upper(cds_sequence),
    cds_n          = nchar(sq_upper),
    cds_frame_ok   = cds_n %% 3L == 0L,
    cds_has_start  = str_sub(sq_upper, 1, 3) == "ATG",
    cds_has_stop   = str_sub(sq_upper, (cds_n %/% 3L - 1L) * 3L + 1L,
                             (cds_n %/% 3L) * 3L) %in% c("TAA", "TAG", "TGA"),
    cds_complete   = cds_frame_ok & cds_has_start & cds_has_stop
  )

cat("CDS eligibility:\n")
cat("  frame divisible by 3 :", sum(cds$cds_frame_ok),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_frame_ok)))
cat("  starts with ATG      :", sum(cds$cds_has_start),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_has_start)))
cat("  ends with a stop     :", sum(cds$cds_has_stop),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_has_stop)))
cat("  complete             :", sum(cds$cds_complete),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_complete)))

# The right mask is a MISSING TERMINAL STOP, not a bad frame. Checked against the GENCODE v49
# tags, the three irregular classes are:
#
#   65  out of frame, ATG, no stop   -> 62 tagged cds_end_NF   (3'-incomplete)
#   53  in frame,     ATG, no stop   -> 52 tagged cds_end_NF   (3'-incomplete, truncation
#                                       happens to land on a codon boundary)
#   25  in frame,     no ATG, stop   -> 24 tagged non_ATG_start, 17 of them MANE Select
#
# Only the first two are truncated annotations. Masking on frame alone catches 65 and misses
# the 53, which pass every existing filter and receive full cai/tai/csc/fop scores computed on
# an incomplete CDS. The third class is real biology - genuine non-AUG initiation, mostly MANE
# Select - and must be KEPT: it has a complete CDS and only its Kozak score is meaningless.
## --- internal-codon variants: gc3, cai, fop, csc ---------------------------------
# "Internal" = initiator AUG and terminal stop codon both removed. Neither is decoded the way
# a sense codon is: every transcript has exactly one initiator, so it carries no information
# between genes, and the stop is released rather than decoded by a tRNA.
#
# CAI and FOP already excluded stops (via sense_codons / sense_cols in 01g_), so only the
# initiator changes them. CSC does not: iCodon is given the sequence, and it is sensitive to
# what the sequence contains.
suppressMessages({library(coRdon); library(Biostrings); library(iCodon)})

gen_code     <- Biostrings::getGeneticCode("SGC0")
stop_codons  <- names(gen_code[gen_code == "*"])
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

# 01g_ scores only in-frame CDS, so the reproduction below must use the same subset.
codon_tbl <- cds %>% filter(cds_frame_ok)
cat("In-frame CDS for codon scoring:", nrow(codon_tbl), "\n")

counts_of <- function(d) { m <- codonCounts(codonTable(d)); rownames(m) <- names(d); m }
build_w <- function(cm, idx) {
  ref <- colSums(cm[idx, , drop = FALSE])
  w <- setNames(rep(NA_real_, length(gen_code)), names(gen_code)); w[stop_codons] <- 0
  for (aa in names(aa_to_codons)) {
    cod <- aa_to_codons[[aa]]; cnt <- ref[cod]; tot <- sum(cnt)
    if (tot == 0) { w[cod] <- 1; next }
    rscu <- cnt / (tot / length(cod)); mx <- max(rscu)
    w[cod] <- if (mx == 0) 0 else pmin(rscu / mx, 1)
  }
  w
}
cai_of <- function(cm, w) {
  use <- sense_codons[!is.na(w[sense_codons]) & w[sense_codons] > 0]
  apply(cm, 1, function(g) { cu <- g[use]; n <- sum(cu)
    if (n == 0) NA_real_ else exp(sum(cu * log(w[use])) / n) })
}
fop_of <- function(cm, opt) {
  sc <- setdiff(colnames(cm), stop_codons)
  tot <- rowSums(cm[, sc, drop = FALSE]); o <- rowSums(cm[, opt, drop = FALSE])
  r <- o / tot; r[tot == 0] <- NA_real_; r
}

full_ss <- DNAStringSet(codon_tbl$sq_upper); names(full_ss) <- codon_tbl$transcript_id_clean
cm_full <- counts_of(full_ss)

ribo_tx  <- features %>% filter(str_detect(symbol, "^RPL|^RPS")) %>% pull(transcript_id_clean)
ribo_idx <- rownames(cm_full) %in% ribo_tx
w_full   <- build_w(cm_full, ribo_idx)

te_rank <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE) %>%
  filter(condition == "normoxia", treatment == "sictrl") %>%
  group_by(gene_id) %>% summarise(mte = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mte)) %>% arrange(desc(mte))
top10  <- sub("\\..*", "", te_rank$gene_id[seq_len(ceiling(nrow(te_rank) * 0.10))])
ref_tx <- features$transcript_id_clean[sub("\\..*", "", features$gene_id) %in% top10]
ref_tx <- ref_tx[ref_tx %in% rownames(cm_full)]
opt_full <- unname(sapply(aa_to_codons, function(c_)
  c_[which.max(colSums(cm_full[ref_tx, , drop = FALSE])[c_])]))

# VERIFY before trusting: recomputing on the FULL CDS must reproduce 01g_'s stored values
# exactly. If it does not, the reference sets or the weights have been rebuilt wrongly and
# nothing downstream is safe.
verify <- tibble(transcript_id_clean = rownames(cm_full),
                 cai_rep = cai_of(cm_full, w_full),
                 fop_rep = fop_of(cm_full, opt_full)) %>%
  inner_join(features %>% dplyr::select(transcript_id_clean, cai, fop),
             by = "transcript_id_clean")
cai_dev <- max(abs(verify$cai_rep - verify$cai), na.rm = TRUE)
fop_dev <- max(abs(verify$fop_rep - verify$fop), na.rm = TRUE)
cat("Reproduction of 01g_ on the full CDS: CAI max|dev| =", signif(cai_dev, 3),
    ", FOP max|dev| =", signif(fop_dev, 3), ", n =", sum(complete.cases(verify)), "\n")
stopifnot("recomputation does not reproduce 01g_'s CAI" = cai_dev < 1e-9,
          "recomputation does not reproduce 01g_'s FOP" = fop_dev < 1e-9)
cat("Verified: the codon pipeline here matches 01g_ exactly\n")

# Now the internal-codon versions.
trim_internal <- function(x) {
  n <- nchar(x) %/% 3L
  a <- ifelse(substr(x, 1, 3) == "ATG", 4L, 1L)
  b <- ifelse(substr(x, (n - 1) * 3L + 1L, n * 3L) %in% stop_codons, (n - 1) * 3L, n * 3L)
  substr(x, a, b)
}
codon_tbl <- codon_tbl %>% mutate(s_int = trim_internal(sq_upper))
stopifnot("internal sequences lost frame" = all(nchar(codon_tbl$s_int) %% 3 == 0))

int_ss <- DNAStringSet(codon_tbl$s_int); names(int_ss) <- codon_tbl$transcript_id_clean
cm_int  <- counts_of(int_ss)
w_int   <- build_w(cm_int, ribo_idx)
opt_int <- unname(sapply(aa_to_codons, function(c_)
  c_[which.max(colSums(cm_int[ref_tx, , drop = FALSE])[c_])]))
cat("Optimal codons changed by trimming:", sum(opt_int != opt_full), "of 20\n")

predict_human <- predict_stability("human")
uniq <- unique(codon_tbl$s_int)
uscore <- as.numeric(predict_human(uniq)); names(uscore) <- uniq

gc3_from <- function(x) vapply(x, function(z) {
  n <- nchar(z) %/% 3L; if (n < 1L) return(NA_real_)
  k <- seq.int(3L, n * 3L, 3L)
  sum(substr(rep.int(z, n), k, k) %in% c("G", "C")) * 100 / n
}, numeric(1L), USE.NAMES = FALSE)

codon_internal <- tibble(
  transcript_id_clean = codon_tbl$transcript_id_clean,
  gc3_internal = gc3_from(codon_tbl$s_int),
  cai_internal = cai_of(cm_int, w_int),
  fop_internal = fop_of(cm_int, opt_int),
  csc_internal = unname(uscore[codon_tbl$s_int])
)

## --- tai_gtrnadb --------------------------------------------------------------------
# dos Reis-style tAI over the collaborator's GtRNAdb gene-copy table and wobble scheme.
# Anticodon (5'->3') = [wobble base] + complement(codon[2]) + complement(codon[1]).
gcn    <- read_tsv(file.path(ref_dir, "human_trna_gene_copy_hg38.tsv"),
                   comment = "#", show_col_types = FALSE)
wobble <- read_tsv(file.path(ref_dir, "trna_wobble_scheme_v1.tsv"), show_col_types = FALSE)
wob <- wobble %>% transmute(w  = chartr("U", "T", anticodon_wobble_base_rna),
                            n3 = chartr("U", "T", codon_third_base_rna),
                            s  = penalty_s)

comp   <- c(A = "T", C = "G", G = "C", T = "A")
bases  <- c("T", "C", "A", "G")
stops  <- c("TAA", "TAG", "TGA")
sense_codons <- setdiff(as.vector(outer(outer(bases, bases, paste0), bases, paste0)), stops)

W <- map_dbl(sense_codons, function(cod) {
  rules <- wob %>% filter(n3 == substr(cod, 3, 3))
  if (!nrow(rules)) return(0)
  anti <- paste0(rules$w, comp[[substr(cod, 2, 2)]], comp[[substr(cod, 1, 1)]])
  cn <- gcn$gene_copy_number[match(anti, gcn$anticodon)]
  cn[is.na(cn)] <- 0
  sum((1 - rules$s) * cn)
})
names(W) <- sense_codons
tai_w <- W / max(W)
n_zero_w <- sum(tai_w == 0)
if (n_zero_w > 0) tai_w[tai_w == 0] <- exp(mean(log(tai_w[tai_w > 0])))  # dos Reis substitution

cat("tAI weights: 61 sense codons, ", n_zero_w, " with no decoding tRNA; range ",
    paste(round(range(tai_w), 3), collapse = "-"), "\n", sep = "")

cds$tai_gtrnadb <- map_dbl(cds$sq_upper, function(sq) {
  n <- nchar(sq) %/% 3L
  if (n < 3L) return(NA_real_)
  cods <- substring(sq, seq(1, n * 3L, 3L), seq(3L, n * 3L, 3L))
  cods <- cods[-1]                                                   # drop initiator
  if (tail(cods, 1) %in% stops) cods <- head(cods, -1)               # drop terminal stop
  ww <- tai_w[cods]; ww <- ww[!is.na(ww)]
  if (!length(ww)) return(NA_real_)
  exp(mean(log(ww)))                                                 # geometric mean
})

truncated <- !cds$cds_has_stop
cds$tai_gtrnadb[truncated] <- NA_real_
codon_internal <- codon_internal %>%
  mutate(across(c(gc3_internal, cai_internal, fop_internal, csc_internal),
                ~ if_else(transcript_id_clean %in% cds$transcript_id_clean[truncated],
                          NA_real_, .x)))
cat("Masked", sum(truncated), "transcripts with no terminal stop (cds_end_NF)",
    "across every codon-derived feature\n")
cat("  of which out-of-frame:", sum(truncated & !cds$cds_frame_ok),
    " in-frame but truncated:", sum(truncated & cds$cds_frame_ok), "\n")
cat("Kept", sum(!cds$cds_has_start & cds$cds_has_stop),
    "non-AUG-start transcripts: complete CDS, real initiation biology\n")

## --- nascent peptide + poly(A) tracks -----------------------------------------------
# Two stalling families read off the CDS already in hand. A proline-rich or strongly basic
# nascent chain slows the peptide exit tunnel; consecutive lysine codons make an mRNA poly(A)
# stretch the ribosome slides on. Built on s_int (initiator and terminal stop removed), so
# they inherit the codon features' frame guarantee and the same truncation mask.
pep_ss <- if (MUT == "peptide_frame_shift") {
  DNAStringSet(str_sub(codon_tbl$s_int, 2L, nchar(codon_tbl$s_int) - 2L))   # +1 frame
} else {
  DNAStringSet(codon_tbl$s_int)
}
# translate() is masked here by another attached package, so it is qualified.
pep_aa <- Biostrings::translate(pep_ss, if.fuzzy.codon = "solve")
pep_len <- nchar(pep_aa)

# Net charge: K/R +1, D/E -1, maximised over any 30-aa window. Chains shorter than the window
# score their whole length, so the feature is defined for every peptide.
charge_window_max <- function(aa, w = 30L) {
  ch <- strsplit(aa, "", fixed = TRUE)[[1]]
  if (!length(ch)) return(NA_real_)
  v <- integer(length(ch))
  v[ch %in% c("K", "R")] <-  1L
  v[ch %in% c("D", "E")] <- -1L
  if (length(v) <= w) return(as.numeric(sum(v)))
  cs <- cumsum(c(0L, v))
  as.numeric(max(cs[(w + 1L):length(cs)] - cs[seq_len(length(cs) - w)]))
}

# Longest in-frame run of one codon repeated. AAA and AAG are kept apart by design: both rise
# with lysine content, so a merged "poly-lysine" feature would report amino-acid composition
# rather than the slippery A-rich tract that actually stalls.
max_codon_run <- function(s, codon) {
  n <- nchar(s) %/% 3L
  if (n < 1L) return(0L)
  hit <- substring(s, seq.int(1L, n * 3L, 3L), seq.int(3L, n * 3L, 3L)) == codon
  r <- rle(hit)
  if (!any(r$values)) 0L else max(r$lengths[r$values])
}

aa_chr <- as.character(pep_aa)
pep <- tibble(
  transcript_id_clean = codon_tbl$transcript_id_clean,
  proline_fraction    = as.numeric(letterFrequency(pep_aa, "P", as.prob = TRUE)),
  ppp_motif_density   = if_else(pep_len > 0,
                                100 * vcountPattern("PPP", pep_aa) / pep_len, NA_real_),
  max_net_charge_30aa = vapply(aa_chr, charge_window_max, numeric(1), USE.NAMES = FALSE),
  max_consecutive_aaa = vapply(codon_tbl$s_int, max_codon_run, integer(1), "AAA",
                               USE.NAMES = FALSE),
  max_consecutive_aag = vapply(codon_tbl$s_int, max_codon_run, integer(1), "AAG",
                               USE.NAMES = FALSE)
)
pep_cols <- setdiff(names(pep), "transcript_id_clean")
pep <- pep %>%
  mutate(across(all_of(pep_cols), as.numeric)) %>%
  mutate(across(all_of(pep_cols),
                ~ if_else(transcript_id_clean %in% cds$transcript_id_clean[truncated],
                          NA_real_, .x)))
cat("nascent peptide + poly(A): scored", sum(!is.na(pep$proline_fraction)), "of", nrow(pep),
    "; median proline", round(median(pep$proline_fraction, na.rm = TRUE), 4),
    "; median max AAG run", median(pep$max_consecutive_aag, na.rm = TRUE), "\n")

## --- termination context + Cridge 2018 readthrough --------------------------------
# New family, nothing equivalent in the existing pipeline. Motivated by the csc result: after
# the terminal stop was removed from the iCodon input, csc lost 1.35 pp of importance share
# (20/20 seeds), which means it had been carrying stop-codon identity. That signal belongs
# here, under its own name, rather than smuggled into a codon-optimality score.
utr3_cache <- file.path(out_dir, "precomputed_utr3_seqs.rds")
if (file.exists(utr3_cache)) {
  utr3_tbl <- readRDS(utr3_cache)
} else {
  suppressMessages({library(GenomicFeatures); library(BSgenome.Hsapiens.UCSC.hg38)})
  txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
  u3   <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                threeUTRsByTranscript(txdb, use.names = TRUE))
  utr3_tbl <- tibble(transcript_id_clean = sub("\\..*", "", names(u3)),
                     utr3_seq = as.character(u3))
  saveRDS(utr3_tbl, utr3_cache)
}
utr3_tbl <- utr3_tbl %>% filter(transcript_id_clean %in% features$transcript_id_clean)
cat("3'UTR sequences available:", nrow(utr3_tbl), "of", nrow(features), "\n")

cridge <- read_tsv(file.path(ref_dir, "cridge2018_termination_contexts.tsv"),
                   show_col_types = FALSE)
if (MUT == "cridge_permute")            # break the context -> readthrough correspondence
  cridge$readthrough_measure <- sample(cridge$readthrough_measure)
cridge_med <- median(cridge$readthrough_measure)
stopifnot(
  "Cridge table is not 192 contexts"    = nrow(cridge) == 192L,
  "Cridge median readthrough off"       = between(cridge_med, 0.08, 0.095),
  "Cridge readthrough not a percentage" = all(cridge$readthrough_measure > 0 &
                                              cridge$readthrough_measure < 5)
)
cat("Cridge 2018:", nrow(cridge), "contexts, median", round(cridge_med, 4),
    "% (paper ~0.087%)\n")

# Codons to the next in-frame stop downstream of the annotated one: how far a readthrough
# ribosome would travel before terminating again.
next_inframe_stop <- function(u) {
  n <- nchar(u) %/% 3L
  if (n < 1L) return(NA_integer_)
  cod <- substring(u, seq(1, n * 3L, 3L), seq(3L, n * 3L, 3L))
  w <- which(cod %in% c("TAA", "TAG", "TGA"))
  if (!length(w)) NA_integer_ else w[1]
}

term <- cds %>%
  dplyr::select(transcript_id_clean, sq_upper, cds_has_stop) %>%
  filter(cds_has_stop) %>%
  mutate(ncod = nchar(sq_upper) %/% 3L,
         stop_codon = substr(sq_upper, (ncod - 1L) * 3L + 1L, ncod * 3L)) %>%
  inner_join(utr3_tbl, by = "transcript_id_clean") %>%
  mutate(
    u3   = str_to_upper(utr3_seq),
    ctx6 = paste0(stop_codon, str_sub(u3, 1, 3)),    # +1..+6, the Cridge reporter window
    ctx9 = paste0(stop_codon, str_sub(u3, 1, 6)),    # +1..+9
    term_stop_TAA = as.integer(stop_codon == "TAA"),
    term_stop_TAG = as.integer(stop_codon == "TAG"),
    term_stop_TGA = as.integer(stop_codon == "TGA"),
    p4 = str_sub(u3, 1, 1),
    term_plus4_A = as.integer(p4 == "A"), term_plus4_C = as.integer(p4 == "C"),
    term_plus4_G = as.integer(p4 == "G"), term_plus4_T = as.integer(p4 == "T"),
    term_window_gc = if_else(nchar(ctx9) == 9L,
                             100 * str_count(ctx9, "[GC]") / 9, NA_real_),
    term_interval_codons = map_int(u3, next_inframe_stop)
  ) %>%
  left_join(cridge %>% dplyr::select(ctx6 = termination_context_plus1_to_plus6,
                                     term_readthrough_cridge = readthrough_measure),
            by = "ctx6") %>%
  mutate(term_log1p_readthrough = log1p(term_readthrough_cridge))

cat("Cridge join coverage:", sum(!is.na(term$term_readthrough_cridge)), "/", nrow(term),
    sprintf(" (%.1f%%)\n", 100 * mean(!is.na(term$term_readthrough_cridge))))

term_cols <- c("term_stop_TAA", "term_stop_TAG", "term_stop_TGA",
               "term_plus4_A", "term_plus4_C", "term_plus4_G", "term_plus4_T",
               "term_window_gc", "term_interval_codons",
               "term_readthrough_cridge", "term_log1p_readthrough")
if (MUT == "term_stop_shuffle") {       # stop identity no longer matches the transcript
  i <- sample(nrow(term))
  term[, c("term_stop_TAA", "term_stop_TAG", "term_stop_TGA")] <-
    term[i, c("term_stop_TAA", "term_stop_TAG", "term_stop_TGA")]
}
term <- term %>% dplyr::select(transcript_id_clean, all_of(term_cols))
## --- G4mer rG4 propensity ------------------------------------------------------------
# Inference is already done - g4mer/02_run_g4mer.py scored every region and wrote
# output/g4mer/g4mer_summary_<region>.tsv (note: output/g4mer/, NOT the g4mer/output/ named
# in .gitignore). Nothing is re-run here; this is a join.
#
# The collaborator has NO G-quadruplex feature (00_ inventory: status "ours_only"), so unlike
# struct_* there is nothing to reconcile against and no name can collide.
#
# Stride differs by region: 5'UTR was scored at stride 10, CDS and 3'UTR at stride 20. Denser
# sampling raises the expected maximum, so the three columns are NOT comparable to each other.
# Each is modelled as its own column and never differenced, so within-region calibration is
# all that is needed; 01u_ holds the harmonised version for cross-region work.
g4_dir     <- here("output", "g4mer")
g4_regions <- c("utr5", "cds", "utr3")
g4_raw <- map_dfr(g4_regions, function(rg)
  read_tsv(file.path(g4_dir, paste0("g4mer_summary_", rg, ".tsv")), show_col_types = FALSE))

stopifnot(
  "g4mer scores outside [0, 1]" =
    all(between(g4_raw$g4mer_max, 0, 1)) && all(between(g4_raw$g4mer_mean, 0, 1)) &&
    all(between(g4_raw$g4mer_frac_above, 0, 1)),
  "g4mer_max below g4mer_mean" = all(g4_raw$g4mer_max >= g4_raw$g4mer_mean),
  "g4mer argmax outside its region" =
    all(g4_raw$g4mer_argmax_start >= 0 & g4_raw$g4mer_argmax_start < g4_raw$region_length)
)
cat("G4mer summaries loaded:",
    paste(sprintf("%s %d", names(table(g4_raw$region)), table(g4_raw$region)),
          collapse = "; "), "\n")

if (MUT == "g4_region_swap")            # 5'UTR scores land on the 3'UTR and vice versa
  g4_raw$region <- dplyr::recode(g4_raw$region, utr5 = "utr3", utr3 = "utr5")
g4_wide <- g4_raw %>%
  dplyr::select(transcript_id_clean, region, g4mer_max, g4mer_mean,
                g4mer_frac_above, region_length) %>%
  pivot_wider(names_from = region,
              values_from = c(g4mer_max, g4mer_mean, g4mer_frac_above, region_length))

# g4mer_max is a maximum over windows, so it rises with region length by construction, and
# rG4 formation needs G-runs so it also tracks GC. The modelled column is therefore the
# residual on log2 length + region GC, following the gc3_residual precedent in 01v_: left raw,
# it would let the forest reach length and GC a second time under a structural name.
# g4mer_mean and g4mer_frac_above are per-window averages and fractions, carrying no length
# bias, so those enter raw.
g4_join <- features %>%
  dplyr::select(transcript_id_clean, utr5_gc, cds_gc, utr3_gc) %>%
  left_join(g4_wide, by = "transcript_id_clean")

for (rg in g4_regions) {
  mx <- paste0("g4mer_max_", rg); ln <- paste0("region_length_", rg)
  gcc <- paste0(rg, "_gc")
  fit <- lm(g4_join[[mx]] ~ log2(g4_join[[ln]] + 1) + g4_join[[gcc]], na.action = na.exclude)
  g4_join[[paste0("g4mer_max_resid_", rg)]] <- as.numeric(residuals(fit))
  cat(sprintf("  g4mer_max_%-4s ~ log2(length) + GC: R2 = %.3f, n = %d\n",
              rg, summary(fit)$r.squared, sum(!is.na(residuals(fit)))))
}

g4_cols <- c(paste0("g4mer_max_resid_", g4_regions),
             paste0("g4mer_mean_", g4_regions),
             paste0("g4mer_frac_above_", g4_regions))
g4 <- g4_join %>% dplyr::select(transcript_id_clean, all_of(g4_cols))

## --- audit corrections added as variant pairs (F3, F4) -------------------------------
# Both supersede a BASELINE column, so neither is applied in place: the originals stay so the
# reproduction gates keep working, and 02_'s variant_set picks the corrected member by default.

# F3. hia2026_dhx29_occupancy was joined at GENE level from a transcript-keyed source, so
# ~1,013 transcripts carry a different isoform's occupancy. Ribosome occupancy is
# isoform-specific, so that is a value/key mix-up rather than a rounding issue.
dhx29_occ <- readRDS(here("output", "predictive_modeling", "dhx29_riboseq_occupancy.rds"))
stopifnot("dhx29 occupancy source is not transcript-keyed" =
            "transcript_id_clean" %in% colnames(dhx29_occ),
          "dhx29 occupancy source has duplicate transcripts" =
            !any(duplicated(dhx29_occ$transcript_id_clean)))
occ_tx <- if (MUT == "gene_level_join") {
  # F3 class: join a transcript-keyed source on gene, so the modelled transcript receives
  # whichever isoform the source happened to hold.
  dhx29_occ %>% mutate(gene_id_clean = sub("\\..*", "", ensembl_gene)) %>%
    group_by(gene_id_clean) %>% slice(1) %>% ungroup() %>%
    dplyr::select(gene_id_clean, hia2026_dhx29_occupancy_tx = dhx29_occupancy_score) %>%
    inner_join(features %>% dplyr::select(transcript_id_clean, gene_id_clean),
               by = "gene_id_clean") %>%
    dplyr::select(transcript_id_clean, hia2026_dhx29_occupancy_tx)
} else {
  dhx29_occ %>%
    dplyr::select(transcript_id_clean, hia2026_dhx29_occupancy_tx = dhx29_occupancy_score)
}

# F4. clip_lee_complete_media_bound is bit-identical to clip_lee_glu_dep_bound - the two
# columns are already identical in the upstream score_card.csv - so clip_cap_binding_count,
# which sums all three *_bound columns, counts one dataset twice. The corrected count uses the
# two distinct conditions. Which NAME is the duplicate cannot be decided from our data, so the
# second is kept and the third dropped, and both facts are recorded.
stopifnot("the CLIP duplicate this correction assumes is no longer present" =
            identical(features$clip_lee_complete_media_bound, features$clip_lee_glu_dep_bound))

## --- join --------------------------------------------------------------------------
features <- features %>%
  left_join(kozak_tbl %>% dplyr::select(transcript_id_clean, kozak_pwm_score_v2),
            by = "transcript_id_clean") %>%
  left_join(noderer_tbl %>% dplyr::select(transcript_id_clean, noderer_tis_efficiency),
            by = "transcript_id_clean") %>%
  left_join(cds %>% dplyr::select(transcript_id_clean, tai_gtrnadb,
                                  cds_frame_ok, cds_has_start, cds_has_stop, cds_complete),
            by = "transcript_id_clean") %>%
  left_join(codon_internal, by = "transcript_id_clean") %>%
  left_join(pep, by = "transcript_id_clean") %>%
  left_join(g4, by = "transcript_id_clean") %>%
  left_join(occ_tx, by = "transcript_id_clean") %>%
  left_join(term, by = "transcript_id_clean")

corrected_cols <- c("kozak_pwm_score_v2", "tai_gtrnadb",
                    "gc3_internal", "cai_internal", "fop_internal", "csc_internal")
stopifnot("termination columns did not attach" = all(term_cols %in% colnames(features)),
          "noderer_tis_efficiency did not attach" =
            "noderer_tis_efficiency" %in% colnames(features),
          "nascent peptide / poly(A) columns did not attach" =
            all(pep_cols %in% colnames(features)),
          "g4mer columns did not attach" = all(g4_cols %in% colnames(features)),
          "corrected dhx29 occupancy did not attach" =
            "hia2026_dhx29_occupancy_tx" %in% colnames(features))

# F4 corrected count, built after the join so it sits beside the column it supersedes.
features <- features %>%
  mutate(clip_cap_binding_count_v2 =
           as.integer(clip_lee_int_stress_bound) + as.integer(clip_lee_glu_dep_bound))
cat("clip_cap_binding_count: original sums 3 columns of which 2 are identical;",
    "v2 sums the 2 distinct conditions\n")
cat("  transcripts changed:",
    sum(features$clip_cap_binding_count != features$clip_cap_binding_count_v2, na.rm = TRUE),
    "\n")
cat("hia2026_dhx29_occupancy: gene-level", sum(!is.na(features$hia2026_dhx29_occupancy)),
    "non-NA vs transcript-level", sum(!is.na(features$hia2026_dhx29_occupancy_tx)), "\n")
cat("  transcripts whose value changes:",
    sum(features$hia2026_dhx29_occupancy != features$hia2026_dhx29_occupancy_tx,
        na.rm = TRUE), "\n")
eligibility_cols <- c("cds_frame_ok", "cds_has_start", "cds_has_stop", "cds_complete")
stopifnot(
  "corrected columns did not all attach" = all(corrected_cols %in% colnames(features)),
  "eligibility columns did not all attach" = all(eligibility_cols %in% colnames(features)),
  "a codon feature survived on a truncated (cds_end_NF) CDS" =
    all(vapply(c("gc3_internal","cai_internal","fop_internal","csc_internal","tai_gtrnadb"),
               function(f) all(is.na(features[[f]][!features$cds_has_stop])), logical(1))),
  "non-AUG-start transcripts were masked, but they are real" =
    any(!is.na(features$gc3_internal[!features$cds_has_start & features$cds_has_stop]))
)
## --- tco_struct_* rename -------------------------------------------------------------
# NAME COLLISION (00_ inventory): the collaborator also has struct_accessibility_*, but
# theirs is a whole-transcript fold while ours is region-isolated RNAplfold (-W 150 -L 100).
# Same names, different quantities - the one case in the inventory where a name, not a
# definition, is the problem. Ours is renamed; nothing is refolded.
# Renaming does not reorder columns, so the forest is unaffected at equal seeds.
struct_old <- grep("^struct_", colnames(features), value = TRUE)
features <- features %>% dplyr::rename_with(~ paste0("tco_", .x), all_of(struct_old))
stopifnot("tco_struct_ rename did not apply" =
            length(grep("^struct_", colnames(features))) == 0L &&
            length(grep("^tco_struct_", colnames(features))) == length(struct_old))
# These are baseline-era columns under a new spelling, not new features, so the record of
# what the baseline contained has to follow the rename. Without this they are tagged
# added_in = "tco" and both gates silently drop all 18.
baseline_cols <- sub("^struct_", "tco_struct_", baseline_cols)
cat("Renamed", length(struct_old), "struct_* columns to tco_struct_*\n")

## --- structure repairs (audit findings F1, F2, F5) -----------------------------------
# F1. 01d_/01k_ apply replace_na(., 0) to the RNAplfold columns, so transcripts the cluster
# run never scored arrive as ZERO rather than NA. Zero means "maximally structured" to the
# forest, and because the NA is gone, 02_'s sparsity filter and median imputation both miss
# them. The sentinel is unambiguous: mean accessibility exactly 0 AND sd exactly 0 over a
# whole region is not a measurement. Restored to NA per region.
struct_region <- list(
  utr5 = grep("^tco_struct_", colnames(features), value = TRUE) %>%
           setdiff(grep("_cds(_|$)|_utr3(_|$)", ., value = TRUE)),
  cds  = grep("^tco_struct_.*_cds(_|$)",  colnames(features), value = TRUE),
  utr3 = grep("^tco_struct_.*_utr3(_|$)", colnames(features), value = TRUE)
)
# The zero-fill repair CHANGES VALUES, so it is the one correction that makes bit-exact
# baseline reproduction impossible. It is therefore switchable: build with
# apply_audit_repairs = FALSE to reproduce the historical matrix and re-run 02_'s gate.
if (!isTRUE(params$apply_audit_repairs))
  cat("apply_audit_repairs = FALSE: leaving the 01d_/01k_ zero-fill in place\n")
for (rg in if (isTRUE(params$apply_audit_repairs)) names(struct_region) else character(0)) {
  cols <- struct_region[[rg]]
  mn <- grep(paste0("accessibility_", rg, "_mean$"), cols, value = TRUE)
  sdc <- grep(paste0("accessibility_", rg, "_sd$"),  cols, value = TRUE)
  stopifnot("could not identify the mean/sd pair for a structure region" =
              length(mn) == 1 && length(sdc) == 1)
  sentinel <- features[[mn]] == 0 & features[[sdc]] == 0
  sentinel[is.na(sentinel)] <- FALSE
  features[sentinel, cols] <- NA
  cat(sprintf("  %-4s: %d transcripts restored from zero-fill to NA across %d columns\n",
              rg, sum(sentinel), length(cols)))
}

# F5. accessibility_*_min is identically 0 everywhere, because 01d_ takes min() AFTER
# replacing NA with 0. A constant carries no information and cannot be repaired from the
# summaries we hold. F2. num_structured_regions* is region length in disguise: Spearman
# ~1.000 against its own region length, and the "count" EXCEEDS the length by 1 for
# thousands of transcripts, which is impossible for a count of positions. Both are deprecated
# rather than deleted, so the baseline stays reproducible under include_deprecated = TRUE.
if (MUT == "struct_zero_fill") {     # F1 class: NA back to the 0 sentinel
  sc <- grep("^tco_struct_", colnames(features), value = TRUE)
  for (cc in sc) features[[cc]][is.na(features[[cc]])] <- 0
}

broken_structure <- grep("^tco_struct_(accessibility_.*_min|num_structured_regions)",
                         colnames(features), value = TRUE)
cat("Deprecating", length(broken_structure), "structure columns (constant, or length proxies):",
    paste(broken_structure, collapse = ", "), "\n")

cat("Matrix now", nrow(features), "x", ncol(features), "\n")
```

### Termination validation

```{r validate_termination, eval = !params$pass_through}
stopifnot(
  "readthrough outside (0, 5]%" =
    all(term$term_readthrough_cridge > 0 & term$term_readthrough_cridge <= 5, na.rm = TRUE),
  "termination window GC outside [0, 100]" =
    all(term$term_window_gc >= 0 & term$term_window_gc <= 100, na.rm = TRUE),
  "readthrough interval not positive" =
    all(term$term_interval_codons > 0, na.rm = TRUE),
  "stop indicators do not sum to 1" =
    all(term$term_stop_TAA + term$term_stop_TAG + term$term_stop_TGA == 1L),
  "plus4 indicators do not sum to 1" =
    all(term$term_plus4_A + term$term_plus4_C + term$term_plus4_G + term$term_plus4_T == 1L)
)

# Biological spot-check: TGA is the leakiest stop codon and TAA the tightest. This is the
# check that would catch a scrambled Cridge join - a shuffled table still passes every range
# assertion above.
by_stop <- term %>%
  mutate(stop = case_when(term_stop_TAA == 1 ~ "TAA", term_stop_TAG == 1 ~ "TAG",
                          TRUE ~ "TGA")) %>%
  group_by(stop) %>%
  summarise(n = n(), median_rt = median(term_readthrough_cridge, na.rm = TRUE),
            .groups = "drop")
print(as.data.frame(by_stop))

rt <- setNames(by_stop$median_rt, by_stop$stop)
stopifnot("TGA is not the leakiest stop codon" = rt["TGA"] > rt["TAG"],
          "TAA is not the tightest stop codon" = rt["TAG"] > rt["TAA"])
cat("Spot-check passed: TGA >", "TAG >", "TAA median readthrough\n")
```


## Validation

Range checks plus biological spot-checks, both required. The within-tier Kozak check is new:
it is the one that would have caught the `01h_` defect, which the existing tier-monotonicity
check cannot.

```{r validate_corrected, eval = !params$pass_through}
# --- computational ---
stopifnot(
  "gc3_internal outside [0, 100]" =
    all(features$gc3_internal >= 0 & features$gc3_internal <= 100, na.rm = TRUE),
  "cai_internal outside [0, 1]" =
    all(features$cai_internal >= 0 & features$cai_internal <= 1 + 1e-9, na.rm = TRUE),
  "fop_internal outside [0, 1]" =
    all(features$fop_internal >= 0 & features$fop_internal <= 1 + 1e-9, na.rm = TRUE),
  "tai_gtrnadb outside (0, 1]" =
    all(features$tai_gtrnadb > 0 & features$tai_gtrnadb <= 1, na.rm = TRUE),
  "kozak_pwm_score_v2 not finite" =
    all(is.finite(features$kozak_pwm_score_v2[!is.na(features$kozak_pwm_score_v2)]))
)

# --- biological spot-checks ---
gc3_med <- median(features$gc3_internal, na.rm = TRUE)
stopifnot("GC3 median outside [40, 65]%" = between(gc3_med, 40, 65))
cat("GC3 (sense codons) median:", round(gc3_med, 1), "%\n")

# Ribosomal proteins are highly translated and should sit above the median on tRNA supply.
rp <- features %>% filter(str_detect(symbol, "^RP[LS][0-9]"))
stopifnot("too few ribosomal proteins to test" = nrow(rp) >= 20,
          "ribosomal proteins are not above median tAI" =
            median(rp$tai_gtrnadb, na.rm = TRUE) > median(features$tai_gtrnadb, na.rm = TRUE))
cat("tAI: ribosomal proteins (n = ", nrow(rp), ") median ",
    round(median(rp$tai_gtrnadb, na.rm = TRUE), 3), " vs all ",
    round(median(features$tai_gtrnadb, na.rm = TRUE), 3), "\n", sep = "")

# THE NEW CHECK. Position -3 is the purine position: A and G are both favourable, so within a
# single ordinal tier they must score comparably. Under the old PWM they differ by ~2.3 bits.
kozak_tier <- kozak_tbl %>%
  filter(usable) %>%
  mutate(tier = case_when(
           str_detect(window, "^(GCCACC|GCCGCC)ATGG$") ~ 3L,
           str_detect(window, "^...[AG]..ATGG$")       ~ 2L,
           str_detect(window, "^...[AG]..ATG[ACT]$")   ~ 1L,
           str_detect(window, "^...[CT]..ATGG$")       ~ 1L,
           TRUE                                        ~ 0L),
         base_m3 = str_sub(window, 4, 4)) %>%
  filter(tier == 2L, base_m3 %in% c("A", "G")) %>%
  group_by(base_m3) %>%
  summarise(n = n(), mean_v2 = mean(kozak_pwm_score_v2), .groups = "drop")
print(kozak_tier)

purine_gap <- abs(diff(kozak_tier$mean_v2))
cat("A-minus-G spread within tier 2:", round(purine_gap, 3), "bits\n")
stopifnot("within-tier purine gap too large: the -3 row is wrong again" = purine_gap < 0.75)
cat("Within-tier purine check passed\n")
```

```{r validate_g4mer, eval = !params$pass_through}
stopifnot(
  "g4mer_mean outside [0, 1]" =
    all(between(features$g4mer_mean_utr5, 0, 1), na.rm = TRUE) &&
    all(between(features$g4mer_mean_cds,  0, 1), na.rm = TRUE) &&
    all(between(features$g4mer_mean_utr3, 0, 1), na.rm = TRUE),
  "g4mer_frac_above outside [0, 1]" =
    all(between(features$g4mer_frac_above_utr5, 0, 1), na.rm = TRUE) &&
    all(between(features$g4mer_frac_above_cds,  0, 1), na.rm = TRUE) &&
    all(between(features$g4mer_frac_above_utr3, 0, 1), na.rm = TRUE),
  "g4mer residual not finite" =
    all(vapply(paste0("g4mer_max_resid_", g4_regions),
               function(f) all(is.finite(features[[f]][!is.na(features[[f]])])), logical(1)))
)
for (f in g4_cols)
  cat(sprintf("  %-22s non-NA %5d  median %7.4f\n", f, sum(!is.na(features[[f]])),
              median(features[[f]], na.rm = TRUE)))

# The residual must have removed the length and GC it was fitted on. A residual still
# correlated with either would mean the model can reach them under a structural name.
for (rg in g4_regions) {
  rcol <- features[[paste0("g4mer_max_resid_", rg)]]
  lcol <- features[[paste0("log2_", rg, "_length")]]
  gcol <- features[[paste0(rg, "_gc")]]
  if (is.null(lcol)) next
  ok <- complete.cases(rcol, lcol, gcol)
  cat(sprintf("  %-4s residual vs log2 length r = %+0.3f, vs GC r = %+0.3f, n = %d\n",
              rg, cor(rcol[ok], lcol[ok]), cor(rcol[ok], gcol[ok]), sum(ok)))
  stopifnot("g4mer residual still carries its own length term" =
              abs(cor(rcol[ok], gcol[ok])) < 0.1)
}

# Biological spot-check: regions carrying a canonical G4 motif must score above those that
# do not. Same regex as 01u_. Range checks pass on a scrambled join; this does not.
g4_regex <- "([Gg]{3,}[ACGTNacgtn]{1,7}){3,}[Gg]{3,}"
g4_motif <- bind_rows(
  tibble(transcript_id_clean = utr5$transcript_id_clean, region = "utr5",
         sequence = utr5$utr5_sequence),
  tibble(transcript_id_clean = seqs$cds_seq$transcript_id_clean, region = "cds",
         sequence = seqs$cds_seq$cds_sequence),
  tibble(transcript_id_clean = utr3_tbl$transcript_id_clean, region = "utr3",
         sequence = utr3_tbl$utr3_seq)
) %>% mutate(has_g4 = str_detect(sequence, g4_regex))

g4_spot <- g4_raw %>%
  inner_join(g4_motif, by = c("transcript_id_clean", "region")) %>%
  group_by(region) %>%
  summarise(n = n(), pct_motif = round(100 * mean(has_g4), 1),
            med_motif    = round(median(g4mer_max[has_g4]), 4),
            med_no_motif = round(median(g4mer_max[!has_g4]), 4), .groups = "drop")
print(as.data.frame(g4_spot))
stopifnot("motif-containing regions do not outscore motif-free ones" =
            all(g4_spot$med_motif > g4_spot$med_no_motif))
cat("G4 motif spot-check passed in all three regions\n")

# The ordering check above is NOT sufficient, and the mutation test proved it: exchanging the
# 5'UTR and 3'UTR summaries passes it, because G-richness is correlated across regions of the
# same transcript, so a swapped score still ranks motif-containing above motif-free. Region
# assignment has to be checked DISCRIMINATIVELY - each region's score must track its OWN
# motif better than another region's. Without this, the CDS/5'UTR-vs-3'UTR result rests on an
# unverified join.
g4_disc <- g4_raw %>%
  dplyr::select(transcript_id_clean, region, g4mer_max) %>%
  inner_join(g4_motif %>% dplyr::select(transcript_id_clean, region, has_g4),
             by = "transcript_id_clean", relationship = "many-to-many") %>%
  dplyr::rename(score_region = region.x, motif_region = region.y) %>%
  group_by(score_region, motif_region) %>%
  summarise(auc_like = mean(has_g4 * g4mer_max) / mean(g4mer_max),
            gap = median(g4mer_max[has_g4]) - median(g4mer_max[!has_g4]),
            .groups = "drop")

own <- g4_disc %>% filter(score_region == motif_region) %>%
  dplyr::select(region = score_region, own_gap = gap)
best_other <- g4_disc %>% filter(score_region != motif_region) %>%
  group_by(region = score_region) %>% summarise(other_gap = max(gap), .groups = "drop")
disc <- inner_join(own, best_other, by = "region")
print(as.data.frame(disc %>% mutate(across(where(is.numeric), ~ round(., 4)))))

stopifnot("a region's G4 score tracks another region's motif better than its own - check the join" =
            all(disc$own_gap > disc$other_gap))
cat("G4 region-assignment check passed: each region tracks its own motif best\n")
```

```{r validate_peptide, eval = !params$pass_through}
stopifnot(
  "proline_fraction outside [0, 1]" =
    all(between(features$proline_fraction, 0, 1), na.rm = TRUE),
  "ppp_motif_density negative" = all(features$ppp_motif_density >= 0, na.rm = TRUE),
  "poly(A) runs negative" =
    all(features$max_consecutive_aaa >= 0, na.rm = TRUE) &&
    all(features$max_consecutive_aag >= 0, na.rm = TRUE),
  "net charge not finite" =
    all(is.finite(features$max_net_charge_30aa[!is.na(features$max_net_charge_30aa)]))
)

# Biological spot-checks. Range checks pass on a scrambled join; these do not.
# 1. Collagens are the canonical proline-rich proteins.
col_p  <- features$proline_fraction[str_detect(features$symbol, "^COL[0-9]")]
med_p  <- median(features$proline_fraction, na.rm = TRUE)
cat("proline_fraction: collagen median", round(median(col_p, na.rm = TRUE), 4),
    "vs transcriptome", round(med_p, 4), ", n =", sum(!is.na(col_p)), "\n")
stopifnot("collagens are not proline-rich - check the translation frame" =
            median(col_p, na.rm = TRUE) > med_p)

# 2. Ribosomal proteins are strongly basic, so they carry high positive windows.
rp_q   <- features$max_net_charge_30aa[str_detect(features$symbol, "^RP[LS][0-9]")]
med_q  <- median(features$max_net_charge_30aa, na.rm = TRUE)
cat("max_net_charge_30aa: ribosomal protein median", median(rp_q, na.rm = TRUE),
    "vs transcriptome", med_q, ", n =", sum(!is.na(rp_q)), "\n")
stopifnot("ribosomal proteins are not basic - check the charge mapping" =
            median(rp_q, na.rm = TRUE) > med_q)

# 3. PPP density must track proline content; if it does not, the motif count is wrong.
#    ~77% of transcripts carry no PPP at all, and that tie block holds the rank correlation
#    near 0.45, so the presence/absence contrast is the sharper statement of the same thing.
okp   <- complete.cases(features$ppp_motif_density, features$proline_fraction)
rho_p <- cor(features$ppp_motif_density[okp], features$proline_fraction[okp],
             method = "spearman")
has_ppp <- features$ppp_motif_density > 0
cat("ppp_motif_density vs proline_fraction: Spearman rho =", round(rho_p, 3),
    ", n =", sum(okp), "; zero-PPP fraction", round(mean(!has_ppp, na.rm = TRUE), 3), "\n")
cat("  median proline with PPP",
    round(median(features$proline_fraction[has_ppp], na.rm = TRUE), 4),
    "vs without", round(median(features$proline_fraction[!has_ppp], na.rm = TRUE), 4), "\n")
stopifnot(
  "PPP density does not track proline content" = rho_p > 0.3,
  "PPP-containing transcripts are not proline-rich - check the motif count" =
    median(features$proline_fraction[has_ppp], na.rm = TRUE) >
    median(features$proline_fraction[!has_ppp], na.rm = TRUE)
)

# 4. AAA and AAG are deliberately separate columns, not a merged lysine count.
cat("poly(A) tracks: max AAA run", max(features$max_consecutive_aaa, na.rm = TRUE),
    ", max AAG run", max(features$max_consecutive_aag, na.rm = TRUE),
    "; transcripts with an AAA run >= 3:",
    sum(features$max_consecutive_aaa >= 3, na.rm = TRUE), "\n")
stopifnot("AAA and AAG runs are identical - the codons were merged" =
            !identical(features$max_consecutive_aaa, features$max_consecutive_aag))
```

```{r validate_noderer, eval = !params$pass_through}
# The coordinate convention is the silent failure mode here: the lookup is exhaustive over
# ACGT, so a window built one nucleotide off still joins - it just returns some other
# context's efficiency. Three checks pin it down.

# 1. Coverage. A frame-shifted window would not carry ATG at 7..9 and would fall out as
#    unusable, so coverage is itself the frame check.
cov_noderer <- mean(!is.na(features$noderer_tis_efficiency))
cat("noderer_tis_efficiency coverage:", sprintf("%.1f%%", 100 * cov_noderer),
    "  median:", round(median(features$noderer_tis_efficiency, na.rm = TRUE), 1), "\n")
stopifnot(
  "Noderer coverage collapsed - check the window offset" = cov_noderer > 0.90,
  "Noderer efficiency outside the published 12-150 range" =
    all(between(features$noderer_tis_efficiency, 12, 150), na.rm = TRUE)
)

# 2. Published determinants: Noderer 2014 reports -3 and +4 as dominant, purine at -3 and G
#    at +4 being favourable. In a -6..+5 window -3 is index 4 and +4 is index 10.
nod_ctx <- noderer %>%
  mutate(purine_minus3 = str_sub(context, 4, 4) %in% c("A", "G"),
         g_plus4       = str_sub(context, 10, 10) == "G")
m3 <- nod_ctx %>% group_by(purine_minus3) %>%
  summarise(n = n(), mean_eff = round(mean(efficiency), 1), .groups = "drop")
g4 <- nod_ctx %>% group_by(g_plus4) %>%
  summarise(n = n(), mean_eff = round(mean(efficiency), 1), .groups = "drop")
print(as.data.frame(m3)); print(as.data.frame(g4))
stopifnot(
  "-3 purine does not raise Noderer efficiency" =
    m3$mean_eff[m3$purine_minus3] > m3$mean_eff[!m3$purine_minus3],
  "+4 G does not raise Noderer efficiency" =
    g4$mean_eff[g4$g_plus4] > g4$mean_eff[!g4$g_plus4]
)

# 3. The same determinants must be visible on OUR transcripts, not just in the reference
#    table - this is what confirms the join is correctly framed on real 5'UTRs.
our_ctx <- noderer_tbl %>% filter(usable, !is.na(noderer_tis_efficiency)) %>%
  mutate(purine_minus3 = str_sub(context, 4, 4) %in% c("A", "G"),
         g_plus4       = str_sub(context, 10, 10) == "G")
om3 <- our_ctx %>% group_by(purine_minus3) %>%
  summarise(n = n(), mean_eff = mean(noderer_tis_efficiency), .groups = "drop")
og4 <- our_ctx %>% group_by(g_plus4) %>%
  summarise(n = n(), mean_eff = mean(noderer_tis_efficiency), .groups = "drop")
cat("our transcripts, -3 purine:", round(om3$mean_eff[om3$purine_minus3], 1), "vs pyrimidine",
    round(om3$mean_eff[!om3$purine_minus3], 1),
    "| +4 G:", round(og4$mean_eff[og4$g_plus4], 1), "vs non-G",
    round(og4$mean_eff[!og4$g_plus4], 1), "\n")
stopifnot(
  "-3 purine effect absent on our transcripts - check the window offset" =
    om3$mean_eff[om3$purine_minus3] > om3$mean_eff[!om3$purine_minus3],
  "+4 G effect absent on our transcripts - check the window offset" =
    og4$mean_eff[og4$g_plus4] > og4$mean_eff[!og4$g_plus4]
)

# 4. Two independent readings of the same quantity - a 1987 frequency table and a FACS-seq
#    measurement - must at least agree in SIGN. They agree only weakly (rho ~0.15), which is
#    itself worth knowing: the inventory's "never model both together" was written assuming
#    near-duplication, and on these transcripts they are far from duplicates.
ok  <- complete.cases(features$noderer_tis_efficiency, features$kozak_pwm_score_v2)
rho <- cor(features$noderer_tis_efficiency[ok], features$kozak_pwm_score_v2[ok],
           method = "spearman")
cat("noderer_tis_efficiency vs kozak_pwm_score_v2: Spearman rho =", round(rho, 3),
    ", n =", sum(ok), "\n")
stopifnot("Noderer and the Kozak PWM disagree in sign" = rho > 0)

# Tier agreement, reported not asserted: efficiency rises steeply across kozak_score tiers
# 0-2, then falls back at tier 3, which holds only ~99 transcripts.
print(features %>% filter(!is.na(noderer_tis_efficiency), !is.na(kozak_score)) %>%
        group_by(kozak_score) %>%
        summarise(n = n(), median_eff = round(median(noderer_tis_efficiency), 1),
                  .groups = "drop") %>% as.data.frame())
```

## How much did each correction move?

```{r correction_deltas, eval = !params$pass_through}
pairs_tbl <- tribble(~current, ~corrected,
  "kozak_pwm_score", "kozak_pwm_score_v2", "gc3", "gc3_internal",
  "tai", "tai_gtrnadb", "cai", "cai_internal",
  "fop", "fop_internal", "csc", "csc_internal")

deltas <- pairs_tbl %>%
  mutate(pearson  = map2_dbl(current, corrected,
           ~ cor(features[[.x]], features[[.y]], use = "complete.obs")),
         spearman = map2_dbl(current, corrected,
           ~ cor(features[[.x]], features[[.y]], use = "complete.obs", method = "spearman")),
         median_current   = map_dbl(current,   ~ median(features[[.x]], na.rm = TRUE)),
         median_corrected = map_dbl(corrected, ~ median(features[[.x]], na.rm = TRUE)),
         n = map2_int(current, corrected,
           ~ sum(complete.cases(features[[.x]], features[[.y]]))))
print(as.data.frame(deltas))

cat("\nCorrelation with gc3_internal:\n")
cat("  tai (ours)     :", round(cor(features$tai,         features$gc3_internal, use = "complete.obs"), 3), "\n")
cat("  tai_gtrnadb    :", round(cor(features$tai_gtrnadb, features$gc3_internal, use = "complete.obs"), 3), "\n")
```

# 2. Annotate every column

`block` decides whether a column can ever be a feature; `family` groups correlated columns so
SHAP can be summed over them exactly.

```{r assign_block}
assign_block <- function(x) {
  case_when(
    # Identifiers. Matches 07's id_cols plus its id_pattern_cols regex.
    str_detect(x, "gene_id|transcript_id|^symbol$")                 ~ "id",
    # Outcome and anything derived from it. Never a feature: these define the target.
    x %in% c("te_lfc", "te_padj", "te_lfc_bin")                     ~ "label",
    str_detect(x, "^sig_|^indiv_te_|^delta_te_")                    ~ "label",
    # eIF4E-derived. Circular for a 4E model, so separable on its own.
    str_detect(x, "4e_resistance|^eif4e_ip|^is_4e_|_4e_|^eif4e_")   ~ "eif4e",
    # Measured in some other experiment. Will not exist for most new datasets.
    str_detect(x, "^clip_|^cnot3_|^zhu2024_|^hia2026_|^karner2026_|^dap5_|^is_dap5_") ~ "external",
    # CDS eligibility flags: QC about whether a codon feature is meaningful, not a feature.
    str_detect(x, "^cds_(frame_ok|has_start|has_stop|complete)$")   ~ "qc",
    # Computable from the transcript sequence alone. These are what transfer.
    TRUE                                                            ~ "intrinsic"
  )
}

assign_family <- function(x) {
  case_when(
    str_detect(x, "gene_id|transcript_id|^symbol$")              ~ "id",
    str_detect(x, "^cds_(frame_ok|has_start|has_stop|complete)$") ~ "cds_eligibility",
    x %in% c("te_lfc", "te_padj", "te_lfc_bin")                  ~ "outcome",
    str_detect(x, "^sig_|^indiv_te_|^delta_te_")                 ~ "outcome_derived",
    str_detect(x, "_length$|^log2_.*_length$")                   ~ "length",
    str_detect(x, "^term_")                                      ~ "termination",
    str_detect(x, "_gc$|^gc3")                                   ~ "gc",
    # Order matters: the CDS and 3'UTR structure columns carry a suffix, the 5'UTR ones
    # do not, so the suffixed patterns must be tested first.
    # The region tag is not always the LAST token: tco_struct_accessibility_cds_mean ends in
    # _mean. Anchoring on "$" filed six CDS/3'UTR columns under structure_utr5, which silently
    # broke the exact family summation feature_annotation.csv exists for.
    str_detect(x, "^tco_struct_.*_cds(_|$)")                     ~ "structure_cds",
    str_detect(x, "^tco_struct_.*_utr3(_|$)")                    ~ "structure_utr3",
    str_detect(x, "^tco_struct_")                                ~ "structure_utr5",
    str_detect(x, "^csc|^cai|^tai|^fop")                         ~ "codon_optimality",
    str_detect(x, "^kozak_|^noderer_")                           ~ "initiation",
    x %in% c("proline_fraction", "ppp_motif_density",
             "max_net_charge_30aa")                               ~ "nascent_peptide",
    x %in% c("max_consecutive_aaa", "max_consecutive_aag")        ~ "polya_track",
    str_detect(x, "^g4mer_")                                     ~ "g4",
    str_detect(x, "^clip_")                                      ~ "clip",
    str_detect(x, "^zhu2024_|^hia2026_|^karner2026_|^cnot3_")    ~ "stability_external",
    str_detect(x, "^dap5_|^is_dap5_")                            ~ "dap5",
    str_detect(x, "4e")                                          ~ "eif4e",
    TRUE                                                         ~ "other"
  )
}

# variant pairs a corrected column with the one it supersedes. 02_ picks one member of each
# pair via variant_set; both together in one forest would split credit between near-duplicates.
variant_pairs <- tribble(
  ~current,           ~corrected,
  "kozak_pwm_score",  "kozak_pwm_score_v2",
  "gc3",              "gc3_internal",
  "tai",              "tai_gtrnadb",
  "cai",              "cai_internal",
  "fop",              "fop_internal",
  "csc",              "csc_internal",
  "hia2026_dhx29_occupancy", "hia2026_dhx29_occupancy_tx",   # F3: gene-level -> transcript
  "clip_cap_binding_count",  "clip_cap_binding_count_v2"     # F4: drops a duplicated dataset
)

annotation <- tibble(feature = colnames(features)) %>%
  mutate(
    block  = assign_block(feature),
    family = assign_family(feature),
    variant = case_when(
      feature %in% variant_pairs$current   ~ "current",
      feature %in% variant_pairs$corrected ~ "corrected",
      TRUE                                 ~ "single"
    ),
    source = if_else(feature %in% variant_pairs$corrected,
                     "01_ (corrected, see 00_)", baseline_rds),
    definition = NA_character_
  )

# Deprecated features stay in the matrix but are not selected by default. Positional CSC is
# dropped by decision: csc_q1-q4 and csc_first75 are five near-collinear arms of one quantity,
# which is the worst case for credit-splitting and the thing SHAP family summation exists to
# avoid. They are flagged rather than deleted so the baseline gate stays runnable - the
# baseline's 51 features include all five.
deprecated_features <- c("csc_q1", "csc_q2", "csc_q3", "csc_q4", "csc_first75",
                         broken_structure,            # F2 + F5, see the structure repairs
                         "clip_lee_complete_media_bound")   # F4, duplicate of glu_dep

annotation <- annotation %>%
  mutate(deprecated = feature %in% deprecated_features,
         added_in   = if_else(feature %in% baseline_cols, "baseline", "tco"))
cat("Columns added by this notebook:", sum(annotation$added_in == "tco"), "\n")
cat("Deprecated (kept in the matrix, not selected by default):",
    sum(annotation$deprecated), "-", paste(deprecated_features, collapse = ", "), "\n")

# A corrected column is only meaningful if its partner is present, and vice versa.
if (!isTRUE(params$pass_through)) {
  stopifnot("a variant pair is incomplete" =
    all(c(variant_pairs$current, variant_pairs$corrected) %in% annotation$feature))
}

annotation %>% count(block, family, name = "n") %>% arrange(block, desc(n)) %>% print(n = 40)
```

```{r annotate_definitions}
# Record the definitions that differ from the collaborator repo, so the unit and the convention
# travel with the column rather than living only in a notebook comment. Audited in 00_.
definitions <- tribble(
  ~feature,           ~definition,
  "utr5_gc",          "percent (0-100); denominator includes ambiguous bases",
  "utr3_gc",          "percent (0-100); denominator includes ambiguous bases",
  "cds_gc",           "percent (0-100); denominator includes ambiguous bases",
  "transcript_gc",    "percent (0-100); length-weighted mean of the three regional GCs",
  "gc3",              "percent (0-100); INCLUDES the stop codon - to be fixed, see 00_",
  "log2_cds_length",  "log2(x + 1); +1 applied uniformly across all three regions",
  "log2_utr5_length", "log2(x + 1)",
  "log2_utr3_length", "log2(x + 1)",
  "kozak_pwm_score",  "Kozak 1987 log-odds; PWM is WRONG at -6 and -3 - to be fixed, see 00_",
  "tai",              "coRdon tAI; collaborator variant to be adopted, see 00_",
  "csc",              "iCodon::predict_stability('human'), whole stop-free CDS",
  "kozak_pwm_score_v2", "Kozak 1987 Table 1 log-odds (imported file); NA unless AUG-anchored",
  "gc3_internal",     "percent (0-100); internal codons: initiator and stop removed",
  "cai_internal",     "Sharp & Li CAI, RPL/RPS reference; internal codons only",
  "fop_internal",     "FOP, top-10%-TE reference; internal codons only",
  "csc_internal",     "iCodon::predict_stability('human') on the internal-codon CDS",
  "tai_gtrnadb",      "dos Reis-style tAI; GtRNAdb hg38 copy number + collaborator wobble v1",
  "noderer_tis_efficiency",
    "Noderer 2014 FACS-seq TIS efficiency (12-150), -6..+5 lookup; NA unless AUG-anchored",
  "proline_fraction",    "fraction (0-1) of internal-codon residues that are proline",
  "ppp_motif_density",   "PPP occurrences per 100 residues; overlapping matches counted",
  "max_net_charge_30aa", "max (K+R) - (D+E) over any 30-aa window; whole chain if shorter",
  "max_consecutive_aaa", "longest in-frame run of AAA lysine codons",
  "max_consecutive_aag", "longest in-frame run of AAG lysine codons",
  "term_stop_TAA",        "1 if the annotated stop codon is TAA, else 0",
  "term_stop_TAG",        "1 if the annotated stop codon is TAG, else 0",
  "term_stop_TGA",        "1 if the annotated stop codon is TGA, else 0",
  "term_plus4_A",         "1 if the base immediately 3' of the stop (+4) is A, else 0",
  "term_plus4_C",         "1 if the base immediately 3' of the stop (+4) is C, else 0",
  "term_plus4_G",         "1 if the base immediately 3' of the stop (+4) is G, else 0",
  "term_plus4_T",         "1 if the base immediately 3' of the stop (+4) is T, else 0",
  "term_window_gc",       "percent GC of the 9 nt stop + 6 nt window (+1..+9)",
  "term_interval_codons", "codons from the annotated stop to the next in-frame stop in the 3'UTR",
  "term_readthrough_cridge",
    "Cridge 2018 Table S1 percent readthrough for this +1..+6 context (192-context lookup)",
  "term_log1p_readthrough", "log1p(term_readthrough_cridge)",
  "g4mer_max_resid_utr5", "G4mer max, residual on log2 5'UTR length + utr5_gc (stride 10)",
  "g4mer_max_resid_cds",  "G4mer max, residual on log2 CDS length + cds_gc (stride 20)",
  "g4mer_max_resid_utr3", "G4mer max, residual on log2 3'UTR length + utr3_gc (stride 20)",
  "g4mer_mean_utr5",      "mean G4mer window score, 5'UTR; no length bias",
  "g4mer_mean_cds",       "mean G4mer window score, CDS; no length bias",
  "g4mer_mean_utr3",      "mean G4mer window score, 3'UTR; no length bias",
  "g4mer_frac_above_utr5", "fraction of 5'UTR windows above the rG4 threshold",
  "g4mer_frac_above_cds",  "fraction of CDS windows above the rG4 threshold",
  "g4mer_frac_above_utr3", "fraction of 3'UTR windows above the rG4 threshold",
  "hia2026_dhx29_occupancy_tx",
    "Hia 2026 DHX29 ribosome occupancy, joined on transcript_id_clean (the source key)",
  "clip_cap_binding_count_v2",
    "cap-binding CLIP datasets with evidence, counting the 2 distinct conditions (0-2)",
  "te_lfc",
    "WARNING: this is the siCTRL hypoxia-vs-normoxia 1hr contrast (translation_categories_1hr.csv), NOT an si3d knockdown contrast. Never read it as the eIF3d TE change."
)

definitions <- definitions %>% filter(feature %in% annotation$feature)
annotation <- annotation %>%
  rows_update(definitions, by = "feature", unmatched = "error")

# Every tco_struct_* column is region-isolated RNAplfold (-W 150 -L 100), NOT the
# collaborator's whole-transcript fold. The tco_ prefix is what keeps the two apart.
# The RNAplfold parameters DIFFER BY REGION, which neither document said. Verified by
# re-running RNAplfold 2.7.2 on ENST00000000233 and reproducing the stored _lunp files
# bit-for-bit: 5'UTR only under -W 80 -L 40 -u 30, CDS and 3'UTR only under -W 150 -L 100
# -u 30. The annotation previously claimed -W 150 -L 100 for all three, and 01d_'s inline
# docs claimed -W 80 -L 40 -u 30 for all three; each was right about its own region.
#
# CONSEQUENCE: a larger window sees more folding context, so accessibility is NOT comparable
# ACROSS regions - the same caveat the g4 family carries for its differing stride. Compare
# tco_struct_*_utr5 with tco_struct_*_utr5, never with _cds.
struct_params <- function(x) if_else(
  str_detect(x, "_cds(_|$)|_utr3(_|$)"),
  "RNAplfold -W 150 -L 100 -u 30, region folded in ISOLATION; NOT comparable to the 5'UTR columns (different window)",
  "RNAplfold -W 80 -L 40 -u 30, region folded in ISOLATION; NOT comparable to the CDS/3'UTR columns (different window)")

annotation <- annotation %>%
  mutate(definition = if_else(str_starts(feature, "tco_struct_"),
                              struct_params(feature), definition))

cat("Columns carrying an explicit definition:", sum(!is.na(annotation$definition)), "\n")
```

# 3. Annotation integrity

The annotation and the matrix must never drift apart. These are the checks that guarantee
`02_` cannot silently select the wrong columns.

```{r validate_annotation}
stopifnot(
  "annotation does not cover the matrix exactly" =
    setequal(annotation$feature, colnames(features)),
  "annotation has duplicate rows" =
    !any(duplicated(annotation$feature)),
  "a column has no block" =
    all(!is.na(annotation$block)),
  "a column has no family" =
    all(!is.na(annotation$family)),
  "unrecognised block" =
    all(annotation$block %in% c("id", "label", "intrinsic", "external", "eif4e",
                                "technical", "qc")),
  "no family may be assigned 'other'" =
    !any(annotation$family == "other"),
  "unrecognised variant" =
    all(annotation$variant %in% c("single", "current", "corrected")),
  # F6 regression guard: the six _cds/_utr3 columns ending in _mean/_min/_sd were filed under
  # structure_utr5 for as long as the anchor was "$". 18 columns, 6 per region, always.
  "structure families are not 6/6/6 - the region regex is wrong again" =
    identical(as.integer(table(factor(
      annotation$family[str_starts(annotation$feature, "tco_struct_")],
      levels = c("structure_utr5", "structure_cds", "structure_utr3")))),
      c(6L, 6L, 6L)),
  "a deprecated feature is missing from the matrix" =
    all(deprecated_features %in% annotation$feature),
  "unrecognised added_in" =
    all(annotation$added_in %in% c("baseline", "tco"))
)
cat("Annotation integrity: OK\n")

# Leakage guard. What matters is not which block an outcome-derived column lands in, but that
# it is not SELECTABLE - i.e. never in intrinsic/external, the only blocks 02_ draws features
# from. (te_4e_resistance_lfc is outcome-shaped but blocked as eif4e, which is also fine.)
selectable_blocks <- c("intrinsic", "external")
leak <- annotation %>%
  filter(block %in% selectable_blocks,
         str_detect(feature, "^te_|^sig_|^indiv_te_|^delta_te_"))
stopifnot("an outcome-derived column is selectable as a feature" = nrow(leak) == 0)
cat("Leakage guard: no outcome-derived column is selectable\n")
```

## Structural invariants

Every bug the independent audit found was a *class* of bug, not a one-off. These turn each
class into a standing check, so the same shape of error cannot return silently. They run on
the modelled columns only (identifiers, labels and deprecated columns are exempt).

```{r feature_invariants, eval = !params$pass_through}
# Mutations that target the structural invariants themselves.
if (MUT == "duplicate_feature")                       # F4 class
  features$g4mer_mean_cds <- features$g4mer_mean_utr5
if (MUT == "constant_feature")                        # F5 class
  features$proline_fraction <- 0.05
if (MUT == "length_proxy_feature")                    # F2 class
  features$term_interval_codons <- features$utr3_length
if (MUT == "golden_drift")                            # silent per-transcript drift
  features$cds_gc <- features$cds_gc + 1e-6

modelled <- annotation %>%
  filter(block %in% c("intrinsic", "external"), !deprecated) %>% pull(feature)
num_mod <- modelled[map_lgl(modelled, ~ is.numeric(features[[.x]]))]
vals <- features[, num_mod]

# A. No two modelled features may be bit-identical. F4: clip_lee_complete_media_bound was a
#    byte-for-byte copy of clip_lee_glu_dep_bound, and a count summing both double-counted it.
dup_pairs <- list()
for (i in seq_along(num_mod)) for (j in seq_len(i - 1)) {
  if (identical(vals[[i]], vals[[j]])) dup_pairs[[length(dup_pairs) + 1]] <-
    paste(num_mod[j], "==", num_mod[i])
}
cat("A. identical modelled columns:", length(dup_pairs), "\n")
if (length(dup_pairs)) cat("   ", paste(unlist(dup_pairs), collapse = "; "), "\n")

# B. No modelled feature may be constant. F5: three accessibility_*_min columns were
#    identically 0, because min() was taken after NA had been replaced with 0.
sds <- map_dbl(num_mod, ~ stats::sd(features[[.x]], na.rm = TRUE))
const_feats <- num_mod[!is.na(sds) & sds == 0]
cat("B. constant modelled columns:", length(const_feats),
    if (length(const_feats)) paste("-", paste(const_feats, collapse = ", ")) else "", "\n")

# C. Nothing outside the length family may be a near-perfect proxy for a region length.
#    F2: num_structured_regions* was Spearman ~1.000 with its own region length, and its
#    "count" exceeded that length, so length entered the forest three more times in disguise.
len_cols <- intersect(c("utr5_length", "cds_length", "utr3_length", "total_length"),
                      colnames(features))
proxies <- list()
for (f in setdiff(num_mod, len_cols)) {
  if (annotation$family[match(f, annotation$feature)] == "length") next
  for (L in len_cols) {
    ok <- stats::complete.cases(features[[f]], features[[L]])
    if (sum(ok) < 100) next
    r <- suppressWarnings(cor(features[[f]][ok], features[[L]][ok], method = "spearman"))
    if (!is.na(r) && abs(r) > 0.999)
      proxies[[length(proxies) + 1]] <- sprintf("%s ~ %s (rho %.5f)", f, L, r)
  }
}
cat("C. length proxies outside the length family:", length(proxies), "\n")
if (length(proxies)) cat("   ", paste(unlist(proxies), collapse = "; "), "\n")

stopifnot(
  "two modelled features are bit-identical"          = length(dup_pairs) == 0,
  "a modelled feature is constant"                   = length(const_feats) == 0,
  "a non-length feature is a near-perfect length proxy" = length(proxies) == 0
)

# D. Sentinel blocks. A continuous feature where one exact value is shared by a large block of
#    rows is how F1 hid: 1,689 transcripts zero-filled rather than NA, which defeats every
#    NA-based guard downstream. Asserted, with an explicit allowlist - a feature is exempt only
#    by being named here with a reason, never by silently slipping under a threshold.
ZERO_HEAVY_OK <- c(
  "ppp_motif_density"     = "PPP is genuinely absent from ~77% of proteins",
  "g4mer_frac_above_utr5" = "most 5'UTRs have no window above the rG4 threshold",
  "g4mer_frac_above_cds"  = "as above, CDS",
  "g4mer_frac_above_utr3" = "as above, 3'UTR",
  "cnot3_ribo_enrichment" = "-log10 FDR, floored at 0 for non-enriched genes"
)
cat("D. continuous columns where one exact value covers >5% of rows:\n")
sentinel_hits <- character(0)
for (f in num_mod) {
  v <- features[[f]]; v <- v[!is.na(v)]
  if (length(unique(v)) < 100) next                 # not continuous
  tb <- sort(table(v), decreasing = TRUE)
  if (tb[1] / length(v) > 0.05) {
    exempt <- f %in% names(ZERO_HEAVY_OK)
    cat(sprintf("     %-42s value %s in %.1f%% of rows%s\n", f, names(tb)[1],
                100 * tb[1] / length(v),
                if (exempt) paste0("  [expected: ", ZERO_HEAVY_OK[[f]], "]") else "  <- UNEXPECTED"))
    if (!exempt) sentinel_hits <- c(sentinel_hits, f)
  }
}
if (!length(sentinel_hits)) cat("     no unexpected sentinel blocks\n")
stopifnot("an unexplained sentinel block appeared - is a NA being filled with a constant?" =
            length(sentinel_hits) == 0)

# E. Join level. F3 was a transcript-keyed source joined at GENE level, so the transcript we
#    model received a different isoform's value.
#
#    The obvious fingerprint - "a gene-level join is constant within gene" - DOES NOT WORK
#    here, and that is worth stating: this matrix holds exactly one transcript per gene, so
#    there are no sibling isoforms to compare. That is precisely why F3 was silent. A
#    gene-level join does not fail or duplicate; it quietly returns whichever isoform the
#    source happened to carry.
#
#    So the only real check is against the SOURCE, which is possible for the external files
#    this notebook reads itself. Inherited external columns are built in
#    code/predictive_modeling/ and cannot be key-checked from here - see FEATURES.md, whose
#    "assigned in" column names the notebook where each one is built.
stopifnot("the matrix is no longer one transcript per gene - re-derive invariant E" =
            !any(duplicated(features$gene_id_clean)))

occ_check <- features %>%
  dplyr::select(transcript_id_clean, stored = hia2026_dhx29_occupancy_tx) %>%
  left_join(dhx29_occ %>% dplyr::select(transcript_id_clean,
                                        src = dhx29_occupancy_score),
            by = "transcript_id_clean")
occ_dev <- max(abs(occ_check$stored - occ_check$src), na.rm = TRUE)
cat("E. hia2026_dhx29_occupancy_tx vs a fresh transcript-level join: max |dev| =",
    signif(occ_dev, 3), ", non-NA", sum(!is.na(occ_check$stored)),
    "(gene-level original had", sum(!is.na(features$hia2026_dhx29_occupancy)), ")\n")
stopifnot(
  "the corrected dhx29 occupancy does not match a transcript-level join" = occ_dev < 1e-12,
  "the corrected dhx29 occupancy has NA where the transcript-level source has a value" =
    !any(is.na(occ_check$stored) & !is.na(occ_check$src))
)
cat("   external columns inherited from predictive_modeling cannot be key-checked here;",
    "FEATURES.md names where each is built\n")
```

## Golden transcripts

A frozen snapshot of 20 transcripts x every modelled feature, committed at
`golden_transcripts.csv`. Range checks pass on a scrambled join and family statistics pass on
a drifted value; this catches "that number is not what it was" whatever the cause. The
transcripts deliberately include the awkward cases: the named spot-check genes, the F1 group
whose structure is NA, truncated CDS and non-AUG starts.

Regenerate **on purpose** with `01d_golden_transcripts.R` when a feature is legitimately
corrected - the commit diff then shows exactly which transcripts and columns moved. A silent
change fails here instead.

```{r golden_transcripts, eval = !params$pass_through}
golden_f <- here("code", "translational_control_overview", "golden_transcripts.csv")
if (!file.exists(golden_f)) {
  cat("No golden file yet - create it with 01d_golden_transcripts.R\n")
} else {
  golden <- read_csv(golden_f, show_col_types = FALSE)
  now <- features %>%
    filter(transcript_id_clean %in% golden$transcript_id_clean) %>%
    dplyr::select(transcript_id_clean, any_of(unique(golden$feature))) %>%
    pivot_longer(-transcript_id_clean, names_to = "feature", values_to = "value_now")

  cmp <- golden %>% left_join(now, by = c("transcript_id_clean", "feature"))
  missing_col <- cmp %>% filter(is.na(value_now) & !is.na(value))
  na_flip     <- cmp %>% filter(!is.na(value_now) & is.na(value))
  moved <- cmp %>% filter(!is.na(value), !is.na(value_now),
                          abs(value - value_now) > 1e-9)

  cat("Golden check:", n_distinct(golden$transcript_id_clean), "transcripts x",
      n_distinct(golden$feature), "features\n")
  cat("  values that moved:", nrow(moved),
      " | became NA:", nrow(missing_col), " | were NA, now present:", nrow(na_flip), "\n")
  if (nrow(moved))
    print(moved %>% arrange(desc(abs(value - value_now))) %>% head(10) %>% as.data.frame())

  stopifnot(
    "a golden feature is missing from the matrix" =
      all(unique(golden$feature) %in% colnames(features)),
    "golden transcript values changed - if this was intended, re-run 01d_golden_transcripts.R" =
      nrow(moved) == 0 && nrow(missing_col) == 0 && nrow(na_flip) == 0
  )
  cat("  golden snapshot matches\n")
}
```

## What `te_lfc` actually is (audit finding F7)

`te_lfc` / `te_padj` / `te_lfc_bin` are **not** an si3d knockdown contrast. They are a
bit-for-bit copy of `output/translation_categories_1hr.csv`, the siCTRL
hypoxia-vs-normoxia 1hr contrast. Two independent signs of it: EIF3D's own `te_lfc` is
**negative** (the documented direction check requires strongly positive under its own
knockdown), and `sig_hypoxia_3d_promotes_TE_1hr_lfc0.5` flags 1,446 transcripts of which only
11 have `te_lfc > 0.5`.

No model is affected - these are `block = label` and `02_` asserts them out of the feature set
- but anything reading `te_lfc` as "the eIF3d TE change" is wrong. The identity is asserted
here so the definition cannot go stale without failing.

```{r verify_te_lfc_provenance, eval = !params$pass_through}
hyp_1hr <- read_csv(here("output", "translation_categories_1hr.csv"), show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::select(gene_id_clean, src_te_lfc = te_lfc)

chk <- features %>% dplyr::select(gene_id_clean, te_lfc) %>%
  inner_join(hyp_1hr, by = "gene_id_clean")
ok  <- stats::complete.cases(chk$te_lfc, chk$src_te_lfc)
dev <- max(abs(chk$te_lfc[ok] - chk$src_te_lfc[ok]))
cat("te_lfc vs translation_categories_1hr.csv: max |dev| =", signif(dev, 3),
    ", n =", sum(ok), "\n")
cat("EIF3D own te_lfc:", round(features$te_lfc[features$symbol == "EIF3D"], 3),
    "(negative confirms this is NOT the si3d contrast)\n")
stopifnot(
  "te_lfc is no longer the hypoxia-response contrast - update its definition" = dev < 1e-9
)
```

# 4. Reproduction gate: does the annotation rebuild the baseline 51?

Notebook 07 arrives at its feature set by a regex cascade. This notebook arrives at one by a
block join. They must agree exactly, or `02_` is not modelling the same thing as the baseline.

```{r reproduce_baseline}
baseline_importance <- read_csv(
  here("output", "predictive_modeling",
       paste0("rf_classification_importance_eif33d_promotes_hypoxia_1hr_lfc0.5_",
              "dhx29_kd_feature_clipaggregate_no_boruta_no4e_noOwnTE.csv")),
  show_col_types = FALSE)

# Selection rule, mirroring the baseline run (no4e, noOwnTE, clip aggregate).
aggregate_clip <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                    "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                    "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")

# The gate reproduces the BASELINE, so it takes the current member of every variant pair.
# The gate reproduces the BASELINE: baseline-era columns only, current variants, and the five
# deprecated positional-CSC features, which the baseline included.
selected <- annotation %>%
  filter(block %in% c("intrinsic", "external"), variant != "corrected",
         added_in == "baseline") %>%
  filter(!(family == "clip" & !feature %in% aggregate_clip)) %>%   # aggregate CLIP mode
  filter(!str_detect(feature, "^(utr5|utr3|cds|total)_length$")) %>%  # log2 versions kept
  pull(feature)

# Sparsity filter is data-dependent, so it is applied on values, not on names.
# 07 computes it on the labelled genes; on this matrix the same 7 columns fall out.
non_na_frac <- map_dbl(selected, ~ mean(!is.na(unlist(features[[.x]]))))
sparse <- selected[non_na_frac < 0.5]
selected <- setdiff(selected, sparse)

cat("Selected by block join:", length(selected), "\n")
cat("Baseline importance CSV:", nrow(baseline_importance), "\n")
cat("Dropped as sparse (<50% non-NA):", length(sparse), "-",
    paste(sparse, collapse = ", "), "\n")

# The frozen baseline CSV predates the tco_struct_* rename, so the gate compares on the
# baseline's spelling. This is a rename and nothing else: the columns, their values and their
# order are untouched, and the count is asserted on both sides.
selected_baseline_names <- sub("^tco_struct_", "struct_", selected)

if (!setequal(selected_baseline_names, baseline_importance$feature)) {
  cat("only in block join:",
      paste(setdiff(selected_baseline_names, baseline_importance$feature), collapse = ", "),
      "\nonly in baseline  :",
      paste(setdiff(baseline_importance$feature, selected_baseline_names), collapse = ", "),
      "\n")
}
stopifnot(
  "the rename changed how many struct columns are selected" =
    sum(str_starts(selected, "tco_struct_")) ==
    sum(str_starts(selected_baseline_names, "struct_")),
  "block join does not reproduce the baseline feature set" =
    setequal(selected_baseline_names, baseline_importance$feature)
)
cat("\nGATE PASSED: the annotation reproduces the baseline 51 features exactly\n")
```

# 5. Write

```{r write_outputs, eval = params$mutate_input == ""}
matrix_out <- features
cat(if (isTRUE(params$pass_through))
      "pass_through = TRUE: baseline copied unchanged\n"
    else
      "pass_through = FALSE: corrected variants added alongside the originals\n")

saveRDS(matrix_out, file.path(out_dir, "feature_matrix_tco.rds"))
write_csv(annotation, file.path(out_dir, "feature_annotation.csv"))

stopifnot("written matrix lost rows" = nrow(matrix_out) == nrow(features))

cat("\nWrote:\n")
cat(" ", file.path(out_dir, "feature_matrix_tco.rds"), "-",
    nrow(matrix_out), "x", ncol(matrix_out), "\n")
cat(" ", file.path(out_dir, "feature_annotation.csv"), "-",
    nrow(annotation), "rows\n")

annotation %>% count(block, name = "columns") %>% arrange(desc(columns)) %>% print()
```

# Summary

The annotation reproduces the baseline feature set exactly by block join, so `02_rf_model.Rmd`
can drop notebook 07's `feature_set` branches without changing what is modelled. `block` makes
the intrinsic-vs-external split a data decision you can revise by editing a CSV, and `family`
is what lets `03_shap.Rmd` sum SHAP over correlated columns exactly.

**State of the build.** Six definitions are corrected (`kozak_pwm_score_v2`, `gc3_internal`,
`cai_internal`, `fop_internal`, `csc_internal`, `tai_gtrnadb`), each kept *alongside* the
column it supersedes so the baseline stays reproducible. The `tco_struct_*` rename is applied.
Five families have been added: `initiation` (Noderer TIS efficiency), `termination` (stop
context + Cridge readthrough), `nascent_peptide`, `polya_track` and `g4` (G4mer, merged from
`output/g4mer/` - no inference re-run).

Every added family must be judged against its own shuffled null via `02b_`/`02c_`, never
against zero: impurity importance is non-negative, so an uninformative family never scores
zero. Of the five, only `g4` clears that floor on AUC, and none beats the no-family arm.

Remaining from `00_`'s inventory: the detectability block
(`control_ribo_frame0_fraction`, `control_ribo_cds_coverage_fraction`), which needs riboseq
coverage rather than a lookup and is deferred.
