---
title: "Build the translational-control feature matrix"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  pass_through: FALSE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Single auditable builder for `code/translational_control_overview/`. Replaces the nine-notebook
chain (`01_` -> `01b_` -> `01d_` -> `22_` -> `23_` -> `01i_` -> `01j_` -> `01k_` -> `87_`), two
of which overwrite `feature_matrix_external_stability.rds` in place. Nothing under
`output/predictive_modeling/` is read-modify-written here; this notebook only reads.

Two artifacts:

| Output | Role |
|---|---|
| `feature_matrix_tco.rds` | the matrix, keyed on `transcript_id_clean` |
| `feature_annotation.csv` | one row per column: `block`, `family`, `source`, `definition` |

**`feature_annotation.csv` is the central artifact.** It replaces notebook 07's eight
`if (feature_set == ...)` branches and its regex cascade with an explicit, inspectable table.
`02_rf_model.Rmd` selects features by joining against it (`include_blocks`), and `03_shap.Rmd`
aggregates SHAP by `family` - which is exact, because SHAP is additive.

## `pass_through`

With `params$pass_through = TRUE` the matrix is copied from the baseline **unchanged** and only
the annotation is new. That is deliberate and is the first thing to run: it lets
`02_rf_model.Rmd` be checked against the frozen baseline before any definition is corrected or
any feature added. Until that reproduction gate passes, no later difference is interpretable -
a fork that silently fails to rebuild the baseline looks exactly like a real feature effect.

Definition fixes (Kozak PWM, GC3 stop codon, `tco_struct_*` rename) and the new feature
families land in later passes, with `pass_through = FALSE`.

```{r libraries}
library(tidyverse)
library(here)
```

```{r load_baseline}
baseline_rds <- "feature_matrix_dhx29_kd_feature.rds"
features <- readRDS(here("output", "predictive_modeling", baseline_rds))

out_dir <- here("output", "translational_control_overview")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Recorded before anything is added, so the gate can always tell baseline-era columns from
# ones this notebook introduced. Self-maintaining: every new family is "tco" automatically.
baseline_cols <- colnames(features)

cat("Baseline matrix:", baseline_rds, "\n")
cat(nrow(features), "transcripts x", ncol(features), "columns\n")

stopifnot(
  "matrix is not keyed on transcript_id_clean" =
    "transcript_id_clean" %in% colnames(features),
  "transcript_id_clean is not unique" =
    !any(duplicated(features$transcript_id_clean)),
  "gene identifier trio is incomplete" =
    all(c("gene_id", "gene_id_clean", "symbol") %in% colnames(features))
)
```

# 1. Corrected feature variants

Three columns are known to be wrong or superseded (audited in `00_`). Each correction is added
as a **parallel column** rather than overwriting the original, so `04_` can measure how much
each one moves the ranking instead of only showing the endpoint. `02_rf_model.Rmd` selects one
of each pair via `variant_set`; the two members of a pair are near-duplicates and must never
enter the same forest, where they would split credit and hide the effect being measured.

| current | corrected | why |
|---|---|---|
| `kozak_pwm_score` | `kozak_pwm_score_v2` | hand-typed PWM is wrong at -6 and -3 |
| `gc3` | `gc3_internal` | initiator and stop codon were both counted |
| `cai` | `cai_internal` | initiator AUG was counted |
| `fop` | `fop_internal` | initiator AUG was counted |
| `csc` | `csc_internal` | iCodon was given the stop-containing CDS |
| `tai` | `tai_gtrnadb` | collaborator variant: explicit wobble table + GtRNAdb copy number |

```{r corrected_features, eval = !params$pass_through}
ref_dir <- here("accessories", "collaborator_reference")
seqs <- readRDS(here("output", "predictive_modeling", "precomputed_kozak_sequences.rds"))

utr5 <- seqs$utr5_seq %>% filter(transcript_id_clean %in% features$transcript_id_clean)
cds  <- seqs$cds_seq  %>% filter(transcript_id_clean %in% features$transcript_id_clean)
stopifnot("sequences missing for some transcripts" =
            nrow(utr5) == nrow(features) && nrow(cds) == nrow(features))

## --- kozak_pwm_score_v2 -------------------------------------------------------------
# Published Kozak 1987 Table 1, imported as a file. Never transcribed: a hand-typed copy is
# what produced the defect this replaces.
kozak_pub <- read_tsv(file.path(ref_dir, "kozak_1987_table1_frequencies.tsv"),
                      show_col_types = FALSE)
win_idx <- c("-6" = 1, "-5" = 2, "-4" = 3, "-3" = 4, "-2" = 5, "-1" = 6, "+4" = 10)
stopifnot("unexpected Kozak context_index set" =
            setequal(kozak_pub$context_index, c(1:6, 10)))

pwm_publ <- kozak_pub %>%
  arrange(match(context_index, win_idx)) %>%
  dplyr::select(A = A_frequency, C = C_frequency, G = G_frequency, T = T_frequency) %>%
  as.matrix()
rownames(pwm_publ) <- names(win_idx)
stopifnot("published Kozak rows do not sum to 1" =
            all(abs(rowSums(pwm_publ) - 1) < 1e-8))

kozak_tbl <- inner_join(utr5, cds, by = "transcript_id_clean") %>%
  mutate(
    utr5_len = nchar(utr5_sequence),
    window   = paste0(str_sub(utr5_sequence, start = pmax(1L, utr5_len - 5L)),
                      str_sub(cds_sequence, 1, 4)),
    # 01h_ scored any >=10 nt window. A window whose CDS does not start with ATG is not a
    # start context at all, so it is NA here rather than silently scored as weak.
    usable   = nchar(window) == 10L & str_sub(window, 7, 9) == "ATG"
  )

logodds <- log2((pwm_publ + 0.001) / 0.25)
score_window <- function(win) {
  mat <- str_split_fixed(win, "", 10)
  rowSums(sapply(seq_along(win_idx), function(k)
    logodds[names(win_idx)[k], mat[, win_idx[k]]]))
}
kozak_tbl$kozak_pwm_score_v2 <- NA_real_
kozak_tbl$kozak_pwm_score_v2[kozak_tbl$usable] <-
  score_window(kozak_tbl$window[kozak_tbl$usable])

cat("kozak_pwm_score_v2: scored", sum(kozak_tbl$usable), "of", nrow(kozak_tbl),
    "; NA (not AUG-anchored or too short):", sum(!kozak_tbl$usable), "\n")

## --- CDS eligibility ----------------------------------------------------------------
# GENCODE carries 3'-incomplete CDS annotations (cds_end_NF): the CDS starts correctly at ATG
# but runs off the end of the annotation, so its length is not a multiple of 3. 01g_ already
# refuses to score these (coRdon requires %%3 == 0), which is why cai/tai/csc/fop are NA for
# them - but gc3 computes on the truncated fragment anyway, giving a median ~8.7 points below
# the rest of the transcriptome. Flag the condition explicitly and apply it uniformly, so a
# codon-derived feature is never computed on a CDS that cannot support codons.
cds <- cds %>%
  mutate(
    sq_upper       = str_to_upper(cds_sequence),
    cds_n          = nchar(sq_upper),
    cds_frame_ok   = cds_n %% 3L == 0L,
    cds_has_start  = str_sub(sq_upper, 1, 3) == "ATG",
    cds_has_stop   = str_sub(sq_upper, (cds_n %/% 3L - 1L) * 3L + 1L,
                             (cds_n %/% 3L) * 3L) %in% c("TAA", "TAG", "TGA"),
    cds_complete   = cds_frame_ok & cds_has_start & cds_has_stop
  )

cat("CDS eligibility:\n")
cat("  frame divisible by 3 :", sum(cds$cds_frame_ok),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_frame_ok)))
cat("  starts with ATG      :", sum(cds$cds_has_start),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_has_start)))
cat("  ends with a stop     :", sum(cds$cds_has_stop),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_has_stop)))
cat("  complete             :", sum(cds$cds_complete),
    sprintf("(%.2f%%)\n", 100 * mean(cds$cds_complete)))

# The right mask is a MISSING TERMINAL STOP, not a bad frame. Checked against the GENCODE v49
# tags, the three irregular classes are:
#
#   65  out of frame, ATG, no stop   -> 62 tagged cds_end_NF   (3'-incomplete)
#   53  in frame,     ATG, no stop   -> 52 tagged cds_end_NF   (3'-incomplete, truncation
#                                       happens to land on a codon boundary)
#   25  in frame,     no ATG, stop   -> 24 tagged non_ATG_start, 17 of them MANE Select
#
# Only the first two are truncated annotations. Masking on frame alone catches 65 and misses
# the 53, which pass every existing filter and receive full cai/tai/csc/fop scores computed on
# an incomplete CDS. The third class is real biology - genuine non-AUG initiation, mostly MANE
# Select - and must be KEPT: it has a complete CDS and only its Kozak score is meaningless.
## --- internal-codon variants: gc3, cai, fop, csc ---------------------------------
# "Internal" = initiator AUG and terminal stop codon both removed. Neither is decoded the way
# a sense codon is: every transcript has exactly one initiator, so it carries no information
# between genes, and the stop is released rather than decoded by a tRNA.
#
# CAI and FOP already excluded stops (via sense_codons / sense_cols in 01g_), so only the
# initiator changes them. CSC does not: iCodon is given the sequence, and it is sensitive to
# what the sequence contains.
suppressMessages({library(coRdon); library(Biostrings); library(iCodon)})

gen_code     <- Biostrings::getGeneticCode("SGC0")
stop_codons  <- names(gen_code[gen_code == "*"])
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

# 01g_ scores only in-frame CDS, so the reproduction below must use the same subset.
codon_tbl <- cds %>% filter(cds_frame_ok)
cat("In-frame CDS for codon scoring:", nrow(codon_tbl), "\n")

counts_of <- function(d) { m <- codonCounts(codonTable(d)); rownames(m) <- names(d); m }
build_w <- function(cm, idx) {
  ref <- colSums(cm[idx, , drop = FALSE])
  w <- setNames(rep(NA_real_, length(gen_code)), names(gen_code)); w[stop_codons] <- 0
  for (aa in names(aa_to_codons)) {
    cod <- aa_to_codons[[aa]]; cnt <- ref[cod]; tot <- sum(cnt)
    if (tot == 0) { w[cod] <- 1; next }
    rscu <- cnt / (tot / length(cod)); mx <- max(rscu)
    w[cod] <- if (mx == 0) 0 else pmin(rscu / mx, 1)
  }
  w
}
cai_of <- function(cm, w) {
  use <- sense_codons[!is.na(w[sense_codons]) & w[sense_codons] > 0]
  apply(cm, 1, function(g) { cu <- g[use]; n <- sum(cu)
    if (n == 0) NA_real_ else exp(sum(cu * log(w[use])) / n) })
}
fop_of <- function(cm, opt) {
  sc <- setdiff(colnames(cm), stop_codons)
  tot <- rowSums(cm[, sc, drop = FALSE]); o <- rowSums(cm[, opt, drop = FALSE])
  r <- o / tot; r[tot == 0] <- NA_real_; r
}

full_ss <- DNAStringSet(codon_tbl$sq_upper); names(full_ss) <- codon_tbl$transcript_id_clean
cm_full <- counts_of(full_ss)

ribo_tx  <- features %>% filter(str_detect(symbol, "^RPL|^RPS")) %>% pull(transcript_id_clean)
ribo_idx <- rownames(cm_full) %in% ribo_tx
w_full   <- build_w(cm_full, ribo_idx)

te_rank <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE) %>%
  filter(condition == "normoxia", treatment == "sictrl") %>%
  group_by(gene_id) %>% summarise(mte = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mte)) %>% arrange(desc(mte))
top10  <- sub("\\..*", "", te_rank$gene_id[seq_len(ceiling(nrow(te_rank) * 0.10))])
ref_tx <- features$transcript_id_clean[sub("\\..*", "", features$gene_id) %in% top10]
ref_tx <- ref_tx[ref_tx %in% rownames(cm_full)]
opt_full <- unname(sapply(aa_to_codons, function(c_)
  c_[which.max(colSums(cm_full[ref_tx, , drop = FALSE])[c_])]))

# VERIFY before trusting: recomputing on the FULL CDS must reproduce 01g_'s stored values
# exactly. If it does not, the reference sets or the weights have been rebuilt wrongly and
# nothing downstream is safe.
verify <- tibble(transcript_id_clean = rownames(cm_full),
                 cai_rep = cai_of(cm_full, w_full),
                 fop_rep = fop_of(cm_full, opt_full)) %>%
  inner_join(features %>% dplyr::select(transcript_id_clean, cai, fop),
             by = "transcript_id_clean")
cai_dev <- max(abs(verify$cai_rep - verify$cai), na.rm = TRUE)
fop_dev <- max(abs(verify$fop_rep - verify$fop), na.rm = TRUE)
cat("Reproduction of 01g_ on the full CDS: CAI max|dev| =", signif(cai_dev, 3),
    ", FOP max|dev| =", signif(fop_dev, 3), ", n =", sum(complete.cases(verify)), "\n")
stopifnot("recomputation does not reproduce 01g_'s CAI" = cai_dev < 1e-9,
          "recomputation does not reproduce 01g_'s FOP" = fop_dev < 1e-9)
cat("Verified: the codon pipeline here matches 01g_ exactly\n")

# Now the internal-codon versions.
trim_internal <- function(x) {
  n <- nchar(x) %/% 3L
  a <- ifelse(substr(x, 1, 3) == "ATG", 4L, 1L)
  b <- ifelse(substr(x, (n - 1) * 3L + 1L, n * 3L) %in% stop_codons, (n - 1) * 3L, n * 3L)
  substr(x, a, b)
}
codon_tbl <- codon_tbl %>% mutate(s_int = trim_internal(sq_upper))
stopifnot("internal sequences lost frame" = all(nchar(codon_tbl$s_int) %% 3 == 0))

int_ss <- DNAStringSet(codon_tbl$s_int); names(int_ss) <- codon_tbl$transcript_id_clean
cm_int  <- counts_of(int_ss)
w_int   <- build_w(cm_int, ribo_idx)
opt_int <- unname(sapply(aa_to_codons, function(c_)
  c_[which.max(colSums(cm_int[ref_tx, , drop = FALSE])[c_])]))
cat("Optimal codons changed by trimming:", sum(opt_int != opt_full), "of 20\n")

predict_human <- predict_stability("human")
uniq <- unique(codon_tbl$s_int)
uscore <- as.numeric(predict_human(uniq)); names(uscore) <- uniq

gc3_from <- function(x) vapply(x, function(z) {
  n <- nchar(z) %/% 3L; if (n < 1L) return(NA_real_)
  k <- seq.int(3L, n * 3L, 3L)
  sum(substr(rep.int(z, n), k, k) %in% c("G", "C")) * 100 / n
}, numeric(1L), USE.NAMES = FALSE)

codon_internal <- tibble(
  transcript_id_clean = codon_tbl$transcript_id_clean,
  gc3_internal = gc3_from(codon_tbl$s_int),
  cai_internal = cai_of(cm_int, w_int),
  fop_internal = fop_of(cm_int, opt_int),
  csc_internal = unname(uscore[codon_tbl$s_int])
)

## --- tai_gtrnadb --------------------------------------------------------------------
# dos Reis-style tAI over the collaborator's GtRNAdb gene-copy table and wobble scheme.
# Anticodon (5'->3') = [wobble base] + complement(codon[2]) + complement(codon[1]).
gcn    <- read_tsv(file.path(ref_dir, "human_trna_gene_copy_hg38.tsv"),
                   comment = "#", show_col_types = FALSE)
wobble <- read_tsv(file.path(ref_dir, "trna_wobble_scheme_v1.tsv"), show_col_types = FALSE)
wob <- wobble %>% transmute(w  = chartr("U", "T", anticodon_wobble_base_rna),
                            n3 = chartr("U", "T", codon_third_base_rna),
                            s  = penalty_s)

comp   <- c(A = "T", C = "G", G = "C", T = "A")
bases  <- c("T", "C", "A", "G")
stops  <- c("TAA", "TAG", "TGA")
sense_codons <- setdiff(as.vector(outer(outer(bases, bases, paste0), bases, paste0)), stops)

W <- map_dbl(sense_codons, function(cod) {
  rules <- wob %>% filter(n3 == substr(cod, 3, 3))
  if (!nrow(rules)) return(0)
  anti <- paste0(rules$w, comp[[substr(cod, 2, 2)]], comp[[substr(cod, 1, 1)]])
  cn <- gcn$gene_copy_number[match(anti, gcn$anticodon)]
  cn[is.na(cn)] <- 0
  sum((1 - rules$s) * cn)
})
names(W) <- sense_codons
tai_w <- W / max(W)
n_zero_w <- sum(tai_w == 0)
if (n_zero_w > 0) tai_w[tai_w == 0] <- exp(mean(log(tai_w[tai_w > 0])))  # dos Reis substitution

cat("tAI weights: 61 sense codons, ", n_zero_w, " with no decoding tRNA; range ",
    paste(round(range(tai_w), 3), collapse = "-"), "\n", sep = "")

cds$tai_gtrnadb <- map_dbl(cds$sq_upper, function(sq) {
  n <- nchar(sq) %/% 3L
  if (n < 3L) return(NA_real_)
  cods <- substring(sq, seq(1, n * 3L, 3L), seq(3L, n * 3L, 3L))
  cods <- cods[-1]                                                   # drop initiator
  if (tail(cods, 1) %in% stops) cods <- head(cods, -1)               # drop terminal stop
  ww <- tai_w[cods]; ww <- ww[!is.na(ww)]
  if (!length(ww)) return(NA_real_)
  exp(mean(log(ww)))                                                 # geometric mean
})

truncated <- !cds$cds_has_stop
cds$tai_gtrnadb[truncated] <- NA_real_
codon_internal <- codon_internal %>%
  mutate(across(c(gc3_internal, cai_internal, fop_internal, csc_internal),
                ~ if_else(transcript_id_clean %in% cds$transcript_id_clean[truncated],
                          NA_real_, .x)))
cat("Masked", sum(truncated), "transcripts with no terminal stop (cds_end_NF)",
    "across every codon-derived feature\n")
cat("  of which out-of-frame:", sum(truncated & !cds$cds_frame_ok),
    " in-frame but truncated:", sum(truncated & cds$cds_frame_ok), "\n")
cat("Kept", sum(!cds$cds_has_start & cds$cds_has_stop),
    "non-AUG-start transcripts: complete CDS, real initiation biology\n")

## --- termination context + Cridge 2018 readthrough --------------------------------
# New family, nothing equivalent in the existing pipeline. Motivated by the csc result: after
# the terminal stop was removed from the iCodon input, csc lost 1.35 pp of importance share
# (20/20 seeds), which means it had been carrying stop-codon identity. That signal belongs
# here, under its own name, rather than smuggled into a codon-optimality score.
utr3_cache <- file.path(out_dir, "precomputed_utr3_seqs.rds")
if (file.exists(utr3_cache)) {
  utr3_tbl <- readRDS(utr3_cache)
} else {
  suppressMessages({library(GenomicFeatures); library(BSgenome.Hsapiens.UCSC.hg38)})
  txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
  u3   <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                threeUTRsByTranscript(txdb, use.names = TRUE))
  utr3_tbl <- tibble(transcript_id_clean = sub("\\..*", "", names(u3)),
                     utr3_seq = as.character(u3))
  saveRDS(utr3_tbl, utr3_cache)
}
utr3_tbl <- utr3_tbl %>% filter(transcript_id_clean %in% features$transcript_id_clean)
cat("3'UTR sequences available:", nrow(utr3_tbl), "of", nrow(features), "\n")

cridge <- read_tsv(file.path(ref_dir, "cridge2018_termination_contexts.tsv"),
                   show_col_types = FALSE)
cridge_med <- median(cridge$readthrough_measure)
stopifnot(
  "Cridge table is not 192 contexts"    = nrow(cridge) == 192L,
  "Cridge median readthrough off"       = between(cridge_med, 0.08, 0.095),
  "Cridge readthrough not a percentage" = all(cridge$readthrough_measure > 0 &
                                              cridge$readthrough_measure < 5)
)
cat("Cridge 2018:", nrow(cridge), "contexts, median", round(cridge_med, 4),
    "% (paper ~0.087%)\n")

# Codons to the next in-frame stop downstream of the annotated one: how far a readthrough
# ribosome would travel before terminating again.
next_inframe_stop <- function(u) {
  n <- nchar(u) %/% 3L
  if (n < 1L) return(NA_integer_)
  cod <- substring(u, seq(1, n * 3L, 3L), seq(3L, n * 3L, 3L))
  w <- which(cod %in% c("TAA", "TAG", "TGA"))
  if (!length(w)) NA_integer_ else w[1]
}

term <- cds %>%
  dplyr::select(transcript_id_clean, sq_upper, cds_has_stop) %>%
  filter(cds_has_stop) %>%
  mutate(ncod = nchar(sq_upper) %/% 3L,
         stop_codon = substr(sq_upper, (ncod - 1L) * 3L + 1L, ncod * 3L)) %>%
  inner_join(utr3_tbl, by = "transcript_id_clean") %>%
  mutate(
    u3   = str_to_upper(utr3_seq),
    ctx6 = paste0(stop_codon, str_sub(u3, 1, 3)),    # +1..+6, the Cridge reporter window
    ctx9 = paste0(stop_codon, str_sub(u3, 1, 6)),    # +1..+9
    term_stop_TAA = as.integer(stop_codon == "TAA"),
    term_stop_TAG = as.integer(stop_codon == "TAG"),
    term_stop_TGA = as.integer(stop_codon == "TGA"),
    p4 = str_sub(u3, 1, 1),
    term_plus4_A = as.integer(p4 == "A"), term_plus4_C = as.integer(p4 == "C"),
    term_plus4_G = as.integer(p4 == "G"), term_plus4_T = as.integer(p4 == "T"),
    term_window_gc = if_else(nchar(ctx9) == 9L,
                             100 * str_count(ctx9, "[GC]") / 9, NA_real_),
    term_interval_codons = map_int(u3, next_inframe_stop)
  ) %>%
  left_join(cridge %>% dplyr::select(ctx6 = termination_context_plus1_to_plus6,
                                     term_readthrough_cridge = readthrough_measure),
            by = "ctx6") %>%
  mutate(term_log1p_readthrough = log1p(term_readthrough_cridge))

cat("Cridge join coverage:", sum(!is.na(term$term_readthrough_cridge)), "/", nrow(term),
    sprintf(" (%.1f%%)\n", 100 * mean(!is.na(term$term_readthrough_cridge))))

term_cols <- c("term_stop_TAA", "term_stop_TAG", "term_stop_TGA",
               "term_plus4_A", "term_plus4_C", "term_plus4_G", "term_plus4_T",
               "term_window_gc", "term_interval_codons",
               "term_readthrough_cridge", "term_log1p_readthrough")
term <- term %>% dplyr::select(transcript_id_clean, all_of(term_cols))
## --- join --------------------------------------------------------------------------
features <- features %>%
  left_join(kozak_tbl %>% dplyr::select(transcript_id_clean, kozak_pwm_score_v2),
            by = "transcript_id_clean") %>%
  left_join(cds %>% dplyr::select(transcript_id_clean, tai_gtrnadb,
                                  cds_frame_ok, cds_has_start, cds_has_stop, cds_complete),
            by = "transcript_id_clean") %>%
  left_join(codon_internal, by = "transcript_id_clean") %>%
  left_join(term, by = "transcript_id_clean")

corrected_cols <- c("kozak_pwm_score_v2", "tai_gtrnadb",
                    "gc3_internal", "cai_internal", "fop_internal", "csc_internal")
stopifnot("termination columns did not attach" = all(term_cols %in% colnames(features)))
eligibility_cols <- c("cds_frame_ok", "cds_has_start", "cds_has_stop", "cds_complete")
stopifnot(
  "corrected columns did not all attach" = all(corrected_cols %in% colnames(features)),
  "eligibility columns did not all attach" = all(eligibility_cols %in% colnames(features)),
  "a codon feature survived on a truncated (cds_end_NF) CDS" =
    all(vapply(c("gc3_internal","cai_internal","fop_internal","csc_internal","tai_gtrnadb"),
               function(f) all(is.na(features[[f]][!features$cds_has_stop])), logical(1))),
  "non-AUG-start transcripts were masked, but they are real" =
    any(!is.na(features$gc3_internal[!features$cds_has_start & features$cds_has_stop]))
)
cat("Matrix now", nrow(features), "x", ncol(features), "\n")
```

### Termination validation

```{r validate_termination, eval = !params$pass_through}
stopifnot(
  "readthrough outside (0, 5]%" =
    all(term$term_readthrough_cridge > 0 & term$term_readthrough_cridge <= 5, na.rm = TRUE),
  "termination window GC outside [0, 100]" =
    all(term$term_window_gc >= 0 & term$term_window_gc <= 100, na.rm = TRUE),
  "readthrough interval not positive" =
    all(term$term_interval_codons > 0, na.rm = TRUE),
  "stop indicators do not sum to 1" =
    all(term$term_stop_TAA + term$term_stop_TAG + term$term_stop_TGA == 1L),
  "plus4 indicators do not sum to 1" =
    all(term$term_plus4_A + term$term_plus4_C + term$term_plus4_G + term$term_plus4_T == 1L)
)

# Biological spot-check: TGA is the leakiest stop codon and TAA the tightest. This is the
# check that would catch a scrambled Cridge join - a shuffled table still passes every range
# assertion above.
by_stop <- term %>%
  mutate(stop = case_when(term_stop_TAA == 1 ~ "TAA", term_stop_TAG == 1 ~ "TAG",
                          TRUE ~ "TGA")) %>%
  group_by(stop) %>%
  summarise(n = n(), median_rt = median(term_readthrough_cridge, na.rm = TRUE),
            .groups = "drop")
print(as.data.frame(by_stop))

rt <- setNames(by_stop$median_rt, by_stop$stop)
stopifnot("TGA is not the leakiest stop codon" = rt["TGA"] > rt["TAG"],
          "TAA is not the tightest stop codon" = rt["TAG"] > rt["TAA"])
cat("Spot-check passed: TGA >", "TAG >", "TAA median readthrough\n")
```


## Validation

Range checks plus biological spot-checks, both required. The within-tier Kozak check is new:
it is the one that would have caught the `01h_` defect, which the existing tier-monotonicity
check cannot.

```{r validate_corrected, eval = !params$pass_through}
# --- computational ---
stopifnot(
  "gc3_internal outside [0, 100]" =
    all(features$gc3_internal >= 0 & features$gc3_internal <= 100, na.rm = TRUE),
  "cai_internal outside [0, 1]" =
    all(features$cai_internal >= 0 & features$cai_internal <= 1 + 1e-9, na.rm = TRUE),
  "fop_internal outside [0, 1]" =
    all(features$fop_internal >= 0 & features$fop_internal <= 1 + 1e-9, na.rm = TRUE),
  "tai_gtrnadb outside (0, 1]" =
    all(features$tai_gtrnadb > 0 & features$tai_gtrnadb <= 1, na.rm = TRUE),
  "kozak_pwm_score_v2 not finite" =
    all(is.finite(features$kozak_pwm_score_v2[!is.na(features$kozak_pwm_score_v2)]))
)

# --- biological spot-checks ---
gc3_med <- median(features$gc3_internal, na.rm = TRUE)
stopifnot("GC3 median outside [40, 65]%" = between(gc3_med, 40, 65))
cat("GC3 (sense codons) median:", round(gc3_med, 1), "%\n")

# Ribosomal proteins are highly translated and should sit above the median on tRNA supply.
rp <- features %>% filter(str_detect(symbol, "^RP[LS][0-9]"))
stopifnot("too few ribosomal proteins to test" = nrow(rp) >= 20,
          "ribosomal proteins are not above median tAI" =
            median(rp$tai_gtrnadb, na.rm = TRUE) > median(features$tai_gtrnadb, na.rm = TRUE))
cat("tAI: ribosomal proteins (n = ", nrow(rp), ") median ",
    round(median(rp$tai_gtrnadb, na.rm = TRUE), 3), " vs all ",
    round(median(features$tai_gtrnadb, na.rm = TRUE), 3), "\n", sep = "")

# THE NEW CHECK. Position -3 is the purine position: A and G are both favourable, so within a
# single ordinal tier they must score comparably. Under the old PWM they differ by ~2.3 bits.
kozak_tier <- kozak_tbl %>%
  filter(usable) %>%
  mutate(tier = case_when(
           str_detect(window, "^(GCCACC|GCCGCC)ATGG$") ~ 3L,
           str_detect(window, "^...[AG]..ATGG$")       ~ 2L,
           str_detect(window, "^...[AG]..ATG[ACT]$")   ~ 1L,
           str_detect(window, "^...[CT]..ATGG$")       ~ 1L,
           TRUE                                        ~ 0L),
         base_m3 = str_sub(window, 4, 4)) %>%
  filter(tier == 2L, base_m3 %in% c("A", "G")) %>%
  group_by(base_m3) %>%
  summarise(n = n(), mean_v2 = mean(kozak_pwm_score_v2), .groups = "drop")
print(kozak_tier)

purine_gap <- abs(diff(kozak_tier$mean_v2))
cat("A-minus-G spread within tier 2:", round(purine_gap, 3), "bits\n")
stopifnot("within-tier purine gap too large: the -3 row is wrong again" = purine_gap < 0.75)
cat("Within-tier purine check passed\n")
```

## How much did each correction move?

```{r correction_deltas, eval = !params$pass_through}
pairs_tbl <- tribble(~current, ~corrected,
  "kozak_pwm_score", "kozak_pwm_score_v2", "gc3", "gc3_internal",
  "tai", "tai_gtrnadb", "cai", "cai_internal",
  "fop", "fop_internal", "csc", "csc_internal")

deltas <- pairs_tbl %>%
  mutate(pearson  = map2_dbl(current, corrected,
           ~ cor(features[[.x]], features[[.y]], use = "complete.obs")),
         spearman = map2_dbl(current, corrected,
           ~ cor(features[[.x]], features[[.y]], use = "complete.obs", method = "spearman")),
         median_current   = map_dbl(current,   ~ median(features[[.x]], na.rm = TRUE)),
         median_corrected = map_dbl(corrected, ~ median(features[[.x]], na.rm = TRUE)),
         n = map2_int(current, corrected,
           ~ sum(complete.cases(features[[.x]], features[[.y]]))))
print(as.data.frame(deltas))

cat("\nCorrelation with gc3_internal:\n")
cat("  tai (ours)     :", round(cor(features$tai,         features$gc3_internal, use = "complete.obs"), 3), "\n")
cat("  tai_gtrnadb    :", round(cor(features$tai_gtrnadb, features$gc3_internal, use = "complete.obs"), 3), "\n")
```

# 2. Annotate every column

`block` decides whether a column can ever be a feature; `family` groups correlated columns so
SHAP can be summed over them exactly.

```{r assign_block}
assign_block <- function(x) {
  case_when(
    # Identifiers. Matches 07's id_cols plus its id_pattern_cols regex.
    str_detect(x, "gene_id|transcript_id|^symbol$")                 ~ "id",
    # Outcome and anything derived from it. Never a feature: these define the target.
    x %in% c("te_lfc", "te_padj", "te_lfc_bin")                     ~ "label",
    str_detect(x, "^sig_|^indiv_te_|^delta_te_")                    ~ "label",
    # eIF4E-derived. Circular for a 4E model, so separable on its own.
    str_detect(x, "4e_resistance|^eif4e_ip|^is_4e_|_4e_|^eif4e_")   ~ "eif4e",
    # Measured in some other experiment. Will not exist for most new datasets.
    str_detect(x, "^clip_|^cnot3_|^zhu2024_|^hia2026_|^karner2026_|^dap5_|^is_dap5_") ~ "external",
    # CDS eligibility flags: QC about whether a codon feature is meaningful, not a feature.
    str_detect(x, "^cds_(frame_ok|has_start|has_stop|complete)$")   ~ "qc",
    # Computable from the transcript sequence alone. These are what transfer.
    TRUE                                                            ~ "intrinsic"
  )
}

assign_family <- function(x) {
  case_when(
    str_detect(x, "gene_id|transcript_id|^symbol$")              ~ "id",
    str_detect(x, "^cds_(frame_ok|has_start|has_stop|complete)$") ~ "cds_eligibility",
    x %in% c("te_lfc", "te_padj", "te_lfc_bin")                  ~ "outcome",
    str_detect(x, "^sig_|^indiv_te_|^delta_te_")                 ~ "outcome_derived",
    str_detect(x, "_length$|^log2_.*_length$")                   ~ "length",
    str_detect(x, "^term_")                                      ~ "termination",
    str_detect(x, "_gc$|^gc3")                                   ~ "gc",
    # Order matters: the CDS and 3'UTR structure columns carry a suffix, the 5'UTR ones
    # do not, so the suffixed patterns must be tested first.
    str_detect(x, "^struct_.*_cds$")                             ~ "structure_cds",
    str_detect(x, "^struct_.*_utr3$")                            ~ "structure_utr3",
    str_detect(x, "^struct_")                                    ~ "structure_utr5",
    str_detect(x, "^csc|^cai|^tai|^fop")                         ~ "codon_optimality",
    str_detect(x, "^kozak_")                                     ~ "initiation",
    str_detect(x, "^clip_")                                      ~ "clip",
    str_detect(x, "^zhu2024_|^hia2026_|^karner2026_|^cnot3_")    ~ "stability_external",
    str_detect(x, "^dap5_|^is_dap5_")                            ~ "dap5",
    str_detect(x, "4e")                                          ~ "eif4e",
    TRUE                                                         ~ "other"
  )
}

# variant pairs a corrected column with the one it supersedes. 02_ picks one member of each
# pair via variant_set; both together in one forest would split credit between near-duplicates.
variant_pairs <- tribble(
  ~current,           ~corrected,
  "kozak_pwm_score",  "kozak_pwm_score_v2",
  "gc3",              "gc3_internal",
  "tai",              "tai_gtrnadb",
  "cai",              "cai_internal",
  "fop",              "fop_internal",
  "csc",              "csc_internal"
)

annotation <- tibble(feature = colnames(features)) %>%
  mutate(
    block  = assign_block(feature),
    family = assign_family(feature),
    variant = case_when(
      feature %in% variant_pairs$current   ~ "current",
      feature %in% variant_pairs$corrected ~ "corrected",
      TRUE                                 ~ "single"
    ),
    source = if_else(feature %in% variant_pairs$corrected,
                     "01_ (corrected, see 00_)", baseline_rds),
    definition = NA_character_
  )

# Deprecated features stay in the matrix but are not selected by default. Positional CSC is
# dropped by decision: csc_q1-q4 and csc_first75 are five near-collinear arms of one quantity,
# which is the worst case for credit-splitting and the thing SHAP family summation exists to
# avoid. They are flagged rather than deleted so the baseline gate stays runnable - the
# baseline's 51 features include all five.
deprecated_features <- c("csc_q1", "csc_q2", "csc_q3", "csc_q4", "csc_first75")

annotation <- annotation %>%
  mutate(deprecated = feature %in% deprecated_features,
         added_in   = if_else(feature %in% baseline_cols, "baseline", "tco"))
cat("Columns added by this notebook:", sum(annotation$added_in == "tco"), "\n")
cat("Deprecated (kept in the matrix, not selected by default):",
    sum(annotation$deprecated), "-", paste(deprecated_features, collapse = ", "), "\n")

# A corrected column is only meaningful if its partner is present, and vice versa.
if (!isTRUE(params$pass_through)) {
  stopifnot("a variant pair is incomplete" =
    all(c(variant_pairs$current, variant_pairs$corrected) %in% annotation$feature))
}

annotation %>% count(block, family, name = "n") %>% arrange(block, desc(n)) %>% print(n = 40)
```

```{r annotate_definitions}
# Record the definitions that differ from the collaborator repo, so the unit and the convention
# travel with the column rather than living only in a notebook comment. Audited in 00_.
definitions <- tribble(
  ~feature,           ~definition,
  "utr5_gc",          "percent (0-100); denominator includes ambiguous bases",
  "utr3_gc",          "percent (0-100); denominator includes ambiguous bases",
  "cds_gc",           "percent (0-100); denominator includes ambiguous bases",
  "transcript_gc",    "percent (0-100); length-weighted mean of the three regional GCs",
  "gc3",              "percent (0-100); INCLUDES the stop codon - to be fixed, see 00_",
  "log2_cds_length",  "log2(x + 1); +1 applied uniformly across all three regions",
  "log2_utr5_length", "log2(x + 1)",
  "log2_utr3_length", "log2(x + 1)",
  "kozak_pwm_score",  "Kozak 1987 log-odds; PWM is WRONG at -6 and -3 - to be fixed, see 00_",
  "tai",              "coRdon tAI; collaborator variant to be adopted, see 00_",
  "csc",              "iCodon::predict_stability('human'), whole stop-free CDS",
  "kozak_pwm_score_v2", "Kozak 1987 Table 1 log-odds (imported file); NA unless AUG-anchored",
  "gc3_internal",     "percent (0-100); internal codons: initiator and stop removed",
  "cai_internal",     "Sharp & Li CAI, RPL/RPS reference; internal codons only",
  "fop_internal",     "FOP, top-10%-TE reference; internal codons only",
  "csc_internal",     "iCodon::predict_stability('human') on the internal-codon CDS",
  "tai_gtrnadb",      "dos Reis-style tAI; GtRNAdb hg38 copy number + collaborator wobble v1"
)

definitions <- definitions %>% filter(feature %in% annotation$feature)
annotation <- annotation %>%
  rows_update(definitions, by = "feature", unmatched = "error")

# Every struct_* column is region-isolated RNAplfold (-W 150 -L 100), NOT the collaborator's
# whole-transcript fold. Renamed to tco_struct_* in a later pass so the two cannot be confused.
annotation <- annotation %>%
  mutate(definition = if_else(
    str_starts(feature, "struct_"),
    "RNAplfold -W 150 -L 100, region folded in ISOLATION (see 00_ section 6)",
    definition))

cat("Columns carrying an explicit definition:", sum(!is.na(annotation$definition)), "\n")
```

# 3. Annotation integrity

The annotation and the matrix must never drift apart. These are the checks that guarantee
`02_` cannot silently select the wrong columns.

```{r validate_annotation}
stopifnot(
  "annotation does not cover the matrix exactly" =
    setequal(annotation$feature, colnames(features)),
  "annotation has duplicate rows" =
    !any(duplicated(annotation$feature)),
  "a column has no block" =
    all(!is.na(annotation$block)),
  "a column has no family" =
    all(!is.na(annotation$family)),
  "unrecognised block" =
    all(annotation$block %in% c("id", "label", "intrinsic", "external", "eif4e",
                                "technical", "qc")),
  "no family may be assigned 'other'" =
    !any(annotation$family == "other"),
  "unrecognised variant" =
    all(annotation$variant %in% c("single", "current", "corrected")),
  "a deprecated feature is missing from the matrix" =
    all(deprecated_features %in% annotation$feature),
  "unrecognised added_in" =
    all(annotation$added_in %in% c("baseline", "tco"))
)
cat("Annotation integrity: OK\n")

# Leakage guard. What matters is not which block an outcome-derived column lands in, but that
# it is not SELECTABLE - i.e. never in intrinsic/external, the only blocks 02_ draws features
# from. (te_4e_resistance_lfc is outcome-shaped but blocked as eif4e, which is also fine.)
selectable_blocks <- c("intrinsic", "external")
leak <- annotation %>%
  filter(block %in% selectable_blocks,
         str_detect(feature, "^te_|^sig_|^indiv_te_|^delta_te_"))
stopifnot("an outcome-derived column is selectable as a feature" = nrow(leak) == 0)
cat("Leakage guard: no outcome-derived column is selectable\n")
```

# 4. Reproduction gate: does the annotation rebuild the baseline 51?

Notebook 07 arrives at its feature set by a regex cascade. This notebook arrives at one by a
block join. They must agree exactly, or `02_` is not modelling the same thing as the baseline.

```{r reproduce_baseline}
baseline_importance <- read_csv(
  here("output", "predictive_modeling",
       paste0("rf_classification_importance_eif33d_promotes_hypoxia_1hr_lfc0.5_",
              "dhx29_kd_feature_clipaggregate_no_boruta_no4e_noOwnTE.csv")),
  show_col_types = FALSE)

# Selection rule, mirroring the baseline run (no4e, noOwnTE, clip aggregate).
aggregate_clip <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                    "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                    "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")

# The gate reproduces the BASELINE, so it takes the current member of every variant pair.
# The gate reproduces the BASELINE: baseline-era columns only, current variants, and the five
# deprecated positional-CSC features, which the baseline included.
selected <- annotation %>%
  filter(block %in% c("intrinsic", "external"), variant != "corrected",
         added_in == "baseline") %>%
  filter(!(family == "clip" & !feature %in% aggregate_clip)) %>%   # aggregate CLIP mode
  filter(!str_detect(feature, "^(utr5|utr3|cds|total)_length$")) %>%  # log2 versions kept
  pull(feature)

# Sparsity filter is data-dependent, so it is applied on values, not on names.
# 07 computes it on the labelled genes; on this matrix the same 7 columns fall out.
non_na_frac <- map_dbl(selected, ~ mean(!is.na(unlist(features[[.x]]))))
sparse <- selected[non_na_frac < 0.5]
selected <- setdiff(selected, sparse)

cat("Selected by block join:", length(selected), "\n")
cat("Baseline importance CSV:", nrow(baseline_importance), "\n")
cat("Dropped as sparse (<50% non-NA):", length(sparse), "-",
    paste(sparse, collapse = ", "), "\n")

stopifnot(
  "block join does not reproduce the baseline feature set" =
    setequal(selected, baseline_importance$feature)
)
cat("\nGATE PASSED: the annotation reproduces the baseline 51 features exactly\n")
```

# 5. Write

```{r write_outputs}
matrix_out <- features
cat(if (isTRUE(params$pass_through))
      "pass_through = TRUE: baseline copied unchanged\n"
    else
      "pass_through = FALSE: corrected variants added alongside the originals\n")

saveRDS(matrix_out, file.path(out_dir, "feature_matrix_tco.rds"))
write_csv(annotation, file.path(out_dir, "feature_annotation.csv"))

stopifnot("written matrix lost rows" = nrow(matrix_out) == nrow(features))

cat("\nWrote:\n")
cat(" ", file.path(out_dir, "feature_matrix_tco.rds"), "-",
    nrow(matrix_out), "x", ncol(matrix_out), "\n")
cat(" ", file.path(out_dir, "feature_annotation.csv"), "-",
    nrow(annotation), "rows\n")

annotation %>% count(block, name = "columns") %>% arrange(desc(columns)) %>% print()
```

# Summary

The annotation reproduces the baseline feature set exactly by block join, so `02_rf_model.Rmd`
can drop notebook 07's `feature_set` branches without changing what is modelled. `block` makes
the intrinsic-vs-external split a data decision you can revise by editing a CSV, and `family`
is what lets `03_shap.Rmd` sum SHAP over correlated columns exactly.

Still pass-through: no definition is corrected and no feature is added yet. Next pass applies
the Kozak PWM fix, the GC3 stop-codon fix and the `tco_struct_*` rename, then merges G4 and the
adopted families.
