---
title: "`r paste0('RF: eIF', params$gene, ' ', params$direction, ' (', params$condition, ', ', params$timepoint, ', LFC ', params$lfc, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  readout: "TE"
  timepoint: "1hr"
  lfc: "0.5"
  include_blocks: "intrinsic,external"
  clip_mode: "aggregate"
  neg_geneset_csv: ""
  matrix_rds: "feature_matrix_tco.rds"
  feature_set_label: "tco"
  variant_set: "corrected"
  initiation_score: "both"
  include_deprecated: FALSE
  exclude_families: ""
  shuffle_families: ""
  neg_seed: 9
  neg_tag: ""
  save_plots: TRUE
  save_model_data: TRUE
  gate_against_baseline: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Fork of `07_rf_classification_eif3d_targets_no_boruta.Rmd` for
`code/translational_control_overview/`. Notebook 07 is frozen and remains the reproducible
baseline; nothing here writes to `output/predictive_modeling/`.

Two deliberate differences from 07, and no others:

1. **One params block.** The YAML above is the only source of truth. Notebook 07 carries a
   second, *disagreeing* set of interactive defaults, so knitting it gives a 56-feature
   `external_stability` model while running it chunk-by-chunk gives a 51-feature
   `dhx29_kd_feature` model. That trap is removed here.
2. **`include_blocks` replaces the `feature_set` branches.** Feature selection is a join
   against `feature_annotation.csv` from `01_`, not a cascade of eight `if (feature_set == ...)`
   clauses and regexes.

Everything about the *model* is deliberately identical to 07: the same negative draw, the same
split, the same ranger call, the same seeds.

## Reproducing the baseline is a constraint, not an aspiration

`gate_against_baseline` asserts that the resulting importance equals the frozen baseline run
exactly. It runs only under `variant_set = "current"`, which is why the superseded columns are
**kept in the matrix rather than deleted**: dropping them would make the baseline permanently
unreproducible. The default is `variant_set = "corrected"`; re-gating needs
`variant_set = "current"` **and** `include_deprecated = TRUE`, since the baseline predates the
positional-CSC deprecation. That requires replicating 07's **RNG call sequence**, not merely its logic:

- Between `set.seed(neg_seed)` and the negative draw, `sample()` must be the only RNG consumer,
  and must be called on the same vector with the same `n`.
- `ranger` sees `train_data[, feature_cols]`, so **column order changes tree construction**.
  `feature_cols` must stay in feature-matrix order.

A tidier rewrite that reorders either would produce a different-but-equally-valid model and
fail the gate for no real reason. Once the gate passes, every later difference is attributable
to the feature change that caused it.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
```

```{r derived_values}
NEG_SEED <- as.integer(params$neg_seed)
stopifnot("neg_tag must be alphanumeric/underscore" =
            params$neg_tag == "" || grepl("^[A-Za-z0-9_]+$", params$neg_tag))

include_blocks <- str_trim(str_split(params$include_blocks, ",")[[1]])
stopifnot("include_blocks must name real blocks" =
            all(include_blocks %in% c("intrinsic", "external", "eif4e", "technical")))

pos_label <- params$direction
neg_label <- paste0("does_not_", sub("s$", "", params$direction))

suffix <- paste0(
  "eif3", params$gene, "_", params$direction, "_",
  if (params$readout != "TE") paste0(tolower(params$readout), "_") else "",
  params$condition, "_", params$timepoint, "_lfc", params$lfc,
  "_", params$feature_set_label,
  if (params$variant_set != "current") paste0("_", params$variant_set) else "",
  if (params$initiation_score != "pwm") paste0("_", params$initiation_score) else "",
  "_clip", params$clip_mode,
  if (params$neg_tag != "") paste0("_", params$neg_tag) else "",
  if (NEG_SEED != 9L) paste0("_negseed", NEG_SEED) else ""
)

out_dir <- here("output", "translational_control_overview")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)

cat("Suffix:", suffix, "\n")
cat("Blocks included:", paste(include_blocks, collapse = ", "), "\n")
```

# Load

```{r load_data}
features   <- readRDS(file.path(out_dir, params$matrix_rds))
annotation <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)

stopifnot("annotation does not match the matrix" =
            setequal(annotation$feature, colnames(features)))
cat("Matrix:", nrow(features), "x", ncol(features), "\n")

geneset_csv <- paste0(params$condition, "_", params$gene, "_", params$direction,
                      "_", params$readout, "_", params$timepoint, "_lfc", params$lfc, ".csv")
positive_genes <- read_csv(here("output", "genesets", geneset_csv), show_col_types = FALSE)
cat("Positive genes:", nrow(positive_genes), "from", geneset_csv, "\n")

if (params$neg_geneset_csv != "") {
  negative_control_genes <- read_csv(here("output", "genesets", params$neg_geneset_csv),
                                     show_col_types = FALSE) %>%
    mutate(gene_id_clean = ensembl_gene)
  neg_file <- file.path("output/genesets", params$neg_geneset_csv)
} else {
  neg_csv  <- if (params$gene == "3e") "negative_control_genes_si3e.csv"
              else "negative_control_genes.csv"
  negative_control_genes <- read_csv(here("output", "predictive_modeling", neg_csv),
                                     show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("\\..*", "", gene_id))
  neg_file <- file.path("output/predictive_modeling", neg_csv)
}
cat("Negative pool:", nrow(negative_control_genes), "from", neg_file, "\n")
```

# Target

```{r create_target}
# RNG contract: set.seed(NEG_SEED) seeds ONLY this draw. The split, the forest and the CV stay
# pinned at 9, so across a seed sweep every difference is negative-class composition. The
# sample() below must remain the only RNG consumer after the seed, in this order.
set.seed(NEG_SEED)

pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
n_before <- nrow(neg_in_features)

# Drop positives from the pool BEFORE sampling. The case_when below resolves the positive
# branch first, so a sampled gene that is also positive would be relabelled positive and the
# negative class would silently end up smaller than number_to_match.
neg_in_features <- neg_in_features %>%
  filter(!gene_id_clean %in% pos_in_features$ensembl_gene)
if (n_before > nrow(neg_in_features))
  cat("Removed", n_before - nrow(neg_in_features), "positives from the negative pool\n")

stopifnot("negative pool still overlaps the positive set" =
  length(intersect(neg_in_features$gene_id_clean, pos_in_features$ensembl_gene)) == 0)

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

if (nrow(neg_in_features) <= nrow(pos_in_features)) {
  cat("NOTE: pool (", nrow(neg_in_features), ") is not larger than the positive set (",
      nrow(pos_in_features), ") - the draw is degenerate and every neg_seed",
      "selects the same genes.\n")
} else {
  cat("Drawing", number_to_match, "of", nrow(neg_in_features),
      "pool genes with neg_seed =", NEG_SEED, "\n")
}

features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

cat("Positive:", sum(model_genes$eif3d_target == pos_label),
    " Negative:", sum(model_genes$eif3d_target == neg_label),
    " Total:", nrow(model_genes), "\n")
```

# Select features

```{r select_features}
# clip_cap_binding_count_v2 is the corrected member of a variant pair, so it must be listed
# here too: this filter drops every clip feature NOT named, which would silently remove the
# correction and leave the clip family with one fewer feature under variant_set = "corrected".
aggregate_clip <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                    "clip_cap_binding_count_v2",
                    "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                    "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")

# One member of each variant pair. The two members are near-duplicates (e.g. kozak_pwm_score
# and kozak_pwm_score_v2); together in one forest they would split credit between themselves
# and hide the very effect the correction is meant to expose.
stopifnot("variant_set must be 'current' or 'corrected'" =
            params$variant_set %in% c("current", "corrected"))
sel <- annotation %>%
  filter(block %in% include_blocks, variant %in% c("single", params$variant_set))

# A gate run reproduces the BASELINE feature set, so it sees baseline-era columns only - the
# same rule 01_'s gate applies via added_in. Without it, every feature the fork adds later
# enters the current-variant run and fails a gate that is about reproduction, not about what
# the fork models today. The termination family did exactly that.
gate_run <- isTRUE(params$gate_against_baseline) &&
            params$variant_set == "current" && isTRUE(params$include_deprecated)
if (gate_run) {
  n_post <- sum(sel$added_in != "baseline")
  sel <- sel %>% filter(added_in == "baseline")
  if (n_post > 0)
    cat("Gate run: excluded", n_post, "features added to the fork after the baseline\n")
}

# kozak_pwm_score_v2 and noderer_tis_efficiency are both start-context strength: one
# estimated from Kozak's 1987 frequency table, the other measured by FACS-seq. 00_'s inventory
# says never to model both, on the assumption they are near-duplicates - but on these
# transcripts they correlate at only rho = 0.15, and swapping one for the other moves test AUC
# by -0.0001 (p = 0.29, 20 paired seeds). They are therefore kept together by default and the
# exclusive modes are retained for that comparison.
stopifnot("initiation_score must be 'both', 'pwm' or 'noderer'" =
            params$initiation_score %in% c("both", "pwm", "noderer"))
drop_init <- switch(params$initiation_score,
  both    = character(0),
  pwm     = "noderer_tis_efficiency",
  noderer = c("kozak_pwm_score", "kozak_pwm_score_v2"))
sel <- sel %>% filter(!feature %in% drop_init)
n_init <- sum(sel$feature %in% c("kozak_pwm_score", "kozak_pwm_score_v2",
                                 "noderer_tis_efficiency"))
stopifnot("an exclusive initiation_score let two start-context scores through" =
            params$initiation_score == "both" || n_init <= 1L)
cat("Start-context scores in the forest:", n_init,
    "(initiation_score =", params$initiation_score, ")\n")

# Deprecated features (positional CSC) are excluded unless explicitly asked for. The baseline
# predates the deprecation, so re-gating needs include_deprecated = TRUE.
# Drop whole families by name, so a new family's contribution can be isolated from other
# changes made in the same pass.
if (nzchar(params$exclude_families)) {
  drop_fams <- str_trim(str_split(params$exclude_families, ",")[[1]])
  stopifnot("exclude_families names an unknown family" = all(drop_fams %in% annotation$family))
  n_drop <- sum(sel$family %in% drop_fams)
  sel <- sel %>% filter(!family %in% drop_fams)
  cat("Excluded", n_drop, "features from families:", paste(drop_fams, collapse = ", "), "\n")
}

if (!isTRUE(params$include_deprecated)) {
  n_dep <- sum(sel$deprecated)
  sel <- sel %>% filter(!deprecated)
  if (n_dep > 0) cat("Excluded", n_dep, "deprecated features (positional CSC)\n")
}

if (params$clip_mode == "aggregate") {
  sel <- sel %>% filter(!(family == "clip" & !feature %in% aggregate_clip))
} else if (params$clip_mode == "individual") {
  sel <- sel %>% filter(!(family == "clip" & feature %in% aggregate_clip))
}

# log2 versions are always present and preferred; keeping both splits importance across two
# near-collinear columns.
sel <- sel %>% filter(!str_detect(feature, "^(utr5|utr3|cds|total)_length$"))

# Preserve feature-matrix column order: ranger's mtry sampling depends on it, so reordering
# here would change the forest even with identical seeds.
feature_cols <- colnames(features)[colnames(features) %in% sel$feature]

# Sparse features inflate impurity importance, because median imputation manufactures a split
# between real and imputed values.
non_na <- map_dbl(feature_cols, ~ mean(!is.na(unlist(model_genes[[.x]]))))
sparse <- feature_cols[non_na < 0.5]
feature_cols <- setdiff(feature_cols, sparse)
if (length(sparse))
  cat("Dropped", length(sparse), "sparse features (<50% non-NA):",
      paste(sparse, collapse = ", "), "\n")

# Constant columns carry no information and cannot be split on, but they still occupy a slot
# in mtry sampling. Three of them reached the model unnoticed (the accessibility_*_min
# columns, identically 0 because 01d_ takes min() after replacing NA with 0), so this is a
# general guard rather than a named exclusion.
# Skipped on a gate run: dropping constants is a forward improvement, and the gate's job is
# to reproduce the historical model, not today's better one.
const <- if (gate_run) character(0) else feature_cols[map_lgl(feature_cols, function(f) {
  v <- unlist(model_genes[[f]]); v <- v[!is.na(v)]
  length(v) == 0 || stats::sd(v) == 0
})]
if (length(const)) {
  feature_cols <- setdiff(feature_cols, const)
  cat("Dropped", length(const), "zero-variance features:", paste(const, collapse = ", "), "\n")
}

list_cols <- feature_cols[map_lgl(feature_cols, ~ is.list(model_genes[[.x]]))]
for (col in list_cols) model_genes[[col]] <- as.numeric(unlist(model_genes[[col]]))

cat("Features:", length(feature_cols), "\n")

stopifnot(
  "an outcome-derived column reached the feature set" =
    !any(str_detect(feature_cols, "^te_lfc$|^te_padj$|^sig_|^indiv_te_|^delta_te_")),
  "feature_cols is not in matrix column order" =
    identical(feature_cols, colnames(features)[colnames(features) %in% feature_cols])
)
```

# Split, impute, train

```{r train_test_split}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

# Shuffled-family null. Permuting a family's values across transcripts destroys its link to
# the target while preserving every marginal distribution, the NA pattern and the correlation
# structure WITHIN the family. Whatever importance the shuffled family still absorbs is what
# a family of that size and shape gets for free - the floor a real family has to clear.
# Impurity importance is non-negative, so an uninformative feature never scores zero.
if (nzchar(params$shuffle_families)) {
  shuf <- str_trim(str_split(params$shuffle_families, ",")[[1]])
  stopifnot("shuffle_families names an unknown family" = all(shuf %in% annotation$family))
  shuf_cols <- annotation$feature[annotation$family %in% shuf]
  shuf_cols <- intersect(feature_cols, shuf_cols)
  set.seed(NEG_SEED)                      # tied to the draw, so the sweep stays paired
  perm <- sample(nrow(model_data))        # one permutation for all of them, preserving
  for (cc in shuf_cols) model_data[[cc]] <- model_data[[cc]][perm]   # within-family structure
  cat("SHUFFLED", length(shuf_cols), "features in families:", paste(shuf, collapse = ", "), "\n")
}

set.seed(9)
train_idx  <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

# Medians from the TRAINING split only, then applied to both.
train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~ median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  df
}
train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

cat("Train:", nrow(train_data), " Test:", nrow(test_data), "\n")
cat("Remaining NAs:", sum(is.na(train_data[, feature_cols])),
    "/", sum(is.na(test_data[, feature_cols])), "\n")
```

```{r train_rf}
# Classes are pre-balanced by the draw, so no case weights.
set.seed(9)
rf_model <- ranger(
  x = train_data[, feature_cols],
  y = train_data$eif3d_target,
  num.trees   = 500,
  importance  = "impurity",
  probability = TRUE,
  seed        = 9
)
print(rf_model)
```

# Evaluate

```{r evaluate}
test_prob <- predict(rf_model, test_data[, feature_cols])$predictions[, pos_label]
roc_obj   <- roc(response = test_data$eif3d_target, predictor = test_prob,
                 levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
test_auc  <- as.numeric(auc(roc_obj))

# Youden threshold from TRAINING OOB predictions, not from the test set.
oob <- rf_model$predictions[, pos_label]
roc_oob <- roc(response = train_data$eif3d_target, predictor = oob,
               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
# coords() can return several tied Youden points; take the first so thresh is scalar.
thresh <- as.numeric(coords(roc_oob, "best", best.method = "youden",
                            ret = "threshold")[[1]])[1]
stopifnot("Youden threshold is not scalar" = length(thresh) == 1L)

cat("Test AUC:", round(test_auc, 4), " | OOB Youden threshold:", round(thresh, 4), "\n")
```

```{r cross_validation}
set.seed(9)
folds <- createFolds(model_data$eif3d_target, k = 5)
cv_auc <- map_dbl(folds, function(idx) {
  tr <- model_data[-idx, ]; te <- model_data[idx, ]
  med <- tr %>% dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~ median(., na.rm = TRUE)))
  tr <- impute_with_medians(tr, med); te <- impute_with_medians(te, med)
  fit <- ranger(x = tr[, feature_cols], y = tr$eif3d_target,
                num.trees = 500, probability = TRUE, seed = 9)
  p <- predict(fit, te[, feature_cols])$predictions[, pos_label]
  as.numeric(auc(roc(te$eif3d_target, p, levels = c(neg_label, pos_label),
                     direction = "<", quiet = TRUE)))
})
cat("5-fold CV AUC: mean", round(mean(cv_auc), 4), " sd", round(sd(cv_auc), 4), "\n")
```

# Importance

```{r importance}
importance_df <- tibble(
  feature    = names(rf_model$variable.importance),
  importance = as.numeric(rf_model$variable.importance)
) %>%
  left_join(annotation %>% dplyr::select(feature, family, block), by = "feature") %>%
  arrange(desc(importance))

print(head(importance_df, 20))
```

```{r fig_roc, fig.width = 10, fig.height = 4.6}
# The AUCs are computed three ways and were never plotted. Held-out and OOB together are the
# useful pair: a large gap between them is overfitting, and the CV spread says whether one
# split flattered the model.
roc_df <- bind_rows(
  tibble(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities,
         curve = sprintf("held-out test (AUC %.3f)", as.numeric(auc(roc_obj)))),
  tibble(fpr = 1 - roc_oob$specificities, tpr = roc_oob$sensitivities,
         curve = sprintf("training OOB (AUC %.3f)", as.numeric(auc(roc_oob))))
)

p_roc <- ggplot(roc_df, aes(fpr, tpr, colour = curve)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_path(linewidth = 0.9) +
  coord_equal() +
  scale_colour_manual(values = c("#E69F00", "#0072B2"), name = NULL) +
  labs(x = "false positive rate", y = "true positive rate", title = "(a) ROC",
       subtitle = sprintf("n = %d test, %d train | %d features",
                          nrow(test_data), nrow(train_data), length(feature_cols))) +
  theme_classic(base_size = 11) +
  theme(legend.position = c(0.98, 0.04), legend.justification = c(1, 0))

# Separation of the two classes, which the AUC summarises in one number.
p_sep <- tibble(prob = test_prob, class = test_data$eif3d_target) %>%
  ggplot(aes(prob, fill = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8, aes(colour = class)) +
  scale_fill_manual(values = c("grey70", "#E69F00"), name = NULL) +
  scale_colour_manual(values = c("grey40", "#E69F00"), name = NULL) +
  geom_vline(xintercept = thresh, linetype = "dashed", colour = "grey30") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("CV AUC %.3f +/- %.3f\nn = %d test genes",
                           mean(cv_auc), sd(cv_auc), nrow(test_data))) +
  labs(x = "predicted probability", y = "density", title = "(b) Class separation",
       subtitle = "dashed line = OOB-chosen threshold") +
  theme_classic(base_size = 11) +
  theme(legend.position = c(0.5, 0.95))

print(cowplot::plot_grid(p_roc, p_sep, nrow = 1, rel_widths = c(1, 1.15)))
ggsave_if(file.path(here("plots"), paste0("02_roc_", suffix, ".pdf")),
          cowplot::plot_grid(p_roc, p_sep, nrow = 1, rel_widths = c(1, 1.15)),
          width = 10, height = 4.6)
```

```{r fig_importance_lollipop, fig.width = 8, fig.height = 12}
# Same lollipop as notebook 07, so the two are visually comparable: grey stems, Okabe-Ito
# dark blue points, flipped, capped at 40 features.
library(cowplot)

imp_plot <- importance_df %>%
  slice_head(n = 40) %>%
  mutate(feature = fct_reorder(feature, importance))

p_imp <- ggplot(imp_plot, aes(x = feature, y = importance)) +
  geom_segment(aes(xend = feature, y = 0, yend = importance),
               colour = "grey60", linewidth = 0.6) +
  geom_point(size = 3, colour = "#0072B2") +
  coord_flip() +
  theme_cowplot() +
  labs(x = NULL, y = "Variable importance (impurity)",
       title = paste0("Feature importance: eIF", params$gene, " ", params$direction),
       subtitle = sprintf("%s | %s | LFC %s | %d features | test AUC %.3f | n = %d features shown",
                          tools::toTitleCase(params$condition), params$timepoint, params$lfc,
                          length(feature_cols), test_auc, nrow(imp_plot)))
print(p_imp)
ggsave_if(file.path(here("plots"), paste0("02_importance_lollipop_", suffix, ".pdf")),
          p_imp, width = 8, height = max(4, nrow(imp_plot) * 0.35))
```

```{r fig_importance_family, fig.width = 7, fig.height = 5}
# The same importance summed by family. Impurity is NOT additive, so this is a descriptive
# grouping, not an exact decomposition - 03_shap.Rmd is where family sums are exact.
fam_imp <- importance_df %>%
  group_by(family) %>%
  summarise(importance = sum(importance), n = n(), .groups = "drop") %>%
  mutate(share = 100 * importance / sum(importance),
         family = fct_reorder(family, share))

p_fam <- ggplot(fam_imp, aes(x = family, y = share)) +
  geom_segment(aes(xend = family, y = 0, yend = share), colour = "grey60", linewidth = 0.6) +
  geom_point(aes(size = n), colour = "#0072B2") +
  coord_flip() +
  scale_size_continuous(range = c(2, 6), name = "features") +
  theme_cowplot() +
  labs(x = NULL, y = "share of total impurity importance (%)",
       title = "Impurity importance by family",
       subtitle = sprintf("descriptive only: impurity does not sum exactly within a family. n = %d features",
                          nrow(importance_df)))
print(p_fam)
ggsave_if(file.path(here("plots"), paste0("02_importance_family_", suffix, ".pdf")),
          p_fam, width = 7, height = 5)
```

# Reproduction gate

```{r gate, eval = params$gate_against_baseline && params$variant_set == "current" && params$include_deprecated}
baseline_path <- here("output", "predictive_modeling",
  paste0("rf_classification_importance_eif33d_promotes_hypoxia_1hr_lfc0.5_",
         "dhx29_kd_feature_clipaggregate_no_boruta_no4e_noOwnTE.csv"))
baseline <- read_csv(baseline_path, show_col_types = FALSE)

# The frozen baseline CSV predates the tco_struct_* rename, so the comparison is made on the
# baseline's spelling. This is a rename and nothing else - same columns, same values, same
# order - and the struct count is asserted on both sides below.
importance_cmp <- importance_df %>%
  mutate(feature = sub("^tco_struct_", "struct_", feature))

cmp <- baseline %>%
  dplyr::select(feature, importance_baseline = importance) %>%
  inner_join(importance_cmp %>% dplyr::select(feature, importance_new = importance),
             by = "feature")

max_dev <- max(abs(cmp$importance_baseline - cmp$importance_new))
cat("Gate: features", nrow(importance_df), "vs baseline", nrow(baseline),
    "| matched", nrow(cmp), "\n")
cat("Max |importance deviation|:", signif(max_dev, 4), "\n")
cat("Spearman rank correlation:",
    round(cor(cmp$importance_baseline, cmp$importance_new, method = "spearman"), 6), "\n")

stopifnot(
  "gate: the rename changed how many struct columns are modelled" =
    sum(str_starts(importance_df$feature, "tco_struct_")) ==
    sum(str_starts(baseline$feature, "struct_")),
  "gate: feature set differs from the baseline" =
    setequal(importance_cmp$feature, baseline$feature),
  "gate: importance values differ from the baseline" = max_dev < 1e-8
)
cat("\nGATE PASSED: this fork reproduces the frozen baseline exactly\n")
```

# Export

```{r export}
saveRDS(rf_model, file.path(out_dir, paste0("rf_classifier_", suffix, ".rds")))
write_csv(importance_df, file.path(out_dir, paste0("rf_importance_", suffix, ".csv")))

write_csv(
  tibble(metric = c("test_auc", "cv_auc_mean", "cv_auc_sd", "n_features",
                    "n_train", "n_test", "n_pos", "n_neg", "oob_threshold"),
         value  = c(test_auc, mean(cv_auc), sd(cv_auc), length(feature_cols),
                    nrow(train_data), nrow(test_data),
                    sum(model_data$eif3d_target == pos_label),
                    sum(model_data$eif3d_target == neg_label), thresh)),
  file.path(out_dir, paste0("rf_performance_", suffix, ".csv")))

write_csv(model_data %>% dplyr::select(gene_id_clean, symbol, eif3d_target),
          file.path(out_dir, paste0("rf_model_gene_sets_", suffix, ".csv")))

# The imputed split itself, so 03_shap.Rmd explains THIS model's data rather than re-deriving
# it from params. 07c_ in the frozen pipeline re-derives, and has to assert its way back to
# agreement; saving removes that whole class of drift. Gated off during sweeps, which would
# otherwise write 60 copies.
if (isTRUE(params$save_model_data)) {
  saveRDS(list(train = train_data, test = test_data, feature_cols = feature_cols,
               pos_label = pos_label, neg_label = neg_label, suffix = suffix),
          file.path(out_dir, paste0("rf_model_data_", suffix, ".rds")))
  cat("Wrote 5 files with suffix:", suffix, "\n")
} else {
  cat("Wrote 4 files with suffix:", suffix, "\n")
}

```
