---
title: "`r paste0('Sweep stability: ', params$preset)`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  preset: "g4"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Notebook 02 keeps every positive and draws a size-matched negative class from the pool, so a
single run reports the AUC and the feature ranking of **one arbitrary draw**. Reading any
difference against zero is meaningless; it has to be read against an arm that differs only in
the thing being tested.

**The sweep is paired by construction, and that is the point.** `neg_seed` seeds *only* the
negative draw; the train/test split, the forest and the CV all stay pinned at 9. So within a
seed every arm sees the **same genes in the same split**. Paired differences isolate the arm
exactly, and are far more sensitive than comparing two distributions of 20 runs.

For a **family** preset the three arms are `none` (family excluded), `real` (as built) and
`shuf` (values permuted across transcripts). `shuf` is the floor, not a formality: impurity
importance is non-negative, so an uninformative family never scores zero. A family that does
not beat its own permuted copy has shown nothing.

Run `02b_neg_subsample_sweep.R <preset>` first. Arms are defined in `02_sweep_presets.R`.

```{r libraries}
library(tidyverse)
library(here)
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
```

```{r load_sweep}
stopifnot("unknown preset" = params$preset %in% names(SWEEP_PRESETS))
arms    <- SWEEP_PRESETS[[params$preset]]
fams    <- PRESET_FAMILIES[[params$preset]]
out_dir <- here("output", "translational_control_overview")

read_run <- function(arm_name, seed, kind) {
  f <- file.path(out_dir, paste0("rf_", kind, "_",
                                 sweep_suffix(arms[[arm_name]], seed), ".csv"))
  if (!file.exists(f)) return(NULL)
  read_csv(f, show_col_types = FALSE) %>% mutate(arm = arm_name, seed = seed)
}

grid <- expand_grid(arm = names(arms), seed = SWEEP_SEEDS)
perf <- pmap_dfr(grid, function(arm, seed) read_run(arm, seed, "performance"))
imp  <- pmap_dfr(grid, function(arm, seed) read_run(arm, seed, "importance"))

stopifnot("no sweep output found - run 02b_neg_subsample_sweep.R <preset> first" =
            nrow(perf) > 0)
cat("Preset:", params$preset, "| arms:", paste(names(arms), collapse = ", "), "\n")
cat("Runs loaded:", n_distinct(paste(perf$arm, perf$seed)), "of", nrow(grid), "\n")
if (n_distinct(paste(perf$arm, perf$seed)) < nrow(grid))
  cat("INCOMPLETE - re-run the preset before reading anything below\n")

auc <- perf %>% filter(metric == "test_auc") %>%
  dplyr::select(arm, seed, test_auc = value)
```

# 1. How much does the draw alone move the AUC?

```{r auc_spread}
auc %>% group_by(arm) %>%
  summarise(n = n(), median = median(test_auc), sd = sd(test_auc),
            min = min(test_auc), max = max(test_auc), .groups = "drop") %>%
  mutate(across(where(is.numeric) & !n, ~ round(., 4))) %>% as.data.frame() %>% print()

cat("\nRange of test AUC attributable to the negative draw alone:",
    round(diff(range(auc$test_auc[auc$arm == names(arms)[1]])), 4), "\n")
```

# 2. Paired differences

```{r paired}
w <- auc %>% pivot_wider(names_from = arm, values_from = test_auc)

pair <- function(a, b) {
  d <- w[[a]] - w[[b]]
  cat(sprintf("%-16s median %+0.4f  mean %+0.4f  sd %0.4f  positive %2d/%d  wilcox p = %.4f\n",
              paste0(a, " - ", b), median(d), mean(d), sd(d), sum(d > 0), length(d),
              suppressWarnings(wilcox.test(w[[a]], w[[b]], paired = TRUE)$p.value)))
}
cmb <- combn(names(arms), 2)
for (i in seq_len(ncol(cmb))) pair(cmb[2, i], cmb[1, i])

cat("\nDraw-to-draw sd:", round(sd(w[[names(arms)[1]]]), 4),
    " vs smallest paired-difference sd:",
    round(min(apply(cmb, 2, function(k) sd(w[[k[2]]] - w[[k[1]]]))), 4), "\n")
```

# 3. Does the family clear its own shuffled floor?

```{r family_share, eval = !all(is.na(fams))}
share <- imp %>% filter(arm %in% c("real", "shuf")) %>%
  group_by(arm, seed) %>% mutate(share = 100 * importance / sum(importance)) %>%
  ungroup() %>% filter(family %in% fams) %>%
  group_by(arm, seed, family) %>% summarise(share = sum(share), .groups = "drop")

share %>% group_by(arm, family) %>%
  summarise(median_share = round(median(share), 3), min = round(min(share), 3),
            max = round(max(share), 3), .groups = "drop") %>% as.data.frame() %>% print()

for (fm in fams) {
  sw <- share %>% filter(family == fm) %>% dplyr::select(arm, seed, share) %>%
    pivot_wider(names_from = arm, values_from = share)
  cat(sprintf("\n%-18s real - shuffled: median %+0.3f pp, positive %2d/%d, wilcox p = %.4f\n",
              fm, median(sw$real - sw$shuf), sum(sw$real > sw$shuf), nrow(sw),
              suppressWarnings(wilcox.test(sw$real, sw$shuf, paired = TRUE)$p.value)))
}
```

# 4. Per-feature rank against its own permuted copy

A feature is only worth ranking if it beats the rank its shuffled twin gets for free.

```{r feature_ranks, eval = !all(is.na(fams))}
ranks <- imp %>% filter(arm %in% c("real", "shuf")) %>%
  group_by(arm, seed) %>% mutate(rk = rank(-importance), nfeat = n()) %>% ungroup() %>%
  filter(family %in% fams) %>%
  group_by(arm, family, feature) %>%
  summarise(median_rank = median(rk), nfeat = first(nfeat), .groups = "drop")

ranks %>%
  pivot_wider(names_from = arm, values_from = median_rank,
              names_prefix = "rank_") %>%
  mutate(gain = rank_shuf - rank_real) %>%     # positive = better than its own null
  arrange(desc(gain)) %>% as.data.frame() %>% print()

cat("\ngain = shuffled rank - real rank. Positive means the real values earn the place;\n")
cat("zero or negative means the column is indistinguishable from noise.\n")
```

# 5. Top-20 stability across draws

```{r top20}
top20 <- imp %>% filter(arm == if ("real" %in% names(arms)) "real" else names(arms)[1]) %>%
  group_by(seed) %>% slice_max(importance, n = 20) %>%
  summarise(fs = list(feature), .groups = "drop")

jac <- combn(nrow(top20), 2, function(k) {
  a <- top20$fs[[k[1]]]; b <- top20$fs[[k[2]]]
  length(intersect(a, b)) / length(union(a, b))
})
cat("Top-20 Jaccard across seed pairs: median", round(median(jac), 3),
    " range", paste(round(range(jac), 3), collapse = " - "), "\n")

always <- Reduce(intersect, top20$fs)
cat("\nTop-20 in EVERY draw (", length(always), "):\n", paste(always, collapse = ", "), "\n")
cat("\nRank a feature off a single run only if it clears that bar.\n")
```
