---
title: "Negative-subsample stability: is the codon correction real?"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  seeds: 20
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Notebook 02 keeps every positive and draws a size-matched negative class from the pool, so a
single run reports the AUC and the feature ranking of **one arbitrary draw**. Reading a
current-vs-corrected difference against zero is therefore meaningless; it has to be read
against the spread produced by the draw itself.

`02b_neg_subsample_sweep.R` renders 02 once per `neg_seed` for both variant sets.

**The sweep is paired by construction, and that is the point.** `neg_seed` seeds *only* the
negative draw; the train/test split, the forest and the CV all stay pinned at 9. So for a given
seed the two variant sets see the **same genes in the same split** and differ only in the six
corrected feature columns. Paired differences therefore isolate the feature change exactly,
and are far more sensitive than comparing two distributions of 20 runs.

```{r libraries}
library(tidyverse)
library(here)
```

```{r load_sweep}
out_dir <- here("output", "translational_control_overview")

# Seed 9 is the default draw and writes no _negseed suffix (02 appends it only when != 9).
suffix_for <- function(variant, seed) {
  paste0("eif33d_promotes_hypoxia_1hr_lfc0.5_tco",
         if (variant != "current") paste0("_", variant) else "",
         "_clipaggregate",
         if (seed != 9L) paste0("_negseed", seed) else "")
}

read_run <- function(variant, seed, kind) {
  f <- file.path(out_dir, paste0("rf_", kind, "_", suffix_for(variant, seed), ".csv"))
  if (!file.exists(f)) return(NULL)
  read_csv(f, show_col_types = FALSE) %>% mutate(variant = variant, seed = seed)
}

grid <- expand_grid(variant = c("current", "corrected"), seed = seq_len(params$seeds))

perf <- pmap_dfr(grid, function(variant, seed) read_run(variant, seed, "performance"))
imp  <- pmap_dfr(grid, function(variant, seed) read_run(variant, seed, "importance"))

stopifnot("no sweep output found - run 02b_neg_subsample_sweep.R first" = nrow(perf) > 0)
cat("Runs loaded:", n_distinct(paste(perf$variant, perf$seed)), "of", nrow(grid), "\n")

auc <- perf %>% filter(metric == "test_auc") %>%
  dplyr::select(variant, seed, test_auc = value) %>%
  left_join(perf %>% filter(metric == "cv_auc_mean") %>%
              dplyr::select(variant, seed, cv_auc = value), by = c("variant", "seed"))
```

# 1. How much does the draw alone move the AUC?

```{r auc_spread}
auc %>%
  group_by(variant) %>%
  summarise(n = n(), median = median(test_auc), sd = sd(test_auc),
            min = min(test_auc), max = max(test_auc), .groups = "drop") %>%
  mutate(across(where(is.numeric) & !n, ~ round(., 4))) %>%
  as.data.frame() %>% print()

cat("\nRange of test AUC attributable to the negative draw alone:",
    round(diff(range(auc$test_auc[auc$variant == "current"])), 4), "\n")
```

# 2. Paired current-vs-corrected difference

```{r paired_auc}
paired <- auc %>%
  dplyr::select(variant, seed, test_auc) %>%
  pivot_wider(names_from = variant, values_from = test_auc) %>%
  mutate(delta = corrected - current)

cat("Paired delta (corrected - current), n =", nrow(paired), "seeds\n")
cat("  median :", round(median(paired$delta), 5), "\n")
cat("  mean   :", round(mean(paired$delta), 5), "\n")
cat("  sd     :", round(sd(paired$delta), 5), "\n")
cat("  range  :", paste(round(range(paired$delta), 5), collapse = " to "), "\n")
cat("  seeds where corrected wins:", sum(paired$delta > 0), "of", nrow(paired), "\n")

wt <- wilcox.test(paired$corrected, paired$current, paired = TRUE)
cat("\nPaired Wilcoxon p =", signif(wt$p.value, 3), "\n")
cat("Draw-to-draw sd within a variant:", round(sd(auc$test_auc[auc$variant=="current"]), 5),
    "vs paired-difference sd:", round(sd(paired$delta), 5), "\n")
```

```{r auc_plot, fig.width = 7, fig.height = 5}
p_auc <- auc %>%
  mutate(variant = factor(variant, levels = c("current", "corrected"))) %>%
  ggplot(aes(x = variant, y = test_auc)) +
  geom_line(aes(group = seed), colour = "grey70", linewidth = 0.4) +
  geom_point(aes(colour = variant), size = 2.5, alpha = 0.85) +
  scale_colour_manual(values = c(current = "grey40", corrected = "#E69F00"), guide = "none") +
  theme_classic(base_size = 12) +
  labs(title = "Test AUC per negative draw, paired by seed",
       subtitle = paste0("n = ", nrow(paired), " seeds; each grey line is one negative draw"),
       x = NULL, y = "Test AUC")
print(p_auc)
ggsave(here("plots", "02c_sweep_auc_paired.pdf"), p_auc, width = 7, height = 5)
```

# 3. Does the `csc` shift survive the draw?

The single run showed `csc` falling from rank 3 to rank 7 (-1.41 percentage points of
importance share). That has to clear the rank movement the draw alone produces.

```{r csc_shift}
canon <- function(x) recode(x,
  kozak_pwm_score_v2 = "kozak_pwm_score", gc3_internal = "gc3", tai_gtrnadb = "tai",
  cai_internal = "cai", fop_internal = "fop", csc_internal = "csc", .default = x)

ranked <- imp %>%
  group_by(variant, seed) %>%
  mutate(pct = 100 * importance / sum(importance),
         rank = rank(-pct), feature = canon(feature)) %>%
  ungroup()

focus <- c("csc", "tai", "gc3", "cai", "fop", "kozak_pwm_score")

ranked %>%
  filter(feature %in% focus) %>%
  group_by(feature, variant) %>%
  summarise(median_rank = median(rank), median_pct = median(pct),
            rank_range = paste(range(rank), collapse = "-"), .groups = "drop") %>%
  pivot_wider(names_from = variant, values_from = c(median_rank, median_pct, rank_range)) %>%
  mutate(across(starts_with("median_pct"), ~ round(., 2))) %>%
  as.data.frame() %>% print()

cat("\nPaired per-seed change in importance share:\n")
ranked %>%
  filter(feature %in% focus) %>%
  dplyr::select(feature, variant, seed, pct) %>%
  pivot_wider(names_from = variant, values_from = pct) %>%
  group_by(feature) %>%
  summarise(median_delta = round(median(corrected - current), 3),
            sd_delta     = round(sd(corrected - current), 3),
            n_negative   = sum(corrected < current), n = n(), .groups = "drop") %>%
  arrange(median_delta) %>% as.data.frame() %>% print()
```

# 4. Feature-ranking stability

```{r top20_stability}
top20 <- ranked %>% filter(rank <= 20) %>%
  group_by(variant, seed) %>% summarise(set = list(feature), .groups = "drop")

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

pairwise <- top20 %>%
  group_by(variant) %>%
  group_modify(~ {
    s <- .x$set; idx <- combn(length(s), 2)
    tibble(j = map_dbl(seq_len(ncol(idx)), ~ jaccard(s[[idx[1, .x]]], s[[idx[2, .x]]])))
  }) %>% ungroup()

pairwise %>% group_by(variant) %>%
  summarise(median_jaccard = round(median(j), 3), min = round(min(j), 3), .groups = "drop") %>%
  as.data.frame() %>% print()

always <- ranked %>% filter(rank <= 20) %>%
  count(variant, feature, name = "n_runs") %>%
  filter(n_runs == params$seeds)

cat("\nFeatures top-20 in EVERY replicate:\n")
always %>% count(variant, name = "n_features") %>% as.data.frame() %>% print()
cat("\ncorrected:\n")
cat(paste(" ", sort(always$feature[always$variant == "corrected"])), sep = "\n")
```

# Interpretation

Rank a feature off a single 02 run only if it clears the bar set here. The paired design is
what makes a small effect readable: comparing two 20-run distributions would be swamped by
draw-to-draw variance, but within a seed both models see identical genes, so the difference is
the feature change alone.
