---
title: "SHAP decomposition of the TCO model"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_set_label: "tco"
  variant_set: "corrected"
  initiation_score: "both"
  clip_mode: "aggregate"
  neg_seed: 9
  explain_n: 200
  explain_sizes: "50,100,200,400"
  run_interactions: FALSE
  cluster_k_max: 6
  cluster_n_null: 50
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Why SHAP here, and what it is not

Impurity importance cannot resolve **credit-splitting across correlated columns**. This matrix
is full of them: `cds_gc` / `gc3_internal` / `csc_internal`, the three G4 regions, the length
family. SHAP is **additive**, so summing within a family is exact rather than approximate, and
family-level attribution is the thing SHAP buys that `02_`'s Gini ranking structurally cannot.

Three things this notebook does **not** claim:

1. **SHAP is not tuned.** TreeSHAP is exact given a tree ensemble; there is no bandwidth or
   sample-size knob. What varies is *which model is explained*, *which genes are explained*,
   and *which summary is read* - so those are what get checked below.
2. **`mean_signed_shap` is not direction.** It averages to ~0 for every feature; its residual
   offset only reflects the gap between explain-set mean prediction and the training baseline.
   The directional quantity is `shap_value_cor` - the correlation of a feature's value with
   its own SHAP value.
3. **One run is one negative draw.** These rankings inherit the draw-to-draw instability that
   `02c_` quantifies for impurity. `neg_seed` is exposed so the same paired sweep can be run
   over SHAP.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(treeshap)
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
set.seed(9)
```

```{r load}
out_dir <- here("output", "translational_control_overview")

# Suffix is built by the SAME helper 02b_/02c_ use, so this cannot drift from 02_'s naming.
suffix <- sweep_suffix(list(feature_set_label = params$feature_set_label,
                            variant_set       = params$variant_set,
                            initiation_score  = params$initiation_score,
                            clip_mode         = params$clip_mode),
                       params$neg_seed)

model_path <- file.path(out_dir, paste0("rf_classifier_", suffix, ".rds"))
data_path  <- file.path(out_dir, paste0("rf_model_data_", suffix, ".rds"))
stopifnot(
  "notebook 02 model not found - knit 02_rf_model.Rmd with these params first" =
    file.exists(model_path),
  "notebook 02 model data not found - re-knit 02_ with save_model_data = TRUE" =
    file.exists(data_path)
)

rf_prob <- readRDS(model_path)
md      <- readRDS(data_path)
stopifnot("saved data does not belong to this model" = identical(md$suffix, suffix))

feature_cols <- md$feature_cols
train_data   <- md$train
test_data    <- md$test
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)

cat("Model:", suffix, "\n")
cat("  treetype:", rf_prob$treetype, "| num.trees:", rf_prob$num.trees,
    "| mtry:", rf_prob$mtry, "| min.node.size:", rf_prob$min.node.size, "\n")
cat("  train:", nrow(train_data), " test:", nrow(test_data),
    " features:", length(feature_cols), "\n")
```

# 1. The surrogate, and Check A

`treeshap` does not support ranger **probability** forests, so what gets explained is a
surrogate *regression* forest on the same 0/1 target and the same rows. Its hyperparameters are
**read off the saved model, never assumed** - ranger's defaults differ by tree type
(`min.node.size` is 10 for probability but 1 for classification and 5 for regression; `mtry` is
`floor(sqrt(p))` for probability but `floor(p/3)` for regression). Assuming them would explain
a different model and every number below would be confidently wrong.

```{r surrogate}
y_train <- as.integer(train_data$eif3d_target == md$pos_label)

fit_surrogate <- function(mtry, min.node.size) {
  set.seed(9)
  ranger(x = train_data[, feature_cols], y = y_train,
         num.trees = rf_prob$num.trees, mtry = mtry, min.node.size = min.node.size,
         importance = "none")
}
rf_shap <- fit_surrogate(rf_prob$mtry, rf_prob$min.node.size)

stopifnot(
  "surrogate is not a regression forest"          = rf_shap$treetype == "Regression",
  "surrogate mtry does not match 02"              = rf_shap$mtry == rf_prob$mtry,
  "surrogate min.node.size does not match 02"     = rf_shap$min.node.size == rf_prob$min.node.size,
  "surrogate num.trees does not match 02"         = rf_shap$num.trees == rf_prob$num.trees
)
cat("Check A passed: surrogate config matches the 02 model structurally\n")
```

# 2. Check B: fidelity, calibrated here rather than inherited

`07c_` in the frozen pipeline uses `fidelity_tol = 0.02` on `max |deviation|`, calibrated on a
**56-feature** set. CLAUDE.md is explicit that this is "a sanity bound, not a tuned one, on
other feature sets". Measuring it here rather than inheriting it shows the design does **not**
transfer, in two ways worth recording.

```{r fidelity}
pred_prob <- predict(rf_prob, data = test_data[, feature_cols])$predictions
pred_prob <- if (is.matrix(pred_prob)) pred_prob[, md$pos_label] else pred_prob

deviation_of <- function(model) {
  pr <- predict(model, data = test_data[, feature_cols])$predictions
  c(max_abs = max(abs(pr - pred_prob)), rmse = sqrt(mean((pr - pred_prob)^2)))
}

p <- length(feature_cols)
configs <- tribble(
  ~label,                ~kind,            ~mtry,        ~min.node.size,
  "matched (correct)",   "correct",        rf_prob$mtry, rf_prob$min.node.size,
  "min.node.size = 1",   "min.node.size",  rf_prob$mtry, 1,
  "min.node.size = 5",   "min.node.size",  rf_prob$mtry, 5,
  "min.node.size = 25",  "min.node.size",  rf_prob$mtry, 25,
  "mtry = floor(p/3)",   "mtry",           floor(p / 3), rf_prob$min.node.size,
  "mtry = 2",            "mtry",           2,            rf_prob$min.node.size
)
devs <- map2_dfr(configs$mtry, configs$min.node.size,
                 ~ as_tibble_row(deviation_of(fit_surrogate(.x, .y))))
configs <- bind_cols(configs, devs)
print(as.data.frame(configs %>% mutate(across(c(max_abs, rmse), ~ round(., 4)))))
```

**Finding 1: `max_abs` is the wrong statistic.** It is one worst-case point and is dominated by
noise, so it ranks deliberately misconfigured surrogates as *better* than the correct one
(`min.node.size = 25` scores 0.069 against the correct 0.111). A tolerance on `max_abs` would
pass a wrong surrogate and fail a right one. RMSE is used below.

**Finding 2: `min.node.size` is not detectable by fidelity on this matrix.** On the 56-feature
set, `min.node.size = 1` deviated 0.0798 against 0.0068 for the correct config - a 12x
separation. Here every value from 1 to 25 lands within a few percent of the correct RMSE. Only
`mtry` moves fidelity materially. So Check B is asserted **only against the misconfigurations
it can actually discriminate**, and `min.node.size` is guarded by Check A's structural equality
instead. Claiming otherwise would be the same unexamined transfer this fork exists to catch.

```{r fidelity_assert}
rmse_of <- function(lbl) configs$rmse[configs$label == lbl]
dev_correct <- rmse_of("matched (correct)")
dev_mtry    <- min(configs$rmse[configs$kind == "mtry"])
dev_mns     <- min(configs$rmse[configs$kind == "min.node.size"])

cat(sprintf("correct RMSE            %.4f\n", dev_correct))
cat(sprintf("nearest mtry error      %.4f  (%.2fx correct)  <- what Check B can detect\n",
            dev_mtry, dev_mtry / dev_correct))
cat(sprintf("nearest min.node.size   %.4f  (%.2fx correct)  <- NOT detectable here\n",
            dev_mns, dev_mns / dev_correct))

stopifnot(
  "surrogate does not track 02's held-out predictions" =
    cor(predict(rf_shap, data = test_data[, feature_cols])$predictions, pred_prob) > 0.95,
  "an mtry misconfiguration is indistinguishable from the correct surrogate" =
    dev_mtry > 1.5 * dev_correct
)
cat("\nCheck B passed: the correct surrogate tracks 02 closely and is clearly separated\n")
cat("from mtry misconfigurations. min.node.size is covered by Check A, not by fidelity.\n")
```

# 3. SHAP

```{r shap}
explain_n  <- min(params$explain_n, nrow(test_data))
set.seed(9)
explain_idx <- sample(nrow(test_data), explain_n)
explain_set <- as.data.frame(test_data[explain_idx, feature_cols])

unified   <- ranger.unify(rf_shap, as.data.frame(train_data[, feature_cols]))
shap_out  <- treeshap(unified, explain_set, interactions = params$run_interactions,
                      verbose = FALSE)
shap_vals <- as.data.frame(shap_out$shaps)
stopifnot("SHAP matrix does not match the explain set" =
            nrow(shap_vals) == explain_n && ncol(shap_vals) == length(feature_cols))
cat("SHAP computed on", explain_n, "held-out genes x", length(feature_cols), "features\n")
```

```{r shap_table}
# shap_value_cor is the DIRECTIONAL summary: a feature's value against its own SHAP value.
shap_value_cor <- map_dbl(feature_cols, function(f) {
  v <- explain_set[[f]]; s <- shap_vals[[f]]
  if (sd(v, na.rm = TRUE) == 0 || sd(s, na.rm = TRUE) == 0) return(NA_real_)
  cor(v, s, use = "complete.obs", method = "spearman")
})

shap_tbl <- tibble(
  feature          = feature_cols,
  mean_abs_shap    = colMeans(abs(shap_vals)),
  mean_signed_shap = colMeans(shap_vals),      # NOT direction - see the header
  shap_value_cor   = shap_value_cor
) %>%
  left_join(ann %>% dplyr::select(feature, family, block), by = "feature") %>%
  arrange(desc(mean_abs_shap)) %>%
  mutate(rank_shap = row_number())

print(as.data.frame(shap_tbl %>% slice_head(n = 20) %>%
        mutate(across(where(is.numeric), ~ round(., 4)))))
```

# 4. Family attribution - the reason to use SHAP at all

SHAP is additive, so a family's contribution is the **exact** sum of its members'. Impurity
importance cannot do this: correlated columns split credit unpredictably between themselves.

```{r family}
fam_tbl <- shap_tbl %>%
  group_by(family) %>%
  summarise(n = n(), family_shap = sum(mean_abs_shap),
            top_feature = feature[which.max(mean_abs_shap)], .groups = "drop") %>%
  mutate(share_pct = round(100 * family_shap / sum(family_shap), 2)) %>%
  arrange(desc(family_shap))
print(as.data.frame(fam_tbl %>% mutate(family_shap = round(family_shap, 4))))
```

```{r vs_impurity}
imp <- read_csv(file.path(out_dir, paste0("rf_importance_", suffix, ".csv")),
                show_col_types = FALSE)
cmp <- shap_tbl %>%
  dplyr::select(feature, family, mean_abs_shap) %>%
  inner_join(imp %>% dplyr::select(feature, impurity = importance), by = "feature")
cat("SHAP vs impurity, feature level: Spearman",
    round(cor(cmp$mean_abs_shap, cmp$impurity, method = "spearman"), 3),
    ", n =", nrow(cmp), "\n")

fam_cmp <- cmp %>% group_by(family) %>%
  summarise(shap = sum(mean_abs_shap), impurity = sum(impurity), .groups = "drop") %>%
  mutate(shap_share = 100 * shap / sum(shap), imp_share = 100 * impurity / sum(impurity),
         delta = shap_share - imp_share) %>% arrange(desc(abs(delta)))
cat("\nFamily share, SHAP vs impurity (delta = where the two disagree):\n")
print(as.data.frame(fam_cmp %>% mutate(across(where(is.numeric), ~ round(., 2))) %>%
                      dplyr::select(family, shap_share, imp_share, delta)))
```

# 5. Is the explain set big enough?

`explain_n` is a free choice, and nothing elsewhere justifies it. If the family ranking is not
stable across sizes, the number is doing work it should not be.

```{r explain_stability}
sizes <- as.integer(str_trim(str_split(params$explain_sizes, ",")[[1]]))
sizes <- sizes[sizes <= nrow(test_data)]

fam_at <- function(n) {
  set.seed(9)
  idx <- sample(nrow(test_data), n)
  sv  <- as.data.frame(treeshap(unified, as.data.frame(test_data[idx, feature_cols]),
                                interactions = FALSE, verbose = FALSE)$shaps)
  tibble(feature = feature_cols, mas = colMeans(abs(sv))) %>%
    left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
    group_by(family) %>% summarise(share = sum(mas), .groups = "drop") %>%
    mutate(share = 100 * share / sum(share), n = n)
}
stab <- map_dfr(sizes, fam_at)

stab %>% pivot_wider(names_from = n, values_from = share, names_prefix = "n") %>%
  arrange(desc(.data[[paste0("n", max(sizes))]])) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>% as.data.frame() %>% print()

rank_cor <- stab %>% group_by(n) %>% mutate(rk = rank(-share)) %>%
  dplyr::select(family, n, rk) %>% pivot_wider(names_from = n, values_from = rk)
cat("\nSpearman of family ranking, smallest vs largest explain set:",
    round(cor(rank_cor[[2]], rank_cor[[ncol(rank_cor)]], method = "spearman"), 3), "\n")
```

# 6. Are there subgroups of targets with different drivers?

Everything above is a **global** ranking: one feature order averaged over all genes. It cannot
say whether there are distinct classes of eIF3d target driven by different features - a
GC/structure-driven set versus a codon-driven set, say. Given that several feature families
turned out real-but-redundant globally, a family that matters for only a subset of genes is
exactly the pattern a global mean hides.

This is the one idea worth borrowing from the `explainer` package (`SHAPclust()`), done here on
our **exact** SHAP values rather than through an mlr3 wrapper with a sampling estimator.

Clustering is on the **positive class only** - including negatives would mostly recover the
class split, which we already know - and on **family-level** profiles, which is exact because
SHAP is additive. Each gene's profile is normalised by its own total absolute contribution, so
clusters reflect *which* families drove a gene, not how confidently it was called.

```{r subgroup_shap}
pos_test <- which(test_data$eif3d_target == md$pos_label)
cat("Positive-class genes in the held-out set:", length(pos_test), "\n")

pos_x    <- as.data.frame(test_data[pos_test, feature_cols])
shap_pos <- as.data.frame(treeshap(unified, pos_x, interactions = FALSE,
                                   verbose = FALSE)$shaps)

fam_of <- ann$family[match(feature_cols, ann$feature)]
fam_mat <- t(rowsum(t(as.matrix(shap_pos)), group = fam_of))   # exact family sums per gene

# Normalise within gene: the profile is the share of that gene's attribution, signed.
denom   <- rowSums(abs(fam_mat)); denom[denom == 0] <- NA
profile <- fam_mat / denom
profile <- profile[stats::complete.cases(profile), , drop = FALSE]
cat("Profiles:", nrow(profile), "genes x", ncol(profile), "families\n")
```

```{r subgroup_k}
library(cluster)

sil_for <- function(m, k) {
  set.seed(9)
  km <- kmeans(m, centers = k, nstart = 25)
  if (length(unique(km$cluster)) < 2) return(NA_real_)
  mean(silhouette(km$cluster, dist(m))[, 3])
}
ks  <- 2:params$cluster_k_max
sil <- map_dbl(ks, ~ sil_for(profile, .x))
print(data.frame(k = ks, mean_silhouette = round(sil, 4)))

# A silhouette is always positive for some k, so it means nothing on its own. The null keeps
# each family's marginal distribution and destroys only the CO-OCCURRENCE between families -
# the same logic as the shuffled-family arm in 02c_: what does structure of this shape score
# when there is no joint structure to find?
null_sil <- map_dbl(seq_len(params$cluster_n_null), function(i) {
  set.seed(1000 + i)
  perm <- apply(profile, 2, sample)
  max(map_dbl(ks, ~ sil_for(perm, .x)), na.rm = TRUE)
})
best_k   <- ks[which.max(sil)]
best_sil <- max(sil, na.rm = TRUE)
p_emp    <- (sum(null_sil >= best_sil) + 1) / (length(null_sil) + 1)

cat(sprintf("\nbest k = %d, silhouette %.4f\n", best_k, best_sil))
cat(sprintf("null (%d permutations): median %.4f, max %.4f\n",
            length(null_sil), median(null_sil), max(null_sil)))
cat(sprintf("empirical p = %.4f  (floor is 1/%d)\n", p_emp, length(null_sil) + 1))
if (p_emp > 0.05)
  cat("\nNOT separable from the null: report this as no detectable subgroup structure,\n",
      "not as k clusters. The cluster means below are then descriptive only.\n")
```

```{r subgroup_characterise}
set.seed(9)
km  <- kmeans(profile, centers = best_k, nstart = 25)
gid <- test_data$gene_id_clean[pos_test][stats::complete.cases(fam_mat / denom)]
sym <- test_data$symbol[pos_test][stats::complete.cases(fam_mat / denom)]

cluster_means <- as_tibble(profile) %>%
  mutate(cluster = km$cluster) %>%
  group_by(cluster) %>%
  summarise(n = n(), across(everything(), mean), .groups = "drop")
print(as.data.frame(cluster_means %>% mutate(across(where(is.numeric), ~ round(., 3)))))

cat("\nTop drivers per cluster (largest mean signed share):\n")
for (k in sort(unique(km$cluster))) {
  v <- colMeans(profile[km$cluster == k, , drop = FALSE])
  top <- sort(v, decreasing = TRUE)[1:3]
  cat(sprintf("  cluster %d (n = %3d): %s\n", k, sum(km$cluster == k),
              paste(sprintf("%s %+.3f", names(top), top), collapse = " | ")))
  cat("      e.g. ", paste(head(sym[km$cluster == k], 6), collapse = ", "), "\n")
}

subgroups <- tibble(gene_id_clean = gid, symbol = sym, cluster = km$cluster) %>%
  bind_cols(as_tibble(profile))
```

```{r export}
write_csv(shap_tbl, file.path(out_dir, paste0("shap_feature_", suffix, ".csv")))
write_csv(fam_tbl,  file.path(out_dir, paste0("shap_family_", suffix, ".csv")))
write_csv(stab,     file.path(out_dir, paste0("shap_explain_stability_", suffix, ".csv")))
write_csv(subgroups, file.path(out_dir, paste0("shap_subgroups_", suffix, ".csv")))
cat("Wrote 4 SHAP files with suffix:", suffix, "\n")
```
