---
title: "SHAP decomposition of the TCO model"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  # The experiment. These must match the 02_ run being explained - 03_ loads that run's saved
  # model and asserts against it, so a mismatch fails loudly rather than explaining the wrong
  # model. One (02_, 03_) pair per experiment is how the cross-experiment SHAP matrix is built.
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  readout: "TE"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set_label: "tco"
  variant_set: "corrected"
  initiation_score: "both"
  clip_mode: "aggregate"
  neg_seed: 9
  explain_n: 200
  explain_sizes: "50,100,200,400"
  run_interactions: FALSE
  save_shap_plots: TRUE
  corr_group_threshold: 0.8
  cluster_k_max: 6
  cluster_n_null: 50
  # Sweep mode skips the two sections that are one-time checks rather than per-run
  # measurements: explain-set size stability (4 extra SHAP computations) and the subgroup
  # null (50 permutations x k values, much the slowest part). Running them per seed or per
  # experiment multiplies cost for no added information.
  sweep_mode: FALSE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Why SHAP here, and what it is not

Impurity importance cannot resolve **credit-splitting across correlated columns**. This matrix
is full of them: `cds_gc` / `gc3_internal` / `csc_internal`, the three G4 regions, the length
family. SHAP is **additive**, so summing within a family is exact rather than approximate, and
family-level attribution is the thing SHAP buys that `02_`'s Gini ranking structurally cannot.

Three things this notebook does **not** claim:

1. **SHAP is not tuned.** TreeSHAP is exact given a tree ensemble; there is no bandwidth or
   sample-size knob. What varies is *which model is explained*, *which genes are explained*,
   and *which summary is read* - so those are what get checked below.
2. **`mean_signed_shap` is not direction.** It averages to ~0 for every feature; its residual
   offset only reflects the gap between explain-set mean prediction and the training baseline.
   The directional quantity is `shap_value_cor` - the correlation of a feature's value with
   its own SHAP value.
3. **One run is one negative draw.** These rankings inherit the draw-to-draw instability that
   `02c_` quantifies for impurity. `neg_seed` is exposed so the same paired sweep can be run
   over SHAP.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(treeshap)
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
set.seed(9)
```

```{r load}
out_dir <- here("output", "translational_control_overview")

# Suffix is built by the SAME helper 02b_/02c_ use, so this cannot drift from 02_'s naming.
suffix <- sweep_suffix(list(condition         = params$condition,
                            gene              = params$gene,
                            direction         = params$direction,
                            readout           = params$readout,
                            timepoint         = params$timepoint,
                            lfc               = params$lfc,
                            feature_set_label = params$feature_set_label,
                            variant_set       = params$variant_set,
                            initiation_score  = params$initiation_score,
                            clip_mode         = params$clip_mode),
                       params$neg_seed)
cat("Experiment: eIF", params$gene, " ", params$direction, " | ", params$condition, " ",
    params$timepoint, " | LFC ", params$lfc, " | readout ", params$readout, "\n", sep = "")

model_path <- file.path(out_dir, paste0("rf_classifier_", suffix, ".rds"))
data_path  <- file.path(out_dir, paste0("rf_model_data_", suffix, ".rds"))
stopifnot(
  "notebook 02 model not found - knit 02_rf_model.Rmd with these params first" =
    file.exists(model_path),
  "notebook 02 model data not found - re-knit 02_ with save_model_data = TRUE" =
    file.exists(data_path)
)

rf_prob <- readRDS(model_path)
md      <- readRDS(data_path)
stopifnot("saved data does not belong to this model" = identical(md$suffix, suffix))

feature_cols <- md$feature_cols
train_data   <- md$train
test_data    <- md$test
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)

cat("Model:", suffix, "\n")
cat("  treetype:", rf_prob$treetype, "| num.trees:", rf_prob$num.trees,
    "| mtry:", rf_prob$mtry, "| min.node.size:", rf_prob$min.node.size, "\n")
cat("  train:", nrow(train_data), " test:", nrow(test_data),
    " features:", length(feature_cols), "\n")
```

# 1. The surrogate, and Check A

`treeshap` does not support ranger **probability** forests, so what gets explained is a
surrogate *regression* forest on the same 0/1 target and the same rows. Its hyperparameters are
**read off the saved model, never assumed** - ranger's defaults differ by tree type
(`min.node.size` is 10 for probability but 1 for classification and 5 for regression; `mtry` is
`floor(sqrt(p))` for probability but `floor(p/3)` for regression). Assuming them would explain
a different model and every number below would be confidently wrong.

```{r surrogate}
y_train <- as.integer(train_data$eif3d_target == md$pos_label)

fit_surrogate <- function(mtry, min.node.size) {
  set.seed(9)
  ranger(x = train_data[, feature_cols], y = y_train,
         num.trees = rf_prob$num.trees, mtry = mtry, min.node.size = min.node.size,
         importance = "none")
}
rf_shap <- fit_surrogate(rf_prob$mtry, rf_prob$min.node.size)

stopifnot(
  "surrogate is not a regression forest"          = rf_shap$treetype == "Regression",
  "surrogate mtry does not match 02"              = rf_shap$mtry == rf_prob$mtry,
  "surrogate min.node.size does not match 02"     = rf_shap$min.node.size == rf_prob$min.node.size,
  "surrogate num.trees does not match 02"         = rf_shap$num.trees == rf_prob$num.trees
)
cat("Check A passed: surrogate config matches the 02 model structurally\n")
```

# 2. Check B: fidelity, calibrated here rather than inherited

`07c_` in the frozen pipeline uses `fidelity_tol = 0.02` on `max |deviation|`, calibrated on a
**56-feature** set. CLAUDE.md is explicit that this is "a sanity bound, not a tuned one, on
other feature sets". Measuring it here rather than inheriting it shows the design does **not**
transfer, in two ways worth recording.

```{r fidelity}
pred_prob <- predict(rf_prob, data = test_data[, feature_cols])$predictions
pred_prob <- if (is.matrix(pred_prob)) pred_prob[, md$pos_label] else pred_prob

deviation_of <- function(model) {
  pr <- predict(model, data = test_data[, feature_cols])$predictions
  c(max_abs = max(abs(pr - pred_prob)), rmse = sqrt(mean((pr - pred_prob)^2)))
}

p <- length(feature_cols)
configs <- tribble(
  ~label,                ~kind,            ~mtry,        ~min.node.size,
  "matched (correct)",   "correct",        rf_prob$mtry, rf_prob$min.node.size,
  "min.node.size = 1",   "min.node.size",  rf_prob$mtry, 1,
  "min.node.size = 5",   "min.node.size",  rf_prob$mtry, 5,
  "min.node.size = 25",  "min.node.size",  rf_prob$mtry, 25,
  "mtry = floor(p/3)",   "mtry",           floor(p / 3), rf_prob$min.node.size,
  "mtry = 2",            "mtry",           2,            rf_prob$min.node.size
)
devs <- map2_dfr(configs$mtry, configs$min.node.size,
                 ~ as_tibble_row(deviation_of(fit_surrogate(.x, .y))))
configs <- bind_cols(configs, devs)
print(as.data.frame(configs %>% mutate(across(c(max_abs, rmse), ~ round(., 4)))))
```

**Finding 1: `max_abs` is the wrong statistic.** It is one worst-case point and is dominated by
noise, so it ranks deliberately misconfigured surrogates as *better* than the correct one
(`min.node.size = 25` scores 0.069 against the correct 0.111). A tolerance on `max_abs` would
pass a wrong surrogate and fail a right one. RMSE is used below.

**Finding 2: `min.node.size` is not detectable by fidelity on this matrix.** On the 56-feature
set, `min.node.size = 1` deviated 0.0798 against 0.0068 for the correct config - a 12x
separation. Here every value from 1 to 25 lands within a few percent of the correct RMSE. Only
`mtry` moves fidelity materially. So Check B is asserted **only against the misconfigurations
it can actually discriminate**, and `min.node.size` is guarded by Check A's structural equality
instead. Claiming otherwise would be the same unexamined transfer this fork exists to catch.

```{r fidelity_assert}
rmse_of <- function(lbl) configs$rmse[configs$label == lbl]
dev_correct <- rmse_of("matched (correct)")
dev_mtry    <- min(configs$rmse[configs$kind == "mtry"])
dev_mns     <- min(configs$rmse[configs$kind == "min.node.size"])

cat(sprintf("correct RMSE            %.4f\n", dev_correct))
cat(sprintf("nearest mtry error      %.4f  (%.2fx correct)  <- what Check B can detect\n",
            dev_mtry, dev_mtry / dev_correct))
cat(sprintf("nearest min.node.size   %.4f  (%.2fx correct)  <- NOT detectable here\n",
            dev_mns, dev_mns / dev_correct))

# What Check B can assert is the ABSOLUTE requirement: the surrogate must reproduce 02's
# held-out predictions closely enough to be worth explaining. The RELATIVE test - "further
# from wrong configurations than from the right one" - is reported but NOT asserted, because
# its power depends on the matrix and on this one it has very little:
#
#   nearest mtry error        1.49x the correct RMSE  (was 1.77x at 72 features)
#   nearest min.node.size     1.00x                   (undetectable, as before)
#
# Lowering a threshold until it passes would be fitting the check to the answer we want, so
# the honest position is that on this matrix fidelity does not discriminate configuration
# errors, and CHECK A's structural equality is the real guard. That is also the strongest
# argument for explaining the actual probability forest with a model-agnostic method
# (kernelshap) and retiring the surrogate altogether.
surr_cor <- cor(predict(rf_shap, data = test_data[, feature_cols])$predictions, pred_prob)
cat(sprintf("\nsurrogate vs 02 held-out predictions: r = %.4f, RMSE = %.4f\n",
            surr_cor, dev_correct))
cat(sprintf("relative separation (reported, not asserted): mtry %.2fx, min.node.size %.2fx\n",
            dev_mtry / dev_correct, dev_mns / dev_correct))
if (dev_mtry / dev_correct < 1.5)
  cat("NOTE: fidelity barely separates a wrong mtry here. Check A is carrying the guard.\n")

stopifnot(
  "surrogate does not track 02's held-out predictions closely enough to explain it" =
    surr_cor > 0.95 && dev_correct < 0.05
)
cat("Check B passed on the absolute requirement.\n")
```

# 3. SHAP

```{r shap}
explain_n  <- min(params$explain_n, nrow(test_data))
set.seed(9)
explain_idx <- sample(nrow(test_data), explain_n)
explain_set <- as.data.frame(test_data[explain_idx, feature_cols])

unified   <- ranger.unify(rf_shap, as.data.frame(train_data[, feature_cols]))
shap_out  <- treeshap(unified, explain_set, interactions = params$run_interactions,
                      verbose = FALSE)
shap_vals <- as.data.frame(shap_out$shaps)
stopifnot("SHAP matrix does not match the explain set" =
            nrow(shap_vals) == explain_n && ncol(shap_vals) == length(feature_cols))
cat("SHAP computed on", explain_n, "held-out genes x", length(feature_cols), "features\n")
```

```{r shap_table}
# shap_value_cor is the DIRECTIONAL summary: a feature's value against its own SHAP value.
shap_value_cor <- map_dbl(feature_cols, function(f) {
  v <- explain_set[[f]]; s <- shap_vals[[f]]
  if (sd(v, na.rm = TRUE) == 0 || sd(s, na.rm = TRUE) == 0) return(NA_real_)
  cor(v, s, use = "complete.obs", method = "spearman")
})

shap_tbl <- tibble(
  feature          = feature_cols,
  mean_abs_shap    = colMeans(abs(shap_vals)),
  mean_signed_shap = colMeans(shap_vals),      # NOT direction - see the header
  shap_value_cor   = shap_value_cor
) %>%
  left_join(ann %>% dplyr::select(feature, family, block), by = "feature") %>%
  arrange(desc(mean_abs_shap)) %>%
  mutate(rank_shap = row_number())

print(as.data.frame(shap_tbl %>% slice_head(n = 20) %>%
        mutate(across(where(is.numeric), ~ round(., 4)))))
```

# 4. Family attribution - the reason to use SHAP at all

SHAP is additive, so a family's contribution is the **exact** sum of its members'. Impurity
importance cannot do this: correlated columns split credit unpredictably between themselves.

```{r family}
fam_tbl <- shap_tbl %>%
  group_by(family) %>%
  summarise(n = n(), family_shap = sum(mean_abs_shap),
            top_feature = feature[which.max(mean_abs_shap)], .groups = "drop") %>%
  mutate(share_pct = round(100 * family_shap / sum(family_shap), 2)) %>%
  arrange(desc(family_shap))
print(as.data.frame(fam_tbl %>% mutate(family_shap = round(family_shap, 4))))
```

```{r vs_impurity}
imp <- read_csv(file.path(out_dir, paste0("rf_importance_", suffix, ".csv")),
                show_col_types = FALSE)
cmp <- shap_tbl %>%
  dplyr::select(feature, family, mean_abs_shap) %>%
  inner_join(imp %>% dplyr::select(feature, impurity = importance), by = "feature")
cat("SHAP vs impurity, feature level: Spearman",
    round(cor(cmp$mean_abs_shap, cmp$impurity, method = "spearman"), 3),
    ", n =", nrow(cmp), "\n")

fam_cmp <- cmp %>% group_by(family) %>%
  summarise(shap = sum(mean_abs_shap), impurity = sum(impurity), .groups = "drop") %>%
  mutate(shap_share = 100 * shap / sum(shap), imp_share = 100 * impurity / sum(impurity),
         delta = shap_share - imp_share) %>% arrange(desc(abs(delta)))
cat("\nFamily share, SHAP vs impurity (delta = where the two disagree):\n")
print(as.data.frame(fam_cmp %>% mutate(across(where(is.numeric), ~ round(., 2))) %>%
                      dplyr::select(family, shap_share, imp_share, delta)))
```

The correlation is the uninteresting part. **The disagreements are the point**: impurity
cannot resolve credit-splitting across correlated columns, so a feature that impurity ranks
far above or below its SHAP attribution is one whose importance was being borrowed from, or
lent to, a correlated neighbour.

```{r fig_impurity_vs_shap, fig.width = 7.5, fig.height = 5.5, eval = !params$sweep_mode}
library(ggrepel)

cmp_p <- cmp %>%
  mutate(shap_share = 100 * mean_abs_shap / sum(mean_abs_shap),
         imp_share  = 100 * impurity / sum(impurity),
         rank_shap  = rank(-mean_abs_shap), rank_imp = rank(-impurity),
         shift      = rank_imp - rank_shap)

lab <- cmp_p %>% slice_max(abs(shift), n = 8)
rho <- cor(cmp_p$mean_abs_shap, cmp_p$impurity, method = "spearman")

p_fs <- ggplot(cmp_p, aes(imp_share, shap_share, colour = family)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 2, alpha = 0.85) +
  geom_text_repel(data = lab, aes(label = feature), size = 2.8, max.overlaps = 20,
                  show.legend = FALSE, min.segment.length = 0) +
  scale_x_log10() + scale_y_log10() +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.6,
           label = sprintf("Spearman = %.3f, n = %d", rho, nrow(cmp_p))) +
  labs(x = "impurity (Gini) share of total (%, log scale)",
       y = "SHAP share of total (%, log scale)",
       title = "Where impurity and SHAP disagree",
       subtitle = "dashed line = perfect agreement; labelled points are the 8 largest rank shifts") +
  theme_classic(base_size = 12)
print(p_fs)
ggsave(here("plots", paste0("03_impurity_vs_shap_", suffix, ".pdf")), p_fs,
       width = 7.5, height = 5.5)

cat("\nLargest rank shifts (positive = impurity ranks it LOWER than SHAP does):\n")
print(as.data.frame(lab %>% arrange(desc(shift)) %>%
        transmute(feature, family, rank_imp, rank_shap, shift,
                  imp_share = round(imp_share, 2), shap_share = round(shap_share, 2))))
```

```{r fig_feature_slope, fig.width = 11, fig.height = 7, eval = !params$sweep_mode}
# The family slope chart at FEATURE resolution, in two cuts.
#   (a) the 20 features SHAP ranks highest - what the model actually leans on
#   (b) the 20 that MOVE most between the two methods - where impurity was borrowing or
#       lending credit. Point size is SHAP share throughout, because a feature can move a
#       long way and still be negligible; without size, (b) would flatter trivial features.
feat_slope <- function(d, ttl, sub) {
  long <- d %>%
    dplyr::select(feature, family, impurity = imp_share, SHAP = shap_share) %>%
    pivot_longer(c(impurity, SHAP), names_to = "method", values_to = "share") %>%
    mutate(method = factor(method, levels = c("impurity", "SHAP")))
  ggplot(long, aes(method, share, group = feature)) +
    geom_line(aes(colour = family), linewidth = 0.7, alpha = 0.85, show.legend = FALSE) +
    geom_point(aes(colour = family, size = share), alpha = 0.9, show.legend = FALSE) +
    ggrepel::geom_text_repel(
      data = long %>% filter(method == "SHAP"),
      aes(label = feature, colour = family), hjust = 0, nudge_x = 0.06, direction = "y",
      size = 2.9, show.legend = FALSE, segment.size = 0.2, max.overlaps = 30) +
    scale_size_continuous(range = c(1.5, 5)) +
    scale_x_discrete(expand = expansion(mult = c(0.10, 0.75))) +
    labs(x = NULL, y = "share of total attribution (%)", title = ttl, subtitle = sub) +
    theme_classic(base_size = 11)
}

top_shap <- cmp_p %>% slice_max(shap_share, n = 20)
movers   <- cmp_p %>% mutate(delta = shap_share - imp_share) %>%
  slice_max(abs(delta), n = 20)

pa <- feat_slope(top_shap, "(a) Top 20 features by SHAP",
                 sprintf("point size = SHAP share. n = %d features in the model",
                         nrow(cmp_p)))
pb <- feat_slope(movers, "(b) Top 20 movers between the two methods",
                 sprintf("point size = SHAP share, so a large move on a small feature stays small. n = %d",
                         nrow(cmp_p)))
print(cowplot::plot_grid(pa, pb, nrow = 1))
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_feature_slope_impurity_vs_shap_", suffix, ".pdf")),
         cowplot::plot_grid(pa, pb, nrow = 1), width = 11, height = 7)

cat("\nThe movers, with their actual importance:\n")
print(as.data.frame(movers %>% arrange(desc(shap_share - imp_share)) %>%
        transmute(feature, family,
                  imp_share = round(imp_share, 2), shap_share = round(shap_share, 2),
                  delta_pp = round(shap_share - imp_share, 2),
                  rank_imp, rank_shap, in_top20_shap = rank_shap <= 20)))
```

```{r fig_family_slope, fig.width = 6.5, fig.height = 5, eval = !params$sweep_mode}
slope <- fam_cmp %>%
  dplyr::select(family, impurity = imp_share, SHAP = shap_share) %>%
  pivot_longer(c(impurity, SHAP), names_to = "method", values_to = "share") %>%
  mutate(method = factor(method, levels = c("impurity", "SHAP")))

p_fam <- ggplot(slope, aes(method, share, group = family)) +
  geom_line(aes(colour = family), linewidth = 0.8, show.legend = FALSE) +
  geom_point(aes(colour = family), size = 2.4, show.legend = FALSE) +
  geom_text_repel(data = slope %>% filter(method == "SHAP"),
                  aes(label = family, colour = family), hjust = 0, nudge_x = 0.08,
                  direction = "y", size = 3, show.legend = FALSE, segment.size = 0.2) +
  scale_x_discrete(expand = expansion(mult = c(0.12, 0.45))) +
  labs(x = NULL, y = "share of total attribution (%)",
       title = "Family attribution: impurity vs SHAP",
       subtitle = sprintf("SHAP sums exactly within a family; impurity does not. n = %d features",
                          nrow(cmp_p))) +
  theme_classic(base_size = 12)
print(p_fam)
ggsave(here("plots", paste0("03_family_impurity_vs_shap_", suffix, ".pdf")), p_fam,
       width = 6.5, height = 5)
```

# 3b. Beeswarm

The canonical SHAP figure, and the only one that shows magnitude, **direction** and
**spread across genes** at once. Every point is one gene; position is that gene's SHAP value
and colour is its feature value, so a clean colour gradient along the x axis means a
monotonic effect and a scrambled one means the feature acts through interactions.

Our tables give these three things separately: `mean_abs_shap` is magnitude,
`shap_value_cor` is direction, and nothing reports the spread.

```{r fig_beeswarm, fig.width = 8, fig.height = 9, eval = !params$sweep_mode}
library(shapviz)
sv <- shapviz(as.matrix(shap_vals), X = explain_set)
p_bee <- sv_importance(sv, kind = "beeswarm", max_display = 25) +
  labs(title = "SHAP beeswarm, top 25 features",
       subtitle = sprintf("one point per gene; colour = feature value. n = %d explained genes",
                          nrow(explain_set))) +
  theme_classic(base_size = 10)
print(p_bee)
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_beeswarm_", suffix, ".pdf")), p_bee, width = 8, height = 9)
```

```{r fig_dependence, fig.width = 10, fig.height = 6.5, eval = !params$sweep_mode}
# Dependence for the strongest features: the SHAPE of the relationship, which a correlation
# cannot show. A threshold effect and a linear one give the same shap_value_cor.
top6 <- shap_tbl$feature[1:6]
deps <- lapply(top6, function(f)
  sv_dependence(sv, v = f, color_var = NULL) + theme_classic(base_size = 9))
print(cowplot::plot_grid(plotlist = deps, nrow = 2))
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_dependence_", suffix, ".pdf")),
         cowplot::plot_grid(plotlist = deps, nrow = 2), width = 10, height = 6.5)
```

# 4a. One signed column per feature

A single column of every feature needs a fill that is **signed, additive and interpretable**.
Only one candidate is all three.

| | signed | additive | comparable across runs |
|---|---|---|---|
| mean \|SHAP\| share (%) | no | yes (sums to 100) | **yes** |
| mean signed SHAP, **all** genes | yes | yes | **useless - cancels to ~0** |
| **mean signed SHAP, positive class only** | **yes** | **yes** | no (raw scale) |
| mean \|SHAP\| x sign(cor) | yes | **no** | yes |

Averaging the signed values over *all* explained genes is the trap: positives and negatives
pull opposite ways and cancel, which is why `mean_signed_shap` averages to ~0 for every feature
and is exported for completeness only. Restricting to the **positive class** removes the
cancellation and keeps additivity exactly - the column sums to the mean lift in predicted
probability above baseline for a target gene, which is asserted below.

Multiplying a magnitude by a sign is the other trap: it looks like the answer, sums to nothing
meaningful, and is simply wrong for any feature whose effect is non-monotonic - and 14 of 69
features here have a direction that flips between negative draws.

`shap_value_cor` is not additive at all. It is a bounded correlation answering "which way",
never "how much".

```{r fig_signed_column, fig.width = 7.5, fig.height = 12, eval = !params$sweep_mode}
pos_rows <- test_data$eif3d_target[explain_idx] == md$pos_label
signed <- tibble(feature = feature_cols,
                 signed_shap = colMeans(shap_vals[pos_rows, , drop = FALSE])) %>%
  left_join(shap_tbl %>% dplyr::select(feature, family, mean_abs_shap), by = "feature") %>%
  mutate(abs_share = 100 * mean_abs_shap / sum(mean_abs_shap),
         # For comparing ACROSS experiments: divide by a common scale, never z-score.
         # See the note below - scaling preserves additivity, centring destroys it.
         signed_share = 100 * signed_shap / sum(abs(signed_shap)))

# Additivity is the claim, so it is asserted rather than assumed: the column must sum to the
# mean of the per-gene row sums, which for TreeSHAP is (prediction - baseline).
row_sums <- rowSums(shap_vals[pos_rows, , drop = FALSE])
stopifnot("the signed column is not additive - do not read it as a decomposition" =
            isTRUE(all.equal(sum(signed$signed_shap), mean(row_sums))))
cat(sprintf("Additive: column sums to %.5f = mean lift above baseline for a target gene (n = %d positives)\n",
            sum(signed$signed_shap), sum(pos_rows)))

lim <- max(abs(signed$signed_shap))
p_sign <- signed %>%
  mutate(feature = fct_reorder(feature, signed_shap)) %>%
  ggplot(aes(x = 1, y = feature, fill = signed_shap)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%+.4f", signed_shap)), size = 2.4) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0,
                       limits = c(-lim, lim), name = "mean signed\nSHAP (positives)") +
  scale_x_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL,
       title = "Signed contribution to a typical eIF3d target",
       subtitle = sprintf(
         "mean signed SHAP over the %d positive-class genes. Vermillion pushes toward target, blue away.\nAdditive: the column sums to %.4f, the mean lift in predicted probability above baseline.",
         sum(pos_rows), sum(signed$signed_shap)),
       # The rationale travels with the figure, so the standalone PDF is self-explanatory.
       caption = paste(
         "Why this quantity: it is the only summary that is signed, additive AND interpretable.",
         "Mean |SHAP| share drops the sign. Mean signed SHAP over ALL explained genes cancels to ~0,",
         "since positives and negatives pull opposite ways. Mean |SHAP| x sign(correlation) is not",
         "additive and is wrong for non-monotonic features (14 of 69 flip sign between negative draws).",
         "To compare ACROSS experiments use signed_share = 100 x value / sum(|values|): scaling by a",
         "common constant keeps additivity, whereas z-scoring subtracts a mean and distorts a group",
         "by its size (termination: raw +0.0004 but z-sum -5.84, from having 11 features).",
         sep = "\n")) +
  theme_classic(base_size = 9) +
  theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(),
        plot.caption = element_text(hjust = 0, size = 6.5, colour = "grey30"),
        plot.caption.position = "plot")
print(p_sign)
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_signed_column_", suffix, ".pdf")), p_sign,
         width = 7.5, height = 12)

write_csv(signed, file.path(out_dir, paste0("shap_signed_positive_", suffix, ".csv")))
```

## Comparing this across experiments: scale, do not z-score

The signed column above is on each run's own probability scale, so it is not directly
comparable between experiments. The fix is to normalise per experiment - but **which**
normalisation matters, and the obvious one is wrong.

**Z-scoring destroys the additivity that made this column worth plotting.** Subtracting a mean
means a group of *k* features loses *k* x mean, so the distortion scales with how many features
a group contains. Measured on this run:

| family | raw signed | `signed_share` | z-score | n |
|---|---|---|---|---|
| `length` | +0.0283 | 25.27 | +7.82 | 4 |
| `termination` | **+0.0004** | 0.37 | **-5.84** | **11** |
| `initiation` | +0.00006 | 0.06 | -2.16 | 4 |

`termination` carries essentially no signed signal yet gets the largest-magnitude z-sum of any
family, purely for having 11 features. Empirically `signed_share / raw` is a **constant across
every family**, while `zscore / raw` ranges from +276 to -34,530.

**Dividing by a common scale keeps everything:** `signed_share = 100 * signed_shap /
sum(|signed_shap|)` is signed, comparable across experiments, and still additive, because
dividing every element by the *same* constant leaves group totals exactly proportional.
Centring is the step that breaks it. Two further conditions for the cross-experiment matrix:
fix the feature universe (a feature dropped in one experiment must not read as "unimportant"),
and carry each run's AUC, since a ranking from a weak model does not mean what a strong one's
does.

A second reason to avoid z-scores here: most features sit near zero and a few are large, so the
sd is set by the tail. On this run 29 of 69 features have |z| < 0.5 - crushed together - while
the maximum is 4.0.

# 4b. Correlated features: grouping by measured redundancy, not by hand

Family sums fix credit-splitting only where the families were drawn correctly in the first
place. `cds_gc` and `gc3_internal` correlate at 0.93 and sit in the same family, so that split
is absorbed - but a hand-drawn boundary cannot catch redundancy it did not anticipate.

The data-driven alternative: cluster features on `1 - |r|`, cut at a threshold, and sum SHAP
within each cluster. **Additivity makes any grouping exact**, so this costs nothing in rigour
and does not require knowing in advance what is redundant. The threshold is a choice, so it is
reported and varied.

```{r corr_groups, eval = !params$sweep_mode}
X <- as.matrix(rbind(train_data[, feature_cols], test_data[, feature_cols]))
Cm <- cor(X, use = "pairwise.complete.obs")
hc <- hclust(as.dist(1 - abs(Cm)), method = "average")

cat("Sensitivity of the grouping to the threshold:\n")
for (thr in c(0.7, 0.8, 0.9)) {
  cl <- cutree(hc, h = 1 - thr)
  cat(sprintf("  |r| >= %.1f : %2d groups from %d features (%d singletons)\n",
              thr, max(cl), length(cl), sum(table(cl) == 1)))
}

cl <- cutree(hc, h = 1 - params$corr_group_threshold)
grp <- tibble(feature = names(cl), grp = cl) %>%
  left_join(shap_tbl %>% dplyr::select(feature, family, mean_abs_shap), by = "feature") %>%
  mutate(share = 100 * mean_abs_shap / sum(mean_abs_shap))

grp_tbl <- grp %>% group_by(grp) %>%
  summarise(n = n(), group_share = sum(share),
            families = paste(sort(unique(family)), collapse = " + "),
            spans_families = n_distinct(family) > 1,
            label = feature[which.max(share)],          # name a group by its strongest member
            members = paste(feature, collapse = ", "), .groups = "drop") %>%
  arrange(desc(group_share))

cat("\n=== multi-feature groups at |r| >=", params$corr_group_threshold, "===\n")
print(as.data.frame(grp_tbl %>% filter(n > 1) %>%
        dplyr::select(n, group_share, families, spans_families, members) %>%
        mutate(group_share = round(group_share, 2))), right = FALSE)

# A group that crosses a family boundary is the case hand-defined families CANNOT catch: the
# family totals understate a shared axis and split it across two rows of the family table.
cross <- grp_tbl %>% filter(n > 1, spans_families)
cat("\nGroups spanning more than one family:", nrow(cross), "\n")
if (nrow(cross)) {
  cat("These are where the family view is misleading - the family totals divide one\n")
  cat("empirical axis between two rows:\n")
  print(as.data.frame(cross %>% dplyr::select(group_share, families, members) %>%
                        mutate(group_share = round(group_share, 2))), right = FALSE)
}
```

```{r fig_corr_groups, fig.width = 8, fig.height = 5.5, eval = !params$sweep_mode}
top_grp <- grp_tbl %>% slice_max(group_share, n = 20) %>%
  mutate(lab = if_else(n > 1, sprintf("%s (+%d)", label, n - 1L), label),
         lab = fct_reorder(lab, group_share))

p_grp <- ggplot(top_grp, aes(lab, group_share)) +
  geom_segment(aes(xend = lab, y = 0, yend = group_share), colour = "grey60",
               linewidth = 0.6) +
  geom_point(aes(colour = spans_families, size = n)) +
  coord_flip() +
  scale_colour_manual(values = c(`FALSE` = "#0072B2", `TRUE` = "#D55E00"),
                      name = "spans >1 family") +
  scale_size_continuous(range = c(2.5, 5.5), name = "features") +
  labs(x = NULL, y = "summed SHAP share (%)",
       title = sprintf("SHAP summed over correlated groups (|r| >= %.1f)",
                       params$corr_group_threshold),
       subtitle = sprintf("groups named by their strongest member; vermillion = crosses a family boundary. n = %d features in %d groups",
                          nrow(grp), max(cl))) +
  theme_classic(base_size = 11)
print(p_grp)
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_shap_corr_groups_", suffix, ".pdf")), p_grp,
         width = 8, height = 5.5)

write_csv(grp_tbl, file.path(out_dir, paste0("shap_corr_groups_", suffix, ".csv")))
```

**Reading it.** Where the correlation grouping and the family grouping agree, the family view is
fine. Where a group crosses a family boundary, the correlation grouping is the honest number:
the family totals are dividing one empirical axis between two rows. For the cross-experiment
comparison, correlated-group totals are also the more robust unit, since individual features
can split credit differently between experiments while their group total stays put.

# 5. Is the explain set big enough?

`explain_n` is a free choice, and nothing elsewhere justifies it. If the family ranking is not
stable across sizes, the number is doing work it should not be.

```{r explain_stability, eval = !params$sweep_mode}
sizes <- as.integer(str_trim(str_split(params$explain_sizes, ",")[[1]]))
sizes <- sizes[sizes <= nrow(test_data)]

fam_at <- function(n) {
  set.seed(9)
  idx <- sample(nrow(test_data), n)
  sv  <- as.data.frame(treeshap(unified, as.data.frame(test_data[idx, feature_cols]),
                                interactions = FALSE, verbose = FALSE)$shaps)
  tibble(feature = feature_cols, mas = colMeans(abs(sv))) %>%
    left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
    group_by(family) %>% summarise(share = sum(mas), .groups = "drop") %>%
    mutate(share = 100 * share / sum(share), n = n)
}
stab <- map_dfr(sizes, fam_at)

stab %>% pivot_wider(names_from = n, values_from = share, names_prefix = "n") %>%
  arrange(desc(.data[[paste0("n", max(sizes))]])) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>% as.data.frame() %>% print()

rank_cor <- stab %>% group_by(n) %>% mutate(rk = rank(-share)) %>%
  dplyr::select(family, n, rk) %>% pivot_wider(names_from = n, values_from = rk)
cat("\nSpearman of family ranking, smallest vs largest explain set:",
    round(cor(rank_cor[[2]], rank_cor[[ncol(rank_cor)]], method = "spearman"), 3), "\n")
```

# 6. Are there subgroups of targets with different drivers?

Everything above is a **global** ranking: one feature order averaged over all genes. It cannot
say whether there are distinct classes of eIF3d target driven by different features - a
GC/structure-driven set versus a codon-driven set, say. Given that several feature families
turned out real-but-redundant globally, a family that matters for only a subset of genes is
exactly the pattern a global mean hides.

This is the one idea worth borrowing from the `explainer` package (`SHAPclust()`), done here on
our **exact** SHAP values rather than through an mlr3 wrapper with a sampling estimator.

Clustering is on the **positive class only** - including negatives would mostly recover the
class split, which we already know - and on **family-level** profiles, which is exact because
SHAP is additive. Each gene's profile is normalised by its own total absolute contribution, so
clusters reflect *which* families drove a gene, not how confidently it was called.

```{r subgroup_shap, eval = !params$sweep_mode}
pos_test <- which(test_data$eif3d_target == md$pos_label)
cat("Positive-class genes in the held-out set:", length(pos_test), "\n")

pos_x    <- as.data.frame(test_data[pos_test, feature_cols])
shap_pos <- as.data.frame(treeshap(unified, pos_x, interactions = FALSE,
                                   verbose = FALSE)$shaps)

sym_all <- test_data$symbol[pos_test]
fam_of <- ann$family[match(feature_cols, ann$feature)]
fam_mat <- t(rowsum(t(as.matrix(shap_pos)), group = fam_of))   # exact family sums per gene

# Normalise within gene: the profile is the share of that gene's attribution, signed.
denom     <- rowSums(abs(fam_mat)); denom[denom == 0] <- NA
profile   <- fam_mat / denom
keep_rows <- stats::complete.cases(profile)      # index, so every later matrix aligns
profile   <- profile[keep_rows, , drop = FALSE]
sym_keep  <- sym_all[keep_rows]
cat("Profiles:", nrow(profile), "genes x", ncol(profile), "families\n")
```

```{r subgroup_k, eval = !params$sweep_mode}
source(here("code", "translational_control_overview", "03_subgroup_null.R"))

# The null permutes each family independently across genes: co-occurrence destroyed, marginals
# preserved. The STATISTIC is what needs care - mean silhouette falls with k, so a raw
# best-over-k comparison pits the null's k = 2 against an observed best at k = 3 and gets the
# answer backwards. subgroup_null() standardises within k first. See 03_subgroup_null.R.
nullres <- subgroup_null(profile, ks = 2:params$cluster_k_max, n_perm = params$cluster_n_null)
print(nullres$per_k %>% mutate(across(where(is.numeric), ~ round(., 4))))

best_k <- nullres$best_k_sil   # describe at the silhouette peak; test uses best_k_z
cat(sprintf("\ntest statistic peaks at k = %d: observed z = %.2f | null median z = %.2f | null max z = %.2f\n",
            nullres$best_k_z, nullres$z_obs, median(nullres$z_null), max(nullres$z_null)))
cat(sprintf("clusters described at k = %d, where the raw silhouette peaks (%.4f)\n",
            nullres$best_k_sil, max(nullres$obs)))
cat(sprintf("empirical p = %.4f (floor 1/%d), n = %d genes\n",
            nullres$p_overall, nullres$n_perm + 1, nrow(profile)))

if (nullres$p_overall > 0.05) {
  cat("\nNOT separable from the null: report no detectable subgroup structure,\n",
      "not k clusters. The cluster means below are then descriptive only.\n")
} else {
  cat("\nSeparable from the null: the profiles carry real co-occurrence structure.\n")
}
```

```{r fig_subgroup_heatmap, fig.width = 8, fig.height = 6, eval = !params$sweep_mode}
# Genes x families, each row a gene's normalised SHAP profile. Rows ordered by cluster, so
# whether the clusters correspond to visible blocks is a question you can answer by looking.
library(pheatmap)

set.seed(9)
km_hm <- kmeans(profile, centers = nullres$best_k_sil, nstart = 25)
ord   <- order(km_hm$cluster)

hm_mat <- profile[ord, , drop = FALSE]                 # a copy, so profile keeps its indexing
rownames(hm_mat) <- make.unique(sym_keep)[ord]
ann_row <- data.frame(cluster = factor(km_hm$cluster[ord]))
rownames(ann_row) <- rownames(hm_mat)

lim <- max(abs(profile))
pheatmap(hm_mat,
         cluster_rows = FALSE, cluster_cols = TRUE,
         show_rownames = FALSE, annotation_row = ann_row,
         # CLAUDE.md diverging convention: Okabe-Ito dark blue to white to vermillion
         color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
         breaks = seq(-lim, lim, length.out = 102),
         main = sprintf("Per-gene SHAP profile by family (n = %d positive-class genes)",
                        nrow(profile)),
         filename = if (params$save_shap_plots)
           here("plots", paste0("03_subgroup_heatmap_", suffix, ".pdf")) else NA,
         width = 8, height = 6)
```

```{r subgroup_resolution, eval = !params$sweep_mode}
# Does the answer depend on RESOLUTION? Family-level profiles are exact sums and denoise
# credit-splitting; per-FEATURE profiles are higher resolution but let correlated columns
# (cds_gc vs gc3_internal) divide a gene's credit arbitrarily between themselves. If
# per-feature clustering finds MORE structure, suspect that split rather than biology.
prof_feat <- as.matrix(shap_pos)[keep_rows, , drop = FALSE]   # same genes as `profile`
den_f <- rowSums(abs(prof_feat)); den_f[den_f == 0] <- 1
prof_feat <- prof_feat / den_f

null_feat <- subgroup_null(prof_feat, ks = 2:params$cluster_k_max,
                           n_perm = params$cluster_n_null)
cat("family-level (", ncol(profile), " dims): z = ", round(nullres$z_obs, 2),
    ", p = ", signif(nullres$p_overall, 3), "\n", sep = "")
cat("per-feature  (", ncol(prof_feat), " dims): z = ", round(null_feat$z_obs, 2),
    ", p = ", signif(null_feat$p_overall, 3), "\n", sep = "")

# Do the two resolutions agree on WHICH genes group together?
set.seed(9); k_f <- kmeans(prof_feat, centers = nullres$best_k_sil, nstart = 25)$cluster
set.seed(9); k_m <- kmeans(profile,     centers = nullres$best_k_sil, nstart = 25)$cluster
tab <- table(family_level = k_m, per_feature = k_f)
print(tab)
agree <- sum(apply(tab, 1, max)) / sum(tab)
cat(sprintf("\nbest-matching agreement between the two resolutions: %.1f%% of genes\n",
            100 * agree))
```

Per-feature reports a far larger z than family-level, and that is a **warning, not a win**:
69 dimensions give correlated columns room to divide a gene's credit between themselves, and
clustering will organise genes by that split. The agreement below is what shows the extra
resolution is not buying a different answer.

```{r fig_resolution_agreement, fig.width = 10, fig.height = 4.5, eval = !params$sweep_mode}
# Each cluster label is arbitrary, so the two partitions are first matched by their best
# overlap - otherwise "agreement" would depend on which cluster kmeans happened to call 1.
best_match <- apply(tab, 1, which.max)
flow <- as.data.frame(tab) %>%
  mutate(family_level = as.integer(as.character(family_level)),
         per_feature  = as.integer(as.character(per_feature)),
         agrees = per_feature == best_match[family_level],
         pct = 100 * Freq / sum(Freq))

AGREE_COL <- "#009E73"; DISAGREE_COL <- "grey70"

# (a) the contingency table itself
p_tile <- ggplot(flow, aes(factor(per_feature), factor(family_level))) +
  geom_tile(aes(fill = agrees), colour = "white", linewidth = 1.2, show.legend = FALSE) +
  geom_text(aes(label = ifelse(Freq > 0, sprintf("%d\n%.1f%%", Freq, pct), "")),
            size = 3.6, lineheight = 0.95) +
  scale_fill_manual(values = c(`TRUE` = alpha(AGREE_COL, 0.45),
                               `FALSE` = alpha(DISAGREE_COL, 0.35))) +
  labs(x = "per-feature cluster (69 dims)", y = "family-level cluster (14 dims)",
       title = "(a) Same genes, two resolutions",
       subtitle = sprintf("green = matched clusters. n = %d genes", sum(tab))) +
  theme_classic(base_size = 12)

# (b) the same thing as flows, where the eye reads agreement as straight-through ribbons
p_flow <- ggplot(flow %>% filter(Freq > 0)) +
  geom_segment(aes(x = 1, xend = 2, y = family_level, yend = per_feature,
                   linewidth = Freq, colour = agrees), alpha = 0.7, show.legend = FALSE) +
  geom_point(aes(x = 1, y = family_level), size = 4, colour = "grey30") +
  geom_point(aes(x = 2, y = per_feature),  size = 4, colour = "grey30") +
  scale_linewidth_continuous(range = c(0.5, 9)) +
  scale_colour_manual(values = c(`TRUE` = AGREE_COL, `FALSE` = DISAGREE_COL)) +
  scale_x_continuous(breaks = c(1, 2), labels = c("family-level", "per-feature"),
                     limits = c(0.85, 2.15)) +
  scale_y_reverse(breaks = sort(unique(c(flow$family_level, flow$per_feature)))) +
  labs(x = NULL, y = "cluster",
       title = "(b) Where the genes go",
       subtitle = sprintf("ribbon width = genes; %.1f%% stay in the matched cluster",
                          100 * agree)) +
  theme_classic(base_size = 12)

print(cowplot::plot_grid(p_tile, p_flow, nrow = 1, rel_widths = c(1, 1)))
if (params$save_shap_plots)
  ggsave(here("plots", paste0("03_resolution_agreement_", suffix, ".pdf")),
         cowplot::plot_grid(p_tile, p_flow, nrow = 1), width = 10, height = 4.5)
```

```{r subgroup_characterise, eval = !params$sweep_mode}
set.seed(9)
km  <- kmeans(profile, centers = best_k, nstart = 25)
gid <- test_data$gene_id_clean[pos_test][stats::complete.cases(fam_mat / denom)]
sym <- test_data$symbol[pos_test][stats::complete.cases(fam_mat / denom)]

cluster_means <- as_tibble(profile) %>%
  mutate(cluster = km$cluster) %>%
  group_by(cluster) %>%
  summarise(n = n(), across(everything(), mean), .groups = "drop")
print(as.data.frame(cluster_means %>% mutate(across(where(is.numeric), ~ round(., 3)))))

cat("\nTop drivers per cluster (largest mean signed share):\n")
for (k in sort(unique(km$cluster))) {
  v <- colMeans(profile[km$cluster == k, , drop = FALSE])
  top <- sort(v, decreasing = TRUE)[1:3]
  cat(sprintf("  cluster %d (n = %3d): %s\n", k, sum(km$cluster == k),
              paste(sprintf("%s %+.3f", names(top), top), collapse = " | ")))
  cat("      e.g. ", paste(head(sym[km$cluster == k], 6), collapse = ", "), "\n")
}

subgroups <- tibble(gene_id_clean = gid, symbol = sym, cluster = km$cluster) %>%
  bind_cols(as_tibble(profile))
```

```{r export}
write_csv(shap_tbl, file.path(out_dir, paste0("shap_feature_", suffix, ".csv")))

# The full per-gene x per-feature matrix. Everything else in this notebook is this object
# collapsed by averaging down the columns; exporting it lets a gene be interrogated directly.
if (!params$sweep_mode)
  write_csv(bind_cols(tibble(gene_id_clean = test_data$gene_id_clean[explain_idx],
                             symbol = test_data$symbol[explain_idx]), shap_vals),
            file.path(out_dir, paste0("shap_per_gene_", suffix, ".csv")))
write_csv(fam_tbl,  file.path(out_dir, paste0("shap_family_", suffix, ".csv")))
if (!params$sweep_mode) {
  write_csv(stab,      file.path(out_dir, paste0("shap_explain_stability_", suffix, ".csv")))
  write_csv(subgroups, file.path(out_dir, paste0("shap_subgroups_", suffix, ".csv")))
  cat("Wrote 4 SHAP files with suffix:", suffix, "\n")
} else {
  cat("Wrote 2 SHAP files (sweep_mode) with suffix:", suffix, "\n")
}
```
