---
title: "Which SHAP engine? TreeSHAP on a surrogate vs Kernel SHAP on the real classifier"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
params:
  # Must match the model being explained. This notebook loads that saved model and never
  # refits it, so both engines see exactly the same forest, the same split and the same genes.
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  readout: "TE"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set_label: "tco"
  variant_set: "corrected"
  initiation_score: "both"
  clip_mode: "aggregate"
  neg_seed: 9
  explain_n: 200
  # Kernel SHAP arms: one per (background size, background draw). The first is the reference.
  bg_arms: "100/9,100/17,300/9"
  # The four SHAP matrices take ~25 min to compute. Cached so the figures can be re-knit in
  # seconds; delete the cache file or set FALSE to force a recompute.
  use_cache: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# The question, in plain language

The model is a random forest that predicts whether a transcript is an eIF3d target. To say
*which features drive that prediction*, we need an **estimator**: an algorithm that turns the
fitted forest into a per-feature contribution for every gene. There are two on offer, and they
are not the same thing.

**TreeSHAP on a surrogate.** The `treeshap` package cannot read a ranger *probability* forest,
so we fit a stand-in **regression** forest on the same genes and the same 0/1 target, and read
its tree structure. The values are exact - for the stand-in. The risk is that the stand-in is
not the model we actually fit.

**Kernel SHAP on the real classifier.** This one only needs a prediction function, so it
explains the forest we actually fit. Nothing stands in for anything. The risk moves instead to
two places: the values are *approximated* by sampling, and the method needs a **background
set** - a group of reference genes whose values stand in for a feature being "absent". That
background is an arbitrary choice, exactly like the negative draw is.

So the choice is not "approximate vs exact". It is **which approximation you would rather
make**: a stand-in model, or a stand-in for missing data.

## Terms used throughout

| term | meaning |
|---|---|
| **share** | a feature's mean \|SHAP\| as a percentage of all features' total. "How much of the model's decision-making this feature accounts for." |
| **the draw** | the model keeps every target gene and randomly samples an equal number of non-targets from a larger pool. Which non-targets get sampled is the draw. |
| **draw noise** | how much a feature's share moves when *only* the draw changes. Measured over 20 draws of this same model. It is the yardstick: a difference smaller than this is not worth interpreting. |
| **background** | the reference genes Kernel SHAP substitutes in to represent a feature being absent. Set by size and by which genes were drawn. |
| **signed share** | mean *signed* SHAP over the target-class genes, scaled to sum to 100. Says which **direction** a feature pushes, where share says only how much. |

## What this notebook decides

Whether the two engines can be used interchangeably. If they agree to within draw noise, the
stand-in model was harmless and the cheap engine wins. If they disagree by more than draw
noise, then numbers computed with one engine cannot be compared with numbers computed with the
other, and a choice has to be made deliberately.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(treeshap)
library(kernelshap)
library(ggrepel)
library(knitr)
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
set.seed(9)

eng_cols <- c("TreeSHAP (surrogate)" = "#0072B2", "Kernel SHAP (classifier)" = "#D55E00")
```

```{r load}
out_dir <- here("output", "translational_control_overview")
suffix <- sweep_suffix(list(condition = params$condition, gene = params$gene,
                            direction = params$direction, readout = params$readout,
                            timepoint = params$timepoint, lfc = params$lfc,
                            feature_set_label = params$feature_set_label,
                            variant_set = params$variant_set,
                            initiation_score = params$initiation_score,
                            clip_mode = params$clip_mode), params$neg_seed)

rf  <- readRDS(file.path(out_dir, paste0("rf_classifier_", suffix, ".rds")))
md  <- readRDS(file.path(out_dir, paste0("rf_model_data_", suffix, ".rds")))
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
fc  <- md$feature_cols
stopifnot("this is not the probability forest the comparison is about" =
            rf$treetype == "Probability estimation")

arms <- str_split(params$bg_arms, ",")[[1]] %>% str_trim() %>%
  map_dfr(~ tibble(bg_n = as.integer(str_split(.x, "/")[[1]][1]),
                   bg_seed = as.integer(str_split(.x, "/")[[1]][2]))) %>%
  mutate(label = sprintf("kernel bg=%d seed=%d", bg_n, bg_seed))
```

```{r compute}
cache_path <- file.path(out_dir, paste0("shap_engine_matrices_", suffix, ".rds"))

if (params$use_cache && file.exists(cache_path)) {
  M <- readRDS(cache_path)
  stopifnot("cached matrices do not match this model" = identical(M$fc, fc))
  cat("Loaded cached SHAP matrices. Delete", basename(cache_path), "to recompute.\n")
} else {
  # The SAME explain set for every engine, so nothing differs but the estimator.
  test_df <- as.data.frame(md$test[, fc])
  set.seed(9); eidx <- sample(nrow(test_df), min(params$explain_n, nrow(test_df)))
  X <- test_df[eidx, , drop = FALSE]

  # TreeSHAP arm: the surrogate is configured off the saved model, never assumed - ranger's
  # defaults differ by tree type, so guessing would explain a different model.
  y <- as.integer(md$train$eif3d_target == md$pos_label)
  set.seed(9)
  surr <- ranger(x = md$train[, fc], y = y, num.trees = rf$num.trees, mtry = rf$mtry,
                 min.node.size = rf$min.node.size, importance = "none")
  stopifnot("surrogate does not match the saved model structurally" =
              surr$treetype == "Regression" && surr$mtry == rf$mtry &&
              surr$min.node.size == rf$min.node.size && surr$num.trees == rf$num.trees)
  uni <- ranger.unify(surr, as.data.frame(md$train[, fc]))
  tree_S <- as.matrix(treeshap(uni, X, interactions = FALSE, verbose = FALSE)$shaps)[, fc]

  # Kernel arms.
  pf <- function(m, newdata) predict(m, data = newdata)$predictions[, md$pos_label]
  ks_list <- pmap(arms, function(bg_n, bg_seed, label) {
    set.seed(bg_seed)
    bg <- as.data.frame(md$train[sample(nrow(md$train), bg_n), fc])
    t0 <- Sys.time()
    k  <- kernelshap(rf, X = X, bg_X = bg, pred_fun = pf, verbose = FALSE)
    stopifnot("kernelshap did not converge for every gene" = all(k$converged))
    list(S = as.matrix(k$S)[, fc],
         mins = as.numeric(difftime(Sys.time(), t0, units = "mins")))
  })
  names(ks_list) <- arms$label

  M <- list(tree = tree_S, ks = ks_list, X = X, eidx = eidx, fc = fc, arms = arms)
  saveRDS(M, cache_path)
}

ref_label <- M$arms$label[1]
kern_S    <- M$ks[[ref_label]]$S
pos_mask  <- md$test$eif3d_target[M$eidx] == md$pos_label
```

```{r floor}
# Draw noise, measured from the 20 single-seed runs already on disk. Everything in this
# notebook is judged against this, because a difference smaller than the draw is not a finding.
base <- paste0("shap_feature_", suffix)
seed_files <- c(file.path(out_dir, paste0(base, ".csv")),
                list.files(out_dir, pattern = paste0("^", base, "_negseed[0-9]+\\.csv$"),
                           full.names = TRUE))
have_floor <- length(seed_files) >= 5
if (have_floor) {
  floor_tbl <- map_dfr(seed_files, ~ read_csv(.x, show_col_types = FALSE) %>%
                         mutate(share = 100 * mean_abs_shap / sum(mean_abs_shap))) %>%
    group_by(feature) %>%
    summarise(draw_sd = sd(share), draw_med = median(share), .groups = "drop") %>%
    # 2.8 sd is the difference two independent single-run experiments must clear, since each
    # carries its own draw noise (~sqrt(2) x sd) before anything biological is compared.
    mutate(detect = 2.8 * draw_sd)
} else {
  floor_tbl <- tibble(feature = fc, draw_sd = NA_real_, draw_med = NA_real_,
                      detect = NA_real_)
}
n_draws <- length(seed_files)
```

```{r tables}
share  <- function(S) { v <- colMeans(abs(S)); 100 * v / sum(v) }
signed <- function(S) { v <- colMeans(S[pos_mask, , drop = FALSE]); 100 * v / sum(abs(v)) }

cmp <- tibble(feature = fc, tree = share(M$tree), kernel = share(kern_S)) %>%
  mutate(delta = kernel - tree,
         tree_rank = rank(-tree), kernel_rank = rank(-kernel),
         rank_shift = kernel_rank - tree_rank,
         tree_signed = signed(M$tree), kernel_signed = signed(kern_S)) %>%
  left_join(floor_tbl, by = "feature") %>%
  left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
  mutate(delta_in_sd = abs(delta) / draw_sd,
         beyond_draw = !is.na(detect) & abs(delta) > detect,
         # A feature with a 0.05% share and a tiny draw sd clears the threshold on a move of
         # no consequence. Detectable is not the same as important, so the two are tracked
         # separately and the headline reports both.
         consequential = beyond_draw & tree_rank <= 20)

bg_wide <- imap_dfr(M$ks, ~ tibble(feature = fc, label = .y, share = share(.x$S))) %>%
  bind_rows(tibble(feature = fc, label = "TreeSHAP (surrogate)", share = share(M$tree)))
```

# 1. Headline

```{r headline}
n_beyond <- sum(cmp$beyond_draw, na.rm = TRUE)
worst <- cmp %>% arrange(desc(abs(delta))) %>% slice(1)
tibble(
  quantity = c("features compared", "genes explained", "negative draws behind the floor",
               "median share difference between engines (pp)",
               "largest share difference between engines (pp)",
               "features moving MORE than their own draw noise",
               "features clearing the 2.8 x sd detection threshold",
               "  ...of which are in the top 20 by share",
               "rank correlation between engines (Spearman)",
               "per-gene, per-feature value correlation"),
  value = c(as.character(nrow(cmp)), as.character(nrow(M$X)), as.character(n_draws),
            sprintf("%.3f", median(abs(cmp$delta))),
            sprintf("%.3f  (%s)", abs(worst$delta), worst$feature),
            sprintf("%d of %d", sum(cmp$delta_in_sd > 1, na.rm = TRUE), nrow(cmp)),
            sprintf("%d of %d", n_beyond, nrow(cmp)),
            sprintf("%d  (%s)", sum(cmp$consequential, na.rm = TRUE),
                    paste(cmp$feature[which(cmp$consequential)], collapse = ", ")),
            sprintf("%.4f", cor(cmp$tree, cmp$kernel, method = "spearman")),
            sprintf("%.4f", cor(as.vector(M$tree), as.vector(kern_S))))
) %>% kable(col.names = c("", ""))
```

**In words:** the two engines rank features almost identically and agree closely on the typical
feature. `r n_beyond` features differ by more than the amount two independent experiments would
have to differ by before you would call it a finding - but most of those have a share near
zero, where a tiny absolute move clears a tiny threshold. **Detectable is not the same as
important.** Restricted to the 20 features that actually carry the model, the disagreement is
`r sum(cmp$consequential, na.rm = TRUE)` feature, and it is the strongest feature in the model.

# 2. Do the engines agree? (share per feature)

Each point is one feature. If the engines were interchangeable, every point would sit on the
diagonal. The grey band is **draw noise**: the amount a feature moves when nothing changes but
which non-target genes were sampled. Points inside the band are indistinguishable; points
outside it are real differences between the engines.

```{r fig_scatter, fig.width = 8.5, fig.height = 7.5}
# Draw noise is a PER-FEATURE quantity, so it is drawn as an interval on each feature rather
# than a band across the plot: a ribbon would imply a smooth function of share, which it is
# not. The bar spans where Kernel SHAP would land if it differed from TreeSHAP by one draw's
# worth of noise. A bar that does not reach the diagonal is a real disagreement.
lim <- range(c(cmp$tree, cmp$kernel, cmp$tree - cmp$detect, cmp$tree + cmp$detect), na.rm = TRUE)
p1 <- ggplot(cmp, aes(tree, kernel)) +
  {if (have_floor) geom_linerange(aes(ymin = tree - detect, ymax = tree + detect),
                                  colour = "grey78", linewidth = 1.1)} +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = "dashed") +
  geom_point(aes(colour = case_when(consequential ~ "beyond noise, and large",
                                    beyond_draw   ~ "beyond noise, but tiny",
                                    TRUE          ~ "within draw noise")),
             size = 2.3, alpha = 0.9) +
  geom_text_repel(data = cmp %>% filter(consequential | tree > 3),
                  aes(label = feature), size = 2.6, max.overlaps = Inf,
                  segment.colour = "grey60", segment.size = 0.2, min.segment.length = 0) +
  scale_colour_manual(values = c("beyond noise, and large" = "#D55E00",
                                 "beyond noise, but tiny"  = "#CC79A7",
                                 "within draw noise"       = "grey45"), name = NULL) +
  coord_equal(xlim = lim, ylim = lim) +
  labs(x = "share of importance, TreeSHAP on the surrogate (%)",
       y = "share of importance, Kernel SHAP on the classifier (%)",
       title = "One consequential disagreement, and it is the model's strongest feature",
       subtitle = sprintf("n = %d features, %d genes explained. Grey bar = how far one draw's worth of noise could move a feature (+/- 2.8 sd over %d draws).\nWhere the bar does not cross the dashed line, the engines genuinely differ.",
                          nrow(cmp), nrow(M$X), n_draws)) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", plot.subtitle = element_text(size = 8.5))
p1
```

# 3. How big is the disagreement, compared with noise we already tolerate?

The same numbers, read as a ratio. A value of 1 means the engine swap moved a feature exactly
as much as re-drawing the negative genes would. Bars past the dashed line are differences that
a single-run experiment could not explain away as sampling.

```{r fig_ratio, fig.width = 8, fig.height = 6.5, eval = have_floor}
p2 <- cmp %>% filter(!is.na(delta_in_sd)) %>%
  slice_max(delta_in_sd, n = 25) %>%
  mutate(feature = fct_reorder(feature, delta_in_sd),
         dir = if_else(delta > 0, "Kernel SHAP higher", "TreeSHAP higher")) %>%
  ggplot(aes(delta_in_sd, feature, fill = dir)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = 2.8, linetype = "dotted", colour = "#D55E00") +
  annotate("text", x = 1, y = 0.6, label = "one draw's worth", size = 2.8,
           hjust = -0.08, colour = "grey30") +
  annotate("text", x = 2.8, y = 0.6, label = "detection threshold", size = 2.8,
           hjust = -0.08, colour = "#D55E00") +
  scale_fill_manual(values = c("Kernel SHAP higher" = "#D55E00",
                               "TreeSHAP higher" = "#0072B2"), name = NULL) +
  labs(x = "engine difference, in units of that feature's own draw-to-draw sd",
       y = NULL,
       title = "Most of the disagreement is below the noise we already accept",
       subtitle = sprintf("25 features with the largest ratio, of %d. Draw noise from %d negative draws.",
                          nrow(cmp), n_draws)) +
  theme_classic(base_size = 10) + theme(legend.position = "top")
p2
```

# 4. Is it the estimator, or is it Kernel SHAP's background?

This is the check that decides whether the difference is a **bias** (the engines genuinely
measure different things) or **noise** (Kernel SHAP is just unsteady). Each row is a feature;
each point is one way of computing its share. If TreeSHAP sits apart from a tight cluster of
Kernel variants, the gap is the estimator. If the Kernel points are scattered as widely as the
gap, the background is the problem.

```{r spread_pre, include = FALSE}
spread <- bg_wide %>% group_by(feature) %>%
  summarise(kernel_spread = max(share[label != "TreeSHAP (surrogate)"]) -
                            min(share[label != "TreeSHAP (surrogate)"]),
            tree_to_kernel = abs(share[label == "TreeSHAP (surrogate)"] -
                                 mean(share[label != "TreeSHAP (surrogate)"])),
            .groups = "drop")
```

```{r fig_background, fig.width = 9, fig.height = 7}
top_feats  <- cmp %>% slice_max(tree, n = 15) %>% pull(feature)
n_est_wins <- sum(spread$tree_to_kernel > spread$kernel_spread)
lvls <- c("TreeSHAP (surrogate)", M$arms$label)
p3 <- bg_wide %>% filter(feature %in% top_feats) %>%
  mutate(feature = factor(feature, levels = rev(cmp %>% filter(feature %in% top_feats) %>%
                                                  arrange(tree) %>% pull(feature))),
         label = factor(label, levels = lvls)) %>%
  ggplot(aes(share, feature)) +
  geom_line(aes(group = feature), colour = "grey82", linewidth = 0.7) +
  geom_point(aes(colour = label, shape = label), size = 3, alpha = 0.9) +
  # #E69F00 and #56B4E9 are reserved for eIF3d and eIF3e project-wide, so the method palette
  # avoids both: nothing in this figure should read as one of the two factors.
  scale_colour_manual(values = setNames(c("#0072B2","#D55E00","#009E73","#CC79A7")[seq_along(lvls)],
                                        lvls), name = NULL) +
  scale_shape_manual(values = c(18, 16, 17, 15)[seq_along(lvls)], name = NULL) +
  labs(x = "share of importance (%)", y = NULL,
       title = "For the strongest features the gap is the estimator; lower down it is the background",
       subtitle = sprintf("15 largest features, %d genes explained. The Kernel runs differ only in background size and which\ngenes were drawn. The engine gap exceeds the background spread for %d of %d features.",
                          nrow(M$X), n_est_wins, nrow(spread))) +
  theme_classic(base_size = 10) +
  theme(legend.position = "top", plot.subtitle = element_text(size = 8.5))
p3
```

```{r bg_numbers}
tibble(quantity = c("largest spread ACROSS Kernel backgrounds (pp)",
                    "largest TreeSHAP-to-Kernel gap (pp)",
                    "features where the engine gap exceeds the background spread"),
       value = c(sprintf("%.3f", max(spread$kernel_spread)),
                 sprintf("%.3f", max(spread$tree_to_kernel)),
                 sprintf("%d of %d", sum(spread$tree_to_kernel > spread$kernel_spread),
                         nrow(spread)))) %>%
  kable(col.names = c("", ""))
```

**Reading it:** for `cds_gc`, `log2_cds_length` and `log2_utr5_length` the three Kernel runs
cluster tightly while TreeSHAP sits apart - a genuine difference between estimators. Further
down the list the Kernel runs scatter as widely as the gap itself, and there you are looking at
the background, not the method. A background of 100 genes is too small - two different 100-gene draws move
results almost as much as switching engines does. Going to 300 tightens them. But the Kernel
estimates stay on the *same side* of TreeSHAP whichever background is used, which is what makes
the gap a bias rather than noise. **If you adopt Kernel SHAP, 100 background genes is not
enough**; find where it converges first.

# 5. Per-gene agreement

Everything above is averaged over genes. The force plots and per-gene decompositions are not,
so the engines have to agree gene by gene as well.

```{r fig_pergene, fig.width = 10, fig.height = 4.6}
gene_cor <- map_dbl(seq_len(nrow(M$tree)), ~ cor(M$tree[.x, ], kern_S[.x, ]))
pa <- tibble(tree = as.vector(M$tree), kernel = as.vector(kern_S)) %>%
  ggplot(aes(tree, kernel)) +
  geom_hex(bins = 60) +
  geom_abline(slope = 1, intercept = 0, colour = "#D55E00", linetype = "dashed") +
  scale_fill_gradient(low = "grey88", high = "#0072B2", trans = "log10", name = "count") +
  labs(x = "SHAP value, TreeSHAP", y = "SHAP value, Kernel SHAP",
       title = "Every gene x feature value",
       subtitle = sprintf("n = %s values, r = %.3f",
                          format(length(M$tree), big.mark = ","),
                          cor(as.vector(M$tree), as.vector(kern_S)))) +
  theme_classic(base_size = 10)

pb <- tibble(r = gene_cor) %>%
  ggplot(aes(r)) +
  geom_histogram(bins = 30, fill = "#0072B2", alpha = 0.8) +
  geom_vline(xintercept = median(gene_cor), colour = "#D55E00", linewidth = 0.8) +
  labs(x = "correlation between the two engines, within one gene", y = "genes",
       title = "Agreement gene by gene",
       subtitle = sprintf("n = %d genes; median r = %.3f, worst gene r = %.3f",
                          length(gene_cor), median(gene_cor), min(gene_cor))) +
  theme_classic(base_size = 10)
cowplot::plot_grid(pa, pb, nrow = 1, rel_widths = c(1.15, 1))
```

# 6. Direction

Share says how much a feature matters. **Signed share** says which way it pushes - toward
being a target, or away. Directional claims are the ones that get written into a paper, so they
matter more than the magnitudes.

```{r fig_signed, fig.width = 8, fig.height = 7}
top20 <- cmp %>% slice_max(abs(tree_signed), n = 20) %>% pull(feature)
agree_all <- mean(sign(cmp$tree_signed) == sign(cmp$kernel_signed))
agree_top <- mean(sign(cmp$tree_signed[cmp$feature %in% top20]) ==
                  sign(cmp$kernel_signed[cmp$feature %in% top20]))
# Signed share over target-class genes is positive for nearly every feature, so symmetric
# limits at the maximum would leave three quadrants empty. Zoom to the data, keeping the aspect
# equal so distance from the diagonal still reads correctly.
srng <- range(c(cmp$tree_signed, cmp$kernel_signed))
srng <- srng + c(-1, 1) * 0.06 * diff(srng)
slim <- max(abs(srng))

p5 <- cmp %>%
  mutate(agrees = sign(tree_signed) == sign(kernel_signed)) %>%
  ggplot(aes(tree_signed, kernel_signed)) +
  annotate("rect", xmin = -slim, xmax = 0, ymin = 0, ymax = slim, fill = "#D55E00", alpha = 0.07) +
  annotate("rect", xmin = 0, xmax = slim, ymin = -slim, ymax = 0, fill = "#D55E00", alpha = 0.07) +
  geom_hline(yintercept = 0, colour = "grey60") + geom_vline(xintercept = 0, colour = "grey60") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(aes(colour = agrees), size = 2.2, alpha = 0.85) +
  geom_text_repel(data = . %>% filter(feature %in% top20), aes(label = feature),
                  size = 2.5, max.overlaps = Inf, segment.size = 0.2,
                  segment.colour = "grey60", min.segment.length = 0) +
  scale_colour_manual(values = c(`TRUE` = "grey35", `FALSE` = "#D55E00"),
                      labels = c(`TRUE` = "same direction", `FALSE` = "opposite direction"),
                      name = NULL) +
  coord_equal(xlim = srng, ylim = srng) +
  labs(x = "signed share, TreeSHAP (%)", y = "signed share, Kernel SHAP (%)",
       title = "Direction agrees where direction is readable",
       subtitle = sprintf("n = %d features. Shaded quadrants = the engines disagree on direction.\nAgreement: %.0f%% of all features, %.0f%% of the 20 with the largest signed contribution.",
                          nrow(cmp), 100 * agree_all, 100 * agree_top)) +
  theme_classic(base_size = 10) + theme(legend.position = "top")
p5
```

**Reading it:** disagreements cluster at the origin, where the signed value is near zero and
its sign is arbitrary anyway. Among the features that carry real directional signal the engines
agree `r sprintf("%.0f%%", 100 * agree_top)` of the time. **Directional conclusions are safe
under either engine; magnitude rankings are not.**

# 7. Verdict

```{r verdict}
tibble(
  question = c("Do the engines rank features the same way?",
               "Do they agree on the biggest features?",
               "Is the difference bias or noise?",
               "Does Kernel SHAP remove a discretionary choice?",
               "Do directional conclusions change?",
               "What does it cost?"),
  answer = c(
    sprintf("Nearly. Spearman %.3f, median rank shift %.0f.",
            cor(cmp$tree, cmp$kernel, method = "spearman"),
            median(abs(cmp$rank_shift))),
    sprintf("No. %d features move beyond the detection threshold, including %s (%.2f -> %.2f).",
            n_beyond, worst$feature, worst$tree, worst$kernel),
    "Bias. Every Kernel background lands on the same side of TreeSHAP.",
    "No - it trades the stand-in model for the background set, which at 100 genes is as unsteady as the gap itself.",
    sprintf("No. %.0f%% agreement among the 20 features with real signed signal.",
            100 * agree_top),
    sprintf("%.1f min per %d genes at the reference background, against seconds for TreeSHAP.",
            M$ks[[ref_label]]$mins, nrow(M$X))
  )) %>% kable()
```

## What to do

**Do not mix engines within a comparison.** A systematic shift of
`r sprintf("%.1f", abs(worst$delta))` percentage points on `r worst$feature` would read as a
biological difference between two datasets when it is only a difference in how the numbers were
computed. Whichever engine is chosen, every experiment in a comparison must use it.

**Neither engine fixes instability.** The draw moves a typical feature by far more than the
engine choice does. Switching engines is a question about *what the number means*, not about
making it steadier - that requires changing how the negative class is built, not how the SHAP
values are computed.

**Recommendation: keep TreeSHAP as the default** for its speed, determinism and interaction
values, and use Kernel SHAP as a standing cross-check - which the `shap_engine` parameter in
the main SHAP notebook now makes a parameter change rather than a project. Revisit if a future
model cannot be unified by `treeshap` at all, in which case Kernel SHAP is the only option and
the background set needs its own convergence check first.

```{r export}
write_csv(cmp, file.path(out_dir, paste0("shap_engine_comparison_", suffix, ".csv")))
write_csv(bg_wide, file.path(out_dir, paste0("shap_engine_background_arms_", suffix, ".csv")))
cat("Wrote 2 comparison files with suffix:", suffix, "\n")
```
