---
title: "`r paste0('RF regression: ', params$outcome_col, ' ~ features (eIF', params$gene, ', ', params$condition, ', ', params$timepoint, ', y=', params$target_transform, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  readout: "TE"
  timepoint: "1hr"
  lfc: "0.5"
  include_blocks: "intrinsic,external"
  clip_mode: "aggregate"
  matrix_rds: "feature_matrix_tco.rds"
  feature_set_label: "tcoreg"
  variant_set: "corrected"
  initiation_score: "both"
  include_deprecated: FALSE
  exclude_families: ""
  exclude_features: ""
  shuffle_families: ""
  outcome_csv: ""
  outcome_col: "te_lfc"
  outcome_padj_col: "te_padj"
  direction_check_csv: ""
  negate_outcome: FALSE
  cell_line: "MDA-MB-231"
  outcome_kind: "te"
  outcome_desc: ""
  kd_symbol: ""
  min_spot_sep: 0.3
  heldout_sep_check: TRUE
  perm_n: 1
  xcell_ref_csv: ""
  neg_lfc: "0.5"
  restrict_to_genes_csv: ""
  parity_ref_suffix: ""
  target_transform: "raw"
  winsorize_y: 0
  split_seed: 9
  forest_seed: 9
  num_trees: 1000
  mtry: 0
  min_node_size: 5
  splitrule: "variance"
  sample_fraction: 0
  weight_mode: "none"
  restrict_to_classifier_genes: FALSE
  gate_feature_parity: TRUE
  clf_suffix_override: ""
  precision_at_k: "50,100,200,500"
  auc_bridge: TRUE
  min_test_pos: 20
  permutation_check: TRUE
  run_cv: TRUE
  save_plots: TRUE
  save_model_data: TRUE
  mutate_target: ""
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
# `params` is a base function outside knitr, so it must be cleared before being defined as a
# list. The defaults below are not re-typed: notebook 07 in the frozen pipeline carries a
# second, disagreeing interactive param set, so knitting it gives a 56-feature model while
# running it chunk-by-chunk gives a 51-feature one. Reading this file's own YAML back is the
# only interactive path that cannot drift from the knit path.
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10_rf_regression.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r direction_facts, include=FALSE}
# The direction of this contrast is DERIVED here, before any prose, so that the sentences in the
# overview cannot state it wrongly. An earlier draft of this notebook claimed the classifier's
# positive set was "genes whose TE went up on knockdown", which is backwards: te_lfc is
# log2(sictrl / si3d), so a positive value means the control is higher and the knockdown LOWER.
# Every directional phrase below is interpolated from these values, and the assertion at the end
# ties them to the `direction` param, so prose and data cannot disagree.
suppressMessages({ library(dplyr); library(readr); library(here) })
.dir_outcome <- if (nzchar(params$outcome_csv)) params$outcome_csv else
  paste0("translation_categories_si", params$gene, "_vs_sictrl_",
         params$condition, "_", params$timepoint, ".csv")
.dir_te  <- read_csv(here("output", .dir_outcome), show_col_types = FALSE) %>%
  mutate(g = sub("[.].*", "", gene_id))
# The positive gene set is needed twice - here to derive the directional wording, and later for
# the biological spot-check and the AUC bridge - so it is defined ONCE as a function and called
# in both places. It was duplicated, and the duplicate is what broke the regressed timepoints:
# this chunk runs before the prose, so it failed on a missing CSV ahead of the branch that
# handles it.
#
# For most timepoints the set is a saved CSV. For the regressed timepoints (1and4hr,
# alltimepoints) no TE-readout CSV exists, so the set is defined inline from the outcome file by
# the same rule those CSVs were built with, mirroring notebook 07's use_full_de_table branch
# rather than inventing a second definition.
positive_set <- function() {
  if (params$readout == "TE" && params$timepoint %in% c("1and4hr", "alltimepoints")) {
    g <- .dir_te %>%
      filter(.data[[params$outcome_col]] > as.numeric(params$lfc),
             .data[[params$outcome_padj_col]] < 0.05) %>% pull(g)
    stopifnot("the inline positive set is empty or implausibly small - check the filter" =
                length(g) >= 50)
    attr(g, "source") <- paste0(.dir_outcome, " (inline: ", params$outcome_col, " > ",
                                params$lfc, " & padj < 0.05)")
    g
  } else {
    csv <- paste0(params$condition, "_", params$gene, "_", params$direction, "_",
                  params$readout, "_", params$timepoint, "_lfc", params$lfc, ".csv")
    pth <- here("output", "genesets", csv)
    stopifnot("positive geneset CSV not found" = file.exists(pth))
    g <- read_csv(pth, show_col_types = FALSE)$ensembl_gene
    attr(g, "source") <- csv
    g
  }
}
.dir_pos <- positive_set()
.dir_p       <- .dir_te %>% filter(g %in% .dir_pos)
.dir_median  <- median(.dir_p[[params$outcome_col]], na.rm = TRUE)
.dir_n       <- nrow(.dir_p)
# Positive outcome = control above knockdown = the knockdown LOWERS it.
.dir_verb    <- if (.dir_median > 0) "falls" else "rises"      # singular: "TE falls"
.dir_verb_pl <- if (.dir_median > 0) "fall"  else "rise"       # plural:  "those genes fall"
.dir_effect  <- if (.dir_median > 0) "promotes" else "inhibits"
# MDA-MB-231 files are written as log2(sictrl / siX) already. MCF7-SIX1 results are computed the
# other way round and are negated into this orientation once, by 10m_mcf7six1_outcome_prep.Rmd,
# so the stored column is promotes-positive in both cell lines but its provenance differs.
# Calling an MCF7-SIX1 outcome "log2(sictrl / si3d)" would name a contrast nobody computed.
.dir_ratio   <- if (params$outcome_kind != "te" || nzchar(params$outcome_desc)) params$outcome_desc else
  if (params$cell_line == "MDA-MB-231")
  paste0("log2(sictrl / si", params$gene, ")") else
  paste0("log2(si", params$gene, " / sictrl) negated, which is the same orientation")
.dir_gene    <- if (grepl("^3", params$gene)) paste0("eIF", params$gene) else  # "3d" -> "eIF3d"
                toupper(params$gene)                                         # "dap5" -> "DAP5"
# The count that belongs in the prose is the number the CLASSIFIER actually modelled, which is the
# gene set intersected with the feature matrix, not the size of the gene set CSV. Read it from 02_'s
# saved metrics when available rather than restating a number from a different denominator.
.dir_clf_suffix <- if (nzchar(params$clf_suffix_override)) params$clf_suffix_override else {
  source(here("code", "translational_control_overview", "02_sweep_presets.R"), local = TRUE)
  sweep_suffix(list(gene = params$gene, direction = params$direction, readout = params$readout,
                    condition = params$condition, timepoint = params$timepoint,
                    lfc = params$lfc, feature_set_label = "tco"), 9L)
}
.dir_clf_perf <- file.path(here("output", "translational_control_overview"),
                           paste0("rf_performance_", .dir_clf_suffix, ".csv"))
.dir_n_clf <- if (file.exists(.dir_clf_perf))
  read_csv(.dir_clf_perf, show_col_types = FALSE) %>% filter(metric == "n_pos") %>% pull(value) else
  NA_real_
.dir_n_phrase <- if (is.na(.dir_n_clf)) paste0("the ", .dir_n, " genes in that set") else
  paste0("the ", format(.dir_n_clf, big.mark = ","), " genes")
# Binding outcomes have no knockdown direction; a stress-ratio hit set can sit near zero (the
# thapsigargin hits barely rise over DMSO), which is a finding, not a sign error.
stopifnot(
  "the positive set's median outcome sign disagrees with the `direction` param" =
    params$outcome_kind != "te" ||
    (params$direction == "promotes" && .dir_median > 0) ||
    (params$direction == "inhibits" && .dir_median < 0)
)
```

# Overview

```{r binding_banner, echo = FALSE, results = "asis", eval = params$outcome_kind != "te"}
# Non-knockdown outcomes (binding enrichment, absolute TE within one condition): the text below
# is written for a knockdown TE contrast, so the banner says what the outcome actually is.
cat(paste0(
  "> **This run does not model a knockdown TE change.** Outcome kind: `", params$outcome_kind,
  "`. The outcome is ", .dir_ratio, " (`", params$outcome_csv, "`, ", params$cell_line, "). ",
  "Wherever the text below says knockdown or \"promotes\", read the outcome above; the ",
  "knockdown direction rulers are skipped. Caveats for each outcome are in its prep script and ",
  "notes file.\n\n"))
```

This notebook predicts each gene's continuous TE log2 fold change from its mRNA sequence and
structure features, using a random forest regression.

It is the regression counterpart to `02_rf_model.Rmd`, which answers a related but different
question. That notebook is a classifier: it takes `r .dir_n_phrase` whose TE **`r .dir_verb`**
when `r .dir_gene` is knocked down - the genes `r .dir_gene` `r .dir_effect` - draws an equal
number from a pool selected for showing no effect, and learns to tell the two apart. This notebook
drops the classes and the pool, and models the TE change itself across every gene in the feature
matrix.

On the sign: the outcome is ``r params$outcome_col``, which is `r .dir_ratio`, so a positive value
means TE was higher in the control than after knockdown. The gene set behind that classifier is
defined by ``r params$outcome_col` > `r params$lfc``, with median
`r sprintf("%+.3f", .dir_median)` across its `r format(.dir_n, big.mark = ",")` genes, so those genes `r .dir_verb_pl` on
knockdown. Positive therefore means `r .dir_gene` `r .dir_effect` translation throughout this
notebook.

Every directional phrase in that paragraph is computed from the outcome file at knit time, not
typed, and a check ties the sign to the `direction` parameter. An earlier draft of this notebook
had the direction backwards in prose while every number in it was correct, which is the failure
this arrangement removes.

**Decisions**

- Model every gene with a measured TE value (10,172) rather than a balanced pair of tails
  (2,872).
- Use the continuous TE log2 fold change as the outcome, not a thresholded label.
- Replace `neg_seed` with `split_seed` as the replication axis. There is no negative draw left
  to seed, so the thing that varies between repeat runs is now which genes are held out. The
  sweep in `10b_`/`10c_` uses that.
- Keep the feature selection identical to `02_`'s, so that when `12_` compares the two models'
  SHAP values the comparison is about the model rather than about the feature sets.
- Write only `rfreg_*` filenames, never `rf_*` or `shap_*`. `assert_no_collision()` in
  `10_regression_presets.R` enforces this so a run here cannot overwrite the frozen
  classification results.

One thing to note when reading the results: because there is no negative pool, no number here
is a property of that pool. That is worth stating because the pool is not a neutral sample of
the transcriptome. Its genes are GC-richer than the rest of the matrix by 2.45 percentage points
(Wilcoxon p = 2.7e-17), so any GC difference measured against it is partly a property of how the
pool was selected.

## Which TE column is the outcome

Three columns in this project have similar names and different meanings, so it is worth being
explicit about which one is being modelled.

| column | where it lives | role here |
|---|---|---|
| `te_lfc` | the outcome CSV for this contrast | **the outcome.** What the forest predicts. |
| `rna_lfc` | the same CSV | checks the file's sign convention. Not modelled. |
| `te_lfc` | the feature matrix | a different contrast. Not used. |

**The outcome** is `te_lfc` from `translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv`.

**`rna_lfc` is used once, to confirm the sign convention.** The filenames read
`si3d_vs_sictrl`, but the DESeq2 coefficient is `Condition_sictrl_vs_si3d`, so the columns are
log2(sictrl / si3d) and positive means eIF3d promotes translation. Since the filename and the
coefficient disagree, the orientation is checked against the data instead: the siRNA destroys
EIF3D's own mRNA, so EIF3D's `rna_lfc` must be strongly positive, and it is (+2.90, about 87%
knockdown).

That check cannot be done on `te_lfc`, because EIF3D's own `te_lfc` is negative (-1.26). EIF3D is
TE-buffered in this contrast: its ribosome footprints rise (+1.45) along with its mRNA (+2.90),
so the ratio moves down. Requiring EIF3D's `te_lfc` to be positive would therefore fail on
correct data.

**The feature matrix's `te_lfc` is a different contrast** and is not the outcome. Audit finding
F7 established that it is the siCTRL hypoxia-versus-normoxia comparison, not a knockdown
comparison: it separates the eIF3d target set by a median of only -0.013 and correlates 0.002
with the outcome used here. `indiv_te_hypoxia_1hr` is also not a fold change, but an absolute TE
level (it correlates 0.933 with the normoxia equivalent).

**Decisions**

- Drop `te_lfc`, `te_padj` and `te_lfc_bin` from the feature matrix before joining the outcome,
  so there is never a second column of that name next to the outcome.
- Check the sign convention on `rna_lfc`, not on `te_lfc`.
- Check the outcome by value against both matrix columns, not by name. Joining either of them
  as the outcome would model the wrong quantity while passing every range and finiteness check,
  so a name-based check would not catch it. The `target_provenance` chunk does this.
- Exclude `rna_lfc`, `ribo_lfc` and `category` from the features along with the outcome columns.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
```

```{r derived_values}
source(here("code", "translational_control_overview", "02_sweep_presets.R"))   # sweep_suffix()
source(here("code", "translational_control_overview", "10_regression_presets.R"))

SPLIT_SEED  <- as.integer(params$split_seed)
FOREST_SEED <- as.integer(params$forest_seed)

stopifnot(
  "target_transform must be raw, rank_ecdf or robust_z" =
    params$target_transform %in% c("raw", "rank_ecdf", "robust_z"),
  "splitrule must be variance, extratrees or maxstat" =
    params$splitrule %in% c("variance", "extratrees", "maxstat"),
  "winsorize_y must be >= 0" = params$winsorize_y >= 0,
  # padj is a monotone function of |lfc|/lfcSE, so weighting by it upweights large |y|. That is
  # weighting by the outcome - a soft target leak that inflates R2, Spearman and every AUC.
  # Blocked rather than documented, because it is the tempting option: padj is the precision
  # column that actually exists in the file, and lfcSE is not saved for this contrast.
  "weight_mode='padj' weights by a function of |y| - that is target leakage" =
    params$weight_mode %in% c("none", "basemean")
)

include_blocks <- str_trim(str_split(params$include_blocks, ",")[[1]])
stopifnot("include_blocks must name real blocks" =
            all(include_blocks %in% c("intrinsic", "external", "eif4e", "technical")))

# Cell line. The only thing that can go badly wrong with a second cell line is the sign, and the
# way it goes wrong is quietly: the MDA-MB-231 outcome files carry rna_lfc and check their own
# direction against it, while the MCF7-SIX1 tables have no RNA column at all, so that check would
# simply print "skipped" and the run would proceed unverified. A non-231 run is therefore
# REQUIRED to name an independent ruler instead of being allowed to skip.
# HeLa = Herrmannova et al. 2024 (eLife), prepared by 10h_herrmannova_outcome_prep.Rmd.
# HEK293T = subunit-seq cap binding (outcome_kind = "binding"), prepared by
# 10s_subunit_seq_outcome_prep.Rmd. There is no knockdown, so no direction ruler applies; the
# prep notebook's replication checks against both papers take its place.
# te_level = absolute TE (log2 ribo/RNA) within one condition, e.g. Teleman empty vector or
# 4E-BP1(4A) (10t_teleman_outcome_prep.R): also no knockdown, so the same rulers are skipped.
stopifnot("outcome_kind must be te, binding or te_level" =
            params$outcome_kind %in% c("te", "binding", "te_level"),
          "a non-knockdown outcome needs outcome_desc saying what it is" =
            params$outcome_kind == "te" || nzchar(params$outcome_desc),
          "cell_line must be MDA-MB-231, MCF7-SIX1, HeLa or HEK293T" =
            params$cell_line %in% c("MDA-MB-231", "MCF7-SIX1", "HeLa", "HEK293T"),
          "a binding outcome is from HEK293T in this project" =
            params$outcome_kind != "binding" || params$cell_line == "HEK293T",
          "a non-knockdown run needs an explicit outcome_csv" =
            params$outcome_kind == "te" || nzchar(params$outcome_csv),
          # A factor with no MDA-MB-231 counterpart (DAP5) has no matched contrast to serve as a
          # ruler; it must instead name its depleted gene, whose own rna_lfc is then asserted.
          "a non-MDA-MB-231 TE run must name xcell_ref_csv (the matched 231 contrast) or kd_symbol (the depleted gene, whose rna_lfc is the ruler)" =
            params$cell_line == "MDA-MB-231" || params$outcome_kind != "te" ||
            nzchar(params$xcell_ref_csv) || nzchar(params$kd_symbol))

# Where this run's negative controls live. Derived from the params rather than passed in, so one
# cell line cannot be scored against the other's controls by forgetting an argument.
neg_csv_path <- if (params$cell_line == "MDA-MB-231") {
  here("output", "predictive_modeling",
       if (params$gene == "3e") "negative_control_genes_si3e.csv" else
                                "negative_control_genes.csv")
} else {
  here("output", "genesets",
       paste0(params$condition, "_", params$gene, "_negative_controls_lfc", params$neg_lfc, ".csv"))
}
stopifnot("the negative control CSV for this cell line and knockdown was not found" =
            file.exists(neg_csv_path))

# Built from the same params as 02_'s, so a regression suffix is string-alignable with the
# classifier suffix it will be compared against in 12_.
suffix <- reg_suffix(list(
  gene = params$gene, direction = params$direction, readout = params$readout,
  condition = params$condition, timepoint = params$timepoint, lfc = params$lfc,
  feature_set_label = params$feature_set_label, variant_set = params$variant_set,
  initiation_score = params$initiation_score, clip_mode = params$clip_mode,
  target_transform = params$target_transform, winsorize_y = params$winsorize_y,
  min_node_size = params$min_node_size, mtry = params$mtry, num_trees = params$num_trees,
  splitrule = params$splitrule, sample_fraction = params$sample_fraction,
  weight_mode = params$weight_mode,
  restrict_to_classifier_genes = params$restrict_to_classifier_genes,
  forest_seed = FOREST_SEED), SPLIT_SEED)

clf_suffix <- if (nzchar(params$clf_suffix_override)) params$clf_suffix_override else
  sweep_suffix(list(gene = params$gene, direction = params$direction, readout = params$readout,
                    condition = params$condition, timepoint = params$timepoint,
                    lfc = params$lfc, feature_set_label = "tco"), 9L)

# "" builds the knockdown contrast filename from params. si3e normoxia 1and4hr is the one
# contrast saved under a different stem, so it needs an explicit outcome_csv.
outcome_file <- if (nzchar(params$outcome_csv)) params$outcome_csv else
  paste0("translation_categories_si", params$gene, "_vs_sictrl_",
         params$condition, "_", params$timepoint, ".csv")

out_dir <- here("output", "translational_control_overview")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
MUTATING <- nzchar(params$mutate_target)
ggsave_if <- function(...) if (isTRUE(params$save_plots) && !MUTATING) ggsave(...)

cat("Cell line:     ", params$cell_line, "\n")
cat("Suffix:        ", suffix, "\n")
cat("Classifier ref:", clf_suffix, "\n")
cat("Outcome file:  ", outcome_file, "  column:", params$outcome_col, "\n")
cat("Neg controls:  ", basename(neg_csv_path), "\n")
if (params$cell_line != "MDA-MB-231")
  cat("Direction ruler:", params$xcell_ref_csv, "\n")
cat("Transform:     ", params$target_transform,
    if (params$winsorize_y > 0) paste0(" (winsorized at +/-", params$winsorize_y, ")") else "", "\n")
cat("Seeds:          split", SPLIT_SEED, "| forest", FOREST_SEED, "\n")
if (MUTATING) cat("*** MUTATION RUN:", params$mutate_target, "- nothing will be written ***\n")
```

# Load

Reads the feature matrix, its annotation table, and the outcome file for this contrast.

**Decisions**

- Take features from `feature_matrix_tco.rds` and outcomes from a separate
  `translation_categories_*` file, rather than expecting the outcome to already be in the matrix.
- Require the annotation table to describe exactly the columns the matrix has, so feature
  selection cannot silently reference a column that is not there.


```{r load_data}
features_raw <- readRDS(file.path(out_dir, params$matrix_rds))
annotation   <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
stopifnot("annotation does not match the matrix" =
            setequal(annotation$feature, colnames(features_raw)))
cat("Matrix:", nrow(features_raw), "x", ncol(features_raw), "\n")

# Drop the label columns before the join. Leaving them would put a second column named te_lfc
# next to the outcome and force dplyr to rename one of them, which is how the wrong column
# could end up being modelled without anyone noticing.
features <- features_raw %>% dplyr::select(-any_of(c("te_lfc", "te_padj", "te_lfc_bin")))

outcome_raw <- read_csv(here("output", outcome_file), show_col_types = FALSE)
stopifnot(
  "outcome CSV lacks the named outcome column" = params$outcome_col %in% names(outcome_raw),
  "outcome CSV lacks the named padj column"    = params$outcome_padj_col %in% names(outcome_raw),
  "outcome CSV has no gene_id column"          = "gene_id" %in% names(outcome_raw)
)
cat("Outcome file:", nrow(outcome_raw), "genes |",
    sum(!is.na(outcome_raw[[params$outcome_col]])), "with a non-NA outcome\n")
```

# Target

Joins the outcome onto the feature matrix, then checks that the joined values are the intended
quantity and the intended direction.

**Decisions**

- Join on `gene_id_clean`. The matrix holds one transcript per gene, so this is one-to-one, and
  the code asserts that rather than assuming it.
- Keep every gene with a non-missing outcome. No filtering on significance: `te_padj` is used
  only for reporting and for the Precision@K labels.
- Do not re-centre the outcome even though its median is 0.304 rather than zero. Centring is an
  affine transform, and a variance-splitting forest is unchanged by it.
  `10d_target_scale_analysis.Rmd` measures and plots this.
- Verify the join reproduces the source file exactly rather than trusting `inner_join`.


```{r create_target}
# sub() on an already-unversioned id is a no-op, so this handles both conventions.
outcome <- outcome_raw %>%
  mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>%
  dplyr::select(gene_id_clean,
                y      = all_of(params$outcome_col),
                y_padj = all_of(params$outcome_padj_col))

# MDA-MB-231 is log2(sictrl/si3d) so positive already means "eIF3d promotes". MCF7-SIX1 runs the
# other way and must be negated to match. Untested on MCF7-SIX1 - the direction check below is
# what would catch a mistake.
if (isTRUE(params$negate_outcome)) {
  outcome$y <- -outcome$y
  cat("NOTE: outcome negated (negate_outcome = TRUE)\n")
}

# --- mutation hooks. Deliberate corruptions, each of which must trip a named check below. ---
if (MUTATING) {
  mt <- params$mutate_target
  if (mt == "target_is_matrix_te_lfc") {
    outcome <- features_raw %>% dplyr::select(gene_id_clean, y = te_lfc, y_padj = te_padj)
  } else if (mt == "target_is_indiv_te") {
    outcome <- features_raw %>% dplyr::select(gene_id_clean, y = indiv_te_hypoxia_1hr) %>%
      mutate(y_padj = 0.5)
  } else if (mt %in% c("flip_target_sign", "permute_target")) {
    # Both of these corrupt the outcome for every gene EXCEPT the knockdown gene itself.
    #
    # Without that exception they trip the contrast anchor, because the anchor compares the
    # knockdown gene's own value against the params-implied file and a negated or permuted value
    # no longer matches. Three of the nine mutations were landing on the anchor for that reason,
    # which meant the geneset-separation spot-check was never demonstrated to fire at all - the
    # "a check never shown to FAIL is uninformative" problem, one level up. Holding the knockdown
    # gene fixed lets these two slip past the anchor so that separation is the only check left
    # that can catch them, which is what proves it works.
    .kd <- paste0("EIF", toupper(params$gene))
    .kd_ids <- if ("symbol" %in% names(outcome_raw))
      sub("[.].*", "", outcome_raw$gene_id[outcome_raw$symbol == .kd]) else character(0)
    .spare <- outcome$gene_id_clean %in% .kd_ids
    .keep  <- outcome$y[.spare]
    if (mt == "flip_target_sign") {
      outcome$y <- -outcome$y
    } else {
      set.seed(9); outcome$y <- sample(outcome$y)
    }
    outcome$y[.spare] <- .keep
    cat(sprintf("  (held %s fixed so the contrast anchor still passes)\n", .kd))
  }
  cat("applied mutation:", mt, "\n")
}

model_genes <- features %>%
  inner_join(outcome, by = "gene_id_clean") %>%
  filter(!is.na(y))

# Optional: restrict to the genes the classifier modelled. This is the arm that separates
# "different model type" from "different gene population" when 12_ compares SHAP - without it,
# agreement between the two models cannot be told apart from two effects cancelling.
if (isTRUE(params$restrict_to_classifier_genes)) {
  clf_genes_path <- file.path(out_dir, paste0("rf_model_gene_sets_", clf_suffix, ".csv"))
  stopifnot("restrict_to_classifier_genes needs 02_'s gene sets - knit 02_rf_model.Rmd first" =
              file.exists(clf_genes_path))
  clf_genes <- read_csv(clf_genes_path, show_col_types = FALSE)$gene_id_clean
  n_before  <- nrow(model_genes)
  model_genes <- model_genes %>% filter(gene_id_clean %in% clf_genes)
  cat("Restricted to the classifier's genes:", n_before, "->", nrow(model_genes), "\n")
}

# Restrict to a supplied gene list. This exists to put two cell lines on one heatmap axis: a SHAP
# share is computed over whichever genes the model explained, so columns built on different gene
# populations are not comparable, and 15_ asserts exactly that. MCF7-SIX1 can score 9,335 of the
# matrix's 10,172 genes, so a combined figure needs the 231 runs restricted to the same 9,335.
if (nzchar(params$restrict_to_genes_csv)) {
  rp <- file.path(out_dir, params$restrict_to_genes_csv)
  stopifnot("restrict_to_genes_csv not found" = file.exists(rp))
  keep_g <- read_csv(rp, show_col_types = FALSE)$gene_id_clean
  stopifnot("restrict_to_genes_csv has no gene_id_clean column" = !is.null(keep_g),
            # Same hazard exclude_features guards: the restriction is not part of the output
            # suffix, so without a distinct label a restricted run silently overwrites the
            # full run's model, importance and SHAP.
            "a restricted run needs a distinct feature_set_label or it overwrites the full run" =
              params$feature_set_label != "tcoreg")
  n_before_r <- nrow(model_genes)
  model_genes <- model_genes %>% filter(gene_id_clean %in% keep_g)
  cat(sprintf("Restricted to %s: %d -> %d genes\n", params$restrict_to_genes_csv,
              n_before_r, nrow(model_genes)))
  stopifnot("the gene restriction removed nearly everything - wrong list?" =
              nrow(model_genes) > 5000)
}

cat("Model genes:", nrow(model_genes), "\n")
```

```{r target_provenance}
# Confirm the outcome is the quantity intended.
#
# The matrix's own te_lfc is the siCTRL hypoxia-versus-normoxia contrast (audit finding F7,
# pinned bit-for-bit by 01_'s verify_te_lfc_provenance), and indiv_te_hypoxia_1hr is an absolute
# TE level. Joining either as the outcome would model the wrong quantity while passing every
# range, finiteness and monotonicity check here, so the outcome is compared against both by
# value rather than by name.
prov <- features_raw %>%
  dplyr::select(gene_id_clean, m_te = te_lfc, m_ind = indiv_te_hypoxia_1hr) %>%
  inner_join(model_genes %>% dplyr::select(gene_id_clean, y), by = "gene_id_clean")
k_te  <- complete.cases(prov$y, prov$m_te)
k_ind <- complete.cases(prov$y, prov$m_ind)
dev_te  <- max(abs(prov$y[k_te] - prov$m_te[k_te]))
r_te    <- cor(prov$y[k_te],  prov$m_te[k_te],  method = "spearman")
r_ind   <- cor(prov$y[k_ind], prov$m_ind[k_ind], method = "spearman")
cat(sprintf("vs matrix te_lfc (hypoxia response): max|dev| %.3f, Spearman %+.3f (n=%d)\n",
            dev_te, r_te, sum(k_te)))
cat(sprintf("vs indiv_te_hypoxia_1hr (abs TE level): Spearman %+.3f\n", r_ind))
stopifnot(
  "the joined outcome equals the matrix's own te_lfc, so the wrong contrast was loaded" = dev_te > 0.5,
  "the target tracks the matrix hypoxia-response te_lfc too closely" = abs(r_te)  < 0.5,
  "the target tracks an absolute TE level too closely"               = abs(r_ind) < 0.5
)

# Never transcribe a reference value: re-parse the source and assert the join reproduced it
# exactly, rather than trusting that inner_join did what was intended.
recheck <- model_genes %>% dplyr::select(gene_id_clean, y) %>%
  inner_join(outcome %>% dplyr::select(gene_id_clean, src = y), by = "gene_id_clean")
stopifnot(
  "the target is not bit-identical to its named source file" =
    nrow(recheck) == nrow(model_genes) && max(abs(recheck$y - recheck$src)) == 0,
  # The matrix is one transcript per gene, so a gene-keyed join must not fan out.
  "the target join was not 1:1 - the matrix is no longer one transcript per gene" =
    !any(duplicated(model_genes$gene_id_clean)),
  "non-finite target" = all(is.finite(model_genes$y)),
  "the join added more than the two expected columns" =
    setequal(setdiff(colnames(model_genes), colnames(features)), c("y", "y_padj"))
)

# A param typo must not silently load normoxia, or 4hr, while every other check passes.
if (!nzchar(params$outcome_csv))
  stopifnot("outcome_csv does not name this contrast" =
    all(vapply(c(paste0("si", params$gene), "sictrl", params$condition, params$timepoint),
               function(t) grepl(t, outcome_file, fixed = TRUE), logical(1))))

cat(sprintf("\nTARGET: n=%d  median %+.3f  mean %+.3f  sd %.3f  MAD %.3f  range [%.2f, %.2f]\n",
            nrow(model_genes), median(model_genes$y), mean(model_genes$y), sd(model_genes$y),
            mad(model_genes$y), min(model_genes$y), max(model_genes$y)))
cat(sprintf("        |y|>1: %d   |y|>2: %d   |y|>3: %d\n",
            sum(abs(model_genes$y) > 1), sum(abs(model_genes$y) > 2),
            sum(abs(model_genes$y) > 3)))
# The median is not zero (0.304 on si3d hypoxia 1hr). Do not "recentre to fix it": centring is
# an affine transform and a variance-splitting forest is invariant to it, so it changes nothing
# the model does and nothing about importance shares. 10d_target_scale_analysis.Rmd measures and
# plots that, and asserts the equivalence.
```

```{r direction_check}
# Confirm the sign convention. This is the only use of rna_lfc in this notebook: it is not
# modelled, not predicted, and is excluded from the features.
#
# te_lfc cannot serve here. EIF3D's own te_lfc is -1.26, because EIF3D is TE-buffered in this
# contrast (its ribo_lfc of +1.45 tracks its rna_lfc of +2.90), so requiring it to be positive
# would fail on correct data. The siRNA destroys EIF3D's mRNA, so rna_lfc gives an unambiguous
# reading: this file is oriented log2(sictrl/si3d), and positive means eIF3d promotes.
# kd_symbol names the depleted gene when it is not EIF<gene>, e.g. EIF4G2 for the Weber DAP5
# knockout (gene = "dap5"); its own rna_lfc is then the ruler, exactly as EIF3D's is here.
kd_symbol <- if (nzchar(params$kd_symbol)) params$kd_symbol else
  paste0("EIF", toupper(params$gene))                        # 3d -> EIF3D, 3e -> EIF3E

# Some contrasts are saved as TE-only tables with no rna_lfc column - te_si3e_vs_sictrl_
# normoxia_1and4hr.csv is one - which would leave the direction unverifiable for that run.
# direction_check_csv points the ruler at the companion RNA file so the check still happens
# rather than being skipped, which is the difference between a verified column and an assumed one.
# Cross-cell-line direction ruler, for any cell line whose outcome file has no RNA column.
#
# This is the check that replaces `rna_lfc > 2`. It is built on independent data - the matched
# MDA-MB-231 contrast - which matters, because a same-cell-line gene set is derived from the very
# file whose sign is in question and so cannot catch an error made consistently upstream. eIF3d
# and eIF3e targets overlap between the two cell lines, so a correctly oriented outcome must
# correlate POSITIVELY with the 231 contrast; an inverted one correlates negatively by the same
# amount.
#
# 10m_mcf7six1_outcome_prep.Rmd runs this same check before it writes the outcome file. Repeating
# it here is deliberate rather than redundant: the prep notebook certifies the file, this
# certifies that THIS RUN loaded that file and not another one.
if (params$cell_line != "MDA-MB-231" && params$outcome_kind == "te" && nzchar(params$xcell_ref_csv)) {
  xref <- read_csv(here("output", params$xcell_ref_csv), show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>%
    dplyr::select(gene_id_clean, ref = all_of(params$outcome_col))
  xj <- model_genes %>% dplyr::select(gene_id_clean, y) %>%
    inner_join(xref, by = "gene_id_clean") %>% filter(is.finite(ref), is.finite(y))
  x_rho <- cor(xj$y, xj$ref, method = "spearman")
  cat(sprintf("cross-cell-line ruler vs %s: Spearman %+.4f (n=%d). An inverted sign would read %+.4f.\n",
              params$xcell_ref_csv, x_rho, nrow(xj), -x_rho))
  stopifnot(
    "the ruler contrast shares too few genes with this run to be informative" = nrow(xj) > 3000,
    "this contrast correlates NEGATIVELY with the matched MDA-MB-231 contrast - the sign is inverted and every direction in this run would be backwards" =
      x_rho > 0.05,
    # The ruler has to point at the SAME experiment. eIF3d and eIF3e targets overlap, so a si3d
    # outcome checked against a si3e reference still correlates positively: the sign would pass
    # while the run was anchored to the wrong contrast.
    "xcell_ref_csv names a different knockdown than this run" =
      grepl(paste0("si", params$gene), params$xcell_ref_csv, fixed = TRUE),
    "xcell_ref_csv names a different oxygen condition than this run" =
      grepl(sub("^[a-z0-9]+_", "", params$condition), params$xcell_ref_csv, fixed = TRUE),
    "xcell_ref_csv must be an MDA-MB-231 contrast - another cell line cannot serve as its own independent ruler" =
      !grepl("mcf7", params$xcell_ref_csv, fixed = TRUE)
  )
}

dir_src <- if (nzchar(params$direction_check_csv)) {
  cat("direction ruler read from the companion file:", params$direction_check_csv, "\n")
  read_csv(here("output", params$direction_check_csv), show_col_types = FALSE)
} else outcome_raw

if ("symbol" %in% names(dir_src) && "rna_lfc" %in% names(dir_src) &&
    !isTRUE(params$negate_outcome)) {
  e <- dir_src %>% filter(symbol == kd_symbol)
  stopifnot("the knockdown target gene is not in the outcome file" = nrow(e) == 1)
  cat(sprintf("%s rna_lfc %+.2f (the ruler: must be strongly positive)%s\n",
              kd_symbol, e$rna_lfc,
              if (params$outcome_col %in% names(e))
                sprintf(" | %s %+.2f", params$outcome_col, e[[params$outcome_col]]) else ""))
  stopifnot(
    "the knockdown gene's rna_lfc is not strongly positive - the contrast direction has flipped" =
      e$rna_lfc > 2
  )
  # Contrast anchor. The rna_lfc ruler above confirms the DIRECTION but not WHICH contrast was
  # loaded: every si3d file has the knockdown gene strongly up, so a normoxia file, or 4hr,
  # sails through it. This pins the specific contrast.
  #
  # No value is transcribed here. An earlier version of this check carried a hand-typed expected
  # value, and the value was simply wrong (0.104 against a true +0.9240 for normoxia 1hr) -
  # exactly the failure mode CLAUDE.md's "never transcribe reference values" rule exists for.
  # Instead the reference is parsed from the file that PARAMS imply, independently of
  # outcome_csv. When outcome_csv overrides to a different contrast the two disagree and this
  # fires; that override path is what this check guards, and 10z_'s wrong_condition_file
  # mutation is what proves it fires.
  implied_file <- paste0("translation_categories_si", params$gene, "_vs_sictrl_",
                         params$condition, "_", params$timepoint, ".csv")
  if (file.exists(here("output", implied_file))) {
    ref <- read_csv(here("output", implied_file), show_col_types = FALSE) %>%
      filter(symbol == kd_symbol)
    exp_y <- if (isTRUE(params$negate_outcome)) -ref[[params$outcome_col]]
             else ref[[params$outcome_col]]
    obs_y <- model_genes$y[model_genes$symbol == kd_symbol]
    # The anchor needs the knockdown gene to still be in the modelled set, and under
    # restrict_to_classifier_genes it cannot be: the classifier's genes are its positive set
    # (te_lfc > 0.5) plus its negative controls (|te_lfc| < 0.5), and EIF3D's own te_lfc of -1.26
    # satisfies neither, so it is absent from 02_'s gene sets by construction. Skipping with a
    # named reason is the honest response - asserting would make that arm unrunnable, and
    # relaxing the assert to length(obs_y) <= 1 would silently retire it for every run. The
    # rna_lfc ruler above still fires, so the DIRECTION is checked either way; only the
    # which-contrast-was-loaded anchor is unavailable.
    if (length(obs_y) != 1) {
      cat(sprintf(paste0("contrast anchor SKIPPED: %s is not among the %d modelled genes",
                         if (isTRUE(params$restrict_to_classifier_genes))
                           " (filtered out by restrict_to_classifier_genes - it is in neither of 02_'s gene sets)"
                         else " (absent from the feature matrix or its target is NA)",
                         ". The anchor cannot be evaluated; the rna_lfc direction ruler above still applies.\n"),
                  kd_symbol, nrow(model_genes)))
    } else {
      cat(sprintf("contrast anchor (%s): %s %s = %+.4f expected, %+.4f observed\n",
                  implied_file, kd_symbol, params$outcome_col, exp_y, obs_y))
      stopifnot(
        "the knockdown gene's own target does not match the contrast params imply - outcome_csv points at the wrong file" =
          length(obs_y) == 1 && length(exp_y) == 1 && abs(obs_y - exp_y) < 1e-6
      )
    }
  } else {
    cat("no file matches the params-implied contrast", implied_file, "- anchor not asserted\n")
  }
} else if (params$outcome_kind != "te") {
  cat("Direction rulers not applicable: this outcome has no knockdown (", params$outcome_kind,
      "). Its prep script holds the checks against the source; the spot-check below tests the ",
      "positive set vs the rest.\n", sep = "")
} else if (params$cell_line != "MDA-MB-231") {
  cat("rna_lfc ruler not applicable: this cell line's tables carry no RNA column.\n",
      "  The direction was checked above against the matched MDA-MB-231 contrast instead.\n", sep = "")
} else {
  cat("NOTE: direction check skipped (no symbol/rna_lfc column, or outcome was negated)\n")
}

# Biological spot-check on the outcome: it must separate the classifier's positive set from its
# negative controls. A sign flip or a scrambled join passes every range check above but fails
# this one.
# Reuses the single definition from the direction_facts chunk, so the two cannot diverge.
pos_genes <- .dir_pos
cat(sprintf("positive set: %d genes from %s\n", length(pos_genes), attr(pos_genes, "source")))
# neg_csv_path is resolved in derived_values from the cell line and knockdown, so an MCF7-SIX1
# run cannot be scored against MDA-MB-231's controls.
cat("negative controls:", basename(neg_csv_path), "\n")
neg_genes <- read_csv(neg_csv_path, show_col_types = FALSE) %>%
  mutate(g = sub("[.].*", "", gene_id)) %>% pull(g)
neg_genes <- setdiff(neg_genes, pos_genes)     # the same guard 02_ applies before sampling

yp <- model_genes$y[model_genes$gene_id_clean %in% pos_genes]
yn <- model_genes$y[model_genes$gene_id_clean %in% neg_genes]
cat(sprintf("target: %s-promoted median %+.3f (n=%d) vs negative controls %+.3f (n=%d), wilcox p=%.2e\n",
            kd_symbol, median(yp), length(yp), median(yn), length(yn),
            wilcox.test(yp, yn)$p.value))
stopifnot(
  "the target does not separate the promoted gene set from the negative controls" =
    median(yp) - median(yn) > as.numeric(params$min_spot_sep)
)
```

# Select features

Chooses which columns the forest sees, by joining against the annotation table.

**Decisions**

- Reuse `02_rf_model.Rmd`'s selection unchanged, which yields the same 69 features. This is what
  makes the SHAP comparison in `12_` a comparison of model type.
- Keep one member of each corrected/superseded feature pair, never both, so they cannot split
  credit between themselves.
- Preserve the matrix's column order, because ranger's feature sampling depends on it and
  reordering would change the forest even with identical seeds.
- Drop features that are more than half missing, and any that are constant.
- Guard against leakage two ways: by name, and by checking no feature correlates above 0.9 with
  the outcome. The second catches a leak that has been renamed, which a name pattern cannot see.
- Assert the feature set matches the classifier's, replacing `02_`'s baseline reproduction gate,
  which has no meaning here.


```{r select_features}
# Ported verbatim from 02_rf_model.Rmd, minus the gate branches: there is no frozen regression
# baseline to reproduce, so gate_against_baseline is replaced by gate_feature_parity below,
# which asserts the same 69 features the classifier uses. Keeping the selection identical is
# what makes the SHAP comparison in 12_ a comparison of model type rather than of feature sets.
aggregate_clip <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                    "clip_cap_binding_count_v2",
                    "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                    "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")

stopifnot("variant_set must be 'current' or 'corrected'" =
            params$variant_set %in% c("current", "corrected"))
sel <- annotation %>%
  filter(block %in% include_blocks, variant %in% c("single", params$variant_set))

stopifnot("initiation_score must be 'both', 'pwm' or 'noderer'" =
            params$initiation_score %in% c("both", "pwm", "noderer"))
drop_init <- switch(params$initiation_score,
  both    = character(0),
  pwm     = "noderer_tis_efficiency",
  noderer = c("kozak_pwm_score", "kozak_pwm_score_v2"))
if (MUTATING && params$mutate_target == "drop_variant_pair_guard")
  drop_init <- character(0)                              # let both PWM variants through
sel <- sel %>% filter(!feature %in% drop_init)
if (MUTATING && params$mutate_target == "drop_variant_pair_guard")
  sel <- annotation %>% filter(block %in% include_blocks)   # both members of every pair
n_init <- sum(sel$feature %in% c("kozak_pwm_score", "kozak_pwm_score_v2",
                                 "noderer_tis_efficiency"))
stopifnot("an exclusive initiation_score let two start-context scores through" =
            params$initiation_score == "both" || n_init <= 1L)

# Drop named individual features. Used to keep one representative per correlated cluster, so a
# signal split across near-duplicates is carried by a single row instead. This CHANGES the model:
# the retained feature absorbs the dropped ones' role, which is the whole point and is also why
# this cannot be done by deleting rows from a heatmap after the fact.
#
# A distinct feature_set_label is required, because exclude_features is not part of the output
# suffix - without it a reduced run would silently overwrite the full run's model, importance and
# SHAP. Same hazard neg_tag guards against in 02_.
if (nzchar(params$exclude_features)) {
  drop_feats <- str_trim(str_split(params$exclude_features, ",")[[1]])
  unknown <- setdiff(drop_feats, annotation$feature)
  stopifnot("exclude_features names a column that is not in the annotation" = !length(unknown),
            "exclude_features needs a distinct feature_set_label or it overwrites the full run" =
              params$feature_set_label != "tcoreg")
  n_before_ef <- nrow(sel)
  sel <- sel %>% filter(!feature %in% drop_feats)
  cat("Excluded", n_before_ef - nrow(sel), "named features:",
      paste(drop_feats, collapse = ", "), "\n")
}

if (nzchar(params$exclude_families)) {
  drop_fams <- str_trim(str_split(params$exclude_families, ",")[[1]])
  stopifnot("exclude_families names an unknown family" = all(drop_fams %in% annotation$family))
  n_drop <- sum(sel$family %in% drop_fams)
  sel <- sel %>% filter(!family %in% drop_fams)
  cat("Excluded", n_drop, "features from families:", paste(drop_fams, collapse = ", "), "\n")
}

if (!isTRUE(params$include_deprecated)) {
  n_dep <- sum(sel$deprecated)
  sel <- sel %>% filter(!deprecated)
  if (n_dep > 0) cat("Excluded", n_dep, "deprecated features\n")
}

if (params$clip_mode == "aggregate") {
  sel <- sel %>% filter(!(family == "clip" & !feature %in% aggregate_clip))
} else if (params$clip_mode == "individual") {
  sel <- sel %>% filter(!(family == "clip" & feature %in% aggregate_clip))
}

sel <- sel %>% filter(!str_detect(feature, "^(utr5|utr3|cds|total)_length$"))

# Preserve feature-matrix column order: ranger's mtry sampling depends on it, so reordering
# would change the forest even with identical seeds.
feature_cols <- colnames(features)[colnames(features) %in% sel$feature]

non_na <- map_dbl(feature_cols, ~ mean(!is.na(unlist(model_genes[[.x]]))))
sparse <- feature_cols[non_na < 0.5]
feature_cols <- setdiff(feature_cols, sparse)
if (length(sparse)) cat("Dropped", length(sparse), "sparse features (<50% non-NA)\n")

const <- feature_cols[map_lgl(feature_cols, function(f) {
  v <- unlist(model_genes[[f]]); v <- v[!is.na(v)]
  length(v) == 0 || stats::sd(v) == 0
})]
if (length(const)) {
  feature_cols <- setdiff(feature_cols, const)
  cat("Dropped", length(const), "zero-variance features:", paste(const, collapse = ", "), "\n")
}

list_cols <- feature_cols[map_lgl(feature_cols, ~ is.list(model_genes[[.x]]))]
for (col in list_cols) model_genes[[col]] <- as.numeric(unlist(model_genes[[col]]))

# Deliberate leak, for the mutation suite: a renamed copy of the outcome, which the name-based
# regex below cannot see and the value-based check must catch.
if (MUTATING && params$mutate_target == "leak_target_as_feature") {
  # Overwrite an EXISTING feature in place rather than appending a new one. Appending changes
  # the column set, which trips the column-order assert first and leaves the value-based
  # collinearity guard untested - the check would then have passed a mutation test without ever
  # having been shown to fire. This version leaves names, order and count untouched, so the
  # collinearity guard is the only thing that can catch it.
  leak_into <- feature_cols[length(feature_cols)]
  model_genes[[leak_into]] <- model_genes$y
  cat("leaked the target into the existing column:", leak_into, "\n")
}

cat("Features:", length(feature_cols), "\n")

stopifnot(
  # Name-based guard, inherited from 02_.
  "an outcome-derived column reached the feature set" =
    !any(str_detect(feature_cols,
      paste0("^te_lfc$|^te_padj$|^te_lfc_bin$|^sig_|^indiv_te_|^delta_te_|",
             "^y$|^y_padj$|^category$|^rna_lfc$|^ribo_lfc$|^rna_padj$|^ribo_padj$"))),
  "feature_cols is not in matrix column order" =
    identical(feature_cols, colnames(features)[colnames(features) %in% feature_cols])
)

# Value-based guard. 02_'s leakage check is purely name-based, so it would pass on a leak that
# had simply been renamed - the same blind spot g4_region_swap exposed in 01_, where a check
# that asserted only a convention passed on a scrambled join. A near-collinear feature is the
# signature a name regex structurally cannot see.
leak_cor <- map_dbl(feature_cols, ~ suppressWarnings(
  cor(as.numeric(model_genes[[.x]]), model_genes$y, use = "complete.obs", method = "spearman")))
names(leak_cor) <- feature_cols
worst <- leak_cor[which.max(abs(leak_cor))]
cat(sprintf("strongest feature-target Spearman: %s %+.3f\n", names(worst), worst))
stopifnot("a feature is near-collinear with the outcome - a renamed leak" =
            max(abs(leak_cor), na.rm = TRUE) < 0.9)

# Parity with the classifier's feature set. This replaces gate_against_baseline: it has content
# (it can fail) and it is the precondition that makes 12_'s comparison about model type.
# Parity cannot hold when features have been deliberately removed, so the gate skips itself
# rather than requiring every caller to remember to switch it off. Its purpose is to guarantee
# that 12_'s classifier comparison is about model type; a reduced-feature run is not for that
# comparison, and saying so here is better than failing on a condition that is false by design.
if (isTRUE(params$gate_feature_parity) && nzchar(params$exclude_features)) {
  cat("Feature-parity gate SKIPPED: exclude_features removed features on purpose, so parity\n",
      "  with the classifier's set cannot hold. This run is not comparable to 02_ in 12_.\n")
}
if (isTRUE(params$gate_feature_parity) && !nzchar(params$exclude_features)) {
  clf_imp_path <- file.path(out_dir, paste0("rf_importance_", clf_suffix, ".csv"))
  if (file.exists(clf_imp_path)) {
    clf_features <- read_csv(clf_imp_path, show_col_types = FALSE)$feature
    if (!setequal(clf_features, feature_cols)) {
      cat("PARITY DIFFERS - only in regression:",
          paste(setdiff(feature_cols, clf_features), collapse = ", "), "\n")
      cat("               only in classifier:",
          paste(setdiff(clf_features, feature_cols), collapse = ", "), "\n")
    }
    stopifnot("feature set differs from the classifier's - 12_'s comparison would not be about model type" =
                setequal(clf_features, feature_cols))
    cat("Feature parity with", clf_suffix, "OK (", length(feature_cols), "features )\n")
  } else {
    cat("NOTE: parity gate skipped -", basename(clf_imp_path), "not found\n")
  }
}

# Cross-run feature parity. The gate above compares against 02_'s classifier, which exists only
# for the MDA-MB-231 contrasts; for any other cell line it prints "not found" and the run goes
# ahead with no parity check at all - which is precisely the guarantee a multi-cell-line heatmap
# depends on. parity_ref_suffix names a REGRESSION run instead, so every column that will share
# an axis can be tied explicitly to one feature set.
if (nzchar(params$parity_ref_suffix)) {
  ref_imp <- file.path(out_dir, paste0("rfreg_importance_", params$parity_ref_suffix, ".csv"))
  stopifnot("parity_ref_suffix names a run with no saved importance - knit that run first" =
              file.exists(ref_imp))
  ref_feats <- read_csv(ref_imp, show_col_types = FALSE)$feature
  if (!setequal(ref_feats, feature_cols)) {
    cat("PARITY DIFFERS - only in this run:",
        paste(setdiff(feature_cols, ref_feats), collapse = ", "), "\n")
    cat("               only in the reference:",
        paste(setdiff(ref_feats, feature_cols), collapse = ", "), "\n")
  }
  stopifnot("feature set differs from parity_ref_suffix - these two runs cannot share a heatmap axis" =
              setequal(ref_feats, feature_cols))
  cat("Cross-run feature parity with", params$parity_ref_suffix, "OK (",
      length(feature_cols), "features )\n")
}
```

# Split, transform, impute

Splits genes into training and held-out sets, optionally transforms the outcome, and fills
missing feature values.

**Decisions**

- Split 80/20 with `createDataPartition`, which bins a numeric outcome into quantiles so the two
  halves have matched distributions.
- Let `split_seed` seed only the split. The forest and the cross-validation use `forest_seed`, so
  across a sweep any difference is attributable to which genes were held out.
- Fit every transform on the training split alone and apply it to the held-out set, so no
  held-out information reaches the map.
- Impute missing features with training-set medians. The code asserts those medians actually
  differ from full-data medians, so the constraint is doing work rather than being vacuous.
- Allow the rank transform to produce ties at the extremes. `ecdf` is a step function, so
  held-out genes beyond the training range all map to 0 or 1. That is lost resolution at the
  tails, and it is reported rather than treated as an error.


```{r train_test_split}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, y, y_padj, all_of(feature_cols))

# Shuffled-family null, unchanged from 02_: permuting a family's values across genes destroys
# its link to the outcome while preserving every marginal, the NA pattern and the within-family
# correlation structure. Impurity importance is non-negative, so an uninformative family never
# scores zero - whatever the permuted copy absorbs is the floor the real one must clear.
if (nzchar(params$shuffle_families)) {
  shuf <- str_trim(str_split(params$shuffle_families, ",")[[1]])
  stopifnot("shuffle_families names an unknown family" = all(shuf %in% annotation$family))
  shuf_cols <- intersect(feature_cols, annotation$feature[annotation$family %in% shuf])
  set.seed(SPLIT_SEED)                    # tied to the split, so the sweep stays paired
  perm <- sample(nrow(model_data))
  for (cc in shuf_cols) model_data[[cc]] <- model_data[[cc]][perm]
  cat("SHUFFLED", length(shuf_cols), "features in:", paste(shuf, collapse = ", "), "\n")
}

# split_seed is the only RNG consumer here, and the only thing it seeds. The forest and the CV
# use forest_seed, so across a sweep every difference is attributable to which genes were held
# out. createDataPartition bins a numeric y into quantiles, so the two splits stay
# distributionally matched rather than merely random.
set.seed(SPLIT_SEED)
train_idx  <- createDataPartition(model_data$y, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]
cat(sprintf("Train %d (median y %+.4f, MAD %.4f) | Test %d (median y %+.4f, MAD %.4f)\n",
            nrow(train_data), median(train_data$y), mad(train_data$y),
            nrow(test_data),  median(test_data$y),  mad(test_data$y)))
```

```{r target_transform}
# Winsorizing first, with bounds from the training split only, then the transform. Both are fitted on the
# training split and applied to test, so no test information reaches either map.
win_bounds <- NULL
if (params$winsorize_y > 0) {
  # The bounds are a FIXED CONSTANT from params, not estimated from the data, so winsorizing
  # has no train/test leakage channel by construction - there is nothing here for a
  # "bounds from all rows" mutation to corrupt. If bounds are ever made quantile-based, they
  # must be computed on train only and that change needs its own mutation.
  b <- params$winsorize_y
  win_bounds <- c(-b, b)
  n_cap_tr <- sum(train_data$y < win_bounds[1] | train_data$y > win_bounds[2])
  train_data$y_win <- pmin(pmax(train_data$y, win_bounds[1]), win_bounds[2])
  test_data$y_win  <- pmin(pmax(test_data$y,  win_bounds[1]), win_bounds[2])
  cat("Winsorized at", win_bounds, "-", n_cap_tr, "training genes capped\n")
} else {
  train_data$y_win <- train_data$y
  test_data$y_win  <- test_data$y
}

apply_transform <- function(y_tr, y_te, mode) switch(mode,
  raw = list(tr = y_tr, te = y_te, inv_scale = 1, ecdf = NULL),
  # AFFINE, and therefore a NO-OP for a variance-splitting forest: the argmax split is unchanged
  # and node impurity scales by 1/MAD^2, so trees, predictions and importance shares are
  # identical. Kept only so the equivalence stays assertable and so 10d_ can plot it. It exists
  # to report SHAP in MAD units, not to change the model.
  robust_z = { m <- median(y_tr); s <- mad(y_tr)
               stopifnot("MAD of the training target is zero" = s > 0)
               list(tr = (y_tr - m) / s, te = (y_te - m) / s, inv_scale = s, ecdf = NULL) },
  # NON-LINEAR, and the only transform that changes the fit. ONE map, built on training rows and
  # applied to both, so train and test share a scale by construction rather than by assertion.
  # (31_ in the frozen pipeline uses rank()/n for train and ecdf for test; those disagree on
  # ties. Using one map removes the class of bug.)
  rank_ecdf = { e <- stats::ecdf(y_tr); list(tr = e(y_tr), te = e(y_te), inv_scale = NA_real_, ecdf = e) },
  stop("unknown target_transform"))

if (MUTATING && params$mutate_target == "ecdf_on_all_rows" &&
    params$target_transform == "rank_ecdf") {
  e_bad <- stats::ecdf(model_data$y)                       # built on train AND test
  tf <- list(tr = e_bad(train_data$y_win), te = e_bad(test_data$y_win),
             inv_scale = NA_real_, ecdf = e_bad)
} else {
  tf <- apply_transform(train_data$y_win, test_data$y_win, params$target_transform)
}
train_data$y_t <- tf$tr
test_data$y_t  <- tf$te

# Every transform here is monotone non-decreasing, which is what lets 10d_ score every arm on one
# yardstick (Spearman against raw held-out y). It is not always strictly increasing: ecdf() is a
# step function, so test genes below the smallest training value all map to 0 and those above the
# largest all map to 1. Those ties are real lost resolution at the tails, not a bug - so they are
# asserted as weak monotonicity and then reported, rather than being allowed to fail a strict
# check or be silently waved through.
is_nondecreasing <- function(x, y) {
  o <- order(x, y); all(diff(y[o]) >= -1e-12)
}
stopifnot(
  "the transform is not monotone non-decreasing in raw y (train)" =
    is_nondecreasing(train_data$y_win, train_data$y_t),
  "the transform is not monotone non-decreasing in raw y (test)"  =
    is_nondecreasing(test_data$y_win,  test_data$y_t)
)
if (params$target_transform %in% c("raw", "robust_z")) {
  # These two are strictly increasing, so they must preserve the ranking exactly.
  stopifnot(
    "an affine transform did not preserve the ranking exactly" =
      isTRUE(all.equal(cor(test_data$y_win, test_data$y_t, method = "spearman"), 1))
  )
} else {
  n_below <- sum(test_data$y_win < min(train_data$y_win))
  n_above <- sum(test_data$y_win > max(train_data$y_win))
  cat(sprintf("rank map: %d test genes clipped to 0 (below train min), %d to 1 (above max); Spearman vs raw y %.6f\n",
              n_below, n_above, cor(test_data$y_win, test_data$y_t, method = "spearman")))
}
if (params$target_transform == "rank_ecdf") {
  stopifnot(
    "the ecdf was built on more rows than the training split" =
      length(environment(tf$ecdf)$x) == nrow(train_data),
    "test ranks escaped [0,1]" = all(test_data$y_t >= 0 & test_data$y_t <= 1),
    # Discriminative: an ecdf built on ALL rows must give DIFFERENT test ranks, or the
    # train-only constraint is doing no work and this check is uninformative.
    # model_data has no y_win column - winsorize_y is asserted to be 0 under rank_ecdf just
    # below, so y is the winsorized target here by definition.
    "train-only and all-rows ecdf agree - the no-leakage constraint is vacuous" =
      max(abs(stats::ecdf(model_data$y)(test_data$y) - test_data$y_t)) > 1e-4,
    # Winsorizing is provably a no-op under a monotone rank map: it only creates ties at the
    # bounds. Combining them would look like an extra arm while testing nothing.
    "winsorize_y has no effect under rank_ecdf - set one or the other" = params$winsorize_y == 0
  )
}
cat(sprintf("Transformed target: train [%.3f, %.3f] median %.3f | test [%.3f, %.3f]\n",
            min(train_data$y_t), max(train_data$y_t), median(train_data$y_t),
            min(test_data$y_t),  max(test_data$y_t)))
```

```{r impute}
# Medians from the training split only, then applied to both.
train_medians <- train_data %>% dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~ median(., na.rm = TRUE)))
all_medians <- model_data %>% dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~ median(., na.rm = TRUE)))
if (MUTATING && params$mutate_target == "impute_from_all_rows")
  train_medians <- all_medians                            # the corruption

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  df
}
train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

# The no-leakage constraint is only meaningful if train-only medians actually differ from
# full-data medians. If they were identical, the constraint would be vacuous and the check
# uninformative - so assert the difference exists, not just that imputation ran.
stopifnot(
  "train-only and full-data medians are identical - the no-leakage constraint is vacuous" =
    any(abs(unlist(train_medians) - unlist(all_medians)) > 0),
  "NAs survived imputation" =
    sum(is.na(train_data[, feature_cols])) + sum(is.na(test_data[, feature_cols])) == 0
)
cat("Imputed; median-difference columns (train vs all):",
    sum(abs(unlist(train_medians) - unlist(all_medians)) > 0), "of", length(feature_cols), "\n")
```

# Train

Fits the regression forest.

**Decisions**

- Set `min.node.size` explicitly rather than relying on the default, which differs between
  regression and probability forests (5 versus 10). `mtry` is the same for both here, 8.
- Use 1000 trees rather than `02_`'s 500, because this forest is grown on 8,140 rows rather than
  2,298.
- Build the argument list so unset parameters are genuinely absent, rather than passed as a value
  that happens to match the current default.
- Do not weight observations. Weighting by `padj` would weight by a function of the outcome's
  own magnitude, which leaks; that option is blocked rather than merely discouraged. `baseMean`
  is available as a proxy, and a properly weighted model would need the standard errors for this
  contrast, which are not saved anywhere.


```{r train_rf}
# Optional precision weighting. baseMean is the only safe proxy available: it is a precision
# proxy that is not a function of |y|. Real lfcSE for this contrast is not saved anywhere (the
# saved CSV has padj but no lfcSE, and condition_level_te.csv's lfcSE belongs to the absolute
# per-condition TE, a different estimate), so a genuinely weighted model needs a small notebook
# that re-runs deFunction("TE", ...) first. Weighting changes the loss, so a weighted arm is not
# comparable to the classifier.
case_wts <- NULL
if (params$weight_mode == "basemean") {
  bm <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>%
    group_by(gene_id_clean) %>% summarise(bm = mean(baseMean, na.rm = TRUE), .groups = "drop")
  w <- train_data %>% dplyr::select(gene_id_clean) %>% left_join(bm, by = "gene_id_clean") %>%
    mutate(w = log10(1 + coalesce(bm, median(bm, na.rm = TRUE)))) %>% pull(w)
  case_wts <- w / mean(w)
  cat("Weighted by log10(1+baseMean); range", round(range(case_wts), 3), "\n")
}

# Built as an arg list so an unset param is genuinely absent rather than passed as a value that
# happens to match today's default. sample.fraction in particular defaults to 1 with replacement
# but 0.632 without, so hardcoding 1 would silently change meaning if replace were ever set.
rf_args <- list(
  x             = train_data[, feature_cols],
  y             = train_data$y_t,
  num.trees     = params$num_trees,
  min.node.size = params$min_node_size,
  splitrule     = params$splitrule,
  # maxstat gives unbiased variable SELECTION but its impurity semantics differ, so that arm
  # reports permutation importance instead and is compared on SHAP, which is engine-agnostic.
  importance    = if (params$splitrule == "maxstat") "permutation" else "impurity",
  seed          = FOREST_SEED
)
if (params$mtry > 0)            rf_args$mtry            <- params$mtry
if (params$sample_fraction > 0) rf_args$sample.fraction <- params$sample_fraction
if (!is.null(case_wts))         rf_args$case.weights    <- case_wts

set.seed(FOREST_SEED)
rf_model <- do.call(ranger, rf_args)

# do.call() records the entire deparsed ranger function in $call - about 35,000 characters
# against 53 for a direct call - so print(rf_model) would dump the whole function body into the
# report. Replace it with a readable record of what was actually run. The data arguments are
# deliberately left out; x and y would be deparsed in full for the same reason.
rf_model$call <- as.call(c(
  quote(ranger),
  list(x = quote(train_data[, feature_cols]), y = quote(y_train)),
  rf_args[setdiff(names(rf_args), c("x", "y"))]
))

# A compact summary, rather than ranger's default print.
cat(sprintf("%-16s %s\n", "type:", rf_model$treetype))
cat(sprintf("%-16s %d trees, mtry %d, min.node.size %d, splitrule %s\n", "forest:",
            rf_model$num.trees, rf_model$mtry, rf_model$min.node.size, params$splitrule))
cat(sprintf("%-16s %d genes x %d features\n", "trained on:",
            rf_model$num.samples, rf_model$num.independent.variables))
cat(sprintf("%-16s %.4f  (MSE %.5f)\n", "OOB R2:",
            rf_model$r.squared, rf_model$prediction.error))
cat(sprintf("%-16s %s\n", "call:", paste(deparse(rf_model$call), collapse = " ")))

stopifnot("this is not a regression forest" = rf_model$treetype == "Regression")
# ranger's default mtry is floor(sqrt(p)) for both probability and regression forests; only
# min.node.size differs by tree type (10 vs 5). The floor(p/3) regression default belongs to
# randomForest, not ranger - 03_shap.Rmd's comment says otherwise and is wrong. Measured here so
# 12_ can state that the SHAP comparison is not confounded by mtry.
if (params$mtry == 0)
  stopifnot("ranger's default mtry is not floor(sqrt(p)) - re-measure before comparing to 02_" =
              rf_model$mtry == floor(sqrt(length(feature_cols))))
cat("mtry", rf_model$mtry, "| min.node.size", rf_model$min.node.size,
    "| num.trees", rf_model$num.trees, "| splitrule", params$splitrule,
    "| OOB R2", round(rf_model$r.squared, 4), "\n")

# The only structural guard against a leak that slipped through both the name and the
# collinearity check - e.g. a linear combination of features reconstructing the outcome.
stopifnot("OOB R2 implausibly high - suspect leakage" = rf_model$r.squared < 0.8)
```

# Evaluate

Scores the model on the held-out genes.

**Decisions**

- Score every variant against the raw held-out outcome, whatever it was trained on. Scoring a
  rank-trained model on ranks and a raw-trained model on raw values compares two different
  yardsticks.
- Report Spearman as the headline. R-squared is reported too but is sensitive to the extreme
  genes, whose values reach about 19 robust standard deviations.
- Re-impute and re-transform inside each cross-validation fold, so the no-leakage rule holds per
  fold as well as on the main split.


```{r evaluate}
# Every arm is scored against raw held-out y, whatever it was trained on. Scoring a rank model
# by R2 on ranks and a raw model by R2 on raw y compares two different yardsticks and makes the
# rank model look better partly by definition (ranks are flatter and more predictable). Spearman
# against raw y is the headline for that reason; R2 is reported too but is tail-sensitive on a
# target whose extremes reach 19 robust-sd.
pred_t    <- predict(rf_model, data = test_data[, feature_cols])$predictions
pred_tr   <- predict(rf_model, data = train_data[, feature_cols])$predictions
spearman_raw <- cor(pred_t, test_data$y, method = "spearman")
pearson_raw  <- cor(pred_t, test_data$y)

# On the trained scale, so R2 is a like-for-like fit statistic there.
ss_res <- sum((test_data$y_t - pred_t)^2)
ss_tot <- sum((test_data$y_t - mean(train_data$y_t))^2)
r2_var <- 1 - ss_res / ss_tot
rmse_t <- sqrt(mean((test_data$y_t - pred_t)^2))

metrics <- tibble(
  metric = c("spearman_raw_y", "pearson_raw_y", "r2_pearson_sq_raw_y",
             "r2_var_explained_trained_scale", "rmse_trained_scale", "oob_r2",
             "spearman_train", "n_genes", "n_train", "n_test", "n_features",
             "mtry", "min_node_size", "num_trees",
             "target_median", "target_mad", "target_sd"),
  value  = c(spearman_raw, pearson_raw, pearson_raw^2, r2_var, rmse_t, rf_model$r.squared,
             cor(pred_tr, train_data$y, method = "spearman"),
             nrow(model_data), nrow(train_data), nrow(test_data), length(feature_cols),
             rf_model$mtry, rf_model$min.node.size, rf_model$num.trees,
             median(model_data$y), mad(model_data$y), sd(model_data$y)))
print(as.data.frame(metrics %>% mutate(value = round(value, 5))), row.names = FALSE)

# Predictions must separate the two gene sets on held-out genes, not just in-sample.
# The held-out separation check needs enough positives for a median to mean anything. At the
# late timepoints there are almost none - si3d hypoxia 24hr puts a single positive gene in the
# 20% test split, si3e 24hr five - so the check would be comparing one gene's prediction against
# 573 negatives and passing or failing on noise. It is therefore gated on having at least
# min_test_pos held-out positives, and says loudly when it skips.
#
# Worth noting the converse: a run with only a handful of positives that PASSES has not been
# meaningfully checked either. The gate applies in both directions.
n_test_pos <- sum(test_data$gene_id_clean %in% pos_genes)
n_test_neg <- sum(test_data$gene_id_clean %in% neg_genes)
sep_pos <- median(pred_t[test_data$gene_id_clean %in% pos_genes])
sep_neg <- median(pred_t[test_data$gene_id_clean %in% neg_genes])
if (!isTRUE(params$heldout_sep_check)) {
  # Off only for cap-binding stress RATIOS, where the positive set is the hit list from one
  # condition and does not define the outcome. There the direction is a result, not a check.
  cat(sprintf("held-out separation REPORTED, not asserted: predicted median %+.3f for %d positives vs %+.3f for %d negatives\n",
              sep_pos, n_test_pos, sep_neg, n_test_neg))
} else if (n_test_pos >= params$min_test_pos) {
  stopifnot(
    "predictions do not separate the promoted set from the negative controls on held-out genes" =
      median(pred_t[test_data$gene_id_clean %in% pos_genes]) >
      median(pred_t[test_data$gene_id_clean %in% neg_genes])
  )
  cat(sprintf("held-out separation check PASSED on %d positives vs %d negatives\n",
              n_test_pos, n_test_neg))
} else {
  cat(sprintf("\n*** held-out separation check SKIPPED: only %d positives in the test split\n",
              n_test_pos))
  cat(sprintf("    (need %d). The positive set has %d genes in the whole matrix, so this\n",
              params$min_test_pos, sum(model_data$gene_id_clean %in% pos_genes)))
  cat("    contrast has too few affected transcripts for the check to have any power. ***\n\n")
}
```

```{r permutation_null, eval = isTRUE(params$permutation_check)}
# A validity check that depends on nothing outside this run: refit on a PERMUTED outcome and
# confirm the signal disappears.
#
# The alternative checks all lean on an external gene set defined by a threshold on the same
# quantity being modelled, which is circular and, at the late timepoints, has almost no members.
# This one asks the only question that matters for a column in the heatmap: is the Spearman we
# just measured attributable to the features, or would any forest on this many genes produce it?
#
# Permuting the outcome while leaving the features untouched destroys the feature-outcome link
# and preserves everything else - the marginal distribution, the sample size, the correlation
# structure among features. Whatever Spearman survives is what this setup yields for free.
set.seed(FOREST_SEED)
y_perm <- sample(train_data$y_t)
perm_args <- rf_args; perm_args$y <- y_perm; perm_args$importance <- "none"
set.seed(FOREST_SEED)
rf_perm  <- do.call(ranger, perm_args)
pred_perm <- predict(rf_perm, data = test_data[, feature_cols])$predictions
sp_perm   <- cor(pred_perm, test_data$y, method = "spearman")

# perm_n > 1 (cap-binding runs): one shuffle is not a null distribution. On small test sets with
# a strongly feature-determined outcome, shuffled forests scatter +/-0.2 around zero (measured on
# the subunit-seq columns), so a fixed |shuffle| < 0.1 bound fails by chance. There the real model
# must instead beat the MAXIMUM of perm_n shuffled forests. perm_n = 1 keeps the original check.
if (as.integer(params$perm_n) > 1) {
  sp_perm_all <- vapply(seq_len(as.integer(params$perm_n)), function(i) {
    set.seed(FOREST_SEED + i)
    pa <- rf_args; pa$y <- sample(train_data$y_t); pa$importance <- "none"; pa$seed <- FOREST_SEED + i
    f <- do.call(ranger, pa)
    cor(predict(f, data = test_data[, feature_cols])$predictions, test_data$y, method = "spearman")
  }, numeric(1))
  cat(sprintf("%d shuffled forests: Spearman mean %+.3f, sd %.3f, range [%+.3f, %+.3f]\n",
              length(sp_perm_all), mean(sp_perm_all), sd(sp_perm_all),
              min(sp_perm_all), max(sp_perm_all)))
}

cat(sprintf("real model      : test Spearman %+.4f | OOB R2 %+.4f\n", spearman_raw, rf_model$r.squared))
cat(sprintf("permuted outcome: test Spearman %+.4f | OOB R2 %+.4f\n", sp_perm, rf_perm$r.squared))
cat(sprintf("difference      : %+.4f\n", spearman_raw - sp_perm))
if (as.integer(params$perm_n) > 1) {
  stopifnot(
    "shuffled forests are centred away from zero - suspect leakage" = abs(mean(sp_perm_all)) < 0.1,
    "the real model does not beat the best of its shuffled forests" = spearman_raw > max(sp_perm_all))
} else stopifnot(
  # The permuted model must not rank held-out genes. If it does, something links the features to
  # the outcome other than biology - a leak, or a split that is not independent.
  "the PERMUTED model still ranks held-out genes - suspect leakage" = abs(sp_perm) < 0.1,
  # And the real model must beat it. This is the floor every column has to clear to be read.
  "the real model does not beat its own permuted null" = spearman_raw > sp_perm
)
if (spearman_raw - sp_perm < 0.1)
  cat("\n*** the real model barely beats its permuted null: this column carries little signal ***\n")
# Exported, so the heatmap can show every column's null rather than each render's log holding it.
perm_metrics <- tibble(
  metric = c("perm_spearman", "perm_oob_r2", "perm_margin"),
  value  = c(sp_perm, rf_perm$r.squared, spearman_raw - sp_perm))
rm(rf_perm, perm_args)
```

```{r cross_validation, eval = isTRUE(params$run_cv)}
# 5-fold CV over all of model_data, re-imputing and re-transforming within each fold so the
# no-leakage discipline holds per fold as well as on the main split.
set.seed(FOREST_SEED)
folds <- createFolds(model_data$y, k = 5)
cv <- map_dfr(seq_along(folds), function(i) {
  idx <- folds[[i]]
  tr <- model_data[-idx, ]; te <- model_data[idx, ]
  b <- params$winsorize_y
  tr$y_win <- if (b > 0) pmin(pmax(tr$y, -b), b) else tr$y
  te$y_win <- if (b > 0) pmin(pmax(te$y, -b), b) else te$y
  tfi <- apply_transform(tr$y_win, te$y_win, params$target_transform)
  med <- tr %>% dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~ median(., na.rm = TRUE)))
  tr <- impute_with_medians(tr, med); te <- impute_with_medians(te, med)
  set.seed(FOREST_SEED)
  fit <- ranger(x = tr[, feature_cols], y = tfi$tr, num.trees = params$num_trees,
                mtry = if (params$mtry > 0) params$mtry else NULL,
                min.node.size = params$min_node_size, splitrule = params$splitrule,
                seed = FOREST_SEED)
  p <- predict(fit, data = te[, feature_cols])$predictions
  tibble(fold = i,
         spearman = cor(p, te$y, method = "spearman"),
         pearson  = cor(p, te$y),
         rmse     = sqrt(mean((tfi$te - p)^2)))
})
cat(sprintf("5-fold CV: Spearman %.4f +/- %.4f | Pearson %.4f +/- %.4f | RMSE %.4f +/- %.4f\n",
            mean(cv$spearman), sd(cv$spearman), mean(cv$pearson), sd(cv$pearson),
            mean(cv$rmse), sd(cv$rmse)))
```

# AUC bridge

Computes an AUC so this model can be put next to the classifier's reported number.

A regression has no threshold, so an AUC only exists against labels supplied from outside. Here
the labels are the classifier's own gene sets, which this model never saw.

**Decisions**

- Treat this as a comparison aid, not a primary metric. Spearman and Precision@K are the primary
  metrics.
- Report the ceiling next to the AUC. For a label that is itself a cut on the outcome, the
  ceiling is 1.0 by construction, so the AUC alone would be misleading.
- Report how much the two models' held-out sets overlap, and the AUC restricted to the genes both
  models held out, with its confidence interval. That subset is around 112 genes, where the
  interval is roughly plus or minus 0.08, so it cannot separate 0.83 from 0.78.
- Leave the real comparison to `12_`, which compares the two models' distributions over their
  respective 20 replicates instead of two single numbers.


```{r auc_bridge, eval = isTRUE(params$auc_bridge)}
# A regression has no threshold, so an AUC exists only against an externally supplied label.
# This is a comparison aid to 02_'s reported test AUC, not a primary metric, and three
# caveats travel with it as columns rather than prose:
#
#  (1) the two models have different test splits, and this regression trained on ~80% of the
#      classifier's test genes, so the two AUCs are not measured on the same genes;
#  (2) the "threshold" label is a cut on the regressed quantity, so its ceiling is ~1.0 by
#      construction - report auc / ceiling, never auc alone;
#  (3) the only fair subset is the genes held out by both models (~115), where the 95% CI
#      half-width is ~0.08 and 0.83 cannot be separated from 0.78. The honest comparison is
#      therefore distribution-vs-distribution over the two sweeps, which 12_ does.
stopifnot("bridge label sets overlap" = length(intersect(pos_genes, neg_genes)) == 0)

br <- test_data %>%
  mutate(pred = pred_t,
         lab_geneset = case_when(gene_id_clean %in% pos_genes ~ 1L,
                                 gene_id_clean %in% neg_genes ~ 0L, TRUE ~ NA_integer_),
         lab_thresh  = as.integer(!is.na(y_padj) & y > as.numeric(params$lfc) & y_padj < 0.05))

auc_of <- function(lab, score) {
  k <- !is.na(lab) & !is.na(score)
  if (length(unique(lab[k])) < 2) return(NA_real_)
  as.numeric(auc(roc(lab[k], score[k], direction = "<", quiet = TRUE)))
}
# The self-contained AUC. Positives are a threshold on this contrast's own outcome and negatives
# are every other held-out gene, so it needs no external gene set and is comparable across
# knockdowns and timepoints. This is the AUC reported.
#
# The screened-comparator version is kept below for continuity with notebook 02, but it is NOT
# the headline: its negatives are genes selected for having no effect, a pool measured to be
# GC-enriched by +2.45 (Wilcoxon p = 2.7e-17), which inflates it (0.806 against 0.721 here).
bridge <- tibble(
  # row 1 = against the screened negative controls; row 2 = against every other held-out gene.
  label   = c("vs_screened_controls", "vs_all_other_genes"),
  n_pos   = c(sum(br$lab_geneset == 1, na.rm = TRUE), sum(br$lab_thresh == 1)),
  n_neg   = c(sum(br$lab_geneset == 0, na.rm = TRUE), sum(br$lab_thresh == 0)),
  auc     = c(auc_of(br$lab_geneset, br$pred), auc_of(br$lab_thresh, br$pred)),
  # The ceiling: the same AUC scored with the TRUE outcome as the predictor.
  ceiling = c(auc_of(br$lab_geneset, br$y),    auc_of(br$lab_thresh, br$y))
) %>% mutate(auc_frac_of_ceiling = auc / ceiling)
print(as.data.frame(bridge %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)

# Precision@K against the random baseline, which is the prevalence.
ks <- as.integer(str_trim(str_split(params$precision_at_k, ",")[[1]]))
prec <- map_dfr(ks, function(k) {
  k <- min(k, nrow(br))
  top <- br %>% arrange(desc(pred)) %>% head(k)
  tibble(K = k, precision = mean(top$lab_thresh == 1),
         baseline = mean(br$lab_thresh == 1)) %>% mutate(enrichment = precision / baseline)
})
print(as.data.frame(prec %>% mutate(across(where(is.double), ~ round(., 4)))), row.names = FALSE)

# Overlap with the classifier's split, and the matched-gene comparison with its CI.
clf_md_path <- file.path(out_dir, paste0("rf_model_data_", clf_suffix, ".rds"))
matched_auc <- NA_real_; n_matched <- NA_integer_
if (file.exists(clf_md_path)) {
  clf_md  <- readRDS(clf_md_path)
  matched <- intersect(clf_md$test$gene_id_clean, test_data$gene_id_clean)
  n_matched <- length(matched)
  trained_on <- sum(clf_md$test$gene_id_clean %in% train_data$gene_id_clean)
  cat(sprintf("\nOverlap with 02_'s split: this regression trained on %d of the classifier's %d test genes (%.0f%%)\n",
              trained_on, nrow(clf_md$test), 100 * trained_on / nrow(clf_md$test)))
  mb <- br %>% filter(gene_id_clean %in% matched, !is.na(lab_geneset))
  if (nrow(mb) > 10 && length(unique(mb$lab_geneset)) == 2) {
    r <- roc(mb$lab_geneset, mb$pred, direction = "<", quiet = TRUE)
    ci <- as.numeric(ci.auc(r))
    matched_auc <- as.numeric(auc(r))
    cat(sprintf("Matched-gene AUC (held out by both models): %.4f [95%% CI %.3f-%.3f], n = %d - underpowered by design\n",
                matched_auc, ci[1], ci[3], nrow(mb)))
  }
} else {
  cat("\nNOTE: 02_'s rf_model_data not found; overlap and matched-gene AUC skipped\n")
}
```

# Importance

Reports which features the forest relied on, and plots the fit.

**Decisions**

- Report impurity importance here, and treat family-level sums as descriptive only. Impurity is
  not additive, so summing it within a family is not exact. `11_shap_regression.Rmd` gives the
  exact split using SHAP.


```{r importance}
importance_df <- tibble(
  feature    = names(rf_model$variable.importance),
  importance = as.numeric(rf_model$variable.importance)
) %>%
  left_join(annotation %>% dplyr::select(feature, family, block), by = "feature") %>%
  arrange(desc(importance)) %>%
  mutate(share_pct = 100 * importance / sum(importance), rank = row_number())
print(as.data.frame(head(importance_df, 20)), row.names = FALSE)

cat("\nFamily shares (impurity is NOT additive - descriptive only; 11_ gives the exact SHAP split):\n")
print(as.data.frame(importance_df %>% group_by(family) %>%
  summarise(n = n(), share_pct = round(sum(share_pct), 2), .groups = "drop") %>%
  arrange(desc(share_pct))), row.names = FALSE)
```

```{r fig_fit, fig.width = 10, fig.height = 4.6, eval = isTRUE(params$auc_bridge)}
# Predicted vs observed on held-out genes, and the same separation the AUC bridge summarises.
# Gated on `auc_bridge` because panel (b) reads `br` and `bridge`, which exist only if the
# auc_bridge chunk above ran. Without the gate a run with auc_bridge = FALSE dies here with
# "object 'br' not found" - an infrastructure crash that a mutation test would misread as a
# check having fired.
p_fit <- tibble(obs = test_data$y, pred = pred_t) %>%
  ggplot(aes(obs, pred)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.18, size = 0.7, colour = "#E69F00") +
  geom_smooth(method = "lm", se = FALSE, colour = "#0072B2", linewidth = 0.8) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = sprintf("Spearman = %.3f\nPearson = %.3f\nn = %d",
                           spearman_raw, pearson_raw, nrow(test_data))) +
  # Derived, not typed. .dir_ratio is built from params in the direction_facts chunk; hardcoding
  # "si3d" here put the wrong knockdown on every si3e figure, and the sibling legend below put
  # the wrong CONDITION on the normoxia figure. Same failure as the directional prose: a label
  # that does not come from the data will eventually disagree with it.
  labs(x = paste0("observed TE LFC, ", .dir_ratio), y = "predicted",
       title = "(a) Held-out fit",
       subtitle = sprintf("%d features | transform: %s", length(feature_cols),
                          params$target_transform)) +
  theme_classic(base_size = 11)

# The legend name for the positive class, built from params rather than written out. CLAUDE.md's
# group_colors block uses the literal "eIF3d hypoxia promotes"; copying that string verbatim is
# what put "hypoxia" on the normoxia figure.
pos_label_txt <- sprintf("%s %s %s", .dir_gene, params$condition, params$direction)
p_sep <- br %>% filter(!is.na(lab_geneset)) %>%
  mutate(cls = if_else(lab_geneset == 1, pos_label_txt, "Negative control")) %>%
  ggplot(aes(pred, fill = cls, colour = cls)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_fill_manual(values = setNames(c("#E69F00", "black"),
                                      c(pos_label_txt, "Negative control")), name = NULL) +
  scale_colour_manual(values = setNames(c("#E69F00", "black"),
                                        c(pos_label_txt, "Negative control")), name = NULL) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("bridge AUC = %.3f\nceiling = %.3f\nn = %d / %d",
                           bridge$auc[1], bridge$ceiling[1], bridge$n_pos[1], bridge$n_neg[1])) +
  labs(x = "predicted TE LFC", y = "density", title = "(b) AUC bridge",
       subtitle = "the regression never saw these labels") +
  theme_classic(base_size = 11) +
  theme(legend.position = c(0.5, 0.96))

pg <- cowplot::plot_grid(p_fit, p_sep, nrow = 1, rel_widths = c(1, 1.15))
print(pg)
ggsave_if(here("plots", paste0("10_fit_", suffix, ".pdf")), pg, width = 10, height = 4.6)
```

# Comparison with the classifier

Four views of how this regression relates to `02_rf_model.Rmd`'s classifier. Everything here
reads saved artifacts and refits nothing, and each part skips cleanly if its inputs are absent.

This is the summary view. `12_regression_vs_classifier.Rmd` carries the full treatment, including
the measured noise floor that decides which differences are real.

**Decisions**

- Compare the classifier's predictions only on genes it never trained on. It modelled 2,872 of
  the 10,172, so most of this notebook's held-out set is genuinely unseen by it.
- Compare importance and SHAP as shares, never as raw values. The two models' SHAP is in
  different units.
- Plot the shrinkage explicitly, since it is the main thing that makes the held-out fit above
  look worse than the ranking actually is.

```{r fig_roc_and_calibration, fig.width = 11, fig.height = 4.6, eval = isTRUE(params$auc_bridge)}
# Gated on `auc_bridge` for the same reason as fig_fit: panel (a) is built from `br`, which the
# auc_bridge chunk creates. Panel (b) needs only `pred_t`, but the two panels share one
# cowplot grid, so the chunk is gated as a unit.
#
# (a) The bridge as an ROC curve rather than two densities, with the ceiling curve drawn. The
# ceiling uses the true outcome as the predictor: for a label that is a threshold on that
# outcome it is perfect by construction, which is why the AUC is never reported on its own.
roc_geneset <- roc(br$lab_geneset[!is.na(br$lab_geneset)],
                   br$pred[!is.na(br$lab_geneset)], direction = "<", quiet = TRUE)
roc_ceiling <- roc(br$lab_geneset[!is.na(br$lab_geneset)],
                   br$y[!is.na(br$lab_geneset)], direction = "<", quiet = TRUE)
roc_thresh  <- roc(br$lab_thresh, br$pred, direction = "<", quiet = TRUE)

roc_df <- bind_rows(
  tibble(fpr = 1 - roc_geneset$specificities, tpr = roc_geneset$sensitivities,
         curve = sprintf("predicted TE vs classifier gene sets (AUC %.3f)",
                         as.numeric(auc(roc_geneset)))),
  tibble(fpr = 1 - roc_thresh$specificities, tpr = roc_thresh$sensitivities,
         curve = sprintf("predicted TE vs a threshold on the outcome (AUC %.3f)",
                         as.numeric(auc(roc_thresh)))),
  tibble(fpr = 1 - roc_ceiling$specificities, tpr = roc_ceiling$sensitivities,
         curve = sprintf("ceiling: TRUE outcome as the predictor (AUC %.3f)",
                         as.numeric(auc(roc_ceiling)))))

p_roc <- ggplot(roc_df, aes(fpr, tpr, colour = curve)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_path(linewidth = 0.9) + coord_equal() +
  scale_colour_manual(values = c("#E69F00", "#0072B2", "grey45"), name = NULL) +
  labs(x = "false positive rate", y = "true positive rate", title = "(a) The AUC bridge as a ROC",
       subtitle = sprintf("held-out genes only | n = %d positives, %d negatives",
                          sum(br$lab_geneset == 1, na.rm = TRUE),
                          sum(br$lab_geneset == 0, na.rm = TRUE))) +
  theme_classic(base_size = 11) +
  theme(legend.position = c(0.98, 0.02), legend.justification = c(1, 0),
        legend.text = element_text(size = 7.5))

# (b) Why the held-out fit looks flat. A forest prediction is an average over leaf values, so
# predictions shrink toward the training mean. Binning the held-out genes by predicted value and
# plotting the mean observed value in each bin shows the ranking is monotone even though the
# magnitudes are compressed.
calib <- tibble(pred = pred_t, obs = test_data$y) %>%
  mutate(bin = ntile(pred, 10)) %>% group_by(bin) %>%
  summarise(pred_mid = mean(pred), obs_mean = mean(obs), obs_se = sd(obs) / sqrt(n()),
            n = n(), .groups = "drop")
sd_ratio <- sd(pred_t) / sd(test_data$y)
p_cal <- ggplot(calib, aes(pred_mid, obs_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbar(aes(ymin = obs_mean - obs_se, ymax = obs_mean + obs_se),
                width = 0, colour = "grey60") +
  geom_point(size = 2.6, colour = "#E69F00") +
  geom_line(colour = "#E69F00", linewidth = 0.6) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.3,
           label = sprintf("prediction sd / outcome sd = %.3f\n(predictions compressed %.1fx)\nn = %d in 10 bins",
                           sd_ratio, 1 / sd_ratio, nrow(test_data))) +
  labs(x = "mean predicted TE LFC in bin", y = "mean observed TE LFC in bin",
       title = "(b) Ranking is monotone; magnitudes are shrunk",
       subtitle = "decile bins of held-out genes. Dashed line is y = x") +
  theme_classic(base_size = 11)

pg_roc <- cowplot::plot_grid(p_roc, p_cal, nrow = 1, rel_widths = c(1.1, 1))
print(pg_roc)
ggsave_if(here("plots", paste0("10_roc_calibration_", suffix, ".pdf")), pg_roc,
          width = 11, height = 4.6)
```

```{r fig_vs_classifier, fig.width = 11, fig.height = 4.8}
clf_model_path <- file.path(out_dir, paste0("rf_classifier_", clf_suffix, ".rds"))
clf_md_path2   <- file.path(out_dir, paste0("rf_model_data_", clf_suffix, ".rds"))
clf_imp_path2  <- file.path(out_dir, paste0("rf_importance_", clf_suffix, ".csv"))
have_clf <- all(file.exists(c(clf_model_path, clf_md_path2, clf_imp_path2)))

if (!have_clf) {
  cat("Classifier artifacts not found; this comparison needs 02_rf_model.Rmd to have been knit.\n")
} else {
  clf      <- readRDS(clf_model_path)
  clf_md2  <- readRDS(clf_md_path2)
  # The classifier can only score these genes if it was fitted on the same columns in the same
  # order, so that is checked - but a mismatch is a MISSING INPUT for this comparison, not a
  # corrupt run, and the section header promises each part skips cleanly when its inputs are
  # absent. It was a stopifnot, which aborted the whole render: that made every arm that
  # deliberately changes the feature set unrunnable, including family_arms_reg()'s `none` arm
  # (exclude_families drops 9 columns for g4, 5 for peptide) and so both family presets.
  # Reported rather than silently dropped, because an UNintended mismatch looks the same.
  same_features <- identical(clf_md2$feature_cols, feature_cols)
}
if (have_clf && !same_features) {
  cat(sprintf(paste0("Panels (c) and (d) SKIPPED: the classifier was fitted on %d features and ",
                     "this run uses %d, so it cannot score these genes.\n"),
              length(clf_md2$feature_cols), length(feature_cols)))
  only_clf <- setdiff(clf_md2$feature_cols, feature_cols)
  only_reg <- setdiff(feature_cols, clf_md2$feature_cols)
  if (length(only_clf)) cat("  only the classifier has:", paste(only_clf, collapse = ", "), "\n")
  if (length(only_reg)) cat("  only this run has:      ", paste(only_reg, collapse = ", "), "\n")
  if (!length(only_clf) && !length(only_reg))
    cat("  the same columns in a DIFFERENT ORDER, which is a different forest\n")
  cat("  Expected when the arm excludes a family; investigate if it was not meant to.\n")
}
if (have_clf && same_features) {
  # Only genes the classifier never trained on. It modelled 2,872 of 10,172, so most of this
  # notebook's held-out set is unseen by it and the comparison is out-of-sample on both sides.
  unseen <- !(test_data$gene_id_clean %in% clf_md2$train$gene_id_clean)
  clf_p  <- predict(clf, test_data[unseen, feature_cols])$predictions[, clf_md2$pos_label]
  cmp_df <- tibble(reg_pred = pred_t[unseen], clf_prob = clf_p,
                   obs = test_data$y[unseen],
                   in_pos = test_data$gene_id_clean[unseen] %in% pos_genes)
  rho_pred <- cor(cmp_df$reg_pred, cmp_df$clf_prob, method = "spearman")

  p_pp <- ggplot(cmp_df, aes(clf_prob, reg_pred)) +
    geom_point(aes(colour = in_pos), size = 0.8, alpha = 0.35) +
    geom_smooth(method = "lm", se = FALSE, colour = "#0072B2", linewidth = 0.8) +
    scale_colour_manual(values = c(`TRUE` = "#E69F00", `FALSE` = "grey55"),
                        labels = c(`TRUE` = "in the positive set", `FALSE` = "not"), name = NULL) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.3,
             label = sprintf("Spearman = %.3f\nn = %d genes\nneither model trained on these",
                             rho_pred, nrow(cmp_df))) +
    labs(x = "classifier predicted probability", y = "regression predicted TE LFC",
         title = "(c) Do the two models order genes the same way?",
         subtitle = "genes held out by this model and never seen by the classifier") +
    theme_classic(base_size = 11) + theme(legend.position = "bottom")

  # (d) Impurity importance as shares. Shares rather than raw values, because the two forests
  # split on different quantities and their raw impurity is not on one scale.
  clf_imp2 <- read_csv(clf_imp_path2, show_col_types = FALSE) %>%
    mutate(clf_share = 100 * importance / sum(importance)) %>%
    dplyr::select(feature, family, clf_share)
  imp_cmp <- importance_df %>% dplyr::select(feature, reg_share = share_pct) %>%
    inner_join(clf_imp2, by = "feature")
  rho_imp <- cor(imp_cmp$clf_share, imp_cmp$reg_share, method = "spearman")

  p_imp <- ggplot(imp_cmp, aes(clf_share, reg_share)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_point(aes(colour = family), size = 2, alpha = 0.85) +
    scale_colour_viridis_d(guide = "none", end = 0.9) +
    ggrepel::geom_text_repel(
      data = imp_cmp %>% filter(pmax(clf_share, reg_share) > 2.5 |
                                abs(clf_share - reg_share) > 1.5),
      aes(label = feature), size = 2.5, max.overlaps = 18, seed = 9) +
    annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
             label = sprintf("Spearman = %.3f\nn = %d features", rho_imp, nrow(imp_cmp))) +
    labs(x = "classifier impurity share (%)", y = "regression impurity share (%)",
         title = "(d) Feature importance, both models",
         subtitle = "shares, since raw impurity is not on a common scale") +
    theme_classic(base_size = 11)

  pg_clf <- cowplot::plot_grid(p_pp, p_imp, nrow = 1, rel_widths = c(1, 1.05))
  print(pg_clf)
  ggsave_if(here("plots", paste0("10_vs_classifier_", suffix, ".pdf")), pg_clf,
            width = 11, height = 4.8)
}
```

```{r fig_shap_compare, fig.width = 11, fig.height = 4.8}
# SHAP is computed in 11_shap_regression.Rmd, not here, so this reads its output if present. That
# keeps this notebook cheap: TreeSHAP on this forest takes far longer than fitting it.
shap_reg_path <- file.path(out_dir, paste0("shapreg_family_", suffix, ".csv"))
shap_clf_path <- file.path(out_dir, paste0("shap_family_", clf_suffix, ".csv"))
feat_reg_path <- file.path(out_dir, paste0("shapreg_feature_", suffix, ".csv"))
feat_clf_path <- file.path(out_dir, paste0("shap_feature_", clf_suffix, ".csv"))
have_shap <- all(file.exists(c(shap_reg_path, shap_clf_path, feat_reg_path, feat_clf_path)))

if (!have_shap) {
  cat("SHAP outputs not found. Run 11_shap_regression.Rmd (and 03_shap.Rmd for the classifier),\n")
  cat("then re-knit this notebook to fill in these panels.\n")
} else {
  fam_r <- read_csv(shap_reg_path, show_col_types = FALSE) %>%
    dplyr::select(family, reg = share_pct)
  fam_c <- read_csv(shap_clf_path, show_col_types = FALSE) %>%
    dplyr::select(family, clf = share_pct)
  fam_j <- inner_join(fam_c, fam_r, by = "family") %>% mutate(delta = reg - clf)

  p_fam <- fam_j %>%
    pivot_longer(c(clf, reg), names_to = "model", values_to = "share") %>%
    mutate(model = recode(model, clf = "classifier", reg = "regression"),
           family = fct_reorder(family, share, .fun = max)) %>%
    ggplot(aes(share, family, fill = model)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.65, alpha = 0.9) +
    scale_fill_manual(values = c(classifier = "#0072B2", regression = "#E69F00"), name = NULL) +
    labs(x = "share of total |SHAP| (%)", y = NULL,
         title = "(e) Family SHAP shares, both models",
         subtitle = sprintf("Spearman %.3f over %d families",
                            cor(fam_j$clf, fam_j$reg, method = "spearman"), nrow(fam_j))) +
    theme_classic(base_size = 11) + theme(legend.position = "bottom")

  ft_r <- read_csv(feat_reg_path, show_col_types = FALSE) %>%
    mutate(reg = 100 * mean_abs_shap / sum(mean_abs_shap)) %>%
    dplyr::select(feature, family, reg, reg_cor = shap_value_cor)
  ft_c <- read_csv(feat_clf_path, show_col_types = FALSE) %>%
    mutate(clf = 100 * mean_abs_shap / sum(mean_abs_shap)) %>%
    dplyr::select(feature, clf, clf_cor = shap_value_cor)
  ft_j <- inner_join(ft_c, ft_r, by = "feature")

  p_ft <- ggplot(ft_j, aes(clf, reg)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_point(aes(colour = family), size = 2, alpha = 0.85) +
    scale_colour_viridis_d(guide = "none", end = 0.9) +
    ggrepel::geom_text_repel(
      data = ft_j %>% filter(pmax(clf, reg) > 3 | abs(clf - reg) > 1.5),
      aes(label = feature), size = 2.5, max.overlaps = 18, seed = 9) +
    annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
             label = sprintf("Spearman = %.3f\nn = %d features",
                             cor(ft_j$clf, ft_j$reg, method = "spearman"), nrow(ft_j))) +
    labs(x = "classifier SHAP share (%)", y = "regression SHAP share (%)",
         title = "(f) Feature SHAP shares",
         subtitle = "shares only: the two models' SHAP is in different units") +
    theme_classic(base_size = 11)

  pg_shap <- cowplot::plot_grid(p_fam, p_ft, nrow = 1, rel_widths = c(1, 1.05))
  print(pg_shap)
  ggsave_if(here("plots", paste0("10_shap_vs_classifier_", suffix, ".pdf")), pg_shap,
            width = 11, height = 4.8)

  cat("\nFamilies where the two models differ most:\n")
  print(as.data.frame(fam_j %>% arrange(desc(abs(delta))) %>% head(6) %>%
    mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
  cat("\nWhether any of these clears replication noise is decided in 12_, not here.\n")
}
```

# Export

Writes the model, its importance, its metrics, its predictions, and the split itself.

**Decisions**

- Write nothing during a mutation run, so a deliberately corrupted run cannot be mistaken for a
  real result.
- Route every path through `assert_no_collision()`.
- Save the imputed and transformed split alongside the model, so `11_` explains this model's
  actual data instead of rebuilding it from parameters. The equivalent notebook in the frozen
  pipeline rebuilds it and then has to assert its way back to agreement.


```{r export, eval = !nzchar(params$mutate_target)}
# A mutation run never writes, so a corrupted matrix cannot be mistaken for a real result.
# Every path goes through assert_no_collision(), so no regression output can land on a
# classification filename and destroy a frozen baseline.
reg_save_rds(rf_model, file.path(out_dir, paste0("rfreg_model_", suffix, ".rds")))
reg_write_csv(importance_df, file.path(out_dir, paste0("rfreg_importance_", suffix, ".csv")))

perf <- bind_rows(
  metrics,
  if (isTRUE(params$run_cv)) tibble(
    metric = c("cv_spearman_mean", "cv_spearman_sd", "cv_pearson_mean", "cv_rmse_mean"),
    value  = c(mean(cv$spearman), sd(cv$spearman), mean(cv$pearson), mean(cv$rmse))) else NULL,
  # Written as a tribble so each metric NAME is bound to its value on the same row. The previous
  # two-parallel-vector form silently desynced - 10 names against 12 values - and the run failed
  # with "Tibble columns must have compatible sizes" rather than writing a wrong number, which
  # was lucky. A tribble cannot desync.
  #
  # Two AUCs, named for their comparator, because that is the whole difference between them.
  # Both define positives identically: outcome > the lfc cut AND padj < 0.05, on held-out genes.
  #   auc                     positives vs every OTHER held-out gene. Self-contained: needs no
  #                           external gene set, so it is comparable across knockdowns and
  #                           timepoints. This is the one reported.
  #   auc_vs_screened_ctrl    positives vs notebook 02's screened negative controls - genes
  #                           selected for having NO effect. That pool is not a neutral sample:
  #                           it is GC-enriched by +2.45 (Wilcoxon p = 2.7e-17), which inflates
  #                           the number (0.806 against 0.721 at si3d hypoxia 1hr). Kept only
  #                           for continuity with 02_; do not compare it across contrasts.
  if (isTRUE(params$auc_bridge)) tibble::tribble(
    ~metric,                    ~value,
    "auc",                       bridge$auc[2],
    "auc_ceiling",               bridge$ceiling[2],
    "auc_n_pos",                 bridge$n_pos[2],
    "auc_n_neg",                 bridge$n_neg[2],
    "auc_vs_screened_ctrl",      bridge$auc[1],
    "auc_vs_screened_ceiling",   bridge$ceiling[1],
    "auc_vs_screened_n_pos",     bridge$n_pos[1],
    "auc_vs_screened_n_neg",     bridge$n_neg[1],
    "matched_gene_auc",          matched_auc,
    "n_matched_genes",           as.numeric(n_matched)
  ) else NULL,
  if (exists("perm_metrics")) perm_metrics else NULL,
  tibble(metric = "split_seed", value = SPLIT_SEED),
  tibble(metric = "forest_seed", value = FOREST_SEED)
)
reg_write_csv(perf, file.path(out_dir, paste0("rfreg_performance_", suffix, ".csv")))

reg_write_csv(
  bind_rows(
    train_data %>% dplyr::select(gene_id_clean, symbol, y, y_padj) %>%
      mutate(pred = pred_tr, split = "train"),
    test_data %>% dplyr::select(gene_id_clean, symbol, y, y_padj) %>%
      mutate(pred = pred_t,  split = "test")),
  file.path(out_dir, paste0("rfreg_predictions_", suffix, ".csv")))

# The imputed, transformed split itself, so 11_shap_regression.Rmd explains this model's data
# rather than re-deriving it from params. 07c_ in the frozen pipeline re-derives and has to
# assert its way back to agreement; saving removes that whole class of drift.
if (isTRUE(params$save_model_data)) {
  reg_save_rds(list(train = train_data, test = test_data, feature_cols = feature_cols,
                    suffix = suffix, clf_suffix = clf_suffix,
                    target_transform = params$target_transform,
                    winsorize_y = params$winsorize_y, inv_scale = tf$inv_scale,
                    split_seed = SPLIT_SEED, forest_seed = FOREST_SEED),
               file.path(out_dir, paste0("rfreg_model_data_", suffix, ".rds")))
}
cat("Wrote outputs with suffix:", suffix, "\n")
```
