---
title: "`r paste0('Regression split-seed sweep: ', params$preset)`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  preset: "transform"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10c_regression_sweep_stability.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It reports one sweep from `10b_regression_split_sweep.R`: the same model refit across 20 different
train/test splits, for each arm of a preset. The point is to show how much a single run's number
moves when nothing about the model or the data changes, only which genes were held out.

**Decisions**

- Vary `split_seed` only. In `02_`'s classifier the equivalent axis was `neg_seed`, which changed
  which negatives were drawn; there is no draw here, so the axis is the split.
- Keep the forest seed fixed, so within a split every arm sees identical rows and arms can be
  compared pairwise.
- Judge arms on paired differences, whose spread is much smaller than the split-to-split spread.
  An effect too small to see in a distribution comparison can be clear when paired.
- Do not reuse the classifier's floor for regression quantities. Two 80/20 splits share about 81%
  of their training rows, where a negative draw shared about 52% of genes, so this floor should be
  tighter and is measured separately.

```{r libraries}
library(tidyverse)
library(here)
library(cowplot)
```

```{r load}
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
source(here("code", "translational_control_overview", "10_regression_presets.R"))
out_dir <- here("output", "translational_control_overview")

stopifnot("unknown preset" = params$preset %in% names(REG_PRESETS))
arms <- REG_PRESETS[[params$preset]]
cat("Preset:", params$preset, "| arms:", paste(names(arms), collapse = ", "),
    "| seeds:", length(REG_SPLIT_SEEDS), "\n")

read_perf <- function(arm, seed) {
  f <- file.path(out_dir, paste0("rfreg_performance_", reg_suffix(arm, seed), ".csv"))
  if (!file.exists(f)) return(NULL)
  d <- read_csv(f, show_col_types = FALSE)
  tibble(split_seed = seed,
         spearman  = d$value[d$metric == "spearman_raw_y"],
         pearson   = d$value[d$metric == "pearson_raw_y"],
         r2        = d$value[d$metric == "r2_pearson_sq_raw_y"],
         oob_r2    = d$value[d$metric == "oob_r2"],
         bridge    = d$value[d$metric == "bridge_auc_geneset"],
         n_feat    = d$value[d$metric == "n_features"])
}
sw <- imap_dfr(arms, function(a, nm)
  map_dfr(REG_SPLIT_SEEDS, ~ read_perf(a, .x)) %>% mutate(arm = nm))
stopifnot("no sweep outputs found - run 10b_regression_split_sweep.R first" = nrow(sw) > 0)
sw$arm <- factor(sw$arm, levels = names(arms))

present <- sw %>% count(arm, name = "n_seeds")
print(as.data.frame(present), row.names = FALSE)
if (any(present$n_seeds < length(REG_SPLIT_SEEDS)))
  cat("WARNING: incomplete arms. Paired statistics use only seeds present in Every arm.\n")
# Pair on the intersection, so a missing render cannot turn a paired test into an unpaired one.
complete_seeds <- sw %>% count(split_seed) %>% filter(n == length(arms)) %>% pull(split_seed)
cat("Seeds complete across all arms:", length(complete_seeds), "\n")
```

# 1. How much one run's number moves

Reports each arm's spread across the 20 splits.

**Decisions**

- Report the full range alongside the standard deviation, since the range is what a single
  unlucky or lucky run could have reported.
- Assert the feature count is identical across the seeds within each arm, so an arm's runs differ
  only by the split. Between arms it may legitimately differ, because an `exclude_families` arm
  drops the family under test.


```{r spread}
spread <- sw %>% group_by(arm) %>%
  summarise(n = n(),
            across(c(spearman, bridge, oob_r2),
                   list(median = median, sd = sd,
                        p05 = ~ quantile(.x, 0.05), p95 = ~ quantile(.x, 0.95))),
            .groups = "drop")
print(as.data.frame(spread %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)

cat("\nThe number a SINGLE run would have reported, per arm:\n")
for (a in levels(sw$arm)) {
  s <- sw %>% filter(arm == a)
  cat(sprintf("  %-8s Spearman range %.4f to %.4f (spread %.4f = %.1fx its own sd)\n",
              a, min(s$spearman), max(s$spearman), diff(range(s$spearman)),
              diff(range(s$spearman)) / sd(s$spearman)))
}
# The feature set must not move across the SEEDS WITHIN an arm: if it does, that arm's runs differ
# by more than the split and its paired differences are confounded.
#
# Between arms a difference is expected, not a fault. family_arms_reg()'s `none` arm passes
# exclude_families, so it deliberately drops the family under test - 9 columns for the g4 preset,
# 5 for peptide - while `real` and `shuf` keep all of them (shuffling permutes values, never drops
# a column). Asserting one global feature count therefore made this report unproducible for 2 of
# the 5 presets, which is the opposite of what the header promises.
feat_by_arm <- sw %>% group_by(arm) %>%
  summarise(n_runs = n(), n_feat = paste(sort(unique(n_feat)), collapse = ", "),
            n_distinct_feat = n_distinct(n_feat), .groups = "drop")
print(as.data.frame(feat_by_arm), row.names = FALSE)
stopifnot("the feature count varies across seeds within an arm - those runs differ by more than the split" =
            all(feat_by_arm$n_distinct_feat == 1))
if (nrow(feat_by_arm) > 1 && n_distinct(feat_by_arm$n_feat) > 1)
  cat("\nFeature counts differ BETWEEN arms, which is expected when an arm excludes a family.\n")
cat("\nFeature count constant within every arm.\n")
```

# 2. Paired differences

Compares each arm against the preset's first arm, pairing within split seed.

**Decisions**

- Pair on the intersection of seeds present in every arm, so a failed render cannot quietly turn a
  paired test into an unpaired one.
- Report the ratio of the paired spread to the unpaired spread, which is what shows why pairing
  was necessary.


```{r paired}
# Everything is compared against the first arm, which is the preset's reference by construction.
ref <- levels(sw$arm)[1]
wide <- sw %>% filter(split_seed %in% complete_seeds) %>%
  dplyr::select(arm, split_seed, spearman) %>%
  pivot_wider(names_from = arm, values_from = spearman)

pair_tbl <- map_dfr(setdiff(levels(sw$arm), ref), function(a) {
  d <- wide[[a]] - wide[[ref]]
  tibble(comparison = paste(a, "-", ref), n = length(d),
         mean_delta = mean(d), sd_delta = sd(d),
         n_positive = sum(d > 0),
         p_wilcox = tryCatch(wilcox.test(d)$p.value, error = function(e) NA_real_),
         # The honest scale: a paired difference is readable only against ITS OWN sd, not
         # against the much larger seed-to-seed sd.
         ratio_to_unpaired_sd = sd(d) / sd(sw$spearman[sw$arm == ref]))
})
print(as.data.frame(pair_tbl %>% mutate(across(where(is.numeric), ~ signif(., 4)))),
      row.names = FALSE)
cat(sprintf("\nseed-to-seed sd of the reference arm: %.4f\n", sd(sw$spearman[sw$arm == ref])))
cat("A paired difference far below that sd can still be real; an UNPAIRED comparison at this\n")
cat("sample size could not resolve it. That is the whole reason the sweep is paired.\n")
```

```{r fig_sweep, fig.width = 11, fig.height = 4.6}
p_pair <- ggplot(sw, aes(arm, spearman)) +
  geom_line(data = sw %>% filter(split_seed %in% complete_seeds),
            aes(group = split_seed), colour = "grey78", linewidth = 0.4) +
  geom_point(aes(colour = arm), size = 2, alpha = 0.85) +
  stat_summary(fun = median, geom = "crossbar", width = 0.45, linewidth = 0.4, colour = "black") +
  scale_colour_viridis_d(guide = "none", end = 0.85) +
  labs(x = NULL, y = "Spearman vs raw held-out y",
       title = "(a) Every arm on one yardstick",
       subtitle = sprintf("grey lines join the same split_seed | %d seeds x %d arms",
                          length(complete_seeds), length(arms))) +
  theme_classic(base_size = 12)

pd <- map_dfr(setdiff(levels(sw$arm), ref), function(a)
  tibble(comparison = paste(a, "-", ref), delta = wide[[a]] - wide[[ref]]))
p_delta <- ggplot(pd, aes(comparison, delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey92") +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.8, colour = "#E69F00") +
  geom_text(data = pair_tbl, aes(comparison, Inf,
            label = sprintf("mean %+.4f\nsd %.4f\n%d/%d > 0\np = %.3g",
                            mean_delta, sd_delta, n_positive, n, p_wilcox)),
            vjust = 1.2, size = 2.9) +
  labs(x = NULL, y = paste("paired difference in Spearman vs", ref),
       title = "(b) Paired differences against their own sd",
       subtitle = "zero is the null; the box is what the draw alone produces") +
  theme_classic(base_size = 12)

p_bridge <- ggplot(sw, aes(arm, bridge)) +
  geom_line(data = sw %>% filter(split_seed %in% complete_seeds),
            aes(group = split_seed), colour = "grey78", linewidth = 0.4) +
  geom_point(aes(colour = arm), size = 2, alpha = 0.85) +
  stat_summary(fun = median, geom = "crossbar", width = 0.45, linewidth = 0.4, colour = "black") +
  scale_colour_viridis_d(guide = "none", end = 0.85) +
  labs(x = NULL, y = "bridge AUC vs the classifier's gene sets",
       title = "(c) The AUC bridge, swept",
       subtitle = "02_'s single reported test AUC was 0.8306") +
  geom_hline(yintercept = 0.8306, linetype = "dotted", colour = "#0072B2") +
  theme_classic(base_size = 12)

pg <- plot_grid(p_pair, p_delta, p_bridge, nrow = 1, rel_widths = c(1, 1, 1))
print(pg)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("10c_sweep_", params$preset, ".pdf")), pg, width = 11, height = 4.6)
```

# 3. Family shares across the sweep

If SHAP was run for these arms, reports how much each family's share moves across splits.

**Decisions**

- Use `2.8 * sd` as the minimum detectable difference, matching how the classifier's floor was
  built so the two can be combined later.
- Skip cleanly and print the command to run when SHAP outputs are absent.


```{r family_shares}
# Only present if 11d_ was run for these arms. Family share is the cross-experiment quantity, so
# its spread over the sweep IS the floor a reported share difference has to clear.
fam <- imap_dfr(arms, function(a, nm)
  map_dfr(REG_SPLIT_SEEDS, function(s) {
    f <- file.path(out_dir, paste0("shapreg_family_", reg_suffix(a, s), ".csv"))
    if (!file.exists(f)) return(NULL)
    read_csv(f, show_col_types = FALSE) %>%
      dplyr::select(family, share_pct) %>% mutate(arm = nm, split_seed = s)
  }))
have_shap <- nrow(fam) > 0
if (!have_shap) {
  cat("No shapreg_family_* files for this preset.\n")
  cat("Run: Rscript code/translational_control_overview/11d_shap_regression_seed_sweep.R", params$preset, "\n")
} else {
  floor_tbl <- fam %>% group_by(arm, family) %>%
    summarise(n = n(), median = median(share_pct), sd = sd(share_pct),
              p05 = quantile(share_pct, 0.05), p95 = quantile(share_pct, 0.95),
              # 2.8 sd, matching shap_noise_floor.csv's construction so the two are comparable.
              min_detectable_diff = 2.8 * sd, .groups = "drop") %>%
    arrange(arm, desc(median))
  print(as.data.frame(floor_tbl %>% mutate(across(where(is.numeric), ~ round(., 3)))),
        row.names = FALSE)
}
```

```{r fig_family, eval = have_shap, fig.width = 9, fig.height = 5}
p_fam <- fam %>%
  mutate(family = fct_reorder(family, share_pct, .fun = median)) %>%
  ggplot(aes(share_pct, family, colour = arm)) +
  geom_point(position = position_dodge(width = 0.5), size = 1.5, alpha = 0.6) +
  scale_colour_viridis_d(end = 0.85, name = NULL) +
  labs(x = "family share of total |SHAP| (%)", y = NULL,
       title = "Family shares across split seeds",
       subtitle = sprintf("one point per seed. The SPREAD is the floor, not the median. n = %d seeds",
                          n_distinct(fam$split_seed))) +
  theme_classic(base_size = 12)
print(p_fam)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("10c_family_shares_", params$preset, ".pdf")), p_fam,
         width = 9, height = 5)
```

```{r export}
reg_write_csv(sw, file.path(out_dir, paste0("rfreg_sweep_", params$preset, ".csv")))
reg_write_csv(pair_tbl, file.path(out_dir, paste0("rfreg_sweep_paired_", params$preset, ".csv")))
if (have_shap)
  reg_write_csv(floor_tbl, file.path(out_dir, paste0("shapreg_noise_floor_", params$preset, ".csv")))
cat("Wrote sweep report for preset:", params$preset, "\n")
```
