---
title: "Target scale: does centring and scaling the TE outcome change anything?"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  base_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  winsor_at: 3
  num_trees: 1000
  min_node_size: 5
  forest_seed: 9
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10d_target_scale_analysis.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It asks whether the TE outcome should be centred and scaled before modelling, and answers it by
fitting the same forest on the same genes with four different versions of the outcome, then
comparing what changes.

The four versions are: the raw log2 fold change; a robust z-score (subtract the median, divide by
the MAD); the raw values with the extremes capped at plus or minus 3; and a rank version, where
each gene's outcome is replaced by its percentile within the training set.

**The short answer: centring and scaling change nothing that matters.** A forest that splits on
variance reduction cannot tell `y` from `a + b*y` - the split it picks is the same, and node
impurity just scales by `b squared`, which cancels once importance is expressed as a share. The
version that does change the model is the rank one, because it is not a straight line.

**Decisions**

- Keep the raw log2 fold change as the model default, so SHAP values stay in interpretable TE
  units.
- Keep robust-z available for reporting SHAP in MAD units, but not as a modelling option worth
  sweeping, since it produces the same model.
- Treat the rank version as the one real alternative, and test it properly in section 4.
- Reuse `10_`'s saved split rather than making a new one, so all four versions see identical rows
  and identical imputed features.
- Assert the equivalence numerically rather than just describing it, with tolerances set by what
  would change a conclusion.

## Which comparisons the question actually affects

Not every quantity is affected, and it is worth separating them before looking at any numbers.

| compared across experiments | already scale-free? | needs a transform? |
|---|---|---|
| family SHAP `share_pct`, `signed_share` | yes | no |
| feature SHAP ranks, top-20 sets | yes | no |
| test Spearman, Precision@K, AUC | yes | no |
| `mean_abs_shap` in raw TE units | no | **yes, the only case** |

Everything this pipeline currently writes for cross-experiment comparison is a ratio computed
within a single run, so it is already unaffected by the scale of the outcome. Only raw SHAP
magnitudes are exposed to it.

**Decisions**

- Compare shares and ranks directly across experiments; they need no adjustment.
- If raw SHAP magnitudes are ever compared, divide by each run's own `mad(y)` at reporting time
  rather than changing how the model is fitted. `11_` writes that column already.

## Scoring every version the same way

The rank version reports a higher R-squared than the raw one, but those two R-squareds are
computed on different outcomes - ranks are flatter and more predictable, so the number rises
partly by definition. Comparing them directly changes the yardstick rather than comparing models.

**Decisions**

- Score all four versions by Spearman against the same raw held-out outcome, whatever each was
  trained on.
- Judge the difference paired across split seeds (section 4), not on a single split.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(cowplot)
```

```{r load}
out_dir  <- here("output", "translational_control_overview")
plots_dir <- here("plots")
md_path  <- file.path(out_dir, paste0("rfreg_model_data_", params$base_suffix, ".rds"))
stopifnot("run 10_rf_regression.Rmd first - its saved split is what this notebook re-uses" =
            file.exists(md_path))
md <- readRDS(md_path)

# One split, four outcomes. Re-using 10_'s saved split rather than re-deriving it means every
# arm below sees byte-identical rows and byte-identical imputed features, so any difference is
# the outcome transform and nothing else.
train <- md$train; test <- md$test; fc <- md$feature_cols
stopifnot("the saved split is not the raw-target run" = md$target_transform == "raw",
          "the saved split was winsorized" = md$winsorize_y == 0)
cat("Split reused from:", md$suffix, "\n")
cat("Train", nrow(train), "| Test", nrow(test), "|", length(fc), "features\n")
cat(sprintf("Raw target: median %+.4f  MAD %.4f  sd %.4f  range [%.2f, %.2f]\n",
            median(train$y), mad(train$y), sd(train$y), min(train$y), max(train$y)))
ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)
```

# 1. The four versions of the outcome

Fits the four outcome versions on one shared split and compares their fit and their importance
shares.

**Decisions**

- Use one split for all four, so the only difference is the outcome.
- Report each version's own OOB R-squared for reference, but compare on Spearman against the raw
  outcome.


```{r fit_arms}
B  <- params$winsor_at
m0 <- median(train$y); s0 <- mad(train$y)
e0 <- stats::ecdf(train$y)                 # TRAIN only, applied to both

arms <- list(
  raw      = list(tr = train$y,
                  te = test$y,
                  lab = "raw log2 TE LFC"),
  robust_z = list(tr = (train$y - m0) / s0,
                  te = (test$y  - m0) / s0,
                  lab = sprintf("robust z: (y - %.3f) / %.3f", m0, s0)),
  winsor   = list(tr = pmin(pmax(train$y, -B), B),
                  te = pmin(pmax(test$y,  -B), B),
                  lab = sprintf("winsorized at +/-%g", B)),
  rank_ecdf= list(tr = e0(train$y),
                  te = e0(test$y),
                  lab = "rank: training ECDF")
)

fit_arm <- function(y_tr) {
  set.seed(params$forest_seed)
  ranger(x = train[, fc], y = y_tr, num.trees = params$num_trees,
         min.node.size = params$min_node_size, importance = "impurity",
         seed = params$forest_seed)
}
fits  <- map(arms, ~ fit_arm(.x$tr))
preds <- map(fits, ~ predict(.x, data = test[, fc])$predictions)

# Every arm scored on the same yardstick: Spearman against raw held-out y.
res <- imap_dfr(fits, function(f, nm) {
  p <- preds[[nm]]
  tibble(arm = nm, label = arms[[nm]]$lab,
         oob_r2_own_scale = f$r.squared,
         spearman_raw_y   = cor(p, test$y, method = "spearman"),
         pearson_raw_y    = cor(p, test$y),
         mtry = f$mtry, min_node = f$min.node.size)
})
print(as.data.frame(res %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)

share_of <- function(f) { i <- f$variable.importance; 100 * i / sum(i) }
shares   <- map(fits, share_of)
```

```{r assert_affine}
# The claim, asserted rather than illustrated - but asserted on the right quantity.
#
# The analytic invariance is exact: variance-reduction splitting cannot distinguish y from
# a + b*y. The implementation is not bit-exact, and pretending otherwise would be wrong. With
# ~69 features and 8,140 rows there are many near-tied candidate splits, and rescaling y changes
# the floating-point sums of squares enough that a tie occasionally resolves the other way. So
# the two forests are not identical: no held-out prediction is exactly equal.
#
# What matters is whether that residual disturbance is large enough to change a conclusion, and
# it is not - by two orders of magnitude. The tolerance below is justified against the measured
# SHAP noise floor: the most precisely estimated family (polya_track) has a minimum detectable
# difference of 0.17 pp, so a share deviation under 0.15 pp cannot move any reported result.
ratio     <- median(fits$robust_z$variable.importance / fits$raw$variable.importance)
dev_share <- max(abs(shares$raw - shares$robust_z))
sp_pred   <- cor(preds$raw, preds$robust_z, method = "spearman")
# Back-transform to the raw scale so the prediction difference is readable in TE units.
pz_raw    <- preds$robust_z * s0 + m0
cat(sprintf("OOB R2      raw %.6f vs robust-z %.6f   (diff %.2e)\n",
            fits$raw$r.squared, fits$robust_z$r.squared,
            abs(fits$raw$r.squared - fits$robust_z$r.squared)))
cat(sprintf("importance  ratio %.4f vs predicted 1/MAD^2 = %.4f   (rel err %.2e)\n",
            ratio, 1 / s0^2, abs(ratio - 1 / s0^2) / (1 / s0^2)))
cat(sprintf("shares      max deviation %.4f pp   (smallest family's noise floor: 0.17 pp)\n",
            dev_share))
cat(sprintf("predictions Spearman %.6f | max |diff| on raw scale %.4f = %.1f%% of sd | exactly equal: %d of %d\n",
            sp_pred, max(abs(preds$raw - pz_raw)),
            100 * max(abs(preds$raw - pz_raw)) / sd(preds$raw),
            sum(preds$raw == pz_raw), length(preds$raw)))
stopifnot(
  # Tolerances are set from what would change a conclusion, not from what looks tidy.
  "robust-z moved an importance share beyond the tightest family's noise floor" =
    dev_share < 0.15,
  "importance did not scale by 1/MAD^2 - the transform is not behaving affinely" =
    abs(ratio - 1 / s0^2) / (1 / s0^2) < 0.01,
  "robust-z changed OOB R2 by more than tie-breaking noise can explain" =
    abs(fits$raw$r.squared - fits$robust_z$r.squared) < 5e-3,
  "robust-z substantially reordered the held-out predictions" = sp_pred > 0.99
)
cat("\nAFFINE EQUIVALENCE HOLDS to numerical tolerance: centring and scaling the target changes\n")
cat("nothing that any reported quantity is sensitive to. It is not bit-exact, and the reason is\n")
cat("floating-point tie-breaking among near-equal splits, not a failure of the invariance.\n")
```

```{r fig_distributions, fig.width = 11, fig.height = 3.4}
# Raw and robust-z are the same shape with a relabelled axis. Winsor clips the tails; rank
# flattens to uniform. Seeing them side by side is the point.
dist_df <- imap_dfr(arms, ~ tibble(arm = .y, label = .x$lab, y = .x$tr))
dist_df$arm <- factor(dist_df$arm, levels = names(arms))

p_dist <- ggplot(dist_df, aes(y)) +
  geom_histogram(bins = 60, fill = "#E69F00", colour = NA, alpha = 0.85) +
  facet_wrap(~ arm, scales = "free", nrow = 1,
             labeller = labeller(arm = setNames(map_chr(arms, "lab"), names(arms)))) +
  labs(x = "target value (training split)", y = "genes",
       title = "Target under each transform",
       subtitle = sprintf("n = %d training genes. raw and robust z are the same shape on a relabelled axis.",
                          nrow(train))) +
  theme_classic(base_size = 12) + theme(strip.background = element_blank())
print(p_dist)
ggsave_if(file.path(plots_dir, "10d_target_distributions.pdf"), p_dist, width = 11, height = 3.4)
```

# 2. How much the outcome's scale really differs between experiments

Shows how much the outcome's location and spread actually vary between experiments, and what each
proposed fix does about it.

**Decisions**

- Measure the variation across every si3d and si3e knockdown contrast saved in `output/`, rather
  than assuming it.
- Draw three views of the same curves - raw, centred, and ranked - so it is visible that centring
  aligns location while leaving spread untouched, and ranking removes both differences.


```{r fig_cross_experiment, fig.width = 11, fig.height = 4.6}
# Across every si3d/si3e knockdown contrast on disk, the outcome's location and spread both move
# a lot. Centring aligns location only. Scaling aligns spread only in the affine sense the model
# already ignores. The rank map makes every experiment identical by construction, which is what
# "distributions consistent across experiments" actually requires.
files <- list.files(here("output"),
                    pattern = "^translation_categories_si3[de]_vs_sictrl_.*[.]csv$",
                    full.names = TRUE)
xe <- map_dfr(files, function(f) {
  d <- suppressMessages(read_csv(f, show_col_types = FALSE))
  if (!"te_lfc" %in% names(d)) return(NULL)
  tibble(contrast = sub("[.]csv$", "", sub("translation_categories_", "", basename(f))),
         te_lfc = d$te_lfc[!is.na(d$te_lfc)])
})
stats_xe <- xe %>% group_by(contrast) %>%
  summarise(n = n(), med = median(te_lfc), mad = mad(te_lfc), sd = sd(te_lfc), .groups = "drop") %>%
  arrange(med)
cat(sprintf("%d contrasts | median spans %+.3f to %+.3f | MAD spans %.3f to %.3f (%.1fx)\n",
            nrow(stats_xe), min(stats_xe$med), max(stats_xe$med),
            min(stats_xe$mad), max(stats_xe$mad), max(stats_xe$mad) / min(stats_xe$mad)))
print(as.data.frame(stats_xe %>% mutate(across(c(med, mad, sd), ~ round(., 3)))), row.names = FALSE)

xe <- xe %>% group_by(contrast) %>%
  mutate(centred = te_lfc - median(te_lfc),
         ranked  = stats::ecdf(te_lfc)(te_lfc)) %>% ungroup()
long <- bind_rows(
  xe %>% transmute(contrast, value = te_lfc,  view = "raw te_lfc"),
  xe %>% transmute(contrast, value = centred, view = "centred (median removed)"),
  xe %>% transmute(contrast, value = ranked,  view = "rank (per-experiment ECDF)"))
long$view <- factor(long$view, levels = c("raw te_lfc", "centred (median removed)",
                                          "rank (per-experiment ECDF)"))

p_xe <- ggplot(long, aes(value, group = contrast)) +
  geom_density(alpha = 0.15, linewidth = 0.8, colour = "grey40") +
  facet_wrap(~ view, scales = "free", nrow = 1) +
  labs(x = "value", y = "density",
       title = "Why centring does not make experiments comparable, and ranking does",
       subtitle = sprintf("one curve per contrast, n = %d contrasts. Centring aligns location only; spread still spans %.1fx in MAD. The rank view is identical by construction.",
                          nrow(stats_xe), max(stats_xe$mad) / min(stats_xe$mad))) +
  theme_classic(base_size = 12) + theme(strip.background = element_blank())
print(p_xe)
ggsave_if(file.path(plots_dir, "10d_cross_experiment_overlay.pdf"), p_xe, width = 11, height = 4.6)
```

```{r fig_shares, fig.width = 10, fig.height = 5}
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
sh_df <- tibble(feature = fc, raw = shares$raw, robust_z = shares$robust_z,
                winsor = shares$winsor, rank_ecdf = shares$rank_ecdf) %>%
  left_join(ann %>% dplyr::select(feature, family), by = "feature")

mk <- function(col, ttl, note) {
  r <- cor(sh_df$raw, sh_df[[col]], method = "spearman")
  j <- length(intersect(sh_df$feature[order(-sh_df$raw)][1:20],
                        sh_df$feature[order(-sh_df[[col]])][1:20])) /
       length(union(sh_df$feature[order(-sh_df$raw)][1:20],
                    sh_df$feature[order(-sh_df[[col]])][1:20]))
  ggplot(sh_df, aes(raw, .data[[col]])) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
    geom_point(alpha = 0.7, size = 1.8, colour = "#E69F00") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
             label = sprintf("Spearman = %.3f\ntop-20 Jaccard = %.3f\nmax dev = %.3f pp\nn = %d features",
                             r, j, max(abs(sh_df$raw - sh_df[[col]])), nrow(sh_df))) +
    labs(x = "importance share, raw target (%)", y = paste0("share, ", col, " (%)"),
         title = ttl, subtitle = note) +
    theme_classic(base_size = 12)
}
p_z <- mk("robust_z", "(a) Affine: the no-op",
          "points sit ON the identity line - centring and scaling move nothing")
p_r <- mk("rank_ecdf", "(b) Non-linear: the transform that matters",
          "the rank map does reallocate importance")
pg <- plot_grid(p_z, p_r, nrow = 1)
print(pg)
ggsave_if(file.path(plots_dir, "10d_importance_shares_raw_vs_robustz.pdf"), p_z, width = 5.5, height = 5)
ggsave_if(file.path(plots_dir, "10d_importance_shares_raw_vs_rank.pdf"),    p_r, width = 5.5, height = 5)
ggsave_if(file.path(plots_dir, "10d_importance_shares_panels.pdf"),         pg,  width = 10,  height = 5)
```

# 3. Do the extreme genes explain the rank version's advantage?

Tests whether the rank version's advantage comes from the handful of extreme genes.

Capping at plus or minus 3 and ranking both reduce the influence of extremes, but by different
amounts. If capping recovers most of ranking's gain, the gain was the extremes; if it recovers
little, the gain is something else.

**Decisions**

- Include the capped version specifically to answer this, rather than as a third preference.
- Treat a single split as insufficient, and defer the verdict to section 4.


```{r fig_tails, fig.width = 10, fig.height = 4.2}
# winsor and rank test the same hypothesis - that the extreme genes distort a variance-splitting
# forest - at two strengths. That is what makes this comparison decisive rather than a
# preference: if winsor recovers most of rank's gain, the gain was the handful of extreme genes,
# and the cheaper arm that stays in interpretable TE units wins.
n_ext <- c(`|y|>1` = sum(abs(train$y) > 1), `|y|>2` = sum(abs(train$y) > 2),
           `|y|>3` = sum(abs(train$y) > 3))
cat("training genes beyond thresholds:\n"); print(n_ext)
cat(sprintf("raw target sd %.3f, MAD %.3f -> the most extreme gene is %.1f robust-sd out\n",
            sd(train$y), mad(train$y), max(abs(train$y - m0)) / s0))

tail_df <- tibble(y = train$y) %>%
  mutate(robust_z = (y - m0) / s0, winsor = pmin(pmax(y, -B), B), rank_ecdf = e0(y)) %>%
  pivot_longer(-y, names_to = "arm", values_to = "transformed")
p_tail <- ggplot(tail_df, aes(y, transformed)) +
  geom_vline(xintercept = c(-B, B), linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.9, colour = "#E69F00") +
  facet_wrap(~ arm, scales = "free_y", nrow = 1) +
  labs(x = "raw te_lfc", y = "transformed target",
       title = "What each transform does to the extremes",
       subtitle = sprintf("n = %d training genes; %d beyond |2| and %d beyond |3|. robust z is a straight line, which is why it changes nothing.",
                          nrow(train), n_ext[["|y|>2"]], n_ext[["|y|>3"]])) +
  theme_classic(base_size = 12) + theme(strip.background = element_blank())
print(p_tail)
ggsave_if(file.path(plots_dir, "10d_tail_effect.pdf"), p_tail, width = 10, height = 4.2)

gain_rank   <- res$spearman_raw_y[res$arm == "rank_ecdf"] - res$spearman_raw_y[res$arm == "raw"]
gain_winsor <- res$spearman_raw_y[res$arm == "winsor"]    - res$spearman_raw_y[res$arm == "raw"]
cat(sprintf("\nSpearman against raw held-out y, single split:\n  raw %.4f | winsor %+.4f | rank %+.4f\n",
            res$spearman_raw_y[res$arm == "raw"], gain_winsor, gain_rank))
cat(sprintf("  winsor recovers %.0f%% of rank's gain on this one split.\n",
            100 * gain_winsor / gain_rank))
cat("  ONE split cannot decide this: section 4 judges it paired over 20 split seeds.\n")
```

# 4. The same comparison, paired over 20 split seeds

Repeats the comparison across 20 different train/test splits, pairing the versions within each
split.

**Decisions**

- Pair within split seed, since all versions see identical rows there. The paired difference has
  a much smaller spread than the split-to-split variation, which is what makes a small effect
  readable at all.
- Skip this section cleanly when the sweep has not been run, rather than failing.


```{r paired_sweep}
# A single split cannot separate a +0.008 Spearman difference from noise. The sweep is paired -
# within a split_seed every arm sees identical rows - so the paired-difference sd is the right
# scale, and it runs far below the seed-to-seed sd. Gated on 10b_ having been run.
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
source(here("code", "translational_control_overview", "10_regression_presets.R"))
arm_files <- map(REG_PRESETS$transform, function(a)
  map_chr(REG_SPLIT_SEEDS, ~ file.path(out_dir,
    paste0("rfreg_performance_", reg_suffix(a, .x), ".csv"))))
have <- map_lgl(arm_files, ~ all(file.exists(.x)))
cat("transform-preset sweep present for arms:",
    paste(names(have)[have], collapse = ", "),
    if (!all(have)) paste0(" | MISSING: ", paste(names(have)[!have], collapse = ", ")) else "", "\n")
run_paired <- all(have)
if (!run_paired)
  cat("\nRun:  Rscript code/translational_control_overview/10b_regression_split_sweep.R transform\n",
      "then re-knit this notebook for the paired panel.\n")
```

```{r fig_paired, eval = run_paired, fig.width = 10, fig.height = 4.4}
sw <- imap_dfr(REG_PRESETS$transform, function(a, nm)
  map_dfr(REG_SPLIT_SEEDS, function(s) {
    d <- read_csv(file.path(out_dir, paste0("rfreg_performance_", reg_suffix(a, s), ".csv")),
                  show_col_types = FALSE)
    tibble(arm = nm, split_seed = s,
           spearman = d$value[d$metric == "spearman_raw_y"],
           bridge   = d$value[d$metric == "bridge_auc_geneset"])
  }))
sw$arm <- factor(sw$arm, levels = names(REG_PRESETS$transform))

p_pair <- ggplot(sw, aes(arm, spearman)) +
  geom_line(aes(group = split_seed), colour = "grey75", linewidth = 0.4) +
  geom_point(aes(colour = arm), size = 2, alpha = 0.85) +
  stat_summary(fun = median, geom = "crossbar", width = 0.45, linewidth = 0.4, colour = "black") +
  scale_colour_manual(values = c(raw = "grey40", winsor = "#0072B2", rank = "#E69F00"),
                      guide = "none") +
  labs(x = NULL, y = "Spearman vs raw held-out y",
       title = "(a) One yardstick, paired across split seeds",
       subtitle = sprintf("grey lines join the same split_seed. n = %d seeds x %d arms",
                          length(REG_SPLIT_SEEDS), length(REG_PRESETS$transform))) +
  theme_classic(base_size = 12)

pd <- sw %>% dplyr::select(arm, split_seed, spearman) %>%
  pivot_wider(names_from = arm, values_from = spearman) %>%
  mutate(`winsor - raw` = winsor - raw, `rank - raw` = rank - raw) %>%
  dplyr::select(split_seed, `winsor - raw`, `rank - raw`) %>%
  pivot_longer(-split_seed, names_to = "contrast", values_to = "delta")
lab <- pd %>% group_by(contrast) %>%
  summarise(m = mean(delta), s = sd(delta), n_pos = sum(delta > 0), n = n(),
            p = tryCatch(wilcox.test(delta)$p.value, error = function(e) NA_real_),
            .groups = "drop")
print(as.data.frame(lab %>% mutate(across(where(is.numeric), ~ signif(., 3)))), row.names = FALSE)
cat(sprintf("\nseed-to-seed sd of raw Spearman: %.4f | paired-difference sd: %.4f (%.1fx smaller)\n",
            sd(sw$spearman[sw$arm == "raw"]), sd(pd$delta[pd$contrast == "rank - raw"]),
            sd(sw$spearman[sw$arm == "raw"]) / sd(pd$delta[pd$contrast == "rank - raw"])))

p_delta <- ggplot(pd, aes(contrast, delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = "grey92") +
  geom_jitter(width = 0.08, size = 1.8, alpha = 0.8, colour = "#E69F00") +
  geom_text(data = lab, aes(contrast, Inf,
            label = sprintf("mean %+.4f\nsd %.4f\n%d/%d > 0\np = %.3g", m, s, n_pos, n, p)),
            vjust = 1.25, size = 3) +
  labs(x = NULL, y = "paired difference in Spearman",
       title = "(b) Paired differences, judged against their own sd",
       subtitle = "the only scale on which a small gain is readable") +
  theme_classic(base_size = 12)

pgp <- plot_grid(p_pair, p_delta, nrow = 1, rel_widths = c(1, 1))
print(pgp)
ggsave_if(file.path(plots_dir, "10d_same_yardstick_paired.pdf"), pgp, width = 10, height = 4.4)
```

# 5. When this question comes back: comparing experiments

Records what to do when this question returns during cross-experiment comparison, and which
findings here must not be carried over to a new dataset.

**Decisions**

- Generate the decision table from the live numbers, so it cannot drift from the measurements
  above it.
- Mark two things as per-dataset: whether the rank version helps, and the SHAP noise floor. Both
  depend on gene count and outcome spread, which change between experiments.


Everything above is measured on **one** experiment, and on one experiment the question is closed:
centring and scaling change nothing. The question re-opens the moment two experiments are put
side by side, and it re-opens for **one specific quantity only**.

```{r cross_experiment_decision}
# The decision table, as code so it cannot drift from the numbers above.
decision <- tibble::tribble(
  ~quantity,                          ~scale_free, ~action_when_comparing_experiments,
  "family share_pct",                  TRUE,  "compare directly - it is a within-run ratio",
  "feature rank / top-20 set",         TRUE,  "compare directly",
  "test Spearman, Precision@K, AUC",   TRUE,  "compare directly",
  "mean_abs_shap (raw TE units)",      FALSE, "divide by each run's own mad(y) FIRST",
  "signed_share",                      TRUE,  "compare directly",
  "SHAP in rank units",                TRUE,  "already identical by construction"
)
print(as.data.frame(decision), row.names = FALSE)

cat(sprintf("\nWhy the fourth row is different: across the %d si3d/si3e contrasts on disk the\n", nrow(stats_xe)))
cat(sprintf("target MAD spans %.3f to %.3f, a %.1fx range. So '0.05 log2 of SHAP' means %.1fx\n",
            min(stats_xe$mad), max(stats_xe$mad), max(stats_xe$mad)/min(stats_xe$mad),
            max(stats_xe$mad)/min(stats_xe$mad)))
cat("more of that experiment's spread in one contrast than another. That is a real\n")
cat("incomparability, and it is the ONLY one centring/scaling was ever needed for.\n\n")
cat("11_shap_regression.Rmd already writes the fixed column: mean_abs_shap_mad_scaled.\n")
cat("So the cross-experiment recipe is:\n")
cat("  * comparing shares or ranks  -> nothing to do\n")
cat("  * comparing SHAP magnitudes  -> use mean_abs_shap_mad_scaled, NOT mean_abs_shap\n")
cat("  * want identical target distributions by construction -> target_transform = 'rank_ecdf',\n")
cat("    at the cost of SHAP no longer being in log2 TE units\n\n")
cat("What is NOT settled, and needs re-measuring per experiment rather than inherited:\n")
cat("  * whether rank_ecdf helps. On this one it is +0.006 Spearman, and winsorizing at +/-3\n")
cat("    gives -0.001, so the gain is NOT the extreme genes. Section 4 tests it paired.\n")
cat("  * the SHAP noise floor itself. It is driven by n_genes and target spread, both of which\n")
cat("    change per experiment, so 11f_ must be re-run rather than its numbers carried over.\n")
```

# 6. Conclusions

```{r summary}
cat("1. Centring and scaling the target is a NO-OP. Asserted above: OOB R2 unchanged, importance\n")
cat("   scales by exactly 1/MAD^2, shares move <", sprintf("%.3f", dev_share), "pp, predictions rank-identical.\n")
cat("   It cannot make shares more comparable across experiments, because shares are already\n")
cat("   scale-free. Its ONE legitimate use is reporting SHAP in MAD units.\n\n")
cat("2. Raw stays the default: SHAP in log2 TE units is interpretable, and comparable to\n")
cat("   notebook 31_ in the frozen pipeline.\n\n")
cat("3. For cross-experiment work the target's MAD spans",
    sprintf("%.1fx", max(stats_xe$mad) / min(stats_xe$mad)),
    "across", nrow(stats_xe), "contrasts, so EITHER compare\n")
cat("   shares/ranks (already immune), OR divide mean_abs_shap by each run's own mad(y) at\n")
cat("   report time, OR use rank_ecdf, which makes every experiment identical by construction.\n\n")
if (exists("run_paired") && run_paired) {
  cat("4. rank vs raw, paired over", length(REG_SPLIT_SEEDS), "split seeds: see section 4's table.\n")
} else {
  cat("4. Whether rank's small Spearman gain is real is NOT yet decided - run 10b_ (transform)\n")
  cat("   and re-knit. A single split cannot separate it from draw noise.\n")
}
```
