---
title: "Herrmannova HeLa outcome prep: eIF3d / eIF3e TE, oriented control over knockdown"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  xlsx: "accessories/herrmannova elife riboseq/elife-95846-supp1-v1.xlsx"
  matrix_rds: "feature_matrix_tco.rds"
  lfc: "0.5"
  neg_lfc: "0.5"
  neg_padj: "0.3"
  min_xcell_rho: 0.05
  min_xcell_sep: 0.03
  write_outputs: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10h_herrmannova_outcome_prep.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

Turns Herrmannova et al. (*eLife* 2024, 95846) ribosome profiling of eIF3d and eIF3e knockdowns in
**HeLa** into outcome files that `10_rf_regression.Rmd` can model exactly like an MDA-MB-231 or
MCF7-SIX1 contrast, so the two knockdowns can become columns in the SHAP heatmap.

The paper reports log2(knockdown / non-targeting). Every value is **negated once, here**, so the
written `te_lfc` / `rna_lfc` are log2(control / knockdown) - the orientation of our own
MDA-MB-231 files - and positive means the factor **promotes**. eIF3h is in the source but not
carried forward.

**Decisions**

- Verify the table against the paper before using it: the published DTEG counts, split by
  direction, and the paper's own padj < 0.05 sheets must be reproduced exactly.
- Keep only genes DESeq2 actually tested (non-NA padj). Genes removed by independent filtering
  have an unshrunk, low-count fold change that is mostly noise.
- Pin the sign three independent ways: each factor's own mRNA, the paper's down/up split, and a
  rank correlation with the matched MDA-MB-231 normoxia contrast. The last is the only one built
  on data outside this file, and it is shown to fail on the un-negated values.
- Gene-level only. The paper reports no transcripts, so the model will use MDA-MB-231's dominant
  isoform for each gene; HeLa's isoforms are unknown.
- Name the outputs by the project's convention (`translation_categories_si{g}_vs_sictrl_*`),
  including its known wart: the filename states the ratio backwards, the contents are
  control over knockdown. `condition = hela_normoxia`, `timepoint = steadystate` (a knockdown
  at steady state, not a hypoxia time course).

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
```

```{r paths}
XLSX <- here(params$xlsx)
OUT  <- here("output")
stopifnot("Herrmannova supplement not found" = file.exists(XLSX))
KD <- c(d = "3d", e = "3e")                       # sheet letter -> project gene code
lfc_cut <- as.numeric(params$lfc)
```

# 1. Load

The `all data` sheet carries four legend rows above the header; the `TE_*` sheets carry one.

```{r load}
all_data <- suppressWarnings(read_excel(XLSX, sheet = "all data", skip = 4))
num_cols <- grep("log2FoldChange$|padj$", names(all_data), value = TRUE)
all_data <- all_data %>% mutate(across(all_of(num_cols), ~ suppressWarnings(as.numeric(.x))))
stopifnot(
  "all data is not 17,520 genes - check the skip" = nrow(all_data) == 17520,
  "all data is not 22 columns"                    = ncol(all_data) == 22,
  "gene_id is not unique"                         = !any(duplicated(all_data$gene_id)),
  "gene_id is versioned - expected bare ENSG"     = !any(grepl("\\.", all_data$gene_id))
)
col <- function(kind, k, what = "log2FoldChange") paste0(kind, "_", k, "_vs_nt.", what)
cat("all data:", nrow(all_data), "genes\n")
```

# 2. Reproduce the paper

## 2a. TE is footprints minus mRNA

If this fails, the TE column being read is not TE.

```{r te_identity}
te_id <- map_dfr(names(KD), function(k) {
  te <- all_data[[col("TE", k)]]; fp <- all_data[[col("FP", k)]]; mr <- all_data[[col("mRNA", k)]]
  ok <- is.finite(te) & is.finite(fp) & is.finite(mr)
  tibble(knockdown = paste0("eIF", KD[[k]]), n = sum(ok), max_abs_diff = max(abs(te - (fp - mr))[ok]))
})
knitr::kable(te_id)
stopifnot("TE column is not FP - mRNA: wrong column" = all(te_id$max_abs_diff < 1e-6))
```

## 2b. The published DTEG counts, split by direction

Herrmannova et al. report the number of differentially translated genes (TE padj < 0.05) per
knockdown, and how many go down vs up. These are the paper's numbers, typed as the expected
values of a replication check; the data are read from the file.

```{r dteg_counts}
PAPER_DTEG <- tribble(
  ~k,  ~paper_total, ~paper_down, ~paper_up,
  "d",        1971L,       1040L,      931L,
  "e",        1180L,        622L,      558L
)
obs <- map_dfr(names(KD), function(k) {
  lf <- all_data[[col("TE", k)]]; pa <- all_data[[col("TE", k, "padj")]]
  sig <- !is.na(pa) & pa < 0.05 & !is.na(lf)
  tibble(k = k, obs_total = sum(sig), obs_down = sum(sig & lf < 0), obs_up = sum(sig & lf > 0))
})
dteg <- left_join(PAPER_DTEG, obs, by = "k") %>% mutate(knockdown = paste0("eIF", KD[k]), .before = 1)
knitr::kable(dteg %>% dplyr::select(-k))
stopifnot("DTEG totals do not match the paper" = all(dteg$paper_total == dteg$obs_total),
          "DTEG down counts do not match the paper" = all(dteg$paper_down == dteg$obs_down),
          "DTEG up counts do not match the paper" = all(dteg$paper_up == dteg$obs_up))
```

## 2c. The paper's own filtered sheets are reproducible from `all data`

```{r sheets}
for (k in names(KD)) {
  sh <- suppressWarnings(read_excel(XLSX, sheet = paste0("TE_", k, " padj 0.05"), skip = 1))
  mine <- all_data$gene_id[!is.na(all_data[[col("TE", k, "padj")]]) &
                             all_data[[col("TE", k, "padj")]] < 0.05]
  stopifnot(setNames(list(setequal(mine, sh$gene_id)),
                     paste0("TE_", k, " sheet is not reproducible from 'all data'"))[[1]])
}
cat("TE_d and TE_e padj < 0.05 sheets reproduced exactly from 'all data'\n")
```

# 3. Direction

## 3a. Ruler one: each factor's own mRNA

The siRNA destroys its own target, so in the source orientation (knockdown / control) EIF3D must
be strongly negative in the eIF3d knockdown and flat in the eIF3e knockdown, and vice versa.

```{r ruler_own}
own <- map_dfr(names(KD), function(k) {
  other <- setdiff(names(KD), k)
  r <- all_data %>% filter(hgnc_symbol == paste0("EIF", toupper(KD[[k]])))
  stopifnot("knockdown target missing or duplicated" = nrow(r) == 1)
  tibble(target = r$hgnc_symbol, own_kd_mRNA = r[[col("mRNA", k)]],
         other_kd_mRNA = r[[col("mRNA", other)]])
})
knitr::kable(own, digits = 2)
stopifnot("a target is not depleted in its own knockdown" = all(own$own_kd_mRNA < -2),
          "a target moves in the knockdown that does not target it" = all(abs(own$other_kd_mRNA) < 1))
```

## 3b. Negate, and keep the tested genes

```{r negate}
neg <- map(setNames(names(KD), KD), function(k) {
  all_data %>%
    transmute(gene_id, gene_id_clean = gene_id, symbol = hgnc_symbol,
              # THE NEGATION. One place. Source is log2(KD / NT); written is log2(NT / KD).
              te_lfc   = -.data[[col("TE", k)]],   te_padj   = .data[[col("TE", k, "padj")]],
              rna_lfc  = -.data[[col("mRNA", k)]], rna_padj  = .data[[col("mRNA", k, "padj")]],
              ribo_lfc = -.data[[col("FP", k)]],   ribo_padj = .data[[col("FP", k, "padj")]],
              source_te_log2FoldChange = .data[[col("TE", k)]])
})
tested <- map(neg, ~ .x %>% filter(!is.na(te_lfc), !is.na(te_padj)))

tco <- readRDS(here("output", "translational_control_overview", params$matrix_rds))
desc <- imap_dfr(tested, function(d, g) tibble(
  knockdown = paste0("eIF", g), tested = nrow(d),
  untested_dropped = sum(!is.na(neg[[g]]$te_lfc)) - nrow(d),
  in_matrix = sum(d$gene_id_clean %in% tco$gene_id_clean),
  median = median(d$te_lfc), MAD = mad(d$te_lfc),
  promotes = sum(d$te_lfc > lfc_cut & d$te_padj < 0.05)))
knitr::kable(desc, digits = 3)

# After negation the paper's "down" genes are exactly the positive significant ones.
down_now <- map_int(names(KD), function(k) {
  d <- tested[[KD[[k]]]]; sum(d$te_padj < 0.05 & d$te_lfc > 0)
})
stopifnot("after negation, the positive significant genes are not the paper's down genes" =
            all(down_now == PAPER_DTEG$paper_down))
own_after <- map_dbl(KD, function(g) tested[[g]] %>% filter(symbol == paste0("EIF", toupper(g))) %>% pull(rna_lfc))
cat(sprintf("own-target rna_lfc after negation: %s (MDA-MB-231 EIF3D is about +2.9)\n",
            paste(sprintf("EIF%s %+.2f", toupper(KD), own_after), collapse = ", ")))
stopifnot("negated own-target rna_lfc must be strongly positive, as in our files" = all(own_after > 2))
```

## 3c. Ruler two: the matched MDA-MB-231 contrast

eIF3d and eIF3e targets overlap between cell lines, so the negated HeLa TE must correlate
**positively** with the same knockdown in MDA-MB-231 normoxia, and 231-defined targets must score
above 231-defined negative controls on HeLa values. Both are also computed for the un-negated
values, which must fail, so the check is shown to discriminate.

```{r ruler_xcell}
ref231 <- c(`3d` = "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv",
            `3e` = "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv")
pos_of <- function(g) read_csv(here("output", "genesets",
    paste0("hypoxia_", g, "_promotes_TE_1hr_lfc0.5.csv")), show_col_types = FALSE)$ensembl_gene
neg_of <- function(g) read_csv(here("output", "predictive_modeling",
    if (g == "3e") "negative_control_genes_si3e.csv" else "negative_control_genes.csv"),
    show_col_types = FALSE) %>% mutate(gg = sub("[.].*", "", gene_id)) %>% pull(gg)

xc <- map_dfr(KD, function(g) {
  m <- tested[[g]] %>% filter(gene_id_clean %in% tco$gene_id_clean)
  r <- read_csv(file.path(OUT, ref231[[g]]), show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>% dplyr::select(gene_id_clean, ref = te_lfc)
  j <- inner_join(m, r, by = "gene_id_clean") %>% filter(is.finite(ref))
  p <- pos_of(g); n <- setdiff(neg_of(g), p)
  yp <- m$te_lfc[m$gene_id_clean %in% p]; yn <- m$te_lfc[m$gene_id_clean %in% n]
  tibble(knockdown = paste0("eIF", g), ref = ref231[[g]], n_paired = nrow(j),
         rho_negated = cor(j$te_lfc, j$ref, method = "spearman"),
         rho_if_raw = -cor(j$te_lfc, j$ref, method = "spearman"),
         n_pos = length(yp), n_neg = length(yn),
         sep_negated = median(yp) - median(yn), sep_if_raw = -(median(yp) - median(yn)))
})
knitr::kable(xc, digits = 4)
stopifnot(
  "negated HeLa TE does not correlate positively with the matched 231 contrast - sign is wrong" =
    all(xc$rho_negated > params$min_xcell_rho),
  "231-defined targets do not score above 231 negative controls on HeLa values" =
    all(xc$sep_negated > params$min_xcell_sep),
  "the un-negated orientation would also have passed - the ruler does not discriminate" =
    all(xc$rho_if_raw < 0) && all(xc$sep_if_raw < 0)
)
```

# 4. Write

```{r write_outcomes, eval = isTRUE(params$write_outputs)}
dest <- function(g) paste0("translation_categories_si", g, "_vs_sictrl_hela_normoxia_steadystate.csv")
for (g in KD) write_csv(tested[[g]] %>% dplyr::select(-gene_id_clean), file.path(OUT, dest(g)))

# Read back and assert the file on disk is the negation of the source, gene for gene.
chk <- map_dfr(names(KD), function(k) {
  a <- read_csv(file.path(OUT, dest(KD[[k]])), show_col_types = FALSE)
  j <- inner_join(a %>% dplyr::select(gene_id, te_lfc, rna_lfc),
                  all_data %>% dplyr::select(gene_id, s_te = all_of(col("TE", k)),
                                             s_rna = all_of(col("mRNA", k))), by = "gene_id")
  tibble(file = dest(KD[[k]]), n = nrow(j), expected_n = nrow(tested[[KD[[k]]]]),
         max_dev_te = max(abs(j$te_lfc + j$s_te)), max_dev_rna = max(abs(j$rna_lfc + j$s_rna), na.rm = TRUE),
         signs_inverted = all(sign(j$te_lfc) == -sign(j$s_te)))
})
knitr::kable(chk)
stopifnot("a written file lost or gained genes" = all(chk$n == chk$expected_n),
          "a written file is not the negation of its source" =
            all(chk$max_dev_te < 1e-12) && all(chk$max_dev_rna < 1e-12),
          "a written file does not have every sign inverted" = all(chk$signs_inverted))
```

# 5. Gene sets for `10_`'s spot-check and AUC bridge

Same rule as the MDA-MB-231 and MCF7-SIX1 sets: positives `te_lfc > 0.5 & padj < 0.05`,
negative controls `|te_lfc| < 0.5 & padj > 0.3`.

```{r write_genesets, eval = isTRUE(params$write_outputs)}
gs <- map_dfr(KD, function(g) {
  d <- tested[[g]]
  nm <- paste0("hela_normoxia_", g, "_promotes_TE_steadystate_lfc", params$lfc)
  p <- d %>% filter(te_lfc > lfc_cut, te_padj < 0.05) %>%
    transmute(db_gene_symbol = symbol, ensembl_gene = gene_id_clean, te_lfc, te_padj,
              gs_name = nm,
              gs_description = paste0("genes whose translation eIF", g, " promotes in HeLa ",
                "(Herrmannova 2024), te_lfc > ", params$lfc, " & padj < 0.05; te_lfc is ",
                "log2(control / knockdown), negated from the paper's log2(KD / NT)"),
              gs_source_species = "homo sapiens")
  write_csv(p, here("output", "genesets", paste0(nm, ".csv")))
  n <- d %>% filter(abs(te_lfc) < as.numeric(params$neg_lfc), te_padj > as.numeric(params$neg_padj)) %>%
    transmute(gene_id, symbol, te_lfc, te_padj)
  nn <- paste0("hela_normoxia_", g, "_negative_controls_lfc", params$neg_lfc, ".csv")
  write_csv(n, here("output", "genesets", nn))
  tibble(knockdown = paste0("eIF", g), positives = nrow(p), negatives = nrow(n),
         pos_in_matrix = sum(p$ensembl_gene %in% tco$gene_id_clean))
})
knitr::kable(gs)
stopifnot("a positive gene set is implausibly small" = all(gs$positives >= 50),
          "a negative control set is smaller than its positive set" = all(gs$negatives > gs$positives))
```

```{r session}
sessionInfo()
```
