---
title: "MCF7-SIX1 outcome prep: establishing the contrast direction once"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  mcf7_repo: "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia"
  matrix_rds: "feature_matrix_tco.rds"
  lfc: "0.5"
  neg_lfc: "0.5"
  neg_padj: "0.3"
  min_xcell_rho: 0.05
  min_xcell_sep: 0.03
  write_outputs: TRUE
  flip_demo: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10m_mcf7six1_outcome_prep.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

MCF7-SIX1 translation-efficiency results live in a different repository and are computed in the
**opposite orientation** from the MDA-MB-231 results this fork models. This notebook is the one
place where that is corrected: it reads the four MCF7-SIX1 contrasts, proves their orientation
three independent ways, negates them, and writes them out in the same shape and the same
orientation as the 231 files.

Everything downstream then reads a promotes-positive table and needs no flag. That is the point:
a per-run `negate_outcome` switch has to be set correctly in every one of eight run
configurations, and a single omission produces a column whose features are read backwards while
every range check still passes. Here the negation happens once, next to the evidence for it.

**Terms used below**

- **promotes-positive** - the sign convention this project uses everywhere: a positive outcome
  means the knockdown *lowered* the quantity, so the factor *promotes* it. The 231 files are
  already promotes-positive; the MCF7-SIX1 files are not.
- **the ruler** - an independent quantity used to check a sign, rather than the quantity being
  checked. Three different rulers are used here, and they do not share a failure mode.
- **cross-cell-line ruler** - the matched 231 contrast. Genuinely independent data, so it cannot
  be fooled by an error made consistently inside the MCF7-SIX1 files.

**Decisions**

- Negate once, here, and write promotes-positive tables. No downstream notebook negates anything.
- Write them under the same `translation_categories_si{gene}_vs_sictrl_{condition}_{timepoint}`
  stem the 231 files use, with `condition = "mcf7six1_{hypoxia|normoxia}"`, so notebook `10_`
  resolves them from params with no new code path and its contrast anchor still applies.
- Keep the same filename wart the 231 files have: the name reads `si3d_vs_sictrl` and the column
  is log2(sictrl / si3d). Matching the existing convention is worth more than fixing the name in
  one corner of the project.
- Verify orientation from the producing source code, not from the filename and not from a value
  typed into this notebook.
- Refuse to write anything if any orientation check fails.

```{r libraries}
library(tidyverse)
library(here)
```

# 1. The four source contrasts

```{r registry}
MCF7 <- file.path(params$mcf7_repo, "counts")
OUT  <- here("output")

# gene / condition -> (source file, the notebook chunk that produced it, the matched 231 ruler)
registry <- tribble(
  ~gene, ~condition,  ~src,                     ~producer,              ~ref231,
  "3d",  "hypoxia",   "hypoxia_si_ctrl_3d.csv", "5_hypoxicresponse.Rmd",
     "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
  "3d",  "normoxia",  "normoxia_sictrl_3d.csv", "3_normoxia.Rmd",
     "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv",
  "3e",  "hypoxia",   "hypoxia_si_ctrl_3e.csv", "5_hypoxicresponse.Rmd",
     "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv",
  "3e",  "normoxia",  "normoxia_sictrl_3e.csv", "3_normoxia.Rmd",
     "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"
) %>%
  mutate(kd_symbol = paste0("EIF", toupper(gene)),
         other_kd  = paste0("EIF", toupper(if_else(gene == "3d", "3e", "3d"))),
         dest      = paste0("translation_categories_si", gene, "_vs_sictrl_mcf7six1_",
                            condition, "_1hr.csv"),
         label     = paste0("si", gene, " ", condition))

stopifnot("a source file is missing" = all(file.exists(file.path(MCF7, registry$src))),
          "a producing notebook is missing" =
            all(file.exists(file.path(params$mcf7_repo, "code", registry$producer))),
          "a matched 231 ruler is missing" = all(file.exists(file.path(OUT, registry$ref231))))
knitr::kable(registry %>% dplyr::select(label, src, producer, dest))
```

# 2. Ruler one: orientation parsed from the producing code

The orientation is a property of how DESeq2 was called, so it is read out of the call rather than
inferred from the output. For each contrast this finds the chunk that writes the file and checks
four things: `treatment` is releveled so **siCTRL is the reference**, `library` is releveled so
**RNA is the reference**, the design is the interaction model, and the coefficient extracted is
the interaction term.

With siCTRL and RNA as the references, that coefficient is
log2 TE(knockdown) - log2 TE(siCTRL) = **log2(KD / siCTRL)** - the opposite of the 231
convention. That is what makes negation necessary, and it is the only reason.

**Decisions**

- Parse the four facts rather than transcribing the conclusion. If the upstream analysis is ever
  re-run with different reference levels, this check fires instead of silently inverting
  every result in the heatmap.
- Require all four. Any one of them alone is compatible with more than one orientation.

```{r parse_orientation}
parse_orientation <- function(producer, src) {
  lines <- readLines(file.path(params$mcf7_repo, "code", producer), warn = FALSE)
  w <- grep(paste0("write_csv\\(.*", src, "\"\\)"), lines, fixed = FALSE)
  stopifnot("could not find the write_csv call that produces this file" = length(w) >= 1)
  w <- w[length(w)]
  h <- grep("^## ", lines[seq_len(w)])                 # nearest preceding section heading
  stopifnot("could not find the enclosing section" = length(h) >= 1)
  blk <- lines[h[length(h)]:w]

  trt  <- grep("treatment\\s*<-\\s*relevel", blk, value = TRUE)
  lib  <- grep("library\\s*<-\\s*relevel",   blk, value = TRUE)
  des  <- grep("design\\s*=\\s*~",           blk, value = TRUE)
  coef <- grep("results\\(.*resultsNames\\(.*\\)\\[[0-9]+\\]", blk, value = TRUE)

  ref_of <- function(x) str_match(x[1], 'ref\\s*=\\s*"([^"]+)"')[, 2]
  list(
    trt_ref   = ref_of(trt),
    lib_ref   = ref_of(lib),
    design    = str_squish(str_replace(des[1], ".*design\\s*=\\s*", "")),
    coef_idx  = as.integer(str_match(coef[1], "\\[([0-9]+)\\]")[, 2]),
    shrunk    = any(grepl("lfcShrink", blk))
  )
}

ori <- pmap_dfr(list(registry$producer, registry$src, registry$label), function(p, s, l) {
  o <- parse_orientation(p, s)
  tibble(label = l, treatment_ref = o$trt_ref, library_ref = o$lib_ref,
         design = o$design, coef_index = o$coef_idx, lfcShrink = o$shrunk)
})
knitr::kable(ori)

stopifnot(
  "treatment is not releveled to siCTRL - the orientation assumed here no longer holds" =
    all(ori$treatment_ref == "sictrl"),
  "library is not releveled to RNA - the interaction term is not a TE contrast" =
    all(ori$library_ref == "rna"),
  "the design is not the treatment-by-library interaction model" =
    all(str_detect(ori$design, "treatment\\s*\\+\\s*library\\s*\\+\\s*treatment:library")),
  "the extracted coefficient is not the 4th, which is the interaction term" =
    all(ori$coef_index == 4L)
)
cat("\nRULER 1 PASSED. With treatment ref = sictrl and library ref = rna, coefficient 4 of\n",
    "~treatment + library + treatment:library is log2 TE(KD) - log2 TE(sictrl) = log2(KD/sictrl).\n",
    "The 231 files are log2(sictrl/KD). Therefore: NEGATE.\n", sep = "")
```

# 3. Load, negate, and describe

```{r load_negate}
raw <- map(seq_len(nrow(registry)), function(i) {
  read_csv(file.path(MCF7, registry$src[i]), show_col_types = FALSE)
})
names(raw) <- registry$label

stopifnot("a source file lacks an expected column" =
            all(map_lgl(raw, ~ all(c("gene_id", "symbol", "log2FoldChange", "padj",
                                     "baseMean", "lfcSE") %in% names(.x)))),
          "a source file has duplicated gene ids" =
            all(map_lgl(raw, ~ !any(duplicated(sub("[.].*", "", .x$gene_id))))))

# THE NEGATION. One line, one place.
neg <- imap(raw, function(d, l) {
  d %>% mutate(gene_id_clean = sub("[.].*", "", gene_id),
               te_lfc  = -log2FoldChange,      # promotes-positive, matching 231
               te_padj = padj)
})

desc <- imap_dfr(neg, function(d, l) tibble(
  contrast = l, n = nrow(d),
  median = median(d$te_lfc, na.rm = TRUE), MAD = mad(d$te_lfc, na.rm = TRUE),
  min = min(d$te_lfc, na.rm = TRUE), max = max(d$te_lfc, na.rm = TRUE),
  sig = sum(d$te_padj < 0.05, na.rm = TRUE)))
knitr::kable(desc, digits = 4)

cat("\nThe medians sit at zero, unlike the 231 contrasts (+0.11 to +0.52). The 231 libraries are\n",
    "spike-in normalised and the knockdown lowers total mRNA per cell, which shifts their TE\n",
    "median up; these are not spike-in normalised. It is a location difference, and a\n",
    "variance-splitting forest is invariant to location and scale, so it does not affect SHAP\n",
    "shares or ranks (measured in 10d_target_scale_analysis.Rmd). It does mean the raw MAD\n",
    "differs between cell lines, so raw SHAP magnitudes are not comparable across them - shares\n",
    "are, and shares are what the heatmap shows.\n", sep = "")
```

# 4. Ruler two: the depleted factor's own TE, with specificity

Each file should show its *own* target moving, and the other factor moving less. This is a weak
ruler on its own - it says the right experiment was loaded rather than which way it points - but
it is the one that would catch two files having been swapped.

**Decisions**

- Check specificity, not just magnitude. "EIF3D is significant" is true in most of these files;
  "EIF3D moves more than EIF3E in the si3d file" is the statement with content.
- Read this ruler in the *raw* orientation, before negation, so it describes the source file.

```{r ruler_kd}
kd <- pmap_dfr(list(seq_len(nrow(registry))), function(i) {
  d <- raw[[registry$label[i]]]
  own_row <- d %>% filter(symbol == registry$kd_symbol[i])
  oth_row <- d %>% filter(symbol == registry$other_kd[i])
  stopifnot("a knockdown target gene is absent from its own file" = nrow(own_row) == 1,
            "the other factor is absent from the file" = nrow(oth_row) >= 1)
  # The row objects are named own_row/oth_row rather than own/other because tibble() evaluates
  # its arguments with data masking, so a column named `other` shadows a variable of that name
  # for every argument after it.
  tibble(contrast = registry$label[i], depleted = registry$kd_symbol[i],
         raw_lfc = own_row$log2FoldChange, padj = own_row$padj,
         other = registry$other_kd[i], other_raw_lfc = oth_row$log2FoldChange[1],
         margin = own_row$log2FoldChange - oth_row$log2FoldChange[1])
})
knitr::kable(kd, digits = 4)

stopifnot(
  "the depleted factor's own raw TE is not clearly positive - the file may be oriented the 231 way" =
    all(kd$raw_lfc > 0.5),
  "the depleted factor's own change is not significant" = all(kd$padj < 0.05),
  "the depleted factor does not move more than the other factor - the files may be swapped" =
    all(kd$margin > 0)
)
cat("\nRULER 2 PASSED. In raw orientation the depleted factor is up in all four files, by more\n",
    "than the other factor in each. Negated, it is down - matching 231, where EIF3D's own\n",
    "te_lfc is -1.26 because it is TE-buffered: the residual mRNA is better translated.\n",
    "Note this ruler is about WHICH experiment, not which direction, which is why ruler 3 exists.\n",
    sep = "")
```

# 5. Ruler three: the matched 231 contrast

This is the only ruler built on independent data, so it is the one that cannot be fooled by a
mistake made consistently throughout the MCF7-SIX1 files. eIF3d and eIF3e targets overlap between
the two cell lines, so the negated MCF7-SIX1 outcome must correlate **positively** with the
matched 231 contrast. If the sign were wrong, it would correlate negatively.

Two forms of it are reported: a transcriptome-wide rank correlation, and the separation of
**231-defined** eIF3d/eIF3e target genes from **231-defined** negative controls, scored on
MCF7-SIX1 values. The second matters because it is the same shape as the spot-check notebook
`10_` runs, so the threshold used there is calibrated here rather than guessed.

**Decisions**

- Report both orientations of the correlation side by side. A check that only ever prints the
  passing number does not show its own discriminating power.
- Restrict to the genes the model will actually see, so the numbers here are the numbers
  `10_` will reproduce.

```{r ruler_xcell}
tco <- readRDS(file.path(here("output", "translational_control_overview"), params$matrix_rds))
pos_of <- function(g) read_csv(here("output", "genesets",
    paste0("hypoxia_", g, "_promotes_TE_1hr_lfc0.5.csv")), show_col_types = FALSE)$ensembl_gene
neg_of <- function(g) read_csv(here("output", "predictive_modeling",
    if (g == "3e") "negative_control_genes_si3e.csv" else "negative_control_genes.csv"),
    show_col_types = FALSE) %>% mutate(gg = sub("[.].*", "", gene_id)) %>% pull(gg)

xc <- pmap_dfr(list(seq_len(nrow(registry))), function(i) {
  g <- registry$gene[i]
  d <- neg[[registry$label[i]]]
  m <- tco %>% dplyr::select(gene_id_clean) %>%
    inner_join(d %>% dplyr::select(gene_id_clean, te_lfc), by = "gene_id_clean") %>%
    filter(!is.na(te_lfc))
  r <- read_csv(file.path(OUT, registry$ref231[i]), show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>%
    dplyr::select(gene_id_clean, ref = te_lfc)
  j <- inner_join(m, r, by = "gene_id_clean") %>% filter(is.finite(ref))
  p <- pos_of(g); n <- setdiff(neg_of(g), p)
  yp <- m$te_lfc[m$gene_id_clean %in% p]; yn <- m$te_lfc[m$gene_id_clean %in% n]
  tibble(contrast = registry$label[i], n_modelled = nrow(m), n_paired = nrow(j),
         rho_negated = cor(j$te_lfc, j$ref, method = "spearman"),
         rho_if_raw  = -cor(j$te_lfc, j$ref, method = "spearman"),
         n_pos = length(yp), n_neg = length(yn),
         sep_negated = median(yp) - median(yn),
         sep_if_raw  = -(median(yp) - median(yn)))
})
knitr::kable(xc, digits = 4)

stopifnot(
  "the negated outcome does not correlate positively with the matched 231 contrast - the sign is wrong" =
    all(xc$rho_negated > params$min_xcell_rho),
  "231-defined targets do not score above 231-defined negative controls on MCF7-SIX1 values" =
    all(xc$sep_negated > params$min_xcell_sep),
  "the raw orientation would have passed too - this ruler has no discriminating power" =
    all(xc$rho_if_raw < 0) && all(xc$sep_if_raw < 0),
  "the four contrasts do not agree on a single gene count" = n_distinct(xc$n_modelled) == 1
)
cat(sprintf(paste0("\nRULER 3 PASSED on all four contrasts. Negated rho spans %+.3f to %+.3f;\n",
                   "the raw orientation would have spanned %+.3f to %+.3f, so the check\n",
                   "distinguishes the two rather than passing either way.\n",
                   "All four score %d of the matrix's %d genes.\n"),
            min(xc$rho_negated), max(xc$rho_negated),
            max(xc$rho_if_raw), min(xc$rho_if_raw),
            xc$n_modelled[1], nrow(tco)))
```

## What a flipped sign would look like

The rulers above pass. A check that has only ever been seen to pass is uninformative, so this
re-runs ruler three on a deliberately un-negated outcome and confirms it fails.

```{r flip_demo, eval = isTRUE(params$flip_demo)}
flip_ok <- map_lgl(seq_len(nrow(registry)), function(i) {
  d <- raw[[registry$label[i]]] %>%                       # NOT negated - the mistake
    mutate(gene_id_clean = sub("[.].*", "", gene_id), te_lfc = log2FoldChange)
  m <- tco %>% dplyr::select(gene_id_clean) %>%
    inner_join(d %>% dplyr::select(gene_id_clean, te_lfc), by = "gene_id_clean") %>%
    filter(!is.na(te_lfc))
  r <- read_csv(file.path(OUT, registry$ref231[i]), show_col_types = FALSE) %>%
    mutate(gene_id_clean = sub("[.].*", "", gene_id)) %>% dplyr::select(gene_id_clean, ref = te_lfc)
  j <- inner_join(m, r, by = "gene_id_clean") %>% filter(is.finite(ref))
  cor(j$te_lfc, j$ref, method = "spearman") > params$min_xcell_rho     # would the check pass?
})
cat("un-negated outcome passes ruler 3 for", sum(flip_ok), "of", nrow(registry), "contrasts\n")
stopifnot("an un-negated outcome would have passed ruler 3 - the check is not discriminative" =
            !any(flip_ok))
cat("CONFIRMED: forgetting the negation fails ruler 3 in every contrast.\n")
```

# 6. Write the promotes-positive outcome tables

```{r write_outcomes, eval = isTRUE(params$write_outputs)}
written <- pmap_chr(list(seq_len(nrow(registry))), function(i) {
  out <- neg[[registry$label[i]]] %>%
    transmute(gene_id, symbol, baseMean, te_lfc, te_padj, lfcSE,
              source_log2FoldChange = log2FoldChange)
  p <- file.path(OUT, registry$dest[i])
  write_csv(out, p)
  p
})
cat(paste0("wrote ", basename(written), collapse = "\n"), "\n")

# Read back and assert the file on disk is the negation of the file on the other side of the
# repo boundary. Writing and then trusting the write is how a half-written file gets modelled.
chk <- pmap_dfr(list(seq_len(nrow(registry))), function(i) {
  a <- read_csv(file.path(OUT, registry$dest[i]), show_col_types = FALSE)
  b <- raw[[registry$label[i]]]
  j <- inner_join(a %>% dplyr::select(gene_id, te_lfc),
                  b %>% dplyr::select(gene_id, log2FoldChange), by = "gene_id")
  tibble(contrast = registry$label[i], n = nrow(j),
         max_dev = max(abs(j$te_lfc + j$log2FoldChange), na.rm = TRUE),
         # Exact, and round-trip-proof: every gene's sign must be inverted.
         signs_inverted = all(sign(j$te_lfc) == -sign(j$log2FoldChange), na.rm = TRUE),
         na_pattern_same = identical(is.na(j$te_lfc), is.na(j$log2FoldChange)))
})
knitr::kable(chk)
# The tolerance is not slack: a CSV round-trip through readr keeps about 15 significant digits,
# so the written value is not bit-identical to the in-memory one. Measured, the worst deviation
# is 7e-18. Asserting == 0 here would be asserting a property of the file format, not of the
# negation, so the exact statement is made on the signs instead and the magnitude gets a bound
# far below anything that could matter.
stopifnot("a written file is not the negation of its source" = all(chk$max_dev < 1e-12),
          "a written file does not have every sign inverted" = all(chk$signs_inverted),
          "a written file's missing-value pattern differs from its source" =
            all(chk$na_pattern_same),
          "a written file lost genes" = all(chk$n == map_int(raw, nrow)))
```

# 7. Gene sets and the common-gene list

Two further artifacts. The gene sets let notebook `10_` run its own same-cell-line spot-check;
they are built from these files, so they cannot catch an error shared with this notebook, which
is exactly why ruler three above exists and is the asserted one.

The common-gene list is the 9,335 genes MCF7-SIX1 can score. Putting MCF7-SIX1 and 231 columns in
one heatmap requires every column to be built on the same genes, because a SHAP share is computed
over whatever genes the model explained. The list is written here so the 231 runs can be
restricted to it rather than each run deciding for itself.

**Decisions**

- Build gene sets for all four contrasts, including si3e, which had none. Otherwise the si3e
  columns would silently lose a check the si3d columns have.
- Define them by the same rule notebook `42_prepare_mcf7six1_genesets.Rmd` used for si3d, so the
  si3d files this notebook writes reproduce the existing ones rather than competing with them.

```{r write_genesets, eval = isTRUE(params$write_outputs)}
lfc_cut <- as.numeric(params$lfc)
gs <- pmap_dfr(list(seq_len(nrow(registry))), function(i) {
  d   <- neg[[registry$label[i]]]
  cnd <- paste0("mcf7six1_", registry$condition[i]); g <- registry$gene[i]
  nm  <- paste0(cnd, "_", g, "_promotes_TE_1hr_lfc", params$lfc, ".csv")
  p <- d %>% filter(te_lfc > lfc_cut, !is.na(te_padj), te_padj < 0.05) %>%
    transmute(db_gene_symbol = symbol, ensembl_gene = gene_id_clean, te_lfc, te_padj,
              gs_name = paste0(cnd, "_", g, "_promotes_TE_1hr_lfc", params$lfc),
              gs_description = paste0("genes whose translation is promoted by eIF", g,
                " in ", registry$condition[i], " at 1hr in MCF7-SIX1 cells, te_lfc > ",
                params$lfc, " (te_lfc is promotes-positive: negated from log2(si", g,
                "/sictrl))"),
              gs_source_species = "homo sapiens")
  write_csv(p, here("output", "genesets", nm))
  n <- d %>% filter(abs(te_lfc) < as.numeric(params$neg_lfc),
                    !is.na(te_padj), te_padj > as.numeric(params$neg_padj)) %>%
    transmute(gene_id, symbol, te_lfc, te_padj)
  nn <- paste0("mcf7six1_", registry$condition[i], "_", g, "_negative_controls_lfc",
               params$neg_lfc, ".csv")
  write_csv(n, here("output", "genesets", nn))
  tibble(contrast = registry$label[i], positives = nrow(p), negatives = nrow(n),
         pos_file = nm, neg_file = nn)
})
knitr::kable(gs %>% dplyr::select(contrast, positives, negatives))
stopifnot("a positive gene set is implausibly small" = all(gs$positives >= 50),
          "a negative control set is smaller than its positive set" = all(gs$negatives > gs$positives))

# Reproduce nb42's existing si3d sets rather than quietly disagreeing with them. nb42 wrote its
# hypoxia/normoxia sets under a name without the cell-line prefix on the condition, so the two
# are compared by content.
for (i in which(registry$gene == "3d")) {
  old <- here("output", "genesets",
              paste0("mcf7six1_", registry$condition[i], "_3d_promotes_TE_1hr_lfc",
                     params$lfc, ".csv"))
  if (file.exists(old)) {
    a <- read_csv(old, show_col_types = FALSE)$ensembl_gene
    b <- read_csv(here("output", "genesets", gs$pos_file[which(registry$gene == "3d" &
           registry$condition == registry$condition[i])]), show_col_types = FALSE)$ensembl_gene
    cat(sprintf("nb42 %s: %d genes; here: %d; identical: %s\n",
                registry$condition[i], length(a), length(b), setequal(a, b)))
  }
}
```

```{r write_common_genes, eval = isTRUE(params$write_outputs)}
common <- tco %>% dplyr::select(gene_id_clean) %>%
  filter(gene_id_clean %in% Reduce(intersect, map(neg, ~ .x$gene_id_clean[!is.na(.x$te_lfc)])))
stopifnot("the common-gene list is not a subset of the feature matrix" =
            all(common$gene_id_clean %in% tco$gene_id_clean),
          "the common-gene list has duplicates" = !any(duplicated(common$gene_id_clean)),
          "the common-gene list is implausibly small" = nrow(common) > 8000)
write_csv(common, file.path(here("output", "translational_control_overview"),
                            "mcf7six1_common_genes.csv"))
cat(sprintf("common genes scorable in every MCF7-SIX1 contrast and present in the matrix: %d of %d (%.1f%%)\n",
            nrow(common), nrow(tco), 100 * nrow(common) / nrow(tco)))
```

# 8. The isoform caveat, measured

A MCF7-SIX1-specific feature matrix is not available, and cannot be built without new compute:
the precomputed G4 scores cover only about 5% of the transcripts MCF7-SIX1 would select that 231
does not. So the MCF7-SIX1 columns are modelled on the **231-selected** transcript per gene.

That is not free, and this section measures the exposure rather than asserting it is small. The
number to know is how often the two cell lines would pick a different dominant transcript for the
same gene; for those genes the feature values describe an isoform that is not the one MCF7-SIX1
mainly expresses, which adds noise to the features and would *understate* the families whose
features are isoform-sensitive.

**Decisions**

- Report the discordance rather than correcting for it. Correcting is not possible without the
  missing features.
- Write the concordant-gene list, so the MCF7-SIX1 model can be re-run on genes where the two
  cell lines agree. Comparing that arm against the full one turns this caveat into a measurement.

```{r isoform_caveat, eval = isTRUE(params$write_outputs)}
m6_path <- here("output", "predictive_modeling", "feature_matrix_mcf7six1_external_stability.rds")
if (file.exists(m6_path)) {
  m6 <- readRDS(m6_path) %>% distinct(gene_id_clean, .keep_all = TRUE)
  cmp <- tco %>% dplyr::select(gene_id_clean, tx_231 = transcript_id_clean) %>%
    inner_join(m6 %>% dplyr::select(gene_id_clean, tx_mcf = transcript_id_clean),
               by = "gene_id_clean") %>%
    filter(gene_id_clean %in% common$gene_id_clean)
  cat(sprintf("genes with a transcript choice in both cell lines: %d\n", nrow(cmp)))
  cat(sprintf("  same dominant transcript:      %5d (%.1f%%)\n",
              sum(cmp$tx_231 == cmp$tx_mcf), 100 * mean(cmp$tx_231 == cmp$tx_mcf)))
  cat(sprintf("  different dominant transcript: %5d (%.1f%%)  <- features describe 231's isoform\n",
              sum(cmp$tx_231 != cmp$tx_mcf), 100 * mean(cmp$tx_231 != cmp$tx_mcf)))
  conc <- cmp %>% filter(tx_231 == tx_mcf) %>% dplyr::select(gene_id_clean)
  write_csv(conc, file.path(here("output", "translational_control_overview"),
                            "mcf7six1_isoform_concordant_genes.csv"))
  cat(sprintf("wrote mcf7six1_isoform_concordant_genes.csv (%d genes)\n", nrow(conc)))
} else {
  cat("NOTE: no MCF7-SIX1 feature matrix on disk, so the discordance cannot be measured here.\n")
}
```

# Summary

```{r summary}
cat("Direction: all three rulers agree that the MCF7-SIX1 files are log2(KD/sictrl) and must be\n")
cat("negated. The un-negated orientation fails ruler 3 in all four contrasts.\n\n")
cat("Written, promotes-positive, ready for notebook 10_ to resolve from params:\n")
cat(paste0("  ", registry$dest, collapse = "\n"), "\n")
cat("\nUse condition = \"mcf7six1_hypoxia\" / \"mcf7six1_normoxia\", timepoint = \"1hr\",\n")
cat("negate_outcome = FALSE (the negation is already in the file).\n")
```
