---
title: "Weber DAP5 outcome prep: HEK293T DAP5 knockout TE, refit with deFunction, on MANE transcripts"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  geo_dir: "accessories/weber_dap5"
  ribodiff_xlsx: "accessories/dap5kd_riboseq_weber2022.xlsx"
  mane_csv: "accessories/human/mane_select_v49.csv"
  min_counts: 40
  min_ribodiff_rho: 0.9
  lfc: "0.5"
  neg_lfc: "0.5"
  neg_padj: "0.3"
  write_outputs: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "10w_weber_dap5_outcome_prep.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

Turns Weber et al. 2022 ribosome profiling of a **DAP5 (EIF4G2) knockout in HEK293T** (GEO
GSE155854) into an outcome file that `10_rf_regression.Rmd` can model, so DAP5 can become a
column in the SHAP, correlation and top-100 heatmaps.

**Decisions (Kate, 2026-10-01)**

- **TE is refit from the GEO read counts with `deFunction(mode = "TE")`**, the same wrapper and
  interaction design (`~ Condition + SeqType + Condition:SeqType`, apeglm shrinkage, no independent
  filtering) that produced every MDA-MB-231 eIF3 column, rather than taking the paper's RiboDiff
  values. The RiboDiff table is used only to show the data reproduce the paper.
- **Orientation: wild type over knockout**, `test_condition = "WT"`, `control_condition = "KO"`, so
  the coefficient is log2(TE WT / TE KO) and **positive = DAP5 promotes**, as in our files. No
  negation is needed anywhere.
- **Features are computed on each gene's MANE Select transcript.** Every other heatmap column uses
  that dataset's own most abundant or representative transcript; the study reports no transcripts
  and no HEK293T isoform data exists for it, so this column is the one stated exception and is
  starred on every heatmap.
- **Genes are joined to MANE by gene symbol, direct matches only.** The GEO tables carry hg19-era
  symbols and no Ensembl IDs. No alias rescue is attempted; the genes lost are counted and written
  out in section 5.
- **Low-count filter: at least `min_counts` reads summed over the four RNA libraries AND over the
  four ribosome libraries.** This is the filter used in the earlier verification of this dataset
  (2026-09-30, r = 0.999 against RiboDiff); it is not a project-wide convention. `deFunction` itself
  does not filter.
- Output names follow the project's `translation_categories_si{gene}_vs_sictrl_*` convention so
  `10_` finds them from params (`gene = dap5`, `condition = hek293t_normoxia`, `timepoint =
  steadystate`), with the convention's known wart: it says "si" and "sictrl" for what is a CRISPR
  knockout against wild type.

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
library(DESeq2)
source(here("code", "functions.R"))
```

# 1. Load

```{r load}
rd_counts <- function(f) read.delim(gzfile(here(params$geo_dir, f)), check.names = FALSE) %>%
  dplyr::select(gene, WT_1, WT_2, DAP5_1, DAP5_2)
rna  <- rd_counts("GSE155854_readcounts_RNA.txt.gz")
ribo <- rd_counts("GSE155854_readcounts_Ribo.txt.gz")
geo_te <- read.delim(gzfile(here(params$geo_dir, "GSE155854_TE_DAP5_vs_wt.txt.gz")),
                     check.names = FALSE) %>%
  dplyr::select(geneIDs, padj, TEcontrol, TEDAP5KO, te = `log2FC_TE(DAP5KO vs control)`)
xl <- suppressMessages(read_excel(here(params$ribodiff_xlsx))) %>%
  dplyr::select(geneIDs, te = `log2FC_TE(DAP5KO vs control)`, padj = `padj TE`)
stopifnot("RNA count genes not unique"  = !any(duplicated(rna$gene)),
          "Ribo count genes not unique" = !any(duplicated(ribo$gene)),
          "GEO TE table genes not unique" = !any(duplicated(geo_te$geneIDs)))
cat("RNA counts:", nrow(rna), "genes | Ribo counts:", nrow(ribo), "genes | GEO TE table:",
    nrow(geo_te), "| supplementary xlsx:", nrow(xl), "\n")
```

# 2. Reproduce the paper

## 2a. 306 DAP5 targets

Weber et al. report **306** mRNAs with reduced TE in the DAP5 knockout (RiboDiff, FDR < 0.005).
That number is typed here as the expected value of a replication check; the data are read from
the files. Both the GEO table and the supplementary xlsx must give it.

```{r targets}
PAPER_N_TARGETS <- 306L
n_geo <- sum(geo_te$padj < 0.005 & geo_te$te < 0, na.rm = TRUE)
n_xl  <- sum(xl$padj < 0.005 & xl$te < 0, na.rm = TRUE)
cat(sprintf("reduced-TE targets at FDR < 0.005: GEO %d | xlsx %d | paper %d\n", n_geo, n_xl, PAPER_N_TARGETS))
stopifnot("GEO TE table does not give the paper's 306 targets" = n_geo == PAPER_N_TARGETS,
          "supplementary xlsx does not give the paper's 306 targets" = n_xl == PAPER_N_TARGETS)
```

## 2b. The TE column is log2(TE knockout / TE control)

```{r te_identity}
dev <- with(geo_te, max(abs(te - log2(TEDAP5KO / TEcontrol)), na.rm = TRUE))
cat("max |te - log2(TEDAP5KO / TEcontrol)|:", signif(dev, 3), "\n")
stopifnot("TE column is not log2(KO / control) - wrong column or orientation" = dev < 1e-6)
```

## 2c. Validated targets go down, GAPDH does not

The paper validates these as DAP5-dependent; in the source orientation (KO / control) each must be
negative. GAPDH is the paper's unaffected control.

```{r validated}
VALIDATED <- c("WNK1", "ROCK1", "AKT1", "INPPL1", "PIK3R2", "RAF1", "CDK12")
v <- geo_te %>% filter(geneIDs %in% c(VALIDATED, "GAPDH"))
knitr::kable(v %>% dplyr::select(geneIDs, te, padj), digits = 3)
stopifnot("a validated target is missing" = all(VALIDATED %in% v$geneIDs),
          "a validated DAP5 target does not fall in the knockout" = all(v$te[v$geneIDs %in% VALIDATED] < 0),
          "GAPDH is significant - it is the paper's unaffected control" = v$padj[v$geneIDs == "GAPDH"] > 0.005)
```

# 3. Refit with `deFunction`

```{r refit}
j <- inner_join(rna, ribo, by = "gene", suffix = c("_rna", "_ribo"))
m <- as.matrix(j[, -1]); rownames(m) <- j$gene
rna_cols <- grep("_rna$", colnames(m)); ribo_cols <- grep("_ribo$", colnames(m))
keep <- rowSums(m[, rna_cols]) >= params$min_counts & rowSums(m[, ribo_cols]) >= params$min_counts
m <- m[keep, ]
cat(sprintf("genes in both count tables: %d | passing >= %d reads in each assay: %d\n",
            nrow(j), params$min_counts, nrow(m)))

meta <- tibble(sample_name = colnames(m)) %>%
  mutate(library = sub(".*_", "", sample_name),
         comparison_variable = ifelse(grepl("^WT_", sample_name), "WT", "KO"))
stopifnot("sample metadata did not parse into 2 WT + 2 KO per assay" =
            all(table(meta$library, meta$comparison_variable) == 2))
norm <- counts(estimateSizeFactors(DESeqDataSetFromMatrix(
  m, data.frame(row.names = colnames(m), x = factor(meta$library)), ~ 1)), normalized = TRUE)

run <- function(mode) deFunction(countsMatrix = m, allCounts_Norm = norm, metadata = meta,
                                 control_condition = "KO", test_condition = "WT", mode = mode)
res <- map(c(TE = "TE", RNA = "RNA", RIBO = "RIBO"), run)
refit <- res$TE %>% dplyr::select(geneIDs = gene_id, te_lfc = log2FoldChange, te_padj = padj) %>%
  left_join(res$RNA  %>% dplyr::select(geneIDs = gene_id, rna_lfc  = log2FoldChange, rna_padj  = padj), by = "geneIDs") %>%
  left_join(res$RIBO %>% dplyr::select(geneIDs = gene_id, ribo_lfc = log2FoldChange, ribo_padj = padj), by = "geneIDs")
cat("refit genes:", nrow(refit), "| non-NA te_padj:", sum(!is.na(refit$te_padj)), "\n")
```

# 4. Direction and agreement with the paper

## 4a. Ruler: DAP5's own mRNA

The knockout destroys EIF4G2's mRNA, so in this orientation (WT / KO) its `rna_lfc` must be
strongly positive - the same ruler as EIF3D's `rna_lfc` in the MDA-MB-231 files. Its TE cannot
serve: ribosomes fall with the mRNA, so the ratio barely moves.

```{r ruler_own}
own <- refit %>% filter(geneIDs == "EIF4G2")
knitr::kable(own, digits = 3)
stopifnot("EIF4G2 is not in the refit" = nrow(own) == 1,
          "EIF4G2 rna_lfc is not strongly positive - the orientation is wrong" = own$rna_lfc > 2)
```

## 4b. Validated targets are positive after the refit, and the refit tracks RiboDiff

Positive here means DAP5 promotes; the paper's validated targets must all be positive. The refit
must rank genes like RiboDiff does, against the **negated** RiboDiff value (theirs is KO / control).

```{r agreement}
vr <- refit %>% filter(geneIDs %in% VALIDATED)
stopifnot("a validated target is positive in the paper's orientation after the refit" =
            nrow(vr) == length(VALIDATED) && all(vr$te_lfc > 0))
a <- inner_join(refit, geo_te %>% dplyr::select(geneIDs, rd = te, rd_padj = padj), by = "geneIDs") %>%
  filter(is.finite(te_lfc), is.finite(rd))
rho <- cor(a$te_lfc, -a$rd, method = "spearman")
shared_targets <- sum(a$te_padj < 0.005 & a$te_lfc > 0 & a$rd_padj < 0.005 & a$rd < 0, na.rm = TRUE)
cat(sprintf("refit vs negated RiboDiff: Spearman %.3f (n = %d) | refit targets (padj < 0.005, te_lfc > 0): %d | shared with the paper's %d: %d\n",
            rho, nrow(a), sum(refit$te_padj < 0.005 & refit$te_lfc > 0, na.rm = TRUE),
            PAPER_N_TARGETS, shared_targets))
stopifnot("the refit does not track the paper's TE" = rho > params$min_ribodiff_rho)

ggplot(a, aes(-rd, te_lfc)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_abline(linetype = 2, colour = "grey50") +
  labs(x = "RiboDiff TE, negated: log2(control / DAP5 KO)",
       y = "deFunction TE refit: log2(WT / KO), apeglm",
       title = sprintf("Weber 2022 DAP5 KO, HEK293T: Spearman %.3f, n = %d", rho, nrow(a))) +
  theme_classic(base_size = 12)
```

# 5. Join to MANE Select by symbol (direct matches only)

```{r mane}
mane <- read_csv(here(params$mane_csv), col_names = c("gene_id_clean", "transcript_id_clean", "symbol"),
                 show_col_types = FALSE)
stopifnot("MANE file is not one transcript per gene" = !any(duplicated(mane$gene_id_clean)),
          "MANE IDs are versioned - expected bare IDs" = !any(grepl("[.]", mane$transcript_id_clean)))
dup_sym <- mane$symbol[duplicated(mane$symbol)]
tested <- refit %>% filter(!is.na(te_lfc), !is.na(te_padj))
joined <- tested %>% inner_join(mane %>% filter(!symbol %in% dup_sym), by = c("geneIDs" = "symbol"))
unmatched <- tested %>% filter(!geneIDs %in% joined$geneIDs)
cat(sprintf("tested genes: %d | matched to a MANE transcript by symbol: %d | unmatched: %d (%.1f%%) | MANE symbols dropped as duplicated: %d\n",
            nrow(tested), nrow(joined), nrow(unmatched), 100 * nrow(unmatched) / nrow(tested), length(unique(dup_sym))))
cat(sprintf("unmatched genes among the refit's DAP5 targets (padj < 0.005, te_lfc > 0): %d of %d\n",
            sum(unmatched$te_padj < 0.005 & unmatched$te_lfc > 0),
            sum(tested$te_padj < 0.005 & tested$te_lfc > 0)))
stopifnot("a symbol matched more than one MANE gene" = !any(duplicated(joined$geneIDs)),
          "EIF4G2 did not match MANE" = "EIF4G2" %in% joined$geneIDs)
```

**Note on the unmatched genes.** These are genes whose hg19-era symbol is not the current HGNC
symbol on any MANE Select transcript (renamed genes, readthroughs, non-coding genes with no MANE
entry). They are dropped rather than rescued through alias tables, by decision; the list is written
to `output/translational_control_overview/weber_dap5_unmatched_symbols.csv` so the loss can be
inspected.

# 6. Write

```{r write, eval = isTRUE(params$write_outputs)}
OUT <- here("output")
outcome <- joined %>% transmute(gene_id = gene_id_clean, symbol = geneIDs, transcript_id_clean,
                                te_lfc, te_padj, rna_lfc, rna_padj, ribo_lfc, ribo_padj)
f_out <- "translation_categories_sidap5_vs_sictrl_hek293t_normoxia_steadystate.csv"
write_csv(outcome, file.path(OUT, f_out))
write_csv(unmatched, file.path(OUT, "translational_control_overview", "weber_dap5_unmatched_symbols.csv"))

# The MANE transcript map (cell set "mane") is built by 01q_mane_tx_map.R from this file and the
# cap-binding outcomes together, so both share one transcript per gene. Run it after this.
stopifnot("an outcome transcript is duplicated" = !any(duplicated(outcome$transcript_id_clean)))

back <- read_csv(file.path(OUT, f_out), show_col_types = FALSE)
stopifnot("written outcome lost or gained genes" = nrow(back) == nrow(joined),
          "written te_lfc differs from the refit" =
            max(abs(back$te_lfc - joined$te_lfc[match(back$symbol, joined$geneIDs)])) < 1e-12)
cat("wrote", f_out, "(", nrow(outcome), "genes ) - now run 01q_mane_tx_map.R\n")
```

# 7. Gene sets for `10_`'s spot-check

Same rule as every other column: positives `te_lfc > 0.5 & padj < 0.05`, negative controls
`|te_lfc| < 0.5 & padj > 0.3`.

```{r genesets, eval = isTRUE(params$write_outputs)}
lfc_cut <- as.numeric(params$lfc)
nm <- paste0("hek293t_normoxia_dap5_promotes_TE_steadystate_lfc", params$lfc)
p <- outcome %>% filter(te_lfc > lfc_cut, te_padj < 0.05) %>%
  transmute(db_gene_symbol = symbol, ensembl_gene = gene_id, te_lfc, te_padj, gs_name = nm,
            gs_description = paste0("genes whose translation DAP5 promotes in HEK293T (Weber 2022, ",
              "GSE155854), deFunction TE refit, te_lfc = log2(WT / KO) > ", params$lfc, " & padj < 0.05"),
            gs_source_species = "homo sapiens")
write_csv(p, here("output", "genesets", paste0(nm, ".csv")))
n <- outcome %>% filter(abs(te_lfc) < as.numeric(params$neg_lfc), te_padj > as.numeric(params$neg_padj)) %>%
  transmute(gene_id, symbol, te_lfc, te_padj)
write_csv(n, here("output", "genesets", paste0("hek293t_normoxia_dap5_negative_controls_lfc", params$neg_lfc, ".csv")))
cat("DAP5-promoted:", nrow(p), "| negative controls:", nrow(n), "\n")
stopifnot("positive gene set is implausibly small" = nrow(p) >= 50,
          "negative control set is smaller than the positive set" = nrow(n) > nrow(p))
```

```{r session}
sessionInfo()
```
