---
title: "`r paste0('SHAP (regression): eIF', params$gene, ' ', params$condition, ' ', params$timepoint, ', y=', params$target_transform)`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  readout: "TE"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set_label: "tcoreg"
  variant_set: "corrected"
  initiation_score: "both"
  clip_mode: "aggregate"
  target_transform: "raw"
  winsorize_y: 0
  min_node_size: 5
  mtry: 0
  num_trees: 1000
  splitrule: "variance"
  weight_mode: "none"
  restrict_to_classifier_genes: FALSE
  exclude_features: ""
  split_seed: 9
  explain_n: 200
  explain_sizes: "50,100,200,400"
  top_frac: 0.10
  corr_group_threshold: 0.8
  cluster_k_max: 6
  cluster_n_null: 50
  save_shap_plots: TRUE
  sweep_mode: FALSE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "11_shap_regression.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook breaks the regression's predictions down per gene and per feature, using SHAP, so
that each prediction can be expressed as a baseline plus one signed contribution from every
feature.

Its main practical advantage over `03_shap.Rmd` is that it explains the model directly.

**Decisions**

- Explain `10_`'s forest itself, with no intermediate model.
- Load the split `10_` saved, rather than rebuilding it from parameters.
- Verify the decomposition by checking that it adds up (below), which is the property that
  matters for any additive explanation.
- Report family-level sums as exact, because SHAP is additive and summing within a family is a
  valid operation.

## Why no second model is needed here

`03_shap.Rmd` cannot explain `02_`'s classifier directly, because `treeshap` does not support
ranger probability forests. It works around that by fitting a substitute regression forest on the
0/1 labels, copying `mtry`, `min.node.size` and `num.trees` off the saved classifier, and then
running two checks to confirm the substitute behaves like the original: one comparing their
configuration, one comparing their held-out predictions.

A regression forest needs none of that, because `treeshap` handles it as-is. Both of those checks
existed to police the substitute, so they are removed rather than carried over. What remains is
the check that actually matters for an additive decomposition: the contributions must sum to the
prediction, with the leftover constant across genes and equal to the forest's mean training
prediction.

**Decisions**

- Remove the substitute model and its two checks.
- Replace them with one check that the decomposition closes to within 1e-8.
- Keep the engine indirection, so Kernel SHAP stays available as a cross-check. With no substitute
  on either side, that comparison is now cleaner than it was.

## Two summaries that had to be redefined

Two things in `03_` were defined in terms of the positive class, which does not exist here.

- The **signed** column averaged SHAP over the positive class. It now averages over the top
  fraction of genes by predicted outcome. The reason for restricting to a subset is unchanged: a
  signed average over all genes cancels to about zero for every feature, which is why
  `mean_signed_shap` is not a measure of direction.
- The **subgroup clustering** clustered held-out positives. It now clusters that same top
  fraction, which keeps the group size comparable to the classifier's 287 genes.

**Decisions**

- Use the top 10% by predicted outcome wherever `03_` used the positive class.
- Keep `shap_value_cor` as the directional summary; it needs no change.
- Keep exporting `mean_signed_shap` for completeness while stating that it is not a direction.

## Cost

SHAP is considerably slower here than on the classifier, because of the model rather than the
method: this forest is grown on 8,140 rows instead of 2,298, so its trees are much deeper, and it
uses 1000 trees instead of 500.

**Decisions**

- Under `sweep_mode = TRUE`, compute one SHAP matrix instead of three by skipping the signed
  column, the correlated groups, the explain-set size check and the subgroups. That is what makes
  the 20-run floor sweep in `11d_` affordable.

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(treeshap)
library(cowplot)
```

```{r load}
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
source(here("code", "translational_control_overview", "10_regression_presets.R"))

suffix <- reg_suffix(list(
  gene = params$gene, direction = params$direction, readout = params$readout,
  condition = params$condition, timepoint = params$timepoint, lfc = params$lfc,
  feature_set_label = params$feature_set_label, variant_set = params$variant_set,
  initiation_score = params$initiation_score, clip_mode = params$clip_mode,
  target_transform = params$target_transform, winsorize_y = params$winsorize_y,
  min_node_size = params$min_node_size, mtry = params$mtry, num_trees = params$num_trees,
  splitrule = params$splitrule, weight_mode = params$weight_mode,
  restrict_to_classifier_genes = params$restrict_to_classifier_genes),
  as.integer(params$split_seed))

out_dir <- here("output", "translational_control_overview")
model_path <- file.path(out_dir, paste0("rfreg_model_", suffix, ".rds"))
md_path    <- file.path(out_dir, paste0("rfreg_model_data_", suffix, ".rds"))
stopifnot(
  "regression model RDS not found - knit 10_rf_regression.Rmd with these params first" =
    file.exists(model_path),
  "regression model data not found - knit 10_ with save_model_data = TRUE" =
    file.exists(md_path))

rf <- readRDS(model_path)
md <- readRDS(md_path)
# The saved split belongs to this model. 07c_ in the frozen pipeline re-derives the split from
# params and has to assert its way back to agreement; 10_ saves it, which removes the drift
# rather than detecting it.
stopifnot("the saved data does not belong to this model" = identical(md$suffix, suffix))

train_data <- md$train; test_data <- md$test; feature_cols <- md$feature_cols
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
stopifnot("annotation does not cover every modelled feature" =
            all(feature_cols %in% ann$feature))

y_units <- if (params$target_transform == "rank_ecdf") "percentile of training y" else
           if (params$target_transform == "robust_z")  "MAD of training y" else "log2 TE LFC"
cat("Suffix:", suffix, "\n")
cat("Model:", rf$treetype, "| mtry", rf$mtry, "| min.node.size", rf$min.node.size,
    "| num.trees", rf$num.trees, "| OOB R2", round(rf$r.squared, 4), "\n")
cat("Train", nrow(train_data), "| Test", nrow(test_data), "|", length(feature_cols), "features\n")
cat("SHAP will be in units of:", y_units, "\n")
```

# 1. Computing SHAP

Computes SHAP values for a sample of held-out genes and checks the decomposition adds up.

**Decisions**

- Explain a random sample of `explain_n` held-out genes rather than all of them, for cost.
- Re-impose the feature order on every call, because `treeshap` returns columns in its own order
  and a silent reorder would attribute values to the wrong features.
- Require the leftover after subtracting all contributions to be constant to within 1e-8, and to
  equal the forest's mean training prediction.


```{r shap_setup}
# No substitute model: what treeshap is handed is notebook 10's forest itself.
stopifnot("this is not a regression forest - treeshap cannot explain a probability forest" =
            rf$treetype == "Regression")

unified <- ranger.unify(rf, as.data.frame(train_data[, feature_cols]))
# Column order is re-imposed on every call: treeshap returns $shaps in its own order, and a
# silent reorder would misattribute every value to the wrong feature.
shap_of    <- function(X) as.data.frame(
  treeshap(unified, as.data.frame(X)[, feature_cols], interactions = FALSE,
           verbose = FALSE)$shaps[, feature_cols])
predict_of <- function(X) predict(rf, data = as.data.frame(X))$predictions

explain_n   <- min(params$explain_n, nrow(test_data))
set.seed(9)
explain_idx <- sample(nrow(test_data), explain_n)
explain_set <- as.data.frame(test_data[explain_idx, feature_cols])
shap_vals   <- shap_of(explain_set)
stopifnot("SHAP matrix does not match the explain set" =
            nrow(shap_vals) == explain_n && ncol(shap_vals) == length(feature_cols))
```

```{r additivity}
# The check that replaces 03_'s two. A decomposition that closes on the predictions, with a
# constant leftover, is by definition a decomposition of this model, so there is no substitute
# model whose fidelity needs arguing.
pred    <- predict_of(explain_set)
implied <- pred - rowSums(shap_vals)
mean_train_pred <- mean(predict_of(train_data[, feature_cols]))
cat(sprintf("baseline %.6f (sd %.2e over %d genes) | mean training prediction %.6f | diff %.2e\n",
            mean(implied), sd(implied), length(implied), mean_train_pred,
            abs(mean(implied) - mean_train_pred)))
stopifnot(
  "SHAP does not close on these predictions - the baseline is not constant" =
    sd(implied) < 1e-8,
  "the recovered baseline is not the forest's mean training prediction" =
    abs(mean(implied) - mean_train_pred) < 1e-6
)
baseline <- mean(implied)
```

```{r shap_table}
shap_value_cor <- map_dbl(feature_cols, function(f) {
  v <- explain_set[[f]]; s <- shap_vals[[f]]
  if (sd(v, na.rm = TRUE) == 0 || sd(s, na.rm = TRUE) == 0) return(NA_real_)
  cor(v, s, use = "complete.obs", method = "spearman")
})

shap_tbl <- tibble(
  feature          = feature_cols,
  mean_abs_shap    = colMeans(abs(shap_vals)),
  mean_signed_shap = colMeans(shap_vals),     # not direction - cancels to ~0. See the header.
  shap_value_cor   = shap_value_cor
) %>%
  left_join(ann %>% dplyr::select(feature, family, block), by = "feature") %>%
  arrange(desc(mean_abs_shap)) %>%
  mutate(share_pct = 100 * mean_abs_shap / sum(mean_abs_shap), rank_shap = row_number())
print(as.data.frame(head(shap_tbl %>%
  mutate(across(where(is.numeric), ~ signif(., 4))), 20)), row.names = FALSE)

# Cross-experiment note: mean_abs_shap is in TE units and so is NOT comparable across
# experiments whose outcome spread differs (MAD spans 7.3x across the 17 si3d/si3e contrasts).
# share_pct is a within-run ratio and is comparable. mad_scaled is the third option: TE units
# divided by this run's own outcome spread. See 10d_target_scale_analysis.Rmd.
shap_tbl$mean_abs_shap_mad_scaled <- shap_tbl$mean_abs_shap / mad(md$train$y)
```

```{r family_table}
# Exact, because SHAP is additive: summing within a family is a legitimate operation, which is
# the reason this decomposition exists at all. Impurity importance structurally cannot do it.
fam_tbl <- shap_tbl %>% group_by(family) %>%
  summarise(n = n(), family_shap = sum(mean_abs_shap),
            top_feature = feature[which.max(mean_abs_shap)], .groups = "drop") %>%
  mutate(share_pct = 100 * family_shap / sum(family_shap)) %>%
  arrange(desc(family_shap))
print(as.data.frame(fam_tbl %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)
stopifnot("family shares do not sum to 100" =
            abs(sum(fam_tbl$share_pct) - 100) < 1e-8)
```

# 2. SHAP versus impurity importance

Compares the SHAP ranking against the impurity ranking from `10_`.

**Decisions**

- Treat disagreement as expected rather than as an error. Impurity favours continuous and
  high-cardinality features, which is the reason for computing SHAP at all.
- Report the comparison at both feature and family level.


```{r impurity_vs_shap}
imp <- read_csv(file.path(out_dir, paste0("rfreg_importance_", suffix, ".csv")),
                show_col_types = FALSE) %>%
  dplyr::select(feature, importance, family) %>%
  mutate(imp_share = 100 * importance / sum(importance),
         rank_imp  = rank(-importance))
cmp <- shap_tbl %>% dplyr::select(feature, family, share_pct, rank_shap) %>%
  inner_join(imp %>% dplyr::select(feature, imp_share, rank_imp), by = "feature")
cat("feature-level Spearman, SHAP share vs impurity share:",
    round(cor(cmp$share_pct, cmp$imp_share, method = "spearman"), 4), "\n")

fam_cmp <- cmp %>% group_by(family) %>%
  summarise(shap = sum(share_pct), imp = sum(imp_share), .groups = "drop") %>%
  mutate(delta = shap - imp) %>% arrange(desc(abs(delta)))
print(as.data.frame(fam_cmp %>% mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
```

```{r fig_shap, fig.width = 11, fig.height = 9, eval = !isTRUE(params$sweep_mode)}
top20 <- shap_tbl %>% head(20) %>% mutate(feature = fct_reorder(feature, mean_abs_shap))
p_lol <- ggplot(top20, aes(mean_abs_shap, feature)) +
  geom_segment(aes(x = 0, xend = mean_abs_shap, yend = feature), colour = "grey70") +
  geom_point(aes(colour = shap_value_cor), size = 3) +
  scale_colour_gradient2(low = "#0072B2", mid = "grey85", high = "#D55E00", midpoint = 0,
                         name = "value vs\nown SHAP") +
  labs(x = paste0("mean |SHAP| (", y_units, ")"), y = NULL,
       title = "(a) Top 20 features",
       subtitle = sprintf("n = %d explained genes | colour is the DIRECTIONAL quantity", explain_n)) +
  theme_classic(base_size = 11)

p_fam <- fam_tbl %>% mutate(family = fct_reorder(family, share_pct)) %>%
  ggplot(aes(share_pct, family)) +
  geom_col(fill = "#E69F00", alpha = 0.9) +
  geom_text(aes(label = sprintf("%.1f%% (%d)", share_pct, n)), hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "share of total |SHAP| (%)", y = NULL, title = "(b) Family shares",
       subtitle = "exact: SHAP is additive, so summing within a family is valid") +
  theme_classic(base_size = 11)

p_cmp <- ggplot(cmp, aes(imp_share, share_pct)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.75, size = 1.8, colour = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.3,
           label = sprintf("Spearman = %.3f\nn = %d features",
                           cor(cmp$share_pct, cmp$imp_share, method = "spearman"), nrow(cmp))) +
  labs(x = "impurity share (%)", y = "SHAP share (%)", title = "(c) SHAP vs impurity",
       subtitle = "impurity is biased toward continuous/high-cardinality features") +
  theme_classic(base_size = 11)

# Beeswarm: the per-gene spread a mean |SHAP| hides.
bees <- shap_vals %>% dplyr::select(all_of(head(shap_tbl$feature, 15))) %>%
  mutate(.row = row_number()) %>%
  pivot_longer(-.row, names_to = "feature", values_to = "shap") %>%
  left_join(explain_set %>% mutate(.row = row_number()) %>%
              pivot_longer(-.row, names_to = "feature", values_to = "value"),
            by = c(".row", "feature")) %>%
  group_by(feature) %>% mutate(v_scaled = if (sd(value, na.rm = TRUE) > 0)
    (value - min(value, na.rm = TRUE)) /
      (max(value, na.rm = TRUE) - min(value, na.rm = TRUE)) else 0.5) %>% ungroup() %>%
  mutate(feature = factor(feature, levels = rev(head(shap_tbl$feature, 15))))
p_bee <- ggplot(bees, aes(shap, feature, colour = v_scaled)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_jitter(height = 0.22, size = 0.7, alpha = 0.55) +
  scale_colour_gradient(low = "#0072B2", high = "#D55E00", name = "feature\nvalue",
                        breaks = c(0, 1), labels = c("low", "high")) +
  labs(x = paste0("SHAP (", y_units, ")"), y = NULL, title = "(d) Per-gene contributions",
       subtitle = "what the mean in (a) averages over") +
  theme_classic(base_size = 11)

pg <- plot_grid(p_lol, p_fam, p_cmp, p_bee, nrow = 2, rel_heights = c(1.15, 1))
print(pg)
if (isTRUE(params$save_shap_plots))
  ggsave(here("plots", paste0("11_shap_", suffix, ".pdf")), pg, width = 11, height = 9)
```

# 3. Signed contributions for the strongly-predicted genes

Reports which features push predictions up or down for the genes the model scores highest.

**Decisions**

- Restrict to the top 10% by predicted outcome, since a signed average over all genes cancels.
- Check the signed values sum to that subset's mean distance above the baseline, so the column is
  a genuine decomposition rather than a set of unrelated averages.


```{r signed, eval = !isTRUE(params$sweep_mode)}
# A signed mean over all genes cancels to ~0 for every feature, which is exactly why
# mean_signed_shap is not a measure of direction. 03_ avoided that by averaging over the
# positive class; there is no class here, so the comparable subset is the top fraction by
# predicted y - the genes the model says are most strongly eIF3d-promoted.
n_top    <- max(20L, floor(params$top_frac * nrow(test_data)))
top_idx  <- order(-predict_of(test_data[, feature_cols]))[seq_len(n_top)]
top_x    <- as.data.frame(test_data[top_idx, feature_cols])
shap_top <- shap_of(top_x)
cat(sprintf("top %.0f%% by predicted y: %d genes (median observed y %+.3f vs %+.3f overall)\n",
            100 * params$top_frac, n_top, median(test_data$y[top_idx]), median(test_data$y)))

signed <- tibble(feature = feature_cols, signed_shap = colMeans(shap_top)) %>%
  left_join(shap_tbl %>% dplyr::select(feature, family, mean_abs_shap), by = "feature") %>%
  mutate(abs_share    = 100 * mean_abs_shap / sum(mean_abs_shap),
         signed_share = 100 * signed_shap / sum(abs(signed_shap))) %>%
  arrange(desc(abs(signed_shap)))

# Additive by construction: the signed means must sum to this subset's mean lift above the
# baseline. If they do not, the column is not a decomposition and must not be read as one.
row_sums <- rowSums(shap_top)
stopifnot("the signed column is not additive - do not read it as a decomposition" =
            isTRUE(all.equal(sum(signed$signed_shap), mean(row_sums))))
cat(sprintf("sum of signed means %.6f = mean total lift %.6f above baseline %.6f\n",
            sum(signed$signed_shap), mean(row_sums), baseline))
print(as.data.frame(head(signed %>% mutate(across(where(is.numeric), ~ signif(., 4))), 15)),
      row.names = FALSE)
```

# 4. Correlated groups

Groups features by how correlated they are with each other, and sums SHAP within those groups.

Families are a grouping decided in advance; this is one read off the data. Where several
near-duplicate features split credit between them, the group recovers the total.

**Decisions**

- Cluster on absolute correlation with average linkage, cutting at `corr_group_threshold`.
- Report the same summary at thresholds 0.7, 0.8 and 0.9, so the grouping is not taken as exact.
- Flag groups that span more than one family, since those are where the prior grouping and the
  data disagree.


```{r corr_groups, eval = !isTRUE(params$sweep_mode)}
# Families are a prior grouping; this is the data-driven one. Where a family splits credit
# across near-duplicate members, the correlated group recovers the total.
Xall <- bind_rows(train_data[, feature_cols], test_data[, feature_cols])
Cm   <- suppressWarnings(cor(Xall, use = "pairwise.complete.obs"))
Cm[is.na(Cm)] <- 0
hc   <- hclust(as.dist(1 - abs(Cm)), method = "average")
grp  <- cutree(hc, h = 1 - params$corr_group_threshold)

cg <- shap_tbl %>% mutate(group = grp[match(feature, names(grp))]) %>%
  group_by(group) %>%
  summarise(n = n(), share = sum(share_pct),
            lead = feature[which.max(mean_abs_shap)],
            families = paste(sort(unique(family)), collapse = "+"),
            spans_families = n_distinct(family) > 1, .groups = "drop") %>%
  arrange(desc(share))
cat("Groups at |r| >=", params$corr_group_threshold, ":", nrow(cg),
    "| multi-feature:", sum(cg$n > 1), "| spanning >1 family:", sum(cg$spans_families), "\n")
print(as.data.frame(head(cg %>% mutate(share = round(share, 2)), 15)), row.names = FALSE)

for (th in c(0.7, 0.8, 0.9)) {
  g2 <- cutree(hc, h = 1 - th)
  cat(sprintf("  threshold %.1f -> %d groups, largest share %.2f%%\n", th, max(g2),
              max(tapply(shap_tbl$share_pct[match(names(g2), shap_tbl$feature)], g2, sum))))
}
```

# 5. Explain-set size, and gene subgroups

Checks whether the family ranking depends on how many genes were explained, then looks for
subgroups of genes driven by different features.

**Decisions**

- Re-run the explain-set size check rather than inheriting `03_`'s conclusion, since the held-out
  set here is about 2,032 genes rather than 574.
- Cluster on family-level profiles normalised within each gene, so groups reflect which families
  drove a gene rather than how strongly it was predicted.
- Test against a null that permutes each family independently across genes, standardising within
  each k before maximising over k.
- Keep two different best-k: one for testing, one for describing. They need not agree.


```{r explain_stability, eval = !isTRUE(params$sweep_mode)}
# 03_'s conclusion was reached on a 574-row test set; this one has ~2,032, so the question is
# re-opened rather than inherited.
sizes <- as.integer(str_trim(str_split(params$explain_sizes, ",")[[1]]))
sizes <- sizes[sizes <= nrow(test_data)]
stab <- map_dfr(sizes, function(n) {
  set.seed(9); idx <- sample(nrow(test_data), n)
  sv <- shap_of(as.data.frame(test_data[idx, feature_cols]))
  tibble(feature = feature_cols, mean_abs = colMeans(abs(sv))) %>%
    left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
    group_by(family) %>% summarise(share = 100 * sum(mean_abs), .groups = "drop") %>%
    mutate(share = 100 * share / sum(share), n = n)
})
wide <- stab %>% pivot_wider(names_from = n, values_from = share, names_prefix = "n") %>%
  arrange(desc(.data[[paste0("n", max(sizes))]]))
print(as.data.frame(wide %>% mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
rk <- stab %>% group_by(n) %>% mutate(rk = rank(-share)) %>%
  dplyr::select(family, n, rk) %>% pivot_wider(names_from = n, values_from = rk)
cat("\nSpearman of family ranking, smallest vs largest explain set:",
    round(cor(rk[[2]], rk[[ncol(rk)]], method = "spearman"), 3), "\n")
```

```{r subgroups, eval = !isTRUE(params$sweep_mode) && exists("shap_top")}
# Clustered on the same top fraction as section 3, so group size stays comparable to the
# classifier's 287 positive-class genes. Normalised within gene, so clusters reflect which
# families drove a gene rather than how strongly it was predicted.
source(here("code", "translational_control_overview", "03_subgroup_null.R"))
fam_of  <- ann$family[match(feature_cols, ann$feature)]
fam_mat <- t(rowsum(t(as.matrix(shap_top)), group = fam_of))   # exact: SHAP is additive
denom   <- rowSums(abs(fam_mat)); denom[denom == 0] <- NA
profile <- fam_mat / denom
keep    <- stats::complete.cases(profile)
profile <- profile[keep, , drop = FALSE]
sym_keep <- test_data$symbol[top_idx][keep]
cat("Profiles:", nrow(profile), "genes x", ncol(profile), "families\n")

# The null permutes each family independently across genes: co-occurrence destroyed, marginals
# preserved. Mean silhouette falls monotonically with k, so a raw best-over-k comparison pits the
# null's k=2 against an observed k=3 and returns the answer backwards. subgroup_null()
# standardises within k first. Two best_k are kept apart deliberately: best_k_z for testing,
# best_k_sil for describing.
nullres <- subgroup_null(profile, ks = 2:params$cluster_k_max, n_perm = params$cluster_n_null)
print(as.data.frame(nullres$per_k %>% mutate(across(where(is.numeric), ~ round(., 4)))),
      row.names = FALSE)
cat(sprintf("\ntest statistic peaks at k = %d: observed z = %.2f | null median %.2f | null max %.2f | p = %.4g\n",
            nullres$best_k_z, nullres$z_obs, median(nullres$z_null), max(nullres$z_null),
            nullres$p_overall))
best_k <- nullres$best_k_sil
set.seed(9)
km <- kmeans(profile, centers = best_k, nstart = 25)
subgroups <- tibble(symbol = sym_keep, cluster = km$cluster) %>% bind_cols(as_tibble(profile))
cat("Described at k =", best_k, "(the raw silhouette peak). Sizes:",
    paste(table(km$cluster), collapse = ", "), "\n")
lead <- apply(km$centers, 1, function(r) paste(names(sort(r, decreasing = TRUE))[1:2], collapse = ", "))
for (i in seq_len(best_k))
  cat(sprintf("  cluster %d (n=%3d): led by %s\n", i, sum(km$cluster == i), lead[i]))
```

```{r fig_null_band, eval = !isTRUE(params$sweep_mode), fig.width = 6.5, fig.height = 4.4}
# Plot the null band per k. In 03_ a single summary number hid that the observed value sat above
# the null at every k of 3 or more; the band is what made that visible.
band <- nullres$per_k
p_band <- ggplot(band, aes(k)) +
  geom_ribbon(aes(ymin = null_min, ymax = null_max), fill = "grey85") +
  geom_line(aes(y = null_median), colour = "grey40", linetype = "dashed", linewidth = 0.7) +
  geom_line(aes(y = observed), colour = "#E69F00", linewidth = 1) +
  geom_point(aes(y = observed), colour = "#E69F00", size = 2.4) +
  labs(x = "k", y = "mean silhouette",
       title = "Observed vs the per-k null band",
       subtitle = sprintf("grey = null min-max over %d permutations, dashed = null median, orange = observed (n = %d genes)",
                          params$cluster_n_null, nrow(profile))) +
  theme_classic(base_size = 12)
print(p_band)
if (isTRUE(params$save_shap_plots))
  ggsave(here("plots", paste0("11_subgroup_null_band_", suffix, ".pdf")), p_band,
         width = 6.5, height = 4.4)
```

# Export

Writes the feature and family tables, and the per-gene values.

**Decisions**

- Always write the feature and family tables, since those are what the floor sweep reads.
- Write the rest only outside sweep mode.


```{r export}
reg_write_csv(shap_tbl, file.path(out_dir, paste0("shapreg_feature_", suffix, ".csv")))
reg_write_csv(fam_tbl,  file.path(out_dir, paste0("shapreg_family_",  suffix, ".csv")))

if (!isTRUE(params$sweep_mode)) {
  reg_write_csv(fam_cmp, file.path(out_dir, paste0("shapreg_vs_impurity_", suffix, ".csv")))
  reg_write_csv(
    bind_cols(tibble(gene_id_clean = test_data$gene_id_clean[explain_idx],
                     symbol = test_data$symbol[explain_idx],
                     y = test_data$y[explain_idx], pred = pred), shap_vals),
    file.path(out_dir, paste0("shapreg_per_gene_", suffix, ".csv")))
  if (exists("signed"))    reg_write_csv(signed, file.path(out_dir, paste0("shapreg_signed_top_", suffix, ".csv")))
  if (exists("cg"))        reg_write_csv(cg, file.path(out_dir, paste0("shapreg_corr_groups_", suffix, ".csv")))
  if (exists("wide"))      reg_write_csv(wide, file.path(out_dir, paste0("shapreg_explain_stability_", suffix, ".csv")))
  if (exists("subgroups")) reg_write_csv(subgroups, file.path(out_dir, paste0("shapreg_subgroups_", suffix, ".csv")))
  if (exists("band"))      reg_write_csv(band, file.path(out_dir, paste0("shapreg_subgroup_null_band_", suffix, ".csv")))
}
cat("Wrote SHAP outputs with suffix:", suffix, "\n")
```
