---
title: "`r paste0('Regression SHAP stability: ', params$preset, ' / ', params$arm)`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  preset: "transform"
  arm: "raw"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "11f_shap_regression_stability.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It reads the SHAP outputs that `11d_shap_regression_seed_sweep.R` produced across 20 splits and
reports how much a family share and a feature rank move when nothing biological changes - the same
10,172 genes every run, a different 80% used for training, the forest seed fixed.

That spread is the threshold a single run's number has to clear before it means anything.

**Decisions**

- Measure this separately from the classifier's floor rather than reusing it. The two come from
  different sources: the classifier's from which negatives were drawn (about 52% gene overlap),
  this one from which rows train (about 81% overlap). This one is expected to be tighter, and the
  ratio is reported rather than assumed.
- Express the threshold as `2.8 * sd`, matching how the classifier's floor was built so `12_` can
  combine them.
- Report the threshold per family, not as one number.

```{r libraries}
library(tidyverse)
library(here)
```

```{r load}
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
source(here("code", "translational_control_overview", "10_regression_presets.R"))
out_dir <- here("output", "translational_control_overview")

stopifnot("unknown preset" = params$preset %in% names(REG_PRESETS))
arm <- REG_PRESETS[[params$preset]][[params$arm]]
stopifnot("unknown arm for this preset" = !is.null(arm))

feat <- map_dfr(REG_SPLIT_SEEDS, function(s) {
  f <- file.path(out_dir, paste0("shapreg_feature_", reg_suffix(arm, s), ".csv"))
  if (!file.exists(f)) return(NULL)
  read_csv(f, show_col_types = FALSE) %>% mutate(split_seed = s)
})
stopifnot("no shapreg_feature_* files found - run 11d_shap_regression_seed_sweep.R first" =
            nrow(feat) > 0)
n_seeds <- n_distinct(feat$split_seed)
cat("Seeds:", n_seeds, "| features per run:", n_distinct(feat$feature), "\n")
stopifnot("the feature set changed across the sweep" =
            n_distinct(table(feat$split_seed)) == 1)
```

# 1. Family share floor

Reports each family's share across the 20 splits, and compares the spread to the classifier's.

**Decisions**

- Prefer the empirical 5th-to-95th percentile interval over a normal-theory one, since shares are
  bounded below at zero and skewed for the small families.


```{r family_floor}
fam <- feat %>% group_by(split_seed, family) %>%
  summarise(share = sum(mean_abs_shap), .groups = "drop_last") %>%
  group_by(split_seed) %>% mutate(share = 100 * share / sum(share)) %>% ungroup()

floor_tbl <- fam %>% group_by(family) %>%
  summarise(n = n(), median = median(share), sd = sd(share),
            lo = quantile(share, 0.05), hi = quantile(share, 0.95),
            cv_pct = 100 * sd / median,
            # 2.8 * sd, matching shap_noise_floor.csv's construction exactly so the two floors
            # can be combined in 12_. Two single-run experiments each carry their own draw
            # noise, so the relevant scale is ~sqrt(2) x sd; 2.8 is that, rounded up.
            min_detectable_diff = 2.8 * sd, .groups = "drop") %>%
  arrange(desc(median))
print(as.data.frame(floor_tbl %>% mutate(across(where(is.numeric), ~ round(., 3)))),
      row.names = FALSE)
cat("\nThe threshold is PER FAMILY, not one number: ",
    sprintf("%s needs %.2f pp, %s only %.2f pp.\n",
            floor_tbl$family[which.max(floor_tbl$min_detectable_diff)],
            max(floor_tbl$min_detectable_diff),
            floor_tbl$family[which.min(floor_tbl$min_detectable_diff)],
            min(floor_tbl$min_detectable_diff)))

# Compare against the classification floor, if present. The expectation is that this one is
# tighter, because the split reuses ~81% of rows where the draw reused ~52% of genes.
clf_f <- file.path(out_dir, "shap_noise_floor.csv")
if (file.exists(clf_f)) {
  cmp <- read_csv(clf_f, show_col_types = FALSE) %>%
    dplyr::select(family, sd_clf = sd, mdd_clf = min_detectable_diff) %>%
    inner_join(floor_tbl %>% dplyr::select(family, sd_reg = sd, mdd_reg = min_detectable_diff),
               by = "family") %>%
    mutate(ratio = sd_reg / sd_clf) %>% arrange(ratio)
  print(as.data.frame(cmp %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)
  cat(sprintf("\nregression sd / classifier sd: median %.2f over %d families (< 1 means tighter)\n",
              median(cmp$ratio), nrow(cmp)))
}
```

```{r fig_floor, fig.width = 9, fig.height = 5}
p <- fam %>% mutate(family = fct_reorder(family, share, .fun = median)) %>%
  ggplot(aes(share, family)) +
  geom_point(size = 1.6, alpha = 0.55, colour = "#E69F00",
             position = position_jitter(height = 0.12, seed = 9)) +
  stat_summary(fun = median, geom = "point", size = 2.6, colour = "black") +
  labs(x = "family share of total |SHAP| (%)", y = NULL,
       title = "What a family share is worth with nothing biological varying",
       subtitle = sprintf("one point per split seed, n = %d seeds | black = median", n_seeds)) +
  theme_classic(base_size = 12)
print(p)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("11f_family_floor_", params$preset, "_", params$arm, ".pdf")),
         p, width = 9, height = 5)
```

# 2. Rank stability and direction

Reports how stable the feature ranking is across splits, and which features' directions can be
read at all.

**Decisions**

- Compute the within-model top-20 overlap, which `12_` needs in order to interpret its cross-model
  number.
- Name the features that are top-20 in every split, since those are the ones that can be ranked
  off a single run.
- Flag features whose direction flips sign across splits. Those have no interpretable direction and
  should be grouped rather than dropped wherever the decomposition has to keep adding up.


```{r rank_stability}
# 12_ needs this number: a CROSS-model top-20 Jaccard means little until you know how well each
# model agrees with itself across its own replication axis.
tops <- feat %>% group_by(split_seed) %>% slice_min(rank_shap, n = 20) %>%
  summarise(f = list(feature), .groups = "drop")
pairs <- combn(nrow(tops), 2)
jac <- apply(pairs, 2, function(ij) {
  a <- tops$f[[ij[1]]]; b <- tops$f[[ij[2]]]
  length(intersect(a, b)) / length(union(a, b))
})
always <- feat %>% filter(rank_shap <= 20) %>% count(feature) %>% filter(n == n_seeds)
cat(sprintf("top-20 Jaccard across seed pairs: median %.3f (range %.3f-%.3f, %d pairs)\n",
            median(jac), min(jac), max(jac), length(jac)))
cat(sprintf("features top-20 in EVERY seed: %d of 20\n", nrow(always)))
cat("  ", paste(always$feature, collapse = ", "), "\n")
cat("\nRank a feature off a single run only if it clears that bar.\n")
```

```{r direction_stability}
# A feature whose shap_value_cor sign flips across seeds has no interpretable direction. 03f_
# found 14 of 69 such features for the classifier; the same test here says which regression
# features can be read directionally at all.
dir_tbl <- feat %>% group_by(feature, family) %>%
  summarise(n = n(), frac_pos = mean(shap_value_cor > 0, na.rm = TRUE),
            median_cor = median(shap_value_cor, na.rm = TRUE), .groups = "drop") %>%
  arrange(abs(frac_pos - 0.5))
unstable <- dir_tbl %>% filter(frac_pos > 0.05, frac_pos < 0.95)
cat(sprintf("sign-unstable features (0.05 < frac_pos < 0.95): %d of %d\n",
            nrow(unstable), nrow(dir_tbl)))
if (nrow(unstable))
  print(as.data.frame(unstable %>% mutate(across(where(is.numeric), ~ round(., 3)))),
        row.names = FALSE)
cat("\nDirection is uninterpretable for these. They must be lumped, not deleted, wherever\n")
cat("additivity has to survive.\n")
```

```{r export}
reg_write_csv(floor_tbl, file.path(out_dir,
  paste0("shapreg_noise_floor_", params$preset, "_", params$arm, ".csv")))
reg_write_csv(dir_tbl, file.path(out_dir,
  paste0("shapreg_direction_stability_", params$preset, "_", params$arm, ".csv")))
reg_write_csv(tibble(arm = params$arm, preset = params$preset,
                     median_jaccard = median(jac), min_jaccard = min(jac),
                     max_jaccard = max(jac), n_pairs = length(jac),
                     always_top20 = nrow(always)),
  file.path(out_dir, paste0("shapreg_top20_jaccard_", params$preset, ".csv")))
# Also write the arm-agnostic name 10c_/12_ look for by default.
reg_write_csv(floor_tbl %>% mutate(arm = params$arm),
  file.path(out_dir, paste0("shapreg_noise_floor_", params$preset, ".csv")))
cat("Wrote the regression SHAP floor for", params$preset, "/", params$arm, "\n")
```
