---
title: "Regression versus classifier: are the SHAP values consistent?"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    # Code is folded by default with an "unfold" control on every block, and a "Code" menu at the
    # top right to expand everything at once. The commentary in these chunks is long on purpose,
    # so the default view should be the prose, the tables and the figures.
    code_folding: hide
params:
  clf_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tco_corrected_both_clipaggregate"
  reg_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  reg_sweep_preset: "transform"
  reg_sweep_arm: "raw"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "12_regression_vs_classifier.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It compares the feature importance the regression produced against the feature importance the
classifier produced, to see whether the two models point at the same features.

It reads saved CSVs from both and fits nothing.

**Decisions**

- Compare shares and ranks only, never raw SHAP magnitudes. The classifier's SHAP is in units of
  probability and the regression's is in log2 TE fold change, so the raw numbers are not
  commensurable. The raw column is not joined at all.
- Require both models to have used the same features, and stop if they did not. Otherwise a
  difference could be a difference in feature sets rather than in models.
- Judge every difference against a measured noise floor rather than against zero.

## What counts as a real difference

Each model's share carries its own run-to-run variation, and the two sources are different: the
classifier's comes from which negatives were drawn (about 52% gene overlap between draws), the
regression's from which genes were held out (about 81% row overlap). So a difference has to clear
both.

**Decisions**

- Combine the two as `2.8 * sqrt(sd_clf^2 + sd_reg^2)`, per family. The saved floor files store
  `2.8 * sd`, so that factor is divided back out before combining.
- If only one floor is available, fall back to it and label the threshold as understating, so a
  verdict based on it is visibly provisional.

## Three things differ between the two models at once

| | classifier | regression |
|---|---|---|
| what SHAP explains | a substitute regression forest on 0/1 labels | the model itself |
| outcome | binary, thresholded | continuous |
| genes | 2,872 balanced (1,436 + 1,436) | 10,172 |

Because all three differ together, agreement could mean the model type does not matter, or it
could mean two separate effects cancelled.

**Decisions**

- Include a third arm that separates them: the regression refit on exactly the classifier's 2,872
  genes. Comparing it to the full regression isolates the change in gene population; comparing it
  to the classifier isolates the change in outcome type.
- Report section 4 as open rather than concluded when that arm has not been run.

```{r libraries}
library(tidyverse)
library(here)
library(cowplot)
```

```{r load}
source(here("code", "translational_control_overview", "02_sweep_presets.R"))
source(here("code", "translational_control_overview", "10_regression_presets.R"))
out_dir <- here("output", "translational_control_overview")

need <- c(clf_fam  = file.path(out_dir, paste0("shap_family_",     params$clf_suffix, ".csv")),
          clf_feat = file.path(out_dir, paste0("shap_feature_",    params$clf_suffix, ".csv")),
          reg_fam  = file.path(out_dir, paste0("shapreg_family_",  params$reg_suffix, ".csv")),
          reg_feat = file.path(out_dir, paste0("shapreg_feature_", params$reg_suffix, ".csv")))
missing <- need[!file.exists(need)]
if (length(missing)) {
  cat("MISSING inputs:\n"); for (m in missing) cat("  ", m, "\n")
  cat("\nRun 03_shap.Rmd (classifier) and 11_shap_regression.Rmd (regression) first.\n")
}
stopifnot("required SHAP outputs are missing - see the list above" = length(missing) == 0)

clf_fam  <- read_csv(need[["clf_fam"]],  show_col_types = FALSE)
clf_feat <- read_csv(need[["clf_feat"]], show_col_types = FALSE)
reg_fam  <- read_csv(need[["reg_fam"]],  show_col_types = FALSE)
reg_feat <- read_csv(need[["reg_feat"]], show_col_types = FALSE)

# The precondition that makes this a comparison of model type rather than of feature sets.
stopifnot(
  "the two models do not share a feature set - this comparison would not be about model type" =
    setequal(clf_feat$feature, reg_feat$feature),
  "the two models' family sets differ" = setequal(clf_fam$family, reg_fam$family))
cat("Both models:", nrow(clf_feat), "features |", nrow(clf_fam), "families\n")
cat("Classifier SHAP units: probability | Regression SHAP units: log2 TE LFC\n")
cat("=> only shares and ranks are joined below; the raw mean_abs_shap column is not.\n")
```

```{r floors}
# Classification floor: measured over 20 neg_seed draws by 03f_.
clf_floor_path <- file.path(out_dir, "shap_noise_floor.csv")
# Regression floor: measured over 20 split_seeds by 10c_ (needs 11d_ to have been run).
reg_floor_path <- file.path(out_dir,
  paste0("shapreg_noise_floor_", params$reg_sweep_preset, ".csv"))

clf_floor <- if (file.exists(clf_floor_path))
  read_csv(clf_floor_path, show_col_types = FALSE) %>%
    dplyr::select(family, sd_clf = sd) else NULL
reg_floor <- if (file.exists(reg_floor_path))
  read_csv(reg_floor_path, show_col_types = FALSE) %>%
    filter(arm == params$reg_sweep_arm) %>% dplyr::select(family, sd_reg = sd) else NULL

cat("classification floor:", if (is.null(clf_floor)) "MISSING" else "present", "\n")
cat("regression floor:    ", if (is.null(reg_floor)) "MISSING (run 11d_ then 10c_)" else "present", "\n")
if (is.null(reg_floor))
  cat("\nWithout the regression floor, deltas are judged against the CLASSIFIER's noise alone,\n",
      "which UNDERSTATES the threshold. Any 'MOVED' verdict below is provisional until 11d_ runs.\n")
```

# 1. Family shares

Compares the two models' family-level shares, with each difference judged against its own floor.

**Decisions**

- Report Spearman and Pearson agreement across families, plus a per-family verdict.
- Draw the floor on both figures as error bars, rather than quoting it in text only.


```{r family_compare}
fam <- clf_fam %>% dplyr::select(family, share_clf = share_pct) %>%
  inner_join(reg_fam %>% dplyr::select(family, share_reg = share_pct), by = "family") %>%
  mutate(delta = share_reg - share_clf)
if (!is.null(clf_floor)) fam <- fam %>% left_join(clf_floor, by = "family")
if (!is.null(reg_floor)) fam <- fam %>% left_join(reg_floor, by = "family")
if (!"sd_clf" %in% names(fam)) fam$sd_clf <- NA_real_
if (!"sd_reg" %in% names(fam)) fam$sd_reg <- NA_real_

fam <- fam %>%
  mutate(
    # Combined floor. With only one side measured, fall back to that side and say so.
    mdd = case_when(
      !is.na(sd_clf) & !is.na(sd_reg) ~ 2.8 * sqrt(sd_clf^2 + sd_reg^2),
      !is.na(sd_clf)                  ~ 2.8 * sd_clf,
      TRUE                            ~ NA_real_),
    floor_basis = case_when(
      !is.na(sd_clf) & !is.na(sd_reg) ~ "combined",
      !is.na(sd_clf)                  ~ "classifier only (understates)",
      TRUE                            ~ "none"),
    verdict = case_when(is.na(mdd)         ~ "no floor",
                        abs(delta) > mdd   ~ "MOVED",
                        TRUE               ~ "within the floor")) %>%
  arrange(desc(share_clf))
print(as.data.frame(fam %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)

rho <- cor(fam$share_clf, fam$share_reg, method = "spearman")
pea <- cor(fam$share_clf, fam$share_reg)
cat(sprintf("\nfamily share agreement: Spearman %.4f | Pearson %.4f | n = %d families\n",
            rho, pea, nrow(fam)))
cat("families beyond their floor:", sum(fam$verdict == "MOVED"), "of",
    sum(fam$verdict != "no floor"), "\n")
if (any(fam$verdict == "MOVED"))
  cat("  ", paste(sprintf("%s %+.2f pp", fam$family[fam$verdict == "MOVED"],
                          fam$delta[fam$verdict == "MOVED"]), collapse = " | "), "\n")
```

```{r fig_family, fig.width = 11, fig.height = 5}
p_sc <- ggplot(fam, aes(share_clf, share_reg)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbar(aes(ymin = share_reg - mdd, ymax = share_reg + mdd),
                width = 0, colour = "grey70", na.rm = TRUE) +
  geom_point(aes(colour = verdict), size = 3) +
  ggrepel::geom_text_repel(aes(label = family), size = 3, max.overlaps = 20, seed = 9) +
  scale_colour_manual(values = c("MOVED" = "#D55E00", "within the floor" = "grey45",
                                 "no floor" = "#0072B2"), name = NULL) +
  labs(x = "classifier SHAP share (%)", y = "regression SHAP share (%)",
       title = "(a) Family shares, with the floor drawn",
       subtitle = sprintf("Spearman %.3f | error bars = combined min detectable difference | n = %d families",
                          rho, nrow(fam))) +
  theme_classic(base_size = 12)

# The slope view: which families gained and which lost, against their own floor.
p_sl <- fam %>% mutate(family = fct_reorder(family, delta)) %>%
  ggplot(aes(delta, family)) +
  geom_vline(xintercept = 0, colour = "grey50") +
  geom_errorbarh(aes(xmin = -mdd, xmax = mdd), height = 0.35, colour = "grey80",
                 linewidth = 3, alpha = 0.5, na.rm = TRUE) +
  geom_point(aes(colour = verdict), size = 3) +
  scale_colour_manual(values = c("MOVED" = "#D55E00", "within the floor" = "grey45",
                                 "no floor" = "#0072B2"), guide = "none") +
  labs(x = "regression share minus classifier share (pp)", y = NULL,
       title = "(b) Change, against each family's own floor",
       subtitle = "grey band is the noise floor; a point inside it has not moved") +
  theme_classic(base_size = 12)

pg <- plot_grid(p_sc, p_sl, nrow = 1, rel_widths = c(1.05, 1))
print(pg)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("12_family_shares_", params$reg_suffix, ".pdf")),
         pg, width = 11, height = 5)
```

# 2. Feature level: ranks and direction

Repeats the comparison one feature at a time, and checks whether the two models agree on the
direction of each feature's effect.

**Decisions**

- Read the cross-model top-20 overlap against how well each model agrees with itself across its
  own repeat runs. A cross-model number alone has nothing to be compared to.
- Test direction agreement twice: over all features, and over only those whose direction is stable
  across repeat runs. Features whose sign flips between runs have no direction to agree about, and
  including them manufactures disagreement.


```{r feature_compare}
ft <- clf_feat %>%
  mutate(share_clf = 100 * mean_abs_shap / sum(mean_abs_shap)) %>%
  dplyr::select(feature, family, share_clf, rank_clf = rank_shap, cor_clf = shap_value_cor) %>%
  inner_join(reg_feat %>%
               mutate(share_reg = 100 * mean_abs_shap / sum(mean_abs_shap)) %>%
               dplyr::select(feature, share_reg, rank_reg = rank_shap, cor_reg = shap_value_cor),
             by = "feature")
cat(sprintf("feature rank Spearman %.4f | share Pearson %.4f | n = %d\n",
            cor(ft$rank_clf, ft$rank_reg, method = "spearman"),
            cor(ft$share_clf, ft$share_reg), nrow(ft)))

jac <- function(a, b, k = 20)
  length(intersect(head(a, k), head(b, k))) / length(union(head(a, k), head(b, k)))
cross_j <- jac(ft$feature[order(ft$rank_clf)], ft$feature[order(ft$rank_reg)])

# The framing that answers the question. A cross-model Jaccard means little alone: it has to be
# read against how well each model agrees with itself across its own replication axis. If
# cross-model agreement is comparable to within-model agreement, the two models agree as well as
# either agrees with itself, which is the strongest form "consistent" can take.
within_clf <- NA_real_; within_reg <- NA_real_
f_clf <- file.path(out_dir, "shap_noise_floor.csv")   # 03f_ reports 0.818 for the classifier
if (file.exists(f_clf)) within_clf <- 0.818            # from 03f_'s top-20 Jaccard
reg_stab <- file.path(out_dir, paste0("shapreg_top20_jaccard_", params$reg_sweep_preset, ".csv"))
if (file.exists(reg_stab))
  within_reg <- read_csv(reg_stab, show_col_types = FALSE) %>%
    filter(arm == params$reg_sweep_arm) %>% pull(median_jaccard) %>% first()

cat(sprintf("\ntop-20 Jaccard:  cross-model %.3f | within-classifier %s | within-regression %s\n",
            cross_j,
            if (is.na(within_clf)) "n/a" else sprintf("%.3f", within_clf),
            if (is.na(within_reg)) "n/a (run 11d_)" else sprintf("%.3f", within_reg)))
if (!is.na(within_clf) && cross_j >= within_clf * 0.9)
  cat("=> cross-model agreement is comparable to the classifier's agreement with itself.\n")
```

```{r direction_compare}
# Direction is uninterpretable for features whose shap_value_cor sign flips across replicates.
# 03f_ found 14 of 69 such features for the classifier. Including them manufactures
# disagreement, so the sign test is reported over all features and over the stable subset.
stab_path <- file.path(out_dir, "shap_direction_stability.csv")
sign_all <- mean(sign(ft$cor_clf) == sign(ft$cor_reg), na.rm = TRUE)
cat(sprintf("shap_value_cor sign agreement, ALL features: %.3f (%d of %d)\n",
            sign_all, sum(sign(ft$cor_clf) == sign(ft$cor_reg), na.rm = TRUE), nrow(ft)))
if (file.exists(stab_path)) {
  ds <- read_csv(stab_path, show_col_types = FALSE)
  unstable <- ds$feature[ds$frac_pos > 0.05 & ds$frac_pos < 0.95]
  cat("sign-unstable in the classifier (03f_):", length(unstable), "features\n")
  fs <- ft %>% filter(!feature %in% unstable)
  agree <- sum(sign(fs$cor_clf) == sign(fs$cor_reg), na.rm = TRUE)
  bt <- binom.test(agree, nrow(fs), p = 0.5)
  cat(sprintf("sign agreement on the STABLE subset: %.3f (%d of %d), binomial p = %.3g\n",
              agree / nrow(fs), agree, nrow(fs), bt$p.value))
  disagree <- fs %>% filter(sign(cor_clf) != sign(cor_reg)) %>%
    dplyr::select(feature, family, cor_clf, cor_reg, rank_clf, rank_reg)
  if (nrow(disagree)) {
    cat("\nstable features whose DIRECTION disagrees between the two models:\n")
    print(as.data.frame(disagree %>% mutate(across(where(is.numeric), ~ round(., 3)))),
          row.names = FALSE)
  }
} else cat("shap_direction_stability.csv not found - stable-subset test skipped\n")
```

```{r fig_feature, fig.width = 11, fig.height = 5}
p_rk <- ggplot(ft, aes(rank_clf, rank_reg)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = family), size = 2, alpha = 0.85) +
  scale_colour_viridis_d(guide = "none", end = 0.9) +
  ggrepel::geom_text_repel(
    data = ft %>% filter(abs(rank_clf - rank_reg) > 15 | pmin(rank_clf, rank_reg) <= 5),
    aes(label = feature), size = 2.6, max.overlaps = 25, seed = 9) +
  labs(x = "classifier SHAP rank", y = "regression SHAP rank",
       title = "(a) Feature ranks",
       subtitle = sprintf("Spearman %.3f | labelled: top 5 or moved >15 ranks | n = %d",
                          cor(ft$rank_clf, ft$rank_reg, method = "spearman"), nrow(ft))) +
  theme_classic(base_size = 12)

p_dir <- ggplot(ft, aes(cor_clf, cor_reg)) +
  geom_hline(yintercept = 0, colour = "grey70") + geom_vline(xintercept = 0, colour = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = sign(cor_clf) == sign(cor_reg)), size = 2, alpha = 0.85) +
  scale_colour_manual(values = c("TRUE" = "grey40", "FALSE" = "#D55E00"),
                      labels = c("TRUE" = "same sign", "FALSE" = "opposite"), name = NULL) +
  labs(x = "classifier shap_value_cor", y = "regression shap_value_cor",
       title = "(b) Direction, the interpretable quantity",
       subtitle = sprintf("sign agreement %.1f%% over all %d features", 100 * sign_all, nrow(ft))) +
  theme_classic(base_size = 12)

pg2 <- plot_grid(p_rk, p_dir, nrow = 1)
print(pg2)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("12_feature_compare_", params$reg_suffix, ".pdf")),
         pg2, width = 11, height = 5)
```

# 3. The AUC comparison, as distributions

Compares the two models' AUCs as distributions over their repeat runs rather than as two numbers.

Only about 112 genes were held out by both models, where the confidence interval is roughly plus
or minus 0.08 - too wide to separate 0.83 from 0.78. Both models do have 20 replicates saved, so
the comparison is made there instead.

**Decisions**

- Compare distributions, not point estimates.
- Leave them unpaired. The two axes are different randomisations, so there is no meaningful
  pairing between a negative draw and a split.
- State plainly that the two AUCs measure related but not identical things: the classifier's is its
  own task, the regression's is against labels it never saw.


```{r auc_distributions}
# A point-vs-point AUC comparison is not answerable at this sample size: only ~112 genes are held
# out by both models, where the 95% CI half-width is ~0.08. Both models do however have a
# 20-replicate distribution on disk, so the comparison is distribution-vs-distribution.
# Unpaired, deliberately: the two axes are different randomisations (negative draw vs split), so
# there is no meaningful pairing between a neg_seed and a split_seed.
clf_perf <- map_dfr(c(9L, 1:20), function(s) {
  suf <- sweep_suffix(list(feature_set_label = "tco"), s)
  f <- file.path(out_dir, paste0("rf_performance_", suf, ".csv"))
  if (!file.exists(f)) return(NULL)
  d <- read_csv(f, show_col_types = FALSE)
  tibble(model = "classifier (neg_seed)", seed = s, auc = d$value[d$metric == "test_auc"])
}) %>% distinct(seed, .keep_all = TRUE)

reg_sweep_path <- file.path(out_dir, paste0("rfreg_sweep_", params$reg_sweep_preset, ".csv"))
reg_perf <- if (file.exists(reg_sweep_path))
  read_csv(reg_sweep_path, show_col_types = FALSE) %>%
    filter(arm == params$reg_sweep_arm) %>%
    transmute(model = "regression (split_seed)", seed = split_seed, auc = bridge) else NULL

if (is.null(reg_perf) || nrow(clf_perf) == 0) {
  cat("Need both sweeps. classifier runs found:", nrow(clf_perf),
      "| regression sweep:", !is.null(reg_perf), "\n")
} else {
  auc_df <- bind_rows(clf_perf, reg_perf)
  print(as.data.frame(auc_df %>% group_by(model) %>%
    summarise(n = n(), median = median(auc), sd = sd(auc),
              p05 = quantile(auc, .05), p95 = quantile(auc, .95), .groups = "drop") %>%
    mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)
  wt <- wilcox.test(auc ~ model, data = auc_df)
  cat(sprintf("\nunpaired Wilcoxon p = %.4g\n", wt$p.value))
  cat("NOTE: the classifier's AUC is its own task (separating its two classes); the regression's\n")
  cat("is the bridge AUC on labels it never saw. Comparable in scale, not identical in meaning.\n")
}
```

```{r fig_auc, eval = exists("auc_df"), fig.width = 7, fig.height = 4.4}
p_auc <- ggplot(auc_df, aes(auc, model, colour = model)) +
  geom_point(position = position_jitter(height = 0.12, seed = 9), size = 2, alpha = 0.75) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, linewidth = 0.4, colour = "black") +
  scale_colour_manual(values = c("classifier (neg_seed)" = "#0072B2",
                                 "regression (split_seed)" = "#E69F00"), guide = "none") +
  labs(x = "AUC", y = NULL, title = "Two replication axes, two distributions",
       subtitle = sprintf("n = %d classifier draws, %d regression splits. Unpaired: the axes are different randomisations.",
                          sum(auc_df$model == "classifier (neg_seed)"),
                          sum(auc_df$model == "regression (split_seed)"))) +
  theme_classic(base_size = 12)
print(p_auc)
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("12_auc_distributions_", params$reg_suffix, ".pdf")),
         p_auc, width = 7, height = 4.4)
```

# 4. Separating gene population from model type

Uses the restricted-gene arm to attribute the family share differences to their causes.

**Decisions**

- Split each difference into a population part and a model part, and report them next to the total.
- Print the commands to build the arm when it is missing, rather than failing.


```{r population_arm}
# The regression refit on exactly the classifier's 2,872 genes. Comparing it to the 10,172-gene
# regression isolates gene population; comparing it to the classifier isolates outcome type plus
# the surrogate. Without this arm, "the two agree" cannot be told from "two effects cancelled".
clfgene_suffix <- reg_suffix(REG_PRESETS$population$clfgene, 9L)
clfgene_path   <- file.path(out_dir, paste0("shapreg_family_", clfgene_suffix, ".csv"))
if (!file.exists(clfgene_path)) {
  cat("The clfgenes arm has not been run. To build it:\n")
  cat("  rmarkdown::render('10_rf_regression.Rmd', params=list(\n")
  cat("      restrict_to_classifier_genes=TRUE, feature_set_label='tcoreg_pop'))\n")
  cat("  rmarkdown::render('11_shap_regression.Rmd', params=list(\n")
  cat("      restrict_to_classifier_genes=TRUE, feature_set_label='tcoreg_pop'))\n")
} else {
  cg_fam <- read_csv(clfgene_path, show_col_types = FALSE) %>%
    dplyr::select(family, share_clfgene = share_pct)
  three <- fam %>% dplyr::select(family, share_clf, share_reg, mdd) %>%
    left_join(cg_fam, by = "family") %>%
    # Oriented the same way as section 1's delta = share_reg - share_clf, so a family that gained
    # share in the regression is positive in BOTH tables. It previously ran the other way
    # (share_clf - share_reg) while the caption claimed to repeat section 1's number, which made
    # total the NEGATION of the quantity it was presented as. This repo has a documented history
    # of contrast signs that disagree with their own labels, so the definition is what moved.
    #
    # Algebraically, reading classifier -> regression:
    #   model_effect = share_clfgene - share_clf   swap the classifier for the regression on the
    #                                              classifier's 2,872 genes (gene set held fixed)
    #   pop_effect   = share_reg     - share_clfgene   then widen from 2,872 to all 10,172 genes
    #   sum          = share_reg     - share_clf   = total = section 1's delta, exactly.
    mutate(model_effect = share_clfgene - share_clf,   # classifier vs regression, same genes
           pop_effect   = share_reg     - share_clfgene, # 2,872 -> 10,172 genes
           total        = share_reg     - share_clf) %>%
    arrange(desc(abs(total)))
  print(as.data.frame(three %>% mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
  # The decomposition is exact by construction, but assert it rather than trusting the arithmetic:
  # a future edit to one component is exactly the kind of change that leaves the caption true and
  # the table wrong.
  stopifnot("the population/model decomposition does not sum to the total" =
              all(abs(three$pop_effect + three$model_effect - three$total) < 1e-8, na.rm = TRUE))
  cat("\nmodel_effect = swapping the classifier for the regression ON THE CLASSIFIER'S GENES\n")
  cat("pop_effect   = then widening from the classifier's genes to all of them\n")
  cat("total        = the raw difference reported in section 1, share_reg - share_clf\n")
  cat("               (model + pop, exactly - asserted above, not merely up to rounding)\n")
  cat(sprintf("\nlargest population effect: %s %+.2f pp | largest model effect: %s %+.2f pp\n",
              three$family[which.max(abs(three$pop_effect))],
              three$pop_effect[which.max(abs(three$pop_effect))],
              three$family[which.max(abs(three$model_effect))],
              three$model_effect[which.max(abs(three$model_effect))]))
}
```

# 5. Summary

States which families moved beyond their floor and which did not.

**Decisions**

- Repeat the caveat about a missing regression floor here, so a provisional verdict is not read as
  settled.


```{r summary}
cat(sprintf("Family shares agree at Spearman %.3f (Pearson %.3f) over %d families.\n",
            rho, pea, nrow(fam)))
cat(sprintf("Feature ranks agree at Spearman %.3f over %d features; top-20 Jaccard %.3f.\n",
            cor(ft$rank_clf, ft$rank_reg, method = "spearman"), nrow(ft), cross_j))
moved <- fam %>% filter(verdict == "MOVED")
if (nrow(moved)) {
  cat(sprintf("\n%d famil%s moved beyond %s floor:\n", nrow(moved),
              if (nrow(moved) == 1) "y" else "ies",
              if (all(moved$floor_basis == "combined")) "the combined" else "its"))
  for (i in seq_len(nrow(moved)))
    cat(sprintf("  %-20s %+6.2f pp  (floor %.2f, basis: %s)\n",
                moved$family[i], moved$delta[i], moved$mdd[i], moved$floor_basis[i]))
  cat("\nThese are NOT noise. Section 4 attributes them to gene population vs model type when\n")
  cat("the clfgenes arm is present; otherwise the attribution is open.\n")
} else {
  cat("\nNo family moved beyond its floor: the two models agree within replication noise.\n")
}
if (any(fam$floor_basis == "classifier only (understates)"))
  cat("\nCAVEAT: the regression floor is missing, so every threshold above is too LENIENT.\n",
      "Run 11d_ then 10c_ before treating any 'MOVED' verdict as settled.\n")
```

```{r export}
reg_write_csv(fam, file.path(out_dir, paste0("regvsclf_family_", params$reg_suffix, ".csv")))
reg_write_csv(ft,  file.path(out_dir, paste0("regvsclf_feature_", params$reg_suffix, ".csv")))
if (exists("auc_df"))
  reg_write_csv(auc_df, file.path(out_dir, paste0("regvsclf_auc_", params$reg_suffix, ".csv")))
if (exists("three"))
  reg_write_csv(three, file.path(out_dir, paste0("regvsclf_population_", params$reg_suffix, ".csv")))
cat("Wrote comparison outputs.\n")
```
