---
title: "`r paste0('Linear regression on TE LFC, and how it compares to the forest (eIF', params$gene, ', ', params$condition, ', ', params$timepoint, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
params:
  rf_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  gene: "3d"
  condition: "hypoxia"
  timepoint: "1hr"
  lfc: "0.5"
  direction: "promotes"
  readout: "TE"
  drop_for_rank: "term_stop_TAA"
  n_spline_features: 8
  spline_df: 4
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "13_linear_regression.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It fits a linear model to the same outcome, on the same genes, with the same features as the
random forest in `10_rf_regression.Rmd`, and asks whether the forest is earning its complexity.

Three arms:

- **OLS** - one coefficient per feature, with standard errors, p-values and VIF. This is the
  readable model.
- **Lasso** - the same features with an L1 penalty, which answers how few of them are actually
  needed.
- **OLS plus splines** - natural splines on the strongest continuous features, which answers
  whether the outcome is non-linear in them.

**Decisions**

- Load `10_`'s saved split rather than rebuilding it, so both models see byte-identical training
  rows, held-out rows and imputed feature values. Any difference in performance is the model.
- Drop one column, ``r params$drop_for_rank``, to make the design full rank (see below). Keep
  every other feature, including the correlated ones, so the comparison with the forest stays a
  comparison of models rather than of feature sets.
- Report coefficients both raw and standardized. Raw reads in natural units; standardized (per SD
  of the feature) is the only version whose magnitudes are comparable between features.
- Print VIF next to every coefficient, so features whose individual estimate is imprecise are
  visible rather than silently over-read.
- Score both models by Spearman against the raw held-out outcome, the same yardstick `10_` uses.

## Why one column has to go

Two feature groups encode related things: `term_plus4_A/C/G/T` (the base after the stop codon)
and `term_stop_TAA/TAG/TGA` (which stop codon). For most genes exactly one dummy in each group is
1, but for the genes whose CDS has no annotated stop **all seven are 0**. Both groups therefore sum
to the same indicator - "has an annotated stop" - which makes one column an exact linear
combination of the others:

```
plus4_T = (stop_TAA + stop_TAG + stop_TGA) - plus4_A - plus4_C - plus4_G
```

A forest is unaffected: a redundant column simply never improves a split, so it is ignored. Linear
regression is not, because it solves for one coefficient per column, and when one column is an
exact combination of the rest there are infinitely many coefficient sets with identical
predictions. The coefficients are not identifiable, and `lm()` silently drops a column and reports
`NA` for it.

Dropping one column removes the dependency and loses nothing, since that column's information is
fully present in the others. `r params$drop_for_rank` is the reference here, so the remaining
dummies in its group read as effects *relative to* it.

```{r libraries}
library(tidyverse)
library(here)
library(glmnet)
library(splines)
library(cowplot)
```

```{r load}
out_dir <- here("output", "translational_control_overview")
md_path <- file.path(out_dir, paste0("rfreg_model_data_", params$rf_suffix, ".rds"))
stopifnot("run 10_rf_regression.Rmd first - its saved split is what this notebook re-uses" =
            file.exists(md_path))
md <- readRDS(md_path)
train <- md$train; test <- md$test
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)

# One column dropped, everything else kept.
stopifnot("drop_for_rank is not in the feature set" =
            params$drop_for_rank %in% md$feature_cols)
fc <- setdiff(md$feature_cols, params$drop_for_rank)

Xtr <- as.matrix(train[, fc]); ytr <- train$y
Xte <- as.matrix(test[, fc]);  yte <- test$y

# The design must now be full rank, or the coefficients are still not identifiable. Asserted
# rather than hoped for, because lm() would otherwise proceed and return NA coefficients.
qr_rank <- qr(cbind(1, Xtr))$rank
cat(sprintf("design: %d columns (incl. intercept), rank %d\n", ncol(Xtr) + 1, qr_rank))
stopifnot("the design is still rank-deficient after the drop - find the remaining dependency" =
            qr_rank == ncol(Xtr) + 1)
cat(sprintf("train %d genes | test %d genes | %d features (dropped %s)\n",
            nrow(Xtr), nrow(Xte), ncol(Xtr), params$drop_for_rank))
ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)
```

# 1. OLS

```{r ols}
ols <- lm(ytr ~ ., data = as.data.frame(Xtr))
stopifnot("OLS still dropped a column - the design is not full rank" =
            !any(is.na(coef(ols))))
pred_ols <- as.numeric(predict(ols, newdata = as.data.frame(Xte)))

# VIF without depending on car: 1 / (1 - R^2) from regressing each feature on the others.
vif_of <- function(X) {
  vapply(seq_len(ncol(X)), function(j) {
    r2 <- summary(lm(X[, j] ~ X[, -j]))$r.squared
    1 / (1 - r2)
  }, numeric(1))
}
vifs <- vif_of(Xtr); names(vifs) <- colnames(Xtr)

# Standardized coefficients: beta * sd(x) / sd(y). This is the only scale on which two features
# with different units can be compared, and it is what lines up with a SHAP share.
sd_x <- apply(Xtr, 2, sd); sd_y <- sd(ytr)
co <- summary(ols)$coefficients
coef_tbl <- tibble(
  feature  = rownames(co),
  beta_raw = co[, 1], se = co[, 2], t = co[, 3], p = co[, 4]
) %>%
  filter(feature != "(Intercept)") %>%
  mutate(beta_std = beta_raw * sd_x[feature] / sd_y,
         vif      = vifs[feature],
         p_adj    = p.adjust(p, method = "BH")) %>%
  left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
  arrange(desc(abs(beta_std)))

cat(sprintf("adjusted R2 (train) %.4f | residual df %d\n",
            summary(ols)$adj.r.squared, ols$df.residual))
cat(sprintf("features with BH-adjusted p < 0.05: %d of %d\n",
            sum(coef_tbl$p_adj < 0.05), nrow(coef_tbl)))
cat(sprintf("features with VIF > 10 (coefficient imprecise, prediction unaffected): %d\n\n",
            sum(coef_tbl$vif > 10)))
print(as.data.frame(coef_tbl %>% head(20) %>%
  mutate(across(c(beta_raw, se, beta_std), ~ signif(., 3)),
         across(c(t, vif), ~ round(., 2)),
         across(c(p, p_adj), ~ signif(., 2)))), row.names = FALSE)
```

# 2. Lasso: how few features are needed?

```{r lasso}
# Standardized internally by glmnet (standardize = TRUE), so the penalty treats features with
# different units comparably. Same seed convention as everywhere else in this project.
set.seed(9)
cvfit <- cv.glmnet(Xtr, ytr, alpha = 1, nfolds = 10, standardize = TRUE)
pred_lasso_min <- as.numeric(predict(cvfit, newx = Xte, s = "lambda.min"))
pred_lasso_1se <- as.numeric(predict(cvfit, newx = Xte, s = "lambda.1se"))

nz <- function(s) {
  b <- as.matrix(coef(cvfit, s = s))
  rownames(b)[b[, 1] != 0 & rownames(b) != "(Intercept)"]
}
nz_min <- nz("lambda.min"); nz_1se <- nz("lambda.1se")
cat(sprintf("lambda.min %.5f -> %d of %d features retained\n",
            cvfit$lambda.min, length(nz_min), ncol(Xtr)))
cat(sprintf("lambda.1se %.5f -> %d of %d features retained\n",
            cvfit$lambda.1se, length(nz_1se), ncol(Xtr)))
cat("\nthe sparse model (lambda.1se) keeps:\n  ",
    paste(nz_1se, collapse = ", "), "\n")

lasso_tbl <- as.matrix(coef(cvfit, s = "lambda.1se")) %>% as.data.frame() %>%
  rownames_to_column("feature") %>% rename(beta_lasso = 2) %>%
  filter(feature != "(Intercept)", beta_lasso != 0) %>%
  mutate(beta_lasso_std = beta_lasso * sd_x[feature] / sd_y) %>%
  left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
  arrange(desc(abs(beta_lasso_std)))
print(as.data.frame(lasso_tbl %>%
  mutate(across(where(is.numeric), ~ signif(., 3)))), row.names = FALSE)
```

# 3. Is the relationship linear? A spline check

```{r splines}
# If the forest beats OLS it must be exploiting non-linearity or interactions. Replacing the
# strongest continuous features with natural splines tests the non-linearity half directly: if
# splines do not beat OLS, the relationship is linear in those features and the forest's extra
# flexibility has nothing to use.
#
# Binary and dummy columns are excluded - a spline on a 0/1 column is the column.
n_unique <- apply(Xtr, 2, function(v) length(unique(v)))
continuous <- names(n_unique)[n_unique > 10]
spline_feats <- coef_tbl %>% filter(feature %in% continuous) %>%
  head(params$n_spline_features) %>% pull(feature)
cat("splined:", paste(spline_feats, collapse = ", "), "\n")
cat(sprintf("(top %d continuous features by |standardized coefficient|, df = %d each)\n\n",
            length(spline_feats), params$spline_df))

mk_formula <- function(feats, sp, df) {
  terms <- vapply(feats, function(f)
    if (f %in% sp) sprintf("ns(`%s`, df = %d)", f, df) else sprintf("`%s`", f), character(1))
  as.formula(paste("ytr ~", paste(terms, collapse = " + ")))
}
dtr <- as.data.frame(Xtr); dte <- as.data.frame(Xte)
fit_sp <- lm(mk_formula(colnames(Xtr), spline_feats, params$spline_df), data = dtr)
pred_sp <- as.numeric(predict(fit_sp, newdata = dte))

# Nested models, so an F test is the direct question: do the extra spline terms explain more than
# chance? It is reported next to the held-out Spearman, which is what actually matters.
an <- anova(ols, fit_sp)
cat(sprintf("extra parameters: %d | F = %.2f | p = %.3g\n",
            abs(diff(an$Df[1:2])), an$F[2], an$`Pr(>F)`[2]))
cat(sprintf("train adj R2: OLS %.4f -> splines %.4f\n",
            summary(ols)$adj.r.squared, summary(fit_sp)$adj.r.squared))
```

# 4. All models on one yardstick

```{r compare}
rf_pred <- read_csv(file.path(out_dir, paste0("rfreg_predictions_", params$rf_suffix, ".csv")),
                    show_col_types = FALSE) %>% filter(split == "test")
stopifnot("the saved RF predictions do not line up with this split" =
            identical(rf_pred$gene_id_clean, test$gene_id_clean))

r2_of <- function(p) 1 - sum((yte - p)^2) / sum((yte - mean(ytr))^2)
models <- tibble::tribble(
  ~model,             ~pred,
  "OLS",               pred_ols,
  "lasso (1se)",       pred_lasso_1se,
  "lasso (min)",       pred_lasso_min,
  "OLS + splines",     pred_sp,
  "random forest",     rf_pred$pred
) %>%
  mutate(spearman = map_dbl(pred, ~ cor(.x, yte, method = "spearman")),
         pearson  = map_dbl(pred, ~ cor(.x, yte)),
         r2       = map_dbl(pred, r2_of),
         rmse     = map_dbl(pred, ~ sqrt(mean((yte - .x)^2))),
         n_param  = c(ncol(Xtr), length(nz_1se), length(nz_min),
                      length(coef(fit_sp)) - 1, NA_integer_))
print(as.data.frame(models %>% dplyr::select(-pred) %>%
  mutate(across(where(is.double), ~ round(., 4)))), row.names = FALSE)

# The differences above are small, so they need the scale that 10c_ measured: the split-to-split
# sd of the forest's own Spearman. A gap smaller than that is not a result.
sw_path <- file.path(out_dir, "rfreg_sweep_transform.csv")
if (file.exists(sw_path)) {
  sd_split <- read_csv(sw_path, show_col_types = FALSE) %>%
    filter(arm == "raw") %>% pull(spearman) %>% sd(na.rm = TRUE)   # seeds rerun later may be blank
  # The sweep was run on MDA-MB-231 si3d hypoxia 1hr; for other contrasts it is a reference scale only.
  if (!grepl("hypoxia_1hr", params$rf_suffix))
    cat("NOTE: the split-noise sd below was measured on si3d hypoxia 1hr, not on this contrast\n")
  if (anyNA(models$spearman)) cat("NOTE: these models returned no held-out Spearman:",
                                   paste(models$model[is.na(models$spearman)], collapse = ", "), "\n")
  gap <- max(models$spearman[models$model != "random forest"], na.rm = TRUE) -
    models$spearman[models$model == "random forest"]
  cat(sprintf("\nbest linear arm minus forest: %+.4f Spearman\n", gap))
  cat(sprintf("split-to-split sd of the forest's Spearman: %.4f (from 10c_'s sweep)\n", sd_split))
  cat(sprintf("=> the gap is %.1f%% of one standard deviation of the split noise.\n",
              100 * abs(gap) / sd_split))
  if (abs(gap) < sd_split)
    cat("   On this evidence the two model families perform the same, and the simpler one wins\n",
        "   on parsimony. A sweep over split seeds would be needed to claim a real difference.\n")
} else {
  cat("\n(run 10b_ to get the split-to-split sd that these differences must be judged against)\n")
}

# The same AUC bridge 10_ reports, so the two model families are compared on that too.
pos_genes <- read_csv(here("output", "genesets",
  paste0(params$condition, "_", params$gene, "_", params$direction, "_",
         params$readout, "_", params$timepoint, "_lfc", params$lfc, ".csv")),
  show_col_types = FALSE)$ensembl_gene
neg_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE) %>% mutate(g = sub("[.].*", "", gene_id)) %>% pull(g)
neg_genes <- setdiff(neg_genes, pos_genes)
lab <- case_when(test$gene_id_clean %in% pos_genes ~ 1L,
                 test$gene_id_clean %in% neg_genes ~ 0L, TRUE ~ NA_integer_)
auc_of <- function(p) {
  k <- !is.na(lab)
  as.numeric(pROC::auc(pROC::roc(lab[k], p[k], direction = "<", quiet = TRUE)))
}
models$bridge_auc <- map_dbl(models$pred, auc_of)
cat(sprintf("\nAUC bridge (n = %d pos / %d neg):\n", sum(lab == 1, na.rm = TRUE),
            sum(lab == 0, na.rm = TRUE)))
print(as.data.frame(models %>% dplyr::select(model, spearman, bridge_auc) %>%
  mutate(across(where(is.double), ~ round(., 4)))), row.names = FALSE)
```

```{r fig_performance, fig.width = 11, fig.height = 4.4}
# A lasso that keeps no features predicts a constant, which has no Spearman (NA). At the weaker
# timepoints lambda.1se can do exactly that; such a model is reported in the text and left off the bar.
p_perf <- models %>% filter(!is.na(spearman)) %>%
  mutate(model = fct_reorder(model, spearman),
         family = if_else(model == "random forest", "forest", "linear")) %>%
  ggplot(aes(spearman, model, fill = family)) +
  geom_col(alpha = 0.9, width = 0.65) +
  geom_text(aes(label = sprintf("%.4f", spearman)), hjust = -0.12, size = 3.2) +
  scale_fill_manual(values = c(linear = "#0072B2", forest = "#E69F00"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(x = "Spearman vs raw held-out outcome", y = NULL,
       title = "(a) Held-out ranking performance",
       subtitle = sprintf("n = %d held-out genes, identical split for every model", nrow(Xte))) +
  theme_classic(base_size = 11) + theme(legend.position = "bottom")

p_sc <- tibble(ols = pred_ols, rf = rf_pred$pred, obs = yte) %>%
  ggplot(aes(rf, ols)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.2, size = 0.7, colour = "#0072B2") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.3,
           label = sprintf("Spearman = %.3f\nn = %d genes",
                           cor(pred_ols, rf_pred$pred, method = "spearman"), nrow(Xte))) +
  labs(x = "forest prediction", y = "OLS prediction",
       title = "(b) Do the two models rank genes the same way?",
       subtitle = "held-out genes") +
  theme_classic(base_size = 11)

pg <- plot_grid(p_perf, p_sc, nrow = 1, rel_widths = c(1.15, 1))
print(pg)
ggsave_if(here("plots", paste0("13_performance_", params$rf_suffix, ".pdf")), pg,
          width = 11, height = 4.4)
```

# 5. Coefficients, and whether they agree with the forest

```{r fig_coefficients, fig.width = 11, fig.height = 6}
top_c <- coef_tbl %>% head(25) %>%
  mutate(feature = fct_reorder(feature, beta_std),
         sig = p_adj < 0.05,
         lo = (beta_raw - 1.96 * se) * sd_x[as.character(feature)] / sd_y,
         hi = (beta_raw + 1.96 * se) * sd_x[as.character(feature)] / sd_y)
p_coef <- ggplot(top_c, aes(beta_std, feature)) +
  geom_vline(xintercept = 0, colour = "grey50") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, colour = "grey65") +
  geom_point(aes(colour = sig, size = pmin(vif, 20))) +
  scale_colour_manual(values = c(`TRUE` = "#0072B2", `FALSE` = "grey65"),
                      labels = c(`TRUE` = "BH p < 0.05", `FALSE` = "not significant"),
                      name = NULL) +
  scale_size_continuous(range = c(1.5, 4.5), name = "VIF (capped at 20)") +
  labs(x = "standardized coefficient (per SD of the feature, in SD of the outcome)", y = NULL,
       title = "(a) OLS coefficients, top 25 by magnitude",
       subtitle = "bars are 95% CI; larger points have more collinearity and a less trustworthy individual estimate") +
  theme_classic(base_size = 11)
print(p_coef)
ggsave_if(here("plots", paste0("13_coefficients_", params$rf_suffix, ".pdf")), p_coef,
          width = 8, height = 6)
```

```{r fig_vs_shap, fig.width = 11, fig.height = 4.8}
# A standardized coefficient and a SHAP share measure different things - one is signed and
# conditional on the other features, the other is an unsigned average contribution - so the
# comparison is between |beta_std| and the SHAP share, on ranks.
shap_path <- file.path(out_dir, paste0("shapreg_feature_", params$rf_suffix, ".csv"))
imp_path  <- file.path(out_dir, paste0("rfreg_importance_", params$rf_suffix, ".csv"))
if (!file.exists(imp_path)) {
  cat("Forest importance not found; run 10_rf_regression.Rmd.\n")
} else {
  imp <- read_csv(imp_path, show_col_types = FALSE) %>%
    dplyr::select(feature, rf_share = share_pct)
  j <- coef_tbl %>% dplyr::select(feature, family, beta_std, p_adj) %>%
    mutate(abs_beta = abs(beta_std)) %>% inner_join(imp, by = "feature")
  p_imp <- ggplot(j, aes(rf_share, abs_beta)) +
    geom_point(aes(colour = family), size = 2, alpha = 0.85) +
    scale_colour_viridis_d(guide = "none", end = 0.9) +
    ggrepel::geom_text_repel(
      data = j %>% filter(rf_share > 2.5 | abs_beta > quantile(abs_beta, 0.92)),
      aes(label = feature), size = 2.5, max.overlaps = 18, seed = 9) +
    annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
             label = sprintf("Spearman = %.3f\nn = %d features",
                             cor(j$rf_share, j$abs_beta, method = "spearman"), nrow(j))) +
    labs(x = "forest impurity share (%)", y = "|standardized OLS coefficient|",
         title = "(a) Coefficient magnitude vs forest importance",
         subtitle = "ranks compared: the two quantities are not on a common scale") +
    theme_classic(base_size = 11)

  if (file.exists(shap_path)) {
    sh <- read_csv(shap_path, show_col_types = FALSE) %>%
      mutate(shap_share = 100 * mean_abs_shap / sum(mean_abs_shap)) %>%
      dplyr::select(feature, shap_share, shap_value_cor)
    j2 <- j %>% inner_join(sh, by = "feature")
    # A coefficient's sign and SHAP's directional summary should agree where both are defined.
    agree <- mean(sign(j2$beta_std) == sign(j2$shap_value_cor), na.rm = TRUE)
    p_shap <- ggplot(j2, aes(shap_share, abs_beta)) +
      geom_point(aes(colour = sign(beta_std) == sign(shap_value_cor)), size = 2, alpha = 0.85) +
      scale_colour_manual(values = c(`TRUE` = "grey40", `FALSE` = "#D55E00"),
                          labels = c(`TRUE` = "sign agrees", `FALSE` = "sign differs"),
                          name = NULL) +
      ggrepel::geom_text_repel(
        data = j2 %>% filter(shap_share > 3 | abs_beta > quantile(abs_beta, 0.92)),
        aes(label = feature), size = 2.5, max.overlaps = 18, seed = 9) +
      annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
               label = sprintf("Spearman = %.3f\nsign agreement %.0f%%\nn = %d",
                               cor(j2$shap_share, j2$abs_beta, method = "spearman"),
                               100 * agree, nrow(j2))) +
      labs(x = "forest SHAP share (%)", y = "|standardized OLS coefficient|",
           title = "(b) Coefficient vs SHAP",
           subtitle = "colour: does the coefficient's sign match SHAP's direction?") +
      theme_classic(base_size = 11) + theme(legend.position = "bottom")
    pg2 <- plot_grid(p_imp, p_shap, nrow = 1)
  } else {
    cat("SHAP not found; run 11_shap_regression.Rmd to fill in panel (b).\n")
    pg2 <- p_imp
  }
  print(pg2)
  ggsave_if(here("plots", paste0("13_coef_vs_forest_", params$rf_suffix, ".pdf")), pg2,
            width = 11, height = 4.8)
}
```

```{r fig_splines, fig.width = 11, fig.height = 6}
# What the spline arm actually fitted: the partial effect of each splined feature, with the
# straight-line OLS fit over it. If the two are close, the relationship is linear and the extra
# parameters bought nothing.
grid_one <- function(f) {
  g <- as.data.frame(matrix(rep(colMeans(Xtr), each = 100), nrow = 100,
                            dimnames = list(NULL, colnames(Xtr))))
  g[[f]] <- seq(quantile(Xtr[, f], 0.01), quantile(Xtr[, f], 0.99), length.out = 100)
  tibble(feature = f, x = g[[f]],
         spline = as.numeric(predict(fit_sp, newdata = g)),
         linear = as.numeric(predict(ols,    newdata = g)))
}
pe <- map_dfr(spline_feats, grid_one) %>%
  pivot_longer(c(spline, linear), names_to = "fit", values_to = "pred")
p_pe <- ggplot(pe, aes(x, pred, colour = fit)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ feature, scales = "free_x", ncol = 4) +
  scale_colour_manual(values = c(linear = "#0072B2", spline = "#E69F00"), name = NULL) +
  labs(x = "feature value (1st to 99th percentile)", y = "predicted TE LFC",
       title = "Partial effects: straight line vs natural spline",
       subtitle = "other features held at their means. Close curves mean the relationship is linear") +
  theme_classic(base_size = 11) + theme(strip.background = element_blank(),
                                        legend.position = "bottom")
print(p_pe)
ggsave_if(here("plots", paste0("13_partial_effects_", params$rf_suffix, ".pdf")), p_pe,
          width = 11, height = 6)
```

# 6. Conclusions

```{r summary}
best_lin <- models %>% filter(model != "random forest") %>% slice_max(spearman, n = 1)
rf_row   <- models %>% filter(model == "random forest")
cat(sprintf("best linear arm: %s, Spearman %.4f (%d parameters)\n",
            best_lin$model, best_lin$spearman, best_lin$n_param))
cat(sprintf("random forest:   Spearman %.4f\n", rf_row$spearman))
cat(sprintf("difference:      %+.4f\n\n", best_lin$spearman - rf_row$spearman))
cat(sprintf("the sparse lasso keeps %d of %d features and reaches Spearman %.4f\n",
            length(nz_1se), ncol(Xtr), models$spearman[models$model == "lasso (1se)"]))
cat(sprintf("splines add %d parameters for a Spearman change of %+.4f\n\n",
            length(coef(fit_sp)) - 1 - ncol(Xtr),
            models$spearman[models$model == "OLS + splines"] - models$spearman[models$model == "OLS"]))
cat("Read these against the split-to-split noise printed in section 4, not against zero.\n")
cat("A single split cannot separate differences of this size; 10b_/10c_ is the machinery for\n")
cat("that, and the same sweep would need to be run for the linear arms to make a firm claim.\n")
```

```{r export}
source(here("code", "translational_control_overview", "10_regression_presets.R"))
reg_write_csv(coef_tbl,  file.path(out_dir, paste0("linreg_coefficients_", params$rf_suffix, ".csv")))
reg_write_csv(lasso_tbl, file.path(out_dir, paste0("linreg_lasso_", params$rf_suffix, ".csv")))
reg_write_csv(models %>% dplyr::select(-pred),
              file.path(out_dir, paste0("linreg_performance_", params$rf_suffix, ".csv")))
reg_write_csv(tibble(gene_id_clean = test$gene_id_clean, symbol = test$symbol, y = yte,
                     pred_ols = pred_ols, pred_lasso_1se = pred_lasso_1se,
                     pred_splines = pred_sp, pred_rf = rf_pred$pred),
              file.path(out_dir, paste0("linreg_predictions_", params$rf_suffix, ".csv")))
cat("Wrote linreg_* outputs.\n")
```
