---
title: "Linear model vs random forest: do they credit the same feature families? (si3d hypoxia 1, 4, 24 hr)"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  stem: "eif33d_promotes_hypoxia_%s_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  timepoints: "1hr,4hr,24hr"
  drop_for_rank: "term_stop_TAA"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "13b_linear_vs_forest_shap.Rmd"))
  params <- lapply(.fm$params, function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

`13_linear_regression.Rmd` showed that ordinary least squares (OLS) predicts the TE change as well
as the forest at 1, 4 and 24 hr. Comparing OLS **coefficients** with forest SHAP feature by
feature is unfair, though: the features are collinear (`cds_gc` has VIF ~51), and OLS can split
credit among near-duplicates arbitrarily (a large positive coefficient cancelling a large negative
one). So here both models are put on the same footing:

- **Linear SHAP is exact.** For a linear model, feature j's SHAP value for gene i is
  `beta_j * (x_ij - mean_j)`, with the mean taken over the training genes (the same background a
  tree explainer uses).
- **Sum within families.** SHAP is additive, so a family's value is the exact sum of its members'.
  The GC family gets its total credit whichever GC column OLS happened to load onto - which is
  what cancels the collinearity problem.
- **Same genes.** Both are evaluated on the 200 held-out genes `11_shap_regression.Rmd` explained
  for the forest, so the two are compared gene for gene.

Two questions per timepoint: do the models give each family a similar **share** of the credit,
and do they attribute it to the **same genes** (per-gene correlation of a family's SHAP between
the two models)?

```{r libraries}
suppressPackageStartupMessages({ library(tidyverse); library(here) })
O <- here("output", "translational_control_overview")
tps <- strsplit(params$timepoints, ",")[[1]]
ann <- read_csv(file.path(O, "feature_annotation.csv"), show_col_types = FALSE)
```

```{r compute}
one_tp <- function(tp) {
  sfx <- sprintf(params$stem, tp)
  md  <- readRDS(file.path(O, paste0("rfreg_model_data_", sfx, ".rds")))
  fc_all <- md$feature_cols; fc <- setdiff(fc_all, params$drop_for_rank)
  Xtr <- as.matrix(md$train[, fc]); stopifnot("OLS design not full rank" = qr(cbind(1, Xtr))$rank == ncol(Xtr) + 1)
  ols <- lm(md$train$y_t ~ Xtr); beta <- coef(ols)[-1]; names(beta) <- fc
  fam <- setNames(ann$family[match(fc_all, ann$feature)], fc_all); stopifnot(!anyNA(fam))
  # forest SHAP for the explained held-out genes, as saved by 11_
  pg <- read_csv(file.path(O, paste0("shapreg_per_gene_", sfx, ".csv")), show_col_types = FALSE)
  stopifnot("per-gene SHAP lacks some features" = all(fc_all %in% names(pg)))
  Sf <- as.matrix(pg[, fc_all])
  left <- pg$pred - rowSums(Sf)
  stopifnot("forest SHAP does not add up to the forest prediction" = diff(range(left)) < 1e-6)
  # linear SHAP on the same genes, from their (imputed) held-out feature values
  te <- md$test[match(pg$gene_id_clean, md$test$gene_id_clean), ]
  stopifnot("explained genes are not all held-out genes" = !anyNA(te$gene_id_clean))
  Sl <- matrix(0, nrow(te), length(fc_all), dimnames = list(NULL, fc_all))
  Sl[, fc] <- sweep(as.matrix(te[, fc]), 2, colMeans(Xtr)) %*% diag(beta)
  pred_lin <- predict(ols, newdata = list(Xtr = as.matrix(te[, fc])))
  stopifnot("linear SHAP does not add up to the OLS prediction" =
              diff(range(pred_lin - rowSums(Sl))) < 1e-6)
  famsum <- function(S) sapply(sort(unique(fam)), function(g) rowSums(S[, fam == g, drop = FALSE]))
  Ff <- famsum(Sf); Fl <- famsum(Sl)
  share <- function(Fm) 100 * colMeans(abs(Fm)) / sum(colMeans(abs(Fm)))
  tibble(timepoint = tp, family = colnames(Ff), forest_share = share(Ff), linear_share = share(Fl),
         per_gene_r = map_dbl(colnames(Ff), ~ suppressWarnings(cor(Ff[, .x], Fl[, .x]))),
         n_genes = nrow(Ff), pred_r = cor(pg$pred, pred_lin, method = "spearman"))
}
res <- map_dfr(tps, one_tp) %>% mutate(timepoint = factor(timepoint, levels = tps))
res %>% distinct(timepoint, n_genes, pred_r) %>% knitr::kable(digits = 3, caption = "Genes compared, and how similarly the two models rank them")
knitr::kable(res %>% dplyr::select(timepoint, family, forest_share, linear_share, per_gene_r) %>% arrange(timepoint, desc(forest_share)), digits = 2)
```

# Family shares, side by side

```{r fig_shares, fig.width = 13, fig.height = 5}
p1 <- res %>% pivot_longer(c(forest_share, linear_share), names_to = "model", values_to = "share") %>%
  mutate(model = recode(model, forest_share = "random forest", linear_share = "linear (OLS)"),
         family = fct_reorder(family, share, .fun = max)) %>%
  ggplot(aes(share, family, fill = model)) + geom_col(position = position_dodge(0.75), width = 0.7) +
  facet_wrap(~ timepoint, labeller = labeller(timepoint = function(x) paste("si3d hypoxia", x))) +
  scale_fill_manual(values = c(`random forest` = "#E69F00", `linear (OLS)` = "#0072B2"), name = NULL) +
  labs(x = "share of SHAP credit (%)", y = NULL, title = "Which feature families each model credits",
       subtitle = sprintf("full model, MDA-MB-231; SHAP summed within family; n = %d held-out genes per timepoint", res$n_genes[1])) +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
p1
if (isTRUE(params$save_plots)) ggsave(here("plots", "13b_linear_vs_forest_family_shares.pdf"), p1, width = 13, height = 5)
```

# Do they credit the same genes?

For each family, the correlation across genes between the forest's and the linear model's SHAP
for that family. High means the two models push the same genes the same way through that family.
Families with tiny shares are noisy here and are shown faded.

```{r fig_agreement, fig.width = 8, fig.height = 5}
p2 <- res %>% mutate(family = fct_reorder(family, forest_share, .fun = max),
                     big = (forest_share + linear_share) / 2 >= 2) %>%
  ggplot(aes(timepoint, family, fill = per_gene_r)) + geom_tile(aes(alpha = big), colour = "white") +
  geom_text(aes(label = sprintf("%.2f", per_gene_r)), size = 3.2) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  labs(x = "si3d hypoxia", y = NULL, title = "Per-gene agreement of family SHAP, forest vs linear",
       subtitle = sprintf("Pearson r across n = %d held-out genes; faded = mean share < 2%%", res$n_genes[1])) +
  theme_classic(base_size = 12)
p2
if (isTRUE(params$save_plots)) ggsave(here("plots", "13b_linear_vs_forest_family_agreement.pdf"), p2, width = 8, height = 5)
```

# Held-out fit, forest vs OLS

Every held-out gene (the 20% the models never trained on), predicted against observed TE change.
R² is computed on these held-out genes as 1 - SSE / SST, so it is the share of the held-out
variance each model explains (it can be below the squared correlation, and below zero for a model
worse than predicting the mean). Both models are scored on the same genes from `13_`'s saved
predictions.

```{r fig_heldout_r2, fig.width = 12, fig.height = 7.5}
ho <- map_dfr(tps, function(tp) {
  read_csv(file.path(O, paste0("linreg_predictions_", sprintf(params$stem, tp), ".csv")), show_col_types = FALSE) %>%
    transmute(timepoint = tp, y, `linear (OLS)` = pred_ols, `random forest` = pred_rf) }) %>%
  pivot_longer(c(`linear (OLS)`, `random forest`), names_to = "model", values_to = "pred") %>%
  mutate(timepoint = factor(timepoint, levels = tps), model = factor(model, levels = c("random forest", "linear (OLS)")))
r2 <- ho %>% group_by(timepoint, model) %>%
  summarise(r2 = 1 - sum((y - pred)^2) / sum((y - mean(y))^2), spearman = cor(y, pred, method = "spearman"),
            n = n(), .groups = "drop")
knitr::kable(r2, digits = 3)
lim <- quantile(ho$y, c(0.002, 0.998))
p3 <- ggplot(ho, aes(pred, y, colour = model)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 0.6, alpha = 0.25) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey40") +
  geom_text(data = r2, aes(x = -Inf, y = Inf, label = sprintf("  R² = %.3f\n  Spearman = %.3f\n  n = %s", r2, spearman, format(n, big.mark = ","))),
            hjust = 0, vjust = 1.2, size = 3.4, colour = "black", inherit.aes = FALSE) +
  scale_colour_manual(values = c(`random forest` = "#E69F00", `linear (OLS)` = "#0072B2"), guide = "none") +
  facet_grid(model ~ timepoint, labeller = labeller(timepoint = function(x) paste("si3d hypoxia", x))) +
  coord_cartesian(ylim = lim) +
  labs(x = "predicted TE change, log2(sictrl / si3d)", y = "observed TE change, log2(sictrl / si3d)",
       title = "Held-out genes: random forest vs linear regression",
       subtitle = "full model, MDA-MB-231; dashed = perfect prediction; y axis clipped to the central 99.6% of genes for display") +
  theme_classic(base_size = 12)
p3
if (isTRUE(params$save_plots)) ggsave(here("plots", "13b_heldout_r2_forest_vs_ols.pdf"), p3, width = 12, height = 7.5)
```

```{r session}
sessionInfo()
```
