---
title: "Feature correlation structure: what is actually independent?"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
params:
  rf_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  matrix_rds: "feature_matrix_tco.rds"
  thresholds: "0.7,0.8,0.9"
  main_threshold: 0.8
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "14_feature_correlation.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It measures how correlated the 69 modelled features are with each other, across every expressed
transcript, and shows where that correlation is large enough to split a SHAP value between
features that are really measuring one thing.

This is a **diagnosis only**. Nothing here is written for another notebook to consume, no feature
is dropped, and no model is refitted.

**Decisions**

- Use all **10,172** transcripts in the feature matrix, not just the training split. Correlation
  between features is a property of the transcripts, not of any particular model fit or split.
- Use **Spearman**, not Pearson. Several of these features are monotone transforms of each other -
  `term_log1p_readthrough` is literally `log1p()` of `term_readthrough_cridge` - and Spearman
  reports those as the perfect redundancy they are, where Pearson would understate it.
- Report at several thresholds rather than one. Where a cluster appears is as informative as the
  cluster itself.
- Annotate every cluster with the SHAP share its members hold, so the cost of the splitting is
  quantified rather than implied.

## Why this matters for the heatmap

When two features carry the same underlying signal, the forest uses whichever is convenient at
each split. One gene's prediction gets attributed to `cds_gc`, another's to `gc3_internal`.
Neither attribution is wrong, but read one row at a time, the *concept* looks half as important as
it is.

Summing SHAP within a correlated group fixes this exactly, because SHAP is additive: the sum of
two features' SHAP values is precisely their joint contribution to each gene's prediction, with
nothing estimated and no refit. Dropping one instead changes the model, and the survivor absorbs
the dropped feature's role by an unknown amount.

The reason to look before deciding is that at these correlations the features are **not
interchangeable**. A pair at rho 0.9 still has about 19% of its variance unshared. Section 5
quantifies exactly how much would be lost by dropping each member.

```{r libraries}
library(tidyverse)
library(here)
library(pheatmap)
library(cowplot)
```

```{r load}
out_dir <- here("output", "translational_control_overview")
md <- readRDS(file.path(out_dir, paste0("rfreg_model_data_", params$rf_suffix, ".rds")))
m   <- readRDS(file.path(out_dir, params$matrix_rds))
ann <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)

fc <- md$feature_cols
stopifnot("the saved model data and the matrix disagree about the features" =
            all(fc %in% colnames(m)))

X <- m[, fc]
for (f in fc) if (is.list(X[[f]])) X[[f]] <- as.numeric(unlist(X[[f]]))
X <- as.matrix(X)
cat(sprintf("%d transcripts x %d features\n", nrow(X), ncol(X)))
cat(sprintf("missing values: %d (%.3f%%)\n", sum(is.na(X)), 100 * mean(is.na(X))))

C <- suppressWarnings(cor(X, use = "pairwise.complete.obs", method = "spearman"))
stopifnot("correlation matrix has missing entries - a feature may be constant" = !anyNA(C))
D <- as.dist(1 - abs(C))
hc <- hclust(D, method = "average")
fam <- setNames(ann$family[match(fc, ann$feature)], fc)
thresholds <- as.numeric(str_trim(str_split(params$thresholds, ",")[[1]]))

shap_path <- file.path(out_dir, paste0("shapreg_feature_", params$rf_suffix, ".csv"))
have_shap <- file.exists(shap_path)
shap <- if (have_shap) read_csv(shap_path, show_col_types = FALSE) %>%
  dplyr::select(feature, share_pct) else NULL
cat("SHAP shares available:", have_shap, "\n")
ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)
```

# 1. The correlation heatmap

```{r fig_heatmap, fig.width = 13, fig.height = 12}
Cd <- C; diag(Cd) <- 0
np <- ncol(C) * (ncol(C) - 1) / 2
cat("pairs above each threshold:\n")
for (th in c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95))
  cat(sprintf("  |rho| > %.2f : %5d of %d pairs (%.2f%%)\n",
              th, sum(abs(Cd) > th) / 2, np, 100 * (sum(abs(Cd) > th) / 2) / np))

fam_cols <- setNames(
  colorRampPalette(c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2",
                     "#D55E00", "#CC79A7", "grey50"))(n_distinct(fam)),
  sort(unique(fam)))
anno <- data.frame(family = fam[fc], row.names = fc)

pheatmap(C,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  clustering_distance_rows = D, clustering_distance_cols = D,
  clustering_method = "average",
  annotation_row = anno, annotation_col = anno,
  annotation_colors = list(family = fam_cols),
  fontsize_row = 6, fontsize_col = 6, treeheight_col = 40, treeheight_row = 40,
  main = sprintf("Spearman correlation, %d features across %d transcripts", ncol(C), nrow(X)))
if (isTRUE(params$save_plots)) {
  # Build silently, then draw the stored gtable into the PDF. Calling pheatmap() directly after
  # pdf() inside a knitr chunk wrote an essentially empty 4 KB file - knitr's own graphics device
  # captured the drawing. Same pattern as 15_shap_heatmap.Rmd's draw_and_save().
  hm_pdf <- pheatmap(C, color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
    breaks = seq(-1, 1, length.out = 102),
    clustering_distance_rows = D, clustering_distance_cols = D,
    clustering_method = "average", annotation_row = anno, annotation_col = anno,
    annotation_colors = list(family = fam_cols), fontsize_row = 6, fontsize_col = 6,
    main = "Spearman correlation between modelled features", silent = TRUE)
  pdf(here("plots", paste0("14_correlation_heatmap_", params$rf_suffix, ".pdf")),
      width = 13, height = 12)
  grid::grid.newpage(); grid::grid.draw(hm_pdf$gtable)
  invisible(dev.off())
}
```

# 2. Clusters at each threshold

```{r clusters}
cluster_at <- function(th) {
  g <- cutree(hc, h = 1 - th)
  tibble(feature = names(g), cluster = as.integer(g)) %>%
    left_join(tibble(feature = fc, family = fam[fc]), by = "feature") %>%
    { if (have_shap) left_join(., shap, by = "feature") else mutate(., share_pct = NA_real_) } %>%
    group_by(cluster) %>% filter(n() > 1) %>%
    arrange(desc(share_pct), .by_group = TRUE) %>%
    mutate(leader = first(feature),
           rho_to_leader = map2_dbl(leader, feature, ~ C[.x, .y])) %>%
    ungroup() %>% mutate(threshold = th)
}
clus <- map_dfr(thresholds, cluster_at)

for (th in thresholds) {
  d <- clus %>% filter(threshold == th)
  cat(sprintf("\n===== |rho| >= %.2f : %d multi-feature clusters covering %d features =====\n",
              th, n_distinct(d$cluster), nrow(d)))
  for (k in unique(d$cluster)) {
    x <- d %>% filter(cluster == k)
    cat(sprintf("\n  cluster %d | %d features | combined SHAP %.2f%% | families: %s\n",
                k, nrow(x), sum(x$share_pct, na.rm = TRUE),
                paste(sort(unique(x$family)), collapse = ", ")))
    for (i in seq_len(nrow(x)))
      cat(sprintf("     %-36s %6.2f%%  %-19s rho=%+.3f\n",
                  x$feature[i], x$share_pct[i], x$family[i], x$rho_to_leader[i]))
  }
}
```

```{r fig_cluster_shap, fig.width = 10, fig.height = 5}
# How much SHAP sits inside a correlated cluster is the size of the problem. A cluster spanning
# more than one family is the case that family-level sums do NOT already solve.
csum <- clus %>% group_by(threshold, cluster) %>%
  summarise(n = n(), shap = sum(share_pct, na.rm = TRUE),
            n_fam = n_distinct(family), leader = first(leader),
            fams = paste(sort(unique(family)), collapse = "+"), .groups = "drop") %>%
  mutate(spans_families = n_fam > 1)
p_cs <- csum %>%
  mutate(lab = paste0(leader, " (+", n - 1, ")")) %>%
  ggplot(aes(shap, reorder(lab, shap), fill = spans_families)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = sprintf("%.1f%%", shap)), hjust = -0.12, size = 3) +
  facet_wrap(~ paste0("|rho| >= ", threshold), scales = "free_y") +
  scale_fill_manual(values = c(`TRUE` = "#D55E00", `FALSE` = "grey60"),
                    labels = c(`TRUE` = "spans families", `FALSE` = "within one family"),
                    name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "combined SHAP share of the cluster (%)", y = NULL,
       title = "How much attribution sits inside a correlated cluster",
       subtitle = "orange clusters cross family lines, so family-level sums do not resolve them") +
  theme_classic(base_size = 11) + theme(legend.position = "bottom",
                                        strip.background = element_blank())
print(p_cs)
ggsave_if(here("plots", paste0("14_cluster_shap_", params$rf_suffix, ".pdf")), p_cs,
          width = 10, height = 5)
```

# 3. How many independent axes are there really?

```{r fig_pca, fig.width = 11, fig.height = 4.6}
# PCA on the correlation matrix. If the 69 features were independent, 69 components would each
# explain about 1.4% and the curve would be a straight line. The gap between that and the real
# curve is the redundancy.
pca <- prcomp(X[complete.cases(X), ], center = TRUE, scale. = TRUE)
ve  <- pca$sdev^2 / sum(pca$sdev^2)
cum <- cumsum(ve)
n_for <- function(p) which(cum >= p)[1]
cat(sprintf("components needed for 50%%: %d | 80%%: %d | 90%%: %d | 95%%: %d  (of %d features)\n",
            n_for(0.5), n_for(0.8), n_for(0.9), n_for(0.95), ncol(X)))
cat(sprintf("PC1 alone explains %.1f%%; the first 5 explain %.1f%%\n", 100*ve[1], 100*cum[5]))

p_scree <- tibble(pc = seq_along(ve), ve = ve, cum = cum) %>% head(40) %>%
  ggplot(aes(pc)) +
  geom_col(aes(y = 100 * ve), fill = "#E69F00", alpha = 0.9) +
  geom_line(aes(y = 100 * cum), colour = "#0072B2", linewidth = 0.9) +
  geom_point(aes(y = 100 * cum), colour = "#0072B2", size = 1.2) +
  geom_hline(yintercept = 100 / ncol(X), linetype = "dashed", colour = "grey50") +
  annotate("text", x = 30, y = 100/ncol(X) + 4, size = 3, colour = "grey40",
           label = sprintf("%.1f%% = what one PC would explain if\nall %d features were independent",
                           100/ncol(X), ncol(X))) +
  labs(x = "principal component", y = "variance explained (%)",
       title = "(a) Scree: bars are per-component, line is cumulative",
       subtitle = sprintf("%d components carry 90%% of the variance in %d features",
                          n_for(0.9), ncol(X))) +
  theme_classic(base_size = 11)

# Features in the space of the first two components, coloured by family: which families occupy
# the same direction, and which are their own axis.
ld <- as_tibble(pca$rotation[, 1:2], rownames = "feature") %>%
  mutate(family = fam[feature])
p_load <- ggplot(ld, aes(PC1, PC2, colour = family)) +
  geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
  geom_point(size = 2, alpha = 0.9) +
  scale_colour_manual(values = fam_cols, name = NULL) +
  ggrepel::geom_text_repel(
    data = ld %>% filter(sqrt(PC1^2 + PC2^2) > quantile(sqrt(ld$PC1^2 + ld$PC2^2), 0.75)),
    aes(label = feature), size = 2.4, max.overlaps = 22, seed = 9, show.legend = FALSE) +
  labs(x = sprintf("PC1 (%.1f%%)", 100*ve[1]), y = sprintf("PC2 (%.1f%%)", 100*ve[2]),
       title = "(b) Feature loadings",
       subtitle = "features pointing the same way are measuring the same axis") +
  theme_classic(base_size = 11) + theme(legend.position = "right",
                                        legend.text = element_text(size = 7))
pg <- plot_grid(p_scree, p_load, nrow = 1, rel_widths = c(1, 1.25))
print(pg)
ggsave_if(here("plots", paste0("14_pca_", params$rf_suffix, ".pdf")), pg, width = 11, height = 4.6)
```

# 4. Is every feature in the right family?

```{r family_audit}
# A systematic version of a spot-check: for every feature, compare its mean |rho| to the other
# members of its own family against its mean |rho| to each other family. A feature that tracks a
# different family more closely than its own is either mislabelled or measuring something other
# than its name suggests.
#
# This was prompted by hia2026_dhx29_occupancy_tx, which is labelled stability_external but is
# built in 23_dhx29_riboseq_slamseq.Rmd as a per-codon enrichment table averaged over the CDS -
# structurally the same construction as csc / cai / fop. Its comment there reads: "Per-gene DHX29
# occupancy = weighted mean of CES values, weighted by codon frequency". The CES table is AU3-
# biased, which is why the score tracks GC so strongly.
fam_of <- fam[fc]
audit <- map_dfr(fc, function(f) {
  own <- setdiff(fc[fam_of == fam_of[[f]]], f)
  by_fam <- map_dbl(sort(unique(fam_of)), function(g) {
    others <- setdiff(fc[fam_of == g], f)
    if (!length(others)) return(NA_real_)
    mean(abs(C[f, others]))
  })
  names(by_fam) <- sort(unique(fam_of))
  best <- names(which.max(by_fam))
  tibble(feature = f, family = fam_of[[f]],
         own_family = if (length(own)) mean(abs(C[f, own])) else NA_real_,
         closest_family = best, closest_value = by_fam[[best]])
}) %>%
  mutate(misplaced = !is.na(own_family) & closest_family != family &
                     closest_value > own_family + 0.1) %>%
  arrange(desc(closest_value - own_family))

cat(sprintf("features whose nearest family is not their own (by > 0.1 mean |rho|): %d of %d\n\n",
            sum(audit$misplaced), nrow(audit)))
print(as.data.frame(audit %>% filter(misplaced) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)

cat("\nThis is a flag, not a verdict: a feature can legitimately correlate with another family\n")
cat("without belonging to it. Check how each one is actually computed before relabelling.\n")
```

# 5. What would dropping a feature actually cost?

```{r unique_variance}
# For each member of a cluster, how much of its variance is NOT explained by the cluster leader.
# This is what would be lost by keeping only the leader. At rho 0.9 that is still about 19%, so
# these features are not interchangeable even when they look redundant.
uv <- clus %>% filter(threshold == params$main_threshold) %>%
  mutate(shared_with_leader = rho_to_leader^2,
         unique_variance    = 1 - rho_to_leader^2) %>%
  dplyr::select(cluster, feature, family, share_pct, rho_to_leader,
                shared_with_leader, unique_variance)
print(as.data.frame(uv %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)

p_uv <- uv %>% filter(unique_variance > 0) %>%
  mutate(feature = reorder(feature, unique_variance)) %>%
  ggplot(aes(100 * unique_variance, feature, fill = family)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = sprintf("%.0f%%", 100 * unique_variance)), hjust = -0.15, size = 3) +
  scale_fill_manual(values = fam_cols, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "variance NOT shared with the cluster leader (%)", y = NULL,
       title = sprintf("What each feature still carries on its own (clusters at |rho| >= %.1f)",
                       params$main_threshold),
       subtitle = "this is what would be discarded by keeping only the leader of each cluster") +
  theme_classic(base_size = 11)
print(p_uv)
ggsave_if(here("plots", paste0("14_unique_variance_", params$rf_suffix, ".pdf")), p_uv,
          width = 9, height = 5)
```

# 6. The decision table

```{r decision_table}
# One row per cluster at the main threshold, with everything needed to make a keep/drop/group
# call, and a blank column for the biological judgement. This notebook deliberately does not make
# that call or write a grouping for anything downstream to consume.
dec <- clus %>% filter(threshold == params$main_threshold) %>%
  group_by(cluster) %>%
  summarise(n_features        = n(),
            members           = paste(feature, collapse = " | "),
            families          = paste(sort(unique(family)), collapse = " + "),
            spans_families    = n_distinct(family) > 1,
            combined_shap_pct = round(sum(share_pct, na.rm = TRUE), 2),
            top_feature       = first(leader),
            top_feature_shap  = round(first(share_pct), 2),
            min_rho           = round(min(abs(rho_to_leader)), 3),
            max_unique_var    = round(max(1 - rho_to_leader^2), 3),
            .groups = "drop") %>%
  arrange(desc(combined_shap_pct)) %>%
  mutate(biological_call = "")
print(as.data.frame(dec %>% dplyr::select(-members)), row.names = FALSE)
cat("\nmembers of each cluster:\n")
for (i in seq_len(nrow(dec)))
  cat(sprintf("  cluster %d: %s\n", dec$cluster[i], dec$members[i]))

cat(sprintf("\nTotal SHAP inside a multi-feature cluster at |rho| >= %.1f: %.1f%%\n",
            params$main_threshold, sum(dec$combined_shap_pct)))
cat(sprintf("Of that, in clusters spanning more than one family: %.1f%%\n",
            sum(dec$combined_shap_pct[dec$spans_families])))
cat("The second number is the part that family-level SHAP sums do not already resolve.\n")
```

```{r export}
# Diagnostic tables only. Nothing here is read by 11_ or 12_, and no grouping is defined.
source(here("code", "translational_control_overview", "10_regression_presets.R"))
reg_write_csv(clus,  file.path(out_dir, paste0("featcorr_clusters_", params$rf_suffix, ".csv")))
reg_write_csv(audit, file.path(out_dir, paste0("featcorr_family_audit_", params$rf_suffix, ".csv")))
reg_write_csv(dec,   file.path(out_dir, paste0("featcorr_decision_table_", params$rf_suffix, ".csv")))
reg_write_csv(as_tibble(C, rownames = "feature"),
              file.path(out_dir, paste0("featcorr_matrix_", params$rf_suffix, ".csv")))
cat("Wrote featcorr_* diagnostics. No grouping file written; nothing downstream changes.\n")
```

# 8. Which features are most correlated, and do they cross families?

The full matrix above is mostly near zero, so it does not show the answer at a glance. These two
figures do: (a) every pair with |rho| > 0.5, ranked, coloured by whether the two features share a
family; (b) only the features in those pairs, with rho printed. A **cross-family** pair is where
the predefined families and the data disagree.

```{r top_pairs, fig.width = 10, fig.height = 11}
pairs_tbl <- as_tibble(as.table(C), .name_repair = ~ c("a", "b", "rho")) %>%
  mutate(a = as.character(a), b = as.character(b)) %>%
  filter(a < b, is.finite(rho)) %>%
  mutate(abs_rho = abs(rho), fam_a = fam[a], fam_b = fam[b],
         same_family = ifelse(fam_a == fam_b, "same family", "different families"),
         label = paste0(a, "  x  ", b),
         families = ifelse(fam_a == fam_b, fam_a, paste(fam_a, "/", fam_b))) %>%
  arrange(desc(abs_rho))
top <- pairs_tbl %>% filter(abs_rho > 0.5)
cat(sprintf("pairs with |rho| > 0.5: %d (same family %d, different families %d)\n", nrow(top),
            sum(top$same_family == "same family"), sum(top$same_family == "different families")))
print(as.data.frame(top %>% transmute(a, b, rho = round(rho, 3), families, same_family)), row.names = FALSE)

pair_cols <- c(`same family` = "#009E73", `different families` = "#CC79A7")  # not the eIF3d/e hues
p_pairs <- ggplot(top, aes(x = abs_rho, y = reorder(label, abs_rho), fill = same_family)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = sprintf("%+.2f", rho)), hjust = -0.1, size = 2.6) +
  geom_vline(xintercept = c(0.7, 0.8, 0.9), linetype = "dashed", colour = "grey55", linewidth = 0.3) +
  scale_fill_manual(values = pair_cols, name = NULL) +
  scale_x_continuous(limits = c(0, 1.08), breaks = seq(0, 1, 0.1), expand = c(0, 0)) +
  labs(x = "|Spearman rho| (printed: signed rho)", y = NULL,
       title = "Most correlated feature pairs",
       subtitle = sprintf("%d features, %d MDA-MB-231 transcripts; %d pairs with |rho| > 0.5; dashed = 0.7 / 0.8 / 0.9",
                          ncol(C), nrow(X), nrow(top))) +
  theme_classic(base_size = 9) + theme(legend.position = "top", axis.text.y = element_text(size = 7))
print(p_pairs)
ggsave_if(here("plots", paste0("14_top_correlated_pairs_", params$rf_suffix, ".pdf")), p_pairs,
          width = 10, height = 11)
```

```{r top_pairs_heatmap, fig.width = 11, fig.height = 10}
inv <- sort(unique(c(top$a, top$b)))
Cs <- C[inv, inv]
hc <- hclust(as.dist(1 - abs(Cs)), method = "average")
lab <- matrix(sprintf("%.2f", Cs), nrow(Cs), dimnames = dimnames(Cs)); diag(lab) <- ""
lab[abs(Cs) < 0.5] <- ""                                   # print only the strong values
fam_pal <- setNames(c("#0072B2", "#009E73", "#CC79A7", "#D55E00", "#F0E442", "#661100", "#999933",
                      "#44AA99", "#882255", "#332288", "#DDCC77", "#AA4499", "#117733", "grey50")[seq_len(n_distinct(fam))],
                    sort(unique(fam)))
an <- data.frame(family = fam[inv], row.names = inv)
hm_top <- pheatmap(Cs, color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102), cluster_rows = hc, cluster_cols = hc,
  display_numbers = lab, number_color = "grey10", fontsize_number = 6.5,
  annotation_row = an, annotation_col = an,
  annotation_colors = list(family = fam_pal[unique(an$family)]),
  fontsize_row = 7.5, fontsize_col = 7.5, treeheight_col = 0,
  main = sprintf("Features in any pair with |rho| > 0.5 (%d of %d); values shown where |rho| >= 0.5", length(inv), ncol(C)),
  silent = TRUE)
grid::grid.newpage(); grid::grid.draw(hm_top$gtable)
if (isTRUE(params$save_plots)) {
  pdf(here("plots", paste0("14_top_correlated_heatmap_", params$rf_suffix, ".pdf")), width = 11, height = 10)
  grid::grid.newpage(); grid::grid.draw(hm_top$gtable); invisible(dev.off())
}
reg_write_csv(top %>% dplyr::select(a, b, rho, abs_rho, fam_a, fam_b, same_family),
              file.path(out_dir, paste0("featcorr_top_pairs_", params$rf_suffix, ".csv")))
```

# 9. Which family members are in each correlated cluster

One block per multi-feature cluster, at each threshold. Bars are the member's SHAP share (si3d
hypoxia 1hr full model), coloured by its **family**, so a cluster that mixes families shows two or
more colours. The label is the member's Spearman rho with the cluster leader (its highest-SHAP
member). Clustering: average linkage on 1 - |rho|, cut at |rho| = threshold (section 2).

```{r cluster_membership, fig.width = 12, fig.height = 9}
cm <- clus %>%
  group_by(threshold, cluster) %>%
  mutate(n_fam = n_distinct(family),
         block = sprintf("%s (+%d)%s", first(leader), n() - 1, ifelse(n_fam > 1, "  [mixed]", ""))) %>%
  ungroup() %>%
  mutate(threshold = factor(sprintf("|rho| >= %.1f", threshold)),
         feature_lab = paste0(feature, "  (", family, ")"))
cat("multi-feature clusters per threshold:\n")
print(as.data.frame(cm %>% distinct(threshold, cluster, n_fam) %>% group_by(threshold) %>%
  summarise(clusters = n(), mixed_family = sum(n_fam > 1), .groups = "drop")), row.names = FALSE)
cm_pal <- fam_pal[sort(unique(cm$family))]
p_cm <- ggplot(cm, aes(x = share_pct, y = reorder(feature_lab, share_pct), fill = family)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = ifelse(feature == leader, "leader", sprintf("rho %+.2f", rho_to_leader))),
            hjust = -0.08, size = 2.5) +
  facet_grid(block ~ threshold, scales = "free_y", space = "free_y") +
  scale_fill_manual(values = cm_pal) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.45))) +
  labs(x = "SHAP share (%), si3d hypoxia 1hr full model", y = NULL,
       title = "Correlated feature clusters and their family members",
       subtitle = sprintf("%d features, %d MDA-MB-231 transcripts; average-linkage clusters on 1 - |rho|; label = rho with the cluster leader",
                          ncol(C), nrow(X))) +
  theme_bw(base_size = 8) +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), legend.position = "bottom",
        panel.grid.major.y = element_blank())
print(p_cm)
ggsave_if(here("plots", paste0("14_cluster_membership_", params$rf_suffix, ".pdf")), p_cm,
          width = 12, height = 9)
```
