---
title: "SHAP heatmap: features by experiment"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
params:
  experiments: "MDA-MB-231 3d hypoxia 1hr=eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3d hypoxia 4hr=eif33d_promotes_hypoxia_4hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3d hypoxia 24hr=eif33d_promotes_hypoxia_24hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3d normoxia 1+4hr=eif33d_promotes_normoxia_1and4hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 1hr=eif33e_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 4hr=eif33e_promotes_hypoxia_4hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 24hr=eif33e_promotes_hypoxia_24hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw;MDA-MB-231 3e normoxia 1+4hr=eif33e_promotes_normoxia_1and4hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  floor_preset: "transform"
  floor_arm: "raw"
  corr_threshold: 0.8
  clusters_suffix: ""
  reduced_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_reduced_corrected_both_clipaggregate_yraw"
  save_plots: TRUE
  out_tag: ""
  allow_mixed_genes: FALSE
  footnote: ""
  dir_cutoff: 0.3
  n_beeswarm: 15
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "15_shap_heatmap.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It builds the SHAP heatmap this whole fork exists to produce: **features as rows, experiments as
columns**, from the random-forest regression's SHAP values.

By default every column must be built on the same genes. `allow_mixed_genes = TRUE` relaxes
that so MDA-MB-231 and MCF7-SIX1 columns can sit in one figure: the feature set is still asserted
identical, but each column's shares are then computed over its own gene population, so a
difference between cell lines mixes biology with gene-population differences. The overlap is
reported below. The contrast is log2(sictrl / si3d) in every column (MCF7-SIX1 and HeLa negated to match),
so a positive outcome means the factor promotes that transcript's translation.

Right now it has **one column**. Everything is written so that adding an experiment is one entry
in the `experiments` param and nothing else changes.

**Decisions**

- The cell value is **`share_pct`**, each feature's percentage of that experiment's total
  \|SHAP\|. This is the only quantity comparable across experiments without adjustment: raw SHAP
  is in log2 TE units, and the outcome's spread varies about 7-fold across contrasts, so raw
  magnitudes are not on one scale. `10d_target_scale_analysis.Rmd` measures this.
- Both views are produced, because they answer different questions. **Individual features** show
  what the model actually used. **Grouped** rows fix the fact that correlated features split a
  single signal between them; summing SHAP within a group is exact, since SHAP is additive.
- Two groupings are shown, not one: by **family** (the prior grouping, which the annotation table
  defines) and by **correlated cluster** (read off the data by `14_feature_correlation.Rmd`).
  They disagree in the one place that matters, and that disagreement is the point.
- Rows are ordered by magnitude while there is one column. With more than one, columns get
  clustered and rows ordered by their mean.
- A feature whose share sits below its own measured noise floor is marked, so a visible cell is
  not mistaken for a real one.

```{r libraries}
library(tidyverse)
library(here)
library(pheatmap)
library(grid)      # grid.newpage / grid.draw, for drawing a captured pheatmap gtable
```

```{r load}
out_dir <- here("output", "translational_control_overview")

# The experiment list. One "label=suffix" pair per experiment, semicolon-separated. Adding a
# second experiment means adding one pair here; no other line in this notebook changes.
parse_experiments <- function(s) {
  parts <- str_trim(str_split(s, ";")[[1]])
  parts <- parts[nzchar(parts)]
  tibble(label  = str_trim(str_replace(parts, "=.*$", "")),
         suffix = str_trim(str_replace(parts, "^[^=]*=", "")))
}
exps <- parse_experiments(params$experiments)
cat("experiments:", nrow(exps), "\n")
print(as.data.frame(exps), row.names = FALSE)

read_one <- function(label, suffix) {
  f <- file.path(out_dir, paste0("shapreg_feature_", suffix, ".csv"))
  if (!file.exists(f)) { warning("missing: ", f); return(NULL) }
  read_csv(f, show_col_types = FALSE) %>%
    transmute(experiment = label, feature, family,
              share = share_pct, direction = shap_value_cor)
}
long <- pmap_dfr(list(exps$label, exps$suffix), read_one)
stopifnot("no SHAP outputs found for any experiment" = nrow(long) > 0)

# Every experiment must describe the same features, or a row means different things in different
# columns. With one experiment this is trivially true; it is asserted now so that it fails loudly
# the first time a second experiment with a different feature set is added.
per_exp <- long %>% count(experiment, name = "n_features")
print(as.data.frame(per_exp), row.names = FALSE)
if (nrow(exps) > 1) {
  # Identical feature SETS, not merely identical counts. Two experiments could each have 69
  # features and not the same 69, in which case a row would mean different things in different
  # columns and the whole comparison would be silently wrong.
  sets <- split(long$feature, long$experiment)
  stopifnot("experiments do not share a feature set - rows would not be comparable" =
              length(unique(map(sets, sort))) == 1)
  # The same applies to the genes behind each column: the shares are computed over whatever
  # transcripts that contrast could score, so a column built on a different gene population is
  # not comparable even with identical features.
  gene_counts <- map_dbl(exps$suffix, function(sf) {
    f <- file.path(out_dir, paste0("rfreg_performance_", sf, ".csv"))
    if (!file.exists(f)) return(NA_real_)
    d <- read_csv(f, show_col_types = FALSE); d$value[d$metric == "n_genes"][1]
  })
  names(gene_counts) <- exps$label
  print(gene_counts)
  if (!isTRUE(params$allow_mixed_genes))
    stopifnot("columns were built on different numbers of genes - shares are not comparable" =
                length(unique(na.omit(gene_counts))) == 1)
  # Every suffix must be distinct, or two columns are the same run under two labels.
  stopifnot("two experiments resolve to the same suffix" = !any(duplicated(exps$suffix)))

  # Identical GENE sets, not just identical counts. Two contrasts could each score 10,172
  # transcripts and not the same 10,172, in which case the shares are computed over different
  # populations and the columns are not comparable even with identical features.
  gene_sets <- map(exps$suffix, function(sf) {
    f <- file.path(out_dir, paste0("rfreg_predictions_", sf, ".csv"))
    if (!file.exists(f)) return(NULL)
    sort(read_csv(f, show_col_types = FALSE)$gene_id_clean)
  })
  names(gene_sets) <- exps$label
  gene_sets <- compact(gene_sets)
  if (length(gene_sets) > 1) {
    same <- length(unique(gene_sets)) == 1
    if (!same) {
      base <- gene_sets[[1]]
      for (nm in names(gene_sets)[-1])
        cat(sprintf("  %s differs from %s by %d genes\n", nm, names(gene_sets)[1],
                    length(setdiff(union(base, gene_sets[[nm]]),
                                   intersect(base, gene_sets[[nm]])))))
    }
    if (isTRUE(params$allow_mixed_genes)) {
      if (!same) {
        shared <- Reduce(intersect, gene_sets)
        cat(sprintf("\n*** MIXED GENE POPULATIONS (allow_mixed_genes = TRUE): columns score %d to %d genes;\n",
                    min(lengths(gene_sets)), max(lengths(gene_sets))))
        cat(sprintf("    %d genes are common to all %d columns. Shares are within-column ratios over\n",
                    length(shared), length(gene_sets)))
        cat("    each column's own genes, so cross-cell-line differences are not purely biological. ***\n")
      }
    } else {
      stopifnot("columns were built on different GENE SETS - shares are not comparable" = same)
      cat(sprintf("gene-set identity: all %d columns share the same %d genes\n",
                  length(gene_sets), length(gene_sets[[1]])))
    }
  }
}
stopifnot("shares do not sum to 100 within an experiment" =
            all(abs(tapply(long$share, long$experiment, sum) - 100) < 1e-6))
# Each column's own performance, so the heatmap says how good the model behind it was. A column
# from a model that barely ranks its own genes should not be read like one that ranks them well.
read_perf <- function(label, suffix) {
  f <- file.path(out_dir, paste0("rfreg_performance_", suffix, ".csv"))
  if (!file.exists(f)) return(tibble(experiment = label, auc = NA_real_, spearman = NA_real_))
  d <- read_csv(f, show_col_types = FALSE)
  g <- function(k) { v <- d$value[d$metric == k]; if (length(v)) v[1] else NA_real_ }
  # An AUC computed against a handful of held-out positives is not a measurement. si3d hypoxia
  # 24hr puts ONE positive gene in its test split. Record it, but do not display it as though it
  # were comparable to a column with 294.
  # The self-contained AUC: positives are a threshold on this contrast's own outcome, negatives
  # are every other held-out gene. No external gene set, so it is comparable across knockdowns.
  np <- g("auc_n_pos"); if (is.na(np)) np <- g("bridge_n_pos")
  a  <- g("auc");       if (is.na(a))  a  <- g("bridge_auc_threshold")
  tibble(experiment = label, auc = a, spearman = g("spearman_raw_y"),
         bridge_n_pos = np,
         auc_usable = !is.na(np) && np >= 20)
}
perf <- pmap_dfr(list(exps$label, exps$suffix), read_perf)
# The outcome's own spread per column. Shares are ratios within a run and so are unaffected by
# it, but raw SHAP is in log2 TE units and is NOT comparable when these differ - which they do,
# by several fold across these contrasts. This table is why the heatmap plots share_pct.
spread <- pmap_dfr(list(exps$label, exps$suffix), function(l, sf) {
  f <- file.path(out_dir, paste0("rfreg_performance_", sf, ".csv"))
  if (!file.exists(f)) return(tibble(experiment = l))
  d <- read_csv(f, show_col_types = FALSE)
  g <- function(k) { v <- d$value[d$metric == k]; if (length(v)) v[1] else NA_real_ }
  tibble(experiment = l, n_genes = g("n_genes"), n_features = g("n_features"),
         outcome_median = g("target_median"), outcome_mad = g("target_mad"))
})
perf <- perf %>% left_join(spread, by = "experiment")
print(as.data.frame(perf %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)
if (sum(!is.na(perf$outcome_mad)) > 1)
  cat(sprintf("\noutcome MAD ranges %.3f to %.3f across columns, a %.1f-fold difference.\n",
              min(perf$outcome_mad, na.rm = TRUE), max(perf$outcome_mad, na.rm = TRUE),
              max(perf$outcome_mad, na.rm = TRUE) / min(perf$outcome_mad, na.rm = TRUE)),
      "Shares are comparable because they are within-run ratios; raw SHAP would not be.\n")

# Column annotation: the AUC bridge, and the held-out Spearman beside it. The colour bars give
# the at-a-glance comparison once there are several columns.
# Named for what they are. "bridge AUC" was my own shorthand and needed explaining, which is a
# sign it should not be on a figure: it is the AUC of the regression's predicted TE scored
# against notebook 02's classifier gene sets, which exist only for si3d. Spearman is the
# regression's own held-out metric and applies to every column.
# Condition and initiation factor, parsed from the SUFFIX rather than the label, so the
# annotation cannot disagree with the run it describes if a label is ever reworded.
meta <- tibble(experiment = exps$label, suffix = exps$suffix) %>%
  # Suffixes are paste0("eif3", gene, ...), so a non-eIF3 gene still carries the "eif3" stem.
  mutate(factor_kd = case_when(str_detect(suffix, "^eif3dhx29_") ~ "DHX29",
                               str_detect(suffix, "^eif3dap5_")  ~ "DAP5",
                               str_detect(suffix, "^eif34e_")   ~ "eIF4E",
                               str_detect(suffix, "^eif33d_") ~ "eIF3d",
                               str_detect(suffix, "^eif33e_") ~ "eIF3e",
                               TRUE ~ NA_character_),
         # Cap-binding columns (subunit-seq, HEK293T) have no oxygen condition; the track says
         # which kind of binding outcome instead. See CAP_BINDING_DATA.md.
         condition = case_when(str_detect(suffix, "_capbind_(glu|tg)vs") ~ "cap binding, stress ratio",
                               str_detect(suffix, "_capbind_")  ~ "cap binding",
                               # Herrmannova HeLa is a steady-state knockdown, not a hypoxia
                               # experiment, so it is labelled "no stress" (a light tint of normoxia's green).
                               # Weber DAP5 (HEK293T) is a steady-state knockout, likewise.
                               str_detect(suffix, "_hela_|_dhx29ip_|_hek293t_normoxia_") ~ "no stress",
                               str_detect(suffix, "_hypoxia_")  ~ "hypoxia",
                               str_detect(suffix, "_normoxia_") ~ "normoxia",
                               TRUE ~ NA_character_),
         # Cell line, ready for the ones still to come. Everything present is MDA-MB-231, so
         # this track is currently one flat colour - it exists so a second line differentiates
         # automatically rather than needing the figure reworked. MCF7-SIX1 carries "mcf7six1"
         # in its condition string by this project's naming convention.
         cell_line = case_when(str_detect(suffix, "_hek293t_")            ~ "HEK293T",
                               str_detect(suffix, "_hela_")               ~ "HeLa",
                               str_detect(suffix, "mcf7six1|mcf7_six1") ~ "MCF7-SIX1",
                               str_detect(suffix, "mcf7")              ~ "MCF7",
                               TRUE                                     ~ "MDA-MB-231"),
         # Which cell line's dominant isoforms the feature matrix was built on. MCF7-SIX1 was
         # fitted twice - on the 231 matrix and on its own (the _mcf7tx runs) - and without this
         # track those two columns would be indistinguishable in a combined figure.
         # The DAP5 column is the one dataset fitted on MANE Select (_manetx) rather than its
         # own most abundant transcript; its label carries a star and the figure a footnote.
         isoforms = case_when(str_detect(suffix, "_mcf7tx")  ~ "MCF7-SIX1",
                              str_detect(suffix, "_helatx")  ~ "HeLa",
                              str_detect(suffix, "_hektx")   ~ "HEK293T",
                              str_detect(suffix, "_manetx")  ~ "MANE Select",
                              TRUE                           ~ "MDA-MB-231"))
stopifnot("a column label carries a star but no footnote explains it" =
            !any(grepl("*", meta$experiment, fixed = TRUE)) || nzchar(params$footnote),
          "a MANE-isoform column is not starred" =
            all(grepl("*", meta$experiment[meta$isoforms == "MANE Select"], fixed = TRUE)))
if (nzchar(params$footnote)) cat(params$footnote, "\n")
stopifnot("could not parse the knockdown or condition from a suffix" =
            !any(is.na(meta$factor_kd)) && !any(is.na(meta$condition)))
cat("cell lines present:", paste(sort(unique(meta$cell_line)), collapse = ", "), "\n")
print(as.data.frame(meta %>% dplyr::select(experiment, factor_kd, condition)), row.names = FALSE)

# Only the two CATEGORICAL tracks are coloured: which factor was knocked down, and which
# condition. Spearman goes in the column label as a number instead of a colour bar - a
# continuous annotation next to a continuous heatmap body reads as a second data layer and
# competes with the values it is meant to contextualise.
#
# AUC is not shown at all. Any AUC needs the continuous outcome thresholded, which reintroduces
# the arbitrary cutoff this regression exists to avoid, and it degenerates exactly where it
# matters: te_lfc > 0.5 selects about half the transcriptome at si3e normoxia 1+4hr and leaves a
# single held-out positive at si3d hypoxia 24hr. Spearman needs no threshold, applies to every
# column identically, and is what the model actually ranks. It is still written to
# rfreg_performance_* for anyone who wants it.
anno_col <- perf %>%
  left_join(meta %>% dplyr::select(experiment, factor_kd, condition, cell_line, isoforms),
            by = "experiment") %>%
  transmute(experiment,
            `cell line`         = cell_line,
            isoforms            = isoforms,
            `initiation factor` = factor_kd,
            condition           = condition,
            rho                 = spearman) %>%
  column_to_rownames("experiment")
# The isoforms track only carries information when some column was fitted on another cell line's
# isoforms. Otherwise it duplicates the cell-line track exactly, so it is dropped.
if (all(meta$isoforms == meta$cell_line)) anno_col$isoforms <- NULL
# With a single column rho has no range, so its colour bar says nothing and its legend degenerates
# to "0.5 ... 0.5"; the value is already printed in the column label.
if (nrow(anno_col) == 1) anno_col$rho <- NULL

# The two reserved project hues, used here for exactly what they are reserved for: eIF3d is
# Okabe-Ito orange and eIF3e Okabe-Ito light blue throughout this project. Condition therefore
# takes two DIFFERENT Okabe-Ito hues, so nothing in the figure can be misread as a knockdown -
# and neither is #0072B2 or #D55E00, which the diverging body scale already uses.
# One colour per meaning across the whole figure, all Okabe-Ito:
#   orange  #E69F00 -> eIF3d      sky     #56B4E9 -> eIF3e   (both reserved project-wide)
#   purple  #CC79A7 -> hypoxia    green   #009E73 -> normoxia
#   black           -> a feature whose direction does not reproduce across seeds
#   #0072B2 and #D55E00 are the diverging body scale, so they are not reused in annotations.
# Cell-line colours are assigned by NAME, so a line added later keeps its colour wherever it
# appears rather than depending on what else is in the figure.
cell_pal <- c(`MDA-MB-231` = "#F0E442",   # the one unused Okabe-Ito hue
              `MCF7-SIX1`  = "#661100",
              `MCF7`       = "#AA4499",
              `HeLa`       = "#999933",   # Tol olive; Herrmannova et al. 2024
              `HEK293T`    = "#44AA99")   # Tol teal; subunit-seq cap binding

anno_col_colors <- list(
  `cell line`         = cell_pal[intersect(names(cell_pal), unique(meta$cell_line))],
  isoforms            = c(`MDA-MB-231` = "grey85", `MCF7-SIX1` = "grey35", `HeLa` = "grey60",
                          `HEK293T` = "grey20", `MANE Select` = "white")[unique(meta$isoforms)],
  `initiation factor` = c(`eIF3d` = "#E69F00", `eIF3e` = "#56B4E9",
                          `DHX29` = "#882255",                           # Tol wine
                          `eIF4E` = "#CC6677",                           # Tol rose
                          `DAP5`  = "#DDCC77")[unique(meta$factor_kd)],  # Tol sand
  condition           = c(hypoxia = "#CC79A7", normoxia = "#009E73", `no stress` = "#8FD3BD",   # light tint of normoxia: unstressed, but distinguishable
                          `cap binding` = "#332288",                 # Tol indigo
                          # lighter indigo: a variant of cap binding, and not DAP5's sand
                          `cap binding, stress ratio` = "#8C7FC7")[unique(meta$condition)],
  # rho is the only CONTINUOUS track, so it gets a single-hue sequential ramp rather than a
  # categorical set. Grey deliberately: the heatmap body is a blue-white-vermillion diverging
  # scale, and any coloured ramp beside it competes with the values it is meant to contextualise.
  # Grey reads as "this is a metric about the column", not as another data layer.
  rho                 = colorRampPalette(c("grey93", "grey20"))(100))

# pheatmap cannot print text inside an annotation bar, so the numbers go into the column label
# instead. That puts them immediately above their own colour bars, which is where they are
# wanted: the colour says how this column compares, the number says what it actually was.
# AUC only where the classifier bridge was run (si3d). Spearman is the regression's own metric
# and is shown for every column.
perf$auc[!perf$auc_usable %in% TRUE] <- NA_real_   # underpowered AUCs are not shown
# ONE line. Column labels are rotated, so a newline stacks the lines sideways and they
# interleave with the neighbouring column - the labels become unreadable. The metrics go in
# parentheses on the same line instead, and the annotation bars carry them as colour too.
# rho only. AUC is deliberately NOT shown, because across these columns it varies with how many
# genes happen to clear the LFC cut rather than with how well the model ranks: si3d hypoxia 24hr
# has ONE held-out positive, so its AUC of 0.41 is a single gene's rank, while si3e normoxia
# 1+4hr has 996 - 49% of the held-out set - so its 0.79 is close to "above median". Putting the
# two side by side invites a comparison that is about the threshold, not the model. That
# threshold is what the regression exists to avoid. rho needs none and means the same thing in
# every column. AUC is still written to rfreg_performance_* for anyone who wants it.
col_labels <- setNames(sprintf("%s  (rho %.2f)", perf$experiment, perf$spearman),
                       perf$experiment)
lab_for <- function(m) unname(col_labels[colnames(m)])

# Did every column's SHAP decomposition actually close? 11_ asserts it per run, but a heatmap
# that puts eight runs side by side should confirm it for all of them at the point of comparison.
#
# EVERY column must be checked - a column with no record fails, rather than being skipped. The
# shapreg_qc_ file was added to 11_ after some runs were made (si3d hypoxia 1hr and 4hr have
# none), so where it is missing the closure is recomputed from shapreg_per_gene_, which holds each
# explained gene's prediction and its per-feature SHAP. That recompute reproduces the stored
# closure_sd and baseline exactly on the columns that have both.
closure_from_per_gene <- function(sf) {
  f <- file.path(out_dir, paste0("shapreg_per_gene_", sf, ".csv"))
  if (!file.exists(f)) return(NULL)
  d <- read_csv(f, show_col_types = FALSE)
  fc <- setdiff(names(d), c("gene_id_clean", "symbol", "y", "pred"))
  implied <- d$pred - rowSums(d[fc])
  tibble(n_features = length(fc), explain_n = nrow(d), baseline = mean(implied),
         closure_sd = sd(implied))
}
qc <- map_dfr(seq_len(nrow(exps)), function(i) {
  f  <- file.path(out_dir, paste0("shapreg_qc_", exps$suffix[i], ".csv"))
  pg <- closure_from_per_gene(exps$suffix[i])
  if (file.exists(f)) {
    q <- read_csv(f, show_col_types = FALSE) %>% mutate(source = "shapreg_qc")
    # Where both exist they must agree, or the per-gene fallback is not a valid substitute.
    if (!is.null(pg))
      stopifnot("per-gene SHAP does not reproduce the stored closure record" =
                  abs(pg$baseline - q$baseline) < 1e-9 && abs(pg$closure_sd - q$closure_sd) < 1e-9)
  } else if (!is.null(pg)) {
    q <- pg %>% mutate(units = NA_character_, target_transform = NA_character_,
                       source = "recomputed from per-gene SHAP")
  } else {
    q <- tibble(closure_sd = NA_real_, source = "none")
  }
  q %>% mutate(experiment = exps$label[i])
})
cat("\nSHAP closure per column (must be ~0; 11_ asserts < 1e-8):\n")
print(as.data.frame(qc %>% dplyr::select(any_of(c("experiment", "n_features", "explain_n",
                                                  "baseline", "closure_sd", "source"))) %>%
  mutate(across(any_of("baseline"), ~ round(., 5)))), row.names = FALSE)
stopifnot("a column has no SHAP closure record (no shapreg_qc_ or shapreg_per_gene_ file)" =
            !any(is.na(qc$closure_sd)))
stopifnot("a column's SHAP decomposition did not close" = all(qc$closure_sd < 1e-8))
# Units must match, or the shares were computed from differently-scaled SHAP. The transform is
# also in every suffix (_yraw), which covers the recomputed columns.
stopifnot("columns were built with different target transforms" =
            length(unique(na.omit(qc$target_transform))) <= 1,
          "columns were built with different target transforms (suffix)" =
            length(unique(str_extract(exps$suffix, "_y[a-z]+"))) == 1)

# Two diagnostics that decide whether a column should be read at all.
diag_tbl <- perf %>%
  mutate(weak_model = spearman < 0.2,
         auc_meaningful = !is.na(auc))
if (any(diag_tbl$weak_model, na.rm = TRUE)) {
  cat("\n*** WEAK COLUMNS: test Spearman < 0.2, so the model found little to learn. Their SHAP\n")
  cat("    is attribution of near-noise and must not be read like an informative column: ***\n")
  for (l in diag_tbl$experiment[which(diag_tbl$weak_model)])
    cat(sprintf("      %s (Spearman %.3f)\n", l, diag_tbl$spearman[diag_tbl$experiment == l]))
}
cat("\nColumn labels carry rho only: Spearman of predicted vs observed TE on held-out genes.\n")
cat("No threshold, identical meaning in every column, and the quantity the forest ranks.\n\n")
cat("AUC is computed and written to rfreg_performance_* but is NOT shown. Across these columns\n")
cat("it tracks how many genes clear the LFC cut rather than model quality - one held-out\n")
cat("positive at si3d hypoxia 24hr against 996 at si3e normoxia 1+4hr - so comparing it between\n")
cat("columns compares thresholds, not models.\n")
if (any(!diag_tbl$auc_meaningful)) {
  cat("\nColumns with no AUC: the bridge compares against notebook 02's classifier gene sets,\n")
  cat("which exist only for si3d. For si3e no classifier comparison is wanted, and a fixed LFC\n")
  cat("cut would be degenerate there anyway - si3e normoxia 1+4hr has median te_lfc +0.619, so\n")
  cat("te_lfc > 0.5 selects about half the transcriptome. That is the case the regression exists\n")
  cat("to handle: the continuous outcome needs no threshold. Spearman is the metric for those.\n")
  for (l in diag_tbl$experiment[!diag_tbl$auc_meaningful]) cat(sprintf("      %s\n", l))
}

# Output paths carry out_tag, so rendering a second feature set (e.g. the intrinsic-only model)
# writes alongside the first instead of overwriting it. Without this the 57-feature run would
# silently replace the 69-feature figures.
TAG <- if (nzchar(params$out_tag)) paste0("_", params$out_tag) else ""
tagf <- function(stem, ext) paste0(stem, TAG, ".", ext)
cat("output tag:", ifelse(nzchar(TAG), TAG, "(none, default feature set)"), "\n")

ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)
```

```{r annotations}
# The direction-stability flags and the family noise floor below were each MEASURED on one
# experiment: the floor_preset / floor_arm split-seed sweep, which is si3d hypoxia 1hr in
# MDA-MB-231 on the full feature set. The floor is driven by gene count and outcome spread, so it
# does not transfer to another contrast, cell line or feature set. They are therefore shown ONLY
# when every column is that experiment, and dropped from the figure otherwise - a bar that looks
# like it covers every column but was measured on one of them is worse than no bar.
source(here("code", "translational_control_overview", "10_regression_presets.R"))
floor_arm_def <- REG_PRESETS[[params$floor_preset]][[params$floor_arm]]
stopifnot("floor_preset / floor_arm not found in REG_PRESETS" = !is.null(floor_arm_def))
# The sweep runs under its own feature_set_label (e.g. tcoreg_tf) only so its outputs do not
# overwrite the main run; its model is otherwise the main tcoreg configuration.
floor_measured_on <- reg_suffix(modifyList(floor_arm_def, list(feature_set_label = "tcoreg")))
floor_applies <- all(exps$suffix == floor_measured_on)
cat("noise floor and direction stability were measured on:", floor_measured_on, "\n")
cat(if (floor_applies) "  every column is that experiment - shown\n" else
      "  not every column is that experiment - both are DROPPED from this figure\n")

# Which features cannot be read directionally: their shap_value_cor sign flips across the 20
# split seeds, so the signed view must not be trusted for them.
dir_path <- file.path(out_dir, paste0("shapreg_direction_stability_",
                                      params$floor_preset, "_", params$floor_arm, ".csv"))
unstable <- if (file.exists(dir_path)) {
  d <- read_csv(dir_path, show_col_types = FALSE)
  d$feature[d$frac_pos > 0.05 & d$frac_pos < 0.95]
} else character(0)
if (!floor_applies) unstable <- character(0)
cat("features with an unstable SHAP direction:", length(unstable), "\n")
if (length(unstable)) cat("  ", paste(unstable, collapse = ", "), "\n")

# The measured per-family floor: how much a family's share moves when nothing biological changes.
floor_path <- file.path(out_dir, paste0("shapreg_noise_floor_",
                                        params$floor_preset, "_", params$floor_arm, ".csv"))
fam_floor <- if (file.exists(floor_path))
  read_csv(floor_path, show_col_types = FALSE) %>%
    dplyr::select(family, floor_sd = sd, floor_mdd = min_detectable_diff) else NULL
if (!floor_applies) fam_floor <- NULL
cat("family noise floor available:", !is.null(fam_floor), "\n")

# The data-driven grouping from 14_. Features not in a multi-feature cluster are their own group.
# Which run's correlated-feature clustering to use. Feature-feature correlation is a property of
# the FEATURE MATRIX, not of the outcome, and every column here is built from the same matrix, so
# one clustering legitimately describes all of them. clusters_suffix makes that reuse explicit:
# without it the grouping is looked up under the first column's suffix, and a figure whose first
# column is a cell line that 14_ was never run for silently loses the panel.
clus_suffix <- if (nzchar(params$clusters_suffix)) params$clusters_suffix else exps$suffix[1]
clus_path <- file.path(out_dir, paste0("featcorr_clusters_", clus_suffix, ".csv"))
if (nzchar(params$clusters_suffix))
  cat("correlated-cluster grouping taken from:", clus_suffix,
      "\n  (valid for every column here: they share one feature matrix, so feature-feature",
      "\n   correlation is identical; only the outcome differs between columns)\n")
corr_group <- if (file.exists(clus_path)) {
  cl <- read_csv(clus_path, show_col_types = FALSE) %>%
    filter(abs(threshold - params$corr_threshold) < 1e-9)
  cl %>% group_by(cluster) %>%
    summarise(group = paste0(first(leader), " +", n() - 1), features = list(feature),
              .groups = "drop") %>%
    unnest(features) %>% dplyr::select(feature = features, group)
} else NULL
cat("correlated-cluster grouping available:", !is.null(corr_group), "\n")
```

# 0. What is in these models, and what was excluded

```{r exclusion_criteria}
# Stated explicitly and derived from the data, so a reader of the figure knows what the columns
# were built from without having to reconstruct it from a param string.
ann_all <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
modelled <- sort(unique(long$feature))
candidate <- ann_all %>% filter(block %in% c("intrinsic", "external"), !deprecated,
                                variant %in% c("single", "corrected"))
excluded <- candidate %>% filter(!feature %in% modelled) %>%
  dplyr::select(feature, family, block)

cat(sprintf("features modelled in every column: %d\n", length(modelled)))
cat(sprintf("blocks represented: %s\n",
            paste(sort(unique(ann_all$block[match(modelled, ann_all$feature)])), collapse = ", ")))

if (nrow(excluded)) {
  cat(sprintf("\nEXCLUDED from these models: %d features\n\n", nrow(excluded)))
  ex_ext <- excluded %>% filter(block == "external")
  if (nrow(ex_ext)) {
    cat("  External measurements - data from other experiments, not computable from the\n")
    cat("  transcript sequence. Excluded so the model rests on intrinsic information only:\n")
    for (fam in sort(unique(ex_ext$family))) {
      fs <- ex_ext$feature[ex_ext$family == fam]
      cat(sprintf("    %-20s %s\n", fam, paste(fs, collapse = ", ")))
    }
  }
  ex_int <- excluded %>% filter(block != "external")
  if (nrow(ex_int)) {
    cat("\n  Other exclusions:\n")
    for (i in seq_len(nrow(ex_int)))
      cat(sprintf("    %-34s (%s, %s)\n", ex_int$feature[i], ex_int$family[i], ex_int$block[i]))
  }
  cat("\n  Note on what was KEPT: several retained features are computed from the sequence but\n")
  cat("  use weight tables estimated elsewhere - csc_internal (mRNA half-lives), cai/fop\n")
  cat("  (reference gene sets), tai_gtrnadb (tRNA gene copy number), g4mer_* (a model trained\n")
  cat("  on rG4-seq), noderer_tis_efficiency (FACS-seq), term_readthrough_cridge (a published\n")
  cat("  192-context lookup). These are sequence features with external weights, not external\n")
  cat("  measurements of these transcripts, which is the line drawn here.\n")
} else {
  cat("\nNo features excluded: these columns use the full candidate set.\n")
}

# Known gap, so it is visible in the report rather than only in a note.
if (!any(grepl("minus9|context_m9", modelled)))
  cat("\n  KNOWN GAP: termination_context_minus9, the one collaborator feature never built,\n",
      " is absent from the matrix entirely. It is pure sequence and belongs in an\n",
      " intrinsic-only model; building it requires returning to 01_ and the genome.\n")
```

# 1. Individual features

```{r fig_features, fig.width = 7, fig.height = 13}
# pheatmap has two failure modes when you want the figure BOTH inline and on disk, and this
# hits both. Re-drawing inside pdf() after an on-screen draw writes a blank page (163 bytes on a
# minimal example), because the second grid draw does not get a fresh page. And pheatmap's own
# filename= closes the active device - measured, device 2 becomes 3 - which clobbers knitr's and
# silently kills every inline figure after it.
#
# Capturing the gtable ONCE with silent = TRUE and drawing it twice avoids both: the device
# survives, the inline figure is captured, and the file is a real figure.
draw_and_save <- function(args, fname = NULL, width = 9, height = NULL,
                          cellheight = 13, scale_title = NULL) {
  # Name the colour scale. pheatmap has no legend title, so the title is the label of the top
  # legend break. Inferred from the scale unless given: every signed scale in this notebook is a
  # direction (correlation of a feature's value with its own SHAP, -1..1) and every unsigned one
  # is a share of total mean |SHAP|. These do NOT sum to the model's rho: per gene, SHAP sums to
  # that gene's prediction minus the mean prediction; shares sum to 100.
  if (!is.null(args$breaks) && is.null(args$legend_breaks)) {
    rng <- range(args$breaks)
    if (is.null(scale_title)) scale_title <- if (rng[1] < 0) "direction\ncor(value, SHAP)" else "SHAP share (%)"
    b <- pretty(rng, n = 5)
    b <- b[b >= rng[1] & b <= rng[2] - 0.15 * diff(rng)]
    args$legend_breaks <- c(b, rng[2])
    args$legend_labels <- c(format(b, trim = TRUE), scale_title)
  }
  # Rows get a FIXED height in points and the page grows to fit, rather than a fixed page
  # squeezing the rows. With 11 families in a 4-inch page the row labels overlapped and the
  # dendrogram took most of the height; a fixed cellheight makes every row legible whether
  # there are 11 of them or 69.
  if (is.null(args$cellheight)) args$cellheight <- cellheight
  # No title on the figure. What each panel shows is stated in the notebook text beside it, and
  # the multi-line titles competed with the column dendrogram for space.
  args$main <- NA
  ph <- do.call(pheatmap, c(args, list(silent = TRUE)))
  # The footnote (e.g. the MANE star) is drawn on the figure itself, so it travels with the PDF.
  foot <- function() if (nzchar(params$footnote))
    grid.text(params$footnote, x = unit(0.1, "in"), y = unit(0.08, "in"),
              just = c("left", "bottom"), gp = gpar(fontsize = 8))
  grid.newpage(); grid.draw(ph$gtable); foot()
  if (!is.null(fname) && isTRUE(params$save_plots)) {
    # Page height is COMPUTED, not assumed. A flat overhead clipped the title once the column
    # labels grew: they are rotated, so their height scales with the longest label's character
    # count rather than being constant.
    if (is.null(height)) {
      lab_txt   <- if (!is.null(args$labels_col)) args$labels_col else colnames(args$mat)
      dendro_in <- if (isTRUE(args$cluster_cols)) 1.2 else 0.2
      n_tracks  <- (if (!is.null(args$annotation_col)) ncol(args$annotation_col) else 0) +
                   (if (!is.null(args$annotation_row)) 1 else 0)
      height <- cellheight * nrow(args$mat) / 72 +   # the matrix itself
                0.075 * max(nchar(lab_txt)) +        # rotated column labels, ~9pt
                dendro_in + 0.22 * n_tracks + 0.9 +  # dendrogram, annotation bars, title
                (if (nzchar(params$footnote)) 0.3 else 0)
    }
    # Width grows with the column count for the same reason. A fixed 9 in page clipped the
    # leftmost columns once a 16-column cross-cell-line figure was drawn.
    cw <- if (!is.null(args$cellwidth)) args$cellwidth else 42
    width <- max(width, cw * ncol(args$mat) / 72 +      # the matrix itself
                        0.07 * max(nchar(rownames(args$mat))) +   # row labels
                        (if (!is.null(args$annotation_row)) 0.5 else 0) + 3.2)  # legends
    # Then never narrower than the drawn figure itself. The estimate above is a floor; legend
    # widths (e.g. a two-line scale title) are absolute and not in it, and a page narrower than
    # the gtable clips BOTH edges because pheatmap centres the figure.
    need <- grid::convertWidth(sum(ph$gtable$widths), "in", valueOnly = TRUE)
    if (is.finite(need)) width <- max(width, need + 0.4)
    pdf(file.path(here("plots"), fname), width = width, height = height)
    grid.newpage(); grid.draw(ph$gtable); foot(); dev.off()
  }
  invisible(ph)
}

mat_of <- function(d, row, val = "share") {
  d %>% dplyr::select(all_of(c(row, "experiment", val))) %>%
    pivot_wider(names_from = experiment, values_from = all_of(val)) %>%
    column_to_rownames(row) %>% as.matrix()
}
ord_rows <- function(m) m[order(rowMeans(m, na.rm = TRUE), decreasing = TRUE), , drop = FALSE]

M <- ord_rows(mat_of(long, "feature"))
fam_of <- long %>% distinct(feature, family) %>% deframe()

# "reproducible" means the feature's direction came out with the same sign in all 20 split-seed
# runs. "flips" means it did not, so the colour in the direction panel is not reproducible for
# that row and should be read as magnitude only. 17 of 69 features flip, carrying 6.3% of the
# total SHAP between them - every large feature is consistent.
anno_row <- data.frame(
  family              = fam_of[rownames(M)],
  `direction repeats` = ifelse(rownames(M) %in% unstable, "flips", "reproducible"),
  row.names = rownames(M), check.names = FALSE)
# Measured on one experiment only; see the annotations chunk. Without this the bar would read
# "reproducible" for every row, which is a claim nobody measured.
if (!floor_applies) anno_row$`direction repeats` <- NULL
fam_cols <- setNames(
  colorRampPalette(c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                     "#0072B2", "#D55E00", "#CC79A7", "grey50"))(n_distinct(anno_row$family)),
  sort(unique(anno_row$family)))

pheat <- function(m, main, annor, annoc = NULL, signed = FALSE, fname = NULL, ...) {
  if (signed) {
    lim <- max(abs(m), na.rm = TRUE)
    col <- colorRampPalette(c("#0072B2", "white", "#D55E00"))(101)
    brk <- seq(-lim, lim, length.out = 102)
  } else {
    col <- colorRampPalette(c("white", "#FDD9A0", "#E69F00", "#D55E00"))(101)
    brk <- seq(0, max(m, na.rm = TRUE), length.out = 102)
  }
  args <- list(mat = m, color = col, breaks = brk,
               cluster_rows = FALSE, cluster_cols = ncol(m) > 2,
               annotation_row = annor, annotation_col = anno_col,
               annotation_colors = c(annoc, anno_col_colors),
               display_numbers = ncol(m) <= 3, number_format = "%.1f", fontsize_number = 7,
               labels_col = lab_for(m),
               fontsize_row = 8, fontsize_col = 9, cellwidth = 42, main = main, ...)
  draw_and_save(args, fname)
}

pheat(M, sprintf("SHAP share (%%) per feature, %d experiment%s",
                 ncol(M), if (ncol(M) == 1) "" else "s"),
      anno_row, list(family = fam_cols,
                     `direction repeats` = c(reproducible = "grey85", flips = "black")),
      fname = tagf("15_heatmap_features_magnitude", "pdf"))
cat(sprintf("\n%d features | top: %s (%.2f%%) | bottom: %s (%.2f%%)\n",
            nrow(M), rownames(M)[1], M[1, 1], rownames(M)[nrow(M)], M[nrow(M), 1]))
cat(sprintf("the top 10 features carry %.1f%% of the total\n", sum(M[1:10, 1])))
```

```{r fig_features_combined, fig.width = 7.5, fig.height = 13}
# One panel carrying both quantities, because they are different measurements and neither
# substitutes for the other:
#
#   share_pct      0 to ~12    how much of the total |SHAP| the feature holds
#   shap_value_cor -1 to +1    which way it pushes, from the correlation of the feature's value
#                              with its own SHAP value
#
# They are independent. karner2026_mda231_log2_ct holds a large share (3.55%) with a weak
# direction (-0.43); g4mer_frac_above_utr5 holds a tiny share (0.64%) with a strong direction
# (+0.81). Multiplying them into a single number would render both as a mid-sized, mid-confident
# feature and lose the distinction, so they get separate visual channels instead.
#
#   COLOUR  = direction. Vermillion means a higher feature value pushes the predicted TE change
#             up (more eIF3d-promoted), dark blue means down, white is 0. Symmetric across the
#             full [-1, 1] range, so colour is comparable between panels and experiments.
#   NUMBER  = the share, printed in the cell.
#
# Rows stay ordered by share, so vertical position is magnitude and colour is direction.
Md <- mat_of(long, "feature", "direction")[rownames(M), , drop = FALSE]

comb_args <- list(
  mat = Md,
  color  = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  cluster_rows = FALSE, cluster_cols = ncol(Md) > 2,
  annotation_row = anno_row, annotation_col = anno_col,
  annotation_colors = c(list(family = fam_cols,
                             `direction repeats` = c(reproducible = "grey85", flips = "black")),
                        anno_col_colors),
  # The printed number is the SHARE, not the value being coloured.
  # Printed shares stop being legible past about four columns.
  display_numbers = if (ncol(Md) <= 4) round(M[rownames(Md), , drop = FALSE], 1) else FALSE,
  fontsize_number = 7, number_color = "black",
  labels_col = lab_for(Md),
  # A feature whose SHAP values are effectively constant has no defined direction - the
  # correlation is undefined, not zero. Those cells are grey, and are counted separately below
  # rather than being folded into the up or down totals.
  na_col = "grey88",
  fontsize_row = 8, fontsize_col = 9, cellwidth = 46,
  main = "Colour = direction (vermillion up, blue down, grey undefined)\nNumber = SHAP share (%), rows ordered by it")
draw_and_save(comb_args, tagf("15_heatmap_features_combined", "pdf"), width = 9,
              scale_title = "colour: direction\nnumber: share (%)")

n_na <- sum(is.na(Md[, 1]))
cat(sprintf("\ndirection: %d push UP, %d push DOWN, %d undefined\n",
            sum(Md[, 1] > 0, na.rm = TRUE), sum(Md[, 1] < 0, na.rm = TRUE), n_na))
cat(sprintf("strongly directional (|cor| > 0.8): %d of %d | ambiguous (|cor| < 0.5): %d\n",
            sum(abs(Md[, 1]) > 0.8, na.rm = TRUE), nrow(Md),
            sum(abs(Md[, 1]) < 0.5, na.rm = TRUE)))
if (n_na)
  cat(sprintf("  undefined: %s - SHAP is effectively constant (mean |SHAP| ~ %.0e), so the\n  correlation has no value rather than a value of zero\n",
              paste(rownames(Md)[is.na(Md[, 1])], collapse = ", "),
              min(long$share[long$feature %in% rownames(Md)[is.na(Md[, 1])]])))
if (length(unstable))
  cat(sprintf("direction does not reproduce across seeds (annotated): %d\n",
              sum(rownames(Md) %in% unstable)))

# The features where the two channels disagree are the reason for keeping both. A large share
# with a weak direction means the model leans on the feature without a consistent sign.
disc <- tibble(feature = rownames(Md), share = M[rownames(Md), 1], cor = Md[, 1]) %>%
  filter(share > median(share), !is.na(cor), abs(cor) < 0.6) %>% arrange(desc(share))
if (nrow(disc)) {
  cat("\nabove-median share but weak direction - the model uses these without a consistent sign:\n")
  print(as.data.frame(disc %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)
}
if (ncol(Md) > 4) {
  cat("\nNOTE: with", ncol(Md), "columns the per-cell numbers are suppressed and this panel is\n")
  cat("direction only. The magnitude-only panel above is the one to read for size.\n")
}
```

# 2. Grouped by family

```{r fig_family, fig.width = 6.5, fig.height = 5}
# Exact: SHAP is additive, so a family's share is the sum of its members' shares.
fam_long <- long %>% group_by(experiment, family) %>%
  summarise(share = sum(share), n = n(), .groups = "drop")
Mf <- ord_rows(mat_of(fam_long, "family"))
annof <- if (!is.null(fam_floor)) {
  d <- fam_floor %>% filter(family %in% rownames(Mf)) %>%
    mutate(`noise floor` = floor_mdd) %>% dplyr::select(family, `noise floor`) %>%
    column_to_rownames("family")
  d[rownames(Mf), , drop = FALSE]
} else NULL
pheat(Mf, sprintf("SHAP share (%%) per family, %d experiment%s",
                  ncol(Mf), if (ncol(Mf) == 1) "" else "s"),
      annof, if (!is.null(annof)) list(`noise floor` = colorRampPalette(c("grey93", "grey20"))(100)),
      fname = tagf("15_heatmap_family_magnitude", "pdf"))
# A family's direction is the share-weighted mean of its members' correlations. When members
# disagree this moves toward 0 and the cell goes white, which is the honest reading: the family
# has no single direction.
# na.rm is required, not optional. is_dap5_repressed has an UNDEFINED shap_value_cor - its SHAP
# is effectively constant (mean |SHAP| 3e-06), so the correlation has no value rather than a
# value of zero - and without na.rm that single member made the whole dap5 family NA, rendering
# it grey in every column. Dropping the undefined members and reporting how many were dropped is
# the honest handling: the family still has a direction, computed from the members that have one.
fam_dir <- long %>% group_by(experiment, family) %>%
  summarise(direction = if (all(is.na(direction))) NA_real_ else
                        weighted.mean(direction, share, na.rm = TRUE),
            n_undefined = sum(is.na(direction)), .groups = "drop")
nd <- fam_dir %>% filter(n_undefined > 0) %>% distinct(family, n_undefined)
if (nrow(nd)) {
  cat("\nfamilies containing a feature with no defined direction (excluded from the weighted mean):\n")
  for (i in seq_len(nrow(nd)))
    cat(sprintf("   %-20s %d of its members\n", nd$family[i], nd$n_undefined[i]))
}
Mfd <- mat_of(fam_dir, "family", "direction")[rownames(Mf), , drop = FALSE]
fd_args <- list(mat = Mfd,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  cluster_rows = FALSE, cluster_cols = ncol(Mfd) > 2, annotation_col = anno_col,
  annotation_colors = anno_col_colors,
  display_numbers = ncol(Mfd) <= 3, number_format = "%.2f", fontsize_number = 8,
  labels_col = lab_for(Mfd),
  fontsize_row = 9, fontsize_col = 9, cellwidth = 42,
  main = "Family direction (share-weighted)\nwhite = members disagree")
draw_and_save(fd_args, tagf("15_heatmap_family_direction", "pdf"), width = 7.5)
mixed <- tibble(family = rownames(Mfd), dir = Mfd[, 1]) %>% filter(abs(dir) < 0.5)
if (nrow(mixed)) {
  cat("\nfamilies with no single direction (members disagree):\n")
  print(as.data.frame(mixed %>% mutate(dir = round(dir, 3))), row.names = FALSE)
}

tab <- fam_long %>% arrange(desc(share))
if (!is.null(fam_floor)) tab <- tab %>% left_join(fam_floor, by = "family")
print(as.data.frame(tab %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)
```

# 2b. Signed share: importance with the direction it pushes

The main view (decided by Kate, 2026-10-04). Each cell is the SHAP share (%) carrying the sign of
the direction - negative means higher feature values push the predicted TE change DOWN. A sign is
only given where the push is consistent: |cor(value, SHAP)| >= `r params$dir_cutoff`, and, where
the 20-seed sweep measured it (si3d hypoxia 1hr, full model only), the sign does not flip across
seeds. Otherwise the cell is grey: the feature carries that share but has no single direction.
For a family, direction is the share-weighted mean over its members, so mixed families go grey.

```{r fig_signed_share, fig.width = max(7, 0.6 * nrow(exps) + 6), fig.height = 13}
signed_mat <- function(share_m, dir_m, unstable_rows = character(0)) {
  ok <- !is.na(dir_m) & abs(dir_m) >= params$dir_cutoff
  ok[rownames(ok) %in% unstable_rows, ] <- FALSE
  v <- share_m * sign(dir_m); v[!ok] <- NA
  list(v = v, lab = matrix(ifelse(ok, sprintf("%+.1f", v), sprintf("%.1f", share_m)),
                           nrow(v), dimnames = dimnames(v)))
}
draw_signed <- function(sm, fname, annor = NULL, annoc = NULL, fontsize_row = 9) {
  lim <- max(abs(sm$v), na.rm = TRUE)
  args <- list(mat = sm$v, color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
               breaks = seq(-lim, lim, length.out = 102), na_col = "grey85",
               cluster_rows = FALSE, cluster_cols = ncol(sm$v) > 2,
               annotation_row = annor, annotation_col = anno_col,
               annotation_colors = c(annoc, anno_col_colors),
               display_numbers = if (ncol(sm$v) <= 4) sm$lab else FALSE,
               fontsize_number = 7, number_color = "black",
               labels_col = lab_for(sm$v), fontsize_row = fontsize_row, fontsize_col = 9, cellwidth = 42)
  draw_and_save(args, fname, width = 7.5,
                scale_title = sprintf("signed SHAP share (%%)\ngrey: |direction| < %.1f", params$dir_cutoff))
}
# Families: rows in the share order of the magnitude panel.
sf <- signed_mat(Mf, Mfd)
draw_signed(sf, tagf("15_heatmap_family_signed_share", "pdf"))
# Features: seed-unstable signs are withheld where they were measured.
sfe <- signed_mat(M, Md[rownames(M), , drop = FALSE], unstable)
draw_signed(sfe, tagf("15_heatmap_features_signed_share", "pdf"),
            annor = anno_row["family"], annoc = list(family = fam_cols), fontsize_row = 8)
cat(sprintf("families with a sign: %d of %d cells | features: %d of %d cells (cutoff |cor| >= %.1f%s)\n",
            sum(!is.na(sf$v)), length(sf$v), sum(!is.na(sfe$v)), length(sfe$v), params$dir_cutoff,
            if (length(unstable)) sprintf(", %d seed-unstable features withheld", length(unstable)) else ""))
```

# 2c. Beeswarm (single experiment only)

The standard SHAP summary plot: one dot per gene, x = that feature's SHAP value for the gene,
colour = the gene's feature value (percentile within the feature). Shows what a single number
cannot - whether the push is monotonic, and how it is spread across genes. Uses the out-of-fold
SHAP for every gene when `23_` has computed it for this run, otherwise the held-out genes `11_`
explained.

```{r fig_beeswarm, fig.width = 8, fig.height = 7, eval = nrow(exps) == 1}
sfx1 <- exps$suffix[1]
oof_f <- file.path(out_dir, paste0("shap_oof_", sfx1, ".rds"))
md1 <- readRDS(file.path(out_dir, paste0("rfreg_model_data_", sfx1, ".rds")))
vals <- bind_rows(md1$train, md1$test)
if (file.exists(oof_f)) {
  o <- readRDS(oof_f); Sb <- o$S; gid <- o$genes$gene_id_clean
  src <- sprintf("out-of-fold SHAP, all %s genes", format(length(gid), big.mark = ","))
} else {
  pg <- read_csv(file.path(out_dir, paste0("shapreg_per_gene_", sfx1, ".csv")), show_col_types = FALSE)
  Sb <- as.matrix(pg[, intersect(md1$feature_cols, names(pg))]); gid <- pg$gene_id_clean
  src <- sprintf("SHAP on %d held-out genes", length(gid))
}
top <- long %>% arrange(desc(share)) %>% filter(feature %in% colnames(Sb)) %>% head(params$n_beeswarm)
bs <- map_dfr(top$feature, function(f) {
  x <- vals[[f]][match(gid, vals$gene_id_clean)]
  tibble(feature = f, shap = Sb[, f], pct = rank(x, na.last = "keep") / sum(!is.na(x)))
}) %>% mutate(feature = factor(feature, levels = rev(top$feature)))
stopifnot("beeswarm: gene IDs did not match the model data" = mean(!is.na(bs$pct)) > 0.95)
lab_y <- setNames(sprintf("%s  (%.1f%%)", top$feature, top$share), top$feature)
p_bs <- ggplot(bs, aes(shap, feature, colour = pct)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  ggbeeswarm::geom_quasirandom(size = 0.35, alpha = 0.6, groupOnX = FALSE, width = 0.4) +
  scale_colour_gradient2(low = "#0072B2", mid = "grey85", high = "#D55E00", midpoint = 0.5,
                         limits = c(0, 1), breaks = c(0, 0.5, 1), labels = c("low", "mid", "high"),
                         name = "feature value\n(percentile)") +
  scale_y_discrete(labels = lab_y) +
  labs(x = "SHAP value (contribution to predicted TE change)", y = NULL,
       title = exps$label[1],
       subtitle = sprintf("top %d features by SHAP share (in brackets); %s", nrow(top), src)) +
  theme_classic(base_size = 11)
p_bs
if (isTRUE(params$save_plots)) ggsave(here("plots", tagf("15_shap_beeswarm", "pdf")), p_bs, width = 8, height = 7)
```

# 3. Grouped by correlated cluster

```{r fig_cluster, fig.width = 6.5, fig.height = 12, eval = !is.null(corr_group)}
# The grouping read off the data rather than assigned in advance. Where a family splits a signal
# across correlated members, or where correlated members sit in DIFFERENT families, this recovers
# the total that either view alone understates.
cg_long <- long %>%
  left_join(corr_group, by = "feature") %>%
  mutate(group = coalesce(group, feature)) %>%
  group_by(experiment, group) %>%
  summarise(share = sum(share), n = n(),
            families = paste(sort(unique(family)), collapse = "+"), .groups = "drop")
Mc <- ord_rows(mat_of(cg_long, "group"))
anno_c <- cg_long %>% distinct(group, n, families) %>%
  transmute(group, size = ifelse(n > 1, "grouped", "single")) %>%
  column_to_rownames("group")
anno_c <- anno_c[rownames(Mc), , drop = FALSE]

# Direction, computed here so the magnitude panel can carry it as a side bar. A cluster's
# direction in one column is the share-weighted mean of its members' shap_value_cor, undefined
# members excluded rather than allowed to blank the row.
cg_dir <- long %>%
  left_join(corr_group, by = "feature") %>%
  mutate(group = coalesce(group, feature)) %>%
  group_by(experiment, group) %>%
  summarise(direction = if (all(is.na(direction))) NA_real_ else
                        weighted.mean(direction, share, na.rm = TRUE), .groups = "drop")
Mcd <- mat_of(cg_dir, "group", "direction")[rownames(Mc), , drop = FALSE]

# Side bar: the row's direction averaged over the columns, each experiment weighted equally.
# BINNED, not continuous: pheatmap stretches a continuous annotation ramp over the observed
# min..max, so white would not sit at 0 and a mildly positive row could render blue. Near 0 means
# either weak everywhere or opposite signs in different experiments - the direction panel below
# says which.
dir_levels <- c("down (< -0.5)", "weak down", "mixed / weak", "weak up", "up (> 0.5)")
dir_cols   <- setNames(c("#0072B2", "#9CC7E4", "grey92", "#F0B48C", "#D55E00"), dir_levels)
mean_dir <- rowMeans(Mcd, na.rm = TRUE)
anno_c_mag <- anno_c
anno_c_mag$`mean direction` <- factor(
  as.character(cut(mean_dir, c(-Inf, -0.5, -0.2, 0.2, 0.5, Inf), labels = dir_levels)),
  levels = dir_levels)
# Grouped is dark grey, not vermillion: vermillion now means "pushes up" in this figure.
size_cols <- c(grouped = "grey30", single = "grey85")
pheat(Mc, sprintf("SHAP share (%%) per correlated cluster (|rho| >= %.1f)", params$corr_threshold),
      anno_c_mag, list(size = size_cols,
                       `mean direction` = dir_cols[levels(droplevels(anno_c_mag$`mean direction`))]),
      fname = tagf("15_heatmap_cluster_magnitude", "pdf"))
cat("\nmean direction bar, rows per bin:\n"); print(table(anno_c_mag$`mean direction`))

# Direction for the clusters too, so all three groupings - features, families, correlated
# clusters - carry both a magnitude and a direction panel. Only the feature and family views had
# one; that was an omission rather than a choice.
#
# A cluster's direction is the share-weighted mean of its members' shap_value_cor, with
# undefined members excluded rather than allowed to blank the row. Because a correlated cluster
# can span families, this is the view that shows whether a single correlated axis pushes one way
# even when its members are filed under different families.
# (cg_dir / Mcd are computed above, before the magnitude panel.)
cd_args <- list(mat = Mcd,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  cluster_rows = FALSE, cluster_cols = ncol(Mcd) > 2,
  annotation_row = anno_c, annotation_col = anno_col,
  annotation_colors = c(list(size = size_cols), anno_col_colors),
  na_col = "grey88", labels_col = lab_for(Mcd),
  display_numbers = ncol(Mcd) <= 3, number_format = "%.2f", fontsize_number = 7,
  fontsize_row = 8, fontsize_col = 9,
  main = sprintf("Direction per correlated cluster (|rho| >= %.1f)\nvermillion up, blue down, white = members disagree",
                 params$corr_threshold))
draw_and_save(cd_args, tagf("15_heatmap_cluster_direction", "pdf"), width = 9)

mixed_c <- tibble(group = rownames(Mcd), dir = Mcd[, 1]) %>%
  filter(!is.na(dir), abs(dir) < 0.4)
if (nrow(mixed_c)) {
  cat("\nclusters with no single direction (members push opposite ways):\n")
  print(as.data.frame(mixed_c %>% mutate(dir = round(dir, 3))), row.names = FALSE)
}

cat("\nmulti-feature clusters, and what each recovers:\n")
print(as.data.frame(cg_long %>% filter(n > 1) %>% arrange(desc(share)) %>%
  mutate(share = round(share, 2))), row.names = FALSE)
```

# 4. Where the two groupings disagree

```{r disagreement, eval = !is.null(corr_group)}
# The reason for showing both. A correlated cluster that stays inside one family is already
# handled by the family view. A cluster that crosses families is not: in the family view its
# signal is split across two or more rows, and no amount of summing within families recovers it.
cross <- cg_long %>% filter(n > 1, str_detect(families, fixed("+")))
if (!nrow(cross)) {
  cat("Every correlated cluster sits within a single family, so the two groupings agree.\n")
} else {
  cat(sprintf("%d correlated cluster%s cross family lines:\n\n", nrow(cross),
              if (nrow(cross) == 1) "" else "s"))
  for (i in seq_len(nrow(cross))) {
    g <- cross$group[i]
    members <- long %>% filter(experiment == cross$experiment[i]) %>%
      left_join(corr_group, by = "feature") %>% filter(group == g) %>%
      arrange(desc(share))
    cat(sprintf("  %s  -> combined %.2f%% across families: %s\n",
                g, cross$share[i], cross$families[i]))
    for (j in seq_len(nrow(members)))
      cat(sprintf("       %-34s %5.2f%%   %s\n",
                  members$feature[j], members$share[j], members$family[j]))
    cat(sprintf("     in the family view this is split across %d rows, the largest being %.2f%%\n\n",
                n_distinct(members$family), max(members$share)))
  }
  cat("Reading the family heatmap alone understates these by that split.\n")
}
```

# 5. Reduced: one representative per correlated cluster

A third view, which does not replace either of the first two. Here the model is **refit** with
only one feature kept from each correlated cluster, so the retained feature absorbs the signal
its near-duplicates were splitting with it.

This has to be a refit rather than a filtered heatmap. Deleting rows after the fact would leave
the survivors holding their old, split-down SHAP values, which understates them by exactly the
amount the exercise is meant to recover.

**Decisions**

- Keep **exactly one** feature per cluster, never zero. `reduce_to_representatives()` in
  `10_regression_presets.R` derives the list from `14_`'s cluster table and asserts that
  invariant, because a feature that looks like noise in this experiment may carry signal in
  another - and if its row is missing here while present there, the columns stop lining up.
- Choose the representative by SHAP share first, then centrality within the cluster, then
  completeness. For a two-member cluster centrality is always tied, so SHAP decides.
- Do not compare shares between this view and the others as absolute numbers. The denominator
  changed: 60 features' shares sum to 100 just as 69 did. Read ordering, and read the gain.

```{r reduced, eval = nzchar(params$reduced_suffix)}
red_path <- file.path(out_dir, paste0("shapreg_feature_", params$reduced_suffix, ".csv"))
have_red <- file.exists(red_path)
if (!have_red) {
  cat("The reduced run is not on disk yet. To build it:\n\n")
  cat("  drop <- reduce_to_representatives(<featcorr_clusters csv>, <shapreg_feature csv>, 0.8)$drop\n")
  cat("  rmarkdown::render('10_rf_regression.Rmd', params = list(\n")
  cat("      exclude_features = paste(drop, collapse = ','),\n")
  cat("      feature_set_label = 'tcoreg_reduced'))\n")
  cat("  then the same params for 11_shap_regression.Rmd\n")
} else {
  red <- read_csv(red_path, show_col_types = FALSE) %>%
    transmute(feature, family, share_red = share_pct, dir_red = shap_value_cor)
  full <- long %>% filter(experiment == exps$label[1]) %>%
    dplyr::select(feature, family, share_full = share, dir_full = direction)
  cat(sprintf("full model: %d features | reduced model: %d features | removed: %d\n",
              nrow(full), nrow(red), nrow(full) - nrow(red)))

  # What each retained feature gained once its near-duplicates were removed.
  cmp <- red %>% inner_join(full %>% dplyr::select(feature, share_full, dir_full), by = "feature") %>%
    mutate(gain = share_red - share_full) %>% arrange(desc(share_red))
  cat("\ntop of the reduced model, with what each feature gained:\n")
  print(as.data.frame(cmp %>% head(15) %>%
    mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)

  # The question this variant exists to answer, per cluster.
  if (!is.null(corr_group)) {
    grp <- corr_group %>% inner_join(full, by = "feature") %>%
      group_by(group) %>%
      summarise(members = n(),
                combined_full = sum(share_full),
                kept = paste(intersect(feature, red$feature), collapse = ", "),
                kept_full = sum(share_full[feature %in% red$feature]), .groups = "drop") %>%
      mutate(kept_red = red$share_red[match(kept, red$feature)],
             recovered = 100 * (kept_red - kept_full) / (combined_full - kept_full)) %>%
      arrange(desc(combined_full))
    cat("\nper cluster: did the survivor absorb what the dropped members were carrying?\n")
    print(as.data.frame(grp %>% dplyr::select(kept, members, combined_full, kept_full,
                                              kept_red, recovered) %>%
      mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
    cat("\n  combined_full = the cluster's total share in the full model\n")
    cat("  kept_full     = the survivor's share there, before the others were removed\n")
    cat("  kept_red      = its share after the refit\n")
    cat("  recovered     = what fraction of the dropped members' share it picked up (%)\n")
    cat("\nNear 100 means the cluster really was one signal. Well under 100 means the dropped\n")
    cat("features were carrying something of their own that is now simply gone.\n")
  }
}
```

```{r fig_reduced, eval = nzchar(params$reduced_suffix) && exists("have_red") && have_red, fig.width = 7, fig.height = 12}
Mr <- red %>% transmute(feature, !!exps$label[1] := share_red) %>%
  column_to_rownames("feature") %>% as.matrix()
Mr <- Mr[order(Mr[, 1], decreasing = TRUE), , drop = FALSE]
anno_r <- data.frame(family = red$family[match(rownames(Mr), red$feature)],
                     row.names = rownames(Mr))
r_args <- list(mat = Mr,
  color = colorRampPalette(c("white", "#FDD9A0", "#E69F00", "#D55E00"))(101),
  breaks = seq(0, max(Mr), length.out = 102),
  cluster_rows = FALSE, cluster_cols = FALSE,
  annotation_row = anno_r, annotation_col = anno_col,
  annotation_colors = c(list(family = fam_cols), anno_col_colors),
  display_numbers = TRUE, number_format = "%.1f", fontsize_number = 7,
  labels_col = lab_for(Mr),
  fontsize_row = 8, fontsize_col = 9, cellwidth = 42,
  main = sprintf("Reduced model: %d features, one per correlated cluster", nrow(Mr)))
draw_and_save(r_args, tagf("15_heatmap_reduced_magnitude", "pdf"), width = 9)

Mrd <- red %>% transmute(feature, !!exps$label[1] := dir_red) %>%
  column_to_rownames("feature") %>% as.matrix()
Mrd <- Mrd[rownames(Mr), , drop = FALSE]
rd_args <- modifyList(r_args, list(mat = Mrd,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102), number_format = "%.2f",
  main = "Reduced model: direction"))
draw_and_save(rd_args, tagf("15_heatmap_reduced_direction", "pdf"), width = 9)
```

# 6. Adding the next experiment

```{r next_steps}
cat("This heatmap currently has", ncol(M), "column(s).\n\n")
cat("To add an experiment, run 10_ and 11_ for it, then add one pair to the experiments param:\n\n")
cat('  params = list(experiments = paste(\n')
cat('    "3d hypoxia 1hr=', exps$suffix[1], '",\n', sep = "")
cat('    "3d normoxia 1hr=<that run\'s suffix>",\n')
cat('    sep = ";"))\n\n')
cat("Nothing else in this notebook changes: rows become the shared feature set, columns get\n")
cat("clustered once there are more than two, and the assert above fails loudly if a new\n")
cat("experiment does not describe the same features.\n\n")
cat("Two things that must hold for a new column to be comparable:\n")
cat("  * the same 69 features, asserted above\n")
cat("  * its own noise floor, measured by 11d_/11f_ for that experiment. The floor is driven by\n")
cat("    gene count and outcome spread, so it does not transfer between experiments.\n")
```

```{r export}
source(here("code", "translational_control_overview", "10_regression_presets.R"))
if (exists("cmp")) reg_write_csv(cmp, file.path(out_dir, tagf("shapreg_heatmap_reduced_vs_full", "csv")))
if (exists("grp")) reg_write_csv(grp, file.path(out_dir, tagf("shapreg_heatmap_cluster_recovery", "csv")))
reg_write_csv(long %>% mutate(signed = share * sign(direction)),
              file.path(out_dir, tagf("shapreg_heatmap_long", "csv")))
reg_write_csv(as_tibble(M, rownames = "feature"),
              file.path(out_dir, tagf("shapreg_heatmap_features", "csv")))
reg_write_csv(as_tibble(Mf, rownames = "family"),
              file.path(out_dir, tagf("shapreg_heatmap_family", "csv")))
if (!is.null(corr_group))
  reg_write_csv(as_tibble(Mc, rownames = "group"),
                file.path(out_dir, tagf("shapreg_heatmap_cluster", "csv")))
cat("Wrote the heatmap matrices in long and wide form.\n")
```
