---
title: "SHAP heatmap: features by experiment"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
params:
  experiments: "si3d hypoxia 1hr=eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_corrected_both_clipaggregate_yraw"
  floor_preset: "transform"
  floor_arm: "raw"
  corr_threshold: 0.8
  reduced_suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_reduced_corrected_both_clipaggregate_yraw"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "15_shap_heatmap.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

It builds the SHAP heatmap this whole fork exists to produce: **features as rows, experiments as
columns**, from the random-forest regression's SHAP values.

Right now it has **one column**. Everything is written so that adding an experiment is one entry
in the `experiments` param and nothing else changes.

**Decisions**

- The cell value is **`share_pct`**, each feature's percentage of that experiment's total
  \|SHAP\|. This is the only quantity comparable across experiments without adjustment: raw SHAP
  is in log2 TE units, and the outcome's spread varies about 7-fold across contrasts, so raw
  magnitudes are not on one scale. `10d_target_scale_analysis.Rmd` measures this.
- Both views are produced, because they answer different questions. **Individual features** show
  what the model actually used. **Grouped** rows fix the fact that correlated features split a
  single signal between them; summing SHAP within a group is exact, since SHAP is additive.
- Two groupings are shown, not one: by **family** (the prior grouping, which the annotation table
  defines) and by **correlated cluster** (read off the data by `14_feature_correlation.Rmd`).
  They disagree in the one place that matters, and that disagreement is the point.
- Rows are ordered by magnitude while there is one column. With more than one, columns get
  clustered and rows ordered by their mean.
- A feature whose share sits below its own measured noise floor is marked, so a visible cell is
  not mistaken for a real one.

```{r libraries}
library(tidyverse)
library(here)
library(pheatmap)
```

```{r load}
out_dir <- here("output", "translational_control_overview")

# The experiment list. One "label=suffix" pair per experiment, semicolon-separated. Adding a
# second experiment means adding one pair here; no other line in this notebook changes.
parse_experiments <- function(s) {
  parts <- str_trim(str_split(s, ";")[[1]])
  parts <- parts[nzchar(parts)]
  tibble(label  = str_trim(str_replace(parts, "=.*$", "")),
         suffix = str_trim(str_replace(parts, "^[^=]*=", "")))
}
exps <- parse_experiments(params$experiments)
cat("experiments:", nrow(exps), "\n")
print(as.data.frame(exps), row.names = FALSE)

read_one <- function(label, suffix) {
  f <- file.path(out_dir, paste0("shapreg_feature_", suffix, ".csv"))
  if (!file.exists(f)) { warning("missing: ", f); return(NULL) }
  read_csv(f, show_col_types = FALSE) %>%
    transmute(experiment = label, feature, family,
              share = share_pct, direction = shap_value_cor)
}
long <- pmap_dfr(list(exps$label, exps$suffix), read_one)
stopifnot("no SHAP outputs found for any experiment" = nrow(long) > 0)

# Every experiment must describe the same features, or a row means different things in different
# columns. With one experiment this is trivially true; it is asserted now so that it fails loudly
# the first time a second experiment with a different feature set is added.
per_exp <- long %>% count(experiment, name = "n_features")
print(as.data.frame(per_exp), row.names = FALSE)
if (nrow(exps) > 1) {
  sets <- split(long$feature, long$experiment)
  stopifnot("experiments do not share a feature set - rows would not be comparable" =
              length(unique(map(sets, sort))) == 1)
}
stopifnot("shares do not sum to 100 within an experiment" =
            all(abs(tapply(long$share, long$experiment, sum) - 100) < 1e-6))
# Each column's own performance, so the heatmap says how good the model behind it was. A column
# from a model that barely ranks its own genes should not be read like one that ranks them well.
read_perf <- function(label, suffix) {
  f <- file.path(out_dir, paste0("rfreg_performance_", suffix, ".csv"))
  if (!file.exists(f)) return(tibble(experiment = label, auc = NA_real_, spearman = NA_real_))
  d <- read_csv(f, show_col_types = FALSE)
  g <- function(k) { v <- d$value[d$metric == k]; if (length(v)) v[1] else NA_real_ }
  tibble(experiment = label, auc = g("bridge_auc_geneset"), spearman = g("spearman_raw_y"))
}
perf <- pmap_dfr(list(exps$label, exps$suffix), read_perf)
print(as.data.frame(perf %>% mutate(across(where(is.numeric), ~ round(., 4)))), row.names = FALSE)

# Column annotation: the AUC bridge, and the held-out Spearman beside it. The colour bars give
# the at-a-glance comparison once there are several columns.
anno_col <- perf %>%
  transmute(experiment, `bridge AUC` = auc, `Spearman` = spearman) %>%
  column_to_rownames("experiment")

# pheatmap cannot print text inside an annotation bar, so the numbers go into the column label
# instead. That puts them immediately above their own colour bars, which is where they are
# wanted: the colour says how this column compares, the number says what it actually was.
col_labels <- setNames(
  sprintf("%s\nAUC %.3f\nrho %.3f", perf$experiment, perf$auc, perf$spearman),
  perf$experiment)
lab_for <- function(m) unname(col_labels[colnames(m)])

ggsave_if <- function(...) if (isTRUE(params$save_plots)) ggsave(...)
```

```{r annotations}
# Which features cannot be read directionally: their shap_value_cor sign flips across the 20
# split seeds, so the signed view must not be trusted for them.
dir_path <- file.path(out_dir, paste0("shapreg_direction_stability_",
                                      params$floor_preset, "_", params$floor_arm, ".csv"))
unstable <- if (file.exists(dir_path)) {
  d <- read_csv(dir_path, show_col_types = FALSE)
  d$feature[d$frac_pos > 0.05 & d$frac_pos < 0.95]
} else character(0)
cat("features with an unstable SHAP direction:", length(unstable), "\n")
if (length(unstable)) cat("  ", paste(unstable, collapse = ", "), "\n")

# The measured per-family floor: how much a family's share moves when nothing biological changes.
floor_path <- file.path(out_dir, paste0("shapreg_noise_floor_",
                                        params$floor_preset, "_", params$floor_arm, ".csv"))
fam_floor <- if (file.exists(floor_path))
  read_csv(floor_path, show_col_types = FALSE) %>%
    dplyr::select(family, floor_sd = sd, floor_mdd = min_detectable_diff) else NULL
cat("family noise floor available:", !is.null(fam_floor), "\n")

# The data-driven grouping from 14_. Features not in a multi-feature cluster are their own group.
clus_path <- file.path(out_dir, paste0("featcorr_clusters_", exps$suffix[1], ".csv"))
corr_group <- if (file.exists(clus_path)) {
  cl <- read_csv(clus_path, show_col_types = FALSE) %>%
    filter(abs(threshold - params$corr_threshold) < 1e-9)
  cl %>% group_by(cluster) %>%
    summarise(group = paste0(first(leader), " +", n() - 1), features = list(feature),
              .groups = "drop") %>%
    unnest(features) %>% dplyr::select(feature = features, group)
} else NULL
cat("correlated-cluster grouping available:", !is.null(corr_group), "\n")
```

# 1. Individual features

```{r fig_features, fig.width = 7, fig.height = 13}
mat_of <- function(d, row, val = "share") {
  d %>% dplyr::select(all_of(c(row, "experiment", val))) %>%
    pivot_wider(names_from = experiment, values_from = all_of(val)) %>%
    column_to_rownames(row) %>% as.matrix()
}
ord_rows <- function(m) m[order(rowMeans(m, na.rm = TRUE), decreasing = TRUE), , drop = FALSE]

M <- ord_rows(mat_of(long, "feature"))
fam_of <- long %>% distinct(feature, family) %>% deframe()

# "reproducible" means the feature's direction came out with the same sign in all 20 split-seed
# runs. "flips" means it did not, so the colour in the direction panel is not reproducible for
# that row and should be read as magnitude only. 17 of 69 features flip, carrying 6.3% of the
# total SHAP between them - every large feature is consistent.
anno_row <- data.frame(
  family              = fam_of[rownames(M)],
  `direction repeats` = ifelse(rownames(M) %in% unstable, "flips", "reproducible"),
  row.names = rownames(M), check.names = FALSE)
fam_cols <- setNames(
  colorRampPalette(c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                     "#0072B2", "#D55E00", "#CC79A7", "grey50"))(n_distinct(anno_row$family)),
  sort(unique(anno_row$family)))

pheat <- function(m, main, annor, annoc = NULL, signed = FALSE, fname = NULL, ...) {
  if (signed) {
    lim <- max(abs(m), na.rm = TRUE)
    col <- colorRampPalette(c("#0072B2", "white", "#D55E00"))(101)
    brk <- seq(-lim, lim, length.out = 102)
  } else {
    col <- colorRampPalette(c("white", "#FDD9A0", "#E69F00", "#D55E00"))(101)
    brk <- seq(0, max(m, na.rm = TRUE), length.out = 102)
  }
  args <- list(mat = m, color = col, breaks = brk,
               cluster_rows = FALSE, cluster_cols = ncol(m) > 2,
               annotation_row = annor, annotation_col = anno_col, annotation_colors = annoc,
               display_numbers = ncol(m) <= 3, number_format = "%.1f", fontsize_number = 7,
               labels_col = lab_for(m),
               fontsize_row = 8, fontsize_col = 9, cellwidth = 42, main = main, ...)
  do.call(pheatmap, args)
  if (!is.null(fname) && isTRUE(params$save_plots)) {
    pdf(file.path(here("plots"), fname), width = 7, height = max(4, 0.17 * nrow(m) + 2))
    do.call(pheatmap, args); dev.off()
  }
}

pheat(M, sprintf("SHAP share (%%) per feature, %d experiment%s",
                 ncol(M), if (ncol(M) == 1) "" else "s"),
      anno_row, list(family = fam_cols,
                     `direction repeats` = c(reproducible = "grey85", flips = "#CC79A7")),
      fname = "15_heatmap_features_magnitude.pdf")
cat(sprintf("\n%d features | top: %s (%.2f%%) | bottom: %s (%.2f%%)\n",
            nrow(M), rownames(M)[1], M[1, 1], rownames(M)[nrow(M)], M[nrow(M), 1]))
cat(sprintf("the top 10 features carry %.1f%% of the total\n", sum(M[1:10, 1])))
```

```{r fig_features_combined, fig.width = 7.5, fig.height = 13}
# One panel carrying both quantities, because they are different measurements and neither
# substitutes for the other:
#
#   share_pct      0 to ~12    how much of the total |SHAP| the feature holds
#   shap_value_cor -1 to +1    which way it pushes, from the correlation of the feature's value
#                              with its own SHAP value
#
# They are independent. karner2026_mda231_log2_ct holds a large share (3.55%) with a weak
# direction (-0.43); g4mer_frac_above_utr5 holds a tiny share (0.64%) with a strong direction
# (+0.81). Multiplying them into a single number would render both as a mid-sized, mid-confident
# feature and lose the distinction, so they get separate visual channels instead.
#
#   COLOUR  = direction. Vermillion means a higher feature value pushes the predicted TE change
#             up (more eIF3d-promoted), dark blue means down, white is 0. Symmetric across the
#             full [-1, 1] range, so colour is comparable between panels and experiments.
#   NUMBER  = the share, printed in the cell.
#
# Rows stay ordered by share, so vertical position is magnitude and colour is direction.
Md <- mat_of(long, "feature", "direction")[rownames(M), , drop = FALSE]

comb_args <- list(
  mat = Md,
  color  = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  cluster_rows = FALSE, cluster_cols = ncol(Md) > 2,
  annotation_row = anno_row, annotation_col = anno_col,
  annotation_colors = list(family = fam_cols,
                           `direction repeats` = c(reproducible = "grey85", flips = "#CC79A7")),
  # The printed number is the SHARE, not the value being coloured.
  display_numbers = round(M[rownames(Md), , drop = FALSE], 1),
  fontsize_number = 7, number_color = "black",
  labels_col = lab_for(Md),
  # A feature whose SHAP values are effectively constant has no defined direction - the
  # correlation is undefined, not zero. Those cells are grey, and are counted separately below
  # rather than being folded into the up or down totals.
  na_col = "grey88",
  fontsize_row = 8, fontsize_col = 9, cellwidth = 46,
  main = "Colour = direction (vermillion up, blue down, grey undefined)\nNumber = SHAP share (%), rows ordered by it")
do.call(pheatmap, comb_args)
if (isTRUE(params$save_plots)) {
  pdf(here("plots", "15_heatmap_features_combined.pdf"),
      width = 7.5, height = max(4, 0.17 * nrow(Md) + 2))
  do.call(pheatmap, comb_args); dev.off()
}

n_na <- sum(is.na(Md[, 1]))
cat(sprintf("\ndirection: %d push UP, %d push DOWN, %d undefined\n",
            sum(Md[, 1] > 0, na.rm = TRUE), sum(Md[, 1] < 0, na.rm = TRUE), n_na))
cat(sprintf("strongly directional (|cor| > 0.8): %d of %d | ambiguous (|cor| < 0.5): %d\n",
            sum(abs(Md[, 1]) > 0.8, na.rm = TRUE), nrow(Md),
            sum(abs(Md[, 1]) < 0.5, na.rm = TRUE)))
if (n_na)
  cat(sprintf("  undefined: %s - SHAP is effectively constant (mean |SHAP| ~ %.0e), so the\n  correlation has no value rather than a value of zero\n",
              paste(rownames(Md)[is.na(Md[, 1])], collapse = ", "),
              min(long$share[long$feature %in% rownames(Md)[is.na(Md[, 1])]])))
if (length(unstable))
  cat(sprintf("direction does not reproduce across seeds (annotated): %d\n",
              sum(rownames(Md) %in% unstable)))

# The features where the two channels disagree are the reason for keeping both. A large share
# with a weak direction means the model leans on the feature without a consistent sign.
disc <- tibble(feature = rownames(Md), share = M[rownames(Md), 1], cor = Md[, 1]) %>%
  filter(share > median(share), !is.na(cor), abs(cor) < 0.6) %>% arrange(desc(share))
if (nrow(disc)) {
  cat("\nabove-median share but weak direction - the model uses these without a consistent sign:\n")
  print(as.data.frame(disc %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)
}
cat("\nWith many columns the printed numbers get cramped; the magnitude-only panel above\n")
cat("remains the one to read at that point.\n")
```

# 2. Grouped by family

```{r fig_family, fig.width = 6.5, fig.height = 5}
# Exact: SHAP is additive, so a family's share is the sum of its members' shares.
fam_long <- long %>% group_by(experiment, family) %>%
  summarise(share = sum(share), n = n(), .groups = "drop")
Mf <- ord_rows(mat_of(fam_long, "family"))
annof <- if (!is.null(fam_floor)) {
  d <- fam_floor %>% filter(family %in% rownames(Mf)) %>%
    mutate(`noise floor` = floor_mdd) %>% dplyr::select(family, `noise floor`) %>%
    column_to_rownames("family")
  d[rownames(Mf), , drop = FALSE]
} else NULL
pheat(Mf, sprintf("SHAP share (%%) per family, %d experiment%s",
                  ncol(Mf), if (ncol(Mf) == 1) "" else "s"),
      annof, NULL, fname = "15_heatmap_family_magnitude.pdf")
# A family's direction is the share-weighted mean of its members' correlations. When members
# disagree this moves toward 0 and the cell goes white, which is the honest reading: the family
# has no single direction.
fam_dir <- long %>% group_by(experiment, family) %>%
  summarise(direction = weighted.mean(direction, share), .groups = "drop")
Mfd <- mat_of(fam_dir, "family", "direction")[rownames(Mf), , drop = FALSE]
fd_args <- list(mat = Mfd,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102),
  cluster_rows = FALSE, cluster_cols = ncol(Mfd) > 2, annotation_col = anno_col,
  display_numbers = ncol(Mfd) <= 3, number_format = "%.2f", fontsize_number = 8,
  labels_col = lab_for(Mfd),
  fontsize_row = 9, fontsize_col = 9, cellwidth = 42,
  main = "Family direction (share-weighted)\nwhite = members disagree")
do.call(pheatmap, fd_args)
if (isTRUE(params$save_plots)) {
  pdf(here("plots", "15_heatmap_family_direction.pdf"), width = 6.5, height = 5)
  do.call(pheatmap, fd_args); dev.off()
}
mixed <- tibble(family = rownames(Mfd), dir = Mfd[, 1]) %>% filter(abs(dir) < 0.5)
if (nrow(mixed)) {
  cat("\nfamilies with no single direction (members disagree):\n")
  print(as.data.frame(mixed %>% mutate(dir = round(dir, 3))), row.names = FALSE)
}

tab <- fam_long %>% arrange(desc(share))
if (!is.null(fam_floor)) tab <- tab %>% left_join(fam_floor, by = "family")
print(as.data.frame(tab %>% mutate(across(where(is.numeric), ~ round(., 3)))), row.names = FALSE)
```

# 3. Grouped by correlated cluster

```{r fig_cluster, fig.width = 6.5, fig.height = 12, eval = !is.null(corr_group)}
# The grouping read off the data rather than assigned in advance. Where a family splits a signal
# across correlated members, or where correlated members sit in DIFFERENT families, this recovers
# the total that either view alone understates.
cg_long <- long %>%
  left_join(corr_group, by = "feature") %>%
  mutate(group = coalesce(group, feature)) %>%
  group_by(experiment, group) %>%
  summarise(share = sum(share), n = n(),
            families = paste(sort(unique(family)), collapse = "+"), .groups = "drop")
Mc <- ord_rows(mat_of(cg_long, "group"))
anno_c <- cg_long %>% distinct(group, n, families) %>%
  transmute(group, size = ifelse(n > 1, "grouped", "single")) %>%
  column_to_rownames("group")
anno_c <- anno_c[rownames(Mc), , drop = FALSE]
pheat(Mc, sprintf("SHAP share (%%) per correlated cluster (|rho| >= %.1f)", params$corr_threshold),
      anno_c, list(size = c(grouped = "#D55E00", single = "grey85")),
      fname = "15_heatmap_cluster_magnitude.pdf")

cat("\nmulti-feature clusters, and what each recovers:\n")
print(as.data.frame(cg_long %>% filter(n > 1) %>% arrange(desc(share)) %>%
  mutate(share = round(share, 2))), row.names = FALSE)
```

# 4. Where the two groupings disagree

```{r disagreement, eval = !is.null(corr_group)}
# The reason for showing both. A correlated cluster that stays inside one family is already
# handled by the family view. A cluster that crosses families is not: in the family view its
# signal is split across two or more rows, and no amount of summing within families recovers it.
cross <- cg_long %>% filter(n > 1, str_detect(families, fixed("+")))
if (!nrow(cross)) {
  cat("Every correlated cluster sits within a single family, so the two groupings agree.\n")
} else {
  cat(sprintf("%d correlated cluster%s cross family lines:\n\n", nrow(cross),
              if (nrow(cross) == 1) "" else "s"))
  for (i in seq_len(nrow(cross))) {
    g <- cross$group[i]
    members <- long %>% filter(experiment == cross$experiment[i]) %>%
      left_join(corr_group, by = "feature") %>% filter(group == g) %>%
      arrange(desc(share))
    cat(sprintf("  %s  -> combined %.2f%% across families: %s\n",
                g, cross$share[i], cross$families[i]))
    for (j in seq_len(nrow(members)))
      cat(sprintf("       %-34s %5.2f%%   %s\n",
                  members$feature[j], members$share[j], members$family[j]))
    cat(sprintf("     in the family view this is split across %d rows, the largest being %.2f%%\n\n",
                n_distinct(members$family), max(members$share)))
  }
  cat("Reading the family heatmap alone understates these by that split.\n")
}
```

# 5. Reduced: one representative per correlated cluster

A third view, which does not replace either of the first two. Here the model is **refit** with
only one feature kept from each correlated cluster, so the retained feature absorbs the signal
its near-duplicates were splitting with it.

This has to be a refit rather than a filtered heatmap. Deleting rows after the fact would leave
the survivors holding their old, split-down SHAP values, which understates them by exactly the
amount the exercise is meant to recover.

**Decisions**

- Keep **exactly one** feature per cluster, never zero. `reduce_to_representatives()` in
  `10_regression_presets.R` derives the list from `14_`'s cluster table and asserts that
  invariant, because a feature that looks like noise in this experiment may carry signal in
  another - and if its row is missing here while present there, the columns stop lining up.
- Choose the representative by SHAP share first, then centrality within the cluster, then
  completeness. For a two-member cluster centrality is always tied, so SHAP decides.
- Do not compare shares between this view and the others as absolute numbers. The denominator
  changed: 60 features' shares sum to 100 just as 69 did. Read ordering, and read the gain.

```{r reduced, eval = nzchar(params$reduced_suffix)}
red_path <- file.path(out_dir, paste0("shapreg_feature_", params$reduced_suffix, ".csv"))
have_red <- file.exists(red_path)
if (!have_red) {
  cat("The reduced run is not on disk yet. To build it:\n\n")
  cat("  drop <- reduce_to_representatives(<featcorr_clusters csv>, <shapreg_feature csv>, 0.8)$drop\n")
  cat("  rmarkdown::render('10_rf_regression.Rmd', params = list(\n")
  cat("      exclude_features = paste(drop, collapse = ','),\n")
  cat("      feature_set_label = 'tcoreg_reduced'))\n")
  cat("  then the same params for 11_shap_regression.Rmd\n")
} else {
  red <- read_csv(red_path, show_col_types = FALSE) %>%
    transmute(feature, family, share_red = share_pct, dir_red = shap_value_cor)
  full <- long %>% filter(experiment == exps$label[1]) %>%
    dplyr::select(feature, family, share_full = share, dir_full = direction)
  cat(sprintf("full model: %d features | reduced model: %d features | removed: %d\n",
              nrow(full), nrow(red), nrow(full) - nrow(red)))

  # What each retained feature gained once its near-duplicates were removed.
  cmp <- red %>% inner_join(full %>% dplyr::select(feature, share_full, dir_full), by = "feature") %>%
    mutate(gain = share_red - share_full) %>% arrange(desc(share_red))
  cat("\ntop of the reduced model, with what each feature gained:\n")
  print(as.data.frame(cmp %>% head(15) %>%
    mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)

  # The question this variant exists to answer, per cluster.
  if (!is.null(corr_group)) {
    grp <- corr_group %>% inner_join(full, by = "feature") %>%
      group_by(group) %>%
      summarise(members = n(),
                combined_full = sum(share_full),
                kept = paste(intersect(feature, red$feature), collapse = ", "),
                kept_full = sum(share_full[feature %in% red$feature]), .groups = "drop") %>%
      mutate(kept_red = red$share_red[match(kept, red$feature)],
             recovered = 100 * (kept_red - kept_full) / (combined_full - kept_full)) %>%
      arrange(desc(combined_full))
    cat("\nper cluster: did the survivor absorb what the dropped members were carrying?\n")
    print(as.data.frame(grp %>% dplyr::select(kept, members, combined_full, kept_full,
                                              kept_red, recovered) %>%
      mutate(across(where(is.numeric), ~ round(., 2)))), row.names = FALSE)
    cat("\n  combined_full = the cluster's total share in the full model\n")
    cat("  kept_full     = the survivor's share there, before the others were removed\n")
    cat("  kept_red      = its share after the refit\n")
    cat("  recovered     = what fraction of the dropped members' share it picked up (%)\n")
    cat("\nNear 100 means the cluster really was one signal. Well under 100 means the dropped\n")
    cat("features were carrying something of their own that is now simply gone.\n")
  }
}
```

```{r fig_reduced, eval = nzchar(params$reduced_suffix) && exists("have_red") && have_red, fig.width = 7, fig.height = 12}
Mr <- red %>% transmute(feature, !!exps$label[1] := share_red) %>%
  column_to_rownames("feature") %>% as.matrix()
Mr <- Mr[order(Mr[, 1], decreasing = TRUE), , drop = FALSE]
anno_r <- data.frame(family = red$family[match(rownames(Mr), red$feature)],
                     row.names = rownames(Mr))
r_args <- list(mat = Mr,
  color = colorRampPalette(c("white", "#FDD9A0", "#E69F00", "#D55E00"))(101),
  breaks = seq(0, max(Mr), length.out = 102),
  cluster_rows = FALSE, cluster_cols = FALSE,
  annotation_row = anno_r, annotation_col = anno_col,
  annotation_colors = list(family = fam_cols),
  display_numbers = TRUE, number_format = "%.1f", fontsize_number = 7,
  labels_col = lab_for(Mr),
  fontsize_row = 8, fontsize_col = 9, cellwidth = 42,
  main = sprintf("Reduced model: %d features, one per correlated cluster", nrow(Mr)))
do.call(pheatmap, r_args)
if (isTRUE(params$save_plots)) {
  pdf(here("plots", "15_heatmap_reduced_magnitude.pdf"), width = 7,
      height = max(4, 0.17 * nrow(Mr) + 2))
  do.call(pheatmap, r_args); dev.off()
}

Mrd <- red %>% transmute(feature, !!exps$label[1] := dir_red) %>%
  column_to_rownames("feature") %>% as.matrix()
Mrd <- Mrd[rownames(Mr), , drop = FALSE]
rd_args <- modifyList(r_args, list(mat = Mrd,
  color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
  breaks = seq(-1, 1, length.out = 102), number_format = "%.2f",
  main = "Reduced model: direction"))
do.call(pheatmap, rd_args)
if (isTRUE(params$save_plots)) {
  pdf(here("plots", "15_heatmap_reduced_direction.pdf"), width = 7,
      height = max(4, 0.17 * nrow(Mrd) + 2))
  do.call(pheatmap, rd_args); dev.off()
}
```

# 6. Adding the next experiment

```{r next_steps}
cat("This heatmap currently has", ncol(M), "column(s).\n\n")
cat("To add an experiment, run 10_ and 11_ for it, then add one pair to the experiments param:\n\n")
cat('  params = list(experiments = paste(\n')
cat('    "si3d hypoxia 1hr=', exps$suffix[1], '",\n', sep = "")
cat('    "si3d normoxia 1hr=<that run\'s suffix>",\n')
cat('    sep = ";"))\n\n')
cat("Nothing else in this notebook changes: rows become the shared feature set, columns get\n")
cat("clustered once there are more than two, and the assert above fails loudly if a new\n")
cat("experiment does not describe the same features.\n\n")
cat("Two things that must hold for a new column to be comparable:\n")
cat("  * the same 69 features, asserted above\n")
cat("  * its own noise floor, measured by 11d_/11f_ for that experiment. The floor is driven by\n")
cat("    gene count and outcome spread, so it does not transfer between experiments.\n")
```

```{r export}
source(here("code", "translational_control_overview", "10_regression_presets.R"))
if (exists("cmp")) reg_write_csv(cmp, file.path(out_dir, "shapreg_heatmap_reduced_vs_full.csv"))
if (exists("grp")) reg_write_csv(grp, file.path(out_dir, "shapreg_heatmap_cluster_recovery.csv"))
reg_write_csv(long %>% mutate(signed = share * sign(direction)),
              file.path(out_dir, "shapreg_heatmap_long.csv"))
reg_write_csv(as_tibble(M, rownames = "feature"),
              file.path(out_dir, "shapreg_heatmap_features.csv"))
reg_write_csv(as_tibble(Mf, rownames = "family"),
              file.path(out_dir, "shapreg_heatmap_family.csv"))
if (!is.null(corr_group))
  reg_write_csv(as_tibble(Mc, rownames = "group"),
                file.path(out_dir, "shapreg_heatmap_cluster.csv"))
cat("Wrote the heatmap matrices in long and wide form.\n")
```
