---
title: "Top-100 feature means heatmap: features by experiment, no model"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
params:
  experiments: "MDA-MB-231 3d hypoxia 1hr=eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3d hypoxia 4hr=eif33d_promotes_hypoxia_4hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3d hypoxia 24hr=eif33d_promotes_hypoxia_24hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3d normoxia 1+4hr=eif33d_promotes_normoxia_1and4hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 1hr=eif33e_promotes_hypoxia_1hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 4hr=eif33e_promotes_hypoxia_4hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3e hypoxia 24hr=eif33e_promotes_hypoxia_24hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MDA-MB-231 3e normoxia 1+4hr=eif33e_promotes_normoxia_1and4hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;MCF7-SIX1 3d hypoxia 1hr=eif33d_promotes_mcf7six1_hypoxia_1hr_lfc0.5_tcoreg_mcf7tx_intrinsic_corrected_both_clipaggregate_yraw;MCF7-SIX1 3d normoxia 1hr=eif33d_promotes_mcf7six1_normoxia_1hr_lfc0.5_tcoreg_mcf7tx_intrinsic_corrected_both_clipaggregate_yraw;MCF7-SIX1 3e hypoxia 1hr=eif33e_promotes_mcf7six1_hypoxia_1hr_lfc0.5_tcoreg_mcf7tx_intrinsic_corrected_both_clipaggregate_yraw;MCF7-SIX1 3e normoxia 1hr=eif33e_promotes_mcf7six1_normoxia_1hr_lfc0.5_tcoreg_mcf7tx_intrinsic_corrected_both_clipaggregate_yraw;HeLa 3d no stress (Herrmannova)=eif33d_promotes_hela_normoxia_steadystate_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;HeLa 3e no stress (Herrmannova)=eif33e_promotes_hela_normoxia_steadystate_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw;HEK293T DHX29 IP enrichment (Hia)=eifdhx29_promotes_hek293t_dhx29ip_ribo_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw"
  # Which feature matrix each column's model used, matched on a suffix substring (first match
  # wins, "" = default). The regression model data does not record the matrix, so this must be
  # stated; the gene-set and feature-set checks below catch a wrong entry.
  matrix_map: "_mcf7tx_=feature_matrix_tco_mcf7six1.rds;=feature_matrix_tco.rds"
  top_n: 100
  padj_max: 0.05
  save_plots: TRUE
  out_tag: ""
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "17_top100_feature_means_heatmap.Rmd"))
  params <- lapply(.fm$params,
                   function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

A third, model-free view of the same features and experiments as `15_shap_heatmap.Rmd` (SHAP) and
`16_feature_correlation_heatmap.Rmd` (Spearman rho). Here each cell is the **mean value of a feature
across the 100 transcripts each experiment most promotes**: what the most strongly regulated
mRNAs look like, in the feature's own units.

**Decisions**

- **Same columns, genes and matrices** as the other two heatmaps (same `experiments` param; genes
  from each regression's saved model data; raw, non-imputed feature values).
- **Top 100 = highest LFC, significant genes first.** LFC is oriented so positive = the factor
  promotes. Genes with padj < 0.05 are taken first, by LFC; if a column has fewer than 100 of
  them, it is filled to exactly 100 with the next-highest-LFC genes regardless of padj. Every
  column is therefore a mean over 100 genes (a 5-gene mean at 3d hypoxia 24hr otherwise took the
  extreme colour in about a third of rows). The column annotation `n_significant` shows how many
  of each column's 100 are significant, so filled columns stay visible.
- **The printed number is the raw mean; the colour is that mean scaled within the feature
  across columns** (z-score of the row). Features have different units, so one scale across
  rows is impossible. Colour therefore answers "in which experiments are the top genes highest
  or lowest for this feature" - **not** whether they differ from a typical gene. Row scaling also
  stretches every row to the full colour range, so a row whose raw means barely differ looks as
  vivid as one that differs a lot; read the printed numbers before the colour.
- **No significance testing** here.
- Intrinsic features, individual rows grouped by family, as in the correlation heatmap.

```{r libraries}
library(tidyverse)
library(here)
library(pheatmap)
library(grid)
```

# Load


```{r load}
out_dir <- here("output", "translational_control_overview")
parse_pairs <- function(s) {
  parts <- str_trim(str_split(s, ";")[[1]]); parts <- parts[nzchar(parts)]
  tibble(key = str_trim(str_replace(parts, "=.*$", "")),
         val = str_trim(str_replace(parts, "^[^=]*=", "")))
}
exps <- parse_pairs(params$experiments) %>% dplyr::rename(label = key, suffix = val)
mmap <- parse_pairs(params$matrix_map)
matrix_for <- function(suffix) {
  hit <- mmap %>% filter(key == "" | str_detect(suffix, fixed(key)))
  hit$val[1]
}
exps <- exps %>% mutate(matrix_rds = map_chr(suffix, matrix_for))
stopifnot("an experiment label is duplicated" = !any(duplicated(exps$label)),
          "a column has no matrix" = !any(is.na(exps$matrix_rds)))
print(as.data.frame(exps %>% dplyr::select(label, matrix_rds)), row.names = FALSE)

annotation <- read_csv(file.path(out_dir, "feature_annotation.csv"), show_col_types = FALSE)
matrices <- set_names(unique(exps$matrix_rds)) %>% map(~ readRDS(file.path(out_dir, .x)))
walk(matrices, ~ stopifnot("a feature matrix is not one transcript per gene" =
                              !any(duplicated(.x$gene_id_clean))))
```

# Top-N means

```{r top_means}
one_column <- function(label, suffix, matrix_rds) {
  md <- readRDS(file.path(out_dir, paste0("rfreg_model_data_", suffix, ".rds")))
  genes <- bind_rows(md$train %>% dplyr::select(gene_id_clean, y, y_padj),
                     md$test  %>% dplyr::select(gene_id_clean, y, y_padj))
  stopifnot("a gene appears twice in a column's model data" = !any(duplicated(genes$gene_id_clean)),
            "this column has no padj to rank within" = any(!is.na(genes$y_padj)))
  N <- as.integer(params$top_n)
  sig  <- genes %>% filter(!is.na(y_padj), y_padj < params$padj_max, y > 0) %>% arrange(desc(y))
  top  <- sig %>% slice_head(n = N)
  if (nrow(top) < N)
    top <- bind_rows(top, genes %>% filter(!gene_id_clean %in% top$gene_id_clean) %>%
                       arrange(desc(y)) %>% slice_head(n = N - nrow(top)))
  n_sig <- min(nrow(sig), N)
  stopifnot("could not fill the top set to N genes" = nrow(top) == N)
  fm <- matrices[[matrix_rds]]
  stopifnot("a modelled feature is missing from this column's matrix" = all(md$feature_cols %in% names(fm)),
            "a top gene is missing from this column's matrix" = all(top$gene_id_clean %in% fm$gene_id_clean))
  x <- fm[match(top$gene_id_clean, fm$gene_id_clean), md$feature_cols, drop = FALSE]
  map_dfr(md$feature_cols, function(f) {
    v <- as.numeric(unlist(x[[f]]))
    tibble(experiment = label, feature = f, mean = mean(v, na.rm = TRUE), n = sum(is.finite(v)))
  }) %>% mutate(n_top = nrow(top), n_significant = n_sig, lfc_min_in_top = min(top$y),
                n_genes = nrow(genes))
}
long <- pmap_dfr(list(exps$label, exps$suffix, exps$matrix_rds), one_column)
sets <- split(long$feature, long$experiment)
stopifnot("columns do not share one feature set" = all(map_lgl(sets, ~ setequal(.x, sets[[1]]))))
long <- long %>% left_join(annotation %>% dplyr::select(feature, family), by = "feature")
stopifnot("a feature has no family" = !any(is.na(long$family)),
          "a mean is not finite" = all(is.finite(long$mean)))

long %>% distinct(experiment, n_genes, n_top, n_significant, lfc_min_in_top) %>%
  arrange(match(experiment, exps$label)) %>% as.data.frame() %>% print(row.names = FALSE)
short <- long %>% distinct(experiment, n_significant) %>% filter(n_significant < params$top_n)
if (nrow(short))
  cat("FILLED beyond padj <", params$padj_max, ":", paste0(short$experiment, " (", short$n_significant,
      " significant)", collapse = "; "), "\n")
```

# Heatmap

Rows grouped by family (same family order as the correlation heatmap uses: by largest spread
here), columns clustered on the row-scaled profile. Printed numbers are raw top-100 means.

```{r heatmap, fig.width = 13, fig.height = 13}
M <- long %>% dplyr::select(feature, experiment, mean) %>%
  pivot_wider(names_from = experiment, values_from = mean) %>% column_to_rownames("feature") %>%
  as.matrix()
M <- M[, exps$label, drop = FALSE]
Z <- t(scale(t(M)))                                  # row z-score across columns
Z[!is.finite(Z)] <- 0                                # a constant row has no spread to show

fam <- long %>% distinct(feature, family)
spread <- tibble(feature = rownames(Z), s = apply(abs(Z), 1, max))
fam_order <- fam %>% left_join(spread, by = "feature") %>% group_by(family) %>%
  summarise(m = max(s), .groups = "drop") %>% arrange(desc(m)) %>% pull(family)
row_order <- fam %>% arrange(match(family, fam_order), feature) %>% pull(feature)
Z <- Z[row_order, , drop = FALSE]; M <- M[row_order, , drop = FALSE]
gaps <- which(diff(match(fam$family[match(row_order, fam$feature)], fam_order)) != 0)

# Print with precision matched to each feature's scale, so 0-1 fractions and log2 lengths both read.
fmt_row <- function(v) { d <- max(0, 2 - floor(log10(max(abs(v), 1e-9)))); formatC(v, format = "f", digits = min(d, 3)) }
lab <- t(apply(M, 1, fmt_row)); dimnames(lab) <- dimnames(M)

meta <- exps %>% transmute(
  experiment = label,
  `cell line` = case_when(str_detect(suffix, "_hek293t_") ~ "HEK293T",
                          str_detect(suffix, "_hela_")    ~ "HeLa",
                          str_detect(suffix, "mcf7six1")  ~ "MCF7-SIX1",
                          TRUE                            ~ "MDA-MB-231"),
  `initiation factor` = case_when(str_detect(suffix, "^eifdhx29_") ~ "DHX29",
                                  str_detect(suffix, "^eif33e_")   ~ "eIF3e", TRUE ~ "eIF3d"),
  condition = case_when(str_detect(suffix, "_hela_|_dhx29ip_") ~ "no stress",
                        str_detect(suffix, "_hypoxia_")  ~ "hypoxia",
                        str_detect(suffix, "_normoxia_") ~ "normoxia",
                        TRUE                             ~ "other")) %>%
  left_join(long %>% distinct(experiment, n_significant), by = "experiment") %>%
  column_to_rownames("experiment")
anno_colors <- list(
  `cell line` = c(`MDA-MB-231` = "#F0E442", `MCF7-SIX1` = "#661100", HeLa = "#999933",
                  HEK293T = "#44AA99")[unique(meta$`cell line`)],
  `initiation factor` = c(eIF3d = "#E69F00", eIF3e = "#56B4E9",
                          DHX29 = "#882255")[unique(meta$`initiation factor`)],
  condition = c(hypoxia = "#CC79A7", normoxia = "#009E73", `no stress` = "#009E73",
                other = "grey60")[unique(meta$condition)],
  n_significant = colorRampPalette(c("grey93", "grey20"))(100))
row_anno <- data.frame(family = fam$family[match(rownames(Z), fam$feature)], row.names = rownames(Z))

lim <- max(abs(Z)); brk <- seq(-lim, lim, length.out = 101)
hm <- pheatmap(Z, color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(100), breaks = brk,
               cluster_rows = FALSE, cluster_cols = if (ncol(Z) > 2) hclust(dist(t(Z))) else FALSE,
               gaps_row = gaps, display_numbers = lab, number_color = "grey15", fontsize_number = 5.5,
               annotation_col = meta, annotation_row = row_anno, annotation_colors = anno_colors,
               main = paste0("Top ", params$top_n, " promoted transcripts: mean feature value\n",
                             "number = raw mean; colour = row z-score across experiments"),
               fontsize_row = 7, fontsize_col = 8, angle_col = 90, silent = TRUE)
grid.newpage(); grid.draw(hm$gtable)
tag <- if (nzchar(params$out_tag)) paste0("_", params$out_tag) else ""
if (isTRUE(params$save_plots))
  ggsave(here("plots", paste0("17_top", params$top_n, "_feature_means_heatmap", tag, ".pdf")),
         hm$gtable, width = 13, height = 13)
```

```{r write}
write_csv(long, file.path(out_dir, paste0("top", params$top_n, "_feature_means", tag, ".csv")))
cat("wrote", paste0("top", params$top_n, "_feature_means", tag, ".csv"), "-", nrow(long), "rows\n")
```

```{r session}
sessionInfo()
```
