---
title: "PCA of experiments on their feature-association profiles"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  corr_threshold: 0.8
  feature_list_csv: "rfreg_importance_eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw.csv"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "22_experiment_pca.Rmd"))
  params <- lapply(.fm$params, function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

Places every experiment (heatmap column) in one space, to see which look alike across all
intrinsic features at once. Model-free (Kate, 2026-10-01, option 1):

- **Profile.** For each experiment, the Spearman rho between each of the 57 intrinsic features
  and the experiment's LFC, over **all** genes in its universe (its own matrix, genes with an
  outcome). All genes, not a top-N set, so the profile does not depend on a few extreme genes.
- **Correlated features count once.** The GC, length and structure blocks are near-duplicates;
  left alone, PC1 would be "GC/length" counted several times. Features are grouped with the
  rule `14_feature_correlation.Rmd` uses (Spearman across MDA-MB-231 transcripts, distance
  1 - |rho|, average linkage, cut at |rho| >= `corr_threshold`), recomputed here on the rebuilt
  matrix. Within a group, members are sign-aligned to the group's first member before their rho
  values are averaged, so anti-correlated members do not cancel.
- **PCA** on the experiments x groups matrix, centred, not scaled (every entry is a rho on the
  same scale, so scaling would inflate groups that barely vary).

Eighteen points: this describes structure, it does not test it.

```{r libraries}
suppressPackageStartupMessages({ library(tidyverse); library(here); library(ggrepel) })
O <- here("output", "translational_control_overview")
kd_cols <- c(eIF3d = "#E69F00", eIF3e = "#56B4E9", eIF4E = "#CC6677", DHX29 = "#882255", DAP5 = "#DDCC77")
line_shapes <- c(`MDA-MB-231` = 16, `MCF7-SIX1` = 17, HeLa = 15, HEK293T = 18)
```

# 1. Experiments

Same columns, outcome files and matrices as the heatmaps.

```{r columns}
tc <- function(g, c_, t) paste0("translation_categories_si", g, "_vs_sictrl_", c_, "_", t, ".csv")
cols <- tribble(
  ~label, ~line, ~set, ~factor, ~outcome, ~lfc,
  "MDA-MB-231 3d hypoxia 1hr",    "MDA-MB-231", "231", "eIF3d", tc("3d", "hypoxia", "1hr"), "te_lfc",
  "MDA-MB-231 3d hypoxia 4hr",    "MDA-MB-231", "231", "eIF3d", tc("3d", "hypoxia", "4hr"), "te_lfc",
  "MDA-MB-231 3d hypoxia 24hr",   "MDA-MB-231", "231", "eIF3d", tc("3d", "hypoxia", "24hr"), "te_lfc",
  "MDA-MB-231 3d normoxia 1+4hr", "MDA-MB-231", "231", "eIF3d", tc("3d", "normoxia", "1and4hr"), "te_lfc",
  "MDA-MB-231 3e hypoxia 1hr",    "MDA-MB-231", "231", "eIF3e", tc("3e", "hypoxia", "1hr"), "te_lfc",
  "MDA-MB-231 3e hypoxia 4hr",    "MDA-MB-231", "231", "eIF3e", tc("3e", "hypoxia", "4hr"), "te_lfc",
  "MDA-MB-231 3e hypoxia 24hr",   "MDA-MB-231", "231", "eIF3e", tc("3e", "hypoxia", "24hr"), "te_lfc",
  "MDA-MB-231 3e normoxia 1+4hr", "MDA-MB-231", "231", "eIF3e", "te_si3e_vs_sictrl_normoxia_1and4hr.csv", "te_lfc",
  "MCF7-SIX1 3d hypoxia 1hr",     "MCF7-SIX1", "mcf7", "eIF3d", tc("3d", "mcf7six1_hypoxia", "1hr"), "te_lfc",
  "MCF7-SIX1 3d normoxia 1hr",    "MCF7-SIX1", "mcf7", "eIF3d", tc("3d", "mcf7six1_normoxia", "1hr"), "te_lfc",
  "MCF7-SIX1 3e hypoxia 1hr",     "MCF7-SIX1", "mcf7", "eIF3e", tc("3e", "mcf7six1_hypoxia", "1hr"), "te_lfc",
  "MCF7-SIX1 3e normoxia 1hr",    "MCF7-SIX1", "mcf7", "eIF3e", tc("3e", "mcf7six1_normoxia", "1hr"), "te_lfc",
  "HeLa 3d no stress (Herrmannova)", "HeLa", "hela", "eIF3d", tc("3d", "hela_normoxia", "steadystate"), "te_lfc",
  "HeLa 3e no stress (Herrmannova)", "HeLa", "hela", "eIF3e", tc("3e", "hela_normoxia", "steadystate"), "te_lfc",
  "HeLa 4E TE level, E.V. (Teleman)",   "HeLa", "hela", "eIF4E", "teleman_absolute_te_ev.csv", "te_level",
  "HeLa 4E TE level, 4E-BP1 (Teleman)", "HeLa", "hela", "eIF4E", "teleman_absolute_te_bp.csv", "te_level",
  "HEK293T DHX29 IP enrichment (Hia)",  "HEK293T", "dhx29", "DHX29", "dhx29_ip_hek293t_enrichment.csv", "enrich_lfc",
  "HEK293T DAP5 no stress (Weber)*",    "HEK293T", "mane", "DAP5", "translation_categories_sidap5_vs_sictrl_hek293t_normoxia_steadystate.csv", "te_lfc")
matrix_of <- c(`231` = "feature_matrix_tco.rds", mcf7 = "feature_matrix_tco_mcf7six1.rds",
               hela = "feature_matrix_tco_hela.rds", dhx29 = "feature_matrix_tco_dhx29.rds",
               mane = "feature_matrix_tco_mane.rds")
stopifnot("an outcome file is missing" = all(file.exists(here("output", cols$outcome))))
feats <- read_csv(file.path(O, params$feature_list_csv), show_col_types = FALSE)$feature
mats  <- map(matrix_of, ~ readRDS(file.path(O, .x)))
stopifnot("an intrinsic feature is missing from a matrix" = all(map_lgl(mats, ~ all(feats %in% names(.x)))))
cat("features:", length(feats), "| experiments:", nrow(cols), "\n")
```

# 2. Feature-association profiles

```{r profiles}
rho_long <- pmap_dfr(cols, function(label, line, set, factor, outcome, lfc) {
  oc <- read_csv(here("output", outcome), show_col_types = FALSE) %>%
    transmute(gene_id_clean = sub("[.].*", "", gene_id), y = .data[[lfc]]) %>% filter(!is.na(y))
  m <- mats[[set]]; j <- m[match(oc$gene_id_clean, m$gene_id_clean), feats]; keep <- !is.na(match(oc$gene_id_clean, m$gene_id_clean))
  y <- oc$y[keep]; j <- j[keep, ]
  map_dfr(feats, function(f) { v <- as.numeric(unlist(j[[f]])); k <- is.finite(v) & is.finite(y)
    tibble(label = label, feature = f, rho = cor(v[k], y[k], method = "spearman"), n = sum(k)) })
})
rho_long %>% group_by(label) %>% summarise(genes = max(n), min_n = min(n), .groups = "drop") %>%
  arrange(match(label, cols$label)) %>% knitr::kable()
stopifnot("a correlation is not finite" = all(is.finite(rho_long$rho)))
```

# 3. Correlated-feature groups

```{r groups}
X <- mats[["231"]][, feats] %>% mutate(across(everything(), ~ as.numeric(unlist(.x))))
C <- cor(X, method = "spearman", use = "pairwise.complete.obs")
hc <- hclust(as.dist(1 - abs(C)), method = "average")
grp <- cutree(hc, h = 1 - params$corr_threshold)
groups <- tibble(feature = names(grp), group_id = grp) %>% group_by(group_id) %>%
  mutate(lead = feature[1],
         sign = sign(C[cbind(feature, lead)]),        # orient each member to the group's first
         size = n(),
         group = if (n() == 1) feature[1] else paste0(feature[1], " (+", n() - 1, ")")) %>% ungroup()
stopifnot("a member could not be sign-aligned" = all(groups$sign %in% c(-1, 1)))
cat(length(feats), "features ->", n_distinct(groups$group), "groups at |rho| >=", params$corr_threshold, "\n")
groups %>% filter(size > 1) %>% group_by(group) %>%
  summarise(members = paste0(ifelse(sign < 0, "-", ""), feature, collapse = ", "), .groups = "drop") %>%
  knitr::kable()
```

# 4. PCA

```{r pca}
prof <- rho_long %>% inner_join(groups, by = "feature") %>%
  group_by(label, group) %>% summarise(rho = mean(sign * rho), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = rho) %>% column_to_rownames("label") %>% as.matrix()
prof <- prof[cols$label, ]
pc <- prcomp(prof, center = TRUE, scale. = FALSE)
ve <- 100 * pc$sdev^2 / sum(pc$sdev^2)
cat(sprintf("variance explained: %s\n", paste(sprintf("PC%d %.1f%%", 1:5, ve[1:5]), collapse = " | ")))
scores <- as_tibble(pc$x[, 1:3], rownames = "label") %>% left_join(cols, by = "label")
```

```{r fig_pca, fig.width = 10, fig.height = 7.5}
p1 <- ggplot(scores, aes(PC1, PC2, colour = factor, shape = line)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 4) +
  geom_text_repel(aes(label = label), size = 3, colour = "grey20", max.overlaps = Inf, seed = 9) +
  scale_colour_manual(values = kd_cols, name = "factor") +
  scale_shape_manual(values = line_shapes, name = "cell line") +
  labs(x = sprintf("PC1 (%.1f%%)", ve[1]), y = sprintf("PC2 (%.1f%%)", ve[2]),
       title = "Experiments by their feature-LFC Spearman profiles",
       subtitle = sprintf("%d intrinsic features in %d correlation groups | n = %d experiments, all genes per experiment",
                          length(feats), ncol(prof), nrow(prof)),
       caption = paste("* features computed on the MANE Select transcript; all other experiments use",
                       "each dataset's own most abundant or representative transcript.")) +
  theme_classic(base_size = 12) +
  theme(plot.caption = element_text(hjust = 0, size = 8), plot.caption.position = "plot")
p1
if (isTRUE(params$save_plots)) ggsave(here("plots", "22_experiment_pca.pdf"), p1, width = 10, height = 7.5)
```

## What the axes mean

Loadings: the feature groups that place experiments along each axis. A group with a positive
loading on PC1 has a higher rho (stronger positive association with the LFC) in experiments on
the right of the plot.

```{r fig_loadings, fig.width = 11, fig.height = 7}
ld <- as_tibble(pc$rotation[, 1:2], rownames = "group") %>%
  left_join(groups %>% distinct(group, size), by = "group") %>%
  pivot_longer(c(PC1, PC2), names_to = "pc", values_to = "loading") %>%
  group_by(pc) %>% slice_max(abs(loading), n = 12) %>% ungroup() %>%
  mutate(group = tidytext_reorder <- paste0(group, "___", pc)) %>%
  mutate(group = fct_reorder(group, loading))
p2 <- ggplot(ld, aes(loading, group, fill = loading > 0)) +
  geom_col(width = 0.7) +
  scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
  scale_fill_manual(values = c(`TRUE` = "#D55E00", `FALSE` = "#0072B2"), guide = "none") +
  facet_wrap(~ pc, scales = "free_y") +
  labs(x = "loading", y = NULL, title = "Top 12 feature groups per axis",
       subtitle = "a group's name is its first member; (+k) = k more correlated features averaged in") +
  theme_classic(base_size = 12)
p2
if (isTRUE(params$save_plots)) ggsave(here("plots", "22_experiment_pca_loadings.pdf"), p2, width = 11, height = 7)
```

```{r fig_scree, fig.width = 5, fig.height = 3.5}
tibble(PC = factor(seq_along(ve)), var = ve) %>% slice_head(n = 10) %>%
  ggplot(aes(PC, var)) + geom_col(fill = "grey50") +
  labs(y = "% variance explained", title = sprintf("Scree, n = %d experiments", nrow(prof))) +
  theme_classic(base_size = 12)
```

```{r session}
sessionInfo()
```
