---
title: "`r paste0('Clustering genes on their SHAP vectors: ', params$label, ', intrinsic model')`"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  suffix: "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw"
  matrix_rds: "feature_matrix_tco.rds"
  n_folds: 5
  ks: "2,3,4,5,6,7,8"
  n_perm: 50
  sil_n: 2000
  save_plots: TRUE
  test_genes: 0
  label: "MDA-MB-231 si3d hypoxia 1hr"
  pos_geneset: "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "23_shap_gene_clustering.Rmd"))
  params <- lapply(.fm$params, function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

Every gene in the MDA-MB-231 si3d hypoxia 1hr regression gets a vector of SHAP values - how much
each intrinsic feature pushed that gene's predicted TE change up or down - and genes are then
clustered on those vectors to find groups the model explains the same way.

**Decisions (Kate, 2026-10-01)**

- **Every gene, each explained by a model that never saw it.** SHAP on a gene the forest was
  trained on overstates the fit. So the 10,172 genes are split into 5 folds (stratified on the
  LFC); for each fold a forest is fitted on the other four - exactly as `10_rf_regression.Rmd`
  fits, with medians from the training folds only - and the held-out fold is explained. Each
  gene's SHAP vector therefore comes from a forest that did not train on it.
- **Intrinsic model** (57 features): external features are other studies' measurements and would
  group genes partly by which datasets cover them.
- **Both value scales**: raw SHAP (how strongly, and which way, each feature moved the gene), and
  normalised within gene (each value divided by the gene's total |SHAP|, so a cluster reflects
  *which* features drove a gene, not how strongly it was called).
- **Both levels**: the 11 feature families (SHAP is additive, so a family's value is the exact sum
  of its members') and the 57 individual features. That makes four clusterings.
- **Choosing k** uses the null that `03_subgroup_null.R` established for the classifier: permute
  each column independently across genes (destroying which features co-occur in a gene, keeping
  every marginal), standardise the observed silhouette **within k** against that null, then take
  the maximum over k. Taking the raw best-over-k compares different quantities and is known to
  return a false "no structure". Adapted to 10,172 genes: k-means is fitted on every gene; the
  silhouette is scored on a fixed random `sil_n` of them, since the full distance matrix (10,172
  squared) does not fit in memory here.

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse); library(here); library(ranger); library(treeshap); library(caret)
  library(cluster); library(pheatmap); library(grid)
})
O <- here("output", "translational_control_overview")
cache_dir <- file.path(O, "shap_oof_cache"); dir.create(cache_dir, showWarnings = FALSE)
ks <- as.integer(strsplit(params$ks, ",")[[1]])
```

# 1. The model's genes and features

Taken from the saved regression run, so the gene universe, LFC and feature set are exactly the
ones that model used. Feature values come unimputed from the matrix; each fold imputes its own.

```{r load}
md <- readRDS(file.path(O, paste0("rfreg_model_data_", params$suffix, ".rds")))
fc <- md$feature_cols
genes <- bind_rows(md$train, md$test) %>% dplyr::select(gene_id_clean, symbol, y = y_t, y_padj)
fm <- readRDS(file.path(O, params$matrix_rds))
X <- fm[match(genes$gene_id_clean, fm$gene_id_clean), fc] %>% mutate(across(everything(), ~ as.numeric(unlist(.x))))
stopifnot("model genes missing from the matrix" = !anyNA(match(genes$gene_id_clean, fm$gene_id_clean)),
          "target is not untransformed - LFC and y_t would differ" = md$target_transform == "raw")
ann <- read_csv(file.path(O, "feature_annotation.csv"), show_col_types = FALSE)
fam <- setNames(ann$family[match(fc, ann$feature)], fc)
stopifnot("a feature has no family" = !anyNA(fam))
# Test mode: a random subset with its own cache and no saved outputs, so the whole notebook can be
# exercised in minutes before the real (hours-long) SHAP run.
if (params$test_genes > 0) {
  set.seed(9); keep <- sort(sample(nrow(genes), params$test_genes))
  genes <- genes[keep, ]; X <- X[keep, ]
  cache_dir <- file.path(O, "shap_oof_cache_test"); dir.create(cache_dir, showWarnings = FALSE)
  cat("TEST MODE:", nrow(genes), "genes; nothing is written to plots/ or the outputs\n")
}
SAVE <- isTRUE(params$save_plots) && params$test_genes == 0
# Plot files carry the contrast (e.g. "hypoxia_4hr") so per-timepoint runs do not overwrite each other.
ptag <- sub("_lfc.*$", "", sub("^eif33d_promotes_", "", params$suffix))
cat(sprintf("genes %d | features %d in %d families\n", nrow(genes), length(fc), n_distinct(fam)))
```

## The refit reproduces the saved model

Fitting on the saved run's own split must give the saved held-out predictions exactly, or the
fold forests below are not the same model.

```{r reproduce}
fit_forest <- function(Xtr, ytr) {
  set.seed(md$forest_seed)
  ranger(x = Xtr, y = ytr, num.trees = 1000, min.node.size = 5, splitrule = "variance",
         importance = "impurity", seed = md$forest_seed)
}
rf0 <- fit_forest(md$train[, fc], md$train$y_t)
saved <- read_csv(file.path(O, paste0("rfreg_predictions_", params$suffix, ".csv")), show_col_types = FALSE)
dev0 <- max(abs(predict(rf0, md$test[, fc])$predictions - saved$pred[match(md$test$gene_id_clean, saved$gene_id_clean)]))
cat("refit vs saved held-out predictions: max |dev|", signif(dev0, 3), "\n")
stopifnot("the refit does not reproduce the saved model" = dev0 < 1e-9)
rm(rf0); invisible(gc())
```

# 2. SHAP for every gene, out of fold

Cached per fold in `shap_oof_cache/`, so an interrupted render resumes. Each fold's SHAP plus
its baseline must add up to that fold forest's prediction for every gene.

```{r shap_oof}
set.seed(9)
folds <- createFolds(genes$y, k = params$n_folds, list = TRUE)
impute <- function(df, med) { for (cc in fc) df[[cc]][is.na(df[[cc]])] <- med[[cc]]; df }
res <- map(seq_along(folds), function(f) {
  cf <- file.path(cache_dir, sprintf("fold%d_%s.rds", f, params$suffix))
  te <- folds[[f]]
  if (file.exists(cf)) { r <- readRDS(cf); stopifnot("cached fold does not match" = identical(r$idx, te)); return(r) }
  tr  <- setdiff(seq_len(nrow(genes)), te)
  med <- X[tr, ] %>% summarise(across(everything(), ~ median(.x, na.rm = TRUE)))
  Xtr <- impute(X[tr, ], med); Xte <- impute(X[te, ], med)
  rf  <- fit_forest(Xtr, genes$y[tr])
  u   <- ranger.unify(rf, as.data.frame(Xtr))
  t0  <- Sys.time()
  sh  <- as.matrix(treeshap(u, as.data.frame(Xte), verbose = FALSE)$shaps[, fc])
  pr  <- predict(rf, Xte)$predictions
  leftover <- pr - rowSums(sh)
  stopifnot("SHAP does not add up to the fold prediction" = diff(range(leftover)) < 1e-6)
  r <- list(idx = te, shap = sh, pred = pr, baseline = mean(leftover),
            minutes = as.numeric(difftime(Sys.time(), t0, units = "mins")))
  saveRDS(r, cf); cat(sprintf("fold %d: %d genes, %.0f min\n", f, length(te), r$minutes)); r
})
S <- matrix(NA_real_, nrow(genes), length(fc), dimnames = list(genes$gene_id_clean, fc))
genes$pred_oof <- NA_real_
for (r in res) { S[r$idx, ] <- r$shap; genes$pred_oof[r$idx] <- r$pred }
stopifnot("a gene has no out-of-fold SHAP" = !anyNA(S))
cat(sprintf("out-of-fold prediction vs LFC: Spearman %.3f (n = %d) | baselines %s\n",
            cor(genes$pred_oof, genes$y, method = "spearman"), nrow(genes),
            paste(round(map_dbl(res, "baseline"), 4), collapse = ", ")))
```

# 3. Four profiles

```{r profiles}
F <- sapply(sort(unique(fam)), function(g) rowSums(S[, fam == g, drop = FALSE]))
norm_rows <- function(M) M / pmax(rowSums(abs(M)), 1e-12)
profiles <- list(`families, raw` = F, `families, normalised` = norm_rows(F),
                 `features, raw` = S, `features, normalised` = norm_rows(S))
stopifnot("family sums do not reproduce the feature SHAP total" = max(abs(rowSums(F) - rowSums(S))) < 1e-9)
map_dfr(profiles, ~ tibble(columns = ncol(.x), sd_mean = mean(apply(.x, 2, sd))), .id = "profile") %>% knitr::kable(digits = 4)
```

# 4. How many clusters: the null band

```{r null}
set.seed(9); sil_idx <- sort(sample(nrow(genes), min(params$sil_n, nrow(genes))))
sil_of <- function(M, k, seed = 9) {
  set.seed(seed); km <- kmeans(M, centers = k, nstart = 10, iter.max = 50)
  if (length(unique(km$cluster[sil_idx])) < 2) return(NA_real_)
  mean(silhouette(km$cluster[sil_idx], dist(M[sil_idx, ]))[, 3])
}
null_test <- function(M) {
  obs <- vapply(ks, function(k) sil_of(M, k), numeric(1))
  nm <- t(vapply(seq_len(params$n_perm), function(i) {
    set.seed(1000 + i); P <- apply(M, 2, sample)            # drawn once per replicate
    vapply(ks, function(k) sil_of(P, k), numeric(1)) }, numeric(length(ks))))
  mu <- colMeans(nm); sdv <- apply(nm, 2, sd)
  stopifnot("null has zero variance" = all(sdv > 0))
  z <- (obs - mu) / sdv
  z_null <- vapply(seq_len(params$n_perm), function(i) { m2 <- nm[-i, , drop = FALSE]
    max((nm[i, ] - colMeans(m2)) / apply(m2, 2, sd)) }, numeric(1))
  list(per_k = tibble(k = ks, observed = obs, null_median = apply(nm, 2, median), null_min = apply(nm, 2, min),
                      null_max = apply(nm, 2, max), z = z),
       p_overall = (sum(z_null >= max(z)) + 1) / (params$n_perm + 1),
       best_k_z = ks[which.max(z)], best_k_sil = ks[which.max(obs)])
}
# Cached: the null is the slow part after SHAP, and it depends only on the profiles and params.
null_cache <- file.path(cache_dir, sprintf("nulls_%s_perm%d_sil%d_k%s.rds", params$suffix, params$n_perm,
                                           params$sil_n, gsub(",", "-", params$ks)))
nulls <- if (SAVE && file.exists(null_cache)) readRDS(null_cache) else map(profiles, null_test)
if (SAVE && !file.exists(null_cache)) saveRDS(nulls, null_cache)
summary_k <- imap_dfr(nulls, ~ tibble(profile = .y, best_k_sil = .x$best_k_sil, best_k_z = .x$best_k_z,
                                      max_z = max(.x$per_k$z), p_overall = .x$p_overall))
knitr::kable(summary_k, digits = 3)
```

```{r fig_null, fig.width = 11, fig.height = 7}
band <- imap_dfr(nulls, ~ .x$per_k %>% mutate(profile = .y))
p_band <- ggplot(band, aes(k)) +
  geom_ribbon(aes(ymin = null_min, ymax = null_max), fill = "grey85") +
  geom_line(aes(y = null_median), colour = "grey50", linetype = 2) +
  geom_line(aes(y = observed), colour = "#E69F00", linewidth = 0.8) + geom_point(aes(y = observed), colour = "#E69F00") +
  facet_wrap(~ profile, scales = "free_y") +
  labs(y = sprintf("mean silhouette (%d-gene sample)", length(sil_idx)),
       title = "Observed silhouette against the column-permuted null, per k",
       subtitle = sprintf("grey band = null min-max over %d permutations; dashed = null median; n = %d genes",
                          params$n_perm, nrow(genes))) +
  theme_classic(base_size = 12)
p_band
if (SAVE) ggsave(here("plots", paste0("23_shap_gene_clusters_", ptag, "_null_band.pdf")), p_band, width = 11, height = 7)
```

# 5. Heatmaps

Genes are columns, ordered by cluster and, within a cluster, by LFC. k is the number of groups.
Each clustering is drawn twice: at the silhouette peak (`best_k_sil`, the cleanest split - k = 2
here) and at the k the null test supports most strongly (`best_k_z`, where the groups are most
clearly real rather than an artefact of clustering), so finer subgroups are visible. Raw
SHAP colours are capped at the 99th percentile of |SHAP| so a few extreme genes do not wash out the
rest; the cap is printed. Written as PNG: 10,172 columns as vector cells make a PDF too large to open.

```{r sets}
# pos_geneset = "" applies the project's rule (LFC > 0.5, padj < 0.05) to this model's own outcome,
# as 10_ does for the regressed timepoints (normoxia 1+4hr), which have no saved set.
gs_pos <- if (nzchar(params$pos_geneset))
  read_csv(here("output", "genesets", params$pos_geneset), show_col_types = FALSE)$ensembl_gene else
  genes$gene_id_clean[genes$y > 0.5 & !is.na(genes$y_padj) & genes$y_padj < 0.05]
gs_neg <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"), show_col_types = FALSE) %>%
  mutate(g = sub("[.].*", "", gene_id)) %>% pull(g)
genes$set <- case_when(genes$gene_id_clean %in% sub("[.].*", "", gs_pos) ~ "eIF3d promoted",
                       genes$gene_id_clean %in% gs_neg ~ "negative control", TRUE ~ "other")
table(genes$set)
```

```{r heatmaps, fig.width = 14, fig.height = 9, results = "asis"}
fam_cols <- setNames(colorRampPalette(c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2",
                                        "#D55E00", "#CC79A7", "grey50"))(n_distinct(fam)), sort(unique(fam)))
assignments <- list()
for (nm in names(profiles)) for (k in unique(c(nulls[[nm]]$best_k_sil, nulls[[nm]]$best_k_z))) {
  M <- profiles[[nm]]
  k_why <- if (k == nulls[[nm]]$best_k_sil) "silhouette peak" else "best supported vs null"
  set.seed(9); cl <- kmeans(M, centers = k, nstart = 25, iter.max = 50)$cluster
  # name clusters by size, largest = 1, so labels are stable across reruns
  cl <- match(cl, as.integer(names(sort(table(cl), decreasing = TRUE))))
  assignments[[paste(nm, k)]] <- tibble(gene_id_clean = genes$gene_id_clean, symbol = genes$symbol,
                                        profile = nm, k = k, cluster = cl)
  ord <- order(cl, -genes$y)
  rows <- if (startsWith(nm, "features")) fc[order(fam, fc)] else colnames(M)
  H <- t(M[ord, rows, drop = FALSE])
  cap <- if (grepl("raw", nm)) quantile(abs(M), 0.99) else max(abs(M))
  H[H > cap] <- cap; H[H < -cap] <- -cap
  acol <- data.frame(cluster = factor(cl[ord]), `LFC (sictrl/si3d)` = genes$y[ord],
                     `predicted (out of fold)` = genes$pred_oof[ord], `gene set` = genes$set[ord],
                     check.names = FALSE, row.names = colnames(H))
  arow <- if (startsWith(nm, "features")) data.frame(family = fam[rows], row.names = rows) else NULL
  acols <- list(`gene set` = c(`eIF3d promoted` = "#E69F00", `negative control` = "black", other = "grey90"),
                cluster = setNames(c("#0072B2", "#CC79A7", "#009E73", "#F0E442", "#D55E00", "#56B4E9", "grey40", "#882255")[seq_len(k)], seq_len(k)),
                `LFC (sictrl/si3d)` = colorRampPalette(c("#0072B2", "white", "#D55E00"))(50),
                `predicted (out of fold)` = colorRampPalette(c("#0072B2", "white", "#D55E00"))(50))
  if (!is.null(arow)) acols$family <- fam_cols
  hm <- pheatmap(H, cluster_rows = FALSE, cluster_cols = FALSE, show_colnames = FALSE,
                 color = colorRampPalette(c("#0072B2", "white", "#D55E00"))(101),
                 breaks = seq(-cap, cap, length.out = 102), annotation_col = acol, annotation_row = arow,
                 annotation_colors = acols, gaps_col = cumsum(table(cl))[-k], fontsize_row = if (ncol(M) > 20) 6 else 9,
                 main = sprintf("SHAP, %s | k = %d (%s) | n = %d genes | colour capped at %.3g",
                                nm, k, k_why, nrow(genes), cap), silent = TRUE)
  cat("\n\n## ", nm, ", k = ", k, "\n\n", sep = ""); grid.newpage(); grid.draw(hm$gtable)
  if (SAVE) {
    f <- here("plots", paste0("23_shap_gene_clusters_", ptag, "_", gsub("[ ,]+", "_", nm), "_k", k, ".png"))
    png(f, width = 14, height = if (ncol(M) > 20) 11 else 6, units = "in", res = 300)
    grid.newpage(); grid.draw(hm$gtable); dev.off()
  }
}
```

# 6. What each cluster is

Per clustering and cluster: size, LFC and out-of-fold prediction, how many eIF3d-promoted genes
it holds, the three families with the largest mean |SHAP| (sign = mean direction), and the
cluster's median sequence values - so a group can be read as "GC-poor, long 5'UTR" rather than
only as "GC SHAP positive".

```{r cluster_summary}
all_assign <- bind_rows(assignments) %>% left_join(genes %>% dplyr::select(gene_id_clean, y, pred_oof, set), by = "gene_id_clean")
fam_long <- as_tibble(F, rownames = "gene_id_clean") %>% pivot_longer(-gene_id_clean, names_to = "family", values_to = "shap")
top_fam <- all_assign %>% inner_join(fam_long, by = "gene_id_clean", relationship = "many-to-many") %>%
  group_by(profile, k, cluster, family) %>% summarise(m = mean(shap), a = mean(abs(shap)), .groups = "drop") %>%
  group_by(profile, k, cluster) %>% slice_max(a, n = 3) %>%
  summarise(top_families = paste0(family, ifelse(m > 0, " (+)", " (-)"), collapse = ", "), .groups = "drop")
seqv <- fm %>% transmute(gene_id_clean, cds_gc, gc3 = gc3_internal, cds_nt = 2^log2_cds_length - 1,
                         utr5_nt = 2^log2_utr5_length - 1, utr5_access = tco_struct_accessibility_utr5_mean)
summ <- all_assign %>% left_join(seqv, by = "gene_id_clean") %>% group_by(profile, k, cluster) %>%
  summarise(n = n(), median_lfc = median(y), median_pred = median(pred_oof),
            eIF3d_promoted = sum(set == "eIF3d promoted"), neg_controls = sum(set == "negative control"),
            cds_gc = median(cds_gc), gc3 = median(gc3, na.rm = TRUE), cds_nt = median(cds_nt),
            utr5_nt = median(utr5_nt), utr5_access = median(utr5_access, na.rm = TRUE), .groups = "drop") %>%
  left_join(top_fam, by = c("profile", "k", "cluster"))
knitr::kable(summ, digits = c(0, 0, 0, 0, 3, 3, 0, 0, 1, 1, 0, 0, 3, 0))
if (SAVE) {
write_csv(all_assign, file.path(O, paste0("shap_gene_clusters_", params$suffix, ".csv")))
write_csv(summ, file.path(O, paste0("shap_gene_cluster_summary_", params$suffix, ".csv")))
saveRDS(list(S = S, genes = genes, folds = folds), file.path(O, paste0("shap_oof_", params$suffix, ".rds")))
}
```

```{r session}
sessionInfo()
```
