---
title: "How k is chosen for the SHAP gene clustering (si3d, MDA-MB-231)"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  stem: "eif33d_promotes_%s_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw"
  timepoints: "hypoxia_1hr,hypoxia_4hr,hypoxia_24hr,normoxia_1and4hr"
  null_tag: "perm50_sil2000_k2-3-4-5-6-7-8"
  ks: "2,3,4,5,6,7,8"
  n_perm_elbow: 20
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(here::here("code", "translational_control_overview", "23b_choosing_k.Rmd"))
  params <- lapply(.fm$params, function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# The procedure

`23_shap_gene_clustering.Rmd` groups genes by their out-of-fold SHAP profiles with k-means. **k is
the number of groups.** Nothing in k-means says what k should be, so every k from 2 to 8 is tried:

1. **Cluster** the genes into k groups (k-means, 10 random starts, `set.seed(9)`).
2. **Score separation with the silhouette.** For each gene: (mean distance to the nearest other
   group - mean distance to its own group) / the larger of the two. It runs from -1 (probably in the
   wrong group) to +1 (clearly in the right one); the score is its mean, computed on a fixed random
   2,000 genes because the full 10,172 x 10,172 distance matrix does not fit in memory here.
3. **Null.** Shuffle each column (family or feature) independently across genes - every column
   keeps its distribution, but which families go together in a gene is destroyed - then cluster
   and score the same way. 50 shuffles.
4. **Compare within the same k:** z = (real silhouette - mean shuffled silhouette) / sd of the
   shuffled silhouettes.

Two k's come out, and the heatmaps are drawn at both:

- **Silhouette peak** (`best_k_sil`): highest raw silhouette. Almost always k = 2, because the
  silhouette falls as k grows even on random data. The cleanest coarse split, not necessarily the
  most real structure.
- **Best supported vs the null** (`best_k_z`): highest z, i.e. where real data beats shuffled data
  by the most. This is the k reported for each timepoint.

**Is there any structure at all?** The best z is compared with the best z each shuffled data set
reaches (which pays for having tried seven k's): p = (shuffles at least as extreme + 1) / 51, so the
smallest possible p is 0.02.

**Why not only the raw silhouette?** Comparing the best raw silhouette over k against the shuffled
best compares two different quantities - real and shuffled data peak at different k - and on the
classifier it returned a confident, wrong "no structure". Standardising within each k first, then
maximising, fixes that (`03_subgroup_null.R`; CLAUDE.md, SHAP subgroups).

**What this notebook adds:** an elbow plot (how much gene-to-gene variation the groups capture as k
rises), drawn for real **and shuffled** data, because on a continuum like this one the real curve
alone bends gently and always rises. The gap between the two curves is the part of each k's gain
that is real structure.

```{r libraries}
suppressPackageStartupMessages({ library(tidyverse); library(here) })
O <- here("output", "translational_control_overview")
ks <- as.integer(strsplit(params$ks, ",")[[1]])
tps <- strsplit(params$timepoints, ",")[[1]]
tp_lab <- function(x) c(hypoxia_1hr = "hypoxia 1hr", hypoxia_4hr = "hypoxia 4hr", hypoxia_24hr = "hypoxia 24hr",
                        normoxia_1and4hr = "normoxia 1+4hr")[x]
have <- tps[file.exists(file.path(O, paste0("shap_oof_", sprintf(params$stem, tps), ".rds"))) &
            file.exists(file.path(O, "shap_oof_cache", paste0("nulls_", sprintf(params$stem, tps), "_", params$null_tag, ".rds")))]
if (length(setdiff(tps, have))) cat("not yet available, skipped:", paste(tp_lab(setdiff(tps, have)), collapse = ", "), "\n")
tps <- have
ann <- read_csv(file.path(O, "feature_annotation.csv"), show_col_types = FALSE)
```

# 1. The null test, per k

From `23_`'s cached null results - the same numbers that chose the k in each heatmap.

```{r null_tables}
nulls <- map(setNames(tps, tps), ~ readRDS(file.path(O, "shap_oof_cache", paste0("nulls_", sprintf(params$stem, .x), "_", params$null_tag, ".rds"))))
per_k <- imap_dfr(nulls, function(nl, tp) imap_dfr(nl, ~ .x$per_k %>% mutate(profile = .y, timepoint = tp_lab(tp),
                                                        best_k_z = .x$best_k_z, best_k_sil = .x$best_k_sil, p_overall = .x$p_overall)))
per_k %>% distinct(timepoint, profile, best_k_sil, best_k_z, p_overall) %>% knitr::kable(digits = 3)
```

```{r fig_z, fig.width = 12, fig.height = 4.5}
p_z <- per_k %>% mutate(timepoint = factor(timepoint, levels = tp_lab(tps))) %>%
  ggplot(aes(k, z, colour = profile)) + geom_hline(yintercept = 0, colour = "grey80") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.8) +
  geom_point(data = ~ filter(.x, k == best_k_z), shape = 21, size = 4, stroke = 1.1, fill = NA) +
  facet_wrap(~ timepoint, nrow = 1) +
  scale_colour_manual(values = c(`families, normalised` = "#0072B2", `families, raw` = "#56B4E9",
                                 `features, normalised` = "#CC79A7", `features, raw` = "#D55E00")) +
  labs(y = "z: real silhouette vs 50 shuffles, same k", title = "How clearly each k beats shuffled data",
       subtitle = "circled = chosen k (highest z). Note how close several k's are: the choice is a preference, not a sharp optimum") +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
p_z
if (isTRUE(params$save_plots)) ggsave(here("plots", "23b_choosing_k_z.pdf"), p_z, width = 12, height = 4.5)
```

# 2. Silhouette against the shuffled band

The family-normalised profile (the one the per-timepoint groups are reported on).

```{r fig_sil, fig.width = 12, fig.height = 4.2}
p_sil <- per_k %>% filter(profile == "families, normalised") %>% mutate(timepoint = factor(timepoint, levels = tp_lab(tps))) %>%
  ggplot(aes(k)) + geom_ribbon(aes(ymin = null_min, ymax = null_max), fill = "grey85") +
  geom_line(aes(y = null_median), linetype = 2, colour = "grey50") +
  geom_line(aes(y = observed), colour = "#0072B2", linewidth = 0.8) + geom_point(aes(y = observed), colour = "#0072B2") +
  geom_vline(aes(xintercept = best_k_z), linetype = 3) + facet_wrap(~ timepoint, nrow = 1) +
  labs(y = "mean silhouette (2,000-gene sample)", title = "Silhouette: real (blue) vs shuffled (grey band = min-max of 50)",
       subtitle = "falls with k on real AND shuffled data - which is why the raw peak (k = 2) cannot choose k; dotted = chosen k") +
  theme_classic(base_size = 12)
p_sil
if (isTRUE(params$save_plots)) ggsave(here("plots", "23b_choosing_k_silhouette.pdf"), p_sil, width = 12, height = 4.2)
```

# 3. Elbow plot, real vs shuffled

Variance explained = 1 - (within-group sum of squares / total sum of squares), on the
family-normalised profile, for k = 1 to 8. Shuffled curves use `n_perm_elbow` shuffles.

```{r elbow}
fam_norm <- function(S, fc) { fam <- setNames(ann$family[match(fc, ann$feature)], fc)
  F <- sapply(sort(unique(fam)), function(g) rowSums(S[, fam == g, drop = FALSE])); F / pmax(rowSums(abs(F)), 1e-12) }
ve <- function(M, k) { if (k == 1) return(0); set.seed(9); km <- kmeans(M, centers = k, nstart = 10, iter.max = 50)
  1 - km$tot.withinss / km$totss }
elbow <- map_dfr(tps, function(tp) {
  o <- readRDS(file.path(O, paste0("shap_oof_", sprintf(params$stem, tp), ".rds")))
  M <- fam_norm(o$S, colnames(o$S))
  real <- tibble(k = c(1, ks), ve = vapply(c(1, ks), function(k) ve(M, k), numeric(1)), set = "real")
  shuf <- map_dfr(seq_len(params$n_perm_elbow), function(i) { set.seed(2000 + i); P <- apply(M, 2, sample)
    tibble(k = c(1, ks), ve = vapply(c(1, ks), function(k) ve(P, k), numeric(1)), rep = i) }) %>%
    group_by(k) %>% summarise(lo = min(ve), hi = max(ve), ve = mean(ve), .groups = "drop") %>% mutate(set = "shuffled")
  bind_rows(real, shuf) %>% mutate(timepoint = tp_lab(tp))
})
elbow %>% dplyr::select(timepoint, set, k, ve) %>% pivot_wider(names_from = set, values_from = ve) %>%
  mutate(real_minus_shuffled = real - shuffled) %>% knitr::kable(digits = 3)
```

```{r fig_elbow, fig.width = 12, fig.height = 4.5}
chosen <- per_k %>% filter(profile == "families, normalised") %>% distinct(timepoint, best_k_z)
p_el <- elbow %>% mutate(timepoint = factor(timepoint, levels = tp_lab(tps))) %>%
  ggplot(aes(k, 100 * ve)) +
  geom_ribbon(data = ~ filter(.x, set == "shuffled"), aes(ymin = 100 * lo, ymax = 100 * hi), fill = "grey85") +
  geom_line(aes(linetype = set, colour = set), linewidth = 0.8) + geom_point(aes(colour = set), size = 1.6) +
  geom_vline(data = chosen %>% mutate(timepoint = factor(timepoint, levels = tp_lab(tps))), aes(xintercept = best_k_z), linetype = 3) +
  scale_colour_manual(values = c(real = "#0072B2", shuffled = "grey45"), name = NULL) +
  scale_linetype_manual(values = c(real = 1, shuffled = 2), name = NULL) +
  facet_wrap(~ timepoint, nrow = 1) + scale_x_continuous(breaks = 1:8) +
  labs(y = "% of gene-to-gene variation captured by the groups", x = "k (number of groups)",
       title = "Elbow plot: real SHAP profiles vs shuffled",
       subtitle = sprintf("families, normalised; shuffled = mean (band = range) of %d shuffles; dotted = chosen k; n = 10,172 genes",
                          params$n_perm_elbow)) +
  theme_classic(base_size = 12) + theme(legend.position = "bottom")
p_el
if (isTRUE(params$save_plots)) ggsave(here("plots", "23b_choosing_k_elbow.pdf"), p_el, width = 12, height = 4.5)
```

# 4. How to read the choice

- The **elbow** shows how much each extra group buys; the gap to the shuffled curve is the part
  of that gain that is real co-variation between families rather than what clustering finds in
  any data.
- The **z curve** is what actually picks k. Where several k's sit at similar z, they are about
  equally supported; the chosen k is the best of near-ties, not a sharp optimum.
- The groups lie along a **continuum** (see the PCA of the profiles in `03b_` / `23_`), so a
  group is better read as a region of a gradient than as a discrete biological class. The k = 2
  split (GC-poor vs GC-rich at 1hr) is the dominant axis; higher k resolves secondary axes such
  as CDS length.

```{r session}
sessionInfo()
```
