---
title: "Verification: GC-rich CDS mRNAs resist eIF4E inhibition (Teleman HeLa)"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  te_change_rds: "output/predictive_modeling/teleman_riboseq_te_fixed.rds"
  absolute_te_rds: "output/predictive_modeling/teleman_absolute_te.rds"
  tx_map_rds: "output/predictive_modeling/precomputed_teleman_tx.rds"
  matrix_rds: "output/translational_control_overview/feature_matrix_tco_hela.rds"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r params_fallback, include=FALSE}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
if (!isTRUE(getOption("knitr.in.progress"))) {
  .fm <- rmarkdown::yaml_front_matter(
    here::here("code", "translational_control_overview", "24_verify_teleman_gc_4e_resistance.Rmd"))
  params <- lapply(.fm$params, function(z) if (is.list(z) && !is.null(z$value)) z$value else z)
  rm(.fm)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# The claim, and how it is checked

**Claim.** In HeLa (Roiuk / Teleman, 4E-BP1(4A) overexpression vs empty vector), mRNAs with a
GC-rich coding sequence keep their translation efficiency better than average when eIF4E is
inhibited - they are relatively resistant.

`code/predictive_modeling/88_lm_gc3_4e_resistance.Rmd` first showed this. This notebook verifies
it independently, from source:

| quantity | source |
|---|---|
| TE change under 4E inhibition | `76_teleman_4e_riboseq_te_validation.Rmd`'s corrected DESeq2 fit; direction re-asserted here two ways |
| the transcript for each gene | HeLa's own dominant transcript (`precomputed_teleman_tx.rds`, Roiuk control RNA) |
| CDS / 5'UTR sequence | GENCODE v49 TxDb + hg38 genome, extracted here |
| GC, GC1-3, lengths | computed here from that sequence; then asserted equal to the HeLa feature matrix |

Then: how strong and how graded the association is, whether length or the 5'UTR explains it,
whether expression level explains it, and whether a known 4E-sensitive class behaves as it should.

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse); library(here); library(GenomicFeatures); library(BSgenome.Hsapiens.UCSC.hg38)
})
```

# 1. The TE change, and its direction

`log2FoldChange` is TE(4E-BP1) relative to TE(empty vector). **Positive = TE held up under eIF4E
inhibition = resistant.** Two independent rulers pin that sign:

1. Teleman / Roiuk name CDKN2B as resistant and GLO1 as sensitive (Fig 1A).
2. The absolute TE levels in each condition (`97_absolute_te_by_4e_class.Rmd`), differenced, must
   track this column positively.

```{r te_change}
fit <- readRDS(here(params$te_change_rds))
te <- fit$fixed %>% transmute(gene_id_clean = sub("[.].*", "", gene_id), symbol, d = log2FoldChange,
                              padj, baseMean) %>% filter(is.finite(d))
abs_te <- readRDS(here(params$absolute_te_rds))
diff_te <- abs_te$inhibitor %>% dplyr::select(gene_id, bp = log2FoldChange) %>%
  inner_join(abs_te$vehicle %>% dplyr::select(gene_id, ev = log2FoldChange), by = "gene_id") %>%
  transmute(gene_id_clean = sub("[.].*", "", gene_id), delta = bp - ev)
chk <- inner_join(te, diff_te, by = "gene_id_clean")
r_anchor <- cor(chk$d, chk$delta)
cdkn2b <- te$d[te$symbol == "CDKN2B"]; glo1 <- te$d[te$symbol == "GLO1"]
cat(sprintf("CDKN2B %+.2f (resistant) | GLO1 %+.2f (sensitive) | vs TE(bp) - TE(ev): r = %.3f (n = %d)\n",
            cdkn2b, glo1, r_anchor, nrow(chk)))
stopifnot("CDKN2B is not resistant - sign flipped" = length(cdkn2b) == 1 && cdkn2b > 0,
          "GLO1 is not sensitive - sign flipped" = length(glo1) == 1 && glo1 < 0,
          "the TE change does not match the difference of the absolute TE levels" = r_anchor > 0.9)
cat(sprintf("genes with a TE change: %d | median %.3f (DESeq2 centres it near 0)\n", nrow(te), median(te$d)))
```

# 2. Sequence features from source

```{r sequences}
txm <- readRDS(here(params$tx_map_rds)) %>%
  transmute(gene_id_clean = as.character(unlist(gene_id_clean)), transcript_id_clean = as.character(unlist(transcript_id_clean)))
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
get_seq <- function(grl) { names(grl) <- sub("[.].*", "", names(grl)); grl <- grl[names(grl) %in% txm$transcript_id_clean]
  s <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38, grl); setNames(toupper(as.character(s)), names(s)) }
cds  <- get_seq(cdsBy(txdb, by = "tx", use.names = TRUE))
utr5 <- get_seq(fiveUTRsByTranscript(txdb, use.names = TRUE))
gcp <- function(s) 100 * str_count(s, "[GC]") / nchar(s)
pos_gc <- function(s, p) {             # GC at codon position p, internal codons (ATG and stop removed)
  n <- nchar(s) %/% 3L
  if (nchar(s) %% 3L != 0L || n < 2L) return(NA_real_)
  cod <- substring(s, seq(1, 3 * n, 3), seq(3, 3 * n, 3))
  if (!tail(cod, 1) %in% c("TAA", "TAG", "TGA")) return(NA_real_)
  cod <- head(cod, -1); if (cod[1] == "ATG") cod <- cod[-1]
  if (!length(cod)) return(NA_real_)
  100 * mean(substr(cod, p, p) %in% c("G", "C"))
}
seqf <- tibble(transcript_id_clean = names(cds), cds = unname(cds)) %>%
  mutate(cds_gc = gcp(cds), gc1 = map_dbl(cds, pos_gc, 1), gc2 = map_dbl(cds, pos_gc, 2), gc3 = map_dbl(cds, pos_gc, 3),
         log2_cds_length = log2(nchar(cds) + 1)) %>%
  left_join(tibble(transcript_id_clean = names(utr5), u5 = unname(utr5)), by = "transcript_id_clean") %>%
  mutate(utr5_gc = gcp(u5), log2_utr5_length = log2(nchar(u5) + 1)) %>% dplyr::select(-cds, -u5)
d <- te %>% inner_join(txm, by = "gene_id_clean") %>% inner_join(seqf, by = "transcript_id_clean")
cat(sprintf("genes with a TE change, a HeLa transcript and a v49 CDS: %d of %d\n", nrow(d), nrow(te)))

# ACTB is GC3-rich; a check on the definition itself.
stopifnot("GC outside [0, 100]" = all(between(d$cds_gc, 0, 100)),
          "ACTB GC3 is not in the GC3-rich range" = d$gc3[d$symbol == "ACTB"] > 60)
```

## The same values are in the HeLa feature matrix

```{r vs_matrix}
m <- readRDS(here(params$matrix_rds))
v <- d %>% inner_join(m %>% dplyr::select(transcript_id_clean, m_cds_gc = cds_gc, m_gc3 = gc3_internal,
                                         m_len = log2_cds_length), by = "transcript_id_clean")
devs <- c(cds_gc = max(abs(v$cds_gc - v$m_cds_gc)), gc3 = max(abs(v$gc3 - v$m_gc3), na.rm = TRUE),
          log2_cds_length = max(abs(v$log2_cds_length - v$m_len)))
print(signif(devs, 3)); cat("transcripts compared:", nrow(v), "\n")
stopifnot("recomputed sequence features differ from the HeLa matrix" = all(devs < 1e-9),
          "GC3 is missing in one source but not the other" = sum(is.na(v$gc3) != is.na(v$m_gc3)) == 0)
```

# 3. The association

```{r association}
rho <- function(x) { k <- is.finite(d$d) & is.finite(d[[x]])
  ct <- cor.test(d$d[k], d[[x]][k], method = "spearman", exact = FALSE)
  tibble(feature = x, rho = unname(ct$estimate), p = ct$p.value, n = sum(k)) }
assoc <- map_dfr(c("cds_gc", "gc1", "gc2", "gc3", "utr5_gc", "log2_cds_length", "log2_utr5_length"), rho)
knitr::kable(assoc, digits = c(0, 3, 3, 0))
```

```{r quintiles}
quint <- d %>% mutate(quintile = ntile(cds_gc, 5)) %>% group_by(quintile) %>%
  summarise(cds_gc_median = median(cds_gc), te_change_median = median(d), te_change_mean = mean(d),
            se = sd(d) / sqrt(n()), frac_resistant = mean(d > 0), n = n(), .groups = "drop")
knitr::kable(quint, digits = 3)
stopifnot("TE change is not monotone across CDS GC quintiles" = all(diff(quint$te_change_median) > 0))
```

```{r figure, fig.width = 11, fig.height = 4.8}
r_cds <- assoc %>% filter(feature == "cds_gc")
p1 <- ggplot(d, aes(cds_gc, d)) +
  geom_point(size = 0.3, alpha = 0.15, colour = "grey30") +
  geom_smooth(method = "loess", colour = "#CC6677", linewidth = 0.9, se = FALSE) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("Spearman rho = %.3f, n = %s", r_cds$rho, format(r_cds$n, big.mark = ","))) +
  labs(x = "CDS GC (%), from GENCODE v49 sequence",
       y = "TE change, log2(4E-BP1(4A) / empty vector)",
       title = "GC-rich CDS mRNAs keep their TE under eIF4E inhibition", subtitle = "HeLa, Roiuk / Teleman") +
  theme_classic(base_size = 12)
p2 <- ggplot(quint, aes(factor(quintile), te_change_mean)) +
  geom_col(fill = "#CC6677", width = 0.7) +
  geom_errorbar(aes(ymin = te_change_mean - se, ymax = te_change_mean + se), width = 0.2) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_text(aes(y = ifelse(te_change_mean > 0, te_change_mean + 0.08, te_change_mean - 0.08),
                label = sprintf("%.0f%% GC\n%.0f%% up", cds_gc_median, 100 * frac_resistant)), size = 3) +
  labs(x = "CDS GC quintile", y = "mean TE change (+/- SE)",
       subtitle = sprintf("n = %s genes per quintile", format(quint$n[1], big.mark = ","))) +
  theme_classic(base_size = 12)
gridExtra::grid.arrange(p1, p2, ncol = 2, widths = c(1.3, 1))
if (isTRUE(params$save_plots))
  ggsave(here("plots", "24_teleman_gc_vs_4e_resistance.pdf"),
         gridExtra::arrangeGrob(p1, p2, ncol = 2, widths = c(1.3, 1)), width = 11, height = 4.8)
```

# 4. Is it something else in disguise?

## Length and the 5'UTR

CDS GC co-varies with CDS length and with the 5'UTR. Partial rank correlation: rank everything,
regress the TE change and CDS GC on the ranks of CDS length, 5'UTR length and 5'UTR GC, and
correlate the residuals.

```{r partial}
pk <- d %>% filter(is.finite(log2_utr5_length), is.finite(utr5_gc))
rr <- function(y) resid(lm(rank(y) ~ rank(pk$log2_cds_length) + rank(pk$log2_utr5_length) + rank(pk$utr5_gc)))
partial <- cor(rr(pk$d), rr(pk$cds_gc))
cat(sprintf("CDS GC vs TE change: rho %.3f raw | %.3f after CDS length, 5'UTR length and 5'UTR GC (n = %d)\n",
            r_cds$rho, partial, nrow(pk)))
stopifnot("the CDS GC association does not survive the length / 5'UTR controls" = partial > 0.3)
```

## Expression level

A GC-dependent library or mapping bias would most likely act through counts. Adding log2
baseMean to a linear model must leave the GC effect intact.

```{r basemean}
lm0 <- lm(d ~ cds_gc + log2_cds_length, data = d)
lm1 <- lm(d ~ cds_gc + log2_cds_length + log2(baseMean), data = d)
kept <- coef(lm1)[["cds_gc"]] / coef(lm0)[["cds_gc"]]
cat(sprintf("CDS GC coefficient: %.4f -> %.4f with log2(baseMean) (%.1f%% kept) | R2 %.3f -> %.3f\n",
            coef(lm0)[["cds_gc"]], coef(lm1)[["cds_gc"]], 100 * kept, summary(lm0)$r.squared, summary(lm1)$r.squared))
stopifnot("expression level explains away the GC effect" = kept > 0.8)
```

## A known 4E-sensitive class behaves as it should

Ribosomal-protein (5'TOP) mRNAs are the textbook eIF4E / mTOR-sensitive class: their TE must fall.

```{r top}
top <- d %>% filter(grepl("^RP[LS][0-9]+[AXYL]?[0-9]*$", symbol))
w <- wilcox.test(top$d, d$d[!d$gene_id_clean %in% top$gene_id_clean])
cat(sprintf("RPL/RPS mRNAs: median TE change %.2f (n = %d) vs all others %.2f | Wilcoxon p = %.2g | median CDS GC %.1f%%\n",
            median(top$d), nrow(top), median(d$d[!d$gene_id_clean %in% top$gene_id_clean]), w$p.value, median(top$cds_gc)))
stopifnot("TOP mRNAs are not 4E-sensitive - the dataset or its sign is off" = median(top$d) < 0)
```

# 5. Conclusion and what it does not show

```{r summary}
tibble(check = c("TE-change direction (CDKN2B, GLO1, absolute-TE difference)",
                 "GC recomputed from GENCODE v49 equals the HeLa matrix",
                 "Spearman CDS GC vs TE change", "Spearman GC3 vs TE change",
                 "Monotone across CDS GC quintiles", "Partial rho after length and 5'UTR",
                 "GC coefficient kept after expression level", "TOP mRNAs fall"),
       result = c("passed", sprintf("passed (max |dev| %.0e)", max(devs)), sprintf("%.3f", r_cds$rho),
                  sprintf("%.3f", assoc$rho[assoc$feature == "gc3"]), "yes",
                  sprintf("%.3f", partial), sprintf("%.1f%%", 100 * kept),
                  sprintf("median %.2f", median(top$d)))) %>% knitr::kable()
```

**Verified:** in this dataset, mRNAs with GC-rich coding sequences hold their TE under eIF4E
inhibition better than the typical mRNA, in a graded way across the whole GC range, and not
because of CDS length, the 5'UTR, or expression level. The 5'UTR itself carries almost none of
it, which is notable for an initiation factor.

**Not shown by these data:**

1. **"Still translated" is relative.** DESeq2 centres the TE change on the median gene, so
   resistant means "lost less than the typical mRNA". If 4E-BP1(4A) lowers global translation,
   GC-rich mRNAs may still drop in absolute terms; that needs a spike-in or a global measure.
2. **A GC-dependent technical bias is reduced, not excluded.** A change between conditions
   cancels any GC bias the two conditions' libraries share, but not one that differs between
   batches; the library-preparation batches would settle it.
3. **One cell line, one way of blocking eIF4E.** A second inhibition dataset would make it firmer.

```{r session}
sessionInfo()
```
