---
title: "Do the NanoLuc codon variants look like eIF3d targets?"
author: "Kate"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  recompute_features: FALSE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# What this notebook does

Four reporter constructs - a parent NanoLuc, a codon-**adapted** variant, an **anti-adapted**
variant, and a **negative control** that is a positional shuffle of the parent's own codons -
are scored by the models built on MDA-MB-231 transcripts, to ask how much each one *looks like*
an mRNA whose translation depends on eIF3d.

**Four models, all intrinsic (57 sequence-derived features).** The external block is 12
measurements made on endogenous genes (eIF3 CLIP, DAP5 sensitivity, published half-lives,
DHX29); a plasmid has none of them, and median-imputing would give all four constructs the
same values while shifting every score toward the average gene. So the intrinsic model is the
only version where every input is real.

| | hypoxia 1hr | normoxia 1+4hr |
|---|---|---|
| **classifier** | promoted vs negative controls, P(target) | same |
| **regression** | predicted TE log2FC on si3d, all genes | same |

**The features are computed from sequence**, because a plasmid has no annotation. That path is
re-implemented in `33b_`/`33c_`/`33d_` and **validated against the stored matrix** by
`33e_`/`33f_`/`33g_`: all 57 reproduce exactly on real transcripts. Sequences are parsed and
spliced by `33a_`.

**What this can and cannot say.** The models were trained on endogenous transcripts. These
constructs sit outside that population - a non-human CDS, vector UTRs, an unusual length
profile - and, more importantly, the adapted and anti-adapted variants were *designed* by
tuning codon usage, which is what several of the strongest features measure. A separation
between them is therefore close to arithmetic, not independent evidence. The informative
comparisons are the ones that were not built in: the size of the separation relative to real
transcripts, whether the ranking agrees across the two conditions and the two model types, and
whether the parent and its shuffled control (identical codon composition) come out together.

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse); library(here); library(ranger); library(treeshap)
})
O <- here("output", "translational_control_overview")
cache <- file.path(O, "plasmid_features.rds")
```

# 1. Features for the four constructs

```{r features}
pl <- read_csv(here("accessories", "transcribed_plasmids_parsed.csv"), show_col_types = FALSE)
stopifnot("expected the four NanoLuc constructs" = nrow(pl) == 4,
          "constructs are not one backbone" = n_distinct(pl$utr5) == 1 && n_distinct(pl$utr3) == 1)

if (params$recompute_features || !file.exists(cache)) {
  source(here("code", "translational_control_overview", "33b_sequence_features.R"))
  source(here("code", "translational_control_overview", "33c_structure_features.R"))
  source(here("code", "translational_control_overview", "33d_g4_features.R"))
  d  <- pl %>% transmute(id = construct, utr5, cds, utr3)
  sq <- seq_features(d)
  f  <- sq %>% left_join(structure_features(d), by = "id") %>%
               left_join(g4_features(d, sq), by = "id")
  saveRDS(f, cache)
} else {
  f <- readRDS(cache)
}

MODELS <- tribble(
  ~label,                  ~kind,         ~suffix,
  "hypoxia 1hr",           "classifier",  "eif33d_promotes_hypoxia_1hr_lfc0.5_tco_intrinsic_corrected_both_clipaggregate",
  "normoxia 1+4hr",        "classifier",  "eif33d_promotes_normoxia_1and4hr_lfc0.5_tco_intrinsic_corrected_both_clipaggregate",
  "hypoxia 1hr",           "regression",  "eif33d_promotes_hypoxia_1hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw",
  "normoxia 1+4hr",        "regression",  "eif33d_promotes_normoxia_1and4hr_lfc0.5_tcoreg_intrinsic_corrected_both_clipaggregate_yraw")

md_of <- function(s, kind) readRDS(file.path(O, paste0(
  if (kind == "classifier") "rf_model_data_" else "rfreg_model_data_", s, ".rds")))
mdl_of <- function(s, kind) readRDS(file.path(O, paste0(
  if (kind == "classifier") "rf_classifier_" else "rfreg_model_", s, ".rds")))

FC <- md_of(MODELS$suffix[1], "classifier")$feature_cols
stopifnot("every model must use the same 57 features" =
            all(map_lgl(seq_len(nrow(MODELS)), ~ identical(md_of(MODELS$suffix[.x], MODELS$kind[.x])$feature_cols, FC))),
          "a feature is missing for the constructs" = all(FC %in% names(f)),
          "a construct feature is NA - the models would see an imputed value" =
            !anyNA(f[, FC]))
X <- as.data.frame(f[, FC]); rownames(X) <- f$id
cat("features:", length(FC), "| constructs:", nrow(X), "| no missing values\n")
```

## Which features can differ at all

The four share a backbone and a protein, so most features are identical by construction. This
table is the honest denominator for everything below: only these columns can separate them,
and between the parent and its shuffled control only the position-dependent ones can.

```{r what_varies}
varies <- f %>% dplyr::select(id, all_of(FC)) %>% pivot_longer(-id) %>%
  group_by(name) %>% summarise(n_distinct = n_distinct(round(value, 10)),
                               spread = diff(range(value)), .groups = "drop") %>%
  filter(n_distinct > 1) %>% arrange(desc(spread))
cat(nrow(varies), "of", length(FC), "features differ between constructs;",
    length(FC) - nrow(varies), "are identical\n")
pn <- f %>% filter(id %in% c("parent_bidir", "negative_control_3d")) %>%
  dplyr::select(all_of(FC)) %>% summarise(across(everything(), ~ diff(range(.x)))) %>%
  pivot_longer(everything()) %>% filter(value > 1e-10)
cat("parent vs negative control (identical codon composition):", nrow(pn), "features differ -",
    paste(pn$name, collapse = ", "), "\n")
print(as.data.frame(varies), digits = 4)
```

# 2. Predictions

A classifier probability is calibrated against a training set deliberately balanced 50/50
between eIF3d-promoted genes and negative controls, so 0.5 is **not** "a typical transcript".
It is reported next to the held-out distribution of each class, which is what makes it
readable. The regression predicts a TE log2 fold change directly and is placed against all
`r format(nrow(readRDS(file.path(O, "feature_matrix_tco.rds"))), big.mark = ",")` transcripts.

```{r predict}
ref_matrix <- readRDS(file.path(O, "feature_matrix_tco.rds"))

predict_one <- function(suffix, kind, label) {
  md <- md_of(suffix, kind); mdl <- mdl_of(suffix, kind)
  # Training medians impute the reference matrix exactly as the model was trained. Imputing with
  # a median leaves that median unchanged, so the saved (already imputed) training split carries
  # the same medians the model was fitted with.
  meds <- md$train %>% dplyr::select(all_of(FC)) %>% summarise(across(everything(), median, na.rm = TRUE))
  ref <- ref_matrix %>% dplyr::select(transcript_id_clean, all_of(FC))
  for (cl in FC) ref[[cl]][is.na(ref[[cl]])] <- meds[[cl]]

  pr <- function(newdata) {
    p <- predict(mdl, data = as.data.frame(newdata))$predictions
    if (is.matrix(p)) p[, md$pos_label] else p
  }
  # Computed before the tibble: inside tibble(), `kind` would resolve to the column being
  # created rather than to the argument.
  is_pos <- if (kind == "classifier") md$test$eif3d_target == md$pos_label else md$test$y > 0.5
  tibble(label = label, kind = kind, suffix = suffix,
         construct = rownames(X), pred = pr(X[, FC]),
         ref_pred = list(pr(ref[, FC])),
         test_pred = list(pr(md$test[, FC])),
         test_is_pos = list(is_pos))
}

preds <- pmap_dfr(list(MODELS$suffix, MODELS$kind, MODELS$label), predict_one) %>%
  rowwise() %>%
  mutate(pct_transcriptome = 100 * mean(ref_pred < pred),
         pct_heldout_pos   = 100 * mean(test_pred[test_is_pos] < pred)) %>%
  ungroup()

print(as.data.frame(preds %>% dplyr::select(kind, label, construct, pred, pct_transcriptome)), digits = 4)
```

```{r pred_table}
preds %>%
  mutate(cell = sprintf("%.3f (%.0f%%)", pred, pct_transcriptome)) %>%
  dplyr::select(construct, kind, label, cell) %>%
  pivot_wider(names_from = c(kind, label), values_from = cell) %>%
  arrange(match(construct, c("3d-adapted", "parent_bidir", "negative_control_3d", "3d-anti_adapted"))) %>%
  as.data.frame() %>% print()
cat("\ncell = prediction (percentile among all transcripts under that model)\n")
```

```{r fig_predictions, fig.width = 12, fig.height = 7.5}
suppressPackageStartupMessages(library(ggrepel))
ord <- c("3d-adapted", "parent_bidir", "negative_control_3d", "3d-anti_adapted")
# The Okabe-Ito orange and sky blue are reserved project-wide for eIF3d and eIF3e, and every
# construct here is an eIF3d experiment, so neither is used to label one construct. The parent
# and its shuffled control get black and grey rather than two near-identical greys.
cols <- c("3d-adapted" = "#D55E00", "parent_bidir" = "black",
          "negative_control_3d" = "grey60", "3d-anti_adapted" = "#0072B2")
short <- c("3d-adapted" = "adapted", "parent_bidir" = "parent",
           "negative_control_3d" = "neg ctrl", "3d-anti_adapted" = "anti-adapted")

bg <- preds %>% dplyr::select(kind, label, ref_pred) %>% distinct(kind, label, .keep_all = TRUE) %>%
  unnest(ref_pred) %>% mutate(panel = paste0(kind, " | ", label))
pt <- preds %>% mutate(panel = paste0(kind, " | ", label),
                       construct = factor(construct, ord), lab = short[as.character(construct)])

# The regression predictions occupy a narrow slice of a wide distribution, so an unzoomed axis
# stacks all four constructs on one tick. Each panel is clipped to the central 98% of its own
# reference distribution, widened if needed to hold every construct.
lims <- bg %>% group_by(panel) %>%
  summarise(lo = quantile(ref_pred, 0.01), hi = quantile(ref_pred, 0.99), .groups = "drop") %>%
  left_join(pt %>% group_by(panel) %>% summarise(pmin = min(pred), pmax = max(pred), .groups = "drop"),
            by = "panel") %>%
  mutate(pad = 0.08 * (pmax(hi, pmax) - pmin(lo, pmin)),
         lo = pmin(lo, pmin) - pad, hi = pmax(hi, pmax) + pad)
bg  <- bg %>% left_join(lims, by = "panel") %>% filter(ref_pred >= lo, ref_pred <= hi)

p <- ggplot(bg, aes(ref_pred)) +
  geom_density(fill = "grey85", colour = "grey55", alpha = 0.7, linewidth = 0.4) +
  geom_vline(data = pt, aes(xintercept = pred, colour = construct), linewidth = 0.9) +
  geom_text_repel(data = pt, aes(x = pred, y = 0, label = lab, colour = construct),
                  angle = 90, hjust = 0, size = 3.1, direction = "x", seed = 9,
                  box.padding = 0.25, min.segment.length = 0.1, segment.size = 0.3,
                  show.legend = FALSE) +
  facet_wrap(~ panel, scales = "free") +
  scale_colour_manual(values = cols, name = NULL, labels = short) +
  expand_limits(y = 0) +
  labs(x = "model prediction", y = "density",
       title = "Where the constructs fall against all transcripts",
       subtitle = sprintf("grey = all %s transcripts scored by the same model; axes clipped to the central 98%% of each distribution",
                          format(nrow(ref_matrix), big.mark = ",")),
       caption = "Classifier: P(eIF3d-promoted), calibrated on a balanced training set, so 0.5 is not 'a typical transcript'.\nRegression: predicted TE log2FC on si3d.") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", plot.caption = element_text(hjust = 0, size = 8, colour = "grey30"))
print(p)
ggsave(here("plots", "33_plasmid_predictions.pdf"), p, width = 12, height = 7.5)
```

The four models use different scales, so percentile against the transcriptome is the one view
that puts them side by side.

```{r fig_percentile, fig.width = 9, fig.height = 4.5}
pp <- preds %>% mutate(construct = factor(construct, rev(ord)),
                       panel = paste0(kind, " | ", label))
ppct <- ggplot(pp, aes(pct_transcriptome, construct, colour = construct)) +
  geom_vline(xintercept = 50, colour = "grey70", linetype = "dashed", linewidth = 0.4) +
  geom_segment(aes(x = 50, xend = pct_transcriptome, yend = construct), linewidth = 0.5) +
  geom_point(size = 3.4) +
  geom_text(aes(label = sprintf("%.0f", pct_transcriptome)), vjust = -1.1, size = 3, show.legend = FALSE) +
  facet_wrap(~ panel, nrow = 1) +
  scale_colour_manual(values = cols, guide = "none") +
  scale_y_discrete(labels = short) +
  scale_x_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)) +
  labs(x = "percentile among all transcripts", y = NULL,
       title = "The same four predictions on one comparable scale",
       subtitle = "dashed = the median transcript; higher = looks more like an eIF3d-dependent mRNA") +
  theme_classic(base_size = 12) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
print(ppct)
ggsave(here("plots", "33_plasmid_percentiles.pdf"), ppct, width = 9, height = 4.5)
```

# 3. Why: per-construct SHAP

SHAP is run on the **regression** models, where the forest is a regression forest and TreeSHAP
applies directly - no surrogate, so no fidelity apparatus is needed. (The classifier is a
probability forest, which `treeshap` cannot take; `03_shap.Rmd` handles that with a surrogate,
which is not worth importing for four sequences.) Values are additive, so a family's
contribution is the exact sum of its members'.

```{r shap, fig.width = 11, fig.height = 7}
ann <- read_csv(file.path(O, "feature_annotation.csv"), show_col_types = FALSE)
shap_one <- function(suffix, label) {
  md <- md_of(suffix, "regression"); mdl <- mdl_of(suffix, "regression")
  uni <- ranger.unify(mdl, as.data.frame(md$train[, FC]))
  s <- treeshap(uni, X[, FC], verbose = FALSE)$shaps
  as_tibble(s) %>% mutate(construct = rownames(X), label = label, .before = 1)
}
shap <- map2_dfr(MODELS$suffix[MODELS$kind == "regression"],
                 MODELS$label[MODELS$kind == "regression"], shap_one)

fam <- shap %>% pivot_longer(all_of(FC), names_to = "feature", values_to = "shap") %>%
  left_join(ann %>% dplyr::select(feature, family), by = "feature") %>%
  group_by(label, construct, family) %>% summarise(shap = sum(shap), .groups = "drop")

pf <- fam %>%
  mutate(construct = factor(construct, ord),
         family = fct_reorder(family, abs(shap), .fun = max)) %>%
  ggplot(aes(shap, family, fill = construct)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  facet_wrap(~ label) +
  scale_fill_manual(values = cols, name = NULL) +
  labs(x = "summed SHAP (predicted TE log2FC)", y = NULL,
       title = "What drives each construct's predicted TE change",
       subtitle = "Regression model, TreeSHAP; family sums are exact because SHAP is additive",
       caption = "Positive = pushes the predicted si3d TE log2FC up (more eIF3d-dependent).") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", plot.caption = element_text(hjust = 0, size = 8, colour = "grey30"))
print(pf)
ggsave(here("plots", "33_plasmid_shap_families.pdf"), pf, width = 11, height = 7)
print(as.data.frame(fam %>% pivot_wider(names_from = construct, values_from = shap)), digits = 3)
```

# 4. Export

```{r export}
write_csv(f, file.path(O, "plasmid_features.csv"))
write_csv(preds %>% dplyr::select(kind, label, construct, pred, pct_transcriptome, pct_heldout_pos),
          file.path(O, "plasmid_predictions.csv"))
write_csv(fam, file.path(O, "plasmid_shap_families.csv"))
cat("wrote plasmid_features.csv, plasmid_predictions.csv, plasmid_shap_families.csv\n")
```
