---
title: "GC3 external sets 01: eIF3d subunit-seq cap-binding enrichment distributions"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  lamper_lfc: 2
  lamper_padj: 0.005
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r drop_stale_params}
# `params` only exists during knit. Interactively it may be a stale list left behind by
# another notebook, which would pass an exists()-only guard and then feed NULLs downstream,
# so drop it first and rebuild it in full below.
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

First notebook of the `gc3_external_sets` folder. It ingests the eIF3d **subunit-seq**
cap-binding tables and does nothing but describe them: how enriched the reported mRNAs are
at the cap, in each condition, before any gene set is cut out of them.

Subunit-seq is the Lee lab eIF3d-TEV assay: a TEV protease domain releases and sequences
only the transcripts held at the 5' cap by eIF3d. Three conditions across two papers:

| condition | source | meaning |
|---|---|---|
| **CM** | Lamper et al. 2020, Table S3 | complete media, unstressed baseline |
| **GLU** | Lamper et al. 2020, Table S3 | glucose deprivation, metabolic stress |
| **ISR** | Lee lab 2023, Table S2 | integrated stress response, chronic stress |

Both files live outside this repo, in `mpra_preliminary/data/`. The loading and the two
replication checks follow `code/predictive_modeling/85_subunit_seq_3d_bound_features.Rmd`,
which consumes the same tables; the thresholds match
`code/clip_binding_sites.Rmd` and `code/ecdf_plots_3e_and_3d.Rmd`.

## What these distributions are and are not

**1. Both tables are gene-level.** `ensembl_gene_id` and `hgnc_symbol` and nothing else.
No transcript IDs, no cluster coordinates, so nothing here can be resolved to an isoform.

**2. These are published tables, not transcriptome-wide results.** Lamper reports 1,141
genes and the ISR table 563. The distributions below therefore describe *the reported
mRNAs*, and there is no unbound comparison group inside these files. A density here says
how strong binding is among genes the papers chose to report, not what fraction of the
transcriptome binds eIF3d.

**3. The ISR table is already filtered.** Every entry has log2FC > 0 by construction, so its
distribution is truncated at zero and its left tail is missing, not empty. Plotted beside
the two Lamper conditions it will always look shifted right. That is the filter, not a
stronger effect, and the panels below are annotated accordingly.

**4. The Lamper file carries less independent information than its 13 columns suggest.**
`IP/IN (AVG)` is *exactly* the mean of the two `RPKM_rep` columns for its condition, to
within floating point (section 1, RPKM identity check). So one of those labels is wrong:
either the per-replicate columns are already IP/input ratios rather than RPKM, or the
averaged column is not a ratio. Nothing in the file resolves which, and there is no
separate input-abundance column either way.

**5. `log2FoldChange` is a shifted copy of the ratio, not an independently shrunk estimate.**
Regressed on `log2(IP/IN)` it gives slope 1.00 and r = 1.000 under GLU, slope 1.03 and
r = 0.987 under CM, with a constant offset of about 2.1 and 1.7 log2 units. That offset is
a normalisation constant between the two libraries. Shrinkage would bend the low end toward
zero and pull the slope below 1; it does not. The column names (`log2FoldChange`, `padj`,
and the full DESeq2 column set in the sister ISR table) point at a DESeq2 origin, but that
is an inference from naming, not something this file establishes.

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    lamper_lfc  = 2,
    lamper_padj = 0.005,
    save_plots  = TRUE
  )
}
stopifnot(
  "params is incomplete" =
    all(c("lamper_lfc", "lamper_padj", "save_plots") %in% names(params)),
  "lamper_lfc must be a single finite number"  = is.numeric(params$lamper_lfc)  &&
    length(params$lamper_lfc) == 1 && is.finite(params$lamper_lfc),
  "lamper_padj must be a single number in (0, 1]" = is.numeric(params$lamper_padj) &&
    length(params$lamper_padj) == 1 && params$lamper_padj > 0 && params$lamper_padj <= 1,
  "save_plots must be TRUE or FALSE" = is.logical(params$save_plots) &&
    length(params$save_plots) == 1
)

LAMPER_LFC  <- params$lamper_lfc
LAMPER_PADJ <- params$lamper_padj
SAVE_PLOTS  <- params$save_plots

SUBUNIT_SEQ_DIR <- "/Users/katematlin/github/mpra_preliminary/data"
LAMPER_XLSX <- file.path(SUBUNIT_SEQ_DIR, "lamper-s3.xlsx")
ISR_XLSX    <- file.path(SUBUNIT_SEQ_DIR, "integrated_stress_response.xlsx")

stopifnot("Lamper Table S3 not found" = file.exists(LAMPER_XLSX),
          "ISR Table S2 not found"    = file.exists(ISR_XLSX))

# All three conditions are eIF3d, so orange alone cannot separate them. GLU keeps eIF3d's
# reserved orange as the headline stress condition; CM and ISR take remaining Okabe-Ito
# hues, avoiding #56B4E9 which is reserved for eIF3e. Same assignment as notebook 85.
cond_colors <- c(
  "CM (complete media)"       = "#CC79A7",
  "GLU (glucose deprivation)" = "#E69F00",
  "ISR (chronic stress)"      = "#009E73"
)
COND_LEVELS <- names(cond_colors)

# n belongs on every panel, so build the legend labels from the plotted rows.
label_with_n <- function(x, n) paste0(x, "  (n = ", format(n, big.mark = ","), ")")
```

---

# Section 1: Load the two tables

```{r load_lamper}
lamper <- read_excel(LAMPER_XLSX) %>%
  clean_names() %>%
  dplyr::mutate(dplyr::across(
    c(log2fold_change_cm, padj_cm, log2fold_change_glu, padj_glu,
      ip_in_cm_avg, ip_in_glu_avg, fold_enrichment_glu_cm),
    ~ suppressWarnings(as.numeric(.))
  )) %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id))

lamper_cols <- c("gene_id_clean", "hgnc_symbol",
                 "log2fold_change_cm", "padj_cm", "log2fold_change_glu", "padj_glu",
                 "ip_in_cm_avg", "ip_in_glu_avg", "fold_enrichment_glu_cm")
stopifnot("Lamper table is missing an expected column" =
            all(lamper_cols %in% names(lamper)))

cat("Lamper Table S3 rows:", nrow(lamper), "\n")
cat("  unique genes:", dplyr::n_distinct(lamper$gene_id_clean), "\n")
cat("  rows with a testable CM  log2FC:", sum(!is.na(lamper$log2fold_change_cm)), "\n")
cat("  rows with a testable GLU log2FC:", sum(!is.na(lamper$log2fold_change_glu)), "\n")

# A duplicated gene key would double-count that gene in every density and quantile below.
stopifnot("Lamper table is not one row per gene" =
            !any(duplicated(lamper$gene_id_clean)))
```

```{r load_isr}
isr <- read_excel(ISR_XLSX) %>%
  clean_names() %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id))

stopifnot("ISR table is missing an expected column" =
            all(c("gene_id_clean", "hgnc_symbol", "log2fold_change", "padj") %in% names(isr)))

cat("ISR Table S2 rows:", nrow(isr), "\n")
cat("  unique genes:", dplyr::n_distinct(isr$gene_id_clean), "\n")
cat("  log2FoldChange range:", round(range(isr$log2fold_change, na.rm = TRUE), 2), "\n")
cat("  max padj:", signif(max(isr$padj, na.rm = TRUE), 3), "\n")

stopifnot("ISR table is not one row per gene" = !any(duplicated(isr$gene_id_clean)))
```

## Replication check: Lamper et al. 2020

The paper's central claim is that eIF3d cap binding is **induced by glucose deprivation**
rather than constitutive. The table has to reproduce that: median IP/input under GLU must
exceed the CM baseline, and the reported per-gene GLU/CM fold enrichment must sit above 1.
Checked before any threshold is applied, so a wrong file or a swapped column pair fails
here rather than silently reversing the biology in every panel below.

```{r replication_check_lamper}
med_ip_cm  <- median(lamper$ip_in_cm_avg,  na.rm = TRUE)
med_ip_glu <- median(lamper$ip_in_glu_avg, na.rm = TRUE)
med_fold   <- median(lamper$fold_enrichment_glu_cm, na.rm = TRUE)

cat("median IP/IN, complete media:      ", round(med_ip_cm, 2), "\n")
cat("median IP/IN, glucose deprivation: ", round(med_ip_glu, 2), "\n")
cat("median fold enrichment (GLU / CM): ", round(med_fold, 2),
    " (expected > 1: cap binding is stress-induced)\n")

stopifnot(
  "Lamper2020: GLU IP/IN does not exceed CM - columns may be swapped" =
    med_ip_glu > med_ip_cm,
  "Lamper2020: median GLU/CM fold enrichment is not above 1" =
    med_fold > 1
)
cat("Lamper 2020 replication check passed\n")
```

```{r plot_check_ip_by_condition, fig.width = 6, fig.height = 4.5}
ip_check <- lamper %>%
  dplyr::select(gene_id_clean,
                `CM (complete media)`       = ip_in_cm_avg,
                `GLU (glucose deprivation)` = ip_in_glu_avg) %>%
  pivot_longer(-gene_id_clean, names_to = "condition", values_to = "ip_in") %>%
  dplyr::filter(!is.na(ip_in), ip_in > 0) %>%
  dplyr::mutate(condition = droplevels(factor(condition, levels = COND_LEVELS)))

ip_check_stats <- ip_check %>%
  group_by(condition) %>%
  summarise(n = n(), median = median(ip_in), .groups = "drop")

p_check_ip <- ggplot(ip_check, aes(condition, ip_in, fill = condition, colour = condition)) +
  geom_violin(alpha = 0.15, linewidth = 0.8) +
  geom_boxplot(width = 0.12, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_text(data = ip_check_stats, inherit.aes = FALSE,
            aes(condition, median, label = paste0("median ", round(median, 1))),
            nudge_x = 0.34, size = 3.2, colour = "grey20") +
  geom_text(data = ip_check_stats, inherit.aes = FALSE,
            aes(condition, y = -Inf, label = paste0("n = ", format(n, big.mark = ","))),
            vjust = -0.6, size = 3.2, colour = "grey30") +
  scale_y_log10() +
  scale_fill_manual(values = cond_colors, guide = "none") +
  scale_colour_manual(values = cond_colors, guide = "none") +
  labs(x = NULL, y = "IP / input enrichment (log10 scale)",
       subtitle = "the replication check, drawn: GLU must sit above CM") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_ip)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_ip_by_condition.pdf"),
                       p_check_ip, width = 6, height = 4.5)
```

The table also carries `Fold Enrichment (GLU/CM)` as its own column. It should be exactly
the ratio of the two IP/IN columns; if it is not, the three columns do not describe the
same pair of samples and the paired panels in section 4 would be comparing different things.

```{r fold_enrichment_identity}
fold_check <- lamper %>%
  dplyr::filter(!is.na(fold_enrichment_glu_cm), !is.na(ip_in_cm_avg), !is.na(ip_in_glu_avg),
                ip_in_cm_avg > 0) %>%
  dplyr::mutate(recomputed = ip_in_glu_avg / ip_in_cm_avg,
                rel_diff   = abs(recomputed - fold_enrichment_glu_cm) /
                             pmax(fold_enrichment_glu_cm, .Machine$double.eps))

cat("genes checked:", nrow(fold_check),
    "| max relative difference:", signif(max(fold_check$rel_diff), 3), "\n")
stopifnot("Fold Enrichment (GLU/CM) is not IP/IN GLU over IP/IN CM" =
            max(fold_check$rel_diff) < 1e-6)
```

```{r plot_check_fold_identity, fig.width = 5.5, fig.height = 5}
p_check_fold <- ggplot(fold_check, aes(recomputed, fold_enrichment_glu_cm)) +
  geom_point(alpha = 0.25, size = 1, colour = "#E69F00") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  scale_x_log10() + scale_y_log10() +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.2,
           label = paste0("n = ", format(nrow(fold_check), big.mark = ","),
                          "\nmax relative difference = ",
                          signif(max(fold_check$rel_diff), 3))) +
  labs(x = "IP/IN GLU divided by IP/IN CM, recomputed here (log10)",
       y = "Fold Enrichment (GLU/CM) as stored (log10)",
       subtitle = "every gene must sit on the dashed identity line") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_fold)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_fold_identity.pdf"),
                       p_check_fold, width = 5.5, height = 5)
```

## Identity check: `IP/IN (AVG)` against the `RPKM` replicate columns

The file presents four `RPKM_*_rep*` columns and two `IP/IN (AVG)` columns, which reads as
per-sample abundance plus a derived ratio. It is not. Each `IP/IN (AVG)` value is the plain
mean of its own two `RPKM` replicate columns, so one of the two labels is wrong and the six
columns carry only two independent quantities. This is asserted rather than described,
because a future revision of the table that made them genuinely different would change what
every panel below is plotting, and should stop the notebook.

```{r rpkm_identity_check}
rpkm_check <- lamper %>%
  dplyr::mutate(
    mean_rpkm_cm  = (rpkm_cm_rep1  + rpkm_cm_rep2)  / 2,
    mean_rpkm_glu = (rpkm_glu_rep1 + rpkm_glu_rep2) / 2
  ) %>%
  dplyr::transmute(
    gene_id_clean,
    `CM (complete media)`       = abs(mean_rpkm_cm  - ip_in_cm_avg)  / pmax(ip_in_cm_avg,  .Machine$double.eps),
    `GLU (glucose deprivation)` = abs(mean_rpkm_glu - ip_in_glu_avg) / pmax(ip_in_glu_avg, .Machine$double.eps)
  ) %>%
  pivot_longer(-gene_id_clean, names_to = "condition", values_to = "rel_diff")

print(rpkm_check %>%
        group_by(condition) %>%
        summarise(n = n(), max_rel_diff = signif(max(rel_diff, na.rm = TRUE), 3),
                  .groups = "drop") %>%
        as.data.frame(), row.names = FALSE)

stopifnot(
  "IP/IN (AVG) is no longer the mean of the two RPKM columns - re-read what the columns are" =
    max(rpkm_check$rel_diff, na.rm = TRUE) < 1e-6
)
cat("IP/IN (AVG) is the mean of the two RPKM replicate columns in both conditions\n")
```

```{r plot_check_rpkm_identity, fig.width = 7.5, fig.height = 4.2}
rpkm_pairs <- lamper %>%
  dplyr::transmute(
    gene_id_clean,
    cm_mean  = (rpkm_cm_rep1  + rpkm_cm_rep2)  / 2, cm_ip  = ip_in_cm_avg,
    glu_mean = (rpkm_glu_rep1 + rpkm_glu_rep2) / 2, glu_ip = ip_in_glu_avg
  ) %>%
  {bind_rows(
    dplyr::transmute(., gene_id_clean, condition = "CM (complete media)",
                     mean_rpkm = cm_mean,  ip_in = cm_ip),
    dplyr::transmute(., gene_id_clean, condition = "GLU (glucose deprivation)",
                     mean_rpkm = glu_mean, ip_in = glu_ip)
  )} %>%
  dplyr::filter(mean_rpkm > 0, ip_in > 0) %>%
  dplyr::mutate(condition = droplevels(factor(condition, levels = COND_LEVELS)))

p_check_rpkm <- ggplot(rpkm_pairs, aes(mean_rpkm, ip_in, colour = condition)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ condition) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = cond_colors, guide = "none") +
  labs(x = "mean of RPKM rep1 and rep2 (log10)",
       y = "IP / IN (AVG) as stored (log10)",
       subtitle = "the two are the same column under two names") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_rpkm)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_rpkm_identity.pdf"),
                       p_check_rpkm, width = 7.5, height = 4.2)
```

## Replication check: Lee lab 2023 ISR table

Published as the set of transcripts **enriched** in the eIF3d-TEV IP, so every entry must
carry a positive log2FoldChange. A negative value would mean the file is a full DESeq2
output rather than the hit table, and every ISR statement below would be describing a
mixture of bound and depleted transcripts.

```{r replication_check_isr}
cat("ISR entries with log2FoldChange > 0:", sum(isr$log2fold_change > 0, na.rm = TRUE),
    "of", nrow(isr), "\n")

stopifnot(
  "ISR table contains non-enriched entries - this is not the filtered hit table" =
    all(isr$log2fold_change > 0, na.rm = TRUE),
  "ISR log2FoldChange must be finite" =
    all(is.finite(isr$log2fold_change[!is.na(isr$log2fold_change)]))
)
cat("ISR 2023 replication check passed\n")
```

The check passes with room to spare, and the margin is the point: the smallest reported ISR
log2FC is 1.145, not something just above zero. The table is cut well inside the positive
range, so its left tail is missing rather than empty.

```{r plot_check_isr_truncation, fig.width = 6.5, fig.height = 4.2}
isr_min <- min(isr$log2fold_change, na.rm = TRUE)

p_check_isr <- ggplot(isr, aes(log2fold_change)) +
  geom_histogram(bins = 60, fill = "#009E73", colour = "#009E73", alpha = 0.15,
                 linewidth = 0.4) +
  geom_vline(xintercept = 0,       linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = isr_min, linewidth = 0.8,     colour = "#009E73") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("n = ", nrow(isr),
                          "\nminimum = ", round(isr_min, 3),
                          "\nentries at or below 0: ",
                          sum(isr$log2fold_change <= 0, na.rm = TRUE))) +
  labs(x = "log2 enrichment, eIF3d-TEV IP vs input (ISR)",
       y = "Genes",
       subtitle = "dashed line: zero. solid line: where the published table actually stops") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_isr)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_isr_truncation.pdf"),
                       p_check_isr, width = 6.5, height = 4.2)
```

---

# Section 2: One row per gene per condition

The three conditions live in two files with different shapes, so everything below is drawn
from a single long table: `condition`, `log2fc`, `padj`, and `ip_in` where the source
provides it (Lamper only).

```{r tidy_long}
enrich <- bind_rows(
  lamper %>%
    dplyr::transmute(gene_id_clean, symbol = hgnc_symbol,
                     condition = "CM (complete media)",
                     log2fc = log2fold_change_cm, padj = padj_cm,
                     ip_in  = ip_in_cm_avg),
  lamper %>%
    dplyr::transmute(gene_id_clean, symbol = hgnc_symbol,
                     condition = "GLU (glucose deprivation)",
                     log2fc = log2fold_change_glu, padj = padj_glu,
                     ip_in  = ip_in_glu_avg),
  isr %>%
    dplyr::transmute(gene_id_clean, symbol = hgnc_symbol,
                     condition = "ISR (chronic stress)",
                     log2fc = log2fold_change, padj = padj,
                     ip_in  = NA_real_)
) %>%
  dplyr::mutate(condition = factor(condition, levels = COND_LEVELS))

stopifnot(
  "an unexpected condition label reached the long table" =
    !any(is.na(enrich$condition)),
  "log2FC must be finite where present" =
    all(is.finite(enrich$log2fc[!is.na(enrich$log2fc)])),
  "padj must lie in [0, 1]" =
    all(enrich$padj >= 0 & enrich$padj <= 1, na.rm = TRUE),
  "IP/IN is a ratio of RPKM and cannot be negative" =
    all(enrich$ip_in >= 0, na.rm = TRUE),
  "each condition must be one row per gene" =
    enrich %>% dplyr::count(condition, gene_id_clean) %>% dplyr::pull(n) %>% max() == 1
)

print(enrich %>%
        group_by(condition) %>%
        summarise(genes = n(),
                  with_log2fc = sum(!is.na(log2fc)),
                  with_padj   = sum(!is.na(padj)),
                  with_ip_in  = sum(!is.na(ip_in)), .groups = "drop") %>%
        as.data.frame())
```

---

The counts in that table are not the same across conditions. All 1,141 Lamper genes carry
an `IP/IN` value in both conditions, but only **770** carry a testable GLU `log2FoldChange`
against 1,133 for CM: DESeq2 returned NA for the rest. So the two log2FC curves in 3b are
drawn over different gene sets, and part of the GLU shift is the 371 genes with no GLU value
rather than stronger binding. The `IP/IN` panel in 3d is the like-for-like marginal comparison, and
section 4 is per gene.

Those NAs are not a sensitivity floor, and they are not random. Every row of this file is a
hit in at least one of the two conditions, so the table is the **union of the two published
hit lists** rather than a results table over a fixed gene universe. The missing values fall
exclusively on genes that entered through the other condition.

```{r union_structure_check}
hit_status <- lamper %>%
  dplyr::transmute(
    gene_id_clean,
    cm_hit  = !is.na(log2fold_change_cm)  & log2fold_change_cm  > LAMPER_LFC &
              padj_cm  < LAMPER_PADJ,
    glu_hit = !is.na(log2fold_change_glu) & log2fold_change_glu > LAMPER_LFC &
              padj_glu < LAMPER_PADJ,
    na_cm   = is.na(log2fold_change_cm),
    na_glu  = is.na(log2fold_change_glu)
  )

cat("hit status of every row:\n")
print(with(hit_status, table(cm_hit = cm_hit, glu_hit = glu_hit)))

cat("\nrows that are a hit in neither condition:",
    sum(!hit_status$cm_hit & !hit_status$glu_hit), "\n")
cat("genes with no GLU log2FC:", sum(hit_status$na_glu),
    "of which CM hits:", sum(hit_status$na_glu & hit_status$cm_hit), "\n")
cat("genes with no CM  log2FC:", sum(hit_status$na_cm),
    "of which GLU hits:", sum(hit_status$na_cm & hit_status$glu_hit), "\n")

# Only assert at the default cut, where the two published hit lists are what the
# thresholds reproduce. A sweep to other values would break the union by construction.
if (LAMPER_LFC == 2 && LAMPER_PADJ == 0.005) {
  stopifnot(
    "some rows are a hit in neither condition - this is not a union of two hit lists" =
      all(hit_status$cm_hit | hit_status$glu_hit),
    "a missing GLU log2FC lands on a gene that is not a CM hit" =
      all(hit_status$cm_hit[hit_status$na_glu]),
    "a missing CM log2FC lands on a gene that is not a GLU hit" =
      all(hit_status$glu_hit[hit_status$na_cm])
  )
  cat("\nconfirmed: the table is the union of the CM and GLU hit lists,",
      "and every NA sits on a gene admitted by the other condition\n")
}
```

Two consequences for everything above and below. The per-condition curves are not "genes
measured in that condition": the GLU curve is its 668 hits plus the CM hits that happened to
retain a GLU value, so its sharp peak is the threshold, not the biology. And this file
contains **no unbound mRNAs at all**, so any later comparison needs a background built from
the expressed transcriptome rather than from non-hit rows here, of which there are none.

---

# Section 3: How enriched are the reported mRNAs?

## 3a: Summary statistics

```{r summary_stats}
enrich_summary <- enrich %>%
  dplyr::filter(!is.na(log2fc)) %>%
  group_by(condition) %>%
  summarise(
    n          = n(),
    min        = round(min(log2fc), 2),
    q25        = round(quantile(log2fc, 0.25), 2),
    median     = round(median(log2fc), 2),
    q75        = round(quantile(log2fc, 0.75), 2),
    max        = round(max(log2fc), 2),
    frac_above_0 = round(mean(log2fc > 0), 3),
    .groups = "drop"
  )
print(as.data.frame(enrich_summary), row.names = FALSE)

ip_summary <- enrich %>%
  dplyr::filter(!is.na(ip_in)) %>%
  group_by(condition) %>%
  summarise(n = n(),
            q25    = round(quantile(ip_in, 0.25), 2),
            median = round(median(ip_in), 2),
            q75    = round(quantile(ip_in, 0.75), 2),
            max    = round(max(ip_in), 1),
            .groups = "drop")
cat("\nIP / input enrichment (Lamper conditions only):\n")
print(as.data.frame(ip_summary), row.names = FALSE)
```

## 3b: log2 enrichment density

The dashed line is the project's established cut for the Lamper table
(`log2FoldChange > 2`, used together with `padj < 0.005` in `clip_binding_sites.Rmd` and
`ecdf_plots_3e_and_3d.Rmd`). It is drawn on all three panels for reference only. The ISR
set takes no further cut, being a hit list already.

```{r plot_log2fc_density, fig.width = 7.5, fig.height = 4.5}
lfc_dat <- enrich %>% dplyr::filter(!is.na(log2fc))
lfc_n   <- lfc_dat %>% dplyr::count(condition)
lfc_lab <- setNames(label_with_n(as.character(lfc_n$condition), lfc_n$n),
                    as.character(lfc_n$condition))

p_lfc_density <- ggplot(lfc_dat, aes(log2fc, fill = condition, colour = condition)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = LAMPER_LFC, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = cond_colors, labels = lfc_lab, name = NULL) +
  scale_colour_manual(values = cond_colors, labels = lfc_lab, name = NULL) +
  labs(x = "log2 enrichment, eIF3d-TEV IP vs input",
       y = "Density",
       subtitle = paste0("dashed line: log2FC = ", LAMPER_LFC,
                         ". ISR is a published hit table, truncated at 0 by construction")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_lfc_density)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_log2fc_density.pdf"),
                       p_lfc_density, width = 7.5, height = 4.5)
```

## 3c: The same distributions as cumulative curves

Densities hide where the mass actually sits relative to a threshold. The ECDF reads the
threshold directly: the height of each curve at the dashed line is the fraction of that
condition's reported genes that fall below the cut.

```{r plot_log2fc_ecdf, fig.width = 7.5, fig.height = 4.5}
p_lfc_ecdf <- ggplot(lfc_dat, aes(log2fc, colour = condition)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = LAMPER_LFC, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = cond_colors, labels = lfc_lab, name = NULL) +
  labs(x = "log2 enrichment, eIF3d-TEV IP vs input",
       y = "Cumulative fraction of reported genes") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_lfc_ecdf)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_log2fc_ecdf.pdf"),
                       p_lfc_ecdf, width = 7.5, height = 4.5)
```

## 3d: Raw IP / input ratios

Plotted on a log10 axis because the ratios span orders of magnitude. Lamper conditions only;
the ISR table reports no IP/input column. Section 3e shows how this scale relates to the
log2FC one.

```{r plot_ip_in_density, fig.width = 7.5, fig.height = 4.5}
ip_dat <- enrich %>% dplyr::filter(!is.na(ip_in), ip_in > 0)
ip_n   <- ip_dat %>% dplyr::count(condition, .drop = TRUE) %>% dplyr::filter(n > 0)
ip_lab <- setNames(label_with_n(as.character(ip_n$condition), ip_n$n),
                   as.character(ip_n$condition))

cat("genes dropped from this panel for a non-positive IP/IN:",
    sum(!is.na(enrich$ip_in) & enrich$ip_in <= 0), "\n")

p_ip <- ggplot(ip_dat, aes(ip_in, fill = condition, colour = condition)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_x_log10() +
  scale_fill_manual(values = cond_colors, labels = ip_lab, name = NULL) +
  scale_colour_manual(values = cond_colors, labels = ip_lab, name = NULL) +
  labs(x = "IP / input enrichment (log10 scale)",
       y = "Density",
       subtitle = "dashed line: IP / input = 1, no enrichment over input") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_ip)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_ip_input_density.pdf"),
                       p_ip, width = 7.5, height = 4.5)
```

## 3e: The two enrichment scales against each other

Whether `log2FoldChange` adds anything over the stored ratio is testable rather than
assumable. If it were a shrunk estimate, the low end would bend toward zero and the slope
would fall below 1. It does not: the relationship is a straight line of slope 1 with a
constant offset, which is a library normalisation constant, and under GLU it is exact.

```{r plot_check_lfc_vs_ratio, fig.width = 7.5, fig.height = 4.2}
scale_pair <- bind_rows(
  lamper %>% dplyr::transmute(gene_id_clean, condition = "CM (complete media)",
                              log2fc = log2fold_change_cm, ip_in = ip_in_cm_avg),
  lamper %>% dplyr::transmute(gene_id_clean, condition = "GLU (glucose deprivation)",
                              log2fc = log2fold_change_glu, ip_in = ip_in_glu_avg)
) %>%
  dplyr::filter(!is.na(log2fc), !is.na(ip_in), ip_in > 0) %>%
  dplyr::mutate(condition = droplevels(factor(condition, levels = COND_LEVELS)),
                log2_ip   = log2(ip_in))

scale_fits <- scale_pair %>%
  group_by(condition) %>%
  summarise(
    n         = n(),
    slope     = coef(lm(log2fc ~ log2_ip))[2],
    intercept = coef(lm(log2fc ~ log2_ip))[1],
    r         = cor(log2fc, log2_ip),
    .groups   = "drop"
  )
print(scale_fits %>%
        dplyr::mutate(dplyr::across(c(slope, intercept, r), ~ round(., 3))) %>%
        as.data.frame(), row.names = FALSE)

p_check_scales <- ggplot(scale_pair, aes(log2_ip, log2fc, colour = condition)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_abline(data = scale_fits, inherit.aes = FALSE,
              aes(slope = slope, intercept = intercept), linewidth = 0.8, colour = "grey20") +
  geom_text(data = scale_fits, inherit.aes = FALSE,
            aes(x = -Inf, y = Inf,
                label = paste0("slope = ", round(slope, 3),
                               "\nintercept = ", round(intercept, 2),
                               "\nr = ", round(r, 3),
                               "\nn = ", format(n, big.mark = ","))),
            hjust = -0.1, vjust = 1.3, size = 3, colour = "grey20") +
  facet_wrap(~ condition) +
  scale_colour_manual(values = cond_colors, guide = "none") +
  labs(x = "log2 of the stored IP / input ratio", y = "stored log2FoldChange",
       subtitle = "dashed: equality. solid: the fitted line, slope 1 with an offset") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_scales)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_lfc_vs_ratio.pdf"),
                       p_check_scales, width = 7.5, height = 4.2)

stopifnot(
  "log2FC is no longer a slope-1 shift of the stored ratio - the columns have changed" =
    all(abs(scale_fits$slope - 1) < 0.1) && all(scale_fits$r > 0.98)
)
```

---

# Section 4: The two Lamper conditions are paired

CM and GLU are measured on the same 1,141 genes, so they can be compared per gene rather
than only as two marginal distributions.

## 4a: Per-gene CM against GLU

```{r plot_cm_vs_glu, fig.width = 6, fig.height = 5.5}
paired <- lamper %>%
  dplyr::filter(!is.na(ip_in_cm_avg), !is.na(ip_in_glu_avg),
                ip_in_cm_avg > 0, ip_in_glu_avg > 0)

frac_up <- mean(paired$ip_in_glu_avg > paired$ip_in_cm_avg)
rho     <- cor(log10(paired$ip_in_cm_avg), log10(paired$ip_in_glu_avg),
               method = "spearman")

p_paired <- ggplot(paired, aes(ip_in_cm_avg, ip_in_glu_avg)) +
  geom_point(alpha = 0.25, size = 1, colour = "#E69F00") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  scale_x_log10() + scale_y_log10() +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.8, size = 3.2,
           label = paste0("Spearman rho = ", round(rho, 3),
                          "\nn = ", format(nrow(paired), big.mark = ","),
                          "\nabove the diagonal: ", round(100 * frac_up), "%")) +
  labs(x = "IP / input, complete media (log10)",
       y = "IP / input, glucose deprivation (log10)",
       subtitle = "dashed line: equal binding in the two conditions") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_paired)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_cm_vs_glu_scatter.pdf"),
                       p_paired, width = 6, height = 5.5)
```

## 4b: Distribution of the induction itself

```{r plot_glu_over_cm, fig.width = 7, fig.height = 4.5}
induction <- lamper %>%
  dplyr::filter(!is.na(fold_enrichment_glu_cm), fold_enrichment_glu_cm > 0) %>%
  dplyr::mutate(log2_glu_over_cm = log2(fold_enrichment_glu_cm))

med_ind <- median(induction$log2_glu_over_cm)

p_induction <- ggplot(induction, aes(log2_glu_over_cm)) +
  geom_density(alpha = 0.15, linewidth = 0.8, fill = "#E69F00", colour = "#E69F00") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = med_ind, linewidth = 0.6, colour = "#E69F00") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("median = ", round(med_ind, 2),
                          "\nn = ", format(nrow(induction), big.mark = ","),
                          "\nabove 0: ", round(100 * mean(induction$log2_glu_over_cm > 0)), "%")) +
  labs(x = "log2 fold enrichment, glucose deprivation over complete media",
       y = "Density",
       subtitle = "dashed line: no change. solid line: median") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_induction)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_glu_over_cm_density.pdf"),
                       p_induction, width = 7, height = 4.5)
```

---

# Section 5: Where the established threshold falls

No gene sets are written here. This section only reports how many genes the existing cut
would take out of each distribution, and checks that number against the counts the rest of
the repo already uses, so a future change of file or threshold surfaces immediately.

```{r threshold_counts}
n_glu <- sum(lamper$log2fold_change_glu > LAMPER_LFC &
               lamper$padj_glu < LAMPER_PADJ, na.rm = TRUE)
n_cm  <- sum(lamper$log2fold_change_cm  > LAMPER_LFC &
               lamper$padj_cm  < LAMPER_PADJ, na.rm = TRUE)
n_isr <- dplyr::n_distinct(isr$gene_id_clean)

threshold_tbl <- tibble(
  condition   = COND_LEVELS,
  reported    = c(nrow(lamper), nrow(lamper), nrow(isr)),
  passing     = c(n_cm, n_glu, n_isr),
  cut_applied = c(paste0("log2FC > ", LAMPER_LFC, " & padj < ", LAMPER_PADJ),
                  paste0("log2FC > ", LAMPER_LFC, " & padj < ", LAMPER_PADJ),
                  "none, published hit table")
) %>%
  dplyr::mutate(frac = round(passing / reported, 3))
print(as.data.frame(threshold_tbl), row.names = FALSE)

# The GLU and CM counts at the default threshold are fixed by clip_binding_sites.Rmd and
# ecdf_plots_3e_and_3d.Rmd. Only assert them when running at that default, so a deliberate
# parameter sweep is not treated as an error.
if (LAMPER_LFC == 2 && LAMPER_PADJ == 0.005) {
  stopifnot(
    "GLU bound-set size disagrees with clip_binding_sites.Rmd (expected 668)" = n_glu == 668,
    "CM bound-set size disagrees with clip_binding_sites.Rmd (expected 664)"  = n_cm  == 664
  )
  cat("bound-set sizes match the established repo counts: GLU 668, CM 664\n")
}

```

```{r plot_check_threshold, fig.width = 8.5, fig.height = 4.2}
# -log10(0) is Inf and would drop points silently, so clamp at the smallest padj the file
# actually reports and say how many were clamped.
padj_floor <- min(enrich$padj[enrich$padj > 0], na.rm = TRUE)
n_clamped  <- sum(enrich$padj == 0, na.rm = TRUE)
cat("padj values clamped at", signif(padj_floor, 3), ":", n_clamped, "\n")

vol <- enrich %>%
  dplyr::filter(!is.na(log2fc), !is.na(padj)) %>%
  dplyr::mutate(
    padj_plot = pmax(padj, padj_floor),
    passes    = if_else(log2fc > LAMPER_LFC & padj < LAMPER_PADJ,
                        "clears the cut", "does not")
  )

vol_counts <- vol %>% dplyr::count(condition, passes)
print(as.data.frame(vol_counts), row.names = FALSE)

pass_colors <- c("clears the cut" = "#0072B2", "does not" = "grey70")

p_check_thresh <- ggplot(vol, aes(log2fc, -log10(padj_plot), colour = passes)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_vline(xintercept = LAMPER_LFC,        linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(LAMPER_PADJ), linetype = "dashed", colour = "grey40") +
  facet_wrap(~ condition) +
  scale_colour_manual(values = pass_colors, name = NULL) +
  labs(x = "log2 enrichment, eIF3d-TEV IP vs input",
       y = "-log10 adjusted p",
       subtitle = paste0("dashed lines: log2FC > ", LAMPER_LFC, " and padj < ", LAMPER_PADJ,
                         ". The ISR panel takes no such cut, it is shown for reference")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 9, colour = "grey40"))

print(p_check_thresh)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_01_check_threshold.pdf"),
                       p_check_thresh, width = 8.5, height = 4.2)
```

```{r write_summaries}
if (SAVE_PLOTS) {
  out_dir <- here("output", "gc3_external_sets")
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
  write_csv(enrich_summary, file.path(out_dir, "01_cap_enrichment_summary.csv"))
  write_csv(threshold_tbl,  file.path(out_dir, "01_threshold_counts.csv"))
  cat("wrote summary tables to", out_dir, "\n")
}
```

---

# Session info

```{r session_info}
sessionInfo()
```
