---
title: "GC3 external sets 02: Okubo et al. iPSC EIF3D riboseq — validation and GC analysis"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Ingest Okubo et al., *Science Advances* 2025;11:eadq5484 — "EIF3D safeguards the homeostasis
of key signaling pathways in human primed pluripotency" — validate it against the paper, then
draw GC for the mRNAs EIF3D **promotes**.

Third independent EIF3D dataset after our MDA-MB-231 (nb89) and Herrmannová HeLa (nb98), and
the **first outside cancer cell lines**: human iPSC, primed pluripotency, CRISPRi rather than
siRNA. If the AT-rich GC3 signature holds here it is a property of EIF3D-dependent translation
rather than of one cell line or knockdown method.

Data: `accessories/okubo sciadv riboseq/` — the Europe PMC supplementary bundle for
PMC11980838, which contains `sciadv.adq5484_data_s1_to_s10.zip`. Science.org returns 403 to
scripted requests; Europe PMC serves the same files openly.

## What makes this dataset different

**1. It is transcript-level.** Unlike Herrmannová (gene-level) and subunit-seq (gene-level),
Data S3/S4 list versioned ENST IDs. That is more information than we have had before — but see
Section 5 for why the figures still use MANE.

**2. The files are pre-split by direction and pre-filtered**, so there is no tested-gene
background. Section 5 states what that costs.

**3. The supplement was corrected after publication.** The v2 PDF cover reads "Data S1 to S10
CORRECTED 27 JUNE 2025". Inside the data zip, `data_s1`/`data_S2` are dated 30 May 2025 while
`data_s3`–`data_s10` are dated 26 June 2025, i.e. S3–S10 were the ones revised. This matters:
the gene counts in the corrected files **do not match the numbers printed in the paper's
text** (Section 3).

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
```

```{r params_fallback}
if (!exists("params") || !is.list(params)) params <- list(save_plots = TRUE)
SAVE_PLOTS <- params$save_plots

OKUBO_DIR <- here("accessories", "okubo sciadv riboseq")
EIF3D_COL <- "#E69F00"   # CLAUDE.md: reserved for eIF3d
REST_COL  <- "grey70"
REST_LAB  <- "Rest of transcriptome"

GC_FEATURES <- c(cds_gc = "CDS GC, all positions",
                 gc1    = "GC1, first codon position",
                 gc2    = "GC2, second codon position",
                 gc3    = "GC3, third codon position")
REGION_FEATURES <- c(utr5_gc = "5' UTR", cds_gc = "CDS", utr3_gc = "3' UTR")

# Paper's reported values, for the replication table.
PAPER_DOWN <- 919L   # dDTEGs, main text
PAPER_UP   <- 402L   # uDTEGs, main text
```

---

# Section 1: Load

`Data S3` = dDTEGs (TE **decreased** on knockdown), `Data S4` = uDTEGs (increased). Legends
taken from the v2 supplement PDF.

```{r load}
stopifnot("Okubo data directory not found" = dir.exists(OKUBO_DIR))

read_okubo <- function(tag) {
  f <- file.path(OKUBO_DIR, paste0("adq5484_data_", tag, ".xlsx"))
  stopifnot(setNames(list(file.exists(f)), paste("missing", basename(f)))[[1]])
  d <- suppressWarnings(read_excel(f))
  stopifnot("unexpected column layout" =
              identical(names(d), c("Gene name", "Transcript ID", "FC", "Adjusted p-value")))
  d %>%
    transmute(symbol        = `Gene name`,
              transcript_id = `Transcript ID`,
              tx_clean      = sub("\\..*", "", `Transcript ID`),
              fc            = suppressWarnings(as.numeric(FC)),
              padj          = suppressWarnings(as.numeric(`Adjusted p-value`)))
}

dDTEG <- read_okubo("s3")   # TE down on KD  -> EIF3D promotes
uDTEG <- read_okubo("s4")   # TE up on KD    -> EIF3D inhibits

cat("Data S3 (dDTEGs):", nrow(dDTEG), "transcripts |",
    n_distinct(dDTEG$symbol), "gene names\n")
cat("Data S4 (uDTEGs):", nrow(uDTEG), "transcripts |",
    n_distinct(uDTEG$symbol), "gene names\n")

stopifnot(
  "dDTEG table is empty"            = nrow(dDTEG) > 100,
  "uDTEG table is empty"            = nrow(uDTEG) > 100,
  "transcript IDs are not unique"   = !any(duplicated(dDTEG$transcript_id)) &&
                                      !any(duplicated(uDTEG$transcript_id)),
  "a transcript appears in both files" =
    length(intersect(dDTEG$transcript_id, uDTEG$transcript_id)) == 0,
  "missing gene names"              = !any(is.na(dDTEG$symbol)) && !any(is.na(uDTEG$symbol))
)
```

---

# Section 2: Validation — thresholds and direction

## 2a. The published thresholds are exactly reproduced

The paper states DTEGs are called at **|log2FC| > 1 and adjusted P < 0.05**. `FC` here is a
**linear ratio**, not a log, so that cut should appear as `FC <= 0.5` in S3 and `FC >= 2` in
S4. If the column were already logged this check would fail immediately.

```{r validate_thresholds}
thr <- bind_rows(
  tibble(file = "Data S3 (dDTEGs)", n = nrow(dDTEG),
         fc_min = min(dDTEG$fc), fc_max = max(dDTEG$fc),
         log2fc_min = min(log2(dDTEG$fc)), log2fc_max = max(log2(dDTEG$fc)),
         padj_max = max(dDTEG$padj)),
  tibble(file = "Data S4 (uDTEGs)", n = nrow(uDTEG),
         fc_min = min(uDTEG$fc), fc_max = max(uDTEG$fc),
         log2fc_min = min(log2(uDTEG$fc)), log2fc_max = max(log2(uDTEG$fc)),
         padj_max = max(uDTEG$padj))
) %>% mutate(across(where(is.numeric), ~ signif(., 4)))
print(as.data.frame(thr), row.names = FALSE)

stopifnot(
  "FC is not a linear ratio - values look already logged" =
    all(dDTEG$fc > 0) && all(uDTEG$fc > 0),
  "Data S3 is not entirely below the FC <= 0.5 cut (|log2FC| > 1)" =
    all(dDTEG$fc <= 0.5 + 1e-9),
  "Data S4 is not entirely above the FC >= 2 cut (|log2FC| > 1)" =
    all(uDTEG$fc >= 2 - 1e-9),
  "an adjusted p-value exceeds 0.05" =
    max(dDTEG$padj) <= 0.05 && max(uDTEG$padj) <= 0.05
)
cat("\nThresholds match the paper exactly: S3 all FC <= 0.5, S4 all FC >= 2, all padj <= 0.05.\n")
```

## 2b. Direction

Methods, quoted: `Fold change_i = TE_i(sgEIF3D) / TE_i(control)` — knockdown over control. So
`FC < 1` means TE **fell** when EIF3D was repressed, i.e. **EIF3D promotes** that transcript.

That is confirmed structurally by the file split: every S3 value is below 1 and every S4 value
above 1, and the files are named d(own)DTEG / u(p)DTEG. Same orientation as Herrmannová
(nb98) and MCF7-SIX1; **opposite** to `translation_categories_si3*_vs_sictrl_*`, where
positive means promotes.

```{r validate_direction}
stopifnot(
  "Data S3 should be entirely FC < 1 (TE down on knockdown)" = all(dDTEG$fc < 1),
  "Data S4 should be entirely FC > 1 (TE up on knockdown)"   = all(uDTEG$fc > 1)
)
cat("Data S3: all", nrow(dDTEG), "transcripts have FC < 1 -> TE fell on knockdown -> EIF3D PROMOTES\n")
cat("Data S4: all", nrow(uDTEG), "transcripts have FC > 1 -> TE rose on knockdown -> EIF3D inhibits\n")
cat("\n=> the EIF3D-promoted set for the GC analysis is Data S3.\n")
```

---

# Section 3: Replication — and a discrepancy with the printed text

```{r validate_counts}
counts_tbl <- tibble(
  set          = c("dDTEGs (Data S3)", "uDTEGs (Data S4)"),
  paper_text   = c(PAPER_DOWN, PAPER_UP),
  obs_transcripts = c(nrow(dDTEG), nrow(uDTEG)),
  obs_genes    = c(n_distinct(dDTEG$symbol), n_distinct(uDTEG$symbol))
) %>% mutate(difference = obs_genes - paper_text)
print(as.data.frame(counts_tbl), row.names = FALSE)

stopifnot(
  "gene counts are not in the neighbourhood of the published values" =
    all(abs(counts_tbl$difference) < 25),
  "gene sets overlap between the two files" =
    length(intersect(unique(dDTEG$symbol), unique(uDTEG$symbol))) == 0
)
```

**The corrected data files give 915 down / 394 up; the paper's text says 919 / 402.**

This is not a parsing error, and it is asserted only loosely above (within 25) for that reason:

- No missing or blank gene names, and no gene or transcript appears in both files.
- Counting distinct GENCODE v49 **gene IDs** instead of gene names gives 911 / 388 — further
  from the published numbers, not closer, because 2.6% / 3.4% of their transcripts are not in
  v49 at all.
- `data_s3`–`data_s10` are dated 26 June 2025 while `data_s1`/`data_S2` are dated 30 May, and
  the supplement carries a "CORRECTED 27 JUNE 2025" notice. The most likely reading is that
  **the text reports pre-correction counts and these files are post-correction.**

Consequence for this repo: the hardcoded `Down 402 / Up 919` in
`code/spikein_literature_comparison/03_bar_te_si3d_si3e_vs_literature.Rmd:304` and `04_...:202`
are the **uncorrected** figures. Worth updating separately; not touched here.

---

# Section 4: Orient to the project convention

```{r orient}
# Gene-level effect size. A gene can list many transcripts (median 2, max 46 in S3) with
# different fold changes, so collapse by the MEDIAN log2FC across its listed transcripts and
# keep the spread for inspection. Negated so positive = EIF3D promotes, matching nb89/nb98.
collapse_genes <- function(d, label) {
  d %>%
    mutate(te_lfc = -log2(fc)) %>%          # negate: positive = promotes
    group_by(symbol) %>%
    summarise(te_lfc     = median(te_lfc),
              te_lfc_min = min(te_lfc), te_lfc_max = max(te_lfc),
              n_tx       = n(), padj_min = min(padj), .groups = "drop") %>%
    mutate(set = label)
}
gene_d <- collapse_genes(dDTEG, "EIF3D-promoted")
gene_u <- collapse_genes(uDTEG, "EIF3D-inhibited")

cat("Gene-level, after negation (positive = EIF3D promotes):\n")
print(as.data.frame(bind_rows(gene_d, gene_u) %>%
  group_by(set) %>%
  summarise(n_genes = n(), median_te_lfc = round(median(te_lfc), 2),
            min = round(min(te_lfc), 2), max = round(max(te_lfc), 2),
            median_tx_per_gene = median(n_tx), .groups = "drop")), row.names = FALSE)

stopifnot(
  "promoted set is not positive after negation" = all(gene_d$te_lfc > 0),
  "inhibited set is not negative after negation" = all(gene_u$te_lfc < 0)
)
cat("\nNegation verified: promoted all > 0, inhibited all < 0.\n")
```

---

# Section 5: Transcript choice — MANE, and why not theirs

Okubo report transcript IDs, which is more than Herrmannová or subunit-seq gave us. The
figures nevertheless use **MANE Select**, for two reasons that only became visible on
inspection:

1. **Their transcripts are mostly not the canonical one** — only **18.3%** of the S3
   transcripts and **11.5%** of S4 are their gene's MANE Select, and a gene lists a median of
   2 (max 46) transcripts. There is no single "their transcript" per gene to use.
2. **There is no tested-gene background in these files.** Data S3/S4 are the significant hits
   only. A background therefore has to come from elsewhere, and if the foreground used their
   isoforms while the background used MANE, the two groups would differ by isoform choice as
   well as by biology.

Using MANE for **both** groups keeps one consistent basis and makes the numbers directly
comparable to nb85 (subunit-seq, also MANE). Their transcript-level calls are retained in the
tables above as a sensitivity record.

**Caveat that stands regardless:** the background is "all other MANE genes", which conflates
*untested* with *unaffected* — the same limitation nb85 carries for published hit lists. It is
stated on every figure.

```{r join_mane}
mane <- readRDS(here("output", "predictive_modeling", "feature_matrix_mane_gc_csc.rds")) %>%
  select(transcript_id_clean, gene_id_clean, symbol,
         all_of(union(names(GC_FEATURES), names(REGION_FEATURES))))
cat("MANE matrix:", nrow(mane), "genes\n")

for (f in union(names(GC_FEATURES), names(REGION_FEATURES))) {
  stopifnot(setNames(list(all(mane[[f]] >= 0 & mane[[f]] <= 100, na.rm = TRUE)),
                     paste(f, "outside [0, 100]"))[[1]])
}
stopifnot("MANE GC3 median outside the expected human range [40, 65]" =
            between(median(mane$gc3, na.rm = TRUE), 40, 65))

# Join on gene SYMBOL: Okubo give no gene IDs, and their transcript IDs are mostly non-MANE
# so they cannot be joined to the MANE matrix directly.
promoted_sym <- gene_d$symbol
inhibited_sym <- gene_u$symbol
cat("promoted genes matched into the MANE matrix:",
    sum(promoted_sym %in% mane$symbol), "of", length(promoted_sym), "\n")
cat("inhibited genes matched into the MANE matrix:",
    sum(inhibited_sym %in% mane$symbol), "of", length(inhibited_sym), "\n")

gc_okubo <- mane %>%
  mutate(group = case_when(symbol %in% promoted_sym  ~ "EIF3D-promoted",
                           symbol %in% inhibited_sym ~ "EIF3D-inhibited",
                           TRUE                      ~ REST_LAB))
print(as.data.frame(count(gc_okubo, group)), row.names = FALSE)
stopifnot("too few promoted genes matched" = sum(gc_okubo$group == "EIF3D-promoted") > 500)
```

---

# Section 6: Figures

```{r plot_fn}
# Same layout as nb85 / nb89 / nb98: unfilled densities stacked over the feature set, dashed
# per-group medians, n in every legend label. `features` selects the family drawn.
gc_density_figure <- function(gc_table, groups, colors, title, subtitle, file = NULL,
                              features = GC_FEATURES) {
  labelled <- unlist(groups, use.names = FALSE)
  stopifnot("groups overlap each other" = !any(duplicated(labelled)))

  d <- gc_table %>% mutate(group = REST_LAB)
  for (nm in names(groups)) d$group[d$symbol %in% groups[[nm]]] <- nm

  lev    <- c(names(groups), REST_LAB)
  counts <- d %>% count(group)
  n_of   <- setNames(counts$n[match(lev, counts$group)], lev); n_of[is.na(n_of)] <- 0L
  lab_of <- setNames(sprintf("%s (n = %s)", lev, format(n_of, big.mark = ",", trim = TRUE)), lev)
  d      <- d %>% mutate(group = factor(lab_of[group], levels = unname(lab_of)))
  pal    <- setNames(unname(colors[lev]), unname(lab_of))
  stopifnot("a group has no colour" = !any(is.na(pal)))

  long <- d %>%
    pivot_longer(all_of(names(features)), names_to = "feature", values_to = "gc") %>%
    filter(!is.na(gc)) %>%
    mutate(feature = factor(features[feature], levels = unname(features)))
  stopifnot("a feature has NAs - legend n would not match the panels" =
              n_distinct(count(long, feature)$n) == 1)

  meds <- long %>% group_by(feature, group) %>%
    summarise(median_gc = median(gc), n = n(), .groups = "drop")

  p <- ggplot(long, aes(x = gc, color = group)) +
    geom_density(fill = NA, linewidth = 0.8) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = group),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    labs(title = title, subtitle = subtitle, x = "GC content (%)", y = "Density") +
    theme_classic(base_size = 15) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(1, "lines"), legend.text = element_text(size = 12),
          plot.subtitle = element_text(size = 11, lineheight = 1.2),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 12))
  print(p)
  if (!is.null(file) && SAVE_PLOTS)
    ggsave(here("plots", file), p, width = 8, height = 2 + 2.25 * length(features))
  print(as.data.frame(meds %>% mutate(median_gc = round(median_gc, 1)) %>%
                        select(feature, group, n, median_gc)), row.names = FALSE)
  invisible(p)
}

OKUBO_NOTE <- paste0(
  "Okubo et al. 2025 (Sci Adv eadq5484), human iPSC, EIF3D CRISPRi day 3\n",
  "Promoted = TE fell on knockdown (their FC = TE(sgEIF3D)/TE(control) < 1), sign flipped here\n",
  "Their cut is |log2FC| > 1 & padj < 0.05, DOUBLE our usual 0.5, so magnitudes are not\n",
  "comparable to nb89 / nb98; direction is\n",
  "Transcript: MANE Select for BOTH groups (only 18% of their listed transcripts are MANE)\n",
  "Background is all other MANE genes, so it conflates untested with unaffected")
```

## 6a. GC by codon position

```{r fig_codon, fig.width = 8, fig.height = 11}
gc_codon <- gc_okubo %>% filter(if_all(all_of(names(GC_FEATURES)), ~ !is.na(.)))
cat("Complete on all four codon-position features:", nrow(gc_codon), "of", nrow(gc_okubo), "\n")

gc_density_figure(
  gc_codon,
  groups = list("EIF3D-promoted" = promoted_sym),
  colors = setNames(c(EIF3D_COL, REST_COL), c("EIF3D-promoted", REST_LAB)),
  title = "GC by codon position: mRNAs promoted by EIF3D (Okubo, human iPSC)",
  subtitle = OKUBO_NOTE,
  file = "gc3ext_02_okubo_gc_by_codon_position_promoted.pdf")
```

## 6b. GC by mRNA region

```{r fig_region, fig.width = 8, fig.height = 9}
gc_region <- gc_okubo %>% filter(if_all(all_of(names(REGION_FEATURES)), ~ !is.na(.)))
cat("Complete on all three regions:", nrow(gc_region), "of", nrow(gc_okubo), "\n")

reg_med <- vapply(names(REGION_FEATURES), function(f) median(gc_region[[f]]), numeric(1))
cat("Background median GC:",
    paste(sprintf("%s %.1f", REGION_FEATURES, reg_med), collapse = " | "), "\n")
stopifnot(
  "5'UTR GC should exceed CDS GC" = reg_med[["utr5_gc"]] > reg_med[["cds_gc"]],
  "CDS GC should exceed 3'UTR GC" = reg_med[["cds_gc"]]  > reg_med[["utr3_gc"]]
)

gc_density_figure(
  gc_region,
  groups = list("EIF3D-promoted" = promoted_sym),
  colors = setNames(c(EIF3D_COL, REST_COL), c("EIF3D-promoted", REST_LAB)),
  title = "GC by mRNA region: mRNAs promoted by EIF3D (Okubo, human iPSC)",
  subtitle = paste0(OKUBO_NOTE, "\nGenes lacking an annotated 5'UTR or 3'UTR excluded"),
  features = REGION_FEATURES,
  file = "gc3ext_02_okubo_gc_by_region_promoted.pdf")
```

---

# Section 7: Summary

```{r summary}
bg <- gc_codon %>% filter(group == REST_LAB)
summ <- bind_rows(
  tibble(set = "MANE background", n = nrow(bg),
         cds_gc = median(bg$cds_gc), gc1 = median(bg$gc1),
         gc2 = median(bg$gc2), gc3 = median(bg$gc3)),
  gc_codon %>% filter(group == "EIF3D-promoted") %>%
    summarise(set = "EIF3D-promoted", n = n(), cds_gc = median(cds_gc),
              gc1 = median(gc1), gc2 = median(gc2), gc3 = median(gc3)),
  gc_codon %>% filter(group == "EIF3D-inhibited") %>%
    summarise(set = "EIF3D-inhibited", n = n(), cds_gc = median(cds_gc),
              gc1 = median(gc1), gc2 = median(gc2), gc3 = median(gc3))
) %>% mutate(across(where(is.numeric), ~ round(., 1)),
             gc3_vs_bg = round(gc3 - gc3[1], 1))
print(as.data.frame(summ), row.names = FALSE)

wil <- map_dfr(c("EIF3D-promoted", "EIF3D-inhibited"), function(g) {
  map_dfr(names(GC_FEATURES), function(f) {
    a <- gc_codon[[f]][gc_codon$group == g]
    b <- gc_codon[[f]][gc_codon$group == REST_LAB]
    tibble(set = g, feature = unname(GC_FEATURES[f]), n = length(a),
           delta = round(median(a) - median(b), 1),
           p_vs_background = format.pval(suppressWarnings(wilcox.test(a, b)$p.value),
                                         digits = 3, eps = 2.2e-16))
  })
})
cat("\nWilcoxon vs MANE background:\n"); print(as.data.frame(wil), row.names = FALSE)

cat("\nCross-dataset context (direction comparable, magnitude NOT - different cell type,\n",
    "transcript basis and LFC cut):\n",
    "  nb89 MDA-MB-231, eIF3d hypoxia-promoted : GC3 50.0 vs bg 57.2  (-7.2)\n",
    "  nb98 HeLa (Herrmannova), eIF3d-promoted : GC3 44.8 vs bg 57.9 (-13.1)\n", sep = "")
```

# Session info

```{r session_info}
sessionInfo()
```
