---
title: "GC3 external sets 03: Herrmannova eLife eIF3 riboseq - validation and GC"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  lfc_cutoff:  0.5
  padj_cutoff: 0.05
  save_plots:  TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Ingest Herrmannová et al., *eLife* 95846 (ribosome profiling of individual eIF3 subunit
knockdowns in **HeLa**), validate it against values reported in the paper, then draw GC by
codon position for the mRNAs each subunit **promotes**.

This is an independent test of the GC3 signal seen in our own MDA-MB-231 data (notebook 89)
and in the published cap-binding / PAR-CLIP sets (notebook 85).

Source: `accessories/herrmannova elife riboseq/elife-95846-supp1-v1.xlsx`.

## Three things that will bite if ignored

**1. The sheets need different `skip` values.** `all data` and the `FP_*` sheets carry four
legend rows above the header; the `TE_*` and gene-set sheets carry one. Reading an `FP_*`
sheet at `skip = 1` silently turns four legend rows into data (4704 rows instead of 4700).

**2. The contrast is inverted relative to our MDA-MB-231 files.** It is
log2(knockdown / non-targeting), so **negative = the factor promotes**. Our
`translation_categories_si3*_vs_sictrl_*.csv` files are the other way round (positive =
promotes). Everything is negated on ingest so that positive = promotes project-wide.

**3. The paper reports no transcripts.** Every sheet is gene-level, so an isoform has to be
chosen for each gene. See Section 4.

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
```

```{r params_fallback}
# `params` only exists during knit; running chunks by hand would otherwise resolve bare
# `params` to an unrelated attached-package function. These match the YAML exactly.
if (!exists("params") || !is.list(params)) {
  params <- list(lfc_cutoff = 0.5, padj_cutoff = 0.05, save_plots = TRUE)
}
LFC_CUT    <- params$lfc_cutoff
PADJ_CUT   <- params$padj_cutoff
SAVE_PLOTS <- params$save_plots

KD        <- c(d = "eIF3d", e = "eIF3e", h = "eIF3h")
# CLAUDE.md reserves #E69F00 for eIF3d and #56B4E9 for eIF3e. eIF3h takes a third
# Okabe-Ito hue; grey70 is the rest of the transcriptome.
KD_COLORS <- c(eIF3d = "#E69F00", eIF3e = "#56B4E9", eIF3h = "#009E73")
REST_LAB  <- "Rest of transcriptome"
REST_COL  <- "grey70"

GC_FEATURES <- c(cds_gc = "CDS GC, all positions",
                 gc1    = "GC1, first codon position",
                 gc2    = "GC2, second codon position",
                 gc3    = "GC3, third codon position")

# Regional GC, ordered 5' -> 3' along the transcript (the nb54 convention). CDS is kept as
# the middle panel so the gradient is readable rather than two disconnected UTR panels.
REGION_FEATURES <- c(utr5_gc = "5' UTR",
                     cds_gc  = "CDS",
                     utr3_gc = "3' UTR")

cat("Params - lfc_cutoff:", LFC_CUT, "| padj_cutoff:", PADJ_CUT,
    "| save_plots:", SAVE_PLOTS, "\n")
```

---

# Section 1: Load

```{r load}
XLSX <- here("accessories", "herrmannova elife riboseq", "elife-95846-supp1-v1.xlsx")
stopifnot("Herrmannova supplement not found" = file.exists(XLSX))

# Columns 4:21 are the 18 log2FoldChange / padj values; some arrive as character.
read_herr <- function(sheet, skip) {
  suppressWarnings(read_excel(XLSX, sheet = sheet, skip = skip)) %>%
    dplyr::mutate(dplyr::across(4:21, ~ suppressWarnings(as.numeric(.))))
}

# skip differs by sheet - see "Three things that will bite" above.
all_data <- read_herr("all data", skip = 4)
te_sheet <- setNames(lapply(names(KD),
                            function(k) read_herr(paste0("TE_", k, " padj 0.05"), skip = 1)),
                     names(KD))
fp_sheet <- setNames(lapply(names(KD),
                            function(k) read_herr(paste0("FP_", k, " padj 0.05"), skip = 4)),
                     names(KD))
gene_sets <- list(
  cyto_rp = read_herr("Cytoplasmic RPs",   skip = 1),
  mito_rp = read_herr("Mitochondrial RPs", skip = 1),
  top     = read_herr("TOP mRNAs",         skip = 1),
  mapk    = read_herr("MAPK signaling pathway hsa04010", skip = 1)
)

cat("all data:", nrow(all_data), "genes,", ncol(all_data), "columns\n")
stopifnot(
  "all data is not 17,520 genes - check the skip"   = nrow(all_data) == 17520,
  "all data is not 22 columns"                      = ncol(all_data) == 22,
  "gene_id column missing - wrong skip"             = "gene_id" %in% names(all_data),
  "gene_id is not unique"                           = !any(duplicated(all_data$gene_id)),
  "gene_id is versioned - expected bare ENSG"       = !any(grepl("\\.", all_data$gene_id))
)
```

---

# Section 2: Validation

Every check here is a `stopifnot`, so a parsing or directionality regression fails the knit
rather than propagating into the figures.

## 2a. These really are the TE columns

The sheet carries three column families — `FP_*` (footprints), `mRNA_*` (mRNA-seq) and
`TE_*`. We use `TE_{d,e,h}_vs_nt.log2FoldChange` / `.padj` and never the other two. Since
TE is defined as footprints over mRNA, the TE column must equal `FP - mRNA` exactly; if it
does not, we are reading the wrong column.

```{r validate_te_columns}
te_col   <- function(k) paste0("TE_",   k, "_vs_nt.log2FoldChange")
te_padj  <- function(k) paste0("TE_",   k, "_vs_nt.padj")
fp_col   <- function(k) paste0("FP_",   k, "_vs_nt.log2FoldChange")
rna_col  <- function(k) paste0("mRNA_", k, "_vs_nt.log2FoldChange")

stopifnot("an expected TE / FP / mRNA column is missing" =
  all(c(vapply(names(KD), te_col, ""),  vapply(names(KD), te_padj, ""),
        vapply(names(KD), fp_col, ""),  vapply(names(KD), rna_col, "")) %in% names(all_data)))

te_identity <- purrr::map_dfr(names(KD), function(k) {
  te <- all_data[[te_col(k)]]; fp <- all_data[[fp_col(k)]]; mr <- all_data[[rna_col(k)]]
  ok <- is.finite(te) & is.finite(fp) & is.finite(mr)
  tibble::tibble(knockdown = KD[[k]], n = sum(ok),
                 cor_te_vs_fp_minus_mrna = round(cor(te[ok], (fp - mr)[ok]), 4),
                 max_abs_diff = max(abs(te - (fp - mr))[ok]))
})
print(as.data.frame(te_identity), row.names = FALSE)
stopifnot("TE column is not FP - mRNA: wrong column picked" =
            all(te_identity$max_abs_diff < 1e-6))
```

## 2b. Direction: the contrast is knockdown vs non-targeting

Each subunit's own mRNA must be strongly **down** in its own knockdown, and flat in the
other two. That fixes the sign of every other column.

```{r validate_direction}
diag_tbl <- purrr::map_dfr(names(KD), function(target) {
  row <- all_data[match(toupper(KD[[target]]), all_data$hgnc_symbol), ]
  tibble::tibble(gene = toupper(KD[[target]]),
                 mRNA_d = row[[rna_col("d")]], mRNA_e = row[[rna_col("e")]],
                 mRNA_h = row[[rna_col("h")]])
}) %>% dplyr::mutate(dplyr::across(where(is.numeric), ~ round(., 2)))
cat("Knockdown diagonal (each factor's own mRNA, log2(KD / non-targeting)):\n")
print(as.data.frame(diag_tbl), row.names = FALSE)

diag_vals <- vapply(names(KD), function(k)
  all_data[[rna_col(k)]][match(toupper(KD[[k]]), all_data$hgnc_symbol)], numeric(1))
offdiag <- vapply(names(KD), function(k) {
  others <- setdiff(names(KD), k)
  max(abs(vapply(others, function(o)
    all_data[[rna_col(o)]][match(toupper(KD[[k]]), all_data$hgnc_symbol)], numeric(1))))
}, numeric(1))

# The paper reports ~16-32 fold knockdown by qPCR; RNA-seq gives ~9 fold. Different assays,
# so assert only that the depletion is unmistakable, not the exact magnitude.
stopifnot(
  "a knockdown target is not depleted in its own KD - direction is flipped" =
    all(diag_vals < -2),
  "a knockdown target moves in a condition that does not target it" =
    all(offdiag < 1)
)
cat("\nDiagonal:", paste(sprintf("%s %.2f", toupper(KD), diag_vals), collapse = ", "),
    "- all < -2, so negative = depleted by the knockdown\n")
cat("=> negative TE_x_vs_nt means TE FELL when the factor was removed, i.e. the factor PROMOTES\n")
```

## 2c. Reproduce the paper's reported gene counts

Herrmannová et al. report DTEG counts split by direction. Reproducing the **split**, not
just the total, is what validates the sign convention above.

```{r validate_dteg_counts}
PAPER_DTEG <- tibble::tribble(
  ~knockdown, ~paper_total, ~paper_down, ~paper_up,
  "eIF3d",           1971L,       1040L,      931L,
  "eIF3e",           1180L,        622L,      558L,
  "eIF3h",             21L,         19L,        2L
)

obs_dteg <- purrr::map_dfr(names(KD), function(k) {
  lf <- all_data[[te_col(k)]]; pa <- all_data[[te_padj(k)]]
  sig <- !is.na(pa) & pa < 0.05 & !is.na(lf)
  tibble::tibble(knockdown = KD[[k]], obs_total = sum(sig),
                 obs_down = sum(sig & lf < 0), obs_up = sum(sig & lf > 0))
})
dteg <- dplyr::left_join(PAPER_DTEG, obs_dteg, by = "knockdown")
print(as.data.frame(dteg), row.names = FALSE)
stopifnot(
  "DTEG totals do not match the paper"          = all(dteg$paper_total == dteg$obs_total),
  "DTEG down counts do not match the paper"     = all(dteg$paper_down  == dteg$obs_down),
  "DTEG up counts do not match the paper"       = all(dteg$paper_up    == dteg$obs_up)
)

# The paper's own filtered sheets must be exactly reproducible from the unfiltered one.
for (k in names(KD)) {
  mine <- all_data$gene_id[!is.na(all_data[[te_padj(k)]]) & all_data[[te_padj(k)]] < 0.05]
  stopifnot(setNames(list(setequal(mine, te_sheet[[k]]$gene_id)),
                     paste0("TE_", k, " sheet is not reproducible from 'all data'"))[[1]])
  fpm <- all_data$gene_id[!is.na(all_data[[paste0("FP_", k, "_vs_nt.padj")]]) &
                            all_data[[paste0("FP_", k, "_vs_nt.padj")]] < 0.05]
  stopifnot(setNames(list(setequal(fpm, fp_sheet[[k]]$gene_id)),
                     paste0("FP_", k, " sheet is not reproducible - check the skip"))[[1]])
}
cat("All six padj<0.05 sheets reproduced exactly from 'all data'\n")
```

## 2d. Reproduce the paper's ribosomal-protein result

The paper reports 14 cytoplasmic and 18 mitochondrial RPs with significantly changed TE in
eIF3eKD, against "only 16" in eIF3dKD.

```{r validate_rp_counts}
rp_counts <- purrr::map_dfr(c(cyto_rp = "cyto_rp", mito_rp = "mito_rp"), function(s) {
  x <- gene_sets[[s]]
  tibble::tibble(set = s, n = nrow(x),
                 sig_d = sum(x$TE_d_vs_nt.padj < 0.05, na.rm = TRUE),
                 sig_e = sum(x$TE_e_vs_nt.padj < 0.05, na.rm = TRUE),
                 sig_h = sum(x$TE_h_vs_nt.padj < 0.05, na.rm = TRUE))
}, .id = "which")
print(as.data.frame(rp_counts), row.names = FALSE)
stopifnot(
  "cytoplasmic RPs significant in eIF3eKD != 14 (paper)" =
    rp_counts$sig_e[rp_counts$set == "cyto_rp"] == 14,
  "mitochondrial RPs significant in eIF3eKD != 18 (paper)" =
    rp_counts$sig_e[rp_counts$set == "mito_rp"] == 18,
  "RPs significant in eIF3dKD != 16 (paper)" = sum(rp_counts$sig_d) == 16
)
cat("Paper: 14 cytoplasmic + 18 mitochondrial RPs in eIF3eKD, 16 total in eIF3dKD - matched\n")
```

## 2e. TOP mRNAs are translationally activated in eIF3dKD and eIF3eKD, not eIF3hKD

The paper's direction claim. "Activated" means TE goes **up** in the knockdown, i.e. the
raw log2FoldChange is positive, i.e. eIF3d/e normally **repress** these. This anchors the
sign from the opposite side to ATF4 below.

```{r validate_top}
top_tbl <- purrr::map_dfr(names(KD), function(k) {
  tv <- gene_sets$top[[te_col(k)]]; bv <- all_data[[te_col(k)]]
  tibble::tibble(knockdown = KD[[k]],
                 top_median = round(median(tv, na.rm = TRUE), 3),
                 background_median = round(median(bv, na.rm = TRUE), 3),
                 wilcox_p = signif(suppressWarnings(wilcox.test(tv, bv)$p.value), 3))
})
print(as.data.frame(top_tbl), row.names = FALSE)
stopifnot(
  "TOP mRNAs are not shifted positive in eIF3dKD / eIF3eKD" =
    all(top_tbl$top_median[top_tbl$knockdown %in% c("eIF3d", "eIF3e")] > 0),
  "TOP mRNAs are not above background in eIF3dKD / eIF3eKD" =
    all((top_tbl$top_median - top_tbl$background_median)[
      top_tbl$knockdown %in% c("eIF3d", "eIF3e")] > 0),
  "TOP mRNAs are not flat-or-down in eIF3hKD" =
    top_tbl$top_median[top_tbl$knockdown == "eIF3h"] <= 0
)
# Reported, not asserted: the eIF3d shift is in the right direction but is not significant
# against this background (the paper calls both d and e significant, likely against a
# different comparator or with a different test).
cat("\nNote: eIF3e clears Wilcoxon comfortably; eIF3d is directionally right but p = ",
    top_tbl$wilcox_p[top_tbl$knockdown == "eIF3d"],
    " against the whole transcriptome. Reported, not asserted.\n", sep = "")
```

## 2f. ATF4 — the opposite-direction anchor

The paper reports ATF4 induction is "almost abolished" in eIF3dKD and eIF3eKD, so eIF3d/e
are *required* for its translation and must come out as promoters: negative raw LFC.

```{r validate_atf4}
atf4 <- all_data[match("ATF4", all_data$hgnc_symbol), ]
atf4_te <- vapply(names(KD), function(k) atf4[[te_col(k)]], numeric(1))
cat("ATF4 raw TE LFC:", paste(sprintf("%s %.2f", KD, atf4_te), collapse = ", "), "\n")
stopifnot(
  "ATF4 is not TE-down in eIF3dKD / eIF3eKD - contradicts the paper" =
    all(atf4_te[c("d", "e")] < -1)
)
cat("ATF4 negative in both => promoted by eIF3d and eIF3e, consistent with the paper.\n",
    "TOP positive, ATF4 negative: the two anchors point opposite ways and both agree with\n",
    "negative = promoted.\n", sep = "")
```

---

# Section 3: Orient to the project convention

Negate so that **positive `te_lfc` = the factor promotes**, matching
`translation_categories_si3*` and MCF7-SIX1 usage elsewhere in this project.

```{r orient}
herr <- all_data %>%
  dplyr::transmute(
    gene_id_clean = gene_id,
    symbol        = hgnc_symbol,
    te_lfc_d  = -.data[[te_col("d")]], te_padj_d = .data[[te_padj("d")]],
    te_lfc_e  = -.data[[te_col("e")]], te_padj_e = .data[[te_padj("e")]],
    te_lfc_h  = -.data[[te_col("h")]], te_padj_h = .data[[te_padj("h")]]
  )

# Post-negation the anchors must flip: ATF4 positive (promoted), TOP negative (repressed).
atf4_post <- herr %>% dplyr::filter(symbol == "ATF4")
top_post  <- median(herr$te_lfc_d[herr$gene_id_clean %in% gene_sets$top$gene_id], na.rm = TRUE)
stopifnot(
  "negation failed: ATF4 should be positive (promoted) after negation" =
    atf4_post$te_lfc_d > 1 && atf4_post$te_lfc_e > 1,
  "negation failed: TOP mRNAs should be negative (repressed) after negation" =
    top_post < 0
)
cat("After negation - ATF4 te_lfc_d =", round(atf4_post$te_lfc_d, 2),
    "(promoted), TOP median te_lfc_d =", round(top_post, 3), "(repressed)\n")
cat("Convention now matches the rest of the project: positive = the factor promotes.\n")
```

---

# Section 4: Transcripts

**The paper supplies none.** All twelve sheets are gene-level (`gene_id`, `ensembl`,
`entrez`, `hgnc_symbol`); there are no transcript identifiers and no cluster coordinates, so
an isoform has to be chosen for each gene before any sequence feature can be computed.

We borrow the HeLa matrix from notebook 78 (`feature_matrix_teleman_4e.rds`) rather than the
MDA-MB-231 one, because this study is HeLa.

**How those transcripts were picked** (notebook 76): the **dominant transcript per gene,
chosen by summed control-condition total-RNA counts** across the control libraries of
**Roiuk et al., GSE243708** HeLa RNA-seq — one row per gene, 98.0% present in the GENCODE
v49 TxDb.

**Caveat.** Same cell line, but the isoform was selected from a *different HeLa experiment*
than Herrmannová's. Notebook 78 measured that this choice matters: of genes shared with the
MDA-MB-231 matrix, only 51.7% get the same isoform. So the transcript assignment is a
modelling decision inherited from another dataset, not something this paper measured.

```{r join_hela}
# Both feature families are pulled here: the codon-position set (Section 6a-c) and the
# regional set (Section 6d). cds_gc is in both, so take the union.
ALL_GC_COLS <- union(names(GC_FEATURES), names(REGION_FEATURES))

hela <- readRDS(here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds")) %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol_hela = symbol,
                dplyr::all_of(ALL_GC_COLS)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

stopifnot("a GC feature is missing from the HeLa matrix" =
            all(ALL_GC_COLS %in% names(hela)))
for (f in ALL_GC_COLS) {
  stopifnot(setNames(list(all(hela[[f]] >= 0 & hela[[f]] <= 100, na.rm = TRUE)),
                     paste(f, "outside [0, 100]"))[[1]])
}
stopifnot("HeLa GC3 median outside the expected human range [40, 65]" =
            dplyr::between(median(hela$gc3, na.rm = TRUE), 40, 65))

gc_herr <- herr %>% dplyr::inner_join(hela, by = "gene_id_clean")
cat("Herrmannova genes:", nrow(herr), "| HeLa matrix:", nrow(hela),
    "| joined:", nrow(gc_herr), "\n")
cat("Lost from Herrmannova (not in the HeLa matrix):", nrow(herr) - nrow(gc_herr), "\n")
cat("HeLa background GC3 median:", round(median(gc_herr$gc3, na.rm = TRUE), 1), "%\n")
stopifnot("join lost too many genes - check the ID convention" = nrow(gc_herr) > 10000)

# A handful of genes have an NA for one GC feature. The legend n is counted before the
# pivot to long form, so leaving them in makes the legend disagree with the panels by a
# few genes. Restrict to complete cases per feature family instead; each figure then has
# one n that is true for all of its panels.
gc_codon <- gc_herr %>%
  dplyr::filter(dplyr::if_all(dplyr::all_of(names(GC_FEATURES)), ~ !is.na(.)))
cat("Complete on all four codon-position features:", nrow(gc_codon), "of", nrow(gc_herr),
    sprintf("(%.1f%%)", 100 * nrow(gc_codon) / nrow(gc_herr)), "\n")
```

---

# Section 5: Gene sets — promoted by each subunit

`padj < PADJ_CUT` **and** `te_lfc > LFC_CUT` (post-negation, so this is the most-depleted-by-
knockdown tail). Both filters, as specified.

```{r gene_sets}
promoted <- function(k, lfc = LFC_CUT, padj = PADJ_CUT) {
  gc_herr$gene_id_clean[
    !is.na(gc_herr[[paste0("te_padj_", k)]]) & gc_herr[[paste0("te_padj_", k)]] < padj &
    !is.na(gc_herr[[paste0("te_lfc_",  k)]]) & gc_herr[[paste0("te_lfc_",  k)]] > lfc]
}
prom_sets <- setNames(lapply(names(KD), promoted), unname(KD))

# Effect-size ranked variants (top N by te_lfc among significant genes), for the two
# knockdowns large enough to support them.
top_n_promoted <- function(k, n) {
  gc_herr %>%
    dplyr::filter(!is.na(.data[[paste0("te_padj_", k)]]),
                  .data[[paste0("te_padj_", k)]] < PADJ_CUT,
                  .data[[paste0("te_lfc_", k)]] > 0) %>%
    dplyr::arrange(dplyr::desc(.data[[paste0("te_lfc_", k)]])) %>%
    head(n) %>% dplyr::pull(gene_id_clean)
}

set_sizes <- purrr::map_dfr(names(KD), function(k) {
  ids <- prom_sets[[KD[[k]]]]
  avail <- sum(!is.na(gc_herr[[paste0("te_padj_", k)]]) &
                 gc_herr[[paste0("te_padj_", k)]] < PADJ_CUT &
                 gc_herr[[paste0("te_lfc_", k)]] > 0)
  tibble::tibble(knockdown = KD[[k]], sig_and_positive = avail,
                 at_lfc_cut = length(ids),
                 can_top200 = avail >= 200, can_top500 = avail >= 500,
                 min_lfc = if (length(ids)) round(min(gc_herr[[paste0("te_lfc_", k)]][
                   gc_herr$gene_id_clean %in% ids]), 2) else NA_real_)
})
cat("Promoted sets in the HeLa matrix (padj <", PADJ_CUT, "& te_lfc >", LFC_CUT, "):\n")
print(as.data.frame(set_sizes), row.names = FALSE)

stopifnot(
  "eIF3d promoted set is implausibly small" = length(prom_sets$eIF3d) > 500,
  "eIF3e promoted set is implausibly small" = length(prom_sets$eIF3e) > 300,
  "eIF3h promoted set should be tiny per the paper" = length(prom_sets$eIF3h) < 100
)
cat("\nNOTE: eIF3h has", length(prom_sets$eIF3h),
    "genes. The paper's own result is that eIF3h knockdown barely changes TE",
    "(21 DTEGs total), so its curve is shown for completeness only and must not be",
    "read as a distribution.\n")
```

---

# Section 6: Figures

```{r plot_fn}
# Stacked density over the four GC features, one or more labelled groups against the rest of
# the transcriptome. Same layout as notebooks 85 and 89: unfilled curves, dashed per-group
# medians, n in every legend label.
gc_density_figure <- function(gc_table, groups, colors, title, subtitle, file = NULL,
                              caution = NULL, features = GC_FEATURES) {
  labelled <- unlist(groups, use.names = FALSE)
  stopifnot("groups overlap each other" = !any(duplicated(labelled)),
            "no groups supplied"        = length(groups) >= 1)

  d <- gc_table %>% dplyr::mutate(group = REST_LAB)
  for (nm in names(groups)) d$group[d$gene_id_clean %in% groups[[nm]]] <- nm

  lev    <- c(names(groups), REST_LAB)
  counts <- d %>% dplyr::count(group)
  n_of   <- setNames(counts$n[match(lev, counts$group)], lev)
  n_of[is.na(n_of)] <- 0L
  lab_of <- setNames(sprintf("%s (n = %s)", lev, format(n_of, big.mark = ",", trim = TRUE)), lev)

  d <- d %>% dplyr::mutate(group = factor(lab_of[group], levels = unname(lab_of)))
  pal <- setNames(unname(colors[lev]), unname(lab_of))
  stopifnot("a group has no colour" = !any(is.na(pal)))

  long <- d %>%
    tidyr::pivot_longer(dplyr::all_of(names(features)),
                        names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc)) %>%
    dplyr::mutate(feature = factor(features[feature], levels = unname(features)))
  # The legend n is counted before this pivot, so it would overstate any panel where a
  # feature is NA. Callers pass complete-case tables; assert that here rather than trust it.
  stopifnot("a feature has NAs - legend n would not match the panels" =
              dplyr::n_distinct(dplyr::count(long, feature)$n) == 1)
  meds <- long %>% dplyr::group_by(feature, group) %>%
    dplyr::summarise(median_gc = median(gc), n = dplyr::n(), .groups = "drop")

  p <- ggplot(long, aes(x = gc, color = group)) +
    geom_density(fill = NA, linewidth = 0.8) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = group),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    labs(title = title,
         subtitle = paste0(subtitle, if (!is.null(caution)) paste0("\n", caution) else ""),
         x = "GC content (%)", y = "Density") +
    theme_classic(base_size = 15) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(1, "lines"),
          legend.text = element_text(size = 12),
          plot.subtitle = element_text(size = 11, lineheight = 1.2),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 12))
  print(p)
  if (!is.null(file) && SAVE_PLOTS)
    ggsave(here("plots", file), p, width = 8, height = 2 + 2.25 * length(features))

  print(as.data.frame(meds %>% dplyr::mutate(median_gc = round(median_gc, 1)) %>%
                        dplyr::select(feature, group, n, median_gc)), row.names = FALSE)
  invisible(p)
}

TRANSCRIPT_NOTE <- paste0(
  "Transcript: the paper is gene-level (no transcript IDs), so each gene uses the HeLa\n",
  "dominant transcript from Roiuk et al. GSE243708, picked by summed control total-RNA\n",
  "counts (nb76). Same cell line, different experiment - an inherited modelling choice")
DIRECTION_NOTE <- "Promoted = TE fell on knockdown (log2(KD / non-targeting) negative), sign flipped here"
```

## 6a. Each subunit separately

```{r fig_per_subunit, fig.width = 8, fig.height = 11}
for (k in names(KD)) {
  nm  <- KD[[k]]
  ids <- prom_sets[[nm]]
  lab <- paste0(nm, "-promoted")
  caution <- if (length(ids) < 100)
    paste0("CAUTION: n = ", length(ids),
           ". eIF3h knockdown barely changes TE (21 DTEGs in the paper) - not a distribution")
  else NULL
  gc_density_figure(
    gc_codon,
    groups = setNames(list(ids), lab),
    colors = setNames(c(KD_COLORS[[nm]], REST_COL), c(lab, REST_LAB)),
    title  = paste0("GC by codon position: mRNAs promoted by ", nm, " (Herrmannova, HeLa)"),
    subtitle = paste0(
      "Positive: ", nm, " TE padj < ", PADJ_CUT, " & |TE LFC| > ", LFC_CUT,
      ", most depleted by knockdown\n", DIRECTION_NOTE, "\n", TRANSCRIPT_NOTE),
    caution = caution,
    file = paste0("gc3ext_03_gc_by_codon_position_", nm, "_promoted.pdf"))
  cat("\n")
}
```

## 6b. Effect-size ranked, top 200 and top 500

```{r fig_topn, fig.width = 8, fig.height = 11}
for (k in c("d", "e")) {
  nm <- KD[[k]]
  for (N in c(200, 500)) {
    ids <- top_n_promoted(k, N)
    if (length(ids) < N) { cat("SKIPPED", nm, "top", N, "- only", length(ids), "available\n"); next }
    lab <- paste0(nm, "-promoted, top ", N)
    floor_lfc <- round(min(gc_herr[[paste0("te_lfc_", k)]][gc_herr$gene_id_clean %in% ids]), 2)
    gc_density_figure(
      gc_codon,
      groups = setNames(list(ids), lab),
      colors = setNames(c(KD_COLORS[[nm]], REST_COL), c(lab, REST_LAB)),
      title  = paste0("GC by codon position: top ", N, " promoted by ", nm,
                      " (Herrmannova, HeLa)"),
      subtitle = paste0(
        "Positive: the ", N, " largest ", nm, " TE increases after sign flip (padj < ",
        PADJ_CUT, "), i.e. most depleted by knockdown. LFC floor ", floor_lfc, "\n",
        DIRECTION_NOTE, "\n", TRANSCRIPT_NOTE),
      file = paste0("gc3ext_03_gc_by_codon_position_", nm, "_promoted_top", N, ".pdf"))
    cat("\n")
  }
}
```

## 6c. All three subunits on one grid

```{r fig_grid, fig.width = 14, fig.height = 12}
grid_long <- purrr::map_dfr(names(KD), function(k) {
  nm  <- KD[[k]]
  ids <- prom_sets[[nm]]
  ids <- intersect(ids, gc_codon$gene_id_clean)
  dplyr::bind_rows(
    gc_codon %>% dplyr::filter(gene_id_clean %in% ids) %>% dplyr::mutate(group = nm),
    gc_codon %>% dplyr::filter(!gene_id_clean %in% ids) %>% dplyr::mutate(group = REST_LAB)
  ) %>% dplyr::mutate(subunit = nm)
}) %>%
  dplyr::select(subunit, group, dplyr::all_of(names(GC_FEATURES))) %>%
  tidyr::pivot_longer(-c(subunit, group), names_to = "feature", values_to = "gc") %>%
  dplyr::filter(!is.na(gc)) %>%
  dplyr::mutate(feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES)),
                subunit = factor(subunit, levels = unname(KD)))

# n per group, counted on ONE column and ONE feature: the background is repeated once per
# subunit, so counting across the whole frame would report three times the real background.
grid_n <- c(vapply(unname(KD), function(nm)
              length(intersect(prom_sets[[nm]], gc_codon$gene_id_clean)), integer(1)),
            setNames(nrow(gc_codon) -
                       length(intersect(prom_sets[[KD[["d"]]]], gc_codon$gene_id_clean)),
                     REST_LAB))
grid_lev <- c(unname(KD), REST_LAB)
grid_lab <- setNames(sprintf("%s (n = %s)", grid_lev,
                             format(grid_n[grid_lev], big.mark = ",", trim = TRUE)), grid_lev)

grid_long <- grid_long %>%
  dplyr::mutate(group = factor(grid_lab[group], levels = unname(grid_lab[grid_lev])))
grid_meds <- grid_long %>% dplyr::group_by(subunit, feature, group) %>%
  dplyr::summarise(median_gc = median(gc), .groups = "drop")
grid_pal <- setNames(c(unname(KD_COLORS[unname(KD)]), REST_COL), unname(grid_lab[grid_lev]))

p_grid <- ggplot(grid_long, aes(x = gc, color = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  geom_vline(data = grid_meds, aes(xintercept = median_gc, color = group),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  facet_grid(feature ~ subunit, scales = "free_y") +
  scale_color_manual(values = grid_pal, name = NULL) +
  labs(title = "GC by codon position: mRNAs promoted by each eIF3 subunit (Herrmannova, HeLa)",
       subtitle = paste0(
         "Promoted: TE padj < ", PADJ_CUT, " & |TE LFC| > ", LFC_CUT,
         ", most depleted by that subunit's knockdown\n", DIRECTION_NOTE, "\n",
         TRANSCRIPT_NOTE, "\nGrey differs slightly per column (each excludes its own ",
         "positive set); legend n is the eIF3d column's"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 15) +
  theme(legend.position = "top", legend.direction = "vertical",
        legend.key.height = unit(1, "lines"),
        legend.text = element_text(size = 12),
        plot.subtitle = element_text(size = 11, lineheight = 1.2),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 12))
print(p_grid)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_03_gc_by_codon_position_by_subunit_grid.pdf"),
                       p_grid, width = 14, height = 12)
```

## 6d. GC by mRNA region: 5'UTR, CDS, 3'UTR

The same sets against the same background, but split by **region of the transcript** rather
than by codon position. Ordered 5' -> 3'; CDS is kept as the middle panel so the gradient
reads as a progression rather than two disconnected UTR panels.

Restricted to genes with all three regions annotated, so the legend n is the n of every
panel. Genes without an annotated UTR are dropped rather than silently shortening one curve.

```{r region_setup}
stopifnot("a regional GC column is missing from the HeLa matrix" =
            all(names(REGION_FEATURES) %in% names(gc_herr)))
for (f in names(REGION_FEATURES)) {
  stopifnot(setNames(list(all(gc_herr[[f]] >= 0 & gc_herr[[f]] <= 100, na.rm = TRUE)),
                     paste(f, "outside [0, 100]"))[[1]])
}

gc_region <- gc_herr %>%
  dplyr::filter(dplyr::if_all(dplyr::all_of(names(REGION_FEATURES)), ~ !is.na(.)))
cat("Genes with all three regions annotated:", nrow(gc_region), "of", nrow(gc_herr),
    sprintf("(%.1f%%)", 100 * nrow(gc_region) / nrow(gc_herr)), "\n")

# Biological spot-check: human mRNAs are GC-rich at the 5' end and AT-rich at the 3' end,
# so the medians must order 5'UTR > CDS > 3'UTR. Catches a column/label swap, which the
# range check above cannot.
reg_med <- vapply(names(REGION_FEATURES), function(f) median(gc_region[[f]]), numeric(1))
cat("Background median GC:",
    paste(sprintf("%s %.1f", REGION_FEATURES, reg_med), collapse = " | "), "\n")
stopifnot(
  "5'UTR GC should exceed CDS GC" = reg_med[["utr5_gc"]] > reg_med[["cds_gc"]],
  "CDS GC should exceed 3'UTR GC" = reg_med[["cds_gc"]]  > reg_med[["utr3_gc"]]
)
```

```{r fig_region_per_subunit, fig.width = 8, fig.height = 9}
for (k in names(KD)) {
  nm  <- KD[[k]]
  ids <- intersect(prom_sets[[nm]], gc_region$gene_id_clean)
  lab <- paste0(nm, "-promoted")
  caution <- if (length(ids) < 100)
    paste0("CAUTION: n = ", length(ids),
           ". eIF3h knockdown barely changes TE (21 DTEGs in the paper) - not a distribution")
  else NULL
  gc_density_figure(
    gc_region,
    groups = setNames(list(ids), lab),
    colors = setNames(c(KD_COLORS[[nm]], REST_COL), c(lab, REST_LAB)),
    title  = paste0("GC by mRNA region: mRNAs promoted by ", nm, " (Herrmannova, HeLa)"),
    subtitle = paste0(
      "Positive: ", nm, " TE padj < ", PADJ_CUT, " & |TE LFC| > ", LFC_CUT,
      ", most depleted by knockdown\n", DIRECTION_NOTE, "\n",
      "Genes lacking an annotated 5'UTR or 3'UTR are excluded, so all panels share one n\n",
      TRANSCRIPT_NOTE),
    caution = caution,
    features = REGION_FEATURES,
    file = paste0("gc3ext_03_gc_by_region_", nm, "_promoted.pdf"))
  cat("\n")
}
```

```{r fig_region_grid, fig.width = 14, fig.height = 9.5}
reg_grid <- purrr::map_dfr(names(KD), function(k) {
  nm  <- KD[[k]]
  ids <- intersect(prom_sets[[nm]], gc_region$gene_id_clean)
  dplyr::bind_rows(
    gc_region %>% dplyr::filter(gene_id_clean %in% ids)  %>% dplyr::mutate(group = nm),
    gc_region %>% dplyr::filter(!gene_id_clean %in% ids) %>% dplyr::mutate(group = REST_LAB)
  ) %>% dplyr::mutate(subunit = nm)
}) %>%
  dplyr::select(subunit, group, dplyr::all_of(names(REGION_FEATURES))) %>%
  tidyr::pivot_longer(-c(subunit, group), names_to = "feature", values_to = "gc") %>%
  dplyr::filter(!is.na(gc)) %>%
  dplyr::mutate(feature = factor(REGION_FEATURES[feature], levels = unname(REGION_FEATURES)),
                subunit = factor(subunit, levels = unname(KD)))

# n counted from the gene sets, not the plotted rows: the background repeats once per column.
reg_n <- c(vapply(unname(KD), function(nm)
             length(intersect(prom_sets[[nm]], gc_region$gene_id_clean)), integer(1)),
           setNames(nrow(gc_region) -
                      length(intersect(prom_sets[[KD[["d"]]]], gc_region$gene_id_clean)),
                    REST_LAB))
reg_lev <- c(unname(KD), REST_LAB)
reg_lab <- setNames(sprintf("%s (n = %s)", reg_lev,
                            format(reg_n[reg_lev], big.mark = ",", trim = TRUE)), reg_lev)
reg_grid <- reg_grid %>%
  dplyr::mutate(group = factor(reg_lab[group], levels = unname(reg_lab[reg_lev])))
reg_meds <- reg_grid %>% dplyr::group_by(subunit, feature, group) %>%
  dplyr::summarise(median_gc = median(gc), .groups = "drop")
reg_pal <- setNames(c(unname(KD_COLORS[unname(KD)]), REST_COL), unname(reg_lab[reg_lev]))

p_reg <- ggplot(reg_grid, aes(x = gc, color = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  geom_vline(data = reg_meds, aes(xintercept = median_gc, color = group),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  facet_grid(feature ~ subunit, scales = "free_y") +
  scale_color_manual(values = reg_pal, name = NULL) +
  labs(title = "GC by mRNA region: mRNAs promoted by each eIF3 subunit (Herrmannova, HeLa)",
       subtitle = paste0(
         "Promoted: TE padj < ", PADJ_CUT, " & |TE LFC| > ", LFC_CUT,
         ", most depleted by that subunit's knockdown\n", DIRECTION_NOTE, "\n",
         "Genes lacking an annotated 5'UTR or 3'UTR excluded (", nrow(gc_region), " of ",
         nrow(gc_herr), " kept)\n", TRANSCRIPT_NOTE,
         "\nGrey differs slightly per column (each excludes its own positive set); ",
         "legend n is the eIF3d column's"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 15) +
  theme(legend.position = "top", legend.direction = "vertical",
        legend.key.height = unit(1, "lines"),
        legend.text = element_text(size = 12),
        plot.subtitle = element_text(size = 11, lineheight = 1.2),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 12))
print(p_reg)
if (SAVE_PLOTS) ggsave(here("plots", "gc3ext_03_gc_by_region_by_subunit_grid.pdf"),
                       p_reg, width = 14, height = 9.5)

cat("\nRegional GC medians (background =", nrow(gc_region), "genes):\n")
print(as.data.frame(reg_meds %>% dplyr::mutate(median_gc = round(median_gc, 1)) %>%
  tidyr::pivot_wider(names_from = feature, values_from = median_gc)), row.names = FALSE)

reg_wilcox <- purrr::map_dfr(names(KD), function(k) {
  nm <- KD[[k]]; ids <- intersect(prom_sets[[nm]], gc_region$gene_id_clean)
  if (length(ids) < 20) return(NULL)
  purrr::map_dfr(names(REGION_FEATURES), function(f) {
    a <- gc_region[[f]][gc_region$gene_id_clean %in% ids]
    b <- gc_region[[f]][!gc_region$gene_id_clean %in% ids]
    tibble::tibble(knockdown = nm, region = unname(REGION_FEATURES[f]), n = length(a),
                   median_set = round(median(a), 1), median_rest = round(median(b), 1),
                   delta = round(median(a) - median(b), 1),
                   p_vs_rest = format.pval(suppressWarnings(wilcox.test(a, b)$p.value),
                                           digits = 3, eps = 2.2e-16))
  })
})
cat("\nRegional GC, promoted set vs rest:\n")
print(as.data.frame(reg_wilcox), row.names = FALSE)
```

---

# Section 7: Summary table

```{r summary}
bg_med <- vapply(names(GC_FEATURES), function(f) median(gc_codon[[f]]), numeric(1))

summary_tbl <- purrr::map_dfr(names(KD), function(k) {
  nm <- KD[[k]]
  purrr::map_dfr(list(list(lab = paste0(nm, " promoted (lfc>", LFC_CUT, ")"),
                           ids = prom_sets[[nm]]),
                      list(lab = paste0(nm, " top 200"), ids = top_n_promoted(k, 200)),
                      list(lab = paste0(nm, " top 500"), ids = top_n_promoted(k, 500))),
                 function(s) {
    if (length(s$ids) < 20) return(NULL)
    x <- gc_codon %>% dplyr::filter(gene_id_clean %in% s$ids)
    tibble::tibble(set = s$lab, n = nrow(x),
                   cds_gc = round(median(x$cds_gc), 1), gc1 = round(median(x$gc1), 1),
                   gc2 = round(median(x$gc2), 1), gc3 = round(median(x$gc3), 1),
                   gc3_vs_bg = round(median(x$gc3) - bg_med[["gc3"]], 1))
  })
})
summary_tbl <- dplyr::bind_rows(
  tibble::tibble(set = "HeLa background (complete GC)", n = nrow(gc_codon),
                 cds_gc = round(bg_med[["cds_gc"]], 1), gc1 = round(bg_med[["gc1"]], 1),
                 gc2 = round(bg_med[["gc2"]], 1), gc3 = round(bg_med[["gc3"]], 1),
                 gc3_vs_bg = 0),
  summary_tbl)
print(as.data.frame(summary_tbl), row.names = FALSE)

cat("\nComparability: this is HeLa, on Roiuk-derived dominant transcripts. Notebook 89's\n",
    "MDA-MB-231 numbers use a different cell line AND a different transcript matrix, so\n",
    "the DIRECTION of any GC3 shift is comparable between them but the MAGNITUDE is not.\n",
    sep = "")

# Wilcoxon of each promoted set against everything else, per feature.
wilcox_tbl <- purrr::map_dfr(names(KD), function(k) {
  nm <- KD[[k]]; ids <- prom_sets[[nm]]
  if (length(ids) < 20) return(NULL)
  purrr::map_dfr(names(GC_FEATURES), function(f) {
    a <- gc_codon[[f]][gc_codon$gene_id_clean %in% ids]
    b <- gc_codon[[f]][!gc_codon$gene_id_clean %in% ids]
    tibble::tibble(knockdown = nm, feature = unname(GC_FEATURES[f]), n = length(a),
                   p_vs_rest = format.pval(suppressWarnings(wilcox.test(a, b)$p.value),
                                           digits = 3, eps = 2.2e-16))
  })
})
cat("\nWilcoxon, promoted set vs rest of transcriptome:\n")
print(as.data.frame(wilcox_tbl), row.names = FALSE)
```

# Session info

```{r session_info}
sessionInfo()
```
