---
title: "Feature Extraction for Hypoxia Response Prediction"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis extracts mRNA sequence and structural features for all genes to build a predictive model of Translation Efficiency (TE) Log2 Fold Change at 1 hour hypoxia.

## Feature Categories

| Category | Features | Count |
|----------|----------|-------|
| **Region Lengths** | 5'UTR, CDS, 3'UTR lengths + log2 transforms | 7 |
| **GC Content** | GC% for 5'UTR, CDS, 3'UTR, full transcript | 4 |
| **K-mer Frequencies** | 4-mers in 5'UTR (256) | 256 |
| **RBP Motifs** | RBNS and CISBP-RNA PWM scores | ~50-100 |
| **Codon Usage** | GC3 content | 1 |

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(tximport)
library(scales)
library(FeatureReachR)
```

# Load Data

## Gene Annotation

```{r load_gene_annotation}
# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

## TE Data (1hr Hypoxia Response)

```{r load_te_data}
# Load TE changes from hypoxia timecourse at 1hr
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))

cat("Loaded TE data:", nrow(te_1hr), "genes\n")

# Add gene symbols
te_1hr <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  filter(!is.na(symbol))

cat("TE data with symbols:", nrow(te_1hr), "genes\n")
```

## Load TxDb and Genome

```{r load_txdb}
# Load pre-built TxDb from Gencode v26
txdb <- loadDb("/Users/katematlin/github/mpra_preliminary/accessories/txdb.gencode26.sqlite")

# Load gene info for transcript-to-symbol mapping
geneInfo <- read_csv("/Users/katematlin/github/mpra_preliminary/accessories/gencode.v26.primary.info.csv.gz",
                     col_names = FALSE) %>%
  filter(!str_detect(X2, "pre_"))

colnames(geneInfo) <- c("gene_id", "transcript_id", "biotype", "symbol")

# Create transcript to gene mapping
tx2gene <- geneInfo[, c("transcript_id", "gene_id")]
gene_symbol_map <- geneInfo %>% dplyr::select(gene_id, symbol) %>% distinct()

# Load genome
genome <- BSgenome.Hsapiens.UCSC.hg38

cat("Loaded TxDb and genome\n")
cat("Transcripts in geneInfo:", nrow(geneInfo), "\n")
```

# Select Most Abundant Transcript Per Gene

```{r load_salmon}
# Get list of salmon quant files
salmon_dir <- here("2025_MDAcells_riboseq", "salmon")
salmon_samples <- list.dirs(salmon_dir, recursive = FALSE, full.names = FALSE)

# Build file paths
salmon_files <- file.path(salmon_dir, salmon_samples, "quant.sf.gz")
names(salmon_files) <- salmon_samples

# Filter to files that exist
salmon_files <- salmon_files[file.exists(salmon_files)]
cat("Found", length(salmon_files), "Salmon quant files\n")

# Import with tximport (transcript-level)
txi_transcripts <- tximport(salmon_files, type = "salmon", tx2gene = tx2gene, txOut = TRUE)

# Get transcript abundances and find most abundant transcript per gene
transcript_abundance <- txi_transcripts$abundance %>%
  as.data.frame() %>%
  rownames_to_column(var = "transcript_id") %>%
  left_join(tx2gene, by = "transcript_id")

# Sum counts across all samples and select most abundant transcript per gene
most_abundant_tx <- transcript_abundance %>%
  mutate(total_abundance = rowSums(across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::select(transcript_id, gene_id, total_abundance) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    # Strip version numbers from transcript IDs for joining with TxDb data
    transcript_id_clean = sub("\\..*", "", transcript_id)
  ) %>%
  group_by(gene_id_clean) %>%
  dplyr::slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

# Add gene symbols
most_abundant_tx <- most_abundant_tx %>%
  left_join(gene_symbol_map %>% mutate(gene_id_clean = sub("\\..*", "", gene_id)),
            by = "gene_id_clean")

cat("Selected most abundant transcript for", nrow(most_abundant_tx), "genes (all)\n")

# Filter to only genes with TE data (reduces from ~50k to ~12k)
te_gene_ids <- te_1hr %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  pull(gene_id_clean) %>%
  unique()

most_abundant_tx <- most_abundant_tx %>%
  filter(gene_id_clean %in% te_gene_ids)

cat("After filtering to genes with TE data:", nrow(most_abundant_tx), "genes\n")

# Check transcript ID formats
cat("\nExample transcript IDs from Salmon:", head(most_abundant_tx$transcript_id, 3), "\n")
cat("Example transcript IDs (cleaned):", head(most_abundant_tx$transcript_id_clean, 3), "\n")
```

# Extract Region Lengths

## 5' UTR, CDS, and 3' UTR Widths

```{r extract_regions}
# Extract regions from TxDb
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cds <- cdsBy(txdb, by = "tx", use.names = TRUE)

# Calculate widths
# Strip version numbers from TxDb transcript IDs to match Salmon IDs

head(utr5_widths)

# utr5_widths <- sum(width(utr5)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   dplyr::mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, utr5_length = `.`)

utr5_widths <- as.data.frame(utr5) %>%
  group_by(group_name) %>%
  summarise(
    utr5_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  dplyr::select(transcript_id_clean, utr5_length)

# utr3_widths <- sum(width(utr3)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, utr3_length = `.`)

# cds_widths <- sum(width(cds)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, cds_length = `.`)

utr3_widths <- as.data.frame(utr3) %>%
  group_by(group_name) %>%
  summarise(
    utr3_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  select(transcript_id_clean, utr3_length)

cds_widths <- as.data.frame(cds) %>%
  group_by(group_name) %>%
  summarise(
    cds_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  select(transcript_id_clean, cds_length)

cat("Extracted region widths:\n")
cat("5' UTRs:", nrow(utr5_widths), "\n")
cat("3' UTRs:", nrow(utr3_widths), "\n")
cat("CDS:", nrow(cds_widths), "\n")
cat("\nExample TxDb transcript IDs (cleaned):", head(utr5_widths$transcript_id_clean, 3), "\n")
```

# Extract Sequences

```{r extract_sequences}
# Extract 5' UTR sequences
# Strip version numbers from TxDb transcript IDs to match Salmon IDs
utr5_sequences <- extractTranscriptSeqs(genome, utr5) %>% as.data.frame()
utr5_seq <- utr5_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr5_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr5_sequence)

# Extract 3' UTR sequences
utr3_sequences <- extractTranscriptSeqs(genome, utr3) %>% as.data.frame()
utr3_seq <- utr3_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr3_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr3_sequence)

# Extract CDS sequences
cds_sequences <- extractTranscriptSeqs(genome, cds) %>% as.data.frame()
cds_seq <- cds_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("cds_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, cds_sequence)

cat("Extracted sequences:\n")
cat("5' UTRs:", nrow(utr5_seq), "\n")
cat("3' UTRs:", nrow(utr3_seq), "\n")
cat("CDS:", nrow(cds_seq), "\n")
```

# Calculate GC Content

```{r calculate_gc}
# GC content function
gc_content <- function(seq) {
  seq <- str_to_upper(seq)
  gc_count <- str_count(seq, "[GC]")
  total_length <- str_length(seq)
  result <- (gc_count / total_length) * 100
  result[total_length == 0] <- NA
  return(result)
}

# Calculate GC content for each region
utr5_gc <- utr5_seq %>%
  mutate(utr5_gc = gc_content(utr5_sequence)) %>%
  dplyr::select(transcript_id_clean, utr5_gc)

utr3_gc <- utr3_seq %>%
  mutate(utr3_gc = gc_content(utr3_sequence)) %>%
  dplyr::select(transcript_id_clean, utr3_gc)

cds_gc <- cds_seq %>%
  mutate(cds_gc = gc_content(cds_sequence)) %>%
  dplyr::select(transcript_id_clean, cds_gc)

cat("Calculated GC content for all regions\n")
```

# Calculate 4-mer Frequencies in 5' UTR

```{r calculate_4mers}
# Get transcripts that we'll use (those with 5' UTR sequences)
# Use transcript_id_clean for joining with TxDb data
valid_transcripts <- most_abundant_tx %>%
  filter(transcript_id_clean %in% utr5_seq$transcript_id_clean)

cat("Transcripts with 5' UTR data:", nrow(valid_transcripts), "out of", nrow(most_abundant_tx), "\n")

# Create DNAStringSet for 5' UTR sequences
utr5_for_kmers <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "")

# Filter out very short sequences (need at least 4nt for 4-mers)
utr5_for_kmers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 4)

utr5_dna <- DNAStringSet(utr5_for_kmers$utr5_sequence)
names(utr5_dna) <- utr5_for_kmers$transcript_id_clean

# Calculate 4-mer frequencies (as proportions)
kmer4_matrix <- oligonucleotideFrequency(utr5_dna, width = 4, as.prob = TRUE)

# Convert to data frame
kmer4_df <- as.data.frame(kmer4_matrix) %>%
  mutate(transcript_id_clean = utr5_for_kmers$transcript_id_clean)

# Add prefix to column names for clarity
colnames(kmer4_df)[1:256] <- paste0("utr5_4mer_", colnames(kmer4_df)[1:256])

cat("Calculated 4-mer frequencies for", nrow(kmer4_df), "transcripts\n")
cat("Number of 4-mer features:", 256, "\n")
```

# Calculate GC3 Content (Third Codon Position GC)

```{r calculate_gc3}
# Function to calculate GC3 content
calculate_gc3 <- function(cds_sequence) {
  if (is.na(cds_sequence) || nchar(cds_sequence) < 3) {
    return(NA)
  }

  # Split into codons
  seq <- str_to_upper(cds_sequence)
  seq_len <- nchar(seq)

  # Only use complete codons
  complete_codons <- seq_len %/% 3
  if (complete_codons == 0) return(NA)

  # Extract third positions
  third_positions <- sapply(1:complete_codons, function(i) {
    substr(seq, (i-1)*3 + 3, (i-1)*3 + 3)
  })

  # Calculate GC percentage at third position
  gc_count <- sum(third_positions %in% c("G", "C"))
  gc3 <- (gc_count / length(third_positions)) * 100

  return(gc3)
}

# Calculate GC3 for all CDS sequences
cds_gc3 <- cds_seq %>%
  mutate(gc3 = sapply(cds_sequence, calculate_gc3)) %>%
  dplyr::select(transcript_id_clean, gc3)

cat("Calculated GC3 content\n")
cat("Median GC3:", round(median(cds_gc3$gc3, na.rm = TRUE), 2), "%\n")
```

# Calculate RBP Motif Scores

```{r rbp_motifs}
# Use FeatureReachR to calculate PWM scores for RBNS motifs

# Prepare 5' UTR sequences as DNAStringSet
utr5_for_motifs <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "",
         str_length(utr5_sequence) >= 6)

utr5_motif_dna <- DNAStringSet(utr5_for_motifs$utr5_sequence)
names(utr5_motif_dna) <- utr5_for_motifs$transcript_id_clean

# Calculate RBNS PWM match counts
# Using a simplified approach - count motif occurrences per kb
cat("Calculating RBP motif scores...\n")

# Get RBNS PWM names
rbns_names <- names(RBNS_PWM)
cat("Number of RBNS motifs:", length(rbns_names), "\n")

# Function to count PWM matches in sequences
count_pwm_matches <- function(pwm, sequences, min_score = 0.8) {
  # Convert to matrix if needed
  pwm_matrix <- as.matrix(pwm)

  # Count matches per sequence
  counts <- sapply(sequences, function(seq) {
    if (length(seq) < ncol(pwm_matrix)) return(0)

    tryCatch({
      matches <- matchPWM(pwm_matrix, seq, min.score = paste0(min_score * 100, "%"))
      length(matches)
    }, error = function(e) 0)
  })

  return(counts)
}

# Calculate motif counts for a subset of key RBPs
# Focus on RBPs known to be relevant to hypoxia/translation
key_rbps <- c("PCBP1", "PCBP2", "PTBP1", "HNRNPA2B1", "ELAVL1", "SRSF1", "SRSF2",
              "SRSF3", "SRSF7", "KHDRBS1", "KHSRP", "CELF1", "MBNL1", "QKI",
              "RBM4", "RBM24", "IGF2BP1", "IGF2BP2", "IGF2BP3", "FMR1", "FXR1",
              "FXR2", "CPEB1", "CPEB4", "TARDBP", "FUS", "HNRNPC", "HNRNPK",
              "HNRNPL", "HNRNPM", "U2AF2", "TIA1", "TIAL1", "ZFP36", "YBX1")

# Find matching RBNS PWMs
available_rbps <- rbns_names[sapply(key_rbps, function(rbp) {
  any(str_detect(rbns_names, fixed(rbp, ignore_case = TRUE)))
})]

cat("Key RBPs with available RBNS PWMs:", length(available_rbps), "\n")

# Initialize motif score dataframe
motif_scores <- data.frame(transcript_id_clean = utr5_for_motifs$transcript_id_clean)

# Calculate scores for each available RBP
for (rbp_pattern in key_rbps) {
  matching_pwms <- rbns_names[str_detect(rbns_names, fixed(rbp_pattern, ignore_case = TRUE))]

  if (length(matching_pwms) > 0) {
    # Use first matching PWM
    pwm_name <- matching_pwms[1]
    pwm <- RBNS_PWM[[pwm_name]]

    # Count matches
    counts <- count_pwm_matches(pwm, utr5_motif_dna)

    # Normalize by sequence length (per kb)
    seq_lengths <- width(utr5_motif_dna)
    normalized_counts <- (counts / seq_lengths) * 1000

    # Add to dataframe
    col_name <- paste0("rbp_", gsub("[^A-Za-z0-9]", "_", rbp_pattern))
    motif_scores[[col_name]] <- normalized_counts

    cat("Processed:", rbp_pattern, "(", pwm_name, ")\n")
  }
}

cat("Calculated motif scores for", ncol(motif_scores) - 1, "RBPs\n")
```

# Assemble Feature Matrix

```{r assemble_features}
# Start with most abundant transcripts and TE data
# Use transcript_id_clean for all joins with TxDb-derived data
head(feature_matrix)


feature_matrix <- most_abundant_tx %>%
  dplyr::select(transcript_id, transcript_id_clean, gene_id = gene_id.x, symbol) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", transcript_id),
    gene_id_clean       = sub("\\..*", "", gene_id)) |> 
  # Add TE data (join by gene_id)
  left_join(te_1hr %>%
              mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
              dplyr::select(gene_id_clean, te_lfc, te_padj),
            by = "gene_id_clean") %>%
  # Add region lengths (join by transcript_id_clean)
  left_join(utr5_widths, by = "transcript_id_clean") %>%
  left_join(utr3_widths, by = "transcript_id_clean") %>%
  left_join(cds_widths, by = "transcript_id_clean") %>%
  # Add log2 transformed lengths
  mutate(
    log2_utr5_length = log2(utr5_length + 1),
    log2_utr3_length = log2(utr3_length + 1),
    log2_cds_length = log2(cds_length + 1),
    total_length = utr5_length + cds_length + utr3_length,
    log2_total_length = log2(total_length + 1)
  ) %>%
  # Add GC content (join by transcript_id_clean)
  left_join(utr5_gc, by = "transcript_id_clean") %>%
  left_join(utr3_gc, by = "transcript_id_clean") %>%
  left_join(cds_gc, by = "transcript_id_clean") %>%
  # Calculate total transcript GC (weighted average)
  mutate(
    transcript_gc = (utr5_gc * utr5_length + cds_gc * cds_length + utr3_gc * utr3_length) /
                    (utr5_length + cds_length + utr3_length)
  ) %>%
  # Add GC3 (join by transcript_id_clean)
  left_join(cds_gc3, by = "transcript_id_clean") %>%
  # Add 4-mer frequencies (join by transcript_id_clean)
  left_join(kmer4_df, by = "transcript_id_clean") %>%
  # Add RBP motif scores (join by transcript_id_clean)
  left_join(motif_scores, by = "transcript_id_clean")

# Check join success
cat("\nJoin diagnostics:\n")
cat("Total transcripts:", nrow(most_abundant_tx), "\n")
cat("With 5' UTR length:", sum(!is.na(feature_matrix$utr5_length)), "\n")
cat("With CDS length:", sum(!is.na(feature_matrix$cds_length)), "\n")
cat("With 3' UTR length:", sum(!is.na(feature_matrix$utr3_length)), "\n")
cat("With TE data:", sum(!is.na(feature_matrix$te_lfc)), "\n")

# Filter to genes with TE data and valid features
feature_matrix <- feature_matrix %>%
  filter(!is.na(te_lfc),
         !is.na(utr5_length),
         !is.na(cds_length))

cat("\nFeature matrix assembled:\n")
cat("Genes:", nrow(feature_matrix), "\n")
cat("Features:", ncol(feature_matrix) - 4, "(excluding gene_id, transcript_id, symbol, te_lfc, te_padj)\n")
```

# Quality Control

```{r quality_control}
# Check for missing values
missing_summary <- feature_matrix %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("\nFeatures with missing values:\n")
print(missing_summary %>% head(20))

# Summary of key features
cat("\n\nFeature summaries:\n")
cat("\n5' UTR length:\n")
print(summary(feature_matrix$utr5_length))

cat("\n3' UTR length:\n")
print(summary(feature_matrix$utr3_length))

cat("\nCDS length:\n")
print(summary(feature_matrix$cds_length))

cat("\n5' UTR GC content:\n")
print(summary(feature_matrix$utr5_gc))

cat("\nTE log2 fold change (target):\n")
print(summary(feature_matrix$te_lfc))
```

# Visualize Feature Distributions

```{r visualize_distributions, fig.width=12, fig.height=10}
# Create histogram of key features
p1 <- ggplot(feature_matrix, aes(x = log2_utr5_length)) +
  geom_histogram(bins = 50, fill = "#0072B2", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "5' UTR Length (log2)", x = "log2(length + 1)", y = "Count")

p2 <- ggplot(feature_matrix, aes(x = log2_cds_length)) +
  geom_histogram(bins = 50, fill = "#E69F00", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "CDS Length (log2)", x = "log2(length + 1)", y = "Count")

p3 <- ggplot(feature_matrix, aes(x = log2_utr3_length)) +
  geom_histogram(bins = 50, fill = "#009E73", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "3' UTR Length (log2)", x = "log2(length + 1)", y = "Count")

p4 <- ggplot(feature_matrix, aes(x = utr5_gc)) +
  geom_histogram(bins = 50, fill = "#CC79A7", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "5' UTR GC Content", x = "GC %", y = "Count")

p5 <- ggplot(feature_matrix, aes(x = cds_gc)) +
  geom_histogram(bins = 50, fill = "#D55E00", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "CDS GC Content", x = "GC %", y = "Count")

p6 <- ggplot(feature_matrix, aes(x = te_lfc)) +
  geom_histogram(bins = 50, fill = "#56B4E9", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_cowplot() +
  labs(title = "TE Log2 Fold Change (Target)", x = "TE LFC", y = "Count")

combined_plot <- plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
print(combined_plot)

ggsave(here("plots", "feature_distributions_predictive_model.pdf"),
       plot = combined_plot, width = 12, height = 10)
```

# Create Binned TE Category for Stratified Sampling

```{r create_bins}
# Create bins for stratified train/test split
feature_matrix <- feature_matrix %>%
  mutate(te_lfc_bin = cut(te_lfc,
                          breaks = quantile(te_lfc, probs = seq(0, 1, 0.2), na.rm = TRUE),
                          include.lowest = TRUE,
                          labels = c("Q1", "Q2", "Q3", "Q4", "Q5")))

cat("\nTE LFC bins for stratified sampling:\n")
print(table(feature_matrix$te_lfc_bin))
```

# Export Feature Matrix

```{r export_features}
# Create output directory if needed
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

# Export as CSV
write_csv(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr.csv\n")

# Export as RDS for faster loading
saveRDS(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr.rds\n")

# Summary
cat("\n\nFinal Feature Matrix Summary:\n")
cat("Total genes:", nrow(feature_matrix), "\n")
cat("Total features:", ncol(feature_matrix), "\n")
cat("  - Identifiers: gene_id, transcript_id, symbol\n")
cat("  - Target: te_lfc, te_padj, te_lfc_bin\n")
cat("  - Region lengths: 7 features\n")
cat("  - GC content: 5 features\n")
cat("  - 4-mer frequencies: 256 features\n")
cat("  - RBP motif scores:", ncol(motif_scores) - 1, "features\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
