---
title: "Feature Extraction for Hypoxia Response Prediction"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis extracts mRNA sequence and structural features for all genes to build a predictive model of Translation Efficiency (TE) Log2 Fold Change at 1 hour hypoxia.

## Feature Categories

| Category | Features | Count |
|----------|----------|-------|
| **Region Lengths** | 5'UTR, CDS, 3'UTR lengths + log2 transforms | 7 |
| **GC Content** | GC% for 5'UTR, CDS, 3'UTR, full transcript | 4 |
| **K-mer Frequencies** | 4-mers in 5'UTR | 256 |
| **K-mer Frequencies** | 5-mers in 5'UTR | 1,024 |
| **K-mer Frequencies** | 6-mers in 5'UTR | 4,096 |
| **RBP Motifs** | RBNS + CISBP-RNA PWM scores (all motifs, 5'UTR/CDS/3'UTR) | ~900 |
| **Codon Usage** | GC3 content | 1 |
| **4E Resistance** | Teleman riboseq 4E-resistant/sensitive status | 2 |
| **DAP5 Sensitivity** | Weber 2022 DAP5 KD up/down status | 2 |
| **Translation Signatures** | eIF3d/3e signature membership (lfc0.5) | 34 |
| **eIF3 CLIP Binding** | Bound/regional binding per CLIP dataset | ~28 |

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(tximport)
library(scales)
library(FeatureReachR)
library(readxl)
```

# Load Data

## Gene Annotation

```{r load_gene_annotation}
# Load gene symbols from annotation file
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

# Keep one entry per gene (prefer protein_coding, then deduplicate)
gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

## TE Data (1hr Hypoxia Response)

```{r load_te_data}
# Load TE changes from hypoxia timecourse at 1hr
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))

cat("Loaded TE data:", nrow(te_1hr), "genes\n")

# Add gene symbols
te_1hr <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  filter(!is.na(symbol))

cat("TE data with symbols:", nrow(te_1hr), "genes\n")
```

## Load TxDb and Genome

```{r load_txdb}
# Load pre-built TxDb from Gencode v49
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))

# Load transcript-to-gene mapping (v49, human only)
tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript_id", "gene_id")) %>%
  filter(str_starts(transcript_id, "ENST"))

# Reuse v49 gene annotation already loaded above
gene_symbol_map <- gene_symbols

# Load genome
genome <- BSgenome.Hsapiens.UCSC.hg38

cat("Loaded TxDb and genome\n")
cat("Transcripts in tx2gene:", nrow(tx2gene), "\n")
```

# Select Most Abundant Transcript Per Gene

```{r load_salmon}
# Get list of salmon quant files
salmon_dir <- here("2025_MDAcells_riboseq", "salmon")
salmon_samples <- list.dirs(salmon_dir, recursive = FALSE, full.names = FALSE)

# Build file paths
salmon_files <- file.path(salmon_dir, salmon_samples, "quant.sf.gz")
names(salmon_files) <- salmon_samples

# Filter to files that exist
salmon_files <- salmon_files[file.exists(salmon_files)]
cat("Found", length(salmon_files), "Salmon quant files\n")

# Import with tximport (transcript-level)
txi_transcripts <- tximport(salmon_files, type = "salmon", tx2gene = tx2gene, txOut = TRUE)

# Get transcript abundances and find most abundant transcript per gene
transcript_abundance <- txi_transcripts$abundance %>%
  as.data.frame() %>%
  rownames_to_column(var = "transcript_id") %>%
  left_join(tx2gene, by = "transcript_id")

# Sum counts across all samples and select most abundant transcript per gene
most_abundant_tx <- transcript_abundance %>%
  mutate(total_abundance = rowSums(across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::select(transcript_id, gene_id, total_abundance) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    # Strip version numbers from transcript IDs for joining with TxDb data
    transcript_id_clean = sub("\\..*", "", transcript_id)
  ) %>%
  group_by(gene_id_clean) %>%
  dplyr::slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

# Add gene symbols
most_abundant_tx <- most_abundant_tx %>%
  left_join(gene_symbol_map %>% mutate(gene_id_clean = sub("\\..*", "", gene_id)),
            by = "gene_id_clean")

cat("Selected most abundant transcript for", nrow(most_abundant_tx), "genes (all)\n")

# Filter to only genes with TE data (reduces from ~50k to ~12k)
te_gene_ids <- te_1hr %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  pull(gene_id_clean) %>%
  unique()

most_abundant_tx <- most_abundant_tx %>%
  filter(gene_id_clean %in% te_gene_ids)

cat("After filtering to genes with TE data:", nrow(most_abundant_tx), "genes\n")

# Check transcript ID formats
cat("\nExample transcript IDs from Salmon:", head(most_abundant_tx$transcript_id, 3), "\n")
cat("Example transcript IDs (cleaned):", head(most_abundant_tx$transcript_id_clean, 3), "\n")
```

# Extract Region Lengths

## 5' UTR, CDS, and 3' UTR Widths

```{r extract_regions}
# Extract regions from TxDb
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cds <- cdsBy(txdb, by = "tx", use.names = TRUE)

# Calculate widths
# Strip version numbers from TxDb transcript IDs to match Salmon IDs

# utr5_widths <- sum(width(utr5)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   dplyr::mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, utr5_length = `.`)

utr5_widths <- as.data.frame(utr5) %>%
  group_by(group_name) %>%
  summarise(
    utr5_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  dplyr::select(transcript_id_clean, utr5_length)

# utr3_widths <- sum(width(utr3)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, utr3_length = `.`)

# cds_widths <- sum(width(cds)) %>%
#   as.data.frame() %>%
#   rownames_to_column("transcript_id_raw") %>%
#   mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
#   dplyr::select(transcript_id_clean, cds_length = `.`)

utr3_widths <- as.data.frame(utr3) %>%
  group_by(group_name) %>%
  summarise(
    utr3_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  dplyr::select(transcript_id_clean, utr3_length)

cds_widths <- as.data.frame(cds) %>%
  group_by(group_name) %>%
  summarise(
    cds_length = sum(width),
    .groups = "drop"
  ) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", group_name)
  ) %>%
  dplyr::select(transcript_id_clean, cds_length)

cat("Extracted region widths:\n")
cat("5' UTRs:", nrow(utr5_widths), "\n")
cat("3' UTRs:", nrow(utr3_widths), "\n")
cat("CDS:", nrow(cds_widths), "\n")
cat("\nExample TxDb transcript IDs (cleaned):", head(utr5_widths$transcript_id_clean, 3), "\n")
```

# Extract Sequences

```{r extract_sequences}
# Extract 5' UTR sequences
# Strip version numbers from TxDb transcript IDs to match Salmon IDs
utr5_sequences <- extractTranscriptSeqs(genome, utr5) %>% as.data.frame()
utr5_seq <- utr5_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr5_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr5_sequence)

# Extract 3' UTR sequences
utr3_sequences <- extractTranscriptSeqs(genome, utr3) %>% as.data.frame()
utr3_seq <- utr3_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr3_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr3_sequence)

# Extract CDS sequences
cds_sequences <- extractTranscriptSeqs(genome, cds) %>% as.data.frame()
cds_seq <- cds_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("cds_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, cds_sequence)

cat("Extracted sequences:\n")
cat("5' UTRs:", nrow(utr5_seq), "\n")
cat("3' UTRs:", nrow(utr3_seq), "\n")
cat("CDS:", nrow(cds_seq), "\n")
```

# Calculate GC Content

```{r calculate_gc}
# GC content function
gc_content <- function(seq) {
  seq <- str_to_upper(seq)
  gc_count <- str_count(seq, "[GC]")
  total_length <- str_length(seq)
  result <- (gc_count / total_length) * 100
  result[total_length == 0] <- NA
  return(result)
}

# Calculate GC content for each region
utr5_gc <- utr5_seq %>%
  mutate(utr5_gc = gc_content(utr5_sequence)) %>%
  dplyr::select(transcript_id_clean, utr5_gc)

utr3_gc <- utr3_seq %>%
  mutate(utr3_gc = gc_content(utr3_sequence)) %>%
  dplyr::select(transcript_id_clean, utr3_gc)

cds_gc <- cds_seq %>%
  mutate(cds_gc = gc_content(cds_sequence)) %>%
  dplyr::select(transcript_id_clean, cds_gc)

cat("Calculated GC content for all regions\n")
```

# Calculate 4-mer Frequencies in 5' UTR

```{r calculate_4mers}
# Get transcripts that we'll use (those with 5' UTR sequences)
# Use transcript_id_clean for joining with TxDb data
valid_transcripts <- most_abundant_tx %>%
  filter(transcript_id_clean %in% utr5_seq$transcript_id_clean)

cat("Transcripts with 5' UTR data:", nrow(valid_transcripts), "out of", nrow(most_abundant_tx), "\n")

# Create DNAStringSet for 5' UTR sequences
utr5_for_kmers <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "")

# Filter out very short sequences (need at least 4nt for 4-mers)
utr5_for_kmers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 4)

utr5_dna <- DNAStringSet(utr5_for_kmers$utr5_sequence)
names(utr5_dna) <- utr5_for_kmers$transcript_id_clean

# Calculate 4-mer frequencies (as proportions)
kmer4_matrix <- oligonucleotideFrequency(utr5_dna, width = 4, as.prob = TRUE)

# Convert to data frame
kmer4_df <- as.data.frame(kmer4_matrix) %>%
  mutate(transcript_id_clean = utr5_for_kmers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

# Add prefix to column names for clarity
colnames(kmer4_df)[1:256] <- paste0("utr5_4mer_", colnames(kmer4_df)[1:256])

cat("Calculated 4-mer frequencies for", nrow(kmer4_df), "transcripts\n")
cat("Number of 4-mer features:", 256, "\n")
```

# Calculate 5-mer Frequencies in 5' UTR

```{r calculate_5mers}
# Filter sequences with at least 5nt
utr5_for_5mers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 5)

utr5_dna_5mer <- DNAStringSet(utr5_for_5mers$utr5_sequence)
names(utr5_dna_5mer) <- utr5_for_5mers$transcript_id_clean

# Calculate 5-mer frequencies (as proportions)
kmer5_matrix <- oligonucleotideFrequency(utr5_dna_5mer, width = 5, as.prob = TRUE)

# Convert to data frame
kmer5_df <- as.data.frame(kmer5_matrix) %>%
  mutate(transcript_id_clean = utr5_for_5mers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

# Add prefix to column names for clarity
colnames(kmer5_df)[1:1024] <- paste0("utr5_5mer_", colnames(kmer5_df)[1:1024])

cat("Calculated 5-mer frequencies for", nrow(kmer5_df), "transcripts\n")
cat("Number of 5-mer features:", 1024, "\n")
```

# Calculate 6-mer Frequencies in 5' UTR

```{r calculate_6mers}
# Filter sequences with at least 6nt
utr5_for_6mers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 6)

utr5_dna_6mer <- DNAStringSet(utr5_for_6mers$utr5_sequence)
names(utr5_dna_6mer) <- utr5_for_6mers$transcript_id_clean

# Calculate 6-mer frequencies (as proportions)
kmer6_matrix <- oligonucleotideFrequency(utr5_dna_6mer, width = 6, as.prob = TRUE)

# Convert to data frame
kmer6_df <- as.data.frame(kmer6_matrix) %>%
  mutate(transcript_id_clean = utr5_for_6mers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

# Add prefix to column names for clarity
colnames(kmer6_df)[1:4096] <- paste0("utr5_6mer_", colnames(kmer6_df)[1:4096])

cat("Calculated 6-mer frequencies for", nrow(kmer6_df), "transcripts\n")
cat("Number of 6-mer features:", 4096, "\n")
```

# Calculate GC3 Content (Third Codon Position GC)

```{r calculate_gc3}
# Function to calculate GC3 content
calculate_gc3 <- function(cds_sequence) {
  if (is.na(cds_sequence) || nchar(cds_sequence) < 3) {
    return(NA)
  }

  # Split into codons
  seq <- str_to_upper(cds_sequence)
  seq_len <- nchar(seq)

  # Only use complete codons
  complete_codons <- seq_len %/% 3
  if (complete_codons == 0) return(NA)

  # Extract third positions
  third_positions <- sapply(1:complete_codons, function(i) {
    substr(seq, (i-1)*3 + 3, (i-1)*3 + 3)
  })

  # Calculate GC percentage at third position
  gc_count <- sum(third_positions %in% c("G", "C"))
  gc3 <- (gc_count / length(third_positions)) * 100

  return(gc3)
}

# Calculate GC3 for all CDS sequences
cds_gc3 <- cds_seq %>%
  mutate(gc3 = sapply(cds_sequence, calculate_gc3)) %>%
  dplyr::select(transcript_id_clean, gc3)

cat("Calculated GC3 content\n")
cat("Median GC3:", round(median(cds_gc3$gc3, na.rm = TRUE), 2), "%\n")
```

# 4E Resistance & DAP5 Sensitivity

```{r 4e_dap5}
# Load Teleman riboseq data (4E-BP1 overexpression)
teleman_data <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv") %>%
  dplyr::filter(!is.na(symbol)) %>%
  dplyr::group_by(symbol) %>%
  dplyr::slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

resistant_4e_symbols <- teleman_data %>% dplyr::filter(log2FoldChange > 1, padj < 0.05) %>% dplyr::pull(symbol)
sensitive_4e_symbols <- teleman_data %>% dplyr::filter(log2FoldChange < -1, padj < 0.05) %>% dplyr::pull(symbol)

cat("4E-resistant genes:", length(resistant_4e_symbols), "\n")
cat("4E-sensitive genes:", length(sensitive_4e_symbols), "\n")

# Load DAP5 data (Weber 2022)
dap5_data <- read_xlsx(here("accessories", "dap5kd_riboseq_weber2022.xlsx")) %>%
  dplyr::filter(!is.na(geneIDs), comment != "no")

dap5_repressed_symbols <- dap5_data %>% dplyr::filter(comment == "up") %>% dplyr::pull(geneIDs)
dap5_promoted_symbols <- dap5_data %>% dplyr::filter(comment == "down") %>% dplyr::pull(geneIDs)

cat("DAP5-repressed genes (up with KD):", length(dap5_repressed_symbols), "\n")
cat("DAP5-promoted genes (down with KD):", length(dap5_promoted_symbols), "\n")
```

# Translation Signature Membership

```{r signature_membership}
# Load all TE signatures
all_sigs <- read_csv(here("output", "genesets", "all_te_signatures.csv"))

# Select lfc0.5 signatures for feature extraction
selected_sig_names <- c(
  # Hypoxia eIF3d per timepoint (lfc0.5)
  "hypoxia_3d_promotes_TE_1hr_lfc0.5", "hypoxia_3d_inhibits_TE_1hr_lfc0.5",
  "hypoxia_3d_promotes_TE_4hr_lfc0.5", "hypoxia_3d_inhibits_TE_4hr_lfc0.5",
  "hypoxia_3d_promotes_TE_24hr_lfc0.5", "hypoxia_3d_inhibits_TE_24hr_lfc0.5",
  # Hypoxia eIF3e per timepoint (lfc0.5)
  "hypoxia_3e_promotes_TE_1hr_lfc0.5", "hypoxia_3e_inhibits_TE_1hr_lfc0.5",
  "hypoxia_3e_promotes_TE_4hr_lfc0.5", "hypoxia_3e_inhibits_TE_4hr_lfc0.5",
  "hypoxia_3e_promotes_TE_24hr_lfc0.5", "hypoxia_3e_inhibits_TE_24hr_lfc0.5",
  # Normoxia eIF3d/3e combined (lfc0.5)
  "normoxia_3d_promotes_TE_combined_lfc0.5", "normoxia_3d_inhibits_TE_combined_lfc0.5",
  "normoxia_3e_promotes_TE_combined_lfc0.5", "normoxia_3e_inhibits_TE_combined_lfc0.5",
  # Hypoxia response (lfc0.5)
  "hypoxia_response_increases_TE_1hr_lfc0.5", "hypoxia_response_decreases_TE_1hr_lfc0.5",
  "hypoxia_response_increases_TE_4hr_lfc0.5", "hypoxia_response_decreases_TE_4hr_lfc0.5",
  "hypoxia_response_increases_TE_24hr_lfc0.5", "hypoxia_response_decreases_TE_24hr_lfc0.5",
  # Hypoxia response inhibited_by (lfc0.5)
  "hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_1hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_1hr_inhibited_by_3e_lfc0.5",
  "hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_4hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_4hr_inhibited_by_3e_lfc0.5",
  "hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_24hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_24hr_inhibited_by_3e_lfc0.5"
)

# Filter to selected signatures and pivot to wide binary format
sig_features <- all_sigs %>%
  dplyr::filter(gs_name %in% selected_sig_names) %>%
  dplyr::distinct(db_gene_symbol, gs_name) %>%
  dplyr::mutate(member = 1L) %>%
  tidyr::pivot_wider(
    names_from = gs_name,
    values_from = member,
    values_fill = 0L,
    names_prefix = "sig_"
  ) %>%
  dplyr::rename(symbol = db_gene_symbol)

cat("Signature features created:", ncol(sig_features) - 1, "signatures\n")
cat("Genes with any signature membership:", nrow(sig_features), "\n")
```

# eIF3 CLIP Binding Features

```{r clip_binding}
# Load CLIP score card data
# Each CLIP column contains a string encoding binding status and region:
#   "not bound", "5UTR", "CDS", "3UTRs", "5UTR & CDS", "3UTRs, cds, 5UTRs",
#   "cap_binder", "other", etc.
# We parse these into binary bound + regional features per condition.
# Note: lee_int_stress, lee_complete_media, lee_glu_dep only have "not bound"
#   and "cap_binder" — no regional info, so only _bound will have 1s for those.
clip_data <- read_csv("/Users/katematlin/github/mpra_preliminary/score_card.csv")

# Define CLIP conditions
clip_conditions <- c("lee_parclip_a_b_d_g", "lee_parclip_d_only", "cate_undiff_npcs",
                      "cate_diff_npcs", "lee_int_stress", "lee_complete_media", "lee_glu_dep")

# Create gene_id_clean for joining
clip_data <- clip_data %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

# Build CLIP features: bound status and regional binding per condition
clip_features <- clip_data %>%
  dplyr::select(gene_id_clean) %>%
  dplyr::distinct()

for (condition in clip_conditions) {
  bound_col <- paste0("clip_", condition, "_bound")
  utr5_col <- paste0("clip_", condition, "_5utr")
  cds_col <- paste0("clip_", condition, "_cds")
  utr3_col <- paste0("clip_", condition, "_3utr")

  # Extract the string column for this condition and parse regions
  # "cap_binder" → bound only (no region assigned)
  # "other" → bound only (no region assigned)
  condition_data <- clip_data %>%
    dplyr::select(gene_id_clean, value = all_of(condition)) %>%
    dplyr::mutate(
      value_lower = tolower(value),
      !!bound_col := as.integer(value_lower != "not bound"),
      !!utr5_col  := as.integer(str_detect(value_lower, "5utr")),
      !!cds_col   := as.integer(str_detect(value_lower, "cds")),
      !!utr3_col  := as.integer(str_detect(value_lower, "3utr"))
    ) %>%
    dplyr::select(gene_id_clean, all_of(c(bound_col, utr5_col, cds_col, utr3_col)))

  clip_features <- clip_features %>%
    dplyr::left_join(condition_data, by = "gene_id_clean")
}

# Fill any remaining NAs with 0
clip_features <- clip_features %>%
  dplyr::mutate(across(-gene_id_clean, ~as.integer(replace_na(., 0L))))

# Drop columns that are all zero (no informative signal)
all_zero_cols <- names(clip_features)[sapply(clip_features, function(x) is.numeric(x) && sum(x) == 0)]
if (length(all_zero_cols) > 0) {
  cat("Dropping", length(all_zero_cols), "all-zero CLIP columns:", paste(all_zero_cols, collapse = ", "), "\n")
  clip_features <- clip_features %>%
    dplyr::select(-all_of(all_zero_cols))
}

cat("CLIP features retained:", ncol(clip_features) - 1, "features\n")
cat("Genes with CLIP data:", nrow(clip_features), "\n")
```

# Calculate RBP Motif Scores (Unbiased - All Motifs, All Regions)

```{r rbp_motifs}
# Score all RBP motifs from both RBNS_PWM and CISBPRNA_hs_PWM databases
# across 5' UTR, CDS, and 3' UTR regions
# This unbiased approach lets the Random Forest determine which motifs are predictive

# --- Prepare sequences for each region ---

# 5' UTR
utr5_motif_seqs <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "",
         str_length(utr5_sequence) >= 6)

utr5_dna_motif <- DNAStringSet(utr5_motif_seqs$utr5_sequence)
names(utr5_dna_motif) <- utr5_motif_seqs$transcript_id_clean

# CDS
cds_motif_seqs <- cds_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(cds_sequence),
         cds_sequence != "",
         str_length(cds_sequence) >= 6)

cds_dna_motif <- DNAStringSet(cds_motif_seqs$cds_sequence)
names(cds_dna_motif) <- cds_motif_seqs$transcript_id_clean

# 3' UTR
utr3_motif_seqs <- utr3_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr3_sequence),
         utr3_sequence != "",
         str_length(utr3_sequence) >= 6)

utr3_dna_motif <- DNAStringSet(utr3_motif_seqs$utr3_sequence)
names(utr3_dna_motif) <- utr3_motif_seqs$transcript_id_clean

cat("Sequences for motif scoring:\n")
cat("  5' UTR:", length(utr5_dna_motif), "\n")
cat("  CDS:", length(cds_dna_motif), "\n")
cat("  3' UTR:", length(utr3_dna_motif), "\n")
cat("RBNS motifs available:", length(RBNS_PWM), "\n")
cat("CISBP-RNA motifs available:", length(CISBPRNA_hs_PWM), "\n")

# --- Helper function to score all motifs from a PWM list against a set of sequences ---
score_pwm_list <- function(pwm_list, sequences, db_prefix, region_prefix) {
  seq_lengths <- width(sequences)
  scores <- data.frame(transcript_id_clean = names(sequences))

  for (motif_name in names(pwm_list)) {
    pwm <- pwm_list[[motif_name]]
    pwm_matrix <- as.matrix(pwm)

    counts <- sapply(sequences, function(seq) {
      if (length(seq) < ncol(pwm_matrix)) return(0)
      tryCatch({
        length(matchPWM(pwm_matrix, seq, min.score = "80%"))
      }, error = function(e) 0)
    })

    col_name <- paste0(db_prefix, "_", region_prefix, "_", gsub("[^A-Za-z0-9]", "_", motif_name))
    scores[[col_name]] <- (counts / seq_lengths) * 1000
  }

  return(scores)
}

# --- Score RBNS motifs in all three regions ---
cat("Scoring RBNS motifs in 5' UTR...\n")
rbns_utr5 <- score_pwm_list(RBNS_PWM, utr5_dna_motif, "rbns", "utr5")
cat("Scoring RBNS motifs in CDS...\n")
rbns_cds <- score_pwm_list(RBNS_PWM, cds_dna_motif, "rbns", "cds")
cat("Scoring RBNS motifs in 3' UTR...\n")
rbns_utr3 <- score_pwm_list(RBNS_PWM, utr3_dna_motif, "rbns", "utr3")

cat("Scored", length(names(RBNS_PWM)), "RBNS motifs x 3 regions\n")

# --- Score CISBP-RNA motifs in all three regions ---
cat("Scoring CISBP-RNA motifs in 5' UTR...\n")
cisbp_utr5 <- score_pwm_list(CISBPRNA_hs_PWM, utr5_dna_motif, "cisbp", "utr5")
cat("Scoring CISBP-RNA motifs in CDS...\n")
cisbp_cds <- score_pwm_list(CISBPRNA_hs_PWM, cds_dna_motif, "cisbp", "cds")
cat("Scoring CISBP-RNA motifs in 3' UTR...\n")
cisbp_utr3 <- score_pwm_list(CISBPRNA_hs_PWM, utr3_dna_motif, "cisbp", "utr3")

cat("Scored", length(names(CISBPRNA_hs_PWM)), "CISBP-RNA motifs x 3 regions\n")

# --- Merge all motif scores ---
# Use full outer joins so transcripts missing a region get NA (filled later)
motif_scores <- rbns_utr5 %>%
  full_join(rbns_cds, by = "transcript_id_clean") %>%
  full_join(rbns_utr3, by = "transcript_id_clean") %>%
  full_join(cisbp_utr5, by = "transcript_id_clean") %>%
  full_join(cisbp_cds, by = "transcript_id_clean") %>%
  full_join(cisbp_utr3, by = "transcript_id_clean")

cat("Total RBP motif features:", ncol(motif_scores) - 1, "\n")
cat("  Per database: RBNS =", length(RBNS_PWM), ", CISBP-RNA =", length(CISBPRNA_hs_PWM), "\n")
cat("  Per region: 5' UTR, CDS, 3' UTR\n")
```

# Assemble Feature Matrix

```{r assemble_features}
# Start with most abundant transcripts and TE data
# Use transcript_id_clean for all joins with TxDb-derived data
feature_matrix <- most_abundant_tx %>%
  dplyr::select(transcript_id, transcript_id_clean, gene_id = gene_id.x, symbol) %>%
  mutate(
    transcript_id_clean = sub("\\..*", "", transcript_id),
    gene_id_clean       = sub("\\..*", "", gene_id)) |> 
  # Add TE data (join by gene_id)
  left_join(te_1hr %>%
              mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
              dplyr::select(gene_id_clean, te_lfc, te_padj),
            by = "gene_id_clean") %>%
  # Add region lengths (join by transcript_id_clean)
  left_join(utr5_widths, by = "transcript_id_clean") %>%
  left_join(utr3_widths, by = "transcript_id_clean") %>%
  left_join(cds_widths, by = "transcript_id_clean") %>%
  # Add log2 transformed lengths
  mutate(
    log2_utr5_length = log2(utr5_length + 1),
    log2_utr3_length = log2(utr3_length + 1),
    log2_cds_length = log2(cds_length + 1),
    total_length = utr5_length + cds_length + utr3_length,
    log2_total_length = log2(total_length + 1)
  ) %>%
  # Add GC content (join by transcript_id_clean)
  left_join(utr5_gc, by = "transcript_id_clean") %>%
  left_join(utr3_gc, by = "transcript_id_clean") %>%
  left_join(cds_gc, by = "transcript_id_clean") %>%
  # Calculate total transcript GC (weighted average)
  mutate(
    transcript_gc = (utr5_gc * utr5_length + cds_gc * cds_length + utr3_gc * utr3_length) /
                    (utr5_length + cds_length + utr3_length)
  ) %>%
  # Add GC3 (join by transcript_id_clean)
  left_join(cds_gc3, by = "transcript_id_clean") %>%
  # Add 4-mer frequencies (join by transcript_id_clean)
  left_join(kmer4_df, by = "transcript_id_clean") %>%
  # Add 5-mer frequencies (join by transcript_id_clean)
  left_join(kmer5_df, by = "transcript_id_clean") %>%
  # Add 6-mer frequencies (join by transcript_id_clean)
  left_join(kmer6_df, by = "transcript_id_clean") %>%
  # Add RBP motif scores (join by transcript_id_clean)
  left_join(motif_scores, by = "transcript_id_clean") %>%
  # Add 4E resistance and DAP5 features
  mutate(
    is_4e_resistant = as.integer(symbol %in% resistant_4e_symbols),
    is_4e_sensitive = as.integer(symbol %in% sensitive_4e_symbols),
    is_dap5_repressed = as.integer(symbol %in% dap5_repressed_symbols),
    is_dap5_promoted = as.integer(symbol %in% dap5_promoted_symbols)
  ) %>%
  # Add signature membership
  left_join(sig_features, by = "symbol") %>%
  # Add CLIP binding
  left_join(clip_features, by = "gene_id_clean")

# Fill NA features with 0
# K-mer frequencies: transcripts without a 5' UTR (or too short for a given k)
# have zero occurrences of every k-mer; utr5_length already captures the absence
feature_matrix <- feature_matrix %>%
  dplyr::mutate(across(starts_with("utr5_4mer_"), ~replace_na(., 0))) %>%
  dplyr::mutate(across(starts_with("utr5_5mer_"), ~replace_na(., 0))) %>%
  dplyr::mutate(across(starts_with("utr5_6mer_"), ~replace_na(., 0))) %>%
  dplyr::mutate(across(starts_with("sig_"), ~replace_na(., 0L))) %>%
  dplyr::mutate(across(starts_with("clip_"), ~replace_na(., 0L)))

# Check join success
cat("\nJoin diagnostics:\n")
cat("Total transcripts:", nrow(most_abundant_tx), "\n")
cat("With 5' UTR length:", sum(!is.na(feature_matrix$utr5_length)), "\n")
cat("With CDS length:", sum(!is.na(feature_matrix$cds_length)), "\n")
cat("With 3' UTR length:", sum(!is.na(feature_matrix$utr3_length)), "\n")
cat("With TE data:", sum(!is.na(feature_matrix$te_lfc)), "\n")

# Filter to genes with TE data and valid features
feature_matrix <- feature_matrix %>%
  filter(!is.na(te_lfc),
         !is.na(utr5_length),
         !is.na(cds_length))

# Drop any binary feature columns (sig_, clip_, is_) that are all zero after filtering
binary_prefixes <- c("sig_", "clip_", "is_")
binary_cols <- names(feature_matrix)[sapply(names(feature_matrix), function(n) any(str_starts(n, binary_prefixes)))]
all_zero_binary <- binary_cols[sapply(binary_cols, function(col) sum(feature_matrix[[col]]) == 0)]
if (length(all_zero_binary) > 0) {
  cat("Dropping", length(all_zero_binary), "all-zero binary columns:\n")
  cat(" ", paste(all_zero_binary, collapse = ", "), "\n")
  feature_matrix <- feature_matrix %>%
    dplyr::select(-all_of(all_zero_binary))
}

cat("\nFeature matrix assembled:\n")
cat("Genes:", nrow(feature_matrix), "\n")
cat("Features:", ncol(feature_matrix) - 4, "(excluding gene_id, transcript_id, symbol, te_lfc, te_padj)\n")
```

# Quality Control

```{r quality_control}
# Check for missing values
missing_summary <- feature_matrix %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("\nFeatures with missing values:\n")
print(missing_summary %>% head(20))

# Summary of key features
cat("\n\nFeature summaries:\n")
cat("\n5' UTR length:\n")
print(summary(feature_matrix$utr5_length))

cat("\n3' UTR length:\n")
print(summary(feature_matrix$utr3_length))

cat("\nCDS length:\n")
print(summary(feature_matrix$cds_length))

cat("\n5' UTR GC content:\n")
print(summary(feature_matrix$utr5_gc))

cat("\nTE log2 fold change (target):\n")
print(summary(feature_matrix$te_lfc))
```

# Visualize Feature Distributions

```{r visualize_distributions, fig.width=12, fig.height=10}
# Create histogram of key features
p1 <- ggplot(feature_matrix, aes(x = log2_utr5_length)) +
  geom_histogram(bins = 50, fill = "#0072B2", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "5' UTR Length (log2)", x = "log2(length + 1)", y = "Count")

p2 <- ggplot(feature_matrix, aes(x = log2_cds_length)) +
  geom_histogram(bins = 50, fill = "#E69F00", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "CDS Length (log2)", x = "log2(length + 1)", y = "Count")

p3 <- ggplot(feature_matrix, aes(x = log2_utr3_length)) +
  geom_histogram(bins = 50, fill = "#009E73", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "3' UTR Length (log2)", x = "log2(length + 1)", y = "Count")

p4 <- ggplot(feature_matrix, aes(x = utr5_gc)) +
  geom_histogram(bins = 50, fill = "#CC79A7", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "5' UTR GC Content", x = "GC %", y = "Count")

p5 <- ggplot(feature_matrix, aes(x = cds_gc)) +
  geom_histogram(bins = 50, fill = "#D55E00", alpha = 0.7) +
  theme_cowplot() +
  labs(title = "CDS GC Content", x = "GC %", y = "Count")

p6 <- ggplot(feature_matrix, aes(x = te_lfc)) +
  geom_histogram(bins = 50, fill = "#56B4E9", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  theme_cowplot() +
  labs(title = "TE Log2 Fold Change (Target)", x = "TE LFC", y = "Count")

combined_plot <- plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
print(combined_plot)

ggsave(here("plots", "feature_distributions_predictive_model.pdf"),
       plot = combined_plot, width = 12, height = 10)
```

# Create Binned TE Category for Stratified Sampling

```{r create_bins}
# Create bins for stratified train/test split
feature_matrix <- feature_matrix %>%
  mutate(te_lfc_bin = cut(te_lfc,
                          breaks = quantile(te_lfc, probs = seq(0, 1, 0.2), na.rm = TRUE),
                          include.lowest = TRUE,
                          labels = c("Q1", "Q2", "Q3", "Q4", "Q5")))

cat("\nTE LFC bins for stratified sampling:\n")
print(table(feature_matrix$te_lfc_bin))
```

# Export Feature Matrix

```{r export_features}
# Create output directory if needed
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

# Export as CSV
write_csv(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr.csv\n")

# Export as RDS for faster loading
saveRDS(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr.rds\n")

# Summary
cat("\n\nFinal Feature Matrix Summary:\n")
cat("Total genes:", nrow(feature_matrix), "\n")
cat("Total features:", ncol(feature_matrix), "\n")
cat("  - Identifiers: gene_id, transcript_id, symbol\n")
cat("  - Target: te_lfc, te_padj, te_lfc_bin\n")
cat("  - Region lengths: 7 features\n")
cat("  - GC content: 5 features\n")
cat("  - 4-mer frequencies: 256 features\n")
cat("  - 5-mer frequencies: 1024 features\n")
cat("  - 6-mer frequencies: 4096 features\n")
cat("  - RBP motif scores (RBNS + CISBP-RNA, unbiased):", ncol(motif_scores) - 1, "features\n")
cat("  - 4E resistance/sensitivity: 4 features\n")
cat("  - Translation signatures:", sum(str_detect(names(feature_matrix), "^sig_")), "features\n")
cat("  - CLIP binding:", sum(str_detect(names(feature_matrix), "^clip_")), "features\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
