---
title: "Targeted Feature Extraction for Hypoxia Response Prediction"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis extracts **biological/functional features only** (no sequence-based features) for a focused predictive model of Translation Efficiency (TE) Log2 Fold Change at 1 hour hypoxia.

## Feature Categories

| Category | Features | Count |
|----------|----------|-------|
| **eIF3 CLIP Binding** | Bound + regional (5utr/cds/3utr) × 7 conditions | ~28 |
| **4E Resistance** | 4E-resistant / 4E-sensitive binary | 2 |
| **DAP5 KD (Weber 2022)** | DAP5-repressed / DAP5-promoted binary | 2 |
| **DAP5 Polysome (GSE115142)** | DAP5-polysome promoted/inhibited binary + continuous LFC | 3 |
| **Target** | te_lfc, te_padj, te_lfc_bin | — |

## Features Excluded (vs original 01_feature_extraction.Rmd)

- Region lengths (5'UTR, CDS, 3'UTR)
- GC content
- K-mer frequencies (4/5/6-mers)
- RBP motif scores (RBNS + CISBP-RNA)
- GC3 codon usage
- Translation signatures (eIF3d/3e membership)

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
```

# Load Gene Annotation

```{r load_gene_annotation}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

# Load TE Data (1hr Hypoxia Response)

```{r load_te_data}
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))

cat("Loaded TE data:", nrow(te_1hr), "genes\n")

te_1hr <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  filter(!is.na(symbol))

cat("TE data with symbols:", nrow(te_1hr), "genes\n")
```

# eIF3 CLIP Binding Features

```{r clip_binding}
clip_data <- read_csv("/Users/katematlin/github/mpra_preliminary/score_card.csv")

clip_conditions <- c("lee_parclip_a_b_d_g", "lee_parclip_d_only", "cate_undiff_npcs",
                      "cate_diff_npcs", "lee_int_stress", "lee_complete_media", "lee_glu_dep")

clip_data <- clip_data %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

clip_features <- clip_data %>%
  dplyr::select(gene_id_clean) %>%
  dplyr::distinct()

for (condition in clip_conditions) {
  bound_col <- paste0("clip_", condition, "_bound")
  utr5_col <- paste0("clip_", condition, "_5utr")
  cds_col <- paste0("clip_", condition, "_cds")
  utr3_col <- paste0("clip_", condition, "_3utr")

  condition_data <- clip_data %>%
    dplyr::select(gene_id_clean, value = all_of(condition)) %>%
    dplyr::mutate(
      value_lower = tolower(value),
      !!bound_col := as.integer(value_lower != "not bound"),
      !!utr5_col  := as.integer(str_detect(value_lower, "5utr")),
      !!cds_col   := as.integer(str_detect(value_lower, "cds")),
      !!utr3_col  := as.integer(str_detect(value_lower, "3utr"))
    ) %>%
    dplyr::select(gene_id_clean, all_of(c(bound_col, utr5_col, cds_col, utr3_col)))

  clip_features <- clip_features %>%
    dplyr::left_join(condition_data, by = "gene_id_clean")
}

clip_features <- clip_features %>%
  dplyr::mutate(across(-gene_id_clean, ~as.integer(replace_na(., 0L))))

# Drop columns that are all zero
all_zero_cols <- names(clip_features)[sapply(clip_features, function(x) is.numeric(x) && sum(x) == 0)]
if (length(all_zero_cols) > 0) {
  cat("Dropping", length(all_zero_cols), "all-zero CLIP columns:", paste(all_zero_cols, collapse = ", "), "\n")
  clip_features <- clip_features %>%
    dplyr::select(-all_of(all_zero_cols))
}

cat("CLIP features retained:", ncol(clip_features) - 1, "features\n")
cat("Genes with CLIP data:", nrow(clip_features), "\n")
```

# 4E Resistance Features

```{r 4e_resistance}
teleman_data <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv") %>%
  dplyr::filter(!is.na(symbol)) %>%
  dplyr::group_by(symbol) %>%
  dplyr::slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

resistant_4e_symbols <- teleman_data %>% dplyr::filter(log2FoldChange > 1, padj < 0.05) %>% dplyr::pull(symbol)
sensitive_4e_symbols <- teleman_data %>% dplyr::filter(log2FoldChange < -1, padj < 0.05) %>% dplyr::pull(symbol)

cat("4E-resistant genes:", length(resistant_4e_symbols), "\n")
cat("4E-sensitive genes:", length(sensitive_4e_symbols), "\n")
```

# DAP5 KD Features (Weber 2022)

```{r dap5_weber}
dap5_data <- read_xlsx(here("accessories", "dap5kd_riboseq_weber2022.xlsx")) %>%
  dplyr::filter(!is.na(geneIDs), comment != "no")

dap5_repressed_symbols <- dap5_data %>% dplyr::filter(comment == "up") %>% dplyr::pull(geneIDs)
dap5_promoted_symbols <- dap5_data %>% dplyr::filter(comment == "down") %>% dplyr::pull(geneIDs)

cat("DAP5-repressed genes (up with KD):", length(dap5_repressed_symbols), "\n")
cat("DAP5-promoted genes (down with KD):", length(dap5_promoted_symbols), "\n")
```

# DAP5 Polysome Features (GSE115142)

```{r dap5_polysome}
dap5_polysome <- read_csv(here("output", "dap5_deseq2_full_results.csv"))

cat("DAP5 polysome DESeq2 results:", nrow(dap5_polysome), "genes\n")

# Define promoted/inhibited based on LFC and significance
dap5_polysome_features <- dap5_polysome %>%
  dplyr::transmute(
    symbol = symbol,
    is_dap5_polysome_promoted  = as.integer(log2FoldChange > 0.5 & padj < 0.05),
    is_dap5_polysome_inhibited = as.integer(log2FoldChange < -0.5 & padj < 0.05),
    dap5_polysome_lfc          = log2FoldChange
  ) %>%
  dplyr::filter(!is.na(symbol))

# Replace NA padj-derived binary features with 0 (genes with NA padj are not significant)
dap5_polysome_features <- dap5_polysome_features %>%
  dplyr::mutate(
    is_dap5_polysome_promoted  = replace_na(is_dap5_polysome_promoted, 0L),
    is_dap5_polysome_inhibited = replace_na(is_dap5_polysome_inhibited, 0L)
  )

cat("DAP5-polysome promoted genes:", sum(dap5_polysome_features$is_dap5_polysome_promoted), "\n")
cat("DAP5-polysome inhibited genes:", sum(dap5_polysome_features$is_dap5_polysome_inhibited), "\n")
```

# Assemble Feature Matrix

```{r assemble_features}
# Start with TE data as the base
feature_matrix <- te_1hr %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::select(gene_id, gene_id_clean, symbol, te_lfc, te_padj) %>%
  # Add CLIP binding features (by gene_id_clean)
  left_join(clip_features, by = "gene_id_clean") %>%
  # Add 4E resistance and DAP5 KD features (by symbol)
  mutate(
    is_4e_resistant  = as.integer(symbol %in% resistant_4e_symbols),
    is_4e_sensitive  = as.integer(symbol %in% sensitive_4e_symbols),
    is_dap5_repressed = as.integer(symbol %in% dap5_repressed_symbols),
    is_dap5_promoted  = as.integer(symbol %in% dap5_promoted_symbols)
  ) %>%
  # Add DAP5 polysome features (by symbol)
  left_join(dap5_polysome_features, by = "symbol")

# Fill NAs with 0 for binary CLIP features
feature_matrix <- feature_matrix %>%
  dplyr::mutate(across(starts_with("clip_"), ~replace_na(., 0L)))

# Fill NAs for DAP5 polysome binary features (genes not in the polysome dataset)
feature_matrix <- feature_matrix %>%
  dplyr::mutate(
    is_dap5_polysome_promoted  = replace_na(is_dap5_polysome_promoted, 0L),
    is_dap5_polysome_inhibited = replace_na(is_dap5_polysome_inhibited, 0L)
  )

# Drop any binary feature columns that are all zero after filtering
binary_prefixes <- c("clip_", "is_")
binary_cols <- names(feature_matrix)[sapply(names(feature_matrix), function(n) any(str_starts(n, binary_prefixes)))]
all_zero_binary <- binary_cols[sapply(binary_cols, function(col) sum(feature_matrix[[col]], na.rm = TRUE) == 0)]
if (length(all_zero_binary) > 0) {
  cat("Dropping", length(all_zero_binary), "all-zero binary columns:\n")
  cat(" ", paste(all_zero_binary, collapse = ", "), "\n")
  feature_matrix <- feature_matrix %>%
    dplyr::select(-all_of(all_zero_binary))
}

# --- Negative control attrition diagnostic ---
neg_ctrl <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
neg_ids <- neg_ctrl$gene_id_clean

cat("\n=== Negative control gene attrition ===\n")
cat("Total negative control genes:", length(neg_ids), "\n")
cat("Unique genes in feature_matrix:", n_distinct(feature_matrix$gene_id_clean), "\n")
cat("Neg ctrl genes in feature_matrix:",
    n_distinct(feature_matrix$gene_id_clean[feature_matrix$gene_id_clean %in% neg_ids]), "\n")
cat("Neg ctrl genes NOT in feature_matrix:",
    sum(!neg_ids %in% feature_matrix$gene_id_clean), "\n")
cat("Duplicate gene_id_clean rows in feature_matrix:",
    nrow(feature_matrix) - n_distinct(feature_matrix$gene_id_clean), "\n")

cat("\nFeature matrix assembled:\n")
cat("Genes:", nrow(feature_matrix), "\n")
cat("Total columns:", ncol(feature_matrix), "\n")
```

# Create Binned TE Category for Stratified Sampling

```{r create_bins}
feature_matrix <- feature_matrix %>%
  mutate(te_lfc_bin = cut(te_lfc,
                          breaks = quantile(te_lfc, probs = seq(0, 1, 0.2), na.rm = TRUE),
                          include.lowest = TRUE,
                          labels = c("Q1", "Q2", "Q3", "Q4", "Q5")))

cat("\nTE LFC bins for stratified sampling:\n")
print(table(feature_matrix$te_lfc_bin))
```

# Quality Control

```{r quality_control}
# Check for missing values
missing_summary <- feature_matrix %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("\nFeatures with missing values:\n")
if (nrow(missing_summary) > 0) {
  print(missing_summary)
} else {
  cat("None (all features complete)\n")
}

# Summarize binary feature counts
cat("\nBinary feature summary (non-zero counts):\n")
binary_cols_final <- names(feature_matrix)[sapply(names(feature_matrix), function(n) any(str_starts(n, c("clip_", "is_"))))]
for (col in binary_cols_final) {
  cat(sprintf("  %s: %d genes\n", col, sum(feature_matrix[[col]], na.rm = TRUE)))
}

# Summary of continuous features
cat("\nDAP5 polysome LFC summary:\n")
print(summary(feature_matrix$dap5_polysome_lfc))

cat("\nTE log2 fold change (target):\n")
print(summary(feature_matrix$te_lfc))
```

# Export Feature Matrix

```{r export_features}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

write_csv(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_targeted.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_targeted.csv\n")

saveRDS(feature_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_targeted.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_targeted.rds\n")

cat("\n\nFinal Feature Matrix Summary:\n")
cat("Total genes:", nrow(feature_matrix), "\n")
cat("Total columns:", ncol(feature_matrix), "\n")
cat("  - Identifiers: gene_id, gene_id_clean, symbol\n")
cat("  - Target: te_lfc, te_padj, te_lfc_bin\n")
cat("  - CLIP binding:", sum(str_detect(names(feature_matrix), "^clip_")), "features\n")
cat("  - 4E resistance/sensitivity: 2 features\n")
cat("  - DAP5 KD (Weber): 2 features\n")
cat("  - DAP5 polysome: 3 features (2 binary + 1 continuous)\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
