---
title: "Pre-compute K-mer Frequencies and RBP Motif Scores"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This helper notebook pre-computes the slow k-mer frequency and RBP motif scoring steps so they can be loaded quickly by `01c_feature_extraction_combined.Rmd`.

**Outputs:**
- `output/predictive_modeling/precomputed_kmer_features.rds`
- `output/predictive_modeling/precomputed_motif_features.rds`
- `output/predictive_modeling/precomputed_most_abundant_tx.rds`

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(tximport)
library(FeatureReachR)
```

# Load Gene Annotation

```{r load_gene_annotation}
gene_anno <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                      col_names = c("gene_id", "symbol", "type", "organism", "source"))

gene_symbols <- gene_anno %>%
  dplyr::select(gene_id, symbol, type) %>%
  arrange(gene_id, type == "protein_coding") %>%
  group_by(gene_id) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  dplyr::select(gene_id, symbol)

cat("Gene annotation loaded:", nrow(gene_symbols), "genes\n")
```

# Load TE Data (1hr Hypoxia Response)

```{r load_te_data}
te_1hr <- read_csv(here("output", "translation_categories_1hr.csv"))

cat("Loaded TE data:", nrow(te_1hr), "genes\n")

te_1hr <- te_1hr %>%
  left_join(gene_symbols, by = "gene_id") %>%
  filter(!is.na(symbol))

cat("TE data with symbols:", nrow(te_1hr), "genes\n")
```

# Load TxDb and Genome

```{r load_txdb}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))

tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript_id", "gene_id")) %>%
  filter(str_starts(transcript_id, "ENST"))

gene_symbol_map <- gene_symbols

genome <- BSgenome.Hsapiens.UCSC.hg38

cat("Loaded TxDb and genome\n")
cat("Transcripts in tx2gene:", nrow(tx2gene), "\n")
```

# Select Most Abundant Transcript Per Gene

```{r load_salmon}
salmon_dir <- here("2025_MDAcells_riboseq", "salmon")
salmon_samples <- list.dirs(salmon_dir, recursive = FALSE, full.names = FALSE)

salmon_files <- file.path(salmon_dir, salmon_samples, "quant.sf.gz")
names(salmon_files) <- salmon_samples

salmon_files <- salmon_files[file.exists(salmon_files)]
cat("Found", length(salmon_files), "Salmon quant files\n")

txi_transcripts <- tximport(salmon_files, type = "salmon", tx2gene = tx2gene, txOut = TRUE)

transcript_abundance <- txi_transcripts$abundance %>%
  as.data.frame() %>%
  rownames_to_column(var = "transcript_id") %>%
  left_join(tx2gene, by = "transcript_id")

most_abundant_tx <- transcript_abundance %>%
  mutate(total_abundance = rowSums(across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::select(transcript_id, gene_id, total_abundance) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    transcript_id_clean = sub("\\..*", "", transcript_id)
  ) %>%
  group_by(gene_id_clean) %>%
  dplyr::slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

most_abundant_tx <- most_abundant_tx %>%
  left_join(gene_symbol_map %>% mutate(gene_id_clean = sub("\\..*", "", gene_id)),
            by = "gene_id_clean")

cat("Selected most abundant transcript for", nrow(most_abundant_tx), "genes (all)\n")

te_gene_ids <- te_1hr %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  pull(gene_id_clean) %>%
  unique()

most_abundant_tx <- most_abundant_tx %>%
  filter(gene_id_clean %in% te_gene_ids)

cat("After filtering to genes with TE data:", nrow(most_abundant_tx), "genes\n")

cat("\nExample transcript IDs from Salmon:", head(most_abundant_tx$transcript_id, 3), "\n")
cat("Example transcript IDs (cleaned):", head(most_abundant_tx$transcript_id_clean, 3), "\n")
```

# Extract Sequences

```{r extract_sequences}
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cds <- cdsBy(txdb, by = "tx", use.names = TRUE)

# Extract 5' UTR sequences
utr5_sequences <- extractTranscriptSeqs(genome, utr5) %>% as.data.frame()
utr5_seq <- utr5_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr5_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr5_sequence)

# Extract 3' UTR sequences
utr3_sequences <- extractTranscriptSeqs(genome, utr3) %>% as.data.frame()
utr3_seq <- utr3_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("utr3_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, utr3_sequence)

# Extract CDS sequences
cds_sequences <- extractTranscriptSeqs(genome, cds) %>% as.data.frame()
cds_seq <- cds_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("cds_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, cds_sequence)

cat("Extracted sequences:\n")
cat("5' UTRs:", nrow(utr5_seq), "\n")
cat("3' UTRs:", nrow(utr3_seq), "\n")
cat("CDS:", nrow(cds_seq), "\n")
```

# Calculate 4-mer Frequencies in 5' UTR

```{r calculate_4mers}
valid_transcripts <- most_abundant_tx %>%
  filter(transcript_id_clean %in% utr5_seq$transcript_id_clean)

cat("Transcripts with 5' UTR data:", nrow(valid_transcripts), "out of", nrow(most_abundant_tx), "\n")

utr5_for_kmers <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "")

# Filter out very short sequences (need at least 4nt for 4-mers)
utr5_for_kmers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 4)

utr5_dna <- DNAStringSet(utr5_for_kmers$utr5_sequence)
names(utr5_dna) <- utr5_for_kmers$transcript_id_clean

kmer4_matrix <- oligonucleotideFrequency(utr5_dna, width = 4, as.prob = TRUE)

kmer4_df <- as.data.frame(kmer4_matrix) %>%
  mutate(transcript_id_clean = utr5_for_kmers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

colnames(kmer4_df)[1:256] <- paste0("utr5_4mer_", colnames(kmer4_df)[1:256])

cat("Calculated 4-mer frequencies for", nrow(kmer4_df), "transcripts\n")
cat("Number of 4-mer features:", 256, "\n")
```

# Calculate 5-mer Frequencies in 5' UTR

```{r calculate_5mers}
utr5_for_5mers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 5)

utr5_dna_5mer <- DNAStringSet(utr5_for_5mers$utr5_sequence)
names(utr5_dna_5mer) <- utr5_for_5mers$transcript_id_clean

kmer5_matrix <- oligonucleotideFrequency(utr5_dna_5mer, width = 5, as.prob = TRUE)

kmer5_df <- as.data.frame(kmer5_matrix) %>%
  mutate(transcript_id_clean = utr5_for_5mers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

colnames(kmer5_df)[1:1024] <- paste0("utr5_5mer_", colnames(kmer5_df)[1:1024])

cat("Calculated 5-mer frequencies for", nrow(kmer5_df), "transcripts\n")
cat("Number of 5-mer features:", 1024, "\n")
```

# Calculate 6-mer Frequencies in 5' UTR

```{r calculate_6mers}
utr5_for_6mers <- utr5_for_kmers %>%
  filter(str_length(utr5_sequence) >= 6)

utr5_dna_6mer <- DNAStringSet(utr5_for_6mers$utr5_sequence)
names(utr5_dna_6mer) <- utr5_for_6mers$transcript_id_clean

kmer6_matrix <- oligonucleotideFrequency(utr5_dna_6mer, width = 6, as.prob = TRUE)

kmer6_df <- as.data.frame(kmer6_matrix) %>%
  mutate(transcript_id_clean = utr5_for_6mers$transcript_id_clean) %>%
  distinct(transcript_id_clean, .keep_all = TRUE)

colnames(kmer6_df)[1:4096] <- paste0("utr5_6mer_", colnames(kmer6_df)[1:4096])

cat("Calculated 6-mer frequencies for", nrow(kmer6_df), "transcripts\n")
cat("Number of 6-mer features:", 4096, "\n")
```

# Calculate RBP Motif Scores (All Motifs, All Regions)

```{r rbp_motifs}
# Score all RBP motifs from both RBNS_PWM and CISBPRNA_hs_PWM databases
# across 5' UTR, CDS, and 3' UTR regions

# --- Prepare sequences for each region ---

# 5' UTR
utr5_motif_seqs <- utr5_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr5_sequence),
         utr5_sequence != "",
         str_length(utr5_sequence) >= 6)

utr5_dna_motif <- DNAStringSet(utr5_motif_seqs$utr5_sequence)
names(utr5_dna_motif) <- utr5_motif_seqs$transcript_id_clean

# CDS
cds_motif_seqs <- cds_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(cds_sequence),
         cds_sequence != "",
         str_length(cds_sequence) >= 6)

cds_dna_motif <- DNAStringSet(cds_motif_seqs$cds_sequence)
names(cds_dna_motif) <- cds_motif_seqs$transcript_id_clean

# 3' UTR
utr3_motif_seqs <- utr3_seq %>%
  filter(transcript_id_clean %in% valid_transcripts$transcript_id_clean,
         !is.na(utr3_sequence),
         utr3_sequence != "",
         str_length(utr3_sequence) >= 6)

utr3_dna_motif <- DNAStringSet(utr3_motif_seqs$utr3_sequence)
names(utr3_dna_motif) <- utr3_motif_seqs$transcript_id_clean

cat("Sequences for motif scoring:\n")
cat("  5' UTR:", length(utr5_dna_motif), "\n")
cat("  CDS:", length(cds_dna_motif), "\n")
cat("  3' UTR:", length(utr3_dna_motif), "\n")
cat("RBNS motifs available:", length(RBNS_PWM), "\n")
cat("CISBP-RNA motifs available:", length(CISBPRNA_hs_PWM), "\n")

# --- Helper function to score all motifs from a PWM list against a set of sequences ---
score_pwm_list <- function(pwm_list, sequences, db_prefix, region_prefix) {
  seq_lengths <- width(sequences)
  scores <- data.frame(transcript_id_clean = names(sequences))

  for (motif_name in names(pwm_list)) {
    pwm <- pwm_list[[motif_name]]
    pwm_matrix <- as.matrix(pwm)

    counts <- sapply(sequences, function(seq) {
      if (length(seq) < ncol(pwm_matrix)) return(0)
      tryCatch({
        length(matchPWM(pwm_matrix, seq, min.score = "80%"))
      }, error = function(e) 0)
    })

    col_name <- paste0(db_prefix, "_", region_prefix, "_", gsub("[^A-Za-z0-9]", "_", motif_name))
    scores[[col_name]] <- (counts / seq_lengths) * 1000
  }

  return(scores)
}

# --- Score RBNS motifs in all three regions ---
cat("Scoring RBNS motifs in 5' UTR...\n")
rbns_utr5 <- score_pwm_list(RBNS_PWM, utr5_dna_motif, "rbns", "utr5")
cat("Scoring RBNS motifs in CDS...\n")
rbns_cds <- score_pwm_list(RBNS_PWM, cds_dna_motif, "rbns", "cds")
cat("Scoring RBNS motifs in 3' UTR...\n")
rbns_utr3 <- score_pwm_list(RBNS_PWM, utr3_dna_motif, "rbns", "utr3")

cat("Scored", length(names(RBNS_PWM)), "RBNS motifs x 3 regions\n")

# --- Score CISBP-RNA motifs in all three regions ---
cat("Scoring CISBP-RNA motifs in 5' UTR...\n")
cisbp_utr5 <- score_pwm_list(CISBPRNA_hs_PWM, utr5_dna_motif, "cisbp", "utr5")
cat("Scoring CISBP-RNA motifs in CDS...\n")
cisbp_cds <- score_pwm_list(CISBPRNA_hs_PWM, cds_dna_motif, "cisbp", "cds")
cat("Scoring CISBP-RNA motifs in 3' UTR...\n")
cisbp_utr3 <- score_pwm_list(CISBPRNA_hs_PWM, utr3_dna_motif, "cisbp", "utr3")

cat("Scored", length(names(CISBPRNA_hs_PWM)), "CISBP-RNA motifs x 3 regions\n")

# --- Merge all motif scores ---
motif_scores <- rbns_utr5 %>%
  full_join(rbns_cds, by = "transcript_id_clean") %>%
  full_join(rbns_utr3, by = "transcript_id_clean") %>%
  full_join(cisbp_utr5, by = "transcript_id_clean") %>%
  full_join(cisbp_cds, by = "transcript_id_clean") %>%
  full_join(cisbp_utr3, by = "transcript_id_clean")

cat("Total RBP motif features:", ncol(motif_scores) - 1, "\n")
cat("  Per database: RBNS =", length(RBNS_PWM), ", CISBP-RNA =", length(CISBPRNA_hs_PWM), "\n")
cat("  Per region: 5' UTR, CDS, 3' UTR\n")
```

# Export Pre-computed Features

```{r export}

\
ir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

# Combine all k-mer dataframes
kmer_features <- kmer4_df %>%
  full_join(kmer5_df, by = "transcript_id_clean") %>%
  full_join(kmer6_df, by = "transcript_id_clean")

cat("K-mer feature matrix:", nrow(kmer_features), "transcripts x", ncol(kmer_features) - 1, "features\n")

# Save k-mer features
saveRDS(kmer_features, here("output", "predictive_modeling", "precomputed_kmer_features.rds"))
cat("Saved: output/predictive_modeling/precomputed_kmer_features.rds\n")

# Save motif scores
saveRDS(motif_scores, here("output", "predictive_modeling", "precomputed_motif_features.rds"))
cat("Saved: output/predictive_modeling/precomputed_motif_features.rds\n")

# Save most abundant transcript selection for consistency
saveRDS(most_abundant_tx, here("output", "predictive_modeling", "precomputed_most_abundant_tx.rds"))
cat("Saved: output/predictive_modeling/precomputed_most_abundant_tx.rds\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
