---
title: "RNA Secondary Structure Feature Extraction for 5' UTRs"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis parses precomputed RNAplfold and RNAfold output (run on the cluster
via `cluster_scripts/viennarna/`) to extract RNA secondary structure features for
5' UTRs, then adds them to the existing combined feature matrix.

## Structure Features Computed (~13 features)

| Source | Feature | Description |
|--------|---------|-------------|
| **RNAplfold** | `struct_accessibility_cap_proximal` | Mean unpaired probability, positions 1-30 |
| **RNAplfold** | `struct_accessibility_aug_context` | Mean unpaired probability, last 30 nt of 5'UTR |
| **RNAplfold** | `struct_accessibility_utr5_mean` | Mean unpaired probability, full 5'UTR |
| **RNAplfold** | `struct_accessibility_utr5_min` | Minimum (most structured region) |
| **RNAplfold** | `struct_accessibility_utr5_sd` | Variability of structure along 5'UTR |
| **RNAplfold** | `struct_num_structured_regions` | Count of 30-nt windows with accessibility < 0.2 |
| **RNAfold** | `struct_mfe_utr5` | Minimum free energy (kcal/mol) |
| **RNAfold** | `struct_mfe_per_nt_utr5` | MFE normalized by 5'UTR length |
| **RNAfold** | `struct_ensemble_free_energy` | Ensemble free energy from partition function |
| **RNAfold** | `struct_ensemble_diversity` | Ensemble diversity (positional entropy) |
| **Dot-bracket** | `struct_frac_paired_utr5` | Fraction of nucleotides in base pairs |
| **Dot-bracket** | `struct_num_hairpins_utr5` | Number of hairpin loops |
| **Dot-bracket** | `struct_longest_stem_utr5` | Length of longest consecutive stem |

## Prerequisites

ViennaRNA outputs were precomputed on the cluster (see `cluster_scripts/viennarna/`):
- `RNAplfold -W 80 -L 40 -u 30` → `_lunp` files in `accessories/plfold_output/rnaplfold_output/`
- `RNAfold -p --noPS` → `rnafold_output.txt` in `accessories/plfold_output/rnafold_output/`

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

# Load Precomputed RNAplfold Output

```{r load_plfold}
plfold_dir <- here("accessories", "plfold_output", "rnaplfold_output", "rnaplfold_output")
plfold_files <- list.files(plfold_dir, pattern = "_lunp$", full.names = TRUE)
cat("RNAplfold _lunp files found:", length(plfold_files), "\n")
```

# Parse RNAplfold Output

```{r parse_rnaplfold}
parse_lunp <- function(filepath) {
  # RNAplfold _lunp files: tab-separated, first column = position, subsequent columns = unpaired prob for length 1..u
  # Header lines start with #
  lines <- readLines(filepath)
  header_end <- max(which(str_starts(lines, "#")))
  data_lines <- lines[(header_end + 1):length(lines)]
  data_lines <- data_lines[nzchar(trimws(data_lines))]

  if (length(data_lines) == 0) return(NULL)

  # Parse: columns are position, u=1, u=2, ..., u=30
  df <- read.table(text = data_lines, header = FALSE, sep = "\t", fill = TRUE)
  n_cols <- ncol(df)

  # Use single-nucleotide unpaired probability (column 2, u=1) for per-position accessibility
  result <- data.frame(
    position = df[[1]],
    unpaired_prob_1nt = df[[2]]
  )

  # Also get 30-nt window accessibility if available (last column for u=30)
  if (n_cols >= 31) {
    result$unpaired_prob_30nt <- df[[31]]  # position + 30 columns
  } else {
    result$unpaired_prob_30nt <- df[[n_cols]]  # use largest available window
  }

  return(result)
}

plfold_features <- lapply(plfold_files, function(f) {
  tx_id <- sub("_lunp$", "", basename(f))
  parsed <- parse_lunp(f)
  if (is.null(parsed)) return(NULL)

  seq_len <- nrow(parsed)

  # Cap-proximal: positions 1-30 (or full sequence if shorter)
  cap_range <- 1:min(30, seq_len)
  # AUG context: last 30 nt of 5'UTR
  aug_range <- max(1, seq_len - 29):seq_len

  # Use single-nt unpaired probability for positional features
  acc_vals <- parsed$unpaired_prob_1nt
  acc_vals[is.na(acc_vals)] <- 0

  # 30-nt window accessibility for structured region counting
  acc_30nt <- parsed$unpaired_prob_30nt
  acc_30nt[is.na(acc_30nt)] <- 0

  # Count structured regions: positions where 30-nt window accessibility < 0.2
  valid_30nt <- acc_30nt[acc_30nt > 0 | !is.na(parsed$unpaired_prob_30nt)]
  num_structured <- sum(valid_30nt < 0.2, na.rm = TRUE)

  tibble(
    transcript_id_clean = tx_id,
    struct_accessibility_cap_proximal = mean(acc_vals[cap_range], na.rm = TRUE),
    struct_accessibility_aug_context = mean(acc_vals[aug_range], na.rm = TRUE),
    struct_accessibility_utr5_mean = mean(acc_vals, na.rm = TRUE),
    struct_accessibility_utr5_min = min(acc_vals, na.rm = TRUE),
    struct_accessibility_utr5_sd = sd(acc_vals, na.rm = TRUE),
    struct_num_structured_regions = num_structured
  )
})

plfold_df <- bind_rows(plfold_features)
cat("RNAplfold features computed for", nrow(plfold_df), "transcripts\n")

cat("\nAccessibility summary (full 5'UTR mean):\n")
print(summary(plfold_df$struct_accessibility_utr5_mean))
```

# Load Precomputed RNAfold Output

```{r load_rnafold}
rnafold_out <- here("accessories", "plfold_output", "rnafold_output", "rnafold_output.txt")
stopifnot("RNAfold output not found — is the cluster job finished?" = file.exists(rnafold_out))
rnafold_lines <- readLines(rnafold_out)
cat("RNAfold output loaded:", length(rnafold_lines), "lines\n")
```

# Parse RNAfold Output

```{r parse_rnafold}
# RNAfold -p output format for each sequence:
# Line 1: >transcript_id
# Line 2: sequence
# Line 3: MFE structure (dot-bracket) ( MFE_value)
# Line 4: ensemble structure {...} [ ensemble_free_energy]
# Line 5: frequency info (may contain ensemble diversity)
# Blank line between sequences

parse_rnafold_output <- function(lines) {
  results <- list()
  i <- 1

  while (i <= length(lines)) {
    # Skip blank lines
    if (nzchar(trimws(lines[i])) == FALSE) {
      i <- i + 1
      next
    }

    # Look for header line
    if (str_starts(lines[i], ">")) {
      tx_id <- sub("^>", "", trimws(lines[i]))

      # Next line: sequence
      i <- i + 1
      if (i > length(lines)) break
      sequence <- trimws(lines[i])
      seq_len <- nchar(sequence)

      # Next line: MFE structure
      i <- i + 1
      if (i > length(lines)) break
      mfe_line <- lines[i]

      # Extract dot-bracket structure and MFE value
      # Format: "...(((...)))... ( -12.34)"
      mfe_match <- str_match(mfe_line, "^([.()]+)\\s+\\(\\s*([\\-\\d.]+)\\)")
      dot_bracket <- NA_character_
      mfe <- NA_real_
      if (!is.na(mfe_match[1, 1])) {
        dot_bracket <- mfe_match[1, 2]
        mfe <- as.numeric(mfe_match[1, 3])
      }

      # Next line: ensemble structure with free energy [...]
      i <- i + 1
      ensemble_fe <- NA_real_
      if (i <= length(lines)) {
        ens_line <- lines[i]
        ens_match <- str_match(ens_line, "\\[\\s*([\\-\\d.]+)\\]")
        if (!is.na(ens_match[1, 1])) {
          ensemble_fe <- as.numeric(ens_match[1, 2])
        }
      }

      # Next line: centroid structure with distance {...}
      # Format: "...(((...)))... {-37.90 d=18.20}"
      i <- i + 1

      # Next line: frequency + ensemble diversity (on the same line)
      # Format: " frequency of mfe structure in ensemble 0.0548151; ensemble diversity 27.37"
      i <- i + 1
      ensemble_div <- NA_real_
      if (i <= length(lines)) {
        freq_line <- lines[i]
        div_match <- str_match(freq_line, "ensemble diversity\\s+([\\d.]+)")
        if (!is.na(div_match[1, 1])) {
          ensemble_div <- as.numeric(div_match[1, 2])
        }
      }

      results[[length(results) + 1]] <- tibble(
        transcript_id_clean = tx_id,
        struct_mfe_utr5 = mfe,
        struct_mfe_per_nt_utr5 = ifelse(seq_len > 0, mfe / seq_len, NA_real_),
        struct_ensemble_free_energy = ensemble_fe,
        struct_ensemble_diversity = ensemble_div,
        dot_bracket = dot_bracket,
        utr5_length_rnafold = seq_len
      )
    }

    i <- i + 1
  }

  bind_rows(results)
}

rnafold_df <- parse_rnafold_output(rnafold_lines)
cat("RNAfold features parsed for", nrow(rnafold_df), "transcripts\n")

cat("\nMFE summary:\n")
print(summary(rnafold_df$struct_mfe_utr5))

cat("\nMFE per nt summary:\n")
print(summary(rnafold_df$struct_mfe_per_nt_utr5))

cat("\nEnsemble free energy summary:\n")
print(summary(rnafold_df$struct_ensemble_free_energy))

cat("\nEnsemble diversity summary:\n")
print(summary(rnafold_df$struct_ensemble_diversity))
```

# Compute Dot-Bracket Features

```{r dot_bracket_features}
parse_dot_bracket <- function(structure) {
  if (is.na(structure)) {
    return(tibble(
      struct_frac_paired_utr5 = NA_real_,
      struct_num_hairpins_utr5 = NA_integer_,
      struct_longest_stem_utr5 = NA_integer_
    ))
  }

  chars <- strsplit(structure, "")[[1]]
  n <- length(chars)

  # Fraction paired: count '(' and ')' characters
  n_paired <- sum(chars %in% c("(", ")"))
  frac_paired <- n_paired / n

  # Count hairpin loops: unpaired regions enclosed by a single base pair
  n_hairpins <- 0L
  stack <- integer(0)
  for (j in seq_along(chars)) {
    if (chars[j] == "(") {
      stack <- c(stack, j)
    } else if (chars[j] == ")") {
      if (length(stack) > 0) {
        open_pos <- stack[length(stack)]
        stack <- stack[-length(stack)]
        if (j - open_pos > 1) {
          inner <- chars[(open_pos + 1):(j - 1)]
          if (all(inner == ".")) {
            n_hairpins <- n_hairpins + 1L
          }
        }
      }
    }
  }

  # Longest stem: longest run of consecutive '(' characters
  rle_result <- rle(chars == "(")
  stem_lengths <- rle_result$lengths[rle_result$values]
  longest_stem <- if (length(stem_lengths) > 0) max(stem_lengths) else 0L

  tibble(
    struct_frac_paired_utr5 = frac_paired,
    struct_num_hairpins_utr5 = n_hairpins,
    struct_longest_stem_utr5 = as.integer(longest_stem)
  )
}

db_features <- map_dfr(seq_len(nrow(rnafold_df)), function(i) {
  parse_dot_bracket(rnafold_df$dot_bracket[i])
})

rnafold_df <- bind_cols(
  rnafold_df %>% dplyr::select(-dot_bracket, -utr5_length_rnafold),
  db_features
)

cat("Dot-bracket features computed\n")
cat("\nFraction paired summary:\n")
print(summary(rnafold_df$struct_frac_paired_utr5))

cat("\nNumber of hairpins summary:\n")
print(summary(rnafold_df$struct_num_hairpins_utr5))

cat("\nLongest stem summary:\n")
print(summary(rnafold_df$struct_longest_stem_utr5))
```

# Combine Structure Features

```{r combine_structure}
structure_features <- plfold_df %>%
  full_join(rnafold_df, by = "transcript_id_clean")

cat("Combined structure features:", nrow(structure_features), "transcripts x",
    ncol(structure_features) - 1, "features\n")

# Check for NAs
na_summary <- structure_features %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0)

cat("\nMissing values:\n")
if (nrow(na_summary) > 0) {
  print(na_summary)
} else {
  cat("None\n")
}
```

# Load Existing Combined Feature Matrix and Join

```{r join_to_combined}
combined_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_combined.rds"))
cat("Loaded combined feature matrix:", nrow(combined_matrix), "genes x", ncol(combined_matrix), "columns\n")

# Join structure features by transcript_id_clean
structure_matrix <- combined_matrix %>%
  left_join(structure_features, by = "transcript_id_clean")

# Fill NAs in structure features with 0 for transcripts without 5'UTR structure data
struct_cols <- names(structure_features)[names(structure_features) != "transcript_id_clean"]
structure_matrix <- structure_matrix %>%
  mutate(across(all_of(struct_cols), ~replace_na(., 0)))

cat("\nStructure-augmented matrix:", nrow(structure_matrix), "genes x", ncol(structure_matrix), "columns\n")

# Count how many genes got structure features
n_with_structure <- sum(combined_matrix$transcript_id_clean %in% structure_features$transcript_id_clean)
cat("Genes with structure features:", n_with_structure, "out of", nrow(combined_matrix), "\n")
```

# Quality Control

```{r quality_control}
cat("Structure feature distributions:\n\n")

for (col in struct_cols) {
  vals <- structure_matrix[[col]]
  cat(col, ":\n")
  print(summary(vals))
  cat("  Non-zero:", sum(vals != 0, na.rm = TRUE), "/", length(vals), "\n\n")
}
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(structure_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_structure.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_structure.rds\n")

write_csv(structure_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_structure.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_structure.csv\n")

cat("\nFinal matrix:", nrow(structure_matrix), "genes x", ncol(structure_matrix), "columns\n")
cat("Structure features added:", length(struct_cols), "\n")
cat("Structure feature names:", paste(struct_cols, collapse = ", "), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
