---
title: "Split CLIP Aggregate Features (eIF3 vs Cap-Binding)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

The original `clip_evidence_count` feature (from 01c) sums binding evidence across
all 6 CLIP conditions into a single count (0–6). Three of those conditions
(`lee_int_stress`, `lee_complete_media`, `lee_glu_dep`) represent **cap-binding**
activity rather than eIF3-specific binding. This notebook splits the aggregate into
two features so the model can learn from them independently:

- `clip_eif3_count` — eIF3-specific binding (0–3)
- `clip_cap_binding_count` — cap-binding activity (0–3)

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
```

# Load Feature Matrix

```{r load_matrix}
structure_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_structure.rds"))
cat("Loaded structure matrix:", nrow(structure_matrix), "genes x", ncol(structure_matrix), "columns\n")
cat("clip_evidence_count present:", "clip_evidence_count" %in% colnames(structure_matrix), "\n")
```

# Compute Split CLIP Features

```{r compute_clip_split}
clip_update <- read_csv("/Users/katematlin/github/mpra_preliminary/scorecard_update.csv") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

eif3_conditions <- c("lee_parclip_a_b_d_g", "cate_undiff_npcs", "cate_diff_npcs")
cap_binding_conditions <- c("lee_int_stress", "lee_complete_media", "lee_glu_dep")

clip_split <- clip_update %>%
  dplyr::select(gene_id_clean, all_of(eif3_conditions), all_of(cap_binding_conditions)) %>%
  dplyr::mutate(
    clip_eif3_count = rowSums(
      across(all_of(eif3_conditions), ~as.integer(tolower(.) != "not bound")),
      na.rm = TRUE
    ),
    clip_cap_binding_count = rowSums(
      across(all_of(cap_binding_conditions), ~as.integer(tolower(.) != "not bound")),
      na.rm = TRUE
    )
  ) %>%
  dplyr::select(gene_id_clean, clip_eif3_count, clip_cap_binding_count)

cat("clip_eif3_count range:", range(clip_split$clip_eif3_count), "\n")
cat("clip_cap_binding_count range:", range(clip_split$clip_cap_binding_count), "\n")

# Join to matrix and drop original aggregate
clip_split_matrix <- structure_matrix %>%
  dplyr::left_join(clip_split, by = "gene_id_clean") %>%
  dplyr::select(-clip_evidence_count)

cat("\nMatrix after split:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
```

# Add Individual Cap-Binding LFC Features

```{r cap_binding_lfc}
# 1. Lee integrated stress response — clip_lee_int_stress_lfc
lee_isr <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/integrated_stress_response.xlsx", sheet = "Table S2") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id)) %>%
  dplyr::select(gene_id_clean, clip_lee_int_stress_lfc = log2FoldChange) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Lee ISR genes:", nrow(lee_isr), "\n")

# 2. Lamper S3 — clip_lee_complete_media_lfc & clip_lee_glu_dep_lfc
lamper_s3 <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/lamper-s3.xlsx", sheet = "Supplementary Table 3") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id)) %>%
  dplyr::select(gene_id_clean,
                clip_lee_complete_media_lfc = log2FoldChange_CM,
                clip_lee_glu_dep_lfc = log2FoldChange_Glu,
                clip_lee_glu_vs_cm_enrichment = `Fold Enrichment (GLU/CM)`) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Lamper S3 genes:", nrow(lamper_s3), "\n")

# 3. Left-join onto clip_split_matrix (unbound genes keep NA)
clip_split_matrix <- clip_split_matrix %>%
  dplyr::left_join(lee_isr, by = "gene_id_clean") %>%
  dplyr::left_join(lamper_s3, by = "gene_id_clean")

cat("\nMatrix after adding LFC features:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
cat("clip_lee_int_stress_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_int_stress_lfc)), "\n")
cat("clip_lee_complete_media_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_complete_media_lfc)), "\n")
cat("clip_lee_glu_dep_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_glu_dep_lfc)), "\n")
cat("clip_lee_glu_vs_cm_enrichment non-NA:", sum(!is.na(clip_split_matrix$clip_lee_glu_vs_cm_enrichment)), "\n")
```

# Quality Control

```{r quality_control}
cat("clip_eif3_count distribution:\n")
print(table(clip_split_matrix$clip_eif3_count, useNA = "ifany"))

cat("\nclip_cap_binding_count distribution:\n")
print(table(clip_split_matrix$clip_cap_binding_count, useNA = "ifany"))

cat("\nSummary — clip_eif3_count:\n")
print(summary(clip_split_matrix$clip_eif3_count))

cat("\nSummary — clip_cap_binding_count:\n")
print(summary(clip_split_matrix$clip_cap_binding_count))

cat("\nSummary — clip_lee_int_stress_lfc:\n")
print(summary(clip_split_matrix$clip_lee_int_stress_lfc))

cat("\nSummary — clip_lee_complete_media_lfc:\n")
print(summary(clip_split_matrix$clip_lee_complete_media_lfc))

cat("\nSummary — clip_lee_glu_dep_lfc:\n")
print(summary(clip_split_matrix$clip_lee_glu_dep_lfc))

cat("\nSummary — clip_lee_glu_vs_cm_enrichment:\n")
print(summary(clip_split_matrix$clip_lee_glu_vs_cm_enrichment))

# Verify the two new features sum to the original clip_evidence_count
original <- structure_matrix %>%
  dplyr::left_join(clip_split, by = "gene_id_clean")
check <- all(
  (original$clip_eif3_count + original$clip_cap_binding_count) == original$clip_evidence_count,
  na.rm = TRUE
)
cat("\nVerification — split sums to original clip_evidence_count:", check, "\n")
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(clip_split_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_clip_split.rds\n")

write_csv(clip_split_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_clip_split.csv\n")

cat("\nFinal matrix:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
cat("New features: clip_eif3_count, clip_cap_binding_count, clip_lee_int_stress_lfc, clip_lee_complete_media_lfc, clip_lee_glu_dep_lfc, clip_lee_glu_vs_cm_enrichment\n")
cat("Removed: clip_evidence_count\n")
```

# Feature Catalog (clip_split + aggregate model)

This section documents every feature that enters the random forest model when
`feature_set = "clip_split"` and `clip_mode = "aggregate"` (as used in
`07_rf_classification_eif3d_targets_no_boruta.Rmd`). The filtering logic below
mirrors the preprocessing in that notebook.

```{r feature_catalog}
library(stringr)

# ── Replicate model-notebook filtering to get the exact feature set ──────────
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin")

all_cols <- colnames(clip_split_matrix)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
all_id_cols <- unique(c(id_cols, id_pattern_cols))
candidate_cols <- setdiff(all_cols, all_id_cols)

# clip_split feature set: remove k-mers and RBP motifs
kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
candidate_cols <- candidate_cols[!str_detect(candidate_cols, kmer_motif_pattern)]

# aggregate clip mode: keep only aggregate CLIP cols, drop regional binary cols
aggregate_clip_cols <- c("clip_eif3_count", "clip_cap_binding_count",
                         "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                         "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")
regional_clip <- candidate_cols[str_detect(candidate_cols, "^clip_") &
                                  !candidate_cols %in% aggregate_clip_cols]
model_feature_cols <- setdiff(candidate_cols, regional_clip)

cat("Model feature count (clip_split + aggregate):", length(model_feature_cols), "\n")

# ── Metadata for known / fixed features ─────────────────────────────────────
known_meta <- tribble(
  ~feature,                              ~category,           ~source,                                    ~type,      ~units,                        ~region,    ~description,

  # CLIP — split aggregate counts (from 01e)
  "clip_eif3_count",                     "CLIP Binding",      "scorecard_update.csv",                     "numeric",  "count (integer 0–3)",         "transcript", "Number of eIF3-specific CLIP datasets with binding evidence (lee_parclip_a_b_d_g, cate_undiff_npcs, cate_diff_npcs); 0 = unbound in all three",
  "clip_cap_binding_count",              "CLIP Binding",      "scorecard_update.csv",                     "numeric",  "count (integer 0–3)",         "transcript", "Number of cap-binding CLIP datasets with binding evidence (lee_int_stress, lee_complete_media, lee_glu_dep); 0 = unbound in all three",
  "clip_lee_int_stress_lfc",             "CLIP Binding",      "integrated_stress_response.xlsx (Lee)",    "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC under integrated stress response (Lee et al.); positive = upregulated under stress; NA for unbound genes",
  "clip_lee_complete_media_lfc",         "CLIP Binding",      "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC in complete media condition (Lamper S3); positive = higher translation",
  "clip_lee_glu_dep_lfc",               "CLIP Binding",      "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC under glucose deprivation (Lamper S3); positive = higher translation under deprivation",
  "clip_lee_glu_vs_cm_enrichment",       "CLIP Binding",      "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "fold enrichment",             "transcript", "Fold enrichment of glucose deprivation vs complete media (GLU/CM); >1 = relatively upregulated under deprivation",

  # Region lengths (from 01c)
  "utr5_length",                         "Region Length",     "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "5'UTR",    "Total 5' UTR length of the most abundant transcript (summed across exons)",
  "utr3_length",                         "Region Length",     "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "3'UTR",    "Total 3' UTR length of the most abundant transcript",
  "cds_length",                          "Region Length",     "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "CDS",      "Total CDS length of the most abundant transcript",

  # GC content (from 01c)
  "utr5_gc",                             "GC Content",        "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "5'UTR",    "GC nucleotide percentage of 5' UTR sequence",
  "utr3_gc",                             "GC Content",        "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "3'UTR",    "GC nucleotide percentage of 3' UTR sequence",
  "cds_gc",                              "GC Content",        "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "CDS",      "GC nucleotide percentage of CDS sequence",
  "gc3",                                 "GC Content",        "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "CDS",      "GC content at third codon positions (GC3); proxy for codon usage bias",

  # 4E / DAP5 (from 01c)
  "4e_resistance_score_lfc",             "4E / DAP5",         "teleman_riboseq_analyzed.csv",             "numeric",  "log2 fold change",            "transcript", "TE LFC under eIF4E inhibition (Teleman ribosome profiling); positive = translation maintained despite 4E inhibition (resistant)",
  "eif4e_ip_vehicle_lfc",               "4E / DAP5",         "res_eif4e_ip_vehicle.csv (Teleman)",       "numeric",  "log2 fold change",            "transcript", "eIF4E-IP enrichment in vehicle condition (baseline cap-binding); positive = enriched in eIF4E-IP pellet",
  "eif4e_ip_inhibitor_only_lfc",        "4E / DAP5",         "res_eif4e_ip_inhibitor_only.csv (Teleman)","numeric",  "log2 fold change",            "transcript", "eIF4E-IP enrichment under 4E inhibitor only; captures residual cap-binding under inhibition",
  "eif4e_ip_inhib_vs_vehicle_lfc",      "4E / DAP5",         "res_eif4e_ip_inhib_vs_vehicle.csv (Teleman)","numeric","log2 fold change",            "transcript", "Differential eIF4E-IP enrichment: inhibitor vs vehicle (interaction term); negative = lost cap-dependence under inhibition",
  "dap5_promoted_score_weber_lfc",      "4E / DAP5",         "dap5kd_riboseq_weber2022.xlsx",            "numeric",  "log2 fold change",            "transcript", "Negated TE LFC from DAP5 KO vs control (Weber 2022); positive = translation promoted by DAP5",
  "dap5_promoted_score_polysome_lfc",   "4E / DAP5",         "dap5_deseq2_full_results.csv (GSE115142)", "numeric",  "log2 fold change",            "transcript", "Polysome LFC from DAP5 KD vs control (GSE115142 DESeq2); positive = translation promoted by DAP5",

  # RNA structure (from 01d)
  "struct_accessibility_cap_proximal",  "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability at positions 1–30 (cap-proximal); higher = more accessible",
  "struct_accessibility_aug_context",   "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability at last 30 nt of 5'UTR (AUG-proximal); higher = more accessible",
  "struct_accessibility_utr5_mean",     "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability across full 5'UTR; higher = more accessible overall",
  "struct_accessibility_utr5_min",      "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Minimum single-nt unpaired probability in 5'UTR; captures the most structured position",
  "struct_accessibility_utr5_sd",       "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "SD of single-nt unpaired probability along 5'UTR; higher = more heterogeneous structure",
  "struct_num_structured_regions",      "RNA Structure",     "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "count (integer >= 0)",        "5'UTR",    "Number of 30-nt windows with 30-nt unpaired probability < 0.2 (highly structured windows)",
  "struct_mfe_utr5",                    "RNA Structure",     "RNAfold -p --noPS",                       "numeric",  "kcal/mol (<= 0)",             "5'UTR",    "Minimum free energy of MFE structure; more negative = more thermodynamically stable",
  "struct_mfe_per_nt_utr5",             "RNA Structure",     "RNAfold -p --noPS",                       "numeric",  "kcal/mol/nt (<= 0)",          "5'UTR",    "MFE normalized by 5'UTR length; enables cross-transcript comparison",
  "struct_ensemble_free_energy",        "RNA Structure",     "RNAfold -p --noPS",                       "numeric",  "kcal/mol (<= 0)",             "5'UTR",    "Ensemble free energy from partition function (-kT log Z); more negative = more structured ensemble",
  "struct_ensemble_diversity",          "RNA Structure",     "RNAfold -p --noPS",                       "numeric",  "nats (>= 0)",                 "5'UTR",    "Positional entropy of structural ensemble; higher = more disordered / heterogeneous set of structures",
  "struct_frac_paired_utr5",            "RNA Structure",     "RNAfold dot-bracket parsing",             "numeric",  "fraction (0–1)",              "5'UTR",    "Fraction of nucleotides in base pairs in MFE structure",
  "struct_num_hairpins_utr5",           "RNA Structure",     "RNAfold dot-bracket parsing",             "numeric",  "count (integer >= 0)",        "5'UTR",    "Number of hairpin loops in MFE structure",
  "struct_longest_stem_utr5",           "RNA Structure",     "RNAfold dot-bracket parsing",             "numeric",  "nucleotides (integer >= 0)",  "5'UTR",    "Length of longest consecutive stem (run of '(' chars) in MFE dot-bracket"
)

# ── Dynamically generate rows for lit_ and orfquant_ features ───────────────
lit_cols <- model_feature_cols[str_detect(model_feature_cols, "^lit_")]
lit_catalog <- tibble(
  feature     = lit_cols,
  category    = "Literature Signatures",
  source      = "translation_signatures_literature_1_9_25.csv",
  type        = "binary",
  units       = "0 / 1",
  region      = NA_character_,
  description = paste0("Binary membership in published gene set: \"",
                       str_replace_all(str_remove(lit_cols, "^lit_"), "_", " "), "\"")
)

orfquant_cols <- model_feature_cols[str_detect(model_feature_cols, "^orfquant_")]
orfquant_catalog <- tibble(
  feature     = orfquant_cols,
  category    = "ORFquant",
  source      = "orfquant_orf_types_normoxia_siCTRL.csv",
  type        = "binary",
  units       = "0 / 1",
  region      = "CDS",
  description = paste0("ORFquant ORF type detected in normoxia siCTRL condition: ",
                       str_remove(orfquant_cols, "^orfquant_"))
)

# ── Combine and verify coverage ──────────────────────────────────────────────
feature_catalog <- bind_rows(known_meta, lit_catalog, orfquant_catalog) %>%
  filter(feature %in% model_feature_cols) %>%
  arrange(category, feature)

# Flag any model features not yet described
missing_from_catalog <- setdiff(model_feature_cols, feature_catalog$feature)
if (length(missing_from_catalog) > 0) {
  warning("Features in model but missing from catalog: ",
          paste(missing_from_catalog, collapse = ", "))
}

cat("Catalog rows:", nrow(feature_catalog), "\n")
cat("Model features covered:", sum(model_feature_cols %in% feature_catalog$feature),
    "/", length(model_feature_cols), "\n")

cat("\nFeatures by category:\n")
print(feature_catalog %>% count(category))

write_csv(feature_catalog,
          here("output", "predictive_modeling", "feature_catalog_clip_split_aggregate.csv"))
cat("\nExported: output/predictive_modeling/feature_catalog_clip_split_aggregate.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
