---
title: "Split CLIP Aggregate Features (eIF3 vs Cap-Binding)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

The original `clip_evidence_count` feature (from 01c) sums binding evidence across
all 6 CLIP conditions into a single count (0–6). Three of those conditions
(`lee_int_stress`, `lee_complete_media`, `lee_glu_dep`) represent **cap-binding**
activity rather than eIF3-specific binding. This notebook splits the aggregate into
two features so the model can learn from them independently:

- `clip_eif3_count` — eIF3-specific binding (0–3)
- `clip_cap_binding_count` — cap-binding activity (0–3)

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
```

# Load Feature Matrix

```{r load_matrix}
structure_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_structure.rds"))
cat("Loaded structure matrix:", nrow(structure_matrix), "genes x", ncol(structure_matrix), "columns\n")
cat("clip_evidence_count present:", "clip_evidence_count" %in% colnames(structure_matrix), "\n")
```

# Compute Split CLIP Features

```{r compute_clip_split}
clip_update <- read_csv("/Users/katematlin/github/mpra_preliminary/scorecard_update.csv") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

eif3_conditions <- c("lee_parclip_a_b_d_g", "cate_undiff_npcs", "cate_diff_npcs")
cap_binding_conditions <- c("lee_int_stress", "lee_complete_media", "lee_glu_dep")

clip_split <- clip_update %>%
  dplyr::select(gene_id_clean, all_of(eif3_conditions), all_of(cap_binding_conditions)) %>%
  dplyr::mutate(
    clip_eif3_count = rowSums(
      across(all_of(eif3_conditions), ~as.integer(tolower(.) != "not bound")),
      na.rm = TRUE
    ),
    clip_cap_binding_count = rowSums(
      across(all_of(cap_binding_conditions), ~as.integer(tolower(.) != "not bound")),
      na.rm = TRUE
    )
  ) %>%
  dplyr::select(gene_id_clean, clip_eif3_count, clip_cap_binding_count)

cat("clip_eif3_count range:", range(clip_split$clip_eif3_count), "\n")
cat("clip_cap_binding_count range:", range(clip_split$clip_cap_binding_count), "\n")

# Join to matrix and drop original aggregate
clip_split_matrix <- structure_matrix %>%
  dplyr::left_join(clip_split, by = "gene_id_clean") %>%
  dplyr::select(-clip_evidence_count)

cat("\nMatrix after split:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
```

# Add Individual Cap-Binding LFC Features

```{r cap_binding_lfc}
# 1. Lee integrated stress response — clip_lee_int_stress_lfc
lee_isr <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/integrated_stress_response.xlsx", sheet = "Table S2") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id)) %>%
  dplyr::select(gene_id_clean, clip_lee_int_stress_lfc = log2FoldChange) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Lee ISR genes:", nrow(lee_isr), "\n")

# 2. Lamper S3 — clip_lee_complete_media_lfc & clip_lee_glu_dep_lfc
lamper_s3 <- read_xlsx("/Users/katematlin/github/mpra_preliminary/data/lamper-s3.xlsx", sheet = "Supplementary Table 3") %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id)) %>%
  dplyr::select(gene_id_clean,
                clip_lee_complete_media_lfc = log2FoldChange_CM,
                clip_lee_glu_dep_lfc = log2FoldChange_Glu,
                clip_lee_glu_vs_cm_enrichment = `Fold Enrichment (GLU/CM)`) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("Lamper S3 genes:", nrow(lamper_s3), "\n")

# 3. Left-join onto clip_split_matrix (unbound genes keep NA)
clip_split_matrix <- clip_split_matrix %>%
  dplyr::left_join(lee_isr, by = "gene_id_clean") %>%
  dplyr::left_join(lamper_s3, by = "gene_id_clean")

cat("\nMatrix after adding LFC features:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
cat("clip_lee_int_stress_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_int_stress_lfc)), "\n")
cat("clip_lee_complete_media_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_complete_media_lfc)), "\n")
cat("clip_lee_glu_dep_lfc non-NA:", sum(!is.na(clip_split_matrix$clip_lee_glu_dep_lfc)), "\n")
cat("clip_lee_glu_vs_cm_enrichment non-NA:", sum(!is.na(clip_split_matrix$clip_lee_glu_vs_cm_enrichment)), "\n")
```

# Quality Control

```{r quality_control}
cat("clip_eif3_count distribution:\n")
print(table(clip_split_matrix$clip_eif3_count, useNA = "ifany"))

cat("\nclip_cap_binding_count distribution:\n")
print(table(clip_split_matrix$clip_cap_binding_count, useNA = "ifany"))

cat("\nSummary — clip_eif3_count:\n")
print(summary(clip_split_matrix$clip_eif3_count))

cat("\nSummary — clip_cap_binding_count:\n")
print(summary(clip_split_matrix$clip_cap_binding_count))

cat("\nSummary — clip_lee_int_stress_lfc:\n")
print(summary(clip_split_matrix$clip_lee_int_stress_lfc))

cat("\nSummary — clip_lee_complete_media_lfc:\n")
print(summary(clip_split_matrix$clip_lee_complete_media_lfc))

cat("\nSummary — clip_lee_glu_dep_lfc:\n")
print(summary(clip_split_matrix$clip_lee_glu_dep_lfc))

cat("\nSummary — clip_lee_glu_vs_cm_enrichment:\n")
print(summary(clip_split_matrix$clip_lee_glu_vs_cm_enrichment))

# Verify the two new features sum to the original clip_evidence_count
original <- structure_matrix %>%
  dplyr::left_join(clip_split, by = "gene_id_clean")
check <- all(
  (original$clip_eif3_count + original$clip_cap_binding_count) == original$clip_evidence_count,
  na.rm = TRUE
)
cat("\nVerification — split sums to original clip_evidence_count:", check, "\n")
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(clip_split_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_clip_split.rds\n")

write_csv(clip_split_matrix, here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_1hr_clip_split.csv\n")

cat("\nFinal matrix:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
cat("New features: clip_eif3_count, clip_cap_binding_count, clip_lee_int_stress_lfc, clip_lee_complete_media_lfc, clip_lee_glu_dep_lfc, clip_lee_glu_vs_cm_enrichment\n")
cat("Removed: clip_evidence_count\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
