---
title: "TE LFC Feature Extraction (clip_split + sictrl TE dynamics)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook extends the `clip_split` feature matrix (from 01e) with 8 new
translational efficiency (TE) features derived from sictrl samples:

**From `output/condition_level_te.csv`** (condition × timepoint TE in sictrl):

- `indiv_te_normoxia_1hr`, `indiv_te_hypoxia_1hr` — log2(ribo/rna) at 1hr
- `indiv_te_normoxia_4hr`, `indiv_te_hypoxia_4hr` — log2(ribo/rna) at 4hr
- `indiv_te_normoxia_24hr`, `indiv_te_hypoxia_24hr` — log2(ribo/rna) at 24hr

**From `output/translation_categories_{timepoint}.csv`** (ΔTE hypoxia vs normoxia):

- `delta_te_4hr`, `delta_te_24hr`

Note: `delta_te_1hr` is excluded — it is identical to `te_lfc` (the model target)
and would constitute target leakage if included as a predictor.

The `te_lfc` feature set is a strict superset of `clip_split` — it inherits all
clip_split features and adds these 8 TE features on top.

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
```

# Load Feature Matrix

```{r load_matrix}
clip_split_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.rds"))
cat("Loaded clip_split matrix:", nrow(clip_split_matrix), "genes x", ncol(clip_split_matrix), "columns\n")
```

# Add Condition-Level TE Features (indiv_te_*)

Six features: absolute TE (log2 ribo/rna) in sictrl for each condition × timepoint.

```{r condition_level_te}
te_wide <- read_csv(here("output", "condition_level_te.csv")) %>%
  dplyr::filter(treatment == "sictrl") %>%
  dplyr::mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    col_name = paste0("indiv_te_", condition, "_", hours, "hr")
  ) %>%
  dplyr::select(gene_id_clean, col_name, log2FoldChange) %>%
  dplyr::distinct(gene_id_clean, col_name, .keep_all = TRUE) %>%
  tidyr::pivot_wider(names_from = col_name, values_from = log2FoldChange)

expected_cols <- c("indiv_te_normoxia_1hr", "indiv_te_hypoxia_1hr",
                   "indiv_te_normoxia_4hr", "indiv_te_hypoxia_4hr",
                   "indiv_te_normoxia_24hr", "indiv_te_hypoxia_24hr")
cat("Condition-level TE genes:", nrow(te_wide), "\n")
cat("Expected columns present:", all(expected_cols %in% colnames(te_wide)), "\n")
cat("Columns:", paste(colnames(te_wide)[-1], collapse = ", "), "\n")
```

# Add Delta TE Features (delta_te_*)

Two features: ΔTE (hypoxia vs normoxia) at 4hr and 24hr from sictrl.
`delta_te_1hr` is intentionally excluded — it is identical to `te_lfc` (the
1hr model target) and would cause target leakage.

```{r delta_te}
delta_te <- map(c("4hr", "24hr"), ~{
  read_csv(here("output", paste0("translation_categories_", .x, ".csv"))) %>%
    dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
    dplyr::select(gene_id_clean, !!paste0("delta_te_", .x) := te_lfc)
}) %>%
  purrr::reduce(full_join, by = "gene_id_clean")

cat("Delta TE genes:", nrow(delta_te), "\n")
cat("Columns:", paste(colnames(delta_te)[-1], collapse = ", "), "\n")
```

# Join to clip_split Matrix

```{r join_features}
te_lfc_matrix <- clip_split_matrix %>%
  left_join(te_wide, by = "gene_id_clean") %>%
  left_join(delta_te, by = "gene_id_clean")

cat("\nMatrix after joining TE features:", nrow(te_lfc_matrix), "genes x", ncol(te_lfc_matrix), "columns\n")
cat("New columns vs clip_split:", ncol(te_lfc_matrix) - ncol(clip_split_matrix), "\n")
```

# Quality Control

```{r quality_control}
new_features <- c(expected_cols, "delta_te_4hr", "delta_te_24hr")

cat("Non-NA counts for new TE features:\n")
for (col in new_features) {
  n_nonNA <- sum(!is.na(te_lfc_matrix[[col]]))
  cat(" ", col, ":", n_nonNA, "non-NA /", nrow(te_lfc_matrix), "\n")
}

cat("\nSummaries:\n")
for (col in new_features) {
  cat("\n", col, ":\n", sep = "")
  print(summary(te_lfc_matrix[[col]]))
}
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(te_lfc_matrix, here("output", "predictive_modeling", "feature_matrix_te_lfc.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_te_lfc.rds\n")

write_csv(te_lfc_matrix, here("output", "predictive_modeling", "feature_matrix_te_lfc.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_te_lfc.csv\n")

cat("\nFinal matrix:", nrow(te_lfc_matrix), "genes x", ncol(te_lfc_matrix), "columns\n")
cat("New features (8 total; delta_te_1hr excluded — identical to te_lfc target):\n")
cat(paste(new_features, collapse = ", "), "\n")
```

# Feature Catalog (te_lfc + aggregate model)

This section documents every feature that enters the random forest model when
`feature_set = "te_lfc"` and `clip_mode = "aggregate"` (as used in
`07_rf_classification_eif3d_targets_no_boruta.Rmd`). The filtering logic below
mirrors the preprocessing in that notebook.

```{r feature_catalog}
library(stringr)

# ── Replicate model-notebook filtering to get the exact feature set ──────────
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin")

all_cols <- colnames(te_lfc_matrix)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
all_id_cols <- unique(c(id_cols, id_pattern_cols))
candidate_cols <- setdiff(all_cols, all_id_cols)

# te_lfc feature set: remove k-mers and RBP motifs
kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
candidate_cols <- candidate_cols[!str_detect(candidate_cols, kmer_motif_pattern)]

# aggregate clip mode: keep only aggregate CLIP cols, drop regional binary cols
aggregate_clip_cols <- c("clip_eif3_count", "clip_cap_binding_count",
                         "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                         "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")
regional_clip <- candidate_cols[str_detect(candidate_cols, "^clip_") &
                                  !candidate_cols %in% aggregate_clip_cols]
model_feature_cols <- setdiff(candidate_cols, regional_clip)

cat("Model feature count (te_lfc + aggregate):", length(model_feature_cols), "\n")

# ── Metadata for known / fixed features ─────────────────────────────────────
known_meta <- tribble(
  ~feature,                              ~category,                    ~source,                                    ~type,      ~units,                        ~region,    ~description,

  # CLIP — split aggregate counts (from 01e)
  "clip_eif3_count",                     "CLIP Binding",               "scorecard_update.csv",                     "numeric",  "count (integer 0–3)",         "transcript", "Number of eIF3-specific CLIP datasets with binding evidence (lee_parclip_a_b_d_g, cate_undiff_npcs, cate_diff_npcs); 0 = unbound in all three",
  "clip_cap_binding_count",              "CLIP Binding",               "scorecard_update.csv",                     "numeric",  "count (integer 0–3)",         "transcript", "Number of cap-binding CLIP datasets with binding evidence (lee_int_stress, lee_complete_media, lee_glu_dep); 0 = unbound in all three",
  "clip_lee_int_stress_lfc",             "CLIP Binding",               "integrated_stress_response.xlsx (Lee)",    "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC under integrated stress response (Lee et al.); positive = upregulated under stress; NA for unbound genes",
  "clip_lee_complete_media_lfc",         "CLIP Binding",               "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC in complete media condition (Lamper S3); positive = higher translation",
  "clip_lee_glu_dep_lfc",               "CLIP Binding",               "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "log2 fold change",            "transcript", "Ribosome footprinting LFC under glucose deprivation (Lamper S3); positive = higher translation under deprivation",
  "clip_lee_glu_vs_cm_enrichment",       "CLIP Binding",               "lamper-s3.xlsx (Lamper/Lee)",              "numeric",  "fold enrichment",             "transcript", "Fold enrichment of glucose deprivation vs complete media (GLU/CM); >1 = relatively upregulated under deprivation",

  # Region lengths (from 01c)
  "utr5_length",                         "Region Length",              "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "5'UTR",    "Total 5' UTR length of the most abundant transcript (summed across exons)",
  "utr3_length",                         "Region Length",              "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "3'UTR",    "Total 3' UTR length of the most abundant transcript",
  "cds_length",                          "Region Length",              "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "nucleotides (integer >= 0)",  "CDS",      "Total CDS length of the most abundant transcript",

  # GC content (from 01c)
  "utr5_gc",                             "GC Content",                 "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "5'UTR",    "GC nucleotide percentage of 5' UTR sequence",
  "utr3_gc",                             "GC Content",                 "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "3'UTR",    "GC nucleotide percentage of 3' UTR sequence",
  "cds_gc",                              "GC Content",                 "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "CDS",      "GC nucleotide percentage of CDS sequence",
  "gc3",                                 "GC Content",                 "GENCODE v49 TxDb / BSgenome.Hsapiens.hg38","numeric",  "percent (0–100)",             "CDS",      "GC content at third codon positions (GC3); proxy for codon usage bias",

  # 4E / DAP5 (from 01c)
  "4e_resistance_score_lfc",             "4E / DAP5",                  "teleman_riboseq_analyzed.csv",             "numeric",  "log2 fold change",            "transcript", "TE LFC under eIF4E inhibition (Teleman ribosome profiling); positive = translation maintained despite 4E inhibition (resistant)",
  "eif4e_ip_vehicle_lfc",               "4E / DAP5",                  "res_eif4e_ip_vehicle.csv (Teleman)",       "numeric",  "log2 fold change",            "transcript", "eIF4E-IP enrichment in vehicle condition (baseline cap-binding); positive = enriched in eIF4E-IP pellet",
  "eif4e_ip_inhibitor_only_lfc",        "4E / DAP5",                  "res_eif4e_ip_inhibitor_only.csv (Teleman)","numeric",  "log2 fold change",            "transcript", "eIF4E-IP enrichment under 4E inhibitor only; captures residual cap-binding under inhibition",
  "eif4e_ip_inhib_vs_vehicle_lfc",      "4E / DAP5",                  "res_eif4e_ip_inhib_vs_vehicle.csv (Teleman)","numeric","log2 fold change",            "transcript", "Differential eIF4E-IP enrichment: inhibitor vs vehicle (interaction term); negative = lost cap-dependence under inhibition",
  "dap5_promoted_score_weber_lfc",      "4E / DAP5",                  "dap5kd_riboseq_weber2022.xlsx",            "numeric",  "log2 fold change",            "transcript", "Negated TE LFC from DAP5 KO vs control (Weber 2022); positive = translation promoted by DAP5",
  "dap5_promoted_score_polysome_lfc",   "4E / DAP5",                  "dap5_deseq2_full_results.csv (GSE115142)", "numeric",  "log2 fold change",            "transcript", "Polysome TE LFC: NS control vs DAP5 shKD (GSE115142 DESeq2, conditionNS.fractionHD); positive = higher TE in control = translation promoted by DAP5",

  # RNA structure (from 01d)
  "struct_accessibility_cap_proximal",  "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability at positions 1–30 (cap-proximal); higher = more accessible",
  "struct_accessibility_aug_context",   "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability at last 30 nt of 5'UTR (AUG-proximal); higher = more accessible",
  "struct_accessibility_utr5_mean",     "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Mean single-nt unpaired probability across full 5'UTR; higher = more accessible overall",
  "struct_accessibility_utr5_min",      "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "Minimum single-nt unpaired probability in 5'UTR; captures the most structured position",
  "struct_accessibility_utr5_sd",       "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "probability (0–1)",           "5'UTR",    "SD of single-nt unpaired probability along 5'UTR; higher = more heterogeneous structure",
  "struct_num_structured_regions",      "RNA Structure",              "RNAplfold (-W 80 -L 40 -u 30)",           "numeric",  "count (integer >= 0)",        "5'UTR",    "Number of 30-nt windows with 30-nt unpaired probability < 0.2 (highly structured windows)",
  "struct_mfe_utr5",                    "RNA Structure",              "RNAfold -p --noPS",                       "numeric",  "kcal/mol (<= 0)",             "5'UTR",    "Minimum free energy of MFE structure; more negative = more thermodynamically stable",
  "struct_mfe_per_nt_utr5",             "RNA Structure",              "RNAfold -p --noPS",                       "numeric",  "kcal/mol/nt (<= 0)",          "5'UTR",    "MFE normalized by 5'UTR length; enables cross-transcript comparison",
  "struct_ensemble_free_energy",        "RNA Structure",              "RNAfold -p --noPS",                       "numeric",  "kcal/mol (<= 0)",             "5'UTR",    "Ensemble free energy from partition function (-kT log Z); more negative = more structured ensemble",
  "struct_ensemble_diversity",          "RNA Structure",              "RNAfold -p --noPS",                       "numeric",  "nats (>= 0)",                 "5'UTR",    "Positional entropy of structural ensemble; higher = more disordered / heterogeneous set of structures",
  "struct_frac_paired_utr5",            "RNA Structure",              "RNAfold dot-bracket parsing",             "numeric",  "fraction (0–1)",              "5'UTR",    "Fraction of nucleotides in base pairs in MFE structure",
  "struct_num_hairpins_utr5",           "RNA Structure",              "RNAfold dot-bracket parsing",             "numeric",  "count (integer >= 0)",        "5'UTR",    "Number of hairpin loops in MFE structure",
  "struct_longest_stem_utr5",           "RNA Structure",              "RNAfold dot-bracket parsing",             "numeric",  "nucleotides (integer >= 0)",  "5'UTR",    "Length of longest consecutive stem (run of '(' chars) in MFE dot-bracket",

  # Translational Efficiency (from 01f — new in te_lfc feature set)
  "indiv_te_normoxia_1hr",              "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl normoxia at 1hr; from DESeq2 model in individual_te_per_sample.Rmd",
  "indiv_te_hypoxia_1hr",               "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl hypoxia at 1hr",
  "indiv_te_normoxia_4hr",              "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl normoxia at 4hr",
  "indiv_te_hypoxia_4hr",               "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl hypoxia at 4hr",
  "indiv_te_normoxia_24hr",             "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl normoxia at 24hr",
  "indiv_te_hypoxia_24hr",              "Translational Efficiency",   "condition_level_te.csv",                   "numeric",  "log2 fold change",            "transcript", "Absolute TE (log2 ribo/rna) in sictrl hypoxia at 24hr",
  "delta_te_4hr",                       "Translational Efficiency",   "translation_categories_4hr.csv",           "numeric",  "log2 fold change",            "transcript", "ΔTE (hypoxia vs normoxia) at 4hr in sictrl; from rna_ribo_te_analysis.Rmd DESeq2 interaction model",
  "delta_te_24hr",                      "Translational Efficiency",   "translation_categories_24hr.csv",          "numeric",  "log2 fold change",            "transcript", "ΔTE (hypoxia vs normoxia) at 24hr in sictrl"
)

# ── Dynamically generate rows for lit_ and orfquant_ features ───────────────
lit_cols <- model_feature_cols[str_detect(model_feature_cols, "^lit_")]
lit_catalog <- tibble(
  feature     = lit_cols,
  category    = "Literature Signatures",
  source      = "translation_signatures_literature_1_9_25.csv",
  type        = "binary",
  units       = "0 / 1",
  region      = NA_character_,
  description = paste0("Binary membership in published gene set: \"",
                       str_replace_all(str_remove(lit_cols, "^lit_"), "_", " "), "\"")
)

orfquant_cols <- model_feature_cols[str_detect(model_feature_cols, "^orfquant_")]
orfquant_catalog <- tibble(
  feature     = orfquant_cols,
  category    = "ORFquant",
  source      = "orfquant_orf_types_normoxia_siCTRL.csv",
  type        = "binary",
  units       = "0 / 1",
  region      = "CDS",
  description = paste0("ORFquant ORF type detected in normoxia siCTRL condition: ",
                       str_remove(orfquant_cols, "^orfquant_"))
)

# ── Combine and verify coverage ──────────────────────────────────────────────
feature_catalog <- bind_rows(known_meta, lit_catalog, orfquant_catalog) %>%
  filter(feature %in% model_feature_cols) %>%
  arrange(category, feature)

# Flag any model features not yet described
missing_from_catalog <- setdiff(model_feature_cols, feature_catalog$feature)
if (length(missing_from_catalog) > 0) {
  warning("Features in model but missing from catalog: ",
          paste(missing_from_catalog, collapse = ", "))
}

cat("Catalog rows:", nrow(feature_catalog), "\n")
cat("Model features covered:", sum(model_feature_cols %in% feature_catalog$feature),
    "/", length(model_feature_cols), "\n")

cat("\nFeatures by category:\n")
print(feature_catalog %>% dplyr::count(category))

write_csv(feature_catalog,
          here("output", "predictive_modeling", "feature_catalog_te_lfc.csv"))
cat("\nExported: output/predictive_modeling/feature_catalog_te_lfc.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
