---
title: "Codon Optimality Feature Extraction (CAI, tAI, CSC)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This notebook adds three mechanistically distinct codon optimality features to the
`te_lfc` feature matrix, producing `feature_matrix_codon_optimality.rds`.

| Feature | Tool | Description |
|---------|------|-------------|
| `cai` | coRdon (Bioconductor) | Codon Adaptive Index — codon usage vs. ribosomal protein reference |
| `tai` | coRdon (Bioconductor) | tRNA Adaptation Index — anticodon supply from GtRNAdb hg38 tRNA GCN |
| `csc` | iCodon (Diez et al. 2022) | Codon Stabilisation Coefficient — codon-level mRNA stability weights |

All three are CDS-level, length-normalized, and orthogonal to `gc3`.

**Prerequisites before knitting:**
- `output/predictive_modeling/feature_matrix_te_lfc.rds` (from 01f)
- `accessories/human/txdb.gencode49.sqlite` (GENCODE v49)
- `accessories/human/human_trna_gcn.csv` — 64-row table from GtRNAdb (hg38);
  columns: `anticodon` (string, e.g. "AAA"), `gcn` (integer gene copy number)
- R packages: `coRdon` (Bioconductor), `iCodon` (CRAN/GitHub)

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(coRdon)
library(iCodon)
```

# Load Feature Matrix

```{r load_feature_matrix}
te_lfc_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_te_lfc.rds"))
cat("Loaded te_lfc matrix:", nrow(te_lfc_matrix), "genes x", ncol(te_lfc_matrix), "columns\n")

keep_tx <- te_lfc_matrix$transcript_id_clean
cat("Target transcripts:", length(keep_tx), "\n")
cat("Example IDs:", head(keep_tx, 3), "\n")
```

# Load TxDb and Genome

```{r load_txdb}
txdb  <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
cat("Loaded TxDb and BSgenome\n")
```

# Extract CDS Sequences

```{r extract_cds}
# Extract CDS for all transcripts
cds <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_sequences_all <- extractTranscriptSeqs(genome, cds)
cat("Total CDS sequences in TxDb:", length(cds_sequences_all), "\n")

# Strip version numbers from TxDb transcript IDs to match feature matrix
cds_names_clean <- sub("\\..*", "", names(cds_sequences_all))

# Filter to transcripts present in the feature matrix
keep_idx <- cds_names_clean %in% keep_tx
cds_sequences <- cds_sequences_all[keep_idx]
names(cds_sequences) <- cds_names_clean[keep_idx]
cat("After filtering to feature matrix transcripts:", length(cds_sequences), "\n")

# coRdon requires CDS lengths divisible by 3
valid_length <- width(cds_sequences) %% 3 == 0
n_invalid <- sum(!valid_length)
if (n_invalid > 0) {
  cat("Removing", n_invalid, "sequences with length not divisible by 3\n")
}
cds_sequences <- cds_sequences[valid_length]
cat("Final CDS sequences for codon scoring:", length(cds_sequences), "\n")
cat("Example names:", head(names(cds_sequences), 3), "\n")
```

# Build Codon Table

```{r codon_table}
ct <- codonTable(cds_sequences)
cat("codonTable built:", nrow(codonCounts(ct)), "sequences\n")
```

# CAI — Codon Adaptive Index

Reference set: ribosomal protein genes (RPL*/RPS*), the standard practice for
computing relative adaptiveness weights in human cells.

Implements Sharp & Li (1987) directly using coRdon's `codonCounts()`:

1. Sum raw codon counts across ribosomal reference genes
2. Per synonymous family: w_i = RSCU_i / max(RSCU_j), capped at 1
3. Per gene: CAI = geometric mean of w_i^count_i, excluding stop codons

This guarantees CAI ∈ [0, 1] regardless of reference set composition.

```{r cai}
# Identify ribosomal protein transcripts
ribo_tx <- te_lfc_matrix %>%
  filter(str_detect(symbol, "^RPL|^RPS")) %>%
  pull(transcript_id_clean)

ribo_subset <- names(cds_sequences) %in% ribo_tx
cat("Ribosomal protein transcripts in reference set:", sum(ribo_subset), "\n")
cat("  (out of", length(ribo_subset), "total sequences)\n")

# Raw codon counts (n_seqs × 64 codons) from coRdon
counts_mat <- codonCounts(ct)          # rows = sequences, cols = codons
cat("Codon count matrix:", nrow(counts_mat), "×", ncol(counts_mat), "\n")

# Aggregate counts across ribosomal reference genes
ref_counts <- colSums(counts_mat[ribo_subset, , drop = FALSE])

# Build relative adaptiveness (w) from standard genetic code
# Biostrings::getGeneticCode returns named char vector codon → amino acid
gen_code <- Biostrings::getGeneticCode("SGC0")  # standard code

# For each synonymous family, w_i = RSCU_i / max(RSCU_j), capped at [0,1]
stop_codons <- names(gen_code[gen_code == "*"])
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

w <- setNames(rep(NA_real_, length(gen_code)), names(gen_code))
w[stop_codons] <- 0  # stop codons excluded

for (aa in names(aa_to_codons)) {
  codons <- aa_to_codons[[aa]]
  counts <- ref_counts[codons]
  total  <- sum(counts)

  if (total == 0) {
    # No reference observations: treat as uniform (w = 1 for all)
    w[codons] <- 1
    next
  }

  n_syn  <- length(codons)
  # RSCU_i = count_i / (total / n_synonymous)
  rscu   <- counts / (total / n_syn)
  max_r  <- max(rscu)

  if (max_r == 0) {
    w[codons] <- 0
  } else {
    w[codons] <- pmin(rscu / max_r, 1)  # cap at 1
  }
}

cat("w values: min =", round(min(w, na.rm = TRUE), 4),
    " max =", round(max(w, na.rm = TRUE), 4), "\n")

# Compute CAI for each gene as geometric mean of w^count
# log(CAI) = sum_i( count_i * log(w_i) ) / sum_i(count_i), over sense codons
compute_cai_gene <- function(gene_counts, w_vals) {
  # Exclude stops and zero-w codons (singletons like Met/Trp have w=1 so they don't inflate)
  use <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use <- gene_counts[use]
  w_use <- w_vals[use]
  n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_use)) / n_total)
}

cai_scores <- apply(counts_mat, 1, compute_cai_gene, w_vals = w)
names(cai_scores) <- names(cds_sequences)

cat("\nCAI summary (Sharp & Li 1987, ribosomal reference):\n")
print(summary(cai_scores))
stopifnot("CAI must be in [0,1]" = all(cai_scores <= 1 + 1e-9, na.rm = TRUE))
cat("Range check passed: all CAI ≤ 1\n")
```

# tAI — tRNA Adaptation Index

Implements dos Reis et al. (2004) tAI directly using:
- `codonCounts(ct)` from coRdon (defined in the CAI chunk)
- `accessories/human/human_trna_gcn.csv`: GtRNAdb hg38 high-confidence anticodon GCN
  (Sudmant et al. 2014, tRNAscan-SE counts; suppressor/SeC anticodons excluded)

Eukaryotic wobble rules (sking = 0, dos Reis 2004):
- A at anticodon position 34 → Inosine in mature tRNA; s = 0 for I:U, I:C, I:A
- G at position 34: s = 0 (G:C WC), s = 0.41 (G:U wobble)
- C or T at position 34: perfect WC only (s = 0)

Amino-acid filtering ensures anticodon-codon contributions stay within synonymous families.

```{r tai}
trna_gcn <- read_csv(here("accessories", "human", "human_trna_gcn.csv"),
                     comment = "#")
cat("tRNA GCN table:", nrow(trna_gcn), "rows\n")
cat("Anticodons:", paste(trna_gcn$anticodon, collapse = ", "), "\n")

# Decode which codons each anticodon recognizes and with what wobble penalty
# anticodon 5'->3' DNA: X1=wobble pos (34), X2, X3
# codon 5'->3': Y1 (WC with X3), Y2 (WC with X2), Y3 (wobble with X1)
decode_anticodon <- function(ac) {
  x1 <- substr(ac, 1, 1)
  x2 <- substr(ac, 2, 2)
  x3 <- substr(ac, 3, 3)
  wc <- c(A = "T", T = "A", G = "C", C = "G")
  y1 <- unname(wc[x3])
  y2 <- unname(wc[x2])
  if (x1 == "A") {          # A → Inosine; I:U(T), I:C, I:A all s=0
    y3  <- c("T", "C", "A")
    svs <- c(0, 0, 0)
  } else if (x1 == "G") {   # G:C WC s=0; G:U(T) wobble s=0.41
    y3  <- c("C", "T")
    svs <- c(0, 0.41)
  } else if (x1 == "C") {   # C:G WC only
    y3  <- "G"
    svs <- 0
  } else {                   # T(=U):A WC only
    y3  <- "A"
    svs <- 0
  }
  data.frame(codon = paste0(y1, y2, y3), s = svs, stringsAsFactors = FALSE)
}

# gen_code and sense_codons defined in CAI chunk above; reuse here
aa_of_codon <- setNames(as.character(gen_code), names(gen_code))

# Accumulate ws(codon) = sum_anticodons (1-s) * GCN, filtered to same amino acid
ws_tai <- setNames(rep(0, length(gen_code)), names(gen_code))

for (i in seq_len(nrow(trna_gcn))) {
  ac  <- trna_gcn$anticodon[i]
  aa  <- trna_gcn$amino_acid[i]   # 1-letter IUPAC, matches gen_code values
  gcn <- trna_gcn$gcn[i]

  decoded <- decode_anticodon(ac)
  for (j in seq_len(nrow(decoded))) {
    codon <- decoded$codon[j]
    s_val <- decoded$s[j]
    # Only credit sense codons encoding the same amino acid as this tRNA
    if (!is.na(aa_of_codon[codon]) && aa_of_codon[codon] == aa) {
      ws_tai[codon] <- ws_tai[codon] + (1 - s_val) * gcn
    }
  }
}

cat("Sense codons with ws=0:", sum(ws_tai[sense_codons] == 0), "\n")
cat("ws range:", round(range(ws_tai[sense_codons]), 3), "\n")

# Normalize: w(codon) = ws / max(ws) across all sense codons
w_tai <- ws_tai / max(ws_tai[sense_codons])
cat("w_tai range:", round(range(w_tai[sense_codons]), 4), "\n")

# tAI for each gene = geometric mean of w_i^count_i (log scale, sense codons only)
compute_tai_gene <- function(gene_counts, w_vals) {
  use     <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use   <- gene_counts[use]
  w_use   <- w_vals[use]
  n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_use)) / n_total)
}

tai_scores <- apply(counts_mat, 1, compute_tai_gene, w_vals = w_tai)
names(tai_scores) <- names(cds_sequences)

cat("\ntAI summary (dos Reis 2004, eukaryotic wobble):\n")
print(summary(tai_scores))
stopifnot("tAI must be in [0,1]" = all(tai_scores <= 1 + 1e-9, na.rm = TRUE))
cat("Range check passed: all tAI ≤ 1\n")
```

# CSC — Codon Stabilisation Coefficient

`iCodon::predict_stability()` is a function factory: call it with `"human"` to
get a predictor, then apply that predictor to a character vector of CDS sequences.
It returns mRNA stability scores derived from human CSC weights (Diez et al. 2022,
*Scientific Reports*). No external file needed.

```{r csc}
# Build human-specific predictor
predict_human <- iCodon::predict_stability("human")

# iCodon requires unique sequences — deduplicate, score, then map back by content
cds_char <- as.character(cds_sequences)
cat("Total sequences:", length(cds_char), "\n")

unique_seqs <- unique(cds_char)
cat("Unique sequences:", length(unique_seqs), "\n")
cat("Duplicates removed:", length(cds_char) - length(unique_seqs), "\n")

# Score unique sequences
unique_scores <- predict_human(unique_seqs)
names(unique_scores) <- unique_seqs   # key by sequence content

# Map back to all transcripts (duplicates get the same score)
csc_scores <- unique_scores[cds_char]
names(csc_scores) <- names(cds_sequences)

cat("CSC summary (iCodon human stability prediction):\n")
print(summary(csc_scores))
cat("Expected range: unbounded (higher = more stable codon usage)\n")
```

# Create Score Table

```{r create_score_table}
codon_scores <- tibble(
  transcript_id_clean = names(cds_sequences),
  cai = cai_scores,
  tai = tai_scores,
  csc = csc_scores
)

cat("Score table:", nrow(codon_scores), "rows\n")
cat("Non-NA: cai =", sum(!is.na(codon_scores$cai)),
    ", tai =", sum(!is.na(codon_scores$tai)),
    ", csc =", sum(!is.na(codon_scores$csc)), "\n")
```

# Join to Feature Matrix

```{r join_matrix}
codon_matrix <- te_lfc_matrix %>%
  left_join(codon_scores, by = "transcript_id_clean")

cat("Matrix after join:", nrow(codon_matrix), "genes x", ncol(codon_matrix), "columns\n")
cat("New columns:", ncol(codon_matrix) - ncol(te_lfc_matrix), "\n")
cat("Row count matches original:", nrow(codon_matrix) == nrow(te_lfc_matrix), "\n")
```

# Quality Control

```{r quality_control}
new_features <- c("cai", "tai", "csc")

cat("Non-NA counts for codon optimality features:\n")
for (col in new_features) {
  n_nonNA <- sum(!is.na(codon_matrix[[col]]))
  cat(" ", col, ":", n_nonNA, "non-NA /", nrow(codon_matrix),
      sprintf("(%.1f%%)\n", 100 * n_nonNA / nrow(codon_matrix)))
}

cat("\nValue ranges:\n")
for (col in new_features) {
  r <- range(codon_matrix[[col]], na.rm = TRUE)
  cat(" ", col, ": [", round(r[1], 4), ",", round(r[2], 4), "]\n")
}

cat("\nSummaries:\n")
for (col in new_features) {
  cat("\n", col, ":\n", sep = "")
  print(summary(codon_matrix[[col]]))
}

# Sanity checks
stopifnot(
  "CAI values must be in [0, 1]" = all(codon_matrix$cai >= 0 & codon_matrix$cai <= 1, na.rm = TRUE),
  "tAI values must be in [0, 1]" = all(codon_matrix$tai >= 0 & codon_matrix$tai <= 1, na.rm = TRUE)
)
cat("\nSanity checks passed: CAI and tAI in [0, 1]\n")
```

# Correlation Among Codon Optimality Features

```{r feature_correlations}
# Correlations with each other and with gc3 proxy
cor_features <- c("cai", "tai", "csc", "gc3")
cor_data <- codon_matrix %>%
  dplyr::select(all_of(cor_features)) %>%
  drop_na()

cat("Pairwise Pearson correlations (n =", nrow(cor_data), ")\n")
print(round(cor(cor_data), 3))
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(codon_matrix,
        here("output", "predictive_modeling", "feature_matrix_codon_optimality.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_codon_optimality.rds\n")

write_csv(codon_matrix,
          here("output", "predictive_modeling", "feature_matrix_codon_optimality.csv"))
cat("Exported: output/predictive_modeling/feature_matrix_codon_optimality.csv\n")

cat("\nFinal matrix:", nrow(codon_matrix), "genes x", ncol(codon_matrix), "columns\n")
cat("New features:", paste(new_features, collapse = ", "), "\n")
```

# Feature Catalog

```{r feature_catalog}
codon_catalog <- tribble(
  ~feature, ~category,          ~source,                                                     ~type,    ~units,                       ~region, ~description,

  "cai",    "Codon Optimality", "coRdon (Bioconductor) + RPL/RPS ribosomal protein ref set", "numeric", "0–1",                        "CDS",   "Codon Adaptive Index (CAI); measures translation efficiency from codon usage relative to highly expressed ribosomal protein genes (RPL*/RPS*); 1 = fully optimal codon usage; computed with coRdon::CAI()",

  "tai",    "Codon Optimality", "coRdon (Bioconductor) + GtRNAdb hg38 tRNA gene copy numbers", "numeric", "0–1",                      "CDS",   "tRNA Adaptation Index (tAI); measures codon-anticodon adaptation using human tRNA gene copy numbers from GtRNAdb (hg38), with wobble base pairing weights; higher = better tRNA supply match; computed with coRdon::tAI()",

  "csc",    "Codon Optimality", "iCodon R package (Diez et al. 2022, Sci Rep)",               "numeric", "unbounded (higher = stable)", "CDS",  "CSC-based mRNA stability prediction; computed with iCodon::predict_stability('human'); derived from human codon stabilization coefficient (CSC) weights; higher = more stable codon usage"
)

cat("Feature catalog rows:", nrow(codon_catalog), "\n")
print(codon_catalog %>% dplyr::select(feature, category, units))

write_csv(codon_catalog,
          here("output", "predictive_modeling", "feature_catalog_codon_optimality.csv"))
cat("\nExported: output/predictive_modeling/feature_catalog_codon_optimality.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
