---
title: "Notebook 01i: Feature Extraction — External mRNA Stability (Hia 2026, Karner 2026)"
output: html_document
date: "`r Sys.Date()`"
---

Extends `feature_matrix_cnot3.rds` with per-gene features from two external datasets:

- **Hia et al. 2026**: DHX29 selective ribosome profiling (occupancy score) and SLAM-seq
  mRNA half-lives (control half-life and DHX29 KO / ctrl log2FC)
- **Karner et al. 2026 (GSE301426)**: MDA-MB-231 SLAM-seq C-to-T conversion ratios
  (transcription rate proxy)

Also renames two existing CNOT3 (Zhu et al. 2024) half-life columns for clarity.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
```

## Load base feature matrix

```{r load_base}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_cnot3.rds"))
cat("Base matrix:", nrow(fm), "rows ×", ncol(fm), "cols\n")
```

## Add CNOT3 half-life columns (Zhu et al. 2024) with standardised names

Always joins fresh from `cnot3_slamseq_halflives.rds`. Drops legacy column names if
they happen to already be present in the base matrix.

```{r add_cnot3_halflives}
cnot3_hl <- readRDS(
  here("output", "predictive_modeling", "cnot3_slamseq_halflives.rds")
) %>%
  dplyr::select(
    ensembl_gene,
    zhu2024_ctrl_halflife          = sgNT_halflife,
    zhu2024_cnot3ko_halflife_logfc = log2_ratio
  )

# Drop old-named columns if present, then join with new names
fm <- fm %>%
  dplyr::select(-any_of(c("cnot3_slamseq_sgNT_halflife", "cnot3_slamseq_log2_ratio"))) %>%
  left_join(cnot3_hl, by = c("gene_id_clean" = "ensembl_gene"))

cat("zhu2024_ctrl_halflife non-NA:",
    sum(!is.na(fm$zhu2024_ctrl_halflife)), "/", nrow(fm), "\n")
cat("zhu2024_cnot3ko_halflife_logfc non-NA:",
    sum(!is.na(fm$zhu2024_cnot3ko_halflife_logfc)), "/", nrow(fm), "\n")
```

## Load DHX29 SLAM-seq features (Hia et al. 2026)

```{r load_dhx29_slam}
dhx29_slam <- readRDS(
  here("output", "predictive_modeling", "dhx29_slamseq_halflives.rds")
) %>%
  dplyr::select(
    ensembl_gene,
    hia2026_ctrl_halflife        = Ctrl_Half_life,
    hia2026_dhx29_halflife_logfc = logFC
  )

cat("DHX29 SLAM-seq genes:", nrow(dhx29_slam), "\n")
cat("hia2026_ctrl_halflife non-NA:",
    sum(!is.na(dhx29_slam$hia2026_ctrl_halflife)), "\n")
cat("hia2026_dhx29_halflife_logfc non-NA:",
    sum(!is.na(dhx29_slam$hia2026_dhx29_halflife_logfc)), "\n")
```

## Load DHX29 riboseq occupancy (Hia et al. 2026) — collapse transcript → gene

```{r load_dhx29_ribo}
dhx29_ribo <- readRDS(
  here("output", "predictive_modeling", "dhx29_riboseq_occupancy.rds")
) %>%
  group_by(ensembl_gene) %>%
  summarise(
    hia2026_dhx29_occupancy = mean(dhx29_occupancy_score, na.rm = TRUE),
    .groups = "drop"
  )

cat("DHX29 riboseq genes (after gene-level collapse):", nrow(dhx29_ribo), "\n")
cat("hia2026_dhx29_occupancy non-NA:",
    sum(!is.na(dhx29_ribo$hia2026_dhx29_occupancy)), "\n")
```

## Load MDA-MB-231 C-to-T ratio (Karner et al. 2026)

```{r load_karner}
karner <- readRDS(
  here("output", "predictive_modeling", "goodarzi_231_ct_ratios.rds")
) %>%
  dplyr::select(ensembl_gene, karner2026_mda231_log2_ct = log2_ct)

cat("Karner 2026 genes:", nrow(karner), "\n")
cat("karner2026_mda231_log2_ct non-NA:",
    sum(!is.na(karner$karner2026_mda231_log2_ct)), "\n")
```

## Join all new features

Feature matrix uses `gene_id_clean` as the versionless Ensembl ID join key.

```{r join_features}
fm_ext <- fm %>%
  left_join(dhx29_slam, by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(dhx29_ribo,  by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(karner,      by = c("gene_id_clean" = "ensembl_gene"))

cat("After join:", nrow(fm_ext), "rows ×", ncol(fm_ext), "cols\n")
cat("New columns added:", ncol(fm_ext) - ncol(fm), "\n\n")

new_cols <- c("hia2026_ctrl_halflife", "hia2026_dhx29_halflife_logfc",
              "hia2026_dhx29_occupancy", "karner2026_mda231_log2_ct")
cat("Coverage within feature matrix genes:\n")
for (col in new_cols) {
  cat(" ", col, ":", sum(!is.na(fm_ext[[col]])), "/", nrow(fm_ext), "\n")
}
```

## Save

```{r save}
saveRDS(fm_ext,
        here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_external_stability.rds\n")
cat("Rows:", nrow(fm_ext), "| Cols:", ncol(fm_ext), "\n")
```

## Session info

```{r session_info}
sessionInfo()
```
