---
title: "Positional CSC Feature Extraction — extends external_stability matrix (first 75 codons + CDS quarters)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Adds five positional CSC (Codon Stabilisation Coefficient) features to the
`external_stability` feature matrix, producing `feature_matrix_csc_positional.rds`.

| Feature | Description |
|---------|-------------|
| `csc_first75` | CSC of first 225 nt (75 codons); NA if CDS < 225 nt |
| `csc_q1` | CSC of first 25% of CDS (codon-aligned) |
| `csc_q2` | CSC of 25–50% of CDS |
| `csc_q3` | CSC of 50–75% of CDS |
| `csc_q4` | CSC of last 25% of CDS |

All scoring uses `iCodon::predict_stability("human")` — the exact same function
used for the full-CDS `csc` column in 01g. Quarter boundaries are codon-aligned
using `floor(n_codons × 0.25/0.5/0.75) × 3` nt.

Reads `feature_matrix_external_stability.rds`, appends the five positional CSC columns,
and writes back to the same file — so `feature_set = "external_stability"` in nb07
automatically includes these features.

**Prerequisites:**
- `output/predictive_modeling/feature_matrix_external_stability.rds` (from 01i)
- `accessories/human/txdb.gencode49.sqlite` (GENCODE v49)
- R packages: `GenomicFeatures`, `BSgenome.Hsapiens.UCSC.hg38`, `iCodon`

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(iCodon)
```

---

# Section 1: Load Feature Matrix and Extract CDS Sequences

```{r load_feature_matrix}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
cat("Feature matrix loaded:", nrow(feature_matrix), "genes x", ncol(feature_matrix), "columns\n")

keep_tx <- feature_matrix$transcript_id_clean
cat("Target transcripts:", length(keep_tx), "\n")
cat("Example IDs:", head(keep_tx, 3), "\n")
```

```{r load_txdb}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
cat("TxDb and genome loaded\n")
```

```{r extract_cds}
# Extract all CDS sequences from TxDb (same pattern as 01g)
cds_all      <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all <- extractTranscriptSeqs(genome, cds_all)
cat("Total CDS sequences in TxDb:", length(cds_seqs_all), "\n")

# Strip version suffixes to match feature matrix transcript_id_clean
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

# Filter to transcripts present in feature matrix
keep_idx    <- cds_names_clean %in% keep_tx
cds_sequences <- cds_seqs_all[keep_idx]
names(cds_sequences) <- cds_names_clean[keep_idx]
cat("After filtering to feature matrix transcripts:", length(cds_sequences), "\n")

# Remove sequences with length not divisible by 3 (as in 01g)
valid_length <- width(cds_sequences) %% 3 == 0
n_invalid    <- sum(!valid_length)
if (n_invalid > 0) {
  cat("Removing", n_invalid, "sequences with length not divisible by 3\n")
}
cds_sequences <- cds_sequences[valid_length]
cat("Final CDS sequences for scoring:", length(cds_sequences), "\n")

# CDS length distribution
cat("\nCDS length summary:\n")
print(summary(as.integer(width(cds_sequences))))

# Count short CDSs that will yield NA for csc_first75
n_short    <- sum(width(cds_sequences) < 225)
n_long     <- sum(width(cds_sequences) >= 225)
cat("\nCDS < 225 nt (< 75 codons) — csc_first75 will be NA:", n_short,
    sprintf("(%.1f%%)\n", 100 * n_short / length(cds_sequences)))
cat("CDS >= 225 nt — csc_first75 will be scored:", n_long, "\n")
```

---

# Section 2: Build iCodon Predictor

```{r build_predictor}
# Build human-specific CSC predictor once; reuse for all five scoring steps
predict_human <- iCodon::predict_stability("human")
cat("Human CSC predictor ready\n")
```

---

# Section 3: Compute `csc_first75`

```{r csc_first75}
# Separate long (>= 225 nt) CDSs from short ones
long_idx  <- width(cds_sequences) >= 225
long_seqs <- cds_sequences[long_idx]
cat("Long CDSs (>= 225 nt):", length(long_seqs), "\n")

# Extract first 225 nt (75 codons) as character strings
seq_75_char <- as.character(subseq(long_seqs, 1, 225))

# Deduplicate → score → restore names (exact 01g pattern)
unique_75  <- unique(seq_75_char)
cat("Unique 75-codon sequences:", length(unique_75), "\n")

scores_75        <- predict_human(unique_75)
names(scores_75) <- unique_75                          # key by sequence content

csc_first75_long        <- scores_75[seq_75_char]
names(csc_first75_long) <- names(long_seqs)            # restore transcript names

# Build full-length vector (NA for short CDSs)
csc_first75 <- rep(NA_real_, length(cds_sequences))
names(csc_first75) <- names(cds_sequences)
csc_first75[names(long_seqs)] <- csc_first75_long

cat("\ncsc_first75 summary:\n")
print(summary(csc_first75))
cat("Non-NA:", sum(!is.na(csc_first75)), "/", length(csc_first75), "\n")
```

---

# Section 4: Compute Quarter CSC (`csc_q1`–`csc_q4`)

```{r quarter_boundaries}
# Codon-aligned boundaries (floor to nearest codon, in nucleotides, 1-based)
n_codons <- as.integer(width(cds_sequences) / 3)

b1 <- pmax(floor(n_codons * 0.25) * 3L, 3L)          # end of Q1 (>= 1 codon)
b2 <- pmax(floor(n_codons * 0.50) * 3L, b1 + 3L)     # end of Q2
b3 <- pmax(floor(n_codons * 0.75) * 3L, b2 + 3L)     # end of Q3
b4 <- as.integer(width(cds_sequences))                 # end of Q4 (= full CDS)

cat("Example boundary sets (first 5 transcripts):\n")
cat(sprintf("  n_codons=%d  Q1=[1,%d]  Q2=[%d,%d]  Q3=[%d,%d]  Q4=[%d,%d]\n",
            n_codons[1], b1[1], b1[1]+1L, b2[1], b2[1]+1L, b3[1], b3[1]+1L, b4[1]))
cat(sprintf("  n_codons=%d  Q1=[1,%d]  Q2=[%d,%d]  Q3=[%d,%d]  Q4=[%d,%d]\n",
            n_codons[2], b1[2], b1[2]+1L, b2[2], b2[2]+1L, b3[2], b3[2]+1L, b4[2]))
```

```{r score_segment_helper}
# Helper: extract one CDS segment per transcript, deduplicate, score, restore names
# Follows exact 01g scoring pattern.
score_segment <- function(seqs, starts, ends, predictor) {
  seg_len <- ends - starts + 1L
  valid   <- seg_len >= 3L          # require >= 1 complete codon
  result  <- rep(NA_real_, length(seqs))
  names(result) <- names(seqs)

  if (!any(valid)) return(result)

  seg_seqs <- as.character(subseq(seqs[valid], starts[valid], ends[valid]))

  unique_segs    <- unique(seg_seqs)
  unique_scores  <- predictor(unique_segs)
  names(unique_scores) <- unique_segs           # key by sequence content
  seg_scores           <- unique_scores[seg_seqs]
  names(seg_scores)    <- names(seqs)[valid]    # restore transcript names

  result[names(seqs)[valid]] <- seg_scores
  result
}
```

```{r compute_quarters}
cat("Computing Q1 (0–25%)...\n")
csc_q1 <- score_segment(cds_sequences, rep(1L, length(cds_sequences)), b1,     predict_human)

cat("Computing Q2 (25–50%)...\n")
csc_q2 <- score_segment(cds_sequences, b1 + 1L,                        b2,     predict_human)

cat("Computing Q3 (50–75%)...\n")
csc_q3 <- score_segment(cds_sequences, b2 + 1L,                        b3,     predict_human)

cat("Computing Q4 (75–100%)...\n")
csc_q4 <- score_segment(cds_sequences, b3 + 1L,                        b4,     predict_human)

cat("\nQuarter CSC summaries:\n")
cat("Q1:\n"); print(summary(csc_q1))
cat("Q2:\n"); print(summary(csc_q2))
cat("Q3:\n"); print(summary(csc_q3))
cat("Q4:\n"); print(summary(csc_q4))

cat("\nNon-NA counts:\n")
cat(" csc_q1:", sum(!is.na(csc_q1)), "/", length(csc_q1), "\n")
cat(" csc_q2:", sum(!is.na(csc_q2)), "/", length(csc_q2), "\n")
cat(" csc_q3:", sum(!is.na(csc_q3)), "/", length(csc_q3), "\n")
cat(" csc_q4:", sum(!is.na(csc_q4)), "/", length(csc_q4), "\n")
```

---

# Section 5: Join to Feature Matrix

```{r build_score_table}
score_tbl <- tibble(
  transcript_id_clean = names(cds_sequences),
  csc_first75         = csc_first75,
  csc_q1              = csc_q1,
  csc_q2              = csc_q2,
  csc_q3              = csc_q3,
  csc_q4              = csc_q4
)

cat("Score table:", nrow(score_tbl), "rows x", ncol(score_tbl), "cols\n")
cat("Transcripts in score table not in feature matrix:",
    sum(!score_tbl$transcript_id_clean %in% feature_matrix$transcript_id_clean), "\n")
```

```{r join_matrix}
fm_new <- feature_matrix %>%
  left_join(score_tbl, by = "transcript_id_clean")

cat("Matrix after join:", nrow(fm_new), "genes x", ncol(fm_new), "columns\n")
cat("New columns added:", ncol(fm_new) - ncol(feature_matrix), "\n")
cat("Row count unchanged:", nrow(fm_new) == nrow(feature_matrix), "\n")
```

---

# Section 6: Quality Control

```{r qc_coverage}
new_features <- c("csc_first75", "csc_q1", "csc_q2", "csc_q3", "csc_q4")

cat("Non-NA coverage for new features:\n")
for (col in new_features) {
  n_nonNA <- sum(!is.na(fm_new[[col]]))
  cat(sprintf("  %-12s : %d / %d (%.1f%%)\n",
              col, n_nonNA, nrow(fm_new),
              100 * n_nonNA / nrow(fm_new)))
}

cat("\nValue ranges:\n")
for (col in new_features) {
  r <- range(fm_new[[col]], na.rm = TRUE)
  cat(sprintf("  %-12s : [%.4f, %.4f]\n", col, r[1], r[2]))
}
```

```{r qc_correlations}
# Pairwise correlations among all CSC features (full-CDS + 5 positional)
cor_cols <- c("csc", new_features)
cor_data <- fm_new %>%
  dplyr::select(all_of(cor_cols)) %>%
  drop_na()

cat("Pairwise Pearson correlations (n =", nrow(cor_data), "):\n")
print(round(cor(cor_data), 3))
```

```{r qc_density, fig.width=10, fig.height=5}
plot_df <- fm_new %>%
  dplyr::select(transcript_id_clean, all_of(cor_cols)) %>%
  pivot_longer(all_of(cor_cols), names_to = "feature", values_to = "csc_value") %>%
  filter(!is.na(csc_value)) %>%
  mutate(feature = factor(feature, levels = cor_cols))

ggplot(plot_df, aes(x = csc_value, color = feature, fill = feature)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x        = "CSC score",
    y        = "Density",
    title    = "Distribution of positional CSC features",
    subtitle = "Full CDS + first-75-codon + 4 quarters",
    color    = NULL, fill = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("output/01j_csc_positional_density.pdf"), width = 10, height = 5)
```

```{r qc_sanity}
# Sanity check: csc_first75 NA exactly for genes with CDS < 225 nt
short_tx <- names(cds_sequences)[width(cds_sequences) < 225]
n_short_na <- sum(is.na(fm_new$csc_first75[fm_new$transcript_id_clean %in% short_tx]))
cat("Short-CDS genes correctly NA in csc_first75:", n_short_na, "/", length(short_tx), "\n")
```

---

# Export

```{r export}
# Write back to feature_matrix_external_stability.rds so that nb07
# feature_set = "external_stability" automatically includes positional CSC.
saveRDS(fm_new,
        here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Exported: output/predictive_modeling/feature_matrix_external_stability.rds\n")
cat("Final matrix:", nrow(fm_new), "genes x", ncol(fm_new), "columns\n")
cat("New features added:", paste(new_features, collapse = ", "), "\n")
```

---

# Section 7: Distribution Plots by Gene Group

Compares all 6 CSC features across three groups:
- **eIF3d hypoxia promotes** — hypoxia 1hr, LFC > 0.5 (`#E69F00`)
- **Negative control** — size-matched to positive set with `set.seed(9)` (`grey40`)
- **All other genes** — remaining genes in feature matrix (`grey80`)

Mirrors the gene set sampling used in the nb07 RF model.

```{r load_genesets}
library(ggrepel)

# Positive: eIF3d hypoxia promotes 1hr LFC 0.5
hyp_promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)

# Negative control pool (never regulated by si3d in any condition)
neg_controls <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# Gene IDs in the feature matrix (versionless)
fm_gene_ids <- sub("\\..*", "", fm_new$gene_id)

pos_ids  <- hyp_promotes$ensembl_gene[hyp_promotes$ensembl_gene %in% fm_gene_ids]
neg_pool <- neg_controls$gene_id_clean[neg_controls$gene_id_clean %in% fm_gene_ids]

cat("Positive genes in feature matrix:", length(pos_ids), "\n")
cat("Negative control pool in feature matrix:", length(neg_pool), "\n")

# Size-match neg ctrl to positive set (replicates nb07 sampling exactly)
set.seed(9)
neg_ids <- sample(neg_pool, min(length(pos_ids), length(neg_pool)))
cat("Negative controls sampled (size-matched):", length(neg_ids), "\n")
```

```{r assign_groups}
group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "grey40",
  "All other genes"        = "grey80"
)

all_csc_feats <- c("csc", "csc_first75", "csc_q1", "csc_q2", "csc_q3", "csc_q4")

grp_df <- fm_new %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    group = case_when(
      gene_id_clean %in% pos_ids ~ "eIF3d hypoxia promotes",
      gene_id_clean %in% neg_ids ~ "Negative control",
      TRUE                       ~ "All other genes"
    ),
    group = factor(group, levels = c("eIF3d hypoxia promotes",
                                     "Negative control",
                                     "All other genes"))
  ) %>%
  dplyr::select(gene_id_clean, symbol, group, all_of(all_csc_feats))

cat("\nGroup sizes:\n")
print(table(grp_df$group))
```

```{r wilcoxon_stats}
# Wilcoxon tests: eIF3d vs Neg ctrl, eIF3d vs All other, for each feature
wx_results <- map_dfr(all_csc_feats, function(feat) {
  eif3d <- grp_df %>% filter(group == "eIF3d hypoxia promotes") %>% pull(!!feat) %>% na.omit()
  neg   <- grp_df %>% filter(group == "Negative control")       %>% pull(!!feat) %>% na.omit()
  other <- grp_df %>% filter(group == "All other genes")         %>% pull(!!feat) %>% na.omit()

  wx_neg   <- wilcox.test(eif3d, neg,   alternative = "two.sided")
  wx_other <- wilcox.test(eif3d, other, alternative = "two.sided")

  tibble(
    feature             = feat,
    n_eif3d             = length(eif3d),
    n_neg               = length(neg),
    n_other             = length(other),
    median_eif3d        = round(median(eif3d), 4),
    median_neg          = round(median(neg),   4),
    median_other        = round(median(other), 4),
    p_eif3d_vs_neg      = wx_neg$p.value,
    p_eif3d_vs_other    = wx_other$p.value
  )
})

cat("=== Wilcoxon test results (eIF3d vs each background) ===\n")
print(wx_results %>% mutate(across(starts_with("p_"), ~ formatC(.x, format = "e", digits = 2))))
```

```{r facet_density, fig.width=12, fig.height=7}
feat_labels <- c(
  csc        = "CSC (full CDS)",
  csc_first75 = "CSC (first 75 codons)",
  csc_q1     = "CSC (Q1: 0–25%)",
  csc_q2     = "CSC (Q2: 25–50%)",
  csc_q3     = "CSC (Q3: 50–75%)",
  csc_q4     = "CSC (Q4: 75–100%)"
)

density_df <- grp_df %>%
  pivot_longer(all_of(all_csc_feats), names_to = "feature", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    feature = factor(feature, levels = all_csc_feats, labels = feat_labels[all_csc_feats])
  )

# Plot All other genes first (grey background), then neg ctrl, then eIF3d on top
density_ordered <- density_df %>%
  arrange(match(group, c("All other genes", "Negative control", "eIF3d hypoxia promotes")))

ggplot(density_ordered, aes(x = value, color = group, fill = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(
    data = density_df %>%
      group_by(feature, group) %>%
      summarise(med = median(value, na.rm = TRUE), .groups = "drop"),
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.5
  ) +
  facet_wrap(~feature, scales = "free", ncol = 3) +
  scale_color_manual(values = group_colors, name = NULL) +
  scale_fill_manual(values  = group_colors, name = NULL) +
  labs(
    x        = "CSC score",
    y        = "Density",
    title    = "Positional CSC distributions by gene group",
    subtitle = "Negative control size-matched to eIF3d set (set.seed(9)) | dashed = group median"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        strip.background = element_blank(),
        strip.text       = element_text(size = 11))

ggsave(here("output/01j_csc_positional_by_group_facet.pdf"), width = 12, height = 7)
```

```{r individual_plots, fig.width=6, fig.height=4.5}
# Individual plot per feature with Wilcoxon p-value annotations
for (feat in all_csc_feats) {
  feat_lab <- feat_labels[feat]
  wx_row   <- wx_results %>% filter(feature == feat)

  p <- ggplot(
    density_df %>% filter(feature == feat_labels[feat]),
    aes(x = value, color = group, fill = group)
  ) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    scale_color_manual(values = group_colors, name = NULL) +
    scale_fill_manual(values  = group_colors, name = NULL) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.3, size = 3.2,
             label = sprintf(
               "eIF3d (n=%d)  med=%.3f\nNeg ctrl (n=%d)  med=%.3f  p=%s\nAll other (n=%d)  med=%.3f  p=%s",
               wx_row$n_eif3d, wx_row$median_eif3d,
               wx_row$n_neg,   wx_row$median_neg,
               formatC(wx_row$p_eif3d_vs_neg,   format = "e", digits = 1),
               wx_row$n_other, wx_row$median_other,
               formatC(wx_row$p_eif3d_vs_other, format = "e", digits = 1)
             )) +
    labs(
      x        = "CSC score",
      y        = "Density",
      title    = feat_lab,
      subtitle = "eIF3d hypoxia 1hr LFC > 0.5 | neg ctrl size-matched set.seed(9)"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")

  print(p)
  ggsave(here(paste0("output/01j_", feat, "_by_group.pdf")), p, width = 6, height = 4.5)
}
```

# Session Info

```{r session_info}
sessionInfo()
```
