---
title: "RNA Secondary Structure Feature Extraction — CDS & 3'UTR (RNAplfold only)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Extends the RNA secondary structure feature set to the **CDS** and **3'UTR** using
RNAplfold accessibility features only. Also removes the 5'UTR RNAfold features
(MFE, ensemble, dot-bracket) that were added in `01d`.

Reads `feature_matrix_external_stability.rds`, drops 5'UTR RNAfold columns, appends
CDS and 3'UTR RNAplfold features, and writes back to the same file.

## Features Added (6 per region)

| Feature | Description |
|---|---|
| `struct_accessibility_start_proximal_cds` | Mean unpaired prob, positions 1–30 (start codon context) |
| `struct_accessibility_stop_proximal_cds` | Mean unpaired prob, last 30 nt (stop codon context) |
| `struct_accessibility_cds_mean` | Mean unpaired prob, full CDS |
| `struct_accessibility_cds_min` | Minimum unpaired prob |
| `struct_accessibility_cds_sd` | SD of unpaired prob |
| `struct_num_structured_regions_cds` | Count of positions with 30-nt accessibility < 0.2 |

3'UTR features use the same set with `_utr3` suffix; positional windows are
`struct_accessibility_stop_proximal_utr3` (positions 1–30, proximal to stop codon)
and `struct_accessibility_distal_utr3` (last 30 nt).

## Features Removed (7)

5'UTR RNAfold features dropped from model:
`struct_mfe_utr5`, `struct_mfe_per_nt_utr5`, `struct_ensemble_free_energy`,
`struct_ensemble_diversity`, `struct_frac_paired_utr5`, `struct_num_hairpins_utr5`,
`struct_longest_stem_utr5`

## Workflow

**Section 1** generates FASTAs locally for cluster submission (run once).
**Section 2** parses CDS RNAplfold output (run when CDS job is complete).
**Section 3** parses 3'UTR RNAplfold output — skipped automatically if files not yet present.
**Section 4** joins features, drops RNAfold columns, and exports.

## Prerequisites

RNAplfold output transferred from cluster to:
- `accessories/plfold_output/rnaplfold_output/cds_rnaplfold_output/` (_lunp files)
- `accessories/plfold_output/rnaplfold_output/utr3_rnaplfold_output/` (_lunp files, when ready)

Cluster commands (bodhi: `/beevol/home/matlinka/timecourse/viennarna/`):
```
sbatch cluster_scripts/viennarna/02b_run_rnaplfold_cds_utr3.bsub
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
```

---

# Section 1: Extract Sequences and Write FASTAs for Cluster

```{r load_feature_matrix}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
cat("Feature matrix loaded:", nrow(feature_matrix), "genes x", ncol(feature_matrix), "columns\n")

keep_tx <- feature_matrix$transcript_id_clean
cat("Target transcripts:", length(keep_tx), "\n")
```

```{r load_txdb}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
cat("TxDb and genome loaded\n")
```

```{r extract_cds_sequences}
cds_all      <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all <- extractTranscriptSeqs(genome, cds_all)
cat("Total CDS sequences in TxDb:", length(cds_seqs_all), "\n")

cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))
keep_idx        <- cds_names_clean %in% keep_tx
cds_seqs        <- cds_seqs_all[keep_idx]
names(cds_seqs) <- cds_names_clean[keep_idx]
cat("CDS sequences matching feature matrix:", length(cds_seqs), "\n")

cat("\nCDS length summary:\n")
print(summary(as.integer(width(cds_seqs))))
```

```{r extract_utr3_sequences}
utr3_all      <- threeUTRsByTranscript(txdb, use.names = TRUE)
utr3_seqs_all <- extractTranscriptSeqs(genome, utr3_all)
cat("Total 3'UTR sequences in TxDb:", length(utr3_seqs_all), "\n")

utr3_names_clean <- sub("\\..*", "", names(utr3_seqs_all))
keep_idx_utr3    <- utr3_names_clean %in% keep_tx
utr3_seqs        <- utr3_seqs_all[keep_idx_utr3]
names(utr3_seqs) <- utr3_names_clean[keep_idx_utr3]
cat("3'UTR sequences matching feature matrix:", length(utr3_seqs), "\n")

n_no_utr3 <- sum(!keep_tx %in% names(utr3_seqs))
cat("Transcripts without annotated 3'UTR:", n_no_utr3,
    sprintf("(%.1f%%)\n", 100 * n_no_utr3 / length(keep_tx)))
```

```{r write_fastas}
dir.create(here("accessories", "plfold_output", "cds"),  showWarnings = FALSE, recursive = TRUE)
dir.create(here("accessories", "plfold_output", "utr3"), showWarnings = FALSE, recursive = TRUE)

cds_fa_path <- here("accessories", "plfold_output", "cds", "sequences.fa")
writeXStringSet(cds_seqs, filepath = cds_fa_path)
cat("Written:", cds_fa_path, "(", length(cds_seqs), "sequences)\n")

utr3_fa_path <- here("accessories", "plfold_output", "utr3", "sequences.fa")
writeXStringSet(utr3_seqs, filepath = utr3_fa_path)
cat("Written:", utr3_fa_path, "(", length(utr3_seqs), "sequences)\n")
```

---

# Parsing Function

```{r define_parser}
parse_lunp <- function(filepath) {
  lines      <- readLines(filepath)
  header_end <- max(which(str_starts(lines, "#")))
  data_lines <- lines[(header_end + 1):length(lines)]
  data_lines <- data_lines[nzchar(trimws(data_lines))]

  if (length(data_lines) == 0) return(NULL)

  df     <- read.table(text = data_lines, header = FALSE, sep = "\t", fill = TRUE)
  n_cols <- ncol(df)

  result <- data.frame(
    position          = df[[1]],
    unpaired_prob_1nt = df[[2]]
  )

  result$unpaired_prob_30nt <- if (n_cols >= 31) df[[31]] else df[[n_cols]]

  return(result)
}
```

---

# Section 2: Parse RNAplfold — CDS

```{r parse_plfold_cds}
plfold_dir_cds <- here("accessories", "plfold_output", "rnaplfold_output", "cds_rnaplfold_output")
stopifnot("CDS RNAplfold output not found" = dir.exists(plfold_dir_cds))

plfold_files_cds <- list.files(plfold_dir_cds, pattern = "_lunp$", full.names = TRUE)
cat("CDS RNAplfold _lunp files found:", length(plfold_files_cds), "\n")

plfold_cds_features <- lapply(plfold_files_cds, function(f) {
  tx_id  <- sub("_lunp$", "", basename(f))
  parsed <- parse_lunp(f)
  if (is.null(parsed)) return(NULL)

  seq_len    <- nrow(parsed)
  cap_range  <- 1:min(30, seq_len)
  stop_range <- max(1, seq_len - 29):seq_len

  acc_vals <- parsed$unpaired_prob_1nt
  acc_vals[is.na(acc_vals)] <- 0

  acc_30nt       <- parsed$unpaired_prob_30nt
  acc_30nt[is.na(acc_30nt)] <- 0
  num_structured <- sum(acc_30nt < 0.2, na.rm = TRUE)

  tibble(
    transcript_id_clean                     = tx_id,
    struct_accessibility_start_proximal_cds = mean(acc_vals[cap_range],  na.rm = TRUE),
    struct_accessibility_stop_proximal_cds  = mean(acc_vals[stop_range], na.rm = TRUE),
    struct_accessibility_cds_mean           = mean(acc_vals, na.rm = TRUE),
    struct_accessibility_cds_min            = min(acc_vals,  na.rm = TRUE),
    struct_accessibility_cds_sd             = sd(acc_vals,   na.rm = TRUE),
    struct_num_structured_regions_cds       = num_structured
  )
})

plfold_cds_df <- bind_rows(plfold_cds_features)
cat("CDS features computed for", nrow(plfold_cds_df), "transcripts\n")

cat("\nAccessibility mean (CDS):\n")
print(summary(plfold_cds_df$struct_accessibility_cds_mean))
```

---

# Section 3: Parse RNAplfold — 3'UTR

```{r parse_plfold_utr3}
plfold_dir_utr3   <- here("accessories", "plfold_output", "rnaplfold_output", "utr3_rnaplfold_output")
plfold_files_utr3 <- if (dir.exists(plfold_dir_utr3)) {
  list.files(plfold_dir_utr3, pattern = "_lunp$", full.names = TRUE)
} else {
  character(0)
}

if (length(plfold_files_utr3) == 0) {
  cat("3'UTR RNAplfold output not yet available — skipping.\n")
  plfold_utr3_df <- NULL
} else {
  cat("3'UTR RNAplfold _lunp files found:", length(plfold_files_utr3), "\n")

  plfold_utr3_features <- lapply(plfold_files_utr3, function(f) {
    tx_id  <- sub("_lunp$", "", basename(f))
    parsed <- parse_lunp(f)
    if (is.null(parsed)) return(NULL)

    seq_len    <- nrow(parsed)
    stop_range <- 1:min(30, seq_len)
    dist_range <- max(1, seq_len - 29):seq_len

    acc_vals <- parsed$unpaired_prob_1nt
    acc_vals[is.na(acc_vals)] <- 0

    acc_30nt       <- parsed$unpaired_prob_30nt
    acc_30nt[is.na(acc_30nt)] <- 0
    num_structured <- sum(acc_30nt < 0.2, na.rm = TRUE)

    tibble(
      transcript_id_clean                     = tx_id,
      struct_accessibility_stop_proximal_utr3 = mean(acc_vals[stop_range], na.rm = TRUE),
      struct_accessibility_distal_utr3        = mean(acc_vals[dist_range], na.rm = TRUE),
      struct_accessibility_utr3_mean          = mean(acc_vals, na.rm = TRUE),
      struct_accessibility_utr3_min           = min(acc_vals,  na.rm = TRUE),
      struct_accessibility_utr3_sd            = sd(acc_vals,   na.rm = TRUE),
      struct_num_structured_regions_utr3      = num_structured
    )
  })

  plfold_utr3_df <- bind_rows(plfold_utr3_features)
  cat("3'UTR features computed for", nrow(plfold_utr3_df), "transcripts\n")

  cat("\nAccessibility mean (3'UTR):\n")
  print(summary(plfold_utr3_df$struct_accessibility_utr3_mean))
}
```

---

# Section 4: Join to Feature Matrix and Export

```{r drop_rnafold_cols}
# Remove 5'UTR RNAfold features (dropped from model)
utr5_rnafold_cols <- c(
  "struct_mfe_utr5", "struct_mfe_per_nt_utr5", "struct_ensemble_free_energy",
  "struct_ensemble_diversity", "struct_frac_paired_utr5",
  "struct_num_hairpins_utr5", "struct_longest_stem_utr5"
)
cols_to_drop <- intersect(utr5_rnafold_cols, names(feature_matrix))
cat("Dropping", length(cols_to_drop), "5'UTR RNAfold columns:", paste(cols_to_drop, collapse = ", "), "\n")

fm_new <- feature_matrix %>% dplyr::select(-all_of(cols_to_drop))
```

```{r join_cds}
cds_cols <- names(plfold_cds_df)[names(plfold_cds_df) != "transcript_id_clean"]

fm_new <- fm_new %>%
  left_join(plfold_cds_df, by = "transcript_id_clean") %>%
  mutate(across(all_of(cds_cols), ~replace_na(., 0)))

n_with_cds <- sum(feature_matrix$transcript_id_clean %in% plfold_cds_df$transcript_id_clean)
cat("CDS structure coverage:", n_with_cds, "/", nrow(feature_matrix), "\n")
cat("CDS features added:", length(cds_cols), "\n")
```

```{r join_utr3}
if (!is.null(plfold_utr3_df)) {
  utr3_cols <- names(plfold_utr3_df)[names(plfold_utr3_df) != "transcript_id_clean"]

  fm_new <- fm_new %>%
    left_join(plfold_utr3_df, by = "transcript_id_clean") %>%
    mutate(across(all_of(utr3_cols), ~replace_na(., 0)))

  n_with_utr3 <- sum(feature_matrix$transcript_id_clean %in% plfold_utr3_df$transcript_id_clean)
  cat("3'UTR structure coverage:", n_with_utr3, "/", nrow(feature_matrix), "\n")
  cat("3'UTR features added:", length(utr3_cols), "\n")
} else {
  cat("3'UTR features not yet added — re-run after cluster job completes.\n")
}
```

```{r export}
cat("\nFinal matrix:", nrow(fm_new), "genes x", ncol(fm_new), "columns\n")
cat("(was:", nrow(feature_matrix), "x", ncol(feature_matrix), ")\n")

saveRDS(fm_new, here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_external_stability.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
