---
title: "01m — Feature Extraction: Codon-Only Feature Matrix"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Subsets `feature_matrix_codon_features.rds` to only the 131 codon-related feature
columns, producing `feature_matrix_codon_only.rds` for use with notebook 07
(`feature_set = "codon_only"`) and notebook 09.

**131 codon features retained:**

- `gc3`, `cai`, `tai`, `csc`, `csc_first75`, `csc_q1`–`csc_q4`, `fop` — 10 cols (from 01g/01j)
- `bg_rscu_*` — 62 cols (from 01l)
- `codon_freq_*` — 59 cols (from 01l)

All non-codon features (UTR lengths, GC%, k-mers, RBP motifs, CLIP counts, RNA
structure, TE features, etc.) are dropped.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Full Feature Matrix

```{r load_feature_matrix}
fm_full <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_codon_features.rds")
)
cat("Full feature matrix:", nrow(fm_full), "genes x", ncol(fm_full), "columns\n")
```

---

# Section 2: Select Codon Features

```{r select_codon_features}
id_cols <- c("transcript_id", "transcript_id_clean", "gene_id", "symbol", "gene_id_clean",
             "transcript_id_raw")
id_cols_present <- id_cols[id_cols %in% names(fm_full)]

codon_pattern <- "^bg_rscu_|^codon_freq_|^gc3$|^cai$|^tai$|^fop$|^csc"
codon_cols    <- names(fm_full)[str_detect(names(fm_full), codon_pattern)]

cat("Identifier columns retained:", length(id_cols_present), "\n")
cat("  ", paste(id_cols_present, collapse = ", "), "\n")
cat("Codon feature columns retained:", length(codon_cols), "(expected 131)\n")

# Sanity: no non-codon features snuck in
bad_cols <- codon_cols[str_detect(codon_cols, "^utr5_|^utr3_|^clip_|^rbns_|^cisbp_|^indiv_te")]
if (length(bad_cols) > 0) {
  stop("Non-codon columns matched: ", paste(bad_cols, collapse = ", "))
}
cat("Non-codon leakage check: passed\n")
stopifnot(length(codon_cols) == 131)
cat("Column count (131) check: passed\n")
```

---

# Section 3: Build and Save Codon-Only Matrix

```{r build_save}
fm_codon <- fm_full %>%
  dplyr::select(all_of(id_cols_present), all_of(codon_cols))

stopifnot(nrow(fm_codon) == nrow(fm_full))
cat("Row count preserved:", nrow(fm_codon), "\n")
cat("Codon-only matrix:", nrow(fm_codon), "genes x", ncol(fm_codon), "columns\n")
cat("  (", length(id_cols_present), "id cols +", length(codon_cols), "codon features)\n")

saveRDS(fm_codon,
        here("output", "predictive_modeling", "feature_matrix_codon_only.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_codon_only.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
