---
title: "01n — Feature Extraction: RIP-seq (IP/IgG, IP/RNA) Feature Matrix"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Adds eIF3d and eIF4E RIP-seq binding features (IP/IgG DESeq2 log2FC and IP/RNA log2
ratio, at each of 4 timepoints) to `feature_matrix_external_stability.rds`, producing
`feature_matrix_ripseq.rds` for use with notebook 07 via
`feature_matrix_path = "feature_matrix_ripseq.rds"` (with `feature_set` set to any
label, e.g. `"ripseq"`, purely for output-file naming).

**16 features added, all gene-level (broadcast to every transcript of the same gene),
NA for genes not covered by the RIP-seq experiment:**

- `ripseq_eif3d_igg_lfc_{timepoint}` — DESeq2 log2FC, FLAG-eIF3d IP vs IgG (4 cols)
- `ripseq_eif3d_rna_lfc_{timepoint}` — log2(IP CPM / spike-in normalized RNA), eIF3d IP (4 cols)
- `eif4e_ip_igg_lfc_{timepoint}` — DESeq2 log2FC, eIF4E IP vs IgG (4 cols)
- `eif4e_ip_rna_lfc_{timepoint}` — log2(IP CPM / spike-in normalized RNA), eIF4E IP (4 cols)

Timepoints: `normoxia_1hr`, `hypoxia_1hr`, `hypoxia_4hr`, `hypoxia_24hr`.

The `eif4e_ip_*` prefix intentionally matches the existing `exclude_4e_features` regex
(`^eif4e_ip`) in notebook 07, so these new eIF4E columns are automatically excluded
when that toggle is set — no change to 07 required.

Source data: `ripseq/code/01_eif3d_rip_deseq.Rmd` (IP/IgG) and
`ripseq/code/18_ip_input_normalization.Rmd` (IP/RNA, spike-in normalized RNA
denominator, fixed 2026-07-01).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load IP/IgG Data (DESeq2 log2FC vs IgG, per timepoint)

```{r load_ip_igg}
timepoints <- c("normoxia_1hr", "hypoxia_1hr", "hypoxia_4hr", "hypoxia_24hr")

read_igg_bait <- function(bait) {
  map_dfr(timepoints, function(tp) {
    read_csv(
      here("ripseq", "output", paste0("ripseq_deseq_", tp, "_", bait, "_vs_igg.csv")),
      show_col_types = FALSE
    ) %>%
      dplyr::select(gene_id_clean, timepoint, log2FoldChange)
  })
}

igg_eif3d <- read_igg_bait("flag_eif3d")
igg_eif4e <- read_igg_bait("eif4e")

cat("eIF3d IP/IgG rows:", nrow(igg_eif3d), "\n")
cat("eIF4E IP/IgG rows:", nrow(igg_eif4e), "\n")

stopifnot(
  "IP/IgG log2FC must be finite" =
    all(is.finite(igg_eif3d$log2FoldChange[!is.na(igg_eif3d$log2FoldChange)])),
  all(is.finite(igg_eif4e$log2FoldChange[!is.na(igg_eif4e$log2FoldChange)]))
)
cat("IP/IgG finiteness check: passed\n")
```

# Section 2: Load IP/RNA Data (log2 ratio vs spike-in normalized RNA, per timepoint)

```{r load_ip_rna}
ip_over_rna <- read_csv(
  here("ripseq", "output", "18_ip_over_rna_enrichment.csv"),
  show_col_types = FALSE
)

stopifnot(
  "18_ip_over_rna_enrichment.csv must contain expected ip_type values" =
    all(c("flag_eif3d", "eif4e") %in% unique(ip_over_rna$ip_type)),
  "18_ip_over_rna_enrichment.csv must contain expected timepoints" =
    all(timepoints %in% unique(ip_over_rna$timepoint))
)

rna_eif3d <- ip_over_rna %>%
  filter(ip_type == "flag_eif3d") %>%
  dplyr::select(gene_id_clean, timepoint, log2_ip_over_rna)

rna_eif4e <- ip_over_rna %>%
  filter(ip_type == "eif4e") %>%
  dplyr::select(gene_id_clean, timepoint, log2_ip_over_rna)

cat("eIF3d IP/RNA rows:", nrow(rna_eif3d), "\n")
cat("eIF4E IP/RNA rows:", nrow(rna_eif4e), "\n")

stopifnot(
  "IP/RNA log2 ratio must be finite" =
    all(is.finite(rna_eif3d$log2_ip_over_rna[!is.na(rna_eif3d$log2_ip_over_rna)])),
  all(is.finite(rna_eif4e$log2_ip_over_rna[!is.na(rna_eif4e$log2_ip_over_rna)]))
)
cat("IP/RNA finiteness check: passed\n")
```

# Section 3: Pivot to Wide, Per-Gene Feature Table

```{r pivot_wide}
igg_eif3d_wide <- igg_eif3d %>%
  pivot_wider(names_from = timepoint, values_from = log2FoldChange,
              names_prefix = "ripseq_eif3d_igg_lfc_")

igg_eif4e_wide <- igg_eif4e %>%
  pivot_wider(names_from = timepoint, values_from = log2FoldChange,
              names_prefix = "eif4e_ip_igg_lfc_")

rna_eif3d_wide <- rna_eif3d %>%
  pivot_wider(names_from = timepoint, values_from = log2_ip_over_rna,
              names_prefix = "ripseq_eif3d_rna_lfc_")

rna_eif4e_wide <- rna_eif4e %>%
  pivot_wider(names_from = timepoint, values_from = log2_ip_over_rna,
              names_prefix = "eif4e_ip_rna_lfc_")

ripseq_features <- igg_eif3d_wide %>%
  full_join(igg_eif4e_wide, by = "gene_id_clean") %>%
  full_join(rna_eif3d_wide, by = "gene_id_clean") %>%
  full_join(rna_eif4e_wide, by = "gene_id_clean")

expected_ripseq_cols <- c(
  paste0("ripseq_eif3d_igg_lfc_", timepoints),
  paste0("eif4e_ip_igg_lfc_", timepoints),
  paste0("ripseq_eif3d_rna_lfc_", timepoints),
  paste0("eif4e_ip_rna_lfc_", timepoints)
)
stopifnot(all(expected_ripseq_cols %in% colnames(ripseq_features)))
cat("RIP-seq feature table:", nrow(ripseq_features), "genes x",
    length(expected_ripseq_cols), "features\n")
```

# Section 4: Join to Feature Matrix

```{r join_to_feature_matrix}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Feature matrix loaded:", nrow(fm), "rows x", ncol(fm), "columns\n")

fm_ripseq <- fm %>%
  left_join(ripseq_features, by = "gene_id_clean")

cat("After join:", nrow(fm_ripseq), "rows x", ncol(fm_ripseq), "columns\n")
cat("New columns added:", ncol(fm_ripseq) - ncol(fm), "\n")

cat("\nCoverage within feature matrix rows:\n")
for (col in expected_ripseq_cols) {
  cat(" ", col, ":", sum(!is.na(fm_ripseq[[col]])), "/", nrow(fm_ripseq), "\n")
}
```

# Section 5: Validation

```{r validation_computational}
stopifnot(
  "row count must be preserved by the join" = nrow(fm_ripseq) == nrow(fm),
  "all expected RIP-seq columns must be present" =
    all(expected_ripseq_cols %in% colnames(fm_ripseq)),
  "RIP-seq LFC/ratio columns must be finite where non-NA" = all(sapply(
    expected_ripseq_cols,
    function(col) all(is.finite(fm_ripseq[[col]][!is.na(fm_ripseq[[col]])]))
  ))
)
cat("Computational checks passed\n")
```

```{r validation_biological}
# EIF3D is the bait in the flag_eif3d IP and should be strongly self-enriched
eif3d_self <- fm_ripseq %>%
  filter(symbol == "EIF3D") %>%
  pull(ripseq_eif3d_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF3D self-enrichment in flag_eif3d IP/IgG must be strongly positive" =
    length(eif3d_self) > 0 && all(eif3d_self > 2, na.rm = TRUE)
)
cat("EIF3D self-enrichment (ripseq_eif3d_igg_lfc_hypoxia_1hr):",
    unique(eif3d_self), "\n")

# EIF4E is the bait in the eif4e IP and should be self-enriched (positive)
eif4e_self <- fm_ripseq %>%
  filter(symbol == "EIF4E") %>%
  pull(eif4e_ip_igg_lfc_hypoxia_1hr)
stopifnot(
  "EIF4E self-enrichment in eif4e IP/IgG must be positive" =
    length(eif4e_self) > 0 && all(eif4e_self > 0, na.rm = TRUE)
)
cat("EIF4E self-enrichment (eif4e_ip_igg_lfc_hypoxia_1hr):",
    unique(eif4e_self), "\n")

cat("Biological spot-checks passed\n")
```

# Section 6: Save Output

```{r save_output}
saveRDS(fm_ripseq, here("output", "predictive_modeling", "feature_matrix_ripseq.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_ripseq.rds\n")
```
