---
title: "01o — Feature Extraction: 5'UTR-Only Feature Matrix"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Subsets `feature_matrix_external_stability.rds` to only 5'UTR-region feature columns,
producing `feature_matrix_utr5_only.rds` for use with notebook 07 (`feature_set = "utr5_only"`,
`feature_matrix_path = "feature_matrix_utr5_only.rds"`).

**16 5'UTR features retained:**

- Sequence/length: `utr5_length`, `log2_utr5_length`, `utr5_gc` — 3 cols (from 01_)
- RNA structure/accessibility: `struct_accessibility_cap_proximal`,
  `struct_accessibility_aug_context`, `struct_accessibility_utr5_mean/min/sd`,
  `struct_num_structured_regions` — 6 cols (from 01d_)
- Kozak context: `kozak_score`, `kozak_optimal`, `kozak_pwm_score` — 3 cols (from 01h_).
  Included even though the Kozak window technically spans a few CDS nucleotides, because it's
  functionally a 5'UTR/translation-initiation-context feature.
- Regional CLIP binding, 5'UTR-restricted: `clip_lee_parclip_a_b_d_g_5utr`,
  `clip_lee_parclip_d_only_5utr`, `clip_cate_undiff_npcs_5utr`, `clip_cate_diff_npcs_5utr` —
  4 cols (from 01_/01b_). Note these only take effect under `clip_mode = "individual"` in
  notebook 07 -- with the usual `clip_mode = "aggregate"` default, they're present in the
  matrix but dropped as "regional" at fit time, same as in every other feature_set.

All non-5'UTR features (CDS/3'UTR/total lengths and GC%, codon optimality, aggregate CLIP
counts, own-TE features, external stability datasets, etc.) are dropped.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Full Feature Matrix

```{r load_feature_matrix}
fm_full <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
cat("Full feature matrix:", nrow(fm_full), "genes x", ncol(fm_full), "columns\n")
```

---

# Section 2: Select 5'UTR Features

```{r select_utr5_features}
id_cols <- c("transcript_id", "transcript_id_clean", "gene_id", "symbol", "gene_id_clean",
             "transcript_id_raw")
id_cols_present <- id_cols[id_cols %in% names(fm_full)]

utr5_pattern <- paste0(
  "^utr5_length$|^log2_utr5_length$|^utr5_gc$|",
  "^struct_accessibility_cap_proximal$|^struct_accessibility_aug_context$|",
  "^struct_accessibility_utr5_|^struct_num_structured_regions$|",
  "^kozak_|_5utr$"
)
utr5_cols <- names(fm_full)[str_detect(names(fm_full), utr5_pattern)]

cat("Identifier columns retained:", length(id_cols_present), "\n")
cat("  ", paste(id_cols_present, collapse = ", "), "\n")
cat("5'UTR feature columns retained:", length(utr5_cols), "(expected 16)\n")
cat("  ", paste(utr5_cols, collapse = ", "), "\n")

# Sanity: no CDS/3'UTR/other-region features snuck in
bad_cols <- utr5_cols[str_detect(utr5_cols, "_cds|_utr3|_3utr|^cds_|^utr3_|^total_")]
if (length(bad_cols) > 0) {
  stop("Non-5'UTR columns matched: ", paste(bad_cols, collapse = ", "))
}
cat("Non-5'UTR leakage check: passed\n")
stopifnot(length(utr5_cols) == 16)
cat("Column count (16) check: passed\n")
```

---

# Section 3: Validation

```{r validation_checks}
fm_check <- fm_full %>% dplyr::select(all_of(utr5_cols))

stopifnot("utr5_length must be positive" =
  all(fm_check$utr5_length > 0, na.rm = TRUE))
stopifnot("utr5_gc must be in [0, 100]" =
  all(fm_check$utr5_gc >= 0 & fm_check$utr5_gc <= 100, na.rm = TRUE))
stopifnot("Accessibility (RNAplfold) must be in [0, 1]" =
  all(fm_check$struct_accessibility_cap_proximal >= 0 & fm_check$struct_accessibility_cap_proximal <= 1, na.rm = TRUE),
  all(fm_check$struct_accessibility_aug_context >= 0 & fm_check$struct_accessibility_aug_context <= 1, na.rm = TRUE),
  all(fm_check$struct_accessibility_utr5_mean >= 0 & fm_check$struct_accessibility_utr5_mean <= 1, na.rm = TRUE))
stopifnot("struct_num_structured_regions must be >= 0" =
  all(fm_check$struct_num_structured_regions >= 0, na.rm = TRUE))
stopifnot("kozak_optimal must be binary" =
  all(fm_check$kozak_optimal %in% c(0, 1, NA)))
cat("Range/computational checks: passed\n")

# Biological spot-check: ACTB has a well-characterized strong Kozak context
actb_row <- fm_full %>% filter(symbol == "ACTB") %>% slice(1)
if (nrow(actb_row) == 1) {
  cat("ACTB kozak_optimal:", actb_row$kozak_optimal, "| kozak_pwm_score:", round(actb_row$kozak_pwm_score, 3), "\n")
}
```

---

# Section 4: Build and Save 5'UTR-Only Matrix

```{r build_save}
fm_utr5 <- fm_full %>%
  dplyr::select(all_of(id_cols_present), all_of(utr5_cols))

stopifnot(nrow(fm_utr5) == nrow(fm_full))
cat("Row count preserved:", nrow(fm_utr5), "\n")
cat("5'UTR-only matrix:", nrow(fm_utr5), "genes x", ncol(fm_utr5), "columns\n")
cat("  (", length(id_cols_present), "id cols +", length(utr5_cols), "5'UTR features)\n")

saveRDS(fm_utr5,
        here("output", "predictive_modeling", "feature_matrix_utr5_only.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_utr5_only.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
