---
title: "01p — Feature Extraction: external_stability minus CDS/3'UTR-specific and Codon-Usage Features"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Subsets `feature_matrix_external_stability.rds` by *removing* everything specific to the CDS
or 3'UTR region, producing `feature_matrix_no_cds_utr3.rds` for use with notebook 07
(`feature_set = "no_cds_utr3"`, `feature_matrix_path = "feature_matrix_no_cds_utr3.rds"`).

**Exclusion scope, per 2026-07-17 conversation:**

- CDS-region-specific: `cds_length`, `log2_cds_length`, `cds_gc`, CDS RNAplfold accessibility
  (`struct_accessibility_*_cds`, `struct_num_structured_regions_cds`), CDS-region regional CLIP
  (`clip_lee_parclip_*_cds`, `clip_cate_*_cds`) — 13 cols
- 3'UTR-region-specific: `utr3_length`, `log2_utr3_length`, `utr3_gc`, 3'UTR RNAplfold
  accessibility (`struct_accessibility_*_utr3`, `struct_num_structured_regions_utr3`),
  3'UTR-region regional CLIP (`clip_cate_*_3utr`) — 11 cols
- Codon-usage/optimality scores, since they can only be computed from CDS codon sequence:
  `cai`, `tai`, `csc`, `fop`, `csc_first75`, `csc_q1`-`csc_q4`, `gc3` — 10 cols. Also
  `cnot3_weighted_codon_score` (a codon-usage-weighted stability score, same family as the
  above even though it wasn't in the originally-discussed list) — 1 col, flagged here rather
  than silently excluded.
- Whole-transcript metrics that blend all three regions: `total_length`, `log2_total_length`,
  `transcript_gc` — 3 cols

**Explicitly kept despite touching the CDS/5'UTR boundary:** Kozak features (`kozak_score`,
`kozak_optimal`, `kozak_pwm_score`) — treated as 5'UTR/initiation-context, consistent with the
`01o_` (utr5_only) matrix.

All other `external_stability` columns (identifiers, own-TE features, `sig_*` flags, aggregate
and non-CDS/non-3'UTR regional CLIP, 5'UTR sequence/GC/structure, Kozak, eIF4E/DAP5 features,
and the external CNOT3/DHX29/Karner stability datasets) are retained.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Full Feature Matrix

```{r load_feature_matrix}
fm_full <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
cat("Full feature matrix:", nrow(fm_full), "genes x", ncol(fm_full), "columns\n")
```

---

# Section 2: Define Exclusion List and Select Remaining Columns

```{r select_columns}
cds_specific <- c(
  "cds_length", "log2_cds_length", "cds_gc",
  "struct_accessibility_start_proximal_cds", "struct_accessibility_stop_proximal_cds",
  "struct_accessibility_cds_mean", "struct_accessibility_cds_min", "struct_accessibility_cds_sd",
  "struct_num_structured_regions_cds",
  "clip_lee_parclip_a_b_d_g_cds", "clip_lee_parclip_d_only_cds",
  "clip_cate_undiff_npcs_cds", "clip_cate_diff_npcs_cds"
)
utr3_specific <- c(
  "utr3_length", "log2_utr3_length", "utr3_gc",
  "struct_accessibility_stop_proximal_utr3", "struct_accessibility_distal_utr3",
  "struct_accessibility_utr3_mean", "struct_accessibility_utr3_min", "struct_accessibility_utr3_sd",
  "struct_num_structured_regions_utr3",
  "clip_cate_undiff_npcs_3utr", "clip_cate_diff_npcs_3utr"
)
codon_usage_derived <- c(
  "cai", "tai", "csc", "fop", "csc_first75", "csc_q1", "csc_q2", "csc_q3", "csc_q4", "gc3",
  "cnot3_weighted_codon_score"
)
whole_transcript <- c("total_length", "log2_total_length", "transcript_gc")

exclude_cols <- c(cds_specific, utr3_specific, codon_usage_derived, whole_transcript)
stopifnot("All exclusion targets must exist in the source matrix" =
  all(exclude_cols %in% names(fm_full)))
cat("Total columns excluded:", length(exclude_cols), "\n")
cat("  CDS-specific:", length(cds_specific), "| 3'UTR-specific:", length(utr3_specific),
    "| Codon-usage-derived:", length(codon_usage_derived),
    "| Whole-transcript:", length(whole_transcript), "\n")

keep_cols <- setdiff(names(fm_full), exclude_cols)
cat("Columns retained:", length(keep_cols), "of", ncol(fm_full), "\n")

# Sanity: confirm Kozak survived (deliberately kept, not CDS-specific per this doc's logic)
stopifnot(all(c("kozak_score", "kozak_optimal", "kozak_pwm_score") %in% keep_cols))
cat("Kozak retained: confirmed\n")

# Sanity: no CDS/3'UTR-region column leaked through
leaked <- keep_cols[str_detect(keep_cols, "^cds_|^utr3_|^log2_cds_|^log2_utr3_") |
                    (str_detect(keep_cols, "^struct_") & str_detect(keep_cols, "cds|utr3")) |
                    str_detect(keep_cols, "_cds$|_3utr$")]
if (length(leaked) > 0) stop("CDS/3'UTR columns leaked into kept set: ", paste(leaked, collapse = ", "))
cat("CDS/3'UTR leakage check: passed\n")
```

---

# Section 3: Build and Save Matrix

```{r build_save}
fm_no_cds_utr3 <- fm_full %>% dplyr::select(all_of(keep_cols))

stopifnot(nrow(fm_no_cds_utr3) == nrow(fm_full))
cat("Row count preserved:", nrow(fm_no_cds_utr3), "\n")
cat("Resulting matrix:", nrow(fm_no_cds_utr3), "genes x", ncol(fm_no_cds_utr3), "columns\n")

saveRDS(fm_no_cds_utr3,
        here("output", "predictive_modeling", "feature_matrix_no_cds_utr3.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_no_cds_utr3.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
