---
title: "01q — Feature Extraction: no_cds_utr3 minus External Stability/Half-life/Occupancy Data"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Subsets `feature_matrix_no_cds_utr3.rds` by removing the remaining external
half-life/occupancy/stability dataset columns, producing `feature_matrix_no_cds_utr3_no_stability.rds`
for use with notebook 07 (`feature_set = "no_cds_utr3_no_stability"`,
`feature_matrix_path = "feature_matrix_no_cds_utr3_no_stability.rds"`).

**Removed, per 2026-07-17 conversation ("delete all halflife codon occupancy stuff...stability"):**

- `cnot3_ribo_enrichment` (Zhu et al. 2024 CNOT3 KO riboseq enrichment)
- `zhu2024_ctrl_halflife`, `zhu2024_cnot3ko_halflife_logfc` (Zhu et al. 2024 SLAM-seq half-lives)
- `hia2026_ctrl_halflife`, `hia2026_dhx29_halflife_logfc` (Hia et al. 2026 SLAM-seq half-lives)
- `hia2026_dhx29_occupancy` (Hia et al. 2026 DHX29 selective ribosome profiling occupancy)
- `karner2026_mda231_log2_ct` (Karner et al. 2026 MDA-MB-231 SLAM-seq transcription-rate proxy)

This is the entire external literature stability dataset bundle (Section 3.9 of the feature
decision log) -- everything else from `no_cds_utr3` (5'UTR sequence/structure, Kozak, own-TE
features, `sig_*` flags, CLIP, eIF4E/DAP5 sensitivity) is retained.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Source Matrix

```{r load_feature_matrix}
fm_full <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_no_cds_utr3.rds")
)
cat("Source matrix:", nrow(fm_full), "genes x", ncol(fm_full), "columns\n")
```

---

# Section 2: Remove External Stability Columns

```{r select_columns}
stability_cols <- c(
  "cnot3_ribo_enrichment",
  "zhu2024_ctrl_halflife", "zhu2024_cnot3ko_halflife_logfc",
  "hia2026_ctrl_halflife", "hia2026_dhx29_halflife_logfc", "hia2026_dhx29_occupancy",
  "karner2026_mda231_log2_ct"
)
stopifnot("All exclusion targets must exist in the source matrix" =
  all(stability_cols %in% names(fm_full)))
cat("Removing", length(stability_cols), "external stability/halflife/occupancy columns:\n")
cat(paste(" ", stability_cols, collapse = "\n"), "\n")

keep_cols <- setdiff(names(fm_full), stability_cols)
cat("Columns retained:", length(keep_cols), "of", ncol(fm_full), "\n")

leaked <- keep_cols[str_detect(keep_cols, "^cnot3_|^zhu2024_|^hia2026_|^karner2026_")]
if (length(leaked) > 0) stop("Stability columns leaked into kept set: ", paste(leaked, collapse = ", "))
cat("Stability-column leakage check: passed\n")
```

---

# Section 3: Build and Save Matrix

```{r build_save}
fm_no_stability <- fm_full %>% dplyr::select(all_of(keep_cols))

stopifnot(nrow(fm_no_stability) == nrow(fm_full))
cat("Row count preserved:", nrow(fm_no_stability), "\n")
cat("Resulting matrix:", nrow(fm_no_stability), "genes x", ncol(fm_no_stability), "columns\n")

saveRDS(fm_no_stability,
        here("output", "predictive_modeling", "feature_matrix_no_cds_utr3_no_stability.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_no_cds_utr3_no_stability.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
