---
title: "01r — Feature Extraction: eIF4E RIP-seq IP/RNA-Only Feature Matrix"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Subsets `feature_matrix_ripseq.rds` to only the project's own eIF4E RIP-seq IP/RNA log2
ratio across the 4 standard conditions, producing `feature_matrix_eif4e_rna_only.rds` for use
with notebook 07 (`feature_set = "eif4e_rna_only"`,
`feature_matrix_path = "feature_matrix_eif4e_rna_only.rds"`).

**4 features retained (own-data eIF4E RIP-seq, IP/RNA ratio only, no IP/IgG):**

- `eif4e_ip_rna_lfc_normoxia_1hr`
- `eif4e_ip_rna_lfc_hypoxia_1hr`
- `eif4e_ip_rna_lfc_hypoxia_4hr`
- `eif4e_ip_rna_lfc_hypoxia_24hr`

Excludes eIF3d RIP-seq (`ripseq_eif3d_*`), eIF4E IP/IgG (`eif4e_ip_igg_lfc_*`), and everything
else in `external_stability`/`ripseq` (all published CLIP data, sequence/structure features,
external stability datasets, etc.) -- this matrix contains *only* the project's own eIF4E
IP/RNA binding evidence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Source Matrix

```{r load_feature_matrix}
fm_full <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_ripseq.rds")
)
cat("Source matrix:", nrow(fm_full), "genes x", ncol(fm_full), "columns\n")
```

---

# Section 2: Select eIF4E IP/RNA Columns

```{r select_columns}
id_cols <- c("transcript_id", "transcript_id_clean", "gene_id", "symbol", "gene_id_clean")
id_cols_present <- id_cols[id_cols %in% names(fm_full)]

eif4e_rna_cols <- c(
  "eif4e_ip_rna_lfc_normoxia_1hr",
  "eif4e_ip_rna_lfc_hypoxia_1hr",
  "eif4e_ip_rna_lfc_hypoxia_4hr",
  "eif4e_ip_rna_lfc_hypoxia_24hr"
)
stopifnot("All 4 eIF4E IP/RNA columns must exist" = all(eif4e_rna_cols %in% names(fm_full)))
cat("Identifier columns retained:", length(id_cols_present), "\n")
cat("eIF4E IP/RNA columns retained:", length(eif4e_rna_cols), "\n")
cat(paste(" ", eif4e_rna_cols, collapse = "\n"), "\n")

# Sanity: confirm no eIF3d RIP-seq or IP/IgG columns leaked in
stopifnot("Must not include eIF3d RIP-seq or IP/IgG columns" =
  !any(c("ripseq_eif3d_igg_lfc_normoxia_1hr", "ripseq_eif3d_rna_lfc_normoxia_1hr",
         "eif4e_ip_igg_lfc_normoxia_1hr") %in% eif4e_rna_cols))
cat("Leakage check: passed\n")
```

---

# Section 3: Validation

```{r validation_checks}
fm_check <- fm_full %>% dplyr::select(all_of(eif4e_rna_cols))
stopifnot("Log-fold-change values must be finite where non-NA" =
  all(sapply(fm_check, function(x) all(is.finite(x[!is.na(x)])))))
cat("Finite-value check: passed\n")

for (col in eif4e_rna_cols) {
  cat(col, "-- non-NA:", sum(!is.na(fm_check[[col]])), "/", nrow(fm_check),
      "| median:", round(median(fm_check[[col]], na.rm = TRUE), 3), "\n")
}
```

---

# Section 4: Build and Save Matrix

```{r build_save}
fm_eif4e_rna <- fm_full %>%
  dplyr::select(all_of(id_cols_present), all_of(eif4e_rna_cols))

stopifnot(nrow(fm_eif4e_rna) == nrow(fm_full))
cat("Row count preserved:", nrow(fm_eif4e_rna), "\n")
cat("Resulting matrix:", nrow(fm_eif4e_rna), "genes x", ncol(fm_eif4e_rna), "columns\n")

saveRDS(fm_eif4e_rna,
        here("output", "predictive_modeling", "feature_matrix_eif4e_rna_only.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_eif4e_rna_only.rds\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
