---
title: "01s — Feature Extraction: no_cds_utr3_no_stability + Own eIF4E RIP-seq IP/RNA"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Extends `feature_matrix_no_cds_utr3_no_stability.rds` (89 cols: `external_stability` minus
CDS/3'UTR-specific features, codon-usage-derived scores, whole-transcript metrics, and the
external half-life/occupancy/stability dataset bundle) by adding the project's own eIF4E
RIP-seq IP/RNA log2 ratio across the 4 standard conditions, producing
`feature_matrix_no_cds_utr3_no_stability_eif4e.rds` for use with notebook 07
(`feature_set = "no_cds_utr3_no_stability_eif4e"`,
`feature_matrix_path = "feature_matrix_no_cds_utr3_no_stability_eif4e.rds"`).

**4 columns added (own-data eIF4E RIP-seq, IP/RNA ratio only, from `feature_matrix_ripseq.rds`):**

- `eif4e_ip_rna_lfc_normoxia_1hr`
- `eif4e_ip_rna_lfc_hypoxia_1hr`
- `eif4e_ip_rna_lfc_hypoxia_4hr`
- `eif4e_ip_rna_lfc_hypoxia_24hr`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load Base Matrix and eIF4E RIP-seq Source

```{r load_matrices}
fm_base <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_no_cds_utr3_no_stability.rds")
)
cat("Base matrix:", nrow(fm_base), "genes x", ncol(fm_base), "columns\n")

fm_ripseq <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_ripseq.rds")
)
cat("RIP-seq source matrix:", nrow(fm_ripseq), "genes x", ncol(fm_ripseq), "columns\n")
```

---

# Section 2: Extract eIF4E IP/RNA Columns and Join

```{r join_columns}
eif4e_rna_cols <- c(
  "eif4e_ip_rna_lfc_normoxia_1hr",
  "eif4e_ip_rna_lfc_hypoxia_1hr",
  "eif4e_ip_rna_lfc_hypoxia_4hr",
  "eif4e_ip_rna_lfc_hypoxia_24hr"
)
stopifnot("All 4 eIF4E IP/RNA columns must exist in ripseq matrix" =
  all(eif4e_rna_cols %in% names(fm_ripseq)))
stopifnot("None of the 4 eIF4E IP/RNA columns should already be in the base matrix" =
  !any(eif4e_rna_cols %in% names(fm_base)))

eif4e_rna_features <- fm_ripseq %>%
  dplyr::select(gene_id_clean, all_of(eif4e_rna_cols)) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

fm_combined <- fm_base %>%
  dplyr::left_join(eif4e_rna_features, by = "gene_id_clean")

cat("After join:", nrow(fm_combined), "genes x", ncol(fm_combined), "columns\n")
cat("New columns added:", ncol(fm_combined) - ncol(fm_base), "\n")
stopifnot("Row count must be preserved (no fan-out from join)" =
  nrow(fm_combined) == nrow(fm_base))

for (col in eif4e_rna_cols) {
  cat(col, "-- non-NA:", sum(!is.na(fm_combined[[col]])), "/", nrow(fm_combined), "\n")
}
```

---

# Section 3: Validation

```{r validation_checks}
stopifnot("Log-fold-change values must be finite where non-NA" =
  all(sapply(fm_combined[eif4e_rna_cols], function(x) all(is.finite(x[!is.na(x)])))))
cat("Finite-value check: passed\n")

# Sanity: values should match the source ripseq matrix exactly for a spot-check gene
check_genes <- intersect(fm_combined$gene_id_clean, fm_ripseq$gene_id_clean)[1:50]
combined_vals <- fm_combined %>% filter(gene_id_clean %in% check_genes) %>%
  arrange(gene_id_clean) %>% pull(eif4e_ip_rna_lfc_hypoxia_1hr)
source_vals <- fm_ripseq %>% filter(gene_id_clean %in% check_genes) %>%
  arrange(gene_id_clean) %>% pull(eif4e_ip_rna_lfc_hypoxia_1hr)
stopifnot("Joined values must match source" = isTRUE(all.equal(combined_vals, source_vals)))
cat("Join fidelity check (50-gene spot check): passed\n")
```

---

# Section 4: Save Matrix

```{r build_save}
saveRDS(fm_combined,
        here("output", "predictive_modeling", "feature_matrix_no_cds_utr3_no_stability_eif4e.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_no_cds_utr3_no_stability_eif4e.rds\n")
cat("Final matrix:", nrow(fm_combined), "genes x", ncol(fm_combined), "columns\n")
cat("  (", ncol(fm_base), "base cols +", length(eif4e_rna_cols), "own eIF4E IP/RNA features)\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
