---
title: "01u - G4mer rG4 features and association with eIF3d hypoxia TE (1hr)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Ingest G4mer RNA G-quadruplex predictions and test whether 5'UTR rG4 propensity is
associated with eIF3d-dependent translation efficiency in hypoxia at 1hr.

G4mer (`Biociphers/g4mer`, fine-tuned from mRNAbert) accepts a maximum of 70 nt per
input and is 6-mer tokenized. Sequences are scanned with a 70 nt sliding window
(stride 10) and collapsed to a per-transcript maximum, which is the aggregation the
model authors recommend. Scores are produced by `g4mer/02_run_g4mer.py`; transcript
selection comes from `precomputed_most_abundant_tx.rds`, the same universe used for
the ViennaRNA structure features.

**Confound to keep in view throughout:** a maximum taken over more windows has a
higher expected value, so `g4mer_max` is mechanically correlated with region length.
rG4 formation also requires G-runs, so the score necessarily tracks GC and G content.
Every association below is therefore reported both raw and after residualizing on
log2 5'UTR length and 5'UTR GC%.

Contrast direction (MDA-MB-231 convention): positive `te_lfc` = eIF3d **promotes**.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)

set.seed(9)
```

```{r conventions}
group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "grey40",
  "All other genes"        = "grey80"
)
```

# Load G4mer Scores

```{r load_g4mer}
g4_dir <- here("output", "g4mer")

summary_path <- file.path(g4_dir, "g4mer_summary_utr5.tsv")
meta_path    <- here("g4mer", "fasta", "utr5_seq_metadata.tsv")

stopifnot(
  "G4mer 5'UTR summary not found - run g4mer/02_run_g4mer.py --region utr5" =
    file.exists(summary_path),
  "5'UTR sequence metadata not found - run g4mer/01_generate_fasta.R" =
    file.exists(meta_path)
)

g4_utr5 <- read_tsv(summary_path, show_col_types = FALSE)
meta_utr5 <- read_tsv(meta_path, show_col_types = FALSE)

cat("G4mer 5'UTR transcripts scored:", nrow(g4_utr5), "\n")
cat("Metadata transcripts:", nrow(meta_utr5), "\n")
```

## Computational Validation

```{r validate_computational}
expected_cols <- c("transcript_id_clean", "region", "region_length", "n_windows",
                   "g4mer_max", "g4mer_mean", "g4mer_frac_above", "g4mer_argmax_start")
stopifnot(all(expected_cols %in% colnames(g4_utr5)))

# G4mer outputs softmax probabilities
stopifnot("g4mer_max outside [0,1]" =
  all(g4_utr5$g4mer_max >= 0 & g4_utr5$g4mer_max <= 1, na.rm = TRUE))
stopifnot("g4mer_mean outside [0,1]" =
  all(g4_utr5$g4mer_mean >= 0 & g4_utr5$g4mer_mean <= 1, na.rm = TRUE))
stopifnot("g4mer_frac_above outside [0,1]" =
  all(g4_utr5$g4mer_frac_above >= 0 & g4_utr5$g4mer_frac_above <= 1, na.rm = TRUE))

# A maximum can never fall below the mean of the same values
stopifnot("g4mer_max < g4mer_mean" = all(g4_utr5$g4mer_max >= g4_utr5$g4mer_mean))

# Every transcript must have been scored at least once
stopifnot("n_windows < 1" = all(g4_utr5$n_windows >= 1))

# argmax must lie inside the sequence
stopifnot("argmax_start out of bounds" =
  all(g4_utr5$g4mer_argmax_start >= 0 &
      g4_utr5$g4mer_argmax_start < g4_utr5$region_length))

# Region lengths must agree with what was written to FASTA
len_check <- g4_utr5 %>%
  dplyr::select(transcript_id_clean, region_length) %>%
  inner_join(meta_utr5 %>% dplyr::select(transcript_id_clean, meta_len = region_length),
             by = "transcript_id_clean")
stopifnot("region_length disagrees with FASTA metadata" =
  all(len_check$region_length == len_check$meta_len))

cat("All computational checks passed.\n")
cat("g4mer_max range: [", round(min(g4_utr5$g4mer_max), 4), ",",
    round(max(g4_utr5$g4mer_max), 4), "]\n")
cat("Median g4mer_max:", round(median(g4_utr5$g4mer_max), 4), "\n")
```

## Biological Spot-Check

Transcripts whose 5'UTR contains a canonical G4 motif (four G-runs of >= 3 separated
by 1-7 nt loops) must score higher than those without. This tests that the model
output tracks G4 sequence grammar, which catches a strand, tokenization, or
window-indexing error that a range check cannot.

```{r validate_biological}
utr5_fa <- here("g4mer", "fasta", "utr5_sequences.fa")
stopifnot("5'UTR FASTA not found" = file.exists(utr5_fa))

fa_lines <- read_lines(utr5_fa)
hdr_idx  <- which(str_starts(fa_lines, ">"))
fa_names <- str_remove(fa_lines[hdr_idx], "^>") %>% word(1)
# writeXStringSet wraps sequence lines, so collapse the block under each header
fa_ends  <- c(hdr_idx[-1] - 1, length(fa_lines))
fa_seqs  <- map2_chr(hdr_idx + 1, fa_ends,
                     ~ str_c(fa_lines[.x:.y], collapse = ""))

g4_regex <- "([Gg]{3,}[ACGTNacgtn]{1,7}){3,}[Gg]{3,}"
motif_df <- tibble(
  transcript_id_clean = fa_names,
  has_g4_motif        = str_detect(fa_seqs, g4_regex)
)

spot <- g4_utr5 %>% inner_join(motif_df, by = "transcript_id_clean")

med_motif    <- median(spot$g4mer_max[spot$has_g4_motif])
med_no_motif <- median(spot$g4mer_max[!spot$has_g4_motif])

cat("5'UTRs with canonical G4 motif:", sum(spot$has_g4_motif),
    sprintf("(%.1f%%)\n", 100 * mean(spot$has_g4_motif)))
cat("Median g4mer_max, motif present:", round(med_motif, 4), "\n")
cat("Median g4mer_max, motif absent: ", round(med_no_motif, 4), "\n")

stopifnot("G4 motif-containing 5'UTRs do not score above motif-free ones" =
  med_motif > med_no_motif)

motif_wilcox <- wilcox.test(g4mer_max ~ has_g4_motif, data = spot)
cat("Wilcoxon p =", format.pval(motif_wilcox$p.value, digits = 3), "\n")

# NRAS carries a well characterized 5'UTR rG4 (Kumari 2007). Informative only:
# which isoform was selected depends on expression, so this is not asserted.
nras <- spot %>%
  inner_join(meta_utr5 %>% dplyr::select(transcript_id_clean, symbol),
             by = "transcript_id_clean") %>%
  filter(symbol == "NRAS")
if (nrow(nras) > 0) {
  pct <- 100 * mean(spot$g4mer_max <= nras$g4mer_max[1])
  cat(sprintf("NRAS 5'UTR g4mer_max = %.4f (%.1fth percentile)\n",
              nras$g4mer_max[1], pct))
}
```

# Build Feature Table

```{r build_features}
g4_features <- g4_utr5 %>%
  inner_join(
    meta_utr5 %>% dplyr::select(transcript_id_clean, gene_id_clean, symbol,
                                gc_pct, g_pct),
    by = "transcript_id_clean"
  ) %>%
  dplyr::rename(
    g4mer_max_utr5        = g4mer_max,
    g4mer_mean_utr5       = g4mer_mean,
    g4mer_frac_above_utr5 = g4mer_frac_above,
    utr5_length           = region_length,
    gc_utr5               = gc_pct,
    g_content_utr5        = g_pct
  ) %>%
  mutate(log2_utr5_length = log2(utr5_length)) %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol,
                g4mer_max_utr5, g4mer_mean_utr5, g4mer_frac_above_utr5,
                utr5_length, log2_utr5_length, gc_utr5, g_content_utr5)

# Leakage guard: no TE column may enter the saved feature table
stopifnot("TE column leaked into feature table" =
  !any(c("te_lfc", "te_padj") %in% colnames(g4_features)))

cat("Feature table:", nrow(g4_features), "transcripts,",
    ncol(g4_features), "columns\n")

dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)
saveRDS(g4_features, here("output", "predictive_modeling", "feature_matrix_g4mer_utr5.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_g4mer_utr5.rds\n")
```

# Association with eIF3d Hypoxia TE (1hr)

```{r load_te}
te_hyp <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("TE genes (si3d vs sictrl, hypoxia 1hr):", nrow(te_hyp), "\n")

# Positive te_lfc = sictrl TE > si3d TE = knockdown reduces TE = eIF3d promotes
dat <- g4_features %>%
  inner_join(te_hyp %>% dplyr::select(gene_id_clean, te_lfc, te_padj, category),
             by = "gene_id_clean") %>%
  filter(!is.na(te_lfc))

cat("Genes with both G4mer score and TE:", nrow(dat), "\n")
```

## Raw Correlation

```{r raw_correlation}
n_dat <- nrow(dat)
sp <- cor.test(dat$g4mer_max_utr5, dat$te_lfc, method = "spearman")

cat(sprintf("Spearman rho = %.4f, p = %s, n = %d\n",
            sp$estimate, format.pval(sp$p.value, digits = 3), n_dat))

# Confounders, for magnitude comparison
for (v in c("log2_utr5_length", "gc_utr5", "g_content_utr5")) {
  ct <- cor.test(dat[[v]], dat$te_lfc, method = "spearman")
  cat(sprintf("  %-18s vs te_lfc: rho = %+.4f, p = %s\n",
              v, ct$estimate, format.pval(ct$p.value, digits = 3)))
}
cat(sprintf("  %-18s vs g4mer_max: rho = %+.4f\n", "log2_utr5_length",
            cor(dat$log2_utr5_length, dat$g4mer_max_utr5, method = "spearman")))
cat(sprintf("  %-18s vs g4mer_max: rho = %+.4f\n", "gc_utr5",
            cor(dat$gc_utr5, dat$g4mer_max_utr5, method = "spearman")))
```

## Partial Correlation, Controlling Length and GC

The raw correlation cannot be interpreted on its own: `g4mer_max` rises with the
number of windows scanned and with G content. Residualizing both variables on log2
5'UTR length and 5'UTR GC% isolates the rG4-specific component.

```{r partial_correlation}
resid_g4 <- residuals(lm(g4mer_max_utr5 ~ log2_utr5_length + gc_utr5, data = dat))
resid_te <- residuals(lm(te_lfc ~ log2_utr5_length + gc_utr5, data = dat))

pc <- cor.test(resid_g4, resid_te, method = "spearman")
cat(sprintf("Partial Spearman rho (length + GC controlled) = %.4f, p = %s, n = %d\n",
            pc$estimate, format.pval(pc$p.value, digits = 3), n_dat))

cat(sprintf("\nRaw rho     = %+.4f\nPartial rho = %+.4f\n",
            sp$estimate, pc$estimate))
```

```{r plot_scatter, fig.width = 6, fig.height = 5}
p_scatter <- ggplot(dat, aes(x = g4mer_max_utr5, y = te_lfc)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "grey30") +
  geom_smooth(method = "lm", se = TRUE, colour = "#E69F00", linewidth = 0.8) +
  labs(
    x = "5'UTR G4mer max rG4 probability",
    y = "TE log2FC (si3d vs siCtrl, hypoxia 1hr)",
    title = "5'UTR rG4 propensity vs eIF3d-dependent TE"
  ) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("Spearman rho = %.3f, p = %s, n = %d",
                           sp$estimate, format.pval(sp$p.value, digits = 3), n_dat)) +
  theme_classic(base_size = 12)

print(p_scatter)
ggsave(here("plots", "01u_g4mer_utr5_vs_te_lfc_scatter.pdf"),
       p_scatter, width = 6, height = 5)
```

## TE by G4mer Quartile

```{r quartile_analysis, fig.width = 6, fig.height = 5}
dat_q <- dat %>%
  mutate(g4_quartile = ntile(g4mer_max_utr5, 4),
         g4_quartile = factor(g4_quartile,
                              labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)")))

quartile_summary <- dat_q %>%
  group_by(g4_quartile) %>%
  summarise(
    n            = dplyr::n(),
    median_te    = median(te_lfc),
    mean_te      = mean(te_lfc),
    median_g4    = median(g4mer_max_utr5),
    median_len   = median(utr5_length),
    median_gc    = median(gc_utr5),
    .groups = "drop"
  )
print(quartile_summary)

q4_vs_q1 <- wilcox.test(
  te_lfc ~ g4_quartile,
  data = dat_q %>% filter(g4_quartile %in% c("Q1 (lowest)", "Q4 (highest)")) %>%
    droplevels()
)
cat(sprintf("\nQ4 vs Q1 Wilcoxon p = %s (n_Q1 = %d, n_Q4 = %d)\n",
            format.pval(q4_vs_q1$p.value, digits = 3),
            sum(dat_q$g4_quartile == "Q1 (lowest)"),
            sum(dat_q$g4_quartile == "Q4 (highest)")))

p_quart <- ggplot(dat_q, aes(x = g4_quartile, y = te_lfc)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, fill = "grey90",
               linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x = "5'UTR G4mer max rG4 probability quartile",
    y = "TE log2FC (si3d vs siCtrl, hypoxia 1hr)",
    title = "eIF3d-dependent TE across 5'UTR rG4 quartiles"
  ) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("Q4 vs Q1 Wilcoxon p = %s, n = %d",
                           format.pval(q4_vs_q1$p.value, digits = 3), n_dat)) +
  theme_classic(base_size = 12)

print(p_quart)
ggsave(here("plots", "01u_g4mer_utr5_te_by_quartile.pdf"),
       p_quart, width = 6, height = 5)
```

## G4mer Score by Gene Group

Compares the eIF3d hypoxia-promoted set against a size-matched negative control set
and all remaining genes, the standard grouping used across this project.

```{r gene_groups, fig.width = 6.5, fig.height = 5}
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE) %>%
  pull(ensembl_gene) %>% unique()

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  pull(gene_id_clean) %>% unique()

# Size-match the negative control set to the positive set
set.seed(9)
neg_pool_avail <- intersect(neg_pool, dat$gene_id_clean)
n_pos_avail    <- length(intersect(promotes, dat$gene_id_clean))
neg_matched    <- sample(neg_pool_avail, min(n_pos_avail, length(neg_pool_avail)))

dat_grp <- dat %>%
  mutate(group = case_when(
    gene_id_clean %in% promotes    ~ "eIF3d hypoxia promotes",
    gene_id_clean %in% neg_matched ~ "Negative control",
    TRUE                           ~ "All other genes"
  )) %>%
  mutate(group = factor(group, levels = names(group_colors)))

group_summary <- dat_grp %>%
  group_by(group) %>%
  summarise(
    n             = dplyr::n(),
    median_g4     = median(g4mer_max_utr5),
    median_len    = median(utr5_length),
    median_gc     = median(gc_utr5),
    .groups = "drop"
  )
print(group_summary)

grp_wilcox <- wilcox.test(
  g4mer_max_utr5 ~ group,
  data = dat_grp %>%
    filter(group %in% c("eIF3d hypoxia promotes", "Negative control")) %>%
    droplevels()
)
n_pos <- sum(dat_grp$group == "eIF3d hypoxia promotes")
n_neg <- sum(dat_grp$group == "Negative control")
cat(sprintf("\nPromoted vs negative control Wilcoxon p = %s (n = %d vs %d)\n",
            format.pval(grp_wilcox$p.value, digits = 3), n_pos, n_neg))

p_group <- ggplot(dat_grp, aes(x = g4mer_max_utr5, fill = group, colour = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_fill_manual(values = group_colors) +
  scale_colour_manual(values = group_colors) +
  labs(
    x = "5'UTR G4mer max rG4 probability",
    y = "Density",
    title = "5'UTR rG4 propensity by gene group",
    fill = NULL, colour = NULL
  ) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = sprintf("promotes vs neg ctrl p = %s\nn = %d vs %d",
                           format.pval(grp_wilcox$p.value, digits = 3),
                           n_pos, n_neg)) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")

print(p_group)
ggsave(here("plots", "01u_g4mer_utr5_by_gene_group.pdf"),
       p_group, width = 6.5, height = 5)
```

## Summary

```{r final_summary}
cat("=== 5'UTR G4mer vs eIF3d hypoxia TE (1hr) ===\n\n")
cat(sprintf("Genes analyzed:            %d\n", n_dat))
cat(sprintf("Raw Spearman rho:          %+.4f (p = %s)\n",
            sp$estimate, format.pval(sp$p.value, digits = 3)))
cat(sprintf("Partial rho (len + GC):    %+.4f (p = %s)\n",
            pc$estimate, format.pval(pc$p.value, digits = 3)))
cat(sprintf("Q4 vs Q1 median TE diff:   %+.4f\n",
            quartile_summary$median_te[4] - quartile_summary$median_te[1]))
cat(sprintf("Promoted vs neg ctrl p:    %s\n",
            format.pval(grp_wilcox$p.value, digits = 3)))
cat("\nInterpretation note: if the partial correlation collapses toward zero, the\n")
cat("raw association is carried by 5'UTR length and GC content rather than by rG4\n")
cat("propensity specifically. Compare against the CDS and 3'UTR runs as a\n")
cat("region-specificity control once the cluster job returns.\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
