---
title: "01v - Positional GC Features (GC1 / GC2 / GC3)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  drop_aggregate_gc: true
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    drop_aggregate_gc = TRUE
  )
}
```

# Overview

Adds **positional GC content** to the feature matrix: GC1, GC2 and GC3, the GC fraction at
each of the three codon positions.

The motivation is that `cds_gc` and `gc3` are 0.935 correlated in this dataset and are
mutually redundant in the model (notebook 07c, ablation section: dropping either alone costs
essentially nothing). Neither can distinguish two very different biological claims:

- *"eIF3d targets have GC-rich coding sequences"* - a bulk composition effect, which is
  partly amino-acid composition and therefore partly protein-level.
- *"eIF3d prefers particular synonymous codons"* - a wobble-position effect, which is free to
  vary without changing the encoded protein.

Splitting CDS GC by codon position separates them. **GC3 is the synonymous position**: it can
change without changing the amino acid in most codon families. **GC1 and GC2 are largely
amino-acid-determined** - changing them usually changes the protein. So if the eIF3d signal
is synonymous codon preference, GC3 should carry it while GC1 and GC2 should not.

## Design note: this notebook drops the aggregate GC columns

With `drop_aggregate_gc = TRUE` (the default) the output matrix **removes `cds_gc` and
`transcript_gc`**. This is deliberate. `cds_gc` is approximately the mean of GC1/GC2/GC3, so
leaving it in gives the forest an aggregate to fall back on and the positional split never
gets tested - it would simply reproduce the redundancy already documented in 07c. Removing it
forces the model to choose among positions, which is the entire point of the comparison.

Set `drop_aggregate_gc = FALSE` to keep them (useful if you want to run the ablation
comparisons in 07c, which reference `cds_gc` by name).

## Inputs and outputs

- Reads `feature_matrix_external_stability.rds` (unmodified)
- Writes `feature_matrix_positional_gc.rds`

**No existing feature matrix is modified.** This notebook only creates a new one.

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
```

# Load Base Feature Matrix

```{r load_base}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Base feature matrix:", nrow(features), "transcripts x", ncol(features), "columns\n")

stopifnot(
  "base matrix is missing transcript_id_clean" = "transcript_id_clean" %in% colnames(features),
  "base matrix is missing gc3 (needed for the validation check)" = "gc3" %in% colnames(features),
  "base matrix is missing cds_gc" = "cds_gc" %in% colnames(features)
)
```

# Extract CDS Sequences

Same construction as `01_feature_extraction.Rmd` so transcript identity matches exactly.

```{r extract_cds}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_sequences <- extractTranscriptSeqs(genome, cds) %>% as.data.frame()

cds_seq <- cds_sequences %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::rename("cds_sequence" = "x") %>%
  dplyr::select(transcript_id_clean, cds_sequence)

cat("Extracted CDS sequences:", nrow(cds_seq), "transcripts\n")
```

# Compute Positional GC

`calculate_gc_at_position()` generalises `calculate_gc3()` from `01_feature_extraction.Rmd`
to any codon position, so GC3 computed here is directly comparable to the existing column.

```{r compute_positional_gc}
calculate_gc_at_position <- function(seqs, position) {
  stopifnot("position must be 1, 2 or 3" = position %in% 1:3)
  seqs <- str_to_upper(seqs)
  vapply(seqs, function(seq) {
    if (is.na(seq) || nchar(seq) < 3L) return(NA_real_)
    n_codons <- nchar(seq) %/% 3L
    starts   <- seq.int(position, n_codons * 3L, 3L)
    bases    <- substr(rep.int(seq, n_codons), starts, starts)
    sum(bases %in% c("G", "C")) * 100 / n_codons
  }, numeric(1L), USE.NAMES = FALSE)
}

positional_gc <- cds_seq %>%
  mutate(
    gc1        = calculate_gc_at_position(cds_sequence, 1),
    gc2        = calculate_gc_at_position(cds_sequence, 2),
    gc3_recomp = calculate_gc_at_position(cds_sequence, 3)
  ) %>%
  dplyr::select(transcript_id_clean, gc1, gc2, gc3_recomp)

cat("Computed positional GC for", nrow(positional_gc), "transcripts\n")
cat("  Median GC1:", round(median(positional_gc$gc1, na.rm = TRUE), 2), "%\n")
cat("  Median GC2:", round(median(positional_gc$gc2, na.rm = TRUE), 2), "%\n")
cat("  Median GC3:", round(median(positional_gc$gc3_recomp, na.rm = TRUE), 2), "%\n")
```

## Computational checks

```{r checks_computational}
stopifnot(
  "gc1 out of [0, 100]" = all(positional_gc$gc1 >= 0 & positional_gc$gc1 <= 100, na.rm = TRUE),
  "gc2 out of [0, 100]" = all(positional_gc$gc2 >= 0 & positional_gc$gc2 <= 100, na.rm = TRUE),
  "gc3 out of [0, 100]" = all(positional_gc$gc3_recomp >= 0 & positional_gc$gc3_recomp <= 100, na.rm = TRUE),
  "gc3 median not in expected range [40, 65]" =
    between(median(positional_gc$gc3_recomp, na.rm = TRUE), 40, 65)
)
cat("Range checks passed\n")
```

## Reproduction check against the existing GC3 column

The strongest available correctness test: GC3 recomputed here must reproduce the `gc3`
column already in the feature matrix, which was computed independently in `01_`. If the
codon-position indexing were off by one, this would fail immediately.

```{r checks_reproduce_gc3}
gc3_compare <- features %>%
  dplyr::select(transcript_id_clean, gc3_existing = gc3) %>%
  inner_join(positional_gc, by = "transcript_id_clean") %>%
  filter(!is.na(gc3_existing), !is.na(gc3_recomp))

max_diff <- max(abs(gc3_compare$gc3_recomp - gc3_compare$gc3_existing))
r_gc3    <- cor(gc3_compare$gc3_recomp, gc3_compare$gc3_existing)

cat("GC3 reproduction check (n =", nrow(gc3_compare), "transcripts):\n")
cat("  max |recomputed - existing|:", format(max_diff, digits = 3), "\n")
cat("  Pearson r:", round(r_gc3, 8), "\n")

stopifnot(
  "recomputed GC3 does not match the existing gc3 column - check codon position indexing" =
    max_diff < 1e-8
)
cat("GC3 reproduction check passed: positional indexing is correct\n")
```

## Biological spot-checks

GC1 > GC2 is a well-established feature of coding sequences: position 1 is enriched for G
in many amino acid families (all four-fold GG* and GA* families start with G), while position
2 carries the hydrophobicity signal and is A/T-rich for hydrophobic residues. GC3 is the most
variable across genes because it is the least constrained.

```{r checks_biological}
med_gc1 <- median(positional_gc$gc1, na.rm = TRUE)
med_gc2 <- median(positional_gc$gc2, na.rm = TRUE)
med_gc3 <- median(positional_gc$gc3_recomp, na.rm = TRUE)
sd_gc1  <- sd(positional_gc$gc1, na.rm = TRUE)
sd_gc2  <- sd(positional_gc$gc2, na.rm = TRUE)
sd_gc3  <- sd(positional_gc$gc3_recomp, na.rm = TRUE)

cat("Median GC by codon position:  GC1 =", round(med_gc1, 1),
    "| GC2 =", round(med_gc2, 1), "| GC3 =", round(med_gc3, 1), "\n")
cat("SD by codon position:         GC1 =", round(sd_gc1, 1),
    "| GC2 =", round(sd_gc2, 1), "| GC3 =", round(sd_gc3, 1), "\n")

stopifnot(
  "GC1 should exceed GC2 in human coding sequence" = med_gc1 > med_gc2,
  "GC3 should be the most variable position (least constrained)" =
    sd_gc3 > sd_gc1 && sd_gc3 > sd_gc2
)
cat("Biological spot-checks passed\n")

# ACTB CDS GC, the established spot-check from 01_, recomputed as a positional mean
actb <- features %>%
  filter(symbol == "ACTB") %>%
  dplyr::select(transcript_id_clean, cds_gc) %>%
  inner_join(positional_gc, by = "transcript_id_clean") %>%
  slice(1)
if (nrow(actb) == 1) {
  actb_mean_pos <- mean(c(actb$gc1, actb$gc2, actb$gc3_recomp))
  cat("ACTB: cds_gc =", round(actb$cds_gc, 1),
      "| mean(GC1,GC2,GC3) =", round(actb_mean_pos, 1), "\n")
  stopifnot(
    "ACTB CDS GC outside expected 50-70%" = between(actb$cds_gc, 50, 70),
    "mean of positional GC should approximate cds_gc" =
      abs(actb_mean_pos - actb$cds_gc) < 2
  )
  cat("ACTB spot-check passed\n")
}
```

# GC3 Residual: the Synonymous Component Beyond Bulk GC

A single-feature version of the same idea. Regressing GC3 on `cds_gc` and keeping the
residual isolates *"GC3 beyond what bulk CDS GC already predicts"*. Because it is orthogonal
to `cds_gc` by construction, it can sit alongside the aggregate columns without the
redundancy problem, so it is retained even when `drop_aggregate_gc = TRUE`.

```{r gc3_residual}
resid_input <- features %>%
  dplyr::select(transcript_id_clean, cds_gc) %>%
  inner_join(positional_gc, by = "transcript_id_clean") %>%
  filter(!is.na(cds_gc), !is.na(gc3_recomp))

gc3_fit <- lm(gc3_recomp ~ cds_gc, data = resid_input)
resid_input$gc3_residual <- residuals(gc3_fit)

cat("GC3 ~ cds_gc:  R^2 =", round(summary(gc3_fit)$r.squared, 4),
    "| residual SD =", round(sd(resid_input$gc3_residual), 3), "\n")
cat("So", round(100 * summary(gc3_fit)$r.squared, 1),
    "% of GC3 variance is explained by bulk CDS GC;",
    "the residual is the remaining synonymous-specific component.\n")

stopifnot(
  "gc3_residual must be orthogonal to cds_gc by construction" =
    abs(cor(resid_input$gc3_residual, resid_input$cds_gc)) < 1e-8,
  "gc3_residual is not finite" = all(is.finite(resid_input$gc3_residual))
)
cat("Orthogonality check passed\n")

positional_gc <- positional_gc %>%
  left_join(resid_input %>% dplyr::select(transcript_id_clean, gc3_residual),
            by = "transcript_id_clean")
```

# Assemble the Output Matrix

```{r assemble}
new_cols <- c("gc1", "gc2", "gc3_residual")

features_out <- features %>%
  left_join(positional_gc %>% dplyr::select(transcript_id_clean, all_of(new_cols)),
            by = "transcript_id_clean")

stopifnot(
  "join changed the row count" = nrow(features_out) == nrow(features),
  "new feature columns are missing" = all(new_cols %in% colnames(features_out))
)

dropped <- character(0)
if (isTRUE(params$drop_aggregate_gc)) {
  dropped <- intersect(c("cds_gc", "transcript_gc"), colnames(features_out))
  features_out <- features_out %>% dplyr::select(-all_of(dropped))
  cat("Dropped aggregate GC columns so the positional split is actually tested:",
      paste(dropped, collapse = ", "), "\n")
} else {
  cat("Kept aggregate GC columns (drop_aggregate_gc = FALSE)\n")
}

cat("\nCoverage of the new features:\n")
for (f in new_cols) {
  cat(sprintf("  %-14s %5.1f%% non-NA\n", f, 100 * mean(!is.na(features_out[[f]]))))
}

cat("\nCorrelations among the CDS GC family (complete cases):\n")
gc_fam <- intersect(c("gc1", "gc2", "gc3", "gc3_residual", "cds_gc", "transcript_gc"),
                    colnames(features_out))
print(round(cor(features_out[, gc_fam], method = "spearman", use = "complete.obs"), 3))

cat("\nOutput matrix:", nrow(features_out), "transcripts x", ncol(features_out), "columns\n")
cat("  (base had", ncol(features), "columns; +", length(new_cols), "added, -",
    length(dropped), "dropped)\n")
```

# Export

```{r export}
out_path <- here("output", "predictive_modeling", "feature_matrix_positional_gc.rds")
saveRDS(features_out, out_path)
cat("Wrote", basename(out_path), "\n")

# Confirm the base matrix was not touched
base_path <- here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
base_check <- readRDS(base_path)
stopifnot(
  "base feature matrix was modified - it must be left untouched" =
    identical(dim(base_check), dim(features)) && identical(colnames(base_check), colnames(features))
)
cat("Verified feature_matrix_external_stability.rds is unchanged\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
