---
title: "Negative Control Gene Set for Random Forest"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis identifies genes that are **not regulated** by eIF3d knockdown in any condition (normoxia or hypoxia). These genes serve as a negative control set for Random Forest classification.

Normoxia is represented by the single time-regressed alltimepoints analysis; hypoxia uses the three individual timepoint comparisons (1hr, 4hr, 24hr).

## Criteria

A gene must satisfy **all** of the following across all 4 eIF3d comparisons (1 normoxia alltimepoints + 3 hypoxia timepoints):

- `|te_lfc| < 0.5` (small effect size)
- `te_padj > 0.3` (clearly non-significant)

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

# Define File List

```{r define_files}
# Normoxia: single time-regressed alltimepoints analysis
normoxia_comparison <- tibble(
  protein = "si3d",
  condition = "normoxia",
  timepoint = "alltimepoints",
  file = here("output", "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv"),
  label = "si3d_normoxia_alltimepoints"
)

# Hypoxia: individual timepoint comparisons
hypoxia_comparisons <- expand.grid(
  protein = "si3d",
  condition = "hypoxia",
  timepoint = c("1hr", "4hr", "24hr"),
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  mutate(
    file = here("output",
                paste0("translation_categories_", protein, "_vs_sictrl_",
                       condition, "_", timepoint, ".csv")),
    label = paste(protein, condition, timepoint, sep = "_")
  )

comparisons <- bind_rows(normoxia_comparison, hypoxia_comparisons)

cat("Defined", nrow(comparisons), "comparisons:\n")
print(comparisons %>% dplyr::select(label, file))
```

# Load All 4 Datasets

```{r load_data}
all_data <- comparisons %>%
  rowwise() %>%
  mutate(data = list(read_csv(file, show_col_types = FALSE))) %>%
  ungroup()

# Verify all files loaded
cat("\nRows per comparison:\n")
all_data %>%
  mutate(n_genes = map_int(data, nrow)) %>%
  dplyr::select(label, n_genes) %>%
  print(n = 4)
```

# Filter Each Dataset

```{r filter_datasets}
# For each comparison, flag genes passing criteria: |te_lfc| < 0.5 AND te_padj > 0.05
passing_genes <- all_data %>%
  mutate(
    passing = map(data, function(df) {
      df %>%
        filter(abs(te_lfc) < 0.5, te_padj > 0.3) %>%
        pull(gene_id)
    })
  )

# Report how many genes pass each individual filter
cat("\nGenes passing criteria per comparison:\n")
passing_genes %>%
  mutate(n_passing = map_int(passing, length)) %>%
  dplyr::select(label, n_passing) %>%
  print(n = 4)
```

# Intersect Across All 4 Comparisons

```{r intersect}
# Find genes present in ALL 4 datasets
all_gene_sets <- map(all_data$data, ~ pull(.x, gene_id))
genes_in_all <- Reduce(intersect, all_gene_sets)
cat("Genes measured in all 4 comparisons:", length(genes_in_all), "\n")

# Find genes passing criteria in ALL 4 comparisons
passing_sets <- passing_genes$passing
negative_control_ids <- Reduce(intersect, passing_sets)

# Also require presence in all 4 datasets
negative_control_ids <- intersect(negative_control_ids, genes_in_all)
cat("Genes passing criteria in all 4 comparisons:", length(negative_control_ids), "\n")
```

# Build Output Table

```{r build_output}
# Get symbol mapping from one of the datasets
symbol_map <- all_data$data[[1]] %>%
  dplyr::select(gene_id, symbol) %>%
  distinct()

# Build output with gene_id, symbol, and count of comparisons measured
negative_control_genes <- tibble(gene_id = negative_control_ids) %>%
  left_join(symbol_map, by = "gene_id") %>%
  mutate(n_comparisons = 6L) %>%
  arrange(gene_id)

nrow(negative_control_genes )

cat("\nNegative control set:", nrow(negative_control_genes), "genes\n")
head(negative_control_genes, 20)
```

# QC & Summary

```{r qc_summary}
# Verify: no gene in the negative set has |te_lfc| >= 0.5 or te_padj <= 0.5 in any condition
cat("=== Verification ===\n\n")

verification <- all_data %>%
  mutate(
    violations = map(data, function(df) {
      df %>%
        filter(gene_id %in% negative_control_ids) %>%
        filter(abs(te_lfc) >= 0.5 | te_padj <= 0.3)
    })
  )

n_violations <- sum(map_int(verification$violations, nrow))
cat("Total violations across all comparisons:", n_violations, "\n")

if (n_violations > 0) {
  warning("Found violations! Inspect the verification object.")
} else {
  cat("PASS: No gene in the negative set violates the criteria in any condition.\n")
}
```

```{r qc_distributions, fig.width=10, fig.height=6}
# Collect TE LFC values for negative control genes across all comparisons
neg_ctrl_te <- all_data %>%
  mutate(
    neg_te = map2(data, label, function(df, lbl) {
      df %>%
        filter(gene_id %in% negative_control_ids) %>%
        dplyr::select(gene_id, te_lfc, te_padj) %>%
        mutate(comparison = lbl)
    })
  ) %>%
  pull(neg_te) %>%
  bind_rows()

cat("\n=== TE LFC distribution in negative control set ===\n")
cat("Overall:\n")
print(summary(neg_ctrl_te$te_lfc))

cat("\nPer comparison:\n")
neg_ctrl_te %>%
  group_by(comparison) %>%
  summarise(
    mean_te_lfc = mean(te_lfc),
    median_te_lfc = median(te_lfc),
    sd_te_lfc = sd(te_lfc),
    min_te_lfc = min(te_lfc),
    max_te_lfc = max(te_lfc),
    .groups = "drop"
  ) %>%
  print(n = 6)

# Plot TE LFC distributions
ggplot(neg_ctrl_te, aes(x = te_lfc)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~comparison, ncol = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "grey40") +
  theme_minimal() +
  labs(title = "TE LFC Distribution of Negative Control Genes",
       subtitle = paste(nrow(negative_control_genes), "genes across 4 eIF3d comparisons"),
       x = "TE Log2 Fold Change", y = "Count")
```

```{r qc_padj, fig.width=10, fig.height=6}
# Plot te_padj distributions
ggplot(neg_ctrl_te, aes(x = te_padj)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  facet_wrap(~comparison, ncol = 3) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "TE padj Distribution of Negative Control Genes",
       x = "TE padj", y = "Count")
```

# Export

```{r export}
# Create output directory if needed
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

# Save as CSV
write_csv(negative_control_genes,
          here("output", "predictive_modeling", "negative_control_genes.csv"))
cat("Exported: output/predictive_modeling/negative_control_genes.csv\n")

# Save as RDS
saveRDS(negative_control_genes,
        here("output", "predictive_modeling", "negative_control_genes.rds"))
cat("Exported: output/predictive_modeling/negative_control_genes.rds\n")

cat("\nFinal negative control set:", nrow(negative_control_genes), "genes\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
