---
title: "Random Forest Model for Hypoxia TE Prediction"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis builds and evaluates a random forest model to predict Translation Efficiency (TE) Log2 Fold Change at 1 hour hypoxia using mRNA sequence and structural features.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ranger)
library(caret)
library(scales)
```

# Load Feature Matrix

```{r load_features}
# Load feature matrix
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr.rds"))

cat("Loaded feature matrix:", nrow(features), "genes x", ncol(features), "features\n")

# Check structure
cat("\nColumn names:\n")
print(head(colnames(features), 30))
cat("...\n")
```

# Preprocessing

## Remove Near-Zero Variance Features

```{r remove_nzv}
# Identify feature columns (exclude identifiers and target)
# Include all possible ID column variants
id_cols <- c("gene_id", "gene_id.x", "gene_id.y", "gene_id_clean",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin")

# Also exclude any column containing "gene_id" or "transcript_id" patterns
all_cols <- colnames(features)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))

feature_cols <- setdiff(colnames(features), id_cols)

cat("Columns excluded as identifiers:\n")
print(intersect(colnames(features), id_cols))
cat("\nInitial feature count:", length(feature_cols), "\n")

# Create feature-only dataframe for preprocessing
feature_data <- features %>%
  dplyr::select(all_of(feature_cols))

# Check for near-zero variance features
nzv <- nearZeroVar(feature_data, saveMetrics = TRUE)
nzv_features <- rownames(nzv)[nzv$nzv]

cat("Near-zero variance features to remove:", length(nzv_features), "\n")

# Remove NZV features
feature_cols_clean <- setdiff(feature_cols, nzv_features)
cat("Features after NZV removal:", length(feature_cols_clean), "\n")
```

## Handle Missing Values

```{r handle_missing}
# Check missing values in remaining features
missing_counts <- features %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:\n")
print(missing_counts)

# Impute missing values with median
features_imputed <- features %>%
  mutate(across(all_of(feature_cols_clean), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Verify no missing values in features
remaining_na <- features_imputed %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  summarise(total_na = sum(is.na(.))) %>%
  pull(total_na)

cat("\nRemaining missing values after imputation:", remaining_na, "\n")
```

## Remove Highly Correlated Features

```{r remove_correlated}
# Calculate correlation matrix (only for numeric features)
feature_matrix_for_cor <- features_imputed %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  dplyr::select(where(is.numeric))

# Find highly correlated pairs (r > 0.9)
cor_matrix <- cor(feature_matrix_for_cor, use = "pairwise.complete.obs")

# Find features to remove based on correlation
high_cor <- findCorrelation(cor_matrix, cutoff = 0.9, names = TRUE)

cat("Highly correlated features to remove:", length(high_cor), "\n")
if (length(high_cor) > 0) {
  cat("Examples:", paste(head(high_cor, 10), collapse = ", "), "\n")
}

# Remove highly correlated features
feature_cols_final <- setdiff(feature_cols_clean, high_cor)
cat("Final feature count:", length(feature_cols_final), "\n")
```

# Prepare Data for Modeling

```{r prepare_data}
# Create final dataset for modeling
model_data <- features_imputed %>%
  dplyr::select(gene_id, transcript_id, transcript_id_clean, symbol, te_lfc, te_lfc_bin, all_of(feature_cols_final)) %>%
  filter(!is.na(te_lfc), !is.na(te_lfc_bin))

cat("Final dataset dimensions:", nrow(model_data), "genes x", length(feature_cols_final), "features\n")

# Summary of target variable
cat("\nTarget variable (te_lfc) summary:\n")
print(summary(model_data$te_lfc))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

# Stratified split by TE LFC quantiles
train_idx <- createDataPartition(model_data$te_lfc_bin, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "genes\n")
cat("Test set:", nrow(test_data), "genes\n")

# Verify stratification
cat("\nTrain set TE bin distribution:\n")
print(prop.table(table(train_data$te_lfc_bin)))

cat("\nTest set TE bin distribution:\n")
print(prop.table(table(test_data$te_lfc_bin)))
```

# Build Random Forest Model

```{r train_model}
# Prepare training data (exclude identifiers and bin column)
train_features <- train_data %>%
  dplyr::select(-gene_id, -transcript_id, -transcript_id_clean, -symbol, -te_lfc_bin)

test_features <- test_data %>%
  dplyr::select(-gene_id, -transcript_id, -transcript_id_clean, -symbol, -te_lfc_bin)

cat("Training random forest model...\n")
cat("Features:", ncol(train_features) - 1, "\n")
cat("Training samples:", nrow(train_features), "\n")

# Train random forest with ranger
set.seed(9)
rf_model <- ranger(
  formula = te_lfc ~ .,
  data = train_features,
  num.trees = 500,
  importance = "impurity",
  seed = 9,
  verbose = TRUE
)

cat("\nModel training complete!\n")
print(rf_model)
```

# Model Evaluation

```{r evaluate_model}
# Make predictions on test set
predictions <- predict(rf_model, test_features)$predictions

# Calculate metrics
metrics <- data.frame(
  RMSE = sqrt(mean((predictions - test_data$te_lfc)^2)),
  MAE = mean(abs(predictions - test_data$te_lfc)),
  R_squared = cor(predictions, test_data$te_lfc)^2,
  Pearson_r = cor(predictions, test_data$te_lfc),
  Spearman_rho = cor(predictions, test_data$te_lfc, method = "spearman")
)

cat("\n========================================\n")
cat("Model Performance on Test Set\n")
cat("========================================\n\n")
print(metrics)
cat("\n")
```

# Visualize Predictions vs Actual

```{r plot_predictions, fig.width=8, fig.height=6}
# Create prediction dataframe
pred_df <- data.frame(
  actual = test_data$te_lfc,
  predicted = predictions,
  symbol = test_data$symbol
)

# Scatter plot of predictions vs actual
p_scatter <- ggplot(pred_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "lm", color = "#0072B2", se = TRUE) +
  theme_cowplot() +
  labs(
    title = "Random Forest: Predicted vs Actual TE LFC",
    subtitle = paste0("RÂ˛ = ", round(metrics$R_squared, 3),
                      ", Pearson r = ", round(metrics$Pearson_r, 3),
                      ", RMSE = ", round(metrics$RMSE, 3)),
    x = "Actual TE log2 Fold Change",
    y = "Predicted TE log2 Fold Change"
  ) +
  coord_equal()

print(p_scatter)

ggsave(here("plots", "rf_predictions_vs_actual_te_1hr.pdf"),
       plot = p_scatter, width = 8, height = 6)
```

# Residual Analysis

```{r residual_analysis, fig.width=10, fig.height=4}
# Calculate residuals
pred_df$residual <- pred_df$actual - pred_df$predicted

# Residual plots
p_resid1 <- ggplot(pred_df, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "#0072B2") +
  theme_cowplot() +
  labs(title = "Residuals vs Predicted",
       x = "Predicted TE LFC", y = "Residual")

p_resid2 <- ggplot(pred_df, aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#0072B2", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  theme_cowplot() +
  labs(title = "Residual Distribution",
       x = "Residual", y = "Count")

p_resid_combined <- plot_grid(p_resid1, p_resid2, ncol = 2)
print(p_resid_combined)

ggsave(here("plots", "rf_residual_analysis_te_1hr.pdf"),
       plot = p_resid_combined, width = 10, height = 4)
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
# Extract feature importance
importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Show top 30 features
cat("\nTop 30 Most Important Features:\n")
print(importance_df %>% head(30))

# Create clean feature names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

# Plot top 30 features
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Top 30 Features for Predicting 1hr Hypoxia TE LFC",
    subtitle = "Random Forest Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", "rf_feature_importance_te_1hr.pdf"),
       plot = p_importance, width = 10, height = 10)
```

# Feature Category Analysis

```{r feature_categories, fig.width=10, fig.height=6}
# Categorize features
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^utr5_4mer_") ~ "5' UTR 4-mers",
      str_detect(feature, "^rbp_") ~ "RBP Motifs",
      str_detect(feature, "length|Length") ~ "Region Lengths",
      str_detect(feature, "_gc$|gc3$|GC") ~ "GC Content",
      TRUE ~ "Other"
    )
  )

# Summarize importance by category
category_importance <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    max_importance = max(importance),
    top_feature = feature[which.max(importance)]
  ) %>%
  arrange(desc(total_importance))

cat("\nFeature Importance by Category:\n")
print(category_importance)

# Plot category importance
p_category <- ggplot(category_importance, aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Importance by Category",
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", "rf_feature_importance_by_category_te_1hr.pdf"),
       plot = p_category, width = 10, height = 6)
```

# Top K-mer Analysis

```{r top_kmers, fig.width=10, fig.height=8}
# Analyze top k-mers
top_kmers <- importance_df %>%
  filter(str_detect(feature, "^utr5_4mer_")) %>%
  mutate(kmer = str_replace(feature, "^utr5_4mer_", "")) %>%
  head(20)

cat("\nTop 20 Most Predictive 4-mers in 5' UTR:\n")
print(top_kmers %>% dplyr::select(kmer, importance))

# Plot top k-mers
p_kmers <- top_kmers %>%
  ggplot(aes(x = reorder(kmer, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#CC79A7") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Top 20 Most Predictive 4-mers in 5' UTR",
    x = "4-mer Sequence",
    y = "Variable Importance"
  )

print(p_kmers)

ggsave(here("plots", "rf_top_kmers_te_1hr.pdf"),
       plot = p_kmers, width = 10, height = 8)
```

# Partial Dependence Plots for Top Features

```{r partial_dependence, fig.width=12, fig.height=8}
# Get top 6 features for partial dependence plots
top_features <- importance_df$feature[1:6]

# Function to calculate partial dependence
calc_partial_dep <- function(model, data, feature, n_points = 50) {
  feature_vals <- seq(min(data[[feature]], na.rm = TRUE),
                      max(data[[feature]], na.rm = TRUE),
                      length.out = n_points)

  pd_values <- sapply(feature_vals, function(val) {
    temp_data <- data
    temp_data[[feature]] <- val
    mean(predict(model, temp_data)$predictions)
  })

  data.frame(
    feature_value = feature_vals,
    predicted = pd_values,
    feature_name = feature
  )
}

# Calculate partial dependence for top features
pd_list <- lapply(top_features, function(f) {
  calc_partial_dep(rf_model, train_features, f)
})

pd_combined <- bind_rows(pd_list)

# Create partial dependence plots
p_pd <- ggplot(pd_combined, aes(x = feature_value, y = predicted)) +
  geom_line(color = "#0072B2", linewidth = 1) +
  facet_wrap(~feature_name, scales = "free_x", ncol = 3) +
  theme_cowplot() +
  labs(
    title = "Partial Dependence Plots: Top 6 Features",
    x = "Feature Value",
    y = "Predicted TE LFC"
  ) +
  theme(strip.text = element_text(size = 8))

print(p_pd)

ggsave(here("plots", "rf_partial_dependence_te_1hr.pdf"),
       plot = p_pd, width = 12, height = 8)
```

# Cross-Validation Performance

```{r cross_validation}
# Perform 5-fold cross-validation
set.seed(9)

# Create folds
folds <- createFolds(model_data$te_lfc_bin, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  test_idx <- folds[[i]]
  train_cv <- model_data[-test_idx, ] %>%
    dplyr::select(-gene_id, -transcript_id, -transcript_id_clean, -symbol, -te_lfc_bin)
  test_cv <- model_data[test_idx, ] %>%
    dplyr::select(-gene_id, -transcript_id, -transcript_id_clean, -symbol, -te_lfc_bin)

  # Train model
  rf_cv <- ranger(
    formula = te_lfc ~ .,
    data = train_cv,
    num.trees = 500,
    seed = 9
  )

  # Predict
  pred_cv <- predict(rf_cv, test_cv)$predictions

  # Calculate metrics
  data.frame(
    fold = i,
    RMSE = sqrt(mean((pred_cv - test_cv$te_lfc)^2)),
    R_squared = cor(pred_cv, test_cv$te_lfc)^2,
    Pearson_r = cor(pred_cv, test_cv$te_lfc)
  )
})

cv_results_df <- bind_rows(cv_results)

cat("\n========================================\n")
cat("5-Fold Cross-Validation Results\n")
cat("========================================\n\n")
print(cv_results_df)

cat("\nMean CV Performance:\n")
cat("RMSE:", round(mean(cv_results_df$RMSE), 4), "+/-", round(sd(cv_results_df$RMSE), 4), "\n")
cat("RÂ˛:", round(mean(cv_results_df$R_squared), 4), "+/-", round(sd(cv_results_df$R_squared), 4), "\n")
cat("Pearson r:", round(mean(cv_results_df$Pearson_r), 4), "+/-", round(sd(cv_results_df$Pearson_r), 4), "\n")
```

# Export Results

```{r export_results}
# Save model
saveRDS(rf_model, here("output", "predictive_modeling", "rf_model_te_1hr.rds"))
cat("Exported: rf_model_te_1hr.rds\n")

# Save importance
write_csv(importance_df, here("output", "predictive_modeling", "rf_importance_te_1hr.csv"))
cat("Exported: rf_importance_te_1hr.csv\n")

# Save metrics
metrics_export <- metrics %>%
  mutate(
    CV_RMSE_mean = mean(cv_results_df$RMSE),
    CV_RMSE_sd = sd(cv_results_df$RMSE),
    CV_R2_mean = mean(cv_results_df$R_squared),
    CV_R2_sd = sd(cv_results_df$R_squared),
    CV_Pearson_mean = mean(cv_results_df$Pearson_r),
    CV_Pearson_sd = sd(cv_results_df$Pearson_r),
    n_train = nrow(train_data),
    n_test = nrow(test_data),
    n_features = length(feature_cols_final)
  )

write_csv(metrics_export, here("output", "predictive_modeling", "rf_performance_te_1hr.csv"))
cat("Exported: rf_performance_te_1hr.csv\n")

# Save CV results
write_csv(cv_results_df, here("output", "predictive_modeling", "rf_cv_results_te_1hr.csv"))
cat("Exported: rf_cv_results_te_1hr.csv\n")

# Save predictions
predictions_export <- pred_df %>%
  mutate(
    gene_id = test_data$gene_id,
    transcript_id = test_data$transcript_id
  ) %>%
  dplyr::select(gene_id, transcript_id, symbol, actual, predicted, residual)

write_csv(predictions_export, here("output", "predictive_modeling", "rf_predictions_te_1hr.csv"))
cat("Exported: rf_predictions_te_1hr.csv\n")
```

# Summary

```{r summary}
cat("\n")
cat("========================================\n")
cat("RANDOM FOREST MODEL SUMMARY\n")
cat("========================================\n\n")

cat("Dataset:\n")
cat("  - Total genes:", nrow(model_data), "\n")
cat("  - Training set:", nrow(train_data), "\n")
cat("  - Test set:", nrow(test_data), "\n")
cat("  - Features:", length(feature_cols_final), "\n\n")

cat("Model Performance (Test Set):\n")
cat("  - RÂ˛:", round(metrics$R_squared, 4), "\n")
cat("  - Pearson r:", round(metrics$Pearson_r, 4), "\n")
cat("  - Spearman rho:", round(metrics$Spearman_rho, 4), "\n")
cat("  - RMSE:", round(metrics$RMSE, 4), "\n")
cat("  - MAE:", round(metrics$MAE, 4), "\n\n")

cat("Cross-Validation (5-fold):\n")
cat("  - Mean RÂ˛:", round(mean(cv_results_df$R_squared), 4), "\n")
cat("  - Mean Pearson r:", round(mean(cv_results_df$Pearson_r), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

cat("\nOutput Files:\n")
cat("  - rf_model_te_1hr.rds\n")
cat("  - rf_importance_te_1hr.csv\n")
cat("  - rf_performance_te_1hr.csv\n")
cat("  - rf_cv_results_te_1hr.csv\n")
cat("  - rf_predictions_te_1hr.csv\n")
cat("========================================\n")
```

# Session Info

```{r session_info}
sessionInfo()
```


Kate notes:
need to add more features, that can predict hte model
the model could not predict anything
could switch to predicting categorical changes like te up, down, or unchanged