---
title: "Negative Control Gene Set for Random Forest — eIF3e"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  neg_lfc: 0.5
  neg_padj: 0.3
  out_genesets_csv: ""
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis identifies genes that are **not regulated** by eIF3e knockdown in any condition (normoxia or hypoxia). These genes serve as a negative control set for Random Forest classification when the positive set is defined by eIF3e activity.

Mirrors nb02 (negative_control_set.Rmd) exactly, using si3e comparisons instead of si3d.
Normoxia is represented by the single time-regressed alltimepoints analysis; hypoxia uses the three individual timepoint comparisons (1hr, 4hr, 24hr).

## Criteria

A gene must satisfy **all** of the following across all 4 eIF3e comparisons (1 normoxia alltimepoints + 3 hypoxia timepoints):

- `|te_lfc| < neg_lfc` (small effect size), default 0.5
- `te_padj > neg_padj` (non-significant), default 0.3

Both are parameters. Defaults reproduce the criteria this notebook has always used.

**Why you would relax `neg_padj`.** Notebook 07 keeps every positive gene and downsamples
only the negative pool, to `min(n_pos, n_neg)`. For eIF3d the negative pool has always been
the larger of the two, so that balanced silently. The eIF3e pool is much smaller and can
invert it: against `hypoxia_3e_promotes_TE_1hr_lfc0.5.csv` (986 positives present in the
feature matrix) the pool sizes that survive into the matrix are

| `neg_padj` | pool | in feature matrix | ratio to 986 positives |
|---|---|---|---|
| > 0.3  | 403  | 260 | 0.26 : 1 |
| > 0.2  | 719  | 502 | 0.51 : 1 |
| > 0.1  | 1111 | 814 | 0.83 : 1 |
| > 0.05 | 1281 | 963 | 0.98 : 1 |

Relax `neg_padj`, never `neg_lfc`. `neg_padj` is the binding constraint at the default
setting (raising `neg_lfc` from 0.5 to 1.0 adds nothing at `neg_padj > 0.3`), and the
`|te_lfc| < 0.5` bound is what makes these genes *unaffected* rather than merely
*underpowered*. Loosening it to 1.0 would admit genes with real two-fold TE changes into a
set whose whole purpose is to contain none.

**Note on the shipped file.** `output/predictive_modeling/negative_control_genes_si3e.csv`
holds 719 genes, which is the `neg_padj > 0.2` result, not the `> 0.3` these defaults
produce (403). The artifact predates a change to the threshold in this notebook. Re-knitting
with defaults will therefore shrink that file to 403 genes and change every existing eIF3e
run that relies on it, so the write is now guarded: the legacy path is only overwritten when
the parameters match the values it was built with.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    neg_lfc          = 0.5,   # |te_lfc| must be below this in all 4 comparisons
    neg_padj         = 0.3,   # te_padj must be above this in all 4 comparisons
    out_genesets_csv = ""     # "" = write only the legacy predictive_modeling/ files.
                              # A filename additionally writes an nb07-ready set to
                              # output/genesets/, e.g.
                              # "si3e_negative_controls_padj0.05.csv"
  )
}
NEG_LFC  <- params$neg_lfc
NEG_PADJ <- params$neg_padj
OUT_GENESETS_CSV <- if (is.null(params$out_genesets_csv)) "" else params$out_genesets_csv

# The shipped negative_control_genes_si3e.csv was built at these values; anything else
# must not overwrite it, or every existing eIF3e run silently changes its negative class.
LEGACY_LFC <- 0.5; LEGACY_PADJ <- 0.2
write_legacy <- isTRUE(all.equal(NEG_LFC, LEGACY_LFC)) &&
                isTRUE(all.equal(NEG_PADJ, LEGACY_PADJ))

stopifnot("neg_lfc must be positive" = NEG_LFC > 0,
          "neg_padj must be in (0, 1)" = NEG_PADJ > 0 && NEG_PADJ < 1)
cat("Criteria: |te_lfc| <", NEG_LFC, "and te_padj >", NEG_PADJ, "in all 4 comparisons\n")
cat("Overwrite legacy predictive_modeling/ files:", write_legacy, "\n")
```

# Define File List

```{r define_files}
# Normoxia: single time-regressed alltimepoints analysis
normoxia_comparison <- tibble(
  protein = "si3e",
  condition = "normoxia",
  timepoint = "alltimepoints",
  file = here("output", "translation_categories_si3e_vs_sictrl_normoxia_alltimepoints.csv"),
  label = "si3e_normoxia_alltimepoints"
)

# Hypoxia: individual timepoint comparisons
hypoxia_comparisons <- expand.grid(
  protein = "si3e",
  condition = "hypoxia",
  timepoint = c("1hr", "4hr", "24hr"),
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  mutate(
    file = here("output",
                paste0("translation_categories_", protein, "_vs_sictrl_",
                       condition, "_", timepoint, ".csv")),
    label = paste(protein, condition, timepoint, sep = "_")
  )

comparisons <- bind_rows(normoxia_comparison, hypoxia_comparisons)

cat("Defined", nrow(comparisons), "comparisons:\n")
print(comparisons %>% dplyr::select(label, file))
```

# Load All 4 Datasets

```{r load_data}
all_data <- comparisons %>%
  rowwise() %>%
  mutate(data = list(read_csv(file, show_col_types = FALSE))) %>%
  ungroup()

# Verify all files loaded
cat("\nRows per comparison:\n")
all_data %>%
  mutate(n_genes = map_int(data, nrow)) %>%
  dplyr::select(label, n_genes) %>%
  print(n = 4)
```

# Filter Each Dataset

```{r filter_datasets}
# For each comparison, flag genes passing criteria: |te_lfc| < 0.5 AND te_padj > 0.3
passing_genes <- all_data %>%
  mutate(
    passing = map(data, function(df) {
      df %>%
        filter(abs(te_lfc) < NEG_LFC, te_padj > NEG_PADJ) %>%
        pull(gene_id)
    })
  )

# Report how many genes pass each individual filter
cat("\nGenes passing criteria per comparison:\n")
passing_genes %>%
  mutate(n_passing = map_int(passing, length)) %>%
  dplyr::select(label, n_passing) %>%
  print(n = 4)
```

# Intersect Across All 4 Comparisons

```{r intersect}
# Find genes present in ALL 4 datasets
all_gene_sets <- map(all_data$data, ~ pull(.x, gene_id))
genes_in_all <- Reduce(intersect, all_gene_sets)
cat("Genes measured in all 4 comparisons:", length(genes_in_all), "\n")

# Find genes passing criteria in ALL 4 comparisons
passing_sets <- passing_genes$passing
negative_control_ids <- Reduce(intersect, passing_sets)

# Also require presence in all 4 datasets
negative_control_ids <- intersect(negative_control_ids, genes_in_all)
cat("Genes passing criteria in all 4 comparisons:", length(negative_control_ids), "\n")
```

# Build Output Table

```{r build_output}
# Get symbol mapping from one of the datasets
symbol_map <- all_data$data[[1]] %>%
  dplyr::select(gene_id, symbol) %>%
  distinct()

# Build output with gene_id, symbol, and count of comparisons measured
negative_control_genes <- tibble(gene_id = negative_control_ids) %>%
  left_join(symbol_map, by = "gene_id") %>%
  mutate(n_comparisons = 4L) %>%
  arrange(gene_id)

cat("\nNegative control set:", nrow(negative_control_genes), "genes\n")
head(negative_control_genes, 20)
```

# QC & Summary

```{r qc_summary}
# Verify: no gene in the negative set violates |te_lfc| < NEG_LFC or te_padj > NEG_PADJ
cat("=== Verification ===\n\n")

verification <- all_data %>%
  mutate(
    violations = map(data, function(df) {
      df %>%
        filter(gene_id %in% negative_control_ids) %>%
        filter(abs(te_lfc) >= NEG_LFC | te_padj <= NEG_PADJ)
    })
  )

n_violations <- sum(map_int(verification$violations, nrow))
cat("Total violations across all comparisons:", n_violations, "\n")

if (n_violations > 0) {
  warning("Found violations! Inspect the verification object.")
} else {
  cat("PASS: No gene in the negative set violates the criteria in any condition.\n")
}
```

```{r qc_distributions, fig.width=10, fig.height=6}
# Collect TE LFC values for negative control genes across all comparisons
neg_ctrl_te <- all_data %>%
  mutate(
    neg_te = map2(data, label, function(df, lbl) {
      df %>%
        filter(gene_id %in% negative_control_ids) %>%
        dplyr::select(gene_id, te_lfc, te_padj) %>%
        mutate(comparison = lbl)
    })
  ) %>%
  pull(neg_te) %>%
  bind_rows()

cat("\n=== TE LFC distribution in negative control set ===\n")
cat("Overall:\n")
print(summary(neg_ctrl_te$te_lfc))

cat("\nPer comparison:\n")
neg_ctrl_te %>%
  group_by(comparison) %>%
  summarise(
    mean_te_lfc   = mean(te_lfc),
    median_te_lfc = median(te_lfc),
    sd_te_lfc     = sd(te_lfc),
    min_te_lfc    = min(te_lfc),
    max_te_lfc    = max(te_lfc),
    .groups = "drop"
  ) %>%
  print(n = 4)

# Plot TE LFC distributions
ggplot(neg_ctrl_te, aes(x = te_lfc)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~comparison, ncol = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "grey40") +
  theme_minimal() +
  labs(title = "TE LFC Distribution of Negative Control Genes (eIF3e)",
       subtitle = paste(nrow(negative_control_genes), "genes across 4 eIF3e comparisons"),
       x = "TE Log2 Fold Change", y = "Count")
```

```{r qc_padj, fig.width=10, fig.height=6}
# Plot te_padj distributions
ggplot(neg_ctrl_te, aes(x = te_padj)) +
  geom_histogram(bins = 50, fill = "coral", alpha = 0.7) +
  facet_wrap(~comparison, ncol = 3) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "TE padj Distribution of Negative Control Genes (eIF3e)",
       x = "TE padj", y = "Count")
```

```{r qc_density_sanity, fig.width=10, fig.height=6}
# Sanity check: where do negative control genes fall in the full TE LFC distribution?
# Build a data frame with all genes from all 4 comparisons, labeled by neg control membership

all_genes_te <- all_data %>%
  mutate(
    all_te = map2(data, label, function(df, lbl) {
      df %>%
        dplyr::select(gene_id, te_lfc) %>%
        mutate(
          comparison  = lbl,
          is_neg_ctrl = gene_id %in% negative_control_ids
        )
    })
  ) %>%
  pull(all_te) %>%
  bind_rows()

n_neg  <- sum(negative_control_ids %in% all_genes_te$gene_id[all_genes_te$comparison == all_genes_te$comparison[[1]]])
n_ctrl <- length(negative_control_ids)

ggplot(all_genes_te, aes(x = te_lfc, color = is_neg_ctrl, fill = is_neg_ctrl)) +
  geom_density(alpha = 0.3, linewidth = 0.7) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dotted", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
  scale_color_manual(values = c("FALSE" = "grey50", "TRUE" = "#d6604d"),
                     labels = c("FALSE" = "All other genes", "TRUE" = paste0("Negative control (n=", n_ctrl, ")"))) +
  scale_fill_manual(values  = c("FALSE" = "grey50", "TRUE" = "#d6604d"),
                    labels  = c("FALSE" = "All other genes", "TRUE" = paste0("Negative control (n=", n_ctrl, ")"))) +
  facet_wrap(~comparison, ncol = 2) +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(
    title    = "Negative control genes in full TE LFC distribution (eIF3e)",
    subtitle = "Negative controls should cluster tightly around 0 in all comparisons",
    x        = "TE Log2 Fold Change",
    y        = "Density",
    color    = NULL, fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Export

```{r export}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

# Legacy path, consumed by notebook 07 whenever gene = "3e" and neg_geneset_csv is empty.
# Only rewritten when the parameters match the values this file was built with, so a
# threshold sweep cannot silently change the negative class of every existing eIF3e run.
if (write_legacy) {
  write_csv(negative_control_genes,
            here("output", "predictive_modeling", "negative_control_genes_si3e.csv"))
  cat("Exported: output/predictive_modeling/negative_control_genes_si3e.csv\n")

  saveRDS(negative_control_genes,
          here("output", "predictive_modeling", "negative_control_genes_si3e.rds"))
  cat("Exported: output/predictive_modeling/negative_control_genes_si3e.rds\n")
} else {
  cat("Skipped the legacy predictive_modeling/ files: parameters differ from the\n",
      "  values they were built with (|te_lfc| <", LEGACY_LFC, ", te_padj >", LEGACY_PADJ,
      ").\n  Pass out_genesets_csv to write this set somewhere notebook 07 can read it.\n")
}

# nb07-ready export. Its neg_geneset_csv param reads from output/genesets/ and keys on an
# `ensembl_gene` column; readout / neg_* columns are the provenance fields nb07 prints in
# its classification summary and checks for a TE-vs-RNA mismatch.
if (nzchar(OUT_GENESETS_CSV)) {
  dir.create(here("output", "genesets"), showWarnings = FALSE, recursive = TRUE)
  geneset_out <- negative_control_genes %>%
    transmute(
      ensembl_gene   = sub("\\..*", "", gene_id),
      db_gene_symbol = symbol,
      gs_name        = tools::file_path_sans_ext(OUT_GENESETS_CSV),
      readout        = "TE",
      neg_contrasts  = paste(comparisons$label, collapse = "; "),
      neg_lfc        = NEG_LFC,
      neg_padj       = NEG_PADJ)
  stopifnot("ensembl_gene must be unique" = !any(duplicated(geneset_out$ensembl_gene)),
            "ensembl_gene still versioned" = !any(grepl("\\.", geneset_out$ensembl_gene)))
  write_csv(geneset_out, here("output", "genesets", OUT_GENESETS_CSV))
  cat("Exported: output/genesets/", OUT_GENESETS_CSV, " (", nrow(geneset_out), " genes)\n", sep = "")
}

cat("\nFinal negative control set:", nrow(negative_control_genes), "genes\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
