---
title: "Expressed background pool for eIF3e target models"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  pos_geneset_csv: "hypoxia_3e_promotes_TE_1hr_lfc0.5.csv"
  de_csv: "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"
  out_csv: "si3e_expressed_background_pool.csv"
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Purpose

Builds the **expressed background pool** used as the negative-class pool for the eIF3e
target models, replacing the screened `si3e_negative_controls_padj0.05.csv` set.

- **Old pool:** genes screened as eIF3e-*protected* (`|te_lfc| < 0.5` & `padj > 0.05` in all
  four si3e contrasts). 1,281 genes, markedly GC-rich relative to the transcriptome.
- **New pool:** every gene DESeq2 could **test** in the si3e hypoxia 1hr contrast, minus
  every gene in the positive set's parent list. No screening on effect size at all.

The claim the model makes changes accordingly: from "eIF3e targets vs eIF3e-protected
mRNAs" to **"eIF3e targets vs a typical expressed mRNA"**. Expect a lower AUC than the
screened-pool runs; those runs are not comparable to these and should not be plotted on
the same axis.

## What "expressed" means here

Two candidate definitions were checked and they agree to within 3 genes:

| Definition | n |
|---|---|
| Present in `feature_matrix_dhx29_kd_feature.rds` (nb01 selects the most abundant isoform from siCTRL normoxia RNA-seq, so matrix membership is already an expression filter) | 10,172 |
| Non-NA `te_padj` in the si3e hypoxia 1hr contrast (DESeq2 independent filtering drops low-count genes) | 11,861 |
| Both | 10,169 |

This notebook writes the **definition** - DESeq2-tested minus 3e-promoted - and leaves the
matrix intersection to each consumer, which reports its own n. Writing a pre-intersected
file would silently bind the pool to one feature matrix, and the model (`dhx29_kd_feature`,
10,172 genes) and the GC figure (`external_stability` joined to `positional_gc`, 9,608)
do not use the same one.

## Which genes are excluded

Excluded: every gene in `hypoxia_3e_promotes_TE_1hr_lfc0.5.csv` (1,083), not merely the
399 eIF3e-*specific* ones. Genes promoted by both eIF3d and eIF3e are known targets and
would be mislabelled as background.

**Not excluded: eIF3e-inhibited genes.** They move in the opposite direction, but removing
them would make the pool "everything eIF3e does not touch", which is the screened set this
pool exists to replace. A background that excludes real biology is not a background.

# Load Libraries

```{r libraries}
library(tidyverse)
library(here)
```

```{r params_default}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    pos_geneset_csv = "hypoxia_3e_promotes_TE_1hr_lfc0.5.csv",
    de_csv          = "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv",
    out_csv         = "si3e_expressed_background_pool.csv"
  )
}
```

# Load the DESeq2 contrast

```{r load_de}
de_path <- here("output", params$de_csv)
stopifnot("DESeq2 results file not found" = file.exists(de_path))

de <- read_csv(de_path, show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("Rows in", params$de_csv, ":", nrow(de), "\n")
stopifnot("te_lfc has non-finite values" = all(is.finite(de$te_lfc[!is.na(de$te_lfc)])))

# Directionality guard (CLAUDE.md): these files are log2(sictrl / si3e) despite the
# filename saying si3e_vs_sictrl, so the knockdown's own target must come out strongly
# POSITIVE. If it is negative the contrast has flipped and the positive set - hence the
# pool's exclusion list - means the opposite of what it says.
eif3e_rna <- de$rna_lfc[de$symbol == "EIF3E"]
cat("EIF3E rna_lfc:", round(eif3e_rna, 2), "(expected ~ +4.9, must be positive)\n")
stopifnot("EIF3E rna_lfc is not strongly positive - contrast direction has flipped" =
  length(eif3e_rna) == 1 && eif3e_rna > 2)
```

# Define the pool

```{r build_pool}
# "Expressed" = DESeq2 could test the gene in this contrast. Independent filtering sets
# padj to NA for genes with too little count support, which is exactly the set we do not
# want drawing negatives from.
expressed <- de %>% filter(!is.na(te_padj))
cat("Expressed (non-NA te_padj):", nrow(expressed), "of", nrow(de), "\n")

pos_path <- here("output", "genesets", params$pos_geneset_csv)
stopifnot("positive gene set not found" = file.exists(pos_path))
pos_genes <- unique(read_csv(pos_path, show_col_types = FALSE)$ensembl_gene)
cat("Positive set (excluded from the pool):", length(pos_genes), "genes\n")

pool <- expressed %>%
  filter(!gene_id_clean %in% pos_genes) %>%
  distinct(gene_id_clean, .keep_all = TRUE)

cat("Background pool:", nrow(pool), "genes\n")

stopifnot("pool overlaps the positive set" =
  length(intersect(pool$gene_id_clean, pos_genes)) == 0)
stopifnot("pool lost more genes than the positive set contains" =
  nrow(expressed) - nrow(pool) <= length(pos_genes))
stopifnot("pool is implausibly small" = nrow(pool) > 5000)
```

The pool is a background, so it must look like the transcriptome rather than like a
screened set. The old screened pool sat well above the transcriptome median on GC3; this
one should not.

```{r pool_distribution}
# Biological spot-check: an unscreened background must be centred on the transcriptome,
# not displaced from it. A drifted median here means the filter is selecting on something.
cat("median te_lfc  - pool:", round(median(pool$te_lfc, na.rm = TRUE), 3),
    " all tested:", round(median(expressed$te_lfc, na.rm = TRUE), 3), "\n")
stopifnot("pool median te_lfc is displaced from the transcriptome median" =
  abs(median(pool$te_lfc, na.rm = TRUE) - median(expressed$te_lfc, na.rm = TRUE)) < 0.1)

# The pool deliberately RETAINS eIF3e-inhibited genes; report how many so the composition
# is on the record rather than assumed.
n_inhibited <- sum(pool$te_lfc < -0.5 & pool$te_padj < 0.05, na.rm = TRUE)
cat("Genes in the pool that eIF3e inhibits (te_lfc < -0.5 & padj < 0.05):",
    n_inhibited, sprintf("(%.1f%%)", 100 * n_inhibited / nrow(pool)), "\n")
```

# Report the intersection with each consumer matrix

```{r matrix_intersections}
# Each consumer intersects the pool with its own matrix. Report both here so a shortfall
# is visible at build time rather than inside a model run.
report_matrix <- function(rds, label) {
  path <- here("output", "predictive_modeling", rds)
  if (!file.exists(path)) {
    cat(label, ": matrix not found, skipped\n"); return(invisible(NULL))
  }
  g <- unique(readRDS(path)$gene_id_clean)
  n <- length(intersect(pool$gene_id_clean, g))
  cat(sprintf("%-28s matrix genes: %6d   pool genes present: %6d\n", label, length(g), n))
  invisible(n)
}
n_model <- report_matrix("feature_matrix_dhx29_kd_feature.rds", "dhx29_kd_feature (models)")
report_matrix("feature_matrix_external_stability.rds", "external_stability")
report_matrix("feature_matrix_positional_gc.rds", "positional_gc (GC figure)")

# The draw is only non-degenerate if the pool exceeds the positive set. The eIF3e-specific
# set is 399 genes in the model matrix.
stopifnot("pool does not exceed the positive set in the model matrix" =
  is.null(n_model) || n_model > 1000)
```

# Write

```{r write}
out <- pool %>%
  transmute(
    ensembl_gene   = gene_id_clean,
    db_gene_symbol = symbol,
    gs_name        = sub("\\.csv$", "", params$out_csv),
    # nb07 stops unless the negative set's readout matches the run's. The pool is drawn
    # from the TE-tested universe, so "TE" is true rather than merely permissive - a file
    # with no readout column would default to "TE" and pass the guard on a missing value.
    readout        = "TE",
    # nb07 prints neg_contrasts/neg_lfc/neg_padj when present. This pool has no per-contrast
    # thresholds - it is defined on expression - so it stamps neg_criteria instead, and nb07
    # reports that rather than claiming the file predates provenance stamping.
    neg_criteria   = paste0("expressed background: non-NA te_padj in ", params$de_csv,
                            ", excluding all ", length(pos_genes), " genes in ",
                            params$pos_geneset_csv, "; NOT screened on effect size"),
    built_by       = "notebook 02c"
  )

stopifnot("output has duplicate genes" = !any(duplicated(out$ensembl_gene)))
stopifnot("output is missing required columns" =
  all(c("ensembl_gene", "db_gene_symbol", "readout") %in% names(out)))

out_path <- here("output", "genesets", params$out_csv)
write_csv(out, out_path)
cat("Wrote", nrow(out), "genes to", out_path, "\n")
```

# Summary

```{r summary}
cat("Pool:              ", nrow(out), "genes\n")
cat("Definition:        expressed (DESeq2-tested) minus all 3e-promoted\n")
cat("Screened on effect size: NO - this is a background, not a control set\n")
cat("Use with notebook 07 as:\n")
cat('  neg_geneset_csv = "', params$out_csv, '"\n', sep = "")
cat('  neg_tag = "exprbg"   # REQUIRED: neg_geneset_csv is not part of the output\n')
cat('                       # suffix, so without a tag this overwrites the runs that\n')
cat('                       # used the screened pool.\n')
```
