---
title: "Random Forest Classification: eIF3d Target Prediction in Hypoxia"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis builds a binary classification model to predict whether a gene is an eIF3d target (promotes TE) vs. non-target in 1 hour hypoxia, based on mRNA sequence features.

**Target Definition:**

- **Positive class ("promotes")**: Genes from `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` (TE padj < 0.05, LFC > 0.5 in siCtrl vs si3d)
- **Negative class ("does_not_promote")**: All other genes

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ranger)
library(caret)
library(pROC)
library(scales)
```

# Load Data

```{r load_data}
# Load feature matrix
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr.rds"))
cat("Loaded feature matrix:", nrow(features), "genes x", ncol(features), "features\n")

# Load target gene list (eIF3d promotes TE)
promotes_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
cat("Loaded", nrow(promotes_genes), "eIF3d target genes (promotes TE)\n")

# Check column names in target file
cat("\nTarget file columns:", paste(colnames(promotes_genes), collapse = ", "), "\n")
```

# Create Binary Target Variable

```{r create_target}
# Create binary target: promotes vs does_not_promote
features <- features %>%
  mutate(eif3d_target = factor(
    ifelse(gene_id_clean %in% promotes_genes$ensembl_gene, "promotes", "does_not_promote"),
    levels = c("does_not_promote", "promotes")
  ))

# Check class distribution
cat("\nClass distribution:\n")
print(table(features$eif3d_target))
cat("\nClass proportions:\n")
print(prop.table(table(features$eif3d_target)))

# Calculate class imbalance ratio
n_promotes <- sum(features$eif3d_target == "promotes")
n_does_not <- sum(features$eif3d_target == "does_not_promote")
cat("\nClass imbalance ratio (negative:positive):", round(n_does_not / n_promotes, 1), ":1\n")
```

# Preprocessing

## Remove Near-Zero Variance Features

```{r remove_nzv}
# Identify feature columns (exclude identifiers and target)
id_cols <- c("gene_id", "gene_id.x", "gene_id.y", "gene_id_clean",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")

# Also exclude any column containing identifier patterns
all_cols <- colnames(features)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))

feature_cols <- setdiff(colnames(features), id_cols)

cat("Columns excluded as identifiers:\n")
print(intersect(colnames(features), id_cols))
cat("\nInitial feature count:", length(feature_cols), "\n")

# Create feature-only dataframe for preprocessing
feature_data <- features %>%
  dplyr::select(all_of(feature_cols))

# Check for near-zero variance features
nzv <- nearZeroVar(feature_data, saveMetrics = TRUE)
nzv_features <- rownames(nzv)[nzv$nzv]

cat("Near-zero variance features to remove:", length(nzv_features), "\n")

# Remove NZV features
feature_cols_clean <- setdiff(feature_cols, nzv_features)
cat("Features after NZV removal:", length(feature_cols_clean), "\n")
```

## Handle Missing Values

```{r handle_missing}
# Check missing values in remaining features
missing_counts <- features %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:\n")
print(missing_counts)

# Impute missing values with median
features_imputed <- features %>%
  mutate(across(all_of(feature_cols_clean), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Verify no missing values in features
remaining_na <- features_imputed %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  summarise(total_na = sum(is.na(.))) %>%
  pull(total_na)

cat("\nRemaining missing values after imputation:", remaining_na, "\n")
```

## Remove Highly Correlated Features

```{r remove_correlated}
# Calculate correlation matrix (only for numeric features)
feature_matrix_for_cor <- features_imputed %>%
  dplyr::select(all_of(feature_cols_clean)) %>%
  dplyr::select(where(is.numeric))

# Find highly correlated pairs (r > 0.9)
cor_matrix <- cor(feature_matrix_for_cor, use = "pairwise.complete.obs")

# Find features to remove based on correlation
high_cor <- findCorrelation(cor_matrix, cutoff = 0.9, names = TRUE)

cat("Highly correlated features to remove:", length(high_cor), "\n")
if (length(high_cor) > 0) {
  cat("Examples:", paste(head(high_cor, 10), collapse = ", "), "\n")
}

# Remove highly correlated features
feature_cols_final <- setdiff(feature_cols_clean, high_cor)
cat("Final feature count:", length(feature_cols_final), "\n")
```

# Prepare Data for Modeling

```{r prepare_data}
# Create final dataset for modeling
model_data <- features_imputed %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols_final)) %>%
  filter(!is.na(eif3d_target))

cat("Final dataset dimensions:", nrow(model_data), "genes x", length(feature_cols_final), "features\n")

# Summary of target variable
cat("\nTarget variable distribution:\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

# Stratified split (80/20)
train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "genes\n")
cat("Test set:", nrow(test_data), "genes\n")

# Verify stratification
cat("\nTrain set class distribution:\n")
print(table(train_data$eif3d_target))
cat("\nTrain set class proportions:\n")
print(prop.table(table(train_data$eif3d_target)))

cat("\nTest set class distribution:\n")
print(table(test_data$eif3d_target))
cat("\nTest set class proportions:\n")
print(prop.table(table(test_data$eif3d_target)))
```

# Calculate Class Weights

```{r class_weights}
# Calculate class weights (inverse frequency weighting)
# This gives higher weight to the minority "promotes" class
class_counts <- table(train_data$eif3d_target)
n_total <- nrow(train_data)

# Weight minority class proportionally higher
weight_promotes <- n_total / (2 * class_counts["promotes"])
weight_does_not_promote <- n_total / (2 * class_counts["does_not_promote"])

# Create case weight vector
case_weights <- ifelse(
  train_data$eif3d_target == "promotes",
  weight_promotes,
  weight_does_not_promote
)

cat("\nClass weights:\n")
cat("  promotes:", round(weight_promotes, 2), "\n")
cat("  does_not_promote:", round(weight_does_not_promote, 2), "\n")
cat("  Weight ratio:", round(weight_promotes / weight_does_not_promote, 1), ":1\n")

cat("\nTraining set size:", nrow(train_data), "(using all samples with class weights)\n")
```

# Train Random Forest Classifier

```{r train_model}
# Prepare full training data (no downsampling)
train_for_model <- train_data %>%
  dplyr::select(eif3d_target, all_of(feature_cols_final))

cat("Training random forest classifier...\n")
cat("Features:", length(feature_cols_final), "\n")
cat("Training samples:", nrow(train_for_model), "\n")

# Train random forest with probability estimation
set.seed(9)
rf_classifier <- ranger(
  formula = eif3d_target ~ .,
  data = train_for_model,
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  case.weights = case_weights,  # Use class weights instead of downsampling
  seed = 9,
  verbose = TRUE
)

cat("\nModel training complete!\n")
print(rf_classifier)
```

# Model Evaluation

## Predictions on Test Set

```{r predictions}
# Prepare test features
test_features <- test_data %>%
  dplyr::select(all_of(feature_cols_final))

# Get probability predictions
pred_probs <- predict(rf_classifier, test_features)$predictions
pred_promotes_prob <- pred_probs[, "promotes"]

# Actual labels
actual <- test_data$eif3d_target

cat("Probability predictions obtained for", length(pred_promotes_prob), "test samples\n")
cat("Actual promotes:", sum(actual == "promotes"), "\n")
cat("Probability range:", round(min(pred_promotes_prob), 4), "-", round(max(pred_promotes_prob), 4), "\n")
```

## ROC Curve and Threshold Optimization

```{r roc_and_threshold, fig.width=8, fig.height=6}
# Calculate ROC curve
roc_obj <- roc(
  response = actual,
  predictor = pred_promotes_prob,
  levels = c("does_not_promote", "promotes"),
  direction = "<"
)

auc_value <- auc(roc_obj)
cat("AUC-ROC:", round(auc_value, 4), "\n")

# Find Youden's J optimal threshold (maximizes sensitivity + specificity - 1)
coords_youden <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")

# Find threshold for 90% sensitivity
target_sensitivity <- 0.90
coords_high_sens <- coords(roc_obj, x = target_sensitivity, input = "sensitivity",
                           ret = c("threshold", "specificity"))

# Use Youden threshold as the optimal threshold
optimal_threshold <- coords_youden$threshold

cat("\n========================================\n")
cat("Threshold Analysis\n")
cat("========================================\n")
cat("Youden optimal threshold:", round(optimal_threshold, 4), "\n")
cat("  -> Expected Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  -> Expected Specificity:", round(coords_youden$specificity, 3), "\n")
cat("\nHigh-sensitivity threshold (90% sens):", round(coords_high_sens$threshold, 4), "\n")
cat("  -> Expected Specificity:", round(coords_high_sens$specificity, 3), "\n")

# Plot ROC curve with optimal threshold marked
roc_df <- data.frame(
  sensitivity = roc_obj$sensitivities,
  specificity = roc_obj$specificities
)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = "ROC Curve: eIF3d Target Classification (Hypoxia)",
    subtitle = paste0("AUC = ", round(auc_value, 3), " | Optimal threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

print(p_roc)

ggsave(here("plots", "rf_classification_roc_eif3d_promotes_hypoxia_1hr.pdf"),
       plot = p_roc, width = 8, height = 6)
```

## Apply Optimal Threshold and Evaluate

```{r confusion_matrix}
# Get class predictions using Youden optimal threshold
pred_class <- factor(
  ifelse(pred_promotes_prob >= optimal_threshold, "promotes", "does_not_promote"),
  levels = c("does_not_promote", "promotes")
)

cat("Using Youden optimal threshold:", round(optimal_threshold, 4), "\n")
cat("Predicted promotes:", sum(pred_class == "promotes"), "\n")
cat("Actual promotes:", sum(actual == "promotes"), "\n")

# Create confusion matrix
cm <- confusionMatrix(pred_class, actual, positive = "promotes")
print(cm)

# Extract key metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]  # Recall
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\n========================================\n")
cat("Key Classification Metrics (Youden Threshold)\n")
cat("========================================\n")
cat("Threshold:", round(optimal_threshold, 4), "\n")
cat("Accuracy:", round(accuracy, 4), "\n")
cat("Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("Sensitivity (Recall):", round(sensitivity, 4), "\n")
cat("Specificity:", round(specificity, 4), "\n")
cat("Precision:", round(precision, 4), "\n")
cat("F1 Score:", round(f1, 4), "\n")
```

## Alternative: High-Sensitivity Threshold

```{r high_sensitivity_threshold}
# Get predictions at high-sensitivity threshold (90% sensitivity)
pred_class_high_sens <- factor(
  ifelse(pred_promotes_prob >= coords_high_sens$threshold, "promotes", "does_not_promote"),
  levels = c("does_not_promote", "promotes")
)

cm_high_sens <- confusionMatrix(pred_class_high_sens, actual, positive = "promotes")

cat("\n========================================\n")
cat("Metrics at High-Sensitivity Threshold (90% target)\n")
cat("========================================\n")
cat("Threshold:", round(coords_high_sens$threshold, 4), "\n")
cat("Sensitivity:", round(cm_high_sens$byClass["Sensitivity"], 4), "\n")
cat("Specificity:", round(cm_high_sens$byClass["Specificity"], 4), "\n")
cat("Precision:", round(cm_high_sens$byClass["Precision"], 4), "\n")
cat("F1 Score:", round(cm_high_sens$byClass["F1"], 4), "\n")
cat("\nConfusion matrix:\n")
print(cm_high_sens$table)
```

## Confusion Matrix Visualization

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
# Create confusion matrix data for plotting
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = "Confusion Matrix (Hypoxia)",
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3), ", F1 = ", round(f1, 3)),
    x = "Actual Class",
    y = "Predicted Class"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold")
  )

print(p_cm)

ggsave(here("plots", "rf_classification_confusion_matrix_eif3d_promotes_hypoxia_1hr.pdf"),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)

# Create stratified folds
folds <- createFolds(model_data$eif3d_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Processing fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- model_data[-test_idx, ]
  test_cv <- model_data[test_idx, ]

  # Calculate class weights for this fold
  fold_counts <- table(train_cv$eif3d_target)
  n_fold <- nrow(train_cv)
  fold_weight_promotes <- n_fold / (2 * fold_counts["promotes"])
  fold_weight_does_not <- n_fold / (2 * fold_counts["does_not_promote"])
  fold_case_weights <- ifelse(train_cv$eif3d_target == "promotes",
                              fold_weight_promotes, fold_weight_does_not)

  # Prepare training data
  train_cv_for_model <- train_cv %>%
    dplyr::select(eif3d_target, all_of(feature_cols_final))

  # Train with class weights
  rf_cv <- ranger(
    formula = eif3d_target ~ .,
    data = train_cv_for_model,
    num.trees = 500,
    probability = TRUE,
    case.weights = fold_case_weights,
    seed = 9
  )

  # Prepare test data
  test_cv_features <- test_cv %>% dplyr::select(all_of(feature_cols_final))

  # Get probability predictions
  pred_probs_cv <- predict(rf_cv, test_cv_features)$predictions[, "promotes"]

  # Calculate ROC and find Youden optimal threshold for this fold
  roc_cv <- roc(test_cv$eif3d_target, pred_probs_cv,
                levels = c("does_not_promote", "promotes"), direction = "<", quiet = TRUE)
  coords_youden_cv <- coords(roc_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                              best.method = "youden")
  fold_threshold <- coords_youden_cv$threshold

  # Apply Youden threshold for predictions
  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= fold_threshold, "promotes", "does_not_promote"),
    levels = c("does_not_promote", "promotes")
  )

  # Calculate metrics
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif3d_target, positive = "promotes")

  data.frame(
    fold = i,
    Threshold = fold_threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n========================================\n")
cat("5-Fold Stratified Cross-Validation Results\n")
cat("(Using Youden optimal threshold per fold)\n")
cat("========================================\n\n")
print(cv_results_df)

# Summary statistics
cat("\nMean CV Performance (+/- SD):\n")
cat("Threshold:", round(mean(cv_results_df$Threshold), 4), "+/-", round(sd(cv_results_df$Threshold), 4), "\n")
cat("AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-", round(sd(cv_results_df$Balanced_Accuracy), 4), "\n")
cat("Sensitivity:", round(mean(cv_results_df$Sensitivity), 4), "+/-", round(sd(cv_results_df$Sensitivity), 4), "\n")
cat("Specificity:", round(mean(cv_results_df$Specificity), 4), "+/-", round(sd(cv_results_df$Specificity), 4), "\n")
cat("Precision:", round(mean(cv_results_df$Precision, na.rm = TRUE), 4), "+/-", round(sd(cv_results_df$Precision, na.rm = TRUE), 4), "\n")
cat("F1:", round(mean(cv_results_df$F1, na.rm = TRUE), 4), "+/-", round(sd(cv_results_df$F1, na.rm = TRUE), 4), "\n")
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
# Extract feature importance
importance_df <- data.frame(
  feature = names(rf_classifier$variable.importance),
  importance = rf_classifier$variable.importance
) %>%
  arrange(desc(importance))

# Show top 30 features
cat("\nTop 30 Most Important Features:\n")
print(importance_df %>% head(30))

# Create clean feature names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

# Plot top 30 features
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Top 30 Features for eIF3d Target Classification (Hypoxia)",
    subtitle = "Random Forest Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", "rf_classification_feature_importance_eif3d_promotes_hypoxia_1hr.pdf"),
       plot = p_importance, width = 10, height = 10)
```

# Feature Category Analysis

```{r feature_categories, fig.width=10, fig.height=6}
# Categorize features
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^utr5_4mer_") ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_") ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_") ~ "CDS 4-mers",
      str_detect(feature, "^rbp_") ~ "RBP Motifs",
      str_detect(feature, "length|Length") ~ "Region Lengths",
      str_detect(feature, "_gc$|gc3$|GC") ~ "GC Content",
      TRUE ~ "Other"
    )
  )

# Summarize importance by category
category_importance <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    max_importance = max(importance),
    top_feature = feature[which.max(importance)]
  ) %>%
  arrange(desc(total_importance))

cat("\nFeature Importance by Category:\n")
print(category_importance)

# Plot category importance
p_category <- ggplot(category_importance, aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Importance by Category (Hypoxia)",
    subtitle = "eIF3d Target Classification",
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", "rf_classification_importance_by_category_eif3d_promotes_hypoxia_1hr.pdf"),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
# Create output directory if needed
output_dir <- here("output", "predictive_modeling")

# Save model
saveRDS(rf_classifier, file.path(output_dir, "rf_classifier_eif3d_promotes_hypoxia_1hr.rds"))
cat("Exported: rf_classifier_eif3d_promotes_hypoxia_1hr.rds\n")

# Save importance
write_csv(importance_df, file.path(output_dir, "rf_classification_importance_eif3d_promotes_hypoxia_1hr.csv"))
cat("Exported: rf_classification_importance_eif3d_promotes_hypoxia_1hr.csv\n")

# Save performance metrics
performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy, sensitivity, specificity, precision, f1),
  CV_Mean = c(
    mean(cv_results_df$AUC),
    NA,
    mean(cv_results_df$Balanced_Accuracy),
    mean(cv_results_df$Sensitivity),
    mean(cv_results_df$Specificity),
    mean(cv_results_df$Precision, na.rm = TRUE),
    mean(cv_results_df$F1, na.rm = TRUE)
  ),
  CV_SD = c(
    sd(cv_results_df$AUC),
    NA,
    sd(cv_results_df$Balanced_Accuracy),
    sd(cv_results_df$Sensitivity),
    sd(cv_results_df$Specificity),
    sd(cv_results_df$Precision, na.rm = TRUE),
    sd(cv_results_df$F1, na.rm = TRUE)
  )
)

write_csv(performance_df, file.path(output_dir, "rf_classification_performance_eif3d_promotes_hypoxia_1hr.csv"))
cat("Exported: rf_classification_performance_eif3d_promotes_hypoxia_1hr.csv\n")

# Save CV results
write_csv(cv_results_df, file.path(output_dir, "rf_classification_cv_results_eif3d_promotes_hypoxia_1hr.csv"))
cat("Exported: rf_classification_cv_results_eif3d_promotes_hypoxia_1hr.csv\n")

# Save predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_promotes = pred_promotes_prob,
  prob_does_not_promote = pred_probs[, "does_not_promote"]
)

write_csv(predictions_export, file.path(output_dir, "rf_classification_predictions_eif3d_promotes_hypoxia_1hr.csv"))
cat("Exported: rf_classification_predictions_eif3d_promotes_hypoxia_1hr.csv\n")
```

# Summary

```{r summary}
cat("\n")
cat("========================================\n")
cat("CLASSIFICATION MODEL SUMMARY (HYPOXIA)\n")
cat("========================================\n\n")

cat("Dataset:\n")
cat("  - Total genes:", nrow(model_data), "\n")
cat("  - Positive class (promotes):", sum(model_data$eif3d_target == "promotes"), "\n")
cat("  - Negative class (does_not_promote):", sum(model_data$eif3d_target == "does_not_promote"), "\n")
cat("  - Class imbalance ratio:", round(n_does_not / n_promotes, 1), ":1\n")
cat("  - Features:", length(feature_cols_final), "\n\n")

cat("Training:\n")
cat("  - Training set:", nrow(train_data), "(using class weights)\n")
cat("  - Weight ratio:", round(weight_promotes / weight_does_not_promote, 1), ":1 (promotes:does_not_promote)\n")
cat("  - Test set:", nrow(test_data), "\n")
cat("  - Decision threshold: Youden optimal (", round(optimal_threshold, 4), ")\n\n")

cat("Model Performance (Test Set @ Youden Threshold):\n")
cat("  - AUC-ROC:", round(as.numeric(auc_value), 4), "\n")
cat("  - Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  - Sensitivity (Recall):", round(sensitivity, 4), "\n")
cat("  - Specificity:", round(specificity, 4), "\n")
cat("  - Precision:", round(precision, 4), "\n")
cat("  - F1 Score:", round(f1, 4), "\n\n")

cat("Cross-Validation (5-fold, Youden threshold per fold):\n")
cat("  - Mean Threshold:", round(mean(cv_results_df$Threshold), 4), "+/-", round(sd(cv_results_df$Threshold), 4), "\n")
cat("  - Mean AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  - Mean Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-", round(sd(cv_results_df$Balanced_Accuracy), 4), "\n")
cat("  - Mean Sensitivity:", round(mean(cv_results_df$Sensitivity), 4), "+/-", round(sd(cv_results_df$Sensitivity), 4), "\n")
cat("  - Mean Specificity:", round(mean(cv_results_df$Specificity), 4), "+/-", round(sd(cv_results_df$Specificity), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

cat("\nInterpretation:\n")
cat("  - AUC > 0.5 indicates better than random classification\n")
cat("  - AUC > 0.7 indicates acceptable discrimination\n")
cat("  - AUC > 0.8 indicates excellent discrimination\n")

cat("\nOutput Files:\n")
cat("  - rf_classifier_eif3d_promotes_hypoxia_1hr.rds\n")
cat("  - rf_classification_importance_eif3d_promotes_hypoxia_1hr.csv\n")
cat("  - rf_classification_performance_eif3d_promotes_hypoxia_1hr.csv\n")
cat("  - rf_classification_cv_results_eif3d_promotes_hypoxia_1hr.csv\n")
cat("  - rf_classification_predictions_eif3d_promotes_hypoxia_1hr.csv\n")
cat("========================================\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
