---
title: "Random Forest Classification: eIF3d Target Prediction in Hypoxia (Negative Control Set)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Binary classification: does a gene's mRNA sequence predict whether eIF3d promotes its translation?

- **Positive class ("promotes")**: Genes from `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- **Negative class ("does_not_promote")**: Genes from `negative_control_genes.csv` (not regulated by eIF3d in any condition/timepoint)

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(Boruta)

```

# Load Data

```{r load_data}
# Feature matrix
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_targeted.rds"))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

# Positive class: eIF3d promotes TE in 1hr hypoxia
promotes_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
cat("Promotes genes:", nrow(promotes_genes), "\n")

# Negative class: genes not regulated by eIF3d in any condition
negative_control_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
cat("Negative control genes:", nrow(negative_control_genes), "\n")
```

# Create Binary Target

```{r create_target}
set.seed(9)
number_to_match <- min(length(promotes_genes$ensembl_gene), nrow(negative_control_genes))

# Label genes
features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% promotes_genes$ensembl_gene ~ "promotes",
    gene_id_clean %in% sample(negative_control_genes$gene_id_clean, number_to_match) ~ "does_not_promote",
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c("does_not_promote", "promotes")))

nrow(model_genes)

cat("Positive (promotes):", sum(model_genes$eif3d_target == "promotes"), "\n")
cat("Negative (does_not_promote):", sum(model_genes$eif3d_target == "does_not_promote"), "\n")
cat("Total:", nrow(model_genes), "\n")
cat("Imbalance ratio (neg:pos):",
    round(sum(model_genes$eif3d_target == "does_not_promote") /
          sum(model_genes$eif3d_target == "promotes"), 1), ":1\n")
```

# Preprocessing

## Remove Near-Zero Variance Features

```{r remove_nzv}
# # Identify feature columns (exclude identifiers and target)
# id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
#              "transcript_id", "transcript_id_clean", "transcript_id_raw",
#              "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")

# all_cols <- colnames(model_genes)
# id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
# id_cols <- unique(c(id_cols, id_pattern_cols))

# feature_cols <- setdiff(colnames(model_genes), id_cols)
# cat("Initial feature count:", length(feature_cols), "\n")

# # Remove near-zero variance
# nzv <- nearZeroVar(model_genes[, feature_cols], saveMetrics = TRUE)
# nzv_to_remove <- rownames(nzv)[nzv$nzv]
# feature_cols <- setdiff(feature_cols, nzv_to_remove)
# cat("Removed", length(nzv_to_remove), "NZV features\n")
# cat("Features after NZV removal:", length(feature_cols), "\n")
```

```{r}

# Identify feature columns (same as before)
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")

all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))

feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count:", length(feature_cols), "\n")

# ---- Boruta feature selection ----
set.seed(9)

boruta_res <- Boruta(
  x = model_genes[, feature_cols],
  y = as.factor(model_genes$eif3d_target),  # classification
  doTrace = 2
)

# Resolve tentative features
boruta_res <- TentativeRoughFix(boruta_res)

# Keep confirmed features only
selected_features <- getSelectedAttributes(boruta_res, withTentative = TRUE)

cat("Selected features after Boruta:", length(selected_features), "\n")

# Update feature set for RF
feature_cols <- selected_features


plot(boruta_res, las = 2, cex.axis = 0.7)

attStats(boruta_res) %>%
  arrange(desc(meanImp))
```




## Median-Impute Missing Values

```{r impute_missing}
missing_counts <- model_genes %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:", nrow(missing_counts), "\n")
if (nrow(missing_counts) > 0) print(missing_counts)

model_genes <- model_genes %>%
  mutate(across(all_of(feature_cols), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

cat("Remaining NAs:", sum(is.na(model_genes[, feature_cols])), "\n")
```



# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
cat("\nClass distribution:\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif3d_target))
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif3d_target))
```

# Train Random Forest

```{r train_rf}
# Class weights (inverse frequency)
class_counts <- table(train_data$eif3d_target)
n_train <- nrow(train_data)
weight_promotes <- n_train / (2 * class_counts["promotes"])
weight_does_not_promote <- n_train / (2 * class_counts["does_not_promote"])

case_weights <- ifelse(train_data$eif3d_target == "promotes",
                       weight_promotes, weight_does_not_promote)

cat("Class weights — promotes:", round(weight_promotes, 2),
    ", does_not_promote:", round(weight_does_not_promote, 2), "\n")

# Train
set.seed(9)
rf_model <- ranger(
  formula = eif3d_target ~ .,
  data = train_data %>% dplyr::select(eif3d_target, all_of(feature_cols)),
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  case.weights = case_weights,
  seed = 9
)

print(rf_model)
```

# Evaluate on Test Set

## ROC Curve and AUC

```{r roc_curve, fig.width=8, fig.height=6}
# Predictions
pred_probs <- predict(rf_model, test_data %>% dplyr::select(all_of(feature_cols)))$predictions
pred_promotes_prob <- pred_probs[, "promotes"]
actual <- test_data$eif3d_target

# ROC
roc_obj <- roc(response = actual, predictor = pred_promotes_prob,
               levels = c("does_not_promote", "promotes"), direction = "<")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

# Youden threshold
coords_youden <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")
optimal_threshold <- coords_youden$threshold

cat("Youden threshold:", round(optimal_threshold, 4), "\n")
cat("  Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  Specificity:", round(coords_youden$specificity, 3), "\n")

# Plot
roc_df <- data.frame(sensitivity = roc_obj$sensitivities,
                     specificity = roc_obj$specificities)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = "ROC Curve: eIF3d Target Classification (Hypoxia)",
    subtitle = paste0("AUC = ", round(auc_value, 3),
                      " | Threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

print(p_roc)

ggsave(here("plots", "rf_classification_roc_eif3d_promotes_hypoxia_1hr_v2.pdf"),
       plot = p_roc, width = 8, height = 6)
```

## Confusion Matrix

```{r confusion_matrix}
pred_class <- factor(
  ifelse(pred_promotes_prob >= optimal_threshold, "promotes", "does_not_promote"),
  levels = c("does_not_promote", "promotes")
)

cm <- confusionMatrix(pred_class, actual, positive = "promotes")
print(cm)

# Extract metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\nKey Metrics (Youden Threshold):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n")
```

## Confusion Matrix Plot

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = "Confusion Matrix (Hypoxia)",
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3),
                      ", F1 = ", round(f1, 3))
  ) +
  theme(legend.position = "none")

print(p_cm)

ggsave(here("plots", "rf_classification_confusion_matrix_eif3d_promotes_hypoxia_1hr_v2.pdf"),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(model_data$eif3d_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- model_data[-test_idx, ]
  test_cv <- model_data[test_idx, ]

  # Class weights for this fold
  fold_counts <- table(train_cv$eif3d_target)
  n_fold <- nrow(train_cv)
  fold_weights <- ifelse(
    train_cv$eif3d_target == "promotes",
    n_fold / (2 * fold_counts["promotes"]),
    n_fold / (2 * fold_counts["does_not_promote"])
  )

  # Train
  rf_cv <- ranger(
    formula = eif3d_target ~ .,
    data = train_cv %>% dplyr::select(eif3d_target, all_of(feature_cols)),
    num.trees = 500,
    probability = TRUE,
    case.weights = fold_weights,
    seed = 9
  )

  # Predict
  pred_probs_cv <- predict(rf_cv, test_cv %>% dplyr::select(all_of(feature_cols)))$predictions[, "promotes"]

  # ROC + Youden threshold
  roc_cv <- roc(test_cv$eif3d_target, pred_probs_cv,
                levels = c("does_not_promote", "promotes"), direction = "<", quiet = TRUE)
  coords_cv <- coords(roc_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                      best.method = "youden")

  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= coords_cv$threshold, "promotes", "does_not_promote"),
    levels = c("does_not_promote", "promotes")
  )
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif3d_target, positive = "promotes")

  data.frame(
    fold = i,
    Threshold = coords_cv$threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n5-Fold CV Results:\n")
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("AUC", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Clean names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

# Bar plot
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: eIF3d Target Classification (Hypoxia)",
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", "rf_classification_feature_importance_eif3d_promotes_hypoxia_1hr_v2.pdf"),
       plot = p_importance, width = 10, height = 10)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^utr5_4mer_") ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_") ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_") ~ "CDS 4-mers",
      str_detect(feature, "^rbp_|^rbns_|^cisbp_") ~ "RBP Motifs",
      str_detect(feature, "^clip_") ~ "CLIP Binding",
      str_detect(feature, "^sig_") ~ "Signatures",
      str_detect(feature, "length|Length") ~ "Region Lengths",
      str_detect(feature, "_gc$|gc3$|GC") ~ "GC Content",
      str_detect(feature, "^is_") ~ "4E/DAP5",
      TRUE ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    top_feature = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nImportance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Importance by Category (Hypoxia)",
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", "rf_classification_importance_by_category_eif3d_promotes_hypoxia_1hr_v2.pdf"),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Model
saveRDS(rf_model, file.path(output_dir, "rf_classifier_eif3d_promotes_hypoxia_1hr_v2.rds"))

# Feature importance
write_csv(importance_df, file.path(output_dir, "rf_classification_importance_eif3d_promotes_hypoxia_1hr_v2.csv"))

# Performance summary
performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy,
               sensitivity, specificity, precision, f1),
  CV_Mean = c(mean(cv_results_df$AUC), NA, mean(cv_results_df$Balanced_Accuracy),
              mean(cv_results_df$Sensitivity), mean(cv_results_df$Specificity),
              mean(cv_results_df$Precision, na.rm = TRUE),
              mean(cv_results_df$F1, na.rm = TRUE)),
  CV_SD = c(sd(cv_results_df$AUC), NA, sd(cv_results_df$Balanced_Accuracy),
            sd(cv_results_df$Sensitivity), sd(cv_results_df$Specificity),
            sd(cv_results_df$Precision, na.rm = TRUE),
            sd(cv_results_df$F1, na.rm = TRUE))
)
write_csv(performance_df, file.path(output_dir, "rf_classification_performance_eif3d_promotes_hypoxia_1hr_v2.csv"))

# CV results
write_csv(cv_results_df, file.path(output_dir, "rf_classification_cv_results_eif3d_promotes_hypoxia_1hr_v2.csv"))

# Predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_promotes = pred_promotes_prob,
  prob_does_not_promote = pred_probs[, "does_not_promote"]
)
write_csv(predictions_export, file.path(output_dir, "rf_classification_predictions_eif3d_promotes_hypoxia_1hr_v2.csv"))

cat("Exported: model RDS, importance CSV, performance CSV, CV results CSV, predictions CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat("CLASSIFICATION SUMMARY (Hypoxia)\n")
cat("========================================\n\n")

cat("Classes:\n")
cat("  Positive (promotes):", sum(model_data$eif3d_target == "promotes"), "\n")
cat("  Negative (does_not_promote):", sum(model_data$eif3d_target == "does_not_promote"), "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("Test Set Performance (Youden threshold =", round(optimal_threshold, 4), "):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n\n")

cat("5-Fold CV:\n")
cat("  AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-",
    round(sd(cv_results_df$Balanced_Accuracy), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}
cat("========================================\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
