---
title: "`r paste0('Random Forest Classification: eIF', params$gene, ' Target Prediction (', tools::toTitleCase(params$condition), ', ', params$direction, ', ', params$timepoint, ', LFC ', params$lfc, ', ', params$feature_set, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "normoxia"
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "combined"
  clip_mode: "individual"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    condition = "hypoxia",
    gene      = "3d",
    direction = "promotes",
    timepoint = "1hr",
    lfc       = "0.5",
    feature_set = "combined", #targeted or combined
    clip_mode   = "aggregate" # "individual", "aggregate", or "both"
  )
}

# Derived file paths and labels from params
geneset_csv <- paste0(params$condition, "_", params$gene, "_", params$direction,
                      "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")
feature_matrix_rds <- paste0("feature_matrix_", params$timepoint, "_", params$feature_set, ".rds")
suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                 params$condition, "_", params$timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode)
pos_label <- params$direction
neg_label <- paste0("does_not_", sub("s$", "", params$direction))
gene_label <- paste0("eIF", params$gene)

cat("Gene set CSV:", geneset_csv, "\n")
cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix:", suffix, "\n")
cat("Positive label:", pos_label, "\n")
cat("Negative label:", neg_label, "\n")
```

# Overview

Binary classification: does a gene's mRNA sequence predict whether `r gene_label` `r params$direction` its translation?

- **Positive class ("`r pos_label`")**: Genes from `` `r geneset_csv` ``
- **Negative class ("`r neg_label`")**: Genes from `negative_control_genes.csv` (not regulated by `r gene_label` in any condition/timepoint)

## Parameters

| Parameter | Values | Description |
|-----------|--------|-------------|
| `condition` | `"hypoxia"`, `"normoxia"` | Experimental condition |
| `gene` | `"3d"`, `"3e"` | eIF3 subunit |
| `direction` | `"promotes"`, `"inhibits"` | Direction of TE regulation |
| `timepoint` | `"1hr"`, `"4hr"`, `"24hr"` | Hypoxia timepoint |
| `lfc` | `"0.5"`, `"1"` | Log2 fold change threshold for gene set |
| `feature_set` | `"targeted"`, `"combined"` | Feature matrix to use: `"targeted"` = biological features only (CLIP, 4E, DAP5; ~36 features from 01b); `"combined"` = biological + UTR lengths + GC + k-mers + RBP motifs (~6,324 features from 01c) |
| `clip_mode` | `"individual"`, `"aggregate"`, `"both"` | Which CLIP features to use: `"individual"` = per-dataset regional binary features (~21 cols: 5utr/cds/3utr × 7 datasets); `"aggregate"` = single numeric count of datasets with any binding evidence (0-7); `"both"` = all CLIP features together |

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(Boruta)
# library(treeshap)
# library(shapviz)

```

# Load Data

```{r load_data}
# Feature matrix
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

# Positive class
positive_genes <- read_csv(here("output", "genesets", geneset_csv))
cat("Positive genes:", nrow(positive_genes), "\n")

# Negative class: genes not regulated by eIF3d in any condition
negative_control_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
cat("Negative control genes:", nrow(negative_control_genes), "\n")
```

# Create Binary Target

```{r create_target}
set.seed(9)
number_to_match <- min(length(positive_genes$ensembl_gene), nrow(negative_control_genes))

# Label genes
features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(negative_control_genes$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

nrow(model_genes)

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
cat("Imbalance ratio (neg:pos):",
    round(sum(model_genes$eif3d_target == neg_label) /
          sum(model_genes$eif3d_target == pos_label), 1), ":1\n")
```

# Preprocessing

## Remove Near-Zero Variance Features

```{r}

# Identify feature columns (same as before)
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")

all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))

feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count:", length(feature_cols), "\n")

# ---- CLIP feature mode ----
if (params$clip_mode == "aggregate") {
  regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & feature_cols != "clip_evidence_count"]
  feature_cols <- setdiff(feature_cols, regional_clip)
  cat("CLIP mode: aggregate — using clip_evidence_count only, dropped", length(regional_clip), "regional features\n")
} else if (params$clip_mode == "individual") {
  feature_cols <- setdiff(feature_cols, "clip_evidence_count")
  cat("CLIP mode: individual — using regional binary features, dropped clip_evidence_count\n")
} else {
  cat("CLIP mode: both — using all CLIP features\n")
}

# ---- Boruta feature selection ----
set.seed(9)

boruta_res <- Boruta(
  x = model_genes[, feature_cols],
  y = as.factor(model_genes$eif3d_target),  # classification
  doTrace = 2
)

# Resolve tentative features
boruta_res <- TentativeRoughFix(boruta_res)

# Keep confirmed features only
selected_features <- getSelectedAttributes(boruta_res, withTentative = TRUE)

cat("Selected features after Boruta:", length(selected_features), "\n")

# Update feature set for RF
feature_cols <- selected_features


# Dynamic width based on number of features (3 shadow attrs + confirmed/rejected)
n_boruta_attrs <- length(boruta_res$finalDecision) + 3  # +3 for shadow min/mean/max
boruta_plot_width <- max(10, n_boruta_attrs * 0.35)

pdf(here("plots", paste0("rf_boruta_importance_", suffix, ".pdf")),
    width = boruta_plot_width, height = 7)
par(mar = c(10, 4, 4, 2))  # extra bottom margin for rotated labels
plot(boruta_res, las = 2, cex.axis = 0.7)
dev.off()

# Also display inline
plot(boruta_res, las = 2, cex.axis = 0.7)

# ggsave(here("plots/barutafeats.pdf"), width=8, height=9)

attStats(boruta_res) %>%
  arrange(desc(meanImp))
```

```{r boruta_diagnostics, fig.width=10, fig.height=8}
# ---- A. Category-level summary table ----
boruta_stats <- attStats(boruta_res) %>%
  tibble::rownames_to_column("feature") %>%
  mutate(
    category = case_when(
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      TRUE                                                    ~ "Other"
    )
  )

category_boruta_summary <- boruta_stats %>%
  group_by(category) %>%
  summarise(
    n_total     = n(),
    n_confirmed = sum(decision == "Confirmed"),
    n_rejected  = sum(decision == "Rejected"),
    n_tentative = sum(decision == "Tentative"),
    pct_confirmed = round(100 * n_confirmed / n_total, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_confirmed))

cat("\n=== Boruta decisions by feature category ===\n")
print(as.data.frame(category_boruta_summary), row.names = FALSE)

# ---- B. Detailed biological feature table ----
shadow_max_imp <- boruta_stats %>%
  filter(str_detect(feature, "^shadow")) %>%
  pull(meanImp) %>%
  max(na.rm = TRUE)

bio_patterns <- "^orfquant_|^clip_|^4e_|^dap5_|^eif4e_ip|^lit_"

bio_features <- boruta_stats %>%
  filter(str_detect(feature, bio_patterns)) %>%
  mutate(
    shadow_max_threshold = shadow_max_imp,
    margin = meanImp - shadow_max_imp,
    narrow_reject = decision == "Rejected" & margin > -1
  ) %>%
  arrange(desc(meanImp))

cat("\n=== Biological feature detail (shadow max threshold =",
    round(shadow_max_imp, 3), ") ===\n")
print(as.data.frame(bio_features %>%
  dplyr::select(feature, decision, meanImp, normHits, shadow_max_threshold,
         margin, narrow_reject)), row.names = FALSE)

# ---- C. Dot plot visualization ----
if (nrow(bio_features) > 0) {
  shadow_stats <- boruta_stats %>%
    filter(str_detect(feature, "^shadow")) %>%
    summarise(
      shadow_min = min(meanImp, na.rm = TRUE),
      shadow_mean = mean(meanImp, na.rm = TRUE),
      shadow_max = max(meanImp, na.rm = TRUE)
    )

  p_bio <- ggplot(bio_features,
                  aes(x = meanImp,
                      y = reorder(feature, meanImp),
                      color = decision)) +
    geom_point(size = 3) +
    geom_vline(xintercept = shadow_stats$shadow_min,
               linetype = "dotted", color = "grey50") +
    geom_vline(xintercept = shadow_stats$shadow_mean,
               linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = shadow_stats$shadow_max,
               linetype = "solid", color = "grey30") +
    scale_color_manual(values = c("Confirmed" = "forestgreen",
                                  "Rejected"  = "firebrick",
                                  "Tentative" = "goldenrod")) +
    labs(
      title = paste0("Boruta: biological features vs shadow threshold (", suffix, ")"),
      x = "Mean Importance",
      y = NULL,
      color = "Decision"
    ) +
    annotate("text", x = shadow_stats$shadow_max, y = 0.5,
             label = "shadow max", hjust = -0.1, size = 3, color = "grey30") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))

  print(p_bio)

  ggsave(here("plots", paste0("rf_boruta_bio_features_", suffix, ".pdf")),
         plot = p_bio, width = 10, height = max(4, nrow(bio_features) * 0.35))
}
```

```{r boruta_shap, fig.width=10, fig.height=max(6, length(feature_cols) * 0.35)}
# # ---- SHAP directionality (Boruta-selected features) ----
# set.seed(9)

# # Balanced class weights (same logic as train_rf chunk)
# shap_counts <- table(model_genes$eif3d_target)
# n_shap <- nrow(model_genes)
# shap_weight_pos <- n_shap / (2 * shap_counts[pos_label])
# shap_weight_neg <- n_shap / (2 * shap_counts[neg_label])
# shap_case_weights <- ifelse(model_genes$eif3d_target == pos_label,
#                             shap_weight_pos, shap_weight_neg)

# # Train temporary ranger on Boruta-selected features
# rf_boruta_shap <- ranger(
#   formula = eif3d_target ~ .,
#   data = model_genes %>% dplyr::select(eif3d_target, all_of(feature_cols)),
#   num.trees = 500,
#   probability = TRUE,
#   case.weights = shap_case_weights,
#   seed = 9
# )

# # Convert to treeshap format
# unified_model <- ranger.unify(rf_boruta_shap,
#                               model_genes %>% dplyr::select(all_of(feature_cols)))

# # Compute SHAP values (subsample if large)
# shap_data <- model_genes %>% dplyr::select(all_of(feature_cols))
# if (nrow(shap_data) > 500) {
#   shap_idx <- sample(nrow(shap_data), 500)
#   shap_data <- shap_data[shap_idx, ]
# }

# shap_res <- treeshap(unified_model, shap_data)

# # Build shapviz object
# sv <- shapviz(shap_res)

# # Beeswarm plot (directionality)
# p_shap_bee <- sv_importance(sv, kind = "beeswarm", max_display = 30L) +
#   theme_cowplot() +
#   labs(title = paste0("SHAP Feature Directionality: Boruta-Selected Features"),
#        subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
#                          " | ", params$timepoint))

# print(p_shap_bee)

# ggsave(here("plots", paste0("rf_boruta_shap_beeswarm_", suffix, ".pdf")),
#        plot = p_shap_bee, width = 10,
#        height = max(6, length(feature_cols) * 0.35))

# # Bar plot (mean |SHAP|)
# p_shap_bar <- sv_importance(sv, kind = "bar", max_display = 30L) +
#   theme_cowplot() +
#   labs(title = paste0("SHAP Feature Importance: Boruta-Selected Features"),
#        subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
#                          " | ", params$timepoint))

# print(p_shap_bar)

# ggsave(here("plots", paste0("rf_boruta_shap_importance_", suffix, ".pdf")),
#        plot = p_shap_bar, width = 10,
#        height = max(6, length(feature_cols) * 0.35))

# # Export SHAP summary
# shap_summary <- data.frame(
#   feature = colnames(shap_res$shaps),
#   mean_abs_shap = colMeans(abs(shap_res$shaps)),
#   mean_shap = colMeans(shap_res$shaps)
# ) %>% arrange(desc(mean_abs_shap))

# write_csv(shap_summary, here("output", "predictive_modeling",
#                               paste0("rf_boruta_shap_summary_", suffix, ".csv")))

# cat("SHAP summary (positive mean_shap = feature pushes toward", pos_label, "class):\n")
# print(shap_summary)
```

```{r correlated_feature_selection, fig.width=10, fig.height=8}
boruta_feats <- getSelectedAttributes(boruta_res, withTentative = TRUE)

cor_mat <- cor(
  model_genes[, boruta_feats],
  use = "pairwise.complete.obs"
)

hc <- hclust(as.dist(1 - abs(cor_mat)))

# Cut tree at correlation threshold of 0.7 (i.e., distance = 0.3)
clusters <- cutree(hc, h = 0.3)

boruta_stats <- attStats(boruta_res)

cluster_df <- boruta_stats %>%
  rownames_to_column("feature") %>%
  filter(feature %in% boruta_feats) %>%
  mutate(cluster = clusters[feature]) %>%
  arrange(cluster, desc(meanImp))

# Show all clusters with >1 member (i.e., correlated groups)
multi_member <- cluster_df %>%
  group_by(cluster) %>%
  filter(n() > 1) %>%
  ungroup()

cat("=== Correlated feature clusters (|r| >= 0.7) ===\n\n")
if (nrow(multi_member) > 0) {
  for (cl in unique(multi_member$cluster)) {
    members <- multi_member %>% filter(cluster == cl)
    cat("Cluster", cl, "(", nrow(members), "features ):\n")
    for (j in seq_len(nrow(members))) {
      cat("  ", members$feature[j],
          "  (Boruta meanImp =", round(members$meanImp[j], 2), ")\n")
    }
    # Show pairwise correlations within this cluster
    cl_feats <- members$feature
    if (length(cl_feats) > 1) {
      cat("  Pairwise correlations:\n")
      for (a in 1:(length(cl_feats) - 1)) {
        for (b in (a + 1):length(cl_feats)) {
          cat("    ", cl_feats[a], " <-> ", cl_feats[b], ": ",
              round(cor_mat[cl_feats[a], cl_feats[b]], 3), "\n")
        }
      }
    }
    cat("\n")
  }
} else {
  cat("No correlated clusters found — all Boruta features are independent.\n\n")
}

# Singletons (uncorrelated features — automatically kept)
singletons <- cluster_df %>%
  group_by(cluster) %>%
  filter(n() == 1) %>%
  ungroup()

cat("=== Uncorrelated features (automatically kept):", nrow(singletons), "===\n")
print(singletons %>% dplyr::select(feature, meanImp) %>% arrange(desc(meanImp)))
```

Review the correlated clusters above and choose which feature to keep from each.
Edit the `selected` vector below — one feature per correlated cluster.

```{r choose_representatives}
# ---- EDIT THIS: pick one feature per correlated cluster ----
# Default: highest Boruta importance per cluster
selected_from_clusters <- cluster_df %>%
  group_by(cluster) %>%
  slice_max(meanImp, n = 1) %>%
  ungroup() %>%
  pull(feature)

# To override, uncomment and replace specific features:
# Example: swap out "utr5_length" for "log2_utr5_length" in cluster 3
# selected_from_clusters[selected_from_clusters == "utr5_length"] <- "log2_utr5_length"

feature_cols <- selected_from_clusters
cat("Final feature count:", length(feature_cols), "\n")
cat("Final features:\n")
cat(paste(" ", feature_cols), sep = "\n")
```






## Median-Impute Missing Values

```{r impute_missing}
missing_counts <- model_genes %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:", nrow(missing_counts), "\n")
if (nrow(missing_counts) > 0) print(missing_counts)

model_genes <- model_genes %>%
  mutate(across(all_of(feature_cols), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

cat("Remaining NAs:", sum(is.na(model_genes[, feature_cols])), "\n")
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
cat("\nClass distribution:\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif3d_target))
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif3d_target))
```

# Train Random Forest

```{r train_rf}
# Class weights (inverse frequency)
class_counts <- table(train_data$eif3d_target)
n_train <- nrow(train_data)
weight_pos <- n_train / (2 * class_counts[pos_label])
weight_neg <- n_train / (2 * class_counts[neg_label])

case_weights <- ifelse(train_data$eif3d_target == pos_label,
                       weight_pos, weight_neg)

cat("Class weights —", pos_label, ":", round(weight_pos, 2),
    ",", neg_label, ":", round(weight_neg, 2), "\n")

# Train
set.seed(9)
rf_model <- ranger(
  formula = eif3d_target ~ .,
  data = train_data %>% dplyr::select(eif3d_target, all_of(feature_cols)),
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  case.weights = case_weights,
  seed = 9
)

print(rf_model)
```

# Evaluate on Test Set

## ROC Curve and AUC

```{r roc_curve, fig.width=8, fig.height=6}
# Predictions
pred_probs <- predict(rf_model, test_data %>% dplyr::select(all_of(feature_cols)))$predictions
pred_pos_prob <- pred_probs[, pos_label]
actual <- test_data$eif3d_target

# ROC
roc_obj <- roc(response = actual, predictor = pred_pos_prob,
               levels = c(neg_label, pos_label), direction = "<")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

# Youden threshold
coords_youden <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")
optimal_threshold <- coords_youden$threshold

cat("Youden threshold:", round(optimal_threshold, 4), "\n")
cat("  Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  Specificity:", round(coords_youden$specificity, 3), "\n")

# Plot
roc_df <- data.frame(sensitivity = roc_obj$sensitivities,
                     specificity = roc_obj$specificities)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = paste0("ROC Curve: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = paste0("AUC = ", round(auc_value, 3),
                      " | Threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

print(p_roc)

ggsave(here("plots", paste0("rf_classification_roc_", suffix, ".pdf")),
       plot = p_roc, width = 8, height = 6)
```

## Confusion Matrix

```{r confusion_matrix}
pred_class <- factor(
  ifelse(pred_pos_prob >= optimal_threshold, pos_label, neg_label),
  levels = c(neg_label, pos_label)
)

cm <- confusionMatrix(pred_class, actual, positive = pos_label)
print(cm)

# Extract metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\nKey Metrics (Youden Threshold):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n")
```

## Confusion Matrix Plot

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = paste0("Confusion Matrix (", tools::toTitleCase(params$condition), ")"),
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3),
                      ", F1 = ", round(f1, 3))
  ) +
  theme(legend.position = "none")

print(p_cm)

ggsave(here("plots", paste0("rf_classification_confusion_matrix_", suffix, ".pdf")),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(model_data$eif3d_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- model_data[-test_idx, ]
  test_cv <- model_data[test_idx, ]

  # Class weights for this fold
  fold_counts <- table(train_cv$eif3d_target)
  n_fold <- nrow(train_cv)
  fold_weights <- ifelse(
    train_cv$eif3d_target == pos_label,
    n_fold / (2 * fold_counts[pos_label]),
    n_fold / (2 * fold_counts[neg_label])
  )

  # Train
  rf_cv <- ranger(
    formula = eif3d_target ~ .,
    data = train_cv %>% dplyr::select(eif3d_target, all_of(feature_cols)),
    num.trees = 500,
    probability = TRUE,
    case.weights = fold_weights,
    seed = 9
  )

  # Predict
  pred_probs_cv <- predict(rf_cv, test_cv %>% dplyr::select(all_of(feature_cols)))$predictions[, pos_label]

  # ROC + Youden threshold
  roc_cv <- roc(test_cv$eif3d_target, pred_probs_cv,
                levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  coords_cv <- coords(roc_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                      best.method = "youden")

  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= coords_cv$threshold, pos_label, neg_label),
    levels = c(neg_label, pos_label)
  )
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif3d_target, positive = pos_label)

  data.frame(
    fold = i,
    Threshold = coords_cv$threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n5-Fold CV Results:\n")
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("AUC", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Clean names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

# Bar plot
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top 30 Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", paste0("rf_classification_feature_importance_", suffix, ".pdf")),
       plot = p_importance, width = 10, height = 10)

# All-features bar plot
all_height <- max(6, nrow(importance_df) * 0.35)

p_importance_all <- importance_df %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("All Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_all)

ggsave(here("plots", paste0("rf_classification_feature_importance_all_", suffix, ".pdf")),
       plot = p_importance_all, width = 10, height = all_height)

# Custom-features-only bar plot (exclude k-mer and RBP motif features)
importance_custom <- importance_df %>%
  filter(!str_detect(feature, "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbp_|^rbns_|^cisbp_"))

custom_height <- max(6, nrow(importance_custom) * 0.35)

p_importance_custom <- importance_custom %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Custom Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity) — CLIP, 4E/DAP5, Signatures, Lengths, GC",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_custom)

ggsave(here("plots", paste0("rf_classification_feature_importance_custom_", suffix, ".pdf")),
       plot = p_importance_custom, width = 10, height = custom_height)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      TRUE                                                    ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    top_feature = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nImportance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance by Category (", tools::toTitleCase(params$condition), ")"),
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", paste0("rf_classification_importance_by_category_", suffix, ".pdf")),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Model
saveRDS(rf_model, file.path(output_dir, paste0("rf_classifier_", suffix, ".rds")))

# Feature importance
write_csv(importance_df, file.path(output_dir, paste0("rf_classification_importance_", suffix, ".csv")))

# Performance summary
performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy,
               sensitivity, specificity, precision, f1),
  CV_Mean = c(mean(cv_results_df$AUC), NA, mean(cv_results_df$Balanced_Accuracy),
              mean(cv_results_df$Sensitivity), mean(cv_results_df$Specificity),
              mean(cv_results_df$Precision, na.rm = TRUE),
              mean(cv_results_df$F1, na.rm = TRUE)),
  CV_SD = c(sd(cv_results_df$AUC), NA, sd(cv_results_df$Balanced_Accuracy),
            sd(cv_results_df$Sensitivity), sd(cv_results_df$Specificity),
            sd(cv_results_df$Precision, na.rm = TRUE),
            sd(cv_results_df$F1, na.rm = TRUE))
)
write_csv(performance_df, file.path(output_dir, paste0("rf_classification_performance_", suffix, ".csv")))

# CV results
write_csv(cv_results_df, file.path(output_dir, paste0("rf_classification_cv_results_", suffix, ".csv")))

# Predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_pos = pred_pos_prob,
  prob_neg = pred_probs[, neg_label]
)
write_csv(predictions_export, file.path(output_dir, paste0("rf_classification_predictions_", suffix, ".csv")))

cat("Exported: model RDS, importance CSV, performance CSV, CV results CSV, predictions CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat(paste0("CLASSIFICATION SUMMARY (", tools::toTitleCase(params$condition), ")\n"))
cat("========================================\n\n")

cat("Parameters:\n")
cat("  Condition:", params$condition, "\n")
cat("  Gene:", paste0("eIF", params$gene), "\n")
cat("  Direction:", params$direction, "\n")
cat("  Timepoint:", params$timepoint, "\n")
cat("  LFC threshold:", params$lfc, "\n")
cat("  Feature set:", params$feature_set, "\n")
cat("  CLIP mode:", params$clip_mode, "\n\n")

cat("Classes:\n")
cat("  Positive (", pos_label, "):", sum(model_data$eif3d_target == pos_label), "\n")
cat("  Negative (", neg_label, "):", sum(model_data$eif3d_target == neg_label), "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("Test Set Performance (Youden threshold =", round(optimal_threshold, 4), "):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n\n")

cat("5-Fold CV:\n")
cat("  AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-",
    round(sd(cv_results_df$Balanced_Accuracy), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

# cat("\nTop 5 Features by SHAP (directional):\n")
# for (i in 1:min(5, nrow(shap_summary))) {
#   direction <- if (shap_summary$mean_shap[i] > 0) paste0("-> ", pos_label) else paste0("-> ", neg_label)
#   cat("  ", i, ". ", shap_summary$feature[i],
#       " (|SHAP| = ", round(shap_summary$mean_abs_shap[i], 4),
#       ", ", direction, ")\n", sep = "")
# }
cat("========================================\n")
```

# Final Feature Importance Plot

```{r final_importance_plot, fig.width=8, fig.height=max(4, length(feature_cols) * 0.35)}
# Lollipop plot of all selected features
p_final_importance <- importance_df %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_segment(aes(xend = reorder(feature_clean, importance), y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(size = 3, color = "#0072B2") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance: ", gene_label, " Target Classification"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | LFC ", params$lfc, " | AUC = ", round(as.numeric(auc_value), 3)),
    x = "",
    y = "Variable Importance (Impurity)"
  )

print(p_final_importance)

ggsave(here("plots", paste0("rf_final_importance_", suffix, ".pdf")),
       plot = p_final_importance, width = 8,
       height = max(4, length(feature_cols) * 0.35))
```

# Session Info

```{r session_info}
sessionInfo()
```


