---
title: "`r paste0('Random Forest Classification (No Boruta): eIF', params$gene, ' Target Prediction (', tools::toTitleCase(params$condition), ', ', params$direction, ', ', params$timepoint, ', LFC ', params$lfc, ', ', params$feature_set, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "normoxia"
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "te_lfc"
  clip_mode: "aggregate"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    condition = "hypoxia",
    gene      = "3d",
    direction = "promotes",
    timepoint = "1hr", # "1hr", "4hr", "24hr", "alltimepoints", or "1and4hr"
    lfc       = "0.5",
    feature_set = "codon_optimality", #targeted, combined, structure, clip_split, te_lfc, codon_optimality, or kozak
    clip_mode   = "aggregate" # "individual", "aggregate", or "both"
  )
}

# Derived file paths and labels from params
geneset_csv <- if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  paste0("translation_categories_si", params$gene, "_vs_sictrl_", params$condition, "_", params$timepoint, ".csv")
} else {
  paste0(params$condition, "_", params$gene, "_", params$direction,
         "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")
}
feature_matrix_rds <- if (params$feature_set == "te_lfc") {
  "feature_matrix_te_lfc.rds"
} else if (params$feature_set == "codon_optimality") {
  "feature_matrix_codon_optimality.rds"
} else if (params$feature_set == "kozak") {
  "feature_matrix_kozak.rds"
} else if (params$feature_set == "clip_split") {
  paste0("feature_matrix_", params$timepoint, "_clip_split.rds")
} else if (params$feature_set == "structure") {
  paste0("feature_matrix_", params$timepoint, "_structure.rds")
} else {
  paste0("feature_matrix_", params$timepoint, "_combined.rds")
}
suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                 params$condition, "_", params$timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")
pos_label <- params$direction
neg_label <- paste0("does_not_", sub("s$", "", params$direction))
gene_label <- paste0("eIF", params$gene)

cat("Gene set CSV:", geneset_csv, "\n")
cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix:", suffix, "\n")
cat("Positive label:", pos_label, "\n")
cat("Negative label:", neg_label, "\n")
```

# Overview

Binary classification without Boruta feature selection: does a gene's mRNA sequence predict whether `r gene_label` `r params$direction` its translation?

- **Positive class ("`r pos_label`")**: Genes from `` `r geneset_csv` ``
- **Negative class ("`r neg_label`")**: Genes from `negative_control_genes.csv` (not regulated by `r gene_label` in any condition/timepoint)

## Parameters

| Parameter | Values | Description |
|-----------|--------|-------------|
| `condition` | `"hypoxia"`, `"normoxia"` | Experimental condition |
| `gene` | `"3d"`, `"3e"` | eIF3 subunit |
| `direction` | `"promotes"`, `"inhibits"` | Direction of TE regulation |
| `timepoint` | `"1hr"`, `"4hr"`, `"24hr"`, `"alltimepoints"` | Timepoint; `"alltimepoints"` uses the time-regressed normoxia model (si3d vs sictrl, all 9 samples) |
| `lfc` | `"0.5"`, `"1"` | Log2 fold change threshold for gene set |
| `feature_set` | `"targeted"`, `"combined"`, `"structure"`, `"clip_split"`, `"te_lfc"`, `"codon_optimality"` | `"targeted"` = combined minus k-mers & RBP motifs (biological + lengths + GC + ORFquant + literature signatures); `"combined"` = all features; `"structure"` = targeted + RNA secondary structure features (RNAplfold accessibility, RNAfold MFE, dot-bracket parsing); `"clip_split"` = structure + split CLIP aggregate into eIF3 vs cap-binding counts; `"te_lfc"` = clip_split + sictrl TE LFCs at all condition × timepoint combinations (from condition_level_te.csv) + delta TE (hypoxia vs normoxia) at each timepoint (from translation_categories); `"codon_optimality"` = te_lfc + CAI, tAI, CSC codon optimality scores (from 01g) |
| `clip_mode` | `"individual"`, `"aggregate"`, `"both"` | Which CLIP features to use: `"individual"` = per-dataset regional binary features (~21 cols: 5utr/cds/3utr × 7 datasets); `"aggregate"` = single numeric count of datasets with any binding evidence (0-7); `"both"` = all CLIP features together |

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)

```

# Load Data

```{r load_data}
# Feature matrix
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

# Positive class
if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  positive_genes <- read_csv(here("output", geneset_csv)) %>%
    filter(te_lfc > as.numeric(params$lfc), te_padj < 0.05) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
} else {
  positive_genes <- read_csv(here("output", "genesets", geneset_csv))
}
cat("Positive genes:", nrow(positive_genes), "\n")

# Negative class: genes not regulated by eIF3d in any condition
negative_control_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
cat("Negative control genes:", nrow(negative_control_genes), "\n")
```

# Create Binary Target

```{r create_target}
set.seed(9)

# Only consider genes present in the feature matrix
pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

# Label genes
features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

nrow(model_genes)

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
cat("Imbalance ratio (neg:pos):",
    round(sum(model_genes$eif3d_target == neg_label) /
          sum(model_genes$eif3d_target == pos_label), 1), ":1\n")
```

# Preprocessing

## Filter Features (No Boruta)

```{r filter_features}
# Identify feature columns (exclude ID/target cols) — same logic as 06
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count (all combined):", length(feature_cols), "\n")

# ---- CLIP feature mode ----
if (params$clip_mode == "aggregate") {
  aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                           "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                           "clip_lee_glu_vs_cm_enrichment")
  regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
  feature_cols <- setdiff(feature_cols, regional_clip)
  cat("CLIP mode: aggregate — using aggregate CLIP counts only, dropped", length(regional_clip), "regional features\n")
} else if (params$clip_mode == "individual") {
  feature_cols <- setdiff(feature_cols, c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                                          "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                                          "clip_lee_glu_vs_cm_enrichment"))
  cat("CLIP mode: individual — using regional binary features, dropped aggregate CLIP counts\n")
} else {
  cat("CLIP mode: both — using all CLIP features\n")
}

# If targeted: remove k-mer and RBP motif columns
if (params$feature_set %in% c("targeted", "structure", "clip_split", "te_lfc", "codon_optimality", "kozak")) {
  kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
  excluded <- feature_cols[str_detect(feature_cols, kmer_motif_pattern)]
  feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]
  cat("Targeted mode: removed", length(excluded), "k-mer/motif columns\n")
}
cat("Feature count after filtering:", length(feature_cols), "\n")
```

## Median-Impute Missing Values

```{r impute_missing}
missing_counts <- model_genes %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:", nrow(missing_counts), "\n")
if (nrow(missing_counts) > 0) print(missing_counts)
cat("(Imputation deferred to after train/test split — medians computed from training set only)\n")
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
cat("\nClass distribution:\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif3d_target))
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif3d_target))

# Compute medians from training set only, then apply to both splits
train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

cat("\nRemaining NAs in train:", sum(is.na(train_data[, feature_cols])), "\n")
cat("Remaining NAs in test:", sum(is.na(test_data[, feature_cols])), "\n")
```

# Train Random Forest

```{r train_rf}
# Classes are already balanced (neg sampled to match pos count), so no case weights needed.
set.seed(9)
rf_model <- ranger(
  x = train_data[, feature_cols],
  y = train_data$eif3d_target,
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  seed = 9
)

print(rf_model)
```

# Evaluate on Test Set

## ROC Curve and AUC

```{r roc_curve, fig.width=8, fig.height=6}
# Predictions
pred_probs <- predict(rf_model, test_data[, feature_cols])$predictions
pred_pos_prob <- pred_probs[, pos_label]
actual <- test_data$eif3d_target

# ROC on test set (threshold-free AUC)
roc_obj <- roc(response = actual, predictor = pred_pos_prob,
               levels = c(neg_label, pos_label), direction = "<")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

# Youden threshold from OOB predictions on training set (avoids test set snooping)
oob_probs <- rf_model$predictions[, pos_label]
roc_oob <- roc(response = train_data$eif3d_target, predictor = oob_probs,
               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
coords_youden <- coords(roc_oob, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")
if (nrow(coords_youden) > 1) coords_youden <- coords_youden[1, ]  # take first if ties
optimal_threshold <- coords_youden$threshold

cat("Youden threshold (from training OOB):", round(optimal_threshold, 4), "\n")
cat("  OOB Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  OOB Specificity:", round(coords_youden$specificity, 3), "\n")

# Plot
roc_df <- data.frame(sensitivity = roc_obj$sensitivities,
                     specificity = roc_obj$specificities)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = paste0("ROC Curve: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = paste0("AUC = ", round(auc_value, 3),
                      " | Threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

print(p_roc)

ggsave(here("plots", paste0("rf_classification_roc_", suffix, ".pdf")),
       plot = p_roc, width = 8, height = 6)
```

## Confusion Matrix

```{r confusion_matrix}
pred_class <- factor(
  ifelse(pred_pos_prob >= optimal_threshold, pos_label, neg_label),
  levels = c(neg_label, pos_label)
)

cm <- confusionMatrix(pred_class, actual, positive = pos_label)
print(cm)

# Extract metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\nKey Metrics (Youden Threshold):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n")
```

## Confusion Matrix Plot

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = paste0("Confusion Matrix (", tools::toTitleCase(params$condition), ")"),
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3),
                      ", F1 = ", round(f1, 3))
  ) +
  theme(legend.position = "none")

print(p_cm)

ggsave(here("plots", paste0("rf_classification_confusion_matrix_", suffix, ".pdf")),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(train_data$eif3d_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- train_data[-test_idx, ]
  test_cv <- train_data[test_idx, ]

  # Impute using this fold's training data medians only
  cv_medians <- train_cv %>%
    dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~median(., na.rm = TRUE)))
  train_cv <- impute_with_medians(train_cv, cv_medians)
  test_cv  <- impute_with_medians(test_cv,  cv_medians)

  # Train (classes already balanced, no case weights needed)
  rf_cv <- ranger(
    x = train_cv[, feature_cols],
    y = train_cv$eif3d_target,
    num.trees = 500,
    probability = TRUE,
    seed = 9
  )

  # Youden threshold from OOB predictions on this fold's training data
  oob_probs_cv <- rf_cv$predictions[, pos_label]
  roc_oob_cv <- roc(response = train_cv$eif3d_target, predictor = oob_probs_cv,
                    levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  coords_cv <- coords(roc_oob_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                      best.method = "youden")
  if (nrow(coords_cv) > 1) coords_cv <- coords_cv[1, ]  # take first if ties

  # Predict on held-out fold
  pred_probs_cv <- predict(rf_cv, test_cv[, feature_cols])$predictions[, pos_label]

  # ROC on test fold (for AUC)
  roc_cv <- roc(test_cv$eif3d_target, pred_probs_cv,
                levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)

  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= coords_cv$threshold, pos_label, neg_label),
    levels = c(neg_label, pos_label)
  )
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif3d_target, positive = pos_label)

  data.frame(
    fold = i,
    Threshold = coords_cv$threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n5-Fold CV Results:\n")
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("AUC", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}

 importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Clean names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

# Bar plot
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top 30 Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", paste0("rf_classification_feature_importance_", suffix, ".pdf")),
       plot = p_importance, width = 10, height = 10)

# All-features bar plot (cap at top 60 to avoid oversized PDFs)
importance_all_plot <- importance_df %>% head(40)
all_height <- max(6, nrow(importance_all_plot) * 0.35)

p_importance_all <- importance_all_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top ", nrow(importance_all_plot), " Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_all)

ggsave(here("plots", paste0("rf_classification_feature_importance_all_", suffix, ".pdf")),
       plot = p_importance_all, width = 10, height = all_height)

# Custom-features-only bar plot (exclude k-mer and RBP motif features, cap at top 60)
importance_custom <- importance_df %>%
  filter(!str_detect(feature, "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbp_|^rbns_|^cisbp_")) %>%
  head(40)

custom_height <- max(6, nrow(importance_custom) * 0.35)

p_importance_custom <- importance_custom %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Custom Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Impurity) — CLIP, 4E/DAP5, Signatures, Lengths, GC",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_custom)

ggsave(here("plots", paste0("rf_classification_feature_importance_custom_", suffix, ".pdf")),
       plot = p_importance_custom, width = 10, height = custom_height)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^struct_")                          ~ "RNA Structure",
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      str_detect(feature, "^indiv_te_|^delta_te_")            ~ "Translational Efficiency",
      str_detect(feature, "^cai$|^tai$|^csc$")               ~ "Codon Optimality",
      TRUE                                                    ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    top_feature = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nImportance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance by Category (", tools::toTitleCase(params$condition), ")"),
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", paste0("rf_classification_importance_by_category_", suffix, ".pdf")),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Model
saveRDS(rf_model, file.path(output_dir, paste0("rf_classifier_", suffix, ".rds")))

# Feature importance
write_csv(importance_df, file.path(output_dir, paste0("rf_classification_importance_", suffix, ".csv")))

# Performance summary
n_pos_total <- sum(model_genes$eif3d_target == pos_label)
n_neg_total <- sum(model_genes$eif3d_target == neg_label)

performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1", "N_pos", "N_neg"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy,
               sensitivity, specificity, precision, f1,
               n_pos_total, n_neg_total),
  CV_Mean = c(mean(cv_results_df$AUC), NA, mean(cv_results_df$Balanced_Accuracy),
              mean(cv_results_df$Sensitivity), mean(cv_results_df$Specificity),
              mean(cv_results_df$Precision, na.rm = TRUE),
              mean(cv_results_df$F1, na.rm = TRUE), NA, NA),
  CV_SD = c(sd(cv_results_df$AUC), NA, sd(cv_results_df$Balanced_Accuracy),
            sd(cv_results_df$Sensitivity), sd(cv_results_df$Specificity),
            sd(cv_results_df$Precision, na.rm = TRUE),
            sd(cv_results_df$F1, na.rm = TRUE), NA, NA)
)
write_csv(performance_df, file.path(output_dir, paste0("rf_classification_performance_", suffix, ".csv")))

# CV results
write_csv(cv_results_df, file.path(output_dir, paste0("rf_classification_cv_results_", suffix, ".csv")))

# Predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_pos = pred_pos_prob,
  prob_neg = pred_probs[, neg_label]
)
write_csv(predictions_export, file.path(output_dir, paste0("rf_classification_predictions_", suffix, ".csv")))

cat("Exported: model RDS, importance CSV, performance CSV, CV results CSV, predictions CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat(paste0("CLASSIFICATION SUMMARY (", tools::toTitleCase(params$condition), ")\n"))
cat("========================================\n\n")

cat("Parameters:\n")
cat("  Condition:", params$condition, "\n")
cat("  Gene:", paste0("eIF", params$gene), "\n")
cat("  Direction:", params$direction, "\n")
cat("  Timepoint:", params$timepoint, "\n")
cat("  LFC threshold:", params$lfc, "\n")
cat("  Feature set:", params$feature_set, "(no Boruta selection)\n")
cat("  CLIP mode:", params$clip_mode, "\n\n")

cat("Classes:\n")
cat("  Positive (", pos_label, "):", sum(model_data$eif3d_target == pos_label), "\n")
cat("  Negative (", neg_label, "):", sum(model_data$eif3d_target == neg_label), "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("Test Set Performance (Youden threshold =", round(optimal_threshold, 4), "):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n\n")

cat("5-Fold CV:\n")
cat("  AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-",
    round(sd(cv_results_df$Balanced_Accuracy), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

cat("========================================\n")
```

# Final Feature Importance Plot

```{r final_importance_plot, fig.width=8, fig.height=max(4, min(40, length(feature_cols)) * 0.35)}
# Lollipop plot of top features (cap at 40)
importance_final_plot <- importance_df %>% head(40)
final_height <- max(4, nrow(importance_final_plot) * 0.35)

p_final_importance <- importance_final_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_segment(aes(xend = reorder(feature_clean, importance), y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(size = 3, color = "#0072B2") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance: ", gene_label, " Target Classification"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | LFC ", params$lfc, " | AUC = ", round(as.numeric(auc_value), 3)),
    x = "",
    y = "Variable Importance (Impurity)"
  )

print(p_final_importance)

ggsave(here("plots", paste0("rf_final_importance_", suffix, ".pdf")),
       plot = p_final_importance, width = 8,
       height = final_height)
```

# Session Info

```{r session_info}
sessionInfo()
```
