---
title: "`r paste0('Random Forest Permutation Importance: eIF', params$gene, ' Target Prediction (', tools::toTitleCase(params$condition), ', ', params$direction, ', ', params$timepoint, ', LFC ', params$lfc, ', ', params$feature_set, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "external_stability"
  clip_mode: "aggregate"
  exclude_4e_features: false
  exclude_own_te_features: true
  neg_geneset_csv: ""
  feature_matrix_path: ""
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    condition = "normoxia",
    gene      = "3d",
    direction = "promotes",
    timepoint = "1and4hr",
    lfc       = "0.5",
    feature_set = "external_stability",
    clip_mode   = "aggregate",
    exclude_4e_features     = FALSE,
    exclude_own_te_features = TRUE,
    neg_geneset_csv = "",
    feature_matrix_path = ""
  )
}

geneset_csv <- if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  paste0("translation_categories_si", params$gene, "_vs_sictrl_", params$condition, "_", params$timepoint, ".csv")
} else {
  paste0(params$condition, "_", params$gene, "_", params$direction,
         "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")
}
feature_matrix_rds <- if (params$feature_set == "te_lfc") {
  "feature_matrix_te_lfc.rds"
} else if (params$feature_set == "codon_optimality") {
  "feature_matrix_codon_optimality.rds"
} else if (params$feature_set == "kozak") {
  "feature_matrix_kozak.rds"
} else if (params$feature_set == "cnot3") {
  "feature_matrix_cnot3.rds"
} else if (params$feature_set == "external_stability") {
  "feature_matrix_external_stability.rds"
} else if (params$feature_set == "codon_features") {
  "feature_matrix_codon_features.rds"
} else if (params$feature_set == "codon_only") {
  "feature_matrix_codon_only.rds"
} else if (params$feature_set == "clip_split") {
  paste0("feature_matrix_", params$timepoint, "_clip_split.rds")
} else if (params$feature_set == "structure") {
  paste0("feature_matrix_", params$timepoint, "_structure.rds")
} else {
  paste0("feature_matrix_", params$timepoint, "_combined.rds")
}
if (!is.null(params$feature_matrix_path) && nchar(params$feature_matrix_path) > 0) {
  feature_matrix_rds <- params$feature_matrix_path
}
suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                 params$condition, "_", params$timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta",
                 if (isTRUE(params$exclude_4e_features)) "_no4e" else "",
                 if (isTRUE(params$exclude_own_te_features)) "_noOwnTE" else "")
suffix_perm <- paste0(suffix, "_perm")

pos_label  <- params$direction
neg_label  <- paste0("does_not_", sub("s$", "", params$direction))
gene_label <- paste0("eIF", params$gene)

cat("Gene set CSV:", geneset_csv, "\n")
cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix (perm):", suffix_perm, "\n")
```

# Overview

Permutation importance validation for the eIF3d/3e RF classifier. A second `ranger` model is trained on identical data (same split, same seed as notebook 07) using `importance = "permutation"` (mean decrease in OOB prediction accuracy when each feature is shuffled). This directly confirms whether the codon features that rank highly under Gini impurity also rank highly under a less biased importance measure.

- **Positive class ("`r pos_label`")**: `` `r geneset_csv` ``
- **Negative class ("`r neg_label`")**: `negative_control_genes.csv` (or custom via `neg_geneset_csv`)

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  positive_genes <- read_csv(here("output", geneset_csv)) %>%
    filter(te_lfc > as.numeric(params$lfc), te_padj < 0.05) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
} else {
  positive_genes <- read_csv(here("output", "genesets", geneset_csv))
}
cat("Positive genes:", nrow(positive_genes), "\n")

if (params$neg_geneset_csv != "") {
  negative_control_genes <- read_csv(
    here("output", "genesets", params$neg_geneset_csv),
    show_col_types = FALSE
  ) %>%
    mutate(gene_id_clean = ensembl_gene)
} else {
  neg_csv <- if (params$gene == "3e") {
    "negative_control_genes_si3e.csv"
  } else {
    "negative_control_genes.csv"
  }
  negative_control_genes <- read_csv(
    here("output", "predictive_modeling", neg_csv),
    show_col_types = FALSE
  ) %>%
    mutate(gene_id_clean = sub("\\..*", "", gene_id))
}
cat("Negative genes:", nrow(negative_control_genes), "\n")
```

# Create Binary Target

```{r create_target}
set.seed(9)

pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
```

# Preprocessing

```{r filter_features}
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count:", length(feature_cols), "\n")

if (params$clip_mode == "aggregate") {
  aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                           "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                           "clip_lee_glu_vs_cm_enrichment")
  regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
  feature_cols <- setdiff(feature_cols, regional_clip)
  cat("CLIP mode: aggregate — dropped", length(regional_clip), "regional features\n")
} else if (params$clip_mode == "individual") {
  feature_cols <- setdiff(feature_cols, c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                                          "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                                          "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment"))
  cat("CLIP mode: individual\n")
} else {
  cat("CLIP mode: both\n")
}

if (params$feature_set %in% c("targeted", "structure", "clip_split", "te_lfc", "codon_optimality",
                               "kozak", "cnot3", "external_stability", "codon_features")) {
  kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
  excluded <- feature_cols[str_detect(feature_cols, kmer_motif_pattern)]
  feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]
  cat("Targeted mode: removed", length(excluded), "k-mer/motif columns\n")
}

if (params$feature_set == "external_stability") {
  codon_detail_cols <- feature_cols[str_detect(feature_cols, "^bg_rscu_|^codon_freq_")]
  if (length(codon_detail_cols) > 0) {
    feature_cols <- setdiff(feature_cols, codon_detail_cols)
    cat("external_stability mode: removed", length(codon_detail_cols), "bg_rscu_/codon_freq_ columns\n")
  }
}

if (isTRUE(params$exclude_4e_features)) {
  circular_regex <- "^4e_resistance|^eif4e_ip"
  circular_lit_exact <- c(
    "lit_eIF4E_activated_translation",
    "lit_phospho_eIF4E_activated_translation",
    "lit_phospho_eIF4E_suppressed_translation",
    "lit_translation_is_sensitive_to_4E_inhibition",
    "lit_translation_is_resistant_to_4E_inhibition"
  )
  circular_cols <- feature_cols[str_detect(feature_cols, circular_regex) |
                                feature_cols %in% circular_lit_exact]
  if (length(circular_cols) > 0) {
    feature_cols <- setdiff(feature_cols, circular_cols)
    cat("Removed", length(circular_cols), "circular eIF4E features\n")
  }
}

if (isTRUE(params$exclude_own_te_features)) {
  own_te_cols <- feature_cols[str_detect(feature_cols, "^indiv_te_|^delta_te_|^sig_")]
  if (length(own_te_cols) > 0) {
    feature_cols <- setdiff(feature_cols, own_te_cols)
    cat("Removed", length(own_te_cols), "own-data TE features\n")
  }
}

raw_length_cols <- feature_cols[str_detect(feature_cols, "^(utr5|utr3|cds|total)_length$")]
if (length(raw_length_cols) > 0) {
  feature_cols <- setdiff(feature_cols, raw_length_cols)
  cat("Removed", length(raw_length_cols), "raw length columns:", paste(raw_length_cols, collapse = ", "), "\n")
}

sparsity_threshold <- 0.5
sparse_cols <- feature_cols[
  sapply(feature_cols, function(f) mean(!is.na(unlist(model_genes[[f]]))) < sparsity_threshold)
]
if (length(sparse_cols) > 0) {
  feature_cols <- setdiff(feature_cols, sparse_cols)
  cat("Removed", length(sparse_cols), "features with <", sparsity_threshold * 100, "% non-NA values:\n")
  cat(paste(" ", sparse_cols, collapse = "\n"), "\n")
}

cat("Feature count after filtering:", length(feature_cols), "\n")

list_cols <- feature_cols[sapply(feature_cols, function(f) is.list(model_genes[[f]]))]
if (length(list_cols) > 0) {
  cat("Coercing", length(list_cols), "list columns to numeric\n")
  for (col in list_cols) model_genes[[col]] <- as.numeric(unlist(model_genes[[col]]))
}
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
cat("Test set:", nrow(test_data), "\n")

train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)
```

# Train Random Forest (Permutation Importance)

```{r train_rf_perm}
set.seed(9)
rf_model_perm <- ranger(
  x          = train_data[, feature_cols],
  y          = train_data$eif3d_target,
  num.trees  = 500,
  importance = "permutation",
  probability = TRUE,
  seed       = 9
)

print(rf_model_perm)
cat("OOB prediction error:", round(rf_model_perm$prediction.error, 4), "\n")
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
importance_df <- data.frame(
  feature    = names(rf_model_perm$variable.importance),
  importance = rf_model_perm$variable.importance
) %>%
  arrange(desc(importance))

importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%") %>%
      str_replace("^bg_rscu_at_mean$", "bg-RSCU: AT-codon mean") %>%
      str_replace("^bg_rscu_gc_mean$", "bg-RSCU: GC-codon mean") %>%
      str_replace("^bg_rscu_deviation$", "bg-RSCU: deviation from background") %>%
      str_replace("^bg_rscu_", "bg-RSCU: ") %>%
      str_replace("^codon_freq_", "Codon freq: ")
  )

cat("Top 30 Features (permutation importance):\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title    = paste0("Top 30 Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Permutation — mean decrease in OOB accuracy)",
    x = "",
    y = "Permutation Importance"
  )

print(p_importance)
ggsave(here("plots", paste0("rf_classification_feature_importance_", suffix_perm, ".pdf")),
       plot = p_importance, width = 10, height = 10)

importance_all_plot <- importance_df %>% head(40)
all_height <- max(6, nrow(importance_all_plot) * 0.35)

p_importance_all <- importance_all_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title    = paste0("Top ", nrow(importance_all_plot), " Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Permutation — mean decrease in OOB accuracy)",
    x = "",
    y = "Permutation Importance"
  )

print(p_importance_all)
ggsave(here("plots", paste0("rf_classification_feature_importance_all_", suffix_perm, ".pdf")),
       plot = p_importance_all, width = 10, height = all_height)

importance_custom <- importance_df %>%
  filter(!str_detect(feature, "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbp_|^rbns_|^cisbp_")) %>%
  head(40)
custom_height <- max(6, nrow(importance_custom) * 0.35)

p_importance_custom <- importance_custom %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title    = paste0("Custom Features: ", gene_label, " Target Classification (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Variable Importance (Permutation) — CLIP, 4E/DAP5, Signatures, Lengths, GC",
    x = "",
    y = "Permutation Importance"
  )

print(p_importance_custom)
ggsave(here("plots", paste0("rf_classification_feature_importance_custom_", suffix_perm, ".pdf")),
       plot = p_importance_custom, width = 10, height = custom_height)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^struct_")                          ~ "RNA Structure",
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      str_detect(feature, "^indiv_te_|^delta_te_")            ~ "Translational Efficiency",
      str_detect(feature, "^cai$|^tai$|^csc$")               ~ "Codon Optimality",
      str_detect(feature, "^bg_rscu_")                        ~ "Background RSCU",
      str_detect(feature, "^codon_freq_")                     ~ "Codon Frequency",
      TRUE                                                    ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features       = n(),
    total_importance = sum(importance),
    mean_importance  = mean(importance),
    top_feature      = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nPermutation Importance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance by Category (", tools::toTitleCase(params$condition), ")"),
    subtitle = "Total Permutation Importance",
    x = "",
    y = "Total Permutation Importance"
  )

print(p_category)
ggsave(here("plots", paste0("rf_classification_importance_by_category_", suffix_perm, ".pdf")),
       plot = p_category, width = 10, height = 6)
```

## Directionality & Signed-Importance Lollipop

```{r compute_directionality}
target_numeric <- as.numeric(model_genes$eif3d_target) - 1

directionality <- importance_df %>%
  rowwise() %>%
  mutate(
    mean_promotes         = mean(model_genes[[feature]][target_numeric == 1], na.rm = TRUE),
    mean_does_not_promote = mean(model_genes[[feature]][target_numeric == 0], na.rm = TRUE),
    diff        = mean_promotes - mean_does_not_promote,
    correlation = tryCatch(
      cor(model_genes[[feature]], target_numeric, use = "complete.obs"),
      error = function(e) NA_real_
    ),
    wilcox_p = tryCatch({
      wt <- wilcox.test(model_genes[[feature]][target_numeric == 1],
                        model_genes[[feature]][target_numeric == 0])
      wt$p.value
    }, error = function(e) NA_real_)
  ) %>%
  ungroup() %>%
  mutate(
    wilcox_p_adj = p.adjust(wilcox_p, method = "BH"),
    direction    = case_when(
      is.na(correlation) ~ NA_character_,
      correlation > 0    ~ "Higher in promotes",
      TRUE               ~ "Lower in promotes"
    )
  ) %>%
  arrange(desc(importance))

cat("Directionality computed for", nrow(directionality), "features\n")
cat("Higher in promotes:", sum(directionality$direction == "Higher in promotes", na.rm = TRUE), "\n")
cat("Lower in promotes: ", sum(directionality$direction == "Lower in promotes",  na.rm = TRUE), "\n")
```

```{r plot_lollipop, fig.width=10, fig.height=10}
lollipop_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  head(30) %>%
  mutate(
    signed_importance = importance * sign(correlation),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    )),
    feature_clean = fct_reorder(feature_clean, signed_importance)
  )

category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

p_lollipop <- ggdotchart(
  lollipop_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 4,
  xlab     = "",
  ylab     = "Signed Permutation Importance (positive = higher in promotes)",
  title    = "Top 30 Features: Signed Permutation Importance",
  subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                    " | ", params$timepoint, " | LFC ", params$lfc)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop)
ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix_perm, ".pdf")),
       plot = p_lollipop, width = 10, height = 10)
```

# Final Feature Importance Plot

```{r final_importance_plot, fig.width=8, fig.height=max(4, min(40, length(feature_cols)) * 0.35)}
importance_final_plot <- importance_df %>% head(40)
final_height <- max(4, nrow(importance_final_plot) * 0.35)

p_final_importance <- importance_final_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_segment(aes(xend = reorder(feature_clean, importance), y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(size = 3, color = "#0072B2") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title    = paste0("Feature Importance: ", gene_label, " Target Classification"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | LFC ", params$lfc, " | Permutation Importance"),
    x = "",
    y = "Permutation Importance"
  )

print(p_final_importance)
ggsave(here("plots", paste0("rf_final_importance_", suffix_perm, ".pdf")),
       plot = p_final_importance, width = 8, height = final_height)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

saveRDS(rf_model_perm,
        file.path(output_dir, paste0("rf_classifier_", suffix_perm, ".rds")))

write_csv(importance_df,
          file.path(output_dir, paste0("rf_classification_importance_", suffix_perm, ".csv")))

directionality_export <- directionality %>%
  mutate(signed_importance = importance * sign(correlation))
write_csv(directionality_export,
          file.path(output_dir, paste0("rf_classification_directionality_", suffix_perm, ".csv")))

cat("Exported: model RDS, importance CSV, directionality CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat(paste0("PERMUTATION IMPORTANCE SUMMARY (", tools::toTitleCase(params$condition), ")\n"))
cat("========================================\n\n")

cat("Parameters:\n")
cat("  Condition:", params$condition, "\n")
cat("  Gene:", gene_label, "\n")
cat("  Direction:", params$direction, "\n")
cat("  Timepoint:", params$timepoint, "\n")
cat("  LFC threshold:", params$lfc, "\n")
cat("  Feature set:", params$feature_set, "\n")
cat("  CLIP mode:", params$clip_mode, "\n\n")

cat("Classes:\n")
cat("  Positive (", pos_label, "):", sum(model_data$eif3d_target == pos_label), "\n")
cat("  Negative (", neg_label, "):", sum(model_data$eif3d_target == neg_label), "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("OOB prediction error:", round(rf_model_perm$prediction.error, 4), "\n\n")

cat("Top 10 Features (permutation importance):\n")
for (i in 1:10) {
  cat("  ", i, ". ", importance_df$feature[i],
      " (", round(importance_df$importance[i], 4), ")\n", sep = "")
}

cat("\nPermutation importance by category:\n")
print(category_summary %>% dplyr::select(category, n_features, total_importance, mean_importance))

cat("========================================\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
