---
title: "`r paste0('Random Forest SHAP Values: eIF', params$gene, ' Target Prediction (', tools::toTitleCase(params$condition), ', ', params$direction, ', ', params$timepoint, ', LFC ', params$lfc, ', ', params$feature_set, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "external_stability"
  clip_mode: "aggregate"
  exclude_4e_features: false
  exclude_own_te_features: true
  neg_geneset_csv: ""
  feature_matrix_path: ""
  waterfall_genes: ""
  run_interactions: true
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# SHAP (Shapley value) decomposition of the notebook 07 classifier.
# Params must match the notebook 07 run being explained — the suffix is built the
# same way, and the saved 07 model is loaded and asserted against (Section 2).
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    condition = "hypoxia",
    gene      = "3d",
    direction = "promotes",
    timepoint = "1hr",
    lfc       = "0.5",
    feature_set = "external_stability",
    clip_mode   = "aggregate",
    exclude_4e_features     = FALSE,
    exclude_own_te_features = TRUE,
    neg_geneset_csv = "",
    feature_matrix_path = "",
    waterfall_genes = "",  # "" = auto-select 1 confident TP, 1 borderline, 1 confident error.
                           # Otherwise a comma-separated list of gene symbols, e.g. "VEGFA,CA9"
    run_interactions = TRUE  # TreeSHAP interaction values scale with the number of features:
                             # ~10 min at 56 features on 200 genes, far longer at 209.
                             # Set FALSE for large feature sets (e.g. codon_features).
  )
}

geneset_csv <- if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  paste0("translation_categories_si", params$gene, "_vs_sictrl_", params$condition, "_", params$timepoint, ".csv")
} else {
  paste0(params$condition, "_", params$gene, "_", params$direction,
         "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")
}
feature_matrix_rds <- if (params$feature_set == "te_lfc") {
  "feature_matrix_te_lfc.rds"
} else if (params$feature_set == "codon_optimality") {
  "feature_matrix_codon_optimality.rds"
} else if (params$feature_set == "kozak") {
  "feature_matrix_kozak.rds"
} else if (params$feature_set == "cnot3") {
  "feature_matrix_cnot3.rds"
} else if (params$feature_set == "ripseq") {
  "feature_matrix_ripseq.rds"
} else if (params$feature_set == "external_stability") {
  "feature_matrix_external_stability.rds"
} else if (params$feature_set == "codon_features") {
  "feature_matrix_codon_features.rds"
} else if (params$feature_set == "codon_only") {
  "feature_matrix_codon_only.rds"
} else if (params$feature_set == "clip_split") {
  paste0("feature_matrix_", params$timepoint, "_clip_split.rds")
} else if (params$feature_set == "structure") {
  paste0("feature_matrix_", params$timepoint, "_structure.rds")
} else {
  paste0("feature_matrix_", params$timepoint, "_combined.rds")
}
if (!is.null(params$feature_matrix_path) && nchar(params$feature_matrix_path) > 0) {
  feature_matrix_rds <- params$feature_matrix_path
}
suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                 params$condition, "_", params$timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta",
                 if (isTRUE(params$exclude_4e_features)) "_no4e" else "",
                 if (isTRUE(params$exclude_own_te_features)) "_noOwnTE" else "")
suffix_shap <- paste0(suffix, "_shap")

pos_label  <- params$direction
neg_label  <- paste0("does_not_", sub("s$", "", params$direction))
gene_label <- paste0("eIF", params$gene)

cat("Gene set CSV:", geneset_csv, "\n")
cat("Feature matrix:", feature_matrix_rds, "\n")
cat("07 model suffix:", suffix, "\n")
cat("Output suffix (shap):", suffix_shap, "\n")
```

# Overview

Shapley (SHAP) decomposition of the notebook 07 Random Forest classifier. Where impurity
importance (07) and permutation importance (07b) give a single non-negative number per
feature, SHAP gives a **signed contribution per gene per feature**, additive to that gene's
predicted probability. That buys three things the other two cannot give:

1. **Direction and magnitude in one view** (beeswarm), without the univariate proxy that
   the 07 signed-importance lollipop uses.
2. **Exact grouping.** SHAP values are additive, so summing them within a feature family is
   an exact operation. This resolves credit-splitting across the correlated blocks in this
   feature set (`csc` / `csc_first75` / `csc_q1`-`csc_q4`, the eight `struct_accessibility_*`
   families, the length and GC families), which impurity importance structurally cannot do.
3. **Per-gene explanations** (waterfall), for interpreting individual candidate genes.

## TreeSHAP and the surrogate forest

`treeshap::ranger.unify()` does not support ranger **probability** forests (it errors on a
missing `pred.1` column). It does support a **regression** forest fit to the 0/1 target,
whose terminal-node predictions are the same class proportion a probability forest stores.

So this notebook fits a surrogate regression forest on the identical training rows, with
`mtry`, `min.node.size` and `num.trees` **read off the saved 07 model** rather than assumed,
and then verifies the surrogate reproduces 07's predictions before explaining it
(Section "Surrogate fidelity"). Gini splitting on a binary target and variance splitting on
the same target as 0/1 select the same splits up to tie-breaking, so the two forests agree
to ~7e-3 in predicted probability but are not bit-identical.

- **Positive class ("`r pos_label`")**: `` `r geneset_csv` ``
- **Negative class ("`r neg_label`")**: `negative_control_genes.csv` (or custom via `neg_geneset_csv`)

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(treeshap)
library(ggbeeswarm)
```

```{r table_helper}
# Summary tables are printed with knitr::kable(format = "simple") so they render as aligned
# tables in the console when chunks are run interactively, and identically in the knit HTML.
# Plain print(as.data.frame(x)) wraps wide tables across several blocks and is hard to read.
print_table <- function(df, digits = 4, caption = NULL) {
  print(knitr::kable(as.data.frame(df), format = "simple", digits = digits, caption = caption))
  invisible(df)
}
```

# Load Data

```{r load_data}
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  positive_genes <- read_csv(here("output", geneset_csv)) %>%
    filter(te_lfc > as.numeric(params$lfc), te_padj < 0.05) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
} else {
  positive_genes <- read_csv(here("output", "genesets", geneset_csv))
}
cat("Positive genes:", nrow(positive_genes), "\n")

if (params$neg_geneset_csv != "") {
  negative_control_genes <- read_csv(
    here("output", "genesets", params$neg_geneset_csv),
    show_col_types = FALSE
  ) %>%
    mutate(gene_id_clean = ensembl_gene)
} else {
  neg_csv <- if (params$gene == "3e") {
    "negative_control_genes_si3e.csv"
  } else {
    "negative_control_genes.csv"
  }
  negative_control_genes <- read_csv(
    here("output", "predictive_modeling", neg_csv),
    show_col_types = FALSE
  ) %>%
    mutate(gene_id_clean = sub("\\..*", "", gene_id))
}
cat("Negative genes:", nrow(negative_control_genes), "\n")
```

# Create Binary Target

Identical to notebook 07: same seed, same sampling, so the labelled set matches that run.

```{r create_target}
set.seed(9)

pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
```

# Preprocessing

```{r filter_features}
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count:", length(feature_cols), "\n")

if (params$clip_mode == "aggregate") {
  aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                           "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                           "clip_lee_glu_vs_cm_enrichment")
  regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
  feature_cols <- setdiff(feature_cols, regional_clip)
  cat("CLIP mode: aggregate — dropped", length(regional_clip), "regional features\n")
} else if (params$clip_mode == "individual") {
  feature_cols <- setdiff(feature_cols, c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                                          "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                                          "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment"))
  cat("CLIP mode: individual\n")
} else {
  cat("CLIP mode: both\n")
}

if (params$feature_set %in% c("targeted", "structure", "clip_split", "te_lfc", "codon_optimality",
                               "kozak", "cnot3", "external_stability", "codon_features",
                               "positional_gc")) {
  kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
  excluded <- feature_cols[str_detect(feature_cols, kmer_motif_pattern)]
  feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]
  cat("Targeted mode: removed", length(excluded), "k-mer/motif columns\n")
}

if (params$feature_set == "external_stability") {
  codon_detail_cols <- feature_cols[str_detect(feature_cols, "^bg_rscu_|^codon_freq_")]
  if (length(codon_detail_cols) > 0) {
    feature_cols <- setdiff(feature_cols, codon_detail_cols)
    cat("external_stability mode: removed", length(codon_detail_cols), "bg_rscu_/codon_freq_ columns\n")
  }
}

if (isTRUE(params$exclude_4e_features)) {
  circular_regex <- "^4e_resistance|^eif4e_ip"
  circular_lit_exact <- c(
    "lit_eIF4E_activated_translation",
    "lit_phospho_eIF4E_activated_translation",
    "lit_phospho_eIF4E_suppressed_translation",
    "lit_translation_is_sensitive_to_4E_inhibition",
    "lit_translation_is_resistant_to_4E_inhibition"
  )
  circular_cols <- feature_cols[str_detect(feature_cols, circular_regex) |
                                feature_cols %in% circular_lit_exact]
  if (length(circular_cols) > 0) {
    feature_cols <- setdiff(feature_cols, circular_cols)
    cat("Removed", length(circular_cols), "circular eIF4E features\n")
  }
}

if (isTRUE(params$exclude_own_te_features)) {
  own_te_cols <- feature_cols[str_detect(feature_cols, "^indiv_te_|^delta_te_|^sig_")]
  if (length(own_te_cols) > 0) {
    feature_cols <- setdiff(feature_cols, own_te_cols)
    cat("Removed", length(own_te_cols), "own-data TE features\n")
  }
}

raw_length_cols <- feature_cols[str_detect(feature_cols, "^(utr5|utr3|cds|total)_length$")]
if (length(raw_length_cols) > 0) {
  feature_cols <- setdiff(feature_cols, raw_length_cols)
  cat("Removed", length(raw_length_cols), "raw length columns:", paste(raw_length_cols, collapse = ", "), "\n")
}

sparsity_threshold <- 0.5
sparse_cols <- feature_cols[
  sapply(feature_cols, function(f) mean(!is.na(unlist(model_genes[[f]]))) < sparsity_threshold)
]
if (length(sparse_cols) > 0) {
  feature_cols <- setdiff(feature_cols, sparse_cols)
  cat("Removed", length(sparse_cols), "features with <", sparsity_threshold * 100, "% non-NA values\n")
}

cat("Feature count after filtering:", length(feature_cols), "\n")

list_cols <- feature_cols[sapply(feature_cols, function(f) is.list(model_genes[[f]]))]
if (length(list_cols) > 0) {
  cat("Coercing", length(list_cols), "list columns to numeric\n")
  for (col in list_cols) model_genes[[col]] <- as.numeric(unlist(model_genes[[col]]))
}

# Leakage guard: the target must never appear among the predictors
stopifnot("target column leaked into feature_cols" = !"eif3d_target" %in% feature_cols)
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

# createDataPartition stratifies by the outcome, so both splits keep the 50/50 balance of
# the labelled set. Printed rather than assumed: the SHAP baseline is the TRAINING base
# rate, so if the test set's balance ever drifted from it, the beeswarm cloud would sit
# off-centre for reasons unrelated to any feature.
cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif3d_target))
cat("  positive fraction:", round(mean(train_data$eif3d_target == pos_label), 4), "\n")
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif3d_target))
cat("  positive fraction:", round(mean(test_data$eif3d_target == pos_label), 4), "\n")

train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

# Full labelled set, imputed with the same training medians (used for export only)
full_data <- impute_with_medians(model_data, train_medians)

# Pre-imputation training rows: NAs here identify which values were median-filled.
# Defined in this ungated chunk because several independently-gated figure chunks use it.
pre_impute_train <- model_data[train_idx, ]

cat("Remaining NAs in train:", sum(is.na(train_data[, feature_cols])), "\n")
cat("Remaining NAs in test:",  sum(is.na(test_data[, feature_cols])),  "\n")
```

# Fit the SHAP-Compatible Surrogate Forest

The surrogate is a regression forest on the 0/1 target. Its hyperparameters are **read off
the saved notebook 07 probability forest** rather than hardcoded, because ranger's defaults
differ by tree type (`min.node.size` is 10 for probability forests but 1 for classification
and 5 for regression; `mtry` is `floor(sqrt(p))` for probability but `floor(p/3)` for
regression). Hardcoding either would silently fit a differently-shaped forest and the SHAP
values would explain the wrong model.

```{r fit_surrogate}
rf_prob_path <- here("output", "predictive_modeling", paste0("rf_classifier_", suffix, ".rds"))
stopifnot("notebook 07 model RDS not found - knit 07 with matching params first" =
            file.exists(rf_prob_path))
rf_prob <- readRDS(rf_prob_path)

cat("Loaded 07 model:", basename(rf_prob_path), "\n")
cat("  treetype:", rf_prob$treetype, "| num.trees:", rf_prob$num.trees,
    "| mtry:", rf_prob$mtry, "| min.node.size:", rf_prob$min.node.size, "\n")

set.seed(9)
rf_shap <- ranger(
  x             = train_data[, feature_cols],
  y             = as.numeric(train_data$eif3d_target == pos_label),
  num.trees     = rf_prob$num.trees,
  mtry          = rf_prob$mtry,           # matches 07's probability-forest setting
  min.node.size = rf_prob$min.node.size,  # matches 07's probability-forest setting
  seed          = 9
)

# --- Check A: structural config match, asserted at the input -------------------
# Deterministic and independent of any prediction-agreement tolerance.
stopifnot(
  "07 model is not a probability forest"          = rf_prob$treetype == "Probability estimation",
  "surrogate is not a regression forest"          = rf_shap$treetype == "Regression",
  "surrogate mtry does not match 07"              = rf_shap$mtry == rf_prob$mtry,
  "surrogate min.node.size does not match 07"     = rf_shap$min.node.size == rf_prob$min.node.size,
  "surrogate num.trees does not match 07"         = rf_shap$num.trees == rf_prob$num.trees,
  "surrogate feature set/order does not match 07" =
    identical(rf_shap$forest$independent.variable.names,
              rf_prob$forest$independent.variable.names)
)
cat("Check A passed: surrogate config matches the 07 model structurally\n")
```

## Surrogate fidelity (test set)

**Check B**, complementary to Check A: Check A guards the configuration, this certifies the
resulting predictions. Both the gate and the reported number are computed on the **held-out
test set**, never on training rows, since training rows were seen by both forests and can
agree more tightly than held-out ones.

The `0.02` bound is calibrated, not arbitrary. Measured max |delta| on this dataset:
matched config `0.0068`; `min.node.size = 1` `0.0798`; `min.node.size = 5` `0.0863`;
`mtry = floor(p/3)` `0.163`. So `0.02` sits ~3x above a correct run and ~4x below the
nearest misconfiguration. The residual `0.0068` is gini-vs-variance split tie-breaking,
not a modelling difference.

```{r check_fidelity}
# Calibrated on the 56-feature external_stability set (see prose above). Other feature sets
# have not been calibrated against a deliberate misconfiguration, so for those this acts as a
# sanity bound rather than a tuned one.
fidelity_tol <- 0.02

p_prob_test <- predict(rf_prob, test_data[, feature_cols])$predictions[, pos_label]
p_reg_test  <- predict(rf_shap, test_data[, feature_cols])$predictions

max_abs_diff_test <- max(abs(p_reg_test - p_prob_test))
r_test <- cor(p_reg_test, p_prob_test)

auc_prob <- as.numeric(auc(roc(test_data$eif3d_target, p_prob_test,
                               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)))
auc_reg  <- as.numeric(auc(roc(test_data$eif3d_target, p_reg_test,
                               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)))

cat("Surrogate fidelity on TEST SET (n =", nrow(test_data), "):\n")
cat("  max |delta prob|:", format(max_abs_diff_test, digits = 3), "  <- gated\n")
cat("  Pearson r:", round(r_test, 6), "(context only)\n")
cat("  AUC 07 probability forest:", round(auc_prob, 5), "(context only)\n")
cat("  AUC surrogate regression forest:", round(auc_reg, 5), "(context only)\n")
cat("  AUC delta:", format(abs(auc_prob - auc_reg), digits = 3), "(context only)\n")

stopifnot("surrogate does not reproduce the 07 model on the test set" =
            max_abs_diff_test < fidelity_tol)
cat("Check B passed: surrogate reproduces 07 test-set predictions within", fidelity_tol, "\n")

# Reported separately, and deliberately not gated: training rows were seen by both
# forests during fitting, so this number is not comparable to the test-set figure above.
p_prob_full <- predict(rf_prob, full_data[, feature_cols])$predictions[, pos_label]
p_reg_full  <- predict(rf_shap, full_data[, feature_cols])$predictions
cat("\nFor reference only, ALL labelled genes (n =", nrow(full_data),
    ", includes training rows): max |delta prob| =",
    format(max(abs(p_reg_full - p_prob_full)), digits = 3), "\n")
```

# Compute Exact TreeSHAP

The unify data is the **training set**, so the SHAP baseline is the mean surrogate
prediction over training rows, i.e. the training base rate. This matters for reading
waterfall plots: the baseline anchor is a train-set number (~0.5 here, since classes are
balanced), not the test-set base rate.

```{r compute_shap}
unified <- ranger.unify(rf_shap, train_data[, feature_cols])

# Baseline the decomposition expands around: mean surrogate output over the unify data
shap_baseline <- mean(predict(rf_shap, train_data[, feature_cols])$predictions)
cat("SHAP baseline (mean surrogate prediction over training rows):",
    round(shap_baseline, 5), "\n")

# Test set: used for all rankings and figures (held out, so no memorisation artefacts)
cat("Computing TreeSHAP on test set (", nrow(test_data), "genes)...\n")
shap_test_obj <- treeshap(unified, test_data[, feature_cols], verbose = FALSE)
shap_test <- as.data.frame(shap_test_obj$shaps)[, feature_cols]

# Full labelled set: export only, so any gene can be looked up later
cat("Computing TreeSHAP on all labelled genes (", nrow(full_data), "genes)...\n")
shap_full_obj <- treeshap(unified, full_data[, feature_cols], verbose = FALSE)
shap_full <- as.data.frame(shap_full_obj$shaps)[, feature_cols]

stopifnot(
  "SHAP matrix dims do not match test set"  = all(dim(shap_test) == c(nrow(test_data), length(feature_cols))),
  "SHAP matrix dims do not match full set"  = all(dim(shap_full) == c(nrow(full_data), length(feature_cols))),
  "SHAP column names do not match features" = identical(colnames(shap_test), feature_cols),
  "test and full SHAP objects are the same size - one was overwritten" =
    nrow(shap_test) != nrow(shap_full)
)

# Every figure below is labelled "test set" and reports n as nrow(shap_test), read from the
# SHAP object itself rather than from test_data. A copy-paste that swapped in shap_full would
# therefore show n = 2872 in the caption instead of silently mislabelling 2872 genes as the
# test set. It is written inline in each figure chunk rather than stored in a variable here,
# so that chunks stay runnable individually in the console.
stopifnot("figure n does not match the test set" = nrow(shap_test) == nrow(test_data))
cat("Figures will report n =", nrow(shap_test), "genes (test set)\n")

# Local accuracy: SHAP values plus baseline must reconstruct the model output exactly
recon_test <- rowSums(shap_test) + shap_baseline
recon_full <- rowSums(shap_full) + shap_baseline
cat("Local accuracy, max |reconstruction error|:\n")
cat("  test set:", format(max(abs(recon_test - p_reg_test)), digits = 3), "\n")
cat("  full set:", format(max(abs(recon_full - p_reg_full)), digits = 3), "\n")
stopifnot(
  "SHAP local accuracy violated on test set" = max(abs(recon_test - p_reg_test)) < 1e-6,
  "SHAP local accuracy violated on full set" = max(abs(recon_full - p_reg_full)) < 1e-6
)
cat("Local accuracy check passed\n")
```

# Feature Categories and Display Names

Every feature is assigned to a named category. Notebook 07's own `case_when` drops
`is_4e_*`, `te_4e_resistance_lfc` and `is_dap5_*` into "Other"; that is not carried over
here, since an "Other" bucket in the grouped-SHAP panel would hide three named feature
families inside what reads as miscellany. The bucket is asserted empty rather than assumed
empty, and membership is printed below so the grouping is auditable.

```{r feature_categories_map}
categorize_feature <- function(f) {
  case_when(
    str_detect(f, "^gc[12]$|^gc3_residual$")               ~ "Positional GC",
    str_detect(f, "^bg_rscu_")                             ~ "Codon usage (RSCU)",
    str_detect(f, "^codon_freq_")                          ~ "Codon frequency",
    str_detect(f, "^log2_.*_length$")                      ~ "Region lengths",
    str_detect(f, "_gc$|^gc3$")                            ~ "GC content",
    str_detect(f, "^struct_") & str_detect(f, "cds")       ~ "RNA structure (CDS)",
    str_detect(f, "^struct_") & str_detect(f, "utr3")      ~ "RNA structure (3'UTR)",
    str_detect(f, "^struct_")                              ~ "RNA structure (5'UTR)",
    str_detect(f, "^clip_")                                ~ "eIF3 CLIP binding",
    str_detect(f, "^is_4e_|^te_4e_|^eif4e_ip_")            ~ "eIF4E sensitivity",
    str_detect(f, "^is_dap5_|^dap5_")                      ~ "DAP5 dependence",
    str_detect(f, "^kozak_")                               ~ "Kozak context",
    str_detect(f, "^cai$|^tai$|^csc|^fop$|weighted_codon") ~ "Codon optimality",
    str_detect(f, "^zhu2024_|^karner2026_")                ~ "mRNA stability",
    str_detect(f, "ribo_enrichment|_occupancy$")           ~ "Ribosome occupancy (external)",
    TRUE                                                   ~ "Other"
  )
}

feature_meta <- tibble(feature = feature_cols) %>%
  mutate(category = categorize_feature(feature))

# Category palette, defined here with the map so every plot that colours by category shares
# it. Deliberately avoids #E69F00 and #56B4E9, which the project reserves for eIF3d/si3d and
# eIF3e/si3e, so no category reads as one of those factors. The three RNA structure
# categories share a purple ramp so related groups look related.
category_colors <- c(
  "GC content"                    = "#0072B2",
  "Codon optimality"              = "#009E73",
  "Region lengths"                = "#CC79A7",
  "RNA structure (5'UTR)"         = "#C2A5CF",
  "RNA structure (CDS)"           = "#9970AB",
  "RNA structure (3'UTR)"         = "#762A83",
  "eIF4E sensitivity"             = "#D55E00",
  "DAP5 dependence"               = "#D668C6",
  "eIF3 CLIP binding"             = "#8C510A",
  "mRNA stability"                = "#35978F",
  "Ribosome occupancy (external)" = "#404040",
  "Kozak context"                 = "#999999",
  "Codon usage (RSCU)"            = "#66A61E",
  "Codon frequency"               = "#B2182B",
  "Positional GC"                 = "#3690C0"
)
stopifnot("a feature category has no assigned colour - extend category_colors" =
            all(unique(feature_meta$category) %in% names(category_colors)))

# Codon-position tests need per-codon features, which only some matrices carry.
run_codon_tests <- any(grepl("^bg_rscu_[A-Z]{3}$", feature_cols))

# Some entanglement diagnostics are written around a specific named trio. They are skipped
# (chunk eval = FALSE) when a feature matrix does not contain all three, so this notebook can
# be pointed at other matrices without editing. The GC3 screen needs only gc3 and runs either way.
pair_diag_feats <- c("cds_gc", "csc", "hia2026_dhx29_occupancy")
run_pair_diag   <- all(pair_diag_feats %in% feature_cols)
run_gc3_screen  <- "gc3" %in% feature_cols
run_interactions <- isTRUE(params$run_interactions) && run_pair_diag
cat("Entanglement diagnostics enabled?  pair diagnostics:", run_pair_diag,
    "| GC3 screen:", run_gc3_screen, "| interactions:", run_interactions, "\n")
if (!run_pair_diag)
  cat("  (missing:", paste(setdiff(pair_diag_feats, feature_cols), collapse = ", "), ")\n")

uncategorized <- feature_meta %>% filter(category == "Other") %>% pull(feature)
if (length(uncategorized) > 0) {
  cat("UNCATEGORIZED FEATURES:\n"); cat(paste(" ", uncategorized, collapse = "\n"), "\n")
}
stopifnot("features fell into the 'Other' bucket - extend categorize_feature()" =
            length(uncategorized) == 0)

cat("Category membership (", n_distinct(feature_meta$category), "categories,",
    nrow(feature_meta), "features):\n")
feature_meta %>%
  group_by(category) %>%
  summarise(n = n(), features = paste(feature, collapse = ", "), .groups = "drop") %>%
  arrange(desc(n)) %>%
  as.data.frame() %>%
  print(right = FALSE)
```

```{r feature_display_names}
clean_feature_name <- function(f) {
  f %>%
    str_replace("^log2_", "log2 ") %>%
    str_replace("_length$", " length") %>%
    str_replace("^utr5", "5'UTR") %>%
    str_replace("^utr3", "3'UTR") %>%
    str_replace("^cds", "CDS") %>%
    str_replace("_gc$", " GC%") %>%
    str_replace("^gc3$", "GC3") %>%
    str_replace("^struct_accessibility_", "Accessibility: ") %>%
    str_replace("^struct_num_structured_regions", "Structured regions") %>%
    str_replace("^clip_eif3_count", "CLIP: eIF3 datasets") %>%
    str_replace("^clip_cap_binding_count", "CLIP: cap-binding datasets") %>%
    str_replace("^cai$", "CAI") %>%
    str_replace("^tai$", "tAI") %>%
    str_replace("^csc$", "CSC (full CDS)") %>%
    str_replace("^fop$", "FOP") %>%
    str_replace("^csc_first75$", "CSC (first 75 codons)") %>%
    str_replace("^csc_q([1-4])$", "CSC (quarter \\1)") %>%
    str_replace("^kozak_", "Kozak: ") %>%
    str_replace("^zhu2024_", "Zhu 2024: ") %>%
    str_replace("^karner2026_", "Karner 2026: ") %>%
    str_replace("^hia2026_", "Hia 2026: ") %>%
    str_replace("^cnot3_", "CNOT3: ") %>%
    str_replace("^is_4e_", "4E-") %>%
    str_replace("^te_4e_resistance_lfc$", "4E resistance TE LFC") %>%
    str_replace("^eif4e_ip_", "eIF4E IP: ") %>%
    str_replace("^is_dap5_", "DAP5-") %>%
    str_replace("^dap5_polysome_lfc$", "DAP5 polysome LFC") %>%
    str_replace_all("_", " ")
}

feature_meta <- feature_meta %>% mutate(feature_clean = clean_feature_name(feature))
```

# SHAP Importance

```{r shap_importance}
# shap_value_cor is the directional summary: the correlation between a feature's VALUE and
# its SHAP value across genes. Positive = high feature values push toward the positive class.
# This is the correct directional quantity; mean_signed_shap is NOT a measure of direction
# (it is the average net push, which is near zero by construction and is dominated by the
# offset between the explain-set mean prediction and the training baseline). It is retained
# in the export for completeness but is deliberately not used for direction anywhere.
shap_value_cor <- map_dbl(feature_cols, function(f) {
  v <- test_data[[f]]; s <- shap_test[[f]]
  if (sd(v, na.rm = TRUE) == 0 || sd(s, na.rm = TRUE) == 0) return(NA_real_)
  cor(v, s, use = "complete.obs")
})

shap_importance <- tibble(
  feature          = feature_cols,
  mean_abs_shap    = colMeans(abs(shap_test)),
  shap_value_cor   = shap_value_cor,
  mean_signed_shap = colMeans(shap_test),
  sd_shap          = apply(shap_test, 2, sd)
) %>%
  left_join(feature_meta, by = "feature") %>%
  arrange(desc(mean_abs_shap)) %>%
  mutate(rank_shap = row_number())

cat("Top 20 features by mean |SHAP| (test set, n =", nrow(shap_test), "):\n")
print_table(shap_importance %>% head(20) %>%
              dplyr::select(rank_shap, feature, mean_abs_shap, shap_value_cor, category),
            caption = "Top 20 features by mean |SHAP|")
```

## Agreement with notebook 07 impurity importance

```{r compare_impurity}
imp07_path <- here("output", "predictive_modeling",
                   paste0("rf_classification_importance_", suffix, ".csv"))

if (file.exists(imp07_path)) {
  imp07 <- read_csv(imp07_path, show_col_types = FALSE) %>%
    dplyr::select(feature, impurity = importance) %>%
    mutate(rank_impurity = rank(-impurity))

  cmp <- shap_importance %>%
    dplyr::select(feature, feature_clean, category, mean_abs_shap, rank_shap) %>%
    inner_join(imp07, by = "feature")

  rho <- cor(cmp$mean_abs_shap, cmp$impurity, method = "spearman")
  cat("Spearman rho, mean |SHAP| vs 07 impurity importance:",
      round(rho, 3), ", n =", nrow(cmp), "\n")

  p_cmp <- ggplot(cmp, aes(x = impurity, y = mean_abs_shap)) +
    geom_point(color = "#E69F00", size = 2, alpha = 0.8) +
    ggrepel::geom_text_repel(
      data = cmp %>% head(10),
      aes(label = feature_clean), size = 3, max.overlaps = 20
    ) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
             label = paste0("Spearman rho = ", round(rho, 3), ", n = ", nrow(cmp))) +
    theme_classic(base_size = 12) +
    labs(
      title = paste0("SHAP vs impurity importance: ", gene_label, " targets"),
      subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                        " | SHAP on held-out test set"),
      x = "Impurity importance (notebook 07)",
      y = "Mean |SHAP| (test set)"
    )

  print(p_cmp)
  ggsave(here("plots", paste0("shap_vs_impurity_", suffix_shap, ".pdf")),
         plot = p_cmp, width = 7, height = 6)
} else {
  cat("07 importance CSV not found, skipping comparison:", basename(imp07_path), "\n")
  rho <- NA_real_
}
```

## Direction agreement with the 07 univariate directionality

Reported, **not asserted**. Direction here is `shap_value_cor` (feature value vs its own
SHAP value), the conditional multivariate analogue of the 07 `correlation` column, which is
univariate. They legitimately disagree where suppression or interaction effects are present,
which is expected given the correlated `csc*`, `struct_accessibility_*` and GC blocks. A flip
is a result to inspect, not a failure.

```{r compare_direction}
dir07_path <- here("output", "predictive_modeling",
                   paste0("rf_classification_directionality_", suffix, ".csv"))

if (file.exists(dir07_path)) {
  dir07 <- read_csv(dir07_path, show_col_types = FALSE) %>%
    dplyr::select(feature, correlation)

  sign_cmp <- shap_importance %>%
    dplyr::select(feature, feature_clean, shap_value_cor, rank_shap) %>%
    inner_join(dir07, by = "feature") %>%
    filter(!is.na(correlation), correlation != 0, !is.na(shap_value_cor), shap_value_cor != 0) %>%
    mutate(agrees = sign(shap_value_cor) == sign(correlation))

  top20 <- sign_cmp %>% arrange(rank_shap) %>% head(20)
  cat("Sign agreement (SHAP vs 07 univariate correlation):\n")
  cat("  all features: ", round(100 * mean(sign_cmp$agrees), 1), "% (n = ", nrow(sign_cmp), ")\n", sep = "")
  cat("  top 20 by |SHAP|: ", round(100 * mean(top20$agrees), 1), "% (n = ", nrow(top20), ")\n", sep = "")

  disagreements <- sign_cmp %>% filter(!agrees) %>% arrange(rank_shap)
  if (nrow(disagreements) > 0) {
    cat("\nFeatures where SHAP direction differs from univariate correlation",
        "(inspect for suppression / interaction effects):\n")
    disagreements %>%
      dplyr::select(rank_shap, feature, shap_value_cor, correlation) %>%
      as.data.frame() %>%
      print(right = FALSE)
  } else {
    cat("\nNo direction disagreements.\n")
  }
} else {
  cat("07 directionality CSV not found, skipping:", basename(dir07_path), "\n")
}
```

# Beeswarm Plot

Each point is one test-set gene. Position on x is that gene's SHAP value for the feature
(signed contribution to predicted probability). Color is the gene's **feature value**, which
is a low-to-high quantity with no meaningful midpoint, so the scale is sequential (viridis),
and is min-max scaled **within each feature** because the 56 features are on different units.

```{r beeswarm, fig.width=10, fig.height=10}
n_display <- 30
top_features <- shap_importance %>% head(n_display) %>% pull(feature)

# Winsorise at the 2nd/98th percentile before scaling so single outliers do not
# compress the colour range for the rest of the genes.
scale01 <- function(x) {
  lo <- quantile(x, 0.02, na.rm = TRUE); hi <- quantile(x, 0.98, na.rm = TRUE)
  if (!is.finite(lo) || !is.finite(hi) || hi <= lo) return(rep(0.5, length(x)))
  pmin(pmax((x - lo) / (hi - lo), 0), 1)
}

bee_df <- map_dfr(top_features, function(f) {
  tibble(
    feature      = f,
    shap         = shap_test[[f]],
    value_scaled = scale01(test_data[[f]])
  )
}) %>%
  left_join(shap_importance %>% dplyr::select(feature, feature_clean, mean_abs_shap),
            by = "feature") %>%
  mutate(feature_clean = fct_reorder(feature_clean, mean_abs_shap))

p_bee <- ggplot(bee_df, aes(x = shap, y = feature_clean, color = value_scaled)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_quasirandom(groupOnX = FALSE, size = 0.7, alpha = 0.6, width = 0.35) +
  scale_color_viridis_c(
    name   = "Feature value",
    breaks = c(0, 1), labels = c("Low", "High")
  ) +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("SHAP beeswarm, top ", n_display, " features: ", gene_label, " targets"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | test set, n = ", nrow(shap_test), " genes"),
    x = paste0("SHAP value (contribution to P(", pos_label, "))"),
    y = ""
  )

print(p_bee)
ggsave(here("plots", paste0("shap_beeswarm_", suffix_shap, ".pdf")),
       plot = p_bee, width = 10, height = 10)
```

# Mean |SHAP| Bar Plot

```{r shap_bar, fig.width=9, fig.height=10}
p_bar <- shap_importance %>%
  head(n_display) %>%
  ggplot(aes(x = reorder(feature_clean, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "#E69F00") +
  coord_flip() +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("Mean |SHAP|, top ", n_display, " features: ", gene_label, " targets"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | test set, n = ", nrow(shap_test), " genes"),
    x = "", y = "Mean |SHAP| (probability units)"
  )

print(p_bar)
ggsave(here("plots", paste0("shap_importance_bar_", suffix_shap, ".pdf")),
       plot = p_bar, width = 9, height = 10)
```

# Directional Importance

Signed importance, following the notebook 07 lollipop convention
(`importance * sign(correlation)`): bar length is mean |SHAP| and the sign comes from
`shap_value_cor`, so bars extend right when high feature values push a gene toward
`r pos_label` and left when they push it away. Bars are coloured by feature category.

Signed this way rather than as `mean_signed_shap`, which is **not** a directional quantity:
it averages to near zero for every feature (contributions cancel across genes) and its
residual offset reflects the gap between the explain-set mean prediction and the training
baseline. It is ~1000x smaller than mean |SHAP| and would render as a flat, near-empty
panel. It remains in the exported CSV for anyone who wants it.

```{r shap_directional_bar, fig.width=10, fig.height=10}
n_display <- 30   # category_colors is defined with the category map

signed_df <- shap_importance %>%
  head(n_display) %>%
  mutate(
    signed_importance = mean_abs_shap * sign(shap_value_cor),
    feature_clean     = fct_reorder(feature_clean, signed_importance)
  )

signed_lim <- max(abs(signed_df$signed_importance)) * 1.05

p_signed <- signed_df %>%
  ggplot(aes(x = feature_clean, y = signed_importance, fill = category)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.4) +
  coord_flip() +
  scale_y_continuous(limits = c(-signed_lim, signed_lim)) +
  scale_fill_manual(values = category_colors, name = NULL) +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("Signed importance, top ", n_display, " features: ",
                      gene_label, " targets"),
    subtitle = paste0("Mean |SHAP| signed by direction; right = higher values push toward ",
                      pos_label, " | test set, n = ", nrow(shap_test), " genes"),
    x = "", y = "Signed mean |SHAP| (probability units)"
  )

print(p_signed)
ggsave(here("plots", paste0("shap_directional_bar_", suffix_shap, ".pdf")),
       plot = p_signed, width = 10, height = 10)
```

# Dependence Plots

```{r dependence, fig.width=11, fig.height=7}
dep_features <- shap_importance %>% head(6) %>% pull(feature)

dep_df <- map_dfr(dep_features, function(f) {
  tibble(feature = f, value = test_data[[f]], shap = shap_test[[f]])
}) %>%
  left_join(shap_importance %>% dplyr::select(feature, feature_clean, rank_shap), by = "feature") %>%
  mutate(feature_clean = fct_reorder(feature_clean, rank_shap))

p_dep <- ggplot(dep_df, aes(x = value, y = shap)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(color = "#E69F00", alpha = 0.5, size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = "#0072B2", linewidth = 0.8) +
  facet_wrap(~ feature_clean, scales = "free_x", nrow = 2) +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("SHAP dependence, top 6 features: ", gene_label, " targets"),
    subtitle = paste0(tools::toTitleCase(params$condition), " | ", params$timepoint,
                      " | test set, n = ", nrow(shap_test), " genes"),
    x = "Feature value", y = paste0("SHAP value (contribution to P(", pos_label, "))")
  )

print(p_dep)
ggsave(here("plots", paste0("shap_dependence_top6_", suffix_shap, ".pdf")),
       plot = p_dep, width = 11, height = 7)
```

# Entanglement Diagnostics: GC Content vs Codon Optimality

The top of the importance ranking contains an apparent opposition: **low** CDS GC% pushes
toward `r pos_label` while **high** CSC pushes toward `r pos_label`. Since GC3 is
mechanistically the bridge between bulk GC content and codon optimality, the two could be
trading credit for one shared signal rather than acting as two effects. Four diagnostics,
cheapest first. Each has a stated decision rule so the answer does not depend on
eyeballing.

## 1. How correlated are they?

```{r entangle_correlation, fig.width=8, fig.height=7, eval=run_pair_diag}
# Derived from categorize_feature() rather than restated as literal lists, so that editing
# the category map cannot leave these diagnostics silently grouping by a stale definition.
gc_block    <- feature_meta %>% filter(category == "GC content")      %>% pull(feature)
codon_block <- feature_meta %>% filter(category == "Codon optimality") %>% pull(feature)
comp_block  <- c(gc_block, codon_block)

stopifnot(
  "GC content category is empty - check categorize_feature()"      = length(gc_block) > 0,
  "Codon optimality category is empty - check categorize_feature()" = length(codon_block) > 0,
  "cds_gc is not in the GC content category"                        = "cds_gc" %in% gc_block,
  "csc is not in the Codon optimality category"                     = "csc" %in% codon_block
)
cat("GC content block (", length(gc_block), "):", paste(gc_block, collapse = ", "), "\n")
cat("Codon optimality block (", length(codon_block), "):", paste(codon_block, collapse = ", "), "\n\n")

cat("Spearman correlation, GC and codon-optimality features (training set, n =",
    nrow(train_data), "):\n\n")
print(round(cor(train_data[, comp_block], method = "spearman"), 2))

rho_gc_csc <- cor(train_data$cds_gc, train_data$csc, method = "spearman")
rho_gc_gc3 <- cor(train_data$cds_gc, train_data$gc3, method = "spearman")
rho_gc3_csc <- cor(train_data$gc3,   train_data$csc, method = "spearman")
# partial correlation of GC and CSC controlling for GC3, the proposed bridge
rho_partial <- (rho_gc_csc - rho_gc_gc3 * rho_gc3_csc) /
  sqrt((1 - rho_gc_gc3^2) * (1 - rho_gc3_csc^2))

cat("\nHeadline pairs:\n")
cat("  rho(cds_gc, csc)        :", round(rho_gc_csc, 3), "\n")
cat("  rho(cds_gc, gc3)        :", round(rho_gc_gc3, 3), "\n")
cat("  rho(gc3, csc)           :", round(rho_gc3_csc, 3), "\n")
cat("  rho(cds_gc, csc | gc3)  :", round(rho_partial, 3), "(partial, controlling GC3)\n")
cat("\nDecision rule: |rho| <~ 0.4 means credit-splitting concern is small;\n")
cat("               |rho| >~ 0.7 means GC and CSC cannot be separated by attribution alone.\n")
```

Each codon-category feature scored against GC3 (the bridge) and against CSC. This is where
the category's internal heterogeneity shows up:

`pct_imputed` is the share of training genes whose value was NA before median imputation.
Those genes all sit at exactly the median, which attenuates the correlations slightly; the
effect here is at most 0.01 (checked against complete cases), so the classification is
unaffected, but it is what produces the flat stripes in the figure below.

The `behaves_as` cutoffs (0.6 / 0.4) are a labelling convenience, not principled constants.
They are safe here only because the data is cleanly bimodal: there is a 0.35 gap below `fop`
and a 0.25 gap below `cnot3_weighted_codon_score`, so the GC-driven call is identical for any
cutoff between 0.50 and 0.80. Note the rule reads `rho_gc3` only, so "GC-independent" means
"not GC-driven", it does not assert the feature is a good CSC proxy.

```{r entangle_codon_membership, eval=run_pair_diag}
codon_profile <- tibble(feature = codon_block) %>%
  mutate(
    rho_gc3     = map_dbl(feature, ~cor(train_data[[.x]], train_data$gc3, method = "spearman")),
    rho_csc     = map_dbl(feature, ~cor(train_data[[.x]], train_data$csc, method = "spearman")),
    pct_imputed = map_dbl(feature, ~100 * mean(is.na(pre_impute_train[[.x]]))),
    behaves_as  = case_when(
      abs(rho_gc3) > 0.6 ~ "GC-driven",
      abs(rho_gc3) > 0.4 ~ "intermediate",
      TRUE               ~ "GC-independent"
    )
  ) %>%
  arrange(desc(rho_gc3))

print_table(codon_profile, caption = "Codon optimality features: GC3 vs CSC entanglement")

cat("\nNote: rho(csc, csc) = 1 is self-correlation, not a result.\n")
```

## 1b. The same comparison, plotted

Reading down a column answers "is this feature measuring GC or measuring codon stability?".
A tight cloud in the top row with a diffuse one below means the feature is really a GC
measure. Grey points are median-imputed and carry no information; they form the flat stripes.

```{r entangle_facets, fig.width=20, fig.height=6.5, eval=run_pair_diag}
codon_ord <- codon_profile$feature

facet_df <- expand_grid(feature = codon_ord, reference = c("gc3", "csc")) %>%
  mutate(
    rho  = map2_dbl(feature, reference,
                    ~cor(train_data[[.x]], train_data[[.y]], method = "spearman")),
    data = map2(feature, reference, ~tibble(
      x        = train_data[[.y]],
      y        = train_data[[.x]],
      imputed  = is.na(pre_impute_train[[.x]]) | is.na(pre_impute_train[[.y]])
    ))
  ) %>%
  left_join(codon_profile %>% dplyr::select(feature, behaves_as), by = "feature") %>%
  mutate(
    ref_lab = ifelse(reference == "gc3", "vs GC3", "vs CSC"),
    f_lab   = feature_meta$feature_clean[match(feature, feature_meta$feature)],
    panel   = factor(paste0(f_lab, "\n", ref_lab),
                     levels = c(paste0(feature_meta$feature_clean[match(codon_ord, feature_meta$feature)], "\nvs GC3"),
                                paste0(feature_meta$feature_clean[match(codon_ord, feature_meta$feature)], "\nvs CSC")))
  ) %>%
  unnest(data)

facet_ann <- facet_df %>%
  distinct(panel, rho, feature, reference) %>%
  mutate(txt = ifelse(feature == "csc" & reference == "csc",
                      "self (rho = 1)", paste0("rho = ", sprintf("%.2f", rho))))

behaves_colors <- c("GC-driven" = "#D55E00", "intermediate" = "#CC79A7",
                    "GC-independent" = "#009E73")

p_facets <- ggplot() +
  geom_point(data = facet_df %>% filter(imputed), aes(x, y),
             color = "grey65", size = 0.35, alpha = 0.35) +
  geom_point(data = facet_df %>% filter(!imputed), aes(x, y, color = behaves_as),
             size = 0.35, alpha = 0.18) +
  geom_smooth(data = facet_df %>% filter(!imputed), aes(x, y),
              method = "lm", se = FALSE, linewidth = 0.7, color = "grey20") +
  geom_text(data = facet_ann, aes(x = Inf, y = Inf, label = txt),
            hjust = 1.08, vjust = 1.5, size = 3) +
  facet_wrap(~panel, scales = "free", nrow = 2) +
  scale_color_manual(values = behaves_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_classic(base_size = 11) +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 9, lineheight = 0.9),
        legend.position = "top") +
  labs(
    title = "Codon optimality features: what are they actually measuring?",
    subtitle = paste0("Top row: each feature vs GC3 (the proposed GC bridge). Bottom row: ",
                      "same feature vs CSC. Grey = median-imputed. Training set, n = ",
                      nrow(train_data), " genes"),
    x = "Reference feature value", y = "Codon feature value"
  )

print(p_facets)
ggsave(here("plots", paste0("shap_codon_gc_facets_", suffix_shap, ".pdf")),
       plot = p_facets, width = 20, height = 6.5)
```

## 1c. GC3 entanglement across every feature

The codon table asks the question of 10 features. GC3 is not specific to codon usage,
though, so the same screen is worth running across the whole feature set: **which features,
in any category, are really restating GC3?**

`gc3` itself is excluded (self-correlation) along with any zero-variance feature.

```{r gc3_screen_all, fig.width=9, fig.height=11, eval=run_gc3_screen}
zero_var <- feature_cols[map_dbl(feature_cols, ~sd(train_data[[.x]])) == 0]
if (length(zero_var) > 0) {
  cat("Excluded", length(zero_var), "zero-variance features (constant, carry no information):\n")
  cat(paste("  ", zero_var, collapse = "\n"), "\n\n")
}
screen_features <- setdiff(feature_cols, c("gc3", zero_var))

gc3_screen <- tibble(feature = screen_features) %>%
  mutate(rho_gc3 = map_dbl(feature, ~cor(train_data[[.x]], train_data$gc3, method = "spearman"))) %>%
  left_join(feature_meta, by = "feature") %>%
  arrange(desc(abs(rho_gc3)))

cat("Features with |rho| > 0.4 against GC3:", sum(abs(gc3_screen$rho_gc3) > 0.4),
    "of", nrow(gc3_screen), "\n\n")
print_table(gc3_screen %>% filter(abs(rho_gc3) > 0.4) %>%
              dplyr::select(feature, rho_gc3, category),
            caption = "Features with |rho| > 0.4 against GC3")

p_gc3_screen <- gc3_screen %>%
  mutate(feature_clean = fct_reorder(feature_clean, rho_gc3)) %>%
  ggplot(aes(rho_gc3, feature_clean, fill = category)) +
  geom_col() +
  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.4) +
  geom_vline(xintercept = c(-0.4, 0.4), linetype = "dashed", color = "grey55") +
  scale_fill_manual(values = category_colors, name = NULL) +
  scale_x_continuous(limits = c(-1, 1)) +
  theme_classic(base_size = 10) +
  labs(
    title    = "GC3 entanglement across every feature",
    subtitle = paste0("Spearman rho vs GC3. Dashed lines at +/-0.4. Training set, n = ",
                      nrow(train_data), " genes"),
    x = "Spearman rho with GC3", y = ""
  )

print(p_gc3_screen)
ggsave(here("plots", paste0("shap_gc3_screen_all_", suffix_shap, ".pdf")),
       plot = p_gc3_screen, width = 9, height = 11)
```

The same screen as scatter panels, one per feature, ordered by |rho|. The bar chart above
gives the ranking; this shows the shape behind each number, which is where you can see that
some low-|rho| features are binary or floored at zero and so a correlation is not really the
right summary for them.

```{r gc3_facets_all, fig.width=19, fig.height=22, eval=run_gc3_screen}
gc3_all_df <- gc3_screen %>%
  mutate(
    panel = factor(paste0(feature_clean, "\nrho = ", sprintf("%.2f", rho_gc3)),
                   levels = paste0(feature_clean, "\nrho = ", sprintf("%.2f", rho_gc3))),
    data  = map(feature, ~tibble(
      x       = train_data$gc3,
      y       = train_data[[.x]],
      imputed = is.na(pre_impute_train[[.x]]) | is.na(pre_impute_train[["gc3"]])
    ))
  ) %>%
  unnest(data)

p_gc3_all <- ggplot() +
  geom_point(data = gc3_all_df %>% filter(imputed), aes(x, y),
             color = "grey70", size = 0.22, alpha = 0.30) +
  geom_point(data = gc3_all_df %>% filter(!imputed), aes(x, y, color = category),
             size = 0.22, alpha = 0.16) +
  geom_smooth(data = gc3_all_df %>% filter(!imputed), aes(x, y),
              method = "lm", se = FALSE, linewidth = 0.5, color = "grey15") +
  facet_wrap(~panel, scales = "free_y", ncol = 7) +
  scale_color_manual(values = category_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 2)) +
  theme_classic(base_size = 9) +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 7, lineheight = 0.9),
        legend.position = "top", axis.text = element_text(size = 6.5)) +
  labs(
    title    = "Every feature vs GC3, ordered by |Spearman rho|",
    subtitle = paste0("Grey points = median-imputed. Training set, n = ", nrow(train_data),
                      " genes. gc3 itself and ", length(zero_var),
                      " constant features excluded."),
    x = "GC3 (%)", y = "Feature value"
  )

print(p_gc3_all)
ggsave(here("plots", paste0("shap_gc3_facets_all_", suffix_shap, ".pdf")),
       plot = p_gc3_all, width = 19, height = 22, limitsize = FALSE)
```

Scatter panels for just the features clearing the threshold, same construction as 1b:

```{r gc3_driven_facets, fig.width=16, fig.height=7, eval=run_gc3_screen}
gc3_driven <- gc3_screen %>% filter(abs(rho_gc3) > 0.4)

gc3_facet_df <- gc3_driven %>%
  mutate(
    panel = factor(paste0(feature_clean, "\nrho = ", sprintf("%.2f", rho_gc3)),
                   levels = paste0(feature_clean, "\nrho = ", sprintf("%.2f", rho_gc3))),
    data  = map(feature, ~tibble(
      x       = train_data$gc3,
      y       = train_data[[.x]],
      imputed = is.na(pre_impute_train[[.x]]) | is.na(pre_impute_train[["gc3"]])
    ))
  ) %>%
  unnest(data)

p_gc3_driven <- ggplot() +
  geom_point(data = gc3_facet_df %>% filter(imputed), aes(x, y),
             color = "grey65", size = 0.35, alpha = 0.35) +
  geom_point(data = gc3_facet_df %>% filter(!imputed), aes(x, y, color = category),
             size = 0.35, alpha = 0.18) +
  geom_smooth(data = gc3_facet_df %>% filter(!imputed), aes(x, y),
              method = "lm", se = FALSE, linewidth = 0.7, color = "grey20") +
  facet_wrap(~panel, scales = "free_y", nrow = 2) +
  scale_color_manual(values = category_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_classic(base_size = 11) +
  theme(strip.background = element_blank(), legend.position = "top") +
  labs(
    title    = "GC3-driven features (|rho| > 0.4)",
    subtitle = paste0("Every feature whose value tracks GC3, any category. ",
                      "Grey = median-imputed. Training set, n = ", nrow(train_data), " genes"),
    x = "GC3 (%)", y = "Feature value"
  )

print(p_gc3_driven)
ggsave(here("plots", paste0("shap_gc3_driven_facets_", suffix_shap, ".pdf")),
       plot = p_gc3_driven, width = 16, height = 7)

# Cross-reference against SHAP importance: a GC3-driven feature that also ranks high is
# where the redundancy actually matters for interpretation.
cat("\nGC3-driven features cross-referenced with SHAP importance rank:\n")
print_table(gc3_driven %>%
              left_join(shap_importance %>% dplyr::select(feature, rank_shap, mean_abs_shap),
                        by = "feature") %>%
              dplyr::select(feature, rho_gc3, rank_shap, mean_abs_shap, category) %>%
              arrange(rank_shap),
            caption = "GC3-driven features vs their SHAP importance rank")
```

### Synthesis: signed importance coloured by GC3 entanglement

The signed importance bar from earlier, but filled by each feature's correlation with GC3
instead of by category. This is the plot that answers "how much of the model's ranking is
GC3 in disguise?" in one view: strongly coloured bars near the top are important features
that are largely restating GC3, pale bars are important features carrying independent signal.

`gc3` itself is shown with rho = 1 by definition. Bars are outlined because a feature with
rho near 0 fills white and would otherwise vanish against the panel.

```{r shap_signed_by_gc3, fig.width=10, fig.height=10, eval=run_gc3_screen}
n_display_gc3 <- 30

rho_gc3_all <- tibble(feature = feature_cols) %>%
  mutate(rho_gc3 = map_dbl(feature, ~{
    if (.x == "gc3") return(1)                       # self-correlation, shown for reference
    if (sd(train_data[[.x]]) == 0) return(NA_real_)  # constant features
    cor(train_data[[.x]], train_data$gc3, method = "spearman")
  }))

signed_gc3_df <- shap_importance %>%
  left_join(rho_gc3_all, by = "feature") %>%
  head(n_display_gc3) %>%
  mutate(
    signed_importance = mean_abs_shap * sign(shap_value_cor),
    feature_clean     = fct_reorder(feature_clean, signed_importance)
  )

gc3_lim <- max(abs(signed_gc3_df$signed_importance)) * 1.05

p_signed_gc3 <- signed_gc3_df %>%
  ggplot(aes(x = feature_clean, y = signed_importance, fill = rho_gc3)) +
  geom_col(color = "grey40", linewidth = 0.2) +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.4) +
  coord_flip() +
  scale_y_continuous(limits = c(-gc3_lim, gc3_lim)) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00",
                       midpoint = 0, limits = c(-1, 1), na.value = "grey85",
                       name = "Spearman rho\nwith GC3") +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("Signed importance coloured by GC3 entanglement: ", gene_label, " targets"),
    subtitle = paste0("Bar = mean |SHAP| signed by direction; right = higher values push toward ",
                      pos_label, "\nFill = how much the feature restates GC3 | test set, n = ",
                      nrow(shap_test), " genes"),
    x = "", y = "Signed mean |SHAP| (probability units)"
  )

print(p_signed_gc3)
ggsave(here("plots", paste0("shap_signed_by_gc3_", suffix_shap, ".pdf")),
       plot = p_signed_gc3, width = 10, height = 10)

cat("Share of top-", n_display_gc3, " signed importance carried by features with |rho(GC3)| > 0.4: ",
    round(100 * sum(abs(signed_gc3_df$mean_abs_shap[abs(signed_gc3_df$rho_gc3) > 0.4]), na.rm = TRUE) /
            sum(signed_gc3_df$mean_abs_shap), 1), "%\n", sep = "")
```

## 2. SHAP interaction values

The most targeted test: TreeSHAP interaction values split each attribution into a main
effect (diagonal) and pairwise cross terms (off-diagonal). If the GC x CSC cross term is
small relative to the two main effects, the opposition is two independent main effects.

This is roughly `length(feature_cols)` times more expensive than plain TreeSHAP (~10 min on
a 200-gene subsample here, ~29 min on the full test set), so it runs on a subsample and is
cached to disk. Delete the cache file or set `force_recompute_interactions = TRUE` to redo it.

```{r shap_interactions, eval=run_interactions}
n_inter_genes <- 200
force_recompute_interactions <- FALSE
inter_path <- here("output", "predictive_modeling",
                   paste0("shap_interactions_", suffix_shap, ".rds"))

set.seed(9)
inter_idx <- sample(seq_len(nrow(shap_test)), min(n_inter_genes, nrow(shap_test)))

if (file.exists(inter_path) && !force_recompute_interactions) {
  cached <- readRDS(inter_path)
  interactions <- cached$interactions
  inter_idx    <- cached$inter_idx
  cat("Loaded cached interaction values:", basename(inter_path), "\n")
} else {
  cat("Computing SHAP interaction values on", length(inter_idx), "genes (expect ~10 min)...\n")
  t_start <- Sys.time()
  ti <- treeshap(unified, test_data[inter_idx, feature_cols], interactions = TRUE, verbose = FALSE)
  interactions <- ti$interactions
  cat("  elapsed:", round(as.numeric(difftime(Sys.time(), t_start, units = "mins")), 1), "min\n")
  saveRDS(list(interactions = interactions, inter_idx = inter_idx), inter_path)
}

# Orientation check: the array is [feature, feature, gene] and each feature's row must sum
# to that gene's SHAP value. Asserted rather than assumed, since indexing a transposed
# array would silently return the wrong cross terms.
recon_inter <- t(apply(interactions, 3, rowSums))
stopifnot(
  "interaction array is not [feature, feature, gene]" =
    all(dim(interactions) == c(length(feature_cols), length(feature_cols), length(inter_idx))),
  "interaction rows do not sum to SHAP values - check array orientation" =
    max(abs(recon_inter - as.matrix(shap_test[inter_idx, dimnames(interactions)[[1]]]))) < 1e-8
)
cat("Orientation confirmed: interactions[i, j, gene], rows sum to SHAP\n\n")

main_gc  <- mean(abs(interactions["cds_gc", "cds_gc", ]))
main_csc <- mean(abs(interactions["csc", "csc", ]))
# the pair's cross term is split across the (i,j) and (j,i) cells, so the total is 2x one cell
cross_gc_csc <- 2 * mean(abs(interactions["cds_gc", "csc", ]))
inter_fraction <- cross_gc_csc / mean(c(main_gc, main_csc))

cat("GC x CSC interaction (n =", length(inter_idx), "genes):\n")
cat("  GC main effect     :", round(main_gc, 6), "\n")
cat("  CSC main effect    :", round(main_csc, 6), "\n")
cat("  GC x CSC cross term:", round(cross_gc_csc, 6), "\n")
cat("  cross / mean(mains):", round(inter_fraction, 4), "\n")
cat("\nDecision rule: small cross term relative to mains = two independent main effects;\n")
cat("               comparable or larger = the two features are trading credit.\n\n")

cat("Largest cross terms involving cds_gc or csc (for context):\n")
cross_tbl <- map_dfr(c("cds_gc", "csc"), function(a) {
  tibble(f1 = a, f2 = setdiff(feature_cols, a),
         cross = map_dbl(setdiff(feature_cols, a), ~2 * mean(abs(interactions[a, .x, ]))))
}) %>% arrange(desc(cross)) %>% head(8)
print_table(cross_tbl, digits = 6, caption = "Largest cross terms involving cds_gc or csc")
```

## 2b. The same pair diagnostics for DHX29 occupancy vs CDS GC%

The GC3 screen flagged `hia2026_dhx29_occupancy` as more GC-entangled than any codon feature
(rho = -0.94 against `cds_gc`) while ranking 4th by SHAP importance, so it gets the same
treatment. The contrast with GC/CSC is the point: two features can both be "entangled with
GC" and yet be entangled in completely different ways.

```{r pair_diagnostics, eval=run_interactions}
pair_diagnostics <- function(a, b) {
  ma <- mean(abs(interactions[a, a, ]))
  mb <- mean(abs(interactions[b, b, ]))
  cr <- 2 * mean(abs(interactions[a, b, ]))
  tibble(
    pair          = paste(a, "x", b),
    rho_value     = cor(train_data[[a]], train_data[[b]], method = "spearman"),
    cor_shap      = cor(shap_test[[a]], shap_test[[b]]),
    main_a        = ma,
    main_b        = mb,
    cross         = cr,
    cross_frac    = cr / mean(c(ma, mb)),
    merged_net    = mean(abs(shap_test[[a]] + shap_test[[b]])),
    sum_of_parts  = mean(abs(shap_test[[a]])) + mean(abs(shap_test[[b]]))
  )
}

pairs_tbl <- bind_rows(
  pair_diagnostics("cds_gc", "csc"),
  pair_diagnostics("gc3",    "csc"),
  pair_diagnostics("cds_gc", "hia2026_dhx29_occupancy"),
  pair_diagnostics("gc3",    "hia2026_dhx29_occupancy"),
  pair_diagnostics("cds_gc", "gc3"),
  pair_diagnostics("hia2026_dhx29_occupancy", "csc")
)
print_table(pairs_tbl, caption = "Pair entanglement diagnostics (interactions on 200 genes)")

cat("\nReading this table:\n")
cat("  cor_shap near 0  -> independent contributions (two real effects)\n")
cat("  cor_shap near +1 -> duplicated contributions (one signal, counted twice)\n")
cat("  merged_net << sum_of_parts -> the pair cancels (credit-trading)\n")
cat("  merged_net ~= sum_of_parts -> the pair adds (same-direction duplication)\n")
```

```{r colored_dependence_helpers, eval=run_pair_diag}
# Defined here because sections 2b and 3 both use it. Colours each point by a second
# feature's value: if the trend stratifies into arms the effect is an interaction, if the
# colour scatters randomly the two features act independently.
colored_dependence <- function(v, color_var) {
  # scale01 is local so this chunk runs standalone in the console
  scale01 <- function(x) {
    lo <- quantile(x, 0.02, na.rm = TRUE); hi <- quantile(x, 0.98, na.rm = TRUE)
    if (!is.finite(lo) || !is.finite(hi) || hi <= lo) return(rep(0.5, length(x)))
    pmin(pmax((x - lo) / (hi - lo), 0), 1)
  }
  v_lab  <- feature_meta$feature_clean[match(v, feature_meta$feature)]
  cv_lab <- feature_meta$feature_clean[match(color_var, feature_meta$feature)]
  tibble(value = test_data[[v]], shap = shap_test[[v]],
         color = scale01(test_data[[color_var]])) %>%
    ggplot(aes(value, shap, color = color)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(size = 1.3, alpha = 0.75) +
    scale_color_viridis_c(name = cv_lab, breaks = c(0, 1), labels = c("Low", "High")) +
    theme_classic(base_size = 12) +
    labs(x = v_lab, y = paste0("SHAP value for ", v_lab))
}

# Residual test: after removing the main trend, does the other feature explain what is left?
resid_cor <- function(v, color_var) {
  fit <- loess(shap_test[[v]] ~ test_data[[v]], span = 0.75)
  cor(residuals(fit), test_data[[color_var]], method = "spearman")
}
```

```{r dhx29_gc_dependence, fig.width=12, fig.height=5.5, eval=run_pair_diag}
cat("Residual correlation after removing the main trend:\n")
cat("  cds_gc residual SHAP vs DHX29 value:", round(resid_cor("cds_gc", "hia2026_dhx29_occupancy"), 3), "\n")
cat("  DHX29 residual SHAP vs cds_gc value:", round(resid_cor("hia2026_dhx29_occupancy", "cds_gc"), 3), "\n")

p_dh <- cowplot::plot_grid(
  colored_dependence("cds_gc", "hia2026_dhx29_occupancy"),
  colored_dependence("hia2026_dhx29_occupancy", "cds_gc"),
  nrow = 1, labels = c("A", "B")
)
print(p_dh)
ggsave(here("plots", paste0("shap_colored_dependence_gc_dhx29_", suffix_shap, ".pdf")),
       plot = p_dh, width = 12, height = 5.5)
```

### Ablation refit: is DHX29 occupancy carrying anything GC does not?

Attribution alone cannot answer this, because both features can receive credit for one
shared signal. Refitting without each feature is the causal test. Each refit is technically a
different model, so this is a redundancy statement, not a re-explanation of the same forest.

```{r ablation_refit, eval=run_pair_diag}
y_train_num <- as.numeric(train_data$eif3d_target == pos_label)

ablation_auc <- function(drop_cols) {
  fs <- setdiff(feature_cols, drop_cols)
  set.seed(9)
  m <- ranger(x = train_data[, fs], y = y_train_num, num.trees = rf_prob$num.trees,
              mtry = min(rf_prob$mtry, length(fs)),
              min.node.size = rf_prob$min.node.size, seed = 9)
  p <- predict(m, test_data[, fs])$predictions
  as.numeric(auc(roc(test_data$eif3d_target, p, levels = c(neg_label, pos_label),
                     direction = "<", quiet = TRUE)))
}

auc_base <- ablation_auc(character(0))
dh <- "hia2026_dhx29_occupancy"

# GC3 is included alongside CDS GC% throughout: the two are themselves 0.94 correlated, so
# dropping one alone leaves the other standing in, and only the joint drops are informative.
ablations <- list(
  "drop DHX29"              = dh,
  "drop cds_gc"             = "cds_gc",
  "drop gc3"                = "gc3",
  "drop cds_gc + gc3"       = c("cds_gc", "gc3"),
  "drop DHX29 + cds_gc"     = c(dh, "cds_gc"),
  "drop DHX29 + gc3"        = c(dh, "gc3"),
  "drop DHX29 + cds_gc + gc3" = c(dh, "cds_gc", "gc3"),
  "drop csc"                = "csc",
  "drop csc + cds_gc"       = c("csc", "cds_gc"),
  "drop csc + gc3"          = c("csc", "gc3"),
  "drop whole GC block"     = gc_block,
  "drop GC block + DHX29"   = c(gc_block, dh)
)

ablation_tbl <- bind_rows(
  tibble(ablation = "full model", n_dropped = 0L, auc = auc_base, delta = NA_real_),
  tibble(ablation = names(ablations)) %>%
    mutate(n_dropped = map_int(ablations, length),
           auc       = map_dbl(ablations, ablation_auc),
           delta     = auc - auc_base)
)

print_table(ablation_tbl,
            caption = paste0("Ablation refit, held-out AUC (test set, n = ", nrow(test_data), ")"))

cat("\nCaution: the AUC bootstrap 95% CI on this test set spans roughly +/-0.03,\n")
cat("so deltas of ~0.005 are within noise. Read the PATTERN (does dropping both cost\n")
cat("the same as dropping one?), not the individual numbers.\n")
```

## 3. Colored dependence

The visual form of the same question. If the trend stratifies into arms by the other
feature's value, the effect is an interaction; if the color scatters randomly across the
trend, the two act independently.

```{r colored_dependence, fig.width=12, fig.height=5, eval=run_pair_diag}
# colored_dependence() and resid_cor() are defined in the helpers chunk above
rc1 <- resid_cor("cds_gc", "csc")
rc2 <- resid_cor("csc", "cds_gc")
cat("Residual correlation after removing the main trend:\n")
cat("  cds_gc residual SHAP vs csc value:", round(rc1, 3), "\n")
cat("  csc residual SHAP vs cds_gc value:", round(rc2, 3), "\n")

p_cd <- cowplot::plot_grid(
  colored_dependence("cds_gc", "csc"),
  colored_dependence("csc", "cds_gc"),
  nrow = 1, labels = c("A", "B")
)
p_cd_titled <- cowplot::plot_grid(
  cowplot::ggdraw() + cowplot::draw_label(
    paste0("GC and CSC entanglement: rho = ", round(rho_gc_csc, 3),
           if (exists("inter_fraction"))
             paste0(", GC x CSC interaction = ", round(100 * inter_fraction, 1),
                    "% of main effects") else "",
           " (n = ", nrow(shap_test), " genes)"),
    fontface = "plain", size = 12, hjust = 0.5),
  p_cd, ncol = 1, rel_heights = c(0.08, 1)
)

print(p_cd_titled)
ggsave(here("plots", paste0("shap_colored_dependence_gc_csc_", suffix_shap, ".pdf")),
       plot = p_cd_titled, width = 12, height = 5.5)
```

## 4. Merged-group net effect

Category grouping fixes credit-splitting *within* a block but not opposition *across* blocks.
If GC and codon optimality were reshuffling one shared signal, merging them should cancel:
the merged net would be far below the sum of the parts.

```{r merged_group_net, eval=run_pair_diag}
net_gc    <- rowSums(shap_test[, gc_block,    drop = FALSE])
net_codon <- rowSums(shap_test[, codon_block, drop = FALSE])
net_comp  <- rowSums(shap_test[, comp_block,  drop = FALSE])

cat("Block net effects (test set, n =", nrow(shap_test), "genes):\n")
cat("  GC block        : mean|net| =", round(mean(abs(net_gc)), 5),
    " sum of parts =", round(sum(colMeans(abs(shap_test[, gc_block]))), 5), "\n")
cat("  Codon block     : mean|net| =", round(mean(abs(net_codon)), 5),
    " sum of parts =", round(sum(colMeans(abs(shap_test[, codon_block]))), 5), "\n")
cat("  MERGED composite: mean|net| =", round(mean(abs(net_comp)), 5),
    " sum of parts =", round(sum(colMeans(abs(shap_test[, comp_block]))), 5), "\n")
cat("  merged vs sum of the two block nets:", round(mean(abs(net_comp)), 5),
    "vs", round(mean(abs(net_gc)) + mean(abs(net_codon)), 5), "\n\n")
cat("  cor(net_gc, net_codon) across genes:", round(cor(net_gc, net_codon), 3), "\n")
cat("  Decision rule: strongly negative = the blocks cancel gene-by-gene (one shared signal);\n")
cat("                 near zero or positive = they are not merely trading credit.\n\n")

# Does reassigning the GC-driven codon features change the category ranking?
gc_driven <- codon_profile %>% filter(behaves_as == "GC-driven") %>% pull(feature)
if (length(gc_driven) > 0) {
  cat("Sensitivity of the category panel to reassigning", paste(gc_driven, collapse = "/"),
      "from Codon optimality to GC content:\n")
  cat("  Codon optimality as grouped :", round(mean(abs(net_codon)), 5), "\n")
  cat("  Codon optimality without them:",
      round(mean(abs(rowSums(shap_test[, setdiff(codon_block, gc_driven), drop = FALSE]))), 5), "\n")
  cat("  GC content as grouped        :", round(mean(abs(net_gc)), 5), "\n")
  cat("  GC content with them         :",
      round(mean(abs(rowSums(shap_test[, c(gc_block, gc_driven), drop = FALSE]))), 5), "\n")
}
```

# Synonymous Codon Preference vs Amino-Acid Composition

Runs only when the feature matrix carries per-codon features (`feature_set = "codon_rscu"`),
skipped otherwise.

Aggregate GC3 cannot answer whether eIF3d prefers particular **synonymous** codons. It is
0.93-correlated with bulk `cds_gc`, 0.66-correlated with GC1 through genome-wide isochore
structure, and it is a net measure across all amino-acid families, so within-family
preferences that run in opposite directions cancel inside it.

Nor can raw codon **frequency**: `codon_freq_AAA` and `codon_freq_AAG` both rise with lysine
content, so their attributions are dominated by amino-acid composition. A family-direction
test built on frequency features will report "same direction" whether or not a synonymous
preference exists, and is therefore uninformative for this question.

**bg-RSCU is the right instrument.** It is normalised within each synonymous family by
construction, so amino-acid content is divided out and the remaining signal is the wobble
choice. The two tests below use it.

```{r codon_setup, eval=run_codon_tests}
genetic_code <- Biostrings::GENETIC_CODE
gc_count <- function(x) map_dbl(x, ~sum(strsplit(.x, "")[[1]] %in% c("G", "C")))

codon_shap <- shap_importance %>%
  filter(str_detect(feature, "^(bg_rscu|codon_freq)_[A-Z]{3}$")) %>%
  mutate(
    kind      = ifelse(str_detect(feature, "^bg_rscu_"), "bg-RSCU (AA-normalised)",
                                                          "codon frequency (raw)"),
    codon     = str_remove(feature, "^(bg_rscu|codon_freq)_"),
    aminoacid = genetic_code[codon],
    gc12      = gc_count(substr(codon, 1, 2)),   # positions 1-2: amino-acid identity
    gc3_pos   = gc_count(substr(codon, 3, 3)),   # position 3: synonymous / wobble
    ending    = ifelse(gc3_pos == 0, "AT-ending", "GC-ending")
  ) %>%
  filter(!is.na(aminoacid), aminoacid != "*")

stopifnot(
  "codon features did not parse into amino acids" = nrow(codon_shap) > 0,
  "third-position classification is wrong" =
    all(codon_shap$ending == ifelse(substr(codon_shap$codon, 3, 3) %in% c("A", "T"),
                                    "AT-ending", "GC-ending"))
)
cat("Parsed", nrow(codon_shap), "per-codon features across",
    n_distinct(codon_shap$aminoacid), "amino acids\n")
```

## Test 1: do AT-ending and GC-ending synonymous codons push opposite ways?

```{r codon_at_gc_test, eval=run_codon_tests}
rscu <- codon_shap %>% filter(str_detect(kind, "bg-RSCU"))

print_table(
  rscu %>% group_by(ending) %>%
    summarise(n = n(),
              median_direction = median(shap_value_cor, na.rm = TRUE),
              frac_positive    = mean(shap_value_cor > 0, na.rm = TRUE),
              median_rank      = median(rank_shap), .groups = "drop"),
  caption = "bg-RSCU SHAP direction by third-codon-position identity")

w_end <- wilcox.test(shap_value_cor ~ ending, data = rscu)
cat("\nWilcoxon, AT-ending vs GC-ending SHAP direction: p =",
    format.pval(w_end$p.value, digits = 3), " n =", nrow(rscu), "\n")
cat("Positive direction = more of this codon (relative to its synonymous family)\n")
cat("pushes a gene toward '", pos_label, "'.\n", sep = "")
```

Within-family view: for each two-fold degenerate amino acid, the AT-ending and GC-ending
codon are the same amino acid, so any direction difference between them is purely
synonymous.

```{r codon_family_pairs, eval=run_codon_tests}
two_fold <- rscu %>%
  group_by(aminoacid) %>%
  filter(n() == 2, n_distinct(ending) == 2) %>%
  summarise(
    at_codon  = codon[ending == "AT-ending"],
    at_dir    = shap_value_cor[ending == "AT-ending"],
    gc_codon  = codon[ending == "GC-ending"],
    gc_dir    = shap_value_cor[ending == "GC-ending"],
    .groups   = "drop"
  ) %>%
  mutate(opposite = sign(at_dir) != sign(gc_dir)) %>%
  arrange(desc(at_dir - gc_dir))

print_table(two_fold, caption = "Two-fold degenerate families: AT-ending vs GC-ending direction")
cat("\nFamilies splitting in opposite directions:", sum(two_fold$opposite),
    "of", nrow(two_fold), "\n")
```

```{r codon_at_gc_plot, fig.width=11, fig.height=8, eval=run_codon_tests}
ending_colors <- c("AT-ending" = "#D55E00", "GC-ending" = "#0072B2")

p_codon_dir <- rscu %>%
  mutate(codon_lab = fct_reorder(paste0(codon, " (", aminoacid, ")"), shap_value_cor)) %>%
  ggplot(aes(shap_value_cor, codon_lab, color = ending)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_segment(aes(x = 0, xend = shap_value_cor, yend = codon_lab), linewidth = 0.4) +
  geom_point(size = 2) +
  scale_color_manual(values = ending_colors, name = NULL) +
  theme_classic(base_size = 10) +
  theme(legend.position = "top") +
  labs(
    title    = "Synonymous codon preference: bg-RSCU SHAP direction",
    subtitle = paste0("Amino-acid normalised, so this is wobble choice only. Right = pushes toward ",
                      pos_label, " | test set, n = ", nrow(shap_test), " genes"),
    x = "SHAP direction (correlation of feature value with its SHAP value)", y = ""
  )
print(p_codon_dir)
ggsave(here("plots", paste0("shap_codon_at_gc_direction_", suffix_shap, ".pdf")),
       plot = p_codon_dir, width = 11, height = 8)

p_pairs <- two_fold %>%
  dplyr::select(aminoacid, at_dir, gc_dir) %>%
  pivot_longer(c(at_dir, gc_dir), names_to = "ending", values_to = "direction") %>%
  mutate(ending = recode(ending, at_dir = "AT-ending", gc_dir = "GC-ending")) %>%
  ggplot(aes(ending, direction, group = aminoacid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(color = "grey60", linewidth = 0.6) +
  geom_point(aes(color = ending), size = 3) +
  ggrepel::geom_text_repel(
    data = ~dplyr::filter(.x, ending == "GC-ending"),
    aes(label = aminoacid), size = 3, nudge_x = 0.15, direction = "y") +
  scale_color_manual(values = ending_colors, guide = "none") +
  theme_classic(base_size = 12) +
  labs(
    title    = "Within-family synonymous flip",
    subtitle = "Each line is one two-fold degenerate amino acid",
    x = "", y = "SHAP direction"
  )
print(p_pairs)
ggsave(here("plots", paste0("shap_codon_family_pairs_", suffix_shap, ".pdf")),
       plot = p_pairs, width = 6, height = 6)
```

## Test 2: positions 1-2 vs position 3 dissociation

Why aggregate GC1, GC2 and GC3 all rank similarly in the `positional_gc` model despite the
synonymous effect being specific to position 3. Two distinct effects run in the same
direction:

- **Positions 1-2 carry amino-acid identity.** If the favoured amino acids are themselves
  AT-coded, GC1 and GC2 drop for a protein-composition reason with no wobble involvement.
- **Position 3 carries synonymous choice**, visible only after amino-acid normalisation.

Raw frequency features should track positions 1-2 (they contain amino-acid identity);
bg-RSCU features should track position 3 only (identity has been divided out).

```{r codon_position_dissociation, eval=run_codon_tests}
dissoc <- codon_shap %>%
  group_by(kind) %>%
  summarise(
    n                       = n(),
    rho_positions_1_2       = cor(gc12,    shap_value_cor, method = "spearman", use = "complete.obs"),
    rho_position_3          = cor(gc3_pos, shap_value_cor, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )
print_table(dissoc, caption = "Which codon position predicts SHAP direction, by feature type")

cat("\nRaw frequency, direction grouped by GC content of positions 1-2:\n")
print_table(
  codon_shap %>% filter(str_detect(kind, "frequency")) %>%
    group_by(gc12) %>%
    summarise(n = n(), median_direction = median(shap_value_cor),
              frac_positive = mean(shap_value_cor > 0), .groups = "drop"),
  caption = "Amino-acid-level composition effect (positions 1-2)")

cat("\nAmino acids ranked by mean direction (from raw frequency features):\n")
print_table(
  codon_shap %>% filter(str_detect(kind, "frequency")) %>%
    group_by(aminoacid) %>%
    summarise(n_codons = n(), mean_direction = mean(shap_value_cor),
              gc_at_positions_1_2 = first(gc12), .groups = "drop") %>%
    arrange(desc(mean_direction)),
  caption = "Amino-acid preference and the GC content of its codons' first two positions")
```

```{r codon_position_plot, fig.width=11, fig.height=5, eval=run_codon_tests}
p_dissoc <- codon_shap %>%
  dplyr::select(kind, shap_value_cor, `positions 1-2` = gc12, `position 3` = gc3_pos) %>%
  pivot_longer(c(`positions 1-2`, `position 3`), names_to = "position", values_to = "gc") %>%
  ggplot(aes(factor(gc), shap_value_cor)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_boxplot(outlier.shape = NA, fill = "grey92", width = 0.6) +
  geom_jitter(width = 0.15, size = 1.1, alpha = 0.6, color = "#0072B2") +
  facet_grid(kind ~ position, scales = "free_x") +
  theme_classic(base_size = 11) +
  theme(strip.background = element_blank()) +
  labs(
    title    = "Amino-acid composition and synonymous choice are separable",
    subtitle = paste0("Raw frequency tracks positions 1-2 (amino-acid identity); ",
                      "bg-RSCU tracks position 3 only | n = ", nrow(shap_test), " genes"),
    x = "Number of G/C bases at these codon positions", y = "SHAP direction"
  )
print(p_dissoc)
ggsave(here("plots", paste0("shap_codon_position_dissociation_", suffix_shap, ".pdf")),
       plot = p_dissoc, width = 11, height = 5)
```

# Grouped SHAP by Category

SHAP values are additive, so summing within a category is exact. This is the one view that
resolves credit-splitting across the correlated blocks: `csc` / `csc_first75` / `csc_q1`-`q4`
each carry a fraction of the codon-optimality signal individually, but their per-gene sum is
the category's true contribution.

```{r grouped_shap, fig.width=9, fig.height=6}
# Per-gene category totals (exact, by additivity)
cat_per_gene <- map_dfr(unique(feature_meta$category), function(cc) {
  fs <- feature_meta %>% filter(category == cc) %>% pull(feature)
  tibble(category = cc, gene_row = seq_len(nrow(shap_test)),
         shap_sum = rowSums(shap_test[, fs, drop = FALSE]))
})

category_importance <- cat_per_gene %>%
  group_by(category) %>%
  summarise(
    mean_abs_shap = mean(abs(shap_sum)),
    .groups = "drop"
  ) %>%
  left_join(feature_meta %>% dplyr::count(category, name = "n_features"), by = "category") %>%
  left_join(
    shap_importance %>% group_by(category) %>%
      slice_max(mean_abs_shap, n = 1) %>%
      dplyr::select(category, top_feature = feature),
    by = "category"
  ) %>%
  arrange(desc(mean_abs_shap))

cat("Grouped SHAP by category (test set, n =", nrow(shap_test), "genes):\n")
print_table(category_importance, caption = "Grouped SHAP by feature category")

# Audit: no category should be masking a high-ranking feature inside a vague bucket
cat("\nTop 15 features by mean |SHAP| and their categories:\n")
print_table(shap_importance %>% head(15) %>% dplyr::select(rank_shap, feature, category),
            caption = "Top 15 features and their categories")

p_cat <- category_importance %>%
  ggplot(aes(x = reorder(category, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "#E69F00") +
  geom_text(aes(label = paste0("n=", n_features)), hjust = -0.2, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_classic(base_size = 12) +
  labs(
    title    = paste0("Grouped SHAP by feature category: ", gene_label, " targets"),
    subtitle = paste0("Per-gene sum within category, exact by additivity | test set, n = ",
                      nrow(shap_test), " genes"),
    x = "", y = "Mean |summed SHAP| (probability units)"
  )

print(p_cat)
ggsave(here("plots", paste0("shap_by_category_", suffix_shap, ".pdf")),
       plot = p_cat, width = 9, height = 6)
```

# Waterfall Plots

Per-gene decompositions. Defaults deliberately span the range of model behaviour rather
than showing three saturated predictions: one confident correct positive, one borderline
call near the decision boundary, and one confident error. Saturated cases tend to have the
least informative decompositions, with every feature pushing the same way.

Note the baseline anchor is the **training** base rate (see the TreeSHAP section), not the
test-set base rate.

```{r select_waterfall_genes}
test_pred_df <- tibble(
  gene_row = seq_len(nrow(test_data)),
  symbol   = test_data$symbol,
  gene_id  = test_data$gene_id_clean,
  actual   = as.character(test_data$eif3d_target),
  prob     = p_reg_test
) %>%
  mutate(predicted = ifelse(prob >= 0.5, pos_label, neg_label),
         correct   = predicted == actual)

if (nchar(params$waterfall_genes) > 0) {
  wanted <- str_trim(str_split(params$waterfall_genes, ",")[[1]])
  waterfall_rows <- test_pred_df %>% filter(symbol %in% wanted)
  missing_genes <- setdiff(wanted, waterfall_rows$symbol)
  if (length(missing_genes) > 0) {
    cat("Requested genes not in the test set:", paste(missing_genes, collapse = ", "), "\n")
  }
  waterfall_rows <- waterfall_rows %>% mutate(role = "requested")
} else {
  confident_tp <- test_pred_df %>%
    filter(actual == pos_label, correct) %>% slice_max(prob, n = 1) %>% mutate(role = "confident correct positive")
  borderline <- test_pred_df %>%
    slice_min(abs(prob - 0.5), n = 1) %>% mutate(role = "borderline call")
  confident_error <- test_pred_df %>%
    filter(!correct) %>% slice_max(abs(prob - 0.5), n = 1) %>% mutate(role = "confident error")
  waterfall_rows <- bind_rows(confident_tp, borderline, confident_error) %>%
    distinct(gene_row, .keep_all = TRUE)
}

cat("Waterfall genes:\n")
print_table(waterfall_rows %>% dplyr::select(symbol, actual, predicted, prob, role),
            caption = "Genes selected for waterfall plots")
```

```{r waterfall_plots, fig.width=9, fig.height=6}
make_waterfall <- function(row_i, info, n_top = 12) {
  contrib <- tibble(feature = feature_cols, shap = as.numeric(shap_test[row_i, ])) %>%
    left_join(feature_meta %>% dplyr::select(feature, feature_clean), by = "feature") %>%
    arrange(desc(abs(shap)))

  top <- contrib %>% head(n_top)
  rest_sum <- sum(contrib$shap[-seq_len(min(n_top, nrow(contrib)))])
  if (abs(rest_sum) > 0) {
    top <- bind_rows(top, tibble(
      feature = "__other__",
      shap = rest_sum,
      feature_clean = paste0(nrow(contrib) - n_top, " other features")
    ))
  }

  # Cumulative positions, starting from the baseline
  top <- top %>%
    mutate(
      order_i = row_number(),
      end     = shap_baseline + cumsum(shap),
      start   = end - shap,
      sign    = ifelse(shap >= 0, paste0("Toward ", pos_label), paste0("Toward ", neg_label)),
      feature_clean = fct_rev(fct_inorder(feature_clean))
    )

  final_pred <- shap_baseline + sum(contrib$shap)

  ggplot(top, aes(y = feature_clean)) +
    geom_vline(xintercept = shap_baseline, linetype = "dashed", color = "grey50") +
    geom_segment(aes(x = start, xend = end, yend = feature_clean, color = sign),
                 linewidth = 4, lineend = "butt") +
    scale_color_manual(values = setNames(c("#D55E00", "#0072B2"),
                                         c(paste0("Toward ", pos_label),
                                           paste0("Toward ", neg_label))),
                       name = NULL) +
    theme_classic(base_size = 12) +
    labs(
      title    = paste0(info$symbol, " (", info$role, ")"),
      subtitle = paste0("actual = ", info$actual, " | predicted P(", pos_label, ") = ",
                        round(final_pred, 3),
                        " | baseline (training base rate) = ", round(shap_baseline, 3)),
      x = paste0("Predicted P(", pos_label, ")"), y = ""
    )
}

for (i in seq_len(nrow(waterfall_rows))) {
  info <- waterfall_rows[i, ]
  p_wf <- make_waterfall(info$gene_row, info)
  print(p_wf)
  ggsave(here("plots", paste0("shap_waterfall_", info$symbol, "_", suffix_shap, ".pdf")),
         plot = p_wf, width = 9, height = 6)
}
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Surrogate model
saveRDS(rf_shap, file.path(output_dir, paste0("rf_classifier_", suffix_shap, ".rds")))

# Per-gene SHAP values, all labelled genes (train + test)
shap_values_export <- bind_cols(
  tibble(
    gene_id   = full_data$gene_id_clean,
    symbol    = full_data$symbol,
    actual    = as.character(full_data$eif3d_target),
    in_train  = seq_len(nrow(full_data)) %in% as.integer(train_idx),
    prob_pos  = p_reg_full,
    baseline  = shap_baseline
  ),
  shap_full
)
write_csv(shap_values_export,
          file.path(output_dir, paste0("shap_values_", suffix_shap, ".csv")))

# Feature-level SHAP importance (test set)
write_csv(shap_importance,
          file.path(output_dir, paste0("shap_importance_", suffix_shap, ".csv")))

# Category-level grouped SHAP (test set)
write_csv(category_importance,
          file.path(output_dir, paste0("shap_category_importance_", suffix_shap, ".csv")))

cat("Exported:\n")
cat("  rf_classifier_", suffix_shap, ".rds\n", sep = "")
cat("  shap_values_", suffix_shap, ".csv (", nrow(shap_values_export), " genes x ",
    length(feature_cols), " features)\n", sep = "")
cat("  shap_importance_", suffix_shap, ".csv\n", sep = "")
cat("  shap_category_importance_", suffix_shap, ".csv\n", sep = "")
```

# Summary

```{r summary}
cat("========================================\n")
cat(paste0("SHAP SUMMARY (", tools::toTitleCase(params$condition), ")\n"))
cat("========================================\n\n")

cat("Parameters:\n")
cat("  Condition:", params$condition, "\n")
cat("  Gene:", gene_label, "\n")
cat("  Direction:", params$direction, "\n")
cat("  Timepoint:", params$timepoint, "\n")
cat("  LFC threshold:", params$lfc, "\n")
cat("  Feature set:", params$feature_set, "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("Surrogate fidelity vs notebook 07 (test set, n =", nrow(shap_test), "):\n")
cat("  max |delta prob|:", format(max_abs_diff_test, digits = 3),
    paste0("(gate: < ", fidelity_tol, ")\n"))
cat("  Pearson r:", round(r_test, 6), "\n")
cat("  AUC:", round(auc_prob, 4), "(07) vs", round(auc_reg, 4), "(surrogate)\n")
if (!is.na(rho)) cat("  Spearman rho vs impurity importance:", round(rho, 3), "\n")
cat("\nSHAP baseline (training base rate):", round(shap_baseline, 4), "\n\n")

cat("Top 10 features by mean |SHAP|:\n")
for (i in seq_len(min(10, nrow(shap_importance)))) {
  cat("  ", i, ". ", shap_importance$feature[i],
      " (", round(shap_importance$mean_abs_shap[i], 5), ", ",
      shap_importance$category[i], ")\n", sep = "")
}

cat("\nTop 5 categories by mean |summed SHAP|:\n")
for (i in 1:min(5, nrow(category_importance))) {
  cat("  ", i, ". ", category_importance$category[i],
      " (", round(category_importance$mean_abs_shap[i], 5), ", n=",
      category_importance$n_features[i], ")\n", sep = "")
}

cat("========================================\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
