---
title: "`r paste0('Feature Directionality Analysis: eIF', params$gene, ' Target Prediction (', tools::toTitleCase(params$condition), ', ', params$direction, ', ', params$timepoint, ', LFC ', params$lfc, ', ', params$feature_set, ', clip_', params$clip_mode, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "hypoxia"
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "kozak"
  clip_mode: "aggregate"
  source_model: "no_boruta"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    condition = "hypoxia",
    gene      = "3d",
    direction = "promotes",
    timepoint = "1hr", # "1hr", "4hr", "24hr", or "alltimepoints" or "1and4hr"
    lfc       = "0.5",
    feature_set = "kozak", #targeted, combined, structure, clip_split, te_lfc, codon_optimality, or kozak
    clip_mode   = "aggregate", # "individual", "aggregate", or "both"
    source_model = "no_boruta" # "no_boruta" (07) or "boruta" (06)
  )
}

# Derived file paths and labels from params
geneset_csv <- if (params$timepoint == "alltimepoints") {
  paste0("translation_categories_si", params$gene, "_vs_sictrl_", params$condition, "_alltimepoints.csv")
} else if (params$timepoint == "1and4hr") {
  paste0("translation_categories_si", params$gene, "_vs_sictrl_", params$condition, "_1and4hr.csv")
} else {
  paste0(params$condition, "_", params$gene, "_", params$direction,
         "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")
}
feature_matrix_rds <- if (params$feature_set == "te_lfc") {
  "feature_matrix_te_lfc.rds"
} else if (params$feature_set == "codon_optimality") {
  "feature_matrix_codon_optimality.rds"
} else if (params$feature_set == "kozak") {
  "feature_matrix_kozak.rds"
} else if (params$feature_set == "clip_split") {
  paste0("feature_matrix_", params$timepoint, "_clip_split.rds")
} else if (params$feature_set == "structure") {
  paste0("feature_matrix_", params$timepoint, "_structure.rds")
} else {
  paste0("feature_matrix_", params$timepoint, "_combined.rds")
}
# Suffix matches 06 (boruta) or 07 (no_boruta) output naming
if (params$source_model == "no_boruta") {
  suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                   params$condition, "_", params$timepoint, "_lfc", params$lfc,
                   "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")
} else {
  suffix <- paste0("eif3", params$gene, "_", params$direction, "_",
                   params$condition, "_", params$timepoint, "_lfc", params$lfc,
                   "_", params$feature_set, "_clip", params$clip_mode)
}
pos_label <- params$direction
neg_label <- paste0("does_not_", sub("s$", "", params$direction))
gene_label <- paste0("eIF", params$gene)

cat("Gene set CSV:", geneset_csv, "\n")
cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix:", suffix, "\n")
cat("Positive label:", pos_label, "\n")
cat("Negative label:", neg_label, "\n")
```

# Overview

Post-hoc directionality analysis of RF features. Source model: **`r params$source_model`** (`r if (params$source_model == "no_boruta") "07" else "06"`). For each feature, we compute point-biserial correlations and Wilcoxon rank-sum tests to determine whether the feature is higher or lower in the `r pos_label` class.

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
# Feature matrix (same as 07 — always use the combined matrix)
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

# Positive class
if (params$timepoint %in% c("alltimepoints", "1and4hr")) {
  positive_genes <- read_csv(here("output", geneset_csv)) %>%
    filter(te_lfc > as.numeric(params$lfc), te_padj < 0.05) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
} else {
  positive_genes <- read_csv(here("output", "genesets", geneset_csv))
}
cat("Positive genes:", nrow(positive_genes), "\n")

# Negative class: genes not regulated by eIF3d in any condition
negative_control_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
cat("Negative control genes:", nrow(negative_control_genes), "\n")

# Importance from 07
importance_file <- here("output", "predictive_modeling",
                        paste0("rf_classification_importance_", suffix, ".csv"))
importance_df <- read_csv(importance_file)
cat("Features in importance table:", nrow(importance_df), "\n")
```

# Recreate Binary Target

Identical logic and seed to 07 so that the same gene sets are used.

```{r create_target}
set.seed(9)

# Only consider genes present in the feature matrix
pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

# Label genes
features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
```

# Compute Directionality

```{r compute_directionality}
# Binary numeric target for correlation (0 = neg, 1 = pos)
target_numeric <- as.numeric(model_genes$eif3d_target) - 1

directionality <- importance_df %>%
  rowwise() %>%
  mutate(
    mean_promotes = mean(model_genes[[feature]][target_numeric == 1], na.rm = TRUE),
    mean_does_not_promote = mean(model_genes[[feature]][target_numeric == 0], na.rm = TRUE),
    diff = mean_promotes - mean_does_not_promote,
    correlation = tryCatch(
      cor(model_genes[[feature]], target_numeric, use = "complete.obs"),
      error = function(e) NA_real_
    ),
    wilcox_p = tryCatch({
      wt <- wilcox.test(model_genes[[feature]][target_numeric == 1],
                        model_genes[[feature]][target_numeric == 0])
      wt$p.value
    }, error = function(e) NA_real_)
  ) %>%
  ungroup() %>%
  mutate(
    wilcox_p_adj = p.adjust(wilcox_p, method = "BH"),
    direction = case_when(
      is.na(correlation) ~ NA_character_,
      correlation >= 0   ~ "Higher in promotes",
      TRUE               ~ "Lower in promotes"
    )
  ) %>%
  arrange(desc(importance))

cat("Directionality computed for", nrow(directionality), "features\n")
cat("Higher in promotes:", sum(directionality$direction == "Higher in promotes", na.rm = TRUE), "\n")
cat("Lower in promotes:", sum(directionality$direction == "Lower in promotes", na.rm = TRUE), "\n")
cat("Significant (Wilcox BH p < 0.05):", sum(directionality$wilcox_p_adj < 0.05, na.rm = TRUE), "\n")
```

# Plots

## 1. Combined Importance + Directionality (Top 30)

```{r plot_importance_directionality, fig.width=10, fig.height=10}
plot_data <- directionality %>%
  filter(!is.na(direction)) %>%
  head(30) %>%
  mutate(
    abs_correlation = abs(correlation),
    feature_clean = fct_reorder(feature_clean, importance)
  )

p_combined <- ggplot(plot_data, aes(x = feature_clean, y = importance)) +
  geom_segment(aes(xend = feature_clean, y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(color = direction, size = abs_correlation)) +
  scale_color_manual(
    values = c("Higher in promotes" = "#009E73", "Lower in promotes" = "#D55E00"),
    name = "Direction"
  ) +
  scale_size_continuous(name = "|Correlation|", range = c(2, 7)) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top 30 Features: Importance + Directionality"),
    subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                      " | ", params$timepoint, " | LFC ", params$lfc),
    x = "",
    y = "Variable Importance (Impurity)"
  )

print(p_combined)

ggsave(here("plots", paste0("rf_directionality_importance_", suffix, ".pdf")),
       plot = p_combined, width = 10, height = 10)
```

## 2. Correlation Bar Plot (Significant Features)

```{r plot_correlation_bars, fig.width=10, fig.height=8}
sig_features <- directionality %>%
  filter(!is.na(correlation), wilcox_p_adj < 0.05) %>%
  arrange(desc(abs(correlation))) %>%
  head(40) %>%
  mutate(feature_clean = fct_reorder(feature_clean, abs(correlation)))

cat("Significant features (BH p < 0.05):", nrow(directionality %>% filter(!is.na(correlation), wilcox_p_adj < 0.05)), "\n")
cat("Plotting top", nrow(sig_features), "by |correlation|\n")

if (nrow(sig_features) > 0) {
  bar_height <- max(6, nrow(sig_features) * 0.35)

  p_corr <- ggplot(sig_features, aes(x = feature_clean, y = correlation, fill = direction)) +
    geom_col() +
    scale_fill_manual(
      values = c("Higher in promotes" = "#009E73", "Lower in promotes" = "#D55E00"),
      name = "Direction"
    ) +
    coord_flip() +
    theme_cowplot() +
    labs(
      title = "Point-Biserial Correlation with Target Class",
      subtitle = paste0("Features with Wilcoxon BH-adjusted p < 0.05 (n = ", nrow(sig_features), ")"),
      x = "",
      y = "Correlation"
    )

  print(p_corr)

  ggsave(here("plots", paste0("rf_directionality_correlation_", suffix, ".pdf")),
         plot = p_corr, width = 10, height = bar_height)
} else {
  cat("No features with BH-adjusted p < 0.05\n")
}
```

## 3. Volcano-Style Plot

```{r plot_volcano, fig.width=10, fig.height=8}
volcano_data <- directionality %>%
  filter(!is.na(correlation), !is.na(wilcox_p)) %>%
  mutate(neg_log10_p = -log10(pmax(wilcox_p, 1e-30)))

# Label top features by importance
top_n_label <- min(10, nrow(volcano_data))
label_data <- volcano_data %>% head(top_n_label)

p_volcano <- ggplot(volcano_data, aes(x = correlation, y = neg_log10_p)) +
  geom_point(aes(size = importance, color = direction), alpha = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = feature_clean),
    size = 3, max.overlaps = 20, segment.color = "grey50"
  ) +
  scale_color_manual(
    values = c("Higher in promotes" = "#009E73", "Lower in promotes" = "#D55E00"),
    name = "Direction"
  ) +
  scale_size_continuous(name = "RF Importance", range = c(1, 8)) +
  theme_cowplot() +
  labs(
    title = "Volcano Plot: Correlation vs. Significance",
    subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                      " | ", params$timepoint, " | LFC ", params$lfc),
    x = "Point-Biserial Correlation",
    y = "-log10(Wilcoxon p-value)"
  )

print(p_volcano)

ggsave(here("plots", paste0("rf_directionality_volcano_", suffix, ".pdf")),
       plot = p_volcano, width = 10, height = 8)
```

## 4. Importance vs. Log2 Fold Change

```{r plot_importance_vs_lfc, fig.width=10, fig.height=8}
lfc_data <- directionality %>%
  filter(!is.na(correlation),
         mean_promotes > 0, mean_does_not_promote > 0) %>%
  mutate(log2fc = log2(mean_promotes / mean_does_not_promote))

# Shared base plot
base_lfc_plot <- ggplot(lfc_data, aes(x = log2fc, y = importance)) +
  geom_point(aes(color = direction), alpha = 0.7, size = 2.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c("Higher in promotes" = "#009E73", "Lower in promotes" = "#D55E00"),
    name = "Direction"
  ) +
  theme_cowplot() +
  labs(
    x = paste0("Log2 Fold Change (", pos_label, " / ", neg_label, ")"),
    y = "Variable Importance (Impurity)"
  )

# Plot A: all features labeled
p_imp_lfc_all <- base_lfc_plot +
  ggrepel::geom_text_repel(
    aes(label = feature_clean),
    size = 2.5, max.overlaps = Inf, segment.color = "grey50",
    segment.size = 0.3, segment.alpha = 0.5,
    min.segment.length = 0, force = 2
  ) +
  labs(
    title = "Feature Importance vs. Log2 Fold Change (All Labeled)",
    subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                      " | ", params$timepoint, " | LFC ", params$lfc)
  )

print(p_imp_lfc_all)

ggsave(here("plots", paste0("rf_directionality_importance_vs_lfc_all_", suffix, ".pdf")),
       plot = p_imp_lfc_all, width = 10, height = 8)

# Plot B: only top features by importance or |log2FC|
top_by_importance <- lfc_data %>% slice_max(importance, n = 10)
top_by_abs_lfc    <- lfc_data %>% slice_max(abs(log2fc), n = 10)
label_data_top    <- bind_rows(top_by_importance, top_by_abs_lfc) %>%
  distinct(feature, .keep_all = TRUE)

p_imp_lfc_top <- base_lfc_plot +
  ggrepel::geom_text_repel(
    data = label_data_top,
    aes(label = feature_clean),
    size = 3, max.overlaps = Inf, segment.color = "grey50",
    segment.size = 0.3, segment.alpha = 0.5,
    min.segment.length = 0, force = 2
  ) +
  labs(
    title = "Feature Importance vs. Log2 Fold Change (Top Features)",
    subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                      " | ", params$timepoint, " | LFC ", params$lfc)
  )

print(p_imp_lfc_top)

ggsave(here("plots", paste0("rf_directionality_importance_vs_lfc_top_", suffix, ".pdf")),
       plot = p_imp_lfc_top, width = 10, height = 8)
```

## 5. Category-Level Summary

```{r plot_category_summary, fig.width=10, fig.height=6}
category_dir <- directionality %>%
  filter(!is.na(correlation)) %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    mean_correlation = mean(correlation, na.rm = TRUE),
    n_higher = sum(direction == "Higher in promotes"),
    n_lower = sum(direction == "Lower in promotes"),
    prop_higher = n_higher / n_features,
    n_significant = sum(wilcox_p_adj < 0.05, na.rm = TRUE),
    mean_importance = mean(importance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_importance))

cat("\nCategory-Level Directionality Summary:\n")
print(category_dir)

# Grouped bar: proportion higher vs lower per category
category_long <- category_dir %>%
  dplyr::select(category, n_higher, n_lower) %>%
  pivot_longer(cols = c(n_higher, n_lower), names_to = "dir", values_to = "count") %>%
  mutate(dir = ifelse(dir == "n_higher", "Higher in promotes", "Lower in promotes"))

p_cat <- ggplot(category_long,
                aes(x = reorder(category, count, sum), y = count, fill = dir)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("Higher in promotes" = "#009E73", "Lower in promotes" = "#D55E00"),
    name = "Direction"
  ) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Direction by Category",
    subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                      " | ", params$timepoint),
    x = "",
    y = "Number of Features"
  )

print(p_cat)

ggsave(here("plots", paste0("rf_directionality_category_", suffix, ".pdf")),
       plot = p_cat, width = 10, height = 6)
```

## 6. Signed-Importance Lollipop (Top 30)

```{r plot_lollipop, fig.width=10, fig.height=10}
lollipop_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  head(30) %>%
  mutate(
    signed_importance = importance * sign(correlation),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    )),
    feature_clean = fct_reorder(feature_clean, signed_importance)
  )

category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

p_lollipop <- ggdotchart(
  lollipop_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 4,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher in promotes)",
  title    = "Top 30 Features: Signed Importance",
  subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                    " | ", params$timepoint, " | LFC ", params$lfc)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop)

ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix, ".pdf")),
       plot = p_lollipop, width = 10, height = 10)
```

## 7. Signed-Importance Lollipop (All Features)

```{r plot_lollipop_all, fig.width=10, fig.height=ceiling(nrow(directionality %>% filter(!is.na(correlation))) * 0.28)}
lollipop_all_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  mutate(
    signed_importance = importance * sign(correlation),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    )),
    feature_clean = fct_reorder(feature_clean, signed_importance)
  )

plot_height_all <- max(10, nrow(lollipop_all_data) * 0.28)

p_lollipop_all <- ggdotchart(
  lollipop_all_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 3,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher in promotes)",
  title    = paste0("All Features: Signed Importance (n = ", nrow(lollipop_all_data), ")"),
  subtitle = paste0(gene_label, " | ", tools::toTitleCase(params$condition),
                    " | ", params$timepoint, " | LFC ", params$lfc)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop_all)

ggsave(here("plots", paste0("rf_directionality_lollipop_all_", suffix, ".pdf")),
       plot = p_lollipop_all, width = 10, height = plot_height_all)
```

# Export

```{r export}
output_dir <- here("output", "predictive_modeling")

# Full directionality table
export_df <- directionality %>%
  dplyr::select(feature, mean_promotes, mean_does_not_promote, diff,
                correlation, wilcox_p, wilcox_p_adj, importance, category, direction)

write_csv(export_df, file.path(output_dir,
          paste0("rf_feature_directionality_", suffix, ".csv")))

cat("Exported: rf_feature_directionality_", suffix, ".csv\n")
cat("  Rows:", nrow(export_df), "\n")
cat("  Significant (BH p < 0.05):", sum(export_df$wilcox_p_adj < 0.05, na.rm = TRUE), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
