---
title: "`r paste0('Cross-Condition Variable Importance Comparison: Hypoxia vs Normoxia (', params$feature_set, ', clip_', params$clip_mode, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  lfc: "0.5"
  feature_set: "te_lfc"
  clip_mode: "aggregate"
  source_model: "no_boruta"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    gene      = "3d",
    direction = "promotes",
    timepoint = "1hr",
    lfc       = "0.5",
    feature_set = "te_lfc", #targeted, combined, structure, clip_split, or te_lfc
    clip_mode   = "aggregate", # "individual", "aggregate", or "both"
    source_model = "no_boruta" # "no_boruta" (07) or "boruta" (06)
  )
}

# Build suffixes for each condition
make_suffix <- function(condition) {
  base <- paste0("eif3", params$gene, "_", params$direction, "_",
                 condition, "_", params$timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode)
  if (params$source_model == "no_boruta") {
    paste0(base, "_no_boruta")
  } else {
    base
  }
}

suffix_hyp <- make_suffix("hypoxia")
suffix_nor <- make_suffix("normoxia")
gene_label <- paste0("eIF", params$gene)
subtitle_label <- paste0(gene_label, " ", params$direction, " | ", params$timepoint,
                         " | LFC ", params$lfc, " | ", params$feature_set,
                         " | clip", params$clip_mode, " | ", params$source_model)

cat("Hypoxia suffix:", suffix_hyp, "\n")
cat("Normoxia suffix:", suffix_nor, "\n")
```

# Overview

Compare RF variable importance between hypoxia and normoxia for `r gene_label` `r params$direction` (`r params$timepoint`, lfc`r params$lfc`, `r params$feature_set`, clip`r params$clip_mode`, `r params$source_model`). Each condition was analyzed independently via 07/08; this notebook loads both sets of results and visualizes the differences.

Raw impurity-based importance is not comparable across independently trained forests (different sample sizes, class distributions, total impurity). We normalize each model's importance to **proportions** (sum to 1) before comparing.

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
# Directionality CSVs (from 08)
dir_hyp <- read_csv(here("output", "predictive_modeling",
  paste0("rf_feature_directionality_", suffix_hyp, ".csv")))
dir_nor <- read_csv(here("output", "predictive_modeling",
  paste0("rf_feature_directionality_", suffix_nor, ".csv")))

cat("Hypoxia features:", nrow(dir_hyp), "\n")
cat("Normoxia features:", nrow(dir_nor), "\n")

# Importance CSVs (from 07) — these have feature_clean and category
imp_hyp <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_importance_", suffix_hyp, ".csv")))
imp_nor <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_importance_", suffix_nor, ".csv")))

# Build feature_clean lookup from both importance files (union in case feature sets differ slightly)
feature_lookup <- bind_rows(
  imp_hyp %>% select(feature, feature_clean, category),
  imp_nor %>% select(feature, feature_clean, category)
) %>%
  distinct(feature, .keep_all = TRUE)

cat("Unique features in lookup:", nrow(feature_lookup), "\n")
```

# Join Conditions

```{r join_data}
# Normalize importance to proportions within each model so that
# values are on a common scale (raw impurity importance is not
# comparable across independently trained forests).
dir_hyp <- dir_hyp %>% mutate(importance = importance / sum(importance))
dir_nor <- dir_nor %>% mutate(importance = importance / sum(importance))

cat("Hypoxia importance sums to:", sum(dir_hyp$importance), "\n")
cat("Normoxia importance sums to:", sum(dir_nor$importance), "\n")

# Inner join directionality tables on feature
joined <- inner_join(
  dir_hyp %>% select(feature, importance_hyp = importance, correlation_hyp = correlation,
                     direction_hyp = direction, diff_hyp = diff,
                     wilcox_p_adj_hyp = wilcox_p_adj),
  dir_nor %>% select(feature, importance_nor = importance, correlation_nor = correlation,
                     direction_nor = direction, diff_nor = diff,
                     wilcox_p_adj_nor = wilcox_p_adj),
  by = "feature"
)

# Add feature_clean and category from lookup
joined <- joined %>%
  left_join(feature_lookup, by = "feature")

# Compute delta importance (hypoxia minus normoxia)
joined <- joined %>%
  mutate(
    delta_importance = importance_hyp - importance_nor,
    delta_sign = ifelse(delta_importance >= 0, "Higher in Hypoxia", "Higher in Normoxia")
  ) %>%
  arrange(desc(importance_hyp + importance_nor))

cat("Features in joined table:", nrow(joined), "\n")
cat("Features only in hypoxia:", sum(!dir_hyp$feature %in% dir_nor$feature), "\n")
cat("Features only in normoxia:", sum(!dir_nor$feature %in% dir_hyp$feature), "\n")
```

# Plots

## 1. Scatter Plot: Hypoxia vs Normoxia Importance

```{r scatter_plot, fig.width=10, fig.height=9}
# Label top features by combined importance
top_n_label <- min(15, nrow(joined))
label_data <- joined %>% head(top_n_label)

p_scatter <- ggplot(joined, aes(x = importance_hyp, y = importance_nor)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = category), size = 3, alpha = 0.7) +
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = feature_clean),
    size = 3, max.overlaps = 20, segment.color = "grey50"
  ) +
  theme_cowplot() +
  labs(
    title = "Variable Importance: Hypoxia vs Normoxia",
    subtitle = subtitle_label,
    x = "Relative Importance (Hypoxia)",
    y = "Relative Importance (Normoxia)",
    color = "Category"
  )

print(p_scatter)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_scatter, width = 10, height = 9)
```

## 2. Dumbbell Plot: Paired Importance by Condition

```{r dumbbell_plot, fig.width=10, fig.height=10}
# Top features by max importance across conditions
dumbbell_data <- joined %>%
  mutate(max_importance = pmax(importance_hyp, importance_nor)) %>%
  arrange(desc(max_importance)) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, max_importance))

p_dumbbell <- ggplot(dumbbell_data) +
  geom_segment(aes(x = importance_hyp, xend = importance_nor,
                   y = feature_clean, yend = feature_clean),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(x = importance_hyp, y = feature_clean, color = "Hypoxia"), size = 3) +
  geom_point(aes(x = importance_nor, y = feature_clean, color = "Normoxia"), size = 3) +
  scale_color_manual(
    values = c("Hypoxia" = "#D55E00", "Normoxia" = "#0072B2"),
    name = "Condition"
  ) +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: Paired Importance by Condition",
    subtitle = subtitle_label,
    x = "Relative Importance (Proportion)",
    y = ""
  )

print(p_dumbbell)

ggsave(here("plots", paste0("rf_importance_comparison_dumbbell_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_dumbbell, width = 10, height = 10)
```

## 3. Delta Importance Bar Plot

```{r delta_bar_plot, fig.width=10, fig.height=10}
delta_data <- joined %>%
  arrange(desc(abs(delta_importance))) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, delta_importance))

p_delta <- ggplot(delta_data, aes(x = feature_clean, y = delta_importance, fill = delta_sign)) +
  geom_col() +
  scale_fill_manual(
    values = c("Higher in Hypoxia" = "#009E73", "Higher in Normoxia" = "#D55E00"),
    name = ""
  ) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Delta Importance: Hypoxia minus Normoxia",
    subtitle = subtitle_label,
    x = "",
    y = "Delta Relative Importance (Hypoxia - Normoxia)"
  )

print(p_delta)

ggsave(here("plots", paste0("rf_importance_comparison_delta_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_delta, width = 10, height = 10)
```

## 4. Signed-Importance Lollipop by Condition (Top 30)

```{r plot_lollipop, fig.width=10, fig.height=10}
category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

make_lollipop <- function(data, cond_label) {
  imp_col  <- if (cond_label == "hypoxia") "importance_hyp"  else "importance_nor"
  corr_col <- if (cond_label == "hypoxia") "correlation_hyp" else "correlation_nor"

  lollipop_data <- data %>%
    filter(!is.na(.data[[corr_col]])) %>%
    mutate(
      importance        = .data[[imp_col]],
      correlation       = .data[[corr_col]],
      signed_importance = importance * sign(correlation),
      broad_category = case_when(
        category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
        category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
        category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
        category == "Translational Efficiency"                   ~ "Translation in hypoxia",
        TRUE                                                     ~ "Intrinsic mRNA features"
      ),
      broad_category = factor(broad_category, levels = c(
        "Intrinsic mRNA features", "eIF3 binding evidence",
        "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
      ))
    ) %>%
    arrange(desc(importance)) %>%
    head(30) %>%
    mutate(feature_clean = fct_reorder(feature_clean, signed_importance))

  ggdotchart(
    lollipop_data,
    x        = "feature_clean",
    y        = "signed_importance",
    color    = "broad_category",
    palette  = category_colors,
    sorting  = "none",
    add      = "segments",
    rotate   = TRUE,
    dot.size = 4,
    xlab     = "",
    ylab     = "Signed Importance (positive = higher in promotes)",
    title    = paste0("Top 30 Features: Signed Importance (", tools::toTitleCase(cond_label), ")"),
    subtitle = subtitle_label
  ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    theme(legend.title = element_blank())
}

p_lollipop_hyp <- make_lollipop(joined, "hypoxia")
p_lollipop_nor <- make_lollipop(joined, "normoxia")

print(p_lollipop_hyp)
print(p_lollipop_nor)

ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix_hyp, ".pdf")),
       plot = p_lollipop_hyp, width = 10, height = 10)
ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix_nor, ".pdf")),
       plot = p_lollipop_nor, width = 10, height = 10)
```

# Export

```{r export}
output_dir <- here("output", "predictive_modeling")

export_df <- joined %>%
  select(feature, feature_clean, category,
         importance_hyp, importance_nor, delta_importance,
         correlation_hyp, correlation_nor,
         direction_hyp, direction_nor,
         diff_hyp, diff_nor,
         wilcox_p_adj_hyp, wilcox_p_adj_nor)

write_csv(export_df, file.path(output_dir,
  paste0("rf_importance_comparison_", suffix_hyp, "_vs_", suffix_nor, ".csv")))

cat("Exported:", nrow(export_df), "features\n")
```

# Gene Set Overlap: Hypoxia vs Normoxia Promotes

```{r venn_genesets, fig.width=6, fig.height=5}
library(ggVennDiagram)

norm_genes <- read_csv(
  here("output", "genesets",
       paste0("normoxia_", params$gene, "_", params$direction,
              "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")),
  show_col_types = FALSE
) %>% pull(db_gene_symbol)

hyp_genes <- read_csv(
  here("output", "genesets",
       paste0("hypoxia_", params$gene, "_", params$direction,
              "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")),
  show_col_types = FALSE
) %>% pull(db_gene_symbol)

cat("Normoxia promotes:", length(norm_genes), "genes\n")
cat("Hypoxia promotes:", length(hyp_genes), "genes\n")
cat("Shared:", length(intersect(norm_genes, hyp_genes)), "genes\n")

venn_sets <- list(
  "Normoxia\npromotes" = norm_genes,
  "Hypoxia\npromotes"  = hyp_genes
)

p_venn <- ggVennDiagram(venn_sets, label_alpha = 0,
                        set_color = c("#D55E00", "#009E73")) +
  scale_fill_gradient(low = "white", high = "white") +
  labs(title = paste0("eIF", params$gene, " ", params$direction,
                      " gene set overlap\n",
                      params$timepoint, ", lfc", params$lfc)) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

print(p_venn)

ggsave(here("plots", paste0("venn_geneset_overlap_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_venn, width = 11, height = 10)
```

# Session Info

```{r session_info}
sessionInfo()
```
