---
title: "`r paste0('Cross-Condition Variable Importance Comparison: Hypoxia vs Normoxia (', params$feature_set, ', clip_', params$clip_mode, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  gene: "3d"
  direction: "promotes"
  timepoint: "1hr"
  timepoint_nor: "1and4hr"
  lfc: "0.5"
  feature_set: "external_stability"  # also accepts "codon_features" (bg-RSCU + codon freq from 01l)
  clip_mode: "aggregate"
  source_model: "no_boruta"
  exclude_own_te: true
  condition_hyp: "hypoxia"
  condition_nor: "normoxia"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    gene      = "3d",
    direction = "promotes",
    timepoint     = "1hr",           # hypoxia timepoint
    timepoint_nor = "1and4hr",       # normoxia timepoint
    lfc       = "0.5",
    feature_set = "codon_features",  # or "codon_features"
    clip_mode   = "aggregate",
    source_model = "no_boruta",
    exclude_own_te = TRUE,           # appends _noOwnTE to match 07 output
    condition_hyp = "hypoxia",       # condition string for the "hypoxia" arm (can be e.g. "mcf7six1_hypoxia")
    condition_nor = "normoxia"       # condition string for the "normoxia" arm (can be e.g. "mcf7six1_normoxia")
  )
}

# Build suffixes for each condition
make_suffix <- function(condition, timepoint = params$timepoint) {
  base <- paste0("eif3", params$gene, "_", params$direction, "_",
                 condition, "_", timepoint, "_lfc", params$lfc,
                 "_", params$feature_set, "_clip", params$clip_mode)
  suffix <- if (params$source_model == "no_boruta") paste0(base, "_no_boruta") else base
  if (isTRUE(params$exclude_own_te)) paste0(suffix, "_noOwnTE") else suffix
}

suffix_hyp <- make_suffix(params$condition_hyp, params$timepoint)
suffix_nor <- make_suffix(params$condition_nor, params$timepoint_nor)
gene_label <- paste0("eIF", params$gene)
subtitle_label <- paste0(gene_label, " ", params$direction,
                         " | hyp: ", params$timepoint, " vs nor: ", params$timepoint_nor,
                         " | LFC ", params$lfc, " | ", params$feature_set,
                         " | clip", params$clip_mode, " | ", params$source_model)

cat("Hypoxia suffix:", suffix_hyp, "\n")
cat("Normoxia suffix:", suffix_nor, "\n")
```

# Overview

Compare RF variable importance between hypoxia and normoxia for `r gene_label` `r params$direction` (`r params$timepoint`, lfc`r params$lfc`, `r params$feature_set`, clip`r params$clip_mode`, `r params$source_model`). Each condition was analyzed independently via 07/08; this notebook loads both sets of results and visualizes the differences.

Raw impurity-based importance is not comparable across independently trained forests (different sample sizes, class distributions, total impurity). We normalize each model's importance to **proportions** (sum to 1) before comparing.

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
# Directionality CSVs (from 07)
dir_hyp <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_directionality_", suffix_hyp, ".csv")))
dir_nor <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_directionality_", suffix_nor, ".csv")))

cat("Hypoxia features:", nrow(dir_hyp), "\n")
cat("Normoxia features:", nrow(dir_nor), "\n")

# Importance CSVs (from 07) — these have feature_clean and category
imp_hyp <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_importance_", suffix_hyp, ".csv")))
imp_nor <- read_csv(here("output", "predictive_modeling",
  paste0("rf_classification_importance_", suffix_nor, ".csv")))

# Build feature_clean lookup from both importance files (union in case feature sets differ slightly)
feature_lookup <- bind_rows(
  imp_hyp %>% dplyr::select(feature, feature_clean, category),
  imp_nor %>% dplyr::select(feature, feature_clean, category)
) %>%
  distinct(feature, .keep_all = TRUE)

cat("Unique features in lookup:", nrow(feature_lookup), "\n")
```

# Join Conditions

```{r join_data}
# Normalize importance to % share within each model so that
# values are on a common scale (raw impurity importance is not
# comparable across independently trained forests).
dir_hyp <- dir_hyp %>% mutate(importance = 100 * importance / sum(importance))
dir_nor <- dir_nor %>% mutate(importance = 100 * importance / sum(importance))

cat("Hypoxia importance sums to:", sum(dir_hyp$importance), "%\n")
cat("Normoxia importance sums to:", sum(dir_nor$importance), "%\n")

# Inner join directionality tables on feature
joined <- inner_join(
  dir_hyp %>% dplyr::select(feature, importance_hyp = importance, correlation_hyp = correlation,
                     direction_hyp = direction, diff_hyp = diff,
                     wilcox_p_adj_hyp = wilcox_p_adj),
  dir_nor %>% dplyr::select(feature, importance_nor = importance, correlation_nor = correlation,
                     direction_nor = direction, diff_nor = diff,
                     wilcox_p_adj_nor = wilcox_p_adj),
  by = "feature"
)

# Add feature_clean and category from lookup
joined <- joined %>%
  left_join(feature_lookup, by = "feature")

# Compute delta importance (hypoxia minus normoxia)
joined <- joined %>%
  mutate(
    delta_importance = importance_hyp - importance_nor,
    delta_sign = ifelse(delta_importance >= 0, "Higher in Hypoxia", "Higher in Normoxia")
  ) %>%
  arrange(desc(importance_hyp + importance_nor))

cat("Features in joined table:", nrow(joined), "\n")
cat("Features only in hypoxia:", sum(!dir_hyp$feature %in% dir_nor$feature), "\n")
cat("Features only in normoxia:", sum(!dir_nor$feature %in% dir_hyp$feature), "\n")
```

# Plots

## 1. Scatter Plot: Hypoxia vs Normoxia Importance

```{r scatter_plot, fig.width=10, fig.height=9}
# Label top features by distance from diagonal (abs delta = residual from y=x)
top_n_label <- 4
label_data <- joined %>%
  slice_max(abs(delta_importance), n = top_n_label)

p_scatter <- ggplot(joined, aes(x = importance_nor, y = importance_hyp)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60") +
  geom_point(aes(color = delta_sign), size = 3, alpha = 0.85) +
  scale_color_manual(
    values = c("Higher in Hypoxia" = "#E69F00", "Higher in Normoxia" = "#0072B2"),
    name = NULL
  ) +
  ggrepel::geom_text_repel(
    data = label_data,
    aes(label = feature_clean),
    size = 3, max.overlaps = 20, segment.color = "grey50"
  ) +
  theme_cowplot() +
  labs(
    title = "Variable Importance: Hypoxia vs Normoxia",
    subtitle = subtitle_label,
    x = "Normoxia: % of total importance",
    y = "Hypoxia: % of total importance"
  )

print(p_scatter)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_scatter, width = 10, height = 9)


```

## 2. Signed Importance Quadrant Plot

```{r signed_importance_quadrant, fig.width=10, fig.height=9}
# Signed importance = % share × sign(correlation)
# Positive = feature higher in eIF3d targets; negative = lower in eIF3d targets
quad_levels <- c(
  "Up in both\n(same direction)",
  "Down in both\n(same direction)",
  "Discordant\n(hyp+, nor−)",
  "Discordant\n(hyp−, nor+)"
)

quad_data <- joined %>%
  mutate(
    signed_hyp = importance_hyp * sign(correlation_hyp),
    signed_nor = importance_nor * sign(correlation_nor),
    quadrant = factor(case_when(
      signed_nor >= 0 & signed_hyp >= 0 ~ "Up in both\n(same direction)",
      signed_nor <  0 & signed_hyp <  0 ~ "Down in both\n(same direction)",
      signed_nor <  0 & signed_hyp >= 0 ~ "Discordant\n(hyp+, nor−)",
      signed_nor >= 0 & signed_hyp <  0 ~ "Discordant\n(hyp−, nor+)"
    ), levels = quad_levels)
  )

quad_colors <- c(
  "Up in both\n(same direction)"  = "#009E73",
  "Down in both\n(same direction)"   = "#D55E00",
  "Discordant\n(hyp+, nor−)"             = "#CC79A7",
  "Discordant\n(hyp−, nor+)"             = "#0072B2"
)

quad_labels <- bind_rows(
  quad_data %>% filter(signed_nor >= 0, signed_hyp >= 0) %>%
    slice_max(sqrt(signed_hyp^2 + signed_nor^2), n = 1),
  quad_data %>% filter(signed_nor <  0, signed_hyp <  0) %>%
    slice_max(sqrt(signed_hyp^2 + signed_nor^2), n = 1),
  quad_data %>% filter(signed_nor <  0, signed_hyp >= 0) %>%
    slice_max(sqrt(signed_hyp^2 + signed_nor^2), n = 1),
  quad_data %>% filter(signed_nor >= 0, signed_hyp <  0) %>%
    slice_max(sqrt(signed_hyp^2 + signed_nor^2), n = 1)
)

p_quad <- ggplot(quad_data, aes(x = signed_nor, y = signed_hyp)) +
  geom_vline(xintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_point(aes(color = quadrant), size = 3, alpha = 0.85) +
  scale_color_manual(values = quad_colors, name = NULL) +
  ggrepel::geom_text_repel(
    data = quad_labels,
    aes(label = feature_clean),
    size = 3, max.overlaps = 20, segment.color = "grey50"
  ) +
  theme_cowplot() +
  labs(
    title = "Signed Importance: Hypoxia vs Normoxia",
    subtitle = subtitle_label,
    x = "Normoxia: signed importance (% share × direction)",
    y = "Hypoxia: signed importance (% share × direction)"
  )

print(p_quad)

ggsave(here("plots", paste0("rf_importance_quadrant_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_quad, width = 10, height = 9)
```

```{r hypoxia_score_barplot, fig.width=8, fig.height=10}
p_hyp_bar <- ggbarplot(
  quad_data,
  x        = "feature_clean",
  y        = "delta_importance",
  fill     = "quadrant",
  palette  = quad_colors,
  sort.val = "asc",
  sort.by.groups = FALSE,
  rotate   = TRUE,
  xlab     = "",
  ylab     = "Hypoxia-Specific Score (% Hypoxia \u2212 % Normoxia)",
  title    = "Feature Importance: Hypoxia-Specific Score",
  subtitle = subtitle_label
) +
  geom_hline(yintercept = 0, color = "grey40", linewidth = 0.4) +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

print(p_hyp_bar)

ggsave(here("plots", paste0("rf_importance_hyp_score_barplot_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_hyp_bar, width = 8, height = 10)
```
 
```{r hypoxia_score_barplot_top30, fig.width=14, fig.height=5}
top30_data <- quad_data %>%
  slice_max(abs(delta_importance), n = 30)

p_hyp_bar_top30 <- ggbarplot(
  top30_data,
  x        = "feature_clean",
  y        = "delta_importance",
  fill     = "quadrant",
  palette  = quad_colors,
  sort.val = "desc",
  sort.by.groups = FALSE,
  rotate   = FALSE,
  xlab     = "",
  ylab     = "Hypoxia-Specific Score (% Hypoxia - % Normoxia)",
  title    = "Feature Importance: Hypoxia-Specific Score (Top 30 by |score|)",
  subtitle = subtitle_label
) +
  geom_hline(yintercept = 0, color = "grey40", linewidth = 0.4) +
  theme(legend.title = element_blank(),
        legend.position = "right",
        legend.key.size = unit(0.5, "cm"),
        legend.key.spacing.y = unit(0.4, "cm"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

print(p_hyp_bar_top30)

ggsave(here("plots", paste0("rf_importance_hyp_score_barplot_top30_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_hyp_bar_top30, width = 12, height = 5)
```

```{r}
# Same scatter, colored by quadrant (directionality-aware)
scatter_quad_labels <- quad_data %>%
  slice_max(abs(delta_importance), n = 8)

p_scatter_quad <- ggplot(quad_data, aes(x = importance_nor, y = importance_hyp)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60") +
  geom_point(aes(color = quadrant), size = 3, alpha = 0.85) +
  scale_color_manual(values = quad_colors, name = NULL) +
  ggrepel::geom_text_repel(
    data = scatter_quad_labels,
    aes(label = feature_clean),
    size = 3, max.overlaps = 20,
    segment.color = "black", segment.linewidth = 0.5,
    segment.alpha = 0.8, min.segment.length = 0,
    box.padding = 0.5
  ) +
  theme_cowplot() +
  labs(
    title = "Variable Importance: Hypoxia vs Normoxia",
    subtitle = subtitle_label,
    x = "Normoxia: % of total importance",
    y = "Hypoxia: % of total importance"
  )

print(p_scatter_quad)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_quad_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_scatter_quad, width = 10, height = 9)

# Same scatter, colored by four broad categories
broad_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1"
)

quad_data <- quad_data %>%
  mutate(broad_category = case_when(
    str_detect(feature, "^clip_")                                          ~ "eIF3 binding evidence",
    str_detect(feature, "^dap5_|^lit_DAP5")                               ~ "Sensitivity to DAP5",
    str_detect(feature, "^4e_resistance|^eif4e_ip|^lit_eIF4E|^lit_eIF4E2|
                          ^lit_mTOR|^lit_phospho_eIF4E|^lit_TOP|
                          ^lit_translation_is|^lit_si_eIF4")               ~ "Sensitivity to eIF4E",
    TRUE                                                                   ~ "Intrinsic mRNA features"
  ))

p_scatter_cat <- ggplot(quad_data, aes(x = importance_nor, y = importance_hyp)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60") +
  geom_point(aes(color = broad_category), size = 3, alpha = 0.85) +
  scale_color_manual(values = broad_colors, name = NULL) +
  ggrepel::geom_text_repel(
    data = scatter_quad_labels,
    aes(label = feature_clean),
    size = 6, max.overlaps = 20,
    segment.color = "black", segment.linewidth = 0.5,
    segment.alpha = 0.8, min.segment.length = 0,
    box.padding = 0.5
  ) +
  theme_cowplot(font_size = 20) +
  labs(
    title = "Variable Importance: Hypoxia vs Normoxia",
    subtitle = subtitle_label,
    x = "Normoxia: % of total importance",
    y = "Hypoxia: % of total importance"
  )

print(p_scatter_cat)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_broad_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_scatter_cat, width = 10, height = 9)
```



## 3. Dumbbell Plot: Paired Importance by Condition

```{r dumbbell_plot, fig.width=10, fig.height=10}
# Top features by max importance across conditions
dumbbell_data <- joined %>%
  mutate(max_importance = pmax(importance_hyp, importance_nor)) %>%
  arrange(desc(max_importance)) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, max_importance))

p_dumbbell <- ggplot(dumbbell_data) +
  geom_segment(aes(x = importance_hyp, xend = importance_nor,
                   y = feature_clean, yend = feature_clean),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(x = importance_hyp, y = feature_clean, color = "Hypoxia"), size = 3) +
  geom_point(aes(x = importance_nor, y = feature_clean, color = "Normoxia"), size = 3) +
  scale_color_manual(
    values = c("Hypoxia" = "#D55E00", "Normoxia" = "#0072B2"),
    name = "Condition"
  ) +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: Paired Importance by Condition",
    subtitle = subtitle_label,
    x = "% of total importance",
    y = ""
  )

print(p_dumbbell)

ggsave(here("plots", paste0("rf_importance_comparison_dumbbell_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_dumbbell, width = 10, height = 10)
```

## 3. Delta Importance Bar Plot

```{r delta_bar_plot, fig.width=10, fig.height=10}
delta_data <- joined %>%
  arrange(desc(abs(delta_importance))) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, delta_importance))

p_delta <- ggplot(delta_data, aes(x = feature_clean, y = delta_importance, fill = delta_sign)) +
  geom_col() +
  scale_fill_manual(
    values = c("Higher in Hypoxia" = "#009E73", "Higher in Normoxia" = "#D55E00"),
    name = ""
  ) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Delta Importance: Hypoxia minus Normoxia",
    subtitle = subtitle_label,
    x = "",
    y = "Delta Relative Importance (Hypoxia - Normoxia)"
  )

print(p_delta)

ggsave(here("plots", paste0("rf_importance_comparison_delta_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_delta, width = 10, height = 10)
```

## 4. Signed-Importance Lollipop by Condition (Top 30)

```{r plot_lollipop, fig.width=10, fig.height=10}
category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

make_lollipop <- function(data, cond_label) {
  imp_col  <- if (cond_label == "hypoxia") "importance_hyp"  else "importance_nor"
  corr_col <- if (cond_label == "hypoxia") "correlation_hyp" else "correlation_nor"

  lollipop_data <- data %>%
    filter(!is.na(.data[[corr_col]])) %>%
    mutate(
      importance        = .data[[imp_col]],
      correlation       = .data[[corr_col]],
      signed_importance = importance * sign(correlation),
      broad_category = case_when(
        category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
        category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
        category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
        category == "Translational Efficiency"                   ~ "Translation in hypoxia",
        TRUE                                                     ~ "Intrinsic mRNA features"
      ),
      broad_category = factor(broad_category, levels = c(
        "Intrinsic mRNA features", "eIF3 binding evidence",
        "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
      ))
    ) %>%
    arrange(desc(importance)) %>%
    head(30) %>%
    mutate(feature_clean = fct_reorder(feature_clean, signed_importance))

  ggdotchart(
    lollipop_data,
    x        = "feature_clean",
    y        = "signed_importance",
    color    = "broad_category",
    palette  = category_colors,
    sorting  = "none",
    add      = "segments",
    rotate   = TRUE,
    dot.size = 4,
    xlab     = "",
    ylab     = "Signed Importance (positive = higher in promotes)",
    title    = paste0("Top 30 Features: Signed Importance (", tools::toTitleCase(cond_label), ")"),
    subtitle = subtitle_label
  ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    theme(legend.title = element_blank())
}

p_lollipop_hyp <- make_lollipop(joined, "hypoxia")
p_lollipop_nor <- make_lollipop(joined, "normoxia")

print(p_lollipop_hyp)
print(p_lollipop_nor)

ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix_hyp, ".pdf")),
       plot = p_lollipop_hyp, width = 10, height = 10)
ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix_nor, ".pdf")),
       plot = p_lollipop_nor, width = 10, height = 10)
```

# Export

```{r export}
output_dir <- here("output", "predictive_modeling")

export_df <- joined %>%
  dplyr::select(feature, feature_clean, category,
         importance_hyp, importance_nor, delta_importance,
         correlation_hyp, correlation_nor,
         direction_hyp, direction_nor,
         diff_hyp, diff_nor,
         wilcox_p_adj_hyp, wilcox_p_adj_nor)

write_csv(export_df, file.path(output_dir,
  paste0("rf_importance_comparison_", suffix_hyp, "_vs_", suffix_nor, ".csv")))

cat("Exported:", nrow(export_df), "features\n")
```

# Gene Set Overlap: Hypoxia vs Normoxia Promotes

```{r venn_genesets, fig.width=6, fig.height=5}
library(ggVennDiagram)

norm_genes <- if (params$timepoint_nor %in% c("1and4hr", "alltimepoints") &&
                  params$condition_nor == "normoxia") {
  read_csv(here("output",
    paste0("translation_categories_si", params$gene, "_vs_sictrl_normoxia_",
           params$timepoint_nor, ".csv")),
    show_col_types = FALSE) %>%
    filter(te_lfc > as.numeric(params$lfc), te_padj < 0.05) %>%
    pull(symbol)
} else {
  read_csv(here("output", "genesets",
    paste0(params$condition_nor, "_", params$gene, "_", params$direction,
           "_TE_", params$timepoint_nor, "_lfc", params$lfc, ".csv")),
    show_col_types = FALSE) %>%
    pull(db_gene_symbol)
}

hyp_genes <- read_csv(
  here("output", "genesets",
       paste0(params$condition_hyp, "_", params$gene, "_", params$direction,
              "_TE_", params$timepoint, "_lfc", params$lfc, ".csv")),
  show_col_types = FALSE
) %>% pull(db_gene_symbol)

cat("Normoxia promotes:", length(norm_genes), "genes\n")
cat("Hypoxia promotes:", length(hyp_genes), "genes\n")
cat("Shared:", length(intersect(norm_genes, hyp_genes)), "genes\n")

venn_sets <- setNames(
  list(norm_genes, hyp_genes),
  c(paste0(params$condition_nor, "\npromotes"),
    paste0(params$condition_hyp, "\npromotes"))
)

p_venn <- ggVennDiagram(venn_sets, label_alpha = 0,
                        set_color = c("#D55E00", "#009E73")) +
  scale_fill_gradient(low = "white", high = "white") +
  labs(title = paste0("eIF", params$gene, " ", params$direction,
                      " gene set overlap\n",
                      params$timepoint, ", lfc", params$lfc)) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

print(p_venn)

ggsave(here("plots", paste0("venn_geneset_overlap_", suffix_hyp, "_vs_", suffix_nor, ".pdf")),
       plot = p_venn, width = 11, height = 10)
```

# Session Info

```{r session_info}
sessionInfo()
```
