---
title: "`r paste0('Random Forest Classification: eIF4E Target Prediction (', params$feature_set, ', clip=', params$clip_mode, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_set: "kozak"
  clip_mode: "aggregate"
  pos_geneset_csv: ""
  neg_geneset_csv: ""
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
# --- COMMON ANALYSIS SETUPS ---
#
# Standard eIF4E target analysis (Teleman riboseq pos/neg):
#   feature_set="external_stability", clip_mode="aggregate",
#   pos_geneset_csv="", neg_geneset_csv=""
#
# eIF3d-promoted vs eIF3e-promoted at 1hr hypoxia:
#   feature_set="external_stability", clip_mode="aggregate",
#   pos_geneset_csv="hypoxia_3d_promotes_TE_1hr_lfc0.5.csv",
#   neg_geneset_csv="hypoxia_3e_promotes_TE_1hr_lfc0.5.csv"
#   Note: eIF4E circular features are NOT removed for custom geneset runs
#         (they are informative features, not circular, for 3d vs 3e)
#
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_set    = "external_stability", # "external_stability" | "kozak" | "te_lfc" | "codon_optimality"
    clip_mode      = "aggregate",          # "aggregate" | "individual" | "both"
    pos_geneset_csv = "",                  # "" = use Teleman eIF4E data (default)
                                           # "filename.csv" = load from output/genesets/
    neg_geneset_csv = ""                   # "" = use Teleman neutral genes (default)
                                           # "filename.csv" = load from output/genesets/
  )
}

feature_matrix_rds <- switch(params$feature_set,
  "te_lfc"             = "feature_matrix_te_lfc.rds",
  "codon_optimality"   = "feature_matrix_codon_optimality.rds",
  "kozak"              = "feature_matrix_kozak.rds",
  "targeted"           = "feature_matrix_te_lfc.rds",
  "external_stability" = "feature_matrix_external_stability.rds",
  stop("feature_set '", params$feature_set, "' is not supported in this notebook")
)

using_custom_genesets <- params$pos_geneset_csv != ""

if (using_custom_genesets) {
  pos_name  <- tools::file_path_sans_ext(params$pos_geneset_csv)
  neg_name  <- tools::file_path_sans_ext(params$neg_geneset_csv)
  suffix    <- paste0(pos_name, "_vs_", neg_name, "_", params$feature_set,
                      "_clip", params$clip_mode, "_no_boruta")
  pos_label <- pos_name
  neg_label <- neg_name
  gene_label <- "custom"
} else {
  suffix    <- paste0("eif4e_promotes_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")
  pos_label <- "promotes_eif4e"
  neg_label <- "does_not_promote_eif4e"
  gene_label <- "eIF4E"
}

cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix:", suffix, "\n")
cat("Positive label:", pos_label, "\n")
cat("Negative label:", neg_label, "\n")
```

# Overview

Binary classification without Boruta feature selection: does a gene's mRNA sequence predict whether it depends on eIF4E for translation?

Training data come from the Teleman riboseq dataset (4E-BP1 overexpression vs. vehicle). 4E-BP1 sequesters eIF4E, so genes that go **down** when 4E-BP1 is overexpressed are **eIF4E-dependent** (positive class).

- **Positive class (`"promotes_eif4e"`)**: Genes with `log2FoldChange < -1` AND `padj < 0.05` in the Teleman dataset — translation is reduced when eIF4E is inhibited, i.e. eIF4E **promotes** their translation (~716 genes).
- **Negative class (`"does_not_promote_eif4e"`)**: Genes unaffected by eIF4E status (`|LFC| < 0.1` AND `padj > 0.5`) — neutral controls that serve as the reference (~2,140 genes, downsampled to match positive class).

**Note on circular features**: The following are excluded to prevent data leakage — they directly measure eIF4E dependence and would circularly predict the training labels: `4e_resistance` scores, `eif4e_ip` binding features, `lit_eIF4E_activated_translation`, `lit_phospho_eIF4E_*`, `lit_translation_is_sensitive_to_4E_inhibition`, and `lit_translation_is_resistant_to_4E_inhibition`.

## Parameters

| Parameter | Values | Description |
|-----------|--------|-------------|
| `feature_set` | `"te_lfc"`, `"codon_optimality"`, `"kozak"` | Which feature matrix to use |
| `clip_mode` | `"individual"`, `"aggregate"`, `"both"` | Which CLIP features to use: `"individual"` = per-dataset regional binary features; `"aggregate"` = single numeric count of datasets with binding evidence (0–7); `"both"` = all CLIP features |

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
# Feature matrix
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

if (using_custom_genesets) {
  # Custom gene sets from output/genesets/
  positive_genes <- read_csv(here("output", "genesets", params$pos_geneset_csv),
                             show_col_types = FALSE) %>%
    rename(ensembl_gene = ensembl_gene)  # already has ensembl_gene column
  cat("Positive genes (", params$pos_geneset_csv, "):", nrow(positive_genes), "\n")

  negative_control_genes <- read_csv(here("output", "genesets", params$neg_geneset_csv),
                                     show_col_types = FALSE) %>%
    rename(gene_id_clean = ensembl_gene)
  cat("Negative genes (", params$neg_geneset_csv, "):", nrow(negative_control_genes), "\n")
} else {
  # Teleman riboseq: 4E-BP1 overexpression vs vehicle
  # LFC = log2(4E-BP1 OE / vehicle)
  # Negative LFC = less translation when eIF4E is inhibited = eIF4E-dependent
  teleman_data <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv")
  cat("Teleman dataset:", nrow(teleman_data), "genes\n")

  # Positive class: eIF4E-sensitive genes (go DOWN when eIF4E is inhibited)
  positive_genes <- teleman_data %>%
    filter(log2FoldChange < -1, padj < 0.05) %>%
    mutate(ensembl_gene = gene_id)  # already clean (no version suffix)
  cat("Positive genes (LFC < -1, padj < 0.05):", nrow(positive_genes), "\n")

  # Negative class: genes unaffected by eIF4E status (middle of distribution)
  negative_control_genes <- teleman_data %>%
    filter(abs(log2FoldChange) < 0.1, padj > 0.5) %>%
    mutate(gene_id_clean = gene_id)
  cat("Negative control genes (|LFC| < 0.1, padj > 0.5):", nrow(negative_control_genes), "\n")
}
```

# Sanity Checks

```{r sanity_checks}
if (!using_custom_genesets) {
  # GLO1: known eIF4E-sensitive gene — should be IN positive set (LFC ≈ -1.29, padj ≈ 8.5e-5)
  glo1 <- teleman_data %>% filter(symbol == "GLO1")
  cat("GLO1:\n")
  print(glo1 %>% dplyr::select(symbol, log2FoldChange, padj))
  cat("GLO1 in positive set:", glo1$gene_id %in% positive_genes$ensembl_gene, "\n\n")

  # CDKN2B: known eIF4E-resistant gene — should NOT be in positive set (LFC ≈ +2.48, padj ≈ 8.7e-6)
  cdkn2b <- teleman_data %>% filter(symbol == "CDKN2B")
  cat("CDKN2B:\n")
  print(cdkn2b %>% dplyr::select(symbol, log2FoldChange, padj))
  cat("CDKN2B in positive set:", cdkn2b$gene_id %in% positive_genes$ensembl_gene, "\n")
} else {
  cat("Custom gene sets — skipping eIF4E sanity checks\n")
  cat("Positive set:", params$pos_geneset_csv, "(", nrow(positive_genes), "genes )\n")
  cat("Negative set:", params$neg_geneset_csv, "(", nrow(negative_control_genes), "genes )\n")
}
```

# Create Binary Target

```{r create_target}
set.seed(9)

# Only consider genes present in the feature matrix
pos_in_features <- positive_genes %>%
  filter(ensembl_gene %in% features$gene_id_clean)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

neg_in_features <- negative_control_genes %>%
  filter(gene_id_clean %in% features$gene_id_clean)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))

# Label genes
features <- features %>%
  mutate(eif4e_target = case_when(
    gene_id_clean %in% positive_genes$ensembl_gene ~ pos_label,
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ neg_label,
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif4e_target)) %>%
  mutate(eif4e_target = factor(eif4e_target, levels = c(neg_label, pos_label)))

nrow(model_genes)

cat("Positive (", pos_label, "):", sum(model_genes$eif4e_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif4e_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
cat("Imbalance ratio (neg:pos):",
    round(sum(model_genes$eif4e_target == neg_label) /
          sum(model_genes$eif4e_target == pos_label), 1), ":1\n")
```

# Preprocessing

## Filter Features (No Boruta)

```{r filter_features}
# Identify feature columns (exclude ID/target cols)
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif4e_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count (all combined):", length(feature_cols), "\n")

# ---- CLIP feature mode ----
if (params$clip_mode == "aggregate") {
  aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                           "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                           "clip_lee_glu_vs_cm_enrichment")
  regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
  feature_cols <- setdiff(feature_cols, regional_clip)
  cat("CLIP mode: aggregate — using aggregate CLIP counts only, dropped", length(regional_clip), "regional features\n")
} else if (params$clip_mode == "individual") {
  feature_cols <- setdiff(feature_cols, c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                                          "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                                          "clip_lee_glu_vs_cm_enrichment"))
  cat("CLIP mode: individual — using regional binary features, dropped aggregate CLIP counts\n")
} else {
  cat("CLIP mode: both — using all CLIP features\n")
}

# If targeted/codon_optimality/kozak/te_lfc: remove k-mer and RBP motif columns
if (params$feature_set %in% c("targeted", "structure", "clip_split", "te_lfc", "codon_optimality", "kozak")) {
  kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
  excluded <- feature_cols[str_detect(feature_cols, kmer_motif_pattern)]
  feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]
  cat("Targeted mode: removed", length(excluded), "k-mer/motif columns\n")
}

# ---- Remove circular features ----
# Only applies when using Teleman eIF4E labels — these features directly measure
# eIF4E dependence and would circularly predict those training labels.
# When using custom gene sets (e.g. 3d vs 3e), eIF4E features are NOT circular
# and are kept as informative predictors.
if (!using_custom_genesets) {
  circular_regex <- "^4e_resistance|^eif4e_ip"
  circular_lit_exact <- c(
    "lit_eIF4E_activated_translation",
    "lit_phospho_eIF4E_activated_translation",
    "lit_phospho_eIF4E_suppressed_translation",
    "lit_translation_is_sensitive_to_4E_inhibition",
    "lit_translation_is_resistant_to_4E_inhibition"
  )
  circular_cols <- feature_cols[str_detect(feature_cols, circular_regex) | feature_cols %in% circular_lit_exact]
  if (length(circular_cols) > 0) {
    feature_cols <- setdiff(feature_cols, circular_cols)
    cat("Removed", length(circular_cols), "circular features:\n")
    cat(paste(" ", circular_cols, collapse = "\n"), "\n")
  } else {
    cat("No circular 4E features found in this feature set\n")
  }
} else {
  cat("Custom gene sets: skipping circular feature removal (eIF4E features are informative here)\n")
}

cat("Feature count after filtering:", length(feature_cols), "\n")
```

## Median-Impute Missing Values

```{r impute_missing}
missing_counts <- model_genes %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:", nrow(missing_counts), "\n")
if (nrow(missing_counts) > 0) print(missing_counts)
cat("(Imputation deferred to after train/test split — medians computed from training set only)\n")
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif4e_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
cat("\nClass distribution:\n")
print(table(model_data$eif4e_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif4e_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif4e_target))
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif4e_target))

# Compute medians from training set only, then apply to both splits
train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

cat("\nRemaining NAs in train:", sum(is.na(train_data[, feature_cols])), "\n")
cat("Remaining NAs in test:", sum(is.na(test_data[, feature_cols])), "\n")
```

# Train Random Forest

```{r train_rf}
# Classes are already balanced (neg sampled to match pos count), so no case weights needed.
set.seed(9)
rf_model <- ranger(
  x = train_data[, feature_cols],
  y = train_data$eif4e_target,
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  seed = 9
)

print(rf_model)
```

# Evaluate on Test Set

## ROC Curve and AUC

```{r roc_curve, fig.width=8, fig.height=6}
# Predictions
pred_probs <- predict(rf_model, test_data[, feature_cols])$predictions
pred_pos_prob <- pred_probs[, pos_label]
actual <- test_data$eif4e_target

# ROC on test set (threshold-free AUC)
roc_obj <- roc(response = actual, predictor = pred_pos_prob,
               levels = c(neg_label, pos_label), direction = "<")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

# Bootstrap 95% CI on AUC (pROC built-in)
ci_auc <- ci.auc(roc_obj, method = "bootstrap", boot.n = 1000)
cat("AUC 95% CI (bootstrap):", round(ci_auc[1], 4), "-", round(ci_auc[3], 4), "\n")

# Youden threshold from OOB predictions on training set (avoids test set snooping)
oob_probs <- rf_model$predictions[, pos_label]
roc_oob <- roc(response = train_data$eif4e_target, predictor = oob_probs,
               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
coords_youden <- coords(roc_oob, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")
if (nrow(coords_youden) > 1) coords_youden <- coords_youden[1, ]  # take first if ties
optimal_threshold <- coords_youden$threshold

cat("Youden threshold (from training OOB):", round(optimal_threshold, 4), "\n")
cat("  OOB Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  OOB Specificity:", round(coords_youden$specificity, 3), "\n")

# Plot
roc_df <- data.frame(sensitivity = roc_obj$sensitivities,
                     specificity = roc_obj$specificities)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = paste0("ROC Curve: ", gene_label, " Target Classification"),
    subtitle = paste0("AUC = ", round(auc_value, 3),
                      " [", round(ci_auc[1], 3), "-", round(ci_auc[3], 3), "]",
                      " | Threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

print(p_roc)

ggsave(here("plots", paste0("rf_classification_roc_", suffix, ".pdf")),
       plot = p_roc, width = 8, height = 6)
```

## Confusion Matrix

```{r confusion_matrix}
pred_class <- factor(
  ifelse(pred_pos_prob >= optimal_threshold, pos_label, neg_label),
  levels = c(neg_label, pos_label)
)

cm <- confusionMatrix(pred_class, actual, positive = pos_label)
print(cm)

# Extract metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\nKey Metrics (Youden Threshold):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n")
```

## Confusion Matrix Plot

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = "Confusion Matrix",
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3),
                      ", F1 = ", round(f1, 3))
  ) +
  theme(legend.position = "none")

print(p_cm)

ggsave(here("plots", paste0("rf_classification_confusion_matrix_", suffix, ".pdf")),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(train_data$eif4e_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- train_data[-test_idx, ]
  test_cv <- train_data[test_idx, ]

  # Impute using this fold's training data medians only
  cv_medians <- train_cv %>%
    dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~median(., na.rm = TRUE)))
  train_cv <- impute_with_medians(train_cv, cv_medians)
  test_cv  <- impute_with_medians(test_cv,  cv_medians)

  # Train (classes already balanced, no case weights needed)
  rf_cv <- ranger(
    x = train_cv[, feature_cols],
    y = train_cv$eif4e_target,
    num.trees = 500,
    probability = TRUE,
    seed = 9
  )

  # Youden threshold from OOB predictions on this fold's training data
  oob_probs_cv <- rf_cv$predictions[, pos_label]
  roc_oob_cv <- roc(response = train_cv$eif4e_target, predictor = oob_probs_cv,
                    levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  coords_cv <- coords(roc_oob_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                      best.method = "youden")
  if (nrow(coords_cv) > 1) coords_cv <- coords_cv[1, ]  # take first if ties

  # Predict on held-out fold
  pred_probs_cv <- predict(rf_cv, test_cv[, feature_cols])$predictions[, pos_label]

  # ROC on test fold (for AUC)
  roc_cv <- roc(test_cv$eif4e_target, pred_probs_cv,
                levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)

  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= coords_cv$threshold, pos_label, neg_label),
    levels = c(neg_label, pos_label)
  )
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif4e_target, positive = pos_label)

  data.frame(
    fold = i,
    Threshold = coords_cv$threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n5-Fold CV Results:\n")
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("AUC", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

# Permutation Test

```{r permutation_test, fig.width=8, fig.height=5}
set.seed(42)
n_perm <- 100
perm_aucs <- numeric(n_perm)

for (i in seq_len(n_perm)) {
  if (i %% 10 == 0) cat("Permutation", i, "/", n_perm, "\n")
  perm_data <- model_data
  perm_data$eif4e_target <- sample(perm_data$eif4e_target)
  perm_train <- impute_with_medians(perm_data[train_idx, ], train_medians)
  perm_test  <- impute_with_medians(perm_data[-train_idx, ], train_medians)
  rf_perm <- ranger(x = perm_train[, feature_cols], y = perm_train$eif4e_target,
                    num.trees = 500, probability = TRUE, seed = i)
  perm_probs <- predict(rf_perm, perm_test[, feature_cols])$predictions[, pos_label]
  roc_perm <- roc(perm_test$eif4e_target, perm_probs,
                  levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  perm_aucs[i] <- as.numeric(auc(roc_perm))
}

perm_pvalue <- mean(perm_aucs >= as.numeric(auc_value))
cat("Permutation p-value:", perm_pvalue, "\n")

p_perm <- ggplot(data.frame(auc = perm_aucs), aes(x = auc)) +
  geom_histogram(bins = 30, fill = "grey70", color = "white") +
  geom_vline(xintercept = as.numeric(auc_value), color = "#E69F00", linewidth = 1.2) +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  annotate("text", x = as.numeric(auc_value), y = Inf,
           label = paste0("Real AUC\n", round(auc_value, 3)),
           vjust = 2, hjust = -0.1, color = "#E69F00", size = 3.5) +
  theme_cowplot() +
  labs(title = "Permutation Test: Null AUC Distribution",
       subtitle = paste0("n=", n_perm, " permutations | p = ", perm_pvalue),
       x = "AUC (shuffled labels)", y = "Count")

print(p_perm)

ggsave(here("plots", paste0("rf_classification_permutation_test_", suffix, ".pdf")),
       plot = p_perm, width = 8, height = 5)
```

# Calibration Plot

```{r calibration_plot, fig.width=6, fig.height=6}
cal_df <- data.frame(
  pred = pred_pos_prob,
  actual = as.numeric(actual == pos_label)
)
cal_df$bin <- cut(cal_df$pred, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
cal_summary <- cal_df %>%
  group_by(bin) %>%
  summarise(mean_pred = mean(pred), fraction_pos = mean(actual), n = n(), .groups = "drop") %>%
  filter(!is.na(bin))

p_cal <- ggplot(cal_summary, aes(x = mean_pred, y = fraction_pos)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(size = n), color = "#0072B2") +
  geom_line(color = "#0072B2") +
  scale_size_continuous(name = "n in bin") +
  xlim(0, 1) + ylim(0, 1) +
  theme_cowplot() +
  labs(title = "Calibration Plot",
       subtitle = paste0(gene_label, " target classifier — test set"),
       x = "Mean Predicted Probability", y = "Observed Fraction Positive")

print(p_cal)

ggsave(here("plots", paste0("rf_classification_calibration_", suffix, ".pdf")),
       plot = p_cal, width = 6, height = 6)
```

# Learning Curve

```{r learning_curve, fig.width=8, fig.height=5}
set.seed(9)
lc_fracs <- seq(0.1, 1.0, by = 0.1)

lc_results <- lapply(lc_fracs, function(frac) {
  if (frac < 1.0) {
    idx <- createDataPartition(train_data$eif4e_target, p = frac, list = FALSE)
    sub_train <- train_data[idx, ]
  } else {
    sub_train <- train_data
  }
  rf_lc <- ranger(x = sub_train[, feature_cols], y = sub_train$eif4e_target,
                  num.trees = 200, probability = TRUE, seed = 9)
  oob_probs_lc <- rf_lc$predictions[, pos_label]
  roc_train_lc <- roc(sub_train$eif4e_target, oob_probs_lc,
                      levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  test_probs_lc <- predict(rf_lc, test_data[, feature_cols])$predictions[, pos_label]
  roc_test_lc <- roc(test_data$eif4e_target, test_probs_lc,
                     levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  data.frame(frac = frac, n_train = nrow(sub_train),
             train_oob_auc = as.numeric(auc(roc_train_lc)),
             test_auc = as.numeric(auc(roc_test_lc)))
})

lc_df <- bind_rows(lc_results)

p_lc <- lc_df %>%
  pivot_longer(c(train_oob_auc, test_auc), names_to = "set", values_to = "auc") %>%
  mutate(set = recode(set, train_oob_auc = "Train (OOB)", test_auc = "Test")) %>%
  ggplot(aes(x = n_train, y = auc, color = set, group = set)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Train (OOB)" = "#E69F00", "Test" = "#0072B2")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  theme_cowplot() +
  labs(title = "Learning Curve", subtitle = paste0(gene_label, " target classifier"),
       x = "Training Set Size", y = "AUC", color = NULL)

print(p_lc)

ggsave(here("plots", paste0("rf_classification_learning_curve_", suffix, ".pdf")),
       plot = p_lc, width = 8, height = 5)
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}

importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Clean names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

# Bar plot
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top 30 Features: ", gene_label, " Target Classification"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", paste0("rf_classification_feature_importance_", suffix, ".pdf")),
       plot = p_importance, width = 10, height = 10)

# All-features bar plot (cap at top 40)
importance_all_plot <- importance_df %>% head(40)
all_height <- max(6, nrow(importance_all_plot) * 0.35)

p_importance_all <- importance_all_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top ", nrow(importance_all_plot), " Features: ", gene_label, " Target Classification"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_all)

ggsave(here("plots", paste0("rf_classification_feature_importance_all_", suffix, ".pdf")),
       plot = p_importance_all, width = 10, height = all_height)

# Custom-features-only bar plot (exclude k-mer and RBP motif features, cap at top 40)
importance_custom <- importance_df %>%
  filter(!str_detect(feature, "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbp_|^rbns_|^cisbp_")) %>%
  head(40)

custom_height <- max(6, nrow(importance_custom) * 0.35)

p_importance_custom <- importance_custom %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Custom Features: ", gene_label, " Target Classification"),
    subtitle = "Variable Importance (Impurity) — CLIP, 4E/DAP5, Signatures, Lengths, GC",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_custom)

ggsave(here("plots", paste0("rf_classification_feature_importance_custom_", suffix, ".pdf")),
       plot = p_importance_custom, width = 10, height = custom_height)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^struct_")                          ~ "RNA Structure",
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      str_detect(feature, "^indiv_te_|^delta_te_")            ~ "Translational Efficiency",
      str_detect(feature, "^cai$|^tai$|^csc$")               ~ "Codon Optimality",
      TRUE                                                    ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    top_feature = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nImportance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance by Category: ", gene_label, " Target Classification"),
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", paste0("rf_classification_importance_by_category_", suffix, ".pdf")),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Model
saveRDS(rf_model, file.path(output_dir, paste0("rf_classifier_", suffix, ".rds")))

# Feature importance
write_csv(importance_df, file.path(output_dir, paste0("rf_classification_importance_", suffix, ".csv")))

# Performance summary
n_pos_total <- sum(model_genes$eif4e_target == pos_label)
n_neg_total <- sum(model_genes$eif4e_target == neg_label)

performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1", "N_pos", "N_neg",
             "AUC_CI_lower", "AUC_CI_upper", "permutation_pvalue"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy,
               sensitivity, specificity, precision, f1,
               n_pos_total, n_neg_total,
               as.numeric(ci_auc[1]), as.numeric(ci_auc[3]), perm_pvalue),
  CV_Mean = c(mean(cv_results_df$AUC), NA, mean(cv_results_df$Balanced_Accuracy),
              mean(cv_results_df$Sensitivity), mean(cv_results_df$Specificity),
              mean(cv_results_df$Precision, na.rm = TRUE),
              mean(cv_results_df$F1, na.rm = TRUE), NA, NA, NA, NA, NA),
  CV_SD = c(sd(cv_results_df$AUC), NA, sd(cv_results_df$Balanced_Accuracy),
            sd(cv_results_df$Sensitivity), sd(cv_results_df$Specificity),
            sd(cv_results_df$Precision, na.rm = TRUE),
            sd(cv_results_df$F1, na.rm = TRUE), NA, NA, NA, NA, NA)
)
write_csv(performance_df, file.path(output_dir, paste0("rf_classification_performance_", suffix, ".csv")))

# CV results
write_csv(cv_results_df, file.path(output_dir, paste0("rf_classification_cv_results_", suffix, ".csv")))

# Predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_pos = pred_pos_prob,
  prob_neg = pred_probs[, neg_label]
)
write_csv(predictions_export, file.path(output_dir, paste0("rf_classification_predictions_", suffix, ".csv")))

# Permutation AUCs
write_csv(data.frame(perm_auc = perm_aucs),
          file.path(output_dir, paste0("rf_classification_permutation_aucs_", suffix, ".csv")))

# Learning curve
write_csv(lc_df, file.path(output_dir, paste0("rf_classification_learning_curve_", suffix, ".csv")))

cat("Exported: model RDS, importance CSV, performance CSV, CV results CSV, predictions CSV, permutation AUCs CSV, learning curve CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat("CLASSIFICATION SUMMARY: eIF4E TARGETS\n")
cat("========================================\n\n")

cat("Parameters:\n")
cat("  Feature set:", params$feature_set, "(no Boruta selection)\n")
cat("  CLIP mode:", params$clip_mode, "\n\n")

cat("Training data: Teleman riboseq (4E-BP1 OE vs vehicle)\n")
cat("  Positive (", pos_label, ") = LFC < -1, padj < 0.05 (eIF4E-dependent)\n")
cat("  Negative (", neg_label, ") = |LFC| < 0.1, padj > 0.5 (unaffected)\n\n")

cat("Classes:\n")
cat("  Positive (", pos_label, "):", sum(model_data$eif4e_target == pos_label), "\n")
cat("  Negative (", neg_label, "):", sum(model_data$eif4e_target == neg_label), "\n")
cat("  Features:", length(feature_cols), "\n\n")

cat("Test Set Performance (Youden threshold =", round(optimal_threshold, 4), "):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  AUC 95% CI (bootstrap):", round(ci_auc[1], 4), "-", round(ci_auc[3], 4), "\n")
cat("  Permutation p-value:", perm_pvalue, "(n=100)\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n\n")

cat("5-Fold CV:\n")
cat("  AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-",
    round(sd(cv_results_df$Balanced_Accuracy), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

cat("========================================\n")
```

# Final Feature Importance Plot

```{r final_importance_plot, fig.width=8, fig.height=max(4, min(40, length(feature_cols)) * 0.35)}
# Lollipop plot of top features (cap at 40)
importance_final_plot <- importance_df %>% head(40)
final_height <- max(4, nrow(importance_final_plot) * 0.35)

p_final_importance <- importance_final_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_segment(aes(xend = reorder(feature_clean, importance), y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(size = 3, color = "#0072B2") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Feature Importance: ", gene_label, " Target Classification"),
    subtitle = paste0(params$feature_set, " | AUC = ", round(as.numeric(auc_value), 3)),
    x = "",
    y = "Variable Importance (Impurity)"
  )

print(p_final_importance)

ggsave(here("plots", paste0("rf_final_importance_", suffix, ".pdf")),
       plot = p_final_importance, width = 8,
       height = final_height)
```

# Feature Directionality

For each feature, computes point-biserial correlations and Wilcoxon rank-sum tests to determine whether the feature is higher or lower in the `promotes_eif4e` class, then renders signed-importance lollipop plots colored by feature category.

## Compute Directionality

```{r compute_directionality}
# Binary numeric target: 0 = does_not_promote_eif4e, 1 = promotes_eif4e
target_numeric <- as.numeric(model_genes$eif4e_target) - 1

directionality <- importance_df %>%
  rowwise() %>%
  mutate(
    mean_promotes         = mean(model_genes[[feature]][target_numeric == 1], na.rm = TRUE),
    mean_does_not_promote = mean(model_genes[[feature]][target_numeric == 0], na.rm = TRUE),
    diff = mean_promotes - mean_does_not_promote,
    correlation = tryCatch(
      cor(model_genes[[feature]], target_numeric, use = "complete.obs"),
      error = function(e) NA_real_
    ),
    wilcox_p = tryCatch({
      wt <- wilcox.test(model_genes[[feature]][target_numeric == 1],
                        model_genes[[feature]][target_numeric == 0])
      wt$p.value
    }, error = function(e) NA_real_)
  ) %>%
  ungroup() %>%
  mutate(
    wilcox_p_adj = p.adjust(wilcox_p, method = "BH"),
    direction = case_when(
      is.na(correlation) ~ NA_character_,
      correlation >= 0   ~ "Higher in promotes",
      TRUE               ~ "Lower in promotes"
    )
  ) %>%
  arrange(desc(importance))

cat("Directionality computed for", nrow(directionality), "features\n")
cat("Higher in promotes:", sum(directionality$direction == "Higher in promotes", na.rm = TRUE), "\n")
cat("Lower in promotes:", sum(directionality$direction == "Lower in promotes", na.rm = TRUE), "\n")
cat("Significant (Wilcox BH p < 0.05):", sum(directionality$wilcox_p_adj < 0.05, na.rm = TRUE), "\n")
```

## Lollipop: Top 30 Features

```{r plot_lollipop, fig.width=10, fig.height=10}
category_colors <- c(
  "Intrinsic mRNA features"  = "#3D9C60",
  "eIF3/cap CLIP binding"    = "#E6A817",
  "DAP5 regulation"          = "#D668C6",
  "Literature signatures"    = "#4A7FC1",
  "Translational efficiency" = "#E07030"
)

make_broad_category <- function(df) {
  df %>%
    mutate(
      broad_category = case_when(
        category == "4E / DAP5"               ~ "DAP5 regulation",
        category == "CLIP Binding"             ~ "eIF3/cap CLIP binding",
        category == "Literature Signatures"    ~ "Literature signatures",
        category == "Translational Efficiency" ~ "Translational efficiency",
        TRUE                                   ~ "Intrinsic mRNA features"
      ),
      broad_category = factor(broad_category, levels = names(category_colors)),
      signed_importance = importance * sign(correlation),
      feature_clean = fct_reorder(feature_clean, signed_importance)
    )
}

lollipop_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  head(30) %>%
  make_broad_category()

p_lollipop <- ggdotchart(
  lollipop_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 4,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher in promotes_eif4e)",
  title    = "Top 30 Features: Signed Importance + Directionality",
  subtitle = paste0(gene_label, " | ", params$feature_set, " | AUC = ", round(as.numeric(auc_value), 3))
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop)

ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix, ".pdf")),
       plot = p_lollipop, width = 10, height = 10)
```

## Lollipop: All Features

```{r plot_lollipop_all, fig.width=10, fig.height=ceiling(nrow(directionality %>% filter(!is.na(correlation))) * 0.28)}
lollipop_all_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  make_broad_category()

plot_height_all <- max(10, nrow(lollipop_all_data) * 0.28)

p_lollipop_all <- ggdotchart(
  lollipop_all_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 3,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher in promotes_eif4e)",
  title    = paste0("All Features: Signed Importance (n = ", nrow(lollipop_all_data), ")"),
  subtitle = paste0(gene_label, " | ", params$feature_set, " | AUC = ", round(as.numeric(auc_value), 3))
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop_all)

ggsave(here("plots", paste0("rf_directionality_lollipop_all_", suffix, ".pdf")),
       plot = p_lollipop_all, width = 10, height = plot_height_all)
```

## Export Directionality Table

```{r export_directionality}
export_dir_df <- directionality %>%
  mutate(signed_importance = importance * sign(correlation)) %>%
  dplyr::select(feature, feature_clean, category, mean_promotes, mean_does_not_promote,
                diff, correlation, wilcox_p, wilcox_p_adj, importance, signed_importance, direction)

write_csv(export_dir_df,
          here("output", "predictive_modeling",
               paste0("rf_feature_directionality_", suffix, ".csv")))

cat("Exported: rf_feature_directionality_", suffix, ".csv\n")
cat("  Rows:", nrow(export_dir_df), "\n")
cat("  Significant (BH p < 0.05):", sum(export_dir_df$wilcox_p_adj < 0.05, na.rm = TRUE), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
