---
title: "Random Forest Classification: eIF3d Hypoxia-Specific vs Normoxia-Specific Targets (1hr, LFC 0.5)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r hardcoded_values}
# Feature set selection: "te_lfc" | "codon_optimality"
feature_set <- "te_lfc"

feature_matrix_rds <- if (feature_set == "te_lfc") {
  "feature_matrix_te_lfc.rds"
} else if (feature_set == "codon_optimality") {
  "feature_matrix_codon_optimality.rds"
}

clip_mode  <- "aggregate"
pos_label  <- "hypoxia_only"
neg_label  <- "normoxia_only"
gene_label <- "eIF3d"
suffix     <- paste0(
  "eif3d_hypoxia_only_vs_normoxia_alltimepoints_promotes_1hr_lfc0.5_",
  feature_set, "_clipaggregate_no_boruta"
)

cat("Feature matrix:", feature_matrix_rds, "\n")
cat("Output suffix:", suffix, "\n")
cat("Positive label:", pos_label, "\n")
cat("Negative label:", neg_label, "\n")
```

# Overview

Binary classification without Boruta feature selection: what mRNA sequence features distinguish genes that eIF3d promotes translation of **only in 1hr hypoxia** from genes eIF3d promotes translation of **only in normoxia**?

This is a hypoxia-selectivity question: both classes are eIF3d targets, but in different conditions.

- **Positive class ("`r pos_label`")**: Genes in `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` AND NOT in the normoxia alltimepoints set
- **Negative class ("`r neg_label`")**: Genes promoted in normoxia (si3d vs sictrl, all timepoints, `te_lfc > 0.5`, `te_padj < 0.05`) AND NOT in `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- Genes present in **both** sets are excluded from both classes

All available hypoxia-only and normoxia-only genes are used (no subsampling).

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(pROC)
library(cowplot)
library(ggpubr)
```

# Load Data

```{r load_data}
# Feature matrix
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")

# Hypoxia promotes set
hypoxia_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
cat("Hypoxia promotes genes:", nrow(hypoxia_genes), "\n")

# Normoxia promotes set (time-regressed: si3d vs sictrl, all 9 normoxia samples)
normoxia_genes <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv")) %>%
  filter(te_lfc > 0.5, te_padj < 0.05) %>%
  mutate(ensembl_gene = sub("\\..*", "", gene_id))
cat("Normoxia promotes genes (alltimepoints):", nrow(normoxia_genes), "\n")

# Compute overlap and exclusive sets
overlap_genes <- intersect(hypoxia_genes$ensembl_gene, normoxia_genes$ensembl_gene)
hypoxia_only_genes <- setdiff(hypoxia_genes$ensembl_gene, normoxia_genes$ensembl_gene)
normoxia_only_genes <- setdiff(normoxia_genes$ensembl_gene, hypoxia_genes$ensembl_gene)

cat("\nGenes in both sets (excluded):", length(overlap_genes), "\n")
cat("Hypoxia-only genes:", length(hypoxia_only_genes), "\n")
cat("Normoxia-only genes:", length(normoxia_only_genes), "\n")
```

# Create Binary Target

```{r create_target}
# Pre-filter to genes present in the feature matrix
hypoxia_only_in_features  <- hypoxia_only_genes[hypoxia_only_genes %in% features$gene_id_clean]
normoxia_only_in_features <- normoxia_only_genes[normoxia_only_genes %in% features$gene_id_clean]
cat("Hypoxia-only in feature matrix:", length(hypoxia_only_in_features), "\n")
cat("Normoxia-only in feature matrix:", length(normoxia_only_in_features), "\n")

# Label genes
features <- features %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% hypoxia_only_in_features  ~ pos_label,
    gene_id_clean %in% normoxia_only_in_features ~ neg_label,
    TRUE ~ NA_character_
  ))

# Keep only labeled genes
model_genes <- features %>%
  filter(!is.na(eif3d_target)) %>%
  mutate(eif3d_target = factor(eif3d_target, levels = c(neg_label, pos_label)))

cat("Positive (", pos_label, "):", sum(model_genes$eif3d_target == pos_label), "\n")
cat("Negative (", neg_label, "):", sum(model_genes$eif3d_target == neg_label), "\n")
cat("Total:", nrow(model_genes), "\n")
cat("Imbalance ratio (neg:pos):",
    round(sum(model_genes$eif3d_target == neg_label) /
          sum(model_genes$eif3d_target == pos_label), 1), ":1\n")
```

# Preprocessing

## Filter Features (No Boruta)

```{r filter_features}
# Identify feature columns (exclude ID/target cols)
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count (all combined):", length(feature_cols), "\n")

# ---- CLIP feature mode: aggregate ----
aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                         "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                         "clip_lee_glu_vs_cm_enrichment")
regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
feature_cols <- setdiff(feature_cols, regional_clip)
cat("CLIP mode: aggregate — using aggregate CLIP counts only, dropped", length(regional_clip), "regional features\n")

# Remove k-mer and RBP motif columns (te_lfc feature set)
kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
excluded <- feature_cols[str_detect(feature_cols, kmer_motif_pattern)]
feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]
cat("Targeted mode: removed", length(excluded), "k-mer/motif columns\n")

# Remove TE features (circular with TE-based outcome definition)
te_pattern <- "^indiv_te_|^delta_te_"
te_excluded <- feature_cols[str_detect(feature_cols, te_pattern)]
feature_cols <- feature_cols[!str_detect(feature_cols, te_pattern)]
cat("Removed", length(te_excluded), "TE features (circular with outcome)\n")

cat("Feature count after filtering:", length(feature_cols), "\n")
```

## Median-Impute Missing Values

```{r impute_missing}
missing_counts <- model_genes %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

cat("Features with missing values:", nrow(missing_counts), "\n")
if (nrow(missing_counts) > 0) print(missing_counts)
cat("(Imputation deferred to after train/test split — medians computed from training set only)\n")
```

# Prepare Modeling Data

```{r prepare_data}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, eif3d_target, all_of(feature_cols))

cat("Final dataset:", nrow(model_data), "genes x", length(feature_cols), "features\n")
cat("\nClass distribution:\n")
print(table(model_data$eif3d_target))
```

# Train/Test Split

```{r train_test_split}
set.seed(9)

train_idx <- createDataPartition(model_data$eif3d_target, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
print(table(train_data$eif3d_target))
cat("\nTest set:", nrow(test_data), "\n")
print(table(test_data$eif3d_target))

# Compute medians from training set only, then apply to both splits
train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

cat("\nRemaining NAs in train:", sum(is.na(train_data[, feature_cols])), "\n")
cat("Remaining NAs in test:", sum(is.na(test_data[, feature_cols])), "\n")
```

# Train Random Forest

```{r train_rf}
set.seed(9)
rf_model <- ranger(
  x = train_data[, feature_cols],
  y = train_data$eif3d_target,
  num.trees = 500,
  importance = "impurity",
  probability = TRUE,
  seed = 9
)

print(rf_model)
```

# Evaluate on Test Set

## ROC Curve and AUC

```{r roc_curve, fig.width=8, fig.height=6}
# Predictions
pred_probs <- predict(rf_model, test_data[, feature_cols])$predictions
pred_pos_prob <- pred_probs[, pos_label]
actual <- test_data$eif3d_target

# ROC on test set (threshold-free AUC)
roc_obj <- roc(response = actual, predictor = pred_pos_prob,
               levels = c(neg_label, pos_label), direction = "<")
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

# Youden threshold from OOB predictions on training set (avoids test set snooping)
oob_probs <- rf_model$predictions[, pos_label]
roc_oob <- roc(response = train_data$eif3d_target, predictor = oob_probs,
               levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
coords_youden <- coords(roc_oob, "best", ret = c("threshold", "sensitivity", "specificity"),
                        best.method = "youden")
optimal_threshold <- coords_youden$threshold

cat("Youden threshold (from training OOB):", round(optimal_threshold, 4), "\n")
cat("  OOB Sensitivity:", round(coords_youden$sensitivity, 3), "\n")
cat("  OOB Specificity:", round(coords_youden$specificity, 3), "\n")

# Plot
roc_df <- data.frame(sensitivity = roc_obj$sensitivities,
                     specificity = roc_obj$specificities)

p_roc <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_point(aes(x = 1 - coords_youden$specificity, y = coords_youden$sensitivity),
             color = "red", size = 3) +
  annotate("text", x = 1 - coords_youden$specificity + 0.05, y = coords_youden$sensitivity,
           label = paste0("Youden\n(", round(optimal_threshold, 3), ")"),
           color = "red", size = 3, hjust = 0) +
  theme_cowplot() +
  labs(
    title = "ROC Curve: eIF3d Hypoxia-Specific vs Normoxia-Specific Target Classification",
    subtitle = paste0("AUC = ", round(auc_value, 3),
                      " | Threshold = ", round(optimal_threshold, 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  coord_equal()

print(p_roc)

ggsave(here("plots", paste0("rf_classification_roc_", suffix, ".pdf")),
       plot = p_roc, width = 8, height = 6)
```

## Confusion Matrix

```{r confusion_matrix}
pred_class <- factor(
  ifelse(pred_pos_prob >= optimal_threshold, pos_label, neg_label),
  levels = c(neg_label, pos_label)
)

cm <- confusionMatrix(pred_class, actual, positive = pos_label)
print(cm)

# Extract metrics
accuracy <- cm$overall["Accuracy"]
balanced_accuracy <- cm$byClass["Balanced Accuracy"]
sensitivity <- cm$byClass["Sensitivity"]
specificity <- cm$byClass["Specificity"]
precision <- cm$byClass["Precision"]
f1 <- cm$byClass["F1"]

cat("\nKey Metrics (Youden Threshold):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n")
```

## Confusion Matrix Plot

```{r confusion_matrix_plot, fig.width=6, fig.height=5}
cm_table <- as.data.frame(cm$table)
colnames(cm_table) <- c("Predicted", "Actual", "Freq")

p_cm <- ggplot(cm_table, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#0072B2") +
  theme_cowplot() +
  labs(
    title = "Confusion Matrix: Hypoxia-Only vs Normoxia-Only",
    subtitle = paste0("Balanced Acc = ", round(balanced_accuracy, 3),
                      ", F1 = ", round(f1, 3))
  ) +
  theme(legend.position = "none")

print(p_cm)

ggsave(here("plots", paste0("rf_classification_confusion_matrix_", suffix, ".pdf")),
       plot = p_cm, width = 6, height = 5)
```

# 5-Fold Stratified Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(train_data$eif3d_target, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")

  test_idx <- folds[[i]]
  train_cv <- train_data[-test_idx, ]
  test_cv <- train_data[test_idx, ]

  # Impute using this fold's training data medians only
  cv_medians <- train_cv %>%
    dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~median(., na.rm = TRUE)))
  train_cv <- impute_with_medians(train_cv, cv_medians)
  test_cv  <- impute_with_medians(test_cv,  cv_medians)

  rf_cv <- ranger(
    x = train_cv[, feature_cols],
    y = train_cv$eif3d_target,
    num.trees = 500,
    probability = TRUE,
    seed = 9
  )

  # Youden threshold from OOB predictions on this fold's training data
  oob_probs_cv <- rf_cv$predictions[, pos_label]
  roc_oob_cv <- roc(response = train_cv$eif3d_target, predictor = oob_probs_cv,
                    levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)
  coords_cv <- coords(roc_oob_cv, "best", ret = c("threshold", "sensitivity", "specificity"),
                      best.method = "youden")

  # Predict on held-out fold
  pred_probs_cv <- predict(rf_cv, test_cv[, feature_cols])$predictions[, pos_label]

  # ROC on test fold (for AUC)
  roc_cv <- roc(test_cv$eif3d_target, pred_probs_cv,
                levels = c(neg_label, pos_label), direction = "<", quiet = TRUE)

  pred_class_cv <- factor(
    ifelse(pred_probs_cv >= coords_cv$threshold, pos_label, neg_label),
    levels = c(neg_label, pos_label)
  )
  cm_cv <- confusionMatrix(pred_class_cv, test_cv$eif3d_target, positive = pos_label)

  data.frame(
    fold = i,
    Threshold = coords_cv$threshold,
    AUC = as.numeric(auc(roc_cv)),
    Balanced_Accuracy = cm_cv$byClass["Balanced Accuracy"],
    Sensitivity = cm_cv$byClass["Sensitivity"],
    Specificity = cm_cv$byClass["Specificity"],
    Precision = cm_cv$byClass["Precision"],
    F1 = cm_cv$byClass["F1"]
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL

cat("\n5-Fold CV Results:\n")
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("AUC", "Balanced_Accuracy", "Sensitivity", "Specificity", "Precision", "F1")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
importance_df <- data.frame(
  feature = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance))

# Clean names for plotting
importance_df <- importance_df %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^utr5_4mer_", "5'UTR 4mer: ") %>%
      str_replace("^utr3_4mer_", "3'UTR 4mer: ") %>%
      str_replace("^cds_4mer_", "CDS 4mer: ") %>%
      str_replace("^rbp_", "RBP: ") %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

# Compute directionality via point-biserial correlation (all labeled genes, pre-split)
target_numeric <- as.numeric(model_data$eif3d_target) - 1  # 0 = normoxia_only, 1 = hypoxia_only

importance_df <- importance_df %>%
  rowwise() %>%
  mutate(
    correlation = tryCatch(
      cor(model_data[[feature]], target_numeric, use = "complete.obs"),
      error = function(e) NA_real_
    ),
    direction = case_when(
      is.na(correlation) ~ NA_character_,
      correlation >= 0   ~ paste0("Higher in ", pos_label),
      TRUE               ~ paste0("Lower in ", pos_label)
    )
  ) %>%
  ungroup()

dir_colors <- c("#009E73", "#D55E00")
names(dir_colors) <- c(paste0("Higher in ", pos_label), paste0("Lower in ", pos_label))

cat("Higher in", pos_label, ":", sum(importance_df$direction == paste0("Higher in ", pos_label), na.rm = TRUE), "\n")
cat("Lower in", pos_label, ":", sum(importance_df$direction == paste0("Lower in ", pos_label), na.rm = TRUE), "\n")

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance, direction))

# Bar plot
p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: eIF3d Hypoxia-Only vs Normoxia-Only Classification",
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance)

ggsave(here("plots", paste0("rf_classification_feature_importance_", suffix, ".pdf")),
       plot = p_importance, width = 10, height = 10)

# Top 40 bar plot
importance_all_plot <- importance_df %>% head(40)
all_height <- max(6, nrow(importance_all_plot) * 0.35)

p_importance_all <- importance_all_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = paste0("Top ", nrow(importance_all_plot), " Features: eIF3d Hypoxia-Only vs Normoxia-Only Classification"),
    subtitle = "Variable Importance (Impurity)",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_all)

ggsave(here("plots", paste0("rf_classification_feature_importance_all_", suffix, ".pdf")),
       plot = p_importance_all, width = 10, height = all_height)

# Custom-features-only bar plot (exclude k-mer and RBP motif features)
importance_custom <- importance_df %>%
  filter(!str_detect(feature, "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbp_|^rbns_|^cisbp_")) %>%
  head(40)

custom_height <- max(6, nrow(importance_custom) * 0.35)

p_importance_custom <- importance_custom %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Custom Features: eIF3d Hypoxia-Only vs Normoxia-Only Classification",
    subtitle = "Variable Importance (Impurity) — CLIP, 4E/DAP5, Signatures, Lengths, GC",
    x = "",
    y = "Variable Importance"
  )

print(p_importance_custom)

ggsave(here("plots", paste0("rf_classification_feature_importance_custom_", suffix, ".pdf")),
       plot = p_importance_custom, width = 10, height = custom_height)
```

## Feature Category Summary

```{r feature_categories, fig.width=10, fig.height=6}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^struct_")                          ~ "RNA Structure",
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^utr5_4mer_")                      ~ "5' UTR 4-mers",
      str_detect(feature, "^utr3_4mer_")                      ~ "3' UTR 4-mers",
      str_detect(feature, "^cds_4mer_")                       ~ "CDS 4-mers",
      str_detect(feature, "^rbns_|^cisbp_")                   ~ "RBP Motifs",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      str_detect(feature, "^indiv_te_|^delta_te_")            ~ "Translational Efficiency",
      str_detect(feature, "^cai$|^tai$|^csc$")               ~ "Codon Optimality",
      TRUE                                                    ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(
    n_features = n(),
    total_importance = sum(importance),
    mean_importance = mean(importance),
    top_feature = feature[which.max(importance)],
    .groups = "drop"
  ) %>%
  arrange(desc(total_importance))

cat("\nImportance by Category:\n")
print(category_summary)

p_category <- ggplot(category_summary,
                     aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Importance by Category: Hypoxia-Only vs Normoxia-Only",
    x = "",
    y = "Total Variable Importance"
  )

print(p_category)

ggsave(here("plots", paste0("rf_classification_importance_by_category_", suffix, ".pdf")),
       plot = p_category, width = 10, height = 6)
```

# Export Results

```{r export_results}
output_dir <- here("output", "predictive_modeling")

# Model
saveRDS(rf_model, file.path(output_dir, paste0("rf_classifier_", suffix, ".rds")))

# Feature importance
write_csv(importance_df, file.path(output_dir, paste0("rf_classification_importance_", suffix, ".csv")))

# Performance summary
performance_df <- data.frame(
  Metric = c("AUC", "Accuracy", "Balanced_Accuracy", "Sensitivity",
             "Specificity", "Precision", "F1"),
  Test_Set = c(as.numeric(auc_value), accuracy, balanced_accuracy,
               sensitivity, specificity, precision, f1),
  CV_Mean = c(mean(cv_results_df$AUC), NA, mean(cv_results_df$Balanced_Accuracy),
              mean(cv_results_df$Sensitivity), mean(cv_results_df$Specificity),
              mean(cv_results_df$Precision, na.rm = TRUE),
              mean(cv_results_df$F1, na.rm = TRUE)),
  CV_SD = c(sd(cv_results_df$AUC), NA, sd(cv_results_df$Balanced_Accuracy),
            sd(cv_results_df$Sensitivity), sd(cv_results_df$Specificity),
            sd(cv_results_df$Precision, na.rm = TRUE),
            sd(cv_results_df$F1, na.rm = TRUE))
)
write_csv(performance_df, file.path(output_dir, paste0("rf_classification_performance_", suffix, ".csv")))

# CV results
write_csv(cv_results_df, file.path(output_dir, paste0("rf_classification_cv_results_", suffix, ".csv")))

# Predictions
predictions_export <- data.frame(
  gene_id = test_data$gene_id_clean,
  symbol = test_data$symbol,
  actual = actual,
  predicted = pred_class,
  prob_hypoxia_only = pred_pos_prob,
  prob_normoxia_only = pred_probs[, neg_label]
)
write_csv(predictions_export, file.path(output_dir, paste0("rf_classification_predictions_", suffix, ".csv")))

cat("Exported: model RDS, importance CSV, performance CSV, CV results CSV, predictions CSV\n")
```

# Summary

```{r summary}
cat("========================================\n")
cat("CLASSIFICATION SUMMARY\n")
cat("eIF3d Hypoxia-Only vs Normoxia-Only (1hr, LFC 0.5)\n")
cat("========================================\n\n")

cat("Gene Sets:\n")
cat("  Hypoxia promotes (raw):", length(hypoxia_genes$ensembl_gene), "\n")
cat("  Normoxia promotes (raw):", length(normoxia_genes$ensembl_gene), "\n")
cat("  Overlap (excluded):", length(overlap_genes), "\n")
cat("  Hypoxia-only (all):", length(hypoxia_only_genes), "\n")
cat("  Normoxia-only (all):", length(normoxia_only_genes), "\n")
cat("  Hypoxia-only (in feature matrix):", length(hypoxia_only_in_features), "\n")
cat("  Normoxia-only (in feature matrix):", length(normoxia_only_in_features), "\n\n")

cat("Model:\n")
cat("  Positive (", pos_label, "):", sum(model_data$eif3d_target == pos_label), "\n")
cat("  Negative (", neg_label, "):", sum(model_data$eif3d_target == neg_label), "\n")
cat("  Features:", length(feature_cols), "\n")
cat("  Feature set:", feature_set, "(no Boruta selection)\n")
cat("  CLIP mode:", clip_mode, "\n\n")

cat("Test Set Performance (Youden threshold =", round(optimal_threshold, 4), "):\n")
cat("  AUC:", round(as.numeric(auc_value), 4), "\n")
cat("  Balanced Accuracy:", round(balanced_accuracy, 4), "\n")
cat("  Sensitivity:", round(sensitivity, 4), "\n")
cat("  Specificity:", round(specificity, 4), "\n")
cat("  Precision:", round(precision, 4), "\n")
cat("  F1:", round(f1, 4), "\n\n")

cat("5-Fold CV:\n")
cat("  AUC:", round(mean(cv_results_df$AUC), 4), "+/-", round(sd(cv_results_df$AUC), 4), "\n")
cat("  Balanced Accuracy:", round(mean(cv_results_df$Balanced_Accuracy), 4), "+/-",
    round(sd(cv_results_df$Balanced_Accuracy), 4), "\n\n")

cat("Top 5 Features:\n")
for (i in 1:5) {
  cat("  ", i, ". ", importance_df$feature[i], " (", round(importance_df$importance[i], 2), ")\n", sep = "")
}

cat("========================================\n")
```

# Final Feature Importance Plot

```{r final_importance_plot, fig.width=8, fig.height=max(4, min(40, length(feature_cols)) * 0.35)}
# Lollipop plot of top features (cap at 40)
importance_final_plot <- importance_df %>% head(40)
final_height <- max(4, nrow(importance_final_plot) * 0.35)

p_final_importance <- importance_final_plot %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_segment(aes(xend = reorder(feature_clean, importance), y = 0, yend = importance),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(color = direction), size = 3) +
  scale_color_manual(values = dir_colors, name = "Direction", na.value = "grey40") +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Feature Importance: eIF3d Hypoxia-Only vs Normoxia-Only Classification",
    subtitle = paste0("1hr | LFC 0.5 | AUC = ", round(as.numeric(auc_value), 3)),
    x = "",
    y = "Variable Importance (Impurity)"
  )

print(p_final_importance)

ggsave(here("plots", paste0("rf_final_importance_", suffix, ".pdf")),
       plot = p_final_importance, width = 8,
       height = final_height)
```

# Signed-Importance Lollipop (Top 30)

```{r plot_lollipop, fig.width=10, fig.height=10}
lollipop_data <- importance_df %>%
  filter(!is.na(correlation)) %>%
  head(30) %>%
  mutate(
    signed_importance = importance * sign(correlation),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    )),
    feature_clean = fct_reorder(feature_clean, signed_importance)
  )

category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

p_lollipop <- ggdotchart(
  lollipop_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 4,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher in hypoxia-only)",
  title    = "Top 30 Features: Signed Importance",
  subtitle = "eIF3d | Hypoxia-Only vs Normoxia-Only | 1hr | LFC 0.5"
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop)

ggsave(here("plots", paste0("rf_directionality_lollipop_", suffix, ".pdf")),
       plot = p_lollipop, width = 10, height = 10)
```

# Session Info

```{r session_info}
sessionInfo()
```
