---
title: "RF Normoxia Timepoint Comparison: eIF3d Promotes (te_lfc)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Compare RF classification performance (AUC) and feature importance across 5 normoxia timepoints for eIF3d promotes, LFC 0.5, te_lfc features, clip aggregate:

- **1hr**: single timepoint
- **4hr**: single timepoint
- **24hr**: single timepoint
- **All (1+4+24hr regressed)**: 9-sample time-regressed model
- **1+4hr regressed**: 6-sample time-regressed model

All models use `feature_matrix_te_lfc.rds` (timepoint-agnostic) and were run via `07_rf_classification_eif3d_targets_no_boruta.Rmd`.

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(pheatmap)
```

```{r define_timepoints}
# Named list: internal suffix → display label
timepoints <- list(
  "1hr"           = "1hr",
  "4hr"           = "4hr",
  "24hr"          = "24hr",
  "alltimepoints" = "All (1+4+24hr)",
  "1and4hr"       = "1+4hr regressed"
)

# Fixed params
gene      <- "3d"
direction <- "promotes"
condition <- "normoxia"
lfc       <- "0.5"
feat      <- "te_lfc"
clip      <- "aggregate"

# Build suffix for each timepoint
make_suffix <- function(tp) {
  paste0("eif3", gene, "_", direction, "_", condition, "_", tp,
         "_lfc", lfc, "_", feat, "_clip", clip, "_no_boruta")
}

suffixes <- setNames(sapply(names(timepoints), make_suffix), names(timepoints))
pos_label <- direction
neg_label <- paste0("does_not_", sub("s$", "", direction))

cat("Suffixes:\n")
print(suffixes)
```

# Load Data

```{r load_performance}
perf_list <- lapply(names(timepoints), function(tp) {
  f <- here("output", "predictive_modeling",
            paste0("rf_classification_performance_", suffixes[[tp]], ".csv"))
  if (!file.exists(f)) { warning("Missing: ", f); return(NULL) }
  read_csv(f, show_col_types = FALSE) %>%
    mutate(timepoint = tp, label = timepoints[[tp]])
})

perf_all <- bind_rows(perf_list)
cat("Performance rows loaded:", nrow(perf_all), "\n")
```

```{r load_cv}
cv_list <- lapply(names(timepoints), function(tp) {
  f <- here("output", "predictive_modeling",
            paste0("rf_classification_cv_results_", suffixes[[tp]], ".csv"))
  if (!file.exists(f)) { warning("Missing CV: ", f); return(NULL) }
  read_csv(f, show_col_types = FALSE) %>%
    mutate(timepoint = tp, label = timepoints[[tp]])
})

cv_all <- bind_rows(cv_list)
cat("CV rows loaded:", nrow(cv_all), "\n")
```

```{r load_importance}
imp_list <- lapply(names(timepoints), function(tp) {
  f <- here("output", "predictive_modeling",
            paste0("rf_classification_importance_", suffixes[[tp]], ".csv"))
  if (!file.exists(f)) { warning("Missing importance: ", f); return(NULL) }
  read_csv(f, show_col_types = FALSE) %>%
    mutate(timepoint = tp, label = timepoints[[tp]])
})

imp_all <- bind_rows(imp_list)
cat("Importance rows loaded:", nrow(imp_all), "\n")
cat("Timepoints with data:", paste(unique(imp_all$timepoint), collapse = ", "), "\n")
```

```{r load_set_sizes}
# Extract total class sizes (train + test) from performance CSVs.
# N_pos and N_neg rows were written by notebook 07.
set_sizes <- perf_all %>%
  filter(Metric %in% c("N_pos", "N_neg")) %>%
  dplyr::select(timepoint, label, Metric, n = Test_Set) %>%
  pivot_wider(names_from = Metric, values_from = n) %>%
  rename(n_pos = N_pos, n_neg = N_neg)

cat("Total class sizes per timepoint:\n")
print(set_sizes)
```

# Section 2: AUC Comparison

```{r auc_data}
tp_order <- names(timepoints)

auc_test <- perf_all %>%
  filter(Metric == "AUC") %>%
  dplyr::select(timepoint, label, auc_test = Test_Set, auc_cv_mean = CV_Mean, auc_cv_sd = CV_SD) %>%
  left_join(set_sizes %>% dplyr::select(timepoint, n_pos, n_neg), by = "timepoint") %>%
  mutate(label = factor(label, levels = unlist(timepoints[tp_order]))) %>%
  arrange(label) %>%
  mutate(axis_label = paste0(label, "\n(", n_pos, "+ / ", n_neg, "-)"),
         axis_label = factor(axis_label, levels = axis_label))

print(auc_test)
```

```{r auc_bar_plot, fig.width=6, fig.height=5}
p_auc <- ggplot(auc_test, aes(x = axis_label, y = auc_test)) +
  geom_col(fill = "#4E79A7", width = 0.6) +
  geom_errorbar(
    aes(ymin = auc_test - auc_cv_sd, ymax = auc_test + auc_cv_sd),
    width = 0.25, color = "black", linewidth = 0.7
  ) +
  geom_point(aes(y = auc_test), shape = 21, fill = "white",
             color = "black", size = 2.5) +
  geom_text(aes(y = auc_test + auc_cv_sd + 0.03, label = round(auc_test, 2)),
            size = 3, color = "grey20", vjust = 0) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_y_continuous(limits = c(0, 1.08), expand = c(0, 0)) +
  labs(
    title    = "eIF3d Promotes: RF Test AUC-ROC by Normoxia Timepoint",
    subtitle = "Bars = test AUC-ROC; error bars = ± cross-validation SD",
    x = NULL, y = "AUC-ROC"
  ) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_auc
ggsave(here("plots/rf_normoxia_timepoint_auc_comparison.pdf"), p_auc, width = 6, height = 5)
```

# Section 3: Feature Importance Heatmap

```{r normalize_importance}
imp_norm <- imp_all %>%
  group_by(timepoint) %>%
  mutate(importance_prop = importance / sum(importance)) %>%
  ungroup()
```

```{r select_top_features}
top_n_features <- 30

top_features <- imp_norm %>%
  group_by(feature) %>%
  summarise(max_prop = max(importance_prop), .groups = "drop") %>%
  slice_max(max_prop, n = top_n_features) %>%
  pull(feature)

cat("Top", top_n_features, "features selected\n")
```

```{r importance_matrix}
imp_matrix <- imp_norm %>%
  filter(feature %in% top_features) %>%
  dplyr::select(feature, timepoint, importance_prop) %>%
  pivot_wider(names_from = timepoint, values_from = importance_prop, values_fill = 0)

if ("feature_clean" %in% colnames(imp_all)) {
  clean_lookup <- imp_all %>%
    dplyr::select(feature, feature_clean) %>%
    distinct(feature, .keep_all = TRUE)
  imp_matrix <- imp_matrix %>%
    left_join(clean_lookup, by = "feature") %>%
    mutate(row_label = coalesce(feature_clean, feature))
} else {
  imp_matrix <- imp_matrix %>% mutate(row_label = feature)
}

mat <- imp_matrix %>%
  dplyr::select(row_label, all_of(names(timepoints))) %>%
  column_to_rownames("row_label") %>%
  as.matrix()

# Rename columns to display labels
colnames(mat) <- unlist(timepoints[colnames(mat)])

cat("Importance matrix:", nrow(mat), "features x", ncol(mat), "timepoints\n")

# Column annotation: n_pos and n_neg per timepoint (test-set counts)
col_ann <- set_sizes %>%
  filter(timepoint %in% names(timepoints)) %>%
  mutate(display = unlist(timepoints[timepoint])) %>%
  dplyr::select(display, `n pos` = n_pos) %>%
  column_to_rownames("display")

# Keep only columns present in mat
col_ann <- col_ann[colnames(mat), , drop = FALSE]
```

```{r importance_heatmap, fig.width=8, fig.height=11}
p_heatmap <- pheatmap(
  mat,
  cluster_rows    = TRUE,
  cluster_cols    = FALSE,
  scale           = "none",
  annotation_col  = col_ann,
  color           = colorRampPalette(c("white", "#4E79A7", "#003366"))(50),
  main            = paste0("Top ", top_n_features, " Features by Importance Proportion\n",
                           "eIF3d Promotes | Normoxia | te_lfc"),
  fontsize_row    = 8,
  fontsize_col    = 9,
  angle_col       = 45,
  border_color    = NA
)

ggsave(here("plots/rf_normoxia_timepoint_importance_heatmap.pdf"), p_heatmap,
       width = 8, height = 11)
```

# Section 4: Feature Importance Correlation

```{r importance_correlation}
imp_wide <- imp_norm %>%
  dplyr::select(feature, timepoint, importance_prop) %>%
  pivot_wider(names_from = timepoint, values_from = importance_prop, values_fill = 0)

tp_cols  <- intersect(names(timepoints), colnames(imp_wide))
cor_mat  <- cor(imp_wide[, tp_cols], method = "spearman")

colnames(cor_mat) <- unlist(timepoints[colnames(cor_mat)])
rownames(cor_mat) <- unlist(timepoints[rownames(cor_mat)])

cat("Correlation matrix:\n")
print(round(cor_mat, 3))

# Row/column annotation: n_pos and n_neg (aligned to cor_mat labels)
cor_ann <- set_sizes %>%
  filter(timepoint %in% names(timepoints)) %>%
  mutate(display = unlist(timepoints[timepoint])) %>%
  dplyr::select(display, `n pos` = n_pos) %>%
  column_to_rownames("display")

cor_ann <- cor_ann[rownames(cor_mat), , drop = FALSE]
```

```{r correlation_heatmap, fig.width=5.5, fig.height=5}
p_cor <- pheatmap(
  cor_mat,
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  display_numbers = TRUE,
  number_format   = "%.2f",
  annotation_row  = cor_ann,
  annotation_col  = cor_ann,
  color           = colorRampPalette(c("white", "#AED6F1", "#2E86C1"))(50),
  breaks          = seq(-1, 1, length.out = 51),
  main            = "Spearman Correlation of Feature Importance\nAcross Normoxia Timepoints",
  fontsize        = 10,
  angle_col       = 45,
  border_color    = "grey80"
)

ggsave(here("plots/rf_normoxia_timepoint_importance_correlation.pdf"), p_cor,
       width = 5.5, height = 5)
```

# Summary

```{r summary_table}
auc_test %>%
  arrange(desc(auc_test)) %>%
  dplyr::select(label, auc_test, auc_cv_mean, auc_cv_sd, n_pos, n_neg) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable(caption = "AUC by Timepoint (test set and CV) with class sizes")
```
