---
title: "12 — Codon Optimality Density Plots: eIF3d Targets"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

Density plots of CSC and FOP codon optimality scores for eIF3d target gene sets
versus all other genes. Style mirrors Roiuk et al., Nat Comm 2024: gray
background distribution for all genes, colored overlay for the gene set of
interest.

Gene sets plotted:

1. **eIF3d hypoxia promotes targets** (1 hr, lfc > 0.5)
2. **eIF3d normoxia promotes targets** (1+4 hr time-regressed, te_lfc > 0.5 & te_padj < 0.05)

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
```

## Load feature matrix

```{r load_features}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_codon_optimality.rds"))

# Strip Ensembl version suffix to match geneset IDs
features <- features %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

glimpse(features)
```

## Collapse to gene level

The feature matrix is transcript-level. Take the mean per gene for CSC and FOP.

```{r gene_level}
gene_features <- features %>%
  group_by(ensembl_gene, symbol) %>%
  summarise(
    csc = mean(csc, na.rm = TRUE),
    fop = mean(fop, na.rm = TRUE),
    .groups = "drop"
  )

cat("Genes in feature matrix:", nrow(gene_features), "\n")
```

## Density plot function

```{r plot_function}
# fg_ids: character vector of ensembl_gene IDs for the foreground set
# fg_label: text label shown on the plot for that set
make_density_plot <- function(df, feature_col, fg_ids, x_label, title,
                              fg_label = "Target set",
                              fg_color = "#E69F00") {
  df <- df %>% mutate(in_fg = ensembl_gene %in% fg_ids)
  bg <- df %>% filter(!is.na(.data[[feature_col]]))
  fg <- bg %>% filter(in_fg)
  n_fg    <- nrow(fg)
  n_other <- sum(!bg$in_fg & !is.na(bg[[feature_col]]))

  ks <- ks.test(
    fg[[feature_col]],
    bg %>% filter(!in_fg) %>% pull(feature_col)
  )
  ks_label <- paste0("KS test p = ", formatC(ks$p.value, format = "e", digits = 2))

  ggplot() +
    geom_density(
      data  = bg,
      aes(x = .data[[feature_col]]),
      fill  = NA, color = "grey60", linewidth = 0.9
    ) +
    geom_density(
      data      = fg,
      aes(x = .data[[feature_col]]),
      fill      = NA, color = fg_color, linewidth = 0.9
    ) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = ks_label, size = 3.5) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
             label = paste0("All other genes (n = ", n_other, ")"),
             size = 3.5, color = "grey50") +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
             label = paste0(fg_label, " (n = ", n_fg, ")"),
             size = 3.5, color = fg_color) +
    scale_x_continuous(expand = expansion(mult = 0.02)) +
    labs(title = title, x = x_label, y = "Density") +
    theme_classic(base_size = 13)
}
```

---

## Gene set 1: eIF3d hypoxia promotes (1 hr, lfc > 0.5)

```{r load_hyp_geneset}
hyp_promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
cat("Genes in hypoxia promotes set:", nrow(hyp_promotes), "\n")
```

### CSC — hypoxia promotes

```{r csc_hyp, fig.width=5, fig.height=4}
p_csc_hyp <- make_density_plot(
  gene_features, "csc",
  fg_ids   = hyp_promotes$ensembl_gene,
  x_label  = "CSC score (codon stabilization coefficient)",
  title    = "CSC: eIF3d hypoxia promotes vs all other genes",
  fg_label = "Promoted by eIF3d in hypoxia 1hr, lfc>0.5"
)
p_csc_hyp

ggsave(here("output", "csc_density_hyp3d_1hr.pdf"), p_csc_hyp, width = 5, height = 4)
```

### FOP — hypoxia promotes

```{r fop_hyp, fig.width=5, fig.height=4}
p_fop_hyp <- make_density_plot(
  gene_features, "fop",
  fg_ids   = hyp_promotes$ensembl_gene,
  x_label  = "FOP (frequency of optimal codons)",
  title    = "FOP: eIF3d hypoxia promotes vs all other genes",
  fg_label = "Promoted by eIF3d in hypoxia 1hr, lfc>0.5"
)
p_fop_hyp

ggsave(here("output", "fop_density_hyp3d_1hr.pdf"), p_fop_hyp, width = 5, height = 4)
```

---

## Gene set 2: eIF3d normoxia promotes (1+4 hr time-regressed, te_lfc > 0.5)

```{r load_normo_geneset}
normo_tc <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
)

normo_promotes <- normo_tc %>%
  filter(te_lfc > 0.5, te_padj < 0.05) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

cat("Genes in normoxia promotes set (1+4hr, te_lfc>0.5, te_padj<0.05):", nrow(normo_promotes), "\n")
```

### CSC — normoxia promotes

```{r csc_normo, fig.width=5, fig.height=4}
p_csc_normo <- make_density_plot(
  gene_features, "csc",
  fg_ids   = normo_promotes$ensembl_gene,
  x_label  = "CSC score (codon stabilization coefficient)",
  title    = "CSC: eIF3d normoxia promotes vs all other genes",
  fg_label = "Promoted by eIF3d in normoxia 1+4hr, lfc>0.5"
)
p_csc_normo

ggsave(here("output", "csc_density_normo3d_1and4hr.pdf"), p_csc_normo, width = 5, height = 4)
```

### FOP — normoxia promotes

```{r fop_normo, fig.width=5, fig.height=4}
p_fop_normo <- make_density_plot(
  gene_features, "fop",
  fg_ids   = normo_promotes$ensembl_gene,
  x_label  = "FOP (frequency of optimal codons)",
  title    = "FOP: eIF3d normoxia promotes vs all other genes",
  fg_label = "Promoted by eIF3d in normoxia 1+4hr, lfc>0.5"
)
p_fop_normo

ggsave(here("output", "fop_density_normo3d_1and4hr.pdf"), p_fop_normo, width = 5, height = 4)
```

---

## Session info

```{r session_info}
sessionInfo()
```
