---
title: "`r paste0('Cross-Model Feature Importance: eIF3d vs eIF4E (', params$feature_set, ', clip=', params$clip_mode, ', lfc', params$lfc, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_set: "kozak"
  clip_mode: "aggregate"
  lfc: "0.5"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_set = "kozak",
    clip_mode   = "aggregate",
    lfc         = "0.5"
  )
}

# Match notebook 07's convention: paste0("eif3", gene, ...) where gene="3d" → "eif33d"
suffix_3d <- paste0("eif3", "3d", "_promotes_normoxia_1and4hr_lfc", params$lfc,
                    "_", params$feature_set, "_clip", params$clip_mode,
                    "_no_boruta_no4e")

suffix_4e <- paste0("eif4e_promotes_", params$feature_set,
                    "_clip", params$clip_mode, "_no_boruta")

# Directionality suffix for eIF3d: base model without _no4e
# (correlations are feature-level, unaffected by which features were excluded from the model)
suffix_3d_dir <- paste0("eif3", "3d", "_promotes_normoxia_1and4hr_lfc", params$lfc,
                        "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")

subtitle_label <- paste0("eIF3d (normoxia, 1+4hr, lfc", params$lfc,
                         ") vs eIF4E | ", params$feature_set,
                         " | clip=", params$clip_mode)

cat("eIF3d suffix:", suffix_3d, "\n")
cat("eIF4E suffix:", suffix_4e, "\n")
```

# Overview

Compare RF variable importance between the eIF3d (normoxia, 1+4hr) model and the eIF4E model to ask: **do the same mRNA sequence features predict eIF3d-promoted translation AND eIF4E-promoted translation, or are different feature classes responsible?**

Both models use an identical feature universe: the circular eIF4E features (`4e_resistance`, `eif4e_ip`, `lit_eIF4E_*`) are excluded from the eIF3d model (via `exclude_4e_features = TRUE` in notebook 07) to enable a fair comparison.

Raw impurity-based importance is not comparable across independently trained forests. We normalize each model's importance to **relative proportions** (sum to 1) before comparing.

**Prerequisites:**
- `output/predictive_modeling/rf_classification_importance_{suffix_3d}.csv` — from notebook 07 with `exclude_4e_features = TRUE`
- `output/predictive_modeling/rf_classification_importance_{suffix_4e}.csv` — from notebook 09

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggrepel)
```

# Load Data

```{r load_data}
imp_3d <- read_csv(here("output", "predictive_modeling",
                        paste0("rf_classification_importance_", suffix_3d, ".csv")))
imp_4e <- read_csv(here("output", "predictive_modeling",
                        paste0("rf_classification_importance_", suffix_4e, ".csv")))

cat("eIF3d features:", nrow(imp_3d), "\n")
cat("eIF4E features:", nrow(imp_4e), "\n")

# Directionality: correlation = point-biserial between feature value and class label
# Positive = higher in "promotes" class, negative = lower
dir_3d <- read_csv(here("output", "predictive_modeling",
                        paste0("rf_feature_directionality_", suffix_3d_dir, ".csv"))) %>%
  select(feature, correlation_3d = correlation)
dir_4e <- read_csv(here("output", "predictive_modeling",
                        paste0("rf_feature_directionality_", suffix_4e, ".csv"))) %>%
  select(feature, correlation_4e = correlation)

cat("eIF3d directionality features:", nrow(dir_3d), "\n")
cat("eIF4E directionality features:", nrow(dir_4e), "\n")
```

# Normalize and Join

```{r join_data}
# Normalize to relative importance (proportions summing to 1)
imp_3d <- imp_3d %>% mutate(rel_importance = importance / sum(importance))
imp_4e <- imp_4e %>% mutate(rel_importance = importance / sum(importance))

cat("eIF3d importance sums to:", sum(imp_3d$rel_importance), "\n")
cat("eIF4E importance sums to:", sum(imp_4e$rel_importance), "\n")

# Inner join on feature (features present in both models)
joined <- inner_join(
  imp_3d %>% select(feature, feature_clean, category, rel_importance_3d = rel_importance),
  imp_4e %>% select(feature, rel_importance_4e = rel_importance),
  by = "feature"
) %>%
  mutate(
    delta = rel_importance_3d - rel_importance_4e,
    total = rel_importance_3d + rel_importance_4e,
    delta_sign = ifelse(delta >= 0, "Higher in eIF3d", "Higher in eIF4E")
  ) %>%
  arrange(desc(total))

# Add directionality (correlation with class label) to joined table
joined <- joined %>%
  left_join(dir_3d, by = "feature") %>%
  left_join(dir_4e, by = "feature") %>%
  mutate(
    signed_3d = rel_importance_3d * sign(correlation_3d),
    signed_4e = rel_importance_4e * sign(correlation_4e),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    ))
  )

cat("Features in both models (inner join):", nrow(joined), "\n")
cat("Features only in eIF3d model:", sum(!imp_3d$feature %in% imp_4e$feature), "\n")
cat("Features only in eIF4E model:", sum(!imp_4e$feature %in% imp_3d$feature), "\n")
cat("Features with direction in both models:", sum(!is.na(joined$signed_3d) & !is.na(joined$signed_4e)), "\n")
```

# Plots

## 1. Scatter Plot: eIF3d vs eIF4E Importance

```{r scatter_plot, fig.width=10, fig.height=9}
label_data <- joined %>%
  arrange(desc(total)) %>%
  head(15)

axis_max <- max(joined$rel_importance_3d, joined$rel_importance_4e) * 1.05

p_scatter <- ggplot(joined, aes(x = rel_importance_4e, y = rel_importance_3d)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = category), size = 3, alpha = 0.7) +
  geom_text_repel(data = label_data, aes(label = feature_clean),
                  size = 3, max.overlaps = 20, segment.color = "grey50") +
  coord_fixed(xlim = c(0, axis_max), ylim = c(0, axis_max)) +
  theme_cowplot() +
  labs(
    title = "Feature Importance: eIF3d vs eIF4E Models",
    subtitle = subtitle_label,
    x = "Relative Importance (eIF4E model)",
    y = "Relative Importance (eIF3d normoxia 1+4hr model)",
    color = "Category"
  )

print(p_scatter)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_", suffix_3d, "_vs_", suffix_4e, ".pdf")),
       plot = p_scatter, width = 10, height = 9)
```

## 2. Signed Importance Scatter: Direction + Importance in One View

```{r signed_scatter, fig.width=10, fig.height=9}
category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

signed_data <- joined %>% filter(!is.na(signed_3d), !is.na(signed_4e))

x_min <- -0.18; x_max <- 0.12   # eIF4E (x-axis)
y_min <- -0.1;  y_max <- 0.1    # eIF3d (y-axis)

label_signed <- signed_data %>%
  mutate(abs_total = abs(signed_3d) + abs(signed_4e)) %>%
  arrange(desc(abs_total)) %>%
  head(20)

p_signed <- ggplot(signed_data, aes(x = signed_4e, y = signed_3d)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_abline(intercept = 0, slope = -1, linetype = "dotted", color = "grey80") +
  geom_point(aes(color = broad_category), size = 3, alpha = 0.7) +
  scale_color_manual(values = category_colors, name = "Category") +
  geom_text_repel(data = label_signed, aes(label = feature_clean),
                  size = 4.5, max.overlaps = Inf, segment.color = "grey50") +
  annotate("text", x = x_min * 0.97, y = y_max * 0.97,
           label = "Higher in eIF3d\nLower in eIF4E",
           hjust = 0, vjust = 1, size = 4, color = "grey45") +
  annotate("text", x = x_max * 0.97, y = y_max * 0.97,
           label = "Higher in both",
           hjust = 1, vjust = 1, size = 4, color = "grey45") +
  annotate("text", x = x_min * 0.97, y = y_min * 0.97,
           label = "Lower in both",
           hjust = 0, vjust = 0, size = 4, color = "grey45") +
  annotate("text", x = x_max * 0.97, y = y_min * 0.97,
           label = "Lower in eIF3d\nHigher in eIF4E",
           hjust = 1, vjust = 0, size = 4, color = "grey45") +
  coord_fixed(xlim = c(x_min, x_max), ylim = c(y_min, y_max)) +
  theme_cowplot(font_size = 14) +
  labs(
    title = "Signed Feature Importance: eIF3d vs eIF4E Models",
    subtitle = paste0(subtitle_label, "  |  axis = importance × direction (+= higher in promotes)"),
    x = "Signed Importance (eIF4E model)",
    y = "Signed Importance (eIF3d normoxia 1+4hr model)"
  )

print(p_signed)

ggsave(here("plots", paste0("rf_importance_comparison_signed_", suffix_3d, "_vs_", suffix_4e, ".pdf")),
       plot = p_signed, width = 10, height = 9)
```

## 3. Dumbbell Plot: Paired Importance by Model

```{r dumbbell_plot, fig.width=10, fig.height=10}
dumbbell_data <- joined %>%
  mutate(max_importance = pmax(rel_importance_3d, rel_importance_4e)) %>%
  arrange(desc(max_importance)) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, max_importance))

p_dumbbell <- ggplot(dumbbell_data) +
  geom_segment(aes(x = rel_importance_3d, xend = rel_importance_4e,
                   y = feature_clean, yend = feature_clean),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(x = rel_importance_3d, y = feature_clean, color = "eIF3d"), size = 3) +
  geom_point(aes(x = rel_importance_4e, y = feature_clean, color = "eIF4E"), size = 3) +
  scale_color_manual(
    values = c("eIF3d" = "#D55E00", "eIF4E" = "#0072B2"),
    name = "Model"
  ) +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: Paired Importance by Model",
    subtitle = subtitle_label,
    x = "Relative Importance (Proportion)",
    y = ""
  )

print(p_dumbbell)

ggsave(here("plots", paste0("rf_importance_comparison_dumbbell_", suffix_3d, "_vs_", suffix_4e, ".pdf")),
       plot = p_dumbbell, width = 10, height = 10)
```

## 4. Delta Importance Bar Plot

```{r delta_bar_plot, fig.width=10, fig.height=10}
delta_data <- joined %>%
  arrange(desc(abs(delta))) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, delta))

p_delta <- ggplot(delta_data, aes(x = feature_clean, y = delta, fill = delta_sign)) +
  geom_col() +
  scale_fill_manual(
    values = c("Higher in eIF3d" = "#D55E00", "Higher in eIF4E" = "#0072B2"),
    name = ""
  ) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Delta Importance: eIF3d minus eIF4E",
    subtitle = subtitle_label,
    x = "",
    y = "Delta Relative Importance (eIF3d - eIF4E)"
  )

print(p_delta)

ggsave(here("plots", paste0("rf_importance_comparison_delta_", suffix_3d, "_vs_", suffix_4e, ".pdf")),
       plot = p_delta, width = 10, height = 10)
```

# Export

```{r export}
output_dir <- here("output", "predictive_modeling")

export_df <- joined %>%
  select(feature, feature_clean, category,
         rel_importance_3d, rel_importance_4e, delta, total, delta_sign,
         correlation_3d, correlation_4e, signed_3d, signed_4e)

write_csv(export_df, file.path(output_dir,
  paste0("rf_importance_comparison_", suffix_3d, "_vs_", suffix_4e, ".csv")))

cat("Exported:", nrow(export_df), "features to\n")
cat(" ", file.path(output_dir, paste0("rf_importance_comparison_", suffix_3d, "_vs_", suffix_4e, ".csv")), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
