---
title: "14 — ECDF: eIF3 CLIP Binding Count for eIF3d-Promoted mRNAs"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

ECDF plot of `clip_eif3_count` (sum of eIF3 PAR-CLIP evidence across Lee et al.,
Cate undiff NPCs, and Cate diff NPCs; range 0–3) for all mRNAs (grey) vs.
mRNAs promoted by eIF3d in normoxia (1+4hr regressed, te_lfc > 0.5 &
te_padj < 0.05; Okabe-Ito orange). A right-shift of the orange ECDF indicates
enrichment of eIF3 CLIP binding in eIF3d-promoted targets.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(cowplot)
library(scales)
```

## Load data

```{r load_data}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_1hr_clip_split.rds"))

targets <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(te_lfc > 0.5, te_padj < 0.05)

cat("Features matrix rows:", nrow(features), "\n")
cat("eIF3d-promoted targets (normoxia 1+4hr):", nrow(targets), "\n")
cat("clip_eif3_count range:", range(features$clip_eif3_count, na.rm = TRUE), "\n")
```

## Label genes

```{r label_genes}
plot_df <- features %>%
  mutate(
    gene_id_clean = str_remove(gene_id, "\\..*"),
    group = if_else(gene_id_clean %in% targets$ensembl_gene,
                    "eIF3d promotes", "All mRNAs")
  )

# clip_eif3_count may be stored as a list-column in the feature matrix; coerce to integer
plot_df[["clip_eif3_count"]] <- as.integer(unlist(plot_df[["clip_eif3_count"]]))

table(plot_df$group)
```

## ECDF plot

```{r ecdf_plot, fig.width=5.5, fig.height=4.5}
n_all    <- sum(plot_df$group == "All mRNAs")
n_target <- sum(plot_df$group == "eIF3d promotes")

p <- ggplot() +
  stat_ecdf(
    data  = filter(plot_df, group == "All mRNAs"),
    aes(x = clip_eif3_count),
    color = "#808080", linewidth = 1
  ) +
  stat_ecdf(
    data  = filter(plot_df, group == "eIF3d promotes"),
    aes(x = clip_eif3_count),
    color = "#E69F00", linewidth = 1
  ) +
  scale_x_continuous(breaks = 0:3, labels = 0:3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  annotate("text", x = Inf, y = 0.05, hjust = 1.05,
           label = paste0("All mRNAs (n = ", n_all, ")"),
           color = "#808080", size = 3.5) +
  annotate("text", x = Inf, y = 0.13, hjust = 1.05,
           label = paste0("eIF3d promotes (n = ", n_target, ")"),
           color = "#E69F00", size = 3.5) +
  labs(
    x        = "eIF3 CLIP binding count (clip_eif3_count)",
    y        = "Cumulative fraction of mRNAs",
    title    = "eIF3 CLIP binding: eIF3d-promoted mRNAs vs. all mRNAs",
    subtitle = "Normoxia 1+4hr regressed | te_lfc > 0.5, te_padj < 0.05"
  ) +
  theme_cowplot()

p
```

## KS test

```{r ks_test}
ks <- ks.test(
  filter(plot_df, group == "eIF3d promotes") %>% pull(clip_eif3_count),
  filter(plot_df, group == "All mRNAs")      %>% pull(clip_eif3_count)
)
cat("KS test D =", round(ks$statistic, 4), "  p =", formatC(ks$p.value, format = "e", digits = 2), "\n")
```

## Save plot

```{r save_plot}
ggsave(
  here("plots", "ecdf_clip_eif3_count_normo_promoted.pdf"),
  p, width = 5.5, height = 4.5
)
ggsave(
  here("plots", "ecdf_clip_eif3_count_normo_promoted.png"),
  p, width = 5.5, height = 4.5, dpi = 300
)
cat("Plots saved.\n")
```

## Grouped proportion bar

```{r grouped_prop_bar, fig.width=5.5, fig.height=4.5}
prop_df <- plot_df %>%
  dplyr::count(group, clip_eif3_count) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n))

p_grouped <- ggplot(prop_df, aes(x = factor(clip_eif3_count), y = prop, fill = group)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("All mRNAs" = "#808080", "eIF3d promotes" = "#E69F00")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.05))) +
  labs(
    x     = "eIF3 CLIP binding count",
    y     = "Proportion of mRNAs",
    fill  = NULL,
    title = "Distribution of eIF3 CLIP binding count",
    subtitle = "Normoxia 1+4hr | te_lfc > 0.5, te_padj < 0.05"
  ) +
  theme_cowplot() +
  theme(legend.position = "top")

p_grouped
```

```{r save_grouped_prop_bar}
ggsave(
  here("plots", "grouped_prop_bar_clip_eif3_count.pdf"),
  p_grouped, width = 5.5, height = 4.5
)
ggsave(
  here("plots", "grouped_prop_bar_clip_eif3_count.png"),
  p_grouped, width = 5.5, height = 4.5, dpi = 300
)
cat("Grouped proportion bar saved.\n")
```

## % with any CLIP binding

```{r pct_any_clip, fig.width=4, fig.height=4.5}
binary_df <- plot_df %>%
  dplyr::select(group, clip_eif3_count) %>%
  mutate(any_clip = clip_eif3_count >= 1) %>%
  group_by(group) %>%
  summarise(pct_bound = mean(any_clip, na.rm = TRUE) * 100, .groups = "drop")

# Fisher test
ct <- table(
  group    = plot_df$group,
  any_clip = plot_df$clip_eif3_count >= 1
)
ft <- fisher.test(ct)
p_label <- paste0("Fisher p = ", formatC(ft$p.value, format = "e", digits = 2))

p_binary <- ggplot(binary_df, aes(x = group, y = pct_bound, fill = group)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("All mRNAs" = "#808080", "eIF3d promotes" = "#E69F00")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  annotate("text", x = 1.5, y = max(binary_df$pct_bound) * 1.07,
           label = p_label, size = 3.5) +
  labs(
    x     = NULL,
    y     = "% with ≥1 eIF3 CLIP dataset",
    fill  = NULL,
    title = "eIF3 CLIP binding (any evidence)",
    subtitle = "Normoxia 1+4hr | te_lfc > 0.5, te_padj < 0.05"
  ) +
  theme_cowplot() +
  theme(legend.position = "none")

p_binary
```

```{r save_pct_any_clip}
ggsave(
  here("plots", "pct_any_clip_eif3_count.pdf"),
  p_binary, width = 4, height = 4.5
)
ggsave(
  here("plots", "pct_any_clip_eif3_count.png"),
  p_binary, width = 4, height = 4.5, dpi = 300
)
cat("% bound bar saved.\n")
cat("Fisher test: OR =", round(ft$estimate, 2), " p =", formatC(ft$p.value, format = "e", digits = 2), "\n")
```

## Stacked proportion bar

```{r stacked_prop_bar, fig.width=4.5, fig.height=4.5}
stacked_df <- plot_df %>%
  dplyr::count(group, clip_eif3_count) %>%
  group_by(group) %>%
  mutate(prop = n / sum(n),
         clip_label = factor(clip_eif3_count, levels = 3:0))  # stack 3 on top

p_stacked <- ggplot(stacked_df, aes(x = group, y = prop, fill = clip_label)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(
    values = c("3" = "#0072B2", "2" = "#56B4E9", "1" = "#F0E442", "0" = "grey85"),
    name   = "CLIP count"
  ) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.02))) +
  labs(
    x     = NULL,
    y     = "Proportion of mRNAs",
    title = "eIF3 CLIP binding count composition",
    subtitle = "Normoxia 1+4hr | te_lfc > 0.5, te_padj < 0.05"
  ) +
  theme_cowplot()

p_stacked
```

```{r save_stacked_prop_bar}
ggsave(
  here("plots", "stacked_prop_bar_clip_eif3_count.pdf"),
  p_stacked, width = 4.5, height = 4.5
)
ggsave(
  here("plots", "stacked_prop_bar_clip_eif3_count.png"),
  p_stacked, width = 4.5, height = 4.5, dpi = 300
)
cat("Stacked proportion bar saved.\n")
```

## ECDF of TE LFC by CLIP binding count

Within eIF3d-promoted mRNAs, does having more eIF3 CLIP binding correlate with
a stronger translation efficiency change? ECDFs of `te_lfc` split by
`clip_eif3_count` (0/1/2/3) — since `te_lfc` is continuous the curves are smooth.

```{r lfc_clip_data}
# Load all mRNAs (no LFC/padj filter) to get te_lfc across the full transcriptome
all_te <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

lfc_clip_df <- all_te %>%
  dplyr::select(ensembl_gene, te_lfc) %>%
  inner_join(
    plot_df %>% dplyr::select(gene_id_clean, clip_eif3_count),
    by = c("ensembl_gene" = "gene_id_clean")
  ) %>%
  mutate(clip_eif3_count = replace_na(clip_eif3_count, 0L),
         clip_label = factor(clip_eif3_count,
                             levels = 0:3,
                             labels = c("0 (none)", "1", "2", "3 (all)")))

# report group sizes
lfc_clip_df %>% dplyr::count(clip_label)
```

```{r lfc_clip_ecdf, fig.width=6, fig.height=4.5}
clip_colors <- c(
  "0 (none)" = "grey70",
  "1"        = "#F0E442",
  "2"        = "#56B4E9",
  "3 (all)"  = "#0072B2"
)

p_lfc_ecdf <- ggplot(lfc_clip_df,
                     aes(x = te_lfc, color = clip_label)) +
  stat_ecdf(linewidth = 1) +
  scale_color_manual(values = clip_colors, name = "eIF3 CLIP\nbinding count") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(
    x        = "TE log2 fold change (si3d vs. sictrl)",
    y        = "Cumulative fraction",
    title    = "TE LFC distribution by eIF3 CLIP binding count",
    subtitle = "All mRNAs | normoxia 1+4hr (si3d vs. sictrl)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")

p_lfc_ecdf
```

```{r save_lfc_ecdf}
ggsave(
  here("plots", "ecdf_te_lfc_by_clip_count_normo_promoted.pdf"),
  p_lfc_ecdf, width = 6, height = 4.5
)
ggsave(
  here("plots", "ecdf_te_lfc_by_clip_count_normo_promoted.png"),
  p_lfc_ecdf, width = 6, height = 4.5, dpi = 300
)
cat("Plot saved.\n")
```

### Collapsed version: 0 / ≥1

```{r lfc_clip_ecdf_collapsed, fig.width=6, fig.height=4.5}
lfc_clip_collapsed <- lfc_clip_df %>%
  mutate(clip_group = if_else(clip_eif3_count >= 1, "≥1 (any)", "0 (none)"),
         clip_group = factor(clip_group, levels = c("0 (none)", "≥1 (any)")))

lfc_clip_collapsed %>% dplyr::count(clip_group)

clip_colors_3 <- c(
  "0 (none)"  = "grey70",
  "≥1 (any)" = "#D55E00"
)

p_lfc_ecdf_collapsed <- ggplot(lfc_clip_collapsed,
                               aes(x = te_lfc, color = clip_group)) +
  stat_ecdf(linewidth = 1) +
  scale_color_manual(values = clip_colors_3, name = "eIF3 CLIP\nbinding count") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(
    x        = "TE log2 fold change (si3d vs. sictrl)",
    y        = "Cumulative fraction",
    title    = "TE LFC distribution by eIF3 CLIP binding count",
    subtitle = "All mRNAs | normoxia 1+4hr (si3d vs. sictrl)"
  ) +
  theme_cowplot() +
  theme(legend.position = "right")+
  xlim(-0.3, 1.1)

p_lfc_ecdf_collapsed
```

```{r ks_test_collapsed}
ks_collapsed <- ks.test(
  filter(lfc_clip_collapsed, clip_group == "≥1 (any)") %>% pull(te_lfc),
  filter(lfc_clip_collapsed, clip_group == "0 (none)")  %>% pull(te_lfc)
)
cat("KS test (≥1 vs 0): D =", round(ks_collapsed$statistic, 4),
    "  p =", formatC(ks_collapsed$p.value, format = "e", digits = 2), "\n")
```

```{r save_lfc_ecdf_collapsed}
ggsave(
  here("plots", "ecdf_te_lfc_by_clip_count_collapsed_normo_promoted.pdf"),
  p_lfc_ecdf_collapsed, width = 6, height = 4.5
)
ggsave(
  here("plots", "ecdf_te_lfc_by_clip_count_collapsed_normo_promoted.png"),
  p_lfc_ecdf_collapsed, width = 6, height = 4.5, dpi = 300
)
cat("Collapsed plot saved.\n")
```

## Session info

```{r session_info}
sessionInfo()
```
