---
title: "`r paste0('Cross-Model Feature Importance: eIF3d Normoxia vs Hypoxia (', params$feature_set, ', clip=', params$clip_mode, ', lfc', params$lfc, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_set: "kozak"
  clip_mode: "aggregate"
  lfc: "0.5"
editor_options:
  chunk_output_type: console
---

```{r}
rm(params)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r derived_values}
# Defaults for interactive use (overwritten by YAML params when knitting)
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_set = "kozak",
    clip_mode   = "aggregate",
    lfc         = "0.5"
  )
}

suffix_norm <- paste0("eif3", "3d", "_promotes_normoxia_1and4hr_lfc", params$lfc,
                      "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")

suffix_hyp  <- paste0("eif3", "3d", "_promotes_hypoxia_1hr_lfc", params$lfc,
                      "_", params$feature_set, "_clip", params$clip_mode, "_no_boruta")

subtitle_label <- paste0("eIF3d normoxia 1+4hr vs hypoxia 1hr | lfc", params$lfc,
                         " | ", params$feature_set, " | clip=", params$clip_mode)

cat("Normoxia suffix:", suffix_norm, "\n")
cat("Hypoxia suffix:", suffix_hyp, "\n")
```

# Overview

Compare RF variable importance between the eIF3d normoxia (1+4hr, time-regressed) model and the eIF3d hypoxia (1hr) model to ask: **do the same mRNA sequence features predict eIF3d-promoted translation in normoxia AND hypoxia, or are condition-specific features responsible?**

Both models are trained independently on the same feature universe (no circular feature exclusion needed since neither model contains features derived from the other's outcome). Raw impurity-based importance is not comparable across independently trained forests. We normalize each model's importance to **relative proportions** (sum to 1) before comparing.

**Prerequisites:**
- `output/predictive_modeling/rf_classification_importance_{suffix_norm}.csv` — from notebook 07
- `output/predictive_modeling/rf_classification_importance_{suffix_hyp}.csv` — from notebook 07

# Setup

```{r load_libraries}
library(tidyverse)
library(here)
library(cowplot)
library(ggrepel)
```

# Load Data

```{r load_data}
imp_norm <- read_csv(here("output", "predictive_modeling",
                          paste0("rf_classification_importance_", suffix_norm, ".csv")))
imp_hyp  <- read_csv(here("output", "predictive_modeling",
                          paste0("rf_classification_importance_", suffix_hyp, ".csv")))

cat("Normoxia features:", nrow(imp_norm), "\n")
cat("Hypoxia features:", nrow(imp_hyp), "\n")

# Directionality: correlation = point-biserial between feature value and class label
# Positive = higher in "promotes" class, negative = lower
dir_norm <- read_csv(here("output", "predictive_modeling",
                          paste0("rf_feature_directionality_", suffix_norm, ".csv"))) %>%
  select(feature, correlation_norm = correlation)
dir_hyp  <- read_csv(here("output", "predictive_modeling",
                          paste0("rf_feature_directionality_", suffix_hyp, ".csv"))) %>%
  select(feature, correlation_hyp = correlation)

cat("Normoxia directionality features:", nrow(dir_norm), "\n")
cat("Hypoxia directionality features:", nrow(dir_hyp), "\n")
```

# Normalize and Join

```{r join_data}
# Normalize to relative importance (proportions summing to 1)
imp_norm <- imp_norm %>% mutate(rel_importance = importance / sum(importance))
imp_hyp  <- imp_hyp  %>% mutate(rel_importance = importance / sum(importance))

cat("Normoxia importance sums to:", sum(imp_norm$rel_importance), "\n")
cat("Hypoxia importance sums to:", sum(imp_hyp$rel_importance), "\n")

# Inner join on feature (features present in both models)
joined <- inner_join(
  imp_norm %>% select(feature, feature_clean, category, rel_importance_norm = rel_importance),
  imp_hyp  %>% select(feature, rel_importance_hyp = rel_importance),
  by = "feature"
) %>%
  mutate(
    delta      = rel_importance_norm - rel_importance_hyp,
    total      = rel_importance_norm + rel_importance_hyp,
    delta_sign = ifelse(delta >= 0, "Higher in normoxia", "Higher in hypoxia")
  ) %>%
  arrange(desc(total))

# Add directionality
joined <- joined %>%
  left_join(dir_norm, by = "feature") %>%
  left_join(dir_hyp,  by = "feature") %>%
  mutate(
    signed_norm = rel_importance_norm * sign(correlation_norm),
    signed_hyp  = rel_importance_hyp  * sign(correlation_hyp),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    ))
  )

cat("Features in both models (inner join):", nrow(joined), "\n")
cat("Features only in normoxia model:", sum(!imp_norm$feature %in% imp_hyp$feature), "\n")
cat("Features only in hypoxia model:", sum(!imp_hyp$feature %in% imp_norm$feature), "\n")
cat("Features with direction in both models:", sum(!is.na(joined$signed_norm) & !is.na(joined$signed_hyp)), "\n")
```

# Plots

## 1. Scatter Plot: Normoxia vs Hypoxia Importance

```{r scatter_plot, fig.width=10, fig.height=9}
label_data <- joined %>%
  arrange(desc(total)) %>%
  head(15)

axis_max <- max(joined$rel_importance_norm, joined$rel_importance_hyp) * 1.05

p_scatter <- ggplot(joined, aes(x = rel_importance_hyp, y = rel_importance_norm)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = category), size = 3, alpha = 0.7) +
  geom_text_repel(data = label_data, aes(label = feature_clean),
                  size = 3, max.overlaps = 20, segment.color = "grey50") +
  coord_fixed(xlim = c(0, axis_max), ylim = c(0, axis_max)) +
  theme_cowplot() +
  labs(
    title = "Feature Importance: eIF3d Normoxia vs Hypoxia Models",
    subtitle = subtitle_label,
    x = "Relative Importance (eIF3d hypoxia 1hr model)",
    y = "Relative Importance (eIF3d normoxia 1+4hr model)",
    color = "Category"
  )

print(p_scatter)

ggsave(here("plots", paste0("rf_importance_comparison_scatter_", suffix_norm, "_vs_", suffix_hyp, ".pdf")),
       plot = p_scatter, width = 10, height = 9)
```

## 2. Signed Importance Scatter: Direction + Importance in One View

```{r signed_scatter, fig.width=10, fig.height=9}
category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

signed_data <- joined %>% filter(!is.na(signed_norm), !is.na(signed_hyp))

axis_lim <- max(abs(c(signed_data$signed_norm, signed_data$signed_hyp))) * 1.1
x_min <- -axis_lim; x_max <- axis_lim
y_min <- -axis_lim; y_max <- axis_lim

label_signed <- signed_data %>%
  mutate(abs_total = abs(signed_norm) + abs(signed_hyp)) %>%
  arrange(desc(abs_total)) %>%
  head(20)

p_signed <- ggplot(signed_data, aes(x = signed_hyp, y = signed_norm)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_abline(intercept = 0, slope = -1, linetype = "dotted", color = "grey80") +
  geom_point(aes(color = broad_category), size = 3, alpha = 0.7) +
  scale_color_manual(values = category_colors, name = "Category") +
  geom_text_repel(data = label_signed, aes(label = feature_clean),
                  size = 4.5, max.overlaps = Inf, segment.color = "grey50") +
  annotate("text", x = x_min * 0.97, y = y_max * 0.97,
           label = "Higher in normoxia\nLower in hypoxia",
           hjust = 0, vjust = 1, size = 4, color = "grey45") +
  annotate("text", x = x_max * 0.97, y = y_max * 0.97,
           label = "Higher in both",
           hjust = 1, vjust = 1, size = 4, color = "grey45") +
  annotate("text", x = x_min * 0.97, y = y_min * 0.97,
           label = "Lower in both",
           hjust = 0, vjust = 0, size = 4, color = "grey45") +
  annotate("text", x = x_max * 0.97, y = y_min * 0.97,
           label = "Lower in normoxia\nHigher in hypoxia",
           hjust = 1, vjust = 0, size = 4, color = "grey45") +
  coord_fixed(xlim = c(x_min, x_max), ylim = c(y_min, y_max)) +
  theme_cowplot(font_size = 14) +
  labs(
    title = "Signed Feature Importance: eIF3d Normoxia vs Hypoxia Models",
    subtitle = paste0(subtitle_label, "  |  axis = importance × direction (+= higher in promotes)"),
    x = "Signed Importance (eIF3d hypoxia 1hr model)",
    y = "Signed Importance (eIF3d normoxia 1+4hr model)"
  )

print(p_signed)

ggsave(here("plots", paste0("rf_importance_comparison_signed_", suffix_norm, "_vs_", suffix_hyp, ".pdf")),
       plot = p_signed, width = 10, height = 9)
```

## 3. Dumbbell Plot: Paired Importance by Condition

```{r dumbbell_plot, fig.width=10, fig.height=10}
dumbbell_data <- joined %>%
  mutate(max_importance = pmax(rel_importance_norm, rel_importance_hyp)) %>%
  arrange(desc(max_importance)) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, max_importance))

p_dumbbell <- ggplot(dumbbell_data) +
  geom_segment(aes(x = rel_importance_norm, xend = rel_importance_hyp,
                   y = feature_clean, yend = feature_clean),
               color = "grey60", linewidth = 0.6) +
  geom_point(aes(x = rel_importance_norm, y = feature_clean, color = "Normoxia 1+4hr"), size = 3) +
  geom_point(aes(x = rel_importance_hyp,  y = feature_clean, color = "Hypoxia 1hr"),    size = 3) +
  scale_color_manual(
    values = c("Normoxia 1+4hr" = "#009E73", "Hypoxia 1hr" = "#E69F00"),
    name = "Condition"
  ) +
  theme_cowplot() +
  labs(
    title = "Top 30 Features: Paired Importance by Condition",
    subtitle = subtitle_label,
    x = "Relative Importance (Proportion)",
    y = ""
  )

print(p_dumbbell)

ggsave(here("plots", paste0("rf_importance_comparison_dumbbell_", suffix_norm, "_vs_", suffix_hyp, ".pdf")),
       plot = p_dumbbell, width = 10, height = 10)
```

## 4. Delta Importance Bar Plot

```{r delta_bar_plot, fig.width=10, fig.height=10}
delta_data <- joined %>%
  arrange(desc(abs(delta))) %>%
  head(30) %>%
  mutate(feature_clean = fct_reorder(feature_clean, delta))

p_delta <- ggplot(delta_data, aes(x = feature_clean, y = delta, fill = delta_sign)) +
  geom_col() +
  scale_fill_manual(
    values = c("Higher in normoxia" = "#009E73", "Higher in hypoxia" = "#E69F00"),
    name = ""
  ) +
  coord_flip() +
  theme_cowplot() +
  labs(
    title = "Delta Importance: eIF3d Normoxia minus Hypoxia",
    subtitle = subtitle_label,
    x = "",
    y = "Delta Relative Importance (normoxia - hypoxia)"
  )

print(p_delta)

ggsave(here("plots", paste0("rf_importance_comparison_delta_", suffix_norm, "_vs_", suffix_hyp, ".pdf")),
       plot = p_delta, width = 10, height = 10)
```

# Export

```{r export}
output_dir <- here("output", "predictive_modeling")

export_df <- joined %>%
  select(feature, feature_clean, category,
         rel_importance_norm, rel_importance_hyp, delta, total, delta_sign,
         correlation_norm, correlation_hyp, signed_norm, signed_hyp)

write_csv(export_df, file.path(output_dir,
  paste0("rf_importance_comparison_", suffix_norm, "_vs_", suffix_hyp, ".csv")))

cat("Exported:", nrow(export_df), "features to\n")
cat(" ", file.path(output_dir, paste0("rf_importance_comparison_", suffix_norm, "_vs_", suffix_hyp, ".csv")), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
