---
title: "Codon Score Frame Controls (CAI, tAI, CSC, FOP)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Negative control: compute CAI, tAI, CSC, and FOP for reading frames 0, 1, and 2.
Frame 0 is the canonical CDS; frames 1 and 2 are out-of-frame shifts (drop first
1 or 2 nt, trim trailing nt to restore divisibility by 3).

Reference weights (CAI relative adaptiveness, tAI w-values, FOP optimal codons) are
derived from frame 0 reference genes and held fixed across frames. If codon optimality
scores reflect real biology, frame 0 should score higher than frames 1/2.

Visualized for eIF3d hypoxia-promoted genes (1 hr, lfc > 0.5).

**Prerequisites:**
- `output/predictive_modeling/feature_matrix_kozak.rds` (from 01h)
- `output/genesets/hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- `output/condition_level_te.csv`
- `accessories/human/txdb.gencode49.sqlite`
- `accessories/human/human_trna_gcn.csv`

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(coRdon)
library(iCodon)
library(ggridges)
library(ggh4x)
```

# Load Feature Matrix and CDS Sequences

```{r load_data}
te_lfc_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_kozak.rds"))
keep_tx <- te_lfc_matrix$transcript_id_clean

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all        <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all   <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx    <- cds_names_clean %in% keep_tx
cds_frame0  <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

# Require frame 0 to be divisible by 3 (same filter as 01g)
valid <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("Frame 0 sequences:", length(cds_frame0), "\n")
```

# Build Per-Frame CDS

Shift by 1 or 2 nt, then trim trailing nt so length is still divisible by 3.

```{r build_frames}
shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)   # drop first `shift` nt
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}

cds_frame1 <- shift_frame(cds_frame0, 1)
cds_frame2 <- shift_frame(cds_frame0, 2)

# Verify all divisible by 3
stopifnot(all(width(cds_frame1) %% 3 == 0),
          all(width(cds_frame2) %% 3 == 0))

cat("Frame 1 sequences:", length(cds_frame1), "\n")
cat("Frame 2 sequences:", length(cds_frame2), "\n")
```

# Build Reference Weights from Frame 0

Reference weights are computed once from frame 0 and applied across all frames.
This is the correct design: optimal codons and relative adaptiveness are defined
biologically, not by the frame being tested.

```{r ref_weights}
# Standard genetic code
gen_code    <- Biostrings::getGeneticCode("SGC0")
stop_codons <- names(gen_code[gen_code == "*"])
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

# Frame 0 codon counts (used to derive reference weights)
# coRdon::codonCounts() returns NULL rownames by default; set them explicitly
# so the FOP reference filter correctly matches transcripts (mirrors 01g fix).
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

# --- CAI weights (ribosomal protein reference) ---
ribo_tx     <- te_lfc_matrix %>%
  filter(str_detect(symbol, "^RPL|^RPS")) %>%
  pull(transcript_id_clean)
ribo_subset <- names(cds_frame0) %in% ribo_tx
cat("Ribosomal reference transcripts:", sum(ribo_subset), "\n")

ref_counts_ribo <- colSums(counts_mat0[ribo_subset, , drop = FALSE])

w_cai <- setNames(rep(NA_real_, length(gen_code)), names(gen_code))
w_cai[stop_codons] <- 0

for (aa in names(aa_to_codons)) {
  codons <- aa_to_codons[[aa]]
  counts <- ref_counts_ribo[codons]
  total  <- sum(counts)
  n_syn  <- length(codons)
  if (total == 0) { w_cai[codons] <- 1; next }
  rscu <- counts / (total / n_syn)
  max_r <- max(rscu)
  w_cai[codons] <- if (max_r == 0) 0 else pmin(rscu / max_r, 1)
}

# --- tAI weights (GtRNAdb hg38) ---
trna_gcn <- read_csv(here("accessories", "human", "human_trna_gcn.csv"), comment = "#")

decode_anticodon <- function(ac) {
  x1 <- substr(ac, 1, 1); x2 <- substr(ac, 2, 2); x3 <- substr(ac, 3, 3)
  wc <- c(A = "T", T = "A", G = "C", C = "G")
  y1 <- unname(wc[x3]); y2 <- unname(wc[x2])
  if (x1 == "A") {
    y3 <- c("T", "C", "A"); svs <- c(0, 0, 0)
  } else if (x1 == "G") {
    y3 <- c("C", "T"); svs <- c(0, 0.41)
  } else if (x1 == "C") {
    y3 <- "G"; svs <- 0
  } else {
    y3 <- "A"; svs <- 0
  }
  data.frame(codon = paste0(y1, y2, y3), s = svs, stringsAsFactors = FALSE)
}

aa_of_codon <- setNames(as.character(gen_code), names(gen_code))
ws_tai <- setNames(rep(0, length(gen_code)), names(gen_code))

for (i in seq_len(nrow(trna_gcn))) {
  ac  <- trna_gcn$anticodon[i]
  aa  <- trna_gcn$amino_acid[i]
  gcn <- trna_gcn$gcn[i]
  decoded <- decode_anticodon(ac)
  for (j in seq_len(nrow(decoded))) {
    codon <- decoded$codon[j]; s_val <- decoded$s[j]
    if (!is.na(aa_of_codon[codon]) && aa_of_codon[codon] == aa)
      ws_tai[codon] <- ws_tai[codon] + (1 - s_val) * gcn
  }
}
w_tai <- ws_tai / max(ws_tai[sense_codons])

# --- FOP optimal codons (top 10% TE genes, sictrl normoxia) ---
te_condition <- read_csv(here("output", "condition_level_te.csv"))
te_ranked <- te_condition %>%
  filter(condition == "normoxia", treatment == "sictrl") %>%
  group_by(gene_id) %>%
  summarise(mean_te = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_te)) %>%
  arrange(desc(mean_te))

top10_cutoff <- ceiling(nrow(te_ranked) * 0.10)
top10_clean  <- sub("\\..*", "", te_ranked$gene_id[seq_len(top10_cutoff)])
gene_id_clean <- sub("\\..*", "", te_lfc_matrix$gene_id)
ref_tx_fop    <- te_lfc_matrix$transcript_id_clean[gene_id_clean %in% top10_clean]
ref_tx_fop    <- ref_tx_fop[ref_tx_fop %in% rownames(counts_mat0)]

ref_counts_fop    <- colSums(counts_mat0[ref_tx_fop, , drop = FALSE])
optimal_codons    <- sapply(aa_to_codons, function(codons) codons[which.max(ref_counts_fop[codons])])
optimal_codon_vec <- unname(optimal_codons)

cat("FOP reference transcripts:", length(ref_tx_fop), "\n")
cat("Optimal codons:", paste(sort(optimal_codon_vec), collapse = " "), "\n")
```

# Score Helper Functions

```{r score_helpers}
compute_cai <- function(counts_mat, w_vals, sense_codons) {
  use <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  apply(counts_mat, 1, function(g) {
    c_use <- g[use]; w_use <- w_vals[use]
    n <- sum(c_use)
    if (n == 0) return(NA_real_)
    exp(sum(c_use * log(w_use)) / n)
  })
}

compute_tai <- function(counts_mat, w_vals, sense_codons) {
  use <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  apply(counts_mat, 1, function(g) {
    c_use <- g[use]; w_use <- w_vals[use]
    n <- sum(c_use)
    if (n == 0) return(NA_real_)
    exp(sum(c_use * log(w_use)) / n)
  })
}

compute_fop <- function(counts_mat, optimal_codon_vec, stop_codons) {
  sense_cols <- setdiff(colnames(counts_mat), stop_codons)
  sense_tot  <- rowSums(counts_mat[, sense_cols, drop = FALSE])
  opti_tot   <- rowSums(counts_mat[, optimal_codon_vec, drop = FALSE])
  fop <- opti_tot / sense_tot
  fop[sense_tot == 0] <- NA
  fop
}

compute_csc <- function(seqs) {
  predict_human <- iCodon::predict_stability("human")
  cds_char <- as.character(seqs)
  unique_seqs <- unique(cds_char)
  unique_scores <- predict_human(unique_seqs)
  names(unique_scores) <- unique_seqs
  scores <- unique_scores[cds_char]
  names(scores) <- names(seqs)
  scores
}
```

# Compute Scores for All Three Frames

```{r score_frames}
frames <- list(
  frame0 = cds_frame0,
  frame1 = cds_frame1,
  frame2 = cds_frame2
)

score_list <- map(names(frames), function(fname) {
  seqs <- frames[[fname]]
  ct   <- codonTable(seqs)
  cm   <- codonCounts(ct)
  rownames(cm) <- names(seqs)

  cat("\nScoring", fname, "...\n")
  tibble(
    transcript_id_clean = names(seqs),
    frame = fname,
    cai   = compute_cai(cm, w_cai, sense_codons),
    tai   = compute_tai(cm, w_tai, sense_codons),
    csc   = compute_csc(seqs),
    fop   = compute_fop(cm, optimal_codon_vec, stop_codons)
  )
})

scores_all <- bind_rows(score_list)
scores_all$frame <- factor(scores_all$frame, levels = c("frame0", "frame1", "frame2"))

cat("\nTotal rows:", nrow(scores_all), "\n")
print(scores_all %>% group_by(frame) %>% summarise(
  cai_med = round(median(cai, na.rm = TRUE), 3),
  tai_med = round(median(tai, na.rm = TRUE), 3),
  csc_med = round(median(csc, na.rm = TRUE), 3),
  fop_med = round(median(fop, na.rm = TRUE), 3)
))
```

# Validate Frame 0 FOP Against Pre-Computed Values

Cross-check: frame 0 FOP computed here should match the `fop` column in
`feature_matrix_codon_optimality.rds` (from 01g). Any systematic discrepancy
would indicate a computation error.

```{r validate_scores}
codon_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_codon_optimality.rds"))

precomputed <- codon_matrix %>%
  dplyr::select(transcript_id_clean, cai_01g = cai, tai_01g = tai,
                csc_01g = csc, fop_01g = fop)

frame0_scores <- scores_all %>%
  filter(frame == "frame0") %>%
  dplyr::select(transcript_id_clean, cai_17 = cai, tai_17 = tai,
                csc_17 = csc, fop_17 = fop)

score_compare <- inner_join(precomputed, frame0_scores, by = "transcript_id_clean")
cat("Transcripts compared:", nrow(score_compare), "\n\n")

for (sc in c("cai", "tai", "csc", "fop")) {
  x <- score_compare[[paste0(sc, "_01g")]]
  y <- score_compare[[paste0(sc, "_17")]]
  r   <- round(cor(x, y, use = "complete.obs"), 4)
  mad <- round(mean(abs(x - y), na.rm = TRUE), 6)
  cat(sc, "— Pearson r:", r, "| mean abs diff:", mad, "\n")
}
```

# Load Gene Sets

Replicates the exact positive and negative sets from the 07 RF model
(hypoxia, 1hr, LFC 0.5, set.seed(9)):
- Positive: genes in feature matrix from `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- Negative: `negative_control_genes.csv` filtered to feature matrix, then randomly
  subsampled to match positive set size with set.seed(9)

Note: scores are restricted to the representative transcript per gene
(the one used in the feature matrix) to match model inputs exactly.

```{r load_gene_sets}
# Annotate transcripts with gene metadata
# Use only the representative transcript per gene (one row per gene in feature matrix)
tx_to_gene <- te_lfc_matrix %>%
  dplyr::select(transcript_id_clean, symbol, gene_id) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# Positive: eIF3d hypoxia promotes, 1hr
# geneset CSV is pre-filtered; matches 07 params$timepoint = "1hr"
hyp_promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
features_gene_ids <- sub("\\..*", "", te_lfc_matrix$gene_id)

pos_in_features <- hyp_promotes %>%
  filter(ensembl_gene %in% features_gene_ids)
cat("Positive genes in feature matrix:", nrow(pos_in_features), "\n")

# Negative: replicate 07 model sampling exactly (set.seed(9))
neg_controls <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

neg_in_features <- neg_controls %>%
  filter(gene_id_clean %in% features_gene_ids)
cat("Negative control genes in feature matrix:", nrow(neg_in_features), "\n")

number_to_match <- min(nrow(pos_in_features), nrow(neg_in_features))
set.seed(9)
neg_sampled_ids <- sample(neg_in_features$gene_id_clean, number_to_match)
cat("Negative controls sampled (matched to positive set):", length(neg_sampled_ids), "\n")

hyp_ids <- pos_in_features$ensembl_gene
neg_ids <- neg_sampled_ids

# tx_to_gene already has one row per representative transcript (from feature matrix)
# so this join gives one score row per gene per frame
plot_scores <- scores_all %>%
  inner_join(tx_to_gene, by = "transcript_id_clean") %>%
  filter(gene_id_clean %in% hyp_ids | gene_id_clean %in% neg_ids) %>%
  mutate(gene_set = case_when(
    gene_id_clean %in% hyp_ids ~ "eIF3d hypoxia promotes",
    gene_id_clean %in% neg_ids ~ "Negative control"
  ))

# Sanity check: no gene should appear in both sets
overlap_check <- intersect(hyp_ids, neg_ids)
cat("Gene IDs in both sets (should be 0):", length(overlap_check), "\n")

cat("\nRows per gene set (all frames):\n")
print(count(plot_scores, gene_set, frame))
```

# Prepare Plot Data

```{r prep_plot}
score_labels <- c(
  cai = "CAI (Codon Adaptive Index)",
  tai = "tAI (tRNA Adaptation Index)",
  csc = "CSC (Codon Stabilisation Coef.)",
  fop = "FOP (Frequency of Optimal Codons)"
)

set_colors   <- c("eIF3d hypoxia promotes" = "#E69F00", "Negative control" = "#000000")
frame_labels <- c(frame0 = "Frame 0 (canonical)", frame1 = "Frame +1", frame2 = "Frame +2")

# Foreground: pos/neg sets only
plot_long <- plot_scores %>%
  pivot_longer(cols = c(cai, tai, csc, fop), names_to = "score", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    score    = factor(score, levels = names(score_labels)),
    frame    = factor(frame, levels = c("frame0", "frame1", "frame2")),
    gene_set = factor(gene_set, levels = c("Negative control", "eIF3d hypoxia promotes"))
  )

# Background: all representative transcripts with scores
plot_long_all <- scores_all %>%
  inner_join(tx_to_gene, by = "transcript_id_clean") %>%
  pivot_longer(cols = c(cai, tai, csc, fop), names_to = "score", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    score = factor(score, levels = names(score_labels)),
    frame = factor(frame, levels = c("frame0", "frame1", "frame2"))
  )
```

# Compute Statistics: Positive vs Negative per Frame × Score

```{r stats}
stat_results <- plot_long %>%
  group_by(score, frame) %>%
  group_modify(~ {
    pos <- .x %>% filter(gene_set == "eIF3d hypoxia promotes") %>% pull(value)
    neg <- .x %>% filter(gene_set == "Negative control")       %>% pull(value)
    wt  <- wilcox.test(pos, neg, exact = FALSE)
    tibble(
      n_pos      = length(pos),
      n_neg      = length(neg),
      median_pos = round(median(pos, na.rm = TRUE), 4),
      median_neg = round(median(neg, na.rm = TRUE), 4),
      p_value    = wt$p.value,
      p_label    = case_when(
        wt$p.value < 0.001 ~ "***",
        wt$p.value < 0.01  ~ "**",
        wt$p.value < 0.05  ~ "*",
        TRUE               ~ "ns"
      )
    )
  }) %>%
  ungroup()

stat_results <- stat_results %>%
  mutate(direction = case_when(
    median_pos > median_neg ~ "promotes > neg_ctrl",
    median_pos < median_neg ~ "promotes < neg_ctrl",
    TRUE                    ~ "equal"
  ))

print(stat_results %>% dplyr::select(score, frame, median_pos, median_neg, direction, p_value, p_label))
```

# Visualize: Boxplots — Positive vs Negative, Faceted by Frame × Score

```{r boxplots, fig.width=12, fig.height=8}
# Join p-value labels for annotation
# Position label at top of each panel
annot_df <- plot_long %>%
  group_by(score, frame) %>%
  summarise(y_pos = max(value, na.rm = TRUE) * 1.05, .groups = "drop") %>%
  left_join(stat_results %>% dplyr::select(score, frame, p_label), by = c("score", "frame")) %>%
  mutate(gene_set = factor("eIF3d hypoxia promotes",
                           levels = c("Negative control", "eIF3d hypoxia promotes")))

p_box <- ggplot(plot_long, aes(x = gene_set, y = value, color = gene_set, fill = gene_set)) +
  geom_boxplot(alpha = 0.2, outlier.size = 0.4, linewidth = 0.7, width = 0.6) +
  geom_text(
    data = annot_df,
    aes(x = 1.5, y = y_pos, label = p_label),
    inherit.aes = FALSE, size = 4, fontface = "bold"
  ) +
  ggh4x::facet_grid2(frame ~ score,
                     scales = "free_y", independent = "y",
                     labeller = labeller(score = score_labels, frame = frame_labels)) +
  scale_x_discrete(labels = c("Neg ctrl", "Promotes")) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values = set_colors) +
  labs(
    title    = "Codon optimality: eIF3d hypoxia promotes vs negative controls",
    subtitle = "Rows = reading frame; *** p<0.001, ** p<0.01, * p<0.05, ns = not significant (Wilcoxon)",
    x = NULL, y = "Score"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "none",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    axis.text.x      = element_text(angle = 30, hjust = 1)
  )

p_box

ggsave(here("plots", "frame_control_codon_scores_boxplot.pdf"),
       p_box, width = 12, height = 8)
cat("Saved: plots/frame_control_codon_scores_boxplot.pdf\n")
```

# Visualize: Density Plots — Grey background + colored overlays, Faceted by Frame × Score

```{r density_plots, fig.width=12, fig.height=8}
# Median lines per gene_set × score × frame for vertical reference
median_lines <- plot_long %>%
  group_by(gene_set, score, frame) %>%
  summarise(med = median(value, na.rm = TRUE), .groups = "drop")

p_density <- ggplot() +
  # Grey background: all expressed genes
  geom_density(
    data = plot_long_all,
    aes(x = value, group = frame),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  # Colored overlays: negative control (black) and promotes (orange)
  geom_density(
    data = plot_long,
    aes(x = value, color = gene_set, fill = gene_set),
    alpha = 0.25, linewidth = 0.8
  ) +
  # Vertical median lines
  geom_vline(
    data = median_lines,
    aes(xintercept = med, color = gene_set),
    linetype = "dashed", linewidth = 0.6
  ) +
  ggh4x::facet_grid2(frame ~ score,
                     scales = "free", independent = "y",
                     labeller = labeller(score = score_labels, frame = frame_labels)) +
  scale_color_manual(
    values = set_colors,
    labels = c("Negative control", "eIF3d hypoxia promotes")
  ) +
  scale_fill_manual(
    values = set_colors,
    labels = c("Negative control", "eIF3d hypoxia promotes")
  ) +
  labs(
    title    = "Codon optimality: eIF3d hypoxia promotes vs negative controls",
    subtitle = "Grey = all expressed genes; Rows = reading frame; Frame 0 = canonical CDS, frames +1/+2 = shifted controls",
    x = "Score", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "top",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )

p_density

ggsave(here("plots", "frame_control_codon_scores_density.pdf"),
       p_density, width = 12, height = 8)
cat("Saved: plots/frame_control_codon_scores_density.pdf\n")
```

## CSC by Frame (standalone)

```{r csc_frame_plot, fig.width=5, fig.height=7}
csc_long     <- plot_long     %>% filter(score == "csc")
csc_long_all <- plot_long_all %>% filter(score == "csc")

median_lines_csc <- csc_long %>%
  group_by(gene_set, frame) %>%
  summarise(med = median(value, na.rm = TRUE), .groups = "drop")

p_csc_frames <- ggplot() +
  geom_density(
    data = csc_long_all,
    aes(x = value, group = frame),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  geom_density(
    data = csc_long,
    aes(x = value, color = gene_set, fill = gene_set),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_vline(
    data = median_lines_csc,
    aes(xintercept = med, color = gene_set),
    linetype = "dashed", linewidth = 0.6
  ) +
  facet_wrap(~ frame, nrow = 3,
             labeller = labeller(frame = frame_labels)) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values  = set_colors) +
  labs(
    title    = "CSC by reading frame",
    subtitle = "Grey = all genes; frame 0 = canonical CDS, +1/+2 = frameshift controls",
    x = "CSC (Codon Stabilisation Coefficient)", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "top",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )

p_csc_frames
ggsave(here("plots", "frame_control_csc_density.pdf"),
       p_csc_frames, width = 5, height = 7)
cat("Saved: plots/frame_control_csc_density.pdf\n")
```

# Statistical Summary Table

```{r stats_print}
print(stat_results)
```

# Session Info

```{r session_info}
sessionInfo()
```
