---
title: "18 — FOP Reference Comparison"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

After fixing the `coRdon::codonCounts()` NULL rownames bug in `01g`, the correct
top-10% TE reference for FOP gives T-ending optimal codons. This reverses the FOP
association: eIF3d hypoxia-promoted genes now score *lower* FOP than expected.

This notebook systematically compares three biologically motivated FOP reference
definitions to understand which, if any, correctly captures the codon preference
relevant to eIF3d-promoted translation.

| Reference | Filter | Aggregation |
|-----------|--------|-------------|
| **FOP_1** | normoxia, sictrl, 1+4+24 hr | mean TE per gene → top 10% |
| **FOP_2** | normoxia, sictrl, 1+4 hr | mean TE per gene → top 10% |
| **FOP_3** | hypoxia, sictrl, 1 hr | single timepoint → top 10% |
| **FOP_4** | normoxia, sictrl, 1+4 hr | time-regressed DESeq2 (~ hours + library) → top 10% |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(coRdon)
library(ggVennDiagram)
library(iCodon)
```

# Load Feature Matrix and CDS Sequences

```{r load_data}
feature_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_kozak.rds"))
keep_tx <- feature_matrix$transcript_id_clean

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

# Compute Codon Counts Matrix

```{r codon_counts}
# Build counts matrix once; set rownames explicitly (critical fix from 01g)
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

# Genetic code helpers
gen_code     <- Biostrings::getGeneticCode("SGC0")
stop_codons  <- names(gen_code[gen_code == "*"])
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

# FOP Scoring Function

```{r compute_fop}
compute_fop <- function(counts_mat, optimal_codon_vec, stop_codons) {
  sense_cols <- setdiff(colnames(counts_mat), stop_codons)
  sense_tot  <- rowSums(counts_mat[, sense_cols, drop = FALSE])
  opti_tot   <- rowSums(counts_mat[, optimal_codon_vec, drop = FALSE])
  fop        <- opti_tot / sense_tot
  fop[sense_tot == 0] <- NA
  fop
}
```

# Define FOP References and Score Transcripts

```{r load_te}
te_condition  <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE)
te_tr         <- read_csv(here("output", "te_time_regressed_normoxia_sictrl_1and4hr.csv"),
                          show_col_types = FALSE)
gene_id_clean <- sub("\\..*", "", feature_matrix$gene_id)
```

## FOP_1: normoxia, sictrl, all timepoints

```{r fop1}
te_ranked_1 <- te_condition %>%
  filter(condition == "normoxia", treatment == "sictrl") %>%
  group_by(gene_id) %>%
  summarise(mean_te = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_te)) %>%
  arrange(desc(mean_te))

top10_clean_1 <- sub("\\..*", "", te_ranked_1$gene_id[seq_len(ceiling(nrow(te_ranked_1) * 0.10))])
ref_tx_1 <- feature_matrix$transcript_id_clean[gene_id_clean %in% top10_clean_1]
ref_tx_1 <- ref_tx_1[ref_tx_1 %in% rownames(counts_mat0)]
cat("FOP_1 reference transcripts:", length(ref_tx_1), "\n")

ref_counts_1     <- colSums(counts_mat0[ref_tx_1, , drop = FALSE])
optimal_codons_1 <- sapply(aa_to_codons, function(codons) codons[which.max(ref_counts_1[codons])])
cat("FOP_1 optimal codons:", paste(sort(unname(optimal_codons_1)), collapse = " "), "\n")

fop_vec1 <- compute_fop(counts_mat0, unname(optimal_codons_1), stop_codons)
```

## FOP_2: normoxia, sictrl, 1+4 hr only

```{r fop2}
te_ranked_2 <- te_condition %>%
  filter(condition == "normoxia", treatment == "sictrl", hours %in% c(1, 4)) %>%
  group_by(gene_id) %>%
  summarise(mean_te = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_te)) %>%
  arrange(desc(mean_te))

top10_clean_2 <- sub("\\..*", "", te_ranked_2$gene_id[seq_len(ceiling(nrow(te_ranked_2) * 0.10))])
ref_tx_2 <- feature_matrix$transcript_id_clean[gene_id_clean %in% top10_clean_2]
ref_tx_2 <- ref_tx_2[ref_tx_2 %in% rownames(counts_mat0)]
cat("FOP_2 reference transcripts:", length(ref_tx_2), "\n")

ref_counts_2     <- colSums(counts_mat0[ref_tx_2, , drop = FALSE])
optimal_codons_2 <- sapply(aa_to_codons, function(codons) codons[which.max(ref_counts_2[codons])])
cat("FOP_2 optimal codons:", paste(sort(unname(optimal_codons_2)), collapse = " "), "\n")

fop_vec2 <- compute_fop(counts_mat0, unname(optimal_codons_2), stop_codons)
```

## FOP_3: hypoxia, sictrl, 1 hr

```{r fop3}
te_ranked_3 <- te_condition %>%
  filter(condition == "hypoxia", treatment == "sictrl", hours == 1) %>%
  group_by(gene_id) %>%
  summarise(mean_te = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_te)) %>%
  arrange(desc(mean_te))

top10_clean_3 <- sub("\\..*", "", te_ranked_3$gene_id[seq_len(ceiling(nrow(te_ranked_3) * 0.10))])
ref_tx_3 <- feature_matrix$transcript_id_clean[gene_id_clean %in% top10_clean_3]
ref_tx_3 <- ref_tx_3[ref_tx_3 %in% rownames(counts_mat0)]
cat("FOP_3 reference transcripts:", length(ref_tx_3), "\n")

ref_counts_3     <- colSums(counts_mat0[ref_tx_3, , drop = FALSE])
optimal_codons_3 <- sapply(aa_to_codons, function(codons) codons[which.max(ref_counts_3[codons])])
cat("FOP_3 optimal codons:", paste(sort(unname(optimal_codons_3)), collapse = " "), "\n")

fop_vec3 <- compute_fop(counts_mat0, unname(optimal_codons_3), stop_codons)
```

## FOP_4: normoxia, sictrl, 1+4 hr (time-regressed DESeq2)

```{r fop4}
te_ranked_4 <- te_tr %>%
  filter(!is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange))

top10_clean_4 <- sub("\\..*", "", te_ranked_4$gene_id[seq_len(ceiling(nrow(te_ranked_4) * 0.10))])
ref_tx_4 <- feature_matrix$transcript_id_clean[gene_id_clean %in% top10_clean_4]
ref_tx_4 <- ref_tx_4[ref_tx_4 %in% rownames(counts_mat0)]
cat("FOP_4 reference transcripts:", length(ref_tx_4), "\n")

ref_counts_4     <- colSums(counts_mat0[ref_tx_4, , drop = FALSE])
optimal_codons_4 <- sapply(aa_to_codons, function(codons) codons[which.max(ref_counts_4[codons])])
cat("FOP_4 optimal codons:", paste(sort(unname(optimal_codons_4)), collapse = " "), "\n")

fop_vec4 <- compute_fop(counts_mat0, unname(optimal_codons_4), stop_codons)
```

# Optimal Codon Overlap Across References

```{r venn_optimal_codons, fig.width=5, fig.height=4}
venn_list <- list(
  FOP_1 = unname(optimal_codons_1),
  FOP_2 = unname(optimal_codons_2),
  FOP_3 = unname(optimal_codons_3),
  FOP_4 = unname(optimal_codons_4)
)

p_venn <- ggVennDiagram(venn_list, label_alpha = 0) +
  scale_fill_gradient(low = "white", high = "#E69F00") +
  labs(title = "Overlap of optimal codons across FOP references") +
  theme(legend.position = "none")
p_venn
ggsave(here("plots", "fop_reference_optimal_codon_venn.pdf"), p_venn, width = 5, height = 4)
```

# Collapse to Gene Level

```{r gene_level}
tx_scores <- tibble(
  transcript_id_clean = names(cds_frame0),
  fop1 = fop_vec1,
  fop2 = fop_vec2,
  fop3 = fop_vec3,
  fop4 = fop_vec4
) %>%
  left_join(
    feature_matrix %>% dplyr::select(transcript_id_clean, gene_id, symbol),
    by = "transcript_id_clean"
  ) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  group_by(ensembl_gene, symbol) %>%
  summarise(
    fop1 = mean(fop1, na.rm = TRUE),
    fop2 = mean(fop2, na.rm = TRUE),
    fop3 = mean(fop3, na.rm = TRUE),
    fop4 = mean(fop4, na.rm = TRUE),
    .groups = "drop"
  )

cat("Gene-level rows:", nrow(tx_scores), "\n")
```

# Load Gene Sets

```{r gene_sets}
promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids <- promotes$ensembl_gene
cat("Promotes genes:", length(promotes_ids), "\n")

neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)   # exclude any overlap before sampling
cat("Neg pool after removing promotes overlap:", nrow(neg_pool), "\n")

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)
cat("Negative control genes (size-matched):", length(neg_ctrl_ids), "\n")
cat("Overlap (should be 0):", length(intersect(promotes_ids, neg_ctrl_ids)), "\n")
```

# Density Plots

```{r fop1_plot, fig.width=5, fig.height=4}
bg1   <- tx_scores %>% filter(!is.na(fop1))
fg_p1 <- bg1 %>% filter(ensembl_gene %in% promotes_ids)
fg_n1 <- bg1 %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks1   <- ks.test(fg_p1$fop1, bg1 %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(fop1))
cat("FOP_1 — promotes n =", nrow(fg_p1), "| neg ctrl n =", nrow(fg_n1),
    "| KS p =", formatC(ks1$p.value, format = "e", digits = 2), "\n")

p1 <- ggplot() +
  geom_density(data = bg1,   aes(x = fop1), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p1, aes(x = fop1), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n1, aes(x = fop1), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks1$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg1), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p1), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n1), ")"), size = 3.5, color = "#000000") +
  labs(title = "FOP_1: normoxia/sictrl all hours", x = "FOP", y = "Density") +
  theme_classic(base_size = 13)
p1
ggsave(here("plots", "fop_ref1_normoxia_all_hours.pdf"), p1, width = 5, height = 4)
```

```{r fop2_plot, fig.width=5, fig.height=4}
bg2   <- tx_scores %>% filter(!is.na(fop2))
fg_p2 <- bg2 %>% filter(ensembl_gene %in% promotes_ids)
fg_n2 <- bg2 %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks2   <- ks.test(fg_p2$fop2, bg2 %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(fop2))
cat("FOP_2 — promotes n =", nrow(fg_p2), "| neg ctrl n =", nrow(fg_n2),
    "| KS p =", formatC(ks2$p.value, format = "e", digits = 2), "\n")

p2 <- ggplot() +
  geom_density(data = bg2,   aes(x = fop2), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p2, aes(x = fop2), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n2, aes(x = fop2), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks2$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg2), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p2), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n2), ")"), size = 3.5, color = "#000000") +
  labs(title = "FOP_2: normoxia/sictrl 1+4 hr", x = "FOP", y = "Density") +
  theme_classic(base_size = 13)
p2
ggsave(here("plots", "fop_ref2_normoxia_1and4hr.pdf"), p2, width = 5, height = 4)
```

```{r fop3_plot, fig.width=5, fig.height=4}
bg3   <- tx_scores %>% filter(!is.na(fop3))
fg_p3 <- bg3 %>% filter(ensembl_gene %in% promotes_ids)
fg_n3 <- bg3 %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks3   <- ks.test(fg_p3$fop3, bg3 %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(fop3))
cat("FOP_3 — promotes n =", nrow(fg_p3), "| neg ctrl n =", nrow(fg_n3),
    "| KS p =", formatC(ks3$p.value, format = "e", digits = 2), "\n")

p3 <- ggplot() +
  geom_density(data = bg3,   aes(x = fop3), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p3, aes(x = fop3), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n3, aes(x = fop3), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks3$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg3), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p3), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n3), ")"), size = 3.5, color = "#000000") +
  labs(title = "FOP_3: hypoxia/sictrl 1 hr", x = "FOP", y = "Density") +
  theme_classic(base_size = 13)
p3
ggsave(here("plots", "fop_ref3_hypoxia_1hr.pdf"), p3, width = 5, height = 4)
```

```{r fop4_plot, fig.width=5, fig.height=4}
bg4   <- tx_scores %>% filter(!is.na(fop4))
fg_p4 <- bg4 %>% filter(ensembl_gene %in% promotes_ids)
fg_n4 <- bg4 %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks4   <- ks.test(fg_p4$fop4, bg4 %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(fop4))
cat("FOP_4 — promotes n =", nrow(fg_p4), "| neg ctrl n =", nrow(fg_n4),
    "| KS p =", formatC(ks4$p.value, format = "e", digits = 2), "\n")

p4 <- ggplot() +
  geom_density(data = bg4,   aes(x = fop4), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p4, aes(x = fop4), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n4, aes(x = fop4), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks4$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg4), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p4), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n4), ")"), size = 3.5, color = "#000000") +
  labs(title = "FOP_4: normoxia/sictrl 1+4 hr (time-regressed)", x = "FOP", y = "Density") +
  theme_classic(base_size = 13)
p4
ggsave(here("plots", "fop_ref4_normoxia_time_regressed.pdf"), p4, width = 5, height = 4)
```

# Codon Usage by Group

```{r codon_usage_groups}
tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")

promotes_tx  <- tx_gene_df %>% filter(ensembl_gene %in% promotes_ids) %>% pull(transcript_id_clean)
neg_ctrl_tx  <- tx_gene_df %>% filter(ensembl_gene %in% neg_ctrl_ids) %>% pull(transcript_id_clean)

promotes_counts <- colSums(counts_mat0[promotes_tx,  , drop = FALSE])
neg_ctrl_counts <- colSums(counts_mat0[neg_ctrl_tx,  , drop = FALSE])
cat("Promotes transcripts:", length(promotes_tx), "| Neg ctrl transcripts:", length(neg_ctrl_tx), "\n")
```

```{r codon_diff}
codon_diff_df <- imap_dfr(aa_to_codons, function(codons, aa) {
  p <- promotes_counts[codons];  p_prop <- p / sum(p)
  n <- neg_ctrl_counts[codons];  n_prop <- n / sum(n)
  tibble(amino_acid = aa, codon = codons,
         promotes_prop = p_prop, neg_ctrl_prop = n_prop,
         delta = p_prop - n_prop)
})

cat("Top 10 codons by |delta| (promotes - neg_ctrl within-AA proportion):\n")
print(codon_diff_df %>% arrange(desc(abs(delta))) %>% head(10))
```

```{r codon_diff_plot, fig.width=10, fig.height=8}
multi_aa <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

plot_df <- codon_diff_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta >= 0, "promotes", "neg_ctrl"),
    codon     = reorder(codon, delta)
  )

p_codon <- ggplot(plot_df, aes(x = codon, y = delta, fill = direction)) +
  geom_col() +
  scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title = "Differential codon usage: eIF3d promotes vs negative control",
    x = "Codon", y = "Δ within-AA proportion (promotes − neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        strip.text = element_text(face = "bold"))
p_codon
ggsave(here("plots", "fop_codon_usage_diff_by_aa.pdf"), p_codon, width = 10, height = 8)
```

```{r frame_shifted_counts}
# Reuse the shift_frame approach from notebook 17
shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}

cds_frame1 <- shift_frame(cds_frame0, 1)
cds_frame2 <- shift_frame(cds_frame0, 2)

make_counts <- function(seqs) {
  ct  <- codonTable(seqs)
  mat <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

# Subset to same promotes/neg_ctrl transcripts (must be in rownames)
p1 <- promotes_tx[promotes_tx %in% rownames(counts_mat1)]
n1 <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat1)]
p2 <- promotes_tx[promotes_tx %in% rownames(counts_mat2)]
n2 <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat2)]

promotes_counts1 <- colSums(counts_mat1[p1, , drop = FALSE])
neg_ctrl_counts1 <- colSums(counts_mat1[n1, , drop = FALSE])
promotes_counts2 <- colSums(counts_mat2[p2, , drop = FALSE])
neg_ctrl_counts2 <- colSums(counts_mat2[n2, , drop = FALSE])

cat("Frame 1 — promotes:", length(p1), "| neg ctrl:", length(n1), "\n")
cat("Frame 2 — promotes:", length(p2), "| neg ctrl:", length(n2), "\n")
```

```{r codon_diff_frame_plot, fig.width=12, fig.height=9}
compute_delta <- function(prom_counts, neg_counts, frame_label) {
  imap_dfr(aa_to_codons, function(codons, aa) {
    p <- prom_counts[codons]; p_prop <- p / sum(p)
    n <- neg_counts[codons];  n_prop <- n / sum(n)
    tibble(amino_acid = aa, codon = codons,
           delta = p_prop - n_prop, frame = frame_label)
  })
}

frame_diff_df <- bind_rows(
  codon_diff_df %>% mutate(frame = "Frame 0 (coding)"),
  compute_delta(promotes_counts1, neg_ctrl_counts1, "Frame 1 (control)"),
  compute_delta(promotes_counts2, neg_ctrl_counts2, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

# Order codons by frame 0 delta for consistent x-axis
codon_order <- codon_diff_df %>%
  filter(amino_acid %in% multi_aa) %>%
  arrange(amino_acid, delta) %>%
  pull(codon)
frame_diff_df <- frame_diff_df %>%
  mutate(codon = factor(codon, levels = codon_order))

# Per-transcript Wilcoxon test for each frame × codon
# Compares promotes vs neg_ctrl within-AA proportions at single-transcript resolution
per_tx_sig <- function(counts_mat, promotes_tx, neg_ctrl_tx, aa_to_codons, frame_label) {
  p_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat)]
  n_tx <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    if (length(codons) < 2) return(NULL)
    sub_p <- counts_mat[p_tx, codons, drop = FALSE]
    sub_n <- counts_mat[n_tx, codons, drop = FALSE]
    rs_p  <- rowSums(sub_p)
    rs_n  <- rowSums(sub_n)
    map_dfr(codons, function(codon) {
      props_p <- sub_p[rs_p > 0, codon] / rs_p[rs_p > 0]
      props_n <- sub_n[rs_n > 0, codon] / rs_n[rs_n > 0]
      if (length(props_p) < 3 || length(props_n) < 3) {
        return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_, frame = frame_label))
      }
      wt <- wilcox.test(props_p, props_n, exact = FALSE)
      tibble(amino_acid = aa, codon = codon, p_val = wt$p.value, frame = frame_label)
    })
  })
}

sig_df <- bind_rows(
  per_tx_sig(counts_mat0, promotes_tx, neg_ctrl_tx, aa_to_codons, "Frame 0 (coding)"),
  per_tx_sig(counts_mat1, promotes_tx, neg_ctrl_tx, aa_to_codons, "Frame 1 (control)"),
  per_tx_sig(counts_mat2, promotes_tx, neg_ctrl_tx, aa_to_codons, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    q_val = p.adjust(p_val, method = "BH"),
    sig   = !is.na(q_val) & q_val < 0.05,
    frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

# Classify frame 0 significance relative to controls:
#   ✦ = frame 0 significant, neither control significant (clean codon signal)
#   · = frame 0 significant, at least one control also significant (composition may confound)
#   "" = frame 0 not significant
f0_sig    <- sig_df %>% filter(frame == "Frame 0 (coding)")    %>% dplyr::select(amino_acid, codon, f0_sig = sig)
ctrl_sig  <- sig_df %>% filter(frame != "Frame 0 (coding)")    %>%
  group_by(amino_acid, codon) %>% summarise(any_ctrl_sig = any(sig), .groups = "drop")

f0_labels <- f0_sig %>%
  left_join(ctrl_sig, by = c("amino_acid", "codon")) %>%
  mutate(sig_label = case_when(
    f0_sig & !any_ctrl_sig ~ "\u2726",   # ✦ clean codon signal
    f0_sig &  any_ctrl_sig ~ "\u00b7",   # · composition may confound
    TRUE                   ~ ""
  )) %>%
  dplyr::select(amino_acid, codon, sig_label) %>%
  mutate(frame = factor("Frame 0 (coding)", levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

# Aggregate to amino-acid level for facet border/label highlighting
aa_any_f0 <- f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
aa_clean  <- f0_labels %>% filter(sig_label == "\u2726") %>% pull(amino_acid) %>% unique()

# Add star to aa_label for any AA with at least one frame-0 significant codon
frame_diff_df <- frame_diff_df %>%
  mutate(aa_label = ifelse(amino_acid %in% aa_any_f0,
                           paste0(aa_label, " *"), aa_label))

aa_label_map <- frame_diff_df %>% distinct(amino_acid, aa_label)
facet_clean  <- aa_label_map %>% filter(amino_acid %in% aa_clean)

p_frames <- ggplot(frame_diff_df, aes(x = codon, y = delta, fill = frame)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_rect(data = facet_clean, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5) +
  scale_fill_manual(
    values = c("Frame 0 (coding)" = "#E69F00",
               "Frame 1 (control)" = "#999999",
               "Frame 2 (control)" = "#CCCCCC")
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Differential codon usage by reading frame: frame 0 vs. frameshift controls",
    subtitle = "Frames 1 & 2 control for nucleotide composition; persistent frame-0 signal = true codon preference\nBlue border: frame 0 significant, controls not (clean codon signal). *: frame 0 significant (controls may also be significant). [q < 0.05, BH-adjusted Wilcoxon]",
    x = "Codon", y = "Δ within-AA proportion (promotes − neg ctrl)",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
    strip.text   = element_text(face = "bold"),
    legend.position = "top"
  )
p_frames
ggsave(here("plots", "fop_codon_diff_by_frame.pdf"), p_frames, width = 12, height = 9)
```

```{r codon_density_ala_leu_ser, fig.width=10, fig.height=4}
# Per-transcript codon proportion densities for Ala, Leu, Ser (frame 0 only)
# Shows the distributions underlying the Wilcoxon tests above

focal_aa <- c("A", "L", "S")

density_df <- imap_dfr(aa_to_codons[focal_aa], function(codons, aa) {
  p_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat0)]
  n_tx <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(counts_mat0)]
  sub_p <- counts_mat0[p_tx, codons, drop = FALSE]
  sub_n <- counts_mat0[n_tx, codons, drop = FALSE]
  rs_p  <- rowSums(sub_p); rs_n <- rowSums(sub_n)
  sub_p <- sub_p[rs_p > 0, , drop = FALSE]; rs_p <- rs_p[rs_p > 0]
  sub_n <- sub_n[rs_n > 0, , drop = FALSE]; rs_n <- rs_n[rs_n > 0]
  bind_rows(
    imap_dfr(as.data.frame(sub_p), function(col, codon)
      tibble(amino_acid = aa, codon = codon, proportion = col / rs_p, group = "promotes")),
    imap_dfr(as.data.frame(sub_n), function(col, codon)
      tibble(amino_acid = aa, codon = codon, proportion = col / rs_n, group = "neg ctrl"))
  )
}) %>%
  left_join(sig_df %>% filter(frame == "Frame 0 (coding)") %>%
              dplyr::select(amino_acid, codon, q_val, sig),
            by = c("amino_acid", "codon")) %>%
  mutate(
    sig         = replace_na(sig, FALSE),
    aa_label    = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon_label = ifelse(sig, paste0(codon, "*"), codon),
    group       = factor(group, levels = c("promotes", "neg ctrl"))
  )

density_colors <- c(promotes = "#E69F00", "neg ctrl" = "#888888")

for (focal in focal_aa) {
  df_aa   <- density_df %>% filter(amino_acid == focal)
  aa_name <- unique(df_aa$aa_label)
  p <- ggplot(df_aa, aes(x = proportion, fill = group, color = group)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    scale_fill_manual(values  = density_colors) +
    scale_color_manual(values = density_colors) +
    facet_wrap(~ codon_label, scales = "free", nrow = 1) +
    labs(
      title    = paste("Per-transcript codon proportions:", aa_name),
      subtitle = "* q < 0.05 in frame 0 (BH-adjusted Wilcoxon). Proportions within amino acid.",
      x = "Within-AA codon proportion (per transcript)", y = "Density",
      fill = NULL, color = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(strip.text = element_text(face = "bold", size = 9), legend.position = "top")
  print(p)
  ggsave(here("plots", paste0("fop_codon_density_", focal, ".pdf")),
         p, width = 10, height = 4)
}
```

```{r codon_density_all_frames, fig.width=10, fig.height=9}
# Same density plots but with one row per frame (frame 0, 1, 2)
# * annotated inside each cell based on that frame x codon's own adj p-value

frame_mats <- list(
  "Frame 0 (coding)"  = counts_mat0,
  "Frame 1 (control)" = counts_mat1,
  "Frame 2 (control)" = counts_mat2
)

density_frames_df <- imap_dfr(frame_mats, function(mat, frame_label) {
  imap_dfr(aa_to_codons[focal_aa], function(codons, aa) {
    p_tx <- promotes_tx[promotes_tx %in% rownames(mat)]
    n_tx <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(mat)]
    sub_p <- mat[p_tx, codons, drop = FALSE]
    sub_n <- mat[n_tx, codons, drop = FALSE]
    rs_p  <- rowSums(sub_p); rs_n <- rowSums(sub_n)
    sub_p <- sub_p[rs_p > 0, , drop = FALSE]; rs_p <- rs_p[rs_p > 0]
    sub_n <- sub_n[rs_n > 0, , drop = FALSE]; rs_n <- rs_n[rs_n > 0]
    bind_rows(
      imap_dfr(as.data.frame(sub_p), function(col, codon)
        tibble(amino_acid = aa, codon = codon, proportion = col / rs_p,
               group = "promotes", frame = frame_label)),
      imap_dfr(as.data.frame(sub_n), function(col, codon)
        tibble(amino_acid = aa, codon = codon, proportion = col / rs_n,
               group = "neg ctrl", frame = frame_label))
    )
  })
}) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    group    = factor(group, levels = c("promotes", "neg ctrl")),
    frame    = factor(frame, levels = names(frame_mats))
  )

for (focal in focal_aa) {
  df_aa   <- density_frames_df %>% filter(amino_acid == focal)
  aa_name <- unique(df_aa$aa_label)

  # Per frame x codon significance annotations for this AA
  sig_annotations <- sig_df %>%
    filter(amino_acid == focal, sig) %>%
    dplyr::select(codon, frame)

  p <- ggplot(df_aa, aes(x = proportion, fill = group, color = group)) +
    geom_density(alpha = 0.35, linewidth = 0.6) +
    geom_text(data = sig_annotations,
              aes(x = 0.5, y = Inf, label = "*"),
              inherit.aes = FALSE, hjust = 0.5, vjust = 1.5, size = 5) +
    scale_fill_manual(values  = density_colors) +
    scale_color_manual(values = density_colors) +
    facet_grid(frame ~ codon, scales = "free") +
    labs(
      title    = paste("Per-transcript codon proportions by frame:", aa_name),
      subtitle = "* q < 0.05 for that frame x codon (BH-adjusted Wilcoxon). Proportions within amino acid.",
      x = "Within-AA codon proportion (per transcript)", y = "Density",
      fill = NULL, color = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(strip.text = element_text(face = "bold", size = 9), legend.position = "top")
  print(p)
  ggsave(here("plots", paste0("fop_codon_density_frames_", focal, ".pdf")),
         p, width = 10, height = 9)
}
```

# Codon Usage Delta vs. CSC Weights

```{r csc_per_codon}
predict_human <- iCodon::predict_stability("human")

# All sense codons from aa_to_codons
all_sense_codons <- unlist(aa_to_codons, use.names = FALSE)

# CSC weight = predict_human on a homopolymer codon sequence (mean of identical weights = that weight)
codon_csc_weights <- sapply(all_sense_codons, function(cod) {
  predict_human(paste(rep(cod, 20), collapse = ""))
})

codon_csc_df <- tibble(codon = names(codon_csc_weights), csc_weight = codon_csc_weights)
cat("Per-codon CSC weights (first 6):\n"); print(head(codon_csc_df))
```

```{r codon_csc_rank, fig.width=5, fig.height=8}
# Rank all sense codons by CSC score
# Highlight codons significantly up (orange) or down (black) in eIF3d promotes vs neg ctrl
codon_rank_df <- codon_csc_df %>%
  left_join(
    imap_dfr(aa_to_codons, function(codons, aa)
      tibble(codon = codons, amino_acid = aa)),
    by = "codon"
  ) %>%
  left_join(sig_df %>% filter(frame == "Frame 0 (coding)") %>%
              dplyr::select(amino_acid, codon, sig),
            by = c("amino_acid", "codon")) %>%
  left_join(codon_diff_df %>% dplyr::select(amino_acid, codon, delta),
            by = c("amino_acid", "codon")) %>%
  mutate(
    sig       = replace_na(sig, FALSE),
    highlight = case_when(
      sig & delta > 0 ~ "up in eIF3d",
      sig & delta < 0 ~ "down in eIF3d",
      TRUE            ~ "not significant"
    ),
    highlight = factor(highlight, levels = c("up in eIF3d", "down in eIF3d", "not significant")),
    codon     = factor(codon, levels = codon[order(csc_weight)])
  )

highlight_colors <- c("up in eIF3d" = "#E69F00", "down in eIF3d" = "black", "not significant" = "grey80")
highlight_sizes  <- c("up in eIF3d" = 3, "down in eIF3d" = 3, "not significant" = 2)

p_codon_csc <- ggplot(codon_rank_df, aes(x = csc_weight, y = codon,
                                          color = highlight, size = highlight)) +
  geom_point() +
  scale_color_manual(values = highlight_colors) +
  scale_size_manual(values  = highlight_sizes) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  labs(
    title    = "Codon stability ranking by CSC score",
    subtitle = "Significant eIF3d promotes vs neg ctrl (frame 0, q < 0.05 BH-adjusted Wilcoxon)",
    x = "CSC weight", y = NULL, color = NULL, size = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(axis.text.y = element_text(size = 7), legend.position = "top")
p_codon_csc
ggsave(here("plots", "fop_codon_csc_rank.pdf"), p_codon_csc, width = 5, height = 8)
```

```{r csc_eif3d_by_aa, fig.width=12, fig.height=9}
# Plot 1: Within-AA faceted dot plot
within_aa_df <- codon_rank_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"))

p_within_aa <- ggplot(within_aa_df, aes(x = delta, y = csc_weight, color = highlight)) +
  geom_point(aes(size = highlight)) +
  geom_text(aes(label = codon), size = 2.5, hjust = -0.2, show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  scale_color_manual(values = highlight_colors) +
  scale_size_manual(values  = highlight_sizes) +
  facet_wrap(~ aa_label, scales = "free_y", nrow = 4) +
  labs(title    = "Within-AA: CSC weight vs eIF3d codon shift",
       subtitle = "Each panel = one amino acid; points colored by frame-0 significance direction",
       x = "\u0394 within-AA proportion (promotes \u2212 neg ctrl)", y = "CSC weight",
       color = NULL, size = NULL) +
  theme_classic(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "top")
p_within_aa
ggsave(here("plots", "fop_csc_by_aa.pdf"), p_within_aa, width = 12, height = 9)

# Plot 2: ECDF of CSC scores by direction group
p_ecdf <- ggplot(codon_rank_df, aes(x = csc_weight, color = highlight)) +
  stat_ecdf(linewidth = 0.8) +
  scale_color_manual(values = highlight_colors) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  labs(title    = "ECDF of CSC scores by eIF3d codon direction",
       subtitle = "Up/down = significant frame-0 Wilcoxon q < 0.05",
       x = "CSC weight", y = "Cumulative proportion", color = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_ecdf
ggsave(here("plots", "fop_csc_ecdf_direction.pdf"), p_ecdf, width = 5, height = 4)
```

```{r codon_diff_csc_plot, fig.width=6, fig.height=5}
scatter_df <- codon_diff_df %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_csc_df, by = "codon")

r_val <- cor(scatter_df$csc_weight, scatter_df$delta, use = "complete.obs")

p_csc <- ggplot(scatter_df, aes(x = csc_weight, y = delta, color = amino_acid, label = codon)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.7,
              aes(group = 1)) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 20, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  labs(
    title = "Codon usage shift vs. CSC weight",
    subtitle = sprintf("Pearson r = %.2f (promotes − neg ctrl within-AA proportion)", r_val),
    x = "Per-codon CSC weight (iCodon, human)",
    y = "Δ within-AA proportion (promotes − neg ctrl)",
    color = "Amino acid"
  ) +
  theme_classic(base_size = 12)
p_csc
ggsave(here("plots", "fop_codon_csc_scatter.pdf"), p_csc, width = 6, height = 5)
```

# Session Info

```{r session_info}
sessionInfo()
```
