---
title: "19 — Nucleotide Composition"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Compare CDS, 5' UTR, and 3' UTR nucleotide composition (A, C, G, T proportions and
GC content) across three groups: eIF3d hypoxia-promoted genes (1 hr, lfc > 0.5),
size-matched negative controls, and all expressed genes (background).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
```

# Load Feature Matrix and Sequences

```{r load_data}
feature_matrix <- readRDS(here("output", "predictive_modeling", "feature_matrix_kozak.rds"))
keep_tx <- feature_matrix$transcript_id_clean

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

# CDS
cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))
keep_idx        <- cds_names_clean %in% keep_tx
cds_frame0      <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]
valid           <- width(cds_frame0) %% 3 == 0
cds_frame0      <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")

# 5' UTR
utr5_all         <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr5_seqs_all    <- extractTranscriptSeqs(genome, utr5_all)
utr5_names_clean <- sub("\\..*", "", names(utr5_seqs_all))
keep_idx5        <- utr5_names_clean %in% keep_tx
utr5_seqs        <- utr5_seqs_all[keep_idx5]
names(utr5_seqs) <- utr5_names_clean[keep_idx5]
cat("5' UTR sequences:", length(utr5_seqs), "\n")

# 3' UTR
utr3_all         <- threeUTRsByTranscript(txdb, use.names = TRUE)
utr3_seqs_all    <- extractTranscriptSeqs(genome, utr3_all)
utr3_names_clean <- sub("\\..*", "", names(utr3_seqs_all))
keep_idx3        <- utr3_names_clean %in% keep_tx
utr3_seqs        <- utr3_seqs_all[keep_idx3]
names(utr3_seqs) <- utr3_names_clean[keep_idx3]
cat("3' UTR sequences:", length(utr3_seqs), "\n")
```

# Compute Nucleotide Composition

```{r nucleotide_composition}
compute_nuc_scores <- function(seqs, label) {
  lf <- Biostrings::letterFrequency(seqs, letters = c("A", "C", "G", "T"), as.prob = TRUE)
  tb <- tibble(
    transcript_id_clean = names(seqs),
    A  = lf[, "A"],
    C  = lf[, "C"],
    G  = lf[, "G"],
    T  = lf[, "T"],
    GC = lf[, "G"] + lf[, "C"]
  )
  row_sums <- tb$A + tb$C + tb$G + tb$T
  cat(label, "row sum check (min, max):",
      round(min(row_sums), 6), round(max(row_sums), 6), "\n")
  tb
}

nuc_cds  <- compute_nuc_scores(cds_frame0, "CDS")
nuc_utr5 <- compute_nuc_scores(utr5_seqs,  "5' UTR")
nuc_utr3 <- compute_nuc_scores(utr3_seqs,  "3' UTR")
```

# Collapse to Gene Level

```{r gene_level}
collapse_to_gene <- function(nuc_scores) {
  nuc_scores %>%
    left_join(
      feature_matrix %>% dplyr::select(transcript_id_clean, gene_id, symbol),
      by = "transcript_id_clean"
    ) %>%
    mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
    group_by(ensembl_gene, symbol) %>%
    summarise(
      A  = mean(A,  na.rm = TRUE),
      C  = mean(C,  na.rm = TRUE),
      G  = mean(G,  na.rm = TRUE),
      T  = mean(T,  na.rm = TRUE),
      GC = mean(GC, na.rm = TRUE),
      .groups = "drop"
    )
}

gene_cds  <- collapse_to_gene(nuc_cds)
gene_utr5 <- collapse_to_gene(nuc_utr5)
gene_utr3 <- collapse_to_gene(nuc_utr3)
cat("CDS gene-level rows:",    nrow(gene_cds),  "\n")
cat("5' UTR gene-level rows:", nrow(gene_utr5), "\n")
cat("3' UTR gene-level rows:", nrow(gene_utr3), "\n")

# Cross-check: GC from nucleotide frequencies vs pre-computed cds_gc in feature_matrix
gc_check <- gene_cds %>%
  left_join(
    feature_matrix %>%
      mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
      group_by(ensembl_gene) %>%
      summarise(cds_gc_ref = mean(cds_gc, na.rm = TRUE), .groups = "drop"),
    by = "ensembl_gene"
  )
cat("CDS GC correlation with feature_matrix cds_gc:",
    round(cor(gc_check$GC, gc_check$cds_gc_ref, use = "complete.obs"), 4), "\n")
```

# Load Gene Sets

```{r gene_sets}
promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids <- promotes$ensembl_gene
cat("Promotes genes:", length(promotes_ids), "\n")

neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)
cat("Neg pool after removing promotes overlap:", nrow(neg_pool), "\n")

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)
cat("Negative control genes (size-matched):", length(neg_ctrl_ids), "\n")
cat("Overlap (should be 0):", length(intersect(promotes_ids, neg_ctrl_ids)), "\n")
```

# CDS Density Plots

## GC Content

```{r cds_gc_plot, fig.width=5, fig.height=4}
bg    <- gene_cds %>% filter(!is.na(GC))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$GC, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(GC))
cat("CDS GC — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

cds_gc_plot <- ggplot() +
  geom_density(data = bg,   aes(x = GC), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = GC), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = GC), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "CDS GC content", x = "GC proportion", y = "Density") +
  theme_classic(base_size = 13)
cds_gc_plot
ggsave(here("plots", "nucleotide_gc_content.pdf"), cds_gc_plot, width = 5, height = 4)
```

## A Frequency

```{r cds_a_plot, fig.width=5, fig.height=4}
bg    <- gene_cds %>% filter(!is.na(A))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$A, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(A))
cat("CDS A — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

cds_a_plot <- ggplot() +
  geom_density(data = bg,   aes(x = A), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = A), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = A), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "CDS adenine (A) frequency", x = "A proportion", y = "Density") +
  theme_classic(base_size = 13)
cds_a_plot
ggsave(here("plots", "nucleotide_a_freq.pdf"), cds_a_plot, width = 5, height = 4)
```

## C Frequency

```{r cds_c_plot, fig.width=5, fig.height=4}
bg    <- gene_cds %>% filter(!is.na(C))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$C, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(C))
cat("CDS C — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

cds_c_plot <- ggplot() +
  geom_density(data = bg,   aes(x = C), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = C), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = C), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "CDS cytosine (C) frequency", x = "C proportion", y = "Density") +
  theme_classic(base_size = 13)
cds_c_plot
ggsave(here("plots", "nucleotide_c_freq.pdf"), cds_c_plot, width = 5, height = 4)
```

## G Frequency

```{r cds_g_plot, fig.width=5, fig.height=4}
bg    <- gene_cds %>% filter(!is.na(G))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$G, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(G))
cat("CDS G — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

cds_g_plot <- ggplot() +
  geom_density(data = bg,   aes(x = G), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = G), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = G), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "CDS guanine (G) frequency", x = "G proportion", y = "Density") +
  theme_classic(base_size = 13)
cds_g_plot
ggsave(here("plots", "nucleotide_g_freq.pdf"), cds_g_plot, width = 5, height = 4)
```

## T Frequency

```{r cds_t_plot, fig.width=5, fig.height=4}
bg    <- gene_cds %>% filter(!is.na(T))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$T, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(T))
cat("CDS T — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

cds_t_plot <- ggplot() +
  geom_density(data = bg,   aes(x = T), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = T), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = T), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "CDS thymine (T) frequency", x = "T proportion", y = "Density") +
  theme_classic(base_size = 13)
cds_t_plot
ggsave(here("plots", "nucleotide_t_freq.pdf"), cds_t_plot, width = 5, height = 4)
```

# 5' UTR Density Plots

## GC Content

```{r utr5_gc_plot, fig.width=5, fig.height=4}
bg    <- gene_utr5 %>% filter(!is.na(GC))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$GC, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(GC))
cat("5'UTR GC — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

utr5_gc_plot <- ggplot() +
  geom_density(data = bg,   aes(x = GC), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = GC), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = GC), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "5' UTR GC content", x = "GC proportion", y = "Density") +
  theme_classic(base_size = 13)
utr5_gc_plot
ggsave(here("plots", "utr5_nucleotide_gc_content.pdf"), utr5_gc_plot, width = 5, height = 4)
```


# 3' UTR Density Plots

## GC Content

```{r utr3_gc_plot, fig.width=5, fig.height=4}
bg    <- gene_utr3 %>% filter(!is.na(GC))
fg_p  <- bg %>% filter(ensembl_gene %in% promotes_ids)
fg_n  <- bg %>% filter(ensembl_gene %in% neg_ctrl_ids)
ks    <- ks.test(fg_p$GC, bg %>% filter(!ensembl_gene %in% promotes_ids) %>% pull(GC))
cat("3'UTR GC — promotes n =", nrow(fg_p), "| neg ctrl n =", nrow(fg_n),
    "| KS p =", formatC(ks$p.value, format = "e", digits = 2), "\n")

utr3_gc_plot <- ggplot() +
  geom_density(data = bg,   aes(x = GC), color = "grey60",  fill = NA, linewidth = 0.9) +
  geom_density(data = fg_p, aes(x = GC), color = "#E69F00", fill = NA, linewidth = 0.9) +
  geom_density(data = fg_n, aes(x = GC), color = "#000000", fill = NA, linewidth = 0.9) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("KS p = ", formatC(ks$p.value, format = "e", digits = 2)), size = 3.5) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 3.2,
           label = paste0("All genes (n = ", nrow(bg), ")"), size = 3.5, color = "grey50") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 4.9,
           label = paste0("eIF3d hypoxia promotes (n = ", nrow(fg_p), ")"), size = 3.5, color = "#E69F00") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 6.6,
           label = paste0("Negative control (n = ", nrow(fg_n), ")"), size = 3.5, color = "#000000") +
  labs(title = "3' UTR GC content", x = "GC proportion", y = "Density") +
  theme_classic(base_size = 13)
utr3_gc_plot
ggsave(here("plots", "utr3_nucleotide_gc_content.pdf"), utr3_gc_plot, width = 5, height = 4)
```

