---
title: "Notebook 20: CSC-Derived mRNA Stability for eIF3d-Promoted Genes"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook computes CSC (Codon Stabilization Coefficient)-based mRNA stability scores
from CDS sequences for eIF3d-promoted genes in hypoxia (1 hr) and a size-matched negative
control set, then produces a density plot comparing the distributions.

**Gene sets:**

- **Promotes**: genes whose translation is promoted by eIF3d in hypoxia at 1hr
  (`te_lfc > 0.5`, `te_padj < 0.05`; source: `output/genesets/hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`)
- **Negative control**: genes not regulated by eIF3d in any condition/timepoint
  (`|te_lfc| < 0.5` AND `te_padj > 0.3` across all 4 comparisons), size-matched to promotes

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(iCodon)
```

```{r load_gene_groups}
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
promotes_ids <- promotes$ensembl_gene

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)

cat("Promotes genes:", length(promotes_ids), "\n")
cat("Neg ctrl genes (size-matched):", length(neg_ctrl_ids), "\n")
```

```{r load_txdb_and_genome}
most_abundant_tx <- readRDS(here("output", "predictive_modeling", "precomputed_most_abundant_tx.rds"))
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
```

```{r extract_cds}
keep_tx <- most_abundant_tx$transcript_id_clean

cds <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_sequences_all <- extractTranscriptSeqs(genome, cds)
cds_names_clean   <- sub("\\..*", "", names(cds_sequences_all))

names(cds_sequences_all) <- cds_names_clean
cds_sequences <- cds_sequences_all[cds_names_clean %in% keep_tx]

valid_length  <- width(cds_sequences) %% 3 == 0
cds_sequences <- cds_sequences[valid_length]
cat("CDS sequences for scoring:", length(cds_sequences), "\n")
```

```{r compute_csc}
predict_human <- iCodon::predict_stability("human")

cds_char    <- as.character(cds_sequences)
unique_seqs <- unique(cds_char)
cat("Unique sequences:", length(unique_seqs), "\n")

unique_scores        <- predict_human(unique_seqs)

# predict_stability() returns a bare vector, so mapping scores back to sequences relies
# entirely on it preserving input order and length. Assert that before naming.
stopifnot(
  "predict_stability() did not return one score per input sequence" =
    length(unique_scores) == length(unique_seqs)
)
names(unique_scores) <- unique_seqs

csc_scores        <- unique_scores[cds_char]
names(csc_scores) <- names(cds_sequences)

cat("CSC summary:\n")
print(summary(csc_scores))

stopifnot(
  "CSC scores must be finite" = all(is.finite(csc_scores)),
  "One CSC score per CDS" = length(csc_scores) == length(cds_sequences)
)

inspect_df <- tibble(
  transcript_id_clean = names(csc_scores),
  csc                 = as.numeric(csc_scores),
  cds_length          = nchar(cds_char),
  cds_start           = substr(cds_char, 1, 30)
) %>%
  left_join(most_abundant_tx %>% dplyr::select(transcript_id_clean, symbol), by = "transcript_id_clean")

# Print a CDS preview, not the full sequences: the previous version dumped 12 complete
# coding sequences into the knitted HTML.
cat("\nLowest CSC (least stable):\n")
print(slice_min(inspect_df, csc, n = 6) %>%
        dplyr::select(symbol, csc, cds_length, cds_start))

cat("\nHighest CSC (most stable):\n")
print(slice_max(inspect_df, csc, n = 6) %>%
        dplyr::select(symbol, csc, cds_length, cds_start))
```

```{r build_table}
tx_gene <- most_abundant_tx %>%
  transmute(
    transcript_id_clean = as.character(unlist(transcript_id_clean)),
    ensembl_gene        = as.character(unlist(gene_id_clean))
  )

csc_table <- tibble(
  transcript_id_clean = names(csc_scores),
  csc                 = as.numeric(csc_scores)
) %>%
  left_join(tx_gene, by = "transcript_id_clean") %>%
  mutate(
    group = case_when(
      ensembl_gene %in% promotes_ids  ~ "eIF3d hypoxia promotes",
      ensembl_gene %in% neg_ctrl_ids  ~ "Negative control",
      TRUE                            ~ "All other genes"
    )
  )

cat("Rows:", nrow(csc_table), "\n")

csc_table |> group_by(group) |> 
  summarise(n())

scored_genes <- csc_table %>% filter(!is.na(ensembl_gene)) %>% pull(ensembl_gene) %>% unique()

promotes_missing <- setdiff(promotes_ids, scored_genes)
neg_ctrl_missing <- setdiff(neg_ctrl_ids, scored_genes)

cat("\nPromotes genes missing CSC:", length(promotes_missing), "of", length(promotes_ids), "\n")
cat("Neg ctrl genes missing CSC:", length(neg_ctrl_missing), "of", length(neg_ctrl_ids), "\n")

# Are the missing ones absent from most_abundant_tx, or absent from TxDb CDS?
in_tx <- promotes_missing %in% tx_gene$ensembl_gene
cat("  Promotes missing: in most_abundant_tx but no CDS:", sum(in_tx),
    "| not in most_abundant_tx:", sum(!in_tx), "\n")
```

```{r save_rds}
saveRDS(csc_table, here("output", "predictive_modeling", "csc_mrna_stability_eif3d_hypoxia_1hr.rds"))
cat("Saved RDS\n")
```

```{r density_plot}
set_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",   # Okabe-Ito orange
  "Negative control"       = "grey40",
  "All other genes"        = "grey80"
)

bg_df <- csc_table %>% dplyr::filter(group == "All other genes")
fg_df <- csc_table %>% dplyr::filter(group != "All other genes")

median_lines <- fg_df %>%
  group_by(group) %>%
  summarise(med = median(csc, na.rm = TRUE), .groups = "drop")

promotes_csc <- csc_table %>% dplyr::filter(group == "eIF3d hypoxia promotes") %>% pull(csc)
neg_ctrl_csc <- csc_table %>% dplyr::filter(group == "Negative control")       %>% pull(csc)

wt <- wilcox.test(promotes_csc, neg_ctrl_csc, alternative = "two.sided")

p_csc <- ggplot() +
  geom_density(
    data = bg_df,
    aes(x = csc),
    color = set_colors[["All other genes"]], fill = NA, linewidth = 0.8
  ) +
  geom_density(
    data = fg_df,
    aes(x = csc, color = group),
    fill = NA, linewidth = 0.8
  ) +
  geom_vline(
    data = median_lines,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  annotate("text",
           x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3,
           label = paste0(
             "Wilcoxon p = ", signif(wt$p.value, 3),
             "\nn promotes = ", sum(!is.na(promotes_csc)),
             "\nn neg ctrl = ", sum(!is.na(neg_ctrl_csc)),
             "\nn other = ",    sum(!is.na(bg_df$csc))
           )) +
  scale_color_manual(values = set_colors) +
  labs(
    title    = "CSC-derived mRNA stability: eIF3d-promoted genes (hypoxia 1hr)",
    subtitle = "Grey = all other genes; dashed = median",
    x = "CSC (iCodon, human)", y = "Density",
    color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_csc
ggsave(here("plots", "csc_mrna_stability_eif3d_hypoxia_1hr.pdf"),
       p_csc, width = 6, height = 5)
cat("Saved: plots/csc_mrna_stability_eif3d_hypoxia_1hr.pdf\n")
```

```{r wilcoxon_test}
# print(), not a bare assignment: the previous version computed `wt` and never
# displayed it, so the notebook's headline comparison shipped with no p-value.
print(wt)
cat("Median promotes:", round(median(promotes_csc, na.rm = TRUE), 4), "\n")
cat("Median neg ctrl:", round(median(neg_ctrl_csc, na.rm = TRUE), 4), "\n")
```

## Validation: iCodon CSC vs ORFeome mRNA Half-Life (Narula et al.)

```{r load_narula_s2}
narula_s2 <- read_csv(
  here("accessories", "csc_data", "narula_rna",
       "supp_073239.119_Supplemental_Table_S2.csv"),
  skip = 7, show_col_types = FALSE
) %>%
  dplyr::select(gene_symbol = Gene.Name, orfeome_halflife = `Pool 1 and 4`) %>%
  filter(!is.na(gene_symbol), !is.na(orfeome_halflife)) %>%
  # gene_symbol is the join key downstream; a duplicate would fan the join out
  distinct(gene_symbol, .keep_all = TRUE)

cat("Narula ORFeome genes with half-life data:", nrow(narula_s2), "\n")

# Narula et al. report ORFeome half-lives in hours (stated in the file header).
stopifnot(
  "Narula S2 half-lives must be positive" =
    all(narula_s2$orfeome_halflife > 0),
  "gene_symbol must be unique in narula_s2" =
    !any(duplicated(narula_s2$gene_symbol))
)
cat("Median ORFeome half-life:", round(median(narula_s2$orfeome_halflife), 2), "h\n")
```

```{r match_icodon_orfeome}
# Add symbol to csc_table (most_abundant_tx already in env)
csc_with_symbol <- csc_table %>%
  left_join(
    most_abundant_tx %>% transmute(
      transcript_id_clean = as.character(unlist(transcript_id_clean)),
      symbol = as.character(unlist(symbol))
    ),
    by = "transcript_id_clean"
  )

# NOTE: this is ALL genes, not the eIF3d-promoted subset. The promotes-only filter is
# deliberately not applied: only 7 promoted genes intersect Narula S2, too few to fit a
# line through. The plot titles, console output and saved filenames below therefore all
# say "all genes" — they previously said "promotes", which misdescribed the figure.
scatter_df <- csc_with_symbol %>%
  filter(!is.na(csc)) %>%
  inner_join(narula_s2, by = c("symbol" = "gene_symbol"),
             relationship = "many-to-one")

cat("Genes matched (all genes × Narula S2):", nrow(scatter_df), "\n")
cat("  of which eIF3d hypoxia promotes:",
    sum(scatter_df$group == "eIF3d hypoxia promotes"), "\n")
```

```{r scatter_icodon_vs_orfeome}
fit <- lm(orfeome_halflife ~ csc, data = scatter_df)
r   <- sign(coef(fit)[["csc"]]) * sqrt(summary(fit)$r.squared)

p_scatter <- ggplot(scatter_df, aes(x = csc, y = orfeome_halflife)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r, 3), "\nn = ", nrow(scatter_df)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs ORFeome mRNA half-life (all genes)",
    subtitle = "ORFeome from Narula et al. Pool 1 & 4",
    x = "iCodon CSC (computational)",
    y = "ORFeome mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_scatter
ggsave(here("plots", "icodon_vs_orfeome_halflife_allgenes.pdf"),
       p_scatter, width = 6, height = 5)
cat("r =", round(r, 4), " n =", nrow(scatter_df), "\n")
cat("Saved: plots/icodon_vs_orfeome_halflife_allgenes.pdf\n")
```

The same relationship with the single extreme half-life removed.

```{r scatter_icodon_vs_orfeome_no_outlier}
# Named threshold rather than a bare magic number. Removes the single ORFeome
# half-life above 1500 h, which otherwise sets the regression slope on its own.
ORFEOME_OUTLIER_CUTOFF <- 1500

scatter_df_no_out <- scatter_df %>%
  filter(orfeome_halflife < ORFEOME_OUTLIER_CUTOFF)

cat("Removed as outliers (>", ORFEOME_OUTLIER_CUTOFF, "h):",
    nrow(scatter_df) - nrow(scatter_df_no_out), "\n")

fit_no_out <- lm(orfeome_halflife ~ csc, data = scatter_df_no_out)
r_no_out   <- sign(coef(fit_no_out)[["csc"]]) * sqrt(summary(fit_no_out)$r.squared)

p_scatter_no_out <- ggplot(scatter_df_no_out, aes(x = csc, y = orfeome_halflife)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_no_out, 3), "\nn = ", nrow(scatter_df_no_out)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs ORFeome mRNA half-life (all genes, no outlier)",
    subtitle = paste0("ORFeome from Narula et al. Pool 1 & 4; half-life < ",
                      ORFEOME_OUTLIER_CUTOFF, " h"),
    x = "iCodon CSC (computational)",
    y = "ORFeome mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_scatter_no_out
ggsave(here("plots", "icodon_vs_orfeome_halflife_allgenes_no_outlier.pdf"),
       p_scatter_no_out, width = 6, height = 5)
cat("r =", round(r_no_out, 4), " n =", nrow(scatter_df_no_out), "\n")
cat("Saved: plots/icodon_vs_orfeome_halflife_allgenes_no_outlier.pdf\n")
```

## Endogenous mRNA half-life validation (Narula et al. Table S1)

```{r load_narula_s1}
narula_s1 <- read_csv(
  here("accessories", "csc_data", "narula_rna",
       "supp_073239.119_Supplemental_Table_S1.csv"),
  skip = 9, show_col_types = FALSE
) %>%
  dplyr::select(
    gene_symbol   = Gene.Name,
    halflife_cl14 = `Cell lines 1 and 4`,
    halflife_cl23 = `Cell lines 2 and 3`,
    halflife_cl56 = `Cell lines 5 and 6`
  ) %>%
  filter(!is.na(gene_symbol)) %>%
  distinct(gene_symbol, .keep_all = TRUE)

# The three cell-line-pair columns sit on very different scales (medians ~8.1 / 1.6 /
# 1.7 h, i.e. ~5x apart). Averaging them with na.rm = TRUE made a gene's half-life
# depend on WHICH columns happened to be non-NA: a cl14-only gene landed ~5x higher
# than a cl23-only gene for purely technical reasons, and the < 100 h outlier cutoff
# below then meant different things for different genes. Restricting the mean to genes
# measured in all three pairs keeps the y-axis a single consistent quantity.
cat("Narula S1 median half-life per cell-line pair:\n")
cat("  cl14:", round(median(narula_s1$halflife_cl14, na.rm = TRUE), 2), "h",
    "| cl23:", round(median(narula_s1$halflife_cl23, na.rm = TRUE), 2), "h",
    "| cl56:", round(median(narula_s1$halflife_cl56, na.rm = TRUE), 2), "h\n")

n_before_complete <- nrow(narula_s1)

narula_s1 <- narula_s1 %>%
  filter(!is.na(halflife_cl14), !is.na(halflife_cl23), !is.na(halflife_cl56)) %>%
  mutate(
    mean_halflife = rowMeans(cbind(halflife_cl14, halflife_cl23, halflife_cl56))
  ) %>%
  filter(is.finite(mean_halflife))

cat("Narula endogenous genes with half-life data:", nrow(narula_s1),
    "(complete cases only; dropped", n_before_complete - nrow(narula_s1),
    "genes missing >= 1 cell-line pair)\n")

stopifnot(
  "Narula S1 half-lives must be positive" =
    all(narula_s1$mean_halflife > 0),
  "gene_symbol must be unique in narula_s1" =
    !any(duplicated(narula_s1$gene_symbol))
)
```

```{r match_icodon_endogenous}
scatter_s1_df <- csc_with_symbol %>%
  filter(!is.na(csc)) %>%
  inner_join(narula_s1, by = c("symbol" = "gene_symbol"),
             relationship = "many-to-one")

cat("Genes matched (all × Narula S1):", nrow(scatter_s1_df), "\n")
```

```{r scatter_icodon_vs_endogenous}
fit_s1 <- lm(mean_halflife ~ csc, data = scatter_s1_df)
r_s1   <- sign(coef(fit_s1)[["csc"]]) * sqrt(summary(fit_s1)$r.squared)

p_scatter_s1 <- ggplot(scatter_s1_df, aes(x = csc, y = mean_halflife)) +
  geom_point(alpha = 0.4, color = "black", size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_s1, 3), "\nn = ", nrow(scatter_s1_df)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs mean endogenous mRNA half-life",
    subtitle = "Endogenous from Narula et al. Table S1 (mean across cell line pairs)",
    x = "iCodon CSC (computational)",
    y = "Endogenous mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_scatter_s1
ggsave(here("plots", "icodon_vs_endogenous_halflife.pdf"),
       p_scatter_s1, width = 6, height = 5)
cat("r =", round(r_s1, 4), "\n")
cat("Saved: plots/icodon_vs_endogenous_halflife.pdf\n")
```

without outliers (mean_halflife < 100 h)

```{r scatter_icodon_vs_endogenous_no_outlier}
scatter_s1_no_out <- scatter_s1_df %>%
  filter(mean_halflife < 100)

fit_s1b <- lm(mean_halflife ~ csc, data = scatter_s1_no_out)
r_s1b   <- sign(coef(fit_s1b)[["csc"]]) * sqrt(summary(fit_s1b)$r.squared)

p_scatter_s1b <- ggplot(scatter_s1_no_out, aes(x = csc, y = mean_halflife)) +
  geom_point(alpha = 0.4, color = "black", size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_s1b, 3), "\nn = ", nrow(scatter_s1_no_out)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs mean endogenous mRNA half-life (no outliers)",
    subtitle = "Endogenous from Narula et al. Table S1 (half-life < 100 h)",
    x = "iCodon CSC (computational)",
    y = "Endogenous mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_scatter_s1b
ggsave(here("plots", "icodon_vs_endogenous_halflife_no_outlier.pdf"),
       p_scatter_s1b, width = 6, height = 5)
cat("r =", round(r_s1b, 4), "\n")
cat("Saved: plots/icodon_vs_endogenous_halflife_no_outlier.pdf\n")
```

### Per cell-line-pair analysis

```{r scatter_icodon_vs_endogenous_per_cl}
cl_vars <- c(
  "Cell lines 1 and 4" = "halflife_cl14",
  "Cell lines 2 and 3" = "halflife_cl23",
  "Cell lines 5 and 6" = "halflife_cl56"
)

for (cl_label in names(cl_vars)) {
  cl_col <- cl_vars[[cl_label]]
  cl_tag <- cl_col  # already a safe filename fragment, e.g. "halflife_cl14"

  df_cl <- scatter_s1_df %>%
    filter(!is.na(.data[[cl_col]]), is.finite(.data[[cl_col]]),
           .data[[cl_col]] < 100)

  fit_cl <- lm(reformulate("csc", cl_col), data = df_cl)
  r_cl   <- sign(coef(fit_cl)[["csc"]]) * sqrt(summary(fit_cl)$r.squared)

  p_cl <- ggplot(df_cl, aes(x = csc, y = .data[[cl_col]])) +
    geom_point(alpha = 0.4, color = "black", size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("r = ", round(r_cl, 3), "\nn = ", nrow(df_cl)),
             hjust = 1.1, vjust = 1.4, size = 4) +
    labs(
      title    = paste0("iCodon CSC vs endogenous mRNA half-life\n(", cl_label, ")"),
      subtitle = "Narula et al. Table S1 (half-life < 100 h)",
      x = "iCodon CSC (computational)",
      y = "Endogenous mRNA half-life (hours)"
    ) +
    theme_classic(base_size = 12)

  out_file <- paste0("plots/icodon_vs_endogenous_halflife_", cl_tag, ".pdf")
  ggsave(here(out_file), p_cl, width = 6, height = 5)
  cat(cl_label, "— n =", nrow(df_cl), "| r =", round(r_cl, 4),
      "| Saved:", out_file, "\n")
  print(p_cl)
}
```

## Validation: iCodon CSC vs Zhu SLAMseq mRNA Half-Life (Zhu et al. Science 2024)

```{r load_zhu_slamseq}
library(readxl)

zhu_raw <- read_xlsx(
  here("accessories", "csc_data", "zhu_slamseq",
       "GSE268322_CNOT3_SLAM-seq_to_C_convertion_ratio.xlsx"),
  sheet = "CNOT3_SLAM-seq"
)

# Select gene columns + sgNT control columns only
sgnt_cols <- grep("sgNT", colnames(zhu_raw), value = TRUE)
zhu_ctrl <- zhu_raw %>%
  dplyr::select(gene_symbol = Gene_Name, gene_id = Gene_ID, all_of(sgnt_cols))

cat("Genes in Zhu SLAMseq:", nrow(zhu_ctrl), "\n")
cat("Control timepoint columns:", length(sgnt_cols), "\n")
```

```{r compute_zhu_halflives}
# Parse timepoint (hours) from column names like "sgNT-chase_0h_rep1_SLAM1"
parse_hours <- function(col) {
  as.numeric(sub(".*chase_(\\d+)h_.*", "\\1", col))
}

# Pivot to long format, compute mean ratio per gene per timepoint
zhu_long <- zhu_ctrl %>%
  pivot_longer(cols = all_of(sgnt_cols), names_to = "col_name", values_to = "tc_ratio") %>%
  mutate(timepoint = parse_hours(col_name)) %>%
  filter(!is.na(tc_ratio)) %>%
  group_by(gene_symbol, gene_id, timepoint) %>%
  summarise(mean_ratio = mean(tc_ratio, na.rm = TRUE), .groups = "drop")

# Normalize each gene to its t=0 value, then fit exponential decay.
#
# Log-linear lm, NOT bounded nls. The previous version used
# nls(..., algorithm = "port", lower = list(lambda = 1e-6)); genes whose decay is flat
# or increasing converged onto that lower bound and emerged as half-lives of
# log(2)/1e-6 = 693,147 h. The downstream filter only removed non-finite and
# non-positive values, so 768 of 12,472 genes (6.2%) entered the analysis as
# "immortal" transcripts and dominated every subsequent lm(). Reported correlation
# against iCodon CSC was r ~ 0.07 instead of ~ 0.3.
#
# This is the same method used in notebook 22 (`slamseq_fit_halflives`): fit
# log(fraction_remaining) ~ timepoint and return NA when the decay rate is not
# positive, so a flat gene is dropped rather than pinned to a bound.
MIN_TIMEPOINTS <- 3

fit_halflife <- function(df) {
  na_result <- tibble(halflife_zhu = NA_real_, fit_r2 = NA_real_, n_points = 0L)

  t0_ratio <- df$mean_ratio[df$timepoint == 0]
  if (length(t0_ratio) == 0 || is.na(t0_ratio) || t0_ratio <= 0) return(na_result)

  df_norm <- df %>%
    mutate(fraction_remaining = mean_ratio / t0_ratio) %>%
    filter(timepoint > 0, fraction_remaining > 0, is.finite(fraction_remaining))

  na_result$n_points <- nrow(df_norm)
  if (nrow(df_norm) < MIN_TIMEPOINTS) return(na_result)

  fit <- tryCatch(
    lm(log(fraction_remaining) ~ timepoint, data = df_norm),
    error = function(e) NULL
  )
  if (is.null(fit)) return(na_result)

  lambda <- -coef(fit)[["timepoint"]]
  r2     <- summary(fit)$r.squared
  if (is.na(lambda) || lambda <= 0) {
    return(tibble(halflife_zhu = NA_real_, fit_r2 = r2, n_points = nrow(df_norm)))
  }
  tibble(halflife_zhu = log(2) / lambda, fit_r2 = r2, n_points = nrow(df_norm))
}

zhu_halflives_raw <- zhu_long %>%
  group_by(gene_symbol, gene_id) %>%
  group_modify(~ fit_halflife(.x)) %>%
  ungroup()

# Exclude half-lives > 100h (near-zero decay rate = unreliable flat fit), matching the
# cutoff used in notebook 22.
MAX_HALFLIFE <- 100

zhu_halflife <- zhu_halflives_raw %>%
  filter(!is.na(halflife_zhu), is.finite(halflife_zhu),
         halflife_zhu > 0, halflife_zhu <= MAX_HALFLIFE) %>%
  # Symbol is the join key for every downstream scatter; a duplicated symbol would
  # fan those joins out.
  distinct(gene_symbol, .keep_all = TRUE)

n_excluded <- sum(zhu_halflives_raw$halflife_zhu > MAX_HALFLIFE, na.rm = TRUE)
cat("Genes fitted:", nrow(zhu_halflives_raw), "\n")
cat("Excluded halflife >", MAX_HALFLIFE, "h:", n_excluded, "\n")
cat("Genes with usable half-life:", nrow(zhu_halflife), "\n")
cat("Half-life summary (hours):\n")
print(summary(zhu_halflife$halflife_zhu))
cat("Fit R-squared summary:\n")
print(summary(zhu_halflife$fit_r2))

# Paper replication check, asserted before these half-lives are joined to anything.
# Zhu 2024 report a median control mRNA half-life of a few hours (Fig 2B). This is the
# check that would have caught the lambda-at-bound defect above: the old code gave a
# median of 6.2 h with a mean of 42,696 h.
med_zhu <- median(zhu_halflife$halflife_zhu, na.rm = TRUE)
cat("Median sgNT half-life:", round(med_zhu, 2), "h\n")
stopifnot(
  "Zhu2024 median sgNT half-life outside expected range" =
    dplyr::between(med_zhu, 2.5, 6.0),
  "Half-lives must be positive and bounded" =
    all(zhu_halflife$halflife_zhu > 0 & zhu_halflife$halflife_zhu <= MAX_HALFLIFE),
  "gene_symbol must be unique (join key for the scatter plots)" =
    !any(duplicated(zhu_halflife$gene_symbol))
)
```

```{r scatter_icodon_vs_zhu}
# Join CSC scores (with symbol) with Zhu half-lives
zhu_scatter_df <- csc_with_symbol %>%
  inner_join(zhu_halflife, by = c("symbol" = "gene_symbol"))

cat("Genes matched (CSC × Zhu):", nrow(zhu_scatter_df), "\n")

fit_zhu <- lm(halflife_zhu ~ csc, data = zhu_scatter_df)
r_zhu   <- sign(coef(fit_zhu)[["csc"]]) * sqrt(summary(fit_zhu)$r.squared)

p_zhu <- ggplot(zhu_scatter_df, aes(x = csc, y = halflife_zhu)) +
  geom_point(alpha = 0.3, size = 0.8, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_zhu, 3), "\nn = ", nrow(zhu_scatter_df)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs. Zhu SLAMseq mRNA half-life",
    subtitle = "Control (sgNT) conditions, Zhu et al. Science 2024",
    x        = "CSC (iCodon)",
    y        = "mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_zhu
ggsave(here("plots", "icodon_vs_zhu_slamseq_halflife.pdf"), p_zhu, width = 6, height = 5)
cat("r =", round(r_zhu, 4), "\n")
cat("Saved: plots/icodon_vs_zhu_slamseq_halflife.pdf\n")
```

```{r scatter_icodon_vs_zhu_no_outlier}
# IQR-based upper cutoff
q3   <- quantile(zhu_scatter_df$halflife_zhu, 0.75)
iqr  <- IQR(zhu_scatter_df$halflife_zhu)
cutoff <- q3 + 3 * iqr

zhu_scatter_no_out <- zhu_scatter_df %>%
  filter(halflife_zhu < cutoff)

fit_zhu_b <- lm(halflife_zhu ~ csc, data = zhu_scatter_no_out)
r_zhu_b   <- sign(coef(fit_zhu_b)[["csc"]]) * sqrt(summary(fit_zhu_b)$r.squared)

p_zhu_b <- ggplot(zhu_scatter_no_out, aes(x = csc, y = halflife_zhu)) +
  geom_point(alpha = 0.3, size = 0.8, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_zhu_b, 3), "\nn = ", nrow(zhu_scatter_no_out)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "iCodon CSC vs. Zhu SLAMseq mRNA half-life (no outliers)",
    subtitle = paste0("Control (sgNT), Zhu et al. 2024; half-life < ", round(cutoff, 1), " h"),
    x        = "CSC (iCodon)",
    y        = "mRNA half-life (hours)"
  ) +
  theme_classic(base_size = 12)

p_zhu_b
ggsave(here("plots", "icodon_vs_zhu_slamseq_halflife_no_outlier.pdf"), p_zhu_b, width = 6, height = 5)
cat("r =", round(r_zhu_b, 4), "\n")
cat("Cutoff:", round(cutoff, 1), "h | n =", nrow(zhu_scatter_no_out), "\n")
cat("Saved: plots/icodon_vs_zhu_slamseq_halflife_no_outlier.pdf\n")
```

## Kate CSC — Per-Codon Arithmetic Mean vs iCodon predict_stability()

The iCodon `predict_stability()` score is the output of a trained elastic-net model
using codon counts, sequence length, and cell-type context. Here we compute a simpler,
fully transparent "Kate CSC": the arithmetic mean of per-codon `human_optimality` weights
across every codon position in the CDS. Because these per-codon weights are signed
(both positive and negative), an arithmetic mean is used rather than a geometric mean.

This correlation shows how much of the iCodon model's predictions are captured by
the raw per-codon CSC values alone, without any length or context adjustments.

```{r kate_csc_weights}
# Load per-codon CSC weights from iCodon (64 codons × optimality score)
w_vec <- setNames(iCodon::human_optimality$optimality,
                  iCodon::human_optimality$codon)

cat("Per-codon weights loaded:", length(w_vec), "codons\n")
cat("Range:", round(range(w_vec), 4), "\n")
cat("Head:\n")
print(head(sort(w_vec)))
print(tail(sort(w_vec)))

# iCodon's human_optimality table is in DNA notation (T, not U), which is what makes
# w_vec[codons] match against BSgenome-derived CDS. If a package update switched it to
# RNA notation every lookup would silently return NA and kate_csc would collapse to a
# mean over only the T-free codons, with no error raised.
stopifnot(
  "iCodon human_optimality must be in DNA notation (T, not U)" =
    !any(grepl("U", names(w_vec))),
  "Expected 64 codons in human_optimality" = length(w_vec) == 64
)
```

```{r compute_kate_csc}
compute_kate_csc <- function(seq_char, w_vec) {
  n        <- nchar(seq_char)
  n_codons <- n %/% 3
  if (n_codons == 0) return(NA_real_)
  starts <- seq(1,        n_codons * 3 - 2, by = 3)
  ends   <- seq(3,        n_codons * 3,     by = 3)
  codons  <- substring(seq_char, starts, ends)
  weights <- w_vec[codons]   # NA for any unrecognised codon (e.g. stop codons)
  mean(weights, na.rm = TRUE)
}

# cds_char is already in the environment from the compute_csc chunk above.
kate_csc  <- vapply(cds_char, compute_kate_csc, numeric(1), w_vec = w_vec)
names(kate_csc) <- names(cds_sequences)

cat("Kate CSC summary:\n")
print(summary(kate_csc))
cat("Non-NA:", sum(!is.na(kate_csc)), "of", length(kate_csc), "\n")

# Guard the codon lookup rate: if weights stopped matching, mean(weights, na.rm = TRUE)
# would quietly average whatever handful of codons still resolved.
match_rate <- mean(!is.na(w_vec[unlist(
  lapply(head(cds_char, 200), function(s) {
    n <- nchar(s) %/% 3
    substring(s, seq(1, n * 3 - 2, by = 3), seq(3, n * 3, by = 3))
  })
)]))
cat("Codon lookup match rate (first 200 CDS):", round(match_rate, 4), "\n")
stopifnot(
  "Codon weight lookup is failing for most codons" = match_rate > 0.9,
  "Kate CSC must be finite where computed" =
    all(is.finite(kate_csc[!is.na(kate_csc)]))
)
```

```{r compare_csc_table}
# Build comparison table: Kate CSC vs iCodon csc_scores per transcript
compare_df <- tibble(
  transcript_id_clean = names(kate_csc),
  kate_csc            = kate_csc,
  icodon_csc          = as.numeric(csc_scores[names(kate_csc)])
) %>%
  left_join(
    most_abundant_tx %>% transmute(
      transcript_id_clean = as.character(unlist(transcript_id_clean)),
      ensembl_gene        = as.character(unlist(gene_id_clean)),
      symbol              = as.character(unlist(symbol))
    ),
    by = "transcript_id_clean"
  ) %>%
  filter(!is.na(kate_csc), !is.na(icodon_csc)) %>%
  mutate(group = if_else(ensembl_gene %in% promotes_ids,
                         "eIF3d hypoxia promotes", "All other genes"))

cat("Genes in comparison:", nrow(compare_df), "\n")
cat("Promotes with both scores:", sum(compare_df$group == "eIF3d hypoxia promotes"), "\n")
```

```{r scatter_kate_vs_icodon}
fit_k <- lm(icodon_csc ~ kate_csc, data = compare_df)
r_k   <- sign(coef(fit_k)[["kate_csc"]]) * sqrt(summary(fit_k)$r.squared)

promotes_df <- compare_df %>% filter(group == "eIF3d hypoxia promotes")
other_df    <- compare_df %>% filter(group == "All other genes")

p_kate <- ggplot() +
  geom_point(data = other_df,
             aes(x = kate_csc, y = icodon_csc),
             alpha = 0.2, size = 0.6, color = "grey60") +
  geom_point(data = promotes_df,
             aes(x = kate_csc, y = icodon_csc),
             alpha = 0.7, size = 1, color = "#E69F00") +
  geom_smooth(data = compare_df,
              aes(x = kate_csc, y = icodon_csc),
              method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_k, 3), "\nn = ", nrow(compare_df)),
           hjust = 1.1, vjust = 1.4, size = 4) +
  labs(
    title    = "Kate CSC vs iCodon predict_stability()",
    subtitle = "Orange = eIF3d hypoxia promotes; Kate CSC = arithmetic mean of per-codon human_optimality",
    x = "Kate CSC (arithmetic mean of per-codon weights)",
    # Higher predict_stability() score = MORE stable, so this is a stability score, not
    # a decay rate. (iCodon's own documentation calls it a decay rate, which inverts the
    # reading of this panel.)
    y = "iCodon CSC (predict_stability stability score)"
  ) +
  theme_classic(base_size = 12)

p_kate
ggsave(here("plots", "kate_csc_vs_icodon_csc.pdf"), p_kate, width = 6, height = 5)
cat("r =", round(r_k, 4), "\n")
cat("Saved: plots/kate_csc_vs_icodon_csc.pdf\n")
```

## Save outputs for downstream notebooks

```{r save_orfeome_csc_weights}
# Per-codon ORFeome CSC: Spearman r(codon frequency, ORFeome half-life) across genes.
# Uses Narula S2 ORFeome half-lives (already loaded as narula_s2).
# cds_char and csc_with_symbol are already in environment.

stop_codons  <- c("TAA", "TAG", "TGA")
sense_codons <- names(w_vec)[!names(w_vec) %in% stop_codons]

# Join ORFeome half-lives to CDS sequences via gene symbol
orfeome_join <- csc_with_symbol %>%
  filter(!is.na(ensembl_gene)) %>%
  inner_join(narula_s2, by = c("symbol" = "gene_symbol"),
             relationship = "many-to-one") %>%
  filter(!is.na(orfeome_halflife), is.finite(orfeome_halflife)) %>%
  mutate(seq_char = cds_char[transcript_id_clean]) %>%
  filter(!is.na(seq_char))

cat("Genes for ORFeome CSC weight computation:", nrow(orfeome_join), "\n")

# These 61 codon weights are estimated from a small intersection (Narula S2 has 355
# genes with Pool 1 and 4 half-lives; only a fraction have a scored CDS here). At this
# n the |rho| needed for nominal p < 0.05 is roughly 0.21, so the saved weights are
# noisy and should not be read as individually significant. p-values and n are saved
# alongside them below so downstream users can judge.
stopifnot(
  "Too few genes to estimate per-codon ORFeome CSC weights" = nrow(orfeome_join) >= 50
)

# Per-gene codon frequency matrix (genes × 61 sense codons)
get_codon_freq <- function(seq_char, codons_list) {
  n      <- nchar(seq_char) %/% 3
  starts <- seq(1, n * 3 - 2, by = 3)
  codons <- substring(seq_char, starts, starts + 2)
  codons <- codons[!codons %in% stop_codons]
  vapply(codons_list, function(c) sum(codons == c) / length(codons), numeric(1))
}

freq_mat <- do.call(rbind, lapply(orfeome_join$seq_char, get_codon_freq,
                                  codons_list = sense_codons))

# get_codon_freq iterates the full sense_codons list, so zero-count codons correctly
# get a frequency of 0 rather than being dropped -- every codon is scored on the same
# set of genes.
stopifnot(
  "Codon frequency matrix must be genes x 61 sense codons" =
    ncol(freq_mat) == length(sense_codons) && nrow(freq_mat) == nrow(orfeome_join),
  "Per-gene sense codon frequencies must sum to 1" =
    all(abs(rowSums(freq_mat) - 1) < 1e-9),
  "Codon frequencies must be in [0, 1]" =
    all(freq_mat >= 0 & freq_mat <= 1)
)

# Spearman r per codon, with p-value and n retained for diagnostics
orfeome_csc_stats <- purrr::map_dfr(seq_along(sense_codons), function(i) {
  ct <- suppressWarnings(cor.test(freq_mat[, i], orfeome_join$orfeome_halflife,
                                  method = "spearman", use = "complete.obs"))
  tibble(codon = sense_codons[i], spearman_r = unname(ct$estimate),
         p_value = ct$p.value, n_genes = nrow(orfeome_join))
}) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

orfeome_csc_weights <- setNames(orfeome_csc_stats$spearman_r, orfeome_csc_stats$codon)

cat("Codons scored:", nrow(orfeome_csc_stats),
    "| n genes:", nrow(orfeome_join),
    "| codons with BH p_adj < 0.05:", sum(orfeome_csc_stats$p_adj < 0.05), "\n")
cat("Top 10 stabilizing codons (highest Spearman r):\n")
print(head(sort(orfeome_csc_weights, decreasing = TRUE), 10))
cat("Top 10 destabilizing codons (lowest Spearman r):\n")
print(head(sort(orfeome_csc_weights, decreasing = FALSE), 10))

stopifnot(
  "All 61 sense codons must be weighted" = length(orfeome_csc_weights) == 61,
  "Spearman weights must lie in [-1, 1]" =
    all(orfeome_csc_weights >= -1 & orfeome_csc_weights <= 1, na.rm = TRUE),
  "Codon names must be unique" = !any(duplicated(names(orfeome_csc_weights)))
)

# Saved as a named numeric vector for backward compatibility -- notebook 23 indexes it
# directly as weight_vec[codons]. Diagnostics go to a separate file.
saveRDS(orfeome_csc_weights,
        here("output", "predictive_modeling", "orfeome_csc_weights.rds"))
saveRDS(orfeome_csc_stats,
        here("output", "predictive_modeling", "orfeome_csc_weights_stats.rds"))
cat("Saved: output/predictive_modeling/orfeome_csc_weights.rds\n")
cat("Saved: output/predictive_modeling/orfeome_csc_weights_stats.rds\n")
```

```{r save_icodon_per_gene}
# iCodon per-gene CSC scores — collapse csc_table to one value per gene.
# most_abundant_tx already provides one transcript per gene, so group_by is a safety check.
icodon_per_gene <- csc_table %>%
  filter(!is.na(ensembl_gene), !is.na(csc)) %>%
  group_by(ensembl_gene) %>%
  summarise(csc = mean(csc, na.rm = TRUE), .groups = "drop")

cat("iCodon per-gene scores:", nrow(icodon_per_gene), "genes\n")

stopifnot(
  "ensembl_gene must be unique in icodon_per_gene (join key downstream)" =
    !any(duplicated(icodon_per_gene$ensembl_gene)),
  "Per-gene CSC scores must be finite" = all(is.finite(icodon_per_gene$csc))
)

saveRDS(icodon_per_gene,
        here("output", "predictive_modeling", "icodon_csc_per_gene.rds"))
cat("Saved: output/predictive_modeling/icodon_csc_per_gene.rds\n")
```