---
title: "Notebook 20: CSC-Derived mRNA Stability for eIF3d-Promoted Genes"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook computes CSC (Codon Stabilization Coefficient)-based mRNA stability scores
from CDS sequences for eIF3d-promoted genes in hypoxia (1 hr) and a size-matched negative
control set, then produces a density plot comparing the distributions.

**Gene sets:**

- **Promotes**: genes whose translation is promoted by eIF3d in hypoxia at 1hr
  (`te_lfc > 0.5`, `te_padj < 0.05`; source: `output/genesets/hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`)
- **Negative control**: genes not regulated by eIF3d in any condition/timepoint
  (`|te_lfc| < 0.5` AND `te_padj > 0.3` across all 4 comparisons), size-matched to promotes

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(iCodon)
```

```{r load_gene_groups}
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
promotes_ids <- promotes$ensembl_gene

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv")) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)

cat("Promotes genes:", length(promotes_ids), "\n")
cat("Neg ctrl genes (size-matched):", length(neg_ctrl_ids), "\n")
```

```{r load_txdb_and_genome}
most_abundant_tx <- readRDS(here("output", "predictive_modeling", "precomputed_most_abundant_tx.rds"))
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
```

```{r extract_cds}
keep_tx <- most_abundant_tx$transcript_id_clean

cds <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_sequences_all <- extractTranscriptSeqs(genome, cds)
cds_names_clean   <- sub("\\..*", "", names(cds_sequences_all))

names(cds_sequences_all) <- cds_names_clean
cds_sequences <- cds_sequences_all[cds_names_clean %in% keep_tx]

valid_length  <- width(cds_sequences) %% 3 == 0
cds_sequences <- cds_sequences[valid_length]
cat("CDS sequences for scoring:", length(cds_sequences), "\n")
```

```{r compute_csc}
predict_human <- iCodon::predict_stability("human")

cds_char    <- as.character(cds_sequences)
unique_seqs <- unique(cds_char)
cat("Unique sequences:", length(unique_seqs), "\n")

unique_scores        <- predict_human(unique_seqs)
names(unique_scores) <- unique_seqs

csc_scores        <- unique_scores[cds_char]
names(csc_scores) <- names(cds_sequences)

cat("CSC summary:\n")
print(summary(csc_scores))

inspect_df <- tibble(
  transcript_id_clean = names(csc_scores),
  csc                 = as.numeric(csc_scores),
  cds_sequence        = cds_char
) %>%
  left_join(most_abundant_tx %>% dplyr::select(transcript_id_clean, symbol), by = "transcript_id_clean")

cat("\nLowest CSC (least stable):\n")
print(slice_min(inspect_df, csc, n = 6) %>% dplyr::select(symbol, csc, cds_sequence))

cat("\nHighest CSC (most stable):\n")
print(slice_max(inspect_df, csc, n = 6) %>% dplyr::select(symbol, csc, cds_sequence))
```

```{r build_table}
tx_gene <- most_abundant_tx %>%
  transmute(
    transcript_id_clean = as.character(unlist(transcript_id_clean)),
    ensembl_gene        = as.character(unlist(gene_id_clean))
  )

csc_table <- tibble(
  transcript_id_clean = names(csc_scores),
  csc                 = as.numeric(csc_scores)
) %>%
  left_join(tx_gene, by = "transcript_id_clean") %>%
  mutate(
    group = case_when(
      ensembl_gene %in% promotes_ids  ~ "eIF3d hypoxia promotes",
      ensembl_gene %in% neg_ctrl_ids  ~ "Negative control",
      TRUE                            ~ "All other genes"
    )
  )

cat("Rows:", nrow(csc_table), "\n")

csc_table |> group_by(group) |> 
  summarise(n())

scored_genes <- csc_table %>% filter(!is.na(ensembl_gene)) %>% pull(ensembl_gene) %>% unique()

promotes_missing <- setdiff(promotes_ids, scored_genes)
neg_ctrl_missing <- setdiff(neg_ctrl_ids, scored_genes)

cat("\nPromotes genes missing CSC:", length(promotes_missing), "of", length(promotes_ids), "\n")
cat("Neg ctrl genes missing CSC:", length(neg_ctrl_missing), "of", length(neg_ctrl_ids), "\n")

# Are the missing ones absent from most_abundant_tx, or absent from TxDb CDS?
in_tx <- promotes_missing %in% tx_gene$ensembl_gene
cat("  Promotes missing: in most_abundant_tx but no CDS:", sum(in_tx),
    "| not in most_abundant_tx:", sum(!in_tx), "\n")
```

```{r save_rds}
saveRDS(csc_table, here("output", "predictive_modeling", "csc_mrna_stability_eif3d_hypoxia_1hr.rds"))
cat("Saved RDS\n")
```

```{r density_plot}
set_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",   # Okabe-Ito orange
  "Negative control"       = "black"
)

bg_df <- csc_table %>% filter(group == "All other genes")
fg_df <- csc_table %>% filter(group != "All other genes")

median_lines <- fg_df %>%
  group_by(group) %>%
  summarise(med = median(csc, na.rm = TRUE), .groups = "drop")

p_csc <- ggplot() +
  geom_density(
    data = bg_df,
    aes(x = csc),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  geom_density(
    data = fg_df,
    aes(x = csc, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_vline(
    data = median_lines,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values  = set_colors) +
  labs(
    title    = "CSC-derived mRNA stability: eIF3d-promoted genes (hypoxia 1hr)",
    subtitle = "Grey = all other genes; dashed = median",
    x = "CSC (iCodon, human)", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position  = "top",
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold")
  )

p_csc
ggsave(here("plots", "csc_mrna_stability_eif3d_hypoxia_1hr.pdf"),
       p_csc, width = 6, height = 5)
cat("Saved: plots/csc_mrna_stability_eif3d_hypoxia_1hr.pdf\n")
```

```{r wilcoxon_test}
promotes_csc <- csc_table %>% filter(group == "eIF3d hypoxia promotes") %>% pull(csc)
neg_ctrl_csc <- csc_table %>% filter(group == "Negative control")       %>% pull(csc)

wt <- wilcox.test(promotes_csc, neg_ctrl_csc, alternative = "two.sided")

```

## Validation: iCodon CSC vs ORFeome mRNA Half-Life (Narula et al.)

```{r load_narula_s2}
narula_s2 <- read_csv(
  here("accessories", "csc_data", "narula_rna",
       "supp_073239.119_Supplemental_Table_S2.csv"),
  skip = 7, show_col_types = FALSE
) %>%
  dplyr::select(gene_symbol = Gene.Name, orfeome_halflife = `Pool 1 and 4`) %>%
  filter(!is.na(gene_symbol), !is.na(orfeome_halflife))

cat("Narula ORFeome genes with half-life data:", nrow(narula_s2), "\n")
```

```{r match_icodon_orfeome}
# Add symbol to csc_table (most_abundant_tx already in env)
csc_with_symbol <- csc_table %>%
  left_join(
    most_abundant_tx %>% transmute(
      transcript_id_clean = as.character(unlist(transcript_id_clean)),
      symbol = as.character(unlist(symbol))
    ),
    by = "transcript_id_clean"
  )

# Filter to promotes group, join with Narula
scatter_df <- csc_with_symbol %>%
  # filter(group == "eIF3d hypoxia promotes", !is.na(csc)) %>%
  inner_join(narula_s2, by = c("symbol" = "gene_symbol"))

cat("Genes matched (promotes × Narula):", nrow(scatter_df), "\n")
```

```{r scatter_icodon_vs_orfeome}
fit <- lm(orfeome_halflife ~ csc, data = scatter_df)
r   <- sign(coef(fit)[["csc"]]) * sqrt(summary(fit)$r.squared)

p_scatter <- ggplot(scatter_df, aes(x = csc, y = orfeome_halflife)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r, 3)),
           hjust = 1.1, vjust = 1.5, size = 4) +
  labs(
    title    = "iCodon CSC vs ORFeome mRNA half-life",
    subtitle = "ORFeome from Narula et al. Pool 1 & 4",
    x = "iCodon CSC (computational)",
    y = "ORFeome mRNA half-life (hours)"
  ) +
  theme_bw(base_size = 12)

p_scatter
ggsave(here("plots", "icodon_vs_orfeome_halflife_promotes.pdf"),
       p_scatter, width = 6, height = 5)
cat("r =", round(r, 4), "\n")
cat("Saved: plots/icodon_vs_orfeome_halflife_promotes.pdf\n")
```

if i remove the one outlier

```{r}
scatter_df_no_out <- scatter_df |> 
  filter(orfeome_halflife < 1500)



fit <- lm(orfeome_halflife ~ csc, data = scatter_df_no_out)
r   <- sign(coef(fit)[["csc"]]) * sqrt(summary(fit)$r.squared)

p_scatter <- ggplot(scatter_df_no_out, aes(x = csc, y = orfeome_halflife, color=group)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r, 3)),
           hjust = 1.1, vjust = 1.5, size = 4) +
  labs(
    title    = "iCodon CSC vs ORFeome mRNA half-life",
    subtitle = "ORFeome from Narula et al. Pool 1 & 4",
    x = "iCodon CSC (computational)",
    y = "ORFeome mRNA half-life (hours)"
  ) +
  theme_bw(base_size = 12)

p_scatter
ggsave(here("plots", "icodon_vs_orfeome_halflife_promotes_no_outlier.pdf"),
       p_scatter, width = 6, height = 5)
cat("r =", round(r, 4), "\n")
cat("Saved: plots/icodon_vs_orfeome_halflife_promotes.pdf\n")
```

## Endogenous mRNA half-life validation (Narula et al. Table S1)

```{r load_narula_s1}
narula_s1 <- read_csv(
  here("accessories", "csc_data", "narula_rna",
       "supp_073239.119_Supplemental_Table_S1.csv"),
  skip = 9, show_col_types = FALSE
) %>%
  dplyr::select(
    gene_symbol   = Gene.Name,
    halflife_cl14 = `Cell lines 1 and 4`,
    halflife_cl23 = `Cell lines 2 and 3`,
    halflife_cl56 = `Cell lines 5 and 6`
  ) %>%
  filter(!is.na(gene_symbol)) %>%
  mutate(
    mean_halflife = rowMeans(cbind(halflife_cl14, halflife_cl23, halflife_cl56), na.rm = TRUE)
  ) %>%
  filter(is.finite(mean_halflife))

cat("Narula endogenous genes with half-life data:", nrow(narula_s1), "\n")
```

```{r match_icodon_endogenous}
scatter_s1_df <- csc_with_symbol %>%
  inner_join(narula_s1, by = c("symbol" = "gene_symbol"))

cat("Genes matched (all × Narula S1):", nrow(scatter_s1_df), "\n")
```

```{r scatter_icodon_vs_endogenous}
fit_s1 <- lm(mean_halflife ~ csc, data = scatter_s1_df)
r_s1   <- sign(coef(fit_s1)[["csc"]]) * sqrt(summary(fit_s1)$r.squared)

p_scatter_s1 <- ggplot(scatter_s1_df, aes(x = csc, y = mean_halflife)) +
  geom_point(alpha = 0.4, color = "black", size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_s1, 3)),
           hjust = 1.1, vjust = 1.5, size = 4) +
  labs(
    title    = "iCodon CSC vs mean endogenous mRNA half-life",
    subtitle = "Endogenous from Narula et al. Table S1 (mean across cell line pairs)",
    x = "iCodon CSC (computational)",
    y = "Endogenous mRNA half-life (hours)"
  ) +
  theme_bw(base_size = 12)

p_scatter_s1
ggsave(here("plots", "icodon_vs_endogenous_halflife.pdf"),
       p_scatter_s1, width = 6, height = 5)
cat("r =", round(r_s1, 4), "\n")
cat("Saved: plots/icodon_vs_endogenous_halflife.pdf\n")
```

without outliers (mean_halflife < 100 h)

```{r scatter_icodon_vs_endogenous_no_outlier}
scatter_s1_no_out <- scatter_s1_df %>%
  filter(mean_halflife < 100)

fit_s1b <- lm(mean_halflife ~ csc, data = scatter_s1_no_out)
r_s1b   <- sign(coef(fit_s1b)[["csc"]]) * sqrt(summary(fit_s1b)$r.squared)

p_scatter_s1b <- ggplot(scatter_s1_no_out, aes(x = csc, y = mean_halflife)) +
  geom_point(alpha = 0.4, color = "black", size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_s1b, 3)),
           hjust = 1.1, vjust = 1.5, size = 4) +
  labs(
    title    = "iCodon CSC vs mean endogenous mRNA half-life (no outliers)",
    subtitle = "Endogenous from Narula et al. Table S1 (half-life < 100 h)",
    x = "iCodon CSC (computational)",
    y = "Endogenous mRNA half-life (hours)"
  ) +
  theme_bw(base_size = 12)

p_scatter_s1b
ggsave(here("plots", "icodon_vs_endogenous_halflife_no_outlier.pdf"),
       p_scatter_s1b, width = 6, height = 5)
cat("r =", round(r_s1b, 4), "\n")
cat("Saved: plots/icodon_vs_endogenous_halflife_no_outlier.pdf\n")
```

### Per cell-line-pair analysis

```{r scatter_icodon_vs_endogenous_per_cl}
cl_vars <- c(
  "Cell lines 1 and 4" = "halflife_cl14",
  "Cell lines 2 and 3" = "halflife_cl23",
  "Cell lines 5 and 6" = "halflife_cl56"
)

for (cl_label in names(cl_vars)) {
  cl_col <- cl_vars[[cl_label]]
  cl_tag <- gsub(" ", "", cl_col)  # e.g. "halflife_cl14"

  df_cl <- scatter_s1_df %>%
    filter(!is.na(.data[[cl_col]]), is.finite(.data[[cl_col]]),
           .data[[cl_col]] < 100)

  fit_cl <- lm(reformulate("csc", cl_col), data = df_cl)
  r_cl   <- sign(coef(fit_cl)[["csc"]]) * sqrt(summary(fit_cl)$r.squared)

  p_cl <- ggplot(df_cl, aes(x = csc, y = .data[[cl_col]])) +
    geom_point(alpha = 0.4, color = "black", size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("r = ", round(r_cl, 3)),
             hjust = 1.1, vjust = 1.5, size = 4) +
    labs(
      title    = paste0("iCodon CSC vs endogenous mRNA half-life\n(", cl_label, ")"),
      subtitle = "Narula et al. Table S1 (half-life < 100 h)",
      x = "iCodon CSC (computational)",
      y = "Endogenous mRNA half-life (hours)"
    ) +
    theme_bw(base_size = 12)

  out_file <- paste0("plots/icodon_vs_endogenous_halflife_", cl_tag, ".pdf")
  ggsave(here(out_file), p_cl, width = 6, height = 5)
  cat(cl_label, "— n =", nrow(df_cl), "| r =", round(r_cl, 4),
      "| Saved:", out_file, "\n")
  print(p_cl)
}
```

