---
title: "Notebook 21: CSC vs GC3 Correlation per mRNA"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook examines the correlation between CSC (Codon Stabilization Coefficient,
from iCodon) and GC3 (GC content at the third codon position) across all mRNAs in
the feature matrix. Both are CDS-level features.

GC3 is expected to correlate with CSC because GC-ending codons tend to have higher
CSC weights in human cells. This notebook quantifies the relationship and checks
whether it differs across eIF3d gene groups.

```{r load_libraries}
library(tidyverse)
library(here)
```

## Load feature matrix

```{r load_features}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_codon_optimality.rds"))

features <- features %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

cat("Rows:", nrow(features), "\n")
cat("Columns:", ncol(features), "\n")
cat("CSC non-NA:", sum(!is.na(features$csc)), "\n")
cat("GC3 non-NA:", sum(!is.na(features$gc3)), "\n")
```

## Collapse to gene level

The feature matrix is transcript-level; collapse to one value per gene using the
mean.

```{r gene_level}
gene_df <- features %>%
  group_by(ensembl_gene, symbol) %>%
  summarise(
    csc = mean(csc, na.rm = TRUE),
    gc3 = mean(gc3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(csc), !is.na(gc3))

cat("Genes with both CSC and GC3:", nrow(gene_df), "\n")
cat("CSC range:", round(range(gene_df$csc), 3), "\n")
cat("GC3 range:", round(range(gene_df$gc3), 3), "\n")
```

## Global CSC vs GC3 scatter

```{r scatter_all, fig.width=5.5, fig.height=4.5}
fit_all <- lm(csc ~ gc3, data = gene_df)
r_all   <- sign(coef(fit_all)[["gc3"]]) * sqrt(summary(fit_all)$r.squared)

p_all <- ggplot(gene_df, aes(x = gc3, y = csc)) +
  geom_point(alpha = 0.2, size = 0.6, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue", linewidth = 0.8) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", round(r_all, 3),
                          "\nn = ", nrow(gene_df)),
           hjust = 1.1, vjust = 1.5, size = 3.5) +
  labs(
    title    = "CSC vs GC3 — all genes",
    subtitle = "Gene-level means; feature_matrix_codon_optimality",
    x = "GC3 (GC content at 3rd codon position)",
    y = "CSC (iCodon mRNA stability score)"
  ) +
  theme_classic(base_size = 12)

p_all
ggsave(here("plots", "csc_vs_gc3_all_genes.pdf"), p_all, width = 5.5, height = 4.5)
cat("r =", round(r_all, 4), "\n")
cat("Saved: plots/csc_vs_gc3_all_genes.pdf\n")
```

## Annotate eIF3d gene groups

Negative control genes are size-matched to the promotes set using the same
`set.seed(9)` and filtering approach as the RF model (notebook 7): restrict both
pools to genes present in `gene_df`, then sample without replacement.

```{r load_gene_groups}
# eIF3d hypoxia promotes (1 hr, lfc > 0.5)
hyp_promotes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)

# Full negative control pool
neg_ctrl_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

# Restrict both to genes present in gene_df (i.e. have CSC and GC3)
promotes_in_df  <- hyp_promotes   %>% filter(ensembl_gene %in% gene_df$ensembl_gene)
neg_pool_in_df  <- neg_ctrl_pool  %>% filter(ensembl_gene %in% gene_df$ensembl_gene)

# Size-match neg ctrl to promotes (set.seed(9) matches notebook 7 / RF model)
set.seed(9)
n_match      <- min(nrow(promotes_in_df), nrow(neg_pool_in_df))
neg_ctrl_ids <- sample(neg_pool_in_df$ensembl_gene, size = n_match, replace = FALSE)

cat("Promotes in gene_df:", nrow(promotes_in_df), "\n")
cat("Neg ctrl (size-matched, set.seed(9)):", length(neg_ctrl_ids), "\n")

gene_df <- gene_df %>%
  mutate(
    group = case_when(
      ensembl_gene %in% promotes_in_df$ensembl_gene ~ "eIF3d hypoxia promotes",
      ensembl_gene %in% neg_ctrl_ids                 ~ "Negative control",
      TRUE                                            ~ "All other genes"
    ),
    group = factor(group, levels = c(
      "eIF3d hypoxia promotes", "Negative control", "All other genes"
    ))
  )

cat("Group counts:\n")
print(table(gene_df$group))
```

## Scatter colored by gene group (faceted)

```{r scatter_groups, fig.width=9, fig.height=4}
group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "black",
  "All other genes"        = "grey70"
)

# Per-facet n and r labels
facet_labels <- gene_df %>%
  group_by(group) %>%
  summarise(
    n = n(),
    r = {
      fit <- lm(csc ~ gc3, data = pick(everything()))
      sign(coef(fit)[["gc3"]]) * sqrt(summary(fit)$r.squared)
    },
    .groups = "drop"
  ) %>%
  mutate(label = paste0("n = ", n, "\nr = ", round(r, 3)))

# bg without group column so it shows in all facets
bg_all <- gene_df %>% dplyr::select(-group)

p_groups <- ggplot(gene_df, aes(x = gc3, y = csc)) +
  geom_point(data = bg_all,
             alpha = 0.1, size = 0.4, color = "grey70") +
  geom_point(aes(color = group), alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black",
              linewidth = 0.7, linetype = "dashed") +
  geom_text(
    data = facet_labels,
    aes(label = label),
    x = Inf, y = -Inf, hjust = 1.05, vjust = -0.3,
    size = 3, inherit.aes = FALSE
  ) +
  scale_color_manual(values = group_colors, name = NULL) +
  facet_wrap(~group) +
  labs(
    title = "CSC vs GC3 by eIF3d gene group",
    x = "GC3 (GC content at 3rd codon position)",
    y = "CSC (iCodon mRNA stability score)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

p_groups
ggsave(here("plots", "csc_vs_gc3_by_group.pdf"), p_groups, width = 9, height = 4)
cat("Saved: plots/csc_vs_gc3_by_group.pdf\n")
print(facet_labels)
```

## Per-group correlations

```{r per_group_correlations}
cat("Per-group Pearson r (CSC ~ GC3):\n")
print(facet_labels %>% dplyr::select(group, n, r))
```

## Residual distribution by group

Residuals from the global CSC ~ GC3 regression. If eIF3d-promoted genes have higher
CSC than expected from GC3 alone, their residuals will be positive.

```{r residuals, fig.width=5.5, fig.height=4}
gene_df <- gene_df %>%
  mutate(csc_resid = residuals(lm(csc ~ gc3, data = gene_df)))

resid_medians <- gene_df %>%
  filter(group != "All other genes") %>%
  group_by(group) %>%
  summarise(med = median(csc_resid, na.rm = TRUE), .groups = "drop")

p_resid <- ggplot() +
  geom_density(
    data = gene_df %>% filter(group == "All other genes"),
    aes(x = csc_resid),
    fill = NA, color = "grey60", linewidth = 0.9
  ) +
  geom_density(
    data = gene_df %>% filter(group != "All other genes"),
    aes(x = csc_resid, color = group),
    fill = NA, linewidth = 0.9
  ) +
  geom_vline(
    data = resid_medians,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  scale_color_manual(
    values = c("eIF3d hypoxia promotes" = "#E69F00",
               "Negative control"       = "black"),
    name = NULL
  ) +
  labs(
    title    = "CSC residuals after GC3 regression",
    subtitle = "Grey = all other genes; dashed = group median",
    x = "CSC residual (observed − predicted from GC3)",
    y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_resid

# Wilcoxon test: promotes vs negative control residuals
promotes_resid  <- gene_df %>% filter(group == "eIF3d hypoxia promotes") %>% pull(csc_resid)
neg_ctrl_resid  <- gene_df %>% filter(group == "Negative control")       %>% pull(csc_resid)

wt <- wilcox.test(promotes_resid, neg_ctrl_resid, alternative = "two.sided")
cat("Wilcoxon test (promotes vs neg ctrl residuals):\n")
cat("  W =", wt$statistic, ", p =", formatC(wt$p.value, format = "e", digits = 2), "\n")
cat("  Promotes median residual:", round(median(promotes_resid, na.rm = TRUE), 4), "\n")
cat("  Neg ctrl median residual:", round(median(neg_ctrl_resid, na.rm = TRUE), 4), "\n")

ggsave(here("plots", "csc_resid_gc3_by_group.pdf"), p_resid, width = 5.5, height = 4)
cat("Saved: plots/csc_resid_gc3_by_group.pdf\n")
```

## Session info

```{r session_info}
sessionInfo()
```
