---
title: "Notebook 23: DHX29 Riboseq & SLAM-seq Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook analyzes two datasets from Hia et al. 2026 examining DHX29's role in mRNA
stability via codon-dependent ribosome stalling:

- **Table S8**: Thor-Ribo-seq — DHX29-IP vs Input codon enrichment scores (CES) at the
  A-site. One row per sense codon; `Fold_Change` = IP_Avg / Input_Avg.
- **Table S6**: SLAM-seq with DHX29 KO — precomputed `Ctrl_Half_life`, `KO_Half_life`, and
  `logFC` (confirmed = log2(KO/Ctrl); spot-check: row 2 Ctrl=4.348, KO=5.364, logFC=0.303,
  log2(5.364/4.348) = 0.304 ✓). Positive logFC = mRNA stabilized upon DHX29 KO. No decay
  fitting needed — use precomputed values directly.

eIF3d-promoted genes (hypoxia 1hr) are overlaid on DHX29 occupancy and SLAM-seq distributions
to test whether 3d-promoted mRNAs are AU3-rich and DHX29-sensitive.

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
library(ggplot2)
library(ggrepel)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
```

---

## Section 1: Gene groups

```{r load_gene_groups}
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE)
promotes_ids <- promotes$ensembl_gene

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)

cat("Promotes genes:", length(promotes_ids), "\n")
cat("Neg ctrl genes (size-matched):", length(neg_ctrl_ids), "\n")
```

---

## Section 2: DHX29 Riboseq Occupancy (Thor-Ribo-seq)

### 2a: Per-gene DHX29 occupancy score

```{r load_s8}
s8_path <- here("accessories", "csc_data", "hia_dhx29",
                "science.adw0288_tables_s1_to_s11",
                "science.adw0288_table_s8.csv")

s8 <- read_csv(s8_path, show_col_types = FALSE) %>%
  mutate(codon = as.character(codon)) %>%
  filter(!codon %in% c("TAA", "TAG", "TGA"))   # keep only sense codons

cat("Sense codons in S8:", nrow(s8), "\n")
cat("Fold_Change range:", round(min(s8$Fold_Change), 3), "-",
    round(max(s8$Fold_Change), 3), "\n")
cat("\nTop 10 codons by CES:\n")
print(s8 %>% arrange(desc(Fold_Change)) %>% head(10) %>%
        dplyr::select(codon, Input_Avg, IP_Avg, Fold_Change))
```

```{r load_cds_for_occupancy}
# Load CDS sequences — same pattern as notebook 22 section 4b
most_abundant_tx <- readRDS(here("output", "predictive_modeling",
                                  "precomputed_most_abundant_tx.rds"))
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

keep_tx <- as.character(unlist(most_abundant_tx$transcript_id_clean))

cds <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_sequences_all <- extractTranscriptSeqs(genome, cds)
cds_names_clean   <- sub("\\..*", "", names(cds_sequences_all))
names(cds_sequences_all) <- cds_names_clean

cds_sequences <- cds_sequences_all[cds_names_clean %in% keep_tx]
valid_length  <- width(cds_sequences) %% 3 == 0
cds_sequences <- cds_sequences[valid_length]
cds_char      <- as.character(cds_sequences)

cat("CDS sequences for scoring:", length(cds_sequences), "\n")
```

```{r compute_dhx29_occupancy}
stop_codons <- c("TAA", "TAG", "TGA")
ces_vec     <- setNames(s8$Fold_Change, s8$codon)

# Per-gene DHX29 occupancy = weighted mean of CES values, weighted by codon frequency
# = sum(CES[c] × count_c) / n_codons  (equivalent to paper's sum(CES × freq) / CDS_length)
compute_occupancy <- function(seq_char) {
  n_codons <- nchar(seq_char) / 3
  if (n_codons < 10) return(NA_real_)
  starts <- seq(1, nchar(seq_char) - 2, by = 3)
  codons <- substring(seq_char, starts, starts + 2)
  codons <- codons[!codons %in% stop_codons]
  sum(ces_vec[codons], na.rm = TRUE) / n_codons
}

occupancy_scores <- vapply(cds_char, compute_occupancy, numeric(1))
names(occupancy_scores) <- names(cds_sequences)

# Map transcript → gene
tx_gene <- most_abundant_tx %>%
  transmute(
    transcript_id_clean = as.character(unlist(transcript_id_clean)),
    ensembl_gene        = as.character(unlist(gene_id_clean)),
    symbol              = as.character(unlist(symbol))
  )

dhx29_occupancy <- tibble(
  transcript_id_clean  = names(occupancy_scores),
  dhx29_occupancy_score = occupancy_scores
) %>%
  left_join(tx_gene, by = "transcript_id_clean") %>%
  filter(!is.na(ensembl_gene), !is.na(dhx29_occupancy_score))

cat("Genes with DHX29 occupancy score:", nrow(dhx29_occupancy), "\n")
cat("Summary:\n")
print(summary(dhx29_occupancy$dhx29_occupancy_score))
```

```{r dhx29_occupancy_annotate}
dhx29_occupancy <- dhx29_occupancy %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                            ~ "All other genes"
  ))

cat("Group counts:\n")
print(dhx29_occupancy %>% count(group))
```

```{r dhx29_occupancy_density_plot}
set_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "black"
)

bg_occ <- dhx29_occupancy %>% filter(group == "All other genes")
fg_occ <- dhx29_occupancy %>% filter(group != "All other genes")

median_occ <- fg_occ %>%
  group_by(group) %>%
  summarise(med = median(dhx29_occupancy_score, na.rm = TRUE), .groups = "drop")

p_occ <- ggplot() +
  geom_density(
    data = bg_occ,
    aes(x = dhx29_occupancy_score),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  geom_density(
    data = fg_occ %>% filter(group == "Negative control"),
    aes(x = dhx29_occupancy_score, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_density(
    data = fg_occ %>% filter(group == "eIF3d hypoxia promotes"),
    aes(x = dhx29_occupancy_score, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_vline(
    data = median_occ,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values  = set_colors) +
  labs(
    title    = "DHX29 ribosome occupancy: eIF3d-promoted genes (hypoxia 1hr)",
    subtitle = "Occupancy = weighted mean CES across CDS codons (S8 Fold_Change weights)\nGrey = all other genes; dashed = median",
    x = "DHX29 occupancy score", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_occ
ggsave(here("plots", "23_dhx29_riboseq_occupancy_by_group.pdf"),
       p_occ, width = 6, height = 5)
cat("Saved: plots/23_dhx29_riboseq_occupancy_by_group.pdf\n")
```

```{r dhx29_occupancy_wilcoxon}
promotes_occ <- dhx29_occupancy %>% filter(group == "eIF3d hypoxia promotes") %>%
  pull(dhx29_occupancy_score)
neg_occ      <- dhx29_occupancy %>% filter(group == "Negative control") %>%
  pull(dhx29_occupancy_score)

wt_occ <- wilcox.test(promotes_occ, neg_occ, alternative = "two.sided")
cat("Wilcoxon promotes vs neg ctrl (occupancy):\n")
cat("  W =", wt_occ$statistic, ", p =", signif(wt_occ$p.value, 3), "\n")
cat("  Median promotes:", round(median(promotes_occ, na.rm = TRUE), 4), "\n")
cat("  Median neg ctrl:", round(median(neg_occ, na.rm = TRUE), 4), "\n")
```

```{r dhx29_top_genes_check}
# Top 10 genes by DHX29 occupancy — expect AU3-rich, low-GC3 genes.
# TASOR is a known DHX29 target highlighted in the paper; appearing near the top is a good sign.
top10_occ <- dhx29_occupancy %>%
  arrange(desc(dhx29_occupancy_score)) %>%
  head(10) %>%
  dplyr::select(symbol, ensembl_gene, dhx29_occupancy_score, group)

cat("Top 10 genes by DHX29 occupancy score:\n")
print(top10_occ)

tasor_rank <- dhx29_occupancy %>%
  arrange(desc(dhx29_occupancy_score)) %>%
  mutate(rank = row_number()) %>%
  filter(str_to_upper(symbol) == "TASOR")

if (nrow(tasor_rank) > 0) {
  cat("\nTASOR rank:", tasor_rank$rank, "of", nrow(dhx29_occupancy),
      "(score:", round(tasor_rank$dhx29_occupancy_score, 4), ")\n")
} else {
  cat("\nTASOR not found in occupancy table",
      "(may lack a matched most-abundant transcript).\n")
}
```

### 2b: Validation — reproduce Figure 4D

Dot plot of all 61 sense codons ranked by CES, colored by AU3 (ending in A/T) vs GC3
(ending in G/C). AU3 codons should dominate the top ranks.

```{r dhx29_codon_ces_plot}
s8_plot <- s8 %>%
  filter(!codon %in% c("TAA", "TAG", "TGA")) %>%  # explicit guard; s8 was filtered at load but re-apply in case chunk was cached
  mutate(
    third_base  = str_sub(codon, 3, 3),
    codon_class = case_when(
      third_base %in% c("A", "T") ~ "AU3",
      third_base %in% c("G", "C") ~ "GC3"
    )
  ) %>%
  arrange(desc(Fold_Change)) %>%
  mutate(rank = row_number())

cat("AU3 codons (n =", sum(s8_plot$codon_class == "AU3"), "):",
    paste(s8_plot$codon[s8_plot$codon_class == "AU3"], collapse = ", "), "\n")
cat("GC3 codons (n =", sum(s8_plot$codon_class == "GC3"), "):",
    paste(s8_plot$codon[s8_plot$codon_class == "GC3"], collapse = ", "), "\n")

ces_colors <- c("AU3" = "#E69F00", "GC3" = "#56B4E9")

p_ces <- ggplot(s8_plot, aes(x = rank, y = Fold_Change, color = codon_class)) +
  geom_hline(yintercept = 1, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(
    aes(label = codon),
    size = 2.8, show.legend = FALSE,
    max.overlaps = 30, box.padding = 0.3
  ) +
  scale_color_manual(values = ces_colors, name = NULL) +
  labs(
    title    = "DHX29 codon enrichment score (CES) by A-site codon",
    subtitle = "Fold_Change = IP_Avg / Input_Avg (Thor-Ribo-seq); reproduces Figure 4D\nAU3 = ends in A/T; GC3 = ends in G/C",
    x = "Rank (1 = highest enrichment)", y = "CES (IP / Input)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_ces
ggsave(here("plots", "23_dhx29_codon_ces_by_class.pdf"),
       p_ces, width = 7, height = 5)
cat("Saved: plots/23_dhx29_codon_ces_by_class.pdf\n")

cat("\nTop 10 codons by CES:\n")
print(s8_plot %>% head(10) %>% dplyr::select(codon, codon_class, Fold_Change, rank))
cat("\nBottom 10 codons by CES:\n")
print(s8_plot %>% tail(10) %>% dplyr::select(codon, codon_class, Fold_Change, rank))
```

---

## Section 3: SLAM-seq Analysis

```{r load_s6}
s6_path <- here("accessories", "csc_data", "hia_dhx29",
                "science.adw0288_tables_s1_to_s11",
                "science.adw0288_table_s6.xlsx")

s6_raw <- read_xlsx(s6_path, sheet = "TableS6") %>%
  mutate(
    gene_name      = as.character(gene_name),
    Ctrl_Half_life = as.numeric(Ctrl_Half_life),
    KO_Half_life   = as.numeric(KO_Half_life),
    logFC          = as.numeric(logFC)
  )

cat("S6 rows:", nrow(s6_raw), "\n")
cat("Columns:", paste(colnames(s6_raw), collapse = ", "), "\n")

# logFC spot-check (confirmed log2 pre-run)
spot <- s6_raw %>%
  filter(!is.na(Ctrl_Half_life), !is.na(KO_Half_life)) %>%
  head(3) %>%
  mutate(log2_computed = round(log2(KO_Half_life / Ctrl_Half_life), 3))
cat("\nlogFC verification (logFC = log2(KO/Ctrl)):\n")
print(spot %>% dplyr::select(gene_name, Ctrl_Half_life, KO_Half_life, logFC, log2_computed))
```

```{r s6_filter}
# Filter: remove rows with missing or zero or >100h half-lives
n_start <- nrow(s6_raw)

s6_filt <- s6_raw %>%
  filter(
    !is.na(Ctrl_Half_life), !is.na(KO_Half_life), !is.na(logFC),
    Ctrl_Half_life > 0,  KO_Half_life > 0,
    Ctrl_Half_life <= 100, KO_Half_life <= 100
  )

cat("Rows before filtering:", n_start, "\n")
cat("Rows after  filtering:", nrow(s6_filt),
    "(removed:", n_start - nrow(s6_filt), ")\n")
```

```{r s6_transcript_mapping}
# Extract versionless ENST ID from QuantSeq transcript name
# Format: ENST00000000412.8_utr3_6_0_chr12_8940361_r  → ENST00000000412
s6_mapped <- s6_filt %>%
  mutate(transcript_id_clean = str_extract(gene_name, "ENST\\d+")) %>%
  filter(!is.na(transcript_id_clean))

# Join to most_abundant_tx to get gene ID
s6_gene <- s6_mapped %>%
  inner_join(
    tx_gene %>% dplyr::select(transcript_id_clean, ensembl_gene, symbol),
    by = "transcript_id_clean"
  )

n_matched  <- nrow(s6_gene)
pct_matched <- round(100 * n_matched / nrow(s6_filt), 1)
cat("Transcripts with ENST match in most_abundant_tx:",
    n_matched, "/", nrow(s6_filt), "(", pct_matched, "%)\n")

if (pct_matched < 50) {
  cat("WARNING: fewer than 50% of S6 transcripts mapped.\n",
      "QuantSeq 3'UTR data was built from a different transcript annotation than\n",
      "most_abundant_tx — low coverage is expected. Downstream analyses will\n",
      "reflect only the subset of genes with matched half-life data.\n")
}

# Where multiple transcripts map to the same gene, take mean half-lives and logFC
halflife_wide <- s6_gene %>%
  group_by(ensembl_gene, symbol) %>%
  summarise(
    Ctrl_Half_life = mean(Ctrl_Half_life, na.rm = TRUE),
    KO_Half_life   = mean(KO_Half_life,   na.rm = TRUE),
    logFC          = mean(logFC,           na.rm = TRUE),
    n_transcripts  = n(),
    .groups = "drop"
  )

cat("Unique genes after deduplication:", nrow(halflife_wide), "\n")
cat("Genes with >1 transcript:", sum(halflife_wide$n_transcripts > 1), "\n")
```

```{r s6_annotate_groups}
halflife_wide <- halflife_wide %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                            ~ "All other genes"
  ))

cat("Gene group counts (SLAM-seq):\n")
print(halflife_wide %>% count(group))
```

### Headline: log2(KO/Ctrl) half-life ratio by group

Positive logFC = mRNA stabilized when DHX29 is knocked out = DHX29 normally promotes its decay.

```{r dhx29_logfc_density_plot}
bg_fc <- halflife_wide %>% filter(group == "All other genes")
fg_fc <- halflife_wide %>% filter(group != "All other genes")

median_fc <- fg_fc %>%
  group_by(group) %>%
  summarise(med = median(logFC, na.rm = TRUE), .groups = "drop")

p_fc <- ggplot() +
  geom_density(
    data = bg_fc,
    aes(x = logFC),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  geom_density(
    data = fg_fc %>% filter(group == "Negative control"),
    aes(x = logFC, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_density(
    data = fg_fc %>% filter(group == "eIF3d hypoxia promotes"),
    aes(x = logFC, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_vline(
    data = median_fc,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values  = set_colors) +
  labs(
    title    = "mRNA stabilization upon DHX29 KO: eIF3d-promoted genes (hypoxia 1hr)",
    subtitle = "logFC = log2(KO / Ctrl) half-life; positive = stabilized upon KO\nGrey = all other genes; dashed = median",
    x = "log2(KO / Ctrl) half-life ratio", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_fc
ggsave(here("plots", "23_dhx29_slamseq_log2ratio_by_group.pdf"),
       p_fc, width = 6, height = 5)
cat("Saved: plots/23_dhx29_slamseq_log2ratio_by_group.pdf\n")
```

```{r dhx29_logfc_wilcoxon}
promotes_fc <- halflife_wide %>% filter(group == "eIF3d hypoxia promotes") %>% pull(logFC)
neg_fc      <- halflife_wide %>% filter(group == "Negative control")       %>% pull(logFC)

wt_fc <- wilcox.test(promotes_fc, neg_fc, alternative = "two.sided")
cat("Wilcoxon logFC (promotes vs neg ctrl):\n")
cat("  W =", wt_fc$statistic, ", p =", signif(wt_fc$p.value, 3), "\n")
cat("  Median promotes:", round(median(promotes_fc, na.rm = TRUE), 3), "\n")
cat("  Median neg ctrl:", round(median(neg_fc, na.rm = TRUE), 3), "\n")
```

### Baseline Ctrl half-life by group

```{r dhx29_ctrl_halflife_density_plot}
bg_hl <- halflife_wide %>% filter(group == "All other genes")
fg_hl <- halflife_wide %>% filter(group != "All other genes")

median_hl <- fg_hl %>%
  group_by(group) %>%
  summarise(med = median(Ctrl_Half_life, na.rm = TRUE), .groups = "drop")

p_hl <- ggplot() +
  geom_density(
    data = bg_hl,
    aes(x = Ctrl_Half_life),
    color = NA, fill = "grey80", alpha = 0.6
  ) +
  geom_density(
    data = fg_hl %>% filter(group == "Negative control"),
    aes(x = Ctrl_Half_life, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_density(
    data = fg_hl %>% filter(group == "eIF3d hypoxia promotes"),
    aes(x = Ctrl_Half_life, color = group, fill = group),
    alpha = 0.25, linewidth = 0.8
  ) +
  geom_vline(
    data = median_hl,
    aes(xintercept = med, color = group),
    linetype = "dashed", linewidth = 0.6
  ) +
  scale_color_manual(values = set_colors) +
  scale_fill_manual(values  = set_colors) +
  coord_cartesian(xlim = c(1, 6.5)) +
  labs(
    title    = "Baseline mRNA half-life (Ctrl): eIF3d-promoted genes (hypoxia 1hr)",
    subtitle = "Ctrl half-life from DHX29 KO SLAM-seq (Hia et al.)\nGrey = all other genes; dashed = median; x-axis clipped at 45h",
    x = "Ctrl mRNA half-life (hours)", y = "Density",
    color = NULL, fill = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_hl
ggsave(here("plots", "23_dhx29_slamseq_halflife_ctrl_by_group.pdf"),
       p_hl, width = 6, height = 5)
cat("Saved: plots/23_dhx29_slamseq_halflife_ctrl_by_group.pdf\n")
```

```{r dhx29_ctrl_halflife_wilcoxon}
promotes_hl <- halflife_wide %>% filter(group == "eIF3d hypoxia promotes") %>%
  pull(Ctrl_Half_life)
neg_hl      <- halflife_wide %>% filter(group == "Negative control") %>%
  pull(Ctrl_Half_life)

wt_hl <- wilcox.test(promotes_hl, neg_hl, alternative = "two.sided")
cat("Wilcoxon Ctrl halflife (promotes vs neg ctrl):\n")
cat("  W =", wt_hl$statistic, ", p =", signif(wt_hl$p.value, 3), "\n")
cat("  Median promotes:", round(median(promotes_hl, na.rm = TRUE), 2), "h\n")
cat("  Median neg ctrl:", round(median(neg_hl, na.rm = TRUE), 2), "h\n")
```

---

## Section 4: Validation

### 4a: GC3 content positive control — reproduce Figure 1H

Per-gene GC3 content should negatively predict logFC: low GC3 (AU3-rich) genes should be
most stabilized upon DHX29 KO. Using paper's exact bin boundaries for direct comparison.

```{r compute_gc3}
compute_gc3 <- function(seq_char) {
  n_codons <- nchar(seq_char) / 3
  if (n_codons < 10) return(NA_real_)
  starts <- seq(1, nchar(seq_char) - 2, by = 3)
  codons <- substring(seq_char, starts, starts + 2)
  codons <- codons[!codons %in% stop_codons]
  third  <- str_sub(codons, 3, 3)
  mean(third %in% c("G", "C"), na.rm = TRUE)
}

gc3_scores <- vapply(cds_char, compute_gc3, numeric(1))
names(gc3_scores) <- names(cds_sequences)

gc3_df <- tibble(
  transcript_id_clean = names(gc3_scores),
  gc3                 = gc3_scores
) %>%
  left_join(tx_gene %>% dplyr::select(transcript_id_clean, ensembl_gene),
            by = "transcript_id_clean") %>%
  filter(!is.na(ensembl_gene), !is.na(gc3))

cat("Genes with GC3 score:", nrow(gc3_df), "\n")
cat("GC3 summary:\n")
print(summary(gc3_df$gc3))
```

```{r gc3_validation_plot}
# Join GC3 to SLAM-seq logFC
gc3_logfc <- gc3_df %>%
  inner_join(halflife_wide %>% dplyr::select(ensembl_gene, logFC),
             by = "ensembl_gene") %>%
  filter(!is.na(logFC))

cat("Genes for GC3 validation:", nrow(gc3_logfc), "\n")

# Paper's bin boundaries
gc3_breaks <- c(0, 0.40, 0.50, 0.60, 0.70, 0.80, 1.0)
gc3_labels <- c("<40%", "40-50%", "50-60%", "60-70%", "70-80%", ">80%")

gc3_binned <- gc3_logfc %>%
  mutate(gc3_bin = cut(gc3, breaks = gc3_breaks, labels = gc3_labels,
                       include.lowest = TRUE, right = FALSE))

cat("\nGenes per GC3 bin:\n")
print(gc3_binned %>% count(gc3_bin))

gc3_medians <- gc3_binned %>%
  group_by(gc3_bin) %>%
  summarise(median_logFC = median(logFC, na.rm = TRUE),
            n = n(), .groups = "drop")

p_gc3 <- ggplot(gc3_binned, aes(x = gc3_bin, y = logFC)) +
  geom_boxplot(fill = "grey80", outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_text(data = gc3_medians,
            aes(x = gc3_bin, y = -Inf, label = paste0("n=", n)),
            vjust = -0.5, size = 3, color = "grey40") +
  labs(
    title    = "DHX29 KO mRNA stabilization by GC3 content",
    subtitle = "Reproduces Figure 1H; low GC3 (AU3-rich) genes expected most stabilized",
    x = "GC3 content bin", y = "log2(KO / Ctrl) half-life"
  ) +
  theme_bw(base_size = 12)

p_gc3
ggsave(here("plots", "23_dhx29_gc3_halflife_validation.pdf"),
       p_gc3, width = 6, height = 5)
cat("Saved: plots/23_dhx29_gc3_halflife_validation.pdf\n")
```

### 4b: CDF plot — top/bottom DHX29 occupancy vs all other genes

```{r dhx29_occupancy_cdf}
# Join occupancy scores with logFC
occ_logfc <- dhx29_occupancy %>%
  dplyr::select(ensembl_gene, dhx29_occupancy_score) %>%
  inner_join(halflife_wide %>% dplyr::select(ensembl_gene, logFC),
             by = "ensembl_gene") %>%
  filter(!is.na(logFC), !is.na(dhx29_occupancy_score))

cat("Genes for CDF plot:", nrow(occ_logfc), "\n")

n_group <- min(1000, floor(nrow(occ_logfc) / 3))

top_dhx29    <- occ_logfc %>% slice_max(dhx29_occupancy_score, n = n_group) %>%
  pull(ensembl_gene)
bottom_dhx29 <- occ_logfc %>% slice_min(dhx29_occupancy_score, n = n_group) %>%
  pull(ensembl_gene)

cat("Top DHX29 occupancy (AU3-rich):", length(top_dhx29), "genes\n")
cat("Bottom DHX29 occupancy (GC3-rich):", length(bottom_dhx29), "genes\n")

cdf_df <- occ_logfc %>%
  mutate(occ_group = case_when(
    ensembl_gene %in% top_dhx29    ~ paste0("Top ", n_group, ": high DHX29\n(AU3-rich)"),
    ensembl_gene %in% bottom_dhx29 ~ paste0("Bottom ", n_group, ": low DHX29\n(GC3-rich)"),
    TRUE                            ~ "All other genes"
  ))

cdf_colors <- c(
  "All other genes" = "grey60"
)
cdf_colors[paste0("Top ", n_group, ": high DHX29\n(AU3-rich)")]    <- "#E69F00"
cdf_colors[paste0("Bottom ", n_group, ": low DHX29\n(GC3-rich)")] <- "#56B4E9"

p_cdf <- ggplot(cdf_df, aes(x = logFC, color = occ_group)) +
  stat_ecdf(linewidth = 0.9) +
  scale_color_manual(values = cdf_colors) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  labs(
    title    = "mRNA stabilization upon DHX29 KO by occupancy score",
    subtitle = paste0("Top vs bottom ", n_group, " genes by DHX29 occupancy score\n",
                      "Rightward shift = stabilized upon KO"),
    x = "log2(KO / Ctrl) half-life ratio", y = "Empirical CDF",
    color = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_cdf
ggsave(here("plots", "23_dhx29_occupancy_cdf_log2ratio.pdf"),
       p_cdf, width = 6.5, height = 5)
cat("Saved: plots/23_dhx29_occupancy_cdf_log2ratio.pdf\n")

# Wilcoxon tests
top_fc    <- cdf_df %>% filter(str_detect(occ_group, "Top"))    %>% pull(logFC)
bot_fc    <- cdf_df %>% filter(str_detect(occ_group, "Bottom")) %>% pull(logFC)
other_fc  <- cdf_df %>% filter(occ_group == "All other genes")  %>% pull(logFC)

wt_tb <- wilcox.test(top_fc, bot_fc,   alternative = "two.sided")
wt_to <- wilcox.test(top_fc, other_fc, alternative = "two.sided")

cat("\nWilcoxon top vs bottom: W =", wt_tb$statistic,
    ", p =", signif(wt_tb$p.value, 3), "\n")
cat("  Median top:", round(median(top_fc, na.rm = TRUE), 3), "\n")
cat("  Median bottom:", round(median(bot_fc, na.rm = TRUE), 3), "\n")
cat("\nWilcoxon top vs all other: W =", wt_to$statistic,
    ", p =", signif(wt_to$p.value, 3), "\n")
```

### 4c: DHX29-dependent codon stability score

Spearman correlation between per-codon CDS frequency and logFC across genes. AU3 codons
should rank at the top (positive correlation), GC3 at the bottom.

```{r dhx29_codon_freq_long}
# Build per-gene codon frequency table — reuses cds_char from section 2
codon_freq_long_dhx <- tibble(
  transcript_id_clean = names(cds_sequences),
  seq                 = cds_char
) %>%
  left_join(dhx29_occupancy %>% dplyr::select(transcript_id_clean, ensembl_gene),
            by = "transcript_id_clean") %>%
  filter(!is.na(ensembl_gene)) %>%
  mutate(
    n_codons = nchar(seq) / 3,
    codons   = map2(seq, n_codons, function(s, n) {
      starts <- seq(1, nchar(s) - 2, by = 3)
      substring(s, starts, starts + 2)
    })
  ) %>%
  dplyr::select(-seq) %>%
  unnest(codons) %>%
  filter(!codons %in% stop_codons) %>%
  count(ensembl_gene, n_codons, codons, name = "codon_count") %>%
  mutate(freq = codon_count / n_codons) %>%
  dplyr::select(ensembl_gene, codons, freq)

# Note: CDS sequences are in DNA notation — CGU (RNA) appears as CGT here.
cat("Codon-gene frequency rows:", nrow(codon_freq_long_dhx), "\n")
cat("Unique sense codons observed:", n_distinct(codon_freq_long_dhx$codons), "\n")
```

```{r dhx29_codon_spearman}
codon_spearman_dhx <- codon_freq_long_dhx %>%
  left_join(halflife_wide %>% dplyr::select(ensembl_gene, logFC),
            by = "ensembl_gene") %>%
  filter(!is.na(logFC)) %>%
  group_by(codons) %>%
  summarise(
    spearman_r = cor(freq, logFC, method = "spearman", use = "complete.obs"),
    n_genes    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(spearman_r)) %>%
  mutate(rank = row_number())

cat("Codons scored:", nrow(codon_spearman_dhx), "\n\n")

# Report ranks of key codons
cat("Representative AU3 codon ranks:\n")
print(codon_spearman_dhx %>% filter(codons %in% c("AAA","AAT","ATA","AGA","TAT","TAA")) %>%
        filter(!codons %in% stop_codons) %>%
        dplyr::select(codons, rank, spearman_r))

cat("\nRepresentative GC3 codon ranks:\n")
print(codon_spearman_dhx %>% filter(codons %in% c("GGC","GCC","CGC","GGG","CCG")) %>%
        dplyr::select(codons, rank, spearman_r))
```

```{r dhx29_codon_stability_plot}
# Codon → amino acid + AU3/GC3 annotation
codon_aa_dhx <- tibble(
  rna_codon = names(GENETIC_CODE),
  aa        = as.character(GENETIC_CODE)
) %>%
  mutate(codon = str_replace_all(rna_codon, "U", "T")) %>%
  filter(!aa %in% c("*")) %>%
  dplyr::select(codon, aa)

codon_plot_dhx <- codon_spearman_dhx %>%
  left_join(codon_aa_dhx, by = c("codons" = "codon")) %>%
  mutate(
    third_base  = str_sub(codons, 3, 3),
    codon_class = case_when(
      third_base %in% c("A", "T") ~ "AU3",
      third_base %in% c("G", "C") ~ "GC3"
    ),
    codon_class = factor(codon_class, levels = c("AU3", "GC3"))
  )

class_colors_dhx <- c("AU3" = "#E69F00", "GC3" = "#56B4E9")

p_codon_dhx <- ggplot(codon_plot_dhx, aes(x = rank, y = spearman_r, color = codon_class)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(
    aes(label = codons),
    size = 2.8, fontface = "bold", show.legend = FALSE,
    max.overlaps = 40, box.padding = 0.3
  ) +
  scale_color_manual(values = class_colors_dhx, name = NULL) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 61)) +
  labs(
    title    = "DHX29-dependent codon stability score",
    subtitle = "Spearman r(codon frequency, log2 KO/Ctrl half-life) across genes\nAU3 = ends in A/T (expect top); GC3 = ends in G/C (expect bottom)",
    x = "Rank (1 = most positive)", y = "Spearman r"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")

p_codon_dhx
ggsave(here("plots", "23_dhx29_codon_stability_scores.pdf"),
       p_codon_dhx, width = 7, height = 5)
cat("Saved: plots/23_dhx29_codon_stability_scores.pdf\n")
```

---

## Section 5: Save outputs

```{r save_outputs}
saveRDS(dhx29_occupancy,    here("output", "predictive_modeling", "dhx29_riboseq_occupancy.rds"))
saveRDS(halflife_wide,      here("output", "predictive_modeling", "dhx29_slamseq_halflives.rds"))
saveRDS(codon_spearman_dhx, here("output", "predictive_modeling", "dhx29_codon_stability_scores.rds"))

cat("Saved:\n")
cat("  output/predictive_modeling/dhx29_riboseq_occupancy.rds\n")
cat("  output/predictive_modeling/dhx29_slamseq_halflives.rds\n")
cat("  output/predictive_modeling/dhx29_codon_stability_scores.rds\n")
```

---

## Section 6: Add DHX29 features to feature matrix

```{r build_dhx29_feature_table}
dhx29_features <- dhx29_occupancy %>%
  dplyr::select(ensembl_gene, dhx29_occupancy_score) %>%
  full_join(
    halflife_wide %>%
      dplyr::select(ensembl_gene,
                    dhx29_slamseq_ctrl_halflife = Ctrl_Half_life,
                    dhx29_slamseq_logFC         = logFC),
    by = "ensembl_gene"
  )

cat("DHX29 feature table:", nrow(dhx29_features), "genes\n")
cat("Genes with occupancy score:", sum(!is.na(dhx29_features$dhx29_occupancy_score)), "\n")
cat("Genes with SLAM-seq data:  ", sum(!is.na(dhx29_features$dhx29_slamseq_logFC)), "\n")
```

```{r join_dhx29_to_feature_matrix}
# Use most recent feature matrix: prefer cnot3 (nb22 output), fall back to kozak
fm_path <- if (file.exists(here("output", "predictive_modeling",
                                 "feature_matrix_cnot3.rds")))
  here("output", "predictive_modeling", "feature_matrix_cnot3.rds") else
  here("output", "predictive_modeling", "feature_matrix_kozak.rds")

cat("Loading feature matrix from:", basename(fm_path), "\n")
fm <- readRDS(fm_path)
cat("Feature matrix:", nrow(fm), "genes ×", ncol(fm), "columns\n")

fm_dhx29 <- fm %>%
  left_join(dhx29_features, by = c("gene_id_clean" = "ensembl_gene"))

cat("After join:", nrow(fm_dhx29), "genes ×", ncol(fm_dhx29), "columns\n")
cat("New columns added:", ncol(fm_dhx29) - ncol(fm), "\n")

cat("\nCoverage within feature matrix genes:\n")
cat("  dhx29_occupancy_score:      ",
    sum(!is.na(fm_dhx29$dhx29_occupancy_score)), "/", nrow(fm_dhx29), "\n")
cat("  dhx29_slamseq_ctrl_halflife:",
    sum(!is.na(fm_dhx29$dhx29_slamseq_ctrl_halflife)), "/", nrow(fm_dhx29), "\n")
cat("  dhx29_slamseq_logFC:        ",
    sum(!is.na(fm_dhx29$dhx29_slamseq_logFC)), "/", nrow(fm_dhx29), "\n")

saveRDS(fm_dhx29, here("output", "predictive_modeling", "feature_matrix_dhx29.rds"))
cat("\nSaved: output/predictive_modeling/feature_matrix_dhx29.rds\n")
```
