---
title: "Notebook 24: Post-2026 Thesis Committee Analyses"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(ggplot2)
library(ggvenn)
```

## Section 1: Load Data

Spike-in normalized DESeq2 results comparing sictrl vs si3d.
Positive LFC = higher TE in sictrl (control); negative LFC = higher TE in si3d knockdown.

```{r load-data}
normoxia <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
hypoxia_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## Section 2: TE LFC Normoxia (1+4hr regressed) vs TE LFC 1hr Hypoxia

```{r join-data}
plot_df <- normoxia %>%
  dplyr::select(gene_id, symbol, te_lfc_normoxia = te_lfc, te_padj_normoxia = te_padj) %>%
  inner_join(
    hypoxia_1hr %>% 
      dplyr::select(gene_id, te_lfc_hypoxia = te_lfc, te_padj_hypoxia = te_padj),
    by = "gene_id"
  ) %>%
  mutate(sig = case_when(
    te_padj_normoxia < 0.05 & te_padj_hypoxia < 0.05 ~ "both",
    te_padj_normoxia < 0.05                           ~ "normoxia only",
    te_padj_hypoxia  < 0.05                           ~ "hypoxia only",
    TRUE                                               ~ "neither"
  ))

sig_colors <- c("both"          = "#8B0000",
                "normoxia only" = "#2166ac",
                "hypoxia only"  = "#d6604d",
                "neither"       = "grey70")

cat("Genes in plot:", nrow(plot_df), "\n")
table(plot_df$sig)

# Pearson correlations for plot annotations
cor_all <- cor.test(plot_df$te_lfc_normoxia, plot_df$te_lfc_hypoxia)
plot_df_sig <- filter(plot_df, sig != "neither")
cor_sig <- cor.test(plot_df_sig$te_lfc_normoxia, plot_df_sig$te_lfc_hypoxia)

fmt_cor <- function(ct) sprintf("r = %.2f, p = %.2e", ct$estimate, ct$p.value)

# Fit regression and extract residuals
lm_fit <- lm(te_lfc_hypoxia ~ te_lfc_normoxia, data = plot_df, na.action = na.exclude)
plot_df <- plot_df %>%
  mutate(residual = residuals(lm_fit))
```

### Plot 1: All genes

```{r plot-all-genes, fig.width=6, fig.height=5}
p <- ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = sig)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6,
              aes(group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  scale_color_manual(values = sig_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = "Significant\n(padj < 0.05)",
    title = "All genes"
  ) +
  theme_classic()

p

ggsave(here("output/24_norm_v_hyp_changes.pdf"), height=4, width=4)

p +
  xlim(-1, 2.5) +
  ylim(-1, 2.5)

ggsave(here("output/all_genes_corr.pdf"), width=5, height=4)

```

### Plot 2: Significant in at least one condition

```{r plot-sig-genes, fig.width=6, fig.height=5}
ggplot(plot_df_sig, aes(te_lfc_normoxia, te_lfc_hypoxia, color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6,
              aes(group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  scale_color_manual(values = sig_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_sig), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = "Significant\n(padj < 0.05)",
    title = "Significant in \u22651 condition"
  ) +
  theme_classic() +
  xlim(-1, 2.5) +
  ylim(-1, 2.5)

ggsave(here("output/sig_genes_corr.pdf"))
```

### Plot 1b: All genes, colored by regression residual

```{r plot-residual, fig.width=6, fig.height=5}
resid_lim <- max(abs(plot_df$residual), na.rm = TRUE)

ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = residual)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#d6604d",
                        midpoint = 0, limits = c(-resid_lim, resid_lim),
                        name = "Residual") +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    title = "All genes \u2014 colored by residual"
  ) +
  theme_classic()
```

## Section 3: Venn Diagram — Normoxia vs Hypoxia Regulated Genes

```{r venn-normoxia-hypoxia, fig.width=5, fig.height=5}
venn_list <- list(
  "Normoxia (1+4hr)" = filter(plot_df, te_padj_normoxia < 0.05)$gene_id,
  "Hypoxia 1hr"      = filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id
)

p_venn <- ggvenn(venn_list,
                 fill_color  = c("#2166ac", "#d6604d"),
                 stroke_size = 0.5,
                 set_name_size = 4) +
  ggtitle("Significant TE changes (padj < 0.05, sictrl vs si3d)")

p_venn

ggsave(here("output/24_venn_normoxia_vs_hypoxia.pdf"), p_venn, width = 5, height = 5)
```

## Section 4: Are Normoxia-Regulated mRNAs a Subset of Hypoxia-Regulated mRNAs?

```{r subset-genesets}
# padj only
norm_sig_padj   <- filter(plot_df, te_padj_normoxia < 0.05)$gene_id
hyp_sig_padj    <- filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id

# padj + |LFC| > 0.5
norm_sig_strict <- filter(plot_df, te_padj_normoxia < 0.05, abs(te_lfc_normoxia) > 0.5)$gene_id
hyp_sig_strict  <- filter(plot_df, te_padj_hypoxia  < 0.05, abs(te_lfc_hypoxia)  > 0.5)$gene_id

universe <- plot_df$gene_id

fisher_overlap <- function(set_a, set_b, universe, label = "") {
  a_and_b <- length(intersect(set_a, set_b))
  a_not_b <- length(setdiff(set_a, set_b))
  b_not_a <- length(setdiff(set_b, set_a))
  neither <- length(universe) - length(union(set_a, set_b))
  m  <- matrix(c(a_and_b, a_not_b, b_not_a, neither), nrow = 2)
  ft <- fisher.test(m, alternative = "greater")
  cat(label, "\n")
  cat(sprintf("  Overlap: %d | OR = %.2f | p = %.2e\n", a_and_b, ft$estimate, ft$p.value))
  cat(sprintf("  %% of normoxia-sig also hypoxia-sig: %.1f%%\n\n",
              100 * a_and_b / length(set_a)))
  invisible(ft)
}

fisher_overlap(norm_sig_padj,   hyp_sig_padj,   universe, "padj < 0.05 only")
fisher_overlap(norm_sig_strict, hyp_sig_strict, universe, "padj < 0.05 & |LFC| > 0.5")
```

### Part 1: Bar plot with LFC + padj cutoff

Universe = normoxia-regulated genes; bars show how many are also hypoxia-regulated vs normoxia only.

```{r barplot-strict, fig.width=4, fig.height=4}
norm_and_hyp <- intersect(norm_sig_strict, hyp_sig_strict)
norm_only    <- setdiff(norm_sig_strict,   hyp_sig_strict)

bar_df <- tibble(
  category = c("Normoxia &\nhypoxia", "Normoxia\nonly"),
  n        = c(length(norm_and_hyp), length(norm_only))
)

ggplot(bar_df, aes(category, n, fill = category)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Normoxia &\nhypoxia" = "#8B0000",
                                "Normoxia\nonly"       = "#2166ac"),
                    guide = "none") +
  labs(x = NULL, y = "Number of genes",
       title = "Normoxia-regulated genes (padj < 0.05 & |LFC| > 0.5)") +
  theme_classic()

ggsave(here("output/24_barplot_subset_strict.pdf"), width = 4, height = 4)
```

### Part 2: Distribution of hypoxia LFC by normoxia significance (no cutoff)

```{r ecdf-normoxia-groups, fig.width=6, fig.height=5}
plot_df <- plot_df %>%
  mutate(norm_group = case_when(
    te_padj_normoxia < 0.05 & te_lfc_normoxia > 0 ~ "Normoxia up",
    te_padj_normoxia < 0.05 & te_lfc_normoxia < 0 ~ "Normoxia down",
    TRUE                                            ~ "Not sig"
  ))

group_colors <- c("Normoxia up"   = "#d6604d",
                  "Normoxia down" = "#2166ac",
                  "Not sig"       = "grey70")

ggplot(plot_df, aes(te_lfc_hypoxia, color = norm_group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = group_colors) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = "Cumulative fraction",
       color = "Normoxia status",
       title = "Hypoxia LFC distribution by normoxia regulation") +
  theme_classic()+
  xlim(-4,1)

ggsave(here("output/24_ecdf_hypoxia_by_normoxia_group.pdf"), width = 6, height = 4)

# Wilcoxon tests vs not-sig background
not_sig_hyp <- filter(plot_df, norm_group == "Not sig")$te_lfc_hypoxia

cat("Normoxia up vs not-sig (greater):\n")
print(wilcox.test(filter(plot_df, norm_group == "Normoxia up")$te_lfc_hypoxia,
                  not_sig_hyp, alternative = "greater"))

cat("\nNormoxia down vs not-sig (less):\n")
print(wilcox.test(filter(plot_df, norm_group == "Normoxia down")$te_lfc_hypoxia,
                  not_sig_hyp, alternative = "less"))
```

## Section 5: Volcano Plot — Hypoxia Changes Colored by Normoxia LFC

```{r volcano-hypoxia, fig.width=6, fig.height=5}
# Clamp color scale to 5th-95th percentile so most points get visible color
norm_lfc_lim <- quantile(abs(plot_df$te_lfc_normoxia), 0.95, na.rm = TRUE)

ggplot(plot_df, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                    color = te_lfc_normoxia)) +
  geom_point(alpha = 0.9, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_gradient2(low = "#053061", mid = "grey85", high = "#67001f",
                        midpoint = 0, limits = c(-norm_lfc_lim, norm_lfc_lim),
                        oob = scales::squish,
                        name = "Normoxia LFC\n(sictrl/si3d)") +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes \u2014 colored by normoxia LFC") +
  theme_classic()+
  ylim(0,8)

ggsave(here("output/24_volcano_hypoxia_colored_normoxia.pdf"), width = 8, height = 5)
```

```{r volcano-hypoxia-normsig, fig.width=6, fig.height=5}
plot_df_v2 <- plot_df %>%
  mutate(norm_sig = te_padj_normoxia < 0.05 & te_lfc_normoxia > 0)

ggplot(plot_df_v2, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                       color = norm_sig)) +
  geom_point(alpha = 0.7, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c("FALSE" = "grey80", "TRUE" = "#009E73"),
                     labels = c("FALSE" = "Not sig in normoxia",
                                "TRUE"  = "Sig and positive in normoxia (padj < 0.05)"),
                     name   = NULL) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes \u2014 normoxia-regulated highlighted") +
  theme_classic() +
  ylim(0, 8)

ggsave(here("output/24_volcano_hypoxia_normsig_highlighted.pdf"), width = 10, height = 5)
```

## Section 6: Venn Diagram — eIF3d vs eIF3e Normoxia Promoted mRNAs (1+4hr)

```{r venn-3d-vs-3e-normoxia, fig.width=5, fig.height=5}
si3d_norm <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

si3e_norm <- read_csv(here("output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

promoted_3d_norm <- filter(si3d_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
promoted_3e_norm <- filter(si3e_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

cat("Promoted by 3d normoxia:", length(promoted_3d_norm), "\n")
cat("Promoted by 3e normoxia:", length(promoted_3e_norm), "\n")
cat("Overlap:", length(intersect(promoted_3d_norm, promoted_3e_norm)), "\n")

venn_norm <- list(
  "eIF3d normoxia" = promoted_3d_norm,
  "eIF3e normoxia" = promoted_3e_norm
)

p_venn_norm <- ggvenn(venn_norm,
                      fill_color  = c("#E69F00", "#56B4E9"),
                      stroke_size = 0.5,
                      set_name_size = 4) +
  ggtitle("Promoted mRNAs: eIF3d vs eIF3e\n(normoxia 1+4hr, padj<0.05 & LFC>0.5)")

p_venn_norm
ggsave(here("output/24_venn_3d_vs_3e_normoxia.pdf"), p_venn_norm, width = 5, height = 5)
```

## Section 7: Venn Diagrams — eIF3d vs eIF3e by Hypoxic Timepoint

```{r venn-3d-vs-3e-hypoxia, fig.width=5, fig.height=5}
make_hyp_venn <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  set_d <- filter(d, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
  set_e <- filter(e, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

  cat("Hypoxia", timepoint, "— 3d:", length(set_d), "| 3e:", length(set_e),
      "| overlap:", length(intersect(set_d, set_e)), "\n")

  ggvenn(list("eIF3d" = set_d, "eIF3e" = set_e),
         fill_color  = c("#E69F00", "#56B4E9"),
         stroke_size = 0.5,
         set_name_size = 4) +
    ggtitle(paste0("Promoted mRNAs: eIF3d vs eIF3e\nhypoxia ", timepoint,
                   " (padj<0.05 & LFC>0.5)"))
}

p_venn_1hr  <- make_hyp_venn("1hr");  p_venn_1hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_1hr.pdf"),  p_venn_1hr,  width = 5, height = 5)

p_venn_4hr  <- make_hyp_venn("4hr");  p_venn_4hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_4hr.pdf"),  p_venn_4hr,  width = 5, height = 5)

p_venn_24hr <- make_hyp_venn("24hr"); p_venn_24hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_24hr.pdf"), p_venn_24hr, width = 5, height = 5)
```

## Section 8: Dot Plot — eIF3d vs eIF3e Normoxia TE LFC

```{r dotplot-3d-vs-3e-normoxia, fig.width=5.5, fig.height=5}
dot_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
  inner_join(
    si3e_norm %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
    by = "gene_id_clean"
  ) %>%
  mutate(sig = case_when(
    te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
    te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
    te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
    TRUE                                                                          ~ "neither"
  ))

cat("Genes in dot plot:", nrow(dot_df), "\n")
cat("Gene counts per sig category:\n")
print(table(dot_df$sig))

dot_colors <- c("both"    = "#8B0000",
                "3d only" = "#E69F00",
                "3e only" = "#56B4E9",
                "neither" = "grey70")

dot_bg <- filter(dot_df, sig == "neither")
dot_fg <- filter(dot_df, sig != "neither")

p_dot <- ggplot() +
  geom_point(data = dot_bg, aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.4, size = 0.7) +
  geom_point(data = dot_fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  labs(
    x        = "TE LFC normoxia \u2014 eIF3d (sictrl/si3d, 1+4hr)",
    y        = "TE LFC normoxia \u2014 eIF3e (sictrl/si3e, 1+4hr)",
    title    = "eIF3d vs eIF3e normoxia TE changes",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = c(-3, 5), ylim = c(-3, 5))

p_dot
ggsave(here("output/24_dotplot_3d_vs_3e_normoxia.pdf"), p_dot, width = 5.5, height = 5)
```

## Section 9: Dot Plots — eIF3d vs eIF3e by Hypoxic Timepoint

```{r dotplot-3d-vs-3e-hypoxia, fig.width=5.5, fig.height=5}
make_hyp_dotplot <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  df <- d %>%
    dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
    inner_join(
      e %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
      by = "gene_id_clean"
    ) %>%
    mutate(sig = case_when(
      te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
      te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
      te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
      TRUE                                                                          ~ "neither"
    ))

  cat("Hypoxia", timepoint, "— genes in plot:", nrow(df), "\n")
  print(table(df$sig))

  dot_colors <- c("both"    = "#8B0000",
                  "3d only" = "#E69F00",
                  "3e only" = "#56B4E9",
                  "neither" = "grey70")

  bg <- filter(df, sig == "neither")
  fg <- filter(df, sig != "neither")

  ggplot() +
    geom_point(data = bg, aes(x = te_lfc_3d, y = te_lfc_3e),
               color = "grey70", alpha = 0.4, size = 0.7) +
    geom_point(data = fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
               alpha = 0.7, size = 1) +
    geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = dot_colors, name = NULL) +
    labs(
      x        = paste0("TE LFC hypoxia ", timepoint, " \u2014 eIF3d (sictrl/si3d)"),
      y        = paste0("TE LFC hypoxia ", timepoint, " \u2014 eIF3e (sictrl/si3e)"),
      title    = paste0("eIF3d vs eIF3e hypoxia ", timepoint, " TE changes"),
      subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
    ) +
    theme_classic(base_size = 12) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5))
}

p_dot_1hr  <- make_hyp_dotplot("1hr");  p_dot_1hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_1hr.pdf"),  p_dot_1hr,  width = 5.5, height = 5)

p_dot_4hr  <- make_hyp_dotplot("4hr");  p_dot_4hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_4hr.pdf"),  p_dot_4hr,  width = 5.5, height = 5)

p_dot_24hr <- make_hyp_dotplot("24hr"); p_dot_24hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_24hr.pdf"), p_dot_24hr, width = 5.5, height = 5)
```

## Section 10: Condition-Specific Gene Sets for Random Forest

Genes promoted specifically in normoxia (1+4hr) OR hypoxia (1hr) by eIF3d.
"Specific" = padj < 0.05 & LFC > threshold in one condition AND NOT meeting threshold in the other.

### 10a: Threshold sweep

```{r condition_specific_threshold_sweep}
hyp1_clean <- hypoxia_1hr %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

gene_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, norm_lfc = te_lfc, norm_padj = te_padj) %>%
  full_join(
    hyp1_clean %>% dplyr::select(gene_id_clean, hyp_lfc = te_lfc, hyp_padj = te_padj),
    by = "gene_id_clean"
  )

lfc_thresholds <- c(0, 0.25, 0.5, 0.75, 1.0)

thresh_table <- map_dfr(lfc_thresholds, function(thresh) {
  norm_set <- filter(gene_df, norm_padj < 0.05, norm_lfc > thresh)$gene_id_clean
  hyp_set  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > thresh)$gene_id_clean
  tibble(
    lfc_thresh        = thresh,
    normoxia_promotes = length(norm_set),
    hypoxia_promotes  = length(hyp_set),
    both              = length(intersect(norm_set, hyp_set)),
    normoxia_specific = length(setdiff(norm_set, hyp_set)),
    hypoxia_specific  = length(setdiff(hyp_set,  norm_set))
  )
})

print(thresh_table)
```

### 10b: Save condition-specific gene sets

Adjust `lfc_thresh` after reviewing the table above, then save both CSVs to `output/genesets/`.
Files are named to match nb07's geneset_csv convention for single-timepoint params.

```{r save_condition_specific_genesets}
lfc_thresh <- 0.5  # adjust after reviewing thresh_table above

norm_promotes <- filter(gene_df, norm_padj < 0.05, norm_lfc > lfc_thresh)$gene_id_clean
hyp_promotes  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > lfc_thresh)$gene_id_clean
norm_specific <- setdiff(norm_promotes, hyp_promotes)
hyp_specific  <- setdiff(hyp_promotes,  norm_promotes)

cat("normoxia-specific:", length(norm_specific), "\n")
cat("hypoxia-specific:", length(hyp_specific), "\n")

# Symbol lookup from si3d_norm (genome-wide test result, has symbols for all genes)
sym_map <- si3d_norm %>%
  dplyr::select(gene_id_clean, symbol) %>%
  distinct()

save_gs <- function(ids, gs_name, gs_desc, fname) {
  df <- tibble(gene_id_clean = ids) %>%
    left_join(sym_map, by = "gene_id_clean") %>%
    transmute(
      db_gene_symbol    = coalesce(symbol, gene_id_clean),
      ensembl_gene      = gene_id_clean,
      gs_name           = gs_name,
      gs_description    = gs_desc,
      gs_source_species = "homo sapiens"
    )
  write_csv(df, here("output", "genesets", fname))
  cat("Saved:", fname, "—", nrow(df), "genes\n")
}

save_gs(
  norm_specific,
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to normoxia (not in hypoxia 1hr), padj<0.05, LFC>", lfc_thresh),
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh, ".csv")
)

save_gs(
  hyp_specific,
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to hypoxia 1hr (not in normoxia), padj<0.05, LFC>", lfc_thresh),
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh, ".csv")
)
```

**To run nb07 with these gene sets**, set params as:

- **Normoxia-specific**: `condition="normoxia"`, `timepoint="normoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`
- **Hypoxia-specific**: `condition="hypoxia"`, `timepoint="hypoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`

Negative control size-matching (set.seed(9)) is handled automatically in nb07.

```{r}

```

## Section 11: Faceted Dot Plot — eIF3d vs eIF3e Across All Timepoints

```{r dotplot-faceted-all-timepoints, fig.width=18, fig.height=5}
load_tp_df <- function(tp_label, d_path, e_path) {
  d <- read_csv(here(d_path), show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(e_path), show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  d %>%
    dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
    inner_join(
      e %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
      by = "gene_id_clean"
    ) %>%
    mutate(
      timepoint = tp_label,
      sig = case_when(
        te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
        te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
        te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
        TRUE                                                                          ~ "neither"
      )
    )
}

all_tp_df <- bind_rows(
  load_tp_df("Normoxia",
             "output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv",
             "output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
  load_tp_df("Hypoxia 1hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
  load_tp_df("Hypoxia 4hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"),
  load_tp_df("Hypoxia 24hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("Normoxia", "Hypoxia 1hr",
                                                   "Hypoxia 4hr", "Hypoxia 24hr")))

dot_colors <- c("both"    = "#8B0000",
                "3d only" = "#E69F00",
                "3e only" = "#56B4E9",
                "neither" = "grey70")

p_faceted <- ggplot() +
  geom_point(data = filter(all_tp_df, sig == "neither"),
             aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.3, size = 0.5) +
  geom_point(data = filter(all_tp_df, sig != "neither"),
             aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  facet_wrap(~ timepoint, nrow = 1) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
  labs(
    x        = "TE LFC eIF3d (sictrl/si3d)",
    y        = "TE LFC eIF3e (sictrl/si3e)",
    title    = "eIF3d vs eIF3e TE changes across timepoints",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 16))

p_faceted
ggsave(here("output/24_dotplot_3d_vs_3e_all_timepoints_faceted.pdf"),
       p_faceted, width = 18, height = 5)
```

## Section 12: Save 3d-specific and 3e-specific Gene Sets for nb07

Positive set for RF: promoted by eIF3d but NOT eIF3e at 1hr hypoxia.
Negative set for RF: promoted by eIF3e but NOT eIF3d at 1hr hypoxia.

```{r save-3d-vs-3e-specific-genesets}
hyp1_df <- all_tp_df %>% filter(timepoint == "Hypoxia 1hr")

# Symbol lookup from 1hr hypoxia si3d data
sym_map_1hr <- read_csv(
  here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  dplyr::select(gene_id_clean, symbol) %>%
  distinct()

save_specific_gs <- function(df, gs_name, gs_desc, fname) {
  out <- df %>%
    left_join(sym_map_1hr, by = "gene_id_clean") %>%
    transmute(
      db_gene_symbol    = coalesce(symbol, gene_id_clean),
      ensembl_gene      = gene_id_clean,
      gs_name           = gs_name,
      gs_description    = gs_desc,
      gs_source_species = "homo sapiens"
    )
  write_csv(out, here("output", "genesets", fname))
  cat("Saved:", fname, "—", nrow(out), "genes\n")
}

# 3d only: promoted by 3d, NOT 3e
gs_3d_only <- hyp1_df %>%
  filter(te_padj_3d < 0.05, te_lfc_3d > 0.5,
         !(te_padj_3e < 0.05 & te_lfc_3e > 0.5))

save_specific_gs(
  gs_3d_only,
  "hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5",
  "mRNAs promoted by eIF3d but not eIF3e at 1hr hypoxia (padj<0.05, LFC>0.5)",
  "hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5.csv"
)

# 3e only: promoted by 3e, NOT 3d
gs_3e_only <- hyp1_df %>%
  filter(te_padj_3e < 0.05, te_lfc_3e > 0.5,
         !(te_padj_3d < 0.05 & te_lfc_3d > 0.5))

save_specific_gs(
  gs_3e_only,
  "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5",
  "mRNAs promoted by eIF3e but not eIF3d at 1hr hypoxia (padj<0.05, LFC>0.5)",
  "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv"
)
```

**To run nb07 with these gene sets**, set params in the `derived_values` chunk as:

Note: after running nb07 for each analysis, re-run the comparison section in nb09 to regenerate the comparison plots.

```r
params <- list(
  condition               = "hypoxia",
  gene                    = "3d",
  direction               = "promotes",
  timepoint               = "1hr_not3e",
  lfc                     = "0.5",
  feature_set             = "external_stability",
  clip_mode               = "aggregate",
  exclude_4e_features     = FALSE,
  exclude_own_te_features = TRUE,
  neg_geneset_csv         = "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv"
)
```

## Section 14: GSEA — KEGG Pathways in 3d-specific vs 3e-specific Hypoxia Targets

Runs GSEA using KEGG pathways on the full TE LFC ranked lists from the si3d and si3e 1hr hypoxia
comparisons. Compares enrichment between the two to identify pathways preferentially regulated
by eIF3d vs eIF3e.

```{r gsea-kegg-setup}
library(fgsea)
library(msigdbr)
library(clusterProfiler)

# KEGG gene sets from MSigDB (TERM2GENE format for clusterProfiler::GSEA)
kegg_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY") %>%
  dplyr::select(gs_name, gene_symbol)
cat("KEGG pathways loaded:", length(unique(kegg_sets$gs_name)), "\n")

# Hallmark gene sets
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)
cat("Hallmark gene sets loaded:", length(unique(hallmark_sets$gs_name)), "\n")
```

```{r gsea-kegg-ranked-lists}
make_ranked_list <- function(csv_path) {
  df <- read_csv(csv_path, show_col_types = FALSE) %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    arrange(desc(te_lfc)) %>%
    distinct(symbol, .keep_all = TRUE)
  rl <- df$te_lfc
  names(rl) <- df$symbol
  rl
}

ranked_3d <- make_ranked_list(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
ranked_3e <- make_ranked_list(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))

cat("3d ranked list:", length(ranked_3d), "genes\n")
cat("3e ranked list:", length(ranked_3e), "genes\n")
```

```{r gsea-kegg-run}
run_gsea <- function(ranked_list, label) {
  result <- GSEA(
    geneList      = ranked_list,
    TERM2GENE     = kegg_sets,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 20,
    maxGSSize     = 1000,
    verbose       = FALSE
  )
  if (nrow(result@result) > 0) {
    result@result %>%
      dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
      mutate(analysis = label,
             leading_edge_count = sapply(strsplit(core_enrichment, "/"), length),
             leading_edge_fraction = leading_edge_count / setSize)
  } else {
    cat(label, "— no pathways met minimum size requirement\n")
    NULL
  }
}

gsea_3d <- run_gsea(ranked_3d, "3d (sictrl/si3d)")
gsea_3e <- run_gsea(ranked_3e, "3e (sictrl/si3e)")

cat("\n3d significant pathways (padj < 0.05):", sum(gsea_3d$p.adjust < 0.05), "\n")
cat("3e significant pathways (padj < 0.05):", sum(gsea_3e$p.adjust < 0.05), "\n")
```

```{r gsea-kegg-comparison-plot, fig.width=10, fig.height=8}
# Combine and show pathways significant in at least one analysis
gsea_combined <- bind_rows(gsea_3d, gsea_3e)

sig_pathways <- gsea_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>% unique()

cat("Pathways significant in at least one analysis:", length(sig_pathways), "\n")

plot_df <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  mutate(
    pathway_clean = str_replace(ID, "^KEGG_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
    sig = p.adjust < 0.05
  ) %>%
  group_by(ID) %>%
  mutate(max_abs_nes = max(abs(NES))) %>%
  ungroup() %>%
  mutate(pathway_clean = fct_reorder(pathway_clean, max_abs_nes))

ggplot(plot_df, aes(x = NES, y = pathway_clean, color = analysis, alpha = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(aes(x = 0, xend = NES, y = pathway_clean, yend = pathway_clean),
               position = position_dodge(width = 0.6), linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 3,
             aes(shape = sig)) +
  scale_color_manual(values = c("3d (sictrl/si3d)" = "#E69F00",
                                "3e (sictrl/si3e)" = "#56B4E9")) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     labels = c("TRUE" = "padj < 0.05", "FALSE" = "ns"),
                     name = NULL) +
  labs(
    x        = "Normalized Enrichment Score (NES)",
    y        = NULL,
    color    = NULL,
    title    = "KEGG pathway GSEA — eIF3d vs eIF3e, 1hr hypoxia",
    subtitle = "Ranked by TE LFC (sictrl/si); solid = padj < 0.05, open = ns"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("output/24_gsea_kegg_3d_vs_3e_hypoxia_1hr.pdf"), width = 10, height = 8)
```

```{r gsea-kegg-export}
write_csv(gsea_3d, here("output/24_gsea_kegg_3d_hypoxia_1hr.csv"))
write_csv(gsea_3e, here("output/24_gsea_kegg_3e_hypoxia_1hr.csv"))
cat("Exported GSEA results for 3d and 3e\n")
```

## Hallmark GSEA

```{r gsea-hallmark-run}
run_gsea_hallmark <- function(ranked_list, label) {
  result <- GSEA(
    geneList      = ranked_list,
    TERM2GENE     = hallmark_sets,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 20,
    maxGSSize     = 1000,
    verbose       = FALSE
  )
  as.data.frame(result) %>% mutate(analysis = label)
}

gsea_hallmark_3d <- run_gsea_hallmark(ranked_3d, "3d (sictrl/si3d)")
gsea_hallmark_3e <- run_gsea_hallmark(ranked_3e, "3e (sictrl/si3e)")

cat("\n3d significant Hallmark pathways (padj < 0.05):", sum(gsea_hallmark_3d$p.adjust < 0.05), "\n")
cat("3e significant Hallmark pathways (padj < 0.05):", sum(gsea_hallmark_3e$p.adjust < 0.05), "\n")
```

```{r gsea-hallmark-comparison-plot, fig.width=10, fig.height=8}
gsea_hallmark_combined <- bind_rows(gsea_hallmark_3d, gsea_hallmark_3e)

sig_hallmark <- gsea_hallmark_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>% unique()

cat("Hallmark pathways significant in at least one analysis:", length(sig_hallmark), "\n")

if (length(sig_hallmark) > 0) {
  plot_hallmark_df <- gsea_hallmark_combined %>%
    filter(ID %in% sig_hallmark) %>%
    mutate(
      pathway_clean = str_replace(ID, "^HALLMARK_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
      sig = p.adjust < 0.05
    ) %>%
    group_by(ID) %>%
    mutate(max_abs_nes = max(abs(NES))) %>%
    ungroup() %>%
    mutate(pathway_clean = fct_reorder(pathway_clean, max_abs_nes))

  ggplot(plot_hallmark_df, aes(x = NES, y = pathway_clean, color = analysis, alpha = sig)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_segment(aes(x = 0, xend = NES, y = pathway_clean, yend = pathway_clean),
                 position = position_dodge(width = 0.6), linewidth = 0.6) +
    geom_point(position = position_dodge(width = 0.6), size = 3,
               aes(shape = sig)) +
    scale_color_manual(values = c("3d (sictrl/si3d)" = "#E69F00",
                                  "3e (sictrl/si3e)" = "#56B4E9")) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
    scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                       labels = c("TRUE" = "padj < 0.05", "FALSE" = "ns"),
                       name = NULL) +
    labs(
      x        = "Normalized Enrichment Score (NES)",
      y        = NULL,
      color    = NULL,
      title    = "Hallmark pathway GSEA — eIF3d vs eIF3e, 1hr hypoxia",
      subtitle = "Ranked by TE LFC (sictrl/si); solid = padj < 0.05, open = ns"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")

  ggsave(here("output/24_gsea_hallmark_3d_vs_3e_hypoxia_1hr.pdf"), width = 10, height = 8)
} else {
  cat("No significant Hallmark pathways in either analysis.\n")
}
```

```{r gsea-hallmark-export}
write_csv(gsea_hallmark_3d, here("output/24_gsea_hallmark_3d_hypoxia_1hr.csv"))
write_csv(gsea_hallmark_3e, here("output/24_gsea_hallmark_3e_hypoxia_1hr.csv"))
cat("Exported Hallmark GSEA results for 3d and 3e\n")
```

## Section 15: Over-Representation Analysis (ORA)

Uses the pre-filtered gene sets from Section 12 (`gs_3d_only` = 3d-specific targets, `gs_3e_only` = 3e-specific targets) against KEGG and Hallmark gene sets. The background is all genes measured in the si3d 1hr hypoxia comparison.

```{r ora-setup}
# Background: all genes measured in the 1hr hypoxia si3d comparison (load symbol from source file)
background_genes <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                              show_col_types = FALSE) %>%
  pull(symbol) %>% na.omit() %>% unique()
cat("Background genes:", length(background_genes), "\n")

# Foreground: join symbol map (built in Section 12) to get gene symbols
ora_3d_genes <- gs_3d_only %>%
  left_join(sym_map_1hr, by = "gene_id_clean") %>%
  mutate(symbol = coalesce(symbol, gene_id_clean)) %>%
  pull(symbol) %>% na.omit() %>% unique()

ora_3e_genes <- gs_3e_only %>%
  left_join(sym_map_1hr, by = "gene_id_clean") %>%
  mutate(symbol = coalesce(symbol, gene_id_clean)) %>%
  pull(symbol) %>% na.omit() %>% unique()

cat("3d-specific genes (ORA foreground):", length(ora_3d_genes), "\n")
cat("3e-specific genes (ORA foreground):", length(ora_3e_genes), "\n")
```

```{r ora-kegg-run}
run_ora <- function(gene_list, term2gene, label, min_gs = 10) {
  enricher(
    gene          = gene_list,
    universe      = background_genes,
    TERM2GENE     = term2gene,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    qvalueCutoff  = 1,
    minGSSize     = min_gs
  ) %>%
    as.data.frame() %>%
    mutate(analysis = label)
}

ora_kegg_3d <- run_ora(ora_3d_genes, kegg_sets, "3d specific")
ora_kegg_3e <- run_ora(ora_3e_genes, kegg_sets, "3e specific")

cat("\n3d KEGG ORA significant terms (padj < 0.05):", sum(ora_kegg_3d$p.adjust < 0.05), "\n")
cat("3e KEGG ORA significant terms (padj < 0.05):", sum(ora_kegg_3e$p.adjust < 0.05), "\n")
```

```{r ora-kegg-ribosome-genes}
# Print genes in ribosome pathway for 3d ORA
ribosome_row <- ora_kegg_3d %>%
  filter(str_detect(ID, regex("RIBOSOM", ignore_case = TRUE)))

if (nrow(ribosome_row) > 0) {
  cat("Ribosome pathway(s) in 3d ORA:\n")
  ribosome_row %>%
    dplyr::select(ID, GeneRatio, BgRatio, p.adjust, geneID) %>%
    print()

  cat("\nOverlapping genes:\n")
  ribosome_row %>%
    pull(geneID) %>%
    str_split("/") %>%
    unlist() %>%
    sort() %>%
    cat(sep = "\n")
} else {
  cat("No ribosome pathway found in 3d KEGG ORA results.\n")
  cat("Available pathways:\n")
  print(ora_kegg_3d %>% dplyr::select(ID, p.adjust) %>% arrange(p.adjust))
}
```

```{r ora-hallmark-run}
ora_hallmark_3d <- run_ora(ora_3d_genes, hallmark_sets, "3d specific")
ora_hallmark_3e <- run_ora(ora_3e_genes, hallmark_sets, "3e specific")

cat("\n3d Hallmark ORA significant terms (padj < 0.05):", sum(ora_hallmark_3d$p.adjust < 0.05), "\n")
cat("3e Hallmark ORA significant terms (padj < 0.05):", sum(ora_hallmark_3e$p.adjust < 0.05), "\n")
```

```{r ora-plot, fig.width=10, fig.height=7}
make_ora_plot <- function(ora_3d, ora_3e, prefix_pattern, title_str) {
  combined <- bind_rows(ora_3d, ora_3e)

  sig_terms <- combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>% unique()

  cat("Significant ORA terms:", length(sig_terms), "\n")

  if (length(sig_terms) == 0) {
    cat("No significant terms to plot.\n")
    return(invisible(NULL))
  }

  plot_df <- combined %>%
    filter(ID %in% sig_terms) %>%
    mutate(
      term_clean = str_replace(ID, prefix_pattern, "") %>% str_replace_all("_", " ") %>% str_to_title(),
      log10_padj = -log10(p.adjust),
      sig        = p.adjust < 0.05
    ) %>%
    group_by(ID) %>%
    mutate(max_log10 = max(log10_padj)) %>%
    ungroup() %>%
    mutate(term_clean = fct_reorder(term_clean, max_log10))

  ggplot(plot_df, aes(x = log10_padj, y = term_clean, fill = analysis, alpha = sig)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    scale_fill_manual(values = c("3d specific" = "#E69F00", "3e specific" = "#56B4E9")) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
    labs(
      x     = "-log10(padj)",
      y     = NULL,
      fill  = NULL,
      title = title_str,
      subtitle = "Dashed line = padj 0.05; background = all measured genes at 1hr hypoxia"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}

p_ora_kegg <- make_ora_plot(ora_kegg_3d, ora_kegg_3e, "^KEGG_",
                             "KEGG ORA — 3d-specific vs 3e-specific targets, 1hr hypoxia")
if (!is.null(p_ora_kegg)) {
  print(p_ora_kegg)
  ggsave(here("output/24_ora_kegg_3d_vs_3e_hypoxia_1hr.pdf"), p_ora_kegg, width = 10, height = 7)
}

p_ora_hallmark <- make_ora_plot(ora_hallmark_3d, ora_hallmark_3e, "^HALLMARK_",
                                 "Hallmark ORA — 3d-specific vs 3e-specific targets, 1hr hypoxia")
if (!is.null(p_ora_hallmark)) {
  print(p_ora_hallmark)
  ggsave(here("output/24_ora_hallmark_3d_vs_3e_hypoxia_1hr.pdf"), p_ora_hallmark, width = 10, height = 7)
}
```

```{r ora-export}
write_csv(ora_kegg_3d,     here("output/24_ora_kegg_3d_hypoxia_1hr.csv"))
write_csv(ora_kegg_3e,     here("output/24_ora_kegg_3e_hypoxia_1hr.csv"))
write_csv(ora_hallmark_3d, here("output/24_ora_hallmark_3d_hypoxia_1hr.csv"))
write_csv(ora_hallmark_3e, here("output/24_ora_hallmark_3e_hypoxia_1hr.csv"))
cat("Exported ORA results\n")
```

# Session Info

```{r session_info}
sessionInfo()
```