---
title: "Notebook 24: Post-2026 Thesis Committee Analyses"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(ggplot2)
library(ggvenn)
library(ggrepel)
```

## Section 1: Load Data

Spike-in normalized DESeq2 results comparing sictrl vs si3d.
Positive LFC = higher TE in sictrl (control); negative LFC = higher TE in si3d knockdown.

```{r load-data}
normoxia <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
hypoxia_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## Section 2: TE LFC Normoxia (1+4hr regressed) vs TE LFC 1hr Hypoxia

```{r join-data}
plot_df <- normoxia %>%
  dplyr::select(gene_id, symbol, te_lfc_normoxia = te_lfc, te_padj_normoxia = te_padj) %>%
  inner_join(
    hypoxia_1hr %>% 
      dplyr::select(gene_id, te_lfc_hypoxia = te_lfc, te_padj_hypoxia = te_padj),
    by = "gene_id"
  ) %>%
  mutate(sig = case_when(
    te_padj_normoxia < 0.05 & te_padj_hypoxia < 0.05 ~ "both",
    te_padj_normoxia < 0.05                           ~ "normoxia only",
    te_padj_hypoxia  < 0.05                           ~ "hypoxia only",
    TRUE                                               ~ "neither"
  ))

sig_colors <- c("both"          = "#8B0000",
                "normoxia only" = "#2166ac",
                "hypoxia only"  = "#d6604d",
                "neither"       = "grey70")

cat("Genes in plot:", nrow(plot_df), "\n")
table(plot_df$sig)

# Pearson correlations for plot annotations
cor_all <- cor.test(plot_df$te_lfc_normoxia, plot_df$te_lfc_hypoxia)
plot_df_sig <- filter(plot_df, sig != "neither")
cor_sig <- cor.test(plot_df_sig$te_lfc_normoxia, plot_df_sig$te_lfc_hypoxia)

fmt_cor <- function(ct) sprintf("r = %.2f, p = %.2e", ct$estimate, ct$p.value)

# Fit regression and extract residuals
lm_fit <- lm(te_lfc_hypoxia ~ te_lfc_normoxia, data = plot_df, na.action = na.exclude)
plot_df <- plot_df %>%
  mutate(residual = residuals(lm_fit))
```

### Plot 1: All genes

```{r plot-all-genes, fig.width=6, fig.height=5}
p <- ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = sig)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6,
              aes(group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  scale_color_manual(values = sig_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = "Significant\n(padj < 0.05)",
    title = "All genes"
  ) +
  theme_classic(base_size = 12)

p

ggsave(here("plots/24_norm_v_hyp_changes.pdf"), height=4, width=4)

# coord_cartesian, not xlim()/ylim(): those dropped points BEFORE geom_smooth fit its
# line, so the drawn slope (0.707) did not match the annotated correlation computed on
# the full data (r = 0.343, slope 0.494). coord_cartesian only zooms.
p_all_zoom <- p + coord_cartesian(xlim = c(-1, 2.5), ylim = c(-1, 2.5))
p_all_zoom

ggsave(here("plots/24_all_genes_corr.pdf"), p_all_zoom, width=5, height=4)

```

### Export: mRNAs significantly promoted in hypoxia OR normoxia

Union of the two promoted sets behind the plot above, with both contrasts' LFC and
padj side by side. "Promoted" uses the project-standard cut in either contrast:
`padj < 0.05 & te_lfc > 0.5`. Both `te_lfc` columns are log2(sictrl / si3d), so
positive = knockdown lowers TE = eIF3d promotes (the filenames state the ratio in
the opposite order; see CLAUDE.md). NA padj (DESeq2 independent filtering) counts
as not significant.

```{r export-promoted-union}
promoted_lfc_thresh  <- 0.5
promoted_padj_thresh <- 0.05

# Direction check: the siRNA depletes its own target, so EIF3D's *rna_lfc* must be
# strongly positive under the log2(sictrl/si3d) convention. Use rna_lfc, not te_lfc —
# TE is a ratio and EIF3D's own te_lfc is negative here (-1.26 at hypoxia 1hr), so it
# carries no directional information.
eif3d_rna <- bind_rows(
  hypoxia_1hr %>% filter(symbol == "EIF3D") %>% transmute(contrast = "hypoxia 1hr", rna_lfc),
  normoxia    %>% filter(symbol == "EIF3D") %>% transmute(contrast = "normoxia 1+4hr", rna_lfc)
)
stopifnot(
  "EIF3D missing from one of the contrasts" = nrow(eif3d_rna) == 2,
  "EIF3D rna_lfc not strongly positive — contrast direction may have flipped" =
    all(eif3d_rna$rna_lfc > 2)
)
cat("EIF3D rna_lfc —",
    paste(eif3d_rna$contrast, round(eif3d_rna$rna_lfc, 2), collapse = " | "), "\n")

promoted_union <- plot_df %>%
  mutate(
    promoted_hypoxia  = !is.na(te_padj_hypoxia)  & te_padj_hypoxia  < promoted_padj_thresh &
                        te_lfc_hypoxia  > promoted_lfc_thresh,
    promoted_normoxia = !is.na(te_padj_normoxia) & te_padj_normoxia < promoted_padj_thresh &
                        te_lfc_normoxia > promoted_lfc_thresh
  ) %>%
  filter(promoted_hypoxia | promoted_normoxia) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    promoted_in = case_when(
      promoted_hypoxia & promoted_normoxia ~ "both",
      promoted_hypoxia                     ~ "hypoxia only",
      TRUE                                 ~ "normoxia only"
    )
  ) %>%
  dplyr::select(
    symbol, gene_id, gene_id_clean, promoted_in,
    te_lfc_hypoxia_1hr   = te_lfc_hypoxia,
    te_padj_hypoxia_1hr  = te_padj_hypoxia,
    te_lfc_normoxia_1and4hr  = te_lfc_normoxia,
    te_padj_normoxia_1and4hr = te_padj_normoxia
  ) %>%
  arrange(desc(pmax(te_lfc_hypoxia_1hr, te_lfc_normoxia_1and4hr, na.rm = TRUE)))

# Every exported gene must clear the cut in at least one contrast.
stopifnot(
  "export contains a gene promoted in neither contrast" =
    all(with(promoted_union,
             (!is.na(te_padj_hypoxia_1hr) & te_padj_hypoxia_1hr < promoted_padj_thresh &
                te_lfc_hypoxia_1hr > promoted_lfc_thresh) |
             (!is.na(te_padj_normoxia_1and4hr) & te_padj_normoxia_1and4hr < promoted_padj_thresh &
                te_lfc_normoxia_1and4hr > promoted_lfc_thresh))),
  "duplicate genes in export" = !any(duplicated(promoted_union$gene_id))
)

cat("Promoted in hypoxia 1hr OR normoxia 1+4hr:", nrow(promoted_union), "genes\n")
print(table(promoted_union$promoted_in))

write_csv(promoted_union, here("output", "24_promoted_hypoxia_or_normoxia_te.csv"))
cat("Saved: output/24_promoted_hypoxia_or_normoxia_te.csv\n")

head(promoted_union, 15)
```

### Plot 2: All genes, hypoxia-specific promoted highlighted

```{r plot-sig-genes, fig.width=7, fig.height=5}
# Every gene is plotted regardless of significance; the padj/lfc thresholds are used
# only to pick which points are colored red.
# Hypoxia-specific promoted genes: promoted in hypoxia at the project-standard
# lfc 0.5 / padj 0.05 threshold, AND in normoxia both non-significant (padj >= 0.05)
# and below the lfc cutoff (lfc < 0.5). Both normoxia conditions must hold — the
# weaker !(sig & lfc>0.5) form would keep genes with a large normoxia effect that
# merely missed significance.
# NA padj (DESeq2 independent filtering) counts as not-significant.
plot_df_hyp <- plot_df %>%
  mutate(
    hyp_promoted = !is.na(te_padj_hypoxia) & te_padj_hypoxia < 0.05 & te_lfc_hypoxia > 0.5,
    nor_notsig   = is.na(te_padj_normoxia) | te_padj_normoxia >= 0.05,
    nor_small    = !is.na(te_lfc_normoxia) & te_lfc_normoxia < 0.5,
    hyp_group    = if_else(hyp_promoted & nor_notsig & nor_small,
                           "Hypoxia-specific promoted", "All other genes")
  ) %>%
  # draw the red points last so they are not buried under the grey cloud
  arrange(hyp_group == "Hypoxia-specific promoted")

hyp_group_colors <- c("Hypoxia-specific promoted" = "#D55E00",  # Okabe-Ito vermillion
                      "All other genes"           = "grey70")

n_hyp_specific <- sum(plot_df_hyp$hyp_group == "Hypoxia-specific promoted")
cat("Genes plotted:", nrow(plot_df_hyp),
    "| hypoxia-specific promoted:", n_hyp_specific, "\n")

# The red group here uses a stricter normoxia exclusion than the canonical geneset CSV
# (which is built in section 10b as setdiff(hyp_promoted, nor_promoted)), so it must be
# a strict SUBSET of that CSV. If it is not, the hypoxia-side thresholds have drifted
# from the ones used to build the CSV.
hyp_specific_ref <- read_csv(
  here("output/genesets/hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv"),
  show_col_types = FALSE)
red_ids <- sub("\\..*", "", plot_df_hyp$gene_id[plot_df_hyp$hyp_group == "Hypoxia-specific promoted"])
ref_ids <- sub("\\..*", "", hyp_specific_ref$ensembl_gene)
stopifnot(
  "red group is not a subset of hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv" =
    all(red_ids %in% ref_ids)
)
cat("canonical hypoxia-specific set:", length(ref_ids),
    "| after requiring normoxia padj >= 0.05 AND lfc < 0.5:", length(red_ids),
    "| dropped:", length(setdiff(ref_ids, red_ids)), "\n")

p_sig <- ggplot(plot_df_hyp, aes(te_lfc_normoxia, te_lfc_hypoxia, color = hyp_group)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = hyp_group_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = sprintf("%s, n = %d", fmt_cor(cor_all), nrow(plot_df_hyp)),
           size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = NULL,
    title = "Identifying mRNAs selectively promoted by 3d in hypoxia"
  ) +
  theme_classic(base_size = 12) +
  # coord_cartesian, not xlim()/ylim(): xlim()/ylim() drop points outside the range
  # rather than zooming, which would silently make the drawn cloud disagree with the
  # point count reported in the annotation.
  coord_cartesian(xlim = c(-1, 3.5), ylim = c(-1, 3.5))

p_sig

ggsave(here("plots/24_sig_genes_corr.pdf"), p_sig, width = 7, height = 5)
```

### Plot 2b: Reporter construct candidates highlighted

Same scatter, but only the 12 reporter constructs are colored: the 9 hypoxia-specific
candidates in vermillion and the 3 controls (PAN2, DXO, ADAT3) in Okabe-Ito dark blue.

```{r plot-construct-genes, fig.width=7, fig.height=5}
construct_candidates <- c("CLIC1", "EIF4G2", "FUT8", "IWS1", "LAPTM4B",
                          "PIGBOS1", "PIP5K1A", "SERTAD2", "STYXL1")
construct_controls   <- c("PAN2", "DXO", "ADAT3")

plot_df_con <- plot_df %>%
  mutate(
    con_group = case_when(
      symbol %in% construct_candidates ~ "Reporter construct",
      symbol %in% construct_controls   ~ "Control construct",
      TRUE                             ~ "All other genes"
    ),
    con_group = factor(con_group,
                       levels = c("All other genes", "Reporter construct", "Control construct"))
  ) %>%
  # draw the highlighted points last so they are not buried under the grey cloud
  arrange(con_group)

# Every construct must resolve to exactly one point; a symbol that fails to match
# (alias, or filtered out of one of the two DESeq2 tables) would silently plot nothing.
con_counts <- plot_df_con %>%
  filter(con_group != "All other genes") %>%
  dplyr::count(symbol)
stopifnot(
  "not all construct symbols matched exactly one gene" =
    setequal(con_counts$symbol, c(construct_candidates, construct_controls)) &&
    all(con_counts$n == 1)
)

con_group_colors <- c("All other genes"    = "grey70",
                      "Reporter construct" = "#D55E00",   # Okabe-Ito vermillion
                      "Control construct"  = "#0072B2")   # Okabe-Ito dark blue

# Base plot has no gene labels; the labeled version adds geom_text_repel on top, so the
# two versions cannot drift apart in points, colors, scales or limits.
p_con_nolab <- ggplot(plot_df_con, aes(te_lfc_normoxia, te_lfc_hypoxia, color = con_group)) +
  geom_point(data = ~ filter(.x, con_group == "All other genes"),
             alpha = 0.5, size = 1) +
  geom_point(data = ~ filter(.x, con_group != "All other genes"),
             size = 2.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = con_group_colors, breaks = names(con_group_colors)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = sprintf("%s, n = %d", fmt_cor(cor_all), nrow(plot_df_con)),
           size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = NULL,
    title = "Reporter construct candidates"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = c(-1, 3.5), ylim = c(-1, 3.5))

p_con <- p_con_nolab +
  geom_text_repel(data = ~ filter(.x, con_group != "All other genes"),
                  aes(label = symbol), size = 3, min.segment.length = 0,
                  box.padding = 0.4, max.overlaps = Inf, show.legend = FALSE)

p_con
p_con_nolab

ggsave(here("plots/24_construct_genes_corr.pdf"), p_con, width = 7, height = 5)
ggsave(here("plots/24_construct_genes_corr_nolabels.pdf"), p_con_nolab, width = 7, height = 5)
```

### Plot 2c: Named condition-specific genes highlighted

Same scatter, highlighting five hand-picked genes: three hypoxia-specific in
Okabe-Ito vermillion and two normoxia-specific in Okabe-Ito bluish green.
(Orange and sky blue are reserved for eIF3d / eIF3e, so they are not used here.)

Highlights are keyed on **`gene_id`, not symbol**. Three of these five symbols
(DDR1, SMIM10L1, TNKS) appear twice in the joined table under different ENSEMBL
IDs, and in each case only one copy carries the signal — matching on symbol would
color a second point sitting near the origin. The IDs below are the ones that
clear the promoted cut; `highlight_gene_ids` records the choice explicitly.

```{r plot-named-specific-genes, fig.width=7, fig.height=5}
# gene_id -> group. Duplicated symbols resolved to the copy carrying the effect:
#   DDR1     ENSG00000230456 (hyp LFC 4.57)  vs ENSG00000204580 (-0.01, canonical ID)
#   SMIM10L1 ENSG00000256537 (hyp LFC 4.88)  vs ENSG00000281930 (-0.20)
#   TNKS     ENSG00000285372 (nor LFC 8.03)  vs ENSG00000173273 (-0.06, canonical ID)
highlight_gene_ids <- tribble(
  ~gene_id_clean,    ~named_group,
  "ENSG00000256537", "Hypoxia-specific",   # SMIM10L1
  "ENSG00000230456", "Hypoxia-specific",   # DDR1
  "ENSG00000215784", "Hypoxia-specific",   # FAM72D
  "ENSG00000285372", "Normoxia-specific",  # TNKS
  "ENSG00000180867", "Normoxia-specific"   # PDIA3P1
)

plot_df_named <- plot_df %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  left_join(highlight_gene_ids, by = "gene_id_clean") %>%
  mutate(
    named_group = factor(coalesce(named_group, "All other genes"),
                         levels = c("All other genes", "Hypoxia-specific", "Normoxia-specific"))
  ) %>%
  # draw the highlighted points last so they are not buried under the grey cloud
  arrange(named_group)

named_hits <- filter(plot_df_named, named_group != "All other genes")

# Each requested ID must resolve to exactly one point, and its symbol must be the
# expected one — an ID that silently dropped out of either DESeq2 table, or a symbol
# that moved to a different ID in a newer annotation, would plot nothing or the wrong point.
stopifnot(
  "not all highlight gene_ids matched exactly one point" =
    nrow(named_hits) == nrow(highlight_gene_ids) &&
    setequal(named_hits$gene_id_clean, highlight_gene_ids$gene_id_clean),
  "highlighted symbols are not the five expected genes" =
    setequal(named_hits$symbol,
             c("SMIM10L1", "DDR1", "FAM72D", "TNKS", "PDIA3P1"))
)

named_group_colors <- c("All other genes"   = "grey70",
                        "Hypoxia-specific"  = "#D55E00",   # Okabe-Ito vermillion
                        "Normoxia-specific" = "#009E73")   # Okabe-Ito bluish green

# Limits are derived from the highlighted points, not hardcoded: TNKS sits at
# normoxia LFC 8.0 and would fall outside the c(-1, 3.5) window used by Plot 2b,
# so a fixed zoom would drop a point the figure exists to show.
pad     <- 0.6
x_range <- c(-1, max(named_hits$te_lfc_normoxia, na.rm = TRUE) + pad)
y_range <- c(-1, max(named_hits$te_lfc_hypoxia,  na.rm = TRUE) + pad)
stopifnot(
  "a highlighted point falls outside the plot limits" =
    all(between(named_hits$te_lfc_normoxia, x_range[1], x_range[2]) &
        between(named_hits$te_lfc_hypoxia,  y_range[1], y_range[2]))
)

p_named_nolab <- ggplot(plot_df_named, aes(te_lfc_normoxia, te_lfc_hypoxia, color = named_group)) +
  geom_point(data = ~ filter(.x, named_group == "All other genes"),
             alpha = 0.5, size = 1) +
  geom_point(data = ~ filter(.x, named_group != "All other genes"),
             size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = named_group_colors, breaks = names(named_group_colors)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = sprintf("%s, n = %d", fmt_cor(cor_all), nrow(plot_df_named)),
           size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = NULL,
    title = "Condition-specific candidate genes"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = x_range, ylim = y_range)

p_named <- p_named_nolab +
  geom_text_repel(data = ~ filter(.x, named_group != "All other genes"),
                  aes(label = symbol), size = 3.2, min.segment.length = 0,
                  box.padding = 0.5, max.overlaps = Inf, show.legend = FALSE)

p_named
p_named_nolab

ggsave(here("plots/24_named_specific_genes_corr.pdf"), p_named, width = 7, height = 5)
ggsave(here("plots/24_named_specific_genes_corr_nolabels.pdf"), p_named_nolab, width = 7, height = 5)

named_hits %>%
  dplyr::select(symbol, gene_id, named_group,
                te_lfc_normoxia, te_padj_normoxia, te_lfc_hypoxia, te_padj_hypoxia) %>%
  arrange(named_group, desc(te_lfc_hypoxia))
```

### Plot 2d: Clonable condition-specific candidates highlighted

Replacement set for Plot 2c. Three of the five genes in 2c (DDR1, TNKS, PDIA3P1)
carry their eIF3d effect on a locus with **no annotated 5'UTR** in GENCODE v49 —
`protein_coding_CDS_not_defined`, `nonsense_mediated_decay` and
`transcribed_processed_pseudogene` respectively — so no 5'UTR reporter can be
built from them (see notebook 65, chunk `isoform_selection`). The genes below were
re-screened from the same two contrasts under the Plot 2 / Plot 2c condition-specific
definitions, with the added requirement that the **MANE Select transcript is the
most-abundant 5'UTR-annotated isoform in both siCTRL normoxia 1hr and siCTRL
hypoxia 1hr**, TPM >= 10, and 5'UTR length in 50–400 nt. Every pick therefore has a
sequence notebook 65 can extract.

Picks are additionally required to be free of internal Eco31I (GGTCTC/GAGACC) and
BsmBI-v2 (CGTCTC/GAGACG) sites, since the bidirectional reporter vector is
linearised with both. That requirement is what removed TMSB15B, which had the
cleanest TE separation of any normoxia-specific candidate (nor 1.93 / hyp 0.10) but
carries an Eco31I site in its 114 bp 5'UTR; GADD45B replaces it. The site check
itself is enforced in notebook 65, chunk `clonable_integrity`.

`te_lfc_3e` (log2(sictrl / si3e), hypoxia 1hr) is reported for information only —
it did not enter the selection.

```{r plot-clonable-specific-genes, fig.width=7, fig.height=5}
# Symbols, not gene IDs: unlike the Plot 2c set none of these map to two ENSG IDs
# in the joined table. The stopifnot below enforces that (one point per symbol), so
# a future annotation change that splits one of them cannot silently colour two points.
clonable_candidates <- tribble(
  ~symbol,     ~named_group,        ~gene_id_expected,
  "MEMO1",     "Hypoxia-specific",  "ENSG00000162959",  # ErbB2/Rac1 cell motility
  "CCN1",      "Hypoxia-specific",  "ENSG00000142871",  # CYR61, matricellular/angiogenesis
  "DNAJC3",    "Hypoxia-specific",  "ENSG00000102580",  # P58IPK, UPR/PERK inhibition
  "GADD45B",   "Normoxia-specific", "ENSG00000099860",  # stress response / NF-kB
  "MAP1LC3B",  "Normoxia-specific", "ENSG00000140941"   # LC3B, autophagosome
)

plot_df_clonable <- plot_df %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  left_join(clonable_candidates %>% dplyr::select(symbol, named_group), by = "symbol") %>%
  mutate(
    named_group = factor(coalesce(named_group, "All other genes"),
                         levels = c("All other genes", "Hypoxia-specific", "Normoxia-specific"))
  ) %>%
  arrange(named_group)

clonable_hits <- filter(plot_df_clonable, named_group != "All other genes")

# Each symbol must resolve to exactly one point, and to the ENSG ID that was screened
# in the UTR pipeline — a symbol that maps to a second locus, or whose ID moved in a
# newer annotation, would colour the wrong point or plot a gene with no clonable UTR.
stopifnot(
  "not all clonable candidates matched exactly one point" =
    nrow(clonable_hits) == nrow(clonable_candidates) &&
    setequal(clonable_hits$symbol, clonable_candidates$symbol),
  "a clonable candidate resolved to an unexpected gene ID" =
    all(clonable_hits$gene_id_clean ==
        clonable_candidates$gene_id_expected[match(clonable_hits$symbol,
                                                   clonable_candidates$symbol)])
)

# Same palette as Plot 2c: orange and sky blue stay reserved for eIF3d / eIF3e.
clonable_group_colors <- c("All other genes"   = "grey70",
                           "Hypoxia-specific"  = "#D55E00",   # Okabe-Ito vermillion
                           "Normoxia-specific" = "#009E73")   # Okabe-Ito bluish green

# Limits derived from the highlighted points, as in Plot 2c — none of these sit at
# TNKS's extreme, so the window is tighter, but it is still not hardcoded.
pad      <- 0.6
xc_range <- c(-1, max(clonable_hits$te_lfc_normoxia, na.rm = TRUE) + pad)
yc_range <- c(-1, max(clonable_hits$te_lfc_hypoxia,  na.rm = TRUE) + pad)
stopifnot(
  "a highlighted point falls outside the plot limits" =
    all(between(clonable_hits$te_lfc_normoxia, xc_range[1], xc_range[2]) &
        between(clonable_hits$te_lfc_hypoxia,  yc_range[1], yc_range[2]))
)

p_clonable_nolab <- ggplot(plot_df_clonable,
                           aes(te_lfc_normoxia, te_lfc_hypoxia, color = named_group)) +
  geom_point(data = ~ filter(.x, named_group == "All other genes"),
             alpha = 0.5, size = 1) +
  geom_point(data = ~ filter(.x, named_group != "All other genes"),
             size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = clonable_group_colors, breaks = names(clonable_group_colors)) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = sprintf("%s, n = %d", fmt_cor(cor_all), nrow(plot_df_clonable)),
           size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = NULL,
    title = "Condition-specific candidates with a clonable 5' UTR"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = xc_range, ylim = yc_range)

p_clonable <- p_clonable_nolab +
  geom_text_repel(data = ~ filter(.x, named_group != "All other genes"),
                  aes(label = symbol), size = 3.2, min.segment.length = 0,
                  box.padding = 0.5, max.overlaps = Inf, show.legend = FALSE)

p_clonable
p_clonable_nolab

ggsave(here("plots/24_clonable_specific_genes_corr.pdf"), p_clonable, width = 7, height = 5)
ggsave(here("plots/24_clonable_specific_genes_corr_nolabels.pdf"), p_clonable_nolab,
       width = 7, height = 5)

# eIF3e status, reported not filtered on. Positive te_lfc = log2(sictrl/si3e) =
# eIF3e promotes. Hypoxia 1hr is the contrast that matters for the hypoxia-specific
# picks; it is shown for all five so the table is read the same way throughout.
si3e_hyp1 <- read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
stopifnot(
  "EIF3E rna_lfc not strongly positive — si3e contrast direction may have flipped" =
    si3e_hyp1$rna_lfc[si3e_hyp1$symbol == "EIF3E"] > 2
)

clonable_hits %>%
  dplyr::select(symbol, gene_id_clean, named_group,
                te_lfc_normoxia, te_padj_normoxia, te_lfc_hypoxia, te_padj_hypoxia) %>%
  left_join(si3e_hyp1 %>% dplyr::select(gene_id_clean,
                                        te_lfc_3e_hyp1hr = te_lfc,
                                        te_padj_3e_hyp1hr = te_padj),
            by = "gene_id_clean") %>%
  mutate(promoted_3e_hyp1hr = !is.na(te_padj_3e_hyp1hr) & te_padj_3e_hyp1hr < 0.05 &
                              te_lfc_3e_hyp1hr > 0.5) %>%
  arrange(named_group, desc(te_lfc_hypoxia))
```

### Plot 2c follow-up: are the five genes also promoted by eIF3e?

Same five genes, scored against every available si3e TE contrast (normoxia 1+4hr
regressed, normoxia 1/4/24hr, hypoxia 1/4/24hr) at the same promoted cut
(`padj < 0.05 & te_lfc > 0.5`). `te_lfc` here is log2(sictrl / si3e), so positive
= eIF3e promotes. Matched on the same `gene_id_clean` values as Plot 2c, so the
duplicate-symbol resolution carries over.

```{r named-genes-si3e-all-conditions}
si3e_files <- tribble(
  ~condition,                 ~path,
  "normoxia 1+4hr regressed", "output/te_si3e_vs_sictrl_normoxia_1and4hr.csv",
  "normoxia 1hr",             "output/translation_categories_si3e_vs_sictrl_normoxia_1hr.csv",
  "normoxia 4hr",             "output/translation_categories_si3e_vs_sictrl_normoxia_4hr.csv",
  "normoxia 24hr",            "output/translation_categories_si3e_vs_sictrl_normoxia_24hr.csv",
  "hypoxia 1hr",              "output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv",
  "hypoxia 4hr",              "output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv",
  "hypoxia 24hr",             "output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"
)
stopifnot("missing si3e result file" = all(file.exists(here(si3e_files$path))))

si3e_all <- si3e_files %>%
  mutate(dat = map(path, ~ read_csv(here(.x), show_col_types = FALSE) %>%
                     mutate(gene_id_clean = str_remove(gene_id, "\\..*")))) %>%
  dplyr::select(condition, dat) %>%
  unnest(dat)

# Direction check: si3e depletes EIF3E, so EIF3E rna_lfc must be strongly positive
# under log2(sictrl/si3e). The 1+4hr regressed file carries only te_lfc/te_padj, so
# it is excluded from the check by the !is.na() filter rather than passing silently —
# hence 6 rows expected, not 7.
eif3e_chk <- si3e_all %>%
  filter(symbol == "EIF3E", !is.na(rna_lfc)) %>%
  dplyr::select(condition, rna_lfc)
stopifnot(
  "EIF3E rna_lfc not strongly positive — si3e contrast direction may have flipped" =
    nrow(eif3e_chk) == 6 && all(eif3e_chk$rna_lfc > 2)
)
cat("EIF3E rna_lfc across si3e conditions:",
    paste(sprintf("%s %.2f", eif3e_chk$condition, eif3e_chk$rna_lfc), collapse = " | "), "\n")

e3_tab <- si3e_all %>%
  inner_join(highlight_gene_ids, by = "gene_id_clean") %>%
  mutate(
    promoted_3e = !is.na(te_padj) & te_padj < 0.05 & te_lfc > 0.5,
    condition   = factor(condition, levels = si3e_files$condition)
  ) %>%
  dplyr::select(symbol, gene_id_clean, named_group, condition,
                te_lfc_3e = te_lfc, te_padj_3e = te_padj, promoted_3e) %>%
  arrange(named_group, symbol, condition)

stopifnot("expected 5 genes x 7 si3e conditions" = nrow(e3_tab) == 35)

# Per-condition detail
e3_tab %>%
  mutate(te_lfc_3e = round(te_lfc_3e, 3), te_padj_3e = signif(te_padj_3e, 3)) %>%
  print(n = 40)

# One row per gene: promoted by eIF3e anywhere?
e3_summary <- e3_tab %>%
  group_by(named_group, symbol) %>%
  summarise(
    n_conditions_promoted = sum(promoted_3e),
    promoted_by_3e_any    = if_else(any(promoted_3e), "YES", "no"),
    which_conditions      = if_else(any(promoted_3e),
                                    paste(condition[promoted_3e], collapse = ", "), "-"),
    max_te_lfc_3e         = round(max(te_lfc_3e, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(named_group, symbol)

e3_summary

write_csv(e3_tab, here("output", "24_named_genes_si3e_all_conditions.csv"))
cat("Saved: output/24_named_genes_si3e_all_conditions.csv\n")
```

### Plot 1b: All genes, colored by regression residual

```{r plot-residual, fig.width=6, fig.height=5}
resid_lim <- max(abs(plot_df$residual), na.rm = TRUE)

ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = residual)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#d6604d",
                        midpoint = 0, limits = c(-resid_lim, resid_lim),
                        name = "Residual") +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    title = "All genes, colored by residual"
  ) +
  theme_classic(base_size = 12)
```

## Section 3: Venn Diagram — Normoxia vs Hypoxia Regulated Genes

```{r venn-normoxia-hypoxia, fig.width=5, fig.height=5}
venn_list <- list(
  "Normoxia (1+4hr)" = filter(plot_df, te_padj_normoxia < 0.05)$gene_id,
  "Hypoxia 1hr"      = filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id
)

p_venn <- ggvenn(venn_list,
                 fill_color  = c("#2166ac", "#d6604d"),
                 stroke_size = 0.5,
                 set_name_size = 4) +
  ggtitle("Significant TE changes (padj < 0.05, sictrl vs si3d)")

p_venn

ggsave(here("plots/24_venn_normoxia_vs_hypoxia.pdf"), p_venn, width = 5, height = 5)
```

## Section 4: Are Normoxia-Regulated mRNAs a Subset of Hypoxia-Regulated mRNAs?

```{r subset-genesets}
# padj only
norm_sig_padj   <- filter(plot_df, te_padj_normoxia < 0.05)$gene_id
hyp_sig_padj    <- filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id

# padj + |LFC| > 0.5
norm_sig_strict <- filter(plot_df, te_padj_normoxia < 0.05, abs(te_lfc_normoxia) > 0.5)$gene_id
hyp_sig_strict  <- filter(plot_df, te_padj_hypoxia  < 0.05, abs(te_lfc_hypoxia)  > 0.5)$gene_id

universe <- plot_df$gene_id

fisher_overlap <- function(set_a, set_b, universe, label = "") {
  a_and_b <- length(intersect(set_a, set_b))
  a_not_b <- length(setdiff(set_a, set_b))
  b_not_a <- length(setdiff(set_b, set_a))
  neither <- length(universe) - length(union(set_a, set_b))
  m  <- matrix(c(a_and_b, a_not_b, b_not_a, neither), nrow = 2)
  ft <- fisher.test(m, alternative = "greater")
  cat(label, "\n")
  cat(sprintf("  Overlap: %d | OR = %.2f | p = %.2e\n", a_and_b, ft$estimate, ft$p.value))
  cat(sprintf("  %% of normoxia-sig also hypoxia-sig: %.1f%%\n\n",
              100 * a_and_b / length(set_a)))
  invisible(ft)
}

fisher_overlap(norm_sig_padj,   hyp_sig_padj,   universe, "padj < 0.05 only")
fisher_overlap(norm_sig_strict, hyp_sig_strict, universe, "padj < 0.05 & |LFC| > 0.5")
```

### Part 1: Bar plot with LFC + padj cutoff

Universe = normoxia-regulated genes; bars show how many are also hypoxia-regulated vs normoxia only.

```{r barplot-strict, fig.width=4, fig.height=4}
norm_and_hyp <- intersect(norm_sig_strict, hyp_sig_strict)
norm_only    <- setdiff(norm_sig_strict,   hyp_sig_strict)

bar_df <- tibble(
  category = c("Normoxia &\nhypoxia", "Normoxia\nonly"),
  n        = c(length(norm_and_hyp), length(norm_only))
)

ggplot(bar_df, aes(category, n, fill = category)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Normoxia &\nhypoxia" = "#8B0000",
                                "Normoxia\nonly"       = "#2166ac"),
                    guide = "none") +
  labs(x = NULL, y = "Number of genes",
       title = "Normoxia-regulated genes (padj < 0.05 & |LFC| > 0.5)") +
  theme_classic(base_size = 12)

ggsave(here("plots/24_barplot_subset_strict.pdf"), width = 4, height = 4)
```

### Part 2: Distribution of hypoxia LFC by normoxia significance (no cutoff)

```{r ecdf-normoxia-groups, fig.width=6, fig.height=5}
plot_df <- plot_df %>%
  mutate(norm_group = case_when(
    te_padj_normoxia < 0.05 & te_lfc_normoxia > 0 ~ "Normoxia up",
    te_padj_normoxia < 0.05 & te_lfc_normoxia < 0 ~ "Normoxia down",
    TRUE                                            ~ "Not sig"
  ))

group_colors <- c("Normoxia up"   = "#d6604d",
                  "Normoxia down" = "#2166ac",
                  "Not sig"       = "grey70")

# Wilcoxon tests vs not-sig background (computed before the plot so the figure can
# carry the statistic it is illustrating)
not_sig_hyp <- filter(plot_df, norm_group == "Not sig")$te_lfc_hypoxia
norm_up_hyp <- filter(plot_df, norm_group == "Normoxia up")$te_lfc_hypoxia
norm_dn_hyp <- filter(plot_df, norm_group == "Normoxia down")$te_lfc_hypoxia

wt_up <- wilcox.test(norm_up_hyp, not_sig_hyp, alternative = "greater")
wt_dn <- wilcox.test(norm_dn_hyp, not_sig_hyp, alternative = "less")

p_ecdf_norm <- ggplot(plot_df, aes(te_lfc_hypoxia, color = norm_group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = group_colors) +
  annotate("text",
           x = Inf, y = -Inf, hjust = 1.1, vjust = -0.4, size = 3,
           label = paste0(
             "Up vs not-sig p = ",   signif(wt_up$p.value, 3),
             "\nDown vs not-sig p = ", signif(wt_dn$p.value, 3),
             "\nn up = ",      length(norm_up_hyp),
             ", n down = ",    length(norm_dn_hyp),
             ", n not sig = ", length(not_sig_hyp)
           )) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = "Cumulative fraction",
       color = "Normoxia status",
       title = "Hypoxia LFC distribution by normoxia regulation") +
  theme_classic(base_size = 12) +
  # coord_cartesian, not xlim(): xlim() removed rows BEFORE stat_ecdf ran, so each
  # curve was renormalised over its surviving subset. It dropped 171 of 430 (40%) of
  # the "Normoxia up" genes -- precisely the high-LFC genes this figure exists to
  # show -- shifting that group's plotted median from 0.906 to 0.702 and understating
  # the separation, while the Wilcoxon tests below used all 430.
  coord_cartesian(xlim = c(-4, 1))

p_ecdf_norm

ggsave(here("plots/24_ecdf_hypoxia_by_normoxia_group.pdf"), p_ecdf_norm,
       width = 6, height = 4)

cat("Normoxia up vs not-sig (greater):\n")
print(wt_up)

cat("\nNormoxia down vs not-sig (less):\n")
print(wt_dn)
```

## Section 5: Volcano Plot — Hypoxia Changes Colored by Normoxia LFC

```{r volcano-hypoxia, fig.width=6, fig.height=5}
# Clamp color scale symmetrically at the 95th percentile of |normoxia LFC| so most
# points get visible color
norm_lfc_lim <- quantile(abs(plot_df$te_lfc_normoxia), 0.95, na.rm = TRUE)

p_volcano_norm <- ggplot(plot_df, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                    color = te_lfc_normoxia)) +
  geom_point(alpha = 0.9, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_gradient2(low = "#053061", mid = "grey85", high = "#67001f",
                        midpoint = 0, limits = c(-norm_lfc_lim, norm_lfc_lim),
                        oob = scales::squish,
                        name = "Normoxia LFC\n(sictrl/si3d)") +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes, colored by normoxia LFC") +
  theme_classic(base_size = 12) +
  # coord_cartesian, not ylim(): ylim() DELETED the three genes above -log10(padj) = 8
  # (max 29.6), including HERPUD1 (te_lfc 3.17, padj 1.2e-25), the most significant
  # eIF3d-promoted gene at 1hr hypoxia. They were absent from the figure with no
  # indication anything had been clipped. Zooming keeps them at the panel edge.
  coord_cartesian(ylim = c(0, 8))

p_volcano_norm

ggsave(here("plots/24_volcano_hypoxia_colored_normoxia.pdf"), p_volcano_norm,
       width = 8, height = 5)

cat("Genes above the y-axis zoom (-log10 padj > 8):\n")
print(plot_df %>%
        dplyr::filter(-log10(te_padj_hypoxia) > 8) %>%
        dplyr::select(any_of(c("symbol", "gene_id_clean")),
                      te_lfc_hypoxia, te_padj_hypoxia) %>%
        arrange(te_padj_hypoxia))
```

```{r volcano-hypoxia-normsig, fig.width=6, fig.height=5}
plot_df_v2 <- plot_df %>%
  mutate(norm_sig = te_padj_normoxia < 0.05 & te_lfc_normoxia > 0)

p_volcano_normsig <- ggplot(plot_df_v2, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                       color = norm_sig)) +
  geom_point(alpha = 0.7, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c("FALSE" = "grey80", "TRUE" = "#009E73"),
                     labels = c("FALSE" = "Not sig in normoxia",
                                "TRUE"  = "Sig and positive in normoxia (padj < 0.05)"),
                     name   = NULL) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes, normoxia-regulated highlighted") +
  theme_classic(base_size = 12) +
  # coord_cartesian, not ylim() -- see the note on the previous volcano.
  coord_cartesian(ylim = c(0, 8))

p_volcano_normsig

ggsave(here("plots/24_volcano_hypoxia_normsig_highlighted.pdf"), p_volcano_normsig,
       width = 10, height = 5)
```

## Section 6: Venn Diagram — eIF3d vs eIF3e Normoxia Promoted mRNAs (1+4hr)

```{r venn-3d-vs-3e-normoxia, fig.width=5, fig.height=5}
si3d_norm <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

si3e_norm <- read_csv(here("output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

promoted_3d_norm <- filter(si3d_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
promoted_3e_norm <- filter(si3e_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

cat("Promoted by 3d normoxia:", length(promoted_3d_norm), "\n")
cat("Promoted by 3e normoxia:", length(promoted_3e_norm), "\n")
cat("Overlap:", length(intersect(promoted_3d_norm, promoted_3e_norm)), "\n")

venn_norm <- list(
  "eIF3d normoxia" = promoted_3d_norm,
  "eIF3e normoxia" = promoted_3e_norm
)

p_venn_norm <- ggvenn(venn_norm,
                      fill_color  = c("#E69F00", "#56B4E9"),
                      stroke_size = 0.5,
                      set_name_size = 4) +
  ggtitle("Promoted mRNAs: eIF3d vs eIF3e\n(normoxia 1+4hr, padj<0.05 & LFC>0.5)")

p_venn_norm
ggsave(here("plots/24_venn_3d_vs_3e_normoxia.pdf"), p_venn_norm, width = 5, height = 5)
```

## Section 7: Venn Diagrams — eIF3d vs eIF3e by Hypoxic Timepoint

```{r venn-3d-vs-3e-hypoxia, fig.width=5, fig.height=5}
make_hyp_venn <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  set_d <- filter(d, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
  set_e <- filter(e, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

  cat("Hypoxia", timepoint, "— 3d:", length(set_d), "| 3e:", length(set_e),
      "| overlap:", length(intersect(set_d, set_e)), "\n")

  ggvenn(list("eIF3d" = set_d, "eIF3e" = set_e),
         fill_color  = c("#E69F00", "#56B4E9"),
         stroke_size = 0.5,
         set_name_size = 4) +
    ggtitle(paste0("Promoted mRNAs: eIF3d vs eIF3e\nhypoxia ", timepoint,
                   " (padj<0.05 & LFC>0.5)"))
}

p_venn_1hr  <- make_hyp_venn("1hr");  p_venn_1hr
ggsave(here("plots/24_venn_3d_vs_3e_hypoxia_1hr.pdf"),  p_venn_1hr,  width = 5, height = 5)

p_venn_4hr  <- make_hyp_venn("4hr");  p_venn_4hr
ggsave(here("plots/24_venn_3d_vs_3e_hypoxia_4hr.pdf"),  p_venn_4hr,  width = 5, height = 5)

p_venn_24hr <- make_hyp_venn("24hr"); p_venn_24hr
ggsave(here("plots/24_venn_3d_vs_3e_hypoxia_24hr.pdf"), p_venn_24hr, width = 5, height = 5)
```

## Section 8: Dot Plot — eIF3d vs eIF3e Normoxia TE LFC

```{r dotplot-3d-vs-3e-normoxia, fig.width=5.5, fig.height=5}
dot_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
  inner_join(
    si3e_norm %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
    by = "gene_id_clean"
  ) %>%
  mutate(sig = case_when(
    te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
    te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
    te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
    TRUE                                                                          ~ "neither"
  ))

cat("Genes in dot plot:", nrow(dot_df), "\n")
cat("Gene counts per sig category:\n")
print(table(dot_df$sig))

dot_colors <- c("both"    = "#8B0000",
                "3d only" = "#E69F00",
                "3e only" = "#56B4E9",
                "neither" = "grey70")

dot_bg <- filter(dot_df, sig == "neither")
dot_fg <- filter(dot_df, sig != "neither")

p_dot <- ggplot() +
  geom_point(data = dot_bg, aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.4, size = 0.7) +
  geom_point(data = dot_fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  labs(
    x        = "TE LFC normoxia, eIF3d (sictrl/si3d, 1+4hr)",
    y        = "TE LFC normoxia, eIF3e (sictrl/si3e, 1+4hr)",
    title    = "eIF3d vs eIF3e normoxia TE changes",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = c(-3, 5), ylim = c(-3, 5))

p_dot
ggsave(here("plots/24_dotplot_3d_vs_3e_normoxia.pdf"), p_dot, width = 5.5, height = 5)
```

## Section 9: Dot Plots — eIF3d vs eIF3e by Hypoxic Timepoint

```{r dotplot-3d-vs-3e-hypoxia, fig.width=5.5, fig.height=5}
make_hyp_dotplot <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  df <- d %>%
    dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
    inner_join(
      e %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
      by = "gene_id_clean"
    ) %>%
    mutate(sig = case_when(
      te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
      te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
      te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
      TRUE                                                                          ~ "neither"
    ))

  cat("Hypoxia", timepoint, "— genes in plot:", nrow(df), "\n")
  print(table(df$sig))

  dot_colors <- c("both"    = "#8B0000",
                  "3d only" = "#E69F00",
                  "3e only" = "#56B4E9",
                  "neither" = "grey70")

  bg <- filter(df, sig == "neither")
  fg <- filter(df, sig != "neither")

  ggplot() +
    geom_point(data = bg, aes(x = te_lfc_3d, y = te_lfc_3e),
               color = "grey70", alpha = 0.4, size = 0.7) +
    geom_point(data = fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
               alpha = 0.7, size = 1) +
    geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = dot_colors, name = NULL) +
    labs(
      x        = paste0("TE LFC hypoxia ", timepoint, ", eIF3d (sictrl/si3d)"),
      y        = paste0("TE LFC hypoxia ", timepoint, ", eIF3e (sictrl/si3e)"),
      title    = paste0("eIF3d vs eIF3e hypoxia ", timepoint, " TE changes"),
      subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
    ) +
    theme_classic(base_size = 12) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5))
}

p_dot_1hr  <- make_hyp_dotplot("1hr");  p_dot_1hr
ggsave(here("plots/24_dotplot_3d_vs_3e_hypoxia_1hr.pdf"),  p_dot_1hr,  width = 5.5, height = 5)

p_dot_4hr  <- make_hyp_dotplot("4hr");  p_dot_4hr
ggsave(here("plots/24_dotplot_3d_vs_3e_hypoxia_4hr.pdf"),  p_dot_4hr,  width = 5.5, height = 5)

p_dot_24hr <- make_hyp_dotplot("24hr"); p_dot_24hr
ggsave(here("plots/24_dotplot_3d_vs_3e_hypoxia_24hr.pdf"), p_dot_24hr, width = 5.5, height = 5)
```

## Section 10: Condition-Specific Gene Sets for Random Forest

Genes promoted specifically in normoxia (1+4hr) OR hypoxia (1hr) by eIF3d.
"Specific" = padj < 0.05 & LFC > threshold in one condition AND NOT meeting threshold in the other.

### 10a: Threshold sweep

```{r condition_specific_threshold_sweep}
hyp1_clean <- hypoxia_1hr %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

gene_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, norm_lfc = te_lfc, norm_padj = te_padj) %>%
  full_join(
    hyp1_clean %>% dplyr::select(gene_id_clean, hyp_lfc = te_lfc, hyp_padj = te_padj),
    by = "gene_id_clean"
  )

lfc_thresholds <- c(0, 0.25, 0.5, 0.75, 1.0)

thresh_table <- map_dfr(lfc_thresholds, function(thresh) {
  norm_set <- filter(gene_df, norm_padj < 0.05, norm_lfc > thresh)$gene_id_clean
  hyp_set  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > thresh)$gene_id_clean
  tibble(
    lfc_thresh        = thresh,
    normoxia_promotes = length(norm_set),
    hypoxia_promotes  = length(hyp_set),
    both              = length(intersect(norm_set, hyp_set)),
    normoxia_specific = length(setdiff(norm_set, hyp_set)),
    hypoxia_specific  = length(setdiff(hyp_set,  norm_set))
  )
})

print(thresh_table)
```

### 10b: Save condition-specific gene sets

Adjust `lfc_thresh` after reviewing the table above, then save both CSVs to `output/genesets/`.
Files are named to match nb07's geneset_csv convention for single-timepoint params.

```{r save_condition_specific_genesets}
lfc_thresh <- 0.5  # adjust after reviewing thresh_table above

norm_promotes <- filter(gene_df, norm_padj < 0.05, norm_lfc > lfc_thresh)$gene_id_clean
hyp_promotes  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > lfc_thresh)$gene_id_clean
norm_specific <- setdiff(norm_promotes, hyp_promotes)
hyp_specific  <- setdiff(hyp_promotes,  norm_promotes)

cat("normoxia-specific:", length(norm_specific), "\n")
cat("hypoxia-specific:", length(hyp_specific), "\n")

# Symbol lookup from si3d_norm (genome-wide test result, has symbols for all genes)
sym_map <- si3d_norm %>%
  dplyr::select(gene_id_clean, symbol) %>%
  distinct()

save_gs <- function(ids, gs_name, gs_desc, fname) {
  df <- tibble(gene_id_clean = ids) %>%
    left_join(sym_map, by = "gene_id_clean") %>%
    transmute(
      db_gene_symbol    = coalesce(symbol, gene_id_clean),
      ensembl_gene      = gene_id_clean,
      gs_name           = gs_name,
      gs_description    = gs_desc,
      gs_source_species = "homo sapiens"
    )
  write_csv(df, here("output", "genesets", fname))
  cat("Saved:", fname, "—", nrow(df), "genes\n")
}

save_gs(
  norm_specific,
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to normoxia (not in hypoxia 1hr), padj<0.05, LFC>", lfc_thresh),
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh, ".csv")
)

save_gs(
  hyp_specific,
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to hypoxia 1hr (not in normoxia), padj<0.05, LFC>", lfc_thresh),
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh, ".csv")
)
```

**To run nb07 with these gene sets**, set params as:

- **Normoxia-specific**: `condition="normoxia"`, `timepoint="normoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`
- **Hypoxia-specific**: `condition="hypoxia"`, `timepoint="hypoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`

Negative control size-matching (set.seed(9)) is handled automatically in nb07.

```{r}

```

## Section 11: Faceted Dot Plot — eIF3d vs eIF3e Across All Timepoints

```{r dotplot-faceted-all-timepoints, fig.width=18, fig.height=5}
load_tp_df <- function(tp_label, d_path, e_path) {
  d <- read_csv(here(d_path), show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(e_path), show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  d %>%
    dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
    inner_join(
      e %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
      by = "gene_id_clean"
    ) %>%
    mutate(
      timepoint = tp_label,
      sig = case_when(
        te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
        te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
        te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
        TRUE                                                                          ~ "neither"
      )
    )
}

all_tp_df <- bind_rows(
  load_tp_df("Normoxia",
             "output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv",
             "output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
  load_tp_df("Hypoxia 1hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
  load_tp_df("Hypoxia 4hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv"),
  load_tp_df("Hypoxia 24hr",
             "output/translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv",
             "output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("Normoxia", "Hypoxia 1hr",
                                                   "Hypoxia 4hr", "Hypoxia 24hr")))

dot_colors <- c("both"    = "#8B0000",
                "3d only" = "#E69F00",
                "3e only" = "#56B4E9",
                "neither" = "grey70")

p_faceted <- ggplot() +
  geom_point(data = filter(all_tp_df, sig == "neither"),
             aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.3, size = 0.5) +
  geom_point(data = filter(all_tp_df, sig != "neither"),
             aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  facet_wrap(~ timepoint, nrow = 1) +
  coord_cartesian(xlim = c(-2, 4), ylim = c(-2, 4)) +
  labs(
    x        = "TE LFC eIF3d (sictrl/si3d)",
    y        = "TE LFC eIF3e (sictrl/si3e)",
    title    = "eIF3d vs eIF3e TE changes across timepoints",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 16))

p_faceted
ggsave(here("plots/24_dotplot_3d_vs_3e_all_timepoints_faceted.pdf"),
       p_faceted, width = 18, height = 5)

p_faceted_2panel <- ggplot() +
  geom_point(data = filter(all_tp_df, sig == "neither",
                            timepoint %in% c("Normoxia", "Hypoxia 1hr")),
             aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.3, size = 0.5) +
  geom_point(data = filter(all_tp_df, sig != "neither",
                            timepoint %in% c("Normoxia", "Hypoxia 1hr")),
             aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  facet_wrap(~ timepoint, nrow = 1) +
  coord_cartesian(xlim = c(-2, 4), ylim = c(-2, 4)) +
  labs(
    x        = "TE LFC eIF3d (sictrl/si3d)",
    y        = "TE LFC eIF3e (sictrl/si3e)",
    title    = "eIF3d vs eIF3e TE changes across timepoints",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 16) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 16))

p_faceted_2panel
ggsave(here("plots/24_dotplot_3d_vs_3e_normoxia_hypoxia1hr.pdf"),
       p_faceted_2panel, width = 9, height = 5)
```

## Section 12: Save 3d-specific and 3e-specific Gene Sets for nb07

Positive set for RF: promoted by eIF3d but NOT eIF3e at 1hr hypoxia.
Negative set for RF: promoted by eIF3e but NOT eIF3d at 1hr hypoxia.

```{r save-3d-vs-3e-specific-genesets}
hyp1_df <- all_tp_df %>% filter(timepoint == "Hypoxia 1hr")

# Symbol lookup from 1hr hypoxia si3d data
sym_map_1hr <- read_csv(
  here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  dplyr::select(gene_id_clean, symbol) %>%
  distinct()

save_specific_gs <- function(df, gs_name, gs_desc, fname) {
  out <- df %>%
    left_join(sym_map_1hr, by = "gene_id_clean") %>%
    transmute(
      db_gene_symbol    = coalesce(symbol, gene_id_clean),
      ensembl_gene      = gene_id_clean,
      gs_name           = gs_name,
      gs_description    = gs_desc,
      gs_source_species = "homo sapiens"
    )
  write_csv(out, here("output", "genesets", fname))
  cat("Saved:", fname, "—", nrow(out), "genes\n")
}

# 3d only: promoted by 3d, NOT 3e
gs_3d_only <- hyp1_df %>%
  filter(te_padj_3d < 0.05, te_lfc_3d > 0.5,
         !(te_padj_3e < 0.05 & te_lfc_3e > 0.5))

save_specific_gs(
  gs_3d_only,
  "hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5",
  "mRNAs promoted by eIF3d but not eIF3e at 1hr hypoxia (padj<0.05, LFC>0.5)",
  "hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5.csv"
)

# 3e only: promoted by 3e, NOT 3d
gs_3e_only <- hyp1_df %>%
  filter(te_padj_3e < 0.05, te_lfc_3e > 0.5,
         !(te_padj_3d < 0.05 & te_lfc_3d > 0.5))

save_specific_gs(
  gs_3e_only,
  "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5",
  "mRNAs promoted by eIF3e but not eIF3d at 1hr hypoxia (padj<0.05, LFC>0.5)",
  "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv"
)
```

**To run nb07 with these gene sets**, set params in the `derived_values` chunk as:

Note: after running nb07 for each analysis, re-run the comparison section in nb09 to regenerate the comparison plots.

```r
params <- list(
  condition               = "hypoxia",
  gene                    = "3d",
  direction               = "promotes",
  timepoint               = "1hr_not3e",
  lfc                     = "0.5",
  feature_set             = "external_stability",
  clip_mode               = "aggregate",
  exclude_4e_features     = FALSE,
  exclude_own_te_features = TRUE,
  neg_geneset_csv         = "hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv"
)
```

## Section 14: GSEA — KEGG Pathways in 3d-specific vs 3e-specific Hypoxia Targets

Runs GSEA using KEGG pathways on the full TE LFC ranked lists from the si3d and si3e 1hr hypoxia
comparisons. Compares enrichment between the two to identify pathways preferentially regulated
by eIF3d vs eIF3e.

```{r gsea-kegg-setup}
library(fgsea)
library(msigdbr)
library(clusterProfiler)

# KEGG gene sets from MSigDB (TERM2GENE format for clusterProfiler::GSEA)
kegg_sets <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY") %>%
  dplyr::select(gs_name, gene_symbol)
cat("KEGG pathways loaded:", length(unique(kegg_sets$gs_name)), "\n")

# Hallmark gene sets
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol)
cat("Hallmark gene sets loaded:", length(unique(hallmark_sets$gs_name)), "\n")
```

```{r gsea-kegg-ranked-lists}
make_ranked_list <- function(csv_path) {
  df <- read_csv(csv_path, show_col_types = FALSE) %>%
    filter(!is.na(symbol), !is.na(te_lfc)) %>%
    arrange(desc(te_lfc)) %>%
    distinct(symbol, .keep_all = TRUE)
  rl <- df$te_lfc
  names(rl) <- df$symbol
  rl
}

ranked_3d <- make_ranked_list(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
ranked_3e <- make_ranked_list(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"))

cat("3d ranked list:", length(ranked_3d), "genes\n")
cat("3e ranked list:", length(ranked_3e), "genes\n")
```

```{r gsea-kegg-run}
run_gsea <- function(ranked_list, label) {
  # GSEA() is permutation-based via fgsea, so p.adjust -- and therefore which pathways
  # appear in the figures and exported CSVs -- is stochastic without a fixed seed.
  # Project standard seed, matching gsea_si3d_vs_sictrl.Rmd.
  set.seed(9)
  result <- GSEA(
    geneList      = ranked_list,
    TERM2GENE     = kegg_sets,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 20,
    maxGSSize     = 1000,
    verbose       = FALSE
  )
  if (nrow(result@result) > 0) {
    result@result %>%
      dplyr::select(ID, NES, pvalue, p.adjust, setSize, core_enrichment) %>%
      mutate(analysis = label,
             leading_edge_count = sapply(strsplit(core_enrichment, "/"), length),
             leading_edge_fraction = leading_edge_count / setSize)
  } else {
    cat(label, "— no pathways met minimum size requirement\n")
    NULL
  }
}

gsea_3d <- run_gsea(ranked_3d, "3d (sictrl/si3d)")
gsea_3e <- run_gsea(ranked_3e, "3e (sictrl/si3e)")

cat("\n3d significant pathways (padj < 0.05):", sum(gsea_3d$p.adjust < 0.05), "\n")
cat("3e significant pathways (padj < 0.05):", sum(gsea_3e$p.adjust < 0.05), "\n")
```

```{r gsea-kegg-comparison-plot, fig.width=10, fig.height=8}
# Combine and show pathways significant in at least one analysis
gsea_combined <- bind_rows(gsea_3d, gsea_3e)

sig_pathways <- gsea_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>% unique()

cat("Pathways significant in at least one analysis:", length(sig_pathways), "\n")

kegg_plot_df <- gsea_combined %>%
  filter(ID %in% sig_pathways) %>%
  mutate(
    pathway_clean = str_replace(ID, "^KEGG_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
    sig = p.adjust < 0.05
  ) %>%
  group_by(ID) %>%
  mutate(max_abs_nes = max(abs(NES))) %>%
  ungroup() %>%
  mutate(pathway_clean = fct_reorder(pathway_clean, max_abs_nes))

ggplot(kegg_plot_df, aes(x = NES, y = pathway_clean, color = analysis, alpha = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_segment(aes(x = 0, xend = NES, y = pathway_clean, yend = pathway_clean),
               position = position_dodge(width = 0.6), linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 3,
             aes(shape = sig)) +
  scale_color_manual(values = c("3d (sictrl/si3d)" = "#E69F00",
                                "3e (sictrl/si3e)" = "#56B4E9")) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     labels = c("TRUE" = "padj < 0.05", "FALSE" = "ns"),
                     name = NULL) +
  labs(
    x        = "Normalized Enrichment Score (NES)",
    y        = NULL,
    color    = NULL,
    title    = "KEGG pathway GSEA, eIF3d vs eIF3e, 1hr hypoxia",
    subtitle = "Ranked by TE LFC (sictrl/si); solid = padj < 0.05, open = ns"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("plots/24_gsea_kegg_3d_vs_3e_hypoxia_1hr.pdf"), width = 10, height = 8)
```

```{r gsea-kegg-export}
write_csv(gsea_3d, here("output/24_gsea_kegg_3d_hypoxia_1hr.csv"))
write_csv(gsea_3e, here("output/24_gsea_kegg_3e_hypoxia_1hr.csv"))
cat("Exported GSEA results for 3d and 3e\n")
```

## Hallmark GSEA

```{r gsea-hallmark-run}
run_gsea_hallmark <- function(ranked_list, label) {
  set.seed(9)   # see note in run_gsea() above
  result <- GSEA(
    geneList      = ranked_list,
    TERM2GENE     = hallmark_sets,
    eps           = 0,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    minGSSize     = 20,
    maxGSSize     = 1000,
    verbose       = FALSE
  )
  as.data.frame(result) %>% mutate(analysis = label)
}

gsea_hallmark_3d <- run_gsea_hallmark(ranked_3d, "3d (sictrl/si3d)")
gsea_hallmark_3e <- run_gsea_hallmark(ranked_3e, "3e (sictrl/si3e)")

cat("\n3d significant Hallmark pathways (padj < 0.05):", sum(gsea_hallmark_3d$p.adjust < 0.05), "\n")
cat("3e significant Hallmark pathways (padj < 0.05):", sum(gsea_hallmark_3e$p.adjust < 0.05), "\n")
```

```{r gsea-hallmark-comparison-plot, fig.width=10, fig.height=8}
gsea_hallmark_combined <- bind_rows(gsea_hallmark_3d, gsea_hallmark_3e)

sig_hallmark <- gsea_hallmark_combined %>%
  filter(p.adjust < 0.05) %>%
  pull(ID) %>% unique()

cat("Hallmark pathways significant in at least one analysis:", length(sig_hallmark), "\n")

if (length(sig_hallmark) > 0) {
  plot_hallmark_df <- gsea_hallmark_combined %>%
    filter(ID %in% sig_hallmark) %>%
    mutate(
      pathway_clean = str_replace(ID, "^HALLMARK_", "") %>% str_replace_all("_", " ") %>% str_to_title(),
      sig = p.adjust < 0.05
    ) %>%
    group_by(ID) %>%
    mutate(max_abs_nes = max(abs(NES))) %>%
    ungroup() %>%
    mutate(pathway_clean = fct_reorder(pathway_clean, max_abs_nes))

  # Assigned and print()ed explicitly: a bare ggplot inside {} only auto-prints when it
  # is the LAST expression in the block. The ggsave() below followed it, so the object
  # was discarded and this panel was missing from the knitted document (the PDF still
  # wrote, via last_plot()).
  p_hallmark <- ggplot(plot_hallmark_df,
                       aes(x = NES, y = pathway_clean, color = analysis, alpha = sig)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_segment(aes(x = 0, xend = NES, y = pathway_clean, yend = pathway_clean),
                 position = position_dodge(width = 0.6), linewidth = 0.6) +
    geom_point(position = position_dodge(width = 0.6), size = 3,
               aes(shape = sig)) +
    scale_color_manual(values = c("3d (sictrl/si3d)" = "#E69F00",
                                  "3e (sictrl/si3e)" = "#56B4E9")) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
    scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                       labels = c("TRUE" = "padj < 0.05", "FALSE" = "ns"),
                       name = NULL) +
    labs(
      x        = "Normalized Enrichment Score (NES)",
      y        = NULL,
      color    = NULL,
      title    = "Hallmark pathway GSEA, eIF3d vs eIF3e, 1hr hypoxia",
      subtitle = "Ranked by TE LFC (sictrl/si); solid = padj < 0.05, open = ns"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")

  print(p_hallmark)

  ggsave(here("plots/24_gsea_hallmark_3d_vs_3e_hypoxia_1hr.pdf"), p_hallmark,
         width = 10, height = 8)
} else {
  cat("No significant Hallmark pathways in either analysis.\n")
}
```

```{r gsea-hallmark-export}
write_csv(gsea_hallmark_3d, here("output/24_gsea_hallmark_3d_hypoxia_1hr.csv"))
write_csv(gsea_hallmark_3e, here("output/24_gsea_hallmark_3e_hypoxia_1hr.csv"))
cat("Exported Hallmark GSEA results for 3d and 3e\n")
```

## Section 15: Over-Representation Analysis (ORA)

Uses the pre-filtered gene sets from Section 12 (`gs_3d_only` = 3d-specific targets, `gs_3e_only` = 3e-specific targets) against KEGG and Hallmark gene sets. The background is all genes measured in the si3d 1hr hypoxia comparison.

```{r ora-setup}
# Background: all genes measured in the 1hr hypoxia si3d comparison (load symbol from source file)
background_genes <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                              show_col_types = FALSE) %>%
  pull(symbol) %>% na.omit() %>% unique()
cat("Background genes:", length(background_genes), "\n")

# Foreground: join symbol map (built in Section 12) to get gene symbols
ora_3d_genes <- gs_3d_only %>%
  left_join(sym_map_1hr, by = "gene_id_clean") %>%
  mutate(symbol = coalesce(symbol, gene_id_clean)) %>%
  pull(symbol) %>% na.omit() %>% unique()

ora_3e_genes <- gs_3e_only %>%
  left_join(sym_map_1hr, by = "gene_id_clean") %>%
  mutate(symbol = coalesce(symbol, gene_id_clean)) %>%
  pull(symbol) %>% na.omit() %>% unique()

cat("3d-specific genes (ORA foreground):", length(ora_3d_genes), "\n")
cat("3e-specific genes (ORA foreground):", length(ora_3e_genes), "\n")
```

```{r ora-kegg-run}
run_ora <- function(gene_list, term2gene, label, min_gs = 10) {
  enricher(
    gene          = gene_list,
    universe      = background_genes,
    TERM2GENE     = term2gene,
    pAdjustMethod = "fdr",
    pvalueCutoff  = 1,
    qvalueCutoff  = 1,
    minGSSize     = min_gs
  ) %>%
    as.data.frame() %>%
    mutate(analysis = label)
}

ora_kegg_3d <- run_ora(ora_3d_genes, kegg_sets, "3d specific")
ora_kegg_3e <- run_ora(ora_3e_genes, kegg_sets, "3e specific")

cat("\n3d KEGG ORA significant terms (padj < 0.05):", sum(ora_kegg_3d$p.adjust < 0.05), "\n")
cat("3e KEGG ORA significant terms (padj < 0.05):", sum(ora_kegg_3e$p.adjust < 0.05), "\n")
```

```{r ora-kegg-ribosome-genes}
# Print genes in ribosome pathway for 3d ORA
ribosome_row <- ora_kegg_3d %>%
  filter(str_detect(ID, regex("RIBOSOM", ignore_case = TRUE)))

if (nrow(ribosome_row) > 0) {
  cat("Ribosome pathway(s) in 3d ORA:\n")
  ribosome_row %>%
    dplyr::select(ID, GeneRatio, BgRatio, p.adjust, geneID) %>%
    print()

  cat("\nOverlapping genes:\n")
  ribosome_row %>%
    pull(geneID) %>%
    str_split("/") %>%
    unlist() %>%
    sort() %>%
    cat(sep = "\n")
} else {
  cat("No ribosome pathway found in 3d KEGG ORA results.\n")
  cat("Available pathways:\n")
  print(ora_kegg_3d %>% dplyr::select(ID, p.adjust) %>% arrange(p.adjust))
}
```

```{r ora-hallmark-run}
ora_hallmark_3d <- run_ora(ora_3d_genes, hallmark_sets, "3d specific")
ora_hallmark_3e <- run_ora(ora_3e_genes, hallmark_sets, "3e specific")

cat("\n3d Hallmark ORA significant terms (padj < 0.05):", sum(ora_hallmark_3d$p.adjust < 0.05), "\n")
cat("3e Hallmark ORA significant terms (padj < 0.05):", sum(ora_hallmark_3e$p.adjust < 0.05), "\n")
```

```{r ora-plot, fig.width=10, fig.height=7}
make_ora_plot <- function(ora_3d, ora_3e, prefix_pattern, title_str) {
  combined <- bind_rows(ora_3d, ora_3e)

  sig_terms <- combined %>%
    filter(p.adjust < 0.05) %>%
    pull(ID) %>% unique()

  cat("Significant ORA terms:", length(sig_terms), "\n")

  if (length(sig_terms) == 0) {
    cat("No significant terms to plot.\n")
    return(invisible(NULL))
  }

  plot_df <- combined %>%
    filter(ID %in% sig_terms) %>%
    mutate(
      term_clean = str_replace(ID, prefix_pattern, "") %>% str_replace_all("_", " ") %>% str_to_title(),
      log10_padj = -log10(p.adjust),
      sig        = p.adjust < 0.05
    ) %>%
    group_by(ID) %>%
    mutate(max_log10 = max(log10_padj)) %>%
    ungroup() %>%
    mutate(term_clean = fct_reorder(term_clean, max_log10))

  ggplot(plot_df, aes(x = log10_padj, y = term_clean, fill = analysis, alpha = sig)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    scale_fill_manual(values = c("3d specific" = "#E69F00", "3e specific" = "#56B4E9")) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.35), guide = "none") +
    labs(
      x     = "-log10(padj)",
      y     = NULL,
      fill  = NULL,
      title = title_str,
      subtitle = "Dashed line = padj 0.05; background = all measured genes at 1hr hypoxia"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}

p_ora_kegg <- make_ora_plot(ora_kegg_3d, ora_kegg_3e, "^KEGG_",
                             "KEGG ORA, 3d-specific vs 3e-specific targets, 1hr hypoxia")
if (!is.null(p_ora_kegg)) {
  print(p_ora_kegg)
  ggsave(here("plots/24_ora_kegg_3d_vs_3e_hypoxia_1hr.pdf"), p_ora_kegg, width = 10, height = 7)
}

p_ora_hallmark <- make_ora_plot(ora_hallmark_3d, ora_hallmark_3e, "^HALLMARK_",
                                 "Hallmark ORA, 3d-specific vs 3e-specific targets, 1hr hypoxia")
if (!is.null(p_ora_hallmark)) {
  print(p_ora_hallmark)
  ggsave(here("plots/24_ora_hallmark_3d_vs_3e_hypoxia_1hr.pdf"), p_ora_hallmark, width = 10, height = 7)
}
```

```{r ora-export}
write_csv(ora_kegg_3d,     here("output/24_ora_kegg_3d_hypoxia_1hr.csv"))
write_csv(ora_kegg_3e,     here("output/24_ora_kegg_3e_hypoxia_1hr.csv"))
write_csv(ora_hallmark_3d, here("output/24_ora_hallmark_3d_hypoxia_1hr.csv"))
write_csv(ora_hallmark_3e, here("output/24_ora_hallmark_3e_hypoxia_1hr.csv"))
cat("Exported ORA results\n")
```

## eIF3d hypoxia-specific promoters (not normoxia, not eIF3e-regulated)

Genes promoted by eIF3d at 1hr hypoxia, excluding:
- genes also promoted by eIF3d in normoxia (1+4hr time-regressed)
- genes regulated by eIF3e in either condition, either direction (hypoxia 1hr;
  normoxia 1+4hr time-regressed)

```{r hyp3d-specific-geneset}
lfc_thresh <- 0.5
padj_thresh <- 0.05

# Use gene_id_clean (version-stripped) for all cross-file matching below, consistent
# with Sections 6-13's convention - raw gene_id happens to match 1:1 across today's
# files, but gene_id_clean is the robust join key if annotation versions ever drift.
hyp_3d <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  filter(te_lfc > lfc_thresh, te_padj < padj_thresh)

nor_3d_promoted <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv")) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  filter(te_lfc > lfc_thresh, te_padj < padj_thresh) %>%
  pull(gene_id_clean)

# eIF3e normoxia uses the 1+4hr TIME-REGRESSED result, matching the eIF3d normoxia file
# above and every other cross-condition comparison in this notebook. This chunk
# previously read the eIF3e normoxia 1hr file, which was the only place in the notebook
# pairing a 1+4hr eIF3d result against a 1hr eIF3e result.
# Note the 1+4hr eIF3e export is `te_si3e_vs_sictrl_normoxia_1and4hr.csv` (TE only, no
# category/rna/ribo columns), so columns are selected explicitly before binding.
# Both files are log2(sictrl / si3e): positive = eIF3e promotes.
eif3e_regulated <- bind_rows(
  read_csv(here("output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
           show_col_types = FALSE) %>%
    dplyr::select(gene_id, te_lfc, te_padj),
  read_csv(here("output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
           show_col_types = FALSE) %>%
    dplyr::select(gene_id, te_lfc, te_padj)
) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
  filter(abs(te_lfc) > lfc_thresh, te_padj < padj_thresh) %>%
  pull(gene_id_clean) %>%
  unique()

cat("eIF3e-regulated exclusion pool:", length(eif3e_regulated), "genes\n")

final_genes <- hyp_3d %>%
  filter(!gene_id_clean %in% nor_3d_promoted,
         !gene_id_clean %in% eif3e_regulated)

cat("Starting set (hyp 3d promotes 1hr):", nrow(hyp_3d), "\n")
cat("After removing normoxia 3d:", sum(!hyp_3d$gene_id_clean %in% nor_3d_promoted), "\n")
cat("Final (also removing 3e-regulated):", nrow(final_genes), "\n")

# Route through the standard geneset schema (db_gene_symbol/ensembl_gene/gs_name/
# gs_description/gs_source_species) matching every other geneset this notebook saves
# (save_gs()/save_specific_gs() above), instead of writing the raw DESeq2 columns.
final_geneset_out <- final_genes %>%
  transmute(
    db_gene_symbol    = coalesce(symbol, gene_id_clean),
    ensembl_gene       = gene_id_clean,
    gs_name           = "hypoxia_3d_promotes_TE_1hr_not_normoxia_not_3e_lfc0.5",
    gs_description    = paste0("eIF3d-promoted mRNAs at 1hr hypoxia, excluding normoxia-promoted ",
                                "(1+4hr time-regressed) and eIF3e-regulated (hypoxia 1hr or ",
                                "normoxia 1+4hr time-regressed, either direction) genes, ",
                                "padj<0.05, LFC>0.5"),
    gs_source_species = "homo sapiens"
  )

write_csv(final_geneset_out, here("output/genesets/hypoxia_3d_promotes_TE_1hr_not_normoxia_not_3e_lfc0.5.csv"))
cat("Saved: hypoxia_3d_promotes_TE_1hr_not_normoxia_not_3e_lfc0.5.csv —",
    nrow(final_geneset_out), "genes\n")
```

```{r hyp3d-specific-validate}
# Cross-check: derive same set by joining plot_df normoxia columns directly
# (different code path: row-level filter vs vector %in% exclusion)
alt_final <- hyp_3d %>%
  left_join(
    plot_df %>%
      mutate(gene_id_clean = str_remove(gene_id, "\\..*")) %>%
      dplyr::select(gene_id_clean, te_lfc_normoxia, te_padj_normoxia),
    by = "gene_id_clean"
  ) %>%
  filter(!(te_lfc_normoxia > lfc_thresh & te_padj_normoxia < padj_thresh)) %>%
  filter(!gene_id_clean %in% eif3e_regulated)

stopifnot("Gene sets from two approaches don't match" =
  setequal(final_genes$gene_id_clean, alt_final$gene_id_clean))
cat("Validation passed: both approaches yield", nrow(final_genes), "genes\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
