---
title: "Notebook 24: Post-2026 Thesis Committee Analyses"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(ggplot2)
library(ggvenn)
```

## Section 1: Load Data

Spike-in normalized DESeq2 results comparing sictrl vs si3d.
Positive LFC = higher TE in sictrl (control); negative LFC = higher TE in si3d knockdown.

```{r load-data}
normoxia <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
hypoxia_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))
```

## Section 2: TE LFC Normoxia (1+4hr regressed) vs TE LFC 1hr Hypoxia

```{r join-data}
plot_df <- normoxia %>%
  dplyr::select(gene_id, symbol, te_lfc_normoxia = te_lfc, te_padj_normoxia = te_padj) %>%
  inner_join(
    hypoxia_1hr %>% 
      dplyr::select(gene_id, te_lfc_hypoxia = te_lfc, te_padj_hypoxia = te_padj),
    by = "gene_id"
  ) %>%
  mutate(sig = case_when(
    te_padj_normoxia < 0.05 & te_padj_hypoxia < 0.05 ~ "both",
    te_padj_normoxia < 0.05                           ~ "normoxia only",
    te_padj_hypoxia  < 0.05                           ~ "hypoxia only",
    TRUE                                               ~ "neither"
  ))

sig_colors <- c("both"          = "#8B0000",
                "normoxia only" = "#2166ac",
                "hypoxia only"  = "#d6604d",
                "neither"       = "grey70")

cat("Genes in plot:", nrow(plot_df), "\n")
table(plot_df$sig)

# Pearson correlations for plot annotations
cor_all <- cor.test(plot_df$te_lfc_normoxia, plot_df$te_lfc_hypoxia)
plot_df_sig <- filter(plot_df, sig != "neither")
cor_sig <- cor.test(plot_df_sig$te_lfc_normoxia, plot_df_sig$te_lfc_hypoxia)

fmt_cor <- function(ct) sprintf("r = %.2f, p = %.2e", ct$estimate, ct$p.value)

# Fit regression and extract residuals
lm_fit <- lm(te_lfc_hypoxia ~ te_lfc_normoxia, data = plot_df, na.action = na.exclude)
plot_df <- plot_df %>%
  mutate(residual = residuals(lm_fit))
```

### Plot 1: All genes

```{r plot-all-genes, fig.width=6, fig.height=5}
p <- ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = sig)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6,
              aes(group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  scale_color_manual(values = sig_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = "Significant\n(padj < 0.05)",
    title = "All genes"
  ) +
  theme_classic()

p

ggsave(here("output/24_norm_v_hyp_changes.pdf"), height=4, width=4)

p +
  xlim(-1, 2.5) +
  ylim(-1, 2.5)

ggsave(here("output/all_genes_corr.pdf"), width=5, height=4)

```

### Plot 2: Significant in at least one condition

```{r plot-sig-genes, fig.width=6, fig.height=5}
ggplot(plot_df_sig, aes(te_lfc_normoxia, te_lfc_hypoxia, color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6,
              aes(group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  scale_color_manual(values = sig_colors) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_sig), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    color = "Significant\n(padj < 0.05)",
    title = "Significant in \u22651 condition"
  ) +
  theme_classic() +
  xlim(-1, 2.5) +
  ylim(-1, 2.5)

ggsave(here("output/sig_genes_corr.pdf"))
```

### Plot 1b: All genes, colored by regression residual

```{r plot-residual, fig.width=6, fig.height=5}
resid_lim <- max(abs(plot_df$residual), na.rm = TRUE)

ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = residual)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#d6604d",
                        midpoint = 0, limits = c(-resid_lim, resid_lim),
                        name = "Residual") +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
           label = fmt_cor(cor_all), size = 3) +
  labs(
    x     = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y     = "TE LFC 1hr hypoxia (sictrl/si3d)",
    title = "All genes \u2014 colored by residual"
  ) +
  theme_classic()
```

## Section 3: Venn Diagram — Normoxia vs Hypoxia Regulated Genes

```{r venn-normoxia-hypoxia, fig.width=5, fig.height=5}
venn_list <- list(
  "Normoxia (1+4hr)" = filter(plot_df, te_padj_normoxia < 0.05)$gene_id,
  "Hypoxia 1hr"      = filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id
)

p_venn <- ggvenn(venn_list,
                 fill_color  = c("#2166ac", "#d6604d"),
                 stroke_size = 0.5,
                 set_name_size = 4) +
  ggtitle("Significant TE changes (padj < 0.05, sictrl vs si3d)")

p_venn

ggsave(here("output/24_venn_normoxia_vs_hypoxia.pdf"), p_venn, width = 5, height = 5)
```

## Section 4: Are Normoxia-Regulated mRNAs a Subset of Hypoxia-Regulated mRNAs?

```{r subset-genesets}
# padj only
norm_sig_padj   <- filter(plot_df, te_padj_normoxia < 0.05)$gene_id
hyp_sig_padj    <- filter(plot_df, te_padj_hypoxia  < 0.05)$gene_id

# padj + |LFC| > 0.5
norm_sig_strict <- filter(plot_df, te_padj_normoxia < 0.05, abs(te_lfc_normoxia) > 0.5)$gene_id
hyp_sig_strict  <- filter(plot_df, te_padj_hypoxia  < 0.05, abs(te_lfc_hypoxia)  > 0.5)$gene_id

universe <- plot_df$gene_id

fisher_overlap <- function(set_a, set_b, universe, label = "") {
  a_and_b <- length(intersect(set_a, set_b))
  a_not_b <- length(setdiff(set_a, set_b))
  b_not_a <- length(setdiff(set_b, set_a))
  neither <- length(universe) - length(union(set_a, set_b))
  m  <- matrix(c(a_and_b, a_not_b, b_not_a, neither), nrow = 2)
  ft <- fisher.test(m, alternative = "greater")
  cat(label, "\n")
  cat(sprintf("  Overlap: %d | OR = %.2f | p = %.2e\n", a_and_b, ft$estimate, ft$p.value))
  cat(sprintf("  %% of normoxia-sig also hypoxia-sig: %.1f%%\n\n",
              100 * a_and_b / length(set_a)))
  invisible(ft)
}

fisher_overlap(norm_sig_padj,   hyp_sig_padj,   universe, "padj < 0.05 only")
fisher_overlap(norm_sig_strict, hyp_sig_strict, universe, "padj < 0.05 & |LFC| > 0.5")
```

### Part 1: Bar plot with LFC + padj cutoff

Universe = normoxia-regulated genes; bars show how many are also hypoxia-regulated vs normoxia only.

```{r barplot-strict, fig.width=4, fig.height=4}
norm_and_hyp <- intersect(norm_sig_strict, hyp_sig_strict)
norm_only    <- setdiff(norm_sig_strict,   hyp_sig_strict)

bar_df <- tibble(
  category = c("Normoxia &\nhypoxia", "Normoxia\nonly"),
  n        = c(length(norm_and_hyp), length(norm_only))
)

ggplot(bar_df, aes(category, n, fill = category)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Normoxia &\nhypoxia" = "#8B0000",
                                "Normoxia\nonly"       = "#2166ac"),
                    guide = "none") +
  labs(x = NULL, y = "Number of genes",
       title = "Normoxia-regulated genes (padj < 0.05 & |LFC| > 0.5)") +
  theme_classic()

ggsave(here("output/24_barplot_subset_strict.pdf"), width = 4, height = 4)
```

### Part 2: Distribution of hypoxia LFC by normoxia significance (no cutoff)

```{r ecdf-normoxia-groups, fig.width=6, fig.height=5}
plot_df <- plot_df %>%
  mutate(norm_group = case_when(
    te_padj_normoxia < 0.05 & te_lfc_normoxia > 0 ~ "Normoxia up",
    te_padj_normoxia < 0.05 & te_lfc_normoxia < 0 ~ "Normoxia down",
    TRUE                                            ~ "Not sig"
  ))

group_colors <- c("Normoxia up"   = "#d6604d",
                  "Normoxia down" = "#2166ac",
                  "Not sig"       = "grey70")

ggplot(plot_df, aes(te_lfc_hypoxia, color = norm_group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = group_colors) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = "Cumulative fraction",
       color = "Normoxia status",
       title = "Hypoxia LFC distribution by normoxia regulation") +
  theme_classic()+
  xlim(-4,1)

ggsave(here("output/24_ecdf_hypoxia_by_normoxia_group.pdf"), width = 6, height = 4)

# Wilcoxon tests vs not-sig background
not_sig_hyp <- filter(plot_df, norm_group == "Not sig")$te_lfc_hypoxia

cat("Normoxia up vs not-sig (greater):\n")
print(wilcox.test(filter(plot_df, norm_group == "Normoxia up")$te_lfc_hypoxia,
                  not_sig_hyp, alternative = "greater"))

cat("\nNormoxia down vs not-sig (less):\n")
print(wilcox.test(filter(plot_df, norm_group == "Normoxia down")$te_lfc_hypoxia,
                  not_sig_hyp, alternative = "less"))
```

## Section 5: Volcano Plot — Hypoxia Changes Colored by Normoxia LFC

```{r volcano-hypoxia, fig.width=6, fig.height=5}
# Clamp color scale to 5th-95th percentile so most points get visible color
norm_lfc_lim <- quantile(abs(plot_df$te_lfc_normoxia), 0.95, na.rm = TRUE)

ggplot(plot_df, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                    color = te_lfc_normoxia)) +
  geom_point(alpha = 0.9, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_gradient2(low = "#053061", mid = "grey85", high = "#67001f",
                        midpoint = 0, limits = c(-norm_lfc_lim, norm_lfc_lim),
                        oob = scales::squish,
                        name = "Normoxia LFC\n(sictrl/si3d)") +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes \u2014 colored by normoxia LFC") +
  theme_classic()+
  ylim(0,8)

ggsave(here("output/24_volcano_hypoxia_colored_normoxia.pdf"), width = 8, height = 5)
```

```{r volcano-hypoxia-normsig, fig.width=6, fig.height=5}
plot_df_v2 <- plot_df %>%
  mutate(norm_sig = te_padj_normoxia < 0.05 & te_lfc_normoxia > 0)

ggplot(plot_df_v2, aes(te_lfc_hypoxia, -log10(te_padj_hypoxia),
                       color = norm_sig)) +
  geom_point(alpha = 0.7, size = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c("FALSE" = "grey80", "TRUE" = "#009E73"),
                     labels = c("FALSE" = "Not sig in normoxia",
                                "TRUE"  = "Sig and positive in normoxia (padj < 0.05)"),
                     name   = NULL) +
  labs(x     = "TE LFC 1hr hypoxia (sictrl/si3d)",
       y     = expression(-log[10](padj)),
       title = "Hypoxia TE changes \u2014 normoxia-regulated highlighted") +
  theme_classic() +
  ylim(0, 8)

ggsave(here("output/24_volcano_hypoxia_normsig_highlighted.pdf"), width = 10, height = 5)
```

## Section 6: Venn Diagram — eIF3d vs eIF3e Normoxia Promoted mRNAs (1+4hr)

```{r venn-3d-vs-3e-normoxia, fig.width=5, fig.height=5}
si3d_norm <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

si3e_norm <- read_csv(here("output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
                      show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

promoted_3d_norm <- filter(si3d_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
promoted_3e_norm <- filter(si3e_norm, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

cat("Promoted by 3d normoxia:", length(promoted_3d_norm), "\n")
cat("Promoted by 3e normoxia:", length(promoted_3e_norm), "\n")
cat("Overlap:", length(intersect(promoted_3d_norm, promoted_3e_norm)), "\n")

venn_norm <- list(
  "eIF3d normoxia" = promoted_3d_norm,
  "eIF3e normoxia" = promoted_3e_norm
)

p_venn_norm <- ggvenn(venn_norm,
                      fill_color  = c("#E69F00", "#56B4E9"),
                      stroke_size = 0.5,
                      set_name_size = 4) +
  ggtitle("Promoted mRNAs: eIF3d vs eIF3e\n(normoxia 1+4hr, padj<0.05 & LFC>0.5)")

p_venn_norm
ggsave(here("output/24_venn_3d_vs_3e_normoxia.pdf"), p_venn_norm, width = 5, height = 5)
```

## Section 7: Venn Diagrams — eIF3d vs eIF3e by Hypoxic Timepoint

```{r venn-3d-vs-3e-hypoxia, fig.width=5, fig.height=5}
make_hyp_venn <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  set_d <- filter(d, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean
  set_e <- filter(e, te_padj < 0.05, te_lfc > 0.5)$gene_id_clean

  cat("Hypoxia", timepoint, "— 3d:", length(set_d), "| 3e:", length(set_e),
      "| overlap:", length(intersect(set_d, set_e)), "\n")

  ggvenn(list("eIF3d" = set_d, "eIF3e" = set_e),
         fill_color  = c("#E69F00", "#56B4E9"),
         stroke_size = 0.5,
         set_name_size = 4) +
    ggtitle(paste0("Promoted mRNAs: eIF3d vs eIF3e\nhypoxia ", timepoint,
                   " (padj<0.05 & LFC>0.5)"))
}

p_venn_1hr  <- make_hyp_venn("1hr");  p_venn_1hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_1hr.pdf"),  p_venn_1hr,  width = 5, height = 5)

p_venn_4hr  <- make_hyp_venn("4hr");  p_venn_4hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_4hr.pdf"),  p_venn_4hr,  width = 5, height = 5)

p_venn_24hr <- make_hyp_venn("24hr"); p_venn_24hr
ggsave(here("output/24_venn_3d_vs_3e_hypoxia_24hr.pdf"), p_venn_24hr, width = 5, height = 5)
```

## Section 8: Dot Plot — eIF3d vs eIF3e Normoxia TE LFC

```{r dotplot-3d-vs-3e-normoxia, fig.width=5.5, fig.height=5}
dot_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
  inner_join(
    si3e_norm %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
    by = "gene_id_clean"
  ) %>%
  mutate(sig = case_when(
    te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
    te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
    te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
    TRUE                                                                          ~ "neither"
  ))

cat("Genes in dot plot:", nrow(dot_df), "\n")
cat("Gene counts per sig category:\n")
print(table(dot_df$sig))

dot_colors <- c("both"    = "#8B0000",
                "3d only" = "#E69F00",
                "3e only" = "#56B4E9",
                "neither" = "grey70")

dot_bg <- filter(dot_df, sig == "neither")
dot_fg <- filter(dot_df, sig != "neither")

p_dot <- ggplot() +
  geom_point(data = dot_bg, aes(x = te_lfc_3d, y = te_lfc_3e),
             color = "grey70", alpha = 0.4, size = 0.7) +
  geom_point(data = dot_fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
             alpha = 0.7, size = 1) +
  geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
  scale_color_manual(values = dot_colors, name = NULL) +
  labs(
    x        = "TE LFC normoxia \u2014 eIF3d (sictrl/si3d, 1+4hr)",
    y        = "TE LFC normoxia \u2014 eIF3e (sictrl/si3e, 1+4hr)",
    title    = "eIF3d vs eIF3e normoxia TE changes",
    subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
  ) +
  theme_classic(base_size = 12) +
  coord_cartesian(xlim = c(-3, 5), ylim = c(-3, 5))

p_dot
ggsave(here("output/24_dotplot_3d_vs_3e_normoxia.pdf"), p_dot, width = 5.5, height = 5)
```

## Section 9: Dot Plots — eIF3d vs eIF3e by Hypoxic Timepoint

```{r dotplot-3d-vs-3e-hypoxia, fig.width=5.5, fig.height=5}
make_hyp_dotplot <- function(timepoint) {
  d <- read_csv(here(paste0("output/translation_categories_si3d_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))
  e <- read_csv(here(paste0("output/translation_categories_si3e_vs_sictrl_hypoxia_",
                            timepoint, ".csv")),
                show_col_types = FALSE) %>%
    mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  df <- d %>%
    dplyr::select(gene_id_clean, te_lfc_3d = te_lfc, te_padj_3d = te_padj) %>%
    inner_join(
      e %>% dplyr::select(gene_id_clean, te_lfc_3e = te_lfc, te_padj_3e = te_padj),
      by = "gene_id_clean"
    ) %>%
    mutate(sig = case_when(
      te_padj_3d < 0.05 & te_lfc_3d > 0.5 & te_padj_3e < 0.05 & te_lfc_3e > 0.5 ~ "both",
      te_padj_3d < 0.05 & te_lfc_3d > 0.5                                         ~ "3d only",
      te_padj_3e < 0.05 & te_lfc_3e > 0.5                                         ~ "3e only",
      TRUE                                                                          ~ "neither"
    ))

  cat("Hypoxia", timepoint, "— genes in plot:", nrow(df), "\n")
  print(table(df$sig))

  dot_colors <- c("both"    = "#8B0000",
                  "3d only" = "#E69F00",
                  "3e only" = "#56B4E9",
                  "neither" = "grey70")

  bg <- filter(df, sig == "neither")
  fg <- filter(df, sig != "neither")

  ggplot() +
    geom_point(data = bg, aes(x = te_lfc_3d, y = te_lfc_3e),
               color = "grey70", alpha = 0.4, size = 0.7) +
    geom_point(data = fg, aes(x = te_lfc_3d, y = te_lfc_3e, color = sig),
               alpha = 0.7, size = 1) +
    geom_hline(yintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0,   linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = dot_colors, name = NULL) +
    labs(
      x        = paste0("TE LFC hypoxia ", timepoint, " \u2014 eIF3d (sictrl/si3d)"),
      y        = paste0("TE LFC hypoxia ", timepoint, " \u2014 eIF3e (sictrl/si3e)"),
      title    = paste0("eIF3d vs eIF3e hypoxia ", timepoint, " TE changes"),
      subtitle = "Promoted = padj<0.05 & LFC>0.5; dotted lines at LFC=0.5"
    ) +
    theme_classic(base_size = 12) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5))
}

p_dot_1hr  <- make_hyp_dotplot("1hr");  p_dot_1hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_1hr.pdf"),  p_dot_1hr,  width = 5.5, height = 5)

p_dot_4hr  <- make_hyp_dotplot("4hr");  p_dot_4hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_4hr.pdf"),  p_dot_4hr,  width = 5.5, height = 5)

p_dot_24hr <- make_hyp_dotplot("24hr"); p_dot_24hr
ggsave(here("output/24_dotplot_3d_vs_3e_hypoxia_24hr.pdf"), p_dot_24hr, width = 5.5, height = 5)
```

## Section 10: Condition-Specific Gene Sets for Random Forest

Genes promoted specifically in normoxia (1+4hr) OR hypoxia (1hr) by eIF3d.
"Specific" = padj < 0.05 & LFC > threshold in one condition AND NOT meeting threshold in the other.

### 10a: Threshold sweep

```{r condition_specific_threshold_sweep}
hyp1_clean <- hypoxia_1hr %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

gene_df <- si3d_norm %>%
  dplyr::select(gene_id_clean, norm_lfc = te_lfc, norm_padj = te_padj) %>%
  full_join(
    hyp1_clean %>% dplyr::select(gene_id_clean, hyp_lfc = te_lfc, hyp_padj = te_padj),
    by = "gene_id_clean"
  )

lfc_thresholds <- c(0, 0.25, 0.5, 0.75, 1.0)

thresh_table <- map_dfr(lfc_thresholds, function(thresh) {
  norm_set <- filter(gene_df, norm_padj < 0.05, norm_lfc > thresh)$gene_id_clean
  hyp_set  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > thresh)$gene_id_clean
  tibble(
    lfc_thresh        = thresh,
    normoxia_promotes = length(norm_set),
    hypoxia_promotes  = length(hyp_set),
    both              = length(intersect(norm_set, hyp_set)),
    normoxia_specific = length(setdiff(norm_set, hyp_set)),
    hypoxia_specific  = length(setdiff(hyp_set,  norm_set))
  )
})

print(thresh_table)
```

### 10b: Save condition-specific gene sets

Adjust `lfc_thresh` after reviewing the table above, then save both CSVs to `output/genesets/`.
Files are named to match nb07's geneset_csv convention for single-timepoint params.

```{r save_condition_specific_genesets}
lfc_thresh <- 0.5  # adjust after reviewing thresh_table above

norm_promotes <- filter(gene_df, norm_padj < 0.05, norm_lfc > lfc_thresh)$gene_id_clean
hyp_promotes  <- filter(gene_df, hyp_padj  < 0.05, hyp_lfc  > lfc_thresh)$gene_id_clean
norm_specific <- setdiff(norm_promotes, hyp_promotes)
hyp_specific  <- setdiff(hyp_promotes,  norm_promotes)

cat("normoxia-specific:", length(norm_specific), "\n")
cat("hypoxia-specific:", length(hyp_specific), "\n")

# Symbol lookup from si3d_norm (genome-wide test result, has symbols for all genes)
sym_map <- si3d_norm %>%
  dplyr::select(gene_id_clean, symbol) %>%
  distinct()

save_gs <- function(ids, gs_name, gs_desc, fname) {
  df <- tibble(gene_id_clean = ids) %>%
    left_join(sym_map, by = "gene_id_clean") %>%
    transmute(
      db_gene_symbol    = coalesce(symbol, gene_id_clean),
      ensembl_gene      = gene_id_clean,
      gs_name           = gs_name,
      gs_description    = gs_desc,
      gs_source_species = "homo sapiens"
    )
  write_csv(df, here("output", "genesets", fname))
  cat("Saved:", fname, "—", nrow(df), "genes\n")
}

save_gs(
  norm_specific,
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to normoxia (not in hypoxia 1hr), padj<0.05, LFC>", lfc_thresh),
  paste0("normoxia_3d_promotes_TE_normoxia_specific_lfc", lfc_thresh, ".csv")
)

save_gs(
  hyp_specific,
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh),
  paste0("eIF3d-promoted mRNAs specific to hypoxia 1hr (not in normoxia), padj<0.05, LFC>", lfc_thresh),
  paste0("hypoxia_3d_promotes_TE_hypoxia_specific_lfc", lfc_thresh, ".csv")
)
```

**To run nb07 with these gene sets**, set params as:

- **Normoxia-specific**: `condition="normoxia"`, `timepoint="normoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`
- **Hypoxia-specific**: `condition="hypoxia"`, `timepoint="hypoxia_specific"`, `lfc="0.5"`, `feature_set="external_stability"`, `exclude_own_te_features=TRUE`

Negative control size-matching (set.seed(9)) is handled automatically in nb07.
