---
title: "Notebook 25: eIF3d vs eIF3e Feature Importance Comparison"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
```

# Overview

Compares signed feature importance across four nb07 RF analyses at 1hr hypoxia:

- **3d standard**: 3d-promoted vs si3d neg control (`1hr`, `external_stability`, `noOwnTE`)
- **3e standard**: 3e-promoted vs si3e neg control (`1hr`, `external_stability`, `noOwnTE`)
- **3d specific**: 3d-only (not 3e) vs si3d neg control (`1hr_not3e`, `external_stability`, `noOwnTE`)
- **3e specific**: 3e-only (not 3d) vs si3e neg control (`1hr_not3d`, `external_stability`, `noOwnTE`)

Re-run nb07 for each analysis to regenerate the directionality CSVs, then re-run this notebook.

# Load Directionality Files

```{r load-directionality}
dir_path <- here("output", "predictive_modeling")

load_dir <- function(fname, label) {
  path <- file.path(dir_path, fname)
  if (!file.exists(path)) {
    message("File not found (run nb07 first): ", fname)
    return(NULL)
  }
  read_csv(path, show_col_types = FALSE) %>% mutate(analysis = label)
}

d_3d_std <- load_dir(
  "rf_classification_directionality_eif33d_promotes_hypoxia_1hr_lfc0.5_external_stability_clipaggregate_no_boruta_noOwnTE.csv",
  "3d standard"
)
d_3e_std <- load_dir(
  "rf_classification_directionality_eif33e_promotes_hypoxia_1hr_lfc0.5_external_stability_clipaggregate_no_boruta_noOwnTE.csv",
  "3e standard"
)
d_3d_sp <- load_dir(
  "rf_classification_directionality_eif33d_promotes_hypoxia_1hr_not3e_lfc0.5_external_stability_clipaggregate_no_boruta_noOwnTE.csv",
  "3d specific (not 3e)"
)
d_3e_sp <- load_dir(
  "rf_classification_directionality_eif33e_promotes_hypoxia_1hr_not3d_lfc0.5_external_stability_clipaggregate_no_boruta_noOwnTE.csv",
  "3e specific (not 3d)"
)

all_dir <- bind_rows(d_3d_std, d_3e_std, d_3d_sp, d_3e_sp)
cat("Analyses loaded:", paste(unique(all_dir$analysis), collapse = ", "), "\n")
cat("Total rows:", nrow(all_dir), "\n")
```

# Comparison Plots

```{r comparison-helper}
make_comparison_plot <- function(analyses, top_n = 30, title = NULL) {
  df <- all_dir %>% filter(analysis %in% analyses)

  top_feats <- df %>%
    group_by(analysis) %>%
    arrange(desc(abs(signed_importance))) %>%
    slice_head(n = top_n) %>%
    ungroup() %>%
    pull(feature) %>% unique()

  plot_df <- df %>%
    filter(feature %in% top_feats) %>%
    group_by(feature) %>%
    mutate(max_abs = max(abs(signed_importance), na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(
      feature_clean = fct_reorder(feature_clean, max_abs),
      analysis      = factor(analysis, levels = analyses)
    )

  ggplot(plot_df, aes(x = signed_importance, y = feature_clean, color = analysis)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_segment(aes(x = 0, xend = signed_importance,
                     y = feature_clean, yend = feature_clean),
                 position = position_dodge(width = 0.6), linewidth = 0.5, alpha = 0.6) +
    geom_point(position = position_dodge(width = 0.6), size = 2.5) +
    scale_color_manual(values = c("3d standard"          = "#E69F00",
                                  "3e standard"          = "#56B4E9",
                                  "3d specific (not 3e)" = "#E69F00",
                                  "3e specific (not 3d)" = "#56B4E9")) +
    labs(x = "Signed Importance (positive = higher in promotes)", y = NULL,
         color = NULL, title = title) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## Plot 1: Specific analyses — 3d-only vs 3e-only

```{r plot-specific, fig.width=12, fig.height=8}
if (!is.null(d_3d_sp) && !is.null(d_3e_sp)) {
  p_sp <- make_comparison_plot(
    c("3d specific (not 3e)", "3e specific (not 3d)"),
    title = "eIF3d vs eIF3e: specific analyses (3d-only vs 3e-only, vs neg control, 1hr hypoxia)"
  )
  print(p_sp)
  ggsave(here("output/25_importance_comparison_specific.pdf"), p_sp, width = 12, height = 8)
}
```

## Plot 1b: Quadrant scatter — 3d-specific vs 3e-specific signed importance

```{r plot-specific-quadrant, fig.width=7, fig.height=7}
if (!is.null(d_3d_sp) && !is.null(d_3e_sp)) {
  quad_df <- d_3d_sp %>%
    dplyr::select(feature, feature_clean, category, signed_importance) %>%
    rename(si_3d = signed_importance) %>%
    inner_join(
      d_3e_sp %>% dplyr::select(feature, signed_importance) %>% rename(si_3e = signed_importance),
      by = "feature"
    ) %>%
    mutate(
      broad_category = case_when(
        str_detect(feature, "^dap5_")                          ~ "Sensitivity to DAP5",
        category == "4E / DAP5"                                ~ "Sensitivity to eIF4E",
        category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
        TRUE                                                   ~ "Intrinsic mRNA features"
      )
    )

  axis_lim <- max(abs(c(quad_df$si_3d, quad_df$si_3e)), na.rm = TRUE) * 1.05

  p_quad <- ggplot(quad_df, aes(x = si_3d, y = si_3e,
                                color = broad_category, label = feature_clean)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.7, size = 2) +
    ggrepel::geom_text_repel(
      data = filter(quad_df, abs(si_3d) > quantile(abs(si_3d), 0.75, na.rm = TRUE) |
                               abs(si_3e) > quantile(abs(si_3e), 0.75, na.rm = TRUE)),
      size = 3, max.overlaps = 20, show.legend = FALSE
    ) +
    scale_color_manual(values = c(
      "Intrinsic mRNA features" = "#009E73",
      "eIF3 binding evidence"   = "#E69F00",
      "Sensitivity to eIF4E"    = "#56B4E9",
      "Sensitivity to DAP5"     = "#CC79A7"
    )) +
    coord_fixed(xlim = c(-axis_lim, axis_lim), ylim = c(-axis_lim, axis_lim)) +
    labs(
      x     = "Signed Importance — eIF3d specific (not 3e)",
      y     = "Signed Importance — eIF3e specific (not 3d)",
      color = NULL,
      title = "Feature importance: eIF3d-specific vs eIF3e-specific",
      subtitle = "Top-right = important for both | left = 3d-specific | top = 3e-specific"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  print(p_quad)
  ggsave(here("output/25_importance_quadrant_specific.pdf"), p_quad, width = 7, height = 7)

  # Version 2: grey out features that agree between 3d and 3e (top-right / bottom-left)
  # Highlight only features that diverge: top-left (3e > 0, 3d < 0) and bottom-right (3d > 0, 3e < 0)
  quad_df2 <- quad_df %>%
    mutate(
      diverges = (si_3d > 0 & si_3e < 0) | (si_3d < 0 & si_3e > 0),
      label_text = if_else(diverges, feature_clean, NA_character_)
    )

  p_quad2 <- ggplot(quad_df2, aes(x = si_3d, y = si_3e)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(data = filter(quad_df2, !diverges),
               color = "grey80", alpha = 0.5, size = 2) +
    geom_point(data = filter(quad_df2, diverges),
               aes(color = broad_category), alpha = 0.8, size = 2.5) +
    ggrepel::geom_text_repel(
      aes(label = label_text),
      data = filter(quad_df2, !is.na(label_text)),
      size = 3, max.overlaps = 20, show.legend = FALSE
    ) +
    scale_color_manual(values = c(
      "Intrinsic mRNA features" = "#009E73",
      "eIF3 binding evidence"   = "#E69F00",
      "Sensitivity to eIF4E"    = "#56B4E9",
      "Sensitivity to DAP5"     = "#CC79A7"
    )) +
    coord_fixed(xlim = c(-axis_lim, axis_lim), ylim = c(-axis_lim, axis_lim)) +
    labs(
      x     = "Signed Importance — eIF3d specific (not 3e)",
      y     = "Signed Importance — eIF3e specific (not 3d)",
      color = NULL,
      title = "Feature importance: divergent features highlighted",
      subtitle = "Grey = agree between 3d & 3e | Colored = diverge"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "bottom")

  print(p_quad2)
  ggsave(here("output/25_importance_quadrant_specific_divergent.pdf"), p_quad2, width = 7, height = 7)
}
```

## Plot 2: Standard analyses — 3d vs 3e

```{r plot-standard, fig.width=12, fig.height=8}
if (!is.null(d_3d_std) && !is.null(d_3e_std)) {
  p_std <- make_comparison_plot(
    c("3d standard", "3e standard"),
    title = "eIF3d vs eIF3e: standard analyses (vs si3d/si3e neg control, 1hr hypoxia)"
  )
  print(p_std)
  ggsave(here("output/25_importance_comparison_standard.pdf"), p_std, width = 12, height = 8)
}
```

## Plot 3: All four analyses

```{r plot-all-four, fig.width=14, fig.height=9}
if (nrow(all_dir) > 0) {
  p_all <- make_comparison_plot(
    c("3d standard", "3e standard", "3d specific (not 3e)", "3e specific (not 3d)"),
    top_n = 25,
    title = "All four analyses: eIF3d vs eIF3e, standard and specific (1hr hypoxia)"
  )
  print(p_all)
  ggsave(here("output/25_importance_comparison_all_four.pdf"), p_all, width = 14, height = 9)
}
```

# Session Info

```{r session_info}
sessionInfo()
```
