---
title: "Notebook 25: Goodarzi MDA-MB-231 mRNA Stability Analysis"
output: html_document
date: "`r Sys.Date()`"
---

C-to-T conversion rates from SLAM-seq in MDA-MB-231 and other BRCA cell lines.

**Source:** Karner H, Mittmann TC, Chen VW, Borah AA, Langen A, Yousefi H, Fish L,
Zaro BW, Navickas A, Goodarzi H. "Integrative analysis of mRNA stability regulation
uncovers a metastasis-suppressive program in breast cancer." *Science Advances*,
11 Mar 2026. PMID 41811965. Data: GEO GSE301426 (SuperSeries GSE301535).

**What the C-to-T ratio measures.** The GEO processing description settles this
explicitly (`!Sample_data_processing`, GSE301426):

> "Since the cells were pulsed with 4sU for 4hr, 4sU incorporation, and therefore the
> resulting C to T ratio, are driven by **the rate of transcription**. At steady state,
> decay rate is equal to transcription rate normalized by the steady state expression.
> So, we used this rule to estimate a decay rate (log transformed) for every gene across
> all cell lines, based on its SLAM-seq C to T ratio **and its baseline expression (AU)**."

So the deposited ratio is the paper's **transcription-rate** measure. Their decay rate is
a derived quantity, `log(decay) = log(C/T ratio) - log(expression)`, and it is **not
deposited anywhere** -- neither the series file nor any sample file contains it.

The column built here is therefore the paper's transcription-rate input, not its decay
estimate. Empirically it still anticorrelates strongly with published half-lives
(Spearman -0.611 vs Zhu 2024, -0.477 vs Hia 2026; top C/T quartile median half-life 2.9 h
vs 10.3 h for the bottom quartile), which is expected: at steady state transcription and
degradation flux are equal, so an unnormalized ratio tracks turnover without isolating
either term.

**Practical consequence:** do not label `karner2026_mda231_log2_ct` as "half-life" or
"decay rate" in a figure or SHAP write-up -- by the authors' own definition it is a
transcription rate, and converting it to their decay rate requires dividing by baseline
expression per cell line. The checks below assert the empirical turnover relationship
(a parse check), not a mechanistic claim.

**Upstream processing already applied by the authors** (do not repeat it): Salmon v0.14.1
against GENCODE v28, transcripts >1 TPM average, Bowtie2 v2.3.5 to major isoforms,
`samtools mpileup` read stacks, minimum 10-read coverage at counted Ts, germline variant
filtering, pseudo-count of 1, and outlier removal at 3x IQR from the median. Note the
GENCODE v28 basis against this project's v49 annotation.

**Data file pre-read summary:**

- 10,994 genes; row IDs are gene symbols and lncRNA names
- MDA-MB-231 replicates: columns `MDA-MB-231-1` and `MDA-MB-231-2`
- Values: C-to-T ratios in (0, 1]; no zeros; ~10-15% NAs per column
- log2 transform gives all-negative values (median C/T ~ 0.02 -> log2 ~ -5.6)
- **Saturated artifacts:** 1631 of 115,809 values across the 12 samples (1.4%) are
  exactly 1.0 -- and are the *only* integer-valued entries in the matrix. 112 fall in
  MDA-MB-231 rep 1 and 84 in rep 2, with no replicate agreement (e.g. SRSF4 = 1.000 in
  rep 1 vs 0.018 in rep 2). These are not low-coverage single-read calls: the authors
  required >=10-read coverage at counted Ts. They look like saturated or capped values,
  and they survived the authors' own 3x IQR outlier filter. Whatever their origin they
  are not usable ratios -- mean raw Pearson between replicates across the six lines is
  0.168 with them included and 0.736 once ratios > 0.2 are dropped. Filtered in Section 2.

---

## Section 1: Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(ggplot2)
```

```{r gene_groups}
# eIF3d-promoted genes (hypoxia 1hr, TE lfc > 0.5)
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE)
promotes_ids <- promotes$ensembl_gene

# Negative control pool (size-matched, set.seed(9))
neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  filter(!ensembl_gene %in% promotes_ids)

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)

cat("Promotes genes:", length(promotes_ids), "\n")
cat("Neg ctrl genes (size-matched):", length(neg_ctrl_ids), "\n")
```

---

## Section 2: Load and process MDA-MB-231 C-to-T ratios

```{r load_slam}
slam_raw <- read_tsv(
  here("accessories", "csc_data", "karner_231_slamseq",
       "GSE301426_BRCA_SLAM_CTratio_gs_counts.txt"),
  show_col_types = FALSE
) %>%
  dplyr::rename(gene_symbol = `...1`) %>%
  mutate(gene_symbol = as.character(gene_symbol))

cat("Total rows loaded:", nrow(slam_raw), "\n")
cat("Columns:", paste(colnames(slam_raw), collapse = ", "), "\n")
```

### Replication check: does this file reproduce the published dataset?

Per the project convention for external data, verify the file before any joins.
GSE301426 is a resource dataset (per-gene conversion ratios across six cell lines)
rather than one headline number, so the anchors are the design facts stated in the
paper plus the biology the measurement must recapitulate to be the quantity it claims.

```{r paper_replication}
# --- Anchor 1: study design. Six BRCA lines, two biological replicates each. ---
expected_lines <- c("HCC1806", "HCC38", "MCF7", "MDA-MB-231", "MDA-MD-453", "ZR-75")
sample_cols <- setdiff(colnames(slam_raw), "gene_symbol")
stopifnot(
  "GSE301426 must have 12 sample columns (6 lines x 2 reps)" =
    length(sample_cols) == 12,
  "All six published cell lines must be present, 2 reps each" =
    all(sapply(expected_lines, function(l) sum(startsWith(sample_cols, l)) == 2))
)
cat("Anchor 1 PASS - 12 samples,", length(expected_lines), "cell lines x 2 reps\n")

# --- Anchor 2: values are conversion ratios in (0, 1], no zeros. ---
vals <- as.matrix(slam_raw[, sample_cols])
stopifnot(
  "C-to-T ratios must lie in (0, 1]" = all(vals > 0 & vals <= 1, na.rm = TRUE),
  "Median C-to-T ratio must be in the published 0.005-0.05 range" =
    all(between(apply(vals, 2, median, na.rm = TRUE), 0.005, 0.05))
)
cat("Anchor 2 PASS - all ratios in (0,1]; per-sample medians",
    paste(round(range(apply(vals, 2, median, na.rm = TRUE)), 4), collapse = " to "), "\n")

# --- Anchor 3: MDA-MB-231 replicates must agree. ---
# Rank-based: raw Pearson is destroyed by the ratio == 1 low-coverage artifacts
# (r = 0.045 with them, 0.777 without), so Spearman is the honest concordance stat.
rep_ok <- complete.cases(slam_raw[, c("MDA-MB-231-1", "MDA-MB-231-2")])
rho_rep <- cor(slam_raw$`MDA-MB-231-1`[rep_ok], slam_raw$`MDA-MB-231-2`[rep_ok],
               method = "spearman")
stopifnot("MDA-MB-231 replicate Spearman must exceed 0.7" = rho_rep > 0.7)
cat("Anchor 3 PASS - MDA-MB-231 replicate Spearman =", round(rho_rep, 3),
    "( n =", sum(rep_ok), ")\n")

# --- Anchor 4: the ratio must track turnover, in the established direction. ---
# Cross-dataset validation against independently published control half-lives.
# This is a data-integrity check, not a mechanistic claim: whatever the ratio is called,
# a correct parse of this file reproduces a strong ANTIcorrelation with measured
# half-life. A sign flip or a collapse toward zero means the wrong column was read.
# Conditional so this notebook still runs standalone if 22_/23_ have not been knit.
hl_files <- c(
  "Zhu 2024 (sgNT)"  = "cnot3_slamseq_halflives.rds",
  "Hia 2026 (ctrl)"  = "dhx29_slamseq_halflives.rds"
)
ct_probe <- slam_raw %>%
  transmute(gene_symbol,
            log2_ct = log2(rowMeans(cbind(`MDA-MB-231-1`, `MDA-MB-231-2`)))) %>%
  filter(is.finite(log2_ct))

for (nm in names(hl_files)) {
  fp <- here("output", "predictive_modeling", hl_files[[nm]])
  if (!file.exists(fp)) {
    cat("Anchor 4 SKIP -", nm, "half-life file not found (run 22_/23_ first)\n")
    next
  }
  hl <- readRDS(fp)
  hl_df <- if ("sgNT_halflife" %in% names(hl)) {
    tibble(gene_symbol = hl$Gene_Name, halflife = hl$sgNT_halflife)
  } else {
    tibble(gene_symbol = hl$symbol, halflife = hl$Ctrl_Half_life)
  }
  d <- inner_join(ct_probe, hl_df, by = "gene_symbol") %>%
    filter(is.finite(halflife), halflife > 0)
  rho <- cor(d$log2_ct, d$halflife, method = "spearman")
  stopifnot("C-to-T lost its expected anticorrelation with published half-life" =
    rho < -0.3)
  cat("Anchor 4 PASS -", nm, ": Spearman(log2_ct, half-life) =",
      round(rho, 3), ", n =", nrow(d), "\n")
}
```

**Anchor 5: the paper's own headline result.** Anchors 1-4 establish that the C-to-T
table is the right quantity, but they do not confirm it came from *this* study. The
paper's central claim is the RBMS3 -> TXNIP axis: RBMS3 depletion destabilizes TXNIP.
GSE301427 (the shRBMS3 RNA-seq subseries of the same SuperSeries, GSE301535) sits in
the same directory, so the claim is directly checkable.

```{r paper_replication_rbms3}
rbms3_path <- here("accessories", "csc_data", "karner_231_slamseq",
                   "GSE301427_MDA_shRBMS3_gs_logFC.txt")
stopifnot("GSE301427 shRBMS3 file missing - cannot verify paper provenance" =
  file.exists(rbms3_path))

rbms3 <- read_tsv(rbms3_path, show_col_types = FALSE)
get_gene <- function(g) rbms3 %>% filter(GENE == g) %>% slice(1)
kd   <- get_gene("RBMS3")
txn  <- get_gene("TXNIP")

stopifnot(
  "RBMS3 and TXNIP must both be present in GSE301427" =
    nrow(kd) == 1 && nrow(txn) == 1,
  # Knockdown efficiency: RBMS3 is the gene being depleted, so it must drop hard.
  "Paper replication failed: RBMS3 not depleted in the shRBMS3 arm" =
    kd$log2FoldChange < -1 && kd$padj < 1e-5,
  # The headline axis: RBMS3 loss destabilizes TXNIP.
  "Paper replication failed: TXNIP not downregulated upon RBMS3 knockdown" =
    txn$log2FoldChange < 0 && txn$padj < 0.05
)
cat("Anchor 5 PASS - Karner 2026 central result reproduced:\n")
cat("  RBMS3 log2FC =", round(kd$log2FoldChange, 3),
    "(padj", format(kd$padj, digits = 3), ") -- knockdown efficiency\n")
cat("  TXNIP log2FC =", round(txn$log2FoldChange, 3),
    "(padj", format(txn$padj, digits = 3), ") -- RBMS3 stabilizes TXNIP\n")
```

```{r select_231}
# Select MDA-MB-231 replicates and compute mean C-to-T ratio.
# na.rm = FALSE: require both replicates to be non-NA (conservative).
slam_231 <- slam_raw %>%
  dplyr::select(gene_symbol, `MDA-MB-231-1`, `MDA-MB-231-2`) %>%
  mutate(
    mean_ct = rowMeans(cbind(`MDA-MB-231-1`, `MDA-MB-231-2`), na.rm = FALSE)
  )

cat("Rows with NA in either replicate:",
    sum(is.na(slam_231$`MDA-MB-231-1`) | is.na(slam_231$`MDA-MB-231-2`)), "\n")

slam_231 <- slam_231 %>% filter(!is.na(mean_ct))
cat("Rows after NA filter:", nrow(slam_231), "\n")

# Drop saturated artifacts: ratios this high are the exactly-1.0 entries (and their
# replicate means) that do not reproduce between replicates (see header). Not a
# coverage issue -- the authors required >=10 reads. Without this filter 29 of them
# reach the saved feature column, 7 of which fall in the eIF3d-promotes set.
max_plausible_ct <- 0.2
n_artifact <- sum(slam_231$`MDA-MB-231-1` > max_plausible_ct |
                  slam_231$`MDA-MB-231-2` > max_plausible_ct)
slam_231 <- slam_231 %>%
  filter(`MDA-MB-231-1` <= max_plausible_ct, `MDA-MB-231-2` <= max_plausible_ct)
cat("Low-coverage artifacts dropped (ratio >", max_plausible_ct, "):", n_artifact, "\n")
cat("Rows after artifact filter:", nrow(slam_231), "\n")
```

```{r map_gene_ids}
# Map gene symbol -> Ensembl gene ID using the GENCODE v49 annotation only.
#
# This deliberately does NOT use precomputed_most_abundant_tx.rds. That object is
# isoform selection derived from this project's own siCTRL normoxia RNA-seq, so routing
# the mapping through it makes an external dataset depend on our measurements -- the one
# thing this notebook is supposed to avoid. The only project-derived inputs here are the
# positive/negative gene set labels.
#
# Practical bonus: the annotation covers far more symbols, so 7956 of 9464 C-to-T genes
# map (84.1%) instead of 7450 (78.7%) -- 563 genes gained, 57 lost.
# lncRNAs with RP*/LINC names still will not map -- expected and documented.
symbol2gene <- read_tsv(
  here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
  col_names = c("gene_id", "symbol", "biotype", "species", "source"),
  show_col_types = FALSE
) %>%
  filter(species == "human") %>%
  transmute(symbol       = as.character(symbol),
            ensembl_gene = sub("\\..*", "", gene_id)) %>%
  distinct(symbol, ensembl_gene) %>%
  distinct(symbol, .keep_all = TRUE)   # one Ensembl ID per symbol

stopifnot(
  "symbol2gene must be a unique symbol -> gene dictionary" =
    !any(duplicated(symbol2gene$symbol)),
  "annotation mapping looks too small - check the gene_anno file" =
    nrow(symbol2gene) > 50000
)

slam_mapped <- slam_231 %>%
  left_join(symbol2gene, by = c("gene_symbol" = "symbol")) %>%
  filter(!is.na(ensembl_gene))

cat("Genes mapping to Ensembl ID:", nrow(slam_mapped), "/", nrow(slam_231),
    "(", round(100 * nrow(slam_mapped) / nrow(slam_231), 1), "%)\n")
cat("Unmapped (lncRNAs/aliases):", nrow(slam_231) - nrow(slam_mapped), "\n")

# distinct(symbol) above guarantees one Ensembl ID per symbol, but NOT the reverse:
# two symbols can share an Ensembl ID (e.g. RN7SL1), which would emit a duplicate
# ensembl_gene row and silently duplicate feature-matrix rows at the 01i_ join.
# Collapse to one row per gene by averaging.
n_dup <- sum(duplicated(slam_mapped$ensembl_gene))
if (n_dup > 0) {
  cat("Collapsing", n_dup, "duplicate ensembl_gene row(s) by mean:",
      paste(unique(slam_mapped$gene_symbol[
        slam_mapped$ensembl_gene %in%
          slam_mapped$ensembl_gene[duplicated(slam_mapped$ensembl_gene)]]),
        collapse = ", "), "\n")
  slam_mapped <- slam_mapped %>%
    group_by(ensembl_gene) %>%
    summarise(gene_symbol = first(gene_symbol),
              mean_ct     = mean(mean_ct),
              .groups     = "drop")
}
stopifnot("ensembl_gene must be unique after collapse" =
  !any(duplicated(slam_mapped$ensembl_gene)))
```

```{r transform_annotate}
# log2 transform: values in (0,1] → log2_ct in (-∞, 0]
# Higher log2_ct (less negative) = higher C-to-T ratio = more nascent RNA
slam_mapped <- slam_mapped %>%
  mutate(log2_ct = log2(mean_ct))

cat("log2_ct range:", round(min(slam_mapped$log2_ct), 2),
    "to", round(max(slam_mapped$log2_ct), 2), "\n")

# Computational checks (project convention: stopifnot, never cat)
stopifnot(
  "mean_ct must be a ratio in (0, 1]" =
    all(slam_mapped$mean_ct > 0 & slam_mapped$mean_ct <= 1),
  "log2_ct must be finite" = all(is.finite(slam_mapped$log2_ct)),
  "log2_ct must be <= 0 by construction" = all(slam_mapped$log2_ct <= 0),
  "Leakage guard: no TE columns may be present" =
    !any(c("te_lfc", "te_padj") %in% colnames(slam_mapped))
)

# Annotate gene groups
slam_mapped <- slam_mapped %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                            ~ "All other genes"
  ))

cat("\nGene counts per group:\n")
print(slam_mapped %>% count(group))
```

---

## Section 3: Density plot

```{r density_plot}
n_promotes <- sum(slam_mapped$group == "eIF3d hypoxia promotes")
n_negctrl  <- sum(slam_mapped$group == "Negative control")

bg     <- slam_mapped %>% filter(group == "All other genes")
fg_neg <- slam_mapped %>% filter(group == "Negative control")
fg_pro <- slam_mapped %>% filter(group == "eIF3d hypoxia promotes")

grp_medians <- slam_mapped %>%
  filter(group != "All other genes") %>%
  group_by(group) %>%
  summarise(med = median(log2_ct, na.rm = TRUE), .groups = "drop")

grp_colors <- c("eIF3d hypoxia promotes" = "#E69F00", "Negative control" = "black")
grp_labels <- c(
  "eIF3d hypoxia promotes" = paste0("eIF3d hypoxia promotes (n=", n_promotes, ")"),
  "Negative control"       = paste0("Negative control (n=",       n_negctrl,  ")")
)

p <- ggplot() +
  geom_density(data = bg,     aes(x = log2_ct),
               fill = "grey70", color = NA, alpha = 0.6) +
  geom_density(data = fg_neg, aes(x = log2_ct),
               fill = NA, color = "black", linewidth = 0.8) +
  geom_density(data = fg_pro, aes(x = log2_ct),
               fill = NA, color = "#E69F00", linewidth = 0.8) +
  geom_vline(data = grp_medians, aes(xintercept = med, color = group),
             linetype = "dashed", linewidth = 0.7, show.legend = FALSE) +
  # Dummy lines for legend with n labels
  geom_line(data = data.frame(x = NA_real_, y = NA_real_, group = "eIF3d hypoxia promotes"),
            aes(x = x, y = y, color = group), linewidth = 0.8, key_glyph = "path") +
  geom_line(data = data.frame(x = NA_real_, y = NA_real_, group = "Negative control"),
            aes(x = x, y = y, color = group), linewidth = 0.8, key_glyph = "path") +
  scale_color_manual(values = grp_colors, labels = grp_labels, name = NULL) +
  labs(
    title    = "MDA-MB-231 SLAM-seq C-to-T conversion ratio",
    subtitle = paste0("Karner et al. 2026; raw ratio, not expression-normalized\n",
                      "Higher = larger labeled fraction, tracks shorter half-life ",
                      "(rho -0.61 vs Zhu 2024)"),
    x        = "log2(C-to-T conversion ratio)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

# Wilcoxon test
wt <- wilcox.test(fg_pro$log2_ct, fg_neg$log2_ct, alternative = "two.sided")
cat("Wilcoxon promotes vs neg ctrl: W =", wt$statistic,
    ", p =", formatC(wt$p.value, format = "e", digits = 2),
    "| median promotes:", round(median(fg_pro$log2_ct, na.rm = TRUE), 3),
    "| median neg ctrl:", round(median(fg_neg$log2_ct, na.rm = TRUE), 3), "\n")

p <- p + annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.3, size = 3.2,
                  label = paste0("Wilcoxon p = ",
                                 formatC(wt$p.value, format = "e", digits = 2),
                                 "\nn = ", n_promotes, " vs ", n_negctrl))

p
ggsave(here("plots", "25_goodarzi_231_ct_ratio_by_group.pdf"), p, width = 7, height = 5)
cat("Saved: plots/25_goodarzi_231_ct_ratio_by_group.pdf\n")
```

```{r positive_controls}
# Biological spot-check (project convention: assert, do not just report).
# Short-lived immediate-early genes must rank ABOVE long-lived ribosomal-protein /
# housekeeping mRNAs on this ratio. Per-gene counterpart to the cross-dataset Anchor 4
# check in Section 2; like it, a check on the parse rather than a mechanistic claim.
ieg_genes <- c("MYC", "JUN", "FOS", "EGR1", "FOSB", "JUNB", "IER2", "DUSP1")
hk_genes  <- c("ACTB", "GAPDH", "RPL13A", "RPLP0", "TUBB", "EEF1A1", "RPS18")

slam_mapped <- slam_mapped %>% mutate(pctile = 100 * rank(log2_ct) / n())

ieg_med <- median(slam_mapped$log2_ct[slam_mapped$gene_symbol %in% ieg_genes])
hk_med  <- median(slam_mapped$log2_ct[slam_mapped$gene_symbol %in% hk_genes])
myc_pct  <- slam_mapped$pctile[slam_mapped$gene_symbol == "MYC"]
actb_pct <- slam_mapped$pctile[slam_mapped$gene_symbol == "ACTB"]

stopifnot(
  "Short-lived IEGs must have higher C-to-T than housekeeping mRNAs" =
    ieg_med > hk_med,
  "MYC (short-lived) must sit in the top decile of C-to-T" = myc_pct > 90,
  "ACTB (long-lived) must sit in the bottom 40% of C-to-T" = actb_pct < 40
)
cat("Spot-check PASS - median log2_ct: IEG =", round(ieg_med, 2),
    "vs housekeeping =", round(hk_med, 2), "\n")
cat("  MYC percentile =", round(myc_pct, 1),
    "| ACTB percentile =", round(actb_pct, 1), "\n")

# MYC and ACTB are the two genes marked on the figure.
pos_ctrl_genes <- c("MYC", "ACTB")

pos_ctrl_df <- slam_mapped %>%
  filter(gene_symbol %in% pos_ctrl_genes) %>%
  dplyr::select(gene_symbol, mean_ct, log2_ct)

cat("\nPositive control gene values:\n")
print(pos_ctrl_df)

if (nrow(pos_ctrl_df) == 0) {
  cat("WARNING: neither MYC nor ACTB found in slam_mapped — check gene_symbol column.\n")
} else {
  # Overlay on density: vertical lines + labels
  p_ctrl <- p +
    geom_vline(data = pos_ctrl_df,
               aes(xintercept = log2_ct),
               color = "firebrick", linetype = "solid", linewidth = 0.8) +
    geom_label(data = pos_ctrl_df,
               aes(x = log2_ct, y = 0, label = gene_symbol),
               color = "firebrick", fill = "white", size = 3.5,
               vjust = -0.3, hjust = 0.5, label.size = 0.3) +
    labs(title = "MDA-MB-231 C-to-T conversion ratio, positive controls marked",
         subtitle = "Red lines = MYC (short-lived) and ACTB (long-lived); Karner et al. 2026")

  p_ctrl
  ggsave(here("plots", "25_goodarzi_231_ct_ratio_positive_controls.pdf"),
         p_ctrl, width = 6, height = 5)
  cat("Saved: plots/25_goodarzi_231_ct_ratio_positive_controls.pdf\n")

  # Report percentile rank of each control gene
  cat("\nPercentile rank of positive controls (among all mapped genes):\n")
  for (i in seq_len(nrow(pos_ctrl_df))) {
    gene  <- pos_ctrl_df$gene_symbol[i]
    val   <- pos_ctrl_df$log2_ct[i]
    pctile <- mean(slam_mapped$log2_ct <= val, na.rm = TRUE) * 100
    cat(" ", gene, ": log2_ct =", round(val, 3),
        "| percentile =", round(pctile, 1), "%\n")
  }
}
```

---

## Section 4: Save output

```{r save_output}
# `group` is deliberately NOT saved. It is derived from the eIF3d TE gene set, i.e.
# from the RF target labels; persisting it next to a feature column invites a future
# join that leaks the label into the feature matrix. Consumers (01i_, 43_) only ever
# take ensembl_gene + log2_ct.
ct_out <- slam_mapped %>%
  dplyr::select(ensembl_gene, symbol = gene_symbol, mean_ct, log2_ct)

stopifnot(
  "Leakage guard: label-derived columns must not be saved" =
    !any(c("group", "te_lfc", "te_padj") %in% colnames(ct_out)),
  "ensembl_gene must be unique in saved output" =
    !any(duplicated(ct_out$ensembl_gene))
)

saveRDS(ct_out, here("output", "predictive_modeling", "goodarzi_231_ct_ratios.rds"))
cat("Saved: output/predictive_modeling/goodarzi_231_ct_ratios.rds\n")
cat("Rows saved:", nrow(ct_out), "\n")
```

---

## Section 5: All six breast cancer cell lines

Karner et al. profiled six lines across three subtypes: luminal (ZR-75-1, MCF7),
HER2-positive (MDA-MB-453), and TNBC (HCC1806, MDA-MB-231, HCC38). Sections 2-4 use
MDA-MB-231 only; this section carries the same gene-group comparison across all six.

Note the deposited column header for MDA-MB-453 is misspelled **`MDA-MD-453`** (D for B).
Matching on the corrected spelling silently drops that line, so the typo is preserved
here and in the Anchor 1 check deliberately.

```{r all_lines_build}
line_meta <- tribble(
  ~line_key,     ~line_label,  ~subtype,
  "ZR-75",       "ZR-75-1",    "Luminal",
  "MCF7",        "MCF7",       "Luminal",
  "MDA-MD-453",  "MDA-MB-453", "HER2+",
  "HCC1806",     "HCC1806",    "TNBC",
  "MDA-MB-231",  "MDA-MB-231", "TNBC",
  "HCC38",       "HCC38",      "TNBC"
)

# Per line: require both replicates, apply the same low-coverage artifact filter.
all_lines <- map_dfr(seq_len(nrow(line_meta)), function(i) {
  k <- line_meta$line_key[i]
  cols <- grep(paste0("^", k, "-[12]$"), colnames(slam_raw), value = TRUE)
  stopifnot("each cell line must have exactly 2 replicate columns" = length(cols) == 2)
  tibble(
    gene_symbol = slam_raw$gene_symbol,
    r1 = slam_raw[[cols[1]]],
    r2 = slam_raw[[cols[2]]]
  ) %>%
    filter(!is.na(r1), !is.na(r2), r1 <= max_plausible_ct, r2 <= max_plausible_ct) %>%
    transmute(gene_symbol,
              log2_ct    = log2((r1 + r2) / 2),
              line_label = line_meta$line_label[i],
              subtype    = line_meta$subtype[i])
}) %>%
  left_join(symbol2gene, by = c("gene_symbol" = "symbol")) %>%
  filter(!is.na(ensembl_gene)) %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  )) %>%
  mutate(line_label = factor(line_label, levels = line_meta$line_label))

stopifnot(
  "all six cell lines must be represented" =
    dplyr::n_distinct(all_lines$line_label) == 6,
  "log2_ct must be finite and <= 0" =
    all(is.finite(all_lines$log2_ct) & all_lines$log2_ct <= 0)
)

cat("Genes per cell line after filtering:\n")
print(all_lines %>% dplyr::count(line_label, subtype))
```

### 5a: Replicate reproducibility across all six lines (paper fig. S1A)

Karner et al. report that "these decay rate measurements were highly reproducible across
biological replicates (average Pearson R ~ 0.9; fig. S1A)."

**That 0.9 is not directly assertable on this file, and it would be wrong to try.** The
paper's R ~ 0.9 describes its *derived decay-rate* estimates, which combine the labeling
ratio with expression; expression is far more reproducible than the raw conversion ratio,
so the derived quantity is smoother. GSE301426's supplementary file contains only the raw
ratio, which reproduces at **Pearson ~ 0.70-0.78 per line (mean ~0.74)**. Asserting 0.9
here would fail on correct data.

So the check below asserts what this file can actually support: every line clears a floor,
and no line is an outlier against the others. The paper's figure is the motivation; the
threshold is calibrated to the deposited quantity.

```{r replicate_reproducibility}
rep_stats <- map_dfr(seq_len(nrow(line_meta)), function(i) {
  k    <- line_meta$line_key[i]
  cols <- grep(paste0("^", k, "-[12]$"), colnames(slam_raw), value = TRUE)
  r1   <- slam_raw[[cols[1]]]
  r2   <- slam_raw[[cols[2]]]

  keep_all <- !is.na(r1) & !is.na(r2)
  keep     <- keep_all & r1 <= max_plausible_ct & r2 <= max_plausible_ct

  tibble(
    line_label   = line_meta$line_label[i],
    subtype      = line_meta$subtype[i],
    n            = sum(keep),
    pearson      = cor(r1[keep], r2[keep]),
    pearson_log2 = cor(log2(r1[keep]), log2(r2[keep])),
    spearman     = cor(r1[keep], r2[keep], method = "spearman"),
    # Same statistic without the artifact filter, to show the filter is load-bearing.
    pearson_unfiltered = cor(r1[keep_all], r2[keep_all])
  )
})

cat("Replicate reproducibility per cell line (artifact-filtered):\n")
print(rep_stats %>% mutate(across(where(is.numeric), ~round(., 3))))
cat("\nMean Pearson across lines:", round(mean(rep_stats$pearson), 3),
    "| paper reports ~0.9 for its derived decay rates (fig. S1A)\n")
cat("Mean Pearson WITHOUT the artifact filter:",
    round(mean(rep_stats$pearson_unfiltered), 3),
    "- the ratio == 1 low-coverage calls destroy replicate agreement\n")

stopifnot(
  "fig S1A: every cell line must clear the replicate-reproducibility floor" =
    all(rep_stats$pearson > 0.6),
  "fig S1A: rank-based replicate agreement must also hold in every line" =
    all(rep_stats$spearman > 0.65),
  "fig S1A: mean replicate Pearson across the six lines is too low" =
    mean(rep_stats$pearson) > 0.65,
  # No single line may be a reproducibility outlier - catches one bad sample pair that
  # a mean-only check would hide.
  "fig S1A: one cell line is far less reproducible than the others" =
    min(rep_stats$pearson) > mean(rep_stats$pearson) - 0.15
)
cat("\nfig S1A PASS - all six lines reproducible, range",
    paste(round(range(rep_stats$pearson), 3), collapse = " to "), "\n")
```

```{r all_lines_stats}
# Per-panel Wilcoxon (promotes vs negative control) plus group sizes.
panel_stats <- all_lines %>%
  group_by(line_label, subtype) %>%
  summarise(
    n_pro   = sum(group == "eIF3d hypoxia promotes"),
    n_neg   = sum(group == "Negative control"),
    n_other = sum(group == "All other genes"),
    med_pro = median(log2_ct[group == "eIF3d hypoxia promotes"]),
    med_neg = median(log2_ct[group == "Negative control"]),
    p_val   = wilcox.test(log2_ct[group == "eIF3d hypoxia promotes"],
                          log2_ct[group == "Negative control"])$p.value,
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Wilcoxon p = ", formatC(p_val, format = "e", digits = 2),
                        "\nn promotes = ", n_pro,
                        "\nn neg ctrl = ", n_neg,
                        "\nn other = ", n_other))

cat("Per-cell-line summary:\n")
print(panel_stats %>% dplyr::select(-label) %>%
        mutate(across(where(is.numeric), ~signif(., 3))))
```

```{r all_lines_density_plot, fig.width=12, fig.height=7}
grp_levels <- c("eIF3d hypoxia promotes", "Negative control", "All other genes")
grp_cols   <- c("eIF3d hypoxia promotes" = "#E69F00",
                "Negative control"       = "black",
                "All other genes"        = "grey70")

plot_df <- all_lines %>% mutate(group = factor(group, levels = grp_levels))

med_df <- plot_df %>%
  group_by(line_label, group) %>%
  summarise(med = median(log2_ct), .groups = "drop")

p_all <- ggplot(plot_df, aes(x = log2_ct, color = group)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = med_df, aes(xintercept = med, color = group),
             linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
  # Convention places this at x = Inf, but in these panels the right side is where the
  # distributions sit and the label collides with the curves. The far-left tail is empty
  # in all six, so the annotation goes there instead.
  geom_text(data = panel_stats, aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.08, vjust = 1.3, size = 2.7) +
  scale_color_manual(values = grp_cols, name = NULL) +
  facet_wrap(~ line_label, ncol = 3, scales = "free_y") +
  labs(
    title    = "SLAM-seq C-to-T conversion ratio by eIF3d gene group, six breast cancer cell lines",
    subtitle = paste0("Karner et al. 2026 (GSE301426); dashed lines = group medians\n",
                      "Higher = larger labeled fraction, which tracks shorter measured ",
                      "half-life (rho = -0.61 vs Zhu 2024)"),
    x        = "log2(C-to-T conversion ratio)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p_all
ggsave(here("plots", "25_karner_ct_ratio_all_six_lines.pdf"), p_all, width = 12, height = 7)
cat("Saved: plots/25_karner_ct_ratio_all_six_lines.pdf\n")
```

### 5b: Decay rate (C-to-T normalized by expression)

The authors' rule, quoted in the header: `decay rate = transcription rate / steady-state
expression`, i.e. `log2_decay = log2_ct - log2(expression)`. This is the quantity behind
Fig. 1D, and it is not deposited -- it has to be rebuilt, and rebuilding it needs baseline
expression **per cell line**.

**Karner deposits no expression data at all.** GSE301426 (the SLAM experiment) has none,
so the denominator the authors actually used is unavailable. The only expression anywhere
in the SuperSeries is `baseMean` in GSE301427 -- and that comes from the **shRBMS3
knockdown experiment**, pooling shCTRL with shRBMS3 samples. It is therefore:

- a different experiment from the one the C-to-T ratios come from,
- MDA-MB-231 only, so this section covers **one cell line, not six**,
- biased for RBMS3 targets, since knockdown samples are averaged in.

Using only Karner-derived inputs (this notebook takes nothing from this project except the
positive/negative gene set labels), that is the sole option. `expr_sources` is kept as a
lookup table so a second denominator can be added later without touching the code.

```{r decay_rate_build}
# Only Karner-derived expression: baseMean from the GSE301427 shRBMS3 DESeq2 table.
# NOTE: deliberately does not use this project's siCTRL RNA-seq. That would be a cleaner,
# unperturbed baseline, but it would fold our own measurements into an external dataset.
expr_karner <- tibble(ensembl_gene = sub("\\..*", "", rbms3$row), expr = rbms3$baseMean) %>%
  filter(expr > 1)

expr_sources <- list(
  "Denominator: Karner baseMean (GSE301427, shRBMS3 experiment)" = expr_karner
)

stopifnot(
  "expression must be positive before log2" =
    all(map_lgl(expr_sources, ~ all(.x$expr > 0))),
  "expression sources must overlap the C-to-T genes" =
    all(map_int(expr_sources, ~ sum(.x$ensembl_gene %in% slam_mapped$ensembl_gene)) > 3000)
)

# MDA-MB-231 C-to-T, already artifact-filtered and gene-mapped in Section 2.
decay_df <- map_dfr(names(expr_sources), function(nm) {
  slam_mapped %>%
    dplyr::select(ensembl_gene, gene_symbol, log2_ct) %>%
    inner_join(expr_sources[[nm]], by = "ensembl_gene") %>%
    mutate(log2_decay = log2_ct - log2(expr), source = nm)
}) %>%
  mutate(group = factor(case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  ), levels = grp_levels),
  source = factor(source, levels = names(expr_sources)))

stopifnot("log2_decay must be finite" = all(is.finite(decay_df$log2_decay)))

decay_stats <- decay_df %>%
  group_by(source) %>%
  summarise(
    n_pro   = sum(group == "eIF3d hypoxia promotes"),
    n_neg   = sum(group == "Negative control"),
    n_other = sum(group == "All other genes"),
    med_pro = median(log2_decay[group == "eIF3d hypoxia promotes"]),
    med_neg = median(log2_decay[group == "Negative control"]),
    p_val   = wilcox.test(log2_decay[group == "eIF3d hypoxia promotes"],
                          log2_decay[group == "Negative control"])$p.value,
    .groups = "drop"
  ) %>%
  mutate(label = paste0("Wilcoxon p = ", formatC(p_val, format = "e", digits = 2),
                        "\nn promotes = ", n_pro,
                        "\nn neg ctrl = ", n_neg,
                        "\nn other = ", n_other))

cat("Decay rate, MDA-MB-231, by expression denominator:\n")
print(decay_stats %>% dplyr::select(-label) %>%
        mutate(across(where(is.numeric), ~signif(., 3))))
```

```{r decay_rate_plot, fig.width=6.5, fig.height=5}
decay_med <- decay_df %>% group_by(source, group) %>%
  summarise(med = median(log2_decay), .groups = "drop")

p_decay <- ggplot(decay_df, aes(x = log2_decay, color = group)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = decay_med, aes(xintercept = med, color = group),
             linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
  geom_text(data = decay_stats, aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.08, vjust = 1.3, size = 2.7) +
  scale_color_manual(values = grp_cols, name = NULL) +
  facet_wrap(~ source, ncol = 2, scales = "free") +
  labs(
    title    = "Estimated mRNA decay rate by eIF3d gene group, MDA-MB-231",
    subtitle = "log2(decay) = log2(C-to-T ratio) - log2(expression), the rule stated in GSE301426 processing\nHigher = faster decay; dashed lines = group medians",
    x        = "log2(estimated decay rate), arbitrary units",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p_decay
ggsave(here("plots", "25_karner_decay_rate_mda231.pdf"), p_decay, width = 6.5, height = 5)
cat("Saved: plots/25_karner_decay_rate_mda231.pdf\n")
```

```{r decay_vs_raw_validation}
# Does expression-normalization actually make the feature better? Test both against
# independently published control half-lives. Reported, not asserted: this is a
# comparison between two candidate features, not a correctness check on either.
hl_ref <- list(
  "Zhu 2024 (sgNT)" = "cnot3_slamseq_halflives.rds",
  "Hia 2026 (ctrl)" = "dhx29_slamseq_halflives.rds"
)
for (nm in names(expr_sources)) {
  d0 <- decay_df %>% filter(source == nm)
  for (hn in names(hl_ref)) {
    fp <- here("output", "predictive_modeling", hl_ref[[hn]])
    if (!file.exists(fp)) next
    hl <- readRDS(fp)
    hl_df <- if ("sgNT_halflife" %in% names(hl)) {
      tibble(ensembl_gene = hl$ensembl_gene, halflife = hl$sgNT_halflife)
    } else {
      tibble(ensembl_gene = hl$ensembl_gene, halflife = hl$Ctrl_Half_life)
    }
    dd <- d0 %>% inner_join(hl_df, by = "ensembl_gene") %>%
      filter(is.finite(halflife), halflife > 0)
    cat(sprintf("%-42s vs %-16s n=%4d  raw C/T rho=%+.3f  decay rho=%+.3f\n",
        nm, hn, nrow(dd),
        cor(dd$log2_ct,    dd$halflife, method = "spearman"),
        cor(dd$log2_decay, dd$halflife, method = "spearman")))
  }
}
cat("\nNote: normalizing by expression makes the correlation with published half-life\n",
    "WEAKER, not stronger, under both denominators. See the confound check below.\n", sep = "")
```

**Read the decay panel with care -- the separation is mostly the denominator.**

The decay plot separates the groups far more dramatically than the raw C/T plot
(p ~ 1e-64 vs 1e-12), which looks like a win. It is not. eIF3d-promoted genes are simply
**more highly expressed** than the negative controls, and expression is the denominator,
so it is subtracted straight into the statistic.

```{r decay_expression_confound}
expr_grouped <- expr_karner %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  )) %>%
  filter(group != "All other genes")

e_pro <- log2(expr_grouped$expr[expr_grouped$group == "eIF3d hypoxia promotes"])
e_neg <- log2(expr_grouped$expr[expr_grouped$group == "Negative control"])
expr_gap <- median(e_pro) - median(e_neg)

d_proj    <- decay_df %>% filter(source == names(expr_sources)[1])
decay_gap <- median(d_proj$log2_decay[d_proj$group == "eIF3d hypoxia promotes"]) -
             median(d_proj$log2_decay[d_proj$group == "Negative control"])
ct_gap    <- median(d_proj$log2_ct[d_proj$group == "eIF3d hypoxia promotes"]) -
             median(d_proj$log2_ct[d_proj$group == "Negative control"])

cat("Median group difference (promotes - neg ctrl), log2 units:\n")
cat("  expression only :", round(expr_gap, 3),
    " (Wilcoxon p =", format.pval(wilcox.test(e_pro, e_neg)$p.value, digits = 3), ")\n")
cat("  raw C-to-T      :", round(ct_gap, 3), "\n")
cat("  derived decay   :", round(decay_gap, 3), "\n")
cat("\n  -> the expression term alone (", round(-expr_gap, 3),
    ") exceeds the entire decay-space gap (", round(decay_gap, 3), ").\n", sep = "")

# Guard: fail loudly if anyone later reports the decay separation as a stability result
# without accounting for the expression difference driving it.
stopifnot(
  "Expression confound check is misconfigured - groups must both be populated" =
    length(e_pro) > 100 && length(e_neg) > 100
)
if (abs(expr_gap) > abs(decay_gap) / 2) {
  cat("\nWARNING: the promotes/neg-ctrl expression difference accounts for the majority\n")
  cat("of the decay-rate separation. Do NOT report the decay panel as evidence that\n")
  cat("eIF3d targets are more stable. Use the raw C-to-T panel, or an expression-matched\n")
  cat("negative control set, for any stability claim.\n")
}
```

Consequences, and why the raw ratio is still the feature that ships:

1. The negative control set is defined by TE criteria only, with no expression matching,
   so it sits `r round(median(log2(expr_karner$expr[expr_karner$ensembl_gene %in% promotes_ids])) - median(log2(expr_karner$expr[expr_karner$ensembl_gene %in% neg_ctrl_ids])), 2)`
   log2 units below the promotes set in abundance.
2. Dividing by expression therefore *adds* a group-discriminating axis that is not decay.
3. Consistent with that, the decay version correlates **less** well with independently
   measured half-lives than the raw ratio does (above).

So `karner2026_mda231_log2_ct` stays as the raw ratio in the feature matrix. Rebuilding
the paper's decay rate is documented here and reproducible, but on this gene-set design it
is a worse feature, not a better one. An expression-matched negative control set would be
the way to make the decay comparison meaningful.

### 5c: The knockdown Karner provides

The only perturbation in this study is **shRBMS3 in MDA-MB-231** (GSE301427); there is no
knockdown for the other five lines. It is also an RNA-seq expression log2FC, not a
half-life ratio, so it is not directly comparable to the DHX29 KO half-life panel: a
shift here means the transcript changed in abundance upon RBMS3 loss, which the paper
attributes to stability but does not measure as a decay rate in this file.

```{r rbms3_kd_plot, fig.width=6.5, fig.height=5}
rbms3_grouped <- rbms3 %>%
  dplyr::select(gene_symbol = GENE, log2FoldChange) %>%
  filter(is.finite(log2FoldChange)) %>%
  left_join(symbol2gene, by = c("gene_symbol" = "symbol")) %>%
  filter(!is.na(ensembl_gene)) %>%
  mutate(group = factor(case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  ), levels = grp_levels))

kd_med <- rbms3_grouped %>% group_by(group) %>%
  summarise(med = median(log2FoldChange), .groups = "drop")

kd_pro <- rbms3_grouped$log2FoldChange[rbms3_grouped$group == "eIF3d hypoxia promotes"]
kd_neg <- rbms3_grouped$log2FoldChange[rbms3_grouped$group == "Negative control"]
kd_wt  <- wilcox.test(kd_pro, kd_neg)

cat("RBMS3 KD, promotes vs neg ctrl: p =", signif(kd_wt$p.value, 3),
    "| median promotes =", round(median(kd_pro), 4),
    "| median neg ctrl =", round(median(kd_neg), 4), "\n")

p_kd <- ggplot(rbms3_grouped, aes(x = log2FoldChange, color = group)) +
  geom_density(linewidth = 0.8) +
  geom_vline(data = kd_med, aes(xintercept = med, color = group),
             linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3,
           label = paste0("Wilcoxon p = ", signif(kd_wt$p.value, 3),
                          "\nn promotes = ", length(kd_pro),
                          "\nn neg ctrl = ", length(kd_neg),
                          "\nn other = ", sum(rbms3_grouped$group == "All other genes"))) +
  scale_color_manual(values = grp_cols, name = NULL) +
  coord_cartesian(xlim = quantile(rbms3_grouped$log2FoldChange, c(0.005, 0.995))) +
  labs(
    title    = "Expression change upon RBMS3 knockdown, MDA-MB-231",
    subtitle = "Karner et al. 2026 (GSE301427)\nNegative = down upon RBMS3 loss; dashed = median",
    x        = "log2FC (shRBMS3 / shCTRL)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_kd
ggsave(here("plots", "25_karner_rbms3_kd_by_group.pdf"), p_kd, width = 6.5, height = 5)
cat("Saved: plots/25_karner_rbms3_kd_by_group.pdf\n")
```

---

## Session info

```{r session_info}
sessionInfo()
```
