---
title: "Notebook 27: TXN Reporter — Codon & Amino Acid Usage"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(iCodon)
```

# Overview

Analyzes codon and amino acid composition of the TXN coding sequence (most abundant
transcript, consistent with the RF feature extraction pipeline) in the context of
eIF3d-mediated translational regulation.

# Configuration

```{r config}
# --- Change this to analyze a different gene ---
query_gene <- "TXN"
# -----------------------------------------------
```

# Section 1: Extract CDS for Most Abundant Transcript

```{r load-transcript}
# Load precomputed most-abundant transcript mapping (same as used in feature extraction)
most_abundant_tx <- readRDS(
  here("output", "predictive_modeling", "precomputed_most_abundant_tx.rds")
)

tx_row <- most_abundant_tx %>% filter(symbol == query_gene)
if (nrow(tx_row) == 0) stop("Gene not found in most_abundant_tx: ", query_gene)

tx_id_clean <- tx_row$transcript_id_clean
gene_id     <- tx_row$gene_id_clean

cat("Gene:         ", query_gene, "\n")
cat("Gene ID:      ", gene_id, "\n")
cat("Transcript:   ", tx_id_clean, "\n")
cat("Total abundance (sum TPM):", round(tx_row$total_abundance, 1), "\n")
```

```{r extract-cds}
# Load TxDb and genome (same as 01_feature_extraction.Rmd)
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

# Extract all CDS by transcript
cds_all <- cdsBy(txdb, by = "tx", use.names = TRUE)

# Match TXN transcript (strip version from TxDb IDs)
cds_names_clean <- str_remove(names(cds_all), "\\..*")
tx_match        <- which(cds_names_clean == tx_id_clean)

if (length(tx_match) == 0) stop("Transcript not found in TxDb: ", tx_id_clean)
cat("Matched TxDb entry:", names(cds_all)[tx_match], "\n")

# Extract CDS sequence
cds_seq <- extractTranscriptSeqs(genome, cds_all[tx_match])[[1]]

cds_len     <- length(cds_seq)
n_codons    <- cds_len / 3

cat("\nCDS length:  ", cds_len, "nt\n")
cat("Codons:      ", floor(n_codons), "(includes stop)\n")
cat("Divisible by 3:", cds_len %% 3 == 0, "\n")
if (cds_len %% 3 != 0) warning("CDS length not divisible by 3!")

cat("\nCDS sequence (first 60 nt):", as.character(subseq(cds_seq, 1, min(60, cds_len))), "...\n")
```

# Section 2: Amino Acid Composition

```{r translate}
# Translate CDS
aa_seq    <- translate(cds_seq)
aa_string <- as.character(aa_seq)[[1]]          # force to plain string
aa_chars  <- strsplit(aa_string, "")[[1]]
aa_chars  <- aa_chars[aa_chars != "*"]

cat("Protein length:", length(aa_chars), "aa (excl. stop)\n")
cat("Sequence:", aa_string, "\n")

aa_counts <- tibble(aa = aa_chars) %>%
  dplyr::count(aa, name = "count") %>%
  mutate(pct = round(100 * count / sum(count), 1))

# Biochemical property groupings
aa_properties <- c(
  # Nonpolar/hydrophobic
  A = "Nonpolar", V = "Nonpolar", I = "Nonpolar", L = "Nonpolar",
  M = "Nonpolar", F = "Nonpolar", W = "Nonpolar", P = "Nonpolar",
  # Polar uncharged
  G = "Polar",    S = "Polar",    T = "Polar",     C = "Polar",
  Y = "Polar",    N = "Polar",    Q = "Polar",
  # Positively charged
  K = "Positive", R = "Positive", H = "Positive",
  # Negatively charged
  D = "Negative", E = "Negative"
)

property_colors <- c(
  Nonpolar = "grey60",
  Polar    = "#009E73",
  Positive = "#0072B2",
  Negative = "#D55E00"
)

aa_counts <- aa_counts %>%
  mutate(
    property = aa_properties[aa],
    aa = fct_reorder(aa, count, .desc = TRUE)
  )

cat("\nAmino acid composition:\n")
aa_counts %>% arrange(desc(count)) %>% print(n = 20)
```

```{r aa-plot, fig.width=9, fig.height=5}
ggplot(aa_counts, aes(x = aa, y = count, fill = property)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(pct, "%")), vjust = -0.4, size = 3) +
  scale_fill_manual(values = property_colors, name = "Property") +
  labs(
    x     = "Amino acid",
    y     = "Count",
    title = paste0(query_gene, " — Amino acid composition"),
    subtitle = paste0(length(aa_chars), " residues (", cds_len, " nt CDS, most abundant transcript: ", tx_id_clean, ")")
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here(paste0("output/27_", query_gene, "_aa_composition.pdf")), width = 9, height = 5)
```

# Section 3: Codon Usage Table

```{r codon-usage}
# Build codon table for TXN
cds_ss  <- DNAStringSet(cds_seq)
names(cds_ss) <- query_gene
ct_txn  <- codonTable(cds_ss)
counts_64 <- codonCounts(ct_txn)[1, ]  # named vector: codon -> count

# Standard genetic code: codon -> amino acid
gc <- getGeneticCode("SGC0")   # standard code, DNA codons

# Build codon usage data frame
codon_df <- tibble(
  codon = names(counts_64),
  count = as.integer(counts_64)
) %>%
  mutate(
    aa = gc[codon],
    aa = if_else(aa == "*", "Stop", aa)
  ) %>%
  filter(aa != "Stop") %>%
  group_by(aa) %>%
  mutate(
    pct_within_aa = round(100 * count / sum(count), 1),
    is_dominant   = count == max(count) & count > 0
  ) %>%
  ungroup() %>%
  arrange(aa, desc(count))

cat("Codon usage table for", query_gene, ":\n")
codon_df %>% filter(count > 0) %>% print(n = 60)
```

```{r codon-usage-plot, fig.width=11, fig.height=9}
# Heatmap-style: amino acid on y, codons on x, fill = pct within AA
# Only show AAs that appear in TXN
used_aas <- codon_df %>% filter(count > 0) %>% pull(aa) %>% unique()

plot_df <- codon_df %>%
  filter(aa %in% used_aas) %>%
  mutate(
    aa = factor(aa),
    label = if_else(count > 0, as.character(count), "")
  )

ggplot(plot_df, aes(x = codon, y = fct_rev(aa), fill = pct_within_aa)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = label), size = 2.8) +
  geom_tile(data = filter(plot_df, is_dominant & count > 0),
            color = "#E69F00", fill = NA, linewidth = 1.1) +
  scale_fill_gradient(low = "white", high = "#0072B2",
                      name = "% within AA", na.value = "grey95") +
  labs(
    x     = "Codon",
    y     = "Amino acid",
    title = paste0(query_gene, " — Codon usage (% within synonymous family)"),
    subtitle = "Orange border = most-used codon per AA in TXN; numbers = raw counts"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, family = "mono", size = 8),
    panel.grid   = element_blank(),
    legend.position = "right"
  )

ggsave(here(paste0("output/27_", query_gene, "_codon_usage_heatmap.pdf")), width = 11, height = 9)
```

# Section 4: Position-Specific Codon Optimality (tAI)

```{r tai-weights}
# Build per-codon tAI weights using the exact same logic as 01g
trna_gcn <- read_csv(here("accessories", "human", "human_trna_gcn.csv"),
                     comment = "#", show_col_types = FALSE)
cat("tRNA GCN table:", nrow(trna_gcn), "rows\n")

# All 64 codons
all_codons  <- names(getGeneticCode("SGC0"))
sense_codons <- all_codons[getGeneticCode("SGC0") != "*"]

# Decode anticodon → codons recognized + wobble penalties (eukaryotic rules)
decode_anticodon <- function(ac) {
  x1 <- substr(ac, 1, 1); x2 <- substr(ac, 2, 2); x3 <- substr(ac, 3, 3)
  wc <- c(A = "T", T = "A", G = "C", C = "G")
  y1 <- wc[x3]; y2 <- wc[x2]
  if (x1 == "A") {
    y3 <- c("T", "C", "A"); svs <- c(0, 0, 0)
  } else if (x1 == "G") {
    y3 <- c("C", "T");      svs <- c(0, 0.41)
  } else if (x1 == "C") {
    y3 <- "G";              svs <- 0
  } else {  # T
    y3 <- "A";              svs <- 0
  }
  data.frame(codon = paste0(y1, y2, y3), s = svs, stringsAsFactors = FALSE)
}

ws_tai <- setNames(rep(0, 64), all_codons)

for (i in seq_len(nrow(trna_gcn))) {
  ac  <- trna_gcn$anticodon[i]
  aa  <- trna_gcn$amino_acid[i]
  gcn <- trna_gcn$gcn[i]
  decoded <- decode_anticodon(ac)
  for (j in seq_len(nrow(decoded))) {
    codon <- decoded$codon[j]; s_val <- decoded$s[j]
    if (!codon %in% all_codons) next
    if (getGeneticCode("SGC0")[codon] != aa) next
    ws_tai[codon] <- ws_tai[codon] + (1 - s_val) * gcn
  }
}

w_tai <- ws_tai / max(ws_tai[sense_codons])
cat("w_tai built. Range:", round(range(w_tai[sense_codons]), 4), "\n")
```

```{r positional-tai, fig.width=10, fig.height=5}
# Split CDS into codons (exclude stop)
cds_char   <- as.character(cds_seq)
codon_pos  <- seq(1, cds_len - 2, by = 3)
codons_vec <- sapply(codon_pos, function(i) substr(cds_char, i, i + 2))
codons_vec <- codons_vec[codons_vec %in% sense_codons]  # drop stop if any

positional_df <- tibble(
  position = seq_along(codons_vec),
  codon    = codons_vec,
  tai_w    = w_tai[codons_vec],
  aa       = gc[codons_vec]
)

cat("Codons analyzed:", nrow(positional_df), "\n")
cat("Mean tAI weight:", round(mean(positional_df$tai_w, na.rm = TRUE), 4), "\n")

ggplot(positional_df, aes(x = position, y = tai_w)) +
  geom_point(alpha = 0.4, size = 1.5, color = "grey50") +
  geom_smooth(method = "loess", span = 0.2, se = TRUE,
              color = "#0072B2", fill = "#0072B2", alpha = 0.15, linewidth = 1) +
  geom_hline(yintercept = mean(positional_df$tai_w, na.rm = TRUE),
             linetype = "dashed", color = "grey40") +
  annotate("text", x = nrow(positional_df) * 0.98,
           y = mean(positional_df$tai_w, na.rm = TRUE) + 0.02,
           label = paste0("mean = ", round(mean(positional_df$tai_w, na.rm = TRUE), 3)),
           hjust = 1, size = 3.5, color = "grey40") +
  scale_x_continuous(
    breaks = c(1, round(nrow(positional_df) / 2), nrow(positional_df)),
    labels = c("5' start", "mid", "3' end")
  ) +
  labs(
    x     = "Codon position",
    y     = "tAI weight",
    title = paste0(query_gene, " — Position-specific tAI weights"),
    subtitle = "Each point = one codon; blue line = loess smooth (span 0.2); dashed = CDS mean"
  ) +
  theme_classic(base_size = 12)

ggsave(here(paste0("output/27_", query_gene, "_positional_tai.pdf")), width = 10, height = 5)
```

# Section 5: Genome-Wide Context (TXN vs Background)

```{r load-background}
# Load feature matrix with CAI, tAI, CSC, FOP
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Feature matrix:", nrow(fm), "genes\n")

# Load eIF3d-promoted gene set
gs_3d <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                   show_col_types = FALSE)

# TXN gene ID
txn_gene_id <- gene_id  # from Section 1

# Annotate
score_df <- fm %>%
  dplyr::select(gene_id_clean, symbol, cai, tai, csc, fop) %>%
  mutate(
    group = case_when(
      gene_id_clean %in% gs_3d$ensembl_gene ~ "eIF3d promoted",
      TRUE                                  ~ "All other genes"
    ),
    is_txn = gene_id_clean == txn_gene_id
  )

txn_scores <- score_df %>% filter(is_txn)
cat("\n", query_gene, "scores:\n")
txn_scores %>% dplyr::select(symbol, cai, tai, csc, fop) %>%
  mutate(across(c(cai, tai, csc, fop), ~ round(.x, 4))) %>%
  print()

# Percentile of TXN within eIF3d-promoted genes
eif3d_scores <- score_df %>% filter(group == "eIF3d promoted")
cat("\nPercentile of", query_gene, "within eIF3d-promoted genes:\n")
for (s in c("cai", "tai", "csc", "fop")) {
  pct <- round(100 * mean(eif3d_scores[[s]] <= txn_scores[[s]][1], na.rm = TRUE), 1)
  cat(sprintf("  %-4s: %.1f%%\n", s, pct))
}
```

```{r context-density-plots, fig.width=11, fig.height=9}
scores_long <- score_df %>%
  filter(!is_txn) %>%
  pivot_longer(c(cai, tai, csc, fop), names_to = "score", values_to = "value")

txn_long <- txn_scores %>%
  pivot_longer(c(cai, tai, csc, fop), names_to = "score", values_to = "value")

score_labels <- c(
  cai = "CAI — Codon Adaptive Index",
  tai = "tAI — tRNA Adaptation Index",
  csc = "CSC — Codon Stabilisation Coefficient",
  fop = "FOP — Frequency of Optimal Codons"
)

ggplot(scores_long, aes(x = value, color = group, fill = group)) +
  geom_density(alpha = 0.2, linewidth = 0.7) +
  geom_vline(data = txn_long,
             aes(xintercept = value),
             color = "#CC79A7", linewidth = 1.2, linetype = "solid") +
  geom_text(data = txn_long,
            aes(x = value, y = Inf, label = paste0(" ", query_gene)),
            color = "#CC79A7", angle = 90, hjust = 1.2, vjust = -0.3,
            size = 3.5, inherit.aes = FALSE) +
  scale_color_manual(values = c("All other genes" = "grey50", "eIF3d promoted" = "#E69F00"),
                     name = NULL) +
  scale_fill_manual(values  = c("All other genes" = "grey50", "eIF3d promoted" = "#E69F00"),
                    name = NULL) +
  facet_wrap(~ score, scales = "free", labeller = labeller(score = score_labels)) +
  labs(
    x     = "Score",
    y     = "Density",
    title = paste0(query_gene, " codon optimality scores in genome-wide context"),
    subtitle = "Orange = eIF3d-promoted (1hr hypoxia, LFC>0.5) | Pink line = TXN"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", strip.background = element_blank())

ggsave(here(paste0("output/27_", query_gene, "_codon_scores_context.pdf")), width = 11, height = 9)
```

# Session Info

```{r session_info}
sessionInfo()
```
