---
title: "Notebook 28: CSC — Full CDS vs First 75 Codons"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(iCodon)
library(ggrepel)
```

# Overview

Compares Codon Stabilisation Coefficient (CSC) computed over the full CDS versus
the first 75 codons (225 nt) of each gene's most abundant transcript. Asks whether
the 5' CDS region CSC correlates with full-CDS CSC genome-wide, and whether
eIF3d-promoted mRNAs show a distinct relationship.

- Full-CDS CSC: loaded from nb20 precomputed output
- 75-codon CSC: computed here from TxDb/BSgenome
- Groups: eIF3d hypoxia promotes / Negative control / All other genes (mirrors nb20)
- Level: gene-level (most abundant transcript per gene)
- Genes with CDS < 225 nt (75 codons) are excluded

---

# Section 1: Load Full-CDS CSC and Group Assignments

```{r load-full-csc}
# Gene-level full-CDS CSC from nb20
csc_full_df <- readRDS(here("output", "predictive_modeling", "icodon_csc_per_gene.rds")) %>%
  dplyr::rename(csc_full = csc)

cat("Full-CDS CSC loaded:", nrow(csc_full_df), "genes\n")

# Group assignments from nb20 (transcript-level → collapse to gene-level)
# Group should be consistent within a gene; take first non-NA value
nb20_groups <- readRDS(
  here("output", "predictive_modeling", "csc_mrna_stability_eif3d_hypoxia_1hr.rds")
) %>%
  dplyr::select(ensembl_gene, group) %>%
  dplyr::distinct(ensembl_gene, .keep_all = TRUE)

cat("Group assignments loaded:", nrow(nb20_groups), "genes\n")
print(table(nb20_groups$group))

# Join
csc_gene_df <- csc_full_df %>%
  left_join(nb20_groups, by = "ensembl_gene") %>%
  mutate(group = replace_na(group, "All other genes"))

cat("\nAfter join:\n")
print(table(csc_gene_df$group))
```

---

# Section 2: Extract First-75-Codon CDS Sequences

```{r load-txdb}
most_abundant_tx <- readRDS(
  here("output", "predictive_modeling", "precomputed_most_abundant_tx.rds")
)
cat("Most abundant transcripts:", nrow(most_abundant_tx), "\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38
cat("TxDb and genome loaded\n")
```

```{r extract-cds-75}
# Extract all CDS by transcript
cds_all        <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_names_clean <- str_remove(names(cds_all), "\\..*")

# Match to most abundant transcripts
tx_ids <- most_abundant_tx$transcript_id_clean
tx_idx <- match(tx_ids, cds_names_clean)

found      <- !is.na(tx_idx)
cat("Transcripts matched to TxDb CDS:", sum(found), "/", length(tx_ids), "\n")

# Extract CDS sequences for matched transcripts
cds_matched <- cds_all[tx_idx[found]]
cds_seqs    <- extractTranscriptSeqs(genome, cds_matched)

# Build per-gene CDS length table
cds_df <- tibble(
  transcript_id_raw   = names(cds_seqs),
  transcript_id_clean = str_remove(transcript_id_raw, "\\..*"),
  cds_length          = width(cds_seqs)
) %>%
  left_join(most_abundant_tx %>% dplyr::select(transcript_id_clean, gene_id_clean, symbol),
            by = "transcript_id_clean")

cat("\nCDS length distribution:\n")
print(summary(cds_df$cds_length))

n_short   <- sum(cds_df$cds_length < 225)
n_retained <- sum(cds_df$cds_length >= 225)
cat("\nGenes excluded (CDS < 225 nt / 75 codons):", n_short, "\n")
cat("Genes retained (CDS >= 225 nt):", n_retained, "\n")

# Filter and take first 225 nt
cds_75_df <- cds_df %>%
  filter(cds_length >= 225) %>%
  mutate(
    seq_75 = as.character(subseq(cds_seqs[transcript_id_raw], 1, 225))
  )

cat("\nFirst 75-codon sequences extracted:", nrow(cds_75_df), "\n")
```

---

# Section 3: Compute 75-Codon CSC

```{r compute-csc-75}
predict_human <- iCodon::predict_stability("human")

cds_75_chars <- cds_75_df$seq_75
unique_seqs   <- unique(cds_75_chars)
cat("Unique 75-codon sequences:", length(unique_seqs), "\n")

unique_scores       <- predict_human(unique_seqs)
names(unique_scores) <- unique_seqs

csc_75_scores <- unique_scores[cds_75_chars]

cds_75_df <- cds_75_df %>%
  mutate(csc_75 = csc_75_scores)

cat("\n75-codon CSC summary:\n")
print(summary(cds_75_df$csc_75))

# Join to full-CDS CSC table via gene ID
csc_combined <- csc_gene_df %>%
  left_join(
    cds_75_df %>% dplyr::select(gene_id_clean, csc_75, symbol, cds_length),
    by = c("ensembl_gene" = "gene_id_clean")
  ) %>%
  filter(!is.na(csc_75), !is.na(csc_full)) %>%
  mutate(
    residual = csc_75 - csc_full,
    group    = factor(group, levels = c("eIF3d hypoxia promotes",
                                         "Negative control",
                                         "All other genes"))
  )

cat("\nFinal analysis set:", nrow(csc_combined), "genes\n")
print(table(csc_combined$group))
```

---

# Section 4: Correlation — All Genes

```{r cor-all, fig.width=7, fig.height=7}
group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "grey40",
  "All other genes"        = "grey80"
)

r_all   <- cor(csc_combined$csc_full, csc_combined$csc_75, method = "pearson")
rho_all <- cor(csc_combined$csc_full, csc_combined$csc_75, method = "spearman")
cat(sprintf("All genes  — Pearson r = %.3f | Spearman rho = %.3f\n", r_all, rho_all))

# Plot: grey first, colored groups on top
plot_ordered <- csc_combined %>%
  arrange(group == "eIF3d hypoxia promotes")

ggplot(plot_ordered, aes(x = csc_full, y = csc_75, color = group)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  annotate("text", x = -Inf, y = Inf,
           label = sprintf("Pearson r = %.3f\nSpearman rho = %.3f\nn = %d", r_all, rho_all, nrow(csc_combined)),
           hjust = -0.1, vjust = 1.2, size = 3.5) +
  scale_color_manual(values = group_colors, name = NULL) +
  labs(
    x        = "CSC — full CDS",
    y        = "CSC — first 75 codons",
    title    = "Full CDS vs 5'-proximal (75 codon) CSC",
    subtitle = "Dashed = identity line (y = x)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))

ggsave(here("output/28_csc_full_vs_75_all_genes.pdf"), width = 7, height = 7)
```

---

# Section 5: Correlation and Residuals by Group

```{r cor-by-group}
cor_table <- csc_combined %>%
  group_by(group) %>%
  summarise(
    n            = n(),
    pearson_r    = cor(csc_full, csc_75, method = "pearson"),
    spearman_rho = cor(csc_full, csc_75, method = "spearman"),
    mean_residual = mean(residual),
    median_residual = median(residual),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

cat("=== Correlation by group ===\n")
print(cor_table)
```

```{r wilcoxon-test}
eif3d_resid <- csc_combined %>% filter(group == "eIF3d hypoxia promotes") %>% pull(residual)
other_resid  <- csc_combined %>% filter(group == "All other genes")       %>% pull(residual)
negct_resid  <- csc_combined %>% filter(group == "Negative control")       %>% pull(residual)

wx_eif3d_vs_other <- wilcox.test(eif3d_resid, other_resid, alternative = "two.sided")
wx_eif3d_vs_neg   <- wilcox.test(eif3d_resid, negct_resid, alternative = "two.sided")

cat(sprintf("\nWilcoxon: eIF3d vs All other  — W = %.0f, p = %.4g\n",
            wx_eif3d_vs_other$statistic, wx_eif3d_vs_other$p.value))
cat(sprintf("Wilcoxon: eIF3d vs Neg ctrl   — W = %.0f, p = %.4g\n",
            wx_eif3d_vs_neg$statistic, wx_eif3d_vs_neg$p.value))
```

```{r residual-density, fig.width=8, fig.height=5}
ggplot(csc_combined, aes(x = residual, color = group, fill = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(data = cor_table,
             aes(xintercept = median_residual, color = group),
             linetype = "dotted", linewidth = 0.9) +
  scale_color_manual(values = group_colors, name = NULL) +
  scale_fill_manual(values  = group_colors, name = NULL) +
  labs(
    x        = "CSC residual (first 75 codons − full CDS)",
    y        = "Density",
    title    = "5'-proximal CSC relative to full-CDS CSC",
    subtitle = "Positive = 5' region more stable than full CDS average | dotted = group median"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("output/28_csc_residual_density_by_group.pdf"), width = 8, height = 5)
```

---

# Section 6: Scatter Highlighting eIF3d Group

```{r scatter-eif3d, fig.width=7, fig.height=7}
# Background: neg ctrl + all other (grey)
bg_df     <- csc_combined %>% filter(group != "eIF3d hypoxia promotes")
eif3d_df  <- csc_combined %>% filter(group == "eIF3d hypoxia promotes")

# Label top/bottom 10 eIF3d genes by residual
label_df <- bind_rows(
  eif3d_df %>% slice_max(residual, n = 10),
  eif3d_df %>% slice_min(residual, n = 10)
)

ggplot() +
  geom_point(data = bg_df,    aes(x = csc_full, y = csc_75),
             color = "grey82", alpha = 0.3, size = 0.8) +
  geom_point(data = eif3d_df, aes(x = csc_full, y = csc_75),
             color = "#E69F00", alpha = 0.7, size = 1.8) +
  geom_text_repel(data = label_df,
                  aes(x = csc_full, y = csc_75, label = symbol),
                  size = 3, color = "#E69F00", max.overlaps = 20) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black", linewidth = 0.6) +
  annotate("text", x = -Inf, y = Inf,
           label = sprintf("eIF3d: r = %.3f, rho = %.3f\nn = %d",
                           cor_table$pearson_r[cor_table$group == "eIF3d hypoxia promotes"],
                           cor_table$spearman_rho[cor_table$group == "eIF3d hypoxia promotes"],
                           nrow(eif3d_df)),
           hjust = -0.1, vjust = 1.2, size = 3.5, color = "#E69F00") +
  labs(
    x        = "CSC — full CDS",
    y        = "CSC — first 75 codons",
    title    = "CSC: full CDS vs first 75 codons — eIF3d targets highlighted",
    subtitle = "Orange = eIF3d-promoted (1hr hypoxia, LFC > 0.5) | labeled: top/bottom 10 by residual"
  ) +
  theme_classic(base_size = 12)

ggsave(here("output/28_csc_scatter_eif3d_highlighted.pdf"), width = 7, height = 7)
```

---

# Export

```{r export}
out_df <- csc_combined %>%
  dplyr::select(ensembl_gene, symbol, group, cds_length, csc_full, csc_75, residual) %>%
  arrange(group, desc(abs(residual)))

write_csv(out_df, here("output", "28_csc_full_vs_75codon.csv"))
cat("Exported:", nrow(out_df), "genes to output/28_csc_full_vs_75codon.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
