---
title: "Notebook 29: eIF3e-Mediated Elongation in the eIF3d Model"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(patchwork)
```

# Overview

Lin et al. (Wolf lab) identified ~2,683 mRNAs with >2-fold ribosome density
accumulation between codons 25–75 upon eIF3e knockdown, implicating eIF3e in
promoting elongation through the early coding region. This notebook asks:

1. Are eIF3e-elongation-dependent genes overrepresented among eIF3d-promoted genes?
2. Do eIF3e-elongation-dependent genes show disproportionate sensitivity to eIF3d
   loss (normoxia 1+4hr, hypoxia 1hr)?
3. Do eIF3e-elongation-dependent genes show disproportionate sensitivity to eIF3e
   loss at the translation level (normoxia 1+4hr, hypoxia 1hr)?
4. Add `eif3e_elongation` as a binary feature to `feature_matrix_external_stability.rds`.

---

# Section 1: Load eIF3e elongation genes

All 2,683 genes in this file are pre-filtered (Lin et al.) to have >2-fold ribosome
density increase between codons 25–75 under si-eIF3e conditions.

```{r load_eif3e_genes}
eif3e_raw <- read_csv(
  here("accessories", "elongation", "elongation_eif3e.csv"),
  skip = 4,
  show_col_types = FALSE
)

# Keep only transcript ID and gene symbol; first two columns
eif3e_genes <- eif3e_raw %>%
  dplyr::select(trans_id = 1, gene_name = 2) %>%
  filter(!is.na(trans_id), trans_id != "") %>%
  mutate(trans_id_clean = str_remove(trans_id, "\\..*"))

cat("eIF3e elongation-dependent genes:", nrow(eif3e_genes), "\n")
cat("Unique gene symbols:", n_distinct(eif3e_genes$gene_name), "\n")

eif3e_symbols <- unique(eif3e_genes$gene_name)
```

---

# Section 2: Overrepresentation in eIF3d-promoted genes

```{r load_eif3d_promoted}
eif3d_promoted <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr.csv"),
  show_col_types = FALSE
)
cat("eIF3d-promoted genes (hypoxia 1hr):", nrow(eif3d_promoted), "\n")

eif3d_symbols <- unique(eif3d_promoted$db_gene_symbol)
```

```{r load_background}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))

# Build background gene table with eIF3e membership
background <- features %>%
  dplyr::select(symbol) %>%
  filter(!is.na(symbol)) %>%
  distinct() %>%
  mutate(
    eif3e_elongation = symbol %in% eif3e_symbols,
    eif3d_promoted   = symbol %in% eif3d_symbols
  )

cat("Background genes:", nrow(background), "\n")
cat("eIF3e-elongation in background:", sum(background$eif3e_elongation), "\n")
cat("eIF3d-promoted in background:", sum(background$eif3d_promoted), "\n")
```

```{r fisher_test}
contingency <- table(
  eif3d_promoted   = background$eif3d_promoted,
  eif3e_elongation = background$eif3e_elongation
)
print(contingency)

fisher_res <- fisher.test(contingency)
cat("\nFisher's exact test:\n")
cat("Odds ratio:", round(fisher_res$estimate, 3), "\n")
cat("p-value:", format.pval(fisher_res$p.value, digits = 3), "\n")
```

```{r overrep_barplot, fig.width=4, fig.height=3.5}
bar_df <- background %>%
  group_by(eif3d_promoted) %>%
  summarise(
    pct_eif3e = mean(eif3e_elongation) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    group = ifelse(eif3d_promoted, "eIF3d-promoted\n(hypoxia 1hr)", "All other genes"),
    group = factor(group, levels = c("All other genes", "eIF3d-promoted\n(hypoxia 1hr)"))
  )

p_bar <- ggplot(bar_df, aes(x = group, y = pct_eif3e, fill = group)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("All other genes" = "grey70",
                               "eIF3d-promoted\n(hypoxia 1hr)" = "#E69F00")) +
  labs(
    x = NULL,
    y = "% eIF3e-elongation dependent",
    title = "eIF3e-elongation genes in eIF3d-promoted set",
    subtitle = paste0(
      "OR = ", round(fisher_res$estimate, 2),
      ", p = ", format.pval(fisher_res$p.value, digits = 2)
    )
  ) +
  theme_classic() +
  theme(legend.position = "none",
        plot.subtitle = element_text(size = 9, color = "grey40"))

print(p_bar)

ggsave(here("plots", "29_eif3e_overrep_in_eif3d_promoted.pdf"),
       p_bar, width = 4, height = 3.5)
```

---

# Section 3a: eIF3d TE dependence by eIF3e elongation status

```{r si3d_normoxia_density, fig.width=5, fig.height=3.5}
si3d_norm <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) %>%
  filter(!is.na(te_lfc)) %>%
  mutate(
    te_lfc = as.numeric(te_lfc),
    eif3e_elongation = symbol %in% eif3e_symbols
  )

wt_si3d_norm <- wilcox.test(te_lfc ~ eif3e_elongation, data = si3d_norm)
n_pos <- sum(si3d_norm$eif3e_elongation)
n_neg <- sum(!si3d_norm$eif3e_elongation)

p_si3d_norm <- ggplot(si3d_norm, aes(x = te_lfc, fill = eif3e_elongation)) +
  geom_density(alpha = 0.6, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#E69F00"),
    labels = c(paste0("Not eIF3e-elongation (n=", n_neg, ")"),
               paste0("eIF3e-elongation (n=", n_pos, ")"))
  ) +
  scale_x_continuous(limits = c(-3, 3)) +
  annotate("text", x = -2.8, y = Inf,
           label = paste0("Wilcoxon p = ", format.pval(wt_si3d_norm$p.value, digits = 2)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  labs(title = "eIF3d dependence — Normoxia 1+4hr (time-regressed)",
       x = "TE log2FC (si3d vs sictrl)", y = "Density", fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8),
        plot.title = element_text(size = 10))

print(p_si3d_norm)
```

```{r si3d_hypoxia_density, fig.width=5, fig.height=3.5}
si3d_hyp <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) %>%
  filter(!is.na(te_lfc)) %>%
  mutate(
    te_lfc = as.numeric(te_lfc),
    eif3e_elongation = symbol %in% eif3e_symbols
  )

wt_si3d_hyp <- wilcox.test(te_lfc ~ eif3e_elongation, data = si3d_hyp)
n_pos <- sum(si3d_hyp$eif3e_elongation)
n_neg <- sum(!si3d_hyp$eif3e_elongation)

p_si3d_hyp <- ggplot(si3d_hyp, aes(x = te_lfc, fill = eif3e_elongation)) +
  geom_density(alpha = 0.6, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#E69F00"),
    labels = c(paste0("Not eIF3e-elongation (n=", n_neg, ")"),
               paste0("eIF3e-elongation (n=", n_pos, ")"))
  ) +
  scale_x_continuous(limits = c(-3, 3)) +
  annotate("text", x = -2.8, y = Inf,
           label = paste0("Wilcoxon p = ", format.pval(wt_si3d_hyp$p.value, digits = 2)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  labs(title = "eIF3d dependence — Hypoxia 1hr",
       x = "TE log2FC (si3d vs sictrl)", y = "Density", fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8),
        plot.title = element_text(size = 10))

print(p_si3d_hyp)
```

```{r si3d_combined_panel, fig.width=10, fig.height=3.5}
p_si3d_panel <- p_si3d_norm + p_si3d_hyp +
  plot_annotation(title = "eIF3d TE sensitivity by eIF3e-elongation status")

print(p_si3d_panel)

ggsave(here("plots", "29_si3d_te_density_by_eif3e_elongation.pdf"),
       p_si3d_panel, width = 10, height = 3.5)
```

---

# Section 3b: eIF3e TE dependence by eIF3e elongation status

```{r si3e_normoxia_density, fig.width=5, fig.height=3.5}
# te_si3e_vs_sictrl_normoxia_1and4hr.csv: spike-in normalized, time-regressed
# cols: gene_id, te_lfc, te_padj, symbol
si3e_norm <- read_csv(
  here("output", "te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
) %>%
  filter(!is.na(te_lfc)) %>%
  mutate(
    te_lfc = as.numeric(te_lfc),
    eif3e_elongation = symbol %in% eif3e_symbols
  )

wt_si3e_norm <- wilcox.test(te_lfc ~ eif3e_elongation, data = si3e_norm)
n_pos <- sum(si3e_norm$eif3e_elongation)
n_neg <- sum(!si3e_norm$eif3e_elongation)

p_si3e_norm <- ggplot(si3e_norm, aes(x = te_lfc, fill = eif3e_elongation)) +
  geom_density(alpha = 0.6, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#E69F00"),
    labels = c(paste0("Not eIF3e-elongation (n=", n_neg, ")"),
               paste0("eIF3e-elongation (n=", n_pos, ")"))
  ) +
  scale_x_continuous(limits = c(-3, 3)) +
  annotate("text", x = -2.8, y = Inf,
           label = paste0("Wilcoxon p = ", format.pval(wt_si3e_norm$p.value, digits = 2)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  labs(title = "eIF3e dependence — Normoxia 1+4hr (time-regressed, spike-in)",
       x = "TE log2FC (si3e vs sictrl)", y = "Density", fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8),
        plot.title = element_text(size = 10))

print(p_si3e_norm)
```

```{r si3e_hypoxia_density, fig.width=5, fig.height=3.5}
si3e_hyp <- read_csv(
  here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) %>%
  filter(!is.na(te_lfc)) %>%
  mutate(
    te_lfc = as.numeric(te_lfc),
    eif3e_elongation = symbol %in% eif3e_symbols
  )

wt_si3e_hyp <- wilcox.test(te_lfc ~ eif3e_elongation, data = si3e_hyp)
n_pos <- sum(si3e_hyp$eif3e_elongation)
n_neg <- sum(!si3e_hyp$eif3e_elongation)

p_si3e_hyp <- ggplot(si3e_hyp, aes(x = te_lfc, fill = eif3e_elongation)) +
  geom_density(alpha = 0.6, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#E69F00"),
    labels = c(paste0("Not eIF3e-elongation (n=", n_neg, ")"),
               paste0("eIF3e-elongation (n=", n_pos, ")"))
  ) +
  scale_x_continuous(limits = c(-3, 3)) +
  annotate("text", x = -2.8, y = Inf,
           label = paste0("Wilcoxon p = ", format.pval(wt_si3e_hyp$p.value, digits = 2)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey30") +
  labs(title = "eIF3e dependence — Hypoxia 1hr",
       x = "TE log2FC (si3e vs sictrl)", y = "Density", fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom", legend.text = element_text(size = 8),
        plot.title = element_text(size = 10))

print(p_si3e_hyp)
```

```{r si3e_combined_panel, fig.width=10, fig.height=3.5}
p_si3e_panel <- p_si3e_norm + p_si3e_hyp +
  plot_annotation(title = "eIF3e TE sensitivity by eIF3e-elongation status")

print(p_si3e_panel)

ggsave(here("plots", "29_si3e_te_density_by_eif3e_elongation.pdf"),
       p_si3e_panel, width = 10, height = 3.5)
```

---

# Section 4: Add eIF3e elongation feature to feature matrix

```{r update_feature_matrix}
features <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))

features <- features %>%
  mutate(eif3e_elongation = symbol %in% eif3e_symbols)

cat("eif3e_elongation column added:\n")
print(table(features$eif3e_elongation, useNA = "ifany"))
cat("Coverage: ", round(mean(features$eif3e_elongation, na.rm = TRUE) * 100, 1), "% of feature matrix genes\n")

saveRDS(features, here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
cat("Saved: feature_matrix_external_stability.rds\n")
```
