---
title: "30 — Condition-Specific Codon Usage"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Differential codon usage analysis for four eIF3d-promoted gene sets vs a
size-matched negative control. Two plots per comparison:

- **Plot A**: Within-AA proportion difference (promotes − neg ctrl), faceted by amino acid
- **Plot B**: Frame 0 vs frameshift controls (frames 1 & 2), with BH-adjusted Wilcoxon significance

Gene sets:

| Label | Source |
|---|---|
| Hypoxia (all promoted) | `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` |
| Normoxia (all promoted) | `normoxia_3d_promotes_TE_combined_lfc0.5.csv` |
| Hypoxia only | `hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv` |
| Normoxia only | `normoxia_3d_promotes_TE_normoxia_specific_lfc0.5.csv` |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
```

---

# Section 1: Setup — CDS Sequences & Codon Count Matrices

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
# Frame 0 codon counts
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

# Frame-shifted matrices for nucleotide composition controls
shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  ct  <- codonTable(seqs)
  mat <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

cds_frame1 <- shift_frame(cds_frame0, 1)
cds_frame2 <- shift_frame(cds_frame0, 2)
counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

# Transcript-to-gene mapping
tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

# Section 2: Helper Functions

```{r helper_functions}
# Safe within-AA proportion: returns 0 for each codon when sum == 0 (avoids NaN)
safe_prop <- function(counts) {
  s <- sum(counts, na.rm = TRUE)
  if (s == 0) rep(0, length(counts)) else counts / s
}

# Compute within-AA proportion delta (promotes - neg_ctrl) at aggregated level
codon_diff_df_fn <- function(promotes_tx, neg_ctrl_tx) {
  promotes_counts <- colSums(counts_mat0[promotes_tx, , drop = FALSE])
  neg_ctrl_counts <- colSums(counts_mat0[neg_ctrl_tx, , drop = FALSE])

  imap_dfr(aa_to_codons, function(codons, aa) {
    p_prop <- safe_prop(promotes_counts[codons])
    n_prop <- safe_prop(neg_ctrl_counts[codons])
    tibble(amino_acid = aa, codon = codons,
           promotes_prop = p_prop, neg_ctrl_prop = n_prop,
           delta = p_prop - n_prop)
  })
}

# Per-transcript Wilcoxon test for one frame
per_tx_sig <- function(counts_mat, promotes_tx, neg_ctrl_tx, frame_label) {
  p_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat)]
  n_tx <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    if (length(codons) < 2) return(NULL)
    sub_p <- counts_mat[p_tx, codons, drop = FALSE]
    sub_n <- counts_mat[n_tx, codons, drop = FALSE]
    rs_p  <- rowSums(sub_p); rs_n <- rowSums(sub_n)
    map_dfr(codons, function(codon) {
      props_p <- sub_p[rs_p > 0, codon] / rs_p[rs_p > 0]
      props_n <- sub_n[rs_n > 0, codon] / rs_n[rs_n > 0]
      if (length(props_p) < 3 || length(props_n) < 3) {
        return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_, frame = frame_label))
      }
      wt <- wilcox.test(props_p, props_n, exact = FALSE)
      tibble(amino_acid = aa, codon = codon, p_val = wt$p.value, frame = frame_label)
    })
  })
}

# Delta for one frame from pre-summed counts
compute_delta <- function(prom_counts, neg_counts, frame_label) {
  imap_dfr(aa_to_codons, function(codons, aa) {
    p_prop <- safe_prop(prom_counts[codons])
    n_prop <- safe_prop(neg_counts[codons])
    tibble(amino_acid = aa, codon = codons,
           delta = p_prop - n_prop, frame = frame_label)
  })
}

# Plot A: differential codon usage bar chart
plot_codon_diff <- function(codon_diff_df, title_label) {
  plot_df <- codon_diff_df %>%
    filter(amino_acid %in% multi_aa) %>%
    mutate(
      aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
      direction = ifelse(delta >= 0, "promotes", "neg_ctrl"),
      codon     = reorder(codon, delta)
    )

  ggplot(plot_df, aes(x = codon, y = delta, fill = direction)) +
    geom_col() +
    scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                      guide = "none") +
    facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    labs(
      title = paste0("Differential codon usage: eIF3d promotes vs negative control\n(", title_label, ")"),
      x = "Codon", y = "\u0394 within-AA proportion (promotes \u2212 neg ctrl)"
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          strip.text  = element_text(face = "bold"))
}

# Plot B: frame control grouped bar chart
plot_frame_ctrl <- function(codon_diff_df, promotes_tx, neg_ctrl_tx, title_label) {
  # Aggregated frame 1 and 2 counts
  p1 <- promotes_tx[promotes_tx %in% rownames(counts_mat1)]
  n1 <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(counts_mat1)]
  p2 <- promotes_tx[promotes_tx %in% rownames(counts_mat2)]
  n2 <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(counts_mat2)]

  pc1 <- colSums(counts_mat1[p1, , drop = FALSE])
  nc1 <- colSums(counts_mat1[n1, , drop = FALSE])
  pc2 <- colSums(counts_mat2[p2, , drop = FALSE])
  nc2 <- colSums(counts_mat2[n2, , drop = FALSE])

  # Combined frame diffs
  frame_diff_df <- bind_rows(
    codon_diff_df %>% mutate(frame = "Frame 0 (coding)"),
    compute_delta(pc1, nc1, "Frame 1 (control)"),
    compute_delta(pc2, nc2, "Frame 2 (control)")
  ) %>%
    filter(amino_acid %in% multi_aa) %>%
    mutate(
      aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
      frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
    )

  # Codon x-axis order by frame 0 delta
  codon_order <- codon_diff_df %>%
    filter(amino_acid %in% multi_aa) %>%
    arrange(amino_acid, delta) %>%
    pull(codon)
  frame_diff_df <- frame_diff_df %>%
    mutate(codon = factor(codon, levels = codon_order))

  # Per-transcript Wilcoxon tests
  sig_df <- bind_rows(
    per_tx_sig(counts_mat0, promotes_tx, neg_ctrl_tx, "Frame 0 (coding)"),
    per_tx_sig(counts_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)"),
    per_tx_sig(counts_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")
  ) %>%
    filter(amino_acid %in% multi_aa) %>%
    mutate(
      q_val = p.adjust(p_val, method = "BH"),
      sig   = !is.na(q_val) & q_val < 0.05,
      frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
    )

  # Classify clean vs. composition-confounded
  f0_sig   <- sig_df %>% filter(frame == "Frame 0 (coding)") %>%
    dplyr::select(amino_acid, codon, f0_sig = sig)
  ctrl_sig <- sig_df %>% filter(frame != "Frame 0 (coding)") %>%
    group_by(amino_acid, codon) %>%
    summarise(any_ctrl_sig = any(sig), .groups = "drop")
  f0_labels <- f0_sig %>%
    left_join(ctrl_sig, by = c("amino_acid", "codon")) %>%
    mutate(sig_label = case_when(
      f0_sig & !any_ctrl_sig ~ "\u2726",
      f0_sig &  any_ctrl_sig ~ "\u00b7",
      TRUE                   ~ ""
    )) %>%
    dplyr::select(amino_acid, codon, sig_label) %>%
    mutate(frame = factor("Frame 0 (coding)",
                          levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

  aa_any_f0 <- f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
  aa_clean  <- f0_labels %>% filter(sig_label == "\u2726") %>% pull(amino_acid) %>% unique()

  frame_diff_df <- frame_diff_df %>%
    mutate(aa_label = ifelse(amino_acid %in% aa_any_f0,
                             paste0(aa_label, " *"), aa_label))

  aa_label_map <- frame_diff_df %>% distinct(amino_acid, aa_label)
  facet_clean  <- aa_label_map %>% filter(amino_acid %in% aa_clean)

  # geom_rect for blue border (only added when there are clean-signal amino acids;
  # passing an empty data frame to geom_rect with facet_wrap causes panel-index errors)
  rect_layer <- if (nrow(facet_clean) > 0) {
    geom_rect(data = facet_clean,
              aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
              inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5)
  } else {
    NULL
  }

  ggplot(frame_diff_df, aes(x = codon, y = delta, fill = frame)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.75) +
    rect_layer +
    scale_fill_manual(
      values = c("Frame 0 (coding)"   = "#E69F00",
                 "Frame 1 (control)"  = "#999999",
                 "Frame 2 (control)"  = "#CCCCCC")
    ) +
    facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    labs(
      title    = paste0("Differential codon usage by reading frame: frame 0 vs. frameshift controls\n(", title_label, ")"),
      subtitle = "Frames 1 & 2 control for nucleotide composition; persistent frame-0 signal = true codon preference\nBlue border: frame 0 significant, controls not (clean codon signal). *: frame 0 significant (controls may also be significant). [q < 0.05, BH-adjusted Wilcoxon]",
      x = "Codon", y = "\u0394 within-AA proportion (promotes \u2212 neg ctrl)",
      fill = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
      strip.text        = element_text(face = "bold"),
      legend.position   = "top"
    )
}
```

---

# Section 3: Negative Control Pool

```{r neg_ctrl_pool}
neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

cat("Negative control pool:", nrow(neg_pool), "genes\n")
```

---

# Section 4a: Hypoxia (All Promoted)

```{r hypoxia_all_load}
promotes_hyp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids_hyp <- promotes_hyp$ensembl_gene
cat("Hypoxia promoted genes:", length(promotes_ids_hyp), "\n")

neg_filtered_hyp <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids_hyp)
set.seed(9)
neg_ctrl_ids_hyp <- sample(neg_filtered_hyp$ensembl_gene,
                            size = length(promotes_ids_hyp), replace = FALSE)
cat("Neg ctrl (size-matched):", length(neg_ctrl_ids_hyp), "\n")

promotes_tx_hyp <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids_hyp) %>%
  pull(transcript_id_clean)
neg_ctrl_tx_hyp <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids_hyp) %>%
  pull(transcript_id_clean)
cat("Promotes transcripts:", length(promotes_tx_hyp),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx_hyp), "\n")
```

```{r hypoxia_all_plota, fig.width=10, fig.height=8}
codon_diff_hyp <- codon_diff_df_fn(promotes_tx_hyp, neg_ctrl_tx_hyp)

cat("Top 10 codons by |delta|:\n")
print(codon_diff_hyp %>% arrange(desc(abs(delta))) %>% head(10))

p_a_hyp <- plot_codon_diff(codon_diff_hyp, "Hypoxia 1hr promoted")
p_a_hyp
ggsave(here("plots", "codon_usage_diff_hypoxia.pdf"), p_a_hyp, width = 10, height = 8)
```

```{r hypoxia_all_plotb, fig.width=12, fig.height=9}
p_b_hyp <- plot_frame_ctrl(codon_diff_hyp, promotes_tx_hyp, neg_ctrl_tx_hyp,
                            "Hypoxia 1hr promoted")
p_b_hyp
ggsave(here("plots", "codon_usage_frame_ctrl_hypoxia.pdf"), p_b_hyp, width = 12, height = 9)
```

---

# Section 4b: Normoxia (All Promoted)

```{r normoxia_all_load}
promotes_nox <- read_csv(
  here("output", "genesets", "normoxia_3d_promotes_TE_combined_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids_nox <- promotes_nox$ensembl_gene
cat("Normoxia promoted genes:", length(promotes_ids_nox), "\n")

neg_filtered_nox <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids_nox)
set.seed(9)
neg_ctrl_ids_nox <- sample(neg_filtered_nox$ensembl_gene,
                            size = length(promotes_ids_nox), replace = FALSE)
cat("Neg ctrl (size-matched):", length(neg_ctrl_ids_nox), "\n")

promotes_tx_nox <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids_nox) %>%
  pull(transcript_id_clean)
neg_ctrl_tx_nox <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids_nox) %>%
  pull(transcript_id_clean)
cat("Promotes transcripts:", length(promotes_tx_nox),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx_nox), "\n")
```

```{r normoxia_all_plota, fig.width=10, fig.height=8}
codon_diff_nox <- codon_diff_df_fn(promotes_tx_nox, neg_ctrl_tx_nox)

cat("Top 10 codons by |delta|:\n")
print(codon_diff_nox %>% arrange(desc(abs(delta))) %>% head(10))

p_a_nox <- plot_codon_diff(codon_diff_nox, "Normoxia (combined) promoted")
p_a_nox
ggsave(here("plots", "codon_usage_diff_normoxia.pdf"), p_a_nox, width = 10, height = 8)
```

```{r normoxia_all_plotb, fig.width=12, fig.height=9}
p_b_nox <- plot_frame_ctrl(codon_diff_nox, promotes_tx_nox, neg_ctrl_tx_nox,
                            "Normoxia (combined) promoted")
p_b_nox
ggsave(here("plots", "codon_usage_frame_ctrl_normoxia.pdf"), p_b_nox, width = 12, height = 9)
```

---

# Section 4c: Hypoxia Only (Hypoxia-Specific)

```{r hypoxia_specific_load}
promotes_hyp_sp <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids_hyp_sp <- promotes_hyp_sp$ensembl_gene
cat("Hypoxia-specific promoted genes:", length(promotes_ids_hyp_sp), "\n")

neg_filtered_hyp_sp <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids_hyp_sp)
set.seed(9)
neg_ctrl_ids_hyp_sp <- sample(neg_filtered_hyp_sp$ensembl_gene,
                               size = length(promotes_ids_hyp_sp), replace = FALSE)
cat("Neg ctrl (size-matched):", length(neg_ctrl_ids_hyp_sp), "\n")

promotes_tx_hyp_sp <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids_hyp_sp) %>%
  pull(transcript_id_clean)
neg_ctrl_tx_hyp_sp <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids_hyp_sp) %>%
  pull(transcript_id_clean)
cat("Promotes transcripts:", length(promotes_tx_hyp_sp),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx_hyp_sp), "\n")
```

```{r hypoxia_specific_plota, fig.width=10, fig.height=8}
codon_diff_hyp_sp <- codon_diff_df_fn(promotes_tx_hyp_sp, neg_ctrl_tx_hyp_sp)

cat("Top 10 codons by |delta|:\n")
print(codon_diff_hyp_sp %>% arrange(desc(abs(delta))) %>% head(10))

p_a_hyp_sp <- plot_codon_diff(codon_diff_hyp_sp, "Hypoxia-specific promoted")
p_a_hyp_sp
ggsave(here("plots", "codon_usage_diff_hypoxia_specific.pdf"), p_a_hyp_sp, width = 10, height = 8)
```

```{r hypoxia_specific_plotb, fig.width=12, fig.height=9}
p_b_hyp_sp <- plot_frame_ctrl(codon_diff_hyp_sp, promotes_tx_hyp_sp, neg_ctrl_tx_hyp_sp,
                               "Hypoxia-specific promoted")
p_b_hyp_sp
ggsave(here("plots", "codon_usage_frame_ctrl_hypoxia_specific.pdf"), p_b_hyp_sp, width = 12, height = 9)
```

---

# Section 4d: Normoxia Only (Normoxia-Specific)

```{r normoxia_specific_load}
promotes_nox_sp <- read_csv(
  here("output", "genesets", "normoxia_3d_promotes_TE_normoxia_specific_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids_nox_sp <- promotes_nox_sp$ensembl_gene
cat("Normoxia-specific promoted genes:", length(promotes_ids_nox_sp), "\n")

neg_filtered_nox_sp <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids_nox_sp)
set.seed(9)
neg_ctrl_ids_nox_sp <- sample(neg_filtered_nox_sp$ensembl_gene,
                               size = length(promotes_ids_nox_sp), replace = FALSE)
cat("Neg ctrl (size-matched):", length(neg_ctrl_ids_nox_sp), "\n")

promotes_tx_nox_sp <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids_nox_sp) %>%
  pull(transcript_id_clean)
neg_ctrl_tx_nox_sp <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids_nox_sp) %>%
  pull(transcript_id_clean)
cat("Promotes transcripts:", length(promotes_tx_nox_sp),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx_nox_sp), "\n")
```

```{r normoxia_specific_plota, fig.width=10, fig.height=8}
codon_diff_nox_sp <- codon_diff_df_fn(promotes_tx_nox_sp, neg_ctrl_tx_nox_sp)

cat("Top 10 codons by |delta|:\n")
print(codon_diff_nox_sp %>% arrange(desc(abs(delta))) %>% head(10))

p_a_nox_sp <- plot_codon_diff(codon_diff_nox_sp, "Normoxia-specific promoted")
p_a_nox_sp
ggsave(here("plots", "codon_usage_diff_normoxia_specific.pdf"), p_a_nox_sp, width = 10, height = 8)
```

```{r normoxia_specific_plotb, fig.width=12, fig.height=9}
p_b_nox_sp <- plot_frame_ctrl(codon_diff_nox_sp, promotes_tx_nox_sp, neg_ctrl_tx_nox_sp,
                               "Normoxia-specific promoted")
p_b_nox_sp
ggsave(here("plots", "codon_usage_frame_ctrl_normoxia_specific.pdf"), p_b_nox_sp, width = 12, height = 9)
```

---

# Section 5: Hypoxia-Specific vs Normoxia-Specific Direct Comparison

Within-AA codon proportions computed from each promotes set directly (no neg ctrl),
then subtracted: Δ = hypoxia-specific proportion − normoxia-specific proportion.
Positive bars = codon enriched in hypoxia-specific; negative bars = enriched in normoxia-specific.

```{r hyp_vs_nox_comparison, fig.width=10, fig.height=8}
# Use promotes_prop columns already computed in sections 4c and 4d
# (requires sections 4c and 4d to have been run)
direct_comp_df <- codon_diff_hyp_sp %>%
  dplyr::select(amino_acid, codon, hyp_prop = promotes_prop) %>%
  left_join(
    codon_diff_nox_sp %>% dplyr::select(amino_acid, codon, nox_prop = promotes_prop),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta = hyp_prop - nox_prop)

cat("Top 10 codons by |delta| (hypoxia-specific vs normoxia-specific):\n")
print(direct_comp_df %>% filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta))) %>% head(10))

plot_df_direct <- direct_comp_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta >= 0, "hypoxia", "normoxia"),
    codon     = reorder(codon, delta)
  )

p_direct <- ggplot(plot_df_direct, aes(x = codon, y = delta, fill = direction)) +
  geom_col() +
  scale_fill_manual(
    values = c(hypoxia = "#E69F00", normoxia = "#5B8DB8"),
    labels = c(hypoxia = "Hypoxia-specific", normoxia = "Normoxia-specific"),
    name   = NULL
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Codon usage: hypoxia-specific vs normoxia-specific eIF3d-promoted genes",
    subtitle = "Within-AA proportion (hypoxia-specific promotes \u2212 normoxia-specific promotes)",
    x = "Codon", y = "\u0394 within-AA proportion"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top"
  )
p_direct
ggsave(here("plots", "codon_usage_hyp_specific_vs_nox_specific.pdf"),
       p_direct, width = 10, height = 8)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
