---
title: "Notebook 32: eIF3 Cap Binding CLIP — Density Plots for eIF3d Targets (Hypoxia 1hr)"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(cowplot)
```

# Overview

Density plots of Lee et al. CLIP cap binding LFC values for genes promoted by eIF3d
at 1hr hypoxia vs a size-matched negative control set.

**Gene sets:**
- **Promoted**: te_lfc > 0.5, padj < 0.05 (si3d vs sictrl, hypoxia 1hr)
- **Negative control**: genes never regulated by eIF3d (`negative_control_genes.csv`), subsampled to match promoted n with CLIP data (seed = 9)
- **All genes**: grey background

# Load Data

```{r load_data}
features_raw <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds")) %>%
  dplyr::select(gene_id_clean, symbol, starts_with("clip_"))

# Coerce list columns to numeric
features <- features_raw %>%
  mutate(
    clip_lee_complete_media_lfc   = as.numeric(unlist(clip_lee_complete_media_lfc)),
    clip_lee_glu_dep_lfc          = as.numeric(unlist(clip_lee_glu_dep_lfc)),
    clip_lee_int_stress_lfc       = as.numeric(unlist(clip_lee_int_stress_lfc)),
    clip_lee_glu_vs_cm_enrichment = as.numeric(unlist(clip_lee_glu_vs_cm_enrichment)),
    clip_eif3_count               = as.numeric(unlist(clip_eif3_count)),
    clip_cap_binding_count        = as.numeric(unlist(clip_cap_binding_count))
  )

cat("Feature matrix genes:", nrow(features), "\n")

te_data <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                    show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::select(gene_id_clean, te_lfc, te_padj)

cat("TE outcome genes:", nrow(te_data), "\n")

neg_controls_raw <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                              show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("Negative control genes (raw):", nrow(neg_controls_raw), "\n")
```

# Define Gene Sets

```{r define_sets}
plot_df <- inner_join(features, te_data, by = "gene_id_clean")
cat("Genes with features + TE outcome:", nrow(plot_df), "\n")

promoted <- plot_df %>%
  filter(te_lfc > 0.5, te_padj < 0.05)
cat("Promoted (LFC > 0.5, padj < 0.05):", nrow(promoted), "\n")
cat("  with non-NA clip_lee_complete_media_lfc:",
    sum(!is.na(promoted$clip_lee_complete_media_lfc)), "\n")

# All negative control genes in the feature matrix — no subsampling needed for density plots
# (density curves normalize to 1 regardless of n)
neg_sampled <- plot_df %>%
  filter(gene_id_clean %in% neg_controls_raw$gene_id_clean)
cat("Negative controls in matrix:", nrow(neg_sampled), "\n")
cat("  with non-NA clip_lee_complete_media_lfc:",
    sum(!is.na(neg_sampled$clip_lee_complete_media_lfc)), "\n")
```

# Density Plots — Cap Binding LFC

```{r make_long}
# Build long-format data frame with group labels for all three CLIP LFC columns
# Only genes with non-NA values per column are included
clip_cols <- c(
  "clip_lee_complete_media_lfc",
  "clip_lee_glu_dep_lfc",
  "clip_lee_int_stress_lfc"
)

long_df <- bind_rows(
  plot_df    %>% mutate(group = "All genes"),
  promoted   %>% mutate(group = "Promoted by eIF3d"),
  neg_sampled %>% mutate(group = "Negative control")
) %>%
  dplyr::select(group, all_of(clip_cols)) %>%
  mutate(group = factor(group, levels = c("All genes", "Negative control", "Promoted by eIF3d"))) %>%
  pivot_longer(all_of(clip_cols), names_to = "clip_col", values_to = "lfc") %>%
  filter(!is.na(lfc))

group_colors <- c(
  "All genes"          = "grey70",
  "Negative control"   = "black",
  "Promoted by eIF3d"  = "#E69F00"
)

# Median lines per group × column
medians_df <- long_df %>%
  group_by(group, clip_col) %>%
  summarise(med = median(lfc), n = n(), .groups = "drop")

cat("Gene counts per group per CLIP column:\n")
print(medians_df %>% dplyr::select(clip_col, group, n, med))
```

## Complete Media

```{r plot_complete_media, fig.width=7, fig.height=5}
col <- "clip_lee_complete_media_lfc"
p1 <- ggplot(filter(long_df, clip_col == col), aes(x = lfc, color = group)) +
  geom_density(linewidth = 0.9, fill = NA) +
  geom_vline(data = filter(medians_df, clip_col == col),
             aes(xintercept = med, color = group), linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = group_colors, name = NULL) +
  theme_cowplot() +
  labs(title = "eIF3 Cap Binding — Complete Media",
       x = "eIF3-cap binding LFC (complete media)", y = "Density")
print(p1)
ggsave(here("plots", "clip_cap_binding_density_complete_media.pdf"), p1, width = 7, height = 5)
```

## Glucose Deprivation

```{r plot_glu_dep, fig.width=7, fig.height=5}
col <- "clip_lee_glu_dep_lfc"
p2 <- ggplot(filter(long_df, clip_col == col), aes(x = lfc, color = group)) +
  geom_density(linewidth = 0.9, fill = NA) +
  geom_vline(data = filter(medians_df, clip_col == col),
             aes(xintercept = med, color = group), linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = group_colors, name = NULL) +
  theme_cowplot() +
  labs(title = "eIF3 Cap Binding — Glucose Deprivation",
       x = "eIF3-cap binding LFC (glucose deprivation)", y = "Density")
print(p2)
ggsave(here("plots", "clip_cap_binding_density_glu_dep.pdf"), p2, width = 7, height = 5)
```

## Integrated Stress

```{r plot_int_stress, fig.width=7, fig.height=5}
col <- "clip_lee_int_stress_lfc"
p3 <- ggplot(filter(long_df, clip_col == col), aes(x = lfc, color = group)) +
  geom_density(linewidth = 0.9, fill = NA) +
  geom_vline(data = filter(medians_df, clip_col == col),
             aes(xintercept = med, color = group), linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = group_colors, name = NULL) +
  theme_cowplot() +
  labs(title = "eIF3 Cap Binding — Integrated Stress",
       x = "eIF3-cap binding LFC (integrated stress)", y = "Density")
print(p3)
ggsave(here("plots", "clip_cap_binding_density_int_stress.pdf"), p3, width = 7, height = 5)
```

# % with eIF3 CLIP Binding Evidence

```{r clip_bound_bar, fig.width=6, fig.height=5}
pct_prom <- 100 * mean(promoted$clip_eif3_count > 0, na.rm = TRUE)
pct_neg  <- 100 * mean(neg_sampled$clip_eif3_count > 0, na.rm = TRUE)
pct_all  <- 100 * mean(plot_df$clip_eif3_count > 0, na.rm = TRUE)

cat("% with eIF3 CLIP binding (clip_eif3_count > 0):\n")
cat("  Promoted:         ", round(pct_prom, 1), "% (n=", nrow(promoted), ")\n")
cat("  Negative control: ", round(pct_neg,  1), "% (n=", nrow(neg_sampled), ")\n")
cat("  All genes:        ", round(pct_all,  1), "% (n=", nrow(plot_df), ")\n")

bound_df <- data.frame(
  group     = factor(c("All genes", "Negative control", "Promoted by eIF3d"),
                     levels = c("All genes", "Negative control", "Promoted by eIF3d")),
  pct_bound = c(pct_all, pct_neg, pct_prom),
  n         = c(nrow(plot_df), nrow(neg_sampled), nrow(promoted))
)

p4 <- ggplot(bound_df, aes(x = group, y = pct_bound, fill = group)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c(
    "All genes"          = "grey70",
    "Negative control"   = "black",
    "Promoted by eIF3d"  = "#E69F00"
  )) +
  geom_text(aes(label = paste0(round(pct_bound, 1), "%\n(n=", n, ")")),
            vjust = -0.3, size = 3.5) +
  theme_cowplot() +
  theme(legend.position = "none") +
  labs(
    title    = "% Genes with eIF3 CLIP Binding Evidence",
    subtitle = "clip_eif3_count > 0",
    x = "",
    y = "% with eIF3 CLIP binding"
  ) +
  ylim(0, max(bound_df$pct_bound) * 1.2)

print(p4)
ggsave(here("plots", "clip_eif3_bound_pct_bar.pdf"), p4, width = 6, height = 5)
```

# Distribution of clip_eif3_count

```{r clip_count_distribution, fig.width=8, fig.height=5}
# Proportion of genes at each count level (0, 1, 2, ...) per group
count_df <- bind_rows(
  plot_df     %>% mutate(group = "All genes"),
  promoted    %>% mutate(group = "Promoted by eIF3d"),
  neg_sampled %>% mutate(group = "Negative control")
) %>%
  mutate(
    group           = factor(group, levels = c("All genes", "Negative control", "Promoted by eIF3d")),
    clip_eif3_count = as.integer(clip_eif3_count)
  ) %>%
  count(group, clip_eif3_count) %>%
  group_by(group) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

print(count_df)

p5 <- ggplot(count_df, aes(x = factor(clip_eif3_count), y = pct, fill = group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "All genes"          = "grey70",
    "Negative control"   = "black",
    "Promoted by eIF3d"  = "#E69F00"
  ), name = NULL) +
  theme_cowplot() +
  labs(
    title = "Distribution of eIF3 CLIP Binding Evidence",
    subtitle = "clip_eif3_count = number of CLIP datasets with binding evidence",
    x = "Number of CLIP datasets (clip_eif3_count)",
    y = "% of genes in group"
  )

print(p5)
ggsave(here("plots", "clip_eif3_count_distribution.pdf"), p5, width = 8, height = 5)
```

# Session Info

```{r session_info}
sessionInfo()
```
