---
title: "Notebook 34: Hypoxia vs Normoxia eIF3d TE Residuals"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(broom)
```

## Load and join data

Spike-in normalized DESeq2 results comparing sictrl vs si3d. Positive LFC = higher TE in sictrl.

```{r load-data}
normoxia    <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"))
hypoxia_1hr <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"))

plot_df <- normoxia %>%
  dplyr::select(gene_id, symbol, te_lfc_normoxia = te_lfc, te_padj_normoxia = te_padj) %>%
  inner_join(
    hypoxia_1hr %>%
      dplyr::select(gene_id, te_lfc_hypoxia = te_lfc, te_padj_hypoxia = te_padj),
    by = "gene_id"
  )

cat("Genes in plot:", nrow(plot_df), "\n")
```

## Fit linear model and extract residuals

```{r fit-model}
lm_fit <- lm(te_lfc_hypoxia ~ te_lfc_normoxia, data = plot_df, na.action = na.exclude)

coefs         <- coef(lm_fit)
formula_label <- sprintf("y = %.3fx + %.3f", coefs["te_lfc_normoxia"], coefs["(Intercept)"])
cat(formula_label, "\n")

plot_df <- augment(lm_fit, data = plot_df)
```

## Plot 1: Scatter with fit line and equation

```{r plot-fit, fig.width=5, fig.height=4}
p1 <- ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia)) +
  geom_point(alpha = 0.4, size = 0.8, color = "grey50") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
           label = formula_label, size = 3) +
  labs(
    x = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y = "TE LFC 1hr hypoxia (sictrl/si3d)"
  ) +
  theme_classic()

p1

ggsave(here("output/34_scatter_fit.pdf"), p1, width = 5, height = 4)
```

## Plot 2: Colored by residual

```{r plot-residuals, fig.width=5.5, fig.height=4}
resid_lim <- max(abs(plot_df$.resid), na.rm = TRUE)

p2 <- ggplot(plot_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = .resid)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#d6604d",
                        midpoint = 0, limits = c(-resid_lim, resid_lim),
                        name = "Residual") +
  labs(
    x = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y = "TE LFC 1hr hypoxia (sictrl/si3d)"
  ) +
  theme_classic()

p2

ggsave(here("output/34_scatter_residuals.pdf"), p2, width = 5.5, height = 4)
```

## Significant mRNA subset

Genes with LFC > 0.5 and padj < 0.05 in normoxia or hypoxia (promoted only).

```{r sig-subset}
sig_df <- plot_df %>%
  filter(
    (te_lfc_normoxia > 0.5 & te_padj_normoxia < 0.05) |
    (te_lfc_hypoxia  > 0.5 & te_padj_hypoxia  < 0.05)
  ) %>%
  dplyr::select(gene_id, symbol,
                te_lfc_normoxia, te_padj_normoxia,
                te_lfc_hypoxia,  te_padj_hypoxia,
                fitted = .fitted, residual = .resid)

cat("Significant genes:", nrow(sig_df), "\n")

write_csv(sig_df, here("output/34_sig_genes_residuals.csv"))
```

## Plot 3: Significant genes with residual segments

Fit line is from the full-dataset model (unchanged). Segments connect each point to its fitted value.

```{r plot-sig-residuals, fig.width=5, fig.height=4}
resid_lim_sig <- max(abs(sig_df$residual), na.rm = TRUE)

p3 <- ggplot(sig_df, aes(te_lfc_normoxia, te_lfc_hypoxia, color = residual)) +
  geom_segment(aes(xend = te_lfc_normoxia, yend = fitted),
               linewidth = 0.3, alpha = 0.6) +
  geom_point(alpha = 0.6, size = 0.9) +
  scale_color_gradient2(low = "#2166ac", mid = "grey90", high = "#d6604d",
                        midpoint = 0, limits = c(-resid_lim_sig, resid_lim_sig),
                        name = "Residual") +
  geom_abline(slope = coefs["te_lfc_normoxia"], intercept = coefs["(Intercept)"],
              color = "black", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    x = "TE LFC normoxia (sictrl/si3d, 1+4hr regressed)",
    y = "TE LFC 1hr hypoxia (sictrl/si3d)",
    title = "Significant promoted genes"
  ) +
  theme_classic()

p3

ggsave(here("output/34_sig_residual_segments.pdf"), p3, width = 5, height = 4)
```
