---
title: "Notebook 37: RF Regression on eIF3d Residuals (relaxed filter: te_lfc > 0)"
output: html_document
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(ranger)
library(caret)
library(cowplot)
library(ggpubr)
library(broom)
```

# Overview

Same approach as notebook 35 but with a configurable gene filter (params: `lfc_min`, `padj_max`).
Default: `te_lfc > 0` in normoxia OR hypoxia, `padj < 1` (no significance filter).
Set `padj_max = 0.3` for a loose filter or `0.05` to match nb35's strict threshold.
Residuals are computed inline from the same normoxia~hypoxia linear regression as nb34.

```{r params}
feature_matrix_rds <- "feature_matrix_external_stability.rds"

# Gene filter parameters
# lfc_min:  minimum te_lfc in normoxia OR hypoxia (use 0 for any positive effect)
# padj_max: maximum padj in normoxia OR hypoxia (use 1 to disable p-value filter)
lfc_min  <- 0
padj_max <- 0.3   # e.g. set to 0.3 for loose filter, 0.05 for strict

suffix <- paste0("eif3d_residuals_lfc", lfc_min,
                 "_padj", padj_max,
                 "_external_stability")
cat("Suffix:", suffix, "\n")
```

# Compute Residuals

```{r compute_residuals}
normoxia    <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                        show_col_types = FALSE)
hypoxia_1hr <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                        show_col_types = FALSE)

plot_df <- normoxia %>%
  dplyr::select(gene_id, symbol, te_lfc_normoxia = te_lfc, te_padj_normoxia = te_padj) %>%
  inner_join(
    hypoxia_1hr %>% dplyr::select(gene_id, te_lfc_hypoxia = te_lfc, te_padj_hypoxia = te_padj),
    by = "gene_id"
  )

cat("Genes measured in both conditions:", nrow(plot_df), "\n")

lm_fit  <- lm(te_lfc_hypoxia ~ te_lfc_normoxia, data = plot_df, na.action = na.exclude)
plot_df <- augment(lm_fit, data = plot_df)

# Filter by lfc_min and padj_max (set padj_max = 1 to disable p-value filter)
residuals_df <- plot_df %>%
  filter(
    (te_lfc_normoxia > lfc_min | te_lfc_hypoxia > lfc_min),
    (te_padj_normoxia < padj_max | te_padj_hypoxia < padj_max)
  ) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  dplyr::select(gene_id_clean, symbol, outcome = .resid,
                te_lfc_normoxia, te_lfc_hypoxia)

cat(sprintf("Genes passing filter (lfc > %g, padj < %g): %d\n", lfc_min, padj_max, nrow(residuals_df)))
cat("Outcome (residual) summary:\n")
print(summary(residuals_df$outcome))
```

# Load Feature Matrix

```{r load_features}
features <- readRDS(here("output", "predictive_modeling", feature_matrix_rds)) %>%
  dplyr::select(-any_of(c("te_lfc", "te_padj", "te_lfc_bin", "symbol")))
cat("Feature matrix:", nrow(features), "genes x", ncol(features), "columns\n")
```

# Join Features and Outcome

```{r create_outcome}
model_genes <- features %>%
  inner_join(residuals_df, by = "gene_id_clean") %>%
  filter(!is.na(outcome))

cat("Genes in model (feature matrix ∩ filtered genes):", nrow(model_genes), "\n")
```

# Filter Features

```{r filter_features}
id_cols <- c("gene_id", "gene_id_clean", "gene_id.x", "gene_id.y",
             "transcript_id", "transcript_id_clean", "transcript_id_raw",
             "symbol", "outcome", "te_lfc", "te_padj", "te_lfc_bin",
             "te_lfc_normoxia", "te_lfc_hypoxia", "eif3d_target", "category")
id_pattern_cols <- colnames(model_genes)[str_detect(colnames(model_genes), "gene_id|transcript_id|^symbol$")]
id_cols      <- unique(c(id_cols, id_pattern_cols))
feature_cols <- setdiff(colnames(model_genes), id_cols)
cat("Initial feature count:", length(feature_cols), "\n")

# Use aggregate CLIP only
aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                         "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc",
                         "clip_lee_glu_dep_lfc", "clip_lee_glu_vs_cm_enrichment")
regional_clip <- feature_cols[str_detect(feature_cols, "^clip_") & !(feature_cols %in% aggregate_clip_cols)]
feature_cols  <- setdiff(feature_cols, regional_clip)

# Remove k-mer / RBP motif columns
kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
feature_cols <- feature_cols[!str_detect(feature_cols, kmer_motif_pattern)]

# Remove own-data TE features
own_te_cols  <- feature_cols[str_detect(feature_cols, "^indiv_te_|^delta_te_|^sig_")]
feature_cols <- setdiff(feature_cols, own_te_cols)

# Remove sparse features (< 50% non-NA)
sparse_cols  <- feature_cols[
  sapply(feature_cols, function(f) mean(!is.na(model_genes[[f]])) < 0.5)
]
if (length(sparse_cols) > 0) {
  feature_cols <- setdiff(feature_cols, sparse_cols)
  cat("Removed", length(sparse_cols), "sparse features\n")
}

cat("Feature count after filtering:", length(feature_cols), "\n")
```

# Train/Test Split and Imputation

```{r train_test_split}
model_data <- model_genes %>%
  dplyr::select(gene_id_clean, symbol, outcome, all_of(feature_cols))

set.seed(9)
train_idx  <- createDataPartition(model_data$outcome, p = 0.8, list = FALSE)
train_data <- model_data[train_idx, ]
test_data  <- model_data[-train_idx, ]

cat("Training set:", nrow(train_data), "\n")
cat("Test set:", nrow(test_data), "\n")

train_medians <- train_data %>%
  dplyr::select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~median(., na.rm = TRUE)))

impute_with_medians <- function(df, medians) {
  for (col in feature_cols) {
    df[[col]] <- ifelse(is.na(df[[col]]), medians[[col]], df[[col]])
  }
  df
}

train_data <- impute_with_medians(train_data, train_medians)
test_data  <- impute_with_medians(test_data,  train_medians)

# Rank outcome: percentile rank within training set; map test set via training ECDF
train_data$outcome_rank <- rank(train_data$outcome, ties.method = "average") / nrow(train_data)
ecdf_train              <- ecdf(train_data$outcome)
test_data$outcome_rank  <- ecdf_train(test_data$outcome)
```

# Train Random Forest (rank outcome)

```{r train_rf}
set.seed(9)
rf_model <- ranger(
  x          = train_data[, feature_cols],
  y          = train_data$outcome_rank,
  num.trees  = 500,
  importance = "impurity",
  seed       = 9
)

print(rf_model)
cat("OOB R²:", round(rf_model$r.squared, 4), "\n")
```

# Test Set Evaluation

```{r test_set_evaluation, fig.width=6, fig.height=5}
pred_test  <- predict(rf_model, test_data[, feature_cols])$predictions
actual     <- test_data$outcome_rank

r2_test    <- cor(pred_test, actual)^2
rmse_test  <- sqrt(mean((pred_test - actual)^2))
pearson_r  <- cor(pred_test, actual, method = "pearson")
spearman_r <- cor(pred_test, actual, method = "spearman")

cat("Test Set Metrics (rank outcome):\n")
cat("  R²:        ", round(r2_test,    4), "\n")
cat("  RMSE:      ", round(rmse_test,  4), "\n")
cat("  Pearson r: ", round(pearson_r,  4), "\n")
cat("  Spearman ρ:", round(spearman_r, 4), "\n")

scatter_df <- data.frame(actual = actual, predicted = pred_test, symbol = test_data$symbol,
                         residual = test_data$outcome)

p_scatter <- ggplot(scatter_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  geom_smooth(method = "lm", se = TRUE, color = "#E69F00", linewidth = 0.8) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5, size = 3.5,
           label = paste0("R² = ", round(r2_test, 3),
                          "  |  Pearson r = ", round(pearson_r, 3),
                          "  |  Spearman ρ = ", round(spearman_r, 3))) +
  labs(
    title = "RF Regression: Predicted vs Observed Residual Rank",
    x     = "Observed residual percentile rank",
    y     = "Predicted residual percentile rank"
  ) +
  theme_cowplot()

print(p_scatter)
ggsave(here("plots", paste0("rf_regression_scatter_", suffix, ".pdf")),
       plot = p_scatter, width = 7, height = 5)
```

# Model Fitness Plots

```{r precision_at_k, fig.width=7, fig.height=5}
resid_threshold <- quantile(scatter_df$residual, 0.75)
baseline <- mean(scatter_df$residual >= resid_threshold)

prec_df <- scatter_df %>%
  arrange(desc(predicted)) %>%
  mutate(
    rank_k    = row_number(),
    precision = cumsum(residual >= resid_threshold) / rank_k
  )

n_positive  <- sum(scatter_df$residual >= resid_threshold)
prec_at_n   <- prec_df$precision[prec_df$rank_k == n_positive]
auc_pk      <- mean(prec_df$precision) / baseline

cat("Positive set size (top quartile):", n_positive, "\n")
cat(sprintf("Precision@%d: %.3f (baseline = %.2f, fold-enrichment = %.2fx)\n",
            n_positive, prec_at_n, baseline, prec_at_n / baseline))
cat(sprintf("Normalized AUC-PK: %.3f (1.0 = random)\n", auc_pk))

p_preck <- ggplot(prec_df %>% filter(rank_k >= 10), aes(x = rank_k, y = precision)) +
  geom_line(color = "#E69F00", alpha = 0.4, linewidth = 0.6) +
  geom_smooth(color = "#E69F00", se = FALSE, linewidth = 1.1, method = "loess", span = 0.3) +
  geom_hline(yintercept = baseline, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = n_positive, linetype = "dotted", color = "grey50") +
  annotate("text", x = Inf, y = baseline, hjust = 1.1, vjust = -0.5,
           label = paste0("Random baseline = ", round(baseline, 2)),
           size = 3.5, color = "grey50") +
  annotate("text", x = n_positive, y = Inf, hjust = -0.1, vjust = 1.5,
           label = paste0("Precision@", n_positive, " = ", round(prec_at_n, 2)),
           size = 3.5, color = "grey30") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = paste0("AUC-PK = ", round(auc_pk, 2), "x baseline"),
           size = 3.5, color = "grey30") +
  labs(title = "Precision@K: RF Regression Residual Rank",
       x = "K (genes ranked by predicted residual rank, descending)",
       y = "Precision (fraction in top quartile of residuals)") +
  theme_classic(base_size = 13)

print(p_preck)
ggsave(here("plots", paste0("rf_regression_precision_at_k_", suffix, ".pdf")),
       plot = p_preck, width = 7, height = 5)
```

```{r quartile_boxplot, fig.width=6, fig.height=5}
scatter_df <- scatter_df %>%
  mutate(pred_quartile = ntile(predicted, 4))

p_quartile <- ggplot(scatter_df, aes(x = factor(pred_quartile), y = actual)) +
  geom_boxplot(fill = "grey90", color = "black", outlier.size = 0.8) +
  labs(title = "Actual residual rank by predicted quartile",
       x = "Predicted quartile (1 = normoxia-specific \u2192 4 = hypoxia-specific)",
       y = "Observed residual percentile rank") +
  theme_classic(base_size = 13)

print(p_quartile)
ggsave(here("plots", paste0("rf_regression_quartile_boxplot_", suffix, ".pdf")),
       plot = p_quartile, width = 6, height = 5)
```

# 5-Fold Cross-Validation

```{r cross_validation}
set.seed(9)
folds <- createFolds(train_data$outcome_rank, k = 5, list = TRUE)

cv_results <- lapply(seq_along(folds), function(i) {
  cat("Fold", i, "...\n")
  test_idx <- folds[[i]]
  train_cv <- train_data[-test_idx, ]
  val_cv   <- train_data[test_idx, ]

  cv_medians <- train_cv %>%
    dplyr::select(all_of(feature_cols)) %>%
    summarise(across(everything(), ~median(., na.rm = TRUE)))
  train_cv <- impute_with_medians(train_cv, cv_medians)
  val_cv   <- impute_with_medians(val_cv,   cv_medians)

  train_cv$outcome_rank <- rank(train_cv$outcome, ties.method = "average") / nrow(train_cv)
  val_cv$outcome_rank   <- ecdf(train_cv$outcome)(val_cv$outcome)

  rf_cv     <- ranger(x = train_cv[, feature_cols], y = train_cv$outcome_rank,
                      num.trees = 500, seed = 9)
  pred_cv   <- predict(rf_cv, val_cv[, feature_cols])$predictions
  actual_cv <- val_cv$outcome_rank

  data.frame(
    fold     = i,
    R2       = cor(pred_cv, actual_cv)^2,
    RMSE     = sqrt(mean((pred_cv - actual_cv)^2)),
    Pearson  = cor(pred_cv, actual_cv, method = "pearson"),
    Spearman = cor(pred_cv, actual_cv, method = "spearman")
  )
})

cv_results_df <- bind_rows(cv_results)
rownames(cv_results_df) <- NULL
print(cv_results_df)

cat("\nMean +/- SD:\n")
for (metric in c("R2", "RMSE", "Pearson", "Spearman")) {
  vals <- cv_results_df[[metric]]
  cat(metric, ":", round(mean(vals, na.rm = TRUE), 4), "+/-",
      round(sd(vals, na.rm = TRUE), 4), "\n")
}
```

```{r cv_spearman_plot, fig.width=5, fig.height=4}
p_cv <- ggplot(cv_results_df, aes(x = factor(fold), y = Spearman)) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = mean(cv_results_df$Spearman),
             linetype = "dashed", color = "#E69F00", linewidth = 0.8) +
  annotate("text", x = Inf, y = mean(cv_results_df$Spearman), hjust = 1.1, vjust = -0.5,
           label = paste0("Mean = ", round(mean(cv_results_df$Spearman), 3),
                          " \u00b1 ", round(sd(cv_results_df$Spearman), 3)),
           size = 3.5, color = "#E69F00") +
  labs(title = "5-fold CV: Spearman \u03c1 per fold",
       x = "Fold", y = "Spearman \u03c1 (predicted vs observed rank)") +
  theme_classic(base_size = 13)

print(p_cv)
ggsave(here("plots", paste0("rf_regression_cv_spearman_", suffix, ".pdf")),
       plot = p_cv, width = 5, height = 4)
```

# Feature Importance

```{r feature_importance, fig.width=10, fig.height=10}
importance_df <- data.frame(
  feature    = names(rf_model$variable.importance),
  importance = rf_model$variable.importance
) %>%
  arrange(desc(importance)) %>%
  mutate(
    feature_clean = feature %>%
      str_replace("^log2_", "log2 ") %>%
      str_replace("_length$", " length") %>%
      str_replace("_gc$", " GC%")
  )

cat("Top 30 Features:\n")
print(importance_df %>% head(30) %>% dplyr::select(feature, importance))

p_importance <- importance_df %>%
  head(30) %>%
  ggplot(aes(x = reorder(feature_clean, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "#E69F00") +
  coord_flip() +
  labs(
    title    = "Top 30 Features: Residual Regression (eIF3d hypoxia vs normoxia)",
    subtitle = "Variable Importance (Impurity) — relaxed filter (te_lfc > 0)",
    x = "", y = "Variable Importance"
  ) +
  theme_cowplot()

print(p_importance)
ggsave(here("plots", paste0("rf_regression_feature_importance_", suffix, ".pdf")),
       plot = p_importance, width = 10, height = 10)
```

## Feature Category Summary

```{r feature_categories, fig.width=8, fig.height=5}
importance_df <- importance_df %>%
  mutate(
    category = case_when(
      str_detect(feature, "^struct_")                          ~ "RNA Structure",
      str_detect(feature, "^orfquant_")                       ~ "ORFquant",
      str_detect(feature, "^clip_")                           ~ "CLIP Binding",
      str_detect(feature, "^lit_")                            ~ "Literature Signatures",
      str_detect(feature, "length|Length")                     ~ "Region Lengths",
      str_detect(feature, "_gc$|^gc3$")                       ~ "GC Content",
      str_detect(feature, "^4e_resistance|^eif4e_ip|^dap5_")  ~ "4E / DAP5",
      str_detect(feature, "^indiv_te_|^delta_te_")            ~ "Translational Efficiency",
      str_detect(feature, "^codon_|^cai_|^tai_|^csc_")        ~ "Codon Optimality",
      TRUE                                                     ~ "Other"
    )
  )

category_summary <- importance_df %>%
  group_by(category) %>%
  summarise(total_importance = sum(importance), n_features = n()) %>%
  arrange(desc(total_importance))

print(category_summary)

p_cat <- category_summary %>%
  ggplot(aes(x = reorder(category, total_importance), y = total_importance)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  coord_flip() +
  labs(
    title = "Feature Category Importance: Residual Regression",
    x = "", y = "Total Importance"
  ) +
  theme_cowplot()

print(p_cat)
ggsave(here("plots", paste0("rf_regression_category_importance_", suffix, ".pdf")),
       plot = p_cat, width = 8, height = 5)
```

## Directionality & Signed-Importance Lollipop

```{r compute_directionality}
directionality <- importance_df %>%
  rowwise() %>%
  mutate(
    correlation = tryCatch(
      cor(train_data[[feature]], train_data$outcome_rank, use = "complete.obs"),
      error = function(e) NA_real_
    )
  ) %>%
  ungroup() %>%
  mutate(
    direction = case_when(
      is.na(correlation) ~ NA_character_,
      correlation > 0    ~ "Positively correlated with residual rank",
      TRUE               ~ "Negatively correlated with residual rank"
    )
  ) %>%
  arrange(desc(importance))

cat("Positively correlated:", sum(directionality$direction == "Positively correlated with residual rank", na.rm = TRUE), "\n")
cat("Negatively correlated:", sum(directionality$direction == "Negatively correlated with residual rank", na.rm = TRUE), "\n")
```

```{r plot_lollipop, fig.width=10, fig.height=10}
lollipop_data <- directionality %>%
  filter(!is.na(correlation)) %>%
  head(30) %>%
  mutate(
    signed_importance = importance * sign(correlation),
    broad_category = case_when(
      category == "4E / DAP5" & str_detect(feature, "^dap5_") ~ "Sensitivity to DAP5",
      category == "4E / DAP5"                                  ~ "Sensitivity to eIF4E",
      category %in% c("CLIP Binding", "Literature Signatures") ~ "eIF3 binding evidence",
      category == "Translational Efficiency"                   ~ "Translation in hypoxia",
      TRUE                                                     ~ "Intrinsic mRNA features"
    ),
    broad_category = factor(broad_category, levels = c(
      "Intrinsic mRNA features", "eIF3 binding evidence",
      "Sensitivity to DAP5", "Sensitivity to eIF4E", "Translation in hypoxia"
    )),
    feature_clean = fct_reorder(feature_clean, signed_importance)
  )

category_colors <- c(
  "Intrinsic mRNA features" = "#3D9C60",
  "eIF3 binding evidence"   = "#E6A817",
  "Sensitivity to DAP5"     = "#D668C6",
  "Sensitivity to eIF4E"    = "#4A7FC1",
  "Translation in hypoxia"  = "#E07030"
)

p_lollipop <- ggdotchart(
  lollipop_data,
  x        = "feature_clean",
  y        = "signed_importance",
  color    = "broad_category",
  palette  = category_colors,
  sorting  = "none",
  add      = "segments",
  rotate   = TRUE,
  dot.size = 4,
  xlab     = "",
  ylab     = "Signed Importance (positive = higher residual rank)",
  title    = "Top 30 Features: Signed Importance",
  subtitle = "Residual rank regression — relaxed filter (te_lfc > 0)"
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.title = element_blank())

print(p_lollipop)
ggsave(here("plots", paste0("rf_regression_lollipop_", suffix, ".pdf")),
       plot = p_lollipop, width = 10, height = 10)
```

```{r}

```