---
title: "38 — RSCU Analysis: eIF3d-Promoted vs Control (Hypoxia 1hr)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

RSCU (Relative Synonymous Codon Usage) for eIF3d-promoted genes vs size-matched
negative controls at 1hr hypoxia.

RSCU_j = count_j * n_syn / AA_total — where n_syn is the number of synonymous
codons for that amino acid. RSCU = 1 means uniform usage; > 1 preferred; < 1
disfavored.

**Key design choice:** RSCU is computed **per transcript** and then averaged
(mean ± SE) across transcripts, giving each gene equal weight regardless of
CDS length. This avoids length-composition confounding if the two gene sets
differ in CDS length distribution (as is likely).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
```

---

# Section 1: Setup — CDS Sequences & Codon Counts

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  ct  <- codonTable(seqs)
  mat <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

cds_frame1  <- shift_frame(cds_frame0, 1)
cds_frame2  <- shift_frame(cds_frame0, 2)
counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

# Section 2: Gene Set Loading

```{r gene_sets}
promotes_raw <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids <- promotes_raw$ensembl_gene
cat("Hypoxia 1hr promoted genes:", length(promotes_ids), "\n")

neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

neg_filtered <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids)
set.seed(9)
neg_ctrl_ids <- sample(neg_filtered$ensembl_gene,
                       size = length(promotes_ids), replace = FALSE)
cat("Neg ctrl (size-matched):", length(neg_ctrl_ids), "\n")

promotes_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids) %>%
  pull(transcript_id_clean)
neg_ctrl_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids) %>%
  pull(transcript_id_clean)

promotes_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat0)]
neg_ctrl_tx <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat0)]
cat("Promotes transcripts:", length(promotes_tx),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx), "\n")
```

---

# Section 3: Per-Transcript RSCU

```{r rscu_function}
# Compute per-transcript RSCU matrix from a codon counts matrix.
# RSCU_j = count_j * n_syn / aa_total (NA when aa_total == 0).
compute_rscu_matrix <- function(counts_mat) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(aa_to_codons)) {
    codons  <- intersect(aa_to_codons[[aa]], colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt  <- counts_mat[, codons, drop = FALSE]
    aa_tot   <- rowSums(sub_cnt)
    n_syn    <- length(codons)
    valid    <- aa_tot > 0
    if (any(valid)) {
      rscu_mat[valid, codons] <- sub_cnt[valid, , drop = FALSE] * n_syn / aa_tot[valid]
    }
  }
  rscu_mat
}
```

```{r compute_rscu}
rscu_mat0 <- compute_rscu_matrix(counts_mat0)
cat("RSCU matrix computed:", nrow(rscu_mat0), "×", ncol(rscu_mat0), "\n")
cat("Range of non-NA RSCU values:",
    round(min(rscu_mat0, na.rm = TRUE), 3), "—",
    round(max(rscu_mat0, na.rm = TRUE), 3), "\n")
```

```{r rscu_summary_function}
# Summarize per-transcript RSCU values into group-level stats.
summarize_rscu <- function(rscu_mat, tx_ids) {
  tx_ids <- tx_ids[tx_ids %in% rownames(rscu_mat)]
  sub    <- rscu_mat[tx_ids, sense_codons, drop = FALSE]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(sub))
    map_dfr(codons, function(codon) {
      vals <- sub[, codon]
      vals <- vals[!is.na(vals)]
      tibble(
        amino_acid  = aa,
        codon       = codon,
        mean_rscu   = mean(vals),
        median_rscu = median(vals),
        se_rscu     = if (length(vals) > 1) sd(vals) / sqrt(length(vals)) else NA_real_,
        n           = length(vals)
      )
    })
  })
}

rscu_promotes <- summarize_rscu(rscu_mat0, promotes_tx)
rscu_neg      <- summarize_rscu(rscu_mat0, neg_ctrl_tx)

cat("RSCU summary — promotes: ", nrow(rscu_promotes), "codon-rows\n")
```

```{r rscu_delta}
rscu_combined <- rscu_promotes %>%
  dplyr::select(amino_acid, codon, promotes_mean = mean_rscu,
                promotes_se = se_rscu, promotes_median = median_rscu,
                promotes_n = n) %>%
  left_join(
    rscu_neg %>% dplyr::select(amino_acid, codon, neg_mean = mean_rscu,
                               neg_se = se_rscu, neg_median = median_rscu,
                               neg_n = n),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta_rscu = promotes_mean - neg_mean)

# Wilcoxon test per codon (per-transcript RSCU)
wilcox_df <- imap_dfr(aa_to_codons, function(codons, aa) {
  codons <- intersect(codons, colnames(rscu_mat0))
  map_dfr(codons, function(codon) {
    p_vals <- rscu_mat0[promotes_tx[promotes_tx %in% rownames(rscu_mat0)], codon]
    n_vals <- rscu_mat0[neg_ctrl_tx[neg_ctrl_tx %in% rownames(rscu_mat0)], codon]
    p_vals <- p_vals[!is.na(p_vals)]
    n_vals <- n_vals[!is.na(n_vals)]
    if (length(p_vals) < 3 || length(n_vals) < 3) {
      return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_))
    }
    wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val = wt$p.value)
  })
}) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))

rscu_combined <- rscu_combined %>%
  left_join(wilcox_df %>% dplyr::select(amino_acid, codon, p_val, p_adj),
            by = c("amino_acid", "codon"))

cat("Top 10 codons by |Δ RSCU|:\n")
print(rscu_combined %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_rscu))) %>%
        dplyr::select(codon, amino_acid, promotes_mean, neg_mean, delta_rscu, p_adj) %>%
        head(10))
```

---

# Section 4: Visualization

```{r plot_a_sidebyside, fig.width=11, fig.height=9}
plot_df_a <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(cols = c(promotes_mean, neg_mean),
               names_to = "group", values_to = "mean_rscu") %>%
  mutate(
    se      = if_else(group == "promotes_mean", promotes_se, neg_se),
    group   = recode(group,
                     promotes_mean = "eIF3d-promoted",
                     neg_mean      = "Neg ctrl"),
    group   = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, mean_rscu)
  )

p_a <- ggplot(plot_df_a, aes(x = codon, y = mean_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = mean_rscu - se, ymax = mean_rscu + se),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "RSCU by group: eIF3d-promoted vs negative control (Hypoxia 1hr)",
    subtitle = "Mean per-transcript RSCU ± SE; dashed line = RSCU 1 (uniform usage)",
    x = "Codon", y = "Mean RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1, size = 8),
    strip.text     = element_text(face = "bold"),
    legend.position = "top"
  )
p_a
ggsave(here("plots", "rscu_sidebyside_hypoxia_1hr.pdf"), p_a, width = 11, height = 9)
```

```{r plot_b_delta, fig.width=10, fig.height=8}
plot_df_b <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta_rscu >= 0, "promotes", "neg_ctrl"),
    sig_label = case_when(
      !is.na(p_adj) & p_adj < 0.05 ~ "*",
      TRUE ~ ""
    ),
    codon = reorder(codon, delta_rscu)
  )

p_b <- ggplot(plot_df_b, aes(x = codon, y = delta_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label,
                y = delta_rscu + sign(delta_rscu) * 0.03),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ RSCU: eIF3d-promoted minus negative control (Hypoxia 1hr)",
    subtitle = "Mean per-transcript RSCU difference; * = BH-adjusted Wilcoxon q < 0.05",
    x = "Codon", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_b
ggsave(here("plots", "rscu_delta_hypoxia_1hr.pdf"), p_b, width = 10, height = 8)
```

```{r frame_ctrl_rscu, fig.width=12, fig.height=9}
rscu_mat1 <- compute_rscu_matrix(counts_mat1)
rscu_mat2 <- compute_rscu_matrix(counts_mat2)

summarize_delta_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      tibble(amino_acid = aa, codon = codon,
             delta = mean(p_vals) - mean(n_vals),
             frame = frame_label)
    })
  })
}

per_tx_sig_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    if (length(codons) < 2) return(NULL)
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      if (length(p_vals) < 3 || length(n_vals) < 3) {
        return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_, frame = frame_label))
      }
      wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
      tibble(amino_acid = aa, codon = codon, p_val = wt$p.value, frame = frame_label)
    })
  })
}

# Frame deltas
f0_delta <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(amino_acid, codon, delta = delta_rscu) %>%
  mutate(frame = "Frame 0 (coding)")
f1_delta <- summarize_delta_rscu(rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)")
f2_delta <- summarize_delta_rscu(rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")

frame_delta_df <- bind_rows(f0_delta, f1_delta, f2_delta) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

# Codon order from frame 0 delta
codon_order <- f0_delta %>%
  filter(amino_acid %in% multi_aa) %>%
  arrange(amino_acid, delta) %>%
  pull(codon)
frame_delta_df <- frame_delta_df %>%
  mutate(codon = factor(codon, levels = codon_order))

# Significance tests
sig_df <- bind_rows(
  per_tx_sig_rscu(rscu_mat0, promotes_tx, neg_ctrl_tx, "Frame 0 (coding)"),
  per_tx_sig_rscu(rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)"),
  per_tx_sig_rscu(rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    q_val = p.adjust(p_val, method = "BH"),
    sig   = !is.na(q_val) & q_val < 0.05,
    frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

f0_sig   <- sig_df %>% filter(frame == "Frame 0 (coding)") %>%
  dplyr::select(amino_acid, codon, f0_sig = sig)
ctrl_sig <- sig_df %>% filter(frame != "Frame 0 (coding)") %>%
  group_by(amino_acid, codon) %>%
  summarise(any_ctrl_sig = any(sig), .groups = "drop")
f0_labels <- f0_sig %>%
  left_join(ctrl_sig, by = c("amino_acid", "codon")) %>%
  mutate(sig_label = case_when(
    f0_sig & !any_ctrl_sig ~ "\u2726",
    f0_sig &  any_ctrl_sig ~ "\u00b7",
    TRUE                   ~ ""
  )) %>%
  dplyr::select(amino_acid, codon, sig_label) %>%
  mutate(frame = factor("Frame 0 (coding)",
                        levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

aa_any_f0 <- f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
aa_clean  <- f0_labels %>% filter(sig_label == "\u2726") %>% pull(amino_acid) %>% unique()

frame_delta_df <- frame_delta_df %>%
  mutate(aa_label = ifelse(amino_acid %in% aa_any_f0,
                           paste0(aa_label, " *"), aa_label))

aa_label_map <- frame_delta_df %>% distinct(amino_acid, aa_label)
facet_clean  <- aa_label_map %>% filter(amino_acid %in% aa_clean)

rect_layer <- if (nrow(facet_clean) > 0) {
  geom_rect(data = facet_clean,
            aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5)
} else {
  NULL
}

p_c <- ggplot(frame_delta_df, aes(x = codon, y = delta, fill = frame)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  rect_layer +
  scale_fill_manual(
    values = c("Frame 0 (coding)"  = "#E69F00",
               "Frame 1 (control)" = "#999999",
               "Frame 2 (control)" = "#CCCCCC")
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ RSCU by reading frame: frame 0 vs frameshift controls (Hypoxia 1hr)",
    subtitle = "Frames 1 & 2 control for nucleotide composition; persistent frame-0 signal = true codon preference\nBlue border: frame 0 significant, controls not (clean codon signal). *: frame 0 significant. [q < 0.05, BH Wilcoxon]",
    x = "Codon", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    strip.text        = element_text(face = "bold"),
    legend.position   = "top"
  )
p_c
ggsave(here("plots", "rscu_frame_ctrl_hypoxia_1hr.pdf"), p_c, width = 12, height = 9)
```

---

# Section 5: Top Codons Summary & GC3 Bias Check

```{r gc3_sanity_check}
# Annotate each codon with its third-position nucleotide
codon_gc3 <- tibble(
  codon          = sense_codons,
  gc3_nucleotide = substr(sense_codons, 3, 3),
  gc3_class      = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending")
)

# Compute per-transcript GC3 from counts_mat0 and compare to feature_matrix$gc3
# gc3 = sum of GC-ending codon counts / total sense codon count per transcript
gc3_codons <- codon_gc3 %>% filter(gc3_class == "GC-ending") %>% pull(codon)
gc3_codons <- intersect(gc3_codons, colnames(counts_mat0))

per_tx_sense_total <- rowSums(counts_mat0[, sense_codons[sense_codons %in% colnames(counts_mat0)]])
per_tx_gc3_count   <- rowSums(counts_mat0[, gc3_codons, drop = FALSE])
per_tx_gc3         <- per_tx_gc3_count / per_tx_sense_total

gc3_check <- tibble(
  transcript_id_clean = names(per_tx_gc3),
  gc3_computed        = per_tx_gc3
) %>%
  left_join(
    feature_matrix %>% dplyr::select(transcript_id_clean, gc3_feature = gc3),
    by = "transcript_id_clean"
  ) %>%
  filter(!is.na(gc3_feature))

cat("GC3 sanity check — correlation with feature matrix gc3:\n")
cat("  Pearson r =", round(cor(gc3_check$gc3_computed, gc3_check$gc3_feature, use = "complete.obs"), 4), "\n")
cat("  n =", nrow(gc3_check), "\n")
```

```{r gc3_scatter, fig.width=4.5, fig.height=4.5}
ggplot(gc3_check, aes(x = gc3_feature, y = gc3_computed)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.6) +
  labs(title = "GC3 sanity check",
       subtitle = "Computed from counts_mat0 vs feature_matrix$gc3",
       x = "GC3 from feature matrix", y = "GC3 computed here") +
  theme_classic(base_size = 11)
```

```{r gc3_by_group}
# Mean GC3 per group
gc3_promotes <- mean(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]], na.rm = TRUE)
gc3_neg      <- mean(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]], na.rm = TRUE)
se_gc3_prom  <- sd(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]], na.rm = TRUE) /
                sqrt(sum(!is.na(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]])))
se_gc3_neg   <- sd(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]], na.rm = TRUE) /
                sqrt(sum(!is.na(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]])))

cat("Mean GC3 — promotes:", round(gc3_promotes, 4), "± SE", round(se_gc3_prom, 4), "\n")
cat("Mean GC3 — neg ctrl:", round(gc3_neg, 4),      "± SE", round(se_gc3_neg, 4), "\n")
cat("Δ GC3 (promotes - neg ctrl):", round(gc3_promotes - gc3_neg, 4), "\n")
```

```{r top_codons_summary}
# Top codons by |Δ RSCU| with GC3 annotation
top_codons <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_nucleotide, gc3_class),
            by = "codon") %>%
  arrange(desc(abs(delta_rscu))) %>%
  dplyr::select(codon, amino_acid, promotes_mean, neg_mean,
                delta_rscu, p_adj, gc3_nucleotide, gc3_class) %>%
  mutate(across(c(promotes_mean, neg_mean, delta_rscu), ~ round(.x, 3)),
         p_adj = signif(p_adj, 3)) %>%
  head(20)

cat("Top 20 codons by |Δ mean RSCU| (multi-synonymous AAs only):\n")
print(top_codons)

# Directional summary for top 10
top10_promotes_enriched <- rscu_combined %>%
  filter(amino_acid %in% multi_aa, delta_rscu > 0) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  arrange(desc(delta_rscu)) %>%
  head(10)

cat("\nOf the top 10 promotes-enriched codons:\n")
print(table(top10_promotes_enriched$gc3_class))
cat("(If predominantly GC-ending → GC bias likely driving signal;\n",
    "mixed/AT-enriched → evidence of translational selection)\n")
```

```{r rscu_vs_gc3, fig.width=5.5, fig.height=5}
# Δ RSCU vs third-position nucleotide — tests whether GC3 bias explains the signal.
# Each point = one codon from a multi-synonymous amino acid.
# Points are labelled for |Δ RSCU| > 0.15 or BH q < 0.05.

plot_df_gc3 <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_nucleotide, gc3_class),
            by = "codon") %>%
  mutate(
    gc3_nucleotide = factor(gc3_nucleotide, levels = c("A", "T", "G", "C")),
    sig            = !is.na(p_adj) & p_adj < 0.05,
    label_codon    = if_else(abs(delta_rscu) > 0.15 | sig, codon, NA_character_)
  )

# Wilcoxon test: GC-ending vs AT-ending Δ RSCU
gc_vals <- plot_df_gc3 %>% filter(gc3_class == "GC-ending") %>% pull(delta_rscu)
at_vals <- plot_df_gc3 %>% filter(gc3_class == "AT-ending") %>% pull(delta_rscu)
wt_gc3  <- wilcox.test(gc_vals, at_vals, exact = FALSE)
gc3_p_label <- paste0("Wilcoxon p = ", signif(wt_gc3$p.value, 2),
                       "\n(GC-ending vs AT-ending)")

# x-axis jitter positions and colours
nt_colours <- c(A = "#56B4E9", T = "#009E73", G = "#E69F00", C = "#CC79A7")

p_gc3 <- ggplot(plot_df_gc3,
                aes(x = gc3_nucleotide, y = delta_rscu, colour = gc3_nucleotide)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_jitter(aes(shape = sig), width = 0.2, size = 2.2, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 20, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = nt_colours, guide = "none") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "ns", `TRUE` = "q < 0.05"),
                     name = "BH Wilcoxon") +
  annotate("text", x = 4.4, y = max(plot_df_gc3$delta_rscu, na.rm = TRUE),
           label = gc3_p_label, hjust = 1, vjust = 1, size = 3, colour = "grey30") +
  labs(
    title    = "\u0394 RSCU vs codon third-position nucleotide (Hypoxia 1hr)",
    subtitle = "Each point = one codon (multi-synonymous AAs only)",
    x = "Third-position nucleotide", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_gc3
ggsave(here("plots", "rscu_delta_vs_gc3_hypoxia_1hr.pdf"), p_gc3, width = 5.5, height = 5)
```

```{r rscu_vs_csc_weight, fig.width=5.5, fig.height=5}
# Load per-codon iCodon CSC weights and check whether AT-ending codons enriched
# in promotes are selectively the stabilizing (high-CSC) ones.
e <- new.env()
data(list = iCodon:::global_vars_data, package = "iCodon", envir = e)
codon_csc_weights <- e$human_optimality %>%
  rename(csc_weight = optimality)

plot_df_csc <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  left_join(codon_csc_weights %>% dplyr::select(codon, csc_weight), by = "codon") %>%
  mutate(
    sig        = !is.na(p_adj) & p_adj < 0.05,
    label_codon = if_else(abs(delta_rscu) > 0.15 | sig, codon, NA_character_)
  )

# Correlation within AT-ending and GC-ending codons separately
cor_at <- cor(
  filter(plot_df_csc, gc3_class == "AT-ending")$csc_weight,
  filter(plot_df_csc, gc3_class == "AT-ending")$delta_rscu,
  use = "complete.obs"
)
cor_gc <- cor(
  filter(plot_df_csc, gc3_class == "GC-ending")$csc_weight,
  filter(plot_df_csc, gc3_class == "GC-ending")$delta_rscu,
  use = "complete.obs"
)
cat("Pearson r (Δ RSCU ~ CSC weight) — AT-ending:", round(cor_at, 3),
    "| GC-ending:", round(cor_gc, 3), "\n")

p_csc <- ggplot(plot_df_csc,
                aes(x = csc_weight, y = delta_rscu, colour = gc3_class)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_point(aes(shape = sig), size = 2.2, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.7, alpha = 0.15) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 20, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(
    values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
    name   = "Third-position"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "ns", `TRUE` = "q < 0.05"),
                     name   = "BH Wilcoxon") +
  annotate("text", x = min(plot_df_csc$csc_weight, na.rm = TRUE),
           y = max(plot_df_csc$delta_rscu, na.rm = TRUE),
           label = paste0("r (AT) = ", round(cor_at, 2),
                          "\nr (GC) = ", round(cor_gc, 2)),
           hjust = 0, vjust = 1, size = 3, colour = "grey30") +
  labs(
    title    = "\u0394 RSCU vs per-codon iCodon CSC weight (Hypoxia 1hr)",
    subtitle = "Tests whether promotes-enriched AT-ending codons are selectively high-CSC\nPositive r (AT-ending) = AT-rich promotes targets use stabilizing codons",
    x = "iCodon CSC weight (human)", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_csc
ggsave(here("plots", "rscu_delta_vs_csc_weight_hypoxia_1hr.pdf"),
       p_csc, width = 5.5, height = 5)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
