---
title: "38 — RSCU Analysis: eIF3d-Promoted vs Control (Hypoxia 1hr)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

RSCU (Relative Synonymous Codon Usage) for eIF3d-promoted genes vs size-matched
negative controls at 1hr hypoxia.

RSCU_j = count_j * n_syn / AA_total — where n_syn is the number of synonymous
codons for that amino acid. RSCU = 1 means uniform usage; > 1 preferred; < 1
disfavored.

**Key design choice:** RSCU is computed **per transcript** and then averaged
(mean ± SE) across transcripts, giving each gene equal weight regardless of
CDS length. This avoids length-composition confounding if the two gene sets
differ in CDS length distribution (as is likely).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(iCodon)
library(ggrepel)
```

---

# Section 1: Setup — CDS Sequences & Codon Counts

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  ct  <- codonTable(seqs)
  mat <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

cds_frame1  <- shift_frame(cds_frame0, 1)
cds_frame2  <- shift_frame(cds_frame0, 2)
counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

codon_gc3 <- tibble(
  codon          = sense_codons,
  gc3_nucleotide = substr(sense_codons, 3, 3),
  gc3_class      = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending")
)

tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

# Section 2: Gene Set Loading

```{r gene_sets}
promotes_raw <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)
promotes_ids <- promotes_raw$ensembl_gene
cat("Hypoxia 1hr promoted genes:", length(promotes_ids), "\n")

neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

neg_filtered <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids)
# Single draw — any codon-level signal that differs between this draw and promotes
# could look real by chance. A follow-up analysis should repeat this sampling
# 100–1000 times and report an empirical p-value (fraction of random draws where
# |Δ RSCU| ≥ observed) to confirm the most significant hits are not seed-specific.
set.seed(9)
neg_ctrl_ids <- sample(neg_filtered$ensembl_gene,
                       size = length(promotes_ids), replace = FALSE)
cat("Neg ctrl (size-matched, single draw seed=9):", length(neg_ctrl_ids), "\n")

promotes_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids) %>%
  pull(transcript_id_clean)
neg_ctrl_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids) %>%
  pull(transcript_id_clean)

promotes_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat0)]
neg_ctrl_tx <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat0)]
cat("Promotes transcripts:", length(promotes_tx),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx), "\n")
```

---

# Section 3: Per-Transcript RSCU

```{r rscu_function}
# Compute per-transcript RSCU matrix from a codon counts matrix.
# RSCU_j = count_j * n_syn / aa_total (NA when aa_total == 0).
compute_rscu_matrix <- function(counts_mat) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(aa_to_codons)) {
    codons  <- intersect(aa_to_codons[[aa]], colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt  <- counts_mat[, codons, drop = FALSE]
    aa_tot   <- rowSums(sub_cnt)
    n_syn    <- length(codons)
    valid    <- aa_tot > 0
    if (any(valid)) {
      rscu_mat[valid, codons] <- sub_cnt[valid, , drop = FALSE] * n_syn / aa_tot[valid]
    }
  }
  rscu_mat
}
```

```{r compute_rscu}
rscu_mat0 <- compute_rscu_matrix(counts_mat0)
cat("RSCU matrix computed:", nrow(rscu_mat0), "×", ncol(rscu_mat0), "\n")
cat("Range of non-NA RSCU values:",
    round(min(rscu_mat0, na.rm = TRUE), 3), "—",
    round(max(rscu_mat0, na.rm = TRUE), 3), "\n")
```

---

# RSCUcaller Formula Verification

RSCUcaller (github.com/Mordziarz/RSCUcaller) could not be installed in this
environment (requires ComplexHeatmap ≥ 2.20.0; installed: 2.18.0). Its source
shows it delegates to `seqinr::uco()` and applies:

`RSCU_i = n_i × (count_i + pseudo) / (total_AA + pseudo × n_i)`

with pseudo = 1 by default. At pseudo = 0 this is identical to
`compute_rscu_matrix()`. The chunk below implements the RSCUcaller formula
directly (pseudo = 1) on a random sample of transcripts and compares per-mRNA
RSCU values with our custom implementation to confirm they are in numerical
agreement (expected r ≈ 1, small offset from the pseudocount).

```{r rscucaller_validation, fig.width=5.5, fig.height=5}
set.seed(42)
val_tx <- sample(rownames(counts_mat0), min(300, nrow(counts_mat0)))

# RSCUcaller formula: RSCU_i = n_i * (count_i + pseudo) / (total_AA + pseudo * n_i)
# At pseudo = 0 this is algebraically identical to compute_rscu_matrix().
# RSCUcaller's default is pseudo = 1 (shrinks sparse amino acids toward 1).
compute_rscu_pkg <- function(counts_mat, pseudo = 0) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(aa_to_codons)) {
    codons  <- intersect(aa_to_codons[[aa]], colnames(counts_mat))
    if (length(codons) == 0) next
    n_syn   <- length(codons)
    sub_cnt <- counts_mat[, codons, drop = FALSE]
    aa_tot  <- rowSums(sub_cnt)
    valid   <- aa_tot > 0
    if (!any(valid)) next
    rscu_mat[valid, codons] <-
      (sub_cnt[valid, , drop = FALSE] + pseudo) * n_syn /
      (aa_tot[valid] + pseudo * n_syn)
  }
  rscu_mat
}

rscu_ours_val  <- rscu_mat0[val_tx, ]
rscu_pkg0_val  <- compute_rscu_pkg(counts_mat0[val_tx, ], pseudo = 0)  # identical formula
rscu_pkg1_val  <- compute_rscu_pkg(counts_mat0[val_tx, ], pseudo = 1)  # RSCUcaller default

# Per-codon Pearson r for both pseudocount settings — extract to plain vectors
sense_in_mat <- intersect(sense_codons, colnames(rscu_ours_val))
codon_cor <- function(mat_a, mat_b, codons) {
  vapply(codons, function(cdn) {
    x    <- as.numeric(mat_a[, cdn])
    y    <- as.numeric(mat_b[, cdn])
    keep <- !is.na(x) & !is.na(y)
    if (sum(keep) < 5 || sd(x[keep]) == 0 || sd(y[keep]) == 0) return(NA_real_)
    cor(x[keep], y[keep])
  }, numeric(1))
}
cor_p0 <- codon_cor(rscu_pkg0_val, rscu_ours_val, sense_in_mat)
cor_p1 <- codon_cor(rscu_pkg1_val, rscu_ours_val, sense_in_mat)

cat("RSCUcaller formula vs custom — per-mRNA RSCU correlation (n =",
    length(val_tx), "transcripts):\n")
cat("  pseudo = 0 (identical formula): median r =",
    round(median(cor_p0, na.rm = TRUE), 4),
    "| min r =", round(min(cor_p0, na.rm = TRUE), 4), "\n")
cat("  pseudo = 1 (RSCUcaller default): median r =",
    round(median(cor_p1, na.rm = TRUE), 4),
    "| min r =", round(min(cor_p1, na.rm = TRUE), 4), "\n")
cat("  pseudo=0 should be r = 1.0 exactly; any deviation is floating-point only.\n")
cat("  pseudo=1 shrinks sparse AAs toward 1 — lower r is the intended smoothing,\n")
cat("  not a formula discrepancy.\n\n")

# Visual: scatter for 6 codons, pseudo=0 (should be identity line) and pseudo=1
example_codons <- intersect(c("CTG", "GAA", "GGC", "AGA", "CAG", "ACC"), sense_in_mat)

# Build scatter data with explicit intermediate vectors — no nested closures
scatter_rows <- vector("list", length(example_codons) * 2)
k <- 0L
for (cdn in example_codons) {
  ours <- as.numeric(rscu_ours_val[, cdn])
  p0   <- as.numeric(rscu_pkg0_val[, cdn])
  p1   <- as.numeric(rscu_pkg1_val[, cdn])
  k <- k + 1L
  scatter_rows[[k]] <- tibble(codon = cdn, our_rscu = ours, pkg_rscu = p0,
                               pseudo = "pseudo = 0 (identical)")
  k <- k + 1L
  scatter_rows[[k]] <- tibble(codon = cdn, our_rscu = ours, pkg_rscu = p1,
                               pseudo = "pseudo = 1 (RSCUcaller default)")
}
scatter_df <- bind_rows(scatter_rows) %>%
  filter(!is.na(our_rscu), !is.na(pkg_rscu))

ggplot(scatter_df, aes(x = our_rscu, y = pkg_rscu, colour = pseudo)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linewidth = 0.5,
              linetype = "dashed") +
  geom_point(alpha = 0.25, size = 0.7) +
  scale_colour_manual(values = c("pseudo = 0 (identical)"          = "#0072B2",
                                  "pseudo = 1 (RSCUcaller default)" = "#E69F00"),
                      name = NULL) +
  facet_wrap(~ codon, scales = "free", nrow = 2) +
  labs(
    title    = "RSCUcaller formula vs custom per-mRNA RSCU",
    subtitle = paste0("n = ", length(val_tx), " transcripts. Dashed = identity.\n",
                      "pseudo=0: algebraically identical (r ≈ 1.0). ",
                      "pseudo=1: RSCUcaller default — shrinks sparse AAs toward 1."),
    x = "Custom RSCU (this notebook)", y = "RSCUcaller RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "top")
```

---

# Section 3b: Background-Corrected RSCU

Per-transcript RSCU relative to the observed codon usage of all ~7 k expressed
mRNAs in the feature matrix. Instead of assuming uniform synonymous codon usage
(1/n_syn), the expected frequency for codon j within amino acid AA is the
fraction of all expressed-mRNA counts of AA that fall on codon j.

bg_RSCU_j = (count_j_tx / aa_total_tx) / (Σ_all_tx count_j / Σ_all_tx aa_total)

bg_RSCU = 1 → same rate as the transcriptome background
bg_RSCU > 1 → over-used relative to background
bg_RSCU < 1 → under-used relative to background

```{r bg_rscu_background_props}
# Background = mean per-transcript codon proportion across all expressed mRNAs.
# Each gene gets equal weight regardless of CDS length, consistent with the
# per-transcript averaging used throughout this notebook. (Pooling raw colSums
# would let long transcripts dominate the background.)
bg_props <- imap(aa_to_codons, function(codons, aa) {
  codons  <- intersect(codons, colnames(counts_mat0))
  if (length(codons) == 0) return(NULL)
  sub_cnt <- counts_mat0[, codons, drop = FALSE]
  aa_tot  <- rowSums(sub_cnt)
  valid   <- aa_tot > 0
  if (!any(valid)) return(NULL)
  # Per-transcript proportion for each codon, then average across transcripts
  prop_mat <- sub_cnt[valid, , drop = FALSE] / aa_tot[valid]
  colMeans(prop_mat)
})
bg_props <- Filter(Negate(is.null), bg_props)

cat("Background proportions computed for", length(bg_props), "amino acids\n")

# Sanity: each AA's props should sum to 1
prop_sums <- sapply(bg_props, sum)
cat("Background proportions sum check (should all be ~1):",
    round(range(prop_sums), 4), "\n")
```

```{r compute_bg_rscu_function}
compute_bg_rscu_matrix <- function(counts_mat, bg_props) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(bg_props)) {
    props  <- bg_props[[aa]]
    codons <- intersect(names(props), colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt <- counts_mat[, codons, drop = FALSE]
    aa_tot  <- rowSums(sub_cnt)
    valid   <- aa_tot > 0
    for (codon in codons) {
      bp <- props[codon]
      if (bp > 0 && any(valid)) {
        rscu_mat[valid, codon] <- (sub_cnt[valid, codon] / aa_tot[valid]) / bp
      }
    }
  }
  rscu_mat
}
```

```{r compute_bg_rscu}
bg_rscu_mat0 <- compute_bg_rscu_matrix(counts_mat0, bg_props)
cat("Background RSCU matrix:", nrow(bg_rscu_mat0), "×", ncol(bg_rscu_mat0), "\n")
cat("Range:", round(min(bg_rscu_mat0, na.rm = TRUE), 3), "—",
    round(max(bg_rscu_mat0, na.rm = TRUE), 3), "\n")

# Sanity: mean bg_RSCU across all transcripts per codon should be ≈ 1
bg_mean_per_codon <- colMeans(bg_rscu_mat0, na.rm = TRUE)
bg_mean_per_codon <- bg_mean_per_codon[sense_codons[sense_codons %in% names(bg_mean_per_codon)]]
cat("Mean bg_RSCU across all expressed mRNAs — range:",
    round(range(bg_mean_per_codon, na.rm = TRUE), 3), "\n")
cat("(Should be ≈ 1 for all codons by construction)\n")
```

```{r std_vs_bg_rscu_corr, fig.width=5.5, fig.height=5}
# Per-codon: mean standard RSCU vs mean background RSCU across all transcripts.
# They differ because standard RSCU assumes uniform usage; background RSCU corrects
# for transcriptome-wide bias. Codons the transcriptome already favours have
# standard RSCU > 1 but background RSCU ≈ 1.
std_mean  <- colMeans(rscu_mat0[, sense_codons[sense_codons %in% colnames(rscu_mat0)],
                                drop = FALSE], na.rm = TRUE)
bg_mean   <- colMeans(bg_rscu_mat0[, sense_codons[sense_codons %in% colnames(bg_rscu_mat0)],
                                   drop = FALSE], na.rm = TRUE)

corr_df <- tibble(
  codon     = names(std_mean),
  std_rscu  = std_mean,
  bg_rscu   = bg_mean[names(std_mean)]
) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  filter(!is.na(bg_rscu))

cat("Pearson r (mean standard RSCU ~ mean background RSCU, per codon):",
    round(cor(corr_df$std_rscu, corr_df$bg_rscu, use = "complete.obs"), 3), "\n")

ggplot(corr_df, aes(x = std_rscu, y = bg_rscu, colour = gc3_class)) +
  geom_hline(yintercept = 1, colour = "grey70", linewidth = 0.4, linetype = "dashed") +
  geom_vline(xintercept = 1, colour = "grey70", linewidth = 0.4, linetype = "dashed") +
  geom_point(size = 2.2, alpha = 0.85) +
  ggrepel::geom_text_repel(
    data = . %>% filter(abs(std_rscu - bg_rscu) > 0.15),
    aes(label = codon), size = 2.5, show.legend = FALSE, max.overlaps = 20
  ) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  labs(
    title    = "Standard RSCU vs background-corrected RSCU (per codon, all expressed mRNAs)",
    subtitle = "Background RSCU uses transcriptome codon frequencies as expected; standard assumes uniform\nDashed lines = 1 (uniform / on-background)",
    x = "Mean standard RSCU (all expressed mRNAs)",
    y = "Mean background RSCU (all expressed mRNAs)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
```

```{r rscu_summary_function}
# Summarize per-transcript RSCU values into group-level stats.
summarize_rscu <- function(rscu_mat, tx_ids) {
  tx_ids <- tx_ids[tx_ids %in% rownames(rscu_mat)]
  sub    <- rscu_mat[tx_ids, sense_codons, drop = FALSE]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(sub))
    map_dfr(codons, function(codon) {
      vals <- sub[, codon]
      vals <- vals[!is.na(vals)]
      tibble(
        amino_acid  = aa,
        codon       = codon,
        mean_rscu   = mean(vals),
        median_rscu = median(vals),
        se_rscu     = if (length(vals) > 1) sd(vals) / sqrt(length(vals)) else NA_real_,
        n           = length(vals)
      )
    })
  })
}

rscu_promotes <- summarize_rscu(rscu_mat0, promotes_tx)
rscu_neg      <- summarize_rscu(rscu_mat0, neg_ctrl_tx)

cat("RSCU summary — promotes: ", nrow(rscu_promotes), "codon-rows\n")
```

```{r rscu_delta}
rscu_combined <- rscu_promotes %>%
  dplyr::select(amino_acid, codon, promotes_mean = mean_rscu,
                promotes_se = se_rscu, promotes_median = median_rscu,
                promotes_n = n) %>%
  left_join(
    rscu_neg %>% dplyr::select(amino_acid, codon, neg_mean = mean_rscu,
                               neg_se = se_rscu, neg_median = median_rscu,
                               neg_n = n),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta_rscu = promotes_mean - neg_mean)

# Wilcoxon test per codon — multi-synonymous AAs only so Met/Trp (RSCU always 1)
# don't inflate the BH denominator.
wilcox_df <- imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
  codons <- intersect(codons, colnames(rscu_mat0))
  map_dfr(codons, function(codon) {
    p_vals <- rscu_mat0[promotes_tx[promotes_tx %in% rownames(rscu_mat0)], codon]
    n_vals <- rscu_mat0[neg_ctrl_tx[neg_ctrl_tx %in% rownames(rscu_mat0)], codon]
    p_vals <- p_vals[!is.na(p_vals)]
    n_vals <- n_vals[!is.na(n_vals)]
    if (length(p_vals) < 3 || length(n_vals) < 3) {
      return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_))
    }
    wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val = wt$p.value)
  })
}) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))

rscu_combined <- rscu_combined %>%
  left_join(wilcox_df %>% dplyr::select(amino_acid, codon, p_val, p_adj),
            by = c("amino_acid", "codon"))

cat("Top 10 codons by |Δ RSCU|:\n")
print(rscu_combined %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_rscu))) %>%
        dplyr::select(codon, amino_acid, promotes_mean, neg_mean, delta_rscu, p_adj) %>%
        head(10))
```

---

# Section 4: Visualization

```{r plot_a_sidebyside, fig.width=11, fig.height=9}
plot_df_a <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(cols = c(promotes_mean, neg_mean),
               names_to = "group", values_to = "mean_rscu") %>%
  mutate(
    se      = if_else(group == "promotes_mean", promotes_se, neg_se),
    group   = recode(group,
                     promotes_mean = "eIF3d-promoted",
                     neg_mean      = "Neg ctrl"),
    group   = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, mean_rscu)
  )

p_a <- ggplot(plot_df_a, aes(x = codon, y = mean_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = mean_rscu - se, ymax = mean_rscu + se),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "RSCU by group: eIF3d-promoted vs negative control (Hypoxia 1hr)",
    subtitle = "Mean per-transcript RSCU ± SE; dashed line = RSCU 1 (uniform usage)",
    x = "Codon", y = "Mean RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1, size = 8),
    strip.text     = element_text(face = "bold"),
    legend.position = "top"
  )
p_a
ggsave(here("plots", "rscu_sidebyside_hypoxia_1hr.pdf"), p_a, width = 11, height = 9)
```

```{r plot_b_delta, fig.width=10, fig.height=8}
plot_df_b <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta_rscu >= 0, "promotes", "neg_ctrl"),
    sig_label = case_when(
      !is.na(p_adj) & p_adj < 0.05 ~ "*",
      TRUE ~ ""
    ),
    codon = reorder(codon, delta_rscu)
  )

p_b <- ggplot(plot_df_b, aes(x = codon, y = delta_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label,
                y = delta_rscu + sign(delta_rscu) * 0.03),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ RSCU: eIF3d-promoted minus negative control (Hypoxia 1hr)",
    subtitle = "Mean per-transcript RSCU difference; * = BH-adjusted Wilcoxon q < 0.05",
    x = "Codon", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_b
ggsave(here("plots", "rscu_delta_hypoxia_1hr.pdf"), p_b, width = 10, height = 8)
```

```{r frame_ctrl_rscu, fig.width=12, fig.height=9}
rscu_mat1 <- compute_rscu_matrix(counts_mat1)
rscu_mat2 <- compute_rscu_matrix(counts_mat2)

summarize_delta_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      tibble(amino_acid = aa, codon = codon,
             delta = mean(p_vals) - mean(n_vals),
             frame = frame_label)
    })
  })
}

per_tx_sig_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    if (length(codons) < 2) return(NULL)
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      if (length(p_vals) < 3 || length(n_vals) < 3) {
        return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_, frame = frame_label))
      }
      wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
      tibble(amino_acid = aa, codon = codon, p_val = wt$p.value, frame = frame_label)
    })
  })
}

# Frame deltas
f0_delta <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(amino_acid, codon, delta = delta_rscu) %>%
  mutate(frame = "Frame 0 (coding)")
f1_delta <- summarize_delta_rscu(rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)")
f2_delta <- summarize_delta_rscu(rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")

frame_delta_df <- bind_rows(f0_delta, f1_delta, f2_delta) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

# Codon order from frame 0 delta
codon_order <- f0_delta %>%
  filter(amino_acid %in% multi_aa) %>%
  arrange(amino_acid, delta) %>%
  pull(codon)
frame_delta_df <- frame_delta_df %>%
  mutate(codon = factor(codon, levels = codon_order))

# Significance tests
sig_df <- bind_rows(
  per_tx_sig_rscu(rscu_mat0, promotes_tx, neg_ctrl_tx, "Frame 0 (coding)"),
  per_tx_sig_rscu(rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)"),
  per_tx_sig_rscu(rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    q_val = p.adjust(p_val, method = "BH"),
    sig   = !is.na(q_val) & q_val < 0.05,
    frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

f0_sig   <- sig_df %>% filter(frame == "Frame 0 (coding)") %>%
  dplyr::select(amino_acid, codon, f0_sig = sig)
ctrl_sig <- sig_df %>% filter(frame != "Frame 0 (coding)") %>%
  group_by(amino_acid, codon) %>%
  summarise(any_ctrl_sig = any(sig), .groups = "drop")
f0_labels <- f0_sig %>%
  left_join(ctrl_sig, by = c("amino_acid", "codon")) %>%
  mutate(sig_label = case_when(
    f0_sig & !any_ctrl_sig ~ "\u2726",
    f0_sig &  any_ctrl_sig ~ "\u00b7",
    TRUE                   ~ ""
  )) %>%
  dplyr::select(amino_acid, codon, sig_label) %>%
  mutate(frame = factor("Frame 0 (coding)",
                        levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

aa_any_f0 <- f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
aa_clean  <- f0_labels %>% filter(sig_label == "\u2726") %>% pull(amino_acid) %>% unique()

frame_delta_df <- frame_delta_df %>%
  mutate(aa_label = ifelse(amino_acid %in% aa_any_f0,
                           paste0(aa_label, " *"), aa_label))

aa_label_map <- frame_delta_df %>% distinct(amino_acid, aa_label)
facet_clean  <- aa_label_map %>% filter(amino_acid %in% aa_clean)

rect_layer <- if (nrow(facet_clean) > 0) {
  geom_rect(data = facet_clean,
            aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5)
} else {
  NULL
}

p_c <- ggplot(frame_delta_df, aes(x = codon, y = delta, fill = frame)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  rect_layer +
  scale_fill_manual(
    values = c("Frame 0 (coding)"  = "#E69F00",
               "Frame 1 (control)" = "#999999",
               "Frame 2 (control)" = "#CCCCCC")
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ RSCU by reading frame: frame 0 vs frameshift controls (Hypoxia 1hr)",
    subtitle = "Frames 1 & 2 share ~2/3 of nucleotides with frame 0 (shifted tiling of the same sequence).\nThey partially control for mononucleotide composition but NOT dinucleotide/trinucleotide bias.\nA signal that collapses in shifted frames is reassuring; persistence in shifted frames does not rule out codon preference.\nBlue border: frame 0 sig, neither control sig. *: frame 0 significant. [q < 0.05, BH Wilcoxon]",
    x = "Codon", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    strip.text        = element_text(face = "bold"),
    legend.position   = "top"
  )
p_c
ggsave(here("plots", "rscu_frame_ctrl_hypoxia_1hr.pdf"), p_c, width = 12, height = 9)
```

---

# Section 5: Top Codons Summary & GC3 Bias Check

```{r gc3_sanity_check}
# Compute per-transcript GC3 from counts_mat0 and compare to feature_matrix$gc3
# gc3 = sum of GC-ending codon counts / total sense codon count per transcript
gc3_codons <- codon_gc3 %>% filter(gc3_class == "GC-ending") %>% pull(codon)
gc3_codons <- intersect(gc3_codons, colnames(counts_mat0))

per_tx_sense_total <- rowSums(counts_mat0[, sense_codons[sense_codons %in% colnames(counts_mat0)]])
per_tx_gc3_count   <- rowSums(counts_mat0[, gc3_codons, drop = FALSE])
per_tx_gc3         <- per_tx_gc3_count / per_tx_sense_total

gc3_check <- tibble(
  transcript_id_clean = names(per_tx_gc3),
  gc3_computed        = per_tx_gc3
) %>%
  left_join(
    feature_matrix %>% dplyr::select(transcript_id_clean, gc3_feature = gc3),
    by = "transcript_id_clean"
  ) %>%
  filter(!is.na(gc3_feature))

cat("GC3 sanity check — correlation with feature matrix gc3:\n")
cat("  Pearson r =", round(cor(gc3_check$gc3_computed, gc3_check$gc3_feature, use = "complete.obs"), 4), "\n")
cat("  n =", nrow(gc3_check), "\n")
```

```{r gc3_scatter, fig.width=4.5, fig.height=4.5}
ggplot(gc3_check, aes(x = gc3_feature, y = gc3_computed)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.6) +
  labs(title = "GC3 sanity check",
       subtitle = "Computed from counts_mat0 vs feature_matrix$gc3",
       x = "GC3 from feature matrix", y = "GC3 computed here") +
  theme_classic(base_size = 11)
```

```{r gc3_by_group}
# Mean GC3 per group
gc3_promotes <- mean(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]], na.rm = TRUE)
gc3_neg      <- mean(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]], na.rm = TRUE)
se_gc3_prom  <- sd(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]], na.rm = TRUE) /
                sqrt(sum(!is.na(per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]])))
se_gc3_neg   <- sd(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]], na.rm = TRUE) /
                sqrt(sum(!is.na(per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]])))

cat("Mean GC3 — promotes:", round(gc3_promotes, 4), "± SE", round(se_gc3_prom, 4), "\n")
cat("Mean GC3 — neg ctrl:", round(gc3_neg, 4),      "± SE", round(se_gc3_neg, 4), "\n")
cat("Δ GC3 (promotes - neg ctrl):", round(gc3_promotes - gc3_neg, 4), "\n")

gc3_p_vals <- per_tx_gc3[promotes_tx[promotes_tx %in% names(per_tx_gc3)]]
gc3_n_vals <- per_tx_gc3[neg_ctrl_tx[neg_ctrl_tx  %in% names(per_tx_gc3)]]
gc3_wt     <- wilcox.test(gc3_p_vals, gc3_n_vals, exact = FALSE)
cat("Wilcoxon test (GC3: promotes vs neg ctrl): W =", gc3_wt$statistic,
    ", p =", signif(gc3_wt$p.value, 3), "\n")
cat("Interpretation: if p < 0.05, GC3 differs between groups and RSCU signals\n",
    "should be interpreted cautiously — AT/GC-ending codon enrichment may partly\n",
    "reflect composition rather than codon selection.\n")
```

```{r top_codons_summary}
# Top codons by |Δ RSCU| with GC3 annotation
top_codons <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_nucleotide, gc3_class),
            by = "codon") %>%
  arrange(desc(abs(delta_rscu))) %>%
  dplyr::select(codon, amino_acid, promotes_mean, neg_mean,
                delta_rscu, p_adj, gc3_nucleotide, gc3_class) %>%
  mutate(across(c(promotes_mean, neg_mean, delta_rscu), ~ round(.x, 3)),
         p_adj = signif(p_adj, 3)) %>%
  head(20)

cat("Top 20 codons by |Δ mean RSCU| (multi-synonymous AAs only):\n")
print(top_codons)

# Directional summary for top 10
top10_promotes_enriched <- rscu_combined %>%
  filter(amino_acid %in% multi_aa, delta_rscu > 0) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  arrange(desc(delta_rscu)) %>%
  head(10)

cat("\nOf the top 10 promotes-enriched codons:\n")
print(table(top10_promotes_enriched$gc3_class))
cat("(If predominantly GC-ending → GC bias likely driving signal;\n",
    "mixed/AT-enriched → evidence of translational selection)\n")
```

```{r rscu_vs_gc3, fig.width=5.5, fig.height=5}
# Δ RSCU vs third-position nucleotide — tests whether GC3 bias explains the signal.
# Each point = one codon from a multi-synonymous amino acid.
# Points are labelled for |Δ RSCU| > 0.15 or BH q < 0.05.

plot_df_gc3 <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_nucleotide, gc3_class),
            by = "codon") %>%
  mutate(
    gc3_nucleotide = factor(gc3_nucleotide, levels = c("A", "T", "G", "C")),
    sig            = !is.na(p_adj) & p_adj < 0.05,
    label_codon    = if_else(abs(delta_rscu) > 0.15 | sig, codon, NA_character_)
  )

# Wilcoxon test: GC-ending vs AT-ending Δ RSCU
gc_vals <- plot_df_gc3 %>% filter(gc3_class == "GC-ending") %>% pull(delta_rscu)
at_vals <- plot_df_gc3 %>% filter(gc3_class == "AT-ending") %>% pull(delta_rscu)
wt_gc3  <- wilcox.test(gc_vals, at_vals, exact = FALSE)
gc3_p_label <- paste0("Wilcoxon p = ", signif(wt_gc3$p.value, 2),
                       "\n(GC-ending vs AT-ending)")

# x-axis jitter positions and colours
nt_colours <- c(A = "#56B4E9", T = "#009E73", G = "#E69F00", C = "#CC79A7")

p_gc3 <- ggplot(plot_df_gc3,
                aes(x = gc3_nucleotide, y = delta_rscu, colour = gc3_nucleotide)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_jitter(aes(shape = sig), width = 0.2, size = 2.2, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 20, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = nt_colours, guide = "none") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "ns", `TRUE` = "q < 0.05"),
                     name = "BH Wilcoxon") +
  annotate("text", x = 4.4, y = max(plot_df_gc3$delta_rscu, na.rm = TRUE),
           label = gc3_p_label, hjust = 1, vjust = 1, size = 3, colour = "grey30") +
  labs(
    title    = "\u0394 RSCU vs codon third-position nucleotide (Hypoxia 1hr)",
    subtitle = "Each point = one codon (multi-synonymous AAs only)",
    x = "Third-position nucleotide", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_gc3
ggsave(here("plots", "rscu_delta_vs_gc3_hypoxia_1hr.pdf"), p_gc3, width = 5.5, height = 5)
```

```{r rscu_vs_csc_weight, fig.width=5.5, fig.height=5}
# Load per-codon iCodon CSC weights and check whether AT-ending codons enriched
# in promotes are selectively the stabilizing (high-CSC) ones.
codon_csc_weights <- dplyr::select(iCodon:::human_optimality, codon, csc_weight = optimality)

plot_df_csc <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  left_join(codon_csc_weights %>% dplyr::select(codon, csc_weight), by = "codon") %>%
  mutate(
    sig        = !is.na(p_adj) & p_adj < 0.05,
    label_codon = if_else(abs(delta_rscu) > 0.15 | sig, codon, NA_character_)
  )

# Correlation within AT-ending and GC-ending codons separately
cor_at <- cor(
  filter(plot_df_csc, gc3_class == "AT-ending")$csc_weight,
  filter(plot_df_csc, gc3_class == "AT-ending")$delta_rscu,
  use = "complete.obs"
)
cor_gc <- cor(
  filter(plot_df_csc, gc3_class == "GC-ending")$csc_weight,
  filter(plot_df_csc, gc3_class == "GC-ending")$delta_rscu,
  use = "complete.obs"
)
cat("Pearson r (Δ RSCU ~ CSC weight) — AT-ending:", round(cor_at, 3),
    "| GC-ending:", round(cor_gc, 3), "\n")

p_csc <- ggplot(plot_df_csc,
                aes(x = csc_weight, y = delta_rscu, colour = gc3_class)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_point(aes(shape = sig), size = 2.2, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.7, alpha = 0.15) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 20, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(
    values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
    name   = "Third-position"
  ) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "ns", `TRUE` = "q < 0.05"),
                     name   = "BH Wilcoxon") +
  annotate("text", x = min(plot_df_csc$csc_weight, na.rm = TRUE),
           y = max(plot_df_csc$delta_rscu, na.rm = TRUE),
           label = paste0("r (AT) = ", round(cor_at, 2),
                          "\nr (GC) = ", round(cor_gc, 2)),
           hjust = 0, vjust = 1, size = 3, colour = "grey30") +
  labs(
    title    = "\u0394 RSCU vs per-codon iCodon CSC weight (Hypoxia 1hr)",
    subtitle = "Tests whether promotes-enriched AT-ending codons are selectively high-CSC\nPositive r (AT-ending) = AT-rich promotes targets use stabilizing codons",
    x = "iCodon CSC weight (human)", y = "\u0394 mean RSCU (promotes \u2212 neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_csc
ggsave(here("plots", "rscu_delta_vs_csc_weight_hypoxia_1hr.pdf"),
       p_csc, width = 5.5, height = 5)
```

---

# Section 6: Background-Corrected RSCU — eIF3d-Promoted vs Control

Using the transcriptome-background-corrected RSCU matrix computed above. This
asks whether promoted mRNAs use codons differently from the background *and*
differently from size-matched controls, after removing the transcriptome-wide
codon bias.

```{r bg_rscu_group_summaries}
bg_rscu_promotes <- summarize_rscu(bg_rscu_mat0, promotes_tx)
bg_rscu_neg      <- summarize_rscu(bg_rscu_mat0, neg_ctrl_tx)

cat("Background RSCU summary — promotes:", nrow(bg_rscu_promotes), "codon-rows\n")
```

```{r bg_rscu_delta}
bg_rscu_combined <- bg_rscu_promotes %>%
  dplyr::select(amino_acid, codon, promotes_mean = mean_rscu,
                promotes_se = se_rscu, promotes_n = n) %>%
  left_join(
    bg_rscu_neg %>% dplyr::select(amino_acid, codon, neg_mean = mean_rscu,
                                  neg_se = se_rscu, neg_n = n),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta_bg_rscu = promotes_mean - neg_mean)

# Wilcoxon test per codon on per-transcript background RSCU (multi_aa only)
bg_wilcox_df <- imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
  codons <- intersect(codons, colnames(bg_rscu_mat0))
  map_dfr(codons, function(codon) {
    p_vals <- bg_rscu_mat0[promotes_tx[promotes_tx %in% rownames(bg_rscu_mat0)], codon]
    n_vals <- bg_rscu_mat0[neg_ctrl_tx[neg_ctrl_tx %in% rownames(bg_rscu_mat0)], codon]
    p_vals <- p_vals[!is.na(p_vals)]
    n_vals <- n_vals[!is.na(n_vals)]
    if (length(p_vals) < 3 || length(n_vals) < 3) {
      return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_))
    }
    wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val = wt$p.value)
  })
}) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))

bg_rscu_combined <- bg_rscu_combined %>%
  left_join(bg_wilcox_df %>% dplyr::select(amino_acid, codon, p_val, p_adj),
            by = c("amino_acid", "codon"))

cat("Top 10 codons by |Δ background RSCU|:\n")
print(bg_rscu_combined %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_bg_rscu))) %>%
        dplyr::select(codon, amino_acid, promotes_mean, neg_mean, delta_bg_rscu, p_adj) %>%
        head(10))
```

```{r bg_rscu_plot_sidebyside, fig.width=11, fig.height=9}
plot_df_bg_a <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(cols = c(promotes_mean, neg_mean),
               names_to = "group", values_to = "mean_bg_rscu") %>%
  mutate(
    se      = if_else(group == "promotes_mean", promotes_se, neg_se),
    group   = recode(group,
                     promotes_mean = "eIF3d-promoted",
                     neg_mean      = "Neg ctrl"),
    group   = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, mean_bg_rscu)
  )

p_bg_a <- ggplot(plot_df_bg_a, aes(x = codon, y = mean_bg_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = mean_bg_rscu - se, ymax = mean_bg_rscu + se),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "Background-corrected RSCU: eIF3d-promoted vs neg ctrl (Hypoxia 1hr)",
    subtitle = "Mean per-transcript bg-RSCU ± SE; dashed = 1 (same rate as transcriptome background)",
    x = "Codon", y = "Mean background-corrected RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_bg_a
ggsave(here("plots", "bg_rscu_sidebyside_hypoxia_1hr.pdf"), p_bg_a, width = 11, height = 9)
```

```{r bg_rscu_plot_delta, fig.width=10, fig.height=8}
plot_df_bg_b <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta_bg_rscu >= 0, "promotes", "neg_ctrl"),
    sig_label = case_when(!is.na(p_adj) & p_adj < 0.05 ~ "*", TRUE ~ ""),
    codon     = reorder(codon, delta_bg_rscu)
  )

p_bg_b <- ggplot(plot_df_bg_b, aes(x = codon, y = delta_bg_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label, y = delta_bg_rscu + sign(delta_bg_rscu) * 0.015),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ background-corrected RSCU: eIF3d-promoted minus neg ctrl (Hypoxia 1hr)",
    subtitle = "Mean per-transcript bg-RSCU difference; * = BH-adjusted Wilcoxon q < 0.05",
    x = "Codon", y = "Δ mean bg-RSCU (promotes − neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_bg_b
ggsave(here("plots", "bg_rscu_delta_hypoxia_1hr.pdf"), p_bg_b, width = 10, height = 8)
```

```{r std_vs_bg_delta_scatter, fig.width=5.5, fig.height=5}
# Does the signal shift when we switch from standard to background RSCU?
# If a codon's Δ standard RSCU ≈ Δ background RSCU, the signal is not explained
# by the transcriptome-wide bias. If they diverge, the standard RSCU signal was
# partly driven by baseline usage differences.
delta_compare <- rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(amino_acid, codon, delta_std = delta_rscu, p_adj_std = p_adj) %>%
  left_join(
    bg_rscu_combined %>% dplyr::select(amino_acid, codon, delta_bg = delta_bg_rscu,
                                       p_adj_bg = p_adj),
    by = c("amino_acid", "codon")
  ) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(
    sig_either = (!is.na(p_adj_std) & p_adj_std < 0.05) |
                 (!is.na(p_adj_bg)  & p_adj_bg  < 0.05),
    label_codon = if_else(sig_either | abs(delta_std - delta_bg) > 0.05,
                          codon, NA_character_)
  )

r_std_bg <- cor(delta_compare$delta_std, delta_compare$delta_bg, use = "complete.obs")
cat("Pearson r (Δ standard RSCU ~ Δ background RSCU):", round(r_std_bg, 3), "\n")

p_comp <- ggplot(delta_compare, aes(x = delta_std, y = delta_bg, colour = gc3_class)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linewidth = 0.5,
              linetype = "dashed") +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_point(aes(shape = sig_either), size = 2.2, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.5,
                           max.overlaps = 20, show.legend = FALSE) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "ns (both)", `TRUE` = "sig in either"),
                     name = NULL) +
  annotate("text", x = min(delta_compare$delta_std, na.rm = TRUE),
           y = max(delta_compare$delta_bg, na.rm = TRUE),
           label = paste0("r = ", round(r_std_bg, 2)),
           hjust = 0, vjust = 1, size = 3.5, colour = "grey30") +
  labs(
    title    = "Δ standard RSCU vs Δ background-corrected RSCU (Hypoxia 1hr)",
    subtitle = "Dashed = identity line. Points off the diagonal: signal partly driven by\ntranscriptome-wide codon bias rather than group-specific preference.",
    x = "Δ standard RSCU (promotes − neg ctrl)",
    y = "Δ background RSCU (promotes − neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")
p_comp
ggsave(here("plots", "rscu_standard_vs_bg_delta_hypoxia_1hr.pdf"), p_comp,
       width = 5.5, height = 5)
```

```{r bg_rscu_violin_by_aa, fig.width=12, fig.height=8}
# Violin + boxplot for amino acids with ≥1 significant codon in background RSCU
sig_aas <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa, !is.na(p_adj), p_adj < 0.05) %>%
  pull(amino_acid) %>%
  unique()

cat("Amino acids with ≥1 significant codon (bg RSCU, BH q < 0.05):",
    length(sig_aas), "\n")
if (length(sig_aas) > 0) cat(paste(sig_aas, collapse = ", "), "\n")

if (length(sig_aas) > 0) {
  # Build per-transcript long-format data for significant amino acids
  sig_codons_by_aa <- bg_rscu_combined %>%
    filter(amino_acid %in% sig_aas, !is.na(p_adj), p_adj < 0.05) %>%
    dplyr::select(amino_acid, codon)

  violin_df <- map_dfr(seq_len(nrow(sig_codons_by_aa)), function(i) {
    aa    <- sig_codons_by_aa$amino_acid[i]
    codon <- sig_codons_by_aa$codon[i]
    tx_promotes <- promotes_tx[promotes_tx %in% rownames(bg_rscu_mat0)]
    tx_neg      <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(bg_rscu_mat0)]
    p_vals <- bg_rscu_mat0[tx_promotes, codon]
    n_vals <- bg_rscu_mat0[tx_neg,      codon]
    bind_rows(
      tibble(codon = codon, amino_acid = aa, group = "eIF3d-promoted",
             bg_rscu = p_vals[!is.na(p_vals)]),
      tibble(codon = codon, amino_acid = aa, group = "Neg ctrl",
             bg_rscu = n_vals[!is.na(n_vals)])
    )
  }) %>%
    mutate(
      group    = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
      aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
      codon_aa = paste0(codon, "\n(", amino_acid, ")")
    ) %>%
    left_join(bg_rscu_combined %>% dplyr::select(codon, p_adj_label = p_adj),
              by = "codon") %>%
    mutate(codon_aa = paste0(codon, "\nq=", signif(p_adj_label, 2)))

  p_viol <- ggplot(violin_df, aes(x = group, y = bg_rscu, fill = group)) +
    geom_violin(alpha = 0.6, linewidth = 0.4, trim = TRUE) +
    geom_boxplot(width = 0.18, outlier.size = 0.5, linewidth = 0.4,
                 fill = "white", alpha = 0.8) +
    geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
    scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                      guide = "none") +
    facet_wrap(~ codon_aa, scales = "free_y", nrow = 2) +
    labs(
      title    = "Per-transcript background RSCU for significant codons (Hypoxia 1hr)",
      subtitle = "Codons with BH q < 0.05 between eIF3d-promoted and neg ctrl.\nDashed = 1 (same usage rate as transcriptome background).",
      x = NULL, y = "Per-transcript background-corrected RSCU"
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1),
      strip.text  = element_text(face = "bold", size = 9)
    )
  p_viol
  ggsave(here("plots", "bg_rscu_violin_by_aa_hypoxia_1hr.pdf"), p_viol,
         width = max(8, length(sig_aas) * 2.5), height = 8)
} else {
  cat("No significant codons in background RSCU comparison — skipping violin plot.\n")
}
```

```{r bg_rscu_frame_ctrl, fig.width=12, fig.height=9}
bg_rscu_mat1 <- compute_bg_rscu_matrix(counts_mat1, bg_props)
bg_rscu_mat2 <- compute_bg_rscu_matrix(counts_mat2, bg_props)

bg_f0_delta <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(amino_acid, codon, delta = delta_bg_rscu) %>%
  mutate(frame = "Frame 0 (coding)")
bg_f1_delta <- summarize_delta_rscu(bg_rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)")
bg_f2_delta <- summarize_delta_rscu(bg_rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")

bg_frame_delta_df <- bind_rows(bg_f0_delta, bg_f1_delta, bg_f2_delta) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

bg_codon_order <- bg_f0_delta %>%
  filter(amino_acid %in% multi_aa) %>%
  arrange(amino_acid, delta) %>%
  pull(codon)
bg_frame_delta_df <- bg_frame_delta_df %>%
  mutate(codon = factor(codon, levels = bg_codon_order))

bg_sig_df <- bind_rows(
  per_tx_sig_rscu(bg_rscu_mat0, promotes_tx, neg_ctrl_tx, "Frame 0 (coding)"),
  per_tx_sig_rscu(bg_rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)"),
  per_tx_sig_rscu(bg_rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    q_val = p.adjust(p_val, method = "BH"),
    sig   = !is.na(q_val) & q_val < 0.05,
    frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

bg_f0_sig   <- bg_sig_df %>% filter(frame == "Frame 0 (coding)") %>%
  dplyr::select(amino_acid, codon, f0_sig = sig)
bg_ctrl_sig <- bg_sig_df %>% filter(frame != "Frame 0 (coding)") %>%
  group_by(amino_acid, codon) %>%
  summarise(any_ctrl_sig = any(sig), .groups = "drop")
bg_f0_labels <- bg_f0_sig %>%
  left_join(bg_ctrl_sig, by = c("amino_acid", "codon")) %>%
  mutate(sig_label = case_when(
    f0_sig & !any_ctrl_sig ~ "✦",
    f0_sig &  any_ctrl_sig ~ "·",
    TRUE                   ~ ""
  )) %>%
  dplyr::select(amino_acid, codon, sig_label) %>%
  mutate(frame = factor("Frame 0 (coding)",
                        levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

bg_aa_any_f0 <- bg_f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
bg_aa_clean  <- bg_f0_labels %>% filter(sig_label == "✦") %>% pull(amino_acid) %>% unique()

bg_frame_delta_df <- bg_frame_delta_df %>%
  mutate(aa_label = ifelse(amino_acid %in% bg_aa_any_f0,
                           paste0(aa_label, " *"), aa_label))

bg_aa_label_map <- bg_frame_delta_df %>% distinct(amino_acid, aa_label)
bg_facet_clean  <- bg_aa_label_map %>% filter(amino_acid %in% bg_aa_clean)

bg_rect_layer <- if (nrow(bg_facet_clean) > 0) {
  geom_rect(data = bg_facet_clean,
            aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5)
} else {
  NULL
}

p_bg_c <- ggplot(bg_frame_delta_df, aes(x = codon, y = delta, fill = frame)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  bg_rect_layer +
  scale_fill_manual(
    values = c("Frame 0 (coding)"  = "#E69F00",
               "Frame 1 (control)" = "#999999",
               "Frame 2 (control)" = "#CCCCCC")
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ background-corrected RSCU by reading frame: frame 0 vs frameshift controls (Hypoxia 1hr)",
    subtitle = "Frames 1 & 2 shifted tiling of the same sequence; background from frame 0 CDS transcriptome proportions.\nBlue border: frame 0 sig, neither control sig. ✦: clean signal. *: frame 0 significant. [q < 0.05, BH Wilcoxon]",
    x = "Codon", y = "Δ mean bg-RSCU (promotes − neg ctrl)",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_bg_c
ggsave(here("plots", "bg_rscu_frame_ctrl_hypoxia_1hr.pdf"), p_bg_c, width = 12, height = 9)
```

# Session Info

```{r session_info}
sessionInfo()
```
