---
title: "39 — SSMART Input FASTA: Full Transcript / te_lfc (si3d vs sictrl, Hypoxia 1hr)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Builds a scored FASTA file for SSMART motif discovery.

- **Score**: `te_lfc` (translational efficiency LFC, si3d vs sictrl, hypoxia 1hr)
- **Sequences**: full spliced transcript (all exons = 5'UTR + CDS + 3'UTR) for the
  most-expressed transcript per gene, derived from Salmon RNA-seq abundance
- **Genes**: all genes with TE LFC data (~7,000)

Output format (one sequence per line, consistent with SSMART expectation):
```
>ENSG00000XXXXXX|1.234
ATCGATCG...
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(tximport)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
```

---

# Load Data

## LFC Data (si3d vs sictrl, Hypoxia 1hr)

```{r load_lfc}
lfc_data_full <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("LFC data loaded:", nrow(lfc_data_full), "genes\n")
cat("te_lfc range: [", round(min(lfc_data_full$te_lfc, na.rm = TRUE), 3),
    ",", round(max(lfc_data_full$te_lfc, na.rm = TRUE), 3), "]\n")
```

## Sanity Check: Significant TE Changes (|LFC| > 0.5, padj < 0.05)

```{r sanity_check}
sig_up <- lfc_data_full %>%
  filter(te_lfc > 0.5, te_padj < 0.05) %>%
  arrange(desc(te_lfc)) %>%
  dplyr::select(symbol, te_lfc, te_padj)

sig_dn <- lfc_data_full %>%
  filter(te_lfc < -0.5, te_padj < 0.05) %>%
  arrange(te_lfc) %>%
  dplyr::select(symbol, te_lfc, te_padj)

cat("Significant TE UP   (lfc > 0.5, padj < 0.05):", nrow(sig_up), "genes\n")
cat("Significant TE DOWN (lfc < -0.5, padj < 0.05):", nrow(sig_dn), "genes\n\n")

cat("--- Top 20 UP ---\n")
print(head(sig_up, 20), n = 20)

cat("\n--- Top 20 DOWN ---\n")
print(head(sig_dn, 20), n = 20)
```

```{r lfc_prep}
lfc_data <- lfc_data_full %>%
  dplyr::select(gene_id_clean, te_lfc)
```

## TxDb, tx2gene, and Genome

```{r load_txdb}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))

tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript_id", "gene_id")) %>%
  filter(str_starts(transcript_id, "ENST"))

genome <- BSgenome.Hsapiens.UCSC.hg38

cat("Loaded TxDb and genome\n")
cat("Transcripts in tx2gene:", nrow(tx2gene), "\n")
```

---

# Select Most Abundant Transcript Per Gene

```{r most_abundant_tx}
salmon_dir  <- here("2025_MDAcells_riboseq", "salmon")
salmon_samples <- list.dirs(salmon_dir, recursive = FALSE, full.names = FALSE)
salmon_files   <- file.path(salmon_dir, salmon_samples, "quant.sf.gz")
names(salmon_files) <- salmon_samples
salmon_files <- salmon_files[file.exists(salmon_files)]
cat("Found", length(salmon_files), "Salmon quant files\n")

txi_transcripts <- tximport(salmon_files, type = "salmon",
                             tx2gene = tx2gene, txOut = TRUE)

transcript_abundance <- txi_transcripts$abundance %>%
  as.data.frame() %>%
  rownames_to_column(var = "transcript_id") %>%
  left_join(tx2gene, by = "transcript_id")

most_abundant_tx <- transcript_abundance %>%
  mutate(total_abundance = rowSums(across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::select(transcript_id, gene_id, total_abundance) %>%
  mutate(
    gene_id_clean      = sub("\\..*", "", gene_id),
    transcript_id_clean = sub("\\..*", "", transcript_id)
  ) %>%
  group_by(gene_id_clean) %>%
  slice_max(total_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

# Filter to genes with LFC data
te_gene_ids <- lfc_data$gene_id_clean
most_abundant_tx <- most_abundant_tx %>%
  filter(gene_id_clean %in% te_gene_ids)

cat("Most abundant transcript selected for", nrow(most_abundant_tx), "genes\n")
```

---

# Extract Full Transcript Sequences

```{r extract_sequences}
# All exons per transcript — concatenated by extractTranscriptSeqs gives the
# full spliced mRNA (5'UTR + CDS + 3'UTR)
exons_by_tx <- exonsBy(txdb, by = "tx", use.names = TRUE)

tx_seq_obj <- extractTranscriptSeqs(genome, exons_by_tx)
tx_seqs <- data.frame(
  transcript_id_raw = names(tx_seq_obj),
  tx_sequence       = as.character(tx_seq_obj),
  stringsAsFactors  = FALSE
) %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw)) %>%
  dplyr::select(transcript_id_clean, tx_sequence)

cat("Extracted sequences for", nrow(tx_seqs), "transcripts\n")
```

---

# Join and Build FASTA Table

```{r join}
fasta_df <- lfc_data %>%
  inner_join(most_abundant_tx, by = "gene_id_clean") %>%
  inner_join(tx_seqs, by = "transcript_id_clean") %>%
  filter(!is.na(tx_sequence), nchar(tx_sequence) > 0) %>%
  dplyr::select(gene_id_clean, te_lfc, transcript_id_clean, tx_sequence)

cat("Genes in final FASTA:", nrow(fasta_df), "\n")
cat("te_lfc range: [", round(min(fasta_df$te_lfc), 3),
    ",", round(max(fasta_df$te_lfc), 3), "]\n")
cat("Transcript length range: [", min(nchar(fasta_df$tx_sequence)),
    ",", max(nchar(fasta_df$tx_sequence)), "] nt\n")
```

---

# Write FASTA

```{r write_fasta}
out_dir <- here("output", "ssmart")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
out_path <- file.path(out_dir, "transcript_te_lfc_hypoxia_1hr.fasta")

fasta_lines <- paste0(
  ">", fasta_df$gene_id_clean, "|", round(fasta_df$te_lfc, 4), "\n",
  fasta_df$tx_sequence
)

writeLines(fasta_lines, out_path)

cat("Written:", out_path, "\n")
cat("Entries:", length(fasta_lines), "\n\n")
cat("--- Preview (first 6 entries) ---\n")
cat(head(fasta_lines, 6), sep = "\n")
```
