---
title: "42 — Prepare MCF7-SIX1 Gene Sets for RF Classification"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Prepares MCF7-SIX1 eIF3d-regulated gene sets (hypoxia + normoxia) and a
size-matched negative control pool for use in notebooks 07 and 09.

**Directionality:** MCF7-SIX1 DESeq2 contrast = log2(si3d / sictrl).
Positive LFC = eIF3d inhibits; negative LFC = eIF3d promotes.
We negate to `te_lfc` where positive = eIF3d promotes (matching 231 convention).

**Outputs** (all saved in output/genesets/ for notebook 07 compatibility):

- `mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- `mcf7six1_normoxia_3d_promotes_TE_1hr_lfc0.5.csv`
- `mcf7six1_3d_negative_controls_lfc0.5.csv`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load MCF7-SIX1 TE Data

```{r load_te_data}
eif3_dir <- "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia"
lfc_cutoff <- 0.5

hyp_raw  <- read_csv(file.path(eif3_dir, "counts/hypoxia_si_ctrl_3d.csv"),
                     show_col_types = FALSE)
norm_raw <- read_csv(file.path(eif3_dir, "counts/normoxia_sictrl_3d.csv"),
                     show_col_types = FALSE)

cat("Columns:", paste(names(hyp_raw), collapse = ", "), "\n")
cat("MCF7-SIX1 hypoxia TE file: ", nrow(hyp_raw),  "genes\n")
cat("MCF7-SIX1 normoxia TE file:", nrow(norm_raw), "genes\n")
```

```{r recode_lfc}
# Negate: positive te_lfc = eIF3d promotes translation
hyp_te <- hyp_raw %>%
  mutate(
    te_lfc        = -log2FoldChange,
    te_padj       = padj,
    gene_id_clean = str_remove(gene_id, "\\..*")
  )

norm_te <- norm_raw %>%
  mutate(
    te_lfc        = -log2FoldChange,
    te_padj       = padj,
    gene_id_clean = str_remove(gene_id, "\\..*")
  )

cat("=== AFTER RE-CODING (positive te_lfc = eIF3d PROMOTES) ===\n")
cat("Hypoxia promotes (te_lfc >", lfc_cutoff, ", padj < 0.05):",
    sum(hyp_te$te_lfc > lfc_cutoff & !is.na(hyp_te$te_padj) & hyp_te$te_padj < 0.05), "\n")
cat("Normoxia promotes (te_lfc >", lfc_cutoff, ", padj < 0.05):",
    sum(norm_te$te_lfc > lfc_cutoff & !is.na(norm_te$te_padj) & norm_te$te_padj < 0.05), "\n")
```

---

# Section 2: Promoted Gene Sets

```{r hypoxia_promotes}
hyp_promotes <- hyp_te %>%
  filter(te_lfc > lfc_cutoff, !is.na(te_padj), te_padj < 0.05) %>%
  transmute(
    db_gene_symbol = symbol,
    ensembl_gene   = gene_id_clean,
    te_lfc,
    te_padj,
    gs_name        = paste0("mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc", lfc_cutoff),
    gs_description = paste0(
      "genes whose translation is promoted by eIF3d in hypoxia at 1hr, abs(LFC)>",
      lfc_cutoff, ", in MCF7-SIX1 cells"
    ),
    gs_source_species = "homo sapiens"
  )

cat("Hypoxia eIF3d-promoted genes:", nrow(hyp_promotes), "\n")
```

```{r normoxia_promotes}
norm_promotes <- norm_te %>%
  filter(te_lfc > lfc_cutoff, !is.na(te_padj), te_padj < 0.05) %>%
  transmute(
    db_gene_symbol = symbol,
    ensembl_gene   = gene_id_clean,
    te_lfc,
    te_padj,
    gs_name        = paste0("mcf7six1_normoxia_3d_promotes_TE_1hr_lfc", lfc_cutoff),
    gs_description = paste0(
      "genes whose translation is promoted by eIF3d in normoxia at 1hr, abs(LFC)>",
      lfc_cutoff, ", in MCF7-SIX1 cells"
    ),
    gs_source_species = "homo sapiens"
  )

cat("Normoxia eIF3d-promoted genes:", nrow(norm_promotes), "\n")
```

---

# Section 3: Negative Control Pool

```{r negative_controls}
# Pool: |te_lfc| < 0.5 AND padj > 0.3 in BOTH hypoxia and normoxia
neg_hyp  <- hyp_te  %>% filter(abs(te_lfc) < lfc_cutoff, !is.na(te_padj), te_padj > 0.3) %>% pull(gene_id_clean)
neg_norm <- norm_te %>% filter(abs(te_lfc) < lfc_cutoff, !is.na(te_padj), te_padj > 0.3) %>% pull(gene_id_clean)

promoted_both <- union(hyp_promotes$ensembl_gene, norm_promotes$ensembl_gene)
neg_pool_ids  <- setdiff(intersect(neg_hyp, neg_norm), promoted_both)

cat("Genes meeting neg ctrl criteria (|LFC| < 0.5, padj > 0.3 in both conditions):", length(neg_pool_ids), "\n")
cat("Excluding", length(promoted_both), "promoted genes\n")
cat("Final neg pool:", length(neg_pool_ids), "genes\n")

# Save the full pool (size-matching happens in notebook 07 with set.seed(9))
neg_ctrl_df <- hyp_te %>%
  filter(gene_id_clean %in% neg_pool_ids) %>%
  transmute(
    db_gene_symbol = symbol,
    ensembl_gene   = gene_id_clean,
    te_lfc,
    te_padj
  )

cat("Rows in neg control CSV:", nrow(neg_ctrl_df), "\n")
```

---

# Section 4: Save Outputs

```{r save_outputs}
dir.create(here("output", "genesets"), showWarnings = FALSE, recursive = TRUE)

write_csv(hyp_promotes,
          here("output", "genesets", "mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
cat("Saved: output/genesets/mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc0.5.csv\n")

write_csv(norm_promotes,
          here("output", "genesets", "mcf7six1_normoxia_3d_promotes_TE_1hr_lfc0.5.csv"))
cat("Saved: output/genesets/mcf7six1_normoxia_3d_promotes_TE_1hr_lfc0.5.csv\n")

write_csv(neg_ctrl_df,
          here("output", "genesets", "mcf7six1_3d_negative_controls_lfc0.5.csv"))
cat("Saved: output/genesets/mcf7six1_3d_negative_controls_lfc0.5.csv\n")
```

---

# Section 5: Validation

```{r validation}
cat("=== FINAL SUMMARY ===\n")
cat("Hypoxia promoted:", nrow(hyp_promotes), "genes\n")
cat("Normoxia promoted:", nrow(norm_promotes), "genes\n")
cat("Neg ctrl pool:", nrow(neg_ctrl_df), "genes\n")
cat("Overlap (hyp promotes ∩ norm promotes):",
    length(intersect(hyp_promotes$ensembl_gene, norm_promotes$ensembl_gene)), "genes\n")

stopifnot(nrow(hyp_promotes)  > 0)
stopifnot(nrow(norm_promotes) > 0)
stopifnot(nrow(neg_ctrl_df)   >= max(nrow(hyp_promotes), nrow(norm_promotes)))
cat("All checks passed.\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
