---
title: "43 — Feature Extraction: MCF7-SIX1 External Stability Features"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Builds the MCF7-SIX1 external stability feature matrix for RF classification
(notebooks 07 and 09). Mirrors the feature scope of MDA-MB-231
`feature_matrix_external_stability.rds` after notebook 07 stripping, but uses
MCF7-SIX1-specific transcript selection and cell-line-specific background RSCU.

**Features included:**

| Group | Columns | Notes |
|-------|---------|-------|
| bg-RSCU | 62 (`bg_rscu_*` × 59 + 3 summaries) | MCF7-SIX1 background |
| Codon frequencies | 59 (`codon_freq_*`) | Length-normalized raw |
| Region lengths + log2 | 9 (`utr5_length`, `utr3_length`, `cds_length`, `total_length`, log2 transforms) | From txdb |
| GC content | 4 (`utr5_gc`, `utr3_gc`, `cds_gc`, `transcript_gc`) | |
| GC3 | 1 (`gc3`) | Fraction of sense codons with GC at 3rd position |
| CAI | 1 | Ribosomal protein reference (MCF7-SIX1 expressed) |
| tAI | 1 | GtRNAdb hg38 tRNA GCN |
| CSC | 1 | iCodon human weights (Diez 2022) |
| FOP | 1 | Top-10% TE reference from MDA-MB-231 sictrl normoxia |
| 4E / DAP5 binary | 4 (`is_4e_resistant`, `is_4e_sensitive`, `is_dap5_repressed`, `is_dap5_promoted`) | Gene symbol lookup |
| Translation signatures | ~34 (`sig_*`) | Same set as notebook 01_ |
| Kozak context | 3 (`kozak_score`, `kozak_optimal`, `kozak_pwm_score`) | MCF7-SIX1 transcript selection |
| CNOT3 external | 2 (`cnot3_ribo_enrichment`, `cnot3_weighted_codon_score`) | Joined from saved RDS |
| External stability | 6 (Zhu2024, Hia2026, Karner2026) | Joined from saved RDS |

**Features NOT included vs MDA-MB-231 `feature_matrix_external_stability.rds`:**
- k-mers / RBP motifs (notebook 07 already removes them; no computational need)
- eIF3 CLIP binding (MDA-MB-231 CLIP-seq data only)
- RNA secondary structure (cluster-computed, not reproducible locally)
- TE LFC features (excluded by `exclude_own_te_features = TRUE` in notebook 07)

**Transcript selection:** most-expressed per gene from MCF7-SIX1 RNA-seq sictrl
hypoxia Salmon (NM2023_0101/0111/0121 — rRNADepletion/KAPA_RNAHyper libraries,
NOT ribosome profiling samples NM2023_0037–0066). One matrix serves both hypoxia
and normoxia RF runs.

**Outputs:**
- `output/predictive_modeling/feature_matrix_mcf7six1_codon_features.rds` (121 codon features only)
- `output/predictive_modeling/feature_matrix_mcf7six1_external_stability.rds` (full matrix)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(AnnotationDbi)
library(iCodon)
library(readxl)
```

---

# Section 1: MCF7-SIX1 TE Data

```{r load_te_data}
eif3_dir <- "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia"

hyp_raw <- read_csv(file.path(eif3_dir, "counts/hypoxia_si_ctrl_3d.csv"),
                    show_col_types = FALSE) %>%
  mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

mcf7_expressed <- unique(hyp_raw$gene_id_clean)
cat("MCF7-SIX1 expressed genes:", length(mcf7_expressed), "\n")
```

---

# Section 2: Transcript Database & CDS Sequences

```{r load_txdb}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

cds_len_df <- tibble(
  transcript_id_clean = cds_names_clean,
  cds_length          = as.integer(width(cds_seqs_all))
)
cat("Total CDS sequences in txdb:", length(cds_seqs_all), "\n")
```

---

# Section 3: Transcript Selection via Salmon RNA-seq

RNA-seq sictrl hypoxia samples: NM2023_0101 (RepA), NM2023_0111 (RepB), NM2023_0121 (RepC).
Confirmed rRNADepletion / KAPA_RNAHyper library prep (NOT ribosome profiling).
Ribo profiling samples (NM2023_0037–0066) are excluded.

```{r salmon_tpm}
rnaseq_ids <- c("NM2023_0101", "NM2023_0111", "NM2023_0121")
salmon_dir <- file.path(eif3_dir, "salmon")

tpm_list <- map(rnaseq_ids, function(sid) {
  f <- file.path(salmon_dir, sid, "quant.sf.gz")
  read_tsv(f, show_col_types = FALSE) %>%
    mutate(
      transcript_id_clean = str_remove(str_extract(Name, "^[^|]+"), "\\..*")
    ) %>%
    dplyr::select(transcript_id_clean, !!sid := TPM)
})

tpm_wide <- purrr::reduce(tpm_list, full_join, by = "transcript_id_clean")
tpm_mat  <- as.matrix(tpm_wide[, rnaseq_ids])
tpm_avg  <- tibble(
  transcript_id_clean = as.character(tpm_wide$transcript_id_clean),
  mean_tpm            = as.numeric(rowMeans(tpm_mat, na.rm = TRUE))
)

cat("Salmon TPM data:", nrow(tpm_avg), "transcripts\n")
cat("Transcripts with mean TPM > 0:", sum(tpm_avg$mean_tpm > 0, na.rm = TRUE), "\n")
```

```{r tx2gene_map}
tx2gene_all <- AnnotationDbi::select(
  txdb,
  keys    = keys(txdb, keytype = "TXNAME"),
  columns = c("TXNAME", "GENEID"),
  keytype = "TXNAME"
) %>%
  as_tibble() %>%
  mutate(
    TXNAME              = as.character(TXNAME),
    GENEID              = as.character(GENEID),
    transcript_id_clean = str_remove(TXNAME,  "\\..*"),
    gene_id_clean       = str_remove(GENEID,  "\\..*")
  ) %>%
  filter(transcript_id_clean %in% cds_names_clean)

cat("Transcripts with CDS in txdb:", nrow(tx2gene_all), "\n")
```

```{r select_primary_tx}
tx_candidates <- tx2gene_all %>%
  dplyr::select(transcript_id_clean, gene_id_clean) %>%
  filter(gene_id_clean %in% mcf7_expressed) %>%
  left_join(tpm_avg,    by = "transcript_id_clean") %>%
  left_join(cds_len_df, by = "transcript_id_clean") %>%
  mutate(
    transcript_id_clean = as.character(transcript_id_clean),
    gene_id_clean       = as.character(gene_id_clean),
    mean_tpm            = as.numeric(replace_na(mean_tpm, 0)),
    cds_length          = as.integer(cds_length)
  )

# Highest mean TPM per gene; longest CDS as tie-break
tx_map <- tx_candidates %>%
  arrange(gene_id_clean, desc(mean_tpm), desc(cds_length)) %>%
  filter(!duplicated(gene_id_clean)) %>%
  dplyr::select(transcript_id_clean, gene_id_clean, mean_tpm, cds_length)

cat("Transcripts selected:", nrow(tx_map), "\n")
cat("  Salmon-covered (mean TPM > 0):", sum(tx_map$mean_tpm > 0), "\n")
cat("  Fallback to longest CDS (mean TPM = 0):", sum(tx_map$mean_tpm == 0), "\n")
```

---

# Section 4: CDS Sequences & Codon Counts

```{r filter_cds}
keep_tx  <- tx_map$transcript_id_clean
keep_idx <- cds_names_clean %in% keep_tx

cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) > 0 & width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_counts}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)
cat("Codon counts matrix:", nrow(counts_mat0), "x", ncol(counts_mat0), "\n")
```

---

# Section 5: Genetic Code Setup

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
stop_codons  <- names(gen_code[gen_code == "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]
multi_syn_codons <- intersect(unlist(aa_to_codons[multi_aa]), colnames(counts_mat0))

codon_gc3 <- tibble(
  codon     = sense_codons,
  gc3_class = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending")
)

cat("Multi-synonymous codons:", length(multi_syn_codons), "(expected 59)\n")
```

---

# Section 6: Background-Corrected RSCU

MCF7-SIX1-specific background: proportions computed from all MCF7-SIX1 expressed
transcripts (not imported from MDA-MB-231 to preserve cell-line context).

```{r bg_props}
bg_props <- imap(aa_to_codons, function(codons, aa) {
  codons  <- intersect(codons, colnames(counts_mat0))
  if (length(codons) == 0) return(NULL)
  sub_cnt <- counts_mat0[, codons, drop = FALSE]
  aa_tot  <- rowSums(sub_cnt)
  valid   <- aa_tot > 0
  if (!any(valid)) return(NULL)
  prop_mat <- sub_cnt[valid, , drop = FALSE] / aa_tot[valid]
  colMeans(prop_mat)
})
bg_props <- Filter(Negate(is.null), bg_props)

prop_sums <- sapply(bg_props, sum)
stopifnot(all(abs(prop_sums - 1) < 1e-6))
cat("bg_props: computed for", length(bg_props), "amino acids; sum-to-1 check: passed\n")
```

```{r compute_bg_rscu}
compute_bg_rscu_matrix <- function(counts_mat, bg_props) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(bg_props)) {
    props  <- bg_props[[aa]]
    codons <- intersect(names(props), colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt <- counts_mat[, codons, drop = FALSE]
    aa_tot  <- rowSums(sub_cnt)
    valid   <- aa_tot > 0
    for (codon in codons) {
      bp <- props[codon]
      if (bp > 0 && any(valid)) {
        rscu_mat[valid, codon] <- (sub_cnt[valid, codon] / aa_tot[valid]) / bp
      }
    }
  }
  rscu_mat
}

bg_rscu_mat0 <- compute_bg_rscu_matrix(counts_mat0, bg_props)
cat("bg-RSCU matrix:", nrow(bg_rscu_mat0), "x", ncol(bg_rscu_mat0), "\n")

stopifnot(min(bg_rscu_mat0, na.rm = TRUE) >= 0)
bg_codon_means <- colMeans(bg_rscu_mat0[, multi_syn_codons], na.rm = TRUE)
cat("bg-RSCU codon mean range:", round(range(bg_codon_means), 4), "(expected ≈ 1)\n")
stopifnot(all(abs(bg_codon_means - 1) < 0.01))
cat("bg-RSCU sanity checks: passed\n")
```

```{r bg_rscu_features}
bg_rscu_df <- as_tibble(bg_rscu_mat0[, multi_syn_codons], rownames = "transcript_id_clean") %>%
  rename_with(~ paste0("bg_rscu_", .), -transcript_id_clean)

stopifnot(!"bg_rscu_ATG" %in% colnames(bg_rscu_df))
stopifnot(!"bg_rscu_TGG" %in% colnames(bg_rscu_df))
cat("bg_rscu_* columns:", ncol(bg_rscu_df) - 1, "(expected 59)\n")
```

```{r bg_rscu_summary}
at_codons <- codon_gc3 %>% filter(gc3_class == "AT-ending") %>%
  pull(codon) %>% intersect(multi_syn_codons)
gc_codons <- codon_gc3 %>% filter(gc3_class == "GC-ending") %>%
  pull(codon) %>% intersect(multi_syn_codons)

bg_rscu_at_cols  <- paste0("bg_rscu_", at_codons)
bg_rscu_gc_cols  <- paste0("bg_rscu_", gc_codons)
bg_rscu_all_cols <- paste0("bg_rscu_", multi_syn_codons)

bg_rscu_summary <- bg_rscu_df %>%
  mutate(
    bg_rscu_at_mean   = rowMeans(across(all_of(bg_rscu_at_cols)),  na.rm = TRUE),
    bg_rscu_gc_mean   = rowMeans(across(all_of(bg_rscu_gc_cols)),  na.rm = TRUE),
    bg_rscu_deviation = rowMeans(
      across(all_of(bg_rscu_all_cols), ~ (. - 1)^2), na.rm = TRUE
    )
  ) %>%
  dplyr::select(transcript_id_clean, bg_rscu_at_mean, bg_rscu_gc_mean, bg_rscu_deviation)

stopifnot(all(bg_rscu_summary$bg_rscu_deviation >= 0, na.rm = TRUE))
stopifnot(all(bg_rscu_summary$bg_rscu_at_mean  >= 0, na.rm = TRUE))
stopifnot(all(bg_rscu_summary$bg_rscu_gc_mean  >= 0, na.rm = TRUE))
cat("bg_rscu summary: computed; non-negative checks: passed\n")
```

---

# Section 7: Per-Codon Raw Frequencies

```{r codon_freq}
sense_in_mat <- sense_codons[sense_codons %in% colnames(counts_mat0)]
total_sense  <- rowSums(counts_mat0[, sense_in_mat, drop = FALSE])
freq_mat     <- counts_mat0[, multi_syn_codons, drop = FALSE] / total_sense

freq_df <- as_tibble(freq_mat, rownames = "transcript_id_clean") %>%
  rename_with(~ paste0("codon_freq_", .), -transcript_id_clean)

stopifnot(!"codon_freq_ATG" %in% colnames(freq_df))
stopifnot(!"codon_freq_TGG" %in% colnames(freq_df))
stopifnot(all(rowSums(freq_mat, na.rm = TRUE) <= 1 + 1e-6))
stopifnot(min(freq_mat, na.rm = TRUE) >= 0)
cat("codon_freq_* columns:", ncol(freq_df) - 1, "(expected 59)\n")
cat("codon_freq range check: passed\n")
```

---

# Section 8: Build Codon Feature Matrix

```{r build_feature_matrix}
gene_anno <- hyp_raw %>%
  dplyr::select(gene_id, symbol, gene_id_clean) %>%
  distinct(gene_id_clean, .keep_all = TRUE)

feature_matrix <- tx_map %>%
  dplyr::select(transcript_id_clean, gene_id_clean) %>%
  inner_join(gene_anno, by = "gene_id_clean") %>%
  left_join(bg_rscu_df,      by = "transcript_id_clean") %>%
  left_join(bg_rscu_summary, by = "transcript_id_clean") %>%
  left_join(freq_df,         by = "transcript_id_clean")

cat("Codon feature matrix dimensions:", nrow(feature_matrix), "genes x", ncol(feature_matrix), "columns\n")

new_feat_cols <- colnames(feature_matrix)[
  !colnames(feature_matrix) %in% c("transcript_id_clean", "gene_id_clean", "gene_id", "symbol")
]
cat("Feature columns added:", length(new_feat_cols), "(expected 121)\n")
cat("  bg_rscu_* :", sum(startsWith(new_feat_cols, "bg_rscu_")), "(expected 62: 59 + 3 summary)\n")
cat("  codon_freq_*:", sum(startsWith(new_feat_cols, "codon_freq_")), "(expected 59)\n")

stopifnot(sum(startsWith(new_feat_cols, "bg_rscu_")) == 62)
stopifnot(sum(startsWith(new_feat_cols, "codon_freq_")) == 59)
cat("Dimension checks: passed\n")
```

---

# Section 9: Save Codon-Only Matrix

```{r save_codon_only}
dir.create(here("output", "predictive_modeling"), showWarnings = FALSE, recursive = TRUE)

saveRDS(feature_matrix,
        here("output", "predictive_modeling", "feature_matrix_mcf7six1_codon_features.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_mcf7six1_codon_features.rds\n")
cat("Rows:", nrow(feature_matrix), "| Columns:", ncol(feature_matrix), "\n")
```

---

# Section 10: Region Lengths and GC Content

Extracts 5'UTR, CDS, and 3'UTR lengths and GC content for MCF7-SIX1 selected
transcripts. Pattern mirrors `01_feature_extraction.Rmd`.

```{r extract_utrs}
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cat("Total 5'UTR transcripts in txdb:", length(utr5), "\n")
cat("Total 3'UTR transcripts in txdb:", length(utr3), "\n")
```

```{r utr_lengths}
utr5_widths <- as.data.frame(utr5) %>%
  mutate(transcript_id_clean = sub("\\..*", "", group_name)) %>%
  group_by(transcript_id_clean) %>%
  summarise(utr5_length = as.integer(sum(width)), .groups = "drop")

utr3_widths <- as.data.frame(utr3) %>%
  mutate(transcript_id_clean = sub("\\..*", "", group_name)) %>%
  group_by(transcript_id_clean) %>%
  summarise(utr3_length = as.integer(sum(width)), .groups = "drop")

cat("5'UTR widths computed:", nrow(utr5_widths), "transcripts\n")
cat("3'UTR widths computed:", nrow(utr3_widths), "transcripts\n")
```

```{r gc_content_function}
gc_content <- function(seq) {
  seq        <- str_to_upper(seq)          # match 01_ behaviour
  gc_count   <- str_count(seq, "[GC]")
  total_len  <- nchar(seq)
  result     <- (gc_count / total_len) * 100
  result[total_len == 0] <- NA_real_
  result
}
```

```{r utr_gc}
utr5_seq_all <- extractTranscriptSeqs(genome, utr5) %>% as.data.frame() %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw),
         utr5_gc = gc_content(x)) %>%
  dplyr::rename(utr5_sequence = x) %>%
  dplyr::select(transcript_id_clean, utr5_sequence, utr5_gc)

utr3_seq_all <- extractTranscriptSeqs(genome, utr3) %>% as.data.frame() %>%
  rownames_to_column("transcript_id_raw") %>%
  mutate(transcript_id_clean = sub("\\..*", "", transcript_id_raw),
         utr3_gc = gc_content(x)) %>%
  dplyr::select(transcript_id_clean, utr3_gc)

cds_gc_df <- tibble(
  transcript_id_clean = names(cds_frame0),
  cds_gc              = gc_content(as.character(cds_frame0))
)

cat("5'UTR GC computed:", nrow(utr5_seq_all), "transcripts\n")
cat("3'UTR GC computed:", nrow(utr3_seq_all), "transcripts\n")
```

```{r join_lengths_gc}
feature_matrix <- feature_matrix %>%
  left_join(tx_map %>% dplyr::select(transcript_id_clean, cds_length),
            by = "transcript_id_clean") %>%
  left_join(utr5_widths,  by = "transcript_id_clean") %>%
  left_join(utr3_widths,  by = "transcript_id_clean") %>%
  left_join(utr5_seq_all %>% dplyr::select(transcript_id_clean, utr5_gc),
            by = "transcript_id_clean") %>%
  left_join(utr3_seq_all, by = "transcript_id_clean") %>%
  left_join(cds_gc_df,    by = "transcript_id_clean") %>%
  mutate(
    utr5_length_safe  = coalesce(as.integer(utr5_length), 0L),
    utr3_length_safe  = coalesce(as.integer(utr3_length), 0L),
    total_length      = utr5_length_safe + cds_length + utr3_length_safe,
    log2_utr5_length  = log2(utr5_length_safe + 1),
    log2_utr3_length  = log2(utr3_length_safe + 1),
    log2_cds_length   = log2(cds_length + 1),
    log2_total_length = log2(total_length + 1),
    transcript_gc     = (coalesce(utr5_gc, 0) * utr5_length_safe +
                         cds_gc             * cds_length +
                         coalesce(utr3_gc, 0) * utr3_length_safe) /
                        pmax(total_length, 1L)
  ) %>%
  dplyr::select(-utr5_length_safe, -utr3_length_safe)

cat("Region length + GC columns added. Matrix now:", nrow(feature_matrix), "x", ncol(feature_matrix), "\n")
cat("Genes with 5'UTR length:", sum(!is.na(feature_matrix$utr5_length)), "\n")
cat("Genes with 3'UTR length:", sum(!is.na(feature_matrix$utr3_length)), "\n")
```

```{r validate_lengths_gc}
# Coding check: GC values must be in biological range
stopifnot(all(feature_matrix$cds_gc  >= 20 & feature_matrix$cds_gc  <= 85, na.rm = TRUE))
stopifnot(all(feature_matrix$utr5_gc >= 10 & feature_matrix$utr5_gc <= 90, na.rm = TRUE))
cat("GC content range checks: passed\n")
cat("cds_gc   range:", round(range(feature_matrix$cds_gc,  na.rm = TRUE), 1), "%\n")
cat("utr5_gc  range:", round(range(feature_matrix$utr5_gc, na.rm = TRUE), 1), "%\n")
cat("utr3_gc  range:", round(range(feature_matrix$utr3_gc, na.rm = TRUE), 1), "%\n")
cat("Median utr5_length:", median(feature_matrix$utr5_length, na.rm = TRUE), "nt (expected ~100–300 nt)\n")
cat("Median utr3_length:", median(feature_matrix$utr3_length, na.rm = TRUE), "nt (expected ~500–1500 nt)\n")
cat("Median cds_length: ", median(feature_matrix$cds_length, na.rm = TRUE), "nt (expected ~800–1500 nt)\n")

# Biological spot-check: ACTB is well-annotated with known UTR lengths
actb <- feature_matrix %>% filter(symbol == "ACTB")
if (nrow(actb) > 0) {
  cat("\nACTB spot-check:\n")
  cat("  utr5_length:", actb$utr5_length, "nt (expected ~200–500 nt)\n")
  cat("  utr3_length:", actb$utr3_length, "nt (expected ~2000–4000 nt)\n")
  cat("  cds_gc:     ", round(actb$cds_gc, 1), "% (expected ~50–55%)\n")
  stopifnot(actb$cds_gc > 40 && actb$cds_gc < 70)
} else {
  cat("ACTB not in MCF7-SIX1 expressed gene set\n")
}
```

---

# Section 11: Codon Optimality (GC3, CAI, tAI, CSC, FOP)

Pattern mirrors `01g_feature_extraction_codon_optimality.Rmd`. Reuses `counts_mat0`,
`cds_frame0`, `gen_code`, `sense_codons`, and `aa_to_codons` from Sections 4–5.

## GC3

```{r gc3}
gc3_codons   <- sense_codons[substr(sense_codons, 3, 3) %in% c("G", "C")]
gc3_codons   <- intersect(gc3_codons, colnames(counts_mat0))
sense_in_cnt <- intersect(sense_codons, colnames(counts_mat0))

gc3_num <- rowSums(counts_mat0[, gc3_codons,   drop = FALSE])
gc3_den <- rowSums(counts_mat0[, sense_in_cnt, drop = FALSE])
gc3_scores <- (gc3_num / gc3_den) * 100   # percent (0–100) to match 01_ calculate_gc3()
gc3_scores[gc3_den == 0] <- NA_real_

cat("GC3 range:", round(range(gc3_scores, na.rm = TRUE), 2), "% (expected ~30–70)\n")
stopifnot(all(gc3_scores >= 0 & gc3_scores <= 100, na.rm = TRUE))
cat("GC3 range check passed\n")
```

## CAI — Codon Adaptation Index

Ribosomal protein transcripts from MCF7-SIX1 hypoxia data as reference.

```{r cai}
ribo_tx <- hyp_raw %>%
  filter(str_detect(symbol, "^RPL|^RPS")) %>%
  mutate(transcript_id_clean = tx_map$transcript_id_clean[match(gene_id_clean, tx_map$gene_id_clean)]) %>%
  filter(!is.na(transcript_id_clean)) %>%
  pull(transcript_id_clean)

ribo_subset <- rownames(counts_mat0) %in% ribo_tx
cat("Ribosomal protein transcripts in codon matrix:", sum(ribo_subset), "\n")

ref_counts_cai <- colSums(counts_mat0[ribo_subset, , drop = FALSE])

w_cai <- setNames(rep(NA_real_, length(gen_code)), names(gen_code))
w_cai[stop_codons] <- 0

for (aa in names(aa_to_codons)) {
  codons <- aa_to_codons[[aa]]
  counts <- ref_counts_cai[codons]
  total  <- sum(counts)
  if (total == 0) {
    w_cai[codons] <- 1
    next
  }
  n_syn <- length(codons)
  rscu  <- counts / (total / n_syn)
  max_r <- max(rscu)
  if (max_r == 0) {
    w_cai[codons] <- 0
  } else {
    w_cai[codons] <- pmin(rscu / max_r, 1)
  }
}

compute_cai_gene <- function(gene_counts, w_vals) {
  use     <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use   <- gene_counts[use]
  w_use   <- w_vals[use]
  n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_use)) / n_total)
}

cai_scores0 <- apply(counts_mat0, 1, compute_cai_gene, w_vals = w_cai)
names(cai_scores0) <- rownames(counts_mat0)

cat("CAI summary:\n")
print(summary(cai_scores0))
stopifnot("CAI must be in [0,1]" = all(cai_scores0 <= 1 + 1e-9, na.rm = TRUE))
cat("CAI range check passed\n")
```

## tAI — tRNA Adaptation Index

```{r tai}
trna_gcn <- read_csv(here("accessories", "human", "human_trna_gcn.csv"),
                     comment = "#", show_col_types = FALSE)
cat("tRNA GCN table:", nrow(trna_gcn), "rows\n")

decode_anticodon <- function(ac) {
  x1 <- substr(ac, 1, 1); x2 <- substr(ac, 2, 2); x3 <- substr(ac, 3, 3)
  wc <- c(A = "T", T = "A", G = "C", C = "G")
  y1 <- unname(wc[x3]); y2 <- unname(wc[x2])
  if (x1 == "A") {
    y3 <- c("T", "C", "A"); svs <- c(0, 0, 0)
  } else if (x1 == "G") {
    y3 <- c("C", "T"); svs <- c(0, 0.41)
  } else if (x1 == "C") {
    y3 <- "G"; svs <- 0
  } else {
    y3 <- "A"; svs <- 0
  }
  data.frame(codon = paste0(y1, y2, y3), s = svs, stringsAsFactors = FALSE)
}

aa_of_codon <- setNames(as.character(gen_code), names(gen_code))
ws_tai <- setNames(rep(0, length(gen_code)), names(gen_code))

for (i in seq_len(nrow(trna_gcn))) {
  ac  <- trna_gcn$anticodon[i]
  aa  <- trna_gcn$amino_acid[i]
  gcn <- trna_gcn$gcn[i]
  decoded <- decode_anticodon(ac)
  for (j in seq_len(nrow(decoded))) {
    codon <- decoded$codon[j]; s_val <- decoded$s[j]
    if (!is.na(aa_of_codon[codon]) && aa_of_codon[codon] == aa)
      ws_tai[codon] <- ws_tai[codon] + (1 - s_val) * gcn
  }
}

w_tai <- ws_tai / max(ws_tai[sense_codons])
cat("w_tai range:", round(range(w_tai[sense_codons]), 4), "\n")

compute_tai_gene <- function(gene_counts, w_vals) {
  use     <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use   <- gene_counts[use]; w_use <- w_vals[use]
  n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_use)) / n_total)
}

tai_scores0 <- apply(counts_mat0, 1, compute_tai_gene, w_vals = w_tai)
names(tai_scores0) <- rownames(counts_mat0)

cat("tAI summary:\n")
print(summary(tai_scores0))
stopifnot("tAI must be in [0,1]" = all(tai_scores0 <= 1 + 1e-9, na.rm = TRUE))
cat("tAI range check passed\n")
```

## CSC — Codon Stabilisation Coefficient

```{r csc}
predict_human <- iCodon::predict_stability("human")

cds_char0     <- as.character(cds_frame0)
unique_seqs0  <- unique(cds_char0)
cat("Unique CDS sequences:", length(unique_seqs0), "\n")

unique_scores0 <- predict_human(unique_seqs0)
names(unique_scores0) <- unique_seqs0

csc_scores0 <- unique_scores0[cds_char0]
names(csc_scores0) <- names(cds_frame0)

cat("CSC summary (iCodon human):\n")
print(summary(csc_scores0))
```

## FOP — Frequency of Optimal Codons

Reference: top 10% of MDA-MB-231 genes by mean TE in sictrl normoxia.

```{r fop}
te_condition <- read_csv(here("output", "condition_level_te.csv"), show_col_types = FALSE)

te_ranked <- te_condition %>%
  filter(condition == "normoxia", treatment == "sictrl") %>%
  group_by(gene_id) %>%
  summarise(mean_te = mean(log2FoldChange, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(mean_te)) %>%
  arrange(desc(mean_te))

top10_cutoff <- ceiling(nrow(te_ranked) * 0.10)
top10_clean  <- sub("\\..*", "", te_ranked$gene_id[seq_len(top10_cutoff)])
cat("Top 10% TE reference genes:", length(top10_clean), "\n")

ref_tx_fop <- tx_map$transcript_id_clean[tx_map$gene_id_clean %in% top10_clean]
ref_tx_fop <- ref_tx_fop[ref_tx_fop %in% rownames(counts_mat0)]
cat("Reference transcripts in codon matrix:", length(ref_tx_fop), "\n")

ref_counts_fop <- colSums(counts_mat0[ref_tx_fop, , drop = FALSE])

optimal_codons <- sapply(aa_to_codons, function(codons) {
  codons[which.max(ref_counts_fop[codons])]
})
optimal_codon_vec <- unname(optimal_codons)

sense_cols_fop  <- setdiff(colnames(counts_mat0), stop_codons)
sense_counts_fop <- rowSums(counts_mat0[, sense_cols_fop, drop = FALSE])
opti_counts_fop  <- rowSums(counts_mat0[, optimal_codon_vec, drop = FALSE])
fop_scores0      <- opti_counts_fop / sense_counts_fop
fop_scores0[sense_counts_fop == 0] <- NA_real_
names(fop_scores0) <- rownames(counts_mat0)

cat("FOP summary:\n")
print(summary(fop_scores0))
stopifnot("FOP must be in [0,1]" = all(fop_scores0 >= 0 & fop_scores0 <= 1, na.rm = TRUE))
cat("FOP range check passed\n")
```

## Join Codon Optimality

```{r join_codon_optimality}
codon_opt_df <- tibble(
  transcript_id_clean = rownames(counts_mat0),
  gc3 = gc3_scores,
  cai = cai_scores0,
  tai = tai_scores0,
  csc = csc_scores0,
  fop = fop_scores0
)

feature_matrix <- feature_matrix %>%
  left_join(codon_opt_df, by = "transcript_id_clean")

cat("Codon optimality joined. Matrix now:", nrow(feature_matrix), "x", ncol(feature_matrix), "\n")
cat("Non-NA: gc3 =", sum(!is.na(feature_matrix$gc3)),
    ", cai =", sum(!is.na(feature_matrix$cai)),
    ", tai =", sum(!is.na(feature_matrix$tai)),
    ", csc =", sum(!is.na(feature_matrix$csc)),
    ", fop =", sum(!is.na(feature_matrix$fop)), "\n")
```

```{r validate_codon_optimality}
# Coding check: value ranges
stopifnot(all(feature_matrix$gc3 >= 0 & feature_matrix$gc3 <= 100, na.rm = TRUE))
stopifnot(all(feature_matrix$cai >= 0 & feature_matrix$cai <= 1,   na.rm = TRUE))
stopifnot(all(feature_matrix$tai >= 0 & feature_matrix$tai <= 1,   na.rm = TRUE))
stopifnot(all(feature_matrix$fop >= 0 & feature_matrix$fop <= 1,   na.rm = TRUE))
cat("Range checks: gc3 [0–100], cai [0–1], tai [0–1], fop [0–1]: passed\n")

# Biological check: ribosomal proteins should have above-median CAI
# (they are the reference set, so their mean w ≈ 1 → high CAI is expected)
ribo_cai  <- feature_matrix %>% filter(str_detect(symbol, "^RPL|^RPS")) %>% pull(cai)
all_cai   <- feature_matrix$cai
cat("Median CAI — RPL/RPS:", round(median(ribo_cai, na.rm = TRUE), 3),
    " vs all:", round(median(all_cai, na.rm = TRUE), 3),
    "(expected: ribo > all)\n")
stopifnot("Ribosomal proteins should have above-median CAI" =
            median(ribo_cai, na.rm = TRUE) > median(all_cai, na.rm = TRUE))

# Biological check: FOP optimal codon for Leu should be CTG (most used in human)
cat("FOP optimal codons:\n"); print(sort(optimal_codon_vec))
stopifnot("CTG expected as optimal Leu codon" = "CTG" %in% optimal_codon_vec)
cat("CTG is optimal Leu codon: confirmed\n")

# Biological check: GC3 should be in the range expected for human coding sequences (~40–65%)
cat("GC3 median:", round(median(feature_matrix$gc3, na.rm = TRUE), 1),
    "% (expected ~45–60%)\n")
stopifnot(median(feature_matrix$gc3, na.rm = TRUE) > 30 &&
          median(feature_matrix$gc3, na.rm = TRUE) < 75)
```

---

# Section 12: Literature and Binary Features

Pattern mirrors `01_feature_extraction.Rmd` 4E/DAP5 and signature membership sections.
Gene symbol lookups are cell-line agnostic.

```{r lit_4e_dap5}
teleman_data <- read_csv("/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv",
                         show_col_types = FALSE) %>%
  filter(!is.na(symbol)) %>%
  group_by(symbol) %>%
  slice_max(order_by = baseMean, n = 1, with_ties = FALSE) %>%
  ungroup()

resistant_4e_symbols <- teleman_data %>% filter(log2FoldChange >  1, padj < 0.05) %>% pull(symbol)
sensitive_4e_symbols  <- teleman_data %>% filter(log2FoldChange < -1, padj < 0.05) %>% pull(symbol)
cat("4E-resistant genes:", length(resistant_4e_symbols), "\n")
cat("4E-sensitive genes:", length(sensitive_4e_symbols), "\n")

dap5_data <- read_xlsx(here("accessories", "dap5kd_riboseq_weber2022.xlsx")) %>%
  filter(!is.na(geneIDs), comment != "no")

dap5_repressed_symbols <- dap5_data %>% filter(comment == "up")   %>% pull(geneIDs)
dap5_promoted_symbols  <- dap5_data %>% filter(comment == "down") %>% pull(geneIDs)
cat("DAP5-repressed (up with KD):", length(dap5_repressed_symbols), "\n")
cat("DAP5-promoted (down with KD):", length(dap5_promoted_symbols), "\n")
```

```{r sig_membership}
all_sigs <- read_csv(here("output", "genesets", "all_te_signatures.csv"),
                     show_col_types = FALSE)

selected_sig_names <- c(
  "hypoxia_3d_promotes_TE_1hr_lfc0.5", "hypoxia_3d_inhibits_TE_1hr_lfc0.5",
  "hypoxia_3d_promotes_TE_4hr_lfc0.5", "hypoxia_3d_inhibits_TE_4hr_lfc0.5",
  "hypoxia_3d_promotes_TE_24hr_lfc0.5", "hypoxia_3d_inhibits_TE_24hr_lfc0.5",
  "hypoxia_3e_promotes_TE_1hr_lfc0.5", "hypoxia_3e_inhibits_TE_1hr_lfc0.5",
  "hypoxia_3e_promotes_TE_4hr_lfc0.5", "hypoxia_3e_inhibits_TE_4hr_lfc0.5",
  "hypoxia_3e_promotes_TE_24hr_lfc0.5", "hypoxia_3e_inhibits_TE_24hr_lfc0.5",
  "normoxia_3d_promotes_TE_combined_lfc0.5", "normoxia_3d_inhibits_TE_combined_lfc0.5",
  "normoxia_3e_promotes_TE_combined_lfc0.5", "normoxia_3e_inhibits_TE_combined_lfc0.5",
  "hypoxia_response_increases_TE_1hr_lfc0.5", "hypoxia_response_decreases_TE_1hr_lfc0.5",
  "hypoxia_response_increases_TE_4hr_lfc0.5", "hypoxia_response_decreases_TE_4hr_lfc0.5",
  "hypoxia_response_increases_TE_24hr_lfc0.5", "hypoxia_response_decreases_TE_24hr_lfc0.5",
  "hypoxia_response_increases_TE_1hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_1hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_1hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_1hr_inhibited_by_3e_lfc0.5",
  "hypoxia_response_increases_TE_4hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_4hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_4hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_4hr_inhibited_by_3e_lfc0.5",
  "hypoxia_response_increases_TE_24hr_inhibited_by_3d_kd_lfc0.5",
  "hypoxia_response_decreases_TE_24hr_inhibited_by_3d_lfc0.5",
  "hypoxia_response_increases_TE_24hr_inhibited_by_3e_kd_lfc0.5",
  "hypoxia_response_decreases_TE_24hr_inhibited_by_3e_lfc0.5"
)

sig_features <- all_sigs %>%
  filter(gs_name %in% selected_sig_names) %>%
  dplyr::select(symbol = db_gene_symbol, gs_name) %>%
  distinct() %>%
  mutate(member = 1L) %>%
  pivot_wider(names_from = gs_name, values_from = member,
              values_fill = 0L, names_prefix = "sig_")

cat("Signature features:", ncol(sig_features) - 1, "signatures\n")
cat("Genes with any membership:", nrow(sig_features), "\n")
```

```{r join_lit_features}
feature_matrix <- feature_matrix %>%
  mutate(
    is_4e_resistant   = as.integer(symbol %in% resistant_4e_symbols),
    is_4e_sensitive   = as.integer(symbol %in% sensitive_4e_symbols),
    is_dap5_repressed = as.integer(symbol %in% dap5_repressed_symbols),
    is_dap5_promoted  = as.integer(symbol %in% dap5_promoted_symbols)
  ) %>%
  left_join(sig_features, by = "symbol") %>%
  mutate(across(starts_with("sig_"), ~replace_na(., 0L)))

# Drop any sig_ columns that are all zero (no MCF7-SIX1 genes present)
sig_cols    <- colnames(feature_matrix)[startsWith(colnames(feature_matrix), "sig_")]
all_zero_sig <- sig_cols[sapply(sig_cols, function(col) sum(feature_matrix[[col]]) == 0)]
if (length(all_zero_sig) > 0) {
  feature_matrix <- feature_matrix %>% dplyr::select(-all_of(all_zero_sig))
  cat("Dropped", length(all_zero_sig), "all-zero sig_ columns\n")
}

cat("Literature features joined. Matrix now:", nrow(feature_matrix), "x", ncol(feature_matrix), "\n")
cat("sig_* columns remaining:", sum(startsWith(colnames(feature_matrix), "sig_")), "\n")
```

---

# Section 13: Kozak Features

Extracts Kozak window (last 6 nt of 5'UTR + first 4 nt of CDS) for MCF7-SIX1
selected transcripts. Pattern mirrors `01h_feature_extraction_kozak.Rmd`.
Reuses `utr5` GRangesList already loaded in Section 10.

```{r kozak_sequences}
# 5'UTR sequences — reuse utr5_seq_all from Section 10 (avoids repeat genome extraction)
utr5_seq_kozak <- utr5_seq_all %>%
  dplyr::select(transcript_id_clean, utr5_sequence) %>%
  filter(transcript_id_clean %in% tx_map$transcript_id_clean)

# CDS sequences (first 4 nt of each CDS in cds_frame0)
cds_seq_kozak <- tibble(
  transcript_id_clean = names(cds_frame0),
  cds_sequence        = as.character(cds_frame0)
) %>%
  filter(transcript_id_clean %in% tx_map$transcript_id_clean)

cat("5'UTR sequences for Kozak:", nrow(utr5_seq_kozak), "\n")
cat("CDS sequences for Kozak:", nrow(cds_seq_kozak), "\n")
```

```{r kozak_window}
seq_joined_koz <- inner_join(utr5_seq_kozak, cds_seq_kozak, by = "transcript_id_clean") %>%
  mutate(
    utr5_len     = nchar(utr5_sequence),
    utr5_tail    = str_sub(utr5_sequence, start = pmax(1L, utr5_len - 5L)),
    cds_head     = str_sub(cds_sequence, 1, 4),
    kozak_window = paste0(utr5_tail, cds_head)
  )

cat("Transcripts with Kozak window:", nrow(seq_joined_koz), "\n")
```

```{r kozak_score}
match_kozak <- function(seq) {
  if (nchar(seq) < 10) return("unknown_short")
  if (grepl("^(GCCACC|GCCGCC)ATGG$", seq)) return("optimal")
  if (grepl("^...[AG]..ATGG$", seq))        return("strong")
  if (grepl("^...[AG]..ATG[ACT]$", seq))    return("moderate")
  if (grepl("^...[CT]..ATGG$", seq))        return("moderate")
  return("unknown")
}

score_map <- c(unknown = 0L, moderate = 1L, strong = 2L, optimal = 3L, unknown_short = NA_integer_)

seq_scored_koz <- seq_joined_koz %>%
  mutate(
    kozak_category = sapply(kozak_window, match_kozak),
    kozak_score    = unname(score_map[kozak_category]),
    kozak_optimal  = as.integer(kozak_score == 3L),
    kozak_category = if_else(kozak_category == "unknown_short", NA_character_, kozak_category)
  )

cat("Kozak categories:\n")
print(table(seq_scored_koz$kozak_category, useNA = "ifany"))
```

```{r kozak_pwm}
kozak_pwm <- matrix(
  c(0.23, 0.22, 0.27, 0.28,
    0.25, 0.28, 0.23, 0.24,
    0.21, 0.34, 0.23, 0.22,
    0.71, 0.09, 0.13, 0.07,
    0.27, 0.28, 0.19, 0.26,
    0.25, 0.39, 0.15, 0.21,
    0.17, 0.12, 0.54, 0.17),
  nrow = 7, ncol = 4, byrow = TRUE,
  dimnames = list(c("m6","m5","m4","m3","m2","m1","p4"), c("A","C","G","T"))
)
log_odds_koz <- log2((kozak_pwm + 0.001) / 0.25)
pos_indices  <- c(1, 2, 3, 4, 5, 6, 10)
pos_names    <- c("m6","m5","m4","m3","m2","m1","p4")

score_pwm <- function(seq) {
  if (nchar(seq) < 10) return(NA_real_)
  nts <- strsplit(seq, "")[[1]]
  s   <- 0
  for (i in seq_along(pos_indices)) {
    nt <- nts[pos_indices[i]]
    if (!nt %in% c("A","C","G","T")) return(NA_real_)
    s <- s + log_odds_koz[pos_names[i], nt]
  }
  s
}

seq_scored_koz <- seq_scored_koz %>%
  mutate(kozak_pwm_score = sapply(kozak_window, score_pwm))

cat("kozak_pwm_score summary:\n")
print(summary(seq_scored_koz$kozak_pwm_score))
```

```{r join_kozak}
kozak_scores_df <- seq_scored_koz %>%
  dplyr::select(transcript_id_clean, kozak_score, kozak_optimal, kozak_pwm_score)

feature_matrix <- feature_matrix %>%
  left_join(kozak_scores_df, by = "transcript_id_clean")

stopifnot(all(na.omit(feature_matrix$kozak_score) %in% 0:3))
cat("Kozak features joined. Matrix now:", nrow(feature_matrix), "x", ncol(feature_matrix), "\n")
cat("kozak_score non-NA:", sum(!is.na(feature_matrix$kozak_score)), "\n")
```

```{r validate_kozak}
# Biological check: mean PWM score must increase monotonically with ordinal tier (0→3)
# — validates that the regex tiers and PWM are internally consistent
kozak_by_tier <- feature_matrix %>%
  filter(!is.na(kozak_score)) %>%
  group_by(kozak_score) %>%
  summarise(mean_pwm = mean(kozak_pwm_score, na.rm = TRUE), n = n(), .groups = "drop") %>%
  arrange(kozak_score)
cat("Mean PWM score by Kozak tier (must increase 0→3):\n")
print(kozak_by_tier)
stopifnot("PWM scores must increase monotonically with kozak_score tier" =
            all(diff(kozak_by_tier$mean_pwm) > 0))
cat("Kozak PWM monotonicity check: passed\n")

# Expected rates (mirrors 01h_ quality control)
na_rate  <- mean(is.na(feature_matrix$kozak_score)) * 100
opt_rate <- mean(feature_matrix$kozak_optimal, na.rm = TRUE) * 100
cat("NA rate (no/short 5'UTR):", round(na_rate, 1), "% (expected ~20–40%)\n")
cat("Optimal Kozak rate:      ", round(opt_rate, 1), "% (expected ~10–25%)\n")
```

---

# Section 14: CNOT3 External Features

Joins CNOT3 selective riboseq enrichment and codon stability scores from Zhu et al. 2024.
These are external datasets keyed by Ensembl gene ID — applicable to any human gene.

```{r cnot3_external}
cnot3_ribo <- readRDS(here("output", "predictive_modeling", "cnot3_riboseq_enrichment.rds")) %>%
  dplyr::select(ensembl_gene, cnot3_ribo_enrichment = neg_log10_min_fdr)

# cnot3_codon_stability_scores.rds is per-codon Spearman r (columns: codons, spearman_r, n_genes, rank)
# Compute per-gene weighted score: sum(spearman_r[codon] * freq[codon]) across all sense codons
codon_spearman <- readRDS(here("output", "predictive_modeling", "cnot3_codon_stability_scores.rds"))
weight_vec    <- setNames(codon_spearman$spearman_r, codon_spearman$codons)
shared_codons <- intersect(names(weight_vec), colnames(counts_mat0))
cat("Codons with Spearman weights:", length(shared_codons), "\n")

sense_in_mat0 <- intersect(sense_codons, colnames(counts_mat0))
total_sense0  <- rowSums(counts_mat0[, sense_in_mat0, drop = FALSE])

w_sub   <- weight_vec[shared_codons]
cnt_sub <- counts_mat0[, shared_codons, drop = FALSE]

cnot3_ws <- as.numeric(cnt_sub %*% w_sub) / total_sense0
cnot3_ws[total_sense0 == 0] <- NA_real_
names(cnot3_ws) <- rownames(counts_mat0)

cat("cnot3_weighted_codon_score summary:\n")
print(summary(cnot3_ws))

cnot3_codon_df <- tibble(
  transcript_id_clean        = rownames(counts_mat0),
  cnot3_weighted_codon_score = cnot3_ws
)

feature_matrix <- feature_matrix %>%
  left_join(cnot3_ribo,     by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(cnot3_codon_df, by = "transcript_id_clean")

cat("CNOT3 features joined. Coverage:\n")
cat("  cnot3_ribo_enrichment:     ", sum(!is.na(feature_matrix$cnot3_ribo_enrichment)),
    "/", nrow(feature_matrix), "\n")
cat("  cnot3_weighted_codon_score:", sum(!is.na(feature_matrix$cnot3_weighted_codon_score)),
    "/", nrow(feature_matrix), "\n")
```

---

# Section 15: External Stability Features

Joins SLAM-seq half-lives and riboseq occupancy from Zhu 2024, Hia 2026, and Karner 2026.
Mirrors `01i_feature_extraction_external_stability.Rmd` exactly.

```{r external_stability}
cnot3_hl <- readRDS(here("output", "predictive_modeling", "cnot3_slamseq_halflives.rds")) %>%
  dplyr::select(
    ensembl_gene,
    zhu2024_ctrl_halflife          = sgNT_halflife,
    zhu2024_cnot3ko_halflife_logfc = log2_ratio
  )

dhx29_slam <- readRDS(here("output", "predictive_modeling", "dhx29_slamseq_halflives.rds")) %>%
  dplyr::select(
    ensembl_gene,
    hia2026_ctrl_halflife        = Ctrl_Half_life,
    hia2026_dhx29_halflife_logfc = logFC
  )

dhx29_ribo <- readRDS(here("output", "predictive_modeling", "dhx29_riboseq_occupancy.rds")) %>%
  group_by(ensembl_gene) %>%
  summarise(hia2026_dhx29_occupancy = mean(dhx29_occupancy_score, na.rm = TRUE),
            .groups = "drop")

karner <- readRDS(here("output", "predictive_modeling", "goodarzi_231_ct_ratios.rds")) %>%
  dplyr::select(ensembl_gene, karner2026_mda231_log2_ct = log2_ct)

feature_matrix_ext <- feature_matrix %>%
  left_join(cnot3_hl,   by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(dhx29_slam, by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(dhx29_ribo, by = c("gene_id_clean" = "ensembl_gene")) %>%
  left_join(karner,     by = c("gene_id_clean" = "ensembl_gene"))

ext_cols <- c("zhu2024_ctrl_halflife", "zhu2024_cnot3ko_halflife_logfc",
              "hia2026_ctrl_halflife", "hia2026_dhx29_halflife_logfc",
              "hia2026_dhx29_occupancy", "karner2026_mda231_log2_ct")
cat("External stability feature coverage:\n")
for (col in ext_cols) {
  cat(" ", col, ":", sum(!is.na(feature_matrix_ext[[col]])), "/", nrow(feature_matrix_ext), "\n")
}
```

---

# Section 16: Save External Stability Matrix

```{r save_external_stability}
saveRDS(feature_matrix_ext,
        here("output", "predictive_modeling", "feature_matrix_mcf7six1_external_stability.rds"))
cat("Saved: output/predictive_modeling/feature_matrix_mcf7six1_external_stability.rds\n")
cat("Rows:", nrow(feature_matrix_ext), "| Columns:", ncol(feature_matrix_ext), "\n")

cat("\n=== FEATURE COLUMN SUMMARY ===\n")
all_cols <- colnames(feature_matrix_ext)
cat("bg_rscu_*        :", sum(startsWith(all_cols, "bg_rscu_")), "\n")
cat("codon_freq_*     :", sum(startsWith(all_cols, "codon_freq_")), "\n")
cat("Region lengths   :", sum(all_cols %in% c("utr5_length","utr3_length","total_length","log2_utr5_length","log2_utr3_length","log2_cds_length","log2_total_length","utr5_gc","utr3_gc","cds_gc","transcript_gc","gc3")), "\n")
cat("Codon optimality :", sum(all_cols %in% c("cai","tai","csc","fop")), "\n")
cat("4E/DAP5 binary   :", sum(all_cols %in% c("is_4e_resistant","is_4e_sensitive","is_dap5_repressed","is_dap5_promoted")), "\n")
cat("sig_*            :", sum(startsWith(all_cols, "sig_")), "\n")
cat("Kozak            :", sum(all_cols %in% c("kozak_score","kozak_optimal","kozak_pwm_score")), "\n")
cat("CNOT3 external   :", sum(all_cols %in% c("cnot3_ribo_enrichment","cnot3_weighted_codon_score")), "\n")
cat("Stability ext.   :", sum(all_cols %in% ext_cols), "\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
