---
title: "44 — Background-Corrected RSCU: eIF3d-Promoted vs Control (Normoxia 1+4hr Time-Regressed)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Background-corrected RSCU for eIF3d-promoted genes vs size-matched negative
controls at normoxia (1hr + 4hr time-regressed DESeq2 results).

bg_RSCU_j = (count_j_tx / aa_total_tx) / (mean per-transcript proportion of
codon j across all expressed mRNAs). bg_RSCU = 1 means same usage as
transcriptome background; > 1 over-used; < 1 under-used.

RSCU is computed **per transcript** and averaged across transcripts, giving each
gene equal weight regardless of CDS length.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(ggrepel)
```

---

# Section 1: Setup — CDS Sequences & Codon Counts

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)

shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  ct  <- codonTable(seqs)
  mat <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

cds_frame1  <- shift_frame(cds_frame0, 1)
cds_frame2  <- shift_frame(cds_frame0, 2)
counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

codon_gc3 <- tibble(
  codon          = sense_codons,
  gc3_nucleotide = substr(sense_codons, 3, 3),
  gc3_class      = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending")
)

tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

# Section 2: Gene Set Loading

```{r gene_sets}
tc <- read_csv(
  here("output", "genesets",
       "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
  show_col_types = FALSE
)

lfc_cutoff   <- 0.5
promotes_ids <- tc %>%
  filter(te_lfc > lfc_cutoff, !is.na(te_padj), te_padj < 0.05) %>%
  mutate(ensembl_gene = sub("\\..*", "", gene_id)) %>%
  pull(ensembl_gene) %>%
  unique()

cat("Normoxia 1+4hr promoted genes (te_lfc >", lfc_cutoff, ", padj < 0.05):",
    length(promotes_ids), "\n")

neg_pool <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

neg_filtered <- neg_pool %>% filter(!ensembl_gene %in% promotes_ids)
set.seed(9)
neg_ctrl_ids <- sample(neg_filtered$ensembl_gene,
                       size = length(promotes_ids), replace = FALSE)
cat("Neg ctrl (size-matched, single draw seed=9):", length(neg_ctrl_ids), "\n")

stopifnot(
  "No promoted genes found"                 = length(promotes_ids) > 0,
  "Neg ctrl pool smaller than promotes set" = nrow(neg_filtered) >= length(promotes_ids),
  "Overlap between promotes and neg ctrl"   = length(intersect(promotes_ids, neg_ctrl_ids)) == 0
)

promotes_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% promotes_ids) %>%
  pull(transcript_id_clean)
neg_ctrl_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% neg_ctrl_ids) %>%
  pull(transcript_id_clean)

promotes_tx <- promotes_tx[promotes_tx %in% rownames(counts_mat0)]
neg_ctrl_tx <- neg_ctrl_tx[neg_ctrl_tx %in% rownames(counts_mat0)]
cat("Promotes transcripts:", length(promotes_tx),
    "| Neg ctrl transcripts:", length(neg_ctrl_tx), "\n")
```

---

# Section 3: Background-Corrected RSCU

```{r bg_rscu_background_props}
# Background = mean per-transcript codon proportion across all expressed mRNAs.
# Each gene gets equal weight regardless of CDS length.
bg_props <- imap(aa_to_codons, function(codons, aa) {
  codons  <- intersect(codons, colnames(counts_mat0))
  if (length(codons) == 0) return(NULL)
  sub_cnt <- counts_mat0[, codons, drop = FALSE]
  aa_tot  <- rowSums(sub_cnt)
  valid   <- aa_tot > 0
  if (!any(valid)) return(NULL)
  prop_mat <- sub_cnt[valid, , drop = FALSE] / aa_tot[valid]
  colMeans(prop_mat)
})
bg_props <- Filter(Negate(is.null), bg_props)

cat("Background proportions computed for", length(bg_props), "amino acids\n")
prop_sums <- sapply(bg_props, sum)
cat("Background proportions sum check (should all be ~1):",
    round(range(prop_sums), 4), "\n")
```

```{r compute_bg_rscu_function}
compute_bg_rscu_matrix <- function(counts_mat, bg_props) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(bg_props)) {
    props  <- bg_props[[aa]]
    codons <- intersect(names(props), colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt <- counts_mat[, codons, drop = FALSE]
    aa_tot  <- rowSums(sub_cnt)
    valid   <- aa_tot > 0
    for (codon in codons) {
      bp <- props[codon]
      if (bp > 0 && any(valid)) {
        rscu_mat[valid, codon] <- (sub_cnt[valid, codon] / aa_tot[valid]) / bp
      }
    }
  }
  rscu_mat
}
```

```{r compute_bg_rscu}
bg_rscu_mat0 <- compute_bg_rscu_matrix(counts_mat0, bg_props)
cat("Background RSCU matrix:", nrow(bg_rscu_mat0), "×", ncol(bg_rscu_mat0), "\n")
cat("Range:", round(min(bg_rscu_mat0, na.rm = TRUE), 3), "—",
    round(max(bg_rscu_mat0, na.rm = TRUE), 3), "\n")

bg_mean_per_codon <- colMeans(bg_rscu_mat0, na.rm = TRUE)
bg_mean_per_codon <- bg_mean_per_codon[sense_codons[sense_codons %in% names(bg_mean_per_codon)]]
cat("Mean bg_RSCU across all expressed mRNAs — range:",
    round(range(bg_mean_per_codon, na.rm = TRUE), 3), "\n")
cat("(Should be ≈ 1 for all codons by construction)\n")
```

```{r rscu_summary_function}
# Summarize per-transcript RSCU/bg-RSCU values into group-level stats.
summarize_rscu <- function(rscu_mat, tx_ids) {
  tx_ids <- tx_ids[tx_ids %in% rownames(rscu_mat)]
  sub    <- rscu_mat[tx_ids, sense_codons, drop = FALSE]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(sub))
    map_dfr(codons, function(codon) {
      vals <- sub[, codon]
      vals <- vals[!is.na(vals)]
      tibble(
        amino_acid  = aa,
        codon       = codon,
        mean_rscu   = mean(vals),
        median_rscu = median(vals),
        se_rscu     = if (length(vals) > 1) sd(vals) / sqrt(length(vals)) else NA_real_,
        n           = length(vals)
      )
    })
  })
}
```

```{r frame_helper_functions}
# Helper functions used by the bg-RSCU frameshift control plot.
summarize_delta_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      tibble(amino_acid = aa, codon = codon,
             delta = mean(p_vals) - mean(n_vals),
             frame = frame_label)
    })
  })
}

per_tx_sig_rscu <- function(rscu_mat, promotes_ids, neg_ctrl_ids, frame_label) {
  p_ids <- promotes_ids[promotes_ids %in% rownames(rscu_mat)]
  n_ids <- neg_ctrl_ids[neg_ctrl_ids  %in% rownames(rscu_mat)]
  imap_dfr(aa_to_codons, function(codons, aa) {
    if (length(codons) < 2) return(NULL)
    codons <- intersect(codons, colnames(rscu_mat))
    map_dfr(codons, function(codon) {
      p_vals <- rscu_mat[p_ids, codon]; p_vals <- p_vals[!is.na(p_vals)]
      n_vals <- rscu_mat[n_ids, codon]; n_vals <- n_vals[!is.na(n_vals)]
      if (length(p_vals) < 3 || length(n_vals) < 3) {
        return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_, frame = frame_label))
      }
      wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
      tibble(amino_acid = aa, codon = codon, p_val = wt$p.value, frame = frame_label)
    })
  })
}
```

---

# Section 4: Group Summaries & Wilcoxon Tests

```{r bg_rscu_group_summaries}
bg_rscu_promotes <- summarize_rscu(bg_rscu_mat0, promotes_tx)
bg_rscu_neg      <- summarize_rscu(bg_rscu_mat0, neg_ctrl_tx)

cat("Background RSCU summary — promotes:", nrow(bg_rscu_promotes), "codon-rows\n")
```

```{r bg_rscu_delta}
bg_rscu_combined <- bg_rscu_promotes %>%
  dplyr::select(amino_acid, codon, promotes_mean = mean_rscu,
                promotes_se = se_rscu, promotes_n = n) %>%
  left_join(
    bg_rscu_neg %>% dplyr::select(amino_acid, codon, neg_mean = mean_rscu,
                                  neg_se = se_rscu, neg_n = n),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta_bg_rscu = promotes_mean - neg_mean)

bg_wilcox_df <- imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
  codons <- intersect(codons, colnames(bg_rscu_mat0))
  map_dfr(codons, function(codon) {
    p_vals <- bg_rscu_mat0[promotes_tx[promotes_tx %in% rownames(bg_rscu_mat0)], codon]
    n_vals <- bg_rscu_mat0[neg_ctrl_tx[neg_ctrl_tx %in% rownames(bg_rscu_mat0)], codon]
    p_vals <- p_vals[!is.na(p_vals)]
    n_vals <- n_vals[!is.na(n_vals)]
    if (length(p_vals) < 3 || length(n_vals) < 3) {
      return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_))
    }
    wt <- wilcox.test(p_vals, n_vals, exact = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val = wt$p.value)
  })
}) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))

bg_rscu_combined <- bg_rscu_combined %>%
  left_join(bg_wilcox_df %>% dplyr::select(amino_acid, codon, p_val, p_adj),
            by = c("amino_acid", "codon"))

cat("Top 10 codons by |Δ background RSCU|:\n")
print(bg_rscu_combined %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_bg_rscu))) %>%
        dplyr::select(codon, amino_acid, promotes_mean, neg_mean, delta_bg_rscu, p_adj) %>%
        head(10))
```

---

# Section 5: Plots

```{r bg_rscu_plot_sidebyside, fig.width=11, fig.height=9}
plot_df_bg_a <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(cols = c(promotes_mean, neg_mean),
               names_to = "group", values_to = "mean_bg_rscu") %>%
  mutate(
    se      = if_else(group == "promotes_mean", promotes_se, neg_se),
    group   = recode(group,
                     promotes_mean = "eIF3d-promoted",
                     neg_mean      = "Neg ctrl"),
    group   = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, mean_bg_rscu)
  )

p_bg_a <- ggplot(plot_df_bg_a, aes(x = codon, y = mean_bg_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = mean_bg_rscu - se, ymax = mean_bg_rscu + se),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "Background-corrected RSCU: eIF3d-promoted vs neg ctrl (Normoxia 1+4hr time-regressed)",
    subtitle = "Mean per-transcript bg-RSCU ± SE; dashed = 1 (same rate as transcriptome background)",
    x = "Codon", y = "Mean background-corrected RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_bg_a
ggsave(here("plots", "bg_rscu_sidebyside_normoxia_1and4hr.pdf"), p_bg_a, width = 11, height = 9)
```

```{r bg_rscu_plot_delta, fig.width=10, fig.height=8}
plot_df_bg_b <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta_bg_rscu >= 0, "promotes", "neg_ctrl"),
    sig_label = case_when(!is.na(p_adj) & p_adj < 0.05 ~ "*", TRUE ~ ""),
    codon     = reorder(codon, delta_bg_rscu)
  )

p_bg_b <- ggplot(plot_df_bg_b, aes(x = codon, y = delta_bg_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label, y = delta_bg_rscu + sign(delta_bg_rscu) * 0.015),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(promotes = "#E69F00", neg_ctrl = "#000000"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ background-corrected RSCU: eIF3d-promoted minus neg ctrl (Normoxia 1+4hr time-regressed)",
    subtitle = "Mean per-transcript bg-RSCU difference; * = BH-adjusted Wilcoxon q < 0.05",
    x = "Codon", y = "Δ mean bg-RSCU (promotes − neg ctrl)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_bg_b
ggsave(here("plots", "bg_rscu_delta_normoxia_1and4hr.pdf"), p_bg_b, width = 10, height = 8)
```

```{r bg_rscu_violin_by_aa, fig.width=12, fig.height=8}
sig_aas <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa, !is.na(p_adj), p_adj < 0.05) %>%
  pull(amino_acid) %>%
  unique()

cat("Amino acids with ≥1 significant codon (bg RSCU, BH q < 0.05):",
    length(sig_aas), "\n")
if (length(sig_aas) > 0) cat(paste(sig_aas, collapse = ", "), "\n")

if (length(sig_aas) > 0) {
  sig_codons_by_aa <- bg_rscu_combined %>%
    filter(amino_acid %in% sig_aas, !is.na(p_adj), p_adj < 0.05) %>%
    dplyr::select(amino_acid, codon)

  violin_df <- map_dfr(seq_len(nrow(sig_codons_by_aa)), function(i) {
    aa    <- sig_codons_by_aa$amino_acid[i]
    codon <- sig_codons_by_aa$codon[i]
    tx_promotes <- promotes_tx[promotes_tx %in% rownames(bg_rscu_mat0)]
    tx_neg      <- neg_ctrl_tx[neg_ctrl_tx  %in% rownames(bg_rscu_mat0)]
    p_vals <- bg_rscu_mat0[tx_promotes, codon]
    n_vals <- bg_rscu_mat0[tx_neg,      codon]
    bind_rows(
      tibble(codon = codon, amino_acid = aa, group = "eIF3d-promoted",
             bg_rscu = p_vals[!is.na(p_vals)]),
      tibble(codon = codon, amino_acid = aa, group = "Neg ctrl",
             bg_rscu = n_vals[!is.na(n_vals)])
    )
  }) %>%
    mutate(
      group    = factor(group, levels = c("eIF3d-promoted", "Neg ctrl")),
      aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")")
    ) %>%
    left_join(bg_rscu_combined %>% dplyr::select(codon, p_adj_label = p_adj),
              by = "codon") %>%
    mutate(codon_aa = paste0(codon, "\nq=", signif(p_adj_label, 2)))

  p_viol <- ggplot(violin_df, aes(x = group, y = bg_rscu, fill = group)) +
    geom_violin(alpha = 0.6, linewidth = 0.4, trim = TRUE) +
    geom_boxplot(width = 0.18, outlier.size = 0.5, linewidth = 0.4,
                 fill = "white", alpha = 0.8) +
    geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
    scale_fill_manual(values = c("eIF3d-promoted" = "#E69F00", "Neg ctrl" = "#000000"),
                      guide = "none") +
    facet_wrap(~ codon_aa, scales = "free_y", nrow = 2) +
    labs(
      title    = "Per-transcript background RSCU for significant codons (Normoxia 1+4hr time-regressed)",
      subtitle = "Codons with BH q < 0.05 between eIF3d-promoted and neg ctrl.\nDashed = 1 (same usage rate as transcriptome background).",
      x = NULL, y = "Per-transcript background-corrected RSCU"
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1),
      strip.text  = element_text(face = "bold", size = 9)
    )
  p_viol
  ggsave(here("plots", "bg_rscu_violin_by_aa_normoxia_1and4hr.pdf"), p_viol,
         width = max(8, length(sig_aas) * 2.5), height = 8)
} else {
  cat("No significant codons in background RSCU comparison — skipping violin plot.\n")
}
```

```{r bg_rscu_frame_ctrl, fig.width=12, fig.height=9}
bg_rscu_mat1 <- compute_bg_rscu_matrix(counts_mat1, bg_props)
bg_rscu_mat2 <- compute_bg_rscu_matrix(counts_mat2, bg_props)

bg_f0_delta <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(amino_acid, codon, delta = delta_bg_rscu) %>%
  mutate(frame = "Frame 0 (coding)")
bg_f1_delta <- summarize_delta_rscu(bg_rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)")
bg_f2_delta <- summarize_delta_rscu(bg_rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")

bg_frame_delta_df <- bind_rows(bg_f0_delta, bg_f1_delta, bg_f2_delta) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    frame    = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

bg_codon_order <- bg_f0_delta %>%
  filter(amino_acid %in% multi_aa) %>%
  arrange(amino_acid, delta) %>%
  pull(codon)
bg_frame_delta_df <- bg_frame_delta_df %>%
  mutate(codon = factor(codon, levels = bg_codon_order))

bg_sig_df <- bind_rows(
  per_tx_sig_rscu(bg_rscu_mat0, promotes_tx, neg_ctrl_tx, "Frame 0 (coding)"),
  per_tx_sig_rscu(bg_rscu_mat1, promotes_tx, neg_ctrl_tx, "Frame 1 (control)"),
  per_tx_sig_rscu(bg_rscu_mat2, promotes_tx, neg_ctrl_tx, "Frame 2 (control)")
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    q_val = p.adjust(p_val, method = "BH"),
    sig   = !is.na(q_val) & q_val < 0.05,
    frame = factor(frame, levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)"))
  )

bg_f0_sig   <- bg_sig_df %>% filter(frame == "Frame 0 (coding)") %>%
  dplyr::select(amino_acid, codon, f0_sig = sig)
bg_ctrl_sig <- bg_sig_df %>% filter(frame != "Frame 0 (coding)") %>%
  group_by(amino_acid, codon) %>%
  summarise(any_ctrl_sig = any(sig), .groups = "drop")
bg_f0_labels <- bg_f0_sig %>%
  left_join(bg_ctrl_sig, by = c("amino_acid", "codon")) %>%
  mutate(sig_label = case_when(
    f0_sig & !any_ctrl_sig ~ "✦",
    f0_sig &  any_ctrl_sig ~ "·",
    TRUE                   ~ ""
  )) %>%
  dplyr::select(amino_acid, codon, sig_label) %>%
  mutate(frame = factor("Frame 0 (coding)",
                        levels = c("Frame 0 (coding)", "Frame 1 (control)", "Frame 2 (control)")))

bg_aa_any_f0 <- bg_f0_labels %>% filter(sig_label != "") %>% pull(amino_acid) %>% unique()
bg_aa_clean  <- bg_f0_labels %>% filter(sig_label == "✦") %>% pull(amino_acid) %>% unique()

bg_frame_delta_df <- bg_frame_delta_df %>%
  mutate(aa_label = ifelse(amino_acid %in% bg_aa_any_f0,
                           paste0(aa_label, " *"), aa_label))

bg_aa_label_map <- bg_frame_delta_df %>% distinct(amino_acid, aa_label)
bg_facet_clean  <- bg_aa_label_map %>% filter(amino_acid %in% bg_aa_clean)

bg_rect_layer <- if (nrow(bg_facet_clean) > 0) {
  geom_rect(data = bg_facet_clean,
            aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = NA, color = "#0072B2", linewidth = 1.5)
} else {
  NULL
}

p_bg_c <- ggplot(bg_frame_delta_df, aes(x = codon, y = delta, fill = frame)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  bg_rect_layer +
  scale_fill_manual(
    values = c("Frame 0 (coding)"  = "#E69F00",
               "Frame 1 (control)" = "#999999",
               "Frame 2 (control)" = "#CCCCCC")
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ background-corrected RSCU by reading frame: frame 0 vs frameshift controls (Normoxia 1+4hr time-regressed)",
    subtitle = "Frames 1 & 2 shifted tiling of the same sequence; background from frame 0 CDS transcriptome proportions.\nBlue border: frame 0 sig, neither control sig. ✦: clean signal. *: frame 0 significant. [q < 0.05, BH Wilcoxon]",
    x = "Codon", y = "Δ mean bg-RSCU (promotes − neg ctrl)",
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_bg_c
ggsave(here("plots", "bg_rscu_frame_ctrl_normoxia_1and4hr.pdf"), p_bg_c, width = 12, height = 9)
```

---

# Save Results

```{r save_bg_rscu}
write_csv(bg_rscu_combined,
          here("output", "predictive_modeling", "bg_rscu_231_normoxia_1and4hr.csv"))
cat("Saved bg_rscu_combined to output/predictive_modeling/bg_rscu_231_normoxia_1and4hr.csv\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
