---
title: "47 — Delta-Delta bg-RSCU: Hypoxia vs Normoxia eIF3d Codon Preference (MCF7-SIX1)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Compares the condition-specificity of eIF3d codon preference between hypoxia
(1hr) and normoxia (1hr) in MCF7-SIX1 cells using pre-computed background-corrected
RSCU summaries from notebooks 41 and 46.

**Delta-delta** = Δbg-RSCU(hypoxia) − Δbg-RSCU(normoxia), where each Δ is
(eIF3d-promoted mean bg-RSCU) − (neg ctrl mean bg-RSCU). Positive delta-delta
= codon more preferred by eIF3d-promoted genes specifically in hypoxia; negative
= normoxia-specific preference.

**Prerequisite:** knit notebook 46 before this one to generate
`bg_rscu_mcf7six1_normoxia_1hr.csv`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
```

---

# Load Data

```{r load_data}
hyp <- read_csv(here("output", "predictive_modeling", "bg_rscu_mcf7six1_hypoxia.csv"),
                show_col_types = FALSE)
nor <- read_csv(here("output", "predictive_modeling", "bg_rscu_mcf7six1_normoxia_1hr.csv"),
                show_col_types = FALSE)

stopifnot(
  "Hypoxia CSV missing expected columns" =
    all(c("codon", "amino_acid", "delta_bg_rscu", "p_adj") %in% colnames(hyp)),
  "Normoxia CSV missing expected columns" =
    all(c("codon", "amino_acid", "delta_bg_rscu", "p_adj") %in% colnames(nor))
)

cat("Hypoxia codons:", nrow(hyp), "| Normoxia codons:", nrow(nor), "\n")
cat("Hypoxia sig (q < 0.05):", sum(hyp$p_adj < 0.05, na.rm = TRUE), "\n")
cat("Normoxia sig (q < 0.05):", sum(nor$p_adj < 0.05, na.rm = TRUE), "\n")
```

---

# Compute Delta-Delta

```{r compute_delta_delta}
# GC3 class from codon string — no external packages needed
codon_gc3 <- tibble(codon = unique(hyp$codon)) %>%
  mutate(gc3_class = if_else(substr(codon, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending"))

# Multi-synonymous amino acids only (Met/Trp have only 1 synonymous codon — RSCU always 1)
multi_aa <- hyp %>% count(amino_acid) %>% filter(n > 1) %>% pull(amino_acid)

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

comp <- hyp %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(codon, amino_acid, delta_hyp = delta_bg_rscu, p_adj_hyp = p_adj) %>%
  inner_join(
    nor %>%
      filter(amino_acid %in% multi_aa) %>%
      dplyr::select(codon, delta_nor = delta_bg_rscu, p_adj_nor = p_adj),
    by = "codon"
  ) %>%
  left_join(codon_gc3, by = "codon") %>%
  mutate(
    delta_delta = delta_hyp - delta_nor,
    sig = case_when(
      !is.na(p_adj_hyp) & p_adj_hyp < 0.05 & !is.na(p_adj_nor) & p_adj_nor < 0.05 ~ "Both",
      !is.na(p_adj_hyp) & p_adj_hyp < 0.05 ~ "Hypoxia only",
      !is.na(p_adj_nor) & p_adj_nor < 0.05 ~ "Normoxia only",
      TRUE ~ "Neither"
    ),
    sig         = factor(sig, levels = c("Both", "Hypoxia only", "Normoxia only", "Neither")),
    label_codon = if_else(sig != "Neither", codon, NA_character_)
  )

r_val <- cor(comp$delta_hyp, comp$delta_nor, use = "complete.obs")
cat("Codons in comparison:", nrow(comp), "\n")
cat("Pearson r (Δ bg-RSCU: hypoxia ~ normoxia):", round(r_val, 3), "\n\n")
cat("Significance breakdown:\n")
print(table(comp$sig))

cat("\nTop 10 codons by |delta-delta| (hypoxia − normoxia preference):\n")
print(comp %>%
  arrange(desc(abs(delta_delta))) %>%
  dplyr::select(codon, amino_acid, gc3_class, delta_hyp, delta_nor, delta_delta, sig) %>%
  head(10))

cat("\nCodons significant in both conditions:\n")
both_sig <- comp %>%
  filter(sig == "Both") %>%
  arrange(desc(abs(delta_delta))) %>%
  dplyr::select(codon, amino_acid, gc3_class, delta_hyp, delta_nor, delta_delta)
if (nrow(both_sig) > 0) print(both_sig) else cat("  None\n")
```

---

# Plots

## Scatter: Δbg-RSCU Hypoxia vs Normoxia

```{r plot_scatter, fig.width=6, fig.height=5.5}
sig_colours <- c(
  "Both"          = "#E69F00",
  "Hypoxia only"  = "#0072B2",
  "Normoxia only" = "#CC79A7",
  "Neither"       = "grey70"
)
sig_sizes <- c("Both" = 2.8, "Hypoxia only" = 2.2, "Normoxia only" = 2.2, "Neither" = 1.6)
sig_alpha  <- c("Both" = 1,   "Hypoxia only" = 0.9, "Normoxia only" = 0.9, "Neither" = 0.5)

axis_lim <- range(c(comp$delta_hyp, comp$delta_nor), na.rm = TRUE)
axis_lim <- axis_lim + diff(axis_lim) * c(-0.05, 0.05)

p_scatter <- ggplot(comp, aes(x = delta_nor, y = delta_hyp, colour = sig)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linewidth = 0.5,
              linetype = "dashed") +
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_point(aes(size = sig, alpha = sig)) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 25, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = sig_colours, name = "Significant (q < 0.05)") +
  scale_size_manual(values = sig_sizes, guide = "none") +
  scale_alpha_manual(values = sig_alpha, guide = "none") +
  annotate("text",
           x = axis_lim[1] + diff(axis_lim) * 0.02,
           y = axis_lim[2] - diff(axis_lim) * 0.02,
           label = paste0("r = ", round(r_val, 2)),
           hjust = 0, vjust = 1, size = 4, colour = "grey30") +
  coord_fixed(xlim = axis_lim, ylim = axis_lim) +
  labs(
    title    = "Δ bg-RSCU: eIF3d codon preference in hypoxia vs normoxia (MCF7-SIX1)",
    subtitle = paste0("Each point = one codon (multi-synonymous AAs). Dashed = identity line.\n",
                      "Points above diagonal: hypoxia-enriched preference. ",
                      "Below diagonal: normoxia-enriched."),
    x = "Δ bg-RSCU (MCF7-SIX1 normoxia 1hr, notebook 46)",
    y = "Δ bg-RSCU (MCF7-SIX1 hypoxia 1hr, notebook 41)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_scatter
ggsave(here("plots", "47_bg_rscu_delta_delta_mcf7six1_scatter.pdf"), p_scatter,
       width = 6, height = 5.5)
```

## Faceted Bar: Delta-Delta by Amino Acid

```{r plot_delta_delta_faceted, fig.width=10, fig.height=8}
plot_df <- comp %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = if_else(delta_delta >= 0, "hypoxia", "normoxia"),
    sig_label = if_else(sig == "Both", "*", ""),
    codon     = reorder(codon, delta_delta)
  )

p_facet <- ggplot(plot_df, aes(x = codon, y = delta_delta, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label,
                y = delta_delta + sign(delta_delta) * 0.01),
            size = 3.5, vjust = 0.5) +
  scale_fill_manual(
    values = c(hypoxia = "#E69F00", normoxia = "#000000"),
    labels = c(hypoxia = "Hypoxia-enriched", normoxia = "Normoxia-enriched"),
    name   = NULL,
    guide  = "none"
  ) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Delta-delta bg-RSCU: hypoxia minus normoxia eIF3d preference (MCF7-SIX1)",
    subtitle = "Positive = more preferred by eIF3d in hypoxia; negative = normoxia-specific.\n* = significant in both conditions (BH q < 0.05 in both notebooks 41 and 46).",
    x = "Codon", y = "Δ(Δ bg-RSCU) = hypoxia − normoxia"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_facet
ggsave(here("plots", "47_bg_rscu_delta_delta_mcf7six1_faceted.pdf"), p_facet,
       width = 10, height = 8)
```

## Ranked Residuals from Identity Line

```{r plot_residuals, fig.width=5.5, fig.height=9}
plot_resid <- comp %>%
  mutate(
    codon     = fct_reorder(codon, delta_delta),
    sig_label = if_else(sig != "Neither", as.character(codon), NA_character_)
  )

p_resid <- ggplot(plot_resid, aes(x = delta_delta, y = codon, colour = gc3_class)) +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_segment(aes(x = 0, xend = delta_delta, yend = codon),
               linewidth = 0.5, alpha = 0.7) +
  geom_point(aes(shape = sig), size = 2) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  scale_shape_manual(
    values = c(Both = 8, "Hypoxia only" = 17, "Normoxia only" = 15, Neither = 16),
    name   = "Significant (q < 0.05)"
  ) +
  labs(
    title    = "Ranked residuals: hypoxia vs normoxia eIF3d codon preference (MCF7-SIX1)",
    subtitle = "Residual = Δbg-RSCU(hypoxia) − Δbg-RSCU(normoxia)\nPositive = stronger eIF3d preference in hypoxia",
    x = "Δ(Δ bg-RSCU) = hypoxia − normoxia",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y     = element_text(size = 7, family = "mono"),
    legend.position = "top",
    legend.box      = "vertical"
  )
p_resid
ggsave(here("plots", "47_bg_rscu_residuals_mcf7six1_ranked.pdf"), p_resid,
       width = 5.5, height = 9)
```

---

# Codon Frequency Scatter: Hypoxia vs Normoxia

```{r codon_freq_scatter, fig.width=6, fig.height=5.5}
freq_hyp <- read_csv(
  here("output", "predictive_modeling", "codon_freq_mcf7six1_hypoxia.csv"),
  show_col_types = FALSE
)
freq_nor <- read_csv(
  here("output", "predictive_modeling", "codon_freq_mcf7six1_normoxia_1hr.csv"),
  show_col_types = FALSE
)

freq_comp <- freq_hyp %>%
  filter(amino_acid %in% multi_aa) %>%
  dplyr::select(codon, amino_acid, gc3_class, delta_hyp = delta_freq) %>%
  inner_join(
    freq_nor %>%
      filter(amino_acid %in% multi_aa) %>%
      dplyr::select(codon, delta_nor = delta_freq),
    by = "codon"
  )

r_freq <- cor(freq_comp$delta_hyp, freq_comp$delta_nor, use = "complete.obs")
cat("Codon frequency — Pearson r (hypoxia ~ normoxia):", round(r_freq, 3), "\n")
cat("Codons in comparison:", nrow(freq_comp), "\n")

freq_colours <- c("GC-ending" = "#009E73", "AT-ending" = "#CC79A7")

freq_axis_lim <- range(c(freq_comp$delta_hyp, freq_comp$delta_nor), na.rm = TRUE)
freq_axis_lim <- freq_axis_lim + diff(freq_axis_lim) * c(-0.05, 0.05)

p_freq_scatter <- ggplot(freq_comp,
                         aes(x = delta_nor, y = delta_hyp, colour = gc3_class)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linewidth = 0.5,
              linetype = "dashed") +
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_point(size = 2.2) +
  ggrepel::geom_text_repel(aes(label = codon), size = 2.8,
                           max.overlaps = 25, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = freq_colours, name = "Third position") +
  annotate("text",
           x = freq_axis_lim[1] + diff(freq_axis_lim) * 0.02,
           y = freq_axis_lim[2] - diff(freq_axis_lim) * 0.02,
           label = paste0("r = ", round(r_freq, 2)),
           hjust = 0, vjust = 1, size = 4, colour = "grey30") +
  coord_fixed(xlim = freq_axis_lim, ylim = freq_axis_lim) +
  labs(
    title    = "Δ codon frequency: eIF3d preference in hypoxia vs normoxia (MCF7-SIX1)",
    subtitle = "Each point = one codon (multi-synonymous AAs). Dashed = identity line.",
    x        = "Δ codon frequency (normoxia 1hr, notebook 46)",
    y        = "Δ codon frequency (hypoxia 1hr, notebook 41)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_freq_scatter
ggsave(here("plots", "47_codon_freq_scatter_mcf7six1.pdf"), p_freq_scatter,
       width = 6, height = 5.5)
```

---

# Save

```{r save}
write_csv(comp,
          here("output", "predictive_modeling", "bg_rscu_delta_delta_mcf7six1_hyp_vs_nor.csv"))
cat("Saved: output/predictive_modeling/bg_rscu_delta_delta_mcf7six1_hyp_vs_nor.csv\n")
cat("Columns:", paste(colnames(comp), collapse = ", "), "\n")
cat("Rows:", nrow(comp), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
