---
title: "49 — TE Quartiles vs Sequence Features (MCF7-SIX1, siCTRL Hypoxia vs Normoxia 1hr)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Examines how mRNA sequence features (GC content and region lengths) vary across
genes stratified by hypoxic translational response in MCF7-SIX1 control cells.
TE LFC = log2(TE hypoxia 1hr / TE normoxia 1hr) in siCTRL cells; genes are split
into four equal-size quartiles from most repressed (Q1) to most induced (Q4).

MCF7-SIX1 TE data from `all_norm_to_hyp_changes.csv` (column `LFC_te_sictrl`).
Feature matrix from `feature_matrix_mcf7six1_external_stability.rds` (nb43).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

---

# Section 1: Load & Join Data

```{r load_data}
eif3_dir <- "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia"

te_raw <- read_csv(
  file.path(eif3_dir, "counts/all_norm_to_hyp_changes.csv"),
  show_col_types = FALSE
)

stopifnot(
  "all_norm_to_hyp_changes.csv missing LFC_te_sictrl column" =
    "LFC_te_sictrl" %in% colnames(te_raw),
  "LFC_te_sictrl contains non-finite values" =
    all(is.finite(te_raw$LFC_te_sictrl[!is.na(te_raw$LFC_te_sictrl)]))
)

te_df <- te_raw %>%
  transmute(
    gene_id_clean = sub("\\..*", "", gene_id),
    te_change     = LFC_te_sictrl  # log2(hypoxia/normoxia) in siCTRL; positive = hypoxia-induced
  ) %>%
  filter(!is.na(te_change))

cat("MCF7-SIX1 TE LFC file:", nrow(te_df), "genes with non-NA values\n")
cat("te_change (LFC) range:", round(range(te_df$te_change), 3), "\n")
```

```{r load_feature_matrix}
fm <- readRDS(here("output", "predictive_modeling",
                   "feature_matrix_mcf7six1_external_stability.rds"))

needed_cols <- c(
  "gene_id_clean", "utr5_gc", "cds_gc", "utr3_gc",
  "utr5_length", "cds_length", "utr3_length", "gc3"
)
stopifnot(
  "Feature matrix missing required columns" =
    all(needed_cols %in% colnames(fm))
)
cat("Feature matrix:", nrow(fm), "rows (transcripts)\n")
```

```{r join_data}
# One representative row per gene — feature matrix is keyed by transcript
fm_genes <- fm %>%
  dplyr::select(gene_id_clean, symbol, utr5_gc, cds_gc, utr3_gc,
                utr5_length, cds_length, utr3_length, gc3) %>%
  distinct(gene_id_clean, .keep_all = TRUE)

joined <- te_df %>%
  inner_join(fm_genes, by = "gene_id_clean") %>%
  filter(!is.na(te_change))

cat("Genes after inner join:", nrow(joined), "\n")

stopifnot(
  "cds_gc out of range"  = all(joined$cds_gc  >= 0 & joined$cds_gc  <= 100, na.rm = TRUE),
  "utr5_gc out of range" = all(joined$utr5_gc >= 0 & joined$utr5_gc <= 100, na.rm = TRUE),
  "utr3_gc out of range" = all(joined$utr3_gc >= 0 & joined$utr3_gc <= 100, na.rm = TRUE),
  "cds_length > 0"       = all(joined$cds_length  > 0, na.rm = TRUE),
  "utr5_length >= 0"     = all(joined$utr5_length >= 0, na.rm = TRUE),
  "utr3_length >= 0"     = all(joined$utr3_length >= 0, na.rm = TRUE)
)
cat("All feature range checks passed.\n")
```

---

# Section 2: Quartile Assignment

```{r assign_quartiles}
plot_df <- joined %>%
  mutate(
    quartile = ntile(te_change, 4),
    quartile_label = factor(
      quartile,
      labels = c("Q1\n(low TE)", "Q2", "Q3", "Q4\n(high TE)")
    )
  )

cat("Genes per quartile:\n")
print(table(plot_df$quartile_label))

q_breaks <- quantile(plot_df$te_change, probs = c(0, 0.25, 0.5, 0.75, 1))
cat("\nTE LFC quartile boundaries:\n")
print(round(q_breaks, 3))
```

## Biological spot-checks

```{r spot_checks}
gene_quartile <- function(sym) {
  row <- plot_df %>% filter(symbol == sym)
  if (nrow(row) == 0) return("not found in joined data")
  paste0("Q", row$quartile[1], " (te_change = ", round(row$te_change[1], 3), ")")
}

cat("TIGAR:    ", gene_quartile("TIGAR"), "\n")
cat("EIF4EBP1: ", gene_quartile("EIF4EBP1"), "\n")
cat("EPAS1:    ", gene_quartile("EPAS1"), "\n")
cat("SLC2A1:   ", gene_quartile("SLC2A1"), "\n")
cat("ATF4:     ", gene_quartile("ATF4"), "\n")
```

## Directionality check: top hypoxia genes in Q4 after join

```{r directionality_check}
top50_orig_ids <- te_df %>%
  slice_max(te_change, n = 50, with_ties = FALSE) %>%
  pull(gene_id_clean)

top50_in_plot <- plot_df %>%
  filter(gene_id_clean %in% top50_orig_ids)

cat("Top 50 hypoxia-increased genes (siCTRL, original ranking):\n")
cat("  Retained after feature matrix join:", nrow(top50_in_plot), "/ 50\n")
cat("  In Q4:", sum(top50_in_plot$quartile == 4), "\n")
cat("  In Q3:", sum(top50_in_plot$quartile == 3), "\n")
cat("  In Q1 (directionality concern):", sum(top50_in_plot$quartile == 1), "\n")

stopifnot(
  "Top hypoxia-increased genes not predominantly in Q4 — check gene ID mismatch" =
    mean(top50_in_plot$quartile == 4) > 0.70,
  "Top hypoxia-increased genes appearing in Q1 — possible directionality reversal" =
    sum(top50_in_plot$quartile == 1) == 0
)
cat("Directionality check passed: top-50 genes are in Q4 with none in Q1.\n")

cat("\nTop 20 genes most increased in hypoxia (siCTRL, in plot_df):\n")
plot_df %>%
  slice_max(te_change, n = 20, with_ties = FALSE) %>%
  dplyr::select(symbol, te_change, quartile_label) %>%
  print(n = 20)
```

---

# Section 3: TE LFC Distribution by Quartile

```{r plot_distribution, fig.width=7, fig.height=4}
quartile_colours <- c(
  "Q1\n(low TE)" = "#2166AC",
  "Q2"           = "#92C5DE",
  "Q3"           = "#F4A582",
  "Q4\n(high TE)" = "#D6604D"
)

vlines <- tibble(
  xintercept = q_breaks[2:4],
  label      = paste0(c(25, 50, 75), "th pct")
)

p_dist <- ggplot(plot_df, aes(x = te_change, fill = quartile_label)) +
  geom_density(alpha = 0.35, linewidth = 0.5) +
  geom_vline(data = vlines, aes(xintercept = xintercept),
             linewidth = 0.4, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = quartile_colours, name = "Quartile") +
  labs(
    title    = "Distribution of TE LFC in siCTRL cells (MCF7-SIX1, hypoxia vs normoxia, 1hr)",
    subtitle = "TE LFC = log2(hypoxia/normoxia). Dashed lines = quartile boundaries.",
    x        = "TE LFC (hypoxia/normoxia, siCTRL)",
    y        = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "right")+
  xlim(-1,1)
p_dist
ggsave(here("plots", "49_te_change_distribution_by_quartile.pdf"),
       p_dist, width = 7, height = 4)
```

---

# Section 4: GC Content by Quartile

```{r plot_gc, fig.width=8, fig.height=4.5}
gc_long <- plot_df %>%
  dplyr::select(gene_id_clean, quartile_label, utr5_gc, cds_gc, utr3_gc) %>%
  pivot_longer(
    cols      = c(utr5_gc, cds_gc, utr3_gc),
    names_to  = "region",
    values_to = "gc_pct"
  ) %>%
  filter(!is.na(gc_pct)) %>%
  mutate(
    region = factor(
      region,
      levels = c("utr5_gc", "cds_gc", "utr3_gc"),
      labels = c("5' UTR", "CDS", "3' UTR")
    )
  )

kw_gc <- gc_long %>%
  group_by(region) %>%
  summarise(
    p_kw  = kruskal.test(gc_pct ~ quartile_label)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = if_else(
      p_kw < 0.001,
      "KW p < 0.001",
      paste0("KW p = ", signif(p_kw, 2))
    )
  )

p_gc <- ggplot(gc_long, aes(x = quartile_label, y = gc_pct, fill = quartile_label)) +
  geom_violin(alpha = 0.5, linewidth = 0.4, trim = TRUE) +
  geom_boxplot(width = 0.18, outlier.size = 0.3, linewidth = 0.4,
               fill = "white", alpha = 0.85) +
  geom_text(
    data        = kw_gc,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05, vjust = 1.4, size = 3, colour = "grey30"
  ) +
  scale_fill_manual(values = quartile_colours, guide = "none") +
  facet_wrap(~ region, scales = "free_y") +
  labs(
    title    = "GC content by TE quartile (MCF7-SIX1, siCTRL hypoxia vs normoxia, 1hr)",
    subtitle = "Q1 = most repressed in hypoxia; Q4 = most induced. KW = Kruskal-Wallis test across all 4 quartiles.",
    x        = "TE quartile",
    y        = "GC content (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
p_gc
ggsave(here("plots", "49_gc_content_by_te_quartile.pdf"),
       p_gc, width = 8, height = 4.5)
```

---

# Section 5: Region Lengths by Quartile

```{r plot_lengths, fig.width=8, fig.height=4.5}
len_long <- plot_df %>%
  dplyr::select(gene_id_clean, quartile_label, utr5_length, cds_length, utr3_length) %>%
  pivot_longer(
    cols      = c(utr5_length, cds_length, utr3_length),
    names_to  = "region",
    values_to = "length_nt"
  ) %>%
  filter(!is.na(length_nt), length_nt > 0) %>%
  mutate(
    region = factor(
      region,
      levels = c("utr5_length", "cds_length", "utr3_length"),
      labels = c("5' UTR", "CDS", "3' UTR")
    ),
    log2_length = log2(length_nt + 1)
  )

kw_len <- len_long %>%
  group_by(region) %>%
  summarise(
    p_kw  = kruskal.test(log2_length ~ quartile_label)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = if_else(
      p_kw < 0.001,
      "KW p < 0.001",
      paste0("KW p = ", signif(p_kw, 2))
    )
  )

p_len <- ggplot(len_long, aes(x = quartile_label, y = log2_length, fill = quartile_label)) +
  geom_violin(alpha = 0.5, linewidth = 0.4, trim = TRUE) +
  geom_boxplot(width = 0.18, outlier.size = 0.3, linewidth = 0.4,
               fill = "white", alpha = 0.85) +
  geom_text(
    data        = kw_len,
    aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05, vjust = 1.4, size = 3, colour = "grey30"
  ) +
  scale_fill_manual(values = quartile_colours, guide = "none") +
  facet_wrap(~ region, scales = "free_y") +
  labs(
    title    = "Region lengths by TE quartile (MCF7-SIX1, siCTRL hypoxia vs normoxia, 1hr)",
    subtitle = "Q1 = most repressed in hypoxia; Q4 = most induced. Lengths log₂-transformed. KW = Kruskal-Wallis.",
    x        = "TE quartile",
    y        = "log₂(length, nt)"
  ) +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
p_len
ggsave(here("plots", "49_region_length_by_te_quartile.pdf"),
       p_len, width = 8, height = 4.5)
```

---

# Section 6: GC3 (Third-Codon-Position GC) by Quartile

```{r plot_gc3, fig.width=4.5, fig.height=4.5}
stopifnot(
  "gc3 out of range" = all(plot_df$gc3 >= 0 & plot_df$gc3 <= 100, na.rm = TRUE)
)

kw_gc3 <- kruskal.test(gc3 ~ quartile_label, data = plot_df)
kw_gc3_label <- if_else(
  kw_gc3$p.value < 0.001,
  "KW p < 0.001",
  paste0("KW p = ", signif(kw_gc3$p.value, 2))
)

p_gc3 <- ggplot(plot_df, aes(x = quartile_label, y = gc3, fill = quartile_label)) +
  geom_violin(alpha = 0.5, linewidth = 0.4, trim = TRUE) +
  geom_boxplot(width = 0.18, outlier.size = 0.3, linewidth = 0.4,
               fill = "white", alpha = 0.85) +
  annotate("text", x = Inf, y = Inf, label = kw_gc3_label,
           hjust = 1.05, vjust = 1.4, size = 3, colour = "grey30") +
  scale_fill_manual(values = quartile_colours, guide = "none") +
  labs(
    title    = "CDS third-codon-position GC (GC3) by TE quartile (MCF7-SIX1)",
    subtitle = "siCTRL hypoxia vs normoxia, 1hr. Q1 = most repressed; Q4 = most induced.\nKW = Kruskal-Wallis test across all 4 quartiles.",
    x        = "TE quartile",
    y        = "GC3 (%)"
  ) +
  theme_classic(base_size = 12)
p_gc3
ggsave(here("plots", "49_gc3_by_te_quartile.pdf"), p_gc3, width = 4.5, height = 4.5)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
