---
title: "50 — Cross-Cell-Line bg-RSCU Delta Correlation (MDA-MB-231 vs MCF7-SIX1)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Compares the hypoxia-vs-normoxia shift in background-corrected RSCU (bg-RSCU delta)
between MDA-MB-231 (nb45) and MCF7-SIX1 (nb47) cell lines. For each cell line the
delta is: mean bg-RSCU(eIF3d-promoted genes) − mean bg-RSCU(negative controls), computed
separately in hypoxia and normoxia. The delta-delta is Δhyp − Δnor.

Positive controls are run before joining to verify that input files are correctly
oriented before interpreting any cross-cell-line correlation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(ggrepel)
library(here)
```

---

# Section 1: Load Data & Positive Control Checks

```{r load_data}
path_231  <- here("output", "predictive_modeling", "bg_rscu_delta_delta_hyp_vs_nor.csv")
path_mcf7 <- here("output", "predictive_modeling", "bg_rscu_delta_delta_mcf7six1_hyp_vs_nor.csv")

stopifnot(
  "bg_rscu_delta_delta_hyp_vs_nor.csv not found — run nb45 first" =
    file.exists(path_231),
  "bg_rscu_delta_delta_mcf7six1_hyp_vs_nor.csv not found — run nb47 first" =
    file.exists(path_mcf7)
)

d231  <- read_csv(path_231,  show_col_types = FALSE)
dmcf7 <- read_csv(path_mcf7, show_col_types = FALSE)

stopifnot(
  "MDA-MB-231 file missing expected columns" =
    all(c("codon", "amino_acid", "gc3_class", "delta_hyp", "delta_nor",
          "delta_delta", "sig") %in% colnames(d231)),
  "MCF7-SIX1 file missing expected columns" =
    all(c("codon", "amino_acid", "gc3_class", "delta_hyp", "delta_nor",
          "delta_delta", "sig") %in% colnames(dmcf7))
)

cat("MDA-MB-231 rows:", nrow(d231), "| MCF7-SIX1 rows:", nrow(dmcf7), "\n")
```

```{r positive_controls}
# ── MDA-MB-231 directionality anchors ──────────────────────────────────────────
# GCG (Ala, GC-ending): extremely rare in human coding sequences due to CpG
# suppression. eIF3d-promoted genes should strongly AVOID this codon relative
# to negative controls → delta_hyp must be strongly negative.
gcg_231 <- d231 %>% filter(codon == "GCG") %>% pull(delta_hyp)
stopifnot(
  "MDA-MB-231 GCG delta_hyp should be < -0.2 (strongly negative)" = gcg_231 < -0.2
)
cat("MDA-MB-231 GCG delta_hyp:", round(gcg_231, 4), " [expect < -0.2] ✓\n")

# GCT (Ala, AT-ending): common, optimal Ala codon. eIF3d-promoted genes should
# prefer this codon → delta_hyp must be positive.
gct_231 <- d231 %>% filter(codon == "GCT") %>% pull(delta_hyp)
stopifnot(
  "MDA-MB-231 GCT delta_hyp should be > 0 (positive)" = gct_231 > 0
)
cat("MDA-MB-231 GCT delta_hyp:", round(gct_231, 4), " [expect > 0] ✓\n")

# ── MCF7-SIX1 report ──────────────────────────────────────────────────────────
# Pipeline verified (nb41 line 91, nb42 lines 65+72): both MCF7-SIX1 hypoxia
# and normoxia gene sets apply te_lfc = -log2FoldChange before filtering, so
# positive te_lfc = eIF3d promotes (matching MDA-MB-231 convention).
# The opposite GCG/GCT signs relative to MDA-MB-231 are genuine biology:
# the specific genes eIF3d promotes in MCF7-SIX1 have different intrinsic
# codon usage than the MDA-MB-231 promotes set.
gcg_mcf7 <- dmcf7 %>% filter(codon == "GCG") %>% pull(delta_hyp)
gct_mcf7 <- dmcf7 %>% filter(codon == "GCT") %>% pull(delta_hyp)

cat("\nMCF7-SIX1 GCG delta_hyp:", round(gcg_mcf7, 4),
    "| GCT delta_hyp:", round(gct_mcf7, 4), "\n")
cat("(Pipeline sign convention verified correct — see nb41:91, nb42:65+72)\n")
```

```{r join_data}
df <- d231 %>%
  dplyr::select(
    codon, amino_acid, gc3_class,
    delta_hyp_231   = delta_hyp,
    delta_nor_231   = delta_nor,
    delta_delta_231 = delta_delta,
    sig_231         = sig
  ) %>%
  inner_join(
    dmcf7 %>% dplyr::select(
      codon,
      delta_hyp_mcf7   = delta_hyp,
      delta_nor_mcf7   = delta_nor,
      delta_delta_mcf7 = delta_delta,
      sig_mcf7         = sig
    ),
    by = "codon"
  )

stopifnot(
  "Join should retain all 59 multi-synonymous codons" = nrow(df) == 59
)
cat("Codons after join:", nrow(df), "\n")
cat("gc3_class breakdown:\n")
print(count(df, gc3_class))
```

---

# Section 2: Scatter Plots

```{r scatter_helpers}
gc3_colors <- c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2")

make_scatter <- function(data, x_col, y_col, x_lab, y_lab, label_n = 15, xlim = NULL, ylim = NULL) {
  x <- data[[x_col]]
  y <- data[[y_col]]
  r_test <- cor.test(x, y)
  r_label <- sprintf("r = %.2f\np = %.3g", r_test$estimate, r_test$p.value)

  label_idx <- order(pmax(abs(x), abs(y)), decreasing = TRUE)[seq_len(label_n)]
  plot_df <- data %>%
    mutate(
      .x    = x,
      .y    = y,
      label = ifelse(row_number() %in% label_idx, codon, NA_character_)
    )

  ggplot(plot_df, aes(.x, .y, colour = gc3_class, label = label)) +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.4) +
    geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 2.5, alpha = 0.75) +
    geom_text_repel(size = 3, max.overlaps = 20, na.rm = TRUE,
                    segment.colour = "grey60", segment.size = 0.3) +
    annotate("text", x = Inf, y = -Inf, label = r_label,
             hjust = 1.1, vjust = -0.4, size = 3.5, colour = "grey30") +
    scale_colour_manual(values = gc3_colors, name = "GC3 class") +
    (if (!is.null(xlim)) scale_x_continuous(limits = xlim) else NULL) +
    (if (!is.null(ylim)) scale_y_continuous(limits = ylim) else NULL) +
    coord_fixed() +
     labs(x = x_lab, y = y_lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## 2a — Delta-Delta (Δhyp − Δnor)

```{r plot_delta_delta, fig.width=6, fig.height=5.5}
p_dd <- make_scatter(
  df,
  x_col = "delta_delta_231",
  y_col = "delta_delta_mcf7",
  x_lab = "bg-RSCU delta (hyp - nor), MDA-MB-231",
  y_lab = "bg-RSCU delta (hyp - nor), MCF7-SIX1"
)
p_dd
ggsave(here("plots", "50_scatter_delta_delta_231_vs_mcf7six1.pdf"),
       p_dd, width = 6, height = 5.5)
```

## 2b — Delta Hypoxia (raw hypoxia preference)

```{r plot_delta_hyp, fig.width=6, fig.height=5.5}
p_hyp <- make_scatter(
  df,
  x_col = "delta_hyp_231",
  y_col = "delta_hyp_mcf7",
  x_lab = "bg-RSCU delta, hypoxia (MDA-MB-231)",
  y_lab = "bg-RSCU delta, hypoxia (MCF7-SIX1)",
  xlim  = c(-0.4, 0.4),
  ylim  = c(-0.4, 0.4)
)
p_hyp
ggsave(here("plots", "50_scatter_delta_hyp_231_vs_mcf7six1.pdf"),
       p_hyp, width = 6, height = 5.5)
```

## 2c — Delta Normoxia (baseline preference)

```{r plot_delta_nor, fig.width=6, fig.height=5.5}
p_nor <- make_scatter(
  df,
  x_col = "delta_nor_231",
  y_col = "delta_nor_mcf7",
  x_lab = "bg-RSCU delta, normoxia (MDA-MB-231)",
  y_lab = "bg-RSCU delta, normoxia (MCF7-SIX1)",
  xlim  = c(-0.4, 0.4),
  ylim  = c(-0.4, 0.4)
)
p_nor
ggsave(here("plots", "50_scatter_delta_nor_231_vs_mcf7six1.pdf"),
       p_nor, width = 6, height = 10.5)
```

---

# Section 3: Summary Statistics

```{r summary_stats}
cat("=== Cross-cell-line bg-RSCU correlations (Pearson) ===\n")
comparisons <- list(
  list(x = df$delta_delta_231, y = df$delta_delta_mcf7, label = "delta_delta (hyp - nor)"),
  list(x = df$delta_hyp_231,   y = df$delta_hyp_mcf7,   label = "delta_hyp (hypoxia only)"),
  list(x = df$delta_nor_231,   y = df$delta_nor_mcf7,   label = "delta_nor (normoxia only)")
)

for (cmp in comparisons) {
  r <- cor.test(cmp$x, cmp$y)
  cat(sprintf("%-32s r = %+.3f  p = %.3g\n",
              cmp$label, r$estimate, r$p.value))
}

cat("\n=== Sign concordance (delta_hyp) by GC3 class ===\n")
df %>%
  mutate(concordant = sign(delta_hyp_231) == sign(delta_hyp_mcf7)) %>%
  count(gc3_class, concordant) %>%
  print()
```
