---
title: "52 — bg-RSCU: Hypoxia-Promoted vs Normoxia-Promoted (MCF7-SIX1, No Negative Control)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Robustness analysis complementing notebooks 41/46/47. Instead of comparing
promoted genes to a negative control (Δ = promoted − neg ctrl), this notebook
compares the bg-RSCU of the hypoxia 1hr promoted set directly to the normoxia
1hr promoted set (Δ = hyp_promoted − nor_promoted). A codon enriched here
is more characteristic of eIF3d-promoted genes in hypoxia than of eIF3d-promoted
genes in normoxia, without the negative control serving as reference.

LFC cutoff: > 0 (any positive TE), padj < 0.05 — matching nb41/46 convention.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(ggrepel)
library(ggVennDiagram)
```

---

# Section 1: Setup — CDS Sequences & Codon Counts

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_mcf7six1_external_stability.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0 & width(cds_frame0) > 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)
cat("Codon counts matrix:", nrow(counts_mat0), "transcripts ×", ncol(counts_mat0), "codons\n")
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

codon_gc3 <- tibble(
  codon          = sense_codons,
  gc3_nucleotide = substr(sense_codons, 3, 3),
  gc3_class      = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"), "GC-ending", "AT-ending")
)

tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*"))

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

```{r cutoffs}
lfc_cutoff  <- 0
padj_cutoff <- 0.05
```

# Section 2: Gene Set Loading

```{r gene_sets}
# Hypoxia 1hr promoted — lfc > 0, padj < 0.05 (matching nb41 convention)
hyp_raw      <- read_csv(
  here("output", "genesets", "mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc0.csv"),
  show_col_types = FALSE
)
hyp_gene_ids <- hyp_raw$ensembl_gene
cat("Hypoxia 1hr promoted genes:", length(hyp_gene_ids), "\n")

# Normoxia 1hr promoted — lfc > 0, padj < 0.05 (matching nb46 convention)
nor_raw      <- read_csv(
  here("output", "genesets", "mcf7six1_normoxia_3d_promotes_TE_1hr_lfc0.csv"),
  show_col_types = FALSE
)
nor_gene_ids <- nor_raw$ensembl_gene
cat("Normoxia 1hr promoted genes:", length(nor_gene_ids), "\n")

stopifnot(
  "No hypoxia promoted genes found"  = length(hyp_gene_ids) > 0,
  "No normoxia promoted genes found" = length(nor_gene_ids) > 0
)

overlap_ids <- intersect(hyp_gene_ids, nor_gene_ids)
cat("Overlap (genes promoted in both conditions):", length(overlap_ids), "\n")
cat("Hypoxia-only:", length(setdiff(hyp_gene_ids, nor_gene_ids)), "\n")
cat("Normoxia-only:", length(setdiff(nor_gene_ids, hyp_gene_ids)), "\n")
```

---

# Section 3: Venn Diagram — Gene Set Overlap

```{r venn_diagram, fig.width=5, fig.height=4.5}
venn_sets <- list(
  "Hypoxia\npromoted"  = hyp_gene_ids,
  "Normoxia\npromoted" = nor_gene_ids
)

p_venn <- ggVennDiagram(venn_sets, label_alpha = 0,
                        set_color = c("#D55E00", "#009E73")) +
  scale_fill_gradient(low = "white", high = "white") +
  labs(title = "eIF3d-promoted gene set overlap\nHypoxia 1hr vs Normoxia 1hr (MCF7-SIX1)",
       subtitle = paste0("lfc > ", lfc_cutoff, ", padj < ", padj_cutoff)) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, size = 11))

print(p_venn)
ggsave(here("plots", "52_venn_hyp_vs_nor_promoted.pdf"), p_venn, width = 8, height = 4.5)
```

---

# Section 4: Background-Corrected RSCU

```{r bg_rscu_background_props}
# Background = mean per-transcript codon proportion across all expressed mRNAs.
# Each gene gets equal weight regardless of CDS length.
bg_props <- imap(aa_to_codons, function(codons, aa) {
  codons  <- intersect(codons, colnames(counts_mat0))
  if (length(codons) == 0) return(NULL)
  sub_cnt <- counts_mat0[, codons, drop = FALSE]
  aa_tot  <- rowSums(sub_cnt)
  valid   <- aa_tot > 0
  if (!any(valid)) return(NULL)
  prop_mat <- sub_cnt[valid, , drop = FALSE] / aa_tot[valid]
  colMeans(prop_mat)
})
bg_props <- Filter(Negate(is.null), bg_props)

cat("Background proportions computed for", length(bg_props), "amino acids\n")
prop_sums <- sapply(bg_props, sum)
cat("Background proportions sum check (should all be ~1):",
    round(range(prop_sums), 4), "\n")
```

```{r compute_bg_rscu_function}
compute_bg_rscu_matrix <- function(counts_mat, bg_props) {
  rscu_mat <- matrix(NA_real_, nrow = nrow(counts_mat), ncol = ncol(counts_mat),
                     dimnames = dimnames(counts_mat))
  for (aa in names(bg_props)) {
    props  <- bg_props[[aa]]
    codons <- intersect(names(props), colnames(counts_mat))
    if (length(codons) == 0) next
    sub_cnt <- counts_mat[, codons, drop = FALSE]
    aa_tot  <- rowSums(sub_cnt)
    valid   <- aa_tot > 0
    for (codon in codons) {
      bp <- props[codon]
      if (bp > 0 && any(valid)) {
        rscu_mat[valid, codon] <- (sub_cnt[valid, codon] / aa_tot[valid]) / bp
      }
    }
  }
  rscu_mat
}
```

```{r compute_bg_rscu}
bg_rscu_mat0 <- compute_bg_rscu_matrix(counts_mat0, bg_props)
cat("Background RSCU matrix:", nrow(bg_rscu_mat0), "×", ncol(bg_rscu_mat0), "\n")
cat("Range:", round(min(bg_rscu_mat0, na.rm = TRUE), 3), "—",
    round(max(bg_rscu_mat0, na.rm = TRUE), 3), "\n")

stopifnot(
  "bg-RSCU contains negative values" = all(bg_rscu_mat0 >= 0, na.rm = TRUE)
)
```

---

# Section 5: Group Transcript Assignment

```{r assign_groups}
hyp_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% hyp_gene_ids) %>%
  pull(transcript_id_clean)
nor_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% nor_gene_ids) %>%
  pull(transcript_id_clean)

hyp_tx <- hyp_tx[hyp_tx %in% rownames(bg_rscu_mat0)]
nor_tx <- nor_tx[nor_tx %in% rownames(bg_rscu_mat0)]

cat("Hypoxia promoted transcripts:", length(hyp_tx), "\n")
cat("Normoxia promoted transcripts:", length(nor_tx), "\n")

stopifnot(
  "No hypoxia transcripts in matrix"  = length(hyp_tx) > 0,
  "No normoxia transcripts in matrix" = length(nor_tx) > 0
)
```

---

# Section 6: Group Summaries & Statistical Testing

```{r summarize_function}
summarize_rscu <- function(rscu_mat, tx_ids) {
  tx_ids <- tx_ids[tx_ids %in% rownames(rscu_mat)]
  sub    <- rscu_mat[tx_ids, sense_codons, drop = FALSE]
  imap_dfr(aa_to_codons, function(codons, aa) {
    codons <- intersect(codons, colnames(sub))
    map_dfr(codons, function(codon) {
      vals <- sub[, codon]
      vals <- vals[!is.na(vals)]
      tibble(
        amino_acid  = aa,
        codon       = codon,
        mean_rscu   = mean(vals),
        se_rscu     = if (length(vals) > 1) sd(vals) / sqrt(length(vals)) else NA_real_,
        n           = length(vals)
      )
    })
  })
}

bg_rscu_hyp <- summarize_rscu(bg_rscu_mat0, hyp_tx)
bg_rscu_nor <- summarize_rscu(bg_rscu_mat0, nor_tx)
cat("bg-RSCU summary — hypoxia promoted:", nrow(bg_rscu_hyp), "codon-rows\n")
```

```{r biological_spotcheck}
# Biological spot-check: CTG (optimal Leu codon) should have bg-RSCU near 1.0
ctg_hyp <- bg_rscu_hyp %>% filter(codon == "CTG") %>% pull(mean_rscu)
ctg_nor <- bg_rscu_nor %>% filter(codon == "CTG") %>% pull(mean_rscu)
cat("CTG (Leu, optimal) mean bg-RSCU — hypoxia:", round(ctg_hyp, 3),
    "| normoxia:", round(ctg_nor, 3), "\n")
cat("(Both should be near 1.0 for an optimally-used reference codon)\n")
```

```{r delta_and_wilcox}
bg_rscu_combined <- bg_rscu_hyp %>%
  dplyr::select(amino_acid, codon, hyp_mean = mean_rscu,
                hyp_se = se_rscu, hyp_n = n) %>%
  left_join(
    bg_rscu_nor %>% dplyr::select(amino_acid, codon, nor_mean = mean_rscu,
                                  nor_se = se_rscu, nor_n = n),
    by = c("amino_acid", "codon")
  ) %>%
  mutate(delta_bg_rscu = hyp_mean - nor_mean)

# Wilcoxon test per codon, multi-synonymous AAs only
wilcox_df <- imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
  codons <- intersect(codons, colnames(bg_rscu_mat0))
  map_dfr(codons, function(codon) {
    h_vals <- bg_rscu_mat0[hyp_tx[hyp_tx %in% rownames(bg_rscu_mat0)], codon]
    n_vals <- bg_rscu_mat0[nor_tx[nor_tx %in% rownames(bg_rscu_mat0)], codon]
    h_vals <- h_vals[!is.na(h_vals)]
    n_vals <- n_vals[!is.na(n_vals)]
    if (length(h_vals) < 3 || length(n_vals) < 3) {
      return(tibble(amino_acid = aa, codon = codon, p_val = NA_real_))
    }
    wt <- wilcox.test(h_vals, n_vals, exact = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val = wt$p.value)
  })
}) %>%
  mutate(p_adj = p.adjust(p_val, method = "BH"))

bg_rscu_combined <- bg_rscu_combined %>%
  left_join(wilcox_df %>% dplyr::select(amino_acid, codon, p_val, p_adj),
            by = c("amino_acid", "codon"))

stopifnot(
  "groups missing from bg_rscu_combined" =
    all(c("hyp_mean", "nor_mean") %in% colnames(bg_rscu_combined))
)

cat("Significant codons (BH q < 0.05):",
    sum(!is.na(bg_rscu_combined$p_adj) & bg_rscu_combined$p_adj < 0.05,
        na.rm = TRUE), "\n")

cat("\nTop 10 codons by |Δ bg-RSCU|:\n")
print(bg_rscu_combined %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_bg_rscu))) %>%
        dplyr::select(codon, amino_acid, hyp_mean, nor_mean, delta_bg_rscu, p_adj) %>%
        head(10))
```

---

# Section 7: Visualization

```{r plot_sidebyside, fig.width=11, fig.height=9}
plot_df_a <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(cols = c(hyp_mean, nor_mean),
               names_to = "group", values_to = "mean_bg_rscu") %>%
  mutate(
    se      = if_else(group == "hyp_mean", hyp_se, nor_se),
    group   = recode(group,
                     hyp_mean = "Hypoxia promoted",
                     nor_mean = "Normoxia promoted"),
    group   = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, mean_bg_rscu)
  )

p_a <- ggplot(plot_df_a, aes(x = codon, y = mean_bg_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = mean_bg_rscu - se, ymax = mean_bg_rscu + se),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("Hypoxia promoted" = "#D55E00",
                                "Normoxia promoted" = "#009E73"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "RSCU: hypoxia vs normoxia eIF3d-promoted (MCF7-SIX1)",
    subtitle = "Mean per-transcript RSCU ± SE; dashed = 1 (same rate as transcriptome background)",
    x = "Codon", y = "Mean RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_a
ggsave(here("plots", "52_bg_rscu_sidebyside_hyp_vs_nor_promoted.pdf"),
       p_a, width = 11, height = 9)
```

```{r plot_delta, fig.width=10, fig.height=8}
plot_df_b <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = ifelse(delta_bg_rscu >= 0, "hypoxia", "normoxia"),
    sig_label = case_when(!is.na(p_adj) & p_adj < 0.05 ~ "*", TRUE ~ ""),
    codon     = reorder(codon, delta_bg_rscu)
  )

p_b <- ggplot(plot_df_b, aes(x = codon, y = delta_bg_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label, y = delta_bg_rscu + sign(delta_bg_rscu) * 0.015),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(hypoxia = "#D55E00", normoxia = "#009E73"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "Δ RSCU: hypoxia-promoted minus normoxia-promoted (MCF7-SIX1)",
    subtitle = "Positive = codon more characteristic of hypoxia-promoted genes;\n* = BH-adjusted Wilcoxon q < 0.05",
    x = "Codon", y = "Δ mean RSCU (hypoxia − normoxia)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_b
ggsave(here("plots", "52_bg_rscu_delta_hyp_vs_nor_promoted.pdf"),
       p_b, width = 10, height = 8)
```

```{r plot_violin, fig.width=12, fig.height=8}
sig_codons <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa, !is.na(p_adj), p_adj < 0.05) %>%
  dplyr::select(amino_acid, codon)

cat("Codons with BH q < 0.05:", nrow(sig_codons), "\n")
if (nrow(sig_codons) > 0) cat(paste(sig_codons$codon, collapse = ", "), "\n")

if (nrow(sig_codons) > 0) {
  violin_df <- map_dfr(seq_len(nrow(sig_codons)), function(i) {
    aa    <- sig_codons$amino_acid[i]
    codon <- sig_codons$codon[i]
    h_vals <- bg_rscu_mat0[hyp_tx[hyp_tx %in% rownames(bg_rscu_mat0)], codon]
    n_vals <- bg_rscu_mat0[nor_tx[nor_tx %in% rownames(bg_rscu_mat0)], codon]
    bind_rows(
      tibble(codon = codon, amino_acid = aa, group = "Hypoxia promoted",
             bg_rscu = h_vals[!is.na(h_vals)]),
      tibble(codon = codon, amino_acid = aa, group = "Normoxia promoted",
             bg_rscu = n_vals[!is.na(n_vals)])
    )
  }) %>%
    mutate(
      group = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted"))
    ) %>%
    left_join(bg_rscu_combined %>% dplyr::select(codon, p_adj),
              by = "codon") %>%
    mutate(codon_label = paste0(codon, "\nq=", signif(p_adj, 2)))

  p_viol <- ggplot(violin_df, aes(x = group, y = bg_rscu, fill = group)) +
    geom_violin(alpha = 0.6, linewidth = 0.4, trim = TRUE) +
    geom_boxplot(width = 0.18, outlier.size = 0.5, linewidth = 0.4,
                 fill = "white", alpha = 0.8) +
    geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
    scale_fill_manual(values = c("Hypoxia promoted"  = "#D55E00",
                                  "Normoxia promoted" = "#009E73"),
                      guide = "none") +
    facet_wrap(~ codon_label, scales = "free_y", nrow = 2) +
    labs(
      title    = "Per-transcript RSCU for significant codons (MCF7-SIX1)",
      subtitle = "Codons with BH q < 0.05 between hypoxia-promoted and normoxia-promoted.\nDashed = 1 (same usage rate as transcriptome background).",
      x = NULL, y = "Per-transcript RSCU"
    ) +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1),
      strip.text  = element_text(face = "bold", size = 9)
    )
  p_viol
  ggsave(here("plots", "52_bg_rscu_violin_hyp_vs_nor_promoted.pdf"), p_viol,
         width = max(8, nrow(sig_codons) * 2.5), height = 8, limitsize = FALSE)
} else {
  cat("No significant codons — skipping violin plot.\n")
}
```

```{r plot_scatter, fig.width=6, fig.height=5.5}
comp_scatter <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(
    sig         = if_else(!is.na(p_adj) & p_adj < 0.05, "Significant", "Neither"),
    label_codon = if_else(sig == "Significant", codon, NA_character_)
  )

r_val    <- cor(comp_scatter$hyp_mean, comp_scatter$nor_mean, use = "complete.obs")
axis_lim <- range(c(comp_scatter$hyp_mean, comp_scatter$nor_mean), na.rm = TRUE)
axis_lim <- axis_lim + diff(axis_lim) * c(-0.05, 0.05)

p_scatter <- ggplot(comp_scatter, aes(x = nor_mean, y = hyp_mean)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linewidth = 0.5,
              linetype = "dashed") +
  geom_hline(yintercept = 1, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 1, colour = "grey85", linewidth = 0.3) +
  geom_point(aes(shape = sig, colour = sig,
                 size  = sig, alpha = sig)) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 25, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = c(Significant = "#0072B2", Neither = "grey70"),
                      name = "Significant (q < 0.05)") +
  scale_shape_manual(values  = c(Significant = 17, Neither = 16),
                     name = "Significant (q < 0.05)") +
  scale_size_manual(values   = c(Significant = 2.5, Neither = 1.6),
                    guide = "none") +
  scale_alpha_manual(values  = c(Significant = 1, Neither = 0.5),
                     guide = "none") +
  annotate("text",
           x = axis_lim[1] + diff(axis_lim) * 0.02,
           y = axis_lim[2] - diff(axis_lim) * 0.02,
           label = paste0("r = ", round(r_val, 2)),
           hjust = 0, vjust = 1, size = 4, colour = "grey30") +
  coord_fixed(xlim = axis_lim, ylim = axis_lim) +
  labs(
    title    = "RSCU: eIF3d codon usage in hypoxia vs normoxia promoted genes (MCF7-SIX1)",
    subtitle = paste0("Each point = one codon (multi-synonymous AAs). Dashed = identity line.\n",
                      "Points above diagonal: higher usage in hypoxia-promoted genes."),
    x = "Mean RSCU (normoxia-promoted)",
    y = "Mean RSCU (hypoxia-promoted)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_scatter
ggsave(here("plots", "52_bg_rscu_scatter_hyp_vs_nor.pdf"), p_scatter, width = 6, height = 5.5)
```

```{r plot_lollipop, fig.width=5.5, fig.height=9}
plot_lollipop <- bg_rscu_combined %>%
  filter(amino_acid %in% multi_aa) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(
    sig = case_when(
      !is.na(p_adj) & p_adj < 0.05 & delta_bg_rscu >= 0 ~ "Hypoxia only",
      !is.na(p_adj) & p_adj < 0.05 & delta_bg_rscu <  0 ~ "Normoxia only",
      TRUE ~ "Neither"
    ),
    sig   = factor(sig, levels = c("Hypoxia only", "Normoxia only", "Neither")),
    codon = fct_reorder(codon, delta_bg_rscu)
  )

p_lollipop <- ggplot(plot_lollipop,
                     aes(x = delta_bg_rscu, y = codon, colour = gc3_class)) +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_segment(aes(x = 0, xend = delta_bg_rscu, yend = codon),
               linewidth = 0.5, alpha = 0.7) +
  geom_point(aes(shape = sig), size = 2) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  scale_shape_manual(
    values = c("Hypoxia only" = 17, "Normoxia only" = 15, "Neither" = 16),
    name   = "Significant (q < 0.05)"
  ) +
  labs(
    title    = "Ranked Δ RSCU: hypoxia vs normoxia eIF3d-promoted (MCF7-SIX1)",
    subtitle = "Positive = stronger codon usage in hypoxia-promoted genes",
    x = "Δ RSCU (hypoxia − normoxia)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y     = element_text(size = 7, family = "mono"),
    legend.position = "top",
    legend.box      = "vertical"
  )
p_lollipop
ggsave(here("plots", "52_bg_rscu_lollipop_ranked.pdf"), p_lollipop, width = 5.5, height = 9)
```

---

# Section 8: Save Results

```{r save_csv}
out_df <- bg_rscu_combined %>%
  dplyr::select(amino_acid, codon,
                hyp_promoted_mean = hyp_mean,
                hyp_promoted_se   = hyp_se,
                hyp_promoted_n    = hyp_n,
                nor_promoted_mean = nor_mean,
                nor_promoted_se   = nor_se,
                nor_promoted_n    = nor_n,
                delta_bg_rscu, p_val, p_adj)

write_csv(out_df,
          here("output", "predictive_modeling", "bg_rscu_mcf7six1_promoted_hyp_vs_nor.csv"))
cat("Saved to output/predictive_modeling/bg_rscu_mcf7six1_promoted_hyp_vs_nor.csv\n")
cat("Rows:", nrow(out_df), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
