---
title: "53 — Cross-Cell-Line RSCU: Promoted Genes (MDA-MB-231 vs MCF7-SIX1)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Compares background-corrected RSCU of eIF3d-promoted genes between MDA-MB-231 (nb51)
and MCF7-SIX1 (nb52). Three scatter plots: mean RSCU in normoxia, mean RSCU in hypoxia,
and the delta (hypoxia − normoxia promoted mean). GC3 class (AT- vs GC-ending codon)
is derived from the third nucleotide of each codon.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(ggrepel)
library(here)
```

---

# Section 1: Load Data

```{r load_data}
path_231  <- here("output", "predictive_modeling", "bg_rscu_231_promoted_hyp_vs_nor.csv")
path_mcf7 <- here("output", "predictive_modeling", "bg_rscu_mcf7six1_promoted_hyp_vs_nor.csv")

stopifnot(
  "bg_rscu_231_promoted_hyp_vs_nor.csv not found — run nb51 first" =
    file.exists(path_231),
  "bg_rscu_mcf7six1_promoted_hyp_vs_nor.csv not found — run nb52 first" =
    file.exists(path_mcf7)
)

d231  <- read_csv(path_231,  show_col_types = FALSE)
dmcf7 <- read_csv(path_mcf7, show_col_types = FALSE)

expected_cols <- c("amino_acid", "codon",
                   "hyp_promoted_mean", "nor_promoted_mean", "delta_bg_rscu")
stopifnot(
  "MDA-MB-231 file missing expected columns" =
    all(expected_cols %in% colnames(d231)),
  "MCF7-SIX1 file missing expected columns" =
    all(expected_cols %in% colnames(dmcf7))
)

cat("MDA-MB-231 rows:", nrow(d231), "| MCF7-SIX1 rows:", nrow(dmcf7), "\n")
```

```{r build_join}
# Derive GC3 class from third nucleotide; filter to multi-synonymous codons
add_gc3 <- function(df) {
  df %>%
    mutate(gc3_class = if_else(
      substr(codon, 3, 3) %in% c("A", "T"), "AT-ending", "GC-ending"
    ))
}

multi_syn_filter <- function(df) {
  df %>% filter(!amino_acid %in% c("*", "M", "W"))
}

d231  <- d231  %>% add_gc3() %>% multi_syn_filter()
dmcf7 <- dmcf7 %>% add_gc3() %>% multi_syn_filter()

df <- d231 %>%
  dplyr::select(
    codon, amino_acid, gc3_class,
    hyp_231   = hyp_promoted_mean,
    nor_231   = nor_promoted_mean,
    delta_231 = delta_bg_rscu
  ) %>%
  inner_join(
    dmcf7 %>% dplyr::select(
      codon,
      hyp_mcf7   = hyp_promoted_mean,
      nor_mcf7   = nor_promoted_mean,
      delta_mcf7 = delta_bg_rscu
    ),
    by = "codon"
  )

stopifnot(
  "Unexpected codon count after join — check input files" =
    between(nrow(df), 55, 63),
  "GC3 class missing a category" =
    all(c("AT-ending", "GC-ending") %in% df$gc3_class)
)

cat("Codons after join:", nrow(df), "\n")
cat("GC3 class breakdown:\n")
print(count(df, gc3_class))
```

---

# Section 2: Scatter Plots

```{r scatter_helpers}
gc3_colors <- c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2")

make_scatter <- function(data, x_col, y_col, x_lab, y_lab,
                         label_n = 10, xlim = NULL, ylim = NULL,
                         origin_lines = TRUE) {
  x <- data[[x_col]]
  y <- data[[y_col]]
  r_test  <- cor.test(x, y)
  r_label <- sprintf("r = %.2f\np = %.3g", r_test$estimate, r_test$p.value)

  label_idx <- order(pmax(abs(x), abs(y)), decreasing = TRUE)[seq_len(label_n)]
  plot_df <- data %>%
    mutate(
      .x    = x,
      .y    = y,
      label = if_else(row_number() %in% label_idx, codon, NA_character_)
    )

  ggplot(plot_df, aes(.x, .y, colour = gc3_class, label = label)) +
    (if (origin_lines) geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.4) else NULL) +
    (if (origin_lines) geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.4) else NULL) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 2.5, alpha = 0.75) +
    geom_text_repel(size = 3, max.overlaps = 20, na.rm = TRUE,
                    segment.colour = "grey60", segment.size = 0.3) +
    annotate("text", x = Inf, y = -Inf, label = r_label,
             hjust = 1.1, vjust = -0.4, size = 3.5, colour = "grey30") +
    scale_colour_manual(values = gc3_colors, name = "GC3 class") +
    (if (!is.null(xlim)) scale_x_continuous(limits = xlim) else NULL) +
    (if (!is.null(ylim)) scale_y_continuous(limits = ylim) else NULL) +
    coord_fixed() +
    labs(x = x_lab, y = y_lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## 2a — Normoxia (mean RSCU of normoxia-promoted genes)

```{r plot_normoxia, fig.width=6, fig.height=5.5}
p_nor <- make_scatter(
  df,
  x_col = "nor_231",
  y_col = "nor_mcf7",
  x_lab = "RSCU, normoxia (MDA-MB-231)",
  y_lab = "RSCU, normoxia (MCF7–SIX1)",
  origin_lines = FALSE
)
p_nor
ggsave(here("plots", "53_rscu_normoxia_231_vs_mcf7six1.pdf"),
       p_nor, width = 6, height = 5.5)
```

## 2b — Hypoxia (mean RSCU of hypoxia-promoted genes)

```{r plot_hypoxia, fig.width=6, fig.height=5.5}
p_hyp <- make_scatter(
  df,
  x_col = "hyp_231",
  y_col = "hyp_mcf7",
  x_lab = "RSCU, hypoxia (MDA-MB-231)",
  y_lab = "RSCU, hypoxia (MCF7–SIX1)",
  origin_lines = FALSE
)
p_hyp
ggsave(here("plots", "53_rscu_hypoxia_231_vs_mcf7six1.pdf"),
       p_hyp, width = 6, height = 5.5)
```

## 2c — Delta (hypoxia − normoxia promoted mean)

```{r plot_delta, fig.width=6, fig.height=5.5}
p_delta <- make_scatter(
  df,
  x_col = "delta_231",
  y_col = "delta_mcf7",
  x_lab = "RSCU delta (hyp − nor), MDA-MB-231",
  y_lab = "RSCU delta (hyp − nor), MCF7–SIX1"
)
p_delta
ggsave(here("plots", "53_rscu_delta_231_vs_mcf7six1.pdf"),
       p_delta, width = 6, height = 5.5)
```

---

# Section 3: Summary Statistics

```{r summary_stats}
cat("=== Cross-cell-line RSCU correlations (Pearson) ===\n")
comparisons <- list(
  list(x = df$nor_231,   y = df$nor_mcf7,   label = "normoxia mean RSCU"),
  list(x = df$hyp_231,   y = df$hyp_mcf7,   label = "hypoxia mean RSCU"),
  list(x = df$delta_231, y = df$delta_mcf7, label = "delta (hyp - nor)")
)

for (cmp in comparisons) {
  r <- cor.test(cmp$x, cmp$y)
  cat(sprintf("%-28s r = %+.3f  p = %.3g\n",
              cmp$label, r$estimate, r$p.value))
}

cat("\n=== Sign concordance (delta) by GC3 class ===\n")
df %>%
  mutate(concordant = sign(delta_231) == sign(delta_mcf7)) %>%
  count(gc3_class, concordant) %>%
  print()
```

```{r}

```