---
title: "55 — Pooled RSCU: Hypoxia-Promoted vs Normoxia-Promoted (MCF7-SIX1)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Mirror of notebook 54 (MDA-MB-231) using MCF7-SIX1 data. Corpus-level (pooled) RSCU
(Sharp & Li 1987): all codon counts from the promoted gene set are summed first, then
a single RSCU value is computed per codon. Length-weighted.

**Sign convention (critical):** Raw MCF7-SIX1 DESeq2 files have
`log2FoldChange = log2(si3d / sictrl)` — positive = eIF3d *inhibits*. All raw LFCs
are negated before use: `te_lfc = -log2FoldChange` so positive = eIF3d promotes
(matching MDA-MB-231 convention). Gene set CSVs from nb42 are already negated —
safe to use directly after re-filtering.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(ggrepel)
library(ggVennDiagram)
```

---

# Section 1: Setup — CDS Sequences & Codon Counts

```{r load_data}
feature_matrix <- readRDS(
  here("output", "predictive_modeling", "feature_matrix_mcf7six1_codon_features.rds")
)
keep_tx <- feature_matrix$transcript_id_clean
cat("Feature matrix:", nrow(feature_matrix), "genes\n")

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% keep_tx
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0 & width(cds_frame0) > 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0), "\n")
```

```{r codon_matrices}
ct0         <- codonTable(cds_frame0)
counts_mat0 <- codonCounts(ct0)
rownames(counts_mat0) <- names(cds_frame0)
cat("Codon counts matrix:", nrow(counts_mat0), "transcripts x", ncol(counts_mat0), "codons\n")
stopifnot("Zero-count rows in codon matrix" = all(rowSums(counts_mat0) > 0))
```

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

aa_three_letter <- c(
  A = "Ala", C = "Cys", D = "Asp", E = "Glu", F = "Phe",
  G = "Gly", H = "His", I = "Ile", K = "Lys", L = "Leu",
  M = "Met", N = "Asn", P = "Pro", Q = "Gln", R = "Arg",
  S = "Ser", T = "Thr", V = "Val", W = "Trp", Y = "Tyr"
)

codon_gc3 <- tibble(
  codon          = sense_codons,
  gc3_nucleotide = substr(sense_codons, 3, 3),
  gc3_class      = if_else(substr(sense_codons, 3, 3) %in% c("G", "C"),
                           "GC-ending", "AT-ending")
)

tx_to_gene <- feature_matrix %>%
  dplyr::select(transcript_id_clean, gene_id_clean) %>%
  mutate(ensembl_gene = gene_id_clean)

tx_gene_df <- tibble(transcript_id_clean = rownames(counts_mat0)) %>%
  left_join(tx_to_gene, by = "transcript_id_clean")
```

---

```{r params}
lfc_cutoff  <- 0
padj_cutoff <- 0.05
n_boot      <- 1000
seed        <- 9
```

# Section 2: Gene Set Loading

```{r gene_sets}
hyp_raw <- read_csv(
  here("output", "genesets", "mcf7six1_hypoxia_3d_promotes_TE_1hr_lfc0.csv"),
  show_col_types = FALSE
)
cat("Hypoxia gene set columns:", paste(colnames(hyp_raw), collapse = ", "), "\n")
stopifnot(
  "te_lfc column missing from hypoxia gene set CSV"  = "te_lfc"  %in% colnames(hyp_raw),
  "te_padj column missing from hypoxia gene set CSV" = "te_padj" %in% colnames(hyp_raw)
)
hyp_lfc_vec  <- hyp_raw[[which(colnames(hyp_raw) == "te_lfc")]]
hyp_padj_vec <- hyp_raw[[which(colnames(hyp_raw) == "te_padj")]]
hyp_raw      <- hyp_raw[hyp_lfc_vec > lfc_cutoff & !is.na(hyp_padj_vec) & hyp_padj_vec < padj_cutoff, ]
hyp_gene_ids <- unique(hyp_raw$ensembl_gene)
cat("Hypoxia 1hr promoted genes (te_lfc >", lfc_cutoff, ", padj <",
    padj_cutoff, "):", length(hyp_gene_ids), "\n")

nor_raw <- read_csv(
  here("output", "genesets", "mcf7six1_normoxia_3d_promotes_TE_1hr_lfc0.csv"),
  show_col_types = FALSE
)
cat("Normoxia gene set columns:", paste(colnames(nor_raw), collapse = ", "), "\n")
stopifnot(
  "te_lfc column missing from normoxia gene set CSV"  = "te_lfc"  %in% colnames(nor_raw),
  "te_padj column missing from normoxia gene set CSV" = "te_padj" %in% colnames(nor_raw)
)
nor_lfc_vec  <- nor_raw[[which(colnames(nor_raw) == "te_lfc")]]
nor_padj_vec <- nor_raw[[which(colnames(nor_raw) == "te_padj")]]
nor_raw      <- nor_raw[nor_lfc_vec > lfc_cutoff & !is.na(nor_padj_vec) & nor_padj_vec < padj_cutoff, ]
nor_gene_ids <- unique(nor_raw$ensembl_gene)
cat("Normoxia 1hr promoted genes (te_lfc >", lfc_cutoff, ", padj <",
    padj_cutoff, "):", length(nor_gene_ids), "\n")

stopifnot(
  "No hypoxia promoted genes found"  = length(hyp_gene_ids) > 0,
  "No normoxia promoted genes found" = length(nor_gene_ids) > 0
)

overlap_ids <- intersect(hyp_gene_ids, nor_gene_ids)
cat("Overlap:", length(overlap_ids),
    "| Hypoxia-only:", length(setdiff(hyp_gene_ids, nor_gene_ids)),
    "| Normoxia-only:", length(setdiff(nor_gene_ids, hyp_gene_ids)), "\n")
```

```{r sign_flip_validation}
# Verify sign convention: raw DESeq2 log2FoldChange = log2(si3d/sictrl).
# Positive = eIF3d inhibits. Must negate before use: te_lfc = -log2FoldChange.
# Gene set CSVs from nb42 already have negated te_lfc; safe to use directly.
#
# Check: the first hypoxia-promoted gene (from hyp_gene_ids) must have
#   raw log2FoldChange < 0  (eIF3d inhibits in raw direction)
#   -log2FoldChange > 0     (promotes after sign flip)

hyp_raw_deseq <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/hypoxia_si_ctrl_3d.csv",
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

known_gene  <- hyp_gene_ids[1]
# Symbol from the raw DESeq2 file — avoids depending on hyp_raw column structure
known_sym   <- hyp_raw_deseq$symbol[hyp_raw_deseq$gene_id_clean == known_gene][1]
raw_lfc_val <- hyp_raw_deseq$log2FoldChange[hyp_raw_deseq$gene_id_clean == known_gene]

cat("Sign flip check -- gene:", known_sym, "(", known_gene, ")\n")
cat("  Raw log2FoldChange (si3d/sictrl):", round(raw_lfc_val, 4),
    " <- should be negative (eIF3d inhibits in raw)\n")
cat("  After negation (te_lfc):", round(-raw_lfc_val, 4),
    " <- should be positive (eIF3d promotes)\n")

stopifnot(
  "Sign flip check: gene not found in DESeq2 file" =
    length(raw_lfc_val) > 0,
  "Sign flip check failed: raw LFC should be negative for a hypoxia-promoted gene" =
    raw_lfc_val < 0
)
cat("Sign flip validation passed.\n")
```

---

# Section 3: Venn Diagram — Gene Set Overlap

```{r venn_diagram, fig.width=5, fig.height=4.5}
venn_sets <- list(
  "Hypoxia\n1hr promoted"  = hyp_gene_ids,
  "Normoxia\n1hr promoted" = nor_gene_ids
)

p_venn <- ggVennDiagram(venn_sets, label_alpha = 0,
                        set_color = c("#D55E00", "#009E73")) +
  scale_fill_gradient(low = "white", high = "white") +
  labs(title = "eIF3d-promoted gene set overlap\nHypoxia 1hr vs Normoxia 1hr (MCF7-SIX1)",
       subtitle = paste0("lfc > ", lfc_cutoff, ", padj < ", padj_cutoff)) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, size = 11))

print(p_venn)
ggsave(here("plots", "55_venn_pooled_hyp_vs_nor.pdf"), p_venn, width = 8, height = 4.5)
```

---

# Section 4: Pooled Background Computation

```{r pooled_background}
# Pool raw codon counts across ALL expressed transcripts (from MCF7-SIX1 feature matrix).
# Consistent with pooled group RSCU computed below.
bg_counts_pooled <- colSums(counts_mat0)
cat("Total pooled sense codons (background):",
    sum(bg_counts_pooled[sense_codons[sense_codons %in% names(bg_counts_pooled)]]), "\n")

bg_props_pooled <- imap(aa_to_codons, function(codons, aa) {
  codons   <- intersect(codons, names(bg_counts_pooled))
  if (length(codons) == 0) return(NULL)
  aa_total <- sum(bg_counts_pooled[codons])
  if (aa_total == 0) return(NULL)
  bg_counts_pooled[codons] / aa_total
})
bg_props_pooled <- Filter(Negate(is.null), bg_props_pooled)

cat("Background proportions computed for", length(bg_props_pooled), "amino acids\n")
prop_sums <- sapply(bg_props_pooled, sum)
cat("Proportions sum check (should all be ~1):", round(range(prop_sums), 6), "\n")
```

---

# Section 5: Group Transcript Assignment

```{r assign_groups}
hyp_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% hyp_gene_ids) %>%
  pull(transcript_id_clean)
nor_tx <- tx_gene_df %>%
  filter(ensembl_gene %in% nor_gene_ids) %>%
  pull(transcript_id_clean)

hyp_tx <- hyp_tx[hyp_tx %in% rownames(counts_mat0)]
nor_tx <- nor_tx[nor_tx %in% rownames(counts_mat0)]

cat("Hypoxia promoted transcripts:", length(hyp_tx), "\n")
cat("Normoxia promoted transcripts:", length(nor_tx), "\n")

stopifnot(
  "No hypoxia transcripts in matrix"  = length(hyp_tx) > 0,
  "No normoxia transcripts in matrix" = length(nor_tx) > 0
)
```

---

# Section 6: Pooled RSCU & Bootstrap Significance

```{r pooled_rscu_function}
# Compute a single pooled RSCU vector from a set of transcript IDs.
# For each codon j in amino acid aa:
#   pooled_RSCU_j = (sum_count_j / sum_aa_total) / bg_proportion_j
compute_pooled_rscu <- function(tx_ids, counts_mat, bg_props) {
  tx_ids  <- intersect(tx_ids, rownames(counts_mat))
  pooled  <- colSums(counts_mat[tx_ids, , drop = FALSE])
  rscu_vec <- set_names(rep(NA_real_, length(sense_codons)), sense_codons)
  for (aa in names(bg_props)) {
    props    <- bg_props[[aa]]
    codons   <- intersect(names(props), names(pooled))
    if (length(codons) == 0) next
    aa_total <- sum(pooled[codons])
    if (aa_total == 0) next
    bp      <- props[codons]
    valid   <- bp > 0
    if (any(valid)) {
      rscu_vec[codons[valid]] <- (pooled[codons[valid]] / aa_total) / bp[valid]
    }
  }
  rscu_vec
}
```

```{r compute_pooled_rscu}
hyp_counts_pooled <- colSums(counts_mat0[hyp_tx, , drop = FALSE])
nor_counts_pooled <- colSums(counts_mat0[nor_tx, , drop = FALSE])

rscu_hyp   <- compute_pooled_rscu(hyp_tx, counts_mat0, bg_props_pooled)
rscu_nor   <- compute_pooled_rscu(nor_tx, counts_mat0, bg_props_pooled)
delta_rscu <- rscu_hyp - rscu_nor

stopifnot(
  "Pooled RSCU contains negative values (hypoxia)"  = all(rscu_hyp >= 0, na.rm = TRUE),
  "Pooled RSCU contains negative values (normoxia)" = all(rscu_nor >= 0, na.rm = TRUE)
)
cat("Hypoxia pooled RSCU range:", round(range(rscu_hyp, na.rm = TRUE), 3), "\n")
cat("Normoxia pooled RSCU range:", round(range(rscu_nor, na.rm = TRUE), 3), "\n")
cat("Total sense codons -- hypoxia pool:",
    sum(hyp_counts_pooled[sense_codons[sense_codons %in% names(hyp_counts_pooled)]]),
    "| normoxia pool:",
    sum(nor_counts_pooled[sense_codons[sense_codons %in% names(nor_counts_pooled)]]), "\n")
```

```{r bg_validation}
# Validation: bg-RSCU computed from the background itself should = 1 for all codons.
rscu_bg_check <- compute_pooled_rscu(rownames(counts_mat0), counts_mat0, bg_props_pooled)
rscu_bg_sense <- rscu_bg_check[sense_codons[sense_codons %in% names(rscu_bg_check)]]
rscu_bg_sense <- rscu_bg_sense[!is.na(rscu_bg_sense)]
cat("Pooled bg-RSCU (should all = 1):",
    "min =", round(min(rscu_bg_sense), 6),
    "max =", round(max(rscu_bg_sense), 6), "\n")
stopifnot(
  "Pooled bg-RSCU deviates from 1 -- background computation error" =
    all(abs(rscu_bg_sense - 1) < 1e-10)
)
```

```{r spot_check}
# Biological spot-check: TTA (AT-ending Leu) — if pattern replicates, should be positive.
tta_delta <- delta_rscu["TTA"]
cat("TTA delta pooled RSCU (expect positive if pattern replicates from MDA-MB-231):",
    round(tta_delta, 4), "\n")

cat("CTG pooled RSCU -- hypoxia:", round(rscu_hyp["CTG"], 3),
    "| normoxia:", round(rscu_nor["CTG"], 3),
    "| delta:", round(rscu_hyp["CTG"] - rscu_nor["CTG"], 3), "\n")
```

```{r bootstrap}
cat("Running bootstrap with", n_boot, "iterations...\n")
set.seed(seed)

boot_list <- lapply(seq_len(n_boot), function(i) {
  b_hyp  <- sample(hyp_tx, length(hyp_tx), replace = TRUE)
  b_nor  <- sample(nor_tx, length(nor_tx), replace = TRUE)
  rscu_h <- compute_pooled_rscu(b_hyp, counts_mat0, bg_props_pooled)
  rscu_n <- compute_pooled_rscu(b_nor, counts_mat0, bg_props_pooled)
  list(hyp = rscu_h, nor = rscu_n, delta = rscu_h - rscu_n)
})

boot_delta_mat <- do.call(cbind, lapply(boot_list, `[[`, "delta"))
boot_hyp_mat   <- do.call(cbind, lapply(boot_list, `[[`, "hyp"))
boot_nor_mat   <- do.call(cbind, lapply(boot_list, `[[`, "nor"))

delta_ci_lo <- apply(boot_delta_mat, 1, quantile, 0.025, na.rm = TRUE)
delta_ci_hi <- apply(boot_delta_mat, 1, quantile, 0.975, na.rm = TRUE)
hyp_ci_lo   <- apply(boot_hyp_mat,   1, quantile, 0.025, na.rm = TRUE)
hyp_ci_hi   <- apply(boot_hyp_mat,   1, quantile, 0.975, na.rm = TRUE)
nor_ci_lo   <- apply(boot_nor_mat,   1, quantile, 0.025, na.rm = TRUE)
nor_ci_hi   <- apply(boot_nor_mat,   1, quantile, 0.975, na.rm = TRUE)

cat("Bootstrap complete -- CIs computed for", length(delta_ci_lo), "codons.\n")
```

```{r chisq_test}
# Per-codon 2x2 chi-squared test on raw pooled counts.
# For codon j in amino acid aa: rows = groups (hyp, nor), cols = (j, all other synonyms).
chi_df <- imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
  codons    <- intersect(codons, names(hyp_counts_pooled))
  if (length(codons) < 2) return(NULL)
  n_aa_hyp  <- sum(hyp_counts_pooled[codons])
  n_aa_nor  <- sum(nor_counts_pooled[codons])
  map_dfr(codons, function(codon) {
    n_j_hyp <- hyp_counts_pooled[codon]
    n_j_nor <- nor_counts_pooled[codon]
    mat <- matrix(
      c(n_j_hyp, n_aa_hyp - n_j_hyp,
        n_j_nor, n_aa_nor - n_j_nor),
      nrow = 2, byrow = TRUE
    )
    test <- if (any(mat < 5)) fisher.test(mat) else chisq.test(mat, correct = FALSE)
    tibble(amino_acid = aa, codon = codon, p_val_chi = test$p.value)
  })
}) %>%
  mutate(p_adj_chi = p.adjust(p_val_chi, method = "BH"))

cat("Chi-squared test complete.\n")
cat("Significant codons (BH q < 0.05):",
    sum(chi_df$p_adj_chi < 0.05, na.rm = TRUE), "\n")
```

```{r build_results}
rscu_df <- tibble(
  codon       = sense_codons,
  amino_acid  = gen_code[sense_codons]
) %>%
  filter(amino_acid != "*") %>%
  mutate(
    rscu_hyp    = rscu_hyp[codon],
    rscu_nor    = rscu_nor[codon],
    delta_rscu  = delta_rscu[codon],
    delta_ci_lo = delta_ci_lo[codon],
    delta_ci_hi = delta_ci_hi[codon],
    hyp_ci_lo   = hyp_ci_lo[codon],
    hyp_ci_hi   = hyp_ci_hi[codon],
    nor_ci_lo   = nor_ci_lo[codon],
    nor_ci_hi   = nor_ci_hi[codon]
  ) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  left_join(chi_df %>% dplyr::select(codon, p_val_chi, p_adj_chi),
            by = "codon") %>%
  mutate(significant = !is.na(p_adj_chi) & p_adj_chi < 0.05)

stopifnot(
  "Result table missing expected columns" =
    all(c("rscu_hyp", "rscu_nor", "delta_rscu", "p_adj_chi", "significant") %in%
          colnames(rscu_df))
)

cat("\nTop 10 codons by |delta pooled RSCU|:\n")
print(rscu_df %>%
        filter(amino_acid %in% multi_aa) %>%
        arrange(desc(abs(delta_rscu))) %>%
        dplyr::select(codon, amino_acid, rscu_hyp, rscu_nor, delta_rscu,
                      p_adj_chi, significant) %>%
        head(10))
```

---

# Section 7: Visualization

```{r plot_sidebyside, fig.width=11, fig.height=9}
plot_df_a <- rscu_df %>%
  filter(amino_acid %in% multi_aa) %>%
  pivot_longer(
    cols      = c(rscu_hyp, rscu_nor),
    names_to  = "group",
    values_to = "pooled_rscu"
  ) %>%
  mutate(
    ci_lo    = if_else(group == "rscu_hyp", hyp_ci_lo, nor_ci_lo),
    ci_hi    = if_else(group == "rscu_hyp", hyp_ci_hi, nor_ci_hi),
    group    = recode(group,
                      rscu_hyp = "Hypoxia promoted",
                      rscu_nor = "Normoxia promoted"),
    group    = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted")),
    aa_label = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    codon    = reorder(codon, pooled_rscu)
  )

p_a <- ggplot(plot_df_a, aes(x = codon, y = pooled_rscu, fill = group)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                position = position_dodge(width = 0.8), width = 0.3,
                linewidth = 0.4) +
  scale_fill_manual(values = c("Hypoxia promoted"  = "#D55E00",
                                "Normoxia promoted" = "#009E73"),
                    name = NULL) +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "Pooled RSCU: hypoxia vs normoxia eIF3d-promoted (MCF7-SIX1)",
    subtitle = "Pooled corpus-level RSCU +/- 95% bootstrap CI; dashed = 1 (same rate as transcriptome background)",
    x = "Codon", y = "Pooled RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    strip.text      = element_text(face = "bold"),
    legend.position = "top"
  )
p_a
ggsave(here("plots", "55_rscu_pooled_sidebyside_hyp_vs_nor.pdf"),
       p_a, width = 11, height = 9)
```

```{r plot_delta, fig.width=10, fig.height=8}
plot_df_b <- rscu_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    aa_label  = paste0(aa_three_letter[amino_acid], " (", amino_acid, ")"),
    direction = if_else(delta_rscu >= 0, "hypoxia", "normoxia"),
    sig_label = if_else(!is.na(significant) & significant, "*", ""),
    codon     = reorder(codon, delta_rscu)
  )

p_b <- ggplot(plot_df_b, aes(x = codon, y = delta_rscu, fill = direction)) +
  geom_col() +
  geom_text(aes(label = sig_label, y = delta_rscu + sign(delta_rscu) * 0.015),
            size = 3, vjust = 0.5) +
  scale_fill_manual(values = c(hypoxia = "#D55E00", normoxia = "#009E73"),
                    guide = "none") +
  facet_wrap(~ aa_label, scales = "free_x", nrow = 4) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(
    title    = "delta Pooled RSCU: hypoxia-promoted minus normoxia-promoted (MCF7-SIX1)",
    subtitle = "Positive = codon more characteristic of hypoxia-promoted genes;\n* = BH-adjusted chi-squared q < 0.05",
    x = "Codon", y = "delta pooled RSCU (hypoxia minus normoxia)"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text  = element_text(face = "bold")
  )
p_b
ggsave(here("plots", "55_rscu_pooled_delta_hyp_vs_nor.pdf"),
       p_b, width = 10, height = 8)
```

```{r plot_scatter, fig.width=6, fig.height=5.5}
comp_scatter <- rscu_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    sig         = if_else(!is.na(significant) & significant, "Significant", "Neither"),
    label_codon = if_else(sig == "Significant", codon, NA_character_)
  )

r_val    <- cor(comp_scatter$rscu_hyp, comp_scatter$rscu_nor, use = "complete.obs")
axis_lim <- range(c(comp_scatter$rscu_hyp, comp_scatter$rscu_nor), na.rm = TRUE)
axis_lim <- axis_lim + diff(axis_lim) * c(-0.05, 0.05)

p_scatter <- ggplot(comp_scatter, aes(x = rscu_nor, y = rscu_hyp)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linewidth = 0.5,
              linetype = "dashed") +
  geom_hline(yintercept = 1, colour = "grey85", linewidth = 0.3) +
  geom_vline(xintercept = 1, colour = "grey85", linewidth = 0.3) +
  geom_point(aes(shape = sig, colour = sig, size = sig, alpha = sig)) +
  ggrepel::geom_text_repel(aes(label = label_codon), size = 2.8,
                           max.overlaps = 25, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = c(Significant = "#0072B2", Neither = "grey70"),
                      name = "Significant (BH q < 0.05)") +
  scale_shape_manual(values  = c(Significant = 17, Neither = 16),
                     name = "Significant (BH q < 0.05)") +
  scale_size_manual(values   = c(Significant = 2.5, Neither = 1.6),
                    guide = "none") +
  scale_alpha_manual(values  = c(Significant = 1, Neither = 0.5),
                     guide = "none") +
  annotate("text",
           x = axis_lim[1] + diff(axis_lim) * 0.02,
           y = axis_lim[2] - diff(axis_lim) * 0.02,
           label = paste0("r = ", round(r_val, 2)),
           hjust = 0, vjust = 1, size = 4, colour = "grey30") +
  coord_fixed(xlim = axis_lim, ylim = axis_lim) +
  labs(
    title    = "Pooled RSCU: eIF3d codon usage in hypoxia vs normoxia promoted genes (MCF7-SIX1)",
    subtitle = paste0("Each point = one codon (multi-synonymous AAs). Dashed = identity line.\n",
                      "Points above diagonal: higher pooled usage in hypoxia-promoted genes."),
    x = "Pooled RSCU (normoxia-promoted)",
    y = "Pooled RSCU (hypoxia-promoted)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_scatter
ggsave(here("plots", "55_rscu_pooled_scatter_hyp_vs_nor.pdf"), p_scatter, width = 6, height = 5.5)

p_scatter_ci <- p_scatter +
  geom_errorbar(aes(ymin = hyp_ci_lo, ymax = hyp_ci_hi),
                width = 0, linewidth = 0.3, alpha = 0.4) +
  geom_errorbarh(aes(xmin = nor_ci_lo, xmax = nor_ci_hi),
                 height = 0, linewidth = 0.3, alpha = 0.4) +
  labs(subtitle = paste0("Each point = one codon (multi-synonymous AAs). Dashed = identity line.\n",
                         "Error bars = 95% bootstrap CI per group (gene-resampling uncertainty)."))
p_scatter_ci
ggsave(here("plots", "55_rscu_pooled_scatter_hyp_vs_nor_ci.pdf"), p_scatter_ci, width = 6, height = 5.5)
```

```{r plot_lollipop, fig.width=5.5, fig.height=9}
plot_lollipop <- rscu_df %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    sig = case_when(
      !is.na(significant) & significant & delta_rscu >= 0 ~ "Hypoxia only",
      !is.na(significant) & significant & delta_rscu <  0 ~ "Normoxia only",
      TRUE ~ "Neither"
    ),
    sig   = factor(sig, levels = c("Hypoxia only", "Normoxia only", "Neither")),
    codon = fct_reorder(codon, delta_rscu)
  )
codon_order_f0 <- levels(plot_lollipop$codon)  # saved for frame-shift plot

lollipop_base <- ggplot(plot_lollipop,
                        aes(x = delta_rscu, y = codon, colour = gc3_class)) +
  geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_point(aes(shape = sig), size = 2) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  scale_shape_manual(
    values = c("Hypoxia only" = 17, "Normoxia only" = 15, "Neither" = 16),
    name   = "Significant (BH q < 0.05)"
  ) +
  labs(x = "delta pooled RSCU (hypoxia minus normoxia)", y = NULL) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y     = element_text(size = 7, family = "mono"),
    legend.position = "top",
    legend.box      = "vertical"
  )

p_lollipop <- lollipop_base +
  labs(
    title    = "Ranked delta Pooled RSCU: hypoxia vs normoxia eIF3d-promoted (MCF7-SIX1)",
    subtitle = "Positive = stronger codon usage in hypoxia-promoted gene pool."
  )
p_lollipop
ggsave(here("plots", "55_rscu_pooled_delta_ranked.pdf"), p_lollipop, width = 5.5, height = 9)

p_lollipop_ci <- lollipop_base +
  geom_errorbarh(aes(xmin = delta_ci_lo, xmax = delta_ci_hi),
                 height = 0.3, linewidth = 0.4, alpha = 0.6) +
  labs(
    title    = "Ranked delta Pooled RSCU: hypoxia vs normoxia eIF3d-promoted (MCF7-SIX1)",
    subtitle = "Positive = stronger codon usage in hypoxia-promoted gene pool.\nError bars = 95% bootstrap CI on delta (gene-resampling uncertainty)."
  )
p_lollipop_ci
ggsave(here("plots", "55_rscu_pooled_delta_ranked_ci.pdf"), p_lollipop_ci, width = 5.5, height = 9)
```

---

# Section 7d: Validation — Is the AT/GC Preference Specific to Synonymous Positions?

## Frame-Shift Control

```{r frame_shift_ctrl}
shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len  <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  seqs <- seqs[width(seqs) > 0]
  ct   <- codonTable(seqs)
  mat  <- codonCounts(ct)
  rownames(mat) <- names(seqs)
  mat
}

cds_frame1  <- shift_frame(cds_frame0, 1)
cds_frame2  <- shift_frame(cds_frame0, 2)
counts_mat1 <- make_counts(cds_frame1)
counts_mat2 <- make_counts(cds_frame2)

rscu_hyp_f1 <- compute_pooled_rscu(hyp_tx, counts_mat1, bg_props_pooled)
rscu_nor_f1 <- compute_pooled_rscu(nor_tx, counts_mat1, bg_props_pooled)
delta_f1    <- rscu_hyp_f1 - rscu_nor_f1

rscu_hyp_f2 <- compute_pooled_rscu(hyp_tx, counts_mat2, bg_props_pooled)
rscu_nor_f2 <- compute_pooled_rscu(nor_tx, counts_mat2, bg_props_pooled)
delta_f2    <- rscu_hyp_f2 - rscu_nor_f2

cat("Frame 0 mean |delta| (multi-synonymous codons):",
    round(mean(abs(delta_rscu[sense_codons[sense_codons %in% names(delta_rscu)]]),
               na.rm = TRUE), 4), "\n")
cat("Frame 1 mean |delta|:",
    round(mean(abs(delta_f1[sense_codons[sense_codons %in% names(delta_f1)]]),
               na.rm = TRUE), 4), "\n")
cat("Frame 2 mean |delta|:",
    round(mean(abs(delta_f2[sense_codons[sense_codons %in% names(delta_f2)]]),
               na.rm = TRUE), 4), "\n")
```

```{r plot_frame_ctrl_scatter, fig.width=6, fig.height=5.5}
frame_ctrl_df <- tibble(
  codon      = sense_codons,
  amino_acid = gen_code[sense_codons]
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    delta_f0 = delta_rscu[codon],
    delta_f1 = delta_f1[codon]
  ) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  filter(!is.na(delta_f0), !is.na(delta_f1))

ax_lim <- range(c(frame_ctrl_df$delta_f0, frame_ctrl_df$delta_f1), na.rm = TRUE)
ax_lim <- ax_lim + diff(ax_lim) * c(-0.05, 0.05)

p_frame_scatter <- ggplot(frame_ctrl_df,
                          aes(x = delta_f1, y = delta_f0, colour = gc3_class)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60", linewidth = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey80") +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey80") +
  geom_point(size = 2.2, alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = codon), size = 2.5,
                           max.overlaps = 20, min.segment.length = 0.2,
                           show.legend = FALSE) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  coord_fixed(xlim = ax_lim, ylim = ax_lim) +
  labs(
    title    = "Frame-shift control: frame-0 vs frame-1 delta pooled RSCU (MCF7-SIX1)",
    subtitle = paste0("x = delta in frame-1 (non-coding); y = frame-0 (real codons).\n",
                      "Points above diagonal: stronger AT/GC effect in real frame than in shifted frame."),
    x = "delta pooled RSCU, frame +1 (non-coding control)",
    y = "delta pooled RSCU, frame 0 (real codons)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_frame_scatter
ggsave(here("plots", "55_rscu_pooled_frame_ctrl_scatter.pdf"),
       p_frame_scatter, width = 6, height = 5.5)
```

```{r plot_frame_ctrl_lollipop, fig.width=7, fig.height=9}
frame_long <- tibble(
  codon      = sense_codons,
  amino_acid = gen_code[sense_codons]
) %>%
  filter(amino_acid %in% multi_aa) %>%
  mutate(
    `Frame 0\n(real codons)`       = delta_rscu[codon],
    `Frame +1\n(non-coding ctrl)`  = delta_f1[codon],
    `Frame +2\n(non-coding ctrl)`  = delta_f2[codon]
  ) %>%
  pivot_longer(cols = starts_with("Frame"),
               names_to = "frame", values_to = "delta") %>%
  mutate(frame = factor(frame,
                        levels = c("Frame 0\n(real codons)",
                                   "Frame +1\n(non-coding ctrl)",
                                   "Frame +2\n(non-coding ctrl)"))) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(codon = factor(codon, levels = codon_order_f0))

p_frame_lollipop <- ggplot(frame_long, aes(x = delta, y = codon, colour = gc3_class)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third-position") +
  facet_wrap(~ frame, nrow = 1) +
  labs(
    title    = "Frame-shift control: delta pooled RSCU across frames (MCF7-SIX1)",
    subtitle = "Codon order fixed to frame-0 ranking. AT/GC pattern should weaken in non-coding frames if real.",
    x = "delta pooled RSCU (hypoxia minus normoxia)",
    y = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y     = element_text(size = 6, family = "mono"),
    legend.position = "top",
    strip.text      = element_text(face = "bold")
  )
p_frame_lollipop
ggsave(here("plots", "55_rscu_pooled_frame_ctrl_lollipop.pdf"),
       p_frame_lollipop, width = 7, height = 9)
```

```{r plot_frame_ctrl_lollipop_star, fig.width=7, fig.height=9}
# Classify codons as "codon-specific" if frame-0 chi-sq significant AND
# |frame-0 delta| > 1.5x max(|frame+1|, |frame+2|) — must beat both shifted frames.
frame_specificity <- tibble(
  codon    = names(delta_rscu),
  f0_delta = as.numeric(delta_rscu),
  f1_delta = as.numeric(delta_f1[names(delta_rscu)]),
  f2_delta = as.numeric(delta_f2[names(delta_rscu)])
) %>%
  mutate(max_shift = pmax(abs(f1_delta), abs(f2_delta))) %>%
  left_join(chi_df %>% dplyr::select(codon, p_adj_chi), by = "codon") %>%
  mutate(codon_specific = !is.na(p_adj_chi) & p_adj_chi < 0.05 &
           abs(f0_delta) > max_shift * 1.5)

cat("Codons with codon-specific preference:",
    sum(frame_specificity$codon_specific, na.rm = TRUE), "of",
    sum(!is.na(frame_specificity$codon_specific)), "multi-synonymous codons\n")

frame_long_star <- frame_long %>%
  left_join(frame_specificity %>% dplyr::select(codon, codon_specific),
            by = "codon") %>%
  mutate(
    codon_specific = replace_na(codon_specific, FALSE),
    codon = factor(codon, levels = codon_order_f0)
  )

x_left <- min(frame_long_star$delta, na.rm = TRUE)
star_data <- frame_long_star %>%
  filter(codon_specific, frame == "Frame 0\n(real codons)") %>%
  mutate(delta = x_left)

p_frame_lollipop_star <- ggplot(
    frame_long_star,
    aes(x = delta, y = codon, colour = gc3_class)
  ) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_point(size = 1.5, alpha = 0.85) +
  geom_point(
    data  = star_data,
    shape = 8, size = 2.5, colour = "black", stroke = 0.6
  ) +
  facet_wrap(~ frame, nrow = 1) +
  scale_colour_manual(
    values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
    name   = "Third-position"
  ) +
  labs(
    title    = "Frame-shift control: star = real codon preference (MCF7-SIX1)",
    subtitle = "star (Frame 0 only): BH-adj chi-sq p<0.05 AND abs(frame-0 delta) > 1.5x max(abs(frame+1), abs(frame+2))",
    x        = "delta pooled RSCU (hypoxia minus normoxia)",
    y        = NULL
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.y     = element_text(size = 6, family = "mono"),
    legend.position = "top",
    strip.text      = element_text(face = "bold")
  )
p_frame_lollipop_star
ggsave(here("plots", "55_rscu_pooled_frame_ctrl_lollipop_star.pdf"),
       p_frame_lollipop_star, width = 7, height = 9)
```

---

## GC Content by Codon Position

```{r gc_by_position, fig.width=5, fig.height=4}
sense_in_mat <- sense_codons[sense_codons %in% colnames(counts_mat0)]

gc_position <- function(counts_vec) {
  counts_vec <- counts_vec[sense_in_mat]
  total <- sum(counts_vec, na.rm = TRUE)
  tibble(
    position = c("GC1", "GC2", "GC3"),
    gc_frac  = c(
      sum(counts_vec[substr(sense_in_mat, 1, 1) %in% c("G","C")], na.rm=TRUE) / total,
      sum(counts_vec[substr(sense_in_mat, 2, 2) %in% c("G","C")], na.rm=TRUE) / total,
      sum(counts_vec[substr(sense_in_mat, 3, 3) %in% c("G","C")], na.rm=TRUE) / total
    )
  )
}

gc_df <- bind_rows(
  gc_position(bg_counts_pooled)  %>% mutate(pool = "Background"),
  gc_position(nor_counts_pooled) %>% mutate(pool = "Normoxia promoted"),
  gc_position(hyp_counts_pooled) %>% mutate(pool = "Hypoxia promoted")
) %>%
  mutate(pool = factor(pool,
                       levels = c("Background", "Normoxia promoted", "Hypoxia promoted")))

stopifnot("GC fractions out of [0,1]" = all(gc_df$gc_frac >= 0 & gc_df$gc_frac <= 1))

cat("GC content by position and pool:\n")
print(gc_df %>% mutate(gc_pct = round(gc_frac * 100, 1)) %>%
        dplyr::select(pool, position, gc_pct) %>%
        pivot_wider(names_from = position, values_from = gc_pct))

p_gc <- ggplot(gc_df, aes(x = position, y = gc_frac * 100, fill = pool)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c(
    "Background"        = "grey60",
    "Normoxia promoted" = "#009E73",
    "Hypoxia promoted"  = "#D55E00"
  ), name = NULL) +
  labs(
    title    = "GC content at each codon position by gene pool (MCF7-SIX1)",
    subtitle = paste0("If AT-enrichment in hypoxia-promoted genes is specific to position 3,\n",
                      "that confirms synonymous codon bias rather than overall gene GC content."),
    x = "Codon position", y = "GC content (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_gc
ggsave(here("plots", "55_rscu_pooled_gc_by_position.pdf"), p_gc, width = 5, height = 4)
```

---

# Section 8: Per-gene GC3 content by gene group

```{r gc3_per_gene}
gc3_codons_in_mat <- sense_in_mat[substr(sense_in_mat, 3, 3) %in% c("G", "C")]

gc3_tx <- tibble(
  transcript_id_clean = rownames(counts_mat0),
  gc3_frac   = apply(counts_mat0[, sense_in_mat, drop = FALSE], 1, function(row) {
    total <- sum(row, na.rm = TRUE)
    if (total == 0) return(NA_real_)
    sum(row[gc3_codons_in_mat], na.rm = TRUE) / total
  }),
  total_codons = rowSums(counts_mat0[, sense_in_mat, drop = FALSE])
) %>%
  left_join(tx_gene_df %>% dplyr::select(transcript_id_clean, ensembl_gene),
            by = "transcript_id_clean") %>%
  filter(!is.na(gc3_frac), total_codons > 0)

# Collapse to gene: keep transcript with most codon counts per gene
gc3_gene <- gc3_tx %>%
  group_by(ensembl_gene) %>%
  slice_max(total_codons, n = 1, with_ties = FALSE) %>%
  ungroup()

stopifnot("GC3 fractions out of [0,1]" =
            all(gc3_gene$gc3_frac >= 0 & gc3_gene$gc3_frac <= 1, na.rm = TRUE))

# Assign group — priority: Overlapping > Hypoxia-only > Normoxia-only > Background
gc3_gene <- gc3_gene %>%
  mutate(group = case_when(
    ensembl_gene %in% overlap_ids                         ~ "Overlapping",
    ensembl_gene %in% setdiff(hyp_gene_ids, overlap_ids) ~ "Hypoxia promoted",
    ensembl_gene %in% setdiff(nor_gene_ids, overlap_ids) ~ "Normoxia promoted",
    TRUE                                                   ~ "Background"
  )) %>%
  mutate(group = factor(group,
    levels = c("Hypoxia promoted", "Overlapping", "Normoxia promoted", "Background")))

cat("GC3 gene-level summary:\n")
gc3_gene %>%
  group_by(group) %>%
  summarise(n = n(), median_gc3 = round(median(gc3_frac * 100, na.rm = TRUE), 1),
            .groups = "drop") %>%
  print()
```

```{r gc3_group_plot, fig.width=6, fig.height=5}
group_colors <- c(
  "Hypoxia promoted"  = "#D55E00",
  "Overlapping"       = "#CC79A7",
  "Normoxia promoted" = "#009E73",
  "Background"        = "grey60"
)

w_hyp_nor <- wilcox.test(
  gc3_gene$gc3_frac[gc3_gene$group == "Hypoxia promoted"],
  gc3_gene$gc3_frac[gc3_gene$group == "Normoxia promoted"]
)$p.value
w_hyp_bg <- wilcox.test(
  gc3_gene$gc3_frac[gc3_gene$group == "Hypoxia promoted"],
  gc3_gene$gc3_frac[gc3_gene$group == "Background"]
)$p.value
w_nor_bg <- wilcox.test(
  gc3_gene$gc3_frac[gc3_gene$group == "Normoxia promoted"],
  gc3_gene$gc3_frac[gc3_gene$group == "Background"]
)$p.value

cat("Wilcoxon p (hyp vs nor):", formatC(w_hyp_nor, format = "e", digits = 2), "\n")
cat("Wilcoxon p (hyp vs bg):", formatC(w_hyp_bg,  format = "e", digits = 2), "\n")
cat("Wilcoxon p (nor vs bg):", formatC(w_nor_bg,  format = "e", digits = 2), "\n")

group_medians <- gc3_gene %>%
  group_by(group) %>%
  summarise(n = n(), med = median(gc3_frac * 100, na.rm = TRUE), .groups = "drop")

n_labels <- group_medians %>%
  mutate(label = paste0(group, "\nn=", n))

p_gc3_group <- ggplot(gc3_gene, aes(x = group, y = gc3_frac * 100, fill = group)) +
  geom_violin(alpha = 0.25, linewidth = 0.5, trim = TRUE) +
  geom_boxplot(width = 0.14, outlier.shape = NA, linewidth = 0.6, fill = "white") +
  geom_text(data = group_medians,
            aes(x = group, y = med, label = round(med, 1)),
            nudge_x = 0.18, size = 3, fontface = "bold", inherit.aes = FALSE) +
  scale_fill_manual(values = group_colors, guide = "none") +
  scale_x_discrete(labels = setNames(n_labels$label, n_labels$group)) +
  annotate("text", x = Inf, y = Inf, hjust = 1.05, vjust = 1.4, size = 3,
           label = paste0("Hyp vs Nor: p=", formatC(w_hyp_nor, format = "e", digits = 1),
                          "\nHyp vs Bg: p=",  formatC(w_hyp_bg,  format = "e", digits = 1),
                          "\nNor vs Bg: p=",  formatC(w_nor_bg,  format = "e", digits = 1))) +
  labs(
    title    = "GC3 content per gene by translational group (MCF7-SIX1)",
    subtitle = "GC% at 3rd codon position; one value per gene (highest-count transcript)",
    x = NULL, y = "GC3 content (%)"
  ) +
  theme_classic(base_size = 12)
p_gc3_group
ggsave(here("plots", "55_gc3_by_group.pdf"), p_gc3_group, width = 6, height = 5)

median_labels <- group_medians %>%
  arrange(med) %>%
  mutate(label_y = seq(0.027, by = -0.003, length.out = n()))

p_gc3_density <- ggplot(gc3_gene, aes(x = gc3_frac * 100, colour = group, fill = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  geom_vline(data = group_medians,
             aes(xintercept = med, colour = group),
             linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
  geom_text(data = median_labels,
            aes(x = med, y = label_y, label = round(med, 1), colour = group),
            hjust = -0.2, size = 3, fontface = "bold",
            show.legend = FALSE, inherit.aes = FALSE) +
  scale_colour_manual(values = group_colors, name = NULL) +
  labs(
    title    = "GC3 content per gene by translational group (MCF7-SIX1)",
    subtitle = "GC% at 3rd codon position; one value per gene (highest-count transcript)",
    x = "GC3 content (%)", y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_gc3_density
ggsave(here("plots", "55_gc3_by_group_density.pdf"), p_gc3_density, width = 6, height = 4)

# Overall CDS GC% (all three codon positions)
gc_per_codon_all <- sapply(sense_in_mat, function(codon) {
  sum(strsplit(codon, "")[[1]] %in% c("G", "C"))
})

gc_all_tx <- tibble(
  transcript_id_clean = rownames(counts_mat0),
  gc_all_frac = apply(counts_mat0[, sense_in_mat, drop = FALSE], 1, function(row) {
    total_nts <- sum(row, na.rm = TRUE) * 3
    if (total_nts == 0) return(NA_real_)
    sum(row * gc_per_codon_all, na.rm = TRUE) / total_nts
  }),
  total_codons = rowSums(counts_mat0[, sense_in_mat, drop = FALSE])
) %>%
  filter(!is.na(gc_all_frac), total_codons > 0)

gc_all_gene <- gc3_gene %>%
  dplyr::select(transcript_id_clean, ensembl_gene, group) %>%
  left_join(gc_all_tx %>% dplyr::select(transcript_id_clean, gc_all_frac),
            by = "transcript_id_clean") %>%
  filter(!is.na(gc_all_frac))

stopifnot("GC_all fractions out of [0,1]" =
            all(gc_all_gene$gc_all_frac >= 0 & gc_all_gene$gc_all_frac <= 1, na.rm = TRUE))

gc_all_medians <- gc_all_gene %>%
  group_by(group) %>%
  summarise(med = median(gc_all_frac * 100, na.rm = TRUE), .groups = "drop")

gc_all_median_labels <- gc_all_medians %>%
  arrange(med) %>%
  mutate(label_y = seq(0.10, by = -0.01, length.out = n()))

cat("Overall GC% medians by group:\n")
print(gc_all_medians %>% mutate(med = round(med, 1)))

p_gc_all_density <- ggplot(gc_all_gene,
                            aes(x = gc_all_frac * 100, colour = group, fill = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  geom_vline(data = gc_all_medians,
             aes(xintercept = med, colour = group),
             linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
  geom_text(data = gc_all_median_labels,
            aes(x = med, y = label_y, label = round(med, 1), colour = group),
            hjust = -0.2, size = 3, fontface = "bold",
            show.legend = FALSE, inherit.aes = FALSE) +
  scale_colour_manual(values = group_colors, name = NULL) +
  labs(
    title    = "Overall CDS GC content per gene by translational group (MCF7-SIX1)",
    subtitle = "GC% across all three codon positions; one value per gene",
    x = "CDS GC content (%)", y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_gc_all_density
ggsave(here("plots", "55_gc_all_by_group_density.pdf"), p_gc_all_density, width = 6, height = 4)
```

```{r gc3_lfc_scatter, fig.width=8, fig.height=4}
# Load MCF7-SIX1 hypoxia TE LFC for all genes.
# Raw log2FoldChange = log2(si3d/sictrl); negate so positive = eIF3d promotes.
hyp_lfc_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/hypoxia_si_ctrl_3d.csv",
  show_col_types = FALSE
) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    te_lfc        = -log2FoldChange  # negate: raw is log2(si3d/sictrl)
  )

nor_lfc_raw <- read_csv(
  "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/normoxia_sictrl_3d.csv",
  show_col_types = FALSE
) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    te_lfc        = -log2FoldChange  # negate
  )

# Verify negation is consistent with gene set (spot-check first gene again)
hyp_lfc_check <- hyp_lfc_raw$te_lfc[hyp_lfc_raw$gene_id_clean == known_gene]
stopifnot("te_lfc sign check in scatter section failed" =
  length(hyp_lfc_check) > 0 && hyp_lfc_check > 0)

lfc_df <- bind_rows(
  hyp_lfc_raw %>% dplyr::select(gene_id_clean, te_lfc, symbol) %>%
    mutate(condition = "Hypoxia 1hr"),
  nor_lfc_raw %>% dplyr::select(gene_id_clean, te_lfc, symbol) %>%
    mutate(condition = "Normoxia 1hr")
) %>%
  inner_join(gc3_gene %>% dplyr::select(ensembl_gene, gc3_frac, group),
             by = c("gene_id_clean" = "ensembl_gene")) %>%
  filter(group %in% c("Hypoxia promoted", "Overlapping", "Normoxia promoted"))

cor_df <- lfc_df %>%
  group_by(condition) %>%
  summarise(
    rho = round(cor(te_lfc, gc3_frac, method = "spearman", use = "complete.obs"), 3),
    .groups = "drop"
  )
cat("Spearman correlation (te_lfc vs GC3%):\n"); print(cor_df)

p_gc3_lfc <- ggplot(lfc_df, aes(x = te_lfc, y = gc3_frac * 100, colour = group)) +
  geom_point(size = 1.2, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey30", linewidth = 0.7) +
  geom_text(data = cor_df,
            aes(x = Inf, y = Inf, label = paste0("rho=", rho)),
            colour = "black", hjust = 1.1, vjust = 1.4, size = 3.5,
            inherit.aes = FALSE) +
  facet_wrap(~ condition, scales = "free_x") +
  scale_colour_manual(values = group_colors, name = NULL) +
  labs(
    title    = "GC3 content vs TE LFC (MCF7-SIX1, eIF3d knockdown)",
    subtitle = "Promoted genes only; Spearman rho per condition",
    x        = "TE LFC (sictrl / si3d, positive = eIF3d promotes)",
    y        = "GC3 content (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_gc3_lfc
ggsave(here("plots", "55_gc3_vs_lfc_scatter.pdf"), p_gc3_lfc, width = 8, height = 4)
```

```{r gc3_lfc_quantile, fig.width=10, fig.height=4}
# Split all genes with LFC + GC3 data into 8 equal quantile bins by TE LFC.
make_quantile_df <- function(lfc_tbl, condition_label) {
  lfc_tbl %>%
    inner_join(gc3_gene %>% dplyr::select(ensembl_gene, gc3_frac),
               by = c("gene_id_clean" = "ensembl_gene")) %>%
    filter(!is.na(te_lfc), !is.na(gc3_frac)) %>%
    mutate(
      quantile_bin = ntile(te_lfc, 8),
      condition    = condition_label
    )
}

hyp_q <- make_quantile_df(hyp_lfc_raw, "Hypoxia 1hr")
nor_q <- make_quantile_df(nor_lfc_raw, "Normoxia 1hr")

bin_medians <- bind_rows(hyp_q, nor_q) %>%
  group_by(condition, quantile_bin) %>%
  summarise(med_lfc = round(median(te_lfc), 2), n = n(), .groups = "drop")
cat("Genes per quantile bin:\n"); print(bin_medians)

lfc_q_all <- bind_rows(hyp_q, nor_q) %>%
  mutate(
    quantile_bin = factor(quantile_bin),
    condition    = factor(condition, levels = c("Hypoxia 1hr", "Normoxia 1hr"))
  )

bin_medians_fct <- bin_medians %>%
  mutate(
    quantile_bin = factor(quantile_bin),
    condition    = factor(condition, levels = c("Hypoxia 1hr", "Normoxia 1hr"))
  )

p_gc3_quantile <- ggplot(lfc_q_all,
                          aes(x = quantile_bin, y = gc3_frac * 100, fill = condition)) +
  geom_violin(alpha = 0.3, linewidth = 0.5, trim = TRUE) +
  geom_boxplot(width = 0.1, outlier.shape = NA, linewidth = 0.5, fill = "white") +
  geom_text(data = bin_medians_fct,
            aes(x = quantile_bin, y = -Inf,
                label = med_lfc),
            vjust = -0.3, size = 2.5, colour = "grey30", inherit.aes = FALSE) +
  facet_wrap(~ condition, nrow = 1) +
  scale_fill_manual(values = c("Hypoxia 1hr" = "#D55E00", "Normoxia 1hr" = "#009E73"),
                    guide = "none") +
  coord_cartesian(clip = "off") +
  labs(
    title    = "GC3 content across TE LFC quantiles (MCF7-SIX1)",
    subtitle = "All expressed genes split into 8 equal bins by TE LFC (Q1 = lowest / most inhibited, Q8 = highest / most promoted)",
    x        = "TE LFC quantile (Q1 to Q8)",
    y        = "GC3 content (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text   = element_text(face = "bold"),
    plot.margin  = margin(t = 5, r = 5, b = 40, l = 5)
  )
p_gc3_quantile
ggsave(here("plots", "55_gc3_lfc_quantile.pdf"), p_gc3_quantile, width = 10, height = 4)
```

```{r aa_codon_proportion, fig.width=7, fig.height=4}
# Arginine (R): check if the large delta_rscu spread replicates in MCF7-SIX1.
target_aa   <- "R"
target_name <- "Arginine (R)"
aa_codons   <- aa_to_codons[[target_aa]]
aa_codons   <- aa_codons[aa_codons %in% names(hyp_counts_pooled)]

prop_df <- bind_rows(
  tibble(codon = aa_codons,
         prop  = hyp_counts_pooled[aa_codons] / sum(hyp_counts_pooled[aa_codons]),
         group = "Hypoxia promoted"),
  tibble(codon = aa_codons,
         prop  = nor_counts_pooled[aa_codons] / sum(nor_counts_pooled[aa_codons]),
         group = "Normoxia promoted"),
  tibble(codon = aa_codons,
         prop  = bg_counts_pooled[aa_codons]  / sum(bg_counts_pooled[aa_codons]),
         group = "Background")
) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(group = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted", "Background")))

codon_order <- prop_df %>%
  filter(group == "Hypoxia promoted") %>%
  arrange(gc3_class, desc(prop)) %>%
  pull(codon)
prop_df <- prop_df %>% mutate(codon = factor(codon, levels = codon_order))

stopifnot("Arginine proportions do not sum to 1" =
  all(abs(tapply(prop_df$prop, prop_df$group, sum) - 1) < 1e-10))

cat("Arginine codon proportions (%):\n")
print(prop_df %>%
  pivot_wider(names_from = group, values_from = prop, id_cols = c(codon, gc3_class)) %>%
  mutate(across(where(is.numeric), ~ round(. * 100, 1))))

p_aa_prop <- ggplot(prop_df, aes(x = codon, y = prop * 100, fill = group)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = round(prop * 100, 1)),
            position = position_dodge(width = 0.75),
            vjust = -0.4, size = 2.5) +
  scale_fill_manual(values = c(
    "Hypoxia promoted"  = "#D55E00",
    "Normoxia promoted" = "#009E73",
    "Background"        = "grey60"
  ), name = NULL) +
  facet_grid(~ gc3_class, scales = "free_x", space = "free_x") +
  labs(
    title    = paste0("Synonymous codon usage: ", target_name, " (MCF7-SIX1)"),
    subtitle = "Proportion of each Arg codon within pooled Arg counts; faceted by 3rd-base class",
    x = NULL, y = "Codon proportion (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", strip.text = element_text(face = "bold"))
p_aa_prop
ggsave(here("plots", "55_arginine_codon_proportion.pdf"), p_aa_prop, width = 7, height = 4)
```

```{r aa_codon_proportion_gly, fig.width=5, fig.height=4}
# Glycine (G): 4 codons, clean AT/GC split in MDA-MB-231 — does it replicate?
target_aa   <- "G"
target_name <- "Glycine (G)"
aa_codons   <- aa_to_codons[[target_aa]]
aa_codons   <- aa_codons[aa_codons %in% names(hyp_counts_pooled)]

prop_df_gly <- bind_rows(
  tibble(codon = aa_codons,
         prop  = hyp_counts_pooled[aa_codons] / sum(hyp_counts_pooled[aa_codons]),
         group = "Hypoxia promoted"),
  tibble(codon = aa_codons,
         prop  = nor_counts_pooled[aa_codons] / sum(nor_counts_pooled[aa_codons]),
         group = "Normoxia promoted"),
  tibble(codon = aa_codons,
         prop  = bg_counts_pooled[aa_codons]  / sum(bg_counts_pooled[aa_codons]),
         group = "Background")
) %>%
  left_join(codon_gc3 %>% dplyr::select(codon, gc3_class), by = "codon") %>%
  mutate(group = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted", "Background")))

codon_order_gly <- prop_df_gly %>%
  filter(group == "Hypoxia promoted") %>%
  arrange(gc3_class, desc(prop)) %>%
  pull(codon)
prop_df_gly <- prop_df_gly %>% mutate(codon = factor(codon, levels = codon_order_gly))

stopifnot("Glycine proportions do not sum to 1" =
  all(abs(tapply(prop_df_gly$prop, prop_df_gly$group, sum) - 1) < 1e-10))

cat("Glycine codon proportions (%):\n")
print(prop_df_gly %>%
  pivot_wider(names_from = group, values_from = prop, id_cols = c(codon, gc3_class)) %>%
  mutate(across(where(is.numeric), ~ round(. * 100, 1))))

p_aa_prop_gly <- ggplot(prop_df_gly, aes(x = codon, y = prop * 100, fill = group)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(aes(label = round(prop * 100, 1)),
            position = position_dodge(width = 0.75),
            vjust = -0.4, size = 2.5) +
  scale_fill_manual(values = c(
    "Hypoxia promoted"  = "#D55E00",
    "Normoxia promoted" = "#009E73",
    "Background"        = "grey60"
  ), name = NULL) +
  facet_grid(~ gc3_class, scales = "free_x", space = "free_x") +
  labs(
    title    = paste0("Synonymous codon usage: ", target_name, " (MCF7-SIX1)"),
    subtitle = "Proportion of each Gly codon within pooled Gly counts; faceted by 3rd-base class",
    x = NULL, y = "Codon proportion (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", strip.text = element_text(face = "bold"))
p_aa_prop_gly
ggsave(here("plots", "55_glycine_codon_proportion.pdf"), p_aa_prop_gly, width = 5, height = 4)
```

---

# Section 9: Save Results

```{r save_csv}
out_df <- rscu_df %>%
  dplyr::select(amino_acid, codon, gc3_class,
                rscu_hyp, hyp_ci_lo, hyp_ci_hi,
                rscu_nor, nor_ci_lo, nor_ci_hi,
                delta_rscu, delta_ci_lo, delta_ci_hi,
                p_val_chi, p_adj_chi, significant)

write_csv(out_df,
          here("output", "predictive_modeling", "bg_rscu_pooled_mcf7six1_hyp_vs_nor.csv"))
cat("Saved to output/predictive_modeling/bg_rscu_pooled_mcf7six1_hyp_vs_nor.csv\n")
cat("Rows:", nrow(out_df), "\n")
cat("Significant codons (BH q < 0.05):", sum(out_df$significant, na.rm = TRUE), "\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
