---
title: "56 — Candidate mRNA Selection and GC3 Content (MDA-MB-231)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Motivation: nb54 found that pooled RSCU in hypoxia-specific eIF3d targets
is enriched for AT-ending (low-GC3) codons. This notebook asks whether that
effect is visible at the level of individual mRNAs. We pick the top
`r params$n_candidates` mRNAs most strongly and specifically promoted by
eIF3d in hypoxia (not promoted in normoxia) in MDA-MB-231 cells, then compare
their per-gene GC3% to background expressed genes and negative controls.

All data: MDA-MB-231.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r params}
n_candidates <- 30
lfc_cutoff   <- 0.5
padj_cutoff  <- 0.05
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(knitr)
```

---

# Section 1: Load Data

```{r load_feature_matrix}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_te_lfc.rds"))
cat("Feature matrix:", nrow(fm), "transcripts x", ncol(fm), "columns\n")

# Dedup to one transcript per gene (take first, as in other notebooks)
fm_gene <- fm |>
  distinct(gene_id_clean, .keep_all = TRUE)
cat("After dedup:", nrow(fm_gene), "genes\n")

# Sanity check: gc3 range
stopifnot("gc3 out of range" = all(fm_gene$gc3 >= 0 & fm_gene$gc3 <= 100, na.rm = TRUE))
cat("GC3 range:", round(range(fm_gene$gc3, na.rm = TRUE), 1), "\n")
```

```{r load_gene_sets}
hyp_specific <- read_csv(
  here("output", "genesets",
       "hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv"),
  show_col_types = FALSE
)
cat("Hypoxia-specific gene set:", nrow(hyp_specific), "genes\n")
# Strip version from ensembl ID for joining
hyp_specific <- hyp_specific |>
  mutate(gene_id_clean = sub("\\..*", "", ensembl_gene))
```

```{r load_te_direct}
# DESeq2 TE results for si3d knockdown in hypoxia 1hr.
# Sign convention (MDA-MB-231): positive te_lfc = eIF3d promotes translation
# (sictrl TE > si3d TE, i.e. knockdown reduces TE).
hyp_te <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, hyp_te_lfc = te_lfc, hyp_te_padj = te_padj)
cat("Hypoxia si3d TE genes:", nrow(hyp_te), "\n")
cat("  promoted (LFC >", params$lfc_cutoff, ", padj <", params$padj_cutoff, "):",
    sum(hyp_te$hyp_te_lfc > params$lfc_cutoff & hyp_te$hyp_te_padj < params$padj_cutoff, na.rm = TRUE), "\n")

# Normoxia 1hr equivalent (same sign convention).
nor_te <- read_csv(
  here("output", "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
  dplyr::select(gene_id_clean, nor_te_lfc = te_lfc, nor_te_padj = te_padj)
cat("Normoxia si3d TE genes:", nrow(nor_te), "\n")
cat("  promoted (LFC >", params$lfc_cutoff, ", padj <", params$padj_cutoff, "):",
    sum(nor_te$nor_te_lfc > params$lfc_cutoff & nor_te$nor_te_padj < params$padj_cutoff, na.rm = TRUE), "\n")
```

```{r load_neg_controls}
neg_ctrl <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) |>
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
cat("Negative control genes:", nrow(neg_ctrl), "\n")
```

---

# Section 2: Identify Top Candidates

Rank all hypoxia-specific eIF3d targets by 1hr hypoxia TE LFC descending,
take top `r params$n_candidates`.

```{r build_candidates}
# Inner join with feature matrix: requires protein-coding gene with GC3 data
# (excludes fusion genes, lncRNAs, and other non-CDS entries)
candidates <- hyp_specific |>
  inner_join(hyp_te, by = "gene_id_clean") |>
  left_join(nor_te,  by = "gene_id_clean") |>
  inner_join(fm_gene |> dplyr::select(gene_id_clean, gc3),
             by = "gene_id_clean") |>
  filter(!is.na(gc3)) |>
  arrange(desc(hyp_te_lfc)) |>
  slice_head(n = params$n_candidates)

stopifnot("Wrong number of candidates" = nrow(candidates) == params$n_candidates)
cat("Top", params$n_candidates, "hypoxia-specific eIF3d targets (protein-coding, with GC3) selected.\n")
```

```{r candidate_table}
candidates |>
  dplyr::select(
    Symbol              = db_gene_symbol,
    `Hyp TE LFC (1hr)` = hyp_te_lfc,
    `Hyp TE padj`       = hyp_te_padj,
    `Nor TE LFC (1hr)` = nor_te_lfc,
    `Nor TE padj`       = nor_te_padj,
    `GC3 (%)`           = gc3
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(caption = paste0(
    "Top ", params$n_candidates,
    " mRNAs most strongly promoted by eIF3d in hypoxia (231s, 1hr, si3d vs sictrl), ",
    "not promoted in normoxia. Ranked by hypoxia TE LFC."
  ))
```

---

# Section 3: Build Comparison Groups

```{r build_groups}
background_gc3 <- fm_gene |>
  dplyr::select(gene_id_clean, gc3)

neg_gc3 <- fm_gene |>
  filter(gene_id_clean %in% neg_ctrl$gene_id_clean) |>
  dplyr::select(gene_id_clean, gc3)

cat("Background genes:", nrow(background_gc3), "\n")
cat("Negative control genes with GC3:", nrow(neg_gc3), "\n")

# Wilcoxon: candidates vs background
wt <- wilcox.test(candidates$gc3, background_gc3$gc3, exact = FALSE)
cat("Wilcoxon candidates vs background: W =", wt$statistic,
    ", p =", signif(wt$p.value, 3), "\n")
```

---

# Section 4: Plot — Per-gene GC3 for Candidates

```{r plot_pergene, fig.width=13, fig.height=4}
bg_median  <- median(background_gc3$gc3, na.rm = TRUE)
bg_q25     <- quantile(background_gc3$gc3, 0.25, na.rm = TRUE)
bg_q75     <- quantile(background_gc3$gc3, 0.75, na.rm = TRUE)
neg_median <- median(neg_gc3$gc3, na.rm = TRUE)

cat("Background median GC3:", round(bg_median, 1), "%\n")
cat("Negative control median GC3:", round(neg_median, 1), "%\n")

cand_plot <- candidates |>
  dplyr::select(symbol = db_gene_symbol, gc3, hyp_te_lfc) |>
  filter(!is.na(gc3)) |>
  mutate(symbol = fct_reorder(symbol, gc3))

p_pergene <- ggplot(cand_plot, aes(x = symbol, y = gc3)) +
  geom_hline(yintercept = bg_median, colour = "grey50",
             linetype = "dashed", linewidth = 0.7) +
  geom_point(colour = "#E69F00", size = 3) +
  # geom_text(aes(label = round(gc3, 1)), vjust = -0.8, size = 3.0) +
  labs(
    x        = NULL,
    y        = "GC3 (%)",
    title    = "GC3% for top 10 hypoxia-specific eIF3d targets",
    subtitle = paste0("Grey dashed = background median (", round(bg_median, 1), "%)")
  ) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

print(p_pergene)
```

---

# Section 5: Save Plots

```{r save_plots}
pdf(here("plots", "56_candidate_gc3_comparison.pdf"), width = 13, height = 4.5)
print(p_pergene)
dev.off()
cat("Plots saved to plots/56_candidate_gc3_comparison.pdf\n")
```

---

# Section 6: Summary Table

```{r summary_table}
summary_df <- candidates |>
  dplyr::select(
    Symbol        = db_gene_symbol,
    `Hyp TE LFC` = hyp_te_lfc,
    `Nor TE LFC` = nor_te_lfc,
    `GC3 (%)`    = gc3
  ) |>
  mutate(
    `GC3 deviation` = round(gc3 - bg_median, 1),
    `GC3 (%)`    = round(gc3, 1),
    `Hyp TE LFC` = round(`Hyp TE LFC`, 2),
    `Nor TE LFC` = round(`Nor TE LFC`, 2)
  )

kable(summary_df,
  caption = paste0(
    "Top ", params$n_candidates, " candidates. ",
    "GC3 deviation = candidate GC3 minus background median (",
    round(bg_median, 1), "%). Negative = AT-enriched relative to background."
  )
)
```
