---
title: "Notebook 58: riboWaltz E/P/A Site Codon Occupancy (MDA-MB-231, 1hr)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook uses riboWaltz to compute per-codon occupancy at the E, P, and A ribosomal
sites for MDA-MB-231 siCTRL and si3d knockdown samples at 1 hour, in both normoxia and
hypoxia. Enrichment is expressed as log2(siCTRL / si3d): positive = codon more occupied
when eIF3d is present; negative = codon less occupied when eIF3d is present.

Inputs:
- `ribowaltz_output/timecourse_231/reads_psite_list_timecourse.rds`
- `ribowaltz_output/timecourse_231/annot_object.rds`
- `accessories/metadata_231_timecourse.csv`

```{r load_libraries}
library(riboWaltz)
library(tidyverse)
library(here)
library(pheatmap)
```

---

## Section 1: Load data and map samples to conditions

```{r load_data}
annot <- readRDS(here("ribowaltz_output", "timecourse_231", "annot_object.rds"))
reads_psite_list <- readRDS(here("ribowaltz_output", "timecourse_231",
                                  "reads_psite_list_timecourse.rds"))

cat("Sample names in reads_psite_list:\n")
print(names(reads_psite_list))
cat("\nAnnotation object dimensions:", nrow(annot), "transcripts\n")
```

```{r filter_samples}
meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE)

meta_1hr <- meta %>%
  dplyr::filter(type == "RiboSeq", hours == 1,
                treatment %in% c("siCTRL", "si3d"))

cat("1hr RiboSeq siCTRL + si3d samples:\n")
print(meta_1hr %>% dplyr::select(id, rep, treatment, condition))

# Match sample names: riboWaltz names may include suffixes after the NM ID
# Strip everything after the first underscore-free NM ID match
psite_names <- names(reads_psite_list)
sample_map <- meta_1hr %>%
  rowwise() %>%
  mutate(psite_name = {
    hit <- psite_names[str_detect(psite_names, fixed(id))]
    if (length(hit) == 1) hit else NA_character_
  }) %>%
  ungroup() %>%
  dplyr::filter(!is.na(psite_name))

cat("\nMatched", nrow(sample_map), "samples to psite list\n")
stopifnot("Expected 12 matched samples (3 reps × 2 treatments × 2 conditions)" =
  nrow(sample_map) == 12)

reads_filtered <- reads_psite_list[sample_map$psite_name]
```

---

## Section 2: Compute codon coverage at E, P, A sites

```{r codon_coverage}
epa_coverage <- codon_coverage(reads_psite_list = reads_filtered,
                                annotation = annot,
                                psite = TRUE,
                                asite = TRUE,
                                esite = TRUE)

cat("codon_coverage output class:", class(epa_coverage), "\n")
cat("Dimensions:", nrow(epa_coverage), "×", ncol(epa_coverage), "\n")
cat("Column names:\n")
print(names(epa_coverage))
```

---

## Section 3: Aggregate and normalize

```{r aggregate}
# Identify site columns: expect pattern <sample>_psite, <sample>_asite, <sample>_esite
psite_cols <- grep("_psite$", names(epa_coverage), value = TRUE)
asite_cols <- grep("_asite$", names(epa_coverage), value = TRUE)
esite_cols <- grep("_esite$", names(epa_coverage), value = TRUE)

stopifnot("P-site columns present" = length(psite_cols) == 12)
stopifnot("A-site columns present" = length(asite_cols) == 12)
stopifnot("E-site columns present" = length(esite_cols) == 12)

# Sum per codon across all transcript positions
codon_counts <- epa_coverage %>%
  group_by(codon) %>%
  summarise(across(all_of(c(psite_cols, asite_cols, esite_cols)), sum),
            .groups = "drop") %>%
  dplyr::filter(!str_detect(codon, "N|X"))   # drop ambiguous codons

# Pivot to long format
codon_long <- codon_counts %>%
  pivot_longer(-codon, names_to = "sample_site", values_to = "count") %>%
  mutate(
    psite_name = str_remove(sample_site, "_(p|a|e)site$"),
    site       = str_extract(sample_site, "(p|a|e)site$")
  ) %>%
  dplyr::select(-sample_site)

# Join with condition metadata
codon_long <- codon_long %>%
  left_join(dplyr::select(sample_map, psite_name, rep, treatment, condition),
            by = "psite_name")

stopifnot("All rows have condition metadata" = !any(is.na(codon_long$condition)))
```

```{r normalize}
# Relative frequency within each (sample, site)
codon_freq <- codon_long %>%
  group_by(psite_name, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup()

stopifnot("Frequencies in [0,1]" = all(codon_freq$freq >= 0 & codon_freq$freq <= 1,
                                        na.rm = TRUE))

# Average across replicates
codon_means <- codon_freq %>%
  group_by(codon, site, condition, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE), .groups = "drop")
```

```{r enrichment}
# log2(siCTRL / si3d): positive = codon more occupied when eIF3d is present
enrichment_df <- codon_means %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(log2_enrich = log2(siCTRL / si3d)) %>%
  dplyr::filter(is.finite(log2_enrich))

# Order codons by P-site siCTRL frequency for consistent axis ordering
psite_order <- codon_means %>%
  dplyr::filter(site == "psite", treatment == "siCTRL", condition == "normoxia") %>%
  arrange(desc(mean_freq)) %>%
  pull(codon)

enrichment_df <- enrichment_df %>%
  mutate(codon = factor(codon, levels = psite_order))
```

---

## Section 4: Validation checks

```{r validation}
# ATG (Met / start codons) should rank in top P-site codons in siCTRL
atg_rank <- which(psite_order == "ATG")
cat("ATG P-site rank (normoxia siCTRL):", atg_rank, "of", length(psite_order), "\n")
stopifnot("ATG in top 20 P-site codons" = atg_rank <= 20)

# Stop codons should have low P-site frequency
stop_codons <- c("TAA", "TAG", "TGA")
stop_pfreq <- codon_means %>%
  dplyr::filter(codon %in% stop_codons, site == "psite", treatment == "siCTRL",
                condition == "normoxia") %>%
  pull(mean_freq)
cat("Stop codon P-site frequencies:", round(stop_pfreq, 5), "\n")
stopifnot("Stop codons rare at P-site (< 0.01)" = all(stop_pfreq < 0.01, na.rm = TRUE))
```

---

## Section 5: Genetic code — codon to amino acid mapping

```{r codon_aa}
genetic_code <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

enrichment_df <- enrichment_df %>%
  mutate(aa = genetic_code[as.character(codon)])

codon_means <- codon_means %>%
  mutate(aa = genetic_code[as.character(codon)])
```

---

## Section 6: Visualizations

### Plot A: Dot plots of log2(siCTRL/si3d) enrichment, per condition

```{r plot_dot_normoxia, fig.width=14, fig.height=5}
site_labels <- c(esite = "E site", psite = "P site", asite = "A site")

plot_dot <- function(cond) {
  enrichment_df %>%
    dplyr::filter(condition == cond, !aa %in% c("Stop", NA)) %>%
    mutate(site = factor(site, levels = c("esite", "psite", "asite"),
                         labels = site_labels)) %>%
    ggplot(aes(x = codon, y = log2_enrich, color = aa)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    facet_wrap(~site, ncol = 3) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5),
      legend.position = "right",
      legend.title = element_text(size = 9),
      legend.text  = element_text(size = 8)
    ) +
    labs(
      title  = paste0("Codon E/P/A site occupancy — ", cond, " (siCTRL vs si3d, 1hr)"),
      x      = "Codon",
      y      = "log2(siCTRL / si3d)",
      color  = "Amino acid"
    )
}

p_nor <- plot_dot("normoxia")
p_hyp <- plot_dot("hypoxia")

print(p_nor)
```

```{r plot_dot_hypoxia, fig.width=14, fig.height=5}
print(p_hyp)
```

```{r save_dot_plots}
ggsave(here("plots", "58_epa_enrichment_dot_normoxia.pdf"), p_nor,
       width = 14, height = 5)
ggsave(here("plots", "58_epa_enrichment_dot_hypoxia.pdf"), p_hyp,
       width = 14, height = 5)
```

### Plot B: Heatmap of log2(siCTRL/si3d) enrichment

```{r plot_heatmap, fig.width=10, fig.height=12}
make_heatmap <- function(cond) {
  mat <- enrichment_df %>%
    dplyr::filter(condition == cond, !aa %in% c("Stop", NA)) %>%
    dplyr::select(codon, site, log2_enrich) %>%
    pivot_wider(names_from = site, values_from = log2_enrich) %>%
    tibble::column_to_rownames("codon") %>%
    as.matrix()

  # Order rows by amino acid, then codon
  row_order <- enrichment_df %>%
    dplyr::filter(condition == cond, !aa %in% c("Stop", NA)) %>%
    distinct(codon, aa) %>%
    arrange(aa, codon) %>%
    pull(codon)
  mat <- mat[row_order[row_order %in% rownames(mat)], , drop = FALSE]

  # Annotation for amino acid grouping
  aa_annot <- enrichment_df %>%
    dplyr::filter(condition == cond, !aa %in% c("Stop", NA)) %>%
    distinct(codon, aa) %>%
    tibble::column_to_rownames("codon")

  lim <- max(abs(mat), na.rm = TRUE)
  pheatmap(
    mat,
    cluster_rows   = FALSE,
    cluster_cols   = FALSE,
    color          = colorRampPalette(c("#2166AC", "white", "#D6604D"))(101),
    breaks         = seq(-lim, lim, length.out = 102),
    annotation_row = aa_annot,
    main           = paste0("log2(siCTRL / si3d) — ", cond, " 1hr"),
    fontsize_row   = 6,
    angle_col      = 45,
    border_color   = NA
  )
}

make_heatmap("normoxia")
```

```{r heatmap_hypoxia}
make_heatmap("hypoxia")
```

```{r save_heatmap}
pdf(here("plots", "58_epa_enrichment_heatmap.pdf"), width = 10, height = 12)
make_heatmap("normoxia")
make_heatmap("hypoxia")
dev.off()
```

### Plot C: Global P-site codon composition in siCTRL (normoxia vs hypoxia)

```{r plot_psite_composition, fig.width=14, fig.height=5}
p_comp <- codon_means %>%
  dplyr::filter(site == "psite", treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  mutate(codon = factor(codon, levels = psite_order)) %>%
  ggplot(aes(x = codon, y = mean_freq, fill = condition)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c(normoxia = "#56B4E9", hypoxia = "#E69F00")) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5)
  ) +
  labs(
    title = "P-site codon composition — siCTRL, normoxia vs hypoxia (1hr)",
    x     = "Codon (ordered by normoxia frequency)",
    y     = "Mean relative frequency",
    fill  = "Condition"
  )

print(p_comp)
ggsave(here("plots", "58_psite_composition_sictrl.pdf"), p_comp,
       width = 14, height = 5)
```

---

## Session info

```{r session_info}
sessionInfo()
```
