---
title: "Notebook 58: riboWaltz E/P/A Site Codon Occupancy (MDA-MB-231, 1hr)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This notebook analyses preferential codon occupancy at the E (exit), P (peptidyl), and
A (aminoacyl) ribosomal sites for MDA-MB-231 samples at 1 hour, using pre-computed
riboWaltz output. Three questions:

1. Are E and P site profiles "as expected" (tracking codon usage) while the A site shows
   differential occupancy reflecting decoding pausing?
2. What is the impact of eIF3d KD (si3d vs siCTRL) on site occupancy?
3. What is the impact of hypoxia on site occupancy?

Inputs:
- `ribowaltz_output/timecourse_231/epa_codon_occupancy_1hr.csv`
- `accessories/metadata_231_timecourse.csv`

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
library(pheatmap)
```

---

## Section 1: Load data and join metadata

```{r load_data}
epa_raw <- read_csv(here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_1hr.csv"),
                    show_col_types = FALSE)

meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE) %>%
  dplyr::filter(type == "RiboSeq", hours == 1)

cat("EPA CSV dimensions:", nrow(epa_raw), "×", ncol(epa_raw), "\n")
cat("Unique samples in CSV:", n_distinct(epa_raw$sample), "\n")
cat("1hr RiboSeq samples in metadata:", nrow(meta), "\n")
```

```{r join_metadata}
epa_df <- epa_raw %>%
  mutate(
    id    = str_remove(sample, "Aligned\\.toTranscriptome\\.out.*"),
    codon = str_replace_all(codon, "U", "T")   # RNA → DNA notation
  ) %>%
  left_join(meta, by = "id")

stopifnot("All rows have treatment metadata" = !any(is.na(epa_df$treatment)))
stopifnot("18 samples present" = n_distinct(epa_df$id) == 18)

cat("\nSamples per treatment × condition:\n")
epa_df %>%
  distinct(id, treatment, condition) %>%
  count(treatment, condition) %>%
  print()
```

---

## Section 2: Normalize to relative frequencies

```{r normalize}
codon_freq <- epa_df %>%
  group_by(id, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup()

stopifnot("Frequencies in [0,1]" =
  all(codon_freq$freq >= 0 & codon_freq$freq <= 1, na.rm = TRUE))

codon_means <- codon_freq %>%
  group_by(codon, site, condition, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE),
            sd_freq   = sd(freq, na.rm = TRUE),
            .groups = "drop")

cat("Codons per site:\n")
codon_means %>% distinct(codon, site) %>% count(site) %>% print()
```

---

## Section 3: Validation

```{r validation}
genetic_code <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

codon_means <- codon_means %>%
  mutate(aa = genetic_code[codon])

# ATG should rank in top 20 P-site codons in siCTRL normoxia
psite_order <- codon_means %>%
  dplyr::filter(site == "psite", treatment == "siCTRL", condition == "normoxia") %>%
  arrange(desc(mean_freq)) %>%
  pull(codon)

atg_rank <- which(psite_order == "ATG")
cat("ATG P-site rank (siCTRL normoxia):", atg_rank, "of", length(psite_order), "\n")
stopifnot("ATG in top 20 P-site codons" = atg_rank <= 20)

# Stop codons should be rare at P-site
stop_pfreq <- codon_means %>%
  dplyr::filter(codon %in% c("TAA", "TAG", "TGA"),
                site == "psite", treatment == "siCTRL", condition == "normoxia") %>%
  pull(mean_freq)
cat("Stop codon P-site frequencies:", round(stop_pfreq, 5), "\n")
stopifnot("Stop codons rare at P-site (< 0.01)" = all(stop_pfreq < 0.01, na.rm = TRUE))
```

---

## Section 4: Helper objects

```{r helpers}
# Codon order (by P-site siCTRL normoxia frequency, most frequent first)
# Used for consistent x-axis ordering across dot plots
codon_order_pfreq <- psite_order

site_labels <- c(esite = "E site", psite = "P site", asite = "A site")

# Wide helper: one row per codon, columns for each site's mean_freq
wide_by_site <- function(df) {
  df %>%
    pivot_wider(names_from = site, values_from = mean_freq,
                names_prefix = "freq_")
}
```

---

## Plot 0: siCTRL normoxia vs hypoxia codon frequency per site

Scatter of mean relative frequency in siCTRL normoxia (x) vs siCTRL hypoxia (y) for each
codon, faceted by site. Codons with the largest residuals from the y=x line are labelled.

```{r plot_0, fig.width=11, fig.height=4}
plot_0_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = condition, values_from = mean_freq) %>%
  mutate(
    residual = hypoxia - normoxia,
    label    = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                      codon, NA_character_),
    site     = factor(site, levels = c("esite", "psite", "asite"),
                      labels = site_labels)
  )

p_0 <- ggplot(plot_0_df, aes(x = normoxia, y = hypoxia, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_wrap(~site, ncol = 3, scales = "free") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL normoxia vs hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per site",
    x        = "siCTRL normoxia — relative frequency",
    y        = "siCTRL hypoxia — relative frequency",
    color    = "Amino acid"
  )

print(p_0)
ggsave(here("plots", "58_codon_freq_normoxia_vs_hypoxia_sictrl.pdf"), p_0,
       width = 11, height = 4)
```

---

## Plot 0b: siCTRL normoxia vs si3d normoxia codon frequency per site

```{r plot_0b, fig.width=11, fig.height=4}
plot_0b_df <- codon_means %>%
  dplyr::filter(condition == "normoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(
    residual = si3d - siCTRL,
    label    = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                      codon, NA_character_),
    site     = factor(site, levels = c("esite", "psite", "asite"),
                      labels = site_labels)
  )

p_0b <- ggplot(plot_0b_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_wrap(~site, ncol = 3, scales = "free") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d normoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per site",
    x        = "siCTRL normoxia — relative frequency",
    y        = "si3d normoxia — relative frequency",
    color    = "Amino acid"
  )

print(p_0b)
ggsave(here("plots", "58_codon_freq_sictrl_vs_si3d_normoxia.pdf"), p_0b,
       width = 11, height = 4)
```

---

## Plot 0c: siCTRL hypoxia vs si3d hypoxia codon frequency per site

```{r plot_0c, fig.width=11, fig.height=4}
plot_0c_df <- codon_means %>%
  dplyr::filter(condition == "hypoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(
    residual = si3d - siCTRL,
    label    = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                      codon, NA_character_),
    site     = factor(site, levels = c("esite", "psite", "asite"),
                      labels = site_labels)
  )

p_0c <- ggplot(plot_0c_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_wrap(~site, ncol = 3, scales = "free") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per site",
    x        = "siCTRL hypoxia — relative frequency",
    y        = "si3d hypoxia — relative frequency",
    color    = "Amino acid"
  )

print(p_0c)
ggsave(here("plots", "58_codon_freq_sictrl_vs_si3d_hypoxia.pdf"), p_0c,
       width = 11, height = 4)
```

---

## Plot A: P-site vs A-site scatter — is A-site different?

```{r plot_a, fig.width=8, fig.height=4}
plot_a_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = site, values_from = mean_freq) %>%
  rename(freq_p = psite, freq_a = asite, freq_e = esite) %>%
  mutate(ap_log2 = log2(freq_a / freq_p),
         label   = ifelse(abs(ap_log2) > 1, codon, NA_character_))

p_a <- ggplot(plot_a_df, aes(x = freq_p, y = freq_a, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 20,
                  show.legend = FALSE) +
  facet_wrap(~condition, ncol = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 12) +
  theme(legend.position = "right",
        legend.title = element_text(size = 9),
        legend.text  = element_text(size = 8)) +
  labs(
    title  = "A-site vs P-site codon occupancy — siCTRL 1hr",
    subtitle = "Diagonal = identical distribution; labelled codons deviate by |log2(A/P)| > 1",
    x      = "P-site relative frequency",
    y      = "A-site relative frequency",
    color  = "Amino acid"
  )

print(p_a)
ggsave(here("plots", "58a_psite_vs_asite_scatter.pdf"), p_a, width = 8, height = 4)
```

---

## Plot B: A/P pausing index — top paused and fastest codons

```{r plot_b, fig.width=8, fig.height=6}
ap_index <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = site, values_from = mean_freq) %>%
  mutate(ap_ratio = log2(asite / psite)) %>%
  dplyr::filter(is.finite(ap_ratio))

# Top 10 most paused + 10 fastest per condition
top_codons <- ap_index %>%
  group_by(condition) %>%
  slice_max(ap_ratio, n = 10) %>%
  bind_rows(
    ap_index %>%
      group_by(condition) %>%
      slice_min(ap_ratio, n = 10)
  ) %>%
  ungroup() %>%
  distinct(codon)

plot_b_df <- ap_index %>%
  dplyr::filter(codon %in% top_codons$codon) %>%
  mutate(codon = reorder(codon, ap_ratio))

p_b <- ggplot(plot_b_df, aes(x = ap_ratio, y = codon, color = aa)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_segment(aes(x = 0, xend = ap_ratio, y = codon, yend = codon),
               color = "grey70", linewidth = 0.5) +
  geom_point(size = 2.2, alpha = 0.9) +
  facet_wrap(~condition, ncol = 2) +
  theme_classic(base_size = 12) +
  theme(legend.position = "right",
        legend.title = element_text(size = 9),
        legend.text  = element_text(size = 8),
        axis.text.y  = element_text(size = 8)) +
  labs(
    title    = "A/P pausing index — siCTRL 1hr (top 10 paused & fastest codons)",
    subtitle = "log2(A-site freq / P-site freq): positive = enriched at A-site (slower decoding)",
    x        = "log2(A-site / P-site)",
    y        = "Codon",
    color    = "Amino acid"
  )

print(p_b)
ggsave(here("plots", "58b_ap_pausing_index.pdf"), p_b, width = 8, height = 6)
```

---

## Plot C: Impact of eIF3d KD on site occupancy

```{r enrich_kd}
enrich_kd <- codon_means %>%
  dplyr::filter(treatment %in% c("siCTRL", "si3d"), !aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, condition, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(log2_enrich = log2(siCTRL / si3d)) %>%
  dplyr::filter(is.finite(log2_enrich)) %>%
  mutate(codon = factor(codon, levels = codon_order_pfreq),
         site  = factor(site, levels = c("esite", "psite", "asite"),
                        labels = site_labels))
```

```{r plot_c_dot, fig.width=14, fig.height=6}
p_c_dot <- ggplot(enrich_kd, aes(x = codon, y = log2_enrich, color = aa)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_grid(condition ~ site) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4.5),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title  = "eIF3d KD effect on codon occupancy — log2(siCTRL / si3d), 1hr",
    x      = "Codon (ordered by P-site siCTRL normoxia frequency)",
    y      = "log2(siCTRL / si3d)",
    color  = "Amino acid"
  )

print(p_c_dot)
ggsave(here("plots", "58c_eif3d_kd_enrichment_dotplot.pdf"), p_c_dot,
       width = 14, height = 6)
```

```{r plot_c_heatmap, fig.width=10, fig.height=12}
make_kd_heatmap <- function(cond) {
  mat <- enrich_kd %>%
    dplyr::filter(condition == cond) %>%
    dplyr::select(codon, site, log2_enrich) %>%
    pivot_wider(names_from = site, values_from = log2_enrich) %>%
    tibble::column_to_rownames("codon") %>%
    as.matrix()

  row_order <- enrich_kd %>%
    dplyr::filter(condition == cond) %>%
    distinct(codon, aa) %>%
    arrange(aa, codon) %>%
    pull(codon)
  mat <- mat[row_order[row_order %in% rownames(mat)], , drop = FALSE]

  aa_annot <- enrich_kd %>%
    dplyr::filter(condition == cond) %>%
    distinct(codon, aa) %>%
    tibble::column_to_rownames("codon")

  lim <- max(abs(mat), na.rm = TRUE)
  pheatmap(
    mat,
    cluster_rows   = FALSE,
    cluster_cols   = FALSE,
    color          = colorRampPalette(c("#2166AC", "white", "#D6604D"))(101),
    breaks         = seq(-lim, lim, length.out = 102),
    annotation_row = aa_annot,
    main           = paste0("log2(siCTRL / si3d) — ", cond, " 1hr"),
    fontsize_row   = 6,
    angle_col      = 45,
    border_color   = NA
  )
}

make_kd_heatmap("normoxia")
```

```{r plot_c_heatmap_hyp}
make_kd_heatmap("hypoxia")
```

```{r save_c_heatmap}
pdf(here("plots", "58c_eif3d_kd_enrichment_heatmap.pdf"), width = 10, height = 12)
make_kd_heatmap("normoxia")
make_kd_heatmap("hypoxia")
dev.off()
```

---

## Plot D: Impact of hypoxia on site occupancy

```{r enrich_hyp}
enrich_hyp <- codon_means %>%
  dplyr::filter(!aa %in% c("Stop", NA)) %>%
  dplyr::select(codon, aa, site, condition, treatment, mean_freq) %>%
  pivot_wider(names_from = condition, values_from = mean_freq) %>%
  mutate(log2_enrich = log2(hypoxia / normoxia)) %>%
  dplyr::filter(is.finite(log2_enrich)) %>%
  mutate(codon = factor(codon, levels = codon_order_pfreq),
         site  = factor(site, levels = c("esite", "psite", "asite"),
                        labels = site_labels))
```

```{r plot_d_dot, fig.width=14, fig.height=6}
p_d_dot <- enrich_hyp %>%
  dplyr::filter(treatment %in% c("siCTRL", "si3d")) %>%
  ggplot(aes(x = codon, y = log2_enrich, color = aa)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_grid(treatment ~ site) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4.5),
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title  = "Hypoxia effect on codon occupancy — log2(hypoxia / normoxia), 1hr",
    x      = "Codon (ordered by P-site siCTRL normoxia frequency)",
    y      = "log2(hypoxia / normoxia)",
    color  = "Amino acid"
  )

print(p_d_dot)
ggsave(here("plots", "58d_hypoxia_enrichment_dotplot.pdf"), p_d_dot,
       width = 14, height = 6)
```

```{r plot_d_heatmap, fig.width=10, fig.height=12}
make_hyp_heatmap <- function(trt) {
  mat <- enrich_hyp %>%
    dplyr::filter(treatment == trt) %>%
    dplyr::select(codon, site, log2_enrich) %>%
    pivot_wider(names_from = site, values_from = log2_enrich) %>%
    tibble::column_to_rownames("codon") %>%
    as.matrix()

  row_order <- enrich_hyp %>%
    dplyr::filter(treatment == trt) %>%
    distinct(codon, aa) %>%
    arrange(aa, codon) %>%
    pull(codon)
  mat <- mat[row_order[row_order %in% rownames(mat)], , drop = FALSE]

  aa_annot <- enrich_hyp %>%
    dplyr::filter(treatment == trt) %>%
    distinct(codon, aa) %>%
    tibble::column_to_rownames("codon")

  lim <- max(abs(mat), na.rm = TRUE)
  pheatmap(
    mat,
    cluster_rows   = FALSE,
    cluster_cols   = FALSE,
    color          = colorRampPalette(c("#2166AC", "white", "#D6604D"))(101),
    breaks         = seq(-lim, lim, length.out = 102),
    annotation_row = aa_annot,
    main           = paste0("log2(hypoxia / normoxia) — ", trt, " 1hr"),
    fontsize_row   = 6,
    angle_col      = 45,
    border_color   = NA
  )
}

make_hyp_heatmap("siCTRL")
```

```{r plot_d_heatmap_si3d}
make_hyp_heatmap("si3d")
```

```{r save_d_heatmap}
pdf(here("plots", "58d_hypoxia_enrichment_heatmap.pdf"), width = 10, height = 12)
make_hyp_heatmap("siCTRL")
make_hyp_heatmap("si3d")
dev.off()
```

---

## Plot E: A-site — eIF3d KD effect vs hypoxia effect

Do the same codons respond to both perturbations at the A site?

```{r plot_e, fig.width=6, fig.height=6}
asite_kd <- enrich_kd %>%
  dplyr::filter(site == "A site", condition == "normoxia") %>%
  dplyr::select(codon, aa, log2_kd = log2_enrich)

asite_hyp <- enrich_hyp %>%
  dplyr::filter(site == "A site", treatment == "siCTRL") %>%
  dplyr::select(codon, aa, log2_hyp = log2_enrich)

plot_e_df <- inner_join(asite_kd, asite_hyp, by = c("codon", "aa")) %>%
  mutate(label = ifelse(abs(log2_kd) > 0.3 | abs(log2_hyp) > 0.3,
                        codon, NA_character_))

p_e <- ggplot(plot_e_df, aes(x = log2_kd, y = log2_hyp, color = aa)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.6) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  theme_classic(base_size = 12) +
  theme(legend.position = "right",
        legend.title = element_text(size = 9),
        legend.text  = element_text(size = 8)) +
  labs(
    title    = "A-site: eIF3d KD vs hypoxia effect (1hr normoxia siCTRL reference)",
    subtitle = "Labelled codons: |log2| > 0.3 in either axis",
    x        = "log2(siCTRL / si3d) A-site [normoxia]",
    y        = "log2(hypoxia / normoxia) A-site [siCTRL]",
    color    = "Amino acid"
  )

print(p_e)
ggsave(here("plots", "58e_asite_kd_vs_hypoxia_scatter.pdf"), p_e, width = 6, height = 6)
```

---

## Session info

```{r session_info}
sessionInfo()
```
