---
title: "Notebook 60: riboWaltz E/P/A Site Codon Occupancy — Positive vs Negative Gene Sets (MDA-MB-231, 1hr)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Replicates the Plot 0 scatter plots from Notebook 58, comparing codon frequency at E/P/A
sites between two transcript subsets:

- **eIF3d promotes** — positive gene set (hypoxia 1hr eIF3d promotes translation)
- **Negative control** — matched negative control transcripts

Inputs:
- `ribowaltz_output/timecourse_231/epa_codon_occupancy_positive_hypoxia_3d_1.csv`
- `ribowaltz_output/timecourse_231/epa_codon_occupancy_matched_negative_hypoxia_3d_1.csv`
- `accessories/metadata_231_timecourse.csv`

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
```

---

## Section 1: Load data and join metadata

```{r load_data}
meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE) %>%
  dplyr::filter(type == "RiboSeq", hours == 1)

pos_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                      show_col_types = FALSE)
neg_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                      show_col_types = FALSE)

cat("Positive set mRNAs:", nrow(pos_genes), "\n")
cat("Negative set mRNAs:", nrow(neg_genes), "\n")

epa_pos <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_positive_hypoxia_3d_1.csv"),
  show_col_types = FALSE
) %>% mutate(gene_set = "eIF3d promotes")

epa_neg <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_matched_negative_hypoxia_3d_1.csv"),
  show_col_types = FALSE
) %>% mutate(gene_set = "Negative control")

cat("1hr RiboSeq samples in metadata:", nrow(meta), "\n")
```

```{r join_metadata}
join_meta <- function(df) {
  df %>%
    mutate(
      id    = str_remove(sample, "Aligned\\.toTranscriptome\\.out.*"),
      codon = str_replace_all(codon, "U", "T")
    ) %>%
    left_join(meta, by = "id")
}

epa_df <- bind_rows(join_meta(epa_pos), join_meta(epa_neg)) %>%
  mutate(gene_set = factor(gene_set, levels = c("eIF3d promotes", "Negative control")))

stopifnot("All rows have treatment metadata" = !any(is.na(epa_df$treatment)))
stopifnot("Positive set: 18 samples" =
  n_distinct(epa_df$id[epa_df$gene_set == "eIF3d promotes"]) == 18)
stopifnot("Negative set: 18 samples" =
  n_distinct(epa_df$id[epa_df$gene_set == "Negative control"]) == 18)

cat("\nSamples per gene_set × treatment × condition:\n")
epa_df %>%
  distinct(gene_set, id, treatment, condition) %>%
  count(gene_set, treatment, condition) %>%
  print()
```

---

## Section 2: Normalize to relative frequencies

```{r normalize}
codon_freq <- epa_df %>%
  group_by(gene_set, id, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup()

stopifnot("Frequencies in [0,1]" =
  all(codon_freq$freq >= 0 & codon_freq$freq <= 1, na.rm = TRUE))

codon_means <- codon_freq %>%
  group_by(gene_set, codon, site, condition, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE),
            sd_freq   = sd(freq, na.rm = TRUE),
            .groups = "drop")

cat("Codons per site × gene_set:\n")
codon_means %>% distinct(gene_set, codon, site) %>% count(gene_set, site) %>% print()
```

---

## Section 3: Validation

```{r validation}
genetic_code <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

gc3_class_map <- tibble(
  codon     = names(genetic_code),
  gc3_class = ifelse(str_sub(names(genetic_code), 3, 3) %in% c("G", "C"),
                     "GC-ending", "AT-ending")
)

codon_means <- codon_means %>%
  mutate(aa = genetic_code[codon])

for (gs in levels(codon_means$gene_set)) {
  psite_order <- codon_means %>%
    dplyr::filter(gene_set == gs, site == "psite",
                  treatment == "siCTRL", condition == "normoxia") %>%
    arrange(desc(mean_freq)) %>%
    pull(codon)

  atg_rank <- which(psite_order == "ATG")
  cat(gs, "— ATG P-site rank:", atg_rank, "of", length(psite_order), "\n")
  stopifnot("ATG in top 20 P-site codons" = atg_rank <= 20)

  stop_pfreq <- codon_means %>%
    dplyr::filter(gene_set == gs, codon %in% c("TAA", "TAG", "TGA"),
                  site == "psite", treatment == "siCTRL", condition == "normoxia") %>%
    pull(mean_freq)
  cat(gs, "— stop codon P-site frequencies:", round(stop_pfreq, 5), "\n")
  stopifnot("Stop codons rare at P-site (< 0.01)" = all(stop_pfreq < 0.01, na.rm = TRUE))
}
```

---

## Section 4: Helper objects

```{r helpers}
site_labels <- c(esite = "E site", psite = "P site", asite = "A site")
```

---

## Plot 0: siCTRL normoxia vs hypoxia codon frequency per site

Scatter of mean relative frequency in siCTRL normoxia (x) vs siCTRL hypoxia (y), faceted
by site (columns) and gene set (rows). Top 15% residuals from y=x are labelled per panel.

```{r plot_0, fig.width=11, fig.height=8}
plot_0_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = condition, values_from = mean_freq) %>%
  mutate(residual = hypoxia - normoxia,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0 <- max(plot_0_df$normoxia, plot_0_df$hypoxia, na.rm = TRUE)

p_0 <- ggplot(plot_0_df, aes(x = normoxia, y = hypoxia, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0)) +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12)
  ) +
  labs(
    title    = "Codon frequency: siCTRL normoxia vs hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL normoxia",
    y        = "siCTRL hypoxia",
    color    = "Amino acid"
  )

print(p_0)
ggsave(here("plots", "60_codon_freq_normoxia_vs_hypoxia_sictrl.pdf"), p_0,
       width = 11, height = 8)
```

---

## Plot 0b: siCTRL normoxia vs si3d normoxia codon frequency per site

```{r plot_0b, fig.width=11, fig.height=8}
plot_0b_df <- codon_means %>%
  dplyr::filter(condition == "normoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(residual = si3d - siCTRL,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0b <- max(plot_0b_df$siCTRL, plot_0b_df$si3d, na.rm = TRUE)

p_0b <- ggplot(plot_0b_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0b)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0b)) +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d normoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL normoxia",
    y        = "si3d normoxia",
    color    = "Amino acid"
  )

print(p_0b)
ggsave(here("plots", "60_codon_freq_sictrl_vs_si3d_normoxia.pdf"), p_0b,
       width = 11, height = 8)
```

---

## Plot 0c: siCTRL hypoxia vs si3d hypoxia codon frequency per site

```{r plot_0c, fig.width=11, fig.height=8}
plot_0c_df <- codon_means %>%
  dplyr::filter(condition == "hypoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(residual = si3d - siCTRL,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0c <- max(plot_0c_df$siCTRL, plot_0c_df$si3d, na.rm = TRUE)

p_0c <- ggplot(plot_0c_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0c)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0c)) +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL hypoxia",
    y        = "si3d hypoxia",
    color    = "Amino acid"
  )

print(p_0c)
ggsave(here("plots", "60_codon_freq_sictrl_vs_si3d_hypoxia.pdf"), p_0c,
       width = 11, height = 8)
```

---

## Section 5: Codon occupancy — eIF3d promotes vs Negative control (siCTRL normoxia)

```{r replicate_agreement}
# For each replicate, does eIF3d promotes have higher codon frequency than negative control?
# Pairs by sample id (same library, different transcript subsets).
replicate_agreement <- codon_freq %>%
  dplyr::filter(treatment == "siCTRL", condition == "normoxia") %>%
  dplyr::select(gene_set, rep, codon, site, freq) %>%
  pivot_wider(names_from = gene_set, values_from = freq) %>%
  dplyr::rename(promotes_freq = `eIF3d promotes`, neg_ctrl_freq = `Negative control`) %>%
  mutate(promotes_higher = promotes_freq > neg_ctrl_freq) %>%
  group_by(codon, site) %>%
  summarise(n_agree = sum(promotes_higher, na.rm = TRUE), .groups = "drop")
```

```{r plot_1_pos_vs_neg, fig.width=11, fig.height=5}
plot_1_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "normoxia",
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, mean_freq) %>%
  pivot_wider(names_from = gene_set, values_from = mean_freq) %>%
  dplyr::rename(
    promotes_mean = `eIF3d promotes`,
    neg_ctrl_mean = `Negative control`
  ) %>%
  left_join(replicate_agreement, by = c("codon", "site")) %>%
  left_join(gc3_class_map, by = "codon") %>%
  mutate(
    site     = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels),
    residual = abs(promotes_mean - neg_ctrl_mean),
    n_agree  = factor(n_agree, levels = 0:3, labels = c("0/3", "1/3", "2/3", "3/3"))
  ) %>%
  mutate(label = ifelse(residual >= quantile(residual, 0.90, na.rm = TRUE),
                        codon, NA_character_))

lim_1 <- max(plot_1_df$promotes_mean, plot_1_df$neg_ctrl_mean, na.rm = TRUE)

p_1 <- ggplot(plot_1_df,
              aes(x = neg_ctrl_mean, y = promotes_mean, color = gc3_class, shape = n_agree)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE, na.rm = TRUE) +
  facet_wrap(~ site, nrow = 1) +
  scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16)) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third position") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_1)) +
  theme_classic(base_size = 16) +
  theme(legend.position = "right",
        legend.title = element_text(size = 13),
        legend.text  = element_text(size = 12)) +
  labs(
    title    = "Codon occupancy: eIF3d promotes vs Negative control (siCTRL normoxia, 1hr)",
    subtitle = "Top 10% by effect size labelled; shape = replicates with promotes > neg ctrl",
    x        = "Negative control mean frequency",
    y        = "eIF3d promotes mean frequency",
    shape    = "Replicates\npromotes > ctrl"
  )

print(p_1)
ggsave(here("plots", "60_codon_freq_pos_vs_neg_sictrl_normoxia.pdf"),
       p_1, width = 11, height = 5)
```

---

## Section 6: Codon occupancy — eIF3d promotes vs Negative control (siCTRL hypoxia)

```{r replicate_agreement_hyp}
replicate_agreement_hyp <- codon_freq %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia") %>%
  dplyr::select(gene_set, rep, codon, site, freq) %>%
  pivot_wider(names_from = gene_set, values_from = freq) %>%
  dplyr::rename(promotes_freq = `eIF3d promotes`, neg_ctrl_freq = `Negative control`) %>%
  mutate(promotes_higher = promotes_freq > neg_ctrl_freq) %>%
  group_by(codon, site) %>%
  summarise(n_agree = sum(promotes_higher, na.rm = TRUE), .groups = "drop")
```

```{r plot_2_pos_vs_neg_hyp, fig.width=11, fig.height=5}
plot_2_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, mean_freq) %>%
  pivot_wider(names_from = gene_set, values_from = mean_freq) %>%
  dplyr::rename(
    promotes_mean = `eIF3d promotes`,
    neg_ctrl_mean = `Negative control`
  ) %>%
  left_join(replicate_agreement_hyp, by = c("codon", "site")) %>%
  left_join(gc3_class_map, by = "codon") %>%
  mutate(
    site     = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels),
    residual = abs(promotes_mean - neg_ctrl_mean),
    n_agree  = factor(n_agree, levels = 0:3, labels = c("0/3", "1/3", "2/3", "3/3"))
  ) %>%
  mutate(label = ifelse(residual >= quantile(residual, 0.90, na.rm = TRUE),
                        codon, NA_character_))

lim_2 <- max(plot_2_df$promotes_mean, plot_2_df$neg_ctrl_mean, na.rm = TRUE)

p_2 <- ggplot(plot_2_df,
              aes(x = neg_ctrl_mean, y = promotes_mean, color = gc3_class, shape = n_agree)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE, na.rm = TRUE) +
  facet_wrap(~ site, nrow = 1) +
  scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16)) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third position") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_2)) +
  theme_classic(base_size = 16) +
  theme(legend.position = "right",
        legend.title = element_text(size = 13),
        legend.text  = element_text(size = 12)) +
  labs(
    title    = "Codon occupancy: eIF3d promotes vs Negative control (siCTRL hypoxia, 1hr)",
    subtitle = "Top 10% by effect size labelled; shape = replicates with promotes > neg ctrl",
    x        = "Negative control mean frequency",
    y        = "eIF3d promotes mean frequency",
    shape    = "Replicates\npromotes > ctrl"
  )

print(p_2)
ggsave(here("plots", "60_codon_freq_pos_vs_neg_sictrl_hypoxia.pdf"),
       p_2, width = 11, height = 5)
```

---

## Section 7: Lollipop — delta codon occupancy (promotes minus neg ctrl)

```{r lollipop_data}
lollipop_agree <- codon_freq %>%
  dplyr::filter(treatment == "siCTRL") %>%
  dplyr::select(gene_set, rep, codon, site, condition, freq) %>%
  pivot_wider(names_from = gene_set, values_from = freq) %>%
  dplyr::rename(promotes_freq = `eIF3d promotes`, neg_ctrl_freq = `Negative control`) %>%
  mutate(promotes_higher = promotes_freq > neg_ctrl_freq) %>%
  group_by(codon, site, condition) %>%
  summarise(n_agree = sum(promotes_higher, na.rm = TRUE), .groups = "drop") %>%
  mutate(n_agree = factor(n_agree, levels = 0:3,
                          labels = c("0/3", "1/3", "2/3", "3/3")))

lollipop_base <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = gene_set, values_from = mean_freq) %>%
  dplyr::rename(promotes_mean = `eIF3d promotes`, neg_ctrl_mean = `Negative control`) %>%
  mutate(delta_freq = promotes_mean - neg_ctrl_mean) %>%
  left_join(gc3_class_map, by = "codon") %>%
  left_join(lollipop_agree, by = c("codon", "site", "condition")) %>%
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia"),
                       labels = c("Normoxia", "Hypoxia")),
    site      = factor(site, levels = c("esite", "psite", "asite"),
                       labels = site_labels)
  )

stopifnot(
  "GC3 class covers all codons" =
    all(lollipop_base$codon %in% gc3_class_map$codon),
  "delta_freq is finite" =
    all(is.finite(lollipop_base$delta_freq[!is.na(lollipop_base$delta_freq)]))
)
```

```{r plot_lollipop_ranked, fig.width=8, fig.height=12}
walk(levels(lollipop_base$site), function(s) {
  codon_order <- lollipop_base %>%
    dplyr::filter(site == s, condition == "Hypoxia") %>%
    arrange(delta_freq) %>%
    pull(codon)

  p <- lollipop_base %>%
    dplyr::filter(site == s) %>%
    mutate(codon = factor(codon, levels = codon_order)) %>%
    ggplot(aes(x = delta_freq, y = codon, colour = gc3_class, shape = n_agree)) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
    geom_segment(aes(x = 0, xend = delta_freq, yend = codon),
                 linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2) +
    facet_wrap(~ condition, nrow = 1) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16),
                       name = "Replicates\npromotes > ctrl") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_classic(base_size = 12) +
    theme(axis.text.y     = element_text(size = 9, family = "mono"),
          legend.position  = "top",
          legend.box       = "vertical") +
    labs(
      title    = paste0("Codon occupancy difference: eIF3d promotes vs Negative control (", s, ")"),
      subtitle = "Ranked by hypoxia delta for each site",
      x        = "Delta codon frequency (promotes minus neg ctrl)",
      y        = NULL
    )

  print(p)
  fname <- paste0("60_lollipop_delta_freq_ranked_", tolower(str_replace(s, " ", "")), ".pdf")
  ggsave(here("plots", fname), p, width = 8, height = 12)
})
```

```{r plot_lollipop_grouped, fig.width=8, fig.height=12}
walk(levels(lollipop_base$site), function(s) {
  codon_order <- lollipop_base %>%
    dplyr::filter(site == s, condition == "Hypoxia") %>%
    arrange(aa, delta_freq) %>%
    pull(codon)

  p <- lollipop_base %>%
    dplyr::filter(site == s) %>%
    mutate(codon = factor(codon, levels = codon_order)) %>%
    ggplot(aes(x = delta_freq, y = codon, colour = aa, shape = n_agree)) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
    geom_segment(aes(x = 0, xend = delta_freq, yend = codon),
                 linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2) +
    facet_wrap(~ condition, nrow = 1) +
    scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16),
                       name = "Replicates\npromotes > ctrl") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_classic(base_size = 12) +
    theme(axis.text.y     = element_text(size = 9, family = "mono"),
          legend.position  = "top",
          legend.box       = "vertical") +
    labs(
      title    = paste0("Codon occupancy difference: eIF3d promotes vs Negative control (", s, ")"),
      subtitle = "Grouped by amino acid, sorted by delta within each group",
      x        = "Delta codon frequency (promotes minus neg ctrl)",
      colour   = "Amino acid",
      y        = NULL
    )

  print(p)
  fname <- paste0("60_lollipop_delta_freq_grouped_by_aa_", tolower(str_replace(s, " ", "")), ".pdf")
  ggsave(here("plots", fname), p, width = 8, height = 12)
})
```

```{r plot_jitter_delta, fig.width=7, fig.height=5}
p_jitter <- ggplot(lollipop_base,
                   aes(x = site, y = delta_freq, colour = gc3_class)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_jitter(width = 0.2, height = 0, size = 2, alpha = 0.7) +
  stat_summary(fun = median, geom = "crossbar",
               width = 0.5, linewidth = 0.5, colour = "black", fatten = 1) +
  facet_grid(gc3_class ~ condition) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third position") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        strip.text = element_text(size = 11)) +
  labs(
    title    = "Codon occupancy difference: eIF3d promotes vs Negative control",
    subtitle = "Each point = one codon; black bar = median",
    x        = NULL,
    y        = "Delta codon frequency (promotes minus neg ctrl)"
  )

print(p_jitter)
ggsave(here("plots", "60_jitter_delta_freq_by_site.pdf"),
       p_jitter, width = 7, height = 5)
```

---

## Section 8: Lollipop — eIF3d knockdown effect on codon occupancy (siCTRL minus si3d)

```{r kd_lollipop_data}
kd_agree <- codon_freq %>%
  dplyr::filter(treatment %in% c("siCTRL", "si3d")) %>%
  dplyr::select(gene_set, rep, codon, site, condition, treatment, freq) %>%
  pivot_wider(names_from = treatment, values_from = freq) %>%
  mutate(sictrl_higher = siCTRL > si3d) %>%
  group_by(gene_set, codon, site, condition) %>%
  summarise(n_agree = sum(sictrl_higher, na.rm = TRUE), .groups = "drop") %>%
  mutate(n_agree = factor(n_agree, levels = 0:3,
                          labels = c("0/3", "1/3", "2/3", "3/3")))

kd_base <- codon_means %>%
  dplyr::filter(treatment %in% c("siCTRL", "si3d"), !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, condition, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  rename(sictrl_mean = siCTRL, si3d_mean = si3d) %>%
  mutate(delta_freq = sictrl_mean - si3d_mean) %>%
  left_join(gc3_class_map, by = "codon") %>%
  left_join(kd_agree, by = c("gene_set", "codon", "site", "condition")) %>%
  mutate(
    condition = factor(condition, levels = c("normoxia", "hypoxia"),
                       labels = c("Normoxia", "Hypoxia")),
    site      = factor(site, levels = c("esite", "psite", "asite"),
                       labels = site_labels),
    gene_set  = factor(gene_set, levels = c("eIF3d promotes", "Negative control"))
  )

stopifnot(
  "delta_freq is finite" =
    all(is.finite(kd_base$delta_freq[!is.na(kd_base$delta_freq)])),
  "both gene sets present" =
    n_distinct(kd_base$gene_set) == 2
)
```

```{r plot_kd_lollipop, fig.width=10, fig.height=14}
walk(levels(kd_base$site), function(s) {
  codon_order <- kd_base %>%
    dplyr::filter(site == s, condition == "Hypoxia",
                  gene_set == "eIF3d promotes") %>%
    arrange(delta_freq) %>%
    pull(codon)

  # Interleaved y-axis: neg ctrl below, promotes above within each codon pair
  y_levels <- as.vector(rbind(
    paste0(codon_order, "__neg"),
    paste0(codon_order, "__pos")
  ))
  # Show codon name only next to the promotes (top) row; blank for neg ctrl row
  y_labels <- ifelse(str_ends(y_levels, "__pos"),
                     str_remove(y_levels, "__pos"), "")

  p <- kd_base %>%
    dplyr::filter(site == s) %>%
    mutate(
      condition = factor(condition, levels = c("Hypoxia", "Normoxia")),
      y_var = paste0(codon, ifelse(gene_set == "eIF3d promotes", "__pos", "__neg")),
      y_var = factor(y_var, levels = y_levels)
    ) %>%
    ggplot(aes(x = delta_freq, y = y_var, colour = gene_set, shape = n_agree)) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
    geom_segment(aes(x = 0, xend = delta_freq, yend = y_var),
                 linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2) +
    facet_wrap(~ condition, nrow = 1) +
    scale_y_discrete(labels = y_labels) +
    scale_colour_manual(values = c("eIF3d promotes" = "#E69F00",
                                   "Negative control" = "grey40"),
                        name = "Gene set") +
    scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16),
                       name = "Replicates\nsiCTRL > si3d") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_classic(base_size = 14) +
    theme(axis.text.y     = element_text(size = 11, family = "mono"),
          legend.position  = "top",
          legend.box       = "vertical") +
    labs(
      title    = paste0("Codon occupancy: eIF3d knockdown effect (", s, ")"),
      subtitle = "Positive = eIF3d promotes occupancy at that codon (siCTRL minus si3d)",
      x        = "Delta codon frequency (siCTRL minus si3d)",
      y        = NULL
    )

  print(p)
  fname <- paste0("60_lollipop_kd_delta_", tolower(str_replace(s, " ", "")), ".pdf")
  ggsave(here("plots", fname), p, width = 10, height = 14)
})
```

---

## Section 9: Raw codon occupancy — siCTRL vs si3d, hypoxia only

```{r plot_kd_raw_freq, fig.width=10, fig.height=14}
walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  # Same codon order as Section 8 (hypoxia delta in eIF3d promotes)
  codon_order <- kd_base %>%
    dplyr::filter(site == s, condition == "Hypoxia",
                  gene_set == "eIF3d promotes") %>%
    arrange(delta_freq) %>%
    pull(codon)

  y_levels <- as.vector(rbind(
    paste0(codon_order, "__neg"),
    paste0(codon_order, "__pos")
  ))
  y_labels <- ifelse(str_ends(y_levels, "__pos"),
                     str_remove(y_levels, "__pos"), "")

  plot_df <- codon_means %>%
    dplyr::filter(site == site_key, condition == "hypoxia",
                  treatment %in% c("siCTRL", "si3d"),
                  !aa %in% c("Stop", NA)) %>%
    left_join(
      kd_agree %>% dplyr::filter(condition == "hypoxia"),
      by = c("gene_set", "codon", "site", "condition")
    ) %>%
    mutate(
      treatment = factor(treatment, levels = c("siCTRL", "si3d")),
      gene_set  = factor(gene_set, levels = c("eIF3d promotes", "Negative control")),
      y_var     = paste0(codon, ifelse(gene_set == "eIF3d promotes", "__pos", "__neg")),
      y_var     = factor(y_var, levels = y_levels)
    )

  p <- ggplot(plot_df, aes(x = mean_freq, y = y_var, colour = gene_set, shape = n_agree)) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
    geom_segment(aes(x = 0, xend = mean_freq, yend = y_var),
                 linewidth = 0.5, alpha = 0.7) +
    geom_point(size = 2) +
    facet_wrap(~ treatment, nrow = 1) +
    scale_y_discrete(labels = y_labels) +
    scale_colour_manual(values = c("eIF3d promotes" = "#E69F00",
                                   "Negative control" = "grey40"),
                        name = "Gene set") +
    scale_shape_manual(values = c("0/3" = 1, "1/3" = 2, "2/3" = 17, "3/3" = 16),
                       name = "Replicates\nsiCTRL > si3d") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_classic(base_size = 14) +
    theme(axis.text.y     = element_text(size = 11, family = "mono"),
          legend.position  = "top",
          legend.box       = "vertical") +
    labs(
      title    = paste0("Codon occupancy: siCTRL vs si3d, hypoxia (", s, ")"),
      subtitle = "Ordered by hypoxia delta (eIF3d promotes); promotes above, neg ctrl below per codon",
      x        = "Mean codon frequency",
      y        = NULL
    )

  print(p)
  fname <- paste0("60_lollipop_raw_freq_hypoxia_", site_key, ".pdf")
  ggsave(here("plots", fname), p, width = 10, height = 14)
})
```

---

## Section 10: Occupancy vs CDS codon frequency correlation (eIF3d promotes, siCTRL hypoxia)

Is high occupancy at a codon driven by that codon being abundant in these genes' CDS, or is it a genuine pausing signal? Codons near the diagonal are composition-explained; outliers off the diagonal are pausing candidates.

```{r cds_codon_freq}
feat_mat <- readRDS(here("output", "predictive_modeling",
                         "feature_matrix_codon_features.rds"))

# One transcript per gene, subset to promotes set
promotes_cds <- feat_mat %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  dplyr::filter(gene_id_clean %in% pos_genes$ensembl_gene) %>%
  dplyr::select(gene_id_clean, starts_with("codon_freq_"))

cat("Promotes transcripts matched in feature matrix:", nrow(promotes_cds),
    "of", nrow(pos_genes), "\n")

# Mean CDS frequency per codon across promotes transcripts -> long format
cds_freq_promotes <- promotes_cds %>%
  summarise(across(starts_with("codon_freq_"), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "codon", values_to = "cds_freq") %>%
  mutate(codon = str_remove(codon, "codon_freq_"))

stopifnot(
  "CDS frequencies sum to < 1 (ATG/TGG excluded)" =
    sum(cds_freq_promotes$cds_freq, na.rm = TRUE) < 1,
  "All CDS frequencies non-negative" =
    all(cds_freq_promotes$cds_freq >= 0, na.rm = TRUE)
)
cat("Codons in CDS frequency table:", nrow(cds_freq_promotes), "\n")
```

```{r plot_occ_vs_cds}
# Base data: all sites
# Stop codons excluded by !aa %in% c("Stop", NA); ATG/TGG dropped by !is.na(cds_freq)
corr_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                gene_set == "eIF3d promotes",
                !aa %in% c("Stop", NA)) %>%
  left_join(cds_freq_promotes, by = "codon") %>%
  left_join(gc3_class_map,     by = "codon") %>%
  dplyr::filter(!is.na(cds_freq))

walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  site_df <- corr_df %>%
    dplyr::filter(site == site_key) %>%
    mutate(
      residual = abs(mean_freq - cds_freq),
      label    = ifelse(residual >= quantile(residual, 0.85, na.rm = TRUE),
                        codon, NA_character_)
    )

  r_val      <- cor(site_df$cds_freq, site_df$mean_freq,
                    method = "spearman", use = "complete.obs")
  corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))
  lim_x      <- max(site_df$cds_freq,  na.rm = TRUE) * 1.05
  lim_y      <- max(site_df$mean_freq, na.rm = TRUE) * 1.05

  p <- ggplot(site_df, aes(x = cds_freq, y = mean_freq, colour = gc3_class)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                    show.legend = FALSE, na.rm = TRUE) +
    geom_text(data = corr_annot,
              aes(x = Inf, y = -Inf, label = label),
              hjust = 1.1, vjust = -0.5,
              parse = TRUE, inherit.aes = FALSE, size = 5, colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_x)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_y)) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    theme_classic(base_size = 16) +
    theme(legend.position = "right",
          legend.title = element_text(size = 13),
          legend.text  = element_text(size = 12)) +
    labs(
      title    = paste0("Occupancy vs CDS frequency - ", s,
                        " (eIF3d promotes, siCTRL hypoxia)"),
      subtitle = "Points on diagonal = composition-explained; outliers = pausing signal. Top 15% residuals labelled.",
      x        = "Mean CDS codon frequency (eIF3d promotes transcripts)",
      y        = paste0("Mean ribosome occupancy frequency (", s, ", siCTRL hypoxia)")
    )

  print(p)
  ggsave(here("plots",
              paste0("60_occupancy_vs_cds_freq_promotes_sictrl_hypoxia_", site_key, ".pdf")),
         p, width = 6, height = 6)
})
```

---

## Section 11: Occupancy vs CDS codon frequency correlation (Negative control, siCTRL hypoxia)

Same diagnostic as Section 10 but for the negative control gene set.

```{r cds_codon_freq_neg}
neg_gene_ids_clean <- sub("\\..*", "", neg_genes$gene_id)

neg_cds <- feat_mat %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  dplyr::filter(gene_id_clean %in% neg_gene_ids_clean) %>%
  dplyr::select(gene_id_clean, starts_with("codon_freq_"))

cat("Negative control transcripts matched in feature matrix:", nrow(neg_cds),
    "of", length(neg_gene_ids_clean), "\n")

cds_freq_neg <- neg_cds %>%
  summarise(across(starts_with("codon_freq_"), ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "codon", values_to = "cds_freq") %>%
  mutate(codon = str_remove(codon, "codon_freq_"))

stopifnot(
  "Neg CDS frequencies sum to < 1 (ATG/TGG excluded)" =
    sum(cds_freq_neg$cds_freq, na.rm = TRUE) < 1,
  "Neg CDS frequencies non-negative" =
    all(cds_freq_neg$cds_freq >= 0, na.rm = TRUE)
)
cat("Codons in neg CDS frequency table:", nrow(cds_freq_neg), "\n")
```

```{r plot_occ_vs_cds_neg}
corr_df_neg <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                gene_set == "Negative control",
                !aa %in% c("Stop", NA)) %>%
  left_join(cds_freq_neg, by = "codon") %>%
  left_join(gc3_class_map, by = "codon") %>%
  dplyr::filter(!is.na(cds_freq))

walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  site_df <- corr_df_neg %>%
    dplyr::filter(site == site_key) %>%
    mutate(
      residual = abs(mean_freq - cds_freq),
      label    = ifelse(residual >= quantile(residual, 0.85, na.rm = TRUE),
                        codon, NA_character_)
    )

  r_val      <- cor(site_df$cds_freq, site_df$mean_freq,
                    method = "spearman", use = "complete.obs")
  corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))
  lim_x      <- max(site_df$cds_freq,  na.rm = TRUE) * 1.05
  lim_y      <- max(site_df$mean_freq, na.rm = TRUE) * 1.05

  p <- ggplot(site_df, aes(x = cds_freq, y = mean_freq, colour = gc3_class)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                    show.legend = FALSE, na.rm = TRUE) +
    geom_text(data = corr_annot,
              aes(x = Inf, y = -Inf, label = label),
              hjust = 1.1, vjust = -0.5,
              parse = TRUE, inherit.aes = FALSE, size = 5, colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_x)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_y)) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    theme_classic(base_size = 16) +
    theme(legend.position = "right",
          legend.title = element_text(size = 13),
          legend.text  = element_text(size = 12)) +
    labs(
      title    = paste0("Occupancy vs CDS frequency - ", s,
                        " (Negative control, siCTRL hypoxia)"),
      subtitle = "Points on diagonal = composition-explained; outliers = pausing signal. Top 15% residuals labelled.",
      x        = "Mean CDS codon frequency (Negative control transcripts)",
      y        = paste0("Mean ribosome occupancy frequency (", s, ", siCTRL hypoxia)")
    )

  print(p)
  ggsave(here("plots",
              paste0("60_occupancy_vs_cds_freq_neg_sictrl_hypoxia_", site_key, ".pdf")),
         p, width = 6, height = 6)
})
```

---

## Section 12: Expression-weighted occupancy vs CDS codon frequency (siCTRL hypoxia) --- TAKE WITH GRAIN OF SALT AND REVISIT

Same as Sections 10–11 but CDS codon frequencies are weighted by each gene's siCTRL hypoxia 1hr
ribosome footprint count, making the x-axis a proper null expectation for occupancy under
no-pausing conditions.

```{r load_ribo_weights}
ribo_raw <- read_csv(here("counts", "raw_filtered_allCounts.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  mutate(total_ribo = hypoxia_sictrl_1_a_ribo_NM2025_0076 +
                      hypoxia_sictrl_1_b_ribo_NM2025_0077 +
                      hypoxia_sictrl_1_c_ribo_NM2025_0078) %>%
  dplyr::select(gene_id_clean, total_ribo)

cat("Genes in count matrix:", nrow(ribo_raw), "\n")

# Expression-weighted CDS freq helper
weighted_cds_freq <- function(gene_ids, label) {
  df <- feat_mat %>%
    distinct(gene_id_clean, .keep_all = TRUE) %>%
    dplyr::filter(gene_id_clean %in% gene_ids) %>%
    dplyr::select(gene_id_clean, starts_with("codon_freq_")) %>%
    left_join(ribo_raw, by = "gene_id_clean") %>%
    mutate(total_ribo = replace_na(total_ribo, 0),
           w = total_ribo / sum(total_ribo, na.rm = TRUE))
  cat(label, "— genes in feature matrix:", nrow(df),
      "| genes with ribo counts:", sum(df$total_ribo > 0), "\n")
  df %>%
    summarise(across(starts_with("codon_freq_"), ~ sum(.x * w, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "codon", values_to = "cds_freq") %>%
    mutate(codon = str_remove(codon, "codon_freq_"))
}

cds_freq_promotes_wt <- weighted_cds_freq(pos_genes$ensembl_gene,  "Promotes (weighted)")
cds_freq_neg_wt      <- weighted_cds_freq(neg_gene_ids_clean,       "Neg ctrl (weighted)")

stopifnot(
  "Promotes weighted CDS freq sums < 1" = sum(cds_freq_promotes_wt$cds_freq) < 1,
  "Neg ctrl weighted CDS freq sums < 1" = sum(cds_freq_neg_wt$cds_freq)      < 1
)
```

```{r plot_occ_vs_cds_promotes_weighted}
corr_df_pos_wt <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                gene_set == "eIF3d promotes",
                !aa %in% c("Stop", NA)) %>%
  left_join(cds_freq_promotes_wt, by = "codon") %>%
  left_join(gc3_class_map, by = "codon") %>%
  dplyr::filter(!is.na(cds_freq))

walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  site_df <- corr_df_pos_wt %>%
    dplyr::filter(site == site_key) %>%
    mutate(
      residual = abs(mean_freq - cds_freq),
      label    = ifelse(residual >= quantile(residual, 0.85, na.rm = TRUE),
                        codon, NA_character_)
    )

  r_val      <- cor(site_df$cds_freq, site_df$mean_freq,
                    method = "spearman", use = "complete.obs")
  corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))
  lim_x      <- max(site_df$cds_freq,  na.rm = TRUE) * 1.05
  lim_y      <- max(site_df$mean_freq, na.rm = TRUE) * 1.05

  p <- ggplot(site_df, aes(x = cds_freq, y = mean_freq, colour = gc3_class)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                    show.legend = FALSE, na.rm = TRUE) +
    geom_text(data = corr_annot,
              aes(x = Inf, y = -Inf, label = label),
              hjust = 1.1, vjust = -0.5,
              parse = TRUE, inherit.aes = FALSE, size = 5, colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_x)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_y)) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    theme_classic(base_size = 16) +
    theme(legend.position = "right",
          legend.title = element_text(size = 13),
          legend.text  = element_text(size = 12)) +
    labs(
      title    = paste0("Occupancy vs expression-weighted CDS frequency - ", s,
                        " (eIF3d promotes, siCTRL hypoxia)"),
      subtitle = "CDS frequency weighted by siCTRL hypoxia 1hr ribosome footprint counts. Top 15% residuals labelled.",
      x        = "Expression-weighted CDS codon frequency (eIF3d promotes)",
      y        = paste0("Mean ribosome occupancy frequency (", s, ", siCTRL hypoxia)")
    )

  print(p)
  ggsave(here("plots",
              paste0("60_occupancy_vs_cds_freq_weighted_promotes_sictrl_hypoxia_", site_key, ".pdf")),
         p, width = 6, height = 6)
})
```

```{r plot_occ_vs_cds_neg_weighted}
corr_df_neg_wt <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                gene_set == "Negative control",
                !aa %in% c("Stop", NA)) %>%
  left_join(cds_freq_neg_wt, by = "codon") %>%
  left_join(gc3_class_map, by = "codon") %>%
  dplyr::filter(!is.na(cds_freq))

walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  site_df <- corr_df_neg_wt %>%
    dplyr::filter(site == site_key) %>%
    mutate(
      residual = abs(mean_freq - cds_freq),
      label    = ifelse(residual >= quantile(residual, 0.85, na.rm = TRUE),
                        codon, NA_character_)
    )

  r_val      <- cor(site_df$cds_freq, site_df$mean_freq,
                    method = "spearman", use = "complete.obs")
  corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))
  lim_x      <- max(site_df$cds_freq,  na.rm = TRUE) * 1.05
  lim_y      <- max(site_df$mean_freq, na.rm = TRUE) * 1.05

  p <- ggplot(site_df, aes(x = cds_freq, y = mean_freq, colour = gc3_class)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                    show.legend = FALSE, na.rm = TRUE) +
    geom_text(data = corr_annot,
              aes(x = Inf, y = -Inf, label = label),
              hjust = 1.1, vjust = -0.5,
              parse = TRUE, inherit.aes = FALSE, size = 5, colour = "grey30") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_x)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                       limits = c(0, lim_y)) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    theme_classic(base_size = 16) +
    theme(legend.position = "right",
          legend.title = element_text(size = 13),
          legend.text  = element_text(size = 12)) +
    labs(
      title    = paste0("Occupancy vs expression-weighted CDS frequency - ", s,
                        " (Negative control, siCTRL hypoxia)"),
      subtitle = "CDS frequency weighted by siCTRL hypoxia 1hr ribosome footprint counts. Top 15% residuals labelled.",
      x        = "Expression-weighted CDS codon frequency (Negative control)",
      y        = paste0("Mean ribosome occupancy frequency (", s, ", siCTRL hypoxia)")
    )

  print(p)
  ggsave(here("plots",
              paste0("60_occupancy_vs_cds_freq_weighted_neg_sictrl_hypoxia_", site_key, ".pdf")),
         p, width = 6, height = 6)
})
```

---

## Section 13: bg-RSCU vs CDS codon frequency and ribosome occupancy (eIF3d promotes, hypoxia)

Checks whether the codon bias seen in the RSCU analyses (GC-ending enrichment in promotes
genes) matches the CDS composition and ribosome occupancy patterns in this notebook.

```{r load_bgrscu}
bgrscu_pooled <- read_csv(
  here("output", "predictive_modeling", "bg_rscu_pooled_231_hyp_vs_nor.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(codon, rscu_hyp, gc3_class)

cat("Pooled bg-RSCU codons:", nrow(bgrscu_pooled), "(expect 59)\n")
stopifnot("bg-RSCU codons match CDS freq codons" =
  all(bgrscu_pooled$codon %in% cds_freq_promotes$codon))
```

```{r plot_bgrscu_vs_cds}
bgrscu_cds_df <- bgrscu_pooled %>%
  left_join(cds_freq_promotes, by = "codon") %>%
  mutate(
    residual = abs(rscu_hyp - mean(rscu_hyp)),
    label    = ifelse(residual >= quantile(residual, 0.85), codon, NA_character_)
  )

r_val      <- cor(bgrscu_cds_df$cds_freq, bgrscu_cds_df$rscu_hyp,
                  method = "spearman", use = "complete.obs")
corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))

p_bgrscu_cds <- ggplot(bgrscu_cds_df, aes(x = cds_freq, y = rscu_hyp, colour = gc3_class)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                  show.legend = FALSE, na.rm = TRUE) +
  geom_text(data = corr_annot, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.1, vjust = -0.5, parse = TRUE,
            inherit.aes = FALSE, size = 5, colour = "grey30") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                      name = "Third position") +
  theme_classic(base_size = 16) +
  theme(legend.position = "right",
        legend.title = element_text(size = 13),
        legend.text  = element_text(size = 12)) +
  labs(
    title    = "Pooled bg-RSCU vs CDS codon frequency (eIF3d promotes, hypoxia)",
    subtitle = "Dashed line: bg-RSCU = 1 (no bias). Top 15% residuals labelled.",
    x        = "Mean CDS codon frequency (eIF3d promotes)",
    y        = "Pooled bg-RSCU (hypoxia promotes)"
  )
print(p_bgrscu_cds)
ggsave(here("plots", "60_bgrscu_vs_cds_freq_promotes_hypoxia.pdf"),
       p_bgrscu_cds, width = 6, height = 6)
```

```{r plot_bgrscu_vs_occupancy}
bgrscu_occ_base <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", condition == "hypoxia",
                gene_set == "eIF3d promotes", !aa %in% c("Stop", NA)) %>%
  left_join(bgrscu_pooled, by = "codon") %>%
  dplyr::filter(!is.na(rscu_hyp))

walk(c("esite", "psite", "asite"), function(site_key) {
  s <- site_labels[site_key]

  site_df <- bgrscu_occ_base %>%
    dplyr::filter(site == site_key) %>%
    mutate(
      residual = abs(mean_freq - median(mean_freq)),
      label    = ifelse(residual >= quantile(residual, 0.85), codon, NA_character_)
    )

  r_val      <- cor(site_df$rscu_hyp, site_df$mean_freq,
                    method = "spearman", use = "complete.obs")
  corr_annot <- tibble(label = paste0("rho == ", round(r_val, 2)))

  p <- ggplot(site_df, aes(x = rscu_hyp, y = mean_freq, colour = gc3_class)) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50", linewidth = 0.6) +
    geom_point(size = 1.8, alpha = 0.8) +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25,
                    show.legend = FALSE, na.rm = TRUE) +
    geom_text(data = corr_annot, aes(x = Inf, y = -Inf, label = label),
              hjust = 1.1, vjust = -0.5, parse = TRUE,
              inherit.aes = FALSE, size = 5, colour = "grey30") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    scale_colour_manual(values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
                        name = "Third position") +
    theme_classic(base_size = 16) +
    theme(legend.position = "right",
          legend.title = element_text(size = 13),
          legend.text  = element_text(size = 12)) +
    labs(
      title    = paste0("Pooled bg-RSCU vs occupancy - ", s,
                        " (eIF3d promotes, siCTRL hypoxia)"),
      subtitle = "Dashed line: bg-RSCU = 1. Top 15% residuals labelled.",
      x        = "Pooled bg-RSCU (hypoxia promotes)",
      y        = paste0("Mean ribosome occupancy frequency (", s, ")")
    )
  print(p)
  ggsave(here("plots",
              paste0("60_bgrscu_vs_occupancy_promotes_hypoxia_", site_key, ".pdf")),
         p, width = 6, height = 6)
})
```

---

## Section 14: GAA codon frequency across gene sets

```{r gaa_freq_groups}
gaa_hyp_rscu <- bgrscu_pooled %>%
  dplyr::filter(codon == "GAA") %>%
  pull(rscu_hyp)

gaa_df <- feat_mat %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  dplyr::select(gene_id_clean, codon_freq_GAA) %>%
  mutate(
    group = case_when(
      gene_id_clean %in% pos_genes$ensembl_gene ~ "eIF3d promotes",
      gene_id_clean %in% neg_gene_ids_clean      ~ "Negative control",
      TRUE                                        ~ "All expressed"
    )
  )

cat("GAA bg-RSCU (hypoxia promotes):", round(gaa_hyp_rscu, 3), "\n")
cat("Group sizes:\n")
print(table(gaa_df$group))
```

```{r plot_gaa_freq}
promotes_gaa <- gaa_df %>% dplyr::filter(group == "eIF3d promotes")   %>% pull(codon_freq_GAA)
neg_gaa      <- gaa_df %>% dplyr::filter(group == "Negative control") %>% pull(codon_freq_GAA)

wt      <- wilcox.test(promotes_gaa, neg_gaa)
p_label <- ifelse(wt$p.value < 0.001, "p < 0.001",
                  paste0("p = ", signif(wt$p.value, 2)))

gaa_density_df <- gaa_df %>%
  mutate(group = factor(group, levels = c("eIF3d promotes", "Negative control", "All expressed")))

median_df <- gaa_density_df %>%
  group_by(group) %>%
  summarise(med = median(codon_freq_GAA, na.rm = TRUE), .groups = "drop")

p_gaa_density <- ggplot(gaa_density_df, aes(x = codon_freq_GAA, colour = group, fill = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(data = median_df, aes(xintercept = med, colour = group),
             linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_colour_manual(values = c("eIF3d promotes"  = "#E69F00",
                                 "Negative control" = "grey40",
                                 "All expressed"    = "grey80"),
                      name = NULL) +
  scale_fill_manual(values = c("eIF3d promotes"  = "#E69F00",
                               "Negative control" = "grey40",
                               "All expressed"    = "grey80"),
                    name = NULL) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = paste0("Wilcoxon promotes vs neg ctrl\n", p_label),
           size = 4.5, colour = "grey30") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 3.8,
           label = paste0("GAA bg-RSCU (hypoxia) = ", round(gaa_hyp_rscu, 3)),
           size = 4.5, colour = "grey30") +
  theme_classic(base_size = 16) +
  theme(legend.position = "top",
        legend.text = element_text(size = 12)) +
  labs(
    title    = "GAA codon frequency by gene set (MDA-MB-231 hypoxia)",
    subtitle = "Dashed lines: group medians. GAA bg-RSCU > 1 = enriched vs transcriptome.",
    x        = "GAA codon frequency per gene",
    y        = "Density"
  )
print(p_gaa_density)
ggsave(here("plots", "60_gaa_codon_freq_by_gene_set.pdf"),
       p_gaa_density, width = 6, height = 5)

gaa_pair_df <- gaa_df %>%
  dplyr::filter(group %in% c("eIF3d promotes", "Negative control")) %>%
  mutate(group = factor(group, levels = c("eIF3d promotes", "Negative control")))

p_gaa_box <- ggplot(gaa_pair_df, aes(x = group, y = codon_freq_GAA, colour = group, fill = group)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA, linewidth = 0.8) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_colour_manual(values = c("eIF3d promotes" = "#E69F00", "Negative control" = "grey40"),
                      guide = "none") +
  scale_fill_manual(values = c("eIF3d promotes" = "#E69F00", "Negative control" = "grey40"),
                    guide = "none") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
           label = p_label, size = 5, colour = "grey30") +
  theme_classic(base_size = 16) +
  labs(
    title = "GAA codon frequency: eIF3d promotes vs negative control",
    x     = NULL,
    y     = "GAA codon frequency per gene"
  )
print(p_gaa_box)
ggsave(here("plots", "60_gaa_codon_freq_promotes_vs_neg.pdf"),
       p_gaa_box, width = 5, height = 5)
```

---

## Session info

```{r session_info}
sessionInfo()
```
