---
title: "Notebook 60: riboWaltz E/P/A Site Codon Occupancy — Positive vs Negative Gene Sets (MDA-MB-231, 1hr)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Replicates the Plot 0 scatter plots from Notebook 58, comparing codon frequency at E/P/A
sites between two transcript subsets:

- **eIF3d promotes** — positive gene set (hypoxia 1hr eIF3d promotes translation)
- **Negative control** — matched negative control transcripts

Inputs:
- `ribowaltz_output/timecourse_231/epa_codon_occupancy_positive_hypoxia_3d_1.csv`
- `ribowaltz_output/timecourse_231/epa_codon_occupancy_matched_negative_hypoxia_3d_1.csv`
- `accessories/metadata_231_timecourse.csv`

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
```

---

## Section 1: Load data and join metadata

```{r load_data}
meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE) %>%
  dplyr::filter(type == "RiboSeq", hours == 1)

pos_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                      show_col_types = FALSE)
neg_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                      show_col_types = FALSE)

cat("Positive set mRNAs:", nrow(pos_genes), "\n")
cat("Negative set mRNAs:", nrow(neg_genes), "\n")

epa_pos <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_positive_hypoxia_3d_1.csv"),
  show_col_types = FALSE
) %>% mutate(gene_set = "eIF3d promotes")

epa_neg <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_matched_negative_hypoxia_3d_1.csv"),
  show_col_types = FALSE
) %>% mutate(gene_set = "Negative control")

cat("1hr RiboSeq samples in metadata:", nrow(meta), "\n")
```

```{r join_metadata}
join_meta <- function(df) {
  df %>%
    mutate(
      id    = str_remove(sample, "Aligned\\.toTranscriptome\\.out.*"),
      codon = str_replace_all(codon, "U", "T")
    ) %>%
    left_join(meta, by = "id")
}

epa_df <- bind_rows(join_meta(epa_pos), join_meta(epa_neg)) %>%
  mutate(gene_set = factor(gene_set, levels = c("eIF3d promotes", "Negative control")))

stopifnot("All rows have treatment metadata" = !any(is.na(epa_df$treatment)))
stopifnot("Positive set: 18 samples" =
  n_distinct(epa_df$id[epa_df$gene_set == "eIF3d promotes"]) == 18)
stopifnot("Negative set: 18 samples" =
  n_distinct(epa_df$id[epa_df$gene_set == "Negative control"]) == 18)

cat("\nSamples per gene_set × treatment × condition:\n")
epa_df %>%
  distinct(gene_set, id, treatment, condition) %>%
  count(gene_set, treatment, condition) %>%
  print()
```

---

## Section 2: Normalize to relative frequencies

```{r normalize}
codon_freq <- epa_df %>%
  group_by(gene_set, id, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup()

stopifnot("Frequencies in [0,1]" =
  all(codon_freq$freq >= 0 & codon_freq$freq <= 1, na.rm = TRUE))

codon_means <- codon_freq %>%
  group_by(gene_set, codon, site, condition, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE),
            sd_freq   = sd(freq, na.rm = TRUE),
            .groups = "drop")

cat("Codons per site × gene_set:\n")
codon_means %>% distinct(gene_set, codon, site) %>% count(gene_set, site) %>% print()
```

---

## Section 3: Validation

```{r validation}
genetic_code <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

codon_means <- codon_means %>%
  mutate(aa = genetic_code[codon])

for (gs in levels(codon_means$gene_set)) {
  psite_order <- codon_means %>%
    dplyr::filter(gene_set == gs, site == "psite",
                  treatment == "siCTRL", condition == "normoxia") %>%
    arrange(desc(mean_freq)) %>%
    pull(codon)

  atg_rank <- which(psite_order == "ATG")
  cat(gs, "— ATG P-site rank:", atg_rank, "of", length(psite_order), "\n")
  stopifnot("ATG in top 20 P-site codons" = atg_rank <= 20)

  stop_pfreq <- codon_means %>%
    dplyr::filter(gene_set == gs, codon %in% c("TAA", "TAG", "TGA"),
                  site == "psite", treatment == "siCTRL", condition == "normoxia") %>%
    pull(mean_freq)
  cat(gs, "— stop codon P-site frequencies:", round(stop_pfreq, 5), "\n")
  stopifnot("Stop codons rare at P-site (< 0.01)" = all(stop_pfreq < 0.01, na.rm = TRUE))
}
```

---

## Section 4: Helper objects

```{r helpers}
site_labels <- c(esite = "E site", psite = "P site", asite = "A site")
```

---

## Plot 0: siCTRL normoxia vs hypoxia codon frequency per site

Scatter of mean relative frequency in siCTRL normoxia (x) vs siCTRL hypoxia (y), faceted
by site (columns) and gene set (rows). Top 15% residuals from y=x are labelled per panel.

```{r plot_0, fig.width=11, fig.height=8}
plot_0_df <- codon_means %>%
  dplyr::filter(treatment == "siCTRL", !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, condition, mean_freq) %>%
  pivot_wider(names_from = condition, values_from = mean_freq) %>%
  mutate(residual = hypoxia - normoxia,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0 <- max(plot_0_df$normoxia, plot_0_df$hypoxia, na.rm = TRUE)

p_0 <- ggplot(plot_0_df, aes(x = normoxia, y = hypoxia, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL normoxia vs hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL normoxia",
    y        = "siCTRL hypoxia",
    color    = "Amino acid"
  )

print(p_0)
ggsave(here("plots", "60_codon_freq_normoxia_vs_hypoxia_sictrl.pdf"), p_0,
       width = 11, height = 8)
```

---

## Plot 0b: siCTRL normoxia vs si3d normoxia codon frequency per site

```{r plot_0b, fig.width=11, fig.height=8}
plot_0b_df <- codon_means %>%
  dplyr::filter(condition == "normoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(residual = si3d - siCTRL,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0b <- max(plot_0b_df$siCTRL, plot_0b_df$si3d, na.rm = TRUE)

p_0b <- ggplot(plot_0b_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0b)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0b)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d normoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL normoxia",
    y        = "si3d normoxia",
    color    = "Amino acid"
  )

print(p_0b)
ggsave(here("plots", "60_codon_freq_sictrl_vs_si3d_normoxia.pdf"), p_0b,
       width = 11, height = 8)
```

---

## Plot 0c: siCTRL hypoxia vs si3d hypoxia codon frequency per site

```{r plot_0c, fig.width=11, fig.height=8}
plot_0c_df <- codon_means %>%
  dplyr::filter(condition == "hypoxia", treatment %in% c("siCTRL", "si3d"),
                !aa %in% c("Stop", NA)) %>%
  dplyr::select(gene_set, codon, aa, site, treatment, mean_freq) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(residual = si3d - siCTRL,
         site = factor(site, levels = c("esite", "psite", "asite"), labels = site_labels)) %>%
  group_by(gene_set, site) %>%
  mutate(label = ifelse(abs(residual) >= quantile(abs(residual), 0.85, na.rm = TRUE),
                        codon, NA_character_)) %>%
  ungroup()

lim_0c <- max(plot_0c_df$siCTRL, plot_0c_df$si3d, na.rm = TRUE)

p_0c <- ggplot(plot_0c_df, aes(x = siCTRL, y = si3d, color = aa)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.6) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  facet_grid(gene_set ~ site) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0c)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), limits = c(0, lim_0c)) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 9),
    legend.text  = element_text(size = 8)
  ) +
  labs(
    title    = "Codon frequency: siCTRL vs si3d hypoxia, 1hr",
    subtitle = "Top 15% residuals from y=x labelled per panel",
    x        = "siCTRL hypoxia",
    y        = "si3d hypoxia",
    color    = "Amino acid"
  )

print(p_0c)
ggsave(here("plots", "60_codon_freq_sictrl_vs_si3d_hypoxia.pdf"), p_0c,
       width = 11, height = 8)
```

---

## Session info

```{r session_info}
sessionInfo()
```
