---
title: "Notebook 61: Glu codon bias (delta pooled RSCU) and A-site occupancy (MDA-MB-231)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Two-panel figure focused on Glutamic acid (Glu; E) codons GAA and GAG:

- **Panel A**: Delta pooled RSCU for Glu — hypoxia-promoted minus normoxia-promoted
  (MDA-MB-231). Reads pre-computed results from notebook 54.
- **Panel B**: A-site codon occupancy scatter — siCTRL vs si3d hypoxia, 1hr, full
  transcriptome. GAA highlighted in orange; all other codons in grey.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
library(patchwork)
```

---

# Section 1: Glu delta pooled RSCU (Panel A)

Reads the nb54 output CSV — no RSCU recomputation.

```{r load_rscu}
rscu_all <- read_csv(
  here("output", "predictive_modeling", "bg_rscu_pooled_231_hyp_vs_nor.csv"),
  show_col_types = FALSE
)

glu_df <- rscu_all %>%
  dplyr::filter(amino_acid == "E") %>%
  mutate(
    gc3_class = factor(gc3_class, levels = c("AT-ending", "GC-ending")),
    sig_label = if_else(!is.na(significant) & significant, "Significant (BH q < 0.05)", "Not significant")
  )

cat("Glu codons:\n")
print(glu_df %>% dplyr::select(codon, gc3_class, rscu_hyp, rscu_nor,
                                delta_rscu, delta_ci_lo, delta_ci_hi,
                                p_adj_chi, significant))
```

```{r validate_glu}
stopifnot(
  "Both Glu codons present" = all(c("GAA", "GAG") %in% glu_df$codon),
  "delta_rscu is finite"    = all(is.finite(glu_df$delta_rscu)),
  "CI columns present"      = all(c("delta_ci_lo", "delta_ci_hi") %in% colnames(glu_df))
)
cat("GAA delta pooled RSCU:", round(glu_df$delta_rscu[glu_df$codon == "GAA"], 4), "\n")
cat("GAG delta pooled RSCU:", round(glu_df$delta_rscu[glu_df$codon == "GAG"], 4), "\n")
```

```{r plot_glu_rscu, fig.width=3, fig.height=4}
p_a <- ggplot(glu_df, aes(x = codon, y = delta_rscu, colour = gc3_class)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_point(aes(shape = sig_label), size = 4) +
  scale_colour_manual(
    values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
    name   = "3rd position"
  ) +
  scale_shape_manual(
    values = c("Significant (BH q < 0.05)" = 17, "Not significant" = 16),
    name   = NULL
  ) +
  scale_x_discrete(limits = c("GAA", "GAG")) +
  labs(
    title    = "Glu codons",
    subtitle = "delta pooled RSCU\n(hypoxia minus normoxia promoted)",
    x        = NULL,
    y        = "delta pooled RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position   = "bottom",
    legend.box        = "vertical",
    legend.key.size   = unit(0.4, "cm"),
    legend.text       = element_text(size = 8),
    legend.title      = element_text(size = 8.5),
    legend.margin     = margin(0, 0, 0, 0),
    legend.box.margin = margin(-5, 0, 0, 0),
    plot.title        = element_text(size = 10, face = "bold"),
    plot.subtitle     = element_text(size = 8, colour = "grey40")
  )
p_a
ggsave(here("plots", "61a_glu_delta_pooled_rscu.pdf"), p_a, width = 3, height = 4)
```

---

# Section 1b: Glu raw pooled RSCU — hypoxia vs normoxia promoted (Panel A2)

Shows the underlying rscu_hyp and rscu_nor values from which Panel A's delta is derived.
Dashed line at 1 = same usage as transcriptome background.

```{r plot_glu_rscu_raw, fig.width=3.5, fig.height=4}
glu_long <- glu_df %>%
  dplyr::select(codon, gc3_class, rscu_hyp, rscu_nor) %>%
  pivot_longer(
    cols      = c(rscu_hyp, rscu_nor),
    names_to  = "group",
    values_to = "rscu"
  ) %>%
  mutate(
    group = recode(group,
                   rscu_hyp = "Hypoxia promoted",
                   rscu_nor = "Normoxia promoted"),
    group = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted"))
  )

p_a2 <- ggplot(glu_long,
               aes(x = codon, y = rscu, colour = group, shape = group)) +
  geom_hline(yintercept = 1, linewidth = 0.4, linetype = "dashed", colour = "grey60") +
  geom_point(position = position_dodge(width = 0.5), size = 3.5) +
  scale_colour_manual(
    values = c("Hypoxia promoted" = "#D55E00", "Normoxia promoted" = "#009E73"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Hypoxia promoted" = 17, "Normoxia promoted" = 16),
    name   = NULL
  ) +
  scale_x_discrete(limits = c("GAA", "GAG")) +
  labs(
    title    = "Glu codons",
    subtitle = "Pooled RSCU +/- 95% bootstrap CI\n(dashed = background level)",
    x        = NULL,
    y        = "Pooled RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position   = "bottom",
    legend.key.size   = unit(0.4, "cm"),
    legend.text       = element_text(size = 8),
    legend.margin     = margin(0, 0, 0, 0),
    plot.title        = element_text(size = 10, face = "bold"),
    plot.subtitle     = element_text(size = 8, colour = "grey40")
  )
p_a2
ggsave(here("plots", "61b_glu_pooled_rscu_hyp_vs_nor.pdf"), p_a2, width = 3.5, height = 4)
```

---

# Section 1c: Pooled Glu codon proportions bar chart

Recomputes raw pooled codon counts for exclusive hypoxia-only and normoxia-only gene sets
(lfc > 0.5; genes in both sets removed) plus the full expressed transcriptome as background.
Shows proportion of GAA vs GAG within pooled Glu codons — values sum to 100% per group.

```{r load_genomic_libs}
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicFeatures)
library(coRdon)
```

```{r build_codon_counts_mat}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))

fm     <- readRDS(here("output", "predictive_modeling",
                       "feature_matrix_external_stability.rds"))
tx_ids <- fm$transcript_id_clean

# Mirror nb54 exactly: extract all, then filter by version-stripped name
cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx <- cds_names_clean %in% tx_ids
cds_seqs <- cds_seqs_all[keep_idx]
names(cds_seqs) <- cds_names_clean[keep_idx]

# Valid CDS: length divisible by 3, non-zero
valid    <- width(cds_seqs) %% 3 == 0 & width(cds_seqs) > 0
cds_seqs <- cds_seqs[valid]

# Drop sequences with ambiguous bases (N) — coRdon validity check fails on these
has_n    <- Biostrings::vcountPattern("N", cds_seqs, fixed = TRUE) > 0
if (any(has_n)) cat("Dropping", sum(has_n), "sequences with ambiguous bases\n")
cds_seqs <- cds_seqs[!has_n]

codon_tbl  <- codonTable(cds_seqs)
counts_mat <- codonCounts(codon_tbl)
rownames(counts_mat) <- names(cds_seqs)

stopifnot("counts_mat has rows" = nrow(counts_mat) > 0)
cat("Expressed transcripts with valid CDS:", nrow(counts_mat), "\n")
```

```{r load_genesets_exclusive}
fm_ids <- fm %>% dplyr::distinct(transcript_id_clean, gene_id_clean)

hyp_genes <- read_csv(
  here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)$ensembl_gene

nor_genes <- read_csv(
  here("output", "genesets", "normoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
  show_col_types = FALSE
)$ensembl_gene

hyp_tx <- fm_ids %>% dplyr::filter(gene_id_clean %in% hyp_genes) %>%
  dplyr::pull(transcript_id_clean)
nor_tx <- fm_ids %>% dplyr::filter(gene_id_clean %in% nor_genes) %>%
  dplyr::pull(transcript_id_clean)

hyp_only_tx <- setdiff(hyp_tx, nor_tx)
nor_only_tx <- setdiff(nor_tx, hyp_tx)
bg_tx       <- rownames(counts_mat)

stopifnot(
  "Hyp-only and nor-only sets are exclusive" =
    length(intersect(hyp_only_tx, nor_only_tx)) == 0
)

cat("Hypoxia-only transcripts:", length(hyp_only_tx),
    "| Normoxia-only:", length(nor_only_tx),
    "| Background:", length(bg_tx), "\n")
cat("Overlap removed:", length(intersect(hyp_tx, nor_tx)), "transcripts\n")
```

```{r pool_glu_proportions}
pool_glu <- function(tx_subset) {
  keep   <- intersect(tx_subset, rownames(counts_mat))
  counts <- colSums(counts_mat[keep, c("GAA", "GAG"), drop = FALSE])
  tibble(codon = names(counts), count = as.integer(counts),
         proportion = counts / sum(counts) * 100)
}

prop_df <- bind_rows(
  pool_glu(hyp_only_tx) %>% mutate(group = "Hypoxia promoted"),
  pool_glu(nor_only_tx) %>% mutate(group = "Normoxia promoted"),
  pool_glu(bg_tx)       %>% mutate(group = "Background")
) %>%
  mutate(
    group     = factor(group, levels = c("Hypoxia promoted", "Normoxia promoted", "Background")),
    gc3_class = if_else(codon == "GAA", "AT-ending", "GC-ending"),
    gc3_class = factor(gc3_class, levels = c("AT-ending", "GC-ending"))
  )

stopifnot(
  "Proportions sum to 100 per group" =
    all(abs(prop_df %>% group_by(group) %>%
              summarise(s = sum(proportion)) %>% pull(s) - 100) < 1e-6),
  "Both codons in all groups" =
    all(c("GAA", "GAG") %in% prop_df$codon)
)

hyp_gaa <- prop_df %>% dplyr::filter(group == "Hypoxia promoted", codon == "GAA") %>%
  dplyr::pull(proportion)
bg_gaa  <- prop_df %>% dplyr::filter(group == "Background", codon == "GAA") %>%
  dplyr::pull(proportion)
stopifnot("Hypoxia promoted has higher GAA proportion than background" = hyp_gaa > bg_gaa)

cat("GAA proportions — Hyp:", round(hyp_gaa, 1),
    "| Nor:", round(prop_df %>% dplyr::filter(group == "Normoxia promoted", codon == "GAA") %>%
                     dplyr::pull(proportion), 1),
    "| Background:", round(bg_gaa, 1), "\n")
```

```{r plot_glu_codon_proportion, fig.width=5, fig.height=4}
p_a3 <- ggplot(prop_df, aes(x = codon, y = proportion, fill = group)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = round(proportion, 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3) +
  facet_wrap(~ gc3_class, scales = "free_x") +
  scale_fill_manual(
    values = c("Hypoxia promoted"  = "#D55E00",
               "Normoxia promoted" = "#009E73",
               "Background"        = "grey65"),
    name = NULL
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Synonymous codon usage: Glutamic acid",
    subtitle = "Proportion of each Glu codon within pooled Glu codons\n(exclusive gene sets, lfc > 0.5)",
    x        = NULL,
    y        = "Codon proportion (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "top",
    legend.text     = element_text(size = 9),
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(size = 11, face = "bold"),
    plot.subtitle   = element_text(size = 8, colour = "grey40")
  )

p_a3
ggsave(here("plots", "61f_glu_pooled_codon_proportion_bar.pdf"), p_a3, width = 5, height = 4)
cat("Saved to plots/61f_glu_pooled_codon_proportion_bar.pdf\n")
```

---

# Section 2: A-site codon occupancy — full transcriptome (Panel B)

Uses the full-transcriptome riboWaltz output (`epa_codon_occupancy_1hr.csv`).
Scatter: siCTRL hypoxia (x) vs si3d hypoxia (y) at the A site.
GAA is highlighted orange; all other codons are grey.

```{r load_metadata}
meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE) %>%
  dplyr::filter(type == "RiboSeq", hours == 1)

cat("1hr RiboSeq samples:", nrow(meta), "\n")
cat("Treatments present:", paste(unique(meta$treatment), collapse = ", "), "\n")
cat("Conditions present:", paste(unique(meta$condition), collapse = ", "), "\n")
```

```{r load_epa}
epa_raw <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_1hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    id    = str_remove(sample, "Aligned\\.toTranscriptome\\.out.*"),
    codon = str_replace_all(codon, "U", "T")
  ) %>%
  left_join(meta, by = "id")

stopifnot(
  "All rows have treatment metadata" = !any(is.na(epa_raw$treatment)),
  "A-site rows present"              = any(epa_raw$site == "asite"),
  "hypoxia rows present"             = any(epa_raw$condition == "hypoxia", na.rm = TRUE)
)

cat("Samples matched:", n_distinct(epa_raw$id), "\n")
cat("Sites present:", paste(unique(epa_raw$site), collapse = ", "), "\n")
```

```{r normalize_asite}
genetic_code_map <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

# Compute relative frequency per sample per site, then average across reps per treatment
asite_freq <- epa_raw %>%
  dplyr::filter(site == "asite",
                condition == "hypoxia",
                treatment %in% c("siCTRL", "si3d")) %>%
  group_by(id, treatment, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  group_by(codon, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE), .groups = "drop") %>%
  mutate(aa = genetic_code_map[codon])

stopifnot(
  "Frequencies in [0,1]" = all(asite_freq$mean_freq >= 0 & asite_freq$mean_freq <= 1),
  "Both treatments present" = all(c("siCTRL", "si3d") %in% asite_freq$treatment)
)

cat("Codons per treatment:", n_distinct(asite_freq$codon), "\n")
```

```{r asite_scatter_df}
asite_wide <- asite_freq %>%
  dplyr::filter(!aa %in% c("Stop", NA)) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(
    highlight = codon == "GAA",
    label     = if_else(highlight, codon, NA_character_)
  )

stopifnot(
  "GAA present in scatter data" = "GAA" %in% asite_wide$codon,
  "siCTRL and si3d columns exist" =
    all(c("siCTRL", "si3d") %in% colnames(asite_wide))
)

gaa_row <- asite_wide %>% dplyr::filter(codon == "GAA")
cat("GAA A-site — siCTRL hypoxia:", scales::percent(gaa_row$siCTRL, accuracy = 0.01),
    "| si3d hypoxia:", scales::percent(gaa_row$si3d, accuracy = 0.01), "\n")
```

```{r plot_asite_scatter, fig.width=5, fig.height=5}
lim <- max(asite_wide$siCTRL, asite_wide$si3d, na.rm = TRUE) * 1.05

p_b <- ggplot(asite_wide,
              aes(x = siCTRL, y = si3d,
                  colour = highlight, alpha = highlight)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60", linewidth = 0.5) +
  geom_point(size = 1.5) +
  geom_text_repel(aes(label = label), size = 3.5,
                  colour = "#E69F00", fontface = "bold",
                  nudge_x = 0.001, max.overlaps = 10,
                  show.legend = FALSE) +
  scale_colour_manual(values = c("FALSE" = "grey65", "TRUE" = "#E69F00"),
                      guide = "none") +
  scale_alpha_manual(values  = c("FALSE" = 0.55, "TRUE" = 1), guide = "none") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  labs(
    title    = "A-site codon occupancy",
    subtitle = "siCTRL vs si3d, hypoxia 1hr\nFull transcriptome",
    x        = "siCTRL hypoxia, relative frequency",
    y        = "si3d hypoxia, relative frequency"
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title    = element_text(size = 10, face = "bold"),
    plot.subtitle = element_text(size = 8, colour = "grey40")
  )
p_b
ggsave(here("plots", "61c_glu_asite_occupancy_sictrl_vs_si3d.pdf"), p_b, width = 5, height = 5)
```

```{r plot_asite_scatter_glu, fig.width=5, fig.height=5}
asite_glu <- asite_wide %>%
  mutate(
    glu_highlight = case_when(
      codon == "GAA" ~ "GAA",
      codon == "GAG" ~ "GAG",
      TRUE           ~ "other"
    ),
    glu_highlight = factor(glu_highlight, levels = c("GAA", "GAG", "other")),
    glu_label     = if_else(codon %in% c("GAA", "GAG"), codon, NA_character_)
  )

p_b2 <- ggplot(asite_glu,
               aes(x = siCTRL, y = si3d,
                   colour = glu_highlight, alpha = glu_highlight)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60", linewidth = 0.5) +
  geom_point(size = 1.5) +
  geom_text_repel(aes(label = glu_label), size = 3.5,
                  nudge_x = 0.001, max.overlaps = 10,
                  show.legend = FALSE) +
  scale_colour_manual(
    values = c("GAA" = "#E69F00", "GAG" = "black", "other" = "grey65"),
    guide  = "none"
  ) +
  scale_alpha_manual(
    values = c("GAA" = 1, "GAG" = 1, "other" = 0.55),
    guide  = "none"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  labs(
    title    = "A-site codon occupancy",
    subtitle = "siCTRL vs si3d, hypoxia 1hr\nFull transcriptome",
    x        = "siCTRL hypoxia, relative frequency",
    y        = "si3d hypoxia, relative frequency"
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title    = element_text(size = 10, face = "bold"),
    plot.subtitle = element_text(size = 8, colour = "grey40")
  )
p_b2
ggsave(here("plots", "61d_glu_asite_occupancy_gaa_gag_highlighted.pdf"),
       p_b2, width = 5, height = 5)
```

---

# Section 2b: Glu A-site frequency bar plot — GAA and GAG (Panel C)

Per-replicate A-site frequencies for GAA and GAG, hypoxia 1hr, siCTRL vs si3d.
Bars = mean; points = individual replicates.

```{r glu_asite_bars, fig.width=4, fig.height=4}
glu_asite_reps <- epa_raw %>%
  dplyr::filter(site == "asite",
                condition == "hypoxia",
                treatment %in% c("siCTRL", "si3d")) %>%
  group_by(id, treatment, rep) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  dplyr::filter(codon %in% c("GAA", "GAG")) %>%
  mutate(
    treatment = factor(treatment, levels = c("siCTRL", "si3d")),
    codon     = factor(codon, levels = c("GAA", "GAG"))
  )

stopifnot(
  "Both codons present"     = all(c("GAA", "GAG") %in% glu_asite_reps$codon),
  "Both treatments present" = all(c("siCTRL", "si3d") %in% glu_asite_reps$treatment),
  "Frequencies in [0,1]"   = all(glu_asite_reps$freq >= 0 & glu_asite_reps$freq <= 1)
)

glu_asite_means <- glu_asite_reps %>%
  group_by(codon, treatment) %>%
  summarise(mean_freq = mean(freq), .groups = "drop")

cat("GAA/GAG A-site means:\n")
print(glu_asite_means %>% mutate(mean_pct = scales::percent(mean_freq, accuracy = 0.01)))

p_c <- ggplot(glu_asite_means,
              aes(x = codon, y = mean_freq, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_point(data = glu_asite_reps,
             aes(x = codon, y = freq, fill = treatment),
             position = position_dodge(width = 0.7),
             shape = 21, size = 1.8, colour = "white", stroke = 0.4) +
  scale_fill_manual(
    values = c("siCTRL" = "grey60", "si3d" = "#E69F00"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title    = "Glu A-site occupancy",
    subtitle = "Hypoxia 1hr, full transcriptome",
    x        = NULL,
    y        = "A-site relative frequency"
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "top",
    legend.text     = element_text(size = 9),
    plot.title      = element_text(size = 10, face = "bold"),
    plot.subtitle   = element_text(size = 8, colour = "grey40")
  )
p_c
ggsave(here("plots", "61e_glu_asite_bar_sictrl_vs_si3d.pdf"), p_c, width = 4, height = 4)
```

---

# Section 3: Combined figure

```{r combined_figure, fig.width=11, fig.height=5}
p_combined <- p_a2 + p_b2 +
  plot_layout(widths = c(1, 2)) +
  plot_annotation(
    title = "Glutamic acid (Glu) — codon bias and A-site occupancy (MDA-MB-231)",
    theme = theme(plot.title = element_text(size = 12, face = "bold"))
  )

p_combined
ggsave(here("plots", "61_glu_rscu_and_asite_occupancy.pdf"),
       p_combined, width = 9, height = 5)
cat("Saved to plots/61_glu_rscu_and_asite_occupancy.pdf\n")
```

---

# Session info

```{r session_info}
sessionInfo()
```
