---
title: "Notebook 61: Glu codon bias (delta pooled RSCU) and A-site occupancy (MDA-MB-231)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Two-panel figure focused on Glutamic acid (Glu; E) codons GAA and GAG:

- **Panel A**: Delta pooled RSCU for Glu — hypoxia-promoted minus normoxia-promoted
  (MDA-MB-231). Reads pre-computed results from notebook 54.
- **Panel B**: A-site codon occupancy scatter — siCTRL vs si3d hypoxia, 1hr, full
  transcriptome. GAA highlighted in orange; all other codons in grey.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
library(patchwork)
```

---

# Section 1: Glu delta pooled RSCU (Panel A)

Reads the nb54 output CSV — no RSCU recomputation.

```{r load_rscu}
rscu_all <- read_csv(
  here("output", "predictive_modeling", "bg_rscu_pooled_231_hyp_vs_nor.csv"),
  show_col_types = FALSE
)

glu_df <- rscu_all %>%
  dplyr::filter(amino_acid == "E") %>%
  mutate(
    gc3_class = factor(gc3_class, levels = c("AT-ending", "GC-ending")),
    sig_label = if_else(!is.na(significant) & significant, "Significant (BH q < 0.05)", "Not significant")
  )

cat("Glu codons:\n")
print(glu_df %>% dplyr::select(codon, gc3_class, rscu_hyp, rscu_nor,
                                delta_rscu, delta_ci_lo, delta_ci_hi,
                                p_adj_chi, significant))
```

```{r validate_glu}
stopifnot(
  "Both Glu codons present" = all(c("GAA", "GAG") %in% glu_df$codon),
  "delta_rscu is finite"    = all(is.finite(glu_df$delta_rscu)),
  "CI columns present"      = all(c("delta_ci_lo", "delta_ci_hi") %in% colnames(glu_df))
)
cat("GAA delta pooled RSCU:", round(glu_df$delta_rscu[glu_df$codon == "GAA"], 4), "\n")
cat("GAG delta pooled RSCU:", round(glu_df$delta_rscu[glu_df$codon == "GAG"], 4), "\n")
```

```{r plot_glu_rscu, fig.width=3, fig.height=4}
p_a <- ggplot(glu_df, aes(x = codon, y = delta_rscu, colour = gc3_class)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey60") +
  geom_errorbar(aes(ymin = delta_ci_lo, ymax = delta_ci_hi),
                width = 0.12, linewidth = 0.7) +
  geom_point(aes(shape = sig_label), size = 4) +
  scale_colour_manual(
    values = c("AT-ending" = "#E69F00", "GC-ending" = "#0072B2"),
    name   = "3rd position"
  ) +
  scale_shape_manual(
    values = c("Significant (BH q < 0.05)" = 17, "Not significant" = 16),
    name   = NULL
  ) +
  scale_x_discrete(limits = c("GAA", "GAG")) +
  labs(
    title    = "Glu codons",
    subtitle = "delta pooled RSCU\n(hypoxia minus normoxia promoted)",
    x        = NULL,
    y        = "delta pooled RSCU"
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position   = "bottom",
    legend.box        = "vertical",
    legend.key.size   = unit(0.4, "cm"),
    legend.text       = element_text(size = 8),
    legend.title      = element_text(size = 8.5),
    legend.margin     = margin(0, 0, 0, 0),
    legend.box.margin = margin(-5, 0, 0, 0),
    plot.title        = element_text(size = 10, face = "bold"),
    plot.subtitle     = element_text(size = 8, colour = "grey40")
  )
p_a
```

---

# Section 2: A-site codon occupancy — full transcriptome (Panel B)

Uses the full-transcriptome riboWaltz output (`epa_codon_occupancy_1hr.csv`).
Scatter: siCTRL hypoxia (x) vs si3d hypoxia (y) at the A site.
GAA is highlighted orange; all other codons are grey.

```{r load_metadata}
meta <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                 show_col_types = FALSE) %>%
  dplyr::filter(type == "RiboSeq", hours == 1)

cat("1hr RiboSeq samples:", nrow(meta), "\n")
cat("Treatments present:", paste(unique(meta$treatment), collapse = ", "), "\n")
cat("Conditions present:", paste(unique(meta$condition), collapse = ", "), "\n")
```

```{r load_epa}
epa_raw <- read_csv(
  here("ribowaltz_output", "timecourse_231", "epa_codon_occupancy_1hr.csv"),
  show_col_types = FALSE
) %>%
  mutate(
    id    = str_remove(sample, "Aligned\\.toTranscriptome\\.out.*"),
    codon = str_replace_all(codon, "U", "T")
  ) %>%
  left_join(meta, by = "id")

stopifnot(
  "All rows have treatment metadata" = !any(is.na(epa_raw$treatment)),
  "A-site rows present"              = any(epa_raw$site == "asite"),
  "hypoxia rows present"             = any(epa_raw$condition == "hypoxia", na.rm = TRUE)
)

cat("Samples matched:", n_distinct(epa_raw$id), "\n")
cat("Sites present:", paste(unique(epa_raw$site), collapse = ", "), "\n")
```

```{r normalize_asite}
genetic_code_map <- c(
  TTT="Phe", TTC="Phe", TTA="Leu", TTG="Leu",
  CTT="Leu", CTC="Leu", CTA="Leu", CTG="Leu",
  ATT="Ile", ATC="Ile", ATA="Ile", ATG="Met",
  GTT="Val", GTC="Val", GTA="Val", GTG="Val",
  TCT="Ser", TCC="Ser", TCA="Ser", TCG="Ser",
  CCT="Pro", CCC="Pro", CCA="Pro", CCG="Pro",
  ACT="Thr", ACC="Thr", ACA="Thr", ACG="Thr",
  GCT="Ala", GCC="Ala", GCA="Ala", GCG="Ala",
  TAT="Tyr", TAC="Tyr", TAA="Stop", TAG="Stop",
  CAT="His", CAC="His", CAA="Gln", CAG="Gln",
  AAT="Asn", AAC="Asn", AAA="Lys", AAG="Lys",
  GAT="Asp", GAC="Asp", GAA="Glu", GAG="Glu",
  TGT="Cys", TGC="Cys", TGA="Stop", TGG="Trp",
  CGT="Arg", CGC="Arg", CGA="Arg", CGG="Arg",
  AGT="Ser", AGC="Ser", AGA="Arg", AGG="Arg",
  GGT="Gly", GGC="Gly", GGA="Gly", GGG="Gly"
)

# Compute relative frequency per sample per site, then average across reps per treatment
asite_freq <- epa_raw %>%
  dplyr::filter(site == "asite",
                condition == "hypoxia",
                treatment %in% c("siCTRL", "si3d")) %>%
  group_by(id, treatment, site) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  group_by(codon, treatment) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE), .groups = "drop") %>%
  mutate(aa = genetic_code_map[codon])

stopifnot(
  "Frequencies in [0,1]" = all(asite_freq$mean_freq >= 0 & asite_freq$mean_freq <= 1),
  "Both treatments present" = all(c("siCTRL", "si3d") %in% asite_freq$treatment)
)

cat("Codons per treatment:", n_distinct(asite_freq$codon), "\n")
```

```{r asite_scatter_df}
asite_wide <- asite_freq %>%
  dplyr::filter(!aa %in% c("Stop", NA)) %>%
  pivot_wider(names_from = treatment, values_from = mean_freq) %>%
  mutate(
    highlight = codon == "GAA",
    label     = if_else(highlight, codon, NA_character_)
  )

stopifnot(
  "GAA present in scatter data" = "GAA" %in% asite_wide$codon,
  "siCTRL and si3d columns exist" =
    all(c("siCTRL", "si3d") %in% colnames(asite_wide))
)

gaa_row <- asite_wide %>% dplyr::filter(codon == "GAA")
cat("GAA A-site — siCTRL hypoxia:", scales::percent(gaa_row$siCTRL, accuracy = 0.01),
    "| si3d hypoxia:", scales::percent(gaa_row$si3d, accuracy = 0.01), "\n")
```

```{r plot_asite_scatter, fig.width=5, fig.height=5}
lim <- max(asite_wide$siCTRL, asite_wide$si3d, na.rm = TRUE) * 1.05

p_b <- ggplot(asite_wide,
              aes(x = siCTRL, y = si3d,
                  colour = highlight, size = highlight, alpha = highlight)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey60", linewidth = 0.5) +
  geom_point() +
  geom_text_repel(aes(label = label), size = 3.5,
                  colour = "#E69F00", fontface = "bold",
                  nudge_x = 0.001, max.overlaps = 10,
                  show.legend = FALSE) +
  scale_colour_manual(values = c("FALSE" = "grey65", "TRUE" = "#E69F00"),
                      guide = "none") +
  scale_size_manual(values   = c("FALSE" = 1.5, "TRUE" = 3.5), guide = "none") +
  scale_alpha_manual(values  = c("FALSE" = 0.55, "TRUE" = 1), guide = "none") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     limits = c(0, lim)) +
  labs(
    title    = "A-site codon occupancy",
    subtitle = "siCTRL vs si3d, hypoxia 1hr\nFull transcriptome",
    x        = "siCTRL hypoxia — relative frequency",
    y        = "si3d hypoxia — relative frequency"
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title    = element_text(size = 10, face = "bold"),
    plot.subtitle = element_text(size = 8, colour = "grey40")
  )
p_b
```

---

# Section 3: Combined figure

```{r combined_figure, fig.width=8, fig.height=5}
p_combined <- p_a + p_b +
  plot_layout(widths = c(1, 2)) +
  plot_annotation(
    title = "Glutamic acid (Glu) — codon bias and A-site occupancy (MDA-MB-231)",
    theme = theme(plot.title = element_text(size = 12, face = "bold"))
  )

p_combined
ggsave(here("plots", "61_glu_rscu_and_asite_occupancy.pdf"),
       p_combined, width = 8, height = 5)
cat("Saved to plots/61_glu_rscu_and_asite_occupancy.pdf\n")
```

---

# Session info

```{r session_info}
sessionInfo()
```
