---
title: "63 - CDS length and GC content: extended negative controls"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Characterize the CDS length and GC content of the **extended negative control** gene set:
genes not regulated by eIF3d OR eIF3e in MDA-MB-231 OR MCF7-SIX1 cells (any condition).

Filtering proceeds in three steps:

1. **MDA-MB-231 si3d baseline** — load `negative_control_genes.csv` from nb02
   (|te_lfc| < 0.5 AND te_padj > 0.3 across normoxia alltimepoints + hypoxia 1hr/4hr/24hr)
2. **MDA-MB-231 si3e** — apply the same criteria to the four si3e comparison files;
   intersect with step 1
3. **MCF7-SIX1 si3d + si3e** — exclude any gene appearing in MCF7-SIX1 si3d/si3e
   regulated gene sets from `all_combined_sigs_1_28_26.csv`

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
```

# Define Extended Negative Controls

## Step 1: MDA-MB-231 si3d baseline (nb02 output)

```{r step1_si3d}
neg_ctrl_si3d <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
)

stopifnot("negative_control_genes.csv is empty" = nrow(neg_ctrl_si3d) > 0)
stopifnot("gene_id column missing" = "gene_id" %in% colnames(neg_ctrl_si3d))

neg_ctrl_si3d <- neg_ctrl_si3d %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("Step 1 - MDA-MB-231 si3d baseline:", nrow(neg_ctrl_si3d), "genes\n")
```

## Step 1b: MDA-MB-231 si3d RNA + ribo filter

Require rna_padj > 0.05 and ribo_padj > 0.05 at hypoxia 1hr and normoxia only
(no LFC cutoff — significance filter only). NA padj values are dropped.
Note: MCF7-SIX1 si3d files contain only TE results — no RNA/ribo level
data is available for that cell line.

```{r step1b_si3d_rna_ribo}
si3d_files <- tibble(
  label = c("si3d_nor_all", "si3d_hyp_1hr", "si3d_hyp_4hr", "si3d_hyp_24hr"),
  file  = here("output", c(
    "translation_categories_si3d_vs_sictrl_normoxia_alltimepoints.csv",
    "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
    "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv",
    "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"
  ))
)

si3d_full_data <- si3d_files %>%
  rowwise() %>%
  mutate(data = list(read_csv(file, show_col_types = FALSE))) %>%
  ungroup()

# si3d_rna_ribo_passing <- si3d_full_data %>%
#   filter(label %in% c("si3d_nor_all", "si3d_hyp_1hr")) %>%
#   mutate(
#     passing = map(data, function(df) {
#       df %>%
#         filter(!is.na(rna_padj), !is.na(ribo_padj),
#                rna_padj  > 0.05,
#                ribo_padj > 0.05) %>%
#         pull(gene_id)
#     })
#   )
#
# cat("Genes passing si3d RNA + ribo criteria per comparison:\n")
# si3d_rna_ribo_passing %>%
#   mutate(n_passing = map_int(passing, length)) %>%
#   dplyr::select(label, n_passing) %>%
#   print(n = 2)
#
# si3d_rna_ribo_ids <- Reduce(intersect, si3d_rna_ribo_passing$passing)
# cat("\nIntersection across hypoxia 1hr + normoxia si3d RNA + ribo:", length(si3d_rna_ribo_ids), "genes\n")
#
# neg_ctrl_si3d <- neg_ctrl_si3d %>%
#   filter(gene_id %in% si3d_rna_ribo_ids)
#
# cat("After si3d RNA + ribo filter:", nrow(neg_ctrl_si3d), "genes\n")
```

## Step 2: MDA-MB-231 si3e filter

```{r step2_si3e}
si3e_files <- tibble(
  label = c("si3e_normoxia_alltimepoints",
            "si3e_hypoxia_1hr",
            "si3e_hypoxia_4hr",
            "si3e_hypoxia_24hr"),
  file = here("output", c(
    "translation_categories_si3e_vs_sictrl_normoxia_alltimepoints.csv",
    "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv",
    "translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv",
    "translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv"
  ))
)

si3e_data <- si3e_files %>%
  rowwise() %>%
  mutate(data = list(read_csv(file, show_col_types = FALSE))) %>%
  ungroup()

si3e_passing <- si3e_data %>%
  mutate(
    passing = map(data, function(df) {
      df %>%
        filter(abs(te_lfc) < 0.5, te_padj > 0.3) %>%
        pull(gene_id)
    })
  )

cat("\nGenes passing si3e criteria per comparison:\n")
si3e_passing %>%
  mutate(n_passing = map_int(passing, length)) %>%
  dplyr::select(label, n_passing) %>%
  print(n = 4)

si3e_neg_ids <- Reduce(intersect, si3e_passing$passing)
cat("\nIntersection across all 4 si3e comparisons:", length(si3e_neg_ids), "genes\n")

n_after_si3e <- sum(neg_ctrl_si3d$gene_id %in% si3e_neg_ids)
cat("After intersecting with si3d baseline:", n_after_si3e, "genes\n")

neg_ctrl_si3d_si3e <- neg_ctrl_si3d %>%
  filter(gene_id %in% si3e_neg_ids)
```

## Step 3: MCF7-SIX1 si3d + si3e filter

Same criteria as MDA-MB-231: |log2FoldChange| < 0.5 AND padj > 0.3 across all four
MCF7-SIX1 comparisons (hypoxia si3d, hypoxia si3e, normoxia si3d, normoxia si3e).
Genes with NA padj are dropped (insufficient counts).

```{r step3_mcf7six1}
mcf7_files <- tibble(
  label = c("mcf7_hyp_si3d", "mcf7_hyp_si3e", "mcf7_nor_si3d", "mcf7_nor_si3e"),
  file = c(
    "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/hypoxia_si_ctrl_3d.csv",
    "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/hypoxia_si_ctrl_3e.csv",
    "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/normoxia_sictrl_3d.csv",
    "/Users/katematlin/github/2024_eIF3e_hypoxia/2024_eIF3e_hypoxia/counts/normoxia_sictrl_3e.csv"
  )
)

mcf7_data <- mcf7_files %>%
  rowwise() %>%
  mutate(data = list(
    read_csv(file, show_col_types = FALSE) %>%
      mutate(gene_id_clean = sub("\\..*", "", gene_id))
  )) %>%
  ungroup()

mcf7_passing <- mcf7_data %>%
  mutate(
    passing = map(data, function(df) {
      df %>%
        filter(!is.na(padj), abs(log2FoldChange) < 0.5, padj > 0.3) %>%
        pull(gene_id_clean)
    })
  )

cat("\nGenes passing MCF7-SIX1 criteria per comparison:\n")
mcf7_passing %>%
  mutate(n_passing = map_int(passing, length)) %>%
  dplyr::select(label, n_passing) %>%
  print(n = 4)

mcf7_neg_ids <- Reduce(intersect, mcf7_passing$passing)
cat("\nIntersection across all 4 MCF7-SIX1 comparisons:", length(mcf7_neg_ids), "genes\n")

n_after_mcf7 <- sum(neg_ctrl_si3d_si3e$gene_id_clean %in% mcf7_neg_ids)
cat("After intersecting with MDA-MB-231 pool:", n_after_mcf7, "genes\n")

neg_ctrl_strict <- neg_ctrl_si3d_si3e %>%
  filter(gene_id_clean %in% mcf7_neg_ids)
```

## Attrition summary

```{r attrition}
n_si3d_baseline <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes.csv"),
  show_col_types = FALSE
) %>% nrow()

cat("=== Attrition Summary ===\n")
cat("MDA-MB-231 si3d TE baseline (nb02):                          ", n_si3d_baseline, "\n")
cat("After MDA-MB-231 si3d RNA + ribo filter:                     ", nrow(neg_ctrl_si3d), "\n")
cat("After MDA-MB-231 si3e TE filter:                             ", nrow(neg_ctrl_si3d_si3e), "\n")
cat("After MCF7-SIX1 si3d + si3e TE filter (|lfc|<0.5, p>0.3):  ", nrow(neg_ctrl_strict), "\n")
```

# Get Transcript IDs from Feature Matrix

```{r load_feature_matrix}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))

tx_all <- fm %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol)

tx_neg <- tx_all %>%
  filter(gene_id_clean %in% neg_ctrl_strict$gene_id_clean)

cat("All expressed genes in feature matrix:", nrow(tx_all), "\n")
cat("Strict negative controls in feature matrix:", nrow(tx_neg), "of", nrow(neg_ctrl_strict), "\n")
```

# Extract CDS Lengths

```{r extract_cds_lengths}
txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds  <- cdsBy(txdb, by = "tx", use.names = TRUE)

cds_len_df <- as.data.frame(cds) %>%
  group_by(group_name) %>%
  summarise(cds_length = sum(width), .groups = "drop") %>%
  mutate(transcript_id_clean = sub("\\..*", "", group_name)) %>%
  dplyr::select(transcript_id_clean, cds_length)

cds_neg <- tx_neg %>%
  left_join(cds_len_df, by = "transcript_id_clean") %>%
  filter(!is.na(cds_length), cds_length > 0, cds_length %% 3 == 0)

cat("Strict negative controls - valid CDS:", nrow(cds_neg), "of", nrow(tx_neg), "\n")
```

# CDS Length Distributions

```{r plot_cds_length, fig.width=6, fig.height=4}
med_len_neg <- median(cds_neg$cds_length)
cat("Median CDS length - strict negative controls:", med_len_neg,
    "nt (", round(med_len_neg / 1000, 2), "kb)\n")

p_len_neg <- ggplot(cds_neg, aes(x = cds_length / 1000)) +
  geom_histogram(binwidth = 0.5, fill = "grey40", color = "white", alpha = 0.85) +
  geom_vline(xintercept = med_len_neg / 1000, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = med_len_neg / 1000 + 0.5, y = Inf,
           label = paste0("median = ", round(med_len_neg / 1000, 2), " kb"),
           hjust = 0, vjust = 1.4, size = 3.5) +
  labs(x = "CDS length (kb)", y = "Number of genes",
       title = "CDS length: strict negative controls") +
  theme_classic(base_size = 12)

print(p_len_neg)
ggsave(here("plots", "63_cds_length_negative_controls.pdf"), p_len_neg, width = 6, height = 4)
```

# Extract CDS Sequences and GC Content

```{r extract_gc}
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_seqs_all  <- extractTranscriptSeqs(genome, cds)
cds_seq_names <- sub("\\..*", "", names(cds_seqs_all))

keep_neg <- cds_seq_names %in% cds_neg$transcript_id_clean
cds_seqs_neg <- cds_seqs_all[keep_neg]
names(cds_seqs_neg) <- cds_seq_names[keep_neg]

gc_neg <- tibble(
  transcript_id_clean = names(cds_seqs_neg),
  gc_pct = as.numeric(letterFrequency(cds_seqs_neg, letters = "GC", as.prob = TRUE)) * 100
)

# Biological spot-check: ACTB CDS GC% expected 50-70%
actb_tx <- cds_neg %>% filter(symbol == "ACTB") %>% pull(transcript_id_clean)
if (length(actb_tx) > 0) {
  actb_gc <- gc_neg %>% filter(transcript_id_clean == actb_tx[1]) %>% pull(gc_pct)
  stopifnot("ACTB CDS GC% outside expected 50-70%" = between(actb_gc, 50, 70))
  cat("ACTB CDS GC%:", round(actb_gc, 2), "\n")
}

stopifnot("GC% out of range" = all(gc_neg$gc_pct >= 0 & gc_neg$gc_pct <= 100, na.rm = TRUE))
cat("GC computed for", nrow(gc_neg), "transcripts\n")
```

# CDS GC Content Distributions

```{r plot_gc, fig.width=6, fig.height=4}
med_gc_neg <- median(gc_neg$gc_pct)
cat("Median CDS GC% - strict negative controls:", round(med_gc_neg, 2), "\n")

p_gc_neg <- ggplot(gc_neg, aes(x = gc_pct)) +
  geom_histogram(binwidth = 1, fill = "grey40", color = "white", alpha = 0.85) +
  geom_vline(xintercept = med_gc_neg, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = med_gc_neg + 1, y = Inf,
           label = paste0("median = ", round(med_gc_neg, 1), "%"),
           hjust = 0, vjust = 1.4, size = 3.5) +
  labs(x = "CDS GC content (%)", y = "Number of genes",
       title = "CDS GC content: strict negative controls") +
  theme_classic(base_size = 12)

print(p_gc_neg)
ggsave(here("plots", "63_cds_gc_negative_controls.pdf"), p_gc_neg, width = 6, height = 4)
```

# Identify Median-Representative Genes

Genes within the 40th-60th percentile for BOTH CDS length and CDS GC content —
the most "typical" negative controls along both dimensions.

```{r median_representative}
cds_gc_combined <- cds_neg %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol, cds_length) %>%
  inner_join(gc_neg, by = "transcript_id_clean")

len_lo  <- quantile(cds_gc_combined$cds_length, 0.40)
len_hi  <- quantile(cds_gc_combined$cds_length, 0.60)
gc_lo   <- quantile(cds_gc_combined$gc_pct,     0.40)
gc_hi   <- quantile(cds_gc_combined$gc_pct,     0.60)

median_rep <- cds_gc_combined %>%
  filter(cds_length >= len_lo, cds_length <= len_hi,
         gc_pct     >= gc_lo,  gc_pct     <= gc_hi)

cat("CDS length window (40th-60th pct):", round(len_lo), "-", round(len_hi), "nt\n")
cat("GC% window        (40th-60th pct):", round(gc_lo, 1), "-", round(gc_hi, 1), "%\n")
cat("Median-representative genes:", nrow(median_rep), "\n\n")

median_rep %>%
  dplyr::select(symbol, gene_id_clean, cds_length, gc_pct) %>%
  arrange(cds_length) %>%
  print(n = 30)
```

# 5' UTR Lengths of Median-Representative Genes

```{r fiveutr_lengths, fig.width=6, fig.height=4}
utrs5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)

utr5_len_df <- as.data.frame(utrs5) %>%
  group_by(group_name) %>%
  summarise(utr5_length = sum(width), .groups = "drop") %>%
  mutate(transcript_id_clean = sub("\\..*", "", group_name)) %>%
  dplyr::select(transcript_id_clean, utr5_length)

median_rep_utr <- median_rep %>%
  left_join(utr5_len_df, by = "transcript_id_clean")

n_with_utr <- sum(!is.na(median_rep_utr$utr5_length))
cat("Median-representative genes with annotated 5' UTR:", n_with_utr, "of", nrow(median_rep), "\n")

utr_plot_df <- median_rep_utr %>% filter(!is.na(utr5_length))
med_utr5 <- median(utr_plot_df$utr5_length)
cat("Median 5' UTR length:", med_utr5, "nt\n")
cat("Range:", min(utr_plot_df$utr5_length), "-", max(utr_plot_df$utr5_length), "nt\n")
print(summary(utr_plot_df$utr5_length))

p_utr5 <- ggplot(utr_plot_df, aes(x = utr5_length)) +
  geom_histogram(binwidth = 25, fill = "grey40", color = "white", alpha = 0.85) +
  geom_vline(xintercept = med_utr5, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = med_utr5 + 10, y = Inf,
           label = paste0("median = ", med_utr5, " nt"),
           hjust = 0, vjust = 1.4, size = 3.5) +
  labs(x = "5' UTR length (nt)", y = "Number of genes",
       title = "5' UTR lengths: median-representative negative controls") +
  theme_classic(base_size = 12)

print(p_utr5)
ggsave(here("plots", "63_utr5_length_median_rep_negative_controls.pdf"),
       p_utr5, width = 6, height = 4)
```

# Annotate Final Candidate Lists

Add CDS length, GC%, 5' UTR length, and per-condition te_lfc/te_padj to both the
full strict negative control set and the median-representative subset.

```{r annotate_candidates}
# si3d_full_data and mcf7_data are loaded in step1b and step3 chunks respectively

# Wide te/rna/ribo LFC+padj for si3d
si3d_wide <- neg_ctrl_strict %>% dplyr::select(gene_id)
for (i in seq_len(nrow(si3d_full_data))) {
  s <- si3d_full_data$label[i]
  si3d_wide <- si3d_wide %>%
    left_join(
      si3d_full_data$data[[i]] %>%
        dplyr::select(gene_id, te_lfc, te_padj, rna_lfc, rna_padj, ribo_lfc, ribo_padj) %>%
        dplyr::rename(!!paste0(s, "_te_lfc")   := te_lfc,
                      !!paste0(s, "_te_padj")  := te_padj,
                      !!paste0(s, "_rna_lfc")  := rna_lfc,
                      !!paste0(s, "_rna_padj") := rna_padj,
                      !!paste0(s, "_ribo_lfc") := ribo_lfc,
                      !!paste0(s, "_ribo_padj"):= ribo_padj),
      by = "gene_id"
    )
}

# Wide LFC/padj for si3e
si3e_shorts <- c("si3e_nor_all", "si3e_hyp_1hr", "si3e_hyp_4hr", "si3e_hyp_24hr")
si3e_wide <- neg_ctrl_strict %>% dplyr::select(gene_id)
for (i in seq_len(nrow(si3e_data))) {
  s <- si3e_shorts[i]
  si3e_wide <- si3e_wide %>%
    left_join(
      si3e_data$data[[i]] %>%
        dplyr::select(gene_id, te_lfc, te_padj) %>%
        dplyr::rename(!!paste0(s, "_te_lfc")  := te_lfc,
                      !!paste0(s, "_te_padj") := te_padj),
      by = "gene_id"
    )
}

# Wide LFC/padj for MCF7-SIX1 (log2FoldChange / padj columns)
mcf7_wide <- neg_ctrl_strict %>% dplyr::select(gene_id_clean)
for (i in seq_len(nrow(mcf7_data))) {
  s <- mcf7_data$label[i]
  mcf7_wide <- mcf7_wide %>%
    left_join(
      mcf7_data$data[[i]] %>%
        dplyr::select(gene_id_clean, log2FoldChange, padj) %>%
        dplyr::rename(!!paste0(s, "_te_lfc")  := log2FoldChange,
                      !!paste0(s, "_te_padj") := padj),
      by = "gene_id_clean"
    )
}

# CDS length, GC%, 5' UTR via transcript
tx_feats <- tx_neg %>%
  left_join(cds_len_df,  by = "transcript_id_clean") %>%
  left_join(gc_neg,      by = "transcript_id_clean") %>%
  left_join(utr5_len_df, by = "transcript_id_clean") %>%
  dplyr::select(gene_id_clean, transcript_id_clean, cds_length, gc_pct, utr5_length)

neg_ctrl_annotated <- neg_ctrl_strict %>%
  left_join(tx_feats,  by = "gene_id_clean") %>%
  left_join(si3d_wide, by = "gene_id") %>%
  left_join(si3e_wide, by = "gene_id") %>%
  left_join(mcf7_wide, by = "gene_id_clean") %>%
  dplyr::select(gene_id, gene_id_clean, symbol, transcript_id_clean,
                cds_length, gc_pct, utr5_length,
                starts_with("si3d_"), starts_with("si3e_"),
                starts_with("mcf7_"))

cat("Annotated table:", nrow(neg_ctrl_annotated), "genes x",
    ncol(neg_ctrl_annotated), "columns\n")
cat("  (si3d: TE + RNA + ribo per condition; si3e + MCF7-SIX1: TE only)\n")
cat("Genes with CDS length:", sum(!is.na(neg_ctrl_annotated$cds_length)), "\n")
cat("Genes with GC%:       ", sum(!is.na(neg_ctrl_annotated$gc_pct)), "\n")
cat("Genes with 5' UTR:    ", sum(!is.na(neg_ctrl_annotated$utr5_length)), "\n")

# Median-representative: join LFC/padj from annotated table
median_rep_annotated <- median_rep_utr %>%
  left_join(
    neg_ctrl_annotated %>%
      dplyr::select(gene_id_clean, starts_with("si3d_"), starts_with("si3e_")),
    by = "gene_id_clean"
  )
```

# Export

```{r export}
write_csv(neg_ctrl_annotated,
          here("output", "predictive_modeling", "negative_control_genes_strict.csv"))
cat("Exported: output/predictive_modeling/negative_control_genes_strict.csv\n")

write_csv(median_rep_annotated,
          here("output", "predictive_modeling", "negative_control_genes_median_representative.csv"))
cat("Exported: output/predictive_modeling/negative_control_genes_median_representative.csv\n")
cat("Final strict negative control set:", nrow(neg_ctrl_annotated), "genes\n")
cat("Median-representative subset:", nrow(median_rep_annotated), "genes\n")
```

# Summary

```{r summary}
cat("=== Summary ===\n\n")
cat("Filter attrition:\n")
cat("  MDA-MB-231 si3d baseline (nb02):     ", nrow(neg_ctrl_si3d), "genes\n")
cat("  After MDA-MB-231 si3e filter:        ", nrow(neg_ctrl_si3d_si3e), "genes\n")
cat("  After MCF7-SIX1 si3d + si3e filter: ", nrow(neg_ctrl_strict), "genes\n\n")
cat("Strict negative controls in feature matrix:", nrow(tx_neg), "genes\n")
cat("  Valid CDS:                   ", nrow(cds_neg), "\n")
cat("  Median CDS length:           ", med_len_neg, "nt (", round(med_len_neg / 1000, 2), "kb)\n")
cat("  Median CDS GC%:              ", round(med_gc_neg, 2), "\n\n")
cat("Median-representative subset (40th-60th pct both length + GC):", nrow(median_rep), "genes\n")
cat("  With annotated 5' UTR:       ", n_with_utr, "\n")
cat("  Median 5' UTR length:        ", med_utr5, "nt\n")
```

# Session Info

```{r session_info}
sessionInfo()
```
