---
title: "65 — Benchling Construct Design: 5' UTR Reporter Constructs"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(tximport)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
```

# Gene lists

```{r define_genes}
# Previous hardcoded list (kept for sanity check below):
# target_symbols <- c(
#   "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
#   "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
# )

# Load candidates from nb11 Benchling export (Exclusive eIF3d hypoxia-specific targets,
# ribo-only changes, RIP-bound, isoform-consistent)
benchling_candidates <- read_csv(
  here("ripseq", "output", "11_benchling_import.csv"),
  show_col_types = FALSE
)
target_symbols <- sub(" 5UTR.*", "", benchling_candidates$Name)

# Sanity check: previously hardcoded targets must all be present
prior_targets <- c(
  "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
  "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
)
missing_from_csv <- setdiff(prior_targets, target_symbols)
stopifnot(
  "All previously hardcoded target symbols must be present in 11_benchling_import.csv" =
    length(missing_from_csv) == 0
)
cat("Loaded", length(target_symbols), "target symbols from 11_benchling_import.csv\n")
cat("All", length(prior_targets), "prior hardcoded symbols confirmed present.\n")

neg_ctrl <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes_strict.csv"),
  show_col_types = FALSE
)
neg_symbols <- neg_ctrl$symbol
cat("Loaded", length(neg_symbols), "negative control candidates from negative_control_genes_strict.csv\n")

# nb24 Plot 2c condition-specific candidates (see notebook 24, chunk
# plot-named-specific-genes). Gene IDs are pinned explicitly here rather than
# resolved by expression: three of these five symbols map to two ENSG IDs each,
# and for SMIM10L1 the expression rule used below picks the WRONG copy — the
# lncRNA locus ENSG00000281930 (higher total RNA) rather than the protein-coding
# locus ENSG00000256537 that actually carries the eIF3d TE effect. DDR1 and TNKS
# likewise carry their signal on the non-canonical copy.
nb24_candidates <- tribble(
  ~symbol,    ~gene_id_forced,   ~nb24_group,
  "DDR1",     "ENSG00000230456", "Hypoxia-specific",
  "FAM72D",   "ENSG00000215784", "Hypoxia-specific",
  "SMIM10L1", "ENSG00000256537", "Hypoxia-specific",
  "PDIA3P1",  "ENSG00000180867", "Normoxia-specific",
  "TNKS",     "ENSG00000285372", "Normoxia-specific"
)
nb24_symbols <- nb24_candidates$symbol

# These must not collide with the existing lists, or the forced-ID join below
# would rewrite the gene ID of an unrelated target/negative-control row.
stopifnot(
  "nb24 candidate symbols must not overlap the target or neg-control lists" =
    length(intersect(nb24_symbols, c(target_symbols, neg_symbols))) == 0
)
cat("Loaded", length(nb24_symbols), "nb24 condition-specific candidates\n")

all_symbols <- c(target_symbols, neg_symbols, nb24_symbols)

gene_groups <- tibble(
  symbol     = all_symbols,
  gene_group = c(rep("target", length(target_symbols)),
                 rep("negative_control", length(neg_symbols)),
                 rep("nb24_candidate", length(nb24_symbols)))
)
```

# Gene annotation lookup

Restrict to gene IDs present in the MDA-MB-231 TE dataset. This resolves ambiguous
symbols like RING1, which maps to multiple ENSG IDs in GENCODE v49 (alternative
loci in the MHC region) but only one of which (ENSG00000204227) is the expressed
canonical locus in our data.

```{r gene_annotation}
# Load TE data to select the expressed gene ID for each symbol.
# complete_te_list_1hr.csv columns: symbol, gene_id, TE columns, biotype,
# RNA count columns ("_rna_"), ribosome footprint columns ("_ribo_").
te_full <- read_csv(here("output", "complete_te_list_1hr.csv"),
                    show_col_types = FALSE) |>
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    total_rna     = rowSums(across(contains("_rna_")), na.rm = TRUE)
  )

# For symbols with multiple expressed ENSG IDs (e.g. RING1 maps to two loci in
# GENCODE v49), keep the one with higher total RNA expression — that is the
# canonical expressed locus (ENSG00000204227 for RING1, which is also the ID
# used in negative_control_genes.csv; the other locus ENSG00000235107 is
# actually upregulated in si3d and is not a suitable negative control).
gene_id_map <- te_full |>
  dplyr::select(symbol, gene_id_clean, total_rna) |>
  group_by(symbol) |>
  dplyr::slice_max(total_rna, n = 1, with_ties = FALSE) |>
  ungroup()

gene_df <- gene_groups |>
  left_join(gene_id_map |> dplyr::select(symbol, gene_id_clean), by = "symbol") |>
  left_join(nb24_candidates |> dplyr::select(symbol, gene_id_forced), by = "symbol") |>
  mutate(gene_id_clean = coalesce(gene_id_forced, gene_id_clean))

# Report where the pinned ID disagrees with the expression rule, so the override is
# visible in the knit rather than silent. SMIM10L1 is the known case.
nb24_override_check <- gene_df |>
  filter(!is.na(gene_id_forced)) |>
  left_join(gene_id_map |> dplyr::select(symbol, gene_id_by_expression = gene_id_clean),
            by = "symbol") |>
  mutate(overridden = gene_id_forced != gene_id_by_expression)
cat("\nnb24 candidates — pinned gene ID vs expression-selected gene ID:\n")
print(nb24_override_check |>
        dplyr::select(symbol, gene_id_forced, gene_id_by_expression, overridden))
stopifnot(
  "pinned nb24 gene IDs must all be present in the TE dataset" =
    all(nb24_candidates$gene_id_forced %in% te_full$gene_id_clean)
)

gene_df <- gene_df |> dplyr::select(-gene_id_forced)

stopifnot(
  "All symbols must be found in TE data" =
    all(all_symbols %in% gene_df$symbol),
  "Each symbol must resolve to exactly one gene ID" =
    nrow(gene_df) == length(all_symbols)
)

cat("Gene IDs resolved:\n")
print(gene_df |> dplyr::select(symbol, gene_group, gene_id_clean))
```

# Transcript isoform selection

Select the most-expressed transcript per gene in siCTRL normoxia 1hr and siCTRL hypoxia 1hr, restricting to transcripts with an annotated 5'UTR in GENCODE v49. This avoids selecting transcripts that lack a 5'UTR annotation (e.g. RING1's most abundant isoform ENST00000436128 has no 5'UTR in the txdb). Use the normoxia-dominant transcript; verify it is also dominant in hypoxia. For any inconsistent gene, fall back to the hypoxia-dominant transcript and report it.

```{r isoform_selection}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

# Transcript IDs (clean, version-stripped) with annotated 5'UTR in GENCODE v49
utr5_all         <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr5_tx_versioned <- names(utr5_all)
utr5_tx_clean    <- sub("\\..*", "", utr5_tx_versioned)

metadata <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                     show_col_types = FALSE)

salmon_dir <- here("2025_MDAcells_riboseq", "salmon")

tx2gene_map <- read_tsv(
  here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
  col_names = c("transcript_id", "gene_id"), show_col_types = FALSE
) |>
  filter(str_starts(transcript_id, "ENST"))

load_top_tx <- function(sample_ids, label) {
  files <- file.path(salmon_dir, sample_ids, "quant.sf.gz")
  names(files) <- sample_ids
  files <- files[file.exists(files)]
  cat(label, "— samples found:", length(files), "\n")
  txi <- tximport(files, type = "salmon", tx2gene = tx2gene_map, txOut = TRUE)
  txi$abundance |>
    as.data.frame() |>
    rownames_to_column("transcript_id") |>
    mutate(
      transcript_id_clean = sub("\\..*", "", transcript_id),
      gene_id_clean       = sub("\\..*", "",
        tx2gene_map$gene_id[match(transcript_id, tx2gene_map$transcript_id)])
    ) |>
    mutate(total_tpm = rowSums(across(where(is.numeric)), na.rm = TRUE))
}

nor_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "normoxia", hours == 1) |>
  pull(id)

hyp_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "hypoxia", hours == 1) |>
  pull(id)

nor_abundance <- load_top_tx(nor_ids, "Normoxia siCTRL 1hr")
hyp_abundance <- load_top_tx(hyp_ids, "Hypoxia siCTRL 1hr")

top_tx_per_gene <- function(abund_df) {
  abund_df |>
    filter(gene_id_clean %in% gene_df$gene_id_clean,
           transcript_id_clean %in% utr5_tx_clean) |>
    group_by(gene_id_clean) |>
    dplyr::slice_max(total_tpm, n = 1, with_ties = FALSE) |>
    ungroup() |>
    dplyr::select(gene_id_clean, transcript_id_clean, total_tpm)
}

nor_top <- top_tx_per_gene(nor_abundance) |>
  dplyr::rename(top_tx_nor = transcript_id_clean, tpm_nor = total_tpm)

hyp_top <- top_tx_per_gene(hyp_abundance) |>
  dplyr::rename(top_tx_hyp = transcript_id_clean, tpm_hyp = total_tpm)

isoform_check <- gene_df |>
  left_join(nor_top, by = "gene_id_clean") |>
  left_join(hyp_top, by = "gene_id_clean") |>
  mutate(consistent = !is.na(top_tx_nor) & !is.na(top_tx_hyp) &
                      top_tx_nor == top_tx_hyp)

cat("\nIsoform dominance consistency (normoxia vs hypoxia siCTRL 1hr):\n")
print(isoform_check |>
        dplyr::select(symbol, gene_group, top_tx_nor, tpm_nor, top_tx_hyp, tpm_hyp, consistent))

switched <- isoform_check |> filter(!consistent)
if (nrow(switched) > 0) {
  cat("\nInconsistent dominant isoform — using hypoxia-dominant transcript for:\n")
  print(switched |> dplyr::select(symbol, top_tx_nor, top_tx_hyp))
}

# Use normoxia-dominant; fall back to hypoxia-dominant if inconsistent
tx_selected <- isoform_check |>
  mutate(
    transcript_id_final = if_else(consistent, top_tx_nor, top_tx_hyp)
  ) |>
  dplyr::select(symbol, gene_group, gene_id_clean, transcript_id_final)

target_missing <- tx_selected |>
  filter(gene_group == "target", is.na(transcript_id_final)) |>
  pull(symbol)
stopifnot(
  "All target genes must have a selected transcript" =
    length(target_missing) == 0
)
neg_missing <- tx_selected |>
  filter(gene_group == "negative_control", is.na(transcript_id_final)) |>
  pull(symbol)
if (length(neg_missing) > 0) {
  cat("Negative controls without a 5'UTR-annotated transcript (excluded from pipeline):\n")
  cat(paste(neg_missing, collapse = ", "), "\n")
}

# nb24 candidates are reported-and-excluded rather than hard-failed: three of the
# five pinned loci have NO transcript with an annotated 5'UTR in GENCODE v49,
# because the copy carrying the eIF3d effect is not a normal coding transcript —
# DDR1 ENSG00000230456 is protein_coding_CDS_not_defined, TNKS ENSG00000285372 is
# nonsense_mediated_decay, and PDIA3P1 is a transcribed_processed_pseudogene. With
# no CDS there is no annotated 5'UTR boundary, so no 5'UTR construct can be built
# from these loci. Their other ENSG copy does have UTRs but shows no TE effect
# (DDR1 ENSG00000204580 te_lfc -0.01, TNKS ENSG00000173273 -0.06), so silently
# substituting it would build a reporter for the wrong transcript.
nb24_missing <- tx_selected |>
  filter(gene_group == "nb24_candidate", is.na(transcript_id_final)) |>
  pull(symbol)
nb24_ok <- tx_selected |>
  filter(gene_group == "nb24_candidate", !is.na(transcript_id_final)) |>
  pull(symbol)
cat("\nnb24 candidates WITH a 5'UTR-annotated transcript:",
    if (length(nb24_ok) > 0) paste(nb24_ok, collapse = ", ") else "(none)", "\n")
cat("nb24 candidates WITHOUT one (cannot build a 5'UTR construct):",
    if (length(nb24_missing) > 0) paste(nb24_missing, collapse = ", ") else "(none)", "\n")

tx_selected <- tx_selected |> filter(!is.na(transcript_id_final))
```

# Extract 5' UTR sequences

```{r extract_utr5}
# Map clean transcript IDs to versioned IDs needed by txdb
# (txdb, genome, utr5_all already loaded in isoform_selection chunk)
tx_version_map <- tx_selected |>
  left_join(
    tx2gene_map |>
      mutate(transcript_id_clean = sub("\\..*", "", transcript_id)) |>
      dplyr::select(transcript_id_clean, transcript_id_versioned = transcript_id) |>
      distinct(transcript_id_clean, .keep_all = TRUE),
    by = c("transcript_id_final" = "transcript_id_clean")
  )

target_utrs <- utr5_all[names(utr5_all) %in% tx_version_map$transcript_id_versioned]
utr5_seqs   <- extractTranscriptSeqs(genome, target_utrs)

# Convert to named character vector so indexing with missing names returns NA
# instead of erroring (DNAStringSet subsetting errors on unknown names)
utr5_char <- setNames(as.character(utr5_seqs), names(utr5_seqs))

seq_df <- tx_version_map |>
  mutate(
    utr5_sequence = utr5_char[transcript_id_versioned],
    utr5_length   = nchar(utr5_sequence)
  ) |>
  arrange(match(symbol, all_symbols))

missing_utr5 <- seq_df |> filter(is.na(utr5_sequence)) |> pull(symbol)
if (length(missing_utr5) > 0) {
  cat("WARNING — no annotated 5'UTR in GENCODE v49 for:", paste(missing_utr5, collapse = ", "), "\n")
}

cat("\n5' UTR lengths (bp):\n")
print(seq_df |> dplyr::select(symbol, gene_group, transcript_id_final, utr5_length))
```

# Validation

```{r validate}
stopifnot(
  "No missing 5'UTR sequences for target genes" =
    !any(is.na(seq_df$utr5_sequence[seq_df$gene_group == "target"])),
  "All 5'UTR sequences non-empty" =
    all(nchar(seq_df$utr5_sequence) > 0, na.rm = TRUE),
  "S100A11 5'UTR length in expected range (~89 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "S100A11"], 75, 105),
  "UBL5 5'UTR length in expected range (~78 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "UBL5"], 65, 95),
  "PTMS 5'UTR length in expected range (~343 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "PTMS"], 300, 380),
  "All construct names unique" =
    !anyDuplicated(seq_df$symbol)
)

cat("All validation checks passed.\n")
```

# Assemble constructs

```{r assemble_constructs}
prefix         <- "GTGAACCGTCAGATCAGATC"
reporter_start <- "atgGTCTTCACACTCGAAGATTTCGTTGGGGACTGGCGACAGACAGCCGGCTACAACCTGGACCAAGTCCTTG"
suffix         <- "AACAGGGAGGTGTGTCCAGT"

constructs_df <- seq_df |>
  mutate(
    Name             = symbol,
    prefix           = prefix,
    reporter_start   = reporter_start,
    suffix           = suffix,
    full_construct   = paste0(prefix, utr5_sequence, reporter_start, suffix),
    construct_length = nchar(full_construct)
  ) |>
  dplyr::select(Name, symbol, gene_group, transcript_id_final,
                utr5_length, utr5_sequence,
                prefix, reporter_start, suffix, full_construct, construct_length)

stopifnot(
  "All full_construct values start with expected prefix" =
    all(str_starts(constructs_df$full_construct, prefix)),
  "Benchling names must be unique" =
    !anyDuplicated(constructs_df$Name)
)

cat("Construct lengths (bp):\n")
print(constructs_df |> dplyr::select(Name, gene_group, utr5_length, construct_length))

# The bidirectional vector is linearised with Eco31I (BsaI, GGTCTC) and BsmBI-v2
# (Esp3I, CGTCTC), so an internal site in an inserted 5'UTR would be re-cut. The
# fixed prefix / reporter_start / suffix are site-free by design; only the variable
# 5'UTR can introduce one. Reported, not filtered — a Gibson-style assembly onto the
# 20 bp homology arms tolerates internal sites, a Golden Gate one does not.
count_sites <- function(x) {
  x <- toupper(x)
  str_count(x, "GGTCTC") + str_count(x, "GAGACC") +   # Eco31I / BsaI, both strands
  str_count(x, "CGTCTC") + str_count(x, "GAGACG")     # Esp3I / BsmBI-v2, both strands
}
stopifnot(
  "fixed construct parts must be free of Eco31I / BsmBI sites" =
    count_sites(paste0(prefix, reporter_start, suffix)) == 0
)
constructs_df <- constructs_df |> mutate(n_typeIIs_sites = count_sites(utr5_sequence))
flagged <- constructs_df |> filter(n_typeIIs_sites > 0)
if (nrow(flagged) > 0) {
  cat("\nWARNING — 5'UTRs containing an internal Eco31I/BsmBI site:\n")
  print(flagged |> dplyr::select(Name, gene_group, utr5_length, n_typeIIs_sites))
} else {
  cat("\nNo internal Eco31I/BsmBI sites in any 5'UTR.\n")
}

# Report what the length filter removes rather than dropping it silently — the
# fixed parts total 113 bp, so this cut requires a 5'UTR of >= 187 bp.
dropped_short <- constructs_df |> filter(construct_length < 300)
if (nrow(dropped_short) > 0) {
  cat("\nDropped by the >= 300 bp construct filter:\n")
  print(dropped_short |> dplyr::select(Name, gene_group, utr5_length, construct_length))
}

constructs_df <- constructs_df |> filter(construct_length >= 300)
cat("\nAfter filtering to >= 300 bp constructs:", nrow(constructs_df), "remain",
    "(", sum(constructs_df$gene_group == "target"), "targets,",
    sum(constructs_df$gene_group == "negative_control"), "neg controls,",
    sum(constructs_df$gene_group == "nb24_candidate"), "nb24 candidates )\n")
```

# Negative control candidate review

Review the table below (printed by `assemble_constructs`), then set `neg_symbols_keep`
to the symbols you want to include in the export. All target genes are always included.

```{r neg_control_review}
cat("Negative control candidates with construct details:\n")
print(
  constructs_df |>
    filter(gene_group == "negative_control") |>
    dplyr::select(symbol, transcript_id_final, utr5_length, construct_length)
)
```

```{r neg_select}
# Update this vector after reviewing the table above.
neg_symbols_keep <- c("ADAT3", "PAN2", "DXO")

neg_candidates <- constructs_df |>
  filter(gene_group == "negative_control") |>
  pull(symbol)
stopifnot(
  "neg_symbols_keep must be a subset of neg controls passing the length filter" =
    all(neg_symbols_keep %in% neg_candidates)
)
cat("Negative controls selected for export:", length(neg_symbols_keep), "\n")
print(neg_symbols_keep)
```

# Final construct selection

```{r final_select}
target_symbols_keep <- c(
  "LAPTM4B", "EIF4G2", "CLIC1", "FUT8",
  "UBE2M", "PIP5K1A", "IWS1", "STYXL1", "PIGBOS1", "SERTAD2", "ZNF146"
)

stopifnot(
  "All target_symbols_keep must be present in constructs_df targets" =
    all(target_symbols_keep %in% constructs_df$symbol[constructs_df$gene_group == "target"])
)

# All nb24 candidates that survived 5'UTR extraction and the length filter are kept.
nb24_symbols_keep <- constructs_df |>
  filter(gene_group == "nb24_candidate") |>
  pull(symbol)

constructs_final <- constructs_df |>
  filter(
    (gene_group == "target" & symbol %in% target_symbols_keep) |
    (gene_group == "negative_control" & symbol %in% neg_symbols_keep) |
    (gene_group == "nb24_candidate" & symbol %in% nb24_symbols_keep)
  )

cat("Final selection:", nrow(constructs_final), "constructs\n")
cat(" ", sum(constructs_final$gene_group == "target"), "targets\n")
cat(" ", sum(constructs_final$gene_group == "negative_control"), "negative controls\n")
cat(" ", sum(constructs_final$gene_group == "nb24_candidate"), "nb24 candidates\n")
print(constructs_final |> dplyr::select(symbol, gene_group, utr5_length, construct_length))

# Account for every one of the five nb24 candidates: built, or dropped with a reason.
nb24_outcome <- nb24_candidates |>
  mutate(
    outcome = case_when(
      symbol %in% constructs_final$symbol                       ~ "construct built",
      symbol %in% nb24_missing                                  ~ "no annotated 5'UTR at pinned locus",
      symbol %in% dropped_short$symbol                          ~ "5'UTR too short (< 187 bp)",
      TRUE                                                      ~ "excluded (see log above)"
    )
  )
cat("\nnb24 candidate outcomes:\n")
print(nb24_outcome)
stopifnot(
  "every nb24 candidate must be accounted for" =
    nrow(nb24_outcome) == length(nb24_symbols)
)
```

# CLIC1 and ADAT3 5' UTR sequences

```{r clic1_adat3_utr5}
focal_genes <- c("CLIC1", "ADAT3")
focal_utrs  <- constructs_final |>
  filter(symbol %in% focal_genes) |>
  dplyr::select(symbol, gene_group, transcript_id_final, utr5_length, utr5_sequence)

stopifnot(
  "CLIC1 and ADAT3 must both be present in constructs_final" =
    all(focal_genes %in% focal_utrs$symbol)
)

for (i in seq_len(nrow(focal_utrs))) {
  cat(focal_utrs$symbol[i], "(", focal_utrs$gene_group[i], ") |",
      focal_utrs$transcript_id_final[i], "| length:", focal_utrs$utr5_length[i], "bp\n")
  cat(focal_utrs$utr5_sequence[i], "\n\n")
}
```

# Export

```{r export}
write_csv(
  constructs_final,
  here("output", "predictive_modeling", "65_constructs_all.csv")
)
cat("Saved construct table:",
    nrow(constructs_final), "rows to output/predictive_modeling/65_constructs_all.csv\n")

benchling_df <- constructs_final |>
  dplyr::select(Name, Bases = full_construct, construct_length)

write_csv(
  benchling_df,
  here("output", "predictive_modeling", "65_benchling_import.csv")
)
cat("Saved Benchling import:", nrow(benchling_df),
    "constructs to output/predictive_modeling/65_benchling_import.csv\n")

cat("\nBenchling import preview:\n")
print(benchling_df |>
        mutate(Bases_preview = paste0(str_sub(Bases, 1, 40), "...")))
```

```{r}

```