---
title: "65 — Benchling Construct Design: 5' UTR Reporter Constructs"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(tximport)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
```

# Gene lists

```{r define_genes}
# Previous hardcoded list (kept for sanity check below):
# target_symbols <- c(
#   "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
#   "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
# )

# Load candidates from nb11 Benchling export (Exclusive eIF3d hypoxia-specific targets,
# ribo-only changes, RIP-bound, isoform-consistent)
benchling_candidates <- read_csv(
  here("ripseq", "output", "11_benchling_import.csv"),
  show_col_types = FALSE
)
target_symbols <- sub(" 5UTR.*", "", benchling_candidates$Name)

# Sanity check: previously hardcoded targets must all be present
prior_targets <- c(
  "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
  "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
)
missing_from_csv <- setdiff(prior_targets, target_symbols)
stopifnot(
  "All previously hardcoded target symbols must be present in 11_benchling_import.csv" =
    length(missing_from_csv) == 0
)
cat("Loaded", length(target_symbols), "target symbols from 11_benchling_import.csv\n")
cat("All", length(prior_targets), "prior hardcoded symbols confirmed present.\n")

neg_ctrl <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes_strict.csv"),
  show_col_types = FALSE
)
neg_symbols <- neg_ctrl$symbol
cat("Loaded", length(neg_symbols), "negative control candidates from negative_control_genes_strict.csv\n")

# nb24 Plot 2c condition-specific candidates (see notebook 24, chunk
# plot-named-specific-genes). Gene IDs are pinned explicitly here rather than
# resolved by expression: three of these five symbols map to two ENSG IDs each,
# and for SMIM10L1 the expression rule used below picks the WRONG copy — the
# lncRNA locus ENSG00000281930 (higher total RNA) rather than the protein-coding
# locus ENSG00000256537 that actually carries the eIF3d TE effect. DDR1 and TNKS
# likewise carry their signal on the non-canonical copy.
nb24_candidates <- tribble(
  ~symbol,    ~gene_id_forced,   ~nb24_group,
  "DDR1",     "ENSG00000230456", "Hypoxia-specific",
  "FAM72D",   "ENSG00000215784", "Hypoxia-specific",
  "SMIM10L1", "ENSG00000256537", "Hypoxia-specific",
  "PDIA3P1",  "ENSG00000180867", "Normoxia-specific",
  "TNKS",     "ENSG00000285372", "Normoxia-specific"
)
nb24_symbols <- nb24_candidates$symbol

# These must not collide with the existing lists, or the forced-ID join below
# would rewrite the gene ID of an unrelated target/negative-control row.
stopifnot(
  "nb24 candidate symbols must not overlap the target or neg-control lists" =
    length(intersect(nb24_symbols, c(target_symbols, neg_symbols))) == 0
)
cat("Loaded", length(nb24_symbols), "nb24 condition-specific candidates\n")

# nb24 Plot 2d replacement set (chunk `plot-clonable-specific-genes`). Three of the
# five Plot 2c genes above carry their eIF3d effect on a locus with no annotated
# 5'UTR, so no construct can be built from them; these five were re-screened under
# the same condition-specific definitions with the added requirement that the MANE
# Select transcript is the most-abundant 5'UTR-annotated isoform in BOTH siCTRL
# normoxia 1hr and siCTRL hypoxia 1hr.
#
# `tx_expected` / `utr5_expected` are the values that screen produced. They are
# pinned here so the isoform-selection and sequence-extraction steps below are
# checked against an independent prior rather than merely re-run: if a GENCODE
# update, a salmon re-quant, or an edit to the selection rule moves any of these,
# the notebook fails instead of silently exporting a different 5'UTR.
nb24_clonable <- tribble(
  ~symbol,     ~gene_id_forced,   ~nb24_group,         ~tx_expected,      ~utr5_expected,
  "MEMO1",     "ENSG00000162959", "Hypoxia-specific",  "ENST00000404530",  84L,
  "DNAJC3",    "ENSG00000102580", "Hypoxia-specific",  "ENST00000602402", 117L,
  "CCN1",      "ENSG00000142871", "Hypoxia-specific",  "ENST00000451137", 224L,
  "GADD45B",   "ENSG00000099860", "Normoxia-specific", "ENST00000215631", 232L,
  "MAP1LC3B",  "ENSG00000140941", "Normoxia-specific", "ENST00000268607",  92L
)
# TMSB15B (ENSG00000158427) was the first normoxia-specific pick on TE grounds
# (nor 1.93 / hyp 0.10, the cleanest separation in the pool) and was dropped here:
# its 114 bp 5'UTR carries an Eco31I site (GGTCTC), which the bidirectional vector
# re-cuts during Golden Gate assembly. GADD45B is the next candidate that is
# site-free, and it has the next-cleanest hypoxia null in the pool (hyp 0.07).
clonable_symbols <- nb24_clonable$symbol

stopifnot(
  "clonable candidate symbols must not overlap any other list" =
    length(intersect(clonable_symbols,
                     c(target_symbols, neg_symbols, nb24_symbols))) == 0,
  "clonable candidate symbols must be unique" =
    !anyDuplicated(clonable_symbols),
  "clonable candidate gene IDs must be unique" =
    !anyDuplicated(nb24_clonable$gene_id_forced),
  "clonable expected transcript IDs must be unique" =
    !anyDuplicated(nb24_clonable$tx_expected)
)
cat("Loaded", length(clonable_symbols), "nb24 Plot 2d clonable candidates\n")

all_symbols <- c(target_symbols, neg_symbols, nb24_symbols, clonable_symbols)

gene_groups <- tibble(
  symbol     = all_symbols,
  gene_group = c(rep("target", length(target_symbols)),
                 rep("negative_control", length(neg_symbols)),
                 rep("nb24_candidate", length(nb24_symbols)),
                 rep("nb24_clonable", length(clonable_symbols)))
)
stopifnot(
  "gene_groups must have one row per symbol with no duplicates" =
    nrow(gene_groups) == length(all_symbols) && !anyDuplicated(gene_groups$symbol)
)
```

# Gene annotation lookup

Restrict to gene IDs present in the MDA-MB-231 TE dataset. This resolves ambiguous
symbols like RING1, which maps to multiple ENSG IDs in GENCODE v49 (alternative
loci in the MHC region) but only one of which (ENSG00000204227) is the expressed
canonical locus in our data.

```{r gene_annotation}
# Load TE data to select the expressed gene ID for each symbol.
# complete_te_list_1hr.csv columns: symbol, gene_id, TE columns, biotype,
# RNA count columns ("_rna_"), ribosome footprint columns ("_ribo_").
te_full <- read_csv(here("output", "complete_te_list_1hr.csv"),
                    show_col_types = FALSE) |>
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    total_rna     = rowSums(across(contains("_rna_")), na.rm = TRUE)
  )

# For symbols with multiple expressed ENSG IDs (e.g. RING1 maps to two loci in
# GENCODE v49), keep the one with higher total RNA expression — that is the
# canonical expressed locus (ENSG00000204227 for RING1, which is also the ID
# used in negative_control_genes.csv; the other locus ENSG00000235107 is
# actually upregulated in si3d and is not a suitable negative control).
gene_id_map <- te_full |>
  dplyr::select(symbol, gene_id_clean, total_rna) |>
  group_by(symbol) |>
  dplyr::slice_max(total_rna, n = 1, with_ties = FALSE) |>
  ungroup()

# Both pinned tables feed one forced-ID column. Kept as a single bind_rows so a
# symbol can never be pinned to two different gene IDs without the uniqueness check
# below firing — a silent duplicate here would make the left_join fan out and
# duplicate that gene's construct row.
forced_ids <- bind_rows(
  nb24_candidates |> dplyr::select(symbol, gene_id_forced),
  nb24_clonable   |> dplyr::select(symbol, gene_id_forced)
)
stopifnot(
  "a symbol is pinned to more than one gene ID" = !anyDuplicated(forced_ids$symbol)
)

gene_df <- gene_groups |>
  left_join(gene_id_map |> dplyr::select(symbol, gene_id_clean), by = "symbol") |>
  left_join(forced_ids, by = "symbol") |>
  mutate(gene_id_clean = coalesce(gene_id_forced, gene_id_clean))

# Report where the pinned ID disagrees with the expression rule, so the override is
# visible in the knit rather than silent. SMIM10L1 is the known case.
nb24_override_check <- gene_df |>
  filter(!is.na(gene_id_forced)) |>
  left_join(gene_id_map |> dplyr::select(symbol, gene_id_by_expression = gene_id_clean),
            by = "symbol") |>
  mutate(overridden = gene_id_forced != gene_id_by_expression)
cat("\nPinned candidates — pinned gene ID vs expression-selected gene ID:\n")
print(nb24_override_check |>
        dplyr::select(symbol, gene_group, gene_id_forced, gene_id_by_expression, overridden))
stopifnot(
  "pinned gene IDs must all be present in the TE dataset" =
    all(forced_ids$gene_id_forced %in% te_full$gene_id_clean)
)

# The Plot 2d set was screened on a single ENSG per symbol, so unlike the Plot 2c
# set the pin must AGREE with the expression rule. A disagreement means the symbol
# has picked up a second expressed locus since the screen and the pick needs redoing.
clonable_override <- nb24_override_check |> filter(gene_group == "nb24_clonable")
stopifnot(
  "every clonable candidate must appear in the override check" =
    nrow(clonable_override) == length(clonable_symbols),
  "a clonable candidate's pinned gene ID disagrees with the expression rule" =
    !any(clonable_override$overridden)
)
cat("All", length(clonable_symbols),
    "clonable candidates: pinned gene ID == expression-selected gene ID.\n")

gene_df <- gene_df |> dplyr::select(-gene_id_forced)

stopifnot(
  "All symbols must be found in TE data" =
    all(all_symbols %in% gene_df$symbol),
  "Each symbol must resolve to exactly one gene ID" =
    nrow(gene_df) == length(all_symbols)
)

cat("Gene IDs resolved:\n")
print(gene_df |> dplyr::select(symbol, gene_group, gene_id_clean))
```

# Transcript isoform selection

Select the most-expressed transcript per gene in siCTRL normoxia 1hr and siCTRL hypoxia 1hr, restricting to transcripts with an annotated 5'UTR in GENCODE v49. This avoids selecting transcripts that lack a 5'UTR annotation (e.g. RING1's most abundant isoform ENST00000436128 has no 5'UTR in the txdb). Use the normoxia-dominant transcript; verify it is also dominant in hypoxia. For any inconsistent gene, fall back to the hypoxia-dominant transcript and report it.

```{r isoform_selection}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

# Transcript IDs (clean, version-stripped) with annotated 5'UTR in GENCODE v49
utr5_all         <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr5_tx_versioned <- names(utr5_all)
utr5_tx_clean    <- sub("\\..*", "", utr5_tx_versioned)

metadata <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                     show_col_types = FALSE)

salmon_dir <- here("2025_MDAcells_riboseq", "salmon")

tx2gene_map <- read_tsv(
  here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
  col_names = c("transcript_id", "gene_id"), show_col_types = FALSE
) |>
  filter(str_starts(transcript_id, "ENST"))

load_top_tx <- function(sample_ids, label) {
  files <- file.path(salmon_dir, sample_ids, "quant.sf.gz")
  names(files) <- sample_ids
  files <- files[file.exists(files)]
  cat(label, "— samples found:", length(files), "\n")
  txi <- tximport(files, type = "salmon", tx2gene = tx2gene_map, txOut = TRUE)
  txi$abundance |>
    as.data.frame() |>
    rownames_to_column("transcript_id") |>
    mutate(
      transcript_id_clean = sub("\\..*", "", transcript_id),
      gene_id_clean       = sub("\\..*", "",
        tx2gene_map$gene_id[match(transcript_id, tx2gene_map$transcript_id)])
    ) |>
    mutate(total_tpm = rowSums(across(where(is.numeric)), na.rm = TRUE))
}

nor_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "normoxia", hours == 1) |>
  pull(id)

hyp_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "hypoxia", hours == 1) |>
  pull(id)

nor_abundance <- load_top_tx(nor_ids, "Normoxia siCTRL 1hr")
hyp_abundance <- load_top_tx(hyp_ids, "Hypoxia siCTRL 1hr")

top_tx_per_gene <- function(abund_df) {
  abund_df |>
    filter(gene_id_clean %in% gene_df$gene_id_clean,
           transcript_id_clean %in% utr5_tx_clean) |>
    group_by(gene_id_clean) |>
    dplyr::slice_max(total_tpm, n = 1, with_ties = FALSE) |>
    ungroup() |>
    dplyr::select(gene_id_clean, transcript_id_clean, total_tpm)
}

nor_top <- top_tx_per_gene(nor_abundance) |>
  dplyr::rename(top_tx_nor = transcript_id_clean, tpm_nor = total_tpm)

hyp_top <- top_tx_per_gene(hyp_abundance) |>
  dplyr::rename(top_tx_hyp = transcript_id_clean, tpm_hyp = total_tpm)

isoform_check <- gene_df |>
  left_join(nor_top, by = "gene_id_clean") |>
  left_join(hyp_top, by = "gene_id_clean") |>
  mutate(consistent = !is.na(top_tx_nor) & !is.na(top_tx_hyp) &
                      top_tx_nor == top_tx_hyp)

cat("\nIsoform dominance consistency (normoxia vs hypoxia siCTRL 1hr):\n")
print(isoform_check |>
        dplyr::select(symbol, gene_group, top_tx_nor, tpm_nor, top_tx_hyp, tpm_hyp, consistent))

switched <- isoform_check |> filter(!consistent)
if (nrow(switched) > 0) {
  cat("\nInconsistent dominant isoform — using hypoxia-dominant transcript for:\n")
  print(switched |> dplyr::select(symbol, top_tx_nor, top_tx_hyp))
}

# Use normoxia-dominant; fall back to hypoxia-dominant if inconsistent
tx_selected <- isoform_check |>
  mutate(
    transcript_id_final = if_else(consistent, top_tx_nor, top_tx_hyp)
  ) |>
  dplyr::select(symbol, gene_group, gene_id_clean, transcript_id_final)

target_missing <- tx_selected |>
  filter(gene_group == "target", is.na(transcript_id_final)) |>
  pull(symbol)
stopifnot(
  "All target genes must have a selected transcript" =
    length(target_missing) == 0
)
neg_missing <- tx_selected |>
  filter(gene_group == "negative_control", is.na(transcript_id_final)) |>
  pull(symbol)
if (length(neg_missing) > 0) {
  cat("Negative controls without a 5'UTR-annotated transcript (excluded from pipeline):\n")
  cat(paste(neg_missing, collapse = ", "), "\n")
}

# nb24 candidates are reported-and-excluded rather than hard-failed: three of the
# five pinned loci have NO transcript with an annotated 5'UTR in GENCODE v49,
# because the copy carrying the eIF3d effect is not a normal coding transcript —
# DDR1 ENSG00000230456 is protein_coding_CDS_not_defined, TNKS ENSG00000285372 is
# nonsense_mediated_decay, and PDIA3P1 is a transcribed_processed_pseudogene. With
# no CDS there is no annotated 5'UTR boundary, so no 5'UTR construct can be built
# from these loci. Their other ENSG copy does have UTRs but shows no TE effect
# (DDR1 ENSG00000204580 te_lfc -0.01, TNKS ENSG00000173273 -0.06), so silently
# substituting it would build a reporter for the wrong transcript.
nb24_missing <- tx_selected |>
  filter(gene_group == "nb24_candidate", is.na(transcript_id_final)) |>
  pull(symbol)
nb24_ok <- tx_selected |>
  filter(gene_group == "nb24_candidate", !is.na(transcript_id_final)) |>
  pull(symbol)
cat("\nnb24 candidates WITH a 5'UTR-annotated transcript:",
    if (length(nb24_ok) > 0) paste(nb24_ok, collapse = ", ") else "(none)", "\n")
cat("nb24 candidates WITHOUT one (cannot build a 5'UTR construct):",
    if (length(nb24_missing) > 0) paste(nb24_missing, collapse = ", ") else "(none)", "\n")

# --- Plot 2d clonable set: isoform selection must be exactly reproducible --------
# These five were screened on the rule "MANE Select transcript is the most-abundant
# 5'UTR-annotated isoform in BOTH siCTRL conditions". All three halves of that are
# re-asserted here against independent sources: the salmon-derived dominance above,
# the transcript IDs pinned in `define_genes`, and MANE Select read off the GENCODE
# v49 companion file. Any one of them drifting is a hard failure, not a warning —
# a construct built on a different isoform is a different experiment.
mane_select <- read_csv(
  here("accessories", "human", "mane_select_v49.csv"),
  col_names = c("gene_id_clean", "mane_tx", "mane_symbol"), show_col_types = FALSE
)
stopifnot(
  "MANE file must be one row per gene" = !anyDuplicated(mane_select$gene_id_clean)
)

clonable_iso <- isoform_check |>
  filter(gene_group == "nb24_clonable") |>
  left_join(nb24_clonable |> dplyr::select(symbol, tx_expected), by = "symbol") |>
  left_join(mane_select |> dplyr::select(gene_id_clean, mane_tx), by = "gene_id_clean")

cat("\nClonable set — isoform selection vs pinned expectation and MANE Select:\n")
print(clonable_iso |>
        dplyr::select(symbol, top_tx_nor, tpm_nor, top_tx_hyp, tpm_hyp,
                      consistent, tx_expected, mane_tx))

stopifnot(
  "every clonable candidate must resolve to a 5'UTR-annotated transcript" =
    nrow(clonable_iso) == length(clonable_symbols) &&
    !any(is.na(clonable_iso$top_tx_nor)) && !any(is.na(clonable_iso$top_tx_hyp)),
  "a clonable candidate's dominant isoform differs between normoxia and hypoxia" =
    all(clonable_iso$consistent),
  "a clonable candidate's selected transcript differs from the screened one" =
    all(clonable_iso$top_tx_nor == clonable_iso$tx_expected),
  "a clonable candidate has no MANE Select transcript" =
    !any(is.na(clonable_iso$mane_tx)),
  "a clonable candidate's most-expressed transcript is not the MANE Select one" =
    all(clonable_iso$top_tx_nor == clonable_iso$mane_tx)
)
cat("Clonable set: dominant isoform == pinned transcript == MANE Select, in both",
    "conditions, for all", nrow(clonable_iso), "genes.\n")

tx_selected <- tx_selected |> filter(!is.na(transcript_id_final))
```

# Extract 5' UTR sequences

```{r extract_utr5}
# Map clean transcript IDs to versioned IDs needed by txdb
# (txdb, genome, utr5_all already loaded in isoform_selection chunk)
tx_version_map <- tx_selected |>
  left_join(
    tx2gene_map |>
      mutate(transcript_id_clean = sub("\\..*", "", transcript_id)) |>
      dplyr::select(transcript_id_clean, transcript_id_versioned = transcript_id) |>
      distinct(transcript_id_clean, .keep_all = TRUE),
    by = c("transcript_id_final" = "transcript_id_clean")
  )

target_utrs <- utr5_all[names(utr5_all) %in% tx_version_map$transcript_id_versioned]
utr5_seqs   <- extractTranscriptSeqs(genome, target_utrs)

# Convert to named character vector so indexing with missing names returns NA
# instead of erroring (DNAStringSet subsetting errors on unknown names)
utr5_char <- setNames(as.character(utr5_seqs), names(utr5_seqs))

seq_df <- tx_version_map |>
  mutate(
    utr5_sequence = utr5_char[transcript_id_versioned],
    utr5_length   = nchar(utr5_sequence)
  ) |>
  arrange(match(symbol, all_symbols))

missing_utr5 <- seq_df |> filter(is.na(utr5_sequence)) |> pull(symbol)
if (length(missing_utr5) > 0) {
  cat("WARNING — no annotated 5'UTR in GENCODE v49 for:", paste(missing_utr5, collapse = ", "), "\n")
}

cat("\n5' UTR lengths (bp):\n")
print(seq_df |> dplyr::select(symbol, gene_group, transcript_id_final, utr5_length))
```

# Validation

```{r validate}
stopifnot(
  "No missing 5'UTR sequences for target genes" =
    !any(is.na(seq_df$utr5_sequence[seq_df$gene_group == "target"])),
  "All 5'UTR sequences non-empty" =
    all(nchar(seq_df$utr5_sequence) > 0, na.rm = TRUE),
  "S100A11 5'UTR length in expected range (~89 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "S100A11"], 75, 105),
  "UBL5 5'UTR length in expected range (~78 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "UBL5"], 65, 95),
  "PTMS 5'UTR length in expected range (~343 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "PTMS"], 300, 380),
  "All construct names unique" =
    !anyDuplicated(seq_df$symbol)
)

# --- Plot 2d clonable set: sequence-level validation -----------------------------
clonable_seq <- seq_df |>
  filter(gene_group == "nb24_clonable") |>
  left_join(nb24_clonable |> dplyr::select(symbol, tx_expected, utr5_expected),
            by = "symbol")

cat("\nClonable set — extracted 5'UTR vs pinned expectation:\n")
print(clonable_seq |>
        dplyr::select(symbol, transcript_id_final, tx_expected,
                      utr5_length, utr5_expected))

stopifnot(
  "every clonable candidate must survive to sequence extraction" =
    nrow(clonable_seq) == length(clonable_symbols),
  "a clonable 5'UTR sequence is missing" =
    !any(is.na(clonable_seq$utr5_sequence)),
  "a clonable transcript ID changed between selection and extraction" =
    all(clonable_seq$transcript_id_final == clonable_seq$tx_expected),
  "a clonable 5'UTR length differs from the screened length" =
    all(clonable_seq$utr5_length == clonable_seq$utr5_expected),
  # extractTranscriptSeqs returns unambiguous DNA; an N would mean an assembly gap
  # inside the UTR, which cannot be ordered as a synthesised fragment.
  "a clonable 5'UTR contains a non-ACGT base" =
    all(str_detect(toupper(clonable_seq$utr5_sequence), "^[ACGT]+$")),
  "a clonable 5'UTR is shorter than 40 bp" =
    all(clonable_seq$utr5_length >= 40)
)

# uAUGs are biology, not an error — but they change how a 5'UTR reporter reads, so
# report the count rather than discovering it after cloning.
clonable_seq <- clonable_seq |>
  mutate(n_uAUG = str_count(toupper(utr5_sequence), "ATG"))
cat("\nUpstream AUGs per clonable 5'UTR (reported, not filtered):\n")
print(clonable_seq |> dplyr::select(symbol, utr5_length, n_uAUG))

cat("\nAll validation checks passed.\n")
```

# Assemble constructs

```{r assemble_constructs}
prefix         <- "GTGAACCGTCAGATCAGATC"
reporter_start <- "atgGTCTTCACACTCGAAGATTTCGTTGGGGACTGGCGACAGACAGCCGGCTACAACCTGGACCAAGTCCTTG"
suffix         <- "AACAGGGAGGTGTGTCCAGT"

constructs_df <- seq_df |>
  mutate(
    Name             = symbol,
    prefix           = prefix,
    reporter_start   = reporter_start,
    suffix           = suffix,
    full_construct   = paste0(prefix, utr5_sequence, reporter_start, suffix),
    construct_length = nchar(full_construct)
  ) |>
  dplyr::select(Name, symbol, gene_group, transcript_id_final,
                utr5_length, utr5_sequence,
                prefix, reporter_start, suffix, full_construct, construct_length)

stopifnot(
  "All full_construct values start with expected prefix" =
    all(str_starts(constructs_df$full_construct, prefix)),
  "Benchling names must be unique" =
    !anyDuplicated(constructs_df$Name)
)

cat("Construct lengths (bp):\n")
print(constructs_df |> dplyr::select(Name, gene_group, utr5_length, construct_length))

# The bidirectional vector is linearised with Eco31I (BsaI, GGTCTC) and BsmBI-v2
# (Esp3I, CGTCTC), so an internal site in an inserted 5'UTR would be re-cut. The
# fixed prefix / reporter_start / suffix are site-free by design; only the variable
# 5'UTR can introduce one. Reported, not filtered — a Gibson-style assembly onto the
# 20 bp homology arms tolerates internal sites, a Golden Gate one does not.
count_sites <- function(x) {
  x <- toupper(x)
  str_count(x, "GGTCTC") + str_count(x, "GAGACC") +   # Eco31I / BsaI, both strands
  str_count(x, "CGTCTC") + str_count(x, "GAGACG")     # Esp3I / BsmBI-v2, both strands
}
stopifnot(
  "fixed construct parts must be free of Eco31I / BsmBI sites" =
    count_sites(paste0(prefix, reporter_start, suffix)) == 0
)
constructs_df <- constructs_df |> mutate(n_typeIIs_sites = count_sites(utr5_sequence))
flagged <- constructs_df |> filter(n_typeIIs_sites > 0)
if (nrow(flagged) > 0) {
  cat("\nWARNING — 5'UTRs containing an internal Eco31I/BsmBI site:\n")
  print(flagged |> dplyr::select(Name, gene_group, utr5_length, n_typeIIs_sites))
} else {
  cat("\nNo internal Eco31I/BsmBI sites in any 5'UTR.\n")
}

# For the hand-picked Plot 2d set an internal site is a hard failure rather than a
# warning: these five are going into the bidirectional vector, which is linearised
# with Eco31I and BsmBI-v2, so a site inside the insert would be re-cut during
# assembly. The five larger groups above are pools that get filtered downstream.
clonable_sites <- constructs_df |> filter(gene_group == "nb24_clonable")
stopifnot(
  "every clonable candidate must reach construct assembly" =
    nrow(clonable_sites) == length(clonable_symbols),
  "a clonable 5'UTR contains an internal Eco31I / BsmBI-v2 site" =
    all(clonable_sites$n_typeIIs_sites == 0)
)
cat("Clonable set: all", nrow(clonable_sites),
    "5'UTRs free of Eco31I (GGTCTC/GAGACC) and BsmBI-v2 (CGTCTC/GAGACG) sites.\n")

# Report what the length filter removes rather than dropping it silently — the
# fixed parts total 113 bp, so this cut requires a 5'UTR of >= 187 bp.
dropped_short <- constructs_df |> filter(construct_length < 300)
if (nrow(dropped_short) > 0) {
  cat("\nDropped by the >= 300 bp construct filter:\n")
  print(dropped_short |> dplyr::select(Name, gene_group, utr5_length, construct_length))
}

# The >= 300 bp cut is applied to the POOLS only. The Plot 2d set is five genes
# chosen by name on their TE behaviour, and the fixed parts total 113 bp, so the cut
# would require a 5'UTR of >= 187 bp and would silently drop MEMO1 (84), MAP1LC3B
# (92) and DNAJC3 (117) — three of the five. Dropping a named pick on a length rule
# it was never screened against is a change to the experiment, so the group is
# exempted and the shortfall is reported.
clonable_short <- constructs_df |>
  filter(gene_group == "nb24_clonable", construct_length < 300)
if (nrow(clonable_short) > 0) {
  cat("\nNOTE — clonable picks below the 300 bp pool cut (exempt, still exported):\n")
  print(clonable_short |> dplyr::select(Name, utr5_length, construct_length))
}

constructs_df <- constructs_df |>
  filter(construct_length >= 300 | gene_group == "nb24_clonable")
stopifnot(
  "the length filter dropped a clonable candidate" =
    sum(constructs_df$gene_group == "nb24_clonable") == length(clonable_symbols)
)
cat("\nAfter filtering to >= 300 bp constructs:", nrow(constructs_df), "remain",
    "(", sum(constructs_df$gene_group == "target"), "targets,",
    sum(constructs_df$gene_group == "negative_control"), "neg controls,",
    sum(constructs_df$gene_group == "nb24_candidate"), "nb24 candidates,",
    sum(constructs_df$gene_group == "nb24_clonable"), "clonable picks (exempt) )\n")
```

# Negative control candidate review

Review the table below (printed by `assemble_constructs`), then set `neg_symbols_keep`
to the symbols you want to include in the export. All target genes are always included.

```{r neg_control_review}
cat("Negative control candidates with construct details:\n")
print(
  constructs_df |>
    filter(gene_group == "negative_control") |>
    dplyr::select(symbol, transcript_id_final, utr5_length, construct_length)
)
```

```{r neg_select}
# Update this vector after reviewing the table above.
neg_symbols_keep <- c("ADAT3", "PAN2", "DXO")

neg_candidates <- constructs_df |>
  filter(gene_group == "negative_control") |>
  pull(symbol)
stopifnot(
  "neg_symbols_keep must be a subset of neg controls passing the length filter" =
    all(neg_symbols_keep %in% neg_candidates)
)
cat("Negative controls selected for export:", length(neg_symbols_keep), "\n")
print(neg_symbols_keep)
```

# Final construct selection

```{r final_select}
target_symbols_keep <- c(
  "LAPTM4B", "EIF4G2", "CLIC1", "FUT8",
  "UBE2M", "PIP5K1A", "IWS1", "STYXL1", "PIGBOS1", "SERTAD2", "ZNF146"
)

stopifnot(
  "All target_symbols_keep must be present in constructs_df targets" =
    all(target_symbols_keep %in% constructs_df$symbol[constructs_df$gene_group == "target"])
)

# All nb24 candidates that survived 5'UTR extraction and the length filter are kept.
nb24_symbols_keep <- constructs_df |>
  filter(gene_group == "nb24_candidate") |>
  pull(symbol)

# All five Plot 2d picks are kept — the set was chosen by name, so there is no
# further selection step for it.
clonable_symbols_keep <- constructs_df |>
  filter(gene_group == "nb24_clonable") |>
  pull(symbol)
stopifnot(
  "all five clonable picks must survive to final selection" =
    setequal(clonable_symbols_keep, clonable_symbols)
)

constructs_final <- constructs_df |>
  filter(
    (gene_group == "target" & symbol %in% target_symbols_keep) |
    (gene_group == "negative_control" & symbol %in% neg_symbols_keep) |
    (gene_group == "nb24_candidate" & symbol %in% nb24_symbols_keep) |
    (gene_group == "nb24_clonable" & symbol %in% clonable_symbols_keep)
  )

cat("Final selection:", nrow(constructs_final), "constructs\n")
cat(" ", sum(constructs_final$gene_group == "target"), "targets\n")
cat(" ", sum(constructs_final$gene_group == "negative_control"), "negative controls\n")
cat(" ", sum(constructs_final$gene_group == "nb24_candidate"), "nb24 candidates\n")
cat(" ", sum(constructs_final$gene_group == "nb24_clonable"), "clonable picks\n")
print(constructs_final |> dplyr::select(symbol, gene_group, utr5_length, construct_length))

# Account for every one of the five nb24 candidates: built, or dropped with a reason.
nb24_outcome <- nb24_candidates |>
  mutate(
    outcome = case_when(
      symbol %in% constructs_final$symbol                       ~ "construct built",
      symbol %in% nb24_missing                                  ~ "no annotated 5'UTR at pinned locus",
      symbol %in% dropped_short$symbol                          ~ "5'UTR too short (< 187 bp)",
      TRUE                                                      ~ "excluded (see log above)"
    )
  )
cat("\nnb24 candidate outcomes:\n")
print(nb24_outcome)
stopifnot(
  "every nb24 candidate must be accounted for" =
    nrow(nb24_outcome) == length(nb24_symbols)
)
```


# Plot 2d clonable picks — construct integrity and export

Every check here is a `stopifnot`. The point is that the exported `Bases` string can
be pasted into Benchling and cloned without re-deriving anything by hand, so each
step from gene symbol to final sequence is re-asserted against an independent source
rather than trusted from the step before it.

```{r clonable_integrity}
clonable_final <- constructs_final |>
  filter(gene_group == "nb24_clonable") |>
  left_join(nb24_clonable |> dplyr::select(symbol, gene_id_forced, nb24_group,
                                           tx_expected, utr5_expected),
            by = "symbol") |>
  arrange(match(symbol, clonable_symbols))

stopifnot(
  "clonable export must contain exactly the five picked genes" =
    nrow(clonable_final) == length(clonable_symbols) &&
    setequal(clonable_final$symbol, clonable_symbols),
  "clonable construct names must be unique" =
    !anyDuplicated(clonable_final$Name)
)

# 1. Provenance: transcript and 5'UTR length still equal the screened values.
stopifnot(
  "transcript ID drifted from the screened one" =
    all(clonable_final$transcript_id_final == clonable_final$tx_expected),
  "5'UTR length drifted from the screened one" =
    all(clonable_final$utr5_length == clonable_final$utr5_expected)
)

# 2. Composition: the construct is exactly prefix + 5'UTR + reporter + suffix.
#    Rebuilt independently here rather than reusing `full_construct`, so an edit to
#    the assembly expression cannot pass this check by construction.
rebuilt <- paste0(prefix, clonable_final$utr5_sequence, reporter_start, suffix)
stopifnot(
  "construct does not equal prefix + 5'UTR + reporter_start + suffix" =
    all(clonable_final$full_construct == rebuilt),
  "construct length does not equal 113 + 5'UTR length" =
    all(clonable_final$construct_length ==
        nchar(prefix) + clonable_final$utr5_length +
        nchar(reporter_start) + nchar(suffix)),
  "construct length disagrees with nchar of the sequence" =
    all(clonable_final$construct_length == nchar(clonable_final$full_construct))
)

# 3. Boundaries: the fixed parts sit where they are supposed to, and the 5'UTR
#    appears exactly once — a UTR that happens to repeat the prefix or the reporter
#    start would make the junction ambiguous for Gibson assembly.
stopifnot(
  "construct does not start with the vector prefix" =
    all(str_starts(clonable_final$full_construct, fixed(prefix))),
  "construct does not end with the vector suffix" =
    all(str_ends(clonable_final$full_construct, fixed(suffix))),
  "reporter start does not appear exactly once" =
    all(str_count(clonable_final$full_construct, fixed(reporter_start)) == 1),
  "5'UTR does not appear exactly once in its own construct" =
    all(mapply(function(cst, u) str_count(cst, fixed(u)) == 1,
               clonable_final$full_construct, clonable_final$utr5_sequence)),
  "5'UTR is not immediately followed by the reporter start" =
    all(mapply(function(cst, u) str_detect(cst, fixed(paste0(u, reporter_start))),
               clonable_final$full_construct, clonable_final$utr5_sequence)),
  "5'UTR is not immediately preceded by the prefix" =
    all(mapply(function(cst, u) str_detect(cst, fixed(paste0(prefix, u))),
               clonable_final$full_construct, clonable_final$utr5_sequence))
)

# 4. Synthesisability: unambiguous bases only, and no internal Eco31I / BsmBI-v2
#    site anywhere in the ordered fragment (not just in the variable 5'UTR — a site
#    could also be created across a junction by the last bases of the UTR).
stopifnot(
  "construct contains a non-ACGT base" =
    all(str_detect(toupper(clonable_final$full_construct), "^[ACGT]+$")),
  "construct contains an internal Eco31I / BsmBI-v2 site" =
    all(count_sites(clonable_final$full_construct) == 0)
)

# 5. Start codon placement. `reporter_start` is a 73 bp overlap into the vector's
#    reporter ORF, which the vector continues, so it is deliberately NOT a whole
#    number of codons and no frame assertion is made on it. What must hold is that
#    its ATG is the first ATG downstream of the inserted 5'UTR and sits at exactly
#    prefix + utr5_length + 1 — that offset is what makes the 5'UTR the 5'UTR.
atg_pos <- str_locate(clonable_final$full_construct,
                      fixed(reporter_start))[, "start"]
stopifnot(
  "reporter start not located in construct" = !any(is.na(atg_pos)),
  "reporter ATG is not immediately after the 5'UTR" =
    all(atg_pos == nchar(prefix) + clonable_final$utr5_length + 1),
  "reporter_start does not begin with ATG" =
    str_starts(toupper(reporter_start), "ATG")
)

# 5b. The vector arms themselves are pinned. An accidental edit to any of the three
#     fixed parts would silently change every construct in this notebook, including
#     the ones already ordered off the earlier export.
stopifnot(
  "vector prefix has changed" =
    prefix == "GTGAACCGTCAGATCAGATC",
  "reporter_start has changed" =
    reporter_start == paste0("atgGTCTTCACACTCGAAGATTTCGTTGGGGACTGGCGACAGACAGCC",
                             "GGCTACAACCTGGACCAAGTCCTTG"),
  "vector suffix has changed" =
    suffix == "AACAGGGAGGTGTGTCCAGT",
  "fixed part lengths have changed" =
    nchar(prefix) == 20 && nchar(reporter_start) == 73 && nchar(suffix) == 20
)

# 6. Cross-check against the biology the picks were made on. Re-read the two
#    contrasts and re-assert the condition-specific definitions from notebook 24,
#    so a construct can never be exported for a gene that no longer qualifies.
#    Direction check first (CLAUDE.md): te_lfc is log2(sictrl / si3d), verified by
#    EIF3D's own rna_lfc being strongly positive.
nor24 <- read_csv(here("output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                  show_col_types = FALSE)
hyp24 <- read_csv(here("output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                  show_col_types = FALSE)
stopifnot(
  "EIF3D rna_lfc not strongly positive — si3d contrast direction may have flipped" =
    hyp24$rna_lfc[hyp24$symbol == "EIF3D"] > 2
)

te_chk <- clonable_final |>
  dplyr::select(symbol, gene_id_clean = gene_id_forced, nb24_group) |>
  left_join(nor24 |> mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
              dplyr::select(gene_id_clean, nor_lfc = te_lfc, nor_padj = te_padj),
            by = "gene_id_clean") |>
  left_join(hyp24 |> mutate(gene_id_clean = sub("\\..*", "", gene_id)) |>
              dplyr::select(gene_id_clean, hyp_lfc = te_lfc, hyp_padj = te_padj),
            by = "gene_id_clean") |>
  mutate(
    qualifies = case_when(
      nb24_group == "Hypoxia-specific" ~
        !is.na(hyp_padj) & hyp_padj < 0.05 & hyp_lfc > 0.5 &
        (is.na(nor_padj) | nor_padj >= 0.05) & !is.na(nor_lfc) & nor_lfc < 0.5,
      nb24_group == "Normoxia-specific" ~
        !is.na(nor_padj) & nor_padj < 0.05 & nor_lfc > 0.5 &
        (is.na(hyp_padj) | hyp_padj >= 0.05) & !is.na(hyp_lfc) & hyp_lfc < 0.5
    )
  )
cat("Condition-specific definition re-checked against the source contrasts:\n")
print(as.data.frame(te_chk |> mutate(across(where(is.numeric), ~round(.x, 4)))))
stopifnot(
  "a clonable pick no longer meets its condition-specific definition" =
    all(te_chk$qualifies)
)

cat("\nAll clonable construct integrity checks passed.\n")

clonable_report <- clonable_final |>
  dplyr::select(Name, nb24_group, gene_id_forced, transcript_id_final,
                utr5_length, construct_length, n_typeIIs_sites)
print(as.data.frame(clonable_report))
```

## Sequences

```{r clonable_sequences}
for (i in seq_len(nrow(clonable_final))) {
  cat(">", clonable_final$Name[i], " | ", clonable_final$nb24_group[i],
      " | ", clonable_final$transcript_id_final[i],
      " | 5'UTR ", clonable_final$utr5_length[i], " bp",
      " | construct ", clonable_final$construct_length[i], " bp\n", sep = "")
  cat(clonable_final$full_construct[i], "\n\n")
}
```

## Export

Written as its own pair of files rather than appended to `65_benchling_import.csv`,
so the constructs already ordered off that file are not disturbed. The same rows are
also present in `65_constructs_all.csv` under `gene_group == "nb24_clonable"`.

```{r clonable_export}
clonable_out <- clonable_final |>
  dplyr::select(Name, nb24_group, gene_id = gene_id_forced,
                transcript_id = transcript_id_final,
                utr5_length, utr5_sequence,
                prefix, reporter_start, suffix,
                full_construct, construct_length, n_typeIIs_sites)

write_csv(clonable_out,
          here("output", "predictive_modeling", "65_constructs_nb24_clonable.csv"))
cat("Saved:", nrow(clonable_out),
    "rows to output/predictive_modeling/65_constructs_nb24_clonable.csv\n")

clonable_benchling <- clonable_final |>
  dplyr::select(Name, Bases = full_construct, construct_length)

# A clonable Name must not equal any other construct's Name, or the two would
# collide as Benchling sequence entries. Checked in memory against the other groups
# in constructs_final rather than against the file on disk: the file is rewritten
# further down and an earlier run of this notebook can have left the clonable rows
# in it, which is not a collision but would look like one.
other_names <- constructs_final |>
  filter(gene_group != "nb24_clonable") |>
  pull(Name)
stopifnot(
  "clonable construct name collides with another construct in this notebook" =
    length(intersect(clonable_benchling$Name, other_names)) == 0
)
cat("No name collisions with", length(other_names), "other constructs.\n")

# Separately: report clonable rows left in the main import file by an earlier run.
# The main export below excludes this group, so they will be cleared on write.
prior_import_path <- here("output", "predictive_modeling", "65_benchling_import.csv")
if (file.exists(prior_import_path)) {
  stale <- intersect(read_csv(prior_import_path, show_col_types = FALSE)$Name,
                     clonable_symbols)
  if (length(stale) > 0) {
    cat("NOTE — clearing stale clonable rows from the main import file:",
        paste(stale, collapse = ", "), "\n")
  }
}

write_csv(clonable_benchling,
          here("output", "predictive_modeling", "65_benchling_import_nb24_clonable.csv"))
cat("Saved Benchling import:", nrow(clonable_benchling),
    "constructs to output/predictive_modeling/65_benchling_import_nb24_clonable.csv\n")

# Round-trip: re-read what was written and confirm it matches what is in memory.
roundtrip <- read_csv(
  here("output", "predictive_modeling", "65_benchling_import_nb24_clonable.csv"),
  show_col_types = FALSE)
stopifnot(
  "written Benchling file does not round-trip" =
    nrow(roundtrip) == nrow(clonable_benchling) &&
    all(roundtrip$Name  == clonable_benchling$Name) &&
    all(roundtrip$Bases == clonable_benchling$Bases) &&
    all(nchar(roundtrip$Bases) == roundtrip$construct_length)
)
cat("Round-trip check passed.\n")

print(clonable_benchling |>
        mutate(Bases_preview = paste0(str_sub(Bases, 1, 40), "...")) |>
        dplyr::select(Name, construct_length, Bases_preview) |>
        as.data.frame())
```

# CLIC1 and ADAT3 5' UTR sequences

```{r clic1_adat3_utr5}
focal_genes <- c("CLIC1", "ADAT3")
focal_utrs  <- constructs_final |>
  filter(symbol %in% focal_genes) |>
  dplyr::select(symbol, gene_group, transcript_id_final, utr5_length, utr5_sequence)

stopifnot(
  "CLIC1 and ADAT3 must both be present in constructs_final" =
    all(focal_genes %in% focal_utrs$symbol)
)

for (i in seq_len(nrow(focal_utrs))) {
  cat(focal_utrs$symbol[i], "(", focal_utrs$gene_group[i], ") |",
      focal_utrs$transcript_id_final[i], "| length:", focal_utrs$utr5_length[i], "bp\n")
  cat(focal_utrs$utr5_sequence[i], "\n\n")
}
```

# Export

```{r export}
write_csv(
  constructs_final,
  here("output", "predictive_modeling", "65_constructs_all.csv")
)
cat("Saved construct table:",
    nrow(constructs_final), "rows to output/predictive_modeling/65_constructs_all.csv\n")

# The Plot 2d picks are excluded from THIS import file and get their own
# (`65_benchling_import_nb24_clonable.csv`, written above). Re-uploading this file
# should not re-import the constructs already in Benchling from the earlier run, so
# it is kept to exactly the rows it had before that group existed. They are still in
# 65_constructs_all.csv, which is the full record rather than an upload manifest.
benchling_df <- constructs_final |>
  filter(gene_group != "nb24_clonable") |>
  dplyr::select(Name, Bases = full_construct, construct_length)
stopifnot(
  "main Benchling import must not contain the clonable picks" =
    !any(benchling_df$Name %in% clonable_symbols)
)

write_csv(
  benchling_df,
  here("output", "predictive_modeling", "65_benchling_import.csv")
)
cat("Saved Benchling import:", nrow(benchling_df),
    "constructs to output/predictive_modeling/65_benchling_import.csv\n")

cat("\nBenchling import preview:\n")
print(benchling_df |>
        mutate(Bases_preview = paste0(str_sub(Bases, 1, 40), "...")))
```

```{r}

```