---
title: "65 — Benchling Construct Design: 5' UTR Reporter Constructs"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(tximport)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
```

# Gene lists

```{r define_genes}
# Previous hardcoded list (kept for sanity check below):
# target_symbols <- c(
#   "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
#   "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
# )

# Load candidates from nb11 Benchling export (Exclusive eIF3d hypoxia-specific targets,
# ribo-only changes, RIP-bound, isoform-consistent)
benchling_candidates <- read_csv(
  here("ripseq", "output", "11_benchling_import.csv"),
  show_col_types = FALSE
)
target_symbols <- sub(" 5UTR.*", "", benchling_candidates$Name)

# Sanity check: previously hardcoded targets must all be present
prior_targets <- c(
  "S100A11", "RAB7A", "BTF3", "CHMP4C", "PTMS",
  "UBL5", "COQ10A", "MAIP1", "EIF1B", "TBC1D8B"
)
missing_from_csv <- setdiff(prior_targets, target_symbols)
stopifnot(
  "All previously hardcoded target symbols must be present in 11_benchling_import.csv" =
    length(missing_from_csv) == 0
)
cat("Loaded", length(target_symbols), "target symbols from 11_benchling_import.csv\n")
cat("All", length(prior_targets), "prior hardcoded symbols confirmed present.\n")

neg_ctrl <- read_csv(
  here("output", "predictive_modeling", "negative_control_genes_strict.csv"),
  show_col_types = FALSE
)
neg_symbols <- neg_ctrl$symbol
cat("Loaded", length(neg_symbols), "negative control candidates from negative_control_genes_strict.csv\n")

all_symbols <- c(target_symbols, neg_symbols)

gene_groups <- tibble(
  symbol     = all_symbols,
  gene_group = c(rep("target", length(target_symbols)),
                 rep("negative_control", length(neg_symbols)))
)
```

# Gene annotation lookup

Restrict to gene IDs present in the MDA-MB-231 TE dataset. This resolves ambiguous
symbols like RING1, which maps to multiple ENSG IDs in GENCODE v49 (alternative
loci in the MHC region) but only one of which (ENSG00000204227) is the expressed
canonical locus in our data.

```{r gene_annotation}
# Load TE data to select the expressed gene ID for each symbol.
# complete_te_list_1hr.csv columns: symbol, gene_id, TE columns, biotype,
# RNA count columns ("_rna_"), ribosome footprint columns ("_ribo_").
te_full <- read_csv(here("output", "complete_te_list_1hr.csv"),
                    show_col_types = FALSE) |>
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    total_rna     = rowSums(across(contains("_rna_")), na.rm = TRUE)
  )

# For symbols with multiple expressed ENSG IDs (e.g. RING1 maps to two loci in
# GENCODE v49), keep the one with higher total RNA expression — that is the
# canonical expressed locus (ENSG00000204227 for RING1, which is also the ID
# used in negative_control_genes.csv; the other locus ENSG00000235107 is
# actually upregulated in si3d and is not a suitable negative control).
gene_id_map <- te_full |>
  dplyr::select(symbol, gene_id_clean, total_rna) |>
  group_by(symbol) |>
  dplyr::slice_max(total_rna, n = 1, with_ties = FALSE) |>
  ungroup()

gene_df <- gene_groups |>
  left_join(gene_id_map |> dplyr::select(symbol, gene_id_clean), by = "symbol")

stopifnot(
  "All symbols must be found in TE data" =
    all(all_symbols %in% gene_df$symbol),
  "Each symbol must resolve to exactly one gene ID" =
    nrow(gene_df) == length(all_symbols)
)

cat("Gene IDs resolved:\n")
print(gene_df |> dplyr::select(symbol, gene_group, gene_id_clean))
```

# Transcript isoform selection

Select the most-expressed transcript per gene in siCTRL normoxia 1hr and siCTRL hypoxia 1hr, restricting to transcripts with an annotated 5'UTR in GENCODE v49. This avoids selecting transcripts that lack a 5'UTR annotation (e.g. RING1's most abundant isoform ENST00000436128 has no 5'UTR in the txdb). Use the normoxia-dominant transcript; verify it is also dominant in hypoxia. For any inconsistent gene, fall back to the hypoxia-dominant transcript and report it.

```{r isoform_selection}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

# Transcript IDs (clean, version-stripped) with annotated 5'UTR in GENCODE v49
utr5_all         <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr5_tx_versioned <- names(utr5_all)
utr5_tx_clean    <- sub("\\..*", "", utr5_tx_versioned)

metadata <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                     show_col_types = FALSE)

salmon_dir <- here("2025_MDAcells_riboseq", "salmon")

tx2gene_map <- read_tsv(
  here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
  col_names = c("transcript_id", "gene_id"), show_col_types = FALSE
) |>
  filter(str_starts(transcript_id, "ENST"))

load_top_tx <- function(sample_ids, label) {
  files <- file.path(salmon_dir, sample_ids, "quant.sf.gz")
  names(files) <- sample_ids
  files <- files[file.exists(files)]
  cat(label, "— samples found:", length(files), "\n")
  txi <- tximport(files, type = "salmon", tx2gene = tx2gene_map, txOut = TRUE)
  txi$abundance |>
    as.data.frame() |>
    rownames_to_column("transcript_id") |>
    mutate(
      transcript_id_clean = sub("\\..*", "", transcript_id),
      gene_id_clean       = sub("\\..*", "",
        tx2gene_map$gene_id[match(transcript_id, tx2gene_map$transcript_id)])
    ) |>
    mutate(total_tpm = rowSums(across(where(is.numeric)), na.rm = TRUE))
}

nor_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "normoxia", hours == 1) |>
  pull(id)

hyp_ids <- metadata |>
  filter(type == "Input_Ribo_RNAseq", treatment == "siCTRL",
         condition == "hypoxia", hours == 1) |>
  pull(id)

nor_abundance <- load_top_tx(nor_ids, "Normoxia siCTRL 1hr")
hyp_abundance <- load_top_tx(hyp_ids, "Hypoxia siCTRL 1hr")

top_tx_per_gene <- function(abund_df) {
  abund_df |>
    filter(gene_id_clean %in% gene_df$gene_id_clean,
           transcript_id_clean %in% utr5_tx_clean) |>
    group_by(gene_id_clean) |>
    dplyr::slice_max(total_tpm, n = 1, with_ties = FALSE) |>
    ungroup() |>
    dplyr::select(gene_id_clean, transcript_id_clean, total_tpm)
}

nor_top <- top_tx_per_gene(nor_abundance) |>
  dplyr::rename(top_tx_nor = transcript_id_clean, tpm_nor = total_tpm)

hyp_top <- top_tx_per_gene(hyp_abundance) |>
  dplyr::rename(top_tx_hyp = transcript_id_clean, tpm_hyp = total_tpm)

isoform_check <- gene_df |>
  left_join(nor_top, by = "gene_id_clean") |>
  left_join(hyp_top, by = "gene_id_clean") |>
  mutate(consistent = !is.na(top_tx_nor) & !is.na(top_tx_hyp) &
                      top_tx_nor == top_tx_hyp)

cat("\nIsoform dominance consistency (normoxia vs hypoxia siCTRL 1hr):\n")
print(isoform_check |>
        dplyr::select(symbol, gene_group, top_tx_nor, tpm_nor, top_tx_hyp, tpm_hyp, consistent))

switched <- isoform_check |> filter(!consistent)
if (nrow(switched) > 0) {
  cat("\nInconsistent dominant isoform — using hypoxia-dominant transcript for:\n")
  print(switched |> dplyr::select(symbol, top_tx_nor, top_tx_hyp))
}

# Use normoxia-dominant; fall back to hypoxia-dominant if inconsistent
tx_selected <- isoform_check |>
  mutate(
    transcript_id_final = if_else(consistent, top_tx_nor, top_tx_hyp)
  ) |>
  dplyr::select(symbol, gene_group, gene_id_clean, transcript_id_final)

target_missing <- tx_selected |>
  filter(gene_group == "target", is.na(transcript_id_final)) |>
  pull(symbol)
stopifnot(
  "All target genes must have a selected transcript" =
    length(target_missing) == 0
)
neg_missing <- tx_selected |>
  filter(gene_group == "negative_control", is.na(transcript_id_final)) |>
  pull(symbol)
if (length(neg_missing) > 0) {
  cat("Negative controls without a 5'UTR-annotated transcript (excluded from pipeline):\n")
  cat(paste(neg_missing, collapse = ", "), "\n")
  tx_selected <- tx_selected |> filter(!is.na(transcript_id_final))
}
```

# Extract 5' UTR sequences

```{r extract_utr5}
# Map clean transcript IDs to versioned IDs needed by txdb
# (txdb, genome, utr5_all already loaded in isoform_selection chunk)
tx_version_map <- tx_selected |>
  left_join(
    tx2gene_map |>
      mutate(transcript_id_clean = sub("\\..*", "", transcript_id)) |>
      dplyr::select(transcript_id_clean, transcript_id_versioned = transcript_id) |>
      distinct(transcript_id_clean, .keep_all = TRUE),
    by = c("transcript_id_final" = "transcript_id_clean")
  )

target_utrs <- utr5_all[names(utr5_all) %in% tx_version_map$transcript_id_versioned]
utr5_seqs   <- extractTranscriptSeqs(genome, target_utrs)

# Convert to named character vector so indexing with missing names returns NA
# instead of erroring (DNAStringSet subsetting errors on unknown names)
utr5_char <- setNames(as.character(utr5_seqs), names(utr5_seqs))

seq_df <- tx_version_map |>
  mutate(
    utr5_sequence = utr5_char[transcript_id_versioned],
    utr5_length   = nchar(utr5_sequence)
  ) |>
  arrange(match(symbol, all_symbols))

missing_utr5 <- seq_df |> filter(is.na(utr5_sequence)) |> pull(symbol)
if (length(missing_utr5) > 0) {
  cat("WARNING — no annotated 5'UTR in GENCODE v49 for:", paste(missing_utr5, collapse = ", "), "\n")
}

cat("\n5' UTR lengths (bp):\n")
print(seq_df |> dplyr::select(symbol, gene_group, transcript_id_final, utr5_length))
```

# Validation

```{r validate}
stopifnot(
  "No missing 5'UTR sequences for target genes" =
    !any(is.na(seq_df$utr5_sequence[seq_df$gene_group == "target"])),
  "All 5'UTR sequences non-empty" =
    all(nchar(seq_df$utr5_sequence) > 0, na.rm = TRUE),
  "S100A11 5'UTR length in expected range (~89 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "S100A11"], 75, 105),
  "UBL5 5'UTR length in expected range (~78 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "UBL5"], 65, 95),
  "PTMS 5'UTR length in expected range (~343 bp)" =
    between(seq_df$utr5_length[seq_df$symbol == "PTMS"], 300, 380),
  "All construct names unique" =
    !anyDuplicated(seq_df$symbol)
)

cat("All validation checks passed.\n")
```

# Assemble constructs

```{r assemble_constructs}
prefix         <- "GTGAACCGTCAGATCAGATC"
reporter_start <- "atgGTCTTCACACTCGAAGATTTCGTTGGGGACTGGCGACAGACAGCCGGCTACAACCTGGACCAAGTCCTTG"
suffix         <- "AACAGGGAGGTGTGTCCAGT"

constructs_df <- seq_df |>
  mutate(
    Name             = symbol,
    prefix           = prefix,
    reporter_start   = reporter_start,
    suffix           = suffix,
    full_construct   = paste0(prefix, utr5_sequence, reporter_start, suffix),
    construct_length = nchar(full_construct)
  ) |>
  dplyr::select(Name, symbol, gene_group, transcript_id_final,
                utr5_length, utr5_sequence,
                prefix, reporter_start, suffix, full_construct, construct_length)

stopifnot(
  "All full_construct values start with expected prefix" =
    all(str_starts(constructs_df$full_construct, prefix)),
  "Benchling names must be unique" =
    !anyDuplicated(constructs_df$Name)
)

cat("Construct lengths (bp):\n")
print(constructs_df |> dplyr::select(Name, gene_group, utr5_length, construct_length))

constructs_df <- constructs_df |> filter(construct_length >= 300)
cat("\nAfter filtering to >= 300 bp constructs:", nrow(constructs_df), "remain",
    "(", sum(constructs_df$gene_group == "target"), "targets,",
    sum(constructs_df$gene_group == "negative_control"), "neg controls )\n")
```

# Negative control candidate review

Review the table below (printed by `assemble_constructs`), then set `neg_symbols_keep`
to the symbols you want to include in the export. All target genes are always included.

```{r neg_control_review}
cat("Negative control candidates with construct details:\n")
print(
  constructs_df |>
    filter(gene_group == "negative_control") |>
    dplyr::select(symbol, transcript_id_final, utr5_length, construct_length)
)
```

```{r neg_select}
# Update this vector after reviewing the table above.
neg_symbols_keep <- c("ADAT3", "PAN2", "DXO")

neg_candidates <- constructs_df |>
  filter(gene_group == "negative_control") |>
  pull(symbol)
stopifnot(
  "neg_symbols_keep must be a subset of neg controls passing the length filter" =
    all(neg_symbols_keep %in% neg_candidates)
)
cat("Negative controls selected for export:", length(neg_symbols_keep), "\n")
print(neg_symbols_keep)
```

# Final construct selection

```{r final_select}
target_symbols_keep <- c(
  "LAPTM4B", "EIF4G2", "CLIC1", "FUT8",
  "UBE2M", "PIP5K1A", "IWS1", "STYXL1", "PIGBOS1", "SERTAD2", "ZNF146"
)

stopifnot(
  "All target_symbols_keep must be present in constructs_df targets" =
    all(target_symbols_keep %in% constructs_df$symbol[constructs_df$gene_group == "target"])
)

constructs_final <- constructs_df |>
  filter(
    (gene_group == "target" & symbol %in% target_symbols_keep) |
    (gene_group == "negative_control" & symbol %in% neg_symbols_keep)
  )

cat("Final selection:", nrow(constructs_final), "constructs\n")
cat(" ", sum(constructs_final$gene_group == "target"), "targets\n")
cat(" ", sum(constructs_final$gene_group == "negative_control"), "negative controls\n")
print(constructs_final |> dplyr::select(symbol, gene_group, utr5_length, construct_length))
```

# Export

```{r export}
write_csv(
  constructs_final,
  here("output", "predictive_modeling", "65_constructs_all.csv")
)
cat("Saved construct table:",
    nrow(constructs_final), "rows to output/predictive_modeling/65_constructs_all.csv\n")

benchling_df <- constructs_final |>
  dplyr::select(Name, Bases = full_construct, construct_length)

write_csv(
  benchling_df,
  here("output", "predictive_modeling", "65_benchling_import.csv")
)
cat("Saved Benchling import:", nrow(benchling_df),
    "constructs to output/predictive_modeling/65_benchling_import.csv\n")

cat("\nBenchling import preview:\n")
print(benchling_df |>
        mutate(Bases_preview = paste0(str_sub(Bases, 1, 40), "...")))
```

```{r}

```