---
title: "66 - RNA-level changes upon eIF3d and eIF3e knockdown"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Visualize the **RNA-level** (not TE) changes caused by eIF3d and eIF3e knockdown in
MDA-MB-231 cells, across both oxygen conditions (normoxia, hypoxia) and all three
timepoints (1hr, 4hr, 24hr). All values are read from the pre-computed DESeq2
outputs `output/translation_categories_si3{d,e}_vs_sictrl_{condition}_{timepoint}.csv`
(written by `code/eif3e_eif3d_normoxia_and_hypoxia.Rmd`); nothing is recomputed here.

**Direction convention.** Despite the file names, the DESeq2 contrast in the source
notebook is run with the knockdown as the reference (`control_condition = "si3d"` /
`"si3e"`, `test_condition = "sictrl"`), so the reported coefficient is
log2(sictrl / knockdown) for both knockdowns:

- `rna_lfc > 0` = sictrl RNA > knockdown RNA = knockdown **reduces** the mRNA level
- `rna_lfc < 0` = knockdown **increases** the mRNA level

The knockdown target itself is therefore expected at strongly positive `rna_lfc`,
which is used as the biological spot-check below.

**Global shift caveat.** These contrasts are spike-in normalized, so the median
`rna_lfc` is not zero (it ranges from ~0.1 to ~0.8 depending on condition and
timepoint) -- eIF3d/eIF3e knockdown lowers total mRNA per cell. The raw
distributions are plotted as-is with the observed median drawn as a dashed line,
and the median is reported in every summary. Between-gene-set comparisons within a
single contrast (the gene-set section) are unaffected by this shift, since both
groups carry it equally.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
```

# Parameters

```{r params}
KNOCKDOWNS <- c("si3d", "si3e")
CONDITIONS <- c("normoxia", "hypoxia")
TIMEPOINTS <- c("1hr", "4hr", "24hr")

LFC_CUTOFF  <- 0.5   # project-wide fold-change standard
PADJ_CUTOFF <- 0.05

# Project convention: eIF3d = Okabe-Ito orange, eIF3e = Okabe-Ito light blue
kd_colors <- c("si3d" = "#E69F00", "si3e" = "#56B4E9")

group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "black",
  "All other genes"        = "grey70"
)

# Up/down direction colors, kept distinct from the eIF3d/eIF3e colors above
dir_colors <- c("mRNA down in knockdown" = "#0072B2",
                "mRNA up in knockdown"   = "#CC79A7")
```

# Load RNA Results

```{r load_rna}
file_grid <- expand_grid(
  knockdown = KNOCKDOWNS,
  condition = CONDITIONS,
  timepoint = TIMEPOINTS
) %>%
  mutate(path = here("output", paste0("translation_categories_", knockdown,
                                      "_vs_sictrl_", condition, "_",
                                      timepoint, ".csv")))

stopifnot("Missing translation_categories input file" = all(file.exists(file_grid$path)))

read_one <- function(p) {
  read_csv(p, show_col_types = FALSE) %>%
    dplyr::select(gene_id, symbol, rna_lfc, rna_padj,
                  ribo_lfc, ribo_padj, te_lfc, te_padj, category)
}

rna_all <- file_grid %>%
  mutate(data = map(path, read_one)) %>%
  dplyr::select(-path) %>%
  unnest(data) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    knockdown     = factor(knockdown, levels = KNOCKDOWNS),
    condition     = factor(condition, levels = CONDITIONS),
    timepoint     = factor(timepoint, levels = TIMEPOINTS),
    contrast      = factor(paste(condition, timepoint),
                           levels = paste(rep(CONDITIONS, each = length(TIMEPOINTS)),
                                          TIMEPOINTS))
  )

cat("Loaded", nrow(rna_all), "rows across",
    n_distinct(rna_all$knockdown) * n_distinct(rna_all$contrast), "contrasts\n")
print(rna_all %>% dplyr::count(knockdown, contrast))
```

# Validation

## Computational checks

```{r checks_computational}
expected_cols <- c("gene_id", "gene_id_clean", "symbol", "rna_lfc", "rna_padj",
                   "ribo_lfc", "te_lfc", "knockdown", "condition", "timepoint")
stopifnot(all(expected_cols %in% colnames(rna_all)))

# Log-fold changes finite where present
stopifnot("Non-finite rna_lfc" = all(is.finite(rna_all$rna_lfc[!is.na(rna_all$rna_lfc)])))
stopifnot("Non-finite ribo_lfc" = all(is.finite(rna_all$ribo_lfc[!is.na(rna_all$ribo_lfc)])))

# Adjusted p-values are probabilities
stopifnot("rna_padj outside [0,1]" =
  all(rna_all$rna_padj >= 0 & rna_all$rna_padj <= 1, na.rm = TRUE))

# Every contrast covers the same gene universe
gene_counts <- rna_all %>% dplyr::count(knockdown, contrast)
stopifnot("Unequal gene universe across contrasts" = n_distinct(gene_counts$n) == 1)

cat("All computational checks passed;", unique(gene_counts$n), "genes per contrast\n")
```

## Biological spot-check: knockdown targets are depleted

`EIF3D` mRNA must be strongly reduced in every si3d contrast and `EIF3E` mRNA in
every si3e contrast. Under the log2(sictrl / knockdown) convention, that means a
large **positive** `rna_lfc` for the targeted gene.

```{r checks_biological}
kd_target <- c(si3d = "EIF3D", si3e = "EIF3E")

target_check <- rna_all %>%
  filter(symbol == kd_target[as.character(knockdown)]) %>%
  dplyr::select(knockdown, contrast, symbol, rna_lfc, rna_padj) %>%
  arrange(knockdown, contrast)

print(target_check)

stopifnot("Knockdown target mRNA not depleted (expected rna_lfc > 2)" =
  all(target_check$rna_lfc > 2))
stopifnot("Knockdown target mRNA depletion not significant" =
  all(target_check$rna_padj < 0.01))

# The non-targeted paralog must be far less affected than the target
paralog_check <- rna_all %>%
  filter(symbol %in% c("EIF3D", "EIF3E")) %>%
  mutate(is_target = symbol == kd_target[as.character(knockdown)]) %>%
  group_by(knockdown, contrast) %>%
  summarise(target_lfc  = rna_lfc[is_target],
            paralog_lfc = rna_lfc[!is_target], .groups = "drop")

stopifnot("Paralog depleted as strongly as the knockdown target" =
  all(paralog_check$target_lfc > paralog_check$paralog_lfc + 1.5))

cat("Knockdown spot-check passed: target mRNA depleted in all",
    nrow(target_check), "contrasts, paralog spared\n")
```

```{r plot_knockdown_validation, fig.width = 7, fig.height = 3}
p_kd <- ggplot(target_check, aes(x = contrast, y = rna_lfc, fill = knockdown)) +
  geom_col(width = 0.7) +
  facet_wrap(~ knockdown, labeller = labeller(
    knockdown = c(si3d = "si3d (EIF3D mRNA)", si3e = "si3e (EIF3E mRNA)"))) +
  scale_fill_manual(values = kd_colors, guide = "none") +
  labs(title = "Knockdown validation: target mRNA depletion",
       x = NULL, y = "RNA log2FC (sictrl / knockdown)") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_kd)
ggsave(here("plots", "66_rna_knockdown_validation.pdf"), p_kd, width = 7, height = 3)
```

# Summary of Significant RNA Changes

```{r summary_table}
rna_summary <- rna_all %>%
  group_by(knockdown, condition, timepoint, contrast) %>%
  summarise(
    n_genes      = sum(!is.na(rna_lfc)),
    median_lfc   = median(rna_lfc, na.rm = TRUE),
    n_down_in_kd = sum(rna_padj < PADJ_CUTOFF & rna_lfc >  LFC_CUTOFF, na.rm = TRUE),
    n_up_in_kd   = sum(rna_padj < PADJ_CUTOFF & rna_lfc < -LFC_CUTOFF, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pct_changed = 100 * (n_down_in_kd + n_up_in_kd) / n_genes)

print(rna_summary, n = 20)

write_csv(rna_summary, here("output", "66_rna_change_summary.csv"))
```

```{r plot_sig_counts, fig.width = 8, fig.height = 4}
counts_long <- rna_summary %>%
  dplyr::select(knockdown, condition, timepoint, n_down_in_kd, n_up_in_kd) %>%
  pivot_longer(c(n_down_in_kd, n_up_in_kd),
               names_to = "direction", values_to = "n") %>%
  mutate(direction = recode(direction,
                            n_down_in_kd = "mRNA down in knockdown",
                            n_up_in_kd   = "mRNA up in knockdown"))

p_counts <- ggplot(counts_long, aes(x = timepoint, y = n, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_text(aes(label = n), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.8) +
  facet_grid(knockdown ~ condition) +
  scale_fill_manual(values = dir_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = paste0("Significant RNA changes (padj < ", PADJ_CUTOFF,
                      ", |log2FC| > ", LFC_CUTOFF, ")"),
       x = "Timepoint", y = "Number of genes", fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_counts)
ggsave(here("plots", "66_rna_sig_gene_counts.pdf"), p_counts, width = 8, height = 4)
```

# Gene Sets

```{r load_genesets}
promotes <- read_csv(here("output", "genesets",
                          "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE)
neg_pool <- read_csv(here("output", "predictive_modeling",
                          "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

promotes_ids <- unique(promotes$ensembl_gene)
neg_pool_ids <- setdiff(unique(neg_pool$gene_id_clean), promotes_ids)

universe <- unique(rna_all$gene_id_clean)
promotes_ids <- intersect(promotes_ids, universe)
neg_pool_ids <- intersect(neg_pool_ids, universe)

# Size-matched negative control set (project convention: set.seed(9))
set.seed(9)
neg_ids <- sample(neg_pool_ids, min(length(promotes_ids), length(neg_pool_ids)))

cat("Promoted genes in universe:", length(promotes_ids), "\n")
cat("Negative control pool:", length(neg_pool_ids),
    "-> size-matched to", length(neg_ids), "\n")

stopifnot("Gene sets overlap" = length(intersect(promotes_ids, neg_ids)) == 0)
stopifnot("Promoted gene set is not the expected ~1604 mRNAs" =
  between(length(promotes_ids), 1500, 1700))
stopifnot("Negative control set not size-matched" =
  length(neg_ids) == length(promotes_ids))
```

# Main Figures: CDF of RNA Changes in Hypoxia 1hr

Cumulative distribution of RNA log2FC at 1hr hypoxia, once with the si3d contrast on
the x axis and once with the si3e contrast. All genes in grey, the eIF3d
hypoxia-promoted set (defined by TE at 1hr hypoxia) in orange, and the size-matched
negative control set in black. The gene sets are identical in both figures -- both are
the eIF3d TE-defined sets -- so the si3e version asks whether those same mRNAs behave
the same way at the RNA level when eIF3e is knocked down instead.

A rightward shift means the mRNA is lower in the knockdown than in sictrl, i.e. the
knockdown reduces its RNA level.

```{r focus_ecdf_builder}
focus_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",  # Okabe-Ito orange = eIF3d gene set
  "Negative control"       = "black",
  "All genes"              = "grey70"
)

# Stack one contrast into the three plotting groups. "All genes" is the full gene
# universe of that contrast (the two gene sets are subsets of it, so this is a
# stacked background rather than a disjoint group). `df` needs gene_id_clean + rna_lfc.
focus_groups <- function(df) {
  base <- df %>% filter(!is.na(rna_lfc))

  bind_rows(
    base %>% mutate(group = "All genes"),
    base %>% filter(gene_id_clean %in% promotes_ids) %>%
      mutate(group = "eIF3d hypoxia promotes"),
    base %>% filter(gene_id_clean %in% neg_ids) %>%
      mutate(group = "Negative control")
  ) %>%
    mutate(group = factor(group, levels = names(focus_colors)))
}

# Convenience wrapper for the per-timepoint contrasts already loaded in rna_all
focus_data <- function(kd, cond = "hypoxia", tp = "1hr") {
  rna_all %>%
    filter(knockdown == kd, condition == cond, timepoint == tp) %>%
    focus_groups()
}

# Wilcoxon p-values plus group sizes, formatted for on-plot annotation
focus_stats <- function(dat) {
  ns <- dat %>%
    group_by(group) %>%
    summarise(n = dplyr::n(), median_lfc = median(rna_lfc), .groups = "drop")

  lfc_of <- function(g) dat$rna_lfc[dat$group == g]
  p_vs_neg <- wilcox.test(lfc_of("eIF3d hypoxia promotes"),
                          lfc_of("Negative control"))$p.value
  p_vs_all <- wilcox.test(lfc_of("eIF3d hypoxia promotes"),
                          lfc_of("All genes"))$p.value

  n_of <- function(g) ns$n[ns$group == g]
  list(
    ns = ns,
    label = sprintf(
      "promoted vs neg ctrl: p = %.2g\npromoted vs all: p = %.2g\nn = %d / %d / %d",
      p_vs_neg, p_vs_all,
      n_of("eIF3d hypoxia promotes"), n_of("Negative control"), n_of("All genes")),
    # Legend labels carry n so group sizes are on the figure itself
    legend_labels = setNames(
      sprintf("%s (n = %s)", ns$group,
              format(ns$n, big.mark = ",", trim = TRUE)),
      as.character(ns$group))
  )
}

focus_ecdf_plot <- function(dat, st, x_lab, plot_title) {
  ggplot(dat, aes(x = rna_lfc, color = group)) +
    stat_ecdf(linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey50",
               linewidth = 0.4) +
    annotate("text", x = Inf, y = -Inf, label = st$label,
             hjust = 1.1, vjust = -0.3, size = 2.8) +
    scale_color_manual(values = focus_colors, labels = st$legend_labels,
                       breaks = names(focus_colors)) +
    coord_cartesian(xlim = c(-1.5, 2.5)) +
    labs(title = plot_title, x = x_lab,
         y = "Cumulative fraction of mRNAs", color = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "inside",
          legend.position.inside = c(0.02, 0.98),
          legend.justification = c(0, 1),
          legend.background = element_blank(),
          legend.key.size = unit(0.8, "lines"))
}
```

## si3d on the x axis

```{r ecdf_si3d_hypoxia_1hr, fig.width = 5.5, fig.height = 4}
ecdf_dat  <- focus_data("si3d")
ecdf_st   <- focus_stats(ecdf_dat)
ecdf_label <- ecdf_st$label          # reused by the density plot below
legend_labels <- ecdf_st$legend_labels

print(ecdf_st$ns)
cat(ecdf_label, "\n")

p_ecdf_focus <- focus_ecdf_plot(
  ecdf_dat, ecdf_st,
  x_lab = "RNA log2FC (sictrl / si3d)",
  plot_title = "RNA changes upon eIF3d knockdown, 1hr hypoxia")

print(p_ecdf_focus)
ggsave(here("plots", "66_ecdf_rna_si3d_hypoxia_1hr.pdf"), p_ecdf_focus,
       width = 5.5, height = 4)
```

## si3e on the x axis

Same eIF3d-defined gene sets, plotted against the si3e RNA contrast.

```{r ecdf_si3e_hypoxia_1hr, fig.width = 5.5, fig.height = 4}
ecdf_dat_3e <- focus_data("si3e")
ecdf_st_3e  <- focus_stats(ecdf_dat_3e)

print(ecdf_st_3e$ns)
cat(ecdf_st_3e$label, "\n")

p_ecdf_focus_3e <- focus_ecdf_plot(
  ecdf_dat_3e, ecdf_st_3e,
  x_lab = "RNA log2FC (sictrl / si3e)",
  plot_title = "RNA changes upon eIF3e knockdown, 1hr hypoxia")

print(p_ecdf_focus_3e)
ggsave(here("plots", "66_ecdf_rna_si3e_hypoxia_1hr.pdf"), p_ecdf_focus_3e,
       width = 5.5, height = 4)
```

# Time-Regressed Normoxia 1+4hr

Same figure for the normoxia 1+4hr contrast, where `hours` is included as a batch
covariate (`deFunction_time_normalized()`, design
`~ hours + Condition + SeqType + Condition:SeqType`) so the two timepoints are
pooled with the time effect regressed out.

**Directionality.** Both contrasts below are oriented log2(sictrl / knockdown), the
same as every other panel in this notebook. This is asserted rather than assumed:
the knockdown target mRNA must come out strongly **positive** (depleted in the
knockdown). Source notebook `eif3e_eif3d_normoxia_and_hypoxia.Rmd` runs these with
`control_condition = "si3d"` / `"si3e"` and `test_condition = "sictrl"`, i.e. the
knockdown is the DESeq2 reference.

```{r load_si3d_1and4hr}
nor14_3d <- read_csv(here("output",
                          "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                     show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# Directionality check: log2(sictrl / si3d) puts the knocked-down gene at positive LFC
eif3d_14 <- nor14_3d %>% filter(symbol == "EIF3D")
cat("si3d normoxia 1+4hr: EIF3D rna_lfc =", round(eif3d_14$rna_lfc, 2),
    " padj =", signif(eif3d_14$rna_padj, 2), "\n")
stopifnot("si3d 1+4hr RNA contrast is not oriented sictrl/si3d" =
  nrow(eif3d_14) == 1 && eif3d_14$rna_lfc > 2 && eif3d_14$rna_padj < 0.01)
```

The si3e 1+4hr time-regressed run in `eif3e_eif3d_normoxia_and_hypoxia.Rmd` saved
**TE only** (`te_si3e_vs_sictrl_normoxia_1and4hr.csv` has just `te_lfc`/`te_padj`),
so the matching RNA contrast does not exist anywhere in `output/`. It is computed
once here with the same function and orientation, then cached to
`output/rna_si3e_vs_sictrl_normoxia_1and4hr.csv` for reuse.

```{r compute_si3e_1and4hr_rna}
si3e_14_cache <- here("output", "rna_si3e_vs_sictrl_normoxia_1and4hr.csv")

if (file.exists(si3e_14_cache)) {
  cat("Using cached", basename(si3e_14_cache), "\n")
  nor14_3e <- read_csv(si3e_14_cache, show_col_types = FALSE)
} else {
  suppressPackageStartupMessages(library(DESeq2))
  source(here("code", "functions.R"))

  allCounts  <- read_csv(here("output", "human_allCounts_raw.csv"),
                         show_col_types = FALSE) %>% column_to_rownames("gene_id")
  normMatrix <- read_csv(here("output", "normalizedCounts.csv"),
                         show_col_types = FALSE) %>% column_to_rownames("gene_id")
  spikein    <- read_csv(here("output", "dros_spikeInCounts_raw.csv"),
                         show_col_types = FALSE) %>% column_to_rownames("gene_id")
  top500_ids <- rownames(spikein)
  metadata   <- read_csv(here("output", "metadata_clean.csv"), show_col_types = FALSE)
  gene_symbols <- read_csv(here("output", "deseq_results_1hr_spike.csv"),
                           show_col_types = FALSE) %>%
    dplyr::select(gene_id, symbol)

  md <- metadata %>%
    filter(condition == "normoxia",
           treatment %in% c("sictrl", "si3e"),
           hours %in% c(1, 4)) %>%
    mutate(comparison_variable = treatment,
           hours = as.factor(hours))

  cat("sictrl vs si3e in normoxia (1+4hr) - Samples:", nrow(md), "\n")
  stopifnot("No si3e normoxia 1+4hr samples found" = nrow(md) > 0)

  res_rna <- deFunction_time_normalized(
    countsMatrix   = allCounts[, md$sample_name],
    allCounts_Norm = normMatrix,
    metadata       = md,
    control_condition = "si3e",     # knockdown is the reference ...
    test_condition    = "sictrl",   # ... so the LFC is log2(sictrl / si3e)
    spikeInMatrix  = spikein[, md$sample_name],
    keepSpikes     = top500_ids,
    mode = "RNA"
  )

  nor14_3e <- res_rna %>%
    dplyr::select(gene_id, log2FoldChange, padj) %>%
    dplyr::rename(rna_lfc = log2FoldChange, rna_padj = padj) %>%
    left_join(gene_symbols, by = "gene_id")

  write_csv(nor14_3e, si3e_14_cache)
  cat("Saved:", basename(si3e_14_cache), "\n")
}

nor14_3e <- nor14_3e %>% mutate(gene_id_clean = sub("\\..*", "", gene_id))

# Same directionality check for si3e
eif3e_14 <- nor14_3e %>% filter(symbol == "EIF3E")
cat("si3e normoxia 1+4hr: EIF3E rna_lfc =", round(eif3e_14$rna_lfc, 2),
    " padj =", signif(eif3e_14$rna_padj, 2), "\n")
stopifnot("si3e 1+4hr RNA contrast is not oriented sictrl/si3e" =
  nrow(eif3e_14) == 1 && eif3e_14$rna_lfc > 2 && eif3e_14$rna_padj < 0.01)
```

```{r ecdf_si3d_normoxia_1and4hr, fig.width = 5.5, fig.height = 4}
ecdf_dat_3d_14 <- focus_groups(nor14_3d)
ecdf_st_3d_14  <- focus_stats(ecdf_dat_3d_14)

print(ecdf_st_3d_14$ns)
cat(ecdf_st_3d_14$label, "\n")

p_ecdf_3d_14 <- focus_ecdf_plot(
  ecdf_dat_3d_14, ecdf_st_3d_14,
  x_lab = "RNA log2FC (sictrl / si3d)",
  plot_title = "RNA changes upon eIF3d knockdown, normoxia 1+4hr")

print(p_ecdf_3d_14)
ggsave(here("plots", "66_ecdf_rna_si3d_normoxia_1and4hr.pdf"), p_ecdf_3d_14,
       width = 5.5, height = 4)
```

```{r ecdf_si3e_normoxia_1and4hr, fig.width = 5.5, fig.height = 4}
ecdf_dat_3e_14 <- focus_groups(nor14_3e)
ecdf_st_3e_14  <- focus_stats(ecdf_dat_3e_14)

print(ecdf_st_3e_14$ns)
cat(ecdf_st_3e_14$label, "\n")

p_ecdf_3e_14 <- focus_ecdf_plot(
  ecdf_dat_3e_14, ecdf_st_3e_14,
  x_lab = "RNA log2FC (sictrl / si3e)",
  plot_title = "RNA changes upon eIF3e knockdown, normoxia 1+4hr")

print(p_ecdf_3e_14)
ggsave(here("plots", "66_ecdf_rna_si3e_normoxia_1and4hr.pdf"), p_ecdf_3e_14,
       width = 5.5, height = 4)
```

# Density View of the Hypoxia 1hr Comparison

Same comparison as densities, for a view of where the mass sits rather than the
cumulative shift.

```{r density_si3d_hypoxia_1hr, fig.width = 5.5, fig.height = 4}
p_dens_focus <- ggplot(ecdf_dat, aes(x = rna_lfc, fill = group, color = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50",
             linewidth = 0.4) +
  annotate("text", x = Inf, y = Inf, label = ecdf_label,
           hjust = 1.1, vjust = 1.4, size = 2.8) +
  scale_color_manual(values = focus_colors, labels = legend_labels,
                     breaks = names(focus_colors)) +
  scale_fill_manual(values = focus_colors, labels = legend_labels,
                    breaks = names(focus_colors)) +
  coord_cartesian(xlim = c(-1.5, 2.5)) +
  labs(title = "RNA changes upon eIF3d knockdown, 1hr hypoxia",
       x = "RNA log2FC (sictrl / si3d)", y = "Density",
       fill = NULL, color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_dens_focus)
ggsave(here("plots", "66_density_rna_si3d_hypoxia_1hr.pdf"), p_dens_focus,
       width = 5.5, height = 4)
```

# Distribution of RNA Changes

Dashed vertical lines mark the observed median per panel (see the global shift
caveat in the overview).

```{r plot_density, fig.width = 9, fig.height = 5}
rna_plot <- rna_all %>% filter(!is.na(rna_lfc))

medians <- rna_plot %>%
  group_by(knockdown, condition, timepoint) %>%
  summarise(median_lfc = median(rna_lfc), n = dplyr::n(), .groups = "drop")

p_dens <- ggplot(rna_plot, aes(x = rna_lfc, fill = knockdown, color = knockdown)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(data = medians, aes(xintercept = median_lfc, color = knockdown),
             linetype = "dashed", linewidth = 0.5) +
  facet_grid(condition ~ timepoint) +
  coord_cartesian(xlim = c(-2, 2)) +
  scale_fill_manual(values = kd_colors) +
  scale_color_manual(values = kd_colors) +
  labs(title = "RNA log2FC distributions upon eIF3d and eIF3e knockdown",
       subtitle = sprintf("n = %d to %d genes per contrast",
                          min(medians$n), max(medians$n)),
       x = "RNA log2FC (sictrl / knockdown)", y = "Density",
       fill = NULL, color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_dens)
ggsave(here("plots", "66_rna_lfc_density.pdf"), p_dens, width = 9, height = 5)
```

```{r plot_ecdf, fig.width = 9, fig.height = 5}
p_ecdf <- ggplot(rna_plot, aes(x = rna_lfc, color = knockdown)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
  facet_grid(condition ~ timepoint) +
  coord_cartesian(xlim = c(-2, 2)) +
  scale_color_manual(values = kd_colors) +
  labs(title = "Cumulative distribution of RNA log2FC",
       subtitle = sprintf("n = %d to %d genes per contrast",
                          min(medians$n), max(medians$n)),
       x = "RNA log2FC (sictrl / knockdown)",
       y = "Cumulative fraction", color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_ecdf)
ggsave(here("plots", "66_rna_lfc_ecdf.pdf"), p_ecdf, width = 9, height = 5)
```

# Volcano Plots

```{r plot_volcano, fig.width = 10, fig.height = 6}
# padj can be exactly 0 for the most significant genes; floor it at the smallest
# representable nonzero value so -log10 stays finite
min_nonzero_padj <- min(rna_all$rna_padj[rna_all$rna_padj > 0], na.rm = TRUE)

volc <- rna_all %>%
  filter(!is.na(rna_lfc), !is.na(rna_padj)) %>%
  mutate(
    padj_floor = pmax(rna_padj, min_nonzero_padj),
    neg_log10p = -log10(padj_floor),
    sig = case_when(
      rna_padj < PADJ_CUTOFF & rna_lfc >  LFC_CUTOFF ~ "mRNA down in knockdown",
      rna_padj < PADJ_CUTOFF & rna_lfc < -LFC_CUTOFF ~ "mRNA up in knockdown",
      TRUE                                           ~ "Not significant"
    )
  )

# Label the 8 most significant genes per panel
set.seed(9)
volc_labels <- volc %>%
  filter(sig != "Not significant") %>%
  group_by(knockdown, contrast) %>%
  slice_min(order_by = rna_padj, n = 8, with_ties = FALSE) %>%
  ungroup()

p_volc <- ggplot(volc, aes(x = rna_lfc, y = neg_log10p, color = sig)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_vline(xintercept = c(-LFC_CUTOFF, LFC_CUTOFF),
             linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_hline(yintercept = -log10(PADJ_CUTOFF),
             linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_text_repel(data = volc_labels, aes(label = symbol),
                  size = 2.2, max.overlaps = 20, show.legend = FALSE,
                  segment.size = 0.2, min.segment.length = 0) +
  facet_grid(knockdown ~ contrast) +
  scale_color_manual(values = c(dir_colors, "Not significant" = "grey80")) +
  labs(title = "RNA changes upon eIF3d and eIF3e knockdown",
       x = "RNA log2FC (sictrl / knockdown)",
       y = "-log10(adjusted p-value)", color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_volc)
ggsave(here("plots", "66_rna_volcano_all_contrasts.pdf"), p_volc, width = 12, height = 6)
```

# Concordance Between si3d and si3e

Are the same mRNAs destabilized by both knockdowns? Pearson correlation of
`rna_lfc` between the two knockdowns within each condition and timepoint.

```{r plot_concordance, fig.width = 9, fig.height = 6}
paired <- rna_all %>%
  dplyr::select(knockdown, condition, timepoint, contrast,
                gene_id, symbol, rna_lfc, rna_padj) %>%
  pivot_wider(id_cols = c(condition, timepoint, contrast, gene_id, symbol),
              names_from = knockdown,
              values_from = c(rna_lfc, rna_padj)) %>%
  filter(!is.na(rna_lfc_si3d), !is.na(rna_lfc_si3e))

cor_stats <- paired %>%
  group_by(condition, timepoint) %>%
  summarise(
    r = cor(rna_lfc_si3d, rna_lfc_si3e, method = "pearson"),
    rho = cor(rna_lfc_si3d, rna_lfc_si3e, method = "spearman"),
    n = dplyr::n(),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("Pearson R = %.2f\nSpearman rho = %.2f\nn = %s",
                         r, rho, format(n, big.mark = ",")))

print(cor_stats)

p_conc <- ggplot(paired, aes(x = rna_lfc_si3d, y = rna_lfc_si3e)) +
  geom_point(size = 0.4, alpha = 0.2, color = "grey30") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "#D55E00", linewidth = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "#0072B2", linewidth = 0.6) +
  geom_text(data = cor_stats, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.1, vjust = -0.3, size = 2.8, inherit.aes = FALSE) +
  facet_grid(condition ~ timepoint) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3)) +
  labs(title = "RNA changes are shared between eIF3d and eIF3e knockdown",
       subtitle = "Dashed line = y = x; blue line = linear fit",
       x = "si3d RNA log2FC (sictrl / si3d)",
       y = "si3e RNA log2FC (sictrl / si3e)") +
  theme_classic(base_size = 12)

print(p_conc)
ggsave(here("plots", "66_rna_lfc_si3d_vs_si3e_scatter.pdf"), p_conc,
       width = 9, height = 6)
```

## Overlap of significantly changed genes

```{r overlap_sig}
overlap_tbl <- paired %>%
  mutate(
    sig_3d = rna_padj_si3d < PADJ_CUTOFF & abs(rna_lfc_si3d) > LFC_CUTOFF,
    sig_3e = rna_padj_si3e < PADJ_CUTOFF & abs(rna_lfc_si3e) > LFC_CUTOFF
  ) %>%
  group_by(condition, timepoint) %>%
  summarise(
    n_3d_only = sum(sig_3d & !sig_3e, na.rm = TRUE),
    n_3e_only = sum(!sig_3d & sig_3e, na.rm = TRUE),
    n_both    = sum(sig_3d & sig_3e, na.rm = TRUE),
    n_same_direction = sum(sig_3d & sig_3e &
                             sign(rna_lfc_si3d) == sign(rna_lfc_si3e), na.rm = TRUE),
    .groups = "drop"
  )

print(overlap_tbl)
write_csv(overlap_tbl, here("output", "66_rna_si3d_si3e_overlap.csv"))
```

# RNA Changes in the eIF3d Hypoxia TE-Promoted Gene Set

Do the mRNAs whose *translation* eIF3d promotes in hypoxia also change at the
**RNA** level upon knockdown? Compared against a size-matched negative control set
and all remaining genes.

Gene sets are the ones loaded above (`promotes_ids`, `neg_ids`).

**Panel columns.** The three independent normoxia timepoints are replaced here by the
single time-regressed normoxia 1+4hr contrast (`hours` as batch covariate), so the
columns are: normoxia 1+4hr, hypoxia 1hr, hypoxia 4hr, hypoxia 24hr. All four are
oriented log2(sictrl / knockdown). The all-gene survey grids earlier in the notebook
still show every timepoint separately.

```{r build_panel_data}
PANEL_LEVELS <- c("normoxia 1+4hr", "hypoxia 1hr", "hypoxia 4hr", "hypoxia 24hr")

panel_cols <- c("knockdown", "contrast", "gene_id", "gene_id_clean", "symbol",
                "rna_lfc", "rna_padj")

rna_panels <- bind_rows(
  rna_all %>%
    filter(condition == "hypoxia") %>%
    mutate(contrast = as.character(contrast)) %>%
    dplyr::select(all_of(panel_cols)),
  nor14_3d %>%
    mutate(knockdown = "si3d", contrast = "normoxia 1+4hr") %>%
    dplyr::select(all_of(panel_cols)),
  nor14_3e %>%
    mutate(knockdown = "si3e", contrast = "normoxia 1+4hr") %>%
    dplyr::select(all_of(panel_cols))
) %>%
  filter(!is.na(rna_lfc)) %>%
  mutate(knockdown = factor(knockdown, levels = KNOCKDOWNS),
         contrast  = factor(contrast, levels = PANEL_LEVELS))

stopifnot("Unexpected panel contrast label" = !any(is.na(rna_panels$contrast)))
print(rna_panels %>% dplyr::count(knockdown, contrast))
```

```{r plot_geneset_density, fig.width = 9, fig.height = 6}
grouped <- rna_panels %>%
  mutate(group = case_when(
    gene_id_clean %in% promotes_ids ~ "eIF3d hypoxia promotes",
    gene_id_clean %in% neg_ids      ~ "Negative control",
    TRUE                            ~ "All other genes"
  )) %>%
  mutate(group = factor(group, levels = names(group_colors)))

group_n <- grouped %>% dplyr::count(knockdown, contrast, group)
print(group_n)

# Wilcoxon: promoted vs size-matched negative control, per panel
wilcox_tbl <- grouped %>%
  filter(group != "All other genes") %>%
  group_by(knockdown, contrast) %>%
  summarise(
    p = wilcox.test(rna_lfc[group == "eIF3d hypoxia promotes"],
                    rna_lfc[group == "Negative control"])$p.value,
    n_pos = sum(group == "eIF3d hypoxia promotes"),
    n_neg = sum(group == "Negative control"),
    med_pos = median(rna_lfc[group == "eIF3d hypoxia promotes"]),
    med_neg = median(rna_lfc[group == "Negative control"]),
    .groups = "drop"
  ) %>%
  mutate(label = sprintf("Wilcoxon p = %.2g\nn = %d vs %d", p, n_pos, n_neg))

print(wilcox_tbl)

p_gs <- ggplot(grouped, aes(x = rna_lfc, fill = group, color = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_text(data = wilcox_tbl, aes(x = Inf, y = Inf, label = label),
            hjust = 1.1, vjust = 1.4, size = 2.6, inherit.aes = FALSE) +
  facet_grid(knockdown ~ contrast) +
  coord_cartesian(xlim = c(-2, 2.5)) +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "RNA changes in eIF3d hypoxia TE-promoted mRNAs",
       subtitle = "Gene set defined by TE (hypoxia 1hr, te_lfc > 0.5); p-values compare promoted vs negative control",
       x = "RNA log2FC (sictrl / knockdown)", y = "Density",
       fill = NULL, color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_gs)
ggsave(here("plots", "66_rna_lfc_te_promoted_geneset.pdf"), p_gs,
       width = 11, height = 6)
```

## Same panels as a CDF

```{r plot_geneset_ecdf, fig.width = 9, fig.height = 6}
p_gs_ecdf <- ggplot(grouped, aes(x = rna_lfc, color = group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50",
             linewidth = 0.4) +
  geom_text(data = wilcox_tbl, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.1, vjust = -0.3, size = 2.6, inherit.aes = FALSE) +
  facet_grid(knockdown ~ contrast) +
  coord_cartesian(xlim = c(-2, 2.5)) +
  scale_color_manual(values = group_colors) +
  labs(title = "RNA changes in eIF3d hypoxia TE-promoted mRNAs",
       subtitle = "Gene set defined by TE (hypoxia 1hr, te_lfc > 0.5); p-values compare promoted vs negative control",
       x = "RNA log2FC (sictrl / knockdown)",
       y = "Cumulative fraction of mRNAs", color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_gs_ecdf)
ggsave(here("plots", "66_ecdf_rna_te_promoted_geneset.pdf"), p_gs_ecdf,
       width = 11, height = 6)
```

```{r plot_geneset_box, fig.width = 8, fig.height = 5}
p_box <- ggplot(grouped, aes(x = contrast, y = rna_lfc, fill = group)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_boxplot(outlier.size = 0.2, outlier.alpha = 0.2, linewidth = 0.3) +
  facet_wrap(~ knockdown, ncol = 1) +
  coord_cartesian(ylim = c(-2, 3)) +
  scale_fill_manual(values = group_colors) +
  labs(title = "RNA changes by gene group",
       subtitle = sprintf("n = %d promoted, %d negative control, %d other genes per panel",
                          length(promotes_ids), length(neg_ids),
                          length(universe) - length(promotes_ids) - length(neg_ids)),
       x = "Contrast", y = "RNA log2FC (sictrl / knockdown)", fill = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_box)
ggsave(here("plots", "66_rna_lfc_geneset_boxplot.pdf"), p_box, width = 8, height = 6)
```

# RNA vs Ribosome Occupancy

Context for whether the RNA changes are forwarded to the ribosome. Points are
colored by the translation category assigned in the source notebook.

```{r plot_rna_vs_ribo, fig.width = 10, fig.height = 6}
rna_ribo <- rna_all %>% filter(!is.na(rna_lfc), !is.na(ribo_lfc))

rr_stats <- rna_ribo %>%
  group_by(knockdown, contrast) %>%
  summarise(r = cor(rna_lfc, ribo_lfc, method = "pearson"),
            n = dplyr::n(), .groups = "drop") %>%
  mutate(label = sprintf("R = %.2f, n = %s", r, format(n, big.mark = ",")))

print(rr_stats)

p_rr <- ggplot(rna_ribo, aes(x = rna_lfc, y = ribo_lfc)) +
  geom_point(size = 0.3, alpha = 0.15, color = "grey30") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "#D55E00", linewidth = 0.4) +
  geom_text(data = rr_stats, aes(x = Inf, y = -Inf, label = label),
            hjust = 1.1, vjust = -0.6, size = 2.8, inherit.aes = FALSE) +
  facet_grid(knockdown ~ contrast) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3)) +
  labs(title = "RNA vs ribosome footprint changes",
       subtitle = "Dashed line = y = x (change fully forwarded to the ribosome)",
       x = "RNA log2FC (sictrl / knockdown)",
       y = "Ribo log2FC (sictrl / knockdown)") +
  theme_classic(base_size = 12)

print(p_rr)
ggsave(here("plots", "66_rna_vs_ribo_scatter.pdf"), p_rr, width = 12, height = 6)
```

# Session Info

```{r session_info}
sessionInfo()
```
