---
title: "eIF3d-promoted genes across all conditions and timepoints"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output: html_document
params:
  lfc_cutoff: 0.5
  padj_cutoff: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)

# knitr builds `params` from the YAML header, but it does not exist when running
# chunks interactively - and bare `params` then resolves to an unrelated function
# from an attached package ("object of type 'closure' is not subsettable").
# Test is.list() rather than exists(), since exists() finds that function.
if (!exists("params") || !is.list(params)) {
  params <- list(lfc_cutoff = 0.5, padj_cutoff = 0.05)
}
```

## Purpose

Build a single collaborator-facing table covering **every tested gene**, with a
per-condition flag marking where its translation is **promoted by eIF3d**. Genes
that are not up in any condition are retained rather than filtered out, so the
file serves as its own background set and the collaborator can compare hits
against non-hits without needing a second export.

Conditions scanned (MDA-MB-231, si3d vs siCTRL, spike-in normalized):

| Condition | Timepoints |
|---|---|
| normoxia | 1hr, 4hr, 24hr |
| hypoxia  | 1hr, 4hr, 24hr |

**Directionality (MDA-MB-231):** the DESeq2 coefficient in these files is
log2(sictrl / si3d), so **positive `te_lfc` = knockdown reduces TE = eIF3d
promotes translation**. This is the convention already used by
`code/generate_signatures.Rmd`; no sign flip is applied here.

**Call:** `te_padj < `r params$padj_cutoff`` AND `te_lfc > `r params$lfc_cutoff``
(the project's standard 0.5 fold-change threshold).

## Load per-condition TE results

```{r load}
conditions <- tribble(
  ~condition,  ~timepoint, ~file,
  "normoxia",  "1hr",      "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv",
  "normoxia",  "4hr",      "translation_categories_si3d_vs_sictrl_normoxia_4hr.csv",
  "normoxia",  "24hr",     "translation_categories_si3d_vs_sictrl_normoxia_24hr.csv",
  "hypoxia",   "1hr",      "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
  "hypoxia",   "4hr",      "translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv",
  "hypoxia",   "24hr",     "translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv"
)

# fail early rather than silently dropping a condition, and name what is missing
missing_files <- conditions$file[!file.exists(here("output", conditions$file))]
if (length(missing_files) > 0) {
  stop("missing TE result file(s): ", paste(missing_files, collapse = ", "))
}

te_long <- conditions %>%
  mutate(data = map(file, ~ read_csv(here("output", .x), show_col_types = FALSE))) %>%
  dplyr::select(-file) %>%
  unnest(data) %>%
  mutate(
    gene_id_clean = sub("\\..*", "", gene_id),
    contrast      = paste0(condition, "_", timepoint),
    contrast      = factor(contrast, levels = paste0(
      rep(c("normoxia", "hypoxia"), each = 3), "_", c("1hr", "4hr", "24hr")))
  )

# computational checks
stopifnot(all(c("gene_id", "symbol", "te_lfc", "te_padj") %in% colnames(te_long)))
stopifnot("te_lfc must be finite where present" =
  all(is.finite(te_long$te_lfc[!is.na(te_long$te_lfc)])))
stopifnot("te_padj must lie in [0, 1]" =
  all(te_long$te_padj >= 0 & te_long$te_padj <= 1, na.rm = TRUE))
stopifnot("all six contrasts must be present" =
  n_distinct(te_long$contrast) == 6)
# one row per gene per contrast: duplicates would inflate n_conditions_promoted
stopifnot("gene x contrast must be unique" =
  nrow(te_long) == n_distinct(te_long$gene_id_clean, te_long$contrast))
# version stripping must not collapse two distinct gene IDs into one
stopifnot("version stripping collapsed distinct gene IDs" =
  n_distinct(te_long$gene_id) == n_distinct(te_long$gene_id_clean))

cat("Rows loaded:", nrow(te_long), "across", n_distinct(te_long$gene_id_clean), "genes\n")
te_long %>% dplyr::count(condition, timepoint)
```

## Call promoted genes per contrast

```{r call}
promoted_long <- te_long %>%
  filter(!is.na(te_padj), te_padj < params$padj_cutoff, te_lfc > params$lfc_cutoff)

promoted_long %>%
  dplyr::count(condition, timepoint, name = "n_promoted") %>%
  arrange(condition, timepoint)

# Anchor check: the si3d hypoxia 1hr promoted set is the project's primary
# positive gene set and is known to contain 1604 genes at lfc > 0.5, padj < 0.05.
# Hard-coded on purpose - if an upstream TE file is ever regenerated, this fires
# rather than silently shipping a different list to a collaborator.
n_hyp_1hr <- sum(promoted_long$contrast == "hypoxia_1hr")
cat("hypoxia 1hr genes up with (promoted by) eIF3d:", n_hyp_1hr, "\n")
if (params$lfc_cutoff == 0.5 && params$padj_cutoff == 0.05) {
  stopifnot("hypoxia 1hr promoted set must contain 1604 genes" = n_hyp_1hr == 1604)
}
```

### Validation: reproduce the existing per-contrast gene sets

`output/genesets/{condition}_3d_promotes_TE_{timepoint}_lfc0.5.csv` were written
by `code/generate_signatures.Rmd` with the same thresholds. At the default params
the sets here must match those files **gene for gene**; this catches a wrong input
file, a flipped sign, or a mis-set threshold.

Caveat: `generate_signatures.Rmd` caps each reference set with `slice_head(n = 2500)`.
All six sets are currently well under that cap (largest is 1604), so the comparison
is exact. If a regenerated TE file ever pushed a contrast past 2500, the reference
would be truncated and this check would fail for that reason rather than a real
discrepancy - the cap-headroom check below makes that case self-explanatory.

```{r validate_against_genesets}
if (params$lfc_cutoff == 0.5 && params$padj_cutoff == 0.05) {
  ref_sets <- conditions %>%
    dplyr::select(condition, timepoint) %>%
    mutate(
      ref_genes = map2(condition, timepoint, ~ read_csv(
        here("output", "genesets",
             paste0(.x, "_3d_promotes_TE_", .y, "_lfc0.5.csv")),
        show_col_types = FALSE)$ensembl_gene)
    )

  here_sets <- promoted_long %>%
    group_by(condition, timepoint) %>%
    summarise(here_genes = list(unique(gene_id_clean)), .groups = "drop")

  comparison <- ref_sets %>%
    full_join(here_sets, by = c("condition", "timepoint")) %>%
    mutate(
      n_ref      = map_int(ref_genes, length),
      n_here     = map_int(here_genes, length),
      identical_set = map2_lgl(ref_genes, here_genes, setequal)
    ) %>%
    dplyr::select(condition, timepoint, n_ref, n_here, identical_set)

  print(comparison)
  # the reference sets are slice_head(2500)-capped upstream; flag if any contrast
  # approaches that cap, which would make the comparison below non-exact
  stopifnot("a contrast now exceeds the reference set's 2500 cap" =
    all(comparison$n_here <= 2500))
  # gene-identity match, not just counts: a flipped sign would produce a
  # same-size but disjoint set and still pass a count-only check
  stopifnot("promoted gene sets disagree with existing geneset CSVs" =
    all(comparison$identical_set))
  cat("All six contrasts reproduce the published gene sets gene-for-gene.\n")
} else {
  cat("Non-default thresholds (lfc =", params$lfc_cutoff,
      ", padj =", params$padj_cutoff, ") - skipping reference comparison.\n")
}
```

### Biological spot-check: directionality

The 1hr hypoxia si3d set is the project's primary positive gene set and by far the
largest. In **every** contrast its median TE LFC must sit **above** the
transcriptome-wide median for that contrast: eIF3d-promoted genes are shifted
positive relative to background. This is a directionality assertion - if the sign
convention were inverted anywhere, the promoted set would fall *below* background.

Note the shift is tested relative to background, not against zero. The hypoxia
effect decays with time (by 24hr the promoted set's median TE LFC is ~0), so an
absolute `> 0` test would fail for a real biological reason rather than a coding
error.

```{r spot_check}
hyp_1hr_genes <- promoted_long %>%
  filter(contrast == "hypoxia_1hr") %>%
  pull(gene_id_clean)

bg_medians <- te_long %>%
  group_by(contrast) %>%
  summarise(background_median = median(te_lfc, na.rm = TRUE), .groups = "drop")

shift_check <- te_long %>%
  filter(gene_id_clean %in% hyp_1hr_genes) %>%
  group_by(contrast) %>%
  summarise(
    set_median   = median(te_lfc, na.rm = TRUE),
    frac_te_lfc_positive = mean(te_lfc > 0, na.rm = TRUE),
    n = n(), .groups = "drop"
  ) %>%
  left_join(bg_medians, by = "contrast") %>%
  mutate(shifted_positive = set_median > background_median)

print(shift_check)
stopifnot("hypoxia 1hr promoted set must sit above background in every contrast" =
  all(shift_check$shifted_positive))

# and in its own calling contrast the shift must be strongly significant
own_contrast <- te_long %>%
  filter(contrast == "hypoxia_1hr") %>%
  mutate(in_set = gene_id_clean %in% hyp_1hr_genes)
wt <- wilcox.test(te_lfc ~ in_set, data = own_contrast, alternative = "less")
cat("hypoxia_1hr promoted vs rest, Wilcoxon p =", format.pval(wt$p.value),
    ", n set =", sum(own_contrast$in_set),
    ", n rest =", sum(!own_contrast$in_set), "\n")
stopifnot("promoted set not significantly shifted positive in its own contrast" =
  wt$p.value < 1e-10)
```

## Build the collaborator table

**One row per tested gene - every gene, not just the promoted ones.** Each of the
six contrasts contributes three columns: `promoted_*` (the up-in-this-condition
flag), plus `te_lfc_*` and `te_padj_*` giving the underlying numbers. Genes never
promoted in any condition are retained with `promoted_any = FALSE`, so the file is
a complete background set as well as a hit list.

```{r build_table}
gene_symbols <- te_long %>%
  filter(!is.na(symbol), symbol != "") %>%
  distinct(gene_id_clean, symbol) %>%
  group_by(gene_id_clean) %>%
  slice_head(n = 1) %>%
  ungroup()

# every gene in the tested universe, promoted or not
all_genes <- te_long %>%
  distinct(gene_id_clean) %>%
  left_join(gene_symbols, by = "gene_id_clean")

promoted_genes <- sort(unique(promoted_long$gene_id_clean))

# per-contrast stats for ALL genes, significant or not
stats_wide <- te_long %>%
  dplyr::select(gene_id_clean, contrast, te_lfc, te_padj) %>%
  pivot_wider(
    names_from  = contrast,
    values_from = c(te_lfc, te_padj),
    names_glue  = "{.value}_{contrast}",
    names_sort  = FALSE
  )

# the per-condition "is it up here" flags - recomputed from the same predicate
# as promoted_long so the two can be cross-checked against each other below
promoted_flags_wide <- te_long %>%
  mutate(promoted = !is.na(te_padj) &
           te_padj < params$padj_cutoff &
           te_lfc  > params$lfc_cutoff) %>%
  dplyr::select(gene_id_clean, contrast, promoted) %>%
  pivot_wider(
    names_from  = contrast,
    values_from = promoted,
    names_glue  = "promoted_{contrast}",
    names_sort  = FALSE
  )

summary_tbl <- promoted_long %>%
  group_by(gene_id_clean) %>%
  summarise(
    # n_distinct(), not n(): n() would silently inflate on duplicate gene x
    # contrast rows (guarded upstream, but keep the summary self-correct)
    n_conditions_promoted = n_distinct(contrast),
    # sort the FACTOR then coerce, so timepoints come out chronologically.
    # sort(as.character(contrast)) sorts alphabetically and puts 24hr before 4hr.
    conditions_promoted   = paste(as.character(sort(contrast)), collapse = "; "),
    promoted_normoxia     = any(condition == "normoxia"),
    promoted_hypoxia      = any(condition == "hypoxia"),
    # NB: max/min are over the contrasts where the gene was PROMOTED, not over
    # all six - hence the explicit `_promoted` suffix. A gene can have a larger
    # te_lfc in a contrast that missed the padj cutoff; that value still appears
    # in its te_lfc_* column but is deliberately not summarised here.
    max_te_lfc_promoted   = max(te_lfc),
    max_te_lfc_condition  = as.character(contrast[which.max(te_lfc)]),
    min_te_padj_promoted  = min(te_padj),
    .groups = "drop"
  ) %>%
  mutate(
    shared_normoxia_hypoxia = promoted_normoxia & promoted_hypoxia,
    condition_class = case_when(
      shared_normoxia_hypoxia ~ "both",
      promoted_hypoxia        ~ "hypoxia_only",
      promoted_normoxia       ~ "normoxia_only"
    )
  )

contrast_levels <- levels(te_long$contrast)

eif3d_te_all <- all_genes %>%
  left_join(summary_tbl,          by = "gene_id_clean") %>%
  left_join(promoted_flags_wide,  by = "gene_id_clean") %>%
  left_join(stats_wide,           by = "gene_id_clean") %>%
  # genes never promoted get 0 / FALSE / "none" rather than NA, so the summary
  # columns stay usable for filtering and never silently drop background genes
  mutate(
    n_conditions_promoted = coalesce(n_conditions_promoted, 0L),
    conditions_promoted   = coalesce(conditions_promoted, "none"),
    promoted_normoxia     = coalesce(promoted_normoxia, FALSE),
    promoted_hypoxia      = coalesce(promoted_hypoxia, FALSE),
    condition_class       = coalesce(condition_class, "not_promoted"),
    promoted_any          = n_conditions_promoted > 0
  ) %>%
  dplyr::select(
    symbol, gene_id_clean,
    promoted_any, condition_class, conditions_promoted, n_conditions_promoted,
    promoted_normoxia, promoted_hypoxia,
    max_te_lfc_promoted, max_te_lfc_condition, min_te_padj_promoted,
    all_of(paste0("promoted_", contrast_levels)),
    all_of(paste0("te_lfc_",   contrast_levels)),
    all_of(paste0("te_padj_",  contrast_levels))
  ) %>%
  arrange(desc(promoted_any), desc(n_conditions_promoted),
          desc(max_te_lfc_promoted))

# the promoted-only subset stays available as a convenience file
eif3d_promoted <- eif3d_te_all %>% filter(promoted_any)

# --- checks on the full table -------------------------------------------------
stopifnot("one row per tested gene" =
  nrow(eif3d_te_all) == n_distinct(te_long$gene_id_clean))
stopifnot("gene_id_clean must be unique" =
  !any(duplicated(eif3d_te_all$gene_id_clean)))
stopifnot("n_conditions_promoted must be 0-6" =
  all(eif3d_te_all$n_conditions_promoted >= 0 &
      eif3d_te_all$n_conditions_promoted <= 6))
stopifnot("missing condition_class" = all(!is.na(eif3d_te_all$condition_class)))

# the per-contrast flags must reproduce n_conditions_promoted exactly. This is
# the key cross-check: the flags come from pivoting the raw predicate, the count
# comes from promoted_long, so agreement means the two paths concur.
flag_mat <- as.matrix(eif3d_te_all[, paste0("promoted_", contrast_levels)])
stopifnot("no NA in the promoted flags" = !any(is.na(flag_mat)))
stopifnot("promoted flags disagree with n_conditions_promoted" =
  all(rowSums(flag_mat) == eif3d_te_all$n_conditions_promoted))
stopifnot("promoted_any disagrees with the flags" =
  all(eif3d_te_all$promoted_any == (rowSums(flag_mat) > 0)))

# per-contrast column totals must equal the per-contrast counts from promoted_long
flag_totals <- colSums(flag_mat)
long_totals <- table(factor(promoted_long$contrast, levels = contrast_levels))
stopifnot("per-contrast flag totals disagree with promoted_long" =
  all(flag_totals == as.integer(long_totals)))

# summary stats present exactly for promoted genes, absent otherwise
stopifnot("max_te_lfc_promoted must be NA iff the gene was never promoted" =
  all(is.na(eif3d_te_all$max_te_lfc_promoted) == !eif3d_te_all$promoted_any))
stopifnot("max_te_lfc_promoted must clear the cutoff where present" =
  all(eif3d_te_all$max_te_lfc_promoted[eif3d_te_all$promoted_any] >
        params$lfc_cutoff))
stopifnot("min_te_padj_promoted must clear the cutoff where present" =
  all(eif3d_te_all$min_te_padj_promoted[eif3d_te_all$promoted_any] <
        params$padj_cutoff))
stopifnot("not_promoted class must match promoted_any" =
  all((eif3d_te_all$condition_class == "not_promoted") ==
        !eif3d_te_all$promoted_any))

# --- checks on the promoted subset (unchanged from the previous version) ------
stopifnot("promoted subset must equal the union of per-contrast sets" =
  setequal(eif3d_promoted$gene_id_clean, promoted_genes))
stopifnot("conditions_promoted disagrees with n_conditions_promoted" =
  all(lengths(strsplit(eif3d_promoted$conditions_promoted, "; ")) ==
        eif3d_promoted$n_conditions_promoted))
stopifnot("promoted_hypoxia flag disagrees with conditions_promoted" =
  all(eif3d_promoted$promoted_hypoxia ==
        grepl("hypoxia", eif3d_promoted$conditions_promoted)))
stopifnot("promoted_normoxia flag disagrees with conditions_promoted" =
  all(eif3d_promoted$promoted_normoxia ==
        grepl("normoxia", eif3d_promoted$conditions_promoted)))

cat("All tested genes:", nrow(eif3d_te_all), "\n")
cat("  promoted in >= 1 condition:", sum(eif3d_te_all$promoted_any), "\n")
cat("  never promoted (background):", sum(!eif3d_te_all$promoted_any), "\n")
cat("  genes with no HGNC symbol:", sum(is.na(eif3d_te_all$symbol)), "\n")
head(eif3d_te_all, 10)
```

## Summary breakdown

```{r summary}
# includes the not_promoted background class
eif3d_te_all %>% dplyr::count(condition_class, name = "n_genes")

eif3d_te_all %>%
  dplyr::count(n_conditions_promoted, name = "n_genes") %>%
  arrange(n_conditions_promoted)

# per-condition tally straight off the exported flag columns
tibble(
  contrast   = contrast_levels,
  n_promoted = as.integer(colSums(flag_mat)),
  n_not      = nrow(eif3d_te_all) - as.integer(colSums(flag_mat))
)

# most frequently recurring genes
eif3d_te_all %>%
  filter(n_conditions_promoted >= 3) %>%
  dplyr::select(symbol, conditions_promoted, n_conditions_promoted,
                max_te_lfc_promoted) %>%
  head(20)

# gene symbols are not unique keys - warn the collaborator explicitly
dup_symbols <- eif3d_te_all %>%
  filter(!is.na(symbol)) %>%
  dplyr::count(symbol, name = "n_gene_ids") %>%
  filter(n_gene_ids > 1)
cat("\nSymbols mapping to >1 gene_id (key on gene_id_clean, not symbol):",
    nrow(dup_symbols), "\n")
```

## Export

```{r export}
# PRIMARY FILE: every tested gene, with an up-in-this-condition flag per contrast
out_all <- here("output", "genesets", "eif3d_te_all_genes_all_conditions.csv")
write_csv(eif3d_te_all, out_all)
cat("Wrote", nrow(eif3d_te_all), "genes (",
    sum(eif3d_te_all$promoted_any), "promoted,",
    sum(!eif3d_te_all$promoted_any), "not ) to", out_all, "\n")

# convenience subset: promoted genes only, same columns
out_wide <- here("output", "genesets", "eif3d_promoted_genes_all_conditions.csv")
write_csv(eif3d_promoted, out_wide)
cat("Wrote", nrow(eif3d_promoted), "promoted genes to", out_wide, "\n")

# long / tidy version of the FULL matrix: one row per gene x contrast, with the
# promoted flag alongside the RNA and ribosome components of each TE call
eif3d_te_long_export <- te_long %>%
  mutate(promoted = !is.na(te_padj) &
           te_padj < params$padj_cutoff &
           te_lfc  > params$lfc_cutoff) %>%
  dplyr::select(gene_id_clean, symbol, condition, timepoint, contrast,
                promoted, te_lfc, te_padj, rna_lfc, ribo_lfc, category) %>%
  arrange(gene_id_clean, contrast)

out_long <- here("output", "genesets", "eif3d_te_all_genes_all_conditions_long.csv")
write_csv(eif3d_te_long_export, out_long)
cat("Wrote", nrow(eif3d_te_long_export), "gene x contrast rows to", out_long, "\n")

# the long file must carry exactly the same promoted calls as the wide file
stopifnot("long/wide promoted calls disagree" =
  sum(eif3d_te_long_export$promoted) == sum(flag_mat))
stopifnot("long file must be gene x contrast complete" =
  nrow(eif3d_te_long_export) == nrow(eif3d_te_all) * length(contrast_levels))
```

## Files written

| File | Contents |
|---|---|
| `eif3d_te_all_genes_all_conditions.csv` | **primary** - every tested gene, one row each, with a promoted flag + TE stats for all 6 contrasts |
| `eif3d_promoted_genes_all_conditions.csv` | the `promoted_any == TRUE` subset, same columns |
| `eif3d_te_all_genes_all_conditions_long.csv` | tidy gene x contrast form, adds `rna_lfc` / `ribo_lfc` / `category` |

## Column dictionary (for the collaborator)

| Column | Meaning |
|---|---|
| `symbol` | HGNC gene symbol (`NA` where GENCODE has none) |
| `gene_id_clean` | ENSEMBL gene ID, version stripped |
| `promoted_any` | TRUE if promoted by eIF3d in at least one of the 6 contrasts |
| `condition_class` | `hypoxia_only` / `normoxia_only` / `both` / `not_promoted` |
| `conditions_promoted` | semicolon-separated contrasts where the gene passed the threshold; `none` if never |
| `n_conditions_promoted` | how many of the 6 contrasts the gene was promoted in (0-6) |
| `promoted_normoxia` / `promoted_hypoxia` | TRUE if promoted in >= 1 timepoint of that condition |
| `max_te_lfc_promoted` | largest TE log2FC **among the contrasts where the gene passed the threshold**; `NA` if never promoted |
| `max_te_lfc_condition` | contrast achieving `max_te_lfc_promoted`; `NA` if never promoted |
| `min_te_padj_promoted` | smallest TE adjusted p-value among those same contrasts; `NA` if never promoted |
| `promoted_{condition}_{timepoint}` | **the per-condition up-flag**: TRUE if `te_padj < 0.05` and `te_lfc > 0.5` in that contrast |
| `te_lfc_{condition}_{timepoint}` | TE log2FC in that contrast, reported whether or not significant |
| `te_padj_{condition}_{timepoint}` | BH-adjusted p-value in that contrast |

Positive `te_lfc` = eIF3d promotes translation of that gene.

**Notes when reading the table:**

1. The file is the **complete tested universe**, so it doubles as its own
   background set. Filter on `promoted_any` for the hit list, or on an individual
   `promoted_*` column for one condition. Genes that are *not* up simply have
   `FALSE` in those columns - their `te_lfc_*` / `te_padj_*` values are still
   reported, so genes going the other way (eIF3d-inhibited, negative `te_lfc`) are
   recoverable from the same file.
2. `max_te_lfc_promoted` summarises only the contrasts where the gene was called
   promoted, so for 65 genes it is **smaller** than the largest value in that row's
   `te_lfc_*` columns. That is not an inconsistency: the larger value sits in a
   contrast that missed the `te_padj < 0.05` cutoff, so the gene was not called
   promoted there. Use the `te_lfc_*` columns for the full picture.
3. `symbol` is **not** a unique key - a handful of symbols map to two ENSEMBL IDs.
   Join and deduplicate on `gene_id_clean`.

`te_padj` is `NA` for a small number of gene x contrast pairs (DESeq2 independent
filtering); those genes were not testable in that contrast and their `promoted_*`
flag is `FALSE`.
