---
title: "71 - GC content vs RNA-level changes across the transcriptome (si3d, hypoxia 1hr)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Does GC content predict how much an mRNA's **abundance** changes when eIF3d is
knocked down? This is the RNA-level counterpart to the TE-level GC analyses in
notebooks 62/63, run across the whole transcriptome rather than within gene sets.

The short answer is no, and that is the point of the notebook. GC content is the
strongest single predictor of eIF3d **TE** dependence (`cds_gc` is the top feature
of the notebook 07 TE classifier) but is close to inert at the RNA level, where
5'UTR *length* dominates instead. This notebook quantifies that dissociation on
the full transcriptome three ways: Spearman correlation against both readouts,
binned medians, and a per-feature TE-vs-RNA fit, plus a gene-set comparison of GC
content against the background transcriptome.

**Contrast.** sictrl vs si3d at hypoxia 1hr, read from the pre-computed
`output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv`. Nothing is refit.
This is the eIF3d-knockdown effect, *not* the hypoxia response (sictrl hypoxia vs
normoxia), which is a different contrast and a different file.

**Direction convention.** The source notebook runs DESeq2 with the knockdown as
the reference (`control_condition = "si3d"`, `test_condition = "sictrl"`), so the
coefficient is log2(sictrl / si3d) despite the file name:

- `rna_lfc > 0` = knockdown **reduces** the mRNA = eIF3d promotes its abundance
- `te_lfc > 0` = knockdown reduces TE = eIF3d promotes its translation

Asserted below against EIF3D itself, whose own mRNA must drop in si3d.

**Global shift.** These contrasts are spike-in normalized and eIF3d knockdown
lowers total mRNA per cell, so the median `rna_lfc` is ~0.41, not 0. Every plot
draws the observed median rather than a zero line, and "up" / "down" throughout
means relative to that median, not to zero.

**Effect sizes, not p-values.** n is ~10,000 genes, so a Spearman rho of 0.03 is
"significant" at p = 0.002 while explaining 0.1% of the variance. Every
correlation is reported with rho and n together, and the binned medians are shown
so the magnitude is visible directly. Do not read the p column as evidence of a
meaningful relationship.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
```

# Parameters

```{r params}
CONTRAST_FILE <- "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"

# GC features to test, in the order they are reported
GC_FEATURES <- c(
  "utr5_gc"       = "5'UTR GC%",
  "cds_gc"        = "CDS GC%",
  "utr3_gc"       = "3'UTR GC%",
  "gc3"           = "GC3",
  "transcript_gc" = "Transcript GC%"
)

# Length features, used as the positive control: the analysis must be able to
# detect a real effect, and 5'UTR length is the one that is actually there.
LEN_FEATURES <- c(
  "log2_utr5_length" = "log2 5'UTR length",
  "log2_cds_length"  = "log2 CDS length",
  "log2_utr3_length" = "log2 3'UTR length"
)

N_BINS <- 10   # deciles for the binned view

# Panel order for every faceted plot. Without an explicit factor, facet_wrap()
# sorts labels alphabetically, which interleaves the GC and Length blocks and
# strands "Transcript GC%" among the length panels. GC first, then Length.
# There are 5 GC and 3 Length features, so faceting at ncol = 5 puts each block
# on its own row.
FEATURE_ORDER <- c(unname(GC_FEATURES), unname(LEN_FEATURES))

# Gene-set names and order. (Colors retained as the canonical group definition;
# the violin plots that consumed them have been removed.)
#   Project convention reserves #E69F00 for eIF3d and #56B4E9 for eIF3e. Both
#   positive sets here are eIF3d-defined, so the established TE set keeps
#   #E69F00 (backward compatible with already-generated figures) and the newer
#   RNA-defined set takes Okabe-Ito vermillion - a distinct hue that cannot be
#   mistaken for the eIF3e blue.
#   "eIF3d-protected (RNA)" is NOT called a negative control here. nb70 defines it
#   as |rna_lfc| < 0.5 & padj > 0.2, i.e. relative to ZERO, while this notebook's
#   null is the global median (~0.43). Those genes therefore RETAINED mRNA while
#   the typical transcript lost ~25-30% - they are a regulated set in their own
#   right, sitting ~0.39 log2 below the transcriptome median, not an unaffected
#   background. nb70 says the same thing ("eIF3d-dependent vs eIF3d-protected");
#   labelling it "negative control" in a GC plot would invite reading its GC
#   shift as "even the control shows the effect".
group_colors <- c(
  "eIF3d TE promoted"      = "#E69F00",
  "eIF3d RNA promoted"     = "#D55E00",
  "Negative control (TE)"  = "black",
  "eIF3d-protected (RNA)"  = "grey40",
  "All other genes"        = "grey70"
)
GROUP_LEVELS <- names(group_colors)
```

# Load Data

```{r load_contrast}
contrast_path <- here("output", CONTRAST_FILE)
stopifnot("contrast file not found" = file.exists(contrast_path))

de <- read_csv(contrast_path, show_col_types = FALSE) %>%
  dplyr::select(gene_id, symbol, rna_lfc, rna_padj, ribo_lfc, ribo_padj,
                te_lfc, te_padj) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("Contrast rows:", nrow(de), "\n")
cat("Median rna_lfc:", round(median(de$rna_lfc, na.rm = TRUE), 3), "\n")
cat("Median te_lfc :", round(median(de$te_lfc,  na.rm = TRUE), 3), "\n")
```

```{r load_features}
features <- readRDS(here("output", "predictive_modeling",
                         "feature_matrix_external_stability.rds"))

# Feature matrices are transcript-keyed in general. This one happens to carry one
# row per gene, but assert it rather than assume - a duplicated gene would double
# -weight it in every correlation below.
stopifnot("feature matrix is not one row per gene" =
  !anyDuplicated(features$gene_id_clean))

keep_cols <- c("gene_id_clean", "symbol", names(GC_FEATURES), names(LEN_FEATURES))
stopifnot("expected feature columns missing from the matrix" =
  all(keep_cols %in% colnames(features)))

feat <- features %>% dplyr::select(all_of(keep_cols)) %>% dplyr::rename(symbol_fm = symbol)

cat("Feature matrix:", nrow(feat), "genes x", ncol(feat), "columns\n")
```

```{r join}
dat <- de %>%
  inner_join(feat, by = "gene_id_clean")

cat("Genes with both a contrast result and features:", nrow(dat), "\n")
cat("  contrast only:", nrow(de) - nrow(dat), "\n")
stopifnot("join produced duplicate genes" = !anyDuplicated(dat$gene_id_clean))
stopifnot("join lost too many genes - check ID versioning" = nrow(dat) > 8000)

GLOBAL_MEDIAN <- median(dat$rna_lfc, na.rm = TRUE)
cat("Global median rna_lfc in the joined set:", round(GLOBAL_MEDIAN, 3), "\n")
```

# Validation

## Computational checks

```{r checks_computational}
# GC percentages must be on a 0-100 scale
for (v in names(GC_FEATURES)) {
  stopifnot(all(dat[[v]] >= 0 & dat[[v]] <= 100, na.rm = TRUE))
}
cat("All GC features within [0, 100]: PASS\n")

# Log-fold changes must be finite where present
for (v in c("rna_lfc", "te_lfc")) {
  stopifnot(all(is.finite(dat[[v]][!is.na(dat[[v]])])))
}
cat("rna_lfc / te_lfc finite: PASS\n")

# Lengths are log2 of a positive quantity, so they must be finite and non-negative
for (v in names(LEN_FEATURES)) {
  stopifnot(all(is.finite(dat[[v]][!is.na(dat[[v]])])))
}
cat("Length features finite: PASS\n")

# Leakage guard. This must run against the SOURCE objects, not against `dat`:
# `dat` is assembled from explicit dplyr::select() calls on both sides, so a
# stray label column could never survive into it and the check would be dead code
# passing on every input.
stopifnot("model label column present in the feature matrix" =
  !"eif3d_target" %in% colnames(features))
stopifnot("model label column present in the contrast table" =
  !"eif3d_target" %in% colnames(de))
cat("No model-label columns in either source table: PASS\n")
```

## Biological spot-checks

```{r checks_biological}
# 1. Direction convention. EIF3D is the knockdown target, so its own mRNA must be
#    depleted in si3d, which under log2(sictrl / si3d) means strongly positive.
eif3d <- dat %>% filter(symbol == "EIF3D")
stopifnot("EIF3D not found in the joined data" = nrow(eif3d) == 1)
stopifnot("EIF3D rna_lfc is not positive - direction convention may have flipped" =
  eif3d$rna_lfc > 0.5)
cat("EIF3D rna_lfc =", round(eif3d$rna_lfc, 2),
    "-> log2(sictrl / si3d) confirmed\n")

# 2. GC3 median must sit in the established 40-65% band (same check as 01_ / 43_)
gc3_med <- median(dat$gc3, na.rm = TRUE)
stopifnot("GC3 median outside the expected 40-65% band" =
  between(gc3_med, 40, 65))
cat("GC3 median:", round(gc3_med, 1), "%\n")

# 3. cds_gc and gc3 are known to be ~0.93 correlated via isochore structure
#    (documented in CLAUDE.md). They are NOT independent tests and must not be
#    counted as two pieces of evidence.
gc_collin <- cor(dat$cds_gc, dat$gc3, use = "complete.obs")
cat("cor(cds_gc, gc3) =", round(gc_collin, 3),
    "- these are one measurement, not two\n")
stopifnot("cds_gc/gc3 collinearity outside the documented range" =
  gc_collin > 0.85)
```

# Continuous Relationship

```{r spearman_table}
# Spearman rho for every GC and length feature against both readouts.
# n is carried alongside rho throughout: at n ~ 10,000 the p-value is not the
# quantity of interest and is reported only for completeness.
cor_one <- function(x, y, feature, label, readout) {
  ok <- complete.cases(x, y)
  ct <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman"))
  tibble(feature = feature, label = label, readout = readout,
         rho = unname(ct$estimate), p = ct$p.value, n = sum(ok))
}

all_feats <- c(GC_FEATURES, LEN_FEATURES)
cor_tbl <- map2_dfr(names(all_feats), unname(all_feats), function(f, lab) {
  bind_rows(
    cor_one(dat[[f]], dat$rna_lfc,  f, lab, "RNA"),
    cor_one(dat[[f]], dat$ribo_lfc, f, lab, "RIBO"),
    cor_one(dat[[f]], dat$te_lfc,   f, lab, "TE")
  )
}) %>%
  mutate(kind = if_else(feature %in% names(GC_FEATURES), "GC", "Length"))

cor_wide <- cor_tbl %>%
  dplyr::select(kind, label, readout, rho, n) %>%
  pivot_wider(names_from = readout, values_from = c(rho, n)) %>%
  arrange(kind, desc(abs(rho_RNA)))

print(cor_wide, n = 20)
```

The comparison that matters is the `rho_RNA` column against `rho_TE` within the GC
block, and then the GC block against the Length block.

```{r binned_medians}
# Binned medians make the effect size visible in the units of the data, which a
# correlation coefficient does not. Bootstrap CI on each bin median.
set.seed(9)
N_BOOT <- 1000

boot_median_ci <- function(x, n_boot = N_BOOT) {
  x <- x[!is.na(x)]
  if (length(x) < 10) return(c(lo = NA_real_, hi = NA_real_))
  bs <- replicate(n_boot, median(sample(x, length(x), replace = TRUE)))
  q <- unname(quantile(bs, c(0.025, 0.975)))
  c(lo = q[1], hi = q[2])
}

# Both bounds must come from ONE resample distribution. Calling boot_median_ci()
# twice (once for [1], once for [2]) would draw the lower bound from one bootstrap
# run and the upper from a different one - not a coherent interval, and in a
# degenerate bin the two can cross and render an inverted errorbar.
binned <- map_dfr(names(all_feats), function(f) {
  d <- dat %>%
    dplyr::select(value = all_of(f), rna_lfc) %>%
    filter(!is.na(value), !is.na(rna_lfc)) %>%
    mutate(bin = ntile(value, N_BINS))
  d %>%
    group_by(bin) %>%
    summarise(
      median_rna_lfc = median(rna_lfc),
      ci = list(boot_median_ci(rna_lfc)),
      n = dplyr::n(),
      .groups = "drop"
    ) %>%
    mutate(ci_lo = map_dbl(ci, "lo"),
           ci_hi = map_dbl(ci, "hi")) %>%
    dplyr::select(-ci) %>%
    mutate(feature = f,
           label = factor(unname(all_feats[f]), levels = FEATURE_ORDER),
           kind = if_else(f %in% names(GC_FEATURES), "GC", "Length"))
})

stopifnot("bootstrap interval bounds are inverted" =
  all(binned$ci_lo <= binned$ci_hi, na.rm = TRUE))

# n per feature, used to annotate panels: utr3_gc / transcript_gc carry 134 NAs,
# so a single transcriptome-wide n would overstate three of the eight panels.
feat_n <- binned %>% group_by(label) %>%
  summarise(n_feat = sum(n), .groups = "drop") %>%
  mutate(txt = paste0("n = ", n_feat))

# Range of the binned medians = the effect size in log2FC units
bin_range <- binned %>%
  group_by(kind, label) %>%
  summarise(span = max(median_rna_lfc) - min(median_rna_lfc), .groups = "drop") %>%
  arrange(desc(span))
print(bin_range)
```

```{r plot_binned, fig.width=10, fig.height=6}
p_binned <- ggplot(binned, aes(x = bin, y = median_rna_lfc)) +
  geom_hline(yintercept = GLOBAL_MEDIAN, linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0.2,
                colour = "grey40", linewidth = 0.4) +
  geom_point(aes(colour = kind), size = 1.8) +
  geom_line(aes(colour = kind), linewidth = 0.8) +
  geom_text(data = feat_n, aes(x = Inf, y = -Inf, label = txt),
            inherit.aes = FALSE, hjust = 1.1, vjust = -0.8,
            size = 2.8, colour = "grey30") +
  scale_colour_manual(values = c("GC" = "#0072B2", "Length" = "#CC79A7"),
                      name = NULL) +
  scale_x_continuous(breaks = seq(1, N_BINS, 2)) +
  # 5 GC + 3 Length features: ncol = 5 puts each block on its own row.
  # x is the decile index in every panel, so it must NOT be free.
  facet_wrap(~ label, ncol = 5) +
  labs(
    title = "RNA change by feature decile (sictrl vs si3d, hypoxia 1hr)",
    subtitle = paste0("Dashed line = observed global median rna_lfc (",
                      round(GLOBAL_MEDIAN, 2),
                      "). Bars = bootstrap 95% CI on the bin median. n per panel."),
    x = paste0("Feature decile (1 = lowest, ", N_BINS, " = highest)"),
    y = "Median rna_lfc, log2(sictrl / si3d)"
  ) +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "top")
p_binned
ggsave(here("plots", "71_rna_change_by_feature_decile.pdf"),
       p_binned, width = 10, height = 6)
```

```{r paired_data}
# Paired TE-vs-RNA view: one row per feature, TE on the left, RNA on the right.
#
# The two readouts have different global medians (te_lfc 0.289, rna_lfc 0.409),
# both non-zero because of the spike-in shift. Plotting raw LFC would put the two
# columns on different baselines and make the within-row comparison misleading.
# Each readout is therefore CENTRED on its own global median, so:
#   y = 0 means "behaves like the typical transcript for this readout"
# and the two columns become directly comparable on one shared axis.
# Panel order left to right. RIBO is appended so the existing TE-left / RNA-right
# layout is preserved. All three legs are spike-in normalized log2(sictrl / si3d)
# from the same producing notebook.
READOUTS <- c("te_lfc" = "TE change", "rna_lfc" = "RNA change",
              "ribo_lfc" = "RIBO change")

readout_medians <- map_dbl(names(READOUTS), ~ median(dat[[.x]], na.rm = TRUE))
names(readout_medians) <- names(READOUTS)
print(round(readout_medians, 3))

# Panel order: feature (rows) x readout (columns), TE before RNA within each row.
PANEL_LEVELS <- as.vector(t(outer(FEATURE_ORDER, unname(READOUTS),
                                  function(a, b) paste0(a, "  |  ", b))))

paired_long <- map_dfr(names(READOUTS), function(rc) {
  dat %>%
    dplyr::select(lfc = all_of(rc), all_of(names(all_feats))) %>%
    pivot_longer(-lfc, names_to = "feature", values_to = "value") %>%
    filter(!is.na(value), !is.na(lfc)) %>%
    mutate(readout_col = rc,
           readout = unname(READOUTS[rc]),
           centered = lfc - readout_medians[[rc]],
           label = unname(all_feats[feature]))
}) %>%
  mutate(panel = factor(paste0(label, "  |  ", readout), levels = PANEL_LEVELS),
         kind  = if_else(feature %in% names(GC_FEATURES), "GC", "Length"))

stopifnot("panel factor has unmatched levels" = !anyNA(paired_long$panel))

# Binned medians per panel, on the centred scale
paired_med <- paired_long %>%
  group_by(panel) %>%
  mutate(bin = ntile(value, N_BINS)) %>%
  group_by(panel, kind, bin) %>%
  summarise(x = median(value), y = median(centered), n = dplyr::n(), .groups = "drop")

# rho + n annotation, matched to the panel it belongs to
# cor_tbl already carries `feature`, which the per-feature plots filter on.
# Map the short readout names in cor_tbl onto the panel labels. A lookup rather
# than nested if_else, so adding a readout cannot silently mislabel a panel.
READOUT_LABEL <- c("TE" = "TE change", "RNA" = "RNA change", "RIBO" = "RIBO change")
rho_lab <- cor_tbl %>%
  mutate(readout = unname(READOUT_LABEL[readout]),
         panel = factor(paste0(label, "  |  ", readout), levels = PANEL_LEVELS),
         txt = paste0("rho = ", sprintf("%+.3f", rho), "\nn = ", n))
stopifnot("rho annotation did not match a panel" = !anyNA(rho_lab$panel))
stopifnot("rho_lab is missing the feature column" = "feature" %in% colnames(rho_lab))

# Shared y limits from the pooled centred values, clipped to 1-99% so a handful
# of outliers cannot stretch the axis and flatten every trend (the within-decile
# SD is ~0.39 while the raw range spans nearly 7 log2 units).
Y_LIM <- quantile(paired_long$centered, c(0.01, 0.99), na.rm = TRUE)
cat("shared y limits (centred):", round(Y_LIM, 2), "\n")
```

```{r plot_per_feature, fig.width=12, fig.height=4}
# One figure per feature, each faceted TE (left) vs RNA (right).
#
# The trend is a straight least-squares fit, not the decile medians: the question
# here is "how does this feature relate to each readout", and a single fitted line
# per panel makes the left-right slope comparison immediate.
#
# coord_cartesian() only zooms the viewport - it does NOT drop rows before the
# fit - so geom_smooth is computed on all genes even though the extreme 1% is off
# screen. (scale_y_continuous(limits=) would have silently refit on the subset.)
kind_colour <- c("GC" = "#D55E00", "Length" = "#CC79A7")

feature_plots <- map(names(all_feats), function(f) {
  lab  <- unname(all_feats[f])
  kind <- if (f %in% names(GC_FEATURES)) "GC" else "Length"

  pd <- paired_long %>%
    filter(feature == f) %>%
    mutate(readout_f = factor(readout, levels = unname(READOUTS)))

  rl <- rho_lab %>%
    filter(feature == f) %>%
    mutate(readout_f = factor(readout, levels = unname(READOUTS)))
  stopifnot("rho annotation missing a readout for this feature" =
    nrow(rl) == length(READOUTS))

  ggplot(pd, aes(x = value, y = centered)) +
    geom_point(alpha = 0.15, size = 0.35, colour = "grey45") +
    geom_hline(yintercept = 0, linetype = "dashed",
               colour = "#0072B2", linewidth = 0.5) +
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
                colour = kind_colour[[kind]], fill = kind_colour[[kind]],
                alpha = 0.25, linewidth = 1) +
    geom_text(data = rl, aes(x = Inf, y = Inf, label = txt),
              inherit.aes = FALSE, hjust = 1.1, vjust = 1.4,
              size = 3.2, colour = "grey20") +
    coord_cartesian(ylim = Y_LIM) +
    facet_wrap(~ readout_f, ncol = 3) +
    labs(
      title = paste0(lab, ": eIF3d-mediated TE change vs RNA change"),
      subtitle = paste0("sictrl vs si3d, hypoxia 1hr. Facets centered on each ",
                        "readout's global median; dashed = median.\n",
                        "Line = linear fit with 95% CI, computed on all genes."),
      x = lab,
      y = "log2FC (sictrl / si3d), 1hr hypoxia"
    ) +
    theme_classic(base_size = 12) +
    theme(strip.text = element_text(face = "bold"))
})
names(feature_plots) <- names(all_feats)

# Save one PDF per feature
iwalk(feature_plots, function(p, f) {
  ggsave(here("plots", paste0("71_te_vs_rna_", f, ".pdf")),
         p, width = 12, height = 4)
})
cat("Wrote", length(feature_plots), "per-feature PDFs to plots/\n")
print(paste0("71_te_vs_rna_", names(feature_plots), ".pdf"))
```

```{r show_per_feature, fig.width=12, fig.height=4, results='asis'}
# Render each feature in turn, GC block first then Length (FEATURE_ORDER).
ordered_feats <- names(all_feats)[order(match(unname(all_feats), FEATURE_ORDER))]
for (f in ordered_feats) {
  cat("\n\n## ", unname(all_feats[f]), "\n\n", sep = "")
  print(feature_plots[[f]])
}
```

```{r fit_slopes}
# The slope behind each fitted line, so the visual comparison has a number.
# Reported per 10 units of the feature for GC (percentage points) and per log2
# unit for the length features.
slope_tbl <- map_dfr(names(all_feats), function(f) {
  map_dfr(names(READOUTS), function(rc) {
    d <- paired_long %>% filter(feature == f, readout_col == rc)
    fit <- lm(centered ~ value, data = d)
    tibble(feature = unname(all_feats[f]),
           readout = unname(READOUTS[rc]),
           slope_per_unit = unname(coef(fit)[2]),
           n = nrow(d))
  })
}) %>%
  pivot_wider(names_from = readout, values_from = c(slope_per_unit, n)) %>%
  dplyr::rename(slope_TE = `slope_per_unit_TE change`,
                slope_RNA = `slope_per_unit_RNA change`) %>%
  mutate(across(c(slope_TE, slope_RNA), ~ signif(.x, 3))) %>%
  dplyr::select(feature, slope_TE, slope_RNA, `n_TE change`)
print(as.data.frame(slope_tbl))
write_csv(slope_tbl, here("output", "71_te_vs_rna_fit_slopes.csv"))
```

# Confounder Check: GC vs Length

GC content and 5'UTR length are not independent, and 5'UTR length carries the real
signal. If the small GC correlations are just length leaking through, they should
vanish once length is partialled out.

```{r partial_correlation}
# Spearman partial correlation: rank-transform, regress out the length ranks, then
# correlate the residuals. Done manually so this does not depend on ppcor.
partial_spearman <- function(x, y, z) {
  ok <- complete.cases(x, y, z)
  rx <- rank(x[ok]); ry <- rank(y[ok]); rz <- rank(z[ok])
  ex <- residuals(lm(rx ~ rz))
  ey <- residuals(lm(ry ~ rz))
  c(rho = unname(cor(ex, ey)), n = sum(ok))
}

# Each n is carried with the rho it belongs to. The partials drop rows on
# complete.cases(x, y, z) while the raw correlation drops only on (x, y), so
# reusing the raw n would overstate the partials' sample size the moment a
# conditioning variable carries NAs (log2_utr3_length has 134).
partial_tbl <- map_dfr(names(GC_FEATURES), function(f) {
  raw <- cor_one(dat[[f]], dat$rna_lfc, f, unname(GC_FEATURES[f]), "RNA")
  pu  <- partial_spearman(dat[[f]], dat$rna_lfc, dat$log2_utr5_length)
  pc  <- partial_spearman(dat[[f]], dat$rna_lfc, dat$log2_cds_length)
  tibble(label = unname(GC_FEATURES[f]),
         rho_raw = raw$rho,                n_raw = raw$n,
         rho_given_utr5_len = pu[["rho"]], n_given_utr5_len = pu[["n"]],
         rho_given_cds_len  = pc[["rho"]], n_given_cds_len  = pc[["n"]])
})
print(partial_tbl)

cat("\ncor(cds_gc, log2_utr5_length)  =",
    round(cor(dat$cds_gc, dat$log2_utr5_length, use = "complete.obs"), 3), "\n")
cat("cor(utr5_gc, log2_utr5_length) =",
    round(cor(dat$utr5_gc, dat$log2_utr5_length, use = "complete.obs"), 3), "\n")
```

# Gene Sets vs Whole Transcriptome

Violin comparison of GC content between the eIF3d-defined gene sets and the rest
of the transcriptome, in the style of notebooks 48/62.

A gene may belong to more than one set (the RNA- and TE-promoted sets overlap),
so groups are built in **long format**: each set is drawn independently from the
full table and a gene contributes a row to every set it belongs to. "All other
genes" is the complement of the union of all four sets. This avoids an arbitrary
priority rule silently reassigning the overlap.

```{r load_genesets}
read_set <- function(path, col) {
  stopifnot("gene set file not found" = file.exists(path))
  read_csv(path, show_col_types = FALSE) %>% pull({{ col }}) %>% unique()
}

set_rna_pos <- read_set(here("output", "genesets",
                             "hypoxia_3d_promotes_RNA_1hr_lfc1.csv"), ensembl_gene)
set_te_pos  <- read_set(here("output", "genesets",
                             "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"), ensembl_gene)
set_rna_neg <- read_set(here("output", "genesets",
                             "si3d_rna_negative_controls_nor1and4_hyp1_hyp4.csv"), ensembl_gene)
set_te_neg  <- read_csv(here("output", "predictive_modeling",
                             "negative_control_genes.csv"), show_col_types = FALSE) %>%
  mutate(g = sub("\\..*", "", gene_id)) %>% pull(g) %>% unique()

set_list <- list(
  "eIF3d RNA promoted"    = set_rna_pos,
  "eIF3d TE promoted"     = set_te_pos,
  "eIF3d-protected (RNA)" = set_rna_neg,
  "Negative control (TE)" = set_te_neg
)

set_sizes <- imap_dfr(set_list, ~ tibble(
  group = .y, n_total = length(.x),
  n_in_data = sum(.x %in% dat$gene_id_clean)
))
print(set_sizes)
stopifnot("a gene set has no overlap with the joined data" = all(set_sizes$n_in_data > 0))
```

## Cross-set overlap

The four sets are **not** mutually exclusive, and the cross-readout overlap is
large enough to change how the violins must be read. Report the full matrix
rather than the positive-positive pair alone.

```{r overlap_matrix}
in_data <- map(set_list, ~ intersect(unique(.x), dat$gene_id_clean))

ov <- outer(seq_along(in_data), seq_along(in_data),
            Vectorize(function(i, j) length(intersect(in_data[[i]], in_data[[j]]))))
dimnames(ov) <- list(names(in_data), names(in_data))
print(ov)

# Within a readout the positive and negative sets are disjoint by construction.
# Assert that, since a violation would mean a gene set was rebuilt incorrectly.
stopifnot("RNA positive and RNA-protected sets overlap" =
  length(intersect(in_data[["eIF3d RNA promoted"]],
                   in_data[["eIF3d-protected (RNA)"]])) == 0)
stopifnot("TE positive and TE negative sets overlap" =
  length(intersect(in_data[["eIF3d TE promoted"]],
                   in_data[["Negative control (TE)"]])) == 0)

cross <- length(intersect(in_data[["eIF3d RNA promoted"]],
                          in_data[["Negative control (TE)"]]))
cat("\nACROSS readouts the sets overlap substantially:\n")
cat("  RNA-promoted genes that are also TE negative controls:", cross,
    sprintf("(%.0f%% of the RNA-promoted set)\n",
            100 * cross / length(in_data[["eIF3d RNA promoted"]])))
cat("  These are genes whose mRNA drops sharply while TE holds constant -\n",
    "  the 'forwarded' category, a real biological class rather than an artifact.\n")
cat("  Consequence: 'Negative control (TE)' median rna_lfc is inflated by them.\n")

te_neg_all <- dat %>% filter(gene_id_clean %in% in_data[["Negative control (TE)"]])
te_neg_excl <- dat %>%
  filter(gene_id_clean %in% setdiff(in_data[["Negative control (TE)"]],
                                    in_data[["eIF3d RNA promoted"]]))
cat("    as drawn                :", round(median(te_neg_all$rna_lfc), 3), "\n")
cat("    minus RNA-promoted genes:", round(median(te_neg_excl$rna_lfc), 3), "\n")
```

```{r build_groups}
in_any <- unique(unlist(set_list))

grouped <- bind_rows(
  imap_dfr(set_list, function(ids, nm) {
    dat %>% filter(gene_id_clean %in% ids) %>% mutate(group = nm)
  }),
  dat %>% filter(!gene_id_clean %in% in_any) %>% mutate(group = "All other genes")
) %>%
  mutate(group = factor(group, levels = GROUP_LEVELS))

group_n <- grouped %>% dplyr::count(group)
print(group_n)

# dplyr::count() defaults to .drop = TRUE, so an empty factor level is ABSENT
# from group_n rather than present with n = 0 - which makes all(group_n$n > 0)
# vacuously true and would let a whole group vanish from every plot silently
# (e.g. if a geneset CSV were regenerated with versioned Ensembl IDs, or a
# set_list name drifted out of sync with GROUP_LEVELS and coerced to NA).
# Check the levels are present, not that the counts are positive.
stopifnot("a group is missing entirely from the plotting data" =
  all(GROUP_LEVELS %in% as.character(group_n$group)))
stopifnot("a group name did not match GROUP_LEVELS and became NA" =
  !anyNA(grouped$group))
```

```{r wilcox_vs_background}
# Each set against "All other genes", which is the comparison the plot invites.
bg <- grouped %>% filter(group == "All other genes")

wilcox_tbl <- map_dfr(names(set_list), function(g) {
  map_dfr(names(GC_FEATURES), function(f) {
    a <- grouped %>% filter(group == g) %>% pull(all_of(f))
    b <- bg %>% pull(all_of(f))
    a <- a[!is.na(a)]; b <- b[!is.na(b)]
    tibble(group = g, feature = unname(GC_FEATURES[f]),
           median_set = median(a), median_bg = median(b),
           delta = median(a) - median(b),
           p_wilcox = suppressWarnings(wilcox.test(a, b)$p.value),
           n_set = length(a), n_bg = length(b))
  })
})

wilcox_tbl <- wilcox_tbl %>% mutate(p_adj = p.adjust(p_wilcox, method = "BH"))
print(wilcox_tbl %>% arrange(desc(abs(delta))), n = 20)
```

`delta` is the quantity to read: it is in GC percentage points. With n in the
thousands the p-values go significant on differences of well under one point.

```{r group_sanity_check}
# Grouping sanity check (kept after the violin was removed): the RNA-promoted set
# is defined as rna_lfc > 1, so it must exceed every other group. If this fails,
# the grouping or the join is wrong.
med_by_group <- grouped %>% group_by(group) %>%
  summarise(m = median(rna_lfc, na.rm = TRUE), .groups = "drop")
print(med_by_group)
stopifnot("RNA-promoted set does not have the highest median rna_lfc" =
  med_by_group$m[med_by_group$group == "eIF3d RNA promoted"] == max(med_by_group$m))
```

# Summary

```{r summary}
summary_tbl <- cor_wide %>%
  mutate(across(c(rho_RNA, rho_TE), ~ round(.x, 3))) %>%
  dplyr::rename(feature = label)
print(summary_tbl, n = 20)

write_csv(cor_tbl,     here("output", "71_gc_vs_rna_correlations.csv"))
write_csv(binned,      here("output", "71_gc_vs_rna_binned_medians.csv"))
write_csv(wilcox_tbl,  here("output", "71_gc_geneset_vs_background.csv"))
write_csv(partial_tbl, here("output", "71_gc_partial_correlations.csv"))

cat("\nWrote 4 summary tables to output/ and 5 plots to plots/\n")
```

```{r headline}
best_gc  <- cor_tbl %>% filter(readout == "RNA", feature %in% names(GC_FEATURES)) %>%
  slice_max(abs(rho), n = 1)
best_len <- cor_tbl %>% filter(readout == "RNA", feature %in% names(LEN_FEATURES)) %>%
  slice_max(abs(rho), n = 1)
gc_te    <- cor_tbl %>% filter(readout == "TE", feature == "cds_gc")

cat("Strongest GC correlate of rna_lfc    :", best_gc$label,
    sprintf("rho = %+.3f (n = %d)", best_gc$rho, best_gc$n), "\n")
cat("Strongest length correlate of rna_lfc:", best_len$label,
    sprintf("rho = %+.3f (n = %d)", best_len$rho, best_len$n), "\n")
cat("cds_gc vs te_lfc, same genes         :",
    sprintf("rho = %+.3f (n = %d)", gc_te$rho, gc_te$n), "\n")
cat("\nGC decile span in rna_lfc units:",
    round(bin_range$span[bin_range$label == unname(GC_FEATURES["cds_gc"])], 3), "\n")
```
