---
title: "72 - Overlap of eIF3d RNA-promoted gene sets: hypoxia vs normoxia"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

How much do the two **RNA-level** eIF3d-promoted gene sets built by notebook 70
share? Both are `rna_lfc > 1 & padj < 0.05` on the sictrl vs si3d contrast, one at
hypoxia 1hr and one at the time-regressed normoxia 1+4hr fit.

- `output/genesets/hypoxia_3d_promotes_RNA_1hr_lfc1.csv`
- `output/genesets/normoxia_3d_promotes_RNA_1and4hr_lfc1.csv`

Nothing is refit or re-thresholded here; the sets are read as written.

**Direction convention.** Both contrasts are log2(sictrl / si3d), so `rna_lfc > 0`
means the knockdown reduced the mRNA, i.e. eIF3d promotes its abundance. EIF3D
itself is excluded from both sets by notebook 70 (it qualifies overwhelmingly, but
that is the siRNA depleting its own target rather than eIF3d-dependent biology).

**The two contrasts are not equally powered, and that shapes the reading.** This
is the point of the last section rather than a footnote:

| | hypoxia 1hr | normoxia 1+4hr |
|---|---|---|
| fit | single timepoint, `deFunction()` | two timepoints, `hours` as batch covariate, `deFunction_time_normalized()` |
| global median `rna_lfc` | 0.409 | 0.505 |

Normoxia draws on more samples and carries a larger global shift, so a gene can
land in the normoxia set and miss the hypoxia set purely on detection power. A
Venn diagram alone cannot separate "condition-specific" from "underpowered", so
the counts are followed by an explicit test of where the non-shared genes actually
sit in the other contrast.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggvenn)
```

# Parameters

```{r params}
SET_FILES <- c(
  "Hypoxia 1hr"     = "hypoxia_3d_promotes_RNA_1hr_lfc1.csv",
  "Normoxia 1+4hr"  = "normoxia_3d_promotes_RNA_1and4hr_lfc1.csv",
  "Normoxia 1hr"    = "normoxia_3d_promotes_RNA_1hr_lfc1.csv"
)

# Source contrasts, used only for the power section
CONTRAST_FILES <- c(
  "Hypoxia 1hr"    = "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
  "Normoxia 1+4hr" = "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv",
  "Normoxia 1hr"   = "translation_categories_si3d_vs_sictrl_normoxia_1hr.csv"
)

# The threshold both sets were built with (notebook 70)
POS_LFC  <- 1.0
POS_PADJ <- 0.05

# Okabe-Ito vermillion (red-orange) for hypoxia, bluish green for normoxia.
# Both sets are eIF3d-defined, so neither uses the reserved eIF3d amber (#E69F00)
# or eIF3e sky blue (#56B4E9) - nothing here should read as one of those factors.
set_colors <- c("Hypoxia 1hr"    = "#D55E00",
                "Normoxia 1+4hr" = "#009E73",
                "Normoxia 1hr"   = "#009E73")
```

# Load Gene Sets

```{r load_sets}
paths <- here("output", "genesets", SET_FILES)
stopifnot("gene set file not found - run notebook 70 for both configurations" =
  all(file.exists(paths)))

sets <- map(set_names(seq_along(SET_FILES), names(SET_FILES)), function(i) {
  read_csv(paths[i], show_col_types = FALSE) %>% pull(ensembl_gene) %>% unique()
})

walk2(names(sets), sets, ~ cat(.x, ":", length(.y), "genes\n"))
```

## Validation

```{r checks}
# Version-stripped ENSEMBL IDs, or the intersect below silently under-counts
walk2(names(sets), sets, function(nm, ids) {
  stopifnot("ids are not version-stripped ENSG" = all(grepl("^ENSG[0-9]+$", ids)))
})
cat("Both sets carry version-stripped ENSG ids: PASS\n")

# unique() above would mask a duplicated file; check the raw column instead
walk2(names(SET_FILES), paths, function(nm, p) {
  raw <- read_csv(p, show_col_types = FALSE)$ensembl_gene
  stopifnot("duplicate gene ids in a gene set file" = !anyDuplicated(raw))
})
cat("No duplicate ids within either file: PASS\n")

# Notebook 70 excludes the knockdown target from both sets
symbols <- map(paths, ~ read_csv(.x, show_col_types = FALSE)$db_gene_symbol)
stopifnot("EIF3D should have been excluded by notebook 70" =
  !any(map_lgl(symbols, ~ "EIF3D" %in% .x)))
cat("EIF3D absent from both sets (excluded upstream): PASS\n")

# Both sets must be non-trivial and must actually overlap
stopifnot("a gene set is empty" = all(lengths(sets) > 0))
stopifnot("the two sets do not overlap at all - check the ID space" =
  length(intersect(sets[[1]], sets[[2]])) > 0)
```

# Overlap

```{r overlap_counts}
H <- sets[["Hypoxia 1hr"]]
N <- sets[["Normoxia 1+4hr"]]

shared       <- intersect(H, N)
hypoxia_only <- setdiff(H, N)
normoxia_only <- setdiff(N, H)

overlap_tbl <- tibble(
  region = c("Shared", "Hypoxia only", "Normoxia only",
             "Hypoxia total", "Normoxia total", "Union"),
  n = c(length(shared), length(hypoxia_only), length(normoxia_only),
        length(H), length(N), length(union(H, N)))
) %>%
  # Percentages only mean something for the three disjoint regions; on the total
  # and union rows they exceed 100 and read as an error.
  mutate(pct_of_hypoxia  = if_else(region %in% c("Shared", "Hypoxia only", "Normoxia only"),
                                   round(100 * n / length(H), 1), NA_real_),
         pct_of_normoxia = if_else(region %in% c("Shared", "Hypoxia only", "Normoxia only"),
                                   round(100 * n / length(N), 1), NA_real_))
print(overlap_tbl)

stopifnot("region counts do not reconcile with the set sizes" =
  length(shared) + length(hypoxia_only) == length(H) &&
  length(shared) + length(normoxia_only) == length(N))

cat("\nThe overlap is asymmetric:\n")
cat(sprintf("  %.0f%% of the hypoxia set is inside the normoxia set\n",
            100 * length(shared) / length(H)))
cat(sprintf("  %.0f%% of the normoxia set is inside the hypoxia set\n",
            100 * length(shared) / length(N)))
cat("  -> the hypoxia set is close to a subset of the normoxia set.\n")
```

```{r venn_helper}
# ggvenn draws list elements left to right, so the normoxia set is passed first to
# put it on the left. fill_color indexes set_colors BY NAME, so each hue stays
# bound to its set rather than to a position - reordering cannot mis-colour it.
make_venn <- function(venn_list, subtitle) {
  stopifnot("a venn set has no colour defined" =
    all(names(venn_list) %in% names(set_colors)))
  ggvenn(
    venn_list,
    fill_color      = unname(set_colors[names(venn_list)]),
    fill_alpha      = 0.45,
    stroke_size     = 0.5,
    stroke_color    = "grey30",
    set_name_size   = 4.5,
    text_size       = 4.5,
    show_percentage = FALSE,
    auto_scale      = FALSE
  ) +
    labs(title = "eIF3d RNA-promoted genes: normoxia vs hypoxia",
         subtitle = subtitle) +
    theme(plot.title    = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 9, colour = "grey30"))
}
```

```{r venn, fig.width=6.5, fig.height=5.5}
venn_list <- list(`Normoxia 1+4hr` = N, `Hypoxia 1hr` = H)

p_venn <- make_venn(
  venn_list,
  subtitle = paste0("sictrl vs si3d, rna_lfc > ", POS_LFC, " & padj < ", POS_PADJ,
                    ". Union = ", length(union(H, N)), " genes.\n",
                    "Contrasts differ in power (see below); set sizes are not ",
                    "directly comparable.")
)
p_venn
ggsave(here("plots", "72_rna_geneset_venn_hypoxia_vs_normoxia.pdf"),
       p_venn, width = 6.5, height = 5.5)
```

## Matched timepoint: normoxia 1hr vs hypoxia 1hr

The comparison above confounds condition with power: the normoxia 1+4hr fit uses
two timepoints and carries a larger global shift (0.505 vs 0.409), so a fixed
`rna_lfc > 1` bar admits more genes there. Repeating it against the **single
timepoint** normoxia 1hr contrast removes that confound - both sides are then one
timepoint, `deFunction()`, with near-identical global medians (0.368 vs 0.409).

```{r venn_matched, fig.width=6.5, fig.height=5.5}
N1 <- sets[["Normoxia 1hr"]]

shared_1hr        <- intersect(H, N1)
hypoxia_only_1hr  <- setdiff(H, N1)
normoxia_only_1hr <- setdiff(N1, H)

matched_tbl <- tibble(
  region = c("Shared", "Hypoxia 1hr only", "Normoxia 1hr only",
             "Hypoxia total", "Normoxia total", "Union"),
  n = c(length(shared_1hr), length(hypoxia_only_1hr), length(normoxia_only_1hr),
        length(H), length(N1), length(union(H, N1)))
)
print(matched_tbl)
stopifnot("matched region counts do not reconcile" =
  length(shared_1hr) + length(hypoxia_only_1hr) == length(H) &&
  length(shared_1hr) + length(normoxia_only_1hr) == length(N1))

p_venn_1hr <- make_venn(
  list(`Normoxia 1hr` = N1, `Hypoxia 1hr` = H),
  subtitle = paste0("Matched single timepoint. sictrl vs si3d, rna_lfc > ", POS_LFC,
                    " & padj < ", POS_PADJ, ". Union = ",
                    length(union(H, N1)), " genes.\n",
                    "Global medians are comparable here (normoxia ",
                    round(median(read_csv(here("output", CONTRAST_FILES[["Normoxia 1hr"]]),
                                          show_col_types = FALSE)$rna_lfc, na.rm = TRUE), 2),
                    ", hypoxia 0.41), so set sizes can be compared directly.")
)
p_venn_1hr
ggsave(here("plots", "72_rna_geneset_venn_normoxia1hr_vs_hypoxia1hr.pdf"),
       p_venn_1hr, width = 6.5, height = 5.5)

write_csv(matched_tbl, here("output", "72_rna_geneset_overlap_matched_1hr.csv"))
```

The direction of the asymmetry flips once the timepoint is matched, which is the
cleanest evidence that the first Venn was reporting a thresholding difference
rather than condition-specific biology.

# Correlation of RNA Changes: Normoxia vs Hypoxia

Gene-by-gene, transcriptome-wide: normoxia `rna_lfc` on x, hypoxia `rna_lfc` on y.
This is the continuous version of the Venn - it uses every gene rather than only
those clearing a threshold, so it is not sensitive to where the cut was placed.

Values are plotted **raw**, not centred, because the offset between the two
contrasts is itself the thing to see: a cloud sitting parallel to but off the
y = x line means the two conditions rank genes the same way while differing in
overall magnitude. Each contrast's global median is drawn, and y = x is shown for
reference.

```{r correlation_data}
de_all <- map(set_names(names(CONTRAST_FILES)), function(nm) {
  read_csv(here("output", CONTRAST_FILES[[nm]]), show_col_types = FALSE) %>%
    dplyr::select(gene_id, symbol, rna_lfc, rna_padj) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
})

# Drop EIF3D from the correlation: it is the siRNA target, sits at rna_lfc ~3 in
# every contrast, and as a single extreme point in the corner it would inflate
# Pearson noticeably while telling us nothing about eIF3d-dependent biology.
build_pair <- function(norm_name) {
  inner_join(
    de_all[[norm_name]] %>% dplyr::select(ensembl_gene, symbol, x = rna_lfc),
    de_all[["Hypoxia 1hr"]] %>% dplyr::select(ensembl_gene, y = rna_lfc),
    by = "ensembl_gene"
  ) %>%
    filter(!is.na(x), !is.na(y), symbol != "EIF3D")
}

pairs_list <- list(
  "Normoxia 1+4hr" = build_pair("Normoxia 1+4hr"),
  "Normoxia 1hr"   = build_pair("Normoxia 1hr")
)
walk2(names(pairs_list), pairs_list, ~ cat(.x, "vs Hypoxia 1hr:", nrow(.y), "genes\n"))
stopifnot("a correlation pair is empty" = all(map_int(pairs_list, nrow) > 8000))
```

```{r correlation_stats}
cor_stats <- map_dfr(names(pairs_list), function(nm) {
  d <- pairs_list[[nm]]
  tibble(
    comparison  = paste0(nm, " vs Hypoxia 1hr"),
    pearson_r   = cor(d$x, d$y, method = "pearson"),
    spearman_rho = cor(d$x, d$y, method = "spearman"),
    slope       = unname(coef(lm(y ~ x, data = d))[2]),
    median_x    = median(d$x),
    median_y    = median(d$y),
    n           = nrow(d)
  )
}) %>%
  mutate(across(c(pearson_r, spearman_rho, slope, median_x, median_y), ~ round(.x, 3)))
print(as.data.frame(cor_stats))
write_csv(cor_stats, here("output", "72_normoxia_vs_hypoxia_correlation.csv"))
```

```{r correlation_plots, fig.width=6, fig.height=5.5}
make_cor_plot <- function(nm) {
  d  <- pairs_list[[nm]]
  st <- cor_stats %>% filter(str_detect(comparison, fixed(nm)))
  lim <- quantile(c(d$x, d$y), c(0.005, 0.995), na.rm = TRUE)

  ggplot(d, aes(x = x, y = y)) +
    geom_abline(slope = 1, intercept = 0, linetype = "solid",
                colour = "grey75", linewidth = 0.5) +
    geom_vline(xintercept = st$median_x, linetype = "dashed",
               colour = "#009E73", linewidth = 0.4) +
    geom_hline(yintercept = st$median_y, linetype = "dashed",
               colour = "#D55E00", linewidth = 0.4) +
    geom_point(alpha = 0.15, size = 0.35, colour = "grey45") +
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
                colour = "#0072B2", fill = "#0072B2", alpha = 0.25,
                linewidth = 1) +
    annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.6, size = 3.2,
             colour = "grey20",
             label = paste0("Pearson R = ", st$pearson_r,
                            "\nSpearman rho = ", st$spearman_rho,
                            "\nslope = ", st$slope,
                            "\nn = ", st$n)) +
    coord_cartesian(xlim = lim, ylim = lim) +
    labs(
      title = paste0("RNA changes on eIF3d knockdown: ", nm, " vs hypoxia 1hr"),
      subtitle = paste0("Grey line = y = x. Dashed = each contrast's global median ",
                        "(x ", st$median_x, ", y ", st$median_y, ").\n",
                        "Blue = linear fit with 95% CI, computed on all genes. ",
                        "EIF3D excluded."),
      x = paste0(nm, " rna_lfc, log2(sictrl / si3d)"),
      y = "Hypoxia 1hr rna_lfc, log2(sictrl / si3d)"
    ) +
    theme_classic(base_size = 12)
}

p_cor_1and4 <- make_cor_plot("Normoxia 1+4hr")
p_cor_1and4
ggsave(here("plots", "72_cor_normoxia1and4hr_vs_hypoxia1hr.pdf"),
       p_cor_1and4, width = 6, height = 5.5)

p_cor_1hr <- make_cor_plot("Normoxia 1hr")
p_cor_1hr
ggsave(here("plots", "72_cor_normoxia1hr_vs_hypoxia1hr.pdf"),
       p_cor_1hr, width = 6, height = 5.5)
```

# Export Gene Lists

```{r export}
sym_map <- map_dfr(paths, ~ read_csv(.x, show_col_types = FALSE) %>%
                     dplyr::select(ensembl_gene, db_gene_symbol)) %>%
  distinct(ensembl_gene, .keep_all = TRUE)

region_tbl <- bind_rows(
  tibble(ensembl_gene = shared,        region = "shared"),
  tibble(ensembl_gene = hypoxia_only,  region = "hypoxia_only"),
  tibble(ensembl_gene = normoxia_only, region = "normoxia_only")
) %>%
  left_join(sym_map, by = "ensembl_gene") %>%
  dplyr::select(ensembl_gene, db_gene_symbol, region) %>%
  arrange(region, db_gene_symbol)

stopifnot("export lost or gained genes" =
  nrow(region_tbl) == length(union(H, N)))
stopifnot("a gene was assigned to more than one region" =
  !anyDuplicated(region_tbl$ensembl_gene))

print(region_tbl %>% dplyr::count(region))
write_csv(overlap_tbl, here("output", "72_rna_geneset_overlap.csv"))
write_csv(region_tbl,  here("output", "72_rna_geneset_overlap_genes.csv"))
cat("\nWrote output/72_rna_geneset_overlap.csv and _genes.csv\n")
```

# Is the Asymmetry Power or Biology?

602 genes are normoxia-only. Two explanations produce that count and the Venn
cannot distinguish them:

- **Power** - the gene behaves the same way in hypoxia but sits under the
  threshold, because the single-timepoint hypoxia fit is noisier.
- **Biology** - the gene is genuinely unaffected by si3d in hypoxia.

These make opposite predictions about where the non-shared genes sit in the *other*
contrast. A power-limited gene should still be shifted up, just short of the cut;
a condition-specific gene should sit at that contrast's global median.

```{r load_contrasts}
de <- map(set_names(seq_along(CONTRAST_FILES), names(CONTRAST_FILES)), function(i) {
  read_csv(here("output", CONTRAST_FILES[i]), show_col_types = FALSE) %>%
    dplyr::select(gene_id, symbol, rna_lfc, rna_padj) %>%
    mutate(ensembl_gene = sub("\\..*", "", gene_id))
})

medians <- map_dbl(de, ~ median(.x$rna_lfc, na.rm = TRUE))
print(round(medians, 3))

# Direction spot-check on both contrasts before drawing any conclusion from them
walk2(names(de), de, function(nm, d) {
  e <- d %>% filter(symbol == "EIF3D")
  stopifnot("EIF3D missing from a contrast" = nrow(e) == 1)
  stopifnot("EIF3D rna_lfc not positive - direction convention may have flipped" =
    e$rna_lfc > 0.5)
  cat(nm, ": EIF3D rna_lfc =", round(e$rna_lfc, 2), "-> log2(sictrl / si3d) confirmed\n")
})
```

```{r power_test}
# Where do the normoxia-only genes sit in the HYPOXIA contrast, and vice versa?
describe_in_other <- function(ids, other, other_name, other_median) {
  d <- other %>% filter(ensembl_gene %in% ids)
  tibble(
    group             = other_name,
    n                 = nrow(d),
    median_rna_lfc    = median(d$rna_lfc, na.rm = TRUE),
    contrast_median   = other_median,
    above_contrast_med = mean(d$rna_lfc > other_median, na.rm = TRUE),
    above_half        = mean(d$rna_lfc > 0.5, na.rm = TRUE),
    above_threshold   = mean(d$rna_lfc > POS_LFC, na.rm = TRUE),
    sig_and_above     = mean(d$rna_lfc > POS_LFC & d$rna_padj < POS_PADJ, na.rm = TRUE)
  )
}

power_tbl <- bind_rows(
  describe_in_other(normoxia_only, de[["Hypoxia 1hr"]],
                    "normoxia-only genes, scored in hypoxia", medians[["Hypoxia 1hr"]]),
  describe_in_other(hypoxia_only, de[["Normoxia 1+4hr"]],
                    "hypoxia-only genes, scored in normoxia", medians[["Normoxia 1+4hr"]]),
  describe_in_other(shared, de[["Hypoxia 1hr"]],
                    "shared genes, scored in hypoxia (reference)", medians[["Hypoxia 1hr"]]),
  # background: all genes, to show what "no effect" looks like in each contrast
  describe_in_other(de[["Hypoxia 1hr"]]$ensembl_gene, de[["Hypoxia 1hr"]],
                    "all genes, hypoxia (background)", medians[["Hypoxia 1hr"]])
) %>%
  mutate(across(c(median_rna_lfc, contrast_median), ~ round(.x, 3)),
         across(c(above_contrast_med, above_half, above_threshold, sig_and_above),
                ~ round(100 * .x, 1)))
print(as.data.frame(power_tbl))
```

`above_contrast_med` is the discriminating column. If the non-shared genes were
condition-specific, roughly 50% of them would sit above the other contrast's
median, matching background. A value well above 50% means they are shifted in the
same direction there too and simply missed the cut.

```{r power_verdict}
norm_only_row <- power_tbl %>% filter(str_detect(group, "^normoxia-only"))
bg_row        <- power_tbl %>% filter(str_detect(group, "background"))

cat("normoxia-only genes above the hypoxia median:",
    norm_only_row$above_contrast_med, "%\n")
cat("background genes above the hypoxia median   :",
    bg_row$above_contrast_med, "% (50% by construction)\n\n")

if (norm_only_row$above_contrast_med > 65) {
  cat("VERDICT: the normoxia-only genes are shifted in the same direction in\n",
      "hypoxia as well, so the asymmetry is mostly DETECTION POWER, not a\n",
      "hypoxia-specific loss of eIF3d dependence.\n")
} else if (norm_only_row$above_contrast_med < 55) {
  cat("VERDICT: the normoxia-only genes sit at the hypoxia background, so they\n",
      "look genuinely NORMOXIA-SPECIFIC rather than underpowered.\n")
} else {
  cat("VERDICT: intermediate - partly power, partly condition-specific.\n")
}
```

```{r plot_power, fig.width=8, fig.height=4.5}
plot_dat <- bind_rows(
  de[["Hypoxia 1hr"]] %>% filter(ensembl_gene %in% normoxia_only) %>%
    mutate(group = "Normoxia only"),
  de[["Hypoxia 1hr"]] %>% filter(ensembl_gene %in% shared) %>%
    mutate(group = "Shared"),
  de[["Hypoxia 1hr"]] %>% mutate(group = "All genes")
) %>%
  mutate(group = factor(group, levels = c("Shared", "Normoxia only", "All genes")))

p_power <- ggplot(plot_dat, aes(x = rna_lfc, colour = group)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = medians[["Hypoxia 1hr"]], linetype = "dashed",
             colour = "grey50", linewidth = 0.5) +
  geom_vline(xintercept = POS_LFC, linetype = "dotted",
             colour = "grey30", linewidth = 0.5) +
  scale_colour_manual(values = c("Shared" = "grey20",
                                 "Normoxia only" = "#009E73",
                                 "All genes" = "grey65"), name = NULL) +
  coord_cartesian(xlim = quantile(de[["Hypoxia 1hr"]]$rna_lfc,
                                  c(0.005, 0.995), na.rm = TRUE)) +
  labs(
    title = "Where the non-shared genes sit in the hypoxia contrast",
    subtitle = paste0("Dashed = hypoxia global median (",
                      round(medians[["Hypoxia 1hr"]], 2),
                      "); dotted = the lfc > ", POS_LFC, " threshold.\n",
                      "If 'Normoxia only' tracks 'All genes' it is condition-specific; ",
                      "if it sits between, it is underpowered."),
    x = "rna_lfc in hypoxia 1hr, log2(sictrl / si3d)",
    y = "Cumulative fraction of genes"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_power
ggsave(here("plots", "72_normoxia_only_genes_in_hypoxia_ecdf.pdf"),
       p_power, width = 8, height = 4.5)
```

# Are the Region-Specific Genes Mechanistically Different?

The notebook 07 RNA classifiers ranked `log2_utr5_length` first in both conditions,
with the GC block rising to ranks 2-6 in normoxia. If the hypoxia-only genes are a
mechanistically distinct class rather than a threshold artifact, they should differ
from the shared genes on those same features.

Read the effect sizes, not the p-values: with hundreds of genes per region a
sub-percentage-point GC difference reaches significance without meaning much.

```{r region_features}
features <- readRDS(here("output", "predictive_modeling",
                         "feature_matrix_external_stability.rds"))
stopifnot("feature matrix is not one row per gene" =
  !anyDuplicated(features$gene_id_clean))

FEATS <- c("log2_utr5_length" = "log2 5'UTR length",
           "log2_cds_length"  = "log2 CDS length",
           "utr5_gc"          = "5'UTR GC%",
           "cds_gc"           = "CDS GC%",
           "gc3"              = "GC3",
           "csc"              = "CSC")
stopifnot("expected feature columns missing" = all(names(FEATS) %in% colnames(features)))

region_feat <- region_tbl %>%
  inner_join(features %>% dplyr::select(ensembl_gene = gene_id_clean, all_of(names(FEATS))),
             by = "ensembl_gene")

# Background = everything in the matrix that is in neither set
bg_feat <- features %>%
  filter(!gene_id_clean %in% union(H, N)) %>%
  dplyr::select(all_of(names(FEATS))) %>%
  mutate(region = "background")

feat_all <- bind_rows(region_feat %>% dplyr::select(region, all_of(names(FEATS))), bg_feat) %>%
  mutate(region = factor(region, levels = c("shared", "hypoxia_only",
                                            "normoxia_only", "background")))
print(feat_all %>% dplyr::count(region))
stopifnot("a region has no genes in the feature matrix" =
  all(dplyr::count(feat_all, region)$n > 0))
```

```{r region_feature_table}
# Median per region, plus each region tested against the shared set (the reference
# class), since "is the region-specific fraction different from the core?" is the
# question. Wilcoxon on medians; BH across all region x feature comparisons.
med_tbl <- feat_all %>%
  pivot_longer(-region, names_to = "feature", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(feature, region) %>%
  summarise(median = median(value), n = dplyr::n(), .groups = "drop") %>%
  mutate(label = unname(FEATS[feature]))

wide_med <- med_tbl %>%
  dplyr::select(label, region, median) %>%
  pivot_wider(names_from = region, values_from = median) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
print(as.data.frame(wide_med))

vs_shared <- map_dfr(names(FEATS), function(f) {
  map_dfr(c("hypoxia_only", "normoxia_only", "background"), function(r) {
    a <- feat_all %>% filter(region == r)        %>% pull(all_of(f))
    b <- feat_all %>% filter(region == "shared") %>% pull(all_of(f))
    a <- a[!is.na(a)]; b <- b[!is.na(b)]
    tibble(feature = unname(FEATS[f]), region = r,
           median_region = median(a), median_shared = median(b),
           delta = median(a) - median(b),
           p = suppressWarnings(wilcox.test(a, b)$p.value),
           n_region = length(a))
  })
}) %>%
  mutate(p_adj = p.adjust(p, method = "BH"),
         across(c(median_region, median_shared, delta), ~ round(.x, 3)),
         across(c(p, p_adj), ~ signif(.x, 3))) %>%
  arrange(desc(abs(delta)))
print(as.data.frame(vs_shared), row.names = FALSE)
write_csv(vs_shared, here("output", "72_region_feature_comparison.csv"))
```

```{r plot_region_features, fig.width=10, fig.height=6}
# Same hue assignment as the Venn: vermillion = hypoxia, green = normoxia.
region_colors <- c("shared" = "grey30", "hypoxia_only" = "#D55E00",
                   "normoxia_only" = "#009E73", "background" = "grey70")

feat_long <- feat_all %>%
  pivot_longer(-region, names_to = "feature", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(label = factor(unname(FEATS[feature]), levels = unname(FEATS)))

facet_n <- feat_long %>%
  dplyr::count(label, region) %>%
  group_by(label) %>%
  summarise(txt = paste0("n = ", min(n), " to ", max(n), " per region"), .groups = "drop")

p_region <- ggplot(feat_long, aes(x = region, y = value, fill = region)) +
  geom_boxplot(outlier.size = 0.3, linewidth = 0.4, alpha = 0.75) +
  geom_text(data = facet_n, aes(x = Inf, y = Inf, label = txt),
            inherit.aes = FALSE, hjust = 1.05, vjust = 1.4,
            size = 2.8, colour = "grey30") +
  scale_fill_manual(values = region_colors, guide = "none") +
  facet_wrap(~ label, ncol = 3, scales = "free_y") +
  labs(
    title = "Sequence features by Venn region",
    subtitle = "Do the condition-specific genes differ from the shared core? Tested against 'shared' in the table above.",
    x = NULL, y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 35, hjust = 1, size = 8))
p_region
ggsave(here("plots", "72_region_sequence_features.pdf"),
       p_region, width = 10, height = 6)
```

# Summary

```{r summary}
cat("Hypoxia 1hr set     :", length(H), "genes\n")
cat("Normoxia 1+4hr set  :", length(N), "genes\n")
cat("Shared              :", length(shared),
    sprintf("(%.0f%% of hypoxia, %.0f%% of normoxia)\n",
            100 * length(shared) / length(H), 100 * length(shared) / length(N)))
cat("Hypoxia only        :", length(hypoxia_only), "\n")
cat("Normoxia only       :", length(normoxia_only), "\n")
cat("Union               :", length(union(H, N)), "\n\n")
cat("Outputs:\n")
cat("  output/72_rna_geneset_overlap.csv\n")
cat("  output/72_rna_geneset_overlap_genes.csv\n")
cat("  output/72_region_feature_comparison.csv\n")
cat("  plots/72_rna_geneset_venn_hypoxia_vs_normoxia.pdf\n")
cat("  plots/72_normoxia_only_genes_in_hypoxia_ecdf.pdf\n")
cat("  plots/72_region_sequence_features.pdf\n")
```
