---
title: "`r paste0('74 - Overlap of eIF3d-promoted genes at the TE vs RNA level (', params$condition, ' ', params$timepoint, ')')`"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  condition: "hypoxia"
  timepoint: "1hr"
  lfc_cut: 0.5
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Do the mRNAs eIF3d promotes **translationally** overlap the ones it promotes at the
level of **abundance**? Both sets come from the same contrast, si3d vs sictrl at
hypoxia 1hr:

- TE set: `te_lfc > 0.5 & te_padj < 0.05`
- RNA set: `rna_lfc > 0.5 & rna_padj < 0.05`

**Both sets use the same cutoff and are derived inline from the same file**, so
neither gets treatment the other does not. The TE derivation is asserted to
reproduce the canonical `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` exactly.

**Headline: the two sets overlap *less* than chance.** 352 shared genes against
~643 expected, Fisher odds ratio 0.37 (depletion p = 3e-61). They are strongly
anti-associated, not two views of one target list.

(Medians quoted in the caveats below are for hypoxia 1hr; the notebook recomputes
them for whichever contrast it is run on.)

**Direction convention.** Every column is log2(sictrl / si3d) - the producing
notebook runs DESeq2 with the knockdown as the reference, so the coefficient is
`Condition_sictrl_vs_si3d` regardless of the `si3d_vs_sictrl` filename. Positive =
knockdown reduces it = eIF3d promotes. Asserted below against EIF3D.

**Two caveats that decide how much the depletion means.** Both are tested rather
than assumed, in their own sections below:

1. **A matched nominal cutoff is not a matched stringency.** The two
   distributions sit at different medians (`te_lfc` 0.289, `rna_lfc` 0.409), so a
   shared 0.5 bar lands 0.211 above the TE median but only 0.091 above the RNA
   median - it is a far laxer cut on RNA, which is why the RNA set comes out at
   ~4,760 genes (40% of the transcriptome; EIF3D is excluded, so the printed
   count is one lower) against the TE set's 1,604. The overlap
   is therefore also reported centred on each median, and swept across a grid of
   both cutoffs.
2. **`te_lfc` and `rna_lfc` are not independent measurements.** They come from
   different DESeq2 fits, but TE is a ribosome-footprint-to-RNA quantity, so noise
   in RNA propagates into TE with the opposite sign. Some mutual depletion is
   expected even under null biology, and the size of that artifact needs bounding
   before the depletion is called biological.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggvenn)
```

# Parameters

```{r params}
# Defaults for interactive use; YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(condition = "hypoxia",  # "hypoxia" | "normoxia"
                 timepoint = "1hr",
                 lfc_cut   = 0.5)
}
CONDITION <- params$condition
TIMEPOINT <- params$timepoint

CONTRAST_FILE <- paste0("translation_categories_si3d_vs_sictrl_",
                        CONDITION, "_", TIMEPOINT, ".csv")
# Canonical TE gene set for this contrast, used only to validate the inline
# derivation below.
TE_SET_FILE <- paste0(CONDITION, "_3d_promotes_TE_", TIMEPOINT, "_lfc0.5.csv")

# Every output carries the contrast, so running a second condition cannot
# overwrite the first.
SUFFIX <- paste0(CONDITION, TIMEPOINT)

# One shared cutoff applied to both readouts, so the comparison is not confounded
# by threshold choice. See the stringency section for why a matched NOMINAL cut is
# still not a matched stringency here.
LFC_CUT <- params$lfc_cut
PADJ    <- 0.05


cat("Contrast :", CONTRAST_FILE, "\n")
cat("TE check :", TE_SET_FILE, "\n")
cat("Suffix   :", SUFFIX, "\n")

# Both sets are eIF3d-defined. The TE set keeps the established eIF3d amber;
# the RNA set takes Okabe-Ito vermillion, matching nb71. Neither uses the eIF3e
# sky blue (#56B4E9), which the project reserves for that factor.
set_colors <- c("TE promoted" = "#E69F00", "RNA promoted" = "#D55E00")
```

# Load Data

```{r load}
de <- read_csv(here("output", CONTRAST_FILE), show_col_types = FALSE) %>%
  dplyr::select(gene_id, symbol, rna_lfc, rna_padj, ribo_lfc, ribo_padj,
                te_lfc, te_padj) %>%
  mutate(ensembl_gene = sub("\\..*", "", gene_id))

# Universe = genes actually testable on BOTH readouts. Rows with NA padj can never
# enter either set, so counting them inflates U, deflates `expected`, and biases
# obs/exp upward - i.e. toward the depletion this notebook reports.
U <- de %>% filter(!is.na(te_padj), !is.na(rna_padj)) %>% pull(ensembl_gene) %>% unique()
cat("Universe restricted to jointly tested genes:", length(U),
    "of", nrow(de), "rows\n")

# EIF3D is the siRNA's own target; nb70 excludes it from the RNA set for that
# reason and the same logic applies here, so it is dropped from BOTH sets rather
# than from one. (It does not qualify for the TE set anyway - te_lfc is negative.)
eif3d_id <- de$ensembl_gene[de$symbol == "EIF3D"]

promoted <- function(lfc_col, padj_col, cut) {
  de %>%
    filter(.data[[lfc_col]] > cut, .data[[padj_col]] < PADJ) %>%
    pull(ensembl_gene) %>%
    setdiff(eif3d_id) %>%
    unique()
}

TE  <- promoted("te_lfc",  "te_padj",  LFC_CUT)
RNA <- promoted("rna_lfc", "rna_padj", LFC_CUT)

cat("Universe:", length(U), "genes\n")
cat("TE promoted  (te_lfc  >", LFC_CUT, "):", length(TE), "\n")
cat("RNA promoted (rna_lfc >", LFC_CUT, "):", length(RNA), "\n")

# The inline TE derivation must reproduce the canonical project gene set, which is
# what makes deriving both sets here rather than reading one from disk safe.
te_path <- here("output", "genesets", TE_SET_FILE)
if (isTRUE(all.equal(LFC_CUT, 0.5)) && file.exists(te_path)) {
  te_canonical <- read_csv(te_path, show_col_types = FALSE)$ensembl_gene %>%
    unique() %>% intersect(U)
  stopifnot("inline TE derivation does not reproduce the canonical TE gene set" =
    setequal(TE, setdiff(te_canonical, eif3d_id)))
  cat("Inline TE set reproduces", TE_SET_FILE, "exactly: PASS\n")
} else {
  cat("No canonical TE set to check against at lfc", LFC_CUT,
      "- inline derivation used as-is.\n")
}
```

## Validation

```{r checks}
stopifnot("duplicate genes in the contrast" = !anyDuplicated(de$ensembl_gene))
stopifnot("ids are not version-stripped" = all(grepl("^ENSG[0-9]+$", c(TE, RNA))))
stopifnot("a set is empty" = length(TE) > 0 && length(RNA) > 0)

# Direction: EIF3D is the knockdown target, so its own mRNA must be depleted
e <- de %>% filter(symbol == "EIF3D")
stopifnot("EIF3D not found" = nrow(e) == 1)
stopifnot("EIF3D rna_lfc not positive - direction may have flipped" = e$rna_lfc > 0.5)
cat("EIF3D rna_lfc =", round(e$rna_lfc, 2), "-> log2(sictrl / si3d) confirmed\n")

# EIF3D is dropped from both sets by construction above
stopifnot("EIF3D should be absent from both sets" =
  !e$ensembl_gene %in% c(TE, RNA))

# The sets must actually satisfy their stated definitions
te_chk  <- de %>% filter(ensembl_gene %in% TE)
rna_chk <- de %>% filter(ensembl_gene %in% RNA)
stopifnot("TE set violates its own threshold" =
  all(te_chk$te_lfc > LFC_CUT & te_chk$te_padj < PADJ, na.rm = TRUE))
stopifnot("RNA set violates its own threshold" =
  all(rna_chk$rna_lfc > LFC_CUT & rna_chk$rna_padj < PADJ, na.rm = TRUE))
cat("Both sets satisfy their stated definitions: PASS\n")
```

# Overlap and Enrichment

```{r overlap}
shared   <- intersect(TE, RNA)
te_only  <- setdiff(TE, RNA)
rna_only <- setdiff(RNA, TE)

# Expected overlap under independence, used in four places below.
overlap_stats <- function(a, b) {
  a <- intersect(a, U); b <- intersect(b, U)
  ov  <- length(intersect(a, b))
  exp <- length(a) * length(b) / length(U)
  list(observed = ov, expected = exp,
       obs_exp = if (exp > 0) ov / exp else NA_real_,
       n_a = length(a), n_b = length(b))
}

expected <- overlap_stats(TE, RNA)$expected

cont <- matrix(c(length(shared),
                 length(RNA) - length(shared),
                 length(TE)  - length(shared),
                 length(U) - length(TE) - length(RNA) + length(shared)),
               nrow = 2)
ft <- fisher.test(cont)

# Hypergeometric tail probabilities. DEPLETION is the alternative of interest,
# so the informative tail is P(X <= observed).
p_enrich  <- phyper(length(shared) - 1, length(TE), length(U) - length(TE),
                    length(RNA), lower.tail = FALSE)
p_deplete <- phyper(length(shared), length(TE), length(U) - length(TE),
                    length(RNA))

overlap_tbl <- tibble(
  metric = c("TE promoted", "RNA promoted", "Shared (observed)",
             "Shared (expected by chance)", "Observed / expected",
             "Fisher odds ratio", "P(overlap >= observed)", "P(overlap <= observed)",
             "TE only", "RNA only", "Universe"),
  value = c(length(TE), length(RNA), length(shared),
            round(expected, 1), round(length(shared) / expected, 3),
            round(unname(ft$estimate), 3), signif(p_enrich, 3), signif(p_deplete, 3),
            length(te_only), length(rna_only), length(U))
)
print(as.data.frame(overlap_tbl), row.names = FALSE)

stopifnot("region counts do not reconcile" =
  length(shared) + length(te_only)  == length(TE) &&
  length(shared) + length(rna_only) == length(RNA))

if (length(shared) < expected) {
  cat("\nThe sets overlap LESS than chance: obs/exp =",
      round(length(shared) / expected, 2), ", depletion p =", signif(p_deplete, 3), "\n")
}
```

```{r venn, fig.width=6.5, fig.height=5.5}
p_venn <- ggvenn(
  list(`TE promoted` = TE, `RNA promoted` = RNA),
  fill_color      = unname(set_colors[c("TE promoted", "RNA promoted")]),
  fill_alpha      = 0.45,
  stroke_size     = 0.5,
  stroke_color    = "grey30",
  set_name_size   = 4.5,
  text_size       = 4.5,
  show_percentage = FALSE,
  auto_scale      = FALSE
) +
  labs(title = paste0("eIF3d-promoted genes: TE level vs RNA level (", CONDITION, " ", TIMEPOINT, ")"),
       subtitle = paste0("sictrl vs si3d. TE: te_lfc > ", LFC_CUT,
                         "; RNA: rna_lfc > ", LFC_CUT, "; both padj < ", PADJ,
                         ".\nShared = ", length(shared), " vs ", round(expected, 0),
                         " expected by chance (Fisher OR ", round(unname(ft$estimate), 2), ").")) +
  theme(plot.title    = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9, colour = "grey30"))
p_venn
ggsave(here("plots", paste0("74_venn_te_vs_rna_promoted_", SUFFIX, ".pdf")),
       p_venn, width = 6.5, height = 5.5)
```

# Stringency: a Shared Cutoff Is Not a Shared Stringency

Both sets are now cut at the same number, which removes threshold choice as a
confounder. It does not make the two cuts equally strict, because the two
distributions are not centred in the same place.

```{r stringency}
med_te  <- median(de$te_lfc,  na.rm = TRUE)
med_rna <- median(de$rna_lfc, na.rm = TRUE)

# Cut each readout at its own median + the same offset, so both select from the
# same position in their own distribution.
offset <- LFC_CUT - med_te
rna_centered <- de %>%
  filter(rna_lfc - med_rna > offset, rna_padj < PADJ) %>%
  pull(ensembl_gene) %>% setdiff(eif3d_id) %>% unique()

stringency_tbl <- tibble(
  readout = c("TE", "RNA (nominal cut)", "RNA (centred to TE stringency)"),
  median  = round(c(med_te, med_rna, med_rna), 3),
  cut     = c(LFC_CUT, LFC_CUT, round(med_rna + offset, 3)),
  above_median = round(c(LFC_CUT - med_te, LFC_CUT - med_rna, offset), 3),
  n_genes = c(length(TE), length(RNA), length(rna_centered))
)
print(as.data.frame(stringency_tbl), row.names = FALSE)

ov_c  <- length(intersect(TE, rna_centered))
exp_c <- length(TE) * length(rna_centered) / length(U)
cat("\nOverlap with the quantile-matched RNA set:", ov_c, "vs", round(exp_c, 1),
    "expected -> obs/exp", round(ov_c / exp_c, 2), "\n")
# Reported, NOT asserted: whether the depletion holds is the finding this section
# exists to measure. Asserting it would abort the knit on any contrast where the
# answer is negative, so the analysis could never report one.
if (!is.na(ov_c / exp_c) && ov_c / exp_c < 1) {
  cat("-> depletion holds under stringency matching as well as nominal matching.\n")
} else {
  cat("-> NOTE: no depletion under stringency matching for this contrast.\n")
}
```

# Caveat 1: Does the Depletion Survive the Thresholds?

The two sets are cut at different stringencies, so the overlap is recomputed
across a grid of both cutoffs. If the depletion is a threshold artifact it should
disappear somewhere in the grid.

```{r threshold_grid}
grid <- expand_grid(te_cut = c(0.25, 0.5, 0.75, 1.0),
                    rna_cut = c(0.5, 0.75, 1.0, 1.5))

sens <- pmap_dfr(grid, function(te_cut, rna_cut) {
  st <- overlap_stats(promoted("te_lfc",  "te_padj",  te_cut),
                      promoted("rna_lfc", "rna_padj", rna_cut))
  tibble(te_cut = te_cut, rna_cut = rna_cut,
         n_te = st$n_a, n_rna = st$n_b, observed = st$observed,
         expected = round(st$expected, 1),
         obs_exp_raw = st$obs_exp, obs_exp = round(st$obs_exp, 2))
})
print(as.data.frame(sens), row.names = FALSE)

cat("\nobs/exp across the whole grid: min =", round(min(sens$obs_exp_raw, na.rm = TRUE), 3),
    " max =", round(max(sens$obs_exp_raw, na.rm = TRUE), 3), "\n")
# Compare the UNROUNDED ratio - a true 0.996 rounds to 1.00 and would read as a
# failure. Reported rather than asserted, for the same reason as above.
n_below <- sum(sens$obs_exp_raw < 1, na.rm = TRUE)
cat(n_below, "of", sum(!is.na(sens$obs_exp_raw)),
    "grid cells are below 1.\n")
if (n_below == sum(!is.na(sens$obs_exp_raw))) {
  cat("-> the depletion is not an artifact of where the cuts sit.\n")
}
```

```{r threshold_heat, fig.width=6.5, fig.height=4.5}
p_sens <- ggplot(sens, aes(x = factor(te_cut), y = factor(rna_cut), fill = obs_exp)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = paste0(obs_exp, "\n(", observed, ")")), size = 3) +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00",
                       midpoint = 1, limits = c(0, 2), name = "obs / exp") +
  labs(title = "Overlap relative to chance across thresholds",
       subtitle = "Value = observed/expected ratio; (n) = observed overlap. Below 1 = depleted.",
       x = "TE lfc cutoff", y = "RNA lfc cutoff") +
  theme_classic(base_size = 12)
p_sens
ggsave(here("plots", paste0("74_threshold_sensitivity_te_vs_rna_", SUFFIX, ".pdf")),
       p_sens, width = 6.5, height = 4.5)
```

# Caveat 2: Are te_lfc and rna_lfc Independent?

This is the load-bearing caveat. `te_lfc` is a ribosome-footprint-to-RNA quantity,
so sampling noise in RNA enters TE with the opposite sign; two sets defined on
"high RNA" and "high TE" would then avoid each other even with no biology at all.

The three columns are fit by **three separate DESeq2 models on different sample
sets** (`mode = "RNA"` on RNA libraries, `"RIBO"` on ribo libraries, `"TE"` on both
with an interaction term). Being separate fits does **not** make them independent:
`te_lfc` still tracks `ribo_lfc - rna_lfc` closely, which is quantified below and
is the reason the RIBO control that follows cannot bound the artifact.

```{r independence}
# NOTE: the three columns come from three DESeq2 fits on different sample sets,
# but that does NOT make them independent - te_lfc tracks ribo_lfc - rna_lfc very
# closely (quantified in the next chunk).
cor_te_rna  <- cor(de$te_lfc,  de$rna_lfc,  use = "complete.obs")
cor_ribo_rna <- cor(de$ribo_lfc, de$rna_lfc, use = "complete.obs")
cor_te_ribo <- cor(de$te_lfc,  de$ribo_lfc, use = "complete.obs")

cat("Transcriptome-wide correlations:\n")
cat("  te_lfc   vs rna_lfc :", round(cor_te_rna, 3), "\n")
cat("  ribo_lfc vs rna_lfc :", round(cor_ribo_rna, 3), "\n")
cat("  te_lfc   vs ribo_lfc:", round(cor_te_ribo, 3), "\n\n")

cat("Interpretation:\n")
cat("  ribo and rna are strongly POSITIVELY correlated (", round(cor_ribo_rna, 2),
    "): mRNA loss is largely forwarded to the ribosome.\n", sep = "")
cat("  te vs rna is NEGATIVE (", round(cor_te_rna, 2),
    "), which is the source of the mutual depletion.\n", sep = "")
```

```{r independence_plot, fig.width=6.5, fig.height=5.5}
plot_df <- de %>%
  mutate(group = case_when(
    ensembl_gene %in% shared   ~ "Shared RNA and TE",
    ensembl_gene %in% te_only  ~ "TE promoted only",
    ensembl_gene %in% rna_only ~ "RNA promoted only",
    TRUE                       ~ "All other genes"
  ) %>% factor(levels = c("All other genes", "TE promoted only",
                          "RNA promoted only", "Shared RNA and TE")))

p_scatter <- ggplot(plot_df %>% arrange(group), aes(x = rna_lfc, y = te_lfc)) +
  geom_hline(yintercept = LFC_CUT,  linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = LFC_CUT, linetype = "dashed", colour = "grey60", linewidth = 0.4) +
  geom_point(aes(colour = group, alpha = group), size = 0.4) +
  scale_colour_manual(values = c("All other genes" = "grey75",
                                 "TE promoted only" = "#E69F00",
                                 "RNA promoted only" = "#D55E00",
                                 "Shared RNA and TE" = "#0072B2"), name = NULL) +
  # One size for every group: the shared genes are the smallest class, so scaling
  # them up would exaggerate how much of the plot they occupy. Colour and the
  # draw order (set via arrange(group)) are enough to pick them out.
  scale_alpha_manual(values = c(0.2, 0.5, 0.5, 0.7), guide = "none") +
  coord_cartesian(xlim = quantile(de$rna_lfc, c(0.005, 0.995), na.rm = TRUE),
                  ylim = quantile(de$te_lfc,  c(0.005, 0.995), na.rm = TRUE)) +
  labs(title = paste0("TE change vs RNA change, ", CONDITION, " ", TIMEPOINT),
       subtitle = paste0("Dashed lines = the two set thresholds. Quadrant overlap is ",
                         "the shared set.\nTranscriptome-wide r(te, rna) = ",
                         round(cor_te_rna, 3), ", n = ", sum(complete.cases(de$te_lfc, de$rna_lfc))),
       x = "rna_lfc, log2(sictrl / si3d)",
       y = "te_lfc, log2(sictrl / si3d)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_scatter
ggsave(here("plots", paste0("74_te_vs_rna_lfc_scatter_", SUFFIX, ".pdf")),
       p_scatter, width = 6.5, height = 5.5)
```

## Bounding the artifact

An earlier version of this section used `ribo_lfc` as a comparator, reasoning that
a set built on it shares the RNA libraries' biology but is fit on the ribo
libraries. **That control is not valid** and the section is kept only to record
why: `te_lfc` is near-algebraically tied to `ribo_lfc - rna_lfc` (measured below),
so an artifact model predicts the very pattern the control was meant to rule out.
The numbers are reported, but they do not bound the artifact.

```{r artifact_bound}
ribo_set <- promoted("ribo_lfc", "ribo_padj", LFC_CUT)

compare_to_te <- function(b, label) {
  st <- overlap_stats(TE, b)
  tibble(comparison = label, n_set = st$n_b, observed = st$observed,
         expected = round(st$expected, 1), obs_exp = round(st$obs_exp, 2))
}
print(as.data.frame(bind_rows(
  compare_to_te(RNA,      "TE set vs RNA-promoted"),
  compare_to_te(ribo_set, "TE set vs RIBO-promoted (same lfc cut)")
)), row.names = FALSE)

# Measure how tightly TE is tied to ribo - rna before drawing any inference.
lk <- de %>% filter(!is.na(te_lfc), !is.na(ribo_lfc), !is.na(rna_lfc))
cat("\ncor(te_lfc, ribo_lfc - rna_lfc) =",
    round(cor(lk$te_lfc, lk$ribo_lfc - lk$rna_lfc), 3),
    " median |te - (ribo - rna)| =",
    round(median(abs(lk$te_lfc - (lk$ribo_lfc - lk$rna_lfc))), 3), "\n\n")

cat("INTERPRETATION - this control does NOT settle the question.\n")
cat("TE is near-algebraically tied to ribo - rna (r ~ 0.9 above), so RNA enters TE\n")
cat("with a negative sign and RIBO with a positive one. A pure-noise model therefore\n")
cat("PREDICTS depletion against the RNA set and enrichment against the RIBO set -\n")
cat("exactly the pattern observed. The RIBO comparison is consistent with the\n")
cat("artifact rather than evidence against it, and the caveat stays open.\n")
cat("Bounding it properly needs a TE estimate that does not share this RNA noise:\n")
cat("split-replicate or permuted TE, not ribo_lfc.\n")
```

# Sequence Features by Quadrant

The four groups from the scatter, profiled on the features the notebook 07
classifiers ranked highest. Unlike the gene-set comparison in nb72, **these four
groups are mutually exclusive by construction**, so an across-group Kruskal-Wallis
test is valid here.

Features are grouped by what they are and where they came up:

- **GC block** - `cds_gc` was the top feature of the TE classifier; `utr5_gc` and
  the GC block ranked 2-6 in the normoxia RNA classifier.
- **Length** - `log2_utr5_length` was the top feature of both RNA classifiers.
- **Codon / stability** - `csc`, `cai`, `zhu2024_ctrl_halflife` (iCodon weights and
  external SLAM-seq half-lives; sequence-derived, not from this experiment).
- **Other** - `hia2026_dhx29_occupancy` was 4th in the TE classifier.

```{r feature_setup}
features <- readRDS(here("output", "predictive_modeling",
                         "feature_matrix_external_stability.rds"))
stopifnot("feature matrix is not one row per gene" =
  !anyDuplicated(features$gene_id_clean))

FEATS <- c(
  # regional lengths
  "log2_utr5_length"              = "log2 5'UTR length",
  "log2_cds_length"               = "log2 CDS length",
  "log2_utr3_length"              = "log2 3'UTR length",
  "log2_total_length"             = "log2 total length",
  # regional GC
  "utr5_gc"                       = "5'UTR GC%",
  "cds_gc"                        = "CDS GC%",
  "utr3_gc"                       = "3'UTR GC%",
  "transcript_gc"                 = "Transcript GC%",
  # coding / occupancy
  "gc3"                           = "GC3",
  "hia2026_dhx29_occupancy"       = "DHX29 occupancy (Hia 2026)",
  "csc"                           = "CSC",
  "struct_num_structured_regions" = "N structured regions",
  # external half-life / stability
  "zhu2024_ctrl_halflife"         = "Half-life, ctrl (Zhu 2024)",
  "zhu2024_cnot3ko_halflife_logfc"= "Half-life logFC, CNOT3 KO (Zhu 2024)",
  "hia2026_ctrl_halflife"         = "Half-life, ctrl (Hia 2026)",
  "hia2026_dhx29_halflife_logfc"  = "Half-life logFC, DHX29 (Hia 2026)",
  "karner2026_mda231_log2_ct"     = "log2 Ct, MDA-231 (Karner 2026)"
)

# Which features go in which figure
GRP_LENGTH   <- c("log2_utr5_length", "log2_cds_length", "log2_utr3_length", "log2_total_length")
GRP_GC       <- c("utr5_gc", "cds_gc", "utr3_gc", "transcript_gc")
GRP_CODING   <- c("cds_gc", "gc3", "hia2026_dhx29_occupancy", "csc")
GRP_HALFLIFE <- c("zhu2024_ctrl_halflife", "zhu2024_cnot3ko_halflife_logfc",
                  "hia2026_ctrl_halflife", "hia2026_dhx29_halflife_logfc",
                  "karner2026_mda231_log2_ct")

stopifnot("expected feature columns missing" = all(names(FEATS) %in% colnames(features)))

# The four scatter groups plus a RIBO-exclusive group.
#
# NAMING, because "only" means something different in each case:
#   TE promoted only   = TE  \ RNA          (defined against RNA, may contain RIBO genes)
#   RNA promoted only  = RNA \ TE           (same)
#   Shared             = TE ∩ RNA
#   RIBO promoted only = RIBO \ (TE ∪ RNA)  (strictly exclusive of BOTH others)
# All five are mutually exclusive, so the Kruskal-Wallis below stays valid. But
# the TE and RNA groups are exclusive only with respect to each other, whereas the
# RIBO group is exclusive of everything - the TE set is 2x enriched for
# RIBO-promoted genes, so many "TE promoted only" transcripts are also RIBO
# promoted. Say "TE not RNA" rather than "TE alone" when describing them.
quad_colors <- c("All other genes"    = "grey70",
                 "TE promoted only"   = "#E69F00",
                 "RIBO promoted only" = "#CC79A7",   # Okabe-Ito reddish purple / pink
                 "RNA promoted only"  = "#D55E00",
                 "Shared RNA and TE"             = "#0072B2")
QUAD_LEVELS <- names(quad_colors)

RIBO_SET  <- promoted("ribo_lfc", "ribo_padj", LFC_CUT)
RIBO_ONLY <- setdiff(RIBO_SET, union(TE, RNA))
cat("RIBO promoted (ribo_lfc >", LFC_CUT, "):",
    length(promoted("ribo_lfc", "ribo_padj", LFC_CUT)),
    "of which exclusive of TE and RNA:", length(RIBO_ONLY), "\n")

quad <- plot_df %>%
  dplyr::select(ensembl_gene, group) %>%
  mutate(group = as.character(group),
         group = if_else(ensembl_gene %in% RIBO_ONLY, "RIBO promoted only", group),
         group = factor(group, levels = QUAD_LEVELS)) %>%
  inner_join(features %>%
               dplyr::select(ensembl_gene = gene_id_clean, all_of(names(FEATS))),
             by = "ensembl_gene")

stopifnot("group coercion produced NA" = !anyNA(quad$group))
# RIBO_ONLY is carved out of "All other genes" only - it must not have absorbed a
# gene that belonged to one of the TE/RNA groups.
stopifnot("RIBO-only overlaps the TE/RNA groups" =
  length(intersect(RIBO_ONLY, union(TE, RNA))) == 0)
print(quad %>% dplyr::count(group))
# .drop = FALSE: an empty factor level is otherwise ABSENT from count(), so
# `all(n >= 50)` passes vacuously for a group with zero genes.
grp_n <- dplyr::count(quad, group, .drop = FALSE)
stopifnot("a quadrant is missing entirely" = nrow(grp_n) == length(QUAD_LEVELS))
stopifnot("a quadrant has too few genes in the feature matrix" = all(grp_n$n >= 50))

# STATISTICAL REFERENCE, separate from the display groups.
#
# The five display groups are mutually exclusive, which is what makes the omnibus
# Kruskal-Wallis valid. But carving RIBO-only genes out of "All other genes" makes
# that group a biased background for the vs-background tests: it removes ~2,150
# translationally-up transcripts, which inflates every delta. Measured on hypoxia
# 1hr, cds_gc for "TE promoted only" reads -7.56 against the carved background but
# -4.78 against the true one, and log2_utr5_length goes from no shift (-0.009) to
# -0.117. The vs-background comparisons below therefore use the UNCARVED
# background: every gene in the feature matrix that is in neither the TE nor the
# RNA set. RIBO-only genes are part of it, as they are part of the transcriptome.
bg_ref <- features %>%
  filter(!gene_id_clean %in% union(TE, RNA)) %>%
  dplyr::select(ensembl_gene = gene_id_clean, all_of(names(FEATS)))
cat("Statistical background (not in TE or RNA sets):", nrow(bg_ref), "genes\n")
cat("  (the displayed 'All other genes' violin excludes RIBO-only:",
    sum(quad$group == "All other genes"), "genes)\n")
```

```{r feature_tests}
# TWO groupings are plotted, because they answer different questions:
#   quad5 - RIBO-only split out as its own class. The five groups partition the
#           universe, but "All other genes" then excludes ribo-promoted genes.
#   quad4 - the original four groups, with RIBO-only folded back into the
#           background so "All other genes" is the full untargeted transcriptome.
# The vs-background statistics below are unaffected: they already use bg_ref,
# which is the uncarved background in both cases.
quad4 <- quad %>%
  mutate(group = factor(if_else(as.character(group) == "RIBO promoted only",
                                "All other genes", as.character(group)),
                        levels = setdiff(QUAD_LEVELS, "RIBO promoted only")))
stopifnot("folding RIBO back produced NA groups" = !anyNA(quad4$group))

to_long <- function(q) {
  q %>%
    pivot_longer(-c(ensembl_gene, group), names_to = "feature", values_to = "value") %>%
    filter(!is.na(value)) %>%
    mutate(label = factor(unname(FEATS[feature]), levels = unname(FEATS)))
}
kw_of <- function(lg) {
  lg %>%
    group_by(label) %>%
    summarise(p_kw = kruskal.test(value ~ group)$p.value, .groups = "drop") %>%
    mutate(p_kw_adj = p.adjust(p_kw, method = "BH"),
           txt = if_else(p_kw_adj < 0.001, "KW p < 0.001",
                         paste0("KW p = ", signif(p_kw_adj, 2))))
}

#   quad6 - splits "TE promoted only" into the genes that move at the TE level and
#           NOWHERE else (truly exclusive) versus those that also move in ribosome
#           footprints. Still a partition, so Kruskal-Wallis stays valid.
#           The exclusive class is small (54 genes at hypoxia 1hr, 3.4% of the TE
#           set) - read its violins with that in mind, especially on the external
#           half-life panels where roughly half the values are NA.
TE_EXCLUSIVE <- setdiff(TE, union(RNA, RIBO_SET))
TE_AND_RIBO  <- setdiff(intersect(TE, RIBO_SET), RNA)

quad6_colors <- c("All other genes"    = "grey70",
                  "TE exclusive"       = "#E69F00",   # eIF3d amber
                  "TE and RIBO"        = "#B3760A",   # darker amber: same family on purpose
                  "RIBO promoted only" = "#CC79A7",
                  "RNA promoted only"  = "#D55E00",
                  "Shared RNA and TE"  = "#0072B2")
QUAD6_LEVELS <- names(quad6_colors)

quad6 <- quad %>%
  mutate(group = as.character(group),
         group = case_when(
           ensembl_gene %in% TE_EXCLUSIVE ~ "TE exclusive",
           ensembl_gene %in% TE_AND_RIBO  ~ "TE and RIBO",
           TRUE                           ~ group),
         group = factor(group, levels = QUAD6_LEVELS))
stopifnot("TE split produced NA groups" = !anyNA(quad6$group))
stopifnot("TE split is not a partition of the old TE-only group" =
  sum(quad6$group %in% c("TE exclusive", "TE and RIBO")) ==
  sum(quad$group == "TE promoted only"))

long  <- to_long(quad);  kw  <- kw_of(long)
long4 <- to_long(quad4); kw4 <- kw_of(long4)
long6 <- to_long(quad6); kw6 <- kw_of(long6)
cat("6-group sizes (TE split by exclusivity):\n")
print(dplyr::count(quad6, group, .drop = FALSE))
cat("5-group sizes:\n"); print(dplyr::count(quad,  group, .drop = FALSE))
cat("4-group sizes:\n"); print(dplyr::count(quad4, group, .drop = FALSE))

# Each labelled group against the "All other genes" background
bg_name <- "All other genes"
vs_bg <- map_dfr(setdiff(QUAD_LEVELS, bg_name), function(g) {
  map_dfr(names(FEATS), function(f) {
    a <- quad   %>% filter(group == g) %>% pull(all_of(f))
    b <- bg_ref %>% pull(all_of(f))            # uncarved reference, see above
    a <- a[!is.na(a)]; b <- b[!is.na(b)]
    # wilcox.test ERRORS (not warns) on an empty vector, so suppressWarnings does
    # not cover it. Sparse external columns can empty a small group: at lfc_cut 1
    # the Shared group drops to ~77 genes and hia2026_ctrl_halflife is ~50% NA.
    p <- if (length(a) < 3 || length(b) < 3) NA_real_
         else suppressWarnings(wilcox.test(a, b)$p.value)
    tibble(group = g, feature = unname(FEATS[f]),
           median_group = if (length(a)) median(a) else NA_real_,
           median_bg = median(b),
           delta = if (length(a)) median(a) - median(b) else NA_real_,
           p = p, n_group = length(a))
  })
}) %>%
  mutate(p_adj = p.adjust(p, method = "BH"),
         across(c(median_group, median_bg, delta), ~ round(.x, 3)),
         across(c(p, p_adj), ~ signif(.x, 3)))

cat("Largest shifts vs the background, by |delta|:\n")
print(as.data.frame(vs_bg %>% arrange(desc(abs(delta))) %>% head(15)), row.names = FALSE)
write_csv(vs_bg, here("output", paste0("74_quadrant_feature_comparison_", SUFFIX, ".csv")))
```

```{r violin_helper}
library(patchwork)

# Shared builder for the three faceted figures. (Per-feature y clipping is not
# available here - facet_wrap cannot take per-panel limits - so the half-life
# figure below is built as separate patchwork panels instead.)
make_quad_violin <- function(feats, title, ncol = 4,
                             dat = long, kw_tbl = kw, lvls = QUAD_LEVELS,
                             fills = quad_colors) {
  d <- dat %>% filter(feature %in% feats) %>%
    mutate(label = factor(unname(FEATS[feature]), levels = unname(FEATS[feats])))
  kw_f <- kw_tbl %>% filter(label %in% levels(d$label))
  # One n per group, comma separated in the same left-to-right order the violins
  # are drawn (QUAD_LEVELS). Counts can differ between panels because some
  # features carry NAs, so this is computed per facet rather than once.
  n_f  <- d %>%
    dplyr::count(label, group, .drop = FALSE) %>%
    complete(label, group = factor(lvls, levels = lvls),
             fill = list(n = 0)) %>%
    arrange(label, group) %>%
    group_by(label) %>%
    summarise(txt = paste0("n = ", paste(n, collapse = ", ")), .groups = "drop")

  ggplot(d, aes(x = group, y = value, fill = group)) +
    geom_violin(alpha = 0.5, linewidth = 0.4, trim = TRUE) +
    geom_boxplot(width = 0.16, outlier.size = 0.2, linewidth = 0.35,
                 fill = "white", alpha = 0.85) +
    geom_text(data = kw_f, aes(x = Inf, y = Inf, label = txt), inherit.aes = FALSE,
              hjust = 1.05, vjust = 1.4, size = 2.6, colour = "grey30") +
    geom_text(data = n_f, aes(x = -Inf, y = Inf, label = txt), inherit.aes = FALSE,
              hjust = -0.08, vjust = 1.4, size = 2.6, colour = "grey30") +
    scale_fill_manual(values = fills, guide = "none") +
    facet_wrap(~ label, ncol = ncol, scales = "free_y") +
    labs(title = title, x = NULL, y = NULL) +
    theme_classic(base_size = 11) +
    theme(strip.text = element_text(face = "bold", size = 9),
          axis.text.x = element_text(angle = 35, hjust = 1, size = 7))
}
```

```{r violin_lengths, fig.width=11, fig.height=4}
p_len <- make_quad_violin(GRP_LENGTH,
  paste0("Regional lengths by TE/RNA quadrant (", CONDITION, " ", TIMEPOINT, ")"))
p_len
ggsave(here("plots", paste0("74_quadrant_lengths_", SUFFIX, ".pdf")),
       p_len, width = 11, height = 4)

# Same panels without the RIBO-only class, background un-carved
p_len4 <- make_quad_violin(GRP_LENGTH,
  paste0("Regional lengths by TE/RNA quadrant, no RIBO class (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long4, kw_tbl = kw4, lvls = levels(quad4$group))
p_len4
ggsave(here("plots", paste0("74_quadrant_lengths_noribo_", SUFFIX, ".pdf")),
       p_len4, width = 11, height = 4)

# TE split into exclusive vs also-RIBO
p_len6 <- make_quad_violin(GRP_LENGTH,
  paste0("Regional lengths, TE split by exclusivity (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long6, kw_tbl = kw6, lvls = QUAD6_LEVELS, fills = quad6_colors)
p_len6
ggsave(here("plots", paste0("74_quadrant_lengths_teexcl_", SUFFIX, ".pdf")),
       p_len6, width = 12, height = 4)
```

```{r violin_gc, fig.width=11, fig.height=4}
p_gc <- make_quad_violin(GRP_GC,
  paste0("Regional GC content by TE/RNA quadrant (", CONDITION, " ", TIMEPOINT, ")"))
p_gc
ggsave(here("plots", paste0("74_quadrant_regional_gc_", SUFFIX, ".pdf")),
       p_gc, width = 11, height = 4)

# Same panels without the RIBO-only class, background un-carved
p_gc4 <- make_quad_violin(GRP_GC,
  paste0("Regional GC content by TE/RNA quadrant, no RIBO class (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long4, kw_tbl = kw4, lvls = levels(quad4$group))
p_gc4
ggsave(here("plots", paste0("74_quadrant_regional_gc_noribo_", SUFFIX, ".pdf")),
       p_gc4, width = 11, height = 4)

# TE split into exclusive vs also-RIBO
p_gc6 <- make_quad_violin(GRP_GC,
  paste0("Regional GC content, TE split by exclusivity (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long6, kw_tbl = kw6, lvls = QUAD6_LEVELS, fills = quad6_colors)
p_gc6
ggsave(here("plots", paste0("74_quadrant_regional_gc_teexcl_", SUFFIX, ".pdf")),
       p_gc6, width = 12, height = 4)
```

```{r violin_coding, fig.width=11, fig.height=4}
p_cod <- make_quad_violin(GRP_CODING,
  paste0("CDS GC, GC3, DHX29 occupancy and CSC (", CONDITION, " ", TIMEPOINT, ")"))
p_cod
ggsave(here("plots", paste0("74_quadrant_coding_features_", SUFFIX, ".pdf")),
       p_cod, width = 11, height = 4)

# Same panels without the RIBO-only class, background un-carved
p_cod4 <- make_quad_violin(GRP_CODING,
  paste0("CDS GC, GC3, DHX29 occupancy and CSC by TE/RNA quadrant, no RIBO class (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long4, kw_tbl = kw4, lvls = levels(quad4$group))
p_cod4
ggsave(here("plots", paste0("74_quadrant_coding_features_noribo_", SUFFIX, ".pdf")),
       p_cod4, width = 11, height = 4)

# TE split into exclusive vs also-RIBO
p_cod6 <- make_quad_violin(GRP_CODING,
  paste0("CDS GC, GC3, DHX29 occupancy and CSC, TE split by exclusivity (", CONDITION, " ", TIMEPOINT, ")"),
  dat = long6, kw_tbl = kw6, lvls = QUAD6_LEVELS, fills = quad6_colors)
p_cod6
ggsave(here("plots", paste0("74_quadrant_coding_features_teexcl_", SUFFIX, ".pdf")),
       p_cod6, width = 12, height = 4)
```

```{r violin_halflife, fig.width=13, fig.height=4.5}
# The half-life features span very different ranges, so each panel gets its own y
# window. coord_cartesian only zooms, so the violins are still computed on all
# genes - the tail is off screen, not dropped.
#
# Zhu control half-life needs a tighter window than the rest: it is strongly right
# skewed (median 4.6 h, 98th percentile 42.3 h, max 99.7 h), so a 2-98% window
# still squashes the bulk into the bottom fifth of the panel. Capping at the 90th
# percentile (15.6 h) keeps ~88% of genes in view with the median centred.
HL_Q_DEFAULT <- c(0.02, 0.98)
HL_Q <- list("zhu2024_ctrl_halflife" = c(0.02, 0.90))

# Unlike the other three figures, n is annotated PER VIOLIN here: the half-life
# features carry thousands of NAs (Hia ~5,800; Karner ~3,200) and the missingness
# is not even across the four groups, so a single range for the panel would hide
# which group is actually thin.

make_hl <- function(dat, lvls, fills = quad_colors) {
  map(GRP_HALFLIFE, function(f) {
    d   <- dat %>% filter(feature == f)
    qs  <- if (!is.null(HL_Q[[f]])) HL_Q[[f]] else HL_Q_DEFAULT
    lim <- quantile(d$value, qs, na.rm = TRUE)
    n_g <- d %>% dplyr::count(group, .drop = FALSE) %>% mutate(y = lim[2])
    ggplot(d, aes(x = group, y = value, fill = group)) +
      geom_violin(alpha = 0.5, linewidth = 0.4, trim = TRUE) +
      geom_boxplot(width = 0.16, outlier.size = 0.2, linewidth = 0.35,
                   fill = "white", alpha = 0.85) +
      geom_text(data = n_g, aes(x = group, y = y, label = n), inherit.aes = FALSE,
                vjust = 1.2, size = 2.5, colour = "grey25") +
      scale_fill_manual(values = fills, guide = "none") +
      coord_cartesian(ylim = lim) +
      labs(title = unname(FEATS[f]), x = NULL, y = NULL) +
      theme_classic(base_size = 10) +
      theme(plot.title = element_text(face = "bold", size = 9),
            axis.text.x = element_text(angle = 35, hjust = 1, size = 7))
  })
}
hl_plots  <- make_hl(long,  QUAD_LEVELS)
hl_plots4 <- make_hl(long4, levels(quad4$group))
hl_plots6 <- make_hl(long6, QUAD6_LEVELS, quad6_colors)

p_hl <- wrap_plots(hl_plots, nrow = 1) +
  plot_annotation(
    title = paste0("mRNA stability features by TE/RNA quadrant (",
                   CONDITION, " ", TIMEPOINT, ")"),
    caption = paste0("Number above each violin = genes with a non-missing value in ",
                     "that group. y axis trimmed per feature; ",
                     "violins computed on all genes."),
    theme = theme(plot.title = element_text(face = "bold", size = 12)))
p_hl
ggsave(here("plots", paste0("74_quadrant_halflife_", SUFFIX, ".pdf")),
       p_hl, width = 13, height = 4.5)

p_hl4 <- wrap_plots(hl_plots4, nrow = 1) +
  plot_annotation(
    title = paste0("mRNA stability features by TE/RNA quadrant, no RIBO class (",
                   CONDITION, " ", TIMEPOINT, ")"),
    caption = paste0("Number above each violin = genes with a non-missing value in ",
                     "that group. y axis trimmed per feature; ",
                     "violins computed on all genes."),
    theme = theme(plot.title = element_text(face = "bold", size = 12)))
p_hl4
ggsave(here("plots", paste0("74_quadrant_halflife_noribo_", SUFFIX, ".pdf")),
       p_hl4, width = 13, height = 4.5)

p_hl6 <- wrap_plots(hl_plots6, nrow = 1) +
  plot_annotation(
    title = paste0("mRNA stability features, TE split by exclusivity (",
                   CONDITION, " ", TIMEPOINT, ")"),
    caption = paste0("Number above each violin = genes with a non-missing value. ",
                     "The TE exclusive class is small - check its n before reading it."),
    theme = theme(plot.title = element_text(face = "bold", size = 12)))
p_hl6
ggsave(here("plots", paste0("74_quadrant_halflife_teexcl_", SUFFIX, ".pdf")),
       p_hl6, width = 14, height = 4.5)
```

# Export

```{r export}
region_tbl <- bind_rows(
  tibble(ensembl_gene = shared,   region = "shared"),
  tibble(ensembl_gene = te_only,  region = "te_only"),
  tibble(ensembl_gene = rna_only, region = "rna_only")
) %>%
  left_join(de %>% dplyr::select(ensembl_gene, symbol, rna_lfc, rna_padj,
                                 ribo_lfc, te_lfc, te_padj), by = "ensembl_gene") %>%
  arrange(region, desc(te_lfc))

stopifnot("export lost or gained genes" =
  nrow(region_tbl) == length(union(TE, RNA)))
stopifnot("a gene landed in two regions" = !anyDuplicated(region_tbl$ensembl_gene))
print(region_tbl %>% dplyr::count(region))

write_csv(overlap_tbl, here("output", paste0("74_te_vs_rna_overlap_summary_", SUFFIX, ".csv")))
write_csv(region_tbl,  here("output", paste0("74_te_vs_rna_overlap_genes_", SUFFIX, ".csv")))
write_csv(sens,        here("output", paste0("74_te_vs_rna_threshold_sensitivity_", SUFFIX, ".csv")))

cat("\nThe", length(shared), "shared genes (highest te_lfc first):\n")
print(region_tbl %>% filter(region == "shared") %>%
        dplyr::select(symbol, rna_lfc, ribo_lfc, te_lfc) %>%
        mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
        head(15))
```

# Summary

```{r summary}
cat("TE promoted (te_lfc >", LFC_CUT, "):", length(TE), "genes\n")
cat("RNA promoted (rna_lfc >", LFC_CUT, "):", length(RNA), "genes\n")
cat("Shared:", length(shared), "vs", round(expected, 1), "expected -> obs/exp",
    round(length(shared) / expected, 2), "\n")
cat("Fisher OR:", round(unname(ft$estimate), 3), " depletion p:", signif(p_deplete, 3), "\n\n")
# Built from SUFFIX so the printed paths are the ones actually written.
cat("Outputs (suffix =", SUFFIX, "):\n")
for (f in c("output/74_te_vs_rna_overlap_summary_%s.csv",
            "output/74_te_vs_rna_overlap_genes_%s.csv",
            "output/74_te_vs_rna_threshold_sensitivity_%s.csv",
            "output/74_quadrant_feature_comparison_%s.csv",
            "plots/74_venn_te_vs_rna_promoted_%s.pdf",
            "plots/74_threshold_sensitivity_te_vs_rna_%s.pdf",
            "plots/74_te_vs_rna_lfc_scatter_%s.pdf",
            "plots/74_quadrant_lengths_%s.pdf",
            "plots/74_quadrant_regional_gc_%s.pdf",
            "plots/74_quadrant_coding_features_%s.pdf",
            "plots/74_quadrant_halflife_%s.pdf")) {
  cat("  ", sprintf(f, SUFFIX), "\n", sep = "")
}
```
