---
title: "75 - SHAP direction, RNA vs TE classifiers"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Per-feature **directional** SHAP summaries from the notebook 07 classifiers,
compared between the RNA-level and TE-level models. One point per feature:
`shap_value_cor` in the RNA model on x, the same feature in the TE model on y.
One figure per oxygen condition.

**Why `shap_value_cor` and not the other columns.** It is the correlation between
a feature's *value* and its own SHAP contribution, so it carries sign: positive
means high values of the feature push a gene toward the promoted class, negative
means they push away. `mean_abs_shap` is magnitude only (always >= 0, so no sign
to compare), and `mean_signed_shap` is explicitly **not** a direction measure -
it averages to ~0 for every feature and its residual offset reflects the gap
between the explain-set mean prediction and the training baseline.

**How to read the quadrants.** Both axes run -1 to +1:

- **upper right / lower left** - the feature pushes the same way in both readouts
- **upper left / lower right** - the feature's direction *flips* between them
- **near an axis** - the feature is directional in one readout and inert in the other

**The two models are not cut at the same threshold.** The TE models use the
project-standard `lfc > 0.5`; the RNA models require `lfc > 1` because the raw RNA
contrast is spike-in shifted (median `rna_lfc` ~0.37-0.41), so a 0.5 cut selects
~40% of the transcriptome and the classes cannot be balanced. Each panel's
subtitle states the exact parameters behind both models rather than leaving this
implicit.

**The two models are also measured against different backgrounds, and that
limits what a flip means.** SHAP direction is defined relative to *each model's
own negative set*, and the two cannot share one: the TE model uses nb02's
`negative_control_genes.csv` (3,319 genes, TE-defined) while the RNA model uses
nb70's `si3d_rna_negative_controls_*` (1,051 genes). They overlap by only 312
genes and differ substantially in composition:

| feature | TE negative set | RNA negative set |
|---|---|---|
| GC3 | 61.8 | 50.7 |
| CDS GC% | 53.8 | 48.8 |
| log2 5'UTR length | 6.81 | 7.55 |

So a sign flip reads as "this feature separates toward promoted in one contrast
and away in the other, *each against its own background*" - part of the contrast
comes from the backgrounds differing, not only from the transcripts. The two sets
cannot be swapped or merged: a TE-defined background is invalid for an RNA model
(307 of its 3,319 genes satisfy the RNA positive definition outright, which is why
notebook 07 now refuses that pairing), so this is inherent to the comparison
rather than a fixable choice.

The dissociation itself does not rest on the gene sets. Notebook 71 finds the same
split transcriptome-wide with no sets involved at all: `cds_gc` correlates
-0.367 with `te_lfc` but +0.031 with `rna_lfc`, and `log2_utr5_length` is the
mirror image (+0.076 vs -0.248). Read the quadrants as corroborating that, not as
a background-independent statement on their own.

**Provenance.** SHAP comes from `07c_rf_shap_eif3d_targets.Rmd`, which fits a
surrogate regression forest on the same training rows (treeshap cannot explain
ranger probability forests), reads `mtry`/`min.node.size`/`num.trees` off the
saved 07 model, and gates on the surrogate reproducing 07's held-out predictions
within 0.02 before explaining it. Nothing is refit here.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
library(cowplot)
```

# Parameters

```{r params}
# Each entry records the FULL parameter set behind that model, so the plot
# subtitles are generated from the same object the files are located with -
# they cannot drift apart.
RUNS <- tribble(
  ~condition,  ~readout, ~lfc,  ~neg_set,
  "hypoxia",   "TE",     "0.5", "negative_control_genes.csv",
  "hypoxia",   "RNA",    "1",   "si3d_rna_negative_controls_nor1and4_hyp1_hyp4.csv",
  "normoxia",  "TE",     "0.5", "negative_control_genes.csv",
  "normoxia",  "RNA",    "1",   "si3d_rna_negative_controls_nor1and4_hyp1_hyp4.csv"
) %>%
  mutate(
    suffix = paste0("eif33d_promotes_",
                    if_else(readout == "TE", "", "rna_"),
                    condition, "_1hr_lfc", lfc,
                    "_external_stability_clipaggregate_no_boruta_noOwnTE_shap"),
    path = here("output", "predictive_modeling",
                paste0("shap_importance_", suffix, ".csv"))
  )
print(RUNS %>% dplyr::select(condition, readout, lfc, suffix))

TIMEPOINT   <- "1hr"
FEATURE_SET <- "external_stability"
TOP_N       <- 30    # features shown in each lollipop
LABEL_N     <- 4     # features labelled per QUADRANT (not per panel), so all
                     # four quadrants get called out rather than only the two
                     # that flip
```

# Load SHAP Summaries

```{r load}
stopifnot("a SHAP importance file is missing - run 07c for that model first" =
  all(file.exists(RUNS$path)))

shap <- RUNS %>%
  mutate(data = map(path, ~ read_csv(.x, show_col_types = FALSE) %>%
                      dplyr::select(feature, feature_clean, category,
                                    mean_abs_shap, shap_value_cor))) %>%
  dplyr::select(condition, readout, data) %>%
  unnest(data)

print(shap %>% dplyr::count(condition, readout))
```

## Validation

```{r checks}
# shap_value_cor is a correlation: it must lie in [-1, 1]
stopifnot("shap_value_cor outside [-1, 1]" =
  all(shap$shap_value_cor >= -1 & shap$shap_value_cor <= 1, na.rm = TRUE))
# mean_abs_shap is a magnitude
stopifnot("mean_abs_shap is negative" = all(shap$mean_abs_shap >= 0, na.rm = TRUE))

# All four models must use the same feature set, or the join below silently drops
# features and the scatter is computed on an unstated subset.
feats <- shap %>% group_split(condition, readout) %>% map(~ sort(.x$feature))
stopifnot("the four models do not share one feature set" =
  all(vapply(feats[-1], identical, logical(1), feats[[1]])))
cat("All four models share", length(feats[[1]]), "features\n")

# One row per feature per model
stopifnot("duplicate features within a model" =
  !any(duplicated(shap %>% dplyr::select(condition, readout, feature))))
```

# RNA vs TE Direction

```{r pair}
paired <- shap %>%
  dplyr::select(condition, readout, feature, feature_clean, category,
                shap_value_cor, mean_abs_shap) %>%
  pivot_wider(names_from = readout,
              values_from = c(shap_value_cor, mean_abs_shap)) %>%
  dplyr::rename(rna = shap_value_cor_RNA, te = shap_value_cor_TE) %>%
  filter(!is.na(rna), !is.na(te)) %>%
  # Distance from the y = x diagonal: how differently a feature behaves between
  # the two readouts. Used to pick which points get labelled.
  mutate(divergence = abs(te - rna),
         # Bubble size / "major" ranking uses the MEAN of the two models'
         # magnitudes. Ranking on the TE model alone would call a feature minor
         # when it is a strong RNA contributor and weak in TE, which is exactly
         # the asymmetry this plot exists to show.
         mean_abs_shap = (mean_abs_shap_RNA + mean_abs_shap_TE) / 2,
         quadrant = case_when(
           rna >= 0 & te >= 0 ~ "both positive",
           rna <  0 & te >= 0 ~ "flips: RNA down, TE up",
           rna <  0 & te <  0 ~ "both negative",
           TRUE               ~ "flips: RNA up, TE down"
         ))

stats <- paired %>%
  group_by(condition) %>%
  summarise(
    pearson  = cor(rna, te, method = "pearson"),
    spearman = cor(rna, te, method = "spearman"),
    n        = dplyr::n(),
    .groups  = "drop"
  ) %>%
  mutate(across(c(pearson, spearman), ~ round(.x, 3)))
print(as.data.frame(stats), row.names = FALSE)
write_csv(paired, here("output", "75_shap_direction_rna_vs_te.csv"))
```

```{r plot_fn}
# Subtitle is built from RUNS, so the parameters shown are the ones the files
# were located with.
make_shap_plot <- function(cond) {
  d  <- paired %>% filter(condition == cond)
  st <- stats  %>% filter(condition == cond)
  r  <- RUNS   %>% filter(condition == cond)
  te_r  <- r %>% filter(readout == "TE")
  rna_r <- r %>% filter(readout == "RNA")

  # Label the largest bubbles in EACH quadrant, so every quadrant is annotated
  # rather than only the high-divergence corners.
  lab <- d %>%
    group_by(quadrant) %>%
    slice_max(mean_abs_shap, n = LABEL_N, with_ties = FALSE) %>%
    ungroup()

  sub <- paste0(
    "TE model:  si3d ", cond, " ", TIMEPOINT, ", te_lfc > ", te_r$lfc,
    " & padj < 0.05  |  neg: ", te_r$neg_set, "\n",
    "RNA model: si3d ", cond, " ", TIMEPOINT, ", rna_lfc > ", rna_r$lfc,
    " & padj < 0.05  |  neg: ", rna_r$neg_set, "\n",
    "Both: feature_set = ", FEATURE_SET,
    ", clip = aggregate, exclude_own_te_features = TRUE, seed 9.  ",
    "Pearson R = ", st$pearson, ", Spearman rho = ", st$spearman,
    ", n = ", st$n, " features."
  )

  ggplot(d, aes(x = rna, y = te)) +
    geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.4) +
    geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey70", linewidth = 0.4) +
    geom_point(aes(size = mean_abs_shap, colour = category), alpha = 0.85) +
    geom_text_repel(data = lab, aes(label = feature_clean),
                    size = 2.8, min.segment.length = 0, max.overlaps = Inf,
                    segment.colour = "grey60", segment.size = 0.3) +
    scale_size_continuous(range = c(1, 6), name = "mean |SHAP|\n(RNA & TE avg)") +
    scale_colour_brewer(palette = "Dark2", name = NULL) +
    coord_equal(xlim = c(-1, 1), ylim = c(-1, 1)) +
    labs(
      title = paste0("SHAP direction, RNA vs TE classifier: ", cond, " ", TIMEPOINT),
      subtitle = sub,
      x = "shap_value_cor, RNA model",
      y = "shap_value_cor, TE model"
    ) +
    theme_classic(base_size = 11) +
    theme(plot.subtitle = element_text(size = 7, colour = "grey30"),
          legend.position = "right",
          legend.text = element_text(size = 7),
          legend.title = element_text(size = 7))
}
```

```{r plot_hypoxia, fig.width=9, fig.height=6.5}
p_hyp <- make_shap_plot("hypoxia")
p_hyp
ggsave(here("plots", "75_shap_direction_rna_vs_te_hypoxia1hr.pdf"),
       p_hyp, width = 9, height = 6.5)
```

```{r plot_normoxia, fig.width=9, fig.height=6.5}
p_nor <- make_shap_plot("normoxia")
p_nor
ggsave(here("plots", "75_shap_direction_rna_vs_te_normoxia1hr.pdf"),
       p_nor, width = 9, height = 6.5)
```

# Per-Model SHAP Lollipops

One plot per model - TE and RNA, hypoxia and normoxia 1hr. Bar length is `mean_abs_shap` (how much the
feature contributes) and its sign is taken from `shap_value_cor` (which way it
pushes), so a single axis carries both - the same construction notebook 07 uses
for `signed_importance`.

Direction is carried by the **sign of the axis**, matching the signed-importance
lollipop in notebook 07: bars extending right mean high feature values favour the
promoted class, bars extending left mean they argue against it. Colour is the
feature category.

Note the axis carries only the *sign* of `shap_value_cor`, not its magnitude, so a
weakly directional feature extends as far as a strongly directional one of the
same mean |SHAP|. In the RNA top 30 that applies to `log2 total length`
(shap_value_cor +0.01) and `Accessibility: cds mean` (-0.07), whose sign is close
to arbitrary - check the exported CSV before reading their direction.

```{r lollipop_fn}
# Signed SHAP importance, matching the directionality lollipop in notebook 07:
# direction goes ON THE AXIS as the sign, rather than being encoded as fill, so
# features that argue against the promoted class extend left of zero. Length is
# still mean |SHAP|, so the magnitude ranking is preserved.
make_shap_lollipop <- function(cond, rdt) {
  r <- RUNS %>% filter(condition == cond, readout == rdt)
  stopifnot("no run matches that condition/readout" = nrow(r) == 1)

  d <- shap %>%
    filter(condition == cond, readout == rdt, !is.na(shap_value_cor)) %>%
    slice_max(mean_abs_shap, n = TOP_N, with_ties = FALSE) %>%
    mutate(signed_shap  = mean_abs_shap * sign(shap_value_cor),
           feature_clean = fct_reorder(feature_clean, signed_shap))

  ggplot(d, aes(x = feature_clean, y = signed_shap)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_segment(aes(xend = feature_clean, y = 0, yend = signed_shap),
                 colour = "grey65", linewidth = 0.6) +
    geom_point(aes(colour = category), size = 3.4) +
    scale_colour_brewer(palette = "Dark2", name = NULL) +
    coord_flip() +
    theme_cowplot(font_size = 11) +
    labs(
      title = paste0("Signed SHAP importance: ", rdt, " model, ", cond, " ", TIMEPOINT),
      subtitle = paste0("si3d ", cond, " ", TIMEPOINT, ", ",
                        tolower(rdt), "_lfc > ", r$lfc, " & padj < 0.05  |  neg: ",
                        r$neg_set, "\n",
                        "feature_set = ", FEATURE_SET,
                        ", clip = aggregate, exclude_own_te_features = TRUE, seed 9.  ",
                        "Top ", nrow(d), " by mean |SHAP|."),
      x = NULL,
      y = "Signed mean |SHAP|  (positive = high values favour promoted)"
    ) +
    theme(plot.subtitle = element_text(size = 7, colour = "grey30"),
          legend.text = element_text(size = 7))
}
```

```{r lollipop_te_hypoxia, fig.width=9, fig.height=9}
p_lolli_te <- make_shap_lollipop("hypoxia", "TE")
p_lolli_te
ggsave(here("plots", "75_shap_lollipop_te_hypoxia1hr.pdf"),
       p_lolli_te, width = 9, height = 9)
```

```{r lollipop_rna_hypoxia, fig.width=9, fig.height=9}
p_lolli_rna <- make_shap_lollipop("hypoxia", "RNA")
p_lolli_rna
ggsave(here("plots", "75_shap_lollipop_rna_hypoxia1hr.pdf"),
       p_lolli_rna, width = 9, height = 9)
```

```{r lollipop_te_normoxia, fig.width=9, fig.height=9}
p_lolli_te_nor <- make_shap_lollipop("normoxia", "TE")
p_lolli_te_nor
ggsave(here("plots", "75_shap_lollipop_te_normoxia1hr.pdf"),
       p_lolli_te_nor, width = 9, height = 9)
```

```{r lollipop_rna_normoxia, fig.width=9, fig.height=9}
p_lolli_rna_nor <- make_shap_lollipop("normoxia", "RNA")
p_lolli_rna_nor
ggsave(here("plots", "75_shap_lollipop_rna_normoxia1hr.pdf"),
       p_lolli_rna_nor, width = 9, height = 9)
```

# Features That Flip Direction

```{r flips}
# A flip = opposite signs in the two readouts, with both contributions
# appreciable rather than one being noise around zero.
flips <- paired %>%
  filter(sign(rna) != sign(te), abs(rna) > 0.2, abs(te) > 0.2) %>%
  arrange(condition, desc(divergence)) %>%
  dplyr::select(condition, feature_clean, category, rna, te, divergence) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
print(as.data.frame(flips), row.names = FALSE)

cat("\nLargest RNA-vs-TE divergence per condition:\n")
print(as.data.frame(
  paired %>% group_by(condition) %>% slice_max(divergence, n = 5) %>%
    dplyr::select(condition, feature_clean, rna, te, divergence) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))), row.names = FALSE)
```

# Summary

```{r summary}
print(as.data.frame(stats), row.names = FALSE)
cat("\nOutputs:\n")
cat("  output/75_shap_direction_rna_vs_te.csv\n")
cat("  plots/75_shap_direction_rna_vs_te_hypoxia1hr.pdf\n")
cat("  plots/75_shap_direction_rna_vs_te_normoxia1hr.pdf\n")
cat("  plots/75_shap_lollipop_te_hypoxia1hr.pdf\n")
cat("  plots/75_shap_lollipop_rna_hypoxia1hr.pdf\n")
cat("  plots/75_shap_lollipop_te_normoxia1hr.pdf\n")
cat("  plots/75_shap_lollipop_rna_normoxia1hr.pdf\n")
```
