---
title: "75 - SHAP direction, RNA vs TE classifiers"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

Per-feature **directional** SHAP summaries from the notebook 07 classifiers,
compared between the RNA-level and TE-level models. One point per feature:
`shap_value_cor` in the RNA model on x, the same feature in the TE model on y.
One figure per oxygen condition.

**Why `shap_value_cor` and not the other columns.** It is the correlation between
a feature's *value* and its own SHAP contribution, so it carries sign: positive
means high values of the feature push a gene toward the promoted class, negative
means they push away. `mean_abs_shap` is magnitude only (always >= 0, so no sign
to compare), and `mean_signed_shap` is explicitly **not** a direction measure -
it averages to ~0 for every feature and its residual offset reflects the gap
between the explain-set mean prediction and the training baseline.

**How to read the quadrants.** Both axes run -1 to +1:

- **upper right / lower left** - the feature pushes the same way in both readouts
- **upper left / lower right** - the feature's direction *flips* between them
- **near an axis** - the feature is directional in one readout and inert in the other

**The two models are not cut at the same threshold.** The TE models use the
project-standard `lfc > 0.5`; the RNA models require `lfc > 1` because the raw RNA
contrast is spike-in shifted (median `rna_lfc` ~0.37-0.41), so a 0.5 cut selects
~40% of the transcriptome and the classes cannot be balanced. Each panel's
subtitle states the exact parameters behind both models rather than leaving this
implicit.

**Provenance.** SHAP comes from `07c_rf_shap_eif3d_targets.Rmd`, which fits a
surrogate regression forest on the same training rows (treeshap cannot explain
ranger probability forests), reads `mtry`/`min.node.size`/`num.trees` off the
saved 07 model, and gates on the surrogate reproducing 07's held-out predictions
within 0.02 before explaining it. Nothing is refit here.

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(ggrepel)
```

# Parameters

```{r params}
# Each entry records the FULL parameter set behind that model, so the plot
# subtitles are generated from the same object the files are located with -
# they cannot drift apart.
RUNS <- tribble(
  ~condition,  ~readout, ~lfc,  ~neg_set,
  "hypoxia",   "TE",     "0.5", "negative_control_genes.csv",
  "hypoxia",   "RNA",    "1",   "si3d_rna_negative_controls_nor1and4_hyp1_hyp4.csv",
  "normoxia",  "TE",     "0.5", "negative_control_genes.csv",
  "normoxia",  "RNA",    "1",   "si3d_rna_negative_controls_nor1and4_hyp1_hyp4.csv"
) %>%
  mutate(
    suffix = paste0("eif33d_promotes_",
                    if_else(readout == "TE", "", "rna_"),
                    condition, "_1hr_lfc", lfc,
                    "_external_stability_clipaggregate_no_boruta_noOwnTE_shap"),
    path = here("output", "predictive_modeling",
                paste0("shap_importance_", suffix, ".csv"))
  )
print(RUNS %>% dplyr::select(condition, readout, lfc, suffix))

TIMEPOINT   <- "1hr"
FEATURE_SET <- "external_stability"
LABEL_N     <- 4     # features labelled per QUADRANT (not per panel), so all
                     # four quadrants get called out rather than only the two
                     # that flip
```

# Load SHAP Summaries

```{r load}
stopifnot("a SHAP importance file is missing - run 07c for that model first" =
  all(file.exists(RUNS$path)))

shap <- RUNS %>%
  mutate(data = map(path, ~ read_csv(.x, show_col_types = FALSE) %>%
                      dplyr::select(feature, feature_clean, category,
                                    mean_abs_shap, shap_value_cor))) %>%
  dplyr::select(condition, readout, data) %>%
  unnest(data)

print(shap %>% dplyr::count(condition, readout))
```

## Validation

```{r checks}
# shap_value_cor is a correlation: it must lie in [-1, 1]
stopifnot("shap_value_cor outside [-1, 1]" =
  all(shap$shap_value_cor >= -1 & shap$shap_value_cor <= 1, na.rm = TRUE))
# mean_abs_shap is a magnitude
stopifnot("mean_abs_shap is negative" = all(shap$mean_abs_shap >= 0, na.rm = TRUE))

# All four models must use the same feature set, or the join below silently drops
# features and the scatter is computed on an unstated subset.
feats <- shap %>% group_split(condition, readout) %>% map(~ sort(.x$feature))
stopifnot("the four models do not share one feature set" =
  all(vapply(feats[-1], identical, logical(1), feats[[1]])))
cat("All four models share", length(feats[[1]]), "features\n")

# One row per feature per model
stopifnot("duplicate features within a model" =
  !any(duplicated(shap %>% dplyr::select(condition, readout, feature))))
```

# RNA vs TE Direction

```{r pair}
paired <- shap %>%
  dplyr::select(condition, readout, feature, feature_clean, category,
                shap_value_cor, mean_abs_shap) %>%
  pivot_wider(names_from = readout,
              values_from = c(shap_value_cor, mean_abs_shap)) %>%
  dplyr::rename(rna = shap_value_cor_RNA, te = shap_value_cor_TE) %>%
  filter(!is.na(rna), !is.na(te)) %>%
  # Distance from the y = x diagonal: how differently a feature behaves between
  # the two readouts. Used to pick which points get labelled.
  mutate(divergence = abs(te - rna),
         # Bubble size / "major" ranking uses the MEAN of the two models'
         # magnitudes. Ranking on the TE model alone would call a feature minor
         # when it is a strong RNA contributor and weak in TE, which is exactly
         # the asymmetry this plot exists to show.
         mean_abs_shap = (mean_abs_shap_RNA + mean_abs_shap_TE) / 2,
         quadrant = case_when(
           rna >= 0 & te >= 0 ~ "both positive",
           rna <  0 & te >= 0 ~ "flips: RNA down, TE up",
           rna <  0 & te <  0 ~ "both negative",
           TRUE               ~ "flips: RNA up, TE down"
         ))

stats <- paired %>%
  group_by(condition) %>%
  summarise(
    pearson  = cor(rna, te, method = "pearson"),
    spearman = cor(rna, te, method = "spearman"),
    n        = dplyr::n(),
    .groups  = "drop"
  ) %>%
  mutate(across(c(pearson, spearman), ~ round(.x, 3)))
print(as.data.frame(stats), row.names = FALSE)
write_csv(paired, here("output", "75_shap_direction_rna_vs_te.csv"))
```

```{r plot_fn}
# Subtitle is built from RUNS, so the parameters shown are the ones the files
# were located with.
make_shap_plot <- function(cond) {
  d  <- paired %>% filter(condition == cond)
  st <- stats  %>% filter(condition == cond)
  r  <- RUNS   %>% filter(condition == cond)
  te_r  <- r %>% filter(readout == "TE")
  rna_r <- r %>% filter(readout == "RNA")

  # Label the largest bubbles in EACH quadrant, so every quadrant is annotated
  # rather than only the high-divergence corners.
  lab <- d %>%
    group_by(quadrant) %>%
    slice_max(mean_abs_shap, n = LABEL_N, with_ties = FALSE) %>%
    ungroup()

  sub <- paste0(
    "TE model:  si3d ", cond, " ", TIMEPOINT, ", te_lfc > ", te_r$lfc,
    " & padj < 0.05  |  neg: ", te_r$neg_set, "\n",
    "RNA model: si3d ", cond, " ", TIMEPOINT, ", rna_lfc > ", rna_r$lfc,
    " & padj < 0.05  |  neg: ", rna_r$neg_set, "\n",
    "Both: feature_set = ", FEATURE_SET,
    ", clip = aggregate, exclude_own_te_features = TRUE, seed 9.  ",
    "Pearson R = ", st$pearson, ", Spearman rho = ", st$spearman,
    ", n = ", st$n, " features."
  )

  ggplot(d, aes(x = rna, y = te)) +
    geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.4) +
    geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey70", linewidth = 0.4) +
    geom_point(aes(size = mean_abs_shap, colour = category), alpha = 0.85) +
    geom_text_repel(data = lab, aes(label = feature_clean),
                    size = 2.8, min.segment.length = 0, max.overlaps = Inf,
                    segment.colour = "grey60", segment.size = 0.3) +
    scale_size_continuous(range = c(1, 6), name = "mean |SHAP|\n(RNA & TE avg)") +
    scale_colour_brewer(palette = "Dark2", name = NULL) +
    coord_equal(xlim = c(-1, 1), ylim = c(-1, 1)) +
    labs(
      title = paste0("SHAP direction, RNA vs TE classifier: ", cond, " ", TIMEPOINT),
      subtitle = sub,
      x = "shap_value_cor, RNA model",
      y = "shap_value_cor, TE model"
    ) +
    theme_classic(base_size = 11) +
    theme(plot.subtitle = element_text(size = 7, colour = "grey30"),
          legend.position = "right",
          legend.text = element_text(size = 7),
          legend.title = element_text(size = 7))
}
```

```{r plot_hypoxia, fig.width=9, fig.height=6.5}
p_hyp <- make_shap_plot("hypoxia")
p_hyp
ggsave(here("plots", "75_shap_direction_rna_vs_te_hypoxia1hr.pdf"),
       p_hyp, width = 9, height = 6.5)
```

```{r plot_normoxia, fig.width=9, fig.height=6.5}
p_nor <- make_shap_plot("normoxia")
p_nor
ggsave(here("plots", "75_shap_direction_rna_vs_te_normoxia1hr.pdf"),
       p_nor, width = 9, height = 6.5)
```

# Features That Flip Direction

```{r flips}
# A flip = opposite signs in the two readouts, with both contributions
# appreciable rather than one being noise around zero.
flips <- paired %>%
  filter(sign(rna) != sign(te), abs(rna) > 0.2, abs(te) > 0.2) %>%
  arrange(condition, desc(divergence)) %>%
  dplyr::select(condition, feature_clean, category, rna, te, divergence) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
print(as.data.frame(flips), row.names = FALSE)

cat("\nLargest RNA-vs-TE divergence per condition:\n")
print(as.data.frame(
  paired %>% group_by(condition) %>% slice_max(divergence, n = 5) %>%
    dplyr::select(condition, feature_clean, rna, te, divergence) %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))), row.names = FALSE)
```

# Summary

```{r summary}
print(as.data.frame(stats), row.names = FALSE)
cat("\nOutputs:\n")
cat("  output/75_shap_direction_rna_vs_te.csv\n")
cat("  plots/75_shap_direction_rna_vs_te_hypoxia1hr.pdf\n")
cat("  plots/75_shap_direction_rna_vs_te_normoxia1hr.pdf\n")
```
