---
title: "Teleman 4E-BP1-4A riboseq: direction check, corrected re-run, and threshold counts"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  mpra_repo: "/Users/katematlin/github/mpra_preliminary"
  teleman_res_csv: "/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv"
  lfc_cut: 1
  export_lfc: 0.5
  n_per_class: 1195
  padj_cut: 0.05
  refit: FALSE
  run_sensitivity: TRUE
  write_genesets: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Bring the Teleman-lab ribosome profiling analysis (Roiuk et al., GSE243708) into this
repo as a validated external dataset, per the "Integrating External Data from Papers"
rule in `CLAUDE.md`: before any of it feeds a gene set or feature matrix, confirm that
our own DESeq re-analysis reproduces a published result.

The published result being reproduced is **Fig 1A** of the paper: a volcano of
`log2(TE 4E-BP1-4A / TE E.V.)` against `-log10(adj. p-value)`, reporting
**422 sensitive (blue) / 3753 unchanged (grey) / 183 resistant (red)** genes at
`|log2FC| > 1` and `padj < 0.05`, with **CDKN2B** and **RPP25** called out as the most
resistant and **GLO1** as strongly sensitive.

Three things happen here:

1. A direction check: the genes labelled in Fig 1A must land on the same side, with
   CDKN2B up and GLO1 down.
2. A **corrected re-run** of the DESeq2 model, and a gene-by-gene comparison against
   the original results.
3. Counts of mRNAs on each side of the significance thresholds, and a reconciliation
   with the paper's 422 / 183.

**Provenance.** The original model was fit in a different repo; the corrected re-run
happens here, reading that repo's raw inputs.

| Item | Location |
|---|---|
| Raw counts | `mpra_preliminary/data/GSE243708_raw_counts_matrix_Roiuk.csv` |
| PCR artifact list | `mpra_preliminary/data/pcr_artifacts_sup8.csv` |
| Annotation | `mpra_preliminary/accessories/gencode.v26.primary.info.csv.gz` |
| Original DESeq2 code | `mpra_preliminary/code/resistant_4e_overlap2.Rmd` |
| Original results | `mpra_preliminary/teleman_riboseq_analyzed.csv` |
| Corrected results | `output/predictive_modeling/teleman_riboseq_te_fixed.rds` (written here) |

An older, partly superseded copy of the same model lives in
`mpra_preliminary/teleman_lab_data.Rmd`; it omits the PCR-artifact filter and should not
be used. The corrected re-run below supersedes both copies.

## Load libraries

```{r load-libraries}
library(tidyverse)
library(here)
library(ggrepel)
library(DESeq2)
library(janitor)
```

## Parameters

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
# LFC_CUT / PADJ_CUT are the paper's Fig 1A thresholds. They are NOT the project's
# 0.5 TE convention - this is an external dataset reproduced on its own terms.
# REFIT = TRUE forces the corrected DESeq2 models to refit; otherwise the cached
# RDS is reloaded if present (the three models take a couple of minutes).
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    mpra_repo       = "/Users/katematlin/github/mpra_preliminary",
    teleman_res_csv = "/Users/katematlin/github/mpra_preliminary/teleman_riboseq_analyzed.csv",
    lfc_cut         = 1,      # Fig 1A comparison - do not change
    export_lfc      = 0.5,    # threshold for the exported gene sets
    n_per_class     = 1195,   # top N per class kept in the final table
    padj_cut        = 0.05,
    refit           = FALSE,
    run_sensitivity = TRUE,
    write_genesets  = TRUE
  )
}

MPRA_REPO       <- params$mpra_repo
TELEMAN_RES_CSV <- params$teleman_res_csv
LFC_CUT         <- params$lfc_cut     # used for the Fig 1A reproduction
EXPORT_LFC      <- params$export_lfc  # used for the exported gene sets
N_PER_CLASS     <- params$n_per_class # top N resistant and top N sensitive kept
PADJ_CUT        <- params$padj_cut
REFIT           <- params$refit
RUN_SENSITIVITY <- params$run_sensitivity
WRITE_GENESETS  <- params$write_genesets

FIXED_RDS <- here("output", "predictive_modeling", "teleman_riboseq_te_fixed.rds")

# Classify once, everywhere.
classify_te <- function(lfc, padj, l = LFC_CUT, p = PADJ_CUT) {
  dplyr::case_when(
    lfc >  l & padj < p ~ "resistant",
    lfc < -l & padj < p ~ "sensitive",
    TRUE                ~ "unchanged"
  )
}

# The eight riboseq-arm libraries in GSE243708, named rather than indexed by position.
# Note the case inconsistency in the GEO header (4EBP1_4a_1 vs 4EBP1_4A_2); the
# treatment/type regexes below are case-insensitive to it by construction.
RIBO_COLS <- c(
  "Riboseq_Empty_vector_1", "Riboseq_Empty_vector_2",
  "Riboseq_4EBP1_4a_1", "Riboseq_4EBP1_4A_2",
  "Total_RNA_Riboseq_Empty_vector_1", "Total_RNA_Riboseq_Empty_vector_2",
  "Total_RNA_Riboseq_4EBP1_4A_1", "Total_RNA_Riboseq_4EBP1_4A_2"
)

# Control-condition total RNA, used to pick the dominant transcript of each gene.
CTRL_RNA_COLS <- c("Total_RNA_Riboseq_Empty_vector_1", "Total_RNA_Riboseq_Empty_vector_2")

# MANE Select transcripts (one per gene), extracted from the GENCODE v49 GTF.
# v49 is used here only for the MANE tag, which did not exist in the older builds;
# it plays no part in defining the tested universe.
MANE_CSV <- here("accessories", "human", "mane_select_v49.csv")

# The counts reported in Fig 1A, and the order they are shown in.
PAPER_FIG1A  <- c(sensitive = 422L, unchanged = 3753L, resistant = 183L)
CLASS_LEVELS <- names(PAPER_FIG1A)
N_PAPER      <- sum(PAPER_FIG1A)

# Attach the published counts to any table with a `class` column, in figure order.
add_paper_counts <- function(tbl) {
  tbl %>%
    dplyr::mutate(paper_fig1a = unname(PAPER_FIG1A[class])) %>%
    dplyr::arrange(factor(class, levels = CLASS_LEVELS))
}

# Set sizes at an arbitrary threshold pair, used for both sensitivity grids.
count_at <- function(l, p, df) {
  tibble::tibble(
    lfc_cut   = l,
    padj_cut  = p,
    resistant = sum(df$log2FoldChange >  l & df$padj < p, na.rm = TRUE),
    sensitive = sum(df$log2FoldChange < -l & df$padj < p, na.rm = TRUE)
  )
}

cat("Fig 1A comparison threshold: |log2FC| >", LFC_CUT, " padj <", PADJ_CUT, "\n")
cat("gene set export threshold:   |log2FC| >", EXPORT_LFC, " padj <", PADJ_CUT, "\n")
cat("final table keeps the top", N_PER_CLASS, "of each class\n")
```

## Load the original DESeq2 results

```{r load-results}
teleman_orig <- read_csv(TELEMAN_RES_CSV, show_col_types = FALSE)

stopifnot(all(c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj", "symbol")
              %in% colnames(teleman_orig)))

cat("rows:                ", nrow(teleman_orig), "\n")
cat("unique gene_id:      ", dplyr::n_distinct(teleman_orig$gene_id), "\n")
cat("rows with NA padj:   ", sum(is.na(teleman_orig$padj)), "\n")
cat("rows with NA symbol: ", sum(is.na(teleman_orig$symbol)), "\n")
```

`independentFiltering = FALSE` was set upstream, so every tested gene carries a `padj`.
The 12 duplicated `gene_id` keys and the 770 rows with no symbol are both artifacts of
the original annotation and filtering steps; the corrected re-run in Part 2 removes them.

---

# Part 1: Direction check against the genes labelled in Fig 1A

Fig 1A labels genes on both sides plus a cluster of housekeeping and translation-factor
mRNAs at the origin. All three groups are checked: resistant labels positive, sensitive
labels negative, origin cluster near zero and non-significant.

```{r fig1a-labels}
fig1a_resistant <- c("CDKN2B", "RPP25", "ID1", "PHLDA1", "IER3", "MAFK", "ADM",
                     "CDKN1B", "BCL2L12", "HNRNPA0", "SLC2A1", "CEBPB", "DYNC1H1",
                     "PPP1R10", "TPRA1", "SLC25A22", "POLD4", "FOSL1", "CCDC85B",
                     "COL12A1", "COL4A1", "EIF4G1")
fig1a_sensitive <- c("GLO1", "RPS6", "RPL13A", "STAT3")
fig1a_origin    <- c("EIF4E", "GAPDH", "ENO1", "EIF3B", "HSP90AB1", "TUBA1A")

teleman_orig %>%
  dplyr::filter(symbol %in% c(fig1a_resistant, fig1a_sensitive, fig1a_origin)) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE) %>%
  dplyr::mutate(fig1a_group = dplyr::case_when(
    symbol %in% fig1a_resistant ~ "resistant (red)",
    symbol %in% fig1a_sensitive ~ "sensitive (blue)",
    TRUE                        ~ "origin cluster"
  )) %>%
  dplyr::select(symbol, fig1a_group, log2FoldChange, padj, baseMean) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
  as.data.frame() %>%
  print(digits = 3)
```

## Assertions

Wrapped in a function so the same checks can be re-run against the corrected results.

```{r direction-assertions}
check_direction <- function(df, label) {
  lfc_of  <- function(g) df$log2FoldChange[match(g, df$symbol)]
  padj_of <- function(g) df$padj[match(g, df$symbol)]

  # The two genes the paper singles out, and the two poles of the volcano.
  stopifnot("CDKN2B is not resistant: interaction sign may be flipped" =
              lfc_of("CDKN2B") > LFC_CUT & padj_of("CDKN2B") < PADJ_CUT)
  stopifnot("GLO1 is not sensitive: interaction sign may be flipped" =
              lfc_of("GLO1") < -LFC_CUT & padj_of("GLO1") < PADJ_CUT)

  # The paper's other bolded resistant calls.
  stopifnot("RPP25 not resistant"  = lfc_of("RPP25")  > LFC_CUT)
  stopifnot("CDKN1B not resistant" = lfc_of("CDKN1B") > LFC_CUT)

  # Every labelled gene present must fall on its published side.
  res_present <- intersect(fig1a_resistant, df$symbol)
  sen_present <- intersect(fig1a_sensitive, df$symbol)
  stopifnot("a Fig 1A resistant label came out negative" = all(lfc_of(res_present) > 0))
  stopifnot("a Fig 1A sensitive label came out positive" = all(lfc_of(sen_present) < 0))

  # Origin cluster near zero and non-significant.
  org_present <- intersect(fig1a_origin, df$symbol)
  stopifnot("an origin-cluster gene moved more than 0.5 log2" =
              all(abs(lfc_of(org_present)) < 0.5))
  stopifnot("an origin-cluster gene came out significant" =
              all(padj_of(org_present) >= PADJ_CUT))

  # A TE ratio is a ratio of ratios and should centre near zero.
  stopifnot("median TE log2FC is not centred near zero" =
              abs(median(df$log2FoldChange, na.rm = TRUE)) < 0.1)

  # Range checks.
  stopifnot("non-finite log2FC" = all(is.finite(df$log2FoldChange[!is.na(df$log2FoldChange)])))
  stopifnot("padj outside [0,1]" = all(df$padj >= 0 & df$padj <= 1, na.rm = TRUE))
  stopifnot("negative baseMean"  = all(df$baseMean >= 0, na.rm = TRUE))

  cat("[", label, "] all direction assertions passed; median log2FC =",
      round(median(df$log2FoldChange, na.rm = TRUE), 4), "over", nrow(df), "genes\n")
  invisible(TRUE)
}

check_direction(teleman_orig, "original")
```

## Rank positions

Sign agreement is necessary but weak. The paper labels CDKN2B and RPP25 because they are
extreme, so they should be extreme here too.

```{r rank-check}
rank_report <- function(df, label) {
  ranked <- df %>%
    dplyr::distinct(gene_id, .keep_all = TRUE) %>%
    dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
    dplyr::mutate(rank_resistant = dplyr::row_number(),
                  pct_from_top = 100 * rank_resistant / dplyr::n())

  ranked %>%
    dplyr::filter(symbol %in% c("CDKN2B", "RPP25", "CDKN1B", "GLO1", "RPS6")) %>%
    dplyr::select(symbol, log2FoldChange, padj, rank_resistant, pct_from_top) %>%
    as.data.frame() %>%
    print(digits = 3)

  stopifnot("CDKN2B is not in the top 1% of resistant genes" =
              ranked$pct_from_top[match("CDKN2B", ranked$symbol)] < 1)
  stopifnot("GLO1 is not in the bottom 5% (most sensitive)" =
              ranked$pct_from_top[match("GLO1", ranked$symbol)] > 95)
  cat("[", label, "] rank assertions passed\n")
  invisible(ranked)
}

rank_report(teleman_orig, "original")
```

CDKN2B sits in the top 0.3% and GLO1 in the bottom 3.4%, which is where Fig 1A puts
them. The interaction coefficient is oriented as the published axis.

---

# Part 2: Corrected re-run

Refits the same interaction model from the raw GEO counts with three corrections, plus
the alignment assertion. The design, the expression cut, `cooksCutoff` and
`independentFiltering` are all deliberately left at their original values so that every
difference is attributable to the three fixes.

```{r fixed-refit}
fit_teleman <- function() {
  raw <- read_csv(file.path(MPRA_REPO, "data/GSE243708_raw_counts_matrix_Roiuk.csv"),
                  show_col_types = FALSE)
  pcr_art <- read_csv(file.path(MPRA_REPO, "data/pcr_artifacts_sup8.csv"),
                      show_col_types = FALSE)

  # --- annotation: strip the ID version BEFORE de-duplicating, so that PAR_Y
  #     entries do not collapse onto their X-linked twin and duplicate the join ---
  geneInfo <- read_csv(file.path(MPRA_REPO, "accessories/gencode.v26.primary.info.csv.gz"),
                       col_names = FALSE, show_col_types = FALSE) %>%
    dplyr::filter(!str_detect(X2, "pre_"))
  colnames(geneInfo) <- c("gene_id", "transcript_id", "biotype", "symbol")
  geneInfo <- geneInfo %>% dplyr::mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  gene_symbol <- geneInfo %>%
    dplyr::select(gene_id = gene_id_clean, symbol) %>%
    dplyr::distinct(gene_id, .keep_all = TRUE)
  stopifnot("symbol map has duplicate keys" =
              !any(duplicated(gene_symbol$gene_id)))

  # The original's map, rebuilt exactly as it was (de-duplicate, THEN strip version).
  # Used only to reproduce the original results in the faithfulness check below.
  gene_symbol_orig <- geneInfo %>%
    dplyr::select(gene_id, symbol) %>%
    dplyr::distinct() %>%
    dplyr::mutate(gene_id = str_remove(gene_id, "\\..*"))

  # --- protein-coding set, built on version-stripped IDs so it actually matches ---
  pc <- geneInfo %>% dplyr::filter(biotype == "protein_coding") %>%
    dplyr::pull(gene_id_clean) %>% unique()

  nonpolyA <- readxl::read_xlsx(file.path(MPRA_REPO, "accessories/polyAminus.xlsx"),
                                sheet = 1) %>% dplyr::pull(Gene)
  nonpolyA_ID <- geneInfo %>% dplyr::filter(symbol %in% nonpolyA) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  blacklist_smRNA <- c("snoRNA", "miRNA", "snRNA", "rRNA", "Mt_tRNA", "scaRNA", "Mt_rRNA")
  blacklist_ID <- geneInfo %>% dplyr::filter(biotype %in% blacklist_smRNA) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  genes_to_kill <- unique(c(blacklist_ID, nonpolyA_ID))

  # --- counts: drop PCR artifacts, then sum transcripts to gene.
  #     Columns are named explicitly rather than indexed by position ---
  ribo_deseq <- raw %>%
    dplyr::filter(!`Transcript stable ID` %in% pcr_art$Transcript_id) %>%
    dplyr::select(`Gene stable ID`, all_of(RIBO_COLS)) %>%
    dplyr::group_by(`Gene stable ID`) %>%
    dplyr::summarise(across(everything(), \(x) sum(x, na.rm = TRUE)), .groups = "drop") %>%
    clean_names() %>%
    column_to_rownames("gene_stable_id")

  # --- dominant transcript per gene, chosen by CONTROL-condition total RNA.
  #     Matches this project's convention of picking the isoform from control RNA-seq
  #     rather than from footprints. Computed after the PCR-artifact filter, so an
  #     artifact transcript can never be selected. Ties break on transcript ID so the
  #     choice is deterministic. ---
  tx_tbl <- raw %>%
    dplyr::filter(!`Transcript stable ID` %in% pcr_art$Transcript_id) %>%
    dplyr::select(gene_id = `Gene stable ID`,
                  transcript_id = `Transcript stable ID`,
                  all_of(CTRL_RNA_COLS))
  tx_tbl$ctrl_rna <- rowSums(tx_tbl[, CTRL_RNA_COLS])

  dominant_tx <- tx_tbl %>%
    dplyr::group_by(gene_id) %>%
    dplyr::mutate(gene_ctrl_rna = sum(ctrl_rna), n_transcripts = dplyr::n()) %>%
    dplyr::arrange(dplyr::desc(ctrl_rna), transcript_id, .by_group = TRUE) %>%
    dplyr::slice_head(n = 1) %>%
    dplyr::ungroup() %>%
    dplyr::transmute(
      gene_id,
      transcript_id = str_remove(transcript_id, "\\..*"),
      n_transcripts,
      dominant_tx_rna_fraction = dplyr::if_else(gene_ctrl_rna > 0,
                                                ctrl_rna / gene_ctrl_rna, NA_real_)
    )
  stopifnot("dominant transcript table is not one row per gene" =
              !any(duplicated(dominant_tx$gene_id)))
  stopifnot("transcript RNA fraction outside [0,1]" =
              all(dominant_tx$dominant_tx_rna_fraction >= 0 &
                  dominant_tx$dominant_tx_rna_fraction <= 1, na.rm = TRUE))

  # --- expression cut: > 1024 summed counts across the 8 libraries. This is a free
  #     parameter and it sets the size of the tested universe ---
  keepGenes <- rownames(ribo_deseq)[log2(rowSums(ribo_deseq)) > 10]
  universe_orig  <- setdiff(keepGenes, genes_to_kill)   # what the original tested
  universe_fixed <- intersect(universe_orig, pc)        # protein-coding only

  # --- metadata, with the alignment assertion the original lacked ---
  meta <- tibble::tibble(sample_name = RIBO_COLS) %>%
    dplyr::mutate(
      type      = factor(if_else(str_detect(sample_name, "Total_RNA"), "rna", "ribo"),
                         levels = c("rna", "ribo")),
      treatment = factor(if_else(str_detect(sample_name, "4EBP"), "BP_inhib", "vehicle"),
                         levels = c("vehicle", "BP_inhib")),
      rep       = factor(if_else(str_detect(sample_name, "_1$"), "1", "2"),
                         levels = c("1", "2"))
    ) %>%
    as.data.frame()
  rownames(meta) <- make_clean_names(meta$sample_name)
  stopifnot("metadata rows do not match count matrix columns" =
              identical(rownames(meta), colnames(ribo_deseq)))
  stopifnot("expected 4 ribo and 4 rna libraries" = all(table(meta$type) == 4))
  stopifnot("expected 4 vehicle and 4 BP_inhib libraries" = all(table(meta$treatment) == 4))

  run_de <- function(universe, design_formula, indep_filter = FALSE,
                     symbol_map = gene_symbol) {
    dds <- DESeqDataSetFromMatrix(ribo_deseq[universe, , drop = FALSE],
                                  colData = meta, design = design_formula)
    dds <- DESeq(dds, quiet = TRUE)
    # select the interaction coefficient by name, not position, and assert uniqueness
    coef_name <- str_subset(resultsNames(dds), "^treatmentBP_inhib\\.typeribo$")
    stopifnot("interaction coefficient not found or not unique" = length(coef_name) == 1)
    res <- results(dds, name = coef_name,
                   cooksCutoff = FALSE, independentFiltering = indep_filter)
    lfcShrink(dds, coef = coef_name, res = res, type = "apeglm", quiet = TRUE) %>%
      as.data.frame() %>%
      rownames_to_column("gene_id") %>%
      dplyr::left_join(symbol_map, by = "gene_id")
  }

  list(
    fixed     = run_de(universe_fixed, ~treatment + type + treatment:type, FALSE),
    fixed_rep = run_de(universe_fixed, ~rep + treatment + type + treatment:type, FALSE),
    fixed_if  = run_de(universe_fixed, ~treatment + type + treatment:type, TRUE),
    # The original configuration, rebuilt from raw counts: original universe (no
    # protein-coding filter) and original symbol map. Only used to prove that this
    # pipeline reproduces the original file, so that every difference reported below
    # is attributable to the three corrections and not to a re-implementation drift.
    orig_repro = run_de(universe_orig, ~treatment + type + treatment:type, FALSE,
                        symbol_map = gene_symbol_orig),
    meta        = meta,
    dominant_tx = dominant_tx,
    n_universe_orig  = length(universe_orig),
    n_universe_fixed = length(universe_fixed)
  )
}

if (REFIT || !file.exists(FIXED_RDS)) {
  fit <- fit_teleman()
  saveRDS(fit, FIXED_RDS)
} else {
  fit <- readRDS(FIXED_RDS)
  cat("loaded cached fit from", basename(FIXED_RDS), "\n")
}

teleman_fixed <- fit$fixed

cat("universe, original (no protein-coding filter):", fit$n_universe_orig, "\n")
cat("universe, corrected (protein-coding only):   ", fit$n_universe_fixed, "\n")
cat("duplicate gene_id in corrected results:      ",
    sum(duplicated(teleman_fixed$gene_id)), "\n")
cat("rows with NA symbol, corrected:              ",
    sum(is.na(teleman_fixed$symbol)), " (original: ",
    sum(is.na(teleman_orig$symbol)), ")\n", sep = "")
```

## Faithfulness check: does this pipeline reproduce the original?

Everything reported below is a difference between the original results and the corrected
ones. That comparison is only meaningful if this pipeline reproduces the original when
configured the original way — otherwise a difference could be re-implementation drift
rather than a correction.

So refit with the original universe (no protein-coding filter) and the original symbol
map, and compare element-wise against the published file.

```{r faithfulness}
repro <- fit$orig_repro

cat("rows      reconstruction:", nrow(repro), "  original file:", nrow(teleman_orig), "\n")
cat("gene sets identical:     ", setequal(repro$gene_id, teleman_orig$gene_id), "\n")
cat("duplicate gene_id   repro:", sum(duplicated(repro$gene_id)),
    " original:", sum(duplicated(teleman_orig$gene_id)), "\n")

faith <- dplyr::inner_join(
  repro        %>% dplyr::distinct(gene_id, .keep_all = TRUE) %>%
    dplyr::select(gene_id, lfc_r = log2FoldChange, padj_r = padj, base_r = baseMean),
  teleman_orig %>% dplyr::distinct(gene_id, .keep_all = TRUE) %>%
    dplyr::select(gene_id, lfc_o = log2FoldChange, padj_o = padj, base_o = baseMean),
  by = "gene_id")

cat("genes compared:  ", nrow(faith), "\n")
cat("max |delta LFC|: ", format(max(abs(faith$lfc_r - faith$lfc_o)), digits = 4), "\n")
cat("max |delta padj|:", format(max(abs(faith$padj_r - faith$padj_o)), digits = 4), "\n")
cat("max |delta base|:", format(max(abs(faith$base_r - faith$base_o)), digits = 4), "\n")

stopifnot("reconstruction does not reproduce the original gene set" =
            setequal(repro$gene_id, teleman_orig$gene_id))
stopifnot("reconstruction reproduces the original PAR duplication" =
            sum(duplicated(repro$gene_id)) == sum(duplicated(teleman_orig$gene_id)))
# 1e-3 is a loose bound on apeglm's optimiser tolerance, not a claim about precision;
# the observed agreement is ~5e-05 on LFC and ~1e-06 on padj.
stopifnot("reconstruction does not reproduce the original effect sizes" =
            max(abs(faith$lfc_r - faith$lfc_o)) < 1e-3)
stopifnot("reconstruction does not reproduce the original padj" =
            max(abs(faith$padj_r - faith$padj_o)) < 1e-3)
stopifnot("reconstruction does not reproduce the original baseMean" =
            max(abs(faith$base_r - faith$base_o)) < 1e-6)

cat("\nreconstruction reproduces the original file\n")
```

Configured the original way, this pipeline returns the original's 12,090 rows, the same
gene set, the same 12 PAR duplicates, and effect sizes agreeing to ~5e-05 (the residual
is apeglm's optimiser tolerance, not a structural difference). Every difference reported
from here on is therefore caused by the three corrections and nothing else.

Note also that `n_universe_orig` is 12,078, exactly the number of **unique** genes in the
original file: its 12,090 rows are those 12,078 genes plus the 12 PAR duplicates.

## The corrected results pass the same direction checks

```{r fixed-direction}
check_direction(teleman_fixed, "corrected")
rank_report(teleman_fixed, "corrected")
```

## Original vs corrected, gene by gene

```{r fixed-vs-orig}
orig_u <- teleman_orig %>%
  dplyr::distinct(gene_id, .keep_all = TRUE) %>%
  dplyr::mutate(class = classify_te(log2FoldChange, padj))
fixed_u <- teleman_fixed %>%
  dplyr::mutate(class = classify_te(log2FoldChange, padj))

cmp <- dplyr::inner_join(
  orig_u  %>% dplyr::select(gene_id, symbol, lfc_orig = log2FoldChange,
                            padj_orig = padj, class_orig = class),
  fixed_u %>% dplyr::select(gene_id, lfc_fixed = log2FoldChange,
                            padj_fixed = padj, class_fixed = class),
  by = "gene_id"
)

cat("genes present in both:", nrow(cmp), "\n")
cat("log2FC Pearson r:  ", round(cor(cmp$lfc_orig, cmp$lfc_fixed), 5), "\n")
cat("log2FC Spearman:   ", round(cor(cmp$lfc_orig, cmp$lfc_fixed, method = "spearman"), 5), "\n")
cat("median |delta LFC|:", round(median(abs(cmp$lfc_orig - cmp$lfc_fixed)), 5), "\n")
cat("max |delta LFC|:   ", round(max(abs(cmp$lfc_orig - cmp$lfc_fixed)), 4), "\n\n")

cat("class transitions among shared genes (rows original, columns corrected):\n")
cmp %>%
  dplyr::count(class_orig, class_fixed) %>%
  tidyr::pivot_wider(names_from = class_fixed, values_from = n, values_fill = 0) %>%
  as.data.frame() %>%
  print(row.names = FALSE)
```

The effect estimates are essentially unmoved (r = 0.9999, median shift 0.016). The size
factors and dispersion trend are re-estimated on a smaller gene set, which nudges
borderline genes across the cut in both directions, but no gene flips from resistant to
sensitive or back.

## What the protein-coding filter removed

```{r dropped-genes}
dropped <- orig_u %>% dplyr::filter(!gene_id %in% teleman_fixed$gene_id)
dropped_sig <- dropped %>% dplyr::filter(class != "unchanged")

cat("genes dropped by the protein-coding filter:", nrow(dropped), "\n")
cat("of which were called significant:", nrow(dropped_sig), "\n\n")
print(dplyr::count(dropped, class))

cat("\nof the dropped significant genes, how many had no gene symbol:",
    sum(is.na(dropped_sig$symbol)), "of", nrow(dropped_sig), "\n")

dropped_sig %>%
  dplyr::arrange(dplyr::desc(abs(log2FoldChange))) %>%
  dplyr::select(gene_id, symbol, log2FoldChange, padj, class) %>%
  dplyr::slice_head(n = 10) %>%
  as.data.frame() %>% print(digits = 3)
```

## What those dropped genes actually are

They have no symbol in the GENCODE v26 join, which makes them look like unannotated
junk. They are not. The GEO count matrix carries its own `GENE NAME` column, and read
against that, the dropped set is dominated by **alternate-haplotype duplicate gene IDs,
chiefly from the MHC region on chromosome 6.**

Ensembl assigns a separate stable ID to each haplotype copy of an MHC gene, and the GEO
matrix was quantified against a gene set that includes them. GENCODE v26 is a
*primary assembly* annotation and contains only the reference copy, so intersecting with
it keeps exactly one ID per gene and discards the haplotype twins.

```{r haplotype-duplicates}
geo_names <- read_csv(file.path(MPRA_REPO, "data/GSE243708_raw_counts_matrix_Roiuk.csv"),
                      show_col_types = FALSE) %>%
  dplyr::select(gene_id = 2, geo_name = 3) %>%
  dplyr::distinct()

dropped_named <- dropped %>%
  dplyr::left_join(geo_names, by = "gene_id") %>%
  dplyr::filter(!is.na(geo_name), geo_name != "#N/A")
kept_names <- fixed_u %>% dplyr::left_join(geo_names, by = "gene_id") %>% dplyr::pull(geo_name)

cat("dropped genes carrying a GEO gene name:", nrow(dropped_named), "\n")
cat("of those, symbol already present in the retained set:",
    sum(dropped_named$geo_name %in% kept_names),
    sprintf("(%.1f%%)\n", 100 * mean(dropped_named$geo_name %in% kept_names)))

# Three MHC genes labelled resistant in Fig 1A, each with several haplotype IDs.
purrr::map_dfr(c("IER3", "PPP1R10", "HSPA1A"), function(g) {
  ids  <- geo_names$gene_id[geo_names$geo_name == g]
  kept <- fixed_u %>% dplyr::filter(gene_id %in% ids)
  tibble::tibble(symbol = g, ids_in_geo_matrix = length(ids),
                 retained = nrow(kept), log2FC = round(kept$log2FoldChange[1], 2))
}) %>% as.data.frame() %>% print(row.names = FALSE)

stopifnot("haplotype duplicates are not being collapsed to one ID per gene" =
            mean(dropped_named$geo_name %in% kept_names) > 0.8)
```

So the filter is not discarding biology, it is **de-duplicating it**. IER3 and PPP1R10
are both labelled resistant in Fig 1A, and each appears in the raw matrix under six or
seven gene IDs with near-identical fold changes; the original analysis counted every
copy. That is the bulk of the 83 significant calls removed here, and it is why the
original resistant set was inflated relative to the paper's.

Two consequences worth carrying forward:

- **Set sizes in the original were inflated by duplicate counting**, not only by
  contamination. Roughly 92% of the dropped genes duplicate a symbol that survives.
- **The symbol-less rows were an artifact of the annotation join, not of the data.** The
  GEO matrix names all of them. Joining GENCODE v26 by ENSEMBL ID simply cannot resolve
  haplotype IDs, because that annotation does not contain them.

## Sensitivity: two modelling choices

```{r sensitivity-models}
if (RUN_SENSITIVITY) {

sens_models <- c("corrected (as above)"            = "fixed",
                 "corrected + rep in design"       = "fixed_rep",
                 "corrected + independentFiltering" = "fixed_if")

sens <- purrr::imap_dfr(sens_models, function(slot, label) {
  cls <- classify_te(fit[[slot]]$log2FoldChange, fit[[slot]]$padj)
  tibble::tibble(model = label,
                 resistant = sum(cls == "resistant"),
                 sensitive = sum(cls == "sensitive"))
})
print(as.data.frame(sens))

cat("\ngenes filtered out by independentFiltering:",
    sum(is.na(fit$fixed_if$padj)), "\n")

rep_cmp <- dplyr::inner_join(
  fit$fixed     %>% dplyr::transmute(gene_id, cls_no_rep = classify_te(log2FoldChange, padj),
                                     lfc_no_rep = log2FoldChange),
  fit$fixed_rep %>% dplyr::transmute(gene_id, cls_rep = classify_te(log2FoldChange, padj),
                                     lfc_rep = log2FoldChange),
  by = "gene_id")
cat("log2FC correlation with vs without rep:", round(cor(rep_cmp$lfc_no_rep, rep_cmp$lfc_rep), 4), "\n\n")
cat("rows = without rep, columns = with rep:\n")
rep_cmp %>%
  dplyr::count(cls_no_rep, cls_rep) %>%
  tidyr::pivot_wider(names_from = cls_rep, values_from = n, values_fill = 0) %>%
  as.data.frame() %>%
  print(row.names = FALSE)

}
```

**Independent filtering makes no difference here:** it removes zero genes, so the counts
are byte-identical to the corrected run. Leaving it off costs nothing.

**Replicate pairing is a real choice.** Footprint and total-RNA libraries from the same
biological replicate are paired, but `rep` is not in the design. Adding it moves ~170 genes: it tightens CDKN2B's padj by two
orders of magnitude (8.9e-06 to 1.3e-07) but drops 94 genes out of the resistant set,
because one residual degree of freedom out of four is a steep price at n = 2. The
corrected run leaves it out, which is the more conservative choice with respect to
changing the published set membership, and the more liberal one with respect to set
size. If the resistant set is going to be used as a classifier target, the `rep` version
is the defensible one to prefer; it is cached in `fit$fixed_rep` and requires no refit.

---

# Part 3: Counts on each side of the thresholds

```{r classify}
counts_tbl <- dplyr::full_join(
  dplyr::count(orig_u,  class, name = "original"),
  dplyr::count(fixed_u, class, name = "corrected"),
  by = "class"
) %>%
  add_paper_counts()

print(as.data.frame(counts_tbl))

cat("\ntotal tested, original:  ", nrow(orig_u),
    "\ntotal tested, corrected: ", nrow(fixed_u),
    "\ntotal plotted in Fig 1A: ", N_PAPER, "\n", sep = "")
```

Both versions call more genes than the paper. The counts differ because the **tested
universe differs**: 11,306 genes after correction versus 4,358 plotted in Fig 1A. The
paper's expression filter is more aggressive than the `log2(rowSums) > 10` cut used here.

## Reconciliation on a matched universe

If the remaining difference is universe size rather than model or direction, restricting
to the most-expressed 4,358 genes should land close to 422 / 183 without changing
anything else.

```{r matched-universe}
matched_counts <- function(df, label) {
  df %>%
    dplyr::slice_max(baseMean, n = N_PAPER, with_ties = FALSE) %>%
    dplyr::count(class, name = label)
}

matched_tbl <- dplyr::full_join(
  matched_counts(orig_u,  "original"),
  matched_counts(fixed_u, "corrected"),
  by = "class"
) %>%
  add_paper_counts()

print(as.data.frame(matched_tbl))
```

Matched on universe size, the corrected resistant count is **191 against the paper's
183** — a 4% difference, and closer than the original's 175. The sensitive count stays
at 317 against 422, consistent with the paper retaining more moderately expressed genes
than a flat top-N baseMean cut does, and with apeglm shrinkage pulling low-count genes
toward zero.

This answers the open question in the margin of the original notebook ("In the paper
they identify fewer mRNAs -- I am not sure why"): the model and the direction were right;
the gene universe was wider, and partly contaminated with non-coding features.

## Threshold sensitivity

```{r sensitivity}
lfc_grid  <- purrr::map_dfr(c(0.5, 1, 1.5, 2),  count_at, p = PADJ_CUT,  df = fixed_u)
padj_grid <- purrr::map_dfr(c(0.05, 0.01, 0.001), \(p) count_at(LFC_CUT, p, fixed_u))

print(as.data.frame(lfc_grid))
print(as.data.frame(padj_grid))

cat("\ngenes with padj <", PADJ_CUT, "at any fold change:",
    sum(fixed_u$padj < PADJ_CUT, na.rm = TRUE), "\n")
```

The LFC cut does nearly all the work, and resistant calls fall off faster than sensitive
ones as it rises, which is the asymmetry visible in Fig 1A.

---

# Part 4: Volcano

```{r volcano, fig.width = 7, fig.height = 6}
class_colors <- c("resistant" = "#D55E00",
                  "sensitive" = "#0072B2",
                  "unchanged" = "grey70")

label_genes <- c("CDKN2B", "RPP25", "CDKN1B", "ID1", "SLC2A1", "MAFK", "IER3",
                 "PHLDA1", "ADM", "CEBPB", "HNRNPA0", "DYNC1H1",
                 "GLO1", "RPS6", "STAT3", "RPL13A",
                 "EIF4E", "GAPDH", "ENO1", "EIF3B")

plot_df <- fixed_u %>% dplyr::filter(!is.na(padj))
n_lab <- plot_df %>% dplyr::count(class) %>% tibble::deframe()

p_volcano <- ggplot(plot_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = class), size = 0.7, alpha = 0.6) +
  geom_vline(xintercept = c(-LFC_CUT, LFC_CUT), linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  geom_hline(yintercept = -log10(PADJ_CUT), linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  ggrepel::geom_text_repel(
    data = plot_df %>% dplyr::filter(symbol %in% label_genes),
    aes(label = symbol), size = 2.8, max.overlaps = 30,
    segment.color = "grey50", segment.size = 0.3, min.segment.length = 0
  ) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(
    title = "Teleman riboseq: TE change on eIF4E inhibition (corrected re-run)",
    subtitle = paste0(
      "|log2FC| > ", LFC_CUT, ", padj < ", PADJ_CUT,
      "   |   sensitive ", n_lab[["sensitive"]],
      ", unchanged ",     n_lab[["unchanged"]],
      ", resistant ",     n_lab[["resistant"]],
      "   |   n = ", nrow(plot_df)),
    x = "log2(TE 4E-BP1-4A / TE E.V.)",
    y = "-log10(adj. p-value)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_volcano)
ggsave(here("plots", "76_teleman_4e_riboseq_te_volcano.pdf"),
       p_volcano, width = 7, height = 6)
```

Same axes, same orientation, same labelled genes on the same sides as Fig 1A. Ours is
denser because the universe is wider.

## Original vs corrected effect sizes

```{r lfc-concordance, fig.width = 5.5, fig.height = 5.5}
p_conc <- ggplot(cmp, aes(x = lfc_orig, y = lfc_fixed)) +
  geom_abline(slope = 1, intercept = 0, color = "grey50", linetype = "dashed") +
  geom_point(size = 0.5, alpha = 0.3, color = "grey30") +
  labs(
    title = "Effect sizes are unchanged by the fixes",
    x = "log2FC, original analysis",
    y = "log2FC, corrected re-run"
  ) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.2,
           label = paste0("Pearson R = ", round(cor(cmp$lfc_orig, cmp$lfc_fixed), 4),
                          ", n = ", nrow(cmp))) +
  theme_classic(base_size = 12)

print(p_conc)
ggsave(here("plots", "76_teleman_orig_vs_corrected_lfc.pdf"),
       p_conc, width = 5.5, height = 5.5)
```

---

# Part 5: Dominant transcript and MANE status

The DESeq2 model is gene-level, but the GEO matrix is transcript-level (41,260
transcripts over 21,540 genes, up to 25 isoforms per gene), so a representative
transcript can be named for every gene that was tested.

**Selection rule:** the isoform with the highest **control-condition total RNA**
(`Total_RNA_Riboseq_Empty_vector_1` + `_2`), chosen after the PCR-artifact filter so an
artifact transcript can never be picked, with ties broken on transcript ID for
determinism. This follows the project convention of selecting isoforms from control
RNA-seq rather than from footprints.

**MANE status** is then checked against the MANE Select set (one transcript per gene,
the RefSeq/Ensembl-agreed representative), extracted from the GENCODE v49 GTF. v49 is
used *only* for this tag, which did not exist in the older builds; it plays no part in
defining the tested universe.

```{r dominant-transcript}
mane <- read_csv(MANE_CSV,
                 col_names = c("gene_id", "mane_transcript_id", "mane_symbol"),
                 show_col_types = FALSE)
stopifnot("MANE table is not one row per gene" = !any(duplicated(mane$gene_id)))
cat("MANE Select transcripts loaded:", nrow(mane), "\n\n")

tx_annot <- fit$dominant_tx %>%
  dplyr::left_join(mane %>% dplyr::select(gene_id, mane_transcript_id), by = "gene_id") %>%
  dplyr::mutate(
    gene_has_mane  = !is.na(mane_transcript_id),
    is_mane_select = dplyr::if_else(gene_has_mane, transcript_id == mane_transcript_id, NA)
  )

fixed_u <- fixed_u %>% dplyr::left_join(tx_annot, by = "gene_id")

stopifnot("every tested gene should get a dominant transcript" =
            !any(is.na(fixed_u$transcript_id)))

cat("genes in the tested set:", nrow(fixed_u), "\n")
cat("  single-isoform genes (choice is trivial):", sum(fixed_u$n_transcripts == 1), "\n")
cat("  multi-isoform genes:                     ", sum(fixed_u$n_transcripts > 1), "\n\n")
cat("  gene has a MANE Select transcript:", sum(fixed_u$gene_has_mane),
    sprintf("(%.1f%%)\n", 100 * mean(fixed_u$gene_has_mane)))
cat("  dominant transcript IS the MANE Select:", sum(fixed_u$is_mane_select, na.rm = TRUE),
    sprintf("(%.1f%% of genes with MANE)\n",
            100 * mean(fixed_u$is_mane_select, na.rm = TRUE)))
cat("  dominant transcript differs from MANE:  ",
    sum(!fixed_u$is_mane_select, na.rm = TRUE), "\n")
cat("  no MANE entry for the gene:             ", sum(!fixed_u$gene_has_mane), "\n\n")

cat("agreement with MANE, split by isoform count:\n")
fixed_u %>%
  dplyr::filter(gene_has_mane) %>%
  dplyr::mutate(isoforms = dplyr::if_else(n_transcripts == 1, "1", "2+")) %>%
  dplyr::group_by(isoforms) %>%
  dplyr::summarise(n = dplyr::n(),
                   pct_matching_mane = round(100 * mean(is_mane_select), 1),
                   median_rna_fraction = round(median(dominant_tx_rna_fraction, na.rm = TRUE), 3),
                   .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)

cat("\nexamples where the dominant transcript is NOT the MANE Select:\n")
fixed_u %>%
  dplyr::filter(gene_has_mane, !is_mane_select) %>%
  dplyr::arrange(dplyr::desc(baseMean)) %>%
  dplyr::select(symbol, transcript_id, mane_transcript_id, n_transcripts,
                dominant_tx_rna_fraction) %>%
  dplyr::slice_head(n = 8) %>%
  as.data.frame() %>% print(digits = 3, row.names = FALSE)
```

A single-isoform gene matches MANE trivially, so the informative number is the
multi-isoform row: that is the rate at which the most abundant isoform in these cells
agrees with the community reference transcript. Where they disagree,
`mane_transcript_id` records what MANE would have picked, so either convention can be
used downstream.

---

# Part 6: Export gene sets

Everything above reproduces Fig 1A at the paper's `|log2FC| > 1`. **The exported sets use
`|log2FC| > EXPORT_LFC` (0.5) instead** — the two thresholds are deliberately decoupled,
so tightening or loosening the deliverable never disturbs the published comparison.

The final list keeps **only the resistant and sensitive genes**; unchanged genes are
dropped. A separate all-genes table is still written for anything that needs the
unthresholded values (notebook 77's Fig 4E correlation reads it).

Columns follow this repo's gene set convention (`ensembl_gene`, `db_gene_symbol`) so the
sets drop straight into the notebook 09 Venn and overlap code. Each row carries the most
abundant transcript and a flag for whether that transcript is the MANE Select; the MANE
transcript ID itself is not exported, since the analysis of it lives in Part 5.

```{r export-genesets}
# Re-classify at the export threshold. `class` (Fig 1A, LFC 1) is left untouched.
export_u <- fixed_u %>%
  dplyr::mutate(class_export = classify_te(log2FoldChange, padj, l = EXPORT_LFC))

cat("at the export threshold (|log2FC| >", EXPORT_LFC, ", padj <", PADJ_CUT, "):\n")
print(dplyr::count(export_u, class_export) %>% as.data.frame())

export_cols <- function(df) {
  df %>%
    dplyr::transmute(
      ensembl_gene   = gene_id,
      db_gene_symbol = symbol,
      transcript_id  = transcript_id,   # most abundant isoform, control RNA
      is_mane_select = is_mane_select,  # flag only; the MANE ID itself is not exported
      n_transcripts  = n_transcripts,
      dominant_tx_rna_fraction = dominant_tx_rna_fraction,
      log2FoldChange = log2FoldChange,
      padj           = padj,
      baseMean       = baseMean,
      class          = class_export
    ) %>%
    dplyr::arrange(dplyr::desc(abs(log2FoldChange)))
}

# Rank within each class by effect size and keep the top N. "Most resistant" is the
# largest positive log2FC; "most sensitive" the most negative. Ties break on padj so
# the cut is deterministic.
make_set <- function(which_class, n_keep = N_PER_CLASS) {
  d <- export_u %>%
    dplyr::filter(class_export == which_class, !is.na(symbol), symbol != "#N/A") %>%
    dplyr::distinct(gene_id, .keep_all = TRUE) %>%
    dplyr::arrange(dplyr::desc(abs(log2FoldChange)), padj)
  n_avail <- nrow(d)
  if (n_avail < n_keep) {
    cat("NOTE: only", n_avail, which_class, "genes available, fewer than the requested",
        n_keep, "- keeping all of them\n")
  }
  d %>% dplyr::slice_head(n = min(n_keep, n_avail)) %>% export_cols()
}

resistant_set <- make_set("resistant")
sensitive_set <- make_set("sensitive")

# The final list: top N resistant + top N sensitive, no unchanged genes.
final_set <- dplyr::bind_rows(resistant_set, sensitive_set) %>%
  dplyr::arrange(factor(class, levels = c("resistant", "sensitive")),
                 dplyr::desc(abs(log2FoldChange)))

cat("\nresistant kept:", nrow(resistant_set),
    " (|log2FC| range", sprintf("%.3f", min(abs(resistant_set$log2FoldChange))),
    "-", sprintf("%.3f", max(abs(resistant_set$log2FoldChange))), ")\n")
cat("sensitive kept:", nrow(sensitive_set),
    " (|log2FC| range", sprintf("%.3f", min(abs(sensitive_set$log2FoldChange))),
    "-", sprintf("%.3f", max(abs(sensitive_set$log2FoldChange))), ")\n")

stopifnot("resistant set contains a non-positive LFC" = all(resistant_set$log2FoldChange > 0))
stopifnot("sensitive set contains a non-negative LFC" = all(sensitive_set$log2FoldChange < 0))
stopifnot("CDKN2B missing from the resistant set" = "CDKN2B" %in% resistant_set$db_gene_symbol)
stopifnot("GLO1 missing from the sensitive set"   = "GLO1"   %in% sensitive_set$db_gene_symbol)
stopifnot("duplicate gene in resistant set" = !any(duplicated(resistant_set$ensembl_gene)))
stopifnot("duplicate gene in sensitive set" = !any(duplicated(sensitive_set$ensembl_gene)))
stopifnot("final list still contains unchanged genes" =
            !any(final_set$class == "unchanged"))
stopifnot("final list contains a class other than resistant/sensitive" =
            all(final_set$class %in% c("resistant", "sensitive")))
stopifnot("final list is not the two sets combined" =
            nrow(final_set) == nrow(resistant_set) + nrow(sensitive_set))
stopifnot("a class exceeds the requested size" =
            nrow(resistant_set) <= N_PER_CLASS && nrow(sensitive_set) <= N_PER_CLASS)
stopifnot("every exported row must name a transcript" =
            !any(is.na(final_set$transcript_id)))

lfc_tag <- sub("\\.", "p", as.character(EXPORT_LFC))

if (WRITE_GENESETS) {

write_csv(resistant_set,
          here("output", "genesets", paste0("teleman_4e_resistant_TE_lfc", lfc_tag, ".csv")))
write_csv(sensitive_set,
          here("output", "genesets", paste0("teleman_4e_sensitive_TE_lfc", lfc_tag, ".csv")))
write_csv(final_set,
          here("output", "genesets", paste0("teleman_4e_resistant_sensitive_TE_lfc", lfc_tag, ".csv")))

# All tested genes, unthresholded. Not the deliverable, but notebook 77 needs the full
# distribution for the Fig 4E correlation, and it keeps the unchanged class available.
write_csv(export_u %>% dplyr::select(gene_id, symbol, transcript_id, is_mane_select,
                                     n_transcripts, dominant_tx_rna_fraction, baseMean,
                                     log2FoldChange, lfcSE, pvalue, padj,
                                     class_lfc1 = class, class_lfc_export = class_export),
          here("output", "teleman_riboseq_te_all_genes.csv"))

cat("\nresistant genes written:", nrow(resistant_set), "\n")
cat("sensitive genes written:", nrow(sensitive_set), "\n")
cat("final combined list:     ", nrow(final_set), "\n")
cat("all-genes table:         ", nrow(export_u), "\n")

}
```

## The final table

Both sets in full, ranked by effect size. Paged so the whole list is browsable without
flooding the document.

```{r print-final-resistant}
cat("TOP", nrow(resistant_set), "MOST RESISTANT (log2FC > ", EXPORT_LFC, ")\n")
rmarkdown::paged_table(resistant_set)
```

```{r print-final-sensitive}
cat("TOP", nrow(sensitive_set), "MOST SENSITIVE (log2FC < -", EXPORT_LFC, ")\n")
rmarkdown::paged_table(sensitive_set)
```



---

# Summary

**Direction is correct and validated, in both the original and the corrected run.** The
interaction coefficient `treatmentBP_inhib.typeribo` is oriented exactly as Fig 1A's x
axis: positive = resistant to eIF4E inhibition. CDKN2B is in the top 0.3% of resistant
genes and GLO1 in the bottom 3.4%, matching the paper's two named extremes; every gene
labelled in Fig 1A falls on its published side, and the housekeeping cluster sits at the
origin and non-significant.

**Counts at `|log2FC| > 1`, `padj < 0.05`:**

| | original | corrected | corrected, matched universe | Fig 1A |
|---|---|---|---|---|
| Resistant (up) | 822 | 796 | **191** | 183 |
| Sensitive (down) | 716 | 664 | 317 | 422 |
| Unchanged | 10,552 | 9,846 | 3,850 | 3,753 |
| Total tested | 12,078 | 11,306 | 4,358 | 4,358 |

**What the fixes changed.** Effect sizes are untouched (Pearson r = 0.9999, median
|Δlog2FC| = 0.016) and no gene flips sign of significance. What changed is set
membership: 83 significant calls were removed, **74 of them alternate-haplotype duplicate
IDs of a gene that is still in the set** (chiefly MHC-region genes such as IER3, PPP1R10
and HSPA1A, which appear under six or seven ENSEMBL IDs each), 59 of which had been in the resistant
set at fold changes up to |log2FC| ~ 4.8. Re-estimating dispersions on the cleaner
universe then moved 62 borderline genes across the cut (34 into resistant, 28 out of
sensitive). The corrected resistant set also reproduces the paper more closely on a
matched universe (191 vs 183, against the original's 175).

**What the corrected run changes, relative to the original.** Three things: the
annotation join no longer duplicates pseudoautosomal genes, the protein-coding filter is
actually applied, and the interaction coefficient is selected by name rather than by
position. The metadata/count-matrix alignment that the original relied on implicitly is
now asserted.

**Left deliberately unchanged**, so that every difference above is attributable to those
three corrections: the design (no `rep` term), the `log2(rowSums) > 10` expression cut,
and `cooksCutoff` / `independentFiltering`. The first and last are quantified in the
sensitivity section; the expression cut is the main remaining reason the tested universe
is wider than the paper's.

**Annotation provenance.** The paper's methods state that reads were mapped to ENSEMBL
transcript assembly **94**, whereas the annotation joined here is **GENCODE v26**
(= Ensembl 88). That mismatch is worth knowing about, but it is not what produces the
symbol-less rows: those are alternate-haplotype IDs, which are absent from *any*
primary-assembly annotation regardless of build. Swapping v26 for the provenance-matched
GENCODE v29 (= Ensembl 94, verified from its own file header) moves the tested universe
by only 14 genes, because both are primary-assembly annotations and the haplotype
collapsing is what actually matters. The gene symbols in the exported tables come from
this annotation, so a handful of names follow v26 conventions rather than current ones;
join on `gene_id` where that matters.
