---
title: "Teleman eIF4E-IP: corrected re-run and Fig 4E validation"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  mpra_repo: "/Users/katematlin/github/mpra_preliminary"
  te_corrected_csv: "output/teleman_riboseq_te_all_genes.csv"
  lfc_cut: 1
  padj_cut: 0.05
  refit: FALSE
  write_outputs: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Companion to notebook 76. That notebook corrected the Teleman **ribosome profiling** TE
analysis; this one applies the same three corrections to the **eIF4E-IP** arm of the same
GEO submission (GSE243708), which was fit by the same upstream code and carries the same
defects.

The three corrections, identical to nb76:

1. Strip the ENSEMBL version **before** de-duplicating the symbol lookup, so `_PAR_Y`
   entries do not collapse onto their X-linked twin and duplicate the annotation join.
2. Actually apply the protein-coding filter. Upstream it was computed into `pc_keep` and
   never referenced, and it was built on versioned IDs while the count matrix rownames
   were version-stripped, so it would have matched nothing had it been used.
3. Select the DESeq2 coefficient by name rather than by positional index.

**Three contrasts are refit**, matching the three the upstream code produced:

| Contrast | Design | Meaning |
|---|---|---|
| Vehicle-only | `~type` on E.V. samples | baseline eIF4E binding (IP vs input) |
| Inhibitor-only | `~type` on 4E-BP1-4A samples | eIF4E binding under inhibition |
| Interaction | `~treatment + type + treatment:type` | **change** in eIF4E binding |

**Validation anchor.** The paper reports (Fig 4E) a negative correlation between the
change in translation efficiency on eIF4E inhibition and eIF4E binding, where the binding
axis is `log2((4E-BP1-4A: IP/input) / (E.V.: IP/input))` — the interaction contrast. That
correlation is asserted below against notebook 76's corrected TE values.

**Note on which binding axis.** Baseline binding and the *change* in binding correlate
with TE change in **opposite** directions (+0.24 and -0.25 respectively). Only the change
in binding reproduces Fig 4E. Do not substitute one for the other.

| Item | Location |
|---|---|
| Raw counts | `mpra_preliminary/data/GSE243708_raw_counts_matrix_Roiuk.csv` |
| Original DESeq2 code | `mpra_preliminary/code/eif4e_ip_clip_analysis.Rmd` |
| Original results | `mpra_preliminary/res_eif4e_ip_{vehicle,inhibitor_only,inhib_vs_vehicle}.csv` |
| Corrected TE (nb76) | `output/teleman_riboseq_te_all_genes.csv` (all tested genes, unthresholded) |
| Corrected IP (here) | `output/teleman_eif4e_ip_{vehicle,inhibitor,interaction}_corrected.csv` |

## Load libraries

```{r load-libraries}
library(tidyverse)
library(here)
library(DESeq2)
library(janitor)
```

## Parameters

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
# REFIT = TRUE forces the three DESeq2 models to refit; otherwise the cached RDS is
# reloaded if present.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    mpra_repo        = "/Users/katematlin/github/mpra_preliminary",
    te_corrected_csv = "output/teleman_riboseq_te_all_genes.csv",
    lfc_cut          = 1,
    padj_cut         = 0.05,
    refit            = FALSE,
    write_outputs    = TRUE
  )
}

MPRA_REPO        <- params$mpra_repo
TE_CORRECTED_CSV <- params$te_corrected_csv
LFC_CUT          <- params$lfc_cut
PADJ_CUT         <- params$padj_cut
REFIT            <- params$refit
WRITE_OUTPUTS    <- params$write_outputs

IP_RDS <- here("output", "predictive_modeling", "teleman_eif4e_ip_fixed.rds")

classify_ip <- function(lfc, padj, l = LFC_CUT, p = PADJ_CUT) {
  dplyr::case_when(
    lfc >  l & padj < p ~ "up",
    lfc < -l & padj < p ~ "down",
    TRUE                ~ "unchanged"
  )
}

cat("thresholds: |log2FC| >", LFC_CUT, " padj <", PADJ_CUT, "\n")
```

## The eIF4E-IP design is unbalanced

Worth knowing before reading any of the numbers: the IP arm has 11 libraries, not 12.
There are three replicates everywhere except the 4E-BP1-4A input, which has two.

```{r design-note}
ip_cols <- c(
  "eIF4E_IPed_RNA_4EBP1_4A_replica1", "eIF4E_IPed_RNA_4EBP1_4A_replica2",
  "eIF4E_IPed_RNA_4EBP1_4A_replica3",
  "eIF4E_IPed_RNA_Empty_vector_replica1", "eIF4E_IPed_RNA_Empty_vector_replica2",
  "eIF4E_IPed_RNA_Empty_vector_replica3",
  "Total_RNA_eIF4E_IP_4EBP1_4A_replica1", "Total_RNA_eIF4E_IP_4EBP1_4A_replica2",
  "Total_RNA_eIF4E_IP_Empty_vector_replica1", "Total_RNA_eIF4E_IP_Empty_vector_replica2",
  "Total_RNA_eIF4E_IP_Empty_vector_replica3"
)
cat("IP-arm libraries:", length(ip_cols), "\n")
```

---

# Part 1: Corrected re-run

```{r ip-refit}
fit_teleman_ip <- function() {
  raw <- read_csv(file.path(MPRA_REPO, "data/GSE243708_raw_counts_matrix_Roiuk.csv"),
                  show_col_types = FALSE)
  pcr_art <- read_csv(file.path(MPRA_REPO, "data/pcr_artifacts_sup8.csv"),
                      show_col_types = FALSE)

  geneInfo <- read_csv(file.path(MPRA_REPO, "accessories/gencode.v26.primary.info.csv.gz"),
                       col_names = FALSE, show_col_types = FALSE) %>%
    dplyr::filter(!str_detect(X2, "pre_"))
  colnames(geneInfo) <- c("gene_id", "transcript_id", "biotype", "symbol")
  geneInfo <- geneInfo %>% dplyr::mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  # correction 1: strip version BEFORE de-duplicating
  gene_symbol <- geneInfo %>%
    dplyr::select(gene_id = gene_id_clean, symbol) %>%
    dplyr::distinct(gene_id, .keep_all = TRUE)
  stopifnot("symbol map has duplicate keys" = !any(duplicated(gene_symbol$gene_id)))

  # correction 2: protein-coding set on version-stripped IDs, and actually applied
  pc <- geneInfo %>% dplyr::filter(biotype == "protein_coding") %>%
    dplyr::pull(gene_id_clean) %>% unique()

  nonpolyA <- readxl::read_xlsx(file.path(MPRA_REPO, "accessories/polyAminus.xlsx"),
                                sheet = 1) %>% dplyr::pull(Gene)
  nonpolyA_ID <- geneInfo %>% dplyr::filter(symbol %in% nonpolyA) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  blacklist_smRNA <- c("snoRNA", "miRNA", "snRNA", "rRNA", "Mt_tRNA", "scaRNA", "Mt_rRNA")
  blacklist_ID <- geneInfo %>% dplyr::filter(biotype %in% blacklist_smRNA) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  genes_to_kill <- unique(c(blacklist_ID, nonpolyA_ID))

  ip_data <- raw %>%
    dplyr::filter(!`Transcript stable ID` %in% pcr_art$Transcript_id) %>%
    dplyr::select(`Gene stable ID`, all_of(ip_cols))

  ip_deseq <- ip_data %>%
    dplyr::group_by(`Gene stable ID`) %>%
    dplyr::summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop") %>%
    clean_names() %>%
    column_to_rownames("gene_stable_id")

  keepGenes <- rownames(ip_deseq)[log2(rowSums(ip_deseq)) > 10]
  universe_orig  <- setdiff(keepGenes, genes_to_kill)
  universe_fixed <- intersect(universe_orig, pc)

  meta <- tibble::tibble(sample_name = colnames(ip_deseq)) %>%
    dplyr::mutate(
      type      = factor(if_else(str_detect(sample_name, "^total_rna"), "input", "IP"),
                         levels = c("input", "IP")),
      treatment = factor(if_else(str_detect(sample_name, "4ebp1_4a"), "BP_inhib", "vehicle"),
                         levels = c("vehicle", "BP_inhib")),
      rep       = factor(str_remove(str_extract(sample_name, "replica[0-9]"), "replica"),
                         levels = c("1", "2", "3"))
    ) %>%
    as.data.frame()
  rownames(meta) <- meta$sample_name

  stopifnot("metadata rows do not match count matrix columns" =
              identical(rownames(meta), colnames(ip_deseq)))
  stopifnot("expected 6 IP and 5 input libraries" =
              all(table(meta$type)[c("input", "IP")] == c(5, 6)))

  run_de <- function(samples, design_formula, coef_regex) {
    m <- droplevels(meta[samples, , drop = FALSE])
    dds <- DESeqDataSetFromMatrix(ip_deseq[universe_fixed, samples, drop = FALSE],
                                  colData = m, design = design_formula)
    dds <- DESeq(dds, quiet = TRUE)
    # correction 3: coefficient by name, asserted unique
    cn <- str_subset(resultsNames(dds), coef_regex)
    stopifnot("coefficient not found or not unique" = length(cn) == 1)
    res <- results(dds, name = cn, cooksCutoff = FALSE, independentFiltering = FALSE)
    lfcShrink(dds, coef = cn, res = res, type = "apeglm", quiet = TRUE) %>%
      as.data.frame() %>%
      rownames_to_column("gene_id") %>%
      dplyr::left_join(gene_symbol, by = "gene_id")
  }

  all_s <- rownames(meta)
  veh_s <- rownames(meta)[meta$treatment == "vehicle"]
  inh_s <- rownames(meta)[meta$treatment == "BP_inhib"]

  list(
    ip_vehicle = run_de(veh_s, ~type, "^type_IP_vs_input$"),
    ip_inhib   = run_de(inh_s, ~type, "^type_IP_vs_input$"),
    ip_inter   = run_de(all_s, ~treatment + type + treatment:type,
                        "^treatmentBP_inhib\\.typeIP$"),
    meta = meta,
    n_universe_orig  = length(universe_orig),
    n_universe_fixed = length(universe_fixed)
  )
}

if (REFIT || !file.exists(IP_RDS)) {
  ip_fit <- fit_teleman_ip()
  saveRDS(ip_fit, IP_RDS)
} else {
  ip_fit <- readRDS(IP_RDS)
  cat("loaded cached fit from", basename(IP_RDS), "\n")
}

print(ip_fit$meta, row.names = FALSE)
print(table(treatment = ip_fit$meta$treatment, type = ip_fit$meta$type))

cat("\nuniverse, original (no protein-coding filter):", ip_fit$n_universe_orig, "\n")
cat("universe, corrected (protein-coding only):   ", ip_fit$n_universe_fixed, "\n")
```

## Integrity checks

```{r ip-integrity}
ip_fit[c("ip_vehicle", "ip_inhib", "ip_inter")] %>%
  purrr::iwalk(function(d, nm) {
    stopifnot("duplicate gene_id"  = !any(duplicated(d$gene_id)))
    stopifnot("missing symbol"     = !any(is.na(d$symbol)))
    stopifnot("non-finite log2FC"  = all(is.finite(d$log2FoldChange[!is.na(d$log2FoldChange)])))
    stopifnot("padj outside [0,1]" = all(d$padj >= 0 & d$padj <= 1, na.rm = TRUE))
    stopifnot("negative baseMean"  = all(d$baseMean >= 0, na.rm = TRUE))
    cat(sprintf("%-11s n=%5d  dup=%d  NA symbol=%d\n", nm, nrow(d),
                sum(duplicated(d$gene_id)), sum(is.na(d$symbol))))
  })
```

---

# Part 2: Fig 4E validation

The published anchor: TE change on eIF4E inhibition should be **negatively** correlated
with the change in eIF4E binding.

```{r fig4e}
te <- read_csv(here(TE_CORRECTED_CSV), show_col_types = FALSE)

fig4e <- te %>%
  dplyr::select(gene_id, symbol, te_lfc = log2FoldChange) %>%
  dplyr::inner_join(ip_fit$ip_vehicle %>% dplyr::select(gene_id, bind_vehicle = log2FoldChange),
                    by = "gene_id") %>%
  dplyr::inner_join(ip_fit$ip_inhib %>% dplyr::select(gene_id, bind_inhib = log2FoldChange),
                    by = "gene_id") %>%
  dplyr::inner_join(ip_fit$ip_inter %>% dplyr::select(gene_id, delta_bind = log2FoldChange),
                    by = "gene_id")

cat("genes with both TE and IP measurements:", nrow(fig4e), "\n\n")

purrr::map_dfr(c("delta_bind", "bind_vehicle", "bind_inhib"), \(v) tibble::tibble(
  binding_axis = v,
  pearson      = round(cor(fig4e$te_lfc, fig4e[[v]]), 3),
  spearman     = round(cor(fig4e$te_lfc, fig4e[[v]], method = "spearman"), 3)
)) %>% as.data.frame() %>% print(row.names = FALSE)

# The published anchor.
stopifnot("Fig 4E not reproduced: TE change vs change in eIF4E binding is not negative" =
            cor(fig4e$te_lfc, fig4e$delta_bind, method = "spearman") < -0.15)

# Categorical form of the same statement: TE-resistant mRNAs should lose eIF4E binding
# relative to TE-sensitive ones.
by_class <- fig4e %>%
  dplyr::mutate(te_class = dplyr::case_when(
    te_lfc >  LFC_CUT ~ "TE resistant",
    te_lfc < -LFC_CUT ~ "TE sensitive",
    TRUE              ~ "TE unchanged")) %>%
  dplyr::group_by(te_class) %>%
  dplyr::summarise(n = dplyr::n(),
                   mean_delta_bind  = mean(delta_bind),
                   mean_bind_vehicle = mean(bind_vehicle), .groups = "drop")
print(as.data.frame(by_class), digits = 3)

stopifnot("TE-resistant mRNAs do not lose eIF4E binding relative to TE-sensitive ones" =
            by_class$mean_delta_bind[by_class$te_class == "TE resistant"] <
            by_class$mean_delta_bind[by_class$te_class == "TE sensitive"])

cat("\nFig 4E anchor reproduced\n")
```

**Baseline binding runs the other way, and that is not an error.** TE-resistant mRNAs are
*more* eIF4E-bound at baseline (mean +0.46) yet *lose* eIF4E occupancy under inhibition,
while TE-sensitive mRNAs start less bound and gain occupancy. Only the change in binding
reproduces Fig 4E; the baseline contrast is a different quantity and is confounded with
transcript abundance.

```{r fig4e-plot, fig.width = 5.5, fig.height = 5.5}
p_fig4e <- ggplot(fig4e, aes(x = delta_bind, y = te_lfc)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_point(size = 0.4, alpha = 0.25, color = "grey30") +
  geom_smooth(method = "lm", se = FALSE, color = "#0072B2", linewidth = 0.8) +
  labs(
    title = "Fig 4E: TE change vs change in eIF4E binding",
    x = "log2 change in eIF4E binding (IP/input, 4E-BP1-4A vs E.V.)",
    y = "log2 TE change (4E-BP1-4A vs E.V.)"
  ) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.4, size = 3.2,
           label = paste0("Pearson R = ", round(cor(fig4e$te_lfc, fig4e$delta_bind), 3),
                          ", n = ", nrow(fig4e))) +
  theme_classic(base_size = 12)

print(p_fig4e)
ggsave(here("plots", "77_teleman_fig4e_te_vs_eif4e_binding.pdf"),
       p_fig4e, width = 5.5, height = 5.5)
```

---

# Part 3: Original vs corrected

```{r ip-compare}
orig_files <- c(vehicle     = "res_eif4e_ip_vehicle.csv",
                inhibitor   = "res_eif4e_ip_inhibitor_only.csv",
                interaction = "res_eif4e_ip_inhib_vs_vehicle.csv")
fixed_names <- c(vehicle = "ip_vehicle", inhibitor = "ip_inhib", interaction = "ip_inter")

compare_tbl <- purrr::map_dfr(names(orig_files), function(k) {
  o <- read_csv(file.path(MPRA_REPO, orig_files[[k]]), show_col_types = FALSE)
  f <- ip_fit[[fixed_names[[k]]]]
  j <- dplyr::inner_join(
    o %>% dplyr::distinct(gene_id, .keep_all = TRUE) %>%
      dplyr::select(gene_id, lfc_o = log2FoldChange),
    f %>% dplyr::select(gene_id, lfc_f = log2FoldChange), by = "gene_id")
  tibble::tibble(
    contrast       = k,
    n_orig         = nrow(o),
    n_fixed        = nrow(f),
    dup_orig       = sum(duplicated(o$gene_id)),
    na_symbol_orig = sum(is.na(o$symbol)),
    up_orig        = sum(classify_ip(o$log2FoldChange, o$padj) == "up",   na.rm = TRUE),
    up_fixed       = sum(classify_ip(f$log2FoldChange, f$padj) == "up",   na.rm = TRUE),
    down_orig      = sum(classify_ip(o$log2FoldChange, o$padj) == "down", na.rm = TRUE),
    down_fixed     = sum(classify_ip(f$log2FoldChange, f$padj) == "down", na.rm = TRUE),
    lfc_r          = round(cor(j$lfc_o, j$lfc_f), 5)
  )
})
print(as.data.frame(compare_tbl))
```

Same pattern as nb76: effect sizes essentially unmoved, set sizes down because the
protein-coding filter collapses alternate-haplotype duplicate gene IDs (chiefly MHC
region) that the original was testing and calling once per haplotype copy. GENCODE v26
is a primary-assembly annotation and contains only the reference copy of each gene, so
intersecting with it keeps exactly one ID per gene. See nb76 for the full accounting.

---

# Part 4: Export

```{r ip-export}
if (WRITE_OUTPUTS) {

out_map <- c(ip_vehicle = "teleman_eif4e_ip_vehicle_corrected.csv",
             ip_inhib   = "teleman_eif4e_ip_inhibitor_corrected.csv",
             ip_inter   = "teleman_eif4e_ip_interaction_corrected.csv")

purrr::iwalk(out_map, function(fname, nm) {
  d <- ip_fit[[nm]] %>%
    dplyr::mutate(class = classify_ip(log2FoldChange, padj)) %>%
    dplyr::select(gene_id, symbol, baseMean, log2FoldChange, lfcSE, pvalue, padj, class)
  write_csv(d, here("output", fname))
  cat("wrote", fname, "-", nrow(d), "genes\n")
})

}
```

---

# Summary

The eIF4E-IP arm carried the same three defects as the riboseq arm and is now corrected
the same way. Across all three contrasts: no duplicate gene IDs, no symbol-less genes,
coefficients selected by name, sample alignment asserted.

**Fig 4E reproduces**, and slightly more strongly after correction: Spearman -0.290
between TE change and change in eIF4E binding, against -0.287 on the original files.

**The two binding axes point opposite ways.** Change in binding correlates -0.25 with TE
change (the published relationship); baseline binding correlates +0.24. Any downstream
analysis using "eIF4E binding" must say which one it means. The interaction contrast is
the one that matches the paper.

**Still unchanged, as in nb76:** the design carries no `rep` term, the `log2(rowSums) > 10`
expression cut is inherited as-is, and `cooksCutoff` / `independentFiltering` are left
off. The IP arm is also unbalanced (5 input vs 6 IP libraries, because the 4E-BP1-4A input
has only two replicates), which is handled correctly by DESeq2 but is worth knowing when
reading the inhibitor-only contrast.
