---
title: "Feature matrix for Teleman 4E-resistance (HeLa dominant transcripts)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  teleman_all_csv: "output/teleman_riboseq_te_all_genes.csv"
  teleman_final_csv: "output/genesets/teleman_4e_resistant_sensitive_TE_lfc0p5.csv"
  structure_dir: ""
  write_outputs: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Builds a feature matrix for the **Teleman 4E-resistance** classification target, keyed on
the **HeLa dominant transcript** selected in notebook 76, so that notebook 07 can be run
with `feature_set = "teleman_4e"`.

## Why this cannot reuse `feature_matrix_external_stability.rds`

That matrix is built on **MDA-MB-231** dominant transcripts; the Teleman ribosome
profiling is **HeLa** (Roiuk et al., Methods). Measured against notebook 76's output:

| | |
|---|---|
| gene overlap with the existing matrix | 83.9% |
| **transcript** overlap | **43.5%** |
| of shared genes, same isoform chosen | **51.7%** |

For roughly half the genes the existing row describes a *different isoform* than the one
HeLa actually translates, so every length, GC, codon and structure value would be for the
wrong molecule. Sequence features are therefore recomputed here from the v49 TxDb on the
HeLa isoform. 98.0% of those transcripts are present in that TxDb.

## Circularity

The target is Teleman 4E resistance, so every eIF4E-derived feature is circular and is
excluded by construction — this matrix never contains one. That includes
`te_4e_resistance_lfc`, which *is* the target variable, and the `is_4e_*` flags.

Notebook 07's `exclude_4e_features` flag alone is **not** sufficient: its regex
(`"^4e_resistance|^eif4e_ip"`) misses `te_4e_resistance_lfc`, `is_4e_resistant` and
`is_4e_sensitive`. That is patched in 07 as part of this work, but the primary defence is
that those columns are never built here. An assertion at the end enforces it.

MDA-MB-231 TE columns (`te_*`, `sig_*`, `delta_te_*`, `indiv_te_*`) are also absent —
wrong cell line, and confounded with the target through the shared TE axis.

## Class design (read before interpreting any model)

Positives are the 1,195 most resistant genes, negatives the 1,195 most sensitive. Both
classes are defined from **the same variable** (`log2FC` of TE) at opposite tails, so a
model trained here learns the *TE axis*, not resistance against a neutral background.
Every SHAP direction is relative to the sensitive set, not to a typical mRNA. This is the
same caveat notebook 75 documents for the RNA vs TE comparison. A resistant-vs-unchanged
run is the complement if that framing is wanted.

## Structure features

Not included in this build. RNAplfold/RNAfold must run on the cluster against the HeLa
isoform set; this notebook writes the FASTA and the transcript mapping those jobs need,
and folds the results back in via a chunk-gated section once `structure_dir` is set.

```{r libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
library(iCodon)   # must be attached, not namespace-qualified: predict_stability()
                  # resolves an internal `codones` object only when on the search path
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
# structure_dir: leave "" until the cluster RNAplfold/RNAfold jobs have run; point it at
# the directory holding their output to enable the structure section.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    teleman_all_csv   = "output/teleman_riboseq_te_all_genes.csv",
    teleman_final_csv = "output/genesets/teleman_4e_resistant_sensitive_TE_lfc0p5.csv",
    structure_dir     = "",
    write_outputs     = TRUE
  )
}

TELEMAN_ALL   <- params$teleman_all_csv
TELEMAN_FINAL <- params$teleman_final_csv
STRUCTURE_DIR <- params$structure_dir
WRITE_OUTPUTS <- params$write_outputs

OUT_RDS <- here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds")
TX_RDS  <- here("output", "predictive_modeling", "precomputed_teleman_tx.rds")

# MANE Select transcripts (one per gene), extracted from the GENCODE v49 GTF by nb76.
# Used only as a fallback when a gene's dominant transcript is absent from the v49 TxDb.
MANE_CSV <- here("accessories", "human", "mane_select_v49.csv")
stopifnot("MANE table not found - build it in notebook 76" = file.exists(MANE_CSV))

set.seed(9)
```

# 1. Transcript set

```{r transcript-set}
teleman <- read_csv(here(TELEMAN_ALL), show_col_types = FALSE)

txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

txdb_tx <- AnnotationDbi::keys(txdb, keytype = "TXNAME")
txdb_clean <- str_remove(txdb_tx, "\\..*")

# The dominant transcript comes from Ensembl 94 (the counts annotation); a few percent
# of those IDs were retired by the time of v49, and dropping them loses whole genes -
# including PHLDA1 and BCL2L12, both labelled resistant in Fig 1A. Where the dominant
# transcript is gone, fall back to the gene's MANE Select transcript, and record which
# rule was used so the substitution is never invisible.
mane_fallback <- read_csv(MANE_CSV,
                          col_names = c("gene_id", "mane_transcript_id", "mane_symbol"),
                          show_col_types = FALSE) %>%
  dplyr::select(gene_id, mane_transcript_id)

tx_set <- teleman %>%
  dplyr::transmute(
    transcript_id_clean = transcript_id,
    gene_id_clean       = gene_id,
    symbol              = symbol,
    teleman_te_lfc      = log2FoldChange,   # target source; dropped before saving
    teleman_te_padj     = padj,
    teleman_class       = class_lfc_export
  ) %>%
  dplyr::left_join(mane_fallback, by = c("gene_id_clean" = "gene_id")) %>%
  dplyr::mutate(
    dominant_ok = transcript_id_clean %in% txdb_clean,
    mane_ok     = !is.na(mane_transcript_id) & mane_transcript_id %in% txdb_clean,
    tx_source   = dplyr::case_when(dominant_ok ~ "dominant",
                                   mane_ok     ~ "mane_fallback",
                                   TRUE        ~ "unavailable"),
    transcript_id_clean = dplyr::if_else(dominant_ok, transcript_id_clean,
                                         mane_transcript_id)
  ) %>%
  dplyr::filter(tx_source != "unavailable") %>%
  dplyr::select(-dominant_ok, -mane_ok, -mane_transcript_id)

cat("genes in the Teleman tested set:", nrow(teleman), "\n")
print(dplyr::count(tx_set, tx_source) %>% as.data.frame())
cat("retained overall:", nrow(tx_set),
    sprintf(" (%.1f%%)\n", 100 * nrow(tx_set) / nrow(teleman)))
cat("dropped (no usable transcript in v49):", nrow(teleman) - nrow(tx_set), "\n")
stopifnot("transcript set is not unique" = !any(duplicated(tx_set$transcript_id_clean)))
stopifnot("gene set is not unique"       = !any(duplicated(tx_set$gene_id_clean)))

if (WRITE_OUTPUTS) {
  saveRDS(tx_set %>% dplyr::select(transcript_id_clean, gene_id_clean, symbol), TX_RDS)
  cat("wrote transcript mapping for the cluster jobs ->", basename(TX_RDS), "\n")
}
```

# 2. Region sequences and lengths

```{r regions}
utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cds  <- cdsBy(txdb, by = "tx", use.names = TRUE)

region_width <- function(grl, label) {
  w <- sum(width(grl))
  tibble::tibble(transcript_id_clean = str_remove(names(w), "\\..*"),
                 !!label := as.integer(w)) %>%
    dplyr::filter(transcript_id_clean %in% tx_set$transcript_id_clean) %>%
    dplyr::distinct(transcript_id_clean, .keep_all = TRUE)
}

lengths_tbl <- region_width(utr5, "utr5_length") %>%
  dplyr::full_join(region_width(cds,  "cds_length"),  by = "transcript_id_clean") %>%
  dplyr::full_join(region_width(utr3, "utr3_length"), by = "transcript_id_clean")

fm <- tx_set %>%
  dplyr::left_join(lengths_tbl, by = "transcript_id_clean") %>%
  dplyr::mutate(
    transcript_length = rowSums(cbind(utr5_length, cds_length, utr3_length), na.rm = TRUE),
    log2_utr5_length  = log2(utr5_length + 1),
    log2_cds_length   = log2(cds_length  + 1),
    log2_utr3_length  = log2(utr3_length + 1)
  )

cat("with CDS length:", sum(!is.na(fm$cds_length)), "of", nrow(fm), "\n")
cat("with 5'UTR:", sum(!is.na(fm$utr5_length)), " with 3'UTR:", sum(!is.na(fm$utr3_length)), "\n")

stopifnot("negative length" =
  all(c(fm$utr5_length, fm$cds_length, fm$utr3_length) >= 0, na.rm = TRUE))
```

# 3. GC content

```{r gc}
seq_of <- function(grl) {
  s <- extractTranscriptSeqs(genome, grl)
  names(s) <- str_remove(names(s), "\\..*")
  s[names(s) %in% tx_set$transcript_id_clean]
}
cds_seqs  <- seq_of(cds)
utr5_seqs <- seq_of(utr5)
utr3_seqs <- seq_of(utr3)
cat("CDS sequences:", length(cds_seqs), " 5'UTR:", length(utr5_seqs), " 3'UTR:", length(utr3_seqs), "\n")

gc_of <- function(seqs, label) {
  tibble::tibble(transcript_id_clean = names(seqs),
                 !!label := as.numeric(letterFrequency(seqs, "GC", as.prob = TRUE) * 100))
}

# Positional GC: GC1/GC2/GC3 across codon positions of the CDS.
cds_ok <- cds_seqs[width(cds_seqs) >= 3 & width(cds_seqs) %% 3 == 0]
pos_gc <- function(seqs, offset) {
  idx <- lapply(width(seqs), \(w) seq(offset, w, by = 3))
  as.numeric(vapply(seq_along(seqs), function(i) {
    b <- as.character(seqs[[i]])[idx[[i]]]
    mean(b %in% c("G", "C")) * 100
  }, numeric(1)))
}
cds_char_split <- lapply(as.character(cds_ok), \(s) strsplit(s, "")[[1]])
gc_pos <- function(offset) {
  vapply(cds_char_split, function(b) mean(b[seq(offset, length(b), by = 3)] %in% c("G", "C")) * 100,
         numeric(1))
}
positional_gc <- tibble::tibble(
  transcript_id_clean = names(cds_ok),
  gc1 = gc_pos(1), gc2 = gc_pos(2), gc3 = gc_pos(3)
)

fm <- fm %>%
  dplyr::left_join(gc_of(cds_seqs,  "cds_gc"),  by = "transcript_id_clean") %>%
  dplyr::left_join(gc_of(utr5_seqs, "utr5_gc"), by = "transcript_id_clean") %>%
  dplyr::left_join(gc_of(utr3_seqs, "utr3_gc"), by = "transcript_id_clean") %>%
  dplyr::left_join(positional_gc, by = "transcript_id_clean")

# gc3 orthogonalised against cds_gc, matching notebook 01v.
ok <- !is.na(fm$gc3) & !is.na(fm$cds_gc)
fm$gc3_residual <- NA_real_
fm$gc3_residual[ok] <- residuals(lm(gc3 ~ cds_gc, data = fm[ok, ]))

for (col in c("cds_gc", "utr5_gc", "utr3_gc", "gc1", "gc2", "gc3")) {
  stopifnot("GC outside [0,100]" = all(fm[[col]] >= 0 & fm[[col]] <= 100, na.rm = TRUE))
}

# Biological spot-checks (CLAUDE.md convention)
actb_gc <- fm$cds_gc[fm$symbol == "ACTB"]
cat("ACTB CDS GC%:", round(actb_gc, 1), "(expected 50-70)\n")
stopifnot("ACTB CDS GC outside 50-70%" =
  length(actb_gc) == 1 && between(actb_gc, 50, 70))
cat("GC3 median:", round(median(fm$gc3, na.rm = TRUE), 1), "(expected 40-65)\n")
stopifnot("GC3 median outside 40-65%" =
  between(median(fm$gc3, na.rm = TRUE), 40, 65))
```

# 4. Codon optimality

```{r codon-optimality}
# --- CSC via iCodon (Diez et al. 2022); no external file needed ---
predict_human <- iCodon::predict_stability("human")
cds_char  <- as.character(cds_ok)
uniq      <- unique(cds_char)
uniq_sc   <- predict_human(uniq)
names(uniq_sc) <- uniq
csc_tbl <- tibble::tibble(transcript_id_clean = names(cds_ok),
                          csc = as.numeric(uniq_sc[cds_char]))

# --- CAI / tAI / FOP. tAI follows notebook 01g exactly (dos Reis 2004 eukaryotic
#     wobble rules over GtRNAdb anticodon gene copy numbers). Ported rather than
#     re-derived, and validated against the existing matrix at the end of this chunk. ---
cts          <- codonTable(cds_ok)
counts_mat   <- codonCounts(cts)
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_of_codon  <- setNames(as.character(gen_code), names(gen_code))

stop_codons  <- names(gen_code[gen_code == "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])

# --- CAI (Sharp & Li 1987) with a ribosomal-protein reference set, as in 01g ---
ribo_subset <- str_detect(tx_set$symbol[match(names(cds_ok), tx_set$transcript_id_clean)],
                          "^RPL|^RPS")
ribo_subset[is.na(ribo_subset)] <- FALSE
cat("ribosomal-protein transcripts in the CAI reference set:", sum(ribo_subset), "\n")
stopifnot("CAI reference set is too small" = sum(ribo_subset) >= 50)

ref_counts <- colSums(counts_mat[ribo_subset, , drop = FALSE])
w <- setNames(rep(NA_real_, length(gen_code)), names(gen_code))
w[stop_codons] <- 0
for (aa in names(aa_to_codons)) {
  codons <- aa_to_codons[[aa]]
  cnt <- ref_counts[codons]; total <- sum(cnt)
  if (total == 0) { w[codons] <- 1; next }
  rscu  <- cnt / (total / length(codons))
  max_r <- max(rscu)
  w[codons] <- if (max_r == 0) 0 else pmin(rscu / max_r, 1)
}

compute_cai_gene <- function(gene_counts, w_vals) {
  use   <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use <- gene_counts[use]; n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_vals[use])) / n_total)
}
cai_vals <- apply(counts_mat, 1, compute_cai_gene, w_vals = w)

trna_gcn <- read_csv(here("accessories", "human", "human_trna_gcn.csv"),
                     comment = "#", show_col_types = FALSE)
cat("tRNA GCN table:", nrow(trna_gcn), "anticodons\n")

# anticodon 5'->3' DNA: X1 = wobble position 34. Codon Y1:X3, Y2:X2, Y3 wobbles with X1.
decode_anticodon <- function(ac) {
  x1 <- substr(ac, 1, 1); x2 <- substr(ac, 2, 2); x3 <- substr(ac, 3, 3)
  wc <- c(A = "T", T = "A", G = "C", C = "G")
  y1 <- unname(wc[x3]); y2 <- unname(wc[x2])
  if (x1 == "A") {        y3 <- c("T", "C", "A"); svs <- c(0, 0, 0)      # A -> inosine
  } else if (x1 == "G") { y3 <- c("C", "T");      svs <- c(0, 0.41)      # G:U wobble
  } else if (x1 == "C") { y3 <- "G";              svs <- 0
  } else {                y3 <- "A";              svs <- 0 }
  data.frame(codon = paste0(y1, y2, y3), s = svs, stringsAsFactors = FALSE)
}

ws_tai <- setNames(rep(0, length(gen_code)), names(gen_code))
for (i in seq_len(nrow(trna_gcn))) {
  decoded <- decode_anticodon(trna_gcn$anticodon[i])
  for (j in seq_len(nrow(decoded))) {
    cd <- decoded$codon[j]
    if (!is.na(aa_of_codon[cd]) && aa_of_codon[cd] == trna_gcn$amino_acid[i]) {
      ws_tai[cd] <- ws_tai[cd] + (1 - decoded$s[j]) * trna_gcn$gcn[i]
    }
  }
}
w_tai <- ws_tai / max(ws_tai[sense_codons])
cat("w_tai range over sense codons:", round(range(w_tai[sense_codons]), 4), "\n")

compute_tai_gene <- function(gene_counts, w_vals) {
  use   <- sense_codons[!is.na(w_vals[sense_codons]) & w_vals[sense_codons] > 0]
  c_use <- gene_counts[use]; n_total <- sum(c_use)
  if (n_total == 0) return(NA_real_)
  exp(sum(c_use * log(w_vals[use])) / n_total)
}
tai_vals <- apply(counts_mat, 1, compute_tai_gene, w_vals = w_tai)

# FOP: fraction of codons drawn from the top tertile of w_tai among sense codons.
opt_codons <- sense_codons[w_tai[sense_codons] >=
                             quantile(w_tai[sense_codons], 2/3, na.rm = TRUE)]
fop_vals <- rowSums(counts_mat[, opt_codons, drop = FALSE]) /
            pmax(rowSums(counts_mat[, sense_codons, drop = FALSE]), 1)

stopifnot("CAI/tAI/FOP length mismatch with the transcript set" =
  length(cai_vals) == length(cds_ok) && length(tai_vals) == length(cds_ok) &&
  length(fop_vals) == length(cds_ok))

codon_tbl <- tibble::tibble(transcript_id_clean = names(cds_ok),
                            cai = as.numeric(cai_vals), tai = as.numeric(tai_vals),
                            fop = as.numeric(fop_vals)) %>%
  dplyr::left_join(csc_tbl, by = "transcript_id_clean")

# Validation against the existing MDA-MB-231 matrix, on transcripts common to both.
#
# tAI weights depend only on the tRNA gene copy number table, not on which transcripts
# were quantified, so for a shared transcript tAI must match EXACTLY. That is a real test
# of the ported wobble rules, which no range check could catch.
#
# CAI cannot match exactly and is not asserted to: its weights come from the RSCU of a
# ribosomal-protein REFERENCE SET, and that reference is drawn from this notebook's
# transcript set (HeLa isoforms) rather than 01g's (MDA-MB-231 isoforms). Using the
# matched reference is the correct choice for internal consistency here; the two should
# still be highly correlated, which is what is checked.
mda <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds")) %>%
  dplyr::select(transcript_id_clean, tai_mda = tai, cai_mda = cai)
chk <- codon_tbl %>% dplyr::inner_join(mda, by = "transcript_id_clean") %>%
  dplyr::filter(!is.na(tai), !is.na(tai_mda), !is.na(cai), !is.na(cai_mda))
cat("shared transcripts for validation:", nrow(chk), "\n")
cat("  max |delta tAI| (must be ~0):", format(max(abs(chk$tai - chk$tai_mda)), digits = 3), "\n")
cat("  CAI correlation vs 01g:", round(cor(chk$cai, chk$cai_mda), 4),
    " max |delta|:", format(max(abs(chk$cai - chk$cai_mda)), digits = 3), "\n")

stopifnot("ported tAI does not reproduce the existing matrix - check the wobble rules" =
  max(abs(chk$tai - chk$tai_mda)) < 1e-6)
stopifnot("CAI is not strongly correlated with the existing matrix" =
  cor(chk$cai, chk$cai_mda) > 0.95)

# --- positional CSC: first 75 codons and CDS quarters (mirrors notebook 01j) ---
# Score an arbitrary codon window. Windows are computed per transcript and any that
# collapse to fewer than one codon are dropped, so short CDSs cannot invert the range.
score_window <- function(seqs, start_codon, end_codon) {
  st <- (start_codon - 1) * 3 + 1
  en <- pmin(end_codon * 3, width(seqs))
  keep <- en - st + 1 >= 3
  s <- subseq(seqs[keep], start = st[keep], end = en[keep])
  s <- subseq(s, 1, width(s) - width(s) %% 3)
  ch <- as.character(s); u <- unique(ch); sc <- predict_human(u); names(sc) <- u
  tibble::tibble(transcript_id_clean = names(s), val = as.numeric(sc[ch]))
}

n_cod <- floor(width(cds_ok) / 3)

# First 75 codons (or the whole CDS if shorter).
csc_first75 <- score_window(cds_ok, rep(1L, length(cds_ok)), pmin(75L, n_cod)) %>%
  dplyr::rename(csc_first75 = val)

# Quarters, only for CDSs long enough for four non-empty quarters.
long_enough <- n_cod >= 8
cds_q <- cds_ok[long_enough]
n_cod_q <- n_cod[long_enough]
cat("CDSs long enough for quartile CSC (>= 8 codons):", length(cds_q),
    "of", length(cds_ok), "\n")

for (i in 1:4) {
  st <- floor((i - 1) / 4 * n_cod_q) + 1
  en <- floor(i / 4 * n_cod_q)
  csc_tbl_i <- score_window(cds_q, st, en) %>%
    dplyr::rename(!!paste0("csc_q", i) := val)
  codon_tbl <- codon_tbl %>% dplyr::left_join(csc_tbl_i, by = "transcript_id_clean")
}
codon_tbl <- codon_tbl %>% dplyr::left_join(csc_first75, by = "transcript_id_clean")

fm <- fm %>% dplyr::left_join(codon_tbl, by = "transcript_id_clean")

stopifnot("CAI outside [0,1]" = all(fm$cai >= 0 & fm$cai <= 1, na.rm = TRUE))
stopifnot("tAI outside [0,1]" = all(fm$tai >= 0 & fm$tai <= 1, na.rm = TRUE))
stopifnot("FOP outside [0,1]" = all(fm$fop >= 0 & fm$fop <= 1, na.rm = TRUE))

# Biological spot-check: ribosomal proteins should have above-median CAI
rp <- str_detect(fm$symbol, "^RP[LS][0-9]")
cat("ribosomal-protein median CAI:", round(median(fm$cai[rp], na.rm = TRUE), 4),
    " vs all:", round(median(fm$cai, na.rm = TRUE), 4), "\n")
stopifnot("ribosomal proteins do not have above-median CAI" =
  median(fm$cai[rp], na.rm = TRUE) > median(fm$cai, na.rm = TRUE))
```

# 5. Kozak context

```{r kozak}
# -6..+4 around the AUG, from the CDS start on the transcript.
tx_seqs_all <- extractTranscriptSeqs(genome, exonsBy(txdb, by = "tx", use.names = TRUE))
names(tx_seqs_all) <- str_remove(names(tx_seqs_all), "\\..*")

utr5_len <- setNames(fm$utr5_length, fm$transcript_id_clean)
kz_ids <- fm$transcript_id_clean[!is.na(fm$utr5_length) & fm$utr5_length >= 6 &
                                 fm$transcript_id_clean %in% names(tx_seqs_all)]
kz_seqs <- tx_seqs_all[kz_ids]
starts  <- utr5_len[kz_ids] - 5
kz_ctx  <- subseq(kz_seqs, start = starts, end = pmin(starts + 9, width(kz_seqs)))
kz_ctx  <- kz_ctx[width(kz_ctx) == 10]

# Kozak 1987 frequency matrix over the 10-nt window, rows A/C/G/T.
kozak_pwm <- matrix(
  c(0.22,0.20,0.26,0.24,0.26,0.23,1,0,0,0.23,
    0.28,0.31,0.22,0.28,0.34,0.36,0,0,0,0.16,
    0.32,0.28,0.35,0.30,0.24,0.24,0,0,1,0.47,
    0.18,0.21,0.17,0.18,0.16,0.17,0,1,0,0.14),
  nrow = 4, byrow = TRUE, dimnames = list(c("A","C","G","T"), NULL))
log_odds <- log2((kozak_pwm + 0.01) / 0.25)

score_kozak <- function(s) {
  b <- strsplit(as.character(s), "")[[1]]
  if (length(b) != 10 || any(!b %in% rownames(log_odds))) return(NA_real_)
  sum(vapply(seq_len(10), \(i) log_odds[b[i], i], numeric(1)))
}
kozak_tbl <- tibble::tibble(
  transcript_id_clean = names(kz_ctx),
  kozak_pwm_score = vapply(kz_ctx, score_kozak, numeric(1)),
  kozak_minus3 = substr(as.character(kz_ctx), 4, 4),
  kozak_plus4  = substr(as.character(kz_ctx), 10, 10)
) %>%
  dplyr::mutate(
    kozak_optimal = as.integer(kozak_minus3 %in% c("A","G") & kozak_plus4 == "G"),
    kozak_score   = as.integer(kozak_minus3 %in% c("A","G")) + as.integer(kozak_plus4 == "G")
  ) %>%
  dplyr::select(-kozak_minus3, -kozak_plus4)

fm <- fm %>% dplyr::left_join(kozak_tbl, by = "transcript_id_clean")

cat("Kozak scored:", sum(!is.na(fm$kozak_pwm_score)), "of", nrow(fm), "\n")

# Biological spot-check: mean PWM score must increase across kozak_score tiers 0..2
tier_means <- fm %>% dplyr::filter(!is.na(kozak_score)) %>%
  dplyr::group_by(kozak_score) %>%
  dplyr::summarise(m = mean(kozak_pwm_score, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(kozak_score)
print(as.data.frame(tier_means), digits = 4)
stopifnot("Kozak PWM score is not monotonic across tiers" =
  all(diff(tier_means$m) > 0))
```

# 6. External stability data (joined by gene)

These are published measurements from other cell lines, joined at gene level. That is
the same assumption the MDA-MB-231 matrix makes; it is no better and no worse for HeLa,
but it is an assumption and not an isoform-specific measurement.

```{r external-stability}
cnot3 <- readRDS(here("output", "predictive_modeling", "cnot3_slamseq_halflives.rds")) %>%
  dplyr::transmute(gene_id_clean = str_remove(Gene_ID, "\\..*"),
                   zhu2024_ctrl_halflife = sgNT_halflife,
                   zhu2024_cnot3ko_halflife_logfc = log2_ratio) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

dhx29 <- readRDS(here("output", "predictive_modeling", "dhx29_slamseq_halflives.rds")) %>%
  dplyr::transmute(gene_id_clean = str_remove(ensembl_gene, "\\..*"),
                   hia2026_ctrl_halflife = Ctrl_Half_life,
                   hia2026_dhx29_halflife_logfc = logFC) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

dhx29_occ <- readRDS(here("output", "predictive_modeling", "dhx29_riboseq_occupancy.rds")) %>%
  dplyr::transmute(gene_id_clean = str_remove(ensembl_gene, "\\..*"),
                   hia2026_dhx29_occupancy = dhx29_occupancy_score) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

karner <- readRDS(here("output", "predictive_modeling", "goodarzi_231_ct_ratios.rds")) %>%
  dplyr::transmute(gene_id_clean = str_remove(ensembl_gene, "\\..*"),
                   karner_ct_ratio = log2_ct) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

fm <- fm %>%
  dplyr::left_join(cnot3,     by = "gene_id_clean") %>%
  dplyr::left_join(dhx29,     by = "gene_id_clean") %>%
  dplyr::left_join(dhx29_occ, by = "gene_id_clean") %>%
  dplyr::left_join(karner,    by = "gene_id_clean")

for (nm in c("zhu2024_ctrl_halflife", "hia2026_ctrl_halflife")) {
  stopifnot("half-life must be positive" = all(fm[[nm]] > 0, na.rm = TRUE))
}
stopifnot("occupancy must be finite" =
  all(is.finite(fm$hia2026_dhx29_occupancy[!is.na(fm$hia2026_dhx29_occupancy)])))

fm %>%
  dplyr::summarise(across(c(zhu2024_ctrl_halflife, zhu2024_cnot3ko_halflife_logfc,
                            hia2026_ctrl_halflife, hia2026_dhx29_halflife_logfc,
                            hia2026_dhx29_occupancy, karner_ct_ratio),
                          ~ sum(!is.na(.x)))) %>%
  tidyr::pivot_longer(everything(), names_to = "feature", values_to = "n_non_missing") %>%
  as.data.frame() %>% print(row.names = FALSE)

# Zhu 2024 reports a median sgNT half-life of ~3.5h (Fig 2B)
cat("\nZhu2024 median control half-life:",
    round(median(fm$zhu2024_ctrl_halflife, na.rm = TRUE), 2), "h\n")
stopifnot("Zhu2024 median half-life outside expected range" =
  between(median(fm$zhu2024_ctrl_halflife, na.rm = TRUE), 2.5, 5.0))
```

# 7. Structure features (chunk-gated)

Enabled once the cluster jobs have run and `structure_dir` points at their output.
See `cluster_scripts/viennarna/README_teleman.md`.

```{r structure}
if (nzchar(STRUCTURE_DIR)) {

struct_files <- list.files(STRUCTURE_DIR, pattern = "\\.rds$", full.names = TRUE)
stopifnot("no structure RDS found in structure_dir" = length(struct_files) > 0)
for (f in struct_files) {
  s <- readRDS(f)
  stopifnot("structure table must be keyed on transcript_id_clean" =
              "transcript_id_clean" %in% names(s))
  fm <- fm %>% dplyr::left_join(s, by = "transcript_id_clean")
  cat("joined", basename(f), "-", ncol(s) - 1, "features\n")
}
acc <- grep("^struct_accessibility", names(fm), value = TRUE)
mfe <- grep("mfe", names(fm), value = TRUE)
for (a in acc) stopifnot("accessibility outside [0,1]" = all(fm[[a]] >= 0 & fm[[a]] <= 1, na.rm = TRUE))
for (m in mfe) stopifnot("MFE must be <= 0" = all(fm[[m]] <= 0, na.rm = TRUE))

} else {
  cat("structure_dir is empty - skipping structure features.\n")
  cat("Run the cluster jobs first (cluster_scripts/viennarna/0[123]c_*_teleman.*),\n")
  cat("then re-knit with structure_dir set to their output directory.\n")
}
```

# 8. Leakage guard and export

```{r export}
# Nothing eIF4E-derived, and no MDA-MB-231 TE, may reach the model.
banned <- grep("4e|4E|^te_|^sig_|^delta_te|^indiv_te", names(fm), value = TRUE)
banned <- setdiff(banned, c("teleman_te_lfc", "teleman_te_padj", "teleman_class"))
stopifnot("a circular eIF4E or MDA TE feature reached the matrix" = length(banned) == 0)

# The target columns MUST NOT be saved into the matrix. Notebook 07 builds its feature
# list as setdiff(colnames, id_cols) where id_cols only matches gene_id/transcript_id/
# symbol - so any other column becomes a feature, including the target. Leaving
# teleman_te_lfc / teleman_class in the RDS leaks the answer and yields AUC = 1.0.
target_cols <- c("teleman_te_lfc", "teleman_te_padj", "teleman_class")
# tx_source is provenance, not biology: it is a character column and 07 would otherwise
# treat it as a predictor.
meta_cols <- c("tx_source")
fm_out <- fm %>% dplyr::select(-dplyr::any_of(c(target_cols, meta_cols)))

feature_cols <- setdiff(names(fm_out),
                        c("transcript_id_clean", "gene_id_clean", "symbol"))
stopifnot("provenance column leaked into the features" =
            !any(meta_cols %in% feature_cols))

cat("rows:", nrow(fm_out), "  feature columns:", length(feature_cols), "\n\n")
cat("features:\n"); print(feature_cols)

cat("\nmissingness (worst 8):\n")
tibble::tibble(feature = feature_cols,
               pct_na = round(100 * vapply(fm_out[feature_cols], \(x) mean(is.na(x)), numeric(1)), 1)) %>%
  dplyr::arrange(dplyr::desc(pct_na)) %>% head(8) %>% as.data.frame() %>% print(row.names = FALSE)

stopifnot("matrix key must be unique" = !any(duplicated(fm_out$transcript_id_clean)))

# The decisive guard: nothing target-derived may survive into the saved matrix.
stopifnot("a target column reached the saved matrix - this leaks and gives AUC 1.0" =
            !any(target_cols %in% names(fm_out)))
stopifnot("no teleman_* column may be saved" =
            length(grep("^teleman_", names(fm_out))) == 0)

if (WRITE_OUTPUTS) {
  saveRDS(fm_out, OUT_RDS)
  cat("\nwrote", basename(OUT_RDS), "with", length(feature_cols), "features",
      "(target columns dropped)\n")
}
```

# 9. Gene sets for notebook 07

```{r genesets}
final <- read_csv(here(TELEMAN_FINAL), show_col_types = FALSE)

pos <- final %>% dplyr::filter(class == "resistant") %>%
  dplyr::transmute(ensembl_gene, db_gene_symbol, log2FoldChange, padj)
neg <- final %>% dplyr::filter(class == "sensitive") %>%
  dplyr::transmute(ensembl_gene, db_gene_symbol, log2FoldChange, padj)

# Only genes that actually have features are usable.
pos_in <- pos %>% dplyr::filter(ensembl_gene %in% fm$gene_id_clean)
neg_in <- neg %>% dplyr::filter(ensembl_gene %in% fm$gene_id_clean)

cat("positives (resistant):", nrow(pos), "-> with features:", nrow(pos_in), "\n")
cat("negatives (sensitive):", nrow(neg), "-> with features:", nrow(neg_in), "\n")

stopifnot("positive and negative sets overlap" =
  length(intersect(pos_in$ensembl_gene, neg_in$ensembl_gene)) == 0)
stopifnot("negative pool smaller than positive set" = nrow(neg_in) >= nrow(pos_in) * 0.5)

# Notebook 07 builds the POSITIVE path from its params as
#   {condition}_{gene}_{direction}_TE_{timepoint}_lfc{lfc}.csv
# so the files must be named to match, or 07 will not find them. `timepoint` is
# meaningless for this dataset (a single timepoint), and "1hr" is used only to satisfy
# the naming convention.
POS_NAME <- "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
NEG_NAME <- "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"

if (WRITE_OUTPUTS) {
  write_csv(pos_in, here("output", "genesets", POS_NAME))
  write_csv(neg_in, here("output", "genesets", NEG_NAME))
  cat("wrote", POS_NAME, "and", NEG_NAME, "to output/genesets/\n")
}
```

# 10. FASTA export by class

Full-length mature transcript sequences (5'UTR + CDS + 3'UTR) for the two modelled
classes, written separately so they can be fed straight into motif or sequence-composition
tools. The sets are the ones notebook 07 actually trains on, so the FASTA and the model
see the same transcripts.

```{r fasta-by-class}
fasta_dir <- here("output", "fasta")
dir.create(fasta_dir, showWarnings = FALSE, recursive = TRUE)

# Full mature transcript sequences, keyed on version-stripped transcript ID.
tx_seqs_full <- extractTranscriptSeqs(genome, exonsBy(txdb, by = "tx", use.names = TRUE))
names(tx_seqs_full) <- str_remove(names(tx_seqs_full), "\\..*")

# Map each class's genes to their dominant transcript.
tx_of_gene <- setNames(fm$transcript_id_clean, fm$gene_id_clean)

write_class_fasta <- function(gene_ids, label) {
  tx <- unname(tx_of_gene[gene_ids])
  tx <- tx[!is.na(tx) & tx %in% names(tx_seqs_full)]
  seqs <- tx_seqs_full[tx]
  seqs <- seqs[width(seqs) > 0]
  out <- file.path(fasta_dir, paste0("teleman_4e_", label, "_transcripts.fa"))
  if (WRITE_OUTPUTS) writeXStringSet(seqs, out)
  tibble::tibble(class = label,
                 genes_in_set = length(gene_ids),
                 transcripts_written = length(seqs),
                 median_width = as.integer(median(width(seqs))),
                 file = basename(out))
}

fasta_summary <- dplyr::bind_rows(
  write_class_fasta(pos_in$ensembl_gene, "resistant"),
  write_class_fasta(neg_in$ensembl_gene, "sensitive")
)
print(as.data.frame(fasta_summary), row.names = FALSE)

cat("\nresistant transcripts:", fasta_summary$transcripts_written[1], "\n")
cat("sensitive transcripts:", fasta_summary$transcripts_written[2], "\n")

stopifnot("a class FASTA came out empty" = all(fasta_summary$transcripts_written > 0))
stopifnot("the two FASTA sets overlap" =
  length(intersect(unname(tx_of_gene[pos_in$ensembl_gene]),
                   unname(tx_of_gene[neg_in$ensembl_gene]))) == 0)
```

Headers are the bare `transcript_id_clean`, matching the region FASTAs written for the
cluster, so results join back on the same key. For 5'UTR-only or CDS-only versions, the
per-region sequences for the full set are already written by
`cluster_scripts/viennarna/01c_generate_fasta_teleman.R`; subset those by the same gene
lists.

# How to run notebook 07 on this

```r
rmarkdown::render(
  "code/predictive_modeling/07_rf_classification_eif3d_targets_no_boruta.Rmd",
  params = list(
    condition               = "teleman",
    gene                    = "4e",
    direction               = "resistant",
    timepoint               = "1hr",     # naming-convention filler; single timepoint
    lfc                     = "0.5",
    feature_set             = "teleman_4e",
    feature_matrix_path     = "feature_matrix_teleman_4e.rds",
    neg_geneset_csv         = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    exclude_4e_features     = TRUE,
    exclude_own_te_features = TRUE
  ))

Those `condition`/`gene`/`direction`/`timepoint`/`lfc` values are what make 07 resolve the
positive set to `teleman_4e_resistant_TE_1hr_lfc0.5.csv`; they are not analysis settings.
```

`feature_set = "teleman_4e"` is a new label, so it cannot overwrite an existing run's
outputs. Both exclusion flags are belt-and-braces: this matrix contains no 4E or MDA TE
columns to begin with, and section 8 asserts it.
