---
title: "GC content of 4E-resistant vs 4E-sensitive mRNAs"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_matrix: "feature_matrix_teleman_4e.rds"
  pos_csv: "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
  neg_csv: "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"
  rf_perf_csv: "rf_classification_performance_eif34e_resistant_teleman_1hr_lfc0.5_teleman_4e_clipaggregate_no_boruta_no4e_noOwnTE.csv"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Two things:

1. Density of **CDS GC%** and **GC3** for the 4E-resistant vs 4E-sensitive sets from
   notebook 76.
2. An explanation for why the notebook 07 classifier reaches AUC ~0.99 on these classes.
   The short version is in the plots: these two sets are almost non-overlapping on GC.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_matrix = "feature_matrix_teleman_4e.rds",
    pos_csv        = "teleman_4e_resistant_TE_1hr_lfc0.5.csv",
    neg_csv        = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    rf_perf_csv    = paste0("rf_classification_performance_eif34e_resistant_teleman_",
                            "1hr_lfc0.5_teleman_4e_clipaggregate_no_boruta_no4e_noOwnTE.csv"),
    save_plots     = TRUE
  )
}

# eIF3d orange and eIF3e blue are reserved for those two factors (CLAUDE.md), so this
# figure uses two of the remaining Okabe-Ito hues, matching the nb76 volcano.
class_colors <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2")
```

# Data

```{r load}
fm  <- readRDS(here("output", "predictive_modeling", params$feature_matrix))
pos <- read_csv(here("output", "genesets", params$pos_csv), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", params$neg_csv), show_col_types = FALSE)

stopifnot(
  "gene sets are empty"  = nrow(pos) > 0 && nrow(neg) > 0,
  "classes overlap"      = length(intersect(pos$ensembl_gene, neg$ensembl_gene)) == 0,
  # nb76 defines resistant as the positive-LFC tail; if this flips, the labels are wrong.
  "class direction flipped" = median(pos$log2FoldChange) > 0 &&
                              median(neg$log2FoldChange) < 0
)

d <- fm %>%
  mutate(class = case_when(
    gene_id_clean %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id_clean %in% neg$ensembl_gene ~ "4E-sensitive")) %>%
  filter(!is.na(class)) %>%
  mutate(class = factor(class, levels = names(class_colors)))

n_class <- dplyr::count(d, class)
print(as.data.frame(n_class), row.names = FALSE)
cat("genes in the CSVs but not in the feature matrix:",
    nrow(pos) + nrow(neg) - nrow(d), "\n")

stopifnot(
  "a class is missing from the feature matrix" = nrow(n_class) == 2 & all(n_class$n > 0),
  # gc3_residual is orthogonalized against cds_gc and is not a percentage, so it is
  # deliberately excluded here.
  "GC values outside [0, 100]" = d %>%
    dplyr::select(any_of(c("cds_gc", "utr5_gc", "utr3_gc", "gc1", "gc2", "gc3"))) %>%
    unlist() %>%
    between(0, 100) %>%
    all(na.rm = TRUE)
)
```

# Density plots

```{r density-fn}
gc_density <- function(df, col, xlab, title) {
  sub <- df %>% filter(!is.na(.data[[col]]))
  wt  <- wilcox.test(reformulate("class", col), data = sub)
  med <- sub %>%
    group_by(class) %>%
    summarise(m = median(.data[[col]]), n = n(), .groups = "drop")

  lab <- paste0(
    "Wilcoxon p = ", format.pval(wt$p.value, digits = 3, eps = 2.2e-16), "\n",
    paste(sprintf("median %s = %.1f (n = %d)", med$class, med$m, med$n),
          collapse = "\n"))

  ggplot(sub, aes(x = .data[[col]], fill = class, color = class)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    geom_vline(data = med, aes(xintercept = m, color = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = class_colors, name = NULL) +
    scale_color_manual(values = class_colors, name = NULL) +
    labs(title = title, x = xlab, y = "Density") +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
             size = 3.2, label = lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## CDS GC content

```{r cds-gc, fig.width = 7, fig.height = 4.5}
p_cds <- gc_density(d, "cds_gc", "CDS GC content (%)",
                    "CDS GC content, 4E-resistant vs 4E-sensitive")
print(p_cds)
if (params$save_plots) ggsave(here("plots", "79_cds_gc_density_4e.pdf"), p_cds,
                              width = 7, height = 4.5)
```

## GC3 (third codon position)

```{r gc3, fig.width = 7, fig.height = 4.5}
p_gc3 <- gc_density(d, "gc3", "GC3, third codon position (%)",
                    "GC3, 4E-resistant vs 4E-sensitive")
print(p_gc3)
if (params$save_plots) ggsave(here("plots", "79_gc3_density_4e.pdf"), p_gc3,
                              width = 7, height = 4.5)
```

## The other GC flavours, for context

```{r other-gc, fig.width = 9, fig.height = 6}
gc_feats <- c("cds_gc", "gc1", "gc2", "gc3", "utr5_gc", "utr3_gc")

gc_long <- d %>%
  dplyr::select(class, all_of(gc_feats)) %>%
  pivot_longer(-class, names_to = "feature", values_to = "gc",
               names_transform = list(feature = ~ factor(.x, levels = gc_feats))) %>%
  filter(!is.na(gc))

p_all <- ggplot(gc_long, aes(x = gc, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "GC content by region and codon position",
       subtitle = paste0("n = ", n_class$n[1], " resistant, ",
                         n_class$n[2], " sensitive"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_all)
if (params$save_plots) ggsave(here("plots", "79_gc_all_regions_density_4e.pdf"), p_all,
                              width = 9, height = 6)
```

Note the contrast between `gc3` and `utr5_gc`: the separation is a **coding-sequence**
property, concentrated at the wobble position. The 5'UTRs of the two sets are nearly
superimposable.

---

# Why the notebook 07 classifier reaches AUC ~0.99

The classifier is not finding something subtle. Each feature is scored here on its own,
as a single-variable classifier, by the rank-based AUC.

```{r single-feature-auc}
# Mann-Whitney AUC. max(a, 1 - a) makes it direction-agnostic, so this measures
# separability, not whether high values mark the resistant class.
auc_single <- function(x, cls, positive = "4E-resistant") {
  ok  <- !is.na(x)
  x   <- x[ok]
  cls <- cls[ok]
  n1  <- sum(cls == positive)
  n0  <- sum(cls != positive)
  a   <- (sum(rank(x)[cls == positive]) - n1 * (n1 + 1) / 2) / (n1 * n0)
  max(a, 1 - a)
}

feats <- c("cds_gc", "gc3", "gc1", "gc2", "utr3_gc", "utr5_gc", "cai", "tai", "fop",
           "csc", "hia2026_dhx29_occupancy", "log2_cds_length", "log2_utr5_length",
           "log2_utr3_length", "kozak_pwm_score")
feats <- intersect(feats, names(d))

auc_tbl <- d %>%
  dplyr::select(class, all_of(feats)) %>%
  pivot_longer(-class, names_to = "feature", values_to = "value") %>%
  group_by(feature) %>%
  summarise(
    auc_alone        = round(auc_single(value, class), 3),
    median_resistant = round(median(value[class == "4E-resistant"], na.rm = TRUE), 2),
    median_sensitive = round(median(value[class == "4E-sensitive"], na.rm = TRUE), 2),
    pct_na           = round(100 * mean(is.na(value)), 1),
    .groups = "drop") %>%
  arrange(desc(auc_alone))

print(as.data.frame(auc_tbl), row.names = FALSE)

auc_of <- function(f) auc_tbl$auc_alone[match(f, auc_tbl$feature)]
med_of <- function(f, cls) auc_tbl[[paste0("median_", cls)]][match(f, auc_tbl$feature)]
n_features <- ncol(dplyr::select(fm, where(is.numeric)))

# Read the forest's AUC from the notebook 07 run rather than hardcoding it: the
# test-set and CV-mean numbers differ and are easy to transpose by hand.
rf_perf <- read_csv(here("output", "predictive_modeling", params$rf_perf_csv),
                    show_col_types = FALSE) %>%
  filter(Metric == "AUC")
stopifnot("no AUC row in the notebook 07 performance file" = nrow(rf_perf) == 1)

rf_auc_test <- round(rf_perf$Test_Set, 3)
rf_auc_cv   <- round(rf_perf$CV_Mean, 3)
```

**`cds_gc` on its own reaches AUC `r auc_of("cds_gc")`.** The full random forest, with
all `r n_features` numeric features, reaches `r rf_auc_test` on the held-out set. The
model is essentially reading GC content and adding very little on top.

## The top features are not independent

```{r feature-correlation}
top <- auc_tbl$feature[auc_tbl$auc_alone > 0.85]
stopifnot("cds_gc is not among the top separating features" = "cds_gc" %in% top)

cm <- d %>%
  dplyr::select(all_of(top)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")
print(round(cm, 2))

cat("\nSpearman correlation with cds_gc:\n")
print(round(sort(abs(cm["cds_gc", setdiff(top, "cds_gc")]), decreasing = TRUE), 3))
```

Everything that separates the classes well is strongly correlated with CDS GC. Even
`hia2026_dhx29_occupancy`, which comes from an entirely different experiment and cell
line, tracks it. There is effectively **one axis** doing the work, not fifteen features.

## What this does and does not mean

**It is not leakage.** The target columns were removed from the feature matrix (notebook
78 asserts it), and no eIF4E-derived feature was ever built.

**The classes make the problem easy by construction.** They are the
`r nrow(pos)` most resistant and `r nrow(neg)` most sensitive genes: opposite tails of
one continuous variable, with a gap in between (resistant have log2FC >=
`r round(min(pos$log2FoldChange), 2)`, sensitive <=
`r round(max(neg$log2FoldChange), 2)`, nothing in the middle).
Separating two extremes is a much easier task than separating resistant mRNAs from
typical ones, and the AUC should not be read as "we can predict 4E resistance".

**The GC difference is large enough to deserve its own scrutiny.** A median CDS GC of
`r med_of("cds_gc", "resistant")`% versus `r med_of("cds_gc", "sensitive")`%, and GC3 of
`r med_of("gc3", "resistant")`% versus `r med_of("gc3", "sensitive")`%, is a very big
compositional split. Two readings, not distinguishable from this data alone:

- *Biological*: GC-rich, structured coding sequences are translated by a route less
  dependent on eIF4E, so they hold their TE when eIF4E is sequestered.
- *Technical*: GC affects library preparation efficiency, and a TE ratio is a ratio of
  two libraries. If footprint and total-RNA libraries have different GC-amplification
  characteristics, a GC-dependent artefact appears in `log2FC` directly. This dataset is
  known to carry PCR artefacts, which is why the authors ship a blocklist
  (Suppl. Data 8) rather than a UMI-deduplicated pipeline.

A useful discriminator would be whether the same GC split shows up in an eIF4E-inhibition
dataset generated with a different library protocol.

## Where the classifier sits relative to GC alone

```{r auc-comparison, fig.width = 6, fig.height = 4}
comp <- auc_tbl %>%
  slice_head(n = 8) %>%
  mutate(feature = fct_reorder(feature, auc_alone))

p_auc <- ggplot(comp, aes(x = auc_alone, y = feature)) +
  geom_col(fill = "grey70", width = 0.7) +
  geom_vline(xintercept = rf_auc_test, linetype = "dashed",
             color = "#D55E00", linewidth = 0.8) +
  annotate("text", x = rf_auc_test, y = 1,
           label = paste0(" full RF, test = ", rf_auc_test),
           hjust = 0, vjust = -0.5, size = 3.2, color = "#D55E00") +
  coord_cartesian(xlim = c(0.5, 1)) +
  labs(title = "Single features vs the full random forest",
       subtitle = paste0("n = ", nrow(d), " genes; single-feature AUCs are ",
                         "direction-agnostic. RF CV mean = ", rf_auc_cv),
       x = "AUC on its own", y = NULL) +
  theme_classic(base_size = 12)

print(p_auc)
if (params$save_plots) ggsave(here("plots", "79_single_feature_auc_vs_rf.pdf"), p_auc,
                              width = 6, height = 4)
```

# Summary

- CDS GC and GC3 separate the two classes almost completely; 5'UTR GC does not.
- `cds_gc` alone gives AUC `r auc_of("cds_gc")` against the full model's
  `r rf_auc_test` (test) / `r rf_auc_cv` (CV mean), so the classifier is
  reading one compositional axis.
- This is a property of the gene sets, not a defect in the model. A resistant-vs-unchanged
  comparison would be the more informative test of predictive power.
