---
title: "GC content of 4E-resistant vs 4E-sensitive mRNAs"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_matrix: "feature_matrix_teleman_4e.rds"
  pos_csv: "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
  neg_csv: "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Two things:

1. Density of **CDS GC%** and **GC3** for the 4E-resistant vs 4E-sensitive sets from
   notebook 76.
2. An explanation for why the notebook 07 classifier reaches AUC 0.99 on these classes.
   The short version is in the plots: these two sets are almost non-overlapping on GC.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_matrix = "feature_matrix_teleman_4e.rds",
    pos_csv        = "teleman_4e_resistant_TE_1hr_lfc0.5.csv",
    neg_csv        = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    save_plots     = TRUE
  )
}
FEATURE_MATRIX <- params$feature_matrix
POS_CSV        <- params$pos_csv
NEG_CSV        <- params$neg_csv
SAVE_PLOTS     <- params$save_plots

# eIF3d orange and eIF3e blue are reserved for those two factors (CLAUDE.md), so this
# figure uses two of the remaining Okabe-Ito hues, matching the nb76 volcano.
class_colors <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2")
```

# Data

```{r load}
fm  <- readRDS(here("output", "predictive_modeling", FEATURE_MATRIX))
pos <- read_csv(here("output", "genesets", POS_CSV), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", NEG_CSV), show_col_types = FALSE)

d <- fm %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id_clean %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id_clean %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE                                ~ NA_character_)) %>%
  dplyr::filter(!is.na(class)) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)))

print(dplyr::count(d, class) %>% as.data.frame(), row.names = FALSE)
stopifnot("classes overlap" =
  length(intersect(pos$ensembl_gene, neg$ensembl_gene)) == 0)
```

# Density plots

```{r density-fn}
gc_density <- function(df, col, xlab, title) {
  sub <- df %>% dplyr::filter(!is.na(.data[[col]]))
  wt  <- wilcox.test(sub[[col]] ~ sub$class)
  med <- sub %>% dplyr::group_by(class) %>%
    dplyr::summarise(m = median(.data[[col]]), .groups = "drop")

  lab <- paste0(
    "Wilcoxon p = ", format.pval(wt$p.value, digits = 3, eps = 2.2e-16),
    "\nmedian ", med$class[1], " = ", round(med$m[1], 1),
    "\nmedian ", med$class[2], " = ", round(med$m[2], 1),
    "\nn = ", sum(sub$class == med$class[1]), " / ", sum(sub$class == med$class[2]))

  ggplot(sub, aes(x = .data[[col]], fill = class, color = class)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    geom_vline(data = med, aes(xintercept = m, color = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = class_colors, name = NULL) +
    scale_color_manual(values = class_colors, name = NULL) +
    labs(title = title, x = xlab, y = "Density") +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
             size = 3.2, label = lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## CDS GC content

```{r cds-gc, fig.width = 7, fig.height = 4.5}
p_cds <- gc_density(d, "cds_gc", "CDS GC content (%)",
                    "CDS GC content, 4E-resistant vs 4E-sensitive")
print(p_cds)
if (SAVE_PLOTS) ggsave(here("plots", "79_cds_gc_density_4e.pdf"), p_cds,
                       width = 7, height = 4.5)
```

## GC3 (third codon position)

```{r gc3, fig.width = 7, fig.height = 4.5}
p_gc3 <- gc_density(d, "gc3", "GC3, third codon position (%)",
                    "GC3, 4E-resistant vs 4E-sensitive")
print(p_gc3)
if (SAVE_PLOTS) ggsave(here("plots", "79_gc3_density_4e.pdf"), p_gc3,
                       width = 7, height = 4.5)
```

## The other GC flavours, for context

```{r other-gc, fig.width = 9, fig.height = 6}
gc_long <- d %>%
  dplyr::select(class, cds_gc, gc1, gc2, gc3, utr5_gc, utr3_gc) %>%
  tidyr::pivot_longer(-class, names_to = "feature", values_to = "gc") %>%
  dplyr::filter(!is.na(gc)) %>%
  dplyr::mutate(feature = factor(feature,
    levels = c("cds_gc", "gc1", "gc2", "gc3", "utr5_gc", "utr3_gc")))

p_all <- ggplot(gc_long, aes(x = gc, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "GC content by region and codon position",
       subtitle = paste0("n = ", sum(d$class == "4E-resistant"), " resistant, ",
                         sum(d$class == "4E-sensitive"), " sensitive"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_all)
if (SAVE_PLOTS) ggsave(here("plots", "79_gc_all_regions_density_4e.pdf"), p_all,
                       width = 9, height = 6)
```

Note the contrast between `gc3` and `utr5_gc`: the separation is a **coding-sequence**
property, concentrated at the wobble position. The 5'UTRs of the two sets are nearly
superimposable.

---

# Why the notebook 07 classifier reaches AUC 0.99

The classifier is not finding something subtle. Each feature is scored here on its own,
as a single-variable classifier, by the rank-based AUC.

```{r single-feature-auc}
auc_single <- function(x, cls, positive = "4E-resistant") {
  ok <- !is.na(x); x <- x[ok]; cls <- cls[ok]
  r  <- rank(x)
  n1 <- sum(cls == positive); n0 <- sum(cls != positive)
  a  <- (sum(r[cls == positive]) - n1 * (n1 + 1) / 2) / (n1 * n0)
  max(a, 1 - a)   # direction-agnostic separability
}

feats <- c("cds_gc", "gc3", "gc1", "gc2", "utr3_gc", "utr5_gc", "cai", "tai", "fop",
           "csc", "hia2026_dhx29_occupancy", "log2_cds_length", "log2_utr5_length",
           "log2_utr3_length", "kozak_pwm_score")
feats <- intersect(feats, names(d))

auc_tbl <- purrr::map_dfr(feats, function(f) {
  tibble::tibble(
    feature = f,
    auc_alone = round(auc_single(d[[f]], d$class), 3),
    median_resistant = round(median(d[[f]][d$class == "4E-resistant"], na.rm = TRUE), 2),
    median_sensitive = round(median(d[[f]][d$class == "4E-sensitive"], na.rm = TRUE), 2),
    pct_na = round(100 * mean(is.na(d[[f]])), 1)
  )
}) %>% dplyr::arrange(dplyr::desc(auc_alone))

print(as.data.frame(auc_tbl), row.names = FALSE)
```

**`cds_gc` on its own reaches AUC `r round(auc_tbl$auc_alone[auc_tbl$feature == "cds_gc"], 3)`.**
The full random forest, with all `r length(setdiff(names(fm), c("transcript_id_clean","gene_id_clean","symbol")))`
features, reaches ~0.99. The model is essentially reading GC content and adding very
little on top.

## The top features are not independent

```{r feature-correlation}
top <- auc_tbl$feature[auc_tbl$auc_alone > 0.85]
cm  <- d %>% dplyr::select(dplyr::all_of(top)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")
print(round(cm, 2))

cat("\nSpearman correlation with cds_gc:\n")
cor_with_gc <- sort(abs(cm["cds_gc", setdiff(top, "cds_gc")]), decreasing = TRUE)
print(round(cor_with_gc, 3))
```

Everything that separates the classes well is strongly correlated with CDS GC. Even
`hia2026_dhx29_occupancy`, which comes from an entirely different experiment and cell
line, tracks it. There is effectively **one axis** doing the work, not fifteen features.

## What this does and does not mean

**It is not leakage.** The target columns were removed from the feature matrix (notebook
78 asserts it), and no eIF4E-derived feature was ever built.

**The classes make the problem easy by construction.** They are the 1,195 most resistant
and 1,195 most sensitive genes: opposite tails of one continuous variable, with a gap in
between (resistant have log2FC >= 0.65, sensitive <= -0.50, nothing in the middle).
Separating two extremes is a much easier task than separating resistant mRNAs from
typical ones, and the AUC should not be read as "we can predict 4E resistance".

**The GC difference is large enough to deserve its own scrutiny.** A median CDS GC of
`r round(auc_tbl$median_resistant[auc_tbl$feature == "cds_gc"], 1)`% versus
`r round(auc_tbl$median_sensitive[auc_tbl$feature == "cds_gc"], 1)`%, and GC3 of
`r round(auc_tbl$median_resistant[auc_tbl$feature == "gc3"], 1)`% versus
`r round(auc_tbl$median_sensitive[auc_tbl$feature == "gc3"], 1)`%, is a very big
compositional split. Two readings, not distinguishable from this data alone:

- *Biological*: GC-rich, structured coding sequences are translated by a route less
  dependent on eIF4E, so they hold their TE when eIF4E is sequestered.
- *Technical*: GC affects library preparation efficiency, and a TE ratio is a ratio of
  two libraries. If footprint and total-RNA libraries have different GC-amplification
  characteristics, a GC-dependent artefact appears in `log2FC` directly. This dataset is
  known to carry PCR artefacts, which is why the authors ship a blocklist
  (Suppl. Data 8) rather than a UMI-deduplicated pipeline.

A useful discriminator would be whether the same GC split shows up in an eIF4E-inhibition
dataset generated with a different library protocol.

## Where the classifier sits relative to GC alone

```{r auc-comparison, fig.width = 6, fig.height = 4}
rf_auc <- 0.992   # test-set AUC from the notebook 07 run on these sets

comp <- auc_tbl %>%
  dplyr::slice_head(n = 8) %>%
  dplyr::mutate(feature = forcats::fct_reorder(feature, auc_alone))

p_auc <- ggplot(comp, aes(x = auc_alone, y = feature)) +
  geom_col(fill = "grey70", width = 0.7) +
  geom_vline(xintercept = rf_auc, linetype = "dashed", color = "#D55E00", linewidth = 0.8) +
  annotate("text", x = rf_auc, y = 1, label = paste0(" full RF = ", rf_auc),
           hjust = 0, vjust = -0.5, size = 3.2, color = "#D55E00") +
  coord_cartesian(xlim = c(0.5, 1)) +
  labs(title = "Single features vs the full random forest",
       subtitle = paste0("n = ", nrow(d), " genes"),
       x = "AUC on its own", y = NULL) +
  theme_classic(base_size = 12)

print(p_auc)
if (SAVE_PLOTS) ggsave(here("plots", "79_single_feature_auc_vs_rf.pdf"), p_auc,
                       width = 6, height = 4)
```

# Summary

- CDS GC and GC3 separate the two classes almost completely; 5'UTR GC does not.
- `cds_gc` alone gives AUC ~0.98 against the full model's ~0.99, so the classifier is
  reading one compositional axis.
- This is a property of the gene sets, not a defect in the model. A resistant-vs-unchanged
  comparison would be the more informative test of predictive power.
