---
title: "Teleman RNA and RIBO separately, and feature densities by 4E class"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  mpra_repo: "/Users/katematlin/github/mpra_preliminary"
  feature_matrix: "feature_matrix_teleman_4e.rds"
  pos_csv: "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
  neg_csv: "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"
  refit: FALSE
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Two things.

**1. Decompose the TE contrast.** Notebooks 76 and 78 use the interaction coefficient,
`log2((ribo/rna in 4E-BP1-4A) / (ribo/rna in E.V.))`. That is a ratio of ratios, so a
systematic effect in *either* library propagates into it. Here the two arms are fit
separately as main effects:

| contrast | libraries | meaning |
|---|---|---|
| RIBO | the four `Riboseq_*` | footprint change on 4E-BP1-4A |
| RNA | the four `Total_RNA_Riboseq_*` | mRNA-level change on 4E-BP1-4A |

This matters because notebook 79 found that 4E-resistant and 4E-sensitive mRNAs differ by
~18 points of CDS GC and ~34 points of GC3, and a compositional split that large could
come from real translational biology **or** from GC-dependent library behaviour. Splitting
the arms says where the GC association actually lives.

**2. Feature densities.** Resistant vs sensitive for every feature built in notebook 78,
not just the GC ones.

```{r libraries}
library(tidyverse)
library(here)
library(DESeq2)
library(janitor)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    mpra_repo      = "/Users/katematlin/github/mpra_preliminary",
    feature_matrix = "feature_matrix_teleman_4e.rds",
    pos_csv        = "teleman_4e_resistant_TE_1hr_lfc0.5.csv",
    neg_csv        = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    refit          = FALSE,
    save_plots     = TRUE
  )
}
MPRA_REPO      <- params$mpra_repo
FEATURE_MATRIX <- params$feature_matrix
POS_CSV        <- params$pos_csv
NEG_CSV        <- params$neg_csv
REFIT          <- params$refit
SAVE_PLOTS     <- params$save_plots

ARMS_RDS <- here("output", "predictive_modeling", "teleman_rna_ribo_arms.rds")

# eIF3d orange / eIF3e blue stay reserved (CLAUDE.md); reuse the nb76/nb79 pair.
class_colors <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2")

RIBO_COLS <- c("Riboseq_Empty_vector_1", "Riboseq_Empty_vector_2",
               "Riboseq_4EBP1_4a_1", "Riboseq_4EBP1_4A_2")
RNA_COLS  <- c("Total_RNA_Riboseq_Empty_vector_1", "Total_RNA_Riboseq_Empty_vector_2",
               "Total_RNA_Riboseq_4EBP1_4A_1", "Total_RNA_Riboseq_4EBP1_4A_2")
```

# Part 1: RNA and RIBO fit separately

Same inputs and filters as notebook 76 - PCR-artifact removal, transcript-to-gene
summing, the `log2(rowSums) > 10` cut and the protein-coding filter - so the gene
universe is comparable. Only the design changes: `~treatment` within one assay type,
rather than the interaction across both.

```{r fit-arms}
fit_arms <- function() {
  raw <- read_csv(file.path(MPRA_REPO, "data/GSE243708_raw_counts_matrix_Roiuk.csv"),
                  show_col_types = FALSE)
  pcr_art <- read_csv(file.path(MPRA_REPO, "data/pcr_artifacts_sup8.csv"),
                      show_col_types = FALSE)

  geneInfo <- read_csv(file.path(MPRA_REPO, "accessories/gencode.v26.primary.info.csv.gz"),
                       col_names = FALSE, show_col_types = FALSE) %>%
    dplyr::filter(!str_detect(X2, "pre_"))
  colnames(geneInfo) <- c("gene_id", "transcript_id", "biotype", "symbol")
  geneInfo <- geneInfo %>% dplyr::mutate(gene_id_clean = str_remove(gene_id, "\\..*"))

  gene_symbol <- geneInfo %>%
    dplyr::select(gene_id = gene_id_clean, symbol) %>%
    dplyr::distinct(gene_id, .keep_all = TRUE)
  pc <- geneInfo %>% dplyr::filter(biotype == "protein_coding") %>%
    dplyr::pull(gene_id_clean) %>% unique()
  nonpolyA <- readxl::read_xlsx(file.path(MPRA_REPO, "accessories/polyAminus.xlsx"),
                                sheet = 1) %>% dplyr::pull(Gene)
  nonpolyA_ID <- geneInfo %>% dplyr::filter(symbol %in% nonpolyA) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  bl <- c("snoRNA", "miRNA", "snRNA", "rRNA", "Mt_tRNA", "scaRNA", "Mt_rRNA")
  bl_ID <- geneInfo %>% dplyr::filter(biotype %in% bl) %>%
    dplyr::pull(gene_id_clean) %>% unique()
  genes_to_kill <- unique(c(bl_ID, nonpolyA_ID))

  counts <- raw %>%
    dplyr::filter(!`Transcript stable ID` %in% pcr_art$Transcript_id) %>%
    dplyr::select(`Gene stable ID`, all_of(c(RIBO_COLS, RNA_COLS))) %>%
    dplyr::group_by(`Gene stable ID`) %>%
    dplyr::summarise(across(everything(), \(x) sum(x, na.rm = TRUE)), .groups = "drop") %>%
    clean_names() %>%
    column_to_rownames("gene_stable_id")

  keep <- rownames(counts)[log2(rowSums(counts)) > 10]
  universe <- intersect(setdiff(keep, genes_to_kill), pc)

  run_arm <- function(cols, label) {
    cn <- make_clean_names(cols)
    m <- data.frame(sample_name = cn,
                    treatment = factor(if_else(str_detect(cn, "4ebp"), "BP_inhib", "vehicle"),
                                       levels = c("vehicle", "BP_inhib")))
    rownames(m) <- cn
    stopifnot("column mismatch" = all(cn %in% colnames(counts)))
    dds <- DESeqDataSetFromMatrix(counts[universe, cn, drop = FALSE], m, ~treatment)
    dds <- DESeq(dds, quiet = TRUE)
    coef_name <- str_subset(resultsNames(dds), "^treatment_BP_inhib_vs_vehicle$")
    stopifnot("coefficient not found" = length(coef_name) == 1)
    res <- results(dds, name = coef_name, cooksCutoff = FALSE, independentFiltering = FALSE)
    lfcShrink(dds, coef = coef_name, res = res, type = "apeglm", quiet = TRUE) %>%
      as.data.frame() %>% rownames_to_column("gene_id") %>%
      dplyr::left_join(gene_symbol, by = "gene_id") %>%
      dplyr::rename_with(~ paste0(label, "_", .x), c("log2FoldChange", "padj", "baseMean"))
  }

  list(ribo = run_arm(RIBO_COLS, "ribo"),
       rna  = run_arm(RNA_COLS,  "rna"),
       n_universe = length(universe))
}

if (REFIT || !file.exists(ARMS_RDS)) {
  arms <- fit_arms(); saveRDS(arms, ARMS_RDS)
} else {
  arms <- readRDS(ARMS_RDS); cat("loaded cached arms from", basename(ARMS_RDS), "\n")
}
cat("universe:", arms$n_universe, "genes\n")
```

## Sanity check: the arms must reconstruct the TE contrast

`te_lfc` is the interaction, so it should be close to `ribo_lfc - rna_lfc`. They will not
be identical - each fit shrinks separately and the interaction model shares dispersion
across all eight libraries - but a weak correlation here would mean something is wrong.

```{r reconstruct-te}
te <- read_csv(here("output", "teleman_riboseq_te_all_genes.csv"), show_col_types = FALSE)

arms_j <- arms$ribo %>%
  dplyr::select(gene_id, ribo_log2FoldChange, ribo_padj, ribo_baseMean) %>%
  dplyr::inner_join(arms$rna %>% dplyr::select(gene_id, rna_log2FoldChange, rna_padj),
                    by = "gene_id") %>%
  dplyr::inner_join(te %>% dplyr::select(gene_id, symbol, te_lfc = log2FoldChange,
                                         te_padj = padj), by = "gene_id") %>%
  dplyr::mutate(ribo_minus_rna = ribo_log2FoldChange - rna_log2FoldChange)

cat("genes with all three:", nrow(arms_j), "\n")
cat("cor(te_lfc, ribo - rna):",
    round(cor(arms_j$te_lfc, arms_j$ribo_minus_rna, use = "complete.obs"), 4), "\n")
stopifnot("te_lfc is not reconstructed by ribo - rna" =
  cor(arms_j$te_lfc, arms_j$ribo_minus_rna, use = "complete.obs") > 0.7)

cat("\nglobal medians (4E-BP1-4A vs E.V.):\n")
cat("  ribo log2FC:", round(median(arms_j$ribo_log2FoldChange, na.rm = TRUE), 3), "\n")
cat("  rna  log2FC:", round(median(arms_j$rna_log2FoldChange,  na.rm = TRUE), 3), "\n")
cat("  te   log2FC:", round(median(arms_j$te_lfc, na.rm = TRUE), 3), "\n")
```

Ribosome profiling normalises away the global drop in translation, so the TE median sits
near zero by construction; the RIBO arm carries whatever global shift survives
size-factor normalisation.

## Where does the GC association live?

The decisive panel. If GC tracks the **RIBO** arm, the compositional split is a
translation phenomenon. If it tracks the **RNA** arm, it is at least partly an
mRNA-abundance or library-composition effect being carried into the ratio.

```{r gc-by-arm}
fm <- readRDS(here("output", "predictive_modeling", FEATURE_MATRIX))
pos <- read_csv(here("output", "genesets", POS_CSV), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", NEG_CSV), show_col_types = FALSE)

gc_arms <- arms_j %>%
  dplyr::inner_join(fm %>% dplyr::select(gene_id_clean, cds_gc, gc3),
                    by = c("gene_id" = "gene_id_clean")) %>%
  dplyr::filter(!is.na(cds_gc))

cor_tbl <- tibble::tibble(
  contrast = c("RIBO (footprints)", "RNA (mRNA level)", "TE (interaction)"),
  cor_cds_gc = c(cor(gc_arms$ribo_log2FoldChange, gc_arms$cds_gc, method = "spearman"),
                 cor(gc_arms$rna_log2FoldChange,  gc_arms$cds_gc, method = "spearman"),
                 cor(gc_arms$te_lfc,              gc_arms$cds_gc, method = "spearman")),
  cor_gc3    = c(cor(gc_arms$ribo_log2FoldChange, gc_arms$gc3, method = "spearman",
                     use = "complete.obs"),
                 cor(gc_arms$rna_log2FoldChange,  gc_arms$gc3, method = "spearman",
                     use = "complete.obs"),
                 cor(gc_arms$te_lfc,              gc_arms$gc3, method = "spearman",
                     use = "complete.obs"))
) %>% dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

cat("Spearman correlation with GC, n =", nrow(gc_arms), "\n")
print(as.data.frame(cor_tbl), row.names = FALSE)
```

```{r arm-medians-by-class}
cls <- gc_arms %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE ~ NA_character_)) %>%
  dplyr::filter(!is.na(class)) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)))

cls %>%
  dplyr::group_by(class) %>%
  dplyr::summarise(n = dplyr::n(),
                   median_ribo = round(median(ribo_log2FoldChange, na.rm = TRUE), 3),
                   median_rna  = round(median(rna_log2FoldChange,  na.rm = TRUE), 3),
                   median_te   = round(median(te_lfc, na.rm = TRUE), 3),
                   .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)
```

```{r arm-density, fig.width = 9, fig.height = 4}
arm_long <- cls %>%
  dplyr::select(class, RIBO = ribo_log2FoldChange, RNA = rna_log2FoldChange, TE = te_lfc) %>%
  tidyr::pivot_longer(-class, names_to = "contrast", values_to = "lfc") %>%
  dplyr::filter(!is.na(lfc)) %>%
  dplyr::mutate(contrast = factor(contrast, levels = c("RNA", "RIBO", "TE")))

p_arm <- ggplot(arm_long, aes(x = lfc, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  facet_wrap(~ contrast, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "4E-BP1-4A vs E.V., each arm fit separately",
       subtitle = paste0("n = ", sum(cls$class == "4E-resistant"), " resistant, ",
                         sum(cls$class == "4E-sensitive"), " sensitive"),
       x = "log2 fold change", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_arm)
if (SAVE_PLOTS) ggsave(here("plots", "80_rna_ribo_te_by_class.pdf"), p_arm,
                       width = 9, height = 4)
```

```{r gc-scatter, fig.width = 9, fig.height = 4}
sc <- gc_arms %>%
  dplyr::select(cds_gc, RIBO = ribo_log2FoldChange, RNA = rna_log2FoldChange, TE = te_lfc) %>%
  tidyr::pivot_longer(-cds_gc, names_to = "contrast", values_to = "lfc") %>%
  dplyr::mutate(contrast = factor(contrast, levels = c("RNA", "RIBO", "TE")))

lab <- sc %>% dplyr::group_by(contrast) %>%
  dplyr::summarise(r = round(cor(cds_gc, lfc, method = "spearman", use = "complete.obs"), 3),
                   .groups = "drop") %>%
  dplyr::mutate(txt = paste0("rho = ", r, "\nn = ", nrow(gc_arms)))

p_sc <- ggplot(sc, aes(x = cds_gc, y = lfc)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_point(size = 0.3, alpha = 0.15, color = "grey30") +
  geom_smooth(method = "lm", se = FALSE, color = "#0072B2", linewidth = 0.8) +
  facet_wrap(~ contrast, nrow = 1) +
  geom_text(data = lab, aes(x = -Inf, y = Inf, label = txt),
            hjust = -0.1, vjust = 1.3, size = 3.2, inherit.aes = FALSE) +
  labs(title = "CDS GC vs each contrast",
       x = "CDS GC content (%)", y = "log2 fold change") +
  theme_classic(base_size = 12)

print(p_sc)
if (SAVE_PLOTS) ggsave(here("plots", "80_cds_gc_vs_each_arm.pdf"), p_sc,
                       width = 9, height = 4)
```

## RIBO and RNA changes, for the TE-defined resistant and sensitive transcripts

The classes here are **exactly** the ones from the TE analysis - the same 1,195 resistant
and 1,195 sensitive transcripts used in notebook 79 - held fixed. What changes between the
two figures is the contrast on the x axis.

GC content is a property of the transcript, not of the assay, so a GC-by-region figure
cannot differ between the arms. The quantity that does differ is the fold change, and
that is what is plotted.

```{r arm-lfc-fixed-classes, fig.width = 7, fig.height = 4.5}
cls_fixed <- arms_j %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE ~ NA_character_)) %>%
  dplyr::filter(!is.na(class)) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)))

cat("n per class (TE-defined, as in nb79):\n")
print(dplyr::count(cls_fixed, class) %>% as.data.frame(), row.names = FALSE)

arm_lfc_plot <- function(col, arm_label, xlab) {
  sub <- cls_fixed %>% dplyr::filter(!is.na(.data[[col]]))
  wt  <- wilcox.test(sub[[col]] ~ sub$class)
  med <- sub %>% dplyr::group_by(class) %>%
    dplyr::summarise(m = median(.data[[col]]), .groups = "drop")

  lab <- paste0(
    "Wilcoxon p = ", format.pval(wt$p.value, digits = 3, eps = 2.2e-16),
    "\nmedian ", med$class[1], " = ", round(med$m[1], 3),
    "\nmedian ", med$class[2], " = ", round(med$m[2], 3),
    "\nn = ", sum(sub$class == med$class[1]), " / ", sum(sub$class == med$class[2]))

  ggplot(sub, aes(x = .data[[col]], fill = class, color = class)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
    geom_vline(data = med, aes(xintercept = m, color = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    scale_fill_manual(values = class_colors, name = NULL) +
    scale_color_manual(values = class_colors, name = NULL) +
    labs(title = paste0(arm_label, " change on 4E-BP1-4A"),
         subtitle = "classes fixed from the TE analysis (same transcripts as nb79)",
         x = xlab, y = "Density") +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2, label = lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}

p_ribo <- arm_lfc_plot("ribo_log2FoldChange", "RIBO (ribosome footprints)",
                       "log2 fold change, footprints (4E-BP1-4A vs E.V.)")
print(p_ribo)
if (SAVE_PLOTS) ggsave(here("plots", "80_ribo_lfc_by_te_class.pdf"), p_ribo,
                       width = 7, height = 4.5)

p_rna <- arm_lfc_plot("rna_log2FoldChange", "RNA (mRNA level)",
                      "log2 fold change, mRNA (4E-BP1-4A vs E.V.)")
print(p_rna)
if (SAVE_PLOTS) ggsave(here("plots", "80_rna_lfc_by_te_class.pdf"), p_rna,
                       width = 7, height = 4.5)
```

Both arms separate the TE-defined classes, in opposite directions: the transcripts called
resistant gain footprints and lose mRNA, and the sensitive ones do the reverse. A story in
which eIF4E sequestration acts on initiation predicts the footprint shift; it does not
predict an mRNA shift of comparable size in the opposite direction.

```{r arm-separation-of-te-classes}
sep <- purrr::map_dfr(
  c(RIBO = "ribo_log2FoldChange", RNA = "rna_log2FoldChange", TE = "te_lfc"),
  function(col) {
    x <- cls_fixed[[col]]; ok <- !is.na(x)
    r <- rank(x[ok]); cl <- cls_fixed$class[ok]
    n1 <- sum(cl == "4E-resistant"); n0 <- sum(cl == "4E-sensitive")
    a  <- (sum(r[cl == "4E-resistant"]) - n1 * (n1 + 1) / 2) / (n1 * n0)
    tibble::tibble(auc = round(max(a, 1 - a), 3),
                   median_resistant = round(median(x[cl == "4E-resistant"], na.rm = TRUE), 3),
                   median_sensitive = round(median(x[cl == "4E-sensitive"], na.rm = TRUE), 3))
  }, .id = "contrast")
print(as.data.frame(sep), row.names = FALSE)
```

TE separates the classes perfectly by construction (they were defined from it), so only
the RIBO and RNA rows carry information.

## GC by region, classes taken from each arm at matched n

The notebook 79 figure defines resistant/sensitive from **TE**. Here the same panel is
drawn twice more, with the classes taken from the **RIBO** contrast alone and from the
**RNA** contrast alone.

Everything else is held constant so the three figures are directly comparable:

- the **same transcript per gene** (nb78's dominant isoform, MANE fallback), so the GC
  values are identical molecules in all three figures;
- the **same class sizes**, 1,195 up and 1,195 down, matching the TE sets.

To reach 1,195 per class the ranking is by effect size alone - the significance filter
used for the TE sets would cap RIBO at 737 and RNA at 465 per class. The count passing
`padj < 0.05` is reported alongside so the difference is visible.

**Which direction is "resistant" in each arm.** `TE = ribo - rna`, so TE rises when
footprints go **up** *or* mRNA goes **down**. The resistant class is therefore the top of
the RIBO distribution but the *bottom* of the RNA distribution. Assigning mRNA-up as
"resistant" would label genes whose median TE is -0.58, i.e. the sensitive ones. Each
figure states its direction in the legend so this cannot be misread.

```{r arm-gc-matched-n, fig.width = 11, fig.height = 7}
gc_cols <- c("cds_gc", "gc1", "gc2", "gc3", "utr5_gc", "utr3_gc")
N_PER_CLASS <- 1195   # matches the TE sets used in nb79

# resistant_dir: which tail of this arm corresponds to HIGHER TE.
#   RIBO -> "up"   (more footprints raises TE)
#   RNA  -> "down" (less mRNA raises TE, since TE = ribo - rna)
arm_gc_matched <- function(lfc_col, padj_col, arm_label, resistant_dir) {
  stopifnot(resistant_dir %in% c("up", "down"))
  d0 <- arms_j %>%
    dplyr::select(gene_id, lfc = dplyr::all_of(lfc_col), padj = dplyr::all_of(padj_col)) %>%
    dplyr::filter(!is.na(lfc))

  hi <- d0 %>% dplyr::arrange(dplyr::desc(lfc)) %>% dplyr::slice_head(n = N_PER_CLASS)
  lo <- d0 %>% dplyr::arrange(lfc) %>% dplyr::slice_head(n = N_PER_CLASS)

  sets <- dplyr::bind_rows(
    (if (resistant_dir == "up") hi else lo) %>% dplyr::mutate(class = "4E-resistant"),
    (if (resistant_dir == "up") lo else hi) %>% dplyr::mutate(class = "4E-sensitive")
  ) %>%
    dplyr::inner_join(fm %>% dplyr::select(gene_id_clean, dplyr::all_of(gc_cols)),
                      by = c("gene_id" = "gene_id_clean")) %>%
    dplyr::mutate(class = factor(class, levels = names(class_colors)))

  cat("\n===", arm_label, " (resistant = ", arm_label, " ",
      if (resistant_dir == "up") "up" else "down", ") ===\n", sep = "")
  print(sets %>% dplyr::group_by(class) %>%
          dplyr::summarise(n = dplyr::n(),
                           n_padj_lt_05 = sum(padj < 0.05, na.rm = TRUE),
                           min_abs_lfc = round(min(abs(lfc)), 3),
                           median_cds_gc = round(median(cds_gc, na.rm = TRUE), 2),
                           median_gc3 = round(median(gc3, na.rm = TRUE), 2),
                           .groups = "drop") %>% as.data.frame(), row.names = FALSE)

  long <- sets %>%
    tidyr::pivot_longer(dplyr::all_of(gc_cols), names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc)) %>%
    dplyr::mutate(feature = factor(feature, levels = gc_cols))

  ggplot(long, aes(x = gc, fill = class, color = class)) +
    geom_density(alpha = 0.15, linewidth = 0.8) +
    facet_wrap(~ feature, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = class_colors, name = NULL) +
    scale_color_manual(values = class_colors, name = NULL) +
    labs(title = paste0("GC content by region and codon position: ", arm_label),
         subtitle = paste0("classes from the ", arm_label, " contrast; resistant = ",
                           arm_label, " ", if (resistant_dir == "up") "up" else "down",
                           " (the direction that raises TE)   |   n = ",
                           N_PER_CLASS, " per class"),
         x = "GC content (%)", y = "Density") +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}

p_gc_ribo <- arm_gc_matched("ribo_log2FoldChange", "ribo_padj", "RIBO", "up")
print(p_gc_ribo)
if (SAVE_PLOTS) ggsave(here("plots", "80_gc_by_region_RIBO.pdf"), p_gc_ribo,
                       width = 11, height = 7)

p_gc_rna <- arm_gc_matched("rna_log2FoldChange", "rna_padj", "RNA", "down")
print(p_gc_rna)
if (SAVE_PLOTS) ggsave(here("plots", "80_gc_by_region_RNA.pdf"), p_gc_rna,
                       width = 11, height = 7)
```

```{r arm-gc-auc-matched}
auc_matched <- function(lfc_col, label) {
  d0 <- arms_j %>% dplyr::select(gene_id, lfc = dplyr::all_of(lfc_col)) %>%
    dplyr::filter(!is.na(lfc))
  sets <- dplyr::bind_rows(
    d0 %>% dplyr::arrange(dplyr::desc(lfc)) %>% dplyr::slice_head(n = N_PER_CLASS) %>%
      dplyr::mutate(cls = "up"),
    d0 %>% dplyr::arrange(lfc) %>% dplyr::slice_head(n = N_PER_CLASS) %>%
      dplyr::mutate(cls = "down")
  ) %>% dplyr::inner_join(fm %>% dplyr::select(gene_id_clean, dplyr::all_of(gc_cols)),
                          by = c("gene_id" = "gene_id_clean"))
  purrr::map_dfr(gc_cols, function(f) {
    x <- sets[[f]]; ok <- !is.na(x)
    r <- rank(x[ok]); cl <- sets$cls[ok]
    n1 <- sum(cl == "up"); n0 <- sum(cl == "down")
    a <- (sum(r[cl == "up"]) - n1 * (n1 + 1) / 2) / (n1 * n0)
    tibble::tibble(definition = label, feature = f, auc = round(max(a, 1 - a), 3))
  })
}

dplyr::bind_rows(
  auc_matched("te_lfc", "TE"),
  auc_matched("ribo_log2FoldChange", "RIBO"),
  auc_matched("rna_log2FoldChange",  "RNA")
) %>%
  tidyr::pivot_wider(names_from = definition, values_from = auc) %>%
  as.data.frame() %>% print(row.names = FALSE)
```

All three use the same molecules and the same class sizes, so the AUCs are directly
comparable: the only thing that changes is which contrast picked the genes.

---

## How confident is the resistant/sensitive call, independent of any model?

The classes come from a single TE coefficient. Whether a given TE change reflects
*translation* is a separate question, and it is answered by looking at the two arms
together - which is what `categorize_translation_changes()` in `code/functions.R` already
does for this project (deltaTE-style classes):

| category | meaning | reading for a TE call |
|---|---|---|
| **Exclusive** | RIBO moves, RNA does not | translational - the confident case |
| **Buffered** | RNA moves; RIBO opposes or does not follow | the TE change is driven by the RNA denominator |
| **Forwarded** | RNA and RIBO move together, TE does not | not a TE change at all |
| **TE_only** | only the TE coefficient is significant | too weak to attribute to either arm |

`lfc_cutoff = 0.0` matches every existing call site in this project, so the categories
reflect significance by `padj` alone.

```{r deltate-categories}
source(here("code", "functions.R"))

cat_out <- categorize_translation_changes(
  res_rna  = arms$rna  %>% dplyr::transmute(gene_id, log2FoldChange = rna_log2FoldChange,
                                            padj = rna_padj),
  res_ribo = arms$ribo %>% dplyr::transmute(gene_id, log2FoldChange = ribo_log2FoldChange,
                                            padj = ribo_padj),
  res_te   = te        %>% dplyr::transmute(gene_id, log2FoldChange, padj),
  padj_cutoff = 0.05, lfc_cutoff = 0.0)

deltate <- cat_out$categorized_table %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE ~ "other")) 

cat("deltaTE categories, all genes:\n")
print(cat_out$summary_table, row.names = FALSE)

cat("\nWithin the 4E sets, % of each category:\n")
in_sets <- deltate %>% dplyr::filter(class != "other")
print(round(100 * prop.table(table(class = in_sets$class,
                                   category = in_sets$category), 1), 1))

cat("\nCounts:\n")
print(table(class = in_sets$class, category = in_sets$category))
```

```{r deltate-scatter, fig.width = 11, fig.height = 5}
# Project convention (code/functions.R): x = RNA log2FC, y = Ribo log2FC,
# alpha 0.6, +/-5 limits, theme_minimal(13), no legend title.
myCols <- c("Not_sig" = "grey80", "Forwarded" = "blue", "Exclusive" = "red",
            "Buffered" = "green", "TE_only" = "purple")
cls_cols <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2", "other" = "grey85")

base_scatter <- function(df, colour_var, cols, title, subtitle) {
  ggplot(df, aes(x = rna_lfc, y = ribo_lfc, color = .data[[colour_var]])) +
    geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
    geom_vline(xintercept = 0, color = "grey70", linewidth = 0.3) +
    geom_point(alpha = 0.6, size = 0.9) +
    scale_color_manual(values = cols) +
    xlim(-5, 5) + ylim(-5, 5) +
    xlab("RNA-seq log2FC") + ylab("Ribo-seq log2FC") +
    labs(title = title, subtitle = subtitle) +
    theme_minimal(base_size = 13) +
    theme(legend.title = element_blank())
}

p_cls <- base_scatter(
  deltate %>% dplyr::arrange(class != "other"), "class", cls_cols,
  "Ribo vs RNA change",
  paste0("n = ", sum(deltate$class == "4E-resistant"), " resistant, ",
         sum(deltate$class == "4E-sensitive"), " sensitive"))

p_cat <- base_scatter(
  deltate %>% dplyr::filter(category != "Not_sig"), "category", myCols,
  "Ribo vs RNA change, coloured by deltaTE category",
  "Not_sig omitted, as in code/functions.R")

print(p_cls); print(p_cat)
if (SAVE_PLOTS) {
  ggsave(here("plots", "80_ribo_vs_rna_by_4e_class.pdf"), p_cls, width = 6.5, height = 5)
  ggsave(here("plots", "80_ribo_vs_rna_by_deltate_category.pdf"), p_cat, width = 6.5, height = 5)
}
```

```{r deltate-scatter-facet, fig.width = 10, fig.height = 5}
p_facet <- base_scatter(
  in_sets, "category", myCols,
  "Where each 4E class sits in the ribo/RNA plane",
  "faceted by 4E class; colour = deltaTE category") +
  facet_wrap(~ class)

print(p_facet)
if (SAVE_PLOTS) ggsave(here("plots", "80_ribo_vs_rna_faceted_by_class.pdf"),
                       p_facet, width = 10, height = 5)
```

**Reading the confidence.** Points high on the y axis with `rna_lfc` near zero are the
clean translational calls. Points that reach a positive TE mainly by moving left along the
x axis owe their TE change to the mRNA denominator. The category table quantifies exactly
that split.

---

## GC by region, split by deltaTE category within the TE classes

The deltaTE categories are defined from the three-way RNA/RIBO/TE comparison, so they
only apply to the **TE-defined** resistant and sensitive sets - the same 1,195 / 1,195
used in notebook 79. Colours are the project's deltaTE palette from `code/functions.R`.

This asks the question the categories exist for: **does the GC separation survive inside
the Exclusive genes**, whose TE change is confidently translational, or does it live in
the Buffered genes, whose TE change is driven by the mRNA arm?

```{r gc-by-te-category, fig.width = 13, fig.height = 6}
cat_cols <- c("Exclusive" = "red", "Buffered" = "green", "TE_only" = "purple")

te_cat_sets <- deltate %>%
  dplyr::filter(class != "other", category %in% names(cat_cols)) %>%
  dplyr::inner_join(fm %>% dplyr::select(gene_id_clean, dplyr::all_of(gc_cols)),
                    by = c("gene_id" = "gene_id_clean")) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)),
                category = factor(category, levels = names(cat_cols)))

cat("n per class x deltaTE category (TE-defined sets):\n")
print(table(class = te_cat_sets$class, category = te_cat_sets$category))

# Forwarded and Not_sig cannot occur here, by construction: both require the gene to be
# NON-significant for TE, while every gene in these sets was selected on te padj < 0.05.
stopifnot("a Forwarded or Not_sig gene appeared in a TE-defined set" =
  !any(deltate$category[deltate$class != "other"] %in% c("Forwarded", "Not_sig")))
cat("\nForwarded / Not_sig are absent by construction: both require TE to be\n",
    "non-significant, and every gene in these sets has te padj < 0.05.\n", sep = "")

cat("\nmedian cds_gc / gc3 per group:\n")
print(te_cat_sets %>% dplyr::group_by(class, category) %>%
        dplyr::summarise(n = dplyr::n(),
                         cds_gc = round(median(cds_gc, na.rm = TRUE), 1),
                         gc3 = round(median(gc3, na.rm = TRUE), 1), .groups = "drop") %>%
        as.data.frame(), row.names = FALSE)

long_cat <- te_cat_sets %>%
  tidyr::pivot_longer(dplyr::all_of(gc_cols), names_to = "feature", values_to = "gc") %>%
  dplyr::filter(!is.na(gc)) %>%
  dplyr::mutate(feature = factor(feature, levels = gc_cols))

# Unfilled lines: three overlapping categories make shaded densities unreadable.
p_gc_cat <- ggplot(long_cat, aes(x = gc, color = category)) +
  geom_density(fill = NA, linewidth = 0.8) +
  facet_grid(class ~ feature, scales = "free_y") +
  scale_color_manual(values = cat_cols, name = NULL) +
  labs(title = "GC content by deltaTE category, within the TE-defined classes",
       subtitle = paste0("colour = deltaTE category   |   n = ",
                         sum(te_cat_sets$class == "4E-resistant"), " resistant, ",
                         sum(te_cat_sets$class == "4E-sensitive"), " sensitive"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", strip.text = element_text(size = 9))

print(p_gc_cat)
if (SAVE_PLOTS) ggsave(here("plots", "80_gc_by_deltate_category.pdf"), p_gc_cat,
                       width = 13, height = 6)
```

## Does the GC split survive inside the translational genes?

Separability of resistant from sensitive on GC, computed **within each deltaTE category
separately**. If the split is translational it should hold inside Exclusive; if it is an
mRNA effect it should be strongest inside Buffered.

```{r gc-auc-within-te-category}
purrr::map_dfr(levels(te_cat_sets$category), function(cc) {
  sub <- te_cat_sets %>% dplyr::filter(category == cc)
  purrr::map_dfr(c("cds_gc", "gc3", "utr3_gc", "utr5_gc"), function(f) {
    x <- sub[[f]]; ok <- !is.na(x)
    r <- rank(x[ok]); cl <- sub$class[ok]
    n1 <- sum(cl == "4E-resistant"); n0 <- sum(cl == "4E-sensitive")
    if (n1 < 20 || n0 < 20) return(NULL)
    a <- (sum(r[cl == "4E-resistant"]) - n1 * (n1 + 1) / 2) / (n1 * n0)
    tibble::tibble(category = cc, n_resistant = n1, n_sensitive = n0,
                   feature = f, auc = round(max(a, 1 - a), 3))
  })
}) %>%
  tidyr::pivot_wider(names_from = feature, values_from = auc) %>%
  as.data.frame() %>% print(row.names = FALSE)
```

---

# Part 2: Feature densities, resistant vs sensitive

Every feature from the notebook 78 matrix, split by class.

```{r feature-densities, fig.width = 11, fig.height = 13}
feat_cols <- setdiff(names(fm), c("transcript_id_clean", "gene_id_clean", "symbol"))

d <- fm %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id_clean %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id_clean %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE ~ NA_character_)) %>%
  dplyr::filter(!is.na(class)) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)))

cat("features plotted:", length(feat_cols), "  genes:", nrow(d), "\n")

long <- d %>%
  dplyr::select(class, dplyr::all_of(feat_cols)) %>%
  tidyr::pivot_longer(-class, names_to = "feature", values_to = "value") %>%
  dplyr::filter(!is.na(value))

p_feat <- ggplot(long, aes(x = value, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.6) +
  facet_wrap(~ feature, scales = "free", ncol = 4) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "All notebook 78 features, 4E-resistant vs 4E-sensitive",
       subtitle = paste0("n = ", sum(d$class == "4E-resistant"), " resistant, ",
                         sum(d$class == "4E-sensitive"), " sensitive"),
       x = NULL, y = "Density") +
  theme_classic(base_size = 9) +
  theme(legend.position = "top",
        strip.text = element_text(size = 7),
        axis.text = element_text(size = 6))

print(p_feat)
if (SAVE_PLOTS) ggsave(here("plots", "80_all_features_density_by_class.pdf"), p_feat,
                       width = 11, height = 13)
```

## Codon optimality: CAI, tAI, FOP and CSC

The four optimality measures, resistant vs sensitive. They do **not** agree, so the
column that matters is the last one: how far each measure is just restating GC3.

```{r optimality-densities, fig.width = 9, fig.height = 6}
opt_cols <- c("cai", "tai", "fop", "csc")

opt_long <- d %>%
  dplyr::select(class, dplyr::all_of(opt_cols)) %>%
  tidyr::pivot_longer(-class, names_to = "measure", values_to = "value") %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::mutate(measure = factor(measure, levels = opt_cols,
                                 labels = c("CAI", "tAI", "FOP", "CSC")))

p_opt <- ggplot(opt_long, aes(x = value, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ measure, scales = "free", nrow = 2) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "Codon optimality measures, 4E-resistant vs 4E-sensitive",
       subtitle = paste0("n = ", sum(d$class == "4E-resistant"), " resistant, ",
                         sum(d$class == "4E-sensitive"), " sensitive"),
       x = NULL, y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_opt)
if (SAVE_PLOTS) ggsave(here("plots", "80_codon_optimality_density_by_class.pdf"),
                       p_opt, width = 9, height = 6)
```

```{r optimality-stats}
opt_stats <- purrr::map_dfr(opt_cols, function(m) {
  x <- d[[m]]; ok <- !is.na(x)
  r <- rank(x[ok]); cl <- d$class[ok]
  n1 <- sum(cl == "4E-resistant"); n0 <- sum(cl == "4E-sensitive")
  a  <- (sum(r[cl == "4E-resistant"]) - n1 * (n1 + 1) / 2) / (n1 * n0)
  tibble::tibble(
    measure = m,
    median_resistant = round(median(x[cl == "4E-resistant"], na.rm = TRUE), 4),
    median_sensitive = round(median(x[cl == "4E-sensitive"], na.rm = TRUE), 4),
    higher_in = ifelse(a > 0.5, "resistant", "sensitive"),
    auc = round(max(a, 1 - a), 3),
    wilcox_p = signif(wilcox.test(x[cl == "4E-resistant"],
                                  x[cl == "4E-sensitive"])$p.value, 3),
    rho_with_gc3 = round(cor(fm[[m]], fm$gc3, method = "spearman",
                             use = "complete.obs"), 3)
  )
})
print(as.data.frame(opt_stats), row.names = FALSE)
```

**Read the `rho_with_gc3` column before the rest.** CAI correlates 0.87 with GC3 in this
dataset, so "resistant have higher CAI" is close to a restatement of "resistant are
GC3-rich" rather than independent evidence. The reason is structural: of the 18
multi-codon amino-acid families, 17 have a C- or G-ending codon as the top-ranked one in
the ribosomal-protein reference, so a gene using the preferred codon everywhere is
GC3-rich by construction.

tAI is built from tRNA gene copy numbers and is independent of the transcripts
(rho with GC3 = -0.15), and CSC comes from iCodon's empirical stability weights
(rho = 0.24). Note that FOP as built in notebook 78 is the fraction of codons in the top
tertile of *tAI* weights, so it is not independent confirmation of tAI.

```{r optimality-agreement}
cat("do the measures agree with each other?\n")
cat("  rho(CAI, tAI) =", round(cor(fm$cai, fm$tai, method = "spearman",
                                   use = "complete.obs"), 3), "\n")
cat("  rho(tAI, FOP) =", round(cor(fm$tai, fm$fop, method = "spearman",
                                   use = "complete.obs"), 3), " (FOP derives from tAI)\n")
cat("  rho(CSC, tAI) =", round(cor(fm$csc, fm$tai, method = "spearman",
                                   use = "complete.obs"), 3), "\n")
```

CAI and tAI are essentially uncorrelated, so they are not two versions of the same
quantity. By the measures not confounded with GC3, the **sensitive** set is the more
optimal one.

## Ranked separation

```{r feature-stats}
auc_single <- function(x, cls, positive = "4E-resistant") {
  ok <- !is.na(x); x <- x[ok]; cls <- cls[ok]
  if (length(unique(cls)) < 2) return(NA_real_)
  r <- rank(x); n1 <- sum(cls == positive); n0 <- sum(cls != positive)
  a <- (sum(r[cls == positive]) - n1 * (n1 + 1) / 2) / (n1 * n0)
  max(a, 1 - a)
}

feat_stats <- purrr::map_dfr(feat_cols, function(f) {
  x <- d[[f]]; ok <- !is.na(x)
  if (sum(ok) < 30 || dplyr::n_distinct(x[ok]) < 3) return(NULL)
  wt <- suppressWarnings(wilcox.test(x[ok] ~ d$class[ok]))
  tibble::tibble(
    feature   = f,
    auc_alone = round(auc_single(x, d$class), 3),
    median_resistant = round(median(x[d$class == "4E-resistant"], na.rm = TRUE), 3),
    median_sensitive = round(median(x[d$class == "4E-sensitive"], na.rm = TRUE), 3),
    wilcox_p  = wt$p.value,
    pct_na    = round(100 * mean(!ok), 1)
  )
}) %>%
  dplyr::mutate(padj = p.adjust(wilcox_p, method = "BH")) %>%
  dplyr::arrange(dplyr::desc(auc_alone))

print(as.data.frame(feat_stats %>%
  dplyr::mutate(wilcox_p = format.pval(wilcox_p, digits = 2, eps = 2.2e-16),
                padj     = format.pval(padj,     digits = 2, eps = 2.2e-16))),
  row.names = FALSE)

if (SAVE_PLOTS) {
  write_csv(feat_stats, here("output", "predictive_modeling",
                             "teleman_4e_feature_separation_stats.csv"))
}
```

# Summary

Read the `cor_tbl` and `arm-density` panels together: they say whether the GC split that
drives the notebook 07 classifier originates in the footprint arm, the mRNA arm, or only
in their ratio. A GC association of similar magnitude in both arms with opposite sign, or
one concentrated in the RNA arm, would argue that the compositional signal is not purely
translational.
