---
title: "GC3 calculated manually, and compared between 4E-resistant and 4E-sensitive"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Why this notebook exists

The GC3 difference between 4E-resistant and 4E-sensitive mRNAs is large (roughly 73%
vs 39%), and large enough to be worth recomputing from scratch rather than trusting the
feature-extraction pipeline.

So this notebook does it the slow, obvious way: read the CDS sequences, walk through
every transcript one at a time, pull out the third base of every codon, and count how
many of those are G or C. No helper functions, no `Biostrings` shortcuts for the counting
itself, nothing clever. Then compare the answer to the pipeline value, and to the two
gene sets.

```{r libraries}
library(tidyverse)
library(Biostrings)
library(here)   # so file paths work whether knitting or running chunks by hand
```

# Step 1: read in the CDS sequences

These are the coding sequences of the dominant transcript of each gene, written by
`cluster_scripts/viennarna/01c_generate_fasta_teleman.R`. The FASTA headers are the
version-stripped transcript IDs.

```{r read-fasta}
cds <- readDNAStringSet(here("output", "viennarna", "teleman_cds_sequences.fa"))

length(cds)
head(names(cds))
head(width(cds))
```

# Step 2: walk through every transcript and count GC at the third codon position

The third position of codon 1 is base 3, of codon 2 is base 6, of codon 3 is base 9, and
so on. So the positions we want are 3, 6, 9, ... up to the end of the sequence.

```{r manual-gc3}
# empty table to fill in as we go
results <- data.frame(
  transcript_id = names(cds),
  n_bases = NA,
  n_codons = NA,
  n_gc_at_pos3 = NA,
  gc3_manual = NA
)

for (i in 1:length(cds)) {

  # the sequence of this one transcript, as plain text
  one_sequence <- as.character(cds[[i]])

  # split it into single letters
  bases <- strsplit(one_sequence, "")[[1]]

  # positions 3, 6, 9, ... are the third base of each codon
  third_positions <- seq(from = 3, to = length(bases), by = 3)

  # the bases sitting at those positions
  third_bases <- bases[third_positions]

  # how many of them are G or C
  n_g <- sum(third_bases == "G")
  n_c <- sum(third_bases == "C")

  # fill in the table
  results$n_bases[i] <- length(bases)
  results$n_codons[i] <- length(third_bases)
  results$n_gc_at_pos3[i] <- n_g + n_c
  results$gc3_manual[i] <- (n_g + n_c) / length(third_bases) * 100
}

head(results)
nrow(results)
```

## A couple of sanity checks on the loop itself

```{r loop-checks}
# every CDS length should divide by 3
table(results$n_bases %% 3)

# number of codons should be the length divided by 3
head(results$n_bases / 3)
head(results$n_codons)

# GC3 has to sit between 0 and 100
range(results$gc3_manual)

# the count of G+C can never exceed the number of codons
sum(results$n_gc_at_pos3 > results$n_codons)
```

## Work one transcript out by hand, completely

Pick a short transcript and print every step, so the arithmetic can be checked with a
calculator.

```{r one-by-hand}
# the shortest CDS in the set
which_short <- which.min(results$n_codons)
results[which_short, ]

short_seq <- as.character(cds[[which_short]])
short_seq

short_bases <- strsplit(short_seq, "")[[1]]
short_third <- short_bases[seq(3, length(short_bases), by = 3)]

# the third base of every codon, in order
short_third

# how many are G or C
sum(short_third == "G")
sum(short_third == "C")

# and the percentage
(sum(short_third == "G") + sum(short_third == "C")) / length(short_third) * 100
```

# Step 3: does this agree with the pipeline?

```{r compare-pipeline}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds"))

compare <- merge(results, fm[, c("transcript_id_clean", "gc3", "symbol")],
                 by.x = "transcript_id", by.y = "transcript_id_clean")

nrow(compare)

# the pipeline leaves gc3 as NA for any CDS whose length is not a multiple of 3,
# so those cannot be compared
sum(is.na(compare$gc3))

# difference between the two, on the transcripts where both exist
compare$difference <- compare$gc3_manual - compare$gc3
both <- compare[!is.na(compare$gc3), ]

nrow(both)
max(abs(both$difference))
sum(abs(both$difference) > 1e-9)
cor(both$gc3_manual, both$gc3)
```

```{r compare-plot, fig.width = 5, fig.height = 5}
plot(both$gc3, both$gc3_manual,
     pch = 16, cex = 0.3, col = rgb(0, 0, 0, 0.2),
     xlab = "GC3 from the pipeline (%)",
     ylab = "GC3 counted by hand (%)",
     main = "Manual GC3 vs pipeline GC3")
abline(0, 1, col = "red")
```

# Step 4: compare GC3 between 4E-resistant and 4E-sensitive

```{r load-genesets}
resistant <- read.csv(here("output", "genesets", "teleman_4e_resistant_TE_1hr_lfc0.5.csv"))
sensitive <- read.csv(here("output", "genesets", "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"))

nrow(resistant)
nrow(sensitive)

# label each transcript by which set its gene belongs to
compare$gene_id <- fm$gene_id_clean[match(compare$transcript_id, fm$transcript_id_clean)]

compare$class <- NA
compare$class[compare$gene_id %in% resistant$ensembl_gene] <- "4E-resistant"
compare$class[compare$gene_id %in% sensitive$ensembl_gene] <- "4E-sensitive"

table(compare$class, useNA = "ifany")
```

```{r compare-classes}
gc3_resistant <- compare$gc3_manual[compare$class == "4E-resistant" & !is.na(compare$class)]
gc3_sensitive <- compare$gc3_manual[compare$class == "4E-sensitive" & !is.na(compare$class)]

length(gc3_resistant)
length(gc3_sensitive)

median(gc3_resistant)
median(gc3_sensitive)

mean(gc3_resistant)
mean(gc3_sensitive)

# spread, not just the centre
quantile(gc3_resistant, c(0.25, 0.75))
quantile(gc3_sensitive, c(0.25, 0.75))

wilcox.test(gc3_resistant, gc3_sensitive)
```

```{r density-plot, fig.width = 7, fig.height = 4.5}
plot_data <- compare[!is.na(compare$class), ]

ggplot(plot_data, aes(x = gc3_manual, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_vline(xintercept = median(gc3_resistant), color = "#D55E00",
             linetype = "dashed", linewidth = 0.5) +
  geom_vline(xintercept = median(gc3_sensitive), color = "#0072B2",
             linetype = "dashed", linewidth = 0.5) +
  scale_fill_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                    name = NULL) +
  scale_color_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                     name = NULL) +
  labs(title = "GC3 counted by hand, 4E-resistant vs 4E-sensitive",
       subtitle = paste0("median ", round(median(gc3_resistant), 1), "% vs ",
                         round(median(gc3_sensitive), 1), "%   |   n = ",
                         length(gc3_resistant), " vs ", length(gc3_sensitive)),
       x = "GC3, third codon position (%)",
       y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("plots", "82_gc3_manual_resistant_vs_sensitive.pdf"), width = 7, height = 4.5)
```

# Step 5: the same thing for the first and second codon positions

A sanity check on the loop itself. If the counting is right, doing it at positions 1 and 2
should give sensible numbers too, and the three positions should average to the overall
CDS GC content. That last check is the strong one: it uses nothing but arithmetic, and it
would fail immediately if the positions were off by one.

```{r manual-gc1-gc2}
results$gc1_manual <- NA
results$gc2_manual <- NA

for (i in 1:length(cds)) {

  bases <- strsplit(as.character(cds[[i]]), "")[[1]]

  # positions 1, 4, 7, ... are the first base of each codon
  first_bases <- bases[seq(from = 1, to = length(bases), by = 3)]

  # positions 2, 5, 8, ... are the second base of each codon
  second_bases <- bases[seq(from = 2, to = length(bases), by = 3)]

  results$gc1_manual[i] <-
    (sum(first_bases == "G") + sum(first_bases == "C")) / length(first_bases) * 100

  results$gc2_manual[i] <-
    (sum(second_bases == "G") + sum(second_bases == "C")) / length(second_bases) * 100
}

head(results[, c("transcript_id", "gc1_manual", "gc2_manual", "gc3_manual")])
```

## Check 1: every position should have the same number of bases

```{r check-counts}
# each codon contributes exactly one base to each position
n_first  <- NA
n_second <- NA
n_third  <- NA

for (i in 1:5) {
  bases <- strsplit(as.character(cds[[i]]), "")[[1]]
  n_first[i]  <- length(bases[seq(1, length(bases), by = 3)])
  n_second[i] <- length(bases[seq(2, length(bases), by = 3)])
  n_third[i]  <- length(bases[seq(3, length(bases), by = 3)])
}

n_first
n_second
n_third
results$n_codons[1:5]
```

## Check 2: the three positions must average to the whole-CDS GC

```{r check-average}
results$gc_mean_of_three <- (results$gc1_manual + results$gc2_manual + results$gc3_manual) / 3

# whole-CDS GC, counted the same plain way
results$cds_gc_manual <- NA
for (i in 1:length(cds)) {
  bases <- strsplit(as.character(cds[[i]]), "")[[1]]
  results$cds_gc_manual[i] <-
    (sum(bases == "G") + sum(bases == "C")) / length(bases) * 100
}

max(abs(results$gc_mean_of_three - results$cds_gc_manual))

# and against the pipeline's cds_gc
check_cds <- merge(results, fm[, c("transcript_id_clean", "cds_gc")],
                   by.x = "transcript_id", by.y = "transcript_id_clean")
max(abs(check_cds$cds_gc_manual - check_cds$cds_gc))
```

## Check 3: the pipeline's gc1 and gc2 should match too

```{r check-gc1-gc2-pipeline}
check2 <- merge(results, fm[, c("transcript_id_clean", "gc1", "gc2")],
                by.x = "transcript_id", by.y = "transcript_id_clean")
check2 <- check2[!is.na(check2$gc1), ]

nrow(check2)
max(abs(check2$gc1_manual - check2$gc1))
max(abs(check2$gc2_manual - check2$gc2))
```

# Step 6: compare all three positions between the two gene sets

```{r compare-all-positions}
compare$gc1_manual <- results$gc1_manual[match(compare$transcript_id, results$transcript_id)]
compare$gc2_manual <- results$gc2_manual[match(compare$transcript_id, results$transcript_id)]

in_sets <- compare[!is.na(compare$class), ]

# medians per position
tapply(in_sets$gc1_manual, in_sets$class, median)
tapply(in_sets$gc2_manual, in_sets$class, median)
tapply(in_sets$gc3_manual, in_sets$class, median)

# and the tests
wilcox.test(in_sets$gc1_manual ~ in_sets$class)
wilcox.test(in_sets$gc2_manual ~ in_sets$class)
wilcox.test(in_sets$gc3_manual ~ in_sets$class)
```

```{r plot-all-positions, fig.width = 9, fig.height = 4}
long <- rbind(
  data.frame(class = in_sets$class, position = "GC1", gc = in_sets$gc1_manual),
  data.frame(class = in_sets$class, position = "GC2", gc = in_sets$gc2_manual),
  data.frame(class = in_sets$class, position = "GC3", gc = in_sets$gc3_manual)
)

ggplot(long, aes(x = gc, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ position, nrow = 1) +
  scale_fill_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                    name = NULL) +
  scale_color_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                     name = NULL) +
  labs(title = "GC counted by hand at each codon position",
       subtitle = paste0("n = ", sum(in_sets$class == "4E-resistant"), " resistant, ",
                         sum(in_sets$class == "4E-sensitive"), " sensitive"),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(here("plots", "82_gc_by_codon_position_manual.pdf"), width = 9, height = 4)
```

```{r threshold-all-positions}
# how well does each position alone sort the two sets, using one cutoff
for (pos in c("gc1_manual", "gc2_manual", "gc3_manual")) {
  r <- in_sets[[pos]][in_sets$class == "4E-resistant"]
  s <- in_sets[[pos]][in_sets$class == "4E-sensitive"]
  th <- (median(r) + median(s)) / 2
  pct <- (sum(r > th) + sum(s < th)) / (length(r) + length(s)) * 100
  cat(pos, ": median", round(median(r), 1), "vs", round(median(s), 1),
      " threshold", round(th, 1), " correct", round(pct, 1), "%\n")
}
```

# Step 7: which nucleotide is actually preferred at each position?

GC3 lumps G and C together, and lumps A and T together as "not GC". That hides the
question of *which* base is preferred. So count all four separately, at each of the three
codon positions, still by hand.

```{r manual-four-bases}
# twelve new columns: each base at each position
for (base in c("A", "C", "G", "T")) {
  for (pos in 1:3) {
    results[[paste0("pos", pos, "_", base)]] <- NA
  }
}

for (i in 1:length(cds)) {

  bases <- strsplit(as.character(cds[[i]]), "")[[1]]

  first  <- bases[seq(from = 1, to = length(bases), by = 3)]
  second <- bases[seq(from = 2, to = length(bases), by = 3)]
  third  <- bases[seq(from = 3, to = length(bases), by = 3)]

  results$pos1_A[i] <- sum(first == "A") / length(first) * 100
  results$pos1_C[i] <- sum(first == "C") / length(first) * 100
  results$pos1_G[i] <- sum(first == "G") / length(first) * 100
  results$pos1_T[i] <- sum(first == "T") / length(first) * 100

  results$pos2_A[i] <- sum(second == "A") / length(second) * 100
  results$pos2_C[i] <- sum(second == "C") / length(second) * 100
  results$pos2_G[i] <- sum(second == "G") / length(second) * 100
  results$pos2_T[i] <- sum(second == "T") / length(second) * 100

  results$pos3_A[i] <- sum(third == "A") / length(third) * 100
  results$pos3_C[i] <- sum(third == "C") / length(third) * 100
  results$pos3_G[i] <- sum(third == "G") / length(third) * 100
  results$pos3_T[i] <- sum(third == "T") / length(third) * 100
}

head(results[, c("transcript_id", "pos3_A", "pos3_C", "pos3_G", "pos3_T")])
```

## Check: the four bases at a position must add up to 100%

```{r check-sums}
sum1 <- results$pos1_A + results$pos1_C + results$pos1_G + results$pos1_T
sum2 <- results$pos2_A + results$pos2_C + results$pos2_G + results$pos2_T
sum3 <- results$pos3_A + results$pos3_C + results$pos3_G + results$pos3_T

max(abs(sum1 - 100))
max(abs(sum2 - 100))
max(abs(sum3 - 100))

# and C + G at position 3 has to equal the GC3 we already counted
max(abs((results$pos3_C + results$pos3_G) - results$gc3_manual))
```

## Density of each nucleotide at each codon position

```{r plot-four-bases, fig.width = 11, fig.height = 7}
# attach the class labels to the per-base table
base_cols <- c("pos1_A","pos1_C","pos1_G","pos1_T",
               "pos2_A","pos2_C","pos2_G","pos2_T",
               "pos3_A","pos3_C","pos3_G","pos3_T")

base_table <- results[, c("transcript_id", base_cols)]
base_table$class <- compare$class[match(base_table$transcript_id, compare$transcript_id)]
base_table <- base_table[!is.na(base_table$class), ]

nrow(base_table)
table(base_table$class)

# long format, one row per transcript per position per base
base_long <- base_table %>%
  pivot_longer(cols = all_of(base_cols),
               names_to = c("position", "nucleotide"),
               names_sep = "_",
               values_to = "percent")

base_long$position <- factor(base_long$position,
                             levels = c("pos1", "pos2", "pos3"),
                             labels = c("codon position 1", "codon position 2",
                                        "codon position 3"))

ggplot(base_long, aes(x = percent, fill = class, color = class)) +
  geom_density(alpha = 0.15, linewidth = 0.7) +
  facet_grid(position ~ nucleotide, scales = "free_y") +
  scale_fill_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                    name = NULL) +
  scale_color_manual(values = c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2"),
                     name = NULL) +
  labs(title = "Which base does each codon position prefer?",
       subtitle = paste0("counted by hand   |   n = ",
                         sum(base_table$class == "4E-resistant"), " resistant, ",
                         sum(base_table$class == "4E-sensitive"), " sensitive"),
       x = "percent of codons with this base at this position",
       y = "Density") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")

ggsave(here("plots", "82_nucleotide_preference_by_position.pdf"), width = 11, height = 7)
```

## The numbers behind those plots

```{r base-medians}
for (pos in c("pos1", "pos2", "pos3")) {
  cat("\n", pos, "\n", sep = "")
  for (base in c("A", "C", "G", "T")) {
    col <- paste0(pos, "_", base)
    r <- base_table[[col]][base_table$class == "4E-resistant"]
    s <- base_table[[col]][base_table$class == "4E-sensitive"]
    cat(sprintf("  %s : resistant %5.1f%%   sensitive %5.1f%%   difference %+5.1f\n",
                base, median(r), median(s), median(r) - median(s)))
  }
}
```

# What this shows

The GC3 values counted base by base match the pipeline exactly, so the difference between
the two gene sets is not an artefact of the feature-extraction code. The size of the
difference, and how little the two distributions overlap, are printed above.
