---
title: "GC content and CSC: eIF3d-promoted vs negative control vs all transcripts (hypoxia 1hr)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  pos_geneset: "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"
  neg_geneset: "negative_control_genes.csv"
  match_negatives: TRUE
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

The same composition analysis run on the Teleman 4E sets (notebooks 79 and 82), but on
**our own MDA-MB-231 data**: mRNAs whose translation eIF3d promotes at hypoxia 1hr,
against the negative control set and against every other transcript.

Features compared: GC content in each region (5'UTR, CDS, 3'UTR, whole transcript),
GC by codon position (GC1/GC2/GC3), and CSC.

Three groups throughout, following the project convention in `CLAUDE.md`:

| group | definition |
|---|---|
| **eIF3d hypoxia promotes** | `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` - te_lfc > 0.5, padj < 0.05 |
| **Negative control** | `negative_control_genes.csv` - \|TE_LFC\| < 0.5 and padj > 0.3 in every si3d condition |
| **All other genes** | everything else in the feature matrix |

**Direction reminder.** The si3d DESeq2 coefficient is log2(sictrl / si3d) despite the
filename, so positive `te_lfc` means the knockdown *reduces* TE, i.e. eIF3d **promotes**
that mRNA. This is checked below rather than assumed.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    pos_geneset     = "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv",
    neg_geneset     = "negative_control_genes.csv",
    match_negatives = TRUE,
    save_plots      = TRUE
  )
}
POS_GENESET     <- params$pos_geneset
NEG_GENESET     <- params$neg_geneset
MATCH_NEGATIVES <- params$match_negatives
SAVE_PLOTS      <- params$save_plots

# eIF3d keeps its reserved orange; negative control black and background grey70 are the
# newer CLAUDE.md convention for new plots.
group_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",
  "Negative control"       = "black",
  "All other genes"        = "grey70"
)
```

# Section 1: Load the feature matrix and the gene sets

`feature_matrix_external_stability.rds` carries the regional GC, GC3 and CSC columns;
GC1 and GC2 come from `feature_matrix_positional_gc.rds`, which is keyed identically.

```{r load}
fm <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
fp <- readRDS(here("output", "predictive_modeling", "feature_matrix_positional_gc.rds"))

features <- fm %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol,
                utr5_gc, cds_gc, utr3_gc, transcript_gc, gc3, csc) %>%
  dplyr::left_join(fp %>% dplyr::select(transcript_id_clean, gc1, gc2),
                   by = "transcript_id_clean")

cat("feature matrix:", nrow(features), "transcripts\n")
stopifnot("matrix key must be unique" = !any(duplicated(features$transcript_id_clean)))

positives <- read_csv(here("output", "genesets", POS_GENESET), show_col_types = FALSE) %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene))

negatives <- read_csv(here("output", "predictive_modeling", NEG_GENESET),
                      show_col_types = FALSE) %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id))

cat("positive set:", nrow(positives), "genes\n")
cat("negative set:", nrow(negatives), "genes\n")

overlap <- intersect(positives$gene_id_clean, negatives$gene_id_clean)
cat("overlap between the two sets:", length(overlap), "\n")
stopifnot("positive and negative sets must be disjoint" = length(overlap) == 0)
```

# Section 2: Direction check

The gene set is only meaningful if positive `te_lfc` really means eIF3d-promoted. Two
checks against the source DESeq2 table: EIF3D's own `rna_lfc` must be strongly positive
(the siRNA depletes its target), and the positive set must have positive `te_lfc`.

```{r direction-check}
de <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
               show_col_types = FALSE) %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", gene_id))

eif3d_rna <- de$rna_lfc[de$symbol == "EIF3D"]
cat("EIF3D rna_lfc =", round(eif3d_rna, 3), " (expected strongly positive, ~+2.9)\n")
stopifnot("EIF3D rna_lfc is not positive - the contrast direction has flipped" =
            eif3d_rna > 1)

pos_te <- de$te_lfc[de$gene_id_clean %in% positives$gene_id_clean]
cat("positive-set te_lfc: median", round(median(pos_te, na.rm = TRUE), 3),
    " min", round(min(pos_te, na.rm = TRUE), 3), "\n")
stopifnot("positive set does not have positive te_lfc" =
            median(pos_te, na.rm = TRUE) > 0.5)

neg_te <- de$te_lfc[de$gene_id_clean %in% negatives$gene_id_clean]
cat("negative-set te_lfc: median", round(median(neg_te, na.rm = TRUE), 3),
    " (expected near 0)\n")
stopifnot("negative set is not centred near zero" =
            abs(median(neg_te, na.rm = TRUE)) < 0.3)

cat("Direction checks passed\n")
```

# Section 3: Assign the three groups

The negative control pool is larger than the promoted set, so with `match_negatives`
(default TRUE) it is **downsampled to the promoted count under `set.seed(9)`**, the way
notebook 07 balances the classifier's input.

This reproduces 07's draw rather than merely its size: the same two gene sets are
restricted to genes present in the feature matrix, in the order they appear in the CSVs,
and `sample()` is called on that vector immediately after `set.seed(9)` with
`min(n_pos, n_neg)` - identical inputs to 07's `create_target` chunk, so the same genes
come out. Unmatched negatives are dropped rather than folded into "All other genes", so
that group keeps its definition (every transcript in neither gene set), which is also
what 07 does with them.

```{r assign-groups}
# Mirrors notebook 07's create_target chunk exactly, including the vector being sampled.
pos_in_features <- positives$gene_id_clean[positives$gene_id_clean %in% features$gene_id_clean]
neg_in_features <- negatives$gene_id_clean[negatives$gene_id_clean %in% features$gene_id_clean]
number_to_match <- min(length(pos_in_features), length(neg_in_features))

cat("positive genes with features:", length(pos_in_features), "of", nrow(positives), "\n")
cat("negative genes with features:", length(neg_in_features), "of", nrow(negatives), "\n")

if (MATCH_NEGATIVES) {
  set.seed(9)  # project-wide seed; same draw as the model input in notebook 07
  neg_keep <- sample(neg_in_features, number_to_match)
  cat("size-matched negative control:", length(neg_in_features), "->",
      length(neg_keep), "(set.seed(9))\n")
} else {
  neg_keep <- neg_in_features
}

d <- features %>%
  dplyr::mutate(group = dplyr::case_when(
    gene_id_clean %in% positives$gene_id_clean ~ "eIF3d hypoxia promotes",
    gene_id_clean %in% neg_keep                ~ "Negative control",
    # unsampled negatives are neither group: drop them, do not call them "other"
    gene_id_clean %in% negatives$gene_id_clean ~ NA_character_,
    TRUE                                       ~ "All other genes")) %>%
  dplyr::filter(!is.na(group)) %>%
  dplyr::mutate(group = factor(group, levels = names(group_colors)))

print(dplyr::count(d, group) %>% as.data.frame(), row.names = FALSE)

n_pos <- sum(d$group == "eIF3d hypoxia promotes")
n_neg <- sum(d$group == "Negative control")

stopifnot("group must have no NA" = !any(is.na(d$group)))
stopifnot("positive group must be non-empty" = n_pos > 0)
stopifnot("negative group must be non-empty" = n_neg > 0)
stopifnot("every positive-set gene in the matrix must be labelled positive" =
            n_pos == length(pos_in_features))
stopifnot("negative control must be size-matched to the positive set" =
            !MATCH_NEGATIVES || n_neg == number_to_match)
stopifnot("no unsampled negative may leak into the background" =
            sum(d$group == "All other genes" & d$gene_id_clean %in% negatives$gene_id_clean) == 0)

# Section 2 checked the full negative pool; re-check the drawn subset, so a pathological
# sample cannot pass unnoticed. Direction is unchanged: te_lfc is log2(sictrl / si3d).
te_by_group <- tibble::tibble(
  group  = c("promoted", "negative control"),
  n      = c(n_pos, n_neg),
  median_te_lfc = c(
    median(de$te_lfc[de$gene_id_clean %in%
                       d$gene_id_clean[d$group == "eIF3d hypoxia promotes"]], na.rm = TRUE),
    median(de$te_lfc[de$gene_id_clean %in%
                       d$gene_id_clean[d$group == "Negative control"]], na.rm = TRUE)))
print(as.data.frame(te_by_group %>% dplyr::mutate(median_te_lfc = round(median_te_lfc, 3))),
      row.names = FALSE)

stopifnot("drawn negative control is not centred near zero" =
            abs(te_by_group$median_te_lfc[2]) < 0.3)
stopifnot("promoted set must sit above the negative control in te_lfc" =
            te_by_group$median_te_lfc[1] > te_by_group$median_te_lfc[2] + 0.5)
```

A single helper builds the legend labels so every plot below carries its own n's.

```{r legend-helper}
# labels of the form "Negative control (n = 1436)", computed on the rows actually plotted
label_with_n <- function(g) {
  tab <- table(droplevels(as.factor(g)))
  stats::setNames(paste0(names(tab), " (n = ", as.integer(tab), ")"), names(tab))
}

# n's make the labels long; wrapping keeps the legend inside the plotting frame
legend_rows <- function(n) guides(color = guide_legend(nrow = n, byrow = TRUE),
                                  fill  = guide_legend(nrow = n, byrow = TRUE))

# Every plot carries the sampling caveat, since the n's alone do not say the set was drawn
match_note <- if (MATCH_NEGATIVES)
  "negative control size-matched to the promoted set, set.seed(9)" else
  "full negative control pool (not size-matched)"
```

# Section 4: Density plots

```{r density-grid, fig.width = 11, fig.height = 6}
gc_cols <- c("utr5_gc", "cds_gc", "utr3_gc", "transcript_gc", "gc1", "gc2", "gc3")

# Complete cases across all GC columns, so the legend's n is exact in every facet rather
# than an upper bound. The missingness is in utr3_gc / transcript_gc (transcripts with no
# annotated 3'UTR); utr5_gc, cds_gc and gc1/gc2/gc3 are complete.
d_gc <- d %>% tidyr::drop_na(dplyr::all_of(gc_cols))
cat("dropped for incomplete GC features:", nrow(d) - nrow(d_gc), "transcripts\n")
print(colSums(is.na(d[, gc_cols])))
print(dplyr::count(d_gc, group) %>% as.data.frame(), row.names = FALSE)

gc_long <- d_gc %>%
  dplyr::select(group, dplyr::all_of(gc_cols)) %>%
  tidyr::pivot_longer(-group, names_to = "feature", values_to = "gc") %>%
  dplyr::mutate(feature = factor(feature, levels = gc_cols))

gc_labels <- label_with_n(d_gc$group)

p_gc <- ggplot(gc_long, aes(x = gc, fill = group, color = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = group_colors, labels = gc_labels, name = NULL) +
  scale_color_manual(values = group_colors, labels = gc_labels, name = NULL) +
  labs(title = "GC content by region and codon position, hypoxia 1hr",
       subtitle = match_note,
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  legend_rows(1)

print(p_gc)
if (SAVE_PLOTS) ggsave(here("plots", "83_gc_by_region_3d_promoted_hypoxia_1hr.pdf"),
                       p_gc, width = 11, height = 6)
```

```{r density-csc, fig.width = 7, fig.height = 4.5}
csc_d <- d %>% dplyr::filter(!is.na(csc))

wt_csc <- wilcox.test(csc_d$csc[csc_d$group == "eIF3d hypoxia promotes"],
                      csc_d$csc[csc_d$group == "Negative control"])

csc_labels <- label_with_n(csc_d$group)

p_csc <- ggplot(csc_d, aes(x = csc, fill = group, color = group)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_fill_manual(values = group_colors, labels = csc_labels, name = NULL) +
  scale_color_manual(values = group_colors, labels = csc_labels, name = NULL) +
  labs(title = "CSC, eIF3d-promoted vs negative control vs all other genes",
       subtitle = match_note,
       x = "CSC (codon stabilisation coefficient)", y = "Density") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("promoted vs negative control\np = ",
                          format.pval(wt_csc$p.value, digits = 3, eps = 2.2e-16),
                          "\nn = ", sum(csc_d$group == "eIF3d hypoxia promotes"), " / ",
                          sum(csc_d$group == "Negative control"))) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  legend_rows(2)

print(p_csc)
if (SAVE_PLOTS) ggsave(here("plots", "83_csc_3d_promoted_hypoxia_1hr.pdf"),
                       p_csc, width = 7, height = 4.5)
```

# Section 5: CDF plots

The project's other three-group comparisons (`ripseq/code/20_*`) use CDFs, which show
distribution shifts more clearly than densities when the groups differ in size.

```{r cdf-grid, fig.width = 11, fig.height = 6}
p_cdf <- ggplot(gc_long, aes(x = gc, color = group)) +
  stat_ecdf(linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_x", ncol = 4) +
  scale_color_manual(values = group_colors, labels = gc_labels, name = NULL) +
  labs(title = "GC content, cumulative distributions",
       subtitle = match_note,
       x = "GC content (%)", y = "Cumulative fraction") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  legend_rows(1)

print(p_cdf)
if (SAVE_PLOTS) ggsave(here("plots", "83_gc_cdf_3d_promoted_hypoxia_1hr.pdf"),
                       p_cdf, width = 11, height = 6)
```

# Section 6: Statistics

For each feature: medians per group, Wilcoxon of the promoted set against each
comparator, and the rank-based AUC of promoted vs negative control.

```{r stats}
all_features <- c(gc_cols, "csc")

stats <- purrr::map_dfr(all_features, function(f) {
  x <- d[[f]]
  ok <- !is.na(x)
  g  <- d$group[ok]; x <- x[ok]

  x_pos <- x[g == "eIF3d hypoxia promotes"]
  x_neg <- x[g == "Negative control"]
  x_oth <- x[g == "All other genes"]

  r  <- rank(c(x_pos, x_neg))
  n1 <- length(x_pos); n0 <- length(x_neg)
  a  <- (sum(r[seq_len(n1)]) - n1 * (n1 + 1) / 2) / (n1 * n0)

  tibble::tibble(
    feature          = f,
    n_promoted       = length(x_pos),
    n_negctrl        = length(x_neg),
    n_other          = length(x_oth),
    median_promoted  = round(median(x_pos), 3),
    median_negctrl   = round(median(x_neg), 3),
    median_other     = round(median(x_oth), 3),
    higher_in        = ifelse(a > 0.5, "promoted", "negative control"),
    auc_vs_negctrl   = round(max(a, 1 - a), 3),
    p_vs_negctrl     = signif(wilcox.test(x_pos, x_neg)$p.value, 3),
    p_vs_other       = signif(wilcox.test(x_pos, x_oth)$p.value, 3)
  )
}) %>%
  dplyr::arrange(dplyr::desc(auc_vs_negctrl))

print(as.data.frame(stats), row.names = FALSE)

if (SAVE_PLOTS) {
  write_csv(stats, here("output", "predictive_modeling",
                        "83_gc_csc_stats_3d_promoted_hypoxia_1hr.csv"))
}
```

# Section 7: Promoted genes split by deltaTE category, against the two backgrounds

The categories come from `categorize_translation_changes()` in `code/functions.R`, which
labels each gene by **which of the three contrasts is significant** - RNA, RIBO and TE.
Every call site in this project passes `lfc_cutoff = 0`, so "significant" here means
`padj < 0.05` alone, with no magnitude filter:

| category | RNA | RIBO | TE | reading |
|---|---|---|---|---|
| **Exclusive** | ns | sig | sig | footprints move, mRNA does not - a pure translational change |
| **Buffered** | sig | sig or ns | sig | mRNA moves and the ribosome does not follow it 1:1, so TE shifts |
| **TE_only** | ns | ns | sig | neither arm clears significance on its own; only the ratio does |
| **Forwarded** | sig | sig | ns | mRNA moves and footprints track it exactly - no TE change |
| **Not_sig** | ns | ns | ns | nothing moves |

Only the **promoted** set is split by category. The negative control and the rest of the
transcriptome are each shown as a single reference curve.

That asymmetry is deliberate, and it is forced by the data: promoted genes were selected
on `te_padj < 0.05`, so they can only be Exclusive, Buffered or TE_only. Genes that are
not promoted have no significant TE change, so they can only be Forwarded or Not_sig -
splitting them by category would compare categories that cannot co-occur.

The question the split answers is whether the promoted set's composition depends on
**which arm drove its TE change**: Exclusive (RIBO moves, RNA does not) against Buffered
(the RNA arm is driving it). TE_only is a residual bucket of ~12 genes here - its curve is
plotted for completeness but is not interpretable, which is exactly what the legend n
makes visible.

```{r deltate-category-gc, fig.width = 13, fig.height = 5}
de_cat <- de %>% dplyr::select(gene_id_clean, category)
stopifnot("category table must be one row per gene" = !any(duplicated(de_cat$gene_id_clean)))

d_cat <- d %>%
  dplyr::left_join(de_cat, by = "gene_id_clean") %>%
  dplyr::mutate(plot_group = dplyr::case_when(
    group == "eIF3d hypoxia promotes" & category == "Exclusive" ~ "Promoted: Exclusive",
    group == "eIF3d hypoxia promotes" & category == "Buffered"  ~ "Promoted: Buffered",
    group == "eIF3d hypoxia promotes" & category == "TE_only"   ~ "Promoted: TE_only",
    group == "Negative control"                                 ~ "Negative control",
    group == "All other genes"                                  ~ "All other genes",
    TRUE                                                        ~ NA_character_)) %>%
  dplyr::filter(!is.na(plot_group))

plot_cols <- c("Promoted: Exclusive" = "red",
               "Promoted: Buffered"  = "green3",
               "Promoted: TE_only"   = "purple",
               "Negative control"    = "black",
               "All other genes"     = "grey70")

d_cat <- d_cat %>%
  dplyr::mutate(plot_group = factor(plot_group, levels = names(plot_cols)))

cat("n per group:\n")
print(dplyr::count(d_cat, plot_group) %>% as.data.frame(), row.names = FALSE)

# The three promoted categories must account for the whole promoted set: a promoted gene
# with a missing or Forwarded/Not_sig category would be dropped silently by the filter above.
stopifnot("promoted genes lost when splitting by category" =
            sum(grepl("^Promoted", d_cat$plot_group)) == n_pos)

# Complete cases, so the legend n holds in every facet
d_cat_gc <- d_cat %>% tidyr::drop_na(dplyr::all_of(gc_cols))
cat("\nn per group with complete GC features:\n")
print(dplyr::count(d_cat_gc, plot_group) %>% as.data.frame(), row.names = FALSE)

# Densities normalise to unit area, so a curve's shape does not reveal how few genes are
# behind it - the per-group n belongs in the legend, next to each colour.
cat_labels <- label_with_n(d_cat_gc$plot_group)

long_cat <- d_cat_gc %>%
  tidyr::pivot_longer(dplyr::all_of(gc_cols), names_to = "feature", values_to = "gc") %>%
  dplyr::mutate(feature = factor(feature, levels = gc_cols))

# unfilled lines: five overlapping groups make shaded densities unreadable
p_cat <- ggplot(long_cat, aes(x = gc, color = plot_group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 4) +
  scale_color_manual(values = plot_cols, labels = cat_labels, name = NULL) +
  labs(title = "GC content: eIF3d-promoted genes by deltaTE category, vs both backgrounds",
       subtitle = paste0("hypoxia 1hr   |   promoted n = ",
                         sum(grepl("^Promoted", d_cat_gc$plot_group)),
                         " total   |   ", match_note),
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 11) +
  theme(legend.position = "top", strip.text = element_text(size = 9),
        legend.text = element_text(size = 9)) +
  legend_rows(2)

print(p_cat)
if (SAVE_PLOTS) ggsave(here("plots", "83_gc_by_deltate_category_3d_promoted.pdf"),
                       p_cat, width = 13, height = 5)
```

```{r deltate-category-csc, fig.width = 7, fig.height = 4.5}
csc_cat <- d_cat %>% dplyr::filter(!is.na(csc))
cat_csc_labels <- label_with_n(csc_cat$plot_group)

p_cat_csc <- ggplot(csc_cat, aes(x = csc, color = plot_group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  scale_color_manual(values = plot_cols, labels = cat_csc_labels, name = NULL) +
  labs(title = "CSC by deltaTE category, vs both backgrounds",
       subtitle = match_note,
       x = "CSC (codon stabilisation coefficient)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", legend.text = element_text(size = 9)) +
  legend_rows(3)

print(p_cat_csc)
if (SAVE_PLOTS) ggsave(here("plots", "83_csc_by_deltate_category_3d_promoted.pdf"),
                       p_cat_csc, width = 7, height = 4.5)
```

## Medians per group

Medians drop NAs per feature, so the group size is an upper bound on what any one column
is computed from. The n's actually behind each median are printed underneath.

```{r category-medians}
cat("medians:\n")
d_cat %>%
  dplyr::group_by(plot_group) %>%
  dplyr::summarise(across(dplyr::all_of(c(gc_cols, "csc")),
                          ~ round(median(.x, na.rm = TRUE), 2)),
                   .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)

cat("\nn behind each median (non-missing values per feature):\n")
d_cat %>%
  dplyr::group_by(plot_group) %>%
  dplyr::summarise(n_group = dplyr::n(),
                   across(dplyr::all_of(c(gc_cols, "csc")), ~ sum(!is.na(.x))),
                   .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)
```

## Within the promoted set: Exclusive vs Buffered

```{r exclusive-vs-buffered}
prom <- d_cat %>%
  dplyr::filter(plot_group %in% c("Promoted: Exclusive", "Promoted: Buffered")) %>%
  dplyr::mutate(cat2 = droplevels(plot_group))

cat("n Exclusive =", sum(prom$cat2 == "Promoted: Exclusive"),
    "  n Buffered =", sum(prom$cat2 == "Promoted: Buffered"), "\n\n")

purrr::map_dfr(c(gc_cols, "csc"), function(f) {
  x <- prom[[f]]; ok <- !is.na(x)
  x <- x[ok]; cl <- prom$cat2[ok]
  n1 <- sum(cl == "Promoted: Exclusive"); n0 <- sum(cl == "Promoted: Buffered")
  if (n1 < 20 || n0 < 20) return(NULL)
  r <- rank(x)
  a <- (sum(r[cl == "Promoted: Exclusive"]) - n1 * (n1 + 1) / 2) / (n1 * n0)
  tibble::tibble(
    feature = f,
    median_exclusive = round(median(x[cl == "Promoted: Exclusive"]), 3),
    median_buffered  = round(median(x[cl == "Promoted: Buffered"]), 3),
    higher_in = ifelse(a > 0.5, "Exclusive", "Buffered"),
    auc = round(max(a, 1 - a), 3),
    wilcox_p = signif(wilcox.test(x[cl == "Promoted: Exclusive"],
                                  x[cl == "Promoted: Buffered"])$p.value, 3))
}) %>% dplyr::arrange(dplyr::desc(auc)) %>% as.data.frame() %>% print(row.names = FALSE)
```

A high AUC would mean the promoted set's composition depends on which arm drove the TE
change - the same diagnostic applied to the Teleman sets in notebook 80.

---

# Section 8: Comparison to the Teleman 4E sets

For scale: in the Teleman eIF4E data the 4E-**resistant** set is GC-**rich** relative to
the 4E-sensitive set - about 19 points of CDS GC and 34 points of GC3 higher (GC3 roughly
73% vs 39%, notebooks 79 and 82) - and a single GC3 threshold classifies 90% of
transcripts correctly.

Both the magnitude and the **sign** are worth reading off the table below. The gap here is
printed as promoted minus negative control, so a negative number means the eIF3d-promoted
set is GC-*poor* relative to transcripts eIF3d does not regulate - the opposite direction
from the 4E-resistant set, at roughly a third of the magnitude. The two comparisons are
not strictly parallel (resistant vs sensitive is extreme-vs-extreme; promoted vs negative
control is regulated-vs-unregulated), so this is a scale check, not a like-for-like
contrast.

```{r scale-comparison}
gap <- stats %>%
  dplyr::filter(feature %in% c("cds_gc", "gc3")) %>%
  dplyr::mutate(gap_promoted_minus_negctrl = round(median_promoted - median_negctrl, 2)) %>%
  dplyr::select(feature, median_promoted, median_negctrl,
                gap_promoted_minus_negctrl, auc_vs_negctrl)

print(as.data.frame(gap), row.names = FALSE)

# Sanity check on the direction claimed in the prose above: eIF3d-promoted mRNAs are the
# GC-poor side of this comparison, unlike the 4E-resistant set.
stopifnot("eIF3d-promoted set is not GC-poor relative to the negative control" =
            all(gap$gap_promoted_minus_negctrl < 0))

cat("\nFor reference, the Teleman gaps (4E-resistant minus 4E-sensitive) were:\n")
cat("  cds_gc  +18.4 points, AUC 0.978\n")
cat("  gc3     +33.9 points, AUC 0.962\n")
cat("i.e. resistant mRNAs are GC-rich; here the promoted mRNAs are GC-poor.\n")
```

Note that `auc_vs_negctrl` is folded to the 0.5-1 range, so it carries magnitude only -
the `higher_in` column in Section 6 carries the direction.

# Summary

The statistics table is the deliverable. Two things about the comparison itself:

- The negative control set is **eIF3d-unaffected**, not a random background, so
  promoted-vs-negative asks whether eIF3d targets differ compositionally from transcripts
  eIF3d demonstrably does not regulate. Promoted-vs-other is the looser test against the
  rest of the transcriptome.
- With `match_negatives = TRUE` that negative set is the **same 1,436 genes the classifier
  in notebook 07 trains on** - same CSVs, same restriction to the feature matrix, same
  `set.seed(9)` draw - so these distributions describe the model's actual input rather
  than a differently sized pool.
