---
title: "Subunit-seq: eIF3d cap-bound mRNAs vs the rest of the transcriptome (MANE transcripts)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  lamper_lfc: 2
  lamper_padj: 0.005
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

The composition analysis of notebook 83, run on the mRNAs that eIF3d binds **at their 5'
cap**, rather than on the mRNAs whose TE eIF3d controls.

The binding data is **subunit-seq**: the Lee lab eIF3d-TEV assay, which uses a TEV
protease domain to release and sequence only the transcripts held at the cap by eIF3d.
Three conditions across two publications:

| group | source | condition |
|---|---|---|
| **complete media (CM)** | Lamper et al. 2020 (Table S3) | unstressed baseline |
| **glucose deprivation (GLU)** | Lamper et al. 2020 (Table S3) | metabolic stress |
| **integrated stress response (ISR)** | Lee lab 2023 (Table S2) | chronic stress |

Features compared: GC content by region (5'UTR, CDS, 3'UTR, whole transcript), GC by
codon position (GC1/GC2/GC3), region lengths, and CSC. Density curves are drawn
**unshaded** (outline only), unlike notebook 83.

## Two things this notebook cannot do, stated up front

**1. Subunit-seq is gene-level.** Both source tables carry `ensembl_gene_id` and
`hgnc_symbol` and nothing else. There are no transcript identifiers, and (unlike the Lee
2015 PAR-CLIP) no cluster coordinates either, so there is no route to transcript
resolution from the data itself. Every binding call below is therefore a **gene** call.
The MANE Select transcript is used as the sequence representative for that gene, which is
a modelling choice made here, not something the assay measured.

**2. "All other genes" conflates untested with unbound.** Lamper Table S3 reports 1,141
genes and the ISR table 563 - these are the published hit tables, not whole-transcriptome
results. A gene absent from them may have been unbound, or may simply not be reported.
The background group is therefore "not called bound in any subunit-seq condition", and
comparisons against it are conservative in an unquantifiable direction.

## Why MANE rather than the pipeline's usual transcript

The `01_` series keys feature matrices on the **most abundant** transcript in siCTRL
normoxia. Only 61.4% of those are the MANE Select for their gene, so
`feature_matrix_external_stability.rds` cannot be reused here - 39% of its rows would be
the wrong transcript for a MANE-keyed analysis. All features below are recomputed from
the GENCODE v49 TxDb on MANE transcripts, using the same helper definitions as `01_`,
`01g_` and `01v_` so the numbers stay comparable to notebook 83.

```{r libraries}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(iCodon)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    lamper_lfc  = 2,
    lamper_padj = 0.005,
    save_plots  = TRUE
  )
}
LAMPER_LFC  <- params$lamper_lfc
LAMPER_PADJ <- params$lamper_padj
SAVE_PLOTS  <- params$save_plots

SUBUNIT_SEQ_DIR <- "/Users/katematlin/github/mpra_preliminary/data"
LAMPER_XLSX <- file.path(SUBUNIT_SEQ_DIR, "lamper-s3.xlsx")
ISR_XLSX    <- file.path(SUBUNIT_SEQ_DIR, "integrated_stress_response.xlsx")

stopifnot("Lamper Table S3 not found" = file.exists(LAMPER_XLSX),
          "ISR Table S2 not found"    = file.exists(ISR_XLSX))

# All three sets are eIF3d, so orange alone cannot separate them. GLU keeps eIF3d's
# reserved orange as the headline stress condition; CM and ISR take remaining Okabe-Ito
# hues, avoiding #56B4E9 which is reserved for eIF3e. Background is grey70 per CLAUDE.md.
group_colors <- c(
  "eIF3d cap-bound (glucose deprivation)" = "#E69F00",
  "eIF3d cap-bound (complete media)"      = "#CC79A7",
  "eIF3d cap-bound (ISR)"                 = "#009E73",
  "All other genes"                       = "grey70"
)
GROUP_LEVELS <- names(group_colors)
```

# Section 1: Load subunit-seq data

Thresholds for the Lamper table follow the ones already used in
`code/clip_binding_sites.Rmd:107` and `code/ecdf_plots_3e_and_3d.Rmd:869`
(`log2FoldChange > 2 & padj < 0.005`), so the sets here agree with those notebooks. The
ISR table is already a filtered hit list and takes no further cut.

```{r load_lamper}
lamper <- read_excel(LAMPER_XLSX) %>%
  clean_names() %>%
  dplyr::mutate(dplyr::across(
    c(log2fold_change_cm, padj_cm, log2fold_change_glu, padj_glu,
      ip_in_cm_avg, ip_in_glu_avg, fold_enrichment_glu_cm),
    ~ suppressWarnings(as.numeric(.))
  )) %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id))

cat("Lamper Table S3 rows:", nrow(lamper), "\n")
cat("  rows with a testable GLU log2FC:", sum(!is.na(lamper$log2fold_change_glu)), "\n")
cat("  rows with a testable CM  log2FC:", sum(!is.na(lamper$log2fold_change_cm)), "\n")
```

## Replication check: Lamper et al. 2020

The paper's central claim is that eIF3d cap binding is **induced by glucose deprivation**,
not constitutive. The table must reproduce that: median IP/input enrichment under GLU has
to exceed the CM baseline, and the reported per-gene GLU/CM fold enrichment has to sit
above 1. This is checked before any threshold is applied, so a wrong file or a swapped
column pair fails here rather than silently reversing the biology downstream.

```{r replication_check_lamper}
med_ip_cm  <- median(lamper$ip_in_cm_avg,  na.rm = TRUE)
med_ip_glu <- median(lamper$ip_in_glu_avg, na.rm = TRUE)
med_fold   <- median(lamper$fold_enrichment_glu_cm, na.rm = TRUE)

cat("median IP/IN, complete media:      ", round(med_ip_cm, 2), "\n")
cat("median IP/IN, glucose deprivation: ", round(med_ip_glu, 2), "\n")
cat("median fold enrichment (GLU / CM): ", round(med_fold, 2),
    " (expected > 1: cap binding is stress-induced)\n")

stopifnot(
  "Lamper2020: GLU IP/IN does not exceed CM - columns may be swapped" =
    med_ip_glu > med_ip_cm,
  "Lamper2020: median GLU/CM fold enrichment is not above 1" =
    med_fold > 1
)
cat("Lamper 2020 replication check passed\n")
```

```{r load_isr}
isr <- read_excel(ISR_XLSX) %>%
  clean_names() %>%
  dplyr::mutate(gene_id_clean = sub("\\..*", "", ensembl_gene_id))

cat("ISR Table S2 rows:", nrow(isr), "\n")
cat("  log2FoldChange range:", round(range(isr$log2fold_change, na.rm = TRUE), 2), "\n")
cat("  max padj:", signif(max(isr$padj, na.rm = TRUE), 3), "\n")
```

## Replication check: Lee lab 2023 ISR table

This table is published as the set of transcripts **enriched** in the eIF3d-TEV IP, so
every entry must have a positive log2FoldChange. A negative value would mean the file is
a full DESeq2 output rather than the hit table, and the "bound" set would then be
contaminated with depleted transcripts.

```{r replication_check_isr}
cat("ISR entries with log2FoldChange > 0:", sum(isr$log2fold_change > 0, na.rm = TRUE),
    "of", nrow(isr), "\n")

stopifnot(
  "ISR table contains non-enriched entries - this is not the filtered hit table" =
    all(isr$log2fold_change > 0, na.rm = TRUE),
  "ISR log2FoldChange must be finite" =
    all(is.finite(isr$log2fold_change[!is.na(isr$log2fold_change)]))
)
cat("ISR 2023 replication check passed\n")
```

## Define the three bound gene sets

```{r define_bound_sets}
bound_glu <- lamper %>%
  dplyr::filter(log2fold_change_glu > LAMPER_LFC, padj_glu < LAMPER_PADJ) %>%
  dplyr::pull(gene_id_clean) %>% unique()

bound_cm <- lamper %>%
  dplyr::filter(log2fold_change_cm > LAMPER_LFC, padj_cm < LAMPER_PADJ) %>%
  dplyr::pull(gene_id_clean) %>% unique()

bound_isr <- isr %>%
  dplyr::pull(gene_id_clean) %>% unique()

bound_sets <- list(
  "eIF3d cap-bound (glucose deprivation)" = bound_glu,
  "eIF3d cap-bound (complete media)"      = bound_cm,
  "eIF3d cap-bound (ISR)"                 = bound_isr
)

cat("bound gene sets (unique ENSG):\n")
for (nm in names(bound_sets)) cat("  ", nm, ":", length(bound_sets[[nm]]), "\n")

cat("\npairwise overlaps:\n")
cat("  GLU & CM :", length(intersect(bound_glu, bound_cm)), "\n")
cat("  GLU & ISR:", length(intersect(bound_glu, bound_isr)), "\n")
cat("  CM  & ISR:", length(intersect(bound_cm,  bound_isr)), "\n")
cat("  any of the three:", length(Reduce(union, bound_sets)), "\n")

stopifnot("every bound set must be non-empty" =
            all(lengths(bound_sets) > 0))
```

The three sets overlap heavily, so they are **not** mutually exclusive groups. Each is
plotted as its own curve against the same shared background below; a gene bound in both
CM and GLU legitimately contributes to both curves.

# Section 2: MANE Select transcript per gene

```{r load_mane}
MANE_CSV <- here("accessories", "human", "mane_select_v49.csv")
stopifnot("MANE Select table not found" = file.exists(MANE_CSV))

mane <- read_csv(MANE_CSV,
                 col_names = c("gene_id_clean", "transcript_id_clean", "symbol"),
                 show_col_types = FALSE)

cat("MANE Select transcripts:", nrow(mane), "\n")
stopifnot(
  "MANE table is not one row per gene" = !any(duplicated(mane$gene_id_clean)),
  "MANE transcript IDs are not unique" = !any(duplicated(mane$transcript_id_clean)),
  "MANE gene IDs must be unversioned"  = !any(grepl("\\.", mane$gene_id_clean))
)

for (nm in names(bound_sets)) {
  n_mapped <- sum(bound_sets[[nm]] %in% mane$gene_id_clean)
  cat(sprintf("  %-38s %4d of %4d genes have a MANE transcript (%.1f%%)\n",
              nm, n_mapped, length(bound_sets[[nm]]),
              100 * n_mapped / length(bound_sets[[nm]])))
}
```

# Section 3: Compute features on MANE transcripts

Region sequences come from the GENCODE v49 TxDb; `gc_content()` and
`calculate_gc_at_position()` are the definitions from `01_feature_extraction.Rmd` and
`01v_feature_extraction_positional_gc.Rmd`, and `transcript_gc` is the same
length-weighted composite of the three regions used at `01_feature_extraction.Rmd:749`.

```{r load_txdb}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

utr5 <- fiveUTRsByTranscript(txdb, use.names = TRUE)
utr3 <- threeUTRsByTranscript(txdb, use.names = TRUE)
cds  <- cdsBy(txdb, by = "tx", use.names = TRUE)

cat("TxDb regions loaded: 5'UTR", length(utr5),
    "| CDS", length(cds), "| 3'UTR", length(utr3), "\n")
```

```{r extract_sequences}
# Extract sequences for one region set, restricted to MANE transcripts, returning a
# tibble keyed by transcript_id_clean.
extract_region <- function(gr_list, seq_col) {
  keep <- sub("\\..*", "", names(gr_list)) %in% mane$transcript_id_clean
  gr_list <- gr_list[keep]
  seqs <- extractTranscriptSeqs(genome, gr_list)
  tibble::tibble(
    transcript_id_clean = sub("\\..*", "", names(gr_list)),
    !!seq_col := as.character(seqs)
  )
}

utr5_seq <- extract_region(utr5, "utr5_sequence")
utr3_seq <- extract_region(utr3, "utr3_sequence")
cds_seq  <- extract_region(cds,  "cds_sequence")

cat("MANE transcripts with a 5'UTR:", nrow(utr5_seq), "\n")
cat("MANE transcripts with a CDS:  ", nrow(cds_seq), "\n")
cat("MANE transcripts with a 3'UTR:", nrow(utr3_seq), "\n")

stopifnot(
  "duplicate transcript in 5'UTR sequences" = !any(duplicated(utr5_seq$transcript_id_clean)),
  "duplicate transcript in CDS sequences"   = !any(duplicated(cds_seq$transcript_id_clean)),
  "duplicate transcript in 3'UTR sequences" = !any(duplicated(utr3_seq$transcript_id_clean))
)
```

```{r gc_helpers}
# From 01_feature_extraction.Rmd
gc_content <- function(seq) {
  seq <- str_to_upper(seq)
  gc_count     <- str_count(seq, "[GC]")
  total_length <- str_length(seq)
  result <- (gc_count / total_length) * 100
  result[total_length == 0] <- NA
  result
}

# From 01v_feature_extraction_positional_gc.Rmd
calculate_gc_at_position <- function(seqs, position) {
  stopifnot("position must be 1, 2 or 3" = position %in% 1:3)
  seqs <- str_to_upper(seqs)
  vapply(seqs, function(seq) {
    if (is.na(seq) || nchar(seq) < 3L) return(NA_real_)
    n_codons <- nchar(seq) %/% 3L
    starts   <- seq.int(position, n_codons * 3L, 3L)
    bases    <- substr(rep.int(seq, n_codons), starts, starts)
    sum(bases %in% c("G", "C")) * 100 / n_codons
  }, numeric(1L), USE.NAMES = FALSE)
}
```

```{r compute_gc}
features <- mane %>%
  dplyr::left_join(utr5_seq, by = "transcript_id_clean") %>%
  dplyr::left_join(cds_seq,  by = "transcript_id_clean") %>%
  dplyr::left_join(utr3_seq, by = "transcript_id_clean") %>%
  dplyr::mutate(
    utr5_length = str_length(utr5_sequence),
    cds_length  = str_length(cds_sequence),
    utr3_length = str_length(utr3_sequence),
    utr5_gc     = gc_content(utr5_sequence),
    cds_gc      = gc_content(cds_sequence),
    utr3_gc     = gc_content(utr3_sequence),
    gc1         = calculate_gc_at_position(cds_sequence, 1),
    gc2         = calculate_gc_at_position(cds_sequence, 2),
    gc3         = calculate_gc_at_position(cds_sequence, 3),
    # Length-weighted composite, matching 01_feature_extraction.Rmd:749
    transcript_gc = (utr5_gc * utr5_length + cds_gc * cds_length + utr3_gc * utr3_length) /
                    (utr5_length + cds_length + utr3_length)
  )

cat("MANE feature table:", nrow(features), "transcripts\n")
cat("non-NA per feature:\n")
for (f in c("utr5_gc", "cds_gc", "utr3_gc", "transcript_gc", "gc1", "gc2", "gc3")) {
  cat(sprintf("  %-14s %5d\n", f, sum(!is.na(features[[f]]))))
}
```

```{r compute_csc}
# iCodon CSC, following the deduplicate-score-restore pattern from 01g_ / 01j_.
predict_human <- iCodon::predict_stability("human")

cds_ok <- features %>%
  dplyr::filter(!is.na(cds_sequence), str_length(cds_sequence) %% 3 == 0,
                str_length(cds_sequence) >= 3)

cat("CDS scorable for CSC (length divisible by 3):", nrow(cds_ok),
    "of", sum(!is.na(features$cds_sequence)), "\n")

unique_seqs   <- unique(cds_ok$cds_sequence)
unique_scores <- predict_human(unique_seqs)
names(unique_scores) <- unique_seqs

features <- features %>%
  dplyr::mutate(csc = dplyr::if_else(
    !is.na(cds_sequence) & str_length(cds_sequence) %% 3 == 0 &
      str_length(cds_sequence) >= 3,
    unname(unique_scores[cds_sequence]),
    NA_real_
  ))

cat("csc non-NA:", sum(!is.na(features$csc)), "\n")
print(summary(features$csc))
```

# Section 4: Validation

## Computational checks

```{r checks_computational}
gc_cols  <- c("utr5_gc", "cds_gc", "utr3_gc", "transcript_gc", "gc1", "gc2", "gc3")
len_cols <- c("utr5_length", "cds_length", "utr3_length")

for (f in gc_cols) {
  stopifnot(setNames(
    list(all(features[[f]] >= 0 & features[[f]] <= 100, na.rm = TRUE)),
    paste(f, "out of [0, 100]")
  )[[1]])
}

stopifnot(
  "region lengths must be positive"  =
    all(unlist(features[len_cols]) > 0, na.rm = TRUE),
  "csc must be finite where present" =
    all(is.finite(features$csc[!is.na(features$csc)])),
  "expected feature columns missing" =
    all(c(gc_cols, len_cols, "csc") %in% colnames(features)),
  "transcript key must be unique"    =
    !any(duplicated(features$transcript_id_clean))
)

# Leakage guard: nothing about eIF3d binding may live in the feature table itself
stopifnot("binding label must not be a feature" =
            !any(c("bound", "group", "log2fold_change_glu") %in% colnames(features)))

cat("Computational checks passed\n")
```

## Biological spot-checks

The established invariants from `01_`, `01g_` and `01v_`, re-run because these features
are computed on a different transcript set (MANE) than those notebooks used.

```{r checks_biological}
actb <- features %>% dplyr::filter(symbol == "ACTB")
cat("ACTB CDS GC:", round(actb$cds_gc, 2), "% (expected 50-70)\n")
stopifnot("ACTB CDS GC outside 50-70%" = between(actb$cds_gc, 50, 70))

med_gc3 <- median(features$gc3, na.rm = TRUE)
cat("median GC3:", round(med_gc3, 2), "% (expected 40-65)\n")
stopifnot("GC3 median outside [40, 65]" = between(med_gc3, 40, 65))

med_gc1 <- median(features$gc1, na.rm = TRUE)
med_gc2 <- median(features$gc2, na.rm = TRUE)
cat("median GC1:", round(med_gc1, 2), "% | median GC2:", round(med_gc2, 2), "%\n")
stopifnot("GC1 should exceed GC2 in human coding sequence" = med_gc1 > med_gc2)
stopifnot("GC3 should be the most variable codon position" =
            sd(features$gc3, na.rm = TRUE) > sd(features$gc1, na.rm = TRUE) &&
            sd(features$gc3, na.rm = TRUE) > sd(features$gc2, na.rm = TRUE))

# Ribosomal proteins are the canonical optimal-codon set, so their CSC must beat the
# transcriptome median (the CAI analogue of the 01g_ check).
rp_csc  <- features$csc[str_detect(features$symbol, "^RP[LS]\\d")]
all_csc <- features$csc
cat("median CSC, ribosomal proteins:", round(median(rp_csc, na.rm = TRUE), 4),
    "| all genes:", round(median(all_csc, na.rm = TRUE), 4),
    "| n RP =", sum(!is.na(rp_csc)), "\n")
stopifnot("ribosomal proteins do not have above-median CSC" =
            median(rp_csc, na.rm = TRUE) > median(all_csc, na.rm = TRUE))

cat("Biological spot-checks passed\n")
```

# Section 5: Assign groups

Each bound set becomes its own curve; the background is every MANE gene called bound in
none of the three conditions. Because the sets overlap, this is built long, one row per
(transcript, group) pair, rather than as a single exclusive factor.

```{r assign_groups}
bound_any <- Reduce(union, bound_sets)

feature_cols <- c(gc_cols, len_cols, "csc")

d_long <- purrr::imap_dfr(bound_sets, function(genes, nm) {
  features %>%
    dplyr::filter(gene_id_clean %in% genes) %>%
    dplyr::mutate(group = nm)
}) %>%
  dplyr::bind_rows(
    features %>%
      dplyr::filter(!gene_id_clean %in% bound_any) %>%
      dplyr::mutate(group = "All other genes")
  ) %>%
  dplyr::mutate(group = factor(group, levels = GROUP_LEVELS))

group_n <- dplyr::count(d_long, group, name = "n")
print(as.data.frame(group_n), row.names = FALSE)

stopifnot(
  "group must have no NA"       = !any(is.na(d_long$group)),
  "every group must be present" = nrow(group_n) == length(GROUP_LEVELS),
  "every group must be non-empty" = all(group_n$n > 0)
)

n_label <- paste0("n = ", paste(sprintf("%s %s", group_n$n,
                                        c("GLU", "CM", "ISR", "other")),
                                collapse = ", "))
cat("\n", n_label, "\n")
```

# Section 6: Density plots (unshaded)

```{r density_gc, fig.width = 11, fig.height = 6}
gc_long <- d_long %>%
  dplyr::select(group, dplyr::all_of(gc_cols)) %>%
  tidyr::pivot_longer(-group, names_to = "feature", values_to = "gc") %>%
  dplyr::filter(!is.na(gc)) %>%
  dplyr::mutate(feature = factor(feature, levels = gc_cols))

p_gc <- ggplot(gc_long, aes(x = gc, color = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 4) +
  scale_color_manual(values = group_colors, name = NULL) +
  labs(title = "GC content by region and codon position, eIF3d cap-bound (subunit-seq)",
       subtitle = n_label,
       x = "GC content (%)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 2))

print(p_gc)
if (SAVE_PLOTS) ggsave(here("plots", "85_gc_by_region_subunit_seq_bound.pdf"),
                       p_gc, width = 11, height = 6)
```

```{r density_length, fig.width = 11, fig.height = 3.5}
len_long <- d_long %>%
  dplyr::select(group, dplyr::all_of(len_cols)) %>%
  tidyr::pivot_longer(-group, names_to = "feature", values_to = "len") %>%
  dplyr::filter(!is.na(len), len > 0) %>%
  dplyr::mutate(feature = factor(feature, levels = len_cols))

p_len <- ggplot(len_long, aes(x = len, color = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  scale_x_log10() +
  scale_color_manual(values = group_colors, name = NULL) +
  labs(title = "Region length, eIF3d cap-bound (subunit-seq)",
       subtitle = n_label,
       x = "Length (nt, log10 scale)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 2))

print(p_len)
if (SAVE_PLOTS) ggsave(here("plots", "85_length_subunit_seq_bound.pdf"),
                       p_len, width = 11, height = 3.5)
```

```{r density_csc, fig.width = 7, fig.height = 4.5}
csc_d <- d_long %>% dplyr::filter(!is.na(csc))

wt_csc <- wilcox.test(
  csc_d$csc[csc_d$group == "eIF3d cap-bound (glucose deprivation)"],
  csc_d$csc[csc_d$group == "All other genes"]
)

p_csc <- ggplot(csc_d, aes(x = csc, color = group)) +
  geom_density(fill = NA, linewidth = 0.8) +
  scale_color_manual(values = group_colors, name = NULL) +
  labs(title = "CSC, eIF3d cap-bound (subunit-seq) vs all other genes",
       x = "CSC (codon stabilisation coefficient)", y = "Density") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3.2,
           label = paste0("GLU-bound vs all other\np = ",
                          format.pval(wt_csc$p.value, digits = 3, eps = 2.2e-16),
                          "\nn = ",
                          sum(csc_d$group == "eIF3d cap-bound (glucose deprivation)"),
                          " / ", sum(csc_d$group == "All other genes"))) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 2))

print(p_csc)
if (SAVE_PLOTS) ggsave(here("plots", "85_csc_subunit_seq_bound.pdf"),
                       p_csc, width = 7, height = 4.5)
```

# Section 7: Cumulative distributions

The bound sets are roughly 20-fold smaller than the background, and CDFs read shifts more
reliably than densities at that size ratio (the same reasoning as notebook 83, Section 5).

```{r cdf_gc, fig.width = 11, fig.height = 6}
p_cdf <- ggplot(gc_long, aes(x = gc, color = group)) +
  stat_ecdf(linewidth = 0.8) +
  facet_wrap(~ feature, scales = "free_x", ncol = 4) +
  scale_color_manual(values = group_colors, name = NULL) +
  labs(title = "GC content, cumulative distributions",
       subtitle = n_label,
       x = "GC content (%)", y = "Cumulative fraction") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top") +
  guides(color = guide_legend(nrow = 2))

print(p_cdf)
if (SAVE_PLOTS) ggsave(here("plots", "85_gc_cdf_subunit_seq_bound.pdf"),
                       p_cdf, width = 11, height = 6)
```

# Section 8: Statistics

For every feature and every bound set: medians, Wilcoxon against the background, and the
rank-based AUC of bound vs background.

```{r stats}
bg <- d_long %>% dplyr::filter(group == "All other genes")

stats <- purrr::map_dfr(setdiff(GROUP_LEVELS, "All other genes"), function(grp) {
  sub <- d_long %>% dplyr::filter(group == grp)
  purrr::map_dfr(feature_cols, function(f) {
    x_pos <- sub[[f]][!is.na(sub[[f]])]
    x_bg  <- bg[[f]][!is.na(bg[[f]])]
    if (length(x_pos) < 3 || length(x_bg) < 3) return(NULL)

    r  <- rank(c(x_pos, x_bg))
    n1 <- length(x_pos); n0 <- length(x_bg)
    a  <- (sum(r[seq_len(n1)]) - n1 * (n1 + 1) / 2) / (n1 * n0)

    tibble::tibble(
      group          = grp,
      feature        = f,
      n_bound        = n1,
      n_background   = n0,
      median_bound   = round(median(x_pos), 3),
      median_bg      = round(median(x_bg), 3),
      higher_in      = ifelse(a > 0.5, "bound", "background"),
      auc            = round(max(a, 1 - a), 3),
      p_vs_bg        = signif(wilcox.test(x_pos, x_bg)$p.value, 3)
    )
  })
}) %>%
  dplyr::arrange(group, dplyr::desc(auc))

print(as.data.frame(stats), row.names = FALSE)
```

# Section 9: Write outputs

The three bound sets are written MANE-keyed, carrying both the gene and its MANE
transcript so downstream notebooks can join on either. Genes without a MANE Select
transcript are retained with `transcript_id_clean = NA` rather than dropped, so the files
stay a faithful record of the published sets.

```{r write_outputs}
if (SAVE_PLOTS) {
  set_files <- c(
    "eIF3d cap-bound (glucose deprivation)" = "subunit_seq_3d_bound_glucose_deprivation_mane.csv",
    "eIF3d cap-bound (complete media)"      = "subunit_seq_3d_bound_complete_media_mane.csv",
    "eIF3d cap-bound (ISR)"                 = "subunit_seq_3d_bound_isr_mane.csv"
  )

  for (nm in names(bound_sets)) {
    out <- tibble::tibble(gene_id_clean = bound_sets[[nm]]) %>%
      dplyr::left_join(mane, by = "gene_id_clean") %>%
      dplyr::mutate(condition = nm)
    write_csv(out, here("output", "genesets", set_files[[nm]]))
    cat("wrote", set_files[[nm]], "-", nrow(out), "genes,",
        sum(!is.na(out$transcript_id_clean)), "with a MANE transcript\n")
  }

  write_csv(stats, here("output", "predictive_modeling",
                        "85_subunit_seq_bound_feature_stats.csv"))
  saveRDS(features, here("output", "predictive_modeling",
                         "feature_matrix_mane_gc_csc.rds"))
  cat("wrote 85_subunit_seq_bound_feature_stats.csv and feature_matrix_mane_gc_csc.rds\n")
}
```

```{r session_info}
sessionInfo()
```
