---
title: "TE change bar plots: hypoxia timecourse + normoxia 1+4hr time-regressed"
author: "Kate"
date: "2026-08-16"
output: html_document
editor_options:
  chunk_output_type: console
---

Re-makes the eIF3d / eIF3e "number of genes with significant TE changes" bar plots
from `code/eif3e_eif3d_normoxia_and_hypoxia.Rmd` (chunks `te_barplot_eif3d` /
`te_barplot_eif3e`), adding a **normoxia 1+4hr time-regressed** bar pair alongside
the three hypoxia timepoints.

Nothing is refit here. All counts are read from the DESeq2 results already saved by
`eif3e_eif3d_normoxia_and_hypoxia.Rmd`:

| Bar | Source file |
|---|---|
| si3d hypoxia 1/4/24hr | `output/translation_categories_si3d_vs_sictrl_hypoxia_{1,4,24}hr.csv` |
| si3d normoxia 1+4hr   | `output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv` |
| si3e hypoxia 1/4/24hr | `output/translation_categories_si3e_vs_sictrl_hypoxia_{1,4,24}hr.csv` |
| si3e normoxia 1+4hr   | `output/te_si3e_vs_sictrl_normoxia_1and4hr.csv` |

The two 1+4hr contrasts come from `deFunction_time_normalized()` (design
`~ hours + Condition + SeqType + Condition:SeqType`), i.e. `hours` regressed out as a
batch covariate - that is the "time-regressed" result requested.

**Directionality.** These contrasts run `control_condition = si3d/si3e` and
`test_condition = sictrl`, so the DESeq2 coefficient is `sictrl_vs_si3d` and every
`te_lfc` is **log2(sictrl / si3d)** despite the `si3d_vs_sictrl` filenames. Positive
`te_lfc` = higher TE in sictrl = knockdown reduces TE = **eIF3d/eIF3e promotes**.
Asserted below on the knockdown target's own `rna_lfc`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggprism)
library(patchwork)
```

# Thresholds

```{r params}
padj_cutoff <- 0.05
lfc_cutoff  <- 0.5
```

# Load saved TE results

```{r load}
te_files <- tribble(
  ~gene,   ~contrast,      ~x_label,        ~path,
  "eIF3d", "hypoxia_1hr",  "1",             "output/translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv",
  "eIF3d", "hypoxia_4hr",  "4",             "output/translation_categories_si3d_vs_sictrl_hypoxia_4hr.csv",
  "eIF3d", "hypoxia_24hr", "24",            "output/translation_categories_si3d_vs_sictrl_hypoxia_24hr.csv",
  "eIF3d", "normoxia_1and4hr", "Norm",      "output/translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv",
  "eIF3e", "hypoxia_1hr",  "1",             "output/translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv",
  "eIF3e", "hypoxia_4hr",  "4",             "output/translation_categories_si3e_vs_sictrl_hypoxia_4hr.csv",
  "eIF3e", "hypoxia_24hr", "24",            "output/translation_categories_si3e_vs_sictrl_hypoxia_24hr.csv",
  "eIF3e", "normoxia_1and4hr", "Norm",      "output/te_si3e_vs_sictrl_normoxia_1and4hr.csv"
)

stopifnot("missing input TE result file" = all(file.exists(here(te_files$path))))

te_list <- set_names(
  map(te_files$path, ~ read_csv(here(.x), show_col_types = FALSE)),
  paste(te_files$gene, te_files$contrast, sep = "_")
)

# every file must carry the TE columns we count on
stopifnot("te_lfc/te_padj/symbol missing" =
  all(map_lgl(te_list, ~ all(c("te_lfc", "te_padj", "symbol") %in% names(.x)))))

# log-fold changes must be finite where present
stopifnot("non-finite te_lfc" =
  all(map_lgl(te_list, ~ all(is.finite(.x$te_lfc[!is.na(.x$te_lfc)])))))

# adjusted p-values must be valid probabilities
stopifnot("te_padj outside [0,1]" =
  all(map_lgl(te_list, ~ all(.x$te_padj >= 0 & .x$te_padj <= 1, na.rm = TRUE))))

walk2(names(te_list), te_list, ~ cat(sprintf("%-24s n = %d genes\n", .x, nrow(.y))))
```

# Directionality check

The siRNA depletes its own target, so the knockdown target's `rna_lfc` must be
strongly **positive** under the log2(sictrl / si3x) coefficient. The si3e 1+4hr TE
file carries TE columns only, so its check uses the companion RNA file written by the
same chunk of the source notebook.

```{r direction_check}
direction_check <- function(df, target_symbol, label) {
  v <- df$rna_lfc[df$symbol == target_symbol & !is.na(df$symbol)]
  stopifnot("knockdown target not found exactly once" = length(v) == 1)
  cat(sprintf("%-24s %s rna_lfc = %+.2f\n", label, target_symbol, v))
  v
}

kd_rna_lfc <- c(
  direction_check(te_list[["eIF3d_hypoxia_1hr"]],      "EIF3D", "si3d hypoxia 1hr"),
  direction_check(te_list[["eIF3d_hypoxia_4hr"]],      "EIF3D", "si3d hypoxia 4hr"),
  direction_check(te_list[["eIF3d_hypoxia_24hr"]],     "EIF3D", "si3d hypoxia 24hr"),
  direction_check(te_list[["eIF3d_normoxia_1and4hr"]], "EIF3D", "si3d normoxia 1+4hr"),
  direction_check(te_list[["eIF3e_hypoxia_1hr"]],      "EIF3E", "si3e hypoxia 1hr"),
  direction_check(te_list[["eIF3e_hypoxia_4hr"]],      "EIF3E", "si3e hypoxia 4hr"),
  direction_check(te_list[["eIF3e_hypoxia_24hr"]],     "EIF3E", "si3e hypoxia 24hr"),
  direction_check(read_csv(here("output/rna_si3e_vs_sictrl_normoxia_1and4hr.csv"),
                           show_col_types = FALSE),
                  "EIF3E", "si3e normoxia 1+4hr")
)

# all knockdowns must show a large positive RNA LFC for their own target
stopifnot("knockdown direction flipped" = all(kd_rna_lfc > 2))
cat("\nDirection confirmed: positive te_lfc = eIF3d/eIF3e promotes\n")
```

# Count significant TE changes

Identical rule to the source notebook's `count_te_changes()`:
`padj < 0.05` and `te_lfc > +0.5` (Up) or `te_lfc < -0.5` (Down).

```{r count}
count_te_changes <- function(df, padj_cutoff = 0.05, lfc_cutoff = 0.5) {
  tibble(
    Direction = c("Up", "Down"),
    Count = c(
      sum(df$te_padj < padj_cutoff & df$te_lfc >  lfc_cutoff, na.rm = TRUE),
      sum(df$te_padj < padj_cutoff & df$te_lfc < -lfc_cutoff, na.rm = TRUE)
    )
  )
}

te_counts <- te_files %>%
  mutate(counts = map(paste(gene, contrast, sep = "_"),
                      ~ count_te_changes(te_list[[.x]], padj_cutoff, lfc_cutoff))) %>%
  unnest(counts) %>%
  mutate(
    x_label   = factor(x_label, levels = c("Norm", "1", "4", "24")),
    Direction = factor(Direction, levels = c("Down", "Up"))
  )

te_counts %>%
  dplyr::select(gene, contrast, Direction, Count) %>%
  pivot_wider(names_from = Direction, values_from = Count) %>%
  print(n = Inf)
```

## Regression check against the published bar plot

The six hypoxia bars must reproduce the values in the existing
`plots/eif3{d,e}_te_changes_barplot_hypoxia.pdf` exactly. If a saved CSV is ever
regenerated with different settings, this fails loudly rather than silently redrawing
a different figure.

```{r regression_check}
expected_hypoxia <- tribble(
  ~gene,   ~contrast,      ~Direction, ~expected,
  "eIF3d", "hypoxia_1hr",  "Up",   1604,
  "eIF3d", "hypoxia_1hr",  "Down",   44,
  "eIF3d", "hypoxia_4hr",  "Up",     42,
  "eIF3d", "hypoxia_4hr",  "Down",   32,
  "eIF3d", "hypoxia_24hr", "Up",     10,
  "eIF3d", "hypoxia_24hr", "Down",   70,
  "eIF3e", "hypoxia_1hr",  "Up",   1083,
  "eIF3e", "hypoxia_1hr",  "Down",   26,
  "eIF3e", "hypoxia_4hr",  "Up",     40,
  "eIF3e", "hypoxia_4hr",  "Down",   26,
  "eIF3e", "hypoxia_24hr", "Up",     31,
  "eIF3e", "hypoxia_24hr", "Down", 5092
)

check <- expected_hypoxia %>%
  left_join(te_counts %>%
              mutate(Direction = as.character(Direction)) %>%
              dplyr::select(gene, contrast, Direction, observed = Count),
            by = c("gene", "contrast", "Direction"))

print(check)
stopifnot("hypoxia bars do not reproduce the published figure" =
  all(check$expected == check$observed))
cat("All 12 hypoxia bars reproduce the published figure.\n")
```

## Global TE shift per contrast (interpretation guard)

A fixed `|LFC| > 0.5` cut only means "a gene-selective change" if the contrast is
centred near zero. It is not for si3e normoxia, where the whole TE distribution is
shifted up. Print the median so the bar heights are read in context.

```{r global_shift}
te_medians <- te_files %>%
  mutate(key           = paste(gene, contrast, sep = "_"),
         median_te_lfc = map_dbl(key, ~ median(te_list[[.x]]$te_lfc, na.rm = TRUE)),
         n_padj_sig    = map_int(key, ~ sum(te_list[[.x]]$te_padj < padj_cutoff, na.rm = TRUE))) %>%
  dplyr::select(gene, contrast, median_te_lfc, n_padj_sig)

print(te_medians)

# si3e normoxia 1+4hr sits above the +0.5 cut on its own median - flag it explicitly
stopifnot("expected si3e normoxia median above the LFC cut" =
  te_medians$median_te_lfc[te_medians$gene == "eIF3e" &
                           te_medians$contrast == "normoxia_1and4hr"] > lfc_cutoff)
cat("\nNOTE: si3e normoxia 1+4hr median te_lfc exceeds the +0.5 cut - that bar is\n",
    "dominated by a transcriptome-wide shift, not gene-selective regulation.\n", sep = "")
```

# Bar plots

```{r plots, fig.width = 6, fig.height = 5}
dir_colors <- c("Up" = "#D55E00", "Down" = "#0072B2")

make_barplot <- function(df, gene_label) {
  ggplot(df, aes(x = x_label, y = Count, fill = Direction)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.9),
              vjust = -0.5, size = 4) +
    scale_fill_manual(values = dir_colors, breaks = c("Down", "Up")) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    theme_prism() +
    theme(legend.position = "top") +
    labs(
      title    = paste0(gene_label, "-induced TE changes"),
      subtitle = "padj < 0.05, |LFC| > 0.5\nUp = TE higher in sictrl (promoted)",
      y        = "Number of genes",
      x        = "Normoxia (time-regressed) / hypoxia hour"
    )
}

p_3d <- make_barplot(filter(te_counts, gene == "eIF3d"), "eIF3d")
p_3e <- make_barplot(filter(te_counts, gene == "eIF3e"), "eIF3e")

print(p_3d)
print(p_3e)
```

```{r combined, fig.width = 6, fig.height = 9}
p_combined <- p_3d / p_3e
print(p_combined)
```

## Faceted panel: eIF3e (left) / eIF3d (right)

Single figure, one facet per knockdown, **eIF3e on the left and eIF3d on the right**.
The y-axis is free because the eIF3e counts run ~3.5x higher than eIF3d, so a shared
axis would flatten the entire eIF3d panel.

```{r faceted, fig.width = 9, fig.height = 5}
te_counts_faceted <- te_counts %>%
  mutate(gene = factor(gene, levels = c("eIF3e", "eIF3d")))  # left = 3e, right = 3d
stopifnot("facet order must be eIF3e then eIF3d" =
  identical(levels(te_counts_faceted$gene), c("eIF3e", "eIF3d")))

p_facet <- ggplot(te_counts_faceted, aes(x = x_label, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5) +
  facet_wrap(~ gene, nrow = 1, scales = "free_y") +
  scale_fill_manual(values = dir_colors, breaks = c("Down", "Up")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  theme_prism() +
  theme(legend.position = "top") +
  labs(
    title    = "eIF3e- and eIF3d-induced TE changes",
    subtitle = "padj < 0.05, |LFC| > 0.5\nUp = TE higher in sictrl (promoted); free y-axis per panel",
    y        = "Number of genes",
    x        = "Normoxia (time-regressed) / hypoxia hour"
  )

print(p_facet)
```

```{r save}
ggsave(here("plots", "86_eif3d_te_changes_barplot_hypoxia_plus_normoxia_regressed.pdf"),
       p_3d, width = 6, height = 5)
ggsave(here("plots", "86_eif3e_te_changes_barplot_hypoxia_plus_normoxia_regressed.pdf"),
       p_3e, width = 6, height = 5)
ggsave(here("plots", "86_te_changes_barplot_hypoxia_plus_normoxia_regressed_combined.pdf"),
       p_combined, width = 6, height = 9)
ggsave(here("plots", "86_te_changes_barplot_faceted_3e_left_3d_right.pdf"),
       p_facet, width = 9, height = 5)

write_csv(te_counts %>% dplyr::select(gene, contrast, x_label, Direction, Count),
          here("output", "86_te_change_counts_hypoxia_plus_normoxia_regressed.csv"))
cat("Saved plots/86_*.pdf and output/86_te_change_counts_*.csv\n")
```

# Notes on reading these bars

- **Up** = `te_lfc > +0.5`, i.e. TE higher in sictrl than in knockdown = the factor
  **promotes** those mRNAs. **Down** = the factor represses them. Legend order is
  Down/Up to match the original figure.
- The normoxia bar is a **single time-regressed contrast fit over the 1hr and 4hr
  samples together** (`hours` as a batch covariate), not an average of two
  per-timepoint contrasts. It has roughly twice the samples of a hypoxia bar, so it is
  better powered and its height is not a like-for-like comparison against a single
  hypoxia timepoint.
- The si3e normoxia 1+4hr bar sits on a distribution whose median `te_lfc` is +0.62,
  above the +0.5 cut, so most of that bar is a transcriptome-wide shift rather than
  gene-selective regulation. si3d normoxia (median +0.16) and the hypoxia 1hr
  contrasts (+0.29 si3d, +0.13 si3e) do not have this problem. Do not compare the si3e
  normoxia bar to the others without centring first.
