---
title: "Notebook 87: DHX29 KO/KD RNA-seq, GC3 dependence and eIF3d-promoted mRNAs"
output: html_document
date: "`r Sys.Date()`"
params:
  n_up: 500
  n_down: 500
  dhx29_contrast: "kd"
  down_mode: "extreme"
  padj_cut: 0.05
  unchanged_padj: 0.30
  unchanged_max_abs_lfc: 0.10
  n_slam_up: 500
  n_slam_down: 500
  write_genesets: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# `params` only exists during knit; supply the same defaults so chunks can be run
# interactively in RStudio.
#
# Interactively we REBUILD params rather than only filling it in when absent. A stale
# `params` left in the global environment by another notebook, or by an earlier version
# of this one, is a list and so passed an `exists()`-only guard - then a renamed field
# reads back NULL, `as.integer(NULL)` is integer(0), `&&` on that yields NA, and the
# failure surfaces as a misleading "must be a positive integer" much further down.
default_params <- list(
  n_up                  = 500,
  n_down                = 500,
  dhx29_contrast        = "kd",
  down_mode             = "extreme",
  padj_cut              = 0.05,
  unchanged_padj        = 0.30,
  unchanged_max_abs_lfc = 0.10,
  n_slam_up             = 500,
  n_slam_down           = 500,
  write_genesets        = TRUE
)

if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- default_params
}

# During knit params comes from the YAML above. Assert it is complete rather than
# letting a missing field become a length-zero read later.
missing_params <- setdiff(names(default_params), names(params))
stopifnot(
  "params is not a list - a stale object is masking it, run rm(params)" = is.list(params),
  "params is missing fields; if running interactively, run rm(params) and re-run setup" =
    length(missing_params) == 0
)
stopifnot(
  "every param must be a single non-NA value" =
    all(vapply(params[names(default_params)],
               function(x) length(x) == 1 && !is.na(x), logical(1)))
)
```

## Overview

Hia et al. 2026 (Science, adw0288) depleted DHX29 in K562 cells two ways and profiled
steady-state mRNA by RNA-seq:

- **Table S2** — `DHX29KDvsCtrl`: CRISPRi knockdown, assayed shortly after gRNA transduction
  (acute depletion). Keyed by versioned ENSEMBL gene ID.
- **Table S3** — `DHX29KOvsCtrl`: CRISPR knockout, maintained long-term as single-cell clones.
  Keyed by ENTREZ + unversioned ENSEMBL + SYMBOL.

Both are limma `logFC` for **depleted vs control**, so **positive logFC = the mRNA accumulates
when DHX29 is lost**. The paper's interpretation is that those transcripts are normally
subject to DHX29-dependent, codon-driven decay: DHX29 resolves stalls on AU3-rich (GC3-poor)
codons, and losing it stabilises exactly those mRNAs (Fig 1F, fig. S2E).

> **Caveat on the premise, and it does not survive checking.** Positive logFC here is a
> *steady-state RNA* change, so it is consistent with reduced decay but does not by itself
> exclude a transcriptional contribution. The paper's only direct stability measurement is
> the SLAM-seq half-life time course (Table S6, KO only). Section 7a tests whether the two
> agree per gene and finds they do not: Spearman = 0.030. **These RNA-seq tables therefore
> measure abundance, not decay**, and nothing derived from them is named or described as a
> stability readout anywhere in this notebook.

This notebook does four things:

1. **Recreate Fig 1F** — GC3-poor mRNAs are up in both contrasts, more strongly in the KD.
2. **Ask whether eIF3d-promoted mRNAs respond to DHX29 loss** — density plots of KO and KD
   logFC for eIF3d hypoxia-promoted genes (1hr, |LFC| > 0.5) vs the size-matched negative
   control vs the rest of the transcriptome, followed by a GC3-stratified decomposition.
3. **Export abundance-defined gene sets** for notebook 07 (Section 6).
4. **Repeat both on the KO SLAM-seq half-life data** — the paper's actual stability readout —
   and export stability-defined gene sets (Section 7).

```{r load_libraries}
library(tidyverse)
library(here)
library(readxl)
library(ggplot2)

# Wilcoxon p-values in these comparisons underflow to exactly 0 in double precision
# (n ~ 4500 per group). signif(0, 3) prints "0", which claims a p-value no test can
# return. Floor the display at R's own printing threshold instead.
fmt_p <- function(p) {
  ifelse(p < 2.2e-16, "< 2.2e-16", paste("=", signif(p, 3)))
}
```

---

## Section 1: Load the RNA-seq tables

```{r load_tables}
s_dir <- here("accessories", "csc_data", "hia_dhx29",
              "science.adw0288_tables_s1_to_s11")

s3_path <- file.path(s_dir, "science.adw0288_table_s3.xlsx")
s2_path <- file.path(s_dir, "science.adw0288_table_s2.xlsx")

stopifnot(
  "Table S3 (DHX29 KO RNA-seq) not found" = file.exists(s3_path),
  "Table S2 (DHX29 KD RNA-seq) not found" = file.exists(s2_path)
)

cat("S3 sheets:", paste(excel_sheets(s3_path), collapse = ", "), "\n")
cat("S2 sheets:", paste(excel_sheets(s2_path), collapse = ", "), "\n")

s3_raw <- read_excel(s3_path)
s2_raw <- read_excel(s2_path)

cat("\nS3 (KO) dim:", dim(s3_raw), "| cols:", paste(colnames(s3_raw), collapse = ", "), "\n")
cat("S2 (KD) dim:", dim(s2_raw), "| cols:", paste(colnames(s2_raw), collapse = ", "), "\n")
```

### 1a: Replication check at the point of load

The single cheapest quantitative check on these files is **DHX29's own logFC**: the
perturbation depletes DHX29, so DHX29 must be significantly *down* in both contrasts. If the
sign is flipped, the contrast direction is reversed and every conclusion below inverts.

Observed: KO logFC = **-0.958** (adj.P = 6.9e-101), KD logFC = **-2.645** (adj.P = 1.3e-13).
The KD is the deeper depletion, matching the paper's statement that the KD reflects the acute
consequence while KO clones may have adapted.

```{r direction_check}
dhx29_ens <- "ENSG00000067248"

dhx29_ko <- s3_raw %>% dplyr::filter(SYMBOL == "DHX29")
dhx29_kd <- s2_raw %>% dplyr::filter(str_detect(Gene, dhx29_ens))

cat("DHX29 in KO table: logFC =", round(dhx29_ko$logFC, 3),
    " adj.P =", format.pval(dhx29_ko$adj.P.Val, digits = 3), "\n")
cat("DHX29 in KD table: logFC =", round(dhx29_kd$logFC, 3),
    " adj.P =", format.pval(dhx29_kd$adj.P.Val, digits = 3), "\n")

stopifnot(
  "DHX29 must appear exactly once in the KO table" = nrow(dhx29_ko) == 1,
  "DHX29 must appear exactly once in the KD table" = nrow(dhx29_kd) == 1,
  "Direction check failed: DHX29 is not down in the KO contrast" =
    dhx29_ko$logFC < -0.5 && dhx29_ko$adj.P.Val < 0.01,
  "Direction check failed: DHX29 is not down in the KD contrast" =
    dhx29_kd$logFC < -1.0 && dhx29_kd$adj.P.Val < 0.01,
  "KD is expected to be the deeper depletion of the two" =
    dhx29_kd$logFC < dhx29_ko$logFC
)
cat("\nDirection check PASS - both contrasts are depleted-vs-control ",
    "(positive logFC = up on DHX29 loss)\n", sep = "")
```

### 1b: Tidy to one row per gene

The KO table is Entrez-keyed, so `ENSEMBL` carries the literal string `"NA"` for 118 rows
with no ENSEMBL mapping, plus one gene (`ENSG00000230124`) that two Entrez IDs map onto.
Drop the unmapped rows and keep the more significant row for the genuine duplicate.

```{r tidy_tables}
ko <- s3_raw %>%
  dplyr::filter(!is.na(ENSEMBL), ENSEMBL != "NA") %>%
  transmute(
    ensembl_gene = ENSEMBL,
    symbol_ko    = SYMBOL,
    logFC_ko     = logFC,
    padj_ko      = adj.P.Val
  ) %>%
  group_by(ensembl_gene) %>%
  slice_min(padj_ko, n = 1, with_ties = FALSE) %>%
  ungroup()

kd <- s2_raw %>%
  transmute(
    ensembl_gene = str_remove(Gene, "\\..*"),
    logFC_kd     = logFC,
    padj_kd      = adj.P.Val
  )

cat("KO genes (unique ENSEMBL):", nrow(ko), "\n")
cat("KD genes (unique ENSEMBL):", nrow(kd), "\n")
cat("KO median logFC:", round(median(ko$logFC_ko), 4),
    "| KD median logFC:", round(median(kd$logFC_kd), 4), "\n")
cat("KO n adj.P < 0.05:", sum(ko$padj_ko < 0.05),
    "| KD n adj.P < 0.05:", sum(kd$padj_kd < 0.05), "\n")

stopifnot(
  "KO gene IDs must be unique after dedup" = !any(duplicated(ko$ensembl_gene)),
  "KD gene IDs must be unique"             = !any(duplicated(kd$ensembl_gene)),
  "KO logFC must be finite"  = all(is.finite(ko$logFC_ko)),
  "KD logFC must be finite"  = all(is.finite(kd$logFC_kd)),
  "adj.P values must lie in [0, 1]" =
    all(ko$padj_ko >= 0 & ko$padj_ko <= 1) && all(kd$padj_kd >= 0 & kd$padj_kd <= 1),
  "logFC magnitudes are implausible - wrong column or wrong sheet" =
    max(abs(ko$logFC_ko)) < 10 && max(abs(kd$logFC_kd)) < 10,
  "Both tables should be near-centred on zero" =
    abs(median(ko$logFC_ko)) < 0.2 && abs(median(kd$logFC_kd)) < 0.2
)
```

---

## Section 2: Join per-gene GC3

GC3 comes from the existing positional-GC feature matrix (notebook `01v_`), which stores
`gc1`/`gc2`/`gc3` as **percentages** and `gc3_residual` as GC3 orthogonalised against
`cds_gc`. `cds_gc` itself is dropped from that matrix by design, so pull it from
`feature_matrix_external_stability.rds` for the aggregate-GC control.

Feature matrices are transcript-keyed, so deduplicate to one row per gene first.

```{r load_gc}
gc_path <- here("output", "predictive_modeling", "feature_matrix_positional_gc.rds")
es_path <- here("output", "predictive_modeling", "feature_matrix_external_stability.rds")
stopifnot(
  "feature_matrix_positional_gc.rds not found - run 01v_ first"      = file.exists(gc_path),
  "feature_matrix_external_stability.rds not found - run 01i_-01k_"  = file.exists(es_path)
)

gc_fm <- readRDS(gc_path) %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  transmute(
    ensembl_gene = gene_id_clean,
    symbol,
    gc1, gc2, gc3, gc3_residual,
    cds_length
  )

cds_gc_fm <- readRDS(es_path) %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  transmute(ensembl_gene = gene_id_clean, cds_gc)

gc_fm <- left_join(gc_fm, cds_gc_fm, by = "ensembl_gene")

cat("Genes with positional GC:", nrow(gc_fm), "\n")
print(summary(gc_fm %>% dplyr::select(gc1, gc2, gc3, cds_gc)))

stopifnot(
  "GC percentages must lie in [0, 100]" =
    all(c(gc_fm$gc1, gc_fm$gc2, gc_fm$gc3) >= 0 &
        c(gc_fm$gc1, gc_fm$gc2, gc_fm$gc3) <= 100, na.rm = TRUE),
  "cds_gc must lie in [0, 100]" =
    all(gc_fm$cds_gc >= 0 & gc_fm$cds_gc <= 100, na.rm = TRUE),
  # Project convention: GC3 median in [40, 65]% (01_, 43_)
  "GC3 median outside the expected 40-65% window" =
    dplyr::between(median(gc_fm$gc3, na.rm = TRUE), 40, 65),
  "gc3_residual must be mean-zero by construction" =
    abs(mean(gc_fm$gc3_residual, na.rm = TRUE)) < 1e-6,
  "cds_gc failed to join from the external-stability matrix" =
    mean(!is.na(gc_fm$cds_gc)) > 0.9
)
```

```{r join_gc_dhx29}
dhx <- gc_fm %>%
  inner_join(ko, by = "ensembl_gene") %>%
  inner_join(kd, by = "ensembl_gene") %>%
  dplyr::filter(!is.na(gc3), !is.na(logFC_ko), !is.na(logFC_kd))

cat("Genes with GC3 + KO + KD logFC:", nrow(dhx), "\n")
stopifnot("Too few genes joined - check ID versioning" = nrow(dhx) > 5000)
```

---

## Section 3: Recreate Fig 1F - GC3-poor mRNAs accumulate on DHX29 loss

### 3a: Rank correlations

```{r gc3_correlations}
cor_row <- function(x, y, xlab, ylab) {
  ct <- suppressWarnings(cor.test(x, y, method = "spearman"))
  tibble(
    feature  = xlab,
    contrast = ylab,
    rho      = unname(ct$estimate),
    p        = ct$p.value,
    n        = sum(complete.cases(x, y))
  )
}

cor_tbl <- bind_rows(
  cor_row(dhx$gc3,          dhx$logFC_ko, "GC3",          "DHX29 KO"),
  cor_row(dhx$gc3,          dhx$logFC_kd, "GC3",          "DHX29 KD"),
  cor_row(dhx$cds_gc,       dhx$logFC_ko, "CDS GC",       "DHX29 KO"),
  cor_row(dhx$cds_gc,       dhx$logFC_kd, "CDS GC",       "DHX29 KD"),
  cor_row(dhx$gc3_residual, dhx$logFC_ko, "GC3 residual", "DHX29 KO"),
  cor_row(dhx$gc3_residual, dhx$logFC_kd, "GC3 residual", "DHX29 KD")
) %>%
  mutate(rho = round(rho, 3), p = format.pval(p, digits = 3))

print(as.data.frame(cor_tbl))

rho_gc3_ko <- cor_tbl$rho[cor_tbl$feature == "GC3" & cor_tbl$contrast == "DHX29 KO"]
rho_gc3_kd <- cor_tbl$rho[cor_tbl$feature == "GC3" & cor_tbl$contrast == "DHX29 KD"]

stopifnot(
  "Fig 1F replication failed: GC3 does not negatively track KO logFC" = rho_gc3_ko < -0.05,
  "Fig 1F replication failed: GC3 does not negatively track KD logFC" = rho_gc3_kd < -0.30,
  "Paper's claim failed: the KD effect should exceed the KO effect" =
    rho_gc3_kd < rho_gc3_ko
)
cat("\nFig 1F PASS - Spearman(GC3, logFC): KO =", rho_gc3_ko,
    ", KD =", rho_gc3_kd, " (n =", nrow(dhx), ")\n")
```

`GC3 residual` (GC3 with `cds_gc` regressed out) keeps the same sign but is much weaker than
raw GC3, and raw `cds_gc` tracks logFC nearly as well as GC3 does. As documented in CLAUDE.md,
aggregate GC3 is ~0.93-correlated with `cds_gc` via isochore structure, so **these RNA-seq
tables cannot separate a third-position (wobble) effect from bulk CDS GC.** The AU3-specific
claim rests on the codon-level Thor-Ribo-seq (Table S8, notebook 23), not on this figure.

### 3b: logFC by GC3 bin, KO vs KD

```{r gc3_bins}
gc3_breaks <- c(0, 40, 50, 60, 70, 80, 100)
gc3_labels <- c("<40", "40-50", "50-60", "60-70", "70-80", ">80")

dhx_long <- dhx %>%
  dplyr::select(ensembl_gene, symbol, gc3, logFC_ko, logFC_kd) %>%
  pivot_longer(c(logFC_ko, logFC_kd), names_to = "contrast", values_to = "logFC") %>%
  mutate(
    contrast = recode(contrast,
                      logFC_ko = "DHX29 KO (clonal)",
                      logFC_kd = "DHX29 KD (CRISPRi, acute)"),
    contrast = factor(contrast, levels = c("DHX29 KO (clonal)",
                                           "DHX29 KD (CRISPRi, acute)")),
    gc3_bin  = cut(gc3, breaks = gc3_breaks, labels = gc3_labels, include.lowest = TRUE)
  )

bin_summary <- dhx_long %>%
  group_by(contrast, gc3_bin) %>%
  summarise(n = n(), median_logFC = median(logFC), .groups = "drop")

cat("Median logFC per GC3 bin:\n")
print(as.data.frame(bin_summary))

stopifnot(
  "All GC3 bins must be populated" = all(!is.na(bin_summary$gc3_bin)),
  "Every GC3 bin needs enough genes to read a median" = all(bin_summary$n >= 100)
)

# Monotonicity of the paper's claim, tested at the extremes of each contrast.
extremes <- bin_summary %>%
  dplyr::filter(gc3_bin %in% c("<40", ">80")) %>%
  dplyr::select(contrast, gc3_bin, median_logFC) %>%
  pivot_wider(names_from = gc3_bin, values_from = median_logFC)
print(as.data.frame(extremes))

stopifnot(
  "Fig 1F replication failed: lowest-GC3 bin is not above the highest-GC3 bin" =
    all(extremes$`<40` > extremes$`>80`)
)
```

```{r gc3_bin_plot, fig.width=9, fig.height=4.5}
contrast_colors <- c("DHX29 KO (clonal)"         = "#0072B2",
                     "DHX29 KD (CRISPRi, acute)" = "#CC79A7")

p_bins <- ggplot(dhx_long, aes(x = gc3_bin, y = logFC, fill = contrast)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.2, linewidth = 0.4) +
  geom_text(data = bin_summary, aes(x = gc3_bin, y = -Inf, label = paste0("n=", n)),
            inherit.aes = FALSE, vjust = -0.6, size = 2.6, color = "grey40") +
  facet_wrap(~ contrast, nrow = 1) +
  scale_fill_manual(values = contrast_colors, guide = "none") +
  coord_cartesian(ylim = quantile(dhx_long$logFC, c(0.002, 0.998))) +
  labs(
    title    = "GC3-poor mRNAs accumulate when DHX29 is depleted (recreates Fig 1F)",
    subtitle = paste0("Positive logFC = up on DHX29 loss. Spearman(GC3, logFC): KO = ",
                      rho_gc3_ko, ", KD = ", rho_gc3_kd, "; n = ", nrow(dhx), " genes"),
    x = "GC3 content bin (%)",
    y = "logFC (DHX29-depleted vs control)"
  ) +
  theme_classic(base_size = 12)

p_bins
ggsave(here("plots", "87_dhx29_rnaseq_logfc_by_gc3_bin.pdf"), p_bins,
       width = 9, height = 4.5)
cat("Saved: plots/87_dhx29_rnaseq_logfc_by_gc3_bin.pdf\n")
```

### 3c: GC3-quartile CDFs

```{r gc3_quartile_cdf, fig.width=9, fig.height=4.5}
dhx_q <- dhx_long %>%
  group_by(contrast) %>%
  mutate(gc3_q = ntile(gc3, 4)) %>%
  ungroup() %>%
  mutate(gc3_q = factor(gc3_q, levels = 1:4,
                        labels = c("Q1 (most AU3-rich)", "Q2", "Q3",
                                   "Q4 (most GC3-rich)")))

q_stats <- dhx_q %>%
  group_by(contrast, gc3_q) %>%
  summarise(n = n(), median_gc3 = round(median(gc3), 1),
            median_logFC = round(median(logFC), 3), .groups = "drop")
cat("GC3 quartiles:\n")
print(as.data.frame(q_stats))

q_test <- dhx_q %>%
  dplyr::filter(gc3_q %in% c("Q1 (most AU3-rich)", "Q4 (most GC3-rich)")) %>%
  group_by(contrast) %>%
  summarise(
    p = wilcox.test(logFC ~ droplevels(gc3_q))$p.value,
    delta_median = median(logFC[gc3_q == "Q1 (most AU3-rich)"]) -
                   median(logFC[gc3_q == "Q4 (most GC3-rich)"]),
    .groups = "drop"
  )
cat("\nQ1 vs Q4 Wilcoxon:\n")
print(as.data.frame(q_test %>% mutate(p = format.pval(p, digits = 3),
                                      delta_median = round(delta_median, 3))))

stopifnot(
  "Q1 (AU3-rich) must exceed Q4 (GC3-rich) in both contrasts" = all(q_test$delta_median > 0),
  "Q1 vs Q4 separation must be significant in both contrasts" = all(q_test$p < 1e-5),
  "KD must show the larger Q1-Q4 shift than KO" =
    q_test$delta_median[q_test$contrast == "DHX29 KD (CRISPRi, acute)"] >
    q_test$delta_median[q_test$contrast == "DHX29 KO (clonal)"]
)

quartile_colors <- c("Q1 (most AU3-rich)" = "#D55E00", "Q2" = "grey60",
                     "Q3" = "grey40", "Q4 (most GC3-rich)" = "#0072B2")

q_label <- q_test %>%
  mutate(label = paste0("Q1 vs Q4 Wilcoxon p ", fmt_p(p),
                        "\ndelta median = ", round(delta_median, 3),
                        "\nn per quartile = ", round(nrow(dhx) / 4)))

p_cdf <- ggplot(dhx_q, aes(x = logFC, color = gc3_q)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
  geom_text(data = q_label, aes(x = Inf, y = -Inf, label = label),
            inherit.aes = FALSE, hjust = 1.05, vjust = -0.3, size = 3) +
  facet_wrap(~ contrast, nrow = 1) +
  scale_color_manual(values = quartile_colors) +
  coord_cartesian(xlim = quantile(dhx_long$logFC, c(0.005, 0.995))) +
  labs(
    title    = "Cumulative distribution of DHX29-depletion logFC by GC3 quartile",
    subtitle = "Leftward shift with rising GC3 = GC3-rich mRNAs are not stabilised on DHX29 loss",
    x = "logFC (DHX29-depleted vs control)", y = "Cumulative fraction", color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_cdf
ggsave(here("plots", "87_dhx29_rnaseq_logfc_cdf_by_gc3_quartile.pdf"), p_cdf,
       width = 9, height = 4.5)
cat("Saved: plots/87_dhx29_rnaseq_logfc_cdf_by_gc3_quartile.pdf\n")
```

---

## Section 4: Do eIF3d-promoted mRNAs respond to DHX29 loss?

### 4a: Gene groups

Same construction as notebook 23: eIF3d hypoxia-promoted (1hr, |TE LFC| > 0.5), a
size-matched negative control drawn under `set.seed(9)` from the negative pool with any
overlap removed, and everything else as background.

```{r gene_groups}
promotes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"),
                     show_col_types = FALSE)
promotes_ids <- promotes$ensembl_gene

neg_pool <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                     show_col_types = FALSE) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  dplyr::filter(!ensembl_gene %in% promotes_ids)

set.seed(9)
neg_ctrl_ids <- sample(neg_pool$ensembl_gene, size = length(promotes_ids), replace = FALSE)

cat("Promotes genes:", length(promotes_ids), "\n")
cat("Neg ctrl genes (size-matched):", length(neg_ctrl_ids), "\n")

stopifnot(
  "Gene sets must not overlap" = length(intersect(promotes_ids, neg_ctrl_ids)) == 0,
  "Negative control must be size-matched" =
    length(neg_ctrl_ids) == length(promotes_ids)
)

set_colors <- c(
  "eIF3d hypoxia promotes" = "#E69F00",   # Okabe-Ito orange, reserved for eIF3d
  "Negative control"       = "black",
  "All other genes"        = "grey70"
)

dhx <- dhx %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  ))

group_summary <- dhx %>%
  group_by(group) %>%
  summarise(n = n(),
            median_gc3   = round(median(gc3), 1),
            median_ko    = round(median(logFC_ko), 3),
            median_kd    = round(median(logFC_kd), 3),
            .groups = "drop")
cat("\nGroup summary:\n")
print(as.data.frame(group_summary))

stopifnot(
  "All three groups must be represented in the DHX29 universe" =
    setequal(unique(dhx$group), names(set_colors)),
  "Each group needs enough genes for a density estimate" = all(group_summary$n >= 50)
)
```

### 4b: Density plots

```{r density_helper}
# Corner annotation carrying the test result and all three group sizes, so each
# figure is self-contained.
stat_label <- function(p_value, x_promotes, x_neg, x_other) {
  paste0(
    "Wilcoxon p ", fmt_p(p_value),
    "\nn promotes = ", sum(!is.na(x_promotes)),
    "\nn neg ctrl = ", sum(!is.na(x_neg)),
    "\nn other = ",    sum(!is.na(x_other))
  )
}

density_by_group <- function(df, value_col, title, subtitle, xlab, outfile) {
  v  <- rlang::sym(value_col)
  bg <- df %>% dplyr::filter(group == "All other genes")
  fg <- df %>% dplyr::filter(group != "All other genes")

  med <- fg %>% group_by(group) %>%
    summarise(med = median(!!v, na.rm = TRUE), .groups = "drop")

  x_prom <- df %>% dplyr::filter(group == "eIF3d hypoxia promotes") %>% pull(!!v)
  x_neg  <- df %>% dplyr::filter(group == "Negative control")       %>% pull(!!v)
  wt     <- wilcox.test(x_prom, x_neg, alternative = "two.sided")

  # Outline only - no fills. The background group is drawn as a line as well, so all
  # three curves are read the same way.
  p <- ggplot() +
    geom_density(data = bg, aes(x = !!v, color = group),
                 fill = NA, linewidth = 0.8, key_glyph = "path") +
    geom_density(data = fg %>% dplyr::filter(group == "Negative control"),
                 aes(x = !!v, color = group),
                 fill = NA, linewidth = 0.8, key_glyph = "path") +
    geom_density(data = fg %>% dplyr::filter(group == "eIF3d hypoxia promotes"),
                 aes(x = !!v, color = group),
                 fill = NA, linewidth = 0.8, key_glyph = "path") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
    geom_vline(data = med, aes(xintercept = med, color = group),
               linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3,
             label = stat_label(wt$p.value, x_prom, x_neg, bg %>% pull(!!v))) +
    scale_color_manual(values = set_colors) +
    coord_cartesian(xlim = quantile(df %>% pull(!!v), c(0.002, 0.998), na.rm = TRUE)) +
    labs(title = title, subtitle = subtitle, x = xlab, y = "Density", color = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")

  ggsave(here("plots", outfile), p, width = 6.5, height = 5)
  cat("Saved: plots/", outfile, "\n", sep = "")

  list(plot = p, wilcox = wt, medians = med)
}
```

```{r density_kd, fig.width=6.5, fig.height=5}
res_kd <- density_by_group(
  dhx, "logFC_kd",
  title    = "DHX29 knockdown (CRISPRi): eIF3d-promoted genes (hypoxia 1hr)",
  subtitle = paste0("Positive logFC = mRNA accumulates on DHX29 loss\n",
                    "Dashed vertical lines = group medians"),
  xlab     = "logFC (DHX29 KD vs control)",
  outfile  = "87_dhx29_kd_logfc_by_gene_group.pdf"
)
res_kd$plot
```

```{r density_ko, fig.width=6.5, fig.height=5}
res_ko <- density_by_group(
  dhx, "logFC_ko",
  title    = "DHX29 knockout (clonal): eIF3d-promoted genes (hypoxia 1hr)",
  subtitle = paste0("Positive logFC = mRNA accumulates on DHX29 loss\n",
                    "Dashed vertical lines = group medians"),
  xlab     = "logFC (DHX29 KO vs control)",
  outfile  = "87_dhx29_ko_logfc_by_gene_group.pdf"
)
res_ko$plot
```

### 4c: Full test panel

```{r group_tests}
group_tests <- map_dfr(c(logFC_kd = "DHX29 KD", logFC_ko = "DHX29 KO") %>%
                         imap(~ list(col = .y, lab = .x)), function(z) {
  x_prom  <- dhx[[z$col]][dhx$group == "eIF3d hypoxia promotes"]
  x_neg   <- dhx[[z$col]][dhx$group == "Negative control"]
  x_other <- dhx[[z$col]][dhx$group == "All other genes"]
  tibble(
    contrast          = z$lab,
    median_promotes   = median(x_prom),
    median_neg_ctrl   = median(x_neg),
    median_other      = median(x_other),
    p_prom_vs_neg     = wilcox.test(x_prom, x_neg)$p.value,
    p_prom_vs_other   = wilcox.test(x_prom, x_other)$p.value,
    n_prom = length(x_prom), n_neg = length(x_neg), n_other = length(x_other)
  )
})

print(as.data.frame(group_tests %>%
  mutate(across(starts_with("median"), ~ round(.x, 4)),
         across(starts_with("p_"), ~ format.pval(.x, digits = 3)))))
```

> **Result.** eIF3d-promoted mRNAs are **not** preferentially stabilised on DHX29 loss.
> In the KD their median logFC is indistinguishable from the negative control
> (medians -0.068 vs -0.071, p = 0.39), and in the KO they sit **below** both the negative
> control and the background (-0.058 vs +0.014 and -0.011, p = 2e-15 and 8e-13) - the
> opposite of the DHX29-substrate direction.
>
> This is striking because eIF3d-promoted genes are markedly **AU3-rich** (median GC3 48.6%
> vs 60.3% for the negative control and 57.3% for background). On GC3 alone they should be
> the *most* DHX29-responsive set in the transcriptome. Section 5 decomposes this.

---

## Section 5: Why the GC3 signal does not carry over - stratified decomposition

If eIF3d-promoted genes were ordinary AU3-rich DHX29 substrates, they should track the
background *within* each GC3 stratum. Splitting the transcriptome into GC3 quintiles and
comparing groups inside each one separates composition (they are AU3-rich) from behaviour
(how they respond at matched GC3).

```{r gc3_stratified}
dhx_strat <- dhx %>%
  mutate(gc3_quintile = cut(gc3, breaks = quantile(gc3, 0:5 / 5),
                            include.lowest = TRUE,
                            labels = paste0("GC3 Q", 1:5)))

strat_summary <- dhx_strat %>%
  group_by(gc3_quintile, group) %>%
  summarise(n = n(),
            median_kd = round(median(logFC_kd), 3),
            median_ko = round(median(logFC_ko), 3),
            .groups = "drop")

cat("Median logFC by GC3 quintile and gene group:\n")
print(as.data.frame(
  strat_summary %>%
    pivot_wider(names_from = group, values_from = c(n, median_kd, median_ko))
))

# Within-stratum sign test: is the promoted set at or below background in every quintile?
strat_wide <- strat_summary %>%
  dplyr::select(gc3_quintile, group, median_kd, median_ko) %>%
  pivot_longer(c(median_kd, median_ko), names_to = "contrast", values_to = "median") %>%
  pivot_wider(names_from = group, values_from = median)

strat_wide <- strat_wide %>%
  mutate(prom_minus_other = `eIF3d hypoxia promotes` - `All other genes`,
         prom_minus_neg   = `eIF3d hypoxia promotes` - `Negative control`)
cat("\nPromoted minus reference, per stratum:\n")
print(as.data.frame(strat_wide %>% mutate(across(where(is.numeric), ~ round(.x, 3)))))

stopifnot(
  "Every GC3 quintile must contain all three groups" =
    all(!is.na(strat_wide$`eIF3d hypoxia promotes`)) &&
    all(!is.na(strat_wide$`Negative control`)),
  "Promoted set must be AU3-richer than the negative control (composition claim)" =
    median(dhx$gc3[dhx$group == "eIF3d hypoxia promotes"]) <
    median(dhx$gc3[dhx$group == "Negative control"])
)

cat("\nStrata where the promoted set sits at or below background:",
    sum(strat_wide$prom_minus_other <= 0), "of", nrow(strat_wide), "\n")
```

```{r stratified_plot, fig.width=9, fig.height=4.5}
strat_plot_df <- strat_summary %>%
  pivot_longer(c(median_kd, median_ko), names_to = "contrast", values_to = "median_logFC") %>%
  mutate(contrast = recode(contrast,
                           median_ko = "DHX29 KO (clonal)",
                           median_kd = "DHX29 KD (CRISPRi, acute)"),
         contrast = factor(contrast, levels = c("DHX29 KO (clonal)",
                                                "DHX29 KD (CRISPRi, acute)")))

p_strat <- ggplot(strat_plot_df,
                  aes(x = gc3_quintile, y = median_logFC, color = group, group = group)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey40", linewidth = 0.5) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ contrast, nrow = 1) +
  scale_color_manual(values = set_colors) +
  labs(
    title    = "eIF3d-promoted mRNAs are AU3-rich but do not respond to DHX29 loss at matched GC3",
    subtitle = paste0("Median logFC within GC3 quintiles; n = ", nrow(dhx), " genes total\n",
                      "GC3 median: promoted ",
                      round(median(dhx$gc3[dhx$group == "eIF3d hypoxia promotes"]), 1),
                      "%, neg ctrl ",
                      round(median(dhx$gc3[dhx$group == "Negative control"]), 1),
                      "%, other ",
                      round(median(dhx$gc3[dhx$group == "All other genes"]), 1), "%"),
    x = "GC3 quintile (low to high)", y = "Median logFC on DHX29 depletion", color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_strat
ggsave(here("plots", "87_dhx29_logfc_by_gc3_quintile_and_group.pdf"), p_strat,
       width = 9, height = 4.5)
cat("Saved: plots/87_dhx29_logfc_by_gc3_quintile_and_group.pdf\n")
```

---
---

## Section 6: Abundance-defined gene sets (DHX29 KD RNA-seq) for notebook 07

Two gene sets are exported to `output/genesets/` in the standard schema
(`db_gene_symbol`, `ensembl_gene`, `gs_name`, `gs_description`, `gs_source_species`), so
notebook 07 can take one as the positive class and the other as the negative class.

**These sets are defined on steady-state abundance, not stability.** They are named
`dhx29_{contrast}_up` / `dhx29_{contrast}_down` for exactly that reason. An earlier draft
called them "sensitive" / "resistant to DHX29-mediated degradation"; Section 7a shows that
claim is not supported by the paper's own half-life data, so the naming was corrected.
Stability-defined sets are built separately in Section 7d.

**Parameters** (all YAML, edit at the top of this notebook):

| Param | Meaning |
|---|---|
| `n_up` / `n_down` | Target set sizes. 07 keeps all positives and downsamples the negative pool to `min(n_pos, n_neg)`, so make `n_down >= n_up` or 07 will silently shrink the positive class. |
| `dhx29_contrast` | `"kd"` (CRISPRi, the stronger and more acute effect - default), `"ko"` (clonal), or `"both"` (require the same sign in both contrasts, rank by KD). |
| `down_mode` | `"extreme"` (default: bottom-N by logFC, `padj < padj_cut` - a symmetric top-vs-bottom contrast) or `"unchanged"` (project negative-control convention: `\|logFC\| < unchanged_max_abs_lfc` and `padj > unchanged_padj`). |
| `padj_cut` | Significance cut applied to both tails in `"extreme"` mode. |

**Definitions.** Positive logFC = the mRNA accumulates when DHX29 is depleted, so:

- **`up`** = top `n_up` genes by logFC, `padj < padj_cut`. More mRNA present without DHX29.
- **`down`** = the most negative-logFC genes (`"extreme"`, default), or genes whose level
  does not depend on DHX29 (`"unchanged"`).

**Why `"extreme"` is the default.** A classifier learns from the contrast it is given, and
extreme-vs-extreme is the sharper, symmetric one: both classes are drawn under the same
significance filter, and the label is monotone in the quantity of interest. Extreme-vs-middle
mixes two different things into the negative class - genes DHX29 does not act on and genes
with too little power to tell - so the model partly learns "was this gene measurable" rather
than "did it respond". The same top-vs-bottom design is what the eIF4E resistant/sensitive
sets in this project use.

```{r geneset_params}
n_up           <- as.integer(params$n_up)
n_down         <- as.integer(params$n_down)
dhx29_contrast <- tolower(params$dhx29_contrast)
down_mode      <- tolower(params$down_mode)

stopifnot(
  "n_up must be a positive integer"   = !is.na(n_up)   && n_up   > 0,
  "n_down must be a positive integer" = !is.na(n_down) && n_down > 0,
  "dhx29_contrast must be one of kd / ko / both" =
    dhx29_contrast %in% c("kd", "ko", "both"),
  "down_mode must be one of unchanged / extreme" =
    down_mode %in% c("unchanged", "extreme")
)

rank_lfc  <- if (dhx29_contrast == "ko") "logFC_ko" else "logFC_kd"
rank_padj <- if (dhx29_contrast == "ko") "padj_ko"  else "padj_kd"

cat("Ranking contrast:", dhx29_contrast, "->", rank_lfc, "\n")
cat("Target sizes: up =", n_up, ", down =", n_down, "\n")
cat("Down mode:", down_mode, "\n")

if (n_down < n_up) {
  warning("n_down < n_up: notebook 07 downsamples the negative pool to min(n_pos, n_neg), ",
          "so the positive class will be silently truncated to ", n_down, ".")
}
```

```{r build_genesets}
pool <- dhx %>%
  dplyr::filter(!is.na(symbol), symbol != "") %>%
  mutate(rank_lfc_val = .data[[rank_lfc]], rank_padj_val = .data[[rank_padj]])

cat("Selection pool:", nrow(pool), "genes\n")

up_pool <- pool %>% dplyr::filter(rank_padj_val < params$padj_cut, rank_lfc_val > 0)
if (dhx29_contrast == "both") {
  up_pool <- up_pool %>% dplyr::filter(logFC_ko > 0, logFC_kd > 0)
}
cat("Up pool (padj <", params$padj_cut,
    if (dhx29_contrast == "both") ", up in both contrasts" else "", "):", nrow(up_pool), "\n")

up_set <- up_pool %>% slice_max(rank_lfc_val, n = n_up, with_ties = FALSE)

if (down_mode == "unchanged") {
  down_pool <- pool %>%
    dplyr::filter(abs(rank_lfc_val) < params$unchanged_max_abs_lfc,
                  rank_padj_val > params$unchanged_padj)
  if (dhx29_contrast == "both") {
    down_pool <- down_pool %>%
      dplyr::filter(abs(logFC_ko) < params$unchanged_max_abs_lfc,
                    abs(logFC_kd) < params$unchanged_max_abs_lfc)
  }
  cat("Down pool (|logFC| <", params$unchanged_max_abs_lfc,
      ", padj >", params$unchanged_padj, "):", nrow(down_pool), "\n")
  down_set <- down_pool %>% slice_min(abs(rank_lfc_val), n = n_down, with_ties = FALSE)
} else {
  down_pool <- pool %>% dplyr::filter(rank_padj_val < params$padj_cut, rank_lfc_val < 0)
  if (dhx29_contrast == "both") {
    down_pool <- down_pool %>% dplyr::filter(logFC_ko < 0, logFC_kd < 0)
  }
  cat("Down pool (down, padj <", params$padj_cut, "):", nrow(down_pool), "\n")
  down_set <- down_pool %>% slice_min(rank_lfc_val, n = n_down, with_ties = FALSE)
}

cat("\nRealised sizes: up =", nrow(up_set), ", down =", nrow(down_set), "\n")
cat("Ceiling under the current criteria: n_up <=", nrow(up_pool),
    ", n_down <=", nrow(down_pool), "\n")

stopifnot(
  "Up pool is smaller than n_up - loosen padj_cut or lower n_up" =
    nrow(up_set) == n_up,
  "Down pool is smaller than n_down - loosen the criteria or lower n_down" =
    nrow(down_set) == n_down,
  "The two sets must be disjoint (notebook 07 requires this)" =
    length(intersect(up_set$ensembl_gene, down_set$ensembl_gene)) == 0,
  "Up genes must all be up on DHX29 depletion" = all(up_set$rank_lfc_val > 0),
  "Gene IDs must be unique within each set" =
    !any(duplicated(up_set$ensembl_gene)) && !any(duplicated(down_set$ensembl_gene))
)
```

### 6a: Set characteristics

```{r geneset_characteristics}
set_profile <- bind_rows(
  up_set   %>% mutate(set = "Up on DHX29 KD"),
  down_set %>% mutate(set = "Down on DHX29 KD")
) %>%
  group_by(set) %>%
  summarise(
    n              = n(),
    median_lfc_kd  = round(median(logFC_kd), 3),
    median_lfc_ko  = round(median(logFC_ko), 3),
    median_gc3     = round(median(gc3), 1),
    median_cds_gc  = round(median(cds_gc, na.rm = TRUE), 1),
    median_cds_len = round(median(cds_length, na.rm = TRUE)),
    .groups = "drop"
  )
print(as.data.frame(set_profile))

sep_test <- wilcox.test(up_set$rank_lfc_val, down_set$rank_lfc_val)
cat("\nUp vs down on the ranking contrast: Wilcoxon p",
    fmt_p(sep_test$p.value), "| n =", nrow(up_set), "vs", nrow(down_set), "\n")

gc3_test <- wilcox.test(up_set$gc3, down_set$gc3)
cat("GC3 difference between the sets: Wilcoxon p", fmt_p(gc3_test$p.value),
    "| medians", round(median(up_set$gc3), 1), "vs", round(median(down_set$gc3), 1), "\n")

cat("\nOverlap with eIF3d-promoted (hypoxia 1hr):",
    length(intersect(up_set$ensembl_gene, promotes_ids)), "of", nrow(up_set), "up |",
    length(intersect(down_set$ensembl_gene, promotes_ids)), "of", nrow(down_set), "down\n")

stopifnot(
  "The two sets must separate on the ranking contrast" = sep_test$p.value < 1e-10,
  "The up set must be more AU3-rich than the down set" =
    median(up_set$gc3) < median(down_set$gc3)
)
```

```{r confound_report}
confound <- set_profile %>% dplyr::select(set, median_gc3, median_cds_gc, median_cds_len)
gc3_up    <- confound$median_gc3[confound$set == "Up on DHX29 KD"]
gc3_down  <- confound$median_gc3[confound$set == "Down on DHX29 KD"]
len_up    <- confound$median_cds_len[confound$set == "Up on DHX29 KD"]
len_down  <- confound$median_cds_len[confound$set == "Down on DHX29 KD"]
len_ratio <- round(max(len_up, len_down) / min(len_up, len_down), 2)

cat("Confound check on the exported sets\n")
cat("  GC3 median:        up", gc3_up, "% vs down", gc3_down, "%\n")
cat("  CDS length median: up", len_up, "nt vs down", len_down, "nt  (", len_ratio, "x )\n")
```

> **Read this before modelling.** The two sets differ sharply on two confounded axes by
> construction:
>
> - **GC.** GC3 median `r gc3_up`% (up) vs `r gc3_down`% (down), and Section 3 shows GC
>   explains most of the DHX29-depletion logFC in these tables.
> - **CDS length.** Median CDS `r len_up` nt (up) vs `r len_down` nt (down), a
>   `r len_ratio`-fold gap that arrives free with the selection, since long CDSs are both
>   AU3-richer and offer more stall sites.
>
> A notebook 07 classifier on these sets will therefore be substantially a GC/length
> classifier, and the top importances will be the GC, codon and length families. That is a
> valid answer to "what distinguishes mRNAs that accumulate on DHX29 knockdown", but it is
> **not** independent evidence for a sequence-specific mechanism beyond base composition and
> size. To ask what predicts the response *beyond* those, run 07c and read the SHAP
> decomposition with the GC and length families summed, or rebuild the down set matched on
> GC3 and `log2_cds_length`.

```{r geneset_gc3_plot, fig.width=6, fig.height=4.5}
geneset_colors <- c("Up on DHX29 KD" = "#D55E00", "Down on DHX29 KD" = "#0072B2")

set_long <- bind_rows(
  up_set   %>% mutate(set = "Up on DHX29 KD"),
  down_set %>% mutate(set = "Down on DHX29 KD")
) %>%
  mutate(set = factor(set, levels = c("Up on DHX29 KD", "Down on DHX29 KD")))

p_set_gc3 <- ggplot(set_long, aes(x = gc3, color = set)) +
  geom_density(fill = NA, linewidth = 0.8, key_glyph = "path") +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4, size = 3,
           label = paste0("Wilcoxon p ", fmt_p(gc3_test$p.value),
                          "\nn up = ", nrow(up_set),
                          "\nn down = ", nrow(down_set))) +
  scale_color_manual(values = geneset_colors) +
  labs(
    title    = "GC3 composition of the exported abundance gene sets",
    subtitle = paste0("Ranked on the ", toupper(dhx29_contrast),
                      " RNA-seq contrast; down mode = ", down_mode),
    x = "GC3 content (%)", y = "Density", color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_set_gc3
ggsave(here("plots", "87_dhx29_kd_geneset_gc3_composition.pdf"), p_set_gc3,
       width = 6, height = 4.5)
cat("Saved: plots/87_dhx29_kd_geneset_gc3_composition.pdf\n")
```

### 6b: Write the gene set CSVs

Filenames follow the pattern notebook 07 builds from its params
(`{condition}_{gene}_{direction}_{readout}_{timepoint}_lfc{lfc}.csv`). No `readout` column is
stamped, so 07 treats both files as `readout = "TE"` and its pairing guard passes.

```{r write_genesets}
geneset_timepoint <- paste0("k562", down_mode)

geneset_stub <- function(direction) {
  paste0("dhx29", dhx29_contrast, "_29_", direction, "_TE_", geneset_timepoint, "_lfc0.csv")
}

as_geneset <- function(df, direction, description) {
  df %>%
    transmute(
      db_gene_symbol    = symbol,
      ensembl_gene      = ensembl_gene,
      gs_name           = paste0("dhx29_", dhx29_contrast, "_", direction),
      gs_description    = description,
      gs_source_species = "homo sapiens",
      logFC_kd, padj_kd, logFC_ko, padj_ko, gc3
    ) %>%
    arrange(desc(abs(logFC_kd)))
}

src_table <- if (dhx29_contrast == "ko") "S3" else "S2"
assay_txt <- paste0("steady-state abundance on DHX29 ", toupper(dhx29_contrast),
                    if (dhx29_contrast == "kd") " CRISPRi knockdown" else " knockout",
                    " in K562 (RNA-seq, not a stability assay)")

up_out <- as_geneset(
  up_set, "up",
  paste0("top ", n_up, " mRNAs with increased ", assay_txt,
         "; padj < ", params$padj_cut,
         ". Hia et al. 2026 Science adw0288 table ", src_table)
)

down_out <- as_geneset(
  down_set, "down",
  paste0(n_down, " mRNAs with ",
         if (down_mode == "unchanged")
           paste0("unchanged ", assay_txt, " (|logFC| < ", params$unchanged_max_abs_lfc,
                  ", padj > ", params$unchanged_padj, ")")
         else
           paste0("decreased ", assay_txt, "; padj < ", params$padj_cut),
         ". Hia et al. 2026 Science adw0288 table ", src_table)
)

up_file   <- geneset_stub("up")
down_file <- geneset_stub("down")

stopifnot(
  "Gene set schema is missing required columns" =
    all(c("db_gene_symbol", "ensembl_gene", "gs_name", "gs_description",
          "gs_source_species") %in% colnames(up_out)),
  "No gene symbol may be blank" =
    all(nzchar(up_out$db_gene_symbol)) && all(nzchar(down_out$db_gene_symbol)),
  "readout column must be absent so 07 treats these as TE-compatible" =
    !"readout" %in% colnames(up_out),
  # Guard against the naming regression this section exists to fix. Matches affirmative
  # decay claims only - the descriptions legitimately contain "not a stability assay".
  "gene set descriptions must not claim a degradation/decay readout" =
    !any(grepl("degrad|decay|clearance|half-life", c(up_out$gs_description,
                                                     down_out$gs_description),
               ignore.case = TRUE)),
  "gene set descriptions must state the abundance readout explicitly" =
    all(grepl("steady-state abundance", c(up_out$gs_description,
                                          down_out$gs_description)))
)

if (isTRUE(params$write_genesets)) {
  write_csv(up_out,   here("output", "genesets", up_file))
  write_csv(down_out, here("output", "genesets", down_file))
  cat("Wrote output/genesets/", up_file,   " (", nrow(up_out),   " genes)\n", sep = "")
  cat("Wrote output/genesets/", down_file, " (", nrow(down_out), " genes)\n", sep = "")
} else {
  cat("write_genesets = FALSE - nothing written\n")
}

cat("\nHead of the up set:\n")
print(as.data.frame(head(up_out %>%
  dplyr::select(db_gene_symbol, ensembl_gene, logFC_kd, padj_kd, gc3), 10)))
```

---

## Section 7: The KO SLAM-seq half-life data - the paper's actual stability readout

Table S6 is a SLAM-seq metabolic-labelling time course (`Ctrl` and `KO` at 0/3/6 h, three
replicates each) with fitted `Ctrl_Half_life` and `KO_Half_life`. It is the only direct decay
measurement in the paper, and it exists **only for the KO** - there are no KD columns in S6.
Notebook 23 already parsed and gene-mapped it; this section reads that saved object rather
than re-deriving the transcript mapping.

`logFC = log2(KO / Ctrl)` half-life, so **positive = stabilised when DHX29 is knocked out**,
i.e. DHX29 normally promotes that mRNA's decay.

```{r load_slamseq}
slam_path <- here("output", "predictive_modeling", "dhx29_slamseq_halflives.rds")
stopifnot(
  "dhx29_slamseq_halflives.rds not found - knit notebook 23 first" = file.exists(slam_path)
)

slam <- readRDS(slam_path) %>%
  dplyr::select(ensembl_gene, symbol, Ctrl_Half_life, KO_Half_life, logFC, n_transcripts) %>%
  dplyr::filter(!is.na(logFC)) %>%
  distinct(ensembl_gene, .keep_all = TRUE)

cat("SLAM-seq genes:", nrow(slam), "\n")
cat("Median Ctrl half-life:", round(median(slam$Ctrl_Half_life), 2),
    "h | median KO half-life:", round(median(slam$KO_Half_life), 2), "h\n")

# Identity check on the logFC column. nb23 aggregates genes with several mapped
# transcripts by averaging half-lives and logFC separately, and the mean of ratios is
# not the ratio of means - so the identity is exact only for single-transcript genes.
# Asserting it separately on the two strata validates the column without failing on
# legitimate aggregation.
dev <- abs(slam$logFC - log2(slam$KO_Half_life / slam$Ctrl_Half_life))
single <- slam$n_transcripts == 1

cat("Single-transcript genes:", sum(single), "| aggregated:", sum(!single), "\n")
cat("Max |logFC - log2(KO/Ctrl)|: single =", format(max(dev[single]), digits = 3),
    "| aggregated =", round(max(dev[!single]), 4), "\n")

stopifnot(
  # Zhu-style replication check: Hia report a median half-life of a few hours.
  "Median control half-life outside the expected range" =
    dplyr::between(median(slam$Ctrl_Half_life), 2.0, 6.0),
  "Half-lives must be positive"  = all(slam$Ctrl_Half_life > 0 & slam$KO_Half_life > 0),
  "logFC must be finite"         = all(is.finite(slam$logFC)),
  "logFC must equal log2(KO/Ctrl) exactly for single-transcript genes" =
    max(dev[single]) < 1e-6,
  "Aggregated genes must still be close to log2(KO/Ctrl)" =
    max(dev[!single]) < 0.1,
  "Aggregation must affect only a small minority of genes" =
    mean(single) > 0.99
)
cat("logFC = log2(KO / Ctrl) half-life: verified\n")
```

### 7a: Abundance and stability do not agree

The check that renamed Section 6's gene sets. If "up on DHX29 depletion" meant "stabilised",
the RNA-seq logFC and the SLAM-seq half-life logFC would track each other.

```{r abundance_vs_stability}
av <- slam %>%
  dplyr::select(ensembl_gene, slam_logfc = logFC) %>%
  inner_join(dhx %>% dplyr::select(ensembl_gene, logFC_kd, logFC_ko, gc3),
             by = "ensembl_gene")

cat("Genes with both readouts:", nrow(av), "\n\n")

agree <- bind_rows(
  cor_row(av$logFC_kd, av$slam_logfc, "KD RNA-seq logFC", "SLAM-seq half-life logFC"),
  cor_row(av$logFC_ko, av$slam_logfc, "KO RNA-seq logFC", "SLAM-seq half-life logFC"),
  cor_row(av$gc3,      av$slam_logfc, "GC3",              "SLAM-seq half-life logFC"),
  cor_row(av$gc3,      av$logFC_kd,   "GC3",              "KD RNA-seq logFC")
) %>%
  mutate(rho = round(rho, 3), p = format.pval(p, digits = 3))
print(as.data.frame(agree))

rho_abund_stab <- as.numeric(agree$rho[agree$feature == "KD RNA-seq logFC"])
rho_gc3_slam   <- as.numeric(agree$rho[agree$feature == "GC3" &
                                       agree$contrast == "SLAM-seq half-life logFC"])

stopifnot(
  "Abundance and stability are expected to be near-orthogonal here" =
    abs(rho_abund_stab) < 0.15,
  "GC3 is expected NOT to predict the SLAM-seq half-life change" =
    abs(rho_gc3_slam) < 0.15
)
```

> **Documented negative result.** The KD abundance change and the KO half-life change
> correlate at rho = `r rho_abund_stab` across `r nrow(av)` genes. GC3 predicts the abundance
> change strongly (rho ~ -0.63) but the half-life change not at all
> (rho = `r rho_gc3_slam`). So the striking GC3 relationship in Section 3 lives entirely in
> the abundance readout and does not appear in the paper's own decay measurement.
>
> Consequences, both applied above: Section 6's sets are named for abundance, and any claim
> that they mark "DHX29-mediated degradation" is unsupported by this data. Notebook 23
> reached the same conclusion from the GC3 side (rho = -0.024, p = 0.118).
>
> This is not a claim the paper is wrong. The KO is the weaker perturbation in the abundance
> data too (rho -0.117 vs -0.633), the SLAM-seq universe is smaller and noisier, and our
> transcript selection and gene-level dedup differ from theirs. It does mean this pipeline
> cannot treat the two readouts as interchangeable.

### 7b: Gene-group density on the SLAM-seq half-life change

Same three groups as Section 4, now on the stability readout.

```{r slam_groups}
slam_gc <- slam %>%
  inner_join(gc_fm %>% dplyr::select(ensembl_gene, gc3, cds_gc, cds_length),
             by = "ensembl_gene") %>%
  mutate(group = case_when(
    ensembl_gene %in% promotes_ids ~ "eIF3d hypoxia promotes",
    ensembl_gene %in% neg_ctrl_ids ~ "Negative control",
    TRUE                           ~ "All other genes"
  ))

cat("SLAM-seq genes with GC3:", nrow(slam_gc), "\n")
print(as.data.frame(
  slam_gc %>% group_by(group) %>%
    summarise(n = n(),
              median_logfc = round(median(logFC), 4),
              median_gc3   = round(median(gc3), 1), .groups = "drop")
))

stopifnot(
  "All three groups must be represented" =
    setequal(unique(slam_gc$group), names(set_colors)),
  "Each group needs enough genes for a density estimate" =
    all(table(slam_gc$group) >= 50)
)
```

```{r density_slam, fig.width=6.5, fig.height=5}
res_slam <- density_by_group(
  slam_gc, "logFC",
  title    = "DHX29 KO mRNA stability: eIF3d-promoted genes (hypoxia 1hr)",
  subtitle = paste0("SLAM-seq log2(KO / Ctrl) half-life; positive = stabilised on KO\n",
                    "Dashed vertical lines = group medians"),
  xlab     = "log2(KO / Ctrl) half-life",
  outfile  = "87_dhx29_slamseq_halflife_by_gene_group.pdf"
)
res_slam$plot
```

```{r slam_group_tests}
x_prom  <- slam_gc$logFC[slam_gc$group == "eIF3d hypoxia promotes"]
x_neg   <- slam_gc$logFC[slam_gc$group == "Negative control"]
x_other <- slam_gc$logFC[slam_gc$group == "All other genes"]

slam_tests <- tibble(
  readout         = "SLAM-seq log2(KO/Ctrl) half-life",
  median_promotes = median(x_prom),
  median_neg_ctrl = median(x_neg),
  median_other    = median(x_other),
  p_prom_vs_neg   = wilcox.test(x_prom, x_neg)$p.value,
  p_prom_vs_other = wilcox.test(x_prom, x_other)$p.value,
  n_prom = length(x_prom), n_neg = length(x_neg), n_other = length(x_other)
)
print(as.data.frame(slam_tests %>%
  mutate(across(starts_with("median"), ~ round(.x, 4)),
         across(starts_with("p_"), ~ format.pval(.x, digits = 3)))))
```

> **Null result, and a cleaner one than Section 4's.** On the stability readout the three
> groups are indistinguishable (both p > 0.6). eIF3d-promoted mRNAs are neither preferentially
> stabilised nor destabilised when DHX29 is knocked out, despite being markedly AU3-rich.
> Section 4 found the same thing on the abundance readout; here there is not even a
> composition effect to decompose, because GC3 does not move this readout at all (7a).

### 7c: SLAM-seq half-life change vs GC3

```{r slam_gc3_plot, fig.width=6, fig.height=4.5}
slam_q <- slam_gc %>%
  mutate(gc3_q = ntile(gc3, 4),
         gc3_q = factor(gc3_q, levels = 1:4,
                        labels = c("Q1 (most AU3-rich)", "Q2", "Q3", "Q4 (most GC3-rich)")))

slam_q_stats <- slam_q %>%
  group_by(gc3_q) %>%
  summarise(n = n(), median_gc3 = round(median(gc3), 1),
            median_logfc = round(median(logFC), 4), .groups = "drop")
print(as.data.frame(slam_q_stats))

slam_q_test <- wilcox.test(
  slam_q$logFC[slam_q$gc3_q == "Q1 (most AU3-rich)"],
  slam_q$logFC[slam_q$gc3_q == "Q4 (most GC3-rich)"]
)

p_slam_gc3 <- ggplot(slam_q, aes(x = logFC, color = gc3_q)) +
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.3, size = 3,
           label = paste0("Q1 vs Q4 Wilcoxon p ", fmt_p(slam_q_test$p.value),
                          "\nSpearman(GC3, logFC) = ", rho_gc3_slam,
                          "\nn per quartile = ", round(nrow(slam_q) / 4))) +
  scale_color_manual(values = quartile_colors) +
  coord_cartesian(xlim = quantile(slam_q$logFC, c(0.005, 0.995))) +
  labs(
    title    = "DHX29 KO half-life change by GC3 quartile",
    subtitle = "Contrast with Section 3: the GC3 effect present in RNA-seq is absent here",
    x = "log2(KO / Ctrl) half-life", y = "Cumulative fraction", color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_slam_gc3
ggsave(here("plots", "87_dhx29_slamseq_halflife_cdf_by_gc3_quartile.pdf"), p_slam_gc3,
       width = 6, height = 4.5)
cat("Saved: plots/87_dhx29_slamseq_halflife_cdf_by_gc3_quartile.pdf\n")
```

### 7d: Stability-defined gene sets

Top/bottom N by SLAM-seq half-life logFC. Table S6 carries no per-gene p-value, so these are
ranked on effect size alone - stated explicitly in the descriptions.

```{r build_slam_genesets}
n_slam_up   <- as.integer(params$n_slam_up)
n_slam_down <- as.integer(params$n_slam_down)

stopifnot(
  "n_slam_up must be a positive integer"   = !is.na(n_slam_up)   && n_slam_up   > 0,
  "n_slam_down must be a positive integer" = !is.na(n_slam_down) && n_slam_down > 0,
  "Not enough SLAM-seq genes for the requested set sizes" =
    nrow(slam_gc) >= n_slam_up + n_slam_down
)

slam_pool <- slam_gc %>% dplyr::filter(!is.na(symbol), symbol != "")

slam_stab   <- slam_pool %>% slice_max(logFC, n = n_slam_up,   with_ties = FALSE)
slam_destab <- slam_pool %>% slice_min(logFC, n = n_slam_down, with_ties = FALSE)

slam_profile <- bind_rows(
  slam_stab   %>% mutate(set = "Stabilised on KO"),
  slam_destab %>% mutate(set = "Destabilised on KO")
) %>%
  group_by(set) %>%
  summarise(n = n(),
            median_logfc   = round(median(logFC), 3),
            median_gc3     = round(median(gc3), 1),
            median_cds_len = round(median(cds_length, na.rm = TRUE)),
            .groups = "drop")
print(as.data.frame(slam_profile))

slam_gc3_test <- wilcox.test(slam_stab$gc3, slam_destab$gc3)
cat("\nGC3 difference between the stability sets: Wilcoxon p",
    fmt_p(slam_gc3_test$p.value), "| medians", round(median(slam_stab$gc3), 1),
    "vs", round(median(slam_destab$gc3), 1), "\n")
cat("Compare the abundance sets, which differ by",
    round(abs(gc3_up - gc3_down), 1), "GC3 points.\n")

stopifnot(
  "Stability sets must be disjoint" =
    length(intersect(slam_stab$ensembl_gene, slam_destab$ensembl_gene)) == 0,
  "Stabilised set must have higher logFC than destabilised" =
    median(slam_stab$logFC) > median(slam_destab$logFC),
  "Gene IDs must be unique within each stability set" =
    !any(duplicated(slam_stab$ensembl_gene)) && !any(duplicated(slam_destab$ensembl_gene))
)
```

```{r write_slam_genesets}
as_slam_geneset <- function(df, direction, description) {
  df %>%
    transmute(
      db_gene_symbol    = symbol,
      ensembl_gene      = ensembl_gene,
      gs_name           = paste0("dhx29_slamseq_", direction),
      gs_description    = description,
      gs_source_species = "homo sapiens",
      Ctrl_Half_life, KO_Half_life, slamseq_logfc = logFC, gc3
    ) %>%
    arrange(desc(abs(slamseq_logfc)))
}

slam_stub <- function(direction) {
  paste0("dhx29slam_29_", direction, "_TE_k562slam_lfc0.csv")
}

stab_out <- as_slam_geneset(
  slam_stab, "stabilized",
  paste0("top ", n_slam_up, " mRNAs by log2(KO / Ctrl) half-life on DHX29 knockout in K562 ",
         "(SLAM-seq); stabilised when DHX29 is lost. Ranked on effect size - table S6 ",
         "reports no per-gene p-value. Hia et al. 2026 Science adw0288 table S6")
)

destab_out <- as_slam_geneset(
  slam_destab, "destabilized",
  paste0("bottom ", n_slam_down, " mRNAs by log2(KO / Ctrl) half-life on DHX29 knockout in ",
         "K562 (SLAM-seq); destabilised when DHX29 is lost. Ranked on effect size - table S6 ",
         "reports no per-gene p-value. Hia et al. 2026 Science adw0288 table S6")
)

stab_file   <- slam_stub("stabilized")
destab_file <- slam_stub("destabilized")

stopifnot(
  "Gene set schema is missing required columns" =
    all(c("db_gene_symbol", "ensembl_gene", "gs_name", "gs_description",
          "gs_source_species") %in% colnames(stab_out)),
  "No gene symbol may be blank" =
    all(nzchar(stab_out$db_gene_symbol)) && all(nzchar(destab_out$db_gene_symbol)),
  "readout column must be absent so 07 treats these as TE-compatible" =
    !"readout" %in% colnames(stab_out)
)

if (isTRUE(params$write_genesets)) {
  write_csv(stab_out,   here("output", "genesets", stab_file))
  write_csv(destab_out, here("output", "genesets", destab_file))
  cat("Wrote output/genesets/", stab_file,   " (", nrow(stab_out),   " genes)\n", sep = "")
  cat("Wrote output/genesets/", destab_file, " (", nrow(destab_out), " genes)\n", sep = "")
} else {
  cat("write_genesets = FALSE - nothing written\n")
}
```

### 7e: Coverage against the feature matrices

```{r geneset_coverage}
check_coverage <- function(matrix_file, sets) {
  path <- here("output", "predictive_modeling", matrix_file)
  if (!file.exists(path)) {
    return(tibble(feature_matrix = matrix_file, set = names(sets), n = NA_integer_))
  }
  ids <- readRDS(path)$gene_id_clean
  tibble(feature_matrix = matrix_file,
         set = names(sets),
         n   = vapply(sets, function(s) sum(s %in% ids), integer(1)))
}

all_sets <- list(
  kd_up            = up_out$ensembl_gene,
  kd_down          = down_out$ensembl_gene,
  slam_stabilized  = stab_out$ensembl_gene,
  slam_destabilized = destab_out$ensembl_gene
)

coverage <- map_dfr(c("feature_matrix_external_stability.rds",
                      "feature_matrix_dhx29_no_hia_perturbation.rds"),
                    check_coverage, sets = all_sets) %>%
  pivot_wider(names_from = set, values_from = n)
print(as.data.frame(coverage))

stopifnot(
  "Too few genes survive the feature-matrix join to train a classifier" =
    min(coverage$slam_stabilized, coverage$slam_destabilized, na.rm = TRUE) >= 100
)
```

---

## Section 8: Feature matrices with the DHX29-derived features removed

`feature_matrix_external_stability.rds` carries three columns derived from the same Hia 2026
study these gene sets come from:

| Column | Source | DHX29 perturbation readout? |
|---|---|---|
| `hia2026_dhx29_occupancy` | Thor-Ribo-seq DHX29-IP vs input (Table S8) | Yes - where DHX29 sits |
| `hia2026_dhx29_halflife_logfc` | SLAM-seq log2(KO / Ctrl) half-life (Table S6) | Yes - response to DHX29 loss |
| `hia2026_ctrl_halflife` | SLAM-seq control half-life (Table S6) | No - unperturbed baseline |

**`hia2026_dhx29_halflife_logfc` IS the label for Section 7d's sets** - it is Table S6's
`logFC`, the exact quantity those sets are ranked on. Using it there would be pure leakage,
not a judgement call. For Section 6's abundance sets it is not circular (it correlates with
the KD logFC at only r = 0.016), but it is dropped in the same variant for consistency.

Two files are written. The occupancy-only file is kept because an earlier 07 run used it;
overwriting it would silently invalidate that run's provenance.

```{r write_reduced_matrices}
fm_full <- readRDS(here("output", "predictive_modeling",
                        "feature_matrix_external_stability.rds"))

drop_sets <- list(
  "feature_matrix_dhx29_no_occupancy.rds"        = "hia2026_dhx29_occupancy",
  "feature_matrix_dhx29_no_hia_perturbation.rds" = c("hia2026_dhx29_occupancy",
                                                     "hia2026_dhx29_halflife_logfc"),
  # Additionally drops the baseline half-life features. hia2026_ctrl_halflife is
  # Table S6's Ctrl_Half_life (rho = 1.000 with it), i.e. the DENOMINATOR of the
  # Section 7d label log2(KO / Ctrl) - it correlates with that label at rho = -0.217
  # purely by construction. zhu2024_ctrl_halflife is an independent measurement of
  # baseline stability (rho = 0.604 with Ctrl_Half_life but only 0.018 with the label,
  # so not itself a leak); it is dropped alongside to test the stability sets without
  # any baseline-decay information at all.
  "feature_matrix_dhx29_no_hia_no_ctrl_halflife.rds" = c("hia2026_dhx29_occupancy",
                                                         "hia2026_dhx29_halflife_logfc",
                                                         "hia2026_ctrl_halflife",
                                                         "zhu2024_ctrl_halflife")
)

for (out_name in names(drop_sets)) {
  drop_cols  <- drop_sets[[out_name]]
  stopifnot("Columns to drop are not all in the matrix" =
              all(drop_cols %in% colnames(fm_full)))

  fm_reduced <- fm_full %>% dplyr::select(-all_of(drop_cols))

  stopifnot(
    "Exactly the named columns must be dropped" =
      ncol(fm_reduced) == ncol(fm_full) - length(drop_cols),
    "Dropped columns must be gone" = !any(drop_cols %in% colnames(fm_reduced)),
    "No rows may be lost"          = nrow(fm_reduced) == nrow(fm_full),
    "The join key must survive"    = "gene_id_clean" %in% colnames(fm_reduced),
    "Only the named columns may be removed" =
      setequal(setdiff(colnames(fm_full), colnames(fm_reduced)), drop_cols)
  )

  saveRDS(fm_reduced, here("output", "predictive_modeling", out_name))
  cat(out_name, ": ", nrow(fm_reduced), " x ", ncol(fm_reduced),
      "  (dropped ", paste(drop_cols, collapse = ", "), ")\n", sep = "")
}

cat("\nBaseline feature_matrix_external_stability.rds is unmodified: ",
    nrow(fm_full), " x ", ncol(fm_full), "\n", sep = "")
```

Each variant needs its **own `feature_set` label**, since 07's output suffix is built from
params and a reused label silently overwrites the previous run's saved model.

| Matrix | Drops | `feature_set` label |
|---|---|---|
| `feature_matrix_external_stability.rds` | nothing | `dhx29_kd_extreme` |
| `feature_matrix_dhx29_no_occupancy.rds` | occupancy | `dhx29_kd_extreme_noocc` |
| `feature_matrix_dhx29_no_hia_perturbation.rds` | both DHX29 features | `*_nohia` |
| `feature_matrix_dhx29_no_hia_no_ctrl_halflife.rds` | + both baseline half-lives | `*_nohia_nohl` |

### 8a: Notebook 07 invocations

```{r print_07_params, echo=FALSE, results='asis'}
emit_call <- function(condition, direction, timepoint, feature_set,
                      matrix_path, neg_csv, out_html) {
  cat("```r\n")
  cat("rmarkdown::render(\n")
  cat("  \"code/predictive_modeling/07_rf_classification_eif3d_targets_no_boruta.Rmd\",\n")
  cat("  params = list(\n")
  cat("    condition               = \"", condition,   "\",\n", sep = "")
  cat("    gene                    = \"29\",\n")
  cat("    direction               = \"", direction,   "\",\n", sep = "")
  cat("    timepoint               = \"", timepoint,   "\",\n", sep = "")
  cat("    lfc                     = \"0\",\n")
  cat("    feature_set             = \"", feature_set, "\",\n", sep = "")
  cat("    feature_matrix_path     = \"", matrix_path, "\",\n", sep = "")
  cat("    neg_geneset_csv         = \"", neg_csv,     "\",\n", sep = "")
  cat("    exclude_4e_features     = TRUE,\n")
  cat("    exclude_own_te_features = TRUE\n")
  cat("  ),\n")
  cat("  output_file = \"", out_html, "\"\n", sep = "")
  cat(")\n")
  cat("```\n\n")
}

cat("**Abundance sets (Section 6), no DHX29-derived features:**\n\n")
emit_call(paste0("dhx29", dhx29_contrast), "up", geneset_timepoint,
          paste0("dhx29_", dhx29_contrast, "_", down_mode, "_nohia"),
          "feature_matrix_dhx29_no_hia_perturbation.rds", down_file,
          paste0("07_rf_dhx29_", dhx29_contrast, "_", down_mode, "_nohia.html"))

cat("**Stability sets (Section 7d), no DHX29-derived features:**\n\n")
emit_call("dhx29slam", "stabilized", "k562slam", "dhx29_slam_nohia",
          "feature_matrix_dhx29_no_hia_perturbation.rds", destab_file,
          "07_rf_dhx29_slam_nohia.html")

cat("**Stability sets, also dropping the baseline half-life features",
    "(removes the label's own denominator):**\n\n")
emit_call("dhx29slam", "stabilized", "k562slam", "dhx29_slam_nohia_nohl",
          "feature_matrix_dhx29_no_hia_no_ctrl_halflife.rds", destab_file,
          "07_rf_dhx29_slam_nohia_nohl.html")
```

Notes:

- `condition` / `gene` / `direction` / `timepoint` / `lfc` exist only to reconstruct the
  positive-set filename; they carry no other meaning for these runs.
- `exclude_own_te_features = TRUE` drops the `indiv_te_*` / `delta_te_*` / `sig_*` columns.
  These are eIF3d TE features and the target here is DHX29, so they are not leakage in the
  usual sense - but leaving them in turns the model into a joint eIF3d/DHX29 question.
- Run `07c_` afterwards with the same params for the SHAP decomposition, which is the only
  way to sum the GC and length families and see how much of each classifier is composition.

---

## Summary

1. **Fig 1F reproduces on the RNA-seq.** GC3 negatively tracks logFC in both contrasts, and
   the acute CRISPRi knockdown is far stronger than the clonal knockout
   (Spearman rho = `r rho_gc3_kd` vs `r rho_gc3_ko`, n = `r nrow(dhx)`), exactly as the paper
   attributes to compensatory adaptation in the KO clones. The Q1-vs-Q4 median shift is
   ~0.46 logFC in the KD vs ~0.06 in the KO.

2. **The GC3 effect is not resolvably third-position.** Raw `cds_gc` predicts logFC almost as
   well as GC3, and GC3 orthogonalised against `cds_gc` retains only a fraction of the
   signal. Steady-state RNA-seq cannot separate wobble choice from bulk CDS GC; the
   codon-resolved evidence for AU3 specificity is Table S8 (notebook 23).

3. **Abundance is not stability in this dataset.** The KD RNA-seq logFC and the KO SLAM-seq
   half-life logFC correlate at rho = `r rho_abund_stab` (n = `r nrow(av)`), and GC3 predicts
   the former strongly but the latter not at all (rho = `r rho_gc3_slam`). Section 6's gene
   sets are therefore named and described for abundance, never for degradation.

4. **eIF3d-promoted mRNAs do not respond to DHX29 loss on either readout.** On abundance
   their KD logFC matches the negative control (p = 0.39) and their KO logFC is significantly
   *lower* than both references; stratifying by GC3 puts them at or below background in 10 of
   10 strata. On stability all three groups are indistinguishable (p > 0.6). They are
   markedly AU3-rich, and it buys them nothing here. Whatever eIF3d does for these transcripts
   in hypoxia is not shared with DHX29-dependent codon-driven decay - consistent with
   notebook 23.

5. **Four gene sets exported**, two per readout, plus two reduced feature matrices with the
   DHX29-derived columns removed. The abundance sets are heavily GC- and length-confounded
   by construction and a classifier on them will largely recover composition; the stability
   sets are not, which makes the pair a useful comparison rather than a redundant one.

---

## Section 9: Add the DHX29 KD RNA-seq logFC as a feature

To ask whether DHX29 responsiveness helps predict **eIF3d promotion** (the standard
`hypoxia / 3d / promotes / 1hr / lfc0.5` contrast), the KD logFC has to be a column in the
feature matrix rather than a label.

This is not circular: the label is an eIF3d TE change in MDA-MB-231, the feature is a DHX29
knockdown RNA abundance change in K562 - different perturbation, cell line and readout.

**A new file is written; `feature_matrix_external_stability.rds` is NOT modified.** Adding a
column to the shared base matrix would change the feature set of every existing
`external_stability` run, so those runs would no longer reproduce on a re-knit, and 07c
asserts a saved model against a rebuilt feature set. Point `feature_matrix_path` at this file
instead. If you do want it in the base matrix, change `out_file` below to
`feature_matrix_external_stability.rds` and re-run every dependent notebook.

```{r build_kd_feature_matrix}
out_file <- "feature_matrix_dhx29_kd_feature.rds"

fm_base <- readRDS(here("output", "predictive_modeling",
                        "feature_matrix_external_stability.rds"))

kd_feature <- kd %>%                       # tidied Table S2 from Section 1b
  transmute(gene_id_clean = ensembl_gene,
            hia2026_dhx29_kd_rna_lfc = logFC_kd)

stopifnot(
  "KD feature key must be unique" = !any(duplicated(kd_feature$gene_id_clean)),
  "Feature name must not already exist in the base matrix" =
    !"hia2026_dhx29_kd_rna_lfc" %in% colnames(fm_base)
)

fm_kd <- fm_base %>% left_join(kd_feature, by = "gene_id_clean")

coverage_pct <- 100 * mean(!is.na(fm_kd$hia2026_dhx29_kd_rna_lfc))
cat("Rows:", nrow(fm_kd), "| columns:", ncol(fm_base), "->", ncol(fm_kd), "\n")
cat(sprintf("DHX29 KD logFC coverage: %.1f%% of genes\n", coverage_pct))

stopifnot(
  "Join must not duplicate or drop rows" = nrow(fm_kd) == nrow(fm_base),
  "Exactly one column may be added" = ncol(fm_kd) == ncol(fm_base) + 1,
  "logFC must be finite where present" =
    all(is.finite(fm_kd$hia2026_dhx29_kd_rna_lfc[!is.na(fm_kd$hia2026_dhx29_kd_rna_lfc)])),
  # 07 drops any feature below 50% non-NA; fail here rather than have it silently vanish.
  "Coverage is below 07's 50% sparsity threshold - the feature would be dropped" =
    coverage_pct > 55,
  "Leakage guard: the eIF3d TE label must not be a feature" =
    !"eif3d_target" %in% colnames(fm_kd)
)

saveRDS(fm_kd, here("output", "predictive_modeling", out_file))
cat("Wrote output/predictive_modeling/", out_file, "\n", sep = "")
cat("Base matrix unmodified:", ncol(fm_base), "columns\n")
```

### 9a: What to expect before running it

Two diagnostics that make the 07 result interpretable rather than surprising.

```{r kd_feature_diagnostics}
diag_df <- fm_kd %>%
  distinct(gene_id_clean, .keep_all = TRUE) %>%
  dplyr::select(gene_id_clean, hia2026_dhx29_kd_rna_lfc, cds_gc) %>%
  left_join(gc_fm %>% dplyr::select(ensembl_gene, gc3),
            by = c("gene_id_clean" = "ensembl_gene")) %>%
  dplyr::filter(!is.na(hia2026_dhx29_kd_rna_lfc))

cat("1. The feature is largely a GC proxy:\n")
print(as.data.frame(bind_rows(
  cor_row(diag_df$cds_gc, diag_df$hia2026_dhx29_kd_rna_lfc, "cds_gc", "DHX29 KD logFC"),
  cor_row(diag_df$gc3,    diag_df$hia2026_dhx29_kd_rna_lfc, "gc3",    "DHX29 KD logFC")
) %>% mutate(rho = round(rho, 3), p = format.pval(p, digits = 3))))

# 2. Marginal separation on the exact contrast 07 will train on.
eif3d_pos <- promotes_ids
eif3d_neg <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                      show_col_types = FALSE) %>%
  mutate(ensembl_gene = str_remove(gene_id, "\\..*")) %>%
  dplyr::filter(!ensembl_gene %in% eif3d_pos) %>%
  pull(ensembl_gene)

x_pos <- diag_df$hia2026_dhx29_kd_rna_lfc[diag_df$gene_id_clean %in% eif3d_pos]
x_neg <- diag_df$hia2026_dhx29_kd_rna_lfc[diag_df$gene_id_clean %in% eif3d_neg]
mw <- wilcox.test(x_pos, x_neg)

cat("\n2. Marginal separation for the eIF3d contrast 07 will train on:\n")
cat("   median (3d promotes) =", round(median(x_pos), 4), ", n =", length(x_pos), "\n")
cat("   median (neg control) =", round(median(x_neg), 4), ", n =", length(x_neg), "\n")
cat("   Wilcoxon p", fmt_p(mw$p.value), "\n")

stopifnot(
  "Both eIF3d classes need enough genes carrying the feature" =
    length(x_pos) >= 100 && length(x_neg) >= 100
)
```

> **Prediction, stated before running.** The feature should rank **low**. It carries almost no
> marginal information about this contrast (the two eIF3d classes have near-identical DHX29 KD
> logFC, Section 4 found the same at p = 0.39), and what signal it does carry is GC, which the
> matrix already represents directly via `cds_gc` / `gc3` / `transcript_gc`. If it ranks
> *high*, the first thing to check is whether it is displacing the GC features rather than
> adding anything - compare its rank against the `external_stability` run without it, and use
> 07c to sum the GC family.

```{r print_07_params_kdfeature, echo=FALSE, results='asis'}
cat("```r\n")
cat("rmarkdown::render(\n")
cat("  \"code/predictive_modeling/07_rf_classification_eif3d_targets_no_boruta.Rmd\",\n")
cat("  params = list(\n")
cat("    condition               = \"hypoxia\",\n")
cat("    gene                    = \"3d\",\n")
cat("    direction               = \"promotes\",\n")
cat("    timepoint               = \"1hr\",\n")
cat("    lfc                     = \"0.5\",\n")
cat("    feature_set             = \"dhx29_kd_feature\",\n")
cat("    feature_matrix_path     = \"", out_file, "\",\n", sep = "")
cat("    exclude_own_te_features = TRUE\n")
cat("  ),\n")
cat("  output_file = \"07_rf_3d_promotes_with_dhx29_kd.html\"\n")
cat(")\n")
cat("```\n")
```
