---
title: "Linear models of 4E resistance from GC3"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  teleman_all_csv: "output/teleman_riboseq_te_all_genes.csv"
  feature_matrix: "feature_matrix_teleman_4e.rds"
  pos_csv: "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
  neg_csv: "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"
  rf_test_auc: 0.992
  train_frac: 0.7
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Notebook 07 reaches a test-set AUC of ~0.99 separating the 4E-resistant from the
4E-sensitive gene sets, and notebook 79 showed that `cds_gc` alone reaches ~0.978 on the
same task. That raises a question the random forest cannot answer: **how much of 4E
resistance does GC3 actually explain, and does it explain anything away from the
extremes?**

This notebook fits explicit linear models, in order of how much they can be trusted:

1. `lm` on the **continuous** TE log2FC across all genes. This is the headline: it is not
   conditioned on a gene being extreme.
2. A held-out split of the same model.
3. **Transfer**: fit on the two tails only, predict the middle. This is the test no
   earlier notebook has run.
4. `glm` (logistic) on the tails, for a like-for-like comparison against the notebook 07
   AUC, plus the calibration curve the forest never reports.
5. A `baseMean` control for the GC library-preparation artefact notebook 79 flagged.

## Direction convention

This is the Teleman/Roiuk HeLa 4E-BP-inhibition dataset (notebook 76), **not** an
MDA-MB-231 si3d contrast. The MDA-MB-231 filename wart described in CLAUDE.md does not
apply here. In `teleman_riboseq_te_all_genes.csv`:

- **positive `log2FoldChange` = TE held up under 4E inhibition = resistant**
- negative = TE dropped = sensitive

Asserted below on the two genes the paper bolds in Fig 1A (CDKN2B resistant, GLO1
sensitive).

## Feature choice

`gc3` plus one more term. The second term was chosen by testing every available feature
as `lfc ~ gc3 + X` over all genes:

| candidate | R2 gain over gc3 alone | Spearman with gc3 |
|---|---|---|
| `log2_cds_length` | **+0.067** | -0.11 |
| `cds_gc` | +0.051 | 0.94 |
| `gc3_residual` | +0.051 | 0.37 |
| `hia2026_dhx29_occupancy` | +0.040 | -0.91 |
| `csc` | +0.031 | 0.24 |
| `cai`, `tai`, `fop`, kozak, UTR GC | <= +0.02 | - |

`log2_cds_length` wins on both counts: the largest gain, and near-orthogonality to GC3, so
the two coefficients stay separately interpretable. `cds_gc`, `gc3_residual` and `gc2`
give a similar gain but are the same compositional axis re-split, and
`hia2026_dhx29_occupancy` is 0.91-correlated with GC3 and missing for ~22% of genes.

```{r libraries}
library(tidyverse)
library(here)
library(pROC)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    teleman_all_csv = "output/teleman_riboseq_te_all_genes.csv",
    feature_matrix  = "feature_matrix_teleman_4e.rds",
    pos_csv         = "teleman_4e_resistant_TE_1hr_lfc0.5.csv",
    neg_csv         = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    rf_test_auc     = 0.992,
    train_frac      = 0.7,
    save_plots      = TRUE
  )
}
TELEMAN_ALL <- params$teleman_all_csv
FEATURE_MATRIX <- params$feature_matrix
POS_CSV <- params$pos_csv
NEG_CSV <- params$neg_csv
RF_TEST_AUC <- params$rf_test_auc
TRAIN_FRAC <- params$train_frac
SAVE_PLOTS <- params$save_plots

# eIF3d orange (#E69F00) and eIF3e blue (#56B4E9) are reserved for those two factors
# (CLAUDE.md), so this notebook uses the same two Okabe-Ito hues as notebooks 76 and 79.
class_colors <- c("4E-resistant" = "#D55E00",
                  "middle"       = "grey70",
                  "4E-sensitive" = "#0072B2")

set.seed(9)
```

# 1. Data

```{r load}
te <- read_csv(here(TELEMAN_ALL), show_col_types = FALSE)
fm <- readRDS(here("output", "predictive_modeling", FEATURE_MATRIX))
pos <- read_csv(here("output", "genesets", POS_CSV), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", NEG_CSV), show_col_types = FALSE)

stopifnot("expected columns missing from the Teleman table" =
  all(c("gene_id", "symbol", "log2FoldChange", "padj", "baseMean") %in% names(te)))
stopifnot("expected features missing from the feature matrix" =
  all(c("gene_id_clean", "gc3", "log2_cds_length") %in% names(fm)))
```

## Leakage guard

The feature matrix is built by notebook 78, which excludes every eIF4E-derived column by
construction. Re-assert it here rather than assume it, since the target is 4E resistance.

```{r leakage-guard}
leaky <- grep("4e_resistance|is_4e|eif4e|^te_|^sig_|^delta_te|^indiv_te",
              names(fm), ignore.case = TRUE, value = TRUE)
stopifnot("a 4E-derived or TE column is present in the feature matrix" =
  length(leaky) == 0)
cat("no 4E-derived or TE columns in the feature matrix\n")
```

## Direction check

```{r direction-check}
lfc_of <- function(g) te$log2FoldChange[match(g, te$symbol)]

stopifnot("CDKN2B is not positive: the resistance sign has flipped" = lfc_of("CDKN2B") > 0)
stopifnot("GLO1 is not negative: the resistance sign has flipped"   = lfc_of("GLO1")   < 0)
cat("CDKN2B (paper: resistant) log2FC =", round(lfc_of("CDKN2B"), 3), "\n")
cat("GLO1   (paper: sensitive) log2FC =", round(lfc_of("GLO1"),   3), "\n")
```

## Join and classes

```{r join}
d <- te %>%
  dplyr::select(gene_id, symbol, log2FoldChange, padj, baseMean) %>%
  dplyr::inner_join(fm, by = c("gene_id" = "gene_id_clean"),
                    suffix = c("", "_fm")) %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE                          ~ "middle"),
    class = factor(class, levels = names(class_colors)))

stopifnot("gene_id is not unique after the join" = !any(duplicated(d$gene_id)))
stopifnot("resistant and sensitive sets overlap" =
  length(intersect(pos$ensembl_gene, neg$ensembl_gene)) == 0)
stopifnot("some positive-set genes were lost in the join" =
  all(pos$ensembl_gene %in% d$gene_id))
stopifnot("some negative-set genes were lost in the join" =
  all(neg$ensembl_gene %in% d$gene_id))

print(dplyr::count(d, class) %>% as.data.frame(), row.names = FALSE)
```

## Value checks

```{r value-checks}
stopifnot("gc3 outside [0, 100]"  = all(d$gc3 >= 0 & d$gc3 <= 100, na.rm = TRUE))
stopifnot("non-finite log2FC"     = all(is.finite(d$log2FoldChange[!is.na(d$log2FoldChange)])))
stopifnot("non-finite cds length" = all(is.finite(d$log2_cds_length[!is.na(d$log2_cds_length)])))
stopifnot("baseMean must be positive to log" = all(d$baseMean > 0, na.rm = TRUE))

# GC3 median in [40, 65]% is the project-wide spot-check (CLAUDE.md); this transcript set
# is HeLa-dominant rather than MDA-MB-231, but the transcriptome-wide value should still
# land in range.
cat("GC3 median over all genes:", round(median(d$gc3, na.rm = TRUE), 1), "%\n")
stopifnot("transcriptome GC3 median out of range" =
  between(median(d$gc3, na.rm = TRUE), 40, 65))
```

## The model frame

```{r model-frame}
md <- d %>%
  dplyr::filter(!is.na(log2FoldChange), !is.na(gc3), !is.na(log2_cds_length))

cat("genes dropped for missing gc3 or CDS length:", nrow(d) - nrow(md), "\n")
cat("genes modelled:", nrow(md), "\n")
print(dplyr::count(md, class) %>% as.data.frame(), row.names = FALSE)
```

## What "middle" actually is

The tails are the top and bottom 1,195 genes **among those passing `padj < 0.05`**, not a
plain magnitude cut. So the middle set is not a clean band around zero: it contains genes
whose log2FC exceeds the tail thresholds but which were not significant. Quantified here,
because it matters for the transfer test in section 4.

```{r middle-composition}
mid <- md %>% dplyr::filter(class == "middle")
res_min <- min(md$log2FoldChange[md$class == "4E-resistant"])
sen_max <- max(md$log2FoldChange[md$class == "4E-sensitive"])

cat("resistant tail starts at log2FC =", round(res_min, 3), "\n")
cat("sensitive tail ends at   log2FC =", round(sen_max, 3), "\n")
cat("middle genes above the resistant threshold:", sum(mid$log2FoldChange > res_min), "\n")
cat("middle genes below the sensitive threshold:", sum(mid$log2FoldChange < sen_max), "\n")
cat("...all of them non-significant:",
    all(mid$padj[abs(mid$log2FoldChange) > 0.5] >= 0.05 |
        is.na(mid$padj[abs(mid$log2FoldChange) > 0.5])), "\n\n")
print(round(quantile(mid$log2FoldChange, c(0, .05, .25, .5, .75, .95, 1)), 3))
```

Section 4 therefore reports the transfer test twice: once on the whole middle set, and
once on a strict band (`|log2FC| < 0.5`) that removes the underpowered large-effect genes.

# 2. Continuous models over all genes

The headline models. No gene is excluded for being unremarkable, so nothing here is
conditioned on the outcome.

```{r fit-continuous}
m_gc3  <- lm(log2FoldChange ~ gc3, data = md)
m_both <- lm(log2FoldChange ~ gc3 + log2_cds_length, data = md)

summary(m_gc3)
summary(m_both)
```

```{r continuous-table}
model_row <- function(m, label) {
  s <- summary(m)
  tibble::tibble(
    model     = label,
    n         = length(m$fitted.values),
    r_squared = round(s$r.squared, 3),
    adj_r2    = round(s$adj.r.squared, 3),
    rmse      = round(sqrt(mean(residuals(m)^2)), 3))
}

fit_tbl <- dplyr::bind_rows(
  model_row(m_gc3,  "lfc ~ gc3"),
  model_row(m_both, "lfc ~ gc3 + log2_cds_length"))
print(as.data.frame(fit_tbl), row.names = FALSE)

cat("\nSpearman gc3 vs log2FC:",
    round(cor(md$gc3, md$log2FoldChange, method = "spearman"), 3),
    "  (n =", nrow(md), ")\n")

cat("\nNested model comparison:\n")
print(anova(m_gc3, m_both))
```

```{r coef-table}
coef_tbl <- function(m, label) {
  broom::tidy(m) %>%
    dplyr::mutate(model = label,
                  across(c(estimate, std.error, statistic), ~ signif(.x, 3)),
                  p.value = format.pval(p.value, digits = 3, eps = 2.2e-16)) %>%
    dplyr::select(model, term, estimate, std.error, statistic, p.value)
}
print(as.data.frame(dplyr::bind_rows(coef_tbl(m_gc3,  "gc3 only"),
                                     coef_tbl(m_both, "gc3 + length"))),
      row.names = FALSE)
```

The `gc3` slope is per **percentage point**, so multiply by 10 to read it as "per 10% GC3".

```{r slope-interpretation}
b <- coef(m_both)
cat("per +10% GC3:        ", round(10 * b[["gc3"]], 3), "log2FC\n")
cat("per doubling of CDS: ", round(b[["log2_cds_length"]], 3), "log2FC\n")
cat("predicted log2FC at GC3 = 40% vs 75%, CDS length held at the median:\n")
med_len <- median(md$log2_cds_length)
print(round(predict(m_both, tibble::tibble(gc3 = c(40, 75), log2_cds_length = med_len)), 3))
```

## Scatter with the fitted relationship

```{r scatter-gc3, fig.width = 7, fig.height = 5}
r2_lab <- paste0("lfc ~ gc3\nR2 = ", round(summary(m_gc3)$r.squared, 3),
                 "\nSpearman = ", round(cor(md$gc3, md$log2FoldChange,
                                            method = "spearman"), 3),
                 "\nn = ", nrow(md))

p_scatter <- ggplot(md, aes(x = gc3, y = log2FoldChange)) +
  geom_point(aes(color = class), size = 0.5, alpha = 0.35) +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linewidth = 0.8,
              se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  scale_color_manual(values = class_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "GC3 vs TE change under 4E inhibition",
       subtitle = "Positive log2FC = TE held up = 4E-resistant",
       x = "GC3, third codon position (%)",
       y = "TE log2FC (4E inhibition vs vehicle)") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.2,
           label = r2_lab) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_scatter)
if (SAVE_PLOTS) ggsave(here("plots", "88_gc3_vs_te_lfc_scatter.pdf"), p_scatter,
                       width = 7, height = 5)
```

## The same plot for every GC flavour

`cds_gc` is exactly the mean of `gc1`, `gc2` and `gc3` (verified to 2e-14), so these four
panels are one quantity and its three-way decomposition, not four independent features.
All panels use the same genes so the fits are directly comparable.

```{r scatter-gc-facets, fig.width = 9.5, fig.height = 7.5}
gc_panel_feats <- c(cds_gc = "CDS GC, all positions (%)",
                    gc1    = "GC1, first codon position (%)",
                    gc2    = "GC2, second codon position (%)",
                    gc3    = "GC3, third codon position (%)")
stopifnot("a GC panel feature is missing" = all(names(gc_panel_feats) %in% names(md)))

gc_long <- md %>%
  dplyr::select(class, log2FoldChange, dplyr::all_of(names(gc_panel_feats))) %>%
  tidyr::pivot_longer(dplyr::all_of(names(gc_panel_feats)),
                      names_to = "feature", values_to = "gc") %>%
  dplyr::mutate(feature = factor(feature, levels = names(gc_panel_feats)))

# One label per panel: R2 and Spearman from that feature's own simple regression.
gc_stats <- gc_long %>%
  dplyr::group_by(feature) %>%
  dplyr::summarise(
    r2       = summary(lm(log2FoldChange ~ gc))$r.squared,
    spearman = cor(gc, log2FoldChange, method = "spearman"),
    n        = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(label = paste0("R2 = ", sprintf("%.3f", r2),
                               "\nSpearman = ", sprintf("%.3f", spearman),
                               "\nn = ", n))
print(as.data.frame(gc_stats %>% dplyr::select(feature, r2, spearman, n) %>%
                      dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))),
      row.names = FALSE)

panel_labeller <- as_labeller(c(cds_gc = "CDS GC (all positions)",
                                gc1 = "GC1 (first codon position)",
                                gc2 = "GC2 (second codon position)",
                                gc3 = "GC3 (third codon position)"))

p_facet <- ggplot(gc_long, aes(x = gc, y = log2FoldChange)) +
  geom_point(aes(color = class), size = 0.4, alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linewidth = 0.8, se = TRUE) +
  geom_text(data = gc_stats, aes(x = -Inf, y = Inf, label = label),
            hjust = -0.1, vjust = 1.3, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ feature, scales = "free_x", ncol = 2, labeller = panel_labeller) +
  scale_color_manual(values = class_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "GC content vs TE change under 4E inhibition, by codon position",
       subtitle = "Positive log2FC = TE held up = 4E-resistant",
       x = "GC content (%)", y = "TE log2FC (4E inhibition vs vehicle)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", strip.background = element_blank())

print(p_facet)
if (SAVE_PLOTS) ggsave(here("plots", "88_gc_by_position_vs_te_lfc_facets.pdf"), p_facet,
                       width = 9.5, height = 7.5)
```

## Binned means

The scatter is dense enough to hide the shape. Binning by GC3 decile shows whether the
relationship is linear across the range or driven by one end.

```{r binned-means, fig.width = 7, fig.height = 4.5}
binned <- md %>%
  dplyr::mutate(gc3_bin = ntile(gc3, 10)) %>%
  dplyr::group_by(gc3_bin) %>%
  dplyr::summarise(gc3_mid = median(gc3),
                   mean_lfc = mean(log2FoldChange),
                   se = sd(log2FoldChange) / sqrt(dplyr::n()),
                   n = dplyr::n(), .groups = "drop")
print(as.data.frame(binned %>% dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))),
      row.names = FALSE)

p_bin <- ggplot(binned, aes(x = gc3_mid, y = mean_lfc)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  geom_errorbar(aes(ymin = mean_lfc - 1.96 * se, ymax = mean_lfc + 1.96 * se),
                width = 0.6, color = "grey40") +
  geom_point(size = 2.5, color = "#D55E00") +
  geom_line(color = "#D55E00", linewidth = 0.8) +
  labs(title = "Mean TE change by GC3 decile",
       subtitle = paste0("error bars = 95% CI on the mean, n = ", nrow(md),
                         " genes in 10 bins"),
       x = "GC3, decile median (%)", y = "Mean TE log2FC") +
  theme_classic(base_size = 12)

print(p_bin)
if (SAVE_PLOTS) ggsave(here("plots", "88_gc3_decile_mean_lfc.pdf"), p_bin,
                       width = 7, height = 4.5)
```

# 3. Held-out evaluation

Routine, and included for completeness. Notebook 07 already evaluates on a held-out 20%
with a bootstrap CI and 5-fold CV, so this is not a correction to that; it just confirms a
two-parameter model is nowhere near capacity-limited.

```{r holdout}
set.seed(9)
train_idx <- sample.int(nrow(md), size = floor(TRAIN_FRAC * nrow(md)))
train <- md[train_idx, ]
test  <- md[-train_idx, ]
cat("train:", nrow(train), " test:", nrow(test), "\n")

eval_holdout <- function(form, label) {
  fit  <- lm(form, data = train)
  pred <- predict(fit, newdata = test)
  ss_res <- sum((test$log2FoldChange - pred)^2)
  ss_tot <- sum((test$log2FoldChange - mean(train$log2FoldChange))^2)
  tibble::tibble(
    model        = label,
    n_train      = nrow(train),
    n_test       = nrow(test),
    train_r2     = round(summary(fit)$r.squared, 3),
    test_r2      = round(1 - ss_res / ss_tot, 3),
    test_spearman = round(cor(pred, test$log2FoldChange, method = "spearman"), 3),
    test_rmse    = round(sqrt(mean((test$log2FoldChange - pred)^2)), 3))
}

ho <- dplyr::bind_rows(
  eval_holdout(log2FoldChange ~ gc3, "lfc ~ gc3"),
  eval_holdout(log2FoldChange ~ gc3 + log2_cds_length, "lfc ~ gc3 + log2_cds_length"))
print(as.data.frame(ho), row.names = FALSE)
```

```{r pred-vs-obs, fig.width = 6, fig.height = 5.5}
fit_ho <- lm(log2FoldChange ~ gc3 + log2_cds_length, data = train)
test_pred <- test %>% dplyr::mutate(pred = predict(fit_ho, newdata = test))

lab <- paste0("test R2 = ", ho$test_r2[ho$model == "lfc ~ gc3 + log2_cds_length"],
              "\nSpearman = ", ho$test_spearman[ho$model == "lfc ~ gc3 + log2_cds_length"],
              "\nn = ", nrow(test))

p_po <- ggplot(test_pred, aes(x = pred, y = log2FoldChange)) +
  geom_point(aes(color = class), size = 0.6, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey40", linewidth = 0.5) +
  scale_color_manual(values = class_colors, name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  coord_equal(xlim = range(test_pred$log2FoldChange),
              ylim = range(test_pred$log2FoldChange)) +
  labs(title = "Held-out prediction, gc3 + CDS length",
       subtitle = "dashed line = perfect prediction",
       x = "Predicted TE log2FC", y = "Observed TE log2FC") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.2, label = lab) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_po)
if (SAVE_PLOTS) ggsave(here("plots", "88_holdout_pred_vs_obs.pdf"), p_po,
                       width = 6, height = 5.5)
```

# 4. Transfer: fit on the tails, predict the middle

This is the test the classifier framing cannot run. Notebook 07 trains and tests entirely
within the two tails, so every gene it has ever scored is an extreme gene. Here the model
is fit **only** on the 1,195 + 1,195 tail genes and then asked to order the middle genes it
has never seen, which is where nearly every real mRNA lives.

```{r transfer}
tails  <- md %>% dplyr::filter(class != "middle")
middle <- md %>% dplyr::filter(class == "middle")
strict <- middle %>% dplyr::filter(abs(log2FoldChange) < 0.5)

cat("tails:", nrow(tails), " middle:", nrow(middle),
    " strict middle (|lfc| < 0.5):", nrow(strict), "\n\n")

m_tail <- lm(log2FoldChange ~ gc3 + log2_cds_length, data = tails)
m_full <- m_both

transfer_row <- function(fit, newdata, fit_label, set_label) {
  pred <- predict(fit, newdata = newdata)
  ss_res <- sum((newdata$log2FoldChange - pred)^2)
  ss_tot <- sum((newdata$log2FoldChange - mean(newdata$log2FoldChange))^2)
  tibble::tibble(
    fit_on   = fit_label,
    evaluated_on = set_label,
    n        = nrow(newdata),
    r2       = round(1 - ss_res / ss_tot, 3),
    spearman = round(cor(pred, newdata$log2FoldChange, method = "spearman"), 3),
    rmse     = round(sqrt(mean((newdata$log2FoldChange - pred)^2)), 3))
}

tr <- dplyr::bind_rows(
  transfer_row(m_tail, middle, "tails only", "middle"),
  transfer_row(m_full, middle, "all genes",  "middle"),
  transfer_row(m_tail, strict, "tails only", "strict middle |lfc| < 0.5"),
  transfer_row(m_full, strict, "all genes",  "strict middle |lfc| < 0.5"))
print(as.data.frame(tr), row.names = FALSE)
```

`r2` here is computed against the middle set's own mean, so a **negative** value means the
model predicts the middle worse than its own average does. Spearman is the more forgiving
read: it asks only whether the ordering carries over.

## How far apart are the two fits?

```{r transfer-coefs}
cmp <- dplyr::bind_rows(
  broom::tidy(m_tail) %>% dplyr::mutate(fit_on = "tails only"),
  broom::tidy(m_full) %>% dplyr::mutate(fit_on = "all genes")) %>%
  dplyr::select(fit_on, term, estimate, std.error) %>%
  dplyr::mutate(across(c(estimate, std.error), ~ signif(.x, 3))) %>%
  tidyr::pivot_wider(names_from = fit_on, values_from = c(estimate, std.error))
print(as.data.frame(cmp), row.names = FALSE)
```

A tail-fitted slope steeper than the full-data slope is the expected signature of fitting
on extremes: the tails exaggerate the effect per unit GC3.

```{r transfer-plot, fig.width = 7, fig.height = 5}
middle_pred <- middle %>% dplyr::mutate(pred_tail = predict(m_tail, newdata = middle))

lab_t <- paste0("fit on tails, evaluated on middle\nSpearman = ",
                tr$spearman[tr$fit_on == "tails only" & tr$evaluated_on == "middle"],
                "\nR2 = ", tr$r2[tr$fit_on == "tails only" & tr$evaluated_on == "middle"],
                "\nn = ", nrow(middle))

p_tr <- ggplot(middle_pred, aes(x = pred_tail, y = log2FoldChange)) +
  geom_point(size = 0.5, alpha = 0.3, color = "grey40") +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "#D55E00", linewidth = 0.6) +
  labs(title = "Tail-trained model applied to the middle genes",
       subtitle = "dashed orange = perfect prediction, black = actual relationship",
       x = "Predicted TE log2FC (model fit on tails only)",
       y = "Observed TE log2FC") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.4, size = 3.2, label = lab_t) +
  theme_classic(base_size = 12)

print(p_tr)
if (SAVE_PLOTS) ggsave(here("plots", "88_tail_fit_predicts_middle.pdf"), p_tr,
                       width = 7, height = 5)
```

# 5. Logistic model on the tails, against the notebook 07 AUC

Like-for-like with the random forest: same two gene sets, same held-out design, two
features instead of ~30.

```{r logistic}
tails_bin <- tails %>%
  dplyr::mutate(resistant = as.integer(class == "4E-resistant"))
stopifnot("class encoding is not 0/1" = all(tails_bin$resistant %in% c(0L, 1L)))

set.seed(9)
tr_idx <- sample.int(nrow(tails_bin), size = floor(TRAIN_FRAC * nrow(tails_bin)))
b_train <- tails_bin[tr_idx, ]
b_test  <- tails_bin[-tr_idx, ]
cat("train:", nrow(b_train), " test:", nrow(b_test), "\n")
print(dplyr::count(b_test, class) %>% as.data.frame(), row.names = FALSE)

g_gc3  <- glm(resistant ~ gc3, data = b_train, family = binomial())
g_both <- glm(resistant ~ gc3 + log2_cds_length, data = b_train, family = binomial())
summary(g_both)

auc_of <- function(fit) {
  p <- predict(fit, newdata = b_test, type = "response")
  as.numeric(pROC::auc(pROC::roc(b_test$resistant, p, quiet = TRUE)))
}
auc_tbl <- tibble::tibble(
  model = c("glm ~ gc3", "glm ~ gc3 + log2_cds_length", "notebook 07 random forest"),
  test_auc = c(round(auc_of(g_gc3), 4), round(auc_of(g_both), 4), RF_TEST_AUC),
  n_features = c(1L, 2L, NA_integer_))
print(as.data.frame(auc_tbl), row.names = FALSE)
```

```{r roc-plot, fig.width = 6, fig.height = 5.5}
p_test_gc3 <- predict(g_gc3,  newdata = b_test, type = "response")
p_test     <- predict(g_both, newdata = b_test, type = "response")

roc_gc3 <- pROC::roc(b_test$resistant, p_test_gc3, quiet = TRUE)
roc_obj <- pROC::roc(b_test$resistant, p_test,     quiet = TRUE)

set.seed(9)
ci_gc3 <- pROC::ci.auc(roc_gc3, method = "bootstrap", boot.n = 1000)
set.seed(9)
ci_auc <- pROC::ci.auc(roc_obj, method = "bootstrap", boot.n = 1000)

roc_points <- function(r) tibble::tibble(fpr = 1 - r$specificities,
                                         tpr = r$sensitivities)
roc_df     <- roc_points(roc_obj)
roc_df_gc3 <- roc_points(roc_gc3)

p_roc <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey60", linewidth = 0.5) +
  geom_line(color = "#D55E00", linewidth = 0.8) +
  coord_equal() +
  labs(title = "Logistic model on GC3 + log2 CDS length, resistant vs sensitive",
       subtitle = paste0("features: gc3, log2_cds_length   AUC = ",
                         round(as.numeric(roc_obj$auc), 3),
                         " [", round(ci_auc[1], 3), "-", round(ci_auc[3], 3),
                         "]   vs random forest ", RF_TEST_AUC,
                         "\nn = ", nrow(b_test), " held-out genes"),
       x = "False positive rate", y = "True positive rate") +
  theme_classic(base_size = 12)

print(p_roc)
if (SAVE_PLOTS) ggsave(here("plots", "88_logistic_roc_vs_rf.pdf"), p_roc,
                       width = 6, height = 5.5)
```

## GC3 on its own

The same held-out genes and the same split, with `log2_cds_length` dropped. This is the
single-feature version of the plot above.

```{r roc-plot-gc3, fig.width = 6, fig.height = 5.5}
p_roc_gc3 <- ggplot(roc_df_gc3, aes(x = fpr, y = tpr)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey60", linewidth = 0.5) +
  geom_line(color = "#D55E00", linewidth = 0.8) +
  coord_equal() +
  labs(title = "Logistic model on GC3 alone, resistant vs sensitive",
       subtitle = paste0("feature: gc3   AUC = ", round(as.numeric(roc_gc3$auc), 3),
                         " [", round(ci_gc3[1], 3), "-", round(ci_gc3[3], 3),
                         "]   vs random forest ", RF_TEST_AUC,
                         "\nn = ", nrow(b_test), " held-out genes"),
       x = "False positive rate", y = "True positive rate") +
  theme_classic(base_size = 12)

print(p_roc_gc3)
if (SAVE_PLOTS) ggsave(here("plots", "88_logistic_roc_gc3_only.pdf"), p_roc_gc3,
                       width = 6, height = 5.5)
```

## Both curves together

What the second feature actually buys, on one panel.

```{r roc-plot-overlay, fig.width = 6.5, fig.height = 5.5}
roc_both <- dplyr::bind_rows(
  roc_df_gc3 %>% dplyr::mutate(model = paste0("gc3 alone (AUC ",
                                              round(as.numeric(roc_gc3$auc), 3), ")")),
  roc_df     %>% dplyr::mutate(model = paste0("gc3 + log2 CDS length (AUC ",
                                              round(as.numeric(roc_obj$auc), 3), ")")))

# #D55E00 and #0072B2 already denote the two classes in this notebook, so the second
# curve uses reddish purple from the remaining Okabe-Ito hues.
roc_colors <- setNames(c("#D55E00", "#CC79A7"), unique(roc_both$model))

p_roc_both <- ggplot(roc_both, aes(x = fpr, y = tpr, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey60", linewidth = 0.5) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = roc_colors, name = NULL) +
  coord_equal() +
  labs(title = "One feature vs two, resistant vs sensitive",
       subtitle = paste0("random forest on ~30 features = ", RF_TEST_AUC,
                         "   n = ", nrow(b_test), " held-out genes"),
       x = "False positive rate", y = "True positive rate") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", legend.direction = "vertical")

print(p_roc_both)
if (SAVE_PLOTS) ggsave(here("plots", "88_logistic_roc_gc3_vs_both.pdf"), p_roc_both,
                       width = 6.5, height = 5.5)
```

## Is codon position 3 special? One curve per GC flavour

Single-feature logistic models, one per codon position, no CDS length term. If 4E
resistance were driven by *wobble* choice, `gc3` should beat `gc1` and `gc2` by a clear
margin and should not be matched by aggregate `cds_gc`. Same split, same held-out genes as
above.

```{r roc-gc-positions, fig.width = 6.5, fig.height = 5.5}
gc_feats <- c("gc1", "gc2", "gc3", "cds_gc")
stopifnot("a GC feature is missing" = all(gc_feats %in% names(b_train)))

fit_one_gc <- function(f) {
  tr <- b_train[!is.na(b_train[[f]]), ]
  te_ <- b_test[!is.na(b_test[[f]]), ]
  fit <- glm(reformulate(f, response = "resistant"), data = tr, family = binomial())
  pr  <- predict(fit, newdata = te_, type = "response")
  r   <- pROC::roc(te_$resistant, pr, quiet = TRUE)
  set.seed(9)
  ci  <- pROC::ci.auc(r, method = "bootstrap", boot.n = 1000)
  list(roc = r,
       row = tibble::tibble(feature = f,
                            n_test = nrow(te_),
                            auc = round(as.numeric(r$auc), 4),
                            ci_low = round(ci[1], 4),
                            ci_high = round(ci[3], 4),
                            slope = signif(coef(fit)[[f]], 3)))
}

gc_fits <- purrr::map(gc_feats, fit_one_gc) %>% setNames(gc_feats)
gc_auc  <- purrr::map_dfr(gc_fits, "row") %>% dplyr::arrange(dplyr::desc(auc))
print(as.data.frame(gc_auc), row.names = FALSE)

gc_roc_df <- purrr::imap_dfr(gc_fits, function(x, f) {
  tibble::tibble(fpr = 1 - x$roc$specificities,
                 tpr = x$roc$sensitivities,
                 feature = f)
}) %>%
  dplyr::left_join(gc_auc %>% dplyr::select(feature, auc), by = "feature") %>%
  dplyr::mutate(label = paste0(feature, " (AUC ", sprintf("%.3f", auc), ")"),
                label = forcats::fct_reorder(label, -auc))

# #E69F00 and #56B4E9 are reserved for eIF3d and eIF3e (CLAUDE.md); these are four of the
# remaining Okabe-Ito hues, with gc3 kept vermillion to match the earlier ROC panels.
gc_colors <- setNames(
  c("#D55E00", "#0072B2", "#009E73", "#CC79A7"),
  levels(gc_roc_df$label)[match(c("gc3", "gc1", "gc2", "cds_gc"),
                                sub(" .*", "", levels(gc_roc_df$label)))])

p_roc_gc <- ggplot(gc_roc_df, aes(x = fpr, y = tpr, color = label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey60", linewidth = 0.5) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = gc_colors, name = NULL) +
  coord_equal() +
  labs(title = "GC by codon position, resistant vs sensitive",
       subtitle = paste0("one logistic model per feature, no CDS length term",
                         "\nrandom forest on ~30 features = ", RF_TEST_AUC,
                         "   n = ", nrow(b_test), " held-out genes"),
       x = "False positive rate", y = "True positive rate") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", legend.direction = "vertical",
        legend.key.height = unit(0.8, "lines"))

print(p_roc_gc)
if (SAVE_PLOTS) ggsave(here("plots", "88_logistic_roc_gc_positions.pdf"), p_roc_gc,
                       width = 6.5, height = 5.5)
```

Read this against the correlations between the features themselves, since `gc1`, `gc2` and
`gc3` are not independent measurements.

```{r gc-position-correlations}
gc_cor <- b_train %>% dplyr::select(dplyr::all_of(gc_feats)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")
print(round(gc_cor, 3))
```

## Calibration

AUC only measures ordering. Calibration asks whether a predicted probability of 0.8 means
8 in 10 of those genes really are resistant. Notebook 07 reports discrimination but never
this.

```{r calibration, fig.width = 6, fig.height = 5.5}
calib <- tibble::tibble(p = p_test, y = b_test$resistant) %>%
  dplyr::mutate(bin = ntile(p, 10)) %>%
  dplyr::group_by(bin) %>%
  dplyr::summarise(mean_pred = mean(p),
                   obs_frac  = mean(y),
                   n = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(se = sqrt(obs_frac * (1 - obs_frac) / n))
print(as.data.frame(calib %>% dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))),
      row.names = FALSE)

# Brier score: mean squared error of the predicted probabilities. 0 is perfect,
# 0.25 is what always predicting 0.5 would give on a balanced set.
brier <- mean((p_test - b_test$resistant)^2)
cat("\nBrier score:", round(brier, 4), " (0.25 = uninformative on a balanced set)\n")

p_cal <- ggplot(calib, aes(x = mean_pred, y = obs_frac)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey60", linewidth = 0.5) +
  geom_errorbar(aes(ymin = pmax(0, obs_frac - 1.96 * se),
                    ymax = pmin(1, obs_frac + 1.96 * se)),
                width = 0.02, color = "grey40") +
  geom_point(size = 2.5, color = "#D55E00") +
  geom_line(color = "#D55E00", linewidth = 0.8) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Calibration, logistic model on GC3 + log2 CDS length",
       subtitle = paste0("decile bins, dashed = perfect calibration, Brier = ",
                         round(brier, 3), ", n = ", nrow(b_test)),
       x = "Mean predicted probability of resistance",
       y = "Observed fraction resistant") +
  theme_classic(base_size = 12)

print(p_cal)
if (SAVE_PLOTS) ggsave(here("plots", "88_logistic_calibration.pdf"), p_cal,
                       width = 6, height = 5.5)
```

# 6. Is it a GC library-preparation artefact?

Notebook 79 raised the alternative reading: TE is a ratio of two libraries, and if
footprint and total-RNA libraries amplify GC-rich templates differently, a GC-dependent
term appears in `log2FoldChange` with no biology behind it. This cannot be settled from
one dataset, but expression level is the handle available here, since amplification bias
acts through it.

```{r basemean-control}
md <- md %>% dplyr::mutate(log2_basemean = log2(baseMean))

cat("Spearman gc3 vs log2(baseMean):",
    round(cor(md$gc3, md$log2_basemean, method = "spearman"), 3), "\n")
cat("Spearman log2FC vs log2(baseMean):",
    round(cor(md$log2FoldChange, md$log2_basemean, method = "spearman"), 3), "\n\n")

m_adj <- lm(log2FoldChange ~ gc3 + log2_cds_length + log2_basemean, data = md)

adj_cmp <- dplyr::bind_rows(
  broom::tidy(m_both) %>% dplyr::mutate(model = "gc3 + length"),
  broom::tidy(m_adj)  %>% dplyr::mutate(model = "gc3 + length + log2(baseMean)")) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(estimate = signif(estimate, 3), std.error = signif(std.error, 3)) %>%
  dplyr::select(model, term, estimate, std.error)
print(as.data.frame(adj_cmp), row.names = FALSE)

cat("\nR2:", round(summary(m_both)$r.squared, 3), "->",
    round(summary(m_adj)$r.squared, 3), "\n")
cat("gc3 coefficient retained:",
    round(100 * coef(m_adj)[["gc3"]] / coef(m_both)[["gc3"]], 1), "% of its unadjusted value\n")
```

A stronger version of the same question: does the GC3 slope hold *within* narrow
expression strata? If the effect were an amplification artefact tied to abundance, it
should weaken in the strata where abundance barely varies.

```{r basemean-strata}
strata <- md %>%
  dplyr::mutate(bm_q = ntile(log2_basemean, 4)) %>%
  dplyr::group_by(bm_q) %>%
  dplyr::group_modify(~ {
    f <- lm(log2FoldChange ~ gc3 + log2_cds_length, data = .x)
    tibble::tibble(n = nrow(.x),
                   median_basemean = round(median(.x$baseMean), 1),
                   gc3_slope = signif(coef(f)[["gc3"]], 3),
                   gc3_se    = signif(summary(f)$coefficients["gc3", "Std. Error"], 3),
                   r2        = round(summary(f)$r.squared, 3))
  }) %>%
  dplyr::ungroup()
print(as.data.frame(strata), row.names = FALSE)
```

```{r strata-plot, fig.width = 6.5, fig.height = 4.5}
p_str <- ggplot(strata, aes(x = factor(bm_q), y = gc3_slope)) +
  geom_hline(yintercept = coef(m_both)[["gc3"]], linetype = "dashed",
             color = "#D55E00", linewidth = 0.6) +
  geom_errorbar(aes(ymin = gc3_slope - 1.96 * gc3_se,
                    ymax = gc3_slope + 1.96 * gc3_se),
                width = 0.15, color = "grey40") +
  geom_point(size = 2.5) +
  labs(title = "GC3 slope within expression quartiles",
       subtitle = paste0("dashed orange = slope over all genes, n = ",
                         paste(strata$n, collapse = " / "), " per quartile"),
       x = "baseMean quartile (1 = lowest)", y = "GC3 coefficient (log2FC per % GC3)") +
  theme_classic(base_size = 12)

print(p_str)
if (SAVE_PLOTS) ggsave(here("plots", "88_gc3_slope_by_basemean_quartile.pdf"), p_str,
                       width = 6.5, height = 4.5)
```

Neither check can rule out a GC bias acting equally at all expression levels. The
discriminating experiment is the one notebook 79 named: the same comparison in a 4E
inhibition dataset built with a different library protocol, ideally UMI-deduplicated.

# 7. Summary

```{r summary-table}
summary_tbl <- tibble::tibble(
  question = c(
    "Variance in TE log2FC explained by GC3, all genes",
    "...adding log2 CDS length",
    "Held-out test R2 (two features)",
    "Spearman on middle genes, model fit on tails only",
    "Held-out AUC, resistant vs sensitive, gc3 alone",
    "Held-out AUC, resistant vs sensitive, two features",
    "Notebook 07 random forest AUC, same sets",
    "GC3 coefficient retained after baseMean adjustment (%)"),
  value = c(
    round(summary(m_gc3)$r.squared, 3),
    round(summary(m_both)$r.squared, 3),
    ho$test_r2[ho$model == "lfc ~ gc3 + log2_cds_length"],
    tr$spearman[tr$fit_on == "tails only" & tr$evaluated_on == "middle"],
    auc_tbl$test_auc[auc_tbl$model == "glm ~ gc3"],
    auc_tbl$test_auc[auc_tbl$model == "glm ~ gc3 + log2_cds_length"],
    RF_TEST_AUC,
    round(100 * coef(m_adj)[["gc3"]] / coef(m_both)[["gc3"]], 1)))
print(as.data.frame(summary_tbl), row.names = FALSE)
```

```{r session-info}
sessionInfo()
```
