---
title: "GC content by codon position across eIF3d, eIF3e and 4E gene sets"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  save_plots: TRUE
  bg_pool_csv: "si3e_expressed_background_pool.csv"
  n_draws: 1000
  n_curves: 50
  draw_seed: 9
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Unfilled density curves of **CDS GC, GC1, GC2 and GC3**, faceted vertically, for four
gene-set comparisons:

1. **eIF3d-specific** targets (hypoxia 1hr, excluding genes also promoted by eIF3e)
2. **eIF3e-specific** targets (hypoxia 1hr, excluding genes also promoted by eIF3d)
3. The **eIF3d + eIF3e intersect** (promoted by both)
4. eIF3d targets **specific to hypoxia** (not promoted in normoxia 1+4hr)
5. eIF3d targets **specific to normoxia 1+4hr** (not promoted in hypoxia 1hr)
6. **4E-resistant vs 4E-sensitive** (Teleman HeLa 4E-BP inhibition)

Each panel set shows the positive class, its negative class, and the rest of the
transcriptome, each with a dashed vertical at its own median. The rest-of-transcriptome
group **excludes** both labelled sets, so the three curves are disjoint and "all other
genes" is literally true.

The negative control is **size-matched to the positive set** with `set.seed(9)`, counting
genes present in the feature matrix rather than rows in the source CSV. Genes dropped by
the match are removed from the figure entirely, not folded into the grey curve, which would
otherwise put negative controls into "all other genes". Matching only shrinks a larger
comparator; where the negative pool is already the smaller of the two (the eIF3d + eIF3e
intersect) no matching is possible and the chunk output says so.

## Where the GC columns come from

The model matrices (`feature_matrix_external_stability.rds`, `..._dhx29_kd_feature.rds`)
carry `cds_gc` and `gc3` but **not** `gc1`/`gc2`; those live in
`feature_matrix_positional_gc.rds`, which by design drops `cds_gc` (notebook 01v, so the
forest cannot fall back on the aggregate). Neither file alone can draw these figures, so
this notebook joins them on `transcript_id_clean` and asserts that the `gc3` column they
share is identical.

The Teleman HeLa matrix (`feature_matrix_teleman_4e.rds`, notebook 78) carries all four
columns already and is keyed on HeLa dominant transcripts, so it is used as-is for figure 4
and never mixed with the MDA-MB-231 rows.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(save_plots  = TRUE,
                 bg_pool_csv = "si3e_expressed_background_pool.csv",
                 n_draws     = 1000,  # draws behind the empirical p (cheap: GC values only)
                 n_curves    = 50,    # draws actually drawn on the figure
                 draw_seed   = 9)
}
SAVE_PLOTS <- params$save_plots

GC_FEATURES <- c(cds_gc = "CDS GC, all positions",
                 gc1    = "GC1, first codon position",
                 gc2    = "GC2, second codon position",
                 gc3    = "GC3, third codon position")

REST_LABEL <- "Rest of transcriptome"
REST_COLOR <- "grey70"   # CLAUDE.md: new plots use grey70 for the rest of the transcriptome
NEG_COLOR  <- "black"    # CLAUDE.md: negative control set is black
```

# 1. Data

```{r load-231}
fm_model <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
fm_pos   <- readRDS(here("output", "predictive_modeling", "feature_matrix_positional_gc.rds"))

stopifnot("cds_gc missing from the model matrix" = "cds_gc" %in% names(fm_model),
          "gc1/gc2 missing from the positional matrix" = all(c("gc1", "gc2") %in% names(fm_pos)))

gc231 <- fm_model %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol, cds_gc, gc3_model = gc3) %>%
  dplyr::inner_join(fm_pos %>% dplyr::select(transcript_id_clean, gc1, gc2, gc3),
                    by = "transcript_id_clean")

# The two matrices must describe the same transcripts, or gc1/gc2 would be attached to a
# different isoform than cds_gc.
stopifnot("gc3 disagrees between the two matrices" =
  max(abs(gc231$gc3_model - gc231$gc3), na.rm = TRUE) == 0)
gc231 <- gc231 %>% dplyr::select(-gc3_model) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("MDA-MB-231 genes:", nrow(gc231), "\n")
for (f in names(GC_FEATURES)) {
  stopifnot("GC outside [0, 100]" = all(gc231[[f]] >= 0 & gc231[[f]] <= 100, na.rm = TRUE))
}
cat("GC3 median:", round(median(gc231$gc3, na.rm = TRUE), 1), "% (expected 40-65)\n")
stopifnot("GC3 median out of range" = between(median(gc231$gc3, na.rm = TRUE), 40, 65))
```

```{r load-teleman}
gc4e <- readRDS(here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds")) %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol,
                dplyr::all_of(names(GC_FEATURES))) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)
cat("HeLa (Teleman) genes:", nrow(gc4e), "\n")
for (f in names(GC_FEATURES)) {
  stopifnot("GC outside [0, 100]" = all(gc4e[[f]] >= 0 & gc4e[[f]] <= 100, na.rm = TRUE))
}
```

## Gene sets

```{r genesets}
read_geneset <- function(f) {
  p <- here("output", "genesets", f)
  stopifnot("gene set not found" = file.exists(p))
  unique(read_csv(p, show_col_types = FALSE)$ensembl_gene)
}

pos_3d_all <- read_geneset("hypoxia_3d_promotes_TE_1hr_lfc0.5.csv")
pos_3e_all <- read_geneset("hypoxia_3e_promotes_TE_1hr_lfc0.5.csv")
pos_3d_only <- read_geneset("hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5.csv")
pos_3e_only <- read_geneset("hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv")
pos_both    <- intersect(pos_3d_all, pos_3e_all)

# The pre-built "_not_" files must be the exact complements of the intersect, or the three
# eIF3 figures would not partition the same universe.
stopifnot("not3e file is not the 3d-only complement" =
  setequal(pos_3d_only, setdiff(pos_3d_all, pos_3e_all)))
stopifnot("not3d file is not the 3e-only complement" =
  setequal(pos_3e_only, setdiff(pos_3e_all, pos_3d_all)))

neg_3d <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                   show_col_types = FALSE) %>%
  dplyr::mutate(g = sub("\\..*", "", gene_id)) %>% dplyr::pull(g) %>% unique()
neg_3e <- read_geneset("si3e_negative_controls_padj0.05.csv")
neg_both <- intersect(neg_3d, neg_3e)

cat("eIF3d promotes (all):", length(pos_3d_all),
    " eIF3e promotes (all):", length(pos_3e_all), "\n")
cat("3d only:", length(pos_3d_only), " 3e only:", length(pos_3e_only),
    " both:", length(pos_both), "\n")
cat("negatives - si3d:", length(neg_3d), " si3e:", length(neg_3e),
    " in both:", length(neg_both), "\n")

stopifnot("a positive set overlaps its negative set" =
  length(intersect(pos_3d_only, neg_3d)) == 0,
  length(intersect(pos_3e_only, neg_3e)) == 0,
  length(intersect(pos_both, neg_both)) == 0)
```

The 4E classes come from notebook 76 and are the two tails of one variable, so their
thresholds are read back off the data rather than assumed.

```{r genesets-4e}
res_4e <- read_geneset("teleman_4e_resistant_TE_1hr_lfc0.5.csv")
sen_4e <- read_geneset("teleman_4e_sensitive_TE_1hr_lfc0.5.csv")
stopifnot("4E classes overlap" = length(intersect(res_4e, sen_4e)) == 0)

teleman_te <- read_csv(here("output", "teleman_riboseq_te_all_genes.csv"),
                       show_col_types = FALSE)
res_cut <- min(teleman_te$log2FoldChange[teleman_te$gene_id %in% res_4e], na.rm = TRUE)
sen_cut <- max(teleman_te$log2FoldChange[teleman_te$gene_id %in% sen_4e], na.rm = TRUE)
cat("4E resistant:", length(res_4e), "genes, TE log2FC >=", round(res_cut, 2), "\n")
cat("4E sensitive:", length(sen_4e), "genes, TE log2FC <=", round(sen_cut, 2), "\n")
stopifnot("4E tails are not on opposite sides of zero" = res_cut > 0 && sen_cut < 0)
```

# 2. Plot function

```{r plot-fn}
# Unfilled density curves, one facet per GC feature, stacked vertically.
# `groups` is an ordered named list: label -> vector of gene ids. Genes in none of them
# become the grey rest-of-transcriptome curve, so the curves are disjoint by construction.
gc_density_figure <- function(gc_table, groups, colors, title, subtitle, file = NULL,
                              size_match = TRUE) {

  labelled <- unlist(groups, use.names = FALSE)
  stopifnot("groups overlap each other" = !any(duplicated(labelled)))
  stopifnot("expected exactly two named groups" = length(groups) == 2)

  # Size-match the second group (the negative / comparator class) to the first, counting
  # only genes actually present in this matrix - sampling the raw CSV would leave the
  # plotted curve short. set.seed(9) is the project-wide seed (CLAUDE.md), set immediately
  # before the draw so the selection is reproducible.
  present <- function(ids) intersect(ids, gc_table$gene_id_clean)
  n_ref <- length(present(groups[[1]]))
  n_cmp <- length(present(groups[[2]]))
  dropped <- character(0)

  if (size_match && n_cmp > n_ref) {
    set.seed(9)
    keep <- sample(present(groups[[2]]), n_ref)
    dropped <- setdiff(present(groups[[2]]), keep)
    groups[[2]] <- keep
    cat("Size-matched", names(groups)[2], ":", n_cmp, "->", n_ref,
        "genes (set.seed(9));", length(dropped), "excluded from the figure entirely\n")
  } else if (size_match) {
    cat("No size-matching applied:", names(groups)[2], "has", n_cmp,
        "genes in the matrix vs", n_ref, "in", names(groups)[1],
        "- the comparator is not the larger set, so downsampling it would not match sizes.\n")
  }

  # Genes dropped by size-matching are removed outright rather than folded into the grey
  # curve, which would otherwise be contaminated with negative-control genes.
  d <- gc_table %>%
    dplyr::filter(!gene_id_clean %in% dropped) %>%
    dplyr::mutate(group = REST_LABEL)
  for (nm in names(groups)) {
    d$group[d$gene_id_clean %in% groups[[nm]]] <- nm
  }
  d <- d %>% dplyr::filter(!is.na(group))

  counts <- d %>% dplyr::count(group)
  lev <- c(names(groups), REST_LABEL)
  n_of <- setNames(counts$n[match(lev, counts$group)], lev)
  n_of[is.na(n_of)] <- 0L

  # n goes in the legend so every curve is labelled with its own size.
  lab_of <- setNames(paste0(lev, " (n = ", format(n_of, big.mark = ","), ")"), lev)
  d <- d %>% dplyr::mutate(group = factor(lab_of[group], levels = unname(lab_of)))
  pal <- setNames(unname(colors[lev]), unname(lab_of))

  long <- d %>%
    tidyr::pivot_longer(dplyr::all_of(names(GC_FEATURES)),
                        names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc)) %>%
    dplyr::mutate(feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES)))

  # Per-group, per-facet medians, drawn as dashed verticals in the matching colour.
  meds <- long %>%
    dplyr::group_by(feature, group) %>%
    dplyr::summarise(median_gc = median(gc), n = dplyr::n(), .groups = "drop")

  p <- ggplot(long, aes(x = gc, color = group)) +
    geom_density(fill = NA, linewidth = 0.8) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = group),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    labs(title = title, subtitle = subtitle,
         x = "GC content (%)", y = "Density") +
    theme_classic(base_size = 12) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(0.85, "lines"),
          plot.subtitle = element_text(size = 8.5, lineheight = 1.15),
          strip.background = element_blank(),
          strip.text = element_text(hjust = 0, face = "bold"))

  print(p)
  if (!is.null(file) && SAVE_PLOTS) {
    ggsave(here("plots", file), p, width = 7.5, height = 10)
  }

  # Medians (the dashed verticals above) and a Wilcoxon of the first group vs the others.
  print(as.data.frame(meds %>%
    dplyr::mutate(median_gc = round(median_gc, 1)) %>%
    dplyr::select(feature, group, n, median_gc)), row.names = FALSE)

  ref <- levels(long$group)[1]
  tests <- long %>%
    dplyr::filter(group != ref) %>%
    dplyr::group_by(feature, group) %>%
    dplyr::group_modify(~ {
      a <- long$gc[long$feature == .y$feature & long$group == ref]
      tibble::tibble(p_vs_reference = format.pval(
        wilcox.test(a, .x$gc)$p.value, digits = 3, eps = 2.2e-16))
    }) %>% dplyr::ungroup()
  cat("\nWilcoxon vs", ref, ":\n")
  print(as.data.frame(tests), row.names = FALSE)

  invisible(p)
}
```

# 3. eIF3d-specific targets

```{r fig-3d, fig.width = 7.5, fig.height = 10}
sub_3d <- paste0(
  "Positive: si3d hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3e\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints)\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d-specific targets" = pos_3d_only,
                "Negative control (si3d)" = neg_3d),
  colors = c("eIF3d-specific targets" = "#E69F00",   # Okabe-Ito orange, reserved for eIF3d
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d-specific targets, hypoxia 1hr",
  subtitle = sub_3d,
  file = "89_gc_position_density_3d_specific.pdf")
```

# 4. eIF3e-specific targets

```{r fig-3e, fig.width = 7.5, fig.height = 10}
sub_3e <- paste0(
  "Positive: si3e hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3d\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.05 in all 4 si3e contrasts (normoxia all-timepoints, hypoxia 1/4/24hr)\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3e-specific targets" = pos_3e_only,
                "Negative control (si3e)" = neg_3e),
  colors = c("eIF3e-specific targets" = "#56B4E9",   # Okabe-Ito light blue, reserved for eIF3e
             "Negative control (si3e)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3e-specific targets, hypoxia 1hr",
  subtitle = sub_3e,
  file = "89_gc_position_density_3e_specific.pdf")
```

# 4b. eIF3e-specific targets vs an expressed background (null band)

Section 4 compares the targets to a **screened** negative set - genes selected for having
no eIF3e effect - which is itself GC-rich, so part of the separation in that figure is a
property of the screen rather than of the targets. This section replaces the comparator
with the **expressed background pool** from notebook 02c: every DESeq2-tested gene except
the 3e-promoted ones, with no screening on effect size.

Because the comparator is now drawn from the same population as the grey curve, the black
curves stop being a contrast and become a **null band**: what 399 randomly chosen expressed
mRNAs look like. The question the figure answers is whether the positive curve falls
outside that band.

The band is computed from `r params$n_draws` draws and *plotted* from the first
`r params$n_curves` (a thousand curves is a black smear). Drawing GC values is free, so the
number of draws here is unrelated to the 20 model replicates in notebooks 90/91, which are
limited by the cost of fitting a forest.

```{r bg-pool}
bg_pool <- read_geneset(params$bg_pool_csv)
cat("Expressed background pool:", length(bg_pool), "genes;",
    length(intersect(bg_pool, gc231$gene_id_clean)), "in the GC matrix\n")

# The pool excludes ALL 3e-promoted genes, so it must be disjoint from the 3e-specific set.
stopifnot("background pool overlaps the eIF3e-specific set" =
  length(intersect(bg_pool, pos_3e_only)) == 0)
```

```{r null-band-fn}
# Positive curve + population curve + one thin curve per random draw from the population.
# Unlike gc_density_figure(), the drawn genes are NOT removed from the population curve:
# they are a sample OF it, not a competing group, and excluding them would misrepresent
# what the band is a band around.
gc_null_band_figure <- function(gc_table, pos_ids, pool_ids, pos_label, pos_color,
                                title, subtitle, file = NULL,
                                n_draws = 1000, n_curves = 50, seed = 9) {

  pos_g  <- intersect(pos_ids,  gc_table$gene_id_clean)
  pool_g <- intersect(pool_ids, gc_table$gene_id_clean)
  stopifnot("positive set and pool overlap" = length(intersect(pos_g, pool_g)) == 0)
  stopifnot("pool is not larger than the positive set" = length(pool_g) > length(pos_g))
  n_pos <- length(pos_g)

  set.seed(seed)   # CLAUDE.md project seed, set once before all draws
  draws <- replicate(n_draws, sample(pool_g, n_pos), simplify = FALSE)

  long_of <- function(ids) gc_table %>%
    dplyr::filter(gene_id_clean %in% ids) %>%
    tidyr::pivot_longer(dplyr::all_of(names(GC_FEATURES)),
                        names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc))

  # Medians of every draw, per GC feature: this is the null distribution the empirical
  # p-value is read off. Computed on all n_draws, not just the plotted ones.
  draw_meds <- purrr::map_dfr(seq_along(draws), function(i) {
    long_of(draws[[i]]) %>%
      dplyr::group_by(feature) %>%
      dplyr::summarise(median_gc = median(gc), .groups = "drop") %>%
      dplyr::mutate(draw = i)
  })

  d_pos  <- long_of(pos_g)  %>% dplyr::mutate(grp = pos_label, id = "pos")
  d_pool <- long_of(pool_g) %>% dplyr::mutate(grp = "Expressed background pool", id = "pool")
  d_draw <- purrr::map_dfr(seq_len(min(n_curves, n_draws)), function(i) {
    long_of(draws[[i]]) %>% dplyr::mutate(grp = "Background draws", id = paste0("d", i))
  })

  lab_pos  <- paste0(pos_label, " (n = ", format(n_pos, big.mark = ","), ")")
  lab_draw <- paste0("Background draws (n = ", format(n_pos, big.mark = ","), " each, ",
                     min(n_curves, n_draws), " of ", format(n_draws, big.mark = ","), " shown)")
  lab_pool <- paste0("Expressed background pool (n = ", format(length(pool_g), big.mark = ","), ")")
  relab <- c(pos_label, "Background draws", "Expressed background pool")
  names(relab) <- relab
  relab[] <- c(lab_pos, lab_draw, lab_pool)

  fac <- function(df) df %>%
    dplyr::mutate(grp = factor(relab[grp], levels = c(lab_pos, lab_draw, lab_pool)),
                  feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES)))
  d_pos <- fac(d_pos); d_pool <- fac(d_pool); d_draw <- fac(d_draw)

  meds <- dplyr::bind_rows(d_pos, d_pool) %>%
    dplyr::group_by(feature, grp) %>%
    dplyr::summarise(median_gc = median(gc), .groups = "drop")

  pal <- setNames(c(pos_color, NEG_COLOR, REST_COLOR), c(lab_pos, lab_draw, lab_pool))

  p <- ggplot(mapping = aes(x = gc, color = grp)) +
    geom_density(data = d_draw, aes(group = id), fill = NA, linewidth = 0.25, alpha = 0.5) +
    geom_density(data = d_pool, fill = NA, linewidth = 0.9) +
    geom_density(data = d_pos,  fill = NA, linewidth = 0.9) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = grp),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    guides(color = guide_legend(override.aes = list(linewidth = 0.9))) +
    labs(title = title, subtitle = subtitle, x = "GC content (%)", y = "Density") +
    theme_classic(base_size = 12) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(0.85, "lines"),
          plot.subtitle = element_text(size = 8.5, lineheight = 1.15),
          strip.background = element_blank(),
          strip.text = element_text(hjust = 0, face = "bold"))
  print(p)
  if (!is.null(file) && SAVE_PLOTS) ggsave(here("plots", file), p, width = 7.5, height = 10)

  # Empirical p: how many random draws have a median at least as extreme as the positive
  # set's, in the direction the positive set actually sits. (k + 1) / (n + 1) so the
  # reported p can never be 0 - with n_draws draws the floor is 1 / (n_draws + 1).
  pos_meds  <- d_pos  %>% dplyr::group_by(feature) %>%
    dplyr::summarise(pos_median = median(gc), .groups = "drop")
  pool_meds <- d_pool %>% dplyr::group_by(feature) %>%
    dplyr::summarise(pool_median = median(gc), .groups = "drop")

  stats <- draw_meds %>%
    dplyr::mutate(feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES))) %>%
    dplyr::group_by(feature) %>%
    dplyr::summarise(draw_mean = mean(median_gc), draw_sd = sd(median_gc),
                     draw_lo = quantile(median_gc, 0.025),
                     draw_hi = quantile(median_gc, 0.975),
                     .groups = "drop") %>%
    dplyr::left_join(pos_meds, by = "feature") %>%
    dplyr::left_join(pool_meds, by = "feature")

  stats <- stats %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      k = if (pos_median <= draw_mean) {
            sum(draw_meds$median_gc[GC_FEATURES[draw_meds$feature] == as.character(feature)] <= pos_median)
          } else {
            sum(draw_meds$median_gc[GC_FEATURES[draw_meds$feature] == as.character(feature)] >= pos_median)
          },
      empirical_p = (k + 1) / (n_draws + 1),
      z = (pos_median - draw_mean) / draw_sd) %>%
    dplyr::ungroup()

  cat("\nPositive-set median vs the null distribution of", n_draws, "draws of", n_pos,
      "genes:\n")
  print(as.data.frame(stats %>% dplyr::transmute(
    feature,
    pos_median  = round(pos_median, 1),
    pool_median = round(pool_median, 1),
    null_mean   = round(draw_mean, 1),
    null_95pct  = paste0(round(draw_lo, 1), "-", round(draw_hi, 1)),
    z           = round(z, 1),
    empirical_p = ifelse(empirical_p <= 1 / (n_draws + 1),
                         paste0("<", signif(1 / (n_draws + 1), 2)),
                         signif(empirical_p, 3)))), row.names = FALSE)

  invisible(list(plot = p, stats = stats))
}
```

```{r fig-3e-nullband, fig.width = 7.5, fig.height = 10}
sub_3e_bg <- paste0(
  "Positive: si3e hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3d\n",
  "Black: ", params$n_curves, " random draws of the same size from the expressed background pool (nb02c);\n",
  "  pool = DESeq2-tested genes minus ALL 3e-promoted genes, not screened on effect size\n",
  "Grey: the whole pool. The draws are a sample OF the grey curve, so they overlap it by design")

nullband_3e <- gc_null_band_figure(
  gc231,
  pos_ids   = pos_3e_only,
  pool_ids  = bg_pool,
  pos_label = "eIF3e-specific targets",
  pos_color = "#56B4E9",                # Okabe-Ito light blue, reserved for eIF3e
  title     = "GC by codon position: eIF3e-specific targets vs an expressed background",
  subtitle  = sub_3e_bg,
  file      = "89_gc_position_nullband_3e_specific.pdf",
  n_draws   = params$n_draws,
  n_curves  = params$n_curves,
  seed      = params$draw_seed)
```

# 5. eIF3d + eIF3e intersect

```{r fig-both, fig.width = 7.5, fig.height = 10}
sub_both <- paste0(
  "Positive: TE LFC > 0.5 & padj < 0.05 in BOTH si3d and si3e, hypoxia 1hr\n",
  "Negative: in both negative pools - si3d |TE LFC| < 0.5 & padj > 0.3, and si3e |TE LFC| < 0.5 & padj > 0.05\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d + eIF3e targets" = pos_both,
                "Negative in both"      = neg_both),
  colors = c("eIF3d + eIF3e targets" = "#009E73",   # Okabe-Ito green
             "Negative in both"      = NEG_COLOR,
             "Rest of transcriptome" = REST_COLOR),
  title = "GC by codon position: mRNAs promoted by both eIF3d and eIF3e, hypoxia 1hr",
  subtitle = sub_both,
  file = "89_gc_position_density_3d_3e_intersect.pdf")
```

# 6. eIF3d targets specific to hypoxia

The two condition-specific sets are pre-built. Their definitions are asserted here rather
than taken from the filenames, because the file descriptions do not record **which**
normoxia contrast the exclusion used: it is the **1+4hr time-regressed** analysis
(`~ hours + Condition + SeqType + Condition:SeqType`), not normoxia 1hr and not
alltimepoints.

```{r condition-specific-sets}
pos_hyp_only <- read_geneset("hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv")
pos_nor_only <- read_geneset("normoxia_3d_promotes_TE_normoxia_specific_lfc0.5.csv")

promoted_in <- function(f) {
  d <- read_csv(here("output", f), show_col_types = FALSE)
  # EIF3D's own rna_lfc must be strongly positive or the contrast direction has flipped
  # (CLAUDE.md: the filenames state the ratio backwards from the coefficient).
  stopifnot("EIF3D rna_lfc is not positive - direction flipped" =
    d$rna_lfc[match("EIF3D", d$symbol)] > 1)
  unique(sub("\\..*", "", d$gene_id[d$te_lfc > 0.5 & d$te_padj < 0.05 & !is.na(d$te_padj)]))
}
hyp_1hr <- promoted_in("translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")
nor_1and4 <- promoted_in("translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv")

stopifnot("hypoxia_specific is not hypoxia 1hr minus normoxia 1+4hr" =
  setequal(pos_hyp_only, setdiff(hyp_1hr, nor_1and4)))
stopifnot("normoxia_specific is not normoxia 1+4hr minus hypoxia 1hr" =
  setequal(pos_nor_only, setdiff(nor_1and4, hyp_1hr)))
stopifnot("the two condition-specific sets overlap" =
  length(intersect(pos_hyp_only, pos_nor_only)) == 0)

cat("hypoxia 1hr promoted:", length(hyp_1hr),
    " normoxia 1+4hr promoted:", length(nor_1and4), "\n")
cat("hypoxia-specific:", length(pos_hyp_only),
    " normoxia-specific:", length(pos_nor_only), "\n")
```

```{r fig-hyp-specific, fig.width = 7.5, fig.height = 10}
sub_hyp <- paste0(
  "Positive: si3d TE LFC > 0.5 & padj < 0.05 in hypoxia 1hr, excluding genes also promoted in normoxia 1+4hr (time-regressed)\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints), size-matched\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d targets, hypoxia-specific" = pos_hyp_only,
                "Negative control (si3d)"         = neg_3d),
  colors = c("eIF3d targets, hypoxia-specific" = "#E69F00",
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d targets specific to hypoxia",
  subtitle = sub_hyp,
  file = "89_gc_position_density_3d_hypoxia_specific.pdf")
```

# 7. eIF3d targets specific to normoxia

Far smaller than its hypoxia counterpart, so the curves are correspondingly rougher; the
group size is printed in the legend.

```{r fig-nor-specific, fig.width = 7.5, fig.height = 10}
sub_nor <- paste0(
  "Positive: si3d TE LFC > 0.5 & padj < 0.05 in normoxia 1+4hr (time-regressed), excluding genes also promoted in hypoxia 1hr\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints), size-matched\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d targets, normoxia-specific" = pos_nor_only,
                "Negative control (si3d)"          = neg_3d),
  colors = c("eIF3d targets, normoxia-specific" = "#E69F00",
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d targets specific to normoxia",
  subtitle = sub_nor,
  file = "89_gc_position_density_3d_normoxia_specific.pdf")
```

# 8. 4E-resistant vs 4E-sensitive

Different dataset and cell line (Roiuk et al. HeLa, 4E-BP inhibition), so this figure is
drawn on the HeLa transcript matrix from notebook 78. Both classes are tails of the same
variable rather than a positive set against unaffected controls, which is why the grey
curve sits between them rather than on top of one.

```{r fig-4e, fig.width = 7.5, fig.height = 10}
sub_4e <- paste0(
  "Resistant: top ", length(res_4e), " genes by TE log2FC on 4E-BP inhibition (>= ",
  round(res_cut, 2), ", padj < 0.05)\n",
  "Sensitive: bottom ", length(sen_4e), " genes (<= ", round(sen_cut, 2), ", padj < 0.05)\n",
  "Grey: HeLa transcriptome excluding both sets. Positive log2FC = TE held up = resistant")

gc_density_figure(
  gc4e,
  groups = list("4E-resistant" = res_4e,
                "4E-sensitive" = sen_4e),
  colors = c("4E-resistant"          = "#D55E00",   # Okabe-Ito vermillion
             "4E-sensitive"          = "#0072B2",   # Okabe-Ito blue
             "Rest of transcriptome" = REST_COLOR),
  title = "GC by codon position: 4E-resistant vs 4E-sensitive mRNAs (HeLa)",
  subtitle = sub_4e,
  file = "89_gc_position_density_4e.pdf")
```

# 9. Session info

```{r session-info}
sessionInfo()
```
