---
title: "GC content by codon position across eIF3d, eIF3e and 4E gene sets"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  save_plots: TRUE
  bg_pool_csv: "si3e_expressed_background_pool.csv"
  n_draws: 1000
  n_curves: 50
  draw_seed: 9
  fig_base_size: 16
  fig_width: 8.5
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Unfilled density curves of **CDS GC, GC1, GC2 and GC3**, faceted vertically, for six
gene-set comparisons:

1. **eIF3d-specific** targets (hypoxia 1hr, excluding genes also promoted by eIF3e)
2. **eIF3e-specific** targets (hypoxia 1hr, excluding genes also promoted by eIF3d)
3. The **eIF3d + eIF3e intersect** (promoted by both)
4. eIF3d targets **specific to hypoxia** (not promoted in normoxia 1+4hr)
5. eIF3d targets **specific to normoxia 1+4hr** (not promoted in hypoxia 1hr)
6. **4E-resistant vs 4E-sensitive** (Teleman HeLa 4E-BP inhibition)

Section 4b adds a seventh figure: the eIF3e-specific set again, but against the unscreened
expressed background as a **null band** rather than against a negative control. It is built
by a separate function and does not follow the layout described next.

Each panel set shows the positive class, its negative class, and the rest of the
transcriptome, each with a dashed vertical at its own median. The rest-of-transcriptome
group **excludes** both labelled sets, so the three curves are disjoint and "all other
genes" is literally true.

The negative control is **size-matched to the positive set** with `set.seed(9)`, counting
genes present in the feature matrix rather than rows in the source CSV. Genes dropped by
the match are removed from the figure entirely, not folded into the grey curve, which would
otherwise put negative controls into "all other genes". Matching only shrinks a larger
comparator; where the negative pool is already the smaller of the two (the eIF3d + eIF3e
intersect) no matching is possible and the chunk output says so.

## Where the GC columns come from

The model matrices (`feature_matrix_external_stability.rds`, `..._dhx29_kd_feature.rds`)
carry `cds_gc` and `gc3` but **not** `gc1`/`gc2`; those live in
`feature_matrix_positional_gc.rds`, which by design drops `cds_gc` (notebook 01v, so the
forest cannot fall back on the aggregate). Neither file alone can draw these figures, so
this notebook joins them on `transcript_id_clean` and asserts that the `gc3` column they
share is identical.

The Teleman HeLa matrix (`feature_matrix_teleman_4e.rds`, notebook 78) carries all four
columns already and is keyed on HeLa dominant transcripts, so it is used as-is for the 4E
figure (section 8) and never mixed with the MDA-MB-231 rows.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(save_plots  = TRUE,
                 bg_pool_csv = "si3e_expressed_background_pool.csv",
                 n_draws     = 1000,  # draws behind the empirical p (cheap: GC values only)
                 n_curves    = 50,    # draws actually drawn on the figure
                 draw_seed   = 9,
                 fig_base_size = 16,  # ggplot base_size for every density figure
                 fig_width     = 8.5) # inches; widened with the text so subtitles still fit
}
SAVE_PLOTS <- params$save_plots

# Text size for every figure in this notebook. The subtitle carries the full gene-set
# definition and is the longest line on the page, so it is scaled at ~0.62 of base_size
# and the canvas is widened with the text - at base_size 16 on the old 7.5in width the
# longest subtitle line (121 chars, section 6) runs past the right edge.
FIG_BASE_SIZE  <- params$fig_base_size
FIG_SUB_SIZE   <- FIG_BASE_SIZE * 0.62
FIG_WIDTH      <- params$fig_width
FIG_HEIGHT     <- 10.5

# Facet order, top to bottom. The three codon positions come first and the
# aggregate sits at the BOTTOM: it is ~the mean of the other three, so reading it
# last makes each panel's contribution to it visible rather than pre-empting them.
# Every other use of this vector (range-check loops, all_of() selection, factor
# levels) is order-independent, so this ordering is purely the figure layout.
GC_FEATURES <- c(gc1    = "GC1, first codon position",
                 gc2    = "GC2, second codon position",
                 gc3    = "GC3, third codon position",
                 cds_gc = "CDS GC, all positions")

REST_LABEL <- "Rest of transcriptome"
REST_COLOR <- "grey70"   # CLAUDE.md: new plots use grey70 for the rest of the transcriptome
NEG_COLOR  <- "black"    # CLAUDE.md: negative control set is black
```

# 1. Data

```{r load-231}
fm_model <- readRDS(here("output", "predictive_modeling", "feature_matrix_external_stability.rds"))
fm_pos   <- readRDS(here("output", "predictive_modeling", "feature_matrix_positional_gc.rds"))

stopifnot("cds_gc missing from the model matrix" = "cds_gc" %in% names(fm_model),
          "gc1/gc2 missing from the positional matrix" = all(c("gc1", "gc2") %in% names(fm_pos)))

gc231 <- fm_model %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol, cds_gc, gc3_model = gc3) %>%
  dplyr::inner_join(fm_pos %>% dplyr::select(transcript_id_clean, gc1, gc2, gc3),
                    by = "transcript_id_clean")

# The two matrices must describe the same transcripts, or gc1/gc2 would be attached to a
# different isoform than cds_gc.
stopifnot("gc3 disagrees between the two matrices" =
  max(abs(gc231$gc3_model - gc231$gc3), na.rm = TRUE) == 0)
gc231 <- gc231 %>% dplyr::select(-gc3_model) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)

cat("MDA-MB-231 genes:", nrow(gc231), "\n")
for (f in names(GC_FEATURES)) {
  stopifnot("GC outside [0, 100]" = all(gc231[[f]] >= 0 & gc231[[f]] <= 100, na.rm = TRUE))
}
cat("GC3 median:", round(median(gc231$gc3, na.rm = TRUE), 1), "% (expected 40-65)\n")
stopifnot("GC3 median out of range" = between(median(gc231$gc3, na.rm = TRUE), 40, 65))
```

```{r load-teleman}
gc4e <- readRDS(here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds")) %>%
  dplyr::select(transcript_id_clean, gene_id_clean, symbol,
                dplyr::all_of(names(GC_FEATURES))) %>%
  dplyr::distinct(gene_id_clean, .keep_all = TRUE)
cat("HeLa (Teleman) genes:", nrow(gc4e), "\n")
for (f in names(GC_FEATURES)) {
  stopifnot("GC outside [0, 100]" = all(gc4e[[f]] >= 0 & gc4e[[f]] <= 100, na.rm = TRUE))
}
```

## Gene sets

```{r genesets}
read_geneset <- function(f) {
  p <- here("output", "genesets", f)
  stopifnot("gene set not found" = file.exists(p))
  unique(read_csv(p, show_col_types = FALSE)$ensembl_gene)
}

pos_3d_all <- read_geneset("hypoxia_3d_promotes_TE_1hr_lfc0.5.csv")
pos_3e_all <- read_geneset("hypoxia_3e_promotes_TE_1hr_lfc0.5.csv")
pos_3d_only <- read_geneset("hypoxia_3d_promotes_TE_1hr_not3e_lfc0.5.csv")
pos_3e_only <- read_geneset("hypoxia_3e_promotes_TE_1hr_not3d_lfc0.5.csv")
pos_both    <- intersect(pos_3d_all, pos_3e_all)

# The pre-built "_not_" files must be the exact complements of the intersect, or the three
# eIF3 figures would not partition the same universe.
stopifnot("not3e file is not the 3d-only complement" =
  setequal(pos_3d_only, setdiff(pos_3d_all, pos_3e_all)))
stopifnot("not3d file is not the 3e-only complement" =
  setequal(pos_3e_only, setdiff(pos_3e_all, pos_3d_all)))

neg_3d <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"),
                   show_col_types = FALSE) %>%
  dplyr::mutate(g = sub("\\..*", "", gene_id)) %>% dplyr::pull(g) %>% unique()
neg_3e <- read_geneset("si3e_negative_controls_padj0.05.csv")
neg_both <- intersect(neg_3d, neg_3e)

cat("eIF3d promotes (all):", length(pos_3d_all),
    " eIF3e promotes (all):", length(pos_3e_all), "\n")
cat("3d only:", length(pos_3d_only), " 3e only:", length(pos_3e_only),
    " both:", length(pos_both), "\n")
cat("negatives - si3d:", length(neg_3d), " si3e:", length(neg_3e),
    " in both:", length(neg_both), "\n")

stopifnot("a positive set overlaps its negative set" =
  length(intersect(pos_3d_only, neg_3d)) == 0,
  length(intersect(pos_3e_only, neg_3e)) == 0,
  length(intersect(pos_both, neg_both)) == 0)
```

The 4E classes come from notebook 76 and are the two tails of one variable, so their
thresholds are read back off the data rather than assumed.

```{r genesets-4e}
res_4e <- read_geneset("teleman_4e_resistant_TE_1hr_lfc0.5.csv")
sen_4e <- read_geneset("teleman_4e_sensitive_TE_1hr_lfc0.5.csv")
stopifnot("4E classes overlap" = length(intersect(res_4e, sen_4e)) == 0)

teleman_te <- read_csv(here("output", "teleman_riboseq_te_all_genes.csv"),
                       show_col_types = FALSE)
res_cut <- min(teleman_te$log2FoldChange[teleman_te$gene_id %in% res_4e], na.rm = TRUE)
sen_cut <- max(teleman_te$log2FoldChange[teleman_te$gene_id %in% sen_4e], na.rm = TRUE)
cat("4E resistant:", length(res_4e), "genes, TE log2FC >=", round(res_cut, 2), "\n")
cat("4E sensitive:", length(sen_4e), "genes, TE log2FC <=", round(sen_cut, 2), "\n")
stopifnot("4E tails are not on opposite sides of zero" = res_cut > 0 && sen_cut < 0)
```

# 2. Plot function

```{r plot-fn}
# Unfilled density curves, one facet per GC feature, stacked vertically.
# `groups` is an ordered named list: label -> vector of gene ids. Genes in none of them
# become the grey rest-of-transcriptome curve, so the curves are disjoint by construction.
gc_density_figure <- function(gc_table, groups, colors, title, subtitle, file = NULL,
                              size_match = TRUE) {

  labelled <- unlist(groups, use.names = FALSE)
  stopifnot("groups overlap each other" = !any(duplicated(labelled)))
  # One group = positive set against the rest of the transcriptome only (no comparator
  # class). Two = positive plus a negative/comparator class, which is then size-matched.
  stopifnot("expected one or two named groups" = length(groups) %in% c(1L, 2L))

  # Size-match the second group (the negative / comparator class) to the first, counting
  # only genes actually present in this matrix - sampling the raw CSV would leave the
  # plotted curve short. set.seed(9) is the project-wide seed (CLAUDE.md), set immediately
  # before the draw so the selection is reproducible.
  present <- function(ids) intersect(ids, gc_table$gene_id_clean)
  n_ref <- length(present(groups[[1]]))
  n_cmp <- if (length(groups) == 2) length(present(groups[[2]])) else NA_integer_
  dropped <- character(0)
  size_match <- size_match && length(groups) == 2   # nothing to match against with one group

  if (size_match && n_cmp > n_ref) {
    set.seed(9)
    keep <- sample(present(groups[[2]]), n_ref)
    dropped <- setdiff(present(groups[[2]]), keep)
    groups[[2]] <- keep
    cat("Size-matched", names(groups)[2], ":", n_cmp, "->", n_ref,
        "genes (set.seed(9));", length(dropped), "excluded from the figure entirely\n")
  } else if (size_match) {
    cat("No size-matching applied:", names(groups)[2], "has", n_cmp,
        "genes in the matrix vs", n_ref, "in", names(groups)[1],
        "- the comparator is not the larger set, so downsampling it would not match sizes.\n")
  }

  # Genes dropped by size-matching are removed outright rather than folded into the grey
  # curve, which would otherwise be contaminated with negative-control genes.
  d <- gc_table %>%
    dplyr::filter(!gene_id_clean %in% dropped) %>%
    dplyr::mutate(group = REST_LABEL)
  for (nm in names(groups)) {
    d$group[d$gene_id_clean %in% groups[[nm]]] <- nm
  }
  d <- d %>% dplyr::filter(!is.na(group))

  counts <- d %>% dplyr::count(group)
  lev <- c(names(groups), REST_LABEL)
  n_of <- setNames(counts$n[match(lev, counts$group)], lev)
  n_of[is.na(n_of)] <- 0L

  # n goes in the legend so every curve is labelled with its own size.
  lab_of <- setNames(paste0(lev, " (n = ", format(n_of, big.mark = ","), ")"), lev)
  d <- d %>% dplyr::mutate(group = factor(lab_of[group], levels = unname(lab_of)))
  pal <- setNames(unname(colors[lev]), unname(lab_of))

  long <- d %>%
    tidyr::pivot_longer(dplyr::all_of(names(GC_FEATURES)),
                        names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc)) %>%
    dplyr::mutate(feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES)))

  # Per-group, per-facet medians, drawn as dashed verticals in the matching colour.
  meds <- long %>%
    dplyr::group_by(feature, group) %>%
    dplyr::summarise(median_gc = median(gc), n = dplyr::n(), .groups = "drop")

  p <- ggplot(long, aes(x = gc, color = group)) +
    geom_density(fill = NA, linewidth = 0.8) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = group),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    labs(title = title, subtitle = subtitle,
         x = "GC content (%)", y = "Density") +
    theme_classic(base_size = FIG_BASE_SIZE) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(0.85, "lines"),
          plot.subtitle = element_text(size = FIG_SUB_SIZE, lineheight = 1.15),
          strip.background = element_blank(),
          strip.text = element_text(hjust = 0, face = "bold"))

  print(p)
  if (!is.null(file) && SAVE_PLOTS) {
    ggsave(here("plots", file), p, width = FIG_WIDTH, height = FIG_HEIGHT)
  }

  # Medians (the dashed verticals above) and a Wilcoxon of the first group vs the others.
  print(as.data.frame(meds %>%
    dplyr::mutate(median_gc = round(median_gc, 1)) %>%
    dplyr::select(feature, group, n, median_gc)), row.names = FALSE)

  ref <- levels(long$group)[1]
  tests <- long %>%
    dplyr::filter(group != ref) %>%
    dplyr::group_by(feature, group) %>%
    dplyr::group_modify(~ {
      a <- long$gc[long$feature == .y$feature & long$group == ref]
      tibble::tibble(p_vs_reference = format.pval(
        wilcox.test(a, .x$gc)$p.value, digits = 3, eps = 2.2e-16))
    }) %>% dplyr::ungroup()
  cat("\nWilcoxon vs", ref, ":\n")
  print(as.data.frame(tests), row.names = FALSE)

  invisible(p)
}
```

# 3. eIF3d-specific targets

```{r fig-3d, fig.width = 7.5, fig.height = 10}
sub_3d <- paste0(
  "Positive: si3d hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3e\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints)\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d-specific targets" = pos_3d_only,
                "Negative control (si3d)" = neg_3d),
  colors = c("eIF3d-specific targets" = "#E69F00",   # Okabe-Ito orange, reserved for eIF3d
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d-specific targets, hypoxia 1hr",
  subtitle = sub_3d,
  file = "89_gc_position_density_3d_specific.pdf")
```

# 4. eIF3e-specific targets

```{r fig-3e, fig.width = 7.5, fig.height = 10}
sub_3e <- paste0(
  "Positive: si3e hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3d\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.05 in all 4 si3e contrasts (normoxia all-timepoints, hypoxia 1/4/24hr)\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3e-specific targets" = pos_3e_only,
                "Negative control (si3e)" = neg_3e),
  colors = c("eIF3e-specific targets" = "#56B4E9",   # Okabe-Ito light blue, reserved for eIF3e
             "Negative control (si3e)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3e-specific targets, hypoxia 1hr",
  subtitle = sub_3e,
  file = "89_gc_position_density_3e_specific.pdf")
```

# 4b. eIF3e-specific targets vs an expressed background (null band)

Section 4 compares the targets to a **screened** negative set - genes selected for having
no eIF3e effect - which is itself GC-rich, so part of the separation in that figure is a
property of the screen rather than of the targets. This section replaces the comparator
with the **expressed background pool** from notebook 02c: every DESeq2-tested gene except
the 3e-promoted ones, with no screening on effect size.

Because the comparator is now drawn from the same population as the grey curve, the black
curves stop being a contrast and become a **null band**: what the same number of randomly
chosen expressed mRNAs look like (399 for this set; the figure legend carries the count). The question the figure answers is whether the positive curve falls
outside that band.

The band is computed from `r params$n_draws` draws and *plotted* from the first
`r params$n_curves` (a thousand curves is a black smear). Drawing GC values is free, so the
number of draws here is unrelated to the 20 model replicates in notebooks 90/91, which are
limited by the cost of fitting a forest.

```{r bg-pool}
bg_pool <- read_geneset(params$bg_pool_csv)
cat("Expressed background pool:", length(bg_pool), "genes;",
    length(intersect(bg_pool, gc231$gene_id_clean)), "in the GC matrix\n")

# The pool excludes ALL 3e-promoted genes, so it must be disjoint from the 3e-specific set.
stopifnot("background pool overlaps the eIF3e-specific set" =
  length(intersect(bg_pool, pos_3e_only)) == 0)
```

```{r null-band-fn}
# Positive curve + population curve + one thin curve per random draw from the population.
# Unlike gc_density_figure(), the drawn genes are NOT removed from the population curve:
# they are a sample OF it, not a competing group, and excluding them would misrepresent
# what the band is a band around.
gc_null_band_figure <- function(gc_table, pos_ids, pool_ids, pos_label, pos_color,
                                title, subtitle, file = NULL,
                                n_draws = 1000, n_curves = 50, seed = 9) {

  pos_g  <- intersect(pos_ids,  gc_table$gene_id_clean)
  pool_g <- intersect(pool_ids, gc_table$gene_id_clean)
  stopifnot("positive set and pool overlap" = length(intersect(pos_g, pool_g)) == 0)
  stopifnot("pool is not larger than the positive set" = length(pool_g) > length(pos_g))
  n_pos <- length(pos_g)

  set.seed(seed)   # CLAUDE.md project seed, set once before all draws
  draws <- replicate(n_draws, sample(pool_g, n_pos), simplify = FALSE)

  long_of <- function(ids) gc_table %>%
    dplyr::filter(gene_id_clean %in% ids) %>%
    tidyr::pivot_longer(dplyr::all_of(names(GC_FEATURES)),
                        names_to = "feature", values_to = "gc") %>%
    dplyr::filter(!is.na(gc))

  # Medians of every draw, per GC feature: this is the null distribution the empirical
  # p-value is read off. Computed on all n_draws, not just the plotted ones.
  draw_meds <- purrr::map_dfr(seq_along(draws), function(i) {
    long_of(draws[[i]]) %>%
      dplyr::group_by(feature) %>%
      dplyr::summarise(median_gc = median(gc), .groups = "drop") %>%
      dplyr::mutate(draw = i)
  })

  d_pos  <- long_of(pos_g)  %>% dplyr::mutate(grp = pos_label, id = "pos")
  d_pool <- long_of(pool_g) %>% dplyr::mutate(grp = "Expressed background pool", id = "pool")
  d_draw <- purrr::map_dfr(seq_len(min(n_curves, n_draws)), function(i) {
    long_of(draws[[i]]) %>% dplyr::mutate(grp = "Background draws", id = paste0("d", i))
  })

  lab_pos  <- paste0(pos_label, " (n = ", format(n_pos, big.mark = ","), ")")
  lab_draw <- paste0("Background draws (n = ", format(n_pos, big.mark = ","), " each, ",
                     min(n_curves, n_draws), " of ", format(n_draws, big.mark = ","), " shown)")
  lab_pool <- paste0("Expressed background pool (n = ", format(length(pool_g), big.mark = ","), ")")
  relab <- c(pos_label, "Background draws", "Expressed background pool")
  names(relab) <- relab
  relab[] <- c(lab_pos, lab_draw, lab_pool)

  fac <- function(df) df %>%
    dplyr::mutate(grp = factor(relab[grp], levels = c(lab_pos, lab_draw, lab_pool)),
                  feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES)))
  d_pos <- fac(d_pos); d_pool <- fac(d_pool); d_draw <- fac(d_draw)

  meds <- dplyr::bind_rows(d_pos, d_pool) %>%
    dplyr::group_by(feature, grp) %>%
    dplyr::summarise(median_gc = median(gc), .groups = "drop")

  pal <- setNames(c(pos_color, NEG_COLOR, REST_COLOR), c(lab_pos, lab_draw, lab_pool))

  p <- ggplot(mapping = aes(x = gc, color = grp)) +
    geom_density(data = d_draw, aes(group = id), fill = NA, linewidth = 0.25, alpha = 0.5) +
    geom_density(data = d_pool, fill = NA, linewidth = 0.9) +
    geom_density(data = d_pos,  fill = NA, linewidth = 0.9) +
    geom_vline(data = meds, aes(xintercept = median_gc, color = grp),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~ feature, ncol = 1, scales = "free_y") +
    scale_color_manual(values = pal, name = NULL) +
    guides(color = guide_legend(override.aes = list(linewidth = 0.9))) +
    labs(title = title, subtitle = subtitle, x = "GC content (%)", y = "Density") +
    theme_classic(base_size = FIG_BASE_SIZE) +
    theme(legend.position = "top", legend.direction = "vertical",
          legend.key.height = unit(0.85, "lines"),
          plot.subtitle = element_text(size = FIG_SUB_SIZE, lineheight = 1.15),
          strip.background = element_blank(),
          strip.text = element_text(hjust = 0, face = "bold"))
  print(p)
  if (!is.null(file) && SAVE_PLOTS) ggsave(here("plots", file), p, width = FIG_WIDTH, height = FIG_HEIGHT)

  # Empirical p: how many random draws have a median at least as extreme as the positive
  # set's, in the direction the positive set actually sits. (k + 1) / (n + 1) so the
  # reported p can never be 0 - with n_draws draws the floor is 1 / (n_draws + 1).
  pos_meds  <- d_pos  %>% dplyr::group_by(feature) %>%
    dplyr::summarise(pos_median = median(gc), .groups = "drop")
  pool_meds <- d_pool %>% dplyr::group_by(feature) %>%
    dplyr::summarise(pool_median = median(gc), .groups = "drop")

  stats <- draw_meds %>%
    dplyr::mutate(feature = factor(GC_FEATURES[feature], levels = unname(GC_FEATURES))) %>%
    dplyr::group_by(feature) %>%
    dplyr::summarise(draw_mean = mean(median_gc), draw_sd = sd(median_gc),
                     draw_lo = quantile(median_gc, 0.025),
                     draw_hi = quantile(median_gc, 0.975),
                     .groups = "drop") %>%
    dplyr::left_join(pos_meds, by = "feature") %>%
    dplyr::left_join(pool_meds, by = "feature")

  stats <- stats %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      k = if (pos_median <= draw_mean) {
            sum(draw_meds$median_gc[GC_FEATURES[draw_meds$feature] == as.character(feature)] <= pos_median)
          } else {
            sum(draw_meds$median_gc[GC_FEATURES[draw_meds$feature] == as.character(feature)] >= pos_median)
          },
      empirical_p = (k + 1) / (n_draws + 1),
      z = (pos_median - draw_mean) / draw_sd) %>%
    dplyr::ungroup()

  cat("\nPositive-set median vs the null distribution of", n_draws, "draws of", n_pos,
      "genes:\n")
  print(as.data.frame(stats %>% dplyr::transmute(
    feature,
    pos_median  = round(pos_median, 1),
    pool_median = round(pool_median, 1),
    null_mean   = round(draw_mean, 1),
    null_95pct  = paste0(round(draw_lo, 1), "-", round(draw_hi, 1)),
    z           = round(z, 1),
    empirical_p = ifelse(empirical_p <= 1 / (n_draws + 1),
                         paste0("<", signif(1 / (n_draws + 1), 2)),
                         signif(empirical_p, 3)))), row.names = FALSE)

  invisible(list(plot = p, stats = stats))
}
```

```{r fig-3e-nullband, fig.width = 7.5, fig.height = 10}
sub_3e_bg <- paste0(
  "Positive: si3e hypoxia 1hr, TE LFC > 0.5 & padj < 0.05, excluding genes also promoted by eIF3d\n",
  "Black: ", params$n_curves, " random draws of the same size from the expressed background pool (nb02c);\n",
  "  pool = DESeq2-tested genes minus ALL 3e-promoted genes, not screened on effect size\n",
  "Grey: the whole pool. The draws are a sample OF the grey curve, so they overlap it by design")

nullband_3e <- gc_null_band_figure(
  gc231,
  pos_ids   = pos_3e_only,
  pool_ids  = bg_pool,
  pos_label = "eIF3e-specific targets",
  pos_color = "#56B4E9",                # Okabe-Ito light blue, reserved for eIF3e
  title     = "GC by codon position: eIF3e-specific targets vs an expressed background",
  subtitle  = sub_3e_bg,
  file      = "89_gc_position_nullband_3e_specific.pdf",
  n_draws   = params$n_draws,
  n_curves  = params$n_curves,
  seed      = params$draw_seed)
```

# 5. eIF3d + eIF3e intersect

```{r fig-both, fig.width = 7.5, fig.height = 10}
sub_both <- paste0(
  "Positive: TE LFC > 0.5 & padj < 0.05 in BOTH si3d and si3e, hypoxia 1hr\n",
  "Negative: in both negative pools - si3d |TE LFC| < 0.5 & padj > 0.3, and si3e |TE LFC| < 0.5 & padj > 0.05\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d + eIF3e targets" = pos_both,
                "Negative in both"      = neg_both),
  colors = c("eIF3d + eIF3e targets" = "#009E73",   # Okabe-Ito green
             "Negative in both"      = NEG_COLOR,
             "Rest of transcriptome" = REST_COLOR),
  title = "GC by codon position: mRNAs promoted by both eIF3d and eIF3e, hypoxia 1hr",
  subtitle = sub_both,
  file = "89_gc_position_density_3d_3e_intersect.pdf")
```

# 6. eIF3d targets specific to hypoxia

The two condition-specific sets are pre-built. Their definitions are asserted here rather
than taken from the filenames, because the file descriptions do not record **which**
normoxia contrast the exclusion used: it is the **1+4hr time-regressed** analysis
(`~ hours + Condition + SeqType + Condition:SeqType`), not normoxia 1hr and not
alltimepoints.

```{r condition-specific-sets}
pos_hyp_only <- read_geneset("hypoxia_3d_promotes_TE_hypoxia_specific_lfc0.5.csv")
pos_nor_only <- read_geneset("normoxia_3d_promotes_TE_normoxia_specific_lfc0.5.csv")

promoted_in <- function(f) {
  d <- read_csv(here("output", f), show_col_types = FALSE)
  # EIF3D's own rna_lfc must be strongly positive or the contrast direction has flipped
  # (CLAUDE.md: the filenames state the ratio backwards from the coefficient).
  stopifnot("EIF3D rna_lfc is not positive - direction flipped" =
    d$rna_lfc[match("EIF3D", d$symbol)] > 1)
  unique(sub("\\..*", "", d$gene_id[d$te_lfc > 0.5 & d$te_padj < 0.05 & !is.na(d$te_padj)]))
}
hyp_1hr <- promoted_in("translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv")
nor_1and4 <- promoted_in("translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv")

stopifnot("hypoxia_specific is not hypoxia 1hr minus normoxia 1+4hr" =
  setequal(pos_hyp_only, setdiff(hyp_1hr, nor_1and4)))
stopifnot("normoxia_specific is not normoxia 1+4hr minus hypoxia 1hr" =
  setequal(pos_nor_only, setdiff(nor_1and4, hyp_1hr)))
stopifnot("the two condition-specific sets overlap" =
  length(intersect(pos_hyp_only, pos_nor_only)) == 0)

cat("hypoxia 1hr promoted:", length(hyp_1hr),
    " normoxia 1+4hr promoted:", length(nor_1and4), "\n")
cat("hypoxia-specific:", length(pos_hyp_only),
    " normoxia-specific:", length(pos_nor_only), "\n")
```

```{r fig-hyp-specific, fig.width = 7.5, fig.height = 10}
sub_hyp <- paste0(
  "Positive: si3d TE LFC > 0.5 & padj < 0.05 in hypoxia 1hr, excluding genes also promoted in normoxia 1+4hr (time-regressed)\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints), size-matched\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d targets, hypoxia-specific" = pos_hyp_only,
                "Negative control (si3d)"         = neg_3d),
  colors = c("eIF3d targets, hypoxia-specific" = "#E69F00",
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d targets specific to hypoxia",
  subtitle = sub_hyp,
  file = "89_gc_position_density_3d_hypoxia_specific.pdf")
```

# 7. eIF3d targets specific to normoxia

Far smaller than its hypoxia counterpart, so the curves are correspondingly rougher; the
group size is printed in the legend.

```{r fig-nor-specific, fig.width = 7.5, fig.height = 10}
sub_nor <- paste0(
  "Positive: si3d TE LFC > 0.5 & padj < 0.05 in normoxia 1+4hr (time-regressed), excluding genes also promoted in hypoxia 1hr\n",
  "Negative: |TE LFC| < 0.5 & padj > 0.3 in all 4 si3d contrasts (hypoxia 1/4/24hr, normoxia all-timepoints), size-matched\n",
  "Grey: MDA-MB-231 transcriptome excluding both sets")

gc_density_figure(
  gc231,
  groups = list("eIF3d targets, normoxia-specific" = pos_nor_only,
                "Negative control (si3d)"          = neg_3d),
  colors = c("eIF3d targets, normoxia-specific" = "#E69F00",
             "Negative control (si3d)" = NEG_COLOR,
             "Rest of transcriptome"   = REST_COLOR),
  title = "GC by codon position: eIF3d targets specific to normoxia",
  subtitle = sub_nor,
  file = "89_gc_position_density_3d_normoxia_specific.pdf")
```

# 7b. eIF3e targets specific to hypoxia / normoxia (background only)

Exploratory counterpart to sections 6 and 7. **No negative control class here** — just the
positive set against the rest of the transcriptome.

That is deliberate. The si3e negative pool is a *screened* set (genes chosen for showing no
eIF3e effect) and is heavily GC-biased: median GC3 **72.9%** against **57.2%** for the whole
matrix, while the unscreened expressed-background pool sits at **57.5%**, i.e. right on the
transcriptome. On an axis that *is* GC content, that comparator would supply much of the
separation by construction. Dropping it also removes two other problems: the si3e negative
pool has only 963 genes in this matrix (too few to size-match a normoxia-specific set), and
a handful of genes fall in both classes because nb02b screens on normoxia *alltimepoints*
while these positives are defined on the normoxia *1+4hr* fit.

**The threshold is not settled.** `te_lfc > 0.5` is the project convention and is what
sections 6 and 7 use for si3d, but every si3e normoxia contrast is shifted well above zero
(median 0.62 at 1+4hr against si3d's 0.16), so 0.5 there selects roughly "above the median".
Both versions are drawn so the shift is visible:

- **raw** — `> 0.5` on both legs, the literal convention
- **centred** — each leg cut the same distance above its own contrast's median that 0.5
  sits above si3d's (+0.211 hypoxia, +0.342 normoxia, derived at knit time)

Nothing is written to `output/genesets/`; these sets live only in this notebook for now.

```{r e3-cond-sets-bg}
si3d_hyp_tbl <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_hypoxia_1hr.csv"),
                         show_col_types = FALSE)
si3d_nor_tbl <- read_csv(here("output", "translation_categories_si3d_vs_sictrl_normoxia_1and4hr.csv"),
                         show_col_types = FALSE)
si3e_hyp_tbl <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_hypoxia_1hr.csv"),
                         show_col_types = FALSE)
# TE-only table (gene_id / te_lfc / te_padj / symbol): the time-regressed 1+4hr fit written
# by eif3e_eif3d_normoxia_and_hypoxia.Rmd. No rna_lfc column, so it is checked indirectly.
si3e_nor_tbl <- read_csv(here("output", "te_si3e_vs_sictrl_normoxia_1and4hr.csv"),
                         show_col_types = FALSE)

assert_kd_rna <- function(tbl, label, sym) {
  stopifnot("rna_lfc absent - cannot verify direction" = "rna_lfc" %in% names(tbl))
  v <- tbl$rna_lfc[match(sym, tbl$symbol)]
  cat(sym, " rna_lfc (", label, "): ", round(v, 2), " - expect strongly positive\n", sep = "")
  stopifnot("knockdown target rna_lfc not positive - direction flipped" = is.finite(v) && v > 1)
}
assert_kd_rna(si3d_hyp_tbl, "si3d hypoxia 1hr",    "EIF3D")
assert_kd_rna(si3d_nor_tbl, "si3d normoxia 1+4hr", "EIF3D")
assert_kd_rna(si3e_hyp_tbl, "si3e hypoxia 1hr",    "EIF3E")

si3e_nor_1hr <- read_csv(here("output", "translation_categories_si3e_vs_sictrl_normoxia_1hr.csv"),
                         show_col_types = FALSE)
assert_kd_rna(si3e_nor_1hr, "si3e normoxia 1hr (sign reference)", "EIF3E")
sign_ref <- si3e_nor_tbl %>%
  dplyr::select(gene_id, te_lfc) %>%
  dplyr::inner_join(si3e_nor_1hr %>% dplyr::select(gene_id, te_lfc_1hr = te_lfc), by = "gene_id")
rho_ref <- cor(sign_ref$te_lfc, sign_ref$te_lfc_1hr, method = "spearman", use = "complete.obs")
cat("si3e normoxia 1+4hr vs 1hr TE LFC, Spearman rho: ", round(rho_ref, 3),
    " (n = ", nrow(sign_ref), ") - expect strongly positive\n", sep = "")
stopifnot("si3e normoxia 1+4hr sign disagrees with the 1hr contrast" = rho_ref > 0.5)

SI3D_CUT <- 0.5
off_hyp  <- SI3D_CUT - median(si3d_hyp_tbl$te_lfc, na.rm = TRUE)
off_nor  <- SI3D_CUT - median(si3d_nor_tbl$te_lfc, na.rm = TRUE)
cat(sprintf("si3d 0.5 cut sits +%.3f above its hypoxia median, +%.3f above its normoxia median\n",
            off_hyp, off_nor))
stopifnot("a si3d offset is not positive" = off_hyp > 0 && off_nor > 0)

promoted_at <- function(tbl, cut) {
  unique(sub("\\..*", "", tbl$gene_id[tbl$te_lfc > cut & tbl$te_padj < 0.05 & !is.na(tbl$te_padj)]))
}
e3_cuts <- list(
  raw     = c(hyp = SI3D_CUT, nor = SI3D_CUT),
  centred = c(hyp = median(si3e_hyp_tbl$te_lfc, na.rm = TRUE) + off_hyp,
              nor = median(si3e_nor_tbl$te_lfc, na.rm = TRUE) + off_nor))

e3_sets <- lapply(names(e3_cuts), function(k) {
  h <- promoted_at(si3e_hyp_tbl, e3_cuts[[k]][["hyp"]])
  n <- promoted_at(si3e_nor_tbl, e3_cuts[[k]][["nor"]])
  list(mode = k, hyp_cut = e3_cuts[[k]][["hyp"]], nor_cut = e3_cuts[[k]][["nor"]],
       hyp = setdiff(h, n), nor = setdiff(n, h), n_h = length(h), n_n = length(n))
})
names(e3_sets) <- names(e3_cuts)

cat("\nsi3e condition-specific set sizes (in-matrix counts appear in each figure legend):\n")
print(as.data.frame(purrr::map_dfr(e3_sets, ~ tibble::tibble(
  mode = .x$mode, hyp_cut = round(.x$hyp_cut, 3), nor_cut = round(.x$nor_cut, 3),
  hyp_promoted = .x$n_h, nor_promoted = .x$n_n,
  hyp_specific = length(.x$hyp), nor_specific = length(.x$nor)))), row.names = FALSE)

stopifnot(
  "condition-specific sets overlap" =
    all(vapply(e3_sets, function(s) length(intersect(s$hyp, s$nor)) == 0, logical(1))),
  "a condition-specific set is empty" =
    all(vapply(e3_sets, function(s) length(s$hyp) > 0 && length(s$nor) > 0, logical(1))))
```

```{r fig-3e-cond-bg, fig.width = 7.5, fig.height = 10}
e3_bg_figure <- function(mode, leg) {
  s   <- e3_sets[[mode]]
  pos <- if (leg == "hyp") s$hyp else s$nor
  cut <- if (leg == "hyp") s$hyp_cut else s$nor_cut
  oth <- if (leg == "hyp") s$nor_cut else s$hyp_cut
  cond     <- if (leg == "hyp") "hypoxia" else "normoxia"
  this_con <- if (leg == "hyp") "hypoxia 1hr" else "normoxia 1+4hr (time-regressed)"
  excl_con <- if (leg == "hyp") "normoxia 1+4hr (time-regressed)" else "hypoxia 1hr"
  pos_lab  <- paste0("eIF3e targets, ", cond, "-specific")

  gc_density_figure(
    gc231,
    groups = setNames(list(pos), pos_lab),
    colors = setNames(c("#56B4E9", REST_COLOR), c(pos_lab, REST_LABEL)),
    title = paste0("GC by codon position: eIF3e targets specific to ", cond,
                   " (", mode, " cut)"),
    subtitle = paste0(
      "Positive: si3e TE LFC > ", round(cut, 2), " & padj < 0.05 in ", this_con,
      ", excluding genes also promoted in ", excl_con, " (TE LFC > ", round(oth, 2), ")\n",
      if (mode == "centred")
        paste0("Cut placed the same distance above this contrast's median that 0.5 sits above si3d's (+",
               round(if (leg == "hyp") off_hyp else off_nor, 2), ")\n")
      else
        "Raw 0.5 cut, the project convention; si3e normoxia median is 0.62, so on that leg this is near-median\n",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class ",
      "(the si3e screened pool is GC-biased, median GC3 72.9 vs 57.2)"),
    file = paste0("89_gc_position_density_3e_", cond, "_specific_", mode, "_bg.pdf"))
}

for (m in c("centred", "raw")) for (l in c("hyp", "nor")) e3_bg_figure(m, l)
```

## Top 200 by effect size, normoxia-specific

A rank-based cut instead of a threshold. This sidesteps the median-shift problem entirely:
taking the N largest TE LFCs does not care where the contrast's centre sits, so it needs no
si3d-derived offset and is directly comparable in *size* to a top-N set from any other
contrast.

Two nested versions, both refilled to exactly `TOP_N` after their exclusions:

1. **normoxia-specific** - drop genes eIF3e also promotes in hypoxia 1hr
2. **+ not 3d-hypoxia** - additionally drop genes eIF3**d** promotes in hypoxia 1hr, so the
   set is enriched for mRNAs that are eIF3e-dependent in normoxia and *not* accounted for by
   the eIF3d hypoxia programme

Both exclusions use the project convention (`te_lfc > 0.5 & padj < 0.05`) on their hypoxia
leg. The eIF3d set is the same `hypoxia_3d_promotes_TE_1hr_lfc0.5.csv` used by section 3.

```{r e3-top200-normoxia, fig.width = 7.5, fig.height = 10}
TOP_N <- 200

# Exclusion legs, both at the project 0.5 convention.
hyp_promoted_3e <- promoted_at(si3e_hyp_tbl, SI3D_CUT)   # eIF3e, hypoxia 1hr
hyp_promoted_3d <- pos_3d_all                            # eIF3d, hypoxia 1hr (section 2)

# Ranked pool of significant normoxia-specific genes, deduplicated to one row per gene
# BEFORE head() so the top-N is N distinct genes rather than N rows.
top_pool <- si3e_nor_tbl %>%
  dplyr::filter(!is.na(te_padj), te_padj < 0.05) %>%
  dplyr::mutate(g = sub("\\..*", "", gene_id)) %>%
  dplyr::arrange(dplyr::desc(te_lfc)) %>%
  dplyr::distinct(g, .keep_all = TRUE)

top_variants <- list(
  list(key = "top200",
       excl = hyp_promoted_3e,
       lab  = paste0("eIF3e normoxia-specific (top ", TOP_N, ")"),
       desc = "excluding genes also promoted by eIF3e in hypoxia 1hr"),
  list(key = "top200_not3dhyp",
       excl = union(hyp_promoted_3e, hyp_promoted_3d),
       lab  = paste0("eIF3e normoxia-specific, not 3d-hypoxia (top ", TOP_N, ")"),
       desc = "excluding genes promoted by eIF3e in hypoxia 1hr OR by eIF3d in hypoxia 1hr")
)

for (v in top_variants) {
  avail <- top_pool %>% dplyr::filter(!g %in% v$excl)
  sel   <- head(avail, TOP_N)
  stopifnot(
    "fewer genes than TOP_N survive the filters" = nrow(sel) == TOP_N,
    "a selected gene is in the exclusion set"    = !any(sel$g %in% v$excl),
    "a selected gene is not significant"         = all(sel$te_padj < 0.05),
    "selected TE LFCs are not all positive"      = all(sel$te_lfc > 0))

  n_mat <- length(intersect(sel$g, gc231$gene_id_clean))
  cat(sprintf("%-16s pool %d available -> top %d | TE LFC %.2f to %.2f | %d in the GC matrix\n",
              v$key, nrow(avail), TOP_N, min(sel$te_lfc), max(sel$te_lfc), n_mat))

  gc_density_figure(
    gc231,
    groups = setNames(list(sel$g), v$lab),
    colors = setNames(c("#56B4E9", REST_COLOR), c(v$lab, REST_LABEL)),
    title = paste0("GC by codon position: ", v$lab),
    subtitle = paste0(
      "Positive: the ", TOP_N, " largest si3e TE LFCs in normoxia 1+4hr (time-regressed) at padj < 0.05,\n",
      "  ", v$desc, " (TE LFC > ", SI3D_CUT, " & padj < 0.05). Implied LFC floor ",
      round(min(sel$te_lfc), 2), "\n",
      "Rank-based, so unaffected by the si3e normoxia median shift that makes a 0.5 cut near-median\n",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class"),
    file = paste0("89_gc_position_density_3e_normoxia_specific_", v$key, "_bg.pdf"))
}
```

## Unfiltered promoted sets (no condition exclusion)

Sections 6, 7 and 7b all subtract one condition's promoted set from the other's. For si3e
that subtraction is destructive: the si3e **normoxia** contrasts call 35-75% of the
transcriptome significant with a strong global GC3 association (rho -0.35 to -0.43), so
removing "also promoted in normoxia" from the si3e hypoxia set deletes 833 of its 986 genes,
and the deleted genes are the AT-rich ones (median GC3 52.0). The 153-gene remainder then
looks GC-rich (73.6) purely as the residue of that cut.

These figures drop the exclusion. Each is simply **promoted in one contrast**
(`te_lfc > 0.5 & padj < 0.05`, the project convention) against the rest of the
transcriptome. The eIF3d pair is included as the reference the eIF3e pair has to be read
against - same threshold, same background, same matrix.

```{r unfiltered-promoted, fig.width = 7.5, fig.height = 10}
unfiltered_sets <- list(
  list(key = "3d_hyp1hr",  tbl = si3d_hyp_tbl, kd = "eIF3d", col = "#E69F00",
       con = "hypoxia 1hr"),
  list(key = "3d_nor1and4", tbl = si3d_nor_tbl, kd = "eIF3d", col = "#E69F00",
       con = "normoxia 1+4hr (time-regressed)"),
  list(key = "3e_hyp1hr",  tbl = si3e_hyp_tbl, kd = "eIF3e", col = "#56B4E9",
       con = "hypoxia 1hr"),
  list(key = "3e_nor1and4", tbl = si3e_nor_tbl, kd = "eIF3e", col = "#56B4E9",
       con = "normoxia 1+4hr (time-regressed)")
)

unfiltered_summary <- purrr::map_dfr(unfiltered_sets, function(u) {
  ids   <- promoted_at(u$tbl, SI3D_CUT)
  inmat <- intersect(ids, gc231$gene_id_clean)
  lab   <- paste0(u$kd, " promoted, ", u$con)

  gc_density_figure(
    gc231,
    groups = setNames(list(ids), lab),
    colors = setNames(c(u$col, REST_COLOR), c(lab, REST_LABEL)),
    title = paste0("GC by codon position: ", lab, " (no exclusion)"),
    subtitle = paste0(
      "Positive: ", u$kd, " TE LFC > ", SI3D_CUT, " & padj < 0.05 in ", u$con,
      ". NO condition exclusion applied\n",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class\n",
      "Contrast-level context: median TE LFC ", round(median(u$tbl$te_lfc, na.rm = TRUE), 3),
      ", ", round(100 * mean(u$tbl$te_padj < 0.05, na.rm = TRUE)), "% of genes at padj < 0.05"),
    file = paste0("89_gc_position_density_unfiltered_", u$key, "_bg.pdf"))

  s <- gc231 %>% dplyr::filter(gene_id_clean %in% inmat)
  tibble::tibble(set = lab, n = nrow(s),
                 cds_gc = round(median(s$cds_gc), 1), gc1 = round(median(s$gc1), 1),
                 gc2 = round(median(s$gc2), 1), gc3 = round(median(s$gc3), 1),
                 pct_sig = round(100 * mean(u$tbl$te_padj < 0.05, na.rm = TRUE)),
                 rho_gc3_te = round(cor(
                   gc231$gc3[match(sub("\\..*", "", u$tbl$gene_id), gc231$gene_id_clean)],
                   u$tbl$te_lfc, method = "spearman", use = "complete.obs"), 3))
})

cat("\nUnfiltered promoted sets vs background (whole matrix medians: cds_gc ",
    round(median(gc231$cds_gc), 1), ", gc1 ", round(median(gc231$gc1), 1),
    ", gc2 ", round(median(gc231$gc2), 1), ", gc3 ", round(median(gc231$gc3), 1), "):\n", sep = "")
print(as.data.frame(unfiltered_summary), row.names = FALSE)
```

## Top 500 by effect size, no exclusions

Same four contrasts, but each set is the **N largest TE LFCs at padj < 0.05**, no condition
exclusion. This equalises set size, which the unfiltered figures above do not: there the
eIF3e normoxia set had 4,961 genes against eIF3e hypoxia's 986, so any difference between
them could have been depth of the list rather than composition.

`TOP_N_UNF` is 500. **eIF3d normoxia 1+4hr cannot reach it** - only 430 genes are
significant with a positive LFC in that contrast - so it is drawn at its maximum and its
true n is reported. The other three are exactly 500.

```{r top500-no-exclusion, fig.width = 7.5, fig.height = 10}
TOP_N_UNF <- 500

top_n_promoted <- function(tbl, n) {
  tbl %>%
    dplyr::filter(!is.na(te_padj), te_padj < 0.05, te_lfc > 0) %>%
    dplyr::mutate(g = sub("\\..*", "", gene_id)) %>%
    dplyr::arrange(dplyr::desc(te_lfc)) %>%
    dplyr::distinct(g, .keep_all = TRUE) %>%
    head(n)
}

top500_ids <- list()
top500_summary <- purrr::map_dfr(unfiltered_sets, function(u) {
  sel <- top_n_promoted(u$tbl, TOP_N_UNF)
  top500_ids[[u$key]] <<- sel$g
  capped <- nrow(sel) < TOP_N_UNF
  lab <- paste0(u$kd, " top ", nrow(sel), ", ", u$con)

  gc_density_figure(
    gc231,
    groups = setNames(list(sel$g), lab),
    colors = setNames(c(u$col, REST_COLOR), c(lab, REST_LABEL)),
    title = paste0("GC by codon position: ", u$kd, " top ", nrow(sel), " by TE LFC, ", u$con),
    subtitle = paste0(
      "Positive: the ", nrow(sel), " largest ", u$kd, " TE LFCs at padj < 0.05 in ", u$con,
      ". NO condition exclusion. Implied LFC floor ", round(min(sel$te_lfc), 2), "\n",
      if (capped) paste0("NOTE: only ", nrow(sel), " genes are significant with LFC > 0 here, so ",
                         TOP_N_UNF, " was not reachable - this set is NOT size-matched to the others\n")
      else "Rank-based and size-matched to the other top-500 sets\n",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class"),
    file = paste0("89_gc_position_density_top", TOP_N_UNF, "_", u$key, "_bg.pdf"))

  s <- gc231 %>% dplyr::filter(gene_id_clean %in% sel$g)
  tibble::tibble(set = lab, n_selected = nrow(sel), n_in_matrix = nrow(s),
                 lfc_floor = round(min(sel$te_lfc), 2),
                 cds_gc = round(median(s$cds_gc), 1), gc1 = round(median(s$gc1), 1),
                 gc2 = round(median(s$gc2), 1), gc3 = round(median(s$gc3), 1))
})

cat("\nTop-", TOP_N_UNF, " sets vs background (matrix medians: cds_gc ",
    round(median(gc231$cds_gc), 1), ", gc1 ", round(median(gc231$gc1), 1),
    ", gc2 ", round(median(gc231$gc2), 1), ", gc3 ", round(median(gc231$gc3), 1), "):\n", sep = "")
print(as.data.frame(top500_summary), row.names = FALSE)
```

### Normoxia / hypoxia overlap

How much of each factor's normoxia set is the same genes as its hypoxia set. Reported both
for the top-N sets above and for the full promoted sets, since the two answer different
questions: the top-N overlap says whether the *strongest* responders are shared, the full
overlap says whether the programmes are shared at all.

```{r norm-hyp-overlap}
overlap_row <- function(kd, a, b, lab_a, lab_b) {
  i <- length(intersect(a, b))
  tibble::tibble(
    kd = kd, comparison = paste0(lab_a, " vs ", lab_b),
    n_normoxia = length(a), n_hypoxia = length(b), overlap = i,
    pct_of_normoxia = round(100 * i / length(a), 1),
    pct_of_hypoxia  = round(100 * i / length(b), 1),
    jaccard = round(i / length(union(a, b)), 3))
}

full_3d_nor <- promoted_at(si3d_nor_tbl, SI3D_CUT); full_3d_hyp <- promoted_at(si3d_hyp_tbl, SI3D_CUT)
full_3e_nor <- promoted_at(si3e_nor_tbl, SI3D_CUT); full_3e_hyp <- promoted_at(si3e_hyp_tbl, SI3D_CUT)

overlap_tbl <- dplyr::bind_rows(
  overlap_row("eIF3d", top500_ids$`3d_nor1and4`, top500_ids$`3d_hyp1hr`, "top-N nor", "top-N hyp"),
  overlap_row("eIF3e", top500_ids$`3e_nor1and4`, top500_ids$`3e_hyp1hr`, "top-N nor", "top-N hyp"),
  overlap_row("eIF3d", full_3d_nor, full_3d_hyp, "all promoted nor", "all promoted hyp"),
  overlap_row("eIF3e", full_3e_nor, full_3e_hyp, "all promoted nor", "all promoted hyp"))
print(as.data.frame(overlap_tbl), row.names = FALSE)

# Cross-factor overlap too: do eIF3d and eIF3e promote the same mRNAs within a condition?
cross_tbl <- dplyr::bind_rows(
  overlap_row("3d vs 3e", full_3d_hyp, full_3e_hyp, "3d hypoxia", "3e hypoxia"),
  overlap_row("3d vs 3e", full_3d_nor, full_3e_nor, "3d normoxia", "3e normoxia"))
cat("\nCross-factor overlap within a condition (first column = eIF3d set):\n")
print(as.data.frame(cross_tbl), row.names = FALSE)
```

### Venn: normoxia 1+4hr vs hypoxia 1hr, eIF3d and eIF3e

The same unfiltered sets drawn above (`te_lfc > 0.5 & padj < 0.05`, no condition exclusion),
as two-circle Venns - one per factor. Each is drawn twice because the two counts answer
different questions and only one of them matches the density figures:

1. **all promoted genes** - every gene passing the cut in each contrast
2. **restricted to the GC matrix** - the n printed in the legends of the unfiltered density
   figures above (the matrix is one row per gene's dominant transcript, so it is a subset)

Within a Venn both circles are the same factor, so the hue separates *condition*, not
factor: the hypoxia circle keeps that factor's reserved colour (eIF3d orange, eIF3e blue)
and normoxia takes Okabe-Ito reddish purple, which is reserved for neither.

**The two factors' Venns are not read the same way.** The eIF3d normoxia contrast calls only
4% of genes significant against hypoxia's 14%, so its normoxia circle is the smaller one and
the overlap is bounded by it. The eIF3e **normoxia** contrast is the opposite extreme - a
0.5 cut there sits essentially at its own median (median TE LFC 0.62), so it calls thousands
of genes and the eIF3e hypoxia circle ends up nearly nested inside it. That asymmetry is a
property of where each contrast's distribution sits, not of how shared the programmes are;
the sizes are printed so it is visible rather than implied.

The overlap is tested against the genes DESeq2 could **test** in both contrasts (non-NA
`te_padj` in each), not against the whole annotation - a hypergeometric p on an inflated
universe is meaningless. A gene promoted in one contrast can be untested in the other, so
those genes are dropped for the test only and the count is reported on the figure.

```{r venn-nor-hyp, fig.width = 8, fig.height = 6.5}
library(ggvenn)

# The circles are the sets exactly as the unfiltered density figures define them. The
# hypergeometric p sits on a stricter footing: the genes DESeq2 could TEST in BOTH
# contrasts (te_padj non-NA in each, i.e. surviving independent filtering).
tested_in <- function(tbl) unique(sub("\\..*", "", tbl$gene_id[!is.na(tbl$te_padj)]))

VENN_NOR_COL <- "#CC79A7"   # Okabe-Ito reddish purple - separates condition, not factor

venn_specs <- list(
  list(kd = "eIF3d", key = "3d", hyp_col = "#E69F00",
       nor_tbl = si3d_nor_tbl, hyp_tbl = si3d_hyp_tbl,
       nor = full_3d_nor, hyp = full_3d_hyp),
  list(kd = "eIF3e", key = "3e", hyp_col = "#56B4E9",
       nor_tbl = si3e_nor_tbl, hyp_tbl = si3e_hyp_tbl,
       nor = full_3e_nor, hyp = full_3e_hyp))

venn_condition <- function(spec, restrict_to_matrix) {
  stopifnot("a promoted gene is not tested in its own contrast" =
              all(spec$nor %in% tested_in(spec$nor_tbl)) &&
              all(spec$hyp %in% tested_in(spec$hyp_tbl)))

  nor <- if (restrict_to_matrix) intersect(spec$nor, gc231$gene_id_clean) else spec$nor
  hyp <- if (restrict_to_matrix) intersect(spec$hyp, gc231$gene_id_clean) else spec$hyp
  ov  <- intersect(nor, hyp)

  # Enrichment of the overlap, on the both-tested universe only.
  uni <- intersect(tested_in(spec$nor_tbl), tested_in(spec$hyp_tbl))
  if (restrict_to_matrix) uni <- intersect(uni, gc231$gene_id_clean)
  nor_t <- intersect(nor, uni); hyp_t <- intersect(hyp, uni); ov_t <- intersect(nor_t, hyp_t)
  n_drop <- length(union(nor, hyp)) - length(union(nor_t, hyp_t))
  expected <- length(nor_t) * length(hyp_t) / length(uni)
  p <- phyper(length(ov_t) - 1L, length(nor_t), length(uni) - length(nor_t),
              length(hyp_t), lower.tail = FALSE)
  stopifnot("overlap exceeds a set size" = length(ov) <= min(length(nor), length(hyp)))

  lst <- list(nor, hyp)
  names(lst) <- c("Normoxia 1+4hr", "Hypoxia 1hr")

  p_venn <- ggvenn(lst, fill_color = c(VENN_NOR_COL, spec$hyp_col),
                   fill_alpha = 0.45, stroke_size = 0.4, stroke_color = "grey30",
                   set_name_size = 4.6, text_size = 4.2, show_percentage = FALSE) +
    labs(title = paste0("mRNAs promoted by ", spec$kd, ": normoxia 1+4hr vs hypoxia 1hr",
                        if (restrict_to_matrix) " (GC matrix)" else ""),
         subtitle = paste0(
           "si", spec$key, " TE LFC > ", SI3D_CUT, " & padj < 0.05 in each contrast; ",
           "normoxia is the 1+4hr time-regressed fit. No condition exclusion\n",
           round(100 * length(ov) / length(nor), 1), "% of the normoxia set and ",
           round(100 * length(ov) / length(hyp), 1),
           "% of the hypoxia set are shared (Jaccard ",
           round(length(ov) / length(union(nor, hyp)), 3), ")\n",
           "Overlap ", length(ov_t), " vs ", round(expected, 1),
           " expected by chance (", round(length(ov_t) / expected, 1),
           "x), hypergeometric p = ", format.pval(p, digits = 2), "\n",
           "Test universe: ", format(length(uni), big.mark = ","),
           " genes tested in both contrasts",
           if (restrict_to_matrix) " and in the GC matrix" else "", "; ", n_drop,
           " set genes untested in the other contrast are excluded from the test only")) +
    theme(plot.title.position = "plot",
          plot.title = element_text(size = 12, face = "bold"),
          plot.subtitle = element_text(size = 7.5, colour = "grey30"),
          plot.margin = margin(10, 14, 10, 14))
  print(p_venn)

  if (SAVE_PLOTS) {
    f <- here("plots", paste0("89_venn_", spec$key, "_promoted_normoxia_vs_hypoxia",
                              if (restrict_to_matrix) "_gcmatrix" else "", ".pdf"))
    ggsave(f, p_venn, width = 8, height = 6.5)
    cat("saved:", basename(f), "\n")
  }

  tibble::tibble(kd = spec$kd,
                 scope = if (restrict_to_matrix) "GC matrix" else "all promoted",
                 normoxia = length(nor), hypoxia = length(hyp), overlap = length(ov),
                 normoxia_only = length(setdiff(nor, hyp)),
                 hypoxia_only = length(setdiff(hyp, nor)),
                 pct_of_normoxia = round(100 * length(ov) / length(nor), 1),
                 pct_of_hypoxia = round(100 * length(ov) / length(hyp), 1),
                 jaccard = round(length(ov) / length(union(nor, hyp)), 3),
                 test_universe = length(uni), expected = round(expected, 1),
                 fold_enrichment = round(length(ov_t) / expected, 2),
                 hypergeom_p = format.pval(p, digits = 3))
}

venn_summary <- purrr::map_dfr(venn_specs, function(s)
  purrr::map_dfr(c(FALSE, TRUE), function(r) venn_condition(s, r)))
cat("\nPromoted normoxia 1+4hr vs hypoxia 1hr, both factors:\n")
print(as.data.frame(venn_summary), row.names = FALSE)

# The GC-matrix rows must reproduce the n printed in the unfiltered density figures above.
for (kd in c("eIF3d", "eIF3e")) {
  r <- venn_summary[venn_summary$kd == kd & venn_summary$scope == "GC matrix", ]
  stopifnot("Venn in-matrix counts disagree with the unfiltered density figures" =
    r$normoxia == unfiltered_summary$n[
      unfiltered_summary$set == paste0(kd, " promoted, normoxia 1+4hr (time-regressed)")] &&
    r$hypoxia == unfiltered_summary$n[
      unfiltered_summary$set == paste0(kd, " promoted, hypoxia 1hr")])
}
```

## Promoted sets with the normoxia/hypoxia overlap removed

The same four contrasts again, but each set now has the genes it **shares with its own
factor's other condition** removed: normoxia-only = promoted in normoxia minus promoted in
hypoxia, and vice versa. Threshold and comparator are unchanged (`te_lfc > 0.5 & padj < 0.05`,
background = rest of transcriptome).

**Read these against the unfiltered versions above, not on their own.** The subtraction is
not symmetric between factors and it is severe for eIF3e: 82.4% of the eIF3e hypoxia set is
also in its normoxia set, so removing the overlap deletes 892 of 1,083 genes. The genes it
deletes are the AT-rich ones, so the 191-gene remainder inverts to look GC-rich. The
before/after table below makes the size of that cut explicit for each panel; where the
"removed" column is most of the set, the resulting median says more about what was taken out
than about what is left.

```{r no-overlap-sets, fig.width = 7.5, fig.height = 10}
med_gc3 <- function(ids) {
  s <- gc231 %>% dplyr::filter(gene_id_clean %in% ids)
  if (nrow(s) == 0) return(NA_real_)
  round(median(s$gc3), 1)
}

no_overlap_summary <- purrr::map_dfr(unfiltered_sets, function(u) {
  self <- promoted_at(u$tbl, SI3D_CUT)
  # the same factor's other condition
  other_tbl <- if (identical(u$kd, "eIF3d")) {
    if (grepl("hypoxia", u$con)) si3d_nor_tbl else si3d_hyp_tbl
  } else {
    if (grepl("hypoxia", u$con)) si3e_nor_tbl else si3e_hyp_tbl
  }
  other  <- promoted_at(other_tbl, SI3D_CUT)
  kept    <- setdiff(self, other)
  removed <- intersect(self, other)
  stopifnot("kept and removed do not partition the set" =
              length(kept) + length(removed) == length(self))

  lab <- paste0(u$kd, " ", if (grepl("hypoxia", u$con)) "hypoxia" else "normoxia", "-only")
  gc_density_figure(
    gc231,
    groups = setNames(list(kept), lab),
    colors = setNames(c(u$col, REST_COLOR), c(lab, REST_LABEL)),
    title = paste0("GC by codon position: ", lab, " (overlap removed)"),
    subtitle = paste0(
      "Positive: ", u$kd, " TE LFC > ", SI3D_CUT, " & padj < 0.05 in ", u$con,
      ", MINUS genes also promoted in the other condition\n",
      "Removed ", length(removed), " of ", length(self), " (",
      round(100 * length(removed) / length(self)), "%); removed genes have median GC3 ",
      med_gc3(removed), " vs ", med_gc3(self), " for the full set\n",
      if (length(removed) / length(self) > 0.5)
        "CAUTION: most of the set was removed - this median reflects the subtraction as much as the biology\n"
      else "",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class"),
    file = paste0("89_gc_position_density_nooverlap_", u$key, "_bg.pdf"))

  tibble::tibble(
    set = lab, n_full = length(self), n_removed = length(removed), n_kept = length(kept),
    pct_removed = round(100 * length(removed) / length(self)),
    gc3_full = med_gc3(self), gc3_removed = med_gc3(removed), gc3_kept = med_gc3(kept))
})

cat("\nEffect of removing the normoxia/hypoxia overlap (background GC3 = ",
    round(median(gc231$gc3), 1), "):\n", sep = "")
print(as.data.frame(no_overlap_summary), row.names = FALSE)
```

## Overlap removed, size-matched to the smallest set

The overlap-removed sets above differ enormously in size (95 to 4,128 genes in the matrix),
so their curves differ in smoothness and their medians in precision. Here every set is cut
to the **top N by TE LFC**, where N is the in-matrix size of the smallest of them - the
eIF3d normoxia-only set. N is derived, not hardcoded, so it tracks the data.

Ranking is done **among genes present in the GC matrix**, not among all selected genes, so
each panel plots exactly N curves rather than N-minus-whatever-was-missing.

This equalises n but not effect size: the LFC floor each set reaches at N is reported, and
they are not comparable (eIF3d normoxia-only takes its whole set to get there).

```{r no-overlap-top-matched, fig.width = 7.5, fig.height = 10}
# Rebuild the four overlap-removed sets, then find N from the smallest.
no_overlap_ids <- lapply(unfiltered_sets, function(u) {
  self <- promoted_at(u$tbl, SI3D_CUT)
  other_tbl <- if (identical(u$kd, "eIF3d")) {
    if (grepl("hypoxia", u$con)) si3d_nor_tbl else si3d_hyp_tbl
  } else {
    if (grepl("hypoxia", u$con)) si3e_nor_tbl else si3e_hyp_tbl
  }
  intersect(setdiff(self, promoted_at(other_tbl, SI3D_CUT)), gc231$gene_id_clean)
})
names(no_overlap_ids) <- vapply(unfiltered_sets, function(u) u$key, character(1))

N_MATCH <- length(no_overlap_ids$`3d_nor1and4`)
cat("Size-matching all overlap-removed sets to the eIF3d normoxia-only set: N =", N_MATCH,
    "\nIn-matrix sizes before matching:\n")
print(vapply(no_overlap_ids, length, integer(1)))
stopifnot("N_MATCH is not the smallest in-matrix set" =
            N_MATCH == min(vapply(no_overlap_ids, length, integer(1))))

matched_summary <- purrr::map_dfr(unfiltered_sets, function(u) {
  ids <- no_overlap_ids[[u$key]]
  sel <- u$tbl %>%
    dplyr::mutate(g = sub("\\..*", "", gene_id)) %>%
    dplyr::filter(g %in% ids) %>%
    dplyr::arrange(dplyr::desc(te_lfc)) %>%
    dplyr::distinct(g, .keep_all = TRUE) %>%
    head(N_MATCH)
  stopifnot("selection is not exactly N_MATCH genes" = nrow(sel) == N_MATCH,
            "a selected gene is not in the GC matrix" = all(sel$g %in% gc231$gene_id_clean))

  cond <- if (grepl("hypoxia", u$con)) "hypoxia" else "normoxia"
  lab  <- paste0(u$kd, " ", cond, "-only, top ", N_MATCH)
  gc_density_figure(
    gc231,
    groups = setNames(list(sel$g), lab),
    colors = setNames(c(u$col, REST_COLOR), c(lab, REST_LABEL)),
    title = paste0("GC by codon position: ", lab, " (size-matched)"),
    subtitle = paste0(
      "Positive: top ", N_MATCH, " by TE LFC among ", u$kd, " ", cond,
      "-only genes present in the GC matrix\n",
      "(", u$kd, " TE LFC > ", SI3D_CUT, " & padj < 0.05 in ", u$con,
      ", minus genes also promoted in the other condition; ", length(ids),
      " available). LFC floor ", round(min(sel$te_lfc), 2), "\n",
      "All four panels in this family are matched at n = ", N_MATCH,
      ", so curve smoothness is comparable; effect size is NOT\n",
      "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class"),
    file = paste0("89_gc_position_density_nooverlap_top", N_MATCH, "_", u$key, "_bg.pdf"))

  s <- gc231 %>% dplyr::filter(gene_id_clean %in% sel$g)
  tibble::tibble(set = lab, n_available = length(ids), n_plotted = nrow(s),
                 lfc_floor = round(min(sel$te_lfc), 2),
                 cds_gc = round(median(s$cds_gc), 1), gc1 = round(median(s$gc1), 1),
                 gc2 = round(median(s$gc2), 1), gc3 = round(median(s$gc3), 1))
})

cat("\nOverlap-removed sets, size-matched to N =", N_MATCH,
    " (background GC3 =", round(median(gc231$gc3), 1), "):\n")
print(as.data.frame(matched_summary), row.names = FALSE)
```

## Overlap removed, matched upward to the largest eIF3d set

The previous family matched *down* to N = 95, which forced eIF3e normoxia-only to be its
extreme tip (LFC floor 1.60) while eIF3d normoxia-only had to use its whole set (floor 0.50).
This family matches *up* instead, to the in-matrix size of the eIF3d hypoxia-only set.

**Only two of the four sets can reach it.** eIF3d normoxia-only has 95 genes in the matrix
and eIF3e hypoxia-only has 153, so for those two "top N" is their entire set and the figure
would be identical to the overlap-removed one already produced above. They are skipped here
rather than written again under a misleading `topN` filename; their rows still appear in the
summary with `reached = FALSE` so the comparison stays honest.

```{r no-overlap-top-matched-up, fig.width = 7.5, fig.height = 10}
N_MATCH_UP <- length(no_overlap_ids$`3d_hyp1hr`)
cat("Matching upward to the eIF3d hypoxia-only in-matrix size: N =", N_MATCH_UP, "\n")
print(vapply(no_overlap_ids, length, integer(1)))

matched_up_summary <- purrr::map_dfr(unfiltered_sets, function(u) {
  ids     <- no_overlap_ids[[u$key]]
  reached <- length(ids) >= N_MATCH_UP
  sel <- u$tbl %>%
    dplyr::mutate(g = sub("\\..*", "", gene_id)) %>%
    dplyr::filter(g %in% ids) %>%
    dplyr::arrange(dplyr::desc(te_lfc)) %>%
    dplyr::distinct(g, .keep_all = TRUE) %>%
    head(min(N_MATCH_UP, length(ids)))

  cond <- if (grepl("hypoxia", u$con)) "hypoxia" else "normoxia"
  lab  <- paste0(u$kd, " ", cond, "-only, top ", N_MATCH_UP)

  if (reached) {
    stopifnot("selection is not exactly N_MATCH_UP genes" = nrow(sel) == N_MATCH_UP)
    gc_density_figure(
      gc231,
      groups = setNames(list(sel$g), lab),
      colors = setNames(c(u$col, REST_COLOR), c(lab, REST_LABEL)),
      title = paste0("GC by codon position: ", lab, " (matched up)"),
      subtitle = paste0(
        "Positive: top ", N_MATCH_UP, " by TE LFC among ", u$kd, " ", cond,
        "-only genes present in the GC matrix\n",
        "(", u$kd, " TE LFC > ", SI3D_CUT, " & padj < 0.05 in ", u$con,
        ", minus genes also promoted in the other condition; ", length(ids),
        " available). LFC floor ", round(min(sel$te_lfc), 2), "\n",
        "Matched to the eIF3d hypoxia-only set size. Only 2 of the 4 sets are large enough to reach it\n",
        "Grey: MDA-MB-231 transcriptome excluding the positive set. No negative control class"),
      file = paste0("89_gc_position_density_nooverlap_top", N_MATCH_UP, "_", u$key, "_bg.pdf"))
  } else {
    cat("SKIPPED figure for", lab, "- only", length(ids),
        "genes available, so top", N_MATCH_UP, "is its entire set",
        "(see the overlap-removed figure for this set)\n")
  }

  s <- gc231 %>% dplyr::filter(gene_id_clean %in% sel$g)
  tibble::tibble(set = paste0(u$kd, " ", cond, "-only"), reached = reached,
                 n_available = length(ids), n_plotted = nrow(s),
                 lfc_floor = round(min(sel$te_lfc), 2),
                 cds_gc = round(median(s$cds_gc), 1), gc1 = round(median(s$gc1), 1),
                 gc2 = round(median(s$gc2), 1), gc3 = round(median(s$gc3), 1))
})

cat("\nMatched upward to N =", N_MATCH_UP,
    " (background GC3 =", round(median(gc231$gc3), 1),
    "); rows with reached = FALSE are the full set, not a top-N cut:\n")
print(as.data.frame(matched_up_summary), row.names = FALSE)
```

## eIF3d normoxia, top 1000 by LFC alone (no padj filter)

The eIF3d normoxia 1+4hr contrast has only 393 genes at `padj < 0.05` in total, so the
overlap-removed set above is stuck at n = 95 and its curves are correspondingly noisy. To
reach n ~ 1000 the **significance filter is dropped entirely and genes are ranked by
`te_lfc` alone**, as requested. Two versions:

1. **including** genes eIF3d also promotes in hypoxia 1hr
2. **excluding** them (`te_lfc > 0.5 & padj < 0.05` in hypoxia 1hr, the standard definition)

Ranking is among genes present in the GC matrix, so each plots exactly `N_NOR`.

**What dropping padj costs.** These are effect-size-ranked sets, not significance-filtered
ones: in the including version 39% of the 1000 clear `padj < 0.05`, in the excluding version
only 10%. The excluding version also has to go below the project 0.5 LFC convention to find
1000 genes (floor 0.371, and 55% of its members sit under 0.5). Both figures state their own
floor and significance fraction so this is never read off the curve alone.

```{r nor-top1000-lfc-only, fig.width = 7.5, fig.height = 10}
N_NOR <- 1000
hyp_promoted_3d <- promoted_at(si3d_hyp_tbl, SI3D_CUT)   # eIF3d, hypoxia 1hr, conventional

nor_pool <- si3d_nor_tbl %>%
  dplyr::mutate(g = sub("\\..*", "", gene_id)) %>%
  dplyr::filter(!is.na(te_lfc), g %in% gc231$gene_id_clean) %>%
  dplyr::arrange(dplyr::desc(te_lfc)) %>%
  dplyr::distinct(g, .keep_all = TRUE)

nor1000_summary <- purrr::map_dfr(
  list(list(key = "incl3dhyp", excl = FALSE,
            lab = paste0("eIF3d normoxia, top ", N_NOR, " (incl. 3d-hypoxia)"),
            desc = "including genes eIF3d also promotes in hypoxia 1hr"),
       list(key = "excl3dhyp", excl = TRUE,
            lab = paste0("eIF3d normoxia, top ", N_NOR, " (excl. 3d-hypoxia)"),
            desc = "excluding genes eIF3d also promotes in hypoxia 1hr (TE LFC > 0.5 & padj < 0.05 there)")),
  function(v) {
    pool <- if (v$excl) nor_pool %>% dplyr::filter(!g %in% hyp_promoted_3d) else nor_pool
    sel  <- head(pool, N_NOR)
    stopifnot("could not reach N_NOR genes"         = nrow(sel) == N_NOR,
              "selection is not ranked by te_lfc"   = !is.unsorted(rev(sel$te_lfc)),
              "a selected gene is not in the matrix" = all(sel$g %in% gc231$gene_id_clean))

    pct_sig <- round(100 * mean(sel$te_padj < 0.05, na.rm = TRUE))
    pct_conv <- round(100 * mean(sel$te_lfc > SI3D_CUT))
    gc_density_figure(
      gc231,
      groups = setNames(list(sel$g), v$lab),
      colors = setNames(c("#E69F00", REST_COLOR), c(v$lab, REST_LABEL)),
      title = paste0("GC by codon position: ", v$lab),
      subtitle = paste0(
        "Positive: top ", N_NOR, " by eIF3d TE LFC in normoxia 1+4hr (time-regressed), ",
        v$desc, "\n",
        "NO padj filter - ranked by LFC alone. LFC floor ", round(min(sel$te_lfc), 3),
        "; ", pct_sig, "% of these clear padj < 0.05, ", pct_conv, "% clear the 0.5 LFC convention\n",
        "Pool available ", nrow(pool), ". Grey: MDA-MB-231 transcriptome excluding the positive set. ",
        "No negative control class"),
      file = paste0("89_gc_position_density_3d_normoxia_top", N_NOR, "_lfconly_", v$key, "_bg.pdf"))

    s <- gc231 %>% dplyr::filter(gene_id_clean %in% sel$g)
    tibble::tibble(set = v$lab, pool = nrow(pool), n = nrow(s),
                   lfc_floor = round(min(sel$te_lfc), 3),
                   lfc_median = round(median(sel$te_lfc), 2),
                   pct_padj05 = pct_sig, pct_lfc_gt_0.5 = pct_conv,
                   cds_gc = round(median(s$cds_gc), 1), gc1 = round(median(s$gc1), 1),
                   gc2 = round(median(s$gc2), 1), gc3 = round(median(s$gc3), 1))
  })

cat("\neIF3d normoxia top-", N_NOR, " by LFC alone (background GC3 = ",
    round(median(gc231$gc3), 1), "):\n", sep = "")
print(as.data.frame(nor1000_summary), row.names = FALSE)
```

# 8. 4E-resistant vs 4E-sensitive

Different dataset and cell line (Roiuk et al. HeLa, 4E-BP inhibition), so this figure is
drawn on the HeLa transcript matrix from notebook 78. Both classes are tails of the same
variable rather than a positive set against unaffected controls, which is why the grey
curve sits between them rather than on top of one.

```{r fig-4e, fig.width = 7.5, fig.height = 10}
sub_4e <- paste0(
  "Resistant: top ", length(res_4e), " genes by TE log2FC on 4E-BP inhibition (>= ",
  round(res_cut, 2), ", padj < 0.05)\n",
  "Sensitive: bottom ", length(sen_4e), " genes (<= ", round(sen_cut, 2), ", padj < 0.05)\n",
  "Grey: HeLa transcriptome excluding both sets. Positive log2FC = TE held up = resistant")

gc_density_figure(
  gc4e,
  groups = list("4E-resistant" = res_4e,
                "4E-sensitive" = sen_4e),
  colors = c("4E-resistant"          = "#D55E00",   # Okabe-Ito vermillion
             "4E-sensitive"          = "#0072B2",   # Okabe-Ito blue
             "Rest of transcriptome" = REST_COLOR),
  title = "GC by codon position: 4E-resistant vs 4E-sensitive mRNAs (HeLa)",
  subtitle = sub_4e,
  file = "89_gc_position_density_4e.pdf")
```

# 9. Session info

```{r session-info}
sessionInfo()
```
