---
title: "Negative-subsample stability of the eIF3e target model"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  neg_tag: "exprbg"
  n_seeds: 20
  top_n: 25
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Purpose

Notebook 07 keeps every positive gene and draws a size-matched negative class from a much
larger pool: 399 of 9,183 for the eIF3e-specific model. One run therefore reports the
performance and the feature ranking of **one arbitrary draw**. This notebook reads the 20
replicate runs produced by `90_replicate_neg_subsample_driver.R` and asks how much of
either is a property of the model rather than of the draw.

Only `neg_seed` varies across the replicates. The split, the forest and the CV are pinned
at seed 9 inside notebook 07, so the spread reported here is attributable to negative-class
composition alone.

**Pool:** `si3e_expressed_background_pool.csv` (notebook 02c) - expressed genes minus all
3e-promoted genes, not screened on effect size. Runs built on the older *screened*
`si3e_negative_controls_padj0.05.csv` pool answer a different question and are not
comparable; they are excluded here by the `neg_tag` filter.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params_default}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(neg_tag = "exprbg", n_seeds = 20, top_n = 25, save_plots = TRUE)
}
SAVE_PLOTS <- params$save_plots
SI3E_COLOR <- "#56B4E9"   # CLAUDE.md: si3e / eIF3e is always Okabe-Ito light blue
```

# 1. Locate the replicate runs

```{r locate}
base_suffix <- paste0("eif33e_promotes_hypoxia_1hr_not3d_lfc0.5_dhx29_kd_feature",
                      "_clipaggregate_no_boruta_no4e_noOwnTE_", params$neg_tag)

# Seed 9 is the un-suffixed run (notebook 07 appends _negseedN only when != 9), so the
# file list has to cover both forms.
suffix_for <- function(seed) {
  paste0(base_suffix, if (seed != 9L) paste0("_negseed", seed) else "")
}
seeds <- seq_len(params$n_seeds)

read_run <- function(prefix, seed) {
  p <- here("output", "predictive_modeling",
            paste0(prefix, "_", suffix_for(seed), ".csv"))
  if (!file.exists(p)) return(NULL)
  read_csv(p, show_col_types = FALSE) %>% mutate(seed = seed)
}

perf <- map_dfr(seeds, ~ read_run("rf_classification_performance", .x))
imp  <- map_dfr(seeds, ~ read_run("rf_classification_importance", .x))
sets <- map_dfr(seeds, ~ read_run("rf_model_gene_sets", .x))

found <- sort(unique(perf$seed))
cat("Replicates found:", length(found), "of", params$n_seeds, "\n")
stopifnot("no replicate runs found - run notebook 90 first" = length(found) > 0)
if (length(found) < params$n_seeds) {
  warning("missing seeds: ", paste(setdiff(seeds, found), collapse = ", "))
}
```

# 2. The replicates must differ in the right way

A sweep that varied nothing would produce a reassuringly tight spread for the wrong
reason. These two checks establish that the positive class is constant and the negative
class genuinely moved before any spread is interpreted.

```{r validate}
stopifnot("model gene sets are missing the columns this check needs" =
  all(c("gene_id_clean", "eif3d_target") %in% names(sets)))

pos_by_seed <- sets %>% filter(eif3d_target == "promotes") %>%
  group_by(seed) %>% summarise(genes = list(sort(gene_id_clean)), .groups = "drop")
neg_by_seed <- sets %>% filter(eif3d_target != "promotes") %>%
  group_by(seed) %>% summarise(genes = list(sort(gene_id_clean)), .groups = "drop")

stopifnot("the positive set is not identical across replicates" =
  length(unique(map_chr(pos_by_seed$genes, ~ paste(.x, collapse = ",")))) == 1)
cat("Positive set identical across all", nrow(pos_by_seed), "replicates:",
    length(pos_by_seed$genes[[1]]), "genes\n")

neg_keys <- map_chr(neg_by_seed$genes, ~ paste(.x, collapse = ","))
stopifnot("two replicates drew the identical negative set - the draw is degenerate" =
  !any(duplicated(neg_keys)))

# How much do the draws actually overlap? Independent 399-of-9,183 draws should share
# about 399 * 399 / 9183 ~ 17 genes, i.e. a Jaccard near 0.02. Much higher would mean the
# pool is far smaller than intended.
jac <- function(a, b) length(intersect(a, b)) / length(union(a, b))
pairs <- combn(seq_len(nrow(neg_by_seed)), 2)
neg_jac <- map_dbl(seq_len(ncol(pairs)),
                   ~ jac(neg_by_seed$genes[[pairs[1, .x]]], neg_by_seed$genes[[pairs[2, .x]]]))
n_neg <- length(neg_by_seed$genes[[1]])
pool_n <- 9183
cat(sprintf("Negative-draw overlap: median Jaccard %.3f (expected ~%.3f for %d of %d)\n",
            median(neg_jac), n_neg / (2 * pool_n - n_neg), n_neg, pool_n))
cat("Distinct genes used as negatives across all replicates:",
    n_distinct(unlist(neg_by_seed$genes)), "\n")
```

# 3. Performance across replicates

```{r performance}
auc <- perf %>% filter(Metric == "AUC") %>%
  transmute(seed, test_auc = Test_Set, cv_mean = CV_Mean, cv_sd = CV_SD)

acc <- perf %>% filter(Metric == "Balanced_Accuracy") %>%
  transmute(seed, bal_acc = Test_Set)

summ <- auc %>% left_join(acc, by = "seed")

cat(sprintf("Test AUC across %d replicates: median %.3f, mean %.3f, sd %.3f, range %.3f-%.3f\n",
            nrow(summ), median(summ$test_auc), mean(summ$test_auc), sd(summ$test_auc),
            min(summ$test_auc), max(summ$test_auc)))
cat(sprintf("Seed 9 (the single run you would otherwise report): %.3f  [percentile %.0f%%]\n",
            summ$test_auc[summ$seed == 9],
            100 * mean(summ$test_auc <= summ$test_auc[summ$seed == 9])))
cat(sprintf("Balanced accuracy: median %.3f, range %.3f-%.3f\n",
            median(summ$bal_acc), min(summ$bal_acc), max(summ$bal_acc)))

print(as.data.frame(summ %>% arrange(seed) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))), row.names = FALSE)
```

```{r plot-auc, fig.width = 5, fig.height = 4.5}
p_auc <- summ %>%
  dplyr::select(seed, `Test AUC` = test_auc, `CV mean AUC` = cv_mean) %>%
  pivot_longer(-seed, names_to = "metric", values_to = "auc") %>%
  ggplot(aes(x = metric, y = auc)) +
  geom_boxplot(width = 0.45, outlier.shape = NA, fill = NA, color = "grey40") +
  geom_jitter(width = 0.12, height = 0, size = 2, alpha = 0.6, color = SI3E_COLOR) +
  geom_point(data = ~ dplyr::filter(.x, seed == 9), size = 3, shape = 21,
             fill = SI3E_COLOR, color = "black", stroke = 0.9) +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey50") +
  labs(title = "eIF3e-specific model: AUC across negative subsamples",
       subtitle = paste0("399 positives held fixed; 399 negatives redrawn from a ",
                         format(pool_n, big.mark = ","), "-gene expressed background\n",
                         "Only neg_seed varies (split/forest/CV pinned at 9). ",
                         "Outlined point = seed 9. n = ", nrow(summ), " replicates"),
       x = NULL, y = "AUC") +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 8.5, lineheight = 1.15))
print(p_auc)
if (SAVE_PLOTS) ggsave(here("plots", paste0("91_auc_across_neg_subsamples_", params$neg_tag, ".pdf")),
                       p_auc, width = 5, height = 4.5)
```

# 4. Feature importance stability

Impurity importance is reported per replicate, so a feature's rank can be summarised the
same way as the AUC. A feature that matters to the *model* holds its rank across draws; a
feature that matters to one *draw* does not.

```{r importance-stability}
stopifnot("importance table is missing expected columns" =
  all(c("feature", "importance") %in% names(imp)))

imp_ranked <- imp %>%
  group_by(seed) %>%
  mutate(rank = rank(-importance, ties.method = "min")) %>%
  ungroup()

feat_summary <- imp_ranked %>%
  group_by(feature) %>%
  summarise(mean_rank = mean(rank), sd_rank = sd(rank),
            best = min(rank), worst = max(rank),
            mean_imp = mean(importance), sd_imp = sd(importance),
            n_top10 = sum(rank <= 10), .groups = "drop") %>%
  arrange(mean_rank)

# ALL features, not just the top of the list: a feature that ranks mid-table on every
# replicate is a different (and more interpretable) situation than one that swings from
# 5th to 40th, and neither is visible from a truncated table.
feat_table <- feat_summary %>%
  transmute(feature,
            mean_rank = round(mean_rank, 1), sd_rank = round(sd_rank, 1),
            best, worst, span = worst - best,
            mean_imp = signif(mean_imp, 3), cv_imp = round(sd_imp / mean_imp, 2),
            times_in_top10 = n_top10,
            always_top20 = worst <= 20)

cat("All", nrow(feat_table), "features by mean rank across", length(found), "replicates:\n")
print(as.data.frame(feat_table), row.names = FALSE)

# `span` is the honest summary: it is the full range a feature's rank covered, so a large
# span means the feature's standing is a property of the draw and not of the model.
out_csv <- here("output", "predictive_modeling",
                paste0("91_feature_rank_stability_", params$neg_tag, ".csv"))
write_csv(feat_table, out_csv)
cat("\nWrote full table to", out_csv, "\n")

cat("\nTop", params$top_n, "shown in the figure below.\n")
```

```{r plot-importance, fig.width = 7.5, fig.height = 7}
top_feats <- feat_summary %>% head(params$top_n) %>% pull(feature)

p_imp <- imp_ranked %>%
  filter(feature %in% top_feats) %>%
  mutate(feature = factor(feature, levels = rev(top_feats))) %>%
  ggplot(aes(x = rank, y = feature)) +
  geom_point(alpha = 0.45, size = 1.8, color = SI3E_COLOR) +
  stat_summary(fun = median, geom = "point", shape = 124, size = 5, color = "black") +
  labs(title = "Feature rank across negative subsamples",
       subtitle = paste0("Each point is one replicate; black bar is the median rank. ",
                         "n = ", length(found), " replicates\n",
                         "Ordered by mean rank; a wide spread means the feature's ",
                         "standing depends on which negatives were drawn"),
       x = "Importance rank (1 = most important)", y = NULL) +
  theme_classic(base_size = 12) +
  theme(plot.subtitle = element_text(size = 8.5, lineheight = 1.15))
print(p_imp)
if (SAVE_PLOTS) ggsave(here("plots", paste0("91_importance_rank_stability_", params$neg_tag, ".pdf")),
                       p_imp, width = 7.5, height = 7)
```

```{r topn-jaccard}
# How reproducible is "the top 20 features" itself? Jaccard between every pair of
# replicates' top-20 lists.
top20 <- imp_ranked %>% filter(rank <= 20) %>%
  group_by(seed) %>% summarise(f = list(feature), .groups = "drop")
tp <- combn(seq_len(nrow(top20)), 2)
top_jac <- map_dbl(seq_len(ncol(tp)), ~ jac(top20$f[[tp[1, .x]]], top20$f[[tp[2, .x]]]))

cat(sprintf("Top-20 agreement between replicate pairs: median Jaccard %.2f (range %.2f-%.2f, %d pairs)\n",
            median(top_jac), min(top_jac), max(top_jac), length(top_jac)))
cat("Features in the top 20 of EVERY replicate:",
    sum(feat_summary$worst <= 20), "\n")
cat("Features in the top 20 of at least one but not all:",
    sum(feat_summary$best <= 20 & feat_summary$worst > 20), "\n")
```

# 5. Top features per replicate, with direction

Notebook 07 writes a `rf_classification_directionality_*.csv` per run, so direction is
already available for every replicate and does not need recomputing. `direction` is the
sign of the feature's correlation with the target: **"Higher in promotes"** means eIF3e
targets carry more of it than the background draw.

Two separate questions, answered below in order:

1. **How often** does a feature reach the top of a run? (`n_top`, out of the replicates)
2. **Does it point the same way** when it gets there? A feature whose direction flips
   between draws is not describing the biology, however often it ranks highly.

```{r load-directionality}
dirs <- map_dfr(seeds, ~ read_run("rf_classification_directionality", .x))
stopifnot("directionality files not found - re-run notebook 90" = nrow(dirs) > 0)
stopifnot("directionality table is missing expected columns" =
  all(c("feature", "importance", "direction", "correlation", "wilcox_p_adj") %in% names(dirs)))

TOP_K <- 10   # "top of a run" = top TOP_K features by impurity importance in that run

dirs <- dirs %>%
  group_by(seed) %>%
  mutate(rank = rank(-importance, ties.method = "min")) %>%
  ungroup() %>%
  mutate(dir_sign = case_when(direction == "Higher in promotes" ~ "+",
                              direction == "Lower in promotes"  ~ "-",
                              TRUE ~ "?"))
cat("Directionality loaded for", n_distinct(dirs$seed), "replicates x",
    n_distinct(dirs$feature), "features\n")
```

## 5.1 Per-replicate top `r TOP_K`, side by side

Rows are every feature that reached the top `r TOP_K` of at least one replicate; columns
are the replicates. A cell shows that run's rank and direction (`3+` = ranked 3rd, higher
in targets); `.` means the feature missed the top `r TOP_K` in that run.

```{r per-run-table}
in_top <- dirs %>% filter(rank <= TOP_K)
feat_order <- in_top %>% dplyr::count(feature, name = "n_top") %>% arrange(desc(n_top))

wide <- dirs %>%
  filter(feature %in% feat_order$feature) %>%
  mutate(cell = ifelse(rank <= TOP_K, paste0(rank, dir_sign), ".")) %>%
  dplyr::select(feature, seed, cell) %>%
  pivot_wider(names_from = seed, values_from = cell, names_prefix = "s") %>%
  left_join(feat_order, by = "feature") %>%
  arrange(desc(n_top)) %>%
  relocate(feature, n_top)

print(as.data.frame(wide), row.names = FALSE)
```

## 5.2 How common, and how consistent

```{r direction-consistency}
dir_summary <- dirs %>%
  group_by(feature) %>%
  summarise(
    n_top          = sum(rank <= TOP_K),
    mean_rank      = round(mean(rank), 1),
    # Consensus direction and how strongly the replicates agree on it. Computed over ALL
    # replicates, not just the ones where the feature ranked highly: a feature can be
    # consistently directional while only occasionally being important.
    n_higher       = sum(dir_sign == "+"),
    n_lower        = sum(dir_sign == "-"),
    consensus      = ifelse(n_higher >= n_lower, "Higher in promotes", "Lower in promotes"),
    pct_agree      = round(100 * max(n_higher, n_lower) / (n_higher + n_lower)),
    median_cor     = round(median(correlation, na.rm = TRUE), 3),
    n_sig          = sum(wilcox_p_adj < 0.05, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(desc(n_top), mean_rank)

cat("All", nrow(dir_summary), "features: how often each reaches the top", TOP_K,
    "and whether its direction holds\n")
cat("(pct_agree = % of the", length(found),
    "replicates agreeing on the consensus direction; n_sig = replicates with BH-adj Wilcoxon p < 0.05)\n\n")
print(as.data.frame(dir_summary), row.names = FALSE)

out_dir_csv <- here("output", "predictive_modeling",
                    paste0("91_feature_direction_consistency_", params$neg_tag, ".csv"))
write_csv(dir_summary, out_dir_csv)
cat("\nWrote", out_dir_csv, "\n")
```

```{r direction-flips}
# A feature that reaches the top of some runs but points in opposite directions between
# them is reporting the draw, not the biology. Reported rather than asserted: with a
# correlation near zero a sign flip is expected and harmless, which is why the magnitude
# of median_cor is printed alongside.
flips <- dir_summary %>% filter(pct_agree < 100)
if (nrow(flips) == 0) {
  cat("No feature changes direction across replicates.\n")
} else {
  cat(nrow(flips), "features change direction between replicates:\n")
  print(as.data.frame(flips %>%
    dplyr::select(feature, n_top, consensus, pct_agree, median_cor, n_sig)), row.names = FALSE)
  cat("\nA flip matters only where |median_cor| is meaningfully above 0;",
      "near zero the sign is noise.\n")
}
```

```{r plot-direction, fig.width = 9, fig.height = 7}
# Direction colours: Okabe-Ito dark blue / reddish purple. Orange and sky blue are reserved
# for eIF3d and eIF3e (CLAUDE.md) and must not be reused for up/down here.
dir_colors <- c("Higher in promotes" = "#0072B2", "Lower in promotes" = "#CC79A7")

tile_feats <- feat_order$feature
p_dir <- dirs %>%
  filter(feature %in% tile_feats) %>%
  mutate(feature = factor(feature, levels = rev(tile_feats)),
         shown = ifelse(rank <= TOP_K, as.character(rank), "")) %>%
  ggplot(aes(x = factor(seed), y = feature)) +
  geom_tile(aes(fill = direction, alpha = rank <= TOP_K), color = "white", linewidth = 0.4) +
  geom_text(aes(label = shown), size = 2.7, color = "white") +
  scale_fill_manual(values = dir_colors, name = NULL, na.value = "grey85") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.18), guide = "none") +
  labs(title = paste0("Top-", TOP_K, " features per replicate, with direction"),
       subtitle = paste0("One column per negative subsample; number = that run's importance rank. ",
                         "Faded = outside the top ", TOP_K, ".\n",
                         "Colour is the sign of the feature's correlation with the target. ",
                         "n = ", length(found), " replicates, ", length(tile_feats),
                         " features reaching the top ", TOP_K, " at least once"),
       x = "neg_seed", y = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 8.5, lineheight = 1.15),
        axis.text.x = element_text(size = 7))
print(p_dir)
if (SAVE_PLOTS) ggsave(here("plots", paste0("91_top_features_direction_", params$neg_tag, ".pdf")),
                       p_dir, width = 9, height = 7)
```

# 6. Summary

```{r summary}
cat("Replicates:                ", length(found), "\n")
cat("Positive class:            ", length(pos_by_seed$genes[[1]]), "genes, identical throughout\n")
cat("Negative class:            ", n_neg, "genes redrawn each replicate from", pool_n, "\n")
cat(sprintf("Test AUC:                   %.3f  (median, sd %.3f, range %.3f-%.3f)\n",
            median(summ$test_auc), sd(summ$test_auc),
            min(summ$test_auc), max(summ$test_auc)))
cat(sprintf("Top-20 feature agreement:   %.2f  (median pairwise Jaccard)\n", median(top_jac)))
cat("Features top-20 in every replicate:", sum(feat_summary$worst <= 20), "\n")
cat("Full per-feature table:      ", basename(out_csv), "\n")
cat("Direction consistency table: ", basename(out_dir_csv), "\n")
cat("Features reaching the top", TOP_K, "in every replicate:",
    sum(dir_summary$n_top == length(found)), "\n")
cat("Features changing direction between replicates:", sum(dir_summary$pct_agree < 100), "\n")
```
