---
title: "5'UTR / CDS / 3'UTR length distributions, 4E-resistant vs 4E-sensitive"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  feature_matrix: "feature_matrix_teleman_4e.rds"
  pos_csv: "teleman_4e_resistant_TE_1hr_lfc0.5.csv"
  neg_csv: "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"
  pos_fasta: "teleman_4e_resistant_transcripts.fa"
  neg_fasta: "teleman_4e_sensitive_transcripts.fa"
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Length distributions of the three transcript regions (5'UTR, CDS, 3'UTR) for the
4E-resistant and 4E-sensitive sets defined in notebook 76 -- the same two groups whose
sequences were exported to `output/fasta/`. Companion to notebook 79, which does the
same for GC content.

Lengths come from the feature matrix (`utr5_length` / `cds_length` / `utr3_length`,
one row per selected transcript) rather than from the FASTA files, which hold whole
transcripts and so cannot be split by region. The chunk below asserts that the
transcripts being plotted are exactly the ones in those FASTA files.

```{r libraries}
library(tidyverse)
library(here)
```

```{r params}
# Defaults for interactive use; the YAML params win when knitting.
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(
    feature_matrix = "feature_matrix_teleman_4e.rds",
    pos_csv        = "teleman_4e_resistant_TE_1hr_lfc0.5.csv",
    neg_csv        = "teleman_4e_sensitive_TE_1hr_lfc0.5.csv",
    pos_fasta      = "teleman_4e_resistant_transcripts.fa",
    neg_fasta      = "teleman_4e_sensitive_transcripts.fa",
    save_plots     = TRUE
  )
}
FEATURE_MATRIX <- params$feature_matrix
POS_CSV        <- params$pos_csv
NEG_CSV        <- params$neg_csv
POS_FASTA      <- params$pos_fasta
NEG_FASTA      <- params$neg_fasta
SAVE_PLOTS     <- params$save_plots

# eIF3d orange and eIF3e blue are reserved for those two factors (CLAUDE.md), so this
# figure uses the same two Okabe-Ito hues as notebooks 76 and 79.
class_colors <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2")

region_labels <- c(
  utr5_length = "5'UTR",
  cds_length  = "CDS",
  utr3_length = "3'UTR"
)
```

# Data

```{r load}
fm  <- readRDS(here("output", "predictive_modeling", FEATURE_MATRIX))
pos <- read_csv(here("output", "genesets", POS_CSV), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", NEG_CSV), show_col_types = FALSE)

stopifnot(all(names(region_labels) %in% colnames(fm)))

d <- fm %>%
  dplyr::mutate(class = dplyr::case_when(
    gene_id_clean %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id_clean %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE                                ~ NA_character_)) %>%
  dplyr::filter(!is.na(class)) %>%
  dplyr::mutate(class = factor(class, levels = names(class_colors)))

print(dplyr::count(d, class) %>% as.data.frame(), row.names = FALSE)
stopifnot("classes overlap" =
  length(intersect(pos$ensembl_gene, neg$ensembl_gene)) == 0)
```

The transcripts plotted here are the ones whose sequences were exported to FASTA.

```{r fasta-check}
fasta_ids <- function(path) {
  sub("^>", "", grep("^>", readLines(path), value = TRUE))
}
fa_pos <- fasta_ids(here("output", "fasta", POS_FASTA))
fa_neg <- fasta_ids(here("output", "fasta", NEG_FASTA))

stopifnot(
  "resistant transcripts differ from exported FASTA" =
    setequal(fa_pos, d$transcript_id_clean[d$class == "4E-resistant"]),
  "sensitive transcripts differ from exported FASTA" =
    setequal(fa_neg, d$transcript_id_clean[d$class == "4E-sensitive"]))
cat("FASTA transcript IDs match the plotted sets:",
    length(fa_pos), "resistant /", length(fa_neg), "sensitive\n")
```

```{r range-checks}
# Lengths are nucleotide counts: non-negative, and the three regions must fit inside
# the transcript.
lens <- d %>% dplyr::select(dplyr::all_of(names(region_labels)))
stopifnot(
  "negative region length" = all(lens >= 0, na.rm = TRUE),
  "region lengths exceed transcript length" =
    all(rowSums(lens) <= d$transcript_length, na.rm = TRUE))

# A complete CDS is a whole number of codons. A handful of GENCODE entries are tagged
# cds_start_NF / cds_end_NF and are not, so this is a tolerance rather than a hard
# assert -- but a large count would mean the CDS ranges are being read wrong.
n_partial <- sum(d$cds_length %% 3 != 0, na.rm = TRUE)
cat("CDS length not a multiple of 3 (incomplete GENCODE CDS):", n_partial,
    "of", sum(!is.na(d$cds_length)), "\n")
stopifnot("too many CDS lengths are not a multiple of 3" =
  n_partial / sum(!is.na(d$cds_length)) < 0.01)

# Biological spot-check: median human 5'UTR is short (~100-300 nt) and the 3'UTR is
# the longest of the three regions on a typical mRNA.
med_all <- vapply(lens, median, numeric(1), na.rm = TRUE)
print(med_all)
stopifnot(
  "median 5'UTR outside 50-400 nt" = dplyr::between(med_all[["utr5_length"]], 50, 400),
  "median 3'UTR not longest region" =
    med_all[["utr3_length"]] > med_all[["utr5_length"]])

# Zero-length regions cannot be shown on a log axis; report how many are dropped.
zero_counts <- lens %>%
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "region", values_to = "length") %>%
  dplyr::filter(length == 0 | is.na(length)) %>%
  dplyr::count(region, name = "n_zero_or_na")
print(as.data.frame(zero_counts), row.names = FALSE)
```

```{r long}
d_long <- d %>%
  dplyr::select(transcript_id_clean, symbol, class,
                dplyr::all_of(names(region_labels))) %>%
  tidyr::pivot_longer(dplyr::all_of(names(region_labels)),
                      names_to = "region", values_to = "length") %>%
  dplyr::filter(!is.na(length), length > 0) %>%
  dplyr::mutate(region = factor(region_labels[region],
                                levels = unname(region_labels)))
```

# Reference: all expressed mRNAs in the Teleman data

The 4E-resistant and 4E-sensitive sets are both tails of the same contrast, so a shift
in one of them is only interpretable against the bulk. "Expressed" here is every gene
DESeq2 could **test** in the Teleman TE contrast: `independentFiltering = FALSE` was set
upstream, so all 11,306 tested genes carry a `padj`, and 11,286 of them have region
annotations in the feature matrix. That whole matrix is the reference -- no filtering on
effect size, so it is a background rather than a third selected group.

Note the two classes are themselves part of this background (2,390 of 11,286, ~21%), so
the reference line is pulled slightly toward them; it is not an independent set.

```{r expressed-background}
expressed_long <- fm %>%
  dplyr::select(transcript_id_clean, dplyr::all_of(names(region_labels))) %>%
  tidyr::pivot_longer(dplyr::all_of(names(region_labels)),
                      names_to = "region", values_to = "length") %>%
  dplyr::filter(!is.na(length), length > 0) %>%
  dplyr::mutate(region = factor(region_labels[region],
                                levels = unname(region_labels)))

med_expressed <- expressed_long %>%
  dplyr::group_by(region) %>%
  dplyr::summarise(n_expressed = dplyr::n(),
                   m_all = median(length), .groups = "drop") %>%
  dplyr::mutate(
    label = paste0("grey dashed = all expressed mRNAs\n(median = ",
                   round(m_all), " nt, n = ", n_expressed, ")"),
    label_short = paste0(scales::comma(round(m_all)), " nt\n(n = ",
                         scales::comma(n_expressed), ")"))
print(as.data.frame(med_expressed), row.names = FALSE)

stopifnot("expressed background is smaller than the two classes combined" =
  all(med_expressed$n_expressed > nrow(d)))
```

# Summary statistics

```{r summary}
len_summary <- d_long %>%
  dplyr::group_by(region, class) %>%
  dplyr::summarise(n = dplyr::n(),
                   median = median(length),
                   mean   = mean(length),
                   q25    = quantile(length, 0.25),
                   q75    = quantile(length, 0.75),
                   .groups = "drop")
print(as.data.frame(len_summary), row.names = FALSE)

len_tests <- d_long %>%
  dplyr::group_by(region) %>%
  dplyr::summarise(
    n_resistant = sum(class == "4E-resistant"),
    n_sensitive = sum(class == "4E-sensitive"),
    median_resistant = median(length[class == "4E-resistant"]),
    median_sensitive = median(length[class == "4E-sensitive"]),
    wilcox_p = wilcox.test(length[class == "4E-resistant"],
                           length[class == "4E-sensitive"])$p.value,
    .groups = "drop")
print(as.data.frame(len_tests), row.names = FALSE)

if (SAVE_PLOTS) {
  write_csv(len_tests, here("output", "92_region_length_stats_4e.csv"))
}
```

Every density panel carries the same three medians: the expressed background and the two
classes. Building them once here keeps the single-region and faceted figures consistent.

```{r panel-medians}
REF_GROUP    <- "all expressed"
median_colors <- c(setNames("grey50", REF_GROUP), class_colors)

panel_medians <- dplyr::bind_rows(
  med_expressed %>%
    dplyr::transmute(region, group = REF_GROUP, m = m_all, n = n_expressed),
  d_long %>%
    dplyr::group_by(region, class) %>%
    dplyr::summarise(m = median(length), n = dplyr::n(), .groups = "drop") %>%
    dplyr::transmute(region, group = as.character(class), m, n)) %>%
  dplyr::mutate(group = factor(group, levels = names(median_colors)),
                label = paste0(group, " median = ",
                               scales::comma(round(m)), " nt")) %>%
  dplyr::arrange(region, group) %>%
  # one text line per group, stacked down from the top of the panel
  dplyr::group_by(region) %>%
  dplyr::mutate(vj = 1.4 + 1.35 * (dplyr::row_number() - 1)) %>%
  dplyr::ungroup()

stopifnot("expected 3 medians per region" =
  all(dplyr::count(panel_medians, region)$n == 3))
print(as.data.frame(panel_medians %>%
                      dplyr::select(region, group, n, m)), row.names = FALSE)
```

# Density plots

Lengths span orders of magnitude, so the x axis is log10 with nucleotide breaks.

```{r density-fn}
len_breaks <- c(10, 30, 100, 300, 1000, 3000, 10000, 30000)

length_density <- function(df, region_name) {
  sub <- df %>% dplyr::filter(region == region_name)
  wt  <- wilcox.test(sub$length[sub$class == "4E-resistant"],
                     sub$length[sub$class == "4E-sensitive"])
  med <- sub %>% dplyr::group_by(class) %>%
    dplyr::summarise(m = median(length), .groups = "drop")

  lab <- paste0(
    "Wilcoxon p = ", format.pval(wt$p.value, digits = 3, eps = 2.2e-16),
    "\nn = ", sum(sub$class == med$class[1]), " / ", sum(sub$class == med$class[2]))

  pm  <- panel_medians %>% dplyr::filter(region == region_name)
  ref <- med_expressed %>% dplyr::filter(region == region_name)

  ggplot(sub, aes(x = length, color = class)) +
    geom_density(fill = NA, linewidth = 0.8) +
    geom_vline(xintercept = ref$m_all, colour = "grey50",
               linetype = "dashed", linewidth = 0.5) +
    geom_vline(data = med, aes(xintercept = m, color = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    # all three medians, colour-matched to their dashed lines
    geom_text(data = pm,
              aes(x = min(sub$length), y = Inf, label = label,
                  colour = group, vjust = vj),
              hjust = 0, size = 3, show.legend = FALSE, inherit.aes = FALSE) +
    scale_x_log10(breaks = len_breaks, labels = scales::comma) +
    # the reference grey is in the scale so geom_text can map to it, but is kept out
    # of `breaks` so the legend still shows only the two classes
    scale_color_manual(values = median_colors, breaks = names(class_colors),
                       name = NULL) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18))) +
    labs(title = paste0(region_name, " length, 4E-resistant vs 4E-sensitive"),
         x = paste0(region_name, " length (nt, log10 scale)"), y = "Density") +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.4,
             size = 3.2, label = lab) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
}
```

## 5'UTR

```{r utr5, fig.width = 7, fig.height = 4.5}
p_utr5 <- length_density(d_long, "5'UTR")
print(p_utr5)
if (SAVE_PLOTS) ggsave(here("plots", "92_utr5_length_density_4e.pdf"), p_utr5,
                       width = 7, height = 4.5)
```

## CDS

```{r cds, fig.width = 7, fig.height = 4.5}
p_cds <- length_density(d_long, "CDS")
print(p_cds)
if (SAVE_PLOTS) ggsave(here("plots", "92_cds_length_density_4e.pdf"), p_cds,
                       width = 7, height = 4.5)
```

## 3'UTR

```{r utr3, fig.width = 7, fig.height = 4.5}
p_utr3 <- length_density(d_long, "3'UTR")
print(p_utr3)
if (SAVE_PLOTS) ggsave(here("plots", "92_utr3_length_density_4e.pdf"), p_utr3,
                       width = 7, height = 4.5)
```

## All three regions together

```{r facet, fig.width = 11, fig.height = 4}
p_facet <- ggplot(d_long, aes(x = length, color = class)) +
  geom_density(fill = NA, linewidth = 0.8) +
  geom_vline(data = med_expressed, aes(xintercept = m_all),
             colour = "grey50", linetype = "dashed", linewidth = 0.5,
             inherit.aes = FALSE) +
  geom_vline(data = panel_medians %>% dplyr::filter(group != REF_GROUP),
             aes(xintercept = m, colour = group),
             linetype = "dashed", linewidth = 0.5, inherit.aes = FALSE,
             show.legend = FALSE) +
  geom_text(data = panel_medians,
            aes(x = min(d_long$length), y = Inf, label = label,
                colour = group, vjust = vj),
            hjust = 0, size = 2.3, show.legend = FALSE, inherit.aes = FALSE) +
  facet_wrap(~ region, scales = "free_y", nrow = 1) +
  scale_x_log10(breaks = len_breaks, labels = scales::comma) +
  # headroom so the three median lines clear the density peak
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.32))) +
  scale_color_manual(values = median_colors, breaks = names(class_colors),
                     name = NULL) +
  labs(title = paste0("Transcript region lengths, 4E-resistant vs 4E-sensitive (n = ",
                      sum(d$class == "4E-resistant"), " / ",
                      sum(d$class == "4E-sensitive"), " transcripts)"),
       subtitle = "dashed lines = medians; grey = all expressed mRNAs in the Teleman data",
       x = "Length (nt, log10 scale)", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 9, colour = "grey40"),
        axis.text.x = element_text(angle = 45, hjust = 1))
print(p_facet)
if (SAVE_PLOTS) ggsave(here("plots", "92_region_length_density_4e_facet.pdf"), p_facet,
                       width = 11, height = 4)
```

# Boxplots

The densities show shape; the boxplots show the medians and spread side by side.

```{r box, fig.width = 8, fig.height = 4.5}
n_lab <- d_long %>%
  dplyr::count(region, class) %>%
  dplyr::mutate(label = paste0("n = ", n))

p_box <- ggplot(d_long, aes(x = class, y = length, fill = class, color = class)) +
  geom_boxplot(alpha = 0.15, linewidth = 0.6, outlier.size = 0.4,
               outlier.alpha = 0.3) +
  geom_text(data = n_lab, aes(x = class, y = Inf, label = label),
            vjust = 1.4, size = 3, show.legend = FALSE, inherit.aes = FALSE,
            colour = "grey30") +
  facet_wrap(~ region, nrow = 1) +
  scale_y_log10(breaks = len_breaks, labels = scales::comma) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_color_manual(values = class_colors, name = NULL) +
  labs(title = "Transcript region lengths, 4E-resistant vs 4E-sensitive",
       x = NULL, y = "Length (nt, log10 scale)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
print(p_box)
if (SAVE_PLOTS) ggsave(here("plots", "92_region_length_boxplot_4e.pdf"), p_box,
                       width = 8, height = 4.5)
```

# Session info

```{r session}
sessionInfo()
```
