---
title: "94 — Frame-Control Pooled RSCU of 4E-Resistant vs 4E-Sensitive mRNAs (Teleman/Roiuk HeLa)"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Codon usage of the **4E-resistant** and **4E-sensitive** sets from notebook 76, drawn in
the frame-control lollipop form of notebooks 54 and 93.

| Figure | x-axis statistic | Reference pool |
|---|---|---|
| `94a_rscu_frame_ctrl_4e_resistant` | pooled bg-RSCU minus 1, **resistant** set | expressed Teleman transcriptome |
| `94b_rscu_frame_ctrl_4e_sensitive` | pooled bg-RSCU minus 1, **sensitive** set | expressed Teleman transcriptome |
| `94c_rscu_frame_ctrl_4e_res_minus_sens` | delta pooled RSCU, **resistant minus sensitive** | the two sets against each other |
| `94d_rscu_resistant_vs_sensitive_scatter` | resistant vs sensitive bg-RSCU minus 1 | continuous, one point per codon |

**Background = every gene the Teleman/Roiuk contrast could test**
(`output/teleman_riboseq_te_all_genes.csv`, 11,306 genes; the ones with a usable v49
transcript are in `feature_matrix_teleman_4e.rds`), with **no screening on effect size**.
The resistant and sensitive genes are themselves part of that background, exactly as in
nb93 — so a single-set bg-RSCU of 1 means "this set uses that codon at the rate of a
typical expressed mRNA". The chi-squared tests use the **complement** of each set rather
than the full background, because a set is not independent of a pool that contains it.

**Each frame gets its own frame-matched background** (nb93 convention, not nb54's): in
frame +1 the pseudo-codon composition of the background differs from frame 0, and
dividing by the frame-0 background would report that difference rather than anything
about the gene set.

**Why this notebook exists.** Notebook 79 showed these two sets are almost separable on
CDS GC and especially GC3, and notebook 88 modelled that. Aggregate GC3 nets out
opposing preferences within synonymous families, so it cannot say whether the split is a
*synonymous-choice* effect. Pooled bg-RSCU is normalized within each amino-acid family
and the frame +1/+2 panels control for raw nucleotide composition: a codon that only
moves in frame 0 is a codon-level signal, one that moves in all three frames is
composition.

**These classes are opposite tails of one variable.** Every statement below is
resistant-relative-to-sensitive or tail-relative-to-background, never
"4E-resistant mRNAs versus a typical mRNA of the same expression".

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Load Libraries

```{r load_libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(coRdon)
```

```{r config}
# Plain constants, no knit `params`: every chunk in this notebook runs as-is in Positron
# (or any interactive R session) as well as under rmarkdown::render(). Edit here to point
# the notebook at a different set of inputs.
FEATURE_MATRIX <- "feature_matrix_teleman_4e.rds"          # output/predictive_modeling/
POS_CSV        <- "teleman_4e_resistant_TE_1hr_lfc0.5.csv" # output/genesets/
NEG_CSV        <- "teleman_4e_sensitive_TE_1hr_lfc0.5.csv" # output/genesets/
TELEMAN_ALL    <- "teleman_riboseq_te_all_genes.csv"       # output/
STAR_RATIO     <- 1.5    # frame-specificity multiple for the star rule
SEED           <- 9      # project-wide standard seed
SAVE_PLOTS     <- TRUE   # FALSE to draw without writing PDFs/CSVs

stopifnot(
  "STAR_RATIO must be a single number >= 1" =
    is.numeric(STAR_RATIO) && length(STAR_RATIO) == 1 && STAR_RATIO >= 1,
  "SEED must be a single number" = is.numeric(SEED) && length(SEED) == 1
)
set.seed(SEED)

# Class colors match nb79 (Okabe-Ito vermillion / dark blue). #E69F00 and #56B4E9 are
# reserved project-wide for eIF3d and eIF3e and must not be reused here.
class_colors <- c("4E-resistant" = "#D55E00", "4E-sensitive" = "#0072B2")

# Third-position coloring, same pair as nb93's scatter panels.
gc3_colors <- c("AT-ending" = "#009E73", "GC-ending" = "#CC79A7")

cat("Config — star ratio:", STAR_RATIO, "| seed:", SEED,
    "| save_plots:", SAVE_PLOTS, "\n")
```

---

# Section 1: Genetic Code Scaffold

```{r genetic_code}
gen_code     <- Biostrings::getGeneticCode("SGC0")
sense_codons <- names(gen_code[gen_code != "*"])
aa_to_codons <- split(sense_codons, gen_code[sense_codons])
multi_aa     <- names(aa_to_codons)[sapply(aa_to_codons, length) > 1]

# ATG (Met) and TGG (Trp) are the only codon for their amino acid: RSCU is 1 in every
# pool by construction and no chi-squared test is defined. Excluded everywhere below,
# matching nb54/nb93.
multi_syn_codons <- sense_codons[gen_code[sense_codons] %in% multi_aa]
stopifnot(
  "single-codon amino acids leaked into multi_syn_codons" =
    !any(c("ATG", "TGG") %in% multi_syn_codons),
  "unexpected number of multi-synonymous sense codons" =
    length(multi_syn_codons) == length(sense_codons) - 2
)

codon_meta <- tibble(
  codon      = sense_codons,
  amino_acid = as.character(gen_code[sense_codons]),
  third_base = substr(sense_codons, 3, 3)
) %>%
  mutate(gc3_class = factor(if_else(third_base %in% c("G", "C"),
                                    "GC-ending", "AT-ending"),
                            levels = names(gc3_colors)))

cat("Sense codons:", length(sense_codons),
    "| multi-synonymous:", length(multi_syn_codons), "\n")
```

---

# Section 2: Gene Sets and the Expressed Background

```{r gene_sets}
fm <- readRDS(here("output", "predictive_modeling", FEATURE_MATRIX))
pos <- read_csv(here("output", "genesets", POS_CSV), show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", NEG_CSV), show_col_types = FALSE)
teleman_all <- read_csv(here("output", TELEMAN_ALL), show_col_types = FALSE)

cat("Feature matrix (HeLa dominant transcripts):", nrow(fm), "genes\n")
cat("Teleman tested universe:", nrow(teleman_all), "genes\n")
cat("Resistant:", nrow(pos), "| sensitive:", nrow(neg), "\n")

# ---- Direction and identity guards ----
# nb76 defines resistant as the positive-LFC tail of log2FC(TE) under eIF4E inhibition
# and asserts CDKN2B / GLO1 as named members of the two classes (its Fig 1A check).
# If either flips, every "resistant prefers X" statement below inverts.
stopifnot(
  "resistant set contains a non-positive LFC" = all(pos$log2FoldChange > 0),
  "sensitive set contains a non-negative LFC" = all(neg$log2FoldChange < 0),
  "classes overlap"        = length(intersect(pos$ensembl_gene, neg$ensembl_gene)) == 0,
  "CDKN2B missing from the resistant set" = "CDKN2B" %in% pos$db_gene_symbol,
  "GLO1 missing from the sensitive set"   = "GLO1"   %in% neg$db_gene_symbol,
  "gene sets are not inside the tested universe" =
    all(c(pos$ensembl_gene, neg$ensembl_gene) %in% teleman_all$gene_id),
  "feature matrix is not gene-unique" = !any(duplicated(fm$gene_id_clean))
)
cat("Resistant log2FC range:", round(range(pos$log2FoldChange), 2),
    "| sensitive:", round(range(neg$log2FoldChange), 2), "\n")

# The background is the whole tested universe, unscreened. The feature matrix is that
# universe minus the genes with no usable v49 transcript, so it IS the background here.
stopifnot("feature matrix contains genes outside the tested universe" =
            all(fm$gene_id_clean %in% teleman_all$gene_id))
cat("Tested genes dropped for lack of a v49 transcript:",
    nrow(teleman_all) - nrow(fm), "\n")

tx_of <- function(gene_ids) {
  fm$transcript_id_clean[fm$gene_id_clean %in% gene_ids]
}
res_tx_req <- tx_of(pos$ensembl_gene)
sen_tx_req <- tx_of(neg$ensembl_gene)
cat("Transcripts available — resistant:", length(res_tx_req),
    "| sensitive:", length(sen_tx_req), "\n")
```

---

# Section 3: CDS Sequences and Per-Frame Codon Counts

```{r load_cds}
txdb   <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
genome <- BSgenome.Hsapiens.UCSC.hg38

cds_all         <- cdsBy(txdb, by = "tx", use.names = TRUE)
cds_seqs_all    <- extractTranscriptSeqs(genome, cds_all)
cds_names_clean <- sub("\\..*", "", names(cds_seqs_all))

keep_idx   <- cds_names_clean %in% fm$transcript_id_clean
cds_frame0 <- cds_seqs_all[keep_idx]
names(cds_frame0) <- cds_names_clean[keep_idx]

valid      <- width(cds_frame0) %% 3 == 0 & width(cds_frame0) > 0
cds_frame0 <- cds_frame0[valid]
cat("CDS sequences (frame 0, divisible by 3):", length(cds_frame0),
    "of", nrow(fm), "feature-matrix transcripts\n")
stopifnot("too few CDS sequences recovered from the v49 TxDb" =
            length(cds_frame0) > 0.9 * nrow(fm))
```

```{r frame_counts}
shift_frame <- function(seqs, shift) {
  shifted <- subseq(seqs, start = shift + 1)
  new_len <- floor(width(shifted) / 3) * 3
  subseq(shifted, start = 1, end = new_len)
}
make_counts <- function(seqs) {
  seqs <- seqs[width(seqs) > 0]
  mat  <- codonCounts(codonTable(seqs))
  rownames(mat) <- names(seqs)
  mat
}

counts_by_frame <- list(
  f0 = make_counts(cds_frame0),
  f1 = make_counts(shift_frame(cds_frame0, 1)),
  f2 = make_counts(shift_frame(cds_frame0, 2))
)
counts_mat0 <- counts_by_frame$f0

for (fr in names(counts_by_frame)) {
  cat(fr, ": ", nrow(counts_by_frame[[fr]]), " transcripts, ",
      ncol(counts_by_frame[[fr]]), " codon columns\n", sep = "")
}
stopifnot(
  "codon count matrices must be non-negative" =
    all(sapply(counts_by_frame, function(m) all(m >= 0))),
  "frame-shifted matrices lost transcripts unexpectedly" =
    all(sapply(counts_by_frame, nrow) > 0.99 * nrow(counts_mat0))
)

res_tx <- intersect(res_tx_req, rownames(counts_mat0))
sen_tx <- intersect(sen_tx_req, rownames(counts_mat0))
all_tx <- rownames(counts_mat0)
n_res  <- length(res_tx); n_sen <- length(sen_tx); n_all <- length(all_tx)

cat("Transcripts scored — resistant:", n_res, "| sensitive:", n_sen,
    "| expressed background:", n_all, "\n")
stopifnot(
  "too few resistant transcripts" = n_res > 100,
  "too few sensitive transcripts" = n_sen > 100,
  "the two sets must not share a transcript" =
    length(intersect(res_tx, sen_tx)) == 0
)
```

---

# Section 4: Per-Frame Pooled bg-RSCU

For codon *j* of amino acid *aa*:

```
pooled_RSCU_j = (set_count_j / set_aa_total) / (bg_count_j / bg_aa_total)
```

with numerator and denominator taken **in the same frame**. A value of 1 means the set
uses codon *j* at exactly the background rate for that amino acid family.

```{r rscu_helpers}
# Pooled background proportions per amino acid family, for one frame.
make_bg_props <- function(counts_mat) {
  bg_counts <- colSums(counts_mat)
  props <- lapply(aa_to_codons[multi_aa], function(codons) {
    codons   <- intersect(codons, names(bg_counts))
    aa_total <- sum(bg_counts[codons])
    if (length(codons) == 0 || aa_total == 0) return(NULL)
    bg_counts[codons] / aa_total
  })
  Filter(Negate(is.null), props)
}

pooled_counts <- function(tx_ids, counts_mat) {
  colSums(counts_mat[intersect(tx_ids, rownames(counts_mat)), , drop = FALSE])
}

compute_pooled_rscu <- function(tx_ids, counts_mat, bg_props) {
  pooled   <- pooled_counts(tx_ids, counts_mat)
  rscu_vec <- set_names(rep(NA_real_, length(sense_codons)), sense_codons)
  for (aa in names(bg_props)) {
    props    <- bg_props[[aa]]
    codons   <- intersect(names(props), names(pooled))
    if (length(codons) == 0) next
    aa_total <- sum(pooled[codons])
    if (aa_total == 0) next
    bp    <- props[codons]
    valid <- bp > 0
    if (any(valid)) {
      rscu_vec[codons[valid]] <- (pooled[codons[valid]] / aa_total) / bp[valid]
    }
  }
  rscu_vec
}
```

```{r compute_rscu_all_frames}
bg_props_by_frame <- lapply(counts_by_frame, make_bg_props)

rscu_of <- function(tx_ids) {
  set_names(
    lapply(names(counts_by_frame), function(fr)
      compute_pooled_rscu(tx_ids, counts_by_frame[[fr]], bg_props_by_frame[[fr]])),
    names(counts_by_frame)
  )
}

rscu_res <- rscu_of(res_tx)
rscu_sen <- rscu_of(sen_tx)
delta_rscu <- set_names(
  lapply(names(counts_by_frame), function(fr) rscu_res[[fr]] - rscu_sen[[fr]]),
  names(counts_by_frame)
)

for (fr in names(counts_by_frame)) {
  cat(fr, " — resistant RSCU range: ",
      paste(round(range(rscu_res[[fr]], na.rm = TRUE), 3), collapse = " to "),
      " | sensitive: ",
      paste(round(range(rscu_sen[[fr]], na.rm = TRUE), 3), collapse = " to "),
      "\n", sep = "")
}
stopifnot("Pooled RSCU contains negative values" =
            all(sapply(c(rscu_res, rscu_sen), function(v) all(v >= 0, na.rm = TRUE))))
```

```{r bg_validation}
# The background scored against itself must give RSCU = 1 for every codon, in every
# frame. This is the check that catches a numerator/denominator frame mismatch.
for (fr in names(counts_by_frame)) {
  chk <- compute_pooled_rscu(all_tx, counts_by_frame[[fr]], bg_props_by_frame[[fr]])
  chk <- chk[!is.na(chk)]
  cat(fr, " bg-RSCU self-check: min = ", round(min(chk), 8),
      " max = ", round(max(chk), 8), " (n = ", length(chk), ")\n", sep = "")
  stopifnot("Pooled bg-RSCU deviates from 1 — frame-matched background is wrong" =
              all(abs(chk - 1) < 1e-10))
}
```

```{r spot_check}
# Cross-notebook consistency with nb79. That notebook reports median GC3 ~63% in the
# resistant set against ~50% in the sensitive set, one of the largest compositional
# splits in the project. If the sets or the frames were mixed up here, the frame-0
# delta would not favour GC-ending codons.
delta_f0 <- tibble(codon = multi_syn_codons,
                   delta = as.numeric(delta_rscu$f0[multi_syn_codons])) %>%
  left_join(codon_meta, by = "codon")

gc3_means <- delta_f0 %>%
  group_by(gc3_class) %>%
  summarise(mean_delta = mean(delta), n = dplyr::n(), .groups = "drop")
print(as.data.frame(gc3_means), row.names = FALSE)

stopifnot(
  "resistant-minus-sensitive delta does not favour GC-ending codons — contradicts nb79" =
    gc3_means$mean_delta[gc3_means$gc3_class == "GC-ending"] >
    gc3_means$mean_delta[gc3_means$gc3_class == "AT-ending"]
)

# Frame-0 deltas must be much larger than the out-of-frame controls if any of this is
# codon-level rather than composition; reported, not asserted.
mean_abs <- function(v) round(mean(abs(v[multi_syn_codons]), na.rm = TRUE), 4)
cat("\nMean |delta| by frame (multi-synonymous codons):\n")
for (fr in names(delta_rscu)) cat("  ", fr, ": ", mean_abs(delta_rscu[[fr]]), "\n", sep = "")
```

---

# Section 5: Significance Tests

Frame-0 only. Per codon, a 2x2 test on raw pooled counts: rows = the two pools, columns
= (this codon, all other synonyms of its amino acid). BH-adjusted. For the single-set
figures the comparator is the **complement** of that set within the expressed
transcriptome, not the full background.

```{r chisq}
chisq_pools <- function(counts_a, counts_b) {
  imap_dfr(aa_to_codons[multi_aa], function(codons, aa) {
    codons <- intersect(codons, names(counts_a))
    if (length(codons) < 2) return(NULL)
    n_aa_a <- sum(counts_a[codons]); n_aa_b <- sum(counts_b[codons])
    if (n_aa_a == 0 || n_aa_b == 0) return(NULL)
    map_dfr(codons, function(cd) {
      mat <- round(matrix(
        c(counts_a[cd], n_aa_a - counts_a[cd],
          counts_b[cd], n_aa_b - counts_b[cd]),
        nrow = 2, byrow = TRUE
      ))
      test <- if (any(mat < 5)) fisher.test(mat) else chisq.test(mat, correct = FALSE)
      tibble(amino_acid = aa, codon = cd, p_val_chi = test$p.value)
    })
  }) %>%
    mutate(p_adj_chi = p.adjust(p_val_chi, method = "BH"))
}

cnt_res <- pooled_counts(res_tx, counts_mat0)
cnt_sen <- pooled_counts(sen_tx, counts_mat0)
cnt_all <- pooled_counts(all_tx, counts_mat0)

cnt_res_comp <- pooled_counts(setdiff(all_tx, res_tx), counts_mat0)
cnt_sen_comp <- pooled_counts(setdiff(all_tx, sen_tx), counts_mat0)

stopifnot(
  "complement counts must reconstruct the background" =
    all(abs((cnt_res + cnt_res_comp) - cnt_all) < 1e-6) &&
    all(abs((cnt_sen + cnt_sen_comp) - cnt_all) < 1e-6)
)

chi_res   <- chisq_pools(cnt_res, cnt_res_comp)
chi_sen   <- chisq_pools(cnt_sen, cnt_sen_comp)
chi_delta <- chisq_pools(cnt_res, cnt_sen)

chi_tables <- list("resistant vs complement" = chi_res,
                   "sensitive vs complement" = chi_sen,
                   "resistant vs sensitive"  = chi_delta)
iwalk(chi_tables, function(tb, nm)
  cat(nm, "— significant codons (BH q < 0.05):",
      sum(tb$p_adj_chi < 0.05, na.rm = TRUE), "of", nrow(tb), "\n"))
stopifnot(
  "chi-squared tables must cover every multi-synonymous codon exactly once" =
    all(sapply(chi_tables, nrow) == length(multi_syn_codons))
)
```

---

# Section 6: Figure Builder

```{r figure_helpers}
frame_levels <- c("Frame 0\n(real codons)",
                  "Frame +1\n(non-coding ctrl)",
                  "Frame +2\n(non-coding ctrl)")

# Star rule (nb54/nb93): frame-0 BH q < 0.05 AND |frame-0 stat| > STAR_RATIO x the larger
# of the two shifted-frame stats. A codon whose out-of-frame pseudo-codon shows the same
# shift is reporting nucleotide composition, not codon choice.
frame_specificity <- function(v0, v1, v2, chi_tbl) {
  padj <- set_names(chi_tbl$p_adj_chi, chi_tbl$codon)
  tibble(
    codon     = multi_syn_codons,
    f0_delta  = as.numeric(v0[multi_syn_codons]),
    f1_delta  = as.numeric(v1[multi_syn_codons]),
    f2_delta  = as.numeric(v2[multi_syn_codons]),
    p_adj_chi = as.numeric(padj[multi_syn_codons])
  ) %>%
    mutate(
      max_shift      = pmax(abs(f1_delta), abs(f2_delta)),
      codon_specific = !is.na(p_adj_chi) & !is.na(f0_delta) & !is.na(max_shift) &
        p_adj_chi < 0.05 & abs(f0_delta) > max_shift * STAR_RATIO
    )
}

make_frame_figure <- function(v0, v1, v2, chi_tbl, title, subtitle, xlab, outfile) {
  spec <- frame_specificity(v0, v1, v2, chi_tbl)
  stopifnot(
    "frame_specificity must cover every multi-synonymous codon exactly once" =
      nrow(spec) == length(multi_syn_codons),
    "codon_specific must be TRUE/FALSE, never NA" = !any(is.na(spec$codon_specific))
  )
  codon_order <- spec %>% arrange(f0_delta) %>% pull(codon)

  plot_df <- spec %>%
    dplyr::select(codon, codon_specific,
                  `Frame 0\n(real codons)`      = f0_delta,
                  `Frame +1\n(non-coding ctrl)` = f1_delta,
                  `Frame +2\n(non-coding ctrl)` = f2_delta) %>%
    pivot_longer(starts_with("Frame"), names_to = "frame", values_to = "delta") %>%
    mutate(frame = factor(frame, levels = frame_levels)) %>%
    left_join(codon_meta, by = "codon") %>%
    mutate(codon = factor(codon, levels = codon_order))

  # Stars pinned to the left edge of the Frame 0 panel as a row indicator.
  star_df <- plot_df %>%
    filter(codon_specific, frame == frame_levels[1]) %>%
    mutate(delta = min(plot_df$delta, na.rm = TRUE))

  # Do GC-ending codons sit on a different side than AT-ending ones?
  f0 <- spec %>% left_join(codon_meta, by = "codon") %>% filter(!is.na(f0_delta))
  wt <- wilcox.test(f0_delta ~ gc3_class, data = f0)
  med_of <- function(cls) median(f0$f0_delta[f0$gc3_class == cls])
  n_of   <- function(cls) sum(f0$gc3_class == cls)
  ann <- sprintf(
    "Frame 0 median: GC-ending = %.3f (n = %d) vs AT-ending = %.3f (n = %d); Wilcoxon p = %.2g",
    med_of("GC-ending"), n_of("GC-ending"),
    med_of("AT-ending"), n_of("AT-ending"), wt$p.value
  )

  p <- ggplot(plot_df, aes(x = delta, y = codon, colour = gc3_class)) +
    geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey60") +
    geom_point(size = 1.5, alpha = 0.85) +
    geom_point(data = star_df, shape = 8, size = 2.5,
               colour = "black", stroke = 0.6, show.legend = FALSE) +
    facet_wrap(~ frame, nrow = 1) +
    scale_colour_manual(values = gc3_colors, name = "Third position") +
    labs(title = title, subtitle = subtitle, x = xlab, y = NULL, caption = ann) +
    theme_classic(base_size = 13) +
    theme(
      axis.text.y     = element_text(size = 9, family = "mono"),
      legend.position = "top",
      strip.text      = element_text(face = "bold"),
      plot.caption    = element_text(hjust = 0, size = 9, colour = "grey30")
    )
  print(p)
  if (SAVE_PLOTS) {
    ggsave(here("plots", outfile), p, width = 9.5, height = 11)
    cat("Saved: plots/", outfile, "\n", sep = "")
  }

  cat("Codon-specific (starred):", sum(spec$codon_specific), "of", nrow(spec),
      "multi-synonymous codons\n")
  cat(ann, "\n")
  invisible(list(plot = p, spec = spec, wilcox = wt,
                 n_gc = n_of("GC-ending"), n_at = n_of("AT-ending")))
}

star_subtitle <- paste0(
  "star (Frame 0 only): BH-adj chi-sq p<0.05 AND abs(frame-0 value) > ",
  STAR_RATIO, "x max(abs(frame+1), abs(frame+2))"
)
```

---

# Section 7: Frame-Control Lollipops

```{r define_conditions}
conditions <- list(
  list(key = "resistant",
       v   = list(rscu_res$f0 - 1, rscu_res$f1 - 1, rscu_res$f2 - 1),
       chi = chi_res,
       title_stem = "4E-resistant codon preference",
       n_note = paste0("n = ", n_res, " transcripts vs ", n_all - n_res,
                       " background transcripts (frame-matched background)"),
       xlab = "pooled bg-RSCU minus 1 (4E-resistant vs expressed background)",
       file_stem = "94a_rscu_frame_ctrl_4e_resistant"),
  list(key = "sensitive",
       v   = list(rscu_sen$f0 - 1, rscu_sen$f1 - 1, rscu_sen$f2 - 1),
       chi = chi_sen,
       title_stem = "4E-sensitive codon preference",
       n_note = paste0("n = ", n_sen, " transcripts vs ", n_all - n_sen,
                       " background transcripts (frame-matched background)"),
       xlab = "pooled bg-RSCU minus 1 (4E-sensitive vs expressed background)",
       file_stem = "94b_rscu_frame_ctrl_4e_sensitive"),
  list(key = "res_minus_sens",
       v   = list(delta_rscu$f0, delta_rscu$f1, delta_rscu$f2),
       chi = chi_delta,
       title_stem = "4E-resistant minus 4E-sensitive codon preference",
       n_note = paste0("n = ", n_res, " resistant vs ", n_sen, " sensitive transcripts"),
       xlab = "delta pooled RSCU (resistant minus sensitive)",
       file_stem = "94c_rscu_frame_ctrl_4e_res_minus_sens")
)
```

```{r fig_lollipops, fig.width=9.5, fig.height=11, results="asis"}
fig_results <- list()
for (cond in conditions) {
  cat("\n\n### ", cond$title_stem, "\n\n", sep = "")
  fig_results[[cond$key]] <- make_frame_figure(
    v0 = cond$v[[1]], v1 = cond$v[[2]], v2 = cond$v[[3]],
    chi_tbl  = cond$chi,
    title    = paste0(cond$title_stem, ", colored by third position"),
    subtitle = paste0(star_subtitle, "\n", cond$n_note),
    xlab     = cond$xlab,
    outfile  = paste0(cond$file_stem, ".pdf")
  )
}
```

---

# Section 8: Resistant vs Sensitive, Codon by Codon

The two single-set statistics against each other. Both are measured against the same
background, so a codon on the diagonal is used identically by the two classes and one
off the diagonal is a genuine class difference. Opposite tails of one variable should
give a **negative** correlation.

```{r scatter, fig.width=7.5, fig.height=6.5}
scatter_df <- tibble(codon = multi_syn_codons) %>%
  mutate(
    resistant = as.numeric((rscu_res$f0 - 1)[codon]),
    sensitive = as.numeric((rscu_sen$f0 - 1)[codon])
  ) %>%
  left_join(codon_meta, by = "codon") %>%
  left_join(fig_results$res_minus_sens$spec %>%
              dplyr::select(codon, codon_specific), by = "codon")

stopifnot(
  "scatter data must have one row per multi-synonymous codon" =
    nrow(scatter_df) == length(multi_syn_codons),
  "RSCU statistics must be finite" =
    all(is.finite(scatter_df$resistant)) && all(is.finite(scatter_df$sensitive))
)

rho <- cor(scatter_df$resistant, scatter_df$sensitive, method = "spearman")
r   <- cor(scatter_df$resistant, scatter_df$sensitive, method = "pearson")
lab <- sprintf("Spearman rho = %.2f (p = %.2g)\nPearson R = %.2f (p = %.2g)\nn = %d codons",
               rho,
               cor.test(scatter_df$resistant, scatter_df$sensitive,
                        method = "spearman", exact = FALSE)$p.value,
               r,
               cor.test(scatter_df$resistant, scatter_df$sensitive)$p.value,
               nrow(scatter_df))

p_scatter <- ggplot(scatter_df, aes(x = sensitive, y = resistant)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted",
              linewidth = 0.4, colour = "grey60") +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30",
              linewidth = 0.6, alpha = 0.15) +
  geom_point(aes(colour = gc3_class, shape = codon_specific), size = 2.2, alpha = 0.9) +
  ggrepel::geom_text_repel(aes(label = codon, colour = gc3_class), size = 2.4,
                           max.overlaps = Inf, min.segment.length = 0.2,
                           segment.size = 0.2, show.legend = FALSE) +
  annotate("text", x = -Inf, y = Inf, label = lab,
           hjust = -0.05, vjust = 1.2, size = 3, colour = "grey30") +
  scale_colour_manual(values = gc3_colors, name = "Third position") +
  scale_shape_manual(values = c(`TRUE` = 8, `FALSE` = 16),
                     name = "Codon-specific (starred)") +
  labs(
    title = "Codon usage of 4E-resistant vs 4E-sensitive mRNAs",
    subtitle = paste0("Frame 0, pooled bg-RSCU minus 1 against the same expressed ",
                      "background (n = ", n_all, " transcripts).",
                      "\nDotted line = identical usage; grey band = linear fit with 95% CI."),
    x = "4E-sensitive (bg-RSCU - 1)",
    y = "4E-resistant (bg-RSCU - 1)"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

print(p_scatter)
if (SAVE_PLOTS) {
  ggsave(here("plots", "94d_rscu_resistant_vs_sensitive_scatter.pdf"),
         p_scatter, width = 7.5, height = 6.5)
  cat("Saved: plots/94d_rscu_resistant_vs_sensitive_scatter.pdf\n")
}
cat(lab, "\n")
```

---

# Section 9: Summary Tables

```{r summary_table}
spec_of <- function(key) fig_results[[key]]$spec

summary_tbl <- codon_meta %>%
  filter(codon %in% multi_syn_codons) %>%
  left_join(spec_of("resistant") %>%
              dplyr::select(codon, res_f0 = f0_delta, res_q = p_adj_chi,
                            res_star = codon_specific), by = "codon") %>%
  left_join(spec_of("sensitive") %>%
              dplyr::select(codon, sen_f0 = f0_delta, sen_q = p_adj_chi,
                            sen_star = codon_specific), by = "codon") %>%
  left_join(spec_of("res_minus_sens") %>%
              dplyr::select(codon, delta_f0 = f0_delta, delta_q = p_adj_chi,
                            delta_star = codon_specific), by = "codon") %>%
  arrange(desc(delta_f0))

if (SAVE_PLOTS) {
  out_csv <- here("output", "predictive_modeling", "94_rscu_4e_resistant_vs_sensitive.csv")
  write_csv(summary_tbl, out_csv)
  cat("Saved:", out_csv, "\n")
}

stopifnot(
  "summary table must have one row per multi-synonymous codon" =
    nrow(summary_tbl) == length(multi_syn_codons),
  "frame-0 statistics must be finite" =
    all(is.finite(c(summary_tbl$res_f0, summary_tbl$sen_f0, summary_tbl$delta_f0)))
)

cat("\nTop 10 codons by resistant-minus-sensitive delta:\n")
print(summary_tbl %>%
        dplyr::select(codon, amino_acid, gc3_class, res_f0, sen_f0, delta_f0,
                      delta_star) %>% head(10))
cat("\nBottom 10:\n")
print(summary_tbl %>%
        dplyr::select(codon, amino_acid, gc3_class, res_f0, sen_f0, delta_f0,
                      delta_star) %>% tail(10))

cat("\nStarred (codon-specific) counts, of", length(multi_syn_codons), "codons:\n")
cat("  resistant:", sum(summary_tbl$res_star),
    "| sensitive:", sum(summary_tbl$sen_star),
    "| resistant-minus-sensitive:", sum(summary_tbl$delta_star), "\n")
```

```{r wilcoxon_summary}
wilcox_tbl <- imap_dfr(fig_results, function(res, nm) {
  tibble(
    comparison  = nm,
    n_gc_ending = res$n_gc,
    n_at_ending = res$n_at,
    wilcox_p    = res$wilcox$p.value
  )
})

# Same test, restricted to the codons that survive the frame control: the composition
# explanation is already excluded for these, so this is the codon-level version.
wilcox_starred <- imap_dfr(fig_results, function(res, nm) {
  d <- res$spec %>% left_join(codon_meta, by = "codon") %>% filter(codon_specific)
  if (n_distinct(d$gc3_class) < 2) {
    return(tibble(comparison = nm, n_starred = nrow(d),
                  n_gc_ending = sum(d$gc3_class == "GC-ending"),
                  n_at_ending = sum(d$gc3_class == "AT-ending"),
                  wilcox_p = NA_real_))
  }
  tibble(comparison  = nm,
         n_starred   = nrow(d),
         n_gc_ending = sum(d$gc3_class == "GC-ending"),
         n_at_ending = sum(d$gc3_class == "AT-ending"),
         wilcox_p    = wilcox.test(f0_delta ~ gc3_class, data = d)$p.value)
})

cat("Wilcoxon of the frame-0 statistic, GC-ending vs AT-ending codons (all codons):\n")
print(as.data.frame(wilcox_tbl), row.names = FALSE)
cat("\nSame, restricted to frame-control-surviving (starred) codons:\n")
print(as.data.frame(wilcox_starred), row.names = FALSE)

if (SAVE_PLOTS) {
  write_csv(wilcox_tbl,
            here("output", "predictive_modeling", "94_rscu_4e_gc3_wilcoxon.csv"))
}
```

```{r session_info}
sessionInfo()
```
