---
title: "95 — Expressed isoforms per gene, siCTRL hypoxia 1hr"
author: "Kate Matlin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
params:
  treatment: "siCTRL"
  condition: "hypoxia"
  timepoint: 1
  tpm_min: 1
  frac_min: 0.10
---

Step 1 of the CDS-3'-end question: **are there even multiple isoforms of the same gene
expressed in the siCTRL hypoxia 1hr data?** If most genes run a single isoform, the
planned within-gene AU-ending vs GC-ending contrast has nothing to compare.

Three questions, in order:

1. How many isoforms per gene are expressed (section 3)?
2. Are the multi-isoform genes real co-expression, or one dominant isoform plus noise
   (section 5)?
3. Do the co-expressed isoforms of a gene differ by **synonymous codon choice**, or do
   they encode different proteins (section 6)? This decides whether a codon or GC
   difference between two isoforms of one gene can be read as a translational signal at
   all, or whether it is just amino-acid composition following an alternative exon.

**Definitions used here** (all set in `params`):

| Choice | Value | Note |
|---|---|---|
| Expression source | salmon TPM, `Input_Ribo_RNAseq`, siCTRL hypoxia 1hr, 3 reps | RiboSeq not used; this is the mRNA pool |
| Eligible transcript | `transcript_type == "protein_coding"` **and** has a CDS in the v49 TxDb | NMD and retained-intron isoforms excluded |
| Expressed isoform | mean TPM >= `tpm_min` **and** isoform fraction >= `frac_min` | fraction denominator is the eligible transcripts of that gene only |
| Where defined | siCTRL hypoxia 1hr only | si3d is not used to define the isoform set |

TPM is reported by salmon over the full index, which includes the Drosophila spike-ins,
so the absolute TPM cut is very slightly conservative. Isoform *fractions* are computed
within a gene and are unaffected.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(here)
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)

if (!exists("params") || !is.list(params)) {
  params <- list(treatment = "siCTRL", condition = "hypoxia", timepoint = 1,
                 tpm_min = 1, frac_min = 0.10)
}
set.seed(9)
```

# Load inputs

```{r samples}
metadata <- read_csv(here("accessories", "metadata_231_timecourse.csv"),
                     show_col_types = FALSE)

sample_ids <- metadata %>%
  filter(type == "Input_Ribo_RNAseq",
         treatment == params$treatment,
         condition == params$condition,
         hours == params$timepoint) %>%
  pull(id)

stopifnot("no RNA-seq samples matched the params" = length(sample_ids) > 0)
cat("Samples:", paste(sample_ids, collapse = ", "), "\n")
```

```{r salmon}
tpm <- sample_ids %>%
  set_names() %>%
  map_dfr(~ read_tsv(here("2025_MDAcells_riboseq", "salmon", .x, "quant.sf.gz"),
                     show_col_types = FALSE) %>%
            dplyr::select(transcript_id = Name, TPM),
          .id = "sample") %>%
  pivot_wider(names_from = sample, values_from = TPM)

tpm <- tpm %>%
  mutate(mean_tpm = rowMeans(across(all_of(sample_ids))),
         n_reps_detected = rowSums(across(all_of(sample_ids)) > 0))

cat("Transcripts quantified:", nrow(tpm), "\n")
```

```{r annotation}
tx2gene <- read_tsv(here("accessories", "human", "tx2gene_hs_dm_v49_r111.tsv.gz"),
                    col_names = c("transcript_id", "gene_id"), show_col_types = FALSE) %>%
  filter(str_starts(transcript_id, "ENST"))

gene_symbols <- read_tsv(here("accessories", "human", "gene_anno_hs_dm_v49_r111.tsv"),
                         col_names = c("gene_id", "symbol", "type", "organism", "source"),
                         show_col_types = FALSE) %>%
  distinct(gene_id, .keep_all = TRUE) %>%
  dplyr::select(gene_id, symbol)

# Transcript biotypes are not stored in the TxDb, so they come from the GTF.
# Built once and cached; rebuilding takes ~2 min.
biotype_path <- here("accessories", "human", "tx_biotype_v49.tsv.gz")
if (!file.exists(biotype_path)) {
  gtf <- read_tsv(here("accessories", "human", "gencode.v49.annotation.gtf.gz"),
                  comment = "#", col_names = FALSE,
                  col_types = cols_only(X3 = col_character(), X9 = col_character())) %>%
    filter(X3 == "transcript")
  tibble(transcript_id   = str_match(gtf$X9, 'transcript_id "([^"]+)"')[, 2],
         transcript_type = str_match(gtf$X9, 'transcript_type "([^"]+)"')[, 2],
         gene_type       = str_match(gtf$X9, 'gene_type "([^"]+)"')[, 2]) %>%
    write_tsv(biotype_path)
}
tx_biotype <- read_tsv(biotype_path, show_col_types = FALSE)

txdb <- loadDb(here("accessories", "human", "txdb.gencode49.sqlite"))
cds_by_tx <- cdsBy(txdb, by = "tx", use.names = TRUE)

mane <- read_csv(here("accessories", "human", "mane_select_v49.csv"),
                 col_names = c("gene_id_clean", "mane_tx_clean", "symbol"),
                 show_col_types = FALSE)

cat("Transcripts with an annotated CDS:", length(cds_by_tx), "\n")
```

# Eligible transcripts and the expressed isoform set

```{r eligible}
eligible <- tpm %>%
  inner_join(tx2gene, by = "transcript_id") %>%
  left_join(tx_biotype, by = "transcript_id") %>%
  filter(transcript_type == "protein_coding",
         gene_type == "protein_coding",
         transcript_id %in% names(cds_by_tx)) %>%
  group_by(gene_id) %>%
  mutate(gene_tpm = sum(mean_tpm),
         iso_frac = if_else(gene_tpm > 0, mean_tpm / gene_tpm, 0)) %>%
  ungroup() %>%
  mutate(gene_id_clean       = sub("\\..*", "", gene_id),
         transcript_id_clean = sub("\\..*", "", transcript_id)) %>%
  left_join(gene_symbols, by = "gene_id")

expressed <- eligible %>%
  filter(mean_tpm >= params$tpm_min, iso_frac >= params$frac_min)

cat("Eligible transcripts:", nrow(eligible), "in", n_distinct(eligible$gene_id), "genes\n")
cat("Expressed isoforms:  ", nrow(expressed), "in", n_distinct(expressed$gene_id), "genes\n")
```

```{r census}
census <- expressed %>%
  group_by(gene_id, gene_id_clean, symbol) %>%
  summarize(n_isoforms = n(),
            gene_tpm   = dplyr::first(gene_tpm),
            top_frac   = max(iso_frac),
            .groups    = "drop")

census %>%
  dplyr::count(n_isoforms, name = "n_genes") %>%
  mutate(pct = round(100 * n_genes / sum(n_genes), 1)) %>%
  knitr::kable(caption = "Genes by number of expressed isoforms")

cat("\nGenes with >= 2 expressed isoforms:", sum(census$n_isoforms >= 2),
    sprintf("(%.1f%%)\n", 100 * mean(census$n_isoforms >= 2)))
cat("Median top-isoform fraction, multi-isoform genes:",
    round(median(census$top_frac[census$n_isoforms >= 2]), 3), "\n")
```

## Threshold sensitivity

The multi-isoform fraction is a function of where the cuts are put, so the chosen
thresholds are shown next to the alternatives rather than on their own.

```{r sensitivity}
threshold_grid <- expand_grid(tpm_cut = c(0.5, 1, 5), frac_cut = c(0, 0.05, 0.10, 0.20))

threshold_grid %>%
  pmap_dfr(function(tpm_cut, frac_cut) {
    g <- eligible %>%
      filter(mean_tpm >= tpm_cut, iso_frac >= frac_cut) %>%
      dplyr::count(gene_id, name = "n_iso")
    tibble(tpm_cut, frac_cut,
           genes = nrow(g),
           pct_multi = round(100 * mean(g$n_iso >= 2), 1),
           max_iso = max(g$n_iso))
  }) %>%
  knitr::kable(caption = "Multi-isoform percentage across thresholds (chosen: tpm 1, frac 0.10)")
```

# Validation

## Computational checks

```{r checks_computational}
stopifnot(
  "negative TPM"                  = all(eligible$mean_tpm >= 0),
  "isoform fraction out of [0,1]" = all(eligible$iso_frac >= 0 & eligible$iso_frac <= 1),
  "expressed set has genes"       = nrow(expressed) > 0,
  "n_isoforms below 1"            = all(census$n_isoforms >= 1),
  "missing expected columns"      = all(c("transcript_id", "transcript_id_clean", "gene_id",
                                          "gene_id_clean", "symbol", "mean_tpm", "iso_frac")
                                        %in% colnames(expressed)),
  "no TE columns should be here"  = !any(c("te_lfc", "te_padj") %in% colnames(expressed))
)

# Isoform fractions must sum to 1 within each expressed gene, over ALL eligible isoforms
frac_sums <- eligible %>%
  filter(gene_tpm > 0) %>%
  group_by(gene_id) %>%
  summarize(total = sum(iso_frac), .groups = "drop")
stopifnot("isoform fractions do not sum to 1" = all(abs(frac_sums$total - 1) < 1e-8))

cat("Computational checks passed\n")
```

## Biological spot-checks

```{r checks_biological}
# 1. Two separate things, deliberately not conflated:
#    (a) the ID join and the index. The MANE Select transcript must be PRESENT in the
#        eligible pool for essentially every gene. Anything below ~95% means versioned
#        vs unversioned IDs, a different annotation, or a salmon index mismatch.
#    (b) whether MANE is actually EXPRESSED. This is biology, not a bug: MANE Select is
#        the reference isoform, not necessarily the one this cell line runs.
mane_genes <- census %>% inner_join(mane, by = "gene_id_clean")

mane_status <- mane_genes %>%
  left_join(eligible %>% dplyr::select(gene_id_clean, transcript_id_clean) %>%
              mutate(in_pool = TRUE),
            by = c("gene_id_clean", "mane_tx_clean" = "transcript_id_clean")) %>%
  left_join(expressed %>% dplyr::select(gene_id_clean, transcript_id_clean) %>%
              mutate(is_expressed = TRUE),
            by = c("gene_id_clean", "mane_tx_clean" = "transcript_id_clean")) %>%
  group_by(gene_id_clean) %>%
  summarize(in_pool        = any(!is.na(in_pool)),
            mane_expressed = any(!is.na(is_expressed)),
            .groups = "drop")

cat(sprintf("MANE Select present in the eligible pool: %.1f%% of genes (n = %d)\n",
            100 * mean(mane_status$in_pool), nrow(mane_status)))
cat(sprintf("MANE Select actually expressed:           %.1f%% of genes (n = %d)\n",
            100 * mean(mane_status$mane_expressed), nrow(mane_status)))
stopifnot("MANE transcripts missing from the pool - ID join or index mismatch" =
            mean(mane_status$in_pool) > 0.95)
# Well above the ~35% a broken join would give, and low enough to allow the real rate,
# which is roughly 73%: in about a quarter of genes this cell line does not run MANE.
stopifnot("MANE Select rarely expressed - check the quantification" =
            mean(mane_status$mane_expressed) > 0.60)

# 2. ACTB is one of the most abundant mRNAs in these cells and is isoform-dominant.
actb <- census %>% filter(symbol == "ACTB")
cat("ACTB: gene TPM", round(actb$gene_tpm, 1),
    "| isoforms", actb$n_isoforms, "| top fraction", round(actb$top_frac, 3), "\n")
stopifnot("ACTB not highly expressed" = nrow(actb) == 1 && actb$gene_tpm > 100,
          "ACTB not isoform-dominant" = actb$top_frac > 0.5)

# 3. Genes with more ANNOTATED coding isoforms should express more of them. Tested as
#    monotonicity across annotation bins, not as a correlation magnitude: n_isoforms is
#    heavily tied at 1, which caps Spearman near 0.2 even though the trend is clean.
annot_vs_expressed <- eligible %>%
  dplyr::count(gene_id, name = "n_annotated") %>%
  inner_join(census, by = "gene_id") %>%
  mutate(annot_bin = cut(n_annotated, c(0, 1, 2, 4, 8, 16, Inf)))

bin_means <- annot_vs_expressed %>%
  group_by(annot_bin) %>%
  summarize(genes = n(), mean_expressed = mean(n_isoforms), .groups = "drop")
print(bin_means)

rho <- cor(annot_vs_expressed$n_annotated, annot_vs_expressed$n_isoforms,
           method = "spearman")
cat("Spearman(annotated coding isoforms, expressed isoforms) =", round(rho, 3),
    ", n =", nrow(annot_vs_expressed), "\n")
stopifnot("expressed isoform count does not rise with annotation" =
            all(diff(bin_means$mean_expressed) > 0),
          "no positive annotation trend" = rho > 0.1)

# 4. Isoform count also rises with expression level, which is a detection confound, not
#    biology. Asserted so the plot in section 5 cannot silently stop being true.
rho_depth <- cor(log2(census$gene_tpm), census$n_isoforms, method = "spearman")
cat("Spearman(log2 gene TPM, expressed isoforms) =", round(rho_depth, 3),
    ", n =", nrow(census), "\n")
stopifnot("expected a positive depth-isoform relationship" = rho_depth > 0.2)
```

# Are the multi-isoform genes real co-expression?

A gene can clear the "2 isoforms" bar with one dominant isoform at 92% and a minor one at
10%. The top-isoform fraction separates that case from genuine co-expression.

```{r plot_census}
p_census <- census %>%
  mutate(n_lab = if_else(n_isoforms >= 5, "5+", as.character(n_isoforms))) %>%
  dplyr::count(n_lab) %>%
  ggplot(aes(n_lab, n)) +
  geom_col(fill = "grey70", colour = "black", linewidth = 0.3) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.2) +
  labs(x = "Expressed isoforms per gene",
       y = "Genes",
       title = sprintf("Expressed coding isoforms per gene, %s %s %shr",
                       params$treatment, params$condition, params$timepoint),
       subtitle = sprintf("mean TPM >= %g and isoform fraction >= %g, n = %d genes",
                          params$tpm_min, params$frac_min, nrow(census))) +
  theme_classic(base_size = 12)
p_census
ggsave(here("plots", "95a_expressed_isoforms_per_gene.pdf"), p_census,
       width = 5.5, height = 4)
```

```{r plot_topfrac}
multi <- census %>% filter(n_isoforms >= 2)

p_top <- ggplot(multi, aes(top_frac)) +
  geom_histogram(binwidth = 0.02, fill = "grey70", colour = "black", linewidth = 0.2) +
  geom_vline(xintercept = median(multi$top_frac), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Fraction of gene TPM in the top isoform",
       y = "Genes",
       title = "Multi-isoform genes are not dominated by one isoform",
       subtitle = sprintf("median = %.2f, dominant (> 0.8) = %d of %d genes",
                          median(multi$top_frac), sum(multi$top_frac > 0.8), nrow(multi))) +
  theme_classic(base_size = 12)
p_top
ggsave(here("plots", "95b_top_isoform_fraction.pdf"), p_top, width = 5.5, height = 4)
```

```{r plot_depth}
# Detection confound: more highly expressed genes can support more detected isoforms.
p_depth <- census %>%
  mutate(n_lab = if_else(n_isoforms >= 5, "5+", as.character(n_isoforms))) %>%
  ggplot(aes(n_lab, log2(gene_tpm))) +
  geom_boxplot(fill = "grey70", outlier.size = 0.4, linewidth = 0.3) +
  labs(x = "Expressed isoforms per gene",
       y = "log2 gene TPM",
       title = "Isoform count rises with expression level",
       subtitle = "Part of the multi-isoform signal is detection power, not biology") +
  theme_classic(base_size = 12)
p_depth
ggsave(here("plots", "95c_isoform_count_vs_expression.pdf"), p_depth, width = 5.5, height = 4)
```

# Do co-expressed isoforms differ by synonymous codon choice?

Two isoforms of one gene can differ in codon usage two ways:

- **Synonymous**: same protein, different codons. Only possible where the isoforms use
  different exons that encode the same amino acids, e.g. duplicated or mutually exclusive
  exons. This is what would let a codon or GC3 difference be read as a translational
  signal with protein identity held constant.
- **Compositional**: different protein. Any codon or GC difference then partly reports
  which amino acids the alternative exon encodes, which is the confound CLAUDE.md flags
  for raw codon frequency.

Each pair of expressed isoforms within a gene is classified by comparing CDS nucleotide
sequences and their translations.

```{r cds_sequences}
multi_tx <- expressed %>% filter(gene_id %in% multi$gene_id)

cds_seqs <- extractTranscriptSeqs(BSgenome.Hsapiens.UCSC.hg38,
                                  cds_by_tx[multi_tx$transcript_id])

# GENCODE CDS annotations can be incomplete at either end; only in-frame CDS can be
# translated or split into codons.
in_frame <- width(cds_seqs) %% 3 == 0 & width(cds_seqs) >= 60
cat("Isoforms of multi-isoform genes:", length(cds_seqs),
    "| in-frame and >= 60 nt:", sum(in_frame), "\n")

cds_seqs <- cds_seqs[in_frame]
prot_seqs <- Biostrings::translate(cds_seqs, if.fuzzy.codon = "solve")
stopifnot("CDS and protein sets out of sync" = length(cds_seqs) == length(prot_seqs))
```

```{r pairs}
# All within-gene pairs of expressed, in-frame isoforms
pairs <- multi_tx %>%
  filter(transcript_id %in% names(cds_seqs)) %>%
  dplyr::select(gene_id, symbol, transcript_id) %>%
  group_by(gene_id, symbol) %>%
  filter(n() >= 2) %>%
  summarize(tx = list(transcript_id), .groups = "drop") %>%
  mutate(combo = map(tx, ~ as_tibble(t(combn(.x, 2)), .name_repair = ~ c("tx_a", "tx_b")))) %>%
  dplyr::select(gene_id, symbol, combo) %>%
  unnest(combo)

cat("Within-gene isoform pairs:", nrow(pairs), "in", n_distinct(pairs$gene_id), "genes\n")

classify_pair <- function(tx_a, tx_b) {
  cds_a <- as.character(cds_seqs[[tx_a]]); cds_b <- as.character(cds_seqs[[tx_b]])
  aa_a  <- as.character(prot_seqs[[tx_a]]); aa_b  <- as.character(prot_seqs[[tx_b]])
  short <- if (nchar(cds_a) <= nchar(cds_b)) cds_a else cds_b
  long  <- if (nchar(cds_a) <= nchar(cds_b)) cds_b else cds_a
  tibble(
    cds_identical     = cds_a == cds_b,
    protein_identical = aa_a == aa_b,
    cds_nested        = !(cds_a == cds_b) && grepl(short, long, fixed = TRUE)
  )
}

pairs <- pairs %>%
  bind_cols(map2_dfr(pairs$tx_a, pairs$tx_b, classify_pair)) %>%
  mutate(class = case_when(
    cds_identical                      ~ "identical CDS",
    protein_identical                  ~ "synonymous (same protein, different CDS)",
    cds_nested                         ~ "nested (one CDS contained in the other)",
    TRUE                               ~ "divergent protein"
  ))

pairs %>%
  dplyr::count(class, name = "n_pairs") %>%
  mutate(pct = round(100 * n_pairs / sum(n_pairs), 1)) %>%
  arrange(desc(n_pairs)) %>%
  knitr::kable(caption = "How pairs of co-expressed isoforms of the same gene differ")
```

```{r codon_distance}
# Codon usage distance per pair, on two scales:
#   codon frequency  - tracks amino-acid composition (CLAUDE.md)
#   RSCU             - normalized within each synonymous family, so it isolates wobble
codon_counts <- trinucleotideFrequency(cds_seqs, step = 3)
# trinucleotideFrequency drops names, so the transcript IDs must be put back before
# anything is looked up by ID.
rownames(codon_counts) <- names(cds_seqs)
stopifnot("codon matrix lost its transcript IDs" =
            identical(rownames(codon_counts), names(cds_seqs)))

codon_aa <- tibble(codon = names(GENETIC_CODE), aa = as.character(GENETIC_CODE)) %>%
  filter(aa != "*") %>%
  group_by(aa) %>%
  filter(n() > 1) %>%          # Met and Trp have no synonymous choice
  ungroup()

codon_long <- codon_counts %>%
  as_tibble(rownames = "transcript_id") %>%
  pivot_longer(-transcript_id, names_to = "codon", values_to = "count") %>%
  inner_join(codon_aa, by = "codon") %>%
  group_by(transcript_id) %>%
  mutate(codon_freq = count / sum(count)) %>%
  group_by(transcript_id, aa) %>%
  mutate(rscu = if (sum(count) > 0) count / mean(count) else NA_real_) %>%
  ungroup()

to_matrix <- function(value_col) {
  codon_long %>%
    dplyr::select(transcript_id, codon, all_of(value_col)) %>%
    pivot_wider(names_from = codon, values_from = all_of(value_col)) %>%
    column_to_rownames("transcript_id") %>%
    as.matrix()
}
freq_mat <- to_matrix("codon_freq")
rscu_mat <- to_matrix("rscu")

euclid <- function(mat, a, b) {
  d <- mat[a, ] - mat[b, ]
  sqrt(sum(d^2, na.rm = TRUE))
}
gc3 <- codon_counts %>%
  as_tibble(rownames = "transcript_id") %>%
  pivot_longer(-transcript_id, names_to = "codon", values_to = "count") %>%
  filter(!codon %in% c("TAA", "TAG", "TGA")) %>%
  group_by(transcript_id) %>%
  summarize(gc3 = 100 * sum(count[str_sub(codon, 3, 3) %in% c("G", "C")]) / sum(count),
            .groups = "drop")

pairs <- pairs %>%
  mutate(freq_dist = map2_dbl(tx_a, tx_b, ~ euclid(freq_mat, .x, .y)),
         rscu_dist = map2_dbl(tx_a, tx_b, ~ euclid(rscu_mat, .x, .y))) %>%
  left_join(gc3 %>% dplyr::rename(gc3_a = gc3), by = c("tx_a" = "transcript_id")) %>%
  left_join(gc3 %>% dplyr::rename(gc3_b = gc3), by = c("tx_b" = "transcript_id")) %>%
  mutate(delta_gc3 = abs(gc3_a - gc3_b))

stopifnot("GC3 outside [0,100]" = all(dplyr::between(gc3$gc3, 0, 100)),
          "negative distance"   = all(pairs$freq_dist >= 0 & pairs$rscu_dist >= 0))
```

```{r null_pairs}
# Null: pairs of isoforms drawn from DIFFERENT genes. Within-gene pairs share most of
# their exons, so they should be far more similar than this.
tx_gene <- multi_tx %>%
  filter(transcript_id %in% rownames(freq_mat)) %>%
  dplyr::select(transcript_id, gene_id)

set.seed(9)
null_pairs <- tibble(
  tx_a = sample(tx_gene$transcript_id, nrow(pairs), replace = TRUE),
  tx_b = sample(tx_gene$transcript_id, nrow(pairs), replace = TRUE)
) %>%
  left_join(tx_gene %>% dplyr::rename(gene_a = gene_id), by = c("tx_a" = "transcript_id")) %>%
  left_join(tx_gene %>% dplyr::rename(gene_b = gene_id), by = c("tx_b" = "transcript_id")) %>%
  filter(gene_a != gene_b) %>%
  mutate(freq_dist = map2_dbl(tx_a, tx_b, ~ euclid(freq_mat, .x, .y)),
         rscu_dist = map2_dbl(tx_a, tx_b, ~ euclid(rscu_mat, .x, .y))) %>%
  left_join(gc3 %>% dplyr::rename(gc3_a = gc3), by = c("tx_a" = "transcript_id")) %>%
  left_join(gc3 %>% dplyr::rename(gc3_b = gc3), by = c("tx_b" = "transcript_id")) %>%
  mutate(delta_gc3 = abs(gc3_a - gc3_b))

dist_long <- bind_rows(
  pairs      %>% mutate(set = "same gene"),
  null_pairs %>% mutate(set = "different genes")
) %>%
  dplyr::select(set, freq_dist, rscu_dist, delta_gc3) %>%
  pivot_longer(-set, names_to = "metric", values_to = "value") %>%
  mutate(metric = dplyr::recode(metric,
                         freq_dist = "codon frequency distance",
                         rscu_dist = "RSCU distance (synonymous only)",
                         delta_gc3 = "abs delta GC3 (%)"),
         set = factor(set, levels = c("same gene", "different genes")))

dist_long %>%
  group_by(metric, set) %>%
  summarize(median = round(median(value, na.rm = TRUE), 3), n = n(), .groups = "drop") %>%
  pivot_wider(names_from = set, values_from = c(median, n)) %>%
  knitr::kable(caption = "Codon usage difference between isoform pairs")

p_dist <- ggplot(dist_long, aes(value, fill = set, colour = set)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ metric, scales = "free") +
  scale_fill_manual(values = c("same gene" = "black", "different genes" = "grey70")) +
  scale_colour_manual(values = c("same gene" = "black", "different genes" = "grey70")) +
  labs(x = "Distance between the two isoforms", y = "Density",
       title = "Isoforms of one gene differ far less in codon usage than unrelated mRNAs",
       subtitle = sprintf("same gene n = %d pairs, different genes n = %d pairs",
                          nrow(pairs), nrow(null_pairs)),
       fill = NULL, colour = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")
p_dist
ggsave(here("plots", "95d_isoform_codon_distance.pdf"), p_dist, width = 8, height = 3.5)
```

The pooled within-gene medians are diluted by the identical-CDS pairs, which are exactly
zero by construction. Split by class, the pairs that actually differ in their CDS are the
only ones that carry any signal for the CDS-3'-end question.

```{r pair_class_distances}
pairs %>%
  group_by(class) %>%
  summarize(n_pairs          = n(),
            median_rscu_dist = round(median(rscu_dist), 3),
            median_abs_dGC3  = round(median(delta_gc3), 2),
            pct_dGC3_over_5  = round(100 * mean(delta_gc3 > 5), 1),
            .groups = "drop") %>%
  arrange(desc(n_pairs)) %>%
  knitr::kable(caption = "Codon usage difference by pair class")

cat("Genes with at least one CDS-differing pair:",
    n_distinct(pairs$gene_id[pairs$class != "identical CDS"]), "\n")
```

```{r synonymous_summary}
syn <- pairs %>% filter(class == "synonymous (same protein, different CDS)")
cat("Isoform pairs that are purely synonymous (identical protein, different CDS):",
    nrow(syn), sprintf("(%.2f%% of %d pairs)\n", 100 * nrow(syn) / nrow(pairs), nrow(pairs)))
if (nrow(syn) > 0) {
  syn %>%
    arrange(desc(rscu_dist)) %>%
    dplyr::select(symbol, tx_a, tx_b, rscu_dist, delta_gc3) %>%
    head(20) %>%
    knitr::kable(caption = "Genes with synonymously distinct co-expressed isoforms")
}
```

# Save

```{r save}
write_rds(expressed, here("output", "predictive_modeling",
                          sprintf("95_expressed_isoforms_%s_%s_%shr.rds",
                                  params$treatment, params$condition, params$timepoint)))
write_csv(census, here("output", "predictive_modeling",
                       sprintf("95_isoform_census_%s_%s_%shr.csv",
                               params$treatment, params$condition, params$timepoint)))
write_csv(pairs %>% dplyr::select(-starts_with("gc3_")),
          here("output", "predictive_modeling", "95_isoform_pair_classes.csv"))
cat("Saved expressed isoform table, gene census and pair classification\n")
```

# What this settles for the next step

```{r conclusion, echo=FALSE, results='asis'}
cat(sprintf("- %d of %d expressed genes (%.1f%%) run more than one coding isoform, so a within-gene contrast is available.\n",
            sum(census$n_isoforms >= 2), nrow(census), 100 * mean(census$n_isoforms >= 2)))
cat(sprintf("- Among those, the median top isoform holds %.0f%% of the gene's mRNA, so the minor isoform is not a trace.\n",
            100 * median(multi$top_frac)))
cat(sprintf("- %.1f%% of within-gene isoform pairs encode a different protein; only %.2f%% differ synonymously.\n",
            100 * mean(pairs$class == "divergent protein" | pairs$class == "nested (one CDS contained in the other)"),
            100 * mean(pairs$class == "synonymous (same protein, different CDS)")))
```
