---
title: "eIF4E RIP enrichment and TE volcano, 4E-resistant vs 4E-sensitive"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Roiuk et al. (GSE243708), reanalysed here: TE from notebook 76, eIF4E-IP from
notebook 77. Sets are the two FASTA files in `output/fasta/`.

```{r libraries}
library(tidyverse)
library(here)
```

```{r settings}
if (!isTRUE(getOption("knitr.in.progress"))) params <- list(save_plots = TRUE)
set.seed(9)

# Same two hues as notebooks 79 and 92, which describe these same sets.
class_colors <- c("4E-resistant"    = "#D55E00",
                  "4E-sensitive"    = "#0072B2",
                  "All other genes" = "grey70")

# Paper-matched pair: the muted red / blue used for factor-dependent vs
# factor-independent transcripts, with the pale tints those figures shade under
# their curves. Outline carries the hue, fill is the tint at near-full opacity,
# so the panel reads the way the published figure does. Used by the vehicle
# violin below; swap the other panels over by pointing their two scales at
# `paper_lines` / `paper_fills` as well.
paper_lines <- c("4E-resistant"    = "#A8322D",
                 "4E-sensitive"    = "#2F6CA8",
                 "All other genes" = "grey55")
paper_fills <- c("4E-resistant"    = "#F2CFC9",
                 "4E-sensitive"    = "#CFDDEC",
                 "All other genes" = "grey88")

# Shared figure styling. Text is sized for a printed panel rather than the HTML
# preview, and the n row gets its own strip of panel: `n_pad()` (for plots whose
# limits come from `coord_cartesian`) and `n_expand` (for plots that keep the data
# limits) push the lower bound down by `n_frac`, so a `y = -Inf` label lands below
# the violins instead of on top of a tail.
fig_theme <- function(base_size = 18) {
  theme_classic(base_size = base_size) +
    theme(axis.text        = element_text(size = base_size - 2, colour = "black"),
          axis.title       = element_text(size = base_size),
          legend.text      = element_text(size = base_size - 2),
          strip.text       = element_text(size = base_size - 2),
          strip.background = element_blank(),
          plot.subtitle    = element_text(size = base_size - 5, colour = "grey30"))
}

# Class on the x axis: angled labels, so the panel can stay narrow at this type size.
theme_class_x <- theme(axis.text.x = element_text(angle = 20, hjust = 1))

n_size <- 4.8   # geom_text size for the n row, ~13.5 pt
n_frac <- 0.16  # share of the y range reserved for it

n_pad <- function(lims, frac = n_frac, top = 0) {
  lims <- as.numeric(lims)
  c(lims[1] - diff(lims) * frac, lims[2] + diff(lims) * top)
}
n_expand <- scale_y_continuous(expand = expansion(mult = c(n_frac, 0.05)))
# Faceted panels also need headroom for the per-facet p, which sits top-right.
n_expand_p <- scale_y_continuous(expand = expansion(mult = c(n_frac, 0.13)))

# One Wilcoxon p per facet. The single-panel plots carry theirs in the subtitle;
# a faceted panel has one test per facet, so it is annotated in each corner from
# the same `*_tests` frame that gets written to CSV.
p_label <- function(tests) {
  dplyr::mutate(tests, label = paste("p =", format.pval(wilcox_p, digits = 3,
                                                        eps = 2.2e-16)))
}
geom_p_label <- function(tests) {
  geom_text(data = p_label(tests), aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.05, vjust = 1.4, size = 4.2,
            colour = "grey30")
}
```

# Data

```{r load}
te_all <- read_csv(here("output", "teleman_riboseq_te_all_genes.csv"),
                   show_col_types = FALSE)
pos    <- read_csv(here("output", "genesets", "teleman_4e_resistant_TE_1hr_lfc0.5.csv"),
                   show_col_types = FALSE)
neg    <- read_csv(here("output", "genesets", "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"),
                   show_col_types = FALSE)
fm     <- readRDS(here("output", "predictive_modeling", "feature_matrix_teleman_4e.rds"))

ip <- read_csv(here("output", "teleman_eif4e_ip_vehicle_corrected.csv"),
               show_col_types = FALSE) %>%
  dplyr::select(gene_id, symbol, vehicle = log2FoldChange) %>%
  inner_join(read_csv(here("output", "teleman_eif4e_ip_inhibitor_corrected.csv"),
                      show_col_types = FALSE) %>%
               dplyr::select(gene_id, inhibitor = log2FoldChange), by = "gene_id") %>%
  inner_join(read_csv(here("output", "teleman_eif4e_ip_interaction_corrected.csv"),
                      show_col_types = FALSE) %>%
               dplyr::select(gene_id, interaction = log2FoldChange,
                             interaction_padj = padj), by = "gene_id")

# The *original* Teleman tables carried 12 PAR duplicate gene_id keys; nb76 and nb77
# both removed them, so all four files read here are already one row per gene. The guard
# stays because a duplicated key in any of them multiplies rows on join and silently
# inflates every n, median and Wilcoxon p below -- and it is guarded on all three joined
# keys, not just `ip`: `te_all` is inner_joined to `ip` for the Fig 4E rho, and `classes`
# is left_joined into every IP, GC and volcano statistic. nb81 guards `ip` the same way.
stopifnot("ip gene_id must be unique"     = !any(duplicated(ip$gene_id)),
          "te_all gene_id must be unique" = !any(duplicated(te_all$gene_id)))

# Gene -> class, from the FASTA headers rather than the CSVs, so the figures describe
# exactly the sequences that were sent out.
classes <- fm %>%
  dplyr::select(gene_id_clean, transcript_id_clean) %>%
  mutate(class = case_when(
    transcript_id_clean %in% sub("^>", "", grep("^>", readLines(
      here("output", "fasta", "teleman_4e_resistant_transcripts.fa")), value = TRUE)) ~ "4E-resistant",
    transcript_id_clean %in% sub("^>", "", grep("^>", readLines(
      here("output", "fasta", "teleman_4e_sensitive_transcripts.fa")), value = TRUE)) ~ "4E-sensitive")) %>%
  filter(!is.na(class)) %>%
  dplyr::select(gene_id = gene_id_clean, class)

# One row per gene: `fm` is gene-deduplicated upstream, but `classes` is left_joined into
# every statistic below, so assert it here rather than trusting that.
stopifnot("classes gene_id must be unique" = !any(duplicated(classes$gene_id)))

print(dplyr::count(classes, class) %>% as.data.frame(), row.names = FALSE)
```

```{r checks}
stopifnot(
  "FASTA sets do not match the gene set CSVs" =
    setequal(classes$gene_id[classes$class == "4E-resistant"], pos$ensembl_gene) &&
    setequal(classes$gene_id[classes$class == "4E-sensitive"], neg$ensembl_gene),
  "class direction flipped" =
    median(pos$log2FoldChange) > 0 && median(neg$log2FoldChange) < 0,
  "CDKN2B not resistant / GLO1 not sensitive" =
    "CDKN2B" %in% pos$db_gene_symbol && "GLO1" %in% neg$db_gene_symbol,
  "padj outside [0,1]" = all(te_all$padj >= 0 & te_all$padj <= 1, na.rm = TRUE),
  # is.finite(NA) is FALSE, not NA, so na.rm would not skip NAs here -- drop them
  # explicitly or the check aborts on any filtered gene.
  "non-finite log2FC" =
    all(is.finite(te_all$log2FoldChange[!is.na(te_all$log2FoldChange)]))
)

# Fig 4E: TE change and change in eIF4E binding are negatively correlated.
rho <- te_all %>%
  inner_join(ip, by = "gene_id") %>%
  summarise(r = cor(log2FoldChange, interaction, method = "spearman",
                    use = "complete.obs"), n = n())
print(as.data.frame(rho), digits = 3, row.names = FALSE)
stopifnot("Fig 4E not reproduced" = rho$r < -0.15)
```

# eIF4E RIP enrichment

## The genes the paper names as losing eIF4E binding

Integrity check on the IP arm at gene level. Roiuk et al. write that "the 'resistant'
mRNAs (shown in yellow) such as RPP25 or CDKN2B were amongst the mRNAs that showed the
largest drop in eIF4E binding upon 4E-BP-4A expression (Fig. 4B, C)", and validate eight
transcripts by Q-RT-PCR on eIF4E IPs in Fig. 4D. Those eight are the labelled points in
Fig. 4B. Their published E.V.-versus-4E-BP1-4A adjusted p-values, transcribed from the
Fig. 4 sheet of the article's Source Data, are the `paper_*` columns below. (The paper
writes ACTB as "ATCB" in that sheet.)

The comparison is not like-for-like: Fig. 4D is Q-RT-PCR normalised to total cellular
mRNA, while `interaction` is the DESeq2 IP/input interaction from notebook 77. Direction
and rank are the meaningful part, not the p-value.

```{r paper-named-genes}
# Roiuk et al. 2024, Nat Commun 15:6692, Fig. 4D source data. paper_call is the
# direction the paper reports for eIF4E binding on 4E-BP1-4A expression.
paper_ip <- tibble::tribble(
  ~symbol,   ~paper_call, ~paper_padj,
  "CDKN2B",  "down",      4.772e-3,
  "RPP25",   "down",      4.772e-3,
  "RPL13A",  "down",      2.730e-2,
  "WARS",    "down",      8.318e-3,
  "TIA1",    "down",      2.730e-2,
  "GAPDH",   "ns",        3.129e-1,
  "ACTB",    "ns",        3.202e-1,
  "GLO1",    "ns",        3.129e-1)

named <- ip %>%
  mutate(pct = 100 * rank(interaction) / n()) %>%
  inner_join(paper_ip, by = "symbol") %>%
  arrange(interaction) %>%
  dplyr::select(symbol, paper_call, paper_padj,
                interaction, padj = interaction_padj, pct)

# inner_join would drop a gene silently, and every assertion below indexes by symbol,
# so a missing row would make its check vacuous rather than failing.
stopifnot("a paper-named gene is missing from the IP table" =
            nrow(named) == nrow(paper_ip))
print(as.data.frame(named), digits = 3, row.names = FALSE)

val <- function(sym, col) named[[col]][match(sym, named$symbol)]

stopifnot(
  # The paper's own text claim: these two are among the largest losses transcriptome-wide.
  "RPP25 / CDKN2B are not among the largest losses of eIF4E binding" =
    all(val(c("RPP25", "CDKN2B"), "pct") < 5) &&
    all(val(c("RPP25", "CDKN2B"), "padj") < 0.05),
  # padj here is 0.043, close enough to the cut that the direction is the durable claim.
  "RPL13A does not lose eIF4E binding" = val("RPL13A", "interaction") < 0,
  "a gene the paper calls unchanged is significant here" =
    all(val(c("GAPDH", "ACTB", "GLO1"), "padj") > 0.05))

cat("\nWARS and TIA1 do not reproduce: WARS", round(val("WARS", "interaction"), 3),
    "(padj", signif(val("WARS", "padj"), 3), "), TIA1",
    round(val("TIA1", "interaction"), 3),
    "(padj", signif(val("TIA1", "padj"), 3), ")\n")
```

**Six of the eight reproduce, and the two that do not fail the same way in the authors'
own output.** RPP25 and CDKN2B, the two named in the text, sit in the bottom 1.5% and 3.8%
of the transcriptome-wide change in eIF4E binding, both significant, which is what
"amongst the mRNAs that showed the largest drop" asserts. RPL13A is down. GAPDH, ACTB and
GLO1 are flat, as the paper reports.

WARS and TIA1 are called significantly down by the Fig. 4D Q-RT-PCR but are at zero here
(-0.04 and +0.02, padj 0.89 and 0.97). This is **not** a defect of the notebook 77
re-run: the authors' own uncorrected DESeq2 table for this contrast,
`mpra_preliminary/res_eif4e_ip_inhib_vs_vehicle.csv`, gives -0.031 (padj 0.92) and +0.021
(padj 0.94) for the same two genes, agreeing with the values here to three decimals. The
divergence is between the paper's Q-RT-PCR and the paper's own RNA-seq, and it exists in
the published data. Both genes are highly expressed (baseMean 31,000 and 3,000), so it is
not a coverage problem.

## Class distributions

`vehicle` and `inhibitor` are separate DESeq2 fits, so the offset between the two
conditions is not anchored. The paired quantity is `interaction`.

The IP arm has its own expression filter, and the sensitive set is the more highly
expressed of the two, so the classes are not equally covered here.

```{r ip-long}
ip_long <- ip %>%
  # padj belongs to the interaction fit alone; carrying it into the long frame would
  # stamp it onto rows labelled Vehicle and Inhibited, where it is not that fit's padj.
  dplyr::select(-interaction_padj) %>%
  left_join(classes, by = "gene_id") %>%
  mutate(class = factor(replace_na(class, "All other genes"),
                        levels = names(class_colors))) %>%
  pivot_longer(c(vehicle, inhibitor, interaction),
               names_to = "contrast", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(contrast = factor(contrast, levels = c("vehicle", "inhibitor", "interaction"),
                           labels = c("Vehicle (E.V.)", "eIF4E inhibited (4E-BP1-4A)",
                                      "Inhibited / vehicle")))

ip_stats <- ip_long %>%
  group_by(contrast, class) %>%
  summarise(n = n(), median = median(value), .groups = "drop")
print(as.data.frame(ip_stats), digits = 3, row.names = FALSE)

ip_tests <- ip_long %>%
  filter(class != "All other genes") %>%
  group_by(contrast) %>%
  summarise(wilcox_p = wilcox.test(value ~ class)$p.value, .groups = "drop")
print(as.data.frame(ip_tests), digits = 3, row.names = FALSE)

classes %>%
  mutate(measured = gene_id %in% ip$gene_id) %>%
  group_by(class) %>%
  summarise(n_set = n(), n_with_ip = sum(measured),
            pct = round(100 * mean(measured), 1), .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)

# Fig 4C: resistant mRNAs lose eIF4E binding relative to the bulk.
stopifnot("Fig 4C not reproduced" =
  ip_stats$median[ip_stats$contrast == "Inhibited / vehicle" &
                  ip_stats$class == "4E-resistant"] <
  median(ip$interaction, na.rm = TRUE))

if (params$save_plots) write_csv(left_join(ip_stats, ip_tests, by = "contrast"),
                                 here("output", "96_eif4e_rip_enrichment_stats.csv"))
```

## Both conditions together

Shown for the two levels side by side, **not** to read a change off. `vehicle` and
`inhibitor` are separate DESeq2 fits with an unanchored offset, so differencing them is not
the same quantity as the interaction: by class median it gives resistant -0.240 /
sensitive +0.261 against the paired interaction's -0.117 / +0.177, overstating the
resistant shift about twofold. The change is the next-but-two section, off `interaction`.

```{r violin-both, fig.width = 9, fig.height = 5.5}
both     <- filter(ip_long,  contrast != "Inhibited / vehicle")
both_lab <- filter(ip_stats, contrast != "Inhibited / vehicle")

# Faceted rather than dodged: at this type size three dodged `n = ` labels ran
# into each other inside one x slot.
p_both <- ggplot(both, aes(class, value, fill = class, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = both_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  geom_p_label(filter(ip_tests, contrast != "Inhibited / vehicle")) +
  facet_wrap(~ contrast, nrow = 1) +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  coord_cartesian(ylim = n_pad(quantile(both$value, c(0.001, 0.999)), top = 0.12)) +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive, per panel.",
                        "Separate fits, unanchored offset: levels only, not a change.",
                        "\nFor the change in binding see the interaction contrast."),
       x = NULL, y = "log2 (eIF4E IP / input)") +
  fig_theme() +
  theme_class_x

print(p_both)
if (params$save_plots) ggsave(here("plots", "96_eif4e_rip_enrichment_violin.pdf"),
                              p_both, width = 9, height = 5.5)
```

```{r ip-density-helper}
# One density panel per contrast: the three classes, their medians, and the
# resistant-vs-sensitive Wilcoxon in the subtitle.
ip_density_plot <- function(contrast_name, xlab) {
  d   <- filter(ip_long,  contrast == contrast_name)
  lab <- filter(ip_stats, contrast == contrast_name) %>% arrange(class)

  ggplot(d, aes(value, colour = class)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    geom_density(fill = NA, linewidth = 0.9) +
    geom_vline(data = lab, aes(xintercept = median, colour = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    scale_colour_manual(values = class_colors, name = NULL,
                        labels = paste0(lab$class, " (n = ", lab$n, ")")) +
    coord_cartesian(xlim = quantile(d$value, c(0.005, 0.995))) +
    labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                          format.pval(ip_tests$wilcox_p[ip_tests$contrast == contrast_name],
                                      digits = 3, eps = 2.2e-16)),
         x = xlab, y = "Density") +
    fig_theme() +
    theme(legend.position = "top")
}
```

## eIF4E IP / input, vehicle

```{r violin-vehicle, fig.width = 4.6, fig.height = 5.5}
veh_d   <- filter(ip_long,  contrast == "Vehicle (E.V.)")
veh_lab <- filter(ip_stats, contrast == "Vehicle (E.V.)")

p_vehicle <- ggplot(veh_d, aes(class, value, fill = class, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_violin(alpha = 0.85, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = veh_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = paper_fills, guide = "none") +
  scale_colour_manual(values = paper_lines, guide = "none") +
  coord_cartesian(ylim = n_pad(quantile(veh_d$value, c(0.001, 0.999)))) +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(ip_tests$wilcox_p[ip_tests$contrast == "Vehicle (E.V.)"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "log2 (eIF4E IP / input), vehicle") +
  fig_theme() +
  theme_class_x

print(p_vehicle)
if (params$save_plots) ggsave(here("plots", "96_eif4e_ip_vehicle_violin.pdf"),
                              p_vehicle, width = 4.6, height = 5.5)
```

```{r density-vehicle, fig.width = 7.5, fig.height = 5.5}
p_vehicle_density <- ip_density_plot("Vehicle (E.V.)",
                                     "log2( eIF4E IP / input ), E.V. cells")
print(p_vehicle_density)
if (params$save_plots) ggsave(here("plots", "96_eif4e_ip_vehicle_density.pdf"),
                              p_vehicle_density, width = 7.5, height = 5.5)
```

## eIF4E IP / input, 4E-BP1-4A

```{r violin-inhibitor, fig.width = 4.6, fig.height = 5.5}
inh_d   <- filter(ip_long,  contrast == "eIF4E inhibited (4E-BP1-4A)")
inh_lab <- filter(ip_stats, contrast == "eIF4E inhibited (4E-BP1-4A)")

p_inhibitor <- ggplot(inh_d, aes(class, value, fill = class, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = inh_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  coord_cartesian(ylim = n_pad(quantile(inh_d$value, c(0.001, 0.999)))) +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(ip_tests$wilcox_p[ip_tests$contrast == "eIF4E inhibited (4E-BP1-4A)"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "log2 (eIF4E IP / input), 4E-BP1-4A") +
  fig_theme() +
  theme_class_x

print(p_inhibitor)
if (params$save_plots) ggsave(here("plots", "96_eif4e_ip_inhibitor_violin.pdf"),
                              p_inhibitor, width = 4.6, height = 5.5)
```

```{r density-inhibitor, fig.width = 7.5, fig.height = 5.5}
p_inhibitor_density <- ip_density_plot("eIF4E inhibited (4E-BP1-4A)",
                                       "log2( eIF4E IP / input ), 4E-BP1-4A cells")
print(p_inhibitor_density)
if (params$save_plots) ggsave(here("plots", "96_eif4e_ip_inhibitor_density.pdf"),
                              p_inhibitor_density, width = 7.5, height = 5.5)
```

## Change in binding: 4E-BP1-4A over vehicle

```{r violin-interaction, fig.width = 4.6, fig.height = 5.5}
int_d   <- filter(ip_long,  contrast == "Inhibited / vehicle")
int_lab <- filter(ip_stats, contrast == "Inhibited / vehicle")

p_interaction <- ggplot(int_d, aes(class, value, fill = class, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = int_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  coord_cartesian(ylim = n_pad(quantile(int_d$value, c(0.001, 0.999)))) +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(ip_tests$wilcox_p[ip_tests$contrast == "Inhibited / vehicle"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "log2( 4E-BP1-4A IP/input  /  E.V. IP/input )") +
  fig_theme() +
  theme_class_x

print(p_interaction)
if (params$save_plots) ggsave(here("plots", "96_eif4e_binding_change_violin.pdf"),
                              p_interaction, width = 4.6, height = 5.5)
```

```{r density-interaction, fig.width = 7.5, fig.height = 5.5}
p_interaction_density <- ip_density_plot(
  "Inhibited / vehicle", "log2( 4E-BP1-4A IP/input  /  E.V. IP/input )")
print(p_interaction_density)
if (params$save_plots) ggsave(here("plots", "96_eif4e_binding_change_density.pdf"),
                              p_interaction_density, width = 7.5, height = 5.5)
```

# TE volcano

Positive log2FC = TE held up when eIF4E is inhibited. Coloured points are the FASTA sets,
cut at padj < 0.05 and |log2FC| > 0.5. That cut leaves 1,258 resistant and 1,195 sensitive
genes, and nb76 caps each class at `N_PER_CLASS = 1,195` -- the sensitive class's own size,
which is what size-matches the two sets. **Only the resistant class is actually truncated.**
So the two colour boundaries have different meanings: the sensitive one is the 0.5 threshold
itself, the resistant one is the truncation, and the 63 genes with 0.5 < log2FC < 0.654 are
resistant by the stated cut but plotted grey. Dashed lines are the paper's Fig 1A thresholds,
for reference.

```{r volcano, fig.width = 8, fig.height = 6.5}
volc <- te_all %>%
  filter(!is.na(padj)) %>%
  left_join(classes, by = "gene_id") %>%
  mutate(class = factor(replace_na(class, "All other genes"),
                        levels = names(class_colors)))

print(dplyr::count(volc, class) %>% as.data.frame(), row.names = FALSE)
stopifnot(
  # all(logical(0)) is TRUE, so the sign checks below pass vacuously if the class
  # join drops out (e.g. a versioned gene_id). Assert the sets landed first.
  "class join did not reach the volcano" =
    sum(volc$class == "4E-resistant") == nrow(pos) &&
    sum(volc$class == "4E-sensitive") == nrow(neg),
  "resistant points are not on the positive side" =
    all(volc$log2FoldChange[volc$class == "4E-resistant"] > 0),
  "sensitive points are not on the negative side" =
    all(volc$log2FoldChange[volc$class == "4E-sensitive"] < 0)
)

# Where each colour boundary actually sits, and which class the 1,195 cap bound. Computed
# rather than quoted, so the caption above cannot drift from the files.
passing <- volc %>%
  dplyr::filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) > 0.5) %>%
  dplyr::mutate(side = dplyr::if_else(log2FoldChange > 0, "resistant", "sensitive")) %>%
  dplyr::count(side, name = "n_passing_cut")

dplyr::tibble(
  side          = c("resistant", "sensitive"),
  n_in_set      = c(sum(volc$class == "4E-resistant"), sum(volc$class == "4E-sensitive")),
  colour_bound  = c(min(volc$log2FoldChange[volc$class == "4E-resistant"]),
                    max(volc$log2FoldChange[volc$class == "4E-sensitive"]))) %>%
  dplyr::left_join(passing, by = "side") %>%
  dplyr::mutate(truncated_by_cap = n_passing_cut > n_in_set,
                n_passing_but_grey = n_passing_cut - n_in_set) %>%
  as.data.frame() %>% print(digits = 3, row.names = FALSE)

p_volcano <- ggplot(volc, aes(log2FoldChange, -log10(padj))) +
  geom_point(data = filter(volc, class == "All other genes"),
             colour = "grey70", size = 0.5, alpha = 0.4) +
  geom_point(data = filter(volc, class != "All other genes"),
             aes(colour = class), size = 0.6, alpha = 0.75) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40",
             linewidth = 0.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40",
             linewidth = 0.4) +
  scale_colour_manual(values = class_colors, breaks = names(class_colors)[1:2],
                      name = NULL) +
  labs(subtitle = paste0(sum(volc$class == "4E-resistant"), " resistant, ",
                         sum(volc$class == "4E-sensitive"), " sensitive, n = ",
                         scales::comma(nrow(volc)), " genes tested"),
       x = "log2( TE in 4E-BP1-4A / TE in E.V. )", y = "-log10(adj. p-value)") +
  fig_theme() +
  theme(legend.position = "top")

print(p_volcano)
if (params$save_plots) ggsave(here("plots", "96_te_volcano_fasta_sets.pdf"),
                              p_volcano, width = 8, height = 6.5)
```

```{r volcano-labelled, fig.width = 8, fig.height = 6.5}
p_volcano_labelled <- p_volcano +
  ggrepel::geom_text_repel(data = filter(volc, symbol %in% c("CDKN2B", "RPP25", "GLO1")),
                           aes(label = symbol), size = 5.5, min.segment.length = 0,
                           segment.size = 0.3)

print(p_volcano_labelled)
if (params$save_plots) ggsave(here("plots", "96_te_volcano_fasta_sets_labelled.pdf"),
                              p_volcano_labelled, width = 8, height = 6.5)
```

# GC content by region

```{r gc-long}
gc_long <- fm %>%
  dplyr::select(gene_id = gene_id_clean, utr5_gc, cds_gc, utr3_gc) %>%
  left_join(classes, by = "gene_id") %>%
  mutate(class = factor(replace_na(class, "All other genes"),
                        levels = names(class_colors))) %>%
  pivot_longer(c(utr5_gc, cds_gc, utr3_gc), names_to = "region", values_to = "gc") %>%
  filter(!is.na(gc)) %>%
  mutate(region = factor(region, levels = c("utr5_gc", "cds_gc", "utr3_gc"),
                         labels = c("5'UTR", "CDS", "3'UTR")))

stopifnot("GC outside [0, 100]" = all(gc_long$gc >= 0 & gc_long$gc <= 100))

gc_stats <- gc_long %>%
  group_by(region, class) %>%
  summarise(n = n(), median = median(gc), .groups = "drop")
print(as.data.frame(gc_stats), digits = 3, row.names = FALSE)

gc_tests <- gc_long %>%
  filter(class != "All other genes") %>%
  group_by(region) %>%
  summarise(wilcox_p = wilcox.test(gc ~ class)$p.value, .groups = "drop")
print(as.data.frame(gc_tests), digits = 3, row.names = FALSE)

if (params$save_plots) write_csv(left_join(gc_stats, gc_tests, by = "region"),
                                 here("output", "96_gc_by_region_stats.csv"))
```

## All three regions together

```{r gc-violin-all, fig.width = 9, fig.height = 5.5}
p_gc_all <- ggplot(gc_long, aes(class, gc, fill = class, colour = class)) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = gc_stats, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  geom_p_label(gc_tests) +
  facet_wrap(~ region, nrow = 1) +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  n_expand_p +
  labs(subtitle = "Wilcoxon, resistant vs sensitive, per region",
       x = NULL, y = "GC content (%)") +
  fig_theme() +
  theme_class_x

print(p_gc_all)
if (params$save_plots) ggsave(here("plots", "96_gc_by_region_violin.pdf"),
                              p_gc_all, width = 9, height = 5.5)
```

## 5'UTR

```{r gc-violin-utr5, fig.width = 4.6, fig.height = 5.5}
utr5_d   <- filter(gc_long,  region == "5'UTR")
utr5_lab <- filter(gc_stats, region == "5'UTR")

p_gc_utr5 <- ggplot(utr5_d, aes(class, gc, fill = class, colour = class)) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = utr5_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  n_expand +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(gc_tests$wilcox_p[gc_tests$region == "5'UTR"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "5'UTR GC content (%)") +
  fig_theme() +
  theme_class_x

print(p_gc_utr5)
if (params$save_plots) ggsave(here("plots", "96_gc_utr5_violin.pdf"),
                              p_gc_utr5, width = 4.6, height = 5.5)
```

## CDS

```{r gc-violin-cds, fig.width = 4.6, fig.height = 5.5}
cds_d   <- filter(gc_long,  region == "CDS")
cds_lab <- filter(gc_stats, region == "CDS")

p_gc_cds <- ggplot(cds_d, aes(class, gc, fill = class, colour = class)) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = cds_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  n_expand +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(gc_tests$wilcox_p[gc_tests$region == "CDS"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "CDS GC content (%)") +
  fig_theme() +
  theme_class_x

print(p_gc_cds)
if (params$save_plots) ggsave(here("plots", "96_gc_cds_violin.pdf"),
                              p_gc_cds, width = 4.6, height = 5.5)
```

## 3'UTR

```{r gc-violin-utr3, fig.width = 4.6, fig.height = 5.5}
utr3_d   <- filter(gc_long,  region == "3'UTR")
utr3_lab <- filter(gc_stats, region == "3'UTR")

p_gc_utr3 <- ggplot(utr3_d, aes(class, gc, fill = class, colour = class)) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = utr3_lab, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  n_expand +
  labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                        format.pval(gc_tests$wilcox_p[gc_tests$region == "3'UTR"],
                                    digits = 3, eps = 2.2e-16)),
       x = NULL, y = "3'UTR GC content (%)") +
  fig_theme() +
  theme_class_x

print(p_gc_utr3)
if (params$save_plots) ggsave(here("plots", "96_gc_utr3_violin.pdf"),
                              p_gc_utr3, width = 4.6, height = 5.5)
```

# GC by codon position in the CDS

```{r gc-position, fig.width = 9.5, fig.height = 5.5}
pos_long <- fm %>%
  dplyr::select(gene_id = gene_id_clean, gc1, gc2, gc3) %>%
  left_join(classes, by = "gene_id") %>%
  mutate(class = factor(replace_na(class, "All other genes"),
                        levels = names(class_colors))) %>%
  pivot_longer(c(gc1, gc2, gc3), names_to = "position", values_to = "gc") %>%
  filter(!is.na(gc)) %>%
  mutate(position = factor(position, levels = c("gc1", "gc2", "gc3"),
                           labels = c("GC1", "GC2", "GC3")))

stopifnot("GC outside [0, 100]" = all(pos_long$gc >= 0 & pos_long$gc <= 100))

pos_stats <- pos_long %>%
  group_by(position, class) %>%
  summarise(n = n(), median = median(gc), .groups = "drop")
print(as.data.frame(pos_stats), digits = 3, row.names = FALSE)

pos_tests <- pos_long %>%
  filter(class != "All other genes") %>%
  group_by(position) %>%
  summarise(wilcox_p = wilcox.test(gc ~ class)$p.value, .groups = "drop")
print(as.data.frame(pos_tests), digits = 3, row.names = FALSE)

if (params$save_plots) write_csv(left_join(pos_stats, pos_tests, by = "position"),
                                 here("output", "96_gc_by_codon_position_stats.csv"))

p_gc_position <- ggplot(pos_long, aes(class, gc, fill = class, colour = class)) +
  geom_violin(alpha = 0.15, linewidth = 0.8, width = 0.75) +
  geom_boxplot(width = 0.1, alpha = 0, linewidth = 0.5, outlier.shape = NA) +
  geom_text(data = pos_stats, aes(y = -Inf, label = paste0("n = ", n)),
            vjust = -0.6, size = n_size, colour = "grey30") +
  geom_p_label(pos_tests) +
  facet_wrap(~ position, nrow = 1) +
  scale_fill_manual(values = class_colors, guide = "none") +
  scale_colour_manual(values = class_colors, guide = "none") +
  n_expand_p +
  labs(subtitle = "Wilcoxon, resistant vs sensitive, per codon position",
       x = NULL, y = "GC content (%)") +
  fig_theme() +
  theme_class_x

print(p_gc_position)
if (params$save_plots) ggsave(here("plots", "96_gc_codon_position_violin.pdf"),
                              p_gc_position, width = 9.5, height = 5.5)
```

```{r session}
sessionInfo()
```
