---
title: "Absolute TE by 4E-resistant / 4E-sensitive class"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
params:
  mpra_repo: "/Users/katematlin/github/mpra_preliminary"
  refit: FALSE
  save_plots: TRUE
editor_options:
  chunk_output_type: console
---

```{r}
if (exists("params") && !isTRUE(getOption("knitr.in.progress"))) rm(params)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

nb96 showed the Teleman 4E-resistant and 4E-sensitive sets separate on eIF4E binding and
GC. The open question is where they sit on the **TE level** axis: are 4E-resistant mRNAs
well translated to begin with, and where do they land once eIF4E is inhibited?

The repo cannot answer that today. The only Teleman TE quantity that exists is
`te_all$log2FoldChange` from nb76, the interaction coefficient
`treatmentBP_inhib.typeribo` -- a TE **change**. The three classes are defined by
thresholding exactly that column, so its distribution by class is three non-overlapping
blocks that restate the cut.

So this notebook computes **absolute TE, `log2(ribo/rna)`, separately in E.V. and in
4E-BP1-4A cells**, following the pattern in `code/individual_te_per_sample.Rmd` (the only
place in the project where a TE *level* rather than a *change* is computed).

**Read section 5 before quoting anything from section 4.** The classes are defined by
`TE_BP - TE_EV`, so selecting on that difference splits the two levels in opposite
directions on its own. Section 5 measures how much of section 4 is that artifact. The
short version: the 4E-BP1-4A separation is real and large, the baseline separation is
mostly artifact.

```{r libraries}
library(tidyverse)
library(here)
library(DESeq2)
library(janitor)
```

```{r settings}
if (!isTRUE(getOption("knitr.in.progress"))) {
  params <- list(mpra_repo = "/Users/katematlin/github/mpra_preliminary",
                 refit = FALSE, save_plots = TRUE)
}
set.seed(9)

MPRA_REPO <- params$mpra_repo
FIT_RDS   <- here("output", "predictive_modeling", "teleman_absolute_te.rds")

# Same hues as nb96 and nb79, which describe these same sets. Okabe-Ito vermillion and
# dark blue; eIF3d amber and eIF3e sky blue stay reserved for those factors.
class_colors <- c("4E-resistant"    = "#D55E00",
                  "4E-sensitive"    = "#0072B2",
                  "All other genes" = "grey70")

# The eight riboseq-arm libraries of GSE243708, named rather than indexed (nb76). The GEO
# header is case-inconsistent (4EBP1_4a_1 vs 4EBP1_4A_2); the regexes below are
# case-insensitive to that by construction.
RIBO_COLS <- c(
  "Riboseq_Empty_vector_1", "Riboseq_Empty_vector_2",
  "Riboseq_4EBP1_4a_1", "Riboseq_4EBP1_4A_2",
  "Total_RNA_Riboseq_Empty_vector_1", "Total_RNA_Riboseq_Empty_vector_2",
  "Total_RNA_Riboseq_4EBP1_4A_1", "Total_RNA_Riboseq_4EBP1_4A_2"
)
```

# 1. Compute absolute TE for the two conditions

The gene universe is taken from nb76's output rather than re-deriving its protein-coding
and `log2(rowSums) > 10` filters. That guarantees this notebook covers exactly the genes
nb76 tested and nb96 plots, and it is far less code.

GSE243708 carries no spike-ins, so size factors are plain DESeq2 defaults -- the
`controlGenes` and spike-removal steps in `individual_te_per_sample.Rmd` do not apply.

```{r fit}
te_all <- read_csv(here("output", "teleman_riboseq_te_all_genes.csv"), show_col_types = FALSE)

fit_absolute_te <- function() {
  raw <- read_csv(file.path(MPRA_REPO, "data/GSE243708_raw_counts_matrix_Roiuk.csv"),
                  show_col_types = FALSE)
  pcr <- read_csv(file.path(MPRA_REPO, "data/pcr_artifacts_sup8.csv"), show_col_types = FALSE)

  cts <- raw %>%
    dplyr::filter(!`Transcript stable ID` %in% pcr$Transcript_id) %>%
    dplyr::select(`Gene stable ID`, all_of(RIBO_COLS)) %>%
    dplyr::group_by(`Gene stable ID`) %>%
    dplyr::summarise(across(everything(), \(x) sum(x, na.rm = TRUE)), .groups = "drop") %>%
    clean_names() %>%
    column_to_rownames("gene_stable_id")

  stopifnot("nb76 universe is not fully present in the count matrix" =
              all(te_all$gene_id %in% rownames(cts)))
  cts <- cts[te_all$gene_id, , drop = FALSE]

  meta <- tibble::tibble(sample_name = RIBO_COLS) %>%
    dplyr::mutate(
      treatment = if_else(str_detect(sample_name, "4EBP"), "BP_inhib", "vehicle"),
      library   = if_else(str_detect(sample_name, "Total_RNA"), "rna", "ribo"),
      rep       = if_else(str_detect(sample_name, "_1$"), "1", "2"),
      group     = factor(paste(treatment, library, sep = "_"))
    ) %>%
    as.data.frame()
  rownames(meta) <- make_clean_names(meta$sample_name)

  stopifnot("metadata rows do not match count matrix columns" =
              identical(rownames(meta), colnames(cts)))
  stopifnot("expected 2 replicates in each of 4 groups" = all(table(meta$group) == 2))

  dds <- DESeq(DESeqDataSetFromMatrix(round(cts), colData = meta, design = ~ group),
               quiet = TRUE)

  # TE level = ribo vs rna WITHIN one condition. Written ribo-first so positive means more
  # footprints per mRNA, i.e. better translated. ashr rather than apeglm because an
  # arbitrary contrast is not a named model coefficient (see individual_te_per_sample.Rmd).
  te_of <- function(treat) {
    ribo <- paste0(treat, "_ribo"); rna <- paste0(treat, "_rna")
    res <- results(dds, contrast = c("group", ribo, rna),
                   independentFiltering = FALSE, cooksCutoff = FALSE)
    lfcShrink(dds, contrast = c("group", ribo, rna), res = res, type = "ashr",
              quiet = TRUE) %>%
      as.data.frame() %>%
      rownames_to_column("gene_id")
  }

  list(vehicle  = te_of("vehicle"),
       inhibitor = te_of("BP_inhib"),
       norm = counts(dds, normalized = TRUE),
       meta = meta,
       size_factors = sizeFactors(dds))
}

if (params$refit || !file.exists(FIT_RDS)) {
  fit <- fit_absolute_te()
  saveRDS(fit, FIT_RDS)
} else {
  fit <- readRDS(FIT_RDS)
  cat("loaded cached fit from", basename(FIT_RDS), "\n")
}

print(fit$meta[, c("treatment", "library", "rep", "group")])
cat("\nsize factors:\n"); print(round(fit$size_factors, 3))
```

# 2. The same TE the naive way, and the correlation between them

The DESeq2 contrast is one estimate of `log2(ribo/rna)`; the direct arithmetic is another.
Computing both checks the contrast is written ribo-first and that the `group` levels map to
the libraries intended. If these disagree the DESeq version is wrong, not subtle.

The `+ 1` pseudocount is on normalized counts. It is what keeps a gene with zero RNA in one
condition finite, and those genes are exactly where the two methods diverge.

```{r naive}
grp_cols <- function(g) rownames(fit$meta)[fit$meta$group == g]

naive <- tibble::tibble(
  gene_id  = rownames(fit$norm),
  ev_naive = log2((rowMeans(fit$norm[, grp_cols("vehicle_ribo"),  drop = FALSE]) + 1) /
                  (rowMeans(fit$norm[, grp_cols("vehicle_rna"),   drop = FALSE]) + 1)),
  bp_naive = log2((rowMeans(fit$norm[, grp_cols("BP_inhib_ribo"), drop = FALSE]) + 1) /
                  (rowMeans(fit$norm[, grp_cols("BP_inhib_rna"),  drop = FALSE]) + 1))
)

te <- fit$vehicle %>%
  dplyr::select(gene_id, ev = log2FoldChange, lfcSE_ev = lfcSE,
                padj_ev = padj, baseMean = baseMean) %>%
  dplyr::inner_join(fit$inhibitor %>%
                      dplyr::select(gene_id, bp = log2FoldChange, lfcSE_bp = lfcSE,
                                    padj_bp = padj), by = "gene_id") %>%
  dplyr::inner_join(naive, by = "gene_id") %>%
  dplyr::inner_join(te_all %>% dplyr::select(gene_id, symbol,
                                             nb76_interaction = log2FoldChange),
                    by = "gene_id") %>%
  dplyr::mutate(delta = bp - ev)

agreement <- purrr::map_dfr(c(ev = "ev", bp = "bp"), function(k) {
  tibble::tibble(condition = k,
                 pearson  = cor(te[[k]], te[[paste0(k, "_naive")]]),
                 spearman = cor(te[[k]], te[[paste0(k, "_naive")]], method = "spearman"),
                 n = nrow(te))
})
print(as.data.frame(agreement), digits = 4, row.names = FALSE)

stopifnot("DESeq and naive TE do not agree in E.V." =
            agreement$pearson[agreement$condition == "ev"] > 0.85,
          "DESeq and naive TE do not agree in 4E-BP1-4A" =
            agreement$pearson[agreement$condition == "bp"] > 0.85)

# Where they diverge: ashr shrinks low-count ratios toward zero, the naive ratio does not.
te %>%
  dplyr::mutate(tertile = ntile(baseMean, 3)) %>%
  dplyr::group_by(tertile) %>%
  dplyr::summarise(n = dplyr::n(),
                   median_baseMean = round(median(baseMean)),
                   r_ev = round(cor(ev, ev_naive), 4),
                   r_bp = round(cor(bp, bp_naive), 4), .groups = "drop") %>%
  as.data.frame() %>% print(row.names = FALSE)
```

```{r naive-scatter, fig.width = 9, fig.height = 4.5}
p_naive <- te %>%
  dplyr::select(gene_id, baseMean, ev, bp, ev_naive, bp_naive) %>%
  tidyr::pivot_longer(c(ev, bp), names_to = "condition", values_to = "deseq") %>%
  dplyr::mutate(naive = if_else(condition == "ev", ev_naive, bp_naive),
                condition = factor(condition, levels = c("ev", "bp"),
                                   labels = c("E.V.", "4E-BP1-4A"))) %>%
  ggplot(aes(naive, deseq)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = "dashed") +
  geom_point(size = 0.4, alpha = 0.25, colour = "grey30") +
  facet_wrap(~ condition) +
  labs(x = "naive log2(ribo / rna), normalized counts + 1",
       y = "DESeq2 group contrast, ashr shrunk") +
  theme_classic(base_size = 14)

print(p_naive)
if (params$save_plots) ggsave(here("plots", "97_deseq_vs_naive_te_scatter.pdf"),
                              p_naive, width = 9, height = 4.5)
```

The two agree closely, and the residual disagreement is confined to the low-expression
tertile exactly as expected. **Everything below uses the DESeq2 values**: they shrink noisy
low-count ratios, carry `lfcSE` and `padj`, and need no pseudocount. The naive values are
kept in the exported table so this comparison stays reproducible.

# 3. Validation

```{r validation}
# Directionality anchor. TE_BP - TE_EV must reproduce nb76's interaction coefficient. The
# shrinkage differs (ashr here, apeglm there) so this is a correlation, not an equality.
anchor <- tibble::tibble(
  pearson  = cor(te$delta, te$nb76_interaction),
  spearman = cor(te$delta, te$nb76_interaction, method = "spearman"),
  sign_agreement = mean(sign(te$delta) == sign(te$nb76_interaction)),
  n = nrow(te))
print(as.data.frame(anchor), digits = 4, row.names = FALSE)

stopifnot("TE levels do not reproduce nb76's TE change: contrast may be inverted" =
            anchor$pearson > 0.9)

# Orientation on the genes nb76 anchors direction with: CDKN2B resistant, GLO1 sensitive.
te %>%
  dplyr::filter(symbol %in% c("CDKN2B", "RPP25", "GLO1", "RPS6")) %>%
  dplyr::select(symbol, ev, bp, delta, nb76_interaction) %>%
  dplyr::arrange(dplyr::desc(delta)) %>%
  as.data.frame() %>% print(digits = 3, row.names = FALSE)

te_of_sym <- function(g, col) te[[col]][match(g, te$symbol)]
stopifnot(
  "CDKN2B TE does not rise under eIF4E inhibition" =
    te_of_sym("CDKN2B", "bp") > te_of_sym("CDKN2B", "ev"),
  "GLO1 TE does not fall under eIF4E inhibition" =
    te_of_sym("GLO1", "bp") < te_of_sym("GLO1", "ev"),
  "non-finite TE" = all(is.finite(te$ev)) && all(is.finite(te$bp)),
  "padj outside [0,1]" = all(te$padj_ev >= 0 & te$padj_ev <= 1, na.rm = TRUE) &&
                         all(te$padj_bp >= 0 & te$padj_bp <= 1, na.rm = TRUE),
  "negative baseMean" = all(te$baseMean >= 0),
  "duplicate gene_id" = !any(duplicated(te$gene_id)),
  "gene set does not match nb76" = setequal(te$gene_id, te_all$gene_id)
)
cat("\nall validation checks passed;", nrow(te), "genes\n")
```

```{r classes}
pos <- read_csv(here("output", "genesets", "teleman_4e_resistant_TE_1hr_lfc0.5.csv"),
                show_col_types = FALSE)
neg <- read_csv(here("output", "genesets", "teleman_4e_sensitive_TE_1hr_lfc0.5.csv"),
                show_col_types = FALSE)

# nb96 asserts these CSVs are setequal to the FASTA headers, so reading them directly gives
# the identical genes in three lines instead of nine.
te <- te %>%
  dplyr::mutate(class = factor(dplyr::case_when(
    gene_id %in% pos$ensembl_gene ~ "4E-resistant",
    gene_id %in% neg$ensembl_gene ~ "4E-sensitive",
    TRUE                          ~ "All other genes"), levels = names(class_colors)))

print(dplyr::count(te, class) %>% as.data.frame(), row.names = FALSE)
stopifnot("class sizes do not match nb96" =
            sum(te$class == "4E-resistant") == nrow(pos) &&
            sum(te$class == "4E-sensitive") == nrow(neg))

if (params$save_plots) {
  write_csv(te %>%
              dplyr::select(gene_id, symbol, class, baseMean,
                            te_ev = ev, te_bp = bp,
                            te_ev_naive = ev_naive, te_bp_naive = bp_naive,
                            lfcSE_ev, lfcSE_bp, padj_ev, padj_bp, delta),
            here("output", "teleman_absolute_te_by_condition.csv"))
}
```

# 4. TE distribution by class, one plot per condition

```{r stats}
te_long <- te %>%
  dplyr::select(gene_id, class, ev, bp) %>%
  tidyr::pivot_longer(c(ev, bp), names_to = "condition", values_to = "te") %>%
  dplyr::mutate(condition = factor(condition, levels = c("ev", "bp"),
                                   labels = c("E.V.", "4E-BP1-4A")))

te_stats <- te_long %>%
  dplyr::group_by(condition, class) %>%
  dplyr::summarise(n = dplyr::n(), median = median(te), .groups = "drop")
print(as.data.frame(te_stats), digits = 3, row.names = FALSE)

te_tests <- te_long %>%
  dplyr::filter(class != "All other genes") %>%
  dplyr::group_by(condition) %>%
  dplyr::summarise(wilcox_p = wilcox.test(te ~ class)$p.value, .groups = "drop")
print(as.data.frame(te_tests), digits = 3, row.names = FALSE)

if (params$save_plots) write_csv(left_join(te_stats, te_tests, by = "condition"),
                                 here("output", "97_absolute_te_by_class_stats.csv"))
```

```{r density-helper}
# One density panel per condition: three classes, their medians, n in the legend, and the
# resistant-vs-sensitive Wilcoxon in the subtitle. Same shape as nb96's ip_density_plot().
te_density_plot <- function(condition_name, xlab) {
  d   <- dplyr::filter(te_long,  condition == condition_name)
  lab <- dplyr::filter(te_stats, condition == condition_name) %>% dplyr::arrange(class)

  ggplot(d, aes(te, colour = class)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
    geom_density(fill = NA, linewidth = 0.9) +
    geom_vline(data = lab, aes(xintercept = median, colour = class),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    scale_colour_manual(values = class_colors, name = NULL,
                        labels = paste0(lab$class, " (n = ", lab$n, ")")) +
    coord_cartesian(xlim = quantile(d$te, c(0.005, 0.995))) +
    labs(subtitle = paste("Wilcoxon, resistant vs sensitive: p =",
                          format.pval(te_tests$wilcox_p[te_tests$condition == condition_name],
                                      digits = 3, eps = 2.2e-16)),
         x = xlab, y = "Density") +
    theme_classic(base_size = 14) +
    theme(legend.position = "top",
          plot.subtitle = element_text(size = 11, colour = "grey40"))
}
```

## E.V. cells

```{r density-ev, fig.width = 7, fig.height = 5}
p_ev <- te_density_plot("E.V.", "Absolute TE, log2( ribo / rna ), E.V. cells")
print(p_ev)
if (params$save_plots) ggsave(here("plots", "97_absolute_te_vehicle_density.pdf"),
                              p_ev, width = 7, height = 5)
```

## 4E-BP1-4A cells

```{r density-bp, fig.width = 7, fig.height = 5}
p_bp <- te_density_plot("4E-BP1-4A", "Absolute TE, log2( ribo / rna ), 4E-BP1-4A cells")
print(p_bp)
if (params$save_plots) ggsave(here("plots", "97_absolute_te_inhibitor_density.pdf"),
                              p_bp, width = 7, height = 5)
```

## Both conditions

```{r violin-both, fig.width = 8, fig.height = 5}
p_violin <- ggplot(te_long, aes(condition, te, fill = class, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  geom_violin(alpha = 0.15, linewidth = 0.8, position = position_dodge(0.9)) +
  geom_boxplot(width = 0.12, alpha = 0, linewidth = 0.5, outlier.shape = NA,
               position = position_dodge(0.9), show.legend = FALSE) +
  geom_text(data = te_stats, aes(y = -Inf, label = paste0("n = ", n)),
            position = position_dodge(0.9), vjust = -0.6, size = 3.5,
            colour = "grey30", show.legend = FALSE) +
  scale_fill_manual(values = class_colors, name = NULL) +
  scale_colour_manual(values = class_colors, name = NULL) +
  coord_cartesian(ylim = quantile(te_long$te, c(0.005, 0.995))) +
  labs(subtitle = "Both conditions are TE levels from the same model, so this axis is anchored.",
       x = NULL, y = "Absolute TE, log2( ribo / rna )") +
  theme_classic(base_size = 14) +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 11, colour = "grey40"))

print(p_violin)
if (params$save_plots) ggsave(here("plots", "97_absolute_te_by_class_violin.pdf"),
                              p_violin, width = 8, height = 5)
```

Unlike nb96's `vehicle` / `inhibitor` IP panels, these two conditions come from a single
`~ group` model on a shared set of size factors, so the offset between them **is** anchored
and the two levels can be compared directly.

# 5. How much of section 4 is selection on the difference?

The classes are the top and bottom of `TE_BP - TE_EV`. Selecting on a difference pushes its
two components apart on its own: the highest-delta genes are enriched for low `ev` and high
`bp` whether or not anything biological distinguishes them. Section 4 therefore cannot be
read as-is, and this section measures the artifact two ways.

## 5a. A null that keeps the marginals but destroys the pairing

```{r selection-null}
set.seed(9)
null_draws <- replicate(200, {
  e <- sample(te$ev); b <- sample(te$bp); d <- b - e
  res <- order(d, decreasing = TRUE)[1:nrow(pos)]
  sen <- order(d)[1:nrow(neg)]
  c(res_ev = median(e[res]), res_bp = median(b[res]),
    sen_ev = median(e[sen]), sen_bp = median(b[sen]))
})

dplyr::bind_rows(
  tibble::tibble(source = "null (random pairing, same selection)",
                 !!!round(apply(null_draws, 1, median), 3)),
  tibble::tibble(source = "observed",
                 res_ev = round(median(te$ev[te$class == "4E-resistant"]), 3),
                 res_bp = round(median(te$bp[te$class == "4E-resistant"]), 3),
                 sen_ev = round(median(te$ev[te$class == "4E-sensitive"]), 3),
                 sen_bp = round(median(te$bp[te$class == "4E-sensitive"]), 3))
) %>% as.data.frame() %>% print(row.names = FALSE)

cat("\ncor(ev, bp) genome-wide:", round(cor(te$ev, te$bp), 3), "\n")
```

Selecting on a difference reproduces a crossover **at least as large as the observed one**.
The observed split is if anything smaller, because `ev` and `bp` are genuinely strongly
correlated, which damps the selection-induced spread relative to the independent null. On
this evidence section 4's crossover is not yet evidence of anything.

## 5b. Split-half: select on one replicate, measure on the other

The decisive test. Both conditions have two replicates, so the classes can be defined from
replicate 1 and the levels measured on replicate 2. Selection noise is then independent of
measurement noise, and the regression-to-the-mean component of the artifact disappears.

Per-replicate TE has no replication left for DESeq2 to model dispersions from, so this
section uses the naive log-ratio, which section 2 showed tracks the DESeq2 values at
r > 0.98.

```{r split-half}
rep_col <- function(treat, lib, rp)
  rownames(fit$meta)[fit$meta$treatment == treat & fit$meta$library == lib &
                     fit$meta$rep == rp]
te_rep <- function(treat, rp)
  log2(fit$norm[, rep_col(treat, "ribo", rp)] + 1) -
  log2(fit$norm[, rep_col(treat, "rna",  rp)] + 1)

ev1 <- te_rep("vehicle", "1"); bp1 <- te_rep("BP_inhib", "1")
ev2 <- te_rep("vehicle", "2"); bp2 <- te_rep("BP_inhib", "2")

cat("cor(delta rep1, delta rep2):", round(cor(bp1 - ev1, bp2 - ev2), 3),
    "   cor(ev rep1, ev rep2):", round(cor(ev1, ev2), 3), "\n\n")

pick <- function(sel) list(res = order(sel, decreasing = TRUE)[1:nrow(pos)],
                           sen = order(sel)[1:nrow(neg)])
sel1 <- pick(bp1 - ev1)

set.seed(9)
sh_null <- replicate(200, {
  s <- pick(sample(bp1 - ev1))
  c(res_ev = median(ev2[s$res]), res_bp = median(bp2[s$res]),
    sen_ev = median(ev2[s$sen]), sen_bp = median(bp2[s$sen]))
})

dplyr::bind_rows(
  tibble::tibble(source = "same replicate (circular)",
                 res_ev = median(ev1[sel1$res]), res_bp = median(bp1[sel1$res]),
                 sen_ev = median(ev1[sel1$sen]), sen_bp = median(bp1[sel1$sen])),
  tibble::tibble(source = "split-half (select rep1, measure rep2)",
                 res_ev = median(ev2[sel1$res]), res_bp = median(bp2[sel1$res]),
                 sen_ev = median(ev2[sel1$sen]), sen_bp = median(bp2[sel1$sen])),
  tibble::tibble(source = "null (shuffled selector, measure rep2)",
                 !!!apply(sh_null, 1, median))
) %>%
  dplyr::mutate(across(where(is.numeric), \(x) round(x, 3)),
                ev_gap = res_ev - sen_ev, bp_gap = res_bp - sen_bp) %>%
  as.data.frame() %>% print(row.names = FALSE)
```

```{r split-half-assertions}
gap <- function(x, y) median(x[sel1$res]) - median(y[sel1$sen])
bp_gap_obs   <- median(bp2[sel1$res]) - median(bp2[sel1$sen])
bp_gap_null  <- median(sh_null["res_bp", ] - sh_null["sen_bp", ])

stopifnot("the 4E-BP1-4A separation does not survive split-half" =
            bp_gap_obs > 10 * abs(bp_gap_null))
cat("4E-BP1-4A gap: observed", round(bp_gap_obs, 3),
    "vs null", round(bp_gap_null, 3), "\n")
```

**Conclusion, and it is not the one section 4 suggests on its own.**

- **Under 4E-BP1-4A the separation is real and large.** Selecting on replicate 1 and
  measuring on replicate 2 still puts the resistant class far above the sensitive class,
  against a null of ~0. This survives independent selection.
- **At baseline it very largely does not.** The apparent E.V. gap collapses once selection
  and measurement are independent, leaving a small residual, and the sign of the residual is
  the opposite of the dramatic crossover in section 4. Most of that crossover was selection
  on the difference, amplified by the fact that `delta` is only moderately reproducible
  between replicates (r ~ 0.4), which is what leaves so much room for regression to the mean.

So the defensible statement is: **4E-resistant mRNAs are distinguished by holding high
translation efficiency when eIF4E is inhibited, not by being unusually poorly translated to
start with.** Section 4's E.V. panel should be read as a description of the selected sets,
not as a claim about baseline biology.

```{r session}
sessionInfo()
```
