---
title: "Feature Set Decision Log: eIF3d Hypoxia RF Classifier (Notebook 07)"
author: "Kate"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
# NOTE: `pm` intentionally shadows lubridate::pm() (checks AM/PM) for the rest of this
# notebook -- it's the shorthand used throughout for output/predictive_modeling/ paths.
pm <- function(f) here("output", "predictive_modeling", f)
```

# Purpose

Notebook 07 (`07_rf_classification_eif3d_targets_no_boruta.Rmd`) is parameterized over a
`feature_set` string that selects one of ~10 different feature matrices built by the
`01_`-`01n_`/`22_`/`23_` extraction notebooks. This document is a decision log for that
feature space: what each feature group is, why it was added, why some things were
deliberately left out, and why `external_stability` is the current default.

**This is a self-checking notebook, not a static writeup.** The tables below are computed
live from the actual feature matrices and saved model files in `output/predictive_modeling/`
every time this is knit — if a documented column disappears or an undocumented one shows up,
the checks in Section 7 will fail or warn rather than silently going stale. That protection
exists because two real staleness bugs were caught and fixed while writing this document
(see the changelog in Section 8) — the pipeline's file-based caching (each notebook reads a
previous notebook's saved `.rds` and overwrites/extends it) has no built-in guarantee that
every step has actually been re-run since its upstream inputs last changed.

---

# Summary: Decision Narrative

The short version, in the order decisions were actually made (not the order notebooks are
numbered). Each row links to the section with the full evidence/rationale.

| Step | New named `feature_set` created here? | Decision | Reasoning |
|---|---|---|---|
| 1. Start simple | **`targeted`** (`01b_`) | Built targeted: lengths, GC%, CLIP, literature signatures | Baseline, cheap and interpretable features first |
| 2. Tried k-mers/RBP motifs | **`combined`** (`01c_`) -- exists, but excluded from the lineage below | Tried adding k-mers/RBP motifs on top of `targeted`; excluded going forward | "Really crowded out signal" -- too many low-signal columns drowned out the real predictors (Sec. 4) |
| 3. Added RNA structure | **`structure`** (`01d_`) | Added 5'UTR accessibility/structure features on top of `targeted` | Incremental addition, not independently compared against `targeted` (Sec. 2) |
| 4. Split the CLIP count | **`clip_split`** (`01e_`) | Split the single aggregate `clip_evidence_count` into eIF3-specific (`clip_eif3_count`) vs cap-binding-specific (`clip_cap_binding_count`) counts | Incremental addition, not independently compared against `structure` (Sec. 2) |
| 5. Added TE features | **`te_lfc`** (`01f_`) | Added own-condition TE LFC features (`indiv_te_*`, `delta_te_*`) | Incremental addition, not independently compared against `clip_split` (Sec. 2) -- this is the feature group flagged circular in step 8 below |
| 6. Added codon optimality | **`codon_optimality`** (`01g_`) | Added CAI/tAI/CSC/FOP | Incremental addition, not independently compared against `te_lfc` (Sec. 2) |
| 7. Added Kozak features | **`kozak`** (`01h_`) | Added Kozak PWM features | Incremental addition, not independently compared against `codon_optimality` (Sec. 2). Steps 3-7 as a whole: **not a designed ablation** -- "otherwise it felt like i was just adding more features in" |
| 8. Deleted circular feature | *(no new name -- decision about `te_lfc`'s existing columns from step 5)* | Recognized `te_lfc`'s own-condition TE features as circular | The TE measurements are close to the same values used to define positive/negative gene sets, which motivated the `exclude_own_te_features` guard in 07 (Sec. 5) |
| 9. Added more CDS features after noticing CDS importance increasing | *(no new name yet -- the observation itself; the resulting build is steps 11-12)* | `cds_gc`/`cds_length` kept surfacing as top-importance features across nearly every fit | Signal, not noise -- worth digging into CDS biology further rather than adding unrelated feature groups |
| 10. **`external_stability` is created here** | **`external_stability`** (`01i_`) -- this is the step where the name/file first exists | `01i_` reads `feature_matrix_cnot3.rds` and adds SLAM-seq half-lives + selective ribosome profiling occupancy from Zhu 2024 (CNOT3 KO), Hia 2026 (DHX29), and Karner 2026 (MDA-MB-231 SLAM-seq) | Literature-grounded, experimentally-measured mRNA stability signal, not just internally-derived features (Sec. 3.9) |
| 11. Acted on the CDS pattern (step 9), part one | *(same name -- `01j_` overwrites **`external_stability`** in place)* | Added positional codon optimality (`01j_`: first-75-codon + CDS-quartile CSC) | Directly follows from step 9 -- doubled down on the codon-optimality feature family that was already showing signal (Sec. 3.3) |
| 12. Acted on the CDS pattern (step 9), part two | *(same name -- `01k_` overwrites **`external_stability`** in place again)* | Added CDS/3'UTR RNA structure (`01k_`), retiring the 5'UTR-only RNAfold columns | A separate decision from step 11 -- extended the same CDS-focused reasoning to structure instead of codon optimality; both landed in the same matrix file but were distinct additions (Sec. 3.5) |
| 13. `external_stability` becomes the default | *(no new name -- the file has been called `external_stability` since step 10; this step is just notebook 07 treating it as the go-to choice)* | Not a separate build chosen over alternatives -- `external_stability` **is** the cumulative endpoint of steps 3-12 stacked in one lineage (structure + clip_split + te_lfc + codon_optimality + kozak + cnot3/dhx29/karner + positional CSC + CDS/3'UTR structure). It became the default simply by being the latest, most complete state of that chain, not by winning a bake-off against `structure`/`clip_split` | Retrospectively, its OOB error isn't actually better than the earlier stages it was built on top of -- but no such comparison drove the decision at the time. **Not driven by OOB error**; biological coverage/signal was the actual reason for continuing to build on this lineage, and the metric is presented only as secondary evidence (Sec. 2 quantitative snapshot) |
| 14. Forked: per-codon detail | **`codon_features`** (`01l_`, reads the by-then-current `external_stability`) | Adds 62 bg-corrected RSCU columns (per-codon usage bias relative to the tRNA-gene-copy-number background, `human_trna_gcn.csv`) + 59 raw codon frequency columns -- built as a separate `feature_set`, not folded into the default. Originally built 2026-04-23 off an older `external_stability` base (which is why it went stale, Sec. 8) and refreshed 2026-07-17 during this document's creation | Shows per-individual-codon signal, not just a single aggregate score -- lets the model test whether specific codon identities matter, beyond the whole-gene CAI/tAI/CSC/FOP already in `external_stability`. Full rationale for not merging still pending (Sec. 6) |
| 15. Forked: own RIP-seq data | **`ripseq`** (`01n_`, reads the by-then-current `external_stability`) | Adds 16 eIF3d/eIF4E RIP-seq IP/IgG and IP/RNA log2FC columns across 4 timepoints -- built as a separate `feature_set`, not folded into the default | This is the lab's **own** RIP-seq data, not a literature dataset like step 10's CNOT3/DHX29/Karner additions -- the goal was to check whether directly-measured own-data binding evidence showed up as strongly as the external literature signal did, once it was available. Full rationale for not merging still pending (Sec. 6) |

**Note on step 13 -- "the default" isn't the whole matrix as-is.** Being `external_stability`'s
"default" status describes which `.rds` file gets loaded, not what actually reaches the model.
Notebook 07 still strips several groups out of it at fit time (Sec. 5 has the live-checked
detail): k-mer/RBP motifs (never present in this matrix to begin with, Sec. 4);
`bg_rscu_`/`codon_freq_` columns (`external_stability`-mode-specific, to stay comparable with
MCF7-SIX1); own-data TE features `indiv_te_*`/`delta_te_*`/`sig_*` when
`exclude_own_te_features = TRUE` (the recommended default, and the direct consequence of step 8);
raw length columns in favor of their log2 versions; and any feature below the 50% non-NA
sparsity threshold. `eif4e_ip_*` is additionally removed if `exclude_4e_features = TRUE`, though
that flag is off by default and, per Sec. 3.7, doesn't catch the rest of the eIF4E/DAP5-related
columns anyway.

**Net effect:** `external_stability` is not the result of a clean "best model wins" ablation.
It's biology-driven feature curation (aggressively cutting noisy feature families like k-mers,
and doubling down on ones showing real signal like CDS-level properties) layered on top of
mostly-incremental feature engineering, with one clearly-identified leakage risk (`te_lfc`)
guarded against via a toggle rather than a hard removal.

---

# Full Feature Inventory

Every column in `feature_matrix_external_stability.rds`, what it is, which notebook
introduced it, and whether it's actually used by notebook 07 today. **Included/Excluded is
not asserted -- it's computed live** by replicating 07's exact default filtering logic
(`feature_set="external_stability"`, `clip_mode="aggregate"`, `exclude_own_te_features=TRUE`,
`exclude_4e_features=FALSE`) against the real positive/negative gene sets for the standard
hypoxia/3d/promotes/1hr/lfc0.5 comparison, including the sparsity check (which depends on the
actual ~2,800-gene modeling subset, not the full 10,172-row matrix). Included features are
listed first; excluded features are grouped at the bottom, with the specific reason each was
dropped.

```{r feature_inventory_setup}
fm_inv <- readRDS(pm("feature_matrix_external_stability.rds"))

# Replicate 07's default positive/negative gene-set join exactly (Sec. "Create Binary Target")
pos_genes <- read_csv(here("output", "genesets", "hypoxia_3d_promotes_TE_1hr_lfc0.5.csv"), show_col_types = FALSE)
neg_genes <- read_csv(here("output", "predictive_modeling", "negative_control_genes.csv"), show_col_types = FALSE) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

set.seed(9)
pos_in_features <- pos_genes %>% filter(ensembl_gene %in% fm_inv$gene_id_clean)
neg_in_features  <- neg_genes %>% filter(gene_id_clean %in% fm_inv$gene_id_clean)
number_to_match  <- min(nrow(pos_in_features), nrow(neg_in_features))

model_genes <- fm_inv %>%
  mutate(eif3d_target = case_when(
    gene_id_clean %in% pos_genes$ensembl_gene ~ "promotes",
    gene_id_clean %in% sample(neg_in_features$gene_id_clean, number_to_match) ~ "does_not_promote",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(eif3d_target))
cat("Replicated 07 default modeling subset:", nrow(model_genes), "genes (",
    sum(model_genes$eif3d_target == "promotes"), "positive,",
    sum(model_genes$eif3d_target == "does_not_promote"), "negative)\n")
```

```{r feature_inventory_filter}
# ---- Replicate 07's filter_features chunk exactly, default params ----
id_cols_base <- c("gene_id", "gene_id_clean", "transcript_id", "transcript_id_clean", "symbol",
                   "te_lfc", "te_padj", "te_lfc_bin", "eif3d_target")
all_cols <- colnames(model_genes)
id_pattern_cols <- all_cols[str_detect(all_cols, "gene_id|transcript_id|^symbol$")]
id_cols <- unique(c(id_cols_base, id_pattern_cols))
step1 <- setdiff(all_cols, id_cols)

aggregate_clip_cols <- c("clip_evidence_count", "clip_eif3_count", "clip_cap_binding_count",
                         "clip_lee_int_stress_lfc", "clip_lee_complete_media_lfc", "clip_lee_glu_dep_lfc",
                         "clip_lee_glu_vs_cm_enrichment")
regional_clip <- step1[str_detect(step1, "^clip_") & !(step1 %in% aggregate_clip_cols)]
step2 <- setdiff(step1, regional_clip)

kmer_motif_pattern <- "^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^utr5_5mer_|^utr5_6mer_|^rbns_|^cisbp_"
kmer_present <- step2[str_detect(step2, kmer_motif_pattern)]
step3 <- setdiff(step2, kmer_present)

codon_detail_cols <- step3[str_detect(step3, "^bg_rscu_|^codon_freq_")]
step4 <- setdiff(step3, codon_detail_cols)

# exclude_4e_features = FALSE (07's actual default) -> nothing removed at this step

own_te_cols <- step4[str_detect(step4, "^indiv_te_|^delta_te_|^sig_")]
step5 <- setdiff(step4, own_te_cols)  # exclude_own_te_features = TRUE (recommended default)

raw_length_cols <- step5[str_detect(step5, "^(utr5|utr3|cds|total)_length$")]
step6 <- setdiff(step5, raw_length_cols)

sparsity_threshold <- 0.5
sparse_cols <- step6[sapply(step6, function(f) mean(!is.na(unlist(model_genes[[f]]))) < sparsity_threshold)]
feature_cols_final <- setdiff(step6, sparse_cols)

cat("Included under default config:", length(feature_cols_final), "of", ncol(fm_inv), "total columns\n")
```

```{r feature_inventory_table, results='asis'}
describe_col <- function(col) {
  case_when(
    col %in% c("transcript_id","transcript_id_clean","gene_id","gene_id_clean","symbol") ~ "Gene/transcript key column",
    col == "te_lfc" ~ "TE log2FC (sictrl/si3d) for the model's own condition -- target-adjacent, always excluded as an id-like column",
    col == "te_padj" ~ "TE adjusted p-value for the model's own condition -- always excluded as an id-like column",
    col == "te_lfc_bin" ~ "Binned TE_LFC category -- always excluded as an id-like column",
    str_detect(col, "^utr5_length$|^utr3_length$|^cds_length$|^total_length$") ~ "Raw region length (nt)",
    str_detect(col, "^log2_") ~ paste0("Log2 of ", sub("^log2_", "", col)),
    col == "utr5_gc" ~ "5'UTR GC%", col == "utr3_gc" ~ "3'UTR GC%", col == "cds_gc" ~ "CDS GC%",
    col == "transcript_gc" ~ "Whole-transcript GC%", col == "gc3" ~ "3rd-codon-position GC%",
    col == "is_4e_resistant" ~ "Literature flag: translation reported eIF4E-inhibition-resistant",
    col == "is_4e_sensitive" ~ "Literature flag: translation reported eIF4E-inhibition-sensitive",
    col == "is_dap5_repressed" ~ "Literature flag: translation reported DAP5-repressed",
    col == "is_dap5_promoted" ~ "Literature flag: translation reported DAP5-promoted",
    str_detect(col, "^sig_") ~ "Own-data TE significance/direction flag for a specific gene/condition/timepoint (see name)",
    str_detect(col, "^clip_lee_parclip_") ~ "Regional eIF3-PAR-CLIP binding flag (Lee et al.), per-dataset",
    str_detect(col, "^clip_cate_") ~ "Regional eIF3-CLIP binding flag (Cate et al. NPC differentiation), per-dataset",
    col == "clip_lee_int_stress_bound" | col == "clip_lee_complete_media_bound" | col == "clip_lee_glu_dep_bound" ~ "Per-dataset CLIP binding flag under a specific metabolic condition (Lee et al.)",
    col == "struct_accessibility_cap_proximal" ~ "RNAplfold unpaired probability, cap-proximal 5'UTR",
    col == "struct_accessibility_aug_context" ~ "RNAplfold unpaired probability, AUG context",
    str_detect(col, "^struct_accessibility_utr5_") ~ "RNAplfold unpaired probability summary, 5'UTR",
    col == "struct_num_structured_regions" ~ "Count of highly-structured 30-nt windows, 5'UTR",
    col == "te_4e_resistance_lfc" ~ "eIF4E-inhibitor resistance score (own/lit-derived LFC)",
    col == "dap5_weber_lfc" ~ "DAP5-dependence LFC (Weber et al.)",
    col == "dap5_polysome_lfc" ~ "DAP5 polysome-association LFC",
    str_detect(col, "^eif4e_ip_") ~ "eIF4E RIP/IP log2FC under a specific drug/vehicle condition",
    str_detect(col, "^clip_eif3_count$|^clip_cap_binding_count$") ~ "Aggregate CLIP evidence count, split by binding-protein category (eIF3 vs cap-binding)",
    str_detect(col, "^clip_lee_.*_lfc$|^clip_lee_glu_vs_cm_enrichment$") ~ "Aggregate CLIP enrichment LFC under a specific metabolic condition (Lee et al.)",
    str_detect(col, "^indiv_te_") ~ "Own-condition TE log2FC at a specific condition/timepoint -- circular relative to TE-defined gene sets",
    str_detect(col, "^delta_te_") ~ "Change in own-condition TE log2FC between timepoints -- circular relative to TE-defined gene sets",
    col == "cai" ~ "Codon Adaptation Index (whole CDS)", col == "tai" ~ "tRNA Adaptation Index (whole CDS)",
    col == "csc" ~ "Codon Stability Coefficient (whole CDS)", col == "fop" ~ "Frequency of Optimal Codons (whole CDS)",
    col == "kozak_score" ~ "Kozak consensus match score", col == "kozak_optimal" ~ "Binary: optimal Kozak context",
    col == "kozak_pwm_score" ~ "Kozak position-weight-matrix score",
    col == "cnot3_ribo_enrichment" ~ "CNOT3 KO riboseq enrichment (Zhu et al. 2024)",
    col == "cnot3_weighted_codon_score" ~ "CNOT3-weighted codon stability score (Zhu et al. 2024)",
    col == "zhu2024_ctrl_halflife" ~ "SLAM-seq control mRNA half-life (Zhu et al. 2024)",
    col == "zhu2024_cnot3ko_halflife_logfc" ~ "SLAM-seq half-life log2FC, CNOT3 KO vs ctrl (Zhu et al. 2024)",
    col == "hia2026_ctrl_halflife" ~ "SLAM-seq control mRNA half-life (Hia et al. 2026)",
    col == "hia2026_dhx29_halflife_logfc" ~ "SLAM-seq half-life log2FC, DHX29 KO vs ctrl (Hia et al. 2026)",
    col == "hia2026_dhx29_occupancy" ~ "Selective ribosome profiling occupancy score, DHX29 (Hia et al. 2026)",
    col == "karner2026_mda231_log2_ct" ~ "SLAM-seq C-to-T conversion ratio, transcription-rate proxy (Karner et al. 2026, MDA-MB-231)",
    str_detect(col, "^csc_q[1-4]$") ~ "Codon Stability Coefficient for a CDS quartile (positional)",
    col == "csc_first75" ~ "Codon Stability Coefficient, first 75 codons only",
    str_detect(col, "^struct_accessibility_") & str_detect(col, "cds") ~ "RNAplfold unpaired probability summary, CDS",
    str_detect(col, "^struct_accessibility_") & str_detect(col, "utr3") ~ "RNAplfold unpaired probability summary, 3'UTR",
    col == "struct_num_structured_regions_cds" ~ "Count of highly-structured 30-nt windows, CDS",
    col == "struct_num_structured_regions_utr3" ~ "Count of highly-structured 30-nt windows, 3'UTR",
    TRUE ~ "(description not yet catalogued -- doc needs updating)"
  )
}

source_map <- c(
  te_lfc="01_ (base)", te_padj="01_ (base)", te_lfc_bin="01_ (base)",
  transcript_id="01_ (base)", transcript_id_clean="01_ (base)", gene_id="01_ (base)",
  gene_id_clean="01_ (base)", symbol="01_ (base)",
  utr5_length="01_ (base)", utr3_length="01_ (base)", cds_length="01_ (base)",
  log2_utr5_length="01_ (base)", log2_utr3_length="01_ (base)", log2_cds_length="01_ (base)",
  total_length="01_ (base)", log2_total_length="01_ (base)",
  utr5_gc="01_ (base)", utr3_gc="01_ (base)", cds_gc="01_ (base)", transcript_gc="01_ (base)", gc3="01_ (base)",
  is_4e_resistant="01_ (base)", is_4e_sensitive="01_ (base)",
  is_dap5_repressed="01_ (base)", is_dap5_promoted="01_ (base)",
  struct_accessibility_cap_proximal="01d_", struct_accessibility_aug_context="01d_",
  struct_accessibility_utr5_mean="01d_", struct_accessibility_utr5_min="01d_", struct_accessibility_utr5_sd="01d_",
  struct_num_structured_regions="01d_",
  te_4e_resistance_lfc="01d_", dap5_weber_lfc="01d_", dap5_polysome_lfc="01d_",
  eif4e_ip_vehicle_lfc="01d_", eif4e_ip_inhibitor_only_lfc="01d_", eif4e_ip_inhib_vs_vehicle_lfc="01d_",
  clip_eif3_count="01e_", clip_cap_binding_count="01e_",
  clip_lee_int_stress_lfc="01e_", clip_lee_complete_media_lfc="01e_", clip_lee_glu_dep_lfc="01e_",
  clip_lee_glu_vs_cm_enrichment="01e_",
  indiv_te_normoxia_1hr="01f_", indiv_te_hypoxia_1hr="01f_", indiv_te_normoxia_4hr="01f_",
  indiv_te_hypoxia_4hr="01f_", indiv_te_normoxia_24hr="01f_", indiv_te_hypoxia_24hr="01f_",
  delta_te_4hr="01f_", delta_te_24hr="01f_",
  cai="01g_", tai="01g_", csc="01g_", fop="01g_",
  kozak_score="01h_", kozak_optimal="01h_", kozak_pwm_score="01h_",
  cnot3_ribo_enrichment="22_", cnot3_weighted_codon_score="22_",
  zhu2024_ctrl_halflife="22_, renamed 01i_", zhu2024_cnot3ko_halflife_logfc="22_, renamed 01i_",
  hia2026_ctrl_halflife="01i_ (from 23_ standalone output)",
  hia2026_dhx29_halflife_logfc="01i_ (from 23_ standalone output)",
  hia2026_dhx29_occupancy="01i_ (from 23_ standalone output)",
  karner2026_mda231_log2_ct="01i_",
  csc_first75="01j_", csc_q1="01j_", csc_q2="01j_", csc_q3="01j_", csc_q4="01j_",
  struct_accessibility_start_proximal_cds="01k_", struct_accessibility_stop_proximal_cds="01k_",
  struct_accessibility_cds_mean="01k_", struct_accessibility_cds_min="01k_", struct_accessibility_cds_sd="01k_",
  struct_num_structured_regions_cds="01k_",
  struct_accessibility_stop_proximal_utr3="01k_", struct_accessibility_distal_utr3="01k_",
  struct_accessibility_utr3_mean="01k_", struct_accessibility_utr3_min="01k_", struct_accessibility_utr3_sd="01k_",
  struct_num_structured_regions_utr3="01k_"
)
source_col <- function(col) {
  if (col %in% names(source_map)) return(unname(source_map[col]))
  case_when(
    str_detect(col, "^sig_") ~ "01_ (base, TE-significance gene-set flags)",
    str_detect(col, "^clip_lee_parclip_|^clip_cate_") ~ "01_/01b_ (base, per-dataset CLIP)",
    col %in% c("clip_lee_int_stress_bound","clip_lee_complete_media_bound","clip_lee_glu_dep_bound") ~ "01_/01b_ (base, per-dataset CLIP)",
    TRUE ~ "(source not yet catalogued)"
  )
}

reason_col <- function(col) {
  case_when(
    col %in% id_cols ~ "Identifier / target-adjacent column -- not a candidate feature at all",
    col %in% kmer_present ~ "k-mer/RBP-motif column (Sec. 4)",
    col %in% codon_detail_cols ~ "bg_rscu_/codon_freq_ detail column (external_stability mode)",
    col %in% regional_clip ~ "Regional per-dataset CLIP column (clip_mode=\"aggregate\" keeps only aggregate counts)",
    col %in% own_te_cols ~ "Own-data TE feature, dropped via exclude_own_te_features=TRUE (Sec. 5)",
    col %in% raw_length_cols ~ "Raw length column, log2 version kept instead (Sec. 5)",
    col %in% sparse_cols ~ paste0("Below 50% non-NA in the current modeling subset (",
                                  round(100 * mean(!is.na(unlist(model_genes[[col]]))), 1), "% coverage)"),
    TRUE ~ ""
  )
}

inventory <- tibble(feature = colnames(fm_inv)) %>%
  mutate(
    what_it_is = map_chr(feature, describe_col),
    source     = map_chr(feature, source_col),
    status     = if_else(feature %in% feature_cols_final, "Included", "Excluded"),
    reason     = if_else(status == "Excluded", map_chr(feature, reason_col), "")
  ) %>%
  arrange(status == "Excluded", feature)

knitr::kable(inventory, col.names = c("Feature", "What it is", "Source", "Status", "Reason if excluded"))
```

```{r feature_inventory_selfcheck}
uncatalogued <- inventory %>% filter(str_detect(what_it_is, "not yet catalogued") | str_detect(source, "not yet catalogued"))
if (nrow(uncatalogued) > 0) {
  cat("WARNING -- these columns are missing a description or source and need the lookup tables above updated:\n")
  print(uncatalogued$feature)
} else {
  cat("OK -- every column in feature_matrix_external_stability.rds has a description and source.\n")
}
```

---

# 1. Pipeline Architecture

The default feature matrix, `feature_matrix_external_stability.rds`, is built by a chain of
notebooks that each read the previous stage's `.rds` and add columns:

```
01_  (lengths, GC%, k-mers, RBP motifs)  ─┐
01b_ (eIF3 CLIP binding, 7 conditions)   ─┴─▶ 01d_ (structure)  *reads 01_ + 01b_ directly,
                                                                  skips 01c_/k-mers — see Sec. 4*
      01d_ ─▶ 01e_ (clip_split: eIF3 vs cap-binding CLIP)
      01e_ ─▶ 01f_ (te_lfc: own-condition TE LFCs — later flagged circular, see Sec. 5)
      01f_ ─▶ 01g_ (codon_optimality: CAI/tAI/CSC/FOP)
      01g_ ─▶ 01h_ (kozak: Kozak PWM features)
      01h_ ─▶ 22_  (cnot3: Zhu 2024 CNOT3 KO half-life/riboseq)      ─▶ feature_matrix_cnot3.rds
      cnot3 ─▶ 01i_ (+ standalone DHX29 half-life/occupancy from 23_, + Karner 2026 stability)
      01i_ ─▶ 01j_ (positional codon stability: first-75-codon + CDS-quartile CSC)  [overwrites in place]
      01j_ ─▶ 01k_ (CDS/3'UTR RNA structure; drops deprecated 5'UTR RNAfold cols)   [overwrites in place]
                                                                 ─▶ feature_matrix_external_stability.rds
```

**The name "external_stability" undersells what this matrix actually is.** It reads as
"just mRNA stability features," but it's really the cumulative superset of nearly every
feature group in the project — sequence, CLIP, structure, codon optimality, Kozak, and
literature-derived stability data — with exactly one deliberate omission: k-mer/RBP motif
features (Section 4).

**A wrinkle in the real ancestry** (not in the CLAUDE.md summary): `01i_` does **not** read
`feature_matrix_dhx29.rds`. It reads `feature_matrix_cnot3.rds` directly, plus three small
standalone files (`dhx29_slamseq_halflives.rds`, `dhx29_riboseq_occupancy.rds`, and the Karner
CSV) that `23_dhx29_riboseq_slamseq.Rmd` also happens to save alongside its own (unused
downstream) full-matrix snapshot. `feature_matrix_dhx29.rds` is an orphaned artifact — no
notebook reads it — and it is currently stale (last built before the `01c_`-skip fix, so it
still carries thousands of k-mer columns). It's harmless because nothing loads it, but it's
confusing to find on disk; see Section 8.

```{r architecture_live_check}
chain_files <- c(
  "feature_matrix_1hr.rds", "feature_matrix_1hr_targeted.rds",
  "feature_matrix_1hr_structure.rds", "feature_matrix_1hr_clip_split.rds",
  "feature_matrix_te_lfc.rds", "feature_matrix_codon_optimality.rds",
  "feature_matrix_kozak.rds", "feature_matrix_cnot3.rds",
  "feature_matrix_external_stability.rds"
)
for (f in chain_files) stopifnot("chain file missing" = file.exists(pm(f)))
cat("All", length(chain_files), "pipeline files present.\n")
```

---

# 2. Chronology of feature_set Options

Per the person who built this (2026-07-17 conversation): the progression from `structure`
through `kozak` was **not** a designed ablation series — "otherwise it felt like i was just
adding more features in." Two decisions in this stretch *were* deliberate, though, and matter
for interpreting the feature set:

- **k-mers/RBP motifs (01c_) were deliberately excluded** — "really crowded out signal."
  This is the one clear subtractive decision in the whole chronology (Section 4).
- **`te_lfc` (01f_) was recognized as adding a circular feature** — the own-condition TE LFC
  values are close to the same measurements used to define the positive/negative gene sets
  in the first place. This is why `exclude_own_te_features` exists in notebook 07 (Section 5);
  the feature was kept in the matrix (for cases where it's *not* circular, e.g. cross-gene
  contrasts) but is excluded by default whenever gene sets are TE-defined.

The one clear *positive* signal-driven decision: CDS-level features (`cds_gc`, `cds_length`)
kept showing up as top-importance features across nearly every fit (see the table in this
section). That empirical pattern motivated deliberately adding more CDS-focused feature
groups afterward — positional codon optimality (`01j_`: first-75-codon and CDS-quartile CSC)
and CDS/3'UTR RNA structure (`01k_`) — rather than continuing to add unrelated feature groups.

## Feature count by stage

```{r chronology_table}
stages <- tribble(
  ~stage,               ~notebook, ~file,
  "targeted",           "01b_",    "feature_matrix_1hr_targeted.rds",
  "structure",          "01d_",    "feature_matrix_1hr_structure.rds",
  "clip_split",         "01e_",    "feature_matrix_1hr_clip_split.rds",
  "te_lfc",             "01f_",    "feature_matrix_te_lfc.rds",
  "codon_optimality",   "01g_",    "feature_matrix_codon_optimality.rds",
  "kozak",              "01h_",    "feature_matrix_kozak.rds",
  "cnot3 (intermediate)","22_",    "feature_matrix_cnot3.rds",
  "external_stability", "01i_/01j_/01k_", "feature_matrix_external_stability.rds"
)
stages <- stages %>%
  mutate(
    exists = file.exists(pm(file)),
    n_cols = map_dbl(file, ~ if (file.exists(pm(.x))) ncol(readRDS(pm(.x))) else NA_real_),
    n_rows = map_dbl(file, ~ if (file.exists(pm(.x))) nrow(readRDS(pm(.x))) else NA_real_)
  )
print(stages %>% dplyr::select(stage, notebook, n_rows, n_cols))
```

## Quantitative snapshot (secondary evidence only — see caveats)

These are **real, already-saved** numbers pulled from `rf_classifier_*.rds` model objects for
the standard hypoxia/3d/promotes/1hr/lfc0.5 comparison — not re-fit for this document (OOB
`prediction.error` is ranger's built-in out-of-bag Brier score; `variable.importance` is
impurity importance). **Caveats, in order of importance:**

1. `exclude_own_te_features` differs across rows (flagged explicitly below) — rows with it
   `TRUE` dropped the circular TE features that rows with it `FALSE` kept, so this is not an
   apples-to-apples comparison of feature *content* alone.
2. OOB Brier error conflates discrimination and calibration and is computed on training data,
   not the held-out test set that 07 reports as its headline AUC.
3. Single 500-tree fits, no repeated CV/seeds — differences under ~0.02 should not be read as
   meaningfully different.
4. **This table did not drive the choice of `external_stability`.** The actual driver was
   biological signal/coverage, not this metric — see the narrative above. It's included for
   transparency, not as justification.

```{r quant_snapshot}
model_files <- tribble(
  ~feature_set,          ~excl_own_te, ~file,
  "targeted",            FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_targeted_clipaggregate_no_boruta.rds",
  "combined",            FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_combined_clipaggregate.rds",
  "structure",           FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_structure_clipaggregate_no_boruta.rds",
  "clip_split",          FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_clip_split_clipaggregate_no_boruta.rds",
  "te_lfc",              FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_te_lfc_clipaggregate_no_boruta.rds",
  "codon_optimality",    FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_codon_optimality_clipaggregate_no_boruta.rds",
  "kozak",               FALSE, "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_kozak_clipaggregate_no_boruta.rds",
  "external_stability",  TRUE,  "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_external_stability_clipaggregate_no_boruta_noOwnTE.rds",
  "codon_features",      TRUE,  "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_codon_features_clipaggregate_no_boruta_noOwnTE.rds",
  "ripseq",              TRUE,  "rf_classifier_eif33d_promotes_hypoxia_1hr_lfc0.5_ripseq_clipaggregate_no_boruta_noOwnTE.rds"
)

snapshot <- model_files %>%
  mutate(exists = file.exists(pm(file))) %>%
  filter(exists) %>%
  mutate(
    model = map(file, ~ readRDS(pm(.x))),
    n_features = map_dbl(model, "num.independent.variables"),
    oob_pred_error = map_dbl(model, "prediction.error"),
    top_feature = map_chr(model, ~ names(sort(.x$variable.importance, decreasing = TRUE))[1])
  ) %>%
  dplyr::select(feature_set, excl_own_te, n_features, oob_pred_error, top_feature) %>%
  arrange(oob_pred_error)

print(snapshot)
cat("\nNote: codon_features and dhx29-derived numbers above predate the",
    Sys.Date(), "data-freshness fixes in Section 8 and should be treated as provisional",
    "until 07 is re-run on the corrected matrix.\n")
```

---

# 3. Current Feature Dictionary (`external_stability`)

```{r load_current_matrix}
fm <- readRDS(pm("feature_matrix_external_stability.rds"))
cat("Current feature_matrix_external_stability.rds:", nrow(fm), "rows x", ncol(fm), "cols\n")
```

## 3.1 Identifiers (not model features)

`transcript_id`, `transcript_id_clean`, `gene_id`, `gene_id_clean`, `symbol` — the standard
gene/transcript key columns. Always excluded from `feature_cols` in 07 via `id_cols`/
`id_pattern_cols` regardless of `feature_set`.

## 3.2 Sequence length & GC content (`01_`)

Raw and log2 UTR5/UTR3/CDS/total length, plus GC% per region and `gc3` (3rd-codon-position
GC). **Raw length columns are dropped by 07 itself** (Section 5) — log2 versions are kept
because length is right-skewed and keeping both splits importance across two correlated
columns.

```{r seq_length_cols}
seq_cols <- grep("length|_gc$|^gc3$|^transcript_gc$", colnames(fm), value = TRUE)
cat(length(seq_cols), "columns:\n"); cat(paste(seq_cols, collapse = ", "), "\n")
stopifnot("ACTB-range GC% sanity check failed" =
  all(fm$cds_gc >= 0 & fm$cds_gc <= 100, na.rm = TRUE))
```

## 3.3 Codon optimality (`01g_`, `01j_`)

Whole-CDS CAI/tAI/CSC/FOP (`01g_`), plus positional codon stability added after CDS features
kept surfacing as top predictors: `csc_first75` (first 75 codons) and `csc_q1`-`csc_q4`
(CDS quartiles).

```{r codon_opt_cols}
codon_cols <- grep("^cai$|^tai$|^csc|^fop$", colnames(fm), value = TRUE)
cat(length(codon_cols), "columns:\n"); cat(paste(codon_cols, collapse = ", "), "\n")
stopifnot("CAI/tAI/FOP must be in [0,1]" =
  all(fm$cai >= 0 & fm$cai <= 1, na.rm = TRUE),
  all(fm$tai >= 0 & fm$tai <= 1, na.rm = TRUE),
  all(fm$fop >= 0 & fm$fop <= 1, na.rm = TRUE))
```

## 3.4 Kozak context (`01h_`)

```{r kozak_cols}
kozak_cols <- grep("^kozak_", colnames(fm), value = TRUE)
cat(length(kozak_cols), "columns:", paste(kozak_cols, collapse = ", "), "\n")
```

## 3.5 RNA secondary structure (`01d_`, `01k_`)

5'UTR accessibility (`01d_`; 3 columns retained — cap-proximal, AUG-context, and UTR5 mean/
min/sd accessibility), plus CDS and 3'UTR accessibility added by `01k_` after the 5'UTR-only
RNAfold thermodynamic columns (`struct_mfe_utr5`, `struct_ensemble_free_energy`, etc.) were
dropped as deprecated.

```{r structure_cols}
struct_cols <- grep("^struct_", colnames(fm), value = TRUE)
cat(length(struct_cols), "columns:\n"); cat(paste(struct_cols, collapse = ", "), "\n")
stopifnot("Deprecated 5'UTR RNAfold columns should be gone" =
  !any(c("struct_mfe_utr5", "struct_ensemble_free_energy") %in% colnames(fm)))
stopifnot("CDS/3'UTR structure columns should be present" =
  all(c("struct_accessibility_cds_mean", "struct_accessibility_utr3_mean") %in% colnames(fm)))
```

## 3.6 CLIP binding — eIF3 (`01b_`, `01e_`)

Aggregate counts (`clip_eif3_count`, `clip_cap_binding_count`) plus per-dataset LFC/binary
columns (Lee et al. glucose/stress conditions, Cate et al. NPC differentiation). Notebook 07's
`clip_mode` param controls whether the aggregate or per-dataset columns are used (Section 5).

```{r clip_cols}
clip_cols <- grep("^clip_", colnames(fm), value = TRUE)
cat(length(clip_cols), "columns:\n"); cat(paste(clip_cols, collapse = ", "), "\n")
```

## 3.7 eIF4E / cap-binding features

`te_4e_resistance_lfc`, `dap5_weber_lfc`, `dap5_polysome_lfc`, `eif4e_ip_*`,
`is_4e_resistant`/`is_4e_sensitive`/`is_dap5_repressed`/`is_dap5_promoted` — all thematically
related to eIF4E/cap-binding biology, grouped here for documentation purposes.

**Only `eif4e_ip_*` is actually removed by `exclude_4e_features = TRUE`.** 07's regex is
`^4e_resistance|^eif4e_ip` plus 5 exact `lit_*eIF4E*` names (Section 5) — `te_4e_resistance_lfc`
doesn't match because it starts with `te_`, not `4e_resistance`, and none of the `dap5_*`/
`is_4e_*`/`is_dap5_*` columns match either pattern. Confirmed live below. Whether that's an
intentional scope decision (only the direct IP-binding data is "circular," not the derived
resistance/DAP5 scores) or a gap in the regex is an open question — flagged in Section 8
rather than assumed either way.

```{r eif4e_cols}
eif4e_cols <- grep("^4e_resistance|^eif4e_ip|^is_4e_|^is_dap5_|^dap5_|^te_4e_", colnames(fm), value = TRUE)
cat(length(eif4e_cols), "columns:", paste(eif4e_cols, collapse = ", "), "\n")

circular_regex_check <- "^4e_resistance|^eif4e_ip"
actually_excluded <- eif4e_cols[str_detect(eif4e_cols, circular_regex_check)]
not_excluded <- setdiff(eif4e_cols, actually_excluded)
cat("Actually removed by exclude_4e_features:", paste(actually_excluded, collapse = ", "), "\n")
cat("NOT removed despite thematic similarity:", paste(not_excluded, collapse = ", "), "\n")
```

## 3.8 Own-data TE features (`01f_`) — excluded by default

`te_lfc`, `te_padj`, `te_lfc_bin`, `indiv_te_*`, `delta_te_*`, and the `sig_*` gene-set flags.
Flagged as circular because these are close to the same TE measurements used to define
positive/negative gene sets. Removed whenever `exclude_own_te_features = TRUE`
(07's recommended default for TE-defined comparisons).

```{r own_te_cols}
own_te_cols <- grep("^indiv_te_|^delta_te_|^sig_|^te_lfc|^te_padj", colnames(fm), value = TRUE)
cat(length(own_te_cols), "columns (", length(grep('^sig_', own_te_cols)), "are sig_* gene-set flags)\n")
```

## 3.9 External literature stability data (`22_`, `23_`/`01i_`)

- **Zhu et al. 2024** (CNOT3 KO): `cnot3_ribo_enrichment`, `cnot3_weighted_codon_score`,
  `zhu2024_ctrl_halflife`, `zhu2024_cnot3ko_halflife_logfc`
- **Hia et al. 2026** (DHX29): `hia2026_ctrl_halflife`, `hia2026_dhx29_halflife_logfc`,
  `hia2026_dhx29_occupancy`
- **Karner et al. 2026** (MDA-MB-231 SLAM-seq): `karner2026_mda231_log2_ct`

```{r external_lit_cols}
ext_cols <- grep("^cnot3_|^zhu2024_|^hia2026_|^karner2026_", colnames(fm), value = TRUE)
cat(length(ext_cols), "columns:", paste(ext_cols, collapse = ", "), "\n")
stopifnot("Half-life columns must be positive" =
  all(fm$zhu2024_ctrl_halflife > 0, na.rm = TRUE),
  all(fm$hia2026_ctrl_halflife > 0, na.rm = TRUE))
```

---

# 4. Excluded Feature Group: k-mers & RBP Motifs

`01_` and `01c_` compute ~900-6,000 k-mer (`utr5_4mer_*`, `utr3_4mer_*`, `cds_4mer_*`,
`utr5_5mer_*`, `utr5_6mer_*`) and RBP motif (`rbns_*`, `cisbp_*`) columns. These are
**deliberately excluded** from `external_stability` and most other feature sets — per the
project owner, "really crowded out signal." `01d_` (the first stage in the
`external_stability` lineage) reads `01_` + `01b_` directly rather than `01c_`'s combined
output, so these columns never enter the chain at all. Notebook 07 also independently strips
any k-mer/motif columns that do make it into a matrix, for the feature_sets listed in its
`kmer_motif_pattern` filter.

```{r kmer_exclusion_check}
kmer_cols_present <- grep("^utr5_4mer_|^utr3_4mer_|^cds_4mer_|^rbns_|^cisbp_", colnames(fm), value = TRUE)
stopifnot("k-mer/RBP motif columns should not be in external_stability" =
  length(kmer_cols_present) == 0)
cat("Confirmed: 0 k-mer/RBP motif columns in the current external_stability matrix.\n")
```

---

# 5. Selection Logic Inside Notebook 07

Beyond which `.rds` file gets loaded, notebook 07 itself applies several filters to
`feature_cols` before fitting — these are documented in-line in the notebook and reproduced
here since they're as much a "feature decision" as anything upstream.

| Param / step | Behavior | Rationale (from 07's own code comments) |
|---|---|---|
| `clip_mode = "aggregate"` (default) | Keeps only the 7 aggregate CLIP columns, drops per-dataset regional ones | Simpler, recommended default |
| `clip_mode = "individual"` | Keeps per-dataset regional binary features, drops aggregates | For dataset-specific analyses |
| k-mer/motif strip | Removes `utr5_4mer_`/`rbns_`/`cisbp_`-pattern columns for a specific list of feature_sets (including `external_stability`) | Redundant safety net on top of Section 4's upstream exclusion |
| `external_stability`-only: drop `bg_rscu_`/`codon_freq_` | Removes per-codon detail columns | "these live in the MCF7-SIX1 external_stability matrix but are not present in the MDA-MB-231 external_stability matrix — keeps feature space consistent across cell lines" |
| `exclude_4e_features = TRUE` | Drops `4e_resistance*`, `eif4e_ip*`, and 5 exact `lit_*eIF4E*` columns | "remove circular eIF4E features (to enable fair cross-model comparison)" |
| `exclude_own_te_features = TRUE` (recommended default) | Drops `indiv_te_*`, `delta_te_*`, `sig_*` | "recommended when pos/neg sets are defined by TE changes" — these are the features flagged circular in Section 2 |
| Raw length drop (always applied) | Drops `utr5_length`/`utr3_length`/`cds_length`/`total_length`, keeps log2 versions | "log2 versions are always present and preferred (right-skewed distributions; keeping both splits feature importance across two correlated cols)" |
| Sparsity filter (always applied, threshold 0.5) | Drops any feature with <50% non-NA values | "sparse features inflate impurity importance via median imputation creating a spurious split between real vs imputed values" |

```{r selection_logic_live_check}
# Confirm the circular eIF4E list and own-TE regex still match real columns in the current matrix
circular_regex <- "^4e_resistance|^eif4e_ip"
circular_lit_exact <- c(
  "lit_eIF4E_activated_translation", "lit_phospho_eIF4E_activated_translation",
  "lit_phospho_eIF4E_suppressed_translation", "lit_translation_is_sensitive_to_4E_inhibition",
  "lit_translation_is_resistant_to_4E_inhibition"
)
matched_circular <- colnames(fm)[str_detect(colnames(fm), circular_regex) | colnames(fm) %in% circular_lit_exact]
cat("exclude_4e_features would currently drop", length(matched_circular), "columns:",
    paste(matched_circular, collapse = ", "), "\n")

own_te_matched <- colnames(fm)[str_detect(colnames(fm), "^indiv_te_|^delta_te_|^sig_")]
cat("exclude_own_te_features would currently drop", length(own_te_matched), "columns.\n")
```

---

# 6. Forked Experimental Extensions: `codon_features` & `ripseq`

Both fork off `external_stability` as separate `feature_set` options rather than being merged
into the default:

- **`codon_features`** (`01l_`): adds 62 bg-corrected RSCU columns (per-codon usage bias
  relative to the tRNA-gene-copy-number background, `human_trna_gcn.csv`) + 59 raw codon
  frequency columns (121 total) on top of `external_stability`'s 134. Where `cai`/`tai`/`csc`/
  `fop` (already in `external_stability`) collapse codon usage into one aggregate score per
  gene, `codon_features` exposes usage at the level of each individual codon — the question
  it's built to answer is whether specific codon identities carry signal beyond what the
  aggregate scores capture, not just "is this gene codon-optimized."
- **`ripseq`** (`01n_`): adds 16 eIF3d/eIF4E RIP-seq IP/IgG and IP/RNA log2FC columns across 4
  timepoints on top of `external_stability`. Unlike step 10's CNOT3/DHX29/Karner additions,
  this is the lab's **own** RIP-seq data, not a literature dataset — it was built to test
  whether directly-measured own-data binding evidence showed up as strongly as the external
  literature signal did, once that own data was available.

Both are currently framed as **experimental extensions, not adopted into the default** — why
neither one displaced `external_stability` as the baseline is still an open question
(Section 8).

---

# 7. Self-Check Summary

```{r final_self_check}
expected_current_cols <- c(
  seq_cols, codon_cols, kozak_cols, struct_cols, clip_cols, eif4e_cols, own_te_cols, ext_cols,
  "transcript_id", "transcript_id_clean", "gene_id", "gene_id_clean", "symbol"
)
undocumented <- setdiff(colnames(fm), expected_current_cols)
if (length(undocumented) > 0) {
  cat("WARNING — columns present in feature_matrix_external_stability.rds but not documented",
      "in any section above (doc needs an update):\n")
  cat(paste(" ", undocumented, collapse = "\n"), "\n")
} else {
  cat("OK — every column in the current matrix is accounted for in this document.\n")
}
missing_documented <- setdiff(expected_current_cols, colnames(fm))
stopifnot("A documented column has disappeared from the matrix" = length(missing_documented) == 0)
```

---

# 8. Open Questions / Known Gaps / Changelog

**Fixed while writing this document (2026-07-17):**

- `feature_matrix_codon_features.rds` was stale — built (2026-04-23) from an
  `external_stability` base that predated later fixes to `feature_matrix_external_stability.rds`
  (2026-04-29), and carried 256 leftover `utr5_4mer_*` k-mer columns (6,537 cols total instead
  of the expected ~250). Fixed by re-running `01l_`; now 250 columns, 0 k-mer columns.
- `01k_` (CDS/3'UTR structure) had apparently never been run against the current
  `feature_matrix_external_stability.rds` — the file still had the deprecated 5'UTR RNAfold
  columns and zero CDS/3'UTR structure columns, even though `01k_`'s cluster-precomputed
  RNAplfold inputs were present and complete. Fixed by running `01k_`; the matrix is now 134
  columns (was 129), with 7 deprecated columns dropped and 12 CDS/3'UTR columns added.

**Not fixed, flagged only:**

- `feature_matrix_dhx29.rds` (6,404 cols, includes stale k-mer columns) is an orphaned
  artifact of `23_dhx29_riboseq_slamseq.Rmd` that no downstream notebook reads. Harmless but
  confusing; candidate for deletion.
- Any `rf_classifier_*.rds` for `codon_features` predates today's fix and should be
  considered stale until 07 is re-run on the corrected matrix.

**Genuinely open (need more input to fill in):**

- Whether `exclude_4e_features` is intentionally scoped to only `eif4e_ip_*` (Section 3.7),
  leaving `dap5_*`/`is_4e_*`/`is_dap5_*`/`te_4e_resistance_lfc` in the model even when the
  flag is set, or whether the regex should be widened to catch those too.
- Why `codon_features` and `ripseq` weren't adopted as the new default over
  `external_stability` (Section 6).
- No formal, controlled ablation exists across `structure` → `clip_split` → `te_lfc` →
  `codon_optimality` → `kozak` — per the project owner, this stretch was incremental
  development, not a designed comparison. Noted here rather than inventing a rationale.

```{r session_info}
sessionInfo()
```
